### 2022-08-05 ms.Rnw
Copied over files, moved header into ms.Rnw

%% Template for the submission to:
%%   The Annals of Applied Statistics [AOAS]
%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% In this template, the places where you   %%
%% need to fill in your information are     %%
%% indicated by '???'.                      %%
%%                                          %%
%% Please do not use \input{...} to include %%
%% other tex files. Submit your LaTeX       %%
%% manuscript as one .tex document.         %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\documentclass[aoas]{imsart}

%% Packages
\RequirePackage{amsthm,amsmath,amsfonts,amssymb,graphicx,enumerate,url,xr,lmodern}
\RequirePackage[authoryear]{natbib}
\usepackage{url}
\usepackage{makecell}
\usepackage{multirow}
\usepackage{multicol}
\usepackage[ruled,noline,linesnumbered]{algorithm2e}
\usepackage{color}
\usepackage[normalem]{ulem}% to use \sout in feedback commands
\usepackage{bm}
\usepackage[mathscr]{euscript}

\startlocaldefs
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% Uncomment next line to change            %%
%% the type of equation numbering           %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\numberwithin{equation}{section}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% For Axiom, Claim, Corollary, Hypothesis, %%
%% Lemma, Theorem, Proposition              %%
%% use \theoremstyle{plain}                 %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\theoremstyle{plain}
%\newtheorem{???}{???}
%\newtheorem*{???}{???}
%\newtheorem{???}{???}[???]
%\newtheorem{???}[???]{???}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% For Assumption, Definition, Example,     %%
%% Notation, Property, Remark, Fact         %%
%% use \theoremstyle{remark}                %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\theoremstyle{remark}
%\newtheorem{???}{???}
%\newtheorem*{???}{???}
%\newtheorem{???}{???}[???]
%\newtheorem{???}[???]{???}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Please put your definitions here:        %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%% parameters %%%%%%%%%%%
\newcommand\Wsat{W_{\mathrm{sat}}}
\newcommand\muIR{\mu_{IR}}
\newcommand\muEI{\mu_{EI}}
\newcommand\transmission{\beta}
\newcommand\seasAmplitude{a}
\newcommand\rainfallExponent{r}
\newcommand\muRS{\mu_{RS}}
\newcommand\vaccineEfficacy{\theta}
\newcommand\muBirth{\mu_S}
\newcommand\muDeath{\delta}
\newcommand\choleraDeath{\delta_{C}}
\newcommand\symptomFrac{f}
\newcommand\asymptomRelativeInfect{\epsilon}
\newcommand\asymptomRelativeShed{\epsilon_{W}}
\newcommand\Wbeta[1]{\beta_{W#1}}
\newcommand\Wremoval{\delta_W}
\newcommand\Wshed{\mu_W}
\newcommand\mixExponent{\nu}
\newcommand\sigmaProc{\sigma_{\mathrm{proc}}}
\newcommand\reportRate{\rho}
\newcommand\obsOverdispersion{\psi}
\newcommand\phaseParm{\phi}
\newcommand\transmissionTrend{\zeta}
\newcommand\vaccClass{Z}
\newcommand\vaccCounter{z}
\newcommand\modelCounter{m}
\newcommand\missing{\hspace{6mm} ---}
\newcommand\fixed{\color{blue}}
\newcommand\demography{D}
\newcommand\code[1]{\texttt{#1}}

\DeclareSymbolFont{matha}{OML}{txmi}{m}{it}% txfonts
\DeclareMathSymbol{\varv}{\mathord}{matha}{118}
\newcommand\myeqref[1]{(\ref{#1})}
\newcommand{\blind}{1}

%% customized math macros
\newcommand\seq[2]{{#1}\!:\!{#2}}
\newcommand\R{\mathbb{R}}
\newcommand\Var{\mathrm{Var}}
\newcommand\var{\Var}
\newcommand\Cov{\mathrm{Cov}}
\newcommand\cov{\Cov}
\newcommand\iid{\mathrm{iid}}
\newcommand\dist{d}
\def\lik{L}
\def\loglik{\ell}

%%%%%% EDITING MACROS %%%%%%%%%
% orange for EI
\definecolor{orange}{rgb}{1,0.5,0}
\newcommand\ei[2]{\sout{#1} \textcolor{orange}{#2}}
\newcommand\eic[1]{\textcolor{orange}{[#1]}}
% green for JW
\definecolor{green}{rgb}{0,0.5,0}
\newcommand\jw[2]{\sout{#1} \textcolor{green}{#2}}
\newcommand\jwc[1]{\textcolor{green}{[#1]}}
% purple for JJ
\definecolor{purple}{rgb}{0.5,0,1}
\newcommand\jj[2]{\sout{#1} \textcolor{purple}{#2}}
\newcommand\jjc[1]{\textcolor{purple}{[#1]}}
% cyan for AR
\definecolor{cyan}{rgb}{0,.5,.5}
\newcommand\ar[2]{\sout{#1} \textcolor{cyan}{#2}}
\newcommand\arc[1]{\textcolor{cyan}{#1}}
% light brown for KT
\definecolor{lightbrown}{rgb}{0.5,0.5,0}
\newcommand\kt[2]{\sout{#1} \textcolor{lightbrown}{#2}}
\newcommand\ktc[1]{\textcolor{lightbrown}{#1}}

\endlocaldefs

\begin{document}

<<Setup, include=FALSE,echo=FALSE,results='hide'>>=
library(knitr)
library(pomp)
library(panelPomp)
library(spatPomp)
library(doParallel)
library(doRNG)
library(haitipkg)
library(tidyverse)

RUN_LEVEL <- 1

opts_knit$set(concordance=TRUE)
opts_chunk$set(
    progress = TRUE,
    concordance = TRUE,
    prompt = TRUE,
    highlight = FALSE,
    tidy = TRUE,
    tidy.opts = list(
        keep.blank.line = FALSE
    ),
    comment = "",
    warning = FALSE,
    message = FALSE,
    error = TRUE,
    echo = FALSE,
    cache = FALSE,
    strip.white = TRUE,
    # results="markup",
    background = "#FFFFFF00",
    size = "normalsize",
    fig.path = "figure/",
    fig.lp = "fig:",
    fig.align = "left",
    fig.show = "asis",
    dev = "pdf",
    dev.args = list(
        bg = "transparent",
        pointsize = 9
    )
)

myround<- function (x, digits = 1) {
  # taken from the broman package
  if (digits < 1)
    stop("This is intended for the case digits >= 1.")
  if (length(digits) > 1) {
    digits <- digits[1]
    warning("Using only digits[1]")
  }
  tmp <- sprintf(paste("%.", digits, "f", sep = ""), x)
  zero <- paste0("0.", paste(rep("0", digits), collapse = ""))
  tmp[tmp == paste0("-", zero)] <- zero
  tmp
}

theme_set(theme_bw() + theme(plot.title = element_text(hjust = 0.5)))

# 40 cores for doob, 8 for ito
doob_cores <- 40
gl_cores <- 36

cores <-  as.numeric(Sys.getenv('SLURM_NTASKS_PER_NODE', unset=NA))
if(is.na(cores)) cores <- detectCores()
registerDoParallel(cores)

dep_labeller <- as_labeller(
  c(
    'Artibonite' = 'Artibonite',
    'Sud_Est' = 'Sud-Est',
    'Sud.Est' = 'Sud-Est',
    'Nippes' = 'Nippes',
    'Nord_Est' = 'Nord-Est',
    'Nord.Est' = 'Nord-Est',
    'Ouest' = 'Ouest',
    'Centre' = 'Centre',
    'Nord' = 'Nord',
    'Sud' = 'Sud',
    'Nord_Ouest' = 'Nord-Ouest',
    'Nord.Ouest' = 'Nord-Ouest',
    'Grande_Anse' = 'Grand\'Anse',
    'Grand.Anse' = 'Grand\'Anse'
  )
)

options(
  scipen = 0,
  help_type = "html",
  stringsAsFactors = FALSE,
  # prompt="R> ",
  continue = "+  ",
  width = 70,
  useFancyQuotes = FALSE,
  reindent.spaces = 2,
  xtable.comment = FALSE
)

true_agg_cases <- haitiCholera %>%
  select(-report) %>%
  pivot_longer(
    data = .,
    cols = -date_saturday,
    values_to = 'cases',
    names_to = 'dep'
  ) %>%
  mutate(
    date = as.Date(date_saturday),
    dep = gsub("\\.", "_", dep)
  ) %>%
  mutate(
    dep = case_when(dep == "Grand_Anse" ~ "Grande_Anse", TRUE ~ dep)
  ) %>%
  tidyr::pivot_wider(
    data = .,
    id_cols = c(date),
    names_from = dep,
    values_from = cases,
    names_prefix = 'cases_'
  ) %>%
  mutate(
    ReportedAll = cases_Artibonite + cases_Centre +
      cases_Grande_Anse + cases_Nippes + cases_Nord +
      cases_Nord_Est + cases_Ouest + cases_Sud +
      cases_Sud_Est + cases_Nord_Ouest
  )

dateToYears <- function(date, origin = as.Date("2014-01-01"), yr_offset = 2014) {
  # This function converts a date to a decimal representation
  #
  # ex: "1976-03-01" -> 1976.163

  julian(date, origin = origin) / 365.25 + yr_offset
}

yearsToDate <- function(year_frac, origin = as.Date("2014-01-01"), yr_offset = 2014.0) {
  # This function is the inverse function of dateToYears; it takes
  # a decimal representation of a date and converts it into a Date.
  #
  # ex: 1976.163 -> "1976-03-01"

  as.Date((year_frac - yr_offset) * 365.25, origin = origin)
}

yearsToDateTime <- function(year_frac, origin = as.Date("2014-01-01"), yr_offset = 2014.0) {
  # Same as the function above, but a DateTime object rather than a Date
  # object.
  #
  # ex: 1976.163 -> "1976-03-01"
  as.POSIXct((year_frac - yr_offset) * 365.25 * 3600 * 24, origin = origin)
}

@

\begin{frontmatter}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% Enter the title of your article here     %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{Informing policy via dynamic models: Eliminating cholera in Haiti}
\runtitle{Eliminating Cholera in Haiti}

\begin{aug}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Only one address is permitted per author. %%
%% Only division, organization and e-mail is %%
%% included in the address.                  %%
%% Additional information can be included in %%
%% the Acknowledgments section if necessary. %%
%% ORCID can be inserted by command:         %%
%% \orcid{0000-0000-0000-0000}               %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\author[A]{\fnms{Jesse}~\snm{Wheeler}\ead[label=e1]{jeswheel@umich.edu}\orcid{0000-0003-3941-3884}},
\author[A]{\fnms{AnnaElaine}~\snm{Rosengart}\ead[label=e2]{aelr@umich.edu}}
\author[A]{\fnms{Zhuoxun}~\snm{Jiang}\ead[label=e3]{zhuoxunj@umich.edu}},
\author[A]{\fnms{Kevin}~\snm{Hao En Tan}\ead[label=e4]{kevtan@umich.edu}},
\author[A]{\fnms{Noah}~\snm{Treutle}\ead[label=e5]{ntreutle@umich.edu}}
\and
\author[A]{\fnms{Edward}~\snm{Ionides}\ead[label=e6]{ionides@umich.edu}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Addresses                                %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\address[A]{Statistics Department, University of Michigan\printead[presep={,\ }]{e1,e2,e3,e4,e5,e6}}
\end{aug}

\begin{abstract}
Public health decisions must be made about when and how to implement interventions to control an infectious disease epidemic.
These decisions should be informed by data on the epidemic as well as current understanding about the transmission dynamics.
Such decisions can be posed as statistical questions about scientifically motivated dynamic models.
\eic{Thus, we encounter the methodological task of building credible, data-informed decisions for stochastic, partially observed, nonlinear dynamic models.
This necessitates addressing the tradeoff between biological fidelity and model simplicity, and the reality of misspecification for models at all levels of complexity. THE CURRENT MS VERSION DOES NOT QUITE AIM AT THESE ISSUES. BUT, THEY ARE HIGHLY RELEVANT , AND GIVE A BROADER STATISTICAL CONTEXT FOR THE EPIDEMIOLOGICAL CASE STUDY. MAYBE WE CAN EDIT IN THAT DIRECTION?}
As a case study, we consider a cholera epidemic in Haiti.
The 2010 introduction of cholera to Haiti led to an extensive outbreak and sustained transmission, eliminated in 2019 with the help of vaccination and other public health measures.
We study three models developed by expert teams to advise on vaccination policies.
We assess methods used for fitting and evaluating these models, leading to recommendations for future studies.
% Methods that lead to better statistical fit include benchmarking mechanistic models against associative statistical models and using state-of-the art tools designed to fit mechanistic models.
Better statistical fit facilitates interpretation of the fitted model, but caution is nevertheless required in drawing policy conclusions based on causal interpretations of the models.
\end{abstract}

\begin{keyword}
  \kwd{Partially observed Markov process}
  \kwd{Hidden Markov model}
  \kwd{infectious disease}
  \kwd{cholera}
  \kwd{sequential Monte Carlo}
\end{keyword}

\end{frontmatter}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Please use \tableofcontents for articles %%
%% with 50 pages and more                   %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\tableofcontents

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%% Main text entry area:

\section{Introduction}

Quantitative models for dynamic systems offer potential for designing effective control measures.
Regulation of biological populations is a fundamental topic in epidemiology, ecology, fisheries and agriculture.
Quantitative models for these population dynamics may be nonlinear and stochastic, with the resulting complexities compounded by incomplete understanding of the underlying biological mechanisms and by partial observability of the system variables.
Developing and testing such models, and assessing their fitness for guiding policy, is a challenging statistical task.
Questions of interest include: What indications should we look for in the data to assess whether the model-based inferences are trustworthy?
What diagnostic tests and model variations can and should be considered in the course of the data analysis?
What are the possible trade-offs of increasing model complexity, such as the inclusion of interactions across spatial units?

This case study investigates the use of dynamic models and spatiotemporal data to inform a policy decision in the context of the cholera outbreak in Haiti, which started in 2010.
We build on a multi-group modeling exercise by \citet{lee20} in which four expert modeling teams developed models to the same dataset with the goal of comparing conclusions on the feasibility of eliminating cholera by a vaccination campaign.
Model~1 is stochastic and describes cholera at the national level;
Model~2 is deterministic with spatial structure, and includes transmission via contaminated water;
Model~3 is stochastic with spatial structure, and accounts for measured rainfall.
Model~4 has an agent-based construction, featuring considerable mechanistic detail but limited ability to calibrate these details to data.
The strengths and weaknesses of the agent-based modeling approach  \citep{tracy18} are outside the scope of this article, and we focus on Models~1--3.

The four independent teams were given the task of projecting the
potential effect of various vaccination deployment strategies on the cholera epidemic in Haiti.
\jwc{TODO: MAYBE DISCUSS MORE ABOUT THE VACCINES, LIKE WHAT EFFECTS THEY ARE KNOWN TO HAVE, NOTE HOW VACCINE SUPPLIES ARE LIMITED, ETC.}
Certain data were shared between the groups, including demography and vaccination history; vaccine efficacy was also fixed at a shared value between groups.
Beyond this, the groups made autonomous decisions on what to include and exclude from their models; this autonomy reduced the possible effect that assumptions about the dynamic system may have on the final conclusion of the study.
Despite this autonomy, and largely adhering to existing guidelines on creating models to inform policy \citep{behrend20,saltelli20}, the consensus across the four models was that an extensive nationwide vaccination campaign would be necessary to eliminate cholera from Haiti.
This conclusion is inconsistent with the fact that there have been no confirmed cases since February, 2019 without additional vaccination efforts \citep{ferguson22}.
The failure of \cite{lee20} to correctly predict the elimination of cholera has been the topic of some discussion \citep{francois20,rebaudetComment20,henrys20,leeReply20}.

Here, we demonstrate that attention to certain statistical considerations could have improved the fit to previous data, as well as leading to forecasts that match more closely the observed outcome.
Based on this retrospective analysis, we offer suggestions on fitting mechanistic models to dynamic systems for future studies.
We proceed by introducing Models 1--3 in Sec.~\ref{sec:models};
in Sec.~\ref{sec:methods}, we present a methodological approach to examining and refining these models.
In Sec.~\ref{sec:results}, we use improved model fits to project cholera incidence in Haiti under various vaccination campaigns.
We then conclude with a discussion on the use of mechanistic models to inform policy decisions in Sec.~\ref{sec:discussion}.

\begin{figure}[ht]
<<Plot_Reported_Cases, echo=FALSE, fig.height=3.5>>=
plot_df <- haitiCholera %>%
  select(-report) %>%
  mutate(date = as.Date(date_saturday)) %>%
  select(-date_saturday) %>%
  pivot_longer(
    data = .,
    cols = -c(date),
    names_to = 'Departement',
    values_to = "Cases",
  )

ggplot(plot_df, aes(x = date, y = Cases + 1)) +
  facet_wrap(~Departement, nrow = 2, labeller = dep_labeller) +
  geom_line() +
  theme(
    axis.title.x = element_blank()
  ) +
  ylab('Reported Cases') +
  scale_y_log10(labels = scales::trans_format("log10", scales::math_format(10^.x))) +
  scale_x_date(date_labels = "'%y", breaks = seq.Date(from = as.Date("2011-01-01"), as.Date("2019-01-01"), by = '2 years'))

@
\caption{\label{CholeraData}
\small
Reported Cholera cases in the outbreak in Haiti from 2010-2019.
}
\end{figure}

% prev_vacc_df <- data.frame(
%   'Departement' = c(
%     'Artibonite',
%     'Centre',
%     'Grand.Anse',
%     'Ouest',
%     'Sud'
%   ),
%   start = as.Date(NA),
%   end = as.Date(NA)
% )
%
% # Grande Anse previous vaccs
% prev_vacc_df[prev_vacc_df$Dep == 'Grand.Anse', 'start'] = as.Date("11-8-2016", format = "%m-%d-%Y")
% prev_vacc_df[prev_vacc_df$Dep == 'Grand.Anse', 'end'] = as.Date("06-02-2017", format = "%m-%d-%Y")
%
% # Sud previous vaccs
% prev_vacc_df[prev_vacc_df$Dep == 'Sud', 'start'] = as.Date("11-8-2016", format = "%m-%d-%Y")
% prev_vacc_df[prev_vacc_df$Dep == 'Sud', 'end'] = as.Date("06-02-2017", format = "%m-%d-%Y")
%
% # Ouest previous vaccs
% prev_vacc_df[prev_vacc_df$Dep == 'Ouest', 'start'] = as.Date("07-25-2017", format = "%m-%d-%Y")
% prev_vacc_df[prev_vacc_df$Dep == 'Ouest', 'end'] = as.Date("08-25-2017", format = "%m-%d-%Y")
%
% # Centre previous vaccs
% prev_vacc_df[prev_vacc_df$Dep == 'Centre', 'start'] = as.Date("11-15-2017", format = "%m-%d-%Y")
% prev_vacc_df[prev_vacc_df$Dep == 'Centre', 'end'] = as.Date("12-16-2017", format = "%m-%d-%Y")
%
% # Artibonite previous vaccs
% prev_vacc_df[prev_vacc_df$Dep == 'Artibonite', 'start'] = as.Date("04-17-2018", format = "%m-%d-%Y")
% prev_vacc_df[prev_vacc_df$Dep == 'Artibonite', 'end'] = as.Date("05-15-2018", format = "%m-%d-%Y")
%

\section{Models}\label{sec:models}
We refer to models that focus on learning relationships between variables in a dataset as associative, and we refer to models that incorporate a known scientific property of the system as causal or mechanistic.
The danger in using forecasting techniques which rely on associative models to predict the consequence of interventions is called the Lucas critique in an econometric context.
\citet{lucas76} pointed out that it is naive to predict the effects of an intervention on a given system based entirely on historical associations.
To successfully quantify the effect of an intervention, a model should therefore both provide a quantitative explanation of existing data and should additionally have a causal interpretation: a manipulation of the system should correspond quantitatively with the corresponding change to the model.
We therefore focus our analysis on mechanistic statistical models, which provides a statistical fit to the available data while also supporting a causal intepretation.

The deliberate limitation of coordination between the groups of \citet{lee20} allows us to treat the models as fairly independently developed expert approaches to understanding cholera transmission.
However, it results in some differences in notation, and choices of data subsets for analysis, that hinder direct comparison.
Here, we have put all three models into a common notational framework.
Translations back to the original notation of \citet{lee20} are given in Table~S-1.

Each model describes the cholera dynamics using a latent state vector $\bm{X}^{(\modelCounter)}(t)$ and an observable response vector $\bm{Y}^{(\modelCounter)}(t)$ for each continuous time-point $t \in \mathcal{T}$, where $\mathcal{T}$ is the set of all time-points and $\modelCounter \in \{1, 2, 3\}$ indexes the model.
The latent state vector $\bm{X}^{(\modelCounter)}$ consists of individuals labeled as susceptible (S), infected (I), asymptomatically infected (A), vaccinated (V), and recovered (R) for $\modelCounter \in \{1, 2, 3\}$, with various sub-divisions sometimes considered across the three models under consideration.
Models~2 and~3 have metapopulation structure, meaning that each individual is a member of a spatial unit, denoted by $u\in \seq{1}{U}$, which corresponds here to membership of the $U=10$ administrative departments.
% Model~1 describes the dynamics aggregated at the national level.

In this study, we focus our attention on important statistical, rather than scientific, details that may improve inference on a dynamic system;
we therefore refrain from commenting on the assumptions in each model or proposing our own alternative description of cholera dynamics, and instead rely on the scientific expertise of \cite{lee20}.

%%%%%%%%% 11111111111 %%%%%%%%%%

\subsection{Model~1}
\label{sec:model1}
$\bm{X}^{(1)}(t) = \big(S_{\vaccCounter}(t),  E_{\vaccCounter}(t), I_{\vaccCounter}(t), A_{\vaccCounter}(t), R_{\vaccCounter}(t), \vaccCounter \in 0:\vaccClass\big)$ describes susceptible, latent (exposed), infected (and symptomatic), asymptomatic, and recovered individuals in vaccine cohort $\vaccCounter$.
Here, $\vaccCounter=0$ corresponds to unvaccinated individuals, and $\vaccCounter \in \seq{1}{\vaccClass}$ describes hypothetical vaccination programs.

The force of infection is
\begin{equation}
\label{model1:lambda}
\lambda(t) = \Big(\sum_{\vaccCounter=0}^{\vaccClass}  I_{\vaccCounter}(t) + \asymptomRelativeInfect \sum_{\vaccCounter=0}^{\vaccClass} A_{\vaccCounter}(t) \Big)^\nu \frac{d\Gamma(t)}{dt} \transmission(t)/N
\end{equation}
where $\transmission(t)$ is a periodic cubic spline representation of seasonality, given in terms of a B-spline basis $\{ s_j(t), j\in \seq{1}{6}\}$ and parameters $\transmission_{1:6}$ as
\begin{equation}
\label{model1:beta}
\log\transmission(t) = \sum_{j=1}^6 \transmission_j s_j(t).
\end{equation}
The process noise $d\Gamma(t)/dt$ is multiplicative Gamma-distributed white noise, with infinitesimal variance parameter $\sigmaProc^2$, which was held constant at $\sigmaProc^2 = 0$ in \cite{lee20}.
The inclusion of gamma white noise in the process model permits the construction of an over-dispersed Markov chain \citep{breto11}, and has previously been found to increase a model's ability to fit data \citep{stocks20}.

Per-capita transition rates are given in Equations~\ref{model1:SE}-\ref{model1:birth}:
\begin{eqnarray}
\label{model1:SE}
\mu_{S_{\vaccCounter}E_{\vaccCounter}} &=& \lambda(t)
\\[-3pt]
\label{model1:EI}
\mu_{E_{\vaccCounter}I_{\vaccCounter}} &=& \muEI\big(1-\symptomFrac_{\vaccCounter}(t)\big)
\\[-3pt]
\label{model1:EA}
\mu_{E_{\vaccCounter}A_{\vaccCounter}} &=& \muEI\, \symptomFrac_{\vaccCounter}(t)
\\[-3pt]
\label{model1:toR}
\mu_{I_{\vaccCounter}R_{\vaccCounter}} &=& \mu_{A_{\vaccCounter}R_{\vaccCounter}} = \muIR
\\[-3pt]
\label{model1:RS}
\mu_{R_{\vaccCounter}S_{\vaccCounter}} &=& \muRS
\\[-3pt]
\label{model1:vacc}
\mu_{S_0S_{\vaccCounter}} &=& \mu_{E_0E_{\vaccCounter}} = \mu_{I_0I_{\vaccCounter}} = \mu_{A_0A_{\vaccCounter}} = \mu_{R_0R_{\vaccCounter}} = \eta_{\vaccCounter}(t)
\\[-3pt]
\label{model1:death}
\mu_{S_{\vaccCounter}\demography} &=& \mu_{E_{\vaccCounter}\demography} = \mu_{I_{\vaccCounter}\demography} = \mu_{A_{\vaccCounter}\demography}=\mu_{R_{\vaccCounter}\demography} = \delta
\\
\label{model1:birth}
\mu_{\demography S_0} &=& \muBirth
\end{eqnarray}
where $\vaccCounter\in \seq{0}{\vaccClass}$.
Here, $\mu_{AB}$ is a transition rate from compartment $A$ to $B$.
We have an additional demographic source and sink compartment $\demography$ modeling entry into the study population due to birth or immigration, and exit from the study population due to death or immigration.
Thus, $\mu_{A\demography}$ is a rate of exiting the study population from compartment $A$ and $\mu_{\demography B}$ is a rate of entering the study population into compartment $B$.
% \eic{TODO: THE LEE ET AL SUPPLEMENTARY EQUATIONS, P14, HAVE A BIRTH RATE INTO THE VACCINE COHORTS, WHICH IS PRESUMABLY A TYPO. IT DOES NOT SHOW IN THE DIAGRAM OR THE CODE.}
% \arc{Yes, I think it is a typo. It is not in the code, and it does not really make much sense.}

In this model, the advantage afforded to vaccinated individuals is an increased probability that an infection is asymptomatic.
Conditional on infection status, vaccinated individuals are also less infectious than their non-vaccinated counterparts by a rate of $\asymptomRelativeInfect = 0.05$ in Eq.~\eqref{model1:lambda}.
In \eqref{model1:EA} and~\eqref{model1:EI} the asymptomatic ratio for non-vaccinated individuals is set $\symptomFrac_0(t)=0$, so that the asymptomatic route is reserved for vaccinated individuals.
For $\vaccCounter\in\seq{1}{\vaccClass}$, the vaccination cohort $\vaccCounter$ is assigned a time $\tau_{\vaccCounter}$, and we take $\symptomFrac_{\vaccCounter}(t) = c \, \theta^*(t-\tau_{\vaccCounter})$
% \begin{equation}
% \label{model1:theta_{\vaccCounter}}
% \symptomFrac_{\vaccCounter}(t) = c \, \theta^*(t-\tau_{\vaccCounter})
% \end{equation}
% \arc{I think we could put this in Table 2 and replace this chunk with a sentence or two defining $\symptomFrac_0(t)$ in words as the proportion of exposed individuals who develop into symptomatic cases and adding a reference to Table 2. This would save a good quarter of a page or so.}
where $\theta^*(t)$ is efficacy at time $t$ since vaccination for adults, taken from \citet{lee20}, Table~S4, and $c=\big(1-(1-0.4688)\times 0.11\big)$ is a correction to allow for reduced efficacy in the 11\% of the population aged under 5 years.
Single and double vaccine doses were modeled by changing the waning of protection; protection was assumed to be equal between single and double dose until 52 weeks after vaccination, at which point the single dose becomes ineffective.

%%%%%%% 222222222222 %%%%%%%%%%

\subsection{Model~2}
\label{sec:model2}
Susceptible individuals are in compartments $S_{u\vaccCounter}(t)$, where $u\in\seq{1}{U}$ corresponds to the $U=10$ departments, and $\vaccCounter\in\seq{0}{4}$ describes vaccination status:
\begin{itemize}
  \item[$\vaccCounter=0$:] Unvaccinated or waned vaccination protection.
  \item[$\vaccCounter=1$:] One dose at age under five years.
  \item[$\vaccCounter=2$:] Two doses at age under five years.
  \item[$\vaccCounter=3$:] One dose at age over five years.
  \item[$\vaccCounter=4$:] Two doses at age over five years.
\end{itemize}

Individuals can progress to a latent infection $E_{u\vaccCounter}$ followed by symptomatic infection $I_{u\vaccCounter}$ with recovery to $R_{u\vaccCounter}$ or asymptomatic infection $A_{u\vaccCounter}$ with recovery to $R^A_{u\vaccCounter}$.
The force of infection depends on both direct transmission and an aquatic reservoir, $W_u(t)$, and is given by
\begin{equation}
\label{model2:lambda}
\lambda_{u}(t) = 0.5\big(1+\seasAmplitude \cos(2\pi t)\big)
\frac{\beta_W\, W_u(t)}{ \Wsat  + W_u(t)} +
\transmission \left\{\sum_{\vaccCounter=0}^4 I_{u\vaccCounter}(t) + \asymptomRelativeInfect \sum_{\vaccCounter=0}^4 A_{u\vaccCounter}(t) \right\}
\end{equation}
The latent state is therefore described by the vector $\bm{X}^{(2)}(t) = \big(S_{u\vaccCounter}(t),\allowbreak E_{u\vaccCounter}(t),\allowbreak I_{u\vaccCounter}(t),\allowbreak A_{u\vaccCounter}(t),\allowbreak R_{u\vaccCounter}(t),\allowbreak R_{u\vaccCounter}^A(t),\allowbreak W_u,\allowbreak u \in \seq{1}{U},\allowbreak \vaccCounter \in \seq{0}{4}\big)$.
A cosine term is included in Eq.~\myeqref{model2:lambda} in order to account for annual seasonality.
Here we note the addition of a phase parameter $\phaseParm$ to allow for a possible shift in the seasonality component in the force of infection; the original implementation of Model~2 in \cite{lee20} sets $\phaseParm = 0$, not allowing for such a shift.
% In Eq.~\myeqref{model2:lambda}, symptomatic individuals are infectious with a rate of $\transmission$, and their asymptotic counterparts are less infectious by a rate of $\asymptomRelativeInfect$.
% Similarly, asymptomatic individuals have a human-to-water shedding rate $\Wshed$ that is lower by a factor of $\asymptomRelativeShed$.
% TODO: I don't think the commented out section above is needed. - Jesse

Individuals move from department $u$ to $v$ at rate $T_{uv}$, and aquatic cholera moves at rate $T^W_{uv}$.
The nonzero transition rates are
\begin{eqnarray}
\label{model2:mu_SE}
\mu_{S_{u\vaccCounter}E_{u\vaccCounter}} &=& \vaccineEfficacy_\vaccCounter \, \lambda
\\[-3pt]
\label{model2:mu_EI}
\mu_{E_{u\vaccCounter}I_{u\vaccCounter}} &=& \symptomFrac\muEI, \quad \mu_{E_{u\vaccCounter}A_{u\vaccCounter}} = (1-\symptomFrac)\muEI
\\[-3pt]
\label{model2:mu_IR}
\mu_{I_{u\vaccCounter}R_{u\vaccCounter}} &=& \mu_{A_{u\vaccCounter}R^A_{u\vaccCounter}} = \muIR
\\[-3pt]
\label{model2:RS}
\mu_{R_{u\vaccCounter}S_{u\vaccCounter}} &=& \mu_{R^A_{u\vaccCounter}S_{u\vaccCounter}} = \muRS
\\[-3pt]
\label{model2:transport}
\mu_{S_{u\vaccCounter}S_{{\varv}\vaccCounter}} &=& \mu_{E_{u\vaccCounter}E_{{\varv}\vaccCounter}} = \mu_{I_{u\vaccCounter} I_{{\varv}\vaccCounter}} = \mu_{A_{u\vaccCounter}A_{{\varv}\vaccCounter}} = \mu_{R_{u\vaccCounter}R_{{\varv}\vaccCounter}} = \mu_{R^A_{u\vaccCounter} R^A_{{\varv}\vaccCounter}} = T_{u\varv}
\\[-3pt]
\label{model2:omega1}
\mu_{S_{u1}S_{u0}} &=& \mu_{S_{u3}S_{u0}} = \omega_1
\\[-3pt]
\label{model2:omega2}
\mu_{S_{u2}S_{u0}} &=& \mu_{S_{u4}S_{u0}} = \omega_2
\\[-3pt]
\label{model2:to_W}
\mu_{\demography W_u} &=& \Wshed \left\{ \sum_{\vaccCounter=0}^4 I_{u\vaccCounter}(t) + \asymptomRelativeShed \sum_{\vaccCounter=0}^4 A_{u\vaccCounter}(t) \right\}
\\[-3pt]
\label{model2:from_W}
\mu_{W_u\demography} &=& \Wremoval
\\[-3pt]
\label{model2:water_transport}
\mu_{W_uW_{\varv}} &=& w_r T^W_{u\varv}
\end{eqnarray}
In \eqref{model2:transport}, we suppose a gravity model:
\begin{equation}
\label{model2:gravity}
T_{u\varv} = v_{\mathrm{rate}} \times \frac{\mathrm{Pop}_u \mathrm{Pop}_{\varv}}{D_{u\varv}^2},
\end{equation}
where $\mathrm{Pop}_u$ is the mean population for department $u$,
$D_{u\varv}$ is a distance measure estimating average road distance between rancomly chosen members of each population, and $v_{\mathrm{rate}}= 10^{-12}$ was treated as a fixed constant.
In \eqref{model2:water_transport}, $T^W_{u\varv}$ is a measure of river flow between departments.
The unit of $W_u(t)$ is cells per ml, with dose response modeled via a saturation constant of $\Wsat$ in \eqref{model2:lambda}.

%%%%%%% 333333333 %%%%%%%%%%

\subsection{Model~3}
\label{sec:model3}

The latent state is described as $\bm{X}^{(3)}(t) = \big(S_{u\vaccCounter}(t),\allowbreak I_{u\vaccCounter}(t),\allowbreak A_{u\vaccCounter}(t),\allowbreak R_{u\vaccCounter k}(t),\allowbreak W_u(t),\allowbreak u \in \seq{0}{U},\allowbreak \vaccCounter \in \seq{0}{4},\allowbreak k \in \seq{1}{3}\big)$.
Here, $\vaccCounter=0$ corresponds to unvaccinated, $\vaccCounter=2j-1$ corresponds to a single dose on the $j$th vaccination campaign in unit $u$ and $\vaccCounter=2j$ corresponds to receiving two doses on the $j$th vaccination campaign.
$k\in\seq{1}{3}$ models non-exponential duration in the recovered class before waning of immunity.
The force of infection is
\begin{eqnarray}
\label{eq:model3:foi}
\lambda_u(t) &=& \Wbeta{_u} \frac{W_u(t)}{\Wsat+2W_u(t)} + \transmission_u \sum_{{\varv}\neq u}\big(I_{\varv 0}(t)+ \asymptomRelativeInfect A_{\varv 0}(t)\big)
\\[-3pt]
\label{eq:model3:SI}
\mu_{S_{u\vaccCounter}I_{u\vaccCounter}} &=& \symptomFrac \,  \lambda_u \big(1-\eta_{u\vaccCounter}(t)\big) \, d\Gamma/dt
\\[-3pt]
\label{eq:model3:SA}
\mu_{S_{u\vaccCounter}A_{u\vaccCounter}} &=& (1-\symptomFrac) \,  \lambda_u \big(1-\eta_{u\vaccCounter}(t)\big) \,  d\Gamma/dt
\\[-3pt]
\label{eq:model3:IR}
\mu_{I_{u\vaccCounter}R_{u\vaccCounter 1}} &=& \mu_{A_{u\vaccCounter}R_{u\vaccCounter 1}} = \muIR
\\[-3pt]
\label{eq:model3:IS}
\mu_{I_{u\vaccCounter}S_{u0}} &=& \muDeath + \choleraDeath
\\[-3pt]
\label{eq:model3:AS}
\mu_{A_{u\vaccCounter}S_{u0}} &=& \muDeath
\\[-3pt]
\label{eq:model3:RRnext}
\mu_{R_{u\vaccCounter 1}R_{u\vaccCounter 2}} &=& \mu_{R_{u\vaccCounter 2}R_{u\vaccCounter 3}} = 3\muRS
\\[-3pt]
\label{eq:model3:RS}
\mu_{R_{u\vaccCounter k}S_{u0}} &=& \muDeath + 3\muRS \, \mathbf{1}_{\{k=3\}}
\\[-3pt]
\label{eq:model3:water}
\mu_{{\demography}W_u} &=& \big[1 + \seasAmplitude \big(J(t))^r \big] D_i \, \Wshed \big[ I_{u0}(t)+ \asymptomRelativeShed A_{u0}(t) \big]
\\[-3pt]
\label{eq:model3:Decay}
\mu_{W_u\demography} &=& \Wremoval
\end{eqnarray}

As with Model~1, $d\Gamma_u(t)/dt$ is multiplicative Gamma-distributed white noise in \myeqref{eq:model3:SI} and \myeqref{eq:model3:SA}.
In \eqref{eq:model3:water}, $J_u(t)$ is a unit-less measurement of precipitation that has been standardized by dividing the observed rainfall at time $t$ by the maximum recorded rainfall in departement $u$ during the epidemic, and $D_u$ is the average population density.
Demographic stochasticity is accounted for by modeling non-cholera related death rate $\muDeath$ in each compartment, along with an additional death rate $\choleraDeath$ in \myeqref{eq:model3:IS} to account for cholera induced deaths among infected individuals.
We note that all deaths are balanced by births into the susceptible compartment in \myeqref{eq:model3:AS} and \myeqref{eq:model3:RS}, thereby maintaining constant population in each department.
% Movement into each vaccination cohort is determined by previous and future vaccination campaigns.
% Here we note that only Model 3 accounted for previous vaccination efforts in the model.
% TODO: The commented lines above seemed unecessary - Jesse

\section{Methods}\label{sec:methods}

<<Load Model Parameters, echo=FALSE, results='hide'>>=
load('model3/output/PanelLocalAll_PF.rda')  # Final model 3 evaluations
load('model3/output/PanelLocalAll.rda')  # Final model 3 parameters (MIF search)
ll_mod3 <- mif_logLik
MIF_mod3 <- local_MIF2_search
rm(mif_logLik, local_MIF2_search, tLocal)

# load('model1/output/trendResultsJoint.rda')  # Parameters and evaluations
load('model1/output/recal_mcap_results.rda')

p1 <- best_params
rm(best_params)

gc()

stew(
  file = "model2/output/model2_fit.rda", {h2_fit <- fit_haiti2()}
)

# load('model2/output/model2_fit_params.rda')
# p2_epi <- h2_epi_params
# h2_fit$end_params <- h2_end_params
# rm(h2_epi_params, h2_end_params)
# gc()

@

<<table-input,echo=FALSE,eval=T, include=FALSE, message=FALSE>>=

best_m3 <- ll_mod3 %>%
  arrange(-logLik) %>%
  slice_head(n = 1) %>%
  pull(which)

p3 <- coef(MIF_mod3[[best_m3]])
rm(MIF_mod3)
gc()

stew(file = "models.rda", {
  h1 <- haiti1_joint()
  coef(h1) <- p1

  h2_epi <- haiti2(region = 'before')
  h2_end <- haiti2(region = 'after', joint = TRUE)

  coef(h2_epi) <- h2_fit$epi_params
  coef(h2_end) <- h2_fit$end_params
  # p2 <- coef(h2)

  h3 <- haiti3_panel(start_time = "2010-10-23", B0 = TRUE)
  coef(h3) <- p3
  pp3 <- pparams(h3)
  p3u <- pp3$specific
  p3s <- pp3$shared
})
@

\subsection{Model Fitting}

Many dynamic systems of scientific and public interest are highly complex by nature; therefore when developing a mathematical description of a dynamic system, it is often necessary to impose assumptions on the system in order to make inference possible.
One common assumption in epidemiology is a homogeneous mixing population \jwc{REF???}.
Other common considerations include whether the proposed model should be stochastic or deterministic; whether the model should have change points in parameter values or should otherwise make adjustments for changes in the dynamic system; and whether the proposed model should include any spatial heterogeneity at a scale permissible by the observed data.
Ultimately, these decisions will result in a different approach to fitting the model.

The models considered in this study represent a diverse selection of possible modelling decisions.
Each of the models independently describe the cholera dynamics as a Partially Observed Markov Process (POMP), since each model is structured in such a way that, given the partially-observable state of the system at time $n$, the future state of the system is independent of its past.
Because of the spatial structure of Models~2 and~3, we refer to these models as SpatPOMPs \citep{asfaw21arxiv}.

In the following subsections, we describe approaches to fitting mechanistic models and discuss potential trade-offs of the selected model complexity.
We begin by fitting the parameters of Model~2, as the deterministic data generating process makes the task of parameter estimation the easiest of the three considered models.

% \eic{USE SOMETHING OTHER THAN $D$ FOR THE NUMBER OF VACCINATION CLASSES, SINCE $D$ IS USED FOR THE DEATH/EMIGRATION/SINK COMPARTMENT. ALSO, MAYBE USE $B$ FOR A BIRTH/IMMIGRATION/SOURCE COMPARTMENT RATHER THAN $D$ FOR BOTH?}
% \jwc{NEW COMMANDS HAVE BEEN MADE FOR VACCINATION CLASSES (\code{vaccClass}) AND THE COUNTERS (\code{vaccCounter}), SO WE CAN JUST CHANGE THE HEADER AND THE WHOLE DOCUMENT WILL UPDATE.}

\subsubsection{Model~2}

Model~2 is a deterministic model, which can be viewed as a special case of a POMP (or a and SpatPOMP) with no randomness in the dynamic model.
By combining a deterministic process model with a simple Gaussian measurement model, Model~2 reduces model fitting to a least squares calculation over parameters in a set of differential equations.
Deterministic compartment models have a long history in the field of infectious disease epidemiology \citep{kermack1927,brauer2017,varghese21},
and can be justified by asymptotic considerations in a large-population limit \citep{dadlani2020,ndii17}.

\citet{lee20} originally fit two versions of model~2 based on a presupposed change in cholera transmission from a epidemic phase to endemic phase that occurred in March, 2014.
We follow their decision to re-estimate model parameters at this break-point, but we include a requirement that the latent state $\bm{X}^{(2)}(t)$ at the start of the endemic period must be the same as the state at the end of the epidemic period.
This additional constraint is sensible for a mechanistic interpretation of the latent state, though it adds difficulty to the task of obtaining a model fit that closely resembles the observed data.
To combat this added difficulty, and to allow for a possible shift in the seasonality component in the force of infection, we introduce an additional phase parameter $\phaseParm$ in Eq.~\ref{model2:lambda}.
% \begin{equation}
% \lambda_{u}(t) = 0.5\big(1+\seasAmplitude \cos(2\pi t + \phaseParm)\big)
% \frac{\beta_W\, W_u(t)}{ \Wsat  + W_u(t)} +
% \beta \left\{\sum_{d=0}^4 I_{ud}(t) + \asymptomRelativeInfect \sum_{d=0}^4 A_{ud}(t) \right\}\label{eq:model2:phase}
% \end{equation}
% \eic{WE COULD SAVE AN EQUATION BY WRITING THE MORE GENERAL ONE INITIALLY AND SAYING THAT THEY ASSUMED $\phaseParm=0$}

% We also note that, without the inclusion of a phase parameter, there is an implicit assumption is that force of infection is greatest around January and weakest around June, and the addition of this parameter allows for the testing of this assumption.
We further increase model flexibility by fitting the parameter $\Wshed$ rather than treating it as fixed.
We implemented this model using the \code{spatPomp} \code{R} package \citep{asfaw21github}.
The model was then fit using the subplex algorithm, implemented in the  \code{subplex} package \citep{king2020Subplex}.
A comparison of the trajectory of the fitted model to the data is given in Fig~\ref{fig:mod2Traj}.

<<Model 2 Trajectory, echo=FALSE>>=
h2_epi_traj <- trajectory(h2_epi, params = h2_fit$epi_params, format = 'data.frame')
h2_epi_traj$Ctotal <- rowSums(h2_epi_traj[, paste0("C", 1:10)]) * h2_fit$epi_params['Rho']

h2_end_traj <- trajectory(h2_end, params = h2_fit$end_params, format = 'data.frame')
h2_end_traj$Ctotal <- rowSums(h2_end_traj[, paste0("C", 1:10)]) * h2_fit$end_params['Rho']

h2_traj <- rbind(
  h2_epi_traj[, c('year', 'Ctotal')],
  h2_end_traj[, c('year', 'Ctotal')]
) %>%
  mutate(
    date = yearsToDate(year)
  )
@

%
% \begin{figure}[ht]
% ggplot() +
%   geom_line(data = true_agg_cases, aes(x = date, y = log(ReportedAll + 1))) +
%   geom_line(data = h2_traj, aes(x = date, y = log(Ctotal + 1)), col = 'blue') +
%   theme(axis.title.x = element_blank()) +
%   ylab("(Log) Reported Cases") +
%   geom_vline(xintercept = yearsToDate(2014.161), linetype = 'dashed')
% @
% \caption{\label{fig:mod2Traj}
% Simulated trajectory of Model~2 (blue curve) compared to reported cholera incidence (black curve). The dashed vertical line shows where the endemic phase starts, and where parameter values are re-estimated.
% }
% \end{figure}
%

\subsubsection{Model~1}

Another approach to modelling a dynamic system is through probabilistic models.
With a probabilistic model,
% rather than modelling the mean behavior of the system through the use of deterministic differential equations,
we suppose the existence of a joint density $f_{\bm{X}^{(\modelCounter)}_{\seq{0}{N}}, \bm{Y}^{(\modelCounter)}_{\seq{1}{N}}}$, with $\bm{X}^{(\modelCounter)}_{\seq{0}{N}}$ denoting the unobservable Markov process of model $\modelCounter$ at times $\seq{0}{N} = \{0, 1, \ldots, N\}$, and $\bm{Y}^{(\modelCounter)}_{\seq{1}{N}}$ denoting the observable process of the system at times $\seq{1}{N}$.
Under this framework, the observed data $y_{\seq{1}{n}}^*$ are assumed to be a single realization of the model $y_{\seq{1}{n}}^* \sim f_{\bm{X}^{(\modelCounter)}_{\seq{0}{N}}, \bm{Y}^{(\modelCounter)}_{\seq{1}{N}}}(x_{\seq{0}{N}}, y_{\seq{1}{N}}; \theta)$, where $\theta$ is a parameter vector that indexes the model.
Using a probabilistic model results in several advantages, including the ability to estimate the variability of the system.
% which is of great interest to scientists and policy makers.
Furthermore, because each draw from the joint distribution represents a potential outcome of the dynamic system, best/worst case scenarios under the assumptions of the model can be easily obtained via simulation.

Once a model for the system has been proposed, the parameter vector $\theta$ needs to be estimated using the observed data.
There exist several modelling techniques that can be used to obtain maximum likelihood estimates (MLE) of the parameters in stochastic dynamic models, including the EM algorithm, Kalman Filter (and extensions) \citep{evensen09}, and iterated filtering algorithms \citep{ionides15}.
In order to use models that are scientifically meaningful rather than only those that are statistically convenient, we restrict ourselves to parameter estimation techniques for POMP models that have the plug-and-play property, which is that the fitting procedure only requires a way to simulate the latent process instead of the ability to evaluate transition densities \citep{breto09,he10}.
Plug-and-play algorithms include ABC and PMCMC \citep{toni09,andrieu10}, but we focus on likelihood maximization via iterated filtering algorithms, which modify the well-known particle filter \citep{arulampalam02} by performing a random walk for each parameter and particle.
These perturbations are carried out iteratively over multiple filtering operations, using the collection of parameters from the previous filtering pass as the parameter initialization for the next, and decreasing the random walk variance at each iteration.

The use of likelihood-based inference allows for the statistical testing of potential improvements to the model.
% Here we need to discuss the inclusion of the white noise in the force of infection:
% TODO: We should address somewhere why we included the process noise in Model 1. Anna suggested this below:
% We first demonstrate this capability by proposing the inclusion of additional stochasticity in the latent process through the force of infection, similar to that in Model~3. Eq.~\myeqref{model1:lambda} becomes
% \begin{equation}
% \label{model1:lambda_wn}
% \lambda(t) = \Big(\sum_{k=0}^{K}  I_{k}(t) + \asymptomRelativeInfect \sum_{k=0}^{K} A_{k}(t) \Big)^\nu d\Gamma_u(t)/dt \transmission(t)/N
% \end{equation}
% where $d\Gamma_u(t)/dt$ is multiplicative Gamma-distributed white noise.
%
% Part of the appeal of mechanistic models lies in their ability to illustrate a dynamical system.
% Inherent is dynamical systems is randomness, and, in the context of epidemiology, this randomness is a crucial component to the system itself.
% Useful and statistically sound incidence forecasting and prediction of vaccine efficacy are not only dependent upon a reasonable reflection of the natural system in the model; it is necessary for the model to also reflect the quantitative characteristics of the processes in question.
% This need, itself, is dependent upon inclusion of sufficient stochasticity in the model \citep{breto09}.
% For now, we motivate this decision by its presence in Model~3 and leave the more rigorous assessment of this adjustment to the supplement.
% The above can no doubt be improved.
% \arc{I think that evaluating the inclusion of both the noise and the trend parameter in the main text may distract from the main point of this manuscript, which is why I mentioned putting it in the supplement.}
% \eic{The dynamic noise issue is important, but is well described elsewhere so can perhaps be handled mostly by references. Maybe we don't need to do more than mention that Lee et al constrained the noise parameter to be zero, but we found considerable advantage from including the parameter (XXX log units of likelihood) consistent with previous investigations (refs).}
% Transition phrase here?
We demonstrate this capability by proposing a linear trend $\transmissionTrend$ in transmission in Eq.~\myeqref{model1:beta}:
\begin{equation}
\label{model1:betat}
\log\transmission(t) = \sum_{j=1}^6 \transmission_s s_j(t) + \transmissionTrend\bar{t}
\end{equation}
Where $\bar{t}$ is the linear mapping $\bar{t}: [0, N] \rightarrow [-1, 1]$ of the time $t$.
The proposal of a linear trend in transmission is a result of observing an apparent decrease in reported cholera infections from 2012-2019 in Fig.~\ref{CholeraData}.
% The inclusion of a possible trend in transmission is due to observing an apparent continual decrease in reported cholera infections from 2012-2019 can be seen in Fig.~\ref{CholeraData}.
While several factors may contribute to this decrease, one explanation is that case-area targeted interventions (CATIs), which included education sessions, increased monitoring, household decontamination, soap distribution, and water chlorination in infected areas \citep{rebaudet19CATI}, may have greatly reduced cholera transmission \citep{rebaudet21}.

We perform a statistical test to determine whether or not the data indicate the presence of a linear trend in transmissibility.
To do this, we perform a profile-likelihood search on the parameter $\transmissionTrend$ and obtain a confidence interval via a Monte Carlo Adjusted Profile (MCAP) \citep{ionides17}.
Similar to Model~2, this model was originally implemented in two parts: an epidemic phase from October 2010 through March 2015, and an endemic phase from March 2015 onward.
As before, we continue to allow the re-estimation of some model parameters at the start of the endemic phase ($\reportRate, \sigmaProc^2$ and $\obsOverdispersion$) but require that the latent Markov process $X(t)$ carry over from one phase into the next.
The resulting confidence interval for $\transmissionTrend$ is $(\Sexpr{myround(mcap_results$ci[1], digits = 3)}, \Sexpr{myround(mcap_results$ci[2], digits = 3)})$, with the full results displayed in Fig.~\ref{fig:betat}.
These results indicate that the inclusion of a trend improves the model fit to the data.
The reported results for Model~1 in the remainder of this article were obtained with the inclusion of the parameter $\transmissionTrend$.

\begin{figure}[ht]
\centering
<<Beta_trend_Figure, fig.height=2.5, fig.width=3.7, fig.align='center'>>=
ggplot() +
  geom_point(data = maxresults, aes(x = betat, y = logLik)) +
  geom_line(data = mcap_results$fit, aes(x = parameter, y = smoothed), col = 'blue') +
  # geom_line(data = mcap_results$fit, aes(x = parameter, y = quadratic), col = 'red') +
  geom_vline(xintercept = mcap_results$ci[1], linetype = 'dashed') +
  geom_vline(xintercept = mcap_results$ci[2], linetype = 'dashed') +
  # geom_vline(xintercept = mcap_results$quadratic_max, col = 'red') +
  geom_vline(xintercept = mcap_results$mle, col = 'blue') +
  labs(x = "Linear Trend in Transmission", y = 'Log Likelihood') +
  theme(axis.text = element_text(size = 8),
        axis.title = element_text(size = 10))
@
\caption{\label{fig:betat}
Monte Carlo adjusted profile of $\transmissionTrend$.
The blue curve is the profile, the blue line indicates the MLE, and the dashed lines indicate the confidence interval.
}
\end{figure}

Model~1 is implemented using the \code{pomp} package \cite{pomp2}, which contains an implementation of the IF2 algorithm that was used to compute the likelihood profile.
% \arc{It may be worth mentioning that the original Model 1 team also implemented their model with pomp. This may be splitting hairs, but this phrasing makes me think we did the entire implementation ourselves which is not true. Maybe just adding in something like: "Following the original methods of the authors"?}
% \jwc{Thanks for this. I agree that the original wording made it sound like we did the entire thing ourselves. To avoid this, and to not increase the length of this statement, I changed it to "Model~1 is implement...", so that we don't imply who did the implementation, but focusing instead on how it is implemented}
% Performing a profile confidence search of a parameter also results in a MLE calculation, as the value that maximizes the profile likelihood also maximizes the entire likelihood surface \citep{murphy2000}.
Simulations from the fitted model compared to the observed data are given in Fig~\ref{fig:mod1fit}.

\begin{figure}[ht]
\centering
<<Model_1_Sims_Figure, fig.height=2.4, fig.width=4.7, fig.align='center'>>=
sims <- simulate(h1, nsim = 500, format = 'data.frame', seed = 3448931)

quants <- sims %>%
  mutate(is_data = .id == 'data') %>%
  filter(!is_data) %>%
  select(.id, week, cases) %>%
  group_by(week) %>%
  summarize(
    q025 = quantile(cases, probs = 0.025, na.rm = TRUE),
    q50  = quantile(cases, probs = 0.500, na.rm = TRUE),
    q975 = quantile(cases, probs = 0.975, na.rm = TRUE)
  ) %>%
  ungroup() %>%
  mutate(date = lubridate::ymd("2010-10-16") + lubridate::weeks(week))

ggplot() +
  geom_line(data = true_agg_cases, aes(x = date, y = ReportedAll + 1)) +
  geom_line(data = quants, aes(x = date, y = q50 + 1), col = 'blue') +
  geom_ribbon(data = quants, aes(x = date, ymin = q025 + 1, ymax = q975 + 1), alpha = 0.5) +
  theme(axis.title.x = element_blank(),
        axis.text = element_text(size = 8),
        axis.title.y = element_text(size = 10))+
  ylab('Reported cholera cases') +
  geom_vline(xintercept = lubridate::weeks(232) + lubridate::ymd("2010-10-16"), linetype = 'dashed') +
  scale_y_log10(labels = scales::trans_format("log10", scales::math_format(10^.x))) +
  scale_x_date(date_labels = "'%y", breaks = seq.Date(from = as.Date("2011-01-01"), as.Date("2019-01-01"), by = '1 year'))
@
\caption{\label{fig:mod1fit}
Simulations from Model~1 compared to reported cholera cases.
The black curve is observed data, the blue curve is median of 500 simulations from the fitted model, and the vertical dashed line represents break-point when parameters are refit.
}
\end{figure}

<<Clean-up Model 1 Simulations, echo=FALSE, include=FALSE>>=
rm(sims, quants)
gc()
@

\subsubsection{Model~3}
Model~3 is also of a probabilistic model.
In this model, both the latent and observable processes can be factored into department specific processes which interact with each other.
We denote the latent and measurement processes for Model~3 as $\bm{X}^{(3)}(t_{0:N}) = X^{(3)}_{1:U, 0:N}$, and $\bm{Y}^{(3)}(t_{1:N}) = Y_{1:U, 1:N}$.
The decision to model the system via metapopulation models versus a model aggregated to a larger spatial scale is one of great scientific interest, and evidence for the former approach has been provided in previous studies \citep{king15}.
Note that this evidence alone does not automatically discredit conclusions drawn via nationally aggregated models, as an argument can easily be made that one should prefer a simple model over a complex one \citep{saltelli20,green15}.
Still, researchers that intend to use a mechanistic model to describe a dynamic system should design the model to be as realistic as their scientific understanding of the system and their computational abilities permit.
% As such, there is strong motivation to include spatial structure in the model designed to approximate the system.
Fitting scientifically flexible metapopulation models, however, remains a challenging statistical problem;
this is due to the fact that the approximation error of particle filters grows exponentially in the dimension of the model \citep{rebeschini15,park20}.
Algorithms that are based on the particle filter therefore become computationally intractable as the number of spatial units increase.

Parameters that must be fit in Model~3 are primarily shared between each departement, the exception to this being the parameters $\beta_{W_u}$, and $\beta_u$, which are unique for each departement $u \in \seq{1}{10}$.
To fit this model, \citet{lee20} simplified the parameter estimation problem by fitting independent departement-level models to the data.
The shared parameters were calibrated using the cholera incidence data from Artibonite, and the departement-specific parameters ($\beta_{W_u}$ and $\beta_u$) were fit using the data from their respective departement.
Reducing a spatially-heterogeneous model to individual units in this fashion requires special treatment of any interactive mechanisms between spatial units, such as found in Eq.~\myeqref{eq:model3:foi}.
In particular, when considering a model for departement $u$, the values $I_\nu(t)$ and $A_\nu(t)$, are unknown for $u \neq v \in \seq{1}{10}$.
A first order approximation of Eq.~\myeqref{eq:model3:foi} for each departement $v \in \seq{1}{10}$ can be obtained using the weekly number of observed cholera cases in each departement:
\begin{equation}
  I_v(t) + A_v(t) \approx \frac{365}{7\reportRate}y^*_v(t)\left(\frac{1}{\muDeath + \choleraDeath + \muIR} + \frac{1 - \symptomFrac}{\symptomFrac(\muDeath + \muIR)}\right)\label{eq:model3:approx}
\end{equation}
This approximation leads to department-specific models that are conditionally independent given the reported number of cholera infections in the remaining departments.
Here, we refer to a collection of POMP models that are independent across units as a PanelPOMP.

In the case of a PanelPOMP, an extension of the IF2 algorithm, known as Panel Iterated Filtering (PIF) \citep{breto20}, can be used to obtain the MLE, which solves the curse of dimensionality for this class of models.
A major advantage of this algorithm over fitting each unit-specific model separately is that PIF can be used to fit both unit-specific parameters and shared parameters; in this way, the calibration of shared parameters involves all of the available data instead of just an arbitrarily chosen subset.

We then fit the panel version of Model~3 using a slight modification of the PIF algorithm, which we call the Block Panel Iterated Filter (BPIF).
The BPIF algorithm is implemented in the \code{panelPomp} package \citep{breto20panelPomp} in the R programming language, and can be used by adding the argument \code{block = TRUE} in the \code{mif2} function.
pseudo-code for this algorithm is provided in Algorithm~S1 in the supplementary material.
% Even with the simplification of Eq.~\myeqref{eq:model3:approx}, a significant amount of computational effort is needed to obtain a proper model fit.
Therefore the reported parameters were obtained using the PanelPOMP version of Model~3, and simulations for the various vaccination campaigns (Sec.~\ref{sec:results}) were obtained using these same parameters in the SpatPOMP version of the model.
% Simulations from the fitted version of the PanelPOMP model are shown in Fig.~\ref{h3panelsims}.

<<Model 3 PanelSims,echo=FALSE>>=
departements <-
  c(
    'Artibonite',
    'Centre',
    'Grande_Anse',
    'Nippes',
    'Nord',
    'Nord-Est',
    'Nord-Ouest',
    'Ouest',
    'Sud',
    'Sud-Est'
  )

registerDoRNG(5317865)

stew(
  file = "model3/output/sims_FinalPanel.rda", {
    foreach(dep=departements, .combine = rbind) %dopar% {
      SIRB <- unitobjects(h3)[[dep]]
      shared <- h3@shared
      specific <- h3@specific[, dep]

      pomp::simulate(SIRB,
                     params = c(shared, specific),
                     nsim = 500, format = 'data.frame',
                     include.data = TRUE) -> sims

      sims$dep <- dep
      sims
    } -> h3_panel_all_sims

    h3_panel_quantiles <- h3_panel_all_sims %>%
      mutate(isdata = .id == 'data') %>%
      # filter(!isdata) %>%
      group_by(isdata, dep, time) %>%
      summarise(
        q05 = quantile(cases, 0.025, na.rm = T),
        mean = mean(cases, na.rm = T),
        q50 = quantile(cases, 0.5, na.rm = T),
        q95 = quantile(cases, 0.975, na.rm = T)
      )

    h3_panel_real_data <- h3_panel_quantiles %>%
      filter(isdata)

    h3_panel_sim_data <- h3_panel_quantiles %>%
      filter(!isdata)

  }
)

h3_panel_sim_data <- h3_panel_sim_data %>%
  mutate(
    dep = gsub("-", "_", dep),
    date = yearsToDateTime(time)
  ) %>%
  filter(date >= yearsToDate(h3@unit.objects$Artibonite@t0)) %>%
  mutate(date = as.Date(lubridate::round_date(date)))

h3_panel_real_data <- h3_panel_real_data %>%
  mutate(
    dep = gsub("-", "_", dep),
    date = yearsToDateTime(time)
  ) %>%
  filter(date >= yearsToDate(h3@unit.objects$Artibonite@t0)) %>%
  mutate(date = as.Date(lubridate::round_date(date)))

plot_order <- c(
  'Artibonite',
  'Sud_Est',
  'Nippes',
  'Nord_Est',
  'Ouest',
  'Centre',
  'Nord',
  'Sud',
  'Nord_Ouest',
  'Grande_Anse'
)

@

\begin{figure}[ht]
<<Model3_PanelSims_Figure, echo=FALSE, fig.height=3>>=

ggplot() +
  geom_line(data = h3_panel_sim_data, aes(x = date, y = q50 + 1), col = 'blue') +
  geom_line(data = h3_panel_real_data, aes(x = date, y = q50 + 1), col = 'black') +
  geom_ribbon(data = h3_panel_sim_data, aes(x = date, ymin = q05 + 1, ymax = q95 + 1),
              alpha = 0.4) +
  facet_wrap(~factor(dep, levels = plot_order), nrow = 2,
             labeller = dep_labeller) +
  ylab('Reported cholera cases') +
  theme(axis.title.x = element_blank()) +
  scale_x_date(date_labels = "'%y", breaks = seq.Date(from = as.Date("2011-01-01"), as.Date("2019-01-01"), by = '2 years')) +
  scale_y_log10(labels = scales::trans_format("log10", scales::math_format(10^.x)))
  # theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1,
  #                                  size = 7),
        # axis.title.x = element_blank())
@
\caption{\label{h3panelsims}
Simulations from initial conditions using panelPOMP version of Model~3.
The black curve represents true case count, the blue curve the median of 500 simulations from the model, and the grey ribbons representing $95\%$ confidence interval.
}
\end{figure}

<<Remove Model3 panelSims, cache=FALSE, include=FALSE, message=FALSE, echo=FALSE>>=
rm(h3_panel_all_sims, h3_panel_sim_data)
gc()
@

Fitting Model~3 simplifies the problem of parameter estimation, but it also introduces additional technicalities that must be addressed.
One concern is that of obtaining model projections, which was the primary goal of the original modeling exercise.
% "primary output" -> "primary goal", based on Anna's suggestion.
The simplified PanelPOMP version of Model~3 relies on the observed cholera cases as a covariate, which are unavailable for use in projections.
To address this, we use the MLE obtained for the PanelPOMP approximation of Model~3 as an estimate for the parameters in fully coupled version of the model, which was implemented using the \code{spatPomp} package.
Simulations from the SpatPOMP version of the model, using the parameters that were fit with the PanelPOMP version of the model, are displayed in Fig.~\ref{h3spatsims}.

<<Model 3 SpatPOMP sims, echo=FALSE>>=
# load('output/FinalPanelAsSpatPomp_PF.rda')

h3Spat <- haiti3_spatPomp()
coef(h3Spat) <- haitipkg:::panelParms_toSpatParms3(p3)

registerDoRNG(6687147)

stew(
  file = "model3/output/SpatPompSims.rda", {
    h3Spat_sims <- simulate(h3Spat, nsim = 500, seed = 784713,
                     format = "data.frame", include.data = TRUE)

    h3Spat_quants <- h3Spat_sims %>%
      mutate(is.data = .id == 'data') %>%
      filter(!is.data) %>%
      rename(dep = unitname) %>%
      group_by(dep, time) %>%
      summarise(q05 = quantile(cases, 0.025, na.rm = T),
                mean = mean(cases, na.rm = T),
                q50 = quantile(cases, 0.5, na.rm = T),
                q95 = quantile(cases, 0.975, na.rm = T)) %>%
      ungroup() %>%
      mutate(date = yearsToDateTime(time)) %>%
      filter(date >= yearsToDate(h3Spat@t0)) %>%
      mutate(date = as.Date(lubridate::round_date(date)))

    h3Spat_truth <- h3Spat_sims %>%
      # select(time, .id, starts_with("cases")) %>%
      mutate(is.data = .id == 'data') %>%
      filter(is.data) %>%
      rename(dep = unitname) %>%
      group_by(dep, time) %>%
      summarise(q05 = quantile(cases, 0.025, na.rm = T),
                mean = mean(cases, na.rm = T),
                q50 = quantile(cases, 0.5, na.rm = T),
                q95 = quantile(cases, 0.975, na.rm = T)) %>%
      ungroup() %>%
      mutate(date = yearsToDateTime(time)) %>%
      filter(date >= yearsToDate(h3Spat@t0)) %>%
      mutate(date = as.Date(lubridate::round_date(date)))

  }
)

@

\begin{figure}[ht]
<<Model3_SpatPOMP_sims_Figure, fig.height=3>>=
ggplot() +
  geom_line(data = h3Spat_quants, aes(x = date, y = q50 + 1), col = 'blue') +
  geom_line(data = h3Spat_truth, aes(x = date, y = q50 + 1), col = 'black') +
  geom_ribbon(data = h3Spat_quants,
              aes(x = date, ymin = q05 + 1, ymax = q95 + 1),
              alpha = 0.4) +
  facet_wrap(~factor(dep, levels = plot_order), nrow = 2,
             labeller = dep_labeller) +
  labs(y = 'Log Reported Cholera Cases') +
  scale_x_date(date_labels = "'%y", breaks = seq.Date(from = as.Date("2011-01-01"), as.Date("2019-01-01"), by = '2 years')) +
  theme(axis.title.x = element_blank()) +
  scale_y_log10(labels = scales::trans_format("log10", scales::math_format(10^.x)))
  # theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1,
  #                                  size = 6),
  #       axis.title.x = element_blank())
@
\caption{\label{h3spatsims}
Simulations from initial conditions using SpatPOMP version of Model~3.
The black curve represents true case count, the blue line the median of 500 simulations from the model, and the grey ribbons representing $95\%$ confidence interval.
}
\end{figure}

<<Remove Model3 spatSims, cache=FALSE, include=FALSE, message=FALSE, echo=FALSE>>=
rm(h3Spat_sims, h3Spat_truth, h3Spat_quants)
gc()
@

While simulations from the SpatPOMP version of the model seem similar to the observed data, we note in Table~\ref{tab:likes} that using the MLE obtained using the approximation in Eq.~\myeqref{eq:model3:approx} results in a lower likelihood value for the SpatPOMP version of the model than the PanelPOMP version.
A natural question to ask is if the MLE obtained using the approximation in Eq.~\myeqref{eq:model3:approx} could result in a drastically different set of parameters than those that would be obtained by estimating the MLE in the fully coupled model.
Unfortunately, the answer to this question would require the ability to accurately and reliably estimate the MLE of the fully coupled model, which is beyond the scope of this article.
% this question is unsatisfactorily left unanswered, because the answer would require knowledge of the the MLE without the simplification, which is an open statistical question and beyond the scope of this article.
% While this approach may provide a reasonable estimate of the model parameters, it is unknown by how much this estimate may be improved.
This highlights the need for advancements in statistical methodology that permit inference on models with coupled metapopulation dynamics.

% \subsection{Comparing mechanistic models with associative benchmarks} %AAAAAAAAAAAAAAA
\subsection{Model Diagnostics}

\jwc{How about this edit?}
Researchers may use fitted models to make inference and draw conclusions about the dynamic system under investigation.
Because of the casual nature of mechanistic models, the validity of these conclusions is dependent on the ability of the proposed model to approximate the system in question.
It is often impossible to know, however, the extent to which a proposed model provides a faithful representation of the system.
In such cases, we rely on the quantitative and statistical ability of the model to match the observed data, which we call the model's goodness-of-fit.
Goodness-of-fit provides evidence supporting the causal interpretation of a model, but cannot by itself rule out the possibility of alternative explanations.

% In the absence of precise scientific knowledge on a dynamic system, a model's goodness-of-fit is often the best tool for determining the model's ability to faithfully represent the dynamic system.
% we provide a detailed discussion on assessing a model's goodness-of-fit in this subsection.

One common approach to assess a mechanistic model's goodness-of-fit is to compare several simulations from the fitted model to the observed data;
% similar to those in Figures~\ref{fig:mod1fit}--\ref{h3panelsims};
% TODO: citation here? \citep{???}
this is done to determine whether or not the observed data are a plausible realization of the fitted model.
While these comparisons are useful, relying only on these qualitative comparisons may lead to over confidence on the goodness-of-fit.
A relevant example is the original parameter estimates provided by \citet{lee20}:
these parameters resulted in simulations that closely resembled the observed data, yet resulted in model likelihoods that were---in some cases---drastically smaller than likelihoods that can be achieved via the likelihood based optimization techniques that were used (see Table~\ref{tab:likes}).
% \eic{TODO: Could we maybe put an example of this here? }
Alternative forms of model validation should therefore be used in conjunction with visual comparisons of simulations to observed data.

The goodness-of-fit of a model can be assessed by comparing a quantitative description of the model fit (such as MSE, predictive accuracy, or model likelihood) to that of other proposed models.
This type of comparison provides insight into how each model performs relative to the others, but does not provide insight on the performance of each model on a given dataset unless the performance of one of the models is widely understood.
% "each other" -> "the others", based on Anna's suggestion.
We refer to such a model as a benchmark, and examples of these models include logistic regression and ordinary least squares.

Comparing associative models using an objective function seems natural, as these models are designed to optimize objective functions given the available data.
In the case of mechanistic modeling, however, comparing models based on an objective measure is less natural because each mechanistic model is designed to provide a scientific description of the dynamic system.
Any comparison between mechanistic models would therefore ideally be based on the scientific validity of each proposed model, which is often impossible to know.
Still, parameters in mechanistic models are fit using techniques that are designed to maximize an objective
% TODO: Anna suggests "objective function". Maybe "objective value" as well, or is it okay to leave as it is?
such as likelihood, and it is expected that a good scientific description of the dynamic system should result in more optimal values of an objective evaluation than would a poor description of the system.
In this sense, one could justify comparing mechanistic models based on some objective measurement such as model likelihoods.
% When performing such a comparison, however,
We do not suggest that the model with the highest likelihood given the data automatically provides a better scientific description of the dynamic system;
still, a comparison based on model likelihoods may help guide researchers to know the plausibility of a proposed model compared to another given their relative ability to describe the observed data.

This comparison still only provides a relative measure of model fit compared to other mechanistic models.
The resulting challenge is that no general benchmark exists for mechanistic models that would allow researchers to assess the goodness-of-fit by comparing to a known standard.
In some cases, a possible benchmark model could be a generally accepted truth, but often no such truth is available.
% In the case of Cholera infections in Haiti, for example, it has been debated whether or not bacteria buildup in the environment is a cause of Cholera infections \cite{???}.
Because of this, we propose the use of an ARMA(2, 1) model as a potential benchmark.
Despite being an associative model, the reason that an ARMA(2, 1) model is an appealing benchmark is because the performance of this class of models is widely understood, and often provides a good description of the data with very limited effort.
% For us, the defining feature of an associative model is that it is assessed primarily on its quantitative ability to explain historical data.
% Mechanistic models, on the other hand, must provide a scientific hypothesis about the system as well as having some quantitative ability to explain observed data.
% By this definition, mechanistic models are a subset of associative models;
% this subset relationship is enough to see that one cannot expect the class of mechanistic models to provide the highest levels of statistical fit.
% In general one should expect a statistical cost to the constraint of requiring mechanistic interpretation of the model.
One question that follows is whether it is useful to assess mechanistic models by their statistical fit, if this is not their primary purpose?
To what extent can we say the data support a scientific model ($A$) over another model ($B$) based on improved fit to data, if neither fit better than an associative model ($C$)?
% An answer is to think in terms of the shortfall between a mechanistic model and the best known associative model.
% If the shortfall is small, on a suitable scale, then the mechanistic model may be considered a valid view of the system from the point of view of these data.
% While the ARMA(2, 1) certainly will not automatically provide the best fit of all associative models,
% This does not demonstrate the correctness of the mechanistic interpretation, but also it does not refute it.
% If the shortfall is large, we conclude that there is some substantial aspect of the system that is not incorporated into the model.
% we cannot necessarily tell which part of the model is invalid, but we know there is a weakness somewhere.
% To assess whether shortfalls are small or large, we may consider using formal statistical hypothesis tests or model selection techniques.
% These can give us a scale for chance variation: a difference small enough to be explained by randomness in the data generating mechanism should not be considered as strong evidence.
% However, the comparison of primary interest is one of practical significance rather than statistical significance.
% We would like a mechanistic model whose statistical fit is close to the best associative models regardless of whether there is enough statistical power to distinguish the difference.

% I like the above addition because it provides a little bit more of the philosophy behind the paragraph below. Perhaps we can combine them? - Anna

When comparing mechanistic models to non-mechanistic benchmarks, we do not require the benchmark model to optimize a measure of statistical fit, nor do we suggest that a mechanistic model must have superior fit to the benchmark in order to be of scientific interest.
It is possible that a successful mechanistic model may have better statistical fit than simple benchmarks such as log-linear models;
still, the specific threshold of beating, or failing to beat, an arbitrary benchmark is similarly arbitrary.
In this case, however, a large shortfall in the fit of a mechanistic model to that of a log-linear model may demonstrate room for substantial improvement.
The use of a common benchmark may also be beneficial when developing models with varying spatial scale, as direct comparisons between models fit to data with different levels of spatial aggregation are meaningless (e.g. comparing Model~1 to Model~3).
Likelihoods of Models~1--3 and their respective ARMA(2, 1) benchmark models are provided in Table~\ref{tab:likes}.
\jwc{WE SAY THAT WE HAVE AN IN-DEPTH DESCRIPTION OFMODEL VALIDATION, BUT ONLY PROPOSE/DEMONSTRATE TWO WAYS TO DO THIS. SHOULD WE MENTION OTHER THINGS, LIKE CONDITIONAL LIKELIHOODS / ESS, CHECKING PLAUSABILITY OF FIT MODEL PARAMETERS, OR COMPARING AUTO-CORRELATIONS BETWEEN SIMULATIONS AND OBSERVATIONS (LIKE AARON KING HAS DONE)? I DON'T THINK WE NECESSARILY HAVE TO SHOW ALL OF THESE, SINCE THE ARTICLE IS ALREADY TOO LONG, BUT MAYBE MENTIONING THEM? WOULD MENTIONING THESE WITHOUT DEMONSTRATING THEM RAISE CONCERNS FROM REVIEWERS?}

<<Table 2 Input, echo=FALSE, eval=TRUE>>=

# Objective function of model 2, epidemic phase
epi_ofun <- traj_objfun(
  h2_epi,
  params = h2_fit$epi_params
)

# Objective function of model 2, endemic phase
end_ofun <- traj_objfun(
  h2_end,
  params = h2_fit$end_params
)

# Evaluate and sum objective functions to get total model likelihood
mod2_ll <- -(epi_ofun(par = h2_fit$epi_params) + end_ofun(par = h2_fit$end_params))

arima_lik2 <- 0

# model 2 AIC
mod2_n_params <- 21  # Number of parameters fit in Model 2
mod2_aic <- 2 * mod2_n_params - 2 * mod2_ll

# Load pre-calculated model 1 likelihood
mod1_ll <- h1_ll['ll']  # Comes from model1/output/recal_mcap_results.rda

# model 1 AIC
mod1_n_params <- 13
mod1_aic <- 2 * mod1_n_params - 2 * mod1_ll

# Load pre-calculated model 3 likelihood
mod3_ll <- ll_mod3 %>%
  arrange(-logLik) %>%
  slice_head(n = 1) %>%
  pull(logLik)

# model 3 AIC
mod3_n_params <- 39
mod3_aic <- 2 * mod3_n_params - 2 * mod3_ll

load('model3/output/FinalPanelAsSpatPomp_PF.rda')
rm(pf, ll)
mod3_spatPomp_ll <- ll_est[1]
rm(ll_est)
mod3_spatPomp_n_params <- 39
mod3_spatPomp_aic <- 2 * mod3_spatPomp_n_params - 2 * mod3_spatPomp_ll
@

<<CalculateARMA, echo=FALSE, message=FALSE, include=FALSE, cache=TRUE>>=

ARMA_benchmarks <- list()

###
### Model 1 ###
###

# Load aggregated data
m1_agg_data <- haiti1_agg_data()

# Fit ARMA(2, 1) on log-cases
m1_log_arma <- arima(log(m1_agg_data$cases + 1), order = c(2, 0, 1))

# Fit ARMA(2, 1) on natural scale of cases
m1_arma <- arima(m1_agg_data$cases, order = c(2, 0, 1))

# Check which one is better, and save benchmark as the better one
if (m1_log_arma$loglik - sum(log(m1_agg_data$cases + 1), na.rm = TRUE) >= m1_arma$loglik) {
  ARMA_benchmarks[['m1']] <- m1_log_arma$loglik - sum(log(m1_agg_data$cases + 1), na.rm = TRUE)
} else {
  ARMA_benchmarks[['m1']] <- m1_arma$loglik
}

rm(m1_agg_data, m1_log_arma, m1_arma)
gc()

###
### Model 2 ###
###

h2_data <- haiti2_data()

# model 2 ARIMA benchmark
h2_dep_names <- h2_data$department %>% unique()
m2_benchmark <- 0
for (i in 1:10){
  cases <- h2_data[h2_data$department == h2_dep_names[i], ]$cases
  log_cases <- log(cases + 1)

  m2_arma_dep_ll <- arima(cases, order=c(2, 0, 1))$loglik
  m2_log_arma_dep_ll <- arima(log_cases, order=c(2, 0, 1))$loglik - sum(log_cases, na.rm = TRUE)

  m2_benchmark <- m2_benchmark + max(m2_arma_dep_ll, m2_log_arma_dep_ll)
}

ARMA_benchmarks[['m2']] <- m2_benchmark

rm(m2_benchmark, h2_dep_names, cases, log_cases, m2_arma_dep_ll, m2_log_arma_dep_ll, h2_data)
gc()

###
### Model 3 ###
###

MODEL3_CASES <- haitiCholera %>%
  dplyr::rename(
    date = date_saturday, Grande_Anse = Grand.Anse,
    Nord_Est = Nord.Est, Nord_Ouest = Nord.Ouest,
    Sud_Est = Sud.Est
  ) %>%
  dplyr::mutate(date = as.Date(date)) %>%
  dplyr::select(-report) %>%
  dplyr::mutate(time_diff = date - lag(date))

jumps <- which(MODEL3_CASES$time_diff > lubridate::make_difftime(day = 7))

temp <- MODEL3_CASES
for (j in 1:length(jumps)) {
  temp <- rbind(
    temp[1:(jumps[j] + j - 2), ],
    rep(NA, ncol(temp)),
    temp[(jumps[j] + j - 1):nrow(temp), ]
  )
}

mod3_cases <- temp %>%
  dplyr::select(-time_diff)

mod3_departements <-
  c(
    'Artibonite',
    'Centre',
    'Grande_Anse',
    'Nippes',
    'Nord',
    'Nord-Est',
    'Nord-Ouest',
    'Ouest',
    'Sud',
    'Sud-Est'
  )

m3_benchmark <- 0
for (dep in mod3_departements) {

  # Get department cases
  cases <- as.numeric(mod3_cases[, gsub("-", "_", dep)])

  # Fit model on natural scale of cases
  m3_a201 <- arima(cases, order = c(2, 0, 1))

  # Fit model on log-scale of cases
  m3_log_a201 <- arima(log(cases + 1), order = c(2, 0, 1))

  m3_benchmark <- m3_benchmark + max(
    m3_log_a201$loglik - sum(log(cases + 1), na.rm = TRUE),
    m3_a201$loglik
  )
}

ARMA_benchmarks[['m3']] <- m3_benchmark
rm(cases, m3_a201, m3_log_a201, m3_benchmark, mod3_departements, dep,
   mod3_cases, temp, jumps, MODEL3_CASES, i, j)
gc()
@

<<Table 2 Lee Input, echo=FALSE, include=FALSE>>=
load('model1/output/lee_mod1_epidemic_fit.rda')
mod1_lee_epi_ll <- max(prof_if$loglik, na.rm = TRUE)
rm(prof_if)
gc()
load('model1/output/lee_mod1_endemic_fit.rda')
mod1_lee_end_ll <- max(prof_if$loglik, na.rm = TRUE)
rm(prof_if)
gc()

mod1_lee_ll <- mod1_lee_epi_ll + mod1_lee_end_ll
mod1_lee_n_params <- 20
mod1_lee_aic <- 2 * mod1_lee_n_params - 2 * mod1_lee_ll

# load('model2/output/mod2_leeFit.rda')
mod2_lee_aic <- 2 * h2_fit$leeFit_n_params - 2 * h2_fit$leeFit_ll
# mod2_lee_aic <- 2 * mod2_leeFit_num_parms - 2 * mod2_leeFit_ll
@

<<Lee3 Sub Calculations, echo=FALSE, include=FALSE, message=FALSE>>=
registerDoRNG(49637421)

# Create Lee et al (2020a) version of model 3 on sub-data
lee3_sub <- haiti3()

# This model is only fit for a subset of the data, so we will set data
# that are not part of this subset as NA. This will allow us to estimate
# the likelihood for only the values that were considered during the
# parameter estimation phase.
set_as_na <- lee3_sub@times < dateToYears(as.Date('2017-06-10'))
temp_ouest_cases <- lee3_sub@data['casesOuest', ]
temp_ouest_cases[set_as_na] <- NA_real_
lee3_sub@data['casesOuest', ] <- temp_ouest_cases

lee3_all <- haiti3(start_time = "2010-10-23")

rm(set_as_na, temp_ouest_cases)
gc()

stew(
  # Save the file
  file = 'model3/output/lee3_results.rda', {

    # Create lists to save results
    lee3_results <- list()

    # Save how long it takes for sub-data
    lee3_results[["sub_pf_time"]] <- system.time({

      # Save all likelihood evaluations
      lee3_results[["sub_ll_evals"]] <- foreach(
        i = 1:36, .combine = c
      ) %dopar% {
        logLik(pfilter(lee3_sub, Np = 5000))  # Get likelihood evaluations
      }
    })

    registerDoRNG(2789131)

    # Save how long it takes for all data
    lee3_results[["all_pf_time"]] <- system.time({

      # Save all likelihood evaluations
      lee3_results[["all_ll_evals"]] <- foreach(
        i = 1:36, .combine = c
      ) %dopar% {
        logLik(pfilter(lee3_all, Np = 5000))  # Get likelihood evaluations
      }
    })

    lee3_results[["n_params"]] <- 29
  }
)

rm(lee3_sub, lee3_all)
gc()
@

<<old TODO: Replace this with the above>>=
# Load Lee et al Model 3 likelihood for entire data
# load('model3/output/lee_mod3_coupled_allData_ll.rda')
# mod3_lee_coupled_ll <- max(lee3_results[["ll_evals_sub_data"]])
# mod3_lee_coupled_n_params <- 29
mod3_lee_coupled_aic <- 2 * lee3_results$n_params - 2 * max(lee3_results$all_ll_evals)

# rm(coupled_ll_evals_all_data, coupled_ll_results_all_data, tPfilter_lee_coupled)
# gc()

# Load Lee et al Model 3 likelihood on partial data
# load('model3/output/lee_mod3_coupled_ll.rda')
# mod3_lee_coupled_partial_ll <- max(coupled_ll_evals_sub_data)
mod3_lee_coupled_partial_aic <- 2 * lee3_results$n_params - 2 * max(lee3_results$sub_ll_evals)

# Load our spatPOMP Model 3 on likelihood on partial data
load('model3/output/mod3_coupled_subset_ll.rda')
mod3_coupled_subset_aic <- 2 * mod3_spatPomp_n_params - 2 * mod3_coupled_subset_ll
@

\begin{table}[ht]
\centering

\begin{tabular}{|c|c|c|c|c|}
\hline
 & \thead{Model~1} & \thead{Model~2} & \thead{Model~3 \\ (panelPOMP)} & \thead{Model~3 \\ (SpatPOMP)}
\\
\hline
\hline
\multirow{2}{*}{Log Likelihood} &
  $\Sexpr{myround(mod1_ll, 1)}$ &
  $\Sexpr{myround(mod2_ll, 1)}$ &
  \multirow{2}{*}{$\Sexpr{myround(mod3_ll[1], 1)}$\footnotemark[2]} &
  $\Sexpr{myround(mod3_spatPomp_ll, 1)}$ \\
    & ($\Sexpr{myround(mod1_lee_ll, 1)}$)\footnotemark[1] &
  ($\Sexpr{myround(h2_fit$leeFit_ll, 1)}$) &
  & ($\Sexpr{myround(max(lee3_results[["all_ll_evals"]]), 1)}$)\footnotemark[3]
\\
\hline
Number of &
  $\Sexpr{as.character(mod1_n_params)}$ & $\Sexpr{as.character(mod2_n_params)}$ & $\Sexpr{as.character(mod3_n_params)}$ & $\Sexpr{mod3_spatPomp_n_params}$ \\
 Fit Parameters & (\Sexpr{as.character(mod1_lee_n_params)}) & ($\Sexpr{as.character(h2_fit$leeFit_n_params)}$) & (29) & ($\Sexpr{as.character(lee3_results$n_params, 1)}$)
\\
\hline
\multirow{2}{*}{AIC} &
  $\Sexpr{myround(mod1_aic, 1)}$ & $\Sexpr{myround(mod2_aic, 1)}$ & \multirow{2}{*}{$\Sexpr{myround(mod3_aic, 1)}$\footnotemark[2]} & $\Sexpr{myround(mod3_spatPomp_aic, 1)}$ \\
  & ($\Sexpr{myround(mod1_lee_aic, 1)}$)\footnotemark[1] & ($\Sexpr{myround(mod2_lee_aic, 1)}$) &  & ($\Sexpr{myround(mod3_lee_coupled_aic, 1)}$)\footnotemark[3]
\\
\hline
\thead{ARMA(2, 1) \\ Log Likelihood} &
  $\Sexpr{myround(ARMA_benchmarks$m1, 1)}$ &
  $\Sexpr{myround(ARMA_benchmarks$m2, 1)}$ &
  $\Sexpr{myround(ARMA_benchmarks$m3, 1)}$ &
  $\Sexpr{myround(ARMA_benchmarks$m3, 1)}$
\\
\hline
\end{tabular}
\caption{\label{tab:likes}Log-likelihood values for each models compared to their ARMA benchmarks.
Values in parenthesis are corresponding values using \citet{lee20} parameter estimates.}
\end{table}

\subsection{Model Projections}\label{sec:filter}

\footnotetext[1]{The reported likelihood is an upper bound of the likelihood of the \citet{lee20} model as it is the largest likelihood obtained using their parameter calibration regime.}
\footnotetext[2]{Parameters to department-specific models not provided by \citet{lee20}, as department fits were only used as an intermediary step to obtain parameter estimates of the coupled model.}
\footnotetext[3]{Model 3 was originally fit to only a subset of the data starting from March 2014 and did not include a large portion of data from Ouest in 2015-2016.
On this subset, the parameters provided by \citet{lee20} achieved a likelihood of $\Sexpr{myround(max(lee3_results$sub_ll_evals), 1)}$.
On this same subset of data, our model achieved a likelihood of $\Sexpr{myround(mod3_coupled_subset_ll, 1)}$.}
As previously noted, simulations from a well-fit mechanistic model may closely resemble the observed data $y_{1:N}^*$.
In such comparisons, these simulations are random draws from the complete estimated joint distribution $f_{\bm{X}^{(\modelCounter)}_{0:N}, \bm{Y}^{(\modelCounter)}_{1:N}}\left(x_{0:N}, y_{1:N}\mid \hat{\theta}\right)$, where $\modelCounter$ identifies the model used for simulations.
It can therefore be tempting to use simulations from this model up to time $N+s$, $f_{\bm{X}^{(\modelCounter)}_{0:N+s}, \bm{Y}^{(\modelCounter)}_{1:N+s}}\left(x_{0:N+s}, y_{1:N+s}\mid \hat{\theta}\right)$, with $s\geq 0$ to project the dynamic system up to a future time $N+s$, as it has been done in previous studies \citep{lee20,???}.
% \eic{TODO: I think it's possible that other people have done this too. Could we find another reference?};
This approach, however, does not take advantage of the information about the state of the system that is contained in the observed data.
Note that each of the models in question (and, more generally, all models that are described as POMPs) are Markovian, that is, the history of the process $\{\bm{X}^{(\modelCounter)}_s, s\leq t\}$ for model $\modelCounter$ is uninformative about the future of the process $\{\bm{X}^{(\modelCounter)}_s, s\geq t\}$, given the current state $\bm{X}^{(\modelCounter)}_t$.
In other words, observing the data $\bm{Y}^{(\modelCounter)}_N = y_N^*$ at time $N$ provides more information about the future state of the system than does the initial conditions.
A more sound projection of the dynamic system into the future could therefore be obtained by simulating using the most recently known state of the system, that is, obtaining draws from the conditional density $f_{\bm{X}^{(\modelCounter)}_{N:s}, \bm{Y}^{\modelCounter}_{N:s}}\big(x_{N:s}, y_{N:s}\mid \hat{\theta}, \bm{X}^{(\modelCounter)}_{N} = \bm{x}^{(\modelCounter)}_N\big)$.

$\bm{X}^{(\modelCounter)}_N$, however, is unobservable and therefore draws from the desired conditional density are unobtainable.
Possible values of $\bm{X}^{(\modelCounter)}_N$ given the observed data $\bm{Y}^{(\modelCounter)}_{1:N} = y_{1:N}^*$ and the proposed model can easily be obtained, however, via the filtering distribution.
Let $\hat{\bm{X}}^{(\modelCounter), i}_N, \, i \in {1, 2, \ldots, J}$ be $\iid$ draws from the filtering distribution at time $N$, with density $\hat{\bm{X}}^{(\modelCounter), i}_N \sim f_{\bm{X}^{(\modelCounter)}_N|\bm{Y}^{(\modelCounter)}_{1:N}}(x_N \mid \hat{\theta}, y_{1:N}^*)$.
A single model projection can then be obtained by simulating from the model $f_{\bm{X}^{(\modelCounter)}_{N:s}, \bm{Y}^{(\modelCounter)}_{N:s}}\big(x_{N:s}, y_{N:s}\mid \hat{\theta}, \bm{X}^{(\modelCounter)}_{N} = \hat{\bm{X}}^{(\modelCounter), i}_N\big)$.
Intuitively, simulating the model starting at the filtering distribution of the most recently available time point is a more appropriate way to project a stochastic dynamic system into the future, as it is not expected that each simulation from initial conditions will result in a latent state at time $N$ consistent with the model and the observation $\bm{Y}^{(\modelCounter)}_N = y_N^*$.
In this particular case study, projecting the future state of the cholera epidemic in Haiti starting from the draws from the filtering distribution allows the proposed models to benefit from the fact that very few cholera cases had been observed between 2018 and 2019, and that cases appear to be decreasing (i.e., the number of susceptible individuals may be small).
% Model projections given in the following section have been obtained using the technique suggested in this subsection.
% Sentence above is unecessary - Jesse

An additional consideration to make when obtaining model forecasts is that of parameter uncertainty.
It has been noted that the uncertainty in just a single parameter can lead to drastically different projections \citep{saltelli20}.
One possible approach to account for parameter uncertainty in model projections is by obtaining confidence intervals for each parameter, sampling parameters from the confidence intervals, simulating the model with the resulting parameter set, and then weighing the resulting model projections based on the likelihood of the given set of parameters, as was done by \citet{king15}.

Note that in order to obtain projections that are consistent with the observed data, one must first be able to sample from the filtering distribution given each set of parameters.
This approach can be done for both deterministic and stochastic models, but requires a large number of computations, especially as the number of observations and model parameters increase.
Because the focus of this study is on model fitting and evaluation, we do not provide model projections accounting for parameter uncertainty.
Instead, we use the projections from point estimates to highlight a major deficiency of deterministic models, which is that the only variability in model projections is a result of parameter uncertainty, which leads to over-confidence in the projections.
This observation is consistent with those made in \citet{king15}, and suggests that stochastic models should be preferred over deterministic models.
\eic{TO THE CREDIT OF MODEL 2, ITS PREDICTIONS OF POLICY IMPLICATIONS WERE AS GOOD AS, OR BETTER THAN, THE OTHER MODELS. PERHAPS THE WEAKNESSES OF DETERMINISTIC MODELS ONLY SHOW UP AS THE DATA ANALYSIS BECOMES MORE REFINED - FOR EXAMPLE, DETERMINISTIC MODELS MAY HAVE SUCH LOW STATISTICAL FIT THAT WE CONCLUDE THEY ARE NOT RELIABLE AS ASSOCIATIVE MODELS, IN WHICH CASE THEY CANNOT BE EXPECTED TO PROVIDE QUANTITATIVELY ACCURATE PREDICTIONS.}

% \arc{Perhaps here (or maybe in the discussion on likelihood-based techniques...or even somewhere else?) we can discuss the utility of point estimates with respect to Model 1. The fitting for Model 1 was done by visual convergence of the parameters, but the plotted simulations were taken from ~100 different sets of parameters values (unsure of exact number because it was not reported). One could argue that this visual convergence is misleading because there is not one set of values — is unclear to me what exactly is actually "converging". Of course, parameter uncertainty is an important topic, and we do discuss it above. But, with respect to interpretability and prediction-informed policy making, a single parameter value has clear benefits. (It may also be worth mentioning that our simulations were done with a single model using one set of parameter values)}
\jwc{Anna: I added some discussion on this in the supplement under Model~1 replication. Does this answer your concern? (see original comment in Rnw file)}

\section{Results}\label{sec:results}

\eic{HERE'S AN INITIAL ATTEMPT TO EXPLAIN MY POINT BELOW FOR THE MS. IT IS HARD TO MAKE CONCRETE POINTS ON THIS SORT OF TOPIC.
A model which aspires to provide quantitative guidance for assessing interventions should provide a quantitative statistical fit for previous data.
However, strong statistical fit does not guarantee a correct causal structure: it does not even necessarily require the model to assert a causal explanation.
A causal interpretation is strengthened by corroborative evidence.
For example, reconstructed latent variables (such as numbers of susceptible and recovered individuals) should make sense in the context of alternative measurments of these variables;
parameter values which fit the data should make sense in the context of alternative lines of evidence about the phenomena being modeled.
The model is necessarily a simplified approximation of a complex process, and so we cannot expect our models (describing and fitted to population-level processes) to perfectly match quantitative understanding of individual-level processes.
Such discrepancies could lead to biases when interventions (modeled as effects on individuals) are included in the population-level, unless there are historical occurrences of the intervation which enable the consequence to be calibrated at the population level.
IF WE WANT TO HEAD IN THIS DIRECTION, WE COULD EXPLAIN HOW THESE PRINCIPLES PLAY OUT IN OUR DATA ANALYSIS.}

\eic{TO AVOID BEING TOO NEGATIVE, WE ALSO COULD EXPLAIN THE LOGIC OF HOW POSITIVE RESULTS CAN LEGITIMATELY BE CONCLUDED.
If a mechanistic model including a feature (such as a representation of a mechanism, or the inclusion of a covariate) fits better than mechanistic models without that feature, and also has competititive fit compared to associative models, this may be taken as evidence supporting the scientific relevance of the feature.
As for any analysis of observational data, we must be alert to the possibility of confounding.
For a covariate, this shows up in a similar way to regression analysis: the covariate under investigation could be a proxy for some other unmodeled or unmeasured covariate.
For a mechanism, the model feature could in principle explain the data by helping to account for some different unmodeled phenomenon.
Assessing potential confounding is part of building an argument for a causal interpretation of a fitted model.
In the context of our analysis, the estimated trend in transmission rate could be explained by any trending variable (such as hygiene improvements, or changes in population behavior), resulting in confounding from colinear covariates.
Alternatively, the trend could be attributed to a decreasing reporting rate rather than decreasing transmission rate, resulting in confounded mechanisms.
The robust statistical conclusion is that a model which allows for change fits better than one which does not---we argue that a decreasing transmission rate is a plausible way to explain this, but the incidence data themselves may not have enough information to pin down the mechanism.
}

Because mechanistic models provide a scientific description of the dynamic system, a properly fit model allows a researcher to perform inference on the system in question. \eic{FIT IS A MEASURE OF ASSOCIATION, AND ASSOCIATION DOES NOT NECESSARILY IMPLY CAUSATION. SO WE HAVE TO WORD THIS CAREFULLY. IT CAN BE A RATHER DELICATE QUESTION TO ESTABLISH THE PROPER CAUSAL INTERPRETATION OF A FITTED MODEL, AND THIS MAY NOT MATCH ONE'S INITIAL GUESS (E.G., THE MEASLES CASE STUDY IN SISMID). IT WOULD BE INTERESTING TO DISCUSS IN PERSON HOW TO THREAD THIS NEEDLE.}
Examples are as diverse as estimating the effective reproductive number ($R_0$) of an infectious disease \citep{he10} or investigating interactions between pedestrians and autonomous vehicles \citep{domeyer22}.
Here, we demonstrate this ability by investigating the seasonal transmission trends in Model~1.
Model~1 includes a flexible cubic spline term in Eq.~\myeqref{model1:lambda}--\myeqref{model1:beta} that allows for seasonality in the force of infection.
This flexibility allows the model to learn any seasonal trends in cholera infections that may be present in the data.
After fitting the model, we explore potential patterns in the seasonal transmission rate by plotting the average value of $\transmission$ in a typical year.
Fig.~\ref{fig:h1SeasRain} shows that the estimated seasonal transmission rate $\transmission$ mimics the rainfall dynamics in Haiti, despite Model~1 not having access to rainfall data.
This result provides evidence that rainfall is an important driver of cholera infections in Haiti.

<<Get Model 1 Seasonality>>=
plot_model_fun <- function(mod) {
  params <- coef(mod)

  beta1 <- params['beta1']
  beta2 <- params['beta2']
  beta3 <- params['beta3']
  beta4 <- params['beta4']
  beta5 <- params['beta5']
  beta6 <- params['beta6']
  betat <- params['betat']

  get_beta <- function(time) {

    covar <- t(mod@covar@table)
    covar_df <- as.data.frame(covar)
    covar_df$times <- mod@covar@times

    betas <- matrix(c(beta1, beta2, beta3, beta4, beta5, beta6), nrow = 1)
    si <- matrix(unlist(covar_df[time, -7]), ncol = 1)
    if (time <= 430) {
      return(as.numeric(exp(betas %*% si + betat * ((time - 215) / (430-215)))))
    } else {
      return(as.numeric(exp(betas %*% si + betat)))
    }
  }

  return(get_beta)
}

get_beta1 <- plot_model_fun(h1)

h1_seas_df <- data.frame(
  'week' = 431:800,
  'date' = lubridate::ymd("2010-10-16") + lubridate::weeks(431:800),
  'trans' = purrr::map_dbl(431:800, get_beta1)
) %>%
  mutate(year = lubridate::year(date)) %>%
  filter(year == 2020) %>%
  mutate(trans_std = (trans - min(trans)) / (max(trans) - min(trans)))

std_rain <- function(x) {
  # This function simply standardizes the rain for us.
  x / max(x)
}

df_rain <- haitiRainfall %>%
  dplyr::summarize(
    date = date, dplyr::across(Artibonite:`Sud-Est`, std_rain)
  ) %>%
  pivot_longer(
    data = .,
    cols = -date,
    names_to = 'dep',
    values_to = 'rainfall'
  ) %>%
  mutate(
    year = lubridate::year(date),
    week = lubridate::week(date)
  ) %>%
  group_by(year, week) %>%
  summarize(national_weekly_rain = sum(rainfall)) %>%
  ungroup() %>% # Not needed, but good practice
  mutate(
    date_week_start = lubridate::ymd(paste0(2014, "-01-01")) + lubridate::weeks(week - 1),
    week_date = as.Date(date_week_start, format = "%m-%d"),
    std_national_rain = (national_weekly_rain - min(national_weekly_rain, na.rm = TRUE)) / (max(national_weekly_rain, na.rm = TRUE) - min(national_weekly_rain, na.rm = TRUE))
  )

mean_rain <- df_rain %>%
  group_by(week_date) %>%
  summarize(mean_rain = mean(std_national_rain)) %>%
  ungroup() %>%
  mutate(mean_rain = (mean_rain - min(mean_rain)) / ((max(mean_rain) - min(mean_rain))))

gg_rain <- df_rain %>%
  filter(week != 1, week != max(week)) %>%
  ggplot(aes(x = week_date, y = std_national_rain, group = year, col = year)) +
  geom_line() +
  theme_bw() +
  guides(color = 'none') +
  ylab('Standardized Weekly\nNational Rain') +
  theme(axis.title.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.text.x = element_blank(),
        axis.text.y = element_text(size = 7.5),
        axis.title.y = element_text(size = 9))

gg_trans <- ggplot() +
  geom_line(data = h1_seas_df, aes(x = date, y = trans_std), linetype = 'dashed') +
  geom_line(
    data = mean_rain,
    aes(x = week_date + lubridate::years(6), y = mean_rain),
    col = '#2166ac'
  ) +
  scale_x_date(date_labels = "%b", date_breaks = '1 month') +
  ylab('Standardized Seasonal\nContact Rate') +
  theme(axis.title.x = element_blank(),
        axis.text = element_text(size = 7.5),
        axis.title.y = element_text(size = 9))

@

\begin{figure}[ht]
\centering
<<Model1_Seasonality_Figure, fig.height=2.9, fig.width=5, fig.align='center'>>=
cowplot::plot_grid(gg_rain, gg_trans, align = "v", ncol = 1)
@
\caption{\label{fig:h1SeasRain}
(Top) weekly rainfall in Haiti, lighter colors representing more recent years.
(Bottom) estimated seasonality in the transmission rate (dashed line) plotted alongside mean rainfall (solid line).
% Observed spikes in rainfall correspond to peaks in the seasonal transmission $\transmission$ in Model~1; similarly, lower values of $\transmission$ correspond to the dry periods.
}
\end{figure}

An additional benefit of mechanistic modelling is that they permit the simulation of potential interventions on a system, as was the primary goal of \citet{lee20}.
Outcomes of their study include estimates for the probability of cholera elimination and cumulative number of cholera infections under several possible vaccination scenarios.
Mimicking their efforts, we define cholera elimination as having less than one infection of cholera over at least 52 consecutive weeks, and provide projections under the following vaccination scenarios:
\arc{Rewrite definition: cholera elimination defined as less than one cholera infection for at least fifty two consecutive weeks in the ten years after the vaccinations?}
\jwc{ Good catch, TODO: is it necessary to say that it stayed away? Does it make a difference?}

\begin{itemize}
  \item[$V0$:] The default vaccination scenario, which is that no additional vaccinations are deployed.
  \item[$V1$:] Vaccination limited to the departements of Centre and Artibonite, and vaccinations deployed over a two-year period.
  \item[$V2$:] Vaccination limited to three departements: Artibonite, Centre, and Ouest over a two-year period.
  \item[$V3$:] Countrywide vaccination implemented over a five-year period.
  \item[$V4$:] Countrywide vaccination implemented over a two-year period.
\end{itemize}

<<Load Model 1 VaccScen Sims, message=FALSE, include=FALSE>>=
load('model1/vacscens/sims_S0.rda')
mod1_V0_sims <- agg_mod1_sims(sims)
mod1_V0_probs <- get_elimProbs(sims, model = 1)
rm(sims)
gc()

load('model1/vacscens/sims_S2.rda')
mod1_V1_sims <- agg_mod1_sims(sims)
mod1_V1_probs <- get_elimProbs(sims, model = 1)
rm(sims)
gc()

load('model1/vacscens/sims_S4.rda')
mod1_V2_sims <- agg_mod1_sims(sims)
mod1_V2_probs <- get_elimProbs(sims, model = 1)
rm(sims)
gc()

load('model1/vacscens/sims_S3.rda')
mod1_V3_sims <- agg_mod1_sims(sims)
mod1_V3_probs <- get_elimProbs(sims, model = 1)
rm(sims)
gc()

load('model1/vacscens/sims_S1.rda')
mod1_V4_sims <- agg_mod1_sims(sims)
mod1_V4_probs <- get_elimProbs(sims, model = 1)
rm(sims)
gc()
@

<<Load Model 2 VaccScenarios, message=FALSE, include=FALSE>>=
# load('model2/output/mod2_vacc_scenarios.rda')

stew(
  file = "model2/output/VaccinationScenarios.rda", {
    mod2_VaccScenarios <- haiti2_vaccScenario(
      end_params = h2_fit$end_params,
      epi_params = h2_fit$epi_params
    )
  }
)
@

<<Load Model 3 Vacc Scenarios, message=FALSE, include=FALSE>>=
load('model3/output/sims_spatPomp_S0.rda')
mod3_V0_sims <- agg_mod3_sims(sims)
mod3_V0_probs <- get_elimProbs(sims, model = 3)
rm(sims)
gc()

load('model3/output/sims_spatPomp_S2.rda')
mod3_V1_sims <- agg_mod3_sims(sims)
mod3_V1_probs <- get_elimProbs(sims, model = 3)
rm(sims)
gc()

load('model3/output/sims_spatPomp_S4.rda')
mod3_V2_sims <- agg_mod3_sims(sims)
mod3_V2_probs <- get_elimProbs(sims, model = 3)
rm(sims)
gc()

load('model3/output/sims_spatPomp_S3.rda')
mod3_V3_sims <- agg_mod3_sims(sims)
mod3_V3_probs <- get_elimProbs(sims, model = 3)
rm(sims)
gc()

load('model3/output/sims_spatPomp_S1.rda')
mod3_V4_sims <- agg_mod3_sims(sims)
mod3_V4_probs <- get_elimProbs(sims, model = 3)
rm(sims)
gc()
@

Simulations from probabilistic models (Models~1 and~3) represent possible trajectories of the dynamic system under the scientific assumptions of the models.
In this case study, estimates of the probability of cholera elimination can therefore be obtained as the proportion of simulations from the fitted model
% starting from draws from the filtering distribution, unecessary - Jesse
that result in cholera elimination.
The results of these projections are summarized in Figs.~\ref{fig:Mod1Scenarios}--\ref{fig:elimProbs}.
These results suggest that cholera elimination was likely, even without intense vaccination efforts.
This conclusion is consistent with the fact that cholera was eradicated from Haiti without additional vaccination efforts \citep{ferguson22}.

In additional to probability of elimination estimates, we are also able to estimate the cumulative number of infections under each vaccination scenario from February 2019 -- February 2024.
Notably, the median number of cumulative cholera infections under the no-vaccination scenario using Models 1 and 3 were $\Sexpr{prettyNum(round(as.numeric(mod1_V0_probs$cumInf['q50'])), big.mark = ",")}$ and $\Sexpr{prettyNum(round(as.numeric(mod3_V0_probs$cumInf['q50'])), big.mark = ",")}$, respectively.
So far, these estimates are far more consistent with the observed number of reported cholera cases than the corresponding estimates from \cite{lee20}, which were approximately $400,000$ and $1,000,000$.

<<Create Model 1 VaccScen plots>>=
gg_m1_V0 <- ggplot() +
  geom_line(data = mod1_V0_sims, aes(x = date, y = q50 + 1), col = 'blue') +
  geom_ribbon(data = mod1_V0_sims, aes(x = date, ymin = q05 + 1, ymax = q95 + 1), alpha = 0.5) +
  geom_line(data = true_agg_cases, aes(x = date, y = ReportedAll + 1)) +
  ylab("Log reported cases") +
  theme(axis.title.x = element_blank()) +
  ggtitle("V0") +
  scale_x_date(date_labels = "'%y") +
  scale_y_log10(labels = scales::trans_format("log10", scales::math_format(10^.x)))

gg_m1_V1 <- ggplot() +
  geom_line(data = mod1_V1_sims, aes(x = date, y = q50 + 1), col = 'blue') +
  geom_ribbon(data = mod1_V1_sims, aes(x = date, ymin = q05 + 1, ymax = q95 + 1), alpha = 0.5) +
  geom_line(data = true_agg_cases, aes(x = date, y = ReportedAll + 1)) +
  ylab("Log reported cases") +
  theme(axis.title = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.y = element_blank()) +
  ggtitle("V1") +
  scale_x_date(date_labels = "'%y") +
  scale_y_log10(labels = scales::trans_format("log10", scales::math_format(10^.x)))

gg_m1_V2 <- ggplot() +
  geom_line(data = mod1_V2_sims, aes(x = date, y = q50 + 1), col = 'blue') +
  geom_ribbon(data = mod1_V2_sims, aes(x = date, ymin = q05 + 1, ymax = q95 + 1), alpha = 0.5) +
  geom_line(data = true_agg_cases, aes(x = date, y = ReportedAll + 1)) +
  ylab("Log reported cases") +
    theme(axis.title = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.y = element_blank()) +
  ggtitle("V2") +
  scale_x_date(date_labels = "'%y") +
  scale_y_log10(labels = scales::trans_format("log10", scales::math_format(10^.x)))

gg_m1_V3 <- ggplot() +
  geom_line(data = mod1_V3_sims, aes(x = date, y = q50 + 1), col = 'blue') +
  geom_ribbon(data = mod1_V3_sims, aes(x = date, ymin = q05 + 1, ymax = q95 + 1), alpha = 0.5) +
  geom_line(data = true_agg_cases, aes(x = date, y = ReportedAll + 1)) +
  ylab("Log reported cases") +
    theme(axis.title = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.y = element_blank()) +
  ggtitle("V3") +
  scale_x_date(date_labels = "'%y") +
  scale_y_log10(labels = scales::trans_format("log10", scales::math_format(10^.x)))

gg_m1_V4 <- ggplot() +
  geom_line(data = mod1_V4_sims, aes(x = date, y = q50 + 1), col = 'blue') +
  geom_ribbon(data = mod1_V4_sims, aes(x = date, ymin = q05 + 1, ymax = q95 + 1), alpha = 0.5) +
  geom_line(data = true_agg_cases, aes(x = date, y = ReportedAll + 1)) +
  ylab("Log reported cases") +
    theme(axis.title = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.y = element_blank()) +
  ggtitle("V4") +
  scale_x_date(date_labels = "'%y") +
  scale_y_log10(labels = scales::trans_format("log10", scales::math_format(10^.x)))

@

\begin{figure}[ht]
<<Plot_Model1_Scenarios, fig.height=2>>=
lay <- rbind(c(1, 1, 1, 1, 2, 2, 2, 3, 3, 3, 4, 4, 4, 5, 5, 5))
gridExtra::grid.arrange(
  gg_m1_V0, gg_m1_V1, gg_m1_V2, gg_m1_V3, gg_m1_V4,
  layout_matrix = lay
)
@
\caption{\label{fig:Mod1Scenarios}
Simulations of Model~1 under each vaccination scenario.
Blue line indicates the median of model simulations, and ribbon represents $95\%$ of simulations.
The various vaccination campaigns made no practical difference in the median scenario, but a drastic difference in the extreme cases.
}
\end{figure}

\begin{figure}[ht]
<<Mod2Fit_and_Scenarios_Figure, fig.height=2, fig.width=5, fig.align='center'>>=
ggplot() +
  geom_line(data = h2_traj, aes(x = date, y = Ctotal + 1), col = 'blue') +
  geom_line(data = true_agg_cases, aes(x = date, y = ReportedAll + 1)) +
  geom_line(data = filter(mod2_VaccScenarios$all_trajs,
                          year >= max(h2_traj$year)),
            aes(x = as.Date(date), y = ReportedCases + 1, color = scenario)) +
  theme(axis.title.x = element_blank(),
        legend.title = element_blank(),
        axis.title.y = element_text(size = 9),
        axis.text = element_text(size = 8)) +
  ylab("Reported cholera cases") +
  scale_color_manual(values = c("V0" = '#f4a582',
                                "V1" = '#d6604d',
                                "V2" = '#b2182b',
                                "V3" = '#92c5de',
                                "V4" = '#4393c3')) +
  geom_vline(xintercept = yearsToDate(2014.161), linetype = 'dashed') +
  scale_y_log10(labels = scales::trans_format("log10", scales::math_format(10^.x))) +
  scale_x_date(date_labels = "'%y", breaks = seq.Date(from = as.Date("2010-01-01"), as.Date("2030-01-01"), by = '2 years'))
@
\caption{\label{fig:mod2Traj}
Simulated trajectory of Model~2 (blue curve) and projections under the various vaccination scenarios.
Reported cholera incidence is shown in black.
The dashed vertical line shows where the endemic phase starts and parameter values are re-estimated.
Note that the various vaccination scenarios had little effect on the projected cholera cases.
This is because the primary effect of vaccines in Model~2 is to remove individuals from the susceptible pool,
yet reported number of cholera infections were low and decreasing at the start of the vaccination campaigns (only $\Sexpr{true_agg_cases$ReportedAll[nrow(true_agg_cases)]}$ cases were reported in all of Haiti),
so this had very little effect on cholera transmission.
}
\end{figure}

<<Create Model 3 VaccScen Plots>>=
gg_m3_V0 <- ggplot() +
  geom_line(data = mod3_V0_sims, aes(x = date, y = q50 + 1), col = 'blue') +
  geom_ribbon(data = mod3_V0_sims, aes(x = date, ymin = q05 + 1, ymax = q95 + 1), alpha = 0.5) +
  geom_line(data = true_agg_cases, aes(x = date, y = ReportedAll + 1)) +
  ylab("Log reported cases") +
  theme(axis.title.x = element_blank()) +
  ggtitle("V0") +
  scale_x_date(date_labels = "'%y") +
  scale_y_log10(labels = scales::trans_format("log10", scales::math_format(10^.x)))

gg_m3_V1 <- ggplot() +
  geom_line(data = mod3_V1_sims, aes(x = date, y = q50 + 1), col = 'blue') +
  geom_ribbon(data = mod3_V1_sims, aes(x = date, ymin = q05 + 1, ymax = q95 + 1), alpha = 0.5) +
  geom_line(data = true_agg_cases, aes(x = date, y = ReportedAll + 1)) +
  ylab("Log reported cases") +
  theme(axis.title = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.y = element_blank()) +
  ggtitle("V1") +
  scale_x_date(date_labels = "'%y") +
  scale_y_log10(labels = scales::trans_format("log10", scales::math_format(10^.x)))

gg_m3_V2 <- ggplot() +
  geom_line(data = mod3_V2_sims, aes(x = date, y = q50 + 1), col = 'blue') +
  geom_ribbon(data = mod3_V2_sims, aes(x = date, ymin = q05 + 1, ymax = q95 + 1), alpha = 0.5) +
  geom_line(data = true_agg_cases, aes(x = date, y = ReportedAll + 1)) +
  ylab("Log reported cases") +
    theme(axis.title = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.y = element_blank()) +
  ggtitle("V2") +
  scale_x_date(date_labels = "'%y") +
  scale_y_log10(labels = scales::trans_format("log10", scales::math_format(10^.x)))

gg_m3_V3 <- ggplot() +
  geom_line(data = mod3_V3_sims, aes(x = date, y = q50 + 1), col = 'blue') +
  geom_ribbon(data = mod3_V3_sims, aes(x = date, ymin = q05 + 1, ymax = q95 + 1), alpha = 0.5) +
  geom_line(data = true_agg_cases, aes(x = date, y = ReportedAll + 1)) +
  ylab("Log reported cases") +
    theme(axis.title = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.y = element_blank()) +
  ggtitle("V3") +
  scale_x_date(date_labels = "'%y") +
  scale_y_log10(labels = scales::trans_format("log10", scales::math_format(10^.x)))

gg_m3_V4 <- ggplot() +
  geom_line(data = mod3_V4_sims, aes(x = date, y = q50 + 1), col = 'blue') +
  geom_ribbon(data = mod3_V4_sims, aes(x = date, ymin = q05 + 1, ymax = q95 + 1), alpha = 0.5) +
  geom_line(data = true_agg_cases, aes(x = date, y = ReportedAll + 1)) +
  ylab("Log reported cases") +
    theme(axis.title = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.y = element_blank()) +
  ggtitle("V4") +
  scale_x_date(date_labels = "'%y") +
  scale_y_log10(labels = scales::trans_format("log10", scales::math_format(10^.x)))
@

\begin{figure}[ht]
<<Plot_Model3_Scenarios, fig.height=2>>=
lay <- rbind(c(1, 1, 1, 1, 2, 2, 2, 3, 3, 3, 4, 4, 4, 5, 5, 5))
gridExtra::grid.arrange(
  gg_m3_V0, gg_m3_V1, gg_m3_V2, gg_m3_V3, gg_m3_V4,
  layout_matrix = lay
)
@
\caption{\label{fig:Mod3Scenarios}
Simulations of Model~3 under each vaccination scenario.
Blue line indicates the median of model simulations, and ribbon represents $95\%$ of simulations.
The various vaccination campaigns made no practical difference in the median scenario, but a drastic difference in the extreme cases.
}
\end{figure}

Probability of elimination estimates of this form are far less meaningful for deterministic models.
This is because the data are not assumed to be draws from the model; the model is instead intended only to provide understanding about the overall trend of the epidemic.
We therefore do not provide probability of elimination estimates under Model~2.
Still, projected trajectories using Model~2 are consistent with those found with Models~1 and ~3, and suggest that cholera was in the process of being eliminated from Haiti.

% \begin{figure}[ht]
% \centering
%
% mod1_cumInf <- data.frame(
%   'scenario' = paste0("V", 0:4),
%   'low' = c(
%     as.numeric(mod1_V0_probs$cumInf['q025']),
%     as.numeric(mod1_V1_probs$cumInf['q025']),
%     as.numeric(mod1_V2_probs$cumInf['q025']),
%     as.numeric(mod1_V3_probs$cumInf['q025']),
%     as.numeric(mod1_V4_probs$cumInf['q025'])
%   ),
%   'high' = c(
%     as.numeric(mod1_V0_probs$cumInf['q975']),
%     as.numeric(mod1_V1_probs$cumInf['q975']),
%     as.numeric(mod1_V2_probs$cumInf['q975']),
%     as.numeric(mod1_V3_probs$cumInf['q975']),
%     as.numeric(mod1_V4_probs$cumInf['q975'])
%   ),
%   'med' = c(
%     as.numeric(mod1_V0_probs$cumInf['q50']),
%     as.numeric(mod1_V1_probs$cumInf['q50']),
%     as.numeric(mod1_V2_probs$cumInf['q50']),
%     as.numeric(mod1_V3_probs$cumInf['q50']),
%     as.numeric(mod1_V4_probs$cumInf['q50'])
%   )
% )
%
% mod3_cumInf <- data.frame(
%   'scenario' = paste0("V", 0:4),
%   'low' = c(
%     as.numeric(mod3_V0_probs$cumInf['q025']),
%     as.numeric(mod3_V1_probs$cumInf['q025']),
%     as.numeric(mod3_V2_probs$cumInf['q025']),
%     as.numeric(mod3_V3_probs$cumInf['q025']),
%     as.numeric(mod3_V4_probs$cumInf['q025'])
%   ),
%   'high' = c(
%     as.numeric(mod3_V0_probs$cumInf['q975']),
%     as.numeric(mod3_V1_probs$cumInf['q975']),
%     as.numeric(mod3_V2_probs$cumInf['q975']),
%     as.numeric(mod3_V3_probs$cumInf['q975']),
%     as.numeric(mod3_V4_probs$cumInf['q975'])
%   ),
%   'med' = c(
%     as.numeric(mod3_V0_probs$cumInf['q50']),
%     as.numeric(mod3_V1_probs$cumInf['q50']),
%     as.numeric(mod3_V2_probs$cumInf['q50']),
%     as.numeric(mod3_V3_probs$cumInf['q50']),
%     as.numeric(mod3_V4_probs$cumInf['q50'])
%   )
% )
%
% mod1_cumInf$model <- "1"
% mod3_cumInf$model <- "3"
%
% all_cumInf <- rbind(mod1_cumInf, mod3_cumInf)
% all_cumInf$scenario <- factor(all_cumInf$scenario, levels = rev(c("V0", "V1", "V2", "V3", "V4")))
%
%
% ggplot(all_cumInf, aes(y = scenario)) +
%   geom_point(aes(x = med)) +
%   geom_linerange(aes(xmin = low, xmax = high)) +
%   scale_x_continuous(
%     labels = scales::unit_format(scale = 1e-6, accuracy = 1),
%     limits = c(0, 5e6)
%   ) +
%   xlab("Cumulative infections (million)") +
%   theme(axis.title.y = element_blank(),
%         axis.title.x = element_text(size = 10),
%         axis.text = element_text(size = 8)) +
%   facet_wrap(~model, ncol = 1,
%              labeller = as_labeller(
%                c("1" = "Model 1", "3" = "Model 3", "2" = "Model 2")
%              ))
%
% \caption{\label{fig:cumInfPlot}
% Simulated cumulative infections under various vaccination scenarios from February, 2019, to February, 2024.
% Point estimates (median cumulative incidence) and $95\%$ error bars are included.
% x-axis scale is intended to ease comparison with Figure~4 of \cite{lee20}.
% \jwc{MAYBE JUST SUMMARIZE THIS FIGURE IN A PARAGRAPH OR MOVE TO SUPPLEMENT TO MAKE SPACE?}
% }
% \end{figure}

Despite these results suggesting that cholera was likely to be eliminated from Haiti without an increase in vaccination effort, we in no way suggest that vaccinations shouldn't be used to combat cholera---or other infectious diseases---in other parts of the world.
In fact, our analysis shows that an increase in vaccination does drastically reduce the negative effects of a worst-case scenario.
% This alone should provide policy makers with enough motivation to continue vaccination efforts around the country.
% Avoid comments on vaccination policy recommendations - Jesse

<<Compute Elimination Probs, echo=FALSE>>=
mod1_V0_probs$ElimTime$scenario = "V0"
mod1_V1_probs$ElimTime$scenario = "V1"
mod1_V2_probs$ElimTime$scenario = "V2"
mod1_V3_probs$ElimTime$scenario = "V3"
mod1_V4_probs$ElimTime$scenario = "V4"

mod1_probElims <- rbind(
  mod1_V0_probs$ElimTime,
  mod1_V1_probs$ElimTime,
  mod1_V2_probs$ElimTime,
  mod1_V3_probs$ElimTime,
  mod1_V4_probs$ElimTime
)

mod3_V0_probs$ElimTime$scenario = "V0"
mod3_V1_probs$ElimTime$scenario = "V1"
mod3_V2_probs$ElimTime$scenario = "V2"
mod3_V3_probs$ElimTime$scenario = "V3"
mod3_V4_probs$ElimTime$scenario = "V4"

mod3_probElims <- rbind(
  mod3_V0_probs$ElimTime,
  mod3_V1_probs$ElimTime,
  mod3_V2_probs$ElimTime,
  mod3_V3_probs$ElimTime,
  mod3_V4_probs$ElimTime
)

mod1_probElims$mod <- "Model 1"
mod3_probElims$mod <- "Model 3"
mod3_probElims$time <- yearsToDate(mod3_probElims$time)

all_prob_elims <- dplyr::bind_rows(mod1_probElims, mod3_probElims)
@

\begin{figure}[ht]
<<Elimination_Probs_Figure, fig.height=2.5>>=
ggplot(tidyr::drop_na(all_prob_elims), aes(x = time, y = elim_prob, col = mod)) +
  geom_line() +
  facet_wrap(~scenario, nrow = 1) +
  scale_color_manual(values = c("Model 1" = "#377eb8", "Model 3" = "#e41a1c")) +
  ylab("Probability of Elimination (%)") +
  theme(axis.title.x = element_blank(),
        legend.position = 'bottom',
        legend.title = element_blank(),
        axis.text.x = element_text(angle = 35, hjust = 1),
        legend.margin=margin(c(-5, 0, -3, 0)),
        legend.box.margin = margin(c(-5, 0, -3, 0))) +
  scale_x_date(date_labels = "'%y", breaks = seq.Date(from = as.Date("2010-01-01"), as.Date("2030-01-01"), by = '2 years'))
@
\caption{\label{fig:elimProbs}
Probability of elimination across simulations for a 10 year period.
Compare to Figure 3A of \cite{lee20}.}
\end{figure}

%%% table tttttttttt

\begin{table}
\small

\hspace{-10mm}\begin{tabular}{|l|l@{\hskip3pt}c@{\hskip3pt}|l@{\hskip3pt}c@{\hskip3pt}|l@{\hskip3pt}c@{\hskip3pt}|}
\hline
Mechanism & Model 1 && Model 2 && Model 3 &
\\
\hline
\hline
Infection (day)
  & {\fixed $\muIR^{-1}=\Sexpr{myround(7/p1["gamma"],1)}$}
  & \eqref{model1:toR}
  & {\fixed $\muIR^{-1}=\Sexpr{myround(365/h2_fit$epi_params["gamma"],1)}$}
  & \eqref{model2:mu_IR}
  & {\fixed $\muIR^{-1}=\Sexpr{myround(365/p3["gamma"],1)}$ }
  & \eqref{eq:model3:IR}
\\
Latency (day)
  & {\fixed $\muEI^{-1}=\Sexpr{myround(7/p1["sigma"],1)}$}
  & \eqref{model1:EA}
  & {\fixed $\muEI^{-1}=\Sexpr{myround(365/h2_fit$epi_params["gammaE"],1)}$}
  & \eqref{model2:mu_EI}
  & \missing
  &
\\
\hline
Seasonality
  & $\begin{array}{l}
    \hspace{-1mm} \transmission_{1:6}=(
      \Sexpr{myround(p1["beta1"],1)},
      \Sexpr{myround(p1["beta2"],1)},
    \\
      \Sexpr{myround(p1["beta3"],1)},
      \Sexpr{myround(p1["beta4"],1)},
      \Sexpr{myround(p1["beta5"],1)},
      \Sexpr{myround(p1["beta6"],1)})
    \end{array}$
  & \eqref{model1:beta}
  & {\fixed $\seasAmplitude=0.4$ }
  & \eqref{model2:lambda}
  & $\begin{array}{l} \seasAmplitude=\Sexpr{myround(p3s["lambdaR"],2)} \\ r=\Sexpr{myround(p3s["r"],2)} \end{array}$
  & \eqref{eq:model3:water}
\\
\hline
Immunity (year)
  & {\fixed $\muRS^{-1}=\Sexpr{myround(1/p1["alpha"]/52,1)}$}
  & \eqref{model1:RS}
  & {\fixed
    $\begin{array}{l}
    \muRS^{-1}= \Sexpr{myround(1/h2_fit$epi_params["sigma"],1)} \\
     \omega_1^{-1} = \Sexpr{myround(1/h2_fit$epi_params["Omega1"],1)} \\
    \omega_2^{-1} = \Sexpr{myround(1/h2_fit$epi_params["Omega2"],1)}
    \end{array}$
    }
  & $\begin{array}{l}
    \eqref{model2:RS} \\
    \eqref{model2:omega1} \\
    \eqref{model2:omega1}
    \end{array} $
  & {\fixed $\muRS^{-1}=\Sexpr{myround(1/p3s["rho"],1)}$}
  & \eqref{eq:model3:RRnext}
\\
\hline
Vaccine efficacy
  & \missing
  &
  & $\begin{array}{l}
      \hspace{-1mm} \fixed{\vaccineEfficacy_{1:4} = (
        \Sexpr{myround(1 - h2_fit$epi_params["VE1"] * 0.4688, 2)},
        \Sexpr{myround(1 - h2_fit$epi_params["VE2"] * 0.4688, 2)}} \\
        \fixed{\Sexpr{myround(1 - h2_fit$epi_params["VE1"], 2)},
        \Sexpr{myround(1 - h2_fit$epi_params["VE2"], 2)}})
        \end{array}$
  & \eqref{model2:mu_SE}
  & \fixed{$\eta_{ud}(t)$}
  &
  \\
Birth/death (year)
  & \fixed{$\begin{array}{l}
    \muBirth^{-1} = \Sexpr{myround(1/p1["mu"]/52,1)} \\
    \muDeath^{-1} = \Sexpr{myround(1/p1["delta"]/52,1)}
    \end{array}$}
  & \eqref{model1:death}
  & \missing
  &
  & {\fixed $\muDeath^{-1} = \Sexpr{myround(1/p3s["mu"],1)}$}
  & \eqref{eq:model3:IS}
\\
Symptomatic frac.
  & {\fixed $\symptomFrac_z(t)=c\theta^*(t-\tau_d)$}
  & (\ref{model1:EI}-\ref{model1:EA})
  & {\fixed $\symptomFrac=\Sexpr{myround(h2_fit$epi_params["k"],1)}$}
  & \eqref{model2:mu_EI}
  & {\fixed $\symptomFrac=\Sexpr{myround(p3s["sigma"],2)}$}
  & \eqref{eq:model3:SA}
\\
Asympt.~infectivity
  & {\fixed $\asymptomRelativeInfect=0.05$ }
  & \eqref{model1:lambda}
  & $\begin{array}{l}
      {\fixed \asymptomRelativeInfect =0.001 } \\
      {\fixed \asymptomRelativeShed = \Sexpr{1e-7} }
    \end{array}$
  & \begin{tabular}{l}
      \eqref{model2:lambda} \\
      \eqref{model2:to_W}
    \end{tabular}
  &  $\begin{array}{l}
      {\fixed \asymptomRelativeInfect =1 } \\
      {\fixed \asymptomRelativeShed = \Sexpr{myround(p3s["XthetaA"],3)} }
    \end{array}$
  & \begin{tabular}{l}
      \eqref{eq:model3:foi} \\
      \eqref{eq:model3:water}
    \end{tabular}
\\
\hline
Human to human
  & $\transmission_{1:6}$ as above
  & \eqref{model1:lambda}
  & $\begin{array}{l}
      \hspace{-1mm} \transmission=(
      \Sexpr{signif(h2_fit$epi_params["Beta"],2)}, \\
      \Sexpr{signif(h2_fit$end_params["Beta"],2)}
      )\end{array}$
  & \eqref{model2:lambda}
  & $\begin{array}{l}
      \hspace{-1mm} \transmission_{1:10}=(
        \Sexpr{myround(p3u["foi_add",1]*1e6,2)},
	\Sexpr{myround(p3u["foi_add",2]*1e6,2)},
      \\
	\Sexpr{myround(p3u["foi_add",3]*1e6,2)},
        \Sexpr{myround(p3u["foi_add",4]*1e6,2)},
	\Sexpr{myround(p3u["foi_add",5]*1e6,2)},
	\Sexpr{myround(p3u["foi_add",6]*1e6,2)},
      \\
	\Sexpr{myround(p3u["foi_add",7]*1e6,2)},
	\Sexpr{myround(p3u["foi_add",8]*1e6,2)},
	\Sexpr{myround(p3u["foi_add",9]*1e6,2)},
	\Sexpr{myround(p3u["foi_add",10]*1e6,2)}
	)
      \\
      \times 10^{-6}
    \end{array}$
  & \eqref{eq:model3:foi}
\\
\hline
Water to human
  & \missing
  &
  & $\begin{array}{l}
    {\fixed \Wsat = \Sexpr{signif(h2_fit$epi_params["Sat"],2)} }
    \\
    \beta_W= (\Sexpr{signif(h2_fit$epi_params["BetaW"],3)}, \Sexpr{signif(h2_fit$end_params["BetaW"],3)})
    \end{array}$
  & \eqref{model2:lambda}
  & $\begin{array}{l}
      \hspace{-1mm} {\fixed \Wsat=1}
        \\
      \hspace{-1mm} \Wbeta{_{1:10}}= (
        \Sexpr{myround(p3u["betaB",1],2)}, \Sexpr{myround(p3u["betaB",2],2)},
       \\
        \Sexpr{myround(p3u["betaB",3],2)}, \Sexpr{myround(p3u["betaB",4],2)},
        \Sexpr{myround(p3u["betaB",5],2)}, \Sexpr{myround(p3u["betaB",6],2)},
       \\
        \Sexpr{myround(p3u["betaB",7],2)}, \Sexpr{myround(p3u["betaB",8],2)},
        \Sexpr{myround(p3u["betaB",9],2)}, \Sexpr{myround(p3u["betaB",10],2)}
      ) \end{array}$
  & \eqref{eq:model3:foi}
\\
\hline
Human to water
  & \missing
  &
  & $\begin{array}{l}
      \hspace{-1mm} \Wshed = (
      \Sexpr{signif(h2_fit$epi_params["Mu"],3)}, \\
      \Sexpr{signif(h2_fit$end_params["Mu"],3)}
      )
      \end{array}$
  & \eqref{model2:to_W}
  & $\Wshed= \Sexpr{signif(p3s["thetaI"],3)}$
  & \eqref{eq:model3:water}
\\
Water survival (wk)
  & \missing
  &
  & $\begin{array}{l}
      \hspace{-1mm} \Wremoval^{-1} = (
      \Sexpr{signif(52/h2_fit$epi_params["Delta"],3)}, \\
      \Sexpr{myround(52/h2_fit$end_params["Delta"],1)}
      )
      \end{array}$
  & \eqref{model2:from_W}
  & $\Wremoval^{-1}=\Sexpr{myround(52/p3s["mu_B"],2)}$
  & \eqref{eq:model3:Decay}
\\
Mixing exponent
  & $\mixExponent=\Sexpr{myround(p1["nu"],2)}$
  & \eqref{model1:lambda}
  & \missing
  &
  & \missing
  &
\\
Proc. noise (wk$^{1/2}$)
  & $\sigmaProc=(\Sexpr{myround(sqrt(p1["sig_sq_epi"]),2)}, \Sexpr{myround(sqrt(p1["sig_sq_end"]),2)})$
  & \eqref{model1:lambda}
  & \missing
  &
  & $\sigmaProc= \Sexpr{myround(p3s["std_W"],3)}$
  & \eqref{eq:model3:SA}
\\
Reporting rate
  & $\reportRate=(\Sexpr{myround(p1["rho_epi"],3)}, \Sexpr{myround(p1["rho_end"],3)})$
  & \eqref{model1:meas}
  & {\fixed $\reportRate=\Sexpr{myround(h2_fit$epi_params["Rho"],2)}$}
  & \eqref{model2:meas}
  & $\reportRate=\Sexpr{myround(p3s["epsilon"],2)}$
  & \eqref{model3:meas}
\\
Obs. overdispersion
  & $\obsOverdispersion=(\Sexpr{round(p1["tau_epi"],2)}, \Sexpr{round(p1["tau_end"],2)})$
  & \eqref{model1:meas}
  &
  &
  & $\obsOverdispersion=\Sexpr{myround(p3s["k"],2)}$
  & \eqref{model3:meas}
\\
\hline
\end{tabular}
\caption{References to the relevant equation are given in parentheses.
Parameters in blue were fixed based on scientific reasoning and not fitted to the data.
[N] denotes parameters added during our re-analysis, not considered by Lee et al.
Translations back into the notation of \citet{lee20} are given in Table~S1.
% \eic{TODO: CHECK PARAMETERIZATION AND UNITS FOR $\obsOverdispersion$ AND $\sigmaProc$ TO MAKE THEM COMPARABLE}
}
\end{table}

\section{Discussion}\label{sec:discussion}

The ongoing global COVID-19 pandemic has provided a clear example on how government policy may be affected by the conclusions of scientific models.
This article demonstrates that fitting appropriate scientific models remains a challenging statistical task, and therefore great care is needed when fitting scientific models for policy recommendations.
We provided a few suggestions that may aid the fitting of mechanistic models such as comparing model likelihoods to a benchmark.
Improved model fits allows for meaningful statistical inference that may provide valuable insight on a dynamic system in question and may improve the accuracy of model-based projections.
% TODO: "Improved model fits", or "Improved model fitting", based on Anna's suggestion.
Caution is nonetheless needed when making policy based on modelling conclusions, as model misspecification may invalidate conclusions.
% TODO: Citation above?

Various  have suggestions been made about why \citet{lee20} failed to accurately predict the eventual eradication of cholera from Haiti.
Proposals include model misspecification, overly difficult elimination criteria, and a potential conflict of interest \citep{rebaudetComment20,henrys20}.
Here, we instead argue that careful attention to important statistical details could have correctly resulted in the conclusion of imminent cholera elimination.
We acknowledge that we have the benefit of hindsight: our demonstration of a statistically principled route to obtain better-fitting models with more accurate predictions does not rule out the possibility of discovering other models that fit well yet predict poorly.

We used the same data and models, and even much of the same code, as \citet{lee20}, and yet ended up with drastically different conclusions.
At a minimum, we have shown that the conclusions are sensitive to details in how the data analysis is carried out, and that attention to statistical fit (including numerical issues such as likelihood maximization) can lead to improved policy guidance.

We  acknowledge there are limitations to this study;
one example was the inability to fit model parameters to the fully-coupled SpatPOMP version of Model~3.
Promising theoretical and methodological developments \citep{ning21ibpf,ionides22} based on the Block Particle Filter \citep{rebeschini15} may potentially be used to fit the SpatPOMP version of Model~3 in future work.

% \arc{Right now, the manuscript's main point/claim is a bit muddled (at least to me). I am not sure how to directly address this, but having a additional paragraph or two in the discussion may help the impact of this article/"drive our points home". I will think more on this in the next week, but maybe we could further emphasize the balance/dependence between policy, prediction, inference, model quality, and model fitting? Well informed policy depends upon reliable prediction which depends upon model quality/performance which depends upon rigorous evaluation and estimation, regardless of correct model specification (though that is also important/of great interest).}
% \jwc{Thanks, I agree that the main point isn't clear, and Ed has suggested as much also. I think that adding additional paragraphs would help this, but we are already 2 pages over the recommended submission length, so re-wording the text may be a better option than adding more to it... I'll also have to think about this more this week.}
% \eic{HERE'S A SUGGESTION FOR A CONCLUDING PARAGRAPH. I THINK IT IS SUPPORTED BY THE RESULTS...}
Inference for mechanistic time series models offers opportunities for understanding and controlling complex dynamic systems. This case study has investigated issues requiring attention when applying powerful new statistical techniques that can enable statistically efficient inference for a general class of partially observed Markov process models. Care must be taken to ensure that the computationally intensive numerical calculations are carried out adequately. Once that is accomplished, care is required to assess what causal conclusions can properly be inferred given the possibility of alternative explanations consistent with the data. Studies that combine model development with thoughtful data analysis, supported by a high standard of reproducibility, build knowledge about the system under investigation. Cautionary warnings about the difficulties inherent in understanding complex systems \citep{saltelli20,ioannidis20} should motivate us to follow best practices in data analysis, rather than avoiding the challenge.

\subsection{Reproducibility and Extendability}

\citet{lee20} published their code and data online, and this reproducibility facilitated our work.
By design, the models were coded and analyzed independently, leading to differing implementation decisions.
Robust data analysis requires not only reproducibility but also extendability: if one wishes to try new model variations, or new approaches to fitting the existing models, or plotting the results in a different way, this should be not excessively burdensome.
Scientific results are only trustworthy so far as they can be critically questioned, and an extendable analysis should facilitate such examination \citep{gentleman07}.

We provide a strong form of reproducibility, as well as extendability, by developing our analysis in the context of an \code{R} package, \code{haitipkg}.
Using a software package mechanism supports documentation, standardization and portability that promote extendability.
In the terminology of \citet{gentleman07}, the source code for this article is a {\it dynamic document} combining code chunks with text.
In addition to reproducing the article, the code can be extended to examine alternative analysis to that presented.
The dynamic document, together with the R packages, form a {\it compendium}, defined by \citet{gentleman07} as a distributable and executable unit which combines data, text and auxiliary software (the latter meaning code written to run in a general-purpose, portable programming environment, which in this case is R).

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Single Appendix:                         %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% \begin{appendix}
% \section*{Block Panel Iterated Filter}
%
% This algorithm is a modification of the panel iterated filter \citep{breto20}.
% The intuition behind this modification is that if the unit-specific
% parameters are independent of one another, we allow these parameters to be
% fit as such by not allowing the particle ancestry in other units to influence
% the fit of unit specific parameter in question.
% Pseudo-code for this algorithm is given in Algorithm~1.
%
% While theoretical properties of this algorithm have yet to be derived, empirical
% results using Model~3 and on simulated data from a simple model suggest that this
% modification may lead to quicker convergence to the model likelihood.
%
% \begin{algorithm}[ht]
%    \caption{\textbf{BPIF}. \\
%     {\bf Inputs}:
%     Simulator of initial density, $f_{X_{u, 0}}(x_{u, 0}; \theta)$ for $u$ in $1:U$.
%     Simulator of transition density, $f_{X_{u, n}|X_{u, n-1}}(x_{u, n}|x_{u, n-1}; \theta)$ for $u$ in $1:U$, $n$ in $1:N_u$.
%     Evaluator of measurement density, $f_{Y_{u, n}|X_{u, n}}(y_{u, n}|x_{u, n} ; \theta)$ for $u$ in $1:U$, $n$ in $1:N_u$.
%     Data $y_{u, n}^*$,for $u$ in $1:U$, $n$ in $1:N_u$.
%     Number of iterations, $M$.
%     Number of particles, $J$.
%     Starting parameter swarm, $\Theta_j^0$ for $j$ in $1:J$.
%     Simulator of perturbation density, $h_{u, n}(\theta | \phi; \sigma)$ for $u$ in $1:U$, $n$ in $0:N_u$
%     Perturbation sequence, $\sigma_m$ for $m$ in $1:M$.
%     {\bf Output:} Final parameter swarm, $\Theta_{j}^{m}$ for $j$ in $1:J$
%     \label{alg:bpif}}
% %\BlankLine
% \For{$m \in 1:M$}{
%   Set $\Theta_{0, j}^m = \Theta_j^{m-1}$ for $j \in 1:J$\;
%     \For{$u \in 1:U$}{
%       Perturb parameters: $\Theta_{u, 0, j}^{F, m} \sim h_{u, 0}(\theta| \Theta_{u - 1, j}^m; \sigma_m)$ for $j \in 1:J$\;
%       Initialize: $X_{u, 0, j}^{F, m} \sim f_{X_{u, 0}}(x_{u, 0}; \Theta_{u, 0, j}^{F, m})$ for $j \in 1:J$\;
%         \For{$n \in 1:N_u$} {
%           Perturb parameters: $\Theta_{u, n, j}^{P, m} \sim h_{u, n}(\theta|\Theta_{u, n-1, j}^{F, m}, \sigma_m)$ for $j \in 1:J$\;
%           Prediction Simulation: $X_{u, n, j}^{P, m} \sim f_{X_{u, n}|X_{u, n-1}}(x_{u, n}|X_{u, n-1, j}^{F, m}; \Theta_{u, n, j}^{P, m})$ for $j \in 1:J$\;
%           Prediction weights: $w_{u, n, j}^m = f_{Y_{u, n}|X_{u, n}}(y_{u, n}^*|X_{u, n, j}^{P, m})$ for $j \in 1:J$\;
%           Draw $k_{1:j}$ with $P(k_j = i) = w_{u, n, i}^m / \sum_{v = 1}^J w_{u, n, v}^m$ for $i, j \in 1:J$\;
%           Set: $\Theta_{u, n, j}^{F, m} = \Theta_{u, n, k_j}^{P, m}$ and $X_{u, n, j}^{F, m} = X_{u, n, k_j}^{P, m}$ for $j \in 1:J$\;
%         }
%       Set $\Theta_{u, j}^m = \Theta_{u, N_u, j}^{F, m}$ for $j \in 1:J$\;
%     }
%   Set $\Theta_j^m = \Theta_{U, j}^m$ for $j \in 1:J$\;
% }
% \end{algorithm}

% \section*{}%% if no title is needed, leave empty \section*{}.
% \begin{table}
%   \begin{tabular}{|c|c|c|c|}\hline
%     \thead{Parameter} & \thead{Our \\ Notation} & \thead{\cite{lee20} \\ Notation} & \thead{Relevant \\ Model(s)} \\
%     \hline
%     \hline
%     % Table begins here
%     \multirow{2}{*}{Reporting Rate} & \multirow{2}{*}{$\reportRate$} & $\rho$ & 1, 2 \\\cline{3-4}
%       & & $\epsilon_1, \epsilon_2$ & 3\\\hline
%     Mixing Coefficient & $\mixExponent$ & $\nu$ & 1 \\\hline
%     \multirow{2}{*}{Measurement Over-Dispersion} & \multirow{2}{*}{$\obsOverdispersion$} & $\tau$ & 1\\\cline{3-4}
%     & & $p$ & 3 \\\hline
%     Birth Rate & $\muBirth$ & $\mu$ & 1 \\\hline
%     \multirow{2}{*}{Natural Mortality Rate} & \multirow{2}{*}{$\muDeath$} & $\delta$ & 1 \\\cline{3-4}
%      & & $\mu$ & 3 \\\hline
%      Cholera Mortality Rate & $\choleraDeath$ & $\alpha$ & 3 \\\hline
%     \multirow{2}{*}{Latent Period} & \multirow{2}{*}{$1/\muEI$} & $1/\sigma$ & 1 \\\cline{3-4}
%                                    & & $1/\gamma_E$ & 2 \\\hline
%     Recovery Rate & $\muIR$ & $\gamma$ & 1, 2, 3 \\\hline
%     \multirow{3}{*}{Loss of Immunity} & \multirow{3}{*}{$\muRS$} & $\alpha$ & 1 \\\cline{3-4}
%                                       & & $\sigma$ & 2 \\\cline{3-4}
%                                       & & $\rho$ & 3 \\\hline
%     \multirow{3}{*}{Symptomatic Ratio} & \multirow{3}{*}{$\symptomFrac$} & $1 - \theta_0$ & 1 \\\cline{3-4}
%     & & $k$ & 2 \\\cline{3-4}
%     & & $\sigma$ & 3\\\hline
%     Asymptomatic & \multirow{2}{*}{$\asymptomRelativeInfect$} & $\kappa$ & 1 \\\cline{3-4}
%     Relative Infectiousness & & $red_\beta$ & 2 \\\hline
%     \multirow{2}{*}{Human-to-Water Shedding} & \multirow{2}{*}{$\Wshed$} & $\mu$ & 2 \\\cline{3-4}
%      & & $\theta_I$ & 3 \\\hline
%     Asymptomatic & \multirow{2}{*}{$\asymptomRelativeShed$} & $red_\mu$ & 2 \\\cline{3-4}
%     Relative Shedding & & $\theta_A/\theta_I$ & 3 \\\hline
%     \multirow{2}{*}{Seasonal Amplitude} & \multirow{2}{*}{$\seasAmplitude$} & $\alpha_s$ & 2 \\\cline{3-4}
%     & & $\lambda$ & 3 \\\hline
%     \multirow{2}{*}{Transmission} & \multirow{2}{*}{$\transmission$} & $\beta$ & 1, 2 \\\cline{3-4}
%     & & $c$ & 3 \\\hline
%     Water-to-Human & \multirow{2}{*}{$\beta_W$} & $\beta_W$ & 2 \\\cline{3-4}
%     Infection Rate & & $\beta$ & 3 \\\hline
%     Bacteria Mortality & \multirow{2}{*}{$\Wremoval$} & $\delta$ & 2 \\\cline{3-4}
%     Rate & & $\mu_\beta$ & 3 \\\hline
%     \multirow{3}{*}{Vaccination Efficacy} & \multirow{3}{*}{$\vaccineEfficacy$} & $\theta_{vk}$ & 1 \\\cline{3-4}
%     & & $\text{VE}_1, \text{VE}_2$ & 2 \\\cline{3-4}
%     & & $\eta_{1d}, \eta_{2d}$ & 3 \\\hline
%     Process Over-dispersion & $\sigmaProc$ & $\sigma_w$ & 3 \\\hline
%     % Table ends here
%   \end{tabular}
%   \caption{
%   \label{tab:translate}Translations between our common notation and notation used in \cite{lee20}\eic{TODO: half saturation constant, 2 and 3. Fraction of susceptible, 2.
%   Initial Value parameters.}
%   }
% \end{table}

% \section*{Initial Values}
%
% To perform inference on POMP models, it is necessary to propose an initial density for the latent process $f_{X_0}(x_0;\theta)$.
% This density is used to obtain initial values of the latent state when fitting and evaluating the model.
% For each of the models considered in this analysis, the initial conditions are derived by enforcing the model dynamics on reported cholera cases.
% For example, in a deterministic model, the total number of Infected and Asymptomatic individuals may be calculated using the observed number of cholera cases at time time of model initialization based on model parameters such as reporting rate $\reportRate$ and fraction of symptomatic cases $\symptomFrac$.
% It is also sometimes necessary to fit some initial value parameters in order to help determine starting values for weakly identifiable compartments.
% In the following subsections, we mention initial value parameters that were fit for each model.
%
% \subsection*{Model~1}
%
% For this model, the number of individuals in the Recovered and Asymptomatic compartments are set to zero, but the initial proportion of Infected and Exposed individuals is estimated as initial value parameters ($I_0$ and $E_0$, respectively) using the MIF2 algorithm.
% Finally, the initial proportion of Susceptible individuals $S_0$ is calculated as $S_0 = 1 - I_0 - E_0$.
%
% \subsection*{Model~3}
%
% We use the reported cases at the start of the pandemic to approximate the number of Asymptomatic, Infectious, and Recovered individuals in each department $u \in 1:U$ using the same approximation as provided in Eq.~\myeqref{eq:model3:approx}.
% The susceptible compartment is initialized so that the sum $S_{u}(0) + I_u(0) + A_u(0) + \sum_k R_{u, k}(0) = \text{population}_u$.
% The bacteria compartment is then initialized using Eq.~\myeqref{eq:model3:b0}:
%
% \begin{equation}
%   B_u(0) = \big[1 + \seasAmplitude \big(\xi_u)^r \big] D_i \, \Wshed \big[ I_{u}(0)+ \asymptomRelativeShed A_{u}(0) \big]\label{eq:model3:b0}
% \end{equation}
%
% Where $\xi_u \in (0, 1)$ are initial value parameters that we introduce in order to allow some flexibility in determining the initial state of the bacteria compartment.

% \section*{Measurement Models}
%
% Each POMP requires specification of a measurement model, which is a statistical description of how observations on the system are obtained.
% In general, we used the same measurement models that were reported in \cite{lee20}.
%
% \subsection*{Model~1}
%
% The only reported cases under this measurement model come from individuals who develop symptoms, as noted in Eq.~\myeqref{model1:meas}.
% \begin{equation}
%   \label{model1:meas}
%   y_{t} \mid \Delta N_{E_{k}I_{k}} = z_{t} \sim \text{NB}\left(\reportRate z_{t}, \obsOverdispersion \right)
% \end{equation}
%
% \subsection*{Model~2}
%
% Model~2 was fit via trajectory matching and therefore does not have an explicit measurement model.
% Implicitly, however, this fitting process is equivalent to having a Gaussian measurement process.
% Therefore the measurement model can be expressed as Eq.~\myeqref{model2:meas}.
% \begin{equation}
%   \label{model2:meas}
%   y_{u, t} \mid \Delta N_{I_{ud}R_{ud}} = z_{u, t} \sim \text{N}\left(\reportRate z_{u, t}, \obsOverdispersion^2 \right)
% \end{equation}
%
% Where $\Delta N_{I_{ud}R_{ud}}$ is the number of individuals who move from compartment $I_{ud}$ to $R_{ud}$, for unit $u \in 1:U$ and vaccination cohort $d \in 0:5$.
% We also note here that the variance parameter $\obsOverdispersion^2$ is separately fit for the epidemic and endemic phases.
%
% \subsection*{Model~3}
%
% In this model, reported cholera cases are assumed to stem from individuals who develop symptoms and seek healthcare.
% Therefore reported cases are assumed to come from an over-dispersed negative binomial model, given the increase in infected individuals:
% \begin{equation}
%   \label{model3:meas}
%   y_{u, t} \mid \Delta N_{S_{ud}I_{ud}} = z_{u, t} \sim \text{NB}\left(\reportRate z_{u, t}, \obsOverdispersion \right)
% \end{equation}
%
% This measurement model is a minor change from that used in \cite{lee20}, which allowed for a change in the reporting rate on January 1st, 2018.
% The fitted values of the reporting rate---before and after January 2018---were $0.97$ and $0.097$, respectively.
% This major change in reporting rate alone could have been the cause that Model~3 originally failed to predict the eradication of cholera, as an overnight change from near perfect to almost non-existent reporting forces the model to explain the observed decrease in cases as a decrease in the reporting of cases rather than of prevalence of cases.
% This shift was justified by a "change of the case definition that occurred on January 1st, 2018";
% this claim was not cited, and we could find no evidence that such a drastic change in the reporting rate would be warranted.

% \end{appendix}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Multiple Appendixes:                     %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Support information, if any,             %%
%% should be provided in the                %%
%% Acknowledgements section.                %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\begin{acks}[Acknowledgments]
% The authors would like to thank ...
%\end{acks}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Funding information, if any,             %%
%% should be provided in the                %%
%% funding section.                         %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{funding}
This work was supported by National Science Foundation grants DMS-1761603 and DMS-1646108.
%
% The second author was supported in part by ...
\end{funding}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Supplementary Material, including data   %%
%% sets and code, should be provided in     %%
%% {supplement} environment with title      %%
%% and short description. It cannot be      %%
%% available exclusively as external link.  %%
%% All Supplementary Material must be       %%
%% available to the reader on Project       %%
%% Euclid with the published article.       %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{supplement}
\stitle{Eliminating cholera in Haiti: Supplement}
\sdescription{
This document contains additional details for Models~1--3, as well as a translation table that facilitates comparisons between these models and those described in \cite{lee20}.
The supplement also demonstrates our capability to faithfully replicate the results
of \cite{lee20}.
}
\end{supplement}

\begin{supplement}
\stitle{\code{haitipkg}}
\sdescription{
This \code{R} package is contained in a public GitHub repository: zjiang2/haitipkg. The package contains all of the data and code used to create and fit the models, as well as other useful functions that were used in this article.
}
\end{supplement}

\begin{supplement}
\stitle{\code{jesseuwheeler/haiti}}
\sdescription{
This GitHub repository contains the \code{.Rnw} files that were used to create this document and the supplement material.
}
\end{supplement}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                  The Bibliography                       %%
%%                                                         %%
%%  imsart-nameyear.bst  will be used to                   %%
%%  create a .BBL file for submission.                     %%
%%                                                         %%
%%  Note that the displayed Bibliography will not          %%
%%  necessarily be rendered by Latex exactly as specified  %%
%%  in the online Instructions for Authors.                %%
%%                                                         %%
%%  MR numbers will be added by VTeX.                      %%
%%                                                         %%
%%  Use \cite{...} to cite references in text.             %%
%%                                                         %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%% if your bibliography is in bibtex format, uncomment commands:
\bibliographystyle{imsart-nameyear} % Style BST file
\bibliography{bib-haiti}       % Bibliography file (usually '*.bib')

\end{document}

### 2022-08-05 ms.Rnw
Refactored Model 1 scenario code

@@
 RUN_LEVEL <- 1
+rl_dir <- paste0("run_level_", RUN_LEVEL, "/")
+
+for (i in 1:3) {
+  if (!dir.exists(paste0("model", i, '/', rl_dir))) {
+    dir.create(paste0("model", i, '/', rl_dir), recursive = TRUE)
+  }
+}
 
@@
 
-myround<- function (x, digits = 1) {
+myround <- function(x, digits = 1) {
   # taken from the broman package
@@
 
+# TODO: Remove these functions and replace with lubridate functions (also in the package).
 dateToYears <- function(date, origin = as.Date("2014-01-01"), yr_offset = 2014) {
@@
 
-stew(
-  file = "model2/output/model2_fit.rda", {h2_fit <- fit_haiti2()}
-)
-
-# load('model2/output/model2_fit_params.rda')
-# p2_epi <- h2_epi_params
-# h2_fit$end_params <- h2_end_params
-# rm(h2_epi_params, h2_end_params)
-# gc()
+# Fit model 2. Note that this takes the same ammount of time at all run
+# levels.
+h2_fit <- bake(
+  file = paste0("model2/", rl_dir, "model2_fit.rds"), {
+    fit_haiti2()
+  },
+  timing = FALSE
+)
 
@@
 <<Load Model 1 VaccScen Sims, message=FALSE, include=FALSE>>=
-load('model1/vacscens/sims_S0.rda')
-mod1_V0_sims <- agg_mod1_sims(sims)
-mod1_V0_probs <- get_elimProbs(sims, model = 1)
-rm(sims)
-gc()
-
-load('model1/vacscens/sims_S2.rda')
-mod1_V1_sims <- agg_mod1_sims(sims)
-mod1_V1_probs <- get_elimProbs(sims, model = 1)
-rm(sims)
-gc()
-
-load('model1/vacscens/sims_S4.rda')
-mod1_V2_sims <- agg_mod1_sims(sims)
-mod1_V2_probs <- get_elimProbs(sims, model = 1)
-rm(sims)
-gc()
-
-load('model1/vacscens/sims_S3.rda')
-mod1_V3_sims <- agg_mod1_sims(sims)
-mod1_V3_probs <- get_elimProbs(sims, model = 1)
-rm(sims)
-gc()
-
-load('model1/vacscens/sims_S1.rda')
-mod1_V4_sims <- agg_mod1_sims(sims)
-mod1_V4_probs <- get_elimProbs(sims, model = 1)
-rm(sims)
+
+# Number of particles and simulations to use for pfilter and project from filter, respectively.
+h1_scen_NP   <- switch(RUN_LEVEL, 50, 500, 5000)
+h1_scen_sims <- switch(RUN_LEVEL, 20, 100,  500)
+
+# Simulate vaccination scenario from filtering distribution. The code
+# below creates a model, samples from the filtering distribution, and then
+# projects the vaccination scenario, under the specified model.
+h1_S0_sims <- bake(
+  file = paste0("model1/", rl_dir, "S0_sims.rds"), {
+
+    # Create the model
+    s0 <- haiti1_joint(vacscen = 'id0')
+
+    # Load coefficients
+    coef(s0) <- p1
+
+    # Get samples from filtering distribution
+    s0_pf <- pfilter(s0, save.state = TRUE, Np = h1_scen_NP)
+
+    # project vacc scenario, with filtering dist, and model.
+    project_from_filter2(mod = s0, PF = s0_pf, nsims = h1_scen_sims)
+  },
+  timing = FALSE
+)
+
+# Clean-up simulations for analysis/visualization
+mod1_V0_sims <- agg_mod1_sims(h1_S0_sims)
+mod1_V0_probs <- get_elimProbs(h1_S0_sims, model = 1)
+rm(h1_S0_sims)
+
+# Simulate vaccination scenario from filtering distribution. The code
+# below creates a model, samples from the filtering distribution, and then
+# projects the vaccination scenario, under the specified model.
+h1_S2_sims <- bake(
+  file = paste0('model1/', rl_dir, 'S2_sims.rds'), {
+
+    # Create the model
+    s2 <- haiti1_joint(vacscen = "id2")
+
+    # Set model parameters that need to be adjusted
+    depts <- 2
+    h1_par_names <- names(p1)
+    h1_params_temp <- c(p1, rep(0.0, 5 * depts))
+    names(h1_params_temp) <- c(
+      h1_par_names,
+      paste0("S", 1:depts, "_0"),
+      paste0("E", 1:depts, "_0"),
+      paste0("I", 1:depts, "_0"),
+      paste0("A", 1:depts, "_0"),
+      paste0("R", 1:depts, "_0")
+    )
+    coef(s2) <- h1_params_temp  # Set model parameters
+
+    # Get samples from filtering distribution for Scenario 2 model:
+    s2_pf <- pfilter(s2, save.state = TRUE, Np = h1_scen_NP)
+
+    # project vacc scenario, with filtering dist, and model.
+    project_from_filter2(mod = s2, PF = s2_pf, nsims = h1_scen_NP)
+  },
+  timing = FALSE
+)
+
+# Clean-up simulations for analysis/visualization
+mod1_V1_sims <- agg_mod1_sims(h1_S2_sims)
+mod1_V1_probs <- get_elimProbs(h1_S2_sims, model = 1)
+rm(h1_S2_sims)
+
+# Simulate vaccination scenario from filtering distribution. The code
+# below creates a model, samples from the filtering distribution, and then
+# projects the vaccination scenario, under the specified model.
+h1_S4_sims <- bake(
+  file = paste0('model1/', rl_dir, 'S4_sims.rds'), {
+
+    # Create the model
+    s4 <- haiti1_joint(vacscen = "id4")
+
+    # Set model parameters that need to be adjusted
+    depts <- 3
+    h1_par_names <- names(p1)
+    h1_params_temp <- c(p1, rep(0.0, 5 * depts))
+    names(h1_params_temp) <- c(
+      h1_par_names,
+      paste0("S", 1:depts, "_0"),
+      paste0("E", 1:depts, "_0"),
+      paste0("I", 1:depts, "_0"),
+      paste0("A", 1:depts, "_0"),
+      paste0("R", 1:depts, "_0")
+    )
+    coef(s4) <- h1_params_temp  # Set model parameters
+
+    # Get samples from filtering distribution for Scenario 2 model:
+    s4_pf <- pfilter(s4, save.state = TRUE, Np = h1_scen_NP)
+
+    # project vacc scenario, with filtering dist, and model.
+    project_from_filter2(mod = s4, PF = s4_pf, nsims = h1_scen_NP)
+  },
+  timing = FALSE
+)
+
+# Clean-up simulations for analysis/visualization
+mod1_V2_sims <- agg_mod1_sims(h1_S4_sims)
+mod1_V2_probs <- get_elimProbs(h1_S4_sims, model = 1)
+rm(h1_S4_sims)
+
+# Simulate vaccination scenario from filtering distribution. The code
+# below creates a model, samples from the filtering distribution, and then
+# projects the vaccination scenario, under the specified model.
+h1_S3_sims <- bake(
+  file = paste0('model1/', rl_dir, 'S3_sims.rds'), {
+
+    # Create the model
+    s3 <- haiti1_joint(vacscen = "id3")
+
+    # Set model parameters that need to be adjusted
+    depts <- 10
+    h1_par_names <- names(p1)
+    h1_params_temp <- c(p1, rep(0.0, 5 * depts))
+    names(h1_params_temp) <- c(
+      h1_par_names,
+      paste0("S", 1:depts, "_0"),
+      paste0("E", 1:depts, "_0"),
+      paste0("I", 1:depts, "_0"),
+      paste0("A", 1:depts, "_0"),
+      paste0("R", 1:depts, "_0")
+    )
+    coef(s3) <- h1_params_temp  # Set model parameters
+
+    # Get samples from filtering distribution for Scenario 2 model:
+    s3_pf <- pfilter(s3, save.state = TRUE, Np = h1_scen_NP)
+
+    # project vacc scenario, with filtering dist, and model.
+    project_from_filter2(mod = s3, PF = s3_pf, nsims = h1_scen_NP)
+  },
+  timing = FALSE
+)
+
+# Clean-up simulations for analysis/visualization
+mod1_V3_sims <- agg_mod1_sims(h1_S3_sims)
+mod1_V3_probs <- get_elimProbs(h1_S3_sims, model = 1)
+rm(h1_S3_sims)
+
+# Simulate vaccination scenario from filtering distribution. The code
+# below creates a model, samples from the filtering distribution, and then
+# projects the vaccination scenario, under the specified model.
+h1_S1_sims <- bake(
+  file = paste0('model1/', rl_dir, 'S1_sims.rds'), {
+
+    # Create the model
+    s1 <- haiti1_joint(vacscen = "id1")
+
+    # Set model parameters that need to be adjusted
+    depts <- 10
+    h1_par_names <- names(p1)
+    h1_params_temp <- c(p1, rep(0.0, 5 * depts))
+    names(h1_params_temp) <- c(
+      h1_par_names,
+      paste0("S", 1:depts, "_0"),
+      paste0("E", 1:depts, "_0"),
+      paste0("I", 1:depts, "_0"),
+      paste0("A", 1:depts, "_0"),
+      paste0("R", 1:depts, "_0")
+    )
+    coef(s1) <- h1_params_temp  # Set model parameters
+
+    # Get samples from filtering distribution for Scenario 2 model:
+    s1_pf <- pfilter(s1, save.state = TRUE, Np = h1_scen_NP)
+
+    # project vacc scenario, with filtering dist, and model.
+    project_from_filter2(mod = s1, PF = s1_pf, nsims = h1_scen_NP)
+  },
+  timing = FALSE
+)
+
+# Clean-up simulations for analysis/visualization
+mod1_V4_sims <- agg_mod1_sims(h1_S1_sims)
+mod1_V4_probs <- get_elimProbs(h1_S1_sims, model = 1)
+rm(h1_S1_sims)
+
 gc()
@@
 
-stew(
-  file = "model2/output/VaccinationScenarios.rda", {
-    mod2_VaccScenarios <- haiti2_vaccScenario(
+# Run Vaccination scenarios for Model 2. Note that the code runs the same
+# speed for each run-level.
+mod2_VaccScenarios <- bake(
+  file = paste0("model2/", rl_dir, "VaccinationScenarios.rds"), {
+    haiti2_vaccScenario(
       end_params = h2_fit$end_params,
@@
     )
-  }
-)
+  },
+  timing = FALSE
+)
+
 @

### 2022-08-05 ms.Rnw
Replaced all stew calls with bakes (except one). Should work on other machines with required .rda files

@@
 
-true_agg_cases <- haitiCholera %>%
+dep_plot_df <- haitiCholera %>%
   select(-report) %>%
@@
     dep = case_when(dep == "Grand_Anse" ~ "Grande_Anse", TRUE ~ dep)
-  ) %>%
+  )
+
+true_agg_cases <- dep_plot_df %>%
   tidyr::pivot_wider(
@@
 <<Model 3 PanelSims,echo=FALSE>>=
+h3_panel_sims <- switch(RUN_LEVEL, 20, 100, 500)
+
 departements <-
@@
 
-stew(
-  file = "model3/output/sims_FinalPanel.rda", {
-    foreach(dep=departements, .combine = rbind) %dopar% {
+h3_panel_sims <- bake(
+  file = paste0('model3/', rl_dir, 'panelPompSims.rds'), {
+    foreach(dep = departements, .combine = rbind) %dopar% {
       SIRB <- unitobjects(h3)[[dep]]
@@
 
-      pomp::simulate(SIRB,
+      pomp::simulate(
+        SIRB,
         params = c(shared, specific),
-                     nsim = 500, format = 'data.frame',
-                     include.data = TRUE) -> sims
+        nsim = h3_panel_sims,
+        format = 'data.frame'
+      ) -> sims
 
@@
       sims
-    } -> h3_panel_all_sims
-
-    h3_panel_quantiles <- h3_panel_all_sims %>%
-      mutate(isdata = .id == 'data') %>%
-      # filter(!isdata) %>%
-      group_by(isdata, dep, time) %>%
+    }
+  }
+)
+
+h3_panel_quantiles <- h3_panel_sims %>%
+  mutate(
+    date = as.Date(
+      lubridate::round_date(lubridate::date_decimal(time), unit = 'day')
+    )
+  ) %>%
+  group_by(dep, date) %>%
   summarise(
@@
     q95 = quantile(cases, 0.975, na.rm = T)
-      )
-
-    h3_panel_real_data <- h3_panel_quantiles %>%
-      filter(isdata)
-
-    h3_panel_sim_data <- h3_panel_quantiles %>%
-      filter(!isdata)
-
-  }
-)
-
-h3_panel_sim_data <- h3_panel_sim_data %>%
+  ) %>%
+  ungroup() %>%
   mutate(
-    dep = gsub("-", "_", dep),
-    date = yearsToDateTime(time)
+    dep = gsub("-", "_", dep)
   ) %>%
-  filter(date >= yearsToDate(h3@unit.objects$Artibonite@t0)) %>%
-  mutate(date = as.Date(lubridate::round_date(date)))
-
-h3_panel_real_data <- h3_panel_real_data %>%
-  mutate(
-    dep = gsub("-", "_", dep),
-    date = yearsToDateTime(time)
-  ) %>%
-  filter(date >= yearsToDate(h3@unit.objects$Artibonite@t0)) %>%
-  mutate(date = as.Date(lubridate::round_date(date)))
+  filter(date >= yearsToDate(h3@unit.objects$Artibonite@t0))
 
@@
 <<Model3_PanelSims_Figure, echo=FALSE, fig.height=3>>=
-
 ggplot() +
-  geom_line(data = h3_panel_sim_data, aes(x = date, y = q50 + 1), col = 'blue') +
-  geom_line(data = h3_panel_real_data, aes(x = date, y = q50 + 1), col = 'black') +
-  geom_ribbon(data = h3_panel_sim_data, aes(x = date, ymin = q05 + 1, ymax = q95 + 1),
+  geom_line(data = h3_panel_quantiles, aes(x = date, y = q50 + 1), col = 'blue') +
+  geom_line(data = dep_plot_df, aes(x = date, y = cases + 1), col = 'black') +
+  geom_ribbon(data = h3_panel_quantiles, aes(x = date, ymin = q05 + 1, ymax = q95 + 1),
               alpha = 0.4) +
@@
   scale_y_log10(labels = scales::trans_format("log10", scales::math_format(10^.x)))
-  # theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1,
-  #                                  size = 7),
-        # axis.title.x = element_blank())
+
 @
@@
 <<Remove Model3 panelSims, cache=FALSE, include=FALSE, message=FALSE, echo=FALSE>>=
-rm(h3_panel_all_sims, h3_panel_sim_data)
+rm(h3_panel_quantiles, h3_panel_sims)
 gc()
@@
 
+h3_spat_nsim <- switch(RUN_LEVEL, 20, 100, 500)
+
+h3Spat_quants <- bake(
+  file = paste0('model3/', rl_dir, 'SpatPompSims.rds'), {
     h3Spat <- haiti3_spatPomp()
@@
 
-registerDoRNG(6687147)
-
-stew(
-  file = "model3/output/SpatPompSims.rda", {
-    h3Spat_sims <- simulate(h3Spat, nsim = 500, seed = 784713,
-                     format = "data.frame", include.data = TRUE)
-
-    h3Spat_quants <- h3Spat_sims %>%
-      mutate(is.data = .id == 'data') %>%
-      filter(!is.data) %>%
+    h3Spat_sims <- simulate(
+      h3Spat, nsim = h3_spat_nsim, format = 'data.frame'
+    )
+
+    h3Spat_sims %>%
       rename(dep = unitname) %>%
       group_by(dep, time) %>%
-      summarise(q05 = quantile(cases, 0.025, na.rm = T),
+      summarise(
+        q05 = quantile(cases, 0.025, na.rm = T),
         mean = mean(cases, na.rm = T),
         q50 = quantile(cases, 0.5, na.rm = T),
-                q95 = quantile(cases, 0.975, na.rm = T)) %>%
+        q95 = quantile(cases, 0.975, na.rm = T)
+      ) %>%
       ungroup() %>%
-      mutate(date = yearsToDateTime(time)) %>%
+      mutate(
+        date = lubridate::date_decimal(time)
+      ) %>%
       filter(date >= yearsToDate(h3Spat@t0)) %>%
-      mutate(date = as.Date(lubridate::round_date(date)))
-
-    h3Spat_truth <- h3Spat_sims %>%
-      # select(time, .id, starts_with("cases")) %>%
-      mutate(is.data = .id == 'data') %>%
-      filter(is.data) %>%
-      rename(dep = unitname) %>%
-      group_by(dep, time) %>%
-      summarise(q05 = quantile(cases, 0.025, na.rm = T),
-                mean = mean(cases, na.rm = T),
-                q50 = quantile(cases, 0.5, na.rm = T),
-                q95 = quantile(cases, 0.975, na.rm = T)) %>%
-      ungroup() %>%
-      mutate(date = yearsToDateTime(time)) %>%
-      filter(date >= yearsToDate(h3Spat@t0)) %>%
-      mutate(date = as.Date(lubridate::round_date(date)))
-
-  }
+      mutate(date = as.Date(lubridate::round_date(date, unit = 'day')))
+  },
+  timing = FALSE
 )
@@
   geom_line(data = h3Spat_quants, aes(x = date, y = q50 + 1), col = 'blue') +
-  geom_line(data = h3Spat_truth, aes(x = date, y = q50 + 1), col = 'black') +
+  geom_line(data = dep_plot_df, aes(x = date, y = cases + 1), col = 'black') +
   geom_ribbon(data = h3Spat_quants,
@@
   scale_y_log10(labels = scales::trans_format("log10", scales::math_format(10^.x)))
-  # theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1,
-  #                                  size = 6),
-  #       axis.title.x = element_blank())
 @
@@
 <<Remove Model3 spatSims, cache=FALSE, include=FALSE, message=FALSE, echo=FALSE>>=
-rm(h3Spat_sims, h3Spat_truth, h3Spat_quants)
+rm(h3Spat_quants, h3_spat_nsim)
 gc()
@@
 <<Lee3 Sub Calculations, echo=FALSE, include=FALSE, message=FALSE>>=
+lee3_NP   <- switch(RUN_LEVEL, 50, 250, 5000)
+lee3_NREP <- switch(RUN_LEVEL,  3,   8,   36)
+
 registerDoRNG(49637421)
@@
 
-stew(
-  # Save the file
-  file = 'model3/output/lee3_results.rda', {
+lee3_results <- bake(
+  file = paste0('model3/', rl_dir, 'lee3_results.rds'), {
 
@@
       lee3_results[["sub_ll_evals"]] <- foreach(
-        i = 1:36, .combine = c
+        i = 1:lee3_NREP, .combine = c
       ) %dopar% {
-        logLik(pfilter(lee3_sub, Np = 5000))  # Get likelihood evaluations
+        logLik(pfilter(lee3_sub, Np = lee3_NP))  # Get likelihood evaluations
       }
@@
       lee3_results[["all_ll_evals"]] <- foreach(
-        i = 1:36, .combine = c
+        i = 1:lee3_NREP, .combine = c
       ) %dopar% {
-        logLik(pfilter(lee3_all, Np = 5000))  # Get likelihood evaluations
+        logLik(pfilter(lee3_all, Np = lee3_NP))  # Get likelihood evaluations
       }
@@
     lee3_results[["n_params"]] <- 29
-  }
+
+    lee3_results
+  },
+  timing = FALSE
 )
@@
 mod1_V4_probs <- get_elimProbs(h1_S1_sims, model = 1)
-rm(h1_S1_sims)
+rm(h1_S1_sims, h1_scen_NP, h1_scen_sims)
 

### 2022-08-13 ms.Rnw
Replaced 'load' statements with 'bake'

@@
 <<Table 2 Input, echo=FALSE, eval=TRUE>>=
+NREPS_EVAL <- switch(RUN_LEVEL,  3,   8,    36)
+NP_EVAL    <- switch(RUN_LEVEL, 50, 500, 10000)
 
@@
 
-load('model3/output/FinalPanelAsSpatPomp_PF.rda')
-rm(pf, ll)
-mod3_spatPomp_ll <- ll_est[1]
-rm(ll_est)
+h3Spat <- haiti3_spatPomp()
+coef(h3Spat) <- haitipkg:::panelParms_toSpatParms3(p3)
+
+# Calculate likelihood via block-particle filter
+mod3_spatPomp_res <- bake(
+  file = paste0('model3/', rl_dir, 'SpatPomp_res.rds'), {
+
+    # Create a list to store results
+    results <- list()
+
+    # Perform block-particle filter
+    h3_bpf <- foreach(i = 1:NREPS_EVAL, .combine = c) %dopar% {
+      bpfilter(h3Spat, Np = NP_EVAL, block_size = 1)
+    }
+
+    ll <- logLik(h3_bpf)
+    results$ll <- logmeanexp(ll[!is.na(ll)], se = FALSE)
+
+    ### Get likelihood for subset of the data
+
+    # Find appropriate sub-set
+    in_subset_cols <- h3_bpf[[1]]@times >= as.numeric(dateToYears(as.Date("2014-03-01")))
+    in_subset_rows <- h3_bpf[[1]]@unit_names != 'Ouest'
+
+    mod3_coupled_subset_evals <- c()
+    for (i in 1:length(pf)) {
+      mod3_coupled_subset_evals <- c(
+        mod3_coupled_subset_evals,
+        sum(h3_bpf[[i]]@block.cond.loglik[in_subset_rows, in_subset_cols])
+      )
+    }
+
+    results$subset_ll <- logmeanexp(mod3_coupled_subset_evals)
+    results
+  },
+  timing = FALSE
+)
+
 mod3_spatPomp_n_params <- 39
-mod3_spatPomp_aic <- 2 * mod3_spatPomp_n_params - 2 * mod3_spatPomp_ll
+mod3_spatPomp_aic <- 2 * mod3_spatPomp_n_params - 2 * mod3_spatPomp_res$ll
+mod3_spatPomp_res$subset_aic <- 2 * mod3_spatPomp_n_params - 2 * mod3_spatPomp_res$subset_ll
+
+rm(NREPS_EVAL, NP_EVAL)
 @
@@
 
-rm(lee3_sub, lee3_all)
+rm(lee3_sub, lee3_all, lee3_NP, lee3_NREP)
 gc()
-@
-
-<<old TODO: Replace this with the above>>=
-# Load Lee et al Model 3 likelihood for entire data
-# load('model3/output/lee_mod3_coupled_allData_ll.rda')
-# mod3_lee_coupled_ll <- max(lee3_results[["ll_evals_sub_data"]])
-# mod3_lee_coupled_n_params <- 29
+
 mod3_lee_coupled_aic <- 2 * lee3_results$n_params - 2 * max(lee3_results$all_ll_evals)
 
-# rm(coupled_ll_evals_all_data, coupled_ll_results_all_data, tPfilter_lee_coupled)
-# gc()
-
-# Load Lee et al Model 3 likelihood on partial data
-# load('model3/output/lee_mod3_coupled_ll.rda')
-# mod3_lee_coupled_partial_ll <- max(coupled_ll_evals_sub_data)
 mod3_lee_coupled_partial_aic <- 2 * lee3_results$n_params - 2 * max(lee3_results$sub_ll_evals)
 
-# Load our spatPOMP Model 3 on likelihood on partial data
-load('model3/output/mod3_coupled_subset_ll.rda')
-mod3_coupled_subset_aic <- 2 * mod3_spatPomp_n_params - 2 * mod3_coupled_subset_ll
 @
@@
   \multirow{2}{*}{$\Sexpr{myround(mod3_ll[1], 1)}$\footnotemark[2]} &
-  $\Sexpr{myround(mod3_spatPomp_ll, 1)}$ \\
+  $\Sexpr{myround(mod3_spatPomp_res$ll, 1)}$ \\
     & ($\Sexpr{myround(mod1_lee_ll, 1)}$)\footnotemark[1] &
@@
 On this subset, the parameters provided by \citet{lee20} achieved a likelihood of $\Sexpr{myround(max(lee3_results$sub_ll_evals), 1)}$.
-On this same subset of data, our model achieved a likelihood of $\Sexpr{myround(mod3_coupled_subset_ll, 1)}$.}
+On this same subset of data, our model achieved a likelihood of $\Sexpr{myround(mod3_spatPomp_res$subset_ll, 1)}$.}
 As previously noted, simulations from a well-fit mechanistic model may closely resemble the observed data $y_{1:N}^*$.
@@
 In this particular case study, projecting the future state of the cholera epidemic in Haiti starting from the draws from the filtering distribution allows the proposed models to benefit from the fact that very few cholera cases had been observed between 2018 and 2019, and that cases appear to be decreasing (i.e., the number of susceptible individuals may be small).
-% Model projections given in the following section have been obtained using the technique suggested in this subsection.
-% Sentence above is unecessary - Jesse
 
@@
 <<Load Model 3 Vacc Scenarios, message=FALSE, include=FALSE>>=
-load('model3/output/sims_spatPomp_S0.rda')
-mod3_V0_sims <- agg_mod3_sims(sims)
-mod3_V0_probs <- get_elimProbs(sims, model = 3)
-rm(sims)
-gc()
-
-load('model3/output/sims_spatPomp_S2.rda')
-mod3_V1_sims <- agg_mod3_sims(sims)
-mod3_V1_probs <- get_elimProbs(sims, model = 3)
-rm(sims)
-gc()
-
-load('model3/output/sims_spatPomp_S4.rda')
-mod3_V2_sims <- agg_mod3_sims(sims)
-mod3_V2_probs <- get_elimProbs(sims, model = 3)
-rm(sims)
-gc()
-
-load('model3/output/sims_spatPomp_S3.rda')
-mod3_V3_sims <- agg_mod3_sims(sims)
-mod3_V3_probs <- get_elimProbs(sims, model = 3)
-rm(sims)
-gc()
-
-load('model3/output/sims_spatPomp_S1.rda')
-mod3_V4_sims <- agg_mod3_sims(sims)
-mod3_V4_probs <- get_elimProbs(sims, model = 3)
-rm(sims)
-gc()
+NP_BPF <- switch(RUN_LEVEL, 50, 200, 1000)
+NSIM   <- switch(RUN_LEVEL, 10,  50,  500)
+
+h3_bpf <- bake(
+  file = paste0("model3/", rl_dir, "h3_bpf.rds"),
+  expr = {
+    bpfilter(h3Spat, Np = NP_BPF, block_size = 1, save_states = TRUE)
+  },
+  timing = FALSE
+)
+
+mod3_V0_res <- bake(
+  file = paste0("model3/", rl_dir, "mod3_V0_res"), {
+
+    # Create a list to store results
+    h3_V0_res <- list()
+
+    # Get "noVacc" scenario parameters
+    h3_S0_par <- get_model3_vacc_scenario_params(scenario = "noVacc")
+    coef(h3Spat)[names(h3_S0_par)] <- h3_S0_par
+
+    # Using parameters and filtered distribution, project system
+    h3_V0_sims <- project_from_filter2(
+      h3Spat, PF = h3_bpf, covarGen = project_rain,
+      nsims = NSIM, seed = 6897154
+    )
+
+    # Store nationally aggregated results
+    h3_V0_res$mod3_V0_sims <- agg_mod3_sims(h3_V0_sims)
+    h3_V0_res$mod3_V0_probs <- get_elimProbs(h3_V0_sims, model = 3)
+
+    # Return results
+    h3_V0_res
+
+  }
+)
+
+mod3_V1_res <- bake(
+  file = paste0("model3/", rl_dir, "mod3_V1_res"), {
+
+    # Create a list to store results
+    h3_V1_res <- list()
+
+    # Get "noVacc" scenario parameters
+    h3_V1_par <- get_model3_vacc_scenario_params(scenario = "2dep")
+    coef(h3Spat)[names(h3_V1_par)] <- h3_V1_par
+
+    # Using parameters and filtered distribution, project system
+    h3_V1_sims <- project_from_filter2(
+      h3Spat, PF = h3_bpf, covarGen = project_rain,
+      nsims = NSIM, seed = 6897154
+    )
+
+    # Store nationally aggregated results
+    h3_V1_res$mod3_V1_sims <- agg_mod3_sims(h3_V1_sims)
+    h3_V1_res$mod3_V1_probs <- get_elimProbs(h3_V1_sims, model = 3)
+
+    # Return results
+    h3_V1_res
+
+  }
+)
+
+mod3_V2_res <- bake(
+  file = paste0("model3/", rl_dir, "mod3_V2_res"), {
+
+    # Create a list to store results
+    h3_V2_res <- list()
+
+    # Get "noVacc" scenario parameters
+    h3_V2_par <- get_model3_vacc_scenario_params(scenario = "3dep")
+    coef(h3Spat)[names(h3_V2_par)] <- h3_V2_par
+
+    # Using parameters and filtered distribution, project system
+    h3_V2_sims <- project_from_filter2(
+      h3Spat, PF = h3_bpf, covarGen = project_rain,
+      nsims = NSIM, seed = 6897154
+    )
+
+    # Store nationally aggregated results
+    h3_V2_res$mod3_V2_sims <- agg_mod3_sims(h3_V2_sims)
+    h3_V2_res$mod3_V2_probs <- get_elimProbs(h3_V2_sims, model = 3)
+
+    # Return results
+    h3_V2_res
+
+  }
+)
+
+mod3_V3_res <- bake(
+  file = paste0("model3/", rl_dir, "mod3_V3_res"), {
+
+    # Create a list to store results
+    h3_V3_res <- list()
+
+    # Get "noVacc" scenario parameters
+    h3_V3_par <- get_model3_vacc_scenario_params(scenario = "slowNation")
+    coef(h3Spat)[names(h3_V3_par)] <- h3_V3_par
+
+    # Using parameters and filtered distribution, project system
+    h3_V3_sims <- project_from_filter2(
+      h3Spat, PF = h3_bpf, covarGen = project_rain,
+      nsims = NSIM, seed = 6897154
+    )
+
+    # Store nationally aggregated results
+    h3_V3_res$mod3_V3_sims <- agg_mod3_sims(h3_V3_sims)
+    h3_V3_res$mod3_V3_probs <- get_elimProbs(h3_V3_sims, model = 3)
+
+    # Return results
+    h3_V3_res
+
+  }
+)
+
+mod3_V4_res <- bake(
+  file = paste0("model3/", rl_dir, "mod3_V4_res"), {
+
+    # Create a list to store results
+    h3_V4_res <- list()
+
+    # Get "noVacc" scenario parameters
+    h3_V4_par <- get_model3_vacc_scenario_params(scenario = "fastNation")
+    coef(h3Spat)[names(h3_V4_par)] <- h3_V4_par
+
+    # Using parameters and filtered distribution, project system
+    h3_V4_sims <- project_from_filter2(
+      h3Spat, PF = h3_bpf, covarGen = project_rain,
+      nsims = NSIM, seed = 6897154
+    )
+
+    # Store nationally aggregated results
+    h3_V4_res$mod3_V4_sims <- agg_mod3_sims(h3_V4_sims)
+    h3_V4_res$mod3_V4_probs <- get_elimProbs(h3_V4_sims, model = 3)
+
+    # Return results
+    h3_V4_res
+
+  }
+)
 @
@@
 In additional to probability of elimination estimates, we are also able to estimate the cumulative number of infections under each vaccination scenario from February 2019 -- February 2024.
-Notably, the median number of cumulative cholera infections under the no-vaccination scenario using Models 1 and 3 were $\Sexpr{prettyNum(round(as.numeric(mod1_V0_probs$cumInf['q50'])), big.mark = ",")}$ and $\Sexpr{prettyNum(round(as.numeric(mod3_V0_probs$cumInf['q50'])), big.mark = ",")}$, respectively.
+Notably, the median number of cumulative cholera infections under the no-vaccination scenario using Models 1 and 3 were $\Sexpr{prettyNum(round(as.numeric(mod1_V0_probs$cumInf['q50'])), big.mark = ",")}$ and $\Sexpr{prettyNum(round(as.numeric(mod3_V0_res$mod3_V0_probs$cumInf['q50'])), big.mark = ",")}$, respectively.
 So far, these estimates are far more consistent with the observed number of reported cholera cases than the corresponding estimates from \cite{lee20}, which were approximately $400,000$ and $1,000,000$.
@@
 gg_m3_V0 <- ggplot() +
-  geom_line(data = mod3_V0_sims, aes(x = date, y = q50 + 1), col = 'blue') +
-  geom_ribbon(data = mod3_V0_sims, aes(x = date, ymin = q05 + 1, ymax = q95 + 1), alpha = 0.5) +
+  geom_line(data = mod3_V0_res$mod3_V0_sims, aes(x = date, y = q50 + 1), col = 'blue') +
+  geom_ribbon(data = mod3_V0_res$mod3_V0_sims, aes(x = date, ymin = q05 + 1, ymax = q95 + 1), alpha = 0.5) +
   geom_line(data = true_agg_cases, aes(x = date, y = ReportedAll + 1)) +
@@
 gg_m3_V1 <- ggplot() +
-  geom_line(data = mod3_V1_sims, aes(x = date, y = q50 + 1), col = 'blue') +
-  geom_ribbon(data = mod3_V1_sims, aes(x = date, ymin = q05 + 1, ymax = q95 + 1), alpha = 0.5) +
+  geom_line(data = mod3_V1_res$mod3_V1_sims, aes(x = date, y = q50 + 1), col = 'blue') +
+  geom_ribbon(data = mod3_V1_res$mod3_V1_sims, aes(x = date, ymin = q05 + 1, ymax = q95 + 1), alpha = 0.5) +
   geom_line(data = true_agg_cases, aes(x = date, y = ReportedAll + 1)) +
@@
 gg_m3_V2 <- ggplot() +
-  geom_line(data = mod3_V2_sims, aes(x = date, y = q50 + 1), col = 'blue') +
-  geom_ribbon(data = mod3_V2_sims, aes(x = date, ymin = q05 + 1, ymax = q95 + 1), alpha = 0.5) +
+  geom_line(data = mod3_V2_res$mod3_V2_sims, aes(x = date, y = q50 + 1), col = 'blue') +
+  geom_ribbon(data = mod3_V2_res$mod3_V2_sims, aes(x = date, ymin = q05 + 1, ymax = q95 + 1), alpha = 0.5) +
   geom_line(data = true_agg_cases, aes(x = date, y = ReportedAll + 1)) +
@@
 gg_m3_V3 <- ggplot() +
-  geom_line(data = mod3_V3_sims, aes(x = date, y = q50 + 1), col = 'blue') +
-  geom_ribbon(data = mod3_V3_sims, aes(x = date, ymin = q05 + 1, ymax = q95 + 1), alpha = 0.5) +
+  geom_line(data = mod3_V3_res$mod3_V3_sims, aes(x = date, y = q50 + 1), col = 'blue') +
+  geom_ribbon(data = mod3_V3_res$mod3_V3_sims, aes(x = date, ymin = q05 + 1, ymax = q95 + 1), alpha = 0.5) +
   geom_line(data = true_agg_cases, aes(x = date, y = ReportedAll + 1)) +
@@
 gg_m3_V4 <- ggplot() +
-  geom_line(data = mod3_V4_sims, aes(x = date, y = q50 + 1), col = 'blue') +
-  geom_ribbon(data = mod3_V4_sims, aes(x = date, ymin = q05 + 1, ymax = q95 + 1), alpha = 0.5) +
+  geom_line(data = mod3_V4_res$mod3_V4_sims, aes(x = date, y = q50 + 1), col = 'blue') +
+  geom_ribbon(data = mod3_V4_res$mod3_V4_sims, aes(x = date, ymin = q05 + 1, ymax = q95 + 1), alpha = 0.5) +
   geom_line(data = true_agg_cases, aes(x = date, y = ReportedAll + 1)) +
@@
 
-mod3_V0_probs$ElimTime$scenario = "V0"
-mod3_V1_probs$ElimTime$scenario = "V1"
-mod3_V2_probs$ElimTime$scenario = "V2"
-mod3_V3_probs$ElimTime$scenario = "V3"
-mod3_V4_probs$ElimTime$scenario = "V4"
+mod3_V0_res$mod3_V0_probs$ElimTime$scenario = "V0"
+mod3_V1_res$mod3_V1_probs$ElimTime$scenario = "V1"
+mod3_V2_res$mod3_V2_probs$ElimTime$scenario = "V2"
+mod3_V3_res$mod3_V3_probs$ElimTime$scenario = "V3"
+mod3_V4_res$mod3_V4_probs$ElimTime$scenario = "V4"
 
 mod3_probElims <- rbind(
-  mod3_V0_probs$ElimTime,
-  mod3_V1_probs$ElimTime,
-  mod3_V2_probs$ElimTime,
-  mod3_V3_probs$ElimTime,
-  mod3_V4_probs$ElimTime
+  mod3_V0_res$mod3_V0_probs$ElimTime,
+  mod3_V1_res$mod3_V1_probs$ElimTime,
+  mod3_V2_res$mod3_V2_probs$ElimTime,
+  mod3_V3_res$mod3_V3_probs$ElimTime,
+  mod3_V4_res$mod3_V4_probs$ElimTime
 )

### 2022-08-22 ms.Rnw
Significant edit. Changed order of Model 2 & Model 1 subsections, added paragraphs and citations

@@
 Such decisions can be posed as statistical questions about scientifically motivated dynamic models.
-\eic{Thus, we encounter the methodological task of building credible, data-informed decisions for stochastic, partially observed, nonlinear dynamic models.
+Thus, we encounter the methodological task of building credible, data-informed decisions for stochastic, partially observed, nonlinear dynamic models.
 This necessitates addressing the tradeoff between biological fidelity and model simplicity, and the reality of misspecification for models at all levels of complexity.
-THE CURRENT MS VERSION DOES NOT QUITE AIM AT THESE ISSUES.
-BUT, THEY ARE HIGHLY RELEVANT , AND GIVE A BROADER STATISTICAL CONTEXT FOR THE EPIDEMIOLOGICAL CASE STUDY.
-MAYBE WE CAN EDIT IN THAT DIRECTION?}
+\eic{EDIT IN THE DIRECTION OF THE PREVIOUS SENTENCE.}
 As a case study, we consider a cholera epidemic in Haiti.
@@
 The four independent teams were given the task of projecting the
-potential effect of various vaccination deployment strategies on the cholera epidemic in Haiti.
-\jwc{TODO: MAYBE DISCUSS MORE ABOUT THE VACCINES, LIKE WHAT EFFECTS THEY ARE KNOWN TO HAVE, NOTE HOW VACCINE SUPPLIES ARE LIMITED, ETC.}
-Certain data were shared between the groups, including demography and vaccination history; vaccine efficacy was also fixed at a shared value between groups.
+potential effect of increased oral cholera vaccine (OCV) efforts across Haiti.
+\jwc{New sentence:}
+OCV is accepted as a safe and effective tool for controlling the spread of cholera, however the available stockpile and production of OCV doses is insufficient to meet global needs \citep{lee20, pezzoli20}.
+In the study, certain data were shared between the groups, including demography and vaccination history; vaccine efficacy was also fixed at a shared value between groups.
 Beyond this, the groups made autonomous decisions on what to include and exclude from their models; this autonomy reduced the possible effect that assumptions about the dynamic system may have on the final conclusion of the study.
@@
 Here, we demonstrate that attention to certain statistical considerations could have improved the fit to previous data, as well as leading to forecasts that match more closely the observed outcome.
+\jwc{NEXT TWO SENTENCES ARE NEW: }
+Because our primary focus is on the statistical tasks of fitting and diagnosing mechanistic models, we refrain from commenting on the feasibility of scientific assumptions implicitly made in \cite{lee20}.
+We demonstrate that model-mispecification does not alone explain the discrepancy between model predictions and reality, as has been suggested \citep{rebaudetComment20}.
 Based on this retrospective analysis, we offer suggestions on fitting mechanistic models to dynamic systems for future studies.
@@
 Many dynamic systems of scientific and public interest are highly complex by nature; therefore when developing a mathematical description of a dynamic system, it is often necessary to impose assumptions on the system in order to make inference possible.
-One common assumption in epidemiology is a homogeneous mixing population \jwc{REF???}.
+For example, one common assumption in epidemiology is a homogeneous mixing population \citep{bansal07}.
 Other common considerations include whether the proposed model should be stochastic or deterministic; whether the model should have change points in parameter values or should otherwise make adjustments for changes in the dynamic system; and whether the proposed model should include any spatial heterogeneity at a scale permissible by the observed data.
-Ultimately, these decisions will result in a different approach to fitting the model.
+Ultimately, these decisions will result in a different approach to calibrating (or fitting) model parameters.
 
@@
 Because of the spatial structure of Models~2 and~3, we refer to these models as SpatPOMPs \citep{asfaw21arxiv}.
-
 In the following subsections, we describe approaches to fitting mechanistic models and discuss potential trade-offs of the selected model complexity.
-We begin by fitting the parameters of Model~2, as the deterministic data generating process makes the task of parameter estimation the easiest of the three considered models.
-
-% \eic{USE SOMETHING OTHER THAN $D$ FOR THE NUMBER OF VACCINATION CLASSES, SINCE $D$ IS USED FOR THE DEATH/EMIGRATION/SINK COMPARTMENT.
-ALSO, MAYBE USE $B$ FOR A BIRTH/IMMIGRATION/SOURCE COMPARTMENT RATHER THAN $D$ FOR BOTH?}
-% \jwc{NEW COMMANDS HAVE BEEN MADE FOR VACCINATION CLASSES (\code{vaccClass}) AND THE COUNTERS (\code{vaccCounter}), SO WE CAN JUST CHANGE THE HEADER AND THE WHOLE DOCUMENT WILL UPDATE.}
-
-\subsubsection{Model~2}
-
-Model~2 is a deterministic model, which can be viewed as a special case of a POMP (or a and SpatPOMP) with no randomness in the dynamic model.
-By combining a deterministic process model with a simple Gaussian measurement model, Model~2 reduces model fitting to a least squares calculation over parameters in a set of differential equations.
-Deterministic compartment models have a long history in the field of infectious disease epidemiology \citep{kermack1927,brauer2017,varghese21},
-and can be justified by asymptotic considerations in a large-population limit \citep{dadlani2020,ndii17}.
-
-\citet{lee20} originally fit two versions of model~2 based on a presupposed change in cholera transmission from a epidemic phase to endemic phase that occurred in March, 2014.
-We follow their decision to re-estimate model parameters at this break-point, but we include a requirement that the latent state $\bm{X}^{(2)}(t)$ at the start of the endemic period must be the same as the state at the end of the epidemic period.
-This additional constraint is sensible for a mechanistic interpretation of the latent state, though it adds difficulty to the task of obtaining a model fit that closely resembles the observed data.
-To combat this added difficulty, and to allow for a possible shift in the seasonality component in the force of infection, we introduce an additional phase parameter $\phaseParm$ in Eq.~\ref{model2:lambda}.
-% \begin{equation}
-% \lambda_{u}(t) = 0.5\big(1+\seasAmplitude \cos(2\pi t + \phaseParm)\big)
-% \frac{\beta_W\, W_u(t)}{ \Wsat  + W_u(t)} +
-% \beta \left\{\sum_{d=0}^4 I_{ud}(t) + \asymptomRelativeInfect \sum_{d=0}^4 A_{ud}(t) \right\}\label{eq:model2:phase}
-% \end{equation}
-% \eic{WE COULD SAVE AN EQUATION BY WRITING THE MORE GENERAL ONE INITIALLY AND SAYING THAT THEY ASSUMED $\phaseParm=0$}
-
-% We also note that, without the inclusion of a phase parameter, there is an implicit assumption is that force of infection is greatest around January and weakest around June, and the addition of this parameter allows for the testing of this assumption.
-We further increase model flexibility by fitting the parameter $\Wshed$ rather than treating it as fixed.
-We implemented this model using the \code{spatPomp} \code{R} package \citep{asfaw21github}.
-The model was then fit using the subplex algorithm, implemented in the  \code{subplex} package \citep{king2020Subplex}.
-A comparison of the trajectory of the fitted model to the data is given in Fig~\ref{fig:mod2Traj}.
-
-<<Model 2 Trajectory, echo=FALSE>>=
-h2_epi_traj <- trajectory(h2_epi, params = h2_fit$epi_params, format = 'data.frame')
-h2_epi_traj$Ctotal <- rowSums(h2_epi_traj[, paste0("C", 1:10)]) * h2_fit$epi_params['Rho']
-
-h2_end_traj <- trajectory(h2_end, params = h2_fit$end_params, format = 'data.frame')
-h2_end_traj$Ctotal <- rowSums(h2_end_traj[, paste0("C", 1:10)]) * h2_fit$end_params['Rho']
-
-h2_traj <- rbind(
-  h2_epi_traj[, c('year', 'Ctotal')],
-  h2_end_traj[, c('year', 'Ctotal')]
-) %>%
-  mutate(
-    date = yearsToDate(year)
-  )
-@
-
-%
-% \begin{figure}[ht]
-% ggplot() +
-%   geom_line(data = true_agg_cases, aes(x = date, y = log(ReportedAll + 1))) +
-%   geom_line(data = h2_traj, aes(x = date, y = log(Ctotal + 1)), col = 'blue') +
-%   theme(axis.title.x = element_blank()) +
-%   ylab("(Log) Reported Cases") +
-%   geom_vline(xintercept = yearsToDate(2014.161), linetype = 'dashed')
-% @
-% \caption{\label{fig:mod2Traj}
-% Simulated trajectory of Model~2 (blue curve) compared to reported cholera incidence (black curve).
-The dashed vertical line shows where the endemic phase starts, and where parameter values are re-estimated.
-% }
-% \end{figure}
-%
 
@@
 
-Another approach to modelling a dynamic system is through probabilistic models.
+One approach to modelling a dynamic system is through probabilistic models.
 With a probabilistic model,
@@
 Once a model for the system has been proposed, the parameter vector $\theta$ needs to be estimated using the observed data.
-There exist several modelling techniques that can be used to obtain maximum likelihood estimates (MLE) of the parameters in stochastic dynamic models, including the EM algorithm, Kalman Filter (and extensions) \citep{evensen09}, and iterated filtering algorithms \citep{ionides15}.
-In order to use models that are scientifically meaningful rather than only those that are statistically convenient, we restrict ourselves to parameter estimation techniques for POMP models that have the plug-and-play property, which is that the fitting procedure only requires a way to simulate the latent process instead of the ability to evaluate transition densities \citep{breto09,he10}.
-Plug-and-play algorithms include ABC and PMCMC \citep{toni09,andrieu10}, but we focus on likelihood maximization via iterated filtering algorithms, which modify the well-known particle filter \citep{arulampalam02} by performing a random walk for each parameter and particle.
-These perturbations are carried out iteratively over multiple filtering operations, using the collection of parameters from the previous filtering pass as the parameter initialization for the next, and decreasing the random walk variance at each iteration.
+There exist several algorithms, both frequentist and Bayesian, that can be used to obtain estimates of the parameters in stochastic dynamic models.
+% including the EM algorithm, Kalman Filter (and extensions) \citep{evensen09}, and iterated filtering algorithms \citep{ionides15}.
+In order to retain the ability to propose models that are scientifically meaningful rather than only those that are simply statistically convenient, we restrict ourselves to parameter estimation techniques that have the plug-and-play property, which is that the fitting procedure only requires the ability to simulate the latent process instead of evaluating transition densities \citep{breto09,he10}.
+Plug-and-play algorithms include ABC and PMCMC \citep{toni09,andrieu10}, but here we focus frequentist methods to maximize model likelihoods.
+To our knowledge, the only plug-and-play frequentist methods that can maximize the complete model likelihood are iterated filtering algorithms, which modify the well-known particle filter \citep{arulampalam02} by performing a random walk for each parameter and particle.
+These perturbations are carried out iteratively over multiple filtering operations, using the collection of parameters from the previous filtering pass as the parameter initialization for the next iteration, and decreasing the random walk variance at each step.
+
+The ability to obtain MLEs for each of the model parameters does not immediately imply that the model itself
 
 The use of likelihood-based inference allows for the statistical testing of potential improvements to the model.
-% Here we need to discuss the inclusion of the white noise in the force of infection:
 % TODO: We should address somewhere why we included the process noise in Model 1. Anna suggested this below:
@@
 Model~1 is implemented using the \code{pomp} package \cite{pomp2}, which contains an implementation of the IF2 algorithm that was used to compute the likelihood profile.
-% \arc{It may be worth mentioning that the original Model 1 team also implemented their model with pomp.
-This may be splitting hairs, but this phrasing makes me think we did the entire implementation ourselves which is not true.
-Maybe just adding in something like: "Following the original methods of the authors"?}
-% \jwc{Thanks for this.
-I agree that the original wording made it sound like we did the entire thing ourselves.
-To avoid this, and to not increase the length of this statement, I changed it to "Model~1 is implement...", so that we don't imply who did the implementation, but focusing instead on how it is implemented}
 % Performing a profile confidence search of a parameter also results in a MLE calculation, as the value that maximizes the profile likelihood also maximizes the entire likelihood surface \citep{murphy2000}.
@@
 gc()
+@
+
+\subsubsection{Model~2}
+
+Model~2 is a deterministic model, which can be viewed as a special case of a POMP (or SpatPOMP) with no randomness in the dynamic model.
+By combining a deterministic process model with a simple Gaussian measurement model, Model~2 reduces model fitting to a least squares calculation over parameters in a set of differential equations.
+Deterministic compartment models have a long history in the field of infectious disease epidemiology \citep{kermack1927,brauer2017,varghese21},
+and can be justified by asymptotic considerations in a large-population limit \citep{dadlani2020,ndii17}.
+
+\citet{lee20} originally fit two versions of model~2 based on a presupposed change in cholera transmission from a epidemic phase to endemic phase that occurred in March, 2014.
+We follow their decision to re-estimate model parameters at this break-point, but we include a requirement that the latent state $\bm{X}^{(2)}(t)$ at the start of the endemic period must be the same as the state at the end of the epidemic period.
+This additional constraint is sensible for a mechanistic interpretation of the latent state, though it adds difficulty to the task of obtaining a model fit that closely resembles the observed data.
+To combat this added difficulty, and to allow for a possible shift in the seasonality component in the force of infection, we introduce an additional phase parameter $\phaseParm$ in Eq.~\ref{model2:lambda}.
+% \begin{equation}
+% \lambda_{u}(t) = 0.5\big(1+\seasAmplitude \cos(2\pi t + \phaseParm)\big)
+% \frac{\beta_W\, W_u(t)}{ \Wsat  + W_u(t)} +
+% \beta \left\{\sum_{d=0}^4 I_{ud}(t) + \asymptomRelativeInfect \sum_{d=0}^4 A_{ud}(t) \right\}\label{eq:model2:phase}
+% \end{equation}
+% \eic{WE COULD SAVE AN EQUATION BY WRITING THE MORE GENERAL ONE INITIALLY AND SAYING THAT THEY ASSUMED $\phaseParm=0$}
+
+% We also note that, without the inclusion of a phase parameter, there is an implicit assumption is that force of infection is greatest around January and weakest around June, and the addition of this parameter allows for the testing of this assumption.
+We further increase model flexibility by fitting the parameter $\Wshed$ rather than treating it as fixed.
+We implemented this model using the \code{spatPomp} \code{R} package \citep{asfaw21github}.
+The model was then fit using the subplex algorithm, implemented in the  \code{subplex} package \citep{king2020Subplex}.
+A comparison of the trajectory of the fitted model to the data is given in Fig~\ref{fig:mod2Traj}.
+
+<<Model 2 Trajectory, echo=FALSE>>=
+h2_epi_traj <- trajectory(h2_epi, params = h2_fit$epi_params, format = 'data.frame')
+h2_epi_traj$Ctotal <- rowSums(h2_epi_traj[, paste0("C", 1:10)]) * h2_fit$epi_params['Rho']
+
+h2_end_traj <- trajectory(h2_end, params = h2_fit$end_params, format = 'data.frame')
+h2_end_traj$Ctotal <- rowSums(h2_end_traj[, paste0("C", 1:10)]) * h2_fit$end_params['Rho']
+
+h2_traj <- rbind(
+  h2_epi_traj[, c('year', 'Ctotal')],
+  h2_end_traj[, c('year', 'Ctotal')]
+) %>%
+  mutate(
+    date = yearsToDate(year)
+  )
 @
@@
 
-\jwc{How about this edit?}
-Researchers may use fitted models to make inference and draw conclusions about the dynamic system under investigation.
-Because of the casual nature of mechanistic models, the validity of these conclusions is dependent on the ability of the proposed model to approximate the system in question.
-It is often impossible to know, however, the extent to which a proposed model provides a faithful representation of the system.
+\jwc{NEW PARAGRAPH/EDIT: }
+The purpose of parameter calibration, irrespective of the approach taken (e.g., frequentist versus Bayesian), is to ensure that the proposed model provides the best possible description of the observed data under the assumptions of the model, as measured by some metric such as model likelihoods.
+Obtaining the best set of parameters for a given model does not, however, guarantee that the model provides an approximate representation of the system in question.
+Model mispecification, which may be thought of as the omission (or inclusion) of a mechanism in the model that is an important (or distracting) feature of the dynamic system, is inevitable at all levels of model complexity.
+In fact, we echo the sentiment of George Box, to whom the aphorism ``all models are wrong, but some are useful" is generally attributed; here, we assume that that some degree of model mispecification is always present, and therefore a researcher must subjectively determine whether or not the model mispecification is severe enough to render the model ``useless".
+
+% Researchers may use fitted models to make inference and draw conclusions about the dynamic system under investigation.
+% Because of the casual nature of mechanistic models, the validity of conclusions drawn using a fitted model is dependent on the ability of the proposed model to approximate the system in question.
+It is often impossible to know, however, the extent to which a model provides a faithful representation of the system.
 In such cases, we rely on the quantitative and statistical ability of the model to match the observed data, which we call the model's goodness-of-fit.
-Goodness-of-fit provides evidence supporting the causal interpretation of a model, but cannot by itself rule out the possibility of alternative explanations.
-
-% In the absence of precise scientific knowledge on a dynamic system, a model's goodness-of-fit is often the best tool for determining the model's ability to faithfully represent the dynamic system.
-% we provide a detailed discussion on assessing a model's goodness-of-fit in this subsection.
+Goodness-of-fit may provide evidence supporting the causal interpretation of one model versus another, but cannot by itself rule out the possibility of alternative explanations.
 
@@
 Likelihoods of Models~1--3 and their respective ARMA(2, 1) benchmark models are provided in Table~\ref{tab:likes}.
-\jwc{WE SAY THAT WE HAVE AN IN-DEPTH DESCRIPTION OFMODEL VALIDATION, BUT ONLY PROPOSE/DEMONSTRATE TWO WAYS TO DO THIS.
-SHOULD WE MENTION OTHER THINGS, LIKE CONDITIONAL LIKELIHOODS / ESS, CHECKING PLAUSABILITY OF FIT MODEL PARAMETERS, OR COMPARING AUTO-CORRELATIONS BETWEEN SIMULATIONS AND OBSERVATIONS (LIKE AARON KING HAS DONE)?
-I DON'T THINK WE NECESSARILY HAVE TO SHOW ALL OF THESE, SINCE THE ARTICLE IS ALREADY TOO LONG, BUT MAYBE MENTIONING THEM?
-WOULD MENTIONING THESE WITHOUT DEMONSTRATING THEM RAISE CONCERNS FROM REVIEWERS?}
 
@@
 \end{table}
+
+\jwc{NEW PARAGRAPH: }
+Similar to comparing likelihoods across models, an additional powerful diagnosis tool is the comparison of conditional likelihoods.
+Conditional likelihoods, defined as the density $f_{Y_k | Y_1, \ldots, Y_{k - 1}}\big(Y_k = y_k^* | y_{1:k-1}^*\big)$, provides a basic description of how well the proposed model can describe each data point, given the previous observations.
+Comparing these results across models---including benchmark models--- can help researchers identify potential deficiencies in the model or errors in the observed data.
+For diagnosing a single model, additional tools for assessing the goodness-of-fit include plotting the effective sample size of each observation \jwc{REF???}, or by comparing any statistic of the observed data to simulations from the model, which is sometimes referred to as diagnostic probes (for example, the autocorrelation function (ACF), as was done in \cite{king15}).
 
@@
 Instead, we use the projections from point estimates to highlight a major deficiency of deterministic models, which is that the only variability in model projections is a result of parameter uncertainty, which leads to over-confidence in the projections.
-This observation is consistent with those made in \citet{king15}, and suggests that stochastic models should be preferred over deterministic models.
-\eic{TO THE CREDIT OF MODEL 2, ITS PREDICTIONS OF POLICY IMPLICATIONS WERE AS GOOD AS, OR BETTER THAN, THE OTHER MODELS.
-PERHAPS THE WEAKNESSES OF DETERMINISTIC MODELS ONLY SHOW UP AS THE DATA ANALYSIS BECOMES MORE REFINED - FOR EXAMPLE, DETERMINISTIC MODELS MAY HAVE SUCH LOW STATISTICAL FIT THAT WE CONCLUDE THEY ARE NOT RELIABLE AS ASSOCIATIVE MODELS, IN WHICH CASE THEY CANNOT BE EXPECTED TO PROVIDE QUANTITATIVELY ACCURATE PREDICTIONS.}
+This observation is consistent with those made in \cite{king15}, and suggests that stochastic models should be preferred over deterministic models.
+
+\jwc{Replaced Ed's comment with this: }
+We note---the credit of deterministic models---that of the four fitted models described in \cite{lee20}, Model~2 provides what seems to be the most accurate model projections.
+This perhaps demonstrates that while deterministic models describe the systems in a less realistic and useful way, the relative ease in fitting these models potentially results in fewer modeling errors.
+As the data analysis becomes more refined, however, the deficiencies of deterministic models become increasingly apparent.
 
@@
 (It may also be worth mentioning that our simulations were done with a single model using one set of parameter values)}
-\jwc{Anna: I added some discussion on this in the supplement under Model~1 replication. Does this answer your concern? (see original comment in Rnw file)}
+% \jwc{Anna: I added some discussion on this in the supplement under Model~1 replication. Does this answer your concern? (see original comment in Rnw file)}
 
@@
 Outcomes of their study include estimates for the probability of cholera elimination and cumulative number of cholera infections under several possible vaccination scenarios.
-Mimicking their efforts, we define cholera elimination as having less than one infection of cholera over at least 52 consecutive weeks, and provide projections under the following vaccination scenarios:
-\arc{Rewrite definition: cholera elimination defined as less than one cholera infection for at least fifty two consecutive weeks in the ten years after the vaccinations?}
-\jwc{ Good catch, TODO: is it necessary to say that it stayed away? Does it make a difference?}
+Mimicking their efforts, we define cholera elimination as having less than one infection of cholera over at least 52 consecutive weeks in the 10-year projection period, and provide projections under the following vaccination scenarios:
 

### 2022-08-23 ms.Rnw
Made some additions to the article

@@
 This necessitates addressing the tradeoff between biological fidelity and model simplicity, and the reality of misspecification for models at all levels of complexity.
-\eic{EDIT IN THE DIRECTION OF THE PREVIOUS SENTENCE.}
+% \eic{EDIT IN THE DIRECTION OF THE PREVIOUS SENTENCE.}
 As a case study, we consider a cholera epidemic in Haiti.
@@
 \jwc{NEW PARAGRAPH/EDIT: }
-The purpose of parameter calibration, irrespective of the approach taken (e.g., frequentist versus Bayesian), is to ensure that the proposed model provides the best possible description of the observed data under the assumptions of the model, as measured by some metric such as model likelihoods.
+The purpose of parameter calibration, irrespective of the approach taken (e.g., frequentist versus Bayesian), is to ensure that the proposed model provides the best possible description of the observed data under the assumptions of the model, as measured by some metric (such as model likelihoods).
 Obtaining the best set of parameters for a given model does not, however, guarantee that the model provides an approximate representation of the system in question.
 Model mispecification, which may be thought of as the omission (or inclusion) of a mechanism in the model that is an important (or distracting) feature of the dynamic system, is inevitable at all levels of model complexity.
-In fact, we echo the sentiment of George Box, to whom the aphorism ``all models are wrong, but some are useful" is generally attributed; here, we assume that that some degree of model mispecification is always present, and therefore a researcher must subjectively determine whether or not the model mispecification is severe enough to render the model ``useless".
+In fact, we echo the sentiment of George Box, to whom the aphorism ``all models are wrong, but some are useful" is often attributed; here, we assume that that some degree of model mispecification is always present, and therefore a researcher must subjectively determine whether or not the model mispecification is severe enough to render the model ``useless".
 
@@
 \jwc{NEW PARAGRAPH: }
-Similar to comparing likelihoods across models, an additional powerful diagnosis tool is the comparison of conditional likelihoods.
-Conditional likelihoods, defined as the density $f_{Y_k | Y_1, \ldots, Y_{k - 1}}\big(Y_k = y_k^* | y_{1:k-1}^*\big)$, provides a basic description of how well the proposed model can describe each data point, given the previous observations.
-Comparing these results across models---including benchmark models--- can help researchers identify potential deficiencies in the model or errors in the observed data.
-For diagnosing a single model, additional tools for assessing the goodness-of-fit include plotting the effective sample size of each observation \jwc{REF???}, or by comparing any statistic of the observed data to simulations from the model, which is sometimes referred to as diagnostic probes (for example, the autocorrelation function (ACF), as was done in \cite{king15}).
-
-\subsection{Model Projections}\label{sec:filter}
+Similar to comparing log-likelihoods across models, an additional powerful diagnosis tool is the comparison of conditional log-likelihoods.
+Conditional likelihoods, defined as the density $f_{Y_k | Y_1, \ldots, Y_{k - 1}}\big(Y_k = y_k^* | y_{1:k-1}^*\big)$, provide a basic description of how well the proposed model can describe each data point, given the previous observations.
+Comparing these results across models---including benchmark models--- can help researchers identify potential model deficiencies, or errors in the observed data.
+Additional tools for assessing the goodness-of-fit of a model include plotting the effective sample size of each observation \jwc{REF???} and comparing any statistic of the observed data to simulations from the model, which is sometimes referred to as diagnostic probes (for example, the autocorrelation function (ACF), as was done in \cite{king15}).
+
+\subsection{Forecasts}\label{sec:filter}
 
@@
 \footnotetext[3]{Model 3 was originally fit to only a subset of the data starting from March 2014 and did not include a large portion of data from Ouest in 2015-2016.
-On this subset, the parameters provided by \citet{lee20} achieved a likelihood of $\Sexpr{myround(max(lee3_results$sub_ll_evals), 1)}$.
+On this subset, the parameters provided by \cite{lee20} achieved a likelihood of $\Sexpr{myround(max(lee3_results$sub_ll_evals), 1)}$.
 On this same subset of data, our model achieved a likelihood of $\Sexpr{myround(mod3_spatPomp_res$subset_ll, 1)}$.}
-As previously noted, simulations from a well-fit mechanistic model may closely resemble the observed data $y_{1:N}^*$.
-In such comparisons, these simulations are random draws from the complete estimated joint distribution $f_{\bm{X}^{(\modelCounter)}_{0:N}, \bm{Y}^{(\modelCounter)}_{1:N}}\left(x_{0:N}, y_{1:N}\mid \hat{\theta}\right)$, where $\modelCounter$ identifies the model used for simulations.
+Forecasting is a powerful tool that has primarily been used to estimate the future state of a system; when a mechanistic model is used, forecasts may provide estimates of the potential effects interventions may have on a system.
+Forecasts can also be used as a way of testing scientific hypothesis in real time by comparing simulations from models with competing descriptions of a system to newly obtained data \citep{lewis22}.
+In order to have any benefit to the researcher, however, forecasts must be obtained with great care.
+In particular, researchers should be careful to account for various forms of uncertainty present in model forecasts, carefully calibrate the proposed model to observed data, and---to the best of their ability---account for the most recently known state of the system.
+
+As an example, we note that simulations from a well-fit mechanistic model may closely resemble the observed data $y_{1:N}^*$.
+In such comparisons, these simulations are random draws from the complete estimated joint distribution $f_{\bm{X}^{(\modelCounter)}_{0:N}, \bm{Y}^{(\modelCounter)}_{1:N}}\left(x_{0:N}, y_{1:N}\mid \hat{\theta}\right)$, where $\modelCounter$ indexes the model used for simulations.
 It can therefore be tempting to use simulations from this model up to time $N+s$, $f_{\bm{X}^{(\modelCounter)}_{0:N+s}, \bm{Y}^{(\modelCounter)}_{1:N+s}}\left(x_{0:N+s}, y_{1:N+s}\mid \hat{\theta}\right)$, with $s\geq 0$ to project the dynamic system up to a future time $N+s$, as it has been done in previous studies \citep{lee20,???}.
@@
 Note that each of the models in question (and, more generally, all models that are described as POMPs) are Markovian, that is, the history of the process $\{\bm{X}^{(\modelCounter)}_s, s\leq t\}$ for model $\modelCounter$ is uninformative about the future of the process $\{\bm{X}^{(\modelCounter)}_s, s\geq t\}$, given the current state $\bm{X}^{(\modelCounter)}_t$.
-In other words, observing the data $\bm{Y}^{(\modelCounter)}_N = y_N^*$ at time $N$ provides more information about the future state of the system than does the initial conditions.
-A more sound projection of the dynamic system into the future could therefore be obtained by simulating using the most recently known state of the system, that is, obtaining draws from the conditional density $f_{\bm{X}^{(\modelCounter)}_{N:s}, \bm{Y}^{\modelCounter}_{N:s}}\big(x_{N:s}, y_{N:s}\mid \hat{\theta}, \bm{X}^{(\modelCounter)}_{N} = \bm{x}^{(\modelCounter)}_N\big)$.
+In other words, observing the data $\bm{Y}^{(\modelCounter)}_N = y_N^*$ at time $N$ provides more information about the future state of the system than the initial conditions.
+In this case, draws from the conditional density $f_{\bm{X}^{(\modelCounter)}_{N:s}, \bm{Y}^{\modelCounter}_{N:s}}\big(x_{N:s}, y_{N:s}\mid \hat{\theta}, \bm{X}^{(\modelCounter)}_{N} = \bm{x}^{(\modelCounter)}_N\big)$ should be preferred, as these simulations account for the most recent known state of the system.
 
 $\bm{X}^{(\modelCounter)}_N$, however, is unobservable and therefore draws from the desired conditional density are unobtainable.
-Possible values of $\bm{X}^{(\modelCounter)}_N$ given the observed data $\bm{Y}^{(\modelCounter)}_{1:N} = y_{1:N}^*$ and the proposed model can easily be obtained, however, via the filtering distribution.
+Informed estimates of $\bm{X}^{(\modelCounter)}_N$ given the observed data $\bm{Y}^{(\modelCounter)}_{1:N} = y_{1:N}^*$ and the proposed model can easily be obtained, however, via the filtering distribution.
 Let $\hat{\bm{X}}^{(\modelCounter), i}_N, \, i \in {1, 2, \ldots, J}$ be $\iid$ draws from the filtering distribution at time $N$, with density $\hat{\bm{X}}^{(\modelCounter), i}_N \sim f_{\bm{X}^{(\modelCounter)}_N|\bm{Y}^{(\modelCounter)}_{1:N}}(x_N \mid \hat{\theta}, y_{1:N}^*)$.
-A single model projection can then be obtained by simulating from the model $f_{\bm{X}^{(\modelCounter)}_{N:s}, \bm{Y}^{(\modelCounter)}_{N:s}}\big(x_{N:s}, y_{N:s}\mid \hat{\theta}, \bm{X}^{(\modelCounter)}_{N} = \hat{\bm{X}}^{(\modelCounter), i}_N\big)$.
+A single model forecast can then be obtained by simulating from the model $f_{\bm{X}^{(\modelCounter)}_{N:s}, \bm{Y}^{(\modelCounter)}_{N:s}}\big(x_{N:s}, y_{N:s}\mid \hat{\theta}, \bm{X}^{(\modelCounter)}_{N} = \hat{\bm{X}}^{(\modelCounter), i}_N\big)$.
 Intuitively, simulating the model starting at the filtering distribution of the most recently available time point is a more appropriate way to project a stochastic dynamic system into the future, as it is not expected that each simulation from initial conditions will result in a latent state at time $N$ consistent with the model and the observation $\bm{Y}^{(\modelCounter)}_N = y_N^*$.
@@
 
-An additional consideration to make when obtaining model forecasts is that of parameter uncertainty.
+Another consideration to make when obtaining model forecasts is that of parameter uncertainty.
 It has been noted that the uncertainty in just a single parameter can lead to drastically different projections \citep{saltelli20}.
@@
 \jwc{Replaced Ed's comment with this: }
-We note---the credit of deterministic models---that of the four fitted models described in \cite{lee20}, Model~2 provides what seems to be the most accurate model projections.
+We note---the credit of deterministic models---that of the four fitted models in \cite{lee20}, Model~2 provides what seems to be the most accurate model projections.
 This perhaps demonstrates that while deterministic models describe the systems in a less realistic and useful way, the relative ease in fitting these models potentially results in fewer modeling errors.
@@
 
-% \arc{Perhaps here (or maybe in the discussion on likelihood-based techniques...or even somewhere else?) we can discuss the utility of point estimates with respect to Model 1.
-The fitting for Model 1 was done by visual convergence of the parameters, but the plotted simulations were taken from ~100 different sets of parameters values (unsure of exact number because it was not reported).
-One could argue that this visual convergence is misleading because there is not one set of values — is unclear to me what exactly is actually "converging".
-Of course, parameter uncertainty is an important topic, and we do discuss it above.
-But, with respect to interpretability and prediction-informed policy making, a single parameter value has clear benefits.
-(It may also be worth mentioning that our simulations were done with a single model using one set of parameter values)}
-% \jwc{Anna: I added some discussion on this in the supplement under Model~1 replication. Does this answer your concern? (see original comment in Rnw file)}
-
 \section{Results}\label{sec:results}
 
-\eic{HERE'S AN INITIAL ATTEMPT TO EXPLAIN MY POINT BELOW FOR THE MS. IT IS HARD TO MAKE CONCRETE POINTS ON THIS SORT OF TOPIC.
+\eic{HERE'S AN INITIAL ATTEMPT TO EXPLAIN MY POINT BELOW FOR THE MS. IT IS HARD TO MAKE CONCRETE POINTS ON THIS SORT OF TOPIC.}
 A model which aspires to provide quantitative guidance for assessing interventions should provide a quantitative statistical fit for previous data.
@@
 A causal interpretation is strengthened by corroborative evidence.
-For example, reconstructed latent variables (such as numbers of susceptible and recovered individuals) should make sense in the context of alternative measurments of these variables;
+For example, reconstructed latent variables (such as numbers of susceptible and recovered individuals) should make sense in the context of alternative measurements of these variables;
 parameter values which fit the data should make sense in the context of alternative lines of evidence about the phenomena being modeled.
-The model is necessarily a simplified approximation of a complex process, and so we cannot expect our models (describing and fitted to population-level processes) to perfectly match quantitative understanding of individual-level processes.
-Such discrepancies could lead to biases when interventions (modeled as effects on individuals) are included in the population-level, unless there are historical occurrences of the intervation which enable the consequence to be calibrated at the population level.
-IF WE WANT TO HEAD IN THIS DIRECTION, WE COULD EXPLAIN HOW THESE PRINCIPLES PLAY OUT IN OUR DATA ANALYSIS.}
-
-\eic{TO AVOID BEING TOO NEGATIVE, WE ALSO COULD EXPLAIN THE LOGIC OF HOW POSITIVE RESULTS CAN LEGITIMATELY BE CONCLUDED.
-If a mechanistic model including a feature (such as a representation of a mechanism, or the inclusion of a covariate) fits better than mechanistic models without that feature, and also has competititive fit compared to associative models, this may be taken as evidence supporting the scientific relevance of the feature.
+\jwc{Perhaps add comment on our final parameter estimates, and remove the following sentences in this paragraph.}
+A model is necessarily a simplified approximation of a complex process, and so it is unreasonable to expect models (describing and fitted to population-level processes) to perfectly match quantitative understanding of individual-level processes.
+Such discrepancies could lead to biases when interventions (modeled as effects on individuals) are included in the population-level, unless there are historical occurrences of the intervention which enable the consequence to be calibrated at the population level.
+
+If a mechanistic model including a feature (such as a representation of a mechanism, or the inclusion of a covariate) fits better than mechanistic models without that feature, and also has competitive fit compared to associative models, this may be taken as evidence supporting the scientific relevance of the feature.
 As for any analysis of observational data, we must be alert to the possibility of confounding.
@@
 For a mechanism, the model feature could in principle explain the data by helping to account for some different unmodeled phenomenon.
-Assessing potential confounding is part of building an argument for a causal interpretation of a fitted model.
+% Assessing potential confounding is part of building an argument for a causal interpretation of a fitted model.
 In the context of our analysis, the estimated trend in transmission rate could be explained by any trending variable (such as hygiene improvements, or changes in population behavior), resulting in confounding from colinear covariates.
 Alternatively, the trend could be attributed to a decreasing reporting rate rather than decreasing transmission rate, resulting in confounded mechanisms.
-The robust statistical conclusion is that a model which allows for change fits better than one which does not---we argue that a decreasing transmission rate is a plausible way to explain this, but the incidence data themselves may not have enough information to pin down the mechanism.
-}
-
-Because mechanistic models provide a scientific description of the dynamic system, a properly fit model allows a researcher to perform inference on the system in question.
-\eic{FIT IS A MEASURE OF ASSOCIATION, AND ASSOCIATION DOES NOT NECESSARILY IMPLY CAUSATION.
-SO WE HAVE TO WORD THIS CAREFULLY.
-IT CAN BE A RATHER DELICATE QUESTION TO ESTABLISH THE PROPER CAUSAL INTERPRETATION OF A FITTED MODEL, AND THIS MAY NOT MATCH ONE'S INITIAL GUESS (E.G., THE MEASLES CASE STUDY IN SISMID).
-IT WOULD BE INTERESTING TO DISCUSS IN PERSON HOW TO THREAD THIS NEEDLE.}
-Examples are as diverse as estimating the effective reproductive number ($R_0$) of an infectious disease \citep{he10} or investigating interactions between pedestrians and autonomous vehicles \citep{domeyer22}.
-Here, we demonstrate this ability by investigating the seasonal transmission trends in Model~1.
-Model~1 includes a flexible cubic spline term in Eq.~\myeqref{model1:lambda}--\myeqref{model1:beta} that allows for seasonality in the force of infection.
-This flexibility allows the model to learn any seasonal trends in cholera infections that may be present in the data.
+The robust statistical conclusion is that a model which allows for change fits better than one which does not---we argue that a decreasing transmission rate is a plausible way to explain this, but the incidence data themselves do not provide enough information to pin down the mechanism.
+
+In a similar fashion, one can take advantage of certain mechanistic features contained in a particular model in order to make inference on a system.
+Examples of this are as diverse as estimating the effective reproductive number ($R_0$) of an infectious disease \citep{he10} or investigating interactions between pedestrians and autonomous vehicles \citep{domeyer22}.
+In our analysis, we demonstrate this ability by examining the results of fitting the flexible cubic spline term in Model~1 (Eq.~\myeqref{model1:lambda}--\myeqref{model1:beta}), which allows for a flexible estimation of seasonality in the force of infection.
 After fitting the model, we explore potential patterns in the seasonal transmission rate by plotting the average value of $\transmission$ in a typical year.
 Fig.~\ref{fig:h1SeasRain} shows that the estimated seasonal transmission rate $\transmission$ mimics the rainfall dynamics in Haiti, despite Model~1 not having access to rainfall data.
-This result provides evidence that rainfall is an important driver of cholera infections in Haiti.
+This result provides evidence that rainfall is a potential driver of cholera infections in Haiti.
 
@@
 
-An additional benefit of mechanistic modelling is that they permit the simulation of potential interventions on a system, as was the primary goal of \citet{lee20}.
+While such inference is useful, one must be careful not to assume the results of a single analysis as a de facto feature of the system.
+It is important instead to recognize and assess the modeling simplifications and assumptions that were used in order to arrive at the conclusions.
+
+An additional benefit of mechanistic modelling is ability to simulate various interventions on a system; this feature is useful to inform policy and was the primary goal of \citet{lee20}.
 Outcomes of their study include estimates for the probability of cholera elimination and cumulative number of cholera infections under several possible vaccination scenarios.
@@
 
-Various  have suggestions been made about why \citet{lee20} failed to accurately predict the eventual eradication of cholera from Haiti.
-Proposals include model misspecification, overly difficult elimination criteria, and a potential conflict of interest \citep{rebaudetComment20,henrys20}.
+Various suggestions been made about why \citet{lee20} failed to accurately predict the eventual eradication of cholera from Haiti, including model misspecification, overly difficult elimination criteria, and a potential conflict of interest \citep{rebaudetComment20,henrys20}.
 Here, we instead argue that careful attention to important statistical details could have correctly resulted in the conclusion of imminent cholera elimination.
-We acknowledge that we have the benefit of hindsight: our demonstration of a statistically principled route to obtain better-fitting models with more accurate predictions does not rule out the possibility of discovering other models that fit well yet predict poorly.
+We acknowledge the benefit of hindsight: our demonstration of a statistically principled route to obtain better-fitting models with more accurate predictions does not rule out the possibility of discovering other models that fit well yet predict poorly.
 

### 2022-08-26 ms.Rnw
Added/removed paragraph, and other minor edits

@@
 \newcommand\code[1]{\texttt{#1}}
+\newcommand\paramVec{\theta}
 
@@
 As a case study, we consider a cholera epidemic in Haiti.
-The 2010 introduction of cholera to Haiti led to an extensive outbreak and sustained transmission, eliminated in 2019 with the help of vaccination and other public health measures.
+The 2010 introduction of cholera to Haiti led to an extensive outbreak and sustained transmission until it was eliminated in 2019.
+% with the help of vaccination and other public health measures.
+% \jwc{Kind of the point here is that it went away without vaccination efforts.}
 We study three models developed by expert teams to advise on vaccination policies.
@@
 
-The four independent teams were given the task of projecting the
+The four independent teams were given the task of estimating the
 potential effect of increased oral cholera vaccine (OCV) efforts across Haiti.
@@
 This conclusion is inconsistent with the fact that there have been no confirmed cases since February, 2019 without additional vaccination efforts \citep{ferguson22}.
+
 The failure of \cite{lee20} to correctly predict the elimination of cholera has been the topic of some discussion \citep{francois20,rebaudetComment20,henrys20,leeReply20}.
-
-Here, we demonstrate that attention to certain statistical considerations could have improved the fit to previous data, as well as leading to forecasts that match more closely the observed outcome.
+We demonstrate that careful attention in model fitting, diagnosing, and forecasting improves each of the proposed model's ability to quantitatively describe observed data and results in forecasts that are more consistent with the observed outcome.
 \jwc{NEXT TWO SENTENCES ARE NEW: }
-Because our primary focus is on the statistical tasks of fitting and diagnosing mechanistic models, we refrain from commenting on the feasibility of scientific assumptions implicitly made in \cite{lee20}.
-We demonstrate that model-mispecification does not alone explain the discrepancy between model predictions and reality, as has been suggested \citep{rebaudetComment20}.
+Because we focus on the statistical tasks of model fitting, diagnosing, and forecasting, we refrain from commenting on the feasibility of scientific assumptions implicitly made in the models under consideration.
+Our results demonstrate that model-mispecification does not alone explain the discrepancy between model predictions and reality, as has been suggested \citep{rebaudetComment20}.
 Based on this retrospective analysis, we offer suggestions on fitting mechanistic models to dynamic systems for future studies.
-We proceed by introducing Models 1--3 in Sec.~\ref{sec:models};
+
+We proceed by introducing Models~1--3 in Sec.~\ref{sec:models};
 in Sec.~\ref{sec:methods}, we present a methodological approach to examining and refining these models.
@@
 Each model describes the cholera dynamics using a latent state vector $\bm{X}^{(\modelCounter)}(t)$ and an observable response vector $\bm{Y}^{(\modelCounter)}(t)$ for each continuous time-point $t \in \mathcal{T}$, where $\mathcal{T}$ is the set of all time-points and $\modelCounter \in \{1, 2, 3\}$ indexes the model.
-The latent state vector $\bm{X}^{(\modelCounter)}$ consists of individuals labeled as susceptible (S), infected (I), asymptomatically infected (A), vaccinated (V), and recovered (R) for $\modelCounter \in \{1, 2, 3\}$, with various sub-divisions sometimes considered across the three models under consideration.
-Models~2 and~3 have metapopulation structure, meaning that each individual is a member of a spatial unit, denoted by $u\in \seq{1}{U}$, which corresponds here to membership of the $U=10$ administrative departments.
+The latent state vector $\bm{X}^{(\modelCounter)}$ consists of individuals labeled as susceptible (S), infected (I), asymptomatically infected (A), vaccinated (V), and recovered (R), with various sub-divisions sometimes considered in each model for $\modelCounter \in \{1, 2, 3\}$.
+Models~2 and~3 have metapopulation structure, meaning that each individual is a member of a spatial unit, denoted by $u\in \seq{1}{U}$, which corresponds to membership of the $U=10$ administrative Haitian departments (anglicized version of departements).
 % Model~1 describes the dynamics aggregated at the national level.
 
-In this study, we focus our attention on important statistical, rather than scientific, details that may improve inference on a dynamic system;
-we therefore refrain from commenting on the assumptions in each model or proposing our own alternative description of cholera dynamics, and instead rely on the scientific expertise of \cite{lee20}.
+% \jwc{Already stated above}
+% In this study, we focus our attention on important statistical, rather than scientific, details that may improve inference on a dynamic system;
+% we therefore refrain from commenting on the assumptions in each model or proposing our own alternative description of cholera dynamics, and instead rely on the scientific expertise of \cite{lee20}.
+\jwc{NEW PARAGRAPH:}
+In the following subsections, detailed descriptions of Models~1--3 are provided.
+While these descriptions are a necessary inclusion for transparency and reproducibility, they do not immediately contribute to our primary goal of model fitting, diagnosing, and forecasting, with the purpose of informing policy.
+A first-time reader may therefore benefit from quickly skimming through Sec.~\ref{sec:methods} and returning to the section later, as needed.
 
@@
 As with Model~1, $d\Gamma_u(t)/dt$ is multiplicative Gamma-distributed white noise in \myeqref{eq:model3:SI} and \myeqref{eq:model3:SA}.
-In \eqref{eq:model3:water}, $J_u(t)$ is a unit-less measurement of precipitation that has been standardized by dividing the observed rainfall at time $t$ by the maximum recorded rainfall in departement $u$ during the epidemic, and $D_u$ is the average population density.
+In \eqref{eq:model3:water}, $J_u(t)$ is a unit-less measurement of precipitation that has been standardized by dividing the observed rainfall at time $t$ by the maximum recorded rainfall in department $u$ during the epidemic, and $D_u$ is the average population density.
 Demographic stochasticity is accounted for by modeling non-cholera related death rate $\muDeath$ in each compartment, along with an additional death rate $\choleraDeath$ in \myeqref{eq:model3:IS} to account for cholera induced deaths among infected individuals.
@@
 
-Many dynamic systems of scientific and public interest are highly complex by nature; therefore when developing a mathematical description of a dynamic system, it is often necessary to impose assumptions on the system in order to make inference possible.
+\jwc{NEW PARAGRAPH}
+% Mechanistic modeling is an attempt to account for known or hypothesized features of a dynamic system.
+Proposed mechanistic structures form a family of statistical models that can be indexed by a parameter vector $\paramVec$.
+Because different values of $\paramVec$ result in drastically different modeling outcomes, researchers must be careful when choosing $\paramVec$.
+Elements of $\paramVec$ are either chosen as constants that reflect our understanding of the system, or calibrated to observed data, with the intent of maximizing a proposed model's ability to quantitatively describe observed data according to some measure.
+While this section is focused on the elements of $\paramVec$ that are calibrated to data, we note that the decision of which elements to fix and which to estimate has consequences for model interpretability, as briefly discussed in Sec.~\ref{sec:results}
+
+The complex nature of the dynamic systems which are of scientific and public interest necessitates the imposition of various modeling assumptions in order to facilitate inference.
+% Many dynamic systems of scientific and public interest are highly complex by nature; therefore when developing a mathematical description of a dynamic system, it is often necessary to impose assumptions on the system in order to make inference possible.
 For example, one common assumption in epidemiology is a homogeneous mixing population \citep{bansal07}.
 Other common considerations include whether the proposed model should be stochastic or deterministic; whether the model should have change points in parameter values or should otherwise make adjustments for changes in the dynamic system; and whether the proposed model should include any spatial heterogeneity at a scale permissible by the observed data.
-Ultimately, these decisions will result in a different approach to calibrating (or fitting) model parameters.
-
-The models considered in this study represent a diverse selection of possible modelling decisions.
-Each of the models independently describe the cholera dynamics as a Partially Observed Markov Process (POMP), since each model is structured in such a way that, given the partially-observable state of the system at time $n$, the future state of the system is independent of its past.
-Because of the spatial structure of Models~2 and~3, we refer to these models as SpatPOMPs \citep{asfaw21arxiv}.
-In the following subsections, we describe approaches to fitting mechanistic models and discuss potential trade-offs of the selected model complexity.
+Ultimately, these decisions may result in a different approach to calibrating model parameters.
+
+\jwc{MAJOR CHANGE TO THIS PARAGRAPH:}
+Each model considered in this study describes cholera dynamics as a Partially Observed Markov Process (POMP), as each model contains unobservable states that evolve sequentially with time.
+Despite their similarities, these models represent a diverse selection of possible modeling assumptions: namely, the use of stochastic (Models~1 and 2) or deterministic (Model~2) equations; spatially-heterogeneous meta-population (Models~2 and 3) or spatially-aggregated (Model~1) structure; and the use of covariates (Model~3) versus mathematical equations (Models~1 and 2) to describe a seasonal mechanism.
+In the following subsections we describe our approach to fitting these mechanistic models.
 
@@
 
-One approach to modelling a dynamic system is through probabilistic models.
+One approach to modeling a dynamic system is through probabilistic models.
 With a probabilistic model,
-% rather than modelling the mean behavior of the system through the use of deterministic differential equations,
+% rather than modeling the mean behavior of the system through the use of deterministic differential equations,
 we suppose the existence of a joint density $f_{\bm{X}^{(\modelCounter)}_{\seq{0}{N}}, \bm{Y}^{(\modelCounter)}_{\seq{1}{N}}}$, with $\bm{X}^{(\modelCounter)}_{\seq{0}{N}}$ denoting the unobservable Markov process of model $\modelCounter$ at times $\seq{0}{N} = \{0, 1, \ldots, N\}$, and $\bm{Y}^{(\modelCounter)}_{\seq{1}{N}}$ denoting the observable process of the system at times $\seq{1}{N}$.
-Under this framework, the observed data $y_{\seq{1}{n}}^*$ are assumed to be a single realization of the model $y_{\seq{1}{n}}^* \sim f_{\bm{X}^{(\modelCounter)}_{\seq{0}{N}}, \bm{Y}^{(\modelCounter)}_{\seq{1}{N}}}(x_{\seq{0}{N}}, y_{\seq{1}{N}}; \theta)$, where $\theta$ is a parameter vector that indexes the model.
-Using a probabilistic model results in several advantages, including the ability to estimate the variability of the system.
+Under this framework, the observed data $y_{\seq{1}{n}}^*$ are assumed to be a single realization of the model $y_{\seq{1}{n}}^* \sim f_{\bm{X}^{(\modelCounter)}_{\seq{0}{N}}, \bm{Y}^{(\modelCounter)}_{\seq{1}{N}}}(x_{\seq{0}{N}}, y_{\seq{1}{N}}; \paramVec)$, where $\paramVec$ is a parameter vector that indexes the model.
+Using a probabilistic model results in several advantages, including the ability to account for variability present in the system.
 % which is of great interest to scientists and policy makers.
@@
 
-Once a model for the system has been proposed, the parameter vector $\theta$ needs to be estimated using the observed data.
+Once a model for the system has been proposed, the parameter vector $\paramVec$ needs to be estimated using the observed data.
 There exist several algorithms, both frequentist and Bayesian, that can be used to obtain estimates of the parameters in stochastic dynamic models.
@@
 In order to retain the ability to propose models that are scientifically meaningful rather than only those that are simply statistically convenient, we restrict ourselves to parameter estimation techniques that have the plug-and-play property, which is that the fitting procedure only requires the ability to simulate the latent process instead of evaluating transition densities \citep{breto09,he10}.
-Plug-and-play algorithms include ABC and PMCMC \citep{toni09,andrieu10}, but here we focus frequentist methods to maximize model likelihoods.
+Plug-and-play algorithms include Bayesian approaches like ABC and PMCMC \citep{toni09,andrieu10}, but here we use frequentist methods to maximize model likelihoods.
 To our knowledge, the only plug-and-play frequentist methods that can maximize the complete model likelihood are iterated filtering algorithms, which modify the well-known particle filter \citep{arulampalam02} by performing a random walk for each parameter and particle.
 These perturbations are carried out iteratively over multiple filtering operations, using the collection of parameters from the previous filtering pass as the parameter initialization for the next iteration, and decreasing the random walk variance at each step.
-
-The ability to obtain MLEs for each of the model parameters does not immediately imply that the model itself
 
@@
 The resulting confidence interval for $\transmissionTrend$ is $(\Sexpr{myround(mcap_results$ci[1], digits = 3)}, \Sexpr{myround(mcap_results$ci[2], digits = 3)})$, with the full results displayed in Fig.~\ref{fig:betat}.
-These results indicate that the inclusion of a trend improves the model fit to the data.
+These results are suggestive that the inclusion of a trend in transmission rate improves the quantitative ability of Model~1 to describe the observed data.
 The reported results for Model~1 in the remainder of this article were obtained with the inclusion of the parameter $\transmissionTrend$.
@@
 
-Parameters that must be fit in Model~3 are primarily shared between each departement, the exception to this being the parameters $\beta_{W_u}$, and $\beta_u$, which are unique for each departement $u \in \seq{1}{10}$.
-To fit this model, \citet{lee20} simplified the parameter estimation problem by fitting independent departement-level models to the data.
-The shared parameters were calibrated using the cholera incidence data from Artibonite, and the departement-specific parameters ($\beta_{W_u}$ and $\beta_u$) were fit using the data from their respective departement.
+Parameters that must be fit in Model~3 are primarily shared between each department, the exception to this being the parameters $\beta_{W_u}$, and $\beta_u$, which are unique for each department $u \in \seq{1}{10}$.
+To fit this model, \citet{lee20} simplified the parameter estimation problem by fitting independent department-level models to the data.
+The shared parameters were calibrated using the cholera incidence data from Artibonite, and the department-specific parameters ($\beta_{W_u}$ and $\beta_u$) were fit using the data from their respective department.
 Reducing a spatially-heterogeneous model to individual units in this fashion requires special treatment of any interactive mechanisms between spatial units, such as found in Eq.~\myeqref{eq:model3:foi}.
-In particular, when considering a model for departement $u$, the values $I_\nu(t)$ and $A_\nu(t)$, are unknown for $u \neq v \in \seq{1}{10}$.
-A first order approximation of Eq.~\myeqref{eq:model3:foi} for each departement $v \in \seq{1}{10}$ can be obtained using the weekly number of observed cholera cases in each departement:
+In particular, when considering a model for department $u$, the values $I_\nu(t)$ and $A_\nu(t)$, are unknown for $u \neq v \in \seq{1}{10}$.
+A first order approximation of Eq.~\myeqref{eq:model3:foi} for each department $v \in \seq{1}{10}$ can be obtained using the weekly number of observed cholera cases in each department:
 \begin{equation}
@@
 The BPIF algorithm is implemented in the \code{panelPomp} package \citep{breto20panelPomp} in the R programming language, and can be used by adding the argument \code{block = TRUE} in the \code{mif2} function.
-pseudo-code for this algorithm is provided in Algorithm~S1 in the supplementary material.
+Pseudo-code for this algorithm is provided in Algorithm~S1 in the supplementary material.
 % Even with the simplification of Eq.~\myeqref{eq:model3:approx}, a significant amount of computational effort is needed to obtain a proper model fit.
-Therefore the reported parameters were obtained using the PanelPOMP version of Model~3, and simulations for the various vaccination campaigns (Sec.~\ref{sec:results}) were obtained using these same parameters in the SpatPOMP version of the model.
+The reported parameters were obtained using the PanelPOMP version of Model~3, and simulations for the various vaccination campaigns (Sec.~\ref{sec:results}) were obtained using these same parameters in the SpatPOMP version of the model.
 % Simulations from the fitted version of the PanelPOMP model are shown in Fig.~\ref{h3panelsims}.
@@
 
-Fitting Model~3 simplifies the problem of parameter estimation, but it also introduces additional technicalities that must be addressed.
-One concern is that of obtaining model projections, which was the primary goal of the original modeling exercise.
+Fitting Model~3 as a PanelPOMP simplifies the problem of parameter estimation, but it also introduces additional technicalities that must be addressed.
+One concern is that of obtaining model forecasts, which was the primary goal of \citet{lee20}.
 % "primary output" -> "primary goal", based on Anna's suggestion.
-The simplified PanelPOMP version of Model~3 relies on the observed cholera cases as a covariate, which are unavailable for use in projections.
+The simplified PanelPOMP version of Model~3 relies on the observed cholera cases as a covariate, which are unavailable for use in forecasts.
 To address this, we use the MLE obtained for the PanelPOMP approximation of Model~3 as an estimate for the parameters in fully coupled version of the model, which was implemented using the \code{spatPomp} package.
@@
 
-While simulations from the SpatPOMP version of the model seem similar to the observed data, we note in Table~\ref{tab:likes} that using the MLE obtained using the approximation in Eq.~\myeqref{eq:model3:approx} results in a lower likelihood value for the SpatPOMP version of the model than the PanelPOMP version.
-A natural question to ask is if the MLE obtained using the approximation in Eq.~\myeqref{eq:model3:approx} could result in a drastically different set of parameters than those that would be obtained by estimating the MLE in the fully coupled model.
+While simulations from the SpatPOMP version of Model~3 resemble the observed data, we note in Table~\ref{tab:likes} that the SpatPOMP version of Model~3 has a lower log-likelihood than the PanelPOMP version using the same model parameters.
+A natural question to ask is if the MLE obtained for the PanelPOMP version of Model~3---obtained using the approximation in Eq.~\myeqref{eq:model3:approx}---could result in a drastically different set of parameters than those that would be obtained by estimating the MLE in the fully coupled model.
 Unfortunately, the answer to this question would require the ability to accurately and reliably estimate the MLE of the fully coupled model, which is beyond the scope of this article.
@@
 \jwc{NEW PARAGRAPH/EDIT: }
-The purpose of parameter calibration, irrespective of the approach taken (e.g., frequentist versus Bayesian), is to ensure that the proposed model provides the best possible description of the observed data under the assumptions of the model, as measured by some metric (such as model likelihoods).
+The purpose of parameter calibration, irrespective of the approach taken, frequentist versus Bayesian), is to ensure that the proposed model provides the best possible description of the observed data under the assumptions of the model, as measured by some metric.
 Obtaining the best set of parameters for a given model does not, however, guarantee that the model provides an approximate representation of the system in question.
 Model mispecification, which may be thought of as the omission (or inclusion) of a mechanism in the model that is an important (or distracting) feature of the dynamic system, is inevitable at all levels of model complexity.
-In fact, we echo the sentiment of George Box, to whom the aphorism ``all models are wrong, but some are useful" is often attributed; here, we assume that that some degree of model mispecification is always present, and therefore a researcher must subjectively determine whether or not the model mispecification is severe enough to render the model ``useless".
-
-% Researchers may use fitted models to make inference and draw conclusions about the dynamic system under investigation.
-% Because of the casual nature of mechanistic models, the validity of conclusions drawn using a fitted model is dependent on the ability of the proposed model to approximate the system in question.
-It is often impossible to know, however, the extent to which a model provides a faithful representation of the system.
-In such cases, we rely on the quantitative and statistical ability of the model to match the observed data, which we call the model's goodness-of-fit.
+In fact, we echo the sentiment of George Box, to whom the aphorism ``all models are wrong, but some are useful" is often attributed: we assume that some degree of model mispecification is always present, and it is necessary to subjectively determine whether or not the model mispecification is severe enough to render the model ``useless".
+It is often impossible to know, however, the extent to which a model provides a faithful representation of the system in question.
+In this section, we discuss some tools and suggestions for diagnosing mechanistic models with the goal of making the subjective assessment of model ``usefulness" more objective by relying on the quantitative and statistical ability of the model to match the observed data, which we call the model's goodness-of-fit.
 Goodness-of-fit may provide evidence supporting the causal interpretation of one model versus another, but cannot by itself rule out the possibility of alternative explanations.
@@
 this is done to determine whether or not the observed data are a plausible realization of the fitted model.
-While these comparisons are useful, relying only on these qualitative comparisons may lead to over confidence on the goodness-of-fit.
-A relevant example is the original parameter estimates provided by \citet{lee20}:
-these parameters resulted in simulations that closely resembled the observed data, yet resulted in model likelihoods that were---in some cases---drastically smaller than likelihoods that can be achieved via the likelihood based optimization techniques that were used (see Table~\ref{tab:likes}).
+While these comparisons are informative, they provide only a naive measure of the goodness-of-fit of a model.
+A relevant example of these comparisons leading to overconfidence in a model's goodness-of-fit is the original parameter estimates provided by \citet{lee20}:
+these parameters resulted in simulations that closely resembled the observed data, yet resulted in model likelihoods that were---in some cases---remarkably smaller than likelihoods that can be achieved via the likelihood based optimization techniques that were used (see Table~\ref{tab:likes}).
 % \eic{TODO: Could we maybe put an example of this here? }
@@
 
-The goodness-of-fit of a model can be assessed by comparing a quantitative description of the model fit (such as MSE, predictive accuracy, or model likelihood) to that of other proposed models.
+Another approach is to compare a quantitative description of the model fit (such as MSE, predictive accuracy, or model likelihood) to that of other proposed models.
 This type of comparison provides insight into how each model performs relative to the others, but does not provide insight on the performance of each model on a given dataset unless the performance of one of the models is widely understood.
@@
 Comparing associative models using an objective function seems natural, as these models are designed to optimize objective functions given the available data.
-In the case of mechanistic modeling, however, comparing models based on an objective measure is less natural because each mechanistic model is designed to provide a scientific description of the dynamic system.
+In the case of mechanistic modeling, however, comparing models based on an objective measure is less natural because each model is designed to provide a scientific description of the dynamic system.
 Any comparison between mechanistic models would therefore ideally be based on the scientific validity of each proposed model, which is often impossible to know.
@@
 % When performing such a comparison, however,
-We do not suggest that the model with the highest likelihood given the data automatically provides a better scientific description of the dynamic system;
-still, a comparison based on model likelihoods may help guide researchers to know the plausibility of a proposed model compared to another given their relative ability to describe the observed data.
-
-This comparison still only provides a relative measure of model fit compared to other mechanistic models.
+We do not suggest that the model with the highest likelihood given the data automatically provides a better scientific description of the dynamic system.
+Still, a comparison based on model likelihoods may help guide researchers to know the plausibility of a proposed model compared to another given their relative ability to describe the observed data.
+
+The comparison across models still only provides a relative measure of model fit compared to other mechanistic models.
 The resulting challenge is that no general benchmark exists for mechanistic models that would allow researchers to assess the goodness-of-fit by comparing to a known standard.
@@
 % In the case of Cholera infections in Haiti, for example, it has been debated whether or not bacteria buildup in the environment is a cause of Cholera infections \cite{???}.
-Because of this, we propose the use of an ARMA(2, 1) model as a potential benchmark.
-Despite being an associative model, the reason that an ARMA(2, 1) model is an appealing benchmark is because the performance of this class of models is widely understood, and often provides a good description of the data with very limited effort.
+Because of this, we propose the use of a log-linear ARMA model as a potential benchmark.
+Despite being an associative model, the reason that an ARMA model is an appealing benchmark is because the performance of this class of models is widely understood, and often provides a good description of the data with very limited effort.
 % For us, the defining feature of an associative model is that it is assessed primarily on its quantitative ability to explain historical data.
@@
 comparing Model~1 to Model~3).
-Likelihoods of Models~1--3 and their respective ARMA(2, 1) benchmark models are provided in Table~\ref{tab:likes}.
+Likelihoods of Models~1--3 and their respective ARMA benchmark models are provided in Table~\ref{tab:likes}.
 
@@
 Conditional likelihoods, defined as the density $f_{Y_k | Y_1, \ldots, Y_{k - 1}}\big(Y_k = y_k^* | y_{1:k-1}^*\big)$, provide a basic description of how well the proposed model can describe each data point, given the previous observations.
-Comparing these results across models---including benchmark models--- can help researchers identify potential model deficiencies, or errors in the observed data.
-Additional tools for assessing the goodness-of-fit of a model include plotting the effective sample size of each observation \jwc{REF???} and comparing any statistic of the observed data to simulations from the model, which is sometimes referred to as diagnostic probes (for example, the autocorrelation function (ACF), as was done in \cite{king15}).
-
-\subsection{Forecasts}\label{sec:filter}
-
-\footnotetext[1]{The reported likelihood is an upper bound of the likelihood of the \citet{lee20} model as it is the largest likelihood obtained using their parameter calibration regime.}
+Comparing these results across models---including benchmark models---can help researchers identify potential model deficiencies, or errors in the observed data.
+Additional tools for assessing the goodness-of-fit of a model include plotting the effective sample size of each observation \jwc{REF???} and comparing any statistic of the observed data to simulations from the model, which is sometimes referred to as diagnostic probes (for example, the autocorrelation function (ACF), as was done in \cite{king15}).\footnotetext[1]{The reported likelihood is an upper bound of the likelihood of the \citet{lee20} model as it is the largest likelihood obtained using their parameter calibration regime.}
 \footnotetext[2]{Parameters to department-specific models not provided by \citet{lee20}, as department fits were only used as an intermediary step to obtain parameter estimates of the coupled model.}
@@
 On this same subset of data, our model achieved a likelihood of $\Sexpr{myround(mod3_spatPomp_res$subset_ll, 1)}$.}
+
+\subsection{Forecasts}\label{sec:filter}
+
+\jwc{NEW PARAGRAPH:}
 Forecasting is a powerful tool that has primarily been used to estimate the future state of a system; when a mechanistic model is used, forecasts may provide estimates of the potential effects interventions may have on a system.
@@
 As an example, we note that simulations from a well-fit mechanistic model may closely resemble the observed data $y_{1:N}^*$.
-In such comparisons, these simulations are random draws from the complete estimated joint distribution $f_{\bm{X}^{(\modelCounter)}_{0:N}, \bm{Y}^{(\modelCounter)}_{1:N}}\left(x_{0:N}, y_{1:N}\mid \hat{\theta}\right)$, where $\modelCounter$ indexes the model used for simulations.
-It can therefore be tempting to use simulations from this model up to time $N+s$, $f_{\bm{X}^{(\modelCounter)}_{0:N+s}, \bm{Y}^{(\modelCounter)}_{1:N+s}}\left(x_{0:N+s}, y_{1:N+s}\mid \hat{\theta}\right)$, with $s\geq 0$ to project the dynamic system up to a future time $N+s$, as it has been done in previous studies \citep{lee20,???}.
+In such comparisons, these simulations are random draws from the complete estimated joint distribution $f_{\bm{X}^{(\modelCounter)}_{0:N}, \bm{Y}^{(\modelCounter)}_{1:N}}\left(x_{0:N}, y_{1:N}\mid \hat{\paramVec}\right)$, where $\modelCounter$ indexes the model used for simulations.
+It can therefore be tempting to use simulations from this model up to time $N+s$, $f_{\bm{X}^{(\modelCounter)}_{0:N+s}, \bm{Y}^{(\modelCounter)}_{1:N+s}}\left(x_{0:N+s}, y_{1:N+s}\mid \hat{\paramVec}\right)$, with $s\geq 0$ to project the dynamic system up to a future time $N+s$, as it has been done in previous studies \citep{lee20,???}.
 % \eic{TODO: I think it's possible that other people have done this too. Could we find another reference?};
@@
 In other words, observing the data $\bm{Y}^{(\modelCounter)}_N = y_N^*$ at time $N$ provides more information about the future state of the system than the initial conditions.
-In this case, draws from the conditional density $f_{\bm{X}^{(\modelCounter)}_{N:s}, \bm{Y}^{\modelCounter}_{N:s}}\big(x_{N:s}, y_{N:s}\mid \hat{\theta}, \bm{X}^{(\modelCounter)}_{N} = \bm{x}^{(\modelCounter)}_N\big)$ should be preferred, as these simulations account for the most recent known state of the system.
+In this case, draws from the conditional density $f_{\bm{X}^{(\modelCounter)}_{N:s}, \bm{Y}^{\modelCounter}_{N:s}}\big(x_{N:s}, y_{N:s}\mid \hat{\paramVec}, \bm{X}^{(\modelCounter)}_{N} = \bm{x}^{(\modelCounter)}_N\big)$ should be preferred as forecasts, as these simulations account for the most recent known state of the system.
 
 $\bm{X}^{(\modelCounter)}_N$, however, is unobservable and therefore draws from the desired conditional density are unobtainable.
-Informed estimates of $\bm{X}^{(\modelCounter)}_N$ given the observed data $\bm{Y}^{(\modelCounter)}_{1:N} = y_{1:N}^*$ and the proposed model can easily be obtained, however, via the filtering distribution.
-Let $\hat{\bm{X}}^{(\modelCounter), i}_N, \, i \in {1, 2, \ldots, J}$ be $\iid$ draws from the filtering distribution at time $N$, with density $\hat{\bm{X}}^{(\modelCounter), i}_N \sim f_{\bm{X}^{(\modelCounter)}_N|\bm{Y}^{(\modelCounter)}_{1:N}}(x_N \mid \hat{\theta}, y_{1:N}^*)$.
-A single model forecast can then be obtained by simulating from the model $f_{\bm{X}^{(\modelCounter)}_{N:s}, \bm{Y}^{(\modelCounter)}_{N:s}}\big(x_{N:s}, y_{N:s}\mid \hat{\theta}, \bm{X}^{(\modelCounter)}_{N} = \hat{\bm{X}}^{(\modelCounter), i}_N\big)$.
+Informed estimates of $\bm{X}^{(\modelCounter)}_N$ given the observed data $\bm{Y}^{(\modelCounter)}_{1:N} = y_{1:N}^*$ can easily be obtained, however, via the filtering distribution.
+Let $\hat{\bm{X}}^{(\modelCounter), i}_N, \, i \in {1, 2, \ldots, J}$ be $\iid$ draws from the filtering distribution at time $N$, with density $\hat{\bm{X}}^{(\modelCounter), i}_N \sim f_{\bm{X}^{(\modelCounter)}_N|\bm{Y}^{(\modelCounter)}_{1:N}}(x_N \mid \hat{\paramVec}, y_{1:N}^*)$.
+A single model forecast can then be obtained by simulating from the model $f_{\bm{X}^{(\modelCounter)}_{N:s}, \bm{Y}^{(\modelCounter)}_{N:s}}\big(x_{N:s}, y_{N:s}\mid \hat{\paramVec}, \bm{X}^{(\modelCounter)}_{N} = \hat{\bm{X}}^{(\modelCounter), i}_N\big)$.
 Intuitively, simulating the model starting at the filtering distribution of the most recently available time point is a more appropriate way to project a stochastic dynamic system into the future, as it is not expected that each simulation from initial conditions will result in a latent state at time $N$ consistent with the model and the observation $\bm{Y}^{(\modelCounter)}_N = y_N^*$.
@@
 It has been noted that the uncertainty in just a single parameter can lead to drastically different projections \citep{saltelli20}.
-One possible approach to account for parameter uncertainty in model projections is by obtaining confidence intervals for each parameter, sampling parameters from the confidence intervals, simulating the model with the resulting parameter set, and then weighing the resulting model projections based on the likelihood of the given set of parameters, as was done by \citet{king15}.
+One possible approach to account for parameter uncertainty in model forecasts is by obtaining confidence intervals for each parameter, sampling parameters from the confidence intervals, simulating the model with the resulting parameter set, and then weighing the resulting model projections based on the likelihood of the given set of parameters, as was done in \citet{king15}.
 
@@
 \jwc{Replaced Ed's comment with this: }
-We note---the credit of deterministic models---that of the four fitted models in \cite{lee20}, Model~2 provides what seems to be the most accurate model projections.
+We note---the credit of deterministic models---that of the four fitted models in \cite{lee20}, Model~2 provides the most apparently accurate forecasts.
 This perhaps demonstrates that while deterministic models describe the systems in a less realistic and useful way, the relative ease in fitting these models potentially results in fewer modeling errors.
@@
 
-\eic{HERE'S AN INITIAL ATTEMPT TO EXPLAIN MY POINT BELOW FOR THE MS. IT IS HARD TO MAKE CONCRETE POINTS ON THIS SORT OF TOPIC.}
-A model which aspires to provide quantitative guidance for assessing interventions should provide a quantitative statistical fit for previous data.
+\jwc{SHOULD THESE FIRST TWO PARAGRAPHS BE MOVED TO THE DISCUSSION?}
+A model which aspires to provide quantitative guidance for assessing interventions should provide a quantitative statistical fit for available data.
 However, strong statistical fit does not guarantee a correct causal structure: it does not even necessarily require the model to assert a causal explanation.
@@
 parameter values which fit the data should make sense in the context of alternative lines of evidence about the phenomena being modeled.
-\jwc{Perhaps add comment on our final parameter estimates, and remove the following sentences in this paragraph.}
-A model is necessarily a simplified approximation of a complex process, and so it is unreasonable to expect models (describing and fitted to population-level processes) to perfectly match quantitative understanding of individual-level processes.
-Such discrepancies could lead to biases when interventions (modeled as effects on individuals) are included in the population-level, unless there are historical occurrences of the intervention which enable the consequence to be calibrated at the population level.
+% \jwc{Perhaps add comment on our final parameter estimates, and remove the following sentences in this paragraph.}
+% A model is necessarily a simplified approximation of a complex process, and so it is unreasonable to expect models (describing and fitted to population-level processes) to perfectly match quantitative understanding of individual-level processes.
+% Such discrepancies could lead to biases when interventions (modeled as effects on individuals) are included in the population-level, unless there are historical occurrences of the intervention which enable the consequence to be calibrated at the population level.
 
@@
 
-An additional benefit of mechanistic modelling is ability to simulate various interventions on a system; this feature is useful to inform policy and was the primary goal of \citet{lee20}.
+An additional benefit of mechanistic modeling is ability to simulate various interventions on a system; this feature is useful to inform policy and was the primary goal of \citet{lee20}.
 Outcomes of their study include estimates for the probability of cholera elimination and cumulative number of cholera infections under several possible vaccination scenarios.
-Mimicking their efforts, we define cholera elimination as having less than one infection of cholera over at least 52 consecutive weeks in the 10-year projection period, and provide projections under the following vaccination scenarios:
+Mimicking their efforts, we define cholera elimination as having less than one infection of cholera over at least 52 consecutive weeks in the 10-year projection period, and provide forecasts under the following vaccination scenarios:
 
 \begin{itemize}
-  \item[$V0$:] The default vaccination scenario, which is that no additional vaccinations are deployed.
-  \item[$V1$:] Vaccination limited to the departements of Centre and Artibonite, and vaccinations deployed over a two-year period.
-  \item[$V2$:] Vaccination limited to three departements: Artibonite, Centre, and Ouest over a two-year period.
+  \item[$V0$:] No additional vaccines are administered.
+  \item[$V1$:] Vaccination limited to the departments of Centre and Artibonite, deployed over a two-year period.
+  \item[$V2$:] Vaccination limited to three departments: Artibonite, Centre, and Ouest deployed over a two-year period.
   \item[$V3$:] Countrywide vaccination implemented over a five-year period.
@@
 The results of these projections are summarized in Figs.~\ref{fig:Mod1Scenarios}--\ref{fig:elimProbs}.
-These results suggest that cholera elimination was likely, even without intense vaccination efforts.
-This conclusion is consistent with the fact that cholera was eradicated from Haiti without additional vaccination efforts \citep{ferguson22}.
-
-In additional to probability of elimination estimates, we are also able to estimate the cumulative number of infections under each vaccination scenario from February 2019 -- February 2024.
+These results suggest that cholera elimination was likely, even without increased vaccination efforts, which is consistent with observed reality \citep{ferguson22}.
+
+Probability of elimination estimates of this form are not meaningful for deterministic models, as the trajectory of these models only represent the mean behavior of the system rather than individual potential outcomes.
+We therefore do not provide probability of elimination estimates under Model~2.
+Still, trajectories obtained by Model~2 are consistent with the simulation results of Models~1 and~3, and suggest that cholera was in the process of being eliminated from Haiti.
+
+In additional to probability of elimination estimates, we provide estimates for the cumulative number of infections under each vaccination scenario from February 2019 -- February 2024.
 Notably, the median number of cumulative cholera infections under the no-vaccination scenario using Models 1 and 3 were $\Sexpr{prettyNum(round(as.numeric(mod1_V0_probs$cumInf['q50'])), big.mark = ",")}$ and $\Sexpr{prettyNum(round(as.numeric(mod3_V0_res$mod3_V0_probs$cumInf['q50'])), big.mark = ",")}$, respectively.
-So far, these estimates are far more consistent with the observed number of reported cholera cases than the corresponding estimates from \cite{lee20}, which were approximately $400,000$ and $1,000,000$.
+While there is remaining time during this projection period in which new cholera infections can be detected, up to this point our estimates are far more consistent with the observed number of reported cholera cases than the corresponding estimates from \cite{lee20}, which were approximately $400,000$ and $1,000,000$.
 
@@
 
-Probability of elimination estimates of this form are far less meaningful for deterministic models.
-This is because the data are not assumed to be draws from the model; the model is instead intended only to provide understanding about the overall trend of the epidemic.
-We therefore do not provide probability of elimination estimates under Model~2.
-Still, projected trajectories using Model~2 are consistent with those found with Models~1 and ~3, and suggest that cholera was in the process of being eliminated from Haiti.
-
 % \begin{figure}[ht]
@@
 
-Despite these results suggesting that cholera was likely to be eliminated from Haiti without an increase in vaccination effort, we in no way suggest that vaccinations shouldn't be used to combat cholera---or other infectious diseases---in other parts of the world.
-In fact, our analysis shows that an increase in vaccination does drastically reduce the negative effects of a worst-case scenario.
-% This alone should provide policy makers with enough motivation to continue vaccination efforts around the country.
+\jwc{PARAGRAPH REMOVED}
+% Despite these results suggesting that cholera was likely to be eliminated from Haiti without an increase in vaccination effort, we in no way suggest that vaccinations shouldn't be used to combat cholera---or other infectious diseases---in other parts of the world.
+% In fact, our analysis shows that an increase in vaccination does drastically reduce the negative effects of a worst-case scenario.
 % Avoid comments on vaccination policy recommendations - Jesse
@@
 % TODO: "Improved model fits", or "Improved model fitting", based on Anna's suggestion.
-Caution is nonetheless needed when making policy based on modelling conclusions, as model misspecification may invalidate conclusions.
+Caution is nonetheless needed when making policy based on modeling conclusions, as model misspecification may invalidate conclusions.
 % TODO: Citation above?

### 2022-08-31 si/si.Rnw
Moved si file over, fixed minor issues

\documentclass[11pt]{article}

%%% JASA. Attention Mac Users. The latest version of the pdfTex compiler for Mac (pdfTeX-1.40.11) outputs files not currently supported by ScholarOne Manuscripts. To resolve this and have a PDF file that will convert properly, the PDF file will need to be distilled to Adobe version 1.4 or earlier. To create a PDF file that uploads correctly, place the command \pdfminorversion=4 in the preamble (before the begin{document}) of your LaTeX file and run pdfLaTeX again.

\pdfminorversion=4

<<debug,echo=F>>=
run_level <- 1

if(run_level=="manual"){
}

@

\newcommand\secTitleSpace{\hspace{3mm}}

<<packages, include=FALSE, cache=FALSE, echo=FALSE>>=
library(tidyverse)
#library(xtable)
library(pomp)
library(haitipkg)
# library(spatPomp)
library(doParallel)
library(doRNG)

cores <-  as.numeric(Sys.getenv('SLURM_NTASKS_PER_NODE', unset = NA))
# if(is.na(cores)) cores <- detectCores()
cores <- 20
registerDoParallel(cores)
theme_set(theme_bw())
@

<<set-opts, include=FALSE, cache=FALSE, echo=FALSE>>=

options(
  scipen = 2,
  help_type = "html",
  stringsAsFactors = FALSE,
  # prompt = "R> ",
  continue = "+  ",
  width = 70,
  useFancyQuotes = FALSE,
  reindent.spaces = 2,
  xtable.comment = FALSE
)
@

<<knitr-opts, include=FALSE, cache=FALSE, purl=FALSE>>=
library("knitr")
opts_knit$set(concordance=TRUE)
opts_chunk$set(
  progress=TRUE,prompt=TRUE,highlight=FALSE,
  tidy=TRUE,
  tidy.opts=list(
    keep.blank.line=FALSE
  ),
  comment="",
  warning=FALSE,
  message=FALSE,
  error=TRUE,
  echo=TRUE,
  cache=FALSE,
  strip.white=TRUE,
  results="markup",
  background="#FFFFFF00",
  size="normalsize",
  fig.path="figure/",
  fig.lp="fig:",
  fig.align="center",
  fig.show="asis",
  #    dpi=300,
  dev="pdf",
  dev.args=list(
    bg="transparent",
    pointsize=9
  )
)

myround <- function (x, digits = 1) {
  # taken from the broman package
  if (digits < 1)
    stop("This is intended for the case digits >= 1.")
  if (length(digits) > 1) {
    digits <- digits[1]
    warning("Using only digits[1]")
  }
  tmp <- sprintf(paste("%.", digits, "f", sep = ""), x)
  zero <- paste0("0.", paste(rep("0", digits), collapse = ""))
  tmp[tmp == paste0("-", zero)] <- zero
  tmp
}

@

%% copied from pnas-new.cls
\usepackage{amsmath,amsfonts,amssymb,graphicx}

\input{header-ms.tex}

\bibliographystyle{apalike}
%% added or modified from pnas-new.cls
\usepackage{natbib}
\usepackage{fullpage}

\newcommand\myeqref[1]{(\ref{#1})}

%\usepackage{relsize}

%\templatetype{pnassupportinginfo}

% \readytosubmit %% Uncomment this line before submitting, so that the instruction page is removed.

%\author{Author1, Author2 and Author3 (Complete author list)}
%\correspondingauthor{Corresponding Author Name.\\E-mail: author.two@email.com}

\renewcommand{\contentsname}{Supplementary Content}
\renewcommand{\refname}{Supplementary References}
\renewcommand\thefigure{S-\arabic{figure}}
\renewcommand\thetable{S-\arabic{table}}
%\renewcommand\thepage{S-\arabic{page}}
\renewcommand\thesection{S\arabic{section}}
\renewcommand\theequation{S\arabic{equation}}
\renewcommand\theprop{S\arabic{prop}}
\renewcommand\thelemma{S\arabic{lemma}}
\renewcommand\thealgocf{S\arabic{algocf}}

\setcounter{tocdepth}{1}

\begin{document}

%% Comment/remove this line before generating final copy for submission
%\instructionspage

\date{\today}
\title{Supplement to ``{\mytitle}''}
  \author{   Provisional author list\\
 Jesse Wheeler, Anna Rosengart, Zhuoxun Jiang, Kevin Hao En Tan, Noah Treutle, Edward L. Ionides
 \\ \hspace{.2cm}\\
    Department of Statistics, University of Michigan\\
}

\newcommand{\blind}{1}

\if1\blind
{
\maketitle
}
\fi

\if0\blind
{
  \bigskip
  \bigskip
  \bigskip
  \begin{center}
    {\LARGE\bf \mytitle}
\end{center}
  \bigskip
  \bigskip
}\fi

%% Adds the main heading for the SI text. Comment out this line if you do not have any supporting information text.

%\SItext

%Draft: \today

\tableofcontents

\newpage

\section{Markov chain and differential equation interpretations of compartment flow rates}

In Sections 2.1, 2.2 and 2.3 we define compartment models in terms of their flow rates.
For a discrete population model, these rates define a Markov chain.
For a continuous and deterministic model, the rates define a system of ordinary differential equations.
Here, we add additional details to clarify the mapping from a collection of rate functions to a fully specified process.
Our treatment follows \citet{breto09}.

A general compartment model is a vector-valued process $X(t)=(X_1(t),\dots,X_c(t))$ denoting the (integer or real-valued) counts in each of $c$ compartments.
The compartments may also have names, but to set up general notation we simply refer to them by their numerical index.
The basic characteristic of a compartment model is that $X(t)$ can be written in terms of the flows $N_{ij}(t)$ from $i$ to $j$, together with flows into and out of each compartment denoted by $N_{\bullet i}(t)$ and $N_{i\bullet}(t)$ respectively.
These flows are required to satisfy a ``conservation of mass'' identity:
\begin{equation}
X_i(t)=X_i(0)+N_{\bullet i}(t) - N_{i\bullet}(t) + \sum_{j\neq
i}N_{ji}(t)-\sum_{j\neq i}N_{ij}(t). \label{eq:conservation}
\end{equation}
Each {\em flow} $N_{ij}(t)$ is associated with a {\em rate} function
$\mu_{ij}=\mu_{ij}(t,X(t))$, where we include the possibility that $i$ or $j$ takes value $\bullet$.

There are different ways to use a collection of rate functions to build a fully specified model.
We proceed to describe the ones we use in this paper: via a system of ordinary differential equations (Sec.~\ref{subsec:ode}), a simple Markov counting system (Sec.~\ref{subsec:smcs}), and an overdispersed Markov counting system (Sec.~\ref{subsec:odmcs}). Other representations include stochastic differential equations driven by Gaussian noise or Gamma noise \citep{bhadra11}.

\subsection{Ordinary differential equation (ODE) interpretation}
\label{subsec:ode}

A basic deterministic specification is
\begin{equation}
\label{eq:ode1}
dN_{ij}/dt = \mu_{ij}\big(t,X(t) \big) X_i(t), \medskip i\in \seq{1}{c}, \medskip j\in \seq{1}{c} \cup\{\bullet\}, \medskip i\neq j,
\end{equation}
where $\mu_{ij}\big(t,X(t)\big)$ is called a per-capita rate or a unit rate.
Flows into the system require special treatment since $X_i(t)$ in \myeqref{eq:ode1} is not defined for $i=\bullet$.
Instead, we specify
\begin{equation}
\label{eq:ode2}
dN_{\bullet i}/dt = \mu_{\bullet i}\big(t,X(t) \big).
\end{equation}

\subsection{Simple Markov counting system interpretation}
\label{subsec:smcs}
A continuous time Markov chain can be specified via its infinitesimal transition probabilities.
A basic approach to this is to define
\begin{eqnarray}
\label{eq:smcs1}
\prob\big[ N_{ij}(t+\delta)-N_{ij}(t)=0 \given X(t)\big]
 &=& 1-\delta \mu_{ij}\big(t,X(t)\big) + o(\delta),
\\
\label{eq:smcs2}
\prob\big[ N_{ij}(t+\delta)-N_{ij}(t)=1 \given X(t)\big]
 &=& \delta \mu_{ij}\big(t,X(t)\big) + o(\delta),
\end{eqnarray}
for $i\in \seq{1}{c}$ and $j\in\seq{1}{c}\cup\{\bullet\}$ with $i\neq j$.
As with the ODE case, we need special attention for flows into the system, and we define
\begin{eqnarray}
\label{eq:smcs3}
\prob\big[ N_{\bullet i}(t+\delta)-N_{\bullet i}(t)=0 \given X(t)\big]
 &=& 1-\delta \mu_{\bullet i}\big(t,X(t)\big) + o(\delta),
\\
\label{eq:smcs4}
\prob\big[ N_{\bullet i}(t+\delta)-N_{\bullet i}(t)=1 \given X(t)\big]
 &=& \delta \mu_{\bullet i}\big(t,X(t)\big) + o(\delta).
\end{eqnarray}
Together with the initial conditions $X(0)$, equations \myeqref{eq:smcs1}--\myeqref{eq:smcs4} define a Markov chain.
Each flow is a simple counting process, meaning a non-decreasing integer-valued process that only has jumps of size one.
We therefore call the Markov chain a simple Markov counting system (SMCS).
The infinitesimal mean of every flow is equal to its infinitesimal variance \citep{breto11} and so an SMCS is called equidispersed.
To permit more general mean-variance relationships for a Markov counting system, we must permit jumps of size greater than one.
The utility of overdispersed models, where the infinitesimal variance of the flow exceeds the infinitesimal mean, has become widely recongized \eic{REFS}.

\subsection{Overdispersed Markov counting system interpretation}
\label{subsec:odmcs}

Including white noise in the rate function enables the possibility of an overdispersed Markov counting system \citep{breto11,breto09,he10}.
Since rates should be non-negative, Gaussian noise is not appropriate and gamma noise is a convenient option that has found various applications \eic{REFS}.
Specifically, we consider a model given by
\begin{equation}
\label{eq:odmcs1}
\mu_{ij}\big(t,X(t)\big) = \bar\mu_{ij}\big(t,X(t)\big) \, d\Gamma_{ij}(t)/dt,
\end{equation}
where $\Gamma_{ij}(t)$ is a stochastic process having independent gamma distributed increments, with
\begin{equation}
\label{eq:odmcs2}
\E\big[\Gamma_{ij}(t)\big] = t, \quad \var\big[\Gamma_{ij}(t)\big] = \sigma_{ij}^2 t.
\end{equation}
Formally interpreting the meaning of \myeqref{eq:odmcs1} is not trivial, and we do so by defining the solution of \myeqref{eq:odmcs1} to be the limit of an Euler scheme.
% This is equivalent to interpreting a stochastic differential equation via its Ito solution \eic{CHECK THIS CLAIM. I THINK IT MIGHT BE TRUE. IF SO, IT WOULD PROVIDE A NICE BRIDGE BETWEEN DEVELOPMENT OF IMPLICIT MODELS (I.E., DEFINING MODELS VIA NUMERICAL SOLUTIONS, OFTEN DONE USING AN EULER METHOD) AND THE SUCCESSFUL THEORY OF ITO SOLUTIONS TO STOCHASTIC DIFFERENTIAL EQUATIONS.}
Therefore, the numerical scheme in Sec.~\ref{sec:numerics} can be taken as a definition of the meaning of \myeqref{eq:odmcs1}.
The Markov chain defined by the limit of this Euler scheme as the step size decreases is an overdispersed Markov counting system, with the possibility of instantaneus jumps of size greater than one \citep{breto11}.

\section{Numerical solutions to compartment models}
\label{sec:numerics}

Models may be fitted and their implications assessed via numerical solutions (i.e., simulations) from the model equations.
All the analyses we consider have this simulation-based property, known as plug-and-play or equation-free or likelihood-free.
The numerical solutions to the model are arguably of more direct scientific interest than the exact solutions to the postulated equations.
For ODE models, numerical methods are well studied and a standard numerical solution package such as deSolve in R is adequate for our purposes.
For SMCS and ODMCS models, exact schemes are feasible when the number of events is small, which may be the case for small populations.
However, for applicability to larger populations, we use instead the following Euler scheme.
Write $\delta$ for an Euler time step, and $\Delta N_{ij}$ for the numerical approximation to $N_{ij}(t+\delta)-N_{ij}(t)$ given $X(t)$.
For each $i$ and $j$ in $\seq{1}{c} \cup \{\bullet\}$ with $i \neq j$, we draw independent Gamma distributed noise increments with mean $\delta$ and variance $\sigma_{ij}^2 \delta$, denoted using a mean-variance parameterization of the gamma distribution as
\begin{equation}
\label{eq:numerics1}
\Delta\Gamma_{ij} \sim \mathrm{gamma}(\delta, \sigma_{ij}^2 \delta).
\end{equation}
In the case of an SMCS model, $\sigma_{ij}=0$ for all $i$ and $j$, so we have $\Delta\Gamma_{ij}=\delta$.
Then, for $i\neq\bullet$ and $j\neq i$, and writing
\begin{equation}
\label{eq:numerics2}
\mu_{ij}=\bar\mu_{ij}\big(t,X(t)\big) \Delta\Gamma_{ij} / \delta,
\end{equation}
we calculate transition probabilities
\begin{eqnarray}
\label{eq:numerics3}
p_{ij} &=& \exp\left\{-\sum_{k\in 1:c \, \cup \{\bullet\}} \mu_{ik} \, \delta \right\}
\frac{\mu_{ij}}{\sum_{k\in 1:c\cup \{\bullet\}} \mu_{ik}},
\\
\label{eq:numerics4}
p_{ii} &=& 1 - \sum_{j\neq i} p_{ij}.
\end{eqnarray}
These probabilities correspond to competing hazards for every individual in compartment $i$ to transition to some compartment $j$, interpreting $j=i$ to mean that the individual remains in $i$.
Then, $\big(\Delta N_{i1},\dots,\Delta N_{ic},\Delta N_{i\bullet}\big)$ has the multinomial distribution where $X_i(t)$ individuals are allocated independently to $\seq{1}{c}\cup\{\bullet\}$ with probabilities given by \eqref{eq:numerics3} and \eqref{eq:numerics4}.
We use the \code{reulermultinom} function in the \code{pomp} package to draw from this multionomial distribution.

For the case $i=\bullet$, one can use
\begin{equation}
\label{eq:numerics6}
\Delta N_{\bullet j} \sim \mathrm{poisson}( \mu_{\bullet j} \delta),
\end{equation}
an independent Poisson random variable with mean $\mu_{\bullet j} \delta$.
To address the issue of making our model consistent with total population as a known quantity, based on census data, in practice we use a deterministic specification of $\Delta N_{\bullet j}$ rather than the Poisson form in \eqref{eq:numerics6}.
\eic{CHECK THIS. DO WE HAVE A CONSTANT POPULATION, OR A POPULATION OF DETERMINISTICALLY GROWING SIZE? DO WE USE BIRTH DATA OR TOTAL POPULATION DATA?}
To make birth/immigration events consistent with death/emigration events and population census, we formally model the death rate as a rate of returning to the susceptible class $S$, and use external transitions from $\bullet$ into $S$ to describe only net population increase.
\jwc{CHECK THIS}.

\section{Initial Values}

To perform inference on POMP models, it is necessary to propose an initial density for the latent process $f_{X_0}(x_0;\theta)$.
This density is used to obtain initial values of the latent state when fitting and evaluating the model.
For each of the models considered in this analysis, the initial conditions are derived by enforcing the model dynamics on reported cholera cases.
For example, in a deterministic model, the total number of Infected and Asymptomatic individuals may be calculated using the observed number of cholera cases at time time of model initialization based on model parameters such as reporting rate $\reportRate$ and fraction of symptomatic cases $\symptomFrac$.
It is also sometimes necessary to fit some initial value parameters in order to help determine starting values for weakly identifiable compartments.
In the following subsections, we mention initial value parameters that were fit for each model.

\subsection{Model~1}

For this model, the number of individuals in the Recovered and Asymptomatic compartments are set to zero, but the initial proportion of Infected and Exposed individuals is estimated as initial value parameters ($I_0$ and $E_0$, respectively) using the MIF2 algorithm.
Finally, the initial proportion of Susceptible individuals $S_0$ is calculated as $S_0 = 1 - I_0 - E_0$.

\subsection{Model~3}

We use the reported cases at the start of the pandemic to approximate the number of Asymptomatic, Infectious, and Recovered individuals in each department $u \in 1:U$ using the same approximation as provided in Eq.~\myeqref{eq:model3:approx}.
The susceptible compartment is initialized so that the sum $S_{u}(0) + I_u(0) + A_u(0) + \sum_k R_{u, k}(0) = \text{population}_u$.
The bacteria compartment is then initialized using Eq.~\myeqref{eq:model3:b0}:

\begin{equation}
  B_u(0) = \big[1 + \seasAmplitude \big(\xi_u)^r \big] D_i \, \Wshed \big[ I_{u}(0)+ \asymptomRelativeShed A_{u}(0) \big]\label{eq:model3:b0}
\end{equation}

Where $\xi_u \in (0, 1)$ are initial value parameters that we introduce in order to allow some flexibility in determining the initial state of the bacteria compartment.

\section{Measurement Models}

Each POMP requires specification of a measurement model, which is a statistical description of how observations on the system are obtained.
In general, we used the same measurement models that were reported in \cite{lee20}.

\subsection{Model~1}

In this model, the advantage afforded by vaccination is an increased probability that an infection is asymptomatic.
Therefore, under the assumptions of this model, all reported cases are assumed to be a fraction of individuals that transition from the exposed to the infected compartment, as noted in Eq.~\myeqref{model1:meas}.
\begin{equation}
  \label{model1:meas}
  y_{t} \mid \Delta N_{E_{k}I_{k}} = z_{t} \sim \text{NB}\left(\reportRate z_{t}, \obsOverdispersion \right)
\end{equation}

\subsection{Model~2}

Model~2 was fit via trajectory matching and therefore does not have an explicit measurement model.
Implicitly, however, this fitting process is equivalent to having a Gaussian measurement process.
Therefore the measurement model can be expressed as Eq.~\myeqref{model2:meas}.
\begin{equation}
  \label{model2:meas}
  y_{u, t} \mid \Delta N_{I_{ud}R_{ud}} = z_{u, t} \sim \text{N}\left(\reportRate z_{u, t}, \obsOverdispersion^2 \right)
\end{equation}

Where $\Delta N_{I_{ud}R_{ud}}$ is the number of individuals who move from compartment $I_{ud}$ to $R_{ud}$, for unit $u \in 1:U$ and vaccination cohort $d \in 0:5$.
We also note here that the variance parameter $\obsOverdispersion^2$ is separately fit for the epidemic and endemic phases.

\subsection{Model~3}

In this model, reported cholera cases are assumed to stem from individuals who develop symptoms and seek healthcare.
Therefore reported cases are assumed to come from an over-dispersed negative binomial model, given the increase in infected individuals:
\begin{equation}
  \label{model3:meas}
  y_{u, t} \mid \Delta N_{S_{ud}I_{ud}} = z_{u, t} \sim \text{NB}\left(\reportRate z_{u, t}, \obsOverdispersion \right)
\end{equation}

This measurement model is a minor change from that used in \cite{lee20}, which allowed for a change in the reporting rate on January 1st, 2018.
The fitted values of the reporting rate---before and after January 2018---were $0.97$ and $0.097$, respectively.
This major change in reporting rate alone could have been the cause that Model~3 originally failed to predict the eradication of cholera, as an overnight change from near perfect to almost non-existent reporting forces the model to explain the observed decrease in cases as a decrease in the reporting of cases rather than of prevalence of cases.
This shift was justified by a "change of the case definition that occurred on January 1st, 2018";
this claim was not cited, and we could find no evidence that such a drastic change in the reporting rate would be warranted.

\section{Block Panel Iterated Filter}

The panel iterated filter (PIF) \citep{breto20} has been successfully used to fit parameters $\theta$ of panelPOMPs in several studies \jwc{TODO: REFs}.
Like other iterated filtering algorithms, PIF iteratively combines particle filter calculations with parameter perturbations.
Theoretical results show that, as the perturbations are reduced over successive iterations, the perturbed parameters converge to the MLE.

In the case of a panelPOMP, the parameter vector $\theta$ may be composed of a set of shared parameters $\phi$, and a set of unit-specific parameters $\psi_{\seq{1}{U}}$, for units $\seq{1}{U}$.
Here, we say that a parameter $\psi_{k}$ for $k \in \seq{1}{U}$ is unit-specific if it is not involved in the one-step transition density or the measurement model of any other unit $l \neq k \in \seq{1}{U}$, and shared parameters are those that are not unit-specific.
The PIF algorithm treats the parameter vector $\theta = (\phi, \psi_{\seq{1}{U}})$ as a complete entity and gives no distinction between shared and unit-specific parameters.
Even though the random walk standard deviation is set to zero for unit-specific parameters when considering data from other units, this lack of distinction may result in particle depletion for unit-specific parameters as the number of units $U$ grows large.

As a solution to this issue, we propose the block panel iterated filter (BPIF).
In this algorithm, shared $\phi$ and unit-specific parameters $\psi_{\seq{1}{U}}$ are handled differently when updating the parameter vector $\theta = (\phi, \psi_{\seq{1}{U}})$.
The BPIF updates $\theta$ by updating the shared vector $\phi$ in the same way that is done with $PIF$, but the indices for weights for unit-specific parameters $\psi_{\seq{1}{U}}$ are only updated using data from the corresponding
unit.
In the case where there are many units, we believe that this modification will lead to less particle depletion.
Pseudo-code for this algorithm is given in Algorithm~\ref{alg:bpif}.
We note that no theoretical properties of this algorithm have yet been derived, but early empirical results suggest that this algorithm is a an improvement to the PIF algorithm.

\begin{algorithm}[ht]
   \caption[BPIF]{\textbf{BPIF}. \\
    {\bf Inputs}:\\\hspace{\textwidth}
    Simulator of initial density, $f_{X_{u, 0}}(x_{u, 0}; \theta)$ for $u$ in $\seq{1}{U}$.\\\hspace{\textwidth}
    Simulator of transition density, $f_{X_{u, n}|X_{u, n-1}}(x_{u, n}|x_{u, n-1}; \theta)$ for $u$ in $\seq{1}{U}$, $n$ in $\seq{1}{N_u}$.\\\hspace{\textwidth}
    Evaluator of measurement density, $f_{Y_{u, n}|X_{u, n}}(y_{u, n}|x_{u, n} ; \theta)$ for $u$ in $\seq{1}{U}$, $n$ in $\seq{1}{N_u}$.\\\hspace{\textwidth}
    Data $y_{u, n}^*$,for $u$ in $\seq{1}{U}$, $n$ in $\seq{1}{N_u}$.\\\hspace{\textwidth}
    Number of iterations, $M$.\\\hspace{\textwidth}
    Number of particles, $J$.\\\hspace{\textwidth}
    Starting parameter swarm, $\Theta_j^0 = \big(\Phi_j^0, \Psi_{1:U, j}^0\big)$ for $j \in \seq{1}{J}$, $u \in \seq{1}{U}$.\\\hspace{\textwidth}
    Simulator of perturbation density, $h_{u, n}(\theta | \phi; \sigma)$ for $u \in \seq{1}{U}$, $n \in 0:N_u$.\\\hspace{\textwidth}
    Perturbation sequence, $\sigma_m$ for $m \in 1:M$.\\\hspace{\textwidth}
    {\bf Output:} \\\hspace{\textwidth}
    Final parameter swarm, $\Theta_{j}^{m} = \big(\Phi_j^m, \Psi_{\seq{1}{U}, j}^m\big)$ for $j \in \seq{1}{J}$, $u \in \seq{1}{U}$.
    \label{alg:bpif}}
%\BlankLine
\For{$m \in 1:M$}{
  Set $\Phi_{0, j}^m = \Phi_j^{m-1}$ for $j \in \seq{1}{J}$\;
    \For{$u \in \seq{1}{U}$}{
      Set $\Phi_{u, 0, j}^{F, m} \sim h_{u, 0}(\phi | \Phi_{0, j}^m; \sigma_m)$ for $j \in \seq{1}{J}$\;
      Set $\Psi_{u, 0, j}^{F, m} \sim h_{u, 0}(\psi | \Psi_{u, j}^{m - 1}; \sigma_m)$ for $j \in \seq{1}{J}$\;
      Set $\Theta_{u, 0, j}^{F, m} = \big(\Phi_{u, 0, j}^{F, m}, \Psi_{u, 0, j}^{F, m}\big)$ for $j \in \seq{1}{J}$\;
      Initialize $X_{u, 0, j}^{F, m} \sim f_{X_{u, 0}}(x_{u, 0}; \Theta_{u, 0, j}^{F, m})$ for $j \in \seq{1}{J}$\;
        \For{$n \in \seq{1}{N_u}$} {
          Set $\Theta_{u, n, j}^{P, m} \sim h_{u, n}(\theta|\Theta_{u, n-1, j}^{F, m}, \sigma_m)$ for $j \in \seq{1}{J}$\;
          $X_{u, n, j}^{P, m} \sim f_{X_{u, n}|X_{u, n-1}}(x_{u, n}|X_{u, n-1, j}^{F, m}; \Theta_{u, n, j}^{P, m})$ for $j \in \seq{1}{J}$\;
          $w_{u, n, j}^m = f_{Y_{u, n}|X_{u, n}}(y_{u, n}^*|X_{u, n, j}^{P, m})$ for $j \in \seq{1}{J}$\;
          Draw $k_{\seq{1}{j}}$ with $P(k_j = i) = w_{u, n, i}^m / \sum_{v = 1}^J w_{u, n, v}^m$ for $i, j \in \seq{1}{J}$\;
          Set $\Theta_{u, n, j}^{F, m} = \Theta_{u, n, k_j}^{P, m} = \big(\Phi_{u, n, k_j}^{P, m}, \Psi_{u, n, k_j}^{P, m}\big)$ and $X_{u, n, j}^{F, m} = X_{u, n, k_j}^{P, m}$ for $j \in \seq{1}{J}$\;
        }
      Set $\Theta_{u, j}^m = \big(\Phi_{u, N_u, j}^{F, m}, \Psi_{u, N_u, j}^{F, m}\big)$ for $j \in \seq{1}{J}$\;
    }
  Set $\Theta_j^m = \big(\Phi_{U, j}^m, \Psi_{1:U, j}^m\big)$ for $j \in \seq{1}{J}$\;
}
\end{algorithm}

\section{Translating to \citet{lee20} notation}

Since the models in \citet{lee20} were developed independently, the choice of notation varies inconsistently between models.
For our reanalysis, we rename parameters to provide a unified notation facilitating comparison between models.
Table~\ref{tab:translate} maps this notation back to the original notations, for reference.

\begin{table}
  \begin{center}
  \begin{tabular}{|c|c|c|c|c|}\hline
    \multirow{2}{*}{Parameter} & Our & \multicolumn{3}{c|}{Lee et al. (2020a)} \\\cline{3-5}
     & Notation & 1 & 2 & 3 \\
    \hline
    \hline
    % Table begins here
    Reporting Rate & $\reportRate$ & $\rho$ & $\rho$ & $\epsilon_1, \epsilon_2$ \\\hline
    Mixing Coefficient & $\mixExponent$ & $\nu$ & --- & --- \\\hline
    Measurement Over-Dispersion & $\obsOverdispersion$ & $\tau$ & --- & $p$ \\\hline
    Birth Rate & $\muBirth$ & $\mu$ & --- & --- \\\hline
    Natural Mortality Rate & $\muDeath$ & $\delta$ & --- & $\mu$ \\\hline
    Cholera Mortality Rate & $\choleraDeath$ & --- & --- & $\alpha$ \\\hline
    Latent Period & $1/\muEI$ & $1/\sigma$ & $1/\gamma_E$ & --- \\\hline
    Recovery Rate & $\muIR$ & $\gamma$ & $\gamma$ & $\gamma$ \\\hline
    Loss of Immunity & $\muRS$ & $\alpha$ & $\sigma$ & $\rho$ \\\hline
    Symptomatic Ratio & $\symptomFrac$ & $1 - \theta_0$ & $k$ & $\sigma$ \\\hline
    Asymptomatic Relative Infectiousness & $\asymptomRelativeInfect$ & $\kappa$ & $red_\beta$ & --- \\\hline
    Human-to-Water Shedding & $\Wshed$ & --- & $\mu$ & $\theta_I$ \\\hline
    Asymptomatic Relative Shedding & $\asymptomRelativeShed$ & --- & $red_\mu$ & $\theta_A/\theta_I$ \\\hline
    Seasonal Amplitude & $\seasAmplitude$ & --- & $\alpha_s$ & $\lambda$ \\\hline
    Transmission & $\transmission$ & $\beta$ & $\beta$ & $c$ \\\hline
    Water-to-Human & $\beta_W$ & --- & $\beta_W$ & $\beta$ \\\hline
    Bacteria Mortality & $\Wremoval$ & --- & $\delta$ & $\mu_\beta$ \\\hline
    Vaccination Efficacy & $\vaccineEfficacy$ & $\theta_{vk}$ & $\theta_1, \theta_2, \theta_{1_5}, \theta_{2_5}$ & $\eta_{1d}, \eta_{2d}$ \\\hline
    Process Over-dispersion & $\sigmaProc$ & --- & --- & $\sigma_w$\\\hline
    % Table ends here
  \end{tabular}
  \end{center}
  \caption{
  \label{tab:translate}Translations between our common notation and notation used in \cite{lee20}\eic{TODO: half saturation constant, 2 and 3. Fraction of susceptible, 2.
  Initial Value parameters.}
  }
\end{table}

\section{Lee et al. (2020) Replication}

In this article we claimed that we were able to obtained better fits to the observed data using the same models that were proposed in \cite{lee20}.
Along with visual comparisons to the data, this claim was supported by comparing likelihoods and AIC values in Table 1.
Because model likelihoods were not provided in \cite{lee20}, it is necessary to replicate these models in order to obtain likelihood estimates.
In the following subsections, we use our \code{R} package \code{haitipkg} to reproduce some of the results of \cite{lee20}.
This reproduction allows us to estimate the likelihoods of the \cite{lee20} version of Models~1--3, and also provides a demonstration on reproducibility.

\subsection{Model~1 Replication}

The model was implemented by a team at Johns Hopkins Bloomberg School of Public Health (hereafter referred to as the Model~1 authors) in the \code{R} programming language using the \code{pomp} package \citep{king16}.
Original source code is publicly available with DOI: 10.5281/zenodo.3360991.
Despite having source code available, no point estimates for model parameters were provided in \cite{lee20}.
According to the supplement materials, this was because model realizations from a single parameter set retained substantial variability, but multiple realizations from a collection of parameter sets resulted in a reasonable visual fit to the data \jwc{REF???}.
We are also inclined to believe that the use of multiple parameter values was in part intended to account for parameter uncertainty (as mentioned in our main text), an effort by the Model~1 authors that we applaud.
We note, however, that simulations from each of the parameter sets were treated with equal importance when being used to diagnose the model fit and make inference on the system.
This practice is problematic given Figures S8 and S9 of the supplement material \jwc{REF???}, which suggest that some parameter sets that were used for inference were several hundred log-likelihood units lower than the best performing sets of parameters.
When accounting for parameter uncertainty, one should instead weigh resulting model projections based on the likelihood of the set of parameters that were used to obtain the model realizations, as was done in \cite{king15}.

Not providing a point estimate for model parameters also has the effect of making model reproduction a more difficult task.
Fortunately, the range and median values for each of the final model parameters are provided in Tables S10 and S11 of the supplement, and bivariate relationships between some of the fitted parameters are displayed in Figures S8 and S9.
We therefore use these basic summaries to reconstruct an approximation to the set of parameters that were used by the Model~1 authors to perform inference.

We begin by constructing 300 different parameter sets based on the basic summaries and figures provided by the Model~1 authors.
This was primarily done using truncated-bivariate normal distributions, where samples that are drawn outside of the desired range of parameters are re-sampled from the distribution.
Because the likelihood surface is complex and almost certainly includes higher than two-level interactions between parameter values, we then perform iterated filtering for the epidemic phase of the data using the 300 different parameter sets as starting points.

<<sampleBetaNu, echo=FALSE, cache=TRUE>>=
set.seed(141837)
sample_beta_nu <- function(mu_beta = 5.8, mu_nu = 0.96) {
  # This function is used to sample starting values for beta and nu.
  #
  # params:
  #   mu_beta: Average of the marginal distribution for beta1. Value was fit visually.
  #   mu_nu: Average of the marginal distribution for nu. Value was fit visually.
  # returns:
  #   data.frame of size 300x2, values of beta1 and nu with truncated bivariate
  #   normal relation.

  # Determine a good value for the marginal standard deviations
  sd_beta <- (12.1-mu_beta) / qnorm(0.995)
  sd_nu <- (0.9-mu_nu) / qnorm(0.005)

  # Create covariance matrix for the bivariate-normal distribution
  cov_mat <- rbind(
    c(sd_beta^2, -0.7 * sd_beta * sd_nu),
    c(-0.7 * sd_beta * sd_nu, sd_nu^2)
  )

  # Sample from the bivariate normal distribution
  X <- MASS::mvrnorm(n = 300, mu = c(mu_beta, mu_nu), Sigma = cov_mat)
  colnames(X) <- c("beta1", "nu")

  # Identify values that fell outside of range, so they can be resampled
  bad_beta <- which(X[, 'beta1'] > 12.1 | X[, 'beta1'] < 0.04)
  bad_nu <- which(X[, 'nu'] > 1 | X[, 'nu'] < 0.9)
  n_resamp <- length(unique(c(bad_beta, bad_nu)))

  # Keep resampling values until all values fall within the range.
  while (n_resamp > 0) {
    to_resamp <- unique(c(bad_beta, bad_nu))
    X[to_resamp, ] <- MASS::mvrnorm(n = n_resamp, mu = c(mu_beta, mu_nu), Sigma = cov_mat)

    bad_beta <- which(X[, 'beta1'] > 12.1 | X[, 'beta1'] < 0.04)
    bad_nu <- which(X[, 'nu'] > 1 | X[, 'nu'] < 0.9)
    n_resamp <- length(unique(c(bad_beta, bad_nu)))
  }

  # Return results as a data.frame
  as.data.frame(X)
}

# Run the function to get values of beta1 and nu
starting_params <- sample_beta_nu()
@

<<sample_rho, echo=FALSE, cache=TRUE>>=
sample_rho <- function(beta) {
  # This function is used to get starting values for the parameter rho, based
  # on it's strong correlation with beta1 and nu. Here, we first sample beta1
  # and nu, and then sample rho based on a regression on beta1. Additionally,
  # it is noted that rho has a bi-modal distribution, with lower values of rho
  # corresponding with high probability to higher values of beta1.
  #
  # params:
  #   beta: previously sampled values of beta1 that will be used to get values
  #         of rho.
  # returns:
  #   vector of length 300 that have an appropriate relationship with the parameter
  #   beta1.

  rhos <- rep(0, 300)

  # Get slope and intercept for the regression
  slope <- (0.4/(5.1 - 6.75))
  intercpt <- -6.75 * slope

  # Determine whice mode should the value of rho belong to, based on regression
  case_when(
    beta > 6.75 ~ 0,  # Large beta --> group "0"
    # Small beta --> more likely to be group "1", with increasing probability as beta gets smaller
    beta <= 6.75 & beta > (0.95 - intercpt) / slope ~ slope * beta + intercpt,
    TRUE ~ 0.95
  ) -> probs

  # Sample the groups that rho belongs to
  rho_group1 <- rbernoulli(300, probs)
  rho_group2 <- !rho_group1

  # Keep track of how many values are in each group, used later for resampling
  n_group1 <- sum(rho_group1)
  n_group2 <- sum(rho_group2)

  # Resample values of rho that are larger than 1
  group1 <- rnorm(n_group1, mean = 1, sd = 0.05)
  n_bad1 <- sum(group1 > 1)
  while(n_bad1 > 0) {
    group1[group1 > 1] <- rnorm(n_bad1, mean = 1, sd = 0.175)
    n_bad1 <- sum(group1 > 1)
  }

  # get standard deviation for regression
  std_beta1 <- (beta[rho_group2] - mean(beta[rho_group2])) / sd(beta[rho_group2])

  group2 <- 0.25 - 0.025 * std_beta1 - (rlnorm(n_group2, sdlog = 0.2) - 1)

  # Resample values of rho in group2 as needed
  which_bad2 <- group2 < .13
  while(sum(which_bad2) > 0) {
    group2[group2 < 0.13] <- 0.25 - 0.1 * std_beta1[which_bad2] - (rlnorm(sum(which_bad2), sdlog = 0.2) - 1)

    which_bad2 <- group2 < .13
  }

  rhos[rho_group1] <- group1
  rhos[rho_group2] <- group2

  # Return the values of rho
  rhos
}

# Sample rho, using already sampled beta1.
starting_params$rho <- sample_rho(starting_params$beta1)
@

<<sample_tau, echo=FALSE, cache=TRUE>>=
sample_tau <- function(mu = 4.1) {
  # Function used to sample tau. Tau is approximately independent of the other
  # parameters.
  #
  # params:
  #   mu: mean of the truncated normal distribution
  # returns:
  #   vector of length 300, sampled values of tau from truncated normal distribution

  # Sample from normal distribution
  X <- rnorm(300, mean = mu, sd = (1.7-mu) / qnorm(0.001))

  # Get how many resamples are needed
  n_resamp <- sum(X > 5.9 | X < 1.7)

  # Keep resampling until all values fall within desired range
  while (n_resamp > 0) {
    X[X > 5.9 | X < 1.7] <- rnorm(n_resamp, mean = mu, sd = (1.7-mu) / qnorm(0.001))

    n_resamp <- sum(X > 5.9 | X < 1.7)
  }

  # Return values of tau
  X
}

# Use function to sample tau
starting_params$tau <- sample_tau()
@

<<initialValueParams, echo=FALSE, cache=TRUE>>=

# Here we note that there are additional parameters that were fit, but their
# relationship between the remaining parameters is not given in Lee et al. (2020).
# Because of this, we simply sample these parameters using a truncated normal distribution:

POP <- 10911819

E_0_probs <- dnorm(seq(1, 1752, 1), mean = 400, sd = (1752-400) / qnorm(0.99))
I_0_probs <- dnorm(seq(1, 2271, 1), mean = 500, sd = (2271-500) / qnorm(0.99))

starting_params$E_0 <- sample(1:1752, size = 300, replace = TRUE, prob = E_0_probs) / POP
starting_params$I_0 <- sample(1:2271, size = 300, replace = TRUE, prob = I_0_probs) / POP

starting_params$S_0 <- 1 - starting_params$E_0 - starting_params$I_0
@

<<RemainingParams, echo=FALSE, cache=FALSE>>=
h1_epi <- haiti1()

bounds <- tribble(
  ~param, ~lower, ~upper,
  "beta2", 1e-08, 10,
  "beta3", 1e-08, 10,
  "beta4", 1e-08, 10,
  "beta5", 1e-08, 10,
  "beta6", 1e-08, 10
)

lower <- bounds$lower
names(lower) <- bounds$param

upper <- bounds$upper
names(upper) <- bounds$param

betas <- sobol_design(
  lower = lower,
  upper = upper,
  nseq  = 300
)

epi_params <- cbind(starting_params, betas)

not_in_start_params <- names(coef(h1_epi))[!names(coef(h1_epi)) %in% colnames(epi_params)]

for (i in 1:length(not_in_start_params)) {
  epi_params[, not_in_start_params[i]] <- coef(h1_epi)[not_in_start_params[i]]
}

epi_params$sig_sq <- 0

# simply re-ordering to match model order
epi_params <- epi_params[, names(coef(h1_epi))]

epi_rw <- rw.sd(
  # Parameters close to desired values
  beta1 = 0.001,
  tau   = 0.001,
  nu    = 0.001,
  rho   = 0.001,
  # Parameters not close to desired values
  beta2 = 0.02,
  beta3 = 0.02,
  beta4 = 0.02,
  beta5 = 0.02,
  beta6 = 0.02,
  E_0   = ivp(0.1),
  I_0   = ivp(0.1)
)

epi_rw <- rw.sd(
  # Parameters close to desired values
  beta1 = 0.005,
  tau   = 0.005,
  nu    = 0.005,
  rho   = 0.005,
  # Parameters not close to desired values
  beta2 = 0.02,
  beta3 = 0.02,
  beta4 = 0.02,
  beta5 = 0.02,
  beta6 = 0.02,
  E_0   = ivp(0.1),
  I_0   = ivp(0.1)
)

registerDoParallel(20)
registerDoRNG(48963587)

# Local MIF near MLE
stew(
  file = "output/reprod_epi_MIF2.rda",
  {
    # Keep track of how long it takes
    tLocal <- system.time({
      foreach(
        i = 1:nrow(epi_params),
        .packages = c('pomp'),
        .combine = c
      ) %dopar% {

        r_params <- unlist(epi_params[i, ])
        coef(h1_epi) <- r_params
        mif2(
          h1_epi,
          Np = 100,
          Nmif = 50,
          cooling.fraction.50 = 0.5,  # Cooling set so that we will reach smaller rw.sd at the end than the end of the unit3 search.
          rw.sd = epi_rw
        ) -> m2
      } -> MIF2_search
    })
  }
)
@

<<PlotEpiDist, echo=FALSE, fig.cap="Compare to Figure S8 of Lee et al. (2020) supplement.", fig.width=5.1, fig.height=5.1, fig.align='center'>>=
epi_results <- as.data.frame(t(coef(MIF2_search)))
epi_results$logLik <- logLik(MIF2_search)

GGally::ggpairs(
  filter(epi_results, logLik > -2200),
  columns = c("rho", "tau", "beta1", "nu", "logLik")
)
@

<<SimulateEpi, echo=FALSE, include=FALSE, message=FALSE, cache=TRUE>>=
epi_complete <- filter(epi_results, logLik > -2200)

# First, we need to simulate the epidemic model using the fitted parameters.
registerDoParallel(20)
registerDoRNG(18599687)
foreach(
  i = 1:nrow(epi_complete),
  .combine = rbind
) %dopar% {
  guess <- unlist(epi_complete[i, ])
  sims <- simulate(h1_epi, nsim = 20, format = 'data.frame',
                   params = guess)  # 20 sims for each set of parameters
  sims$param_set <- i
  sims
} -> all_sims

quants <- all_sims %>%
  select(.id, week, cases) %>%
  group_by(week) %>%
  summarize(
    q025 = quantile(cases, probs = 0.025, na.rm = TRUE),
    q50  = quantile(cases, probs = 0.500, na.rm = TRUE),
    q975 = quantile(cases, probs = 0.975, na.rm = TRUE)
  ) %>%
  ungroup() %>%
  mutate(date = lubridate::ymd("2010-10-16") + lubridate::weeks(week))

true_agg_cases <- haitiCholera %>%
  select(-report) %>%
  pivot_longer(
    data = .,
    cols = -date_saturday,
    values_to = 'cases',
    names_to = 'dep'
  ) %>%
  mutate(
    date = as.Date(date_saturday),
    dep = gsub("\\.", "_", dep)
  ) %>%
  mutate(
    dep = case_when(dep == "Grand_Anse" ~ "Grande_Anse", TRUE ~ dep)
  ) %>%
  tidyr::pivot_wider(
    data = .,
    id_cols = c(date),
    names_from = dep,
    values_from = cases,
    names_prefix = 'cases_'
  ) %>%
  mutate(
    ReportedAll = cases_Artibonite + cases_Centre +
      cases_Grande_Anse + cases_Nippes + cases_Nord +
      cases_Nord_Est + cases_Ouest + cases_Sud +
      cases_Sud_Est + cases_Nord_Ouest
  )
@

Following the model fitting scheme used by the Model~1 authors, we then use these resulting parameters as starting values for the endemic phase, and refit parameters using MIF2.
The resulting bivariate relations between $\rho, \tau, \beta_1$ and $\nu$ are given in Figure TODO.

<<FitEnd, echo=FALSE, message=FALSE, include=FALSE>>=
h1_end <- haiti1(period = "endemic")
start_states <- all_sims %>% filter(week == max(week))

end_rw <- rw.sd(
  beta1 = 0.05,
  tau   = 0.005,
  nu    = 0.04,
  rho   = 0.01,
  beta2 = 0.05,
  beta3 = 0.05,
  beta4 = 0.05,
  beta5 = 0.05,
  beta6 = 0.05
)

registerDoParallel(20)
registerDoRNG(21489587)

end_start <- epi_complete %>%
  select(-logLik)

# Local MIF near MLE
stew(
  file = "output/reprod_end_MIF2.rda",
  {
    foreach(
      i = 1:nrow(end_start),
      .packages = c('pomp'),
      .combine = c
    ) %dopar% {

      output <- list()
      r_params <- unlist(end_start[i, ])

      start_states %>%
        filter(param_set == i) %>%
        select(-.id, -week, -param_set) %>%
        sapply(., median, na.rm = TRUE) -> rinit_parms

      # rinit_parms["S"] <- POP - (rinit_parms["E"] + rinit_parms["I"] + rinit_parms["A"] + rinit_parms["R"])

      # Set the inital state
      new_rinit <- Csnippet(
        sprintf(
          "
          S = nearbyint(%f);
          E = nearbyint(%f);
          I = nearbyint(%f);
          A = nearbyint(%f);
          R = nearbyint(%f);
          incid = nearbyint(%f);
          foival = %f;
          Str0 = nearbyint(%f);
          Sout = nearbyint(%f);
          Sin = nearbyint(%f);
          ",
          rinit_parms['S'],
          rinit_parms['E'],
          rinit_parms['I'],
          rinit_parms['A'],
          rinit_parms['R'],
          rinit_parms['incid'],
          rinit_parms['foival'],
          rinit_parms['Str0'],
          rinit_parms['Sout'],
          rinit_parms['Sin']
        )
      )

      h1_end <- h1_end %>% pomp(
        rinit = new_rinit,
        statenames = c("S", "E", "I", "A", "R", "incid", "foival", "Str0", "Sout", "Sin")
      )

      coef(h1_end) <- r_params
      mif2(
        h1_end,
        Np = 100,
        Nmif = 100,
        cooling.type = "hyperbolic",
        cooling.fraction.50 = 0.05,  # Cooling set so that we will reach smaller rw.sd at the end than the end of the unit3 search.
        rw.sd = epi_rw
      ) -> m2

      output[['m2']] <- m2

      coef(h1_end) <- coef(m2)
      sims <- simulate(
        h1_end, nsim = 20, format = 'data.frame',
        params = coef(m2)
      )
      sims$param_set <- i

      output[['sims']] <- sims
      output
    } -> MIF2_search_end
  }
)
@

<<plotEndParams, fig.width=5.1, fig.height=5.1, fig.align='center', fig.cap="Bivariate relationships between variables after fitting endemic period. Compare to S9 of Lee et al. (2020) supplement.", cache=TRUE, echo=FALSE>>=
MIF_params <- matrix(nrow = length(MIF2_search_end) / 2, ncol = length(coef(MIF2_search_end[[1]])))
all_sims_end <- data.frame()
liks <- rep(0, nrow(end_start))
for (i in 1:nrow(end_start)) {
  MIF_params[i, ] <- coef(MIF2_search_end[[(2 * i) - 1]])
  all_sims_end <- rbind(all_sims_end, MIF2_search_end[[(2 * i)]])
  liks[i] <- logLik(MIF2_search_end[[(2 * i) - 1]])
}

colnames(MIF_params) <- names(coef(MIF2_search_end[[1]]))
end_results <- as.data.frame(MIF_params)
end_results$logLik <- liks

GGally::ggpairs(
  filter(end_results, logLik > -2200),
  columns = c("rho", "tau", "beta1", "nu", "logLik")
)
@

Now that we have fit the parameters, we can simulate from the models.

<<plotMod1Sims, echo=FALSE, fig.cap="Simulations from various parameter sets. Simulations from parameter sets with likelihood $<$ -2200 are removed.", cache=TRUE, fig.width=4.5, fig.height=4.25>>=
end_complete <- filter(end_results, logLik > -2200)

bad_end_pars <- which(end_results$logLik < -2200)

quants_end <- all_sims_end %>%
  filter(!param_set %in% bad_end_pars) %>%
  select(.id, week, cases) %>%
  group_by(week) %>%
  summarize(
    q025 = quantile(cases, probs = 0.025, na.rm = TRUE),
    q50  = quantile(cases, probs = 0.500, na.rm = TRUE),
    q975 = quantile(cases, probs = 0.975, na.rm = TRUE)
  ) %>%
  ungroup() %>%
  mutate(date = lubridate::ymd("2010-10-16") + lubridate::weeks(week))

ggplot() +
  geom_point(data = true_agg_cases, aes(x = date, y = ReportedAll)) +
  geom_line(data = quants, aes(x = date, y = q50), col = '#45ac5f') +
  geom_ribbon(data = quants, aes(x = date, ymin = q025, ymax = q975), alpha = 0.5,
              fill = '#45ac5f') +
  geom_line(data = quants_end, aes(x = date, y = q50), col = '#882871') +
  geom_ribbon(data = quants_end, aes(x = date, ymin = q025, ymax = q975), alpha = 0.5, fill = '#882871') +
  ylab("Reported, Symptomatic Incidence") +
  scale_x_date(date_breaks = "1 year") +
  theme(axis.title.x = element_blank(),
        axis.text.x = element_text(hjust = 1, angle = 60))
@

\subsection{Model~2 Replication}

The source code that was used to implement Model~2 in \cite{lee20} was written in the \code{Python} programming language, and is publicly available at \url{https://github.com/lulelita/HaitiCholeraMultiModelingProject}.
In order to perform our analysis in a unified framework, we re-implemented this model in the \code{R} programming language using the \code{spatPomp} package \citep{asfaw21github}.

\subsection{Model~3 Replication}

The code that was originally used to implement Model~3 is archived with the DOI: 10.5281/zenodo.3360723, and also available in the public GitHub repository: \code{jcblemai/haiti-mass-ocv-campaign}.
Because the code was made publicly available, and final model parameters were reported in the supplementary material of \cite{lee20}, we were able to reproduce Model~3 by directly using the source code.
In Fig.~\ref{fig:mod3rep}, we plot simulations from this model.
This figure should be compared to Figure S18 of \cite{lee20}.
We note that slight differences may be accounted for due to variance in the model simulations and the difference in programming language used to produce the figure.

<<Model3Replication, echo=FALSE, cache=TRUE>>=
plot_order <- c(
  'Artibonite',
  'Sud_Est',
  'Nippes',
  'Nord_Est',
  'Ouest',
  'Centre',
  'Nord',
  'Sud',
  'Nord_Ouest',
  'Grande_Anse'
)

dep_labeller <- as_labeller(
  c(
    'Artibonite' = 'Artibonite',
    'Sud_Est' = 'Sud-Est',
    'Sud.Est' = 'Sud-Est',
    'Nippes' = 'Nippes',
    'Nord_Est' = 'Nord-Est',
    'Nord.Est' = 'Nord-Est',
    'Ouest' = 'Ouest',
    'Centre' = 'Centre',
    'Nord' = 'Nord',
    'Sud' = 'Sud',
    'Nord_Ouest' = 'Nord-Ouest',
    'Nord.Ouest' = 'Nord-Ouest',
    'Grande_Anse' = 'Grand\'Anse',
    'Grand.Anse' = 'Grand\'Anse'
  )
)

lee3 <- haiti3()
lee3_sims <- simulate(lee3, nsim = 1000, seed = 321, format = 'data.frame')

sub_cases <- haitiCholera %>%
  dplyr::mutate(date = as.Date(date_saturday)) %>%
  dplyr::select(-report, -date_saturday) %>%
  dplyr::filter(lubridate::decimal_date(date) >= min(lee3_sims$time)) %>%
  tidyr::pivot_longer(
    data = .,
    cols = -date,
    names_to = "dep",
    values_to = "cases"
  ) %>%
  dplyr::mutate(
    dep = dplyr::case_when(
      dep == "Grand.Anse" ~ "Grande_Anse",
      dep == "Nord.Est" ~ "Nord_Est",
      dep == "Nord.Ouest" ~ "Nord_Ouest",
      dep == "Sud.Est" ~ "Sud_Est",
      TRUE ~ dep
    )
  )

quants <- lee3_sims %>%
  dplyr::select(-CasesAll) %>%
  tidyr::pivot_longer(
    data = .,
    cols = starts_with("cases"),
    names_prefix = "cases",
    names_to = "dep",
    values_to = "cases"
  ) %>%
  dplyr::group_by(dep, time) %>%
  dplyr::summarise(
    q05 = quantile(cases, 0.025, na.rm = T),
    mean = mean(cases, na.rm = T),
    q50 = quantile(cases, 0.5, na.rm = T),
    q95 = quantile(cases, 0.975, na.rm = T)
  ) %>%
  dplyr::ungroup() %>%
  dplyr::mutate(
    date = lubridate::date_decimal(time),
    date = as.Date(lubridate::round_date(date))
  )

quants <- quants %>%
  dplyr::filter(
    !(dep == "Ouest" & date < '2017-06-10')
  )
@

\begin{figure}[ht]
<<PlotMod3Rep, echo=FALSE, message=FALSE, fig.height=3.75>>=
ggplot() +
  geom_ribbon(data = quants, aes(x = date, ymin = q05, ymax = q95),
              fill = 'darkblue', alpha = 0.5) +
  geom_line(data = quants, aes(x = date, y = q50), col = 'darkblue') +
  geom_point(data = dplyr::filter(sub_cases, cases < 500), aes(x = date, y = cases), size = 0.2) +
  facet_wrap(~factor(dep, levels = plot_order), nrow = 2, labeller = dep_labeller) +
  labs(y = 'Reported Cholera Cases') +
  scale_x_date(date_breaks = "6 months", date_labels = "%Y-%m") +
  theme(
    axis.text.x = element_text(angle = 40, hjust = 1, vjust = 1, size = 7),
    axis.title.x = element_blank()
  )
@
\caption{\label{fig:mod3rep}
Simulations from Model~3.
Compare to Figure S18 of \cite{lee20}.
}
\end{figure}

<<lee3Cleanup, echo=FALSE, include=FALSE, message=FALSE>>=
rm(lee3, lee3_sims, quants, sub_cases)
gc()
@

% \renewcommand\missing{---}
%
% \begin{table}
% \begin{center}
% \begin{tabular}{|l|c|c|c|c|}
% \hline
% Mechanism & Unified & Old model 1 & Old model 2 & Old model 3
% \\
% \hline
% \hline
% Recovery rate
%   & $\muIR$
%   & $\gamma$
%   & $\gamma$
%   & $\gamma$
% \\
% Latency
%   & $\muEI$
%   & $\sigma$
%   & $\sigma$
%   & \missing
% \\
% Direct transmission
%   & $\transmission$
%   & $\beta_{1:6}$
%   & $\beta$
%   & $c^i$
% \\
% Rainfall exponent
%   & \rainfallExponent
%   & \missing
%   & \missing
%   & $r$
% \\
% Lost immunity
%   & $\muRS$
%   & $\alpha$
%   & $\alpha$
%   & $\rho$
% \\
%   & $\omega_1$, $\omega_2$
%   & \missing
%   & $\omega_1$, $\omega_2$
%   & \missing
% \\
% Vaccine efficacy
%   & $\vaccineEfficacy_{1:2}$
%   & \missing
%   & $\theta_{1:2}$
%   & \missing
% \\
% Birth rate
%   & $\muBirth$
%   & $\mu$
%   &
%   & $\mu$
% \\
% Death rate
%   & $\muDeath$
%   & $\delta$
%   &
%   & $\mu$
% \\
% Symptomatic frac.
%   & $\symptomFrac$
%   & $\theta_k(t)=c\theta^*(t-\tau_k)$
%   & $k$
%   & $\theta$
% \\
% Asympt.~infectivity
%   & $\asymptomRelativeInfect$
%   & $(1-\kappa)$
%   & $\beta_A/\beta$
%   & 1
% \\
% Asympt.~shedding
%   & $\asymptomRelativeShed$
%   & \missing
%   & $\mu_A/\mu$
%   & $\theta_A/\theta_I$
% \\
% Seasonality amplitude
%   & $\seasAmplitude$
%   & \missing
%   & $\alpha_s$
%   & $a$
% \\
% Water to human
%   & $\Wbeta{}$
%   & \missing
%   & $\beta_W$
%   & $\beta^i$
% \\
%   & $\Wsat$
%   & \missing
%   & $\mathrm{sat}$
%   & $1$
% \\
% Cholera decay in water
%   & $\Wremoval$
%   & \missing
%   & $\delta$
%   & $\mu_B$
% \\
% Human to water
%   & $\Wshed$
%   & \missing
%   & $\mu$
%   & $\theta_I$
% \\
% Mixing exponent
%   & $\mixExponent$
%   & $\nu$
%   & \missing
%   & \missing
% \\
% Process noise
%   & $\sigmaProc$
%   & \missing
%   & \missing
%   & $\sigma$
% \\
% Reporting rate
%   & $\reportRate$
%   & $\rho$
%   & $\rho$
%   & $\epsilon$
% \\
% Observation overdispersion
%   & $\obsOverdispersion$
%   & $\tau$
%   & \missing
%   & $k$
% \\
% \hline
% \end{tabular}
% \end{center}
% \caption{Translation from the unified notation of this article back into the notations used for Lee et al.~(2020)}
% \label{tab:translate}
% \end{table}

\section{Investigation into model 2}

As part of our case study into the Haiti paper, model 2 was recreated using the spatPomp package in R. To confirm the validity of our model, we recreated Figure S15 from the supplement. In the original model, the authors reinitialized the state values for the endemic phase starting on March 2014. In particular, the group assumed that 75\% of the population was susceptible to cholera. We believe this choice to be unnatural. Having the state values change in the middle of a simulation has no causal interpretation. Instead, the most logical thing to do when fitting an epidemic and endemic model would be to carry over the state values from the former to the latter. This keeps the integrity of the model intact.

In Figure S-2, we created the same endemic model with the only difference being the state values are carried over from the end of the epidemic phase. Here, we see that the model predicts a much different trajectory than in Figure S-1. This suggests that the decision to reinitialize state values in the middle of a simulation, as in the original model, is likely a poor choice since it has no cause and does not reflect the nature of the epidemic. One consideration is the choice of measurement component. In the original model, we use a linear measurement component. To avoid a situation where the trajectory goes to zero, we can implement a logarithmic measurement component. This is a more natural measure since the fit should be relative to the magnitude of the reported data.

<<cache=T, echo=F, fig.cap="Recreation of model 2 using the pomp package.", fig.height=4>>=
library(haitipkg)
library("pomp")
library("dplyr")
library("grid")
library("gridExtra")
library("lattice")
theme_set(theme_bw())

h2_data <- haiti2_data()
h2_epi <- haiti2(region="before")
h2_end <- haiti2(region="after")
p2_epi <- coef(h2_epi)

# temporary fix. need to figure out why i cant install panelpomp
p2_end <- coef(h2_end)
p2_end[c("VR","Mu","v")] <- c(1.304e-9, 1.897e2, 8.201e1)

traj1 <- trajectory(h2_epi,params=p2_epi,format="data.frame")
traj1$Ctotal <- rowSums(traj1[,paste0("C",1:10)])*0.2
traj2 <- trajectory(h2_end,params=p2_end,format="data.frame")
traj2$Ctotal <- rowSums(traj2[,paste0("C",1:10)])*0.2

traj_agg <- rbind(traj1[,c("year","Ctotal")], traj2[,c("year","Ctotal")])
names(traj_agg) <- c("year","cases")
traj_agg$group <- "trajectory"

haiti_agg <- aggregate(h2_data$cases,
                       by = list(h2_data$year),
                       FUN = sum)
names(haiti_agg) <- c("year","cases")
haiti_agg$group <- "reported data"

final <- rbind(traj_agg, haiti_agg)

pp1 <- ggplot(data=final %>% filter(year<2014.161),
             aes(x=year,y=cases,group=group)) +
  geom_line(aes(linetype=group,color=group)) +
  scale_linetype_manual(values=c("dashed","solid")) +
  scale_x_continuous(breaks=seq(2010,2020,by=1)) +
  theme(legend.position=c(0.7,0.7))

pp2 <- ggplot(data=final %>% filter(year>2014.161),
             aes(x=year,y=cases,group=group)) +
  geom_line(aes(linetype=group,color=group)) +
  scale_linetype_manual(values=c("dashed","solid")) +
  scale_x_continuous(breaks=seq(2010,2020,by=1)) +
  theme(legend.position=c(0.7,0.7)) +
  ylim(0,7000)

grid.arrange(pp1,pp2)
@

<<cache=T, echo=F, fig.cap="Endemic trajectory with state values carried over from epidemic model.", fig.height=2>>=
library("subplex")
theme_set(theme_bw())

h2_end_joint <- haiti2(region="after",joint=TRUE) %>% pomp()
sim <- trajectory(h2_epi,params=p2_epi,format="data.frame")
end_IVPS <- sim[nrow(sim),-c(321,322)] %>% as.numeric()
names(end_IVPS) <- names(h2_end_joint@params[24:343])

joint_params <- c(p2_end[1:23],end_IVPS)
ofun <- traj_objfun(h2_end_joint,est=c("Mu","v","VR"),
                     params=joint_params)
theta <- c(Mu=1e3,v=100,VR=2e-11)
fit <- subplex(par=log(theta), fn=ofun)
joint_params[c("Mu","v","VR")] <- exp(fit$par)

traj3 <- trajectory(h2_end_joint,params=joint_params, format="data.frame")
traj3$Ctotal <- rowSums(traj3[,paste0("C",1:10)])*joint_params["Rho"]

traj_agg <- rbind(traj3[,c("year","Ctotal")])
names(traj_agg) <- c("year","cases")
traj_agg$group <- "trajectory"
final <- rbind(traj_agg, haiti_agg)

pp3 <- ggplot(data=final %>% filter(year>2014.161),
             aes(x=year,y=cases,group=group)) +
  geom_line(aes(linetype=group,color=group)) +
  scale_linetype_manual(values=c("dashed","solid")) +
  scale_x_continuous(breaks=seq(2010,2020,by=1)) +
  theme(legend.position=c(0.7,0.7)) +
  ylim(0,7000)

pp3
@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\clearpage

\bibliography{../bib-haiti}

\end{document}

### 2022-09-01 si/si.Rnw
Major updates to the supplemental text

@@
 
-\input{header-ms.tex}
+%%%% parameters %%%%%%%%%%5
+\newcommand\Wsat{W_{\mathrm{sat}}}
+\newcommand\muIR{\mu_{IR}}
+\newcommand\muEI{\mu_{EI}}
+\newcommand\transmission{\beta}
+\newcommand\seasAmplitude{a}
+\newcommand\rainfallExponent{r}
+\newcommand\muRS{\mu_{RS}}
+\newcommand\vaccineEfficacy{\theta}
+\newcommand\muBirth{\mu_S}
+\newcommand\muDeath{\delta}
+\newcommand\choleraDeath{\delta_{C}}
+\newcommand\symptomFrac{f}
+\newcommand\asymptomRelativeInfect{\epsilon}
+\newcommand\asymptomRelativeShed{\epsilon_{W}}
+\newcommand\Wbeta[1]{\beta_{W#1}}
+\newcommand\Wremoval{\delta_W}
+\newcommand\Wshed{\mu_W}
+\newcommand\mixExponent{\nu}
+\newcommand\sigmaProc{\sigma_{\mathrm{proc}}}
+\newcommand\reportRate{\rho}
+\newcommand\obsOverdispersion{\psi}
+\newcommand\phaseParm{\phi}
+\newcommand\transmissionTrend{\zeta}
+\newcommand\vaccClass{Z}
+\newcommand\vaccCounter{z}
+\newcommand\modelCounter{m}
+
+\newcommand\missing{\hspace{6mm} ---}
+\newcommand\fixed{\color{blue}}
+\newcommand\demography{D}
+
+\newcommand\mytitle{Informing policy via dynamic models: Eliminating cholera in Haiti}
+\newcommand\comma{{\hspace{-0.25mm},\hspace{-0.2mm}}}
+
+%%%%%%%%%%%%%%%% START OF USER DEFINITIONS %%%%%%%%%%%%%%%5
+
+\newcommand\siOnly[1]{} %% redefined for the supplement
+\newcommand\msOnly[1]{#1} %% redefined for the supplement
+
+\newcommand\code[1]{\texttt{#1}}
+\usepackage{xurl}
+% \usepackage[breaklinks]{hyperref}
+\usepackage{makecell}
+\usepackage{multirow}
+\usepackage{multicol}
+\usepackage[ruled,noline,linesnumbered]{algorithm2e}
+% \usepackage{setspace}
+
+%% EDITING MACROS
+\usepackage{color}
+\usepackage[normalem]{ulem}% to use \sout in feedback commands
+\usepackage{bm}
+% orange for EI
+\definecolor{orange}{rgb}{1,0.5,0}
+\newcommand\ei[2]{\sout{#1} \textcolor{orange}{#2}}
+\newcommand\eic[1]{\textcolor{orange}{[#1]}}
+% green for JW
+\definecolor{green}{rgb}{0,0.5,0}
+\newcommand\jw[2]{\sout{#1} \textcolor{green}{#2}}
+\newcommand\jwc[1]{\textcolor{green}{[#1]}}
+% purple for JJ
+\definecolor{purple}{rgb}{0.5,0,1}
+\newcommand\jj[2]{\sout{#1} \textcolor{purple}{#2}}
+\newcommand\jjc[1]{\textcolor{purple}{[#1]}}
+% cyan for AR
+\definecolor{cyan}{rgb}{0,.5,.5}
+\newcommand\ar[2]{\sout{#1} \textcolor{cyan}{#2}}
+\newcommand\arc[1]{\textcolor{cyan}{#1}}
+% light brown for KT
+\definecolor{lightbrown}{rgb}{0.5,0.5,0}
+\newcommand\kt[2]{\sout{#1} \textcolor{lightbrown}{#2}}
+\newcommand\ktc[1]{\textcolor{lightbrown}{#1}}
+
+%% basic POMP definitions
+\newcommand\Xspace{{\mathbb X}}
+\newcommand\Yspace{{\mathbb Y}}
+\newcommand\Thetaspace{\R^{\Thetadim}}
+\newcommand\Xunitspace{{\scriptsize{\mathcal X}}}
+\newcommand\Yunitspace{{\tiny{\mathcal Y}}}
+\newcommand\hatTheta{\widehat{\Theta}}
+\newcommand\Xdim{{\mathrm{dim}}(\Xspace)}
+\newcommand\Ydim{{\mathrm{dim}}(\Yspace)}
+\newcommand\Thetadim{P}
+\newcommand\thetadim{p}
+\newcommand\timeSet{{\mathbb T}}
+\newcommand\data[1]{#1^*}
+
+\newcommand\unitSpecific{\hspace{0.1mm}\mathrm{us}}
+\newcommand\shared{\hspace{0.15mm}\mathrm{sh}}
+
+% for algorithm pseudocode
+\SetKwFor{For}{for}{:}{end}
+
+%% for particle filters
+\newcommand\unit{u}
+\newcommand\altUnit{\tilde u}
+\newcommand\Unit{U}
+\newcommand\rep{i}
+\newcommand\Rep{\mathcal{I}}
+\newcommand\island{\rep}
+\newcommand\Island{\Rep}
+%\newcommand\txtisland{replicate}
+\newcommand\altIsland{\tilde i}
+\usepackage[mathscr]{euscript}
+
+\renewcommand\time{n}
+\newcommand\myvec[1]{\boldsymbol{#1}}
+\newcommand\altTime{\tilde n}
+\newcommand\Time{N}
+\newcommand\Np{J}
+\newcommand\np{j}
+\newcommand\altNp{k}
+\newcommand\altAltNp{\tilde j}
+\newcommand\resampleIndex{r}
+\newcommand\unitWeight{w}
+
+%% for all iterated filtering algorithms
+\newcommand\Nit{M}
+\newcommand\nit{m}
+
+%% customized math macros
+\newcommand\prob{\mathbb{P}}
+\newcommand\E{\mathbb{E}}
+\newcommand\dd[1]{\mathrm{d}{#1}}
+\newcommand\given{{\,\vert\,}}
+\newcommand\equals{{{\,}={\,}}}
+\newcommand\myequals{\hspace{0.5mm}{=}\hspace{0.5mm}}
+\newcommand\myto{{\;:\;}}
+\newcommand\seq[2]{{#1}\!:\!{#2}}
+\newcommand\mydot{{\,\cdot\,}}
+\newcommand\cp[2]{N_{\mathrm{#1}\mathrm{#2}}}
+\newcommand\giventh{{\hspace{0.5mm};\hspace{0.5mm}}}
+%\newcommand\normal{{\mathrm{Normal}}}
+\newcommand\normal{\mathcal{N}}
+\newcommand\argequals{{\,=\,}}
+\newcommand\lags{c}
+\newcommand\maxlag{\overline{c}}
+\newcommand\nlfList{C}
+\newcommand\bigO{\mathcal{O}}
+\newcommand\loglik{\lambda}
+\newcommand\loglikMC{\MC{\loglik}}
+\newcommand\R{\mathbb{R}}
+\newcommand\param{\,;}
+\newcommand\mycolon{{\hspace{0.6mm}:\hspace{0.6mm}}}
+\newcommand\MC[1]{#1^{\,\mbox{\tiny MC}}}
+%\newcommand\EMC{\MC{\E}}
+\newcommand\EMC{{\E}}
+\newcommand\Var{\mathrm{Var}}
+\newcommand\var{\Var}
+\newcommand\Cov{\mathrm{Cov}}
+\newcommand\cov{\Cov}
+\newcommand\iid{\mathrm{iid}}
+%\newcommand\dist{\mathrm{dist}}
+\newcommand\dist{d}
+\newcommand\transpose{\top}
+
+\newcommand\gnsep{,}
+
+\def\lik{L}
+\def\loglik{\ell}
+
+%% THEOREM-LIKE ENVIRONMENTS
+\usepackage{amsthm}
+\newtheorem{prop}{Proposition}
+\newtheorem{theorem}{Theorem}
+\newtheorem{lemma}{Lemma}
+\newtheorem{example}{Example}
+\newtheorem{remark}{Remark}
+\newtheorem{cor}{Corrolary}
+\newtheorem{fact}{Fact}
+\newtheorem{assumption}{Assumption}
+\newtheorem{assumptionB}{Assumption}
+% \newtheorem{procedure}{Procedure}
+\renewcommand\theassumption{A\arabic{assumption}}
+\renewcommand\theassumptionB{B\arabic{assumptionB}}
+
+%\usepackage{float}
+%\floatstyle{ruled}
+%\newfloat{textbox}{t}{tbx}
+%\floatname{textbox}{Box}
+%\renewcommand\thetextbox{\arabic{textbox}}
+
+%% FOR PSEUDOCODE SETUP
+\newcommand\mystretch{\rule[-2mm]{0mm}{5mm} }
+\newcommand\asp{\hspace{4mm}}
+\newcommand\codeIndent{\hspace{4mm}}
+
+%%%%%%%%%%%%%%%% END OF USER DEFINITIONSS %%%%%%%%%%%%%%%
 
@@
 \title{Supplement to ``{\mytitle}''}
-  \author{   Provisional author list\\
- Jesse Wheeler, Anna Rosengart, Zhuoxun Jiang, Kevin Hao En Tan, Noah Treutle, Edward L. Ionides
+  \author{Jesse Wheeler, Anna Rosengart, Zhuoxun Jiang, \\ Kevin Hao En Tan, Noah Treutle, Edward L. Ionides
  \\ \hspace{.2cm}\\
@@
 \end{equation}
+
+This is the the interpretation and implementation used for Model~2 in our study.
 
@@
 The infinitesimal mean of every flow is equal to its infinitesimal variance \citep{breto11} and so an SMCS is called equidispersed.
+We note that the special case of Model~1 used in \cite{lee20} (with $\sigmaProc = 0$) is an SMCS.
 To permit more general mean-variance relationships for a Markov counting system, we must permit jumps of size greater than one.
-The utility of overdispersed models, where the infinitesimal variance of the flow exceeds the infinitesimal mean, has become widely recongized \eic{REFS}.
+The utility of overdispersed models, where the infinitesimal variance of the flow exceeds the infinitesimal mean, has become widely recognized \citep{stocks20,he10}.
 
@@
 Including white noise in the rate function enables the possibility of an overdispersed Markov counting system \citep{breto11,breto09,he10}.
-Since rates should be non-negative, Gaussian noise is not appropriate and gamma noise is a convenient option that has found various applications \eic{REFS}.
+Since rates should be non-negative, Gaussian noise is not appropriate and gamma noise is a convenient option that has found various applications \citep{romero-severson15, subramanian20}.
 Specifically, we consider a model given by
@@
 The numerical solutions to the model are arguably of more direct scientific interest than the exact solutions to the postulated equations.
-For ODE models, numerical methods are well studied and a standard numerical solution package such as deSolve in R is adequate for our purposes.
+For ODE models, numerical methods are well studied and a standard numerical solution package such as \code{deSolve} in \code{R} is adequate for our purposes.
 For SMCS and ODMCS models, exact schemes are feasible when the number of events is small, which may be the case for small populations.
@@
 \end{equation}
-an independent Poisson random variable with mean $\mu_{\bullet j} \delta$.
-To address the issue of making our model consistent with total population as a known quantity, based on census data, in practice we use a deterministic specification of $\Delta N_{\bullet j}$ rather than the Poisson form in \eqref{eq:numerics6}.
-\eic{CHECK THIS. DO WE HAVE A CONSTANT POPULATION, OR A POPULATION OF DETERMINISTICALLY GROWING SIZE? DO WE USE BIRTH DATA OR TOTAL POPULATION DATA?}
-To make birth/immigration events consistent with death/emigration events and population census, we formally model the death rate as a rate of returning to the susceptible class $S$, and use external transitions from $\bullet$ into $S$ to describe only net population increase.
-\jwc{CHECK THIS}.
+an independent Poisson random variable with mean $\mu_{\bullet j} \delta$, as was done in Model~1.
+
+Another common approach is to balance the total number of flows in and out of the compartment, i.e., $\sum_{i}N_{\bullet i}(t) = \sum_{i}N_{i \bullet}(t)$, in order to make the model consistent with the known total population, as was done in Models~2 and~3.
+In this case, we formally model the death rate as a rate of returning to the susceptible class $S$, and use external transitions from $\bullet$ into $S$ to describe only net population increase.
 
@@
 For each of the models considered in this analysis, the initial conditions are derived by enforcing the model dynamics on reported cholera cases.
-For example, in a deterministic model, the total number of Infected and Asymptomatic individuals may be calculated using the observed number of cholera cases at time time of model initialization based on model parameters such as reporting rate $\reportRate$ and fraction of symptomatic cases $\symptomFrac$.
 It is also sometimes necessary to fit some initial value parameters in order to help determine starting values for weakly identifiable compartments.
@@
 
-For this model, the number of individuals in the Recovered and Asymptomatic compartments are set to zero, but the initial proportion of Infected and Exposed individuals is estimated as initial value parameters ($I_0$ and $E_0$, respectively) using the MIF2 algorithm.
-Finally, the initial proportion of Susceptible individuals $S_0$ is calculated as $S_0 = 1 - I_0 - E_0$.
+For this model, the number of individuals in the Recovered and Asymptomatic compartments are set to zero, but the initial proportion of Infected and Exposed individuals is estimated as initial value parameters ($I(0)$ and $E(0)$, respectively) using the MIF2 algorithm.
+Finally, the initial proportion of Susceptible individuals $S_0$ is calculated as $S(0) = 1 - I(0) - E(0)$.
+
+\subsection{Model~2}
+
+Model~2 assumes that the initial vales are constant, and no parameters need to be estimated.
+Starting value for latent state compartments are chosen so as to enforce the model dynamics on the observed number of cases.
+Specifically, the model sets $I_{u}(0) = y^*_{u}(0) / \reportRate$ for each unit $u \in 1:10$, where $I_{u}(t)$ is the number of infected individuals in unit $u$ at time $t$, $y^*_{u}(t)$ is the reported number of cases, and $\reportRate$ is the reporting rate.
 
@@
 
-We use the reported cases at the start of the pandemic to approximate the number of Asymptomatic, Infectious, and Recovered individuals in each department $u \in 1:U$ using the same approximation as provided in Eq.~\myeqref{eq:model3:approx}.
+We use the reported cases at the start of the pandemic to approximate the number of Asymptomatic, Infectious, and Recovered individuals in each department $u \in 1:U$ using the same approximation as provided in Eq.~34.
+% TODO: make sure it aligns with \myeqref{eq:model3:approx}.
 The susceptible compartment is initialized so that the sum $S_{u}(0) + I_u(0) + A_u(0) + \sum_k R_{u, k}(0) = \text{population}_u$.
@@
 
-The panel iterated filter (PIF) \citep{breto20} has been successfully used to fit parameters $\theta$ of panelPOMPs in several studies \jwc{TODO: REFs}.
+The panel iterated filter (PIF) \citep{breto20} has been successfully used to fit parameters of panelPOMP models in several studies \citep{domeyer22,wale2019,breto20}.
 Like other iterated filtering algorithms, PIF iteratively combines particle filter calculations with parameter perturbations.
@@
 
-In the case of a panelPOMP, the parameter vector $\theta$ may be composed of a set of shared parameters $\phi$, and a set of unit-specific parameters $\psi_{\seq{1}{U}}$, for units $\seq{1}{U}$.
-Here, we say that a parameter $\psi_{k}$ for $k \in \seq{1}{U}$ is unit-specific if it is not involved in the one-step transition density or the measurement model of any other unit $l \neq k \in \seq{1}{U}$, and shared parameters are those that are not unit-specific.
-The PIF algorithm treats the parameter vector $\theta = (\phi, \psi_{\seq{1}{U}})$ as a complete entity and gives no distinction between shared and unit-specific parameters.
+In the case of a panelPOMP, the parameter vector $\theta$ may be composed of a set of shared parameters $\phi$, and a set of unit-specific parameters $\psi_{u}$, for units $u \in \seq{1}{U}$.
+Here, we say that a parameter $\psi_{u}$ for $u \in \seq{1}{U}$ is unit-specific if it is not involved in the one-step transition density or the measurement model of any other unit $k \neq u \in \seq{1}{U}$, and shared parameters are those that are not unit-specific.
+The PIF algorithm treats the parameter vector $\theta = (\phi, \psi_{1:U})$ as a complete entity and gives no distinction between shared and unit-specific parameters.
 Even though the random walk standard deviation is set to zero for unit-specific parameters when considering data from other units, this lack of distinction may result in particle depletion for unit-specific parameters as the number of units $U$ grows large.
+This is because it is likely that, when updating the parameters for a given unit, several parameter vector $\theta$ particles will not be resampled due to having low weights.
+When the number of units is large, this will lead to a decreasing number of ancestral particles to sample unit-specific parameters $\psi_k$ as $k$ increases.
 
 As a solution to this issue, we propose the block panel iterated filter (BPIF).
-In this algorithm, shared $\phi$ and unit-specific parameters $\psi_{\seq{1}{U}}$ are handled differently when updating the parameter vector $\theta = (\phi, \psi_{\seq{1}{U}})$.
-The BPIF updates $\theta$ by updating the shared vector $\phi$ in the same way that is done with $PIF$, but the indices for weights for unit-specific parameters $\psi_{\seq{1}{U}}$ are only updated using data from the corresponding
-unit.
+In this algorithm, shared $\phi$ and unit-specific parameters $\psi_{u}$ are handled differently when updating the parameter vector $\theta = (\phi, \psi_{1:U})$.
+The BPIF updates $\theta$ by updating the shared vector $\phi$ in the same way that is done with PIF, but index resampling for unit-specific parameters $\psi_{1:U}$ only occurs when data from the corresponding
+unit are being considered.
 In the case where there are many units, we believe that this modification will lead to less particle depletion.
@@
     Transmission & $\transmission$ & $\beta$ & $\beta$ & $c$ \\\hline
-    Water-to-Human & $\beta_W$ & --- & $\beta_W$ & $\beta$ \\\hline
+    Water-to-Human & $\Wbeta{}$ & --- & $\beta_W$ & $\beta$ \\\hline
     Bacteria Mortality & $\Wremoval$ & --- & $\delta$ & $\mu_\beta$ \\\hline
@@
     Process Over-dispersion & $\sigmaProc$ & --- & --- & $\sigma_w$\\\hline
-    % Table ends here
+    % Half Saturation Constant & $\Wsat$ & --- & $Sat$ & $K$ \\\hline
   \end{tabular}
@@
   \caption{
-  \label{tab:translate}Translations between our common notation and notation used in \cite{lee20}\eic{TODO: half saturation constant, 2 and 3. Fraction of susceptible, 2.
-  Initial Value parameters.}
+  \label{tab:translate}Translations between our common notation and notation used in \cite{lee20}
   }
@@
 
-In this article we claimed that we were able to obtained better fits to the observed data using the same models that were proposed in \cite{lee20}.
-Along with visual comparisons to the data, this claim was supported by comparing likelihoods and AIC values in Table 1.
+In this article we claimed that we were able to obtain better fits to the observed data using the same models that were proposed in \cite{lee20}.
+Along with visual comparisons to the data, this claim was supported by comparing likelihoods and AIC values in Table~1.
 Because model likelihoods were not provided in \cite{lee20}, it is necessary to replicate these models in order to obtain likelihood estimates.
+Here we would like to thank the authors of \cite{lee20}, who provided detailed descriptions of their models, which enabled us to build on their work.
 In the following subsections, we use our \code{R} package \code{haitipkg} to reproduce some of the results of \cite{lee20}.
@@
 Despite having source code available, no point estimates for model parameters were provided in \cite{lee20}.
-According to the supplement materials, this was because model realizations from a single parameter set retained substantial variability, but multiple realizations from a collection of parameter sets resulted in a reasonable visual fit to the data \jwc{REF???}.
+According to the supplement materials, this was because model realizations from a single parameter set retained substantial variability, but multiple realizations from a collection of parameter sets resulted in a reasonable visual fit to the data.
 We are also inclined to believe that the use of multiple parameter values was in part intended to account for parameter uncertainty (as mentioned in our main text), an effort by the Model~1 authors that we applaud.
 We note, however, that simulations from each of the parameter sets were treated with equal importance when being used to diagnose the model fit and make inference on the system.
-This practice is problematic given Figures S8 and S9 of the supplement material \jwc{REF???}, which suggest that some parameter sets that were used for inference were several hundred log-likelihood units lower than the best performing sets of parameters.
+This practice is problematic given Figures S8 and S9 of the supplement material, which suggest that some parameter sets that were used for inference were several hundred log-likelihood units lower than the best performing sets of parameters.
 When accounting for parameter uncertainty, one should instead weigh resulting model projections based on the likelihood of the set of parameters that were used to obtain the model realizations, as was done in \cite{king15}.
@@
 
-<<PlotEpiDist, echo=FALSE, fig.cap="Compare to Figure S8 of Lee et al. (2020) supplement.", fig.width=5.1, fig.height=5.1, fig.align='center'>>=
+<<PlotEpiDist, echo=FALSE, fig.cap="Compare to Figure S8 of Lee et al. (2020) supplement.", fig.width=7, fig.height=7, fig.align='center'>>=
 epi_results <- as.data.frame(t(coef(MIF2_search)))
@@
 
-<<plotEndParams, fig.width=5.1, fig.height=5.1, fig.align='center', fig.cap="Bivariate relationships between variables after fitting endemic period.
+<<plotEndParams, fig.width=7, fig.height=7, fig.align='center', fig.cap="Bivariate relationships between variables after fitting endemic period.
 Compare to S9 of Lee et al.
@@
 
-<<plotMod1Sims, echo=FALSE, fig.cap="Simulations from various parameter sets. Simulations from parameter sets with likelihood $<$ -2200 are removed.", cache=TRUE, fig.width=4.5, fig.height=4.25>>=
+<<plotMod1Sims, echo=FALSE, fig.cap="Simulations from various parameter sets.
+Simulations from parameter sets with likelihood $<$ -2200 are removed.
+Compare to Figure S7 of \\cite{lee20sup}.", cache=TRUE, fig.width=6.5, fig.height=3.75>>=
 end_complete <- filter(end_results, logLik > -2200)
-
 bad_end_pars <- which(end_results$logLik < -2200)
@@
   geom_point(data = true_agg_cases, aes(x = date, y = ReportedAll)) +
-  geom_line(data = quants, aes(x = date, y = q50), col = '#45ac5f') +
-  geom_ribbon(data = quants, aes(x = date, ymin = q025, ymax = q975), alpha = 0.5,
-              fill = '#45ac5f') +
+  geom_line(data = quants, aes(x = date, y = q50), col = '#0b8a2b', size = 1) +
+  geom_ribbon(
+    data = quants,
+    aes(x = date, ymin = q025, ymax = q975),
+    alpha = 0.5, fill = '#0b9c30'
+  ) +
   geom_line(data = quants_end, aes(x = date, y = q50), col = '#882871') +
@@
   theme(axis.title.x = element_blank(),
-        axis.text.x = element_text(hjust = 1, angle = 60))
+        axis.text.x = element_text(hjust = 1, angle = 45))
 @
@@
 
-The source code that was used to implement Model~2 in \cite{lee20} was written in the \code{Python} programming language, and is publicly available at \url{https://github.com/lulelita/HaitiCholeraMultiModelingProject}.
-In order to perform our analysis in a unified framework, we re-implemented this model in the \code{R} programming language using the \code{spatPomp} package \citep{asfaw21github}.
+Model~2 was developed by a team that consisted of members from the Fred Hutchinson Cancer Research Center and the University of Florida (hereafter referred to as the Model~2 authors).
+While Model~2 is the only deterministic model we considered in our analysis, it contains perhaps the most complex descriptions of cholera in Haiti: Model~2 accounts for movement between spatial units; human-to-human and environment-to-human cholera infections; and transfer of water between spatial units based on elevations charts and river flows.
+
+The source code that was used to implement Model~2 in \cite{lee20} was written in the \code{Python} programming language, and is publicly available at \url{10.5281/zenodo.3360857} and its accompanying GitHub repository \url{https://github.com/lulelita/HaitiCholeraMultiModelingProject}.
+In order to perform our analysis in a unified framework, we re-implemented this model in the \code{R} programming language using the \code{spatPomp} package \citep{asfaw21github}, which facilitates the creation of meta-population models.
+We note that the travel and water matrices used to implement the complex dynamics in Model~2 \citep{lee20sup} are not available in either the Zenodo archive or the GitHub repository;
+instead, we obtained these matrices via personal correspondence with the Model~2 authors.
+
+Using these matrices, and the point estimates for model parameters listed in \citep{lee20sup}, we created trajectories of the cholera dynamics and compared this to available data.
+These trajectories, shown in Figure~\ref{fig:mod2rep}, are very similar to the trajectories shown in Figure~S15 of \cite{lee20sup}, suggesting that our re-implementation of Model~2 in the \code{haitipkg} is a faithful approximation to the original model used in \cite{lee20}.
+
+<<Model2Replication, cache=TRUE, echo=FALSE>>=
+h2_epi <- haiti2()
+h2_end <- haiti2(region = 'after')
+
+h2_epi_traj <- trajectory(h2_epi, format = 'data.frame')
+h2_end_traj <- trajectory(h2_end, format = 'data.frame')
+
+h2_epi_traj$Ctotal <- rowSums(h2_epi_traj[, paste0("C", 1:10)]) * 0.2
+h2_end_traj$Ctotal <- rowSums(h2_end_traj[, paste0("C", 1:10)]) * 0.2
+
+dep_plot_df <- haitipkg::haitiCholera %>%
+  select(-report) %>%
+  tidyr::pivot_longer(
+    data = .,
+    cols = -date_saturday,
+    values_to = 'cases',
+    names_to = 'dep'
+  ) %>%
+  mutate(
+    date = as.Date(date_saturday),
+    dep = gsub("\\.", "_", dep)
+  ) %>%
+  mutate(
+    dep = case_when(dep == "Grand_Anse" ~ "Grande_Anse", TRUE ~ dep)
+  )
+
+true_agg_cases <- dep_plot_df %>%
+  tidyr::pivot_wider(
+    data = .,
+    id_cols = c(date),
+    names_from = dep,
+    values_from = cases,
+    names_prefix = 'cases_'
+  ) %>%
+  mutate(
+    ReportedAll = cases_Artibonite + cases_Centre +
+      cases_Grande_Anse + cases_Nippes + cases_Nord +
+      cases_Nord_Est + cases_Ouest + cases_Sud +
+      cases_Sud_Est + cases_Nord_Ouest
+  ) %>%
+  select(date, ReportedAll)
+
+h2_epi_traj$date <- as.Date(
+  lubridate::round_date(
+    lubridate::date_decimal(h2_epi_traj$year),
+    unit = 'day'
+  )
+)
+
+h2_end_traj$date <- as.Date(
+  lubridate::round_date(
+    lubridate::date_decimal(h2_end_traj$year),
+    unit = 'day'
+  )
+)
+
+all_traj <- bind_rows(
+  select(h2_epi_traj, date, Ctotal),
+  select(h2_end_traj, date, Ctotal)
+)
+
+projections <- left_join(
+  x = true_agg_cases,
+  y = all_traj,
+  by = 'date'
+)
+@
+
+\begin{figure}[ht]
+<<plotModel2Rep, cache=TRUE, echo=FALSE, fig.width=6.5, fig.height=4.5>>=
+gg_epi <- projections %>%
+  filter(date < as.Date("03-01-2014", format = '%m-%d-%Y')) %>%
+  ggplot(aes(x = date)) +
+  geom_line(aes(y = ReportedAll), col = 'black', linetype = 'dashed') +
+  geom_line(aes(y = Ctotal), col = 'red', size = 1) +
+  theme_bw() +
+  scale_y_continuous(limits = c(0, 30000), breaks = seq(0, 30000, 5000)) +
+  theme(axis.title.x = element_blank()) +
+  ylab("Reported number of cases")
+
+gg_end <- projections %>%
+  filter(date >= as.Date("03-01-2014", format = '%m-%d-%Y')) %>%
+  ggplot(aes(x = date)) +
+  geom_line(aes(y = ReportedAll), col = 'black', linetype = 'dashed') +
+  geom_line(aes(y = Ctotal), col = 'red', size = 1) +
+  theme_bw() +
+  scale_y_continuous(limits = c(0, 7000), breaks = seq(0, 7000, 1000)) +
+  theme(axis.title.x = element_blank()) +
+  ylab("Reported number of cases") +
+  scale_x_date(date_breaks = "1 year", date_labels = "%Y")
+
+cowplot::plot_grid(gg_epi, gg_end, ncol = 1)
+@
+\caption{\label{fig:mod2rep}
+Model 2 trajectories using the \code{haitipkg}. Compare to Figure S15 of \cite{lee20sup}.
+}
+\end{figure}
+
+We do note, however, several discrepancies exit between Figure~\ref{fig:mod2rep} and Figure~S15 of \cite{lee20sup}, such as a delayed second wave in the epidemic period and a larger initial wave during the endemic period.
+While these are only minor discrepancies, because Model~2 is deterministic one may expect to be able to obtain perfect replication in model trajectories.
+We believe these discrepancies are a result of slight difference in model parameters.
+For example, we note that the parameters $\transmission$, $\Wbeta{}$ (See Table~\ref{tab:translate}) had reported values of $9.9 \times 10^{-7}$ and $4.03 \times 10^{-2}$, respectively (Table~S13 of \cite{lee20sup}), but were actually fit to data and therefore likely had values that were far more precise.
+Additionally, our implementation of Model~2 used a time scale of years and many of the parameters were reported on a weekly scale, so small differences may result due to unit conversions.
+The collective effect of these small differences in model parameters likely will result in differences in model trajectories, as seen here.
 
@@
 
-The code that was originally used to implement Model~3 is archived with the DOI: 10.5281/zenodo.3360723, and also available in the public GitHub repository: \code{jcblemai/haiti-mass-ocv-campaign}.
+Model~3 was developed by a team of researchers at the Laboratory of the Swiss Federal Institute of Technology in Lausanne, hereafter referred to as the Model~3 authors.
+The code that was originally used to implement Model~3 is archived with the DOI: \url{10.5281/zenodo.3360723}, and also available in the public GitHub repository: \code{jcblemai/haiti-mass-ocv-campaign}.
 Because the code was made publicly available, and final model parameters were reported in the supplementary material of \cite{lee20}, we were able to reproduce Model~3 by directly using the source code.
@@
 
-\section{Investigation into model 2}
-
-As part of our case study into the Haiti paper, model 2 was recreated using the spatPomp package in R.
+% \section{Investigation into model 2}
+%
+% As part of our case study into the Haiti paper, model 2 was recreated using the spatPomp package in R.
 To confirm the validity of our model, we recreated Figure S15 from the supplement.
@@
 This keeps the integrity of the model intact.
-
-In Figure S-2, we created the same endemic model with the only difference being the state values are carried over from the end of the epidemic phase.
+%
+% In Figure S-2, we created the same endemic model with the only difference being the state values are carried over from the end of the epidemic phase.
 Here, we see that the model predicts a much different trajectory than in Figure S-1.
@@
 
-<<cache=T, echo=F, fig.cap="Recreation of model 2 using the pomp package.", fig.height=4>>=
-library(haitipkg)
-library("pomp")
-library("dplyr")
-library("grid")
-library("gridExtra")
-library("lattice")
-theme_set(theme_bw())
-
-h2_data <- haiti2_data()
-h2_epi <- haiti2(region="before")
-h2_end <- haiti2(region="after")
-p2_epi <- coef(h2_epi)
-
-# temporary fix. need to figure out why i cant install panelpomp
-p2_end <- coef(h2_end)
-p2_end[c("VR","Mu","v")] <- c(1.304e-9, 1.897e2, 8.201e1)
-
-traj1 <- trajectory(h2_epi,params=p2_epi,format="data.frame")
-traj1$Ctotal <- rowSums(traj1[,paste0("C",1:10)])*0.2
-traj2 <- trajectory(h2_end,params=p2_end,format="data.frame")
-traj2$Ctotal <- rowSums(traj2[,paste0("C",1:10)])*0.2
-
-traj_agg <- rbind(traj1[,c("year","Ctotal")], traj2[,c("year","Ctotal")])
-names(traj_agg) <- c("year","cases")
-traj_agg$group <- "trajectory"
-
-haiti_agg <- aggregate(h2_data$cases,
-                       by = list(h2_data$year),
-                       FUN = sum)
-names(haiti_agg) <- c("year","cases")
-haiti_agg$group <- "reported data"
-
-final <- rbind(traj_agg, haiti_agg)
-
-pp1 <- ggplot(data=final %>% filter(year<2014.161),
-             aes(x=year,y=cases,group=group)) +
-  geom_line(aes(linetype=group,color=group)) +
-  scale_linetype_manual(values=c("dashed","solid")) +
-  scale_x_continuous(breaks=seq(2010,2020,by=1)) +
-  theme(legend.position=c(0.7,0.7))
-
-pp2 <- ggplot(data=final %>% filter(year>2014.161),
-             aes(x=year,y=cases,group=group)) +
-  geom_line(aes(linetype=group,color=group)) +
-  scale_linetype_manual(values=c("dashed","solid")) +
-  scale_x_continuous(breaks=seq(2010,2020,by=1)) +
-  theme(legend.position=c(0.7,0.7)) +
-  ylim(0,7000)
-
-grid.arrange(pp1,pp2)
-@
-
-<<cache=T, echo=F, fig.cap="Endemic trajectory with state values carried over from epidemic model.", fig.height=2>>=
-library("subplex")
-theme_set(theme_bw())
-
-h2_end_joint <- haiti2(region="after",joint=TRUE) %>% pomp()
-sim <- trajectory(h2_epi,params=p2_epi,format="data.frame")
-end_IVPS <- sim[nrow(sim),-c(321,322)] %>% as.numeric()
-names(end_IVPS) <- names(h2_end_joint@params[24:343])
-
-joint_params <- c(p2_end[1:23],end_IVPS)
-ofun <- traj_objfun(h2_end_joint,est=c("Mu","v","VR"),
-                     params=joint_params)
-theta <- c(Mu=1e3,v=100,VR=2e-11)
-fit <- subplex(par=log(theta), fn=ofun)
-joint_params[c("Mu","v","VR")] <- exp(fit$par)
-
-traj3 <- trajectory(h2_end_joint,params=joint_params, format="data.frame")
-traj3$Ctotal <- rowSums(traj3[,paste0("C",1:10)])*joint_params["Rho"]
-
-traj_agg <- rbind(traj3[,c("year","Ctotal")])
-names(traj_agg) <- c("year","cases")
-traj_agg$group <- "trajectory"
-final <- rbind(traj_agg, haiti_agg)
-
-pp3 <- ggplot(data=final %>% filter(year>2014.161),
-             aes(x=year,y=cases,group=group)) +
-  geom_line(aes(linetype=group,color=group)) +
-  scale_linetype_manual(values=c("dashed","solid")) +
-  scale_x_continuous(breaks=seq(2010,2020,by=1)) +
-  theme(legend.position=c(0.7,0.7)) +
-  ylim(0,7000)
-
-pp3
-@
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2022-09-02 fit_model1.R
Created new file to fit model 1

@@ -1,0 +1,40 @@
+# This script is used to fit Model 1.
+#
+# @created: Sep 2, 2022
+
+library(haitipkg)
+library(pomp)
+library(foreach)
+library(doParallel)
+library(doRNG)
+library(dplyr)
+
+cores <-  as.numeric(Sys.getenv('SLURM_NTASKS_PER_NODE', unset=NA))
+# cores <- 20
+if(is.na(cores)) cores <- detectCores()
+registerDoParallel(cores)
+
+set.seed(636813)
+
+# Set Run Level for debug, timing, and full computation
+RUN_LEVEL = 3
+
+NP_LOCAL     <- switch(RUN_LEVEL, 100, 1000,   2000)
+NMIF         <- switch(RUN_LEVEL,  20,   50,    200)
+NUM_TREND    <- switch(RUN_LEVEL,   5,   10,     90)
+NPROF        <- switch(RUN_LEVEL,   4,    4,     24)
+NREPS_EVAL   <- switch(RUN_LEVEL,  20,   20,  cores)
+NP_EVAL      <- switch(RUN_LEVEL, 200, 1000,   5000)
+
+results <- fit_haiti1(
+  NP = NP_LOCAL, NMIF = NMIF, NUM_TREND = NUM_TREND, NPROF = NPROF,
+  NREPS_EVAL = NREPS_EVAL, NP_EVAL = NP_EVAL, ncores = cores
+)
+
+save_dir <- paste0("model1/run_level_", RUN_LEVEL)
+
+if (!dir.exists(save_dir)) {
+  dir.create(save_dir, recursive = TRUE)
+}
+
+save(results, file = paste0(save_dir, '/haiti1_fit.rda'))

### 2022-09-02 ms.Rnw
Minor changes
Created new file to fit model 1
refactored replication of lee et al model 1

@@
 \newcommand\modelCounter{m}
-\newcommand\missing{\hspace{6mm} ---}
+\newcommand\missing{---}
 \newcommand\fixed{\color{blue}}
@@
 
+@
+
+<<ReplicateModel1, echo=FALSE, message=FALSE, include=FALSE>>=
+# In this chunk, we replicate the results in Lee et al for model 1.
+# We do this replication here in the ms.Rnw file because we need to get
+# estimates for the likelihood of the models used by lee et al. The chunk has
+# the following steps:
+#   (1) Initialize some starting values close to those used in lee et al
+#   (2) do small MIF, similar to what was done in lee et al, to get a fit for
+#       epidemic period.
+#   (3) Simulate from model
+#   (4) Use simulations to initialize endemic phase of the model, and
+#       perform MIF, similar to what lee et al did for this phase
+#   (5) Simulate from model
+#   (6) Calculate the likelihood of each possible set of parameters.
+
+N_PARAMS   <- switch(RUN_LEVEL,  10,  20,  300)
+N_LEE_SIMS <- switch(RUN_LEVEL,  10,  20,   20)
+NMIF       <- switch(RUN_LEVEL,   5,  20,   50)
+NP_MIF     <- switch(RUN_LEVEL,  20,  50,  100)  # Too few, but matches Lee et al
+NP_EVAL    <- switch(RUN_LEVEL,  50, 200, 2000)
+NREPS_EVAL <- switch(RUN_LEVEL,   3,   5,   36)
+
+#-------------------------- Step 1 ----------------------
+
+set.seed(141837)
+sample_beta_nu <- function(mu_beta = 5.8, mu_nu = 0.96, n_params = N_PARAMS) {
+  # This function is used to sample starting values for beta and nu.
+  #
+  # params:
+  #   mu_beta: Average of the marginal distribution for beta1. Value was fit visually.
+  #   mu_nu: Average of the marginal distribution for nu. Value was fit visually.
+  # returns:
+  #   data.frame of size 300x2, values of beta1 and nu with truncated bivariate
+  #   normal relation.
+
+  # Determine a good value for the marginal standard deviations
+  sd_beta <- (12.1-mu_beta) / qnorm(0.995)
+  sd_nu <- (0.9-mu_nu) / qnorm(0.005)
+
+  # Create covariance matrix for the bivariate-normal distribution
+  cov_mat <- rbind(
+    c(sd_beta^2, -0.7 * sd_beta * sd_nu),
+    c(-0.7 * sd_beta * sd_nu, sd_nu^2)
+  )
+
+  # Sample from the bivariate normal distribution
+  X <- MASS::mvrnorm(n = n_params, mu = c(mu_beta, mu_nu), Sigma = cov_mat)
+  colnames(X) <- c("beta1", "nu")
+
+  # Identify values that fell outside of range, so they can be resampled
+  bad_beta <- which(X[, 'beta1'] > 12.1 | X[, 'beta1'] < 0.04)
+  bad_nu <- which(X[, 'nu'] > 1 | X[, 'nu'] < 0.9)
+  n_resamp <- length(unique(c(bad_beta, bad_nu)))
+
+  # Keep resampling values until all values fall within the range.
+  while (n_resamp > 0) {
+    to_resamp <- unique(c(bad_beta, bad_nu))
+    X[to_resamp, ] <- MASS::mvrnorm(n = n_resamp, mu = c(mu_beta, mu_nu), Sigma = cov_mat)
+
+    bad_beta <- which(X[, 'beta1'] > 12.1 | X[, 'beta1'] < 0.04)
+    bad_nu <- which(X[, 'nu'] > 1 | X[, 'nu'] < 0.9)
+    n_resamp <- length(unique(c(bad_beta, bad_nu)))
+  }
+
+  # Return results as a data.frame
+  as.data.frame(X)
+}
+# Run the function to get values of beta1 and nu
+starting_params <- sample_beta_nu()
+
+sample_rho <- function(beta, n_params = N_PARAMS) {
+  # This function is used to get starting values for the parameter rho, based
+  # on it's strong correlation with beta1 and nu. Here, we first sample beta1
+  # and nu, and then sample rho based on a regression on beta1. Additionally,
+  # it is noted that rho has a bi-modal distribution, with lower values of rho
+  # corresponding with high probability to higher values of beta1.
+  #
+  # params:
+  #   beta: previously sampled values of beta1 that will be used to get values
+  #         of rho.
+  # returns:
+  #   vector of length 300 that have an appropriate relationship with the parameter
+  #   beta1.
+
+  rhos <- rep(0, N_PARAMS)
+
+  # Get slope and intercept for the regression
+  slope <- (0.4/(5.1 - 6.75))
+  intercpt <- -6.75 * slope
+
+  # Determine whice mode should the value of rho belong to, based on regression
+  case_when(
+    beta > 6.75 ~ 0,  # Large beta --> group "0"
+    # Small beta --> more likely to be group "1", with increasing probability as beta gets smaller
+    beta <= 6.75 & beta > (0.95 - intercpt) / slope ~ slope * beta + intercpt,
+    TRUE ~ 0.95
+  ) -> probs
+
+  # Sample the groups that rho belongs to
+  rho_group1 <- rbernoulli(N_PARAMS, probs)
+  rho_group2 <- !rho_group1
+
+  # Keep track of how many values are in each group, used later for resampling
+  n_group1 <- sum(rho_group1)
+  n_group2 <- sum(rho_group2)
+
+  # Resample values of rho that are larger than 1
+  group1 <- rnorm(n_group1, mean = 1, sd = 0.05)
+  n_bad1 <- sum(group1 > 1)
+  while(n_bad1 > 0) {
+    group1[group1 > 1] <- rnorm(n_bad1, mean = 1, sd = 0.175)
+    n_bad1 <- sum(group1 > 1)
+  }
+
+  # get standard deviation for regression
+  std_beta1 <- (beta[rho_group2] - mean(beta[rho_group2])) / sd(beta[rho_group2])
+
+  group2 <- 0.25 - 0.025 * std_beta1 - (rlnorm(n_group2, sdlog = 0.2) - 1)
+
+  # Resample values of rho in group2 as needed
+  which_bad2 <- group2 < .13
+  while(sum(which_bad2) > 0) {
+    group2[group2 < 0.13] <- 0.25 - 0.1 * std_beta1[which_bad2] - (rlnorm(sum(which_bad2), sdlog = 0.2) - 1)
+
+    which_bad2 <- group2 < .13
+  }
+
+  rhos[rho_group1] <- group1
+  rhos[rho_group2] <- group2
+
+  # Return the values of rho
+  rhos
+}
+# Sample rho, using already sampled beta1.
+starting_params$rho <- sample_rho(starting_params$beta1)
+
+sample_tau <- function(mu = 4.1, n_params = N_PARAMS) {
+  # Function used to sample tau. Tau is approximately independent of the other
+  # parameters.
+  #
+  # params:
+  #   mu: mean of the truncated normal distribution
+  # returns:
+  #   vector of length 300, sampled values of tau from truncated normal distribution
+
+  # Sample from normal distribution
+  X <- rnorm(n_params, mean = mu, sd = (1.7-mu) / qnorm(0.001))
+
+  # Get how many resamples are needed
+  n_resamp <- sum(X > 5.9 | X < 1.7)
+
+  # Keep resampling until all values fall within desired range
+  while (n_resamp > 0) {
+    X[X > 5.9 | X < 1.7] <- rnorm(n_resamp, mean = mu, sd = (1.7-mu) / qnorm(0.001))
+
+    n_resamp <- sum(X > 5.9 | X < 1.7)
+  }
+
+  # Return values of tau
+  X
+}
+# Use function to sample tau
+starting_params$tau <- sample_tau()
+
+# Here we note that there are additional parameters that were fit, but their
+# relationship between the remaining parameters is not given in Lee et al. (2020).
+# Because of this, we simply sample these parameters using a truncated normal distribution:
+
+POP <- 10911819
+E_0_probs <- dnorm(seq(1, 1752, 1), mean = 400, sd = (1752-400) / qnorm(0.99))
+I_0_probs <- dnorm(seq(1, 2271, 1), mean = 500, sd = (2271-500) / qnorm(0.99))
+starting_params$E_0 <- sample(1:1752, size = N_PARAMS, replace = TRUE, prob = E_0_probs) / POP
+starting_params$I_0 <- sample(1:2271, size = N_PARAMS, replace = TRUE, prob = I_0_probs) / POP
+starting_params$S_0 <- 1 - starting_params$E_0 - starting_params$I_0
+
+rm(sample_beta_nu, sample_rho, sample_tau, POP, I_0_probs, E_0_probs)
+gc()
+
+#-------------------------- Step 2 ----------------------
+
+# Create epi model for haiti1
+h1_epi <- haiti1()
+
+bounds <- tribble(
+  ~param, ~lower, ~upper,
+  "beta2", 1e-08, 10,
+  "beta3", 1e-08, 10,
+  "beta4", 1e-08, 10,
+  "beta5", 1e-08, 10,
+  "beta6", 1e-08, 10
+)
+
+lower <- bounds$lower
+names(lower) <- bounds$param
+
+upper <- bounds$upper
+names(upper) <- bounds$param
+
+betas <- sobol_design(
+  lower = lower,
+  upper = upper,
+  nseq  = N_PARAMS
+)
+
+epi_params <- cbind(starting_params, betas)
+
+not_in_start_params <- names(coef(h1_epi))[!names(coef(h1_epi)) %in% colnames(epi_params)]
+
+for (i in 1:length(not_in_start_params)) {
+  epi_params[, not_in_start_params[i]] <- coef(h1_epi)[not_in_start_params[i]]
+}
+
+rm(i)
+
+epi_params$sig_sq <- 0
+
+# simply re-ordering to match model order
+epi_params <- epi_params[, names(coef(h1_epi))]
+
+epi_rw <- rw.sd(
+  # Parameters close to desired values
+  beta1 = 0.001,
+  tau   = 0.001,
+  nu    = 0.001,
+  rho   = 0.001,
+  # Parameters not close to desired values
+  beta2 = 0.02,
+  beta3 = 0.02,
+  beta4 = 0.02,
+  beta5 = 0.02,
+  beta6 = 0.02,
+  E_0   = ivp(0.1),
+  I_0   = ivp(0.1)
+)
+
+registerDoRNG(48963587)
+
+lee1_epi_mif <- bake(
+  file = paste0("model1/", rl_dir, "lee1_epi_fit.rds"),
+  {
+    foreach(
+      i = 1:nrow(epi_params),
+      .packages = c('pomp'),
+      .combine = c
+    ) %dopar% {
+      r_params <- unlist(epi_params[i, ])
+      coef(h1_epi) <- r_params
+      mif2(
+        h1_epi,
+        Np = NP_MIF,
+        Nmif = NMIF,
+        cooling.fraction.50 = 0.5,  # Cooling set so that we will reach smaller rw.sd at the end than the end of the unit3 search.
+        rw.sd = epi_rw
+      )
+    }
+  }
+)
+
+rm(epi_rw, epi_params, not_in_start_params,
+   lower, upper, betas, bounds, starting_params)
+gc()
+
+# epi_complete <- as.data.frame(t(coef(lee1_epi_mif)))
+# epi_complete$logLik <- logLik(lee1_epi_mif)
+
+# GGally::ggpairs(
+#   filter(epi_results, logLik > -2200),
+#   columns = c("rho", "tau", "beta1", "nu", "logLik")
+# )
+
+# epi_complete <- filter(epi_complete, logLik > -2200)
+
+lee_epi_ll <- bake(
+  file = paste0("model1/", rl_dir, "lee1_epi_evals.rds"), {
+
+    # Create data.frame to save results
+    lee_epi_ll <- data.frame(
+      "pfLL" = rep(0, length(lee1_epi_mif)),
+      "pfse" = rep(0, length(lee1_epi_mif))
+    )
+
+    for (j in 1:nrow(lee_epi_ll)) {
+      # Get parameters of interest
+      pf_params <- coef(lee1_epi_mif[[j]])
+
+      # make results reproducible
+      registerDoRNG((j * 687383921) %% 7919)
+
+      # Calculate log likelihoods
+      ll_evals <- foreach(i=1:NREPS_EVAL, .combine = c) %dopar% {
+        logLik(pfilter(h1_epi, params = pf_params, Np = NP_EVAL))
+      }
+      lee_epi_ll[j, 'pfLL'] <- logmeanexp(ll_evals, se = TRUE)[1]
+      lee_epi_ll[j, 'pfse'] <- logmeanexp(ll_evals, se = TRUE)[2]
+    }
+
+    lee_epi_ll
+  }
+)
+
+# First, we need to simulate the epidemic model using the fitted parameters.
+registerDoRNG(18599687)
+foreach(
+  i = 1:length(lee1_epi_mif),
+  .combine = rbind
+) %dopar% {
+  guess <- coef(lee1_epi_mif)
+  sims <- simulate(h1_epi, nsim = N_LEE_SIMS, format = 'data.frame',
+                   params = guess)  # 20 sims for each set of parameters
+  sims$param_set <- i
+  sims
+} -> all_sims
+
+#-------------------------- Step 3 ----------------------
+
+# quants <- all_sims %>%
+#   select(.id, week, cases) %>%
+#   group_by(week) %>%
+#   summarize(
+#     q025 = quantile(cases, probs = 0.025, na.rm = TRUE),
+#     q50  = quantile(cases, probs = 0.500, na.rm = TRUE),
+#     q975 = quantile(cases, probs = 0.975, na.rm = TRUE)
+#   ) %>%
+#   ungroup() %>%
+#   mutate(date = lubridate::ymd("2010-10-16") + lubridate::weeks(week))
+
+#-------------------------- Step 4 ----------------------
+
+h1_end <- haiti1(period = "endemic")
+start_states <- all_sims %>% filter(week == max(week))
+
+end_rw <- rw.sd(
+  beta1 = 0.05,
+  tau   = 0.005,
+  nu    = 0.04,
+  rho   = 0.01,
+  beta2 = 0.05,
+  beta3 = 0.05,
+  beta4 = 0.05,
+  beta5 = 0.05,
+  beta6 = 0.05
+)
+
+registerDoParallel(20)
+registerDoRNG(21489587)
+
+end_start <- t(coef(lee1_epi_mif))
+
+# end_start <- epi_complete %>%
+#   select(-logLik)
+
+lee1_end_mif <- bake(
+  file = paste0("model1/", rl_dir, "lee1_end_fit.rds"),
+  {
+    foreach(
+      i = 1:nrow(end_start),
+      .packages = c('pomp'),
+      .combine = c
+    ) %dopar% {
+
+      output <- list()
+      r_params <- end_start[i, ]
+
+      start_states %>%
+        filter(param_set == i) %>%
+        select(-.id, -week, -param_set) %>%
+        sapply(., median, na.rm = TRUE) -> rinit_parms
+
+      # Set the inital state
+      new_rinit <- Csnippet(
+        sprintf(
+          "
+          S = nearbyint(%f);
+          E = nearbyint(%f);
+          I = nearbyint(%f);
+          A = nearbyint(%f);
+          R = nearbyint(%f);
+          incid = nearbyint(%f);
+          foival = %f;
+          Str0 = nearbyint(%f);
+          Sout = nearbyint(%f);
+          Sin = nearbyint(%f);
+          ",
+          rinit_parms['S'],
+          rinit_parms['E'],
+          rinit_parms['I'],
+          rinit_parms['A'],
+          rinit_parms['R'],
+          rinit_parms['incid'],
+          rinit_parms['foival'],
+          rinit_parms['Str0'],
+          rinit_parms['Sout'],
+          rinit_parms['Sin']
+        )
+      )
+
+      h1_end <- h1_end %>% pomp(
+        rinit = new_rinit,
+        statenames = c(
+          "S", "E", "I", "A", "R", "incid",
+          "foival", "Str0", "Sout", "Sin"
+        )
+      )
+
+      coef(h1_end) <- r_params
+      mif2(
+        h1_end,
+        Np = NP_MIF,
+        Nmif = NMIF,
+        cooling.type = "hyperbolic",
+        cooling.fraction.50 = 0.05,  # Cooling set so that we will reach smaller rw.sd at the end than the end of the unit3 search.
+        rw.sd = end_rw
+      )
+      # output[['m2']] <- m2
+      #
+      # coef(h1_end) <- coef(m2)
+      # sims <- simulate(  # Step 5
+      #   h1_end, nsim = N_LEE_SIMS, format = 'data.frame',
+      #   params = coef(m2)
+      # )
+      # sims$param_set <- i
+      #
+      # output[['sims']] <- sims
+      # output
+    }
+  }
+)
+
+rm(end_rw, start_states)
+gc()
+
+lee_end_ll <- bake(
+  file = paste0("model1/", rl_dir, "lee1_end_evals.rds"), {
+
+    # Create data.frame to save results
+    lee_end_ll <- data.frame(
+      "pfLL" = rep(0, length(lee1_end_mif)),
+      "pfse" = rep(0, length(lee1_end_mif))
+    )
+
+    # Perform particle filter to get likelihood estimates
+    for (j in 1:nrow(lee_end_ll)) {
+      # Get parameters of interest
+      pf_params <- coef(lee1_end_mif[[j]])
+
+      # make results reproducible
+      registerDoRNG((j * 687383921) %% 7919)
+
+      # Calculate log likelihoods
+      ll_evals <- foreach(i=1:NREPS_EVAL, .combine = c) %dopar% {
+        logLik(pfilter(h1_end, params = pf_params, Np = NP_EVAL))
+      }
+
+      lee_end_ll[j, 'pfLL'] <- logmeanexp(ll_evals, se = TRUE)[1]
+      lee_end_ll[j, 'pfse'] <- logmeanexp(ll_evals, se = TRUE)[2]
+    }
+
+    lee_end_ll
+  }
+)
+
+rm(lee1_end_mif, lee1_epi_mif)
+gc()
+
+# MIF_params <- matrix(nrow = length(lee1_end_mif$m2) / 2, ncol = length(coef(lee1_end_mif[[1]])))
+# all_sims_end <- data.frame()
+# liks <- rep(0, nrow(end_start))
+# for (i in 1:nrow(end_start)) {
+#   MIF_params[i, ] <- coef(lee1_end_mif[[(2 * i) - 1]])
+#   all_sims_end <- rbind(all_sims_end, lee1_end_mif[[(2 * i)]])
+#   liks[i] <- logLik(lee1_end_mif[[(2 * i) - 1]])
+# }
+#
+# colnames(MIF_params) <- names(coef(lee1_end_mif[[1]]))
+# end_results <- as.data.frame(MIF_params)
+# end_results$logLik <- liks
+
+# GGally::ggpairs(
+#   filter(end_results, logLik > -2200),
+#   columns = c("rho", "tau", "beta1", "nu", "logLik")
+# )
+
+#-------------------------- Step 6 ----------------------
+
+mod1_lee_ll <- max(lee_end_ll$pfLL, na.rm = TRUE) + max(lee_epi_ll$pfLL, na.rm = TRUE)
+
+rm(
+  lee_end_ll, lee_epi_ll, h1_end, h1_epi, all_sims, end_start,
+  N_LEE_SIMS, N_PARAMS, NMIF, NP_EVAL, NP_MIF, NREPS_EVAL
+)
+gc()
 @
@@
 \label{eq:model3:foi}
-\lambda_u(t) &=& \Wbeta{_u} \frac{W_u(t)}{\Wsat+2W_u(t)} + \transmission_u \sum_{{\varv}\neq u}\big(I_{\varv 0}(t)+ \asymptomRelativeInfect A_{\varv 0}(t)\big)
+\lambda_u(t) &=& \Wbeta{_u} \frac{W_u(t)}{1+W_u(t)} + \transmission_u \sum_{{\varv}\neq u}\big(I_{\varv 0}(t)+ \asymptomRelativeInfect A_{\varv 0}(t)\big)
 \\[-3pt]
@@
 
-The use of likelihood-based inference allows for the statistical testing of potential improvements to the model.
+The ability to maximize the likelihood allows for likelihood-based inference, like performing statistical tests for potential improvements to the model.
 % TODO: We should address somewhere why we included the process noise in Model 1. Anna suggested this below:
@@
 <<Table 2 Lee Input, echo=FALSE, include=FALSE>>=
-load('model1/output/lee_mod1_epidemic_fit.rda')
-mod1_lee_epi_ll <- max(prof_if$loglik, na.rm = TRUE)
-rm(prof_if)
-gc()
-load('model1/output/lee_mod1_endemic_fit.rda')
-mod1_lee_end_ll <- max(prof_if$loglik, na.rm = TRUE)
-rm(prof_if)
-gc()
-
-mod1_lee_ll <- mod1_lee_epi_ll + mod1_lee_end_ll
 mod1_lee_n_params <- 20
@@
 \hline
-Immunity (year)
+$\begin{array}{l}
+\text{Immunity} \\
+\text{(year)}
+\end{array}$
   & {\fixed $\muRS^{-1}=\Sexpr{myround(1/p1["alpha"]/52,1)}$}
@@
   \\
-Birth/death (year)
+Birth/death (yr)
   & \fixed{$\begin{array}{l}
@@
 \\
-Asympt.~infectivity
+$\begin{array}{l}
+\text{Asymptomatic} \\
+\text{infectivity}
+\end{array}$
   & {\fixed $\asymptomRelativeInfect=0.05$ }
@@
 \hline
-Human to human
+$\begin{array}{l}
+\text{Human to}\\
+\text{human}\end{array}$
   & $\transmission_{1:6}$ as above
@@
 \hline
-Water to human
+$\begin{array}{l}
+\text{Water to}\\ \text{human}
+\end{array}$
   & \missing
@@
   & $\begin{array}{l}
-      \hspace{-1mm} {\fixed \Wsat=1}
-        \\
       \hspace{-1mm} \Wbeta{_{1:10}}= (
@@
 \hline
-Human to water
+$\begin{array}{l}
+\text{Human to}\\
+\text{water}\end{array}$
   & \missing
@@
 \\
-Water survival (wk)
+$\begin{array}{l}
+\text{Water} \\
+\text{survival (wk)}
+\end{array}$
   & \missing
@@
 \\
-Proc. noise (wk$^{1/2}$)
+$\begin{array}{l}
+\text{Process} \\
+\text{noise} (\text{wk}^{1/2})
+\end{array}$
   & $\sigmaProc=(\Sexpr{myround(sqrt(p1["sig_sq_epi"]),2)}, \Sexpr{myround(sqrt(p1["sig_sq_end"]),2)})$
@@
   & $\reportRate=(\Sexpr{myround(p1["rho_epi"],3)}, \Sexpr{myround(p1["rho_end"],3)})$
-  & \eqref{model1:meas}
+  & (S16) % TODO: Make sure manual equations match!
   & {\fixed $\reportRate=\Sexpr{myround(h2_fit$epi_params["Rho"],2)}$}
-  & \eqref{model2:meas}
+  & (S17)
   & $\reportRate=\Sexpr{myround(p3s["epsilon"],2)}$
-  & \eqref{model3:meas}
+  & (S18)
 \\
-Obs. overdispersion
+$\begin{array}{l}
+\text{Observation} \\
+\text{overdispersion}
+\end{array}$
   & $\obsOverdispersion=(\Sexpr{round(p1["tau_epi"],2)}, \Sexpr{round(p1["tau_end"],2)})$
-  & \eqref{model1:meas}
+  & (S16)
   &
@@
   & $\obsOverdispersion=\Sexpr{myround(p3s["k"],2)}$
-  & \eqref{model3:meas}
+  & (S18)
 \\

### 2022-09-05 si/si.Rnw
Moved calculations to main document

@@
 Not providing a point estimate for model parameters also has the effect of making model reproduction a more difficult task.
-Fortunately, the range and median values for each of the final model parameters are provided in Tables S10 and S11 of the supplement, and bivariate relationships between some of the fitted parameters are displayed in Figures S8 and S9.
-We therefore use these basic summaries to reconstruct an approximation to the set of parameters that were used by the Model~1 authors to perform inference.
+We therefore rely on the range and median values for each of the final model parameters are provided in Tables S10 and S11 of the supplement, and bivariate relationships between some of the fitted parameters (Figures S8 and S9) in order to replicate their results.
 
@@
 
-<<sampleBetaNu, echo=FALSE, cache=TRUE>>=
-set.seed(141837)
-sample_beta_nu <- function(mu_beta = 5.8, mu_nu = 0.96) {
-  # This function is used to sample starting values for beta and nu.
-  #
-  # params:
-  #   mu_beta: Average of the marginal distribution for beta1. Value was fit visually.
-  #   mu_nu: Average of the marginal distribution for nu. Value was fit visually.
-  # returns:
-  #   data.frame of size 300x2, values of beta1 and nu with truncated bivariate
-  #   normal relation.
-
-  # Determine a good value for the marginal standard deviations
-  sd_beta <- (12.1-mu_beta) / qnorm(0.995)
-  sd_nu <- (0.9-mu_nu) / qnorm(0.005)
-
-  # Create covariance matrix for the bivariate-normal distribution
-  cov_mat <- rbind(
-    c(sd_beta^2, -0.7 * sd_beta * sd_nu),
-    c(-0.7 * sd_beta * sd_nu, sd_nu^2)
-  )
-
-  # Sample from the bivariate normal distribution
-  X <- MASS::mvrnorm(n = 300, mu = c(mu_beta, mu_nu), Sigma = cov_mat)
-  colnames(X) <- c("beta1", "nu")
-
-  # Identify values that fell outside of range, so they can be resampled
-  bad_beta <- which(X[, 'beta1'] > 12.1 | X[, 'beta1'] < 0.04)
-  bad_nu <- which(X[, 'nu'] > 1 | X[, 'nu'] < 0.9)
-  n_resamp <- length(unique(c(bad_beta, bad_nu)))
-
-  # Keep resampling values until all values fall within the range.
-  while (n_resamp > 0) {
-    to_resamp <- unique(c(bad_beta, bad_nu))
-    X[to_resamp, ] <- MASS::mvrnorm(n = n_resamp, mu = c(mu_beta, mu_nu), Sigma = cov_mat)
-
-    bad_beta <- which(X[, 'beta1'] > 12.1 | X[, 'beta1'] < 0.04)
-    bad_nu <- which(X[, 'nu'] > 1 | X[, 'nu'] < 0.9)
-    n_resamp <- length(unique(c(bad_beta, bad_nu)))
-  }
-
-  # Return results as a data.frame
-  as.data.frame(X)
-}
-
-# Run the function to get values of beta1 and nu
-starting_params <- sample_beta_nu()
-@
-
-<<sample_rho, echo=FALSE, cache=TRUE>>=
-sample_rho <- function(beta) {
-  # This function is used to get starting values for the parameter rho, based
-  # on it's strong correlation with beta1 and nu. Here, we first sample beta1
-  # and nu, and then sample rho based on a regression on beta1. Additionally,
-  # it is noted that rho has a bi-modal distribution, with lower values of rho
-  # corresponding with high probability to higher values of beta1.
-  #
-  # params:
-  #   beta: previously sampled values of beta1 that will be used to get values
-  #         of rho.
-  # returns:
-  #   vector of length 300 that have an appropriate relationship with the parameter
-  #   beta1.
-
-  rhos <- rep(0, 300)
-
-  # Get slope and intercept for the regression
-  slope <- (0.4/(5.1 - 6.75))
-  intercpt <- -6.75 * slope
-
-  # Determine whice mode should the value of rho belong to, based on regression
-  case_when(
-    beta > 6.75 ~ 0,  # Large beta --> group "0"
-    # Small beta --> more likely to be group "1", with increasing probability as beta gets smaller
-    beta <= 6.75 & beta > (0.95 - intercpt) / slope ~ slope * beta + intercpt,
-    TRUE ~ 0.95
-  ) -> probs
-
-  # Sample the groups that rho belongs to
-  rho_group1 <- rbernoulli(300, probs)
-  rho_group2 <- !rho_group1
-
-  # Keep track of how many values are in each group, used later for resampling
-  n_group1 <- sum(rho_group1)
-  n_group2 <- sum(rho_group2)
-
-  # Resample values of rho that are larger than 1
-  group1 <- rnorm(n_group1, mean = 1, sd = 0.05)
-  n_bad1 <- sum(group1 > 1)
-  while(n_bad1 > 0) {
-    group1[group1 > 1] <- rnorm(n_bad1, mean = 1, sd = 0.175)
-    n_bad1 <- sum(group1 > 1)
-  }
-
-  # get standard deviation for regression
-  std_beta1 <- (beta[rho_group2] - mean(beta[rho_group2])) / sd(beta[rho_group2])
-
-  group2 <- 0.25 - 0.025 * std_beta1 - (rlnorm(n_group2, sdlog = 0.2) - 1)
-
-  # Resample values of rho in group2 as needed
-  which_bad2 <- group2 < .13
-  while(sum(which_bad2) > 0) {
-    group2[group2 < 0.13] <- 0.25 - 0.1 * std_beta1[which_bad2] - (rlnorm(sum(which_bad2), sdlog = 0.2) - 1)
-
-    which_bad2 <- group2 < .13
-  }
-
-  rhos[rho_group1] <- group1
-  rhos[rho_group2] <- group2
-
-  # Return the values of rho
-  rhos
-}
-
-# Sample rho, using already sampled beta1.
-starting_params$rho <- sample_rho(starting_params$beta1)
-@
-
-<<sample_tau, echo=FALSE, cache=TRUE>>=
-sample_tau <- function(mu = 4.1) {
-  # Function used to sample tau. Tau is approximately independent of the other
-  # parameters.
-  #
-  # params:
-  #   mu: mean of the truncated normal distribution
-  # returns:
-  #   vector of length 300, sampled values of tau from truncated normal distribution
-
-  # Sample from normal distribution
-  X <- rnorm(300, mean = mu, sd = (1.7-mu) / qnorm(0.001))
-
-  # Get how many resamples are needed
-  n_resamp <- sum(X > 5.9 | X < 1.7)
-
-  # Keep resampling until all values fall within desired range
-  while (n_resamp > 0) {
-    X[X > 5.9 | X < 1.7] <- rnorm(n_resamp, mean = mu, sd = (1.7-mu) / qnorm(0.001))
-
-    n_resamp <- sum(X > 5.9 | X < 1.7)
-  }
-
-  # Return values of tau
-  X
-}
-
-# Use function to sample tau
-starting_params$tau <- sample_tau()
-@
-
-<<initialValueParams, echo=FALSE, cache=TRUE>>=
-
-# Here we note that there are additional parameters that were fit, but their
-# relationship between the remaining parameters is not given in Lee et al. (2020).
-# Because of this, we simply sample these parameters using a truncated normal distribution:
-
-POP <- 10911819
-
-E_0_probs <- dnorm(seq(1, 1752, 1), mean = 400, sd = (1752-400) / qnorm(0.99))
-I_0_probs <- dnorm(seq(1, 2271, 1), mean = 500, sd = (2271-500) / qnorm(0.99))
-
-starting_params$E_0 <- sample(1:1752, size = 300, replace = TRUE, prob = E_0_probs) / POP
-starting_params$I_0 <- sample(1:2271, size = 300, replace = TRUE, prob = I_0_probs) / POP
-
-starting_params$S_0 <- 1 - starting_params$E_0 - starting_params$I_0
-@
-
-<<RemainingParams, echo=FALSE, cache=FALSE>>=
+<<Load lee Model1 Params, echo=FALSE>>=
 h1_epi <- haiti1()
 
-bounds <- tribble(
-  ~param, ~lower, ~upper,
-  "beta2", 1e-08, 10,
-  "beta3", 1e-08, 10,
-  "beta4", 1e-08, 10,
-  "beta5", 1e-08, 10,
-  "beta6", 1e-08, 10
-)
-
-lower <- bounds$lower
-names(lower) <- bounds$param
-
-upper <- bounds$upper
-names(upper) <- bounds$param
-
-betas <- sobol_design(
-  lower = lower,
-  upper = upper,
-  nseq  = 300
-)
-
-epi_params <- cbind(starting_params, betas)
-
-not_in_start_params <- names(coef(h1_epi))[!names(coef(h1_epi)) %in% colnames(epi_params)]
-
-for (i in 1:length(not_in_start_params)) {
-  epi_params[, not_in_start_params[i]] <- coef(h1_epi)[not_in_start_params[i]]
-}
-
-epi_params$sig_sq <- 0
-
-# simply re-ordering to match model order
-epi_params <- epi_params[, names(coef(h1_epi))]
-
-epi_rw <- rw.sd(
-  # Parameters close to desired values
-  beta1 = 0.001,
-  tau   = 0.001,
-  nu    = 0.001,
-  rho   = 0.001,
-  # Parameters not close to desired values
-  beta2 = 0.02,
-  beta3 = 0.02,
-  beta4 = 0.02,
-  beta5 = 0.02,
-  beta6 = 0.02,
-  E_0   = ivp(0.1),
-  I_0   = ivp(0.1)
-)
-
-epi_rw <- rw.sd(
-  # Parameters close to desired values
-  beta1 = 0.005,
-  tau   = 0.005,
-  nu    = 0.005,
-  rho   = 0.005,
-  # Parameters not close to desired values
-  beta2 = 0.02,
-  beta3 = 0.02,
-  beta4 = 0.02,
-  beta5 = 0.02,
-  beta6 = 0.02,
-  E_0   = ivp(0.1),
-  I_0   = ivp(0.1)
-)
-
-registerDoParallel(20)
-registerDoRNG(48963587)
-
-# Local MIF near MLE
-stew(
-  file = "output/reprod_epi_MIF2.rda",
-  {
-    # Keep track of how long it takes
-    tLocal <- system.time({
-      foreach(
-        i = 1:nrow(epi_params),
-        .packages = c('pomp'),
-        .combine = c
-      ) %dopar% {
-
-        r_params <- unlist(epi_params[i, ])
-        coef(h1_epi) <- r_params
-        mif2(
-          h1_epi,
-          Np = 100,
-          Nmif = 50,
-          cooling.fraction.50 = 0.5,  # Cooling set so that we will reach smaller rw.sd at the end than the end of the unit3 search.
-          rw.sd = epi_rw
-        ) -> m2
-      } -> MIF2_search
-    })
-  }
-)
-@
-
-<<PlotEpiDist, echo=FALSE, fig.cap="Compare to Figure S8 of Lee et al. (2020) supplement.", fig.width=7, fig.height=7, fig.align='center'>>=
-epi_results <- as.data.frame(t(coef(MIF2_search)))
-epi_results$logLik <- logLik(MIF2_search)
-
-GGally::ggpairs(
-  filter(epi_results, logLik > -2200),
-  columns = c("rho", "tau", "beta1", "nu", "logLik")
-)
-@
-
-<<SimulateEpi, echo=FALSE, include=FALSE, message=FALSE, cache=TRUE>>=
-epi_complete <- filter(epi_results, logLik > -2200)
-
-# First, we need to simulate the epidemic model using the fitted parameters.
-registerDoParallel(20)
+RL_dir <- paste0("run_level_", run_level, "/")
+lee1_epi_mif <- readRDS(file = paste0("../model1/", RL_dir, "lee1_epi_fit.rds"))
+@
+
+<<Simulate Lee Model 1, echo=FALSE, cache=TRUE>>=
+N_LEE_SIMS <- switch(run_level,  10,  20,   20)
+
 registerDoRNG(18599687)
 foreach(
-  i = 1:nrow(epi_complete),
+  i = 1:length(lee1_epi_mif),
   .combine = rbind
 ) %dopar% {
-  guess <- unlist(epi_complete[i, ])
-  sims <- simulate(h1_epi, nsim = 20, format = 'data.frame',
+  guess <- coef(lee1_epi_mif)
+  sims <- simulate(h1_epi, nsim = N_LEE_SIMS, format = 'data.frame',
                    params = guess)  # 20 sims for each set of parameters
@@
   mutate(date = lubridate::ymd("2010-10-16") + lubridate::weeks(week))
+
+@
+
+<<PlotEpiDist, echo=FALSE, fig.cap="Compare to Figure S8 of Lee et al. (2020) supplement.", fig.width=7, fig.height=7, fig.align='center'>>=
+epi_results <- as.data.frame(t(coef(lee1_epi_mif)))
+epi_results$logLik <- logLik(lee1_epi_mif)
+
+GGally::ggpairs(
+  filter(epi_results, logLik > -2200),
+  columns = c("rho", "tau", "beta1", "nu", "logLik")
+)
+@
+
+<<SimulateEpi, echo=FALSE, include=FALSE, message=FALSE, cache=TRUE>>=
+h1_end <- haiti1(period = "endemic")
+
+lee1_end_mif <- readRDS(
+  file = paste0("../model1/", RL_dir, "lee1_end_fit.rds")
+)
 
@@
 Following the model fitting scheme used by the Model~1 authors, we then use these resulting parameters as starting values for the endemic phase, and refit parameters using MIF2.
-The resulting bivariate relations between $\rho, \tau, \beta_1$ and $\nu$ are given in Figure TODO.
+To do this, we use the same number of particles and MIF iterations that were used in \citet{lee20}.
+The resulting bivariate relations between $\rho, \tau, \beta_1$ and $\nu$ are given in Figure~\ref{fig:plotEndParams}.
 
 <<FitEnd, echo=FALSE, message=FALSE, include=FALSE>>=
-h1_end <- haiti1(period = "endemic")
-start_states <- all_sims %>% filter(week == max(week))
-
-end_rw <- rw.sd(
-  beta1 = 0.05,
-  tau   = 0.005,
-  nu    = 0.04,
-  rho   = 0.01,
-  beta2 = 0.05,
-  beta3 = 0.05,
-  beta4 = 0.05,
-  beta5 = 0.05,
-  beta6 = 0.05
-)
-
-registerDoParallel(20)
-registerDoRNG(21489587)
-
-end_start <- epi_complete %>%
-  select(-logLik)
-
-# Local MIF near MLE
-stew(
-  file = "output/reprod_end_MIF2.rda",
-  {
 foreach(
-      i = 1:nrow(end_start),
-      .packages = c('pomp'),
-      .combine = c
+  i = 1:length(lee1_end_mif),
+  .combine = rbind
 ) %dopar% {
-
-      output <- list()
-      r_params <- unlist(end_start[i, ])
-
-      start_states %>%
-        filter(param_set == i) %>%
-        select(-.id, -week, -param_set) %>%
-        sapply(., median, na.rm = TRUE) -> rinit_parms
-
-      # rinit_parms["S"] <- POP - (rinit_parms["E"] + rinit_parms["I"] + rinit_parms["A"] + rinit_parms["R"])
-
-      # Set the inital state
-      new_rinit <- Csnippet(
-        sprintf(
-          "
-          S = nearbyint(%f);
-          E = nearbyint(%f);
-          I = nearbyint(%f);
-          A = nearbyint(%f);
-          R = nearbyint(%f);
-          incid = nearbyint(%f);
-          foival = %f;
-          Str0 = nearbyint(%f);
-          Sout = nearbyint(%f);
-          Sin = nearbyint(%f);
-          ",
-          rinit_parms['S'],
-          rinit_parms['E'],
-          rinit_parms['I'],
-          rinit_parms['A'],
-          rinit_parms['R'],
-          rinit_parms['incid'],
-          rinit_parms['foival'],
-          rinit_parms['Str0'],
-          rinit_parms['Sout'],
-          rinit_parms['Sin']
-        )
-      )
-
-      h1_end <- h1_end %>% pomp(
-        rinit = new_rinit,
-        statenames = c("S", "E", "I", "A", "R", "incid", "foival", "Str0", "Sout", "Sin")
-      )
-
-      coef(h1_end) <- r_params
-      mif2(
-        h1_end,
-        Np = 100,
-        Nmif = 100,
-        cooling.type = "hyperbolic",
-        cooling.fraction.50 = 0.05,  # Cooling set so that we will reach smaller rw.sd at the end than the end of the unit3 search.
-        rw.sd = epi_rw
-      ) -> m2
-
-      output[['m2']] <- m2
-
-      coef(h1_end) <- coef(m2)
+  guess <- coef(lee1_end_mif)
   sims <- simulate(
-        h1_end, nsim = 20, format = 'data.frame',
-        params = coef(m2)
+    lee1_end_mif[[i]], nsim = N_LEE_SIMS,
+    format = 'data.frame',
+    params = guess
   )
   sims$param_set <- i
-
-      output[['sims']] <- sims
-      output
-    } -> MIF2_search_end
-  }
-)
+  sims
+} -> end_sims
+
+# quants_end <- end_sims %>%
+#   select(.id, week, cases) %>%
+#   group_by(week) %>%
+#   summarize(
+#     q025 = quantile(cases, probs = 0.025, na.rm = TRUE),
+#     q50  = quantile(cases, probs = 0.500, na.rm = TRUE),
+#     q975 = quantile(cases, probs = 0.975, na.rm = TRUE)
+#   ) %>%
+#   ungroup() %>%
+#   mutate(date = lubridate::ymd("2010-10-16") + lubridate::weeks(week))
 @
@@
 (2020) supplement.", cache=TRUE, echo=FALSE>>=
-MIF_params <- matrix(nrow = length(MIF2_search_end) / 2, ncol = length(coef(MIF2_search_end[[1]])))
-all_sims_end <- data.frame()
-liks <- rep(0, nrow(end_start))
-for (i in 1:nrow(end_start)) {
-  MIF_params[i, ] <- coef(MIF2_search_end[[(2 * i) - 1]])
-  all_sims_end <- rbind(all_sims_end, MIF2_search_end[[(2 * i)]])
-  liks[i] <- logLik(MIF2_search_end[[(2 * i) - 1]])
-}
-
-colnames(MIF_params) <- names(coef(MIF2_search_end[[1]]))
-end_results <- as.data.frame(MIF_params)
-end_results$logLik <- liks
+
+end_results <- as.data.frame(t(coef(lee1_end_mif)))
+end_results$logLik <- logLik(lee1_end_mif)
 
@@
 Now that we have fit the parameters, we can simulate from the models.
+The results of these simulations are shown in Figure~\ref{fig:plotMod1Sims}.
+While not a perfect replication of the results given in \citet{lee20sup}, the similarities between the model simulations, and the fact that we based our implementation of Model~1 primarily on the source code that was provided, makes us confident that we have been able to obtain a reliable representation of the model described in \citet{lee20}.
 
@@
 
-quants_end <- all_sims_end %>%
+quants_end <- end_sims %>%
   filter(!param_set %in% bad_end_pars) %>%

### 2022-09-05 si/si.Rnw
Changed renamed functions

@@
 <<Model3Replication, echo=FALSE, cache=TRUE>>=
+NSIM <- switch(run_level, 50, 100, 1000)
+
 plot_order <- c(
@@
 
-lee3 <- haiti3()
-lee3_sims <- simulate(lee3, nsim = 1000, seed = 321, format = 'data.frame')
+lee3 <- haiti3_lee_et_al()
+lee3_sims <- simulate(lee3, nsim = NSIM, seed = 321, format = 'data.frame')
 

### 2022-09-05 ms.Rnw
Changed renamed functions
Changed the way model 2 is fit

@@
 
+mod2_epi_ll <- -epi_ofun(par = h2_fit$epi_params)
+mod2_end_ll <- -end_ofun(par = h2_fit$end_params)
+
+mod2_ll <- mod2_epi_ll - sum(apply(log(h2_epi@data + 1), 1, sum)) + (mod2_end_ll - sum(apply(log(h2_end@data + 1), 1, sum, na.rm = TRUE)))
+
 # Evaluate and sum objective functions to get total model likelihood
-mod2_ll <- -(epi_ofun(par = h2_fit$epi_params) + end_ofun(par = h2_fit$end_params))
+# mod2_ll <- -(epi_ofun(par = h2_fit$epi_params) + end_ofun(par = h2_fit$end_params))
 
@@
 # Create Lee et al (2020a) version of model 3 on sub-data
-lee3_sub <- haiti3()
+lee3_sub <- haiti3_lee_et_al()
 
@@
 
-lee3_all <- haiti3(start_time = "2010-10-23")
+lee3_all <- haiti3_lee_et_al(start_time = "2010-10-23")
 
@@
       \Sexpr{signif(52/h2_fit$epi_params["Delta"],3)}, \\
-      \Sexpr{myround(52/h2_fit$end_params["Delta"],1)}
+      \Sexpr{signif(52/h2_fit$end_params["Delta"],1)}
       )

### 2022-09-08 ms.Rnw
various text edits

@@
 Such decisions can be posed as statistical questions about scientifically motivated dynamic models.
-Thus, we encounter the methodological task of building credible, data-informed decisions for stochastic, partially observed, nonlinear dynamic models.
+Thus, we encounter the methodological task of building credible, data-informed decisions based on stochastic, partially observed, nonlinear dynamic models.
 This necessitates addressing the tradeoff between biological fidelity and model simplicity, and the reality of misspecification for models at all levels of complexity.
-% \eic{EDIT IN THE DIRECTION OF THE PREVIOUS SENTENCE.}
 As a case study, we consider a cholera epidemic in Haiti.
 The 2010 introduction of cholera to Haiti led to an extensive outbreak and sustained transmission until it was eliminated in 2019.
-% with the help of vaccination and other public health measures.
-% \jwc{Kind of the point here is that it went away without vaccination efforts.}
 We study three models developed by expert teams to advise on vaccination policies.
 We assess methods used for fitting and evaluating these models, leading to recommendations for future studies.
-% Methods that lead to better statistical fit include benchmarking mechanistic models against associative statistical models and using state-of-the art tools designed to fit mechanistic models.
-Better statistical fit facilitates interpretation of the fitted model, but caution is nevertheless required in drawing policy conclusions based on causal interpretations of the models.
+Diagnosis of model misspecification and development of alternative models can lead to improved statistical fit, but caution is nevertheless required in drawing policy conclusions based on causal interpretations of the models.
 \end{abstract}
@@
 The four independent teams were given the task of estimating the
-potential effect of increased oral cholera vaccine (OCV) efforts across Haiti.
-\jwc{New sentence:}
+potential effect of prospective oral cholera vaccine (OCV) programs.
 OCV is accepted as a safe and effective tool for controlling the spread of cholera, however the available stockpile and production of OCV doses is insufficient to meet global needs \citep{lee20, pezzoli20}.
@@
 
-The failure of \cite{lee20} to correctly predict the elimination of cholera has been the topic of some discussion \citep{francois20,rebaudetComment20,henrys20,leeReply20}.
-We demonstrate that careful attention in model fitting, diagnosing, and forecasting improves each of the proposed model's ability to quantitatively describe observed data and results in forecasts that are more consistent with the observed outcome.
-\jwc{NEXT TWO SENTENCES ARE NEW: }
-Because we focus on the statistical tasks of model fitting, diagnosing, and forecasting, we refrain from commenting on the feasibility of scientific assumptions implicitly made in the models under consideration.
-Our results demonstrate that model-mispecification does not alone explain the discrepancy between model predictions and reality, as has been suggested \citep{rebaudetComment20}.
+The failure of \cite{lee20} to correctly predict the elimination of cholera has been debated \citep{francois20,rebaudetComment20,henrys20,leeReply20}.
+\citet{rebaudetComment20} suggested that the models proposed by \citet{lee20} were too unrealistic.
+We find a more nuanced conclusion: attention to methodological details in model fitting, diagnosis and forecasting can improve each of the proposed model's ability to quantitatively describe observed data.
+These improvements results in forecasts that are more consistent with the observed outcome, without requiring major changes to the model structures.
 Based on this retrospective analysis, we offer suggestions on fitting mechanistic models to dynamic systems for future studies.
@@
 
-\section{Models}\label{sec:models}
+\section{Mechanistic models for cholera in Haiti}\label{sec:models}
 We refer to models that focus on learning relationships between variables in a dataset as associative, and we refer to models that incorporate a known scientific property of the system as causal or mechanistic.
@@
 The deliberate limitation of coordination between the groups of \citet{lee20} allows us to treat the models as fairly independently developed expert approaches to understanding cholera transmission.
-However, it results in some differences in notation, and choices of data subsets for analysis, that hinder direct comparison.
+However, it led to differences in notation, and in subsets of the data chosen for analysis, that hinder direct comparison.
 Here, we have put all three models into a common notational framework.
@@
 
-Each model describes the cholera dynamics using a latent state vector $\bm{X}^{(\modelCounter)}(t)$ and an observable response vector $\bm{Y}^{(\modelCounter)}(t)$ for each continuous time-point $t \in \mathcal{T}$, where $\mathcal{T}$ is the set of all time-points and $\modelCounter \in \{1, 2, 3\}$ indexes the model.
+Each model describes the cholera dynamics using a latent state vector $\bm{X}^{(\modelCounter)}(t)$ for each continuous time-point $t \in \mathcal{T}$, where $\mathcal{T}$ is the set of all time-points and $\modelCounter \in \{1, 2, 3\}$ indexes the model.
+The observation at time $t_n$ is modeled by a response vector $\bm{Y}^{(\modelCounter)}_n$.
 The latent state vector $\bm{X}^{(\modelCounter)}$ consists of individuals labeled as susceptible (S), infected (I), asymptomatically infected (A), vaccinated (V), and recovered (R), with various sub-divisions sometimes considered in each model for $\modelCounter \in \{1, 2, 3\}$.
 Models~2 and~3 have metapopulation structure, meaning that each individual is a member of a spatial unit, denoted by $u\in \seq{1}{U}$, which corresponds to membership of the $U=10$ administrative Haitian departments (anglicized version of departements).
-% Model~1 describes the dynamics aggregated at the national level.
-
-% \jwc{Already stated above}
-% In this study, we focus our attention on important statistical, rather than scientific, details that may improve inference on a dynamic system;
-% we therefore refrain from commenting on the assumptions in each model or proposing our own alternative description of cholera dynamics, and instead rely on the scientific expertise of \cite{lee20}.
-\jwc{NEW PARAGRAPH:}
-In the following subsections, detailed descriptions of Models~1--3 are provided.
-While these descriptions are a necessary inclusion for transparency and reproducibility, they do not immediately contribute to our primary goal of model fitting, diagnosing, and forecasting, with the purpose of informing policy.
+
+In the following subsections, complete descriptions of Models~1--3 are provided.
+While the model description is scientifically critical, as well as being necessary for transparency and reproducibility, the model details are not essential to our methodological discussions of how to diagnose and address model misspecification with the purpose of informing policy.
 A first-time reader may therefore benefit from quickly skimming through Sec.~\ref{sec:methods} and returning to the section later, as needed.
@@
 Here, $\vaccCounter=0$ corresponds to unvaccinated individuals, and $\vaccCounter \in \seq{1}{\vaccClass}$ describes hypothetical vaccination programs.
-
 The force of infection is
@@
 \end{equation}
-The process noise $d\Gamma(t)/dt$ is multiplicative Gamma-distributed white noise, with infinitesimal variance parameter $\sigmaProc^2$, which was held constant at $\sigmaProc^2 = 0$ in \cite{lee20}.
-The inclusion of gamma white noise in the process model permits the construction of an over-dispersed Markov chain \citep{breto11}, and has previously been found to increase a model's ability to fit data \citep{stocks20}.
-
+The process noise $d\Gamma(t)/dt$ is multiplicative Gamma-distributed white noise, with infinitesimal variance parameter $\sigmaProc^2$.
+\cite{lee20} included process noise in Model~3 but not in Model~1, i.e., they set $\sigmaProc^2 = 0$.
+Gamma white noise in the transmission rate gives rise to an over-dispersed latent Markov process \citep{breto11} which has been found to improve the statistical fit of disease transmission models \citep{stocks20,he10}.
 Per-capita transition rates are given in Equations~\ref{model1:SE}-\ref{model1:birth}:
@@
 Thus, $\mu_{A\demography}$ is a rate of exiting the study population from compartment $A$ and $\mu_{\demography B}$ is a rate of entering the study population into compartment $B$.
-% \eic{TODO: THE LEE ET AL SUPPLEMENTARY EQUATIONS, P14, HAVE A BIRTH RATE INTO THE VACCINE COHORTS, WHICH IS PRESUMABLY A TYPO. IT DOES NOT SHOW IN THE DIAGRAM OR THE CODE.}
-% \arc{Yes, I think it is a typo. It is not in the code, and it does not really make much sense.}
-
-In this model, the advantage afforded to vaccinated individuals is an increased probability that an infection is asymptomatic.
+
+In Model~1, the advantage afforded to vaccinated individuals is an increased probability that an infection is asymptomatic.
 Conditional on infection status, vaccinated individuals are also less infectious than their non-vaccinated counterparts by a rate of $\asymptomRelativeInfect = 0.05$ in Eq.~\eqref{model1:lambda}.
@@
 
-\jwc{NEW PARAGRAPH}
-% Mechanistic modeling is an attempt to account for known or hypothesized features of a dynamic system.
-Proposed mechanistic structures form a family of statistical models that can be indexed by a parameter vector $\paramVec$.
-Because different values of $\paramVec$ result in drastically different modeling outcomes, researchers must be careful when choosing $\paramVec$.
-Elements of $\paramVec$ are either chosen as constants that reflect our understanding of the system, or calibrated to observed data, with the intent of maximizing a proposed model's ability to quantitatively describe observed data according to some measure.
-While this section is focused on the elements of $\paramVec$ that are calibrated to data, we note that the decision of which elements to fix and which to estimate has consequences for model interpretability, as briefly discussed in Sec.~\ref{sec:results}
-
-The complex nature of the dynamic systems which are of scientific and public interest necessitates the imposition of various modeling assumptions in order to facilitate inference.
-% Many dynamic systems of scientific and public interest are highly complex by nature; therefore when developing a mathematical description of a dynamic system, it is often necessary to impose assumptions on the system in order to make inference possible.
-For example, one common assumption in epidemiology is a homogeneous mixing population \citep{bansal07}.
-Other common considerations include whether the proposed model should be stochastic or deterministic; whether the model should have change points in parameter values or should otherwise make adjustments for changes in the dynamic system; and whether the proposed model should include any spatial heterogeneity at a scale permissible by the observed data.
-Ultimately, these decisions may result in a different approach to calibrating model parameters.
-
-\jwc{MAJOR CHANGE TO THIS PARAGRAPH:}
-Each model considered in this study describes cholera dynamics as a Partially Observed Markov Process (POMP), as each model contains unobservable states that evolve sequentially with time.
+Proposed mechanistic structures form a family of statistical models indexed by a parameter vector $\paramVec$.
+Different values of $\paramVec$ can result in qualitative differences in the predicted behavior of the system.
+The complex nature of biological systems necessitates a search for modeling assumptions that combine insightful simplicity with fidelity to biological reality.
+For example, many models commonly used in epidemiology are motivated by reasoning about a homogeneous mixing population \citep{bansal07} which is simultaneously an avenue for powerful simplification and a source of model misspecification.
+Other common considerations include whether the proposed model should be stochastic or deterministic; whether the model should have change points in parameter values or should otherwise make adjustments for changes through time in the dynamic system; and whether the proposed model should include any spatial heterogeneity at a scale permissible by the observed data.
+In addition, elements of $\paramVec$ can either be chosen as constants, based on scientific reasoning and previous knowledge, or calibrated to observed data.
+Suitable methodology for calibrating model parameters may depend on other modeling decisions.
+While this section is focused on the elements of $\paramVec$ that are calibrated to data, we note that the decision of which elements to fix and which to estimate has consequences for model interpretability, as discussed in Sec.~\ref{sec:results}.
+
+All three model considered in this study describes cholera dynamics via unobservable states that evolve dynamically with time.
 Despite their similarities, these models represent a diverse selection of possible modeling assumptions: namely, the use of stochastic (Models~1 and 2) or deterministic (Model~2) equations; spatially-heterogeneous meta-population (Models~2 and 3) or spatially-aggregated (Model~1) structure; and the use of covariates (Model~3) versus mathematical equations (Models~1 and 2) to describe a seasonal mechanism.
+All these structures can be described in the framework of partially observed Markov process (POMP) models, with the understanding that the deterministic Model~2 is a degenerate case of a stochastic model.
 In the following subsections we describe our approach to fitting these mechanistic models.
@@
 One approach to modeling a dynamic system is through probabilistic models.
-With a probabilistic model,
-% rather than modeling the mean behavior of the system through the use of deterministic differential equations,
-we suppose the existence of a joint density $f_{\bm{X}^{(\modelCounter)}_{\seq{0}{N}}, \bm{Y}^{(\modelCounter)}_{\seq{1}{N}}}$, with $\bm{X}^{(\modelCounter)}_{\seq{0}{N}}$ denoting the unobservable Markov process of model $\modelCounter$ at times $\seq{0}{N} = \{0, 1, \ldots, N\}$, and $\bm{Y}^{(\modelCounter)}_{\seq{1}{N}}$ denoting the observable process of the system at times $\seq{1}{N}$.
+With a probabilistic model, we suppose the existence of a joint density $f_{\bm{X}^{(\modelCounter)}_{\seq{0}{N}}, \bm{Y}^{(\modelCounter)}_{\seq{1}{N}}}$, with $\bm{X}^{(\modelCounter)}_{\seq{0}{N}}$ denoting the unobservable Markov process of model $\modelCounter$ at times $\seq{0}{N} = \{0, 1, \ldots, N\}$, and $\bm{Y}^{(\modelCounter)}_{\seq{1}{N}}$ denoting the observable process of the system at times $\seq{1}{N}$.
 Under this framework, the observed data $y_{\seq{1}{n}}^*$ are assumed to be a single realization of the model $y_{\seq{1}{n}}^* \sim f_{\bm{X}^{(\modelCounter)}_{\seq{0}{N}}, \bm{Y}^{(\modelCounter)}_{\seq{1}{N}}}(x_{\seq{0}{N}}, y_{\seq{1}{N}}; \paramVec)$, where $\paramVec$ is a parameter vector that indexes the model.
@@
 
-\jwc{NEW PARAGRAPH:}
-Forecasting is a powerful tool that has primarily been used to estimate the future state of a system; when a mechanistic model is used, forecasts may provide estimates of the potential effects interventions may have on a system.
-Forecasts can also be used as a way of testing scientific hypothesis in real time by comparing simulations from models with competing descriptions of a system to newly obtained data \citep{lewis22}.
-In order to have any benefit to the researcher, however, forecasts must be obtained with great care.
-In particular, researchers should be careful to account for various forms of uncertainty present in model forecasts, carefully calibrate the proposed model to observed data, and---to the best of their ability---account for the most recently known state of the system.
-
+The central goal of a forecast is to provide an accurate estimate of the future state of a system based on currently available data.
+When a mechanistic model is used, forecasts may also provide estimates of the future effects of potential interventions.
+Forecasting models are built using available scientific understanding, but forecasting can also be a way of testing new scientific hypotheses in real time \citep{lewis22}.
+In order to provide trustworthy information, however, the reliability of the forecast should be undersood.
+In particular, researchers should account for various forms of uncertainty present in model forecasts, and  calibrate the proposed model to observed data.
+
+\eic{I THINK WE WILL HAVE TO EXPLAIN THIS CAREFULLY IF THE POINT IS TO APPEAR NONTRIVIAL.
+IT IS WORTH NOTING THAT LEE ET AL DID NOT PROPERLY FORECAST FROM THE BEST ESTIMATE OF THE CURRENT STATE.
+HOWEVER, IT WILL BE OBVIOUS TO MOST SATISTICAL READERS THAT THIS IS NOT THE BEST THING TO DO.
+IT MAY NOT BE SO OBVIOUS THAT THE DIFFERENCE IS IMPORTANT IF AND ONLY IF A DETERMINISTIC MODEL IS INADEQUATE.
+THE QUESTION OF WHETHER TO USE A DETERMINISTIC MODEL REMAINS OF CURRENT INTEREST...}
 As an example, we note that simulations from a well-fit mechanistic model may closely resemble the observed data $y_{1:N}^*$.
@@
 It can therefore be tempting to use simulations from this model up to time $N+s$, $f_{\bm{X}^{(\modelCounter)}_{0:N+s}, \bm{Y}^{(\modelCounter)}_{1:N+s}}\left(x_{0:N+s}, y_{1:N+s}\mid \hat{\paramVec}\right)$, with $s\geq 0$ to project the dynamic system up to a future time $N+s$, as it has been done in previous studies \citep{lee20,???}.
-% \eic{TODO: I think it's possible that other people have done this too. Could we find another reference?};
 This approach, however, does not take advantage of the information about the state of the system that is contained in the observed data.
@@
 One possible approach to account for parameter uncertainty in model forecasts is by obtaining confidence intervals for each parameter, sampling parameters from the confidence intervals, simulating the model with the resulting parameter set, and then weighing the resulting model projections based on the likelihood of the given set of parameters, as was done in \citet{king15}.
+\eic{THE ``OBVIOUS'' THING TO DO HERE IS A BAYESIAN APPROACH.
+IN SOME SENSE, KING15 IS EMPIRICAL BAYES.
+IT MAY TAKE CARE TO DISCUSS THIS WITHOUT OPENING UP A CAN OF WORMS - I CAN HAVE A GO AT MAKING SUGGESTIONS, NEXT TIME I'M WORKING ON THE MS.}
 

### 2022-09-09 ms.Rnw
Minor aesthetic changes to table
minor adjustments based on model 2 changes

@@
                                 "V4" = '#4393c3')) +
-  geom_vline(xintercept = yearsToDate(2014.161), linetype = 'dashed') +
+  # geom_vline(xintercept = yearsToDate(2014.161), linetype = 'dashed') +
   scale_y_log10(labels = scales::trans_format("log10", scales::math_format(10^.x))) +
@@
 Simulated trajectory of Model~2 (blue curve) and projections under the various vaccination scenarios.
-Reported cholera incidence is shown in black.
-The dashed vertical line shows where the endemic phase starts and parameter values are re-estimated.
-Note that the various vaccination scenarios had little effect on the projected cholera cases.
-This is because the primary effect of vaccines in Model~2 is to remove individuals from the susceptible pool,
-yet reported number of cholera infections were low and decreasing at the start of the vaccination campaigns (only $\Sexpr{true_agg_cases$ReportedAll[nrow(true_agg_cases)]}$ cases were reported in all of Haiti),
-so this had very little effect on cholera transmission.
-}
+Reported cholera incidence is shown in black.}
 \end{figure}
@@
   & \eqref{model2:mu_EI}
-  & \missing
+  & \multicolumn{1}{c}{\hspace{16pt}\missing}
   &
@@
   & \eqref{model2:lambda}
-  & $\begin{array}{l} \seasAmplitude=\Sexpr{myround(p3s["lambdaR"],2)} \\ r=\Sexpr{myround(p3s["r"],2)} \end{array}$
+  & $\begin{array}{l} \seasAmplitude=\Sexpr{myround(p3s["lambdaR"],2)} \\ r=\Sexpr{signif(p3s["r"],3)} \end{array}$
   & \eqref{eq:model3:water}
@@
     $\begin{array}{l}
-    \muRS^{-1}= \Sexpr{myround(1/h2_fit$epi_params["sigma"],1)} \\
+    \muRS^{-1}= \Sexpr{signif(1/h2_fit$epi_params["sigma"],2)} \\
      \omega_1^{-1} = \Sexpr{myround(1/h2_fit$epi_params["Omega1"],1)} \\
@@
 Vaccine efficacy
-  & \missing
+  & \multicolumn{1}{c}{\hspace{16pt}\missing}
   &
@@
   & \eqref{model1:death}
-  & \missing
+  & \multicolumn{1}{c}{\hspace{16pt}\missing}
   &
@@
   & \eqref{model1:lambda}
-  & $\begin{array}{l}
-      \hspace{-1mm} \transmission=(
-      \Sexpr{signif(h2_fit$epi_params["Beta"],2)}, \\
-      \Sexpr{signif(h2_fit$end_params["Beta"],2)}
-      )\end{array}$
+  & $\transmission=$\Sexpr{signif(h2_fit$epi_params["Beta"],3)}
   & \eqref{model2:lambda}
@@
 \end{array}$
-  & \missing
+  & \multicolumn{1}{c}{\hspace{16pt}\missing}
   &
@@
     \\
-    \beta_W= (\Sexpr{signif(h2_fit$epi_params["BetaW"],3)}, \Sexpr{signif(h2_fit$end_params["BetaW"],3)})
+    \beta_W= \Sexpr{signif(h2_fit$epi_params["BetaW"],3)}
     \end{array}$
@@
 \text{water}\end{array}$
-  & \missing
+  & \multicolumn{1}{c}{\hspace{16pt}\missing}
   &
-  & $\begin{array}{l}
-      \hspace{-1mm} \Wshed = (
-      \Sexpr{signif(h2_fit$epi_params["Mu"],3)}, \\
-      \Sexpr{signif(h2_fit$end_params["Mu"],3)}
-      )
-      \end{array}$
+  & $\Wshed = $ \Sexpr{signif(h2_fit$epi_params["Mu"],3)}
   & \eqref{model2:to_W}
@@
 \end{array}$
-  & \missing
+  & \multicolumn{1}{c}{\hspace{16pt}\missing}
   &
-  & $\begin{array}{l}
-      \hspace{-1mm} \Wremoval^{-1} = (
-      \Sexpr{signif(52/h2_fit$epi_params["Delta"],3)}, \\
-      \Sexpr{signif(52/h2_fit$end_params["Delta"],1)}
-      )
-      \end{array}$
+  & $\Wremoval^{-1} = $ \Sexpr{signif(52/h2_fit$epi_params["Delta"],3)}
   & \eqref{model2:from_W}
@@
   & \eqref{model1:lambda}
-  & \missing
+  & \multicolumn{1}{c}{\hspace{16pt}\missing}
   &
-  & \missing
+  & \multicolumn{1}{c}{\hspace{16pt}\missing}
   &
@@
   & \eqref{model1:lambda}
-  & \missing
+  & \multicolumn{1}{c}{\hspace{16pt}\missing}
   &

### 2022-09-10 ms.Rnw
some edits... more to follow

@@
 potential effect of prospective oral cholera vaccine (OCV) programs.
-OCV is accepted as a safe and effective tool for controlling the spread of cholera, however the available stockpile and production of OCV doses is insufficient to meet global needs \citep{lee20, pezzoli20}.
+OCV is accepted as a safe and effective tool for controlling the spread of cholera, however the available stockpile and production of OCV doses is insufficient to meet global needs \citep{lee20, pezzoli20} \eic{NEW CHOLERA VACCINES, NEW STUDIES OF EXISTING VACCINES, AND INCREASED PRODUCTION VOLUMES ALL AROSE DURING 2010-2019.
+MAYBE SOMETHING LIKE:
+Advances in OCV technology and vaccine availability raised the possibility of planning a national vaccination program \citep{lee20, pezzoli20}}
 In the study, certain data were shared between the groups, including demography and vaccination history; vaccine efficacy was also fixed at a shared value between groups.
@@
 
-% prev_vacc_df <- data.frame(
-%   'Departement' = c(
-%     'Artibonite',
-%     'Centre',
-%     'Grand.Anse',
-%     'Ouest',
-%     'Sud'
-%   ),
-%   start = as.Date(NA),
-%   end = as.Date(NA)
-% )
-%
-% # Grande Anse previous vaccs
-% prev_vacc_df[prev_vacc_df$Dep == 'Grand.Anse', 'start'] = as.Date("11-8-2016", format = "%m-%d-%Y")
-% prev_vacc_df[prev_vacc_df$Dep == 'Grand.Anse', 'end'] = as.Date("06-02-2017", format = "%m-%d-%Y")
-%
-% # Sud previous vaccs
-% prev_vacc_df[prev_vacc_df$Dep == 'Sud', 'start'] = as.Date("11-8-2016", format = "%m-%d-%Y")
-% prev_vacc_df[prev_vacc_df$Dep == 'Sud', 'end'] = as.Date("06-02-2017", format = "%m-%d-%Y")
-%
-% # Ouest previous vaccs
-% prev_vacc_df[prev_vacc_df$Dep == 'Ouest', 'start'] = as.Date("07-25-2017", format = "%m-%d-%Y")
-% prev_vacc_df[prev_vacc_df$Dep == 'Ouest', 'end'] = as.Date("08-25-2017", format = "%m-%d-%Y")
-%
-% # Centre previous vaccs
-% prev_vacc_df[prev_vacc_df$Dep == 'Centre', 'start'] = as.Date("11-15-2017", format = "%m-%d-%Y")
-% prev_vacc_df[prev_vacc_df$Dep == 'Centre', 'end'] = as.Date("12-16-2017", format = "%m-%d-%Y")
-%
-% # Artibonite previous vaccs
-% prev_vacc_df[prev_vacc_df$Dep == 'Artibonite', 'start'] = as.Date("04-17-2018", format = "%m-%d-%Y")
-% prev_vacc_df[prev_vacc_df$Dep == 'Artibonite', 'end'] = as.Date("05-15-2018", format = "%m-%d-%Y")
-%
-
 \section{Mechanistic models for cholera in Haiti}\label{sec:models}
-We refer to models that focus on learning relationships between variables in a dataset as associative, and we refer to models that incorporate a known scientific property of the system as causal or mechanistic.
+Models that focus on learning relationships between variables in a dataset are called {\it associative}, whereas models that incorporate a known scientific property of the system are called {\it causal} or {\it mechanistic}.
 The danger in using forecasting techniques which rely on associative models to predict the consequence of interventions is called the Lucas critique in an econometric context.
 \citet{lucas76} pointed out that it is naive to predict the effects of an intervention on a given system based entirely on historical associations.
-To successfully quantify the effect of an intervention, a model should therefore both provide a quantitative explanation of existing data and should additionally have a causal interpretation: a manipulation of the system should correspond quantitatively with the corresponding change to the model.
-We therefore focus our analysis on mechanistic statistical models, which provides a statistical fit to the available data while also supporting a causal intepretation.
+To successfully predict the effect of an intervention, a model should therefore both provide a quantitative explanation of existing data and should have a causal interpretation: a manipulation of the system should correspond quantitatively with the corresponding change to the model.
+This motivates the development of mechanistic statistical models, which provides a statistical fit to the available data while also supporting a causal intepretation.
 
@@
 The observation at time $t_n$ is modeled by a response vector $\bm{Y}^{(\modelCounter)}_n$.
-The latent state vector $\bm{X}^{(\modelCounter)}$ consists of individuals labeled as susceptible (S), infected (I), asymptomatically infected (A), vaccinated (V), and recovered (R), with various sub-divisions sometimes considered in each model for $\modelCounter \in \{1, 2, 3\}$.
-Models~2 and~3 have metapopulation structure, meaning that each individual is a member of a spatial unit, denoted by $u\in \seq{1}{U}$, which corresponds to membership of the $U=10$ administrative Haitian departments (anglicized version of departements).
+The latent state vector $\bm{X}^{(\modelCounter)}(t)$ consists of individuals labeled as susceptible (S), infected (I), asymptomatically infected (A), vaccinated (V), and recovered (R), with various sub-divisions sometimes considered in each model for $\modelCounter \in \{1, 2, 3\}$.
+Models~2 and~3 have metapopulation structure, meaning that each individual is a member of a spatial unit, denoted by a subscript $u\in \seq{1}{U}$.
+Here, the spatial units are the $U=10$ Haitian administrative d\'{e}partements (henceforth anglicized as departments).
 
@@
 While the model description is scientifically critical, as well as being necessary for transparency and reproducibility, the model details are not essential to our methodological discussions of how to diagnose and address model misspecification with the purpose of informing policy.
-A first-time reader may therefore benefit from quickly skimming through Sec.~\ref{sec:methods} and returning to the section later, as needed.
+A first-time reader may choose to skim through the rest of this section, and return later.
 
@@
 The process noise $d\Gamma(t)/dt$ is multiplicative Gamma-distributed white noise, with infinitesimal variance parameter $\sigmaProc^2$.
-\cite{lee20} included process noise in Model~3 but not in Model~1, i.e., they set $\sigmaProc^2 = 0$.
+\cite{lee20} included process noise in Model~3 but not in Model~1, i.e., they fixed $\sigmaProc^2 = 0$.
 Gamma white noise in the transmission rate gives rise to an over-dispersed latent Markov process \citep{breto11} which has been found to improve the statistical fit of disease transmission models \citep{stocks20,he10}.
+
 Per-capita transition rates are given in Equations~\ref{model1:SE}-\ref{model1:birth}:
@@
 The latent state is therefore described by the vector $\bm{X}^{(2)}(t) = \big(S_{u\vaccCounter}(t),\allowbreak E_{u\vaccCounter}(t),\allowbreak I_{u\vaccCounter}(t),\allowbreak A_{u\vaccCounter}(t),\allowbreak R_{u\vaccCounter}(t),\allowbreak R_{u\vaccCounter}^A(t),\allowbreak W_u,\allowbreak u \in \seq{1}{U},\allowbreak \vaccCounter \in \seq{0}{4}\big)$.
-A cosine term is included in Eq.~\myeqref{model2:lambda} in order to account for annual seasonality.
-Here we note the addition of a phase parameter $\phaseParm$ to allow for a possible shift in the seasonality component in the force of infection; the original implementation of Model~2 in \cite{lee20} sets $\phaseParm = 0$, not allowing for such a shift.
-% In Eq.~\myeqref{model2:lambda}, symptomatic individuals are infectious with a rate of $\transmission$, and their asymptotic counterparts are less infectious by a rate of $\asymptomRelativeInfect$.
-% Similarly, asymptomatic individuals have a human-to-water shedding rate $\Wshed$ that is lower by a factor of $\asymptomRelativeShed$.
-% TODO: I don't think the commented out section above is needed. - Jesse
+The cosine term in Eq.~\myeqref{model2:lambda} accounts for annual seasonality, with a phase parameter $\phaseParm$.
+The original implementation of Model~2 in \cite{lee20} fixes the phase at $\phaseParm = 0$.
 
@@
 \end{eqnarray}
-In \eqref{model2:transport}, we suppose a gravity model:
+In \eqref{model2:transport} the spatial coupling is specified by a gravity model,
 \begin{equation}
@@
 As with Model~1, $d\Gamma_u(t)/dt$ is multiplicative Gamma-distributed white noise in \myeqref{eq:model3:SI} and \myeqref{eq:model3:SA}.
-In \eqref{eq:model3:water}, $J_u(t)$ is a unit-less measurement of precipitation that has been standardized by dividing the observed rainfall at time $t$ by the maximum recorded rainfall in department $u$ during the epidemic, and $D_u$ is the average population density.
+In \eqref{eq:model3:water}, $J_u(t)$ is a dimensionless measurement of precipitation that has been standardized by dividing the observed rainfall at time $t$ by the maximum recorded rainfall in department $u$ during the epidemic, and $D_u$ is the average population density.
 Demographic stochasticity is accounted for by modeling non-cholera related death rate $\muDeath$ in each compartment, along with an additional death rate $\choleraDeath$ in \myeqref{eq:model3:IS} to account for cholera induced deaths among infected individuals.
 We note that all deaths are balanced by births into the susceptible compartment in \myeqref{eq:model3:AS} and \myeqref{eq:model3:RS}, thereby maintaining constant population in each department.
-% Movement into each vaccination cohort is determined by previous and future vaccination campaigns.
-% Here we note that only Model 3 accounted for previous vaccination efforts in the model.
-% TODO: The commented lines above seemed unecessary - Jesse
-
-\section{Methods}\label{sec:methods}
+
+\section{Statistical Analysis}\label{sec:methods}
+
+We consider model fitting (Sec.~\ref{model_fitting}) followed by diagnostic investigations (Sec.~\ref{model_diagnostics}) and forecasting (Sec.~\ref{sec:filter}).
 
@@
 
-\subsection{Model Fitting}
+\subsection{Model Fitting}\label{model_fitting}
 
@@
 % \subsection{Comparing mechanistic models with associative benchmarks} %AAAAAAAAAAAAAAA
-\subsection{Model Diagnostics}
-
-\jwc{NEW PARAGRAPH/EDIT: }
-The purpose of parameter calibration, irrespective of the approach taken, frequentist versus Bayesian), is to ensure that the proposed model provides the best possible description of the observed data under the assumptions of the model, as measured by some metric.
-Obtaining the best set of parameters for a given model does not, however, guarantee that the model provides an approximate representation of the system in question.
-Model mispecification, which may be thought of as the omission (or inclusion) of a mechanism in the model that is an important (or distracting) feature of the dynamic system, is inevitable at all levels of model complexity.
-In fact, we echo the sentiment of George Box, to whom the aphorism ``all models are wrong, but some are useful" is often attributed: we assume that some degree of model mispecification is always present, and it is necessary to subjectively determine whether or not the model mispecification is severe enough to render the model ``useless".
-It is often impossible to know, however, the extent to which a model provides a faithful representation of the system in question.
-In this section, we discuss some tools and suggestions for diagnosing mechanistic models with the goal of making the subjective assessment of model ``usefulness" more objective by relying on the quantitative and statistical ability of the model to match the observed data, which we call the model's goodness-of-fit.
+\subsection{Model Diagnostics} \label{model_diagnostics}
+
+Parameter calibration (whether Bayesian or frequentist) aims to find the best description of the observed data under the assumptions of the model.
+Obtaining the best fitting set of parameters for a given model does not, however, guarantee that the model provides an accurate representation of the system in question.
+Model misspecification, which may be thought of as the omission of a mechanism in the model that is an important feature of the dynamic system, is inevitable at all levels of model complexity.
+To make progress, while accepting proper limitations, one must bear in mind the much-quoted observation of \citet{box79} that ``all models are wrong but some are useful.''
+\eic{I WAS WONDERING WHETHER TO SAY SOMETHING LIKE, This is not just good practical advice for applied statistics, but matches broader appreciation of the limitations of all scientific knowledge (REF).
+DO YOU THINK THIS IS USEFUL, OR A DISTRACTION?}
+In this section, we discuss some tools and suggestions for diagnosing mechanistic models with the goal of making the subjective assessment of model ``usefulness" more objective by relying on the quantitative and statistical ability of the model to match the observed data, which we call the model's goodness-of-fit
+\eic{GOODNESS-OF-FIT SEEMED INITIALLY TO ME A MEASURE OF "RIGHT VS WRONG" RATHER THAN "USEFUL VS USELESS".
+I THINK NOW I SEE YOUR POINT: GOODNESS OF FIT CAN GIVE US SOME INDICATION OF HOW WRONG A MODEL IS, WITH THE GUIDING PRINCIPLE THAT A MODEL WHICH IS TOO FAR WRONG IS PROBABLY NOT RELIABLE FOR USEFUL PURPOSES.
+MAYBE WE HAVE TO EXPLAIN THIS TO THE READER?}
 Goodness-of-fit may provide evidence supporting the causal interpretation of one model versus another, but cannot by itself rule out the possibility of alternative explanations.
@@
 One common approach to assess a mechanistic model's goodness-of-fit is to compare several simulations from the fitted model to the observed data;
-% similar to those in Figures~\ref{fig:mod1fit}--\ref{h3panelsims};
-% TODO: citation here? \citep{???}
 this is done to determine whether or not the observed data are a plausible realization of the fitted model.

### 2022-09-11 ms.Rnw
some sunday morning tinkering

@@
 
-One common approach to assess a mechanistic model's goodness-of-fit is to compare several simulations from the fitted model to the observed data;
-this is done to determine whether or not the observed data are a plausible realization of the fitted model.
-While these comparisons are informative, they provide only a naive measure of the goodness-of-fit of a model.
-A relevant example of these comparisons leading to overconfidence in a model's goodness-of-fit is the original parameter estimates provided by \citet{lee20}:
-these parameters resulted in simulations that closely resembled the observed data, yet resulted in model likelihoods that were---in some cases---remarkably smaller than likelihoods that can be achieved via the likelihood based optimization techniques that were used (see Table~\ref{tab:likes}).
-% \eic{TODO: Could we maybe put an example of this here? }
+One common approach to assess a mechanistic model's goodness-of-fit is to compare simulations from the fitted model to the observed data.
+Visual inspection may indicate defects in the model, or may  suggest that the observed data are a plausible realization of the fitted model.
+While visual comparisons can be informative, they provide only a weak and informal measure of the goodness-of-fit of a model.
+The study by \citet{lee20} provides an example of this: their models and parameter estimates resulted in simulations that visually resembled the observed data, yet resulted in model likelihoods that were---in some cases---remarkably smaller than likelihoods that can be achieved via the likelihood based optimization techniques that were used (see Table~\ref{tab:likes}).
 Alternative forms of model validation should therefore be used in conjunction with visual comparisons of simulations to observed data.
 
-Another approach is to compare a quantitative description of the model fit (such as MSE, predictive accuracy, or model likelihood) to that of other proposed models.
-This type of comparison provides insight into how each model performs relative to the others, but does not provide insight on the performance of each model on a given dataset unless the performance of one of the models is widely understood.
-% "each other" -> "the others", based on Anna's suggestion.
-We refer to such a model as a benchmark, and examples of these models include logistic regression and ordinary least squares.
-
-Comparing associative models using an objective function seems natural, as these models are designed to optimize objective functions given the available data.
-In the case of mechanistic modeling, however, comparing models based on an objective measure is less natural because each model is designed to provide a scientific description of the dynamic system.
-Any comparison between mechanistic models would therefore ideally be based on the scientific validity of each proposed model, which is often impossible to know.
-Still, parameters in mechanistic models are fit using techniques that are designed to maximize an objective
-% TODO: Anna suggests "objective function". Maybe "objective value" as well, or is it okay to leave as it is?
-such as likelihood, and it is expected that a good scientific description of the dynamic system should result in more optimal values of an objective evaluation than would a poor description of the system.
-In this sense, one could justify comparing mechanistic models based on some objective measurement such as model likelihoods.
-% When performing such a comparison, however,
-We do not suggest that the model with the highest likelihood given the data automatically provides a better scientific description of the dynamic system.
-Still, a comparison based on model likelihoods may help guide researchers to know the plausibility of a proposed model compared to another given their relative ability to describe the observed data.
-
-The comparison across models still only provides a relative measure of model fit compared to other mechanistic models.
-The resulting challenge is that no general benchmark exists for mechanistic models that would allow researchers to assess the goodness-of-fit by comparing to a known standard.
-In some cases, a possible benchmark model could be a generally accepted truth, but often no such truth is available.
-% In the case of Cholera infections in Haiti, for example, it has been debated whether or not bacteria buildup in the environment is a cause of Cholera infections \cite{???}.
-Because of this, we propose the use of a log-linear ARMA model as a potential benchmark.
-Despite being an associative model, the reason that an ARMA model is an appealing benchmark is because the performance of this class of models is widely understood, and often provides a good description of the data with very limited effort.
-% For us, the defining feature of an associative model is that it is assessed primarily on its quantitative ability to explain historical data.
-% Mechanistic models, on the other hand, must provide a scientific hypothesis about the system as well as having some quantitative ability to explain observed data.
-% By this definition, mechanistic models are a subset of associative models;
-% this subset relationship is enough to see that one cannot expect the class of mechanistic models to provide the highest levels of statistical fit.
-% In general one should expect a statistical cost to the constraint of requiring mechanistic interpretation of the model.
-One question that follows is whether it is useful to assess mechanistic models by their statistical fit, if this is not their primary purpose?
-To what extent can we say the data support a scientific model ($A$) over another model ($B$) based on improved fit to data, if neither fit better than an associative model ($C$)?
-% An answer is to think in terms of the shortfall between a mechanistic model and the best known associative model.
-% If the shortfall is small, on a suitable scale, then the mechanistic model may be considered a valid view of the system from the point of view of these data.
-% While the ARMA(2, 1) certainly will not automatically provide the best fit of all associative models,
-% This does not demonstrate the correctness of the mechanistic interpretation, but also it does not refute it.
-% If the shortfall is large, we conclude that there is some substantial aspect of the system that is not incorporated into the model.
-% we cannot necessarily tell which part of the model is invalid, but we know there is a weakness somewhere.
-% To assess whether shortfalls are small or large, we may consider using formal statistical hypothesis tests or model selection techniques.
-% These can give us a scale for chance variation: a difference small enough to be explained by randomness in the data generating mechanism should not be considered as strong evidence.
-% However, the comparison of primary interest is one of practical significance rather than statistical significance.
-% We would like a mechanistic model whose statistical fit is close to the best associative models regardless of whether there is enough statistical power to distinguish the difference.
-
-% I like the above addition because it provides a little bit more of the philosophy behind the paragraph below. Perhaps we can combine them? - Anna
-
-When comparing mechanistic models to non-mechanistic benchmarks, we do not require the benchmark model to optimize a measure of statistical fit, nor do we suggest that a mechanistic model must have superior fit to the benchmark in order to be of scientific interest.
-It is possible that a successful mechanistic model may have better statistical fit than simple benchmarks such as log-linear models;
-still, the specific threshold of beating, or failing to beat, an arbitrary benchmark is similarly arbitrary.
-In this case, however, a large shortfall in the fit of a mechanistic model to that of a log-linear model may demonstrate room for substantial improvement.
+Another approach is to compare a quantitative measure of the model fit (such as MSE, predictive accuracy, or model likelihood) among all proposed models.
+These comparisons provide insight into how each model performs relative to the others.
+To calibrate relative measures of fit, it is useful to compare against a model that has well-understood statistical ability to fit data, and we call this model a {\it benchmarks}.
+Standard statistical models, interpreted as associative models without requiring any mechanistic interpretation of their parameters, provide suitable benchmarks. 
+Examples include linear regression, auto-regressive moving average time series models, or even independent and identically distributed measurements.
+The benchmarks enable us to evaluate the goodness of fit that can be expected of a suitable mechanistic model.
+
+Goodness-of-fit alone does not guarantee that a model provides a correct causal interpretation of the model.
+Indeed, associative models are not constrained to have a causal interpretation, and typically are designed with the sole goal of providing a statistical fit to data.
+Consequently, we should not require a candidate mechanistic model to beat all benchmarks.
+However, a mechanistic model which falls far short against benchmarks is evidently failing to explain some substantial aspect of the data.
+A convenient measure of fit should have interpretable differences that help to operationalize the meaning of far short.
+Ideally, the measure should also have favorable theoretical properties.
+Consequently, we focus on log-likelihood as a measure of goodness of fit, and we adjust for the degrees of freedom of the models to be compared by using Aka-Ike's information criterion (AIC).
+
+It should be universal practice to present measures of goodness of fit for published models, and mechanistic models should be compared against benchmarks.
+This alone would assist authors and readers to confront any major statistical limitations of the proposed mechanistic models.
+In addition, the published goodness of fit provides a concrete measure for subsequent research to identify and remedy limitations in the analysis, or to update the investigation based on new data or new scientific understanding.
+When combined with online availability of data and code, objective measures of fit provide a powerful tool to accelerate scientific progress, following the paradigm of the {\it common task framework} \citep[][Sec.~6]{donoho17}.
+In our literature review of the Haiti cholera epidemic \eic{CHECK THIS.
+HOW/WHERE SHOULD WE DESCRIBE THIS?} no quantitative measures of goodness of fit, and no benchmark models, were considered in any of the ??
+papers which calibrated a mechanistic model to data in order to obtain scientific conclusions.
+
+In some cases, a possible benchmark model could be a generally accepted mechanistic model, but often no such model is available.
+Because of this, we use a log-linear Gaussian ARMA model as an associative benchmark, as recommended by \citet{he10}.
+The theory and practice of ARMA models is well developed, but the exponential growth and decay characteristic of biological dynamics suggests applying these linear models on a log scale.
+
 The use of a common benchmark may also be beneficial when developing models with varying spatial scale, as direct comparisons between models fit to data with different levels of spatial aggregation are meaningless (e.g.
 comparing Model~1 to Model~3).
+\eic{I DON'T QUITE UNDERSTAND THIS YET.
+BENCHMARKS TEND TO APPLY TO SPECIFIC DATA SETS, SO CANNOT READILY BE USED TO COMPARE DIFFERENT LEVELS OF AGGREGATION.
+LIKELIHOOD BENCHMARKS CAN BE USED TO STUDY THE VALUE OF EXPLICIT SPATIAL COUPLING.
+HOW TO BENCHMARK ACROSS AGGREGATION LEVELS IS, I THINK, SOMETHING OF AN OPEN PROBLEM.}
 Likelihoods of Models~1--3 and their respective ARMA benchmark models are provided in Table~\ref{tab:likes}.

### 2022-09-11 ms.Rnw
typo

@@
 Ideally, the measure should also have favorable theoretical properties.
-Consequently, we focus on log-likelihood as a measure of goodness of fit, and we adjust for the degrees of freedom of the models to be compared by using Aka-Ike's information criterion (AIC).
+Consequently, we focus on log-likelihood as a measure of goodness of fit, and we adjust for the degrees of freedom of the models to be compared by using the Akaike information criterion (AIC) \eic{REF}.
 

### 2022-09-12 ms.Rnw
finishing this round of edits

@@
 In this model, both the latent and observable processes can be factored into department specific processes which interact with each other.
-We denote the latent and measurement processes for Model~3 as $\bm{X}^{(3)}(t_{0:N}) = X^{(3)}_{1:U, 0:N}$, and $\bm{Y}^{(3)}(t_{1:N}) = Y_{1:U, 1:N}$.
+We denote the latent and measurement processes for Model~3 as $\bm{X}^{(3)}(t_{0:N}) = X^{(3)}_{1:U, 0:N}$, and $\bm{Y}^{(3)}(t_{1:N}) = Y_{1:U, 1:N}$ \eic{IS THIS NEEDED HERE?}.
 The decision to model the system via metapopulation models versus a model aggregated to a larger spatial scale is one of great scientific interest, and evidence for the former approach has been provided in previous studies \citep{king15}.
-Note that this evidence alone does not automatically discredit conclusions drawn via nationally aggregated models, as an argument can easily be made that one should prefer a simple model over a complex one \citep{saltelli20,green15}.
-Still, researchers that intend to use a mechanistic model to describe a dynamic system should design the model to be as realistic as their scientific understanding of the system and their computational abilities permit.
-% As such, there is strong motivation to include spatial structure in the model designed to approximate the system.
+Note that this evidence alone does not automatically discredit conclusions drawn via nationally aggregated models, as there are also good reasons to prefer a simple model over a complex one \citep{saltelli20,green15}.
+Still, researchers that intend to use a mechanistic model to describe a dynamic system should design the model to be as realistic as their scientific understanding of the system and their computational abilities permit \eic{THIS SENTENCE SEEMS TO DISMISS THE ARGUMENTS IN THE PREVIOUS SENTENCE.
+PERHAPS THIS SENTENCE CAN JUST BE DELETED}.
 Fitting scientifically flexible metapopulation models, however, remains a challenging statistical problem;
@@
 The shared parameters were calibrated using the cholera incidence data from Artibonite, and the department-specific parameters ($\beta_{W_u}$ and $\beta_u$) were fit using the data from their respective department.
-Reducing a spatially-heterogeneous model to individual units in this fashion requires special treatment of any interactive mechanisms between spatial units, such as found in Eq.~\myeqref{eq:model3:foi}.
+Reducing a spatially coupled \eic{not spatially heterogeneous, as previously written} model to individual units in this fashion requires special treatment of any interactive mechanisms between spatial units, such as found in Eq.~\myeqref{eq:model3:foi}.
 In particular, when considering a model for department $u$, the values $I_\nu(t)$ and $A_\nu(t)$, are unknown for $u \neq v \in \seq{1}{10}$.
@@
 \hline
-\multirow{2}{*}{Log Likelihood} &
+\multirow{2}{*}{Log-likelihood} &
   $\Sexpr{myround(mod1_ll, 1)}$ &
@@
 \hline
-\thead{ARMA(2, 1) \\ Log Likelihood} &
+\thead{Log-ARMA(2,1) \\ Log-likelihood} &
   $\Sexpr{myround(ARMA_benchmarks$m1, 1)}$ &
@@
 \caption{\label{tab:likes}Log-likelihood values for each models compared to their ARMA benchmarks.
-Values in parenthesis are corresponding values using \citet{lee20} parameter estimates.}
+Values in parenthesis are corresponding values using \citet{lee20} parameter estimates.
+\eic{I THINK THAT THE NOTES SHOULD GO IN THE CAPTION RATHER THAN BEING FOOTNOTES.}
+\eic{WE SHOULD REVIEW THIS TABLE IN A MEETING.
+RIGHT NOW, I'M SURPRISED BY A FEW THINGS.
+WHY IS THE LEE ET AL MODEL 2 LIKELIHOOD HIGHER THAN OURS?
+IS THE (29) IN THE PANELPOMP COLUMN A TYPO, SINCE WE DON'T COMPARE TO LEE ET AL IN THIS COLUMN?
+NESTING RELATIONSHIP FOR THE SPATPOMP VERSION OF MODEL 3 MAKES IT HARD TO EXPLAIN THE RATHER LARGE DECREASE IN LOGLIK.
+AT FACE VALUE, THIS SEEMS TO MEAN THAT THE COUPLING IS SUFFICIENTLY POORLY SPECIFIED THAT WE ARE BETTER OFF WITHOUT IT.
+I SEEM TO RECALL THAT OUR UNDERSTANDING INVOLVES THE WAY PANELPOMP USES A COVARIATE THAT DEPENDS ON CASES IN OTHER DEPARTMENTS.
+SINCE LEE ET AL ALSO DO THAT, COULD IT BE THAT THEIR ANALYSIS IS MORE CLOSELY RELATED TO THE PANELPOMP VERSION THAN THE SPATPOMP VERSION?}
+}
 \end{table}
@@
 \jwc{SHOULD THESE FIRST TWO PARAGRAPHS BE MOVED TO THE DISCUSSION?}
+\eic{YES.
+ALSO, THERE ARE RESULTS IN SEC.~3.
+PERHAPS, THE RAINFALL ANALYSIS IN SEC 4 COULD BE ANOTHER ASPECT OF THE SEC 3 ANALYSIS: COMPARISON OF MODEL FIT WITH OTHER LINES OF SCIENTIFIC EVIDENCE.
+THEN THE CURRENT SEC 4 COULD BE DISBANDED.
+SOMETIMES THE METHODS-RESULTS-DISCUSSION FORMAT DOES NOT FIT SO NATURALLY WITH APPLIED STATS.
+WE CAN DISCUSS THIS.
+ONE IDEA IS TO HAVE A SECTION NAMED SOMETHING LIKE "LESSONS LEARNED FROM MODELING THE HAITI EPIDEMIC" OR "ADVICE FOR FUTURE STUDIES".
+THE LATTER CAN SEEM PRESUMPTUOUS, AND SHOULD BE DONE WITH DUE CAUTION, BUT FROM ANOTHER PERSPECTIVE IF WE ARE NOT PROVIDING CONCRETE GUIDANCE FOR FUTURE WORK THEN WHAT ARE WE DOING?
+OUR RESPONSIBILITY IS TO MAKE THE STRONGEST CLAIM WE THINK WE CAN PROPERLY SUPPORT; THEN THE REFEREES CAN HAVE A GO AT PULLING THE CLAIM DOWN.}
+\eic{I THINK IT IS INTERESTING THAT THE MODEL-SUPPORTED SUGGESTION BY \citep{andrews11} HAS LARGELY PROVED ACCURATE (VACCINATION, NON-PHARMACEUTICAL INTERVENTIONS, AND INFRASTRUCTURE IMPROVEMENTS ALL HAVE A ROLE AND SHOULD BE FOLLOWED SIMULTANEOUSLY)DESPITE LEGITIMATE CONCERNS ABOUT WHETHER THE SCIENTIFIC SUPPORT FOR THE MODEL IS STRONG ENOUGH TO ROBUSTLY GUIDE POLICY \citep{grad12}.
+THE MAIN CLAIM OF ANDREWS11 IS THAT FORECASTS AND POLICY NOT SUPPORTED BY A MODEL SEEM EVEN MORE UN-ROOTED THAN RECOMMENDATIONS BASED ON A WEAKLY SUPPORTED MODEL.
+THE CONCLUSION FROM ALL THIS, WHICH I THINK IS SUPPORTED BY OUR RE-ANALYSIS OF LEE ET AL, IS THAT MECHANISTIC MODELS FITTED TO DATA ARE A STRONG WAY TO ENFORCE CONSISTENT THINKING, BUT WEAKER AS A WAY TO CHECK THAT YOU ARE ACTUALLY RIGHT DUE TO THE LIMITATIONS OF INFERRING CAUSAL MECHANISMS FROM OBSERVATIONAL DATA.
+POLICY-MAKERS SHOULD FOLLOW AN APPROACH WHICH IS SELF-CONSISTENT AND CONSISTENT WITH PREVIOUS EVIDENCE THAT MIGHT WORK (WHICH MODELS CAN HELP WITH) BUT THEY MUST BE AGILE AS NEW EVIDENCE EMERGES.
+MODELS, TOGETHER WITH DATA ANALYSIS, CAN ALSO HELP REVEAL WHEN THE EVIDENCE SUPPORTING A P0LICY IS BREAKING DOWN, FORCING A RE-THINK.
+}
+
 A model which aspires to provide quantitative guidance for assessing interventions should provide a quantitative statistical fit for available data.
@@
 Simulations from probabilistic models (Models~1 and~3) represent possible trajectories of the dynamic system under the scientific assumptions of the models.
-In this case study, estimates of the probability of cholera elimination can therefore be obtained as the proportion of simulations from the fitted model
-% starting from draws from the filtering distribution, unecessary - Jesse
-that result in cholera elimination.
+In this case study, estimates of the probability of cholera elimination can therefore be obtained as the proportion of simulations from the fitted model that result in cholera elimination.
 The results of these projections are summarized in Figs.~\ref{fig:Mod1Scenarios}--\ref{fig:elimProbs}.
 These results suggest that cholera elimination was likely, even without increased vaccination efforts, which is consistent with observed reality \citep{ferguson22}.
+\eic{I LIKE THE IDEA OF REFERENCING POST-EPIDEMIC REFLECTIONS.
+UNLESS THERE ARE NO ALTERNATIVES, IT IS BEST TO AVOID CITING A NON-PEER-REVIEWED ARTICLE SUCH AS THIS.
+HOW ABOUT \citep{piarroux22} OR \citep{trevisin22}?}
 

### 2022-09-12 ms.Rnw
IMPORTANT: haitipkg update needed

--- 2022-09-11 ms.Rnw
@@
 
-  h2_epi <- haiti2(region = 'before')
-  h2_end <- haiti2(region = 'after', joint = TRUE)
-
-  coef(h2_epi) <- h2_fit$epi_params
-  coef(h2_end) <- h2_fit$end_params
-  # p2 <- coef(h2)
+  h2_epi <- haiti2(region = 'before', cutoff = 10000, measure = 'log')
+
+  coef(h2_epi) <- h2_fit$h2_params
 
@@
 
-Model~2 is a deterministic model, which can be viewed as a special case of a POMP (or SpatPOMP) with no randomness in the dynamic model.
+Model~2 is a deterministic model, which can be viewed as a special case of a POMP (or SpatPOMP) with no randomness in the dynamic process or measurement.
 By combining a deterministic process model with a simple Gaussian measurement model, Model~2 reduces model fitting to a least squares calculation over parameters in a set of differential equations.
@@
 <<Model 2 Trajectory, echo=FALSE>>=
-h2_epi_traj <- trajectory(h2_epi, params = h2_fit$epi_params, format = 'data.frame')
-h2_epi_traj$Ctotal <- rowSums(h2_epi_traj[, paste0("C", 1:10)]) * h2_fit$epi_params['Rho']
-
-h2_end_traj <- trajectory(h2_end, params = h2_fit$end_params, format = 'data.frame')
-h2_end_traj$Ctotal <- rowSums(h2_end_traj[, paste0("C", 1:10)]) * h2_fit$end_params['Rho']
-
-h2_traj <- rbind(
-  h2_epi_traj[, c('year', 'Ctotal')],
-  h2_end_traj[, c('year', 'Ctotal')]
-) %>%
+h2_epi_traj <- trajectory(h2_epi, params = h2_fit$h2_params, format = 'data.frame')
+h2_epi_traj$Ctotal <- rowSums(h2_epi_traj[, paste0("C", 1:10)]) * h2_fit$h2_params['Rho']
+
+h2_traj <- h2_epi_traj %>%
+  select(year, Ctotal) %>%
   mutate(
@@
 
-One common approach to assess a mechanistic model's goodness-of-fit is to compare simulations from the fitted model to the observed data.
-Visual inspection may indicate defects in the model, or may  suggest that the observed data are a plausible realization of the fitted model.
-While visual comparisons can be informative, they provide only a weak and informal measure of the goodness-of-fit of a model.
-The study by \citet{lee20} provides an example of this: their models and parameter estimates resulted in simulations that visually resembled the observed data, yet resulted in model likelihoods that were---in some cases---remarkably smaller than likelihoods that can be achieved via the likelihood based optimization techniques that were used (see Table~\ref{tab:likes}).
+One common approach to assess a mechanistic model's goodness-of-fit is to compare several simulations from the fitted model to the observed data;
+this is done to determine whether or not the observed data are a plausible realization of the fitted model.
+While these comparisons are informative, they provide only a naive measure of the goodness-of-fit of a model.
+A relevant example of these comparisons leading to overconfidence in a model's goodness-of-fit is the original parameter estimates provided by \citet{lee20}:
+these parameters resulted in simulations that closely resembled the observed data, yet resulted in model likelihoods that were---in some cases---remarkably smaller than likelihoods that can be achieved via the likelihood based optimization techniques that were used (see Table~\ref{tab:likes}).
+% \eic{TODO: Could we maybe put an example of this here? }
 Alternative forms of model validation should therefore be used in conjunction with visual comparisons of simulations to observed data.
 
-Another approach is to compare a quantitative measure of the model fit (such as MSE, predictive accuracy, or model likelihood) among all proposed models.
-These comparisons provide insight into how each model performs relative to the others.
-To calibrate relative measures of fit, it is useful to compare against a model that has well-understood statistical ability to fit data, and we call this model a {\it benchmarks}.
-Standard statistical models, interpreted as associative models without requiring any mechanistic interpretation of their parameters, provide suitable benchmarks. 
-Examples include linear regression, auto-regressive moving average time series models, or even independent and identically distributed measurements.
-The benchmarks enable us to evaluate the goodness of fit that can be expected of a suitable mechanistic model.
-
-Goodness-of-fit alone does not guarantee that a model provides a correct causal interpretation of the model.
-Indeed, associative models are not constrained to have a causal interpretation, and typically are designed with the sole goal of providing a statistical fit to data.
-Consequently, we should not require a candidate mechanistic model to beat all benchmarks.
-However, a mechanistic model which falls far short against benchmarks is evidently failing to explain some substantial aspect of the data.
-A convenient measure of fit should have interpretable differences that help to operationalize the meaning of far short.
-Ideally, the measure should also have favorable theoretical properties.
-Consequently, we focus on log-likelihood as a measure of goodness of fit, and we adjust for the degrees of freedom of the models to be compared by using Aka-Ike's information criterion (AIC).
-
-It should be universal practice to present measures of goodness of fit for published models, and mechanistic models should be compared against benchmarks.
-This alone would assist authors and readers to confront any major statistical limitations of the proposed mechanistic models.
-In addition, the published goodness of fit provides a concrete measure for subsequent research to identify and remedy limitations in the analysis, or to update the investigation based on new data or new scientific understanding.
-When combined with online availability of data and code, objective measures of fit provide a powerful tool to accelerate scientific progress, following the paradigm of the {\it common task framework} \citep[][Sec.~6]{donoho17}.
-In our literature review of the Haiti cholera epidemic \eic{CHECK THIS.
-HOW/WHERE SHOULD WE DESCRIBE THIS?} no quantitative measures of goodness of fit, and no benchmark models, were considered in any of the ??
-papers which calibrated a mechanistic model to data in order to obtain scientific conclusions.
-
-In some cases, a possible benchmark model could be a generally accepted mechanistic model, but often no such model is available.
-Because of this, we use a log-linear Gaussian ARMA model as an associative benchmark, as recommended by \citet{he10}.
-The theory and practice of ARMA models is well developed, but the exponential growth and decay characteristic of biological dynamics suggests applying these linear models on a log scale.
-
+Another approach is to compare a quantitative description of the model fit (such as MSE, predictive accuracy, or model likelihood) to that of other proposed models.
+This type of comparison provides insight into how each model performs relative to the others, but does not provide insight on the performance of each model on a given dataset unless the performance of one of the models is widely understood.
+% "each other" -> "the others", based on Anna's suggestion.
+We refer to such a model as a benchmark, and examples of these models include logistic regression and ordinary least squares.
+
+Comparing associative models using an objective function seems natural, as these models are designed to optimize objective functions given the available data.
+In the case of mechanistic modeling, however, comparing models based on an objective measure is less natural because each model is designed to provide a scientific description of the dynamic system.
+Any comparison between mechanistic models would therefore ideally be based on the scientific validity of each proposed model, which is often impossible to know.
+Still, parameters in mechanistic models are fit using techniques that are designed to maximize an objective
+% TODO: Anna suggests "objective function". Maybe "objective value" as well, or is it okay to leave as it is?
+such as likelihood, and it is expected that a good scientific description of the dynamic system should result in more optimal values of an objective evaluation than would a poor description of the system.
+In this sense, one could justify comparing mechanistic models based on some objective measurement such as model likelihoods.
+% When performing such a comparison, however,
+We do not suggest that the model with the highest likelihood given the data automatically provides a better scientific description of the dynamic system.
+Still, a comparison based on model likelihoods may help guide researchers to know the plausibility of a proposed model compared to another given their relative ability to describe the observed data.
+
+The comparison across models still only provides a relative measure of model fit compared to other mechanistic models.
+The resulting challenge is that no general benchmark exists for mechanistic models that would allow researchers to assess the goodness-of-fit by comparing to a known standard.
+In some cases, a possible benchmark model could be a generally accepted truth, but often no such truth is available.
+% In the case of Cholera infections in Haiti, for example, it has been debated whether or not bacteria buildup in the environment is a cause of Cholera infections \cite{???}.
+Because of this, we propose the use of a log-linear ARMA model as a potential benchmark.
+Despite being an associative model, the reason that an ARMA model is an appealing benchmark is because the performance of this class of models is widely understood, and often provides a good description of the data with very limited effort.
+% For us, the defining feature of an associative model is that it is assessed primarily on its quantitative ability to explain historical data.
+% Mechanistic models, on the other hand, must provide a scientific hypothesis about the system as well as having some quantitative ability to explain observed data.
+% By this definition, mechanistic models are a subset of associative models;
+% this subset relationship is enough to see that one cannot expect the class of mechanistic models to provide the highest levels of statistical fit.
+% In general one should expect a statistical cost to the constraint of requiring mechanistic interpretation of the model.
+One question that follows is whether it is useful to assess mechanistic models by their statistical fit, if this is not their primary purpose?
+To what extent can we say the data support a scientific model ($A$) over another model ($B$) based on improved fit to data, if neither fit better than an associative model ($C$)?
+% An answer is to think in terms of the shortfall between a mechanistic model and the best known associative model.
+% If the shortfall is small, on a suitable scale, then the mechanistic model may be considered a valid view of the system from the point of view of these data.
+% While the ARMA(2, 1) certainly will not automatically provide the best fit of all associative models,
+% This does not demonstrate the correctness of the mechanistic interpretation, but also it does not refute it.
+% If the shortfall is large, we conclude that there is some substantial aspect of the system that is not incorporated into the model.
+% we cannot necessarily tell which part of the model is invalid, but we know there is a weakness somewhere.
+% To assess whether shortfalls are small or large, we may consider using formal statistical hypothesis tests or model selection techniques.
+% These can give us a scale for chance variation: a difference small enough to be explained by randomness in the data generating mechanism should not be considered as strong evidence.
+% However, the comparison of primary interest is one of practical significance rather than statistical significance.
+% We would like a mechanistic model whose statistical fit is close to the best associative models regardless of whether there is enough statistical power to distinguish the difference.
+
+% I like the above addition because it provides a little bit more of the philosophy behind the paragraph below. Perhaps we can combine them? - Anna
+
+When comparing mechanistic models to non-mechanistic benchmarks, we do not require the benchmark model to optimize a measure of statistical fit, nor do we suggest that a mechanistic model must have superior fit to the benchmark in order to be of scientific interest.
+It is possible that a successful mechanistic model may have better statistical fit than simple benchmarks such as log-linear models;
+still, the specific threshold of beating, or failing to beat, an arbitrary benchmark is similarly arbitrary.
+In this case, however, a large shortfall in the fit of a mechanistic model to that of a log-linear model may demonstrate room for substantial improvement.
 The use of a common benchmark may also be beneficial when developing models with varying spatial scale, as direct comparisons between models fit to data with different levels of spatial aggregation are meaningless (e.g.
 comparing Model~1 to Model~3).
-\eic{I DON'T QUITE UNDERSTAND THIS YET.
-BENCHMARKS TEND TO APPLY TO SPECIFIC DATA SETS, SO CANNOT READILY BE USED TO COMPARE DIFFERENT LEVELS OF AGGREGATION.
-LIKELIHOOD BENCHMARKS CAN BE USED TO STUDY THE VALUE OF EXPLICIT SPATIAL COUPLING.
-HOW TO BENCHMARK ACROSS AGGREGATION LEVELS IS, I THINK, SOMETHING OF AN OPEN PROBLEM.}
 Likelihoods of Models~1--3 and their respective ARMA benchmark models are provided in Table~\ref{tab:likes}.
@@
   h2_epi,
-  params = h2_fit$epi_params
-)
-
-# Objective function of model 2, endemic phase
-end_ofun <- traj_objfun(
-  h2_end,
-  params = h2_fit$end_params
-)
-
-mod2_epi_ll <- -epi_ofun(par = h2_fit$epi_params)
-mod2_end_ll <- -end_ofun(par = h2_fit$end_params)
-
-mod2_ll <- mod2_epi_ll - sum(apply(log(h2_epi@data + 1), 1, sum)) + (mod2_end_ll - sum(apply(log(h2_end@data + 1), 1, sum, na.rm = TRUE)))
+  params = h2_fit$h2_params
+)
+
+mod2_epi_ll <- -epi_ofun(par = h2_fit$h2_params)
+# mod2_end_ll <- -end_ofun(par = h2_fit$end_params)
+
+mod2_ll <- mod2_epi_ll - sum(log(h2_epi@data + 1), na.rm = TRUE)
 
@@
 # model 2 AIC
-mod2_n_params <- 21  # Number of parameters fit in Model 2
+mod2_n_params <- 21  # TODO: adjust Number of parameters fit in Model 2
 mod2_aic <- 2 * mod2_n_params - 2 * mod2_ll
@@
     haiti2_vaccScenario(
-      end_params = h2_fit$end_params,
-      epi_params = h2_fit$epi_params
+      h2_params = h2_fit$h2_params
     )
@@
 
-\hspace{-10mm}\begin{tabular}{|l|l@{\hskip3pt}c@{\hskip3pt}|l@{\hskip3pt}c@{\hskip3pt}|l@{\hskip3pt}c@{\hskip3pt}|}
+\hspace{-10mm}\begin{tabular}{|l@{}|l@{}c@{}|l@{}c@{}|l@{}c@{}|}
 \hline
@@
   & \eqref{model1:toR}
-  & {\fixed $\muIR^{-1}=\Sexpr{myround(365/h2_fit$epi_params["gamma"],1)}$}
+  & {\fixed $\muIR^{-1}=\Sexpr{myround(365/h2_fit$h2_params["gamma"],1)}$}
   & \eqref{model2:mu_IR}
@@
   & \eqref{model1:EA}
-  & {\fixed $\muEI^{-1}=\Sexpr{myround(365/h2_fit$epi_params["gammaE"],1)}$}
+  & {\fixed $\muEI^{-1}=\Sexpr{myround(365/h2_fit$h2_params["gammaE"],1)}$}
   & \eqref{model2:mu_EI}
@@
     $\begin{array}{l}
-    \muRS^{-1}= \Sexpr{signif(1/h2_fit$epi_params["sigma"],2)} \\
-     \omega_1^{-1} = \Sexpr{myround(1/h2_fit$epi_params["Omega1"],1)} \\
-    \omega_2^{-1} = \Sexpr{myround(1/h2_fit$epi_params["Omega2"],1)}
+    \muRS^{-1}= \Sexpr{signif(1/h2_fit$h2_params["sigma"],2)} \\
+     \omega_1^{-1} = \Sexpr{myround(1/h2_fit$h2_params["Omega1"],1)} \\
+    \omega_2^{-1} = \Sexpr{myround(1/h2_fit$h2_params["Omega2"],1)}
     \end{array}$
@@
       \hspace{-1mm} \fixed{\vaccineEfficacy_{1:4} = (
-        \Sexpr{myround(1 - h2_fit$epi_params["VE1"] * 0.4688, 2)},
-        \Sexpr{myround(1 - h2_fit$epi_params["VE2"] * 0.4688, 2)}} \\
-        \fixed{\Sexpr{myround(1 - h2_fit$epi_params["VE1"], 2)},
-        \Sexpr{myround(1 - h2_fit$epi_params["VE2"], 2)}})
+        \Sexpr{myround(1 - h2_fit$h2_params["VE1"] * 0.4688, 2)},
+        \Sexpr{myround(1 - h2_fit$h2_params["VE2"] * 0.4688, 2)}} \\
+        \fixed{\Sexpr{myround(1 - h2_fit$h2_params["VE1"], 2)},
+        \Sexpr{myround(1 - h2_fit$h2_params["VE2"], 2)}})
         \end{array}$
@@
   & (\ref{model1:EI}-\ref{model1:EA})
-  & {\fixed $\symptomFrac=\Sexpr{myround(h2_fit$epi_params["k"],1)}$}
+  & {\fixed $\symptomFrac=\Sexpr{myround(h2_fit$h2_params["k"],1)}$}
   & \eqref{model2:mu_EI}
@@
   & \eqref{model1:lambda}
-  & $\transmission=$\Sexpr{signif(h2_fit$epi_params["Beta"],3)}
+  & $\transmission=$\Sexpr{signif(h2_fit$h2_params["Beta"],3)}
   & \eqref{model2:lambda}
@@
   & $\begin{array}{l}
-    {\fixed \Wsat = \Sexpr{signif(h2_fit$epi_params["Sat"],2)} }
+    {\fixed \Wsat = \Sexpr{signif(h2_fit$h2_params["Sat"],2)} }
     \\
-    \beta_W= \Sexpr{signif(h2_fit$epi_params["BetaW"],3)}
+    \beta_W= \Sexpr{signif(h2_fit$h2_params["BetaW"],3)}
     \end{array}$
@@
   &
-  & $\Wshed = $ \Sexpr{signif(h2_fit$epi_params["Mu"],3)}
+  & $\Wshed = $ \Sexpr{signif(h2_fit$h2_params["Mu"],3)}
   & \eqref{model2:to_W}
@@
   &
-  & $\Wremoval^{-1} = $ \Sexpr{signif(52/h2_fit$epi_params["Delta"],3)}
+  & $\Wremoval^{-1} = $ \Sexpr{signif(52/h2_fit$h2_params["Delta"],3)}
   & \eqref{model2:from_W}
@@
   & (S16) % TODO: Make sure manual equations match!
-  & {\fixed $\reportRate=\Sexpr{myround(h2_fit$epi_params["Rho"],2)}$}
+  & {\fixed $\reportRate=\Sexpr{myround(h2_fit$h2_params["Rho"],2)}$}
   & (S17)

### 2022-09-12 ms.Rnw
Compiled merge
minor edits

@@
 
+\newcolumntype{t}{>{\tiny}c}
+
 \endlocaldefs
@@
 potential effect of prospective oral cholera vaccine (OCV) programs.
-OCV is accepted as a safe and effective tool for controlling the spread of cholera, however the available stockpile and production of OCV doses is insufficient to meet global needs \citep{lee20, pezzoli20} \eic{NEW CHOLERA VACCINES, NEW STUDIES OF EXISTING VACCINES, AND INCREASED PRODUCTION VOLUMES ALL AROSE DURING 2010-2019.
-MAYBE SOMETHING LIKE:
-Advances in OCV technology and vaccine availability raised the possibility of planning a national vaccination program \citep{lee20, pezzoli20}}
+While OCV is accepted as a safe and effective tool for controlling the spread of cholera, the global stockpile of OCV doses remains limited \citep{pezzoli20}.
+Advances in OCV technology and vaccine availability, however, raised the possibility of planning a national vaccination program \citep{lee20}.
 In the study, certain data were shared between the groups, including demography and vaccination history; vaccine efficacy was also fixed at a shared value between groups.
@@
 Number of &
-  $\Sexpr{as.character(mod1_n_params)}$ & $\Sexpr{as.character(mod2_n_params)}$ & $\Sexpr{as.character(mod3_n_params)}$ & $\Sexpr{mod3_spatPomp_n_params}$ \\
- Fit Parameters & (\Sexpr{as.character(mod1_lee_n_params)}) & ($\Sexpr{as.character(h2_fit$leeFit_n_params)}$) & (29) & ($\Sexpr{as.character(lee3_results$n_params, 1)}$)
+  $\Sexpr{as.character(mod1_n_params)}$ & $\Sexpr{as.character(mod2_n_params)}$ & \multirow{2}{*}{$\Sexpr{as.character(mod3_n_params)}$} & $\Sexpr{mod3_spatPomp_n_params}$ \\
+ Fit Parameters & (\Sexpr{as.character(mod1_lee_n_params)}) & ($\Sexpr{as.character(h2_fit$leeFit_n_params)}$) && ($\Sexpr{as.character(lee3_results$n_params, 1)}$)
 \\
@@
 
-\eic{I THINK WE WILL HAVE TO EXPLAIN THIS CAREFULLY IF THE POINT IS TO APPEAR NONTRIVIAL.
+% \eic{I THINK WE WILL HAVE TO EXPLAIN THIS CAREFULLY IF THE POINT IS TO APPEAR NONTRIVIAL.
 IT IS WORTH NOTING THAT LEE ET AL DID NOT PROPERLY FORECAST FROM THE BEST ESTIMATE OF THE CURRENT STATE.
 HOWEVER, IT WILL BE OBVIOUS TO MOST SATISTICAL READERS THAT THIS IS NOT THE BEST THING TO DO.
-IT MAY NOT BE SO OBVIOUS THAT THE DIFFERENCE IS IMPORTANT IF AND ONLY IF A DETERMINISTIC MODEL IS INADEQUATE.
-THE QUESTION OF WHETHER TO USE A DETERMINISTIC MODEL REMAINS OF CURRENT INTEREST...}
-As an example, we note that simulations from a well-fit mechanistic model may closely resemble the observed data $y_{1:N}^*$.
-In such comparisons, these simulations are random draws from the complete estimated joint distribution $f_{\bm{X}^{(\modelCounter)}_{0:N}, \bm{Y}^{(\modelCounter)}_{1:N}}\left(x_{0:N}, y_{1:N}\mid \hat{\paramVec}\right)$, where $\modelCounter$ indexes the model used for simulations.
-It can therefore be tempting to use simulations from this model up to time $N+s$, $f_{\bm{X}^{(\modelCounter)}_{0:N+s}, \bm{Y}^{(\modelCounter)}_{1:N+s}}\left(x_{0:N+s}, y_{1:N+s}\mid \hat{\paramVec}\right)$, with $s\geq 0$ to project the dynamic system up to a future time $N+s$, as it has been done in previous studies \citep{lee20,???}.
-This approach, however, does not take advantage of the information about the state of the system that is contained in the observed data.
-Note that each of the models in question (and, more generally, all models that are described as POMPs) are Markovian, that is, the history of the process $\{\bm{X}^{(\modelCounter)}_s, s\leq t\}$ for model $\modelCounter$ is uninformative about the future of the process $\{\bm{X}^{(\modelCounter)}_s, s\geq t\}$, given the current state $\bm{X}^{(\modelCounter)}_t$.
-In other words, observing the data $\bm{Y}^{(\modelCounter)}_N = y_N^*$ at time $N$ provides more information about the future state of the system than the initial conditions.
-In this case, draws from the conditional density $f_{\bm{X}^{(\modelCounter)}_{N:s}, \bm{Y}^{\modelCounter}_{N:s}}\big(x_{N:s}, y_{N:s}\mid \hat{\paramVec}, \bm{X}^{(\modelCounter)}_{N} = \bm{x}^{(\modelCounter)}_N\big)$ should be preferred as forecasts, as these simulations account for the most recent known state of the system.
-
-$\bm{X}^{(\modelCounter)}_N$, however, is unobservable and therefore draws from the desired conditional density are unobtainable.
-Informed estimates of $\bm{X}^{(\modelCounter)}_N$ given the observed data $\bm{Y}^{(\modelCounter)}_{1:N} = y_{1:N}^*$ can easily be obtained, however, via the filtering distribution.
+IT MAY NOT BE SO OBVIOUS THAT THE DIFFERENCE IS IMPORTANT IF AND ONLY IF A DETERMINISTIC MODEL IS INADEQUATE.}
+
+% As an example, we note that simulations from a well-fit mechanistic model may closely resemble the observed data $y_{1:N}^*$.
+% In such comparisons, these simulations are random draws from the complete estimated joint distribution $f_{\bm{X}^{(\modelCounter)}_{0:N}, \bm{Y}^{(\modelCounter)}_{1:N}}\left(x_{0:N}, y_{1:N}\mid \hat{\paramVec}\right)$, where $\modelCounter$ indexes the model used for simulations.
+% It can therefore be tempting to use simulations from this model up to time $N+s$, $f_{\bm{X}^{(\modelCounter)}_{0:N+s}, \bm{Y}^{(\modelCounter)}_{1:N+s}}\left(x_{0:N+s}, y_{1:N+s}\mid \hat{\paramVec}\right)$, with $s\geq 0$ to project the dynamic system up to a future time $N+s$, as it has been done in previous studies \citep{lee20,???}.
+% This approach, however, does not take advantage of the information about the state of the system that is contained in the observed data.
+% Note that each of the models in question (and, more generally, all models that are described as POMPs) are Markovian, that is, the history of the process $\{\bm{X}^{(\modelCounter)}_s, s\leq t\}$ for model $\modelCounter$ is uninformative about the future of the process $\{\bm{X}^{(\modelCounter)}_s, s\geq t\}$, given the current state $\bm{X}^{(\modelCounter)}_t$.
+% In other words, observing the data $\bm{Y}^{(\modelCounter)}_N = y_N^*$ at time $N$ provides more information about the future state of the system than the initial conditions.
+% In this case, draws from the conditional density $f_{\bm{X}^{(\modelCounter)}_{N:s}, \bm{Y}^{\modelCounter}_{N:s}}\big(x_{N:s}, y_{N:s}\mid \hat{\paramVec}, \bm{X}^{(\modelCounter)}_{N} = \bm{x}^{(\modelCounter)}_N\big)$ should be preferred as forecasts, as these simulations account for the most recent known state of the system.
+The most recently available information about a dynamic system should be used to obtain reliable forecasts.
+Using state space model $m$, this means that in order to obtain a forecast up to time $N+s$, where $s\geq 0$ and $N$ is the index for the last available data, we should simulate from the conditional density $f_{\bm{X}^{(\modelCounter)}_{N:s}, \bm{Y}^{\modelCounter}_{N:s}}\big(x_{N:s}, y_{N:s}\mid \hat{\paramVec}, \bm{X}^{(\modelCounter)}_{N} = \bm{x}^{(\modelCounter)}_N\big)$.
+We note, however, that $\bm{X}^{(\modelCounter)}_N$ is unobservable and therefore draws from the desired conditional density are unobtainable.
+Informed estimates of $\bm{X}^{(\modelCounter)}_N$ given the observed data $\bm{Y}^{(\modelCounter)}_{1:N} = y_{1:N}^*$ can easily be obtained using the filtering distribution.
 Let $\hat{\bm{X}}^{(\modelCounter), i}_N, \, i \in {1, 2, \ldots, J}$ be $\iid$ draws from the filtering distribution at time $N$, with density $\hat{\bm{X}}^{(\modelCounter), i}_N \sim f_{\bm{X}^{(\modelCounter)}_N|\bm{Y}^{(\modelCounter)}_{1:N}}(x_N \mid \hat{\paramVec}, y_{1:N}^*)$.
 A single model forecast can then be obtained by simulating from the model $f_{\bm{X}^{(\modelCounter)}_{N:s}, \bm{Y}^{(\modelCounter)}_{N:s}}\big(x_{N:s}, y_{N:s}\mid \hat{\paramVec}, \bm{X}^{(\modelCounter)}_{N} = \hat{\bm{X}}^{(\modelCounter), i}_N\big)$.
-Intuitively, simulating the model starting at the filtering distribution of the most recently available time point is a more appropriate way to project a stochastic dynamic system into the future, as it is not expected that each simulation from initial conditions will result in a latent state at time $N$ consistent with the model and the observation $\bm{Y}^{(\modelCounter)}_N = y_N^*$.
+Obtaining forecasts in this fashion allows us to get results that are consistent with the model assumptions and the observed data.
+% Intuitively, simulating the model starting at the filtering distribution of the most recently available time point is a more appropriate way to project a stochastic dynamic system into the future, as it is not expected that each simulation from initial conditions will result in a latent state at time $N$ consistent with the model and the observation $\bm{Y}^{(\modelCounter)}_N = y_N^*$.
 In this particular case study, projecting the future state of the cholera epidemic in Haiti starting from the draws from the filtering distribution allows the proposed models to benefit from the fact that very few cholera cases had been observed between 2018 and 2019, and that cases appear to be decreasing (i.e., the number of susceptible individuals may be small).
+This alone partially explains the unrealistic forecasts of \cite{lee20}, which were obtained by simulating the model from initial conditions;
+Table~S7 of their supplement material shows that of their simulations that were consistent with observing zero cases in 2019, all forecasts resulted in the elimination of cholera.
+\eic{THE QUESTION OF WHETHER TO USE A DETERMINISTIC MODEL REMAINS OF CURRENT INTEREST...}
 
@@
 Fig.~\ref{fig:h1SeasRain} shows that the estimated seasonal transmission rate $\transmission$ mimics the rainfall dynamics in Haiti, despite Model~1 not having access to rainfall data.
-This result provides evidence that rainfall is a potential driver of cholera infections in Haiti.
+This result provides evidence that rainfall plays an important role in cholera transmission in Haiti, which is consistent with previous studies \citep{eisenberg13,rinaldo12}.
+We also note the increase in the transmission rate during the Fall, which was also noticed at an earlier stage of the epidemic \citep{rinaldo12}.
 
@@
 The results of these projections are summarized in Figs.~\ref{fig:Mod1Scenarios}--\ref{fig:elimProbs}.
-These results suggest that cholera elimination was likely, even without increased vaccination efforts, which is consistent with observed reality \citep{ferguson22}.
-\eic{I LIKE THE IDEA OF REFERENCING POST-EPIDEMIC REFLECTIONS.
-UNLESS THERE ARE NO ALTERNATIVES, IT IS BEST TO AVOID CITING A NON-PEER-REVIEWED ARTICLE SUCH AS THIS.
-HOW ABOUT \citep{piarroux22} OR \citep{trevisin22}?}
+These results suggest that cholera elimination was likely, even without increased vaccination efforts, which is consistent with observed reality \citep{trevisin22}.
 
@@
 \begin{table}
-\small
-
-\hspace{-10mm}\begin{tabular}{|l@{}|l@{}c@{}|l@{}c@{}|l@{}c@{}|}
+% \small
+\resizebox{\textwidth}{!}{
+\hspace{-10mm}\begin{tabular}{|l@{}|l@{}t@{}|l@{}t@{}|l@{}t@{}|}
 \hline
@@
 \hline
-\end{tabular}
+\end{tabular}}
 \caption{References to the relevant equation are given in parentheses.

### 2022-09-13 ms.Rnw
Changes to Model2 description, some responses to Ed, other minor edits

@@
 \label{model2:lambda}
-\lambda_{u}(t) = 0.5\big(1+\seasAmplitude \cos(2\pi t)\big)
+\lambda_{u}(t) = 0.5\big(1+\seasAmplitude \cos(2\pi t + \phaseParm)\big)
 \frac{\beta_W\, W_u(t)}{ \Wsat  + W_u(t)} +
@@
 The cosine term in Eq.~\myeqref{model2:lambda} accounts for annual seasonality, with a phase parameter $\phaseParm$.
-The original implementation of Model~2 in \cite{lee20} fixes the phase at $\phaseParm = 0$.
+The implementation of Model~2 in \cite{lee20} fixes $\phaseParm = 0$.
 
@@
 Furthermore, because each draw from the joint distribution represents a potential outcome of the dynamic system, best/worst case scenarios under the assumptions of the model can be easily obtained via simulation.
-
-Once a model for the system has been proposed, the parameter vector $\paramVec$ needs to be estimated using the observed data.
+The added benefits of a stochastic model are coupled with statistical difficulty of fitting the parameter vector $\paramVec$ to observed data.
+
 There exist several algorithms, both frequentist and Bayesian, that can be used to obtain estimates of the parameters in stochastic dynamic models.
@@
 To do this, we perform a profile-likelihood search on the parameter $\transmissionTrend$ and obtain a confidence interval via a Monte Carlo Adjusted Profile (MCAP) \citep{ionides17}.
-Similar to Model~2, this model was originally implemented in two parts: an epidemic phase from October 2010 through March 2015, and an endemic phase from March 2015 onward.
-As before, we continue to allow the re-estimation of some model parameters at the start of the endemic phase ($\reportRate, \sigmaProc^2$ and $\obsOverdispersion$) but require that the latent Markov process $X(t)$ carry over from one phase into the next.
+Model~1 was implemented in \cite{lee20} by fitting two distinct phases: an epidemic phase from October 2010 through March 2015, and an endemic phase from March 2015 onward.
+We similarly allow the re-estimation of process and measurement overdispersion parameters ($\sigmaProc^2$ and $\obsOverdispersion$), and require that the latent Markov process $X(t)$ carry over from one phase into the next.
 The resulting confidence interval for $\transmissionTrend$ is $(\Sexpr{myround(mcap_results$ci[1], digits = 3)}, \Sexpr{myround(mcap_results$ci[2], digits = 3)})$, with the full results displayed in Fig.~\ref{fig:betat}.
@@
 
-\citet{lee20} originally fit two versions of model~2 based on a presupposed change in cholera transmission from a epidemic phase to endemic phase that occurred in March, 2014.
-We follow their decision to re-estimate model parameters at this break-point, but we include a requirement that the latent state $\bm{X}^{(2)}(t)$ at the start of the endemic period must be the same as the state at the end of the epidemic period.
-This additional constraint is sensible for a mechanistic interpretation of the latent state, though it adds difficulty to the task of obtaining a model fit that closely resembles the observed data.
-To combat this added difficulty, and to allow for a possible shift in the seasonality component in the force of infection, we introduce an additional phase parameter $\phaseParm$ in Eq.~\ref{model2:lambda}.
-% \begin{equation}
-% \lambda_{u}(t) = 0.5\big(1+\seasAmplitude \cos(2\pi t + \phaseParm)\big)
-% \frac{\beta_W\, W_u(t)}{ \Wsat  + W_u(t)} +
-% \beta \left\{\sum_{d=0}^4 I_{ud}(t) + \asymptomRelativeInfect \sum_{d=0}^4 A_{ud}(t) \right\}\label{eq:model2:phase}
-% \end{equation}
-% \eic{WE COULD SAVE AN EQUATION BY WRITING THE MORE GENERAL ONE INITIALLY AND SAYING THAT THEY ASSUMED $\phaseParm=0$}
-
-% We also note that, without the inclusion of a phase parameter, there is an implicit assumption is that force of infection is greatest around January and weakest around June, and the addition of this parameter allows for the testing of this assumption.
-We further increase model flexibility by fitting the parameter $\Wshed$ rather than treating it as fixed.
+\cite{lee20} fit two versions of model~2 based on a presupposed change in cholera transmission from a epidemic phase to endemic phase that occurred in March, 2014.
+The inclusion of a change-point in model states and parameters increases the flexibility of the model and hence the ability to fit the observed data.
+The increase in model flexibility, however, results in hidden states that are inconsistent between model phases.
+A naive solution to this would be to estimate only model parameters in separate phases and not the hidden states, but this may also result in hidden states in one phase of the model that are inconsistent with the other phase of the outbreak, since only the data from one phase is used to fit model parameters.
+
+The inclusion of a model break-point in \cite{lee20} is perhaps due to a challenging feature of fitting a deterministic model via least squares: discrepancies between model trajectories and observed case counts in highly infectious periods of a disease outbreak will result in greater penalty than the discrepancies between model trajectories and observed case counts in times of relatively low infectiousness, resulting in a bias towards accurately describing periods of high infectiousness.
+This issue is particularly troublesome for this case study: the inability to accurately fit times of low infectiousness may result in poor model forecasts, as few cases of cholera were observed the last few years epidemic.
+
+To combat this issue, we fit the model to log-transformed case counts, where the difference between periods of high and low infectiousness is less drastic.
+Other solutions include changing the measurement model to include overdispersion, as was done in Models~1 and~3.
+Here we chose to fit the model to transformed case counts rather than adding overdispersion to the measurement model with the goal of making the fewest changes to the model in \cite{lee20} as possible.
+In practice, however, the inclusion of overdispersion in the measurement process may be preferred \jwc{REFS???}.
+
 We implemented this model using the \code{spatPomp} \code{R} package \citep{asfaw21github}.
@@
 In this model, both the latent and observable processes can be factored into department specific processes which interact with each other.
-We denote the latent and measurement processes for Model~3 as $\bm{X}^{(3)}(t_{0:N}) = X^{(3)}_{1:U, 0:N}$, and $\bm{Y}^{(3)}(t_{1:N}) = Y_{1:U, 1:N}$ \eic{IS THIS NEEDED HERE?}.
+% We denote the latent and measurement processes for Model~3 as $\bm{X}^{(3)}(t_{0:N}) = X^{(3)}_{1:U, 0:N}$, and $\bm{Y}^{(3)}(t_{1:N}) = Y_{1:U, 1:N}$ \eic{IS THIS NEEDED HERE?}.
 The decision to model the system via metapopulation models versus a model aggregated to a larger spatial scale is one of great scientific interest, and evidence for the former approach has been provided in previous studies \citep{king15}.
 Note that this evidence alone does not automatically discredit conclusions drawn via nationally aggregated models, as there are also good reasons to prefer a simple model over a complex one \citep{saltelli20,green15}.
-Still, researchers that intend to use a mechanistic model to describe a dynamic system should design the model to be as realistic as their scientific understanding of the system and their computational abilities permit \eic{THIS SENTENCE SEEMS TO DISMISS THE ARGUMENTS IN THE PREVIOUS SENTENCE.
+% Still, researchers that intend to use a mechanistic model to describe a dynamic system should design the model to be as realistic as their scientific understanding of the system and their computational abilities permit \eic{THIS SENTENCE SEEMS TO DISMISS THE ARGUMENTS IN THE PREVIOUS SENTENCE.
 PERHAPS THIS SENTENCE CAN JUST BE DELETED}.
-Fitting scientifically flexible metapopulation models, however, remains a challenging statistical problem;
-this is due to the fact that the approximation error of particle filters grows exponentially in the dimension of the model \citep{rebeschini15,park20}.
-Algorithms that are based on the particle filter therefore become computationally intractable as the number of spatial units increase.
+
+Fitting scientifically flexible metapopulation models is a challenging statistical problem.
+In particular, parameter estimation techniques based on the particle filter become computationally intractable as the number of spatial units increase.
+This is a result of the approximation error of particle filters growing exponentially in the dimension of the model \citep{rebeschini15,park20}.
 
 Parameters that must be fit in Model~3 are primarily shared between each department, the exception to this being the parameters $\beta_{W_u}$, and $\beta_u$, which are unique for each department $u \in \seq{1}{10}$.
-To fit this model, \citet{lee20} simplified the parameter estimation problem by fitting independent department-level models to the data.
+To avoid the parameter estimation issue in high-dimensional models, \citet{lee20} simplified the problem by fitting independent department-level models to the data.
 The shared parameters were calibrated using the cholera incidence data from Artibonite, and the department-specific parameters ($\beta_{W_u}$ and $\beta_u$) were fit using the data from their respective department.
-Reducing a spatially coupled \eic{not spatially heterogeneous, as previously written} model to individual units in this fashion requires special treatment of any interactive mechanisms between spatial units, such as found in Eq.~\myeqref{eq:model3:foi}.
+Reducing a spatially coupled model to individual units in this fashion requires special treatment of any interactive mechanisms between spatial units, such as found in Eq.~\myeqref{eq:model3:foi}.
 In particular, when considering a model for department $u$, the values $I_\nu(t)$ and $A_\nu(t)$, are unknown for $u \neq v \in \seq{1}{10}$.
@@
 # model 2 AIC
-mod2_n_params <- 21  # TODO: adjust Number of parameters fit in Model 2
+mod2_n_params <- 16  # TODO: adjust Number of parameters fit in Model 2
 mod2_aic <- 2 * mod2_n_params - 2 * mod2_ll
@@
 A single model forecast can then be obtained by simulating from the model $f_{\bm{X}^{(\modelCounter)}_{N:s}, \bm{Y}^{(\modelCounter)}_{N:s}}\big(x_{N:s}, y_{N:s}\mid \hat{\paramVec}, \bm{X}^{(\modelCounter)}_{N} = \hat{\bm{X}}^{(\modelCounter), i}_N\big)$.
-Obtaining forecasts in this fashion allows us to get results that are consistent with the model assumptions and the observed data.
+Obtaining forecasts in this fashion allows us to obtain results that are consistent with the model assumptions and the observed data.
 % Intuitively, simulating the model starting at the filtering distribution of the most recently available time point is a more appropriate way to project a stochastic dynamic system into the future, as it is not expected that each simulation from initial conditions will result in a latent state at time $N$ consistent with the model and the observation $\bm{Y}^{(\modelCounter)}_N = y_N^*$.
@@
 % \end{figure}
-
-\jwc{PARAGRAPH REMOVED}
-% Despite these results suggesting that cholera was likely to be eliminated from Haiti without an increase in vaccination effort, we in no way suggest that vaccinations shouldn't be used to combat cholera---or other infectious diseases---in other parts of the world.
-% In fact, our analysis shows that an increase in vaccination does drastically reduce the negative effects of a worst-case scenario.
-% Avoid comments on vaccination policy recommendations - Jesse
 

### 2022-09-13 si/si.Rnw
Changes to Model2 description, some responses to Ed, other minor edits

@@
   \label{model1:meas}
-  y_{t} \mid \Delta N_{E_{k}I_{k}} = z_{t} \sim \text{NB}\left(\reportRate z_{t}, \obsOverdispersion \right)
+  y_{t} \mid \Delta N_{E_{\vaccCounter}I_{\vaccCounter}}(t) = z(t) \sim \text{NB}\left(\reportRate z_(t), \obsOverdispersion \right)
 \end{equation}
 
+Where $y_t$ is the reported cholera cases at time $t \in 1:N$ and $\Delta N_{E_{\vaccCounter}I_{\vaccCounter}}(t)$ is the number of individuals who moved from compartment $E_\vaccCounter$ to $I_\vaccCounter$ since observation $t-1 \in 0:N-1$.
+
 \subsection{Model~2}
 
-Model~2 was fit via trajectory matching and therefore does not have an explicit measurement model.
-Implicitly, however, this fitting process is equivalent to having a Gaussian measurement process.
-Therefore the measurement model can be expressed as Eq.~\myeqref{model2:meas}.
+As mentioned in the main text, Model~2 was fit using log-transformed observed cases.
+We fit Model~2 using the subplex algorithm in the \code{subplex} package, using
+a Gaussian measurement model (Eq.~\myeqref{model2:meas}) on the log-transformed cases as the loss function.
+
 \begin{equation}
   \label{model2:meas}
-  y_{u, t} \mid \Delta N_{I_{ud}R_{ud}} = z_{u, t} \sim \text{N}\left(\reportRate z_{u, t}, \obsOverdispersion^2 \right)
+  \log\big(y_{u, t} + 1\big) \mid \bm{X}^{(2)}(t) = \bm{x}^{(2)}(t) \sim \text{N}\left(\log \big(\reportRate \sum_{\vaccCounter} I_{u, \vaccCounter}(t) + 1\big), \obsOverdispersion^2 \right)
 \end{equation}
 
-Where $\Delta N_{I_{ud}R_{ud}}$ is the number of individuals who move from compartment $I_{ud}$ to $R_{ud}$, for unit $u \in 1:U$ and vaccination cohort $d \in 0:5$.
-We also note here that the variance parameter $\obsOverdispersion^2$ is separately fit for the epidemic and endemic phases.
+Where $\sum_{\vaccCounter} I_{u, \vaccCounter}(t)$ is the total number of individuals who are infected at time $t$ in unit $u \in 1:U$, for all vaccination campaigns $\vaccCounter$.
+We note that fitting a model with this loss function is equivalent to fitting using least squares, with $\log(y_{u, t} + 1)$ as the response variable.
 
@@
   \label{model3:meas}
-  y_{u, t} \mid \Delta N_{S_{ud}I_{ud}} = z_{u, t} \sim \text{NB}\left(\reportRate z_{u, t}, \obsOverdispersion \right)
+  y_{u, t} \mid \Delta N_{S_{u \vaccCounter}I_{u \vaccCounter}}(t) = z_{u}(t) \sim \text{NB}\left(\reportRate z_{u}(t), \obsOverdispersion \right)
 \end{equation}
+
+Where $\Delta N_{S_{u \vaccCounter}I_{u \vaccCounter}}(t)$ is the number of individuals who moved from compartment $S_{u \vaccCounter}$ to $I_{u \vaccCounter}$ since observation $t-1 \in 0:N-1$.
 
@@
 The fitted values of the reporting rate---before and after January 2018---were $0.97$ and $0.097$, respectively.
-This major change in reporting rate alone could have been the cause that Model~3 originally failed to predict the eradication of cholera, as an overnight change from near perfect to almost non-existent reporting forces the model to explain the observed decrease in cases as a decrease in the reporting of cases rather than of prevalence of cases.
+This major change in reporting rate alone could explain why Model~3 in \cite{lee20} failed to predict the eradication of cholera.
+An overnight change from near perfect to almost non-existent reporting forces the model to explain the observed reduction in reported cases as a decrease in the reporting of cases, rather than a decrease in the prevalence of cholera.
 This shift was justified by a "change of the case definition that occurred on January 1st, 2018";
 this claim was not cited, and we could find no evidence that such a drastic change in the reporting rate would be warranted.
+We therefore do not allow a change in reporting rate when fitting Model~3.
 

### 2022-09-15 ms.Rnw
Major structural changes

@@
 
+# Check if RUN_LEVEL == 3 results exists, if so, use them.
+# Include a flag to check if RUN_LEVEL == 3. If so, we might want to recompute.
+# TODO: Only save parameters and likelihoods, not entire MIF objects.
+if (RUN_LEVEL != 3 & file.exists("model1/run_level_3/lee1_epi_fit.rds")) {
+  lee1_epi_mif <- readRDS("model1/run_level_3/lee1_epi_fit.rds")
+} else {
   lee1_epi_mif <- bake(
@@
   )
+}
 
@@
 
+if (RUN_LEVEL != 3 & file.exists("model1/run_level_3/lee1_epi_evals.rds")) {
+  lee_epi_ll <- readRDS("model1/run_level_3/lee1_epi_evals.rds")
+} else {
   lee_epi_ll <- bake(
@@
 
+# TODO: Only save parameter values and "likelihoods".
+if (RUN_LEVEL != 3 & file.exists("model1/run_level_3/lee1_end_fit.rds")) {
+  lee1_end_mif <- readRDS("model1/run_level_3/lee1_end_fit.rds")
+} else {
   lee1_end_mif <- bake(
@@
 
+if (RUN_LEVEL != 3 & file.exists("model1/run_level_3/lee1_end_evals.rds")) {
+  lee_end_ll <- readRDS("model1/run_level_3/lee1_end_evals.rds")
+} else {
   lee_end_ll <- bake(
@@
 We proceed by introducing Models~1--3 in Sec.~\ref{sec:models};
-in Sec.~\ref{sec:methods}, we present a methodological approach to examining and refining these models.
-In Sec.~\ref{sec:results}, we use improved model fits to project cholera incidence in Haiti under various vaccination campaigns.
-We then conclude with a discussion on the use of mechanistic models to inform policy decisions in Sec.~\ref{sec:discussion}.
+in Sec.~\ref{sec:methods}, we present a methodological approach to examining and refining these models, and then use improved model fits to project cholera incidence in Haiti under various vaccination scenarios.
+We then conclude with recommendations on the use of mechanistic models to inform policy decisions in Sec.~\ref{sec:advice}.
 
@@
 Suitable methodology for calibrating model parameters may depend on other modeling decisions.
-While this section is focused on the elements of $\paramVec$ that are calibrated to data, we note that the decision of which elements to fix and which to estimate has consequences for model interpretability, as discussed in Sec.~\ref{sec:results}.
+% While this section is focused on the elements of $\paramVec$ that are calibrated to data, we note that the decision of which elements to fix and which to estimate has consequences for model interpretability, as discussed in Sec.~\ref{sec:science}.
 
@@
 % We denote the latent and measurement processes for Model~3 as $\bm{X}^{(3)}(t_{0:N}) = X^{(3)}_{1:U, 0:N}$, and $\bm{Y}^{(3)}(t_{1:N}) = Y_{1:U, 1:N}$ \eic{IS THIS NEEDED HERE?}.
-The decision to model the system via metapopulation models versus a model aggregated to a larger spatial scale is one of great scientific interest, and evidence for the former approach has been provided in previous studies \citep{king15}.
+The decision to model the system via metapopulation models versus a model aggregated to a larger spatial scale is one of great scientific interest.
+Evidence for the former approach has been provided in previous studies \citep{king15}, including the specific case of heterogeneity between Haitian departments in cholera transmission \citep{collins14}.
 Note that this evidence alone does not automatically discredit conclusions drawn via nationally aggregated models, as there are also good reasons to prefer a simple model over a complex one \citep{saltelli20,green15}.
-% Still, researchers that intend to use a mechanistic model to describe a dynamic system should design the model to be as realistic as their scientific understanding of the system and their computational abilities permit \eic{THIS SENTENCE SEEMS TO DISMISS THE ARGUMENTS IN THE PREVIOUS SENTENCE.
-PERHAPS THIS SENTENCE CAN JUST BE DELETED}.
 
@@
 
-We then fit the panel version of Model~3 using a slight modification of the PIF algorithm, which we call the Block Panel Iterated Filter (BPIF).
+We then fit the spatially decoupled version of Model~3 using a slight modification of the PIF algorithm, which we call the Block Panel Iterated Filter (BPIF).
 The BPIF algorithm is implemented in the \code{panelPomp} package \citep{breto20panelPomp} in the R programming language, and can be used by adding the argument \code{block = TRUE} in the \code{mif2} function.
 Pseudo-code for this algorithm is provided in Algorithm~S1 in the supplementary material.
-% Even with the simplification of Eq.~\myeqref{eq:model3:approx}, a significant amount of computational effort is needed to obtain a proper model fit.
-The reported parameters were obtained using the PanelPOMP version of Model~3, and simulations for the various vaccination campaigns (Sec.~\ref{sec:results}) were obtained using these same parameters in the SpatPOMP version of the model.
+While the reported parameters were obtained using the decoupled representation of Model~3, simulations for the various vaccination campaigns were obtained using fully coupled version of the model.
 % Simulations from the fitted version of the PanelPOMP model are shown in Fig.~\ref{h3panelsims}.
@@
 
-\begin{figure}[ht]
-<<Model3_PanelSims_Figure, echo=FALSE, fig.height=3>>=
-ggplot() +
-  geom_line(data = h3_panel_quantiles, aes(x = date, y = q50 + 1), col = 'blue') +
-  geom_line(data = dep_plot_df, aes(x = date, y = cases + 1), col = 'black') +
-  geom_ribbon(data = h3_panel_quantiles, aes(x = date, ymin = q05 + 1, ymax = q95 + 1),
-              alpha = 0.4) +
-  facet_wrap(~factor(dep, levels = plot_order), nrow = 2,
-             labeller = dep_labeller) +
-  ylab('Reported cholera cases') +
-  theme(axis.title.x = element_blank()) +
-  scale_x_date(date_labels = "'%y", breaks = seq.Date(from = as.Date("2011-01-01"), as.Date("2019-01-01"), by = '2 years')) +
-  scale_y_log10(labels = scales::trans_format("log10", scales::math_format(10^.x)))
-
-@
-\caption{\label{h3panelsims}
-Simulations from initial conditions using panelPOMP version of Model~3.
-The black curve represents true case count, the blue curve the median of 500 simulations from the model, and the grey ribbons representing $95\%$ confidence interval.
-}
-\end{figure}
+% \begin{figure}[ht]
+% ggplot() +
+%   geom_line(data = h3_panel_quantiles, aes(x = date, y = q50 + 1), col = 'blue') +
+%   geom_line(data = dep_plot_df, aes(x = date, y = cases + 1), col = 'black') +
+%   geom_ribbon(data = h3_panel_quantiles, aes(x = date, ymin = q05 + 1, ymax = q95 + 1),
+%               alpha = 0.4) +
+%   facet_wrap(~factor(dep, levels = plot_order), nrow = 2,
+%              labeller = dep_labeller) +
+%   ylab('Reported cholera cases') +
+%   theme(axis.title.x = element_blank()) +
+%   scale_x_date(date_labels = "'%y", breaks = seq.Date(from = as.Date("2011-01-01"), as.Date("2019-01-01"), by = '2 years')) +
+%   scale_y_log10(labels = scales::trans_format("log10", scales::math_format(10^.x)))
+%
+% \caption{\label{h3panelsims}
+% Simulations from initial conditions using panelPOMP version of Model~3.
+% The black curve represents true case count, the blue curve the median of 500 simulations from the model, and the grey ribbons representing $95\%$ confidence interval.
+% }
+% \end{figure}
 
@@
 
+In some cases, a possible benchmark model could be a generally accepted mechanistic model, but often no such model is available.
+Because of this, we use a log-linear Gaussian ARMA model as an associative benchmark, as recommended by \citet{he10}.
+The theory and practice of ARMA models is well developed, but the exponential growth and decay characteristic of biological dynamics suggests applying these linear models on a log scale.
+Likelihoods of Models~1--3 and their respective ARMA benchmark models are provided in Table~\ref{tab:likes}.
+
 It should be universal practice to present measures of goodness of fit for published models, and mechanistic models should be compared against benchmarks.
@@
 
-In some cases, a possible benchmark model could be a generally accepted mechanistic model, but often no such model is available.
-Because of this, we use a log-linear Gaussian ARMA model as an associative benchmark, as recommended by \citet{he10}.
-The theory and practice of ARMA models is well developed, but the exponential growth and decay characteristic of biological dynamics suggests applying these linear models on a log scale.
-
-The use of a common benchmark may also be beneficial when developing models with varying spatial scale, as direct comparisons between models fit to data with different levels of spatial aggregation are meaningless (e.g.
-comparing Model~1 to Model~3).
-\eic{I DON'T QUITE UNDERSTAND THIS YET.
-BENCHMARKS TEND TO APPLY TO SPECIFIC DATA SETS, SO CANNOT READILY BE USED TO COMPARE DIFFERENT LEVELS OF AGGREGATION.
-LIKELIHOOD BENCHMARKS CAN BE USED TO STUDY THE VALUE OF EXPLICIT SPATIAL COUPLING.
-HOW TO BENCHMARK ACROSS AGGREGATION LEVELS IS, I THINK, SOMETHING OF AN OPEN PROBLEM.}
-Likelihoods of Models~1--3 and their respective ARMA benchmark models are provided in Table~\ref{tab:likes}.
+The use of benchmarks may also be beneficial when developing models with varying spatial scale, where a direct comparison between models likelihoods is meaningless.
+In such a case, a benchmark model could be fit to each spatial resolution being considered, and each model compared to their respective benchmark.
+Large advantages (shortcomings) in model likelihood relative to the benchmark for a given spatial scale that are not present in other spatial scales may provide weak evidence for (against) the spatial resolution under consideration.
+While these comparisons don't provide a formal statistical test of the amount of spatial resolution needed for a given model, they can be quite informative.
 
@@
 
-\begin{tabular}{|c|c|c|c|c|}
+\begin{tabular}{|c|c|c|c|}
 \hline
- & \thead{Model~1} & \thead{Model~2} & \thead{Model~3 \\ (panelPOMP)} & \thead{Model~3 \\ (SpatPOMP)}
+ & \thead{Model~1} & \thead{Model~2} & \thead{Model~3}
 \\
@@
   $\Sexpr{myround(mod2_ll, 1)}$ &
-  \multirow{2}{*}{$\Sexpr{myround(mod3_ll[1], 1)}$\footnotemark[2]} &
   $\Sexpr{myround(mod3_spatPomp_res$ll, 1)}$ \\
@@
   ($\Sexpr{myround(h2_fit$leeFit_ll, 1)}$) &
-  & ($\Sexpr{myround(max(lee3_results[["all_ll_evals"]]), 1)}$)\footnotemark[3]
+  ($\Sexpr{myround(max(lee3_results[["all_ll_evals"]]), 1)}$)\footnotemark[2]
 \\
@@
 Number of &
-  $\Sexpr{as.character(mod1_n_params)}$ & $\Sexpr{as.character(mod2_n_params)}$ & \multirow{2}{*}{$\Sexpr{as.character(mod3_n_params)}$} & $\Sexpr{mod3_spatPomp_n_params}$ \\
- Fit Parameters & (\Sexpr{as.character(mod1_lee_n_params)}) & ($\Sexpr{as.character(h2_fit$leeFit_n_params)}$) && ($\Sexpr{as.character(lee3_results$n_params, 1)}$)
+  $\Sexpr{as.character(mod1_n_params)}$ &
+  $\Sexpr{as.character(mod2_n_params)}$ &
+  $\Sexpr{mod3_spatPomp_n_params}$ \\
+ Fit Parameters &
+ (\Sexpr{as.character(mod1_lee_n_params)}) &
+ ($\Sexpr{as.character(h2_fit$leeFit_n_params)}$) &
+ ($\Sexpr{as.character(lee3_results$n_params, 1)}$)
 \\
@@
 \multirow{2}{*}{AIC} &
-  $\Sexpr{myround(mod1_aic, 1)}$ & $\Sexpr{myround(mod2_aic, 1)}$ & \multirow{2}{*}{$\Sexpr{myround(mod3_aic, 1)}$\footnotemark[2]} & $\Sexpr{myround(mod3_spatPomp_aic, 1)}$ \\
-  & ($\Sexpr{myround(mod1_lee_aic, 1)}$)\footnotemark[1] & ($\Sexpr{myround(mod2_lee_aic, 1)}$) &  & ($\Sexpr{myround(mod3_lee_coupled_aic, 1)}$)\footnotemark[3]
+  $\Sexpr{myround(mod1_aic, 1)}$ &
+  $\Sexpr{myround(mod2_aic, 1)}$ &
+  $\Sexpr{myround(mod3_spatPomp_aic, 1)}$ \\
+  & ($\Sexpr{myround(mod1_lee_aic, 1)}$)\footnotemark[1] &
+  ($\Sexpr{myround(mod2_lee_aic, 1)}$) &
+  ($\Sexpr{myround(mod3_lee_coupled_aic, 1)}$)\footnotemark[2]
 \\
@@
   $\Sexpr{myround(ARMA_benchmarks$m2, 1)}$ &
-  $\Sexpr{myround(ARMA_benchmarks$m3, 1)}$ &
   $\Sexpr{myround(ARMA_benchmarks$m3, 1)}$
@@
 Values in parenthesis are corresponding values using \citet{lee20} parameter estimates.
-\eic{I THINK THAT THE NOTES SHOULD GO IN THE CAPTION RATHER THAN BEING FOOTNOTES.}
-\eic{WE SHOULD REVIEW THIS TABLE IN A MEETING.
-RIGHT NOW, I'M SURPRISED BY A FEW THINGS.
-WHY IS THE LEE ET AL MODEL 2 LIKELIHOOD HIGHER THAN OURS?
-IS THE (29) IN THE PANELPOMP COLUMN A TYPO, SINCE WE DON'T COMPARE TO LEE ET AL IN THIS COLUMN?
-NESTING RELATIONSHIP FOR THE SPATPOMP VERSION OF MODEL 3 MAKES IT HARD TO EXPLAIN THE RATHER LARGE DECREASE IN LOGLIK.
-AT FACE VALUE, THIS SEEMS TO MEAN THAT THE COUPLING IS SUFFICIENTLY POORLY SPECIFIED THAT WE ARE BETTER OFF WITHOUT IT.
-I SEEM TO RECALL THAT OUR UNDERSTANDING INVOLVES THE WAY PANELPOMP USES A COVARIATE THAT DEPENDS ON CASES IN OTHER DEPARTMENTS.
-SINCE LEE ET AL ALSO DO THAT, COULD IT BE THAT THEIR ANALYSIS IS MORE CLOSELY RELATED TO THE PANELPOMP VERSION THAN THE SPATPOMP VERSION?}
+$\textsuperscript{1}$The reported likelihood is an upper bound of the likelihood of the \citet{lee20} model as it is the largest likelihood obtained using their parameter calibration regime.
+$\textsuperscript{2}$\cite{lee20} fit Model~3 to only a subset of the data (March 2014 onward) and did not include data from Ouest in 2015-2016.
+On this subset, the \cite{lee20} fit has a likelihood of $\Sexpr{myround(max(lee3_results$sub_ll_evals), 1)}$.
+On this same subset, our model has a likelihood of $\Sexpr{myround(mod3_spatPomp_res$subset_ll, 1)}$.
 }
@@
 
-\jwc{NEW PARAGRAPH: }
 Similar to comparing log-likelihoods across models, an additional powerful diagnosis tool is the comparison of conditional log-likelihoods.
@@
 Comparing these results across models---including benchmark models---can help researchers identify potential model deficiencies, or errors in the observed data.
-Additional tools for assessing the goodness-of-fit of a model include plotting the effective sample size of each observation \jwc{REF???} and comparing any statistic of the observed data to simulations from the model, which is sometimes referred to as diagnostic probes (for example, the autocorrelation function (ACF), as was done in \cite{king15}).\footnotetext[1]{The reported likelihood is an upper bound of the likelihood of the \citet{lee20} model as it is the largest likelihood obtained using their parameter calibration regime.}
-\footnotetext[2]{Parameters to department-specific models not provided by \citet{lee20}, as department fits were only used as an intermediary step to obtain parameter estimates of the coupled model.}
-\footnotetext[3]{Model 3 was originally fit to only a subset of the data starting from March 2014 and did not include a large portion of data from Ouest in 2015-2016.
-On this subset, the parameters provided by \cite{lee20} achieved a likelihood of $\Sexpr{myround(max(lee3_results$sub_ll_evals), 1)}$.
-On this same subset of data, our model achieved a likelihood of $\Sexpr{myround(mod3_spatPomp_res$subset_ll, 1)}$.}
+Additional tools for assessing the goodness-of-fit of a model include plotting the effective sample size of each observation \jwc{REF???} and comparing any statistic of the observed data to simulations from the model, which is sometimes referred to as diagnostic probes (for example, the autocorrelation function (ACF), as was done in \cite{king15}).
 
@@
 
-\jwc{Replaced Ed's comment with this: }
 We note---the credit of deterministic models---that of the four fitted models in \cite{lee20}, Model~2 provides the most apparently accurate forecasts.
@@
 
-\section{Results}\label{sec:results}
-
-\jwc{SHOULD THESE FIRST TWO PARAGRAPHS BE MOVED TO THE DISCUSSION?}
-\eic{YES.
-ALSO, THERE ARE RESULTS IN SEC.~3.
-PERHAPS, THE RAINFALL ANALYSIS IN SEC 4 COULD BE ANOTHER ASPECT OF THE SEC 3 ANALYSIS: COMPARISON OF MODEL FIT WITH OTHER LINES OF SCIENTIFIC EVIDENCE.
-THEN THE CURRENT SEC 4 COULD BE DISBANDED.
-SOMETIMES THE METHODS-RESULTS-DISCUSSION FORMAT DOES NOT FIT SO NATURALLY WITH APPLIED STATS.
-WE CAN DISCUSS THIS.
-ONE IDEA IS TO HAVE A SECTION NAMED SOMETHING LIKE "LESSONS LEARNED FROM MODELING THE HAITI EPIDEMIC" OR "ADVICE FOR FUTURE STUDIES".
-THE LATTER CAN SEEM PRESUMPTUOUS, AND SHOULD BE DONE WITH DUE CAUTION, BUT FROM ANOTHER PERSPECTIVE IF WE ARE NOT PROVIDING CONCRETE GUIDANCE FOR FUTURE WORK THEN WHAT ARE WE DOING?
-OUR RESPONSIBILITY IS TO MAKE THE STRONGEST CLAIM WE THINK WE CAN PROPERLY SUPPORT; THEN THE REFEREES CAN HAVE A GO AT PULLING THE CLAIM DOWN.}
-\eic{I THINK IT IS INTERESTING THAT THE MODEL-SUPPORTED SUGGESTION BY \citep{andrews11} HAS LARGELY PROVED ACCURATE (VACCINATION, NON-PHARMACEUTICAL INTERVENTIONS, AND INFRASTRUCTURE IMPROVEMENTS ALL HAVE A ROLE AND SHOULD BE FOLLOWED SIMULTANEOUSLY)DESPITE LEGITIMATE CONCERNS ABOUT WHETHER THE SCIENTIFIC SUPPORT FOR THE MODEL IS STRONG ENOUGH TO ROBUSTLY GUIDE POLICY \citep{grad12}.
-THE MAIN CLAIM OF ANDREWS11 IS THAT FORECASTS AND POLICY NOT SUPPORTED BY A MODEL SEEM EVEN MORE UN-ROOTED THAN RECOMMENDATIONS BASED ON A WEAKLY SUPPORTED MODEL.
-THE CONCLUSION FROM ALL THIS, WHICH I THINK IS SUPPORTED BY OUR RE-ANALYSIS OF LEE ET AL, IS THAT MECHANISTIC MODELS FITTED TO DATA ARE A STRONG WAY TO ENFORCE CONSISTENT THINKING, BUT WEAKER AS A WAY TO CHECK THAT YOU ARE ACTUALLY RIGHT DUE TO THE LIMITATIONS OF INFERRING CAUSAL MECHANISMS FROM OBSERVATIONAL DATA.
-POLICY-MAKERS SHOULD FOLLOW AN APPROACH WHICH IS SELF-CONSISTENT AND CONSISTENT WITH PREVIOUS EVIDENCE THAT MIGHT WORK (WHICH MODELS CAN HELP WITH) BUT THEY MUST BE AGILE AS NEW EVIDENCE EMERGES.
-MODELS, TOGETHER WITH DATA ANALYSIS, CAN ALSO HELP REVEAL WHEN THE EVIDENCE SUPPORTING A P0LICY IS BREAKING DOWN, FORCING A RE-THINK.
-}
-
-A model which aspires to provide quantitative guidance for assessing interventions should provide a quantitative statistical fit for available data.
-However, strong statistical fit does not guarantee a correct causal structure: it does not even necessarily require the model to assert a causal explanation.
-A causal interpretation is strengthened by corroborative evidence.
-For example, reconstructed latent variables (such as numbers of susceptible and recovered individuals) should make sense in the context of alternative measurements of these variables;
-parameter values which fit the data should make sense in the context of alternative lines of evidence about the phenomena being modeled.
-% \jwc{Perhaps add comment on our final parameter estimates, and remove the following sentences in this paragraph.}
-% A model is necessarily a simplified approximation of a complex process, and so it is unreasonable to expect models (describing and fitted to population-level processes) to perfectly match quantitative understanding of individual-level processes.
-% Such discrepancies could lead to biases when interventions (modeled as effects on individuals) are included in the population-level, unless there are historical occurrences of the intervention which enable the consequence to be calibrated at the population level.
-
-If a mechanistic model including a feature (such as a representation of a mechanism, or the inclusion of a covariate) fits better than mechanistic models without that feature, and also has competitive fit compared to associative models, this may be taken as evidence supporting the scientific relevance of the feature.
-As for any analysis of observational data, we must be alert to the possibility of confounding.
-For a covariate, this shows up in a similar way to regression analysis: the covariate under investigation could be a proxy for some other unmodeled or unmeasured covariate.
-For a mechanism, the model feature could in principle explain the data by helping to account for some different unmodeled phenomenon.
-% Assessing potential confounding is part of building an argument for a causal interpretation of a fitted model.
-In the context of our analysis, the estimated trend in transmission rate could be explained by any trending variable (such as hygiene improvements, or changes in population behavior), resulting in confounding from colinear covariates.
-Alternatively, the trend could be attributed to a decreasing reporting rate rather than decreasing transmission rate, resulting in confounded mechanisms.
-The robust statistical conclusion is that a model which allows for change fits better than one which does not---we argue that a decreasing transmission rate is a plausible way to explain this, but the incidence data themselves do not provide enough information to pin down the mechanism.
+\subsection{Comparison of Model Fit to Oustide Scientific Evidence}\label{sec:science}
 
@@
 It is important instead to recognize and assess the modeling simplifications and assumptions that were used in order to arrive at the conclusions.
+
+\subsection{Predicting the Effect of Interventions}\label{sec:interventions}
 
@@
 % x-axis scale is intended to ease comparison with Figure~4 of \cite{lee20}.
-% \jwc{MAYBE JUST SUMMARIZE THIS FIGURE IN A PARAGRAPH OR MOVE TO SUPPLEMENT TO MAKE SPACE?}
 % }
@@
 Translations back into the notation of \citet{lee20} are given in Table~S1.
-% \eic{TODO: CHECK PARAMETERIZATION AND UNITS FOR $\obsOverdispersion$ AND $\sigmaProc$ TO MAKE THEM COMPARABLE}
 }
 \end{table}
+
+\section{Advising policy via dynamic modeling}\label{sec:advice}
+
+\eic{I THINK IT IS INTERESTING THAT THE MODEL-SUPPORTED SUGGESTION BY \citep{andrews11} HAS LARGELY PROVED ACCURATE (VACCINATION, NON-PHARMACEUTICAL INTERVENTIONS, AND INFRASTRUCTURE IMPROVEMENTS ALL HAVE A ROLE AND SHOULD BE FOLLOWED SIMULTANEOUSLY)DESPITE LEGITIMATE CONCERNS ABOUT WHETHER THE SCIENTIFIC SUPPORT FOR THE MODEL IS STRONG ENOUGH TO ROBUSTLY GUIDE POLICY \citep{grad12}.
+THE MAIN CLAIM OF ANDREWS11 IS THAT FORECASTS AND POLICY NOT SUPPORTED BY A MODEL SEEM EVEN MORE UN-ROOTED THAN RECOMMENDATIONS BASED ON A WEAKLY SUPPORTED MODEL.
+THE CONCLUSION FROM ALL THIS, WHICH I THINK IS SUPPORTED BY OUR RE-ANALYSIS OF LEE ET AL, IS THAT MECHANISTIC MODELS FITTED TO DATA ARE A STRONG WAY TO ENFORCE CONSISTENT THINKING, BUT WEAKER AS A WAY TO CHECK THAT YOU ARE ACTUALLY RIGHT DUE TO THE LIMITATIONS OF INFERRING CAUSAL MECHANISMS FROM OBSERVATIONAL DATA.
+POLICY-MAKERS SHOULD FOLLOW AN APPROACH WHICH IS SELF-CONSISTENT AND CONSISTENT WITH PREVIOUS EVIDENCE THAT MIGHT WORK (WHICH MODELS CAN HELP WITH) BUT THEY MUST BE AGILE AS NEW EVIDENCE EMERGES.
+MODELS, TOGETHER WITH DATA ANALYSIS, CAN ALSO HELP REVEAL WHEN THE EVIDENCE SUPPORTING A P0LICY IS BREAKING DOWN, FORCING A RE-THINK.
+}
+
+A model which aspires to provide quantitative guidance for assessing interventions should provide a quantitative statistical fit for available data.
+However, strong statistical fit does not guarantee a correct causal structure: it does not even necessarily require the model to assert a causal explanation.
+A causal interpretation is strengthened by corroborative evidence.
+For example, reconstructed latent variables (such as numbers of susceptible and recovered individuals) should make sense in the context of alternative measurements of these variables;
+parameter values which fit the data should make sense in the context of alternative lines of evidence about the phenomena being modeled.
+% \jwc{Perhaps add comment on our final parameter estimates, and remove the following sentences in this paragraph.}
+% A model is necessarily a simplified approximation of a complex process, and so it is unreasonable to expect models (describing and fitted to population-level processes) to perfectly match quantitative understanding of individual-level processes.
+% Such discrepancies could lead to biases when interventions (modeled as effects on individuals) are included in the population-level, unless there are historical occurrences of the intervention which enable the consequence to be calibrated at the population level.
+
+If a mechanistic model including a feature (such as a representation of a mechanism, or the inclusion of a covariate) fits better than mechanistic models without that feature, and also has competitive fit compared to associative models, this may be taken as evidence supporting the scientific relevance of the feature.
+As for any analysis of observational data, we must be alert to the possibility of confounding.
+For a covariate, this shows up in a similar way to regression analysis: the covariate under investigation could be a proxy for some other unmodeled or unmeasured covariate.
+For a mechanism, the model feature could in principle explain the data by helping to account for some different unmodeled phenomenon.
+% Assessing potential confounding is part of building an argument for a causal interpretation of a fitted model.
+In the context of our analysis, the estimated trend in transmission rate could be explained by any trending variable (such as hygiene improvements, or changes in population behavior), resulting in confounding from colinear covariates.
+Alternatively, the trend could be attributed to a decreasing reporting rate rather than decreasing transmission rate, resulting in confounded mechanisms.
+The robust statistical conclusion is that a model which allows for change fits better than one which does not---we argue that a decreasing transmission rate is a plausible way to explain this, but the incidence data themselves do not provide enough information to pin down the mechanism.
 
@@
 We provided a few suggestions that may aid the fitting of mechanistic models such as comparing model likelihoods to a benchmark.
-Improved model fits allows for meaningful statistical inference that may provide valuable insight on a dynamic system in question and may improve the accuracy of model-based projections.
+Improved model fits allows for meaningful statistical inference that may provide valuable insight on a dynamic system and may improve the accuracy of model-based projections.
 % TODO: "Improved model fits", or "Improved model fitting", based on Anna's suggestion.
@@
 We acknowledge there are limitations to this study;
-one example was the inability to fit model parameters to the fully-coupled SpatPOMP version of Model~3.
-Promising theoretical and methodological developments \citep{ning21ibpf,ionides22} based on the Block Particle Filter \citep{rebeschini15} may potentially be used to fit the SpatPOMP version of Model~3 in future work.
-
-% \arc{Right now, the manuscript's main point/claim is a bit muddled (at least to me).
-I am not sure how to directly address this, but having a additional paragraph or two in the discussion may help the impact of this article/"drive our points home".
-I will think more on this in the next week, but maybe we could further emphasize the balance/dependence between policy, prediction, inference, model quality, and model fitting?
-Well informed policy depends upon reliable prediction which depends upon model quality/performance which depends upon rigorous evaluation and estimation, regardless of correct model specification (though that is also important/of great interest).}
-% \jwc{Thanks, I agree that the main point isn't clear, and Ed has suggested as much also.
-I think that adding additional paragraphs would help this, but we are already 2 pages over the recommended submission length, so re-wording the text may be a better option than adding more to it...
-I'll also have to think about this more this week.}
-% \eic{HERE'S A SUGGESTION FOR A CONCLUDING PARAGRAPH. I THINK IT IS SUPPORTED BY THE RESULTS...}
+one example was the inability to fit model parameters to the fully-coupled version of Model~3.
+Promising theoretical and methodological developments \citep{ning21ibpf,ionides22} based on the Block Particle Filter \citep{rebeschini15} may potentially be used to obtain better parameter estimates for Model~3 in future work.
+
 Inference for mechanistic time series models offers opportunities for understanding and controlling complex dynamic systems.
@@
 The dynamic document, together with the R packages, form a {\it compendium}, defined by \citet{gentleman07} as a distributable and executable unit which combines data, text and auxiliary software (the latter meaning code written to run in a general-purpose, portable programming environment, which in this case is R).
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% Single Appendix:                         %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% \begin{appendix}
-% \section*{Block Panel Iterated Filter}
-%
-% This algorithm is a modification of the panel iterated filter \citep{breto20}.
-% The intuition behind this modification is that if the unit-specific
-% parameters are independent of one another, we allow these parameters to be
-% fit as such by not allowing the particle ancestry in other units to influence
-% the fit of unit specific parameter in question.
-% Pseudo-code for this algorithm is given in Algorithm~1.
-%
-% While theoretical properties of this algorithm have yet to be derived, empirical
-% results using Model~3 and on simulated data from a simple model suggest that this
-% modification may lead to quicker convergence to the model likelihood.
-%
-% \begin{algorithm}[ht]
-%    \caption{\textbf{BPIF}. \\
-%     {\bf Inputs}:
-%     Simulator of initial density, $f_{X_{u, 0}}(x_{u, 0}; \theta)$ for $u$ in $1:U$.
-%     Simulator of transition density, $f_{X_{u, n}|X_{u, n-1}}(x_{u, n}|x_{u, n-1}; \theta)$ for $u$ in $1:U$, $n$ in $1:N_u$.
-%     Evaluator of measurement density, $f_{Y_{u, n}|X_{u, n}}(y_{u, n}|x_{u, n} ; \theta)$ for $u$ in $1:U$, $n$ in $1:N_u$.
-%     Data $y_{u, n}^*$,for $u$ in $1:U$, $n$ in $1:N_u$.
-%     Number of iterations, $M$.
-%     Number of particles, $J$.
-%     Starting parameter swarm, $\Theta_j^0$ for $j$ in $1:J$.
-%     Simulator of perturbation density, $h_{u, n}(\theta | \phi; \sigma)$ for $u$ in $1:U$, $n$ in $0:N_u$
-%     Perturbation sequence, $\sigma_m$ for $m$ in $1:M$.
-%     {\bf Output:} Final parameter swarm, $\Theta_{j}^{m}$ for $j$ in $1:J$
-%     \label{alg:bpif}}
-% %\BlankLine
-% \For{$m \in 1:M$}{
-%   Set $\Theta_{0, j}^m = \Theta_j^{m-1}$ for $j \in 1:J$\;
-%     \For{$u \in 1:U$}{
-%       Perturb parameters: $\Theta_{u, 0, j}^{F, m} \sim h_{u, 0}(\theta| \Theta_{u - 1, j}^m; \sigma_m)$ for $j \in 1:J$\;
-%       Initialize: $X_{u, 0, j}^{F, m} \sim f_{X_{u, 0}}(x_{u, 0}; \Theta_{u, 0, j}^{F, m})$ for $j \in 1:J$\;
-%         \For{$n \in 1:N_u$} {
-%           Perturb parameters: $\Theta_{u, n, j}^{P, m} \sim h_{u, n}(\theta|\Theta_{u, n-1, j}^{F, m}, \sigma_m)$ for $j \in 1:J$\;
-%           Prediction Simulation: $X_{u, n, j}^{P, m} \sim f_{X_{u, n}|X_{u, n-1}}(x_{u, n}|X_{u, n-1, j}^{F, m}; \Theta_{u, n, j}^{P, m})$ for $j \in 1:J$\;
-%           Prediction weights: $w_{u, n, j}^m = f_{Y_{u, n}|X_{u, n}}(y_{u, n}^*|X_{u, n, j}^{P, m})$ for $j \in 1:J$\;
-%           Draw $k_{1:j}$ with $P(k_j = i) = w_{u, n, i}^m / \sum_{v = 1}^J w_{u, n, v}^m$ for $i, j \in 1:J$\;
-%           Set: $\Theta_{u, n, j}^{F, m} = \Theta_{u, n, k_j}^{P, m}$ and $X_{u, n, j}^{F, m} = X_{u, n, k_j}^{P, m}$ for $j \in 1:J$\;
-%         }
-%       Set $\Theta_{u, j}^m = \Theta_{u, N_u, j}^{F, m}$ for $j \in 1:J$\;
-%     }
-%   Set $\Theta_j^m = \Theta_{U, j}^m$ for $j \in 1:J$\;
-% }
-% \end{algorithm}
-
-% \section*{}%% if no title is needed, leave empty \section*{}.
-% \begin{table}
-%   \begin{tabular}{|c|c|c|c|}\hline
-%     \thead{Parameter} & \thead{Our \\ Notation} & \thead{\cite{lee20} \\ Notation} & \thead{Relevant \\ Model(s)} \\
-%     \hline
-%     \hline
-%     % Table begins here
-%     \multirow{2}{*}{Reporting Rate} & \multirow{2}{*}{$\reportRate$} & $\rho$ & 1, 2 \\\cline{3-4}
-%       & & $\epsilon_1, \epsilon_2$ & 3\\\hline
-%     Mixing Coefficient & $\mixExponent$ & $\nu$ & 1 \\\hline
-%     \multirow{2}{*}{Measurement Over-Dispersion} & \multirow{2}{*}{$\obsOverdispersion$} & $\tau$ & 1\\\cline{3-4}
-%     & & $p$ & 3 \\\hline
-%     Birth Rate & $\muBirth$ & $\mu$ & 1 \\\hline
-%     \multirow{2}{*}{Natural Mortality Rate} & \multirow{2}{*}{$\muDeath$} & $\delta$ & 1 \\\cline{3-4}
-%      & & $\mu$ & 3 \\\hline
-%      Cholera Mortality Rate & $\choleraDeath$ & $\alpha$ & 3 \\\hline
-%     \multirow{2}{*}{Latent Period} & \multirow{2}{*}{$1/\muEI$} & $1/\sigma$ & 1 \\\cline{3-4}
-%                                    & & $1/\gamma_E$ & 2 \\\hline
-%     Recovery Rate & $\muIR$ & $\gamma$ & 1, 2, 3 \\\hline
-%     \multirow{3}{*}{Loss of Immunity} & \multirow{3}{*}{$\muRS$} & $\alpha$ & 1 \\\cline{3-4}
-%                                       & & $\sigma$ & 2 \\\cline{3-4}
-%                                       & & $\rho$ & 3 \\\hline
-%     \multirow{3}{*}{Symptomatic Ratio} & \multirow{3}{*}{$\symptomFrac$} & $1 - \theta_0$ & 1 \\\cline{3-4}
-%     & & $k$ & 2 \\\cline{3-4}
-%     & & $\sigma$ & 3\\\hline
-%     Asymptomatic & \multirow{2}{*}{$\asymptomRelativeInfect$} & $\kappa$ & 1 \\\cline{3-4}
-%     Relative Infectiousness & & $red_\beta$ & 2 \\\hline
-%     \multirow{2}{*}{Human-to-Water Shedding} & \multirow{2}{*}{$\Wshed$} & $\mu$ & 2 \\\cline{3-4}
-%      & & $\theta_I$ & 3 \\\hline
-%     Asymptomatic & \multirow{2}{*}{$\asymptomRelativeShed$} & $red_\mu$ & 2 \\\cline{3-4}
-%     Relative Shedding & & $\theta_A/\theta_I$ & 3 \\\hline
-%     \multirow{2}{*}{Seasonal Amplitude} & \multirow{2}{*}{$\seasAmplitude$} & $\alpha_s$ & 2 \\\cline{3-4}
-%     & & $\lambda$ & 3 \\\hline
-%     \multirow{2}{*}{Transmission} & \multirow{2}{*}{$\transmission$} & $\beta$ & 1, 2 \\\cline{3-4}
-%     & & $c$ & 3 \\\hline
-%     Water-to-Human & \multirow{2}{*}{$\beta_W$} & $\beta_W$ & 2 \\\cline{3-4}
-%     Infection Rate & & $\beta$ & 3 \\\hline
-%     Bacteria Mortality & \multirow{2}{*}{$\Wremoval$} & $\delta$ & 2 \\\cline{3-4}
-%     Rate & & $\mu_\beta$ & 3 \\\hline
-%     \multirow{3}{*}{Vaccination Efficacy} & \multirow{3}{*}{$\vaccineEfficacy$} & $\theta_{vk}$ & 1 \\\cline{3-4}
-%     & & $\text{VE}_1, \text{VE}_2$ & 2 \\\cline{3-4}
-%     & & $\eta_{1d}, \eta_{2d}$ & 3 \\\hline
-%     Process Over-dispersion & $\sigmaProc$ & $\sigma_w$ & 3 \\\hline
-%     % Table ends here
-%   \end{tabular}
-%   \caption{
-%   \label{tab:translate}Translations between our common notation and notation used in \cite{lee20}\eic{TODO: half saturation constant, 2 and 3. Fraction of susceptible, 2.
-%   Initial Value parameters.}
-%   }
-% \end{table}
-
-% \section*{Initial Values}
-%
-% To perform inference on POMP models, it is necessary to propose an initial density for the latent process $f_{X_0}(x_0;\theta)$.
-% This density is used to obtain initial values of the latent state when fitting and evaluating the model.
-% For each of the models considered in this analysis, the initial conditions are derived by enforcing the model dynamics on reported cholera cases.
-% For example, in a deterministic model, the total number of Infected and Asymptomatic individuals may be calculated using the observed number of cholera cases at time time of model initialization based on model parameters such as reporting rate $\reportRate$ and fraction of symptomatic cases $\symptomFrac$.
-% It is also sometimes necessary to fit some initial value parameters in order to help determine starting values for weakly identifiable compartments.
-% In the following subsections, we mention initial value parameters that were fit for each model.
-%
-% \subsection*{Model~1}
-%
-% For this model, the number of individuals in the Recovered and Asymptomatic compartments are set to zero, but the initial proportion of Infected and Exposed individuals is estimated as initial value parameters ($I_0$ and $E_0$, respectively) using the MIF2 algorithm.
-% Finally, the initial proportion of Susceptible individuals $S_0$ is calculated as $S_0 = 1 - I_0 - E_0$.
-%
-% \subsection*{Model~3}
-%
-% We use the reported cases at the start of the pandemic to approximate the number of Asymptomatic, Infectious, and Recovered individuals in each department $u \in 1:U$ using the same approximation as provided in Eq.~\myeqref{eq:model3:approx}.
-% The susceptible compartment is initialized so that the sum $S_{u}(0) + I_u(0) + A_u(0) + \sum_k R_{u, k}(0) = \text{population}_u$.
-% The bacteria compartment is then initialized using Eq.~\myeqref{eq:model3:b0}:
-%
-% \begin{equation}
-%   B_u(0) = \big[1 + \seasAmplitude \big(\xi_u)^r \big] D_i \, \Wshed \big[ I_{u}(0)+ \asymptomRelativeShed A_{u}(0) \big]\label{eq:model3:b0}
-% \end{equation}
-%
-% Where $\xi_u \in (0, 1)$ are initial value parameters that we introduce in order to allow some flexibility in determining the initial state of the bacteria compartment.
-
-% \section*{Measurement Models}
-%
-% Each POMP requires specification of a measurement model, which is a statistical description of how observations on the system are obtained.
-% In general, we used the same measurement models that were reported in \cite{lee20}.
-%
-% \subsection*{Model~1}
-%
-% The only reported cases under this measurement model come from individuals who develop symptoms, as noted in Eq.~\myeqref{model1:meas}.
-% \begin{equation}
-%   \label{model1:meas}
-%   y_{t} \mid \Delta N_{E_{k}I_{k}} = z_{t} \sim \text{NB}\left(\reportRate z_{t}, \obsOverdispersion \right)
-% \end{equation}
-%
-% \subsection*{Model~2}
-%
-% Model~2 was fit via trajectory matching and therefore does not have an explicit measurement model.
-% Implicitly, however, this fitting process is equivalent to having a Gaussian measurement process.
-% Therefore the measurement model can be expressed as Eq.~\myeqref{model2:meas}.
-% \begin{equation}
-%   \label{model2:meas}
-%   y_{u, t} \mid \Delta N_{I_{ud}R_{ud}} = z_{u, t} \sim \text{N}\left(\reportRate z_{u, t}, \obsOverdispersion^2 \right)
-% \end{equation}
-%
-% Where $\Delta N_{I_{ud}R_{ud}}$ is the number of individuals who move from compartment $I_{ud}$ to $R_{ud}$, for unit $u \in 1:U$ and vaccination cohort $d \in 0:5$.
-% We also note here that the variance parameter $\obsOverdispersion^2$ is separately fit for the epidemic and endemic phases.
-%
-% \subsection*{Model~3}
-%
-% In this model, reported cholera cases are assumed to stem from individuals who develop symptoms and seek healthcare.
-% Therefore reported cases are assumed to come from an over-dispersed negative binomial model, given the increase in infected individuals:
-% \begin{equation}
-%   \label{model3:meas}
-%   y_{u, t} \mid \Delta N_{S_{ud}I_{ud}} = z_{u, t} \sim \text{NB}\left(\reportRate z_{u, t}, \obsOverdispersion \right)
-% \end{equation}
-%
-% This measurement model is a minor change from that used in \cite{lee20}, which allowed for a change in the reporting rate on January 1st, 2018.
-% The fitted values of the reporting rate---before and after January 2018---were $0.97$ and $0.097$, respectively.
-% This major change in reporting rate alone could have been the cause that Model~3 originally failed to predict the eradication of cholera, as an overnight change from near perfect to almost non-existent reporting forces the model to explain the observed decrease in cases as a decrease in the reporting of cases rather than of prevalence of cases.
-% This shift was justified by a "change of the case definition that occurred on January 1st, 2018";
-% this claim was not cited, and we could find no evidence that such a drastic change in the reporting rate would be warranted.
-
-% \end{appendix}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% Multiple Appendixes:                     %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 

### 2022-09-15 ms.Rnw
HAITIPKG UPDATE, moved estimation of Model 1 parameters to ms

@@
 library(spatPomp)
+library(foreach)
 library(doParallel)
@@
 # 40 cores for doob, 8 for ito
-doob_cores <- 40
-gl_cores <- 36
+# doob_cores <- 40
+# gl_cores <- 36
 
@@
 Despite this autonomy, and largely adhering to existing guidelines on creating models to inform policy \citep{behrend20,saltelli20}, the consensus across the four models was that an extensive nationwide vaccination campaign would be necessary to eliminate cholera from Haiti.
-This conclusion is inconsistent with the fact that there have been no confirmed cases since February, 2019 without additional vaccination efforts \citep{ferguson22}.
+This conclusion is inconsistent with the fact that there have been no confirmed cases since February, 2019 \citep{trevisin22}, despite the lack of a concentrated vaccination effort.
 
@@
 
+<<FitModel1, echo=FALSE, message=FALSE>>=
+set.seed(636813)
+
+NP_H1        <- switch(RUN_LEVEL, 100, 1000,   2000)
+NMIF_H1      <- switch(RUN_LEVEL,  10,   50,    200)
+NUM_TREND    <- switch(RUN_LEVEL,   5,   10,     90)
+NPROF        <- switch(RUN_LEVEL,   4,    4,     24)
+NP_EVAL      <- switch(RUN_LEVEL, 200, 1000,   5000)
+NREPS_EVAL   <- switch(RUN_LEVEL, cores, 20,  cores)
+
+if (RUN_LEVEL != 3 & file.exists("model1/run_level_3/haiti1_fit.rds")) {
+  h1_MIF <- readRDS("model1/run_level_3/haiti1_fit.rds")
+} else {
+  h1_MIF <- bake(
+    file = paste0("model1/", rl_dir, "haiti1_fit.rds"), {
+      fit_haiti1(
+        NP = NP_H1, 
+        NMIF = NMIF_H1, 
+        NUM_TREND = NUM_TREND, 
+        NPROF = NPROF,
+        NREPS_EVAL = NREPS_EVAL, 
+        NP_EVAL = NP_EVAL, 
+        ncores = cores
+      )
+    }
+  )
+}
+
+h1_prof_res <- h1_MIF %>%
+  group_by(betat) %>%
+  summarize(logLik = max(ll))
+
+mcap_results <- pomp::mcap(h1_prof_res$logLik, h1_prof_res$betat, span = 1)
+
+# Get the best set of parameters from MIF2 search. 
+p1 <- h1_MIF %>%
+  filter(ll == max(ll)) %>%
+  select(-ll, -ll.se) %>%
+  unlist()
+
+mod1_ll <- h1_MIF %>%
+  filter(ll == max(ll)) %>%
+  select(ll) %>%
+  unlist()
+
+rm(NP_H1, NMIF_H1, NUM_TREND, NPROF, NREPS_EVAL, NP_EVAL, h1_MIF)
+gc()
+
+@
+
 <<Load Model Parameters, echo=FALSE, results='hide'>>=
@@
 
-# load('model1/output/trendResultsJoint.rda')  # Parameters and evaluations
-load('model1/output/recal_mcap_results.rda')
-
-p1 <- best_params
-rm(best_params)
-
-gc()
-
-# Fit model 2. Note that this takes the same ammount of time at all run
+# Fit model 2. Note that this takes the same amount of time at all run
 # levels.
@@
 ggplot() +
-  geom_point(data = maxresults, aes(x = betat, y = logLik)) +
+  geom_point(data = h1_prof_res, aes(x = betat, y = logLik)) +
   geom_line(data = mcap_results$fit, aes(x = parameter, y = smoothed), col = 'blue') +
@@
 
+if (RUN_LEVEL != 3 & file.exists("model3/run_level_3/panelPompSims.rds")) {
+  h3_panel_sims <- readRDS("model3/run_level_3/panelPompSims.rds")
+} else {
   h3_panel_sims <- bake(
@@
 
+if (RUN_LEVEL != 3 & file.exists("model3/run_level_3/SpatPompSims.rds")) {
+  h3Spat_quants <- readRDS("model3/run_level_3/SpatPompSims.rds")
+} else {
   h3Spat_quants <- bake(
@@
 
-# Load pre-calculated model 1 likelihood
-mod1_ll <- h1_ll['ll']  # Comes from model1/output/recal_mcap_results.rda
-
 # model 1 AIC
@@
 # Calculate likelihood via block-particle filter
+if (RUN_LEVEL != 3 & file.exists("model3/run_level_3/SpatPomp_res.rds")) {
+  mod3_spatPomp_res <- readRDS("model3/run_level_3/SpatPomp_res.rds")
+} else{
   mod3_spatPomp_res <- bake(
@@
 
+if (RUN_LEVEL != 3 & file.exists("model3/run_level_3/lee3_results.rds")) {
+  lee3_results <- readRDS("model3/run_level_3/lee3_results.rds")
+} else {
   lee3_results <- bake(
@@
 # projects the vaccination scenario, under the specified model.
+if (RUN_LEVEL != 3 & file.exists("model1/run_level_3/S0_sims.rds")) {
+  h1_S0_sims <- readRDS("model1/run_level_3/S0_sims.rds")
+} else {
   h1_S0_sims <- bake(
@@
 # projects the vaccination scenario, under the specified model.
+if (RUN_LEVEL != 3 & file.exists("model1/run_level_3/S2_sims.rds")) {
+  h1_S2_sims <- readRDS("model1/run_level_3/S2_sims.rds")
+} else {
   h1_S2_sims <- bake(
@@
 # projects the vaccination scenario, under the specified model.
+if (RUN_LEVEL != 3 & file.exists("model1/run_level_3/S4_sims.rds")) {
+  h1_S4_sims <- readRDS("model1/run_level_3/S4_sims.rds")
+} else {
   h1_S4_sims <- bake(
@@
 # projects the vaccination scenario, under the specified model.
+if (RUN_LEVEL != 3 & file.exists("model1/run_level_3/S3_sims.rds")) {
+  h1_S3_sims <- readRDS("model1/run_level_3/S3_sims.rds")
+} else {
   h1_S3_sims <- bake(
@@
 # projects the vaccination scenario, under the specified model.
+if (RUN_LEVEL != 3 & file.exists("model1/run_level_3/S1_sims.rds")) {
+  h1_S1_sims <- readRDS("model1/run_level_3/S1_sims.rds")
+} else {
   h1_S1_sims <- bake(
@@
 
+if (RUN_LEVEL != 3 & file.exists("model3/run_level_3/h3_bpf.rds")) {
+  h3_bpf <- readRDS("model3/run_level_3/h3_bpf.rds")
+} else {
   h3_bpf <- bake(
@@
   )
-
+}
+
+if (RUN_LEVEL != 3 & file.exists("model3/run_level_3/mod3_V0_res.rds")) {
+  mod3_V0_res <- readRDS("model3/run_level_3/mod3_V0_res.rds")
+} else {
   mod3_V0_res <- bake(
-  file = paste0("model3/", rl_dir, "mod3_V0_res"), {
+    file = paste0("model3/", rl_dir, "mod3_V0_res.rds"), {
 
@@
   )
-
+}
+
+if (RUN_LEVEL != 3 & file.exists("model3/run_level_3/mod3_V1_res.rds")) {
+  mod3_V1_res <- readRDS("model3/run_level_3/mod3_V1_res.rds")
+} else {
   mod3_V1_res <- bake(
-  file = paste0("model3/", rl_dir, "mod3_V1_res"), {
+    file = paste0("model3/", rl_dir, "mod3_V1_res.rds"), {
 
@@
   )
-
+}
+
+if (RUN_LEVEL != 3 & file.exists("model3/run_level_3/mod3_V2_res.rds")) {
+  mod3_V2_res <- readRDS("model3/run_level_3/mod3_V2_res.rds")
+} else {
   mod3_V2_res <- bake(
-  file = paste0("model3/", rl_dir, "mod3_V2_res"), {
+    file = paste0("model3/", rl_dir, "mod3_V2_res.rds"), {
 
@@
   )
-
+}
+
+if (RUN_LEVEL != 3 & file.exists("model3/run_level_3/mod3_V3_res.rds")) {
+  mod3_V3_res <- readRDS("model3/run_level_3/mod3_V3_res.rds")
+} else {
   mod3_V3_res <- bake(
-  file = paste0("model3/", rl_dir, "mod3_V3_res"), {
+    file = paste0("model3/", rl_dir, "mod3_V3_res.rds"), {
 
@@
   )
-
+}
+
+if (RUN_LEVEL != 3 & file.exists("model3/run_level_3/mod3_V4_res.rds")) {
+  mod3_V4_res <- readRDS("model3/run_level_3/mod3_V4_res.rds")
+} else {
   mod3_V4_res <- bake(
-  file = paste0("model3/", rl_dir, "mod3_V4_res"), {
+    file = paste0("model3/", rl_dir, "mod3_V4_res.rds"), {
 
@@
   )
+}
 @

### 2022-09-16 ms.Rnw
Minor edits

@@
 \DeclareMathSymbol{\varv}{\mathord}{matha}{118}
+\DeclareMathOperator*{\argmax}{arg\,max}
 \newcommand\myeqref[1]{(\ref{#1})}
@@
   )
-
-# TODO: Remove these functions and replace with lubridate functions (also in the package).
-dateToYears <- function(date, origin = as.Date("2014-01-01"), yr_offset = 2014) {
-  # This function converts a date to a decimal representation
-  #
-  # ex: "1976-03-01" -> 1976.163
-
-  julian(date, origin = origin) / 365.25 + yr_offset
-}
-
-yearsToDate <- function(year_frac, origin = as.Date("2014-01-01"), yr_offset = 2014.0) {
-  # This function is the inverse function of dateToYears; it takes
-  # a decimal representation of a date and converts it into a Date.
-  #
-  # ex: 1976.163 -> "1976-03-01"
-
-  as.Date((year_frac - yr_offset) * 365.25, origin = origin)
-}
-
-yearsToDateTime <- function(year_frac, origin = as.Date("2014-01-01"), yr_offset = 2014.0) {
-  # Same as the function above, but a DateTime object rather than a Date
-  # object.
-  #
-  # ex: 1976.163 -> "1976-03-01"
-  as.POSIXct((year_frac - yr_offset) * 365.25 * 3600 * 24, origin = origin)
-}
-
 @
@@
 Demographic stochasticity is accounted for by modeling non-cholera related death rate $\muDeath$ in each compartment, along with an additional death rate $\choleraDeath$ in \myeqref{eq:model3:IS} to account for cholera induced deaths among infected individuals.
-We note that all deaths are balanced by births into the susceptible compartment in \myeqref{eq:model3:AS} and \myeqref{eq:model3:RS}, thereby maintaining constant population in each department.
+All deaths are balanced by births into the susceptible compartment in \myeqref{eq:model3:AS} and \myeqref{eq:model3:RS}, thereby maintaining constant population in each department.
 
@@
 
-We consider model fitting (Sec.~\ref{model_fitting}) followed by diagnostic investigations (Sec.~\ref{model_diagnostics}) and forecasting (Sec.~\ref{sec:filter}).
-
-<<FitModel1, echo=FALSE, message=FALSE>>=
+We consider model fitting (Sec.~\ref{sec:model_fitting}) followed by diagnostic investigations (Sec.~\ref{sec:model_diagnostics}) and forecasting (Sec.~\ref{sec:filter}).
+
+<<FitModel1, echo=FALSE, message=FALSE, include=FALSE>>=
 set.seed(636813)
@@
 
-\subsection{Model Fitting}\label{model_fitting}
+\subsection{Model Fitting}\label{sec:model_fitting}
 
@@
   mutate(
-    date = yearsToDate(year)
+    date = as.Date(lubridate::round_date(lubridate::date_decimal(year), unit = 'day'))
   )
@@
 Evidence for the former approach has been provided in previous studies \citep{king15}, including the specific case of heterogeneity between Haitian departments in cholera transmission \citep{collins14}.
-Note that this evidence alone does not automatically discredit conclusions drawn via nationally aggregated models, as there are also good reasons to prefer a simple model over a complex one \citep{saltelli20,green15}.
+Note that this evidence alone does not automatically discredit conclusions drawn via nationally aggregated models, as there are also reasons to prefer a simple model over a complex one \citep{saltelli20,green15}.
 
@@
 In the case of a PanelPOMP, an extension of the IF2 algorithm, known as Panel Iterated Filtering (PIF) \citep{breto20}, can be used to obtain the MLE, which solves the curse of dimensionality for this class of models.
-A major advantage of this algorithm over fitting each unit-specific model separately is that PIF can be used to fit both unit-specific parameters and shared parameters; in this way, the calibration of shared parameters involves all of the available data instead of just an arbitrarily chosen subset.
-
-We then fit the spatially decoupled version of Model~3 using a slight modification of the PIF algorithm, which we call the Block Panel Iterated Filter (BPIF).
-The BPIF algorithm is implemented in the \code{panelPomp} package \citep{breto20panelPomp} in the R programming language, and can be used by adding the argument \code{block = TRUE} in the \code{mif2} function.
+A major advantage of this algorithm is that PIF can be used to fit both unit-specific parameters and shared parameters; in this way, the calibration of shared parameters involves all of the available data instead of just an arbitrarily chosen subset.
+
+We fit the PanelPOMP version of Model~3 using a slight modification of the PIF algorithm, which we call the Block Panel Iterated Filter (BPIF).
+Let $l_{\text{P}}(y_{1:N}^*; \paramVec)$ denote the log-likelihood function of the PanelPOMP version of Model~3.
+The output of the BPIF algorithm is the MLE $\tilde{\paramVec} = \argmax_{\paramVec} l_{\text{P}}(y_{1:N}^*; \paramVec)$.
+The BPIF algorithm is implemented in the \code{panelPomp} package \citep{breto20panelPomp} as the argument \code{block = TRUE} in the \code{mif2} function.
 Pseudo-code for this algorithm is provided in Algorithm~S1 in the supplementary material.
-While the reported parameters were obtained using the decoupled representation of Model~3, simulations for the various vaccination campaigns were obtained using fully coupled version of the model.
-% Simulations from the fitted version of the PanelPOMP model are shown in Fig.~\ref{h3panelsims}.
+
+% While the reported parameters were obtained using the decoupled representation of Model~3, simulations for the various vaccination campaigns were obtained using fully coupled version of the model, implemented using the \code{spatPomp} package.
 
@@
   ) %>%
-  filter(date >= yearsToDate(h3@unit.objects$Artibonite@t0))
+  filter(date >= as.Date(
+    lubridate::round_date(
+      lubridate::date_decimal(h3@unit.objects$Artibonite@t0), unit = 'day'
+    )
+  )
+  )
 
@@
 
-Fitting Model~3 as a PanelPOMP simplifies the problem of parameter estimation, but it also introduces additional technicalities that must be addressed.
+Fitting Model~3 in this fashion simplifies the problem of parameter estimation, but it also introduces additional technicalities that must be addressed.
 One concern is that of obtaining model forecasts, which was the primary goal of \citet{lee20}.
@@
 The simplified PanelPOMP version of Model~3 relies on the observed cholera cases as a covariate, which are unavailable for use in forecasts.
-To address this, we use the MLE obtained for the PanelPOMP approximation of Model~3 as an estimate for the parameters in fully coupled version of the model, which was implemented using the \code{spatPomp} package.
-Simulations from the SpatPOMP version of the model, using the parameters that were fit with the PanelPOMP version of the model, are displayed in Fig.~\ref{h3spatsims}.
+To address this, we use the MLE obtained for the PanelPOMP approximation of Model~3 as an estimate for the parameters in fully coupled version of the model, which was implemented using the \code{spatPomp} package, similar to what was done in \cite{lee20}.
+Simulations from the fully coupled version of the model using the estimated parameter vector $\tilde{\paramVec}$, are displayed in Fig.~\ref{h3spatsims}.
 
@@
         ) %>%
-        filter(date >= yearsToDate(h3Spat@t0)) %>%
+        filter(date >= as.Date(lubridate::round_date(lubridate::date_decimal(h3Spat@t0), unit = 'day'))) %>%
         mutate(date = as.Date(lubridate::round_date(date, unit = 'day')))
@@
 \caption{\label{h3spatsims}
-Simulations from initial conditions using SpatPOMP version of Model~3.
+Simulations from initial conditions using the spatially coupled version of Model~3.
 The black curve represents true case count, the blue line the median of 500 simulations from the model, and the grey ribbons representing $95\%$ confidence interval.
@@
 
-While simulations from the SpatPOMP version of Model~3 resemble the observed data, we note in Table~\ref{tab:likes} that the SpatPOMP version of Model~3 has a lower log-likelihood than the PanelPOMP version using the same model parameters.
-A natural question to ask is if the MLE obtained for the PanelPOMP version of Model~3---obtained using the approximation in Eq.~\myeqref{eq:model3:approx}---could result in a drastically different set of parameters than those that would be obtained by estimating the MLE in the fully coupled model.
-Unfortunately, the answer to this question would require the ability to accurately and reliably estimate the MLE of the fully coupled model, which is beyond the scope of this article.
-% this question is unsatisfactorily left unanswered, because the answer would require knowledge of the the MLE without the simplification, which is an open statistical question and beyond the scope of this article.
-% While this approach may provide a reasonable estimate of the model parameters, it is unknown by how much this estimate may be improved.
-This highlights the need for advancements in statistical methodology that permit inference on models with coupled metapopulation dynamics.
-
-% \subsection{Comparing mechanistic models with associative benchmarks} %AAAAAAAAAAAAAAA
-\subsection{Model Diagnostics} \label{model_diagnostics}
+% While simulations from the SpatPOMP version of Model~3 resemble the observed data, we note in Table~\ref{tab:likes} that the SpatPOMP version of Model~3 has a lower log-likelihood than the PanelPOMP version using the same model parameters.
+Let $l_{\text{C}}(y_{1:N}^*; \paramVec)$ denote the log-likelihood of the fully coupled model, and let $\hat{\paramVec} = \argmax_{\paramVec} l_{\text{C}}(y_{1:N}^*; \paramVec)$.
+If the goal of parameter estimation is to maximize the likelihood of the fully coupled model, the approach of estimating parameters using a decoupled model is clearly sub optimal, as $l_{\text{C}}(y_{1:N}^*; \hat{\paramVec}) \geq l_{\text{C}}(y_{1:N}^*; \tilde{\paramVec})$.
+A natural question that arises is if the difference between $\hat{\paramVec}$ and $\tilde{\paramVec}$ is scientifically and statistically meaningful.
+The answer to this question requires the ability to reliably estimate $\hat{\paramVec}$, which is model specific and beyond the scope of this article.
+
+In this case study, a brief analysis of the fitted parameters $\tilde{\paramVec}$ provides some interesting insight. 
+Consider the unexpectedly large exponent on rainfall: $r = \Sexpr{signif(p3s["r"],3)}$ (Eq.~\ref{eq:model3:water}).
+Because the value of rainfall $J(t) \in (0, 1)$, the large coefficient results in $J(t)^r \approx 0$. 
+This result may initially be surprising given the importance of rainfall as a seasonal diver of cholera infections in Haiti (see Sec.~\ref{sec:science}).
+One plausible explanation of this result is that the fitting procedure---used to obtain parameter values in the independent department model given the reported cases in other departments---has learned an over-dependence on the reported cholera cases in other departments, and thus the seasonality of cholera incidence in department $u$ depends on the highly correlated covariate of the reported cholera cases in departments $1:U\setminus \{u\}$.
+Whatever the case may be, this fitting procedure results in model likelihoods unable to improve upon a log-linear benchmark (see Sec.~\ref{sec:model_diagnostics} and Tab.~\ref{tab:likes}), and highlights the need for advancements in statistical methodology that permit inference on models with coupled metapopulation dynamics.
+
+\subsection{Model Diagnostics} \label{sec:model_diagnostics}
 
@@
 To make progress, while accepting proper limitations, one must bear in mind the much-quoted observation of \citet{box79} that ``all models are wrong but some are useful.''
-\eic{I WAS WONDERING WHETHER TO SAY SOMETHING LIKE, This is not just good practical advice for applied statistics, but matches broader appreciation of the limitations of all scientific knowledge (REF).
-DO YOU THINK THIS IS USEFUL, OR A DISTRACTION?}
-In this section, we discuss some tools and suggestions for diagnosing mechanistic models with the goal of making the subjective assessment of model ``usefulness" more objective by relying on the quantitative and statistical ability of the model to match the observed data, which we call the model's goodness-of-fit
-\eic{GOODNESS-OF-FIT SEEMED INITIALLY TO ME A MEASURE OF "RIGHT VS WRONG" RATHER THAN "USEFUL VS USELESS".
-I THINK NOW I SEE YOUR POINT: GOODNESS OF FIT CAN GIVE US SOME INDICATION OF HOW WRONG A MODEL IS, WITH THE GUIDING PRINCIPLE THAT A MODEL WHICH IS TOO FAR WRONG IS PROBABLY NOT RELIABLE FOR USEFUL PURPOSES.
-MAYBE WE HAVE TO EXPLAIN THIS TO THE READER?}
+This is not just practical advice for applied statistics, but matches broader appreciation of the limitations of all scientific knowledge \eic{REF???}.
+In this section, we provide some tools and suggestions for diagnosing mechanistic models with the goal of making the subjective assessment of model ``usefulness" more objective.
+To do this, we will rely on the quantitative and statistical ability of the model to match the observed data, which we call the model's {\it goodness-of-fit}, with the guiding principle that a model which cannot adequately describe observed data may not be reliable for useful purposes.
 Goodness-of-fit may provide evidence supporting the causal interpretation of one model versus another, but cannot by itself rule out the possibility of alternative explanations.
@@
 Ideally, the measure should also have favorable theoretical properties.
-Consequently, we focus on log-likelihood as a measure of goodness of fit, and we adjust for the degrees of freedom of the models to be compared by using the Akaike information criterion (AIC) \eic{REF}.
+Consequently, we focus on log-likelihood as a measure of goodness of fit, and we adjust for the degrees of freedom of the models to be compared by using the Akaike information criterion (AIC) \citep{aic74}.
 
@@
       # Find appropriate sub-set
-      in_subset_cols <- h3_bpf[[1]]@times >= as.numeric(dateToYears(as.Date("2014-03-01")))
+      in_subset_cols <- h3_bpf[[1]]@times >= as.numeric(lubridate::decimal_date(as.Date("2014-03-01")))
       in_subset_rows <- h3_bpf[[1]]@unit_names != 'Ouest'
@@
       mod3_coupled_subset_evals <- c()
-      for (i in 1:length(pf)) {
+      for (i in 1:length(h3_bpf)) {
         mod3_coupled_subset_evals <- c(
@@
 # parameter estimation phase.
-set_as_na <- lee3_sub@times < dateToYears(as.Date('2017-06-10'))
+set_as_na <- lee3_sub@times < lubridate::decimal_date(as.Date('2017-06-10'))
 temp_ouest_cases <- lee3_sub@data['casesOuest', ]
@@
 $\textsuperscript{1}$The reported likelihood is an upper bound of the likelihood of the \citet{lee20} model as it is the largest likelihood obtained using their parameter calibration regime.
-$\textsuperscript{2}$\cite{lee20} fit Model~3 to only a subset of the data (March 2014 onward) and did not include data from Ouest in 2015-2016.
-On this subset, the \cite{lee20} fit has a likelihood of $\Sexpr{myround(max(lee3_results$sub_ll_evals), 1)}$.
+$\textsuperscript{2}$\cite{lee20} fit Model~3 to a subset of the data (March 2014 onward, excluding data from Ouest in 2015-2016).
+On this subset, their model has a likelihood of $\Sexpr{myround(max(lee3_results$sub_ll_evals), 1)}$.
 On this same subset, our model has a likelihood of $\Sexpr{myround(mod3_spatPomp_res$subset_ll, 1)}$.
@@
 
-% \eic{I THINK WE WILL HAVE TO EXPLAIN THIS CAREFULLY IF THE POINT IS TO APPEAR NONTRIVIAL.
-IT IS WORTH NOTING THAT LEE ET AL DID NOT PROPERLY FORECAST FROM THE BEST ESTIMATE OF THE CURRENT STATE.
-HOWEVER, IT WILL BE OBVIOUS TO MOST SATISTICAL READERS THAT THIS IS NOT THE BEST THING TO DO.
-IT MAY NOT BE SO OBVIOUS THAT THE DIFFERENCE IS IMPORTANT IF AND ONLY IF A DETERMINISTIC MODEL IS INADEQUATE.}
-
-% As an example, we note that simulations from a well-fit mechanistic model may closely resemble the observed data $y_{1:N}^*$.
-% In such comparisons, these simulations are random draws from the complete estimated joint distribution $f_{\bm{X}^{(\modelCounter)}_{0:N}, \bm{Y}^{(\modelCounter)}_{1:N}}\left(x_{0:N}, y_{1:N}\mid \hat{\paramVec}\right)$, where $\modelCounter$ indexes the model used for simulations.
-% It can therefore be tempting to use simulations from this model up to time $N+s$, $f_{\bm{X}^{(\modelCounter)}_{0:N+s}, \bm{Y}^{(\modelCounter)}_{1:N+s}}\left(x_{0:N+s}, y_{1:N+s}\mid \hat{\paramVec}\right)$, with $s\geq 0$ to project the dynamic system up to a future time $N+s$, as it has been done in previous studies \citep{lee20,???}.
-% This approach, however, does not take advantage of the information about the state of the system that is contained in the observed data.
-% Note that each of the models in question (and, more generally, all models that are described as POMPs) are Markovian, that is, the history of the process $\{\bm{X}^{(\modelCounter)}_s, s\leq t\}$ for model $\modelCounter$ is uninformative about the future of the process $\{\bm{X}^{(\modelCounter)}_s, s\geq t\}$, given the current state $\bm{X}^{(\modelCounter)}_t$.
-% In other words, observing the data $\bm{Y}^{(\modelCounter)}_N = y_N^*$ at time $N$ provides more information about the future state of the system than the initial conditions.
-% In this case, draws from the conditional density $f_{\bm{X}^{(\modelCounter)}_{N:s}, \bm{Y}^{\modelCounter}_{N:s}}\big(x_{N:s}, y_{N:s}\mid \hat{\paramVec}, \bm{X}^{(\modelCounter)}_{N} = \bm{x}^{(\modelCounter)}_N\big)$ should be preferred as forecasts, as these simulations account for the most recent known state of the system.
 The most recently available information about a dynamic system should be used to obtain reliable forecasts.
 Using state space model $m$, this means that in order to obtain a forecast up to time $N+s$, where $s\geq 0$ and $N$ is the index for the last available data, we should simulate from the conditional density $f_{\bm{X}^{(\modelCounter)}_{N:s}, \bm{Y}^{\modelCounter}_{N:s}}\big(x_{N:s}, y_{N:s}\mid \hat{\paramVec}, \bm{X}^{(\modelCounter)}_{N} = \bm{x}^{(\modelCounter)}_N\big)$.
-We note, however, that $\bm{X}^{(\modelCounter)}_N$ is unobservable and therefore draws from the desired conditional density are unobtainable.
+The latent state $\bm{X}^{(\modelCounter)}_N$ is by definition unobservable, however, and therefore draws from the desired conditional density are unobtainable.
 Informed estimates of $\bm{X}^{(\modelCounter)}_N$ given the observed data $\bm{Y}^{(\modelCounter)}_{1:N} = y_{1:N}^*$ can easily be obtained using the filtering distribution.
@@
 Obtaining forecasts in this fashion allows us to obtain results that are consistent with the model assumptions and the observed data.
-% Intuitively, simulating the model starting at the filtering distribution of the most recently available time point is a more appropriate way to project a stochastic dynamic system into the future, as it is not expected that each simulation from initial conditions will result in a latent state at time $N$ consistent with the model and the observation $\bm{Y}^{(\modelCounter)}_N = y_N^*$.
 In this particular case study, projecting the future state of the cholera epidemic in Haiti starting from the draws from the filtering distribution allows the proposed models to benefit from the fact that very few cholera cases had been observed between 2018 and 2019, and that cases appear to be decreasing (i.e., the number of susceptible individuals may be small).
@@
 Table~S7 of their supplement material shows that of their simulations that were consistent with observing zero cases in 2019, all forecasts resulted in the elimination of cholera.
-\eic{THE QUESTION OF WHETHER TO USE A DETERMINISTIC MODEL REMAINS OF CURRENT INTEREST...}
+% \eic{THE QUESTION OF WHETHER TO USE A DETERMINISTIC MODEL REMAINS OF CURRENT INTEREST...}
 
@@
 
-We note---the credit of deterministic models---that of the four fitted models in \cite{lee20}, Model~2 provides the most apparently accurate forecasts.
-This perhaps demonstrates that while deterministic models describe the systems in a less realistic and useful way, the relative ease in fitting these models potentially results in fewer modeling errors.
+To the credit of deterministic models, of the four fitted models in \cite{lee20}, Model~2 provides the most apparently accurate forecasts.
+This perhaps demonstrates that while deterministic models describe the systems in a less realistic way, the relative ease in fitting these models potentially results in fewer modeling errors.
 As the data analysis becomes more refined, however, the deficiencies of deterministic models become increasingly apparent.
 
-\subsection{Comparison of Model Fit to Oustide Scientific Evidence}\label{sec:science}
-
-In a similar fashion, one can take advantage of certain mechanistic features contained in a particular model in order to make inference on a system.
-Examples of this are as diverse as estimating the effective reproductive number ($R_0$) of an infectious disease \citep{he10} or investigating interactions between pedestrians and autonomous vehicles \citep{domeyer22}.
-In our analysis, we demonstrate this ability by examining the results of fitting the flexible cubic spline term in Model~1 (Eq.~\myeqref{model1:lambda}--\myeqref{model1:beta}), which allows for a flexible estimation of seasonality in the force of infection.
-After fitting the model, we explore potential patterns in the seasonal transmission rate by plotting the average value of $\transmission$ in a typical year.
-Fig.~\ref{fig:h1SeasRain} shows that the estimated seasonal transmission rate $\transmission$ mimics the rainfall dynamics in Haiti, despite Model~1 not having access to rainfall data.
-This result provides evidence that rainfall plays an important role in cholera transmission in Haiti, which is consistent with previous studies \citep{eisenberg13,rinaldo12}.
-We also note the increase in the transmission rate during the Fall, which was also noticed at an earlier stage of the epidemic \citep{rinaldo12}.
-
-<<Get Model 1 Seasonality>>=
-plot_model_fun <- function(mod) {
-  params <- coef(mod)
-
-  beta1 <- params['beta1']
-  beta2 <- params['beta2']
-  beta3 <- params['beta3']
-  beta4 <- params['beta4']
-  beta5 <- params['beta5']
-  beta6 <- params['beta6']
-  betat <- params['betat']
-
-  get_beta <- function(time) {
-
-    covar <- t(mod@covar@table)
-    covar_df <- as.data.frame(covar)
-    covar_df$times <- mod@covar@times
-
-    betas <- matrix(c(beta1, beta2, beta3, beta4, beta5, beta6), nrow = 1)
-    si <- matrix(unlist(covar_df[time, -7]), ncol = 1)
-    if (time <= 430) {
-      return(as.numeric(exp(betas %*% si + betat * ((time - 215) / (430-215)))))
-    } else {
-      return(as.numeric(exp(betas %*% si + betat)))
-    }
-  }
-
-  return(get_beta)
-}
-
-get_beta1 <- plot_model_fun(h1)
-
-h1_seas_df <- data.frame(
-  'week' = 431:800,
-  'date' = lubridate::ymd("2010-10-16") + lubridate::weeks(431:800),
-  'trans' = purrr::map_dbl(431:800, get_beta1)
-) %>%
-  mutate(year = lubridate::year(date)) %>%
-  filter(year == 2020) %>%
-  mutate(trans_std = (trans - min(trans)) / (max(trans) - min(trans)))
-
-std_rain <- function(x) {
-  # This function simply standardizes the rain for us.
-  x / max(x)
-}
-
-df_rain <- haitiRainfall %>%
-  dplyr::summarize(
-    date = date, dplyr::across(Artibonite:`Sud-Est`, std_rain)
-  ) %>%
-  pivot_longer(
-    data = .,
-    cols = -date,
-    names_to = 'dep',
-    values_to = 'rainfall'
-  ) %>%
-  mutate(
-    year = lubridate::year(date),
-    week = lubridate::week(date)
-  ) %>%
-  group_by(year, week) %>%
-  summarize(national_weekly_rain = sum(rainfall)) %>%
-  ungroup() %>% # Not needed, but good practice
-  mutate(
-    date_week_start = lubridate::ymd(paste0(2014, "-01-01")) + lubridate::weeks(week - 1),
-    week_date = as.Date(date_week_start, format = "%m-%d"),
-    std_national_rain = (national_weekly_rain - min(national_weekly_rain, na.rm = TRUE)) / (max(national_weekly_rain, na.rm = TRUE) - min(national_weekly_rain, na.rm = TRUE))
-  )
-
-mean_rain <- df_rain %>%
-  group_by(week_date) %>%
-  summarize(mean_rain = mean(std_national_rain)) %>%
-  ungroup() %>%
-  mutate(mean_rain = (mean_rain - min(mean_rain)) / ((max(mean_rain) - min(mean_rain))))
-
-gg_rain <- df_rain %>%
-  filter(week != 1, week != max(week)) %>%
-  ggplot(aes(x = week_date, y = std_national_rain, group = year, col = year)) +
-  geom_line() +
-  theme_bw() +
-  guides(color = 'none') +
-  ylab('Standardized Weekly\nNational Rain') +
-  theme(axis.title.x = element_blank(),
-        axis.ticks.x = element_blank(),
-        axis.text.x = element_blank(),
-        axis.text.y = element_text(size = 7.5),
-        axis.title.y = element_text(size = 9))
-
-gg_trans <- ggplot() +
-  geom_line(data = h1_seas_df, aes(x = date, y = trans_std), linetype = 'dashed') +
-  geom_line(
-    data = mean_rain,
-    aes(x = week_date + lubridate::years(6), y = mean_rain),
-    col = '#2166ac'
-  ) +
-  scale_x_date(date_labels = "%b", date_breaks = '1 month') +
-  ylab('Standardized Seasonal\nContact Rate') +
-  theme(axis.title.x = element_blank(),
-        axis.text = element_text(size = 7.5),
-        axis.title.y = element_text(size = 9))
-
-@
-
-\begin{figure}[ht]
-\centering
-<<Model1_Seasonality_Figure, fig.height=2.9, fig.width=5, fig.align='center'>>=
-cowplot::plot_grid(gg_rain, gg_trans, align = "v", ncol = 1)
-@
-\caption{\label{fig:h1SeasRain}
-(Top) weekly rainfall in Haiti, lighter colors representing more recent years.
-(Bottom) estimated seasonality in the transmission rate (dashed line) plotted alongside mean rainfall (solid line).
-% Observed spikes in rainfall correspond to peaks in the seasonal transmission $\transmission$ in Model~1; similarly, lower values of $\transmission$ correspond to the dry periods.
-}
-\end{figure}
-
-While such inference is useful, one must be careful not to assume the results of a single analysis as a de facto feature of the system.
-It is important instead to recognize and assess the modeling simplifications and assumptions that were used in order to arrive at the conclusions.
-
-\subsection{Predicting the Effect of Interventions}\label{sec:interventions}
-
-An additional benefit of mechanistic modeling is ability to simulate various interventions on a system; this feature is useful to inform policy and was the primary goal of \citet{lee20}.
+Forecasting can be used to simulate the simulate various interventions on a system; this feature is useful to inform policy and was the primary goal of \citet{lee20}.
 Outcomes of their study include estimates for the probability of cholera elimination and cumulative number of cholera infections under several possible vaccination scenarios.
@@
 mod3_probElims$mod <- "Model 3"
-mod3_probElims$time <- yearsToDate(mod3_probElims$time)
+mod3_probElims$time <- as.Date(lubridate::round_date(lubridate::date_decimal(mod3_probElims$time), unit = 'day'))
 
@@
 
+\subsection{Reckoning Scientific Understanding With Fitted Models}\label{sec:science}
+
+The resulting mechanisms in a fitted model can be compared to current scientific knowledge about a system. 
+Synonymity between model based inference and our current understanding of a system may taken as a confirmation of both model based conclusions and our understanding of the system \jwc{REF???}.
+On the other hand, discrepancies found in these comparisons may have the potential to spark new scientific breakthroughs \jwc{REFS???}.
+The potential for new discoveries based on dynamic modeling outcomes is rooted in a model's ability to effectively approximate the system in question, and is therefore inexorably linked to a models goodness-of-fit.
+As such, the perception of scientific conclusions made via dynamic modeling should be weighed by the model's goodness-of-fit.
+
+In our analysis, we demonstrate how one may compare model mechanisms with current scientific understanding by examining the results of fitting the flexible cubic spline term in Model~1 (Eq.~\myeqref{model1:lambda}--\myeqref{model1:beta});
+the cubic splines permit flexible estimation of seasonality in the force of infection.
+After fitting the model, we explore potential patterns in the seasonal transmission rate by plotting the average value of $\transmission$ in a typical year.
+Fig.~\ref{fig:h1SeasRain} shows that the estimated seasonal transmission rate $\transmission$ mimics the rainfall dynamics in Haiti, despite Model~1 not having access to rainfall data.
+This result provides evidence that rainfall played an important role in cholera transmission in Haiti, consistent with previous studies \citep{lemaitre19,eisenberg13};
+the estimated seasonality also features an increased transmission rate during the Fall, which was noticed at an earlier stage of the epidemic \citep{rinaldo12}.
+
+<<Get Model 1 Seasonality>>=
+plot_model_fun <- function(mod) {
+  params <- coef(mod)
+
+  beta1 <- params['beta1']
+  beta2 <- params['beta2']
+  beta3 <- params['beta3']
+  beta4 <- params['beta4']
+  beta5 <- params['beta5']
+  beta6 <- params['beta6']
+  betat <- params['betat']
+
+  get_beta <- function(time) {
+
+    covar <- t(mod@covar@table)
+    covar_df <- as.data.frame(covar)
+    covar_df$times <- mod@covar@times
+
+    betas <- matrix(c(beta1, beta2, beta3, beta4, beta5, beta6), nrow = 1)
+    si <- matrix(unlist(covar_df[time, -7]), ncol = 1)
+    if (time <= 430) {
+      return(as.numeric(exp(betas %*% si + betat * ((time - 215) / (430-215)))))
+    } else {
+      return(as.numeric(exp(betas %*% si + betat)))
+    }
+  }
+
+  return(get_beta)
+}
+
+get_beta1 <- plot_model_fun(h1)
+
+h1_seas_df <- data.frame(
+  'week' = 431:800,
+  'date' = lubridate::ymd("2010-10-16") + lubridate::weeks(431:800),
+  'trans' = purrr::map_dbl(431:800, get_beta1)
+) %>%
+  mutate(year = lubridate::year(date)) %>%
+  filter(year == 2020) %>%
+  mutate(trans_std = (trans - min(trans)) / (max(trans) - min(trans)))
+
+std_rain <- function(x) {
+  # This function simply standardizes the rain for us.
+  x / max(x)
+}
+
+df_rain <- haitiRainfall %>%
+  dplyr::summarize(
+    date = date, dplyr::across(Artibonite:`Sud-Est`, std_rain)
+  ) %>%
+  pivot_longer(
+    data = .,
+    cols = -date,
+    names_to = 'dep',
+    values_to = 'rainfall'
+  ) %>%
+  mutate(
+    year = lubridate::year(date),
+    week = lubridate::week(date)
+  ) %>%
+  group_by(year, week) %>%
+  summarize(national_weekly_rain = sum(rainfall)) %>%
+  ungroup() %>% # Not needed, but good practice
+  mutate(
+    date_week_start = lubridate::ymd(paste0(2014, "-01-01")) + lubridate::weeks(week - 1),
+    week_date = as.Date(date_week_start, format = "%m-%d"),
+    std_national_rain = (national_weekly_rain - min(national_weekly_rain, na.rm = TRUE)) / (max(national_weekly_rain, na.rm = TRUE) - min(national_weekly_rain, na.rm = TRUE))
+  )
+
+mean_rain <- df_rain %>%
+  group_by(week_date) %>%
+  summarize(mean_rain = mean(std_national_rain)) %>%
+  ungroup() %>%
+  mutate(mean_rain = (mean_rain - min(mean_rain)) / ((max(mean_rain) - min(mean_rain))))
+
+gg_rain <- df_rain %>%
+  filter(week != 1, week != max(week)) %>%
+  ggplot(aes(x = week_date, y = std_national_rain, group = year, col = year)) +
+  geom_line() +
+  theme_bw() +
+  guides(color = 'none') +
+  ylab('Standardized Weekly\nNational Rain') +
+  theme(axis.title.x = element_blank(),
+        axis.ticks.x = element_blank(),
+        axis.text.x = element_blank(),
+        axis.text.y = element_text(size = 7.5),
+        axis.title.y = element_text(size = 9))
+
+gg_trans <- ggplot() +
+  geom_line(data = h1_seas_df, aes(x = date, y = trans_std), linetype = 'dashed') +
+  geom_line(
+    data = mean_rain,
+    aes(x = week_date + lubridate::years(6), y = mean_rain),
+    col = '#2166ac'
+  ) +
+  scale_x_date(date_labels = "%b", date_breaks = '1 month') +
+  ylab('Standardized Seasonal\nContact Rate') +
+  theme(axis.title.x = element_blank(),
+        axis.text = element_text(size = 7.5),
+        axis.title.y = element_text(size = 9))
+
+@
+
+\begin{figure}[ht]
+\centering
+<<Model1_Seasonality_Figure, fig.height=2.9, fig.width=5, fig.align='center'>>=
+cowplot::plot_grid(gg_rain, gg_trans, align = "v", ncol = 1)
+@
+\caption{\label{fig:h1SeasRain}
+(Top) weekly rainfall in Haiti, lighter colors representing more recent years.
+(Bottom) estimated seasonality in the transmission rate (dashed line) plotted alongside mean rainfall (solid line).
+}
+\end{figure}
+
+While dynamic models may potentially result in meaningful scientific insights, one must be careful not to assume the results of a single analysis as a de-facto feature of the system.
+It is important instead to recognize and assess the modeling simplifications and assumptions that were used in order to arrive at the conclusions.
+In epidemiological studies, for example, it is unreasonable to expect that our quantitative understanding of individual-level processes will perfectly match model parameters that were fit to population-level case counts \jwc{REF??}.
+This makes the direct interpretation of such parameters difficult.
+
+Our case study provides an example of this in the parameter estimate for the duration of natural immunity due to cholera infection $\muRS^{-1}$. 
+Under the framework of Model~2, the best estimate for this parameter is $\Sexpr{signif(1/h2_fit$h2_params["sigma"],2)}$, suggesting that individuals have effectively permanent immunity to cholera once infected, and thus the pool of susceptible individuals decreases over time.
+The interpretation of this result is muddled due to the fact that the data ranged from 2010-2019, and therefore estimates of immunity longer than 10 years---the upper end of previous estimates of natural immunity \citep{king08}---effectively result in the same model dynamics.
+The depletion of susceptible individuals may also be attributed to confounding mechanisms that are unaccounted for in the model---such as vaccination and other non-pharmaceutical interventions that reduce cholera transmission \citep{trevisin22, rebaudet21}---that were not accounted for in the model.
+Perhaps the best interpretation of this results, then, is that the model most adequately describes the observed data by having a steady decrease in the number of susceptible individuals; in this case, the fact that the parameter estimate $\muRS$ is inconsistent with the expected result can be taken as evidence of model mispecification.
+
 \section{Advising policy via dynamic modeling}\label{sec:advice}

### 2022-09-16 ms.Rnw
Document is now fully dynamic

@@
 
+<<FitModel3, echo=FALSE, include=FALSE, message=FALSE>>=
+set.seed(636813)
+
+if (RUN_LEVEL == 1){
+
+  # Global search, All params
+  SEARCH1 <- list(
+    NP = 50, NMIF = 3, NREPS = 3,
+    NREPS_EVAL = 3, NP_EVAL = 100
+  )
+
+  # Local search, unit specific params
+  SEARCH2 <- list(
+    TOP_N = 1, NP = 50, NMIF = 3, NREPS = 3, NREPS_EVAL = 3,
+    NP_EVAL = 100
+  )
+
+  # Number of searches
+  N_SEARCHES <- 2L
+
+  # THESE SEARCHES AREN'T CONDUCTED
+  SEARCH3 <- list(
+    TOP_N = 1, NP = 20, NMIF = 3, NREPS = 3, NREPS_EVAL = 3, NP_EVAL = 25
+  )
+
+  SEARCH4 <- list(
+    TOP_N = 1, NP = 20, NMIF = 3, NREPS = 3, NREPS_EVAL = 3, NP_EVAL = 25
+  )
+
+  SEARCH5 <- list(
+    TOP_N = 1, NP = 20, NMIF = 3, NREPS = 3, NREPS_EVAL = 3, NP_EVAL = 25
+  )
+} else if (RUN_LEVEL == 2) {
+
+  # Global search, all parameters
+  SEARCH1 <- list(
+    NP = 200, NMIF = 10, NREPS = 36, NREPS_EVAL = 36, NP_EVAL = 500
+  )
+
+  # Local search, unit specific
+  SEARCH2 <- list(
+    TOP_N = 2, NP = 200, NMIF = 20, NREPS = 18, NREPS_EVAL = 36, NP_EVAL = 500
+  )
+
+  # Local search, shared parameters
+  SEARCH3 <- list(
+    TOP_N = 2, NP = 200, NMIF = 10, NREPS = 18, NREPS_EVAL = 36, NP_EVAL = 500
+  )
+
+  # Local search, unit specific
+  SEARCH4 <- list(
+    TOP_N = 2, NP = 200, NMIF = 10, NREPS = 18, NREPS_EVAL = 36, NP_EVAL = 500
+  )
+
+  # Local search, all parameters
+  SEARCH5 <- list(
+    TOP_N = 2, NP = 200, NMIF =  5, NREPS = 18, NREPS_EVAL = 36, NP_EVAL = 500
+  )
+
+  # Number of searches
+  N_SEARCHES <- 5L
+} else if (RUN_LEVEL == 3) {
+
+  # Global search, all parameters
+  SEARCH1 <- list(
+    NP = 1500, NMIF = 100, NREPS = 144, NREPS_EVAL = 36, NP_EVAL = 2000
+  )
+
+  # Local search, unit specific parameters
+  SEARCH2 <- list(
+    TOP_N = 4, NP = 2000, NMIF = 100, NREPS = 18, NREPS_EVAL = 36, NP_EVAL = 4000
+  )
+
+  # Local search, shared parameters
+  SEARCH3 <- list(
+    TOP_N = 3, NP = 4000, NMIF = 50, NREPS = 24, NREPS_EVAL = 36, NP_EVAL = 5000
+  )
+
+  # Local search, unit specific parameters
+  SEARCH4 <- list(
+    TOP_N = 3, NP = 4000, NMIF = 50, NREPS = 24, NREPS_EVAL = 36, NP_EVAL = 5000
+  )
+
+  # Local search, all parameters
+  SEARCH5 <- list(
+    TOP_N = 3, NP = 4000, NMIF = 25, NREPS = 24, NREPS_EVAL = 36, NP_EVAL = 5000
+  )
+
+  # Total number of searches
+  N_SEARCHES <- 5L
+} else {
+  stop("Run level must be 1, 2 or 3.")
+}
+
+haiti3_fit <- bake(
+  file = paste0("model3/", rl_dir, "haiti3_fit.rds"), {
+    fit_haiti3(
+      search1 = SEARCH1,
+      search2 = SEARCH2,
+      search3 = SEARCH3,
+      search4 = SEARCH4,
+      search5 = SEARCH5,
+      n_searches = N_SEARCHES,
+      ncores = cores
+    )
+  }
+)
+
+rm(SEARCH1, SEARCH2, SEARCH3, SEARCH4, SEARCH5, N_SEARCHES)
+gc()
+@
+
 <<Load Model Parameters, echo=FALSE, results='hide'>>=
-load('model3/output/PanelLocalAll_PF.rda')  # Final model 3 evaluations
-load('model3/output/PanelLocalAll.rda')  # Final model 3 parameters (MIF search)
-ll_mod3 <- mif_logLik
-MIF_mod3 <- local_MIF2_search
-rm(mif_logLik, local_MIF2_search, tLocal)
+# load('model3/output/PanelLocalAll_PF.rda')  # Final model 3 evaluations
+# load('model3/output/PanelLocalAll.rda')  # Final model 3 parameters (MIF search)
+
+# haiti3_fit contains ll estimate for many parameter sets, so we want to find
+# the highest value
+mod3_ll <- max(
+  haiti3_fit[[length(haiti3_fit)]]$logLiks[, 'logLik'],
+  na.rm = TRUE
+)
+
+# MIF_mod3 <- local_MIF2_search
+# rm(mif_logLik, local_MIF2_search, tLocal)
 
@@
 
-best_m3 <- ll_mod3 %>%
-  arrange(-logLik) %>%
-  slice_head(n = 1) %>%
+# best_m3 <- ll_mod3 %>%
+#   arrange(-logLik) %>%
+#   slice_head(n = 1) %>%
+#   pull(which)
+
+best_m3 <- haiti3_fit[[length(haiti3_fit)]]$logLiks %>%
+  filter(logLik == max(logLik, na.rm = TRUE)) %>%
   pull(which)
 
-p3 <- coef(MIF_mod3[[best_m3]])
-rm(MIF_mod3)
+p3 <- haiti3_fit[[length(haiti3_fit)]]$params[best_m3, ]
 gc()
@@
 NREPS_EVAL <- switch(RUN_LEVEL,   3,   8,    36)
-NP_EVAL    <- switch(RUN_LEVEL, 50, 500, 10000)
+NP_EVAL    <- switch(RUN_LEVEL, 100, 500, 10000)
 
@@
 # Load pre-calculated model 3 likelihood
-mod3_ll <- ll_mod3 %>%
-  arrange(-logLik) %>%
-  slice_head(n = 1) %>%
-  pull(logLik)
+# mod3_ll <- ll_mod3 %>%
+#   arrange(-logLik) %>%
+#   slice_head(n = 1) %>%
+#   pull(logLik)
 
@@
       h3_bpf <- foreach(i = 1:NREPS_EVAL, .combine = c) %dopar% {
-        bpfilter(h3Spat, Np = NP_EVAL, block_size = 1)
+        bpfilter(h3Spat, Np = NP_EVAL, block_size = 1,
+                 params = haitipkg:::panelParms_toSpatParms3(p3))
       }

### 2022-09-16 ms.Rnw
Changed n parameters fit model 3

@@
 # model 3 AIC
-mod3_n_params <- 39
+mod3_n_params <- 29
 mod3_aic <- 2 * mod3_n_params - 2 * mod3_ll
@@
 
-mod3_spatPomp_n_params <- 39
+mod3_spatPomp_n_params <- 29
 mod3_spatPomp_aic <- 2 * mod3_spatPomp_n_params - 2 * mod3_spatPomp_res$ll

### 2022-09-17 ms.Rnw
Reduced size of saved data

@@
   guess <- coef(lee1_epi_mif)
-  sims <- simulate(h1_epi, nsim = N_LEE_SIMS, format = 'data.frame',
-                   params = guess)  # 20 sims for each set of parameters
+  sims <- simulate(
+    h1_epi,
+    nsim = N_LEE_SIMS,
+    format = 'data.frame',
+    params = guess
+  )  %>%
+    dplyr::filter(week == max(week)) %>%
+    dplyr::select(-.id, -week)
+
   sims$param_set <- i
   sims
-} -> all_sims
+} -> start_states
 
@@
 h1_end <- haiti1(period = "endemic")
-start_states <- all_sims %>% filter(week == max(week))
 
@@
           filter(param_set == i) %>%
-          select(-.id, -week, -param_set) %>%
+          select(-param_set) %>%
           sapply(., median, na.rm = TRUE) -> rinit_parms
@@
 rm(
-  lee_end_ll, lee_epi_ll, h1_end, h1_epi, all_sims, end_start,
+  lee_end_ll, lee_epi_ll, h1_end, h1_epi, end_start,
   N_LEE_SIMS, N_PARAMS, NMIF, NP_EVAL, NP_MIF, NREPS_EVAL

### 2022-09-20 ms.Rnw
one more reference

@@
 To make progress, while accepting proper limitations, one must bear in mind the much-quoted observation of \citet{box79} that ``all models are wrong but some are useful.''
-This is not just practical advice for applied statistics, but matches broader appreciation of the limitations of all scientific knowledge \eic{REF???}.
+Beyond being good practical advice for applied statistics, this assertion is relevant for the philosophical justification of statistical inference as severe testing \citep[][Sec.~4.8]{mayo18}.
 In this section, we provide some tools and suggestions for diagnosing mechanistic models with the goal of making the subjective assessment of model ``usefulness" more objective.

### 2022-09-21 ms.Rnw
Major revision

--- 2022-09-17 ms.Rnw
@@
 
-Quantitative models for dynamic systems offer potential for designing effective control measures.
 Regulation of biological populations is a fundamental topic in epidemiology, ecology, fisheries and agriculture.
-Quantitative models for these population dynamics may be nonlinear and stochastic, with the resulting complexities compounded by incomplete understanding of the underlying biological mechanisms and by partial observability of the system variables.
+Population dynamics may be nonlinear and stochastic, with the resulting complexities compounded by incomplete understanding of the underlying biological mechanisms and by partial observability of the system variables.
+Quantitative models for these dynamic systems offer potential for designing effective control measures.
 Developing and testing such models, and assessing their fitness for guiding policy, is a challenging statistical task.
@@
 
-The failure of \cite{lee20} to correctly predict the elimination of cholera has been debated \citep{francois20,rebaudetComment20,henrys20,leeReply20}.
-\citet{rebaudetComment20} suggested that the models proposed by \citet{lee20} were too unrealistic.
+The discrepancy between the model based conclusions of \cite{lee20} and the subsequent elimination of cholera cholera has been debated \citep{francois20,rebaudetComment20,henrys20,leeReply20}.
+Suggested origins of this discrepancy include the use of unrealistic models \citep{rebaudetComment20} and having unrealistic criteria for cholera elimination \citep{henrys20}.
 We find a more nuanced conclusion: attention to methodological details in model fitting, diagnosis and forecasting can improve each of the proposed model's ability to quantitatively describe observed data.
@@
 
+While this work is focused on the models proposed in \cite{lee20}, we believe that our suggestions are relevant to a larger class of problems. 
+In order to determine the extent to which this work generalizes, we preformed a literature review by searching PubMed with keywords: Haiti, cholera, and model.
+The search resulted in 66 papers, of which 32 proposed or used dynamic models to describe the cholera epidemic in Haiti.
+This literature review confirmed that close attention to details in fitting, diagnosing, and forecasting is of great importance when modeling cholera in Haiti.
+% Although the literature review focused on cholera and Haiti, 
+The wide relevance of our findings in these papers---supplemented with studies of other epidemiological and dynamic systems---suggests that this work may also be relevant to a much broader class of problems.
+
 We proceed by introducing Models~1--3 in Sec.~\ref{sec:models};
 in Sec.~\ref{sec:methods}, we present a methodological approach to examining and refining these models, and then use improved model fits to project cholera incidence in Haiti under various vaccination scenarios.
-We then conclude with recommendations on the use of mechanistic models to inform policy decisions in Sec.~\ref{sec:advice}.
+This is followed by a consideration of the robustness of model based policy recommendations in Sec.~\ref{sec:advice}, and a discussion in Sec.~\ref{sec:discussion}.
 
@@
 The ability to maximize the likelihood allows for likelihood-based inference, like performing statistical tests for potential improvements to the model.
-% TODO: We should address somewhere why we included the process noise in Model 1. Anna suggested this below:
-% We first demonstrate this capability by proposing the inclusion of additional stochasticity in the latent process through the force of infection, similar to that in Model~3.
-Eq.~\myeqref{model1:lambda} becomes
-% \begin{equation}
-% \label{model1:lambda_wn}
-% \lambda(t) = \Big(\sum_{k=0}^{K}  I_{k}(t) + \asymptomRelativeInfect \sum_{k=0}^{K} A_{k}(t) \Big)^\nu d\Gamma_u(t)/dt \transmission(t)/N
-% \end{equation}
-% where $d\Gamma_u(t)/dt$ is multiplicative Gamma-distributed white noise.
-%
 % Part of the appeal of mechanistic models lies in their ability to illustrate a dynamical system.
@@
 % This need, itself, is dependent upon inclusion of sufficient stochasticity in the model \citep{breto09}.
-% For now, we motivate this decision by its presence in Model~3 and leave the more rigorous assessment of this adjustment to the supplement.
-% The above can no doubt be improved.
-% \arc{I think that evaluating the inclusion of both the noise and the trend parameter in the main text may distract from the main point of this manuscript, which is why I mentioned putting it in the supplement.}
-% \eic{The dynamic noise issue is important, but is well described elsewhere so can perhaps be handled mostly by references.
-Maybe we don't need to do more than mention that Lee et al constrained the noise parameter to be zero, but we found considerable advantage from including the parameter (XXX log units of likelihood) consistent with previous investigations (refs).}
-% Transition phrase here?
 We demonstrate this capability by proposing a linear trend $\transmissionTrend$ in transmission in Eq.~\myeqref{model1:beta}:
@@
 When combined with online availability of data and code, objective measures of fit provide a powerful tool to accelerate scientific progress, following the paradigm of the {\it common task framework} \citep[][Sec.~6]{donoho17}.
-In our literature review of the Haiti cholera epidemic \eic{CHECK THIS.
-HOW/WHERE SHOULD WE DESCRIBE THIS?} no quantitative measures of goodness of fit, and no benchmark models, were considered in any of the ??
-papers which calibrated a mechanistic model to data in order to obtain scientific conclusions.
+In our literature review of the Haiti cholera epidemic \eic{CHECK THIS} no quantitative measures of goodness of fit, and no benchmark models, were considered in any of the 24 papers which calibrated a mechanistic model to data in order to obtain scientific conclusions.
 
@@
 
-Another consideration to make when obtaining model forecasts is that of parameter uncertainty.
-It has been noted that the uncertainty in just a single parameter can lead to drastically different projections \citep{saltelli20}.
-One possible approach to account for parameter uncertainty in model forecasts is by obtaining confidence intervals for each parameter, sampling parameters from the confidence intervals, simulating the model with the resulting parameter set, and then weighing the resulting model projections based on the likelihood of the given set of parameters, as was done in \citet{king15}.
-\eic{THE ``OBVIOUS'' THING TO DO HERE IS A BAYESIAN APPROACH.
-IN SOME SENSE, KING15 IS EMPIRICAL BAYES.
-IT MAY TAKE CARE TO DISCUSS THIS WITHOUT OPENING UP A CAN OF WORMS - I CAN HAVE A GO AT MAKING SUGGESTIONS, NEXT TIME I'M WORKING ON THE MS.}
-
-Note that in order to obtain projections that are consistent with the observed data, one must first be able to sample from the filtering distribution given each set of parameters.
-This approach can be done for both deterministic and stochastic models, but requires a large number of computations, especially as the number of observations and model parameters increase.
-Because the focus of this study is on model fitting and evaluation, we do not provide model projections accounting for parameter uncertainty.
-Instead, we use the projections from point estimates to highlight a major deficiency of deterministic models, which is that the only variability in model projections is a result of parameter uncertainty, which leads to over-confidence in the projections.
-This observation is consistent with those made in \cite{king15}, and suggests that stochastic models should be preferred over deterministic models.
-
+It has been noted that the uncertainty in just a single parameter can lead to drastically different forecasts \citep{saltelli20};
+parameter uncertainty should therefore be considered when obtaining model forecasts to influence policy. 
+If a Bayesian technique is used for parameter estimation, the most natural way to account for parameter uncertainty is to obtain multiple simulations from the model where each simulation is obtained using parameters $\paramVec$ drawn from the posterior distribution $\paramVec \sim f_{\Theta}\big(\paramVec | \bm{Y}^{(\modelCounter)}_{1:N} = y_{1:N}^*\big)$.
+In the frequentest case, one approach is a quasi-empirical Bayes: endow to set $\Theta_K = \{\theta_1, \theta_2, \ldots, \theta_K\}$ with a discrete uniform distribution, such that $P\big(\paramVec = \theta_k\big) = \frac{1}{K}$ for all values $k \in \seq{1}{K}$.
+In this case, the posterior distribution of $\paramVec | \bm{Y}^{(\modelCounter)}_{1:N}$ can be expressed as: 
+$P\big(\paramVec = \theta_k | \bm{Y}^{(\modelCounter)}_{1:N} = y_{1:N}^*\big) = \frac{l(y_{1:N}^*; \theta_k)}{\sum_{j = 1}^K l(y_{1:N}^*; \theta_j)}$; to complete the formulation, we then choose the set $\Theta_K$ such that each $\theta_k \in \Theta_K$ is a single output of the MIF2 algorithm, which has been shown to be within a neighborhood of the MLE \cite{ionides15}.
+This approach has been taken in previous studies \citep{king15}, and the result is simply weighing parameter sets based on their corresponding likelihoods: parameters with higher likelihoods should be used more frequently in model forecasts.
+This can be done for both deterministic and stochastic models, but requires a large number of computations, especially as the number of observations and model parameters increase.
+Because our direct aim is not to provide forecasts for policy recommendation, we do not include parameter uncertainty in our forecasts.
+Instead, we use the projections from point estimates to highlight a major deficiency of deterministic models, which is that the only variability in model projections is a result of parameter uncertainty, which leads to over-confidence in forecasts \cite{king15}.
 To the credit of deterministic models, of the four fitted models in \cite{lee20}, Model~2 provides the most apparently accurate forecasts.
@@
 Reporting rate
-  & $\reportRate=(\Sexpr{myround(p1["rho_epi"],3)}, \Sexpr{myround(p1["rho_end"],3)})$
+  & $\reportRate=\Sexpr{myround(p1["rho"], 3)}$
   & (S16) % TODO: Make sure manual equations match!
@@
 The resulting mechanisms in a fitted model can be compared to current scientific knowledge about a system.
-Synonymity between model based inference and our current understanding of a system may taken as a confirmation of both model based conclusions and our understanding of the system \jwc{REF???}.
-On the other hand, discrepancies found in these comparisons may have the potential to spark new scientific breakthroughs \jwc{REFS???}.
-The potential for new discoveries based on dynamic modeling outcomes is rooted in a model's ability to effectively approximate the system in question, and is therefore inexorably linked to a models goodness-of-fit.
-As such, the perception of scientific conclusions made via dynamic modeling should be weighed by the model's goodness-of-fit.
+Agreement between model based inference and our current understanding of a system may taken as a confirmation of both model based conclusions and our understanding of the system.
+On the other hand, discrepancies found in these comparisons may have the potential to spark new scientific knowledge \citep{ganusov16}.
+The potential for new discoveries based on dynamic modeling outcomes, however, is rooted in a model's ability to effectively approximate the system in question, and is therefore inexorably linked to a models goodness-of-fit.
+As such, conclusions based on dynamic models that poorly fit observed data should not be considered as much as those resulting from models that describe observed data well.
 
@@
 the cubic splines permit flexible estimation of seasonality in the force of infection.
-After fitting the model, we explore potential patterns in the seasonal transmission rate by plotting the average value of $\transmission$ in a typical year.
+After fitting the model parameters, we explore potential patterns in the seasonal transmission rate by plotting the average value of $\transmission$ in a typical year.
 Fig.~\ref{fig:h1SeasRain} shows that the estimated seasonal transmission rate $\transmission$ mimics the rainfall dynamics in Haiti, despite Model~1 not having access to rainfall data.
@@
 The depletion of susceptible individuals may also be attributed to confounding mechanisms that are unaccounted for in the model---such as vaccination and other non-pharmaceutical interventions that reduce cholera transmission \citep{trevisin22, rebaudet21}---that were not accounted for in the model.
-Perhaps the best interpretation of this results, then, is that the model most adequately describes the observed data by having a steady decrease in the number of susceptible individuals; in this case, the fact that the parameter estimate $\muRS$ is inconsistent with the expected result can be taken as evidence of model mispecification.
-
-\section{Advising policy via dynamic modeling}\label{sec:advice}
-
-\eic{I THINK IT IS INTERESTING THAT THE MODEL-SUPPORTED SUGGESTION BY \citep{andrews11} HAS LARGELY PROVED ACCURATE (VACCINATION, NON-PHARMACEUTICAL INTERVENTIONS, AND INFRASTRUCTURE IMPROVEMENTS ALL HAVE A ROLE AND SHOULD BE FOLLOWED SIMULTANEOUSLY)DESPITE LEGITIMATE CONCERNS ABOUT WHETHER THE SCIENTIFIC SUPPORT FOR THE MODEL IS STRONG ENOUGH TO ROBUSTLY GUIDE POLICY \citep{grad12}.
-THE MAIN CLAIM OF ANDREWS11 IS THAT FORECASTS AND POLICY NOT SUPPORTED BY A MODEL SEEM EVEN MORE UN-ROOTED THAN RECOMMENDATIONS BASED ON A WEAKLY SUPPORTED MODEL.
-THE CONCLUSION FROM ALL THIS, WHICH I THINK IS SUPPORTED BY OUR RE-ANALYSIS OF LEE ET AL, IS THAT MECHANISTIC MODELS FITTED TO DATA ARE A STRONG WAY TO ENFORCE CONSISTENT THINKING, BUT WEAKER AS A WAY TO CHECK THAT YOU ARE ACTUALLY RIGHT DUE TO THE LIMITATIONS OF INFERRING CAUSAL MECHANISMS FROM OBSERVATIONAL DATA.
-POLICY-MAKERS SHOULD FOLLOW AN APPROACH WHICH IS SELF-CONSISTENT AND CONSISTENT WITH PREVIOUS EVIDENCE THAT MIGHT WORK (WHICH MODELS CAN HELP WITH) BUT THEY MUST BE AGILE AS NEW EVIDENCE EMERGES.
-MODELS, TOGETHER WITH DATA ANALYSIS, CAN ALSO HELP REVEAL WHEN THE EVIDENCE SUPPORTING A P0LICY IS BREAKING DOWN, FORCING A RE-THINK.
-}
+Perhaps the best interpretation of this results, then, is that the model most adequately describes the observed data by having a steady decrease in the number of susceptible individuals.
+In this case, the fact that the parameter estimate $\muRS$ is inconsistent with the expected result can be taken as evidence of model mispecification.
+
+\section{Robustness of model based conclusions}\label{sec:advice}
+
+% \eic{I THINK IT IS INTERESTING THAT THE MODEL-SUPPORTED SUGGESTION BY \citep{andrews11} HAS LARGELY PROVED ACCURATE (VACCINATION, NON-PHARMACEUTICAL INTERVENTIONS, AND INFRASTRUCTURE IMPROVEMENTS ALL HAVE A ROLE AND SHOULD BE FOLLOWED SIMULTANEOUSLY)DESPITE LEGITIMATE CONCERNS ABOUT WHETHER THE SCIENTIFIC SUPPORT FOR THE MODEL IS STRONG ENOUGH TO ROBUSTLY GUIDE POLICY \citep{grad12}.
+% THE MAIN CLAIM OF ANDREWS11 IS THAT FORECASTS AND POLICY NOT SUPPORTED BY A MODEL SEEM EVEN MORE UN-ROOTED THAN RECOMMENDATIONS BASED ON A WEAKLY SUPPORTED MODEL.
+% THE CONCLUSION FROM ALL THIS, WHICH I THINK IS SUPPORTED BY OUR RE-ANALYSIS OF LEE ET AL, IS THAT MECHANISTIC MODELS FITTED TO DATA ARE A STRONG WAY TO ENFORCE CONSISTENT THINKING, BUT WEAKER AS A WAY TO CHECK THAT YOU ARE ACTUALLY RIGHT DUE TO THE LIMITATIONS OF INFERRING CAUSAL MECHANISMS FROM OBSERVATIONAL DATA.
+% POLICY-MAKERS SHOULD FOLLOW AN APPROACH WHICH IS SELF-CONSISTENT AND CONSISTENT WITH PREVIOUS EVIDENCE THAT MIGHT WORK (WHICH MODELS CAN HELP WITH) BUT THEY MUST BE AGILE AS NEW EVIDENCE EMERGES.
+% MODELS, TOGETHER WITH DATA ANALYSIS, CAN ALSO HELP REVEAL WHEN THE EVIDENCE SUPPORTING A P0LICY IS BREAKING DOWN, FORCING A RE-THINK.
+% }
 
@@
 We provided a few suggestions that may aid the fitting of mechanistic models such as comparing model likelihoods to a benchmark.
-Improved model fits allows for meaningful statistical inference that may provide valuable insight on a dynamic system and may improve the accuracy of model-based projections.
-% TODO: "Improved model fits", or "Improved model fitting", based on Anna's suggestion.
+Improved model fits allows for meaningful statistical inference that may provide valuable insight on a dynamic system and may improve the accuracy of model forecasts.
 Caution is nonetheless needed when making policy based on modeling conclusions, as model misspecification may invalidate conclusions.
-% TODO: Citation above?
-
-Various suggestions been made about why \citet{lee20} failed to accurately predict the eventual eradication of cholera from Haiti, including model misspecification, overly difficult elimination criteria, and a potential conflict of interest \citep{rebaudetComment20,henrys20}.
-Here, we instead argue that careful attention to important statistical details could have correctly resulted in the conclusion of imminent cholera elimination.
+
+In this article we argue that careful attention to important statistical details could have led the models in \cite{lee20} to correctly predict the imminent cholera elimination.
 We acknowledge the benefit of hindsight: our demonstration of a statistically principled route to obtain better-fitting models with more accurate predictions does not rule out the possibility of discovering other models that fit well yet predict poorly.
-
 We used the same data and models, and even much of the same code, as \citet{lee20}, and yet ended up with drastically different conclusions.
@@
 
-We acknowledge there are limitations to this study;
-one example was the inability to fit model parameters to the fully-coupled version of Model~3.
+We acknowledge there are limitations to this study, such as the inability to fit model parameters to the fully-coupled version of Model~3.
 Promising theoretical and methodological developments \citep{ning21ibpf,ionides22} based on the Block Particle Filter \citep{rebeschini15} may potentially be used to obtain better parameter estimates for Model~3 in future work.
@@
 Studies that combine model development with thoughtful data analysis, supported by a high standard of reproducibility, build knowledge about the system under investigation.
-Cautionary warnings about the difficulties inherent in understanding complex systems \citep{saltelli20,ioannidis20} should motivate us to follow best practices in data analysis, rather than avoiding the challenge.
+Cautionary warnings about the difficulties inherent in understanding complex systems \citep{saltelli20,ioannidis20,ganusov16} should motivate us to follow best practices in data analysis, rather than avoiding the challenge.
 

### 2022-09-21 ms.Rnw
Minor edits

@@
 Here we chose to fit the model to transformed case counts rather than adding overdispersion to the measurement model with the goal of making the fewest changes to the model in \cite{lee20} as possible.
-In practice, however, the inclusion of overdispersion in the measurement process may be preferred \jwc{REFS???}.
+In practice, however, the inclusion of overdispersion in the measurement process may be preferred.
 
@@
 Comparing these results across models---including benchmark models---can help researchers identify potential model deficiencies, or errors in the observed data.
-Additional tools for assessing the goodness-of-fit of a model include plotting the effective sample size of each observation \jwc{REF???} and comparing any statistic of the observed data to simulations from the model, which is sometimes referred to as diagnostic probes (for example, the autocorrelation function (ACF), as was done in \cite{king15}).
+Additional tools for assessing the goodness-of-fit of a model include plotting the effective sample size of each observation, and comparing any statistic of the observed data to simulations from the model---which is sometimes referred to as diagnostic probes.
 
@@
 If a Bayesian technique is used for parameter estimation, the most natural way to account for parameter uncertainty is to obtain multiple simulations from the model where each simulation is obtained using parameters $\paramVec$ drawn from the posterior distribution $\paramVec \sim f_{\Theta}\big(\paramVec | \bm{Y}^{(\modelCounter)}_{1:N} = y_{1:N}^*\big)$.
-In the frequentest case, one approach is a quasi-empirical Bayes: endow to set $\Theta_K = \{\theta_1, \theta_2, \ldots, \theta_K\}$ with a discrete uniform distribution, such that $P\big(\paramVec = \theta_k\big) = \frac{1}{K}$ for all values $k \in \seq{1}{K}$.
+In the frequentist case, one approach is a quasi-empirical Bayes: endow to set $\Theta_K = \{\theta_1, \theta_2, \ldots, \theta_K\}$ with a discrete uniform distribution, such that $P\big(\paramVec = \theta_k\big) = \frac{1}{K}$ for all values $k \in \seq{1}{K}$.
 In this case, the posterior distribution of $\paramVec | \bm{Y}^{(\modelCounter)}_{1:N}$ can be expressed as: 
@@
 It is important instead to recognize and assess the modeling simplifications and assumptions that were used in order to arrive at the conclusions.
-In epidemiological studies, for example, it is unreasonable to expect that our quantitative understanding of individual-level processes will perfectly match model parameters that were fit to population-level case counts \jwc{REF??}.
-This makes the direct interpretation of such parameters difficult.
+In epidemiological studies, for example, it is unreasonable to expect that our quantitative understanding of individual-level processes will perfectly match model parameters that were fit to population-level case counts \jwc{REF??},
+making direct interpretation estimated parameters difficult.
 
@@
 The depletion of susceptible individuals may also be attributed to confounding mechanisms that are unaccounted for in the model---such as vaccination and other non-pharmaceutical interventions that reduce cholera transmission \citep{trevisin22, rebaudet21}---that were not accounted for in the model.
-Perhaps the best interpretation of this results, then, is that the model most adequately describes the observed data by having a steady decrease in the number of susceptible individuals.
-In this case, the fact that the parameter estimate $\muRS$ is inconsistent with the expected result can be taken as evidence of model mispecification.
+Perhaps the best interpretation of the estimated parameter, then, is that under model assumptions, the model most adequately describes the observed data by having a steady decrease in the number of susceptible individuals.
+The poor model fit compared to a log-linear benchmark (see Table~\ref{tab:likes}), and the large inconsistenty between model parameters and our understanding of the system, may be taken as evidence of model mispecificaiton.
 

### 2022-09-22 ms.Rnw
Minor edit

@@
 It is important instead to recognize and assess the modeling simplifications and assumptions that were used in order to arrive at the conclusions.
-In epidemiological studies, for example, it is unreasonable to expect that our quantitative understanding of individual-level processes will perfectly match model parameters that were fit to population-level case counts \jwc{REF??},
+In epidemiological studies, for example, it is unreasonable to expect that our quantitative understanding of individual-level processes will perfectly match model parameters that were fit to population-level case counts \citep{he10} \jwc{ED: ARE THERE BETTER REFERENCES FOR THIS?},
 making direct interpretation estimated parameters difficult.

### 2022-09-22 ms.Rnw
minor edits

@@
 In order to determine the extent to which this work generalizes, we preformed a literature review by searching PubMed with keywords: Haiti, cholera, and model.
-The search resulted in 66 papers, of which 32 proposed or used dynamic models to describe the cholera epidemic in Haiti.
-This literature review confirmed that close attention to details in fitting, diagnosing, and forecasting is of great importance when modeling cholera in Haiti.
+The search resulted in 66 papers, of which 32 used dynamic models to describe the cholera epidemic in Haiti.
+This literature review confirmed that close attention to details in fitting, diagnosing, and forecasting is of great importance when modeling an infectious disease outbreak.
 % Although the literature review focused on cholera and Haiti,
@@
 The observation at time $t_n$ is modeled by a response vector $\bm{Y}^{(\modelCounter)}_n$.
-The latent state vector $\bm{X}^{(\modelCounter)}(t)$ consists of individuals labeled as susceptible (S), infected (I), asymptomatically infected (A), vaccinated (V), and recovered (R), with various sub-divisions sometimes considered in each model for $\modelCounter \in \{1, 2, 3\}$.
+The latent state vector $\bm{X}^{(\modelCounter)}(t)$ consists of individuals labeled as susceptible (S), infected (I), asymptomatically infected (A), vaccinated (V), and recovered (R), with various sub-divisions sometimes considered in each model $\modelCounter$.
 Models~2 and~3 have metapopulation structure, meaning that each individual is a member of a spatial unit, denoted by a subscript $u\in \seq{1}{U}$.
@@
 In the following subsections, complete descriptions of Models~1--3 are provided.
-While the model description is scientifically critical, as well as being necessary for transparency and reproducibility, the model details are not essential to our methodological discussions of how to diagnose and address model misspecification with the purpose of informing policy.
+While the model description is scientifically critical, as well as necessary for transparency and reproducibility, the model details are not essential to our methodological discussions of how to diagnose and address model misspecification with the purpose of informing policy.
 A first-time reader may choose to skim through the rest of this section, and return later.
@@
 
-We consider model fitting (Sec.~\ref{sec:model_fitting}) followed by diagnostic investigations (Sec.~\ref{sec:model_diagnostics}) and forecasting (Sec.~\ref{sec:filter}).
+We consider model fitting (Sec.~\ref{sec:model_fitting}) followed by diagnostic investigations (Sec.~\ref{sec:model_diagnostics}), forecasting (Sec.~\ref{sec:filter}) and interpreting model fit (Sec.~\ref{sec:science}).
 
@@
 
-All three model considered in this study describes cholera dynamics via unobservable states that evolve dynamically with time.
+Each of the three models considered in this study describes cholera dynamics via unobservable states that evolve dynamically with time.
 Despite their similarities, these models represent a diverse selection of possible modeling assumptions: namely, the use of stochastic (Models~1 and 2) or deterministic (Model~2) equations; spatially-heterogeneous meta-population (Models~2 and 3) or spatially-aggregated (Model~1) structure; and the use of covariates (Model~3) versus mathematical equations (Models~1 and 2) to describe a seasonal mechanism.
@@
 One approach to modeling a dynamic system is through probabilistic models.
-With a probabilistic model, we suppose the existence of a joint density $f_{\bm{X}^{(\modelCounter)}_{\seq{0}{N}}, \bm{Y}^{(\modelCounter)}_{\seq{1}{N}}}$, with $\bm{X}^{(\modelCounter)}_{\seq{0}{N}}$ denoting the unobservable Markov process of model $\modelCounter$ at times $\seq{0}{N} = \{0, 1, \ldots, N\}$, and $\bm{Y}^{(\modelCounter)}_{\seq{1}{N}}$ denoting the observable process of the system at times $\seq{1}{N}$.
-Under this framework, the observed data $y_{\seq{1}{n}}^*$ are assumed to be a single realization of the model $y_{\seq{1}{n}}^* \sim f_{\bm{X}^{(\modelCounter)}_{\seq{0}{N}}, \bm{Y}^{(\modelCounter)}_{\seq{1}{N}}}(x_{\seq{0}{N}}, y_{\seq{1}{N}}; \paramVec)$, where $\paramVec$ is a parameter vector that indexes the model.
-Using a probabilistic model results in several advantages, including the ability to account for variability present in the system.
-% which is of great interest to scientists and policy makers.
+With such a model, the existence of a joint density $f_{\bm{X}^{(\modelCounter)}_{\seq{0}{N}}, \bm{Y}^{(\modelCounter)}_{\seq{1}{N}}}$ is supposed, with $\bm{X}^{(\modelCounter)}_{\seq{0}{N}}$ denoting the unobservable Markov process of model $\modelCounter$ at times $\seq{0}{N} = \{0, 1, \ldots, N\}$, and $\bm{Y}^{(\modelCounter)}_{\seq{1}{N}}$ denoting the observable process of the system at times $\seq{1}{N}$.
+Under this framework, the observed data $y_{\seq{1}{N}}^*$---along with some unobserved value $\bm{X}^{(\modelCounter)}_{\seq{0}{N}} = x^*_{\seq{0}{N}}$---are assumed to be a single realization of the model $\big(x^*_{\seq{0}{N}}, y_{\seq{1}{N}}^*\big) \sim f_{\bm{X}^{(\modelCounter)}_{\seq{0}{N}}, \bm{Y}^{(\modelCounter)}_{\seq{1}{N}}}(x_{\seq{0}{N}}, y_{\seq{1}{N}}; \paramVec)$, where $\paramVec$ is a parameter vector that indexes the model.
+This modeling framework results in several advantages, including the ability to account for variability present in the system.
 Furthermore, because each draw from the joint distribution represents a potential outcome of the dynamic system, best/worst case scenarios under the assumptions of the model can be easily obtained via simulation.
-The added benefits of a stochastic model are coupled with statistical difficulty of fitting the parameter vector $\paramVec$ to observed data.
-
-There exist several algorithms, both frequentist and Bayesian, that can be used to obtain estimates of the parameters in stochastic dynamic models.
-% including the EM algorithm, Kalman Filter (and extensions) \citep{evensen09}, and iterated filtering algorithms \citep{ionides15}.
+% The added benefits of a stochastic model are coupled with statistical difficulty of fitting the parameter vector $\paramVec$ to observed data.
+
+Several algorithms exist, both frequentist and Bayesian, that can be used to obtain estimates of the parameters for this class of models.
 In order to retain the ability to propose models that are scientifically meaningful rather than only those that are simply statistically convenient, we restrict ourselves to parameter estimation techniques that have the plug-and-play property, which is that the fitting procedure only requires the ability to simulate the latent process instead of evaluating transition densities \citep{breto09,he10}.
@@
 The proposal of a linear trend in transmission is a result of observing an apparent decrease in reported cholera infections from 2012-2019 in Fig.~\ref{CholeraData}.
-% The inclusion of a possible trend in transmission is due to observing an apparent continual decrease in reported cholera infections from 2012-2019 can be seen in Fig.~\ref{CholeraData}.
 While several factors may contribute to this decrease, one explanation is that case-area targeted interventions (CATIs), which included education sessions, increased monitoring, household decontamination, soap distribution, and water chlorination in infected areas \citep{rebaudet19CATI}, may have greatly reduced cholera transmission \citep{rebaudet21}.
@@
 
-Model~1 is implemented using the \code{pomp} package \cite{pomp2}, which contains an implementation of the IF2 algorithm that was used to compute the likelihood profile.
-% Performing a profile confidence search of a parameter also results in a MLE calculation, as the value that maximizes the profile likelihood also maximizes the entire likelihood surface \citep{murphy2000}.
-Simulations from the fitted model compared to the observed data are given in Fig~\ref{fig:mod1fit}.
+Model~1 is implemented using the \code{pomp} package \cite{pomp2}, relying heavily on the source code provided by \cite{lee20}, who also used the MIF2 algorithm to estimate $\paramVec$.
+Aside from the introduction of the parameter $\transmissionTrend$ and the multiplicative Gamma-distributed white noise $d\Gamma(t)/dt$ in the process model (Eq.~\ref{model1:lambda}), the only change made that could increase the likelihood of the model was using a substantially increased computational effort to maximize the likelihood.
+While flexible, plug-and-play methods like MIF2 require a large number of computations, and the theoretic ability to maximize the likelihood depends on asymptotics in both the number of particles used and the number of MIF iterations.
+The large difference in likelihoods that were obtain by these two approaches (see Table~\ref{tab:likes}) highlights the importance of carefully determining the necessary computational effort needed to maximize model likelihoods and acting accordingly.
+
+% Simulations from the fitted model compared to the observed data are given in Fig~\ref{fig:mod1fit}.
 
@@
 
-Model~2 is a deterministic model, which can be viewed as a special case of a POMP (or SpatPOMP) with no randomness in the dynamic process or measurement.
+Model~2 is a deterministic model, which are degenerate probabilistic models that contain no randomness in the dynamic process or measurement.
 By combining a deterministic process model with a simple Gaussian measurement model, Model~2 reduces model fitting to a least squares calculation over parameters in a set of differential equations.
@@
 The increase in model flexibility, however, results in hidden states that are inconsistent between model phases.
-A naive solution to this would be to estimate only model parameters in separate phases and not the hidden states, but this may also result in hidden states in one phase of the model that are inconsistent with the other phase of the outbreak, since only the data from one phase is used to fit model parameters.
-
+% A naive solution to this would be to estimate only model parameters in separate phases and not the hidden states, but this may also result in hidden states in one phase of the model that are inconsistent with the other phase of the outbreak, since only the data from one phase is used to fit model parameters.
 The inclusion of a model break-point in \cite{lee20} is perhaps due to a challenging feature of fitting a deterministic model via least squares: discrepancies between model trajectories and observed case counts in highly infectious periods of a disease outbreak will result in greater penalty than the discrepancies between model trajectories and observed case counts in times of relatively low infectiousness, resulting in a bias towards accurately describing periods of high infectiousness.
-This issue is particularly troublesome for this case study: the inability to accurately fit times of low infectiousness may result in poor model forecasts, as few cases of cholera were observed the last few years epidemic.
+This issue is particularly troublesome for this case study: the inability to accurately fit times of low infectiousness may result in poor model forecasts, as few cases of cholera were observed the last few years of the epidemic.
 
@@
 \subsubsection{Model~3}
-Model~3 is also of a probabilistic model.
-In this model, both the latent and observable processes can be factored into department specific processes which interact with each other.
-% We denote the latent and measurement processes for Model~3 as $\bm{X}^{(3)}(t_{0:N}) = X^{(3)}_{1:U, 0:N}$, and $\bm{Y}^{(3)}(t_{1:N}) = Y_{1:U, 1:N}$ \eic{IS THIS NEEDED HERE?}.
-The decision to model the system via metapopulation models versus a model aggregated to a larger spatial scale is one of great scientific interest.
+Both the latent and observable processes in Model~3 can be be factored into department specific processes which interact with each other.
+The decision to model a dynamic system via metapopulation models versus a model aggregated to a larger spatial scale is one of great scientific interest.
 Evidence for the former approach has been provided in previous studies \citep{king15}, including the specific case of heterogeneity between Haitian departments in cholera transmission \citep{collins14}.
@@
 
-Parameters that must be fit in Model~3 are primarily shared between each department, the exception to this being the parameters $\beta_{W_u}$, and $\beta_u$, which are unique for each department $u \in \seq{1}{10}$.
-To avoid the parameter estimation issue in high-dimensional models, \citet{lee20} simplified the problem by fitting independent department-level models to the data.
-The shared parameters were calibrated using the cholera incidence data from Artibonite, and the department-specific parameters ($\beta_{W_u}$ and $\beta_u$) were fit using the data from their respective department.
+Parameters that must be fit in Model~3 are shared between each department, the exception to this being the parameters $\beta_{W_u}$, and $\beta_u$, which are unique for each department $u \in \seq{1}{10}$.
+To avoid the parameter estimation issue in high-dimensional models, \citet{lee20} simplified the problem by using the MIF2 algorithm to separately estimate parameters for independent department-level models;
+the shared parameters were calibrated using the cholera incidence data from Artibonite, and the department-specific parameters ($\beta_{W_u}$ and $\beta_u$) were fit using the data from their respective department.
 Reducing a spatially coupled model to individual units in this fashion requires special treatment of any interactive mechanisms between spatial units, such as found in Eq.~\myeqref{eq:model3:foi}.
@@
 
-We fit the PanelPOMP version of Model~3 using a slight modification of the PIF algorithm, which we call the Block Panel Iterated Filter (BPIF).
+We propose and use a slight modification of the PIF algorithm, which we call the Block Panel Iterated Filter (BPIF), to estimate parameters for a PanelPOMP version of Model~3.
 Let $l_{\text{P}}(y_{1:N}^*; \paramVec)$ denote the log-likelihood function of the PanelPOMP version of Model~3.
@@
 The simplified PanelPOMP version of Model~3 relies on the observed cholera cases as a covariate, which are unavailable for use in forecasts.
-To address this, we use the MLE obtained for the PanelPOMP approximation of Model~3 as an estimate for the parameters in fully coupled version of the model, which was implemented using the \code{spatPomp} package, similar to what was done in \cite{lee20}.
+Similar to what was done in \cite{lee20}, we use the parameter estimates obtained with the independent departmental approximation of Model~3 as an estimate for the parameters in the fully coupled version of the model, which was implemented using the \code{spatPomp} package.
 Simulations from the fully coupled version of the model using the estimated parameter vector $\tilde{\paramVec}$, are displayed in Fig.~\ref{h3spatsims}.
@@
 A natural question that arises is if the difference between $\hat{\paramVec}$ and $\tilde{\paramVec}$ is scientifically and statistically meaningful.
-The answer to this question requires the ability to reliably estimate $\hat{\paramVec}$, which is model specific and beyond the scope of this article.
-
-In this case study, a brief analysis of the fitted parameters $\tilde{\paramVec}$ provides some interesting insight.
+The answer to this question requires the ability to reliably estimate $\hat{\paramVec}$, the answer to which is model specific and beyond the scope of this case study.
+
+A brief analysis of the fitted parameters $\tilde{\paramVec}$ provides some interesting insight.
 Consider the unexpectedly large exponent on rainfall: $r = \Sexpr{signif(p3s["r"],3)}$ (Eq.~\ref{eq:model3:water}).
@@
 These comparisons provide insight into how each model performs relative to the others.
-To calibrate relative measures of fit, it is useful to compare against a model that has well-understood statistical ability to fit data, and we call this model a {\it benchmarks}.
+To calibrate relative measures of fit, it is useful to compare against a model that has well-understood statistical ability to fit data, and we call this model a {\it benchmark}.
 Standard statistical models, interpreted as associative models without requiring any mechanistic interpretation of their parameters, provide suitable benchmarks.
@@
 Comparing these results across models---including benchmark models---can help researchers identify potential model deficiencies, or errors in the observed data.
-Additional tools for assessing the goodness-of-fit of a model include plotting the effective sample size of each observation, and comparing any statistic of the observed data to simulations from the model---which is sometimes referred to as diagnostic probes.
+Additional tools for assessing the goodness-of-fit of a model include plotting the effective sample size of each observation \citep{liu01}, and comparing any statistic of the observed data to simulations from the model \citep{wood10}---which is sometimes referred to as diagnostic probes \citep{king16}.
 
@@
 Forecasting models are built using available scientific understanding, but forecasting can also be a way of testing new scientific hypotheses in real time \citep{lewis22}.
-In order to provide trustworthy information, however, the reliability of the forecast should be undersood.
+In order to provide trustworthy information, however, the reliability of the forecast should be understood.
 In particular, researchers should account for various forms of uncertainty present in model forecasts, and  calibrate the proposed model to observed data.
@@
 The most recently available information about a dynamic system should be used to obtain reliable forecasts.
-Using state space model $m$, this means that in order to obtain a forecast up to time $N+s$, where $s\geq 0$ and $N$ is the index for the last available data, we should simulate from the conditional density $f_{\bm{X}^{(\modelCounter)}_{N:s}, \bm{Y}^{\modelCounter}_{N:s}}\big(x_{N:s}, y_{N:s}\mid \hat{\paramVec}, \bm{X}^{(\modelCounter)}_{N} = \bm{x}^{(\modelCounter)}_N\big)$.
-The latent state $\bm{X}^{(\modelCounter)}_N$ is by definition unobservable, however, and therefore draws from the desired conditional density are unobtainable.
+Using state space model $m$, this means that in order to obtain a forecast up to time $N+s$, where $s\geq 0$ and $N$ is the index for the last available data, we should simulate from the conditional density $f_{\bm{X}^{(\modelCounter)}_{N:s}, \bm{Y}^{\modelCounter}_{N:s}}\big(x_{N:s}, y_{N:s}\mid \hat{\paramVec}, \bm{X}^{(\modelCounter)}_{N} = \bm{x}^{(\modelCounter)}_N\big)$, where $\bm{x}^{(\modelCounter)}_N$ is the value of the latent state at time $N$.
+Because the latent state $\bm{X}^{(\modelCounter)}_N$ is unobservable, draws from the desired conditional density are unobtainable.
 Informed estimates of $\bm{X}^{(\modelCounter)}_N$ given the observed data $\bm{Y}^{(\modelCounter)}_{1:N} = y_{1:N}^*$ can easily be obtained using the filtering distribution.
@@
 Obtaining forecasts in this fashion allows us to obtain results that are consistent with the model assumptions and the observed data.
-In this particular case study, projecting the future state of the cholera epidemic in Haiti starting from the draws from the filtering distribution allows the proposed models to benefit from the fact that very few cholera cases had been observed between 2018 and 2019, and that cases appear to be decreasing (i.e., the number of susceptible individuals may be small).
+In this particular case study, projecting the future state of the cholera epidemic starting from latent states that are draws from the filtering distribution allows the proposed models to benefit from the fact that very few cholera cases had been observed between 2018 and 2019, and that cases appear to be decreasing.
 This alone partially explains the unrealistic forecasts of \cite{lee20}, which were obtained by simulating the model from initial conditions;
@@
 parameter uncertainty should therefore be considered when obtaining model forecasts to influence policy.
-If a Bayesian technique is used for parameter estimation, the most natural way to account for parameter uncertainty is to obtain multiple simulations from the model where each simulation is obtained using parameters $\paramVec$ drawn from the posterior distribution $\paramVec \sim f_{\Theta}\big(\paramVec | \bm{Y}^{(\modelCounter)}_{1:N} = y_{1:N}^*\big)$.
-In the frequentist case, one approach is a quasi-empirical Bayes: endow to set $\Theta_K = \{\theta_1, \theta_2, \ldots, \theta_K\}$ with a discrete uniform distribution, such that $P\big(\paramVec = \theta_k\big) = \frac{1}{K}$ for all values $k \in \seq{1}{K}$.
-In this case, the posterior distribution of $\paramVec | \bm{Y}^{(\modelCounter)}_{1:N}$ can be expressed as: 
-$P\big(\paramVec = \theta_k | \bm{Y}^{(\modelCounter)}_{1:N} = y_{1:N}^*\big) = \frac{l(y_{1:N}^*; \theta_k)}{\sum_{j = 1}^K l(y_{1:N}^*; \theta_j)}$; to complete the formulation, we then choose the set $\Theta_K$ such that each $\theta_k \in \Theta_K$ is a single output of the MIF2 algorithm, which has been shown to be within a neighborhood of the MLE \cite{ionides15}.
-This approach has been taken in previous studies \citep{king15}, and the result is simply weighing parameter sets based on their corresponding likelihoods: parameters with higher likelihoods should be used more frequently in model forecasts.
+If a Bayesian technique is used for parameter estimation, the most natural way to account for parameter uncertainty is to obtain $J$ simulations from the model where each simulation is obtained using parameters $\Theta_{1:J}$ drawn from the posterior distribution $\Theta_{1:J} \sim f_{\Theta}\big(\paramVec | \bm{Y}^{(\modelCounter)}_{1:N} = y_{1:N}^*\big)$.
+In the frequentist case, one approach is empirical Bayes: endow to set $\Theta_K = \{\theta_1, \theta_2, \ldots, \theta_K\}$ with a discrete uniform distribution, such that $P\big(\Theta_{j} = \theta_k\big) = \frac{1}{K}$ for all values $k \in \seq{1}{K}$.
+In this case, the posterior distribution of $\Theta_{j} | \bm{Y}^{(\modelCounter)}_{1:N}$ can be expressed as:
+$P\big(\Theta_{j} = \theta_k | \bm{Y}^{(\modelCounter)}_{1:N} = y_{1:N}^*\big) = \frac{l(y_{1:N}^*; \theta_k)}{\sum_{l = 1}^K l(y_{1:N}^*; \theta_l)}$.
+To complete the formulation, we then choose the set $\Theta_K$ such that each $\theta_k \in \Theta_K$ is a single output of the MIF2 algorithm, which has been shown to be within a neighborhood of the MLE \cite{ionides15}.
+This approach has been taken in previous studies \citep{king15}, and the result is simply weighing parameter sets based on their corresponding likelihoods: parameters with higher likelihoods should be used more frequently in model forecasts than those that correspond to lower likelihoods.
 This can be done for both deterministic and stochastic models, but requires a large number of computations, especially as the number of observations and model parameters increase.
@@
 
-Forecasting can be used to simulate the simulate various interventions on a system; this feature is useful to inform policy and was the primary goal of \citet{lee20}.
+Forecasting can be used to simulate various interventions on a system; this feature is useful to inform policy and was the primary goal of \citet{lee20}.
 Outcomes of their study include estimates for the probability of cholera elimination and cumulative number of cholera infections under several possible vaccination scenarios.
-Mimicking their efforts, we define cholera elimination as having less than one infection of cholera over at least 52 consecutive weeks in the 10-year projection period, and provide forecasts under the following vaccination scenarios:
+Mimicking their efforts, we define cholera elimination as having less than one infection of cholera over at least 52 consecutive weeks, and provide forecasts under the following vaccination scenarios:
 
@@
 Simulations from probabilistic models (Models~1 and~3) represent possible trajectories of the dynamic system under the scientific assumptions of the models.
-In this case study, estimates of the probability of cholera elimination can therefore be obtained as the proportion of simulations from the fitted model that result in cholera elimination.
-The results of these projections are summarized in Figs.~\ref{fig:Mod1Scenarios}--\ref{fig:elimProbs}.
-These results suggest that cholera elimination was likely, even without increased vaccination efforts, which is consistent with observed reality \citep{trevisin22}.
+Estimates of the probability of cholera elimination can therefore be obtained as the proportion of simulations from these models that result in cholera elimination.
+The results of these projections are summarized in Figs.~\ref{fig:Mod1Scenarios}--\ref{fig:elimProbs}, and suggest that cholera elimination was likely even without increased vaccination efforts---consistent with observed reality \citep{trevisin22}.
 
@@
 The resulting mechanisms in a fitted model can be compared to current scientific knowledge about a system.
-Agreement between model based inference and our current understanding of a system may taken as a confirmation of both model based conclusions and our understanding of the system.
+Agreement between model based inference and our current understanding of a system may taken as a confirmation of both model based conclusions and our scientific understanding.
 On the other hand, discrepancies found in these comparisons may have the potential to spark new scientific knowledge \citep{ganusov16}.
@@
 the cubic splines permit flexible estimation of seasonality in the force of infection.
-After fitting the model parameters, we explore potential patterns in the seasonal transmission rate by plotting the average value of $\transmission$ in a typical year.
+After calibrating model parameters to observed data, we explore potential patterns in the seasonal transmission rate by plotting the average value of $\transmission$ in a typical year.
 Fig.~\ref{fig:h1SeasRain} shows that the estimated seasonal transmission rate $\transmission$ mimics the rainfall dynamics in Haiti, despite Model~1 not having access to rainfall data.
-This result provides evidence that rainfall played an important role in cholera transmission in Haiti, consistent with previous studies \citep{lemaitre19,eisenberg13};
-the estimated seasonality also features an increased transmission rate during the Fall, which was noticed at an earlier stage of the epidemic \citep{rinaldo12}.
+This result provides evidence that rainfall played an important role in cholera transmission in Haiti, consistent with previous studies \citep{lemaitre19,eisenberg13}.
+The estimated seasonality also features an increased transmission rate during the Fall, which was noticed at an earlier stage of the epidemic \citep{rinaldo12}.
 
@@
 Our case study provides an example of this in the parameter estimate for the duration of natural immunity due to cholera infection $\muRS^{-1}$.
-Under the framework of Model~2, the best estimate for this parameter is $\Sexpr{signif(1/h2_fit$h2_params["sigma"],2)}$, suggesting that individuals have effectively permanent immunity to cholera once infected, and thus the pool of susceptible individuals decreases over time.
+Under the framework of Model~2, the best estimate for this parameter is $\Sexpr{signif(1/h2_fit$h2_params["sigma"],3)}$, suggesting that individuals have effectively permanent immunity to cholera once infected, and thus the pool of susceptible individuals decreases over time.
 The interpretation of this result is muddled due to the fact that the data ranged from 2010-2019, and therefore estimates of immunity longer than 10 years---the upper end of previous estimates of natural immunity \citep{king08}---effectively result in the same model dynamics.
@@
 
-\section{Robustness of model based conclusions}\label{sec:advice}
+\section{Robustness of model based conclusions}\label{sec:advice}\jwc{BETTER NAME FOR THIS?}
 
@@
 For example, reconstructed latent variables (such as numbers of susceptible and recovered individuals) should make sense in the context of alternative measurements of these variables;
-parameter values which fit the data should make sense in the context of alternative lines of evidence about the phenomena being modeled.
+parameters that have been calibrated to data should make sense in the context of alternative lines of evidence about the phenomena being modeled.
 % \jwc{Perhaps add comment on our final parameter estimates, and remove the following sentences in this paragraph.}
@@
 
-If a mechanistic model including a feature (such as a representation of a mechanism, or the inclusion of a covariate) fits better than mechanistic models without that feature, and also has competitive fit compared to associative models, this may be taken as evidence supporting the scientific relevance of the feature.
+If a mechanistic model including a feature (such as a representation of a mechanism, or the inclusion of a covariate) fits better than mechanistic models without that feature, and also has competitive fit compared to associative benchmarks, this may be taken as evidence supporting the scientific relevance of the feature.
 As for any analysis of observational data, we must be alert to the possibility of confounding.
@@
 \sdescription{
-This \code{R} package is contained in a public GitHub repository: zjiang2/haitipkg.
+This \code{R} package is contained in a public GitHub repository: jeswheel/haitipkg.
 The package contains all of the data and code used to create and fit the models, as well as other useful functions that were used in this article.

### 2022-09-22 ms.Rnw
Compiled results from RL3

@@
 A brief analysis of the fitted parameters $\tilde{\paramVec}$ provides some interesting insight.
-Consider the unexpectedly large exponent on rainfall: $r = \Sexpr{signif(p3s["r"],3)}$ (Eq.~\ref{eq:model3:water}).
+Consider the large estimated exponent on rainfall: $r = \Sexpr{signif(p3s["r"],3)}$ (Eq.~\ref{eq:model3:water}).
 Because the value of rainfall $J(t) \in (0, 1)$, the large coefficient results in $J(t)^r \approx 0$.
@@
 One plausible explanation of this result is that the fitting procedure---used to obtain parameter values in the independent department model given the reported cases in other departments---has learned an over-dependence on the reported cholera cases in other departments, and thus the seasonality of cholera incidence in department $u$ depends on the highly correlated covariate of the reported cholera cases in departments $1:U\setminus \{u\}$.
-Whatever the case may be, this fitting procedure results in model likelihoods unable to improve upon a log-linear benchmark (see Sec.~\ref{sec:model_diagnostics} and Tab.~\ref{tab:likes}), and highlights the need for advancements in statistical methodology that permit inference on models with coupled metapopulation dynamics.
+Whatever the case may be, this fitting procedure results in a model that is unable to replicate the observed number of cases by simulating from initial conditions (see Fig.~\ref{h3spatsims}), and highlights the need for advancements in statistical methodology that permit inference on models with coupled metapopulation dynamics.
 
@@
 Simulations of Model~1 under each vaccination scenario.
-Blue line indicates the median of model simulations, and ribbon represents $95\%$ of simulations.
-The various vaccination campaigns made no practical difference in the median scenario, but a drastic difference in the extreme cases.
+Blue line indicates the simulated median of reported cases, and the ribbon represents $95\%$ of simulations.
+% The various vaccination campaigns made no practical difference in the median scenario, but a drastic difference in the extreme cases.
 }
@@
 Simulations of Model~3 under each vaccination scenario.
-Blue line indicates the median of model simulations, and ribbon represents $95\%$ of simulations.
-The various vaccination campaigns made no practical difference in the median scenario, but a drastic difference in the extreme cases.
+Blue line indicates the simulated median of reported cases, and ribbon represents $95\%$ of simulations.
+% The various vaccination campaigns made no practical difference in the median scenario, but a drastic difference in the extreme cases.
 }

### 2022-09-23 ms.Rnw
a few more edits

@@
 When combined with online availability of data and code, objective measures of fit provide a powerful tool to accelerate scientific progress, following the paradigm of the {\it common task framework} \citep[][Sec.~6]{donoho17}.
-In our literature review of the Haiti cholera epidemic \eic{CHECK THIS} no quantitative measures of goodness of fit, and no benchmark models, were considered in any of the 24 papers which calibrated a mechanistic model to data in order to obtain scientific conclusions.
+In our literature review of the Haiti cholera epidemic, no benchmark models were considered in any of the 24 papers which calibrated a mechanistic model to data in order to obtain scientific conclusions.
 
@@
 In such a case, a benchmark model could be fit to each spatial resolution being considered, and each model compared to their respective benchmark.
-Large advantages (shortcomings) in model likelihood relative to the benchmark for a given spatial scale that are not present in other spatial scales may provide weak evidence for (against) the spatial resolution under consideration.
-While these comparisons don't provide a formal statistical test of the amount of spatial resolution needed for a given model, they can be quite informative.
+Large advantages (shortcomings) in model likelihood relative to the benchmark for a given spatial scale that are not present in other spatial scales may provide weak evidence for (against) the statistical fit of models across a range of spatial resolutions.
 
@@
 Comparing these results across models---including benchmark models---can help researchers identify potential model deficiencies, or errors in the observed data.
-Additional tools for assessing the goodness-of-fit of a model include plotting the effective sample size of each observation \citep{liu01}, and comparing any statistic of the observed data to simulations from the model \citep{wood10}---which is sometimes referred to as diagnostic probes \citep{king16}.
+Additional tools for assessing the goodness-of-fit of a model include plotting the effective sample size of each observation \citep{liu01}, and comparing summary statistics of the observed data to simulations from the model \citep{wood10}.
+These summary statistics are sometimes called diagnostic probes \citep{king16}.
 
@@
 Table~S7 of their supplement material shows that of their simulations that were consistent with observing zero cases in 2019, all forecasts resulted in the elimination of cholera.
-% \eic{THE QUESTION OF WHETHER TO USE A DETERMINISTIC MODEL REMAINS OF CURRENT INTEREST...}
 
@@
 
-While dynamic models may potentially result in meaningful scientific insights, one must be careful not to assume the results of a single analysis as a de-facto feature of the system.
-It is important instead to recognize and assess the modeling simplifications and assumptions that were used in order to arrive at the conclusions.
-In epidemiological studies, for example, it is unreasonable to expect that our quantitative understanding of individual-level processes will perfectly match model parameters that were fit to population-level case counts \citep{he10} \jwc{ED: ARE THERE BETTER REFERENCES FOR THIS?},
-making direct interpretation estimated parameters difficult.
+For any model-based inference, it is important to recognize and assess the modeling simplifications and assumptions that were used in order to arrive at the conclusions.
+In epidemiological studies, for example, quantitative understanding of individual-level processes may not perfectly match model parameters that were fit to population-level case counts, even when the model provides a strong statistical fit \citep{he10}.
+This makes direct interpretation of estimated parameters delicate.
 
@@
 Under the framework of Model~2, the best estimate for this parameter is $\Sexpr{signif(1/h2_fit$h2_params["sigma"],3)}$, suggesting that individuals have effectively permanent immunity to cholera once infected, and thus the pool of susceptible individuals decreases over time.
-The interpretation of this result is muddled due to the fact that the data ranged from 2010-2019, and therefore estimates of immunity longer than 10 years---the upper end of previous estimates of natural immunity \citep{king08}---effectively result in the same model dynamics.
+To interpret this result, we bear in mind that the data ranged from 2010-2019, and therefore estimates of immunity longer than 10 years---the upper end of previous estimates of natural immunity \citep{king08}---effectively result in the same model dynamics.
 The depletion of susceptible individuals may also be attributed to confounding mechanisms that are unaccounted for in the model---such as vaccination and other non-pharmaceutical interventions that reduce cholera transmission \citep{trevisin22, rebaudet21}---that were not accounted for in the model.
 Perhaps the best interpretation of the estimated parameter, then, is that under model assumptions, the model most adequately describes the observed data by having a steady decrease in the number of susceptible individuals.
-The poor model fit compared to a log-linear benchmark (see Table~\ref{tab:likes}), and the large inconsistenty between model parameters and our understanding of the system, may be taken as evidence of model mispecificaiton.
-
-\section{Robustness of model based conclusions}\label{sec:advice}\jwc{BETTER NAME FOR THIS?}
-
-% \eic{I THINK IT IS INTERESTING THAT THE MODEL-SUPPORTED SUGGESTION BY \citep{andrews11} HAS LARGELY PROVED ACCURATE (VACCINATION, NON-PHARMACEUTICAL INTERVENTIONS, AND INFRASTRUCTURE IMPROVEMENTS ALL HAVE A ROLE AND SHOULD BE FOLLOWED SIMULTANEOUSLY)DESPITE LEGITIMATE CONCERNS ABOUT WHETHER THE SCIENTIFIC SUPPORT FOR THE MODEL IS STRONG ENOUGH TO ROBUSTLY GUIDE POLICY \citep{grad12}.
-% THE MAIN CLAIM OF ANDREWS11 IS THAT FORECASTS AND POLICY NOT SUPPORTED BY A MODEL SEEM EVEN MORE UN-ROOTED THAN RECOMMENDATIONS BASED ON A WEAKLY SUPPORTED MODEL.
-% THE CONCLUSION FROM ALL THIS, WHICH I THINK IS SUPPORTED BY OUR RE-ANALYSIS OF LEE ET AL, IS THAT MECHANISTIC MODELS FITTED TO DATA ARE A STRONG WAY TO ENFORCE CONSISTENT THINKING, BUT WEAKER AS A WAY TO CHECK THAT YOU ARE ACTUALLY RIGHT DUE TO THE LIMITATIONS OF INFERRING CAUSAL MECHANISMS FROM OBSERVATIONAL DATA.
-% POLICY-MAKERS SHOULD FOLLOW AN APPROACH WHICH IS SELF-CONSISTENT AND CONSISTENT WITH PREVIOUS EVIDENCE THAT MIGHT WORK (WHICH MODELS CAN HELP WITH) BUT THEY MUST BE AGILE AS NEW EVIDENCE EMERGES.
-% MODELS, TOGETHER WITH DATA ANALYSIS, CAN ALSO HELP REVEAL WHEN THE EVIDENCE SUPPORTING A P0LICY IS BREAKING DOWN, FORCING A RE-THINK.
-% }
+The weak statistical fit of Model~2 compared to a log-linear benchmark (see Table~\ref{tab:likes}) cautions us aginst drawing quantitative conclusions from this model.
+A model that has a poor statistical fit may nevertheless provide a useful conceptual framework for thinking about the system under investigation.
+However, a claim that the model has been validated against data should be reserved for situations where the model provides a statistical fit that is competitive against alternative explanations.
+
+\section{Robust interpretation of model based conclusions}\label{sec:advice}
 
@@
 A causal interpretation is strengthened by corroborative evidence.
-For example, reconstructed latent variables (such as numbers of susceptible and recovered individuals) should make sense in the context of alternative measurements of these variables;
-parameters that have been calibrated to data should make sense in the context of alternative lines of evidence about the phenomena being modeled.
-% \jwc{Perhaps add comment on our final parameter estimates, and remove the following sentences in this paragraph.}
-% A model is necessarily a simplified approximation of a complex process, and so it is unreasonable to expect models (describing and fitted to population-level processes) to perfectly match quantitative understanding of individual-level processes.
-% Such discrepancies could lead to biases when interventions (modeled as effects on individuals) are included in the population-level, unless there are historical occurrences of the intervention which enable the consequence to be calibrated at the population level.
+For example, reconstructed latent variables (such as numbers of susceptible and recovered individuals) should make sense in the context of alternative measurements of these variables \citep{grad12}.
+In addition, parameters that have been calibrated to data should make sense in the context of alternative lines of evidence about the phenomena being modeled, while making allowance for the possibility that the interpretations of parameters may vary when modeling across differing spatial scales.
 
@@
 For a mechanism, the model feature could in principle explain the data by helping to account for some different unmodeled phenomenon.
-% Assessing potential confounding is part of building an argument for a causal interpretation of a fitted model.
 In the context of our analysis, the estimated trend in transmission rate could be explained by any trending variable (such as hygiene improvements, or changes in population behavior), resulting in confounding from colinear covariates.
@@
 \citet{lee20} published their code and data online, and this reproducibility facilitated our work.
-By design, the models were coded and analyzed independently, leading to differing implementation decisions.
 Robust data analysis requires not only reproducibility but also extendability: if one wishes to try new model variations, or new approaches to fitting the existing models, or plotting the results in a different way, this should be not excessively burdensome.
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%\begin{acks}[Acknowledgments]
-% The authors would like to thank ...
-%\end{acks}
+\begin{acks}[Acknowledgments]
+The authors would like to thank Mercedes Pascual and Betz Halloran for helpful discussions. Laura Matrajt provided additional data for the Model~2 analysis. 
+\end{acks}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2022-09-23 ms.Rnw
a few typos

@@
 
-  # Determine whice mode should the value of rho belong to, based on regression
+  # Determine which mode should the value of rho belong to, based on regression
   case_when(
@@
 
-        # Set the inital state
+        # Set the initial state
         new_rinit <- Csnippet(
@@
 To successfully predict the effect of an intervention, a model should therefore both provide a quantitative explanation of existing data and should have a causal interpretation: a manipulation of the system should correspond quantitatively with the corresponding change to the model.
-This motivates the development of mechanistic statistical models, which provides a statistical fit to the available data while also supporting a causal intepretation.
+This motivates the development of mechanistic statistical models, which provides a statistical fit to the available data while also supporting a causal interpretation.
 
@@
 where $\mathrm{Pop}_u$ is the mean population for department $u$,
-$D_{u\varv}$ is a distance measure estimating average road distance between rancomly chosen members of each population, and $v_{\mathrm{rate}}= 10^{-12}$ was treated as a fixed constant.
+$D_{u\varv}$ is a distance measure estimating average road distance between randomly chosen members of each population, and $v_{\mathrm{rate}}= 10^{-12}$ was treated as a fixed constant.
 In \eqref{model2:water_transport}, $T^W_{u\varv}$ is a measure of river flow between departments.
@@
 % Simulations from initial conditions using panelPOMP version of Model~3.
-% The black curve represents true case count, the blue curve the median of 500 simulations from the model, and the grey ribbons representing $95\%$ confidence interval.
+% The black curve represents true case count, the blue curve the median of 500 simulations from the model, and the gray ribbons representing $95\%$ confidence interval.
 % }
@@
 Simulations from initial conditions using the spatially coupled version of Model~3.
-The black curve represents true case count, the blue line the median of 500 simulations from the model, and the grey ribbons representing $95\%$ confidence interval.
+The black curve represents true case count, the blue line the median of 500 simulations from the model, and the gray ribbons representing $95\%$ confidence interval.
 }
@@
 Perhaps the best interpretation of the estimated parameter, then, is that under model assumptions, the model most adequately describes the observed data by having a steady decrease in the number of susceptible individuals.
-The weak statistical fit of Model~2 compared to a log-linear benchmark (see Table~\ref{tab:likes}) cautions us aginst drawing quantitative conclusions from this model.
+The weak statistical fit of Model~2 compared to a log-linear benchmark (see Table~\ref{tab:likes}) cautions us against drawing quantitative conclusions from this model.
 A model that has a poor statistical fit may nevertheless provide a useful conceptual framework for thinking about the system under investigation.
@@
 For a mechanism, the model feature could in principle explain the data by helping to account for some different unmodeled phenomenon.
-In the context of our analysis, the estimated trend in transmission rate could be explained by any trending variable (such as hygiene improvements, or changes in population behavior), resulting in confounding from colinear covariates.
+In the context of our analysis, the estimated trend in transmission rate could be explained by any trending variable (such as hygiene improvements, or changes in population behavior), resulting in confounding from collinear covariates.
 Alternatively, the trend could be attributed to a decreasing reporting rate rather than decreasing transmission rate, resulting in confounded mechanisms.

### 2022-09-26 ms.Rnw
Started Clean-up of saved files

@@
 
+plot_order <- c(
+  'Artibonite',
+  'Sud_Est',
+  'Nippes',
+  'Nord_Est',
+  'Ouest',
+  'Centre',
+  'Nord',
+  'Sud',
+  'Nord_Ouest',
+  'Grande_Anse'
+)
+
 options(
@@
     {
+      lee1_epi_mif_results <- list()
+
       foreach(
@@
         )
-      }
+      } -> lee1_epi_mif_out
+
+      foreach(i = 1:length(lee1_epi_mif_out), .combine = rbind) %do% {
+        guess <- coef(lee1_epi_mif_out[[i]])
+        sims <- simulate(
+          h1_epi,
+          n = N_LEE_SIMS,
+          format = 'data.frame',
+          params = guess
+        )  %>%
+          dplyr::filter(week == max(week)) %>%
+          dplyr::select(-.id, -week)
+
+          sims$param_set <- i
+          sims
+      } -> start_states
+
+      lee1_epi_mif_results$lee1_epi_mif_params <- t(coef(lee1_epi_mif_out))
+      lee1_epi_mif_results$start_states <- start_states
+      lee1_epi_mif_results
     }
@@
 
-      for (j in 1:nrow(lee_epi_ll)) {
+      for (j in 1:nrow(lee1_epi_mif$lee1_epi_mif_params)) {
         # Get parameters of interest
-        pf_params <- coef(lee1_epi_mif[[j]])
+        pf_params <- lee1_epi_mif$lee1_epi_mif_params[j, ]
 
@@
 }
-
-# First, we need to simulate the epidemic model using the fitted parameters.
-registerDoRNG(18599687)
-foreach(
-  i = 1:length(lee1_epi_mif),
-  .combine = rbind
-) %dopar% {
-  guess <- coef(lee1_epi_mif)
-  sims <- simulate(
-    h1_epi,
-    nsim = N_LEE_SIMS,
-    format = 'data.frame',
-    params = guess
-  )  %>%
-    dplyr::filter(week == max(week)) %>%
-    dplyr::select(-.id, -week)
-
-  sims$param_set <- i
-  sims
-} -> start_states
 
@@
 
-end_start <- t(coef(lee1_epi_mif))
+end_start <- lee1_epi_mif$lee1_epi_mif_params
 
@@
         .packages = c('pomp'),
-        .combine = c
+        .combine = rbind
       ) %dopar% {
@@
 
-        start_states %>%
+        lee1_epi_mif$start_states %>%
           filter(param_set == i) %>%
@@
         coef(h1_end) <- r_params
-        mif2(
+        mf_end <- mif2(
           h1_end,
@@
         )
-        # output[['m2']] <- m2
-        #
-        # coef(h1_end) <- coef(m2)
-        # sims <- simulate(  # Step 5
-        #   h1_end, nsim = N_LEE_SIMS, format = 'data.frame',
-        #   params = coef(m2)
-        # )
-        # sims$param_set <- i
-        #
-        # output[['sims']] <- sims
-        # output
-      }
-    }
-  )
-}
-
-rm(end_rw, start_states)
+
+        coef(mf_end)
+      }
+    }
+  )
+}
+
+rm(end_rw)
 gc()
@@
       lee_end_ll <- data.frame(
-        "pfLL" = rep(0, length(lee1_end_mif)),
-        "pfse" = rep(0, length(lee1_end_mif))
+        "pfLL" = rep(0, nrow(lee1_end_mif)),
+        "pfse" = rep(0, nrow(lee1_end_mif))
       )
@@
         # Get parameters of interest
-        pf_params <- coef(lee1_end_mif[[j]])
+        pf_params <- lee1_end_mif[j, ]
 
@@
 
+if (RUN_LEVEL != 3 & file.exists("model3/run_level_3/haiti3_fit.rds")) {
+  haiti3_fit <- readRDS("model3/run_level_3/haiti3_fit.rds")
+} else {
   haiti3_fit <- bake(
@@
 <<Load Model Parameters, echo=FALSE, results='hide'>>=
-# load('model3/output/PanelLocalAll_PF.rda')  # Final model 3 evaluations
-# load('model3/output/PanelLocalAll.rda')  # Final model 3 parameters (MIF search)
 
@@
 h2_fit <- bake(
-  file = paste0("model2/", rl_dir, "model2_fit.rds"), {
+  file = paste0("model2/model2_fit.rds"), {
     fit_haiti2()
@@
 
-<<Model 3 PanelSims,echo=FALSE>>=
-h3_panel_sims <- switch(RUN_LEVEL, 20, 100, 500)
-
-departements <-
-  c(
-    'Artibonite',
-    'Centre',
-    'Grande_Anse',
-    'Nippes',
-    'Nord',
-    'Nord-Est',
-    'Nord-Ouest',
-    'Ouest',
-    'Sud',
-    'Sud-Est'
-  )
-
-registerDoRNG(5317865)
-
-if (RUN_LEVEL != 3 & file.exists("model3/run_level_3/panelPompSims.rds")) {
-  h3_panel_sims <- readRDS("model3/run_level_3/panelPompSims.rds")
-} else {
-  h3_panel_sims <- bake(
-    file = paste0('model3/', rl_dir, 'panelPompSims.rds'), {
-      foreach(dep = departements, .combine = rbind) %dopar% {
-        SIRB <- unitobjects(h3)[[dep]]
-        shared <- h3@shared
-        specific <- h3@specific[, dep]
-
-        pomp::simulate(
-          SIRB,
-          params = c(shared, specific),
-          nsim = h3_panel_sims,
-          format = 'data.frame'
-        ) -> sims
-
-        sims$dep <- dep
-        sims
-      }
-    }
-  )
-}
-
-h3_panel_quantiles <- h3_panel_sims %>%
-  mutate(
-    date = as.Date(
-      lubridate::round_date(lubridate::date_decimal(time), unit = 'day')
-    )
-  ) %>%
-  group_by(dep, date) %>%
-  summarise(
-    q05 = quantile(cases, 0.025, na.rm = T),
-    mean = mean(cases, na.rm = T),
-    q50 = quantile(cases, 0.5, na.rm = T),
-    q95 = quantile(cases, 0.975, na.rm = T)
-  ) %>%
-  ungroup() %>%
-  mutate(
-    dep = gsub("-", "_", dep)
-  ) %>%
-  filter(date >= as.Date(
-    lubridate::round_date(
-      lubridate::date_decimal(h3@unit.objects$Artibonite@t0), unit = 'day'
-    )
-  )
-  )
-
-plot_order <- c(
-  'Artibonite',
-  'Sud_Est',
-  'Nippes',
-  'Nord_Est',
-  'Ouest',
-  'Centre',
-  'Nord',
-  'Sud',
-  'Nord_Ouest',
-  'Grande_Anse'
-)
-
-@
-
-% \begin{figure}[ht]
-% ggplot() +
-%   geom_line(data = h3_panel_quantiles, aes(x = date, y = q50 + 1), col = 'blue') +
-%   geom_line(data = dep_plot_df, aes(x = date, y = cases + 1), col = 'black') +
-%   geom_ribbon(data = h3_panel_quantiles, aes(x = date, ymin = q05 + 1, ymax = q95 + 1),
-%               alpha = 0.4) +
-%   facet_wrap(~factor(dep, levels = plot_order), nrow = 2,
-%              labeller = dep_labeller) +
-%   ylab('Reported cholera cases') +
-%   theme(axis.title.x = element_blank()) +
-%   scale_x_date(date_labels = "'%y", breaks = seq.Date(from = as.Date("2011-01-01"), as.Date("2019-01-01"), by = '2 years')) +
-%   scale_y_log10(labels = scales::trans_format("log10", scales::math_format(10^.x)))
-%
-% \caption{\label{h3panelsims}
-% Simulations from initial conditions using panelPOMP version of Model~3.
-% The black curve represents true case count, the blue curve the median of 500 simulations from the model, and the gray ribbons representing $95\%$ confidence interval.
-% }
-% \end{figure}
-
-<<Remove Model3 panelSims, cache=FALSE, include=FALSE, message=FALSE, echo=FALSE>>=
-rm(h3_panel_quantiles, h3_panel_sims)
-gc()
-@
-
 Fitting Model~3 in this fashion simplifies the problem of parameter estimation, but it also introduces additional technicalities that must be addressed.
 One concern is that of obtaining model forecasts, which was the primary goal of \citet{lee20}.
-% "primary output" -> "primary goal", based on Anna's suggestion.
 The simplified PanelPOMP version of Model~3 relies on the observed cholera cases as a covariate, which are unavailable for use in forecasts.
@@
 <<Model 3 SpatPOMP sims, echo=FALSE>>=
-# load('output/FinalPanelAsSpatPomp_PF.rda')
-
 h3_spat_nsim <- switch(RUN_LEVEL, 20, 100, 500)
@@
 mod2_epi_ll <- -epi_ofun(par = h2_fit$h2_params)
-# mod2_end_ll <- -end_ofun(par = h2_fit$end_params)
-
 mod2_ll <- mod2_epi_ll - sum(log(h2_epi@data + 1), na.rm = TRUE)
-
-# Evaluate and sum objective functions to get total model likelihood
-# mod2_ll <- -(epi_ofun(par = h2_fit$epi_params) + end_ofun(par = h2_fit$end_params))
 
@@
 mod1_aic <- 2 * mod1_n_params - 2 * mod1_ll
-
-# Load pre-calculated model 3 likelihood
-# mod3_ll <- ll_mod3 %>%
-#   arrange(-logLik) %>%
-#   slice_head(n = 1) %>%
-#   pull(logLik)
 
@@
 
-# load('model2/output/mod2_leeFit.rda')
 mod2_lee_aic <- 2 * h2_fit$leeFit_n_params - 2 * h2_fit$leeFit_ll
-# mod2_lee_aic <- 2 * mod2_leeFit_num_parms - 2 * mod2_leeFit_ll
 @

### 2022-09-27 ms.Rnw
Saving only results of haiti 1Vaccination scenario sims

@@
 
-# TODO: Only save parameter values and "likelihoods".
 if (RUN_LEVEL != 3 & file.exists("model1/run_level_3/lee1_end_fit.rds")) {
@@
 if (RUN_LEVEL != 3 & file.exists("model1/run_level_3/S0_sims.rds")) {
-  h1_S0_sims <- readRDS("model1/run_level_3/S0_sims.rds")
+  h1_S0_results <- readRDS("model1/run_level_3/S0_sims.rds")
 } else {
-  h1_S0_sims <- bake(
+  h1_S0_results <- bake(
     file = paste0("model1/", rl_dir, "S0_sims.rds"), {
@@
       # project vacc scenario, with filtering dist, and model.
-      project_from_filter2(mod = s0, PF = s0_pf, nsims = h1_scen_sims)
+      h1_S0_sims <- project_from_filter2(mod = s0, PF = s0_pf, nsims = h1_scen_sims)
+
+      results <- list()
+      results$mod1_V0_sims <- agg_mod1_sims(h1_S0_sims)
+      results$mod1_V0_probs <- get_elimProbs(h1_S0_sims, model = 1)
+
+      results
     },
@@
 }
-
-# Clean-up simulations for analysis/visualization
-mod1_V0_sims <- agg_mod1_sims(h1_S0_sims)
-mod1_V0_probs <- get_elimProbs(h1_S0_sims, model = 1)
-rm(h1_S0_sims)
 
@@
 if (RUN_LEVEL != 3 & file.exists("model1/run_level_3/S2_sims.rds")) {
-  h1_S2_sims <- readRDS("model1/run_level_3/S2_sims.rds")
+  h1_S2_results <- readRDS("model1/run_level_3/S2_sims.rds")
 } else {
-  h1_S2_sims <- bake(
+  h1_S2_results <- bake(
     file = paste0('model1/', rl_dir, 'S2_sims.rds'), {
@@
       # project vacc scenario, with filtering dist, and model.
-      project_from_filter2(mod = s2, PF = s2_pf, nsims = h1_scen_NP)
+      h1_S2_sims <- project_from_filter2(mod = s2, PF = s2_pf, nsims = h1_scen_NP)
+
+      results <- list()
+      results$mod1_V1_sims <- agg_mod1_sims(h1_S2_sims)
+      results$mod1_V1_probs <- get_elimProbs(h1_S2_sims, model = 1)
+
+      results
     },
@@
 }
-
-# Clean-up simulations for analysis/visualization
-mod1_V1_sims <- agg_mod1_sims(h1_S2_sims)
-mod1_V1_probs <- get_elimProbs(h1_S2_sims, model = 1)
-rm(h1_S2_sims)
 
@@
 if (RUN_LEVEL != 3 & file.exists("model1/run_level_3/S4_sims.rds")) {
-  h1_S4_sims <- readRDS("model1/run_level_3/S4_sims.rds")
+  h1_S4_results <- readRDS("model1/run_level_3/S4_sims.rds")
 } else {
-  h1_S4_sims <- bake(
+  h1_S4_results <- bake(
     file = paste0('model1/', rl_dir, 'S4_sims.rds'), {
@@
       # project vacc scenario, with filtering dist, and model.
-      project_from_filter2(mod = s4, PF = s4_pf, nsims = h1_scen_NP)
+      h1_S4_sims <- project_from_filter2(mod = s4, PF = s4_pf, nsims = h1_scen_NP)
+
+      results <- list()
+      results$mod1_V2_sims <- agg_mod1_sims(h1_S4_sims)
+      results$mod1_V2_probs <- get_elimProbs(h1_S4_sims, model = 1)
+
+      results
     },
@@
 }
-
-# Clean-up simulations for analysis/visualization
-mod1_V2_sims <- agg_mod1_sims(h1_S4_sims)
-mod1_V2_probs <- get_elimProbs(h1_S4_sims, model = 1)
-rm(h1_S4_sims)
 
@@
 if (RUN_LEVEL != 3 & file.exists("model1/run_level_3/S3_sims.rds")) {
-  h1_S3_sims <- readRDS("model1/run_level_3/S3_sims.rds")
+  h1_S3_results <- readRDS("model1/run_level_3/S3_sims.rds")
 } else {
-  h1_S3_sims <- bake(
+  h1_S3_results <- bake(
     file = paste0('model1/', rl_dir, 'S3_sims.rds'), {
@@
       # project vacc scenario, with filtering dist, and model.
-      project_from_filter2(mod = s3, PF = s3_pf, nsims = h1_scen_NP)
+      h1_S3_sims <- project_from_filter2(mod = s3, PF = s3_pf, nsims = h1_scen_NP)
+
+      results <- list()
+      results$mod1_V3_sims <- agg_mod1_sims(h1_S3_sims)
+      results$mod1_V3_probs <- get_elimProbs(h1_S3_sims, model = 1)
+
+      results
     },
@@
 }
-
-# Clean-up simulations for analysis/visualization
-mod1_V3_sims <- agg_mod1_sims(h1_S3_sims)
-mod1_V3_probs <- get_elimProbs(h1_S3_sims, model = 1)
-rm(h1_S3_sims)
 
@@
 if (RUN_LEVEL != 3 & file.exists("model1/run_level_3/S1_sims.rds")) {
-  h1_S1_sims <- readRDS("model1/run_level_3/S1_sims.rds")
+  h1_S1_results <- readRDS("model1/run_level_3/S1_sims.rds")
 } else {
-  h1_S1_sims <- bake(
+  h1_S1_results <- bake(
     file = paste0('model1/', rl_dir, 'S1_sims.rds'), {
@@
       # project vacc scenario, with filtering dist, and model.
-      project_from_filter2(mod = s1, PF = s1_pf, nsims = h1_scen_NP)
+      h1_S1_sims <- project_from_filter2(mod = s1, PF = s1_pf, nsims = h1_scen_NP)
+
+      results <- list()
+      results$mod1_V4_sims <- agg_mod1_sims(h1_S1_sims)
+      results$mod1_V4_probs <- get_elimProbs(h1_S1_sims, model = 1)
+
+      results
     },
@@
 }
-
-# Clean-up simulations for analysis/visualization
-mod1_V4_sims <- agg_mod1_sims(h1_S1_sims)
-mod1_V4_probs <- get_elimProbs(h1_S1_sims, model = 1)
-rm(h1_S1_sims, h1_scen_NP, h1_scen_sims)
-
-gc()
 @
@@
 <<Load Model 2 VaccScenarios, message=FALSE, include=FALSE>>=
-# load('model2/output/mod2_vacc_scenarios.rda')
 
@@
 mod2_VaccScenarios <- bake(
-  file = paste0("model2/", rl_dir, "VaccinationScenarios.rds"), {
+  file = paste0("model2/VaccinationScenarios.rds"), {
     haiti2_vaccScenario(
@@
 )
-
 @

### 2022-09-27 ms.Rnw
various edits; more to follow

--- 2022-09-26 ms.Rnw
@@
 Despite this autonomy, and largely adhering to existing guidelines on creating models to inform policy \citep{behrend20,saltelli20}, the consensus across the four models was that an extensive nationwide vaccination campaign would be necessary to eliminate cholera from Haiti.
-This conclusion is inconsistent with the fact that there have been no confirmed cases since February, 2019 \citep{trevisin22}, despite the lack of a concentrated vaccination effort.
+This conclusion is inconsistent with the fact that there have been no confirmed cases since February, 2019 \citep{trevisin22}, despite the lack of a national vaccination program.
 
@@
 We find a more nuanced conclusion: attention to methodological details in model fitting, diagnosis and forecasting can improve each of the proposed model's ability to quantitatively describe observed data.
-These improvements results in forecasts that are more consistent with the observed outcome, without requiring major changes to the model structures.
+These improvements result in forecasts that are more consistent with the observed outcome, without requiring major changes to the model structures.
 Based on this retrospective analysis, we offer suggestions on fitting mechanistic models to dynamic systems for future studies.
 
-While this work is focused on the models proposed in \cite{lee20}, we believe that our suggestions are relevant to a larger class of problems.
-In order to determine the extent to which this work generalizes, we preformed a literature review by searching PubMed with keywords: Haiti, cholera, and model.
+While this work is focused on the models proposed in \cite{lee20}, our suggestions have broader relevance.
+To investigate the extent to which \citet{lee20} is typical of the substantial body of work on Haiti cholera dynamics, we preformed a literature review by searching PubMed with keywords: Haiti, cholera, model.
 The search resulted in 66 papers, of which 32 used dynamic models to describe the cholera epidemic in Haiti.
-This literature review confirmed that close attention to details in fitting, diagnosing, and forecasting is of great importance when modeling an infectious disease outbreak.
-% Although the literature review focused on cholera and Haiti,
-The wide relevance of our findings in these papers---supplemented with studies of other epidemiological and dynamic systems---suggests that this work may also be relevant to a much broader class of problems.
+The models make various choices on the dichotomies considered by \citet{lee20}; deterministic versus stochastic; compartment model versus agent-based; aggregated versus spatially explicit.
+This is no accident, since \citet{lee20} purposefully designed their study to cover the range of current modeling practice.
 
 We proceed by introducing Models~1--3 in Sec.~\ref{sec:models};
-in Sec.~\ref{sec:methods}, we present a methodological approach to examining and refining these models, and then use improved model fits to project cholera incidence in Haiti under various vaccination scenarios.
+in Sec.~\ref{sec:methods}, we present a systematic approach to examining and refining these models, and then use improved model fits to project cholera incidence in Haiti under various vaccination scenarios.
 This is followed by a consideration of the robustness of model based policy recommendations in Sec.~\ref{sec:advice}, and a discussion in Sec.~\ref{sec:discussion}.
@@
 
-The deliberate limitation of coordination between the groups of \citet{lee20} allows us to treat the models as fairly independently developed expert approaches to understanding cholera transmission.
+The four mechanistic models of \citet{lee20} were deliberately developed with limited coordination.
+This allows us to treat the models as fairly independently developed expert approaches to understanding cholera transmission.
 However, it led to differences in notation, and in subsets of the data chosen for analysis, that hinder direct comparison.
@@
 
-Each model describes the cholera dynamics using a latent state vector $\bm{X}^{(\modelCounter)}(t)$ for each continuous time-point $t \in \mathcal{T}$, where $\mathcal{T}$ is the set of all time-points and $\modelCounter \in \{1, 2, 3\}$ indexes the model.
+Each model describes the cholera dynamics using a latent state vector $\bm{X}^{(\modelCounter)}(t)$ for each continuous time-point $t$ and $\modelCounter \in \{1, 2, 3\}$ indexes the model.
+The $N$ observation times are denoted by $t_1,\dots,t_N$ which is also written as $t_{1:N}$.
 The observation at time $t_n$ is modeled by a response vector $\bm{Y}^{(\modelCounter)}_n$.
@@
 
-In the following subsections, complete descriptions of Models~1--3 are provided.
+In the following subsections, complete descriptions of Models~1--3 are provided \eic{WE NEED INITIAL VALUES AND A MEASUREMENT MODEL TO COMPLETE THE DESCRIPTION.
+MAYBE WE JUST SAY WE'RE DESCRIBING THE DYNAMICS HERE, OR MAYBE WE ADD THE MISSING STUFF, WHICH HAS TO BE SOMEWHERE ANYHOW.
+SOME OF THIS IS DONE IN SEC 3.1.1 FOR MODEL 1, BUT PERHAPS BETTER TO DO IT SYSTEMATICALLY WHEN THE MODEL IS ORIGINALLY DEFINED.}.
 While the model description is scientifically critical, as well as necessary for transparency and reproducibility, the model details are not essential to our methodological discussions of how to diagnose and address model misspecification with the purpose of informing policy.
@@
 
+\eic{PERHAPS THIS PARAGRAPH COULD MOVE TO A LATER SECTION?
+THE FOLLOWING PARAGRAPH IS MORE WHAT I EXPECTED READING THROUGH WHEN I REACHED THIS POINT IN THE MS.
+PERHAPS PARTS OF THIS FIRST PARAGRAPH ARE EVEN REDUNDANT, IF THEY ARE COVERED IN THE ORIGINAL DISCUSSION OF THE MODELS.}
 Proposed mechanistic structures form a family of statistical models indexed by a parameter vector $\paramVec$.
@@
 The complex nature of biological systems necessitates a search for modeling assumptions that combine insightful simplicity with fidelity to biological reality.
-For example, many models commonly used in epidemiology are motivated by reasoning about a homogeneous mixing population \citep{bansal07} which is simultaneously an avenue for powerful simplification and a source of model misspecification.
+For example, many models commonly used in epidemiology are motivated by reasoning about a homogeneous mixing population which is simultaneously an avenue for powerful simplification and a source of model misspecification \citep{bansal07}.
 Other common considerations include whether the proposed model should be stochastic or deterministic; whether the model should have change points in parameter values or should otherwise make adjustments for changes through time in the dynamic system; and whether the proposed model should include any spatial heterogeneity at a scale permissible by the observed data.
@@
 
+\eic{ALL THE MODELS ARE PROBABILISTIC IN SOME SENSE, EVEN THOSE WITH DETERMINISTIC DYNAMICS.
+HAVING POINTED OUT THAT THEY ARE ALL POMP MODELS ABOVE, MAYBE WE SHOULD DEFINE THE CORRESPONDING NOTATION ABOVE TOO, RATHER THAN SPECIFICALLY FOR MODEL 1}
 One approach to modeling a dynamic system is through probabilistic models.
@@
 
+\eic{MAYBE THE KEY THING HERE IS THAT MODEL 1 IS A ``STANDARD'' POMP MODEL, FOR WHICH MUCH SUCCESSFUL METHODOLOGY IS WIDELY AVAILABLE. THAT IS BASICALLY WHAT WE PROCEED TO SAY.}
 Several algorithms exist, both frequentist and Bayesian, that can be used to obtain estimates of the parameters for this class of models.
@@
 
-The ability to maximize the likelihood allows for likelihood-based inference, like performing statistical tests for potential improvements to the model.
-% Part of the appeal of mechanistic models lies in their ability to illustrate a dynamical system.
-% Inherent is dynamical systems is randomness, and, in the context of epidemiology, this randomness is a crucial component to the system itself.
-% Useful and statistically sound incidence forecasting and prediction of vaccine efficacy are not only dependent upon a reasonable reflection of the natural system in the model; it is necessary for the model to also reflect the quantitative characteristics of the processes in question.
-% This need, itself, is dependent upon inclusion of sufficient stochasticity in the model \citep{breto09}.
+The ability to maximize the likelihood allows for likelihood-based inference, like \eic{I THINK like IS TOO INFORMAL AND PERHAPS ALSO VAGUE; e.g.
+OR for example WOULD NOT WORK HERE; MAYBE including?} performing statistical tests for potential improvements to the model.
 We demonstrate this capability by proposing a linear trend $\transmissionTrend$ in transmission in Eq.~\myeqref{model1:beta}:
@@
 Where $\bar{t}$ is the linear mapping $\bar{t}: [0, N] \rightarrow [-1, 1]$ of the time $t$.
+\eic{BEST TO AVOID NOTATION MORE MATHEMATICALLY ABSTRACT THAN NECESSARY - WE WANT THIS TO BE ACCESSIBLE TO PEOPLE WITHOUT MUCH ANALYSIS EXPERIENCE.
+ALSO, THIS IS NOT QUITE CORRECT; PERHAPS WE MEAN $[t_0,t_N]$}
 The proposal of a linear trend in transmission is a result of observing an apparent decrease in reported cholera infections from 2012-2019 in Fig.~\ref{CholeraData}.
@@
 
-Model~1 is implemented using the \code{pomp} package \cite{pomp2}, relying heavily on the source code provided by \cite{lee20}, who also used the MIF2 algorithm to estimate $\paramVec$.
-Aside from the introduction of the parameter $\transmissionTrend$ and the multiplicative Gamma-distributed white noise $d\Gamma(t)/dt$ in the process model (Eq.~\ref{model1:lambda}), the only change made that could increase the likelihood of the model was using a substantially increased computational effort to maximize the likelihood.
-While flexible, plug-and-play methods like MIF2 require a large number of computations, and the theoretic ability to maximize the likelihood depends on asymptotics in both the number of particles used and the number of MIF iterations.
+We implemented Model~1 using the \code{pomp} package \cite{pomp2}, relying heavily on the \code{pomp} source code provided by \cite{lee20}.
+Both analyses used the \code{mif2} implementation of the IF2 algorithm to estimate $\paramVec$ by maximum likelihood.
+Aside from the introduction of the parameter $\transmissionTrend$ and the multiplicative Gamma-distributed white noise $d\Gamma(t)/dt$ in the process model (Eq.~\ref{model1:lambda}), the another change we made to increase the likelihood of the fitted model was using a substantially increased computational effort in the numerical maximization.
+\eic{LET'S DISCUSS HOW THIS SENTENCE CAN BE MISREAD AT FIRST PASS (AND ARGUABLY IS AMBIGUOUS EVEN AFTER MORE THOUGHT).
+I THINK THIS IS COMMONLY THE CASE IN SENTENCES WITH MULITPLE SUBORDINATE CLAUSES} While flexible, plug-and-play methods like IF2 require a large number of computations, and the theoretic ability to maximize the likelihood depends on asymptotics in both the number of particles used and the number of filtering iterations.
 The large difference in likelihoods that were obtain by these two approaches (see Table~\ref{tab:likes}) highlights the importance of carefully determining the necessary computational effort needed to maximize model likelihoods and acting accordingly.
-
-% Simulations from the fitted model compared to the observed data are given in Fig~\ref{fig:mod1fit}.
 
@@
 
-Model~2 is a deterministic model, which are degenerate probabilistic models that contain no randomness in the dynamic process or measurement.
+Model~2 is a deterministic model, which are degenerate probabilistic models that contain no randomness in the dynamic process or measurement \eic{HERE, WE USE A MEMBER OF A CLASS AS A SHORT-CUT FOR THE CLASS ITSELF, WHICH READS STRANGELY AND FORMALLY HAS A PLURALITY MIS-MATCH.}.
 By combining a deterministic process model with a simple Gaussian measurement model, Model~2 reduces model fitting to a least squares calculation over parameters in a set of differential equations.
@@
 Other solutions include changing the measurement model to include overdispersion, as was done in Models~1 and~3.
+This permits the consideration of demographic stochasticity, which is dominant for small infected populations, togehter with log scale stochasticity (also called multiplicative, or environmental, or extra-demographic) which is dominant at high population counts.
 Here we chose to fit the model to transformed case counts rather than adding overdispersion to the measurement model with the goal of making the fewest changes to the model in \cite{lee20} as possible.
-In practice, however, the inclusion of overdispersion in the measurement process may be preferred.
 

### 2022-09-27 ms.Rnw
changed code to match updated haitipkg

@@
 
-  # Global search, All params
-  SEARCH1 <- list(
-    NP = 50, NMIF = 3, NREPS = 3,
-    NREPS_EVAL = 3, NP_EVAL = 100
-  )
-
-  # Local search, unit specific params
+  # Global search, All parameters
+  SEARCH1 <- list(NP = 20, NMIF = 3, NREPS = 3, NREPS_EVAL = 3, NP_EVAL = 25)
+
+  # Local search, all parameters
   SEARCH2 <- list(
-    TOP_N = 1, NP = 50, NMIF = 3, NREPS = 3, NREPS_EVAL = 3,
-    NP_EVAL = 100
-  )
+    TOP_N = 1, NP = 20, NMIF = 3, NREPS = 3, NREPS_EVAL = 3, NP_EVAL = 25
+  )
+
+  # Local search, unit specific parameters
+  SEARCH3 <- list(TOP_N = 1, NP = 20, NMIF = 3, NREPS = 3, NREPS_EVAL = 3, NP_EVAL = 25)
 
   # Number of searches
-  N_SEARCHES <- 2L
-
-  # THESE SEARCHES AREN'T CONDUCTED
-  SEARCH3 <- list(
-    TOP_N = 1, NP = 20, NMIF = 3, NREPS = 3, NREPS_EVAL = 3, NP_EVAL = 25
-  )
-
-  SEARCH4 <- list(
-    TOP_N = 1, NP = 20, NMIF = 3, NREPS = 3, NREPS_EVAL = 3, NP_EVAL = 25
-  )
-
-  SEARCH5 <- list(
-    TOP_N = 1, NP = 20, NMIF = 3, NREPS = 3, NREPS_EVAL = 3, NP_EVAL = 25
-  )
+  N_SEARCHES <- 3L
+
 } else if (RUN_LEVEL == 2) {
@@
 
-  # Local search, unit specific
+  # Local search, all parameters
   SEARCH2 <- list(
@@
 
-  # Local search, shared parameters
+  # Local search, unit specific parameters
   SEARCH3 <- list(
@@
 
-  # Local search, unit specific
-  SEARCH4 <- list(
-    TOP_N = 2, NP = 200, NMIF = 10, NREPS = 18, NREPS_EVAL = 36, NP_EVAL = 500
-  )
-
-  # Local search, all parameters
-  SEARCH5 <- list(
-    TOP_N = 2, NP = 200, NMIF =  5, NREPS = 18, NREPS_EVAL = 36, NP_EVAL = 500
-  )
-
   # Number of searches
-  N_SEARCHES <- 5L
+  N_SEARCHES <- 3L
 } else if (RUN_LEVEL == 3) {
@@
   SEARCH1 <- list(
-    NP = 1500, NMIF = 100, NREPS = 144, NREPS_EVAL = 36, NP_EVAL = 2000
+    NP = 1500, NMIF = 100, NREPS = 144, NREPS_EVAL = 36, NP_EVAL = 2000,
+    RW_SD = rw.sd(
+      betaB = 0.02,   # Unit Specific
+      foi_add = 0.02, # Unit Specific
+      mu_B = 0.02,
+      thetaI = 0.02,
+      XthetaA = 0.02,
+      lambdaR = 0.02,
+      r = 0.01,
+      std_W = 0.02,
+      epsilon = 0.02,
+      k = 0.02,
+      sigma = 0.02
+    )
+  )
+
+  # Local search, all parameters
+  SEARCH2 <- list(
+    TOP_N = 12, NP = 2000, NMIF = 100, NREPS = 6, NREPS_EVAL = 36, NP_EVAL = 3000,
+    RW_SD = rw.sd(
+      betaB = 0.0075,   # Unit Specific
+      foi_add = 0.0075, # Unit Specific
+      mu_B = 0.0075,
+      thetaI = 0.0075,
+      XthetaA = 0.0075,
+      lambdaR = 0.0075,
+      r = 0.005,
+      std_W = 0.0075,
+      epsilon = 0.0075,
+      k = 0.0075,
+      sigma = 0.0075
+    )
   )
@@
   # Local search, unit specific parameters
-  SEARCH2 <- list(
-    TOP_N = 4, NP = 2000, NMIF = 100, NREPS = 18, NREPS_EVAL = 36, NP_EVAL = 4000
-  )
-
-  # Local search, shared parameters
   SEARCH3 <- list(
-    TOP_N = 3, NP = 4000, NMIF = 50, NREPS = 24, NREPS_EVAL = 36, NP_EVAL = 5000
-  )
-
-  # Local search, unit specific parameters
-  SEARCH4 <- list(
-    TOP_N = 3, NP = 4000, NMIF = 50, NREPS = 24, NREPS_EVAL = 36, NP_EVAL = 5000
-  )
-
-  # Local search, all parameters
-  SEARCH5 <- list(
-    TOP_N = 3, NP = 4000, NMIF = 25, NREPS = 24, NREPS_EVAL = 36, NP_EVAL = 5000
+    TOP_N = 9, NP = 2000, NMIF = 100, NREPS = 8, NREPS_EVAL = 36, NP_EVAL = 4000,
+    RW_SD = rw.sd(
+      betaB = 0.005,
+      foi_add = 0.005
+    )
   )
@@
   # Total number of searches
-  N_SEARCHES <- 5L
+  N_SEARCHES <- 3L
 } else {
@@
         search3 = SEARCH3,
-        search4 = SEARCH4,
-        search5 = SEARCH5,
         n_searches = N_SEARCHES,
@@
 
-rm(SEARCH1, SEARCH2, SEARCH3, SEARCH4, SEARCH5, N_SEARCHES)
+rm(SEARCH1, SEARCH2, SEARCH3, N_SEARCHES)
 gc()
@@
       s0_pf <- pfilter(s0, save.state = TRUE, Np = h1_scen_NP)
+      end_states <- saved.states(s0_pf)
+      end_states <- end_states[[length(end_states)]]
 
       # project vacc scenario, with filtering dist, and model.
-      h1_S0_sims <- project_from_filter2(mod = s0, PF = s0_pf, nsims = h1_scen_sims)
+      h1_S0_sims <- project_from_filter(mod = s0, end_states = end_states, nsims = h1_scen_sims)
 
@@
       s2_pf <- pfilter(s2, save.state = TRUE, Np = h1_scen_NP)
+      end_states <- saved.states(s2_pf)
+      end_states <- end_states[[length(end_states)]]
 
       # project vacc scenario, with filtering dist, and model.
-      h1_S2_sims <- project_from_filter2(mod = s2, PF = s2_pf, nsims = h1_scen_NP)
+      h1_S2_sims <- project_from_filter(mod = s2, end_states = end_states, nsims = h1_scen_NP)
 
@@
       s4_pf <- pfilter(s4, save.state = TRUE, Np = h1_scen_NP)
+      end_states <- saved.states(s4_pf)
+      end_states <- end_states[[length(end_states)]]
 
       # project vacc scenario, with filtering dist, and model.
-      h1_S4_sims <- project_from_filter2(mod = s4, PF = s4_pf, nsims = h1_scen_NP)
+      h1_S4_sims <- project_from_filter(mod = s4, end_states = end_states, nsims = h1_scen_NP)
 
@@
       s3_pf <- pfilter(s3, save.state = TRUE, Np = h1_scen_NP)
+      end_states <- saved.states(s3_pf)
+      end_states <- end_states[[length(end_states)]]
 
       # project vacc scenario, with filtering dist, and model.
-      h1_S3_sims <- project_from_filter2(mod = s3, PF = s3_pf, nsims = h1_scen_NP)
+      h1_S3_sims <- project_from_filter(mod = s3, end_states = end_states, nsims = h1_scen_NP)
 
@@
       s1_pf <- pfilter(s1, save.state = TRUE, Np = h1_scen_NP)
+      end_states <- saved.states(s1_pf)
+      end_states <- end_states[[length(end_states)]]
 
       # project vacc scenario, with filtering dist, and model.
-      h1_S1_sims <- project_from_filter2(mod = s1, PF = s1_pf, nsims = h1_scen_NP)
+      h1_S1_sims <- project_from_filter(mod = s1, end_states = end_states, nsims = h1_scen_NP)
 
@@
 
-if (RUN_LEVEL != 3 & file.exists("model3/run_level_3/h3_bpf.rds")) {
-  h3_bpf <- readRDS("model3/run_level_3/h3_bpf.rds")
+if (RUN_LEVEL != 3 & file.exists("model3/run_level_3/h3_bpf_end_states.rds")) {
+  h3_bpf_end_states <- readRDS("model3/run_level_3/h3_bpf_end_states.rds")
 } else {
-  h3_bpf <- bake(
-    file = paste0("model3/", rl_dir, "h3_bpf.rds"),
+  h3_bpf_end_states <- bake(
+    file = paste0("model3/", rl_dir, "h3_bpf_end_states.rds"),
     expr = {
-      bpfilter(h3Spat, Np = NP_BPF, block_size = 1, save_states = TRUE)
+      h3_bpf <- bpfilter(h3Spat, Np = NP_BPF, block_size = 1, save_states = TRUE)
+      ss <- h3_bpf@saved.states
+      ss[[length(ss)]]
     },
@@
       # Using parameters and filtered distribution, project system
-      h3_V0_sims <- project_from_filter2(
-        h3Spat, PF = h3_bpf, covarGen = project_rain,
+      h3_V0_sims <- project_from_filter(
+        h3Spat, end_states = h3_bpf_end_states, covarGen = project_rain,
         nsims = NSIM, seed = 6897154
@@
       # Using parameters and filtered distribution, project system
-      h3_V1_sims <- project_from_filter2(
-        h3Spat, PF = h3_bpf, covarGen = project_rain,
+      h3_V1_sims <- project_from_filter(
+        h3Spat, end_states = h3_bpf_end_states, covarGen = project_rain,
         nsims = NSIM, seed = 6897154
@@
       # Using parameters and filtered distribution, project system
-      h3_V2_sims <- project_from_filter2(
-        h3Spat, PF = h3_bpf, covarGen = project_rain,
+      h3_V2_sims <- project_from_filter(
+        h3Spat, end_states = h3_bpf_end_states, covarGen = project_rain,
         nsims = NSIM, seed = 6897154
@@
       # Using parameters and filtered distribution, project system
-      h3_V3_sims <- project_from_filter2(
-        h3Spat, PF = h3_bpf, covarGen = project_rain,
+      h3_V3_sims <- project_from_filter(
+        h3Spat, end_states = h3_bpf_end_states, covarGen = project_rain,
         nsims = NSIM, seed = 6897154
@@
       # Using parameters and filtered distribution, project system
-      h3_V4_sims <- project_from_filter2(
-        h3Spat, PF = h3_bpf, covarGen = project_rain,
+      h3_V4_sims <- project_from_filter(
+        h3Spat, end_states = h3_bpf_end_states, covarGen = project_rain,
         nsims = NSIM, seed = 6897154

### 2022-09-27 ms.Rnw
some more edits

--- 2022-09-27 ms.Rnw
@@
 
-To combat this issue, we fit the model to log-transformed case counts, where the difference between periods of high and low infectiousness is less drastic.
+To combat this issue, we fit the model to log-transformed case counts, where the difference between periods of high and low infectiousness is less drastic \eic{, since the log scale stabilizes the variation during periods of high and low infectiousness??
+I'M NOT ENTIRELY HAPPY WITH THAT EITHER}.
 Other solutions include changing the measurement model to include overdispersion, as was done in Models~1 and~3.
-This permits the consideration of demographic stochasticity, which is dominant for small infected populations, togehter with log scale stochasticity (also called multiplicative, or environmental, or extra-demographic) which is dominant at high population counts.
+This permits the consideration of demographic stochasticity, which is dominant for small infected populations, together with log scale stochasticity (also called multiplicative, or environmental, or extra-demographic) which is dominant at high population counts.
 Here we chose to fit the model to transformed case counts rather than adding overdispersion to the measurement model with the goal of making the fewest changes to the model in \cite{lee20} as possible.
@@
 Both the latent and observable processes in Model~3 can be be factored into department specific processes which interact with each other.
-The decision to model a dynamic system via metapopulation models versus a model aggregated to a larger spatial scale is one of great scientific interest.
+This decision to address metapopulation dynamics via a spatially explicit model, rather than to aggregate over space, is double-edged.
 Evidence for the former approach has been provided in previous studies \citep{king15}, including the specific case of heterogeneity between Haitian departments in cholera transmission \citep{collins14}.
-Note that this evidence alone does not automatically discredit conclusions drawn via nationally aggregated models, as there are also reasons to prefer a simple model over a complex one \citep{saltelli20,green15}.
+A legitimate preference for simplicity \citep{saltelli20,green15} can support a decision to consider nationally aggregated models.
 
@@
 
-Parameters that must be fit in Model~3 are shared between each department, the exception to this being the parameters $\beta_{W_u}$, and $\beta_u$, which are unique for each department $u \in \seq{1}{10}$.
-To avoid the parameter estimation issue in high-dimensional models, \citet{lee20} simplified the problem by using the MIF2 algorithm to separately estimate parameters for independent department-level models;
+Parameters that must be fit in Model~3 are shared between each department, the exception to this being the parameters $\beta_{W_u}$, and $\beta_u$, which take department-specific values for each $u \in \seq{1}{10}$.
+To avoid the parameter estimation issue in high-dimensional models, \citet{lee20} simplified the problem by using the MIF2 \eic{IF2} algorithm to separately estimate parameters for independent department-level models;
 the shared parameters were calibrated using the cholera incidence data from Artibonite, and the department-specific parameters ($\beta_{W_u}$ and $\beta_u$) were fit using the data from their respective department.
@@
 In particular, when considering a model for department $u$, the values $I_\nu(t)$ and $A_\nu(t)$, are unknown for $u \neq v \in \seq{1}{10}$.
-A first order approximation of Eq.~\myeqref{eq:model3:foi} for each department $v \in \seq{1}{10}$ can be obtained using the weekly number of observed cholera cases in each department:
+A first order \eic{IS THIS BEING USED IN A TECHNICAL SENSE HERE?} approximation of Eq.~\myeqref{eq:model3:foi} for each department $v \in \seq{1}{10}$ can be obtained using the weekly number of observed cholera cases in each department:
 \begin{equation}
@@
 
-In the case of a PanelPOMP, an extension of the IF2 algorithm, known as Panel Iterated Filtering (PIF) \citep{breto20}, can be used to obtain the MLE, which solves the curse of dimensionality for this class of models.
+In the case of a PanelPOMP, an extension of the IF2 algorithm, known as Panel Iterated Filtering (PIF) \citep{breto20}, can be used to obtain the MLE, which solves the curse of dimensionality for this class of models \eic{WHICH CONVENTIONALLY REFERS TO THE PRECEDING OBJECT, HERE THE MLE, WHICH IS NOT WHAT WE WANT}.
 A major advantage of this algorithm is that PIF can be used to fit both unit-specific parameters and shared parameters; in this way, the calibration of shared parameters involves all of the available data instead of just an arbitrarily chosen subset.
 
+\eic{READERS MAY BE CONFUSED HERE. IS THIS A NEW METHOD? IT SEEMS TO BE IMPLEMENTED BY BRETO ET AL IN 2020. THIS IS DELICATE.}
 We propose and use a slight modification of the PIF algorithm, which we call the Block Panel Iterated Filter (BPIF), to estimate parameters for a PanelPOMP version of Model~3.
@@
 The output of the BPIF algorithm is the MLE $\tilde{\paramVec} = \argmax_{\paramVec} l_{\text{P}}(y_{1:N}^*; \paramVec)$.
-The BPIF algorithm is implemented in the \code{panelPomp} package \citep{breto20panelPomp} as the argument \code{block = TRUE} in the \code{mif2} function.
+The BPIF algorithm is implemented in the \code{panelPomp} package \citep{breto20panelPomp} as the argument \code{block = TRUE} in the \code{mif2} function \eic{THIS LEVEL OF CODING DETAIL IS UNUSUAL IN A CASE STUDY}.
 Pseudo-code for this algorithm is provided in Algorithm~S1 in the supplementary material.
@@
 
-Fitting Model~3 in this fashion simplifies the problem of parameter estimation, but it also introduces additional technicalities that must be addressed.
+Fitting Model~3 in this fashion simplifies the problem of parameter estimation, but it also introduces additional technicalities that must be addressed \eic{THIS MAKES IT SEEM LIKE THE FITTING METHOD CAUSES THE PROBLEM, BUT THE LEE ET AL APPROACH ALSO HAS THIS ISSUE.
+ARGUABLY, IT IS A PART OF THE MODEL NOT THE METHOD, THOUGH THE TWO ARE RELATED.
+MODEL 3 IS WRITTEN AS FULLY COUPLED IN SEC 2.3, SO FITTING THE NON-COUPLED VERSION IS ARGUABLY FITTING SOMETHING ELSE ENTIRELY, THOUGH IT CAN ALWAYS BE VIEWED AS AN APPROXIMATION.}
 One concern is that of obtaining model forecasts, which was the primary goal of \citet{lee20}.
@@
 Additional tools for assessing the goodness-of-fit of a model include plotting the effective sample size of each observation \citep{liu01}, and comparing summary statistics of the observed data to simulations from the model \citep{wood10}.
-These summary statistics are sometimes called diagnostic probes \citep{king16}.
+These summary statistics are sometimes called diagnostic probes \citep{king15,king16}.
 

### 2022-09-27 ms.Rnw
Fixed errors due to package update

--- 2022-09-27 ms.Rnw
@@
 In additional to probability of elimination estimates, we provide estimates for the cumulative number of infections under each vaccination scenario from February 2019 -- February 2024.
-Notably, the median number of cumulative cholera infections under the no-vaccination scenario using Models 1 and 3 were $\Sexpr{prettyNum(round(as.numeric(mod1_V0_probs$cumInf['q50'])), big.mark = ",")}$ and $\Sexpr{prettyNum(round(as.numeric(mod3_V0_res$mod3_V0_probs$cumInf['q50'])), big.mark = ",")}$, respectively.
+Notably, the median number of cumulative cholera infections under the no-vaccination scenario using Models 1 and 3 were $\Sexpr{prettyNum(round(as.numeric(h1_S0_results$mod1_V0_probs$cumInf['q50'])), big.mark = ",")}$ and $\Sexpr{prettyNum(round(as.numeric(mod3_V0_res$mod3_V0_probs$cumInf['q50'])), big.mark = ",")}$, respectively.
 While there is remaining time during this projection period in which new cholera infections can be detected, up to this point our estimates are far more consistent with the observed number of reported cholera cases than the corresponding estimates from \cite{lee20}, which were approximately $400,000$ and $1,000,000$.
@@
 gg_m1_V0 <- ggplot() +
-  geom_line(data = mod1_V0_sims, aes(x = date, y = q50 + 1), col = 'blue') +
-  geom_ribbon(data = mod1_V0_sims, aes(x = date, ymin = q05 + 1, ymax = q95 + 1), alpha = 0.5) +
+  geom_line(data = h1_S0_results$mod1_V0_sims, aes(x = date, y = q50 + 1), col = 'blue') +
+  geom_ribbon(data = h1_S0_results$mod1_V0_sims, aes(x = date, ymin = q05 + 1, ymax = q95 + 1), alpha = 0.5) +
   geom_line(data = true_agg_cases, aes(x = date, y = ReportedAll + 1)) +
@@
 gg_m1_V1 <- ggplot() +
-  geom_line(data = mod1_V1_sims, aes(x = date, y = q50 + 1), col = 'blue') +
-  geom_ribbon(data = mod1_V1_sims, aes(x = date, ymin = q05 + 1, ymax = q95 + 1), alpha = 0.5) +
+  geom_line(data = h1_S2_results$mod1_V1_sims, aes(x = date, y = q50 + 1), col = 'blue') +
+  geom_ribbon(data = h1_S2_results$mod1_V1_sims, aes(x = date, ymin = q05 + 1, ymax = q95 + 1), alpha = 0.5) +
   geom_line(data = true_agg_cases, aes(x = date, y = ReportedAll + 1)) +
@@
 gg_m1_V2 <- ggplot() +
-  geom_line(data = mod1_V2_sims, aes(x = date, y = q50 + 1), col = 'blue') +
-  geom_ribbon(data = mod1_V2_sims, aes(x = date, ymin = q05 + 1, ymax = q95 + 1), alpha = 0.5) +
+  geom_line(data = h1_S4_results$mod1_V2_sims, aes(x = date, y = q50 + 1), col = 'blue') +
+  geom_ribbon(data = h1_S4_results$mod1_V2_sims, aes(x = date, ymin = q05 + 1, ymax = q95 + 1), alpha = 0.5) +
   geom_line(data = true_agg_cases, aes(x = date, y = ReportedAll + 1)) +
@@
 gg_m1_V3 <- ggplot() +
-  geom_line(data = mod1_V3_sims, aes(x = date, y = q50 + 1), col = 'blue') +
-  geom_ribbon(data = mod1_V3_sims, aes(x = date, ymin = q05 + 1, ymax = q95 + 1), alpha = 0.5) +
+  geom_line(data = h1_S3_results$mod1_V3_sims, aes(x = date, y = q50 + 1), col = 'blue') +
+  geom_ribbon(data = h1_S3_results$mod1_V3_sims, aes(x = date, ymin = q05 + 1, ymax = q95 + 1), alpha = 0.5) +
   geom_line(data = true_agg_cases, aes(x = date, y = ReportedAll + 1)) +
@@
 gg_m1_V4 <- ggplot() +
-  geom_line(data = mod1_V4_sims, aes(x = date, y = q50 + 1), col = 'blue') +
-  geom_ribbon(data = mod1_V4_sims, aes(x = date, ymin = q05 + 1, ymax = q95 + 1), alpha = 0.5) +
+  geom_line(data = h1_S1_results$mod1_V4_sims, aes(x = date, y = q50 + 1), col = 'blue') +
+  geom_ribbon(data = h1_S1_results$mod1_V4_sims, aes(x = date, ymin = q05 + 1, ymax = q95 + 1), alpha = 0.5) +
   geom_line(data = true_agg_cases, aes(x = date, y = ReportedAll + 1)) +
@@
 
-% \begin{figure}[ht]
-% \centering
-%
-% mod1_cumInf <- data.frame(
-%   'scenario' = paste0("V", 0:4),
-%   'low' = c(
-%     as.numeric(mod1_V0_probs$cumInf['q025']),
-%     as.numeric(mod1_V1_probs$cumInf['q025']),
-%     as.numeric(mod1_V2_probs$cumInf['q025']),
-%     as.numeric(mod1_V3_probs$cumInf['q025']),
-%     as.numeric(mod1_V4_probs$cumInf['q025'])
-%   ),
-%   'high' = c(
-%     as.numeric(mod1_V0_probs$cumInf['q975']),
-%     as.numeric(mod1_V1_probs$cumInf['q975']),
-%     as.numeric(mod1_V2_probs$cumInf['q975']),
-%     as.numeric(mod1_V3_probs$cumInf['q975']),
-%     as.numeric(mod1_V4_probs$cumInf['q975'])
-%   ),
-%   'med' = c(
-%     as.numeric(mod1_V0_probs$cumInf['q50']),
-%     as.numeric(mod1_V1_probs$cumInf['q50']),
-%     as.numeric(mod1_V2_probs$cumInf['q50']),
-%     as.numeric(mod1_V3_probs$cumInf['q50']),
-%     as.numeric(mod1_V4_probs$cumInf['q50'])
-%   )
-% )
-%
-% mod3_cumInf <- data.frame(
-%   'scenario' = paste0("V", 0:4),
-%   'low' = c(
-%     as.numeric(mod3_V0_probs$cumInf['q025']),
-%     as.numeric(mod3_V1_probs$cumInf['q025']),
-%     as.numeric(mod3_V2_probs$cumInf['q025']),
-%     as.numeric(mod3_V3_probs$cumInf['q025']),
-%     as.numeric(mod3_V4_probs$cumInf['q025'])
-%   ),
-%   'high' = c(
-%     as.numeric(mod3_V0_probs$cumInf['q975']),
-%     as.numeric(mod3_V1_probs$cumInf['q975']),
-%     as.numeric(mod3_V2_probs$cumInf['q975']),
-%     as.numeric(mod3_V3_probs$cumInf['q975']),
-%     as.numeric(mod3_V4_probs$cumInf['q975'])
-%   ),
-%   'med' = c(
-%     as.numeric(mod3_V0_probs$cumInf['q50']),
-%     as.numeric(mod3_V1_probs$cumInf['q50']),
-%     as.numeric(mod3_V2_probs$cumInf['q50']),
-%     as.numeric(mod3_V3_probs$cumInf['q50']),
-%     as.numeric(mod3_V4_probs$cumInf['q50'])
-%   )
-% )
-%
-% mod1_cumInf$model <- "1"
-% mod3_cumInf$model <- "3"
-%
-% all_cumInf <- rbind(mod1_cumInf, mod3_cumInf)
-% all_cumInf$scenario <- factor(all_cumInf$scenario, levels = rev(c("V0", "V1", "V2", "V3", "V4")))
-%
-%
-% ggplot(all_cumInf, aes(y = scenario)) +
-%   geom_point(aes(x = med)) +
-%   geom_linerange(aes(xmin = low, xmax = high)) +
-%   scale_x_continuous(
-%     labels = scales::unit_format(scale = 1e-6, accuracy = 1),
-%     limits = c(0, 5e6)
-%   ) +
-%   xlab("Cumulative infections (million)") +
-%   theme(axis.title.y = element_blank(),
-%         axis.title.x = element_text(size = 10),
-%         axis.text = element_text(size = 8)) +
-%   facet_wrap(~model, ncol = 1,
-%              labeller = as_labeller(
-%                c("1" = "Model 1", "3" = "Model 3", "2" = "Model 2")
-%              ))
-%
-% \caption{\label{fig:cumInfPlot}
-% Simulated cumulative infections under various vaccination scenarios from February, 2019, to February, 2024.
-% Point estimates (median cumulative incidence) and $95\%$ error bars are included.
-% x-axis scale is intended to ease comparison with Figure~4 of \cite{lee20}.
-% }
-% \end{figure}
-
 <<Compute Elimination Probs, echo=FALSE>>=
-mod1_V0_probs$ElimTime$scenario = "V0"
-mod1_V1_probs$ElimTime$scenario = "V1"
-mod1_V2_probs$ElimTime$scenario = "V2"
-mod1_V3_probs$ElimTime$scenario = "V3"
-mod1_V4_probs$ElimTime$scenario = "V4"
+h1_S0_results$mod1_V0_probs$ElimTime$scenario = "V0"
+h1_S2_results$mod1_V1_probs$ElimTime$scenario = "V1"
+h1_S4_results$mod1_V2_probs$ElimTime$scenario = "V2"
+h1_S3_results$mod1_V3_probs$ElimTime$scenario = "V3"
+h1_S1_results$mod1_V4_probs$ElimTime$scenario = "V4"
 
 mod1_probElims <- rbind(
-  mod1_V0_probs$ElimTime,
-  mod1_V1_probs$ElimTime,
-  mod1_V2_probs$ElimTime,
-  mod1_V3_probs$ElimTime,
-  mod1_V4_probs$ElimTime
+  h1_S0_results$mod1_V0_probs$ElimTime,
+  h1_S2_results$mod1_V1_probs$ElimTime,
+  h1_S4_results$mod1_V2_probs$ElimTime,
+  h1_S3_results$mod1_V3_probs$ElimTime,
+  h1_S1_results$mod1_V4_probs$ElimTime
 )

### 2022-09-27 ms.Rnw
a few more edits; didn't have time to figure out why knitting failed

--- 2022-09-27 ms.Rnw
@@
 
-We consider model fitting (Sec.~\ref{sec:model_fitting}) followed by diagnostic investigations (Sec.~\ref{sec:model_diagnostics}), forecasting (Sec.~\ref{sec:filter}) and interpreting model fit (Sec.~\ref{sec:science}).
+We consider model fitting (Sec.~\ref{sec:model_fitting}) followed by diagnostic investigations (Sec.~\ref{sec:model_diagnostics}), forecasting (Sec.~\ref{sec:filter}) and \eic{PERHAPS external scientific corroboration of model fit RATHER THAN interpreting model fit} (Sec.~\ref{sec:science}).
 
@@
 Beyond being good practical advice for applied statistics, this assertion is relevant for the philosophical justification of statistical inference as severe testing \citep[][Sec.~4.8]{mayo18}.
-In this section, we provide some tools and suggestions for diagnosing mechanistic models with the goal of making the subjective assessment of model ``usefulness" more objective.
+In this section, we provide some tools and suggestions for diagnosing mechanistic models with the goal of making the subjective assessment of model ``usefulness'' more objective.
 To do this, we will rely on the quantitative and statistical ability of the model to match the observed data, which we call the model's {\it goodness-of-fit}, with the guiding principle that a model which cannot adequately describe observed data may not be reliable for useful purposes.
@@
 While visual comparisons can be informative, they provide only a weak and informal measure of the goodness-of-fit of a model.
-The study by \citet{lee20} provides an example of this: their models and parameter estimates resulted in simulations that visually resembled the observed data, yet resulted in model likelihoods that were---in some cases---remarkably smaller than likelihoods that can be achieved via the likelihood based optimization techniques that were used (see Table~\ref{tab:likes}).
+The study by \citet{lee20} provides an example of this: their models and parameter estimates resulted in simulations that visually resembled the observed data, yet resulted in model likelihoods that were---in some cases---considerably smaller than likelihoods that can be achieved via the likelihood based optimization techniques that were used (see Table~\ref{tab:likes}).
 Alternative forms of model validation should therefore be used in conjunction with visual comparisons of simulations to observed data.
@@
 Because of this, we use a log-linear Gaussian ARMA model as an associative benchmark, as recommended by \citet{he10}.
-The theory and practice of ARMA models is well developed, but the exponential growth and decay characteristic of biological dynamics suggests applying these linear models on a log scale.
+The theory and practice of ARMA models is well developed, and these linear models are appropriate on a log scale due to the exponential growth and decay characteristic of biological dynamics.
 Likelihoods of Models~1--3 and their respective ARMA benchmark models are provided in Table~\ref{tab:likes}.
@@
 It should be universal practice to present measures of goodness of fit for published models, and mechanistic models should be compared against benchmarks.
-This alone would assist authors and readers to confront any major statistical limitations of the proposed mechanistic models.
-In addition, the published goodness of fit provides a concrete measure for subsequent research to identify and remedy limitations in the analysis, or to update the investigation based on new data or new scientific understanding.
+In our literature review of the Haiti cholera epidemic, no benchmark models were considered in any of the 24 papers which calibrated a mechanistic model to data in order to obtain scientific conclusions.
+Including benchmarks would help authors and readers to detect and confront any major statistical limitations of the proposed mechanistic models.
+In addition, the published goodness of fit provides a concrete point of comparison for subsequent scientific investigations.
 When combined with online availability of data and code, objective measures of fit provide a powerful tool to accelerate scientific progress, following the paradigm of the {\it common task framework} \citep[][Sec.~6]{donoho17}.
-In our literature review of the Haiti cholera epidemic, no benchmark models were considered in any of the 24 papers which calibrated a mechanistic model to data in order to obtain scientific conclusions.
-
-The use of benchmarks may also be beneficial when developing models with varying spatial scale, where a direct comparison between models likelihoods is meaningless.
-In such a case, a benchmark model could be fit to each spatial resolution being considered, and each model compared to their respective benchmark.
+
+The use of benchmarks may also be beneficial when developing models at differing spatial scales, where a direct comparison between models likelihoods is meaningless.
+In such a case, a benchmark model can be fit to each spatial resolution being considered, and each model compared to their respective benchmark.
 Large advantages (shortcomings) in model likelihood relative to the benchmark for a given spatial scale that are not present in other spatial scales may provide weak evidence for (against) the statistical fit of models across a range of spatial resolutions.
@@
 
-\subsection{Reckoning Scientific Understanding With Fitted Models}\label{sec:science}
+\subsection{Corroborating Fitted Models with Previous Scientific Knowledge}\label{sec:science}
 
@@
 Agreement between model based inference and our current understanding of a system may taken as a confirmation of both model based conclusions and our scientific understanding.
-On the other hand, discrepancies found in these comparisons may have the potential to spark new scientific knowledge \citep{ganusov16}.
-The potential for new discoveries based on dynamic modeling outcomes, however, is rooted in a model's ability to effectively approximate the system in question, and is therefore inexorably linked to a models goodness-of-fit.
-As such, conclusions based on dynamic models that poorly fit observed data should not be considered as much as those resulting from models that describe observed data well.
-
-In our analysis, we demonstrate how one may compare model mechanisms with current scientific understanding by examining the results of fitting the flexible cubic spline term in Model~1 (Eq.~\myeqref{model1:lambda}--\myeqref{model1:beta});
-the cubic splines permit flexible estimation of seasonality in the force of infection.
-After calibrating model parameters to observed data, we explore potential patterns in the seasonal transmission rate by plotting the average value of $\transmission$ in a typical year.
+On the other hand, comparisons may generate unexpected results that have the potential to spark new scientific knowledge \citep{ganusov16}.
+
+%% The potential for new discoveries based on fitting a dynamic model, however, is rooted in a model's ability to effectively approximate the system in question, and is therefore inexorably linked to a model's goodness-of-fit.
+%% As such, conclusions based on dynamic models that poorly fit observed data should not be considered as much as those resulting from models that describe observed data well.
+THIS DUPLICATES LATER SENTENCES, SO I THINK IT CAN BE REMOVED HERE.
+
+In the context of our case study, we demonstrate how the fit of Model~1 corroborates other evidence concerning the role of rainfall in cholera epidemics.
+Specifically, we examine the results of fitting the flexible cubic spline term in Model~1 (Eq.~\myeqref{model1:lambda}--\myeqref{model1:beta}).
+The cubic splines permit flexible estimation of seasonality in the force of infection, $\transmission(t)$.
 Fig.~\ref{fig:h1SeasRain} shows that the estimated seasonal transmission rate $\transmission$ mimics the rainfall dynamics in Haiti, despite Model~1 not having access to rainfall data.
-This result provides evidence that rainfall played an important role in cholera transmission in Haiti, consistent with previous studies \citep{lemaitre19,eisenberg13}.
+This is consistent with previous studies finding that rainfall played an important role in cholera transmission in Haiti \citep{lemaitre19,eisenberg13}.
 The estimated seasonality also features an increased transmission rate during the Fall, which was noticed at an earlier stage of the epidemic \citep{rinaldo12}.

### 2022-09-28 si/si.Rnw

@@
 ) %dopar% {
-  guess <- coef(lee1_epi_mif)
+  guess <- lee1_epi_mif$lee1_epi_mif_params[i, ] 
   sims <- simulate(h1_epi, nsim = N_LEE_SIMS, format = 'data.frame',
@@
   sims$param_set <- i
-  sims
+  sims %>%
+    dplyr::select(.id, week, cases, param_set)
 } -> all_sims
 
-quants <- all_sims %>%
-  select(.id, week, cases) %>%
-  group_by(week) %>%
-  summarize(
-    q025 = quantile(cases, probs = 0.025, na.rm = TRUE),
-    q50  = quantile(cases, probs = 0.500, na.rm = TRUE),
-    q975 = quantile(cases, probs = 0.975, na.rm = TRUE)
-  ) %>%
-  ungroup() %>%
-  mutate(date = lubridate::ymd("2010-10-16") + lubridate::weeks(week))
-
 @
@@
 <<PlotEpiDist, echo=FALSE, fig.cap="Compare to Figure S8 of Lee et al. (2020) supplement.", fig.width=7, fig.height=7, fig.align='center'>>=
-epi_results <- as.data.frame(t(coef(lee1_epi_mif)))
-epi_results$logLik <- logLik(lee1_epi_mif)
+lee1_epi_evals <- readRDS(file = paste0("../model1/", RL_dir, "lee1_epi_evals.rds"))
+
+epi_results <- as.data.frame(lee1_epi_mif$lee1_epi_mif_params)
+epi_results$logLik <- lee1_epi_evals$pfLL
 
 GGally::ggpairs(
-  filter(epi_results, logLik > -2200),
+  dplyr::filter(epi_results, logLik > -20000),
   columns = c("rho", "tau", "beta1", "nu", "logLik")
@@
 foreach(
-  i = 1:length(lee1_end_mif),
+  i = 1:nrow(lee1_end_mif),
   .combine = rbind
 ) %dopar% {
-  guess <- coef(lee1_end_mif)
+  r_params <- lee1_end_mif[i, ]
+
+  lee1_epi_mif$start_states %>%
+    filter(param_set == i) %>%
+    select(-param_set) %>%
+    sapply(., median, na.rm = TRUE) -> rinit_parms
+
+  # Set the initial state
+  new_rinit <- Csnippet(
+    sprintf(
+      "
+          S = nearbyint(%f);
+          E = nearbyint(%f);
+          I = nearbyint(%f);
+          A = nearbyint(%f);
+          R = nearbyint(%f);
+          incid = nearbyint(%f);
+          foival = %f;
+          Str0 = nearbyint(%f);
+          Sout = nearbyint(%f);
+          Sin = nearbyint(%f);
+          ",
+      rinit_parms['S'],
+      rinit_parms['E'],
+      rinit_parms['I'],
+      rinit_parms['A'],
+      rinit_parms['R'],
+      rinit_parms['incid'],
+      rinit_parms['foival'],
+      rinit_parms['Str0'],
+      rinit_parms['Sout'],
+      rinit_parms['Sin']
+    )
+  )
+
+  h1_end <- h1_end %>% pomp(
+    rinit = new_rinit,
+    statenames = c(
+      "S", "E", "I", "A", "R", "incid",
+      "foival", "Str0", "Sout", "Sin"
+    )
+  )
+
+  coef(h1_end) <- r_params
+
   sims <- simulate(
-    lee1_end_mif[[i]], nsim = N_LEE_SIMS,
+    h1_end, nsim = N_LEE_SIMS,
     format = 'data.frame',
-    params = guess
+    params = r_params
   )
   sims$param_set <- i
-  sims
+  sims %>%
+    dplyr::select(.id, week, cases, param_set)
 } -> end_sims
 
-# quants_end <- end_sims %>%
-#   select(.id, week, cases) %>%
-#   group_by(week) %>%
-#   summarize(
-#     q025 = quantile(cases, probs = 0.025, na.rm = TRUE),
-#     q50  = quantile(cases, probs = 0.500, na.rm = TRUE),
-#     q975 = quantile(cases, probs = 0.975, na.rm = TRUE)
-#   ) %>%
-#   ungroup() %>%
-#   mutate(date = lubridate::ymd("2010-10-16") + lubridate::weeks(week))
 @
@@
 (2020) supplement.", cache=TRUE, echo=FALSE>>=
-
-end_results <- as.data.frame(t(coef(lee1_end_mif)))
-end_results$logLik <- logLik(lee1_end_mif)
+lee1_end_evals <- readRDS(file = paste0("../model1/", RL_dir, "lee1_end_evals.rds"))
+
+end_results <- as.data.frame(lee1_end_mif)
+end_results$logLik <- lee1_epi_evals$pfLL
 
 GGally::ggpairs(
-  filter(end_results, logLik > -2200),
+  filter(end_results, logLik > -20000),
   columns = c("rho", "tau", "beta1", "nu", "logLik")
@@
 Compare to Figure S7 of \\cite{lee20sup}.", cache=TRUE, fig.width=6.5, fig.height=3.75>>=
-end_complete <- filter(end_results, logLik > -2200)
-bad_end_pars <- which(end_results$logLik < -2200)
+
+epi_complete <- filter(epi_results, logLik > -20000)
+bad_epi_pars <- which(epi_results$logLik < -20000)
+
+quants <- all_sims %>%
+  filter(!param_set %in% bad_epi_pars) %>%
+  group_by(week) %>%
+  summarize(
+    q025 = quantile(cases, probs = 0.025, na.rm = TRUE),
+    q50  = quantile(cases, probs = 0.500, na.rm = TRUE),
+    q975 = quantile(cases, probs = 0.975, na.rm = TRUE)
+  ) %>%
+  ungroup() %>%
+  mutate(date = lubridate::ymd("2010-10-16") + lubridate::weeks(week))
+
+end_complete <- filter(end_results, logLik > -20000)
+bad_end_pars <- which(end_results$logLik < -20000)
 
@@
   filter(!param_set %in% bad_end_pars) %>%
-  select(.id, week, cases) %>%
   group_by(week) %>%
@@
 
-% \renewcommand\missing{---}
-%
-% \begin{table}
-% \begin{center}
-% \begin{tabular}{|l|c|c|c|c|}
-% \hline
-% Mechanism & Unified & Old model 1 & Old model 2 & Old model 3
-% \\
-% \hline
-% \hline
-% Recovery rate
-%   & $\muIR$
-%   & $\gamma$
-%   & $\gamma$
-%   & $\gamma$
-% \\
-% Latency
-%   & $\muEI$
-%   & $\sigma$
-%   & $\sigma$
-%   & \missing
-% \\
-% Direct transmission
-%   & $\transmission$
-%   & $\beta_{1:6}$
-%   & $\beta$
-%   & $c^i$
-% \\
-% Rainfall exponent
-%   & \rainfallExponent
-%   & \missing
-%   & \missing
-%   & $r$
-% \\
-% Lost immunity
-%   & $\muRS$
-%   & $\alpha$
-%   & $\alpha$
-%   & $\rho$
-% \\
-%   & $\omega_1$, $\omega_2$
-%   & \missing
-%   & $\omega_1$, $\omega_2$
-%   & \missing
-% \\
-% Vaccine efficacy
-%   & $\vaccineEfficacy_{1:2}$
-%   & \missing
-%   & $\theta_{1:2}$
-%   & \missing
-% \\
-% Birth rate
-%   & $\muBirth$
-%   & $\mu$
-%   &
-%   & $\mu$
-% \\
-% Death rate
-%   & $\muDeath$
-%   & $\delta$
-%   &
-%   & $\mu$
-% \\
-% Symptomatic frac.
-%   & $\symptomFrac$
-%   & $\theta_k(t)=c\theta^*(t-\tau_k)$
-%   & $k$
-%   & $\theta$
-% \\
-% Asympt.~infectivity
-%   & $\asymptomRelativeInfect$
-%   & $(1-\kappa)$
-%   & $\beta_A/\beta$
-%   & 1
-% \\
-% Asympt.~shedding
-%   & $\asymptomRelativeShed$
-%   & \missing
-%   & $\mu_A/\mu$
-%   & $\theta_A/\theta_I$
-% \\
-% Seasonality amplitude
-%   & $\seasAmplitude$
-%   & \missing
-%   & $\alpha_s$
-%   & $a$
-% \\
-% Water to human
-%   & $\Wbeta{}$
-%   & \missing
-%   & $\beta_W$
-%   & $\beta^i$
-% \\
-%   & $\Wsat$
-%   & \missing
-%   & $\mathrm{sat}$
-%   & $1$
-% \\
-% Cholera decay in water
-%   & $\Wremoval$
-%   & \missing
-%   & $\delta$
-%   & $\mu_B$
-% \\
-% Human to water
-%   & $\Wshed$
-%   & \missing
-%   & $\mu$
-%   & $\theta_I$
-% \\
-% Mixing exponent
-%   & $\mixExponent$
-%   & $\nu$
-%   & \missing
-%   & \missing
-% \\
-% Process noise
-%   & $\sigmaProc$
-%   & \missing
-%   & \missing
-%   & $\sigma$
-% \\
-% Reporting rate
-%   & $\reportRate$
-%   & $\rho$
-%   & $\rho$
-%   & $\epsilon$
-% \\
-% Observation overdispersion
-%   & $\obsOverdispersion$
-%   & $\tau$
-%   & \missing
-%   & $k$
-% \\
-% \hline
-% \end{tabular}
-% \end{center}
-% \caption{Translation from the unified notation of this article back into the notations used for Lee et al.~(2020)}
-% \label{tab:translate}
-% \end{table}
-
-% \section{Investigation into model 2}
-%
-% As part of our case study into the Haiti paper, model 2 was recreated using the spatPomp package in R.
-To confirm the validity of our model, we recreated Figure S15 from the supplement.
-In the original model, the authors reinitialized the state values for the endemic phase starting on March 2014.
-In particular, the group assumed that 75\% of the population was susceptible to cholera.
-We believe this choice to be unnatural.
-Having the state values change in the middle of a simulation has no causal interpretation.
-Instead, the most logical thing to do when fitting an epidemic and endemic model would be to carry over the state values from the former to the latter.
-This keeps the integrity of the model intact.
-%
-% In Figure S-2, we created the same endemic model with the only difference being the state values are carried over from the end of the epidemic phase.
-Here, we see that the model predicts a much different trajectory than in Figure S-1.
-This suggests that the decision to reinitialize state values in the middle of a simulation, as in the original model, is likely a poor choice since it has no cause and does not reflect the nature of the epidemic.
-One consideration is the choice of measurement component.
-In the original model, we use a linear measurement component.
-To avoid a situation where the trajectory goes to zero, we can implement a logarithmic measurement component.
-This is a more natural measure since the fit should be relative to the magnitude of the reported data.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \clearpage

### 2022-09-28 ms.Rnw
minor edits

@@
 
-for (i in 1:3) {
+# run_level_* for model 2 is not needed.
+for (i in c(1, 3)) {
   if (!dir.exists(paste0("model", i, '/', rl_dir))) {
@@
 }
+
+if (!dir.exists('model2')) dir.create("model2/")
 
@@
   # Global search, all parameters
-  SEARCH1 <- list(
-    NP = 1500, NMIF = 100, NREPS = 144, NREPS_EVAL = 36, NP_EVAL = 2000,
-    RW_SD = rw.sd(
-      betaB = 0.02,   # Unit Specific
-      foi_add = 0.02, # Unit Specific
-      mu_B = 0.02,
-      thetaI = 0.02,
-      XthetaA = 0.02,
-      lambdaR = 0.02,
-      r = 0.01,
-      std_W = 0.02,
-      epsilon = 0.02,
-      k = 0.02,
-      sigma = 0.02
-    )
-  )
-
-  # Local search, all parameters
-  SEARCH2 <- list(
-    TOP_N = 12, NP = 2000, NMIF = 100, NREPS = 6, NREPS_EVAL = 36, NP_EVAL = 3000,
-    RW_SD = rw.sd(
-      betaB = 0.0075,   # Unit Specific
-      foi_add = 0.0075, # Unit Specific
-      mu_B = 0.0075,
-      thetaI = 0.0075,
-      XthetaA = 0.0075,
-      lambdaR = 0.0075,
-      r = 0.005,
-      std_W = 0.0075,
-      epsilon = 0.0075,
-      k = 0.0075,
-      sigma = 0.0075
-    )
-  )
-
-  # Local search, unit specific parameters
-  SEARCH3 <- list(
-    TOP_N = 9, NP = 2000, NMIF = 100, NREPS = 8, NREPS_EVAL = 36, NP_EVAL = 4000,
-    RW_SD = rw.sd(
-      betaB = 0.005,
-      foi_add = 0.005
-    )
+
+  SEARCH1 = list(
+    NBPF = 100,
+    NP = 1500,
+    SPAT_REGRESSION = 0.5,
+    NREPS = 144,
+    NP_EVAL = 2000,
+    NREPS_EVAL = 36
+  )
+
+  SEARCH2 = list(
+    TOP_N = 12,
+    NBPF = 100,
+    NP = 2000,
+    SPAT_REGRESSION = 0.5,
+    NREPS = 6,
+    NP_EVAL = 3000,
+    NREPS_EVAL = 36
+  )
+
+  SEARCH3 = list(
+    TOP_N = 9,
+    NBPF = 100,
+    NP = 2500,
+    SPAT_REGRESSION = 0.5,
+    NREPS = 8,
+    NP_EVAL = 5000,
+    NREPS_EVAL = 36
   )

### 2022-09-28 si/si.Rnw
si edits

@@
 
+\eic{CHECK: EVERYTHING IN SI SHOULD BE REFERENCED IN THE MAIN TEXT.}
+
 \section{Markov chain and differential equation interpretations of compartment flow rates}
 
-In Sections 2.1, 2.2 and 2.3 we define compartment models in terms of their flow rates.
+In Sections 2.1, 2.2 and 2.3 of the main article, we define compartment models in terms of their flow rates.
 For a discrete population model, these rates define a Markov chain.
@@
 \end{equation}
+\eic{medskip IS VERTICAL SPACE. I USUALLY USE hspace}
 where $\mu_{ij}\big(t,X(t)\big)$ is called a per-capita rate or a unit rate.
@@
 \end{equation}
-
 This is the the interpretation and implementation used for Model~2 in our study.
@@
 The infinitesimal mean of every flow is equal to its infinitesimal variance \citep{breto11} and so an SMCS is called equidispersed.
+\eic{HERE AND ELSEWHERE, in Lee et al SHOULD BE by Lee et al. CONVENTIONALLY, THE TEXT CITATION REFERS TO THE PEOPLE RATHER THAN THE ARTICLE.}
 We note that the special case of Model~1 used in \cite{lee20} (with $\sigmaProc = 0$) is an SMCS.
@@
 Formally interpreting the meaning of \myeqref{eq:odmcs1} is not trivial, and we do so by defining the solution of \myeqref{eq:odmcs1} to be the limit of an Euler scheme.
-% This is equivalent to interpreting a stochastic differential equation via its Ito solution \eic{CHECK THIS CLAIM.
-I THINK IT MIGHT BE TRUE.
-IF SO, IT WOULD PROVIDE A NICE BRIDGE BETWEEN DEVELOPMENT OF IMPLICIT MODELS (I.E., DEFINING MODELS VIA NUMERICAL SOLUTIONS, OFTEN DONE USING AN EULER METHOD) AND THE SUCCESSFUL THEORY OF ITO SOLUTIONS TO STOCHASTIC DIFFERENTIAL EQUATIONS.}
 Therefore, the numerical scheme in Sec.~\ref{sec:numerics} can be taken as a definition of the meaning of \myeqref{eq:odmcs1}.
@@
 
-Model~2 assumes that the initial vales are constant, and no parameters need to be estimated.
-Starting value for latent state compartments are chosen so as to enforce the model dynamics on the observed number of cases.
+Model~2 assumes that the initial vales are known and constant, and so no initial value parameters need to be estimated.
+Starting values for latent state compartments are chosen so as to enforce the model dynamics on the observed number of cases.
 Specifically, the model sets $I_{u}(0) = y^*_{u}(0) / \reportRate$ for each unit $u \in 1:10$, where $I_{u}(t)$ is the number of infected individuals in unit $u$ at time $t$, $y^*_{u}(t)$ is the reported number of cases, and $\reportRate$ is the reporting rate.
@@
 The bacteria compartment is then initialized using Eq.~\myeqref{eq:model3:b0}:
-
 \begin{equation}
-  B_u(0) = \big[1 + \seasAmplitude \big(\xi_u)^r \big] D_i \, \Wshed \big[ I_{u}(0)+ \asymptomRelativeShed A_{u}(0) \big]\label{eq:model3:b0}
+  B_u(0) = \big[1 + \seasAmplitude \big(\xi_u)^r \big] D_i \, \Wshed \big[ I_{u}(0)+ \asymptomRelativeShed A_{u}(0) \big]\label{eq:model3:b0},
 \end{equation}
-
-Where $\xi_u \in (0, 1)$ are initial value parameters that we introduce in order to allow some flexibility in determining the initial state of the bacteria compartment.
+where $\xi_u \in (0, 1)$ is an initial value parameter that allows some flexibility in determining the initial state of the bacteria compartment.
 
@@
 Where $y_t$ is the reported cholera cases at time $t \in 1:N$ and $\Delta N_{E_{\vaccCounter}I_{\vaccCounter}}(t)$ is the number of individuals who moved from compartment $E_\vaccCounter$ to $I_\vaccCounter$ since observation $t-1 \in 0:N-1$.
+\eic{HERE AND ELSEHWERE: PUNCTUATION. NO NEED FOR A NEW PARAGRAPH HERE AFTER THE DISPLAY EQUATION. ALSO, PROBABLY THERE SHOULD BE A COMMA AT THE END OF THE EQUATION AND Where THEN IS NOT CAPITALIZED}
+
+\eic{AS A PEDANTIC POINT WHICH IS NOT UNIVERSALLY IMPLEMENTED BUT I HAPPEN TO THINK IS WORTHWHILE: THE NOTATION IN S16 USUALLY CORRESPONDS TO A DISTRIBUTION, BUT $y_n$ (WHICH I SUPPOSE HERE IS WRITTEN $y_t$) IS DATA, AND THEREFORE A NUMBER NOT A RANDOM VARIABLE.
+WE MODEL IT BY A RANDOM VARIABLE $Y_n$.
+THE DISTINCTION MAY NOT BE CONFUSING FOR MANY, BUT WHEN WE ARE ENCOURAGING OURSELVES AND OTHERS TO THINK ABOUT THE DIFFERENCES BETWEEN DATA AND MODELS FOR DATA, IT IS COUNTER-PRODUCTIVE TO HAVE NOTATION WHICH SUPPRESSES THIS DIFFERENCE.}
+
+\eic{DUE TO DIFFERENT PARAMETERIZATIONS OF NB (AS WELL AS TO CLARIFY THE MODEL) I THINK IT IS GOOD PRACTICE TO GIVE THE MOMENTS, E.G., ``, and $\text{NB}\left(\reportRate z, \obsOverdispersion \right)$ is the negative binomial distribution with mean $\reportRate z$ and variance $\reportRate z+ \psi^2\reportRate^2 z^2)$'' IF THIS IS CORRECT.}
 
@@
 
-The panel iterated filter (PIF) \citep{breto20} has been successfully used to fit parameters of panelPOMP models in several studies \citep{domeyer22,wale2019,breto20}.
+The panel iterated filter (PIF) algorithm of \citet{breto20} has been used to fit parameters of panelPOMP models in several studies \citep{domeyer22,wale2019,ranjeva17,ranjeva19}.
 Like other iterated filtering algorithms, PIF iteratively combines particle filter calculations with parameter perturbations.
@@
 
-As a solution to this issue, we propose the block panel iterated filter (BPIF).
+As a solution to this issue, we propose the block panel iterated filter (BPIF), described in Algorithm~\ref{alg:bpif}.
 In this algorithm, shared $\phi$ and unit-specific parameters $\psi_{u}$ are handled differently when updating the parameter vector $\theta = (\phi, \psi_{1:U})$.
@@
 In the case where there are many units, we believe that this modification will lead to less particle depletion.
-Pseudo-code for this algorithm is given in Algorithm~\ref{alg:bpif}.
-We note that no theoretical properties of this algorithm have yet been derived, but early empirical results suggest that this algorithm is a an improvement to the PIF algorithm.
+\eic{USUALLY BEST TO AVOID believe IN SCIENTIFIC WRITING.
+CAN WE SAY SOMETHING SOLID ABOUT WHAT WE HAVE FOUND.
+EVEN, I THINK THE FOLLOWING IS TRUE: We have found that this modification leads to improved likelihood maximization performance, especially as the number of units increases.}
+No theoretical properties of this algorithm have yet been derived, but the approach has similarities to the block particle filter of \citet{rebeschini15} and its extension to parameter estimation \citep{ionides22,ning21ibpf}.
 

### 2022-09-28 ms.Rnw
Addressing a few of Ed's comments

@@
 
-The ability to maximize the likelihood allows for likelihood-based inference, like \eic{I THINK like IS TOO INFORMAL AND PERHAPS ALSO VAGUE; e.g.
-OR for example WOULD NOT WORK HERE; MAYBE including?} performing statistical tests for potential improvements to the model.
+The ability to maximize the likelihood allows for likelihood-based inference, such as performing statistical tests for potential model improvements.
 We demonstrate this capability by proposing a linear trend $\transmissionTrend$ in transmission in Eq.~\myeqref{model1:beta}:
@@
 \end{equation}
-Where $\bar{t}$ is the linear mapping $\bar{t}: [0, N] \rightarrow [-1, 1]$ of the time $t$.
-\eic{BEST TO AVOID NOTATION MORE MATHEMATICALLY ABSTRACT THAN NECESSARY - WE WANT THIS TO BE ACCESSIBLE TO PEOPLE WITHOUT MUCH ANALYSIS EXPERIENCE.
-ALSO, THIS IS NOT QUITE CORRECT; PERHAPS WE MEAN $[t_0,t_N]$}
+Where $\bar{t} = \frac{t - (t_N + t_0) / 2}{t_N - (t_N + t_0) / 2}$, so that $\bar{t} \in [-1, 1]$.
 The proposal of a linear trend in transmission is a result of observing an apparent decrease in reported cholera infections from 2012-2019 in Fig.~\ref{CholeraData}.
@@
 Both analyses used the \code{mif2} implementation of the IF2 algorithm to estimate $\paramVec$ by maximum likelihood.
-Aside from the introduction of the parameter $\transmissionTrend$ and the multiplicative Gamma-distributed white noise $d\Gamma(t)/dt$ in the process model (Eq.~\ref{model1:lambda}), the another change we made to increase the likelihood of the fitted model was using a substantially increased computational effort in the numerical maximization.
-\eic{LET'S DISCUSS HOW THIS SENTENCE CAN BE MISREAD AT FIRST PASS (AND ARGUABLY IS AMBIGUOUS EVEN AFTER MORE THOUGHT).
-I THINK THIS IS COMMONLY THE CASE IN SENTENCES WITH MULITPLE SUBORDINATE CLAUSES} While flexible, plug-and-play methods like IF2 require a large number of computations, and the theoretic ability to maximize the likelihood depends on asymptotics in both the number of particles used and the number of filtering iterations.
+One change we made in the statistical analysis that led to larger model likelihoods was increasing the computational effort in the numerical maximization.
+While IF2 is a powerful method that can be used to fit parameters for a large class of models, the theoretic ability to maximize the likelihood depends on asymptotics in both the number of particles and the number of filtering iterations and therefore requires a large number of computations.
 The large difference in likelihoods that were obtain by these two approaches (see Table~\ref{tab:likes}) highlights the importance of carefully determining the necessary computational effort needed to maximize model likelihoods and acting accordingly.
@@
 
-Model~2 is a deterministic model, which are degenerate probabilistic models that contain no randomness in the dynamic process or measurement \eic{HERE, WE USE A MEMBER OF A CLASS AS A SHORT-CUT FOR THE CLASS ITSELF, WHICH READS STRANGELY AND FORMALLY HAS A PLURALITY MIS-MATCH.}.
-By combining a deterministic process model with a simple Gaussian measurement model, Model~2 reduces model fitting to a least squares calculation over parameters in a set of differential equations.
+Model~2 is a system of deterministic differential equations and can therefore be considered a degenerate probabilistic model with no randomness in the dynamic process.
+By combining a deterministic process model with a simple Gaussian measurement model, the parameter estimation problem for a set of differential equations reduces to a least squares calculation.
 Deterministic compartment models have a long history in the field of infectious disease epidemiology \citep{kermack1927,brauer2017,varghese21},
@@
 Parameters that must be fit in Model~3 are shared between each department, the exception to this being the parameters $\beta_{W_u}$, and $\beta_u$, which take department-specific values for each $u \in \seq{1}{10}$.
-To avoid the parameter estimation issue in high-dimensional models, \citet{lee20} simplified the problem by using the MIF2 \eic{IF2} algorithm to separately estimate parameters for independent department-level models;
+To avoid the parameter estimation issue in high-dimensional models, \citet{lee20} simplified the problem by using the IF2 algorithm to separately estimate parameters for independent department-level models;
 the shared parameters were calibrated using the cholera incidence data from Artibonite, and the department-specific parameters ($\beta_{W_u}$ and $\beta_u$) were fit using the data from their respective department.
@@
 In particular, when considering a model for department $u$, the values $I_\nu(t)$ and $A_\nu(t)$, are unknown for $u \neq v \in \seq{1}{10}$.
-A first order \eic{IS THIS BEING USED IN A TECHNICAL SENSE HERE?} approximation of Eq.~\myeqref{eq:model3:foi} for each department $v \in \seq{1}{10}$ can be obtained using the weekly number of observed cholera cases in each department:
+An approximation of Eq.~\myeqref{eq:model3:foi} for each department $v \in \seq{1}{10}$ can be obtained using the weekly number of observed cholera cases in each department:
 \begin{equation}
@@
 $P\big(\Theta_{j} = \theta_k | \bm{Y}^{(\modelCounter)}_{1:N} = y_{1:N}^*\big) = \frac{l(y_{1:N}^*; \theta_k)}{\sum_{l = 1}^K l(y_{1:N}^*; \theta_l)}$.
-To complete the formulation, we then choose the set $\Theta_K$ such that each $\theta_k \in \Theta_K$ is a single output of the MIF2 algorithm, which has been shown to be within a neighborhood of the MLE \cite{ionides15}.
+To complete the formulation, we then choose the set $\Theta_K$ such that each $\theta_k \in \Theta_K$ is a single output of the IF2 algorithm, which has been shown to be within a neighborhood of the MLE \cite{ionides15}.
 This approach has been taken in previous studies \citep{king15}, and the result is simply weighing parameter sets based on their corresponding likelihoods: parameters with higher likelihoods should be used more frequently in model forecasts than those that correspond to lower likelihoods.

### 2022-09-28 ms.Rnw
Addressed a few comments

@@
 
-Each model describes the cholera dynamics using a latent state vector $\bm{X}^{(\modelCounter)}(t)$ for each continuous time-point $t$ and $\modelCounter \in \{1, 2, 3\}$ indexes the model.
-The $N$ observation times are denoted by $t_1,\dots,t_N$ which is also written as $t_{1:N}$.
-The observation at time $t_n$ is modeled by a response vector $\bm{Y}^{(\modelCounter)}_n$.
-The latent state vector $\bm{X}^{(\modelCounter)}(t)$ consists of individuals labeled as susceptible (S), infected (I), asymptomatically infected (A), vaccinated (V), and recovered (R), with various sub-divisions sometimes considered in each model $\modelCounter$.
+Each model describes the cholera dynamics as a partially observed Markov process (POMP) with a latent state vector $\bm{X}^{(\modelCounter)}(t)$ for each continuous time point $t$ and model $\modelCounter \in \{1, 2, 3\}$.
+$N$ observations on the system are collected at time points $t_1,\dots,t_N$, written as $t_{1:N}$.
+The observation at time $t_n$ is modeled by the random vector $\bm{Y}^{(\modelCounter)}_n$.
+While the latent process exists between observation times, the value of the latent state at observations times is of particular interest.
+We therefore write $\bm{X}^{(\modelCounter)}_n = \bm{X}^{(\modelCounter)}(t_n)$ to denote the value of the latent process at the $n$th observation time, and $\bm{X}^{(\modelCounter)}_{1:N}$ is the collection of latent state values for all observed time points.
+The observable random variables $\bm{Y}^{(\modelCounter)}_{1:N}$ are assumed to be conditionally independent given $\bm{X}^{(\modelCounter)}_{0:N}$.
+Together, with the density for the initial value of the latent state $\bm{X}^{(\modelCounter)}_{0} = \bm{X}^{(\modelCounter)}(t_0)$, each model assumes a joint density $f_{\bm{X}^{(\modelCounter)}_{0:N}, \bm{Y}^{(\modelCounter)}_{1:N}}\big(\bm{x}^{(\modelCounter)}_{0:N}, \bm{y}^{(\modelCounter)}_{1:N};\paramVec\big)$, where $\paramVec$ is a parameter vector that indexes the model.
+The observed data $\bm{y}_{1:N}^*$ are modeled as a realization of the observation processes.
+\jwc{SHOULD WE SAY SOMETHING ABOUT HOW THE JOINT DENSITY IS INSTRACTABLE, OR CAN BE FACTORED USING THE MARKOV PROPERTY?}
+
+For each model $\modelCounter \in \{1, 2, 3\}$, the latent state vector $\bm{X}^{(\modelCounter)}(t)$ consists of individuals labeled as susceptible (S), infected (I), asymptomatically infected (A), vaccinated (V), and recovered (R), with various sub-divisions sometimes considered.
 Models~2 and~3 have metapopulation structure, meaning that each individual is a member of a spatial unit, denoted by a subscript $u\in \seq{1}{U}$.
@@
 
-\eic{ALL THE MODELS ARE PROBABILISTIC IN SOME SENSE, EVEN THOSE WITH DETERMINISTIC DYNAMICS.
-HAVING POINTED OUT THAT THEY ARE ALL POMP MODELS ABOVE, MAYBE WE SHOULD DEFINE THE CORRESPONDING NOTATION ABOVE TOO, RATHER THAN SPECIFICALLY FOR MODEL 1}
-One approach to modeling a dynamic system is through probabilistic models.
-With such a model, the existence of a joint density $f_{\bm{X}^{(\modelCounter)}_{\seq{0}{N}}, \bm{Y}^{(\modelCounter)}_{\seq{1}{N}}}$ is supposed, with $\bm{X}^{(\modelCounter)}_{\seq{0}{N}}$ denoting the unobservable Markov process of model $\modelCounter$ at times $\seq{0}{N} = \{0, 1, \ldots, N\}$, and $\bm{Y}^{(\modelCounter)}_{\seq{1}{N}}$ denoting the observable process of the system at times $\seq{1}{N}$.
-Under this framework, the observed data $y_{\seq{1}{N}}^*$---along with some unobserved value $\bm{X}^{(\modelCounter)}_{\seq{0}{N}} = x^*_{\seq{0}{N}}$---are assumed to be a single realization of the model $\big(x^*_{\seq{0}{N}}, y_{\seq{1}{N}}^*\big) \sim f_{\bm{X}^{(\modelCounter)}_{\seq{0}{N}}, \bm{Y}^{(\modelCounter)}_{\seq{1}{N}}}(x_{\seq{0}{N}}, y_{\seq{1}{N}}; \paramVec)$, where $\paramVec$ is a parameter vector that indexes the model.
-This modeling framework results in several advantages, including the ability to account for variability present in the system.
-Furthermore, because each draw from the joint distribution represents a potential outcome of the dynamic system, best/worst case scenarios under the assumptions of the model can be easily obtained via simulation.
-% The added benefits of a stochastic model are coupled with statistical difficulty of fitting the parameter vector $\paramVec$ to observed data.
-
-\eic{MAYBE THE KEY THING HERE IS THAT MODEL 1 IS A ``STANDARD'' POMP MODEL, FOR WHICH MUCH SUCCESSFUL METHODOLOGY IS WIDELY AVAILABLE. THAT IS BASICALLY WHAT WE PROCEED TO SAY.}
-Several algorithms exist, both frequentist and Bayesian, that can be used to obtain estimates of the parameters for this class of models.
+Model~1 describes cholera dynamics at the nationally aggregated scale so that $\bm{Y}^{(1)}_n \in \R$ for each observation time $n \in 1:N$, and the latent state vector $\bm{X}^{(1)}(t)$ is comprised of national level population counts.
+Several algorithms exist, both frequentist and Bayesian, that can be used to obtain estimates of the parameters for the class of models.
 In order to retain the ability to propose models that are scientifically meaningful rather than only those that are simply statistically convenient, we restrict ourselves to parameter estimation techniques that have the plug-and-play property, which is that the fitting procedure only requires the ability to simulate the latent process instead of evaluating transition densities \citep{breto09,he10}.

### 2022-10-04 ms.Rnw
addressed some of Ed's comments, along with some additional edits

@@
 Together, with the density for the initial value of the latent state $\bm{X}^{(\modelCounter)}_{0} = \bm{X}^{(\modelCounter)}(t_0)$, each model assumes a joint density $f_{\bm{X}^{(\modelCounter)}_{0:N}, \bm{Y}^{(\modelCounter)}_{1:N}}\big(\bm{x}^{(\modelCounter)}_{0:N}, \bm{y}^{(\modelCounter)}_{1:N};\paramVec\big)$, where $\paramVec$ is a parameter vector that indexes the model.
-The observed data $\bm{y}_{1:N}^*$ are modeled as a realization of the observation processes.
-\jwc{SHOULD WE SAY SOMETHING ABOUT HOW THE JOINT DENSITY IS INSTRACTABLE, OR CAN BE FACTORED USING THE MARKOV PROPERTY?}
+The observed data $\bm{y}_{1:N}^*$, along with the unobserved true value of the latent state, are modeled as a realization of this joint distribution.
 
@@
 
-We consider model fitting (Sec.~\ref{sec:model_fitting}) followed by diagnostic investigations (Sec.~\ref{sec:model_diagnostics}), forecasting (Sec.~\ref{sec:filter}) and \eic{PERHAPS external scientific corroboration of model fit RATHER THAN interpreting model fit} (Sec.~\ref{sec:science}).
+We consider model fitting (Sec.~\ref{sec:model_fitting}) followed by diagnostic investigations (Sec.~\ref{sec:model_diagnostics}), forecasting (Sec.~\ref{sec:filter}) and external scientific corroboration of model fit (Sec.~\ref{sec:science}).
 
@@
 
-\eic{PERHAPS THIS PARAGRAPH COULD MOVE TO A LATER SECTION?
-THE FOLLOWING PARAGRAPH IS MORE WHAT I EXPECTED READING THROUGH WHEN I REACHED THIS POINT IN THE MS.
-PERHAPS PARTS OF THIS FIRST PARAGRAPH ARE EVEN REDUNDANT, IF THEY ARE COVERED IN THE ORIGINAL DISCUSSION OF THE MODELS.}
-Proposed mechanistic structures form a family of statistical models indexed by a parameter vector $\paramVec$.
+% \eic{PERHAPS THIS PARAGRAPH COULD MOVE TO A LATER SECTION?}
+% The complex nature of biological systems necessitates a search for modeling assumptions that combine insightful simplicity with fidelity to biological reality.
+% For example, many models commonly used in epidemiology are motivated by reasoning about a homogeneous mixing population which is simultaneously an avenue for powerful simplification and a source of model misspecification \citep{bansal07}.
+% Other common considerations include whether the proposed model should be stochastic or deterministic; whether the model should have change points in parameter values or should otherwise make adjustments for changes through time in the dynamic system; and whether the proposed model should include any spatial heterogeneity at a scale permissible by the observed data.
+% While this section is focused on the elements of $\paramVec$ that are calibrated to data, we note that the decision of which elements to fix and which to estimate has consequences for model interpretability, as discussed in Sec.~\ref{sec:science}.
+Each of the three models considered in this study describes cholera dynamics as partially observed Markov processes (POMP), with the understanding that the deterministic Model~2 is a degenerate case of a stochastic model.
+% Despite their similarities, these models represent a diverse selection of possible modeling assumptions: namely, the use of stochastic (Models~1 and 2) or deterministic (Model~2) equations; spatially-heterogeneous meta-population (Models~2 and 3) or spatially-aggregated (Model~1) structure; and the use of covariates (Model~3) versus mathematical equations (Models~1 and 2) to describe a seasonal mechanism.
+As such, each proposed mechanistic structure forms a family of statistical models indexed by a parameter vector $\paramVec$.
 Different values of $\paramVec$ can result in qualitative differences in the predicted behavior of the system.
-The complex nature of biological systems necessitates a search for modeling assumptions that combine insightful simplicity with fidelity to biological reality.
-For example, many models commonly used in epidemiology are motivated by reasoning about a homogeneous mixing population which is simultaneously an avenue for powerful simplification and a source of model misspecification \citep{bansal07}.
-Other common considerations include whether the proposed model should be stochastic or deterministic; whether the model should have change points in parameter values or should otherwise make adjustments for changes through time in the dynamic system; and whether the proposed model should include any spatial heterogeneity at a scale permissible by the observed data.
-In addition, elements of $\paramVec$ can either be chosen as constants, based on scientific reasoning and previous knowledge, or calibrated to observed data.
-Suitable methodology for calibrating model parameters may depend on other modeling decisions.
-% While this section is focused on the elements of $\paramVec$ that are calibrated to data, we note that the decision of which elements to fix and which to estimate has consequences for model interpretability, as discussed in Sec.~\ref{sec:science}.
-
-Each of the three models considered in this study describes cholera dynamics via unobservable states that evolve dynamically with time.
-Despite their similarities, these models represent a diverse selection of possible modeling assumptions: namely, the use of stochastic (Models~1 and 2) or deterministic (Model~2) equations; spatially-heterogeneous meta-population (Models~2 and 3) or spatially-aggregated (Model~1) structure; and the use of covariates (Model~3) versus mathematical equations (Models~1 and 2) to describe a seasonal mechanism.
-All these structures can be described in the framework of partially observed Markov process (POMP) models, with the understanding that the deterministic Model~2 is a degenerate case of a stochastic model.
-In the following subsections we describe our approach to fitting these mechanistic models.
+Therefore, the choice of $\paramVec$ used to make inference about the system can greatly affect model based conclusions.
+Specific elements of $\paramVec$ can be chosen as constants based on scientific understanding of the system, but parameters can also be calibrated to data in a way that maximizes some measure of congruency between the observed data and the assumed mechanistic structure.
+Mechanistic models are used as a way of approximating real-world mechanisms present in a dynamic system.
+While calibrating model parameters does not directly achieve this goal---as model assumptions may be incorrect, and do not change as the model is calibrated to data---the congruency between the model and observed data works as a proxy for the congruency between the model and the true underlying dynamic system, since the observed data are a result of the system.
+% All these structures can be described in the framework of partially observed Markov process (POMP) models, with the understanding that the deterministic Model~2 is a degenerate case of a stochastic model.
+In the following subsections we describe our approach to fitting the three proposed mechanistic models by calibrating $\paramVec$ to observed cholera incidence data.
 
@@
 
-Model~2 is a system of deterministic differential equations and can therefore be considered a degenerate probabilistic model with no randomness in the dynamic process.
-By combining a deterministic process model with a simple Gaussian measurement model, the parameter estimation problem for a set of differential equations reduces to a least squares calculation.
-Deterministic compartment models have a long history in the field of infectious disease epidemiology \citep{kermack1927,brauer2017,varghese21},
+As previously stated, Model~2 is a deterministic compartmental model defined by a set of coupled differential equations.
+The use of deterministic compartment models have a long history in the field of infectious disease epidemiology \citep{kermack1927,brauer2017,varghese21},
 and can be justified by asymptotic considerations in a large-population limit \citep{dadlani2020,ndii17}.
-
+Because the process model of Model~2 is deterministic, the parameter estimation problem for Model~2 reduces to a least squares calculation when combined with a Gaussian measurement model (see supplement material for details).
 \cite{lee20} fit two versions of model~2 based on a presupposed change in cholera transmission from a epidemic phase to endemic phase that occurred in March, 2014.
@@
 The inclusion of a model break-point in \cite{lee20} is perhaps due to a challenging feature of fitting a deterministic model via least squares: discrepancies between model trajectories and observed case counts in highly infectious periods of a disease outbreak will result in greater penalty than the discrepancies between model trajectories and observed case counts in times of relatively low infectiousness, resulting in a bias towards accurately describing periods of high infectiousness.
-This issue is particularly troublesome for this case study: the inability to accurately fit times of low infectiousness may result in poor model forecasts, as few cases of cholera were observed the last few years of the epidemic.
-
-To combat this issue, we fit the model to log-transformed case counts, where the difference between periods of high and low infectiousness is less drastic \eic{, since the log scale stabilizes the variation during periods of high and low infectiousness??
-I'M NOT ENTIRELY HAPPY WITH THAT EITHER}.
+This issue is particularly troublesome for modeling cholera dynamics in Haiti: the inability to accurately fit times of low infectiousness may result in poor model forecasts, as few cases of cholera were observed the last few years of the epidemic.
+
+To combat this issue, we fit the model to log-transformed case counts, where the difference between periods of high and low infectiousness is less drastic, since the log scale stabilizes the variation during periods of high and low infectiousness.
 Other solutions include changing the measurement model to include overdispersion, as was done in Models~1 and~3.
@@
 
+% It has been noted that the uncertainty in just a single parameter can lead to drastically different forecasts \citep{saltelli20};
+% parameter uncertainty should therefore be considered when obtaining model forecasts to influence policy.
+% If a Bayesian technique is used for parameter estimation, the most natural way to account for parameter uncertainty is to obtain $J$ simulations from the model where each simulation is obtained using parameters $\Theta_{1:J}$ drawn from the posterior distribution $\Theta_{1:J} \sim f_{\Theta}\big(\paramVec | \bm{Y}^{(\modelCounter)}_{1:N} = y_{1:N}^*\big)$.
+% In the frequentist case, one approach is empirical Bayes: endow to set $\Theta_K = \{\theta_1, \theta_2, \ldots, \theta_K\}$ with a discrete uniform distribution, such that $P\big(\Theta_{j} = \theta_k\big) = \frac{1}{K}$ for all values $k \in \seq{1}{K}$.
+% In this case, the posterior distribution of $\Theta_{j} | \bm{Y}^{(\modelCounter)}_{1:N}$ can be expressed as:
+% $P\big(\Theta_{j} = \theta_k | \bm{Y}^{(\modelCounter)}_{1:N} = y_{1:N}^*\big) = \frac{l(y_{1:N}^*; \theta_k)}{\sum_{l = 1}^K l(y_{1:N}^*; \theta_l)}$.
+To complete the formulation, we then choose the set $\Theta_K$ such that each $\theta_k \in \Theta_K$ is a single output of the IF2 algorithm, which has been shown to be within a neighborhood of the MLE \cite{ionides15}.
+% This approach has been taken in previous studies \citep{king15}, and the result is simply weighing parameter sets based on their corresponding likelihoods: parameters with higher likelihoods should be used more frequently in model forecasts than those that correspond to lower likelihoods.
+% This can be done for both deterministic and stochastic models, but requires a large number of computations, especially as the number of observations and model parameters increase.
+% Because our direct aim is not to provide forecasts for policy recommendation, we do not include parameter uncertainty in our forecasts.
+% Instead, we use the projections from point estimates to highlight a major deficiency of deterministic models, which is that the only variability in model projections is a result of parameter uncertainty, which leads to over-confidence in forecasts \cite{king15}.
+% To the credit of deterministic models, of the four fitted models in \cite{lee20}, Model~2 provides the most apparently accurate forecasts.
+% This perhaps demonstrates that while deterministic models describe the systems in a less realistic way, the relative ease in fitting these models potentially results in fewer modeling errors.
+% As the data analysis becomes more refined, however, the deficiencies of deterministic models become increasingly apparent.
+
 It has been noted that the uncertainty in just a single parameter can lead to drastically different forecasts \citep{saltelli20};
 parameter uncertainty should therefore be considered when obtaining model forecasts to influence policy.
-If a Bayesian technique is used for parameter estimation, the most natural way to account for parameter uncertainty is to obtain $J$ simulations from the model where each simulation is obtained using parameters $\Theta_{1:J}$ drawn from the posterior distribution $\Theta_{1:J} \sim f_{\Theta}\big(\paramVec | \bm{Y}^{(\modelCounter)}_{1:N} = y_{1:N}^*\big)$.
-In the frequentist case, one approach is empirical Bayes: endow to set $\Theta_K = \{\theta_1, \theta_2, \ldots, \theta_K\}$ with a discrete uniform distribution, such that $P\big(\Theta_{j} = \theta_k\big) = \frac{1}{K}$ for all values $k \in \seq{1}{K}$.
-In this case, the posterior distribution of $\Theta_{j} | \bm{Y}^{(\modelCounter)}_{1:N}$ can be expressed as:
-$P\big(\Theta_{j} = \theta_k | \bm{Y}^{(\modelCounter)}_{1:N} = y_{1:N}^*\big) = \frac{l(y_{1:N}^*; \theta_k)}{\sum_{l = 1}^K l(y_{1:N}^*; \theta_l)}$.
-To complete the formulation, we then choose the set $\Theta_K$ such that each $\theta_k \in \Theta_K$ is a single output of the IF2 algorithm, which has been shown to be within a neighborhood of the MLE \cite{ionides15}.
-This approach has been taken in previous studies \citep{king15}, and the result is simply weighing parameter sets based on their corresponding likelihoods: parameters with higher likelihoods should be used more frequently in model forecasts than those that correspond to lower likelihoods.
-This can be done for both deterministic and stochastic models, but requires a large number of computations, especially as the number of observations and model parameters increase.
+If a Bayesian technique is used for parameter estimation, the most natural way to account for parameter uncertainty is to obtain simulations from the model where each simulation is obtained using parameters drawn from the posterior distribution.
+In the frequentist case, one approach is empirical Bayes, as was done in \cite{king15} (see supplement for details).
+Both of these approaches share the similarity that parameters with higher likelihoods should be used more frequently in model forecasts than those that correspond to lower likelihoods.
+
 Because our direct aim is not to provide forecasts for policy recommendation, we do not include parameter uncertainty in our forecasts.
@@
 This perhaps demonstrates that while deterministic models describe the systems in a less realistic way, the relative ease in fitting these models potentially results in fewer modeling errors.
-As the data analysis becomes more refined, however, the deficiencies of deterministic models become increasingly apparent.
+As the data analysis becomes more refined, however, the deficiencies of deterministic models become increasingly apparent. \jwc{CHANGE THIS BASED ON DISCUSSION WITH ED.}
 
@@
 \begin{supplement}
-\stitle{\code{jesseuwheeler/haiti}}
+\stitle{\code{jeswheel/haiti\_article}}
 \sdescription{

### 2022-10-04 si/si.Rnw
addressed some of Ed's comments, along with some additional edits

@@
 
-%% Comment/remove this line before generating final copy for submission
-%\instructionspage
-
 \date{\today}
@@
 
-\eic{CHECK: EVERYTHING IN SI SHOULD BE REFERENCED IN THE MAIN TEXT.}
-
 \section{Markov chain and differential equation interpretations of compartment flow rates}
@@
 \label{eq:ode1}
-dN_{ij}/dt = \mu_{ij}\big(t,X(t) \big) X_i(t), \medskip i\in \seq{1}{c}, \medskip j\in \seq{1}{c} \cup\{\bullet\}, \medskip i\neq j,
+dN_{ij}/dt = \mu_{ij}\big(t,X(t) \big) X_i(t), \hspace{3mm} i\in \seq{1}{c}, \hspace{3mm} j\in \seq{1}{c} \cup\{\bullet\}, \hspace{3mm} i\neq j,
 \end{equation}
-\eic{medskip IS VERTICAL SPACE. I USUALLY USE hspace}
 where $\mu_{ij}\big(t,X(t)\big)$ is called a per-capita rate or a unit rate.
@@
 The infinitesimal mean of every flow is equal to its infinitesimal variance \citep{breto11} and so an SMCS is called equidispersed.
-\eic{HERE AND ELSEWHERE, in Lee et al SHOULD BE by Lee et al. CONVENTIONALLY, THE TEXT CITATION REFERS TO THE PEOPLE RATHER THAN THE ARTICLE.}
-We note that the special case of Model~1 used in \cite{lee20} (with $\sigmaProc = 0$) is an SMCS.
+We note that the special case of Model~1 used by \cite{lee20} (with $\sigmaProc = 0$) is an SMCS.
 To permit more general mean-variance relationships for a Markov counting system, we must permit jumps of size greater than one.
@@
   \label{model1:meas}
-  y_{t} \mid \Delta N_{E_{\vaccCounter}I_{\vaccCounter}}(t) = z(t) \sim \text{NB}\left(\reportRate z_(t), \obsOverdispersion \right)
+  Y_{n} \mid \Delta N_{E_{\vaccCounter}I_{\vaccCounter}}(n) = z(n) \sim \text{NB}\left(\reportRate z(n), \obsOverdispersion \right),
 \end{equation}
-
-Where $y_t$ is the reported cholera cases at time $t \in 1:N$ and $\Delta N_{E_{\vaccCounter}I_{\vaccCounter}}(t)$ is the number of individuals who moved from compartment $E_\vaccCounter$ to $I_\vaccCounter$ since observation $t-1 \in 0:N-1$.
-\eic{HERE AND ELSEHWERE: PUNCTUATION. NO NEED FOR A NEW PARAGRAPH HERE AFTER THE DISPLAY EQUATION. ALSO, PROBABLY THERE SHOULD BE A COMMA AT THE END OF THE EQUATION AND Where THEN IS NOT CAPITALIZED}
-
-\eic{AS A PEDANTIC POINT WHICH IS NOT UNIVERSALLY IMPLEMENTED BUT I HAPPEN TO THINK IS WORTHWHILE: THE NOTATION IN S16 USUALLY CORRESPONDS TO A DISTRIBUTION, BUT $y_n$ (WHICH I SUPPOSE HERE IS WRITTEN $y_t$) IS DATA, AND THEREFORE A NUMBER NOT A RANDOM VARIABLE.
-WE MODEL IT BY A RANDOM VARIABLE $Y_n$.
-THE DISTINCTION MAY NOT BE CONFUSING FOR MANY, BUT WHEN WE ARE ENCOURAGING OURSELVES AND OTHERS TO THINK ABOUT THE DIFFERENCES BETWEEN DATA AND MODELS FOR DATA, IT IS COUNTER-PRODUCTIVE TO HAVE NOTATION WHICH SUPPRESSES THIS DIFFERENCE.}
-
-\eic{DUE TO DIFFERENT PARAMETERIZATIONS OF NB (AS WELL AS TO CLARIFY THE MODEL) I THINK IT IS GOOD PRACTICE TO GIVE THE MOMENTS, E.G., ``, and $\text{NB}\left(\reportRate z, \obsOverdispersion \right)$ is the negative binomial distribution with mean $\reportRate z$ and variance $\reportRate z+ \psi^2\reportRate^2 z^2)$'' IF THIS IS CORRECT.}
+where $Y_n$ is the reported cholera cases at time $n \in 1:N$ and $\Delta N_{E_{\vaccCounter}I_{\vaccCounter}}(n)$ is the number of individuals who moved from compartment $E_\vaccCounter$ to $I_\vaccCounter$ since observation $n-1 \in 0:N-1$, and $\text{NB}\left(\reportRate z, \obsOverdispersion \right)$ is the negative binomial distribution with mean $\reportRate z$ and variance $\reportRate z+ \psi^2\reportRate^2 z^2)$.
+\jwc{CHECK THIS}.
 
@@
   \label{model2:meas}
-  \log\big(y_{u, t} + 1\big) \mid \bm{X}^{(2)}(t) = \bm{x}^{(2)}(t) \sim \text{N}\left(\log \big(\reportRate \sum_{\vaccCounter} I_{u, \vaccCounter}(t) + 1\big), \obsOverdispersion^2 \right)
+  \log\big(Y_{u, n} + 1\big) \mid \bm{X}^{(2)}(n) = \bm{x}^{(2)}(n) \sim \text{N}\left(\log \big(\reportRate \sum_{\vaccCounter} I_{u, \vaccCounter}(n) + 1\big), \obsOverdispersion^2 \right)
 \end{equation}
 
-Where $\sum_{\vaccCounter} I_{u, \vaccCounter}(t)$ is the total number of individuals who are infected at time $t$ in unit $u \in 1:U$, for all vaccination campaigns $\vaccCounter$.
-We note that fitting a model with this loss function is equivalent to fitting using least squares, with $\log(y_{u, t} + 1)$ as the response variable.
+Where $\sum_{\vaccCounter} I_{u, \vaccCounter}(n)$ is the total number of individuals who are infected at time $n$ in unit $u \in 1:U$, for all vaccination campaigns $\vaccCounter$.
+We note that fitting a model with this loss function is equivalent to fitting using least squares, with $\log(y_{u, n} + 1)$ as the response variable.
 
@@
   \label{model3:meas}
-  y_{u, t} \mid \Delta N_{S_{u \vaccCounter}I_{u \vaccCounter}}(t) = z_{u}(t) \sim \text{NB}\left(\reportRate z_{u}(t), \obsOverdispersion \right)
+  Y_{u, n} \mid \Delta N_{S_{u \vaccCounter}I_{u \vaccCounter}}(t) = z_{u}(n) \sim \text{NB}\left(\reportRate z_{u}(n), \obsOverdispersion \right)
 \end{equation}
 
-Where $\Delta N_{S_{u \vaccCounter}I_{u \vaccCounter}}(t)$ is the number of individuals who moved from compartment $S_{u \vaccCounter}$ to $I_{u \vaccCounter}$ since observation $t-1 \in 0:N-1$.
+Where $\Delta N_{S_{u \vaccCounter}I_{u \vaccCounter}}(n)$ is the number of individuals who moved from compartment $S_{u \vaccCounter}$ to $I_{u \vaccCounter}$ since observation $n-1 \in 0:N-1$.
 
@@
 unit are being considered.
-In the case where there are many units, we believe that this modification will lead to less particle depletion.
-\eic{USUALLY BEST TO AVOID believe IN SCIENTIFIC WRITING.
-CAN WE SAY SOMETHING SOLID ABOUT WHAT WE HAVE FOUND.
-EVEN, I THINK THE FOLLOWING IS TRUE: We have found that this modification leads to improved likelihood maximization performance, especially as the number of units increases.}
+% In the case where there are many units, we believe that this modification will lead to less particle depletion.
+We have found that this modification leads to improved likelihood maximization performance, especially as the number of units increases.
 No theoretical properties of this algorithm have yet been derived, but the approach has similarities to the block particle filter of \citet{rebeschini15} and its extension to parameter estimation \citep{ionides22,ning21ibpf}.

### 2022-10-08 ms.Rnw
added measurement and ivp parameters

@@
 \newcommand\transmissionTrend{\zeta}
+\newcommand\Binit{\xi}
 \newcommand\vaccClass{Z}
@@
 
-In the following subsections, complete descriptions of Models~1--3 are provided \eic{WE NEED INITIAL VALUES AND A MEASUREMENT MODEL TO COMPLETE THE DESCRIPTION.
-MAYBE WE JUST SAY WE'RE DESCRIBING THE DYNAMICS HERE, OR MAYBE WE ADD THE MISSING STUFF, WHICH HAS TO BE SOMEWHERE ANYHOW.
-SOME OF THIS IS DONE IN SEC 3.1.1 FOR MODEL 1, BUT PERHAPS BETTER TO DO IT SYSTEMATICALLY WHEN THE MODEL IS ORIGINALLY DEFINED.}.
+In the following subsections, descriptions of Models~1--3 are provided.
 While the model description is scientifically critical, as well as necessary for transparency and reproducibility, the model details are not essential to our methodological discussions of how to diagnose and address model misspecification with the purpose of informing policy.
@@
 For $\vaccCounter\in\seq{1}{\vaccClass}$, the vaccination cohort $\vaccCounter$ is assigned a time $\tau_{\vaccCounter}$, and we take $\symptomFrac_{\vaccCounter}(t) = c \, \theta^*(t-\tau_{\vaccCounter})$
-% \begin{equation}
-% \label{model1:theta_{\vaccCounter}}
-% \symptomFrac_{\vaccCounter}(t) = c \, \theta^*(t-\tau_{\vaccCounter})
-% \end{equation}
-% \arc{I think we could put this in Table 2 and replace this chunk with a sentence or two defining $\symptomFrac_0(t)$ in words as the proportion of exposed individuals who develop into symptomatic cases and adding a reference to Table 2.
-This would save a good quarter of a page or so.}
 where $\theta^*(t)$ is efficacy at time $t$ since vaccination for adults, taken from \citet{lee20}, Table~S4, and $c=\big(1-(1-0.4688)\times 0.11\big)$ is a correction to allow for reduced efficacy in the 11\% of the population aged under 5 years.
 Single and double vaccine doses were modeled by changing the waning of protection; protection was assumed to be equal between single and double dose until 52 weeks after vaccination, at which point the single dose becomes ineffective.
+The latent state vector $\bm{X}^{(1)}(t)$ is initialized by setting each compartment count for each vaccination scenario $\vaccCounter \neq 0$ as zero, and introducing initial-value parameters $I_{0,0}$ and $E_{0, 0}$ such that $R_{0}(0) = 0$, $I_{0}(0) = \mathrm{Pop} \times I_{0, 0}$, $E_{0}(0) = \mathrm{Pop} \times E_{0, 0}$ and $S_{0}(0) = \mathrm{Pop} \times (1 - I_{0, 0} - E_{0, 0})$, where $\mathrm{Pop}$ is the total population of Haiti.
+Reported cholera cases at time point $n$ ($\bm{Y}^{(1)}_n$) are assumed to come from a negative binomial measurement model, where only a fraction ($\reportRate$) of new weekly cases are reported.
+See Sec.~S4 of the supplement material for more details.
 
@@
 The unit of $W_u(t)$ is cells per ml, with dose response modeled via a saturation constant of $\Wsat$ in \eqref{model2:lambda}.
+The starting value for each element of the latent state vector $\bm{X}^{(2)}(0)$ are set to zero except for $I_{u0}(0) = y_u^*(0)/\reportRate$ and $R_{u0}(0) = \mathrm{Pop}_u - I_{u0}(0)$, where $y_u^*(0)$ is the reported number of cholera cases in department $u$ at time $t = 0$.
+Reported cases are assumed to come from a log-normal distribution, with the log-scale mean equal to the reporting rate $\reportRate$ times the number of newly infected individuals.
+See Sec.~\ref{sec:fit2} and Sec.~S4.2 of the supplement for more details.
 
@@
 \label{eq:model3:water}
-\mu_{{\demography}W_u} &=& \big[1 + \seasAmplitude \big(J(t))^r \big] D_i \, \Wshed \big[ I_{u0}(t)+ \asymptomRelativeShed A_{u0}(t) \big]
+\mu_{{\demography}W_u} &=& \big[1 + \seasAmplitude \big(J_u(t))^r \big] \mathrm{Den}_u \, \Wshed \big[ I_{u0}(t)+ \asymptomRelativeShed A_{u0}(t) \big]
 \\[-3pt]
@@
 As with Model~1, $d\Gamma_u(t)/dt$ is multiplicative Gamma-distributed white noise in \myeqref{eq:model3:SI} and \myeqref{eq:model3:SA}.
-In \eqref{eq:model3:water}, $J_u(t)$ is a dimensionless measurement of precipitation that has been standardized by dividing the observed rainfall at time $t$ by the maximum recorded rainfall in department $u$ during the epidemic, and $D_u$ is the average population density.
+In \eqref{eq:model3:water}, $J_u(t)$ is a dimensionless measurement of precipitation that has been standardized by dividing the observed rainfall at time $t$ by the maximum recorded rainfall in department $u$ during the epidemic, and $\mathrm{Den}_u$ is the population density.
 Demographic stochasticity is accounted for by modeling non-cholera related death rate $\muDeath$ in each compartment, along with an additional death rate $\choleraDeath$ in \myeqref{eq:model3:IS} to account for cholera induced deaths among infected individuals.
 All deaths are balanced by births into the susceptible compartment in \myeqref{eq:model3:AS} and \myeqref{eq:model3:RS}, thereby maintaining constant population in each department.
+
+Latent states are initialized using an approximation of the instantaneous number of infected, asymptomatic, and recovered individuals at time $t=0$ by using the first week of cholera incidence data.
+Specifically, we set $I_{u0}(0) = \frac{y^*_{u0}}{\reportRate(\muDeath + \choleraDeath + \muIR)}$, $A_{u0}(0) = \frac{1 - \symptomFrac}{\symptomFrac}I_{u0}(0)$, and $R_{u0k} = y^*_{u0} - I_{u0}(0) - A_{u0}(0)$, and all other compartments that represent population counts are set to zero at time $t=0$.
+We replace $J_u(0)$ in \eqref{eq:model3:water} with initial value parameters $\Binit_u$ to allow for some flexibility in initializing the bacterial compartment $W_{u}(0)$ while still maintaining the primary dynamics of the model.
+Reported cholera cases are assumed to come from a negative binomial measurement model with mean equal to a fraction ($\reportRate$) of individuals in each unit who develop symptoms seek healthcare (see Sec.~S4 of the supplement material for more details).
 
@@
 
-\subsubsection{Model~2}
+\subsubsection{Model~2}\label{sec:fit2}
 
@@
 Both the latent and observable processes in Model~3 can be be factored into department specific processes which interact with each other.
-This decision to address metapopulation dynamics via a spatially explicit model, rather than to aggregate over space, is double-edged.
+The decision to address metapopulation dynamics via a spatially explicit model, rather than to aggregate over space, is double-edged.
 Evidence for the former approach has been provided in previous studies \citep{king15}, including the specific case of heterogeneity between Haitian departments in cholera transmission \citep{collins14}.

### 2022-10-10 ms.Rnw
Changed M3 -> IBPF fitting

@@
 
-Parameters that must be fit in Model~3 are shared between each department, the exception to this being the parameters $\beta_{W_u}$, and $\beta_u$, which take department-specific values for each $u \in \seq{1}{10}$.
+Model~3 parameters that are calibrated to data are primarily shared between each department, meaning that the estimated parameter value for department $u$ is the same value as the parameter value in department $v$, with $u, v \in \seq{1}{U}$.
+The exception to this being the parameters $\beta_{W_u}$, and $\beta_u$, which take department-specific values for each department $u$.
 To avoid the parameter estimation issue in high-dimensional models, \citet{lee20} simplified the problem by using the IF2 algorithm to separately estimate parameters for independent department-level models;
@@
 Reducing a spatially coupled model to individual units in this fashion requires special treatment of any interactive mechanisms between spatial units, such as found in Eq.~\myeqref{eq:model3:foi}.
-In particular, when considering a model for department $u$, the values $I_\nu(t)$ and $A_\nu(t)$, are unknown for $u \neq v \in \seq{1}{10}$.
-An approximation of Eq.~\myeqref{eq:model3:foi} for each department $v \in \seq{1}{10}$ can be obtained using the weekly number of observed cholera cases in each department:
-\begin{equation}
-  I_v(t) + A_v(t) \approx \frac{365}{7\reportRate}y^*_v(t)\left(\frac{1}{\muDeath + \choleraDeath + \muIR} + \frac{1 - \symptomFrac}{\symptomFrac(\muDeath + \muIR)}\right)\label{eq:model3:approx}
-\end{equation}
-This approximation leads to department-specific models that are conditionally independent given the reported number of cholera infections in the remaining departments.
-Here, we refer to a collection of POMP models that are independent across units as a PanelPOMP.
-
-In the case of a PanelPOMP, an extension of the IF2 algorithm, known as Panel Iterated Filtering (PIF) \citep{breto20}, can be used to obtain the MLE, which solves the curse of dimensionality for this class of models \eic{WHICH CONVENTIONALLY REFERS TO THE PRECEDING OBJECT, HERE THE MLE, WHICH IS NOT WHAT WE WANT}.
-A major advantage of this algorithm is that PIF can be used to fit both unit-specific parameters and shared parameters; in this way, the calibration of shared parameters involves all of the available data instead of just an arbitrarily chosen subset.
-
-\eic{READERS MAY BE CONFUSED HERE. IS THIS A NEW METHOD? IT SEEMS TO BE IMPLEMENTED BY BRETO ET AL IN 2020. THIS IS DELICATE.}
-We propose and use a slight modification of the PIF algorithm, which we call the Block Panel Iterated Filter (BPIF), to estimate parameters for a PanelPOMP version of Model~3.
-Let $l_{\text{P}}(y_{1:N}^*; \paramVec)$ denote the log-likelihood function of the PanelPOMP version of Model~3.
-The output of the BPIF algorithm is the MLE $\tilde{\paramVec} = \argmax_{\paramVec} l_{\text{P}}(y_{1:N}^*; \paramVec)$.
-The BPIF algorithm is implemented in the \code{panelPomp} package \citep{breto20panelPomp} as the argument \code{block = TRUE} in the \code{mif2} function \eic{THIS LEVEL OF CODING DETAIL IS UNUSUAL IN A CASE STUDY}.
-Pseudo-code for this algorithm is provided in Algorithm~S1 in the supplementary material.
+In particular, when considering a model for department $u$, the values $I_\nu(t)$ and $A_\nu(t)$, are unknown for $u \neq v$.
+To address this issue, \cite{lee20} conditioned each department level model on reported cholera cases in other departments and approximated the values of $I_\nu(t)$ and $A_\nu(t)$ using the respective reported cases.
+
+% Fitting Model~3 in this fashion simplifies the problem of parameter estimation, but it also introduces additional technicalities that must be addressed.
+% One concern is that of obtaining model forecasts, which was the primary goal of \citet{lee20}.
+The simplified, spatially-decoupled version of Model~3 relies on the observed cholera cases for each department to describe cholera dynamics, making it impossible to obtain forecasts for future time points from the fitted model.
+In order to obtain model forecasts, the parameters fit for the conditionally independent department level models were used in the fully coupled version of Model~3.
+This approach of model calibration and forecasting avoids the issue of particle depletion by simplifying the parameter estimation problem, but it may result in some undesirable outcomes.
+One concern is that cholera dynamics in department $u$ are highly related to the dynamics in the remaining departments;
+calibrating model parameters using observed cases in other departments as a model covariate may therefore lead to an over-dependence on observed cholera cases.
+Another concern is that the two versions of the model are not the same, resulting in sub-optimal parameter estimates for the spatially coupled model.
+This is because the likelihood of the coupled model at its maximum likelihood estimate (MLE) is greater than or equal to the likelihood of the coupled model at the MLE of the decoupled model.
+These two concerns may explain the unrealistic forecasts and low-likelihood of the model (see Table~\ref{tab:likes}).
+
+For our analysis, we calibrate the parameters of the spatially coupled version of Model~3 using the novel iterated block particle filter (IBPF) algorithm of \cite{ionides22}.
+This algorithm extends the work of \cite{ning21ibpf}, who provided theoretic justification for the version of the algorithm that only estimates unit-specific parameters.
+While computationally intensive, the IBPF algorithm enables us to directly estimate the parameters of models describing high-dimensional partially-observed nonlinear dynamic systems via likelihood maximization.
+The ability to directly estimate parameters of Model~3 is responsible for the large increase in model likelihoods reported in Table~\ref{tab:likes}, as no algorithm with similar capabilities existed when \cite{lee20} published their results.
+Simulations from the fitted model are displayed in Fig~.\ref{fig:h3spatsims}.
+
+% An approximation of Eq.~\myeqref{eq:model3:foi} for each department $v \in \seq{1}{10}$ can be obtained using the weekly number of observed cholera cases in each department:
+% \begin{equation}
+%   I_v(t) + A_v(t) \approx \frac{365}{7\reportRate}y^*_v(t)\left(\frac{1}{\muDeath + \choleraDeath + \muIR} + \frac{1 - \symptomFrac}{\symptomFrac(\muDeath + \muIR)}\right)\label{eq:model3:approx}
+% \end{equation}
+% This approximation leads to department-specific models that are conditionally independent given the reported number of cholera infections in the remaining departments.
+% Here, we refer to a collection of POMP models that are independent across units as a PanelPOMP.
+%
+% In the case of a PanelPOMP, an extension of the IF2 algorithm, known as Panel Iterated Filtering (PIF) \citep{breto20}, can be used to obtain the MLE, which solves the curse of dimensionality for this class of models \eic{WHICH CONVENTIONALLY REFERS TO THE PRECEDING OBJECT, HERE THE MLE, WHICH IS NOT WHAT WE WANT}.
+% A major advantage of this algorithm is that PIF can be used to fit both unit-specific parameters and shared parameters; in this way, the calibration of shared parameters involves all of the available data instead of just an arbitrarily chosen subset.
+%
+% \eic{READERS MAY BE CONFUSED HERE. IS THIS A NEW METHOD? IT SEEMS TO BE IMPLEMENTED BY BRETO ET AL IN 2020. THIS IS DELICATE.}
+% We propose and use a slight modification of the PIF algorithm, which we call the Block Panel Iterated Filter (BPIF), to estimate parameters for a PanelPOMP version of Model~3.
+% Let $l_{\text{P}}(y_{1:N}^*; \paramVec)$ denote the log-likelihood function of the PanelPOMP version of Model~3.
+% The output of the BPIF algorithm is the MLE $\tilde{\paramVec} = \argmax_{\paramVec} l_{\text{P}}(y_{1:N}^*; \paramVec)$.
+% The BPIF algorithm is implemented in the \code{panelPomp} package \citep{breto20panelPomp} as the argument \code{block = TRUE} in the \code{mif2} function \eic{THIS LEVEL OF CODING DETAIL IS UNUSUAL IN A CASE STUDY}.
+% Pseudo-code for this algorithm is provided in Algorithm~S1 in the supplementary material.
 
@@
 
-Fitting Model~3 in this fashion simplifies the problem of parameter estimation, but it also introduces additional technicalities that must be addressed \eic{THIS MAKES IT SEEM LIKE THE FITTING METHOD CAUSES THE PROBLEM, BUT THE LEE ET AL APPROACH ALSO HAS THIS ISSUE.
-ARGUABLY, IT IS A PART OF THE MODEL NOT THE METHOD, THOUGH THE TWO ARE RELATED.
-MODEL 3 IS WRITTEN AS FULLY COUPLED IN SEC 2.3, SO FITTING THE NON-COUPLED VERSION IS ARGUABLY FITTING SOMETHING ELSE ENTIRELY, THOUGH IT CAN ALWAYS BE VIEWED AS AN APPROXIMATION.}
-One concern is that of obtaining model forecasts, which was the primary goal of \citet{lee20}.
-The simplified PanelPOMP version of Model~3 relies on the observed cholera cases as a covariate, which are unavailable for use in forecasts.
-Similar to what was done in \cite{lee20}, we use the parameter estimates obtained with the independent departmental approximation of Model~3 as an estimate for the parameters in the fully coupled version of the model, which was implemented using the \code{spatPomp} package.
-Simulations from the fully coupled version of the model using the estimated parameter vector $\tilde{\paramVec}$, are displayed in Fig.~\ref{h3spatsims}.
+% Fitting Model~3 in this fashion simplifies the problem of parameter estimation, but it also introduces additional technicalities that must be addressed.
+% One concern is that of obtaining model forecasts, which was the primary goal of \citet{lee20}.
+% The simplified PanelPOMP version of Model~3 relies on the observed cholera cases as a covariate, which are unavailable for use in forecasts.
+% Similar to what was done in \cite{lee20}, we use the parameter estimates obtained with the independent departmental approximation of Model~3 as an estimate for the parameters in the fully coupled version of the model, which was implemented using the \code{spatPomp} package.
+% Simulations from the fully coupled version of the model using the estimated parameter vector $\tilde{\paramVec}$, are displayed in Fig.~\ref{h3spatsims}.
 
@@
 % While simulations from the SpatPOMP version of Model~3 resemble the observed data, we note in Table~\ref{tab:likes} that the SpatPOMP version of Model~3 has a lower log-likelihood than the PanelPOMP version using the same model parameters.
-Let $l_{\text{C}}(y_{1:N}^*; \paramVec)$ denote the log-likelihood of the fully coupled model, and let $\hat{\paramVec} = \argmax_{\paramVec} l_{\text{C}}(y_{1:N}^*; \paramVec)$.
-If the goal of parameter estimation is to maximize the likelihood of the fully coupled model, the approach of estimating parameters using a decoupled model is clearly sub optimal, as $l_{\text{C}}(y_{1:N}^*; \hat{\paramVec}) \geq l_{\text{C}}(y_{1:N}^*; \tilde{\paramVec})$.
-A natural question that arises is if the difference between $\hat{\paramVec}$ and $\tilde{\paramVec}$ is scientifically and statistically meaningful.
-The answer to this question requires the ability to reliably estimate $\hat{\paramVec}$, the answer to which is model specific and beyond the scope of this case study.
-
-A brief analysis of the fitted parameters $\tilde{\paramVec}$ provides some interesting insight.
-Consider the large estimated exponent on rainfall: $r = \Sexpr{signif(p3s["r"],3)}$ (Eq.~\ref{eq:model3:water}).
-Because the value of rainfall $J(t) \in (0, 1)$, the large coefficient results in $J(t)^r \approx 0$.
-This result may initially be surprising given the importance of rainfall as a seasonal diver of cholera infections in Haiti (see Sec.~\ref{sec:science}).
-One plausible explanation of this result is that the fitting procedure---used to obtain parameter values in the independent department model given the reported cases in other departments---has learned an over-dependence on the reported cholera cases in other departments, and thus the seasonality of cholera incidence in department $u$ depends on the highly correlated covariate of the reported cholera cases in departments $1:U\setminus \{u\}$.
-Whatever the case may be, this fitting procedure results in a model that is unable to replicate the observed number of cases by simulating from initial conditions (see Fig.~\ref{h3spatsims}), and highlights the need for advancements in statistical methodology that permit inference on models with coupled metapopulation dynamics.
+% Let $l_{\text{C}}(y_{1:N}^*; \paramVec)$ denote the log-likelihood of the fully coupled model, and let $\hat{\paramVec} = \argmax_{\paramVec} l_{\text{C}}(y_{1:N}^*; \paramVec)$.
+% If the goal of parameter estimation is to maximize the likelihood of the fully coupled model, the approach of estimating parameters using a decoupled model is clearly sub optimal, as $l_{\text{C}}(y_{1:N}^*; \hat{\paramVec}) \geq l_{\text{C}}(y_{1:N}^*; \tilde{\paramVec})$.
+% A natural question that arises is if the difference between $\hat{\paramVec}$ and $\tilde{\paramVec}$ is scientifically and statistically meaningful.
+% The answer to this question requires the ability to reliably estimate $\hat{\paramVec}$, the answer to which is model specific and beyond the scope of this case study.
+
+% A brief analysis of the fitted parameters $\tilde{\paramVec}$ provides some interesting insight.
+% Consider the large estimated exponent on rainfall: $r = \Sexpr{signif(p3s["r"],3)}$ (Eq.~\ref{eq:model3:water}).
+% Because the value of rainfall $J(t) \in (0, 1)$, the large coefficient results in $J(t)^r \approx 0$.
+% This result may initially be surprising given the importance of rainfall as a seasonal diver of cholera infections in Haiti (see Sec.~\ref{sec:science}).
+% One plausible explanation of this result is that the fitting procedure---used to obtain parameter values in the independent department model given the reported cases in other departments---has learned an over-dependence on the reported cholera cases in other departments, and thus the seasonality of cholera incidence in department $u$ depends on the highly correlated covariate of the reported cholera cases in departments $1:U\setminus \{u\}$.
+% Whatever the case may be, this fitting procedure results in a model that is unable to replicate the observed number of cases by simulating from initial conditions (see Fig.~\ref{h3spatsims}), and highlights the need for advancements in statistical methodology that permit inference on models with coupled metapopulation dynamics.
 
@@
 Instead, we use the projections from point estimates to highlight a major deficiency of deterministic models, which is that the only variability in model projections is a result of parameter uncertainty, which leads to over-confidence in forecasts \cite{king15}.
-To the credit of deterministic models, of the four fitted models in \cite{lee20}, Model~2 provides the most apparently accurate forecasts.
-This perhaps demonstrates that while deterministic models describe the systems in a less realistic way, the relative ease in fitting these models potentially results in fewer modeling errors.
-As the data analysis becomes more refined, however, the deficiencies of deterministic models become increasingly apparent. \jwc{CHANGE THIS BASED ON DISCUSSION WITH ED.}
+% To the credit of deterministic models, of the four fitted models in \cite{lee20}, Model~2 provides the most apparently accurate forecasts.
+% This perhaps demonstrates that while deterministic models describe the systems in a less realistic way, the relative ease in fitting these models potentially results in fewer modeling errors.
+% As the data analysis becomes more refined, however, the deficiencies of deterministic models become increasingly apparent. \jwc{CHANGE THIS BASED ON DISCUSSION WITH ED.}
 
@@
 
-We acknowledge there are limitations to this study, such as the inability to fit model parameters to the fully-coupled version of Model~3.
-Promising theoretical and methodological developments \citep{ning21ibpf,ionides22} based on the Block Particle Filter \citep{rebeschini15} may potentially be used to obtain better parameter estimates for Model~3 in future work.
+% We acknowledge there are limitations to this study, such as the inability to fit model parameters to the fully-coupled version of Model~3.
+% Promising theoretical and methodological developments \citep{ning21ibpf,ionides22} based on the Block Particle Filter \citep{rebeschini15} may potentially be used to obtain better parameter estimates for Model~3 in future work.
 

### 2022-10-10 si/si.Rnw
Updated measurement / rinit models

@@
 \begin{equation}
-  B_u(0) = \big[1 + \seasAmplitude \big(\xi_u)^r \big] D_i \, \Wshed \big[ I_{u}(0)+ \asymptomRelativeShed A_{u}(0) \big]\label{eq:model3:b0},
-\end{equation}
-where $\xi_u \in (0, 1)$ is an initial value parameter that allows some flexibility in determining the initial state of the bacteria compartment.
+  B_u(0) = \big[1 + \seasAmplitude \big(\Binit_u)^r \big] D_i \, \Wshed \big[ I_{u}(0)+ \asymptomRelativeShed A_{u}(0) \big]\label{eq:model3:b0},
+\end{equation}
+where $\Binit_u \in (0, 1)$ is an initial value parameter that allows some flexibility in determining the initial state of the bacteria compartment.
 
@@
   \label{model1:meas}
-  Y_{n} \mid \Delta N_{E_{\vaccCounter}I_{\vaccCounter}}(n) = z(n) \sim \text{NB}\left(\reportRate z(n), \obsOverdispersion \right),
-\end{equation}
-where $Y_n$ is the reported cholera cases at time $n \in 1:N$ and $\Delta N_{E_{\vaccCounter}I_{\vaccCounter}}(n)$ is the number of individuals who moved from compartment $E_\vaccCounter$ to $I_\vaccCounter$ since observation $n-1 \in 0:N-1$, and $\text{NB}\left(\reportRate z, \obsOverdispersion \right)$ is the negative binomial distribution with mean $\reportRate z$ and variance $\reportRate z+ \psi^2\reportRate^2 z^2)$.
+  Y_{n} \mid \Delta N_{EI}(n) = z(n) \sim \text{NB}\left(\reportRate z(n), \obsOverdispersion \right),
+\end{equation}
+where $Y_n$ is the reported cholera cases at time $n \in 1:N$ and $\Delta N_{EI}(n)$ is the sum total of individuals across vaccination compartment $\vaccCounter$ who moved from compartment $E_\vaccCounter$ to $I_\vaccCounter$ since observation $n-1 \in 0:N-1$, and $\text{NB}\left(\reportRate z, \obsOverdispersion \right)$ is the negative binomial distribution with mean $\reportRate z$ and variance $\reportRate z+ \psi^2\reportRate^2 z^2)$.
 \jwc{CHECK THIS}.
 
-\subsection{Model~2}
-
-As mentioned in the main text, Model~2 was fit using log-transformed observed cases.
+\subsection{Model~2}\label{sec:mod2Meas}
+
+As mentioned in the main text, Model~2 was fit using reported case counts that were transformed using the natural logarithm.
 We fit Model~2 using the subplex algorithm in the \code{subplex} package, using
-a Gaussian measurement model (Eq.~\myeqref{model2:meas}) on the log-transformed cases as the loss function.
-
+a Gaussian measurement model (Eq.~\myeqref{model2:meas}) on the log transformed cases as the loss function.
 \begin{equation}
   \label{model2:meas}
-  \log\big(Y_{u, n} + 1\big) \mid \bm{X}^{(2)}(n) = \bm{x}^{(2)}(n) \sim \text{N}\left(\log \big(\reportRate \sum_{\vaccCounter} I_{u, \vaccCounter}(n) + 1\big), \obsOverdispersion^2 \right)
-\end{equation}
-
-Where $\sum_{\vaccCounter} I_{u, \vaccCounter}(n)$ is the total number of individuals who are infected at time $n$ in unit $u \in 1:U$, for all vaccination campaigns $\vaccCounter$.
-We note that fitting a model with this loss function is equivalent to fitting using least squares, with $\log(y_{u, n} + 1)$ as the response variable.
+  \log\big(Y_{u, n} + 1\big) \mid \Delta N_{EI}(n) = z(n) \sim \text{N}\left(\log \big(\reportRate z(n) + 1\big), \obsOverdispersion^2 \right),
+\end{equation}
+where $\Delta N_{EI}(n)$ is the sum total of individuals across vaccination compartment $\vaccCounter$ and unit $u$ who moved from compartment $E_{u\vaccCounter}$ to $I_{u\vaccCounter}$ since observation $n-1 \in 0:N-1$.
+Therefore, because the natural logarithm of observed case counts (plus one, to avoid taking the logarithm of zero) has a normal distribution, $Y_{u, n} + 1$ is assumed to follow a log-normal distribution with log-mean parameter $\mu = \log\big(\Delta N_{EI}(n) + 1\big)$ and log-variance $\obsOverdispersion^2$.
+We note that fitting a model with this measurement model is equivalent to fitting using least squares, with $\log(Y_{u, n} + 1)$ as the response variable.
+
+This measurement model differs from that used by \citet{lee20}, who fit the model in two stages: epidemic and endemic phases.
+Although their text and supplement material do not explicitly describe the measurement model used, inspection of the code provided with their submission suggests a change in measurement model between the epidemic and endemic phases.
+In the file \code{choleraEqsPublic.py}, \citet{lee20} create several functions, where each function represents a set of coupled differential equations that could potentially be used to model cholera incidence data.
+Each function returns a vector (or \code{numpy} array) that represents the change in each state variable for a single time step, including the variable \code{dC}, which tracks the number of new infections and is used to obtain the reported case counts.
+Following their comments and code, it appears that the function \code{choleraEqs10WithoutVaccinationNetwork} was used to describe the dynamics of the epidemic phase, and \code{choleraEqs11WithoutVaccinationNetwork} was used for the endemic stage (see Sec~\ref{sec:mod2rep} for more details).
+Because their models were fit using least squares, the code in these functions suggest that the measurement model for the epidemic phase is
+\begin{equation}
+  \label{model2:meas}
+  Y_{u, n} \mid \Delta N_{EI}(n) = z(n) \sim \text{N}\left(\reportRate z(n), \obsOverdispersion^2 \right),
+\end{equation}
+which is similar to our measurement model, the primary difference being that the measurement model is applied to raw case counts rather than log-transformed case counts.
+The measurement model for the endemic phase modifies the epidemic model by counting both asymptomatically infected (A) and symptomatically infected (I) individuals in the case counts:
+\begin{equation}
+  \label{model2:meas}
+  Y_{u, n} \mid \Delta N_{EI}(n) = z_1(n), \Delta N_{EA}(n) = z_2(n) \sim \text{N}\left(\reportRate (z_1(n) + z_2), \obsOverdispersion^2 \right),
+\end{equation}
+where the notation for $\Delta N_{EA}(n)$ is similar to $\Delta N_{EI}(n)$, described above.
 
@@
 
-\subsection{Model~2 Replication}
+\subsection{Model~2 Replication}\label{sec:mod2rep}
 
 Model~2 was developed by a team that consisted of members from the Fred Hutchinson Cancer Research Center and the University of Florida (hereafter referred to as the Model~2 authors).
-While Model~2 is the only deterministic model we considered in our analysis, it contains perhaps the most complex descriptions of cholera in Haiti: Model~2 accounts for movement between spatial units; human-to-human and environment-to-human cholera infections; and transfer of water between spatial units based on elevations charts and river flows.
-
-The source code that was used to implement Model~2 in \cite{lee20} was written in the \code{Python} programming language, and is publicly available at \url{10.5281/zenodo.3360857} and its accompanying GitHub repository \url{https://github.com/lulelita/HaitiCholeraMultiModelingProject}.
+While Model~2 is the only deterministic model we considered in our analysis, it contains perhaps the most complex descriptions of cholera in Haiti: Model~2 accounts for movement between spatial units; human-to-human and environment-to-human cholera infections; and transfer of water between spatial units based on elevation charts and river flows.
+
+The source code that the Model~2 authors used to generate their results was written in the \code{Python} programming language, and is publicly available at \url{10.5281/zenodo.3360857} and its accompanying GitHub repository \url{https://github.com/lulelita/HaitiCholeraMultiModelingProject}.
 In order to perform our analysis in a unified framework, we re-implemented this model in the \code{R} programming language using the \code{spatPomp} package \citep{asfaw21github}, which facilitates the creation of meta-population models.
@@
 instead, we obtained these matrices via personal correspondence with the Model~2 authors.
-
 Using these matrices, and the point estimates for model parameters listed in \citep{lee20sup}, we created trajectories of the cholera dynamics and compared this to available data.
-These trajectories, shown in Figure~\ref{fig:mod2rep}, are very similar to the trajectories shown in Figure~S15 of \cite{lee20sup}, suggesting that our re-implementation of Model~2 in the \code{haitipkg} is a faithful approximation to the original model used in \cite{lee20}.
+These trajectories, shown in Figure~\ref{fig:mod2rep}, are very similar to the trajectories shown in Figure~S15 of \cite{lee20sup}.
 
@@
 
-We do note, however, several discrepancies exit between Figure~\ref{fig:mod2rep} and Figure~S15 of \cite{lee20sup}, such as a delayed second wave in the epidemic period and a larger initial wave during the endemic period.
-While these are only minor discrepancies, because Model~2 is deterministic one may expect to be able to obtain perfect replication in model trajectories.
-We believe these discrepancies are a result of slight difference in model parameters.
-For example, we note that the parameters $\transmission$, $\Wbeta{}$ (See Table~\ref{tab:translate}) had reported values of $9.9 \times 10^{-7}$ and $4.03 \times 10^{-2}$, respectively (Table~S13 of \cite{lee20sup}), but were actually fit to data and therefore likely had values that were far more precise.
+There are minor differences between Figure~\ref{fig:mod2rep} and Figure~S15 of \cite{lee20sup}.
+% such as a delayed second wave in the epidemic period and a larger initial wave during the endemic period.
+While the discrepancy appears minor, the deterministic nature of Model~2 implies that an exact replication of model trajectories should be possible.
+In this case, these discrepancies may be attributed to model implementation and small differences in model parameters.
+Details about the measurement models and how latent states were initialized for the epidemic model were not provided by \citet{lee20sup} and therefore these details must be inferred by looking at the provided source code.
+According to repository comments, the files \code{fitInPieces3paramsCleanMay2019Public.py} and \code{fitInPiecesMuWithFracSusFixedAllInfectionsPublic.py} were used to fit the epidemic and endemic phases of the model respectively, although it is apparent that these exact files were not used to obtain the reported results.
+These files contain errors that making them impossible to run, and therefore these files cannot be the code that generated the reported model fits\footnote{One example of why the code cannot be run that the file loads functions from a non-extant file named \code{choleraEqs.py} in line 13 rather than \code{choleraEqsPublic.py}.}.
+The inability to replicate the results by \citet{lee20} by running the provided source code makes checking whether or not a our numeric implementation faithfully represents their results very difficult.
+% The difficulty of reproducing the result of \citet{lee20} is coupled with a misunderstanding of how these results were originally obtained.
+
+\jwc{PROBABLY DELETE THIS NEXT BIT, BUT IT'S USEFUL TO DOCUMENT IT FOR NOW...}
+One concern that arises when inspecting the files \code{fitInPieces3paramsCleanMay2019Public.py} and \code{fitInPiecesMuWithFracSusFixedAllInfectionsPublic.py} is that the output of these files (which contains the estimated parameters) was not referenced/used in either the file that generates the sensitivity analysis or the file that generates the model simulations; instead, alternative (non-extant) files with similar names are loaded, suggesting parameters that were used to generate the final results were not a direct result of the estimated parameters.
+Furthermore, inspection of the script \code{simulateHaitiFullTimeInPiecesAllVaccinationScenariosSingleParameterVectorAllInfectionsPublic.py}---which is used to obtain simulations for each of the vaccination scenarios---results in a number of concerns.
+First, the script uses parameters from the pickle file \code{leastSquaresDataUpTo2019/fitEpidemicPiece\_resultsMatrix\_18Mar2019.pickle}, which only appears in the repository as a commented out line of code in the file \code{generateSetsOfParametersForSensitivityAnalysisPublic.py}, suggesting that the set of parameters used to simulate each vaccination scenario was not the most recent set of parameters generated by the sensitivity analysis.
+Second, the script appears to use the set of differential equations from the function \code{choleraEqs11WithVaccinationNetwork} for the epidemic phase, even though parameter values for this phase were estimated using \code{choleraEqs10WithVaccinationNetwork}.
+It is possible, however, that we simply are interpreting the code incorrectly and that these concerns are unwarranted.
+Furthermore, because the provided code is non-functional, it is possible that the code that was actually used to obtain the reported results does not include these potential errors.
+This highlights the importance of providing the exact source code that was used to obtain the reported results of any analysis.
+
+Another potential explanation for the discrepancy between our simulations from Model~2 and those of \citet{lee20} is that the parameters that we used are only approximately the same as those used by \citet{lee20sup}.
+For example, the parameters $\transmission$, $\Wbeta{}$ (See Table~\ref{tab:translate}) had reported values of $9.9 \times 10^{-7}$ and $4.03 \times 10^{-2}$, respectively (Table~S13 of \cite{lee20sup}), but were actually fit to data and therefore likely had values that were far more precise.
 Additionally, our implementation of Model~2 used a time scale of years and many of the parameters were reported on a weekly scale, so small differences may result due to unit conversions.
-The collective effect of these small differences in model parameters likely will result in differences in model trajectories, as seen here.
+The collective effect of these small differences in model parameters likely will result in differences in model trajectories.
+This issue could have been mitigated if the saved files containing estimated model parameters were included as part of the code submission.
 

### 2022-10-10 ms.Rnw
Updated R code to spatPomp fitting

@@
 <<FitModel3, echo=FALSE, include=FALSE, message=FALSE>>=
-set.seed(636813)
-
+set.seed(1859687)
+
+# Create vectors for the unit and shared parameters
+unit_specific_names <- c("betaB", "foi_add", "Binit")
+shared_param_names <- c(
+  "mu_B", "XthetaA", "thetaI", "lambdaR", "r", "std_W",
+  "epsilon", "k", "sigma"
+)
+est_param_names <- c(
+  unit_specific_names, shared_param_names
+)
+
+# Add unit numbers to each parameter
+est_param_names_expanded <- paste0(rep(est_param_names, each = 10), 1:10)
+
+setrw1 <- function(x) {
+  if (grepl('^r[[:digit:]]+$', x)) {
+    return(0.01 / 2)
+  } else if (grepl('^Binit[[:digit:]]+$', x)) {
+    return(expression(ivp(0.02)))
+  } else {
+    return(0.01)
+  }
+}
+
+setrw2 <- function(x) {
+  if (grepl('^r[[:digit:]]+$', x)) {
+    return(0.0025 / 2)
+  } else if (grepl('^Binit[[:digit:]]+$', x)) {
+    return(expression(ivp(0.005)))
+  } else {
+    return(0.0025)
+  }
+}
+
+setrw3 <- function(x) {
+  if (grepl('^r[[:digit:]]+$', x)) {
+    return(0.001 / 2)
+  } else if (grepl('^Binit[[:digit:]]+$', x)) {
+    return(expression(ivp(0.002)))
+  } else {
+    return(0.001)
+  }
+}
+
+# First Search RW
+reg_rw_1.sd <- lapply(est_param_names_expanded, setrw1)
+names(reg_rw_1.sd) <- est_param_names_expanded
+chol_rw_1.sd <- do.call(rw.sd, reg_rw_1.sd)
+
+# Second Search RW
+reg_rw_2.sd <- lapply(est_param_names_expanded, setrw2)
+names(reg_rw_2.sd) <- est_param_names_expanded
+chol_rw_2.sd <- do.call(rw.sd, reg_rw_2.sd)
+
+# Third Search RW
+reg_rw_3.sd <- lapply(est_param_names_expanded, setrw3)
+names(reg_rw_3.sd) <- est_param_names_expanded
+chol_rw_3.sd <- do.call(rw.sd, reg_rw_3.sd)
+
+# Search hyper-parameters for each run-level and search
 if (RUN_LEVEL == 1){
-
-  # Global search, All parameters
-  SEARCH1 <- list(NP = 20, NMIF = 3, NREPS = 3, NREPS_EVAL = 3, NP_EVAL = 25)
-
-  # Local search, all parameters
-  SEARCH2 <- list(
-    TOP_N = 1, NP = 20, NMIF = 3, NREPS = 3, NREPS_EVAL = 3, NP_EVAL = 25
-  )
-
-  # Local search, unit specific parameters
-  SEARCH3 <- list(TOP_N = 1, NP = 20, NMIF = 3, NREPS = 3, NREPS_EVAL = 3, NP_EVAL = 25)
-
-  # Number of searches
-  N_SEARCHES <- 3L
-
+  SEARCH1 = list(
+    NBPF = 5,
+    NP = 50,
+    SPAT_REGRESSION = 0.5,
+    NREPS = 6,
+    NP_EVAL = 100,
+    NREPS_EVAL = 6,
+    RW_SD = chol_rw_1.sd,
+    COOLING = 0.5,
+    KEEP_TRACES = TRUE
+  )
+
+  SEARCH2 = list(
+    TOP_N = 2,
+    NBPF = 5,
+    NP = 50,
+    SPAT_REGRESSION = 0.5,
+    NREPS = 3,
+    NP_EVAL = 100,
+    NREPS_EVAL = 6,
+    RW_SD = chol_rw_2.sd,
+    COOLING = 0.5,
+    KEEP_TRACES = TRUE
+  )
+
+  SEARCH3 = list(
+    TOP_N = 2,
+    NBPF = 5,
+    NP = 100,
+    SPAT_REGRESSION = 0.5,
+    NREPS = 3,
+    NP_EVAL = 100,
+    NREPS_EVAL = 6,
+    RW_SD = chol_rw_3.sd,
+    COOLING = 0.5,
+    KEEP_TRACES = TRUE
+  )
+
+  N_SEARCHES <- 2L
 } else if (RUN_LEVEL == 2) {
-
-  # Global search, all parameters
-  SEARCH1 <- list(
-    NP = 200, NMIF = 10, NREPS = 36, NREPS_EVAL = 36, NP_EVAL = 500
-  )
-
-  # Local search, all parameters
-  SEARCH2 <- list(
-    TOP_N = 2, NP = 200, NMIF = 20, NREPS = 18, NREPS_EVAL = 36, NP_EVAL = 500
-  )
-
-  # Local search, unit specific parameters
-  SEARCH3 <- list(
-    TOP_N = 2, NP = 200, NMIF = 10, NREPS = 18, NREPS_EVAL = 36, NP_EVAL = 500
-  )
-
-  # Number of searches
+  SEARCH1 = list(
+    NBPF = 50,
+    NP = 1000,
+    SPAT_REGRESSION = 0.5,
+    NREPS = 36,
+    NP_EVAL = 1000,
+    NREPS_EVAL = 36,
+    RW_SD = chol_rw_1.sd,
+    COOLING = 0.36,
+    KEEP_TRACES = TRUE
+  )
+
+  SEARCH2 = list(
+    TOP_N = 6,
+    NBPF = 50,
+    NP = 1000,
+    SPAT_REGRESSION = 0.5,
+    NREPS = 6,
+    NP_EVAL = 1000,
+    NREPS_EVAL = 36,
+    RW_SD = chol_rw_2.sd,
+    COOLING = 0.5,
+    KEEP_TRACES = TRUE
+  )
+
+  SEARCH3 = list(
+    TOP_N = 6,
+    NBPF = 50,
+    NP = 1000,
+    SPAT_REGRESSION = 0.5,
+    NREPS = 6,
+    NP_EVAL = 1000,
+    NREPS_EVAL = 36,
+    RW_SD = chol_rw_3.sd,
+    COOLING = 0.5,
+    KEEP_TRACES = TRUE
+  )
+
   N_SEARCHES <- 3L
@@
 
-  # Global search, all parameters
-
   SEARCH1 = list(
     NBPF = 100,
-    NP = 1500,
+    NP = 10000,
     SPAT_REGRESSION = 0.5,
-    NREPS = 144,
-    NP_EVAL = 2000,
-    NREPS_EVAL = 36
+    NREPS = 72,
+    NP_EVAL = 10000,
+    NREPS_EVAL = 36,
+    RW_SD = chol_rw_1.sd,
+    COOLING = 0.5,
+    KEEP_TRACES = TRUE
   )
@@
   SEARCH2 = list(
-    TOP_N = 12,
+    TOP_N = 9,
     NBPF = 100,
-    NP = 2000,
+    NP = 10000,
+    SPAT_REGRESSION = 0.5,
+    NREPS = 4,
+    NP_EVAL = 10000,
+    NREPS_EVAL = 36,
+    RW_SD = chol_rw_2.sd,
+    COOLING = 0.5,
+    KEEP_TRACES = TRUE
+  )
+
+  SEARCH3 = list(
+    TOP_N = 6,
+    NBPF = 100,
+    NP = 10000,
     SPAT_REGRESSION = 0.5,
     NREPS = 6,
-    NP_EVAL = 3000,
-    NREPS_EVAL = 36
-  )
-
-  SEARCH3 = list(
-    TOP_N = 9,
-    NBPF = 100,
-    NP = 2500,
-    SPAT_REGRESSION = 0.5,
-    NREPS = 8,
-    NP_EVAL = 5000,
-    NREPS_EVAL = 36
-  )
-
-  # Total number of searches
+    NP_EVAL = 10000,
+    NREPS_EVAL = 36,
+    RW_SD = chol_rw_3.sd,
+    COOLING = 0.5,
+    KEEP_TRACES = TRUE
+  )
+
   N_SEARCHES <- 3L
-} else {
-  stop("Run level must be 1, 2 or 3.")
-}
-
-if (RUN_LEVEL != 3 & file.exists("model3/run_level_3/haiti3_fit.rds")) {
-  haiti3_fit <- readRDS("model3/run_level_3/haiti3_fit.rds")
+}
+
+if (RUN_LEVEL != 3 & file.exists("model3/run_level_3/haiti3_coupled_fit.rds")) {
+  haiti3_fit <- readRDS("model3/run_level_3/haiti3_coupled_fit.rds")
 } else {
   haiti3_fit <- bake(
-    file = paste0("model3/", rl_dir, "haiti3_fit.rds"), {
-      fit_haiti3(
+    file = paste0("model3/", rl_dir, "haiti3_coupled_fit.rds"), {
+      fit_coupled_haiti3(
+        nsearches = N_SEARCHES,
         search1 = SEARCH1,
@@
         search3 = SEARCH3,
-        n_searches = N_SEARCHES,
         ncores = cores
@@
 
-rm(SEARCH1, SEARCH2, SEARCH3, N_SEARCHES)
+rm(
+  SEARCH1, SEARCH2, SEARCH3, N_SEARCHES, unit_specific_names, 
+  shared_param_names, est_param_names, est_param_names_expanded, 
+  setrw1, setrw2, setrw3, reg_rw_1.sd, reg_rw_2.sd, reg_rw_3.sd, 
+  chol_rw_1.sd, chol_rw_2.sd, chol_rw_3.sd
+)
 gc()
+
 @
@@
 )
-
-# MIF_mod3 <- local_MIF2_search
-# rm(mif_logLik, local_MIF2_search, tLocal)
 
@@
 
-# best_m3 <- ll_mod3 %>%
-#   arrange(-logLik) %>%
-#   slice_head(n = 1) %>%
-#   pull(which)
-
-best_m3 <- haiti3_fit[[length(haiti3_fit)]]$logLiks %>%
-  filter(logLik == max(logLik, na.rm = TRUE)) %>%
-  pull(which)
+best_m3 <- which.max(haiti3_fit[[length(haiti3_fit)]]$logLiks$logLik)
 
 p3 <- haiti3_fit[[length(haiti3_fit)]]$params[best_m3, ]
-gc()
+
+p3_specific <- c("foi_add", "betaB", "Binit")
+p3_shared <- c(
+  "mu_B", "XthetaA", "thetaI", "lambdaR", "r", "std_W",
+  "epsilon", "k", "sigma"
+)
+
+p3u <- matrix(nrow = length(p3_specific), ncol = 10)
+rownames(p3u) <- p3_specific
+p3s <- numeric(length(p3_shared))
+names(p3s) <- p3_shared
+
+for (p in p3_shared) {
+  pat <- paste0("^", p, "[[:digit:]]+$")
+  p3s[p] <-  mean(p3[grepl(pat, names(p3))])
+}
+
+for (p in p3_specific) {
+  pat <- paste0("^", p, "[[:digit:]]+$")
+  p3u[p, ] <- p3[grepl(pat, names(p3))]
+}
+
+rm(p3_specific, p3_shared, p, pat)
 
@@
   h2_epi <- haiti2(region = 'before', cutoff = 10000, measure = 'log')
-
   coef(h2_epi) <- h2_fit$h2_params
 
-  h3 <- haiti3_panel(start_time = "2010-10-23", B0 = TRUE)
-  coef(h3) <- p3
-  pp3 <- pparams(h3)
-  p3u <- pp3$specific
-  p3s <- pp3$shared
+  h3Spat <- haiti3_spatPomp()
+  coef(h3Spat) <- p3
 })
@@
     file = paste0('model3/', rl_dir, 'SpatPompSims.rds'), {
-      h3Spat <- haiti3_spatPomp()
-      coef(h3Spat) <- haitipkg:::panelParms_toSpatParms3(p3)
+      # h3Spat <- haiti3_spatPomp()
+      # coef(h3Spat) <- haitipkg:::panelParms_toSpatParms3(p3)
 
@@
 
-h3Spat <- haiti3_spatPomp()
-coef(h3Spat) <- haitipkg:::panelParms_toSpatParms3(p3)
+# h3Spat <- haiti3_spatPomp()
+# coef(h3Spat) <- haitipkg:::panelParms_toSpatParms3(p3)
 

### 2022-10-11 ms.Rnw
Minor modifications

@@
 
-While this work is focused on the models proposed in \cite{lee20}, our suggestions have broader relevance.
+While this work is focused on the models proposed by \cite{lee20}, our suggestions have broader relevance.
 To investigate the extent to which \citet{lee20} is typical of the substantial body of work on Haiti cholera dynamics, we preformed a literature review by searching PubMed with keywords: Haiti, cholera, model.
@@
 \label{model1:lambda}
-\lambda(t) = \Big(\sum_{\vaccCounter=0}^{\vaccClass}  I_{\vaccCounter}(t) + \asymptomRelativeInfect \sum_{\vaccCounter=0}^{\vaccClass} A_{\vaccCounter}(t) \Big)^\nu \frac{d\Gamma(t)}{dt} \transmission(t)/N
+\lambda(t) = \Big(\sum_{\vaccCounter=0}^{\vaccClass}  I_{\vaccCounter}(t) + \asymptomRelativeInfect \sum_{\vaccCounter=0}^{\vaccClass} A_{\vaccCounter}(t) \Big)^\nu \frac{d\Gamma(t)}{dt} \transmission(t)/N,
 \end{equation}
@@
 \label{model1:SE}
-\mu_{S_{\vaccCounter}E_{\vaccCounter}} &=& \lambda(t)
+\mu_{S_{\vaccCounter}E_{\vaccCounter}} &=& \lambda(t),
 \\[-3pt]
 \label{model1:EI}
-\mu_{E_{\vaccCounter}I_{\vaccCounter}} &=& \muEI\big(1-\symptomFrac_{\vaccCounter}(t)\big)
+\mu_{E_{\vaccCounter}I_{\vaccCounter}} &=& \muEI\big(1-\symptomFrac_{\vaccCounter}(t)\big),
 \\[-3pt]
 \label{model1:EA}
-\mu_{E_{\vaccCounter}A_{\vaccCounter}} &=& \muEI\, \symptomFrac_{\vaccCounter}(t)
+\mu_{E_{\vaccCounter}A_{\vaccCounter}} &=& \muEI\, \symptomFrac_{\vaccCounter}(t),
 \\[-3pt]
 \label{model1:toR}
-\mu_{I_{\vaccCounter}R_{\vaccCounter}} &=& \mu_{A_{\vaccCounter}R_{\vaccCounter}} = \muIR
+\mu_{I_{\vaccCounter}R_{\vaccCounter}} &=& \mu_{A_{\vaccCounter}R_{\vaccCounter}} = \muIR,
 \\[-3pt]
 \label{model1:RS}
-\mu_{R_{\vaccCounter}S_{\vaccCounter}} &=& \muRS
+\mu_{R_{\vaccCounter}S_{\vaccCounter}} &=& \muRS,
 \\[-3pt]
 \label{model1:vacc}
-\mu_{S_0S_{\vaccCounter}} &=& \mu_{E_0E_{\vaccCounter}} = \mu_{I_0I_{\vaccCounter}} = \mu_{A_0A_{\vaccCounter}} = \mu_{R_0R_{\vaccCounter}} = \eta_{\vaccCounter}(t)
+\mu_{S_0S_{\vaccCounter}} &=& \mu_{E_0E_{\vaccCounter}} = \mu_{I_0I_{\vaccCounter}} = \mu_{A_0A_{\vaccCounter}} = \mu_{R_0R_{\vaccCounter}} = \eta_{\vaccCounter}(t),
 \\[-3pt]
 \label{model1:death}
-\mu_{S_{\vaccCounter}\demography} &=& \mu_{E_{\vaccCounter}\demography} = \mu_{I_{\vaccCounter}\demography} = \mu_{A_{\vaccCounter}\demography}=\mu_{R_{\vaccCounter}\demography} = \delta
+\mu_{S_{\vaccCounter}\demography} &=& \mu_{E_{\vaccCounter}\demography} = \mu_{I_{\vaccCounter}\demography} = \mu_{A_{\vaccCounter}\demography}=\mu_{R_{\vaccCounter}\demography} = \delta,
 \\
 \label{model1:birth}
-\mu_{\demography S_0} &=& \muBirth
+\mu_{\demography S_0} &=& \muBirth,
 \end{eqnarray}
@@
 For $\vaccCounter\in\seq{1}{\vaccClass}$, the vaccination cohort $\vaccCounter$ is assigned a time $\tau_{\vaccCounter}$, and we take $\symptomFrac_{\vaccCounter}(t) = c \, \theta^*(t-\tau_{\vaccCounter})$
-where $\theta^*(t)$ is efficacy at time $t$ since vaccination for adults, taken from \citet{lee20}, Table~S4, and $c=\big(1-(1-0.4688)\times 0.11\big)$ is a correction to allow for reduced efficacy in the 11\% of the population aged under 5 years.
+where $\theta^*(t)$ is efficacy at time $t$ since vaccination for adults, taken from \cite{lee20}, Table~S4, and $c=\big(1-(1-0.4688)\times 0.11\big)$ is a correction to allow for reduced efficacy in the 11\% of the population aged under 5 years.
 Single and double vaccine doses were modeled by changing the waning of protection; protection was assumed to be equal between single and double dose until 52 weeks after vaccination, at which point the single dose becomes ineffective.
@@
 \frac{\beta_W\, W_u(t)}{ \Wsat  + W_u(t)} +
-\transmission \left\{\sum_{\vaccCounter=0}^4 I_{u\vaccCounter}(t) + \asymptomRelativeInfect \sum_{\vaccCounter=0}^4 A_{u\vaccCounter}(t) \right\}
+\transmission \left\{\sum_{\vaccCounter=0}^4 I_{u\vaccCounter}(t) + \asymptomRelativeInfect \sum_{\vaccCounter=0}^4 A_{u\vaccCounter}(t) \right\}.
 \end{equation}
@@
 The cosine term in Eq.~\myeqref{model2:lambda} accounts for annual seasonality, with a phase parameter $\phaseParm$.
-The implementation of Model~2 in \cite{lee20} fixes $\phaseParm = 0$.
+The \cite{lee20} implementation of Model~2 fixes $\phaseParm = 0$.
 
@@
 \label{model2:mu_SE}
-\mu_{S_{u\vaccCounter}E_{u\vaccCounter}} &=& \vaccineEfficacy_\vaccCounter \, \lambda
+\mu_{S_{u\vaccCounter}E_{u\vaccCounter}} &=& \vaccineEfficacy_\vaccCounter \, \lambda,
 \\[-3pt]
 \label{model2:mu_EI}
-\mu_{E_{u\vaccCounter}I_{u\vaccCounter}} &=& \symptomFrac\muEI, \quad \mu_{E_{u\vaccCounter}A_{u\vaccCounter}} = (1-\symptomFrac)\muEI
+\mu_{E_{u\vaccCounter}I_{u\vaccCounter}} &=& \symptomFrac\muEI, \quad \mu_{E_{u\vaccCounter}A_{u\vaccCounter}} = (1-\symptomFrac)\muEI,
 \\[-3pt]
 \label{model2:mu_IR}
-\mu_{I_{u\vaccCounter}R_{u\vaccCounter}} &=& \mu_{A_{u\vaccCounter}R^A_{u\vaccCounter}} = \muIR
+\mu_{I_{u\vaccCounter}R_{u\vaccCounter}} &=& \mu_{A_{u\vaccCounter}R^A_{u\vaccCounter}} = \muIR,
 \\[-3pt]
 \label{model2:RS}
-\mu_{R_{u\vaccCounter}S_{u\vaccCounter}} &=& \mu_{R^A_{u\vaccCounter}S_{u\vaccCounter}} = \muRS
+\mu_{R_{u\vaccCounter}S_{u\vaccCounter}} &=& \mu_{R^A_{u\vaccCounter}S_{u\vaccCounter}} = \muRS,
 \\[-3pt]
 \label{model2:transport}
-\mu_{S_{u\vaccCounter}S_{{\varv}\vaccCounter}} &=& \mu_{E_{u\vaccCounter}E_{{\varv}\vaccCounter}} = \mu_{I_{u\vaccCounter} I_{{\varv}\vaccCounter}} = \mu_{A_{u\vaccCounter}A_{{\varv}\vaccCounter}} = \mu_{R_{u\vaccCounter}R_{{\varv}\vaccCounter}} = \mu_{R^A_{u\vaccCounter} R^A_{{\varv}\vaccCounter}} = T_{u\varv}
+\mu_{S_{u\vaccCounter}S_{{\varv}\vaccCounter}} &=& \mu_{E_{u\vaccCounter}E_{{\varv}\vaccCounter}} = \mu_{I_{u\vaccCounter} I_{{\varv}\vaccCounter}} = \mu_{A_{u\vaccCounter}A_{{\varv}\vaccCounter}} = \mu_{R_{u\vaccCounter}R_{{\varv}\vaccCounter}} = \mu_{R^A_{u\vaccCounter} R^A_{{\varv}\vaccCounter}} = T_{u\varv},
 \\[-3pt]
 \label{model2:omega1}
-\mu_{S_{u1}S_{u0}} &=& \mu_{S_{u3}S_{u0}} = \omega_1
+\mu_{S_{u1}S_{u0}} &=& \mu_{S_{u3}S_{u0}} = \omega_1,
 \\[-3pt]
 \label{model2:omega2}
-\mu_{S_{u2}S_{u0}} &=& \mu_{S_{u4}S_{u0}} = \omega_2
+\mu_{S_{u2}S_{u0}} &=& \mu_{S_{u4}S_{u0}} = \omega_2,
 \\[-3pt]
 \label{model2:to_W}
-\mu_{\demography W_u} &=& \Wshed \left\{ \sum_{\vaccCounter=0}^4 I_{u\vaccCounter}(t) + \asymptomRelativeShed \sum_{\vaccCounter=0}^4 A_{u\vaccCounter}(t) \right\}
+\mu_{\demography W_u} &=& \Wshed \left\{ \sum_{\vaccCounter=0}^4 I_{u\vaccCounter}(t) + \asymptomRelativeShed \sum_{\vaccCounter=0}^4 A_{u\vaccCounter}(t) \right\},
 \\[-3pt]
 \label{model2:from_W}
-\mu_{W_u\demography} &=& \Wremoval
+\mu_{W_u\demography} &=& \Wremoval,
 \\[-3pt]
 \label{model2:water_transport}
-\mu_{W_uW_{\varv}} &=& w_r T^W_{u\varv}
+\mu_{W_uW_{\varv}} &=& w_r T^W_{u\varv}.
 \end{eqnarray}
@@
 The force of infection is
+\begin{equation}
+\label{eq:model3:foi}
+\lambda_u(t) = \Wbeta{_u} \frac{W_u(t)}{1+W_u(t)} + \transmission_u \sum_{{\varv}\neq u}\big(I_{\varv 0}(t)+ \asymptomRelativeInfect A_{\varv 0}(t)\big),
+\end{equation}
+and Per-capita transition rates are given in Equations~\ref{eq:model3:SI}--\ref{eq:model3:Decay}.
 \begin{eqnarray}
-\label{eq:model3:foi}
-\lambda_u(t) &=& \Wbeta{_u} \frac{W_u(t)}{1+W_u(t)} + \transmission_u \sum_{{\varv}\neq u}\big(I_{\varv 0}(t)+ \asymptomRelativeInfect A_{\varv 0}(t)\big)
-\\[-3pt]
 \label{eq:model3:SI}
-\mu_{S_{u\vaccCounter}I_{u\vaccCounter}} &=& \symptomFrac \,  \lambda_u \big(1-\eta_{u\vaccCounter}(t)\big) \, d\Gamma/dt
+\mu_{S_{u\vaccCounter}I_{u\vaccCounter}} &=& \symptomFrac \,  \lambda_u \big(1-\eta_{u\vaccCounter}(t)\big) \, d\Gamma/dt,
 \\[-3pt]
 \label{eq:model3:SA}
-\mu_{S_{u\vaccCounter}A_{u\vaccCounter}} &=& (1-\symptomFrac) \,  \lambda_u \big(1-\eta_{u\vaccCounter}(t)\big) \,  d\Gamma/dt
+\mu_{S_{u\vaccCounter}A_{u\vaccCounter}} &=& (1-\symptomFrac) \,  \lambda_u \big(1-\eta_{u\vaccCounter}(t)\big) \,  d\Gamma/dt,
 \\[-3pt]
 \label{eq:model3:IR}
-\mu_{I_{u\vaccCounter}R_{u\vaccCounter 1}} &=& \mu_{A_{u\vaccCounter}R_{u\vaccCounter 1}} = \muIR
+\mu_{I_{u\vaccCounter}R_{u\vaccCounter 1}} &=& \mu_{A_{u\vaccCounter}R_{u\vaccCounter 1}} = \muIR,
 \\[-3pt]
 \label{eq:model3:IS}
-\mu_{I_{u\vaccCounter}S_{u0}} &=& \muDeath + \choleraDeath
+\mu_{I_{u\vaccCounter}S_{u0}} &=& \muDeath + \choleraDeath,
 \\[-3pt]
 \label{eq:model3:AS}
-\mu_{A_{u\vaccCounter}S_{u0}} &=& \muDeath
+\mu_{A_{u\vaccCounter}S_{u0}} &=& \muDeath,
 \\[-3pt]
 \label{eq:model3:RRnext}
-\mu_{R_{u\vaccCounter 1}R_{u\vaccCounter 2}} &=& \mu_{R_{u\vaccCounter 2}R_{u\vaccCounter 3}} = 3\muRS
+\mu_{R_{u\vaccCounter 1}R_{u\vaccCounter 2}} &=& \mu_{R_{u\vaccCounter 2}R_{u\vaccCounter 3}} = 3\muRS,
 \\[-3pt]
 \label{eq:model3:RS}
-\mu_{R_{u\vaccCounter k}S_{u0}} &=& \muDeath + 3\muRS \, \mathbf{1}_{\{k=3\}}
+\mu_{R_{u\vaccCounter k}S_{u0}} &=& \muDeath + 3\muRS \, \mathbf{1}_{\{k=3\}},
 \\[-3pt]
 \label{eq:model3:water}
-\mu_{{\demography}W_u} &=& \big[1 + \seasAmplitude \big(J_u(t))^r \big] \mathrm{Den}_u \, \Wshed \big[ I_{u0}(t)+ \asymptomRelativeShed A_{u0}(t) \big]
+\mu_{{\demography}W_u} &=& \big[1 + \seasAmplitude \big(J_u(t))^r \big] \mathrm{Den}_u \, \Wshed \big[ I_{u0}(t)+ \asymptomRelativeShed A_{u0}(t) \big],
 \\[-3pt]
 \label{eq:model3:Decay}
-\mu_{W_u\demography} &=& \Wremoval
+\mu_{W_u\demography} &=& \Wremoval.
 \end{eqnarray}
-
 As with Model~1, $d\Gamma_u(t)/dt$ is multiplicative Gamma-distributed white noise in \myeqref{eq:model3:SI} and \myeqref{eq:model3:SA}.
@@
 
-\subsubsection{Model~1}
+\subsubsection{Calibrating Model~1 Parameters}\label{sec:fit1}
 
@@
 \label{model1:betat}
-\log\transmission(t) = \sum_{j=1}^6 \transmission_s s_j(t) + \transmissionTrend\bar{t}
+\log\transmission(t) = \sum_{j=1}^6 \transmission_s s_j(t) + \transmissionTrend\bar{t},
 \end{equation}
-Where $\bar{t} = \frac{t - (t_N + t_0) / 2}{t_N - (t_N + t_0) / 2}$, so that $\bar{t} \in [-1, 1]$.
+where $\bar{t} = \frac{t - (t_N + t_0) / 2}{t_N - (t_N + t_0) / 2}$, so that $\bar{t} \in [-1, 1]$.
 The proposal of a linear trend in transmission is a result of observing an apparent decrease in reported cholera infections from 2012-2019 in Fig.~\ref{CholeraData}.
@@
 To do this, we perform a profile-likelihood search on the parameter $\transmissionTrend$ and obtain a confidence interval via a Monte Carlo Adjusted Profile (MCAP) \citep{ionides17}.
-Model~1 was implemented in \cite{lee20} by fitting two distinct phases: an epidemic phase from October 2010 through March 2015, and an endemic phase from March 2015 onward.
+\cite{lee20} implemented Model~1 by fitting two distinct phases: an epidemic phase from October 2010 through March 2015, and an endemic phase from March 2015 onward.
 We similarly allow the re-estimation of process and measurement overdispersion parameters ($\sigmaProc^2$ and $\obsOverdispersion$), and require that the latent Markov process $X(t)$ carry over from one phase into the next.
@@
 
-\subsubsection{Model~2}\label{sec:fit2}
+\subsubsection{Calibrating Model~2 Parameters}\label{sec:fit2}
 
@@
 % A naive solution to this would be to estimate only model parameters in separate phases and not the hidden states, but this may also result in hidden states in one phase of the model that are inconsistent with the other phase of the outbreak, since only the data from one phase is used to fit model parameters.
-The inclusion of a model break-point in \cite{lee20} is perhaps due to a challenging feature of fitting a deterministic model via least squares: discrepancies between model trajectories and observed case counts in highly infectious periods of a disease outbreak will result in greater penalty than the discrepancies between model trajectories and observed case counts in times of relatively low infectiousness, resulting in a bias towards accurately describing periods of high infectiousness.
+The inclusion of a model break-point by \cite{lee20} is perhaps due to a challenging feature of fitting a deterministic model via least squares: discrepancies between model trajectories and observed case counts in highly infectious periods of a disease outbreak will result in greater penalty than the discrepancies between model trajectories and observed case counts in times of relatively low infectiousness, resulting in a bias towards accurately describing periods of high infectiousness.
 This issue is particularly troublesome for modeling cholera dynamics in Haiti: the inability to accurately fit times of low infectiousness may result in poor model forecasts, as few cases of cholera were observed the last few years of the epidemic.
@@
 This permits the consideration of demographic stochasticity, which is dominant for small infected populations, together with log scale stochasticity (also called multiplicative, or environmental, or extra-demographic) which is dominant at high population counts.
-Here we chose to fit the model to transformed case counts rather than adding overdispersion to the measurement model with the goal of making the fewest changes to the model in \cite{lee20} as possible.
+Here we chose to fit the model to transformed case counts rather than adding overdispersion to the measurement model with the goal of making the fewest changes to the model proposed by \cite{lee20} as possible.
 
@@
 
-\subsubsection{Model~3}
+\subsubsection{Calibrating Model~3 Parameters}\label{sec:fit3}
 Both the latent and observable processes in Model~3 can be be factored into department specific processes which interact with each other.
@@
 The ability to directly estimate parameters of Model~3 is responsible for the large increase in model likelihoods reported in Table~\ref{tab:likes}, as no algorithm with similar capabilities existed when \cite{lee20} published their results.
-Simulations from the fitted model are displayed in Fig~.\ref{fig:h3spatsims}.
+Simulations from the fitted model are displayed in Fig.~\ref{fig:h3spatsims}.
 
@@
 @
-\caption{\label{h3spatsims}
+\caption{\label{fig:h3spatsims}
 Simulations from initial conditions using the spatially coupled version of Model~3.
@@
 If a Bayesian technique is used for parameter estimation, the most natural way to account for parameter uncertainty is to obtain simulations from the model where each simulation is obtained using parameters drawn from the posterior distribution.
-In the frequentist case, one approach is empirical Bayes, as was done in \cite{king15} (see supplement for details).
+In the frequentist case, one approach is empirical Bayes, as was done by \cite{king15} (see supplement for details). % TODO: \jwc{ADD THIS TO SUPPLEMENT!}
 Both of these approaches share the similarity that parameters with higher likelihoods should be used more frequently in model forecasts than those that correspond to lower likelihoods.
@@
 Notably, the median number of cumulative cholera infections under the no-vaccination scenario using Models 1 and 3 were $\Sexpr{prettyNum(round(as.numeric(h1_S0_results$mod1_V0_probs$cumInf['q50'])), big.mark = ",")}$ and $\Sexpr{prettyNum(round(as.numeric(mod3_V0_res$mod3_V0_probs$cumInf['q50'])), big.mark = ",")}$, respectively.
-While there is remaining time during this projection period in which new cholera infections can be detected, up to this point our estimates are far more consistent with the observed number of reported cholera cases than the corresponding estimates from \cite{lee20}, which were approximately $400,000$ and $1,000,000$.
+While there is remaining time during this projection period in which new cholera infections can be detected, up to this point our estimates are far more consistent with the observed number of reported cholera cases than the corresponding estimates of \cite{lee20}, which were approximately $400,000$ and $1,000,000$.
 
@@
 
-In this article we argue that careful attention to important statistical details could have led the models in \cite{lee20} to correctly predict the imminent cholera elimination.
+In this article we argue that careful attention to important statistical details could have led the models proposed by \cite{lee20} to correctly predict the imminent cholera elimination.
 We acknowledge the benefit of hindsight: our demonstration of a statistically principled route to obtain better-fitting models with more accurate predictions does not rule out the possibility of discovering other models that fit well yet predict poorly.
@@
 \sdescription{
-This document contains additional details for Models~1--3, as well as a translation table that facilitates comparisons between these models and those described in \cite{lee20}.
+This document contains additional details for Models~1--3, as well as a translation table that facilitates comparisons between these models and those described by \cite{lee20}.
 The supplement also demonstrates our capability to faithfully replicate the results

### 2022-10-11 si/si.Rnw
Minor modifications

@@
 run_level <- 1
-
-if(run_level=="manual"){
-}
-
 @
@@
 library(tidyverse)
-#library(xtable)
 library(pomp)
 library(haitipkg)
-# library(spatPomp)
+library(spatPomp)
 library(doParallel)
@@
 \newcommand\giventh{{\hspace{0.5mm};\hspace{0.5mm}}}
-%\newcommand\normal{{\mathrm{Normal}}}
 \newcommand\normal{\mathcal{N}}
@@
 \newcommand\MC[1]{#1^{\,\mbox{\tiny MC}}}
-%\newcommand\EMC{\MC{\E}}
 \newcommand\EMC{{\E}}
@@
 \newcommand\iid{\mathrm{iid}}
-%\newcommand\dist{\mathrm{dist}}
 \newcommand\dist{d}
@@
 \newtheorem{assumptionB}{Assumption}
-% \newtheorem{procedure}{Procedure}
 \renewcommand\theassumption{A\arabic{assumption}}
 \renewcommand\theassumptionB{B\arabic{assumptionB}}
-
-%\usepackage{float}
-%\floatstyle{ruled}
-%\newfloat{textbox}{t}{tbx}
-%\floatname{textbox}{Box}
-%\renewcommand\thetextbox{\arabic{textbox}}
 
@@
 \bibliographystyle{apalike}
-%% added or modified from pnas-new.cls
 \usepackage{natbib}
@@
 \newcommand\myeqref[1]{(\ref{#1})}
-
-%\usepackage{relsize}
-
-%\templatetype{pnassupportinginfo}
-
-% \readytosubmit %% Uncomment this line before submitting, so that the instruction page is removed.
-
-%\author{Author1, Author2 and Author3 (Complete author list)}
-%\correspondingauthor{Corresponding Author Name.\\E-mail: author.two@email.com}
 
@@
 \renewcommand\thetable{S-\arabic{table}}
-%\renewcommand\thepage{S-\arabic{page}}
 \renewcommand\thesection{S\arabic{section}}
@@
 }\fi
-
-%% Adds the main heading for the SI text. Comment out this line if you do not have any supporting information text.
-
-%\SItext
 
@@
 Each POMP requires specification of a measurement model, which is a statistical description of how observations on the system are obtained.
-In general, we used the same measurement models that were reported in \cite{lee20}.
+In general, we used the same measurement models that were reported by \cite{lee20}.
 
@@
 In this model, the advantage afforded by vaccination is an increased probability that an infection is asymptomatic.
-Therefore, under the assumptions of this model, all reported cases are assumed to be a fraction of individuals that transition from the exposed to the infected compartment, as noted in Eq.~\myeqref{model1:meas}.
+Therefore, under the assumptions of this model, all reported cases are assumed to be a fraction of individuals that transition from the exposed to the infected compartment, as noted in Eq.~\myeqref{model1:meas}:
 \begin{equation}
@@
   \label{model3:meas}
-  Y_{u, n} \mid \Delta N_{S_{u \vaccCounter}I_{u \vaccCounter}}(t) = z_{u}(n) \sim \text{NB}\left(\reportRate z_{u}(n), \obsOverdispersion \right)
-\end{equation}
-
-Where $\Delta N_{S_{u \vaccCounter}I_{u \vaccCounter}}(n)$ is the number of individuals who moved from compartment $S_{u \vaccCounter}$ to $I_{u \vaccCounter}$ since observation $n-1 \in 0:N-1$.
-
-This measurement model is a minor change from that used in \cite{lee20}, which allowed for a change in the reporting rate on January 1st, 2018.
+  Y_{u, n} \mid \Delta N_{S_{u \vaccCounter}I_{u \vaccCounter}}(t) = z_{u}(n) \sim \text{NB}\left(\reportRate z_{u}(n), \obsOverdispersion \right),
+\end{equation}
+where $\Delta N_{S_{u \vaccCounter}I_{u \vaccCounter}}(n)$ is the number of individuals who moved from compartment $S_{u \vaccCounter}$ to $I_{u \vaccCounter}$ since observation $n-1 \in 0:N-1$.
+
+This measurement model is a minor change from that used by \cite{lee20}, which allowed for a change in the reporting rate on January 1st, 2018.
 The fitted values of the reporting rate---before and after January 2018---were $0.97$ and $0.097$, respectively.
-This major change in reporting rate alone could explain why Model~3 in \cite{lee20} failed to predict the eradication of cholera.
+This major change in reporting rate alone could explain why Model~3 of \cite{lee20} failed to predict the eradication of cholera.
 An overnight change from near perfect to almost non-existent reporting forces the model to explain the observed reduction in reported cases as a decrease in the reporting of cases, rather than a decrease in the prevalence of cholera.
@@
 
-\section{Block Panel Iterated Filter}
-
-The panel iterated filter (PIF) algorithm of \citet{breto20} has been used to fit parameters of panelPOMP models in several studies \citep{domeyer22,wale2019,ranjeva17,ranjeva19}.
-Like other iterated filtering algorithms, PIF iteratively combines particle filter calculations with parameter perturbations.
-Theoretical results show that, as the perturbations are reduced over successive iterations, the perturbed parameters converge to the MLE.
-
-In the case of a panelPOMP, the parameter vector $\theta$ may be composed of a set of shared parameters $\phi$, and a set of unit-specific parameters $\psi_{u}$, for units $u \in \seq{1}{U}$.
-Here, we say that a parameter $\psi_{u}$ for $u \in \seq{1}{U}$ is unit-specific if it is not involved in the one-step transition density or the measurement model of any other unit $k \neq u \in \seq{1}{U}$, and shared parameters are those that are not unit-specific.
-The PIF algorithm treats the parameter vector $\theta = (\phi, \psi_{1:U})$ as a complete entity and gives no distinction between shared and unit-specific parameters.
-Even though the random walk standard deviation is set to zero for unit-specific parameters when considering data from other units, this lack of distinction may result in particle depletion for unit-specific parameters as the number of units $U$ grows large.
-This is because it is likely that, when updating the parameters for a given unit, several parameter vector $\theta$ particles will not be resampled due to having low weights.
-When the number of units is large, this will lead to a decreasing number of ancestral particles to sample unit-specific parameters $\psi_k$ as $k$ increases.
-
-As a solution to this issue, we propose the block panel iterated filter (BPIF), described in Algorithm~\ref{alg:bpif}.
-In this algorithm, shared $\phi$ and unit-specific parameters $\psi_{u}$ are handled differently when updating the parameter vector $\theta = (\phi, \psi_{1:U})$.
-The BPIF updates $\theta$ by updating the shared vector $\phi$ in the same way that is done with PIF, but index resampling for unit-specific parameters $\psi_{1:U}$ only occurs when data from the corresponding
-unit are being considered.
-% In the case where there are many units, we believe that this modification will lead to less particle depletion.
-We have found that this modification leads to improved likelihood maximization performance, especially as the number of units increases.
-No theoretical properties of this algorithm have yet been derived, but the approach has similarities to the block particle filter of \citet{rebeschini15} and its extension to parameter estimation \citep{ionides22,ning21ibpf}.
-
-\begin{algorithm}[ht]
-   \caption[BPIF]{\textbf{BPIF}. \\
-    {\bf Inputs}:\\\hspace{\textwidth}
-    Simulator of initial density, $f_{X_{u, 0}}(x_{u, 0}; \theta)$ for $u$ in $\seq{1}{U}$.\\\hspace{\textwidth}
-    Simulator of transition density, $f_{X_{u, n}|X_{u, n-1}}(x_{u, n}|x_{u, n-1}; \theta)$ for $u$ in $\seq{1}{U}$, $n$ in $\seq{1}{N_u}$.\\\hspace{\textwidth}
-    Evaluator of measurement density, $f_{Y_{u, n}|X_{u, n}}(y_{u, n}|x_{u, n} ; \theta)$ for $u$ in $\seq{1}{U}$, $n$ in $\seq{1}{N_u}$.\\\hspace{\textwidth}
-    Data $y_{u, n}^*$,for $u$ in $\seq{1}{U}$, $n$ in $\seq{1}{N_u}$.\\\hspace{\textwidth}
-    Number of iterations, $M$.\\\hspace{\textwidth}
-    Number of particles, $J$.\\\hspace{\textwidth}
-    Starting parameter swarm, $\Theta_j^0 = \big(\Phi_j^0, \Psi_{1:U, j}^0\big)$ for $j \in \seq{1}{J}$, $u \in \seq{1}{U}$.\\\hspace{\textwidth}
-    Simulator of perturbation density, $h_{u, n}(\theta | \phi; \sigma)$ for $u \in \seq{1}{U}$, $n \in 0:N_u$.\\\hspace{\textwidth}
-    Perturbation sequence, $\sigma_m$ for $m \in 1:M$.\\\hspace{\textwidth}
-    {\bf Output:} \\\hspace{\textwidth}
-    Final parameter swarm, $\Theta_{j}^{m} = \big(\Phi_j^m, \Psi_{\seq{1}{U}, j}^m\big)$ for $j \in \seq{1}{J}$, $u \in \seq{1}{U}$.
-    \label{alg:bpif}}
-%\BlankLine
-\For{$m \in 1:M$}{
-  Set $\Phi_{0, j}^m = \Phi_j^{m-1}$ for $j \in \seq{1}{J}$\;
-    \For{$u \in \seq{1}{U}$}{
-      Set $\Phi_{u, 0, j}^{F, m} \sim h_{u, 0}(\phi | \Phi_{0, j}^m; \sigma_m)$ for $j \in \seq{1}{J}$\;
-      Set $\Psi_{u, 0, j}^{F, m} \sim h_{u, 0}(\psi | \Psi_{u, j}^{m - 1}; \sigma_m)$ for $j \in \seq{1}{J}$\;
-      Set $\Theta_{u, 0, j}^{F, m} = \big(\Phi_{u, 0, j}^{F, m}, \Psi_{u, 0, j}^{F, m}\big)$ for $j \in \seq{1}{J}$\;
-      Initialize $X_{u, 0, j}^{F, m} \sim f_{X_{u, 0}}(x_{u, 0}; \Theta_{u, 0, j}^{F, m})$ for $j \in \seq{1}{J}$\;
-        \For{$n \in \seq{1}{N_u}$} {
-          Set $\Theta_{u, n, j}^{P, m} \sim h_{u, n}(\theta|\Theta_{u, n-1, j}^{F, m}, \sigma_m)$ for $j \in \seq{1}{J}$\;
-          $X_{u, n, j}^{P, m} \sim f_{X_{u, n}|X_{u, n-1}}(x_{u, n}|X_{u, n-1, j}^{F, m}; \Theta_{u, n, j}^{P, m})$ for $j \in \seq{1}{J}$\;
-          $w_{u, n, j}^m = f_{Y_{u, n}|X_{u, n}}(y_{u, n}^*|X_{u, n, j}^{P, m})$ for $j \in \seq{1}{J}$\;
-          Draw $k_{\seq{1}{j}}$ with $P(k_j = i) = w_{u, n, i}^m / \sum_{v = 1}^J w_{u, n, v}^m$ for $i, j \in \seq{1}{J}$\;
-          Set $\Theta_{u, n, j}^{F, m} = \Theta_{u, n, k_j}^{P, m} = \big(\Phi_{u, n, k_j}^{P, m}, \Psi_{u, n, k_j}^{P, m}\big)$ and $X_{u, n, j}^{F, m} = X_{u, n, k_j}^{P, m}$ for $j \in \seq{1}{J}$\;
+% \section{Block Panel Iterated Filter}
+% 
+% The panel iterated filter (PIF) algorithm of \citet{breto20} has been used to fit parameters of panelPOMP models in several studies \citep{domeyer22,wale2019,ranjeva17,ranjeva19}.
+% Like other iterated filtering algorithms, PIF iteratively combines particle filter calculations with parameter perturbations.
+% Theoretical results show that, as the perturbations are reduced over successive iterations, the perturbed parameters converge to the MLE.
+% 
+% In the case of a panelPOMP, the parameter vector $\theta$ may be composed of a set of shared parameters $\phi$, and a set of unit-specific parameters $\psi_{u}$, for units $u \in \seq{1}{U}$.
+% Here, we say that a parameter $\psi_{u}$ for $u \in \seq{1}{U}$ is unit-specific if it is not involved in the one-step transition density or the measurement model of any other unit $k \neq u \in \seq{1}{U}$, and shared parameters are those that are not unit-specific.
+% The PIF algorithm treats the parameter vector $\theta = (\phi, \psi_{1:U})$ as a complete entity and gives no distinction between shared and unit-specific parameters.
+% Even though the random walk standard deviation is set to zero for unit-specific parameters when considering data from other units, this lack of distinction may result in particle depletion for unit-specific parameters as the number of units $U$ grows large.
+% This is because it is likely that, when updating the parameters for a given unit, several parameter vector $\theta$ particles will not be resampled due to having low weights.
+% When the number of units is large, this will lead to a decreasing number of ancestral particles to sample unit-specific parameters $\psi_k$ as $k$ increases.
+% 
+% As a solution to this issue, we propose the block panel iterated filter (BPIF), described in Algorithm~\ref{alg:bpif}.
+% In this algorithm, shared $\phi$ and unit-specific parameters $\psi_{u}$ are handled differently when updating the parameter vector $\theta = (\phi, \psi_{1:U})$.
+% The BPIF updates $\theta$ by updating the shared vector $\phi$ in the same way that is done with PIF, but index resampling for unit-specific parameters $\psi_{1:U}$ only occurs when data from the corresponding
+% unit are being considered.
+% We have found that this modification leads to improved likelihood maximization performance, especially as the number of units increases.
+% No theoretical properties of this algorithm have yet been derived, but the approach has similarities to the block particle filter of \citet{rebeschini15} and its extension to parameter estimation \citep{ionides22,ning21ibpf}.
+% 
+% \begin{algorithm}[ht]
+%    \caption[BPIF]{\textbf{BPIF}. \\
+%     {\bf Inputs}:\\\hspace{\textwidth}
+%     Simulator of initial density, $f_{X_{u, 0}}(x_{u, 0}; \theta)$ for $u$ in $\seq{1}{U}$.\\\hspace{\textwidth}
+%     Simulator of transition density, $f_{X_{u, n}|X_{u, n-1}}(x_{u, n}|x_{u, n-1}; \theta)$ for $u$ in $\seq{1}{U}$, $n$ in $\seq{1}{N_u}$.\\\hspace{\textwidth}
+%     Evaluator of measurement density, $f_{Y_{u, n}|X_{u, n}}(y_{u, n}|x_{u, n} ; \theta)$ for $u$ in $\seq{1}{U}$, $n$ in $\seq{1}{N_u}$.\\\hspace{\textwidth}
+%     Data $y_{u, n}^*$,for $u$ in $\seq{1}{U}$, $n$ in $\seq{1}{N_u}$.\\\hspace{\textwidth}
+%     Number of iterations, $M$.\\\hspace{\textwidth}
+%     Number of particles, $J$.\\\hspace{\textwidth}
+%     Starting parameter swarm, $\Theta_j^0 = \big(\Phi_j^0, \Psi_{1:U, j}^0\big)$ for $j \in \seq{1}{J}$, $u \in \seq{1}{U}$.\\\hspace{\textwidth}
+%     Simulator of perturbation density, $h_{u, n}(\theta | \phi; \sigma)$ for $u \in \seq{1}{U}$, $n \in 0:N_u$.\\\hspace{\textwidth}
+%     Perturbation sequence, $\sigma_m$ for $m \in 1:M$.\\\hspace{\textwidth}
+%     {\bf Output:} \\\hspace{\textwidth}
+%     Final parameter swarm, $\Theta_{j}^{m} = \big(\Phi_j^m, \Psi_{\seq{1}{U}, j}^m\big)$ for $j \in \seq{1}{J}$, $u \in \seq{1}{U}$.
+%     \label{alg:bpif}}
+% %\BlankLine
+% \For{$m \in 1:M$}{
+%   Set $\Phi_{0, j}^m = \Phi_j^{m-1}$ for $j \in \seq{1}{J}$\;
+%     \For{$u \in \seq{1}{U}$}{
+%       Set $\Phi_{u, 0, j}^{F, m} \sim h_{u, 0}(\phi | \Phi_{0, j}^m; \sigma_m)$ for $j \in \seq{1}{J}$\;
+%       Set $\Psi_{u, 0, j}^{F, m} \sim h_{u, 0}(\psi | \Psi_{u, j}^{m - 1}; \sigma_m)$ for $j \in \seq{1}{J}$\;
+%       Set $\Theta_{u, 0, j}^{F, m} = \big(\Phi_{u, 0, j}^{F, m}, \Psi_{u, 0, j}^{F, m}\big)$ for $j \in \seq{1}{J}$\;
+%       Initialize $X_{u, 0, j}^{F, m} \sim f_{X_{u, 0}}(x_{u, 0}; \Theta_{u, 0, j}^{F, m})$ for $j \in \seq{1}{J}$\;
+%         \For{$n \in \seq{1}{N_u}$} {
+%           Set $\Theta_{u, n, j}^{P, m} \sim h_{u, n}(\theta|\Theta_{u, n-1, j}^{F, m}, \sigma_m)$ for $j \in \seq{1}{J}$\;
+%           $X_{u, n, j}^{P, m} \sim f_{X_{u, n}|X_{u, n-1}}(x_{u, n}|X_{u, n-1, j}^{F, m}; \Theta_{u, n, j}^{P, m})$ for $j \in \seq{1}{J}$\;
+%           $w_{u, n, j}^m = f_{Y_{u, n}|X_{u, n}}(y_{u, n}^*|X_{u, n, j}^{P, m})$ for $j \in \seq{1}{J}$\;
+%           Draw $k_{\seq{1}{j}}$ with $P(k_j = i) = w_{u, n, i}^m / \sum_{v = 1}^J w_{u, n, v}^m$ for $i, j \in \seq{1}{J}$\;
+%           Set $\Theta_{u, n, j}^{F, m} = \Theta_{u, n, k_j}^{P, m} = \big(\Phi_{u, n, k_j}^{P, m}, \Psi_{u, n, k_j}^{P, m}\big)$ and $X_{u, n, j}^{F, m} = X_{u, n, k_j}^{P, m}$ for $j \in \seq{1}{J}$\;
+%         }
+%       Set $\Theta_{u, j}^m = \big(\Phi_{u, N_u, j}^{F, m}, \Psi_{u, N_u, j}^{F, m}\big)$ for $j \in \seq{1}{J}$\;
+%     }
+%   Set $\Theta_j^m = \big(\Phi_{U, j}^m, \Psi_{1:U, j}^m\big)$ for $j \in \seq{1}{J}$\;
+% }
+% \end{algorithm}
+
+\section{Lee et al. (2020) Replication}
+
+In this article we claimed that we were able to obtain better fits to the observed data using the same models that were proposed by \cite{lee20}.
+Along with visual comparisons to the data, this claim was supported by comparing likelihoods and AIC values in Table~1.
+Because model likelihoods were not provided by \cite{lee20}, it is necessary to replicate these models in order to obtain likelihood estimates.
+Here we would like to thank the authors of \cite{lee20}, who provided detailed descriptions of their models, which enabled us to build on their work.
+In the following subsections, we use our \code{R} package \code{haitipkg} to reproduce some of the results of \cite{lee20}.
+This reproduction allows us to estimate the likelihoods of the \cite{lee20} version of Models~1--3, and also provides a demonstration on reproducibility.
+
+\subsection{Model~1 Replication}
+
+The model was implemented by a team at Johns Hopkins Bloomberg School of Public Health (hereafter referred to as the Model~1 authors) in the \code{R} programming language using the \code{pomp} package \citep{king16}.
+Original source code is publicly available with DOI: 10.5281/zenodo.3360991.
+Despite having source code available, no point estimates for model parameters were provided by \cite{lee20}.
+According to the supplement materials, this was because model realizations from a single parameter set retained substantial variability, but multiple realizations from a collection of parameter sets resulted in a reasonable visual fit to the data.
+We are also inclined to believe that the use of multiple parameter values was in part intended to account for parameter uncertainty (as mentioned in our main text), an effort by the Model~1 authors that we applaud.
+We note, however, that simulations from each of the parameter sets were treated with equal importance when being used to diagnose the model fit and make inference on the system.
+This practice is problematic given Figures S8 and S9 of the supplement material, which suggest that some parameter sets that were used for inference were several hundred log-likelihood units lower than the best performing sets of parameters.
+When accounting for parameter uncertainty, one should instead weigh resulting model projections based on the likelihood of the set of parameters that were used to obtain the model realizations, as was done by \cite{king15}.
+
+Not providing a point estimate for model parameters also has the effect of making model reproduction a more difficult task.
+We therefore rely on the range and median values for each of the final model parameters are provided in Tables S10 and S11 of the supplement, and bivariate relationships between some of the fitted parameters (Figures S8 and S9) in order to replicate their results.
+
+We begin by constructing 300 different parameter sets based on the basic summaries and figures provided by the Model~1 authors.
+This was primarily done using truncated-bivariate normal distributions, where samples that are drawn outside of the desired range of parameters are re-sampled from the distribution.
+Because the likelihood surface is complex and almost certainly includes higher than two-level interactions between parameter values, we then perform iterated filtering for the epidemic phase of the data using the 300 different parameter sets as starting points.
+
+<<Load lee Model1 Params, echo=FALSE>>=
+h1_epi <- haiti1()
+
+RL_dir <- paste0("run_level_", run_level, "/")
+lee1_epi_mif <- readRDS(file = paste0("../model1/", RL_dir, "lee1_epi_fit.rds"))
+@
+
+<<Simulate Lee Model 1, echo=FALSE, cache=TRUE>>=
+N_LEE_SIMS <- switch(run_level,  10,  15,   20)
+
+registerDoRNG(18599687)
+foreach(
+  i = 1:length(lee1_epi_mif),
+  .combine = rbind
+) %dopar% {
+  guess <- lee1_epi_mif$lee1_epi_mif_params[i, ]
+  sims <- simulate(h1_epi, nsim = N_LEE_SIMS, format = 'data.frame',
+                   params = guess)  # 20 sims for each set of parameters
+  sims$param_set <- i
+  sims %>%
+    dplyr::select(.id, week, cases, param_set)
+} -> all_sims
+
+@
+
+<<PlotEpiDist, echo=FALSE, fig.cap="Compare to Figure S8 of Lee et al. (2020) supplement.", fig.width=7, fig.height=7, fig.align='center'>>=
+lee1_epi_evals <- readRDS(file = paste0("../model1/", RL_dir, "lee1_epi_evals.rds"))
+
+epi_results <- as.data.frame(lee1_epi_mif$lee1_epi_mif_params)
+epi_results$logLik <- lee1_epi_evals$pfLL
+
+GGally::ggpairs(
+  dplyr::filter(epi_results, logLik > -20000),
+  columns = c("rho", "tau", "beta1", "nu", "logLik")
+)
+@
+
+<<SimulateEpi, echo=FALSE, include=FALSE, message=FALSE, cache=TRUE>>=
+h1_end <- haiti1(period = "endemic")
+
+lee1_end_mif <- readRDS(
+  file = paste0("../model1/", RL_dir, "lee1_end_fit.rds")
+)
+
+true_agg_cases <- haitiCholera %>%
+  select(-report) %>%
+  pivot_longer(
+    data = .,
+    cols = -date_saturday,
+    values_to = 'cases',
+    names_to = 'dep'
+  ) %>%
+  mutate(
+    date = as.Date(date_saturday),
+    dep = gsub("\\.", "_", dep)
+  ) %>%
+  mutate(
+    dep = case_when(dep == "Grand_Anse" ~ "Grande_Anse", TRUE ~ dep)
+  ) %>%
+  tidyr::pivot_wider(
+    data = .,
+    id_cols = c(date),
+    names_from = dep,
+    values_from = cases,
+    names_prefix = 'cases_'
+  ) %>%
+  mutate(
+    ReportedAll = cases_Artibonite + cases_Centre +
+      cases_Grande_Anse + cases_Nippes + cases_Nord +
+      cases_Nord_Est + cases_Ouest + cases_Sud +
+      cases_Sud_Est + cases_Nord_Ouest
+  )
+@
+
+Following the model fitting scheme used by the Model~1 authors, we then use these resulting parameters as starting values for the endemic phase, and refit parameters using MIF2.
+To do this, we use the same number of particles and MIF iterations that were used by \citet{lee20}.
+The resulting bivariate relations between $\rho, \tau, \beta_1$ and $\nu$ are given in Figure~\ref{fig:plotEndParams}.
+
+<<FitEnd, echo=FALSE, message=FALSE, include=FALSE>>=
+foreach(
+  i = 1:nrow(lee1_end_mif),
+  .combine = rbind
+) %dopar% {
+  r_params <- lee1_end_mif[i, ]
+
+  lee1_epi_mif$start_states %>%
+    filter(param_set == i) %>%
+    select(-param_set) %>%
+    sapply(., median, na.rm = TRUE) -> rinit_parms
+
+  # Set the initial state
+  new_rinit <- Csnippet(
+    sprintf(
+      "
+          S = nearbyint(%f);
+          E = nearbyint(%f);
+          I = nearbyint(%f);
+          A = nearbyint(%f);
+          R = nearbyint(%f);
+          incid = nearbyint(%f);
+          foival = %f;
+          Str0 = nearbyint(%f);
+          Sout = nearbyint(%f);
+          Sin = nearbyint(%f);
+          ",
+      rinit_parms['S'],
+      rinit_parms['E'],
+      rinit_parms['I'],
+      rinit_parms['A'],
+      rinit_parms['R'],
+      rinit_parms['incid'],
+      rinit_parms['foival'],
+      rinit_parms['Str0'],
+      rinit_parms['Sout'],
+      rinit_parms['Sin']
+    )
+  )
+
+  h1_end <- h1_end %>% pomp(
+    rinit = new_rinit,
+    statenames = c(
+      "S", "E", "I", "A", "R", "incid",
+      "foival", "Str0", "Sout", "Sin"
+    )
+  )
+
+  coef(h1_end) <- r_params
+
+  sims <- simulate(
+    h1_end, nsim = N_LEE_SIMS,
+    format = 'data.frame',
+    params = r_params
+  )
+  sims$param_set <- i
+  sims %>%
+    dplyr::select(.id, week, cases, param_set)
+} -> end_sims
+
+@
+
+<<plotEndParams, fig.width=7, fig.height=7, fig.align='center', fig.cap="Bivariate relationships between variables after fitting endemic period.
+Compare to S9 of Lee et al.
+(2020) supplement.", cache=TRUE, echo=FALSE>>=
+lee1_end_evals <- readRDS(file = paste0("../model1/", RL_dir, "lee1_end_evals.rds"))
+
+end_results <- as.data.frame(lee1_end_mif)
+end_results$logLik <- lee1_epi_evals$pfLL
+
+GGally::ggpairs(
+  filter(end_results, logLik > -20000),
+  columns = c("rho", "tau", "beta1", "nu", "logLik")
+)
+@
+
+Now that we have fit the parameters, we can simulate from the models.
+The results of these simulations are shown in Figure~\ref{fig:plotMod1Sims}.
+While not a perfect replication of the results given by \citet{lee20sup}, the similarities between the model simulations, and the fact that we based our implementation of Model~1 primarily on the source code that was provided, makes us confident that we have been able to obtain a reliable representation of the model described by \citet{lee20}.
+
+<<plotMod1Sims, echo=FALSE, fig.cap="Simulations from various parameter sets.
+Simulations from parameter sets with likelihood $<$ -2200 are removed.
+Compare to Figure S7 of \\cite{lee20sup}.", cache=TRUE, fig.width=6.5, fig.height=3.75>>=
+
+epi_complete <- filter(epi_results, logLik > -20000)
+bad_epi_pars <- which(epi_results$logLik < -20000)
+
+quants <- all_sims %>%
+  filter(!param_set %in% bad_epi_pars) %>%
+  group_by(week) %>%
+  summarize(
+    q025 = quantile(cases, probs = 0.025, na.rm = TRUE),
+    q50  = quantile(cases, probs = 0.500, na.rm = TRUE),
+    q975 = quantile(cases, probs = 0.975, na.rm = TRUE)
+  ) %>%
+  ungroup() %>%
+  mutate(date = lubridate::ymd("2010-10-16") + lubridate::weeks(week))
+
+end_complete <- filter(end_results, logLik > -20000)
+bad_end_pars <- which(end_results$logLik < -20000)
+
+quants_end <- end_sims %>%
+  filter(!param_set %in% bad_end_pars) %>%
+  group_by(week) %>%
+  summarize(
+    q025 = quantile(cases, probs = 0.025, na.rm = TRUE),
+    q50  = quantile(cases, probs = 0.500, na.rm = TRUE),
+    q975 = quantile(cases, probs = 0.975, na.rm = TRUE)
+  ) %>%
+  ungroup() %>%
+  mutate(date = lubridate::ymd("2010-10-16") + lubridate::weeks(week))
+
+ggplot() +
+  geom_point(data = true_agg_cases, aes(x = date, y = ReportedAll)) +
+  geom_line(data = quants, aes(x = date, y = q50), col = '#0b8a2b', size = 1) +
+  geom_ribbon(
+    data = quants,
+    aes(x = date, ymin = q025, ymax = q975),
+    alpha = 0.5, fill = '#0b9c30'
+  ) +
+  geom_line(data = quants_end, aes(x = date, y = q50), col = '#882871') +
+  geom_ribbon(data = quants_end, aes(x = date, ymin = q025, ymax = q975), alpha = 0.5, fill = '#882871') +
+  ylab("Reported, Symptomatic Incidence") +
+  scale_x_date(date_breaks = "1 year") +
+  theme(axis.title.x = element_blank(),
+        axis.text.x = element_text(hjust = 1, angle = 45))
+@
+
+\subsection{Model~2 Replication}\label{sec:mod2rep}
+
+Model~2 was developed by a team that consisted of members from the Fred Hutchinson Cancer Research Center and the University of Florida (hereafter referred to as the Model~2 authors).
+While Model~2 is the only deterministic model we considered in our analysis, it contains perhaps the most complex descriptions of cholera in Haiti: Model~2 accounts for movement between spatial units; human-to-human and environment-to-human cholera infections; and transfer of water between spatial units based on elevation charts and river flows.
+
+The source code that the Model~2 authors used to generate their results was written in the \code{Python} programming language, and is publicly available at \url{10.5281/zenodo.3360857} and its accompanying GitHub repository \url{https://github.com/lulelita/HaitiCholeraMultiModelingProject}.
+In order to perform our analysis in a unified framework, we re-implemented this model in the \code{R} programming language using the \code{spatPomp} package \citep{asfaw21github}, which facilitates the creation of meta-population models.
+We note that the travel and water matrices used to implement the complex dynamics in Model~2 \citep{lee20sup} are not available in either the Zenodo archive or the GitHub repository;
+instead, we obtained these matrices via personal correspondence with the Model~2 authors.
+Using these matrices, and the point estimates for model parameters provided by \citep{lee20sup}, we created trajectories of the cholera dynamics and compared this to available data.
+These trajectories, shown in Figure~\ref{fig:mod2rep}, are very similar to the trajectories shown in Figure~S15 of \cite{lee20sup}.
+
+<<Model2Replication, cache=TRUE, echo=FALSE>>=
+h2_epi <- haiti2()
+h2_end <- haiti2(region = 'after')
+
+h2_epi_traj <- trajectory(h2_epi, format = 'data.frame')
+h2_end_traj <- trajectory(h2_end, format = 'data.frame')
+
+h2_epi_traj$Ctotal <- rowSums(h2_epi_traj[, paste0("C", 1:10)]) * 0.2
+h2_end_traj$Ctotal <- rowSums(h2_end_traj[, paste0("C", 1:10)]) * 0.2
+
+dep_plot_df <- haitipkg::haitiCholera %>%
+  select(-report) %>%
+  tidyr::pivot_longer(
+    data = .,
+    cols = -date_saturday,
+    values_to = 'cases',
+    names_to = 'dep'
+  ) %>%
+  mutate(
+    date = as.Date(date_saturday),
+    dep = gsub("\\.", "_", dep)
+  ) %>%
+  mutate(
+    dep = case_when(dep == "Grand_Anse" ~ "Grande_Anse", TRUE ~ dep)
+  )
+
+true_agg_cases <- dep_plot_df %>%
+  tidyr::pivot_wider(
+    data = .,
+    id_cols = c(date),
+    names_from = dep,
+    values_from = cases,
+    names_prefix = 'cases_'
+  ) %>%
+  mutate(
+    ReportedAll = cases_Artibonite + cases_Centre +
+      cases_Grande_Anse + cases_Nippes + cases_Nord +
+      cases_Nord_Est + cases_Ouest + cases_Sud +
+      cases_Sud_Est + cases_Nord_Ouest
+  ) %>%
+  select(date, ReportedAll)
+
+h2_epi_traj$date <- as.Date(
+  lubridate::round_date(
+    lubridate::date_decimal(h2_epi_traj$year),
+    unit = 'day'
+  )
+)
+
+h2_end_traj$date <- as.Date(
+  lubridate::round_date(
+    lubridate::date_decimal(h2_end_traj$year),
+    unit = 'day'
+  )
+)
+
+all_traj <- bind_rows(
+  select(h2_epi_traj, date, Ctotal),
+  select(h2_end_traj, date, Ctotal)
+)
+
+projections <- left_join(
+  x = true_agg_cases,
+  y = all_traj,
+  by = 'date'
+)
+@
+
+\begin{figure}[ht]
+<<plotModel2Rep, cache=TRUE, echo=FALSE, fig.width=6.5, fig.height=4.5>>=
+gg_epi <- projections %>%
+  filter(date < as.Date("03-01-2014", format = '%m-%d-%Y')) %>%
+  ggplot(aes(x = date)) +
+  geom_line(aes(y = ReportedAll), col = 'black', linetype = 'dashed') +
+  geom_line(aes(y = Ctotal), col = 'red', size = 1) +
+  theme_bw() +
+  scale_y_continuous(limits = c(0, 30000), breaks = seq(0, 30000, 5000)) +
+  theme(axis.title.x = element_blank()) +
+  ylab("Reported number of cases")
+
+gg_end <- projections %>%
+  filter(date >= as.Date("03-01-2014", format = '%m-%d-%Y')) %>%
+  ggplot(aes(x = date)) +
+  geom_line(aes(y = ReportedAll), col = 'black', linetype = 'dashed') +
+  geom_line(aes(y = Ctotal), col = 'red', size = 1) +
+  theme_bw() +
+  scale_y_continuous(limits = c(0, 7000), breaks = seq(0, 7000, 1000)) +
+  theme(axis.title.x = element_blank()) +
+  ylab("Reported number of cases") +
+  scale_x_date(date_breaks = "1 year", date_labels = "%Y")
+
+cowplot::plot_grid(gg_epi, gg_end, ncol = 1)
+@
+\caption{\label{fig:mod2rep}
+Model 2 trajectories using the \code{haitipkg}. Compare to Figure S15 of \cite{lee20sup}.
 }
-      Set $\Theta_{u, j}^m = \big(\Phi_{u, N_u, j}^{F, m}, \Psi_{u, N_u, j}^{F, m}\big)$ for $j \in \seq{1}{J}$\;
+\end{figure}
+
+There are minor differences between Figure~\ref{fig:mod2rep} and Figure~S15 of \cite{lee20sup}.
+% such as a delayed second wave in the epidemic period and a larger initial wave during the endemic period.
+While the discrepancy appears minor, the deterministic nature of Model~2 implies that an exact replication of model trajectories should be possible.
+In this case, these discrepancies may be attributed to model implementation and small differences in model parameters.
+Details about the measurement models and how latent states were initialized for the epidemic model were not provided by \citet{lee20sup} and therefore these details must be inferred by looking at the provided source code.
+According to repository comments, the files \code{fitInPieces3paramsCleanMay2019Public.py} and \code{fitInPiecesMuWithFracSusFixedAllInfectionsPublic.py} were used to fit the epidemic and endemic phases of the model respectively, although it is apparent that these exact files were not used to obtain the reported results.
+These files contain errors that making them impossible to run, and therefore these files cannot be the code that generated the reported model fits\footnote{One example of why the code cannot be run that the file loads functions from a non-extant file named \code{choleraEqs.py} in line 13 rather than \code{choleraEqsPublic.py}.}.
+The inability to replicate the results by \citet{lee20} by running the provided source code makes checking whether or not a our numeric implementation faithfully represents their results very difficult.
+% The difficulty of reproducing the result of \citet{lee20} is coupled with a misunderstanding of how these results were originally obtained.
+
+\jwc{PROBABLY DELETE THIS NEXT BIT, BUT IT'S USEFUL TO DOCUMENT IT FOR NOW...}
+One concern that arises when inspecting the files \code{fitInPieces3paramsCleanMay2019Public.py} and \code{fitInPiecesMuWithFracSusFixedAllInfectionsPublic.py} is that the output of these files (which contains the estimated parameters) was not referenced/used in either the file that generates the sensitivity analysis or the file that generates the model simulations; instead, alternative (non-extant) files with similar names are loaded, suggesting parameters that were used to generate the final results were not a direct result of the estimated parameters.
+Furthermore, inspection of the script \code{simulateHaitiFullTimeInPiecesAllVaccinationScenariosSingleParameterVectorAllInfectionsPublic.py}---which is used to obtain simulations for each of the vaccination scenarios---results in a number of concerns.
+First, the script uses parameters from the pickle file \code{leastSquaresDataUpTo2019/fitEpidemicPiece\_resultsMatrix\_18Mar2019.pickle}, which only appears in the repository as a commented out line of code in the file \code{generateSetsOfParametersForSensitivityAnalysisPublic.py}, suggesting that the set of parameters used to simulate each vaccination scenario was not the most recent set of parameters generated by the sensitivity analysis.
+Second, the script appears to use the set of differential equations from the function \code{choleraEqs11WithVaccinationNetwork} for the epidemic phase, even though parameter values for this phase were estimated using \code{choleraEqs10WithVaccinationNetwork}.
+It is possible, however, that we simply are interpreting the code incorrectly and that these concerns are unwarranted.
+Furthermore, because the provided code is non-functional, it is possible that the code that was actually used to obtain the reported results does not include these potential errors.
+This highlights the importance of providing the exact source code that was used to obtain the reported results of any analysis.
+
+Another potential explanation for the discrepancy between our simulations from Model~2 and those of \citet{lee20} is that the parameters that we used are only approximately the same as those used by \citet{lee20sup}.
+For example, the parameters $\transmission$, $\Wbeta{}$ (See Table~\ref{tab:translate}) had reported values of $9.9 \times 10^{-7}$ and $4.03 \times 10^{-2}$, respectively (Table~S13 of \cite{lee20sup}), but were actually fit to data and therefore likely had values that were far more precise.
+Additionally, our implementation of Model~2 used a time scale of years and many of the parameters were reported on a weekly scale, so small differences may result due to unit conversions.
+The collective effect of these small differences in model parameters likely will result in differences in model trajectories.
+This issue could have been mitigated if the saved files containing estimated model parameters were included as part of the code submission.
+
+\subsection{Model~3 Replication}
+
+Model~3 was developed by a team of researchers at the Laboratory of the Swiss Federal Institute of Technology in Lausanne, hereafter referred to as the Model~3 authors.
+The code that was originally used to implement Model~3 is archived with the DOI: \url{10.5281/zenodo.3360723}, and also available in the public GitHub repository: \code{jcblemai/haiti-mass-ocv-campaign}.
+Because the code was made publicly available, and final model parameters were reported in the supplementary material of \cite{lee20}, we were able to reproduce Model~3 by directly using the source code.
+In Fig.~\ref{fig:mod3rep}, we plot simulations from this model.
+This figure should be compared to Figure S18 of \cite{lee20}.
+We note that slight differences may be accounted for due to variance in the model simulations and the difference in programming language used to produce the figure.
+
+<<Model3Replication, echo=FALSE, cache=TRUE>>=
+NSIM <- switch(run_level, 50, 100, 1000)
+
+plot_order <- c(
+  'Artibonite',
+  'Sud_Est',
+  'Nippes',
+  'Nord_Est',
+  'Ouest',
+  'Centre',
+  'Nord',
+  'Sud',
+  'Nord_Ouest',
+  'Grande_Anse'
+)
+
+dep_labeller <- as_labeller(
+  c(
+    'Artibonite' = 'Artibonite',
+    'Sud_Est' = 'Sud-Est',
+    'Sud.Est' = 'Sud-Est',
+    'Nippes' = 'Nippes',
+    'Nord_Est' = 'Nord-Est',
+    'Nord.Est' = 'Nord-Est',
+    'Ouest' = 'Ouest',
+    'Centre' = 'Centre',
+    'Nord' = 'Nord',
+    'Sud' = 'Sud',
+    'Nord_Ouest' = 'Nord-Ouest',
+    'Nord.Ouest' = 'Nord-Ouest',
+    'Grande_Anse' = 'Grand\'Anse',
+    'Grand.Anse' = 'Grand\'Anse'
+  )
+)
+
+lee3 <- haiti3_lee_et_al()
+lee3_sims <- simulate(lee3, nsim = NSIM, seed = 321, format = 'data.frame')
+
+sub_cases <- haitiCholera %>%
+  dplyr::mutate(date = as.Date(date_saturday)) %>%
+  dplyr::select(-report, -date_saturday) %>%
+  dplyr::filter(lubridate::decimal_date(date) >= min(lee3_sims$time)) %>%
+  tidyr::pivot_longer(
+    data = .,
+    cols = -date,
+    names_to = "dep",
+    values_to = "cases"
+  ) %>%
+  dplyr::mutate(
+    dep = dplyr::case_when(
+      dep == "Grand.Anse" ~ "Grande_Anse",
+      dep == "Nord.Est" ~ "Nord_Est",
+      dep == "Nord.Ouest" ~ "Nord_Ouest",
+      dep == "Sud.Est" ~ "Sud_Est",
+      TRUE ~ dep
+    )
+  )
+
+quants <- lee3_sims %>%
+  dplyr::select(-CasesAll) %>%
+  tidyr::pivot_longer(
+    data = .,
+    cols = starts_with("cases"),
+    names_prefix = "cases",
+    names_to = "dep",
+    values_to = "cases"
+  ) %>%
+  dplyr::group_by(dep, time) %>%
+  dplyr::summarise(
+    q05 = quantile(cases, 0.025, na.rm = T),
+    mean = mean(cases, na.rm = T),
+    q50 = quantile(cases, 0.5, na.rm = T),
+    q95 = quantile(cases, 0.975, na.rm = T)
+  ) %>%
+  dplyr::ungroup() %>%
+  dplyr::mutate(
+    date = lubridate::date_decimal(time),
+    date = as.Date(lubridate::round_date(date))
+  )
+
+quants <- quants %>%
+  dplyr::filter(
+    !(dep == "Ouest" & date < '2017-06-10')
+  )
+@
+
+\begin{figure}[ht]
+<<PlotMod3Rep, echo=FALSE, message=FALSE, fig.height=3.75>>=
+ggplot() +
+  geom_ribbon(data = quants, aes(x = date, ymin = q05, ymax = q95),
+              fill = 'darkblue', alpha = 0.5) +
+  geom_line(data = quants, aes(x = date, y = q50), col = 'darkblue') +
+  geom_point(data = dplyr::filter(sub_cases, cases < 500), aes(x = date, y = cases), size = 0.2) +
+  facet_wrap(~factor(dep, levels = plot_order), nrow = 2, labeller = dep_labeller) +
+  labs(y = 'Reported Cholera Cases') +
+  scale_x_date(date_breaks = "6 months", date_labels = "%Y-%m") +
+  theme(
+    axis.text.x = element_text(angle = 40, hjust = 1, vjust = 1, size = 7),
+    axis.title.x = element_blank()
+  )
+@
+\caption{\label{fig:mod3rep}
+Simulations from Model~3.
+Compare to Figure S18 of \cite{lee20}.
 }
-  Set $\Theta_j^m = \big(\Phi_{U, j}^m, \Psi_{1:U, j}^m\big)$ for $j \in \seq{1}{J}$\;
-}
-\end{algorithm}
+\end{figure}
+
+<<lee3Cleanup, echo=FALSE, include=FALSE, message=FALSE>>=
+rm(lee3, lee3_sims, quants, sub_cases)
+gc()
+@
 
@@
 
-Since the models in \citet{lee20} were developed independently, the choice of notation varies inconsistently between models.
+Since the models of \citet{lee20} were developed independently, the choice of notation varies inconsistently between models.
 For our reanalysis, we rename parameters to provide a unified notation facilitating comparison between models.
@@
   \caption{
-  \label{tab:translate}Translations between our common notation and notation used in \cite{lee20}
+  \label{tab:translate}Translations between our common notation and notation used by \cite{lee20}
   }
@@
 
-\section{Lee et al. (2020) Replication}
-
-In this article we claimed that we were able to obtain better fits to the observed data using the same models that were proposed in \cite{lee20}.
-Along with visual comparisons to the data, this claim was supported by comparing likelihoods and AIC values in Table~1.
-Because model likelihoods were not provided in \cite{lee20}, it is necessary to replicate these models in order to obtain likelihood estimates.
-Here we would like to thank the authors of \cite{lee20}, who provided detailed descriptions of their models, which enabled us to build on their work.
-In the following subsections, we use our \code{R} package \code{haitipkg} to reproduce some of the results of \cite{lee20}.
-This reproduction allows us to estimate the likelihoods of the \cite{lee20} version of Models~1--3, and also provides a demonstration on reproducibility.
-
-\subsection{Model~1 Replication}
-
-The model was implemented by a team at Johns Hopkins Bloomberg School of Public Health (hereafter referred to as the Model~1 authors) in the \code{R} programming language using the \code{pomp} package \citep{king16}.
-Original source code is publicly available with DOI: 10.5281/zenodo.3360991.
-Despite having source code available, no point estimates for model parameters were provided in \cite{lee20}.
-According to the supplement materials, this was because model realizations from a single parameter set retained substantial variability, but multiple realizations from a collection of parameter sets resulted in a reasonable visual fit to the data.
-We are also inclined to believe that the use of multiple parameter values was in part intended to account for parameter uncertainty (as mentioned in our main text), an effort by the Model~1 authors that we applaud.
-We note, however, that simulations from each of the parameter sets were treated with equal importance when being used to diagnose the model fit and make inference on the system.
-This practice is problematic given Figures S8 and S9 of the supplement material, which suggest that some parameter sets that were used for inference were several hundred log-likelihood units lower than the best performing sets of parameters.
-When accounting for parameter uncertainty, one should instead weigh resulting model projections based on the likelihood of the set of parameters that were used to obtain the model realizations, as was done in \cite{king15}.
-
-Not providing a point estimate for model parameters also has the effect of making model reproduction a more difficult task.
-We therefore rely on the range and median values for each of the final model parameters are provided in Tables S10 and S11 of the supplement, and bivariate relationships between some of the fitted parameters (Figures S8 and S9) in order to replicate their results.
-
-We begin by constructing 300 different parameter sets based on the basic summaries and figures provided by the Model~1 authors.
-This was primarily done using truncated-bivariate normal distributions, where samples that are drawn outside of the desired range of parameters are re-sampled from the distribution.
-Because the likelihood surface is complex and almost certainly includes higher than two-level interactions between parameter values, we then perform iterated filtering for the epidemic phase of the data using the 300 different parameter sets as starting points.
-
-<<Load lee Model1 Params, echo=FALSE>>=
-h1_epi <- haiti1()
-
-RL_dir <- paste0("run_level_", run_level, "/")
-lee1_epi_mif <- readRDS(file = paste0("../model1/", RL_dir, "lee1_epi_fit.rds"))
-@
-
-<<Simulate Lee Model 1, echo=FALSE, cache=TRUE>>=
-N_LEE_SIMS <- switch(run_level,  10,  20,   20)
-
-registerDoRNG(18599687)
-foreach(
-  i = 1:length(lee1_epi_mif),
-  .combine = rbind
-) %dopar% {
-  guess <- lee1_epi_mif$lee1_epi_mif_params[i, ]
-  sims <- simulate(h1_epi, nsim = N_LEE_SIMS, format = 'data.frame',
-                   params = guess)  # 20 sims for each set of parameters
-  sims$param_set <- i
-  sims %>%
-    dplyr::select(.id, week, cases, param_set)
-} -> all_sims
-
-@
-
-<<PlotEpiDist, echo=FALSE, fig.cap="Compare to Figure S8 of Lee et al. (2020) supplement.", fig.width=7, fig.height=7, fig.align='center'>>=
-lee1_epi_evals <- readRDS(file = paste0("../model1/", RL_dir, "lee1_epi_evals.rds"))
-
-epi_results <- as.data.frame(lee1_epi_mif$lee1_epi_mif_params)
-epi_results$logLik <- lee1_epi_evals$pfLL
-
-GGally::ggpairs(
-  dplyr::filter(epi_results, logLik > -20000),
-  columns = c("rho", "tau", "beta1", "nu", "logLik")
-)
-@
-
-<<SimulateEpi, echo=FALSE, include=FALSE, message=FALSE, cache=TRUE>>=
-h1_end <- haiti1(period = "endemic")
-
-lee1_end_mif <- readRDS(
-  file = paste0("../model1/", RL_dir, "lee1_end_fit.rds")
-)
-
-true_agg_cases <- haitiCholera %>%
-  select(-report) %>%
-  pivot_longer(
-    data = .,
-    cols = -date_saturday,
-    values_to = 'cases',
-    names_to = 'dep'
-  ) %>%
-  mutate(
-    date = as.Date(date_saturday),
-    dep = gsub("\\.", "_", dep)
-  ) %>%
-  mutate(
-    dep = case_when(dep == "Grand_Anse" ~ "Grande_Anse", TRUE ~ dep)
-  ) %>%
-  tidyr::pivot_wider(
-    data = .,
-    id_cols = c(date),
-    names_from = dep,
-    values_from = cases,
-    names_prefix = 'cases_'
-  ) %>%
-  mutate(
-    ReportedAll = cases_Artibonite + cases_Centre +
-      cases_Grande_Anse + cases_Nippes + cases_Nord +
-      cases_Nord_Est + cases_Ouest + cases_Sud +
-      cases_Sud_Est + cases_Nord_Ouest
-  )
-@
-
-Following the model fitting scheme used by the Model~1 authors, we then use these resulting parameters as starting values for the endemic phase, and refit parameters using MIF2.
-To do this, we use the same number of particles and MIF iterations that were used in \citet{lee20}.
-The resulting bivariate relations between $\rho, \tau, \beta_1$ and $\nu$ are given in Figure~\ref{fig:plotEndParams}.
-
-<<FitEnd, echo=FALSE, message=FALSE, include=FALSE>>=
-foreach(
-  i = 1:nrow(lee1_end_mif),
-  .combine = rbind
-) %dopar% {
-  r_params <- lee1_end_mif[i, ]
-
-  lee1_epi_mif$start_states %>%
-    filter(param_set == i) %>%
-    select(-param_set) %>%
-    sapply(., median, na.rm = TRUE) -> rinit_parms
-
-  # Set the initial state
-  new_rinit <- Csnippet(
-    sprintf(
-      "
-          S = nearbyint(%f);
-          E = nearbyint(%f);
-          I = nearbyint(%f);
-          A = nearbyint(%f);
-          R = nearbyint(%f);
-          incid = nearbyint(%f);
-          foival = %f;
-          Str0 = nearbyint(%f);
-          Sout = nearbyint(%f);
-          Sin = nearbyint(%f);
-          ",
-      rinit_parms['S'],
-      rinit_parms['E'],
-      rinit_parms['I'],
-      rinit_parms['A'],
-      rinit_parms['R'],
-      rinit_parms['incid'],
-      rinit_parms['foival'],
-      rinit_parms['Str0'],
-      rinit_parms['Sout'],
-      rinit_parms['Sin']
-    )
-  )
-
-  h1_end <- h1_end %>% pomp(
-    rinit = new_rinit,
-    statenames = c(
-      "S", "E", "I", "A", "R", "incid",
-      "foival", "Str0", "Sout", "Sin"
-    )
-  )
-
-  coef(h1_end) <- r_params
-
-  sims <- simulate(
-    h1_end, nsim = N_LEE_SIMS,
-    format = 'data.frame',
-    params = r_params
-  )
-  sims$param_set <- i
-  sims %>%
-    dplyr::select(.id, week, cases, param_set)
-} -> end_sims
-
-@
-
-<<plotEndParams, fig.width=7, fig.height=7, fig.align='center', fig.cap="Bivariate relationships between variables after fitting endemic period.
-Compare to S9 of Lee et al.
-(2020) supplement.", cache=TRUE, echo=FALSE>>=
-lee1_end_evals <- readRDS(file = paste0("../model1/", RL_dir, "lee1_end_evals.rds"))
-
-end_results <- as.data.frame(lee1_end_mif)
-end_results$logLik <- lee1_epi_evals$pfLL
-
-GGally::ggpairs(
-  filter(end_results, logLik > -20000),
-  columns = c("rho", "tau", "beta1", "nu", "logLik")
-)
-@
-
-Now that we have fit the parameters, we can simulate from the models.
-The results of these simulations are shown in Figure~\ref{fig:plotMod1Sims}.
-While not a perfect replication of the results given in \citet{lee20sup}, the similarities between the model simulations, and the fact that we based our implementation of Model~1 primarily on the source code that was provided, makes us confident that we have been able to obtain a reliable representation of the model described in \citet{lee20}.
-
-<<plotMod1Sims, echo=FALSE, fig.cap="Simulations from various parameter sets.
-Simulations from parameter sets with likelihood $<$ -2200 are removed.
-Compare to Figure S7 of \\cite{lee20sup}.", cache=TRUE, fig.width=6.5, fig.height=3.75>>=
-
-epi_complete <- filter(epi_results, logLik > -20000)
-bad_epi_pars <- which(epi_results$logLik < -20000)
-
-quants <- all_sims %>%
-  filter(!param_set %in% bad_epi_pars) %>%
-  group_by(week) %>%
-  summarize(
-    q025 = quantile(cases, probs = 0.025, na.rm = TRUE),
-    q50  = quantile(cases, probs = 0.500, na.rm = TRUE),
-    q975 = quantile(cases, probs = 0.975, na.rm = TRUE)
-  ) %>%
-  ungroup() %>%
-  mutate(date = lubridate::ymd("2010-10-16") + lubridate::weeks(week))
-
-end_complete <- filter(end_results, logLik > -20000)
-bad_end_pars <- which(end_results$logLik < -20000)
-
-quants_end <- end_sims %>%
-  filter(!param_set %in% bad_end_pars) %>%
-  group_by(week) %>%
-  summarize(
-    q025 = quantile(cases, probs = 0.025, na.rm = TRUE),
-    q50  = quantile(cases, probs = 0.500, na.rm = TRUE),
-    q975 = quantile(cases, probs = 0.975, na.rm = TRUE)
-  ) %>%
-  ungroup() %>%
-  mutate(date = lubridate::ymd("2010-10-16") + lubridate::weeks(week))
-
-ggplot() +
-  geom_point(data = true_agg_cases, aes(x = date, y = ReportedAll)) +
-  geom_line(data = quants, aes(x = date, y = q50), col = '#0b8a2b', size = 1) +
-  geom_ribbon(
-    data = quants,
-    aes(x = date, ymin = q025, ymax = q975),
-    alpha = 0.5, fill = '#0b9c30'
-  ) +
-  geom_line(data = quants_end, aes(x = date, y = q50), col = '#882871') +
-  geom_ribbon(data = quants_end, aes(x = date, ymin = q025, ymax = q975), alpha = 0.5, fill = '#882871') +
-  ylab("Reported, Symptomatic Incidence") +
-  scale_x_date(date_breaks = "1 year") +
-  theme(axis.title.x = element_blank(),
-        axis.text.x = element_text(hjust = 1, angle = 45))
-@
-
-\subsection{Model~2 Replication}\label{sec:mod2rep}
-
-Model~2 was developed by a team that consisted of members from the Fred Hutchinson Cancer Research Center and the University of Florida (hereafter referred to as the Model~2 authors).
-While Model~2 is the only deterministic model we considered in our analysis, it contains perhaps the most complex descriptions of cholera in Haiti: Model~2 accounts for movement between spatial units; human-to-human and environment-to-human cholera infections; and transfer of water between spatial units based on elevation charts and river flows.
-
-The source code that the Model~2 authors used to generate their results was written in the \code{Python} programming language, and is publicly available at \url{10.5281/zenodo.3360857} and its accompanying GitHub repository \url{https://github.com/lulelita/HaitiCholeraMultiModelingProject}.
-In order to perform our analysis in a unified framework, we re-implemented this model in the \code{R} programming language using the \code{spatPomp} package \citep{asfaw21github}, which facilitates the creation of meta-population models.
-We note that the travel and water matrices used to implement the complex dynamics in Model~2 \citep{lee20sup} are not available in either the Zenodo archive or the GitHub repository;
-instead, we obtained these matrices via personal correspondence with the Model~2 authors.
-Using these matrices, and the point estimates for model parameters listed in \citep{lee20sup}, we created trajectories of the cholera dynamics and compared this to available data.
-These trajectories, shown in Figure~\ref{fig:mod2rep}, are very similar to the trajectories shown in Figure~S15 of \cite{lee20sup}.
-
-<<Model2Replication, cache=TRUE, echo=FALSE>>=
-h2_epi <- haiti2()
-h2_end <- haiti2(region = 'after')
-
-h2_epi_traj <- trajectory(h2_epi, format = 'data.frame')
-h2_end_traj <- trajectory(h2_end, format = 'data.frame')
-
-h2_epi_traj$Ctotal <- rowSums(h2_epi_traj[, paste0("C", 1:10)]) * 0.2
-h2_end_traj$Ctotal <- rowSums(h2_end_traj[, paste0("C", 1:10)]) * 0.2
-
-dep_plot_df <- haitipkg::haitiCholera %>%
-  select(-report) %>%
-  tidyr::pivot_longer(
-    data = .,
-    cols = -date_saturday,
-    values_to = 'cases',
-    names_to = 'dep'
-  ) %>%
-  mutate(
-    date = as.Date(date_saturday),
-    dep = gsub("\\.", "_", dep)
-  ) %>%
-  mutate(
-    dep = case_when(dep == "Grand_Anse" ~ "Grande_Anse", TRUE ~ dep)
-  )
-
-true_agg_cases <- dep_plot_df %>%
-  tidyr::pivot_wider(
-    data = .,
-    id_cols = c(date),
-    names_from = dep,
-    values_from = cases,
-    names_prefix = 'cases_'
-  ) %>%
-  mutate(
-    ReportedAll = cases_Artibonite + cases_Centre +
-      cases_Grande_Anse + cases_Nippes + cases_Nord +
-      cases_Nord_Est + cases_Ouest + cases_Sud +
-      cases_Sud_Est + cases_Nord_Ouest
-  ) %>%
-  select(date, ReportedAll)
-
-h2_epi_traj$date <- as.Date(
-  lubridate::round_date(
-    lubridate::date_decimal(h2_epi_traj$year),
-    unit = 'day'
-  )
-)
-
-h2_end_traj$date <- as.Date(
-  lubridate::round_date(
-    lubridate::date_decimal(h2_end_traj$year),
-    unit = 'day'
-  )
-)
-
-all_traj <- bind_rows(
-  select(h2_epi_traj, date, Ctotal),
-  select(h2_end_traj, date, Ctotal)
-)
-
-projections <- left_join(
-  x = true_agg_cases,
-  y = all_traj,
-  by = 'date'
-)
-@
-
-\begin{figure}[ht]
-<<plotModel2Rep, cache=TRUE, echo=FALSE, fig.width=6.5, fig.height=4.5>>=
-gg_epi <- projections %>%
-  filter(date < as.Date("03-01-2014", format = '%m-%d-%Y')) %>%
-  ggplot(aes(x = date)) +
-  geom_line(aes(y = ReportedAll), col = 'black', linetype = 'dashed') +
-  geom_line(aes(y = Ctotal), col = 'red', size = 1) +
-  theme_bw() +
-  scale_y_continuous(limits = c(0, 30000), breaks = seq(0, 30000, 5000)) +
-  theme(axis.title.x = element_blank()) +
-  ylab("Reported number of cases")
-
-gg_end <- projections %>%
-  filter(date >= as.Date("03-01-2014", format = '%m-%d-%Y')) %>%
-  ggplot(aes(x = date)) +
-  geom_line(aes(y = ReportedAll), col = 'black', linetype = 'dashed') +
-  geom_line(aes(y = Ctotal), col = 'red', size = 1) +
-  theme_bw() +
-  scale_y_continuous(limits = c(0, 7000), breaks = seq(0, 7000, 1000)) +
-  theme(axis.title.x = element_blank()) +
-  ylab("Reported number of cases") +
-  scale_x_date(date_breaks = "1 year", date_labels = "%Y")
-
-cowplot::plot_grid(gg_epi, gg_end, ncol = 1)
-@
-\caption{\label{fig:mod2rep}
-Model 2 trajectories using the \code{haitipkg}. Compare to Figure S15 of \cite{lee20sup}.
-}
-\end{figure}
-
-There are minor differences between Figure~\ref{fig:mod2rep} and Figure~S15 of \cite{lee20sup}.
-% such as a delayed second wave in the epidemic period and a larger initial wave during the endemic period.
-While the discrepancy appears minor, the deterministic nature of Model~2 implies that an exact replication of model trajectories should be possible.
-In this case, these discrepancies may be attributed to model implementation and small differences in model parameters.
-Details about the measurement models and how latent states were initialized for the epidemic model were not provided by \citet{lee20sup} and therefore these details must be inferred by looking at the provided source code.
-According to repository comments, the files \code{fitInPieces3paramsCleanMay2019Public.py} and \code{fitInPiecesMuWithFracSusFixedAllInfectionsPublic.py} were used to fit the epidemic and endemic phases of the model respectively, although it is apparent that these exact files were not used to obtain the reported results.
-These files contain errors that making them impossible to run, and therefore these files cannot be the code that generated the reported model fits\footnote{One example of why the code cannot be run that the file loads functions from a non-extant file named \code{choleraEqs.py} in line 13 rather than \code{choleraEqsPublic.py}.}.
-The inability to replicate the results by \citet{lee20} by running the provided source code makes checking whether or not a our numeric implementation faithfully represents their results very difficult.
-% The difficulty of reproducing the result of \citet{lee20} is coupled with a misunderstanding of how these results were originally obtained.
-
-\jwc{PROBABLY DELETE THIS NEXT BIT, BUT IT'S USEFUL TO DOCUMENT IT FOR NOW...}
-One concern that arises when inspecting the files \code{fitInPieces3paramsCleanMay2019Public.py} and \code{fitInPiecesMuWithFracSusFixedAllInfectionsPublic.py} is that the output of these files (which contains the estimated parameters) was not referenced/used in either the file that generates the sensitivity analysis or the file that generates the model simulations; instead, alternative (non-extant) files with similar names are loaded, suggesting parameters that were used to generate the final results were not a direct result of the estimated parameters.
-Furthermore, inspection of the script \code{simulateHaitiFullTimeInPiecesAllVaccinationScenariosSingleParameterVectorAllInfectionsPublic.py}---which is used to obtain simulations for each of the vaccination scenarios---results in a number of concerns.
-First, the script uses parameters from the pickle file \code{leastSquaresDataUpTo2019/fitEpidemicPiece\_resultsMatrix\_18Mar2019.pickle}, which only appears in the repository as a commented out line of code in the file \code{generateSetsOfParametersForSensitivityAnalysisPublic.py}, suggesting that the set of parameters used to simulate each vaccination scenario was not the most recent set of parameters generated by the sensitivity analysis.
-Second, the script appears to use the set of differential equations from the function \code{choleraEqs11WithVaccinationNetwork} for the epidemic phase, even though parameter values for this phase were estimated using \code{choleraEqs10WithVaccinationNetwork}.
-It is possible, however, that we simply are interpreting the code incorrectly and that these concerns are unwarranted.
-Furthermore, because the provided code is non-functional, it is possible that the code that was actually used to obtain the reported results does not include these potential errors.
-This highlights the importance of providing the exact source code that was used to obtain the reported results of any analysis.
-
-Another potential explanation for the discrepancy between our simulations from Model~2 and those of \citet{lee20} is that the parameters that we used are only approximately the same as those used by \citet{lee20sup}.
-For example, the parameters $\transmission$, $\Wbeta{}$ (See Table~\ref{tab:translate}) had reported values of $9.9 \times 10^{-7}$ and $4.03 \times 10^{-2}$, respectively (Table~S13 of \cite{lee20sup}), but were actually fit to data and therefore likely had values that were far more precise.
-Additionally, our implementation of Model~2 used a time scale of years and many of the parameters were reported on a weekly scale, so small differences may result due to unit conversions.
-The collective effect of these small differences in model parameters likely will result in differences in model trajectories.
-This issue could have been mitigated if the saved files containing estimated model parameters were included as part of the code submission.
-
-\subsection{Model~3 Replication}
-
-Model~3 was developed by a team of researchers at the Laboratory of the Swiss Federal Institute of Technology in Lausanne, hereafter referred to as the Model~3 authors.
-The code that was originally used to implement Model~3 is archived with the DOI: \url{10.5281/zenodo.3360723}, and also available in the public GitHub repository: \code{jcblemai/haiti-mass-ocv-campaign}.
-Because the code was made publicly available, and final model parameters were reported in the supplementary material of \cite{lee20}, we were able to reproduce Model~3 by directly using the source code.
-In Fig.~\ref{fig:mod3rep}, we plot simulations from this model.
-This figure should be compared to Figure S18 of \cite{lee20}.
-We note that slight differences may be accounted for due to variance in the model simulations and the difference in programming language used to produce the figure.
-
-<<Model3Replication, echo=FALSE, cache=TRUE>>=
-NSIM <- switch(run_level, 50, 100, 1000)
-
-plot_order <- c(
-  'Artibonite',
-  'Sud_Est',
-  'Nippes',
-  'Nord_Est',
-  'Ouest',
-  'Centre',
-  'Nord',
-  'Sud',
-  'Nord_Ouest',
-  'Grande_Anse'
-)
-
-dep_labeller <- as_labeller(
-  c(
-    'Artibonite' = 'Artibonite',
-    'Sud_Est' = 'Sud-Est',
-    'Sud.Est' = 'Sud-Est',
-    'Nippes' = 'Nippes',
-    'Nord_Est' = 'Nord-Est',
-    'Nord.Est' = 'Nord-Est',
-    'Ouest' = 'Ouest',
-    'Centre' = 'Centre',
-    'Nord' = 'Nord',
-    'Sud' = 'Sud',
-    'Nord_Ouest' = 'Nord-Ouest',
-    'Nord.Ouest' = 'Nord-Ouest',
-    'Grande_Anse' = 'Grand\'Anse',
-    'Grand.Anse' = 'Grand\'Anse'
-  )
-)
-
-lee3 <- haiti3_lee_et_al()
-lee3_sims <- simulate(lee3, nsim = NSIM, seed = 321, format = 'data.frame')
-
-sub_cases <- haitiCholera %>%
-  dplyr::mutate(date = as.Date(date_saturday)) %>%
-  dplyr::select(-report, -date_saturday) %>%
-  dplyr::filter(lubridate::decimal_date(date) >= min(lee3_sims$time)) %>%
-  tidyr::pivot_longer(
-    data = .,
-    cols = -date,
-    names_to = "dep",
-    values_to = "cases"
-  ) %>%
-  dplyr::mutate(
-    dep = dplyr::case_when(
-      dep == "Grand.Anse" ~ "Grande_Anse",
-      dep == "Nord.Est" ~ "Nord_Est",
-      dep == "Nord.Ouest" ~ "Nord_Ouest",
-      dep == "Sud.Est" ~ "Sud_Est",
-      TRUE ~ dep
-    )
-  )
-
-quants <- lee3_sims %>%
-  dplyr::select(-CasesAll) %>%
-  tidyr::pivot_longer(
-    data = .,
-    cols = starts_with("cases"),
-    names_prefix = "cases",
-    names_to = "dep",
-    values_to = "cases"
-  ) %>%
-  dplyr::group_by(dep, time) %>%
-  dplyr::summarise(
-    q05 = quantile(cases, 0.025, na.rm = T),
-    mean = mean(cases, na.rm = T),
-    q50 = quantile(cases, 0.5, na.rm = T),
-    q95 = quantile(cases, 0.975, na.rm = T)
-  ) %>%
-  dplyr::ungroup() %>%
-  dplyr::mutate(
-    date = lubridate::date_decimal(time),
-    date = as.Date(lubridate::round_date(date))
-  )
-
-quants <- quants %>%
-  dplyr::filter(
-    !(dep == "Ouest" & date < '2017-06-10')
-  )
-@
-
-\begin{figure}[ht]
-<<PlotMod3Rep, echo=FALSE, message=FALSE, fig.height=3.75>>=
-ggplot() +
-  geom_ribbon(data = quants, aes(x = date, ymin = q05, ymax = q95),
-              fill = 'darkblue', alpha = 0.5) +
-  geom_line(data = quants, aes(x = date, y = q50), col = 'darkblue') +
-  geom_point(data = dplyr::filter(sub_cases, cases < 500), aes(x = date, y = cases), size = 0.2) +
-  facet_wrap(~factor(dep, levels = plot_order), nrow = 2, labeller = dep_labeller) +
-  labs(y = 'Reported Cholera Cases') +
-  scale_x_date(date_breaks = "6 months", date_labels = "%Y-%m") +
-  theme(
-    axis.text.x = element_text(angle = 40, hjust = 1, vjust = 1, size = 7),
-    axis.title.x = element_blank()
-  )
-@
-\caption{\label{fig:mod3rep}
-Simulations from Model~3.
-Compare to Figure S18 of \cite{lee20}.
-}
-\end{figure}
-
-<<lee3Cleanup, echo=FALSE, include=FALSE, message=FALSE>>=
-rm(lee3, lee3_sims, quants, sub_cases)
-gc()
-@
-
 \clearpage

### 2022-10-12 ms.Rnw
minor edits and some tidying

@@
 %%   The Annals of Applied Statistics [AOAS]
-%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% In this template, the places where you   %%
-%% need to fill in your information are     %%
-%% indicated by '???'.                      %%
-%%                                          %%
-%% Please do not use \input{...} to include %%
-%% other tex files. Submit your LaTeX       %%
-%% manuscript as one .tex document.         %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\documentclass[aoas]{imsart}
+
+%\documentclass[aoas]{imsart}
+\documentclass[aoas,preprint]{imsart}
 
@@
 \startlocaldefs
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                                          %%
-%% Uncomment next line to change            %%
-%% the type of equation numbering           %%
-%%                                          %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%\numberwithin{equation}{section}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                                          %%
-%% For Axiom, Claim, Corollary, Hypothesis, %%
-%% Lemma, Theorem, Proposition              %%
-%% use \theoremstyle{plain}                 %%
-%%                                          %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%\theoremstyle{plain}
-%\newtheorem{???}{???}
-%\newtheorem*{???}{???}
-%\newtheorem{???}{???}[???]
-%\newtheorem{???}[???]{???}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                                          %%
-%% For Assumption, Definition, Example,     %%
-%% Notation, Property, Remark, Fact         %%
-%% use \theoremstyle{remark}                %%
-%%                                          %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%\theoremstyle{remark}
-%\newtheorem{???}{???}
-%\newtheorem*{???}{???}
-%\newtheorem{???}{???}[???]
-%\newtheorem{???}[???]{???}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% Please put your definitions here:        %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
   stringsAsFactors = FALSE,
-  # prompt="R> ",
   continue = "+  ",
@@
 \begin{abstract}
+\begin{center}
+Working draft, \today
+\end{center}
+
+\vspace{3mm}
+
 Public health decisions must be made about when and how to implement interventions to control an infectious disease epidemic.
@@
 \end{frontmatter}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% Please use \tableofcontents for articles %%
-%% with 50 pages and more                   %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%\tableofcontents
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%% Main text entry area:
 
@@
 In the study, certain data were shared between the groups, including demography and vaccination history; vaccine efficacy was also fixed at a shared value between groups.
-Beyond this, the groups made autonomous decisions on what to include and exclude from their models; this autonomy reduced the possible effect that assumptions about the dynamic system may have on the final conclusion of the study.
-Despite this autonomy, and largely adhering to existing guidelines on creating models to inform policy \citep{behrend20,saltelli20}, the consensus across the four models was that an extensive nationwide vaccination campaign would be necessary to eliminate cholera from Haiti.
-This conclusion is inconsistent with the fact that there have been no confirmed cases since February, 2019 \citep{trevisin22}, despite the lack of a national vaccination program.
+Beyond this, the groups made autonomous decisions on what to include and exclude from their models.
+The groups largely adhered to existing guidelines on creating models to inform policy \citep{behrend20,saltelli20} and, despite their autonomy, obtained a consensus that an extensive nationwide vaccination campaign would be necessary to eliminate cholera from Haiti.
+This conclusion is inconsistent with the fact that there have been no confirmed cases since February, 2019 \citep{trevisin22} without the implementation of a national vaccination program.
 
 The discrepancy between the model based conclusions of \cite{lee20} and the subsequent elimination of cholera cholera has been debated \citep{francois20,rebaudetComment20,henrys20,leeReply20}.
-Suggested origins of this discrepancy include the use of unrealistic models \citep{rebaudetComment20} and having unrealistic criteria for cholera elimination \citep{henrys20}.
+Suggested origins of this discrepancy include the use of unrealistic models \citep{rebaudetComment20} and unrealistic criteria for cholera elimination \citep{henrys20}.
 We find a more nuanced conclusion: attention to methodological details in model fitting, diagnosis and forecasting can improve each of the proposed model's ability to quantitatively describe observed data.
@@
 
-% \eic{PERHAPS THIS PARAGRAPH COULD MOVE TO A LATER SECTION?}
-% The complex nature of biological systems necessitates a search for modeling assumptions that combine insightful simplicity with fidelity to biological reality.
-% For example, many models commonly used in epidemiology are motivated by reasoning about a homogeneous mixing population which is simultaneously an avenue for powerful simplification and a source of model misspecification \citep{bansal07}.
-% Other common considerations include whether the proposed model should be stochastic or deterministic; whether the model should have change points in parameter values or should otherwise make adjustments for changes through time in the dynamic system; and whether the proposed model should include any spatial heterogeneity at a scale permissible by the observed data.
-% While this section is focused on the elements of $\paramVec$ that are calibrated to data, we note that the decision of which elements to fix and which to estimate has consequences for model interpretability, as discussed in Sec.~\ref{sec:science}.
-Each of the three models considered in this study describes cholera dynamics as partially observed Markov processes (POMP), with the understanding that the deterministic Model~2 is a degenerate case of a stochastic model.
-% Despite their similarities, these models represent a diverse selection of possible modeling assumptions: namely, the use of stochastic (Models~1 and 2) or deterministic (Model~2) equations; spatially-heterogeneous meta-population (Models~2 and 3) or spatially-aggregated (Model~1) structure; and the use of covariates (Model~3) versus mathematical equations (Models~1 and 2) to describe a seasonal mechanism.
-As such, each proposed mechanistic structure forms a family of statistical models indexed by a parameter vector $\paramVec$.
-Different values of $\paramVec$ can result in qualitative differences in the predicted behavior of the system.
+Each of the three models considered in this study describes cholera dynamics as a partially observed Markov process (POMP), with the understanding that the deterministic Model~2 is a degenerate case of a stochastic model.
+Each model is indexed by a parameter vector, $\paramVec$, and different values of $\paramVec$ can result in qualitative differences in the predicted behavior of the system.
 Therefore, the choice of $\paramVec$ used to make inference about the system can greatly affect model based conclusions.
-Specific elements of $\paramVec$ can be chosen as constants based on scientific understanding of the system, but parameters can also be calibrated to data in a way that maximizes some measure of congruency between the observed data and the assumed mechanistic structure.
-Mechanistic models are used as a way of approximating real-world mechanisms present in a dynamic system.
-While calibrating model parameters does not directly achieve this goal---as model assumptions may be incorrect, and do not change as the model is calibrated to data---the congruency between the model and observed data works as a proxy for the congruency between the model and the true underlying dynamic system, since the observed data are a result of the system.
-% All these structures can be described in the framework of partially observed Markov process (POMP) models, with the understanding that the deterministic Model~2 is a degenerate case of a stochastic model.
+Elements of $\paramVec$ can be fixed at a constant value based on scientific understanding of the system, but parameters can also be calibrated to data by maximizing a measure of congruency between the observed data and the assumed mechanistic structure.
+Calibrating model parameters does not guarantee that the resulting model successfully approximates real-world mechanisms, since model assumptions may be incorrect, and do not change as the model is calibrated to data
+However, the congruency between the model and observed data serves as a proxy for the congruency between the model and the true underlying dynamic system, since the observed data are a result of the system.
 In the following subsections we describe our approach to fitting the three proposed mechanistic models by calibrating $\paramVec$ to observed cholera incidence data.
@@
   geom_line(data = mcap_results$fit, aes(x = parameter, y = smoothed), col = 'blue') +
-  # geom_line(data = mcap_results$fit, aes(x = parameter, y = quadratic), col = 'red') +
   geom_vline(xintercept = mcap_results$ci[1], linetype = 'dashed') +
   geom_vline(xintercept = mcap_results$ci[2], linetype = 'dashed') +
-  # geom_vline(xintercept = mcap_results$quadratic_max, col = 'red') +
   geom_vline(xintercept = mcap_results$mle, col = 'blue') +
@@
 
-% Fitting Model~3 in this fashion simplifies the problem of parameter estimation, but it also introduces additional technicalities that must be addressed.
-% One concern is that of obtaining model forecasts, which was the primary goal of \citet{lee20}.
 The simplified, spatially-decoupled version of Model~3 relies on the observed cholera cases for each department to describe cholera dynamics, making it impossible to obtain forecasts for future time points from the fitted model.
@@
 Simulations from the fitted model are displayed in Fig.~\ref{fig:h3spatsims}.
-
-% An approximation of Eq.~\myeqref{eq:model3:foi} for each department $v \in \seq{1}{10}$ can be obtained using the weekly number of observed cholera cases in each department:
-% \begin{equation}
-%   I_v(t) + A_v(t) \approx \frac{365}{7\reportRate}y^*_v(t)\left(\frac{1}{\muDeath + \choleraDeath + \muIR} + \frac{1 - \symptomFrac}{\symptomFrac(\muDeath + \muIR)}\right)\label{eq:model3:approx}
-% \end{equation}
-% This approximation leads to department-specific models that are conditionally independent given the reported number of cholera infections in the remaining departments.
-% Here, we refer to a collection of POMP models that are independent across units as a PanelPOMP.
-%
-% In the case of a PanelPOMP, an extension of the IF2 algorithm, known as Panel Iterated Filtering (PIF) \citep{breto20}, can be used to obtain the MLE, which solves the curse of dimensionality for this class of models \eic{WHICH CONVENTIONALLY REFERS TO THE PRECEDING OBJECT, HERE THE MLE, WHICH IS NOT WHAT WE WANT}.
-% A major advantage of this algorithm is that PIF can be used to fit both unit-specific parameters and shared parameters; in this way, the calibration of shared parameters involves all of the available data instead of just an arbitrarily chosen subset.
-%
-% \eic{READERS MAY BE CONFUSED HERE. IS THIS A NEW METHOD? IT SEEMS TO BE IMPLEMENTED BY BRETO ET AL IN 2020. THIS IS DELICATE.}
-% We propose and use a slight modification of the PIF algorithm, which we call the Block Panel Iterated Filter (BPIF), to estimate parameters for a PanelPOMP version of Model~3.
-% Let $l_{\text{P}}(y_{1:N}^*; \paramVec)$ denote the log-likelihood function of the PanelPOMP version of Model~3.
-% The output of the BPIF algorithm is the MLE $\tilde{\paramVec} = \argmax_{\paramVec} l_{\text{P}}(y_{1:N}^*; \paramVec)$.
-% The BPIF algorithm is implemented in the \code{panelPomp} package \citep{breto20panelPomp} as the argument \code{block = TRUE} in the \code{mif2} function \eic{THIS LEVEL OF CODING DETAIL IS UNUSUAL IN A CASE STUDY}.
-% Pseudo-code for this algorithm is provided in Algorithm~S1 in the supplementary material.
-
-% While the reported parameters were obtained using the decoupled representation of Model~3, simulations for the various vaccination campaigns were obtained using fully coupled version of the model, implemented using the \code{spatPomp} package.
-
-% Fitting Model~3 in this fashion simplifies the problem of parameter estimation, but it also introduces additional technicalities that must be addressed.
-% One concern is that of obtaining model forecasts, which was the primary goal of \citet{lee20}.
-% The simplified PanelPOMP version of Model~3 relies on the observed cholera cases as a covariate, which are unavailable for use in forecasts.
-% Similar to what was done in \cite{lee20}, we use the parameter estimates obtained with the independent departmental approximation of Model~3 as an estimate for the parameters in the fully coupled version of the model, which was implemented using the \code{spatPomp} package.
-% Simulations from the fully coupled version of the model using the estimated parameter vector $\tilde{\paramVec}$, are displayed in Fig.~\ref{h3spatsims}.
 
@@
 @
-
-% While simulations from the SpatPOMP version of Model~3 resemble the observed data, we note in Table~\ref{tab:likes} that the SpatPOMP version of Model~3 has a lower log-likelihood than the PanelPOMP version using the same model parameters.
-% Let $l_{\text{C}}(y_{1:N}^*; \paramVec)$ denote the log-likelihood of the fully coupled model, and let $\hat{\paramVec} = \argmax_{\paramVec} l_{\text{C}}(y_{1:N}^*; \paramVec)$.
-% If the goal of parameter estimation is to maximize the likelihood of the fully coupled model, the approach of estimating parameters using a decoupled model is clearly sub optimal, as $l_{\text{C}}(y_{1:N}^*; \hat{\paramVec}) \geq l_{\text{C}}(y_{1:N}^*; \tilde{\paramVec})$.
-% A natural question that arises is if the difference between $\hat{\paramVec}$ and $\tilde{\paramVec}$ is scientifically and statistically meaningful.
-% The answer to this question requires the ability to reliably estimate $\hat{\paramVec}$, the answer to which is model specific and beyond the scope of this case study.
-
-% A brief analysis of the fitted parameters $\tilde{\paramVec}$ provides some interesting insight.
-% Consider the large estimated exponent on rainfall: $r = \Sexpr{signif(p3s["r"],3)}$ (Eq.~\ref{eq:model3:water}).
-% Because the value of rainfall $J(t) \in (0, 1)$, the large coefficient results in $J(t)^r \approx 0$.
-% This result may initially be surprising given the importance of rainfall as a seasonal diver of cholera infections in Haiti (see Sec.~\ref{sec:science}).
-% One plausible explanation of this result is that the fitting procedure---used to obtain parameter values in the independent department model given the reported cases in other departments---has learned an over-dependence on the reported cholera cases in other departments, and thus the seasonality of cholera incidence in department $u$ depends on the highly correlated covariate of the reported cholera cases in departments $1:U\setminus \{u\}$.
-% Whatever the case may be, this fitting procedure results in a model that is unable to replicate the observed number of cases by simulating from initial conditions (see Fig.~\ref{h3spatsims}), and highlights the need for advancements in statistical methodology that permit inference on models with coupled metapopulation dynamics.
 
@@
 
-% It has been noted that the uncertainty in just a single parameter can lead to drastically different forecasts \citep{saltelli20};
-% parameter uncertainty should therefore be considered when obtaining model forecasts to influence policy.
-% If a Bayesian technique is used for parameter estimation, the most natural way to account for parameter uncertainty is to obtain $J$ simulations from the model where each simulation is obtained using parameters $\Theta_{1:J}$ drawn from the posterior distribution $\Theta_{1:J} \sim f_{\Theta}\big(\paramVec | \bm{Y}^{(\modelCounter)}_{1:N} = y_{1:N}^*\big)$.
-% In the frequentist case, one approach is empirical Bayes: endow to set $\Theta_K = \{\theta_1, \theta_2, \ldots, \theta_K\}$ with a discrete uniform distribution, such that $P\big(\Theta_{j} = \theta_k\big) = \frac{1}{K}$ for all values $k \in \seq{1}{K}$.
-% In this case, the posterior distribution of $\Theta_{j} | \bm{Y}^{(\modelCounter)}_{1:N}$ can be expressed as:
-% $P\big(\Theta_{j} = \theta_k | \bm{Y}^{(\modelCounter)}_{1:N} = y_{1:N}^*\big) = \frac{l(y_{1:N}^*; \theta_k)}{\sum_{l = 1}^K l(y_{1:N}^*; \theta_l)}$.
-To complete the formulation, we then choose the set $\Theta_K$ such that each $\theta_k \in \Theta_K$ is a single output of the IF2 algorithm, which has been shown to be within a neighborhood of the MLE \cite{ionides15}.
-% This approach has been taken in previous studies \citep{king15}, and the result is simply weighing parameter sets based on their corresponding likelihoods: parameters with higher likelihoods should be used more frequently in model forecasts than those that correspond to lower likelihoods.
-% This can be done for both deterministic and stochastic models, but requires a large number of computations, especially as the number of observations and model parameters increase.
-% Because our direct aim is not to provide forecasts for policy recommendation, we do not include parameter uncertainty in our forecasts.
-% Instead, we use the projections from point estimates to highlight a major deficiency of deterministic models, which is that the only variability in model projections is a result of parameter uncertainty, which leads to over-confidence in forecasts \cite{king15}.
-% To the credit of deterministic models, of the four fitted models in \cite{lee20}, Model~2 provides the most apparently accurate forecasts.
-% This perhaps demonstrates that while deterministic models describe the systems in a less realistic way, the relative ease in fitting these models potentially results in fewer modeling errors.
-% As the data analysis becomes more refined, however, the deficiencies of deterministic models become increasingly apparent.
-
 It has been noted that the uncertainty in just a single parameter can lead to drastically different forecasts \citep{saltelli20};
@@
 If a Bayesian technique is used for parameter estimation, the most natural way to account for parameter uncertainty is to obtain simulations from the model where each simulation is obtained using parameters drawn from the posterior distribution.
-In the frequentist case, one approach is empirical Bayes, as was done by \cite{king15} (see supplement for details). % TODO: \jwc{ADD THIS TO SUPPLEMENT!}
+In the frequentist case, one approach is empirical Bayes, as was done by \cite{king15} (see supplement for details). 
 Both of these approaches share the similarity that parameters with higher likelihoods should be used more frequently in model forecasts than those that correspond to lower likelihoods.
@@
 Instead, we use the projections from point estimates to highlight a major deficiency of deterministic models, which is that the only variability in model projections is a result of parameter uncertainty, which leads to over-confidence in forecasts \cite{king15}.
-% To the credit of deterministic models, of the four fitted models in \cite{lee20}, Model~2 provides the most apparently accurate forecasts.
-% This perhaps demonstrates that while deterministic models describe the systems in a less realistic way, the relative ease in fitting these models potentially results in fewer modeling errors.
-% As the data analysis becomes more refined, however, the deficiencies of deterministic models become increasingly apparent. \jwc{CHANGE THIS BASED ON DISCUSSION WITH ED.}
 
@@
 
-%% The potential for new discoveries based on fitting a dynamic model, however, is rooted in a model's ability to effectively approximate the system in question, and is therefore inexorably linked to a model's goodness-of-fit.
-%% As such, conclusions based on dynamic models that poorly fit observed data should not be considered as much as those resulting from models that describe observed data well.
-THIS DUPLICATES LATER SENTENCES, SO I THINK IT CAN BE REMOVED HERE.
-
 In the context of our case study, we demonstrate how the fit of Model~1 corroborates other evidence concerning the role of rainfall in cholera epidemics.
@@
 To interpret this result, we bear in mind that the data ranged from 2010-2019, and therefore estimates of immunity longer than 10 years---the upper end of previous estimates of natural immunity \citep{king08}---effectively result in the same model dynamics.
-The depletion of susceptible individuals may also be attributed to confounding mechanisms that are unaccounted for in the model---such as vaccination and other non-pharmaceutical interventions that reduce cholera transmission \citep{trevisin22, rebaudet21}---that were not accounted for in the model.
+The depletion of susceptible individuals may also be attributed to confounding mechanisms---such as localized vaccination programs and non-pharmaceutical interventions that reduce cholera transmission \citep{trevisin22, rebaudet21}---that were not accounted for in the model.
 Perhaps the best interpretation of the estimated parameter, then, is that under model assumptions, the model most adequately describes the observed data by having a steady decrease in the number of susceptible individuals.
@@
 
-The ongoing global COVID-19 pandemic has provided a clear example on how government policy may be affected by the conclusions of scientific models.
+The ongoing global COVID-19 pandemic has provided a clear example on how government policy may be affected by the conclusions of scientific models \citep{saltelli20}.
 This article demonstrates that fitting appropriate scientific models remains a challenging statistical task, and therefore great care is needed when fitting scientific models for policy recommendations.
@@
 At a minimum, we have shown that the conclusions are sensitive to details in how the data analysis is carried out, and that attention to statistical fit (including numerical issues such as likelihood maximization) can lead to improved policy guidance.
-
-% We acknowledge there are limitations to this study, such as the inability to fit model parameters to the fully-coupled version of Model~3.
-% Promising theoretical and methodological developments \citep{ning21ibpf,ionides22} based on the Block Particle Filter \citep{rebeschini15} may potentially be used to obtain better parameter estimates for Model~3 in future work.
 
@@
 
-We provide a strong form of reproducibility, as well as extendability, by developing our analysis in the context of an \code{R} package, \code{haitipkg}.
+We provide a strong form of reproducibility, as well as extendability, by developing our analysis in the context of a software package, \code{haitipkg}, written in the R language \citep{r}.
 Using a software package mechanism supports documentation, standardization and portability that promote extendability.
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% Support information, if any,             %%
-%% should be provided in the                %%
-%% Acknowledgements section.                %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \begin{acks}[Acknowledgments]
@@
 \end{acks}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% Funding information, if any,             %%
-%% should be provided in the                %%
-%% funding section.                         %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \begin{funding}
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% Supplementary Material, including data   %%
-%% sets and code, should be provided in     %%
-%% {supplement} environment with title      %%
-%% and short description. It cannot be      %%
-%% available exclusively as external link.  %%
-%% All Supplementary Material must be       %%
-%% available to the reader on Project       %%
-%% Euclid with the published article.       %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \begin{supplement}
@@
 \begin{supplement}
-\stitle{\code{haitipkg}}
+\stitle{Auxiliary software: \code{haitipkg}}
 \sdescription{
-This \code{R} package is contained in a public GitHub repository: jeswheel/haitipkg.
+The \code{haitipkg} R package is maintained in a GitHub repository, \code{jeswheel/haitipkg}. The submitted version of this package will be archived on Zenodo.
 The package contains all of the data and code used to create and fit the models, as well as other useful functions that were used in this article.
@@
 \begin{supplement}
-\stitle{\code{jeswheel/haiti\_article}}
+\stitle{Dynamic document: \code{haiti\_article}}
 \sdescription{
-This GitHub repository contains the \code{.Rnw} files that were used to create this document and the supplement material.
+The Rnw \citep[R noweb,][]{ramsey94} files generating this article and its supplement are maintained in a GitHub repository, \code{jeswheel/haiti\_article}.
+The submitted version will be archived on Zenodo.
 }
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                  The Bibliography                       %%
-%%                                                         %%
-%%  imsart-nameyear.bst  will be used to                   %%
-%%  create a .BBL file for submission.                     %%
-%%                                                         %%
-%%  Note that the displayed Bibliography will not          %%
-%%  necessarily be rendered by Latex exactly as specified  %%
-%%  in the online Instructions for Authors.                %%
-%%                                                         %%
-%%  MR numbers will be added by VTeX.                      %%
-%%                                                         %%
-%%  Use \cite{...} to cite references in text.             %%
-%%                                                         %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%% if your bibliography is in bibtex format, uncomment commands:
-\bibliographystyle{imsart-nameyear} % Style BST file
-\bibliography{bib-haiti}       % Bibliography file (usually '*.bib')
+\bibliographystyle{imsart-nameyear} 
+\bibliography{bib-haiti}
 

### 2022-10-12 ms.Rnw
Compiled Ed's changes, pushing results

@@
 Elements of $\paramVec$ can be fixed at a constant value based on scientific understanding of the system, but parameters can also be calibrated to data by maximizing a measure of congruency between the observed data and the assumed mechanistic structure.
-Calibrating model parameters does not guarantee that the resulting model successfully approximates real-world mechanisms, since model assumptions may be incorrect, and do not change as the model is calibrated to data
+Calibrating model parameters does not guarantee that the resulting model successfully approximates real-world mechanisms, since model assumptions may be incorrect, and do not change as the model is calibrated to data.
 However, the congruency between the model and observed data serves as a proxy for the congruency between the model and the true underlying dynamic system, since the observed data are a result of the system.

### 2022-10-12 ms.Rnw
Worked on references

@@
 As previously stated, Model~2 is a deterministic compartmental model defined by a set of coupled differential equations.
-The use of deterministic compartment models have a long history in the field of infectious disease epidemiology \citep{kermack1927,brauer2017,varghese21},
+The use of deterministic compartment models have a long history in the field of infectious disease epidemiology \citep{kermack1927,brauer2017,giordano20},
 and can be justified by asymptotic considerations in a large-population limit \citep{dadlani2020,ndii17}.

### 2022-10-12 si/si.Rnw
Worked on references

@@
 \end{equation}
-where $Y_n$ is the reported cholera cases at time $n \in 1:N$ and $\Delta N_{EI}(n)$ is the sum total of individuals across vaccination compartment $\vaccCounter$ who moved from compartment $E_\vaccCounter$ to $I_\vaccCounter$ since observation $n-1 \in 0:N-1$, and $\text{NB}\left(\reportRate z, \obsOverdispersion \right)$ is the negative binomial distribution with mean $\reportRate z$ and variance $\reportRate z+ \psi^2\reportRate^2 z^2)$.
-\jwc{CHECK THIS}.
+where $Y_n$ is the reported cholera cases at time $n \in 1:N$ and $\Delta N_{EI}(n)$ is the sum total of individuals across vaccination compartment $\vaccCounter$ who moved from compartment $E_\vaccCounter$ to $I_\vaccCounter$ since observation $n-1 \in 0:N-1$, and $\text{NB}\left(\reportRate z, \obsOverdispersion \right)$ is the negative binomial distribution with mean $\reportRate z$ and variance $\reportRate z \big(1 + \frac{\reportRate z}{\obsOverdispersion}\big)$.
 

### 2022-10-13 ms.Rnw
edits on forecasting etc

@@
 \small
-Reported Cholera cases in the outbreak in Haiti from 2010-2019.
+Weekly reported cholera cases in Haiti from October 2010 to January 2019.
 }
@@
 Plug-and-play algorithms include Bayesian approaches like ABC and PMCMC \citep{toni09,andrieu10}, but here we use frequentist methods to maximize model likelihoods.
-To our knowledge, the only plug-and-play frequentist methods that can maximize the complete model likelihood are iterated filtering algorithms, which modify the well-known particle filter \citep{arulampalam02} by performing a random walk for each parameter and particle.
+To our knowledge, the only plug-and-play frequentist methods that can maximize the likelihood for POMP models of this complexity are iterated filtering algorithms, which modify the well-known particle filter \citep{arulampalam02} by performing a random walk for each parameter and particle.
 These perturbations are carried out iteratively over multiple filtering operations, using the collection of parameters from the previous filtering pass as the parameter initialization for the next iteration, and decreasing the random walk variance at each step.
@@
 
-To combat this issue, we fit the model to log-transformed case counts, where the difference between periods of high and low infectiousness is less drastic, since the log scale stabilizes the variation during periods of high and low infectiousness.
+To combat this issue, we fit the model to log-transformed case counts, since the log scale stabilizes the variation during periods of high and low incidence.
 Other solutions include changing the measurement model to include overdispersion, as was done in Models~1 and~3.
@@
 Evidence for the former approach has been provided in previous studies \citep{king15}, including the specific case of heterogeneity between Haitian departments in cholera transmission \citep{collins14}.
-A legitimate preference for simplicity \citep{saltelli20,green15} can support a decision to consider nationally aggregated models.
+However, a legitimate preference for simplicity \citep{saltelli20,green15} can support a decision to consider nationally aggregated models.
 
@@
 The exception to this being the parameters $\beta_{W_u}$, and $\beta_u$, which take department-specific values for each department $u$.
-To avoid the parameter estimation issue in high-dimensional models, \citet{lee20} simplified the problem by using the IF2 algorithm to separately estimate parameters for independent department-level models;
+To avoid the parameter estimation issue in high-dimensional models, \citet{lee20} simplified the problem by using the IF2 algorithm to estimate parameters separately for independent department-level models;
 the shared parameters were calibrated using the cholera incidence data from Artibonite, and the department-specific parameters ($\beta_{W_u}$ and $\beta_u$) were fit using the data from their respective department.
@@
 In order to obtain model forecasts, the parameters fit for the conditionally independent department level models were used in the fully coupled version of Model~3.
-This approach of model calibration and forecasting avoids the issue of particle depletion by simplifying the parameter estimation problem, but it may result in some undesirable outcomes.
+This approach of model calibration and forecasting avoids the issue of particle depletion by simplifying the parameter estimation problem, but this advantage raises new issues.
 One concern is that cholera dynamics in department $u$ are highly related to the dynamics in the remaining departments;
@@
 This is because the likelihood of the coupled model at its maximum likelihood estimate (MLE) is greater than or equal to the likelihood of the coupled model at the MLE of the decoupled model.
-These two concerns may explain the unrealistic forecasts and low-likelihood of the model (see Table~\ref{tab:likes}).
-
-For our analysis, we calibrate the parameters of the spatially coupled version of Model~3 using the novel iterated block particle filter (IBPF) algorithm of \cite{ionides22}.
+These two concerns may explain the unrealistic forecasts and low likelihood of the model (see Table~\ref{tab:likes}).
+
+For our analysis, we calibrate the parameters of the spatially coupled version of Model~3 using the iterated block particle filter (IBPF) algorithm of \cite{ionides22}.
 This algorithm extends the work of \cite{ning21ibpf}, who provided theoretic justification for the version of the algorithm that only estimates unit-specific parameters.
@@
 
-The central goal of a forecast is to provide an accurate estimate of the future state of a system based on currently available data.
-When a mechanistic model is used, forecasts may also provide estimates of the future effects of potential interventions.
-Forecasting models are built using available scientific understanding, but forecasting can also be a way of testing new scientific hypotheses in real time \citep{lewis22}.
-In order to provide trustworthy information, however, the reliability of the forecast should be understood.
-In particular, researchers should account for various forms of uncertainty present in model forecasts, and  calibrate the proposed model to observed data.
-
-The most recently available information about a dynamic system should be used to obtain reliable forecasts.
-Using state space model $m$, this means that in order to obtain a forecast up to time $N+s$, where $s\geq 0$ and $N$ is the index for the last available data, we should simulate from the conditional density $f_{\bm{X}^{(\modelCounter)}_{N:s}, \bm{Y}^{\modelCounter}_{N:s}}\big(x_{N:s}, y_{N:s}\mid \hat{\paramVec}, \bm{X}^{(\modelCounter)}_{N} = \bm{x}^{(\modelCounter)}_N\big)$, where $\bm{x}^{(\modelCounter)}_N$ is the value of the latent state at time $N$.
-Because the latent state $\bm{X}^{(\modelCounter)}_N$ is unobservable, draws from the desired conditional density are unobtainable.
-Informed estimates of $\bm{X}^{(\modelCounter)}_N$ given the observed data $\bm{Y}^{(\modelCounter)}_{1:N} = y_{1:N}^*$ can easily be obtained using the filtering distribution.
-Let $\hat{\bm{X}}^{(\modelCounter), i}_N, \, i \in {1, 2, \ldots, J}$ be $\iid$ draws from the filtering distribution at time $N$, with density $\hat{\bm{X}}^{(\modelCounter), i}_N \sim f_{\bm{X}^{(\modelCounter)}_N|\bm{Y}^{(\modelCounter)}_{1:N}}(x_N \mid \hat{\paramVec}, y_{1:N}^*)$.
-A single model forecast can then be obtained by simulating from the model $f_{\bm{X}^{(\modelCounter)}_{N:s}, \bm{Y}^{(\modelCounter)}_{N:s}}\big(x_{N:s}, y_{N:s}\mid \hat{\paramVec}, \bm{X}^{(\modelCounter)}_{N} = \hat{\bm{X}}^{(\modelCounter), i}_N\big)$.
-Obtaining forecasts in this fashion allows us to obtain results that are consistent with the model assumptions and the observed data.
-In this particular case study, projecting the future state of the cholera epidemic starting from latent states that are draws from the filtering distribution allows the proposed models to benefit from the fact that very few cholera cases had been observed between 2018 and 2019, and that cases appear to be decreasing.
-This alone partially explains the unrealistic forecasts of \cite{lee20}, which were obtained by simulating the model from initial conditions;
-Table~S7 of their supplement material shows that of their simulations that were consistent with observing zero cases in 2019, all forecasts resulted in the elimination of cholera.
-
-It has been noted that the uncertainty in just a single parameter can lead to drastically different forecasts \citep{saltelli20};
-parameter uncertainty should therefore be considered when obtaining model forecasts to influence policy.
-If a Bayesian technique is used for parameter estimation, the most natural way to account for parameter uncertainty is to obtain simulations from the model where each simulation is obtained using parameters drawn from the posterior distribution.
-In the frequentist case, one approach is empirical Bayes, as was done by \cite{king15} (see supplement for details). 
-Both of these approaches share the similarity that parameters with higher likelihoods should be used more frequently in model forecasts than those that correspond to lower likelihoods.
-
-Because our direct aim is not to provide forecasts for policy recommendation, we do not include parameter uncertainty in our forecasts.
-Instead, we use the projections from point estimates to highlight a major deficiency of deterministic models, which is that the only variability in model projections is a result of parameter uncertainty, which leads to over-confidence in forecasts \cite{king15}.
-
-Forecasting can be used to simulate various interventions on a system; this feature is useful to inform policy and was the primary goal of \citet{lee20}.
+Forecasts aim to provide an accurate estimate of the future state of a system based on currently available data, together with an assessment of uncertainty.
+Mechanistic forecasting models, compatible with current scientific understanding, may also provide estimates of the future effects of potential interventions.
+Further, they may enable real-time testing of new scientific hypotheses \citep{lewis22}.
+
+Recent information about a dynamic system should be more relevant for a forecast than older information.
+This assertion may seem self-evident, but it is not the case for deterministic models, for which the initial conditions together with the parameters are sufficient for forecasting, and so recent data do not have special importance.
+Epidemiological forecasts based on deterministic models are not uncommon in practice, despite their limitations \citep{king15}.
+That may explain why \citet{lee20} chose to obtain forecasts by simulating the calibrated models forward from initial conditions.
+Here, we compare with a forecast projected from the filtered distribution: to obtain a forecast using POMP model $m$ at a collection of times $t_{N+1:N+s}$, where $N$ is the index for the last available data, we simulate forward from a draw from
+$f^{(m)}_{\bm{X}_N|\bm{Y}_{1:N}}(\bm{x}_{N} | \bm{y}^*_{1:N} ; \hat\theta)$
+where $\hat\theta$ is a vector of calibrated parameters.
+The decision not to do this partially explains the unsuccessful forecasts of \cite{lee20}: their Table~S7 shows that the subset of their simulations which were consistent with observing zero cases in 2019 predicted the elimination of cholera.
+In this case study, projecting the future state of the cholera epidemic starting from latent states that are draws from the filtering distribution allows the model-based forecasts to benefit from the fact that very few cholera cases were observed in 2018 and January 2019.
+
+Uncertainty in just a single parameter can lead to drastically different forecasts \citep{saltelli20}.
+Therefore, parameter uncertainty should also be considered when obtaining model forecasts to influence policy.
+If a Bayesian technique is used for parameter estimation, a natural way to account for parameter uncertainty is to obtain simulations from the model where each simulation is obtained using parameters drawn from the estimated posterior distribution.
+For frequentist inference, an empirical Bayes approach leads to similar methodology \citep{king15}; details are provided in the supplement. 
+Both of these approaches share the similarity that parameters are chosen for the forecast approximately in proportion to their corresponding value of the likelihood function, $f^{(m)}_{\bm{Y}_{1:N}}(\bm{y}_{1:N}^*; \theta)$.
+
+The primary forecasting goal of \citet{lee20} was to investigate the potential consequences of vaccination interventions on a system to inform policy.
 Outcomes of their study include estimates for the probability of cholera elimination and cumulative number of cholera infections under several possible vaccination scenarios.

### 2022-10-13 ms.Rnw
a few more edits to the discussion

@@
 This case study has investigated issues requiring attention when applying powerful new statistical techniques that can enable statistically efficient inference for a general class of partially observed Markov process models.
-Care must be taken to ensure that the computationally intensive numerical calculations are carried out adequately.
+Researchers should check that the computationally intensive numerical calculations are carried out adequately.
+Comparison against benchmarks and alternative model specifications should be considered to evaluate the statistical goodness-of-fit.
 Once that is accomplished, care is required to assess what causal conclusions can properly be inferred given the possibility of alternative explanations consistent with the data.

### 2022-10-13 ms.Rnw
some tidying/commenting code. haitipkg update needed

@@
 <<Setup, include=FALSE,echo=FALSE,results='hide'>>=
+
+# Load Required packages
 library(knitr)
@@
 
+# Set run level:
+#   1: Used primarily for debugging
+#   2: Used to test various results in a way that provides meaningful results, yet doesn't require an extensive amount of time to run.
+#   3: Used to get final results.
+#
+# Throughout the document, if run_level_3 results exist already exist,
+# then these results are used.
 RUN_LEVEL <- 1
@@
 
-# run_level_* for model 2 is not needed.
+# Check if run_level_* directories exist; if not, create them.
 for (i in c(1, 3)) {
@@
 
+# Model 2 doesn't need a run_level, since all computations are the same.
 if (!dir.exists('model2')) dir.create("model2/")
@@
 
+# Setting black and white ggplot2 theme for entire document.
 theme_set(theme_bw() + theme(plot.title = element_text(hjust = 0.5)))
 
-# 40 cores for doob, 8 for ito
-# doob_cores <- 40
-# gl_cores <- 36
-
+# Get the number of cores. The high performance computing cluster that we
+# used has 36 cores.
 cores <-  as.numeric(Sys.getenv('SLURM_NTASKS_PER_NODE', unset=NA))
@@
 
+# Create a mapping between department names and what should be shown in figures
 dep_labeller <- as_labeller(
@@
 
+# Vector used to arrange figures in the same order that was used in Lee et al (2020)
 plot_order <- c(
@@
 
+# Create "long" version of the observed cholera cases, for plotting purposes.
 dep_plot_df <- haitiCholera %>%
@@
 
+# Calculate total number of nationally aggregated cholera cases, using previous
+# data.
 true_agg_cases <- dep_plot_df %>%
@@
 set.seed(141837)
-sample_beta_nu <- function(mu_beta = 5.8, mu_nu = 0.96, n_params = N_PARAMS) {
-  # This function is used to sample starting values for beta and nu.
-  #
-  # params:
-  #   mu_beta: Average of the marginal distribution for beta1. Value was fit visually.
-  #   mu_nu: Average of the marginal distribution for nu. Value was fit visually.
-  # returns:
-  #   data.frame of size 300x2, values of beta1 and nu with truncated bivariate
-  #   normal relation.
-
-  # Determine a good value for the marginal standard deviations
-  sd_beta <- (12.1-mu_beta) / qnorm(0.995)
-  sd_nu <- (0.9-mu_nu) / qnorm(0.005)
-
-  # Create covariance matrix for the bivariate-normal distribution
-  cov_mat <- rbind(
-    c(sd_beta^2, -0.7 * sd_beta * sd_nu),
-    c(-0.7 * sd_beta * sd_nu, sd_nu^2)
-  )
-
-  # Sample from the bivariate normal distribution
-  X <- MASS::mvrnorm(n = n_params, mu = c(mu_beta, mu_nu), Sigma = cov_mat)
-  colnames(X) <- c("beta1", "nu")
-
-  # Identify values that fell outside of range, so they can be resampled
-  bad_beta <- which(X[, 'beta1'] > 12.1 | X[, 'beta1'] < 0.04)
-  bad_nu <- which(X[, 'nu'] > 1 | X[, 'nu'] < 0.9)
-  n_resamp <- length(unique(c(bad_beta, bad_nu)))
-
-  # Keep resampling values until all values fall within the range.
-  while (n_resamp > 0) {
-    to_resamp <- unique(c(bad_beta, bad_nu))
-    X[to_resamp, ] <- MASS::mvrnorm(n = n_resamp, mu = c(mu_beta, mu_nu), Sigma = cov_mat)
-
-    bad_beta <- which(X[, 'beta1'] > 12.1 | X[, 'beta1'] < 0.04)
-    bad_nu <- which(X[, 'nu'] > 1 | X[, 'nu'] < 0.9)
-    n_resamp <- length(unique(c(bad_beta, bad_nu)))
-  }
-
-  # Return results as a data.frame
-  as.data.frame(X)
-}
-# Run the function to get values of beta1 and nu
-starting_params <- sample_beta_nu()
-
-sample_rho <- function(beta, n_params = N_PARAMS) {
-  # This function is used to get starting values for the parameter rho, based
-  # on it's strong correlation with beta1 and nu. Here, we first sample beta1
-  # and nu, and then sample rho based on a regression on beta1. Additionally,
-  # it is noted that rho has a bi-modal distribution, with lower values of rho
-  # corresponding with high probability to higher values of beta1.
-  #
-  # params:
-  #   beta: previously sampled values of beta1 that will be used to get values
-  #         of rho.
-  # returns:
-  #   vector of length 300 that have an appropriate relationship with the parameter
-  #   beta1.
-
-  rhos <- rep(0, N_PARAMS)
-
-  # Get slope and intercept for the regression
-  slope <- (0.4/(5.1 - 6.75))
-  intercpt <- -6.75 * slope
-
-  # Determine which mode should the value of rho belong to, based on regression
-  case_when(
-    beta > 6.75 ~ 0,  # Large beta --> group "0"
-    # Small beta --> more likely to be group "1", with increasing probability as beta gets smaller
-    beta <= 6.75 & beta > (0.95 - intercpt) / slope ~ slope * beta + intercpt,
-    TRUE ~ 0.95
-  ) -> probs
-
-  # Sample the groups that rho belongs to
-  rho_group1 <- rbernoulli(N_PARAMS, probs)
-  rho_group2 <- !rho_group1
-
-  # Keep track of how many values are in each group, used later for resampling
-  n_group1 <- sum(rho_group1)
-  n_group2 <- sum(rho_group2)
-
-  # Resample values of rho that are larger than 1
-  group1 <- rnorm(n_group1, mean = 1, sd = 0.05)
-  n_bad1 <- sum(group1 > 1)
-  while(n_bad1 > 0) {
-    group1[group1 > 1] <- rnorm(n_bad1, mean = 1, sd = 0.175)
-    n_bad1 <- sum(group1 > 1)
-  }
-
-  # get standard deviation for regression
-  std_beta1 <- (beta[rho_group2] - mean(beta[rho_group2])) / sd(beta[rho_group2])
-
-  group2 <- 0.25 - 0.025 * std_beta1 - (rlnorm(n_group2, sdlog = 0.2) - 1)
-
-  # Resample values of rho in group2 as needed
-  which_bad2 <- group2 < .13
-  while(sum(which_bad2) > 0) {
-    group2[group2 < 0.13] <- 0.25 - 0.1 * std_beta1[which_bad2] - (rlnorm(sum(which_bad2), sdlog = 0.2) - 1)
-
-    which_bad2 <- group2 < .13
-  }
-
-  rhos[rho_group1] <- group1
-  rhos[rho_group2] <- group2
-
-  # Return the values of rho
-  rhos
-}
-# Sample rho, using already sampled beta1.
-starting_params$rho <- sample_rho(starting_params$beta1)
-
-sample_tau <- function(mu = 4.1, n_params = N_PARAMS) {
-  # Function used to sample tau. Tau is approximately independent of the other
-  # parameters.
-  #
-  # params:
-  #   mu: mean of the truncated normal distribution
-  # returns:
-  #   vector of length 300, sampled values of tau from truncated normal distribution
-
-  # Sample from normal distribution
-  X <- rnorm(n_params, mean = mu, sd = (1.7-mu) / qnorm(0.001))
-
-  # Get how many resamples are needed
-  n_resamp <- sum(X > 5.9 | X < 1.7)
-
-  # Keep resampling until all values fall within desired range
-  while (n_resamp > 0) {
-    X[X > 5.9 | X < 1.7] <- rnorm(n_resamp, mean = mu, sd = (1.7-mu) / qnorm(0.001))
-
-    n_resamp <- sum(X > 5.9 | X < 1.7)
-  }
-
-  # Return values of tau
-  X
-}
-# Use function to sample tau
-starting_params$tau <- sample_tau()
-
-# Here we note that there are additional parameters that were fit, but their
-# relationship between the remaining parameters is not given in Lee et al. (2020).
-# Because of this, we simply sample these parameters using a truncated normal distribution:
-
-POP <- 10911819
-E_0_probs <- dnorm(seq(1, 1752, 1), mean = 400, sd = (1752-400) / qnorm(0.99))
-I_0_probs <- dnorm(seq(1, 2271, 1), mean = 500, sd = (2271-500) / qnorm(0.99))
-starting_params$E_0 <- sample(1:1752, size = N_PARAMS, replace = TRUE, prob = E_0_probs) / POP
-starting_params$I_0 <- sample(1:2271, size = N_PARAMS, replace = TRUE, prob = I_0_probs) / POP
-starting_params$S_0 <- 1 - starting_params$E_0 - starting_params$I_0
-
-rm(sample_beta_nu, sample_rho, sample_tau, POP, I_0_probs, E_0_probs)
-gc()
+
+# Get starting parameters close to those that are provided
+# in Lee et al (2020)
+starting_params <- get_h1_startVals(N_PARAMS)
 
@@
 
+# Do a small global search for starting parameters for beta2-beta6
 bounds <- tribble(
@@
 
+# Get a sequence of starting values for the betas
 betas <- sobol_design(
@@
 
+# Combine betas with the starting_params data.frame object.
 epi_params <- cbind(starting_params, betas)
 
+# Get the names of all of the remaining coefficients (fixed parameters).
 not_in_start_params <- names(coef(h1_epi))[!names(coef(h1_epi)) %in% colnames(epi_params)]
 
+# Add fixed parameters to the matrix of starting parameters
 for (i in 1:length(not_in_start_params)) {
@@
 }
-
-rm(i)
 
@@
 # Include a flag to check if RUN_LEVEL == 3. If so, we might want to recompute.
-# TODO: Only save parameters and likelihoods, not entire MIF objects.
 if (RUN_LEVEL != 3 & file.exists("model1/run_level_3/lee1_epi_fit.rds")) {
@@
     {
+      # Create a list to save all of the results.
       lee1_epi_mif_results <- list()
 
+      # Loop through each of the starting parameters
       foreach(
@@
       ) %dopar% {
+
+        # Get the ith set of starting parameters
         r_params <- unlist(epi_params[i, ])
+
+        # Set the coefficients of model 1 to the ith starting values
         coef(h1_epi) <- r_params
+
+        # Perform short IF2 from the starting value
         mif2(
@@
           Nmif = NMIF,
-          cooling.fraction.50 = 0.5,  # Cooling set so that we will reach smaller rw.sd at the end than the end of the unit3 search.
+          cooling.fraction.50 = 0.5,
           rw.sd = epi_rw
         )
-      } -> lee1_epi_mif_out
-
+      } -> lee1_epi_mif_out  # Contains nrow(epi_params) mif2d objects
+
+      # For each output of the MIF2 objects, simulate Model 1 with the IF2
+      # parameters and save the states from the final week of the simulation
       foreach(i = 1:length(lee1_epi_mif_out), .combine = rbind) %do% {
+
+        # Get IF2 parameter set
         guess <- coef(lee1_epi_mif_out[[i]])
+
+        # Simulate with the parameter set
         sims <- simulate(
@@
         )  %>%
-          dplyr::filter(week == max(week)) %>%
+          dplyr::filter(week == max(week)) %>%  # Just keep final states, to use later.
           dplyr::select(-.id, -week)
@@
 rm(epi_rw, epi_params, not_in_start_params,
-   lower, upper, betas, bounds, starting_params)
+   lower, upper, betas, bounds, starting_params, i)
 gc()
@@
 \begin{frontmatter}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                                          %%
-%% Enter the title of your article here     %%
-%%                                          %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \title{Informing policy via dynamic models: Eliminating cholera in Haiti}
@@
 \begin{aug}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% Only one address is permitted per author. %%
-%% Only division, organization and e-mail is %%
-%% included in the address.                  %%
-%% Additional information can be included in %%
-%% the Acknowledgments section if necessary. %%
-%% ORCID can be inserted by command:         %%
-%% \orcid{0000-0000-0000-0000}               %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \author[A]{\fnms{Jesse}~\snm{Wheeler}\ead[label=e1]{jeswheel@umich.edu}\orcid{0000-0003-3941-3884}},
@@
 \author[A]{\fnms{Edward}~\snm{Ionides}\ead[label=e6]{ionides@umich.edu}}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% Addresses                                %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \address[A]{Statistics Department, University of Michigan\printead[presep={,\ }]{e1,e2,e3,e4,e5,e6}}
+
 \end{aug}
@@
 <<Plot_Reported_Cases, echo=FALSE, fig.height=3.5>>=
-plot_df <- haitiCholera %>%
-  select(-report) %>%
-  mutate(date = as.Date(date_saturday)) %>%
-  select(-date_saturday) %>%
-  pivot_longer(
-    data = .,
-    cols = -c(date),
-    names_to = 'Departement',
-    values_to = "Cases",
-  )
-
-ggplot(plot_df, aes(x = date, y = Cases + 1)) +
-  facet_wrap(~Departement, nrow = 2, labeller = dep_labeller) +
+
+# Plot reported cholera cases, by department. Creates Figure 1.
+ggplot(dep_plot_df, aes(x = date, y = cases + 1)) +
+  facet_wrap(~dep, nrow = 2, labeller = dep_labeller) +
   geom_line() +
-  theme(
-    axis.title.x = element_blank()
+  theme(axis.title.x = element_blank()) +
+  ylab('Reported Cases') +
+  scale_y_log10(
+    labels = scales::trans_format("log10", scales::math_format(10^.x))
   ) +
-  ylab('Reported Cases') +
-  scale_y_log10(labels = scales::trans_format("log10", scales::math_format(10^.x))) +
-  scale_x_date(date_labels = "'%y", breaks = seq.Date(from = as.Date("2011-01-01"), as.Date("2019-01-01"), by = '2 years'))
-
+  scale_x_date(
+    date_labels = "'%y",
+    breaks = seq.Date(
+      from = as.Date("2011-01-01"),
+      to = as.Date("2019-01-01"),
+      by = '2 years'
+    )
+  )
 @
@@
 <<FitModel1, echo=FALSE, message=FALSE, include=FALSE>>=
+
+# This code chunk is used to fit Model 1 using the IF2 algorithm.
 set.seed(636813)
 
+# Check RUN_LEVEL, and set hyper-parameters to IF2 and pfilter accordingly.
+#
+#                      RUN_LEVEL =    1,    2,      3
 NP_H1        <- switch(RUN_LEVEL,   100, 1000,   2000)
@@
 
+# Check first if RUN_LEVEL_3 file exists in directory, if not, create the
+# RUN_LEVEL version of the results and save them.
 if (RUN_LEVEL != 3 & file.exists("model1/run_level_3/haiti1_fit.rds")) {
@@
 
+# The result is a data.frame containing many parameter values and their
+# corresponding likelihood evaluation. To perform a likelihood profile,
+# we calculate the maximum likelihood for each value of betat.
 h1_prof_res <- h1_MIF %>%
@@
 
+# Save the likelihood of the best parameters, for use later.
 mod1_ll <- h1_MIF %>%
@@
 
+# Just a bit of tidying up: removing parameters that we won't use again.
 rm(NP_H1, NMIF_H1, NUM_TREND, NPROF, NREPS_EVAL, NP_EVAL, h1_MIF)
@@
 
+<<FitModel2, echo=FALSE, include=FALSE, message=FALSE>>=
+
+# This code chunk fits Model 2. This is a quick caclulation compared to models
+# 1 and 3, so no run_level is used.
+
+h2_fit <- bake(
+  file = paste0("model2/model2_fit.rds"), {
+    fit_haiti2()
+  },
+  timing = FALSE
+)
+@
+
 <<FitModel3, echo=FALSE, include=FALSE, message=FALSE>>=
+
+# This code chunk is used to fit Model 3 using the IBPF algorithm.
 set.seed(1859687)
@@
 
+# haiti3_fit contains ll estimate for many parameter sets, so we want to find
+# the highest value
+mod3_ll <- max(
+  haiti3_fit[[length(haiti3_fit)]]$logLiks[, 'logLik'],
+  na.rm = TRUE
+)
+
+# Just a bit of tidying up: removing parameters that we won't use again.
 rm(
@@
 
-<<Load Model Parameters, echo=FALSE, results='hide'>>=
-
-# haiti3_fit contains ll estimate for many parameter sets, so we want to find
-# the highest value
-mod3_ll <- max(
-  haiti3_fit[[length(haiti3_fit)]]$logLiks[, 'logLik'],
-  na.rm = TRUE
-)
-
-# Fit model 2. Note that this takes the same amount of time at all run
-# levels.
-h2_fit <- bake(
-  file = paste0("model2/model2_fit.rds"), {
-    fit_haiti2()
-  },
-  timing = FALSE
-)
-
-@
-
 <<table-input,echo=FALSE,eval=T, include=FALSE, message=FALSE>>=
 
+# Find the index for parameters of model 3 that correspond to the MLE.
 best_m3 <- which.max(haiti3_fit[[length(haiti3_fit)]]$logLiks$logLik)
 
+# Save the parameters that correspond to the MLE of Model 3.
 p3 <- haiti3_fit[[length(haiti3_fit)]]$params[best_m3, ]
@@
 <<Model_1_Sims_Figure, fig.height=2.4, fig.width=4.7, fig.align='center'>>=
+# This code chunk creates the figure containing simulations from Model 1
+
+# Simulate Model 1 from initial conditions.
 sims <- simulate(h1, nsim = 500, format = 'data.frame', seed = 3448931)
 
+# Calculate 0.025 and 0.975 percentiles of simulations, at each time point.
 quants <- sims %>%
@@
 
+# Plot the data, median value of the simulation, and estimated quantiles.
 ggplot() +
@@
   geom_line(data = quants, aes(x = date, y = q50 + 1), col = 'blue') +
-  geom_ribbon(data = quants, aes(x = date, ymin = q025 + 1, ymax = q975 + 1), alpha = 0.5) +
-  theme(axis.title.x = element_blank(),
+  geom_ribbon(  # Plot confidence region
+    data = quants,
+    aes(x = date, ymin = q025 + 1, ymax = q975 + 1),
+    alpha = 0.5
+  ) +
+  theme(
+    axis.title.x = element_blank(),
     axis.text = element_text(size = 8),
-        axis.title.y = element_text(size = 10))+
+    axis.title.y = element_text(size = 10)
+  )+
   ylab('Reported cholera cases') +
-  geom_vline(xintercept = lubridate::weeks(232) + lubridate::ymd("2010-10-16"), linetype = 'dashed') +
-  scale_y_log10(labels = scales::trans_format("log10", scales::math_format(10^.x))) +
-  scale_x_date(date_labels = "'%y", breaks = seq.Date(from = as.Date("2011-01-01"), as.Date("2019-01-01"), by = '1 year'))
+  geom_vline(  # Show change from epidemic -> endemic models.
+    xintercept = lubridate::weeks(232) + lubridate::ymd("2010-10-16"),
+    linetype = 'dashed'
+  ) +
+  scale_y_log10(  # Plot on the log-scale
+    labels = scales::trans_format("log10", scales::math_format(10^.x))
+  ) +
+  scale_x_date(  # Format the date on x-axis.
+    date_labels = "'%y",
+    breaks = seq.Date(
+      from = as.Date("2011-01-01"),
+      as.Date("2019-01-01"),
+      by = '1 year'
+    )
+  )
 @
@@
 The increase in model flexibility, however, results in hidden states that are inconsistent between model phases.
-% A naive solution to this would be to estimate only model parameters in separate phases and not the hidden states, but this may also result in hidden states in one phase of the model that are inconsistent with the other phase of the outbreak, since only the data from one phase is used to fit model parameters.
 The inclusion of a model break-point by \cite{lee20} is perhaps due to a challenging feature of fitting a deterministic model via least squares: discrepancies between model trajectories and observed case counts in highly infectious periods of a disease outbreak will result in greater penalty than the discrepancies between model trajectories and observed case counts in times of relatively low infectiousness, resulting in a bias towards accurately describing periods of high infectiousness.
@@
 <<Model 2 Trajectory, echo=FALSE>>=
-h2_epi_traj <- trajectory(h2_epi, params = h2_fit$h2_params, format = 'data.frame')
-h2_epi_traj$Ctotal <- rowSums(h2_epi_traj[, paste0("C", 1:10)]) * h2_fit$h2_params['Rho']
+
+# This code chunk calculates the "trajectory" of the differential equations
+# for Model 2.
+
+h2_epi_traj <- trajectory(
+  h2_epi,  # Model
+  params = h2_fit$h2_params,  # estimated parameters
+  format = 'data.frame'
+)
+
+# Display the output as nationally aggregated data, as was done in Lee et al (2020)
+h2_epi_traj$Ctotal <- rowSums(
+  h2_epi_traj[, paste0("C", 1:10)]  # Infections in each department
+) * h2_fit$h2_params['Rho']  # multiply by reporting rate, to compare to reported cases.
 
@@
 <<Model 3 SpatPOMP sims, echo=FALSE>>=
+
+# Number of simulations from Model 3. Because it's high-dimensional, we include
+# a run level to avoid time/memory issues when editing. Final results shown
+# with RUN_LEVEL = 3.
+
 h3_spat_nsim <- switch(RUN_LEVEL, 20, 100, 500)
 
+# Check if run_level_3 results exit, and use them if they do.
 if (RUN_LEVEL != 3 & file.exists("model3/run_level_3/SpatPompSims.rds")) {
@@
     file = paste0('model3/', rl_dir, 'SpatPompSims.rds'), {
-      # h3Spat <- haiti3_spatPomp()
-      # coef(h3Spat) <- haitipkg:::panelParms_toSpatParms3(p3)
-
+
+      # Simulate from the fitted version of Model 3
       h3Spat_sims <- simulate(
@@
 
+      # For each department and time-point, calculate percentiles of the
+      # simulations
       h3Spat_sims %>%
@@
         ) %>%
-        filter(date >= as.Date(lubridate::round_date(lubridate::date_decimal(h3Spat@t0), unit = 'day'))) %>%
+        # filter(
+        #   date >= as.Date(lubridate::round_date(lubridate::date_decimal(h3Spat@t0), unit = 'day'))
+        #   ) %>%
         mutate(date = as.Date(lubridate::round_date(date, unit = 'day')))
@@
 <<Model3_SpatPOMP_sims_Figure, fig.height=3>>=
+
+# Plot the reported cases and confidence region for each department.
+
 ggplot() +
@@
   geom_line(data = dep_plot_df, aes(x = date, y = cases + 1), col = 'black') +
-  geom_ribbon(data = h3Spat_quants,
+  geom_ribbon(  # plot confidence region.
+    data = h3Spat_quants,
     aes(x = date, ymin = q05 + 1, ymax = q95 + 1),
-              alpha = 0.4) +
-  facet_wrap(~factor(dep, levels = plot_order), nrow = 2,
-             labeller = dep_labeller) +
+    alpha = 0.4
+  ) +
+  facet_wrap(
+    ~factor(dep, levels = plot_order), nrow = 2, labeller = dep_labeller
+  ) +
   labs(y = 'Log Reported Cholera Cases') +
-  scale_x_date(date_labels = "'%y", breaks = seq.Date(from = as.Date("2011-01-01"), as.Date("2019-01-01"), by = '2 years')) +
+  scale_x_date(  # Format the date on the X-axis.
+    date_labels = "'%y",
+    breaks = seq.Date(
+      from = as.Date("2011-01-01"),
+      to = as.Date("2019-01-01"),
+      by = '2 years'
+    )
+  ) +
   theme(axis.title.x = element_blank()) +
-  scale_y_log10(labels = scales::trans_format("log10", scales::math_format(10^.x)))
+  scale_y_log10(  # Plot on log-scale
+    labels = scales::trans_format("log10", scales::math_format(10^.x))
+  )
 @
@@
 <<Table 2 Input, echo=FALSE, eval=TRUE>>=
+
+# TODO: Not needed after fitting Model 3.
 NREPS_EVAL <- switch(RUN_LEVEL,   3,   8,    36)
@@
 
+# TODO: The code below will not be needed once we update the fitting of model 3.
 # Calculate likelihood via block-particle filter
@@
 <<Get Model 1 Seasonality>>=
-plot_model_fun <- function(mod) {
-  params <- coef(mod)
-
-  beta1 <- params['beta1']
-  beta2 <- params['beta2']
-  beta3 <- params['beta3']
-  beta4 <- params['beta4']
-  beta5 <- params['beta5']
-  beta6 <- params['beta6']
-  betat <- params['betat']
-
-  get_beta <- function(time) {
-
-    covar <- t(mod@covar@table)
-    covar_df <- as.data.frame(covar)
-    covar_df$times <- mod@covar@times
-
-    betas <- matrix(c(beta1, beta2, beta3, beta4, beta5, beta6), nrow = 1)
-    si <- matrix(unlist(covar_df[time, -7]), ncol = 1)
-    if (time <= 430) {
-      return(as.numeric(exp(betas %*% si + betat * ((time - 215) / (430-215)))))
-    } else {
-      return(as.numeric(exp(betas %*% si + betat)))
-    }
-  }
-
-  return(get_beta)
-}
-
-get_beta1 <- plot_model_fun(h1)
-
-h1_seas_df <- data.frame(
-  'week' = 431:800,
-  'date' = lubridate::ymd("2010-10-16") + lubridate::weeks(431:800),
-  'trans' = purrr::map_dbl(431:800, get_beta1)
-) %>%
-  mutate(year = lubridate::year(date)) %>%
-  filter(year == 2020) %>%
-  mutate(trans_std = (trans - min(trans)) / (max(trans) - min(trans)))
-
-std_rain <- function(x) {
-  # This function simply standardizes the rain for us.
-  x / max(x)
-}
+# plot_model_fun <- function(mod) {
+#   params <- coef(mod)
+#
+#   beta1 <- params['beta1']
+#   beta2 <- params['beta2']
+#   beta3 <- params['beta3']
+#   beta4 <- params['beta4']
+#   beta5 <- params['beta5']
+#   beta6 <- params['beta6']
+#   betat <- params['betat']
+#
+#   get_beta <- function(time) {
+#
+#     covar <- t(mod@covar@table)
+#     covar_df <- as.data.frame(covar)
+#     covar_df$times <- mod@covar@times
+#
+#     betas <- matrix(c(beta1, beta2, beta3, beta4, beta5, beta6), nrow = 1)
+#     si <- matrix(unlist(covar_df[time, -7]), ncol = 1)
+#     if (time <= 430) {
+#       return(as.numeric(exp(betas %*% si + betat * ((time - 215) / (430-215)))))
+#     } else {
+#       return(as.numeric(exp(betas %*% si + betat)))
+#     }
+#   }
+#
+#   return(get_beta)
+# }
+#
+# get_beta1 <- plot_model_fun(h1)
+#
+# h1_seas_df <- data.frame(
+#   'week' = 431:800,
+#   'date' = lubridate::ymd("2010-10-16") + lubridate::weeks(431:800),
+#   'trans' = purrr::map_dbl(431:800, get_beta1)
+# ) %>%
+#   mutate(year = lubridate::year(date)) %>%
+#   filter(year == 2020) %>%
+#   mutate(trans_std = (trans - min(trans)) / (max(trans) - min(trans)))
+
+h1_seas_df <- get_h1_seasonality(h1)
+
+# std_rain <- function(x) {
+#   # This function simply standardizes the rain for us.
+#   x / max(x)
+# }
 
 df_rain <- haitiRainfall %>%
-  dplyr::summarize(
-    date = date, dplyr::across(Artibonite:`Sud-Est`, std_rain)
-  ) %>%
-  pivot_longer(
+  # dplyr::summarize(
+  #   # Standardize rainfall value using the function above
+  #   date = date, dplyr::across(Artibonite:`Sud-Est`, std_rain)
+  # ) %>%
+  pivot_longer(  # Convert data to "long" format
     data = .,
@@
   ) %>%
-  mutate(
-    year = lubridate::year(date),
-    week = lubridate::week(date)
-  ) %>%
+  mutate(year = lubridate::year(date), week = lubridate::week(date)) %>%
   group_by(year, week) %>%
-  summarize(national_weekly_rain = sum(rainfall)) %>%
+  summarize(nwr = sum(rainfall)) %>%
   ungroup() %>%  # Not needed, but good practice
   mutate(
-    date_week_start = lubridate::ymd(paste0(2014, "-01-01")) + lubridate::weeks(week - 1),
+    date_week_start = lubridate::ymd(paste0(2014, "-01-01")) +
+      lubridate::weeks(week - 1),
     week_date = as.Date(date_week_start, format = "%m-%d"),
-    std_national_rain = (national_weekly_rain - min(national_weekly_rain, na.rm = TRUE)) / (max(national_weekly_rain, na.rm = TRUE) - min(national_weekly_rain, na.rm = TRUE))
-  )
-
+    std_national_rain = (nwr - min(nwr)) / (max(nwr) - min(nwr))
+  )
+
+# Calculate the average national weekly rainfall.
 mean_rain <- df_rain %>%
@@
 
+# Create a figure showing annual rainfal, scaled from 0-1. This is top figure
 gg_rain <- df_rain %>%
@@
   ylab('Standardized Weekly\nNational Rain') +
-  theme(axis.title.x = element_blank(),
+  theme(  # Remove elements that are un-unsed becuase this is top figure
+    axis.title.x = element_blank(),
     axis.ticks.x = element_blank(),
@@
     axis.text.y = element_text(size = 7.5),
-        axis.title.y = element_text(size = 9))
-
+    axis.title.y = element_text(size = 9)
+  )
+
+# Create plot that shows the seasonality of Model 1, alongside the average
+# weekly national rainfall.
 gg_trans <- ggplot() +

### 2022-10-14 ms.Rnw
Comments/cleaning up

@@
 
-<<ReplicateModel1, echo=FALSE, message=FALSE, include=FALSE>>=
-# In this chunk, we replicate the results in Lee et al for model 1.
-# We do this replication here in the ms.Rnw file because we need to get
-# estimates for the likelihood of the models used by lee et al. The chunk has
-# the following steps:
-#   (1) Initialize some starting values close to those used in lee et al
-#   (2) do small MIF, similar to what was done in lee et al, to get a fit for
-#       epidemic period.
-#   (3) Simulate from model
-#   (4) Use simulations to initialize endemic phase of the model, and
-#       perform MIF, similar to what lee et al did for this phase
-#   (5) Simulate from model
-#   (6) Calculate the likelihood of each possible set of parameters.
-
-N_PARAMS   <- switch(RUN_LEVEL,  10,  20,  300)
-N_LEE_SIMS <- switch(RUN_LEVEL,  10,  20,   20)
-NMIF       <- switch(RUN_LEVEL,   5,  20,   50)
-NP_MIF     <- switch(RUN_LEVEL,  20,  50,  100)  # Too few, but matches Lee et al
-NP_EVAL    <- switch(RUN_LEVEL,  50, 200, 2000)
-NREPS_EVAL <- switch(RUN_LEVEL,   3,   5,   36)
-
-#-------------------------- Step 1 ----------------------
+<<CalculateLee1Likelihoods, echo=FALSE, message=FALSE, include=FALSE>>=
+
+# In this code chunk, we attempt to estimate the likelihood of Model 1 that 
+# is described by Lee et al (2020). Because this chunk produces some 
+# computationally intensive calculations that must also be done in the 
+# supplement, we save some additional information in this chunk that is only 
+# used in the supplement (si/si.Rnw) and not in the manuscript.
+
+# Setting algorithmic hyper-parameters based on RUN_LEVEL. 
+# 
+#                    RUN_LEVEL =  1,   2,    3
+N_PARAMS   <- switch(RUN_LEVEL,  10,  20,  300)  # Number of parameters to sample
+N_LEE_SIMS <- switch(RUN_LEVEL,  10,  20,   20)  # Number of simulations per parameter
+NMIF       <- switch(RUN_LEVEL,   5,  20,   50)  # Number of MIF iterations
+NP_MIF     <- switch(RUN_LEVEL,  20,  50,  100)  # Number of particles for IF2. This is too few for proper optimization, but matches Lee et al
+NP_EVAL    <- switch(RUN_LEVEL,  50, 200, 2000)  # Number of particles used to evaluate the likelihood 
+NREPS_EVAL <- switch(RUN_LEVEL,   3,   5,   36)  # Number of replicates of likelihood evaluation.
 
@@
 starting_params <- get_h1_startVals(N_PARAMS)
-
-#-------------------------- Step 2 ----------------------
 
@@
 
+# The process model of Lee et al (2020) didn't include over-dispersion
 epi_params$sig_sq <- 0
 
-# simply re-ordering to match model order
+# re-ordering starting parameters to match model order
 epi_params <- epi_params[, names(coef(h1_epi))]
 
+# Setting the random walk standard deviation. 
 epi_rw <- rw.sd(
@@
 
+# Some clean-up of unused variables, to avoid "Environment clutter".
 rm(epi_rw, epi_params, not_in_start_params,
@@
 
-# epi_complete <- as.data.frame(t(coef(lee1_epi_mif)))
-# epi_complete$logLik <- logLik(lee1_epi_mif)
-
-# GGally::ggpairs(
-#   filter(epi_results, logLik > -2200),
-#   columns = c("rho", "tau", "beta1", "nu", "logLik")
-# )
-
-# epi_complete <- filter(epi_complete, logLik > -2200)
-
+# Check if the run_level_3 results exist. If they do, load these results. 
+# If not, compute the run_level_<RUN_LEVEL> results and save them. 
 if (RUN_LEVEL != 3 & file.exists("model1/run_level_3/lee1_epi_evals.rds")) {
@@
 
-#-------------------------- Step 3 ----------------------
-
-# quants <- all_sims %>%
-#   select(.id, week, cases) %>%
-#   group_by(week) %>%
-#   summarize(
-#     q025 = quantile(cases, probs = 0.025, na.rm = TRUE),
-#     q50  = quantile(cases, probs = 0.500, na.rm = TRUE),
-#     q975 = quantile(cases, probs = 0.975, na.rm = TRUE)
-#   ) %>%
-#   ungroup() %>%
-#   mutate(date = lubridate::ymd("2010-10-16") + lubridate::weeks(week))
-
-#-------------------------- Step 4 ----------------------
-
+# Create the endemic version of the model
 h1_end <- haiti1(period = "endemic")
 
+# Our parameter values are now very similar to those used by Lee et al (2020). 
+# We therefore use the same rw.sd that was used in their source code. 
 end_rw <- rw.sd(
@@
 
-registerDoParallel(20)
-registerDoRNG(21489587)
-
+registerDoRNG(21489587)  # For reproducible computing
+
+# Store the parameter values from the end of the IF2 search for the epidemic 
+# model as the start values for the IF2 search for the endemic model, as was 
+# described by Lee et al (2020)
 end_start <- lee1_epi_mif$lee1_epi_mif_params
 
-# end_start <- epi_complete %>%
-#   select(-logLik)
-
+# Check if the run_level_3 results exist. If they do, load these results. 
+# If not, compute the run_level_<RUN_LEVEL> results and save them. 
 if (RUN_LEVEL != 3 & file.exists("model1/run_level_3/lee1_end_fit.rds")) {
@@
     {
-      foreach(
+      foreach(  # Parallel execution of the IF2 algorith. 
         i = 1:nrow(end_start),
@@
 
-        output <- list()
+        # Get the ith set of starting parameters for the endemic model. These 
+        # correspond to the parameters resulting from the ith execution of the 
+        # IF2 algorithm on the epidemic model
         r_params <- end_start[i, ]
 
+        # Calculate the starting states for each set of IF2 parameters. Because 
+        # there are <N_LEE_SIMS> simulations for each set of parameters, we 
+        # take the median value for each set of parameters as the starting 
+        # value for the endemic model. 
         lee1_epi_mif$start_states %>%
@@
 
-        # Set the initial state
+        # Set the initial state by creating a new rinit function for endemic 
+        # model. 
         new_rinit <- Csnippet(
@@
 
+        # Create a new pomp object representing the endemic model, where the 
+        # starting state is the median of the latent states from the epidemic 
+        # model for the parameter set i. 
         h1_end <- h1_end %>% pomp(
@@
 
+        # Set model coefficients to be the parameter set i.
         coef(h1_end) <- r_params
@@
           cooling.type = "hyperbolic",
-          cooling.fraction.50 = 0.05,  # Cooling set so that we will reach smaller rw.sd at the end than the end of the unit3 search.
+          cooling.fraction.50 = 0.05,  # Matches Lee et al (2020)
           rw.sd = end_rw

### 2022-10-14 si/si.Rnw
Comments/cleaning up

@@
 Details about the measurement models and how latent states were initialized for the epidemic model were not provided by \citet{lee20sup} and therefore these details must be inferred by looking at the provided source code.
-According to repository comments, the files \code{fitInPieces3paramsCleanMay2019Public.py} and \code{fitInPiecesMuWithFracSusFixedAllInfectionsPublic.py} were used to fit the epidemic and endemic phases of the model respectively, although it is apparent that these exact files were not used to obtain the reported results.
-These files contain errors that making them impossible to run, and therefore these files cannot be the code that generated the reported model fits\footnote{One example of why the code cannot be run that the file loads functions from a non-extant file named \code{choleraEqs.py} in line 13 rather than \code{choleraEqsPublic.py}.}.
+According to repository comments, the files \code{fit\allowbreak In\allowbreak Pieces\allowbreak 3params\allowbreak Clean\allowbreak May2019\allowbreak Public.py} and \code{fit\allowbreak In\allowbreak Pieces\allowbreak Mu\allowbreak With\allowbreak Frac\allowbreak Sus\allowbreak Fixed\allowbreak All\allowbreak Infections\allowbreak Public.py} were used to fit the epidemic and endemic phases of the model respectively, although it is apparent that these exact files were not used to obtain the reported results.
+These files contain errors that making them impossible to run, and therefore these files cannot be the code that generated the reported model fits\footnote{One example of why the code cannot be run that the file loads functions from a non-extant file named \code{choleraEqs.py} in line 13 rather than \code{cholera\allowbreak Eqs\allowbreak Public.py}.}.
 The inability to replicate the results by \citet{lee20} by running the provided source code makes checking whether or not a our numeric implementation faithfully represents their results very difficult.
@@
 \jwc{PROBABLY DELETE THIS NEXT BIT, BUT IT'S USEFUL TO DOCUMENT IT FOR NOW...}
-One concern that arises when inspecting the files \code{fitInPieces3paramsCleanMay2019Public.py} and \code{fitInPiecesMuWithFracSusFixedAllInfectionsPublic.py} is that the output of these files (which contains the estimated parameters) was not referenced/used in either the file that generates the sensitivity analysis or the file that generates the model simulations; instead, alternative (non-extant) files with similar names are loaded, suggesting parameters that were used to generate the final results were not a direct result of the estimated parameters.
-Furthermore, inspection of the script \code{simulateHaitiFullTimeInPiecesAllVaccinationScenariosSingleParameterVectorAllInfectionsPublic.py}---which is used to obtain simulations for each of the vaccination scenarios---results in a number of concerns.
-First, the script uses parameters from the pickle file \code{leastSquaresDataUpTo2019/fitEpidemicPiece\_resultsMatrix\_18Mar2019.pickle}, which only appears in the repository as a commented out line of code in the file \code{generateSetsOfParametersForSensitivityAnalysisPublic.py}, suggesting that the set of parameters used to simulate each vaccination scenario was not the most recent set of parameters generated by the sensitivity analysis.
-Second, the script appears to use the set of differential equations from the function \code{choleraEqs11WithVaccinationNetwork} for the epidemic phase, even though parameter values for this phase were estimated using \code{choleraEqs10WithVaccinationNetwork}.
+One concern that arises when inspecting the files \code{fit\allowbreak In\allowbreak Pieces3\allowbreak params\allowbreak Clean\allowbreak May\allowbreak 2019\allowbreak Public.py} and \code{fit\allowbreak In\allowbreak Pieces\allowbreak Mu\allowbreak With\allowbreak Frac\allowbreak Sus\allowbreak Fixed\allowbreak All\allowbreak Infections\allowbreak Public.py} is that the output of these files (which contains the estimated parameters) was not referenced/used in either the file that generates the sensitivity analysis or the file that generates the model simulations; instead, alternative (non-extant) files with similar names are loaded, suggesting parameters that were used to generate the final results were not a direct result of the estimated parameters.
+Furthermore, inspection of the script \code{simulate\allowbreak Haiti\allowbreak Full\allowbreak Time\allowbreak In\allowbreak Pieces\allowbreak All\allowbreak Vaccination\allowbreak Scenarios\allowbreak Single\allowbreak Parameter\allowbreak Vector\allowbreak All\allowbreak Infections\allowbreak Public.py}---which is used to obtain simulations for each of the vaccination scenarios---results in a number of concerns.
+First, the script uses parameters from the pickle file \code{least\allowbreak Squares\allowbreak Data\allowbreak UpTo\allowbreak 2019\allowbreak /fit\allowbreak Epidemic\allowbreak Piece\_results\allowbreak Matrix\_18\allowbreak Mar\allowbreak 2019.pickle}, which only appears in the repository as a commented out line of code in the file \code{generate\allowbreak Sets\allowbreak Of\allowbreak Parameters\allowbreak For\allowbreak Sensitivity\allowbreak Analysis\allowbreak Public.py}, suggesting that the set of parameters used to simulate each vaccination scenario was not the most recent set of parameters generated by the sensitivity analysis.
+Second, the script appears to use the set of differential equations from the function \code{cholera\allowbreak Eqs11\allowbreak With\allowbreak Vaccination\allowbreak Network} for the epidemic phase, even though parameter values for this phase were estimated using \code{cholera\allowbreak Eqs10\allowbreak With\allowbreak Vaccination\allowbreak Network}.
 It is possible, however, that we simply are interpreting the code incorrectly and that these concerns are unwarranted.

### 2022-10-17 ms.Rnw
Updated how Model 1 is replicated.

@@
 #                    RUN_LEVEL =  1,   2,    3
-N_PARAMS   <- switch(RUN_LEVEL,  10,  20,  300)  # Number of parameters to sample
+N_PARAMS   <- switch(RUN_LEVEL,  10,  20,  nrow(h1LeeStartsEpi))  # Number of parameters to sample
 N_LEE_SIMS <- switch(RUN_LEVEL,  10,  20,   20)  # Number of simulations per parameter
-NMIF       <- switch(RUN_LEVEL,   5,  20,   50)  # Number of MIF iterations
-NP_MIF     <- switch(RUN_LEVEL,  20,  50,  100)  # Number of particles for IF2. This is too few for proper optimization, but matches Lee et al
 NP_EVAL    <- switch(RUN_LEVEL,  50, 200, 2000)  # Number of particles used to evaluate the likelihood
@@
 
+h1_epi <- haiti1()
+
 set.seed(141837)
-
-# Get starting parameters close to those that are provided
-# in Lee et al (2020)
-starting_params <- get_h1_startVals(N_PARAMS)
-
-# Create epi model for haiti1
-h1_epi <- haiti1()
-
-# Do a small global search for starting parameters for beta2-beta6
-bounds <- tribble(
-  ~param, ~lower, ~upper,
-  "beta2", 1e-08, 10,
-  "beta3", 1e-08, 10,
-  "beta4", 1e-08, 10,
-  "beta5", 1e-08, 10,
-  "beta6", 1e-08, 10
-)
-
-lower <- bounds$lower
-names(lower) <- bounds$param
-
-upper <- bounds$upper
-names(upper) <- bounds$param
-
-# Get a sequence of starting values for the betas
-betas <- sobol_design(
-  lower = lower,
-  upper = upper,
-  nseq  = N_PARAMS
-)
-
-# Combine betas with the starting_params data.frame object.
-epi_params <- cbind(starting_params, betas)
-
-# Get the names of all of the remaining coefficients (fixed parameters).
-not_in_start_params <- names(coef(h1_epi))[!names(coef(h1_epi)) %in% colnames(epi_params)]
-
-# Add fixed parameters to the matrix of starting parameters
-for (i in 1:length(not_in_start_params)) {
-  epi_params[, not_in_start_params[i]] <- coef(h1_epi)[not_in_start_params[i]]
-}
-
-# The process model of Lee et al (2020) didn't include over-dispersion
-epi_params$sig_sq <- 0
-
-# re-ordering starting parameters to match model order
-epi_params <- epi_params[, names(coef(h1_epi))]
-
-# Setting the random walk standard deviation. 
-epi_rw <- rw.sd(
-  # Parameters close to desired values
-  beta1 = 0.001,
-  tau   = 0.001,
-  nu    = 0.001,
-  rho   = 0.001,
-  # Parameters not close to desired values
-  beta2 = 0.02,
-  beta3 = 0.02,
-  beta4 = 0.02,
-  beta5 = 0.02,
-  beta6 = 0.02,
-  E_0   = ivp(0.1),
-  I_0   = ivp(0.1)
-)
-
-registerDoRNG(48963587)
-
-# Check if RUN_LEVEL == 3 results exists, if so, use them.
-# Include a flag to check if RUN_LEVEL == 3. If so, we might want to recompute.
-if (RUN_LEVEL != 3 & file.exists("model1/run_level_3/lee1_epi_fit.rds")) {
-  lee1_epi_mif <- readRDS("model1/run_level_3/lee1_epi_fit.rds")
-} else {
-  lee1_epi_mif <- bake(
-    file = paste0("model1/", rl_dir, "lee1_epi_fit.rds"),
-    {
-      # Create a list to save all of the results.
-      lee1_epi_mif_results <- list()
-
-      # Loop through each of the starting parameters
-      foreach(
-        i = 1:nrow(epi_params),
-        .packages = c('pomp'),
-        .combine = c
-      ) %dopar% {
-
-        # Get the ith set of starting parameters
-        r_params <- unlist(epi_params[i, ])
-
-        # Set the coefficients of model 1 to the ith starting values
-        coef(h1_epi) <- r_params
-
-        # Perform short IF2 from the starting value
-        mif2(
-          h1_epi,
-          Np = NP_MIF,
-          Nmif = NMIF,
-          cooling.fraction.50 = 0.5,
-          rw.sd = epi_rw
-        )
-      } -> lee1_epi_mif_out  # Contains nrow(epi_params) mif2d objects
-
-      # For each output of the MIF2 objects, simulate Model 1 with the IF2
-      # parameters and save the states from the final week of the simulation
-      foreach(i = 1:length(lee1_epi_mif_out), .combine = rbind) %do% {
-
-        # Get IF2 parameter set
-        guess <- coef(lee1_epi_mif_out[[i]])
-
-        # Simulate with the parameter set
-        sims <- simulate(
-          h1_epi,
-          n = N_LEE_SIMS,
-          format = 'data.frame',
-          params = guess
-        )  %>%
-          dplyr::filter(week == max(week)) %>%  # Just keep final states, to use later.
-          dplyr::select(-.id, -week)
-
-        sims$param_set <- i
-        sims
-      } -> start_states
-
-      lee1_epi_mif_results$lee1_epi_mif_params <- t(coef(lee1_epi_mif_out))
-      lee1_epi_mif_results$start_states <- start_states
-      lee1_epi_mif_results
-    }
-  )
-}
-
-# Some clean-up of unused variables, to avoid "Environment clutter".
-rm(epi_rw, epi_params, not_in_start_params,
-   lower, upper, betas, bounds, starting_params, i)
-gc()
 
@@
 
+      which_params <- sample(
+        1:nrow(h1LeeStartsEpi),
+        size = N_PARAMS,
+        replace = FALSE
+      )
+      epi_params <- h1LeeStartsEpi[which_params, ]
+
       # Create data.frame to save results
       lee_epi_ll <- data.frame(
-        "pfLL" = rep(0, length(lee1_epi_mif)),
-        "pfse" = rep(0, length(lee1_epi_mif))
-      )
-
-      for (j in 1:nrow(lee1_epi_mif$lee1_epi_mif_params)) {
+        "pfLL" = NA_real_,
+        "pfse" = NA_real_,
+        "parid" = epi_params$parid
+      )
+
+      for (j in 1:nrow(lee_epi_ll)) {
         # Get parameters of interest
-        pf_params <- lee1_epi_mif$lee1_epi_mif_params[j, ]
+        pf_params <- epi_params[j, -1][names(coef(h1_epi))]
 
@@
 h1_end <- haiti1(period = "endemic")
-
-# Our parameter values are now very similar to those used by Lee et al (2020). 
-# We therefore use the same rw.sd that was used in their source code. 
-end_rw <- rw.sd(
-  beta1 = 0.05,
-  tau   = 0.005,
-  nu    = 0.04,
-  rho   = 0.01,
-  beta2 = 0.05,
-  beta3 = 0.05,
-  beta4 = 0.05,
-  beta5 = 0.05,
-  beta6 = 0.05
-)
-
-registerDoRNG(21489587)  # For reproducible computing
-
-# Store the parameter values from the end of the IF2 search for the epidemic 
-# model as the start values for the IF2 search for the endemic model, as was 
-# described by Lee et al (2020)
-end_start <- lee1_epi_mif$lee1_epi_mif_params
-
-# Check if the run_level_3 results exist. If they do, load these results. 
-# If not, compute the run_level_<RUN_LEVEL> results and save them. 
-if (RUN_LEVEL != 3 & file.exists("model1/run_level_3/lee1_end_fit.rds")) {
-  lee1_end_mif <- readRDS("model1/run_level_3/lee1_end_fit.rds")
-} else {
-  lee1_end_mif <- bake(
-    file = paste0("model1/", rl_dir, "lee1_end_fit.rds"),
-    {
-      foreach(  # Parallel execution of the IF2 algorith. 
-        i = 1:nrow(end_start),
-        .packages = c('pomp'),
-        .combine = rbind
-      ) %dopar% {
-
-        # Get the ith set of starting parameters for the endemic model. These 
-        # correspond to the parameters resulting from the ith execution of the 
-        # IF2 algorithm on the epidemic model
-        r_params <- end_start[i, ]
-
-        # Calculate the starting states for each set of IF2 parameters. Because 
-        # there are <N_LEE_SIMS> simulations for each set of parameters, we 
-        # take the median value for each set of parameters as the starting 
-        # value for the endemic model. 
-        lee1_epi_mif$start_states %>%
-          filter(param_set == i) %>%
-          select(-param_set) %>%
-          sapply(., median, na.rm = TRUE) -> rinit_parms
-
-        # Set the initial state by creating a new rinit function for endemic 
-        # model. 
-        new_rinit <- Csnippet(
-          sprintf(
-            "
-          S = nearbyint(%f);
-          E = nearbyint(%f);
-          I = nearbyint(%f);
-          A = nearbyint(%f);
-          R = nearbyint(%f);
-          incid = nearbyint(%f);
-          foival = %f;
-          Str0 = nearbyint(%f);
-          Sout = nearbyint(%f);
-          Sin = nearbyint(%f);
-          ",
-          rinit_parms['S'],
-          rinit_parms['E'],
-          rinit_parms['I'],
-          rinit_parms['A'],
-          rinit_parms['R'],
-          rinit_parms['incid'],
-          rinit_parms['foival'],
-          rinit_parms['Str0'],
-          rinit_parms['Sout'],
-          rinit_parms['Sin']
-          )
-        )
-
-        # Create a new pomp object representing the endemic model, where the 
-        # starting state is the median of the latent states from the epidemic 
-        # model for the parameter set i. 
-        h1_end <- h1_end %>% pomp(
-          rinit = new_rinit,
-          statenames = c(
-            "S", "E", "I", "A", "R", "incid",
-            "foival", "Str0", "Sout", "Sin"
-          )
-        )
-
-        # Set model coefficients to be the parameter set i.
-        coef(h1_end) <- r_params
-        mf_end <- mif2(
-          h1_end,
-          Np = NP_MIF,
-          Nmif = NMIF,
-          cooling.type = "hyperbolic",
-          cooling.fraction.50 = 0.05,  # Matches Lee et al (2020)
-          rw.sd = end_rw
-        )
-
-        coef(mf_end)
-      }
-    }
-  )
-}
-
-rm(end_rw)
-gc()
 
@@
 
+      end_params <- h1LeeStartsEnd[h1LeeStartsEnd$parid %in% lee_epi_ll$parid, ]
+
       # Create data.frame to save results
       lee_end_ll <- data.frame(
-        "pfLL" = rep(0, nrow(lee1_end_mif)),
-        "pfse" = rep(0, nrow(lee1_end_mif))
+        "pfLL" = NA_real_,
+        "pfse" = NA_real_,
+        'parid' = end_params$parid
       )
@@
         # Get parameters of interest
-        pf_params <- lee1_end_mif[j, ]
+        pf_params <- end_params[j, ][names(coef(h1_end))]
 
@@
 
-rm(lee1_end_mif, lee1_epi_mif)
-gc()
-
-# MIF_params <- matrix(nrow = length(lee1_end_mif$m2) / 2, ncol = length(coef(lee1_end_mif[[1]])))
-# all_sims_end <- data.frame()
-# liks <- rep(0, nrow(end_start))
-# for (i in 1:nrow(end_start)) {
-#   MIF_params[i, ] <- coef(lee1_end_mif[[(2 * i) - 1]])
-#   all_sims_end <- rbind(all_sims_end, lee1_end_mif[[(2 * i)]])
-#   liks[i] <- logLik(lee1_end_mif[[(2 * i) - 1]])
-# }
-#
-# colnames(MIF_params) <- names(coef(lee1_end_mif[[1]]))
-# end_results <- as.data.frame(MIF_params)
-# end_results$logLik <- liks
-
-# GGally::ggpairs(
-#   filter(end_results, logLik > -2200),
-#   columns = c("rho", "tau", "beta1", "nu", "logLik")
-# )
-
-#-------------------------- Step 6 ----------------------
-
-mod1_lee_ll <- max(lee_end_ll$pfLL, na.rm = TRUE) + max(lee_epi_ll$pfLL, na.rm = TRUE)
+# Combine epidemic and endemic likelihoods by parid to get the likelihood of
+# the complete model.
+mod1_all_likes <- dplyr::inner_join(
+  x = lee_epi_ll,
+  y = lee_end_ll,
+  by = "parid"
+  ) %>%
+  dplyr::mutate(
+    joint_ll = pfLL.x + pfLL.y
+  )
+
+# Approximate the likelihood of the Lee et al (2020) version of Model 1 by
+# using the parameter sets that resulted in the maximum likelihood.
+mod1_lee_ll <- max(mod1_all_likes$joint_ll, na.rm = TRUE)
 
 rm(
-  lee_end_ll, lee_epi_ll, h1_end, h1_epi, end_start,
-  N_LEE_SIMS, N_PARAMS, NMIF, NP_EVAL, NP_MIF, NREPS_EVAL
+  lee_end_ll, lee_epi_ll, h1_end, h1_epi, i, mod1_all_likes,
+  N_LEE_SIMS, N_PARAMS, NP_EVAL, NREPS_EVAL, end_params, epi_params,
+  pf_params, j, ll_evals, which_params
 )
@@
 Values in parenthesis are corresponding values using \citet{lee20} parameter estimates.
-$\textsuperscript{1}$The reported likelihood is an upper bound of the likelihood of the \citet{lee20} model as it is the largest likelihood obtained using their parameter calibration regime.
+$\textsuperscript{1}$The reported likelihood is an upper bound of the likelihood of the \citet{lee20} model.
 $\textsuperscript{2}$\cite{lee20} fit Model~3 to a subset of the data (March 2014 onward, excluding data from Ouest in 2015-2016).
@@
 On this same subset, our model has a likelihood of $\Sexpr{myround(mod3_spatPomp_res$subset_ll, 1)}$.
+See Sec.~S5 of the supplement material for more details on estimating the likelihood of the \cite{lee20} models.
 }

### 2022-10-17 si/si.Rnw
Updated how Model 1 is replicated.

@@
 Original source code is publicly available with DOI: 10.5281/zenodo.3360991.
-Despite having source code available, no point estimates for model parameters were provided by \cite{lee20}.
+The final results reported by the Model~1 authors were obtained by using several different parameter sets rather than a single point estimate.
 According to the supplement materials, this was because model realizations from a single parameter set retained substantial variability, but multiple realizations from a collection of parameter sets resulted in a reasonable visual fit to the data.
 We are also inclined to believe that the use of multiple parameter values was in part intended to account for parameter uncertainty (as mentioned in our main text), an effort by the Model~1 authors that we applaud.
-We note, however, that simulations from each of the parameter sets were treated with equal importance when being used to diagnose the model fit and make inference on the system.
-This practice is problematic given Figures S8 and S9 of the supplement material, which suggest that some parameter sets that were used for inference were several hundred log-likelihood units lower than the best performing sets of parameters.
-When accounting for parameter uncertainty, one should instead weigh resulting model projections based on the likelihood of the set of parameters that were used to obtain the model realizations, as was done by \cite{king15}.
-
-Not providing a point estimate for model parameters also has the effect of making model reproduction a more difficult task.
-We therefore rely on the range and median values for each of the final model parameters are provided in Tables S10 and S11 of the supplement, and bivariate relationships between some of the fitted parameters (Figures S8 and S9) in order to replicate their results.
-
-We begin by constructing 300 different parameter sets based on the basic summaries and figures provided by the Model~1 authors.
-This was primarily done using truncated-bivariate normal distributions, where samples that are drawn outside of the desired range of parameters are re-sampled from the distribution.
-Because the likelihood surface is complex and almost certainly includes higher than two-level interactions between parameter values, we then perform iterated filtering for the epidemic phase of the data using the 300 different parameter sets as starting points.
+Simulations from each of the parameter sets, however, were treated with equal importance when being used to diagnose the model fit and make inference on the system.
+This is problematic given Figures S8 and S9 of the supplement material, which suggest that some parameter sets that were used for inference were several hundred log-likelihood units lower than other parameter sets.
+Such a large difference in log-likelihoods is well beyond the threshold of statistical uncertainty determined by Wilks' theorem.
+
+To fully reproduce the results of the Model~1 authors, it is necessary to use the exact same set of model parameters.
+Because the parameters used to obtain the result given by \cite{lee20} are not publicly available, we relied on the source code provided by the Model~1 authors to approximately recreate the parameter set.
+Due to software updates since the publication of the source code, we were unable to produce the exact same set of parameters.
+Still, running the publicly available source code resulted in a set of parameters that are visually similar to those used by the Model~1 authors (See Figures~\ref{fig:PlotEpiDist} and \ref{fig:plotEndParams}).
+Furthermore, simulations using the set of parameters produced by the source code appear practically equivalent to those displayed by \cite{lee20} (See Figure~\ref{fig:plotMod1Sims}).
 
@@
 h1_epi <- haiti1()
-
 RL_dir <- paste0("run_level_", run_level, "/")
-lee1_epi_mif <- readRDS(file = paste0("../model1/", RL_dir, "lee1_epi_fit.rds"))
 @
@@
 <<Simulate Lee Model 1, echo=FALSE, cache=TRUE>>=
-N_LEE_SIMS <- switch(run_level,  10,  15,   20)
+N_LEE_SIMS <- 20
 
@@
 foreach(
-  i = 1:length(lee1_epi_mif),
+  i = 1:length(h1LeeStartsEpi),
   .combine = rbind
 ) %dopar% {
-  guess <- lee1_epi_mif$lee1_epi_mif_params[i, ]
+  guess <- h1LeeStartsEpi[i, ]
   sims <- simulate(h1_epi, nsim = N_LEE_SIMS, format = 'data.frame',
@@
 } -> all_sims
-
 @
@@
 <<PlotEpiDist, echo=FALSE, fig.cap="Compare to Figure S8 of Lee et al. (2020) supplement.", fig.width=7, fig.height=7, fig.align='center'>>=
-lee1_epi_evals <- readRDS(file = paste0("../model1/", RL_dir, "lee1_epi_evals.rds"))
-
-epi_results <- as.data.frame(lee1_epi_mif$lee1_epi_mif_params)
-epi_results$logLik <- lee1_epi_evals$pfLL
-
 GGally::ggpairs(
-  dplyr::filter(epi_results, logLik > -20000),
-  columns = c("rho", "tau", "beta1", "nu", "logLik")
+  h1LeeStartsEpi,
+  columns = c("rho", "tau", "beta1", "nu", "loglik")
 )
@@
 h1_end <- haiti1(period = "endemic")
-
-lee1_end_mif <- readRDS(
-  file = paste0("../model1/", RL_dir, "lee1_end_fit.rds")
-)
 
@@
 
-Following the model fitting scheme used by the Model~1 authors, we then use these resulting parameters as starting values for the endemic phase, and refit parameters using MIF2.
-To do this, we use the same number of particles and MIF iterations that were used by \citet{lee20}.
-The resulting bivariate relations between $\rho, \tau, \beta_1$ and $\nu$ are given in Figure~\ref{fig:plotEndParams}.
-
 <<FitEnd, echo=FALSE, message=FALSE, include=FALSE>>=
 foreach(
-  i = 1:nrow(lee1_end_mif),
+  i = 1:nrow(h1LeeStartsEnd),
   .combine = rbind
 ) %dopar% {
-  r_params <- lee1_end_mif[i, ]
-
-  lee1_epi_mif$start_states %>%
-    filter(param_set == i) %>%
-    select(-param_set) %>%
-    sapply(., median, na.rm = TRUE) -> rinit_parms
-
-  # Set the initial state
-  new_rinit <- Csnippet(
-    sprintf(
-      "
-          S = nearbyint(%f);
-          E = nearbyint(%f);
-          I = nearbyint(%f);
-          A = nearbyint(%f);
-          R = nearbyint(%f);
-          incid = nearbyint(%f);
-          foival = %f;
-          Str0 = nearbyint(%f);
-          Sout = nearbyint(%f);
-          Sin = nearbyint(%f);
-          ",
-      rinit_parms['S'],
-      rinit_parms['E'],
-      rinit_parms['I'],
-      rinit_parms['A'],
-      rinit_parms['R'],
-      rinit_parms['incid'],
-      rinit_parms['foival'],
-      rinit_parms['Str0'],
-      rinit_parms['Sout'],
-      rinit_parms['Sin']
-    )
-  )
-
-  h1_end <- h1_end %>% pomp(
-    rinit = new_rinit,
-    statenames = c(
-      "S", "E", "I", "A", "R", "incid",
-      "foival", "Str0", "Sout", "Sin"
-    )
-  )
-
+  r_params <- h1LeeStartsEnd[i, ][names(coef(h1_end))]
   coef(h1_end) <- r_params
@@
 (2020) supplement.", cache=TRUE, echo=FALSE>>=
-lee1_end_evals <- readRDS(file = paste0("../model1/", RL_dir, "lee1_end_evals.rds"))
-
-end_results <- as.data.frame(lee1_end_mif)
-end_results$logLik <- lee1_epi_evals$pfLL
-
 GGally::ggpairs(
-  filter(end_results, logLik > -20000),
-  columns = c("rho", "tau", "beta1", "nu", "logLik")
-)
-@
-
-Now that we have fit the parameters, we can simulate from the models.
-The results of these simulations are shown in Figure~\ref{fig:plotMod1Sims}.
-While not a perfect replication of the results given by \citet{lee20sup}, the similarities between the model simulations, and the fact that we based our implementation of Model~1 primarily on the source code that was provided, makes us confident that we have been able to obtain a reliable representation of the model described by \citet{lee20}.
-
-<<plotMod1Sims, echo=FALSE, fig.cap="Simulations from various parameter sets.
-Simulations from parameter sets with likelihood $<$ -2200 are removed.
-Compare to Figure S7 of \\cite{lee20sup}.", cache=TRUE, fig.width=6.5, fig.height=3.75>>=
-
-epi_complete <- filter(epi_results, logLik > -20000)
-bad_epi_pars <- which(epi_results$logLik < -20000)
+  h1LeeStartsEnd,
+  columns = c("rho", "tau", "beta1", "nu", "loglik")
+)
+@
+
+<<echo=FALSE, include=FALSE, message=FALSE>>=
+h1_epi_evals <- readRDS(paste0("../model1/", RL_dir, "lee1_epi_evals.rds"))
+h1_end_evals <- readRDS(paste0("../model1/", RL_dir, "lee1_end_evals.rds"))
+
+mod1_all_likes <- dplyr::inner_join(
+  x = h1_epi_evals,
+  y = h1_end_evals,
+  by = "parid"
+  ) %>%
+  dplyr::mutate(
+    joint_ll = pfLL.x + pfLL.y
+  )
+
+# Approximate the likelihood of the Lee et al (2020) version of Model 1 by
+# using the parameter sets that resulted in the maximum likelihood.
+mod1_lee_ll <- max(mod1_all_likes$joint_ll, na.rm = TRUE)
+@
+
+<<plotMod1Sims, echo=FALSE, fig.cap=paste0("Simulations using parameter sets that were generated by running source code provided by Lee et al (2020).
+Compare to Figure S7 of \\cite{lee20sup}.
+The upper bound for the likelihood of this model is ", round(mod1_lee_ll), "."), cache=TRUE, fig.width=6.5, fig.height=3.75>>=
 
 quants <- all_sims %>%
-  filter(!param_set %in% bad_epi_pars) %>%
+  filter(!param_set %in% c(1, 2)) %>%  # Remove parameter sets inconsistent with Lee et al (2020).
   group_by(week) %>%
@@
 
-end_complete <- filter(end_results, logLik > -20000)
-bad_end_pars <- which(end_results$logLik < -20000)
-
 quants_end <- end_sims %>%
-  filter(!param_set %in% bad_end_pars) %>%
   group_by(week) %>%
@@
 @
+
+To calculate an upper-bound on the likelihood of this model---which uses a large number of parameters that each have a different likelihood value---we estimate the likelihood corresponding to each parameter set using a particle filter, and report the maximum of these likelihoods.
 

### 2022-10-18 ms.Rnw
Minor updates, code cleaning

@@
 \label{eq:model3:IS}
-\mu_{I_{u\vaccCounter}S_{u0}} &=& \muDeath + \choleraDeath,
-\\[-3pt]
-\label{eq:model3:AS}
-\mu_{A_{u\vaccCounter}S_{u0}} &=& \muDeath,
+\mu_{I_{u\vaccCounter}S_{u0}} &=& \muDeath + \choleraDeath, \hspace{2mm} \mu_{A_{u\vaccCounter}S_{u0}} = \muDeath
 \\[-3pt]
@@
 Demographic stochasticity is accounted for by modeling non-cholera related death rate $\muDeath$ in each compartment, along with an additional death rate $\choleraDeath$ in \myeqref{eq:model3:IS} to account for cholera induced deaths among infected individuals.
-All deaths are balanced by births into the susceptible compartment in \myeqref{eq:model3:AS} and \myeqref{eq:model3:RS}, thereby maintaining constant population in each department.
+All deaths are balanced by births into the susceptible compartment in \myeqref{eq:model3:IS} and \myeqref{eq:model3:RS}, thereby maintaining constant population in each department.
 
@@
   "mu_B", "XthetaA", "thetaI", "lambdaR", "r", "std_W",
-  "epsilon", "k", "sigma"
-)
+  "epsilon", "k", "sigma", "rho", "gamma"
+)
+
+mod3_n_params <- 10 * length(unit_specific_names) + length(shared_param_names)
+
 est_param_names <- c(
@@
 
-setrw1 <- function(x) {
-  if (grepl('^r[[:digit:]]+$', x)) {
+set_rw_1 <- function(x) {
+  if (grepl('^foi_add[[:digit:]]+$', x) || grepl('^betaB[[:digit:]]+$', x)) {
     return(0.01 / 2)
   } else if (grepl('^Binit[[:digit:]]+$', x)) {
-    return(expression(ivp(0.02)))
+    return(expression(ivp(0.025)))
   } else {
@@
 
-setrw2 <- function(x) {
-  if (grepl('^r[[:digit:]]+$', x)) {
-    return(0.0025 / 2)
+set_rw_2 <- function(x) {
+  if (grepl('^foi_add[[:digit:]]+$', x) || grepl('^betaB[[:digit:]]+$', x)) {
+    return(0.003 / 2)
   } else if (grepl('^Binit[[:digit:]]+$', x)) {
-    return(expression(ivp(0.005)))
+    return(expression(ivp(0.0075)))
   } else {
-    return(0.0025)
-  }
-}
-
-setrw3 <- function(x) {
-  if (grepl('^r[[:digit:]]+$', x)) {
+    return(0.003)
+  }
+}
+
+set_rw_3 <- function(x) {
+  if (grepl('^foi_add[[:digit:]]+$', x) || grepl('^betaB[[:digit:]]+$', x)) {
     return(0.001 / 2)
@@
 # First Search RW
-reg_rw_1.sd <- lapply(est_param_names_expanded, setrw1)
+reg_rw_1.sd <- lapply(est_param_names_expanded, set_rw_1)
 names(reg_rw_1.sd) <- est_param_names_expanded
@@
 # Second Search RW
-reg_rw_2.sd <- lapply(est_param_names_expanded, setrw2)
+reg_rw_2.sd <- lapply(est_param_names_expanded, set_rw_2)
 names(reg_rw_2.sd) <- est_param_names_expanded
@@
 # Third Search RW
-reg_rw_3.sd <- lapply(est_param_names_expanded, setrw3)
+reg_rw_3.sd <- lapply(est_param_names_expanded, set_rw_3)
 names(reg_rw_3.sd) <- est_param_names_expanded
@@
     NP = 50,
-    SPAT_REGRESSION = 0.5,
+    SPAT_REGRESSION = 0.05,
     NREPS = 6,
     NP_EVAL = 100,
-    NREPS_EVAL = 6,
+    NREPS_EVAL = 5,
     RW_SD = chol_rw_1.sd,
@@
     NP = 50,
-    SPAT_REGRESSION = 0.5,
+    SPAT_REGRESSION = 0.05,
     NREPS = 3,
     NP_EVAL = 100,
-    NREPS_EVAL = 6,
+    NREPS_EVAL = 5,
     RW_SD = chol_rw_2.sd,
@@
     NP = 100,
-    SPAT_REGRESSION = 0.5,
+    SPAT_REGRESSION = 0.05,
     NREPS = 3,
@@
   SEARCH1 = list(
-    NBPF = 50,
-    NP = 1000,
-    SPAT_REGRESSION = 0.5,
+    NBPF = 20,
+    NP = 500,
+    SPAT_REGRESSION = 0.05,
     NREPS = 36,
     NP_EVAL = 1000,
-    NREPS_EVAL = 36,
+    NREPS_EVAL = 12,
     RW_SD = chol_rw_1.sd,
-    COOLING = 0.36,
+    COOLING = 0.5,
     KEEP_TRACES = TRUE
@@
     TOP_N = 6,
-    NBPF = 50,
-    NP = 1000,
-    SPAT_REGRESSION = 0.5,
+    NBPF = 20,
+    NP = 500,
+    SPAT_REGRESSION = 0.05,
     NREPS = 6,
     NP_EVAL = 1000,
-    NREPS_EVAL = 36,
+    NREPS_EVAL = 12,
     RW_SD = chol_rw_2.sd,
@@
     TOP_N = 6,
-    NBPF = 50,
-    NP = 1000,
-    SPAT_REGRESSION = 0.5,
+    NBPF = 20,
+    NP = 500,
+    SPAT_REGRESSION = 0.05,
     NREPS = 6,
     NP_EVAL = 1000,
-    NREPS_EVAL = 36,
+    NREPS_EVAL = 12,
     RW_SD = chol_rw_3.sd,
@@
 
-  N_SEARCHES <- 3L
+  N_SEARCHES <- 1L
 } else if (RUN_LEVEL == 3) {
@@
     NBPF = 100,
-    NP = 10000,
-    SPAT_REGRESSION = 0.5,
-    NREPS = 72,
-    NP_EVAL = 10000,
-    NREPS_EVAL = 36,
+    NP = 5000,
+    SPAT_REGRESSION = 0.05,
+    NREPS = 144,
+    NP_EVAL = 5000,
+    NREPS_EVAL = 18,
     RW_SD = chol_rw_1.sd,
@@
   SEARCH2 = list(
-    TOP_N = 9,
+    TOP_N = 12,
     NBPF = 100,
-    NP = 10000,
-    SPAT_REGRESSION = 0.5,
-    NREPS = 4,
-    NP_EVAL = 10000,
-    NREPS_EVAL = 36,
+    NP = 5000,
+    SPAT_REGRESSION = 0.05,
+    NREPS = 6,
+    NP_EVAL = 5000,
+    NREPS_EVAL = 18,
     RW_SD = chol_rw_2.sd,
@@
   SEARCH3 = list(
-    TOP_N = 6,
+    TOP_N = 9,
     NBPF = 100,
-    NP = 10000,
-    SPAT_REGRESSION = 0.5,
-    NREPS = 6,
-    NP_EVAL = 10000,
-    NREPS_EVAL = 36,
+    NP = 5000,
+    SPAT_REGRESSION = 0.05,
+    NREPS = 8,
+    NP_EVAL = 5000,
+    NREPS_EVAL = 18,
     RW_SD = chol_rw_3.sd,
@@
   N_SEARCHES <- 3L
+} else {
+  stop("Run level must be 1, 2 or 3.")
 }
@@
         search3 = SEARCH3,
-        ncores = cores
+        ncores = cores,
+        search_rho = TRUE,
+        search_gamma = TRUE
       )
@@
 rm(
-  SEARCH1, SEARCH2, SEARCH3, N_SEARCHES, unit_specific_names,
-  shared_param_names, est_param_names, est_param_names_expanded,
-  setrw1, setrw2, setrw3, reg_rw_1.sd, reg_rw_2.sd, reg_rw_3.sd,
+  SEARCH1, SEARCH2, SEARCH3, N_SEARCHES,
+  est_param_names, est_param_names_expanded,
+  set_rw_1, set_rw_2, set_rw_3, reg_rw_1.sd, reg_rw_2.sd, reg_rw_3.sd,
   chol_rw_1.sd, chol_rw_2.sd, chol_rw_3.sd
@@
 
-p3_specific <- c("foi_add", "betaB", "Binit")
-p3_shared <- c(
-  "mu_B", "XthetaA", "thetaI", "lambdaR", "r", "std_W",
-  "epsilon", "k", "sigma"
-)
-
-p3u <- matrix(nrow = length(p3_specific), ncol = 10)
-rownames(p3u) <- p3_specific
-p3s <- numeric(length(p3_shared))
-names(p3s) <- p3_shared
-
-for (p in p3_shared) {
+p3u <- matrix(nrow = length(unit_specific_names), ncol = 10)
+rownames(p3u) <- unit_specific_names
+p3s <- numeric(length(shared_param_names))
+names(p3s) <- shared_param_names
+
+for (p in shared_param_names) {
   pat <- paste0("^", p, "[[:digit:]]+$")
@@
 
-for (p in p3_specific) {
+for (p in unit_specific_names) {
   pat <- paste0("^", p, "[[:digit:]]+$")
@@
 
-rm(p3_specific, p3_shared, p, pat)
+rm(unit_specific_names, shared_param_names, p, pat)
 
@@
 
-h3_spat_nsim <- switch(RUN_LEVEL, 20, 100, 500)
+h3_nsim <- switch(RUN_LEVEL, 20, 100, 500)
 
 # Check if run_level_3 results exit, and use them if they do.
-if (RUN_LEVEL != 3 & file.exists("model3/run_level_3/SpatPompSims.rds")) {
-  h3Spat_quants <- readRDS("model3/run_level_3/SpatPompSims.rds")
+if (RUN_LEVEL != 3 & file.exists("model3/run_level_3/h3Sims.rds")) {
+  h3_quants <- readRDS("model3/run_level_3/h3Sims.rds")
 } else {
-  h3Spat_quants <- bake(
-    file = paste0('model3/', rl_dir, 'SpatPompSims.rds'), {
+  h3_quants <- bake(
+    file = paste0('model3/', rl_dir, 'h3Sims.rds'), {
 
       # Simulate from the fitted version of Model 3
-      h3Spat_sims <- simulate(
-        h3Spat, nsim = h3_spat_nsim, format = 'data.frame'
+      h3_sims <- simulate(
+        h3Spat, nsim = h3_nsim, format = 'data.frame'
       )
@@
       # simulations
-      h3Spat_sims %>%
+      h3_sims %>%
         rename(dep = unitname) %>%
@@
         ) %>%
-        # filter(
-        #   date >= as.Date(lubridate::round_date(lubridate::date_decimal(h3Spat@t0), unit = 'day'))
-        #   ) %>%
         mutate(date = as.Date(lubridate::round_date(date, unit = 'day')))
@@
 ggplot() +
-  geom_line(data = h3Spat_quants, aes(x = date, y = q50 + 1), col = 'blue') +
+  geom_line(data = h3_quants, aes(x = date, y = q50 + 1), col = 'blue') +
   geom_line(data = dep_plot_df, aes(x = date, y = cases + 1), col = 'black') +
   geom_ribbon(  # plot confidence region.
-    data = h3Spat_quants,
+    data = h3_quants,
     aes(x = date, ymin = q05 + 1, ymax = q95 + 1),
@@
 <<Remove Model3 spatSims, cache=FALSE, include=FALSE, message=FALSE, echo=FALSE>>=
-rm(h3Spat_quants, h3_spat_nsim)
+rm(h3_quants, h3_nsim)
 gc()
@@
 
-# TODO: Not needed after fitting Model 3.
 NREPS_EVAL <- switch(RUN_LEVEL,   3,   8,    36)
-NP_EVAL    <- switch(RUN_LEVEL, 100, 500, 10000)
+NP_EVAL    <- switch(RUN_LEVEL, 100, 500,  5000)
 
@@
 # model 2 AIC
-mod2_n_params <- 16  # TODO: adjust Number of parameters fit in Model 2
+mod2_n_params <- h2_fit$n_fit_params
 mod2_aic <- 2 * mod2_n_params - 2 * mod2_ll
@@
 # model 3 AIC
-mod3_n_params <- 29
 mod3_aic <- 2 * mod3_n_params - 2 * mod3_ll
@@
 
-# TODO: The code below will not be needed once we update the fitting of model 3.
 # Calculate likelihood via block-particle filter
-if (RUN_LEVEL != 3 & file.exists("model3/run_level_3/SpatPomp_res.rds")) {
-  mod3_spatPomp_res <- readRDS("model3/run_level_3/SpatPomp_res.rds")
+if (RUN_LEVEL != 3 & file.exists("model3/run_level_3/mod3_eval_res.rds")) {
+  mod3_eval_res <- readRDS("model3/run_level_3/mod3_eval_res.rds")
 } else{
-  mod3_spatPomp_res <- bake(
-    file = paste0('model3/', rl_dir, 'SpatPomp_res.rds'), {
+  mod3_eval_res <- bake(
+    file = paste0('model3/', rl_dir, 'mod3_eval_res.rds'), {
 
@@
       h3_bpf <- foreach(i = 1:NREPS_EVAL, .combine = c) %dopar% {
-        bpfilter(h3Spat, Np = NP_EVAL, block_size = 1,
-                 params = haitipkg:::panelParms_toSpatParms3(p3))
+        bpfilter(h3Spat, Np = NP_EVAL, block_size = 1)
       }
@@
 
-mod3_spatPomp_n_params <- 29
-mod3_spatPomp_aic <- 2 * mod3_spatPomp_n_params - 2 * mod3_spatPomp_res$ll
-mod3_spatPomp_res$subset_aic <- 2 * mod3_spatPomp_n_params - 2 * mod3_spatPomp_res$subset_ll
+mod3_aic <- 2 * mod3_n_params - 2 * mod3_eval_res$ll
+mod3_eval_res$subset_aic <- 2 * mod3_n_params - 2 * mod3_eval_res$subset_ll
 
@@
   $\Sexpr{myround(mod2_ll, 1)}$ &
-  $\Sexpr{myround(mod3_spatPomp_res$ll, 1)}$ \\
+  $\Sexpr{myround(mod3_eval_res$ll, 1)}$ \\
     & ($\Sexpr{myround(mod1_lee_ll, 1)}$)\footnotemark[1] &
@@
   $\Sexpr{as.character(mod2_n_params)}$ &
-  $\Sexpr{mod3_spatPomp_n_params}$ \\
+  $\Sexpr{as.character(mod3_n_params)}$ \\
  Fit Parameters &
@@
   $\Sexpr{myround(mod2_aic, 1)}$ &
-  $\Sexpr{myround(mod3_spatPomp_aic, 1)}$ \\
+  $\Sexpr{myround(mod3_aic, 1)}$ \\
   & ($\Sexpr{myround(mod1_lee_aic, 1)}$)\footnotemark[1] &
@@
 On this subset, their model has a likelihood of $\Sexpr{myround(max(lee3_results$sub_ll_evals), 1)}$.
-On this same subset, our model has a likelihood of $\Sexpr{myround(mod3_spatPomp_res$subset_ll, 1)}$.
+On this same subset, our model has a likelihood of $\Sexpr{myround(mod3_eval_res$subset_ll, 1)}$.
 See Sec.~S5 of the supplement material for more details on estimating the likelihood of the \cite{lee20} models.
@@
   & \eqref{model2:mu_IR}
-  & {\fixed $\muIR^{-1}=\Sexpr{myround(365/p3["gamma"],1)}$ }
+  & {\fixed $\muIR^{-1}=\Sexpr{myround(365/p3s["gamma"],1)}$ }
   & \eqref{eq:model3:IR}
@@
   &
-  & {\fixed $\muDeath^{-1} = \Sexpr{myround(1/p3s["mu"],1)}$}
+  & {\fixed $\muDeath^{-1} = \Sexpr{myround(1/p3["mu1"],1)}$}
   & \eqref{eq:model3:IS}

### 2022-10-19 ms.Rnw
Minor changes

@@
 
-mod3_n_params <- 10 * length(unit_specific_names) + length(shared_param_names)
+mod3_n_params <- length(unit_specific_names) * 10 + length(shared_param_names)
 
@@
   SEARCH1 = list(
-    NBPF = 20,
+    NBPF = 25,
     NP = 500,
@@
     COOLING = 0.5,
-    KEEP_TRACES = TRUE
+    KEEP_TRACES = TRUE,
+    KEEP_LIK_MAT = TRUE
   )
@@
     TOP_N = 6,
-    NBPF = 20,
+    NBPF = 25,
     NP = 500,
@@
         search_rho = TRUE,
-        search_gamma = TRUE
+        search_gamma = TRUE,
+        search_Binit = FALSE
       )
@@
 <<Table 2 Input, echo=FALSE, eval=TRUE>>=
-
 NREPS_EVAL <- switch(RUN_LEVEL,   3,   8,    36)
@@
 mod3_aic <- 2 * mod3_n_params - 2 * mod3_ll
-
-# h3Spat <- haiti3_spatPomp()
-# coef(h3Spat) <- haitipkg:::panelParms_toSpatParms3(p3)
 

### 2022-10-19 ms.Rnw
minor change, ready to compile from GL

@@
 # Create vectors for the unit and shared parameters
-unit_specific_names <- c("betaB", "foi_add", "Binit")
+unit_specific_names <- c("betaB", "foi_add")
 shared_param_names <- c(

### 2022-10-19 ms.Rnw
Don't save traces

@@
     COOLING = 0.5,
-    KEEP_TRACES = TRUE
+    KEEP_TRACES = FALSE
   )
@@
     COOLING = 0.5,
-    KEEP_TRACES = TRUE,
+    KEEP_TRACES = FALSE,
     KEEP_LIK_MAT = TRUE

### 2022-10-25 ms.Rnw
Made setting model parameters more explicit

@@
 
+      coef(h3Spat) <- p3
       # Simulate from the fitted version of Model 3
@@
 
+      coef(h3Spat) <- p3
+
       # Perform block-particle filter
@@
 
-mod3_aic <- 2 * mod3_n_params - 2 * mod3_eval_res$ll
 mod3_eval_res$subset_aic <- 2 * mod3_n_params - 2 * mod3_eval_res$subset_ll
@@
   $\Sexpr{myround(mod2_ll, 1)}$ &
-  $\Sexpr{myround(mod3_eval_res$ll, 1)}$ \\
+  $\Sexpr{myround(mod3_ll, 1)}$ \\
     & ($\Sexpr{myround(mod1_lee_ll, 1)}$)\footnotemark[1] &
@@
     expr = {
+      coef(h3Spat) <- p3
       h3_bpf <- bpfilter(h3Spat, Np = NP_BPF, block_size = 1, save_states = TRUE)
@@
       h3_S0_par <- get_model3_vacc_scenario_params(scenario = "noVacc")
+      coef(h3Spat) <- p3
       coef(h3Spat)[names(h3_S0_par)] <- h3_S0_par
@@
       h3_V1_par <- get_model3_vacc_scenario_params(scenario = "2dep")
+      coef(h3Spat) <- p3
       coef(h3Spat)[names(h3_V1_par)] <- h3_V1_par
@@
       h3_V2_par <- get_model3_vacc_scenario_params(scenario = "3dep")
+      coef(h3Spat) <- p3
       coef(h3Spat)[names(h3_V2_par)] <- h3_V2_par
@@
       h3_V3_par <- get_model3_vacc_scenario_params(scenario = "slowNation")
+      coef(h3Spat) <- p3
       coef(h3Spat)[names(h3_V3_par)] <- h3_V3_par
@@
       h3_V4_par <- get_model3_vacc_scenario_params(scenario = "fastNation")
+      coef(h3Spat) <- p3
       coef(h3Spat)[names(h3_V4_par)] <- h3_V4_par

### 2022-11-21 ms.Rnw
Improved replication of model 3

@@
 # Create Lee et al (2020a) version of model 3 on sub-data
-lee3_sub <- haiti3_lee_et_al()
+lee3_sub <- lee3_spatPomp()
 
@@
 # parameter estimation phase.
-set_as_na <- lee3_sub@times < lubridate::decimal_date(as.Date('2017-06-10'))
-temp_ouest_cases <- lee3_sub@data['casesOuest', ]
-temp_ouest_cases[set_as_na] <- NA_real_
-lee3_sub@data['casesOuest', ] <- temp_ouest_cases
-
-lee3_all <- haiti3_lee_et_al(start_time = "2010-10-23")
+# set_as_na <- lee3_sub@times < lubridate::decimal_date(as.Date('2017-06-10'))
+# temp_ouest_cases <- lee3_sub@data['casesOuest', ]
+# temp_ouest_cases[set_as_na] <- NA_real_
+# lee3_sub@data['casesOuest', ] <- temp_ouest_cases
+
+lee3_all <- lee3_spatPomp(start_date = "2010-10-23")
 
@@
       # Save how long it takes for sub-data
-      lee3_results[["sub_pf_time"]] <- system.time({
+      lee3_results[["sub_bpf_time"]] <- system.time({
+
+        lee3_sub_ll_evals <- foreach(
+          i = 1:lee3_NREP, .combine = rbind
+        ) %dopar% {
+
+          liks <- numeric(10L)
+
+          lee_bpf_out <- bpfilter(lee3_sub, block_size = 1, Np = lee3_NP)
+          Ouest_ind <- which(lee_bpf_out@unit_names == "Ouest")
+
+          for (u in 1:10) {
+            if (u == Ouest_ind) {
+              liks[u] <- sum(lee_bpf_out@block.cond.loglik[u, lee3_sub@times >= lubridate::decimal_date(as.Date('2017-06-10'))])
+            } else {
+              liks[u] <- sum(lee_bpf_out@block.cond.loglik[u, ])
+            }
+          }
+
+          liks
+        }
+      })
 
       # Save all likelihood evaluations
-        lee3_results[["sub_ll_evals"]] <- foreach(
+      lee3_results[["sub_ll_eval"]] <- logmeanexp(rowSums(lee3_sub_ll_evals)[!is.na(rowSums(lee3_sub_ll_evals))])
+
+      registerDoRNG(2789131)
+
+      # Save how long it takes for all data
+      lee3_results[["all_pf_time"]] <- system.time({
+        # Save all likelihood evaluations
+
+        lee3_all_ll_evals <- foreach(
           i = 1:lee3_NREP, .combine = c
         ) %dopar% {
-          logLik(pfilter(lee3_sub, Np = lee3_NP))  # Get likelihood evaluations
-        }
+          logLik(bpfilter(lee3_all, block_size = 1, Np = lee3_NP))  # Get likelihood evaluations
+        }
+
+        lee3_results[["all_ll_eval"]] <- logmeanexp(lee3_all_ll_evals[!is.na(lee3_all_ll_evals)])
+
       })
 
-      registerDoRNG(2789131)
-
-      # Save how long it takes for all data
-      lee3_results[["all_pf_time"]] <- system.time({
-
-        # Save all likelihood evaluations
-        lee3_results[["all_ll_evals"]] <- foreach(
-          i = 1:lee3_NREP, .combine = c
-        ) %dopar% {
-          logLik(pfilter(lee3_all, Np = lee3_NP))  # Get likelihood evaluations
-        }
-      })
-
-      lee3_results[["n_params"]] <- 29
+      lee3_results[["n_params"]] <- 29  # TODO: Check to make sure this is the correct number.
 
@@
 
-mod3_lee_coupled_aic <- 2 * lee3_results$n_params - 2 * max(lee3_results$all_ll_evals)
-
-mod3_lee_coupled_partial_aic <- 2 * lee3_results$n_params - 2 * max(lee3_results$sub_ll_evals)
+mod3_lee_coupled_aic <- 2 * lee3_results$n_params - 2 * lee3_results$all_ll_eval
+
+mod3_lee_coupled_partial_aic <- 2 * lee3_results$n_params - 2 * lee3_results$sub_ll_eval
 
@@
   ($\Sexpr{myround(h2_fit$leeFit_ll, 1)}$) &
-  ($\Sexpr{myround(max(lee3_results[["all_ll_evals"]]), 1)}$)\footnotemark[2]
+  ($\Sexpr{myround(lee3_results[["all_ll_eval"]], 1)}$)\footnotemark[2]
 \\
@@
 $\textsuperscript{2}$\cite{lee20} fit Model~3 to a subset of the data (March 2014 onward, excluding data from Ouest in 2015-2016).
-On this subset, their model has a likelihood of $\Sexpr{myround(max(lee3_results$sub_ll_evals), 1)}$.
+On this subset, their model has a likelihood of $\Sexpr{myround(lee3_results$sub_ll_eval, 1)}$.
 On this same subset, our model has a likelihood of $\Sexpr{myround(mod3_eval_res$subset_ll, 1)}$.

### 2022-11-21 si/si.Rnw
Improved replication of model 3

@@
 <<Model3Replication, echo=FALSE, cache=TRUE>>=
-NSIM <- switch(run_level, 50, 100, 1000)
+NSIM <- switch(run_level, 50, 200, 500)
 
@@
 
-lee3 <- haiti3_lee_et_al()
+lee3 <- lee3_spatPomp()
 lee3_sims <- simulate(lee3, nsim = NSIM, seed = 321, format = 'data.frame')
@@
 quants <- lee3_sims %>%
-  dplyr::select(-CasesAll) %>%
-  tidyr::pivot_longer(
-    data = .,
-    cols = starts_with("cases"),
-    names_prefix = "cases",
-    names_to = "dep",
-    values_to = "cases"
-  ) %>%
-  dplyr::group_by(dep, time) %>%
+  dplyr::group_by(unitname, time) %>%
   dplyr::summarise(
@@
     date = as.Date(lubridate::round_date(date))
-  )
+  ) %>%
+  dplyr::rename(dep = unitname)
 

### 2022-11-21 ms.Rnw
minor edits; more major edits to follow

@@
 Model~4 has an agent-based construction, featuring considerable mechanistic detail but limited ability to calibrate these details to data.
-The strengths and weaknesses of the agent-based modeling approach  \citep{tracy18} are outside the scope of this article, and we focus on Models~1--3.
+We focus on Models~1--3, as the strengths and weaknesses of the agent-based modeling approach  \citep{tracy18} are outside the scope of this article, and we focus on Models~1--3.
 
@@
 The groups largely adhered to existing guidelines on creating models to inform policy \citep{behrend20,saltelli20} and, despite their autonomy, obtained a consensus that an extensive nationwide vaccination campaign would be necessary to eliminate cholera from Haiti.
-This conclusion is inconsistent with the fact that there have been no confirmed cases since February, 2019 \citep{trevisin22} without the implementation of a national vaccination program.
-
-The discrepancy between the model based conclusions of \cite{lee20} and the subsequent elimination of cholera cholera has been debated \citep{francois20,rebaudetComment20,henrys20,leeReply20}.
+Although several cholera chases have recently been detected in Haiti (\citep{TODO}),
+their conclusion is inconsistent with the prolonged period with no confirmed cholera cases between February, 2019 \citep{trevisin22} and \jwc{TODO}.
+
+The discrepancy between the model based conclusions of \cite{lee20} and the prolonged absence of cholera in Haiti has been debated \citep{francois20,rebaudetComment20,henrys20,leeReply20}.
 Suggested origins of this discrepancy include the use of unrealistic models \citep{rebaudetComment20} and unrealistic criteria for cholera elimination \citep{henrys20}.
@@
 Elements of $\paramVec$ can be fixed at a constant value based on scientific understanding of the system, but parameters can also be calibrated to data by maximizing a measure of congruency between the observed data and the assumed mechanistic structure.
-Calibrating model parameters does not guarantee that the resulting model successfully approximates real-world mechanisms, since model assumptions may be incorrect, and do not change as the model is calibrated to data.
-However, the congruency between the model and observed data serves as a proxy for the congruency between the model and the true underlying dynamic system, since the observed data are a result of the system.
-In the following subsections we describe our approach to fitting the three proposed mechanistic models by calibrating $\paramVec$ to observed cholera incidence data.
+Calibrating model parameters to observed data does not guarantee that the resulting model successfully approximates real-world mechanisms, since model assumptions may be incorrect, and do not change as the model is calibrated to data.
+However, the congruency between the model and observed data serves as a proxy for the congruency between the model and the true underlying dynamic system, since the observed data are a result of the generating system.
+As such, the best possible fit of the proposed mechanistic structure to the observed data is desirable.
+In the following subsections we describe our approach to calibrating the three proposed mechanistic models to observed cholera incidence data.
 
@@
 Model~1 describes cholera dynamics at the nationally aggregated scale so that $\bm{Y}^{(1)}_n \in \R$ for each observation time $n \in 1:N$, and the latent state vector $\bm{X}^{(1)}(t)$ is comprised of national level population counts.
-Several algorithms exist, both frequentist and Bayesian, that can be used to obtain estimates of the parameters for the class of models.
+Several algorithms exist, both frequentist and Bayesian, that can be used to obtain estimates of the parameters for this class of models.
 In order to retain the ability to propose models that are scientifically meaningful rather than only those that are simply statistically convenient, we restrict ourselves to parameter estimation techniques that have the plug-and-play property, which is that the fitting procedure only requires the ability to simulate the latent process instead of evaluating transition densities \citep{breto09,he10}.
 Plug-and-play algorithms include Bayesian approaches like ABC and PMCMC \citep{toni09,andrieu10}, but here we use frequentist methods to maximize model likelihoods.
-To our knowledge, the only plug-and-play frequentist methods that can maximize the likelihood for POMP models of this complexity are iterated filtering algorithms, which modify the well-known particle filter \citep{arulampalam02} by performing a random walk for each parameter and particle.
+To our knowledge, the only plug-and-play, frequentist methods that can maximize the likelihood for POMP models of this complexity are iterated filtering algorithms \citep{ionides15}, which modify the well-known particle filter \citep{arulampalam02} by performing a random walk for each parameter and particle.
 These perturbations are carried out iteratively over multiple filtering operations, using the collection of parameters from the previous filtering pass as the parameter initialization for the next iteration, and decreasing the random walk variance at each step.
@@
 
-As previously stated, Model~2 is a deterministic compartmental model defined by a set of coupled differential equations.
+Model~2 is a deterministic compartmental model defined by a set of coupled differential equations.
 The use of deterministic compartment models have a long history in the field of infectious disease epidemiology \citep{kermack1927,brauer2017,giordano20},
 and can be justified by asymptotic considerations in a large-population limit \citep{dadlani2020,ndii17}.
-Because the process model of Model~2 is deterministic, the parameter estimation problem for Model~2 reduces to a least squares calculation when combined with a Gaussian measurement model (see supplement material for details).
+Because the process model of Model~2 is deterministic, the parameter estimation problem for Model~2 reduces to a least squares calculation when combined with a Gaussian measurement model (see supplement material for details) \jwc{TODO: cite which part of supplement}.
 \cite{lee20} fit two versions of model~2 based on a presupposed change in cholera transmission from a epidemic phase to endemic phase that occurred in March, 2014.
@@
 Both the latent and observable processes in Model~3 can be be factored into department specific processes which interact with each other.
-The decision to address metapopulation dynamics via a spatially explicit model, rather than to aggregate over space, is double-edged.
+The decision to address metapopulation dynamics using a spatially explicit model, rather than to aggregate over space, is double-edged.
 Evidence for the former approach has been provided in previous studies \citep{king15}, including the specific case of heterogeneity between Haitian departments in cholera transmission \citep{collins14}.
 However, a legitimate preference for simplicity \citep{saltelli20,green15} can support a decision to consider nationally aggregated models.
-
-Fitting scientifically flexible metapopulation models is a challenging statistical problem.
-In particular, parameter estimation techniques based on the particle filter become computationally intractable as the number of spatial units increase.
+One issue that arises when fitting spatially explicit models is that parameter estimation techniques based on the particle filter become computationally intractable as the number of spatial units increase.
 This is a result of the approximation error of particle filters growing exponentially in the dimension of the model \citep{rebeschini15,park20}.
@@
 
-The simplified, spatially-decoupled version of Model~3 relies on the observed cholera cases for each department to describe cholera dynamics, making it impossible to obtain forecasts for future time points from the fitted model.
-In order to obtain model forecasts, the parameters fit for the conditionally independent department level models were used in the fully coupled version of Model~3.
+The simplified, spatially-decoupled version of Model~3 implemented by \cite{lee20} relies on the observed cholera cases for each department to describe cholera dynamics, making it impossible to obtain forecasts for future time points from the fitted model.
+Therefore, in order to obtain model forecasts, the parameters of the calibrated conditionally independent department level models were used in the fully coupled version of Model~3.
 This approach of model calibration and forecasting avoids the issue of particle depletion by simplifying the parameter estimation problem, but this advantage raises new issues.
@@
 
-Goodness-of-fit alone does not guarantee that a model provides a correct causal interpretation of the model.
-Indeed, associative models are not constrained to have a causal interpretation, and typically are designed with the sole goal of providing a statistical fit to data.
+Associative models are not constrained to have a causal interpretation, and typically are designed with the sole goal of providing a statistical fit to data.
 Consequently, we should not require a candidate mechanistic model to beat all benchmarks.
@@
 
-It should be universal practice to present measures of goodness of fit for published models, and mechanistic models should be compared against benchmarks.
+\jwc{TODO: CHECK} It should be universal practice to present measures of goodness of fit for published models, and mechanistic models should be compared against benchmarks.
 In our literature review of the Haiti cholera epidemic, no benchmark models were considered in any of the 24 papers which calibrated a mechanistic model to data in order to obtain scientific conclusions.
@@
 lee3_all <- lee3_spatPomp(start_date = "2010-10-23")
-
-rm(set_as_na, temp_ouest_cases)
-gc()
 
@@
 
-Similar to comparing log-likelihoods across models, an additional powerful diagnosis tool is the comparison of conditional log-likelihoods.
-Conditional likelihoods, defined as the density $f_{Y_k | Y_1, \ldots, Y_{k - 1}}\big(Y_k = y_k^* | y_{1:k-1}^*\big)$, provide a basic description of how well the proposed model can describe each data point, given the previous observations.
-Comparing these results across models---including benchmark models---can help researchers identify potential model deficiencies, or errors in the observed data.
-Additional tools for assessing the goodness-of-fit of a model include plotting the effective sample size of each observation \citep{liu01}, and comparing summary statistics of the observed data to simulations from the model \citep{wood10}.
-These summary statistics are sometimes called diagnostic probes \citep{king15,king16}.
+Table~\ref{tab:likes} shows that despite the significant improvements that were made by using a contemporary approach of parameter calibration for Model~3, this model still fails to beat a log-linear ARMA benchmark.
+This suggests that Model~3 is failing to adequately explain some substantial aspect of the data.
+As our goal for this article is the estimation and diagnosis of the models at hand, we explore the shortcoming in log-likelihood further in section \jwc{TODO} of the supplement material.
+There, we use additional tools like conditional log-likelihoods to understand how well the model describes each data point, given the previous observations.
+Additional tools for assessing the goodness-of-fit of a model include plotting the effective sample size of each observation \citep{liu01}, and comparing summary statistics of the observed data to simulations from the model \citep{wood10}; these summary statistics are sometimes called diagnostic probes \citep{king15,king16}.
 
@@
 If a Bayesian technique is used for parameter estimation, a natural way to account for parameter uncertainty is to obtain simulations from the model where each simulation is obtained using parameters drawn from the estimated posterior distribution.
-For frequentist inference, an empirical Bayes approach leads to similar methodology \citep{king15}; details are provided in the supplement.
+For frequentist inference, one possible approach obtaining model forecasts from various values of $\theta$, and then weighing the resulting forecasts by their corresponding log-likelihoods \citep{king15}; see section \jwc{TODO} of the supplement material for more details.
 Both of these approaches share the similarity that parameters are chosen for the forecast approximately in proportion to their corresponding value of the likelihood function, $f^{(m)}_{\bm{Y}_{1:N}}(\bm{y}_{1:N}^*; \theta)$.
@@
 The ongoing global COVID-19 pandemic has provided a clear example on how government policy may be affected by the conclusions of scientific models \citep{saltelli20}.
-This article demonstrates that fitting appropriate scientific models remains a challenging statistical task, and therefore great care is needed when fitting scientific models for policy recommendations.
+This article demonstrates that fitting appropriate scientific models remains a challenging statistical task, and therefore great care is needed when fitting scientific models with the purpose of making policy recommendations.
 We provided a few suggestions that may aid the fitting of mechanistic models such as comparing model likelihoods to a benchmark.
@@
 
-In this article we argue that careful attention to important statistical details could have led the models proposed by \cite{lee20} to correctly predict the imminent cholera elimination.
+In this article we argue that careful attention to important statistical details could have led the models proposed by \cite{lee20} to correctly predict the initial elimination of cholera from Haiti in 2019.
 We acknowledge the benefit of hindsight: our demonstration of a statistically principled route to obtain better-fitting models with more accurate predictions does not rule out the possibility of discovering other models that fit well yet predict poorly.
 We used the same data and models, and even much of the same code, as \citet{lee20}, and yet ended up with drastically different conclusions.
-At a minimum, we have shown that the conclusions are sensitive to details in how the data analysis is carried out, and that attention to statistical fit (including numerical issues such as likelihood maximization) can lead to improved policy guidance.
+At a minimum, we have shown that the conclusions are sensitive to details in how the data analysis is carried out, and that attention to statistical fit (including numerical consideration such as likelihood maximization) can lead to improved policy guidance.
 
@@
 \sdescription{
-This document contains additional details for Models~1--3, as well as a translation table that facilitates comparisons between these models and those described by \cite{lee20}.
-The supplement also demonstrates our capability to faithfully replicate the results
-of \cite{lee20}.
+This document contains additional details about the numerical implementations of Models~1--3, as well as a translation table that facilitates comparisons between these models and those described by \cite{lee20}.
+The supplement also demonstrates our capability to faithfully estimate the likelihood of the models of \cite{lee20}.
 }

### 2022-11-30 ms.Rnw
minor edits

@@
 library(pomp)
-library(panelPomp)
 library(spatPomp)
@@
 The groups largely adhered to existing guidelines on creating models to inform policy \citep{behrend20,saltelli20} and, despite their autonomy, obtained a consensus that an extensive nationwide vaccination campaign would be necessary to eliminate cholera from Haiti.
-Although several cholera chases have recently been detected in Haiti (\citep{TODO}),
-their conclusion is inconsistent with the prolonged period with no confirmed cholera cases between February, 2019 \citep{trevisin22} and \jwc{TODO}.
+Although several cholera cases have recently been detected in Haiti (\citep{world2010outbreak}),
+their conclusion is inconsistent with the prolonged period with no confirmed cholera cases between February, 2019 \citep{trevisin22} and October, 2022.
 
@@
 
-\jwc{TODO: CHECK} It should be universal practice to present measures of goodness of fit for published models, and mechanistic models should be compared against benchmarks.
-In our literature review of the Haiti cholera epidemic, no benchmark models were considered in any of the 24 papers which calibrated a mechanistic model to data in order to obtain scientific conclusions.
+It should be universal practice to present measures of goodness of fit for published models, and mechanistic models should be compared against benchmarks.
+In our literature review of the Haiti cholera epidemic, no benchmark models were considered in any of the 32 papers that used dynamic models to describe the cholera in order to obtain scientific conclusions.
 Including benchmarks would help authors and readers to detect and confront any major statistical limitations of the proposed mechanistic models.

### 2022-11-30 ms.Rnw
Change model3 model to fit Iinit and hurricane parameters

@@
-%% Template for the submission to:
+%% Based on the template for the submission to:
 %%   The Annals of Applied Statistics [AOAS]
@@
 if (!dir.exists('model2')) dir.create("model2/")
+
+if (RUN_LEVEL == 3) {
+  # Print session info into a txt file. Only update if RUN_LEVEL == 3, which
+  # is when the final results are calculated.
+  cat(
+    utils::capture.output(sessionInfo()),
+    file = "sessionInfo.txt",
+    sep="\n"
+  )
+}
 
@@
 # Create vectors for the unit and shared parameters
-unit_specific_names <- c("betaB", "foi_add")
+unit_specific_names <- c("betaB", "foi_add", "aHur", "hHur", "Iinit")
 shared_param_names <- c(
   "mu_B", "XthetaA", "thetaI", "lambdaR", "r", "std_W",
-  "epsilon", "k", "sigma", "rho", "gamma"
-)
-
-mod3_n_params <- length(unit_specific_names) * 10 + length(shared_param_names)
+  "epsilon", "k"
+)
+
+# betaB + foi_add for all 10 units +
+# aHur3, aHur9, hHur3, hHur9 +
+# Iinit3, Iinit7, Iinit9 +
+# shared_parameters
+mod3_n_params <- 2 * 10 + 7 + length(shared_param_names)
 
@@
 set_rw_1 <- function(x) {
-  if (grepl('^foi_add[[:digit:]]+$', x) || grepl('^betaB[[:digit:]]+$', x)) {
-    return(0.01 / 2)
-  } else if (grepl('^Binit[[:digit:]]+$', x)) {
-    return(expression(ivp(0.025)))
+  if (gsub("[[:digit:]]+$", "", x) %in% shared_param_names) {
+    return(0.01)
+  } else if (x %in% paste0(rep(c("aHur", 'hHur'), each = 2), c(3, 9))) {
+    return(expression(ifelse(time >= 2016.754 & time <= 2017, 0.015, 0)))
+  } else if (x %in% c("Iinit3", "Iinit7", "Iinit10")) {
+    return(expression(ivp(0.25)))
+  } else if (grepl('^foi_add[[:digit:]]+$', x)) {
+    return(0.015)
   } else {
-    return(0.01)
+    return(0)
   }
@@
 set_rw_2 <- function(x) {
-  if (grepl('^foi_add[[:digit:]]+$', x) || grepl('^betaB[[:digit:]]+$', x)) {
-    return(0.003 / 2)
-  } else if (grepl('^Binit[[:digit:]]+$', x)) {
-    return(expression(ivp(0.0075)))
+  if (gsub("[[:digit:]]+$", "", x) %in% shared_param_names) {
+    return(0.003)
+  } else if (x %in% paste0(rep(c("aHur", 'hHur'), each = 2), c(3, 9))) {
+    return(expression(ifelse(time >= 2016.754 & time <= 2017, 0.005, 0)))
+  } else if (x %in% c("Iinit3", "Iinit7", "Iinit10")) {
+    return(expression(ivp(0.125)))
+  } else if (grepl('^foi_add[[:digit:]]+$', x)) {
+    return(0.005)
   } else {
+    return(0)
+  }
+}
+
+set_rw_3 <- function(x) {
+  if (gsub("[[:digit:]]+$", "", x) %in% shared_param_names) {
+    return(0.0025)
+  } else if (x %in% paste0(rep(c("aHur", 'hHur'), each = 2), c(3, 9))) {
+    return(expression(ifelse(time >= 2016.754 & time <= 2017, 0.005, 0)))
+  } else if (x %in% c("Iinit3", "Iinit7", "Iinit10")) {
+    return(expression(ivp(0.125)))
+  } else if (grepl('^foi_add[[:digit:]]+$', x)) {
     return(0.003)
-  }
-}
-
-set_rw_3 <- function(x) {
-  if (grepl('^foi_add[[:digit:]]+$', x) || grepl('^betaB[[:digit:]]+$', x)) {
-    return(0.001 / 2)
-  } else if (grepl('^Binit[[:digit:]]+$', x)) {
-    return(expression(ivp(0.002)))
   } else {
-    return(0.001)
+    return(0)
   }
@@
   SEARCH1 = list(
-    NBPF = 5,
-    NP = 50,
+    NBPF = 15,
+    NP = 500,
     SPAT_REGRESSION = 0.05,
-    NREPS = 6,
-    NP_EVAL = 100,
-    NREPS_EVAL = 5,
+    NREPS = 36,
+    NP_EVAL = 500,
+    NREPS_EVAL = 3,
     RW_SD = chol_rw_1.sd,
@@
   SEARCH2 = list(
-    TOP_N = 2,
-    NBPF = 5,
-    NP = 50,
+    TOP_N = 9,
+    NBPF = 10,
+    NP = 500,
     SPAT_REGRESSION = 0.05,
-    NREPS = 3,
-    NP_EVAL = 100,
-    NREPS_EVAL = 5,
+    NREPS = 4,
+    NP_EVAL = 500,
+    NREPS_EVAL = 3,
     RW_SD = chol_rw_2.sd,
@@
   SEARCH3 = list(
-    TOP_N = 2,
-    NBPF = 5,
-    NP = 100,
+    TOP_N = 6,
+    NBPF = 10,
+    NP = 500,
     SPAT_REGRESSION = 0.05,
-    NREPS = 3,
-    NP_EVAL = 100,
-    NREPS_EVAL = 6,
+    NREPS = 6,
+    NP_EVAL = 500,
+    NREPS_EVAL = 3,
     RW_SD = chol_rw_3.sd,
@@
 
-  N_SEARCHES <- 2L
+  N_SEARCHES <- 3L
 } else if (RUN_LEVEL == 2) {
   SEARCH1 = list(
-    NBPF = 25,
-    NP = 500,
+    NBPF = 100,
+    NP = 1000,
     SPAT_REGRESSION = 0.05,
-    NREPS = 36,
+    NREPS = 72,
     NP_EVAL = 1000,
-    NREPS_EVAL = 12,
+    NREPS_EVAL = 8,
     RW_SD = chol_rw_1.sd,
@@
   SEARCH2 = list(
-    TOP_N = 6,
-    NBPF = 25,
-    NP = 500,
+    TOP_N = 9,
+    NBPF = 100,
+    NP = 1000,
     SPAT_REGRESSION = 0.05,
-    NREPS = 6,
+    NREPS = 4,
     NP_EVAL = 1000,
-    NREPS_EVAL = 12,
+    NREPS_EVAL = 9,
     RW_SD = chol_rw_2.sd,
@@
     TOP_N = 6,
-    NBPF = 25,
-    NP = 500,
+    NBPF = 100,
+    NP = 1000,
     SPAT_REGRESSION = 0.05,
@@
     NP_EVAL = 1000,
-    NREPS_EVAL = 12,
+    NREPS_EVAL = 9,
     RW_SD = chol_rw_3.sd,
@@
 
-  N_SEARCHES <- 1L
+  N_SEARCHES <- 3L
 } else if (RUN_LEVEL == 3) {
@@
     NBPF = 100,
-    NP = 5000,
+    NP = 2000,
     SPAT_REGRESSION = 0.05,
     NREPS = 144,
-    NP_EVAL = 5000,
+    NP_EVAL = 2000,
     NREPS_EVAL = 18,
@@
     COOLING = 0.5,
-    KEEP_TRACES = FALSE
+    KEEP_TRACES = FALSE,
+    KEEP_LIKE_MAT = FALSE
   )
@@
     NBPF = 100,
-    NP = 5000,
+    NP = 2000,
     SPAT_REGRESSION = 0.05,
     NREPS = 6,
-    NP_EVAL = 5000,
+    NP_EVAL = 2000,
     NREPS_EVAL = 18,
@@
     COOLING = 0.5,
-    KEEP_TRACES = FALSE
+    KEEP_TRACES = FALSE,
+    KEEP_LIK_MAT = FALSE
   )
@@
     NBPF = 100,
-    NP = 5000,
+    NP = 2000,
     SPAT_REGRESSION = 0.05,
     NREPS = 8,
-    NP_EVAL = 5000,
+    NP_EVAL = 2000,
     NREPS_EVAL = 18,
@@
   N_SEARCHES <- 3L
-} else {
-  stop("Run level must be 1, 2 or 3.")
 }
@@
         search3 = SEARCH3,
-        ncores = cores,
-        search_rho = TRUE,
-        search_gamma = TRUE,
-        search_Binit = FALSE
+        search_rho = FALSE,
+        search_gamma = FALSE,
+        search_Iinit = TRUE,
+        search_hur = TRUE,
+        ncores = cores
       )
@@
 for (p in shared_param_names) {
-  pat <- paste0("^", p, "[[:digit:]]+$")
-  p3s[p] <-  mean(p3[grepl(pat, names(p3))])
+  p3s[p] <- p3[paste0(p, 1)]  # They are the same for 1:10
 }

### 2022-12-05 ms.Rnw
Fixed model3 fitting rs.sd typo

@@
     return(0.015)
+  } else if (grepl('^betaB[[:digit:]]+$', x)) {
+    return(0.01)
   } else {
@@
     return(0.005)
+  } else if (grepl('^betaB[[:digit:]]+$', x)) {
+    return(0.003)
   } else {
@@
     return(0.003)
+  } else if (grepl('^betaB[[:digit:]]+$', x)) {
+    return(0.0025)
   } else {

### 2022-12-14 ms.Rnw
Pushing all files used to compile results
Minor edits to the ms

@@
 # then these results are used.
-RUN_LEVEL <- 1
+RUN_LEVEL <- 3
 rl_dir <- paste0("run_level_", RUN_LEVEL, "/")
@@
 We find a more nuanced conclusion: attention to methodological details in model fitting, diagnosis and forecasting can improve each of the proposed model's ability to quantitatively describe observed data.
-These improvements result in forecasts that are more consistent with the observed outcome, without requiring major changes to the model structures.
+This improved ability may result in more accurate forecasts and facilitates the exploration of model assumptions.
 Based on this retrospective analysis, we offer suggestions on fitting mechanistic models to dynamic systems for future studies.
@@
 Together, with the density for the initial value of the latent state $\bm{X}^{(\modelCounter)}_{0} = \bm{X}^{(\modelCounter)}(t_0)$, each model assumes a joint density $f_{\bm{X}^{(\modelCounter)}_{0:N}, \bm{Y}^{(\modelCounter)}_{1:N}}\big(\bm{x}^{(\modelCounter)}_{0:N}, \bm{y}^{(\modelCounter)}_{1:N};\paramVec\big)$, where $\paramVec$ is a parameter vector that indexes the model.
-The observed data $\bm{y}_{1:N}^*$, along with the unobserved true value of the latent state, are modeled as a realization of this joint distribution.
+The observed data $\bm{y}_{1:N}^* = y_{1:N,1:U}^*$, along with the unobserved true value of the latent state, are modeled as a realization of this joint distribution.
 
@@
 
-Latent states are initialized using an approximation of the instantaneous number of infected, asymptomatic, and recovered individuals at time $t=0$ by using the first week of cholera incidence data.
-Specifically, we set $I_{u0}(0) = \frac{y^*_{u0}}{\reportRate(\muDeath + \choleraDeath + \muIR)}$, $A_{u0}(0) = \frac{1 - \symptomFrac}{\symptomFrac}I_{u0}(0)$, and $R_{u0k} = y^*_{u0} - I_{u0}(0) - A_{u0}(0)$, and all other compartments that represent population counts are set to zero at time $t=0$.
-We replace $J_u(0)$ in \eqref{eq:model3:water} with initial value parameters $\Binit_u$ to allow for some flexibility in initializing the bacterial compartment $W_{u}(0)$ while still maintaining the primary dynamics of the model.
-Reported cholera cases are assumed to come from a negative binomial measurement model with mean equal to a fraction ($\reportRate$) of individuals in each unit who develop symptoms seek healthcare (see Sec.~S4 of the supplement material for more details).
+Latent states are initialized using an approximation of the instantaneous number of infected, asymptomatic, and recovered individuals at time $t_0$ by using the first week of cholera incidence data.
+Specifically, we set $I_{u0}(0) = \frac{y^*_{1u}}{\reportRate(\muDeath + \choleraDeath + \muIR)}$, $A_{u0}(0) = \frac{1 - \symptomFrac}{\symptomFrac}I_{u0}(0)$, $R_{u0k} = y^*_{1u} - I_{u0}(0) - A_{u0}(0)$, and initialize $B_{u}(0)$ by enforcing the rainfall dynamics supposed by the one step transition model;
+all other compartments that represent population counts are set to zero at time $t_0$.
+For each unit $u$ with zero case counts at time $t_1$, this initialization scheme results in having zero individuals in the Infected and Asymptomatic compartments, as well as having no bacteria in the aquatic reservoir.
+In reality, it's plausible that some bacteria or infected individuals were present in unit $u$ but simply went unreported. 
+As such, for departments with zero case counts in week 1, we consider estimating the number of infected individuals rather then treating this value as a constant. 
+Finally, reported cholera cases are assumed to come from a negative binomial measurement model with mean equal to a fraction ($\reportRate$) of individuals in each unit who develop symptoms seek healthcare (see Sec.~S4 of the supplement material for more details).
 
@@
 
-h2_data <- haiti2_data()
+h2_data <- haiti2_data(recreate = FALSE)
 

### 2022-12-14 ms.Rnw
Minor edits

@@
 \newcommand\Wbeta[1]{\beta_{W#1}}
+\newcommand\Whur[1]{\beta_{W#1}^{hm}}
+\newcommand\hHur[1]{h_{#1}^{hm}}
+\newcommand\tHur{t_{hm}}
 \newcommand\Wremoval{\delta_W}
@@
 \label{eq:model3:foi}
-\lambda_u(t) = \Wbeta{_u} \frac{W_u(t)}{1+W_u(t)} + \transmission_u \sum_{{\varv}\neq u}\big(I_{\varv 0}(t)+ \asymptomRelativeInfect A_{\varv 0}(t)\big),
+\lambda_u(t) = \left(\Wbeta{_u} + 1_{(t \geq \tHur)}\Whur{_u} e^{-\hHur{u}\left(t - \tHur\right)}\right) \frac{W_u(t)}{1+W_u(t)} + \transmission_u \sum_{{\varv}\neq u}\big(I_{\varv 0}(t)+ \asymptomRelativeInfect A_{\varv 0}(t)\big),
 \end{equation}
-and Per-capita transition rates are given in Equations~\ref{eq:model3:SI}--\ref{eq:model3:Decay}.
+where $\tHur$ is the time Hurricane Matthew struck Haiti \jwc{TODO Cite}, and $1_{(A)}$ is the indicator function for event $A$.
+In \cite{lee20}, $\Whur{_u}$ and $\hHur{u}$ were set to zero for all $u$;
+the need to account for the effect Hurricane Matthew had on cholera transmission for this model is explored in Sec. \jwc{TODO} of the supplement.
+
+Per-capita transition rates are given in Equations~\ref{eq:model3:SI}--\ref{eq:model3:Decay}.
 \begin{eqnarray}
@@
 In reality, it's plausible that some bacteria or infected individuals were present in unit $u$ but simply went unreported.
-As such, for departments with zero case counts in week 1, we consider estimating the number of infected individuals rather then treating this value as a constant. 
-Finally, reported cholera cases are assumed to come from a negative binomial measurement model with mean equal to a fraction ($\reportRate$) of individuals in each unit who develop symptoms seek healthcare (see Sec.~S4 of the supplement material for more details).
+As such, for departments with zero case counts in week 1, we consider estimating the number of infected individuals rather then treating this value as a constant (see supplement Sec.
+\jwc{TODO} for more details).
+Finally, reported cholera cases are assumed to come from a negative binomial measurement model with mean equal to a fraction ($\reportRate$) of individuals in each unit who develop symptoms seek healthcare
+(see Sec. \jwc{TODO} of the supplement).
 
@@
 \centering
-<<Beta_trend_Figure, fig.height=2.5, fig.width=3.7, fig.align='center'>>=
+<<Beta_trend_Figure, fig.height=2.4, fig.width=3.6, fig.align='center'>>=
 ggplot() +
@@
 \centering
-<<Model_1_Sims_Figure, fig.height=2.4, fig.width=4.7, fig.align='center'>>=
+<<Model_1_Sims_Figure, fig.height=2.35, fig.width=4.6, fig.align='center'>>=
 # This code chunk creates the figure containing simulations from Model 1
@@
 and can be justified by asymptotic considerations in a large-population limit \citep{dadlani2020,ndii17}.
-Because the process model of Model~2 is deterministic, the parameter estimation problem for Model~2 reduces to a least squares calculation when combined with a Gaussian measurement model (see supplement material for details) \jwc{TODO: cite which part of supplement}.
+Because the process model of Model~2 is deterministic, the parameter estimation problem for Model~2 reduces to a least squares calculation when combined with a Gaussian measurement model (see supplement material for details)
+\jwc{TODO: cite which part of supplement}.
 \cite{lee20} fit two versions of model~2 based on a presupposed change in cholera transmission from a epidemic phase to endemic phase that occurred in March, 2014.
@@
 
-Model~3 parameters that are calibrated to data are primarily shared between each department, meaning that the estimated parameter value for department $u$ is the same value as the parameter value in department $v$, with $u, v \in \seq{1}{U}$.
-The exception to this being the parameters $\beta_{W_u}$, and $\beta_u$, which take department-specific values for each department $u$.
-To avoid the parameter estimation issue in high-dimensional models, \citet{lee20} simplified the problem by using the IF2 algorithm to estimate parameters separately for independent department-level models;
-the shared parameters were calibrated using the cholera incidence data from Artibonite, and the department-specific parameters ($\beta_{W_u}$ and $\beta_u$) were fit using the data from their respective department.
+% Model~3 parameters that are calibrated to data are primarily shared between each department, meaning that the estimated parameter value for department $u$ is the same value as the parameter value in department $v$, with $u, v \in \seq{1}{U}$.
+% The exception to this being the parameters $\beta_{W_u}$, and $\beta_u$, which take department-specific values for each department $u$.
+To avoid the approximation error present in high-dimensional models, \citet{lee20} simplified the problem of estimating the parameters of Model~3 by creating an approximate version of the model where the units are independent given the observed data.
+% and fitting using the IF2 algorithm to estimate parameters separately for independent department-level models;
+% the shared parameters were calibrated using the cholera incidence data from Artibonite, and the department-specific parameters ($\beta_{W_u}$ and $\beta_u$) were fit using the data from their respective department.
 Reducing a spatially coupled model to individual units in this fashion requires special treatment of any interactive mechanisms between spatial units, such as found in Eq.~\myeqref{eq:model3:foi}.
-In particular, when considering a model for department $u$, the values $I_\nu(t)$ and $A_\nu(t)$, are unknown for $u \neq v$.
-To address this issue, \cite{lee20} conditioned each department level model on reported cholera cases in other departments and approximated the values of $I_\nu(t)$ and $A_\nu(t)$ using the respective reported cases.
-
-The simplified, spatially-decoupled version of Model~3 implemented by \cite{lee20} relies on the observed cholera cases for each department to describe cholera dynamics, making it impossible to obtain forecasts for future time points from the fitted model.
-Therefore, in order to obtain model forecasts, the parameters of the calibrated conditionally independent department level models were used in the fully coupled version of Model~3.
-This approach of model calibration and forecasting avoids the issue of particle depletion by simplifying the parameter estimation problem, but this advantage raises new issues.
+% In particular, when considering a model for department $u$, the values $I_\nu(t)$ and $A_\nu(t)$, are unknown for $u \neq v$.
+% To address this issue, \cite{lee20} conditioned each department level model on reported cholera cases in other departments and approximated the values of $I_\nu(t)$ and $A_\nu(t)$ using the respective reported cases.
+Because the simplified, spatially-decoupled version of Model~3 implemented by \cite{lee20} relies on the observed cholera cases it is impossible to obtain forecasts for future time points from the fitted model.
+Therefore, in order to obtain model forecasts, \cite{lee20} used the parameters estimates from the spatially-decoupled approximation of Model~3 to obtain forecasts using the fully coupled version of the model.
+This approach of model calibration and forecasting avoids the issue of particle depletion, but may also be problematic.
 One concern is that cholera dynamics in department $u$ are highly related to the dynamics in the remaining departments;
-calibrating model parameters using observed cases in other departments as a model covariate may therefore lead to an over-dependence on observed cholera cases.
-Another concern is that the two versions of the model are not the same, resulting in sub-optimal parameter estimates for the spatially coupled model.
-This is because the likelihood of the coupled model at its maximum likelihood estimate (MLE) is greater than or equal to the likelihood of the coupled model at the MLE of the decoupled model.
+calibrating model parameters while conditioning on the observed cases in other departments may therefore lead to an over-dependence on observed cholera cases.
+Another concern is that the two versions of the model are not the same, resulting in sub-optimal parameter estimates for the spatially coupled model,
+as set of parameters that maximize the likelihood of the decoupled model almost certainly do not maximize the likelihood of the fully coupled model.
 These two concerns may explain the unrealistic forecasts and low likelihood of the model (see Table~\ref{tab:likes}).
@@
 Beyond being good practical advice for applied statistics, this assertion is relevant for the philosophical justification of statistical inference as severe testing \citep[][Sec.~4.8]{mayo18}.
-In this section, we provide some tools and suggestions for diagnosing mechanistic models with the goal of making the subjective assessment of model ``usefulness'' more objective.
+In this section, we discuss some tools for diagnosing mechanistic models with the goal of making the subjective assessment of model ``usefulness'' more objective.
 To do this, we will rely on the quantitative and statistical ability of the model to match the observed data, which we call the model's {\it goodness-of-fit}, with the guiding principle that a model which cannot adequately describe observed data may not be reliable for useful purposes.
@@
 Another approach is to compare a quantitative measure of the model fit (such as MSE, predictive accuracy, or model likelihood) among all proposed models.
-These comparisons provide insight into how each model performs relative to the others.
+These comparisons, which provide insight into how each model performs relative to the others, are quite common \jwc{CITE SOME HAITI PAPERS}.
 To calibrate relative measures of fit, it is useful to compare against a model that has well-understood statistical ability to fit data, and we call this model a {\it benchmark}.
@@
 # model 1 AIC
-mod1_n_params <- 13
+mod1_n_params <- length(
+  c('beta1', 'beta2', 'beta3', 'beta4', 'beta5', 'beta6',
+    'rho', 'nu', 'E_0', 'I_0', 'tau_epi',
+    'tau_end', 'sig_sq_epi', 'sig_sq_end', 'betat')
+)
 mod1_aic <- 2 * mod1_n_params - 2 * mod1_ll
@@
 \hline
-\thead{Log-ARMA(2,1) \\ Log-likelihood} &
-  $\Sexpr{myround(ARMA_benchmarks$m1, 1)}$ &
-  $\Sexpr{myround(ARMA_benchmarks$m2, 1)}$ &
-  $\Sexpr{myround(ARMA_benchmarks$m3, 1)}$
+\thead{Log-ARMA(2,1) \\ AIC} &
+  $\Sexpr{myround(-2 * ARMA_benchmarks$m1 + 2 * 5, 1)}$ &
+  $\Sexpr{myround(-2 * ARMA_benchmarks$m2 + 2 * 5, 1)}$ &
+  $\Sexpr{myround(-2 * ARMA_benchmarks$m3 + 2 * 5, 1)}$
 \\
@@
 \end{tabular}
-\caption{\label{tab:likes}Log-likelihood values for each models compared to their ARMA benchmarks.
+\caption{\label{tab:likes}AIC values for each models compared to their ARMA benchmarks.
 Values in parenthesis are corresponding values using \citet{lee20} parameter estimates.
@@
 
-Table~\ref{tab:likes} shows that despite the significant improvements that were made by using a contemporary approach of parameter calibration for Model~3, this model still fails to beat a log-linear ARMA benchmark.
-This suggests that Model~3 is failing to adequately explain some substantial aspect of the data.
-As our goal for this article is the estimation and diagnosis of the models at hand, we explore the shortcoming in log-likelihood further in section \jwc{TODO} of the supplement material.
-There, we use additional tools like conditional log-likelihoods to understand how well the model describes each data point, given the previous observations.
-Additional tools for assessing the goodness-of-fit of a model include plotting the effective sample size of each observation \citep{liu01}, and comparing summary statistics of the observed data to simulations from the model \citep{wood10}; these summary statistics are sometimes called diagnostic probes \citep{king15,king16}.
+While a useful practice, comparing model log-likelihoods to a suitable benchmark alone may not be sufficient to diagnosing the goodness-of-fit of a given model.
+Additional techniques include the inspection of conditional log-likelihoods of each observation given the previous observations in order to understand how well the model describes each data point.
+Other tools include plotting the effective sample size of each observation \citep{liu01}; plotting the values of the hidden states from simulations (see supplement Sec.\jwc{TODO});
+and comparing summary statistics of the observed data to simulations from the model \citep{wood10,king15}.
 
@@
 If a Bayesian technique is used for parameter estimation, a natural way to account for parameter uncertainty is to obtain simulations from the model where each simulation is obtained using parameters drawn from the estimated posterior distribution.
-For frequentist inference, one possible approach obtaining model forecasts from various values of $\theta$, and then weighing the resulting forecasts by their corresponding log-likelihoods \citep{king15}; see section \jwc{TODO} of the supplement material for more details.
+For frequentist inference, one possible approach obtaining model forecasts from various values of $\theta$, and then weighing the resulting forecasts by their corresponding log-likelihoods \citep{king15}; see Sec.
+\jwc{TODO} of the supplement material for more details.
 Both of these approaches share the similarity that parameters are chosen for the forecast approximately in proportion to their corresponding value of the likelihood function, $f^{(m)}_{\bm{Y}_{1:N}}(\bm{y}_{1:N}^*; \theta)$.
@@
 \begin{figure}[ht]
-<<Mod2Fit_and_Scenarios_Figure, fig.height=2, fig.width=5, fig.align='center'>>=
+<<Mod2Fit_and_Scenarios_Figure, fig.height=1.9, fig.width=4.8, fig.align='center'>>=
 ggplot() +
@@
 \begin{figure}[ht]
-<<Elimination_Probs_Figure, fig.height=2.5>>=
+<<Elimination_Probs_Figure, fig.height=2.2>>=
 ggplot(tidyr::drop_na(all_prob_elims), aes(x = time, y = elim_prob, col = mod)) +
@@
   & \eqref{model1:beta}
-  & {\fixed $\seasAmplitude=0.4$ }
+  & $\begin{array}{l}
+     {\fixed \seasAmplitude=0.4} \\
+     \phaseParm = \Sexpr{myround(h2_fit$h2_params['phase'], 2)}
+     \end{array}$
   & \eqref{model2:lambda}
@@
   & \eqref{model1:RS}
-  & {\fixed
+  &
     $\begin{array}{l}
     \muRS^{-1}= \Sexpr{signif(1/h2_fit$h2_params["sigma"],2)} \\
-     \omega_1^{-1} = \Sexpr{myround(1/h2_fit$h2_params["Omega1"],1)} \\
-    \omega_2^{-1} = \Sexpr{myround(1/h2_fit$h2_params["Omega2"],1)}
+    {\fixed \omega_1^{-1} = \Sexpr{myround(1/h2_fit$h2_params["Omega1"],1)}} \\
+    {\fixed \omega_2^{-1} = \Sexpr{myround(1/h2_fit$h2_params["Omega2"],1)}}
     \end{array}$
-    }
   & $\begin{array}{l}
@@
     \end{array} $
-  & {\fixed $\muRS^{-1}=\Sexpr{myround(1/p3s["rho"],1)}$}
+  & {\fixed $\muRS^{-1}=\Sexpr{myround(1/p3['rho1'],1)}$}
   & \eqref{eq:model3:RRnext}
@@
   & \eqref{model2:mu_EI}
-  & {\fixed $\symptomFrac=\Sexpr{myround(p3s["sigma"],2)}$}
+  & {\fixed $\symptomFrac=\Sexpr{myround(p3["sigma1"],2)}$}
   & \eqref{eq:model3:SA}
@@
   &
-  & $\Wremoval^{-1} = $ \Sexpr{signif(52/h2_fit$h2_params["Delta"],3)}
+  & {\fixed $\Wremoval^{-1} = $ \Sexpr{signif(52/h2_fit$h2_params["Delta"],3)}}
   & \eqref{model2:from_W}
@@
 (Bottom) estimated seasonality in the transmission rate (dashed line) plotted alongside mean rainfall (solid line).
+The outsided effect of rainfall in the fall may be due to Hurricane Matthew, which struck Haiti in October of 2016 and resulted in an increase of cholera cases in the nation.
 }

### 2022-12-14 si/si.Rnw
Minor edits

@@
 \pdfminorversion=4
-
-<<debug,echo=F>>=
-run_level <- 1
-@
 
@@
 
-<<Load lee Model1 Params, echo=FALSE>>=
+<<Simulate Lee Model 1, echo=FALSE, cache=TRUE>>=
 h1_epi <- haiti1()
-RL_dir <- paste0("run_level_", run_level, "/")
-@
-
-<<Simulate Lee Model 1, echo=FALSE, cache=TRUE>>=
 N_LEE_SIMS <- 20
@@
 <<echo=FALSE, include=FALSE, message=FALSE>>=
-h1_epi_evals <- readRDS(paste0("../model1/", RL_dir, "lee1_epi_evals.rds"))
-h1_end_evals <- readRDS(paste0("../model1/", RL_dir, "lee1_end_evals.rds"))
+h1_epi_evals <- readRDS("../model1/run_level_3/lee1_epi_evals.rds")
+h1_end_evals <- readRDS("../model1/run_level_3/lee1_end_evals.rds")
 
@@
 <<Model3Replication, echo=FALSE, cache=TRUE>>=
-NSIM <- switch(run_level, 50, 200, 500)
+NSIM <- 500
 
@@
 lee3 <- lee3_spatPomp()
-lee3_sims <- simulate(lee3, nsim = NSIM, seed = 321, format = 'data.frame')
+lee3_sims <- bake("../model3/si_lee3Sims.rds", {
+  simulate(lee3, nsim = NSIM, seed = 321, format = 'data.frame')
+})
 
@@
 @
+
+\section{Calibrating Model~3 to observed cases}
+
+<<CalibrateOrig3, echo=FALSE>>=
+set.seed(8636712)
+
+# Set Run Level for debug, timing, and full computation
+RUN_LEVEL = 2
+
+# Create vectors for the unit and shared parameters
+unit_specific_names <- c("betaB", "foi_add")
+shared_param_names <- c(
+  "mu_B", "XthetaA", "thetaI", "lambdaR", "r", "std_W",
+  "epsilon", "k"
+)
+est_param_names <- c(
+  unit_specific_names, shared_param_names
+)
+
+# Add unit numbers to each parameter
+est_param_names_expanded <- paste0(rep(est_param_names, each = 10), 1:10)
+
+set_rw_1 <- function(x) {
+  if (gsub("[[:digit:]]+$", "", x) %in% shared_param_names) {
+    return(0.01)
+  } else if (x %in% paste0("k", 1:10)) {
+    return(0.01)
+  } else if (grepl('^foi_add[[:digit:]]+$', x) || grepl('^betaB[[:digit:]]+$', x)) {
+    return(0.01)
+  } else {
+    return(0)
+  }
+}
+
+set_rw_2 <- function(x) {
+  if (gsub("[[:digit:]]+$", "", x) %in% shared_param_names) {
+    return(0.003)
+  } else if (x %in% paste0("k", 1:10)) {
+    return(0.003)
+  } else if (grepl('^foi_add[[:digit:]]+$', x) || grepl('^betaB[[:digit:]]+$', x)) {
+    return(0.003)
+  } else {
+    return(0)
+  }
+}
+
+set_rw_3 <- function(x) {
+  if (gsub("[[:digit:]]+$", "", x) %in% shared_param_names) {
+    return(0.0025)
+  } else if (x %in% paste0("k", 1:10)) {
+    return(0.0025)
+  } else if (grepl('^foi_add[[:digit:]]+$', x) || grepl('^betaB[[:digit:]]+$', x)) {
+    return(0.0025)
+  } else {
+    return(0)
+  }
+}
+
+# First Search RW
+reg_rw_1.sd <- lapply(est_param_names_expanded, set_rw_1)
+names(reg_rw_1.sd) <- est_param_names_expanded
+chol_rw_1.sd <- do.call(rw.sd, reg_rw_1.sd)
+
+# Second Search RW
+reg_rw_2.sd <- lapply(est_param_names_expanded, set_rw_2)
+names(reg_rw_2.sd) <- est_param_names_expanded
+chol_rw_2.sd <- do.call(rw.sd, reg_rw_2.sd)
+
+# Third Search RW
+reg_rw_3.sd <- lapply(est_param_names_expanded, set_rw_3)
+names(reg_rw_3.sd) <- est_param_names_expanded
+chol_rw_3.sd <- do.call(rw.sd, reg_rw_3.sd)
+
+# Search hyper-parameters for each run-level and search
+SEARCH1 = list(
+  NBPF = 100,
+  NP = 1000,
+  SPAT_REGRESSION = 0.05,
+  NREPS = 72,
+  NP_EVAL = 1000,
+  NREPS_EVAL = 8,
+  RW_SD = chol_rw_1.sd,
+  COOLING = 0.5,
+  KEEP_TRACES = TRUE,
+  KEEP_LIK_MAT = TRUE
+)
+
+SEARCH2 = list(
+  TOP_N = 9,
+  NBPF = 100,
+  NP = 1000,
+  SPAT_REGRESSION = 0.05,
+  NREPS = 4,
+  NP_EVAL = 1000,
+  NREPS_EVAL = 9,
+  RW_SD = chol_rw_2.sd,
+  COOLING = 0.5,
+  KEEP_TRACES = TRUE,
+  KEEP_LIK_MAT = TRUE
+)
+
+SEARCH3 = list(
+  TOP_N = 6,
+  NBPF = 100,
+  NP = 1000,
+  SPAT_REGRESSION = 0.05,
+  NREPS = 6,
+  NP_EVAL = 1000,
+  NREPS_EVAL = 9,
+  RW_SD = chol_rw_3.sd,
+  COOLING = 0.5,
+  KEEP_TRACES = TRUE,
+  KEEP_LIK_MAT = TRUE
+)
+
+N_SEARCHES <- 3L
+
+save_file <- "../model3/run_level_2/fit_orig_model.rds"
+
+# TODO: replace with bake
+haiti3_orig_fit <- readRDS(save_file)
+
+# haiti3_orig_fit <- bake(
+#   file = save_file, {
+#     fit_coupled_haiti3(
+#       nsearches = N_SEARCHES,
+#       search1 = SEARCH1,
+#       search2 = SEARCH2,
+#       search3 = SEARCH3,
+#       search_rho = FALSE,
+#       search_gamma = FALSE,
+#       ncores = cores
+#     )
+#   }
+# )
+
+best_m <- which.max(haiti3_orig_fit$search3$logLiks$logLik)
+
+best_params <- haiti3_orig_fit$search3$params[best_m, ]
+
+all_liks <- as.data.frame(haiti3_orig_fit$search3$likMat)
+all_liks$which <- rep(1:36, each = 9)
+best_mod <- filter(all_liks, which == best_m)
+
+unitLikesM3 <- best_mod %>%
+  pivot_longer(
+    cols = -which,
+    names_to = "unit",
+    names_prefix = "V",
+    values_to = "logLik"
+  ) %>%
+  select(-which) %>%
+  group_by(unit) %>%
+  summarize(
+    ll_est = logmeanexp(logLik),
+    ll_se = logmeanexp(logLik, se = TRUE)[2]
+  ) %>%
+  ungroup() %>%
+  mutate(unit = as.numeric(unit)) %>%
+  arrange(unit)
+@
+
+<<M3_bench, echo=FALSE, message=FALSE, include=FALSE>>=
+MODEL3_CASES <- haitiCholera %>%
+  dplyr::rename(
+    date = date_saturday, Grande_Anse = Grand.Anse,
+    Nord_Est = Nord.Est, Nord_Ouest = Nord.Ouest,
+    Sud_Est = Sud.Est
+  ) %>%
+  dplyr::mutate(date = as.Date(date)) %>%
+  dplyr::select(-report) %>%
+  dplyr::mutate(time_diff = date - lag(date))
+
+jumps <- which(MODEL3_CASES$time_diff > lubridate::make_difftime(day = 7))
+
+temp <- MODEL3_CASES
+for (j in 1:length(jumps)) {
+  temp <- rbind(
+    temp[1:(jumps[j] + j - 2), ],
+    rep(NA, ncol(temp)),
+    temp[(jumps[j] + j - 1):nrow(temp), ]
+  )
+}
+
+mod3_cases <- temp %>%
+  dplyr::select(-time_diff)
+
+mod3_departements <-
+  c(
+    'Artibonite',
+    'Centre',
+    'Grande_Anse',
+    'Nippes',
+    'Nord',
+    'Nord-Est',
+    'Nord-Ouest',
+    'Ouest',
+    'Sud',
+    'Sud-Est'
+  )
+
+m3_benchmark <- c()
+for (dep in mod3_departements) {
+
+  # Get department cases
+  cases <- as.numeric(mod3_cases[, gsub("-", "_", dep)])
+
+  # Fit model on natural scale of cases
+  m3_a201 <- arima(cases, order = c(2, 0, 1))
+
+  # Fit model on log-scale of cases
+  m3_log_a201 <- arima(log(cases + 1), order = c(2, 0, 1))
+
+  m3_benchmark <- c(m3_benchmark, max(
+    m3_log_a201$loglik - sum(log(cases + 1), na.rm = TRUE),
+    m3_a201$loglik
+  )
+  )
+}
+
+rm(cases, m3_a201, m3_log_a201, mod3_departements, dep,
+   mod3_cases, temp, jumps, MODEL3_CASES, j)
+
+unitLikesARMA <- data.frame(
+  unit = 1:10,
+  ll = m3_benchmark
+) %>%
+  as_tibble()
+@
+
+<<combineARMA_M3, echo=FALSE>>=
+h3 <- haiti3_spatPomp()
+deps <- h3@unit_names
+ARMA_M3_joined <- dplyr::left_join(
+  x = unitLikesM3,
+  y = unitLikesARMA,
+  by = "unit"
+) %>%
+  select(-ll_se) %>%
+  mutate(unit = as.character(unit)) %>%
+  pivot_longer(
+    cols = -unit,
+    names_to = "type",
+    values_to = "ll"
+  ) %>%
+  mutate(unit = paste0(unit, ": ", deps[as.numeric(unit)]))
+@
+
+<<M3bpf, echo=FALSE, message=FALSE, include=FALSE>>=
+
+coef(h3)[names(best_params)] <- best_params
+
+registerDoRNG(40765101)
+
+bpf_out <- bake(
+  file = "../model3/run_level_3/si_bpf.rds", {
+    h3 %>% bpfilter(
+      Np = 2000,
+      block_size = 1,
+      save_states = TRUE
+    )
+  }
+)
+
+@
+
+In this section, we provide more detail on the process that was used to estimate the coefficients used in Model~3.
+In particular, we discuss why we decided to include additional model parameters that were not considered by \cite{lee20} (parameters related to initial conditions and Hurricane Matthew).
+To calibrate this model, we used the iterated block particle filter (IBPF) method of \cite{ionides22}.
+Due to the novelty of this algorithm, there does not exist published examples the IBPF algorithm used for data analysis outside of the papers in which the algorithm was introduced \citep{ning21ibpf,ionides22}, which at least partially motivates the creation of this section.
+
+\cite{lee20} were only able to estimate model parameters to a simplified version of Model~3 on a subset of the available data, as no method existed at the time of their publication to fit a fully coupled meta-population model to disease incidence data.
+Building on their results, we fit the fully coupled version of Model~3 to all available data.
+Our preliminary results (not shown) suggest that a single search for the MLE using the IBPF algorithm is insufficient;
+this result is consistent with the findings of \cite{ionides22}, who used a sequence of successive searches for the MLE in order to properly maximize the model likelihood.
+We therefore use a similar approach by performing three successive searches for the MLE.
+The first search is performed by obtaining initial values for the parameters by uniformly sampling values from a predefined hypercube.
+Subsequent searches use the resulting parameter values that correspond the largest model likelihoods.
+
+We use the technique described above to fit the fully coupled version of Model~3 proposed by \cite{lee20} to all observed cholera data.
+The maximum likelihood we obtained with the model was $\Sexpr{round(max(haiti3_orig_fit$search3$logLiks$logLik))}$, which is higher than the log-ARMA benchmark ($\Sexpr{round(sum(unitLikesARMA$ll))}$).
+For meta-population models, it is worth considering how well the model fits the data to each spatial unit.
+This can be done by looking at conditional log-likelihoods, which is part of the output of the \texttt{bpfilter} algorithm in the \texttt{spatPomp} package.
+The likelihoods for each department, compared to the corresponding log-ARMA benchmark, are displayed in Fig.~\ref{fig:h3UnitLikes}.
+The figure demonstrates that while the log-likelihood of the fitted model outperforms the log-ARMA benchmark, the model has lower likelihoods than the benchmark for some departments.
+This result warrants further investigation of the model fit.
+
+\begin{figure}[ht]
+<<h3UnitLikes, fig.height=3, fig.width=5.5, echo=FALSE>>=
+ggplot(ARMA_M3_joined, aes(x = ll, y = reorder(unit, ll))) +
+  geom_point(aes(col = type)) +
+  ylab("Unit") +
+  xlab("Log Likelihood") +
+  scale_color_manual(
+    labels = c("ll" = "ARMA", "ll_est" = "Model 3"),
+    values = c("ll" = "#1b9e77", "ll_est" = "#d95f02")
+  ) +
+  theme(legend.title = element_blank(),
+        axis.title.y = element_blank())
+@
+\caption{\label{fig:h3UnitLikes}Log-likelihoods of Model~3 for each department compared to the corresponding log-ARMA benchmark.}
+\end{figure}
+
+To investigate the goodness-of-fit of Model~3 further and attempt to determine why the model is underperforming in some departments, we plot the conditional log likelihoods of each observation given the previous observations for each department in Fig.~\ref{fig:condLL}.
+This figure reveals that the fitted model has difficult time explaining certain features of the data.
+For example, many departments (in particular Sud and Ouest) have observations with lower conditional log-likelihoods near October 2016 than at other time points.
+Further investigation of the model output reveals that the model is struggling to explain the sudden surge in cholera cases that occurred at this time, which coincides with the time that Hurricane Matthew struck Haiti.
+While the model does include a mechanism to account for increased cholera transmission due to large rainfall events, the mechanism does not appear to be sufficient to capture the damaging effects of the hurricane, which was most impactful in the the Sud and Ouest departments \jwc{TODO: CITE}.
+This result led us to include parameters that would increase the transmission between environmental cholera and humans for Sud and Ouest \jwc{TODO: just check these are the correct departments}.
+Other \jwc{TODO: IVPs}.
+
+\begin{figure}[ht]
+<<CondLLFig, echo=FALSE, message=FALSE, fig.height=3>>=
+deps[grepl("_", deps)] <- gsub("_", "-", deps[grepl("_", deps)])
+deps[deps == "Grande-Anse"] <- "Grand\'Anse"
+names(deps) <- as.character(1:10)
+
+myLabeller <- as_labeller(deps)
+
+t(bpf_out@block.cond.loglik) %>%
+  as.data.frame() %>%
+  mutate(t = bpf_out@times) %>%
+  pivot_longer(
+    cols = -t,
+    values_to = 'likelihood',
+    names_prefix = "V",
+    names_to = 'Unit'
+  ) %>%
+  mutate(date = as.Date(lubridate::date_decimal(t))) %>%
+  ggplot(aes(x = date, y = likelihood)) +
+  geom_line() +
+  geom_point() +
+  facet_wrap(~Unit, labeller = myLabeller, scales = 'free_y',
+             nrow = 2) +
+  theme_bw() +
+  theme(axis.title.x = element_blank()) +
+  ylab("Conditional log-likelihoods")
+@
+\caption{\label{fig:condLL}Conditional log-likelihoods of Model~3 prior to the inclusion of parameters for initial values and Hurricane Matthew.}
+\end{figure}
 

### 2022-12-15 ms.Rnw
Minor Edits to ms and si

@@
 \subsubsection{Calibrating Model~3 Parameters}\label{sec:fit3}
-Both the latent and observable processes in Model~3 can be be factored into department specific processes which interact with each other.
+Model~3 describes cholera dynamics in Haiti using a metapopulation model, where the hidden states in each administrative department has an effect on the dymanics in other departments.
 The decision to address metapopulation dynamics using a spatially explicit model, rather than to aggregate over space, is double-edged.
@@
 That may explain why \citet{lee20} chose to obtain forecasts by simulating the calibrated models forward from initial conditions.
-Here, we compare with a forecast projected from the filtered distribution: to obtain a forecast using POMP model $m$ at a collection of times $t_{N+1:N+s}$, where $N$ is the index for the last available data, we simulate forward from a draw from
+Here, we compare with a forecast projected from the filtering distribution: to obtain a forecast using POMP model $m$ at a collection of times $t_{N+1:N+s}$, where $N$ is the index for the last available data, we simulate forward from a draw from
 $f^{(m)}_{\bm{X}_N|\bm{Y}_{1:N}}(\bm{x}_{N} | \bm{y}^*_{1:N} ; \hat\theta)$

### 2022-12-15 si/si.Rnw
Minor Edits to ms and si

@@
 \newcommand\modelCounter{m}
+\newcommand\Whur[1]{\beta_{W#1}^{hm}}
+\newcommand\hHur[1]{h_{#1}^{hm}}
 
@@
 @
-\caption{\label{fig:h3UnitLikes}Log-likelihoods of Model~3 for each department compared to the corresponding log-ARMA benchmark.}
+\caption{\label{fig:h3UnitLikes}Log-likelihoods of Model~3 for each department compared to the corresponding log-ARMA benchmark prior to the inclusion of parameters for initial values and Hurricane Matthew.}
 \end{figure}
 
-To investigate the goodness-of-fit of Model~3 further and attempt to determine why the model is underperforming in some departments, we plot the conditional log likelihoods of each observation given the previous observations for each department in Fig.~\ref{fig:condLL}.
+In an attempt to determine why the model is underperforming in some departments, we plot the conditional log likelihoods of each observation given the previous observations for each department in Fig.~\ref{fig:condLL}.
 This figure reveals that the fitted model has difficult time explaining certain features of the data.
-For example, many departments (in particular Sud and Ouest) have observations with lower conditional log-likelihoods near October 2016 than at other time points.
+For example, many departments (in particular Sud) have observations with lower conditional log-likelihoods near October 2016 than at other time points.
 Further investigation of the model output reveals that the model is struggling to explain the sudden surge in cholera cases that occurred at this time, which coincides with the time that Hurricane Matthew struck Haiti.
-While the model does include a mechanism to account for increased cholera transmission due to large rainfall events, the mechanism does not appear to be sufficient to capture the damaging effects of the hurricane, which was most impactful in the the Sud and Ouest departments \jwc{TODO: CITE}.
-This result led us to include parameters that would increase the transmission between environmental cholera and humans for Sud and Ouest \jwc{TODO: just check these are the correct departments}.
-Other \jwc{TODO: IVPs}.
+While the model does include a mechanism to account for increased cholera transmission due to large rainfall events, the mechanism does not appear to be sufficient to capture the damaging effects of the hurricane, which had the greatest impact in the the Sud and Grand'Anse departments \jwc{TODO: CITE}.
+This result led us to include parameters $\Whur{u}$ and $\hHur{u}$ in Eq.~23 that would increase the transmission between environmental cholera and humans for Sud and Grand'Anse.
+The effect of the hurricane on cholera transmission is assumed to have an exponential decay, where the magnitude is determined by $\Whur{u}$ and the duration of the effect determined by $\hHur{u}$.
 
@@
 \end{figure}
+
+Other points of concern include the observations at or near the start of the recorded cases, for example, in Nord-Ouest and Sud-Est.
+These low conditional log-likelihoods are a result of having zero or near zero confirmed cases in the first week of observed data, followed by a rapid growth of cases.
+The model cannot explain the rapid growth of cholera in the absence of starting cases because the initialization scheme forces departments with no recorded cases in the first week to have hidden states $I_{u0}(0) = A_{u0}(0) = B_{u}(0) = 0$.
+In reality, however, it is possible to have some infected or asymptomatic individuals in a department when no cases are recorded due to under-reporting.
+This may be particularly true at the onset of an infectious disease outbreak, as lower public awareness and inadequate testing infrastructure may result in many cases going undetected \jwc{TODO: CITE}.
+We therefore include and estimate $I_{0u}$ that determines the initial number of Infected individuals at the start of the pandemic for the departments Grande'Anse, Nord-Ouest, and Sud-Est;
+each of these departments have both low starting case counts followed by rapid growth, and low initial values of the conditinoal log-likelihood.
+
+<<loadFinalModel3, echo=FALSE, include=FALSE, message=FALSE>>=
+h3_final_results <- readRDS("../model3/run_level_3/haiti3_coupled_fit.rds")
+
+best_m <- which.max(h3_final_results$search3$logLiks$logLik)
+best_params <- h3_final_results$search3$params[best_m, ]
+all_liks <- as.data.frame(h3_final_results$search3$likMat)
+all_liks$which <- rep(1:72, each = 18)
+best_mod <- dplyr::filter(all_liks, which == best_m)
+
+unitLikesM3 <- best_mod %>%
+  pivot_longer(
+    cols = -which,
+    names_to = "unit",
+    names_prefix = "V",
+    values_to = "logLik"
+  ) %>%
+  select(-which) %>%
+  group_by(unit) %>%
+  summarize(
+    ll_est = pomp::logmeanexp(logLik),
+    ll_se = pomp::logmeanexp(logLik, se = TRUE)[2]
+  ) %>%
+  ungroup() %>%
+  mutate(unit = as.numeric(unit)) %>%
+  arrange(unit)
+
+ARMA_M3_joined <- dplyr::left_join(
+  x = unitLikesM3,
+  y = unitLikesARMA,
+  by = "unit"
+) %>%
+  select(-ll_se) %>%
+  mutate(unit = as.character(unit)) %>%
+  pivot_longer(
+    cols = -unit,
+    names_to = "type",
+    values_to = "ll"
+  ) %>%
+  mutate(unit = paste0(unit, ": ", deps[as.numeric(unit)]))
+@
+
+We refit Model~3 after introducing these seven parameters---three initial value parameters and four parameters related to Hurricane Matthew---using the same successive search scheme described above.
+The resulting model has a log-likelihood value of $\Sexpr{round(max(h3_final_results$search3$logLiks$logLik), 1)}$.
+We also provide figures for department specific log-likelihoods and conditional log-likelihoods in Figs.~\ref{fig:finalUnitLL} and \ref{fig:finalCondLL}.
+
+\begin{figure}[ht]
+<<finalUnitLLFig, fig.height=3, fig.width=5.5, echo=FALSE>>=
+ggplot(ARMA_M3_joined, aes(x = ll, y = reorder(unit, ll))) +
+  geom_point(aes(col = type)) +
+  ylab("Unit") +
+  xlab("Log Likelihood") +
+  scale_color_manual(
+    labels = c("ll" = "ARMA", "ll_est" = "Model 3"),
+    values = c("ll" = "#1b9e77", "ll_est" = "#d95f02")
+  ) +
+  theme(legend.title = element_blank(),
+        axis.title.y = element_blank())
+@
+\caption{\label{fig:finalUnitLL}Log-likelihoods of Model~3 for each department compared to the corresponding log-ARMA benchmark after adding and estimating parameters related to initial values and Hurricane Matthew.}
+\end{figure}
+
+<<finalModel3BPF, echo=FALSE, message=FALSE, include=FALSE>>=
+coef(h3)[names(best_params)] <- best_params
+
+bpf_out <- bake(
+  file = "../model3/run_level_3/si_h3_final_bpf.rds", {
+    h3 %>% bpfilter(
+      Np = 2000,
+      block_size = 1,
+      save_states = TRUE
+    )
+  },
+  seed = 786321
+)
+@
+
+\begin{figure}[ht]
+<<finalCondLLFig, echo=FALSE, message=FALSE, fig.height=3>>=
+t(bpf_out@block.cond.loglik) %>%
+  as.data.frame() %>%
+  mutate(t = bpf_out@times) %>%
+  pivot_longer(
+    cols = -t,
+    values_to = 'likelihood',
+    names_prefix = "V",
+    names_to = 'Unit'
+  ) %>%
+  mutate(date = as.Date(lubridate::date_decimal(t))) %>%
+  ggplot(aes(x = date, y = likelihood)) +
+  geom_line() +
+  geom_point() +
+  facet_wrap(~Unit, labeller = myLabeller, scales = 'free_y',
+             nrow = 2) +
+  theme_bw() +
+  theme(axis.title.x = element_blank()) +
+  ylab("Conditional log-likelihoods")
+@
+\caption{\label{fig:finalCondLL}Conditional log-likelihoods of Model~3 after adding and estimating the parameters for initial values and Hurricane Matthew.}
+\end{figure}
+
+While the addition of the Hurricane parameters seems to have resulted in a dramatic increase of conditional likelihoods around and after October 2016, the inclusion of the initial value parameters does not appear to have fixed the low conditional log-likelihood values in Sud-Est and Nord-Ouest.
+This could be due to many factors, including the possibility that the coupling between units forces low initial value parameters in a given department in order for the model to fit the data in the remaining departments.
+Whatever the case may be, the inclusion of the seven parameters resulted in a $\Sexpr{round(max(h3_final_results$search3$logLiks$logLik - haiti3_orig_fit$search3$logLiks$logLik), 1)}$, which is a highly statistically significant difference.
+One could similarly test whether the inclusion of the initial value parameters resulted in a significantly better fit to the observed data, but we do not do this here due to the large computational effort that would be required.
+
+\subsection{Examining the Hidden States of the Calibrated Model}
+
+For mechanistic models, beating a suitable statistical benchmark does not alone guarantee that the model provides an accurate description of a dynamic process.
+Indeed, a good statistical fit does not require the model to assert a causal explanation.
+For example, reconstructed latent variables should make sense in the context of alternative measurements of these variables \citep{grad12}.
+We demonstrate this principle by examining the latent state of the calibrated model.
+In particular, we examine the compartment of susceptible individuals under various scenarios, as this compartment drives transmission dynamics in SIR models for infectious disease outbreaks.
+
+Recall that the filtering distribution for the calibrated version of Model~3 at time $t_k$ is defined as the distribution of the latent state at time $t_k$ given the data from times $t_{1}:t_{k}$, i.e.
+$f^{(3)}_{\bm{X}_k|\bm{Y}_{1:k}}(\bm{x}_{k} | \bm{y}^*_{1:k} ; \hat\theta)$.
+In general, one may expect simulations from the filtering distribution of a model with a good statistical fit to result in hidden states that are highly consistent with the observed data because the filtering distribution is conditioned on the observed data.
+Fig.~\ref{fig:h3Sus} shows the number of susceptible individuals from various simulations of the model:
+simulations from Model~3 under initial conditions are displayed in red; simulations from the filtering distribution of model are displayed in blue.
 

### 2022-12-15 ms.Rnw
minor edits

@@
       h3_V0_res$mod3_V0_probs <- get_elimProbs(h3_V0_sims, model = 3)
+
+      # For the supplement material, save quantiles of the Susceptible individuals
+      h3_V0_res$SQuants <- h3_V0_sims %>%
+        mutate(
+          date = as.Date(lubridate::round_date(lubridate::date_decimal(time), unit = 'day'))
+        ) %>%
+        group_by(unitname, date) %>%
+        mutate(
+          pop = h3Spat@params[paste0("H", which(first(unitname) == unit_names(h3Spat)))]
+        ) %>%
+        summarize(
+          Q025 = quantile((S + VSd + VSdd + VSdd_alt + VSd_alt) / pop, probs = 0.025),
+          Q50 = quantile((S + VSd + VSdd + VSdd_alt + VSd_alt) / pop, probs = 0.5),
+          Q975 = quantile((S + VSd + VSdd + VSdd_alt + VSd_alt) / pop, probs = 0.975)
+        ) %>%
+        ungroup() %>%
+        rename(dep = unitname)
 

### 2022-12-15 si/si.Rnw
minor edits

@@
 In general, one may expect simulations from the filtering distribution of a model with a good statistical fit to result in hidden states that are highly consistent with the observed data because the filtering distribution is conditioned on the observed data.
-Fig.~\ref{fig:h3Sus} shows the number of susceptible individuals from various simulations of the model:
+Fig.~\ref{fig:h3Sus} shows the percentage of individuals that are in the susceptible compartment from various simulations of the model:
 simulations from Model~3 under initial conditions are displayed in red; simulations from the filtering distribution of model are displayed in blue.
+This figure shows that simulations from initial conditions tends to result in a much more rapid depletion of susceptible individuals at the start of the epidemic than simulations from the filtering distribution, suggesting the calibrated model has a propensity to predict larger outbreaks than what is typically seen in the data.
+This also appears true in later stages of the time period of observed data, where the data suggests that there is a replenishment of the susceptible compartment, but the Model simulations typically retain a smaller proportion of susceptible individuals.
+This result demonstrates that the calibrated model favors a more rapid growth in cholera cases than what is typically seen in the observed data, which explains why the model fails to predict the absence of cases between February, 2019 and October, 2022
+
+<<h3SusFigSetup, echo=FALSE, message=FALSE, include=FALSE>>=
+init_conditions_results <- bake(
+  file = "../model3/run_level_3/si_InitSimsQuants.rds", {
+
+    results <- list()
+
+    h3Spat_sims <- simulate(
+      h3, nsim = 200, format = 'data.frame'
+    )
+
+    h3_spat_end <- h3Spat_sims %>%
+      filter(time == max(time))
+
+    results[["end_conditions"]] <- h3_spat_end %>%
+      select(-time, -cases) %>%
+      mutate(
+        unit = case_when(
+          unitname == "Artibonite" ~ 1,
+          unitname == "Centre" ~ 2,
+          unitname == "Grande_Anse" ~ 3,
+          unitname == "Nippes" ~ 4,
+          unitname == "Nord" ~ 5,
+          unitname == "Nord_Est" ~ 6,
+          unitname == "Nord_Ouest" ~ 7,
+          unitname == "Ouest" ~ 8,
+          unitname == "Sud" ~ 9,
+          unitname == "Sud_Est" ~ 10
+        )
+      ) %>%
+      select(-unitname) %>%
+      pivot_wider(
+        data = .,
+        names_from = unit,
+        values_from = c(S, I, A, R_one, R_two, R_three, VSd, VR1d, VR2d,
+                        VR3d, VSdd, VR1dd, VR2dd, VR3dd, VSd_alt, VR1d_alt,
+                        VR2d_alt, VR3d_alt, VSdd_alt, VR1dd_alt, VR2dd_alt,
+                        VR3dd_alt, C, B, Doses, totInc),
+        names_sep = ""
+      ) %>%
+      select(-.id) %>%
+      as.matrix() %>%
+      t()
+
+    results[["SQuants"]] <- h3Spat_sims %>%
+      rename(dep = unitname) %>%
+      group_by(dep, time) %>%
+      summarise(
+        Q025 = quantile(S + VSd + VSdd + VSd_alt + VSdd_alt, probs = 0.025),
+        Q50  = quantile(S + VSd + VSdd + VSd_alt + VSdd_alt, probs = 0.5),
+        Q975 = quantile(S + VSd + VSdd + VSd_alt + VSdd_alt, probs = 0.975)
+      ) %>%
+      ungroup() %>%
+      rowwise() %>%
+      mutate(
+        unit_num = which(h3@unit_names == dep),
+        date = lubridate::date_decimal(time),
+        Q025 = Q025 / h3@params[paste0("H", unit_num)],
+        Q50  = Q50  / h3@params[paste0("H", unit_num)],
+        Q975 = Q975 / h3@params[paste0("H", unit_num)]
+      ) %>%
+      filter(date >= as.Date(lubridate::round_date(lubridate::date_decimal(h3@t0), unit = 'day'))) %>%
+      mutate(date = as.Date(lubridate::round_date(date, unit = 'day')))
+
+    results
+  },
+  seed = 834687
+)
+
+h3Spat_quants <- init_conditions_results[["SQuants"]]
+
+mod3_V0_initcond_res <- bake(
+  file = '../model3/run_level_3/si_vacc_sim_from_init.rds', {
+    h3_V0_res <- list()
+
+    # Get "noVacc" scenario parameters
+    h3_S0_par <- get_model3_vacc_scenario_params(scenario = "noVacc")
+    h3Spat <- h3
+    p3 <- best_params
+    coef(h3Spat) <- p3
+    coef(h3Spat)[names(h3_S0_par)] <- h3_S0_par
+
+    h3_bpf_end_states <- init_conditions_results$end_conditions
+
+    # Using parameters and filtered distribution, project system
+    h3_V0_sims <- project_from_filter(
+      h3Spat, end_states = h3_bpf_end_states, covarGen = project_rain,
+      nsims = 500, seed = 184697
+    )
+
+    # Store nationally aggregated results
+    h3_V0_res$mod3_V0_sims <- agg_mod3_sims(h3_V0_sims)
+    h3_V0_res$mod3_V0_probs <- get_elimProbs(h3_V0_sims, model = 3)
+
+    h3_V0_res$UnitQuants <- h3_V0_sims %>%
+      mutate(
+        date = as.Date(lubridate::round_date(lubridate::date_decimal(time), unit = 'day'))
+      ) %>%
+      group_by(unitname, date) %>%
+      summarize(
+        Q025 = quantile(cases, probs = 0.025),
+        Q50 = quantile(cases, probs = 0.5),
+        Q975 = quantile(cases, probs = 0.975)
+      ) %>%
+      ungroup() %>%
+      rename(dep = unitname)
+
+    h3_V0_res$SQuants <- h3_V0_sims %>%
+      mutate(
+        date = as.Date(lubridate::round_date(lubridate::date_decimal(time), unit = 'day'))
+      ) %>%
+      group_by(unitname, date) %>%
+      mutate(
+        pop = h3@params[
+          paste0("H", which(first(unitname) == unit_names(h3)))
+        ]
+      ) %>%
+      summarize(
+        Q025 = quantile((S + VSd + VSdd + VSdd_alt + VSd_alt) / pop, probs = 0.025),
+        Q50 = quantile((S + VSd + VSdd + VSdd_alt + VSd_alt) / pop, probs = 0.5),
+        Q975 = quantile((S + VSd + VSdd + VSdd_alt + VSd_alt) / pop, probs = 0.975)
+      ) %>%
+      ungroup() %>%
+      rename(dep = unitname)
+
+    h3_V0_res
+  }
+)
+
+Sus_df <- bake(
+  file = "../model3/run_level_3/si_S_quants", {
+    foreach(i=1:length(h3@times), .combine = bind_rows) %do% {
+      get_h3_S_quants(bpf_out, i)
+    }
+  }
+)
+
+mod3_V0_res <- readRDS("../model3/run_level_3/mod3_V0_res.rds")
+@
+
+\begin{figure}[ht]
+<<h3SusFig, echo=FALSE, message=FALSE, fig.height=3>>=
+ggplot() +
+  # Sims from initial condition
+  geom_line(data = h3Spat_quants, aes(x = date, y = Q50), col = 'red') +
+  geom_ribbon(
+    data = h3Spat_quants,
+    aes(x = date, ymin = Q025, ymax = Q975),
+    fill = 'red', alpha = 0.5
+  ) +
+  # projections from initial condition, using filtered sims function
+  geom_line(data = mod3_V0_initcond_res$SQuants, aes(x = date, y = Q50), col = 'red') +
+  geom_ribbon(
+    data = mod3_V0_initcond_res$SQuants,
+    aes(x = date, ymin = Q025, ymax = Q975),
+    fill = 'red', alpha = 0.5
+  ) +
+  # Sims from filtered distribution
+  geom_line(
+    data = Sus_df,
+    aes(x = as.Date(lubridate::date_decimal(time)), y = Q50),
+    col = 'blue'
+  ) +
+  geom_ribbon(
+    data = Sus_df,
+    aes(x = as.Date(lubridate::date_decimal(time)), ymin = Q025, ymax = Q975),
+    fill = 'blue', alpha = 0.5
+  ) +
+  # Projections from filtered distribution
+  geom_line(data = mod3_V0_res$SQuants, aes(x = date, y = Q50), col = 'blue') +
+  geom_ribbon(
+    data = mod3_V0_res$SQuants,
+    aes(x = date, ymin = Q025, ymax = Q975),
+    fill = 'blue', alpha = 0.5
+  ) +
+  facet_wrap(~dep, scales = 'free_y', nrow = 2) +
+  theme_bw() +
+  theme(axis.title.x = element_blank()) +
+  ylab("Total Susceptible (%)")
+@
+\caption{\label{fig:h3Sus}Percentage of individuals that are in the susceptible compartment.
+Simulations from Model~3 under initial conditions are displayed in red; simulations from the filtering distribution of model are displayed in blue.}
+\end{figure}
 

### 2022-12-15 si/si.Rnw
re-estimating model3 in SI for reproducibility

@@
 
-# Set Run Level for debug, timing, and full computation
-RUN_LEVEL = 2
-
 # Create vectors for the unit and shared parameters
@@
 # TODO: replace with bake
-haiti3_orig_fit <- readRDS(save_file)
-
-# haiti3_orig_fit <- bake(
-#   file = save_file, {
-#     fit_coupled_haiti3(
-#       nsearches = N_SEARCHES,
-#       search1 = SEARCH1,
-#       search2 = SEARCH2,
-#       search3 = SEARCH3,
-#       search_rho = FALSE,
-#       search_gamma = FALSE,
-#       ncores = cores
-#     )
-#   }
-# )
+# haiti3_orig_fit <- readRDS(save_file)
+
+haiti3_orig_fit <- bake(
+  file = save_file, {
+    fit_coupled_haiti3(
+      nsearches = N_SEARCHES,
+      search1 = SEARCH1,
+      search2 = SEARCH2,
+      search3 = SEARCH3,
+      search_rho = FALSE,
+      search_gamma = FALSE,
+      ncores = cores
+    )
+  }
+)
 

### 2022-12-16 ms.Rnw
a few edits; consider removing elimiation from the title

@@
 
-\title{Informing policy via dynamic models: Eliminating cholera in Haiti}
-\runtitle{Eliminating Cholera in Haiti}
+\title{Informing policy via dynamic models: Cholera in Haiti}
+\runtitle{Cholera in Haiti}
 
@@
 As a case study, we consider a cholera epidemic in Haiti.
-The 2010 introduction of cholera to Haiti led to an extensive outbreak and sustained transmission until it was eliminated in 2019.
+The 2010 introduction of cholera to Haiti led to an extensive outbreak and sustained transmission.
+Cholera was eliminated in 2019, though it returned in 2022.
 We study three models developed by expert teams to advise on vaccination policies.
@@
 The groups largely adhered to existing guidelines on creating models to inform policy \citep{behrend20,saltelli20} and, despite their autonomy, obtained a consensus that an extensive nationwide vaccination campaign would be necessary to eliminate cholera from Haiti.
-Although several cholera cases have recently been detected in Haiti (\citep{world2010outbreak}),
-their conclusion is inconsistent with the prolonged period with no confirmed cholera cases between February, 2019 \citep{trevisin22} and October, 2022.
+Their conclusion is inconsistent with the prolonged period with no confirmed cholera cases between February, 2019 and September, 2022 \citep{trevisin22} though cholera has recently reemerged in Haiti \citep{rubin22}.
 
@@
 The \code{haitipkg} R package is maintained in a GitHub repository, \code{jeswheel/haitipkg}. The submitted version of this package will be archived on Zenodo.
-The package contains all of the data and code used to create and fit the models, as well as other useful functions that were used in this article.
+The package contains all of the data and code used to create, fit and explore the models presented in this article.
 }

### 2022-12-17 ms.Rnw
Changed table parameters, worked on reference and supp

--- 2022-12-15 ms.Rnw
@@
 The groups largely adhered to existing guidelines on creating models to inform policy \citep{behrend20,saltelli20} and, despite their autonomy, obtained a consensus that an extensive nationwide vaccination campaign would be necessary to eliminate cholera from Haiti.
-Although several cholera cases have recently been detected in Haiti (\citep{world2010outbreak}),
+Although several cholera cases have recently been detected in Haiti \citep{world2010outbreak,rubin22},
 their conclusion is inconsistent with the prolonged period with no confirmed cholera cases between February, 2019 \citep{trevisin22} and October, 2022.
@@
 A first-time reader may choose to skim through the rest of this section, and return later.
+Additional details about the numeric implementation of these models are provided in Secs.~S1 and S2 of the supplement.
 
@@
 Single and double vaccine doses were modeled by changing the waning of protection; protection was assumed to be equal between single and double dose until 52 weeks after vaccination, at which point the single dose becomes ineffective.
+
 The latent state vector $\bm{X}^{(1)}(t)$ is initialized by setting each compartment count for each vaccination scenario $\vaccCounter \neq 0$ as zero, and introducing initial-value parameters $I_{0,0}$ and $E_{0, 0}$ such that $R_{0}(0) = 0$, $I_{0}(0) = \mathrm{Pop} \times I_{0, 0}$, $E_{0}(0) = \mathrm{Pop} \times E_{0, 0}$ and $S_{0}(0) = \mathrm{Pop} \times (1 - I_{0, 0} - E_{0, 0})$, where $\mathrm{Pop}$ is the total population of Haiti.
 Reported cholera cases at time point $n$ ($\bm{Y}^{(1)}_n$) are assumed to come from a negative binomial measurement model, where only a fraction ($\reportRate$) of new weekly cases are reported.
-See Sec.~S4 of the supplement material for more details.
+See Sec.~S3 of the supplement material for more details.
 
@@
 Reported cases are assumed to come from a log-normal distribution, with the log-scale mean equal to the reporting rate $\reportRate$ times the number of newly infected individuals.
-See Sec.~\ref{sec:fit2} and Sec.~S4.2 of the supplement for more details.
+See Sec.~\ref{sec:fit2} and Sec.~S3.2 of the supplement for more details.
 
@@
 \end{equation}
-where $\tHur$ is the time Hurricane Matthew struck Haiti \jwc{TODO Cite}, and $1_{(A)}$ is the indicator function for event $A$.
+where $\tHur$ is the time Hurricane Matthew struck Haiti \citep{ferreirai16}, and $1_{(A)}$ is the indicator function for event $A$.
 In \cite{lee20}, $\Whur{_u}$ and $\hHur{u}$ were set to zero for all $u$;
-the need to account for the effect Hurricane Matthew had on cholera transmission for this model is explored in Sec. \jwc{TODO} of the supplement.
+the need to account for the effect Hurricane Matthew had on cholera transmission for this model is explored in Sec.~S5 of the supplement.
 
@@
 In reality, it's plausible that some bacteria or infected individuals were present in unit $u$ but simply went unreported.
-As such, for departments with zero case counts in week 1, we consider estimating the number of infected individuals rather then treating this value as a constant (see supplement Sec.
-\jwc{TODO} for more details).
+As such, for departments with zero case counts in week 1, we consider estimating the number of infected individuals rather then treating this value as a constant (see supplement Sec.~S5 for more details).
 Finally, reported cholera cases are assumed to come from a negative binomial measurement model with mean equal to a fraction ($\reportRate$) of individuals in each unit who develop symptoms seek healthcare
-(see Sec. \jwc{TODO} of the supplement).
+(see Sec.~S3.3 of the supplement).
 
@@
 and can be justified by asymptotic considerations in a large-population limit \citep{dadlani2020,ndii17}.
-Because the process model of Model~2 is deterministic, the parameter estimation problem for Model~2 reduces to a least squares calculation when combined with a Gaussian measurement model (see supplement material for details)
-\jwc{TODO: cite which part of supplement}.
+Because the process model of Model~2 is deterministic, the parameter estimation problem for Model~2 reduces to a least squares calculation when combined with a Gaussian measurement model (see Sec.~S3.2 of the supplement material for details)
 \cite{lee20} fit two versions of model~2 based on a presupposed change in cholera transmission from a epidemic phase to endemic phase that occurred in March, 2014.
@@
 Another approach is to compare a quantitative measure of the model fit (such as MSE, predictive accuracy, or model likelihood) among all proposed models.
-These comparisons, which provide insight into how each model performs relative to the others, are quite common \jwc{CITE SOME HAITI PAPERS}.
+These comparisons, which provide insight into how each model performs relative to the others, are quite common \citep{rinaldo12,sallah17}.
 To calibrate relative measures of fit, it is useful to compare against a model that has well-understood statistical ability to fit data, and we call this model a {\it benchmark}.
@@
 
-      lee3_results[["n_params"]] <- 29  # TODO: Check to make sure this is the correct number.
+      lee3_results[["n_params"]] <- 29
 
@@
 On this same subset, our model has a likelihood of $\Sexpr{myround(mod3_eval_res$subset_ll, 1)}$.
-See Sec.~S5 of the supplement material for more details on estimating the likelihood of the \cite{lee20} models.
+See Sec.~S6 of the supplement material for more details on estimating the likelihood of the \cite{lee20} models.
 }
@@
 Additional techniques include the inspection of conditional log-likelihoods of each observation given the previous observations in order to understand how well the model describes each data point.
-Other tools include plotting the effective sample size of each observation \citep{liu01}; plotting the values of the hidden states from simulations (see supplement Sec.\jwc{TODO});
+Other tools include plotting the effective sample size of each observation \citep{liu01}; plotting the values of the hidden states from simulations (supplement Sec.~5.1);
 and comparing summary statistics of the observed data to simulations from the model \citep{wood10,king15}.
@@
 
-Forecasts aim to provide an accurate estimate of the future state of a system based on currently available data, together with an assessment of uncertainty.
+Forecasts are an attempt to provide an accurate estimate of the future state of a system based on currently available data, together with an assessment of uncertainty.
 Mechanistic forecasting models, compatible with current scientific understanding, may also provide estimates of the future effects of potential interventions.
@@
 If a Bayesian technique is used for parameter estimation, a natural way to account for parameter uncertainty is to obtain simulations from the model where each simulation is obtained using parameters drawn from the estimated posterior distribution.
-For frequentist inference, one possible approach obtaining model forecasts from various values of $\theta$, and then weighing the resulting forecasts by their corresponding log-likelihoods \citep{king15}; see Sec.
-\jwc{TODO} of the supplement material for more details.
+For frequentist inference, one possible approach obtaining model forecasts from various values of $\theta$, where the values of $\theta$ are sampled proportionally their corresponding likelihoods \citep{king15}; see Sec.~S7 of the supplement material for more details.
 Both of these approaches share the similarity that parameters are chosen for the forecast approximately in proportion to their corresponding value of the likelihood function, $f^{(m)}_{\bm{Y}_{1:N}}(\bm{y}_{1:N}^*; \theta)$.
+In this analysis, our forecasts do not account for parameter uncertainty as our goal is to comment on the estimation and diagnosis of mechanistic models, rather than suggesting our results to be used to influence vaccination policy in Haiti.
+Furthermore, we use the projections from a single point estimate to highlight a deficiency of deterministic models, which is that the only variability in model projections is a result of parameter uncertainty, which leads to over-confidence in forecasts \cite{king15}.
 
@@
   & \eqref{model2:mu_IR}
-  & {\fixed $\muIR^{-1}=\Sexpr{myround(365/p3s["gamma"],1)}$ }
+  & {\fixed $\muIR^{-1}=\Sexpr{myround(365/p3["gamma1"],1)}$ }
   & \eqref{eq:model3:IR}
@@
 \hline
-Seasonality
+\multirow{2}{*}{Seasonality}
   & $\begin{array}{l}
@@
   & \eqref{model1:beta}
-  & $\begin{array}{l}
-     {\fixed \seasAmplitude=0.4} \\
-     \phaseParm = \Sexpr{myround(h2_fit$h2_params['phase'], 2)}
-     \end{array}$
+  & {\fixed \seasAmplitude=0.4}
   & \eqref{model2:lambda}
-  & $\begin{array}{l} \seasAmplitude=\Sexpr{myround(p3s["lambdaR"],2)} \\ r=\Sexpr{signif(p3s["r"],3)} \end{array}$
+  & $\seasAmplitude = \Sexpr{myround(p3s["lambdaR"],2)}$
+  & \eqref{eq:model3:water} \\
+  & $\transmissionTrend = \Sexpr{myround(p1["betat"],2)}$ [N]
+  & \eqref{model1:betat}
+  & $\phaseParm = \Sexpr{myround(h2_fit$h2_params['phase'], 2)}$ [N]
+  & \eqref{model2:lambda}
+  & $r = \Sexpr{signif(p3s["r"],3)}$
   & \eqref{eq:model3:water}
-\\
-\hline
+  \\\hline
 $\begin{array}{l}
@@
 \hline
-Vaccine efficacy
-  & \multicolumn{1}{c}{\hspace{16pt}\missing}
-  &
-  & $\begin{array}{l}
-      \hspace{-1mm} \fixed{\vaccineEfficacy_{1:4} = (
-        \Sexpr{myround(1 - h2_fit$h2_params["VE1"] * 0.4688, 2)},
-        \Sexpr{myround(1 - h2_fit$h2_params["VE2"] * 0.4688, 2)}} \\
-        \fixed{\Sexpr{myround(1 - h2_fit$h2_params["VE1"], 2)},
-        \Sexpr{myround(1 - h2_fit$h2_params["VE2"], 2)}})
-        \end{array}$
-  & \eqref{model2:mu_SE}
-  & \fixed{$\eta_{ud}(t)$}
-  &
-  \\
+% Vaccine efficacy
+%   & \multicolumn{1}{c}{\hspace{16pt}\missing}
+%   &
+%   & $\begin{array}{l}
+%       \hspace{-1mm} \fixed{\vaccineEfficacy_{1:4} = (
+%         \Sexpr{myround(1 - h2_fit$h2_params["VE1"] * 0.4688, 2)},
+%         \Sexpr{myround(1 - h2_fit$h2_params["VE2"] * 0.4688, 2)}} \\
+%         \fixed{\Sexpr{myround(1 - h2_fit$h2_params["VE1"], 2)},
+%         \Sexpr{myround(1 - h2_fit$h2_params["VE2"], 2)}})
+%         \end{array}$
+%   & \eqref{model2:mu_SE}
+%   & \fixed{$\eta_{ud}(t)$}
+%   &
+%   \\
 Birth/death (yr)
@@
       {\fixed \asymptomRelativeInfect =1 } \\
-      {\fixed \asymptomRelativeShed = \Sexpr{myround(p3s["XthetaA"],3)} }
+      \asymptomRelativeShed = \Sexpr{myround(p3s["XthetaA"],3)}
     \end{array}$
@@
 \end{array}$
-  & $\sigmaProc=(\Sexpr{myround(sqrt(p1["sig_sq_epi"]),2)}, \Sexpr{myround(sqrt(p1["sig_sq_end"]),2)})$
+  & $\sigmaProc=(\Sexpr{myround(sqrt(p1["sig_sq_epi"]),2)}, \Sexpr{myround(sqrt(p1["sig_sq_end"]),2)})$ [N]
   & \eqref{model1:lambda}
@@
   & $\reportRate=\Sexpr{myround(p1["rho"], 3)}$
-  & (S16) % TODO: Make sure manual equations match!
+  & (S15)
   & {\fixed $\reportRate=\Sexpr{myround(h2_fit$h2_params["Rho"],2)}$}
-  & (S17)
+  & (S16)
   & $\reportRate=\Sexpr{myround(p3s["epsilon"],2)}$
-  & (S18)
+  & (S19)
 \\
@@
 \text{Observation} \\
-\text{overdispersion}
+\text{variance}
 \end{array}$
-  & $\obsOverdispersion=(\Sexpr{round(p1["tau_epi"],2)}, \Sexpr{round(p1["tau_end"],2)})$
+  & $\obsOverdispersion = (\Sexpr{myround(p1["tau_epi"],2)}, \Sexpr{round(p1["tau_end"],2)})$
+  & (S15)
+  & $\obsOverdispersion = \Sexpr{signif(h2_fit$h2_params['sigma'], 3)}$
   & (S16)
-  &
-  &
-  & $\obsOverdispersion=\Sexpr{myround(p3s["k"],2)}$
-  & (S18)
+  & $\obsOverdispersion = \Sexpr{myround(p3s["k"],2)}$
+  & (S19)
 \\
 \hline
+\multirow{2}{*}{Initial Values} & $I_{0,0} = \Sexpr{round(prod(p1[c('pop_0', 'I_0')]))}$ & & \multicolumn{1}{c}{\multirow{2}{*}{\hspace{16pt}\missing}} & & \multirow{2}{*}{$I_{3,7,10;0}(0)= (
+        \Sexpr{round(prod(p3[c('H3', 'Iinit3')]))}, \Sexpr{round(prod(p3[c('H7', 'Iinit7')]))}, \Sexpr{round(prod(p3[c('H10', 'Iinit10')]))})$ [N]} & \\
+& $E_{0,0} = \Sexpr{round(prod(p1[c('pop_0', 'E_0')]))}$ & & \multicolumn{1}{c}{} & & & \\\hline
+Hurricane & \multicolumn{1}{c}{\multirow{2}{*}{\hspace{16pt}\missing}} & & \multicolumn{1}{c}{\multirow{2}{*}{\hspace{16pt}\missing}} & & $\Whur{3, 9} = (\Sexpr{myround(p3['aHur3'], 2)}, \Sexpr{myround(p3['aHur9'], 2)})$ [N] & \multirow{2}{*}{\eqref{eq:model3:foi}} \\
+Parameters & \multicolumn{1}{c}{} & & \multicolumn{1}{c}{} & & $\hHur{3, 9} = (\Sexpr{myround(p3['hHur3'], 2)}, \Sexpr{myround(p3['hHur9'], 2)})$ [N] & \\\hline
+
 \end{tabular}}

### 2022-12-17 si/si.Rnw
Changed table parameters, worked on reference and supp

@@
 \documentclass[11pt]{article}
-
-%%% JASA.
-Attention Mac Users.
-The latest version of the pdfTex compiler for Mac (pdfTeX-1.40.11) outputs files not currently supported by ScholarOne Manuscripts.
-To resolve this and have a PDF file that will convert properly, the PDF file will need to be distilled to Adobe version 1.4 or earlier.
-To create a PDF file that uploads correctly, place the command \pdfminorversion=4 in the preamble (before the begin{document}) of your LaTeX file and run pdfLaTeX again.
 
@@
 
-%Draft: \today
-
 \tableofcontents
@@
 
-\section{Initial Values}
-
-To perform inference on POMP models, it is necessary to propose an initial density for the latent process $f_{X_0}(x_0;\theta)$.
-This density is used to obtain initial values of the latent state when fitting and evaluating the model.
-For each of the models considered in this analysis, the initial conditions are derived by enforcing the model dynamics on reported cholera cases.
-It is also sometimes necessary to fit some initial value parameters in order to help determine starting values for weakly identifiable compartments.
-In the following subsections, we mention initial value parameters that were fit for each model.
-
-\subsection{Model~1}
-
-For this model, the number of individuals in the Recovered and Asymptomatic compartments are set to zero, but the initial proportion of Infected and Exposed individuals is estimated as initial value parameters ($I(0)$ and $E(0)$, respectively) using the MIF2 algorithm.
-Finally, the initial proportion of Susceptible individuals $S_0$ is calculated as $S(0) = 1 - I(0) - E(0)$.
-
-\subsection{Model~2}
-
-Model~2 assumes that the initial vales are known and constant, and so no initial value parameters need to be estimated.
-Starting values for latent state compartments are chosen so as to enforce the model dynamics on the observed number of cases.
-Specifically, the model sets $I_{u}(0) = y^*_{u}(0) / \reportRate$ for each unit $u \in 1:10$, where $I_{u}(t)$ is the number of infected individuals in unit $u$ at time $t$, $y^*_{u}(t)$ is the reported number of cases, and $\reportRate$ is the reporting rate.
-
-\subsection{Model~3}
-
-We use the reported cases at the start of the pandemic to approximate the number of Asymptomatic, Infectious, and Recovered individuals in each department $u \in 1:U$ using the same approximation as provided in Eq.~34.
-% TODO: make sure it aligns with \myeqref{eq:model3:approx}.
-The susceptible compartment is initialized so that the sum $S_{u}(0) + I_u(0) + A_u(0) + \sum_k R_{u, k}(0) = \text{population}_u$.
-The bacteria compartment is then initialized using Eq.~\myeqref{eq:model3:b0}:
-\begin{equation}
-  B_u(0) = \big[1 + \seasAmplitude \big(\Binit_u)^r \big] D_i \, \Wshed \big[ I_{u}(0)+ \asymptomRelativeShed A_{u}(0) \big]\label{eq:model3:b0},
-\end{equation}
-where $\Binit_u \in (0, 1)$ is an initial value parameter that allows some flexibility in determining the initial state of the bacteria compartment.
-
 \section{Measurement Models}
@@
 \begin{equation}
-  \label{model2:meas}
+  \label{model2:measEpi}
   Y_{u, n} \mid \Delta N_{EI}(n) = z(n) \sim \text{N}\left(\reportRate z(n), \obsOverdispersion^2 \right),
@@
 \begin{equation}
-  \label{model2:meas}
-  Y_{u, n} \mid \Delta N_{EI}(n) = z_1(n), \Delta N_{EA}(n) = z_2(n) \sim \text{N}\left(\reportRate (z_1(n) + z_2), \obsOverdispersion^2 \right),
+  \label{model2:measEnd}
+  Y_{u, n} \mid \Delta N_{EI}(n) = z_1(n), \Delta N_{EA}(n) = z_2(n) \sim \text{N}\left(\reportRate \big(z_1(n) + z_2(n)\big), \obsOverdispersion^2 \right),
 \end{equation}
@@
 The fitted values of the reporting rate---before and after January 2018---were $0.97$ and $0.097$, respectively.
-This major change in reporting rate alone could explain why Model~3 of \cite{lee20} failed to predict the eradication of cholera.
-An overnight change from near perfect to almost non-existent reporting forces the model to explain the observed reduction in reported cases as a decrease in the reporting of cases, rather than a decrease in the prevalence of cholera.
+An instantaneous change from near perfect to almost non-existent reporting can be problematic, as it forces the model to explain the observed reduction in reported cases as a decrease in the reporting of cases, rather than a decrease in the prevalence of cholera.
 This shift was justified by a "change of the case definition that occurred on January 1st, 2018";
@@
 We therefore do not allow a change in reporting rate when fitting Model~3.
+
+\section{Initial Values}
+
+To perform inference on POMP models, it is necessary to propose an initial density for the latent process $f_{X_0}(x_0;\theta)$.
+This density is used to obtain initial values of the latent state when fitting and evaluating the model.
+For each of the models considered in this analysis, the initial conditions are derived by enforcing the model dynamics on reported cholera cases.
+It is also sometimes necessary to fit some initial value parameters in order to help determine starting values for weakly identifiable compartments.
+In the following subsections, we mention initial value parameters that were fit for each model.
+
+\subsection{Model~1}
+
+For this model, the number of individuals in the Recovered and Asymptomatic compartments are set to zero, but the initial proportion of Infected and Exposed individuals is estimated as initial value parameters ($I(0)$ and $E(0)$, respectively) using the MIF2 algorithm.
+Finally, the initial proportion of Susceptible individuals $S_0$ is calculated as $S(0) = 1 - I(0) - E(0)$.
+
+\subsection{Model~2}
+
+Model~2 assumes that the initial vales are known and constant, and so no initial value parameters need to be estimated.
+Starting values for latent state compartments are chosen so as to enforce the model dynamics on the observed number of cases.
+Specifically, the model sets $I_{u}(0) = y^*_{u}(0) / \reportRate$ for each unit $u \in 1:10$, where $I_{u}(t)$ is the number of infected individuals in unit $u$ at time $t$, $y^*_{u}(t)$ is the reported number of cases, and $\reportRate$ is the reporting rate.
+
+\subsection{Model~3}
+
+As described in the main test, the latent states are initialized by estimating of the number of infected, asymptomatic, and recovered individuals at time $t_0$ by using observed case counts at time $t_1$.
+The initial value for the bacteria compartment is then determined using the rainfall dynamics of Eq.~23.
+For the departments of Grande'Anse, Nord-Ouest, and Sud-Est, estimating the latent states at time $t_0$ using the observed data at time $t_1$ is problematic (see Sec.~\ref{sec:mod3Cal}), so we instead estimate initial value parameters $I_{u0}(0)$ for the integers $u \in \{3, 7, 9\}$, which correspond to the aforementioned departments, and the remaining hidden states are estimated by enforcing model dynamics using this result.
 
@@
 
-\section{Lee et al. (2020) Replication}
-
-In this article we claimed that we were able to obtain better fits to the observed data using the same models that were proposed by \cite{lee20}.
-Along with visual comparisons to the data, this claim was supported by comparing likelihoods and AIC values in Table~1.
-Because model likelihoods were not provided by \cite{lee20}, it is necessary to replicate these models in order to obtain likelihood estimates.
-Here we would like to thank the authors of \cite{lee20}, who provided detailed descriptions of their models, which enabled us to build on their work.
-In the following subsections, we use our \code{R} package \code{haitipkg} to reproduce some of the results of \cite{lee20}.
-This reproduction allows us to estimate the likelihoods of the \cite{lee20} version of Models~1--3, and also provides a demonstration on reproducibility.
-
-\subsection{Model~1 Replication}
-
-The model was implemented by a team at Johns Hopkins Bloomberg School of Public Health (hereafter referred to as the Model~1 authors) in the \code{R} programming language using the \code{pomp} package \citep{king16}.
-Original source code is publicly available with DOI: 10.5281/zenodo.3360991.
-The final results reported by the Model~1 authors were obtained by using several different parameter sets rather than a single point estimate.
-According to the supplement materials, this was because model realizations from a single parameter set retained substantial variability, but multiple realizations from a collection of parameter sets resulted in a reasonable visual fit to the data.
-We are also inclined to believe that the use of multiple parameter values was in part intended to account for parameter uncertainty (as mentioned in our main text), an effort by the Model~1 authors that we applaud.
-Simulations from each of the parameter sets, however, were treated with equal importance when being used to diagnose the model fit and make inference on the system.
-This is problematic given Figures S8 and S9 of the supplement material, which suggest that some parameter sets that were used for inference were several hundred log-likelihood units lower than other parameter sets.
-Such a large difference in log-likelihoods is well beyond the threshold of statistical uncertainty determined by Wilks' theorem.
-
-To fully reproduce the results of the Model~1 authors, it is necessary to use the exact same set of model parameters.
-Because the parameters used to obtain the result given by \cite{lee20} are not publicly available, we relied on the source code provided by the Model~1 authors to approximately recreate the parameter set.
-Due to software updates since the publication of the source code, we were unable to produce the exact same set of parameters.
-Still, running the publicly available source code resulted in a set of parameters that are visually similar to those used by the Model~1 authors (See Figures~\ref{fig:PlotEpiDist} and \ref{fig:plotEndParams}).
-Furthermore, simulations using the set of parameters produced by the source code appear practically equivalent to those displayed by \cite{lee20} (See Figure~\ref{fig:plotMod1Sims}).
-
-<<Simulate Lee Model 1, echo=FALSE, cache=TRUE>>=
-h1_epi <- haiti1()
-N_LEE_SIMS <- 20
-
-registerDoRNG(18599687)
-foreach(
-  i = 1:length(h1LeeStartsEpi),
-  .combine = rbind
-) %dopar% {
-  guess <- h1LeeStartsEpi[i, ]
-  sims <- simulate(h1_epi, nsim = N_LEE_SIMS, format = 'data.frame',
-                   params = guess)  # 20 sims for each set of parameters
-  sims$param_set <- i
-  sims %>%
-    dplyr::select(.id, week, cases, param_set)
-} -> all_sims
-@
-
-<<PlotEpiDist, echo=FALSE, fig.cap="Compare to Figure S8 of Lee et al. (2020) supplement.", fig.width=7, fig.height=7, fig.align='center'>>=
-GGally::ggpairs(
-  h1LeeStartsEpi,
-  columns = c("rho", "tau", "beta1", "nu", "loglik")
-)
-@
-
-<<SimulateEpi, echo=FALSE, include=FALSE, message=FALSE, cache=TRUE>>=
-h1_end <- haiti1(period = "endemic")
-
-true_agg_cases <- haitiCholera %>%
-  select(-report) %>%
-  pivot_longer(
-    data = .,
-    cols = -date_saturday,
-    values_to = 'cases',
-    names_to = 'dep'
-  ) %>%
-  mutate(
-    date = as.Date(date_saturday),
-    dep = gsub("\\.", "_", dep)
-  ) %>%
-  mutate(
-    dep = case_when(dep == "Grand_Anse" ~ "Grande_Anse", TRUE ~ dep)
-  ) %>%
-  tidyr::pivot_wider(
-    data = .,
-    id_cols = c(date),
-    names_from = dep,
-    values_from = cases,
-    names_prefix = 'cases_'
-  ) %>%
-  mutate(
-    ReportedAll = cases_Artibonite + cases_Centre +
-      cases_Grande_Anse + cases_Nippes + cases_Nord +
-      cases_Nord_Est + cases_Ouest + cases_Sud +
-      cases_Sud_Est + cases_Nord_Ouest
-  )
-@
-
-<<FitEnd, echo=FALSE, message=FALSE, include=FALSE>>=
-foreach(
-  i = 1:nrow(h1LeeStartsEnd),
-  .combine = rbind
-) %dopar% {
-  r_params <- h1LeeStartsEnd[i, ][names(coef(h1_end))]
-  coef(h1_end) <- r_params
-
-  sims <- simulate(
-    h1_end, nsim = N_LEE_SIMS,
-    format = 'data.frame',
-    params = r_params
-  )
-  sims$param_set <- i
-  sims %>%
-    dplyr::select(.id, week, cases, param_set)
-} -> end_sims
-
-@
-
-<<plotEndParams, fig.width=7, fig.height=7, fig.align='center', fig.cap="Bivariate relationships between variables after fitting endemic period.
-Compare to S9 of Lee et al.
-(2020) supplement.", cache=TRUE, echo=FALSE>>=
-GGally::ggpairs(
-  h1LeeStartsEnd,
-  columns = c("rho", "tau", "beta1", "nu", "loglik")
-)
-@
-
-<<echo=FALSE, include=FALSE, message=FALSE>>=
-h1_epi_evals <- readRDS("../model1/run_level_3/lee1_epi_evals.rds")
-h1_end_evals <- readRDS("../model1/run_level_3/lee1_end_evals.rds")
-
-mod1_all_likes <- dplyr::inner_join(
-  x = h1_epi_evals,
-  y = h1_end_evals,
-  by = "parid"
-  ) %>%
-  dplyr::mutate(
-    joint_ll = pfLL.x + pfLL.y
-  )
-
-# Approximate the likelihood of the Lee et al (2020) version of Model 1 by
-# using the parameter sets that resulted in the maximum likelihood.
-mod1_lee_ll <- max(mod1_all_likes$joint_ll, na.rm = TRUE)
-@
-
-<<plotMod1Sims, echo=FALSE, fig.cap=paste0("Simulations using parameter sets that were generated by running source code provided by Lee et al (2020).
-Compare to Figure S7 of \\cite{lee20sup}.
-The upper bound for the likelihood of this model is ", round(mod1_lee_ll), "."), cache=TRUE, fig.width=6.5, fig.height=3.75>>=
-
-quants <- all_sims %>%
-  filter(!param_set %in% c(1, 2)) %>%  # Remove parameter sets inconsistent with Lee et al (2020).
-  group_by(week) %>%
-  summarize(
-    q025 = quantile(cases, probs = 0.025, na.rm = TRUE),
-    q50  = quantile(cases, probs = 0.500, na.rm = TRUE),
-    q975 = quantile(cases, probs = 0.975, na.rm = TRUE)
-  ) %>%
-  ungroup() %>%
-  mutate(date = lubridate::ymd("2010-10-16") + lubridate::weeks(week))
-
-quants_end <- end_sims %>%
-  group_by(week) %>%
-  summarize(
-    q025 = quantile(cases, probs = 0.025, na.rm = TRUE),
-    q50  = quantile(cases, probs = 0.500, na.rm = TRUE),
-    q975 = quantile(cases, probs = 0.975, na.rm = TRUE)
-  ) %>%
-  ungroup() %>%
-  mutate(date = lubridate::ymd("2010-10-16") + lubridate::weeks(week))
-
-ggplot() +
-  geom_point(data = true_agg_cases, aes(x = date, y = ReportedAll)) +
-  geom_line(data = quants, aes(x = date, y = q50), col = '#0b8a2b', size = 1) +
-  geom_ribbon(
-    data = quants,
-    aes(x = date, ymin = q025, ymax = q975),
-    alpha = 0.5, fill = '#0b9c30'
-  ) +
-  geom_line(data = quants_end, aes(x = date, y = q50), col = '#882871') +
-  geom_ribbon(data = quants_end, aes(x = date, ymin = q025, ymax = q975), alpha = 0.5, fill = '#882871') +
-  ylab("Reported, Symptomatic Incidence") +
-  scale_x_date(date_breaks = "1 year") +
-  theme(axis.title.x = element_blank(),
-        axis.text.x = element_text(hjust = 1, angle = 45))
-@
-
-To calculate an upper-bound on the likelihood of this model---which uses a large number of parameters that each have a different likelihood value---we estimate the likelihood corresponding to each parameter set using a particle filter, and report the maximum of these likelihoods.
-
-\subsection{Model~2 Replication}\label{sec:mod2rep}
-
-Model~2 was developed by a team that consisted of members from the Fred Hutchinson Cancer Research Center and the University of Florida (hereafter referred to as the Model~2 authors).
-While Model~2 is the only deterministic model we considered in our analysis, it contains perhaps the most complex descriptions of cholera in Haiti: Model~2 accounts for movement between spatial units; human-to-human and environment-to-human cholera infections; and transfer of water between spatial units based on elevation charts and river flows.
-
-The source code that the Model~2 authors used to generate their results was written in the \code{Python} programming language, and is publicly available at \url{10.5281/zenodo.3360857} and its accompanying GitHub repository \url{https://github.com/lulelita/HaitiCholeraMultiModelingProject}.
-In order to perform our analysis in a unified framework, we re-implemented this model in the \code{R} programming language using the \code{spatPomp} package \citep{asfaw21github}, which facilitates the creation of meta-population models.
-We note that the travel and water matrices used to implement the complex dynamics in Model~2 \citep{lee20sup} are not available in either the Zenodo archive or the GitHub repository;
-instead, we obtained these matrices via personal correspondence with the Model~2 authors.
-Using these matrices, and the point estimates for model parameters provided by \citep{lee20sup}, we created trajectories of the cholera dynamics and compared this to available data.
-These trajectories, shown in Figure~\ref{fig:mod2rep}, are very similar to the trajectories shown in Figure~S15 of \cite{lee20sup}.
-
-<<Model2Replication, cache=TRUE, echo=FALSE>>=
-h2_epi <- haiti2()
-h2_end <- haiti2(region = 'after')
-
-h2_epi_traj <- trajectory(h2_epi, format = 'data.frame')
-h2_end_traj <- trajectory(h2_end, format = 'data.frame')
-
-h2_epi_traj$Ctotal <- rowSums(h2_epi_traj[, paste0("C", 1:10)]) * 0.2
-h2_end_traj$Ctotal <- rowSums(h2_end_traj[, paste0("C", 1:10)]) * 0.2
-
-dep_plot_df <- haitipkg::haitiCholera %>%
-  select(-report) %>%
-  tidyr::pivot_longer(
-    data = .,
-    cols = -date_saturday,
-    values_to = 'cases',
-    names_to = 'dep'
-  ) %>%
-  mutate(
-    date = as.Date(date_saturday),
-    dep = gsub("\\.", "_", dep)
-  ) %>%
-  mutate(
-    dep = case_when(dep == "Grand_Anse" ~ "Grande_Anse", TRUE ~ dep)
-  )
-
-true_agg_cases <- dep_plot_df %>%
-  tidyr::pivot_wider(
-    data = .,
-    id_cols = c(date),
-    names_from = dep,
-    values_from = cases,
-    names_prefix = 'cases_'
-  ) %>%
-  mutate(
-    ReportedAll = cases_Artibonite + cases_Centre +
-      cases_Grande_Anse + cases_Nippes + cases_Nord +
-      cases_Nord_Est + cases_Ouest + cases_Sud +
-      cases_Sud_Est + cases_Nord_Ouest
-  ) %>%
-  select(date, ReportedAll)
-
-h2_epi_traj$date <- as.Date(
-  lubridate::round_date(
-    lubridate::date_decimal(h2_epi_traj$year),
-    unit = 'day'
-  )
-)
-
-h2_end_traj$date <- as.Date(
-  lubridate::round_date(
-    lubridate::date_decimal(h2_end_traj$year),
-    unit = 'day'
-  )
-)
-
-all_traj <- bind_rows(
-  select(h2_epi_traj, date, Ctotal),
-  select(h2_end_traj, date, Ctotal)
-)
-
-projections <- left_join(
-  x = true_agg_cases,
-  y = all_traj,
-  by = 'date'
-)
-@
-
-\begin{figure}[ht]
-<<plotModel2Rep, cache=TRUE, echo=FALSE, fig.width=6.5, fig.height=4.5>>=
-gg_epi <- projections %>%
-  filter(date < as.Date("03-01-2014", format = '%m-%d-%Y')) %>%
-  ggplot(aes(x = date)) +
-  geom_line(aes(y = ReportedAll), col = 'black', linetype = 'dashed') +
-  geom_line(aes(y = Ctotal), col = 'red', size = 1) +
-  theme_bw() +
-  scale_y_continuous(limits = c(0, 30000), breaks = seq(0, 30000, 5000)) +
-  theme(axis.title.x = element_blank()) +
-  ylab("Reported number of cases")
-
-gg_end <- projections %>%
-  filter(date >= as.Date("03-01-2014", format = '%m-%d-%Y')) %>%
-  ggplot(aes(x = date)) +
-  geom_line(aes(y = ReportedAll), col = 'black', linetype = 'dashed') +
-  geom_line(aes(y = Ctotal), col = 'red', size = 1) +
-  theme_bw() +
-  scale_y_continuous(limits = c(0, 7000), breaks = seq(0, 7000, 1000)) +
-  theme(axis.title.x = element_blank()) +
-  ylab("Reported number of cases") +
-  scale_x_date(date_breaks = "1 year", date_labels = "%Y")
-
-cowplot::plot_grid(gg_epi, gg_end, ncol = 1)
-@
-\caption{\label{fig:mod2rep}
-Model 2 trajectories using the \code{haitipkg}. Compare to Figure S15 of \cite{lee20sup}.
-}
-\end{figure}
-
-There are minor differences between Figure~\ref{fig:mod2rep} and Figure~S15 of \cite{lee20sup}.
-% such as a delayed second wave in the epidemic period and a larger initial wave during the endemic period.
-While the discrepancy appears minor, the deterministic nature of Model~2 implies that an exact replication of model trajectories should be possible.
-In this case, these discrepancies may be attributed to model implementation and small differences in model parameters.
-Details about the measurement models and how latent states were initialized for the epidemic model were not provided by \citet{lee20sup} and therefore these details must be inferred by looking at the provided source code.
-According to repository comments, the files \code{fit\allowbreak In\allowbreak Pieces\allowbreak 3params\allowbreak Clean\allowbreak May2019\allowbreak Public.py} and \code{fit\allowbreak In\allowbreak Pieces\allowbreak Mu\allowbreak With\allowbreak Frac\allowbreak Sus\allowbreak Fixed\allowbreak All\allowbreak Infections\allowbreak Public.py} were used to fit the epidemic and endemic phases of the model respectively, although it is apparent that these exact files were not used to obtain the reported results.
-These files contain errors that making them impossible to run, and therefore these files cannot be the code that generated the reported model fits\footnote{One example of why the code cannot be run that the file loads functions from a non-extant file named \code{choleraEqs.py} in line 13 rather than \code{cholera\allowbreak Eqs\allowbreak Public.py}.}.
-The inability to replicate the results by \citet{lee20} by running the provided source code makes checking whether or not a our numeric implementation faithfully represents their results very difficult.
-% The difficulty of reproducing the result of \citet{lee20} is coupled with a misunderstanding of how these results were originally obtained.
-
-\jwc{PROBABLY DELETE THIS NEXT BIT, BUT IT'S USEFUL TO DOCUMENT IT FOR NOW...}
-One concern that arises when inspecting the files \code{fit\allowbreak In\allowbreak Pieces3\allowbreak params\allowbreak Clean\allowbreak May\allowbreak 2019\allowbreak Public.py} and \code{fit\allowbreak In\allowbreak Pieces\allowbreak Mu\allowbreak With\allowbreak Frac\allowbreak Sus\allowbreak Fixed\allowbreak All\allowbreak Infections\allowbreak Public.py} is that the output of these files (which contains the estimated parameters) was not referenced/used in either the file that generates the sensitivity analysis or the file that generates the model simulations; instead, alternative (non-extant) files with similar names are loaded, suggesting parameters that were used to generate the final results were not a direct result of the estimated parameters.
-Furthermore, inspection of the script \code{simulate\allowbreak Haiti\allowbreak Full\allowbreak Time\allowbreak In\allowbreak Pieces\allowbreak All\allowbreak Vaccination\allowbreak Scenarios\allowbreak Single\allowbreak Parameter\allowbreak Vector\allowbreak All\allowbreak Infections\allowbreak Public.py}---which is used to obtain simulations for each of the vaccination scenarios---results in a number of concerns.
-First, the script uses parameters from the pickle file \code{least\allowbreak Squares\allowbreak Data\allowbreak UpTo\allowbreak 2019\allowbreak /fit\allowbreak Epidemic\allowbreak Piece\_results\allowbreak Matrix\_18\allowbreak Mar\allowbreak 2019.pickle}, which only appears in the repository as a commented out line of code in the file \code{generate\allowbreak Sets\allowbreak Of\allowbreak Parameters\allowbreak For\allowbreak Sensitivity\allowbreak Analysis\allowbreak Public.py}, suggesting that the set of parameters used to simulate each vaccination scenario was not the most recent set of parameters generated by the sensitivity analysis.
-Second, the script appears to use the set of differential equations from the function \code{cholera\allowbreak Eqs11\allowbreak With\allowbreak Vaccination\allowbreak Network} for the epidemic phase, even though parameter values for this phase were estimated using \code{cholera\allowbreak Eqs10\allowbreak With\allowbreak Vaccination\allowbreak Network}.
-It is possible, however, that we simply are interpreting the code incorrectly and that these concerns are unwarranted.
-Furthermore, because the provided code is non-functional, it is possible that the code that was actually used to obtain the reported results does not include these potential errors.
-This highlights the importance of providing the exact source code that was used to obtain the reported results of any analysis.
-
-Another potential explanation for the discrepancy between our simulations from Model~2 and those of \citet{lee20} is that the parameters that we used are only approximately the same as those used by \citet{lee20sup}.
-For example, the parameters $\transmission$, $\Wbeta{}$ (See Table~\ref{tab:translate}) had reported values of $9.9 \times 10^{-7}$ and $4.03 \times 10^{-2}$, respectively (Table~S13 of \cite{lee20sup}), but were actually fit to data and therefore likely had values that were far more precise.
-Additionally, our implementation of Model~2 used a time scale of years and many of the parameters were reported on a weekly scale, so small differences may result due to unit conversions.
-The collective effect of these small differences in model parameters likely will result in differences in model trajectories.
-This issue could have been mitigated if the saved files containing estimated model parameters were included as part of the code submission.
-
-\subsection{Model~3 Replication}
-
-Model~3 was developed by a team of researchers at the Laboratory of the Swiss Federal Institute of Technology in Lausanne, hereafter referred to as the Model~3 authors.
-The code that was originally used to implement Model~3 is archived with the DOI: \url{10.5281/zenodo.3360723}, and also available in the public GitHub repository: \code{jcblemai/haiti-mass-ocv-campaign}.
-Because the code was made publicly available, and final model parameters were reported in the supplementary material of \cite{lee20}, we were able to reproduce Model~3 by directly using the source code.
-In Fig.~\ref{fig:mod3rep}, we plot simulations from this model.
-This figure should be compared to Figure S18 of \cite{lee20}.
-We note that slight differences may be accounted for due to variance in the model simulations and the difference in programming language used to produce the figure.
-
-<<Model3Replication, echo=FALSE, cache=TRUE>>=
-NSIM <- 500
-
-plot_order <- c(
-  'Artibonite',
-  'Sud_Est',
-  'Nippes',
-  'Nord_Est',
-  'Ouest',
-  'Centre',
-  'Nord',
-  'Sud',
-  'Nord_Ouest',
-  'Grande_Anse'
-)
-
-dep_labeller <- as_labeller(
-  c(
-    'Artibonite' = 'Artibonite',
-    'Sud_Est' = 'Sud-Est',
-    'Sud.Est' = 'Sud-Est',
-    'Nippes' = 'Nippes',
-    'Nord_Est' = 'Nord-Est',
-    'Nord.Est' = 'Nord-Est',
-    'Ouest' = 'Ouest',
-    'Centre' = 'Centre',
-    'Nord' = 'Nord',
-    'Sud' = 'Sud',
-    'Nord_Ouest' = 'Nord-Ouest',
-    'Nord.Ouest' = 'Nord-Ouest',
-    'Grande_Anse' = 'Grand\'Anse',
-    'Grand.Anse' = 'Grand\'Anse'
-  )
-)
-
-lee3 <- lee3_spatPomp()
-lee3_sims <- bake("../model3/si_lee3Sims.rds", {
-  simulate(lee3, nsim = NSIM, seed = 321, format = 'data.frame')
-})
-
-sub_cases <- haitiCholera %>%
-  dplyr::mutate(date = as.Date(date_saturday)) %>%
-  dplyr::select(-report, -date_saturday) %>%
-  dplyr::filter(lubridate::decimal_date(date) >= min(lee3_sims$time)) %>%
-  tidyr::pivot_longer(
-    data = .,
-    cols = -date,
-    names_to = "dep",
-    values_to = "cases"
-  ) %>%
-  dplyr::mutate(
-    dep = dplyr::case_when(
-      dep == "Grand.Anse" ~ "Grande_Anse",
-      dep == "Nord.Est" ~ "Nord_Est",
-      dep == "Nord.Ouest" ~ "Nord_Ouest",
-      dep == "Sud.Est" ~ "Sud_Est",
-      TRUE ~ dep
-    )
-  )
-
-quants <- lee3_sims %>%
-  dplyr::group_by(unitname, time) %>%
-  dplyr::summarise(
-    q05 = quantile(cases, 0.025, na.rm = T),
-    mean = mean(cases, na.rm = T),
-    q50 = quantile(cases, 0.5, na.rm = T),
-    q95 = quantile(cases, 0.975, na.rm = T)
-  ) %>%
-  dplyr::ungroup() %>%
-  dplyr::mutate(
-    date = lubridate::date_decimal(time),
-    date = as.Date(lubridate::round_date(date))
-  ) %>%
-  dplyr::rename(dep = unitname)
-
-quants <- quants %>%
-  dplyr::filter(
-    !(dep == "Ouest" & date < '2017-06-10')
-  )
-@
-
-\begin{figure}[ht]
-<<PlotMod3Rep, echo=FALSE, message=FALSE, fig.height=3.75>>=
-ggplot() +
-  geom_ribbon(data = quants, aes(x = date, ymin = q05, ymax = q95),
-              fill = 'darkblue', alpha = 0.5) +
-  geom_line(data = quants, aes(x = date, y = q50), col = 'darkblue') +
-  geom_point(data = dplyr::filter(sub_cases, cases < 500), aes(x = date, y = cases), size = 0.2) +
-  facet_wrap(~factor(dep, levels = plot_order), nrow = 2, labeller = dep_labeller) +
-  labs(y = 'Reported Cholera Cases') +
-  scale_x_date(date_breaks = "6 months", date_labels = "%Y-%m") +
-  theme(
-    axis.text.x = element_text(angle = 40, hjust = 1, vjust = 1, size = 7),
-    axis.title.x = element_blank()
-  )
-@
-\caption{\label{fig:mod3rep}
-Simulations from Model~3.
-Compare to Figure S18 of \cite{lee20}.
-}
-\end{figure}
-
-<<lee3Cleanup, echo=FALSE, include=FALSE, message=FALSE>>=
-rm(lee3, lee3_sims, quants, sub_cases)
-gc()
-@
-
-\section{Calibrating Model~3 to observed cases}
+\section{Calibrating Model~3 to observed cases}\label{sec:mod3Cal}
 
@@
 
+# TODO: Move to run_level_3 folder...
+
 save_file <- "../model3/run_level_2/fit_orig_model.rds"
-
-# TODO: replace with bake
-# haiti3_orig_fit <- readRDS(save_file)
 
@@
 Further investigation of the model output reveals that the model is struggling to explain the sudden surge in cholera cases that occurred at this time, which coincides with the time that Hurricane Matthew struck Haiti.
-While the model does include a mechanism to account for increased cholera transmission due to large rainfall events, the mechanism does not appear to be sufficient to capture the damaging effects of the hurricane, which had the greatest impact in the the Sud and Grand'Anse departments \jwc{TODO: CITE}.
+While the model does include a mechanism to account for increased cholera transmission due to large rainfall events, the mechanism does not appear to be sufficient to capture the damaging effects of the hurricane, which had the greatest impact in the the Sud and Grand'Anse departments \citep{ferreirai16}.
 This result led us to include parameters $\Whur{u}$ and $\hHur{u}$ in Eq.~23 that would increase the transmission between environmental cholera and humans for Sud and Grand'Anse.
@@
   theme_bw() +
-  theme(axis.title.x = element_blank()) +
+  theme(axis.title.x = element_blank(),
+        axis.text.x = element_text(angle = -45, hjust = 0)) +
   ylab("Conditional log-likelihoods")
@@
 In reality, however, it is possible to have some infected or asymptomatic individuals in a department when no cases are recorded due to under-reporting.
-This may be particularly true at the onset of an infectious disease outbreak, as lower public awareness and inadequate testing infrastructure may result in many cases going undetected \jwc{TODO: CITE}.
+This may be particularly true at the onset of an infectious disease outbreak, as lower public awareness and inadequate testing infrastructure may result in many cases going undetected.
 We therefore include and estimate $I_{0u}$ that determines the initial number of Infected individuals at the start of the pandemic for the departments Grande'Anse, Nord-Ouest, and Sud-Est;
@@
 This also appears true in later stages of the time period of observed data, where the data suggests that there is a replenishment of the susceptible compartment, but the Model simulations typically retain a smaller proportion of susceptible individuals.
-This result demonstrates that the calibrated model favors a more rapid growth in cholera cases than what is typically seen in the observed data, which explains why the model fails to predict the absence of cases between February, 2019 and October, 2022
+This result demonstrates that the calibrated model favors a more rapid growth in cholera cases than what is typically seen in the observed data, which explains why the model fails to predict the absence of cases between February, 2019 and October, 2022.
 
@@
 Sus_df <- bake(
-  file = "../model3/run_level_3/si_S_quants", {
+  file = "../model3/run_level_3/si_S_quants.rds", {
     foreach(i=1:length(h3@times), .combine = bind_rows) %do% {
@@
   ) +
+  geom_vline(xintercept = max(h3Spat_quants$date), linetype = 'dashed') +
   facet_wrap(~dep, scales = 'free_y', nrow = 2) +
   theme_bw() +
+  theme(axis.title.x = element_blank(),
+        axis.text.x = element_text(angle = -45, hjust = 0)) +
+  ylab("Total Susceptible (%)")
+@
+\caption{\label{fig:h3Sus}Percentage of individuals that are in the susceptible compartment.
+Simulations from Model~3 under initial conditions are displayed in red; simulations from the filtering distribution of model are displayed in blue.
+The dashed line represents the end of the observed data.}
+\end{figure}
+
+\section{Lee et al. (2020) Replication}
+
+In this article we claimed that we were able to obtain better fits to the observed data using the same models that were proposed by \cite{lee20}.
+Along with visual comparisons to the data, this claim was supported by comparing likelihoods and AIC values in Table~1.
+Because model likelihoods were not provided by \cite{lee20}, it is necessary to replicate these models in order to obtain likelihood estimates.
+Here we would like to thank the authors of \cite{lee20}, who provided detailed descriptions of their models, which enabled us to build on their work.
+In the following subsections, we use our \code{R} package \code{haitipkg} to reproduce some of the results of \cite{lee20}.
+This reproduction allows us to estimate the likelihoods of the \cite{lee20} version of Models~1--3, and also provides a demonstration on reproducibility.
+
+\subsection{Model~1 Replication}
+
+The model was implemented by a team at Johns Hopkins Bloomberg School of Public Health (hereafter referred to as the Model~1 authors) in the \code{R} programming language using the \code{pomp} package \citep{king16}.
+Original source code is publicly available with DOI: 10.5281/zenodo.3360991.
+The final results reported by the Model~1 authors were obtained by using several different parameter sets rather than a single point estimate.
+According to the supplement materials, this was because model realizations from a single parameter set retained substantial variability, but multiple realizations from a collection of parameter sets resulted in a reasonable visual fit to the data.
+We are also inclined to believe that the use of multiple parameter values was in part intended to account for parameter uncertainty (as mentioned in our main text), an effort by the Model~1 authors that we applaud.
+Simulations from each of the parameter sets, however, were treated with equal importance when being used to diagnose the model fit and make inference on the system.
+This is problematic given Figures S8 and S9 of the supplement material, which suggest that some parameter sets that were used for inference were several hundred log-likelihood units lower than other parameter sets.
+Such a large difference in log-likelihoods is well beyond the threshold of statistical uncertainty determined by Wilks' theorem.
+
+To fully reproduce the results of the Model~1 authors, it is necessary to use the exact same set of model parameters.
+Because the parameters used to obtain the result given by \cite{lee20} are not publicly available, we relied on the source code provided by the Model~1 authors to approximately recreate the parameter set.
+Due to software updates since the publication of the source code, we were unable to produce the exact same set of parameters.
+Still, running the publicly available source code resulted in a set of parameters that are visually similar to those used by the Model~1 authors (See Figures~\ref{fig:PlotEpiDist} and \ref{fig:plotEndParams}).
+Furthermore, simulations using the set of parameters produced by the source code appear practically equivalent to those displayed by \cite{lee20} (See Figure~\ref{fig:plotMod1Sims}).
+
+<<Simulate Lee Model 1, echo=FALSE, cache=TRUE>>=
+h1_epi <- haiti1()
+N_LEE_SIMS <- 20
+
+registerDoRNG(18599687)
+foreach(
+  i = 1:length(h1LeeStartsEpi),
+  .combine = rbind
+) %dopar% {
+  guess <- h1LeeStartsEpi[i, ]
+  sims <- simulate(h1_epi, nsim = N_LEE_SIMS, format = 'data.frame',
+                   params = guess)  # 20 sims for each set of parameters
+  sims$param_set <- i
+  sims %>%
+    dplyr::select(.id, week, cases, param_set)
+} -> all_sims
+@
+
+<<PlotEpiDist, echo=FALSE, fig.cap="Compare to Figure S8 of Lee et al. (2020) supplement.", fig.width=7, fig.height=7, fig.align='center'>>=
+GGally::ggpairs(
+  h1LeeStartsEpi,
+  columns = c("rho", "tau", "beta1", "nu", "loglik")
+)
+@
+
+<<SimulateEpi, echo=FALSE, include=FALSE, message=FALSE, cache=TRUE>>=
+h1_end <- haiti1(period = "endemic")
+
+true_agg_cases <- haitiCholera %>%
+  select(-report) %>%
+  pivot_longer(
+    data = .,
+    cols = -date_saturday,
+    values_to = 'cases',
+    names_to = 'dep'
+  ) %>%
+  mutate(
+    date = as.Date(date_saturday),
+    dep = gsub("\\.", "_", dep)
+  ) %>%
+  mutate(
+    dep = case_when(dep == "Grand_Anse" ~ "Grande_Anse", TRUE ~ dep)
+  ) %>%
+  tidyr::pivot_wider(
+    data = .,
+    id_cols = c(date),
+    names_from = dep,
+    values_from = cases,
+    names_prefix = 'cases_'
+  ) %>%
+  mutate(
+    ReportedAll = cases_Artibonite + cases_Centre +
+      cases_Grande_Anse + cases_Nippes + cases_Nord +
+      cases_Nord_Est + cases_Ouest + cases_Sud +
+      cases_Sud_Est + cases_Nord_Ouest
+  )
+@
+
+<<FitEnd, echo=FALSE, message=FALSE, include=FALSE>>=
+foreach(
+  i = 1:nrow(h1LeeStartsEnd),
+  .combine = rbind
+) %dopar% {
+  r_params <- h1LeeStartsEnd[i, ][names(coef(h1_end))]
+  coef(h1_end) <- r_params
+
+  sims <- simulate(
+    h1_end, nsim = N_LEE_SIMS,
+    format = 'data.frame',
+    params = r_params
+  )
+  sims$param_set <- i
+  sims %>%
+    dplyr::select(.id, week, cases, param_set)
+} -> end_sims
+
+@
+
+<<plotEndParams, fig.width=7, fig.height=7, fig.align='center', fig.cap="Bivariate relationships between variables after fitting endemic period.
+Compare to S9 of Lee et al.
+(2020) supplement.", cache=TRUE, echo=FALSE>>=
+GGally::ggpairs(
+  h1LeeStartsEnd,
+  columns = c("rho", "tau", "beta1", "nu", "loglik")
+)
+@
+
+<<echo=FALSE, include=FALSE, message=FALSE>>=
+h1_epi_evals <- readRDS("../model1/run_level_3/lee1_epi_evals.rds")
+h1_end_evals <- readRDS("../model1/run_level_3/lee1_end_evals.rds")
+
+mod1_all_likes <- dplyr::inner_join(
+  x = h1_epi_evals,
+  y = h1_end_evals,
+  by = "parid"
+  ) %>%
+  dplyr::mutate(
+    joint_ll = pfLL.x + pfLL.y
+  )
+
+# Approximate the likelihood of the Lee et al (2020) version of Model 1 by
+# using the parameter sets that resulted in the maximum likelihood.
+mod1_lee_ll <- max(mod1_all_likes$joint_ll, na.rm = TRUE)
+@
+
+<<plotMod1Sims, echo=FALSE, fig.cap=paste0("Simulations using parameter sets that were generated by running source code provided by Lee et al (2020).
+Compare to Figure S7 of \\cite{lee20sup}.
+The upper bound for the likelihood of this model is ", round(mod1_lee_ll), "."), cache=TRUE, fig.width=6.5, fig.height=3.75>>=
+
+quants <- all_sims %>%
+  filter(!param_set %in% c(1, 2)) %>%  # Remove parameter sets inconsistent with Lee et al (2020).
+  group_by(week) %>%
+  summarize(
+    q025 = quantile(cases, probs = 0.025, na.rm = TRUE),
+    q50  = quantile(cases, probs = 0.500, na.rm = TRUE),
+    q975 = quantile(cases, probs = 0.975, na.rm = TRUE)
+  ) %>%
+  ungroup() %>%
+  mutate(date = lubridate::ymd("2010-10-16") + lubridate::weeks(week))
+
+quants_end <- end_sims %>%
+  group_by(week) %>%
+  summarize(
+    q025 = quantile(cases, probs = 0.025, na.rm = TRUE),
+    q50  = quantile(cases, probs = 0.500, na.rm = TRUE),
+    q975 = quantile(cases, probs = 0.975, na.rm = TRUE)
+  ) %>%
+  ungroup() %>%
+  mutate(date = lubridate::ymd("2010-10-16") + lubridate::weeks(week))
+
+ggplot() +
+  geom_point(data = true_agg_cases, aes(x = date, y = ReportedAll)) +
+  geom_line(data = quants, aes(x = date, y = q50), col = '#0b8a2b', size = 1) +
+  geom_ribbon(
+    data = quants,
+    aes(x = date, ymin = q025, ymax = q975),
+    alpha = 0.5, fill = '#0b9c30'
+  ) +
+  geom_line(data = quants_end, aes(x = date, y = q50), col = '#882871') +
+  geom_ribbon(data = quants_end, aes(x = date, ymin = q025, ymax = q975), alpha = 0.5, fill = '#882871') +
+  ylab("Reported, Symptomatic Incidence") +
+  scale_x_date(date_breaks = "1 year") +
+  theme(axis.title.x = element_blank(),
+        axis.text.x = element_text(hjust = 1, angle = 45))
+@
+
+To calculate an upper-bound on the likelihood of this model---which uses a large number of parameters that each have a different likelihood value---we estimate the likelihood corresponding to each parameter set using a particle filter, and report the maximum of these likelihoods.
+
+\subsection{Model~2 Replication}\label{sec:mod2rep}
+
+Model~2 was developed by a team that consisted of members from the Fred Hutchinson Cancer Research Center and the University of Florida (hereafter referred to as the Model~2 authors).
+While Model~2 is the only deterministic model we considered in our analysis, it contains perhaps the most complex descriptions of cholera in Haiti: Model~2 accounts for movement between spatial units; human-to-human and environment-to-human cholera infections; and transfer of water between spatial units based on elevation charts and river flows.
+
+The source code that the Model~2 authors used to generate their results was written in the \code{Python} programming language, and is publicly available at \url{10.5281/zenodo.3360857} and its accompanying GitHub repository \url{https://github.com/lulelita/HaitiCholeraMultiModelingProject}.
+In order to perform our analysis in a unified framework, we re-implemented this model in the \code{R} programming language using the \code{spatPomp} package \citep{asfaw21github}, which facilitates the creation of meta-population models.
+We note that the travel and water matrices used to implement the complex dynamics in Model~2 \citep{lee20sup} are not available in either the Zenodo archive or the GitHub repository;
+instead, we obtained these matrices via personal correspondence with the Model~2 authors.
+Using these matrices, and the point estimates for model parameters provided by \citep{lee20sup}, we created trajectories of the cholera dynamics and compared this to available data.
+These trajectories, shown in Figure~\ref{fig:mod2rep}, are very similar to the trajectories shown in Figure~S15 of \cite{lee20sup}.
+
+<<Model2Replication, cache=TRUE, echo=FALSE>>=
+h2_epi <- haiti2()
+h2_end <- haiti2(region = 'after')
+
+h2_epi_traj <- trajectory(h2_epi, format = 'data.frame')
+h2_end_traj <- trajectory(h2_end, format = 'data.frame')
+
+h2_epi_traj$Ctotal <- rowSums(h2_epi_traj[, paste0("C", 1:10)]) * 0.2
+h2_end_traj$Ctotal <- rowSums(h2_end_traj[, paste0("C", 1:10)]) * 0.2
+
+dep_plot_df <- haitipkg::haitiCholera %>%
+  select(-report) %>%
+  tidyr::pivot_longer(
+    data = .,
+    cols = -date_saturday,
+    values_to = 'cases',
+    names_to = 'dep'
+  ) %>%
+  mutate(
+    date = as.Date(date_saturday),
+    dep = gsub("\\.", "_", dep)
+  ) %>%
+  mutate(
+    dep = case_when(dep == "Grand_Anse" ~ "Grande_Anse", TRUE ~ dep)
+  )
+
+true_agg_cases <- dep_plot_df %>%
+  tidyr::pivot_wider(
+    data = .,
+    id_cols = c(date),
+    names_from = dep,
+    values_from = cases,
+    names_prefix = 'cases_'
+  ) %>%
+  mutate(
+    ReportedAll = cases_Artibonite + cases_Centre +
+      cases_Grande_Anse + cases_Nippes + cases_Nord +
+      cases_Nord_Est + cases_Ouest + cases_Sud +
+      cases_Sud_Est + cases_Nord_Ouest
+  ) %>%
+  select(date, ReportedAll)
+
+h2_epi_traj$date <- as.Date(
+  lubridate::round_date(
+    lubridate::date_decimal(h2_epi_traj$year),
+    unit = 'day'
+  )
+)
+
+h2_end_traj$date <- as.Date(
+  lubridate::round_date(
+    lubridate::date_decimal(h2_end_traj$year),
+    unit = 'day'
+  )
+)
+
+all_traj <- bind_rows(
+  select(h2_epi_traj, date, Ctotal),
+  select(h2_end_traj, date, Ctotal)
+)
+
+projections <- left_join(
+  x = true_agg_cases,
+  y = all_traj,
+  by = 'date'
+)
+@
+
+\begin{figure}[ht]
+<<plotModel2Rep, cache=TRUE, echo=FALSE, fig.width=6.5, fig.height=4.5>>=
+gg_epi <- projections %>%
+  filter(date < as.Date("03-01-2014", format = '%m-%d-%Y')) %>%
+  ggplot(aes(x = date)) +
+  geom_line(aes(y = ReportedAll), col = 'black', linetype = 'dashed') +
+  geom_line(aes(y = Ctotal), col = 'red', size = 1) +
+  theme_bw() +
+  scale_y_continuous(limits = c(0, 30000), breaks = seq(0, 30000, 5000)) +
   theme(axis.title.x = element_blank()) +
-  ylab("Total Susceptible (%)")
-@
-\caption{\label{fig:h3Sus}Percentage of individuals that are in the susceptible compartment.
-Simulations from Model~3 under initial conditions are displayed in red; simulations from the filtering distribution of model are displayed in blue.}
+  ylab("Reported number of cases")
+
+gg_end <- projections %>%
+  filter(date >= as.Date("03-01-2014", format = '%m-%d-%Y')) %>%
+  ggplot(aes(x = date)) +
+  geom_line(aes(y = ReportedAll), col = 'black', linetype = 'dashed') +
+  geom_line(aes(y = Ctotal), col = 'red', size = 1) +
+  theme_bw() +
+  scale_y_continuous(limits = c(0, 7000), breaks = seq(0, 7000, 1000)) +
+  theme(axis.title.x = element_blank()) +
+  ylab("Reported number of cases") +
+  scale_x_date(date_breaks = "1 year", date_labels = "%Y")
+
+cowplot::plot_grid(gg_epi, gg_end, ncol = 1)
+@
+\caption{\label{fig:mod2rep}
+Model 2 trajectories using the \code{haitipkg}. Compare to Figure S15 of \cite{lee20sup}.
+}
 \end{figure}
+
+There are minor differences between Figure~\ref{fig:mod2rep} and Figure~S15 of \cite{lee20sup}.
+While the discrepancy appears minor, the deterministic nature of Model~2 implies that an exact replication of model trajectories should be possible.
+In this case, these discrepancies may be attributed to implementing the model in two different programing languages.
+Another potential explanation for the discrepancy is that the parameters that we used are only approximately the same as those used by \citet{lee20sup}.
+For example, the parameters $\transmission$, $\Wbeta{}$ (See Table~\ref{tab:translate}) had reported values of $9.9 \times 10^{-7}$ and $4.03 \times 10^{-2}$, respectively (Table~S13 of \cite{lee20sup}), but were actually fit to data and therefore likely had values that were far more precise.
+Additionally, our implementation of Model~2 used a time scale of years and many of the parameters were reported on a weekly scale, so small differences may result due to unit conversions.
+The collective effect of these small differences in model parameters likely will result in small differences in model trajectories.
+
+Some additional concerns about being able to accurately replicate the results of \cite{lee20} are valid.
+Details about the measurement models and how latent states were initialized for the epidemic model were not provided by \citet{lee20sup} and therefore these details must be inferred by looking at the provided source code.
+According to repository comments, the files \code{fit\allowbreak In\allowbreak Pieces\allowbreak 3params\allowbreak Clean\allowbreak May2019\allowbreak Public.py} and \code{fit\allowbreak In\allowbreak Pieces\allowbreak Mu\allowbreak With\allowbreak Frac\allowbreak Sus\allowbreak Fixed\allowbreak All\allowbreak Infections\allowbreak Public.py} were used to fit the epidemic and endemic phases of the model respectively, although it is apparent that these exact files were not used to obtain the reported results since the files contain some variable-naming errors that make it impossible to run the files without making modifications \footnote{One example of why the code cannot be run that the file loads functions from a non-extant file named \code{choleraEqs.py} in line 13 rather than \code{cholera\allowbreak Eqs\allowbreak Public.py}.}.
+The inability to replicate the results by \citet{lee20} by running the provided source code makes checking whether or not a our numeric implementation faithfully represents their results very difficult.
+Additionally, the script that was said to been used to obtain the results reported by \cite{lee20} appears to use a different measurement model than what was described in the supplemental material, again making it difficult to fully replicate the result of \cite{lee20} without being able to easily run the provided source code.
+In this case, we chose to use measurement model described by Eq.~\ref{model2:measEpi} for both phases of the epidemic, as this seemed to visually match the results of \cite{lee20} most closely.
+
+\subsection{Model~3 Replication}
+
+Model~3 was developed by a team of researchers at the Laboratory of the Swiss Federal Institute of Technology in Lausanne, hereafter referred to as the Model~3 authors.
+The code that was originally used to implement Model~3 is archived with the DOI: \url{10.5281/zenodo.3360723}, and also available in the public GitHub repository: \code{jcblemai/haiti-mass-ocv-campaign}.
+Because the code was made publicly available, and final model parameters were reported in the supplementary material of \cite{lee20}, we were able to reproduce Model~3 by directly using the source code.
+In Fig.~\ref{fig:mod3rep}, we plot simulations from this model.
+This figure should be compared to Figure S18 of \cite{lee20}.
+We note that slight differences may be accounted for due to variance in the model simulations and the difference in programming language used to produce the figure.
+
+<<Model3Replication, echo=FALSE, cache=TRUE>>=
+NSIM <- 500
+
+plot_order <- c(
+  'Artibonite',
+  'Sud_Est',
+  'Nippes',
+  'Nord_Est',
+  'Ouest',
+  'Centre',
+  'Nord',
+  'Sud',
+  'Nord_Ouest',
+  'Grande_Anse'
+)
+
+dep_labeller <- as_labeller(
+  c(
+    'Artibonite' = 'Artibonite',
+    'Sud_Est' = 'Sud-Est',
+    'Sud.Est' = 'Sud-Est',
+    'Nippes' = 'Nippes',
+    'Nord_Est' = 'Nord-Est',
+    'Nord.Est' = 'Nord-Est',
+    'Ouest' = 'Ouest',
+    'Centre' = 'Centre',
+    'Nord' = 'Nord',
+    'Sud' = 'Sud',
+    'Nord_Ouest' = 'Nord-Ouest',
+    'Nord.Ouest' = 'Nord-Ouest',
+    'Grande_Anse' = 'Grand\'Anse',
+    'Grand.Anse' = 'Grand\'Anse'
+  )
+)
+
+lee3 <- lee3_spatPomp()
+lee3_sims <- bake("../model3/run_level_3/si_lee3Sims.rds", {
+  simulate(lee3, nsim = NSIM, seed = 321, format = 'data.frame')
+})
+
+sub_cases <- haitiCholera %>%
+  dplyr::mutate(date = as.Date(date_saturday)) %>%
+  dplyr::select(-report, -date_saturday) %>%
+  dplyr::filter(lubridate::decimal_date(date) >= min(lee3_sims$time)) %>%
+  tidyr::pivot_longer(
+    data = .,
+    cols = -date,
+    names_to = "dep",
+    values_to = "cases"
+  ) %>%
+  dplyr::mutate(
+    dep = dplyr::case_when(
+      dep == "Grand.Anse" ~ "Grande_Anse",
+      dep == "Nord.Est" ~ "Nord_Est",
+      dep == "Nord.Ouest" ~ "Nord_Ouest",
+      dep == "Sud.Est" ~ "Sud_Est",
+      TRUE ~ dep
+    )
+  )
+
+quants <- lee3_sims %>%
+  dplyr::group_by(unitname, time) %>%
+  dplyr::summarise(
+    q05 = quantile(cases, 0.025, na.rm = T),
+    mean = mean(cases, na.rm = T),
+    q50 = quantile(cases, 0.5, na.rm = T),
+    q95 = quantile(cases, 0.975, na.rm = T)
+  ) %>%
+  dplyr::ungroup() %>%
+  dplyr::mutate(
+    date = lubridate::date_decimal(time),
+    date = as.Date(lubridate::round_date(date))
+  ) %>%
+  dplyr::rename(dep = unitname)
+
+quants <- quants %>%
+  dplyr::filter(
+    !(dep == "Ouest" & date < '2017-06-10')
+  )
+@
+
+\begin{figure}[ht]
+<<PlotMod3Rep, echo=FALSE, message=FALSE, fig.height=3.75>>=
+ggplot() +
+  geom_ribbon(data = quants, aes(x = date, ymin = q05, ymax = q95),
+              fill = 'darkblue', alpha = 0.5) +
+  geom_line(data = quants, aes(x = date, y = q50), col = 'darkblue') +
+  geom_point(data = dplyr::filter(sub_cases, cases < 500), aes(x = date, y = cases), size = 0.2) +
+  facet_wrap(~factor(dep, levels = plot_order), nrow = 2, labeller = dep_labeller) +
+  labs(y = 'Reported Cholera Cases') +
+  scale_x_date(date_breaks = "6 months", date_labels = "%Y-%m") +
+  theme(
+    axis.text.x = element_text(angle = 40, hjust = 1, vjust = 1, size = 7),
+    axis.title.x = element_blank()
+  )
+@
+\caption{\label{fig:mod3rep}
+Simulations from Model~3.
+Compare to Figure S18 of \cite{lee20}.
+}
+\end{figure}
+
+<<lee3Cleanup, echo=FALSE, include=FALSE, message=FALSE>>=
+rm(lee3, lee3_sims, quants, sub_cases)
+gc()
+@
+
+\section{Forecasting with parameter uncertainty}
+
+Let $f_{Y_{1:N}}(y_{1:N} | \theta)$ denote the pdf of the model under consideration, were $\theta$ is a parameter vector that indexes the model.
+Furthermore, denote the observed data as $y_{1:N}^*$.
+Because the uncertainty in just a single parameter can lead to drastically different forecasts \citep{saltelli20},
+parameter uncertainty should be considered when obtaining model forecasts when the goal is to influencing policy.
+In a Bayesian modelling paradigm, the most natural way to account for parameter uncertainty in model forecasts is to suppose that $\theta$ comes from a distribution $f_{\Theta}$, and then to obtain $J$ forecasts from the model where each forecast is obtained using parameters drawn from the posterior distribution $\Theta_{1:J} \mid Y_{1:N} = y_{1:N}^* \sim f_{\Theta}\big(\theta | Y_{1:N} = y_{1:N}^*\big)$.
+
+When frequentist methods are used, however, there does not exist a posterior distribution from which one could sample.
+A common approach could be to obtain a weighted average of the simulations from various models \citep{hoeting99}, but this can be problematic when forecasts from each model are very different from each other \citep{grueber11}.
+A similar approach that has been taken \citep{king15} is to obtain model forecasts using multiple sets of parameter values and then sample from the resulting forecasts using weights proportional to the corresponding likelihoods of the parameter values.
+This approach could be considered a quasi-Empirical Bayes, as it is equivalent to using a discrete uniform prior where the set of values in the prior distribution is determined by a stochastic routine applied to the observed data, as discussed below.
+
+For each $i \in 1:k$, let $\Theta_i$ be a random vector of model parameters.
+Then, letting $\Theta$ denote the true model parameters, we endow the set $\{\Theta_1, \Theta_2, \ldots, \Theta_k\}$ with a discrete uniform distribution, such that $P\big(\Theta = \Theta_i\big) = \frac{1}{K}$ for all values $i \in \seq{1}{k}$.
+Using this as a prior distribution, the posterior distribution of $\Theta | Y_{1:N} = y_{1:N}^*$ can be expressed as:
+$P\big(\Theta = \Theta_k | Y_{1:N} = y_{1:N}^*\big) = \frac{f_{Y_{1:N}}(y_{1:N}^*| \Theta_k)}{\sum_{l = 1}^K f_{Y_{1:N}}(y_{1:N}^*| \theta_l)}$.
+In a true Empirical Bayes analysis, the values $\Theta_1, \ldots, \Theta_k$ of the prior distribution would be chosen using the observed data, resulting in a posterior distribution that weighs the prior parameter vectors proportional to their corresponding likelihoods.
+Instead, we choose $\Theta_i$ to be the output of a stochastic routine applied to the observed data by setting $\Theta_i$ to be the output of an iterated filtering algorithm.
+In practice, because the likelihood maximization routines of iterated filtering methods are stochastic, it is common to run the iterated filtering method multiple times for each model in order to obtain a maximum likelihood estimate for model parameters.
+This results in a natural set of parameters near the MLE that could be used in in the discrete prior distribution.
 

### 2022-12-17 ms.Rnw

@@
 The groups largely adhered to existing guidelines on creating models to inform policy \citep{behrend20,saltelli20} and, despite their autonomy, obtained a consensus that an extensive nationwide vaccination campaign would be necessary to eliminate cholera from Haiti.
-Although several cholera cases have recently been detected in Haiti \citep{world2010outbreak,rubin22},
-their conclusion is inconsistent with the prolonged period with no confirmed cholera cases between February, 2019 \citep{trevisin22} and October, 2022.
+Their conclusion is inconsistent with the prolonged period with no confirmed cholera cases between February, 2019 and September, 2022 \citep{trevisin22} though cholera has recently reemerged in Haiti \citep{rubin22}.
 

### 2022-12-19 ms.Rnw
Refactored files, addressed some of Ed's comments

@@
 h1_scen_NP   <- switch(RUN_LEVEL, 50, 500, 5000)
-h1_scen_sims <- switch(RUN_LEVEL, 20, 100,  500)
+h1_scen_sims <- switch(RUN_LEVEL, 20, 100, 1000)
 
@@
       # project vacc scenario, with filtering dist, and model.
-      h1_S2_sims <- project_from_filter(mod = s2, end_states = end_states, nsims = h1_scen_NP)
+      h1_S2_sims <- project_from_filter(mod = s2, end_states = end_states, nsims = h1_scen_sims)
 
@@
       # project vacc scenario, with filtering dist, and model.
-      h1_S4_sims <- project_from_filter(mod = s4, end_states = end_states, nsims = h1_scen_NP)
+      h1_S4_sims <- project_from_filter(mod = s4, end_states = end_states, nsims = h1_scen_sims)
 
@@
       # project vacc scenario, with filtering dist, and model.
-      h1_S3_sims <- project_from_filter(mod = s3, end_states = end_states, nsims = h1_scen_NP)
+      h1_S3_sims <- project_from_filter(mod = s3, end_states = end_states, nsims = h1_scen_sims)
 
@@
       # project vacc scenario, with filtering dist, and model.
-      h1_S1_sims <- project_from_filter(mod = s1, end_states = end_states, nsims = h1_scen_NP)
+      h1_S1_sims <- project_from_filter(mod = s1, end_states = end_states, nsims = h1_scen_sims)
 
@@
 <<Load Model 3 Vacc Scenarios, message=FALSE, include=FALSE>>=
-NP_BPF <- switch(RUN_LEVEL, 50, 200, 1000)
-NSIM   <- switch(RUN_LEVEL, 10,  50,  500)
+NP_BPF <- switch(RUN_LEVEL, 50, 200, 2000)
+NSIM   <- switch(RUN_LEVEL, 10,  50, 1000)
 

### 2022-12-19 si/si.Rnw
Refactored files, addressed some of Ed's comments

@@
 
-<<packages, include=FALSE, cache=FALSE, echo=FALSE>>=
+<<packages, include=FALSE>>=
 library(tidyverse)
@@
 cores <-  as.numeric(Sys.getenv('SLURM_NTASKS_PER_NODE', unset = NA))
-# if(is.na(cores)) cores <- detectCores()
-cores <- 20
+if(is.na(cores)) cores <- detectCores()
+# cores <- 20
 registerDoParallel(cores)
 theme_set(theme_bw())
-@
-
-<<set-opts, include=FALSE, cache=FALSE, echo=FALSE>>=
-
+
+if (!dir.exists("../model3/si")) {
+  dir.create("../model3/si", recursive = TRUE)
+}
+@
+
+<<set-opts, include=FALSE>>=
 options(
@@
   stringsAsFactors = FALSE,
-  # prompt = "R> ",
   continue = "+  ",
@@
 
-<<knitr-opts, include=FALSE, cache=FALSE, purl=FALSE>>=
-library("knitr")
+<<knitr-opts, include=FALSE, purl=FALSE>>=
+library(knitr)
 opts_knit$set(concordance=TRUE)
@@
   error=TRUE,
-  echo=TRUE,
+  echo=FALSE,
   cache=FALSE,
@@
   results="markup",
-  background="#FFFFFF00",
-  size="normalsize",
   fig.path="figure/",
@@
   fig.show="asis",
-  #    dpi=300,
   dev="pdf",
@@
 
-\newcommand\mytitle{Informing policy via dynamic models: Eliminating cholera in Haiti}
+\newcommand\mytitle{Informing policy via dynamic models: Cholera in Haiti}
 \newcommand\comma{{\hspace{-0.25mm},\hspace{-0.2mm}}}
@@
 Here, we add additional details to clarify the mapping from a collection of rate functions to a fully specified process.
-Our treatment follows \citet{breto09}.
-
-A general compartment model is a vector-valued process $X(t)=(X_1(t),\dots,X_c(t))$ denoting the (integer or real-valued) counts in each of $c$ compartments.
+Our treatment follows \citet{breto09} \jwc{TODO: Cite the short course}.
+
+A general compartment model is a vector-valued process $X(t)=(X_1(t),\dots,X_c(t))$ denoting the (integer or real-valued) counts in each of $c$ compartments, where $t$ is any continuous value in the interval  $\left[t_0, \infty\right)$ for some real valued starting time $t_0$.
 The compartments may also have names, but to set up general notation we simply refer to them by their numerical index.
-The basic characteristic of a compartment model is that $X(t)$ can be written in terms of the flows $N_{ij}(t)$ from $i$ to $j$, together with flows into and out of each compartment denoted by $N_{\bullet i}(t)$ and $N_{i\bullet}(t)$ respectively.
+The basic characteristic of a compartment model is that $X(t)$ can be written in terms of the flows $N_{ij}(t)$ from $i$ to $j$.
+A flow into compartment $i$ from outside the system is denoted by $N_{\circ i}$, and a flow out of the system from compartment $i$ is denoted by $N_{i\circ}$.
+We call $\circ$ a source/sink compartment, though it is an irregular compartment since $X_{circ}(t)$ is not defined.
 These flows are required to satisfy a ``conservation of mass'' identity:
 \begin{equation}
-X_i(t)=X_i(0)+N_{\bullet i}(t) - N_{i\bullet}(t) + \sum_{j\neq
+X_i(t)=X_i(t_0)+N_{\bullet i}(t) - N_{i\bullet}(t) + \sum_{j\neq
 i}N_{ji}(t)-\sum_{j\neq i}N_{ij}(t). \label{eq:conservation}
@@
 \prob\big[ N_{ij}(t+\delta)-N_{ij}(t)=0 \given X(t)\big]
- &=& 1-\delta \mu_{ij}\big(t,X(t)\big) + o(\delta),
+ &=& 1-\delta \mu_{ij}\big(t,X(t)\big)X_i(t) + o(\delta),
 \\
@@
 \prob\big[ N_{ij}(t+\delta)-N_{ij}(t)=1 \given X(t)\big]
- &=& \delta \mu_{ij}\big(t,X(t)\big) + o(\delta),
+ &=& \delta \mu_{ij}\big(t,X(t)\big)X_i(t) + o(\delta),
 \end{eqnarray}
@@
 \end{equation}
-Formally interpreting the meaning of \myeqref{eq:odmcs1} is not trivial, and we do so by defining the solution of \myeqref{eq:odmcs1} to be the limit of an Euler scheme.
+Formally interpreting the meaning of \myeqref{eq:odmcs1} is not trivial, and we do so by constructing a Markov process $X(t)$ as the limit of the Euler scheme descrived in Section~\ref{sec:numerics}, below.
 Therefore, the numerical scheme in Sec.~\ref{sec:numerics} can be taken as a definition of the meaning of \myeqref{eq:odmcs1}.
-The Markov chain defined by the limit of this Euler scheme as the step size decreases is an overdispersed Markov counting system, with the possibility of instantaneus jumps of size greater than one \citep{breto11}.
+The Markov chain defined by the limit of this Euler scheme as the step size decreases is an overdispersed Markov counting system, with the possibility of instantaneous jumps of size greater than one \citep{breto11}.
 
@@
 
-For the case $i=\bullet$, one can use
+Different treatments of demographic flows---such as birth, death, immigration and emigration---are possible.
+For the case $i=\bullet$, the treatment used by Model~1 is to set
 \begin{equation}
@@
 \end{equation}
-an independent Poisson random variable with mean $\mu_{\bullet j} \delta$, as was done in Model~1.
-
-Another common approach is to balance the total number of flows in and out of the compartment, i.e., $\sum_{i}N_{\bullet i}(t) = \sum_{i}N_{i \bullet}(t)$, in order to make the model consistent with the known total population, as was done in Models~2 and~3.
+an independent Poisson random variable with mean $\mu_{\bullet j} \delta$.
+
+Models~2 and 3 used an alternative approach, balancing the total number of flows in and out of the compartment, i.e., $\sum_{i}N_{\bullet i}(t) = \sum_{i}N_{i \bullet}(t)$, in order to make the model consistent with the known total population.
 In this case, we formally model the death rate as a rate of returning to the susceptible class $S$, and use external transitions from $\bullet$ into $S$ to describe only net population increase.
@@
 An instantaneous change from near perfect to almost non-existent reporting can be problematic, as it forces the model to explain the observed reduction in reported cases as a decrease in the reporting of cases, rather than a decrease in the prevalence of cholera.
-This shift was justified by a "change of the case definition that occurred on January 1st, 2018";
+This shift was justified by a ``change of the case definition that occurred on January 1st, 2018";
 this claim was not cited, and we could find no evidence that such a drastic change in the reporting rate would be warranted.
@@
 
-To perform inference on POMP models, it is necessary to propose an initial density for the latent process $f_{X_0}(x_0;\theta)$.
+To perform inference on POMP models, it is necessary to propose an initial probability density for the latent process $f_{X_0}(x_0;\theta)$, including the possibility that the initial values could be known, or could be a non-random function of the unknown parameter vector, $\theta$.
 This density is used to obtain initial values of the latent state when fitting and evaluating the model.
 For each of the models considered in this analysis, the initial conditions are derived by enforcing the model dynamics on reported cholera cases.
-It is also sometimes necessary to fit some initial value parameters in order to help determine starting values for weakly identifiable compartments.
+It is also sometimes necessary to fit some initial value parameters in order to help determine initial values for weakly identifiable compartments.
 In the following subsections, we mention initial value parameters that were fit for each model.
@@
 
-For this model, the number of individuals in the Recovered and Asymptomatic compartments are set to zero, but the initial proportion of Infected and Exposed individuals is estimated as initial value parameters ($I(0)$ and $E(0)$, respectively) using the MIF2 algorithm.
+For this model, the number of individuals in the Recovered and Asymptomatic compartments are set to zero, but the initial proportion of Infected and Exposed individuals is estimated as initial value parameters ($I(0)$ and $E(0)$, respectively) using the IF2 algorithm, implemented as \texttt{mif2} in the \texttt{pomp} package.
 Finally, the initial proportion of Susceptible individuals $S_0$ is calculated as $S(0) = 1 - I(0) - E(0)$.
@@
 
-Model~2 assumes that the initial vales are known and constant, and so no initial value parameters need to be estimated.
-Starting values for latent state compartments are chosen so as to enforce the model dynamics on the observed number of cases.
-Specifically, the model sets $I_{u}(0) = y^*_{u}(0) / \reportRate$ for each unit $u \in 1:10$, where $I_{u}(t)$ is the number of infected individuals in unit $u$ at time $t$, $y^*_{u}(t)$ is the reported number of cases, and $\reportRate$ is the reporting rate.
+Model~2 assumes that the initial values are a deterministic function of the reporting rate and the initial case reports, and so no initial value parameters need to be estimated.
+Initial values for latent state compartments are chosen so as to enforce the model dynamics on the observed number of cases.
+Specifically, the model sets $I_{u0}(0) = y^*_{1u} / \reportRate$ for each unit $u \in 1:10$, where $I_{u0}(t)$ is the number of infected individuals in unit $u$ at time $t$, vaccination scenario $z = 0$, $y^*_{tu}$ is the reported number of cases, and $\reportRate$ is the reporting rate.
 
@@
 The initial value for the bacteria compartment is then determined using the rainfall dynamics of Eq.~23.
-For the departments of Grande'Anse, Nord-Ouest, and Sud-Est, estimating the latent states at time $t_0$ using the observed data at time $t_1$ is problematic (see Sec.~\ref{sec:mod3Cal}), so we instead estimate initial value parameters $I_{u0}(0)$ for the integers $u \in \{3, 7, 9\}$, which correspond to the aforementioned departments, and the remaining hidden states are estimated by enforcing model dynamics using this result.
+For the departments of Grande'Anse, Nord-Ouest, and Sud-Est, estimating the latent states at time $t_0$ using the observed data at time $t_1$ is problematic (see Sec.~\ref{sec:mod3Cal}), so we instead estimate initial value of the latent state $I_{u0}(t_0)$ for the integers $u \in \{3, 7, 9\}$, which correspond to the aforementioned departments.
+The remaining hidden states are estimated by enforcing model dynamics using this result.
 
@@
 
-<<CalibrateOrig3, echo=FALSE>>=
+<<CalibrateOrig3>>=
 set.seed(8636712)
@@
 )
+
+# Vector containing all of the estimate parameter names.
 est_param_names <- c(
@@
 
+# This function sets the rw.sd for named parameters, first search.
 set_rw_1 <- function(x) {
@@
 
+# This function sets the rw.sd for named parameters, second search.
 set_rw_2 <- function(x) {
@@
 
+# This function sets the rw.sd for named parameters, final search.
 set_rw_3 <- function(x) {
@@
 
-# Search hyper-parameters for each run-level and search
+# Search1 hyper-parameters
 SEARCH1 = list(
@@
   COOLING = 0.5,
-  KEEP_TRACES = TRUE,
-  KEEP_LIK_MAT = TRUE
-)
-
+  KEEP_TRACES = FALSE,
+  KEEP_LIK_MAT = FALSE
+)
+
+# Search2 hyper-parameters
 SEARCH2 = list(
@@
   COOLING = 0.5,
-  KEEP_TRACES = TRUE,
-  KEEP_LIK_MAT = TRUE
-)
-
+  KEEP_TRACES = FALSE,
+  KEEP_LIK_MAT = FALSE
+)
+
+# Search3 hyper-parameters
 SEARCH3 = list(
@@
   COOLING = 0.5,
-  KEEP_TRACES = TRUE,
+  KEEP_TRACES = FALSE,
   KEEP_LIK_MAT = TRUE
@@
 
-# TODO: Move to run_level_3 folder...
-
-save_file <- "../model3/run_level_2/fit_orig_model.rds"
-
+save_file <- "../model3/si/fit_orig_model.rds"
+
+# Fit the model and save the results
 haiti3_orig_fit <- bake(
@@
 
+# Find model index with best log-likelihood
 best_m <- which.max(haiti3_orig_fit$search3$logLiks$logLik)
 
+# Find parameters using best_m index
 best_params <- haiti3_orig_fit$search3$params[best_m, ]
 
+# Get unit log-likelihoods for the best model.
 all_liks <- as.data.frame(haiti3_orig_fit$search3$likMat)
-all_liks$which <- rep(1:36, each = 9)
+all_liks$which <- rep(1:(SEARCH3$TOP_N * SEARCH3$NREPS), each = SEARCH3$NREPS_EVAL)
 best_mod <- filter(all_liks, which == best_m)
 
+# Pivot longer to get estimate of log-likelihood for each unit.
 unitLikesM3 <- best_mod %>%
@@
 
-<<M3_bench, echo=FALSE, message=FALSE, include=FALSE>>=
+<<M3_bench, echo=FALSE, message=FALSE, include=FALSE, cache=TRUE>>=
+
+# Create dataset that will be used to fit ARMA models. To do this, we
+# need to make sure that missing values are marked as NA, so we check
+# for any jumps in time.
 MODEL3_CASES <- haitiCholera %>%
@@
 
+# Find jumps
 jumps <- which(MODEL3_CASES$time_diff > lubridate::make_difftime(day = 7))
 
+# If there is a jump, insert NA between observations
 temp <- MODEL3_CASES
@@
 
-mod3_cases <- temp %>%
-  dplyr::select(-time_diff)
+mod3_cases <- temp %>% dplyr::select(-time_diff)
 
@@
 deps <- h3@unit_names
+
+# Combine tables with model 3 and benchmark likelihoods.
 ARMA_M3_joined <- dplyr::left_join(
@@
 
+# Change model parameters to the best set of parameters from previous
+# search
 coef(h3)[names(best_params)] <- best_params
 
-registerDoRNG(40765101)
-
-bpf_out <- bake(
-  file = "../model3/run_level_3/si_bpf.rds", {
-    h3 %>% bpfilter(
+registerDoRNG(40765101)  # for modular, reproducible results
+
+# Perform block partical filter with fitted model, to get conditional
+# log-likelihoods.
+bpf_orig3_out <- bake(
+  file = "../model3/si/bpf_orig3_cond_LL.rds", {
+
+    results <- list()
+
+    bpf_out <- h3 %>% bpfilter(
       Np = 2000,
@@
     )
+
+    results$Sus_df <- foreach(i=1:length(h3@times), .combine = bind_rows) %do% {
+      get_h3_S_quants(bpf_out, i)
     }
-)
-
+
+    results$block.cond.loglik <- bpf_out@block.cond.loglik
+    results$times <- bpf_out@times  # Saved for convenience
+    results
+  },
+  seed = 432235
+)
 @
@@
 While the model does include a mechanism to account for increased cholera transmission due to large rainfall events, the mechanism does not appear to be sufficient to capture the damaging effects of the hurricane, which had the greatest impact in the the Sud and Grand'Anse departments \citep{ferreirai16}.
-This result led us to include parameters $\Whur{u}$ and $\hHur{u}$ in Eq.~23 that would increase the transmission between environmental cholera and humans for Sud and Grand'Anse.
+This result led us to include parameters $\Whur{u}$ and $\hHur{u}$ in Eq.~23 of the main text, which allows for an increase in the transmission rate between environmental cholera and humans for Sud and Grand'Anse during and after the hurricane.
 The effect of the hurricane on cholera transmission is assumed to have an exponential decay, where the magnitude is determined by $\Whur{u}$ and the duration of the effect determined by $\hHur{u}$.
@@
 <<CondLLFig, echo=FALSE, message=FALSE, fig.height=3>>=
+
+# Changing labels to human-readable labels.
 deps[grepl("_", deps)] <- gsub("_", "-", deps[grepl("_", deps)])
@@
 names(deps) <- as.character(1:10)
-
 myLabeller <- as_labeller(deps)
 
-t(bpf_out@block.cond.loglik) %>%
+# Plot conditional log-likelihoods for each unit
+t(bpf_orig3_out$block.cond.loglik) %>%
   as.data.frame() %>%
-  mutate(t = bpf_out@times) %>%
+  mutate(t = bpf_orig3_out$times) %>%
   pivot_longer(
@@
 <<loadFinalModel3, echo=FALSE, include=FALSE, message=FALSE>>=
+
+# This result was used in the manuscript, so it doesn't need to be recalculated.
 h3_final_results <- readRDS("../model3/run_level_3/haiti3_coupled_fit.rds")
 
+# Save best_m as index for the model parameters with the best likelihood.
+# Then, save the model parameter values and the likelihood evaluations
 best_m <- which.max(h3_final_results$search3$logLiks$logLik)
@@
 
+# Get repeated estimates of unit log-likelihoods
 unitLikesM3 <- best_mod %>%
@@
 
+# Combine model likelihood table to benchmarks
 ARMA_M3_joined <- dplyr::left_join(
@@
 <<finalModel3BPF, echo=FALSE, message=FALSE, include=FALSE>>=
+
+# Set the model parameters to the best results from previous search
 coef(h3)[names(best_params)] <- best_params
 
-bpf_out <- bake(
-  file = "../model3/run_level_3/si_h3_final_bpf.rds", {
-    h3 %>% bpfilter(
+# block particle filter to get conditional log-likelihoods.
+bpf_final3_out <- bake(
+  file = "../model3/si/h3_final_bpf_cond_ll.rds", {
+
+    results <- list()
+
+    bpf_out <- h3 %>% bpfilter(
       Np = 2000,
@@
     )
+
+    results$block.cond.loglik <- bpf_out@block.cond.loglik
+    results$times <- bpf_out@times  # saved for convenience
+    results
   },
@@
 <<finalCondLLFig, echo=FALSE, message=FALSE, fig.height=3>>=
-t(bpf_out@block.cond.loglik) %>%
+t(bpf_final3_out$block.cond.loglik) %>%
   as.data.frame() %>%
-  mutate(t = bpf_out@times) %>%
+  mutate(t = bpf_final3_out$times) %>%
   pivot_longer(
@@
 This could be due to many factors, including the possibility that the coupling between units forces low initial value parameters in a given department in order for the model to fit the data in the remaining departments.
-Whatever the case may be, the inclusion of the seven parameters resulted in a $\Sexpr{round(max(h3_final_results$search3$logLiks$logLik - haiti3_orig_fit$search3$logLiks$logLik), 1)}$, which is a highly statistically significant difference.
+Whatever the case may be, the inclusion of the seven parameters resulted in a $\Sexpr{round(max(h3_final_results$search3$logLiks$logLik) - max(haiti3_orig_fit$search3$logLiks$logLik), 1)}$ increase in log-likelihood, which is a highly statistically significant difference.
 One could similarly test whether the inclusion of the initial value parameters resulted in a significantly better fit to the observed data, but we do not do this here due to the large computational effort that would be required.
@@
 <<h3SusFigSetup, echo=FALSE, message=FALSE, include=FALSE>>=
+
+# This bake perfoms the following:
+#  (1) simulates the model from initial conditions
+#  (2) get the values of the hidden states at the end of each simulation, and
+#      saves these for later use.
+#  (3) Obtain quantiles for the Susceptible pool from the simulations.
 init_conditions_results <- bake(
-  file = "../model3/run_level_3/si_InitSimsQuants.rds", {
-
-    results <- list()
-
+  file = "../model3/si/InitSimsQuants.rds", {
+
+    results <- list()  # To return final output of bake
+
+    # Simulate from the fitted model
     h3Spat_sims <- simulate(
-      h3, nsim = 200, format = 'data.frame'
-    )
-
+      h3, nsim = 2000, format = 'data.frame'
+    )
+
+    # Get states at the end of the simulations
     h3_spat_end <- h3Spat_sims %>%
@@
 
+    # Save hidden state values at the end of the simulation. Saved as a matrix
+    # in the same format as the output from a block-particle filter, so that
+    # it can be used as the input to project_from_filter.
     results[["end_conditions"]] <- h3_spat_end %>%
@@
 
+    # Get the quantiles for the Susceptible pool from the simulations
     results[["SQuants"]] <- h3Spat_sims %>%
@@
       ) %>%
-      filter(date >= as.Date(lubridate::round_date(lubridate::date_decimal(h3@t0), unit = 'day'))) %>%
+      filter(
+        date >= as.Date(
+          lubridate::round_date(lubridate::date_decimal(h3@t0), unit = 'day')
+        )
+      ) %>%
       mutate(date = as.Date(lubridate::round_date(date, unit = 'day')))
@@
 
+# Get the output of the "bake" call for plotting.
 h3Spat_quants <- init_conditions_results[["SQuants"]]
 
+# This bake call uses the ending hidden states from above to forecast cholera
+# cases in the future under the V0 vaccination scenario
 mod3_V0_initcond_res <- bake(
-  file = '../model3/run_level_3/si_vacc_sim_from_init.rds', {
+  file = '../model3/si/vacc_sim_from_init.rds', {
     h3_V0_res <- list()
@@
 
-    # Using parameters and filtered distribution, project system
+    # Using parameters and end state values, obtain projections
     h3_V0_sims <- project_from_filter(
       h3Spat, end_states = h3_bpf_end_states, covarGen = project_rain,
-      nsims = 500, seed = 184697
+      nsims = 1000, seed = 184697
     )
@@
       summarize(
-        Q025 = quantile((S + VSd + VSdd + VSdd_alt + VSd_alt) / pop, probs = 0.025),
-        Q50 = quantile((S + VSd + VSdd + VSdd_alt + VSd_alt) / pop, probs = 0.5),
-        Q975 = quantile((S + VSd + VSdd + VSdd_alt + VSd_alt) / pop, probs = 0.975)
+        Q025 = quantile(
+          (S + VSd + VSdd + VSdd_alt + VSd_alt) / pop,
+          probs = 0.025
+        ),
+        Q50 = quantile(
+          (S + VSd + VSdd + VSdd_alt + VSd_alt) / pop,
+          probs = 0.5
+        ),
+        Q975 = quantile(
+          (S + VSd + VSdd + VSdd_alt + VSd_alt) / pop,
+          probs = 0.975
+        )
       ) %>%
@@
 
-Sus_df <- bake(
-  file = "../model3/run_level_3/si_S_quants.rds", {
-    foreach(i=1:length(h3@times), .combine = bind_rows) %do% {
-      get_h3_S_quants(bpf_out, i)
-    }
-  }
-)
-
+# This result was already used in the manuscript; it doesn't need to
+# be recalculated.
 mod3_V0_res <- readRDS("../model3/run_level_3/mod3_V0_res.rds")
@@
   geom_line(
-    data = Sus_df,
+    data = bpf_orig3_out$Sus_df,
     aes(x = as.Date(lubridate::date_decimal(time)), y = Q50),
@@
   geom_ribbon(
-    data = Sus_df,
+    data = bpf_orig3_out$Sus_df,
     aes(x = as.Date(lubridate::date_decimal(time)), ymin = Q025, ymax = Q975),
@@
 
-\section{Lee et al. (2020) Replication}
+\section{Replication of \cite{lee20}}
 
@@
 In the following subsections, we use our \code{R} package \code{haitipkg} to reproduce some of the results of \cite{lee20}.
-This reproduction allows us to estimate the likelihoods of the \cite{lee20} version of Models~1--3, and also provides a demonstration on reproducibility.
+This reproduction allows us to estimate the likelihoods of the \cite{lee20} version of Models~1--3, and also provides a demonstration of reproducibility.
 
@@
 
-To calculate an upper-bound on the likelihood of this model---which uses a large number of parameters that each have a different likelihood value---we estimate the likelihood corresponding to each parameter set using a particle filter, and report the maximum of these likelihoods.
+Because the model forecasts provided by \cite{lee20} come from various sets of parameters---which each correspond to a unique log-likelihood value---it is not obvious how one would obtain an estimate for the log-likelihood of the model that was used for simulations by the Model~1 authors.
+One approach could be to calculate the logarithm of the weighted mean of the likelihoods for each parameter sets used to obtain the forecasts, where the weights are proportional to the number of times the parameter set was used.
+However, in an effort to not underestimate the likelihood of the model of the Model~1 authors, we report the estimated log-likelihood as the log-likelihood value corresponding to the parameter set with the largest likelihood value, even though the majority of simulations were obtained using parameter sets with lower likelihood values.
+In this sense, we consider the log-likelihood reported in Table~1 of the main text to be an upper-bound of the log-likelihood of the model used by \cite{lee20}.
+For each parameter set, the log-likelihood was estimated using a particle filter, implemented as the \texttt{pfilter} function in the \texttt{pomp} package.
 
@@
 lee3 <- lee3_spatPomp()
-lee3_sims <- bake("../model3/run_level_3/si_lee3Sims.rds", {
-  simulate(lee3, nsim = NSIM, seed = 321, format = 'data.frame')
+lee3_sims_quants <- bake("../model3/si/lee3Sims.rds", {
+
+  # Simulate from lee et al 2020 model
+  sims_temp <- simulate(
+    lee3, nsim = NSIM,
+    seed = 321, format = 'data.frame'
+  )
+
+  # Get desired quantiles of reported cases from simulations
+  quants <- sims_temp %>%
+    dplyr::group_by(unitname, time) %>%
+    dplyr::summarise(
+      q05 = quantile(cases, 0.025, na.rm = T),
+      mean = mean(cases, na.rm = T),
+      q50 = quantile(cases, 0.5, na.rm = T),
+      q95 = quantile(cases, 0.975, na.rm = T)
+    ) %>%
+    dplyr::ungroup() %>%
+    dplyr::mutate(
+      date = lubridate::date_decimal(time),
+      date = as.Date(lubridate::round_date(date))
+    ) %>%
+    dplyr::rename(dep = unitname)
+
+  # Remove cases from Ouest that were not considered by Lee et al. (2020)
+  quants %>%
+    dplyr::filter(
+      !(dep == "Ouest" & date < '2017-06-10')
+    )
+
 })
 
+# Get the subset of cholera cases that match simulations from Lee et al. (2020) model.
 sub_cases <- haitiCholera %>%
@@
   dplyr::select(-report, -date_saturday) %>%
-  dplyr::filter(lubridate::decimal_date(date) >= min(lee3_sims$time)) %>%
+  dplyr::filter(lubridate::decimal_date(date) >= min(lee3@times)) %>%
   tidyr::pivot_longer(
@@
   )
-
-quants <- lee3_sims %>%
-  dplyr::group_by(unitname, time) %>%
-  dplyr::summarise(
-    q05 = quantile(cases, 0.025, na.rm = T),
-    mean = mean(cases, na.rm = T),
-    q50 = quantile(cases, 0.5, na.rm = T),
-    q95 = quantile(cases, 0.975, na.rm = T)
-  ) %>%
-  dplyr::ungroup() %>%
-  dplyr::mutate(
-    date = lubridate::date_decimal(time),
-    date = as.Date(lubridate::round_date(date))
-  ) %>%
-  dplyr::rename(dep = unitname)
-
-quants <- quants %>%
-  dplyr::filter(
-    !(dep == "Ouest" & date < '2017-06-10')
-  )
 @
@@
 ggplot() +
-  geom_ribbon(data = quants, aes(x = date, ymin = q05, ymax = q95),
-              fill = 'darkblue', alpha = 0.5) +
-  geom_line(data = quants, aes(x = date, y = q50), col = 'darkblue') +
-  geom_point(data = dplyr::filter(sub_cases, cases < 500), aes(x = date, y = cases), size = 0.2) +
-  facet_wrap(~factor(dep, levels = plot_order), nrow = 2, labeller = dep_labeller) +
+  # Region for upper and lower simulation bounds
+  geom_ribbon(
+    data = lee3_sims_quants,
+    aes(x = date, ymin = q05, ymax = q95),
+    fill = 'darkblue', alpha = 0.5
+  ) +
+  # Median line
+  geom_line(
+    data = lee3_sims_quants,
+    aes(x = date, y = q50),
+    col = 'darkblue'
+  ) +
+  # Observed data
+  geom_point(
+    data = dplyr::filter(sub_cases, cases < 500),
+    aes(x = date, y = cases), size = 0.2
+  ) +
+  facet_wrap(
+    ~factor(dep, levels = plot_order),
+    nrow = 2, labeller = dep_labeller
+  ) +
   labs(y = 'Reported Cholera Cases') +
@@
 <<lee3Cleanup, echo=FALSE, include=FALSE, message=FALSE>>=
-rm(lee3, lee3_sims, quants, sub_cases)
+rm(lee3, lee3_sims_quants, sub_cases)
 gc()
@@
 A similar approach that has been taken \citep{king15} is to obtain model forecasts using multiple sets of parameter values and then sample from the resulting forecasts using weights proportional to the corresponding likelihoods of the parameter values.
-This approach could be considered a quasi-Empirical Bayes, as it is equivalent to using a discrete uniform prior where the set of values in the prior distribution is determined by a stochastic routine applied to the observed data, as discussed below.
+This approach could be considered as empirical Bayes, as it is equivalent to using a discrete uniform prior where the set of values in the prior distribution is determined by a stochastic routine applied to the observed data, as discussed below.
 
@@
 $P\big(\Theta = \Theta_k | Y_{1:N} = y_{1:N}^*\big) = \frac{f_{Y_{1:N}}(y_{1:N}^*| \Theta_k)}{\sum_{l = 1}^K f_{Y_{1:N}}(y_{1:N}^*| \theta_l)}$.
-In a true Empirical Bayes analysis, the values $\Theta_1, \ldots, \Theta_k$ of the prior distribution would be chosen using the observed data, resulting in a posterior distribution that weighs the prior parameter vectors proportional to their corresponding likelihoods.
+In a standard empirical Bayes analysis, the values $\Theta_1, \ldots, \Theta_k$ of the prior distribution would be chosen using the observed data, resulting in a posterior distribution that weighs the prior parameter vectors proportional to their corresponding likelihoods.
 Instead, we choose $\Theta_i$ to be the output of a stochastic routine applied to the observed data by setting $\Theta_i$ to be the output of an iterated filtering algorithm.

### 2022-12-20 si/si.Rnw
Pushing all supplement files

@@
 
-    results$Sus_df <- foreach(i=1:length(h3@times), .combine = bind_rows) %do% {
-      get_h3_S_quants(bpf_out, i)
-    }
-
     results$block.cond.loglik <- bpf_out@block.cond.loglik
@@
     )
+
+    results$Sus_df <- foreach(i=1:length(h3@times), .combine = bind_rows) %do% {
+      get_h3_S_quants(bpf_out, i)
+    }
 
@@
 
+    # store the end states that will be used as initial values
     h3_bpf_end_states <- init_conditions_results$end_conditions
@@
 
+    # Get quantiles of observed cases for each department and time
     h3_V0_res$UnitQuants <- h3_V0_sims %>%
       mutate(
-        date = as.Date(lubridate::round_date(lubridate::date_decimal(time), unit = 'day'))
+        date = as.Date(  # Get date object from decimal time
+          lubridate::round_date(lubridate::date_decimal(time), unit = 'day')
+        )
       ) %>%
@@
 
+    # Get quantiles of susceptible individuals for each department and time
     h3_V0_res$SQuants <- h3_V0_sims %>%
       mutate(
-        date = as.Date(lubridate::round_date(lubridate::date_decimal(time), unit = 'day'))
+        date = as.Date(
+          lubridate::round_date(lubridate::date_decimal(time), unit = 'day')
+        )
       ) %>%
@@
       mutate(
-        pop = h3@params[
-          paste0("H", which(first(unitname) == unit_names(h3)))
-        ]
-      ) %>%
-      summarize(
+        # Get the population for each unit
+        pop = h3@params[paste0("H", which(first(unitname) == unit_names(h3)))]
+      ) %>%
+      summarize(  # Calculate the quantiles
         Q025 = quantile(
@@
   geom_line(
-    data = bpf_orig3_out$Sus_df,
+    data = bpf_final3_out$Sus_df,
     aes(x = as.Date(lubridate::date_decimal(time)), y = Q50),
@@
   geom_ribbon(
-    data = bpf_orig3_out$Sus_df,
+    data = bpf_final3_out$Sus_df,
     aes(x = as.Date(lubridate::date_decimal(time)), ymin = Q025, ymax = Q975),

### 2022-12-20 ms.Rnw
AR comments

@@
 The groups largely adhered to existing guidelines on creating models to inform policy \citep{behrend20,saltelli20} and, despite their autonomy, obtained a consensus that an extensive nationwide vaccination campaign would be necessary to eliminate cholera from Haiti.
-Their conclusion is inconsistent with the prolonged period with no confirmed cholera cases between February, 2019 and September, 2022 \citep{trevisin22} though cholera has recently reemerged in Haiti \citep{rubin22}.
+Their conclusion is inconsistent with the prolonged period with no confirmed cholera cases between February, 2019 and September, 2022 \citep{trevisin22}, though cholera has recently reemerged in Haiti \citep{rubin22}.
 
@@
 We proceed by introducing Models~1--3 in Sec.~\ref{sec:models};
-in Sec.~\ref{sec:methods}, we present a systematic approach to examining and refining these models, and then use improved model fits to project cholera incidence in Haiti under various vaccination scenarios.
+in Sec.~\ref{sec:methods}, we present a systematic approach to examining and refining these models and then use improved model fits to project cholera incidence in Haiti under various vaccination scenarios.
 This is followed by a consideration of the robustness of model based policy recommendations in Sec.~\ref{sec:advice}, and a discussion in Sec.~\ref{sec:discussion}.
@@
 Together, with the density for the initial value of the latent state $\bm{X}^{(\modelCounter)}_{0} = \bm{X}^{(\modelCounter)}(t_0)$, each model assumes a joint density $f_{\bm{X}^{(\modelCounter)}_{0:N}, \bm{Y}^{(\modelCounter)}_{1:N}}\big(\bm{x}^{(\modelCounter)}_{0:N}, \bm{y}^{(\modelCounter)}_{1:N};\paramVec\big)$, where $\paramVec$ is a parameter vector that indexes the model.
+\arc{(Perhaps have this next sentence later when $U$ has been defined?)}
 The observed data $\bm{y}_{1:N}^* = y_{1:N,1:U}^*$, along with the unobserved true value of the latent state, are modeled as a realization of this joint distribution.
@@
 Latent states are initialized using an approximation of the instantaneous number of infected, asymptomatic, and recovered individuals at time $t_0$ by using the first week of cholera incidence data.
-Specifically, we set $I_{u0}(0) = \frac{y^*_{1u}}{\reportRate(\muDeath + \choleraDeath + \muIR)}$, $A_{u0}(0) = \frac{1 - \symptomFrac}{\symptomFrac}I_{u0}(0)$, $R_{u0k} = y^*_{1u} - I_{u0}(0) - A_{u0}(0)$, and initialize $B_{u}(0)$ by enforcing the rainfall dynamics supposed by the one step transition model;
+Specifically, we set $I_{u0}(0) = \frac{y^*_{1u}}{\reportRate(\muDeath + \choleraDeath + \muIR)}$, $A_{u0}(0) = \frac{1 - \symptomFrac}{\symptomFrac}I_{u0}(0)$, $R_{u0k} = y^*_{1u} - I_{u0}(0) - A_{u0}(0)$, and initialize $B_{u}(0)$ \arc{(define $B_{u}$?)} by enforcing the rainfall dynamics supposed by the one step transition model;
 all other compartments that represent population counts are set to zero at time $t_0$.
@@
 As such, for departments with zero case counts in week 1, we consider estimating the number of infected individuals rather then treating this value as a constant (see supplement Sec.~S5 for more details).
-Finally, reported cholera cases are assumed to come from a negative binomial measurement model with mean equal to a fraction ($\reportRate$) of individuals in each unit who develop symptoms seek healthcare
+Finally, reported cholera cases are assumed to come from a negative binomial measurement model with mean equal to a fraction ($\reportRate$) of individuals in each unit who develop symptoms and seek healthcare
 (see Sec.~S3.3 of the supplement).
@@
 We perform a statistical test to determine whether or not the data indicate the presence of a linear trend in transmissibility.
-To do this, we perform a profile-likelihood search on the parameter $\transmissionTrend$ and obtain a confidence interval via a Monte Carlo Adjusted Profile (MCAP) \citep{ionides17}.
+To do this, we perform a profile-likelihood search on the parameter $\transmissionTrend$ and obtain a confidence interval \arc{(confidence level?)} via a Monte Carlo Adjusted Profile (MCAP) \citep{ionides17}.
 \cite{lee20} implemented Model~1 by fitting two distinct phases: an epidemic phase from October 2010 through March 2015, and an endemic phase from March 2015 onward.
@@
 Monte Carlo adjusted profile of $\transmissionTrend$.
-The blue curve is the profile, the blue line indicates the MLE, and the dashed lines indicate the confidence interval.
+The blue curve is the profile, the blue line indicates the MLE, and the dashed lines indicate the confidence interval. \arc{level?}
 }
@@
 Because the process model of Model~2 is deterministic, the parameter estimation problem for Model~2 reduces to a least squares calculation when combined with a Gaussian measurement model (see Sec.~S3.2 of the supplement material for details)
-\cite{lee20} fit two versions of model~2 based on a presupposed change in cholera transmission from a epidemic phase to endemic phase that occurred in March, 2014.
+\cite{lee20} fit two versions of Model~2 based on a presupposed change in cholera transmission from a epidemic phase to endemic phase that occurred in March, 2014.
 The inclusion of a change-point in model states and parameters increases the flexibility of the model and hence the ability to fit the observed data.
@@
 The inclusion of a model break-point by \cite{lee20} is perhaps due to a challenging feature of fitting a deterministic model via least squares: discrepancies between model trajectories and observed case counts in highly infectious periods of a disease outbreak will result in greater penalty than the discrepancies between model trajectories and observed case counts in times of relatively low infectiousness, resulting in a bias towards accurately describing periods of high infectiousness.
-This issue is particularly troublesome for modeling cholera dynamics in Haiti: the inability to accurately fit times of low infectiousness may result in poor model forecasts, as few cases of cholera were observed the last few years of the epidemic.
+This issue is particularly troublesome for modeling cholera dynamics in Haiti: the inability to accurately fit times of low infectiousness may result in poor model forecasts, as few cases of cholera were observed in the last few years of the epidemic.
 
@@
 \subsubsection{Calibrating Model~3 Parameters}\label{sec:fit3}
-Model~3 describes cholera dynamics in Haiti using a metapopulation model, where the hidden states in each administrative department has an effect on the dymanics in other departments.
+Model~3 describes cholera dynamics in Haiti using a metapopulation model, where the hidden states in each administrative department has an effect on the dynamics in other departments.
 The decision to address metapopulation dynamics using a spatially explicit model, rather than to aggregate over space, is double-edged.
@@
 
-The use of benchmarks may also be beneficial when developing models at differing spatial scales, where a direct comparison between models likelihoods is meaningless.
+The use of benchmarks may also be beneficial when developing models at differing spatial scales, where a direct comparison between model\ar{s} likelihoods is meaningless.
 In such a case, a benchmark model can be fit to each spatial resolution being considered, and each model compared to their respective benchmark.
@@
 \end{tabular}
-\caption{\label{tab:likes}AIC values for each models compared to their ARMA benchmarks.
-Values in parenthesis are corresponding values using \citet{lee20} parameter estimates.
+\caption{\label{tab:likes}AIC values for each model compared to their ARMA benchmark.
+Values in parentheses are corresponding values using \citet{lee20} parameter estimates.
 $\textsuperscript{1}$The reported likelihood is an upper bound of the likelihood of the \citet{lee20} model.
@@
 If a Bayesian technique is used for parameter estimation, a natural way to account for parameter uncertainty is to obtain simulations from the model where each simulation is obtained using parameters drawn from the estimated posterior distribution.
-For frequentist inference, one possible approach obtaining model forecasts from various values of $\theta$, where the values of $\theta$ are sampled proportionally their corresponding likelihoods \citep{king15}; see Sec.~S7 of the supplement material for more details.
+For frequentist inference, one possible approach is obtaining model forecasts from various values of $\theta$, where the values of $\theta$ are sampled proportionally according to their corresponding likelihoods \citep{king15}; see Sec.~S7 of the supplement material for more details.
 Both of these approaches share the similarity that parameters are chosen for the forecast approximately in proportion to their corresponding value of the likelihood function, $f^{(m)}_{\bm{Y}_{1:N}}(\bm{y}_{1:N}^*; \theta)$.
@@
 While there is remaining time during this projection period in which new cholera infections can be detected, up to this point our estimates are far more consistent with the observed number of reported cholera cases than the corresponding estimates of \cite{lee20}, which were approximately $400,000$ and $1,000,000$.
+\arc{(I don't fully understand how both of our Model 1 and 3 cumulative case estimates can be more consistent with observed cases if our Model 1 has a lower estimate than theirs and our Model 3 has a higher estimate than theirs.)}
 
@@
 The resulting mechanisms in a fitted model can be compared to current scientific knowledge about a system.
-Agreement between model based inference and our current understanding of a system may taken as a confirmation of both model based conclusions and our scientific understanding.
+Agreement between model based inference and our current understanding of a system may be taken as a confirmation of both model based conclusions and our scientific understanding.
 On the other hand, comparisons may generate unexpected results that have the potential to spark new scientific knowledge \citep{ganusov16}.
@@
 We provided a few suggestions that may aid the fitting of mechanistic models such as comparing model likelihoods to a benchmark.
-Improved model fits allows for meaningful statistical inference that may provide valuable insight on a dynamic system and may improve the accuracy of model forecasts.
+Improved model fits allow for meaningful statistical inference that may provide valuable insight on a dynamic system and may improve the accuracy of model forecasts.
 Caution is nonetheless needed when making policy based on modeling conclusions, as model misspecification may invalidate conclusions.
@@
 \citet{lee20} published their code and data online, and this reproducibility facilitated our work.
-Robust data analysis requires not only reproducibility but also extendability: if one wishes to try new model variations, or new approaches to fitting the existing models, or plotting the results in a different way, this should be not excessively burdensome.
+Robust data analysis requires not only reproducibility but also extendability: if one wishes to try new model variations, or new approaches to fitting the existing models, or plotting the results in a different way, this should \ar{be not} not be excessively burdensome.
 Scientific results are only trustworthy so far as they can be critically questioned, and an extendable analysis should facilitate such examination \citep{gentleman07}.

### 2022-12-21 ms.Rnw
tweaks to the abstract

@@
 The 2010 introduction of cholera to Haiti led to an extensive outbreak and sustained transmission.
-Cholera was eliminated in 2019, though it returned in 2022.
-We study three models developed by expert teams to advise on vaccination policies.
-We assess methods used for fitting and evaluating these models, leading to recommendations for future studies.
-Diagnosis of model misspecification and development of alternative models can lead to improved statistical fit, but caution is nevertheless required in drawing policy conclusions based on causal interpretations of the models.
+Cholera was considered eliminated in 2019, though it reemerged in 2022.
+We study three dynamic models developed by expert teams to advise on vaccination policies.
+We assess previous methods used for fitting and evaluating these models, and we develop data analysis strategies leading to improved statistical fit.
+Specifically, we present approaches to diagnosis of model misspecification, development of alternative models, and computational improvements in optimization, in the context of likelihood-based inference on nonlinear dynamic systems.
+Our workflow is reproducible and extendable, facilitating future investigations of this disease system.
 \end{abstract}

### 2023-01-04 ms.Rnw
Implemented changes from the break

@@
 \newcommand\fixed{\color{blue}}
-\newcommand\demography{D}
+\newcommand\demography{\bullet}
 \newcommand\code[1]{\texttt{#1}}
@@
 This necessitates addressing the tradeoff between biological fidelity and model simplicity, and the reality of misspecification for models at all levels of complexity.
-As a case study, we consider a cholera epidemic in Haiti.
-The 2010 introduction of cholera to Haiti led to an extensive outbreak and sustained transmission.
-Cholera was considered eliminated in 2019, though it reemerged in 2022.
+As a case study, we consider the 2010-2019 cholera epidemic in Haiti.
 We study three dynamic models developed by expert teams to advise on vaccination policies.
@@
 Model~4 has an agent-based construction, featuring considerable mechanistic detail but limited ability to calibrate these details to data.
-We focus on Models~1--3, as the strengths and weaknesses of the agent-based modeling approach  \citep{tracy18} are outside the scope of this article, and we focus on Models~1--3.
+We focus on Models~1--3, as the strengths and weaknesses of the agent-based modeling approach  \citep{tracy18} are outside the scope of this article.
 
@@
 To successfully predict the effect of an intervention, a model should therefore both provide a quantitative explanation of existing data and should have a causal interpretation: a manipulation of the system should correspond quantitatively with the corresponding change to the model.
-This motivates the development of mechanistic statistical models, which provides a statistical fit to the available data while also supporting a causal interpretation.
+This motivates the development of mechanistic models, which provides a statistical fit to the available data while also supporting a causal interpretation.
 
@@
 Here, we have put all three models into a common notational framework.
-Translations back to the original notation of \citet{lee20} are given in Table~S-1.
+Translations back to the original notation of \citet{lee20} are given in Table~S-1 of the supplement.
 
@@
 Together, with the density for the initial value of the latent state $\bm{X}^{(\modelCounter)}_{0} = \bm{X}^{(\modelCounter)}(t_0)$, each model assumes a joint density $f_{\bm{X}^{(\modelCounter)}_{0:N}, \bm{Y}^{(\modelCounter)}_{1:N}}\big(\bm{x}^{(\modelCounter)}_{0:N}, \bm{y}^{(\modelCounter)}_{1:N};\paramVec\big)$, where $\paramVec$ is a parameter vector that indexes the model.
-\arc{(Perhaps have this next sentence later when $U$ has been defined?)}
-The observed data $\bm{y}_{1:N}^* = y_{1:N,1:U}^*$, along with the unobserved true value of the latent state, are modeled as a realization of this joint distribution.
+The observed data $\bm{y}_{1:N}^*$, along with the unobserved true value of the latent state, are modeled as a realization of this joint distribution.
 
 For each model $\modelCounter \in \{1, 2, 3\}$, the latent state vector $\bm{X}^{(\modelCounter)}(t)$ consists of individuals labeled as susceptible (S), infected (I), asymptomatically infected (A), vaccinated (V), and recovered (R), with various sub-divisions sometimes considered.
-Models~2 and~3 have metapopulation structure, meaning that each individual is a member of a spatial unit, denoted by a subscript $u\in \seq{1}{U}$.
+Models~2 and~3 have metapopulation structure, meaning that each individual is a member of a spatial unit, denoted by a subscript $u\in \seq{1}{U}$, in which case we denote the observed data for each unit using $\bm{y}_{1:N}^* = y_{1:N,1:U}^*$.
 Here, the spatial units are the $U=10$ Haitian administrative d\'{e}partements (henceforth anglicized as departments).
@@
 A first-time reader may choose to skim through the rest of this section, and return later.
-Additional details about the numeric implementation of these models are provided in Secs.~S1 and S2 of the supplement.
+Additional details about the numeric implementation of these models are provided in Sec.~S1 and S2 of the supplement.
 
@@
 Latent states are initialized using an approximation of the instantaneous number of infected, asymptomatic, and recovered individuals at time $t_0$ by using the first week of cholera incidence data.
-Specifically, we set $I_{u0}(0) = \frac{y^*_{1u}}{\reportRate(\muDeath + \choleraDeath + \muIR)}$, $A_{u0}(0) = \frac{1 - \symptomFrac}{\symptomFrac}I_{u0}(0)$, $R_{u0k} = y^*_{1u} - I_{u0}(0) - A_{u0}(0)$, and initialize $B_{u}(0)$ \arc{(define $B_{u}$?)} by enforcing the rainfall dynamics supposed by the one step transition model;
+Specifically, we set $I_{u0}(0) = \frac{y^*_{1u}}{\reportRate(\muDeath + \choleraDeath + \muIR)}$, $A_{u0}(0) = \frac{1 - \symptomFrac}{\symptomFrac}I_{u0}(0)$, $R_{u0k} = y^*_{1u} - I_{u0}(0) - A_{u0}(0)$, and initialize $W_{u}(0)$ by enforcing the rainfall dynamics supposed by the one step transition model;
 all other compartments that represent population counts are set to zero at time $t_0$.
@@
 We perform a statistical test to determine whether or not the data indicate the presence of a linear trend in transmissibility.
-To do this, we perform a profile-likelihood search on the parameter $\transmissionTrend$ and obtain a confidence interval \arc{(confidence level?)} via a Monte Carlo Adjusted Profile (MCAP) \citep{ionides17}.
+To do this, we perform a profile-likelihood search on the parameter $\transmissionTrend$ and obtain a $95\%$ confidence interval via a Monte Carlo Adjusted Profile (MCAP) \citep{ionides17}.
 \cite{lee20} implemented Model~1 by fitting two distinct phases: an epidemic phase from October 2010 through March 2015, and an endemic phase from March 2015 onward.
@@
 Monte Carlo adjusted profile of $\transmissionTrend$.
-The blue curve is the profile, the blue line indicates the MLE, and the dashed lines indicate the confidence interval. \arc{level?}
+The blue curve is the profile, the blue line indicates the MLE, and the dashed lines indicate the $95\%$ confidence interval.
 }
@@
 % To address this issue, \cite{lee20} conditioned each department level model on reported cholera cases in other departments and approximated the values of $I_\nu(t)$ and $A_\nu(t)$ using the respective reported cases.
-Because the simplified, spatially-decoupled version of Model~3 implemented by \cite{lee20} relies on the observed cholera cases it is impossible to obtain forecasts for future time points from the fitted model.
+Because the simplified, spatially-decoupled version of Model~3 implemented by \cite{lee20} relies on the observed cholera cases, the calibrated model cannot be used to obtain forecasts.
 Therefore, in order to obtain model forecasts, \cite{lee20} used the parameters estimates from the spatially-decoupled approximation of Model~3 to obtain forecasts using the fully coupled version of the model.
@@
 
-The use of benchmarks may also be beneficial when developing models at differing spatial scales, where a direct comparison between model\ar{s} likelihoods is meaningless.
+The use of benchmarks may also be beneficial when developing models at differing spatial scales, where a direct comparison between model likelihoods is meaningless.
 In such a case, a benchmark model can be fit to each spatial resolution being considered, and each model compared to their respective benchmark.
@@
 While a useful practice, comparing model log-likelihoods to a suitable benchmark alone may not be sufficient to diagnosing the goodness-of-fit of a given model.
-Additional techniques include the inspection of conditional log-likelihoods of each observation given the previous observations in order to understand how well the model describes each data point.
+Additional techniques include the inspection of conditional log-likelihoods of each observation given the previous observations in order to understand how well the model describes each data point (Sec.~S5).
 Other tools include plotting the effective sample size of each observation \citep{liu01}; plotting the values of the hidden states from simulations (supplement Sec.~5.1);
@@
 Forecasts are an attempt to provide an accurate estimate of the future state of a system based on currently available data, together with an assessment of uncertainty.
-Mechanistic forecasting models, compatible with current scientific understanding, may also provide estimates of the future effects of potential interventions.
+Forecasts from mechanistic models that are compatible with current scientific understanding may also provide estimates of the future effects of potential interventions.
 Further, they may enable real-time testing of new scientific hypotheses \citep{lewis22}.
@@
 where $\hat\theta$ is a vector of calibrated parameters.
-The decision not to do this partially explains the unsuccessful forecasts of \cite{lee20}: their Table~S7 shows that the subset of their simulations which were consistent with observing zero cases in 2019 predicted the elimination of cholera.
-In this case study, projecting the future state of the cholera epidemic starting from latent states that are draws from the filtering distribution allows the model-based forecasts to benefit from the fact that very few cholera cases were observed in 2018 and January 2019.
+The decision not to do this partially explains the unsuccessful forecasts of \cite{lee20}: their Table~S7 shows that the subset of their simulations which were consistent with observing zero cases in 2019 also accurately predicted the prolonged absence of detected cholera.
 
@@
 Both of these approaches share the similarity that parameters are chosen for the forecast approximately in proportion to their corresponding value of the likelihood function, $f^{(m)}_{\bm{Y}_{1:N}}(\bm{y}_{1:N}^*; \theta)$.
-In this analysis, our forecasts do not account for parameter uncertainty as our goal is to comment on the estimation and diagnosis of mechanistic models, rather than suggesting our results to be used to influence vaccination policy in Haiti.
+In this analysis, our forecasts do not account for parameter uncertainty as our goal is to comment on the estimation and diagnosis of mechanistic models, rather than influencing vaccination policy in Haiti.
 Furthermore, we use the projections from a single point estimate to highlight a deficiency of deterministic models, which is that the only variability in model projections is a result of parameter uncertainty, which leads to over-confidence in forecasts \cite{king15}.
@@
 The primary forecasting goal of \citet{lee20} was to investigate the potential consequences of vaccination interventions on a system to inform policy.
-Outcomes of their study include estimates for the probability of cholera elimination and cumulative number of cholera infections under several possible vaccination scenarios.
-Mimicking their efforts, we define cholera elimination as having less than one infection of cholera over at least 52 consecutive weeks, and provide forecasts under the following vaccination scenarios:
+One outcome of their study include estimates for the probability of cholera elimination under several possible vaccination scenarios.
+Mimicking their efforts, we define cholera elimination as having no cholera infections over at least 52 consecutive weeks, and provide forecasts under the following vaccination scenarios:
 
@@
 Estimates of the probability of cholera elimination can therefore be obtained as the proportion of simulations from these models that result in cholera elimination.
-The results of these projections are summarized in Figs.~\ref{fig:Mod1Scenarios}--\ref{fig:elimProbs}, and suggest that cholera elimination was likely even without increased vaccination efforts---consistent with observed reality \citep{trevisin22}.
+The results of these projections are summarized in Figs.~\ref{fig:Mod1Scenarios}--\ref{fig:elimProbs}.
 
 Probability of elimination estimates of this form are not meaningful for deterministic models, as the trajectory of these models only represent the mean behavior of the system rather than individual potential outcomes.
-We therefore do not provide probability of elimination estimates under Model~2.
-Still, trajectories obtained by Model~2 are consistent with the simulation results of Models~1 and~3, and suggest that cholera was in the process of being eliminated from Haiti.
-
-In additional to probability of elimination estimates, we provide estimates for the cumulative number of infections under each vaccination scenario from February 2019 -- February 2024.
-Notably, the median number of cumulative cholera infections under the no-vaccination scenario using Models 1 and 3 were $\Sexpr{prettyNum(round(as.numeric(h1_S0_results$mod1_V0_probs$cumInf['q50'])), big.mark = ",")}$ and $\Sexpr{prettyNum(round(as.numeric(mod3_V0_res$mod3_V0_probs$cumInf['q50'])), big.mark = ",")}$, respectively.
-While there is remaining time during this projection period in which new cholera infections can be detected, up to this point our estimates are far more consistent with the observed number of reported cholera cases than the corresponding estimates of \cite{lee20}, which were approximately $400,000$ and $1,000,000$.
-\arc{(I don't fully understand how both of our Model 1 and 3 cumulative case estimates can be more consistent with observed cases if our Model 1 has a lower estimate than theirs and our Model 3 has a higher estimate than theirs.)}
+We therefore do not provide probability of elimination estimates under Model~2, but show trajectories under the various vaccination scenarios using this model (Fig.~\ref{fig:mod2Traj}).
 
@@
 This is consistent with previous studies finding that rainfall played an important role in cholera transmission in Haiti \citep{lemaitre19,eisenberg13}.
-The estimated seasonality also features an increased transmission rate during the Fall, which was noticed at an earlier stage of the epidemic \citep{rinaldo12}.
+The estimated seasonality also features an increased transmission rate during the Fall, which was noticed at an earlier stage of the epidemic \citep{rinaldo12}; the high transimission rate in the fall may also be a result of the increase transission that occured in the fall of 2016, when hurricane Matthew struck Haiti.
 
@@
 Our case study provides an example of this in the parameter estimate for the duration of natural immunity due to cholera infection $\muRS^{-1}$.
-Under the framework of Model~2, the best estimate for this parameter is $\Sexpr{signif(1/h2_fit$h2_params["sigma"],3)}$, suggesting that individuals have effectively permanent immunity to cholera once infected, and thus the pool of susceptible individuals decreases over time.
+Under the framework of Model~2, the best estimate for this parameter is $\Sexpr{signif(1/h2_fit$h2_params["sigma"],3)}$, suggesting that individuals have effectively permanent immunity to cholera once infected.
 To interpret this result, we bear in mind that the data ranged from 2010-2019, and therefore estimates of immunity longer than 10 years---the upper end of previous estimates of natural immunity \citep{king08}---effectively result in the same model dynamics.
@@
 For example, reconstructed latent variables (such as numbers of susceptible and recovered individuals) should make sense in the context of alternative measurements of these variables \citep{grad12}.
-In addition, parameters that have been calibrated to data should make sense in the context of alternative lines of evidence about the phenomena being modeled, while making allowance for the possibility that the interpretations of parameters may vary when modeling across differing spatial scales.
+In the context of our case study, by reconstructing the latent states of Model~3, we notice that the calibrated model favors higher levels of cholera transmission than what was typically observed in the incidence data (Sec.~S5.1 of the supplement).
+This result hints at the possibility of model mispecification, and should prompt us to be sceptical of the reliability of forecasts from this model.
+Similarly, parameters that have been calibrated to data should make sense in the context of alternative lines of evidence about the phenomena being modeled, while making allowance for the possibility that the interpretations of parameters may vary when modeling across differing spatial scales.
 
@@
 
-In this article we argue that careful attention to important statistical details could have led the models proposed by \cite{lee20} to correctly predict the initial elimination of cholera from Haiti in 2019.
-We acknowledge the benefit of hindsight: our demonstration of a statistically principled route to obtain better-fitting models with more accurate predictions does not rule out the possibility of discovering other models that fit well yet predict poorly.
+In this article we argue that careful attention to important statistical details could have resulted in improved statistical fits to the observed incidence data.
+We acknowledge the benefit of hindsight: our demonstration of a statistically principled route to obtain better-fitting models resulting in more robust insights does not rule out the possibility of discovering other models that fit well yet predict poorly.
 We used the same data and models, and even much of the same code, as \citet{lee20}, and yet ended up with drastically different conclusions.
@@
 \citet{lee20} published their code and data online, and this reproducibility facilitated our work.
-Robust data analysis requires not only reproducibility but also extendability: if one wishes to try new model variations, or new approaches to fitting the existing models, or plotting the results in a different way, this should \ar{be not} not be excessively burdensome.
+Robust data analysis requires not only reproducibility but also extendability: if one wishes to try new model variations, or new approaches to fitting the existing models, or plotting the results in a different way, this should not be excessively burdensome.
 Scientific results are only trustworthy so far as they can be critically questioned, and an extendable analysis should facilitate such examination \citep{gentleman07}.

### 2023-01-04 si/si.Rnw
Implemented changes from the break

@@
 \newcommand\fixed{\color{blue}}
-\newcommand\demography{D}
+\newcommand\demography{\bullet}
 
@@
 The basic characteristic of a compartment model is that $X(t)$ can be written in terms of the flows $N_{ij}(t)$ from $i$ to $j$.
-A flow into compartment $i$ from outside the system is denoted by $N_{\circ i}$, and a flow out of the system from compartment $i$ is denoted by $N_{i\circ}$.
-We call $\circ$ a source/sink compartment, though it is an irregular compartment since $X_{circ}(t)$ is not defined.
+A flow into compartment $i$ from outside the system is denoted by $N_{\demography i}$, and a flow out of the system from compartment $i$ is denoted by $N_{i\demography}$.
+We call $\demography$ a source/sink compartment, though it is an irregular compartment since $X_{\demography}(t)$ is not defined.
 These flows are required to satisfy a ``conservation of mass'' identity:
 \begin{equation}
-X_i(t)=X_i(t_0)+N_{\bullet i}(t) - N_{i\bullet}(t) + \sum_{j\neq
+X_i(t)=X_i(t_0)+N_{\demography i}(t) - N_{i\demography}(t) + \sum_{j\neq
 i}N_{ji}(t)-\sum_{j\neq i}N_{ij}(t). \label{eq:conservation}
@@
 Each {\em flow} $N_{ij}(t)$ is associated with a {\em rate} function
-$\mu_{ij}=\mu_{ij}(t,X(t))$, where we include the possibility that $i$ or $j$ takes value $\bullet$.
+$\mu_{ij}=\mu_{ij}(t,X(t))$, where we include the possibility that $i$ or $j$ takes value $\demography$.
 
@@
 \label{eq:ode1}
-dN_{ij}/dt = \mu_{ij}\big(t,X(t) \big) X_i(t), \hspace{3mm} i\in \seq{1}{c}, \hspace{3mm} j\in \seq{1}{c} \cup\{\bullet\}, \hspace{3mm} i\neq j,
+dN_{ij}/dt = \mu_{ij}\big(t,X(t) \big) X_i(t), \hspace{3mm} i\in \seq{1}{c}, \hspace{3mm} j\in \seq{1}{c} \cup\{\demography\}, \hspace{3mm} i\neq j,
 \end{equation}
 where $\mu_{ij}\big(t,X(t)\big)$ is called a per-capita rate or a unit rate.
-Flows into the system require special treatment since $X_i(t)$ in \myeqref{eq:ode1} is not defined for $i=\bullet$.
+Flows into the system require special treatment since $X_i(t)$ in \myeqref{eq:ode1} is not defined for $i=\demography$.
 Instead, we specify
@@
 \label{eq:ode2}
-dN_{\bullet i}/dt = \mu_{\bullet i}\big(t,X(t) \big).
+dN_{\demography i}/dt = \mu_{\demography i}\big(t,X(t) \big).
 \end{equation}
@@
 \end{eqnarray}
-for $i\in \seq{1}{c}$ and $j\in\seq{1}{c}\cup\{\bullet\}$ with $i\neq j$.
+for $i\in \seq{1}{c}$ and $j\in\seq{1}{c}\cup\{\demography\}$ with $i\neq j$.
 As with the ODE case, we need special attention for flows into the system, and we define
@@
 \label{eq:smcs3}
-\prob\big[ N_{\bullet i}(t+\delta)-N_{\bullet i}(t)=0 \given X(t)\big]
- &=& 1-\delta \mu_{\bullet i}\big(t,X(t)\big) + o(\delta),
+\prob\big[ N_{\demography i}(t+\delta)-N_{\demography i}(t)=0 \given X(t)\big]
+ &=& 1-\delta \mu_{\demography i}\big(t,X(t)\big) + o(\delta),
 \\
 \label{eq:smcs4}
-\prob\big[ N_{\bullet i}(t+\delta)-N_{\bullet i}(t)=1 \given X(t)\big]
- &=& \delta \mu_{\bullet i}\big(t,X(t)\big) + o(\delta).
+\prob\big[ N_{\demography i}(t+\delta)-N_{\demography i}(t)=1 \given X(t)\big]
+ &=& \delta \mu_{\demography i}\big(t,X(t)\big) + o(\delta).
 \end{eqnarray}
@@
 Write $\delta$ for an Euler time step, and $\Delta N_{ij}$ for the numerical approximation to $N_{ij}(t+\delta)-N_{ij}(t)$ given $X(t)$.
-For each $i$ and $j$ in $\seq{1}{c} \cup \{\bullet\}$ with $i \neq j$, we draw independent Gamma distributed noise increments with mean $\delta$ and variance $\sigma_{ij}^2 \delta$, denoted using a mean-variance parameterization of the gamma distribution as
+For each $i$ and $j$ in $\seq{1}{c} \cup \{\demography\}$ with $i \neq j$, we draw independent Gamma distributed noise increments with mean $\delta$ and variance $\sigma_{ij}^2 \delta$, denoted using a mean-variance parameterization of the gamma distribution as
 \begin{equation}
@@
 In the case of an SMCS model, $\sigma_{ij}=0$ for all $i$ and $j$, so we have $\Delta\Gamma_{ij}=\delta$.
-Then, for $i\neq\bullet$ and $j\neq i$, and writing
+Then, for $i\neq \demography$ and $j\neq i$, and writing
 \begin{equation}
@@
 \label{eq:numerics3}
-p_{ij} &=& \exp\left\{-\sum_{k\in 1:c \, \cup \{\bullet\}} \mu_{ik} \, \delta \right\}
-\frac{\mu_{ij}}{\sum_{k\in 1:c\cup \{\bullet\}} \mu_{ik}},
+p_{ij} &=& \exp\left\{-\sum_{k\in 1:c \, \cup \{\demography\}} \mu_{ik} \, \delta \right\}
+\frac{\mu_{ij}}{\sum_{k\in 1:c\cup \{\demography\}} \mu_{ik}},
 \\
@@
 These probabilities correspond to competing hazards for every individual in compartment $i$ to transition to some compartment $j$, interpreting $j=i$ to mean that the individual remains in $i$.
-Then, $\big(\Delta N_{i1},\dots,\Delta N_{ic},\Delta N_{i\bullet}\big)$ has the multinomial distribution where $X_i(t)$ individuals are allocated independently to $\seq{1}{c}\cup\{\bullet\}$ with probabilities given by \eqref{eq:numerics3} and \eqref{eq:numerics4}.
+Then, $\big(\Delta N_{i1},\dots,\Delta N_{ic},\Delta N_{i\demography}\big)$ has the multinomial distribution where $X_i(t)$ individuals are allocated independently to $\seq{1}{c}\cup\{\demography\}$ with probabilities given by \eqref{eq:numerics3} and \eqref{eq:numerics4}.
 We use the \code{reulermultinom} function in the \code{pomp} package to draw from this multionomial distribution.
@@
 Different treatments of demographic flows---such as birth, death, immigration and emigration---are possible.
-For the case $i=\bullet$, the treatment used by Model~1 is to set
+For the case $i=\demography$, the treatment used by Model~1 is to set
 \begin{equation}
 \label{eq:numerics6}
-\Delta N_{\bullet j} \sim \mathrm{poisson}( \mu_{\bullet j} \delta),
+\Delta N_{\demography j} \sim \mathrm{poisson}( \mu_{\demography j} \delta),
 \end{equation}
-an independent Poisson random variable with mean $\mu_{\bullet j} \delta$.
-
-Models~2 and 3 used an alternative approach, balancing the total number of flows in and out of the compartment, i.e., $\sum_{i}N_{\bullet i}(t) = \sum_{i}N_{i \bullet}(t)$, in order to make the model consistent with the known total population.
-In this case, we formally model the death rate as a rate of returning to the susceptible class $S$, and use external transitions from $\bullet$ into $S$ to describe only net population increase.
+an independent Poisson random variable with mean $\mu_{\demography j} \delta$.
+
+Models~2 and 3 used an alternative approach, balancing the total number of flows in and out of the compartment, i.e., $\sum_{i}N_{\demography i}(t) = \sum_{i}N_{i \demography}(t)$, in order to make the model consistent with the known total population.
+In this case, we formally model the death rate as a rate of returning to the susceptible class $S$, and use external transitions from $\demography$ into $S$ to describe only net population increase.
 

### 2023-01-05 ms.Rnw
some minor edits

@@
 where $\mathrm{Pop}_u$ is the mean population for department $u$,
-$D_{u\varv}$ is a distance measure estimating average road distance between randomly chosen members of each population, and $v_{\mathrm{rate}}= 10^{-12}$ was treated as a fixed constant.
+$D_{u\varv}$ is a distance measure estimating average road distance between randomly chosen members of each population, and $v_{\mathrm{rate}}= 10^{-12}$ \eic{UNITS?} was treated as a fixed constant.
 In \eqref{model2:water_transport}, $T^W_{u\varv}$ is a measure of river flow between departments.
@@
 Latent states are initialized using an approximation of the instantaneous number of infected, asymptomatic, and recovered individuals at time $t_0$ by using the first week of cholera incidence data.
-Specifically, we set $I_{u0}(0) = \frac{y^*_{1u}}{\reportRate(\muDeath + \choleraDeath + \muIR)}$, $A_{u0}(0) = \frac{1 - \symptomFrac}{\symptomFrac}I_{u0}(0)$, $R_{u0k} = y^*_{1u} - I_{u0}(0) - A_{u0}(0)$, and initialize $W_{u}(0)$ by enforcing the rainfall dynamics supposed by the one step transition model;
+Specifically, we set $I_{u0}(0) = \frac{y^*_{1u}}{\reportRate(\muDeath + \choleraDeath + \muIR)}$, $A_{u0}(0) = \frac{1 - \symptomFrac}{\symptomFrac}I_{u0}(0)$, $R_{u0k} = y^*_{1u} - I_{u0}(0) - A_{u0}(0)$, and we initialize $W_{u}(0)$ by enforcing the rainfall dynamics supposed by the one step transition model;
 all other compartments that represent population counts are set to zero at time $t_0$.
 For each unit $u$ with zero case counts at time $t_1$, this initialization scheme results in having zero individuals in the Infected and Asymptomatic compartments, as well as having no bacteria in the aquatic reservoir.
-In reality, it's plausible that some bacteria or infected individuals were present in unit $u$ but simply went unreported.
-As such, for departments with zero case counts in week 1, we consider estimating the number of infected individuals rather then treating this value as a constant (see supplement Sec.~S5 for more details).
+In reality, it is plausible that some bacteria or infected individuals were present in unit $u$ but went unreported.
+Therefore, for departments with zero case counts in week 1, we consider estimating the number of infected individuals rather then treating this value as a constant (see supplement Sec.~S5 for more details).
 Finally, reported cholera cases are assumed to come from a negative binomial measurement model with mean equal to a fraction ($\reportRate$) of individuals in each unit who develop symptoms and seek healthcare
@@
 Calibrating model parameters to observed data does not guarantee that the resulting model successfully approximates real-world mechanisms, since model assumptions may be incorrect, and do not change as the model is calibrated to data.
-However, the congruency between the model and observed data serves as a proxy for the congruency between the model and the true underlying dynamic system, since the observed data are a result of the generating system.
+However, the congruency between the model and observed data serves as a proxy for the congruency between the model and the true underlying dynamic system.
 As such, the best possible fit of the proposed mechanistic structure to the observed data is desirable.
@@
 The proposal of a linear trend in transmission is a result of observing an apparent decrease in reported cholera infections from 2012-2019 in Fig.~\ref{CholeraData}.
-While several factors may contribute to this decrease, one explanation is that case-area targeted interventions (CATIs), which included education sessions, increased monitoring, household decontamination, soap distribution, and water chlorination in infected areas \citep{rebaudet19CATI}, may have greatly reduced cholera transmission \citep{rebaudet21}.
+While several factors may contribute to this decrease, one explanation is that case-area targeted interventions (CATIs), which included education sessions, increased monitoring, household decontamination, soap distribution, and water chlorination in infected areas \citep{rebaudet19CATI}, may have substantially reduced cholera transmission \citep{rebaudet21}.
 
@@
 
-We implemented Model~1 using the \code{pomp} package \cite{pomp2}, relying heavily on the \code{pomp} source code provided by \cite{lee20}.
+We implemented Model~1 using the \code{pomp} package \citep{pomp2}, relying heavily on the \code{pomp} source code provided by \cite{lee20}.
 Both analyses used the \code{mif2} implementation of the IF2 algorithm to estimate $\paramVec$ by maximum likelihood.
 One change we made in the statistical analysis that led to larger model likelihoods was increasing the computational effort in the numerical maximization.
-While IF2 is a powerful method that can be used to fit parameters for a large class of models, the theoretic ability to maximize the likelihood depends on asymptotics in both the number of particles and the number of filtering iterations and therefore requires a large number of computations.
+While IF2 enables parameter estimation for a large class of models, the theoretic ability to maximize the likelihood depends on asymptotics in both the number of particles and the number of filtering iterations.
+Many Monte Carlo replications are then required to quantify and further reduce the error. 
 The large difference in likelihoods that were obtain by these two approaches (see Table~\ref{tab:likes}) highlights the importance of carefully determining the necessary computational effort needed to maximize model likelihoods and acting accordingly.
@@
 and can be justified by asymptotic considerations in a large-population limit \citep{dadlani2020,ndii17}.
-Because the process model of Model~2 is deterministic, the parameter estimation problem for Model~2 reduces to a least squares calculation when combined with a Gaussian measurement model (see Sec.~S3.2 of the supplement material for details)
+Because the process model of Model~2 is deterministic, the parameter estimation problem for Model~2 reduces to a least squares calculation when combined with a Gaussian measurement model (see Sec.~S3.2 of the supplement material for details).
 \cite{lee20} fit two versions of Model~2 based on a presupposed change in cholera transmission from a epidemic phase to endemic phase that occurred in March, 2014.
@@
 To combat this issue, we fit the model to log-transformed case counts, since the log scale stabilizes the variation during periods of high and low incidence.
-Other solutions include changing the measurement model to include overdispersion, as was done in Models~1 and~3.
+An alternative solution is to change the measurement model to include overdispersion, as was done in Models~1 and~3.
 This permits the consideration of demographic stochasticity, which is dominant for small infected populations, together with log scale stochasticity (also called multiplicative, or environmental, or extra-demographic) which is dominant at high population counts.
-Here we chose to fit the model to transformed case counts rather than adding overdispersion to the measurement model with the goal of making the fewest changes to the model proposed by \cite{lee20} as possible.
+Here we chose to fit the model to transformed case counts rather than adding overdispersion to the measurement model with the goal of minimizing the changes to the model proposed by \cite{lee20}.
 
@@
 However, a legitimate preference for simplicity \citep{saltelli20,green15} can support a decision to consider nationally aggregated models.
-One issue that arises when fitting spatially explicit models is that parameter estimation techniques based on the particle filter become computationally intractable as the number of spatial units increase.
+One issue that arises when fitting spatially explicit models is that parameter estimation techniques based on the particle filter become computationally intractable as the number of spatial units increases.
 This is a result of the approximation error of particle filters growing exponentially in the dimension of the model \citep{rebeschini15,park20}.
 
-% Model~3 parameters that are calibrated to data are primarily shared between each department, meaning that the estimated parameter value for department $u$ is the same value as the parameter value in department $v$, with $u, v \in \seq{1}{U}$.
-% The exception to this being the parameters $\beta_{W_u}$, and $\beta_u$, which take department-specific values for each department $u$.
 To avoid the approximation error present in high-dimensional models, \citet{lee20} simplified the problem of estimating the parameters of Model~3 by creating an approximate version of the model where the units are independent given the observed data.
-% and fitting using the IF2 algorithm to estimate parameters separately for independent department-level models;
-% the shared parameters were calibrated using the cholera incidence data from Artibonite, and the department-specific parameters ($\beta_{W_u}$ and $\beta_u$) were fit using the data from their respective department.
 Reducing a spatially coupled model to individual units in this fashion requires special treatment of any interactive mechanisms between spatial units, such as found in Eq.~\myeqref{eq:model3:foi}.
-% In particular, when considering a model for department $u$, the values $I_\nu(t)$ and $A_\nu(t)$, are unknown for $u \neq v$.
-% To address this issue, \cite{lee20} conditioned each department level model on reported cholera cases in other departments and approximated the values of $I_\nu(t)$ and $A_\nu(t)$ using the respective reported cases.
-Because the simplified, spatially-decoupled version of Model~3 implemented by \cite{lee20} relies on the observed cholera cases, the calibrated model cannot be used to obtain forecasts.
+Because the simplified, spatially-decoupled version of Model~3 implemented by \cite{lee20} relies on the observed cholera cases, the calibrated model cannot readily be used to obtain forecasts.
 Therefore, in order to obtain model forecasts, \cite{lee20} used the parameters estimates from the spatially-decoupled approximation of Model~3 to obtain forecasts using the fully coupled version of the model.
@@
 Another concern is that the two versions of the model are not the same, resulting in sub-optimal parameter estimates for the spatially coupled model,
-as set of parameters that maximize the likelihood of the decoupled model almost certainly do not maximize the likelihood of the fully coupled model.
+as parameters that maximize the likelihood of the decoupled model almost certainly do not maximize the likelihood of the fully coupled model.
 These two concerns may explain the unrealistic forecasts and low likelihood of the model (see Table~\ref{tab:likes}).
@@
 Associative models are not constrained to have a causal interpretation, and typically are designed with the sole goal of providing a statistical fit to data.
-Consequently, we should not require a candidate mechanistic model to beat all benchmarks.
+Therefore, we should not require a candidate mechanistic model to beat all benchmarks.
 However, a mechanistic model which falls far short against benchmarks is evidently failing to explain some substantial aspect of the data.
@@
 It should be universal practice to present measures of goodness of fit for published models, and mechanistic models should be compared against benchmarks.
-In our literature review of the Haiti cholera epidemic, no benchmark models were considered in any of the 32 papers that used dynamic models to describe the cholera in order to obtain scientific conclusions.
+In our literature review of the Haiti cholera epidemic, no non-mechanistic benchmark models were considered in any of the 32 papers that used dynamic models to describe cholera in order to obtain scientific conclusions.
 Including benchmarks would help authors and readers to detect and confront any major statistical limitations of the proposed mechanistic models.
@@
 In such a case, a benchmark model can be fit to each spatial resolution being considered, and each model compared to their respective benchmark.
-Large advantages (shortcomings) in model likelihood relative to the benchmark for a given spatial scale that are not present in other spatial scales may provide weak evidence for (against) the statistical fit of models across a range of spatial resolutions.
+Large advantages (or shortcomings) in model likelihood relative to the benchmark for a given spatial scale that are not present in other spatial scales may provide weak evidence for (or against) the statistical fit of models across a range of spatial resolutions.
 
@@
 
-While a useful practice, comparing model log-likelihoods to a suitable benchmark alone may not be sufficient to diagnosing the goodness-of-fit of a given model.
+Comparing model log-likelihoods to a suitable benchmark may not be sufficient to identify all the strengths and weaknesses of a given model.
 Additional techniques include the inspection of conditional log-likelihoods of each observation given the previous observations in order to understand how well the model describes each data point (Sec.~S5).
@@
 Both of these approaches share the similarity that parameters are chosen for the forecast approximately in proportion to their corresponding value of the likelihood function, $f^{(m)}_{\bm{Y}_{1:N}}(\bm{y}_{1:N}^*; \theta)$.
-In this analysis, our forecasts do not account for parameter uncertainty as our goal is to comment on the estimation and diagnosis of mechanistic models, rather than influencing vaccination policy in Haiti.
-Furthermore, we use the projections from a single point estimate to highlight a deficiency of deterministic models, which is that the only variability in model projections is a result of parameter uncertainty, which leads to over-confidence in forecasts \cite{king15}.
+In this analysis, we do not construct forecasts accounting for parameter uncertainty as our focus is on the estimation and diagnosis of mechanistic models.
+Furthermore, we use the projections from a single point estimate to highlight the deficiency of deterministic models that the only variability in model projections is a result of parameter uncertainty, which can lead to over-confidence in forecasts \cite{king15}.
 
@@
 One outcome of their study include estimates for the probability of cholera elimination under several possible vaccination scenarios.
-Mimicking their efforts, we define cholera elimination as having no cholera infections over at least 52 consecutive weeks, and provide forecasts under the following vaccination scenarios:
+Mimicking their approach, we define cholera elimination as an absence of cholera infections for at least 52 consecutive weeks, and we provide forecasts under the following vaccination scenarios:
 
@@
 This is consistent with previous studies finding that rainfall played an important role in cholera transmission in Haiti \citep{lemaitre19,eisenberg13}.
-The estimated seasonality also features an increased transmission rate during the Fall, which was noticed at an earlier stage of the epidemic \citep{rinaldo12}; the high transimission rate in the fall may also be a result of the increase transission that occured in the fall of 2016, when hurricane Matthew struck Haiti.
+The estimated seasonality also features an increased transmission rate during the fall, which was noticed at an earlier stage of the epidemic \citep{rinaldo12}; the high transimission rate in the fall may also be a result of the increase transission that occured in the fall of 2016, when hurricane Matthew struck Haiti.
 
@@
 
-Our case study provides an example of this in the parameter estimate for the duration of natural immunity due to cholera infection $\muRS^{-1}$.
-Under the framework of Model~2, the best estimate for this parameter is $\Sexpr{signif(1/h2_fit$h2_params["sigma"],3)}$, suggesting that individuals have effectively permanent immunity to cholera once infected.
-To interpret this result, we bear in mind that the data ranged from 2010-2019, and therefore estimates of immunity longer than 10 years---the upper end of previous estimates of natural immunity \citep{king08}---effectively result in the same model dynamics.
+Our case study provides an example of this in the parameter estimate for the duration of natural immunity due to cholera infection, $\muRS^{-1}$.
+Under the framework of Model~2, the best estimate for this parameter is $\Sexpr{signif(1/h2_fit$h2_params["sigma"],3)} \, \mathrm{yr}$, suggesting that individuals have effectively permanent immunity to cholera once infected.
+To interpret this result, we bear in mind that the data ranged from 2010-2019, and therefore estimates of immunity longer than $10\, \mathrm{yr}$---the upper end of previous estimates of natural immunity \citep{king08}---effectively result in the same model dynamics.
 The depletion of susceptible individuals may also be attributed to confounding mechanisms---such as localized vaccination programs and non-pharmaceutical interventions that reduce cholera transmission \citep{trevisin22, rebaudet21}---that were not accounted for in the model.
@@
 
+\eic{I THINK THERE IS SOME REDUNDANCY IN THIS SECTION; IDEAS MENTIONED MORE THAN ONCE. LET'S DISCUSS POSSIBLE EDITS.}
+
 The ongoing global COVID-19 pandemic has provided a clear example on how government policy may be affected by the conclusions of scientific models \citep{saltelli20}.
-This article demonstrates that fitting appropriate scientific models remains a challenging statistical task, and therefore great care is needed when fitting scientific models with the purpose of making policy recommendations.
-We provided a few suggestions that may aid the fitting of mechanistic models such as comparing model likelihoods to a benchmark.
+This article demonstrates that fitting appropriate scientific models remains a challenging statistical task, and therefore care is needed when fitting scientific models with the purpose of making policy recommendations.
+We have provided suggestions that may aid the fitting of mechanistic models such as comparing model likelihoods to a benchmark.
 Improved model fits allow for meaningful statistical inference that may provide valuable insight on a dynamic system and may improve the accuracy of model forecasts.
@@
 
-In this article we argue that careful attention to important statistical details could have resulted in improved statistical fits to the observed incidence data.
+In our case study, we found that careful attention to statistical details could have resulted in improved statistical fits to the observed incidence data.
 We acknowledge the benefit of hindsight: our demonstration of a statistically principled route to obtain better-fitting models resulting in more robust insights does not rule out the possibility of discovering other models that fit well yet predict poorly.
 We used the same data and models, and even much of the same code, as \citet{lee20}, and yet ended up with drastically different conclusions.
-At a minimum, we have shown that the conclusions are sensitive to details in how the data analysis is carried out, and that attention to statistical fit (including numerical consideration such as likelihood maximization) can lead to improved policy guidance.
+At a minimum, we have shown that the conclusions are sensitive to details in how the data analysis is carried out, and that attention to statistical fit (including numerical considerations such as likelihood maximization) can lead to improved policy guidance.
 

### 2023-01-05 ms.Rnw
minor edits to references and si

@@
 
-We implemented Model~1 using the \code{pomp} package \citep{pomp2}, relying heavily on the \code{pomp} source code provided by \cite{lee20}.
+We implemented Model~1 using the \code{pomp} package \citep{king16}, relying heavily on the \code{pomp} source code provided by \cite{lee20}.
 Both analyses used the \code{mif2} implementation of the IF2 algorithm to estimate $\paramVec$ by maximum likelihood.
@@
 
-We implemented this model using the \code{spatPomp} \code{R} package \citep{asfaw21github}.
+We implemented this model using the \code{spatPomp} \code{R} package \citep{asfaw22cran}.
 The model was then fit using the subplex algorithm, implemented in the  \code{subplex} package \citep{king2020Subplex}.

### 2023-01-05 si/si.Rnw
minor edits to references and si

@@
 Here, we add additional details to clarify the mapping from a collection of rate functions to a fully specified process.
-Our treatment follows \citet{breto09} \jwc{TODO: Cite the short course}.
+Our treatment follows \citet{breto09}.
 
@@
 Building on their results, we fit the fully coupled version of Model~3 to all available data.
-Our preliminary results (not shown) suggest that a single search for the MLE using the IBPF algorithm is insufficient;
-this result is consistent with the findings of \cite{ionides22}, who used a sequence of successive searches for the MLE in order to properly maximize the model likelihood.
-We therefore use a similar approach by performing three successive searches for the MLE.
-The first search is performed by obtaining initial values for the parameters by uniformly sampling values from a predefined hypercube.
-Subsequent searches use the resulting parameter values that correspond the largest model likelihoods.
+In order to properly maximize the model likelihood, we found that it was necessary to use multiple searches for the MLE, periodically pruning away less successful searches.
+The first collection of searches was performed by obtaining initial values for the parameters by uniformly sampling values from a predefined hypercube.
+Two subsequent refinement searches used starting values corresponding to the largest model likelihoods obtained in the previous round.
+This is similar to the approach used by \cite{ionides22}.
 
@@
 The source code that the Model~2 authors used to generate their results was written in the \code{Python} programming language, and is publicly available at \url{10.5281/zenodo.3360857} and its accompanying GitHub repository \url{https://github.com/lulelita/HaitiCholeraMultiModelingProject}.
-In order to perform our analysis in a unified framework, we re-implemented this model in the \code{R} programming language using the \code{spatPomp} package \citep{asfaw21github}, which facilitates the creation of meta-population models.
+In order to perform our analysis in a unified framework, we re-implemented this model in the \code{R} programming language using the \code{spatPomp} package \citep{asfaw22cran}, which facilitates the creation of meta-population models.
 We note that the travel and water matrices used to implement the complex dynamics in Model~2 \citep{lee20sup} are not available in either the Zenodo archive or the GitHub repository;
@@
 Another potential explanation for the discrepancy is that the parameters that we used are only approximately the same as those used by \citet{lee20sup}.
-For example, the parameters $\transmission$, $\Wbeta{}$ (See Table~\ref{tab:translate}) had reported values of $9.9 \times 10^{-7}$ and $4.03 \times 10^{-2}$, respectively (Table~S13 of \cite{lee20sup}), but were actually fit to data and therefore likely had values that were far more precise.
+For example, the parameters $\transmission$, $\Wbeta{}$ (See Table~\ref{tab:translate}) had reported values of $9.9 \times 10^{-7}$ and $4.03 \times 10^{-2}$, respectively (Table~S13 of \cite{lee20sup}), but were actually fit to data and therefore likely these values have been rounded.
 Additionally, our implementation of Model~2 used a time scale of years and many of the parameters were reported on a weekly scale, so small differences may result due to unit conversions.

### 2023-01-05 ms.Rnw
version ready for circulating

@@
 where $\mathrm{Pop}_u$ is the mean population for department $u$,
-$D_{u\varv}$ is a distance measure estimating average road distance between randomly chosen members of each population, and $v_{\mathrm{rate}}= 10^{-12}$ \eic{UNITS?} was treated as a fixed constant.
+$D_{u\varv}$ is a distance measure estimating average road distance between randomly chosen members of each population, and $v_{\mathrm{rate}}= 10^{-12} \,\mbox{km$^2$yr$^{-1}$}$  was treated as a fixed constant.
 In \eqref{model2:water_transport}, $T^W_{u\varv}$ is a measure of river flow between departments.
@@
 
-\eic{I THINK THERE IS SOME REDUNDANCY IN THIS SECTION; IDEAS MENTIONED MORE THAN ONCE. LET'S DISCUSS POSSIBLE EDITS.}
-
 The ongoing global COVID-19 pandemic has provided a clear example on how government policy may be affected by the conclusions of scientific models \citep{saltelli20}.
-This article demonstrates that fitting appropriate scientific models remains a challenging statistical task, and therefore care is needed when fitting scientific models with the purpose of making policy recommendations.
-We have provided suggestions that may aid the fitting of mechanistic models such as comparing model likelihoods to a benchmark.
-Improved model fits allow for meaningful statistical inference that may provide valuable insight on a dynamic system and may improve the accuracy of model forecasts.
-Caution is nonetheless needed when making policy based on modeling conclusions, as model misspecification may invalidate conclusions.
-
-In our case study, we found that careful attention to statistical details could have resulted in improved statistical fits to the observed incidence data.
+This article demonstrates that fitting appropriate scientific models to guide policy is a challenging statistical task.
+In our case study, we found that additional attention to statistical details could have resulted in improved statistical fits to the observed incidence data, leading to improvements in the accuracy of the resulting policy guidance.
+We used the same data and models, and even much of the same code, as \citet{lee20}, and yet ended up with drastically different conclusions.
 We acknowledge the benefit of hindsight: our demonstration of a statistically principled route to obtain better-fitting models resulting in more robust insights does not rule out the possibility of discovering other models that fit well yet predict poorly.
-We used the same data and models, and even much of the same code, as \citet{lee20}, and yet ended up with drastically different conclusions.
-At a minimum, we have shown that the conclusions are sensitive to details in how the data analysis is carried out, and that attention to statistical fit (including numerical considerations such as likelihood maximization) can lead to improved policy guidance.
 

### 2023-01-17 ms.Rnw
names, addresses

@@
 \author[A]{\fnms{Jesse}~\snm{Wheeler}\ead[label=e1]{jeswheel@umich.edu}\orcid{0000-0003-3941-3884}},
-\author[A]{\fnms{AnnaElaine}~\snm{Rosengart}\ead[label=e2]{aelr@umich.edu}}
+\author[B]{\fnms{AnnaElaine L.}~\snm{Rosengart}\ead[label=e2]{arosenga@andrew.cmu.edu}}
 \author[A]{\fnms{Zhuoxun}~\snm{Jiang}\ead[label=e3]{zhuoxunj@umich.edu}},
-\author[A]{\fnms{Kevin}~\snm{Hao En Tan}\ead[label=e4]{kevtan@umich.edu}},
+\author[A]{\fnms{Kevin}~\snm{Tan}\ead[label=e4]{kevtan@umich.edu}},
 \author[A]{\fnms{Noah}~\snm{Treutle}\ead[label=e5]{ntreutle@umich.edu}}
 \and
-\author[A]{\fnms{Edward}~\snm{Ionides}\ead[label=e6]{ionides@umich.edu}}
-
-\address[A]{Statistics Department, University of Michigan\printead[presep={,\ }]{e1,e2,e3,e4,e5,e6}}
+\author[A]{\fnms{Edward L.}~\snm{Ionides}\ead[label=e6]{ionides@umich.edu}}
+
+\address[A]{Statistics Department, University of Michigan\printead[presep={,\ }]{e1,e3,e4,e5,e6}}
+
+\address[B]{Statistics \& Data Science Department, Carnegie Mellon University\printead[presep={,\ }]{e2}}
 
@@
 
-We implemented this model using the \code{spatPomp} \code{R} package \citep{asfaw22cran}.
+We implemented this model using the \code{spatPomp} \code{R} package \citep{asfaw23arxiv}.
 The model was then fit using the subplex algorithm, implemented in the  \code{subplex} package \citep{king2020Subplex}.

### 2023-01-17 si/si.Rnw
version ready for circulating
names, addresses

@@
 \title{Supplement to ``{\mytitle}''}
-  \author{Jesse Wheeler, Anna Rosengart, Zhuoxun Jiang, \\ Kevin Hao En Tan, Noah Treutle, Edward L. Ionides
+  \author{Jesse Wheeler, AnnaElaine L. Rosengart, Zhuoxun Jiang, \\ Kevin Tan, Noah Treutle and Edward L. Ionides
  \\ \hspace{.2cm}\\
@@
 
+\clearpage
+
 \section{Replication of \cite{lee20}}
@@
 The source code that the Model~2 authors used to generate their results was written in the \code{Python} programming language, and is publicly available at \url{10.5281/zenodo.3360857} and its accompanying GitHub repository \url{https://github.com/lulelita/HaitiCholeraMultiModelingProject}.
-In order to perform our analysis in a unified framework, we re-implemented this model in the \code{R} programming language using the \code{spatPomp} package \citep{asfaw22cran}, which facilitates the creation of meta-population models.
+In order to perform our analysis in a unified framework, we re-implemented this model in the \code{R} programming language using the \code{spatPomp} package \citep{asfaw23arxiv}, which facilitates the creation of meta-population models.
 We note that the travel and water matrices used to implement the complex dynamics in Model~2 \citep{lee20sup} are not available in either the Zenodo archive or the GitHub repository;

### 2023-01-21 si/si.Rnw
prepared documents for submission to Arxiv

@@
 
-<<echo=FALSE, include=FALSE, message=FALSE>>=
+<<calcMod1LeeLikes, echo=FALSE, include=FALSE, message=FALSE>>=
 h1_epi_evals <- readRDS("../model1/run_level_3/lee1_epi_evals.rds")

### 2023-01-21 ms.Rnw
prepared documents for submission to Arxiv
Added haitipkg zenodo doi

@@
 \begin{abstract}
-\begin{center}
-Working draft, \today
-\end{center}
+% \begin{center}
+% Working draft, \today
+% \end{center}
 
@@
 \sdescription{
-The \code{haitipkg} R package is maintained in a GitHub repository, \code{jeswheel/haitipkg}. The submitted version of this package will be archived on Zenodo.
+The \code{haitipkg} R package is maintained in a GitHub repository, \code{jeswheel/haitipkg}. The submitted version of this package is archived on Zenodo: \code{10.5281/zenodo.7557099}
 The package contains all of the data and code used to create, fit and explore the models presented in this article.

### 2023-01-26 ms.Rnw
Updated to meet AoAS submission requirements

@@
 %\documentclass[aoas]{imsart}
-\documentclass[aoas,preprint]{imsart}
+\documentclass[aoas]{imsart}
 
@@
 % \end{center}
-
-\vspace{3mm}
+% \vspace{3mm}
 
@@
 
-\begin{figure}[ht]
+\begin{figure}
 <<Plot_Reported_Cases, echo=FALSE, fig.height=3.5>>=
@@
 
-\begin{figure}[ht]
+\begin{figure}
 \centering
@@
 
-\begin{figure}[ht]
+\begin{figure}
 <<Model3_SpatPOMP_sims_Figure, fig.height=3>>=
@@
 
-\begin{table}[ht]
+\begin{table}
 \centering
@@
 
-\begin{figure}[ht]
+\begin{figure}
 <<Plot_Model1_Scenarios, fig.height=2>>=
@@
 
-\begin{figure}[ht]
+\begin{figure}
 <<Mod2Fit_and_Scenarios_Figure, fig.height=1.9, fig.width=4.8, fig.align='center'>>=
@@
 
-\begin{figure}[ht]
+\begin{figure}
 <<Plot_Model3_Scenarios, fig.height=2>>=
@@
 
-\begin{figure}[ht]
+\begin{figure}
 <<Elimination_Probs_Figure, fig.height=2.2>>=
@@
 \begin{table}
+\centering
 % \small
-\resizebox{\textwidth}{!}{
-\hspace{-10mm}\begin{tabular}{|l@{}|l@{}t@{}|l@{}t@{}|l@{}t@{}|}
+\begin{tabular}{|l@{}|l@{}t@{}|l@{}t@{}|l@{}t@{}|}
 \hline
@@
   & \eqref{eq:model3:water} \\
-  & $\transmissionTrend = \Sexpr{myround(p1["betat"],2)}$ [N]
+  & $\transmissionTrend = \Sexpr{myround(p1["betat"],2)}^*$
   & \eqref{model1:betat}
-  & $\phaseParm = \Sexpr{myround(h2_fit$h2_params['phase'], 2)}$ [N]
+  & $\phaseParm = \Sexpr{myround(h2_fit$h2_params['phase'], 2)}^*$
   & \eqref{model2:lambda}
@@
   \\\hline
-$\begin{array}{l}
-\text{Immunity} \\
-\text{(year)}
-\end{array}$
+  Immunity (yr)
   & {\fixed $\muRS^{-1}=\Sexpr{myround(1/p1["alpha"]/52,1)}$}
@@
 \hline
-% Vaccine efficacy
-%   & \multicolumn{1}{c}{\hspace{16pt}\missing}
-%   &
-%   & $\begin{array}{l}
-%       \hspace{-1mm} \fixed{\vaccineEfficacy_{1:4} = (
-%         \Sexpr{myround(1 - h2_fit$h2_params["VE1"] * 0.4688, 2)},
-%         \Sexpr{myround(1 - h2_fit$h2_params["VE2"] * 0.4688, 2)}} \\
-%         \fixed{\Sexpr{myround(1 - h2_fit$h2_params["VE1"], 2)},
-%         \Sexpr{myround(1 - h2_fit$h2_params["VE2"], 2)}})
-%         \end{array}$
-%   & \eqref{model2:mu_SE}
-%   & \fixed{$\eta_{ud}(t)$}
-%   &
-%   \\
 Birth/death (yr)
@@
 \\
-Symptomatic frac.
+Sympt. frac.
   & {\fixed $\symptomFrac_z(t)=c\theta^*(t-\tau_d)$}
@@
 $\begin{array}{l}
-\text{Asymptomatic} \\
+\text{Asympt.} \\
 \text{infectivity}
@@
 \end{array}$
-  & $\sigmaProc=(\Sexpr{myround(sqrt(p1["sig_sq_epi"]),2)}, \Sexpr{myround(sqrt(p1["sig_sq_end"]),2)})$ [N]
+  & $\sigmaProc=(\Sexpr{myround(sqrt(p1["sig_sq_epi"]),2)}, \Sexpr{myround(sqrt(p1["sig_sq_end"]),2)})^*$
   & \eqref{model1:lambda}
@@
 \multirow{2}{*}{Initial Values} & $I_{0,0} = \Sexpr{round(prod(p1[c('pop_0', 'I_0')]))}$ & & \multicolumn{1}{c}{\multirow{2}{*}{\hspace{16pt}\missing}} & & \multirow{2}{*}{$I_{3,7,10;0}(0)= (
-        \Sexpr{round(prod(p3[c('H3', 'Iinit3')]))}, \Sexpr{round(prod(p3[c('H7', 'Iinit7')]))}, \Sexpr{round(prod(p3[c('H10', 'Iinit10')]))})$ [N]} & \\
+        \Sexpr{round(prod(p3[c('H3', 'Iinit3')]))}, \Sexpr{round(prod(p3[c('H7', 'Iinit7')]))}, \Sexpr{round(prod(p3[c('H10', 'Iinit10')]))})^*$} & \\
 & $E_{0,0} = \Sexpr{round(prod(p1[c('pop_0', 'E_0')]))}$ & & \multicolumn{1}{c}{} & & & \\\hline
-Hurricane & \multicolumn{1}{c}{\multirow{2}{*}{\hspace{16pt}\missing}} & & \multicolumn{1}{c}{\multirow{2}{*}{\hspace{16pt}\missing}} & & $\Whur{3, 9} = (\Sexpr{myround(p3['aHur3'], 2)}, \Sexpr{myround(p3['aHur9'], 2)})$ [N] & \multirow{2}{*}{\eqref{eq:model3:foi}} \\
-Parameters & \multicolumn{1}{c}{} & & \multicolumn{1}{c}{} & & $\hHur{3, 9} = (\Sexpr{myround(p3['hHur3'], 2)}, \Sexpr{myround(p3['hHur9'], 2)})$ [N] & \\\hline
-
-\end{tabular}}
+Hurricane & \multicolumn{1}{c}{\multirow{2}{*}{\hspace{16pt}\missing}} & & \multicolumn{1}{c}{\multirow{2}{*}{\hspace{16pt}\missing}} & & $\Whur{3, 9} = (\Sexpr{myround(p3['aHur3'], 2)}, \Sexpr{myround(p3['aHur9'], 2)})^*$ & \multirow{2}{*}{\eqref{eq:model3:foi}} \\
+Parameters & \multicolumn{1}{c}{} & & \multicolumn{1}{c}{} & & $\hHur{3, 9} = (\Sexpr{myround(p3['hHur3'], 2)}, \Sexpr{myround(p3['hHur9'], 2)})^*$ & \\\hline
+
+\end{tabular}
 \caption{References to the relevant equation are given in parentheses.
 Parameters in blue were fixed based on scientific reasoning and not fitted to the data.
-[N] denotes parameters added during our re-analysis, not considered by Lee et al.
+$^*$ denotes parameters added during our re-analysis, not considered by Lee et al.
 Translations back into the notation of \citet{lee20} are given in Table~S1.

### 2023-01-26 fit_model1.R
removed unused file

@@ -1,40 +1,0 @@
-# This script is used to fit Model 1.
-#
-# @created: Sep 2, 2022
-
-library(haitipkg)
-library(pomp)
-library(foreach)
-library(doParallel)
-library(doRNG)
-library(dplyr)
-
-cores <-  as.numeric(Sys.getenv('SLURM_NTASKS_PER_NODE', unset=NA))
-# cores <- 20
-if(is.na(cores)) cores <- detectCores()
-registerDoParallel(cores)
-
-set.seed(636813)
-
-# Set Run Level for debug, timing, and full computation
-RUN_LEVEL = 3
-
-NP_LOCAL     <- switch(RUN_LEVEL, 100, 1000,   2000)
-NMIF         <- switch(RUN_LEVEL,  20,   50,    200)
-NUM_TREND    <- switch(RUN_LEVEL,   5,   10,     90)
-NPROF        <- switch(RUN_LEVEL,   4,    4,     24)
-NREPS_EVAL   <- switch(RUN_LEVEL,  20,   20,  cores)
-NP_EVAL      <- switch(RUN_LEVEL, 200, 1000,   5000)
-
-results <- fit_haiti1(
-  NP = NP_LOCAL, NMIF = NMIF, NUM_TREND = NUM_TREND, NPROF = NPROF,
-  NREPS_EVAL = NREPS_EVAL, NP_EVAL = NP_EVAL, ncores = cores
-)
-
-save_dir <- paste0("model1/run_level_", RUN_LEVEL)
-
-if (!dir.exists(save_dir)) {
-  dir.create(save_dir, recursive = TRUE)
-}
-
-save(results, file = paste0(save_dir, '/haiti1_fit.rda'))

### 2023-06-12 ms.Rnw
started revisions

@@
 \newcommand\muRS{\mu_{RS}}
-\newcommand\vaccineEfficacy{\theta}
+\newcommand\vaccineEfficacy{\vartheta}
 \newcommand\muBirth{\mu_S}
@@
 \newcommand\paramVec{\theta}
+\newcommand\childReduce{q}
 
@@
 \label{model1:EI}
-\mu_{E_{\vaccCounter}I_{\vaccCounter}} &=& \muEI\big(1-\symptomFrac_{\vaccCounter}(t)\big),
+\mu_{E_{\vaccCounter}I_{\vaccCounter}} &=& \muEI\big(1-\vaccineEfficacy_\vaccCounter(t)\big),
 \\[-3pt]
 \label{model1:EA}
-\mu_{E_{\vaccCounter}A_{\vaccCounter}} &=& \muEI\, \symptomFrac_{\vaccCounter}(t),
+\mu_{E_{\vaccCounter}A_{\vaccCounter}} &=& \muEI\, \vaccineEfficacy_\vaccCounter(t),
 \\[-3pt]
@@
 In \eqref{model1:EA} and~\eqref{model1:EI} the asymptomatic ratio for non-vaccinated individuals is set $\symptomFrac_0(t)=0$, so that the asymptomatic route is reserved for vaccinated individuals.
-For $\vaccCounter\in\seq{1}{\vaccClass}$, the vaccination cohort $\vaccCounter$ is assigned a time $\tau_{\vaccCounter}$, and we take $\symptomFrac_{\vaccCounter}(t) = c \, \theta^*(t-\tau_{\vaccCounter})$
-where $\theta^*(t)$ is efficacy at time $t$ since vaccination for adults, taken from \cite{lee20}, Table~S4, and $c=\big(1-(1-0.4688)\times 0.11\big)$ is a correction to allow for reduced efficacy in the 11\% of the population aged under 5 years.
+For $\vaccCounter\in\seq{1}{\vaccClass}$, the vaccination cohort $\vaccCounter$ is assigned a time $\tau_{\vaccCounter}$, and we take $\vaccineEfficacy_\vaccCounter(t) = c \, \vaccineEfficacy^*(t-\tau_{\vaccCounter})$
+where $\vaccineEfficacy^*(t)$ is efficacy at time $t$ since vaccination for adults, taken from \cite{lee20}, Table~S4, and $c=\big(1-(1-0.4688)\times 0.11\big)$ is a correction to allow for reduced efficacy in the 11\% of the population aged under 5 years.
 Single and double vaccine doses were modeled by changing the waning of protection; protection was assumed to be equal between single and double dose until 52 weeks after vaccination, at which point the single dose becomes ineffective.
@@
 The unit of $W_u(t)$ is cells per ml, with dose response modeled via a saturation constant of $\Wsat$ in \eqref{model2:lambda}.
+In \eqref{model2:mu_SE}, $\vaccineEfficacy_\vaccCounter$ denotes the vaccine efficacy for each vaccination campaign $\vaccCounter \in \vaccClass$, with $\vaccineEfficacy_0 = 1$, $\vaccineEfficacy_1 = 1 - 0.429 \childReduce$, $\vaccineEfficacy_2 = 1 - 0.519 \childReduce$, $\vaccineEfficacy_3 = 1 - 0.429$, and $\vaccineEfficacy_4 = 1 - 0.519$
+Here, $\childReduce = 0.4688$ represents the reduced efficacy of the vaccination for children under the age of five years, and the values $0.429$ and $0.519$ are the median effectiveness of one and two doses over their effective period respectively, according to Table S4 in the supplement material of \citet{lee20}.
+Because vaccine efficacy remains constant, individuals in this model transition from a vaccinated compartment to the susceptible compartment at the end of the vaccine coverage period.
+
 The starting value for each element of the latent state vector $\bm{X}^{(2)}(0)$ are set to zero except for $I_{u0}(0) = y_u^*(0)/\reportRate$ and $R_{u0}(0) = \mathrm{Pop}_u - I_{u0}(0)$, where $y_u^*(0)$ is the reported number of cholera cases in department $u$ at time $t = 0$.
@@
 
-The latent state is described as $\bm{X}^{(3)}(t) = \big(S_{u\vaccCounter}(t),\allowbreak I_{u\vaccCounter}(t),\allowbreak A_{u\vaccCounter}(t),\allowbreak R_{u\vaccCounter k}(t),\allowbreak W_u(t),\allowbreak u \in \seq{0}{U},\allowbreak \vaccCounter \in \seq{0}{4},\allowbreak k \in \seq{1}{3}\big)$.
+The latent state is described as $\bm{X}^{(3)}(t) = \big(S_{u\vaccCounter}(t),\allowbreak I_{u}(t),\allowbreak A_{u}(t),\allowbreak R_{u\vaccCounter k}(t),\allowbreak W_u(t),\allowbreak u \in \seq{0}{U},\allowbreak \vaccCounter \in \seq{0}{4},\allowbreak k \in \seq{1}{3}\big)$.
 Here, $\vaccCounter=0$ corresponds to unvaccinated, $\vaccCounter=2j-1$ corresponds to a single dose on the $j$th vaccination campaign in unit $u$ and $\vaccCounter=2j$ corresponds to receiving two doses on the $j$th vaccination campaign.
@@
 \label{eq:model3:foi}
-\lambda_u(t) = \left(\Wbeta{_u} + 1_{(t \geq \tHur)}\Whur{_u} e^{-\hHur{u}\left(t - \tHur\right)}\right) \frac{W_u(t)}{1+W_u(t)} + \transmission_u \sum_{{\varv}\neq u}\big(I_{\varv 0}(t)+ \asymptomRelativeInfect A_{\varv 0}(t)\big),
+\lambda_u(t) = \left(\Wbeta{_u} + 1_{(t \geq \tHur)}\Whur{_u} e^{-\hHur{u}\left(t - \tHur\right)}\right) \frac{W_u(t)}{1+W_u(t)} + \transmission_u \sum_{{\varv}\neq u}\big(I_{\varv}(t)+ \asymptomRelativeInfect A_{\varv}(t)\big),
 \end{equation}
@@
 \label{eq:model3:SI}
-\mu_{S_{u\vaccCounter}I_{u\vaccCounter}} &=& \symptomFrac \,  \lambda_u \big(1-\eta_{u\vaccCounter}(t)\big) \, d\Gamma/dt,
+\mu_{S_{u\vaccCounter}I_{u}} &=& \symptomFrac \,  \lambda_u \big(1-\vaccineEfficacy_{u\vaccCounter}(t)\big) \, d\Gamma/dt,
 \\[-3pt]
 \label{eq:model3:SA}
-\mu_{S_{u\vaccCounter}A_{u\vaccCounter}} &=& (1-\symptomFrac) \,  \lambda_u \big(1-\eta_{u\vaccCounter}(t)\big) \,  d\Gamma/dt,
+\mu_{S_{u\vaccCounter}A_{u}} &=& (1-\symptomFrac) \,  \lambda_u \big(1-\vaccineEfficacy_{u\vaccCounter}(t)\big) \,  d\Gamma/dt,
 \\[-3pt]
 \label{eq:model3:IR}
-\mu_{I_{u\vaccCounter}R_{u\vaccCounter 1}} &=& \mu_{A_{u\vaccCounter}R_{u\vaccCounter 1}} = \muIR,
+\mu_{I_{u}R_{u\vaccCounter 1}} &=& \mu_{A_{u}R_{u\vaccCounter 1}} = \muIR,
 \\[-3pt]
 \label{eq:model3:IS}
-\mu_{I_{u\vaccCounter}S_{u0}} &=& \muDeath + \choleraDeath, \hspace{2mm} \mu_{A_{u\vaccCounter}S_{u0}} = \muDeath
+\mu_{I_{u}S_{u0}} &=& \muDeath + \choleraDeath, \hspace{2mm} \mu_{A_{u}S_{u0}} = \muDeath
 \\[-3pt]
@@
 \label{eq:model3:water}
-\mu_{{\demography}W_u} &=& \big[1 + \seasAmplitude \big(J_u(t))^r \big] \mathrm{Den}_u \, \Wshed \big[ I_{u0}(t)+ \asymptomRelativeShed A_{u0}(t) \big],
+\mu_{{\demography}W_u} &=& \big[1 + \seasAmplitude \big(J_u(t))^r \big] \mathrm{Den}_u \, \Wshed \big[ I_{u}(t)+ \asymptomRelativeShed A_{u}(t) \big],
 \\[-3pt]
@@
 All deaths are balanced by births into the susceptible compartment in \myeqref{eq:model3:IS} and \myeqref{eq:model3:RS}, thereby maintaining constant population in each department.
+
+Similar to Model~1, there are no distinct compartments for individuals under five years of age, and the vaccination efficacy is taken as a age adjusted weighted average of the efficacy for individuals both over and under five years of age:  $\vaccineEfficacy_{u \vaccCounter}(t) = c\vaccineEfficacy^*(t - \tau_{uz})$, where $\tau_{uz}$ is the vaccination time for unit $u$ and vaccination campaign $z$.
+The value $c$ and the function $\vaccineEfficacy^*$ are equivalent to those described in Sec.~\ref{sec:model1}.
 
@@
 names(reg_rw_1.sd) <- est_param_names_expanded
-chol_rw_1.sd <- do.call(rw.sd, reg_rw_1.sd)
+chol_rw_1.sd <- do.call(rw_sd, reg_rw_1.sd)
 
@@
 names(reg_rw_2.sd) <- est_param_names_expanded
-chol_rw_2.sd <- do.call(rw.sd, reg_rw_2.sd)
+chol_rw_2.sd <- do.call(rw_sd, reg_rw_2.sd)
 
@@
 names(reg_rw_3.sd) <- est_param_names_expanded
-chol_rw_3.sd <- do.call(rw.sd, reg_rw_3.sd)
+chol_rw_3.sd <- do.call(rw_sd, reg_rw_3.sd)
 
@@
 Here, we compare with a forecast projected from the filtering distribution: to obtain a forecast using POMP model $m$ at a collection of times $t_{N+1:N+s}$, where $N$ is the index for the last available data, we simulate forward from a draw from
-$f^{(m)}_{\bm{X}_N|\bm{Y}_{1:N}}(\bm{x}_{N} | \bm{y}^*_{1:N} ; \hat\theta)$
-where $\hat\theta$ is a vector of calibrated parameters.
+$f^{(m)}_{\bm{X}_N|\bm{Y}_{1:N}}(\bm{x}_{N} | \bm{y}^*_{1:N} ; \hat\paramVec)$
+where $\hat\paramVec$ is a vector of calibrated parameters.
 The decision not to do this partially explains the unsuccessful forecasts of \cite{lee20}: their Table~S7 shows that the subset of their simulations which were consistent with observing zero cases in 2019 also accurately predicted the prolonged absence of detected cholera.
@@
 If a Bayesian technique is used for parameter estimation, a natural way to account for parameter uncertainty is to obtain simulations from the model where each simulation is obtained using parameters drawn from the estimated posterior distribution.
-For frequentist inference, one possible approach is obtaining model forecasts from various values of $\theta$, where the values of $\theta$ are sampled proportionally according to their corresponding likelihoods \citep{king15}; see Sec.~S7 of the supplement material for more details.
-Both of these approaches share the similarity that parameters are chosen for the forecast approximately in proportion to their corresponding value of the likelihood function, $f^{(m)}_{\bm{Y}_{1:N}}(\bm{y}_{1:N}^*; \theta)$.
+For frequentist inference, one possible approach is obtaining model forecasts from various values of $\paramVec$, where the values of $\paramVec$ are sampled proportionally according to their corresponding likelihoods \citep{king15}; see Sec.~S7 of the supplement material for more details.
+Both of these approaches share the similarity that parameters are chosen for the forecast approximately in proportion to their corresponding value of the likelihood function, $f^{(m)}_{\bm{Y}_{1:N}}(\bm{y}_{1:N}^*; \paramVec)$.
 In this analysis, we do not construct forecasts accounting for parameter uncertainty as our focus is on the estimation and diagnosis of mechanistic models.
@@
 Sympt. frac.
-  & {\fixed $\symptomFrac_z(t)=c\theta^*(t-\tau_d)$}
+  & {\fixed $\symptomFrac_z(t)=c\vaccineEfficacy^*(t-\tau_d)$}
   & (\ref{model1:EI}-\ref{model1:EA})

### 2023-06-15 ms.Rnw
MLE details

@@
 
-Each model describes the cholera dynamics as a partially observed Markov process (POMP) with a latent state vector $\bm{X}^{(\modelCounter)}(t)$ for each continuous time point $t$ and model $\modelCounter \in \{1, 2, 3\}$.
+Each model describes the cholera dynamics as a partially observed Markov process (POMP) with a latent state vector $\bm{X}(t)$ for each continuous time point $t$.
 $N$ observations on the system are collected at time points $t_1,\dots,t_N$, written as $t_{1:N}$.
-The observation at time $t_n$ is modeled by the random vector $\bm{Y}^{(\modelCounter)}_n$.
+The observation at time $t_n$ is modeled by the random vector $\bm{Y}_n$.
 While the latent process exists between observation times, the value of the latent state at observations times is of particular interest.
-We therefore write $\bm{X}^{(\modelCounter)}_n = \bm{X}^{(\modelCounter)}(t_n)$ to denote the value of the latent process at the $n$th observation time, and $\bm{X}^{(\modelCounter)}_{1:N}$ is the collection of latent state values for all observed time points.
-The observable random variables $\bm{Y}^{(\modelCounter)}_{1:N}$ are assumed to be conditionally independent given $\bm{X}^{(\modelCounter)}_{0:N}$.
-Together, with the density for the initial value of the latent state $\bm{X}^{(\modelCounter)}_{0} = \bm{X}^{(\modelCounter)}(t_0)$, each model assumes a joint density $f_{\bm{X}^{(\modelCounter)}_{0:N}, \bm{Y}^{(\modelCounter)}_{1:N}}\big(\bm{x}^{(\modelCounter)}_{0:N}, \bm{y}^{(\modelCounter)}_{1:N};\paramVec\big)$, where $\paramVec$ is a parameter vector that indexes the model.
+We therefore write $\bm{X}_n = \bm{X}(t_n)$ to denote the value of the latent process at the $n$th observation time, and $\bm{X}_{1:N}$ is the collection of latent state values for all observed time points.
+The observable random variables $\bm{Y}_{1:N}$ are assumed to be conditionally independent given $\bm{X}_{0:N}$.
+Together, with the density for the initial value of the latent state $\bm{X}_{0} = \bm{X}(t_0)$, each model assumes a joint density $f_{\bm{X}_{0:N}, \bm{Y}_{1:N}}\big(\bm{x}_{0:N}, \bm{y}_{1:N};\paramVec\big)$, where $\paramVec$ is a parameter vector that indexes the model.
 The observed data $\bm{y}_{1:N}^*$, along with the unobserved true value of the latent state, are modeled as a realization of this joint distribution.
 
-For each model $\modelCounter \in \{1, 2, 3\}$, the latent state vector $\bm{X}^{(\modelCounter)}(t)$ consists of individuals labeled as susceptible (S), infected (I), asymptomatically infected (A), vaccinated (V), and recovered (R), with various sub-divisions sometimes considered.
+Because of the probabilistic nature of both the unobserved latent state and the observable random variables, it is possible to consider various marginal and conditional densities of these two jointly random vectors.
+An important example is the marginal density of the observed random vector $\bm{Y}_{1:N}$, evaluated at the observed data $\bm{y}_{1:N}^*$: $f_{\bm{Y}_{1:N}}\big(\bm{y}_{1:N}^*; \paramVec\big) = \int f_{\bm{X}_{0:N}, \bm{Y}_{1:N}}\big(\bm{x}_{0:N}, \bm{y}_{1:N};\paramVec\big) d\bm{x}_{0:N}$.
+When treated as a function of the parameter vector $\paramVec$, this is called the \emph{likelihood function}, which is the basis of likelihood based statistical inference.
+
+Using the conditional independence of $\bm{Y}_{1:N}$ given $\bm{X}_{0:N}$ and the Markov property of $\bm{X}_{0:N}$, the joint density can be refactored as: $f_{\bm{X}_{0:N}, \bm{Y}_{1:N}}\big(\bm{x}_{0:N}, \bm{y}_{1:N};\paramVec\big) = f_{\bm{X}_0}\big(\bm{x}_0;\paramVec\big)\prod_{n = 1}^N f_{\bm{X}_n|\bm{X}_{n-1}}\big(\bm{x}_{n}|\bm{x}_{n-1}; \paramVec\big)f_{\bm{Y}_n|\bm{X}_{n}}\big(\bm{y}_n|\bm{x}_{n}\big)$.
+This factorization is useful because it demonstrates that POMP models may be completely described using three parts: the \emph{initialization model} for the latent states $f_{\bm{X}_0}\big(\bm{x}_0;\paramVec\big)$; the \emph{one-step transition density}, or \emph{the process model} $f_{\bm{X}_n|\bm{X}_{n-1}}\big(\bm{x}_{n}|\bm{x}_{n-1}; \paramVec\big)$; and the \emph{measurement model} $f_{\bm{Y}_n|\bm{X}_{n}}\big(\bm{y}_n|\bm{x}_{n}\big)$.
+In the following subsections, we describe Models 1--3 in terms of these three components.
+The latent state vector $\bm{X}(t)$ for each model consists of individuals labeled as susceptible (S), infected (I), asymptomatically infected (A), vaccinated (V), and recovered (R), with various sub-divisions sometimes considered.
+The observable random vector $\bm{Y}_{1:N}$ represents the random vector of cholera incidence data for each model;
 Models~2 and~3 have metapopulation structure, meaning that each individual is a member of a spatial unit, denoted by a subscript $u\in \seq{1}{U}$, in which case we denote the observed data for each unit using $\bm{y}_{1:N}^* = y_{1:N,1:U}^*$.
@@
 
-In the following subsections, descriptions of Models~1--3 are provided.
-While the model description is scientifically critical, as well as necessary for transparency and reproducibility, the model details are not essential to our methodological discussions of how to diagnose and address model misspecification with the purpose of informing policy.
+While the complete model description is scientifically critical, as well as necessary for transparency and reproducibility, the model details are not essential to our methodological discussions of how to diagnose and address model misspecification with the purpose of informing policy.
 A first-time reader may choose to skim through the rest of this section, and return later.
@@
 \label{sec:model1}
-$\bm{X}^{(1)}(t) = \big(S_{\vaccCounter}(t),  E_{\vaccCounter}(t), I_{\vaccCounter}(t), A_{\vaccCounter}(t), R_{\vaccCounter}(t), \vaccCounter \in 0:\vaccClass\big)$ describes susceptible, latent (exposed), infected (and symptomatic), asymptomatic, and recovered individuals in vaccine cohort $\vaccCounter$.
+The latent state vector $\bm{X}(t) = \big(S_{\vaccCounter}(t),  E_{\vaccCounter}(t), I_{\vaccCounter}(t), A_{\vaccCounter}(t), R_{\vaccCounter}(t), \vaccCounter \in 0:\vaccClass\big)$ describes susceptible, latent (exposed), infected (and symptomatic), asymptomatic, and recovered individuals in vaccine cohort $\vaccCounter$ at time $t$.
 Here, $\vaccCounter=0$ corresponds to unvaccinated individuals, and $\vaccCounter \in \seq{1}{\vaccClass}$ describes hypothetical vaccination programs.
@@
 
+For any time point in $t_{1:N}$, the process model describes the movement of individuals from one latent state compartment to another.
 Per-capita transition rates are given in Equations~\ref{model1:SE}-\ref{model1:birth}:
@@
 
-The latent state vector $\bm{X}^{(1)}(t)$ is initialized by setting each compartment count for each vaccination scenario $\vaccCounter \neq 0$ as zero, and introducing initial-value parameters $I_{0,0}$ and $E_{0, 0}$ such that $R_{0}(0) = 0$, $I_{0}(0) = \mathrm{Pop} \times I_{0, 0}$, $E_{0}(0) = \mathrm{Pop} \times E_{0, 0}$ and $S_{0}(0) = \mathrm{Pop} \times (1 - I_{0, 0} - E_{0, 0})$, where $\mathrm{Pop}$ is the total population of Haiti.
-Reported cholera cases at time point $n$ ($\bm{Y}^{(1)}_n$) are assumed to come from a negative binomial measurement model, where only a fraction ($\reportRate$) of new weekly cases are reported.
+The latent state vector $\bm{X}(t)$ is initialized by setting each compartment count for each vaccination scenario $\vaccCounter \neq 0$ as zero, and introducing initial-value parameters $I_{0,0}$ and $E_{0, 0}$ such that $R_{0}(0) = 0$, $I_{0}(0) = \mathrm{Pop} \times I_{0, 0}$, $E_{0}(0) = \mathrm{Pop} \times E_{0, 0}$ and $S_{0}(0) = \mathrm{Pop} \times (1 - I_{0, 0} - E_{0, 0})$, where $\mathrm{Pop}$ is the total population of Haiti.
+The measurement model assumes reported cholera cases at time point $n$ come from a negative binomial distribution, where only a fraction ($\reportRate$) of new weekly cases are reported.
 See Sec.~S3 of the supplement material for more details.
@@
 \end{equation}
-The latent state is therefore described by the vector $\bm{X}^{(2)}(t) = \big(S_{u\vaccCounter}(t),\allowbreak E_{u\vaccCounter}(t),\allowbreak I_{u\vaccCounter}(t),\allowbreak A_{u\vaccCounter}(t),\allowbreak R_{u\vaccCounter}(t),\allowbreak R_{u\vaccCounter}^A(t),\allowbreak W_u,\allowbreak u \in \seq{1}{U},\allowbreak \vaccCounter \in \seq{0}{4}\big)$.
+The latent state is therefore described by the vector $\bm{X}(t) = \big(S_{u\vaccCounter}(t),\allowbreak E_{u\vaccCounter}(t),\allowbreak I_{u\vaccCounter}(t),\allowbreak A_{u\vaccCounter}(t),\allowbreak R_{u\vaccCounter}(t),\allowbreak R_{u\vaccCounter}^A(t),\allowbreak W_u,\allowbreak u \in \seq{1}{U},\allowbreak \vaccCounter \in \seq{0}{4}\big)$.
 The cosine term in Eq.~\myeqref{model2:lambda} accounts for annual seasonality, with a phase parameter $\phaseParm$.
@@
 
-The starting value for each element of the latent state vector $\bm{X}^{(2)}(0)$ are set to zero except for $I_{u0}(0) = y_u^*(0)/\reportRate$ and $R_{u0}(0) = \mathrm{Pop}_u - I_{u0}(0)$, where $y_u^*(0)$ is the reported number of cholera cases in department $u$ at time $t = 0$.
+The starting value for each element of the latent state vector $\bm{X}(0)$ are set to zero except for $I_{u0}(0) = y_u^*(0)/\reportRate$ and $R_{u0}(0) = \mathrm{Pop}_u - I_{u0}(0)$, where $y_u^*(0)$ is the reported number of cholera cases in department $u$ at time $t = 0$.
 Reported cases are assumed to come from a log-normal distribution, with the log-scale mean equal to the reporting rate $\reportRate$ times the number of newly infected individuals.
@@
 
-The latent state is described as $\bm{X}^{(3)}(t) = \big(S_{u\vaccCounter}(t),\allowbreak I_{u}(t),\allowbreak A_{u}(t),\allowbreak R_{u\vaccCounter k}(t),\allowbreak W_u(t),\allowbreak u \in \seq{0}{U},\allowbreak \vaccCounter \in \seq{0}{4},\allowbreak k \in \seq{1}{3}\big)$.
+The latent state is described as $\bm{X}(t) = \big(S_{u\vaccCounter}(t),\allowbreak I_{u}(t),\allowbreak A_{u}(t),\allowbreak R_{u\vaccCounter k}(t),\allowbreak W_u(t),\allowbreak u \in \seq{0}{U},\allowbreak \vaccCounter \in \seq{0}{4},\allowbreak k \in \seq{1}{3}\big)$.
 Here, $\vaccCounter=0$ corresponds to unvaccinated, $\vaccCounter=2j-1$ corresponds to a single dose on the $j$th vaccination campaign in unit $u$ and $\vaccCounter=2j$ corresponds to receiving two doses on the $j$th vaccination campaign.
@@
 However, the congruency between the model and observed data serves as a proxy for the congruency between the model and the true underlying dynamic system.
-As such, the best possible fit of the proposed mechanistic structure to the observed data is desirable.
+As such, it is desirable to obtain the best possible fit of the proposed mechanistic structure to the observed data.
 In the following subsections we describe our approach to calibrating the three proposed mechanistic models to observed cholera incidence data.
@@
 
-Model~1 describes cholera dynamics at the nationally aggregated scale so that $\bm{Y}^{(1)}_n \in \R$ for each observation time $n \in 1:N$, and the latent state vector $\bm{X}^{(1)}(t)$ is comprised of national level population counts.
-Several algorithms exist, both frequentist and Bayesian, that can be used to obtain estimates of the parameters for this class of models.
+One of the most frequently used approaches to calibrate the parameters of a statistical model is maximum likelihood estimation.
+In the case of a POMP model, the likelihood function---as described in Sec.~\ref{sec:models}---is written as $\mathcal{L}(\paramVec) = \int f_{\bm{X}_{0:N}, \bm{Y}_{1:N}}\big(\bm{x}_{0:N}, \bm{y}_{1:N};\paramVec\big) d\bm{x}_{0:N}$.
+The high-dimensional integral in the likelihood, and the potentially non-linear nature of the process model, makes it impossible to obtain a closed form expression of the likelihood function of POMP models, except in the most basic cases.
+For example, the process model that defines Model~1 was developed with the intent of providing a plausible description of cholera dynamics rather than a providing statistically convenient description.
+Because of this, a closed form expression of the likelihood of Model~1 is not readily available.
+
 In order to retain the ability to propose models that are scientifically meaningful rather than only those that are simply statistically convenient, we restrict ourselves to parameter estimation techniques that have the plug-and-play property, which is that the fitting procedure only requires the ability to simulate the latent process instead of evaluating transition densities \citep{breto09,he10}.
-Plug-and-play algorithms include Bayesian approaches like ABC and PMCMC \citep{toni09,andrieu10}, but here we use frequentist methods to maximize model likelihoods.
-To our knowledge, the only plug-and-play, frequentist methods that can maximize the likelihood for POMP models of this complexity are iterated filtering algorithms \citep{ionides15}, which modify the well-known particle filter \citep{arulampalam02} by performing a random walk for each parameter and particle.
-These perturbations are carried out iteratively over multiple filtering operations, using the collection of parameters from the previous filtering pass as the parameter initialization for the next iteration, and decreasing the random walk variance at each step.
+Plug-and-play algorithms include Bayesian approaches like ABC and PMCMC \citep{toni09,andrieu10}, but here we use algorithms that enable maximum likelihood estimation.
+To our knowledge, the only plug-and-play methods that can maximize the likelihood for POMP models of this complexity are iterated filtering algorithms \citep{ionides15}, which modify the well-known \emph{particle filter} \citep{arulampalam02}.
+The particle filter is often used to obtain values of latent states, but can also be used to obtain likelihood estimates of a POMP model.
+Iterated filtering algorithms, like IF2 \citep{ionides15}, extend the particle filter by performing a random walk for each parameter and particle;
+these perturbations are carried out iteratively over multiple filtering operations, using the collection of parameters from the previous filtering pass as the parameter initialization for the next iteration, and decreasing the random walk variance at each step.
 
@@
 and can be justified by asymptotic considerations in a large-population limit \citep{dadlani2020,ndii17}.
-Because the process model of Model~2 is deterministic, the parameter estimation problem for Model~2 reduces to a least squares calculation when combined with a Gaussian measurement model (see Sec.~S3.2 of the supplement material for details).
+Because the process model of Model~2 is deterministic, maximum likelihood estimation reduces to a least squares calculation when combined with a Gaussian measurement model (see Sec.~S3.2 of the supplement material for details).
 \cite{lee20} fit two versions of Model~2 based on a presupposed change in cholera transmission from a epidemic phase to endemic phase that occurred in March, 2014.
@@
 This algorithm extends the work of \cite{ning21ibpf}, who provided theoretic justification for the version of the algorithm that only estimates unit-specific parameters.
-While computationally intensive, the IBPF algorithm enables us to directly estimate the parameters of models describing high-dimensional partially-observed nonlinear dynamic systems via likelihood maximization.
+The IBPF algorithm enables us to directly estimate the parameters of models describing high-dimensional partially-observed nonlinear dynamic systems via likelihood maximization.
 The ability to directly estimate parameters of Model~3 is responsible for the large increase in model likelihoods reported in Table~\ref{tab:likes}, as no algorithm with similar capabilities existed when \cite{lee20} published their results.
@@
 Here, we compare with a forecast projected from the filtering distribution: to obtain a forecast using POMP model $m$ at a collection of times $t_{N+1:N+s}$, where $N$ is the index for the last available data, we simulate forward from a draw from
-$f^{(m)}_{\bm{X}_N|\bm{Y}_{1:N}}(\bm{x}_{N} | \bm{y}^*_{1:N} ; \hat\paramVec)$
+$f_{\bm{X}_N|\bm{Y}_{1:N}}(\bm{x}_{N} | \bm{y}^*_{1:N} ; \hat\paramVec)$
 where $\hat\paramVec$ is a vector of calibrated parameters.
@@
 For frequentist inference, one possible approach is obtaining model forecasts from various values of $\paramVec$, where the values of $\paramVec$ are sampled proportionally according to their corresponding likelihoods \citep{king15}; see Sec.~S7 of the supplement material for more details.
-Both of these approaches share the similarity that parameters are chosen for the forecast approximately in proportion to their corresponding value of the likelihood function, $f^{(m)}_{\bm{Y}_{1:N}}(\bm{y}_{1:N}^*; \paramVec)$.
+Both of these approaches share the similarity that parameters are chosen for the forecast approximately in proportion to their corresponding value of the likelihood function, $f_{\bm{Y}_{1:N}}(\bm{y}_{1:N}^*; \paramVec)$.
 In this analysis, we do not construct forecasts accounting for parameter uncertainty as our focus is on the estimation and diagnosis of mechanistic models.

### 2023-06-29 ms.Rnw
No longer check for existing LVL 3 runs

@@
 
-# Check if the run_level_3 results exist. If they do, load these results.
-# If not, compute the run_level_<RUN_LEVEL> results and save them.
-if (RUN_LEVEL != 3 & file.exists("model1/run_level_3/lee1_epi_evals.rds")) {
-  lee_epi_ll <- readRDS("model1/run_level_3/lee1_epi_evals.rds")
-} else {
+# Compute the run_level_<RUN_LEVEL> results and save them.
 lee_epi_ll <- bake(
@@
 
+    # Randomly sample from starting params used by Lee et al (2020).
+    # If RUN_LEVEL == 3, all starting parameters are used.
     which_params <- sample(
@@
       }
-        lee_epi_ll[j, 'pfLL'] <- logmeanexp(ll_evals, se = TRUE)[1]
-        lee_epi_ll[j, 'pfse'] <- logmeanexp(ll_evals, se = TRUE)[2]
+      lee_epi_ll[j, 'pfLL'] <- logmeanexp(ll_evals, se = TRUE)[1]  # Log-lik
+      lee_epi_ll[j, 'pfse'] <- logmeanexp(ll_evals, se = TRUE)[2]  # SE
     }
@@
 )
-}
 
@@
 
-if (RUN_LEVEL != 3 & file.exists("model1/run_level_3/lee1_end_evals.rds")) {
-  lee_end_ll <- readRDS("model1/run_level_3/lee1_end_evals.rds")
-} else {
 lee_end_ll <- bake(
@@
 # Check RUN_LEVEL, and set hyper-parameters to IF2 and pfilter accordingly.
-#
 #                      RUN_LEVEL =    1,    2,      3
@@
 
-# Check first if RUN_LEVEL_3 file exists in directory, if not, create the
-# RUN_LEVEL version of the results and save them.
-if (RUN_LEVEL != 3 & file.exists("model1/run_level_3/haiti1_fit.rds")) {
-  h1_MIF <- readRDS("model1/run_level_3/haiti1_fit.rds")
-} else {
+# Create the <RUN_LEVEL> version of the results and save them.
   h1_MIF <- bake(
     file = paste0("model1/", rl_dir, "haiti1_fit.rds"), {
-      fit_haiti1(
+      fit_haiti1(  # haitipkg function used to fit Model 1
         NP = NP_H1,
@@
 
-# Create vectors for the unit and shared parameters
+# Create vectors for the unit and shared parameters that are estimated
 unit_specific_names <- c("betaB", "foi_add", "aHur", "hHur", "Iinit")
 shared_param_names <- c(
-  "mu_B", "XthetaA", "thetaI", "lambdaR", "r", "std_W",
-  "epsilon", "k"
-)
-
-# betaB + foi_add for all 10 units +
-# aHur3, aHur9, hHur3, hHur9 +
-# Iinit3, Iinit7, Iinit9 +
-# shared_parameters
-mod3_n_params <- 2 * 10 + 7 + length(shared_param_names)
-
-est_param_names <- c(
-  unit_specific_names, shared_param_names
-)
+  "mu_B", "XthetaA", "thetaI", "lambdaR", "r", "std_W", "epsilon", "k"
+)
+
+est_param_names <- c(unit_specific_names, shared_param_names)
 
@@
 
+# Simple function that is used to set the rw_sd for the first search
+# This function is convenient so that we don't have to type out the rw_sd
+# for each unit parameter.
 set_rw_1 <- function(x) {
@@
 
+# Simple function that is used to set the rw_sd for the second search
+# This function is convenient so that we don't have to type out the rw_sd
+# for each unit parameter.
 set_rw_2 <- function(x) {
@@
 
+# Simple function that is used to set the rw_sd for the third search
+# This function is convenient so that we don't have to type out the rw_sd
+# for each unit parameter.
 set_rw_3 <- function(x) {
@@
 
-# Search hyper-parameters for each run-level and search
-if (RUN_LEVEL == 1){
-  SEARCH1 = list(
-    NBPF = 15,
-    NP = 500,
-    SPAT_REGRESSION = 0.05,
-    NREPS = 36,
-    NP_EVAL = 500,
-    NREPS_EVAL = 3,
-    RW_SD = chol_rw_1.sd,
-    COOLING = 0.5,
-    KEEP_TRACES = TRUE
-  )
-
-  SEARCH2 = list(
-    TOP_N = 9,
-    NBPF = 10,
-    NP = 500,
-    SPAT_REGRESSION = 0.05,
-    NREPS = 4,
-    NP_EVAL = 500,
-    NREPS_EVAL = 3,
-    RW_SD = chol_rw_2.sd,
-    COOLING = 0.5,
-    KEEP_TRACES = TRUE
-  )
-
-  SEARCH3 = list(
-    TOP_N = 6,
-    NBPF = 10,
-    NP = 500,
-    SPAT_REGRESSION = 0.05,
-    NREPS = 6,
-    NP_EVAL = 500,
-    NREPS_EVAL = 3,
-    RW_SD = chol_rw_3.sd,
-    COOLING = 0.5,
-    KEEP_TRACES = TRUE
-  )
-
-  N_SEARCHES <- 3L
-} else if (RUN_LEVEL == 2) {
-  SEARCH1 = list(
-    NBPF = 100,
-    NP = 1000,
-    SPAT_REGRESSION = 0.05,
-    NREPS = 72,
-    NP_EVAL = 1000,
-    NREPS_EVAL = 8,
-    RW_SD = chol_rw_1.sd,
-    COOLING = 0.5,
-    KEEP_TRACES = TRUE,
-    KEEP_LIK_MAT = TRUE
-  )
-
-  SEARCH2 = list(
-    TOP_N = 9,
-    NBPF = 100,
-    NP = 1000,
-    SPAT_REGRESSION = 0.05,
-    NREPS = 4,
-    NP_EVAL = 1000,
-    NREPS_EVAL = 9,
-    RW_SD = chol_rw_2.sd,
-    COOLING = 0.5,
-    KEEP_TRACES = TRUE,
-    KEEP_LIK_MAT = TRUE
-  )
-
-  SEARCH3 = list(
-    TOP_N = 6,
-    NBPF = 100,
-    NP = 1000,
-    SPAT_REGRESSION = 0.05,
-    NREPS = 6,
-    NP_EVAL = 1000,
-    NREPS_EVAL = 9,
-    RW_SD = chol_rw_3.sd,
-    COOLING = 0.5,
-    KEEP_TRACES = TRUE,
-    KEEP_LIK_MAT = TRUE
-  )
-
-  N_SEARCHES <- 3L
-} else if (RUN_LEVEL == 3) {
-
-  SEARCH1 = list(
-    NBPF = 100,
-    NP = 2000,
-    SPAT_REGRESSION = 0.05,
-    NREPS = 144,
-    NP_EVAL = 2000,
-    NREPS_EVAL = 18,
-    RW_SD = chol_rw_1.sd,
-    COOLING = 0.5,
-    KEEP_TRACES = FALSE,
-    KEEP_LIKE_MAT = FALSE
-  )
-
-  SEARCH2 = list(
-    TOP_N = 12,
-    NBPF = 100,
-    NP = 2000,
-    SPAT_REGRESSION = 0.05,
-    NREPS = 6,
-    NP_EVAL = 2000,
-    NREPS_EVAL = 18,
-    RW_SD = chol_rw_2.sd,
-    COOLING = 0.5,
-    KEEP_TRACES = FALSE,
-    KEEP_LIK_MAT = FALSE
-  )
-
-  SEARCH3 = list(
-    TOP_N = 9,
-    NBPF = 100,
-    NP = 2000,
-    SPAT_REGRESSION = 0.05,
-    NREPS = 8,
-    NP_EVAL = 2000,
-    NREPS_EVAL = 18,
-    RW_SD = chol_rw_3.sd,
-    COOLING = 0.5,
-    KEEP_TRACES = FALSE,
-    KEEP_LIK_MAT = TRUE
-  )
-
-  N_SEARCHES <- 3L
-}
-
-if (RUN_LEVEL != 3 & file.exists("model3/run_level_3/haiti3_coupled_fit.rds")) {
-  haiti3_fit <- readRDS("model3/run_level_3/haiti3_coupled_fit.rds")
-} else {
+# Search hyper-parameters for each run-level and search. This sets
+# parameters like number of particles, number of BPF iterations, and
+# number of replications for each search.
+h3_fit_parms <- set_h3_fit_parms(
+  RUN_LEVEL = RUN_LEVEL,
+  rw_1.sd = chol_rw_1.sd,
+  rw_2.sd = chol_rw_2.sd,
+  rw_3.sd = chol_rw_3.sd
+)
+
+# Calculate the number of parameters that are estimated:
+# This is calculated by the number of non-zero rw_sd values (including expressions), but then subtracting of repeated shared parameters.
+mod3_n_params <- sum(reg_rw_1.sd != 0, na.rm = TRUE) + sum(sapply(reg_rw_1.sd, is.expression)) - length(shared_param_names) * 9
+
+# Fit Model 3
 haiti3_fit <- bake(
@@
     fit_coupled_haiti3(
-        nsearches = N_SEARCHES,
-        search1 = SEARCH1,
-        search2 = SEARCH2,
-        search3 = SEARCH3,
+      nsearches = h3_fit_parms$N_SEARCHES,
+      search1 = h3_fit_parms$SEARCH1,
+      search2 = h3_fit_parms$SEARCH2,
+      search3 = h3_fit_parms$SEARCH3,
       search_rho = FALSE,
@@
 
-# Check if run_level_3 results exit, and use them if they do.
-if (RUN_LEVEL != 3 & file.exists("model3/run_level_3/h3Sims.rds")) {
-  h3_quants <- readRDS("model3/run_level_3/h3Sims.rds")
-} else {
+# Get quantiles of simulations from the calibrated Model 3.
 h3_quants <- bake(
@@
 # Plot the reported cases and confidence region for each department.
-
 ggplot() +
@@
 # Calculate likelihood via block-particle filter
-if (RUN_LEVEL != 3 & file.exists("model3/run_level_3/mod3_eval_res.rds")) {
-  mod3_eval_res <- readRDS("model3/run_level_3/mod3_eval_res.rds")
-} else{
 mod3_eval_res <- bake(
@@
 
-MODEL3_CASES <- haitiCholera %>%
-  dplyr::rename(
-    date = date_saturday, Grande_Anse = Grand.Anse,
-    Nord_Est = Nord.Est, Nord_Ouest = Nord.Ouest,
-    Sud_Est = Sud.Est
-  ) %>%
-  dplyr::mutate(date = as.Date(date)) %>%
-  dplyr::select(-report) %>%
-  dplyr::mutate(time_diff = date - lag(date))
-
-jumps <- which(MODEL3_CASES$time_diff > lubridate::make_difftime(day = 7))
-
-temp <- MODEL3_CASES
-for (j in 1:length(jumps)) {
-  temp <- rbind(
-    temp[1:(jumps[j] + j - 2), ],
-    rep(NA, ncol(temp)),
-    temp[(jumps[j] + j - 1):nrow(temp), ]
-  )
-}
-
-mod3_cases <- temp %>%
-  dplyr::select(-time_diff)
-
-mod3_departements <-
-  c(
-    'Artibonite',
-    'Centre',
-    'Grande_Anse',
-    'Nippes',
-    'Nord',
-    'Nord-Est',
-    'Nord-Ouest',
-    'Ouest',
-    'Sud',
-    'Sud-Est'
-  )
+mod3_cases <- as.data.frame(t(h3Spat@data))
 
 m3_benchmark <- 0
-for (dep in mod3_departements) {
+for (dep in 1:10) {
 
   # Get department cases
-  cases <- as.numeric(mod3_cases[, gsub("-", "_", dep)])
+  cases <- as.numeric(mod3_cases[, dep])
 
@@
 ARMA_benchmarks[['m3']] <- m3_benchmark
-rm(cases, m3_a201, m3_log_a201, m3_benchmark, mod3_departements, dep,
+rm(cases, m3_a201, m3_log_a201, m3_benchmark, dep,
    mod3_cases, temp, jumps, MODEL3_CASES, i, j)
@@
 
-if (RUN_LEVEL != 3 & file.exists("model3/run_level_3/lee3_results.rds")) {
-  lee3_results <- readRDS("model3/run_level_3/lee3_results.rds")
-} else {
 lee3_results <- bake(
@@
 mod3_lee_coupled_partial_aic <- 2 * lee3_results$n_params - 2 * lee3_results$sub_ll_eval
-
 @
@@
 # projects the vaccination scenario, under the specified model.
-if (RUN_LEVEL != 3 & file.exists("model1/run_level_3/S0_sims.rds")) {
-  h1_S0_results <- readRDS("model1/run_level_3/S0_sims.rds")
-} else {
 h1_S0_results <- bake(
@@
 # projects the vaccination scenario, under the specified model.
-if (RUN_LEVEL != 3 & file.exists("model1/run_level_3/S2_sims.rds")) {
-  h1_S2_results <- readRDS("model1/run_level_3/S2_sims.rds")
-} else {
 h1_S2_results <- bake(
@@
 # projects the vaccination scenario, under the specified model.
-if (RUN_LEVEL != 3 & file.exists("model1/run_level_3/S4_sims.rds")) {
-  h1_S4_results <- readRDS("model1/run_level_3/S4_sims.rds")
-} else {
 h1_S4_results <- bake(
@@
 # projects the vaccination scenario, under the specified model.
-if (RUN_LEVEL != 3 & file.exists("model1/run_level_3/S3_sims.rds")) {
-  h1_S3_results <- readRDS("model1/run_level_3/S3_sims.rds")
-} else {
 h1_S3_results <- bake(
@@
 # projects the vaccination scenario, under the specified model.
-if (RUN_LEVEL != 3 & file.exists("model1/run_level_3/S1_sims.rds")) {
-  h1_S1_results <- readRDS("model1/run_level_3/S1_sims.rds")
-} else {
 h1_S1_results <- bake(
@@
 )
-}
 @
@@
 
-if (RUN_LEVEL != 3 & file.exists("model3/run_level_3/h3_bpf_end_states.rds")) {
-  h3_bpf_end_states <- readRDS("model3/run_level_3/h3_bpf_end_states.rds")
-} else {
 h3_bpf_end_states <- bake(
@@
 )
-}
-
-if (RUN_LEVEL != 3 & file.exists("model3/run_level_3/mod3_V0_res.rds")) {
-  mod3_V0_res <- readRDS("model3/run_level_3/mod3_V0_res.rds")
-} else {
+
 mod3_V0_res <- bake(
@@
 )
-}
-
-if (RUN_LEVEL != 3 & file.exists("model3/run_level_3/mod3_V1_res.rds")) {
-  mod3_V1_res <- readRDS("model3/run_level_3/mod3_V1_res.rds")
-} else {
+
 mod3_V1_res <- bake(
@@
 )
-}
-
-if (RUN_LEVEL != 3 & file.exists("model3/run_level_3/mod3_V2_res.rds")) {
-  mod3_V2_res <- readRDS("model3/run_level_3/mod3_V2_res.rds")
-} else {
+
 mod3_V2_res <- bake(
@@
     h3_V2_res
-
-    }
-  )
-}
-
-if (RUN_LEVEL != 3 & file.exists("model3/run_level_3/mod3_V3_res.rds")) {
-  mod3_V3_res <- readRDS("model3/run_level_3/mod3_V3_res.rds")
-} else {
+  }
+)
+
 mod3_V3_res <- bake(
@@
 )
-}
-
-if (RUN_LEVEL != 3 & file.exists("model3/run_level_3/mod3_V4_res.rds")) {
-  mod3_V4_res <- readRDS("model3/run_level_3/mod3_V4_res.rds")
-} else {
+
 mod3_V4_res <- bake(
@@
 <<Get Model 1 Seasonality>>=
-# plot_model_fun <- function(mod) {
-#   params <- coef(mod)
-#
-#   beta1 <- params['beta1']
-#   beta2 <- params['beta2']
-#   beta3 <- params['beta3']
-#   beta4 <- params['beta4']
-#   beta5 <- params['beta5']
-#   beta6 <- params['beta6']
-#   betat <- params['betat']
-#
-#   get_beta <- function(time) {
-#
-#     covar <- t(mod@covar@table)
-#     covar_df <- as.data.frame(covar)
-#     covar_df$times <- mod@covar@times
-#
-#     betas <- matrix(c(beta1, beta2, beta3, beta4, beta5, beta6), nrow = 1)
-#     si <- matrix(unlist(covar_df[time, -7]), ncol = 1)
-#     if (time <= 430) {
-#       return(as.numeric(exp(betas %*% si + betat * ((time - 215) / (430-215)))))
-#     } else {
-#       return(as.numeric(exp(betas %*% si + betat)))
-#     }
-#   }
-#
-#   return(get_beta)
-# }
-#
-# get_beta1 <- plot_model_fun(h1)
-#
-# h1_seas_df <- data.frame(
-#   'week' = 431:800,
-#   'date' = lubridate::ymd("2010-10-16") + lubridate::weeks(431:800),
-#   'trans' = purrr::map_dbl(431:800, get_beta1)
-# ) %>%
-#   mutate(year = lubridate::year(date)) %>%
-#   filter(year == 2020) %>%
-#   mutate(trans_std = (trans - min(trans)) / (max(trans) - min(trans)))
 
@@
 
-# std_rain <- function(x) {
-#   # This function simply standardizes the rain for us.
-#   x / max(x)
-# }
-
 df_rain <- haitiRainfall %>%
-  # dplyr::summarize(
-  #   # Standardize rainfall value using the function above
-  #   date = date, dplyr::across(Artibonite:`Sud-Est`, std_rain)
-  # ) %>%
   pivot_longer(  # Convert data to "long" format

### 2023-06-30 ms.Rnw
minor edit

@@
 mod3_ll <- max(
-  haiti3_fit[[length(haiti3_fit)]]$logLiks[, 'logLik'],
+  haiti3_fit[[length(haiti3_fit)]]$logLiks$logLik,
   na.rm = TRUE

### 2023-07-12 ms-plos.Rnw
started working on plos paper

--- 2023-06-30 ms.Rnw
@@
-%% Based on the template for the submission to:
-%%   The Annals of Applied Statistics [AOAS]
-
-%\documentclass[aoas]{imsart}
-\documentclass[aoas]{imsart}
-
-%% Packages
+% Template for PLoS
+% Version 3.6 Aug 2022
+%
+% % % % % % % % % % % % % % % % % % % % % %
+%
+% -- IMPORTANT NOTE
+%
+% This template contains comments intended
+% to minimize problems and delays during our production
+% process. Please follow the template instructions
+% whenever possible.
+%
+% % % % % % % % % % % % % % % % % % % % % % %
+%
+% Once your paper is accepted for publication,
+% PLEASE REMOVE ALL TRACKED CHANGES in this file
+% and leave only the final text of your manuscript.
+% PLOS recommends the use of latexdiff to track changes during review, as this will help to maintain a clean tex file.
+% Visit https://www.ctan.org/pkg/latexdiff?lang=en for info or contact us at latex@plos.org.
+%
+%
+% There are no restrictions on package use within the LaTeX files except that no packages listed in the template may be deleted.
+%
+% Please do not include colors or graphics in the text.
+%
+% The manuscript LaTeX source should be contained within a single file (do not use \input, \externaldocument, or similar commands).
+%
+% % % % % % % % % % % % % % % % % % % % % % %
+%
+% -- FIGURES AND TABLES
+%
+% Please include tables/figure captions directly after the paragraph where they are first cited in the text.
+%
+% DO NOT INCLUDE GRAPHICS IN YOUR MANUSCRIPT
+% - Figures should be uploaded separately from your manuscript file.
+% - Figures generated using LaTeX should be extracted and removed from the PDF before submission.
+% - Figures containing multiple panels/subfigures must be combined into one image file before submission.
+% For figure citations, please use "Fig" instead of "Figure".
+% See http://journals.plos.org/plosone/s/figures for PLOS figure guidelines.
+%
+% Tables should be cell-based and may not contain:
+% - spacing/line breaks within cells to alter layout or alignment
+% - do not nest tabular environments (no tabular environments within tabular environments)
+% - no graphics or colored text (cell background color/shading OK)
+% See http://journals.plos.org/plosone/s/tables for table guidelines.
+%
+% For tables that exceed the width of the text column, use the adjustwidth environment as illustrated in the example table in text below.
+%
+% % % % % % % % % % % % % % % % % % % % % % % %
+%
+% -- EQUATIONS, MATH SYMBOLS, SUBSCRIPTS, AND SUPERSCRIPTS
+%
+% IMPORTANT
+% Below are a few tips to help format your equations and other special characters according to our specifications.
+For more tips to help reduce the possibility of formatting errors during conversion, please see our LaTeX guidelines at http://journals.plos.org/plosone/s/latex
+%
+% For inline equations, please be sure to include all portions of an equation in the math environment.
+For example, x$^2$ is incorrect; this should be formatted as $x^2$ (or $\mathrm{x}^2$ if the romanized font is desired).
+%
+% Do not include text that is not math in the math environment. For example, CO2 should be written as CO\textsubscript{2} instead of CO$_2$.
+%
+% Please add line breaks to long display equations when possible in order to fit size of the column.
+%
+% For inline equations, please do not include punctuation (commas, etc) within the math environment unless this is part of the equation.
+%
+% When adding superscript or subscripts outside of brackets/braces, please group using {}.  For example, change "[U(D,E,\gamma)]^2" to "{[U(D,E,\gamma)]}^2".
+%
+% Do not use \cal for caligraphic font.  Instead, use \mathcal{}
+%
+% % % % % % % % % % % % % % % % % % % % % % % %
+%
+% Please contact latex@plos.org with any questions.
+%
+% % % % % % % % % % % % % % % % % % % % % % % %
+
+\documentclass[10pt,letterpaper]{article}
+\usepackage[top=0.85in,left=2.75in,footskip=0.75in]{geometry}
+
+% amsmath and amssymb packages, useful for mathematical formulas and symbols
+\usepackage{amsmath,amssymb}
+
+% Use adjustwidth environment to exceed column width (see example table in text)
+\usepackage{changepage}
+
+% textcomp package and marvosym package for additional characters
+\usepackage{textcomp,marvosym}
+
+% cite package, to clean up citations in the main text. Do not remove.
+\usepackage{cite}
+
+% Use nameref to cite supporting information files (see Supporting Information section for more info)
+\usepackage{nameref,hyperref}
+
+% line numbers
+\usepackage[right]{lineno}
+
+% ligatures disabled
+\usepackage[nopatch=eqnum]{microtype}
+\DisableLigatures[f]{encoding = *, family = * }
+
+% color can be used to apply background shading to table cells only
+\usepackage[table]{xcolor}
+
+% array package and thick rules for tables
+\usepackage{array}
+
+% create "+" rule type for thick vertical lines
+\newcolumntype{+}{!{\vrule width 2pt}}
+
+% create \thickcline for thick horizontal lines of variable length
+\newlength\savedwidth
+\newcommand\thickcline[1]{%
+  \noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
+  \cline{#1}%
+  \noalign{\vskip\arrayrulewidth}%
+  \noalign{\global\arrayrulewidth\savedwidth}%
+}
+
+% \thickhline command for thick horizontal lines that span the table
+\newcommand\thickhline{\noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
+\hline
+\noalign{\global\arrayrulewidth\savedwidth}}
+
+% Remove comment for double spacing
+%\usepackage{setspace}
+%\doublespacing
+
+% Text layout
+\raggedright
+\setlength{\parindent}{0.5cm}
+\textwidth 5.25in
+\textheight 8.75in
+
+% Bold the 'Figure #' in the caption and separate it from the title/caption with a period
+% Captions will be left justified
+\usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,justification=raggedright,singlelinecheck=off]{caption}
+\renewcommand{\figurename}{Fig}
+
+% Use the PLoS provided BiBTeX style
+\bibliographystyle{plos2015}
+
+% Remove brackets from numbering in List of References
+\makeatletter
+\renewcommand{\@biblabel}[1]{\quad#1.}
+\makeatother
+
+% Header and Footer with logo
+\usepackage{lastpage,fancyhdr,graphicx}
+\usepackage{epstopdf}
+%\pagestyle{myheadings}
+\pagestyle{fancy}
+\fancyhf{}
+%\setlength{\headheight}{27.023pt}
+%\lhead{\includegraphics[width=2.0in]{PLOS-submission.eps}}
+\rfoot{\thepage/\pageref{LastPage}}
+\renewcommand{\headrulewidth}{0pt}
+\renewcommand{\footrule}{\hrule height 2pt \vspace{2mm}}
+\fancyheadoffset[L]{2.25in}
+\fancyfootoffset[L]{2.25in}
+\lfoot{\today}
+
+%% Include all macros below
+
+\newcommand{\lorem}{{\bf LOREM}}
+\newcommand{\ipsum}{{\bf IPSUM}}
+
+%% END MACROS SECTION
+
 \RequirePackage{amsthm,amsmath,amsfonts,amssymb,graphicx,enumerate,url,xr,lmodern}
-\RequirePackage[authoryear]{natbib}
 \usepackage{url}
@@
 \usepackage[mathscr]{euscript}
-
-\startlocaldefs
 
@@
 
-\endlocaldefs
+%% START CUSTOM MACROS
+
+%% END CUSTOM MACROS
 
 \begin{document}
+\vspace*{0.2in}
+
+% Title must be 250 characters or less.
+\begin{flushleft}
+{\Large
+\textbf\newline{Title of submission to PLOS journals} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
+}
+\newline
+% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
+\\
+Jesse Wheeler\textsuperscript{1*},
+AnnaElaine L. Rosengart\textsuperscript{2},
+Zhuoxun Jiang\textsuperscript{1},
+Kevin Tan\textsuperscript{3},
+Noah Treutle\textsuperscript{1},
+Edward L. Ionides\textsuperscript{1}
+\\
+\bigskip
+\textbf{1} Statistics Department, University of Michigan, Ann Arbor, Michigan, USA
+\\
+\textbf{2} Affiliation Dept/Program/Center, Institution Name, City, State, Country
+\\
+\textbf{3} Affiliation Dept/Program/Center, Institution Name, City, State, Country
+\\
+\bigskip
+
+% Insert additional author notes using the symbols described below. Insert symbol callouts after author names as necessary.
+%
+% Remove or comment out the author notes below if they aren't used.
+%
+% Primary Equal Contribution Note
+% \Yinyang These authors contributed equally to this work.
+
+% Additional Equal Contribution Note
+% Also use this double-dagger symbol for special authorship notes, such as senior authorship.
+% \ddag These authors also contributed equally to this work.
+
+% Current address notes
+% \textcurrency Current Address: Dept/Program/Center, Institution Name, City, State, Country % change symbol to "\textcurrency a" if more than one current address note
+% \textcurrency b Insert second current address
+% \textcurrency c Insert third current address
+
+% Deceased author note
+% \dag Deceased
+
+% Group/Consortium Author Note
+% \textpilcrow Membership list can be found in the Acknowledgments section.
+
+% Use the asterisk to denote corresponding authorship and provide email address in note below.
+* jeswheel@umich.edu
+
+\end{flushleft}
 
@@
 opts_chunk$set(
-    progress = TRUE,
     concordance = TRUE,
-    prompt = TRUE,
-    highlight = FALSE,
     tidy = TRUE,
+    message = FALSE,
     tidy.opts = list(
@@
     comment = "",
-    warning = FALSE,
-    message = FALSE,
-    error = TRUE,
     echo = FALSE,
-    cache = FALSE,
-    strip.white = TRUE,
-    # results="markup",
-    background = "#FFFFFF00",
-    size = "normalsize",
-    fig.path = "figure/",
-    fig.lp = "fig:",
-    fig.align = "left",
-    fig.show = "asis",
-    dev = "pdf",
     dev.args = list(
@@
         pointsize = 9
-    )
+    ),
+    fig.path = "testFigs/"
 )
@@
 
-\begin{frontmatter}
-
-\title{Informing policy via dynamic models: Cholera in Haiti}
-\runtitle{Cholera in Haiti}
-
-\begin{aug}
-
-\author[A]{\fnms{Jesse}~\snm{Wheeler}\ead[label=e1]{jeswheel@umich.edu}\orcid{0000-0003-3941-3884}},
-\author[B]{\fnms{AnnaElaine L.}~\snm{Rosengart}\ead[label=e2]{arosenga@andrew.cmu.edu}}
-\author[A]{\fnms{Zhuoxun}~\snm{Jiang}\ead[label=e3]{zhuoxunj@umich.edu}},
-\author[A]{\fnms{Kevin}~\snm{Tan}\ead[label=e4]{kevtan@umich.edu}},
-\author[A]{\fnms{Noah}~\snm{Treutle}\ead[label=e5]{ntreutle@umich.edu}}
-\and
-\author[A]{\fnms{Edward L.}~\snm{Ionides}\ead[label=e6]{ionides@umich.edu}}
-
-\address[A]{Statistics Department, University of Michigan\printead[presep={,\ }]{e1,e3,e4,e5,e6}}
-
-\address[B]{Statistics \& Data Science Department, Carnegie Mellon University\printead[presep={,\ }]{e2}}
-
-\end{aug}
-
-\begin{abstract}
-% \begin{center}
-% Working draft, \today
-% \end{center}
-% \vspace{3mm}
-
+% Please keep the abstract below 300 words
+\section*{Abstract}
 Public health decisions must be made about when and how to implement interventions to control an infectious disease epidemic.
@@
 Our workflow is reproducible and extendable, facilitating future investigations of this disease system.
-\end{abstract}
-
-\begin{keyword}
-  \kwd{Partially observed Markov process}
-  \kwd{Hidden Markov model}
-  \kwd{infectious disease}
-  \kwd{cholera}
-  \kwd{sequential Monte Carlo}
-\end{keyword}
-
-\end{frontmatter}
-
-\section{Introduction}
+
+% Please keep the Author Summary between 150 and 200 words
+% Use first person. PLOS ONE authors please skip this step.
+% Author Summary not valid for PLOS ONE submissions.
+\section*{Author summary}
+Lorem ipsum dolor sit amet, consectetur adipiscing elit.
+Curabitur eget porta erat.
+Morbi consectetur est vel gravida pretium.
+Suspendisse ut dui eu ante cursus gravida non sed sem.
+Nullam sapien tellus, commodo id velit id, eleifend volutpat quam.
+Phasellus mauris velit, dapibus finibus elementum vel, pulvinar non tellus.
+Nunc pellentesque pretium diam, quis maximus dolor faucibus id.
+Nunc convallis sodales ante, ut ullamcorper est egestas vitae.
+Nam sit amet enim ultrices, ultrices elit pulvinar, volutpat risus.
+
+\linenumbers
+
+% Use "Eq" instead of "Equation" for equation citations.
+\section*{Introduction}
 
@@
 This case study investigates the use of dynamic models and spatiotemporal data to inform a policy decision in the context of the cholera outbreak in Haiti, which started in 2010.
-We build on a multi-group modeling exercise by \citet{lee20} in which four expert modeling teams developed models to the same dataset with the goal of comparing conclusions on the feasibility of eliminating cholera by a vaccination campaign.
+We build on a multi-group modeling exercise by \cite{lee20} in which four expert modeling teams developed models to the same dataset with the goal of comparing conclusions on the feasibility of eliminating cholera by a vaccination campaign.
 Model~1 is stochastic and describes cholera at the national level;
@@
 Model~4 has an agent-based construction, featuring considerable mechanistic detail but limited ability to calibrate these details to data.
-We focus on Models~1--3, as the strengths and weaknesses of the agent-based modeling approach  \citep{tracy18} are outside the scope of this article.
+We focus on Models~1--3, as the strengths and weaknesses of the agent-based modeling approach \cite{tracy18} are outside the scope of this article.
 
@@
 potential effect of prospective oral cholera vaccine (OCV) programs.
-While OCV is accepted as a safe and effective tool for controlling the spread of cholera, the global stockpile of OCV doses remains limited \citep{pezzoli20}.
-Advances in OCV technology and vaccine availability, however, raised the possibility of planning a national vaccination program \citep{lee20}.
+While OCV is accepted as a safe and effective tool for controlling the spread of cholera, the global stockpile of OCV doses remains limited \cite{pezzoli20}.
+Advances in OCV technology and vaccine availability, however, raised the possibility of planning a national vaccination program \cite{lee20}.
 In the study, certain data were shared between the groups, including demography and vaccination history; vaccine efficacy was also fixed at a shared value between groups.
 Beyond this, the groups made autonomous decisions on what to include and exclude from their models.
-The groups largely adhered to existing guidelines on creating models to inform policy \citep{behrend20,saltelli20} and, despite their autonomy, obtained a consensus that an extensive nationwide vaccination campaign would be necessary to eliminate cholera from Haiti.
-Their conclusion is inconsistent with the prolonged period with no confirmed cholera cases between February, 2019 and September, 2022 \citep{trevisin22}, though cholera has recently reemerged in Haiti \citep{rubin22}.
-
-The discrepancy between the model based conclusions of \cite{lee20} and the prolonged absence of cholera in Haiti has been debated \citep{francois20,rebaudetComment20,henrys20,leeReply20}.
-Suggested origins of this discrepancy include the use of unrealistic models \citep{rebaudetComment20} and unrealistic criteria for cholera elimination \citep{henrys20}.
+The groups largely adhered to existing guidelines on creating models to inform policy \cite{behrend20,saltelli20} and, despite their autonomy, obtained a consensus that an extensive nationwide vaccination campaign would be necessary to eliminate cholera from Haiti.
+Their conclusion is inconsistent with the prolonged period with no confirmed cholera cases between February, 2019 and September, 2022 \cite{trevisin22}, though cholera has recently reemerged in Haiti \cite{rubin22}.
+
+The discrepancy between the model based conclusions of \cite{lee20} and the prolonged absence of cholera in Haiti has been debated \cite{francois20,rebaudetComment20,henrys20,leeReply20}.
+Suggested origins of this discrepancy include the use of unrealistic models \cite{rebaudetComment20} and unrealistic criteria for cholera elimination \cite{henrys20}.
 We find a more nuanced conclusion: attention to methodological details in model fitting, diagnosis and forecasting can improve each of the proposed model's ability to quantitatively describe observed data.
@@
 While this work is focused on the models proposed by \cite{lee20}, our suggestions have broader relevance.
-To investigate the extent to which \citet{lee20} is typical of the substantial body of work on Haiti cholera dynamics, we preformed a literature review by searching PubMed with keywords: Haiti, cholera, model.
+To investigate the extent to which \cite{lee20} is typical of the substantial body of work on Haiti cholera dynamics, we preformed a literature review by searching PubMed with keywords: Haiti, cholera, model.
 The search resulted in 66 papers, of which 32 used dynamic models to describe the cholera epidemic in Haiti.
-The models make various choices on the dichotomies considered by \citet{lee20}; deterministic versus stochastic; compartment model versus agent-based; aggregated versus spatially explicit.
-This is no accident, since \citet{lee20} purposefully designed their study to cover the range of current modeling practice.
+The models make various choices on the dichotomies considered by \cite{lee20}; deterministic versus stochastic; compartment model versus agent-based; aggregated versus spatially explicit.
+This is no accident, since \cite{lee20} purposefully designed their study to cover the range of current modeling practice.
 
@@
 
-\section{Mechanistic models for cholera in Haiti}\label{sec:models}
+\section*{Materials and methods}
+\subsection*{Mechanistic models for disease modeling}
 Models that focus on learning relationships between variables in a dataset are called {\it associative}, whereas models that incorporate a known scientific property of the system are called {\it causal} or {\it mechanistic}.
 The danger in using forecasting techniques which rely on associative models to predict the consequence of interventions is called the Lucas critique in an econometric context.
-\citet{lucas76} pointed out that it is naive to predict the effects of an intervention on a given system based entirely on historical associations.
+\cite{lucas76} pointed out that it is naive to predict the effects of an intervention on a given system based entirely on historical associations.
 To successfully predict the effect of an intervention, a model should therefore both provide a quantitative explanation of existing data and should have a causal interpretation: a manipulation of the system should correspond quantitatively with the corresponding change to the model.
@@
 
-The four mechanistic models of \citet{lee20} were deliberately developed with limited coordination.
+The four mechanistic models of \cite{lee20} were deliberately developed with limited coordination.
 This allows us to treat the models as fairly independently developed expert approaches to understanding cholera transmission.
@@
 Here, we have put all three models into a common notational framework.
-Translations back to the original notation of \citet{lee20} are given in Table~S-1 of the supplement.
+Translations back to the original notation of \cite{lee20} are given in Table~S-1 of the supplement.
 
@@
 
-\subsection{Model~1}
+\subsubsection*{Model~1}
 \label{sec:model1}
@@
 \cite{lee20} included process noise in Model~3 but not in Model~1, i.e., they fixed $\sigmaProc^2 = 0$.
-Gamma white noise in the transmission rate gives rise to an over-dispersed latent Markov process \citep{breto11} which has been found to improve the statistical fit of disease transmission models \citep{stocks20,he10}.
+Gamma white noise in the transmission rate gives rise to an over-dispersed latent Markov process \cite{breto11} which has been found to improve the statistical fit of disease transmission models \cite{stocks20,he10}.
 
 For any time point in $t_{1:N}$, the process model describes the movement of individuals from one latent state compartment to another.
-Per-capita transition rates are given in Equations~\ref{model1:SE}-\ref{model1:birth}:
+Per-capita transition rates are given in Eqs.~\ref{model1:SE}-\ref{model1:birth}:
 \begin{eqnarray}
@@
 
-\subsection{Model~2}
+\subsubsection*{Model~2}
 \label{sec:model2}
@@
 In \eqref{model2:mu_SE}, $\vaccineEfficacy_\vaccCounter$ denotes the vaccine efficacy for each vaccination campaign $\vaccCounter \in \vaccClass$, with $\vaccineEfficacy_0 = 1$, $\vaccineEfficacy_1 = 1 - 0.429 \childReduce$, $\vaccineEfficacy_2 = 1 - 0.519 \childReduce$, $\vaccineEfficacy_3 = 1 - 0.429$, and $\vaccineEfficacy_4 = 1 - 0.519$
-Here, $\childReduce = 0.4688$ represents the reduced efficacy of the vaccination for children under the age of five years, and the values $0.429$ and $0.519$ are the median effectiveness of one and two doses over their effective period respectively, according to Table S4 in the supplement material of \citet{lee20}.
+Here, $\childReduce = 0.4688$ represents the reduced efficacy of the vaccination for children under the age of five years, and the values $0.429$ and $0.519$ are the median effectiveness of one and two doses over their effective period respectively, according to Table S4 in the supplement material of \cite{lee20}.
 Because vaccine efficacy remains constant, individuals in this model transition from a vaccinated compartment to the susceptible compartment at the end of the vaccine coverage period.
@@
 
-\subsection{Model~3}
+\subsubsection*{Model~3}
 \label{sec:model3}
@@
 \end{equation}
-where $\tHur$ is the time Hurricane Matthew struck Haiti \citep{ferreirai16}, and $1_{(A)}$ is the indicator function for event $A$.
+where $\tHur$ is the time Hurricane Matthew struck Haiti \cite{ferreirai16}, and $1_{(A)}$ is the indicator function for event $A$.
 In \cite{lee20}, $\Whur{_u}$ and $\hHur{u}$ were set to zero for all $u$;
@@
 
-\section{Statistical Analysis}\label{sec:methods}
-
-We consider model fitting (Sec.~\ref{sec:model_fitting}) followed by diagnostic investigations (Sec.~\ref{sec:model_diagnostics}), forecasting (Sec.~\ref{sec:filter}) and external scientific corroboration of model fit (Sec.~\ref{sec:science}).
-
 <<FitModel1, echo=FALSE, message=FALSE, include=FALSE>>=
@@
 
+haiti3_fit <- readRDS("model3/run_level_3/h3_revision.rds")
+
 # Fit Model 3
-haiti3_fit <- bake(
-  file = paste0("model3/", rl_dir, "haiti3_coupled_fit.rds"), {
-    fit_coupled_haiti3(
-      nsearches = h3_fit_parms$N_SEARCHES,
-      search1 = h3_fit_parms$SEARCH1,
-      search2 = h3_fit_parms$SEARCH2,
-      search3 = h3_fit_parms$SEARCH3,
-      search_rho = FALSE,
-      search_gamma = FALSE,
-      search_Iinit = TRUE,
-      search_hur = TRUE,
-      ncores = cores
-    )
-  }
-)
+# haiti3_fit <- bake(
+#   file = paste0("model3/", rl_dir, "haiti3_coupled_fit.rds"), {
+#     fit_coupled_haiti3(
+#       nsearches = h3_fit_parms$N_SEARCHES,
+#       search1 = h3_fit_parms$SEARCH1,
+#       search2 = h3_fit_parms$SEARCH2,
+#       search3 = h3_fit_parms$SEARCH3,
+#       search_rho = FALSE,
+#       search_gamma = FALSE,
+#       search_Iinit = TRUE,
+#       search_hur = TRUE,
+#       ncores = cores
+#     )
+#   }
+# )
 
@@
 
-\subsection{Model Fitting}\label{sec:model_fitting}
+\subsection*{Model Fitting}\label{sec:model_fitting}
 
@@
 
-\subsubsection{Calibrating Model~1 Parameters}\label{sec:fit1}
+\subsubsection*{Calibrating Model~1 Parameters}\label{sec:fit1}
 
@@
 
-In order to retain the ability to propose models that are scientifically meaningful rather than only those that are simply statistically convenient, we restrict ourselves to parameter estimation techniques that have the plug-and-play property, which is that the fitting procedure only requires the ability to simulate the latent process instead of evaluating transition densities \citep{breto09,he10}.
-Plug-and-play algorithms include Bayesian approaches like ABC and PMCMC \citep{toni09,andrieu10}, but here we use algorithms that enable maximum likelihood estimation.
-To our knowledge, the only plug-and-play methods that can maximize the likelihood for POMP models of this complexity are iterated filtering algorithms \citep{ionides15}, which modify the well-known \emph{particle filter} \citep{arulampalam02}.
+In order to retain the ability to propose models that are scientifically meaningful rather than only those that are simply statistically convenient, we restrict ourselves to parameter estimation techniques that have the plug-and-play property, which is that the fitting procedure only requires the ability to simulate the latent process instead of evaluating transition densities \cite{breto09,he10}.
+Plug-and-play algorithms include Bayesian approaches like ABC and PMCMC \cite{toni09,andrieu10}, but here we use algorithms that enable maximum likelihood estimation.
+To our knowledge, the only plug-and-play methods that can maximize the likelihood for POMP models of this complexity are iterated filtering algorithms \cite{ionides15}, which modify the well-known \emph{particle filter} \cite{arulampalam02}.
 The particle filter is often used to obtain values of latent states, but can also be used to obtain likelihood estimates of a POMP model.
-Iterated filtering algorithms, like IF2 \citep{ionides15}, extend the particle filter by performing a random walk for each parameter and particle;
+Iterated filtering algorithms, like IF2 \cite{ionides15}, extend the particle filter by performing a random walk for each parameter and particle;
 these perturbations are carried out iteratively over multiple filtering operations, using the collection of parameters from the previous filtering pass as the parameter initialization for the next iteration, and decreasing the random walk variance at each step.
@@
 The proposal of a linear trend in transmission is a result of observing an apparent decrease in reported cholera infections from 2012-2019 in Fig.~\ref{CholeraData}.
-While several factors may contribute to this decrease, one explanation is that case-area targeted interventions (CATIs), which included education sessions, increased monitoring, household decontamination, soap distribution, and water chlorination in infected areas \citep{rebaudet19CATI}, may have substantially reduced cholera transmission \citep{rebaudet21}.
+While several factors may contribute to this decrease, one explanation is that case-area targeted interventions (CATIs), which included education sessions, increased monitoring, household decontamination, soap distribution, and water chlorination in infected areas \cite{rebaudet19CATI}, may have substantially reduced cholera transmission \cite{rebaudet21}.
 
 We perform a statistical test to determine whether or not the data indicate the presence of a linear trend in transmissibility.
-To do this, we perform a profile-likelihood search on the parameter $\transmissionTrend$ and obtain a $95\%$ confidence interval via a Monte Carlo Adjusted Profile (MCAP) \citep{ionides17}.
+To do this, we perform a profile-likelihood search on the parameter $\transmissionTrend$ and obtain a $95\%$ confidence interval via a Monte Carlo Adjusted Profile (MCAP) \cite{ionides17}.
 \cite{lee20} implemented Model~1 by fitting two distinct phases: an epidemic phase from October 2010 through March 2015, and an endemic phase from March 2015 onward.
@@
 
-We implemented Model~1 using the \code{pomp} package \citep{king16}, relying heavily on the \code{pomp} source code provided by \cite{lee20}.
+We implemented Model~1 using the \code{pomp} package \cite{king16}, relying heavily on the \code{pomp} source code provided by \cite{lee20}.
 Both analyses used the \code{mif2} implementation of the IF2 algorithm to estimate $\paramVec$ by maximum likelihood.
@@
 
-\subsubsection{Calibrating Model~2 Parameters}\label{sec:fit2}
+\subsubsection*{Calibrating Model~2 Parameters}\label{sec:fit2}
 
 Model~2 is a deterministic compartmental model defined by a set of coupled differential equations.
-The use of deterministic compartment models have a long history in the field of infectious disease epidemiology \citep{kermack1927,brauer2017,giordano20},
-and can be justified by asymptotic considerations in a large-population limit \citep{dadlani2020,ndii17}.
+The use of deterministic compartment models have a long history in the field of infectious disease epidemiology \cite{kermack1927,brauer2017,giordano20},
+and can be justified by asymptotic considerations in a large-population limit \cite{dadlani2020,ndii17}.
 Because the process model of Model~2 is deterministic, maximum likelihood estimation reduces to a least squares calculation when combined with a Gaussian measurement model (see Sec.~S3.2 of the supplement material for details).
@@
 
-We implemented this model using the \code{spatPomp} \code{R} package \citep{asfaw23arxiv}.
-The model was then fit using the subplex algorithm, implemented in the  \code{subplex} package \citep{king2020Subplex}.
+We implemented this model using the \code{spatPomp} \code{R} package \cite{asfaw23arxiv}.
+The model was then fit using the subplex algorithm, implemented in the  \code{subplex} package \cite{king2020Subplex}.
 A comparison of the trajectory of the fitted model to the data is given in Fig~\ref{fig:mod2Traj}.
@@
 
-\subsubsection{Calibrating Model~3 Parameters}\label{sec:fit3}
+\subsubsection*{Calibrating Model~3 Parameters}\label{sec:fit3}
 Model~3 describes cholera dynamics in Haiti using a metapopulation model, where the hidden states in each administrative department has an effect on the dynamics in other departments.
 The decision to address metapopulation dynamics using a spatially explicit model, rather than to aggregate over space, is double-edged.
-Evidence for the former approach has been provided in previous studies \citep{king15}, including the specific case of heterogeneity between Haitian departments in cholera transmission \citep{collins14}.
-However, a legitimate preference for simplicity \citep{saltelli20,green15} can support a decision to consider nationally aggregated models.
+Evidence for the former approach has been provided in previous studies \cite{king15}, including the specific case of heterogeneity between Haitian departments in cholera transmission \cite{collins14}.
+However, a legitimate preference for simplicity \cite{saltelli20,green15} can support a decision to consider nationally aggregated models.
 One issue that arises when fitting spatially explicit models is that parameter estimation techniques based on the particle filter become computationally intractable as the number of spatial units increases.
-This is a result of the approximation error of particle filters growing exponentially in the dimension of the model \citep{rebeschini15,park20}.
-
-To avoid the approximation error present in high-dimensional models, \citet{lee20} simplified the problem of estimating the parameters of Model~3 by creating an approximate version of the model where the units are independent given the observed data.
+This is a result of the approximation error of particle filters growing exponentially in the dimension of the model \cite{rebeschini15,park20}.
+
+To avoid the approximation error present in high-dimensional models, \cite{lee20} simplified the problem of estimating the parameters of Model~3 by creating an approximate version of the model where the units are independent given the observed data.
 Reducing a spatially coupled model to individual units in this fashion requires special treatment of any interactive mechanisms between spatial units, such as found in Eq.~\myeqref{eq:model3:foi}.
@@
 
-\subsection{Model Diagnostics} \label{sec:model_diagnostics}
+\subsection*{Model Diagnostics}\label{sec:model_diagnostics}
 
@@
 Model misspecification, which may be thought of as the omission of a mechanism in the model that is an important feature of the dynamic system, is inevitable at all levels of model complexity.
-To make progress, while accepting proper limitations, one must bear in mind the much-quoted observation of \citet{box79} that ``all models are wrong but some are useful.''
-Beyond being good practical advice for applied statistics, this assertion is relevant for the philosophical justification of statistical inference as severe testing \citep[][Sec.~4.8]{mayo18}.
+To make progress, while accepting proper limitations, one must bear in mind the much-quoted observation of \cite{box79} that ``all models are wrong but some are useful.''
+Beyond being good practical advice for applied statistics, this assertion is relevant for the philosophical justification of statistical inference as severe testing \cite{mayo18} % \citep[][Sec.~4.8]{mayo18}.
 In this section, we discuss some tools for diagnosing mechanistic models with the goal of making the subjective assessment of model ``usefulness'' more objective.
@@
 While visual comparisons can be informative, they provide only a weak and informal measure of the goodness-of-fit of a model.
-The study by \citet{lee20} provides an example of this: their models and parameter estimates resulted in simulations that visually resembled the observed data, yet resulted in model likelihoods that were---in some cases---considerably smaller than likelihoods that can be achieved via the likelihood based optimization techniques that were used (see Table~\ref{tab:likes}).
+The study by \cite{lee20} provides an example of this: their models and parameter estimates resulted in simulations that visually resembled the observed data, yet resulted in model likelihoods that were---in some cases---considerably smaller than likelihoods that can be achieved via the likelihood based optimization techniques that were used (see Table~\ref{tab:likes}).
 Alternative forms of model validation should therefore be used in conjunction with visual comparisons of simulations to observed data.
@@
 Another approach is to compare a quantitative measure of the model fit (such as MSE, predictive accuracy, or model likelihood) among all proposed models.
-These comparisons, which provide insight into how each model performs relative to the others, are quite common \citep{rinaldo12,sallah17}.
+These comparisons, which provide insight into how each model performs relative to the others, are quite common \cite{rinaldo12,sallah17}.
 To calibrate relative measures of fit, it is useful to compare against a model that has well-understood statistical ability to fit data, and we call this model a {\it benchmark}.
@@
 Ideally, the measure should also have favorable theoretical properties.
-Consequently, we focus on log-likelihood as a measure of goodness of fit, and we adjust for the degrees of freedom of the models to be compared by using the Akaike information criterion (AIC) \citep{aic74}.
+Consequently, we focus on log-likelihood as a measure of goodness of fit, and we adjust for the degrees of freedom of the models to be compared by using the Akaike information criterion (AIC) \cite{aic74}.
 
 In some cases, a possible benchmark model could be a generally accepted mechanistic model, but often no such model is available.
-Because of this, we use a log-linear Gaussian ARMA model as an associative benchmark, as recommended by \citet{he10}.
+Because of this, we use a log-linear Gaussian ARMA model as an associative benchmark, as recommended by \cite{he10}.
 The theory and practice of ARMA models is well developed, and these linear models are appropriate on a log scale due to the exponential growth and decay characteristic of biological dynamics.
@@
 In addition, the published goodness of fit provides a concrete point of comparison for subsequent scientific investigations.
-When combined with online availability of data and code, objective measures of fit provide a powerful tool to accelerate scientific progress, following the paradigm of the {\it common task framework} \citep[][Sec.~6]{donoho17}.
+When combined with online availability of data and code, objective measures of fit provide a powerful tool to accelerate scientific progress, following the paradigm of the {\it common task framework} \cite{donoho17} % \citep[][Sec.~6]{donoho17}.
 
@@
 \caption{\label{tab:likes}AIC values for each model compared to their ARMA benchmark.
-Values in parentheses are corresponding values using \citet{lee20} parameter estimates.
-$\textsuperscript{1}$The reported likelihood is an upper bound of the likelihood of the \citet{lee20} model.
+Values in parentheses are corresponding values using \cite{lee20} parameter estimates.
+$\textsuperscript{1}$The reported likelihood is an upper bound of the likelihood of the \cite{lee20} model.
 $\textsuperscript{2}$\cite{lee20} fit Model~3 to a subset of the data (March 2014 onward, excluding data from Ouest in 2015-2016).
@@
 Additional techniques include the inspection of conditional log-likelihoods of each observation given the previous observations in order to understand how well the model describes each data point (Sec.~S5).
-Other tools include plotting the effective sample size of each observation \citep{liu01}; plotting the values of the hidden states from simulations (supplement Sec.~5.1);
-and comparing summary statistics of the observed data to simulations from the model \citep{wood10,king15}.
-
-\subsection{Forecasts}\label{sec:filter}
+Other tools include plotting the effective sample size of each observation \cite{liu01}; plotting the values of the hidden states from simulations (supplement Sec.~5.1);
+and comparing summary statistics of the observed data to simulations from the model \cite{wood10,king15}.
+
+\subsection*{Corroborating Fitted Models with Scientific Knowledge}
+
+The resulting mechanisms in a fitted model can be compared to current scientific knowledge about a system.
+Agreement between model based inference and our current understanding of a system may be taken as a confirmation of both model based conclusions and our scientific understanding.
+On the other hand, comparisons may generate unexpected results that have the potential to spark new scientific knowledge \cite{ganusov16}.
+
+In the context of our case study, we demonstrate how the fit of Model~1 corroborates other evidence concerning the role of rainfall in cholera epidemics.
+Specifically, we examine the results of fitting the flexible cubic spline term in Model~1 (Eq.~\myeqref{model1:lambda}--\myeqref{model1:beta}).
+The cubic splines permit flexible estimation of seasonality in the force of infection, $\transmission(t)$.
+Fig.~\ref{fig:h1SeasRain} shows that the estimated seasonal transmission rate $\transmission$ mimics the rainfall dynamics in Haiti, despite Model~1 not having access to rainfall data.
+This is consistent with previous studies finding that rainfall played an important role in cholera transmission in Haiti \cite{lemaitre19,eisenberg13}.
+The estimated seasonality also features an increased transmission rate during the fall, which was noticed at an earlier stage of the epidemic \cite{rinaldo12}; the high transimission rate in the fall may also be a result of the increase transission that occured in the fall of 2016, when hurricane Matthew struck Haiti.
+
+<<Get Model 1 Seasonality>>=
+
+h1_seas_df <- get_h1_seasonality(h1)
+
+df_rain <- haitiRainfall %>%
+  pivot_longer(  # Convert data to "long" format
+    data = .,
+    cols = -date,
+    names_to = 'dep',
+    values_to = 'rainfall'
+  ) %>%
+  mutate(year = lubridate::year(date), week = lubridate::week(date)) %>%
+  group_by(year, week) %>%
+  summarize(nwr = sum(rainfall)) %>%
+  ungroup() %>%  # Not needed, but good practice
+  mutate(
+    date_week_start = lubridate::ymd(paste0(2014, "-01-01")) +
+      lubridate::weeks(week - 1),
+    week_date = as.Date(date_week_start, format = "%m-%d"),
+    std_national_rain = (nwr - min(nwr)) / (max(nwr) - min(nwr))
+  )
+
+# Calculate the average national weekly rainfall.
+mean_rain <- df_rain %>%
+  group_by(week_date) %>%
+  summarize(mean_rain = mean(std_national_rain)) %>%
+  ungroup() %>%
+  mutate(mean_rain = (mean_rain - min(mean_rain)) / ((max(mean_rain) - min(mean_rain))))
+
+# Create a figure showing annual rainfal, scaled from 0-1. This is top figure
+gg_rain <- df_rain %>%
+  filter(week != 1, week != max(week)) %>%
+  ggplot(aes(x = week_date, y = std_national_rain, group = year, col = year)) +
+  geom_line() +
+  theme_bw() +
+  guides(color = 'none') +
+  ylab('Standardized Weekly\nNational Rain') +
+  theme(  # Remove elements that are un-unsed becuase this is top figure
+    axis.title.x = element_blank(),
+    axis.ticks.x = element_blank(),
+    axis.text.x = element_blank(),
+    axis.text.y = element_text(size = 7.5),
+    axis.title.y = element_text(size = 9)
+  )
+
+# Create plot that shows the seasonality of Model 1, alongside the average
+# weekly national rainfall.
+gg_trans <- ggplot() +
+  geom_line(data = h1_seas_df, aes(x = date, y = trans_std), linetype = 'dashed') +
+  geom_line(
+    data = mean_rain,
+    aes(x = week_date + lubridate::years(6), y = mean_rain),
+    col = '#2166ac'
+  ) +
+  scale_x_date(date_labels = "%b", date_breaks = '1 month') +
+  ylab('Standardized Seasonal\nContact Rate') +
+  theme(axis.title.x = element_blank(),
+        axis.text = element_text(size = 7.5),
+        axis.title.y = element_text(size = 9))
+
+@
+
+\begin{figure}
+\centering
+<<Model1_Seasonality_Figure, fig.height=2.9, fig.width=5, fig.align='center'>>=
+cowplot::plot_grid(gg_rain, gg_trans, align = "v", ncol = 1)
+@
+\caption{\label{fig:h1SeasRain}
+(Top) weekly rainfall in Haiti, lighter colors representing more recent years.
+(Bottom) estimated seasonality in the transmission rate (dashed line) plotted alongside mean rainfall (solid line).
+The outsided effect of rainfall in the fall may be due to Hurricane Matthew, which struck Haiti in October of 2016 and resulted in an increase of cholera cases in the nation.
+}
+\end{figure}
+
+For any model-based inference, it is important to recognize and assess the modeling simplifications and assumptions that were used in order to arrive at the conclusions.
+In epidemiological studies, for example, quantitative understanding of individual-level processes may not perfectly match model parameters that were fit to population-level case counts, even when the model provides a strong statistical fit \cite{he10}.
+This makes direct interpretation of estimated parameters delicate.
+
+Our case study provides an example of this in the parameter estimate for the duration of natural immunity due to cholera infection, $\muRS^{-1}$.
+Under the framework of Model~2, the best estimate for this parameter is $\Sexpr{signif(1/h2_fit$h2_params["sigma"],3)} \, \mathrm{yr}$, suggesting that individuals have effectively permanent immunity to cholera once infected.
+To interpret this result, we bear in mind that the data ranged from 2010-2019, and therefore estimates of immunity longer than $10\, \mathrm{yr}$---the upper end of previous estimates of natural immunity \cite{king08}---effectively result in the same model dynamics.
+The depletion of susceptible individuals may also be attributed to confounding mechanisms---such as localized vaccination programs and non-pharmaceutical interventions that reduce cholera transmission \cite{trevisin22, rebaudet21}---that were not accounted for in the model.
+Perhaps the best interpretation of the estimated parameter, then, is that under model assumptions, the model most adequately describes the observed data by having a steady decrease in the number of susceptible individuals.
+The weak statistical fit of Model~2 compared to a log-linear benchmark (see Table~\ref{tab:likes}) cautions us against drawing quantitative conclusions from this model.
+A model that has a poor statistical fit may nevertheless provide a useful conceptual framework for thinking about the system under investigation.
+However, a claim that the model has been validated against data should be reserved for situations where the model provides a statistical fit that is competitive against alternative explanations.
+
+% Place figure captions after the first paragraph in which they are cited.
+\begin{figure}[!h]
+\caption{{\bf Bold the figure title.}
+Figure caption text here, please use this space for the figure panel descriptions instead of using subfigure commands. A: Lorem ipsum dolor sit amet. B: Consectetur adipiscing elit.}
+\label{fig1}
+\end{figure}
+
+% Results and Discussion can be combined.
+\section*{Results}
+
+\subsection*{Forecasts}\label{sec:filter}
 
@@
 Forecasts from mechanistic models that are compatible with current scientific understanding may also provide estimates of the future effects of potential interventions.
-Further, they may enable real-time testing of new scientific hypotheses \citep{lewis22}.
+Further, they may enable real-time testing of new scientific hypotheses \cite{lewis22}.
 
@@
 This assertion may seem self-evident, but it is not the case for deterministic models, for which the initial conditions together with the parameters are sufficient for forecasting, and so recent data do not have special importance.
-Epidemiological forecasts based on deterministic models are not uncommon in practice, despite their limitations \citep{king15}.
-That may explain why \citet{lee20} chose to obtain forecasts by simulating the calibrated models forward from initial conditions.
+Epidemiological forecasts based on deterministic models are not uncommon in practice, despite their limitations \cite{king15}.
+That may explain why \cite{lee20} chose to obtain forecasts by simulating the calibrated models forward from initial conditions.
 Here, we compare with a forecast projected from the filtering distribution: to obtain a forecast using POMP model $m$ at a collection of times $t_{N+1:N+s}$, where $N$ is the index for the last available data, we simulate forward from a draw from
@@
 
-Uncertainty in just a single parameter can lead to drastically different forecasts \citep{saltelli20}.
+Uncertainty in just a single parameter can lead to drastically different forecasts \cite{saltelli20}.
 Therefore, parameter uncertainty should also be considered when obtaining model forecasts to influence policy.
 If a Bayesian technique is used for parameter estimation, a natural way to account for parameter uncertainty is to obtain simulations from the model where each simulation is obtained using parameters drawn from the estimated posterior distribution.
-For frequentist inference, one possible approach is obtaining model forecasts from various values of $\paramVec$, where the values of $\paramVec$ are sampled proportionally according to their corresponding likelihoods \citep{king15}; see Sec.~S7 of the supplement material for more details.
+For frequentist inference, one possible approach is obtaining model forecasts from various values of $\paramVec$, where the values of $\paramVec$ are sampled proportionally according to their corresponding likelihoods \cite{king15}; see Sec.~S7 of the supplement material for more details.
 Both of these approaches share the similarity that parameters are chosen for the forecast approximately in proportion to their corresponding value of the likelihood function, $f_{\bm{Y}_{1:N}}(\bm{y}_{1:N}^*; \paramVec)$.
@@
 
-The primary forecasting goal of \citet{lee20} was to investigate the potential consequences of vaccination interventions on a system to inform policy.
+The primary forecasting goal of \cite{lee20} was to investigate the potential consequences of vaccination interventions on a system to inform policy.
 One outcome of their study include estimates for the probability of cholera elimination under several possible vaccination scenarios.
@@
 \begin{table}
+\begin{adjustwidth}{-2.25in}{0in}
 \centering
@@
 $^*$ denotes parameters added during our re-analysis, not considered by Lee et al.
-Translations back into the notation of \citet{lee20} are given in Table~S1.
-}
+Translations back into the notation of \cite{lee20} are given in Table~S1.
+}
+\end{adjustwidth}
 \end{table}
 
-\subsection{Corroborating Fitted Models with Previous Scientific Knowledge}\label{sec:science}
-
-The resulting mechanisms in a fitted model can be compared to current scientific knowledge about a system.
-Agreement between model based inference and our current understanding of a system may be taken as a confirmation of both model based conclusions and our scientific understanding.
-On the other hand, comparisons may generate unexpected results that have the potential to spark new scientific knowledge \citep{ganusov16}.
-
-In the context of our case study, we demonstrate how the fit of Model~1 corroborates other evidence concerning the role of rainfall in cholera epidemics.
-Specifically, we examine the results of fitting the flexible cubic spline term in Model~1 (Eq.~\myeqref{model1:lambda}--\myeqref{model1:beta}).
-The cubic splines permit flexible estimation of seasonality in the force of infection, $\transmission(t)$.
-Fig.~\ref{fig:h1SeasRain} shows that the estimated seasonal transmission rate $\transmission$ mimics the rainfall dynamics in Haiti, despite Model~1 not having access to rainfall data.
-This is consistent with previous studies finding that rainfall played an important role in cholera transmission in Haiti \citep{lemaitre19,eisenberg13}.
-The estimated seasonality also features an increased transmission rate during the fall, which was noticed at an earlier stage of the epidemic \citep{rinaldo12}; the high transimission rate in the fall may also be a result of the increase transission that occured in the fall of 2016, when hurricane Matthew struck Haiti.
-
-<<Get Model 1 Seasonality>>=
-
-h1_seas_df <- get_h1_seasonality(h1)
-
-df_rain <- haitiRainfall %>%
-  pivot_longer(  # Convert data to "long" format
-    data = .,
-    cols = -date,
-    names_to = 'dep',
-    values_to = 'rainfall'
-  ) %>%
-  mutate(year = lubridate::year(date), week = lubridate::week(date)) %>%
-  group_by(year, week) %>%
-  summarize(nwr = sum(rainfall)) %>%
-  ungroup() %>%  # Not needed, but good practice
-  mutate(
-    date_week_start = lubridate::ymd(paste0(2014, "-01-01")) +
-      lubridate::weeks(week - 1),
-    week_date = as.Date(date_week_start, format = "%m-%d"),
-    std_national_rain = (nwr - min(nwr)) / (max(nwr) - min(nwr))
-  )
-
-# Calculate the average national weekly rainfall.
-mean_rain <- df_rain %>%
-  group_by(week_date) %>%
-  summarize(mean_rain = mean(std_national_rain)) %>%
-  ungroup() %>%
-  mutate(mean_rain = (mean_rain - min(mean_rain)) / ((max(mean_rain) - min(mean_rain))))
-
-# Create a figure showing annual rainfal, scaled from 0-1. This is top figure
-gg_rain <- df_rain %>%
-  filter(week != 1, week != max(week)) %>%
-  ggplot(aes(x = week_date, y = std_national_rain, group = year, col = year)) +
-  geom_line() +
-  theme_bw() +
-  guides(color = 'none') +
-  ylab('Standardized Weekly\nNational Rain') +
-  theme(  # Remove elements that are un-unsed becuase this is top figure
-    axis.title.x = element_blank(),
-    axis.ticks.x = element_blank(),
-    axis.text.x = element_blank(),
-    axis.text.y = element_text(size = 7.5),
-    axis.title.y = element_text(size = 9)
-  )
-
-# Create plot that shows the seasonality of Model 1, alongside the average
-# weekly national rainfall.
-gg_trans <- ggplot() +
-  geom_line(data = h1_seas_df, aes(x = date, y = trans_std), linetype = 'dashed') +
-  geom_line(
-    data = mean_rain,
-    aes(x = week_date + lubridate::years(6), y = mean_rain),
-    col = '#2166ac'
-  ) +
-  scale_x_date(date_labels = "%b", date_breaks = '1 month') +
-  ylab('Standardized Seasonal\nContact Rate') +
-  theme(axis.title.x = element_blank(),
-        axis.text = element_text(size = 7.5),
-        axis.title.y = element_text(size = 9))
-
-@
-
-\begin{figure}
+% Place tables after the first paragraph in which they are cited.
+\begin{table}[!ht]
+\begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
 \centering
-<<Model1_Seasonality_Figure, fig.height=2.9, fig.width=5, fig.align='center'>>=
-cowplot::plot_grid(gg_rain, gg_trans, align = "v", ncol = 1)
-@
-\caption{\label{fig:h1SeasRain}
-(Top) weekly rainfall in Haiti, lighter colors representing more recent years.
-(Bottom) estimated seasonality in the transmission rate (dashed line) plotted alongside mean rainfall (solid line).
-The outsided effect of rainfall in the fall may be due to Hurricane Matthew, which struck Haiti in October of 2016 and resulted in an increase of cholera cases in the nation.
-}
-\end{figure}
-
-For any model-based inference, it is important to recognize and assess the modeling simplifications and assumptions that were used in order to arrive at the conclusions.
-In epidemiological studies, for example, quantitative understanding of individual-level processes may not perfectly match model parameters that were fit to population-level case counts, even when the model provides a strong statistical fit \citep{he10}.
-This makes direct interpretation of estimated parameters delicate.
-
-Our case study provides an example of this in the parameter estimate for the duration of natural immunity due to cholera infection, $\muRS^{-1}$.
-Under the framework of Model~2, the best estimate for this parameter is $\Sexpr{signif(1/h2_fit$h2_params["sigma"],3)} \, \mathrm{yr}$, suggesting that individuals have effectively permanent immunity to cholera once infected.
-To interpret this result, we bear in mind that the data ranged from 2010-2019, and therefore estimates of immunity longer than $10\, \mathrm{yr}$---the upper end of previous estimates of natural immunity \citep{king08}---effectively result in the same model dynamics.
-The depletion of susceptible individuals may also be attributed to confounding mechanisms---such as localized vaccination programs and non-pharmaceutical interventions that reduce cholera transmission \citep{trevisin22, rebaudet21}---that were not accounted for in the model.
-Perhaps the best interpretation of the estimated parameter, then, is that under model assumptions, the model most adequately describes the observed data by having a steady decrease in the number of susceptible individuals.
-The weak statistical fit of Model~2 compared to a log-linear benchmark (see Table~\ref{tab:likes}) cautions us against drawing quantitative conclusions from this model.
-A model that has a poor statistical fit may nevertheless provide a useful conceptual framework for thinking about the system under investigation.
-However, a claim that the model has been validated against data should be reserved for situations where the model provides a statistical fit that is competitive against alternative explanations.
-
-\section{Robust interpretation of model based conclusions}\label{sec:advice}
+\caption{
+{\bf Table caption Nulla mi mi, venenatis sed ipsum varius, volutpat euismod diam.}}
+\begin{tabular}{|l+l|l|l|l|l|l|l|}
+\hline
+\multicolumn{4}{|l|}{\bf Heading1} & \multicolumn{4}{|l|}{\bf Heading2}\\ \thickhline
+$cell1 row1$ & cell2 row 1 & cell3 row 1 & cell4 row 1 & cell5 row 1 & cell6 row 1 & cell7 row 1 & cell8 row 1\\ \hline
+$cell1 row2$ & cell2 row 2 & cell3 row 2 & cell4 row 2 & cell5 row 2 & cell6 row 2 & cell7 row 2 & cell8 row 2\\ \hline
+$cell1 row3$ & cell2 row 3 & cell3 row 3 & cell4 row 3 & cell5 row 3 & cell6 row 3 & cell7 row 3 & cell8 row 3\\ \hline
+\end{tabular}
+\begin{flushleft} Table notes Phasellus venenatis, tortor nec vestibulum mattis, massa tortor interdum felis, nec pellentesque metus tortor nec nisl.
+Ut ornare mauris tellus, vel dapibus arcu suscipit sed.
+\end{flushleft}
+\label{table1}
+\end{adjustwidth}
+\end{table}
+
+\section*{Discussion}\label{sec:discussion}
+
+The ongoing global COVID-19 pandemic has provided a clear example on how government policy may be affected by the conclusions of scientific models \cite{saltelli20}.
+This article demonstrates that fitting appropriate scientific models to guide policy is a challenging statistical task.
+In our case study, we found that additional attention to statistical details could have resulted in improved statistical fits to the observed incidence data, leading to improvements in the accuracy of the resulting policy guidance.
+We used the same data and models, and even much of the same code, as \cite{lee20}, and yet ended up with drastically different conclusions.
+We acknowledge the benefit of hindsight: our demonstration of a statistically principled route to obtain better-fitting models resulting in more robust insights does not rule out the possibility of discovering other models that fit well yet predict poorly.
+
+Inference for mechanistic time series models offers opportunities for understanding and controlling complex dynamic systems.
+This case study has investigated issues requiring attention when applying powerful new statistical techniques that can enable statistically efficient inference for a general class of partially observed Markov process models.
+Researchers should check that the computationally intensive numerical calculations are carried out adequately.
+Comparison against benchmarks and alternative model specifications should be considered to evaluate the statistical goodness-of-fit.
+Once that is accomplished, care is required to assess what causal conclusions can properly be inferred given the possibility of alternative explanations consistent with the data.
+Studies that combine model development with thoughtful data analysis, supported by a high standard of reproducibility, build knowledge about the system under investigation.
+Cautionary warnings about the difficulties inherent in understanding complex systems \cite{saltelli20,ioannidis20,ganusov16} should motivate us to follow best practices in data analysis, rather than avoiding the challenge.
+
+\subsection*{Robust interpretation of model based conclusions}\label{sec:advice}
 
@@
 A causal interpretation is strengthened by corroborative evidence.
-For example, reconstructed latent variables (such as numbers of susceptible and recovered individuals) should make sense in the context of alternative measurements of these variables \citep{grad12}.
+For example, reconstructed latent variables (such as numbers of susceptible and recovered individuals) should make sense in the context of alternative measurements of these variables \cite{grad12}.
 In the context of our case study, by reconstructing the latent states of Model~3, we notice that the calibrated model favors higher levels of cholera transmission than what was typically observed in the incidence data (Sec.~S5.1 of the supplement).
@@
 
-\section{Discussion}\label{sec:discussion}
-
-The ongoing global COVID-19 pandemic has provided a clear example on how government policy may be affected by the conclusions of scientific models \citep{saltelli20}.
-This article demonstrates that fitting appropriate scientific models to guide policy is a challenging statistical task.
-In our case study, we found that additional attention to statistical details could have resulted in improved statistical fits to the observed incidence data, leading to improvements in the accuracy of the resulting policy guidance.
-We used the same data and models, and even much of the same code, as \citet{lee20}, and yet ended up with drastically different conclusions.
-We acknowledge the benefit of hindsight: our demonstration of a statistically principled route to obtain better-fitting models resulting in more robust insights does not rule out the possibility of discovering other models that fit well yet predict poorly.
-
-Inference for mechanistic time series models offers opportunities for understanding and controlling complex dynamic systems.
-This case study has investigated issues requiring attention when applying powerful new statistical techniques that can enable statistically efficient inference for a general class of partially observed Markov process models.
-Researchers should check that the computationally intensive numerical calculations are carried out adequately.
-Comparison against benchmarks and alternative model specifications should be considered to evaluate the statistical goodness-of-fit.
-Once that is accomplished, care is required to assess what causal conclusions can properly be inferred given the possibility of alternative explanations consistent with the data.
-Studies that combine model development with thoughtful data analysis, supported by a high standard of reproducibility, build knowledge about the system under investigation.
-Cautionary warnings about the difficulties inherent in understanding complex systems \citep{saltelli20,ioannidis20,ganusov16} should motivate us to follow best practices in data analysis, rather than avoiding the challenge.
-
-\subsection{Reproducibility and Extendability}
-
-\citet{lee20} published their code and data online, and this reproducibility facilitated our work.
+\subsection*{Reproducibility and Extendability}
+
+\cite{lee20} published their code and data online, and this reproducibility facilitated our work.
 Robust data analysis requires not only reproducibility but also extendability: if one wishes to try new model variations, or new approaches to fitting the existing models, or plotting the results in a different way, this should not be excessively burdensome.
-Scientific results are only trustworthy so far as they can be critically questioned, and an extendable analysis should facilitate such examination \citep{gentleman07}.
-
-We provide a strong form of reproducibility, as well as extendability, by developing our analysis in the context of a software package, \code{haitipkg}, written in the R language \citep{r}.
+Scientific results are only trustworthy so far as they can be critically questioned, and an extendable analysis should facilitate such examination \cite{gentleman07}.
+
+We provide a strong form of reproducibility, as well as extendability, by developing our analysis in the context of a software package, \code{haitipkg}, written in the R language \cite{r}.
 Using a software package mechanism supports documentation, standardization and portability that promote extendability.
-In the terminology of \citet{gentleman07}, the source code for this article is a {\it dynamic document} combining code chunks with text.
+In the terminology of \cite{gentleman07}, the source code for this article is a {\it dynamic document} combining code chunks with text.
 In addition to reproducing the article, the code can be extended to examine alternative analysis to that presented.
-The dynamic document, together with the R packages, form a {\it compendium}, defined by \citet{gentleman07} as a distributable and executable unit which combines data, text and auxiliary software (the latter meaning code written to run in a general-purpose, portable programming environment, which in this case is R).
-
-\begin{acks}[Acknowledgments]
-The authors would like to thank Mercedes Pascual and Betz Halloran for helpful discussions. Laura Matrajt provided additional data for the Model~2 analysis.
-\end{acks}
-
-\begin{funding}
-This work was supported by National Science Foundation grants DMS-1761603 and DMS-1646108.
-%
-% The second author was supported in part by ...
-\end{funding}
-
-\begin{supplement}
-\stitle{Eliminating cholera in Haiti: Supplement}
-\sdescription{
-This document contains additional details about the numerical implementations of Models~1--3, as well as a translation table that facilitates comparisons between these models and those described by \cite{lee20}.
-The supplement also demonstrates our capability to faithfully estimate the likelihood of the models of \cite{lee20}.
-}
-\end{supplement}
-
-\begin{supplement}
-\stitle{Auxiliary software: \code{haitipkg}}
-\sdescription{
-The \code{haitipkg} R package is maintained in a GitHub repository, \code{jeswheel/haitipkg}. The submitted version of this package is archived on Zenodo: \code{10.5281/zenodo.7557099}
-The package contains all of the data and code used to create, fit and explore the models presented in this article.
-}
-\end{supplement}
-
-\begin{supplement}
-\stitle{Dynamic document: \code{haiti\_article}}
-\sdescription{
-The Rnw \citep[R noweb,][]{ramsey94} files generating this article and its supplement are maintained in a GitHub repository, \code{jeswheel/haiti\_article}.
-The submitted version will be archived on Zenodo.
-}
-\end{supplement}
-
-\bibliographystyle{imsart-nameyear}
+The dynamic document, together with the R packages, form a {\it compendium}, defined by \cite{gentleman07} as a distributable and executable unit which combines data, text and auxiliary software (the latter meaning code written to run in a general-purpose, portable programming environment, which in this case is R).
+
+\section*{Supporting information}
+
+% Include only the SI item label in the paragraph heading. Use the \nameref{label} command to cite SI items in the text.
+\paragraph*{S1 Fig.}
+\label{S1_Fig}
+{\bf Bold the title sentence.} Add descriptive text after the title of the item (optional).
+
+\paragraph*{S2 Fig.}
+\label{S2_Fig}
+{\bf Lorem ipsum.} Maecenas convallis mauris sit amet sem ultrices gravida.
+Etiam eget sapien nibh.
+Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula.
+Curabitur fringilla pulvinar lectus consectetur pellentesque.
+
+\paragraph*{S1 File.}
+\label{S1_File}
+{\bf Lorem ipsum.}  Maecenas convallis mauris sit amet sem ultrices gravida.
+Etiam eget sapien nibh.
+Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula.
+Curabitur fringilla pulvinar lectus consectetur pellentesque.
+
+\paragraph*{S1 Video.}
+\label{S1_Video}
+{\bf Lorem ipsum.}  Maecenas convallis mauris sit amet sem ultrices gravida.
+Etiam eget sapien nibh.
+Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula.
+Curabitur fringilla pulvinar lectus consectetur pellentesque.
+
+\paragraph*{S1 Appendix.}
+\label{S1_Appendix}
+{\bf Lorem ipsum.} Maecenas convallis mauris sit amet sem ultrices gravida.
+Etiam eget sapien nibh.
+Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula.
+Curabitur fringilla pulvinar lectus consectetur pellentesque.
+
+\paragraph*{S1 Table.}
+\label{S1_Table}
+{\bf Lorem ipsum.} Maecenas convallis mauris sit amet sem ultrices gravida.
+Etiam eget sapien nibh.
+Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula.
+Curabitur fringilla pulvinar lectus consectetur pellentesque.
+
+\section*{Acknowledgments}
+Cras egestas velit mauris, eu mollis turpis pellentesque sit amet.
+Interdum et malesuada fames ac ante ipsum primis in faucibus.
+Nam id pretium nisi.
+Sed ac quam id nisi malesuada congue.
+Sed interdum aliquet augue, at pellentesque quam rhoncus vitae.
+
+\nolinenumbers
+
+% Either type in your references using
+% \begin{thebibliography}{}
+% \bibitem{}
+% Text
+% \end{thebibliography}
+%
+% or
+%
+% Compile your BiBTeX database using our plos2015.bst
+% style file and paste the contents of your .bbl file
+% here. See http://journals.plos.org/plosone/s/latex for
+% step-by-step instructions.
+%
+% \begin{thebibliography}{10}
+%
+% \bibitem{bib1}
+% Conant GC, Wolfe KH.
+% \newblock {{T}urning a hobby into a job: how duplicated genes find new
+%   functions}.
+% \newblock Nat Rev Genet. 2008 Dec;9(12):938--950.
+%
+% \bibitem{bib2}
+% Ohno S.
+% \newblock Evolution by gene duplication.
+% \newblock London: George Alien \& Unwin Ltd. Berlin, Heidelberg and New York:
+%   Springer-Verlag.; 1970.
+%
+% \bibitem{bib3}
+% Magwire MM, Bayer F, Webster CL, Cao C, Jiggins FM.
+% \newblock {{S}uccessive increases in the resistance of {D}rosophila to viral
+%   infection through a transposon insertion followed by a {D}uplication}.
+% \newblock PLoS Genet. 2011 Oct;7(10):e1002337.
+%
+% \end{thebibliography}
+
 \bibliography{bib-haiti}

### 2023-07-12 checkRes.R

@@ -1,0 +1,83 @@
+# Checking RL2 results.
+#
+# Jull 11
+
+library(pomp)
+library(spatPomp)
+library(tidyverse)
+
+results <- readRDS("model3/run_level_2/h3_revision.rds")
+# results3 <- readRDS("model3/run_level_3/h3_revision.rds")
+
+s1_traces <- melt(results$search1$traces) %>%
+  rename(L1 = `.L1`) %>%
+  mutate(iteration = as.numeric(iteration))
+
+s2_traces <- melt(results$search2$traces) %>%
+  rename(L1 = `.L1`) %>%
+  # filter(as.numeric(iteration) != 0) %>%
+  mutate(iteration = as.numeric(iteration) + 100)
+
+s3_traces <- melt(results$search3$traces) %>%
+  rename(L1 = `.L1`) %>%
+  # filter(as.numeric(iteration) != 0) %>%
+  mutate(iteration = as.numeric(iteration) + 200)
+
+unit_specific <- c("betaB", "foi_add", "aHur", "hHur")
+
+shared <- c(paste0(c(
+  "mu_B", "XthetaA", "thetaI", "lambdaR", "r", "std_W",
+  "epsilon", "k"
+), "1"), 'loglik')
+
+s1_topn <- order(-results$search1$logLiks$logLik)[1:9]
+s2_topn <- order(-results$search2$logLiks$logLik)[1:6]
+s3_top10 <- order(-results$search3$logLiks$logLik)[1:10]
+
+s2_ancestor <- function(id) {
+  s1_topn[((id - 1) %/% 4) + 1]
+}
+
+s3_ancestor2 <- function(id) {
+  s2_topn[((id - 1) %/% 6) + 1]
+}
+
+s3_ancestor1 <- function(id) {
+  s2_ancestor(s3_ancestor2(id))
+}
+
+s2_traces %>%
+  mutate(ancest_1 = s2_ancestor(L1)) -> s2_traces
+
+s3_traces %>%
+  mutate(ancest_2 = s3_ancestor2(L1),
+         ancest_1 = s3_ancestor1(L1)) -> s3_traces
+
+merged_results <- bind_rows(
+  filter(s1_traces, L1 %in% unique(s3_traces$ancest_1)) %>%
+    mutate(L2 = L1),
+  filter(s2_traces, L1 %in% unique(s3_traces$ancest_2)) %>%
+    mutate(L2 = ancest_1) %>%
[…]

### 2023-07-14 ms-plos.Rnw
syntactical changes + model 3 code update

@@
 # then these results are used.
-RUN_LEVEL <- 3
+RUN_LEVEL <- 1
 rl_dir <- paste0("run_level_", RUN_LEVEL, "/")
@@
     message = FALSE,
+    warning = FALSE,
     tidy.opts = list(
@@
 This case study investigates the use of dynamic models and spatiotemporal data to inform a policy decision in the context of the cholera outbreak in Haiti, which started in 2010.
-We build on a multi-group modeling exercise by \cite{lee20} in which four expert modeling teams developed models to the same dataset with the goal of comparing conclusions on the feasibility of eliminating cholera by a vaccination campaign.
+We build on a multi-group modeling exercise by Lee et al.~\cite{lee20} in which four expert modeling teams developed models to the same dataset with the goal of comparing conclusions on the feasibility of eliminating cholera by a vaccination campaign.
 Model~1 is stochastic and describes cholera at the national level;
@@
 The groups largely adhered to existing guidelines on creating models to inform policy \cite{behrend20,saltelli20} and, despite their autonomy, obtained a consensus that an extensive nationwide vaccination campaign would be necessary to eliminate cholera from Haiti.
-Their conclusion is inconsistent with the prolonged period with no confirmed cholera cases between February, 2019 and September, 2022 \cite{trevisin22}, though cholera has recently reemerged in Haiti \cite{rubin22}.
-
-The discrepancy between the model based conclusions of \cite{lee20} and the prolonged absence of cholera in Haiti has been debated \cite{francois20,rebaudetComment20,henrys20,leeReply20}.
+Their forecasts, however, were inconsistent with the prolonged period with no confirmed cholera cases between February, 2019 and September, 2022 \cite{trevisin22}.
+Though cholera has recently reemerged in Haiti \cite{rubin22}, the inability to accurately forecast cholera incidence from 2019-2022---while diligently adhering to existing scientific standards for infectious disease modeling---demonstrates that existing standards of disease modeling and forecasting may be insufficient.
+
+The discrepancy between the model based conclusions of Lee et al.~\cite{lee20} and the prolonged absence of cholera in Haiti has been debated \cite{francois20,rebaudetComment20,henrys20,leeReply20}.
 Suggested origins of this discrepancy include the use of unrealistic models \cite{rebaudetComment20} and unrealistic criteria for cholera elimination \cite{henrys20}.
@@
 
-While this work is focused on the models proposed by \cite{lee20}, our suggestions have broader relevance.
+While this work is focused on the models proposed by Lee et al.~\cite{lee20}, our suggestions have broader relevance.
 To investigate the extent to which \cite{lee20} is typical of the substantial body of work on Haiti cholera dynamics, we preformed a literature review by searching PubMed with keywords: Haiti, cholera, model.
 The search resulted in 66 papers, of which 32 used dynamic models to describe the cholera epidemic in Haiti.
-The models make various choices on the dichotomies considered by \cite{lee20}; deterministic versus stochastic; compartment model versus agent-based; aggregated versus spatially explicit.
-This is no accident, since \cite{lee20} purposefully designed their study to cover the range of current modeling practice.
-
-We proceed by introducing Models~1--3 in Sec.~\ref{sec:models};
-in Sec.~\ref{sec:methods}, we present a systematic approach to examining and refining these models and then use improved model fits to project cholera incidence in Haiti under various vaccination scenarios.
-This is followed by a consideration of the robustness of model based policy recommendations in Sec.~\ref{sec:advice}, and a discussion in Sec.~\ref{sec:discussion}.
+The models make various choices on the dichotomies considered by Lee et al.~\cite{lee20}; deterministic versus stochastic; compartment model versus agent-based; aggregated versus spatially explicit.
+This is no accident, since Lee et al.~\cite{lee20} purposefully designed their study to cover the range of current modeling practice.
+
+We proceed by introducing the general modeling scheme employed by Models~1--3 and provide details of each individual model;
+we then describe how each model is calibrated to data, and present a systematic approach to examining and refining these models.
+We then use improved model fits to project cholera incidence in Haiti under various vaccination scenarios considered by Lee et al.~\cite{lee20}.
+Finally, we conclude with a discussion of the results.
+% This is followed by a consideration of the robustness of model based policy recommendations in Sec.~\ref{sec:advice}, and a discussion in Sec.~\ref{sec:discussion}.
 
@@
 The danger in using forecasting techniques which rely on associative models to predict the consequence of interventions is called the Lucas critique in an econometric context.
-\cite{lucas76} pointed out that it is naive to predict the effects of an intervention on a given system based entirely on historical associations.
+Lucas et al.~\cite{lucas76} pointed out that it is naive to predict the effects of an intervention on a given system based entirely on historical associations.
 To successfully predict the effect of an intervention, a model should therefore both provide a quantitative explanation of existing data and should have a causal interpretation: a manipulation of the system should correspond quantitatively with the corresponding change to the model.
@@
 
-The four mechanistic models of \cite{lee20} were deliberately developed with limited coordination.
+The four mechanistic models of Lee et al.~\cite{lee20} were deliberately developed with limited coordination.
 This allows us to treat the models as fairly independently developed expert approaches to understanding cholera transmission.
@@
 Here, we have put all three models into a common notational framework.
-Translations back to the original notation of \cite{lee20} are given in Table~S-1 of the supplement.
+Translations back to the original notation of Lee et al.~\cite{lee20} are given in Table~S-1 of the supplement.
 
@@
 Because of the probabilistic nature of both the unobserved latent state and the observable random variables, it is possible to consider various marginal and conditional densities of these two jointly random vectors.
-An important example is the marginal density of the observed random vector $\bm{Y}_{1:N}$, evaluated at the observed data $\bm{y}_{1:N}^*$: $f_{\bm{Y}_{1:N}}\big(\bm{y}_{1:N}^*; \paramVec\big) = \int f_{\bm{X}_{0:N}, \bm{Y}_{1:N}}\big(\bm{x}_{0:N}, \bm{y}_{1:N};\paramVec\big) d\bm{x}_{0:N}$.
-When treated as a function of the parameter vector $\paramVec$, this is called the \emph{likelihood function}, which is the basis of likelihood based statistical inference.
-
-Using the conditional independence of $\bm{Y}_{1:N}$ given $\bm{X}_{0:N}$ and the Markov property of $\bm{X}_{0:N}$, the joint density can be refactored as: $f_{\bm{X}_{0:N}, \bm{Y}_{1:N}}\big(\bm{x}_{0:N}, \bm{y}_{1:N};\paramVec\big) = f_{\bm{X}_0}\big(\bm{x}_0;\paramVec\big)\prod_{n = 1}^N f_{\bm{X}_n|\bm{X}_{n-1}}\big(\bm{x}_{n}|\bm{x}_{n-1}; \paramVec\big)f_{\bm{Y}_n|\bm{X}_{n}}\big(\bm{y}_n|\bm{x}_{n}\big)$.
-This factorization is useful because it demonstrates that POMP models may be completely described using three parts: the \emph{initialization model} for the latent states $f_{\bm{X}_0}\big(\bm{x}_0;\paramVec\big)$; the \emph{one-step transition density}, or \emph{the process model} $f_{\bm{X}_n|\bm{X}_{n-1}}\big(\bm{x}_{n}|\bm{x}_{n-1}; \paramVec\big)$; and the \emph{measurement model} $f_{\bm{Y}_n|\bm{X}_{n}}\big(\bm{y}_n|\bm{x}_{n}\big)$.
+An important example is the marginal density of the observed random vector $\bm{Y}_{1:N}$, evaluated at the observed data $\bm{y}_{1:N}^*$, as shown in Eq.~\myeqref{eq:likedef}:
+
+\begin{eqnarray}
+\label{eq:likedef}
+f_{\bm{Y}_{1:N}}\big(\bm{y}_{1:N}^*; \paramVec\big) = \int f_{\bm{X}_{0:N}, \bm{Y}_{1:N}}\big(\bm{x}_{0:N}, \bm{y}_{1:N};\paramVec\big) d\bm{x}_{0:N}.
+\end{eqnarray}
+
+\noindent When treated as a function of the parameter vector $\paramVec$, this marginal density is called the \emph{likelihood function}, which is the basis of likelihood based statistical inference.
+
+Using the conditional independence of $\bm{Y}_{1:N}$ given $\bm{X}_{0:N}$ and the Markov property of $\bm{X}_{0:N}$, the joint density can be refactored into the useful form given in Eq.~\myeqref{eq:jointLik}:
+
+\begin{eqnarray}
+\label{eq:jointLik}
+f_{\bm{X}_{0:N}, \bm{Y}_{1:N}}\big(\bm{x}_{0:N}, \bm{y}_{1:N};\paramVec\big) = f_{\bm{X}_0}\big(\bm{x}_0;\paramVec\big)\prod_{n = 1}^N f_{\bm{X}_n|\bm{X}_{n-1}}\big(\bm{x}_{n}|\bm{x}_{n-1}; \paramVec\big)f_{\bm{Y}_n|\bm{X}_{n}}\big(\bm{y}_n|\bm{x}_{n}\big).
+\end{eqnarray}
+
+\noindent This factorization is useful because it demonstrates that POMP models may be completely described using three parts: the \emph{initialization model} for the latent states $f_{\bm{X}_0}\big(\bm{x}_0;\paramVec\big)$; the \emph{one-step transition density}, or \emph{the process model} $f_{\bm{X}_n|\bm{X}_{n-1}}\big(\bm{x}_{n}|\bm{x}_{n-1}; \paramVec\big)$; and the \emph{measurement model} $f_{\bm{Y}_n|\bm{X}_{n}}\big(\bm{y}_n|\bm{x}_{n}\big)$.
 In the following subsections, we describe Models 1--3 in terms of these three components.
@@
 The process noise $d\Gamma(t)/dt$ is multiplicative Gamma-distributed white noise, with infinitesimal variance parameter $\sigmaProc^2$.
-\cite{lee20} included process noise in Model~3 but not in Model~1, i.e., they fixed $\sigmaProc^2 = 0$.
+Lee et al.~\cite{lee20} included process noise in Model~3 but not in Model~1, i.e., they fixed $\sigmaProc^2 = 0$.
 Gamma white noise in the transmission rate gives rise to an over-dispersed latent Markov process \cite{breto11} which has been found to improve the statistical fit of disease transmission models \cite{stocks20,he10}.
@@
 For any time point in $t_{1:N}$, the process model describes the movement of individuals from one latent state compartment to another.
-Per-capita transition rates are given in Eqs.~\ref{model1:SE}-\ref{model1:birth}:
+Per-capita transition rates are given in Eqs.~\myeqref{model1:SE}-\myeqref{model1:birth}:
 \begin{eqnarray}
@@
 \mu_{S_{\vaccCounter}E_{\vaccCounter}} &=& \lambda(t),
-\\[-3pt]
+\\%[-3pt]
 \label{model1:EI}
 \mu_{E_{\vaccCounter}I_{\vaccCounter}} &=& \muEI\big(1-\vaccineEfficacy_\vaccCounter(t)\big),
-\\[-3pt]
+\\%[-3pt]
 \label{model1:EA}
 \mu_{E_{\vaccCounter}A_{\vaccCounter}} &=& \muEI\, \vaccineEfficacy_\vaccCounter(t),
-\\[-3pt]
+\\%[-3pt]
 \label{model1:toR}
 \mu_{I_{\vaccCounter}R_{\vaccCounter}} &=& \mu_{A_{\vaccCounter}R_{\vaccCounter}} = \muIR,
-\\[-3pt]
+\\%[-3pt]
 \label{model1:RS}
 \mu_{R_{\vaccCounter}S_{\vaccCounter}} &=& \muRS,
-\\[-3pt]
+\\%[-3pt]
 \label{model1:vacc}
 \mu_{S_0S_{\vaccCounter}} &=& \mu_{E_0E_{\vaccCounter}} = \mu_{I_0I_{\vaccCounter}} = \mu_{A_0A_{\vaccCounter}} = \mu_{R_0R_{\vaccCounter}} = \eta_{\vaccCounter}(t),
-\\[-3pt]
+\\%[-3pt]
 \label{model1:death}
@@
 Conditional on infection status, vaccinated individuals are also less infectious than their non-vaccinated counterparts by a rate of $\asymptomRelativeInfect = 0.05$ in Eq.~\eqref{model1:lambda}.
-In \eqref{model1:EA} and~\eqref{model1:EI} the asymptomatic ratio for non-vaccinated individuals is set $\symptomFrac_0(t)=0$, so that the asymptomatic route is reserved for vaccinated individuals.
+In Eqs.~\myeqref{model1:EA} and~\myeqref{model1:EI} the asymptomatic ratio for non-vaccinated individuals is set $\symptomFrac_0(t)=0$, so that the asymptomatic route is reserved for vaccinated individuals.
 For $\vaccCounter\in\seq{1}{\vaccClass}$, the vaccination cohort $\vaccCounter$ is assigned a time $\tau_{\vaccCounter}$, and we take $\vaccineEfficacy_\vaccCounter(t) = c \, \vaccineEfficacy^*(t-\tau_{\vaccCounter})$
@@
 The cosine term in Eq.~\myeqref{model2:lambda} accounts for annual seasonality, with a phase parameter $\phaseParm$.
-The \cite{lee20} implementation of Model~2 fixes $\phaseParm = 0$.
+The Lee et al.~\cite{lee20} implementation of Model~2 fixes $\phaseParm = 0$.
 
@@
 \mu_{S_{u\vaccCounter}E_{u\vaccCounter}} &=& \vaccineEfficacy_\vaccCounter \, \lambda,
-\\[-3pt]
+\\%[-3pt]
 \label{model2:mu_EI}
 \mu_{E_{u\vaccCounter}I_{u\vaccCounter}} &=& \symptomFrac\muEI, \quad \mu_{E_{u\vaccCounter}A_{u\vaccCounter}} = (1-\symptomFrac)\muEI,
-\\[-3pt]
+\\%[-3pt]
 \label{model2:mu_IR}
 \mu_{I_{u\vaccCounter}R_{u\vaccCounter}} &=& \mu_{A_{u\vaccCounter}R^A_{u\vaccCounter}} = \muIR,
-\\[-3pt]
+\\%[-3pt]
 \label{model2:RS}
 \mu_{R_{u\vaccCounter}S_{u\vaccCounter}} &=& \mu_{R^A_{u\vaccCounter}S_{u\vaccCounter}} = \muRS,
-\\[-3pt]
+\\%[-3pt]
 \label{model2:transport}
 \mu_{S_{u\vaccCounter}S_{{\varv}\vaccCounter}} &=& \mu_{E_{u\vaccCounter}E_{{\varv}\vaccCounter}} = \mu_{I_{u\vaccCounter} I_{{\varv}\vaccCounter}} = \mu_{A_{u\vaccCounter}A_{{\varv}\vaccCounter}} = \mu_{R_{u\vaccCounter}R_{{\varv}\vaccCounter}} = \mu_{R^A_{u\vaccCounter} R^A_{{\varv}\vaccCounter}} = T_{u\varv},
-\\[-3pt]
+\\%[-3pt]
 \label{model2:omega1}
 \mu_{S_{u1}S_{u0}} &=& \mu_{S_{u3}S_{u0}} = \omega_1,
-\\[-3pt]
+\\%[-3pt]
 \label{model2:omega2}
 \mu_{S_{u2}S_{u0}} &=& \mu_{S_{u4}S_{u0}} = \omega_2,
-\\[-3pt]
+\\%[-3pt]
 \label{model2:to_W}
 \mu_{\demography W_u} &=& \Wshed \left\{ \sum_{\vaccCounter=0}^4 I_{u\vaccCounter}(t) + \asymptomRelativeShed \sum_{\vaccCounter=0}^4 A_{u\vaccCounter}(t) \right\},
-\\[-3pt]
+\\%[-3pt]
 \label{model2:from_W}
 \mu_{W_u\demography} &=& \Wremoval,
-\\[-3pt]
+\\%[-3pt]
 \label{model2:water_transport}
@@
 \end{eqnarray}
-In \eqref{model2:transport} the spatial coupling is specified by a gravity model,
+In Eq.~\myeqref{model2:transport} the spatial coupling is specified by a gravity model,
 \begin{equation}
@@
 $D_{u\varv}$ is a distance measure estimating average road distance between randomly chosen members of each population, and $v_{\mathrm{rate}}= 10^{-12} \,\mbox{km$^2$yr$^{-1}$}$  was treated as a fixed constant.
-In \eqref{model2:water_transport}, $T^W_{u\varv}$ is a measure of river flow between departments.
-The unit of $W_u(t)$ is cells per ml, with dose response modeled via a saturation constant of $\Wsat$ in \eqref{model2:lambda}.
-In \eqref{model2:mu_SE}, $\vaccineEfficacy_\vaccCounter$ denotes the vaccine efficacy for each vaccination campaign $\vaccCounter \in \vaccClass$, with $\vaccineEfficacy_0 = 1$, $\vaccineEfficacy_1 = 1 - 0.429 \childReduce$, $\vaccineEfficacy_2 = 1 - 0.519 \childReduce$, $\vaccineEfficacy_3 = 1 - 0.429$, and $\vaccineEfficacy_4 = 1 - 0.519$
-Here, $\childReduce = 0.4688$ represents the reduced efficacy of the vaccination for children under the age of five years, and the values $0.429$ and $0.519$ are the median effectiveness of one and two doses over their effective period respectively, according to Table S4 in the supplement material of \cite{lee20}.
+In Eq.~\myeqref{model2:water_transport}, $T^W_{u\varv}$ is a measure of river flow between departments.
+The unit of $W_u(t)$ is cells per ml, with dose response modeled via a saturation constant of $\Wsat$ in Eq.~\myeqref{model2:lambda}.
+In Eq.~\myeqref{model2:mu_SE}, $\vaccineEfficacy_\vaccCounter$ denotes the vaccine efficacy for each vaccination campaign $\vaccCounter \in \vaccClass$, with $\vaccineEfficacy_0 = 1$, $\vaccineEfficacy_1 = 1 - 0.429 \childReduce$, $\vaccineEfficacy_2 = 1 - 0.519 \childReduce$, $\vaccineEfficacy_3 = 1 - 0.429$, and $\vaccineEfficacy_4 = 1 - 0.519$
+Here, $\childReduce = 0.4688$ represents the reduced efficacy of the vaccination for children under the age of five years, and the values $0.429$ and $0.519$ are the median effectiveness of one and two doses over their effective period respectively, according to Table S4 in the supplement material of Lee et al.~\cite{lee20}.
 Because vaccine efficacy remains constant, individuals in this model transition from a vaccinated compartment to the susceptible compartment at the end of the vaccine coverage period.
@@
 Reported cases are assumed to come from a log-normal distribution, with the log-scale mean equal to the reporting rate $\reportRate$ times the number of newly infected individuals.
-See Sec.~\ref{sec:fit2} and Sec.~S3.2 of the supplement for more details.
+See the section describing how Model~2 is fit to data and Sec.~S3.2 of the supplement for more details.
 
@@
 
-Per-capita transition rates are given in Equations~\ref{eq:model3:SI}--\ref{eq:model3:Decay}.
+Per-capita transition rates are given in Eqs.~\myeqref{eq:model3:SI}--\myeqref{eq:model3:Decay}.
 \begin{eqnarray}
@@
 \mu_{S_{u\vaccCounter}I_{u}} &=& \symptomFrac \,  \lambda_u \big(1-\vaccineEfficacy_{u\vaccCounter}(t)\big) \, d\Gamma/dt,
-\\[-3pt]
+\\%[-3pt]
 \label{eq:model3:SA}
 \mu_{S_{u\vaccCounter}A_{u}} &=& (1-\symptomFrac) \,  \lambda_u \big(1-\vaccineEfficacy_{u\vaccCounter}(t)\big) \,  d\Gamma/dt,
-\\[-3pt]
+\\%[-3pt]
 \label{eq:model3:IR}
 \mu_{I_{u}R_{u\vaccCounter 1}} &=& \mu_{A_{u}R_{u\vaccCounter 1}} = \muIR,
-\\[-3pt]
+\\%[-3pt]
 \label{eq:model3:IS}
 \mu_{I_{u}S_{u0}} &=& \muDeath + \choleraDeath, \hspace{2mm} \mu_{A_{u}S_{u0}} = \muDeath
-\\[-3pt]
+\\%[-3pt]
 \label{eq:model3:RRnext}
 \mu_{R_{u\vaccCounter 1}R_{u\vaccCounter 2}} &=& \mu_{R_{u\vaccCounter 2}R_{u\vaccCounter 3}} = 3\muRS,
-\\[-3pt]
+\\%[-3pt]
 \label{eq:model3:RS}
 \mu_{R_{u\vaccCounter k}S_{u0}} &=& \muDeath + 3\muRS \, \mathbf{1}_{\{k=3\}},
-\\[-3pt]
+\\%[-3pt]
 \label{eq:model3:water}
 \mu_{{\demography}W_u} &=& \big[1 + \seasAmplitude \big(J_u(t))^r \big] \mathrm{Den}_u \, \Wshed \big[ I_{u}(t)+ \asymptomRelativeShed A_{u}(t) \big],
-\\[-3pt]
+\\%[-3pt]
 \label{eq:model3:Decay}
@@
 \end{eqnarray}
-As with Model~1, $d\Gamma_u(t)/dt$ is multiplicative Gamma-distributed white noise in \myeqref{eq:model3:SI} and \myeqref{eq:model3:SA}.
-In \eqref{eq:model3:water}, $J_u(t)$ is a dimensionless measurement of precipitation that has been standardized by dividing the observed rainfall at time $t$ by the maximum recorded rainfall in department $u$ during the epidemic, and $\mathrm{Den}_u$ is the population density.
-Demographic stochasticity is accounted for by modeling non-cholera related death rate $\muDeath$ in each compartment, along with an additional death rate $\choleraDeath$ in \myeqref{eq:model3:IS} to account for cholera induced deaths among infected individuals.
-All deaths are balanced by births into the susceptible compartment in \myeqref{eq:model3:IS} and \myeqref{eq:model3:RS}, thereby maintaining constant population in each department.
+As with Model~1, $d\Gamma_u(t)/dt$ is multiplicative Gamma-distributed white noise in Eqs.~\myeqref{eq:model3:SI} and \myeqref{eq:model3:SA}.
+In Eq.~\myeqref{eq:model3:water}, $J_u(t)$ is a dimensionless measurement of precipitation that has been standardized by dividing the observed rainfall at time $t$ by the maximum recorded rainfall in department $u$ during the epidemic, and $\mathrm{Den}_u$ is the population density.
+Demographic stochasticity is accounted for by modeling non-cholera related death rate $\muDeath$ in each compartment, along with an additional death rate $\choleraDeath$ in Eq.~\myeqref{eq:model3:IS} to account for cholera induced deaths among infected individuals.
+All deaths are balanced by births into the susceptible compartment in Eqs.~\myeqref{eq:model3:IS} and \myeqref{eq:model3:RS}, thereby maintaining constant population in each department.
 
 Similar to Model~1, there are no distinct compartments for individuals under five years of age, and the vaccination efficacy is taken as a age adjusted weighted average of the efficacy for individuals both over and under five years of age:  $\vaccineEfficacy_{u \vaccCounter}(t) = c\vaccineEfficacy^*(t - \tau_{uz})$, where $\tau_{uz}$ is the vaccination time for unit $u$ and vaccination campaign $z$.
-The value $c$ and the function $\vaccineEfficacy^*$ are equivalent to those described in Sec.~\ref{sec:model1}.
+The value $c$ and the function $\vaccineEfficacy^*$ are equivalent to those described in the Model~1 description.
 
@@
 # This code chunk is used to fit Model 3 using the IBPF algorithm.
-set.seed(1859687)
-
-# Create vectors for the unit and shared parameters that are estimated
+set.seed(8636712)
+
+# Create vectors for the unit and shared parameters
 unit_specific_names <- c("betaB", "foi_add", "aHur", "hHur", "Iinit")
 shared_param_names <- c(
-  "mu_B", "XthetaA", "thetaI", "lambdaR", "r", "std_W", "epsilon", "k"
-)
-
-est_param_names <- c(unit_specific_names, shared_param_names)
+  "mu_B", "XthetaA", "thetaI", "lambdaR", "r", "std_W",
+  "epsilon", "k"
+)
+est_param_names <- c(
+  unit_specific_names, shared_param_names
+)
 
@@
     return(expression(ifelse(time >= 2016.754 & time <= 2017, 0.015, 0)))
-  } else if (x %in% c("Iinit3", "Iinit7", "Iinit10")) {
+  } else if (x %in% c("Iinit3", "Iinit4")) {
     return(expression(ivp(0.25)))
@@
 
-# Simple function that is used to set the rw_sd for the second search
-# This function is convenient so that we don't have to type out the rw_sd
-# for each unit parameter.
 set_rw_2 <- function(x) {
@@
     return(expression(ifelse(time >= 2016.754 & time <= 2017, 0.005, 0)))
-  } else if (x %in% c("Iinit3", "Iinit7", "Iinit10")) {
+  } else if (x %in% c("Iinit3", "Iinit4")) {
     return(expression(ivp(0.125)))
@@
 
-# Simple function that is used to set the rw_sd for the third search
-# This function is convenient so that we don't have to type out the rw_sd
-# for each unit parameter.
-set_rw_3 <- function(x) {
-  if (gsub("[[:digit:]]+$", "", x) %in% shared_param_names) {
-    return(0.0025)
-  } else if (x %in% paste0(rep(c("aHur", 'hHur'), each = 2), c(3, 9))) {
-    return(expression(ifelse(time >= 2016.754 & time <= 2017, 0.005, 0)))
-  } else if (x %in% c("Iinit3", "Iinit7", "Iinit10")) {
-    return(expression(ivp(0.125)))
-  } else if (grepl('^foi_add[[:digit:]]+$', x)) {
-    return(0.003)
-  } else if (grepl('^betaB[[:digit:]]+$', x)) {
-    return(0.0025)
-  } else {
-    return(0)
-  }
-}
-
 # First Search RW
@@
 
-# Third Search RW
-reg_rw_3.sd <- lapply(est_param_names_expanded, set_rw_3)
-names(reg_rw_3.sd) <- est_param_names_expanded
-chol_rw_3.sd <- do.call(rw_sd, reg_rw_3.sd)
+if (RUN_LEVEL == 1){
+  SEARCH1 = list(
+    NBPF = 10,
+    NP = 200,
+    SPAT_REGRESSION = 0.05,
+    NREPS = 20,
+    NP_EVAL = 200,
+    NREPS_EVAL = 3,
+    RW_SD = chol_rw_1.sd,
+    COOLING = 0.5,
+    KEEP_TRACES = TRUE
+  )
+
+  SEARCH2 = list(
+    TOP_N = 8,
+    NBPF = 10,
+    NP = 200,
+    SPAT_REGRESSION = 0.05,
+    NREPS = 2,
+    NP_EVAL = 200,
+    NREPS_EVAL = 3,
+    RW_SD = chol_rw_2.sd,
+    COOLING = 0.5,
+    KEEP_TRACES = TRUE
+  )
+
+} else if (RUN_LEVEL == 2) {
+  SEARCH1 = list(
+    NBPF = 100,
+    NP = 1000,
+    SPAT_REGRESSION = 0.05,
+    NREPS = 72,
+    NP_EVAL = 1000,
+    NREPS_EVAL = 8,
+    RW_SD = chol_rw_1.sd,
+    COOLING = 0.5,
+    KEEP_TRACES = TRUE,
+    KEEP_LIK_MAT = TRUE
+  )
+
+  SEARCH2 = list(
+    TOP_N = 9,
+    NBPF = 100,
+    NP = 1000,
+    SPAT_REGRESSION = 0.05,
+    NREPS = 4,
+    NP_EVAL = 1000,
+    NREPS_EVAL = 9,
+    RW_SD = chol_rw_2.sd,
+    COOLING = 0.5,
+    KEEP_TRACES = TRUE,
+    KEEP_LIK_MAT = TRUE
+  )
+
+} else if (RUN_LEVEL == 3) {
+
+  SEARCH1 = list(
+    NBPF = 100,
+    NP = 2500,
+    SPAT_REGRESSION = 0.05,
+    NREPS = 144,
+    NP_EVAL = 2500,
+    NREPS_EVAL = 18,
+    RW_SD = chol_rw_1.sd,
+    COOLING = 0.5,
+    KEEP_TRACES = FALSE,
+    KEEP_LIKE_MAT = FALSE
+  )
+
+  SEARCH2 = list(
+    TOP_N = 12,
+    NBPF = 100,
+    NP = 2500,
+    SPAT_REGRESSION = 0.05,
+    NREPS = 6,
+    NP_EVAL = 2500,
+    NREPS_EVAL = 18,
+    RW_SD = chol_rw_2.sd,
+    COOLING = 0.5,
+    KEEP_TRACES = FALSE,
+    KEEP_LIK_MAT = FALSE
+  )
+
+  # SEARCH3 = list(  # Not used, but was used for some preliminary results.
+  #   TOP_N = 9,
+  #   NBPF = 100,
+  #   NP = 2000,
+  #   SPAT_REGRESSION = 0.05,
+  #   NREPS = 8,
+  #   NP_EVAL = 2000,
+  #   NREPS_EVAL = 18,
+  #   RW_SD = rw_3.sd,
+  #   COOLING = 0.5,
+  #   KEEP_TRACES = FALSE,
+  #   KEEP_LIK_MAT = TRUE
+  # )
+}
 
@@
 # number of replications for each search.
-h3_fit_parms <- set_h3_fit_parms(
-  RUN_LEVEL = RUN_LEVEL,
-  rw_1.sd = chol_rw_1.sd,
-  rw_2.sd = chol_rw_2.sd,
-  rw_3.sd = chol_rw_3.sd
-)
+# h3_fit_parms <- set_h3_fit_parms(
+#   RUN_LEVEL = RUN_LEVEL,
+#   rw_1.sd = chol_rw_1.sd,
+#   rw_2.sd = chol_rw_2.sd
+# )
 
@@
 
-haiti3_fit <- readRDS("model3/run_level_3/h3_revision.rds")
+# haiti3_fit <- readRDS("model3/run_level_3/h3_revision.rds")
 
 # Fit Model 3
-# haiti3_fit <- bake(
-#   file = paste0("model3/", rl_dir, "haiti3_coupled_fit.rds"), {
-#     fit_coupled_haiti3(
-#       nsearches = h3_fit_parms$N_SEARCHES,
-#       search1 = h3_fit_parms$SEARCH1,
-#       search2 = h3_fit_parms$SEARCH2,
-#       search3 = h3_fit_parms$SEARCH3,
-#       search_rho = FALSE,
-#       search_gamma = FALSE,
-#       search_Iinit = TRUE,
-#       search_hur = TRUE,
-#       ncores = cores
-#     )
-#   }
-# )
+haiti3_fit <- bake(
+  file = paste0("model3/", rl_dir, "haiti3_coupled_fit.rds"), {
+    fit_haiti3(
+      nsearches = 2L,
+      search1 = SEARCH1,
+      search2 = SEARCH2,
+      search_rho = FALSE,
+      search_gamma = FALSE,
+      search_Iinit = TRUE,
+      search_hur = TRUE,
+      ncores = cores
+    )
+  }
+)
 
@@
 As such, it is desirable to obtain the best possible fit of the proposed mechanistic structure to the observed data.
-In the following subsections we describe our approach to calibrating the three proposed mechanistic models to observed cholera incidence data.
+
+In this article we calibrate the parameters of each of our models by maximizing the likelihood, as described in Eq.~\myeqref{eq:likedef}.
+In the following subsections we describe in detail our approach to calibrating the three proposed mechanistic models to observed cholera incidence data.
 
@@
 One of the most frequently used approaches to calibrate the parameters of a statistical model is maximum likelihood estimation.
-In the case of a POMP model, the likelihood function---as described in Sec.~\ref{sec:models}---is written as $\mathcal{L}(\paramVec) = \int f_{\bm{X}_{0:N}, \bm{Y}_{1:N}}\big(\bm{x}_{0:N}, \bm{y}_{1:N};\paramVec\big) d\bm{x}_{0:N}$.
+In the case of a POMP model, the likelihood function---as described in Eq.~\myeqref{eq:likedef}---is written as $\mathcal{L}(\paramVec) = \int f_{\bm{X}_{0:N}, \bm{Y}_{1:N}}\big(\bm{x}_{0:N}, \bm{y}_{1:N};\paramVec\big) d\bm{x}_{0:N}$.
 The high-dimensional integral in the likelihood, and the potentially non-linear nature of the process model, makes it impossible to obtain a closed form expression of the likelihood function of POMP models, except in the most basic cases.
@@
 
-In order to retain the ability to propose models that are scientifically meaningful rather than only those that are simply statistically convenient, we restrict ourselves to parameter estimation techniques that have the plug-and-play property, which is that the fitting procedure only requires the ability to simulate the latent process instead of evaluating transition densities \cite{breto09,he10}.
+In order to retain the ability to propose models that are scientifically meaningful rather than only those that are simply statistically convenient, we restrict ourselves to parameter estimation techniques that have the plug-and-play property, which is that the fitting procedure only requires the ability to simulate the latent process instead of evaluating transition densities \cite{breto09,he10};
+in the context of the notation and definitions employed in this article, this means that we only require the ability to simulate from $f_{\bm{X}_0}\big(\bm{x}_0;\paramVec\big)$ and $f_{\bm{X}_n|\bm{X}_{n-1}}\big(\bm{x}_{n}|\bm{x}_{n-1}; \paramVec\big)$ rather than needing to evaluate these densities.
 Plug-and-play algorithms include Bayesian approaches like ABC and PMCMC \cite{toni09,andrieu10}, but here we use algorithms that enable maximum likelihood estimation.
-To our knowledge, the only plug-and-play methods that can maximize the likelihood for POMP models of this complexity are iterated filtering algorithms \cite{ionides15}, which modify the well-known \emph{particle filter} \cite{arulampalam02}.
+To our knowledge, the only plug-and-play methods that can be used to maximize the likelihood for POMP models of this complexity are iterated filtering algorithms \cite{ionides15}, which modify the well-known \emph{particle filter} \cite{arulampalam02}.
 The particle filter is often used to obtain values of latent states, but can also be used to obtain likelihood estimates of a POMP model.
@@
 To do this, we perform a profile-likelihood search on the parameter $\transmissionTrend$ and obtain a $95\%$ confidence interval via a Monte Carlo Adjusted Profile (MCAP) \cite{ionides17}.
-\cite{lee20} implemented Model~1 by fitting two distinct phases: an epidemic phase from October 2010 through March 2015, and an endemic phase from March 2015 onward.
+Lee et al.~\cite{lee20} implemented Model~1 by fitting two distinct phases: an epidemic phase from October 2010 through March 2015, and an endemic phase from March 2015 onward.
 We similarly allow the re-estimation of process and measurement overdispersion parameters ($\sigmaProc^2$ and $\obsOverdispersion$), and require that the latent Markov process $X(t)$ carry over from one phase into the next.
@@
 
-We implemented Model~1 using the \code{pomp} package \cite{king16}, relying heavily on the \code{pomp} source code provided by \cite{lee20}.
+We implemented Model~1 using the \code{pomp} package \cite{king16}, relying heavily on the \code{pomp} source code provided by Lee et al.~\cite{lee20}.
 Both analyses used the \code{mif2} implementation of the IF2 algorithm to estimate $\paramVec$ by maximum likelihood.
@@
 Many Monte Carlo replications are then required to quantify and further reduce the error.
-The large difference in likelihoods that were obtain by these two approaches (see Table~\ref{tab:likes}) highlights the importance of carefully determining the necessary computational effort needed to maximize model likelihoods and acting accordingly.
+The large difference in likelihoods that were obtain by these two approaches (see Table~\ref{tab:likes}) can primarily be attributed to increasing the computational effort in fitting the model; this result highlights the importance of carefully determining the necessary computational effort needed to maximize model likelihoods and acting accordingly.
 
@@
 Because the process model of Model~2 is deterministic, maximum likelihood estimation reduces to a least squares calculation when combined with a Gaussian measurement model (see Sec.~S3.2 of the supplement material for details).
-\cite{lee20} fit two versions of Model~2 based on a presupposed change in cholera transmission from a epidemic phase to endemic phase that occurred in March, 2014.
+Lee et al.~\cite{lee20} fit two versions of Model~2 based on a presupposed change in cholera transmission from a epidemic phase to endemic phase that occurred in March, 2014.
 The inclusion of a change-point in model states and parameters increases the flexibility of the model and hence the ability to fit the observed data.
 The increase in model flexibility, however, results in hidden states that are inconsistent between model phases.
-The inclusion of a model break-point by \cite{lee20} is perhaps due to a challenging feature of fitting a deterministic model via least squares: discrepancies between model trajectories and observed case counts in highly infectious periods of a disease outbreak will result in greater penalty than the discrepancies between model trajectories and observed case counts in times of relatively low infectiousness, resulting in a bias towards accurately describing periods of high infectiousness.
+The inclusion of a model break-point by Lee et al.~\cite{lee20} is perhaps due to a challenging feature of fitting a deterministic model via least squares: discrepancies between model trajectories and observed case counts in highly infectious periods of a disease outbreak will result in greater penalty than the discrepancies between model trajectories and observed case counts in times of relatively low infectiousness, resulting in a bias towards accurately describing periods of high infectiousness.
 This issue is particularly troublesome for modeling cholera dynamics in Haiti: the inability to accurately fit times of low infectiousness may result in poor model forecasts, as few cases of cholera were observed in the last few years of the epidemic.
@@
 This permits the consideration of demographic stochasticity, which is dominant for small infected populations, together with log scale stochasticity (also called multiplicative, or environmental, or extra-demographic) which is dominant at high population counts.
-Here we chose to fit the model to transformed case counts rather than adding overdispersion to the measurement model with the goal of minimizing the changes to the model proposed by \cite{lee20}.
+Here we chose to fit the model to transformed case counts rather than adding overdispersion to the measurement model with the goal of minimizing the changes to the model proposed by Lee et al.~\cite{lee20}.
 
@@
 The model was then fit using the subplex algorithm, implemented in the  \code{subplex} package \cite{king2020Subplex}.
-A comparison of the trajectory of the fitted model to the data is given in Fig~\ref{fig:mod2Traj}.
+A comparison of the trajectory of the fitted model to the data is given in Fig.~\ref{fig:mod2Traj}.
 
@@
 Evidence for the former approach has been provided in previous studies \cite{king15}, including the specific case of heterogeneity between Haitian departments in cholera transmission \cite{collins14}.
-However, a legitimate preference for simplicity \cite{saltelli20,green15} can support a decision to consider nationally aggregated models.
+However, a legitimate preference for simplicity can support a decision to consider nationally aggregated models \cite{saltelli20,green15}.
 One issue that arises when fitting spatially explicit models is that parameter estimation techniques based on the particle filter become computationally intractable as the number of spatial units increases.
@@
 
-To avoid the approximation error present in high-dimensional models, \cite{lee20} simplified the problem of estimating the parameters of Model~3 by creating an approximate version of the model where the units are independent given the observed data.
+To avoid the approximation error present in high-dimensional models, Lee et al.~\cite{lee20} simplified the problem of estimating the parameters of Model~3 by creating an approximate version of the model where the units are independent given the observed data.
 Reducing a spatially coupled model to individual units in this fashion requires special treatment of any interactive mechanisms between spatial units, such as found in Eq.~\myeqref{eq:model3:foi}.
-Because the simplified, spatially-decoupled version of Model~3 implemented by \cite{lee20} relies on the observed cholera cases, the calibrated model cannot readily be used to obtain forecasts.
-Therefore, in order to obtain model forecasts, \cite{lee20} used the parameters estimates from the spatially-decoupled approximation of Model~3 to obtain forecasts using the fully coupled version of the model.
+Because the simplified, spatially-decoupled version of Model~3 implemented in \cite{lee20} relies on the observed cholera cases, the calibrated model cannot readily be used to obtain forecasts.
+Therefore, in order to obtain model forecasts, Lee et al.~\cite{lee20} used the parameters estimates from the spatially-decoupled approximation of Model~3 to obtain forecasts using the fully coupled version of the model.
 This approach of model calibration and forecasting avoids the issue of particle depletion, but may also be problematic.
@@
 
-For our analysis, we calibrate the parameters of the spatially coupled version of Model~3 using the iterated block particle filter (IBPF) algorithm of \cite{ionides22}.
-This algorithm extends the work of \cite{ning21ibpf}, who provided theoretic justification for the version of the algorithm that only estimates unit-specific parameters.
+For our analysis, we calibrate the parameters of the spatially coupled version of Model~3 using the iterated block particle filter (IBPF) algorithm \cite{ionides22}.
+This algorithm extends the work of Ning and Ionides \cite{ning21ibpf}, who provided theoretic justification for the version of the algorithm that only estimates unit-specific parameters.
 The IBPF algorithm enables us to directly estimate the parameters of models describing high-dimensional partially-observed nonlinear dynamic systems via likelihood maximization.
-The ability to directly estimate parameters of Model~3 is responsible for the large increase in model likelihoods reported in Table~\ref{tab:likes}, as no algorithm with similar capabilities existed when \cite{lee20} published their results.
+The ability to directly estimate parameters of Model~3 is responsible for the large increase in model likelihoods reported in Table~\ref{tab:likes}, as no algorithm with similar capabilities existed when Lee et al.~\cite{lee20} published their results.
 Simulations from the fitted model are displayed in Fig.~\ref{fig:h3spatsims}.
@@
 
-Parameter calibration (whether Bayesian or frequentist) aims to find the best description of the observed data under the assumptions of the model.
+The goal of parameter calibration---whether done using Bayesian or frequentist methods---is to find the best description of the observed data under the assumptions of the model.
 Obtaining the best fitting set of parameters for a given model does not, however, guarantee that the model provides an accurate representation of the system in question.
 Model misspecification, which may be thought of as the omission of a mechanism in the model that is an important feature of the dynamic system, is inevitable at all levels of model complexity.
-To make progress, while accepting proper limitations, one must bear in mind the much-quoted observation of \cite{box79} that ``all models are wrong but some are useful.''
-Beyond being good practical advice for applied statistics, this assertion is relevant for the philosophical justification of statistical inference as severe testing \cite{mayo18} % \citep[][Sec.~4.8]{mayo18}.
+To make progress, while accepting proper limitations, one must bear in mind the much-quoted observation of George Box \cite{box79} that ``all models are wrong but some are useful.''
+Beyond being good practical advice for applied statistics, this assertion is relevant for the philosophical justification of statistical inference as severe testing \cite{mayo18}.
+% \citep[][Sec.~4.8]{mayo18}.
 In this section, we discuss some tools for diagnosing mechanistic models with the goal of making the subjective assessment of model ``usefulness'' more objective.
@@
 While visual comparisons can be informative, they provide only a weak and informal measure of the goodness-of-fit of a model.
-The study by \cite{lee20} provides an example of this: their models and parameter estimates resulted in simulations that visually resembled the observed data, yet resulted in model likelihoods that were---in some cases---considerably smaller than likelihoods that can be achieved via the likelihood based optimization techniques that were used (see Table~\ref{tab:likes}).
+The study by Lee et al.~\cite{lee20} provides an example of this: their models and parameter estimates resulted in simulations that visually resembled the observed data, yet resulted in model likelihoods that were---in some cases---considerably smaller than likelihoods that can be achieved via the likelihood based optimization techniques that were used (see Table~\ref{tab:likes}).
 Alternative forms of model validation should therefore be used in conjunction with visual comparisons of simulations to observed data.
@@
 Examples include linear regression, auto-regressive moving average time series models, or even independent and identically distributed measurements.
-The benchmarks enable us to evaluate the goodness of fit that can be expected of a suitable mechanistic model.
+Benchmarks enable us to evaluate the goodness of fit that can be expected of a suitable mechanistic model.
 
@@
 In addition, the published goodness of fit provides a concrete point of comparison for subsequent scientific investigations.
-When combined with online availability of data and code, objective measures of fit provide a powerful tool to accelerate scientific progress, following the paradigm of the {\it common task framework} \cite{donoho17} % \citep[][Sec.~6]{donoho17}.
+When combined with online availability of data and code, objective measures of fit provide a powerful tool to accelerate scientific progress, following the paradigm of the {\it common task framework} \cite{donoho17}.
+% \citep[][Sec.~6]{donoho17}.
 
@@
 In the context of our case study, we demonstrate how the fit of Model~1 corroborates other evidence concerning the role of rainfall in cholera epidemics.
-Specifically, we examine the results of fitting the flexible cubic spline term in Model~1 (Eq.~\myeqref{model1:lambda}--\myeqref{model1:beta}).
+Specifically, we examine the results of fitting the flexible cubic spline term in Model~1 (Eqs.~\myeqref{model1:lambda}--\myeqref{model1:beta}).
 The cubic splines permit flexible estimation of seasonality in the force of infection, $\transmission(t)$.
@@
 This is consistent with previous studies finding that rainfall played an important role in cholera transmission in Haiti \cite{lemaitre19,eisenberg13}.
-The estimated seasonality also features an increased transmission rate during the fall, which was noticed at an earlier stage of the epidemic \cite{rinaldo12}; the high transimission rate in the fall may also be a result of the increase transission that occured in the fall of 2016, when hurricane Matthew struck Haiti.
+The estimated seasonality also features an increased transmission rate during the fall, which was noticed at an earlier stage of the epidemic \cite{rinaldo12}; the high transmission rate in the fall may also be a result of the increase transmission that occurred in the fall of 2016, when hurricane Matthew struck Haiti \cite{ferreirai16}.
 
@@
 
+\subsection*{Robust interpretation of model based conclusions}\label{sec:advice}
+
+A model which aspires to provide quantitative guidance for assessing interventions should provide a quantitative statistical fit for available data.
+However, strong statistical fit does not guarantee a correct causal structure: it does not even necessarily require the model to assert a causal explanation.
+A causal interpretation is strengthened by corroborative evidence.
+For example, reconstructed latent variables (such as numbers of susceptible and recovered individuals) should make sense in the context of alternative measurements of these variables \cite{grad12}.
+In the context of our case study, by reconstructing the latent states of Model~3, we notice that the calibrated model favors higher levels of cholera transmission than what was typically observed in the incidence data (Sec.~S5.1 of the supplement).
+This result hints at the possibility of model mispecification, and should prompt us to be sceptical of the reliability of forecasts from this model.
+Similarly, parameters that have been calibrated to data should make sense in the context of alternative lines of evidence about the phenomena being modeled, while making allowance for the possibility that the interpretations of parameters may vary when modeling across differing spatial scales.
+
+If a mechanistic model including a feature (such as a representation of a mechanism, or the inclusion of a covariate) fits better than mechanistic models without that feature, and also has competitive fit compared to associative benchmarks, this may be taken as evidence supporting the scientific relevance of the feature.
+As for any analysis of observational data, we must be alert to the possibility of confounding.
+For a covariate, this shows up in a similar way to regression analysis: the covariate under investigation could be a proxy for some other unmodeled or unmeasured covariate.
+For a mechanism, the model feature could in principle explain the data by helping to account for some different unmodeled phenomenon.
+In the context of our analysis, the estimated trend in transmission rate could be explained by any trending variable (such as hygiene improvements, or changes in population behavior), resulting in confounding from collinear covariates.
+Alternatively, the trend could be attributed to a decreasing reporting rate rather than decreasing transmission rate, resulting in confounded mechanisms.
+The robust statistical conclusion is that a model which allows for change fits better than one which does not---we argue that a decreasing transmission rate is a plausible way to explain this, but the incidence data themselves do not provide enough information to pin down the mechanism.
+
 % Place figure captions after the first paragraph in which they are cited.
@@
 Epidemiological forecasts based on deterministic models are not uncommon in practice, despite their limitations \cite{king15}.
-That may explain why \cite{lee20} chose to obtain forecasts by simulating the calibrated models forward from initial conditions.
+That may explain why Lee et al.~\cite{lee20} chose to obtain forecasts by simulating the calibrated models forward from initial conditions.
 Here, we compare with a forecast projected from the filtering distribution: to obtain a forecast using POMP model $m$ at a collection of times $t_{N+1:N+s}$, where $N$ is the index for the last available data, we simulate forward from a draw from
@@
 
-The primary forecasting goal of \cite{lee20} was to investigate the potential consequences of vaccination interventions on a system to inform policy.
+The primary forecasting goal of Lee et al.~\cite{lee20} was to investigate the potential consequences of vaccination interventions on a system to inform policy.
 One outcome of their study include estimates for the probability of cholera elimination under several possible vaccination scenarios.
@@
     s0_pf <- pfilter(s0, save.state = TRUE, Np = h1_scen_NP)
-    end_states <- saved.states(s0_pf)
+    end_states <- saved_states(s0_pf)
     end_states <- end_states[[length(end_states)]]
@@
     s2_pf <- pfilter(s2, save.state = TRUE, Np = h1_scen_NP)
-    end_states <- saved.states(s2_pf)
+    end_states <- saved_states(s2_pf)
     end_states <- end_states[[length(end_states)]]
@@
     s4_pf <- pfilter(s4, save.state = TRUE, Np = h1_scen_NP)
-    end_states <- saved.states(s4_pf)
+    end_states <- saved_states(s4_pf)
     end_states <- end_states[[length(end_states)]]
@@
     s3_pf <- pfilter(s3, save.state = TRUE, Np = h1_scen_NP)
-    end_states <- saved.states(s3_pf)
+    end_states <- saved_states(s3_pf)
     end_states <- end_states[[length(end_states)]]
@@
     s1_pf <- pfilter(s1, save.state = TRUE, Np = h1_scen_NP)
-    end_states <- saved.states(s1_pf)
+    end_states <- saved_states(s1_pf)
     end_states <- end_states[[length(end_states)]]
@@
 Probability of elimination across simulations for a 10 year period.
-Compare to Figure 3A of \cite{lee20}.}
+Compare to Fig.~3A of \cite{lee20}.}
 \end{figure}
@@
 In our case study, we found that additional attention to statistical details could have resulted in improved statistical fits to the observed incidence data, leading to improvements in the accuracy of the resulting policy guidance.
-We used the same data and models, and even much of the same code, as \cite{lee20}, and yet ended up with drastically different conclusions.
+We used the same data and models, and even much of the same code, as Lee et al.~\cite{lee20}, and yet ended up with drastically different conclusions.
 We acknowledge the benefit of hindsight: our demonstration of a statistically principled route to obtain better-fitting models resulting in more robust insights does not rule out the possibility of discovering other models that fit well yet predict poorly.
@@
 
-\subsection*{Robust interpretation of model based conclusions}\label{sec:advice}
-
-A model which aspires to provide quantitative guidance for assessing interventions should provide a quantitative statistical fit for available data.
-However, strong statistical fit does not guarantee a correct causal structure: it does not even necessarily require the model to assert a causal explanation.
-A causal interpretation is strengthened by corroborative evidence.
-For example, reconstructed latent variables (such as numbers of susceptible and recovered individuals) should make sense in the context of alternative measurements of these variables \cite{grad12}.
-In the context of our case study, by reconstructing the latent states of Model~3, we notice that the calibrated model favors higher levels of cholera transmission than what was typically observed in the incidence data (Sec.~S5.1 of the supplement).
-This result hints at the possibility of model mispecification, and should prompt us to be sceptical of the reliability of forecasts from this model.
-Similarly, parameters that have been calibrated to data should make sense in the context of alternative lines of evidence about the phenomena being modeled, while making allowance for the possibility that the interpretations of parameters may vary when modeling across differing spatial scales.
-
-If a mechanistic model including a feature (such as a representation of a mechanism, or the inclusion of a covariate) fits better than mechanistic models without that feature, and also has competitive fit compared to associative benchmarks, this may be taken as evidence supporting the scientific relevance of the feature.
-As for any analysis of observational data, we must be alert to the possibility of confounding.
-For a covariate, this shows up in a similar way to regression analysis: the covariate under investigation could be a proxy for some other unmodeled or unmeasured covariate.
-For a mechanism, the model feature could in principle explain the data by helping to account for some different unmodeled phenomenon.
-In the context of our analysis, the estimated trend in transmission rate could be explained by any trending variable (such as hygiene improvements, or changes in population behavior), resulting in confounding from collinear covariates.
-Alternatively, the trend could be attributed to a decreasing reporting rate rather than decreasing transmission rate, resulting in confounded mechanisms.
-The robust statistical conclusion is that a model which allows for change fits better than one which does not---we argue that a decreasing transmission rate is a plausible way to explain this, but the incidence data themselves do not provide enough information to pin down the mechanism.
-
 \subsection*{Reproducibility and Extendability}
@@
 In addition to reproducing the article, the code can be extended to examine alternative analysis to that presented.
-The dynamic document, together with the R packages, form a {\it compendium}, defined by \cite{gentleman07} as a distributable and executable unit which combines data, text and auxiliary software (the latter meaning code written to run in a general-purpose, portable programming environment, which in this case is R).
+The dynamic document, together with the R packages, form a {\it compendium}, defined by Gentleman and Temple Lang~\cite{gentleman07} as a distributable and executable unit which combines data, text and auxiliary software (the latter meaning code written to run in a general-purpose, portable programming environment, which in this case is R).
 
@@
 
+% TODO: we have to remove this and copy the output from the .bbl file.
 \bibliography{bib-haiti}

### 2023-07-14 ms-plos.Rnw
Changed model3 save file

@@
 haiti3_fit <- bake(
-  file = paste0("model3/", rl_dir, "haiti3_coupled_fit.rds"), {
+  file = paste0("model3/", rl_dir, "haiti3_fit.rds"), {
     fit_haiti3(

### 2023-07-14 ms-plos.Rnw
Keep track of likelihood matrix

@@
     KEEP_TRACES = FALSE,
-    KEEP_LIK_MAT = FALSE
+    KEEP_LIK_MAT = TRUE
   )

### 2023-08-15 ms-plos.Rnw
changed run_level
minor edits + updated TODO list

@@
 {\Large
-\textbf\newline{Title of submission to PLOS journals} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
+\textbf\newline{Informing policy via dynamic models: Cholera in Haiti} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
 }
@@
 
-For any time point in $t_{1:N}$, the process model describes the movement of individuals from one latent state compartment to another.
+For any time point in $t_{1:N}$, the process model $f_{\bm{X}_n|\bm{X}_{n-1}}\big(\bm{x}_{n}|\bm{x}_{n-1}; \paramVec\big)$ is defined by the describing how individuals move from one latent state compartment to another.
 Per-capita transition rates are given in Eqs.~\myeqref{model1:SE}-\myeqref{model1:birth}:
@@
 
-The latent state vector $\bm{X}(t)$ is initialized by setting each compartment count for each vaccination scenario $\vaccCounter \neq 0$ as zero, and introducing initial-value parameters $I_{0,0}$ and $E_{0, 0}$ such that $R_{0}(0) = 0$, $I_{0}(0) = \mathrm{Pop} \times I_{0, 0}$, $E_{0}(0) = \mathrm{Pop} \times E_{0, 0}$ and $S_{0}(0) = \mathrm{Pop} \times (1 - I_{0, 0} - E_{0, 0})$, where $\mathrm{Pop}$ is the total population of Haiti.
+The latent state vector $\bm{X}(t)$ is initialized by setting the counts for each compartment and vaccination scenario $\vaccCounter \neq 0$ as zero, and introducing initial-value parameters $I_{0,0}$ and $E_{0, 0}$ such that $R_{0}(0) = 0$, $I_{0}(0) = \mathrm{Pop} \times I_{0, 0}$, $E_{0}(0) = \mathrm{Pop} \times E_{0, 0}$ and $S_{0}(0) = \mathrm{Pop} \times (1 - I_{0, 0} - E_{0, 0})$, where $\mathrm{Pop}$ is the total population of Haiti.
 The measurement model assumes reported cholera cases at time point $n$ come from a negative binomial distribution, where only a fraction ($\reportRate$) of new weekly cases are reported.
@@
 
+Like Model~1, the process model $f_{\bm{X}_n|\bm{X}_{n-1}}\big(\bm{x}_{n}|\bm{x}_{n-1}; \paramVec\big)$ is primarily defined via the description of movement of individuals between compartments, however Model~2 also includes a dynamic description of a latent bacterial compartment as well.
 Individuals can progress to a latent infection $E_{u\vaccCounter}$ followed by symptomatic infection $I_{u\vaccCounter}$ with recovery to $R_{u\vaccCounter}$ or asymptomatic infection $A_{u\vaccCounter}$ with recovery to $R^A_{u\vaccCounter}$.
@@
 
-Per-capita transition rates are given in Eqs.~\myeqref{eq:model3:SI}--\myeqref{eq:model3:Decay}.
+Per-capita transition rates are used for both compartments representing human counts and the aquatic reservoir of bacteria; these rates  are given in Eqs.~\myeqref{eq:model3:SI}--\myeqref{eq:model3:Decay}.
 \begin{eqnarray}
@@
 The resulting confidence interval for $\transmissionTrend$ is $(\Sexpr{myround(mcap_results$ci[1], digits = 3)}, \Sexpr{myround(mcap_results$ci[2], digits = 3)})$, with the full results displayed in Fig.~\ref{fig:betat}.
-These results are suggestive that the inclusion of a trend in transmission rate improves the quantitative ability of Model~1 to describe the observed data.
+These results are suggestive that the inclusion of a trend in the transmission rate improves the quantitative ability of Model~1 to describe the observed data.
 The reported results for Model~1 in the remainder of this article were obtained with the inclusion of the parameter $\transmissionTrend$.
@@
 Many Monte Carlo replications are then required to quantify and further reduce the error.
-The large difference in likelihoods that were obtain by these two approaches (see Table~\ref{tab:likes}) can primarily be attributed to increasing the computational effort in fitting the model; this result highlights the importance of carefully determining the necessary computational effort needed to maximize model likelihoods and acting accordingly.
+The large increase in the log-likelihood for Model~1 (see Table~\ref{tab:likes}) can primarily be attributed to increasing the computational effort in fitting the model; this result highlights the importance of carefully determining the necessary computational effort needed to maximize model likelihoods and acting accordingly.
 

### 2023-08-21 ms-plos.Rnw
minor edits after read-through

@@
 Quantitative models for these dynamic systems offer potential for designing effective control measures.
-Developing and testing such models, and assessing their fitness for guiding policy, is a challenging statistical task.
+Developing and testing models for dynamic systems, and assessing their fitness for guiding policy, is a challenging statistical task.
 Questions of interest include: What indications should we look for in the data to assess whether the model-based inferences are trustworthy?
@@
 Beyond this, the groups made autonomous decisions on what to include and exclude from their models.
-The groups largely adhered to existing guidelines on creating models to inform policy \cite{behrend20,saltelli20} and, despite their autonomy, obtained a consensus that an extensive nationwide vaccination campaign would be necessary to eliminate cholera from Haiti.
+The groups largely adhered to existing guidelines for creating models to inform policy \cite{behrend20,saltelli20} and, despite their autonomy, obtained a consensus that an extensive nationwide vaccination campaign would be necessary to eliminate cholera from Haiti.
 Their forecasts, however, were inconsistent with the prolonged period with no confirmed cholera cases between February, 2019 and September, 2022 \cite{trevisin22}.
@@
 $k\in\seq{1}{3}$ models non-exponential duration in the recovered class before waning of immunity.
+The processes model $f_{\bm{X}_n|\bm{X}_{n-1}}\big(\bm{x}_n|\bm{x}_{n-1};\paramVec\big)$ describes the movement of individuals between latent compartments, as well as the birth and death process of local, unobserved bacterial compartments $W_{u}(t)$.
 The force of infection is
@@
 In the case of a POMP model, the likelihood function---as described in Eq.~\myeqref{eq:likedef}---is written as $\mathcal{L}(\paramVec) = \int f_{\bm{X}_{0:N}, \bm{Y}_{1:N}}\big(\bm{x}_{0:N}, \bm{y}_{1:N};\paramVec\big) d\bm{x}_{0:N}$.
-The high-dimensional integral in the likelihood, and the potentially non-linear nature of the process model, makes it impossible to obtain a closed form expression of the likelihood function of POMP models, except in the most basic cases.
-For example, the process model that defines Model~1 was developed with the intent of providing a plausible description of cholera dynamics rather than a providing statistically convenient description.
+The high-dimensional integral in the likelihood, and the potentially non-linear nature of the process model, make it impossible to obtain a closed form expression of the likelihood function of POMP models except in the most basic cases.
+The process model that defines Model~1 was developed with the intent of providing a plausible description of cholera dynamics rather than a providing statistically convenient description.
 Because of this, a closed form expression of the likelihood of Model~1 is not readily available.
@@
 The proposal of a linear trend in transmission is a result of observing an apparent decrease in reported cholera infections from 2012-2019 in Fig.~\ref{CholeraData}.
-While several factors may contribute to this decrease, one explanation is that case-area targeted interventions (CATIs), which included education sessions, increased monitoring, household decontamination, soap distribution, and water chlorination in infected areas \cite{rebaudet19CATI}, may have substantially reduced cholera transmission \cite{rebaudet21}.
+While several factors may contribute to this decrease, one explanation is that case-area targeted interventions (CATIs), which included education sessions, increased monitoring, household decontamination, soap distribution, and water chlorination in infected areas \cite{rebaudet19CATI}, may have substantially reduced cholera transmission over time \cite{rebaudet21}.
 
@@
 as parameters that maximize the likelihood of the decoupled model almost certainly do not maximize the likelihood of the fully coupled model.
-These two concerns may explain the unrealistic forecasts and low likelihood of the model (see Table~\ref{tab:likes}).
-
+These two concerns may explain the unrealistic forecasts and low likelihood of Model~3 from Lee et al.~\cite{lee20} (see Table~\ref{tab:likes}).
+
+At the time Lee et al.~\ref{lee20} was published, there was no known algorithm that could readily be used to maximize the likelihood of an arbitrary meta-population POMP model with coupled spatial dynamics, which justifies the spatial decoupling approximation that was used to calibrate model parameters.
 For our analysis, we calibrate the parameters of the spatially coupled version of Model~3 using the iterated block particle filter (IBPF) algorithm \cite{ionides22}.
@@
 The IBPF algorithm enables us to directly estimate the parameters of models describing high-dimensional partially-observed nonlinear dynamic systems via likelihood maximization.
-The ability to directly estimate parameters of Model~3 is responsible for the large increase in model likelihoods reported in Table~\ref{tab:likes}, as no algorithm with similar capabilities existed when Lee et al.~\cite{lee20} published their results.
+The ability to directly estimate parameters of Model~3 is responsible for the large increase in model likelihoods reported in Table~\ref{tab:likes}.
 Simulations from the fitted model are displayed in Fig.~\ref{fig:h3spatsims}.
@@
 The goal of parameter calibration---whether done using Bayesian or frequentist methods---is to find the best description of the observed data under the assumptions of the model.
-Obtaining the best fitting set of parameters for a given model does not, however, guarantee that the model provides an accurate representation of the system in question.
+Obtaining the best fitting set of parameters for a given model does not, however, guarantee that the model provides an accurate representation of the system under investigation.
 Model misspecification, which may be thought of as the omission of a mechanism in the model that is an important feature of the dynamic system, is inevitable at all levels of model complexity.
@@
 In this section, we discuss some tools for diagnosing mechanistic models with the goal of making the subjective assessment of model ``usefulness'' more objective.
-To do this, we will rely on the quantitative and statistical ability of the model to match the observed data, which we call the model's {\it goodness-of-fit}, with the guiding principle that a model which cannot adequately describe observed data may not be reliable for useful purposes.
+To do this, we will rely on the quantitative ability of the model to match the observed data, which we call the model's {\it goodness-of-fit}, with the guiding principle that a model which cannot adequately describe observed data may not be reliable for useful purposes.
 Goodness-of-fit may provide evidence supporting the causal interpretation of one model versus another, but cannot by itself rule out the possibility of alternative explanations.
@@
 While visual comparisons can be informative, they provide only a weak and informal measure of the goodness-of-fit of a model.
-The study by Lee et al.~\cite{lee20} provides an example of this: their models and parameter estimates resulted in simulations that visually resembled the observed data, yet resulted in model likelihoods that were---in some cases---considerably smaller than likelihoods that can be achieved via the likelihood based optimization techniques that were used (see Table~\ref{tab:likes}).
+The study by Lee et al.~\cite{lee20} provides an example of this: their models and parameter estimates resulted in simulations that visually resembled the observed data, yet resulted in model likelihoods that were considerably smaller than likelihoods that can be achieved via the likelihood based optimization techniques that were used (see Table~\ref{tab:likes}).
 Alternative forms of model validation should therefore be used in conjunction with visual comparisons of simulations to observed data.
@@
 In some cases, a possible benchmark model could be a generally accepted mechanistic model, but often no such model is available.
-Because of this, we use a log-linear Gaussian ARMA model as an associative benchmark, as recommended by \cite{he10}.
+Because of this, we use a log-linear Gaussian ARMA model as an associative benchmark, as recommended by He et al.~\cite{he10}.
 The theory and practice of ARMA models is well developed, and these linear models are appropriate on a log scale due to the exponential growth and decay characteristic of biological dynamics.
@@
 For a mechanism, the model feature could in principle explain the data by helping to account for some different unmodeled phenomenon.
-In the context of our analysis, the estimated trend in transmission rate could be explained by any trending variable (such as hygiene improvements, or changes in population behavior), resulting in confounding from collinear covariates.
+In the context of our analysis, the statistical evidence of a trend in transmission rate in Model~1 could be explained by any trending variable (such as hygiene improvements, or changes in population behavior), resulting in confounding from colinear covariates.
 Alternatively, the trend could be attributed to a decreasing reporting rate rather than decreasing transmission rate, resulting in confounded mechanisms.
@@
 That may explain why Lee et al.~\cite{lee20} chose to obtain forecasts by simulating the calibrated models forward from initial conditions.
-Here, we compare with a forecast projected from the filtering distribution: to obtain a forecast using POMP model $m$ at a collection of times $t_{N+1:N+s}$, where $N$ is the index for the last available data, we simulate forward from a draw from
-$f_{\bm{X}_N|\bm{Y}_{1:N}}(\bm{x}_{N} | \bm{y}^*_{1:N} ; \hat\paramVec)$
+Here, we obtain forecasts by simulating future values using latent states that are consistent with the most recent observations by drawing latent states at time the last observation time $t_N$ from the filtering distribution $f_{\bm{X}_N|\bm{Y}_{1:N}}(\bm{x}_{N} | \bm{y}^*_{1:N} ; \hat\paramVec)$,
 where $\hat\paramVec$ is a vector of calibrated parameters.
-The decision not to do this partially explains the unsuccessful forecasts of \cite{lee20}: their Table~S7 shows that the subset of their simulations which were consistent with observing zero cases in 2019 also accurately predicted the prolonged absence of detected cholera.
+The decision not to do this partially explains the unsuccessful forecasts of Lee et al.~\cite{lee20}: their Table~S7 shows that the subset of their simulations which were consistent with observing zero cases in 2019 also accurately predicted the prolonged absence of detected cholera.
 
@@
 
-\cite{lee20} published their code and data online, and this reproducibility facilitated our work.
+Lee et al.~\cite{lee20} published their code and data online, and this reproducibility facilitated our work.
 Robust data analysis requires not only reproducibility but also extendability: if one wishes to try new model variations, or new approaches to fitting the existing models, or plotting the results in a different way, this should not be excessively burdensome.
@@
 Using a software package mechanism supports documentation, standardization and portability that promote extendability.
-In the terminology of \cite{gentleman07}, the source code for this article is a {\it dynamic document} combining code chunks with text.
+In the terminology of Gentleman and Temple Lang~\cite{gentleman07}, the source code for this article is a {\it dynamic document} combining code chunks with text.
 In addition to reproducing the article, the code can be extended to examine alternative analysis to that presented.

### 2023-08-21 ms-plos.Rnw
First draft of author summary

@@
 \section*{Author summary}
-Lorem ipsum dolor sit amet, consectetur adipiscing elit.
-Curabitur eget porta erat.
-Morbi consectetur est vel gravida pretium.
-Suspendisse ut dui eu ante cursus gravida non sed sem.
-Nullam sapien tellus, commodo id velit id, eleifend volutpat quam.
-Phasellus mauris velit, dapibus finibus elementum vel, pulvinar non tellus.
-Nunc pellentesque pretium diam, quis maximus dolor faucibus id.
-Nunc convallis sodales ante, ut ullamcorper est egestas vitae.
-Nam sit amet enim ultrices, ultrices elit pulvinar, volutpat risus.
+Scientists often use data from an infectious disease outbreak to build statistical models.
+Among other things, these models are often used to guide governmental policy, as the models can provide insight into the dynamics of an outbreak.
+Despite recent computational and methodological advancements, fitting dynamic models to observed incidence data remains a challenging task.
+In this article, we discuss common mistakes that may lead to models that do not adequately describe reality, and make suggestions that can lead to more useful statistical models.
+We do this by conducting a re-analysis of the 2012-2019 cholera outbreak in Haiti.
+We focus on the importance of appropriately diagnosing how well a model describes observed data, correctly interpreting model based results, and stressing the importance of reproducibility.
+An important contribution of this article is demonstrating a use-case of a newly developed computational algorithm to fit spatially coupled meta-population models.
+We assert that raising the currently accepted standards of infectious disease modeling will result in a greater ability of scientists and policy makers to understand and respond to future infectious disease outbreaks.
 
@@
 On this same subset, our model has a likelihood of $\Sexpr{myround(mod3_eval_res$subset_ll, 1)}$.
-See Sec.~S6 of the supplement material for more details on estimating the likelihood of the \cite{lee20} models.
+See Sec.~S6 of the supplement material for more details on estimating the likelihood of the Lee et al.~\cite{lee20} models.
 }

### 2023-08-21 ms-plos.Rnw
Formatted the parameter table

@@
 % \small
-\begin{tabular}{|l@{}|l@{}t@{}|l@{}t@{}|l@{}t@{}|}
+\begin{tabular}{|p{0.15\linewidth}|p{0.14\linewidth}t|p{0.13\linewidth}t|p{0.17\linewidth}t|}
 \hline
@@
 Infection (day)
-  & {\fixed $\muIR^{-1}=\Sexpr{myround(7/p1["gamma"],1)}$}
+  & ${\fixed \muIR^{-1}=\Sexpr{myround(7/p1["gamma"],1)}}$
   & \eqref{model1:toR}
-  & {\fixed $\muIR^{-1}=\Sexpr{myround(365/h2_fit$h2_params["gamma"],1)}$}
+  & ${\fixed \muIR^{-1}=\Sexpr{myround(365/h2_fit$h2_params["gamma"],1)}}$
   & \eqref{model2:mu_IR}
-  & {\fixed $\muIR^{-1}=\Sexpr{myround(365/p3["gamma1"],1)}$ }
+  & ${\fixed \muIR^{-1}=\Sexpr{myround(365/p3["gamma1"],1)}}$
   & \eqref{eq:model3:IR}
@@
 Latency (day)
-  & {\fixed $\muEI^{-1}=\Sexpr{myround(7/p1["sigma"],1)}$}
+  & ${\fixed \muEI^{-1}=\Sexpr{myround(7/p1["sigma"],1)}}$
   & \eqref{model1:EA}
-  & {\fixed $\muEI^{-1}=\Sexpr{myround(365/h2_fit$h2_params["gammaE"],1)}$}
+  & ${\fixed \muEI^{-1}=\Sexpr{myround(365/h2_fit$h2_params["gammaE"],1)}}$
   & \eqref{model2:mu_EI}
@@
 \hline
-\multirow{2}{*}{Seasonality}
-  & $\begin{array}{l}
-    \hspace{-1mm} \transmission_{1:6}=(
-      \Sexpr{myround(p1["beta1"],1)},
-      \Sexpr{myround(p1["beta2"],1)},
-    \\
-      \Sexpr{myround(p1["beta3"],1)},
-      \Sexpr{myround(p1["beta4"],1)},
-      \Sexpr{myround(p1["beta5"],1)},
-      \Sexpr{myround(p1["beta6"],1)})
-    \end{array}$
+\multirow{3}{*}{Seasonality}
+  & $\transmission_{1:6}=(
+      \Sexpr{myround(p1["beta1"],1)},\allowbreak
+      \Sexpr{myround(p1["beta2"],1)},\allowbreak
+      \Sexpr{myround(p1["beta3"],1)},\allowbreak
+      \Sexpr{myround(p1["beta4"],1)},\allowbreak
+      \Sexpr{myround(p1["beta5"],1)},\allowbreak
+      \Sexpr{myround(p1["beta6"],1)})$
   & \eqref{model1:beta}
-  & {\fixed \seasAmplitude=0.4}
-  & \eqref{model2:lambda}
-  & $\seasAmplitude = \Sexpr{myround(p3s["lambdaR"],2)}$
-  & \eqref{eq:model3:water} \\
+  & \multirow{2}{*}{${\fixed \seasAmplitude=0.4}$}
+  & \multirow{2}{*}{\eqref{model2:lambda}}
+  & \multirow{2}{*}{$\seasAmplitude = \Sexpr{myround(p3s["lambdaR"],2)}$}
+  & \multirow{2}{*}{\eqref{eq:model3:water}}
+  \\
   & $\transmissionTrend = \Sexpr{myround(p1["betat"],2)}^*$
   & \eqref{model1:betat}
-  & $\phaseParm = \Sexpr{myround(h2_fit$h2_params['phase'], 2)}^*$
+  & \multirow{-2}{*}{$\phaseParm = \Sexpr{myround(h2_fit$h2_params['phase'], 2)}^*$}
+  & \multirow{-2}{*}{\eqref{model2:lambda}}
+  & \multirow{-2}{*}{$\rainfallExponent = \Sexpr{signif(p3s["r"],3)}$}
+  & \multirow{-2}{*}{\eqref{eq:model3:water}}
+  \\\hline
+  \multirow{3}{*}{Immunity (yr)}
+  & \multirow{3}{*}{${\fixed \muRS^{-1}=\Sexpr{myround(1/p1["alpha"]/52,1)}}$}
+  & \multirow{3}{*}{\eqref{model1:RS}}
+  & $\muRS^{-1}= \Sexpr{signif(1/h2_fit$h2_params["sigma"],2)}$
+  & \eqref{model2:RS}
+  & \multirow{3}{*}{${\fixed \muRS^{-1}=\Sexpr{myround(1/p3['rho1'],1)}}$}
+  & \multirow{3}{*}{\eqref{eq:model3:RRnext}}
+\\
+& & & ${\fixed \omega_1^{-1} = \Sexpr{myround(1/h2_fit$h2_params["Omega1"],1)}}$ & \eqref{model2:omega1} & &\\
+& & & ${\fixed \omega_2^{-1} = \Sexpr{myround(1/h2_fit$h2_params["Omega2"],1)}}$ & \eqref{model2:omega2} & &\\
+\hline
+\multirow{2}{*}{Birth/death (yr)}
+  & ${\fixed \muBirth^{-1} = \Sexpr{myround(1/p1["mu"]/52,1)}}$
+    %
+  & \eqref{model1:death}
+  & \multirow{2}{*}{\missing}
+  &
+  & \multirow{2}{*}{${\fixed \muDeath^{-1} = \Sexpr{myround(1/p3["mu1"],1)}}$}
+  & \multirow{2}{*}{\eqref{eq:model3:IS}}
+\\
+& ${\fixed \muDeath^{-1} = \Sexpr{myround(1/p1["delta"]/52,1)}}$
+& \eqref{model1:death}
+&
+&
+& &
+\\
+Sympt. frac.
+  & ${\fixed \symptomFrac_z(t)=c\vaccineEfficacy^*(t-\tau_d)}$
+  & (\ref{model1:EI}-\ref{model1:EA})
+  & ${\fixed \symptomFrac=\Sexpr{myround(h2_fit$h2_params["k"],1)}}$
+  & \eqref{model2:mu_EI}
+  & ${\fixed \symptomFrac=\Sexpr{myround(p3["sigma1"],2)}}$
+  & \eqref{eq:model3:SA}
+\\
+Asympt.
+  & \multirow{2}{*}{${\fixed \asymptomRelativeInfect=0.05}$}
+  & \multirow{2}{*}{\eqref{model1:lambda}}
+  & ${\fixed \asymptomRelativeInfect = 0.001}$
   & \eqref{model2:lambda}
-  & $r = \Sexpr{signif(p3s["r"],3)}$
+  &  ${\fixed \asymptomRelativeInfect = 1}$
+  & \eqref{eq:model3:foi}
+\\
+infectivity
+&
+&
+& ${\fixed \asymptomRelativeShed = \Sexpr{1e-7} }$
+& \eqref{model2:to_W}
+& $\asymptomRelativeShed = \Sexpr{myround(p3s["XthetaA"],3)}$
 & \eqref{eq:model3:water}
-  \\\hline
-  Immunity (yr)
-  & {\fixed $\muRS^{-1}=\Sexpr{myround(1/p1["alpha"]/52,1)}$}
-  & \eqref{model1:RS}
+\\
+\hline
+\multirow{4}{*}{Human to human}
+  & \multirow{4}{*}{$\transmission_{1:6}$ as above}
+  & \multirow{4}{*}{\eqref{model1:lambda}}
+  & \multirow{4}{*}{$\transmission=$\Sexpr{signif(h2_fit$h2_params["Beta"],3)}}
+  & \multirow{4}{*}{\eqref{model2:lambda}}
+  & $\transmission_{1:10}=(
+  \Sexpr{myround(p3u["foi_add",1]*1e6,2)}, \allowbreak
+	\Sexpr{myround(p3u["foi_add",2]*1e6,2)}, \allowbreak
+	\Sexpr{myround(p3u["foi_add",3]*1e6,2)}, \allowbreak
+  \Sexpr{myround(p3u["foi_add",4]*1e6,2)}, \allowbreak
+	\Sexpr{myround(p3u["foi_add",5]*1e6,2)}, \allowbreak
+	\Sexpr{myround(p3u["foi_add",6]*1e6,2)}, \allowbreak
+	\Sexpr{myround(p3u["foi_add",7]*1e6,2)}, \allowbreak
+	\Sexpr{myround(p3u["foi_add",8]*1e6,2)}, \allowbreak
+	\Sexpr{myround(p3u["foi_add",9]*1e6,2)}, \allowbreak
+	\Sexpr{myround(p3u["foi_add",10]*1e6,2)}
+	) \allowbreak \times 10^{-6}$
+  & \multirow{4}{*}{\eqref{eq:model3:foi}}
+\\
+\hline
+\multirow{5}{*}{Water to human}
+  & \multirow{5}{*}{\missing}
+  % & \multicolumn{1}{c}{\hspace{16pt}\missing}
   &
-    $\begin{array}{l}
-    \muRS^{-1}= \Sexpr{signif(1/h2_fit$h2_params["sigma"],2)} \\
-    {\fixed \omega_1^{-1} = \Sexpr{myround(1/h2_fit$h2_params["Omega1"],1)}} \\
-    {\fixed \omega_2^{-1} = \Sexpr{myround(1/h2_fit$h2_params["Omega2"],1)}}
-    \end{array}$
-  & $\begin{array}{l}
-    \eqref{model2:RS} \\
-    \eqref{model2:omega1} \\
-    \eqref{model2:omega1}
-    \end{array} $
-  & {\fixed $\muRS^{-1}=\Sexpr{myround(1/p3['rho1'],1)}$}
-  & \eqref{eq:model3:RRnext}
+  & \multirow{3}{*}{${\fixed \Wsat = \Sexpr{signif(h2_fit$h2_params["Sat"],2)}}$}
+  & \multirow{3}{*}{\eqref{model2:lambda}}
+  % \eqref{model2:lambda}
+  & $\Wbeta{_{1:10}}= (
+        \Sexpr{myround(p3u["betaB",1],2)}, \allowbreak
+        \Sexpr{myround(p3u["betaB",2],2)}, \allowbreak
+        \Sexpr{myround(p3u["betaB",3],2)}, \allowbreak
+        \Sexpr{myround(p3u["betaB",4],2)}, \allowbreak
+        \Sexpr{myround(p3u["betaB",5],2)}, \allowbreak
+        \Sexpr{myround(p3u["betaB",6],2)}, \allowbreak
+        \Sexpr{myround(p3u["betaB",7],2)}, \allowbreak
+        \Sexpr{myround(p3u["betaB",8],2)}, \allowbreak
+        \Sexpr{myround(p3u["betaB",9],2)}, \allowbreak
+        \Sexpr{myround(p3u["betaB",10],2)})$
+  & \multirow{4}{*}{\eqref{eq:model3:foi}}
+\\
+& & & \multirow{-3}{*}{$\beta_W= \Sexpr{signif(h2_fit$h2_params["BetaW"],3)}$} & \multirow{-3}{*}{\eqref{model2:lambda}} & &
 \\
 \hline
-Birth/death (yr)
-  & \fixed{$\begin{array}{l}
-    \muBirth^{-1} = \Sexpr{myround(1/p1["mu"]/52,1)} \\
-    \muDeath^{-1} = \Sexpr{myround(1/p1["delta"]/52,1)}
-    \end{array}$}
-  & \eqref{model1:death}
-  & \multicolumn{1}{c}{\hspace{16pt}\missing}
-  &
-  & {\fixed $\muDeath^{-1} = \Sexpr{myround(1/p3["mu1"],1)}$}
-  & \eqref{eq:model3:IS}
-\\
-Sympt. frac.
-  & {\fixed $\symptomFrac_z(t)=c\vaccineEfficacy^*(t-\tau_d)$}
-  & (\ref{model1:EI}-\ref{model1:EA})
-  & {\fixed $\symptomFrac=\Sexpr{myround(h2_fit$h2_params["k"],1)}$}
-  & \eqref{model2:mu_EI}
-  & {\fixed $\symptomFrac=\Sexpr{myround(p3["sigma1"],2)}$}
-  & \eqref{eq:model3:SA}
-\\
-$\begin{array}{l}
-\text{Asympt.} \\
-\text{infectivity}
-\end{array}$
-  & {\fixed $\asymptomRelativeInfect=0.05$ }
-  & \eqref{model1:lambda}
-  & $\begin{array}{l}
-      {\fixed \asymptomRelativeInfect =0.001 } \\
-      {\fixed \asymptomRelativeShed = \Sexpr{1e-7} }
-    \end{array}$
-  & \begin{tabular}{l}
-      \eqref{model2:lambda} \\
-      \eqref{model2:to_W}
-    \end{tabular}
-  &  $\begin{array}{l}
-      {\fixed \asymptomRelativeInfect =1 } \\
-      \asymptomRelativeShed = \Sexpr{myround(p3s["XthetaA"],3)}
-    \end{array}$
-  & \begin{tabular}{l}
-      \eqref{eq:model3:foi} \\
-      \eqref{eq:model3:water}
-    \end{tabular}
-\\
-\hline
-$\begin{array}{l}
-\text{Human to}\\
-\text{human}\end{array}$
-  & $\transmission_{1:6}$ as above
-  & \eqref{model1:lambda}
-  & $\transmission=$\Sexpr{signif(h2_fit$h2_params["Beta"],3)}
-  & \eqref{model2:lambda}
-  & $\begin{array}{l}
-      \hspace{-1mm} \transmission_{1:10}=(
-        \Sexpr{myround(p3u["foi_add",1]*1e6,2)},
-	\Sexpr{myround(p3u["foi_add",2]*1e6,2)},
-      \\
-	\Sexpr{myround(p3u["foi_add",3]*1e6,2)},
-        \Sexpr{myround(p3u["foi_add",4]*1e6,2)},
-	\Sexpr{myround(p3u["foi_add",5]*1e6,2)},
-	\Sexpr{myround(p3u["foi_add",6]*1e6,2)},
-      \\
-	\Sexpr{myround(p3u["foi_add",7]*1e6,2)},
-	\Sexpr{myround(p3u["foi_add",8]*1e6,2)},
-	\Sexpr{myround(p3u["foi_add",9]*1e6,2)},
-	\Sexpr{myround(p3u["foi_add",10]*1e6,2)}
-	)
-      \\
-      \times 10^{-6}
-    \end{array}$
-  & \eqref{eq:model3:foi}
-\\
-\hline
-$\begin{array}{l}
-\text{Water to}\\ \text{human}
-\end{array}$
-  & \multicolumn{1}{c}{\hspace{16pt}\missing}
-  &
-  & $\begin{array}{l}
-    {\fixed \Wsat = \Sexpr{signif(h2_fit$h2_params["Sat"],2)} }
-    \\
-    \beta_W= \Sexpr{signif(h2_fit$h2_params["BetaW"],3)}
-    \end{array}$
-  & \eqref{model2:lambda}
-  & $\begin{array}{l}
-      \hspace{-1mm} \Wbeta{_{1:10}}= (
-        \Sexpr{myround(p3u["betaB",1],2)}, \Sexpr{myround(p3u["betaB",2],2)},
-       \\
-        \Sexpr{myround(p3u["betaB",3],2)}, \Sexpr{myround(p3u["betaB",4],2)},
-        \Sexpr{myround(p3u["betaB",5],2)}, \Sexpr{myround(p3u["betaB",6],2)},
-       \\
-        \Sexpr{myround(p3u["betaB",7],2)}, \Sexpr{myround(p3u["betaB",8],2)},
-        \Sexpr{myround(p3u["betaB",9],2)}, \Sexpr{myround(p3u["betaB",10],2)}
-      ) \end{array}$
-  & \eqref{eq:model3:foi}
-\\
-\hline
-$\begin{array}{l}
-\text{Human to}\\
-\text{water}\end{array}$
-  & \multicolumn{1}{c}{\hspace{16pt}\missing}
+Human to water
+  & \missing
+  % & \multicolumn{1}{c}{\hspace{16pt}\missing}
   &
@@
 \\
-$\begin{array}{l}
-\text{Water} \\
-\text{survival (wk)}
-\end{array}$
-  & \multicolumn{1}{c}{\hspace{16pt}\missing}
+Water survival \newline (wk)
+  & \missing
+  % & \multicolumn{1}{c}{\hspace{16pt}\missing}
   &
-  & {\fixed $\Wremoval^{-1} = $ \Sexpr{signif(52/h2_fit$h2_params["Delta"],3)}}
+  & ${\fixed \Wremoval^{-1} = \Sexpr{signif(52/h2_fit$h2_params["Delta"],3)}}$
   & \eqref{model2:from_W}
@@
   & \eqref{model1:lambda}
-  & \multicolumn{1}{c}{\hspace{16pt}\missing}
+  & \missing
+  % & \multicolumn{1}{c}{\hspace{16pt}\missing}
   &
-  & \multicolumn{1}{c}{\hspace{16pt}\missing}
+  & \missing
+  % & \multicolumn{1}{c}{\hspace{16pt}\missing}
   &
 \\
-$\begin{array}{l}
-\text{Process} \\
-\text{noise} (\text{wk}^{1/2})
-\end{array}$
-  & $\sigmaProc=(\Sexpr{myround(sqrt(p1["sig_sq_epi"]),2)}, \Sexpr{myround(sqrt(p1["sig_sq_end"]),2)})^*$
-  & \eqref{model1:lambda}
-  & \multicolumn{1}{c}{\hspace{16pt}\missing}
+\hline
+Process noise \newline ($\text{wk}^{1/2}$)
+  & $\sigmaProc =$ \newline $(\Sexpr{myround(sqrt(p1["sig_sq_epi"]),2)},
+  \Sexpr{myround(sqrt(p1["sig_sq_end"]),2)})^*$
+  & \multirow{2}{*}{\eqref{model1:lambda}}
+  & \multirow{2}{*}{\missing}
+  % & \multicolumn{1}{c}{\hspace{16pt}\missing}
   &
-  & $\sigmaProc= \Sexpr{myround(p3s["std_W"],3)}$
-  & \eqref{eq:model3:SA}
+  & \multirow{2}{*}{$\sigmaProc= \Sexpr{myround(p3s["std_W"],3)}$}
+  & \multirow{2}{*}{\eqref{eq:model3:SA}}
 \\
@@
   & $\reportRate=\Sexpr{myround(p1["rho"], 3)}$
-  & (S15)
-  & {\fixed $\reportRate=\Sexpr{myround(h2_fit$h2_params["Rho"],2)}$}
-  & (S16)
+  & (S15) % TODO: Check
+  & ${\fixed \reportRate=\Sexpr{myround(h2_fit$h2_params["Rho"],2)}}$
+  & (S16) % TODO: Check
   & $\reportRate=\Sexpr{myround(p3s["epsilon"],2)}$
-  & (S19)
-\\
-$\begin{array}{l}
-\text{Observation} \\
-\text{variance}
-\end{array}$
-  & $\obsOverdispersion = (\Sexpr{myround(p1["tau_epi"],2)}, \Sexpr{round(p1["tau_end"],2)})$
-  & (S15)
-  & $\obsOverdispersion = \Sexpr{signif(h2_fit$h2_params['sigma'], 3)}$
-  & (S16)
-  & $\obsOverdispersion = \Sexpr{myround(p3s["k"],2)}$
-  & (S19)
+  & (S19) % TODO: Check
+\\
+Observation variance
+  & $\obsOverdispersion =$ \newline $(\Sexpr{myround(p1["tau_epi"],2)}, \Sexpr{round(p1["tau_end"],2)})$
+  & \multirow{2}{*}{(S15)} % TODO: Check
+  & \multirow{2}{*}{$\obsOverdispersion = \Sexpr{signif(h2_fit$h2_params['sigma'], 3)}$}
+  & \multirow{2}{*}{(S16)} % TODO: Check
+  & \multirow{2}{*}{$\obsOverdispersion = \Sexpr{myround(p3s["k"],2)}$}
+  & \multirow{2}{*}{(S19)} % TODO: Check
 \\
 \hline
-\multirow{2}{*}{Initial Values} & $I_{0,0} = \Sexpr{round(prod(p1[c('pop_0', 'I_0')]))}$ & & \multicolumn{1}{c}{\multirow{2}{*}{\hspace{16pt}\missing}} & & \multirow{2}{*}{$I_{3,7,10;0}(0)= (
-        \Sexpr{round(prod(p3[c('H3', 'Iinit3')]))}, \Sexpr{round(prod(p3[c('H7', 'Iinit7')]))}, \Sexpr{round(prod(p3[c('H10', 'Iinit10')]))})^*$} & \\
-& $E_{0,0} = \Sexpr{round(prod(p1[c('pop_0', 'E_0')]))}$ & & \multicolumn{1}{c}{} & & & \\\hline
-Hurricane & \multicolumn{1}{c}{\multirow{2}{*}{\hspace{16pt}\missing}} & & \multicolumn{1}{c}{\multirow{2}{*}{\hspace{16pt}\missing}} & & $\Whur{3, 9} = (\Sexpr{myround(p3['aHur3'], 2)}, \Sexpr{myround(p3['aHur9'], 2)})^*$ & \multirow{2}{*}{\eqref{eq:model3:foi}} \\
-Parameters & \multicolumn{1}{c}{} & & \multicolumn{1}{c}{} & & $\hHur{3, 9} = (\Sexpr{myround(p3['hHur3'], 2)}, \Sexpr{myround(p3['hHur9'], 2)})^*$ & \\\hline
+\multirow{2}{*}{Initial Values} & $I_{0,0} = \Sexpr{round(prod(p1[c('pop_0', 'I_0')]))}$
+  &
+  & \multirow{2}{*}{\missing}
+  % & \multicolumn{1}{c}{\multirow{2}{*}{\hspace{16pt}\missing}}
+  &
+  & \multirow{2}{*}{$I_{3,7,10;0}(0)= (
+        \Sexpr{round(prod(p3[c('H3', 'Iinit3')]))}, \Sexpr{round(prod(p3[c('H7', 'Iinit7')]))}, \Sexpr{round(prod(p3[c('H10', 'Iinit10')]))})^*$}
+  &  % TODO: \multirow{2}{*}{EQUATION NUMBER}
+  \\
+  & $E_{0,0} = \Sexpr{round(prod(p1[c('pop_0', 'E_0')]))}$ & & \multicolumn{1}{c}{}
+  &
+  &
+  &
+\\\hline
+\multirow{3}{*}{Hurricane}
+  & \multirow{4}{*}{\missing}
+% & \multicolumn{1}{c}{\multirow{2}{*}{\hspace{16pt}\missing}}
+  &
+  & \multirow{4}{*}{\missing}
+  % & \multicolumn{1}{c}{\multirow{2}{*}{\hspace{16pt}\missing}}
+  &
+  & $\Whur{3, 9} =$\newline $(\Sexpr{myround(p3['aHur3'], 2)}, \Sexpr{myround(p3['aHur9'], 2)})^*$
+  & \multirow{2}{*}{\eqref{eq:model3:foi}}
+\\
+ \multirow{-1}{*}{Parameters} &
+  % & \multicolumn{1}{c}{}
+  &
+  &
+  % & \multicolumn{1}{c}{}
+  &
+  & $\hHur{3, 9} =$\newline $(\Sexpr{myround(p3['hHur3'], 2)}, \Sexpr{myround(p3['hHur9'], 2)})^*$
+  & \multirow{2}{*}{\eqref{eq:model3:foi}}
+\\\hline
 

### 2023-08-22 ms-plos.Rnw
Removed Vacc Sim Figures

@@
 We build on a multi-group modeling exercise by Lee et al.~\cite{lee20} in which four expert modeling teams developed models to the same dataset with the goal of comparing conclusions on the feasibility of eliminating cholera by a vaccination campaign.
+The data that was used for this study, and that is the subject of our reanalysis, is displayed in Fig.~\ref{fig:CholeraData}.
 Model~1 is stochastic and describes cholera at the national level;
@@
 
-The four independent teams were given the task of estimating the
-potential effect of prospective oral cholera vaccine (OCV) programs.
-While OCV is accepted as a safe and effective tool for controlling the spread of cholera, the global stockpile of OCV doses remains limited \cite{pezzoli20}.
-Advances in OCV technology and vaccine availability, however, raised the possibility of planning a national vaccination program \cite{lee20}.
-In the study, certain data were shared between the groups, including demography and vaccination history; vaccine efficacy was also fixed at a shared value between groups.
-Beyond this, the groups made autonomous decisions on what to include and exclude from their models.
-The groups largely adhered to existing guidelines for creating models to inform policy \cite{behrend20,saltelli20} and, despite their autonomy, obtained a consensus that an extensive nationwide vaccination campaign would be necessary to eliminate cholera from Haiti.
-Their forecasts, however, were inconsistent with the prolonged period with no confirmed cholera cases between February, 2019 and September, 2022 \cite{trevisin22}.
-Though cholera has recently reemerged in Haiti \cite{rubin22}, the inability to accurately forecast cholera incidence from 2019-2022---while diligently adhering to existing scientific standards for infectious disease modeling---demonstrates that existing standards of disease modeling and forecasting may be insufficient.
-
-The discrepancy between the model based conclusions of Lee et al.~\cite{lee20} and the prolonged absence of cholera in Haiti has been debated \cite{francois20,rebaudetComment20,henrys20,leeReply20}.
-Suggested origins of this discrepancy include the use of unrealistic models \cite{rebaudetComment20} and unrealistic criteria for cholera elimination \cite{henrys20}.
-We find a more nuanced conclusion: attention to methodological details in model fitting, diagnosis and forecasting can improve each of the proposed model's ability to quantitatively describe observed data.
-This improved ability may result in more accurate forecasts and facilitates the exploration of model assumptions.
-Based on this retrospective analysis, we offer suggestions on fitting mechanistic models to dynamic systems for future studies.
-
-While this work is focused on the models proposed by Lee et al.~\cite{lee20}, our suggestions have broader relevance.
-To investigate the extent to which \cite{lee20} is typical of the substantial body of work on Haiti cholera dynamics, we preformed a literature review by searching PubMed with keywords: Haiti, cholera, model.
-The search resulted in 66 papers, of which 32 used dynamic models to describe the cholera epidemic in Haiti.
-The models make various choices on the dichotomies considered by Lee et al.~\cite{lee20}; deterministic versus stochastic; compartment model versus agent-based; aggregated versus spatially explicit.
-This is no accident, since Lee et al.~\cite{lee20} purposefully designed their study to cover the range of current modeling practice.
-
-We proceed by introducing the general modeling scheme employed by Models~1--3 and provide details of each individual model;
-we then describe how each model is calibrated to data, and present a systematic approach to examining and refining these models.
-We then use improved model fits to project cholera incidence in Haiti under various vaccination scenarios considered by Lee et al.~\cite{lee20}.
-Finally, we conclude with a discussion of the results.
-% This is followed by a consideration of the robustness of model based policy recommendations in Sec.~\ref{sec:advice}, and a discussion in Sec.~\ref{sec:discussion}.
-
-\begin{figure}
+\begin{figure}[!h]
 <<Plot_Reported_Cases, echo=FALSE, fig.height=3.5>>=
@@
 @
-\caption{\label{CholeraData}
+\caption{\label{fig:CholeraData}
 \small
@@
 \end{figure}
+
+The four independent teams were given the task of estimating the
+potential effect of prospective oral cholera vaccine (OCV) programs.
+While OCV is accepted as a safe and effective tool for controlling the spread of cholera, the global stockpile of OCV doses remains limited \cite{pezzoli20}.
+Advances in OCV technology and vaccine availability, however, raised the possibility of planning a national vaccination program \cite{lee20}.
+In the study, certain data were shared between the groups, including demography and vaccination history; vaccine efficacy was also fixed at a shared value between groups.
+Beyond this, the groups made autonomous decisions on what to include and exclude from their models.
+The groups largely adhered to existing guidelines for creating models to inform policy \cite{behrend20,saltelli20} and, despite their autonomy, obtained a consensus that an extensive nationwide vaccination campaign would be necessary to eliminate cholera from Haiti.
+Their forecasts, however, were inconsistent with the prolonged period with no confirmed cholera cases between February, 2019 and September, 2022 \cite{trevisin22}.
+Though cholera has recently reemerged in Haiti \cite{rubin22}, the inability to accurately forecast cholera incidence from 2019-2022---while diligently adhering to existing scientific standards for infectious disease modeling---demonstrates that existing standards of disease modeling and forecasting may be insufficient.
+
+The discrepancy between the model based conclusions of Lee et al.~\cite{lee20} and the prolonged absence of cholera in Haiti has been debated \cite{francois20,rebaudetComment20,henrys20,leeReply20}.
+Suggested origins of this discrepancy include the use of unrealistic models \cite{rebaudetComment20} and unrealistic criteria for cholera elimination \cite{henrys20}.
+We find a more nuanced conclusion: attention to methodological details in model fitting, diagnosis and forecasting can improve each of the proposed model's ability to quantitatively describe observed data.
+This improved ability may result in more accurate forecasts and facilitates the exploration of model assumptions.
+Based on this retrospective analysis, we offer suggestions on fitting mechanistic models to dynamic systems for future studies.
+
+While this work is focused on the models proposed by Lee et al.~\cite{lee20}, our suggestions have broader relevance.
+To investigate the extent to which \cite{lee20} is typical of the substantial body of work on Haiti cholera dynamics, we preformed a literature review by searching PubMed with keywords: Haiti, cholera, model.
+The search resulted in 66 papers, of which 32 used dynamic models to describe the cholera epidemic in Haiti.
+The models make various choices on the dichotomies considered by Lee et al.~\cite{lee20}; deterministic versus stochastic; compartment model versus agent-based; aggregated versus spatially explicit.
+This is no accident, since Lee et al.~\cite{lee20} purposefully designed their study to cover the range of current modeling practice.
+
+We proceed by introducing the general modeling scheme employed by Models~1--3 and provide details of each individual model;
+we then describe how each model is calibrated to data, and present a systematic approach to examining and refining these models.
+We then use improved model fits to project cholera incidence in Haiti under various vaccination scenarios considered by Lee et al.~\cite{lee20}.
+Finally, we conclude with a discussion of the results.
+% This is followed by a consideration of the robustness of model based policy recommendations in Sec.~\ref{sec:advice}, and a discussion in Sec.~\ref{sec:discussion}.
 
@@
 In the case of a POMP model, the likelihood function---as described in Eq.~\myeqref{eq:likedef}---is written as $\mathcal{L}(\paramVec) = \int f_{\bm{X}_{0:N}, \bm{Y}_{1:N}}\big(\bm{x}_{0:N}, \bm{y}_{1:N};\paramVec\big) d\bm{x}_{0:N}$.
-The high-dimensional integral in the likelihood, and the potentially non-linear nature of the process model, make it impossible to obtain a closed form expression of the likelihood function of POMP models except in the most basic cases.
+The high-dimensional integral in the likelihood, and the potentially non-linear nature of the process model, make it impossible to obtain a closed form expression of the likelihood function of POMP models except in the most simple cases.
 The process model that defines Model~1 was developed with the intent of providing a plausible description of cholera dynamics rather than a providing statistically convenient description.
@@
 where $\bar{t} = \frac{t - (t_N + t_0) / 2}{t_N - (t_N + t_0) / 2}$, so that $\bar{t} \in [-1, 1]$.
-The proposal of a linear trend in transmission is a result of observing an apparent decrease in reported cholera infections from 2012-2019 in Fig.~\ref{CholeraData}.
+The proposal of a linear trend in transmission is a result of observing an apparent decrease in reported cholera infections from 2012-2019 in Fig.~\ref{fig:CholeraData}.
 While several factors may contribute to this decrease, one explanation is that case-area targeted interventions (CATIs), which included education sessions, increased monitoring, household decontamination, soap distribution, and water chlorination in infected areas \cite{rebaudet19CATI}, may have substantially reduced cholera transmission over time \cite{rebaudet21}.
@@
 
-\begin{figure}
+\begin{figure}[!h]
 \centering
@@
 The large increase in the log-likelihood for Model~1 (see Table~\ref{tab:likes}) can primarily be attributed to increasing the computational effort in fitting the model; this result highlights the importance of carefully determining the necessary computational effort needed to maximize model likelihoods and acting accordingly.
-
-\begin{figure}
+Simulations from the initial conditions of our fitted model are plotted against the observed incidence data in Fig.~\ref{fig:mod1fit}.
+
+\begin{figure}[!h]
 \centering
@@
 
-\subsubsection*{Calibrating Model~3 Parameters}\label{sec:fit3}
-Model~3 describes cholera dynamics in Haiti using a metapopulation model, where the hidden states in each administrative department has an effect on the dynamics in other departments.
-The decision to address metapopulation dynamics using a spatially explicit model, rather than to aggregate over space, is double-edged.
-Evidence for the former approach has been provided in previous studies \cite{king15}, including the specific case of heterogeneity between Haitian departments in cholera transmission \cite{collins14}.
-However, a legitimate preference for simplicity can support a decision to consider nationally aggregated models \cite{saltelli20,green15}.
-One issue that arises when fitting spatially explicit models is that parameter estimation techniques based on the particle filter become computationally intractable as the number of spatial units increases.
-This is a result of the approximation error of particle filters growing exponentially in the dimension of the model \cite{rebeschini15,park20}.
-
-To avoid the approximation error present in high-dimensional models, Lee et al.~\cite{lee20} simplified the problem of estimating the parameters of Model~3 by creating an approximate version of the model where the units are independent given the observed data.
-Reducing a spatially coupled model to individual units in this fashion requires special treatment of any interactive mechanisms between spatial units, such as found in Eq.~\myeqref{eq:model3:foi}.
-Because the simplified, spatially-decoupled version of Model~3 implemented in \cite{lee20} relies on the observed cholera cases, the calibrated model cannot readily be used to obtain forecasts.
-Therefore, in order to obtain model forecasts, Lee et al.~\cite{lee20} used the parameters estimates from the spatially-decoupled approximation of Model~3 to obtain forecasts using the fully coupled version of the model.
-This approach of model calibration and forecasting avoids the issue of particle depletion, but may also be problematic.
-One concern is that cholera dynamics in department $u$ are highly related to the dynamics in the remaining departments;
-calibrating model parameters while conditioning on the observed cases in other departments may therefore lead to an over-dependence on observed cholera cases.
-Another concern is that the two versions of the model are not the same, resulting in sub-optimal parameter estimates for the spatially coupled model,
-as parameters that maximize the likelihood of the decoupled model almost certainly do not maximize the likelihood of the fully coupled model.
-These two concerns may explain the unrealistic forecasts and low likelihood of Model~3 from Lee et al.~\cite{lee20} (see Table~\ref{tab:likes}).
-
-At the time Lee et al.~\ref{lee20} was published, there was no known algorithm that could readily be used to maximize the likelihood of an arbitrary meta-population POMP model with coupled spatial dynamics, which justifies the spatial decoupling approximation that was used to calibrate model parameters.
-For our analysis, we calibrate the parameters of the spatially coupled version of Model~3 using the iterated block particle filter (IBPF) algorithm \cite{ionides22}.
-This algorithm extends the work of Ning and Ionides \cite{ning21ibpf}, who provided theoretic justification for the version of the algorithm that only estimates unit-specific parameters.
-The IBPF algorithm enables us to directly estimate the parameters of models describing high-dimensional partially-observed nonlinear dynamic systems via likelihood maximization.
-The ability to directly estimate parameters of Model~3 is responsible for the large increase in model likelihoods reported in Table~\ref{tab:likes}.
-Simulations from the fitted model are displayed in Fig.~\ref{fig:h3spatsims}.
-
-<<Model 3 SpatPOMP sims, echo=FALSE>>=
-
-# Number of simulations from Model 3. Because it's high-dimensional, we include
-# a run level to avoid time/memory issues when editing. Final results shown
-# with RUN_LEVEL = 3.
-
-h3_nsim <- switch(RUN_LEVEL, 20, 100, 500)
-
-# Get quantiles of simulations from the calibrated Model 3.
-h3_quants <- bake(
-  file = paste0('model3/', rl_dir, 'h3Sims.rds'), {
-
-    coef(h3Spat) <- p3
-    # Simulate from the fitted version of Model 3
-    h3_sims <- simulate(
-      h3Spat, nsim = h3_nsim, format = 'data.frame'
-    )
-
-    # For each department and time-point, calculate percentiles of the
-    # simulations
-    h3_sims %>%
-      rename(dep = unitname) %>%
-      group_by(dep, time) %>%
-      summarise(
-        q05 = quantile(cases, 0.025, na.rm = T),
-        mean = mean(cases, na.rm = T),
-        q50 = quantile(cases, 0.5, na.rm = T),
-        q95 = quantile(cases, 0.975, na.rm = T)
-      ) %>%
-      ungroup() %>%
-      mutate(
-        date = lubridate::date_decimal(time)
-      ) %>%
-      mutate(date = as.Date(lubridate::round_date(date, unit = 'day')))
-  },
-  timing = FALSE
-)
-
-@
-
-\begin{figure}
-<<Model3_SpatPOMP_sims_Figure, fig.height=3>>=
-
-# Plot the reported cases and confidence region for each department.
-ggplot() +
-  geom_line(data = h3_quants, aes(x = date, y = q50 + 1), col = 'blue') +
-  geom_line(data = dep_plot_df, aes(x = date, y = cases + 1), col = 'black') +
-  geom_ribbon(  # plot confidence region.
-    data = h3_quants,
-    aes(x = date, ymin = q05 + 1, ymax = q95 + 1),
-    alpha = 0.4
-  ) +
-  facet_wrap(
-    ~factor(dep, levels = plot_order), nrow = 2, labeller = dep_labeller
-  ) +
-  labs(y = 'Log Reported Cholera Cases') +
-  scale_x_date(  # Format the date on the X-axis.
-    date_labels = "'%y",
-    breaks = seq.Date(
-      from = as.Date("2011-01-01"),
-      to = as.Date("2019-01-01"),
-      by = '2 years'
-    )
-  ) +
-  theme(axis.title.x = element_blank()) +
-  scale_y_log10(  # Plot on log-scale
-    labels = scales::trans_format("log10", scales::math_format(10^.x))
-  )
-@
-\caption{\label{fig:h3spatsims}
-Simulations from initial conditions using the spatially coupled version of Model~3.
-The black curve represents true case count, the blue line the median of 500 simulations from the model, and the gray ribbons representing $95\%$ confidence interval.
-}
-\end{figure}
-
-<<Remove Model3 spatSims, cache=FALSE, include=FALSE, message=FALSE, echo=FALSE>>=
-rm(h3_quants, h3_nsim)
-gc()
-@
-
-\subsection*{Model Diagnostics}\label{sec:model_diagnostics}
-
-The goal of parameter calibration---whether done using Bayesian or frequentist methods---is to find the best description of the observed data under the assumptions of the model.
-Obtaining the best fitting set of parameters for a given model does not, however, guarantee that the model provides an accurate representation of the system under investigation.
-Model misspecification, which may be thought of as the omission of a mechanism in the model that is an important feature of the dynamic system, is inevitable at all levels of model complexity.
-To make progress, while accepting proper limitations, one must bear in mind the much-quoted observation of George Box \cite{box79} that ``all models are wrong but some are useful.''
-Beyond being good practical advice for applied statistics, this assertion is relevant for the philosophical justification of statistical inference as severe testing \cite{mayo18}.
-% \citep[][Sec.~4.8]{mayo18}.
-In this section, we discuss some tools for diagnosing mechanistic models with the goal of making the subjective assessment of model ``usefulness'' more objective.
-To do this, we will rely on the quantitative ability of the model to match the observed data, which we call the model's {\it goodness-of-fit}, with the guiding principle that a model which cannot adequately describe observed data may not be reliable for useful purposes.
-Goodness-of-fit may provide evidence supporting the causal interpretation of one model versus another, but cannot by itself rule out the possibility of alternative explanations.
-
-One common approach to assess a mechanistic model's goodness-of-fit is to compare simulations from the fitted model to the observed data.
-Visual inspection may indicate defects in the model, or may  suggest that the observed data are a plausible realization of the fitted model.
-While visual comparisons can be informative, they provide only a weak and informal measure of the goodness-of-fit of a model.
-The study by Lee et al.~\cite{lee20} provides an example of this: their models and parameter estimates resulted in simulations that visually resembled the observed data, yet resulted in model likelihoods that were considerably smaller than likelihoods that can be achieved via the likelihood based optimization techniques that were used (see Table~\ref{tab:likes}).
-Alternative forms of model validation should therefore be used in conjunction with visual comparisons of simulations to observed data.
-
-Another approach is to compare a quantitative measure of the model fit (such as MSE, predictive accuracy, or model likelihood) among all proposed models.
-These comparisons, which provide insight into how each model performs relative to the others, are quite common \cite{rinaldo12,sallah17}.
-To calibrate relative measures of fit, it is useful to compare against a model that has well-understood statistical ability to fit data, and we call this model a {\it benchmark}.
-Standard statistical models, interpreted as associative models without requiring any mechanistic interpretation of their parameters, provide suitable benchmarks.
-Examples include linear regression, auto-regressive moving average time series models, or even independent and identically distributed measurements.
-Benchmarks enable us to evaluate the goodness of fit that can be expected of a suitable mechanistic model.
-
-Associative models are not constrained to have a causal interpretation, and typically are designed with the sole goal of providing a statistical fit to data.
-Therefore, we should not require a candidate mechanistic model to beat all benchmarks.
-However, a mechanistic model which falls far short against benchmarks is evidently failing to explain some substantial aspect of the data.
-A convenient measure of fit should have interpretable differences that help to operationalize the meaning of far short.
-Ideally, the measure should also have favorable theoretical properties.
-Consequently, we focus on log-likelihood as a measure of goodness of fit, and we adjust for the degrees of freedom of the models to be compared by using the Akaike information criterion (AIC) \cite{aic74}.
-
-In some cases, a possible benchmark model could be a generally accepted mechanistic model, but often no such model is available.
-Because of this, we use a log-linear Gaussian ARMA model as an associative benchmark, as recommended by He et al.~\cite{he10}.
-The theory and practice of ARMA models is well developed, and these linear models are appropriate on a log scale due to the exponential growth and decay characteristic of biological dynamics.
-Likelihoods of Models~1--3 and their respective ARMA benchmark models are provided in Table~\ref{tab:likes}.
-
-It should be universal practice to present measures of goodness of fit for published models, and mechanistic models should be compared against benchmarks.
-In our literature review of the Haiti cholera epidemic, no non-mechanistic benchmark models were considered in any of the 32 papers that used dynamic models to describe cholera in order to obtain scientific conclusions.
-Including benchmarks would help authors and readers to detect and confront any major statistical limitations of the proposed mechanistic models.
-In addition, the published goodness of fit provides a concrete point of comparison for subsequent scientific investigations.
-When combined with online availability of data and code, objective measures of fit provide a powerful tool to accelerate scientific progress, following the paradigm of the {\it common task framework} \cite{donoho17}.
-% \citep[][Sec.~6]{donoho17}.
-
-The use of benchmarks may also be beneficial when developing models at differing spatial scales, where a direct comparison between model likelihoods is meaningless.
-In such a case, a benchmark model can be fit to each spatial resolution being considered, and each model compared to their respective benchmark.
-Large advantages (or shortcomings) in model likelihood relative to the benchmark for a given spatial scale that are not present in other spatial scales may provide weak evidence for (or against) the statistical fit of models across a range of spatial resolutions.
-
-<<Table 2 Input, echo=FALSE, eval=TRUE>>=
-NREPS_EVAL <- switch(RUN_LEVEL,   3,   8,    36)
-NP_EVAL    <- switch(RUN_LEVEL, 100, 500,  5000)
-
-# Objective function of model 2, epidemic phase
-epi_ofun <- traj_objfun(
-  h2_epi,
-  params = h2_fit$h2_params
-)
-
-mod2_epi_ll <- -epi_ofun(par = h2_fit$h2_params)
-mod2_ll <- mod2_epi_ll - sum(log(h2_epi@data + 1), na.rm = TRUE)
-
-arima_lik2 <- 0
-
-# model 2 AIC
-mod2_n_params <- h2_fit$n_fit_params
-mod2_aic <- 2 * mod2_n_params - 2 * mod2_ll
-
-# model 1 AIC
-mod1_n_params <- length(
-  c('beta1', 'beta2', 'beta3', 'beta4', 'beta5', 'beta6',
-    'rho', 'nu', 'E_0', 'I_0', 'tau_epi',
-    'tau_end', 'sig_sq_epi', 'sig_sq_end', 'betat')
-)
-mod1_aic <- 2 * mod1_n_params - 2 * mod1_ll
-
-# model 3 AIC
-mod3_aic <- 2 * mod3_n_params - 2 * mod3_ll
-
-# Calculate likelihood via block-particle filter
-mod3_eval_res <- bake(
-  file = paste0('model3/', rl_dir, 'mod3_eval_res.rds'), {
-
-    # Create a list to store results
-    results <- list()
-
-    coef(h3Spat) <- p3
-
-    # Perform block-particle filter
-    h3_bpf <- foreach(i = 1:NREPS_EVAL, .combine = c) %dopar% {
-      bpfilter(h3Spat, Np = NP_EVAL, block_size = 1)
-    }
-
-    ll <- logLik(h3_bpf)
-    results$ll <- logmeanexp(ll[!is.na(ll)], se = FALSE)
-
-    ### Get likelihood for subset of the data
-
-    # Find appropriate sub-set
-    in_subset_cols <- h3_bpf[[1]]@times >= as.numeric(lubridate::decimal_date(as.Date("2014-03-01")))
-    in_subset_rows <- h3_bpf[[1]]@unit_names != 'Ouest'
-
-    mod3_coupled_subset_evals <- c()
-    for (i in 1:length(h3_bpf)) {
-      mod3_coupled_subset_evals <- c(
-        mod3_coupled_subset_evals,
-        sum(h3_bpf[[i]]@block.cond.loglik[in_subset_rows, in_subset_cols])
-      )
-    }
-
-    results$subset_ll <- logmeanexp(mod3_coupled_subset_evals)
-    results
-  },
-  timing = FALSE
-)
-
-mod3_eval_res$subset_aic <- 2 * mod3_n_params - 2 * mod3_eval_res$subset_ll
-
-rm(NREPS_EVAL, NP_EVAL)
-@
-
-<<CalculateARMA, echo=FALSE, message=FALSE, include=FALSE, cache=TRUE>>=
-
-ARMA_benchmarks <- list()
-
-###
-### Model 1 ###
-###
-
-# Load aggregated data
-m1_agg_data <- haiti1_agg_data()
-
-# Fit ARMA(2, 1) on log-cases
-m1_log_arma <- arima(log(m1_agg_data$cases + 1), order = c(2, 0, 1))
-
-# Fit ARMA(2, 1) on natural scale of cases
-m1_arma <- arima(m1_agg_data$cases, order = c(2, 0, 1))
-
-# Check which one is better, and save benchmark as the better one
-if (m1_log_arma$loglik - sum(log(m1_agg_data$cases + 1), na.rm = TRUE) >= m1_arma$loglik) {
-  ARMA_benchmarks[['m1']] <- m1_log_arma$loglik - sum(log(m1_agg_data$cases + 1), na.rm = TRUE)
-} else {
-  ARMA_benchmarks[['m1']] <- m1_arma$loglik
-}
-
-rm(m1_agg_data, m1_log_arma, m1_arma)
-gc()
-
-###
-### Model 2 ###
-###
-
-h2_data <- haiti2_data(recreate = FALSE)
-
-# model 2 ARIMA benchmark
-h2_dep_names <- h2_data$department %>% unique()
-m2_benchmark <- 0
-for (i in 1:10){
-  cases <- h2_data[h2_data$department == h2_dep_names[i], ]$cases
-  log_cases <- log(cases + 1)
-
-  m2_arma_dep_ll <- arima(cases, order=c(2, 0, 1))$loglik
-  m2_log_arma_dep_ll <- arima(log_cases, order=c(2, 0, 1))$loglik - sum(log_cases, na.rm = TRUE)
-
-  m2_benchmark <- m2_benchmark + max(m2_arma_dep_ll, m2_log_arma_dep_ll)
-}
-
-ARMA_benchmarks[['m2']] <- m2_benchmark
-
-rm(m2_benchmark, h2_dep_names, cases, log_cases, m2_arma_dep_ll, m2_log_arma_dep_ll, h2_data)
-gc()
-
-###
-### Model 3 ###
-###
-
-mod3_cases <- as.data.frame(t(h3Spat@data))
-
-m3_benchmark <- 0
-for (dep in 1:10) {
-
-  # Get department cases
-  cases <- as.numeric(mod3_cases[, dep])
-
-  # Fit model on natural scale of cases
-  m3_a201 <- arima(cases, order = c(2, 0, 1))
-
-  # Fit model on log-scale of cases
-  m3_log_a201 <- arima(log(cases + 1), order = c(2, 0, 1))
-
-  m3_benchmark <- m3_benchmark + max(
-    m3_log_a201$loglik - sum(log(cases + 1), na.rm = TRUE),
-    m3_a201$loglik
-  )
-}
-
-ARMA_benchmarks[['m3']] <- m3_benchmark
-rm(cases, m3_a201, m3_log_a201, m3_benchmark, dep,
-   mod3_cases, temp, jumps, MODEL3_CASES, i, j)
-gc()
-@
-
-<<Table 2 Lee Input, echo=FALSE, include=FALSE>>=
-mod1_lee_n_params <- 20
-mod1_lee_aic <- 2 * mod1_lee_n_params - 2 * mod1_lee_ll
-
-mod2_lee_aic <- 2 * h2_fit$leeFit_n_params - 2 * h2_fit$leeFit_ll
-@
-
-<<Lee3 Sub Calculations, echo=FALSE, include=FALSE, message=FALSE>>=
-lee3_NP   <- switch(RUN_LEVEL, 50, 250, 5000)
-lee3_NREP <- switch(RUN_LEVEL,  3,   8,   36)
-
-registerDoRNG(49637421)
-
-# Create Lee et al (2020a) version of model 3 on sub-data
-lee3_sub <- lee3_spatPomp()
-
-# This model is only fit for a subset of the data, so we will set data
-# that are not part of this subset as NA. This will allow us to estimate
-# the likelihood for only the values that were considered during the
-# parameter estimation phase.
-# set_as_na <- lee3_sub@times < lubridate::decimal_date(as.Date('2017-06-10'))
-# temp_ouest_cases <- lee3_sub@data['casesOuest', ]
-# temp_ouest_cases[set_as_na] <- NA_real_
-# lee3_sub@data['casesOuest', ] <- temp_ouest_cases
-
-lee3_all <- lee3_spatPomp(start_date = "2010-10-23")
-
-lee3_results <- bake(
-  file = paste0('model3/', rl_dir, 'lee3_results.rds'), {
-
-    # Create lists to save results
-    lee3_results <- list()
-
-    # Save how long it takes for sub-data
-    lee3_results[["sub_bpf_time"]] <- system.time({
-
-      lee3_sub_ll_evals <- foreach(
-        i = 1:lee3_NREP, .combine = rbind
-      ) %dopar% {
-
-        liks <- numeric(10L)
-
-        lee_bpf_out <- bpfilter(lee3_sub, block_size = 1, Np = lee3_NP)
-        Ouest_ind <- which(lee_bpf_out@unit_names == "Ouest")
-
-        for (u in 1:10) {
-          if (u == Ouest_ind) {
-            liks[u] <- sum(lee_bpf_out@block.cond.loglik[u, lee3_sub@times >= lubridate::decimal_date(as.Date('2017-06-10'))])
-          } else {
-            liks[u] <- sum(lee_bpf_out@block.cond.loglik[u, ])
-          }
-        }
-
-        liks
-      }
-    })
-
-    # Save all likelihood evaluations
-    lee3_results[["sub_ll_eval"]] <- logmeanexp(rowSums(lee3_sub_ll_evals)[!is.na(rowSums(lee3_sub_ll_evals))])
-
-    registerDoRNG(2789131)
-
-    # Save how long it takes for all data
-    lee3_results[["all_pf_time"]] <- system.time({
-      # Save all likelihood evaluations
-
-      lee3_all_ll_evals <- foreach(
-        i = 1:lee3_NREP, .combine = c
-      ) %dopar% {
-        logLik(bpfilter(lee3_all, block_size = 1, Np = lee3_NP))  # Get likelihood evaluations
-      }
-
-      lee3_results[["all_ll_eval"]] <- logmeanexp(lee3_all_ll_evals[!is.na(lee3_all_ll_evals)])
-
-    })
-
-    lee3_results[["n_params"]] <- 29
-
-    lee3_results
-  },
-  timing = FALSE
-)
-
-rm(lee3_sub, lee3_all, lee3_NP, lee3_NREP)
-gc()
-
-mod3_lee_coupled_aic <- 2 * lee3_results$n_params - 2 * lee3_results$all_ll_eval
-
-mod3_lee_coupled_partial_aic <- 2 * lee3_results$n_params - 2 * lee3_results$sub_ll_eval
-@
-
-\begin{table}
-\centering
-
-\begin{tabular}{|c|c|c|c|}
-\hline
- & \thead{Model~1} & \thead{Model~2} & \thead{Model~3}
-\\
-\hline
-\hline
-\multirow{2}{*}{Log-likelihood} &
-  $\Sexpr{myround(mod1_ll, 1)}$ &
-  $\Sexpr{myround(mod2_ll, 1)}$ &
-  $\Sexpr{myround(mod3_ll, 1)}$ \\
-    & ($\Sexpr{myround(mod1_lee_ll, 1)}$)\footnotemark[1] &
-  ($\Sexpr{myround(h2_fit$leeFit_ll, 1)}$) &
-  ($\Sexpr{myround(lee3_results[["all_ll_eval"]], 1)}$)\footnotemark[2]
-\\
-\hline
-Number of &
-  $\Sexpr{as.character(mod1_n_params)}$ &
-  $\Sexpr{as.character(mod2_n_params)}$ &
-  $\Sexpr{as.character(mod3_n_params)}$ \\
- Fit Parameters &
- (\Sexpr{as.character(mod1_lee_n_params)}) &
- ($\Sexpr{as.character(h2_fit$leeFit_n_params)}$) &
- ($\Sexpr{as.character(lee3_results$n_params, 1)}$)
-\\
-\hline
-\multirow{2}{*}{AIC} &
-  $\Sexpr{myround(mod1_aic, 1)}$ &
-  $\Sexpr{myround(mod2_aic, 1)}$ &
-  $\Sexpr{myround(mod3_aic, 1)}$ \\
-  & ($\Sexpr{myround(mod1_lee_aic, 1)}$)\footnotemark[1] &
-  ($\Sexpr{myround(mod2_lee_aic, 1)}$) &
-  ($\Sexpr{myround(mod3_lee_coupled_aic, 1)}$)\footnotemark[2]
-\\
-\hline
-\thead{Log-ARMA(2,1) \\ AIC} &
-  $\Sexpr{myround(-2 * ARMA_benchmarks$m1 + 2 * 5, 1)}$ &
-  $\Sexpr{myround(-2 * ARMA_benchmarks$m2 + 2 * 5, 1)}$ &
-  $\Sexpr{myround(-2 * ARMA_benchmarks$m3 + 2 * 5, 1)}$
-\\
-\hline
-\end{tabular}
-\caption{\label{tab:likes}AIC values for each model compared to their ARMA benchmark.
-Values in parentheses are corresponding values using \cite{lee20} parameter estimates.
-$\textsuperscript{1}$The reported likelihood is an upper bound of the likelihood of the \cite{lee20} model.
-$\textsuperscript{2}$\cite{lee20} fit Model~3 to a subset of the data (March 2014 onward, excluding data from Ouest in 2015-2016).
-On this subset, their model has a likelihood of $\Sexpr{myround(lee3_results$sub_ll_eval, 1)}$.
-On this same subset, our model has a likelihood of $\Sexpr{myround(mod3_eval_res$subset_ll, 1)}$.
-See Sec.~S6 of the supplement material for more details on estimating the likelihood of the Lee et al.~\cite{lee20} models.
-}
-\end{table}
-
-Comparing model log-likelihoods to a suitable benchmark may not be sufficient to identify all the strengths and weaknesses of a given model.
-Additional techniques include the inspection of conditional log-likelihoods of each observation given the previous observations in order to understand how well the model describes each data point (Sec.~S5).
-Other tools include plotting the effective sample size of each observation \cite{liu01}; plotting the values of the hidden states from simulations (supplement Sec.~5.1);
-and comparing summary statistics of the observed data to simulations from the model \cite{wood10,king15}.
-
-\subsection*{Corroborating Fitted Models with Scientific Knowledge}
-
-The resulting mechanisms in a fitted model can be compared to current scientific knowledge about a system.
-Agreement between model based inference and our current understanding of a system may be taken as a confirmation of both model based conclusions and our scientific understanding.
-On the other hand, comparisons may generate unexpected results that have the potential to spark new scientific knowledge \cite{ganusov16}.
-
-In the context of our case study, we demonstrate how the fit of Model~1 corroborates other evidence concerning the role of rainfall in cholera epidemics.
-Specifically, we examine the results of fitting the flexible cubic spline term in Model~1 (Eqs.~\myeqref{model1:lambda}--\myeqref{model1:beta}).
-The cubic splines permit flexible estimation of seasonality in the force of infection, $\transmission(t)$.
-Fig.~\ref{fig:h1SeasRain} shows that the estimated seasonal transmission rate $\transmission$ mimics the rainfall dynamics in Haiti, despite Model~1 not having access to rainfall data.
-This is consistent with previous studies finding that rainfall played an important role in cholera transmission in Haiti \cite{lemaitre19,eisenberg13}.
-The estimated seasonality also features an increased transmission rate during the fall, which was noticed at an earlier stage of the epidemic \cite{rinaldo12}; the high transmission rate in the fall may also be a result of the increase transmission that occurred in the fall of 2016, when hurricane Matthew struck Haiti \cite{ferreirai16}.
-
-<<Get Model 1 Seasonality>>=
-
-h1_seas_df <- get_h1_seasonality(h1)
-
-df_rain <- haitiRainfall %>%
-  pivot_longer(  # Convert data to "long" format
-    data = .,
-    cols = -date,
-    names_to = 'dep',
-    values_to = 'rainfall'
-  ) %>%
-  mutate(year = lubridate::year(date), week = lubridate::week(date)) %>%
-  group_by(year, week) %>%
-  summarize(nwr = sum(rainfall)) %>%
-  ungroup() %>%  # Not needed, but good practice
-  mutate(
-    date_week_start = lubridate::ymd(paste0(2014, "-01-01")) +
-      lubridate::weeks(week - 1),
-    week_date = as.Date(date_week_start, format = "%m-%d"),
-    std_national_rain = (nwr - min(nwr)) / (max(nwr) - min(nwr))
-  )
-
-# Calculate the average national weekly rainfall.
-mean_rain <- df_rain %>%
-  group_by(week_date) %>%
-  summarize(mean_rain = mean(std_national_rain)) %>%
-  ungroup() %>%
-  mutate(mean_rain = (mean_rain - min(mean_rain)) / ((max(mean_rain) - min(mean_rain))))
-
-# Create a figure showing annual rainfal, scaled from 0-1. This is top figure
-gg_rain <- df_rain %>%
-  filter(week != 1, week != max(week)) %>%
-  ggplot(aes(x = week_date, y = std_national_rain, group = year, col = year)) +
-  geom_line() +
-  theme_bw() +
-  guides(color = 'none') +
-  ylab('Standardized Weekly\nNational Rain') +
-  theme(  # Remove elements that are un-unsed becuase this is top figure
-    axis.title.x = element_blank(),
-    axis.ticks.x = element_blank(),
-    axis.text.x = element_blank(),
-    axis.text.y = element_text(size = 7.5),
-    axis.title.y = element_text(size = 9)
-  )
-
-# Create plot that shows the seasonality of Model 1, alongside the average
-# weekly national rainfall.
-gg_trans <- ggplot() +
-  geom_line(data = h1_seas_df, aes(x = date, y = trans_std), linetype = 'dashed') +
-  geom_line(
-    data = mean_rain,
-    aes(x = week_date + lubridate::years(6), y = mean_rain),
-    col = '#2166ac'
-  ) +
-  scale_x_date(date_labels = "%b", date_breaks = '1 month') +
-  ylab('Standardized Seasonal\nContact Rate') +
-  theme(axis.title.x = element_blank(),
-        axis.text = element_text(size = 7.5),
-        axis.title.y = element_text(size = 9))
-
-@
-
-\begin{figure}
-\centering
-<<Model1_Seasonality_Figure, fig.height=2.9, fig.width=5, fig.align='center'>>=
-cowplot::plot_grid(gg_rain, gg_trans, align = "v", ncol = 1)
-@
-\caption{\label{fig:h1SeasRain}
-(Top) weekly rainfall in Haiti, lighter colors representing more recent years.
-(Bottom) estimated seasonality in the transmission rate (dashed line) plotted alongside mean rainfall (solid line).
-The outsided effect of rainfall in the fall may be due to Hurricane Matthew, which struck Haiti in October of 2016 and resulted in an increase of cholera cases in the nation.
-}
-\end{figure}
-
-For any model-based inference, it is important to recognize and assess the modeling simplifications and assumptions that were used in order to arrive at the conclusions.
-In epidemiological studies, for example, quantitative understanding of individual-level processes may not perfectly match model parameters that were fit to population-level case counts, even when the model provides a strong statistical fit \cite{he10}.
-This makes direct interpretation of estimated parameters delicate.
-
-Our case study provides an example of this in the parameter estimate for the duration of natural immunity due to cholera infection, $\muRS^{-1}$.
-Under the framework of Model~2, the best estimate for this parameter is $\Sexpr{signif(1/h2_fit$h2_params["sigma"],3)} \, \mathrm{yr}$, suggesting that individuals have effectively permanent immunity to cholera once infected.
-To interpret this result, we bear in mind that the data ranged from 2010-2019, and therefore estimates of immunity longer than $10\, \mathrm{yr}$---the upper end of previous estimates of natural immunity \cite{king08}---effectively result in the same model dynamics.
-The depletion of susceptible individuals may also be attributed to confounding mechanisms---such as localized vaccination programs and non-pharmaceutical interventions that reduce cholera transmission \cite{trevisin22, rebaudet21}---that were not accounted for in the model.
-Perhaps the best interpretation of the estimated parameter, then, is that under model assumptions, the model most adequately describes the observed data by having a steady decrease in the number of susceptible individuals.
-The weak statistical fit of Model~2 compared to a log-linear benchmark (see Table~\ref{tab:likes}) cautions us against drawing quantitative conclusions from this model.
-A model that has a poor statistical fit may nevertheless provide a useful conceptual framework for thinking about the system under investigation.
-However, a claim that the model has been validated against data should be reserved for situations where the model provides a statistical fit that is competitive against alternative explanations.
-
-\subsection*{Robust interpretation of model based conclusions}\label{sec:advice}
-
-A model which aspires to provide quantitative guidance for assessing interventions should provide a quantitative statistical fit for available data.
-However, strong statistical fit does not guarantee a correct causal structure: it does not even necessarily require the model to assert a causal explanation.
-A causal interpretation is strengthened by corroborative evidence.
-For example, reconstructed latent variables (such as numbers of susceptible and recovered individuals) should make sense in the context of alternative measurements of these variables \cite{grad12}.
-In the context of our case study, by reconstructing the latent states of Model~3, we notice that the calibrated model favors higher levels of cholera transmission than what was typically observed in the incidence data (Sec.~S5.1 of the supplement).
-This result hints at the possibility of model mispecification, and should prompt us to be sceptical of the reliability of forecasts from this model.
-Similarly, parameters that have been calibrated to data should make sense in the context of alternative lines of evidence about the phenomena being modeled, while making allowance for the possibility that the interpretations of parameters may vary when modeling across differing spatial scales.
-
-If a mechanistic model including a feature (such as a representation of a mechanism, or the inclusion of a covariate) fits better than mechanistic models without that feature, and also has competitive fit compared to associative benchmarks, this may be taken as evidence supporting the scientific relevance of the feature.
-As for any analysis of observational data, we must be alert to the possibility of confounding.
-For a covariate, this shows up in a similar way to regression analysis: the covariate under investigation could be a proxy for some other unmodeled or unmeasured covariate.
-For a mechanism, the model feature could in principle explain the data by helping to account for some different unmodeled phenomenon.
-In the context of our analysis, the statistical evidence of a trend in transmission rate in Model~1 could be explained by any trending variable (such as hygiene improvements, or changes in population behavior), resulting in confounding from colinear covariates.
-Alternatively, the trend could be attributed to a decreasing reporting rate rather than decreasing transmission rate, resulting in confounded mechanisms.
-The robust statistical conclusion is that a model which allows for change fits better than one which does not---we argue that a decreasing transmission rate is a plausible way to explain this, but the incidence data themselves do not provide enough information to pin down the mechanism.
-
-% Place figure captions after the first paragraph in which they are cited.
-\begin{figure}[!h]
-\caption{{\bf Bold the figure title.}
-Figure caption text here, please use this space for the figure panel descriptions instead of using subfigure commands. A: Lorem ipsum dolor sit amet. B: Consectetur adipiscing elit.}
-\label{fig1}
-\end{figure}
-
-% Results and Discussion can be combined.
-\section*{Results}
-
-\subsection*{Forecasts}\label{sec:filter}
-
-Forecasts are an attempt to provide an accurate estimate of the future state of a system based on currently available data, together with an assessment of uncertainty.
-Forecasts from mechanistic models that are compatible with current scientific understanding may also provide estimates of the future effects of potential interventions.
-Further, they may enable real-time testing of new scientific hypotheses \cite{lewis22}.
-
-Recent information about a dynamic system should be more relevant for a forecast than older information.
-This assertion may seem self-evident, but it is not the case for deterministic models, for which the initial conditions together with the parameters are sufficient for forecasting, and so recent data do not have special importance.
-Epidemiological forecasts based on deterministic models are not uncommon in practice, despite their limitations \cite{king15}.
-That may explain why Lee et al.~\cite{lee20} chose to obtain forecasts by simulating the calibrated models forward from initial conditions.
-Here, we obtain forecasts by simulating future values using latent states that are consistent with the most recent observations by drawing latent states at time the last observation time $t_N$ from the filtering distribution $f_{\bm{X}_N|\bm{Y}_{1:N}}(\bm{x}_{N} | \bm{y}^*_{1:N} ; \hat\paramVec)$,
-where $\hat\paramVec$ is a vector of calibrated parameters.
-The decision not to do this partially explains the unsuccessful forecasts of Lee et al.~\cite{lee20}: their Table~S7 shows that the subset of their simulations which were consistent with observing zero cases in 2019 also accurately predicted the prolonged absence of detected cholera.
-
-Uncertainty in just a single parameter can lead to drastically different forecasts \cite{saltelli20}.
-Therefore, parameter uncertainty should also be considered when obtaining model forecasts to influence policy.
-If a Bayesian technique is used for parameter estimation, a natural way to account for parameter uncertainty is to obtain simulations from the model where each simulation is obtained using parameters drawn from the estimated posterior distribution.
-For frequentist inference, one possible approach is obtaining model forecasts from various values of $\paramVec$, where the values of $\paramVec$ are sampled proportionally according to their corresponding likelihoods \cite{king15}; see Sec.~S7 of the supplement material for more details.
-Both of these approaches share the similarity that parameters are chosen for the forecast approximately in proportion to their corresponding value of the likelihood function, $f_{\bm{Y}_{1:N}}(\bm{y}_{1:N}^*; \paramVec)$.
-In this analysis, we do not construct forecasts accounting for parameter uncertainty as our focus is on the estimation and diagnosis of mechanistic models.
-Furthermore, we use the projections from a single point estimate to highlight the deficiency of deterministic models that the only variability in model projections is a result of parameter uncertainty, which can lead to over-confidence in forecasts \cite{king15}.
-
-The primary forecasting goal of Lee et al.~\cite{lee20} was to investigate the potential consequences of vaccination interventions on a system to inform policy.
-One outcome of their study include estimates for the probability of cholera elimination under several possible vaccination scenarios.
-Mimicking their approach, we define cholera elimination as an absence of cholera infections for at least 52 consecutive weeks, and we provide forecasts under the following vaccination scenarios:
-
-\begin{itemize}
-  \item[$V0$:] No additional vaccines are administered.
-  \item[$V1$:] Vaccination limited to the departments of Centre and Artibonite, deployed over a two-year period.
-  \item[$V2$:] Vaccination limited to three departments: Artibonite, Centre, and Ouest deployed over a two-year period.
-  \item[$V3$:] Countrywide vaccination implemented over a five-year period.
-  \item[$V4$:] Countrywide vaccination implemented over a two-year period.
-\end{itemize}
-
-<<Load Model 1 VaccScen Sims, message=FALSE, include=FALSE>>=
-
-# Number of particles and simulations to use for pfilter and project from filter, respectively.
-h1_scen_NP   <- switch(RUN_LEVEL, 50, 500, 5000)
-h1_scen_sims <- switch(RUN_LEVEL, 20, 100, 1000)
-
-# Simulate vaccination scenario from filtering distribution. The code
-# below creates a model, samples from the filtering distribution, and then
-# projects the vaccination scenario, under the specified model.
-h1_S0_results <- bake(
-  file = paste0("model1/", rl_dir, "S0_sims.rds"), {
-
-    # Create the model
-    s0 <- haiti1_joint(vacscen = 'id0')
-
-    # Load coefficients
-    coef(s0) <- p1
-
-    # Get samples from filtering distribution
-    s0_pf <- pfilter(s0, save.state = TRUE, Np = h1_scen_NP)
-    end_states <- saved_states(s0_pf)
-    end_states <- end_states[[length(end_states)]]
-
-    # project vacc scenario, with filtering dist, and model.
-    h1_S0_sims <- project_from_filter(mod = s0, end_states = end_states, nsims = h1_scen_sims)
-
-    results <- list()
-    results$mod1_V0_sims <- agg_mod1_sims(h1_S0_sims)
-    results$mod1_V0_probs <- get_elimProbs(h1_S0_sims, model = 1)
-
-    results
-  },
-  timing = FALSE
-)
-
-# Simulate vaccination scenario from filtering distribution. The code
-# below creates a model, samples from the filtering distribution, and then
-# projects the vaccination scenario, under the specified model.
-h1_S2_results <- bake(
-  file = paste0('model1/', rl_dir, 'S2_sims.rds'), {
-
-    # Create the model
-    s2 <- haiti1_joint(vacscen = "id2")
-
-    # Set model parameters that need to be adjusted
-    depts <- 2
-    h1_par_names <- names(p1)
-    h1_params_temp <- c(p1, rep(0.0, 5 * depts))
-    names(h1_params_temp) <- c(
-      h1_par_names,
-      paste0("S", 1:depts, "_0"),
-      paste0("E", 1:depts, "_0"),
-      paste0("I", 1:depts, "_0"),
-      paste0("A", 1:depts, "_0"),
-      paste0("R", 1:depts, "_0")
-    )
-    coef(s2) <- h1_params_temp  # Set model parameters
-
-    # Get samples from filtering distribution for Scenario 2 model:
-    s2_pf <- pfilter(s2, save.state = TRUE, Np = h1_scen_NP)
-    end_states <- saved_states(s2_pf)
-    end_states <- end_states[[length(end_states)]]
-
-    # project vacc scenario, with filtering dist, and model.
-    h1_S2_sims <- project_from_filter(mod = s2, end_states = end_states, nsims = h1_scen_sims)
-
-    results <- list()
-    results$mod1_V1_sims <- agg_mod1_sims(h1_S2_sims)
-    results$mod1_V1_probs <- get_elimProbs(h1_S2_sims, model = 1)
-
-    results
-  },
-  timing = FALSE
-)
-
-# Simulate vaccination scenario from filtering distribution. The code
-# below creates a model, samples from the filtering distribution, and then
-# projects the vaccination scenario, under the specified model.
-h1_S4_results <- bake(
-  file = paste0('model1/', rl_dir, 'S4_sims.rds'), {
-
-    # Create the model
-    s4 <- haiti1_joint(vacscen = "id4")
-
-    # Set model parameters that need to be adjusted
-    depts <- 3
-    h1_par_names <- names(p1)
-    h1_params_temp <- c(p1, rep(0.0, 5 * depts))
-    names(h1_params_temp) <- c(
-      h1_par_names,
-      paste0("S", 1:depts, "_0"),
-      paste0("E", 1:depts, "_0"),
-      paste0("I", 1:depts, "_0"),
-      paste0("A", 1:depts, "_0"),
-      paste0("R", 1:depts, "_0")
-    )
-    coef(s4) <- h1_params_temp  # Set model parameters
-
-    # Get samples from filtering distribution for Scenario 2 model:
-    s4_pf <- pfilter(s4, save.state = TRUE, Np = h1_scen_NP)
-    end_states <- saved_states(s4_pf)
-    end_states <- end_states[[length(end_states)]]
-
-    # project vacc scenario, with filtering dist, and model.
-    h1_S4_sims <- project_from_filter(mod = s4, end_states = end_states, nsims = h1_scen_sims)
-
-    results <- list()
-    results$mod1_V2_sims <- agg_mod1_sims(h1_S4_sims)
-    results$mod1_V2_probs <- get_elimProbs(h1_S4_sims, model = 1)
-
-    results
-  },
-  timing = FALSE
-)
-
-# Simulate vaccination scenario from filtering distribution. The code
-# below creates a model, samples from the filtering distribution, and then
-# projects the vaccination scenario, under the specified model.
-h1_S3_results <- bake(
-  file = paste0('model1/', rl_dir, 'S3_sims.rds'), {
-
-    # Create the model
-    s3 <- haiti1_joint(vacscen = "id3")
-
-    # Set model parameters that need to be adjusted
-    depts <- 10
-    h1_par_names <- names(p1)
-    h1_params_temp <- c(p1, rep(0.0, 5 * depts))
-    names(h1_params_temp) <- c(
-      h1_par_names,
-      paste0("S", 1:depts, "_0"),
-      paste0("E", 1:depts, "_0"),
-      paste0("I", 1:depts, "_0"),
-      paste0("A", 1:depts, "_0"),
-      paste0("R", 1:depts, "_0")
-    )
-    coef(s3) <- h1_params_temp  # Set model parameters
-
-    # Get samples from filtering distribution for Scenario 2 model:
-    s3_pf <- pfilter(s3, save.state = TRUE, Np = h1_scen_NP)
-    end_states <- saved_states(s3_pf)
-    end_states <- end_states[[length(end_states)]]
-
-    # project vacc scenario, with filtering dist, and model.
-    h1_S3_sims <- project_from_filter(mod = s3, end_states = end_states, nsims = h1_scen_sims)
-
-    results <- list()
-    results$mod1_V3_sims <- agg_mod1_sims(h1_S3_sims)
-    results$mod1_V3_probs <- get_elimProbs(h1_S3_sims, model = 1)
-
-    results
-  },
-  timing = FALSE
-)
-
-# Simulate vaccination scenario from filtering distribution. The code
-# below creates a model, samples from the filtering distribution, and then
-# projects the vaccination scenario, under the specified model.
-h1_S1_results <- bake(
-  file = paste0('model1/', rl_dir, 'S1_sims.rds'), {
-
-    # Create the model
-    s1 <- haiti1_joint(vacscen = "id1")
-
-    # Set model parameters that need to be adjusted
-    depts <- 10
-    h1_par_names <- names(p1)
-    h1_params_temp <- c(p1, rep(0.0, 5 * depts))
-    names(h1_params_temp) <- c(
-      h1_par_names,
-      paste0("S", 1:depts, "_0"),
-      paste0("E", 1:depts, "_0"),
-      paste0("I", 1:depts, "_0"),
-      paste0("A", 1:depts, "_0"),
-      paste0("R", 1:depts, "_0")
-    )
-    coef(s1) <- h1_params_temp  # Set model parameters
-
-    # Get samples from filtering distribution for Scenario 2 model:
-    s1_pf <- pfilter(s1, save.state = TRUE, Np = h1_scen_NP)
-    end_states <- saved_states(s1_pf)
-    end_states <- end_states[[length(end_states)]]
-
-    # project vacc scenario, with filtering dist, and model.
-    h1_S1_sims <- project_from_filter(mod = s1, end_states = end_states, nsims = h1_scen_sims)
-
-    results <- list()
-    results$mod1_V4_sims <- agg_mod1_sims(h1_S1_sims)
-    results$mod1_V4_probs <- get_elimProbs(h1_S1_sims, model = 1)
-
-    results
-  },
-  timing = FALSE
-)
-@
-
 <<Load Model 2 VaccScenarios, message=FALSE, include=FALSE>>=
@@
 
-<<Load Model 3 Vacc Scenarios, message=FALSE, include=FALSE>>=
-NP_BPF <- switch(RUN_LEVEL, 50, 200, 2000)
-NSIM   <- switch(RUN_LEVEL, 10,  50, 1000)
-
-h3_bpf_end_states <- bake(
-  file = paste0("model3/", rl_dir, "h3_bpf_end_states.rds"),
-  expr = {
-    coef(h3Spat) <- p3
-    h3_bpf <- bpfilter(h3Spat, Np = NP_BPF, block_size = 1, save_states = TRUE)
-    ss <- h3_bpf@saved.states
-    ss[[length(ss)]]
-  },
-  timing = FALSE
-)
-
-mod3_V0_res <- bake(
-  file = paste0("model3/", rl_dir, "mod3_V0_res.rds"), {
-
-    # Create a list to store results
-    h3_V0_res <- list()
-
-    # Get "noVacc" scenario parameters
-    h3_S0_par <- get_model3_vacc_scenario_params(scenario = "noVacc")
-    coef(h3Spat) <- p3
-    coef(h3Spat)[names(h3_S0_par)] <- h3_S0_par
-
-    # Using parameters and filtered distribution, project system
-    h3_V0_sims <- project_from_filter(
-      h3Spat, end_states = h3_bpf_end_states, covarGen = project_rain,
-      nsims = NSIM, seed = 6897154
-    )
-
-    # Store nationally aggregated results
-    h3_V0_res$mod3_V0_sims <- agg_mod3_sims(h3_V0_sims)
-    h3_V0_res$mod3_V0_probs <- get_elimProbs(h3_V0_sims, model = 3)
-
-    # For the supplement material, save quantiles of the Susceptible individuals
-    h3_V0_res$SQuants <- h3_V0_sims %>%
-      mutate(
-        date = as.Date(lubridate::round_date(lubridate::date_decimal(time), unit = 'day'))
-      ) %>%
-      group_by(unitname, date) %>%
-      mutate(
-        pop = h3Spat@params[paste0("H", which(first(unitname) == unit_names(h3Spat)))]
-      ) %>%
-      summarize(
-        Q025 = quantile((S + VSd + VSdd + VSdd_alt + VSd_alt) / pop, probs = 0.025),
-        Q50 = quantile((S + VSd + VSdd + VSdd_alt + VSd_alt) / pop, probs = 0.5),
-        Q975 = quantile((S + VSd + VSdd + VSdd_alt + VSd_alt) / pop, probs = 0.975)
-      ) %>%
-      ungroup() %>%
-      rename(dep = unitname)
-
-    # Return results
-    h3_V0_res
-
-  }
-)
-
-mod3_V1_res <- bake(
-  file = paste0("model3/", rl_dir, "mod3_V1_res.rds"), {
-
-    # Create a list to store results
-    h3_V1_res <- list()
-
-    # Get "noVacc" scenario parameters
-    h3_V1_par <- get_model3_vacc_scenario_params(scenario = "2dep")
-    coef(h3Spat) <- p3
-    coef(h3Spat)[names(h3_V1_par)] <- h3_V1_par
-
-    # Using parameters and filtered distribution, project system
-    h3_V1_sims <- project_from_filter(
-      h3Spat, end_states = h3_bpf_end_states, covarGen = project_rain,
-      nsims = NSIM, seed = 6897154
-    )
-
-    # Store nationally aggregated results
-    h3_V1_res$mod3_V1_sims <- agg_mod3_sims(h3_V1_sims)
-    h3_V1_res$mod3_V1_probs <- get_elimProbs(h3_V1_sims, model = 3)
-
-    # Return results
-    h3_V1_res
-
-  }
-)
-
-mod3_V2_res <- bake(
-  file = paste0("model3/", rl_dir, "mod3_V2_res.rds"), {
-
-    # Create a list to store results
-    h3_V2_res <- list()
-
-    # Get "noVacc" scenario parameters
-    h3_V2_par <- get_model3_vacc_scenario_params(scenario = "3dep")
-    coef(h3Spat) <- p3
-    coef(h3Spat)[names(h3_V2_par)] <- h3_V2_par
-
-    # Using parameters and filtered distribution, project system
-    h3_V2_sims <- project_from_filter(
-      h3Spat, end_states = h3_bpf_end_states, covarGen = project_rain,
-      nsims = NSIM, seed = 6897154
-    )
-
-    # Store nationally aggregated results
-    h3_V2_res$mod3_V2_sims <- agg_mod3_sims(h3_V2_sims)
-    h3_V2_res$mod3_V2_probs <- get_elimProbs(h3_V2_sims, model = 3)
-
-    # Return results
-    h3_V2_res
-  }
-)
-
-mod3_V3_res <- bake(
-  file = paste0("model3/", rl_dir, "mod3_V3_res.rds"), {
-
-    # Create a list to store results
-    h3_V3_res <- list()
-
-    # Get "noVacc" scenario parameters
-    h3_V3_par <- get_model3_vacc_scenario_params(scenario = "slowNation")
-    coef(h3Spat) <- p3
-    coef(h3Spat)[names(h3_V3_par)] <- h3_V3_par
-
-    # Using parameters and filtered distribution, project system
-    h3_V3_sims <- project_from_filter(
-      h3Spat, end_states = h3_bpf_end_states, covarGen = project_rain,
-      nsims = NSIM, seed = 6897154
-    )
-
-    # Store nationally aggregated results
-    h3_V3_res$mod3_V3_sims <- agg_mod3_sims(h3_V3_sims)
-    h3_V3_res$mod3_V3_probs <- get_elimProbs(h3_V3_sims, model = 3)
-
-    # Return results
-    h3_V3_res
-
-  }
-)
-
-mod3_V4_res <- bake(
-  file = paste0("model3/", rl_dir, "mod3_V4_res.rds"), {
-
-    # Create a list to store results
-    h3_V4_res <- list()
-
-    # Get "noVacc" scenario parameters
-    h3_V4_par <- get_model3_vacc_scenario_params(scenario = "fastNation")
-    coef(h3Spat) <- p3
-    coef(h3Spat)[names(h3_V4_par)] <- h3_V4_par
-
-    # Using parameters and filtered distribution, project system
-    h3_V4_sims <- project_from_filter(
-      h3Spat, end_states = h3_bpf_end_states, covarGen = project_rain,
-      nsims = NSIM, seed = 6897154
-    )
-
-    # Store nationally aggregated results
-    h3_V4_res$mod3_V4_sims <- agg_mod3_sims(h3_V4_sims)
-    h3_V4_res$mod3_V4_probs <- get_elimProbs(h3_V4_sims, model = 3)
-
-    # Return results
-    h3_V4_res
-
-  }
-)
-@
-
-Simulations from probabilistic models (Models~1 and~3) represent possible trajectories of the dynamic system under the scientific assumptions of the models.
-Estimates of the probability of cholera elimination can therefore be obtained as the proportion of simulations from these models that result in cholera elimination.
-The results of these projections are summarized in Figs.~\ref{fig:Mod1Scenarios}--\ref{fig:elimProbs}.
-
-Probability of elimination estimates of this form are not meaningful for deterministic models, as the trajectory of these models only represent the mean behavior of the system rather than individual potential outcomes.
-We therefore do not provide probability of elimination estimates under Model~2, but show trajectories under the various vaccination scenarios using this model (Fig.~\ref{fig:mod2Traj}).
-
-<<Create Model 1 VaccScen plots>>=
-gg_m1_V0 <- ggplot() +
-  geom_line(data = h1_S0_results$mod1_V0_sims, aes(x = date, y = q50 + 1), col = 'blue') +
-  geom_ribbon(data = h1_S0_results$mod1_V0_sims, aes(x = date, ymin = q05 + 1, ymax = q95 + 1), alpha = 0.5) +
-  geom_line(data = true_agg_cases, aes(x = date, y = ReportedAll + 1)) +
-  ylab("Log reported cases") +
-  theme(axis.title.x = element_blank()) +
-  ggtitle("V0") +
-  scale_x_date(date_labels = "'%y") +
-  scale_y_log10(labels = scales::trans_format("log10", scales::math_format(10^.x)))
-
-gg_m1_V1 <- ggplot() +
-  geom_line(data = h1_S2_results$mod1_V1_sims, aes(x = date, y = q50 + 1), col = 'blue') +
-  geom_ribbon(data = h1_S2_results$mod1_V1_sims, aes(x = date, ymin = q05 + 1, ymax = q95 + 1), alpha = 0.5) +
-  geom_line(data = true_agg_cases, aes(x = date, y = ReportedAll + 1)) +
-  ylab("Log reported cases") +
-  theme(axis.title = element_blank(),
-        axis.ticks.y = element_blank(),
-        axis.text.y = element_blank()) +
-  ggtitle("V1") +
-  scale_x_date(date_labels = "'%y") +
-  scale_y_log10(labels = scales::trans_format("log10", scales::math_format(10^.x)))
-
-gg_m1_V2 <- ggplot() +
-  geom_line(data = h1_S4_results$mod1_V2_sims, aes(x = date, y = q50 + 1), col = 'blue') +
-  geom_ribbon(data = h1_S4_results$mod1_V2_sims, aes(x = date, ymin = q05 + 1, ymax = q95 + 1), alpha = 0.5) +
-  geom_line(data = true_agg_cases, aes(x = date, y = ReportedAll + 1)) +
-  ylab("Log reported cases") +
-    theme(axis.title = element_blank(),
-        axis.ticks.y = element_blank(),
-        axis.text.y = element_blank()) +
-  ggtitle("V2") +
-  scale_x_date(date_labels = "'%y") +
-  scale_y_log10(labels = scales::trans_format("log10", scales::math_format(10^.x)))
-
-gg_m1_V3 <- ggplot() +
-  geom_line(data = h1_S3_results$mod1_V3_sims, aes(x = date, y = q50 + 1), col = 'blue') +
-  geom_ribbon(data = h1_S3_results$mod1_V3_sims, aes(x = date, ymin = q05 + 1, ymax = q95 + 1), alpha = 0.5) +
-  geom_line(data = true_agg_cases, aes(x = date, y = ReportedAll + 1)) +
-  ylab("Log reported cases") +
-    theme(axis.title = element_blank(),
-        axis.ticks.y = element_blank(),
-        axis.text.y = element_blank()) +
-  ggtitle("V3") +
-  scale_x_date(date_labels = "'%y") +
-  scale_y_log10(labels = scales::trans_format("log10", scales::math_format(10^.x)))
-
-gg_m1_V4 <- ggplot() +
-  geom_line(data = h1_S1_results$mod1_V4_sims, aes(x = date, y = q50 + 1), col = 'blue') +
-  geom_ribbon(data = h1_S1_results$mod1_V4_sims, aes(x = date, ymin = q05 + 1, ymax = q95 + 1), alpha = 0.5) +
-  geom_line(data = true_agg_cases, aes(x = date, y = ReportedAll + 1)) +
-  ylab("Log reported cases") +
-    theme(axis.title = element_blank(),
-        axis.ticks.y = element_blank(),
-        axis.text.y = element_blank()) +
-  ggtitle("V4") +
-  scale_x_date(date_labels = "'%y") +
-  scale_y_log10(labels = scales::trans_format("log10", scales::math_format(10^.x)))
-
-@
-
-\begin{figure}
-<<Plot_Model1_Scenarios, fig.height=2>>=
-lay <- rbind(c(1, 1, 1, 1, 2, 2, 2, 3, 3, 3, 4, 4, 4, 5, 5, 5))
-gridExtra::grid.arrange(
-  gg_m1_V0, gg_m1_V1, gg_m1_V2, gg_m1_V3, gg_m1_V4,
-  layout_matrix = lay
-)
-@
-\caption{\label{fig:Mod1Scenarios}
-Simulations of Model~1 under each vaccination scenario.
-Blue line indicates the simulated median of reported cases, and the ribbon represents $95\%$ of simulations.
-% The various vaccination campaigns made no practical difference in the median scenario, but a drastic difference in the extreme cases.
-}
-\end{figure}
-
-\begin{figure}
+\begin{figure}[!h]
 <<Mod2Fit_and_Scenarios_Figure, fig.height=1.9, fig.width=4.8, fig.align='center'>>=
@@
 
-<<Create Model 3 VaccScen Plots>>=
-gg_m3_V0 <- ggplot() +
-  geom_line(data = mod3_V0_res$mod3_V0_sims, aes(x = date, y = q50 + 1), col = 'blue') +
-  geom_ribbon(data = mod3_V0_res$mod3_V0_sims, aes(x = date, ymin = q05 + 1, ymax = q95 + 1), alpha = 0.5) +
-  geom_line(data = true_agg_cases, aes(x = date, y = ReportedAll + 1)) +
-  ylab("Log reported cases") +
+\subsubsection*{Calibrating Model~3 Parameters}\label{sec:fit3}
+Model~3 describes cholera dynamics in Haiti using a metapopulation model, where the hidden states in each administrative department has an effect on the dynamics in other departments.
+The decision to address metapopulation dynamics using a spatially explicit model, rather than to aggregate over space, is double-edged.
+Evidence for the former approach has been provided in previous studies \cite{king15}, including the specific case of heterogeneity between Haitian departments in cholera transmission \cite{collins14}.
+However, a legitimate preference for simplicity can support a decision to consider nationally aggregated models \cite{saltelli20,green15}.
+One issue that arises when fitting spatially explicit models is that parameter estimation techniques based on the particle filter become computationally intractable as the number of spatial units increases.
+This is a result of the approximation error of particle filters growing exponentially in the dimension of the model \cite{rebeschini15,park20}.
+
+To avoid the approximation error present in high-dimensional models, Lee et al.~\cite{lee20} simplified the problem of estimating the parameters of Model~3 by creating an approximate version of the model where the units are independent given the observed data.
+Reducing a spatially coupled model to individual units in this fashion requires special treatment of any interactive mechanisms between spatial units, such as found in Eq.~\myeqref{eq:model3:foi}.
+Because the simplified, spatially-decoupled version of Model~3 implemented in \cite{lee20} relies on the observed cholera cases, the calibrated model cannot readily be used to obtain forecasts.
+Therefore, in order to obtain model forecasts, Lee et al.~\cite{lee20} used the parameters estimates from the spatially-decoupled approximation of Model~3 to obtain forecasts using the fully coupled version of the model.
+This approach of model calibration and forecasting avoids the issue of particle depletion, but may also be problematic.
+One concern is that cholera dynamics in department $u$ are highly related to the dynamics in the remaining departments;
+calibrating model parameters while conditioning on the observed cases in other departments may therefore lead to an over-dependence on observed cholera cases.
+Another concern is that the two versions of the model are not the same, resulting in sub-optimal parameter estimates for the spatially coupled model,
+as parameters that maximize the likelihood of the decoupled model almost certainly do not maximize the likelihood of the fully coupled model.
+These two concerns may explain the unrealistic forecasts and low likelihood of Model~3 from Lee et al.~\cite{lee20} (see Table~\ref{tab:likes}).
+
+At the time Lee et al.~\cite{lee20} was published, there was no known algorithm that could readily be used to maximize the likelihood of an arbitrary meta-population POMP model with coupled spatial dynamics, which justifies the spatial decoupling approximation that was used to calibrate model parameters.
+For our analysis, we calibrate the parameters of the spatially coupled version of Model~3 using the iterated block particle filter (IBPF) algorithm \cite{ionides22}.
+This algorithm extends the work of Ning and Ionides \cite{ning21ibpf}, who provided theoretic justification for the version of the algorithm that only estimates unit-specific parameters.
+The IBPF algorithm enables us to directly estimate the parameters of models describing high-dimensional partially-observed nonlinear dynamic systems via likelihood maximization.
+The ability to directly estimate parameters of Model~3 is responsible for the large increase in model likelihoods reported in Table~\ref{tab:likes}.
+Simulations from the fitted model are displayed in Fig.~\ref{fig:h3spatsims}.
+
+<<Model 3 SpatPOMP sims, echo=FALSE>>=
+
+# Number of simulations from Model 3. Because it's high-dimensional, we include
+# a run level to avoid time/memory issues when editing. Final results shown
+# with RUN_LEVEL = 3.
+
+h3_nsim <- switch(RUN_LEVEL, 20, 100, 500)
+
+# Get quantiles of simulations from the calibrated Model 3.
+h3_quants <- bake(
+  file = paste0('model3/', rl_dir, 'h3Sims.rds'), {
+
+    coef(h3Spat) <- p3
+    # Simulate from the fitted version of Model 3
+    h3_sims <- simulate(
+      h3Spat, nsim = h3_nsim, format = 'data.frame'
+    )
+
+    # For each department and time-point, calculate percentiles of the
+    # simulations
+    h3_sims %>%
+      rename(dep = unitname) %>%
+      group_by(dep, time) %>%
+      summarise(
+        q05 = quantile(cases, 0.025, na.rm = T),
+        mean = mean(cases, na.rm = T),
+        q50 = quantile(cases, 0.5, na.rm = T),
+        q95 = quantile(cases, 0.975, na.rm = T)
+      ) %>%
+      ungroup() %>%
+      mutate(
+        date = lubridate::date_decimal(time)
+      ) %>%
+      mutate(date = as.Date(lubridate::round_date(date, unit = 'day')))
+  },
+  timing = FALSE
+)
+
+@
+
+\begin{figure}[!h]
+<<Model3_SpatPOMP_sims_Figure, fig.height=3>>=
+
+# Plot the reported cases and confidence region for each department.
+ggplot() +
+  geom_line(data = h3_quants, aes(x = date, y = q50 + 1), col = 'blue') +
+  geom_line(data = dep_plot_df, aes(x = date, y = cases + 1), col = 'black') +
+  geom_ribbon(  # plot confidence region.
+    data = h3_quants,
+    aes(x = date, ymin = q05 + 1, ymax = q95 + 1),
+    alpha = 0.4
+  ) +
+  facet_wrap(
+    ~factor(dep, levels = plot_order), nrow = 2, labeller = dep_labeller
+  ) +
+  labs(y = 'Log Reported Cholera Cases') +
+  scale_x_date(  # Format the date on the X-axis.
+    date_labels = "'%y",
+    breaks = seq.Date(
+      from = as.Date("2011-01-01"),
+      to = as.Date("2019-01-01"),
+      by = '2 years'
+    )
+  ) +
   theme(axis.title.x = element_blank()) +
-  ggtitle("V0") +
-  scale_x_date(date_labels = "'%y") +
-  scale_y_log10(labels = scales::trans_format("log10", scales::math_format(10^.x)))
-
-gg_m3_V1 <- ggplot() +
-  geom_line(data = mod3_V1_res$mod3_V1_sims, aes(x = date, y = q50 + 1), col = 'blue') +
-  geom_ribbon(data = mod3_V1_res$mod3_V1_sims, aes(x = date, ymin = q05 + 1, ymax = q95 + 1), alpha = 0.5) +
-  geom_line(data = true_agg_cases, aes(x = date, y = ReportedAll + 1)) +
-  ylab("Log reported cases") +
-  theme(axis.title = element_blank(),
-        axis.ticks.y = element_blank(),
-        axis.text.y = element_blank()) +
-  ggtitle("V1") +
-  scale_x_date(date_labels = "'%y") +
-  scale_y_log10(labels = scales::trans_format("log10", scales::math_format(10^.x)))
-
-gg_m3_V2 <- ggplot() +
-  geom_line(data = mod3_V2_res$mod3_V2_sims, aes(x = date, y = q50 + 1), col = 'blue') +
-  geom_ribbon(data = mod3_V2_res$mod3_V2_sims, aes(x = date, ymin = q05 + 1, ymax = q95 + 1), alpha = 0.5) +
-  geom_line(data = true_agg_cases, aes(x = date, y = ReportedAll + 1)) +
-  ylab("Log reported cases") +
-    theme(axis.title = element_blank(),
-        axis.ticks.y = element_blank(),
-        axis.text.y = element_blank()) +
-  ggtitle("V2") +
-  scale_x_date(date_labels = "'%y") +
-  scale_y_log10(labels = scales::trans_format("log10", scales::math_format(10^.x)))
-
-gg_m3_V3 <- ggplot() +
-  geom_line(data = mod3_V3_res$mod3_V3_sims, aes(x = date, y = q50 + 1), col = 'blue') +
-  geom_ribbon(data = mod3_V3_res$mod3_V3_sims, aes(x = date, ymin = q05 + 1, ymax = q95 + 1), alpha = 0.5) +
-  geom_line(data = true_agg_cases, aes(x = date, y = ReportedAll + 1)) +
-  ylab("Log reported cases") +
-    theme(axis.title = element_blank(),
-        axis.ticks.y = element_blank(),
-        axis.text.y = element_blank()) +
-  ggtitle("V3") +
-  scale_x_date(date_labels = "'%y") +
-  scale_y_log10(labels = scales::trans_format("log10", scales::math_format(10^.x)))
-
-gg_m3_V4 <- ggplot() +
-  geom_line(data = mod3_V4_res$mod3_V4_sims, aes(x = date, y = q50 + 1), col = 'blue') +
-  geom_ribbon(data = mod3_V4_res$mod3_V4_sims, aes(x = date, ymin = q05 + 1, ymax = q95 + 1), alpha = 0.5) +
-  geom_line(data = true_agg_cases, aes(x = date, y = ReportedAll + 1)) +
-  ylab("Log reported cases") +
-    theme(axis.title = element_blank(),
-        axis.ticks.y = element_blank(),
-        axis.text.y = element_blank()) +
-  ggtitle("V4") +
-  scale_x_date(date_labels = "'%y") +
-  scale_y_log10(labels = scales::trans_format("log10", scales::math_format(10^.x)))
+  scale_y_log10(  # Plot on log-scale
+    labels = scales::trans_format("log10", scales::math_format(10^.x))
+  )
 @
-
-\begin{figure}
-<<Plot_Model3_Scenarios, fig.height=2>>=
-lay <- rbind(c(1, 1, 1, 1, 2, 2, 2, 3, 3, 3, 4, 4, 4, 5, 5, 5))
-gridExtra::grid.arrange(
-  gg_m3_V0, gg_m3_V1, gg_m3_V2, gg_m3_V3, gg_m3_V4,
-  layout_matrix = lay
-)
+\caption{\label{fig:h3spatsims}
+Simulations from initial conditions using the spatially coupled version of Model~3.
+The black curve represents true case count, the blue line the median of 500 simulations from the model, and the gray ribbons representing $95\%$ confidence interval.
+}
+\end{figure}
+
+<<Remove Model3 spatSims, cache=FALSE, include=FALSE, message=FALSE, echo=FALSE>>=
+rm(h3_quants, h3_nsim)
+gc()
 @
-\caption{\label{fig:Mod3Scenarios}
-Simulations of Model~3 under each vaccination scenario.
-Blue line indicates the simulated median of reported cases, and ribbon represents $95\%$ of simulations.
-% The various vaccination campaigns made no practical difference in the median scenario, but a drastic difference in the extreme cases.
+
+\subsection*{Model Diagnostics}\label{sec:model_diagnostics}
+
+The goal of parameter calibration---whether done using Bayesian or frequentist methods---is to find the best description of the observed data under the assumptions of the model.
+Obtaining the best fitting set of parameters for a given model does not, however, guarantee that the model provides an accurate representation of the system under investigation.
+Model misspecification, which may be thought of as the omission of a mechanism in the model that is an important feature of the dynamic system, is inevitable at all levels of model complexity.
+To make progress, while accepting proper limitations, one must bear in mind the much-quoted observation of George Box \cite{box79} that ``all models are wrong but some are useful.''
+Beyond being good practical advice for applied statistics, this assertion is relevant for the philosophical justification of statistical inference as severe testing \cite{mayo18}.
+% \citep[][Sec.~4.8]{mayo18}.
+In this section, we discuss some tools for diagnosing mechanistic models with the goal of making the subjective assessment of model ``usefulness'' more objective.
+To do this, we will rely on the quantitative ability of the model to match the observed data, which we call the model's {\it goodness-of-fit}, with the guiding principle that a model which cannot adequately describe observed data may not be reliable for useful purposes.
+Goodness-of-fit may provide evidence supporting the causal interpretation of one model versus another, but cannot by itself rule out the possibility of alternative explanations.
+
+One common approach to assess a mechanistic model's goodness-of-fit is to compare simulations from the fitted model to the observed data.
+Visual inspection may indicate defects in the model, or may  suggest that the observed data are a plausible realization of the fitted model.
+While visual comparisons can be informative, they provide only a weak and informal measure of the goodness-of-fit of a model.
+The study by Lee et al.~\cite{lee20} provides an example of this: their models and parameter estimates resulted in simulations that visually resembled the observed data, yet resulted in model likelihoods that were considerably smaller than likelihoods that can be achieved via the likelihood based optimization techniques that were used (see Table~\ref{tab:likes}).
+Alternative forms of model validation should therefore be used in conjunction with visual comparisons of simulations to observed data.
+
+Another approach is to compare a quantitative measure of the model fit (such as MSE, predictive accuracy, or model likelihood) among all proposed models.
+These comparisons, which provide insight into how each model performs relative to the others, are quite common \cite{rinaldo12,sallah17}.
+To calibrate relative measures of fit, it is useful to compare against a model that has well-understood statistical ability to fit data, and we call this model a {\it benchmark}.
+Standard statistical models, interpreted as associative models without requiring any mechanistic interpretation of their parameters, provide suitable benchmarks.
+Examples include linear regression, auto-regressive moving average time series models, or even independent and identically distributed measurements.
+Benchmarks enable us to evaluate the goodness of fit that can be expected of a suitable mechanistic model.
+
+Associative models are not constrained to have a causal interpretation, and typically are designed with the sole goal of providing a statistical fit to data.
+Therefore, we should not require a candidate mechanistic model to beat all benchmarks.
+However, a mechanistic model which falls far short against benchmarks is evidently failing to explain some substantial aspect of the data.
+A convenient measure of fit should have interpretable differences that help to operationalize the meaning of far short.
+Ideally, the measure should also have favorable theoretical properties.
+Consequently, we focus on log-likelihood as a measure of goodness of fit, and we adjust for the degrees of freedom of the models to be compared by using the Akaike information criterion (AIC) \cite{aic74}.
+
+In some cases, a possible benchmark model could be a generally accepted mechanistic model, but often no such model is available.
+Because of this, we use a log-linear Gaussian ARMA model as an associative benchmark, as recommended by He et al.~\cite{he10}.
+The theory and practice of ARMA models is well developed, and these linear models are appropriate on a log scale due to the exponential growth and decay characteristic of biological dynamics.
+Likelihoods of Models~1--3 and their respective ARMA benchmark models are provided in Table~\ref{tab:likes}.
+
+It should be universal practice to present measures of goodness of fit for published models, and mechanistic models should be compared against benchmarks.
+In our literature review of the Haiti cholera epidemic, no non-mechanistic benchmark models were considered in any of the 32 papers that used dynamic models to describe cholera in order to obtain scientific conclusions.
+Including benchmarks would help authors and readers to detect and confront any major statistical limitations of the proposed mechanistic models.
+In addition, the published goodness of fit provides a concrete point of comparison for subsequent scientific investigations.
+When combined with online availability of data and code, objective measures of fit provide a powerful tool to accelerate scientific progress, following the paradigm of the {\it common task framework} \cite{donoho17}.
+% \citep[][Sec.~6]{donoho17}.
+
+The use of benchmarks may also be beneficial when developing models at differing spatial scales, where a direct comparison between model likelihoods is meaningless.
+In such a case, a benchmark model can be fit to each spatial resolution being considered, and each model compared to their respective benchmark.
+Large advantages (or shortcomings) in model likelihood relative to the benchmark for a given spatial scale that are not present in other spatial scales may provide weak evidence for (or against) the statistical fit of models across a range of spatial resolutions.
+
+<<Table 2 Input, echo=FALSE, eval=TRUE>>=
+NREPS_EVAL <- switch(RUN_LEVEL,   3,   8,    36)
+NP_EVAL    <- switch(RUN_LEVEL, 100, 500,  5000)
+
+# Objective function of model 2, epidemic phase
+epi_ofun <- traj_objfun(
+  h2_epi,
+  params = h2_fit$h2_params
+)
+
+mod2_epi_ll <- -epi_ofun(par = h2_fit$h2_params)
+mod2_ll <- mod2_epi_ll - sum(log(h2_epi@data + 1), na.rm = TRUE)
+
+arima_lik2 <- 0
+
+# model 2 AIC
+mod2_n_params <- h2_fit$n_fit_params
+mod2_aic <- 2 * mod2_n_params - 2 * mod2_ll
+
+# model 1 AIC
+mod1_n_params <- length(
+  c('beta1', 'beta2', 'beta3', 'beta4', 'beta5', 'beta6',
+    'rho', 'nu', 'E_0', 'I_0', 'tau_epi',
+    'tau_end', 'sig_sq_epi', 'sig_sq_end', 'betat')
+)
+mod1_aic <- 2 * mod1_n_params - 2 * mod1_ll
+
+# model 3 AIC
+mod3_aic <- 2 * mod3_n_params - 2 * mod3_ll
+
+# Calculate likelihood via block-particle filter
+mod3_eval_res <- bake(
+  file = paste0('model3/', rl_dir, 'mod3_eval_res.rds'), {
+
+    # Create a list to store results
+    results <- list()
+
+    coef(h3Spat) <- p3
+
+    # Perform block-particle filter
+    h3_bpf <- foreach(i = 1:NREPS_EVAL, .combine = c) %dopar% {
+      bpfilter(h3Spat, Np = NP_EVAL, block_size = 1)
+    }
+
+    ll <- logLik(h3_bpf)
+    results$ll <- logmeanexp(ll[!is.na(ll)], se = FALSE)
+
+    ### Get likelihood for subset of the data
+
+    # Find appropriate sub-set
+    in_subset_cols <- h3_bpf[[1]]@times >= as.numeric(lubridate::decimal_date(as.Date("2014-03-01")))
+    in_subset_rows <- h3_bpf[[1]]@unit_names != 'Ouest'
+
+    mod3_coupled_subset_evals <- c()
+    for (i in 1:length(h3_bpf)) {
+      mod3_coupled_subset_evals <- c(
+        mod3_coupled_subset_evals,
+        sum(h3_bpf[[i]]@block.cond.loglik[in_subset_rows, in_subset_cols])
+      )
+    }
+
+    results$subset_ll <- logmeanexp(mod3_coupled_subset_evals)
+    results
+  },
+  timing = FALSE
+)
+
+mod3_eval_res$subset_aic <- 2 * mod3_n_params - 2 * mod3_eval_res$subset_ll
+
+rm(NREPS_EVAL, NP_EVAL)
+@
+
+<<CalculateARMA, echo=FALSE, message=FALSE, include=FALSE, cache=TRUE>>=
+
+ARMA_benchmarks <- list()
+
+###
+### Model 1 ###
+###
+
+# Load aggregated data
+m1_agg_data <- haiti1_agg_data()
+
+# Fit ARMA(2, 1) on log-cases
+m1_log_arma <- arima(log(m1_agg_data$cases + 1), order = c(2, 0, 1))
+
+# Fit ARMA(2, 1) on natural scale of cases
+m1_arma <- arima(m1_agg_data$cases, order = c(2, 0, 1))
+
+# Check which one is better, and save benchmark as the better one
+if (m1_log_arma$loglik - sum(log(m1_agg_data$cases + 1), na.rm = TRUE) >= m1_arma$loglik) {
+  ARMA_benchmarks[['m1']] <- m1_log_arma$loglik - sum(log(m1_agg_data$cases + 1), na.rm = TRUE)
+} else {
+  ARMA_benchmarks[['m1']] <- m1_arma$loglik
+}
+
+rm(m1_agg_data, m1_log_arma, m1_arma)
+gc()
+
+###
+### Model 2 ###
+###
+
+h2_data <- haiti2_data(recreate = FALSE)
+
+# model 2 ARIMA benchmark
+h2_dep_names <- h2_data$department %>% unique()
+m2_benchmark <- 0
+for (i in 1:10){
+  cases <- h2_data[h2_data$department == h2_dep_names[i], ]$cases
+  log_cases <- log(cases + 1)
+
+  m2_arma_dep_ll <- arima(cases, order=c(2, 0, 1))$loglik
+  m2_log_arma_dep_ll <- arima(log_cases, order=c(2, 0, 1))$loglik - sum(log_cases, na.rm = TRUE)
+
+  m2_benchmark <- m2_benchmark + max(m2_arma_dep_ll, m2_log_arma_dep_ll)
+}
+
+ARMA_benchmarks[['m2']] <- m2_benchmark
+
+rm(m2_benchmark, h2_dep_names, cases, log_cases, m2_arma_dep_ll, m2_log_arma_dep_ll, h2_data)
+gc()
+
+###
+### Model 3 ###
+###
+
+mod3_cases <- as.data.frame(t(h3Spat@data))
+
+m3_benchmark <- 0
+for (dep in 1:10) {
+
+  # Get department cases
+  cases <- as.numeric(mod3_cases[, dep])
+
+  # Fit model on natural scale of cases
+  m3_a201 <- arima(cases, order = c(2, 0, 1))
+
+  # Fit model on log-scale of cases
+  m3_log_a201 <- arima(log(cases + 1), order = c(2, 0, 1))
+
+  m3_benchmark <- m3_benchmark + max(
+    m3_log_a201$loglik - sum(log(cases + 1), na.rm = TRUE),
+    m3_a201$loglik
+  )
+}
+
+ARMA_benchmarks[['m3']] <- m3_benchmark
+rm(cases, m3_a201, m3_log_a201, m3_benchmark, dep,
+   mod3_cases, temp, jumps, MODEL3_CASES, i, j)
+gc()
+@
+
+<<Table 2 Lee Input, echo=FALSE, include=FALSE>>=
+mod1_lee_n_params <- 20
+mod1_lee_aic <- 2 * mod1_lee_n_params - 2 * mod1_lee_ll
+
+mod2_lee_aic <- 2 * h2_fit$leeFit_n_params - 2 * h2_fit$leeFit_ll
+@
+
+<<Lee3 Sub Calculations, echo=FALSE, include=FALSE, message=FALSE>>=
+lee3_NP   <- switch(RUN_LEVEL, 50, 250, 5000)
+lee3_NREP <- switch(RUN_LEVEL,  3,   8,   36)
+
+registerDoRNG(49637421)
+
+# Create Lee et al (2020a) version of model 3 on sub-data
+lee3_sub <- lee3_spatPomp()
+
+# This model is only fit for a subset of the data, so we will set data
+# that are not part of this subset as NA. This will allow us to estimate
+# the likelihood for only the values that were considered during the
+# parameter estimation phase.
+# set_as_na <- lee3_sub@times < lubridate::decimal_date(as.Date('2017-06-10'))
+# temp_ouest_cases <- lee3_sub@data['casesOuest', ]
+# temp_ouest_cases[set_as_na] <- NA_real_
+# lee3_sub@data['casesOuest', ] <- temp_ouest_cases
+
+lee3_all <- lee3_spatPomp(start_date = "2010-10-23")
+
+lee3_results <- bake(
+  file = paste0('model3/', rl_dir, 'lee3_results.rds'), {
+
+    # Create lists to save results
+    lee3_results <- list()
+
+    # Save how long it takes for sub-data
+    lee3_results[["sub_bpf_time"]] <- system.time({
+
+      lee3_sub_ll_evals <- foreach(
+        i = 1:lee3_NREP, .combine = rbind
+      ) %dopar% {
+
+        liks <- numeric(10L)
+
+        lee_bpf_out <- bpfilter(lee3_sub, block_size = 1, Np = lee3_NP)
+        Ouest_ind <- which(lee_bpf_out@unit_names == "Ouest")
+
+        for (u in 1:10) {
+          if (u == Ouest_ind) {
+            liks[u] <- sum(lee_bpf_out@block.cond.loglik[u, lee3_sub@times >= lubridate::decimal_date(as.Date('2017-06-10'))])
+          } else {
+            liks[u] <- sum(lee_bpf_out@block.cond.loglik[u, ])
+          }
+        }
+
+        liks
+      }
+    })
+
+    # Save all likelihood evaluations
+    lee3_results[["sub_ll_eval"]] <- logmeanexp(rowSums(lee3_sub_ll_evals)[!is.na(rowSums(lee3_sub_ll_evals))])
+
+    registerDoRNG(2789131)
+
+    # Save how long it takes for all data
+    lee3_results[["all_pf_time"]] <- system.time({
+      # Save all likelihood evaluations
+
+      lee3_all_ll_evals <- foreach(
+        i = 1:lee3_NREP, .combine = c
+      ) %dopar% {
+        logLik(bpfilter(lee3_all, block_size = 1, Np = lee3_NP))  # Get likelihood evaluations
+      }
+
+      lee3_results[["all_ll_eval"]] <- logmeanexp(lee3_all_ll_evals[!is.na(lee3_all_ll_evals)])
+
+    })
+
+    lee3_results[["n_params"]] <- 29
+
+    lee3_results
+  },
+  timing = FALSE
+)
+
+rm(lee3_sub, lee3_all, lee3_NP, lee3_NREP)
+gc()
+
+mod3_lee_coupled_aic <- 2 * lee3_results$n_params - 2 * lee3_results$all_ll_eval
+
+mod3_lee_coupled_partial_aic <- 2 * lee3_results$n_params - 2 * lee3_results$sub_ll_eval
+@
+
+\begin{table}
+\centering
+
+\begin{tabular}{|c|c|c|c|}
+\hline
+ & \thead{Model~1} & \thead{Model~2} & \thead{Model~3}
+\\
+\hline
+\hline
+\multirow{2}{*}{Log-likelihood} &
+  $\Sexpr{myround(mod1_ll, 1)}$ &
+  $\Sexpr{myround(mod2_ll, 1)}$ &
+  $\Sexpr{myround(mod3_ll, 1)}$ \\
+    & ($\Sexpr{myround(mod1_lee_ll, 1)}$)\footnotemark[1] &
+  ($\Sexpr{myround(h2_fit$leeFit_ll, 1)}$) &
+  ($\Sexpr{myround(lee3_results[["all_ll_eval"]], 1)}$)\footnotemark[2]
+\\
+\hline
+Number of &
+  $\Sexpr{as.character(mod1_n_params)}$ &
+  $\Sexpr{as.character(mod2_n_params)}$ &
+  $\Sexpr{as.character(mod3_n_params)}$ \\
+ Fit Parameters &
+ (\Sexpr{as.character(mod1_lee_n_params)}) &
+ ($\Sexpr{as.character(h2_fit$leeFit_n_params)}$) &
+ ($\Sexpr{as.character(lee3_results$n_params, 1)}$)
+\\
+\hline
+\multirow{2}{*}{AIC} &
+  $\Sexpr{myround(mod1_aic, 1)}$ &
+  $\Sexpr{myround(mod2_aic, 1)}$ &
+  $\Sexpr{myround(mod3_aic, 1)}$ \\
+  & ($\Sexpr{myround(mod1_lee_aic, 1)}$)\footnotemark[1] &
+  ($\Sexpr{myround(mod2_lee_aic, 1)}$) &
+  ($\Sexpr{myround(mod3_lee_coupled_aic, 1)}$)\footnotemark[2]
+\\
+\hline
+\thead{Log-ARMA(2,1) \\ AIC} &
+  $\Sexpr{myround(-2 * ARMA_benchmarks$m1 + 2 * 5, 1)}$ &
+  $\Sexpr{myround(-2 * ARMA_benchmarks$m2 + 2 * 5, 1)}$ &
+  $\Sexpr{myround(-2 * ARMA_benchmarks$m3 + 2 * 5, 1)}$
+\\
+\hline
+\end{tabular}
+\caption{\label{tab:likes}AIC values for each model compared to their ARMA benchmark.
+Values in parentheses are corresponding values using \cite{lee20} parameter estimates.
+$\textsuperscript{1}$The reported likelihood is an upper bound of the likelihood of the \cite{lee20} model.
+$\textsuperscript{2}$\cite{lee20} fit Model~3 to a subset of the data (March 2014 onward, excluding data from Ouest in 2015-2016).
+On this subset, their model has a likelihood of $\Sexpr{myround(lee3_results$sub_ll_eval, 1)}$.
+On this same subset, our model has a likelihood of $\Sexpr{myround(mod3_eval_res$subset_ll, 1)}$.
+See Sec.~S6 of the supplement material for more details on estimating the likelihood of the Lee et al.~\cite{lee20} models.
+}
+\end{table}
+
+Comparing model log-likelihoods to a suitable benchmark may not be sufficient to identify all the strengths and weaknesses of a given model.
+Additional techniques include the inspection of conditional log-likelihoods of each observation given the previous observations in order to understand how well the model describes each data point (Sec.~S5).
+Other tools include plotting the effective sample size of each observation \cite{liu01}; plotting the values of the hidden states from simulations (supplement Sec.~5.1);
+and comparing summary statistics of the observed data to simulations from the model \cite{wood10,king15}.
+
+\subsection*{Corroborating Fitted Models with Scientific Knowledge}
+
+The resulting mechanisms in a fitted model can be compared to current scientific knowledge about a system.
+Agreement between model based inference and our current understanding of a system may be taken as a confirmation of both model based conclusions and our scientific understanding.
+On the other hand, comparisons may generate unexpected results that have the potential to spark new scientific knowledge \cite{ganusov16}.
+
+In the context of our case study, we demonstrate how the fit of Model~1 corroborates other evidence concerning the role of rainfall in cholera epidemics.
+Specifically, we examine the results of fitting the flexible cubic spline term in Model~1 (Eqs.~\myeqref{model1:lambda}--\myeqref{model1:beta}).
+The cubic splines permit flexible estimation of seasonality in the force of infection, $\transmission(t)$.
+Fig.~\ref{fig:h1SeasRain} shows that the estimated seasonal transmission rate $\transmission$ mimics the rainfall dynamics in Haiti, despite Model~1 not having access to rainfall data.
+This is consistent with previous studies finding that rainfall played an important role in cholera transmission in Haiti \cite{lemaitre19,eisenberg13}.
+The estimated seasonality also features an increased transmission rate during the fall, which was noticed at an earlier stage of the epidemic \cite{rinaldo12}; the high transmission rate in the fall may also be a result of the increase transmission that occurred in the fall of 2016, when hurricane Matthew struck Haiti \cite{ferreirai16}.
+
+<<Get Model 1 Seasonality>>=
+h1_seas_df <- get_h1_seasonality(h1)
+
+df_rain <- haitiRainfall %>%
+  pivot_longer(  # Convert data to "long" format
+    data = .,
+    cols = -date,
+    names_to = 'dep',
+    values_to = 'rainfall'
+  ) %>%
+  mutate(year = lubridate::year(date), week = lubridate::week(date)) %>%
+  group_by(year, week) %>%
+  summarize(nwr = sum(rainfall)) %>%
+  ungroup() %>%  # Not needed, but good practice
+  mutate(
+    date_week_start = lubridate::ymd(paste0(2014, "-01-01")) +
+      lubridate::weeks(week - 1),
+    week_date = as.Date(date_week_start, format = "%m-%d"),
+    std_national_rain = (nwr - min(nwr)) / (max(nwr) - min(nwr))
+  )
+
+# Calculate the average national weekly rainfall.
+mean_rain <- df_rain %>%
+  group_by(week_date) %>%
+  summarize(mean_rain = mean(std_national_rain)) %>%
+  ungroup() %>%
+  mutate(mean_rain = (mean_rain - min(mean_rain)) / ((max(mean_rain) - min(mean_rain))))
+
+# Create a figure showing annual rainfal, scaled from 0-1. This is top figure
+gg_rain <- df_rain %>%
+  filter(week != 1, week != max(week)) %>%
+  ggplot(aes(x = week_date, y = std_national_rain, group = year, col = year)) +
+  geom_line() +
+  theme_bw() +
+  guides(color = 'none') +
+  ylab('Standardized Weekly\nNational Rain') +
+  theme(  # Remove elements that are un-unsed becuase this is top figure
+    axis.title.x = element_blank(),
+    axis.ticks.x = element_blank(),
+    axis.text.x = element_blank(),
+    axis.text.y = element_text(size = 7.5),
+    axis.title.y = element_text(size = 9)
+  )
+
+# Create plot that shows the seasonality of Model 1, alongside the average
+# weekly national rainfall.
+gg_trans <- ggplot() +
+  geom_line(data = h1_seas_df, aes(x = date, y = trans_std), linetype = 'dashed') +
+  geom_line(
+    data = mean_rain,
+    aes(x = week_date + lubridate::years(6), y = mean_rain),
+    col = '#2166ac'
+  ) +
+  scale_x_date(date_labels = "%b", date_breaks = '1 month') +
+  ylab('Standardized Seasonal\nContact Rate') +
+  theme(axis.title.x = element_blank(),
+        axis.text = element_text(size = 7.5),
+        axis.title.y = element_text(size = 9))
+
+@
+
+\begin{figure}[!h]
+\centering
+<<Model1_Seasonality_Figure, fig.height=2.9, fig.width=5, fig.align='center'>>=
+cowplot::plot_grid(gg_rain, gg_trans, align = "v", ncol = 1)
+@
+\caption{\label{fig:h1SeasRain}
+(Top) weekly rainfall in Haiti, lighter colors representing more recent years.
+(Bottom) estimated seasonality in the transmission rate (dashed line) plotted alongside mean rainfall (solid line).
+The outsided effect of rainfall in the fall may be due to Hurricane Matthew, which struck Haiti in October of 2016 and resulted in an increase of cholera cases in the nation.
 }
 \end{figure}
+
+For any model-based inference, it is important to recognize and assess the modeling simplifications and assumptions that were used in order to arrive at the conclusions.
+In epidemiological studies, for example, quantitative understanding of individual-level processes may not perfectly match model parameters that were fit to population-level case counts, even when the model provides a strong statistical fit \cite{he10}.
+This makes direct interpretation of estimated parameters delicate.
+
+Our case study provides an example of this in the parameter estimate for the duration of natural immunity due to cholera infection, $\muRS^{-1}$.
+Under the framework of Model~2, the best estimate for this parameter is $\Sexpr{signif(1/h2_fit$h2_params["sigma"],3)} \, \mathrm{yr}$, suggesting that individuals have effectively permanent immunity to cholera once infected.
+To interpret this result, we bear in mind that the data ranged from 2010-2019, and therefore estimates of immunity longer than $10\, \mathrm{yr}$---the upper end of previous estimates of natural immunity \cite{king08}---effectively result in the same model dynamics.
+The depletion of susceptible individuals may also be attributed to confounding mechanisms---such as localized vaccination programs and non-pharmaceutical interventions that reduce cholera transmission \cite{trevisin22, rebaudet21}---that were not accounted for in the model.
+Perhaps the best interpretation of the estimated parameter, then, is that under model assumptions, the model most adequately describes the observed data by having a steady decrease in the number of susceptible individuals.
+The weak statistical fit of Model~2 compared to a log-linear benchmark (see Table~\ref{tab:likes}) cautions us against drawing quantitative conclusions from this model.
+A model that has a poor statistical fit may nevertheless provide a useful conceptual framework for thinking about the system under investigation.
+However, a claim that the model has been validated against data should be reserved for situations where the model provides a statistical fit that is competitive against alternative explanations.
+
+\subsection*{Robust interpretation of model based conclusions}\label{sec:advice}
+
+A model which aspires to provide quantitative guidance for assessing interventions should provide a quantitative statistical fit for available data.
+However, strong statistical fit does not guarantee a correct causal structure: it does not even necessarily require the model to assert a causal explanation.
+A causal interpretation is strengthened by corroborative evidence.
+For example, reconstructed latent variables (such as numbers of susceptible and recovered individuals) should make sense in the context of alternative measurements of these variables \cite{grad12}.
+In the context of our case study, by reconstructing the latent states of Model~3, we notice that the calibrated model favors higher levels of cholera transmission than what was typically observed in the incidence data (Sec.~S5.1 of the supplement).
+This result hints at the possibility of model mispecification, and should prompt us to be sceptical of the reliability of forecasts from this model.
+Similarly, parameters that have been calibrated to data should make sense in the context of alternative lines of evidence about the phenomena being modeled, while making allowance for the possibility that the interpretations of parameters may vary when modeling across differing spatial scales.
+
+If a mechanistic model including a feature (such as a representation of a mechanism, or the inclusion of a covariate) fits better than mechanistic models without that feature, and also has competitive fit compared to associative benchmarks, this may be taken as evidence supporting the scientific relevance of the feature.
+As for any analysis of observational data, we must be alert to the possibility of confounding.
+For a covariate, this shows up in a similar way to regression analysis: the covariate under investigation could be a proxy for some other unmodeled or unmeasured covariate.
+For a mechanism, the model feature could in principle explain the data by helping to account for some different unmodeled phenomenon.
+In the context of our analysis, the statistical evidence of a trend in transmission rate in Model~1 could be explained by any trending variable (such as hygiene improvements, or changes in population behavior), resulting in confounding from colinear covariates.
+Alternatively, the trend could be attributed to a decreasing reporting rate rather than decreasing transmission rate, resulting in confounded mechanisms.
+The robust statistical conclusion is that a model which allows for change fits better than one which does not---we argue that a decreasing transmission rate is a plausible way to explain this, but the incidence data themselves do not provide enough information to pin down the mechanism.
+
+% Place figure captions after the first paragraph in which they are cited.
+\begin{figure}[!h]
+\caption{{\bf Bold the figure title.}
+Figure caption text here, please use this space for the figure panel descriptions instead of using subfigure commands. A: Lorem ipsum dolor sit amet. B: Consectetur adipiscing elit.}
+\label{fig1}
+\end{figure}
+
+% Results and Discussion can be combined.
+\section*{Results}
+
+\subsection*{Forecasts}\label{sec:filter}
+
+Forecasts are an attempt to provide an accurate estimate of the future state of a system based on currently available data, together with an assessment of uncertainty.
+Forecasts from mechanistic models that are compatible with current scientific understanding may also provide estimates of the future effects of potential interventions.
+Further, they may enable real-time testing of new scientific hypotheses \cite{lewis22}.
+
+Recent information about a dynamic system should be more relevant for a forecast than older information.
+This assertion may seem self-evident, but it is not the case for deterministic models, for which the initial conditions together with the parameters are sufficient for forecasting, and so recent data do not have special importance.
+Epidemiological forecasts based on deterministic models are not uncommon in practice, despite their limitations \cite{king15}.
+That may explain why Lee et al.~\cite{lee20} chose to obtain forecasts by simulating the calibrated models forward from initial conditions.
+Here, we obtain forecasts by simulating future values using latent states that are consistent with the most recent observations by drawing latent states at time the last observation time $t_N$ from the filtering distribution $f_{\bm{X}_N|\bm{Y}_{1:N}}(\bm{x}_{N} | \bm{y}^*_{1:N} ; \hat\paramVec)$,
+where $\hat\paramVec$ is a vector of calibrated parameters.
+The decision not to do this partially explains the unsuccessful forecasts of Lee et al.~\cite{lee20}: their Table~S7 shows that the subset of their simulations which were consistent with observing zero cases in 2019 also accurately predicted the prolonged absence of detected cholera.
+
+Uncertainty in just a single parameter can lead to drastically different forecasts \cite{saltelli20}.
+Therefore, parameter uncertainty should also be considered when obtaining model forecasts to influence policy.
+If a Bayesian technique is used for parameter estimation, a natural way to account for parameter uncertainty is to obtain simulations from the model where each simulation is obtained using parameters drawn from the estimated posterior distribution.
+For frequentist inference, one possible approach is obtaining model forecasts from various values of $\paramVec$, where the values of $\paramVec$ are sampled proportionally according to their corresponding likelihoods \cite{king15}; see Sec.~S7 of the supplement material for more details.
+Both of these approaches share the similarity that parameters are chosen for the forecast approximately in proportion to their corresponding value of the likelihood function, $f_{\bm{Y}_{1:N}}(\bm{y}_{1:N}^*; \paramVec)$.
+In this analysis, we do not construct forecasts accounting for parameter uncertainty as our focus is on the estimation and diagnosis of mechanistic models.
+Furthermore, we use the projections from a single point estimate to highlight the deficiency of deterministic models that the only variability in model projections is a result of parameter uncertainty, which can lead to over-confidence in forecasts \cite{king15}.
+
+The primary forecasting goal of Lee et al.~\cite{lee20} was to investigate the potential consequences of vaccination interventions on a system to inform policy.
+One outcome of their study include estimates for the probability of cholera elimination under several possible vaccination scenarios.
+Mimicking their approach, we define cholera elimination as an absence of cholera infections for at least 52 consecutive weeks, and we provide forecasts under the following vaccination scenarios:
+
+\begin{itemize}
+  \item[$V0$:] No additional vaccines are administered.
+  \item[$V1$:] Vaccination limited to the departments of Centre and Artibonite, deployed over a two-year period.
+  \item[$V2$:] Vaccination limited to three departments: Artibonite, Centre, and Ouest deployed over a two-year period.
+  \item[$V3$:] Countrywide vaccination implemented over a five-year period.
+  \item[$V4$:] Countrywide vaccination implemented over a two-year period.
+\end{itemize}
+
+<<Load Model 1 VaccScen Sims, message=FALSE, include=FALSE>>=
+
+# Number of particles and simulations to use for pfilter and project from filter, respectively.
+h1_scen_NP   <- switch(RUN_LEVEL, 50, 500, 5000)
+h1_scen_sims <- switch(RUN_LEVEL, 20, 100, 1000)
+
+# Simulate vaccination scenario from filtering distribution. The code
+# below creates a model, samples from the filtering distribution, and then
+# projects the vaccination scenario, under the specified model.
+h1_S0_results <- bake(
+  file = paste0("model1/", rl_dir, "S0_sims.rds"), {
+
+    # Create the model
+    s0 <- haiti1_joint(vacscen = 'id0')
+
+    # Load coefficients
+    coef(s0) <- p1
+
+    # Get samples from filtering distribution
+    s0_pf <- pfilter(s0, save.state = TRUE, Np = h1_scen_NP)
+    end_states <- saved_states(s0_pf)
+    end_states <- end_states[[length(end_states)]]
+
+    # project vacc scenario, with filtering dist, and model.
+    h1_S0_sims <- project_from_filter(mod = s0, end_states = end_states, nsims = h1_scen_sims)
+
+    results <- list()
+    results$mod1_V0_sims <- agg_mod1_sims(h1_S0_sims)
+    results$mod1_V0_probs <- get_elimProbs(h1_S0_sims, model = 1)
+
+    results
+  },
+  timing = FALSE
+)
+
+# Simulate vaccination scenario from filtering distribution. The code
+# below creates a model, samples from the filtering distribution, and then
+# projects the vaccination scenario, under the specified model.
+h1_S2_results <- bake(
+  file = paste0('model1/', rl_dir, 'S2_sims.rds'), {
+
+    # Create the model
+    s2 <- haiti1_joint(vacscen = "id2")
+
+    # Set model parameters that need to be adjusted
+    depts <- 2
+    h1_par_names <- names(p1)
+    h1_params_temp <- c(p1, rep(0.0, 5 * depts))
+    names(h1_params_temp) <- c(
+      h1_par_names,
+      paste0("S", 1:depts, "_0"),
+      paste0("E", 1:depts, "_0"),
+      paste0("I", 1:depts, "_0"),
+      paste0("A", 1:depts, "_0"),
+      paste0("R", 1:depts, "_0")
+    )
+    coef(s2) <- h1_params_temp  # Set model parameters
+
+    # Get samples from filtering distribution for Scenario 2 model:
+    s2_pf <- pfilter(s2, save.state = TRUE, Np = h1_scen_NP)
+    end_states <- saved_states(s2_pf)
+    end_states <- end_states[[length(end_states)]]
+
+    # project vacc scenario, with filtering dist, and model.
+    h1_S2_sims <- project_from_filter(mod = s2, end_states = end_states, nsims = h1_scen_sims)
+
+    results <- list()
+    results$mod1_V1_sims <- agg_mod1_sims(h1_S2_sims)
+    results$mod1_V1_probs <- get_elimProbs(h1_S2_sims, model = 1)
+
+    results
+  },
+  timing = FALSE
+)
+
+# Simulate vaccination scenario from filtering distribution. The code
+# below creates a model, samples from the filtering distribution, and then
+# projects the vaccination scenario, under the specified model.
+h1_S4_results <- bake(
+  file = paste0('model1/', rl_dir, 'S4_sims.rds'), {
+
+    # Create the model
+    s4 <- haiti1_joint(vacscen = "id4")
+
+    # Set model parameters that need to be adjusted
+    depts <- 3
+    h1_par_names <- names(p1)
+    h1_params_temp <- c(p1, rep(0.0, 5 * depts))
+    names(h1_params_temp) <- c(
+      h1_par_names,
+      paste0("S", 1:depts, "_0"),
+      paste0("E", 1:depts, "_0"),
+      paste0("I", 1:depts, "_0"),
+      paste0("A", 1:depts, "_0"),
+      paste0("R", 1:depts, "_0")
+    )
+    coef(s4) <- h1_params_temp  # Set model parameters
+
+    # Get samples from filtering distribution for Scenario 2 model:
+    s4_pf <- pfilter(s4, save.state = TRUE, Np = h1_scen_NP)
+    end_states <- saved_states(s4_pf)
+    end_states <- end_states[[length(end_states)]]
+
+    # project vacc scenario, with filtering dist, and model.
+    h1_S4_sims <- project_from_filter(mod = s4, end_states = end_states, nsims = h1_scen_sims)
+
+    results <- list()
+    results$mod1_V2_sims <- agg_mod1_sims(h1_S4_sims)
+    results$mod1_V2_probs <- get_elimProbs(h1_S4_sims, model = 1)
+
+    results
+  },
+  timing = FALSE
+)
+
+# Simulate vaccination scenario from filtering distribution. The code
+# below creates a model, samples from the filtering distribution, and then
+# projects the vaccination scenario, under the specified model.
+h1_S3_results <- bake(
+  file = paste0('model1/', rl_dir, 'S3_sims.rds'), {
+
+    # Create the model
+    s3 <- haiti1_joint(vacscen = "id3")
+
+    # Set model parameters that need to be adjusted
+    depts <- 10
+    h1_par_names <- names(p1)
+    h1_params_temp <- c(p1, rep(0.0, 5 * depts))
+    names(h1_params_temp) <- c(
+      h1_par_names,
+      paste0("S", 1:depts, "_0"),
+      paste0("E", 1:depts, "_0"),
+      paste0("I", 1:depts, "_0"),
+      paste0("A", 1:depts, "_0"),
+      paste0("R", 1:depts, "_0")
+    )
+    coef(s3) <- h1_params_temp  # Set model parameters
+
+    # Get samples from filtering distribution for Scenario 2 model:
+    s3_pf <- pfilter(s3, save.state = TRUE, Np = h1_scen_NP)
+    end_states <- saved_states(s3_pf)
+    end_states <- end_states[[length(end_states)]]
+
+    # project vacc scenario, with filtering dist, and model.
+    h1_S3_sims <- project_from_filter(mod = s3, end_states = end_states, nsims = h1_scen_sims)
+
+    results <- list()
+    results$mod1_V3_sims <- agg_mod1_sims(h1_S3_sims)
+    results$mod1_V3_probs <- get_elimProbs(h1_S3_sims, model = 1)
+
+    results
+  },
+  timing = FALSE
+)
+
+# Simulate vaccination scenario from filtering distribution. The code
+# below creates a model, samples from the filtering distribution, and then
+# projects the vaccination scenario, under the specified model.
+h1_S1_results <- bake(
+  file = paste0('model1/', rl_dir, 'S1_sims.rds'), {
+
+    # Create the model
+    s1 <- haiti1_joint(vacscen = "id1")
+
+    # Set model parameters that need to be adjusted
+    depts <- 10
+    h1_par_names <- names(p1)
+    h1_params_temp <- c(p1, rep(0.0, 5 * depts))
+    names(h1_params_temp) <- c(
+      h1_par_names,
+      paste0("S", 1:depts, "_0"),
+      paste0("E", 1:depts, "_0"),
+      paste0("I", 1:depts, "_0"),
+      paste0("A", 1:depts, "_0"),
+      paste0("R", 1:depts, "_0")
+    )
+    coef(s1) <- h1_params_temp  # Set model parameters
+
+    # Get samples from filtering distribution for Scenario 2 model:
+    s1_pf <- pfilter(s1, save.state = TRUE, Np = h1_scen_NP)
+    end_states <- saved_states(s1_pf)
+    end_states <- end_states[[length(end_states)]]
+
+    # project vacc scenario, with filtering dist, and model.
+    h1_S1_sims <- project_from_filter(mod = s1, end_states = end_states, nsims = h1_scen_sims)
+
+    results <- list()
+    results$mod1_V4_sims <- agg_mod1_sims(h1_S1_sims)
+    results$mod1_V4_probs <- get_elimProbs(h1_S1_sims, model = 1)
+
+    results
+  },
+  timing = FALSE
+)
+@
+
+<<Load Model 3 Vacc Scenarios, message=FALSE, include=FALSE>>=
+NP_BPF <- switch(RUN_LEVEL, 50, 200, 2000)
+NSIM   <- switch(RUN_LEVEL, 10,  50, 1000)
+
+h3_bpf_end_states <- bake(
+  file = paste0("model3/", rl_dir, "h3_bpf_end_states.rds"),
+  expr = {
+    coef(h3Spat) <- p3
+    h3_bpf <- bpfilter(h3Spat, Np = NP_BPF, block_size = 1, save_states = TRUE)
+    ss <- h3_bpf@saved.states
+    ss[[length(ss)]]
+  },
+  timing = FALSE
+)
+
+mod3_V0_res <- bake(
+  file = paste0("model3/", rl_dir, "mod3_V0_res.rds"), {
+
+    # Create a list to store results
+    h3_V0_res <- list()
+
+    # Get "noVacc" scenario parameters
+    h3_S0_par <- get_model3_vacc_scenario_params(scenario = "noVacc")
+    coef(h3Spat) <- p3
+    coef(h3Spat)[names(h3_S0_par)] <- h3_S0_par
+
+    # Using parameters and filtered distribution, project system
+    h3_V0_sims <- project_from_filter(
+      h3Spat, end_states = h3_bpf_end_states, covarGen = project_rain,
+      nsims = NSIM, seed = 6897154
+    )
+
+    # Store nationally aggregated results
+    h3_V0_res$mod3_V0_sims <- agg_mod3_sims(h3_V0_sims)
+    h3_V0_res$mod3_V0_probs <- get_elimProbs(h3_V0_sims, model = 3)
+
+    # For the supplement material, save quantiles of the Susceptible individuals
+    h3_V0_res$SQuants <- h3_V0_sims %>%
+      mutate(
+        date = as.Date(lubridate::round_date(lubridate::date_decimal(time), unit = 'day'))
+      ) %>%
+      group_by(unitname, date) %>%
+      mutate(
+        pop = h3Spat@params[paste0("H", which(first(unitname) == unit_names(h3Spat)))]
+      ) %>%
+      summarize(
+        Q025 = quantile((S + VSd + VSdd + VSdd_alt + VSd_alt) / pop, probs = 0.025),
+        Q50 = quantile((S + VSd + VSdd + VSdd_alt + VSd_alt) / pop, probs = 0.5),
+        Q975 = quantile((S + VSd + VSdd + VSdd_alt + VSd_alt) / pop, probs = 0.975)
+      ) %>%
+      ungroup() %>%
+      rename(dep = unitname)
+
+    # Return results
+    h3_V0_res
+
+  }
+)
+
+mod3_V1_res <- bake(
+  file = paste0("model3/", rl_dir, "mod3_V1_res.rds"), {
+
+    # Create a list to store results
+    h3_V1_res <- list()
+
+    # Get "noVacc" scenario parameters
+    h3_V1_par <- get_model3_vacc_scenario_params(scenario = "2dep")
+    coef(h3Spat) <- p3
+    coef(h3Spat)[names(h3_V1_par)] <- h3_V1_par
+
+    # Using parameters and filtered distribution, project system
+    h3_V1_sims <- project_from_filter(
+      h3Spat, end_states = h3_bpf_end_states, covarGen = project_rain,
+      nsims = NSIM, seed = 6897154
+    )
+
+    # Store nationally aggregated results
+    h3_V1_res$mod3_V1_sims <- agg_mod3_sims(h3_V1_sims)
+    h3_V1_res$mod3_V1_probs <- get_elimProbs(h3_V1_sims, model = 3)
+
+    # Return results
+    h3_V1_res
+
+  }
+)
+
+mod3_V2_res <- bake(
+  file = paste0("model3/", rl_dir, "mod3_V2_res.rds"), {
+
+    # Create a list to store results
+    h3_V2_res <- list()
+
+    # Get "noVacc" scenario parameters
+    h3_V2_par <- get_model3_vacc_scenario_params(scenario = "3dep")
+    coef(h3Spat) <- p3
+    coef(h3Spat)[names(h3_V2_par)] <- h3_V2_par
+
+    # Using parameters and filtered distribution, project system
+    h3_V2_sims <- project_from_filter(
+      h3Spat, end_states = h3_bpf_end_states, covarGen = project_rain,
+      nsims = NSIM, seed = 6897154
+    )
+
+    # Store nationally aggregated results
+    h3_V2_res$mod3_V2_sims <- agg_mod3_sims(h3_V2_sims)
+    h3_V2_res$mod3_V2_probs <- get_elimProbs(h3_V2_sims, model = 3)
+
+    # Return results
+    h3_V2_res
+  }
+)
+
+mod3_V3_res <- bake(
+  file = paste0("model3/", rl_dir, "mod3_V3_res.rds"), {
+
+    # Create a list to store results
+    h3_V3_res <- list()
+
+    # Get "noVacc" scenario parameters
+    h3_V3_par <- get_model3_vacc_scenario_params(scenario = "slowNation")
+    coef(h3Spat) <- p3
+    coef(h3Spat)[names(h3_V3_par)] <- h3_V3_par
+
+    # Using parameters and filtered distribution, project system
+    h3_V3_sims <- project_from_filter(
+      h3Spat, end_states = h3_bpf_end_states, covarGen = project_rain,
+      nsims = NSIM, seed = 6897154
+    )
+
+    # Store nationally aggregated results
+    h3_V3_res$mod3_V3_sims <- agg_mod3_sims(h3_V3_sims)
+    h3_V3_res$mod3_V3_probs <- get_elimProbs(h3_V3_sims, model = 3)
+
+    # Return results
+    h3_V3_res
+
+  }
+)
+
+mod3_V4_res <- bake(
+  file = paste0("model3/", rl_dir, "mod3_V4_res.rds"), {
+
+    # Create a list to store results
+    h3_V4_res <- list()
+
+    # Get "noVacc" scenario parameters
+    h3_V4_par <- get_model3_vacc_scenario_params(scenario = "fastNation")
+    coef(h3Spat) <- p3
+    coef(h3Spat)[names(h3_V4_par)] <- h3_V4_par
+
+    # Using parameters and filtered distribution, project system
+    h3_V4_sims <- project_from_filter(
+      h3Spat, end_states = h3_bpf_end_states, covarGen = project_rain,
+      nsims = NSIM, seed = 6897154
+    )
+
+    # Store nationally aggregated results
+    h3_V4_res$mod3_V4_sims <- agg_mod3_sims(h3_V4_sims)
+    h3_V4_res$mod3_V4_probs <- get_elimProbs(h3_V4_sims, model = 3)
+
+    # Return results
+    h3_V4_res
+
+  }
+)
+@
+
+Simulations from probabilistic models (Models~1 and~3) represent possible trajectories of the dynamic system under the scientific assumptions of the models.
+Estimates of the probability of cholera elimination can therefore be obtained as the proportion of simulations from these models that result in cholera elimination.
+The results of these projections are summarized in Figs. \ref{fig:elimProbs}.
 
@@
 
-\begin{figure}
+\begin{figure}[!h]
 <<Elimination_Probs_Figure, fig.height=2.2>>=
@@
 Probability of elimination across simulations for a 10 year period.
-Compare to Fig.~3A of \cite{lee20}.}
+Compare to Fig.~3A of Lee et al.~\cite{lee20}.}
 \end{figure}
+
+Probability of elimination estimates of this form are not meaningful for deterministic models, as the trajectory of these models only represent the mean behavior of the system rather than individual potential outcomes.
+We therefore do not provide probability of elimination estimates under Model~2, but show trajectories under the various vaccination scenarios using this model (Fig.~\ref{fig:mod2Traj}).
 
@@
   & \eqref{model2:mu_EI}
-  & \multicolumn{1}{c}{\hspace{16pt}\missing}
+  & \missing
+  % & \multicolumn{1}{c}{\hspace{16pt}\missing}
   &
@@
 Water survival \newline (wk)
-  & \missing
+  & \multirow{2}{*}{\missing}
   % & \multicolumn{1}{c}{\hspace{16pt}\missing}
   &
-  & ${\fixed \Wremoval^{-1} = \Sexpr{signif(52/h2_fit$h2_params["Delta"],3)}}$
-  & \eqref{model2:from_W}
-  & $\Wremoval^{-1}=\Sexpr{myround(52/p3s["mu_B"],2)}$
-  & \eqref{eq:model3:Decay}
+  & \multirow{2}{*}{${\fixed \Wremoval^{-1} = \Sexpr{signif(52/h2_fit$h2_params["Delta"],3)}}$}
+  & \multirow{2}{*}{\eqref{model2:from_W}}
+  & \multirow{2}{*}{$\Wremoval^{-1}=\Sexpr{myround(52/p3s["mu_B"],2)}$}
+  & \multirow{2}{*}{\eqref{eq:model3:Decay}}
 \\
@@
 \hline
-Process noise \newline ($\text{wk}^{1/2}$)
+Process noise \newline (wk\textsuperscript{1/2})
   & $\sigmaProc =$ \newline $(\Sexpr{myround(sqrt(p1["sig_sq_epi"]),2)},
@@
 \\
-Observation variance
+Observation \newline variance
   & $\obsOverdispersion =$ \newline $(\Sexpr{myround(p1["tau_epi"],2)}, \Sexpr{round(p1["tau_end"],2)})$
@@
 \\\hline
-
 \end{tabular}
-\caption{References to the relevant equation are given in parentheses.
+\caption{\label{tab:params}References to the relevant equation are given in parentheses.
 Parameters in blue were fixed based on scientific reasoning and not fitted to the data.
@@
 
-% Place tables after the first paragraph in which they are cited.
-\begin{table}[!ht]
-\begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
-\centering
-\caption{
-{\bf Table caption Nulla mi mi, venenatis sed ipsum varius, volutpat euismod diam.}}
-\begin{tabular}{|l+l|l|l|l|l|l|l|}
-\hline
-\multicolumn{4}{|l|}{\bf Heading1} & \multicolumn{4}{|l|}{\bf Heading2}\\ \thickhline
-$cell1 row1$ & cell2 row 1 & cell3 row 1 & cell4 row 1 & cell5 row 1 & cell6 row 1 & cell7 row 1 & cell8 row 1\\ \hline
-$cell1 row2$ & cell2 row 2 & cell3 row 2 & cell4 row 2 & cell5 row 2 & cell6 row 2 & cell7 row 2 & cell8 row 2\\ \hline
-$cell1 row3$ & cell2 row 3 & cell3 row 3 & cell4 row 3 & cell5 row 3 & cell6 row 3 & cell7 row 3 & cell8 row 3\\ \hline
-\end{tabular}
-\begin{flushleft} Table notes Phasellus venenatis, tortor nec vestibulum mattis, massa tortor interdum felis, nec pellentesque metus tortor nec nisl.
+% TODO: Remove the unused code below.
+% % Place tables after the first paragraph in which they are cited.
+% \begin{table}[!ht]
+% \begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
+% \centering
+% \caption{
+% {\bf Table caption Nulla mi mi, venenatis sed ipsum varius, volutpat euismod diam.}}
+% \begin{tabular}{|l+l|l|l|l|l|l|l|}
+% \hline
+% \multicolumn{4}{|l|}{\bf Heading1} & \multicolumn{4}{|l|}{\bf Heading2}\\ \thickhline
+% $cell1 row1$ & cell2 row 1 & cell3 row 1 & cell4 row 1 & cell5 row 1 & cell6 row 1 & cell7 row 1 & cell8 row 1\\ \hline
+% $cell1 row2$ & cell2 row 2 & cell3 row 2 & cell4 row 2 & cell5 row 2 & cell6 row 2 & cell7 row 2 & cell8 row 2\\ \hline
+% $cell1 row3$ & cell2 row 3 & cell3 row 3 & cell4 row 3 & cell5 row 3 & cell6 row 3 & cell7 row 3 & cell8 row 3\\ \hline
+% \end{tabular}
+% \begin{flushleft} Table notes Phasellus venenatis, tortor nec vestibulum mattis, massa tortor interdum felis, nec pellentesque metus tortor nec nisl.
 Ut ornare mauris tellus, vel dapibus arcu suscipit sed.
-\end{flushleft}
-\label{table1}
-\end{adjustwidth}
-\end{table}
+% \end{flushleft}
+% \label{table1}
+% \end{adjustwidth}
+% \end{table}
 

### 2023-08-22 ms-plos.Rnw
Moved the figures and tables, made figure titles

@@
 \newcommand\code[1]{\texttt{#1}}
+\newcommand\figTitle{\bf}
 \newcommand\paramVec{\theta}
@@
 \caption{\label{fig:CholeraData}
-\small
-Weekly reported cholera cases in Haiti from October 2010 to January 2019.
+{\figTitle Weekly cholera cases.}
+Weekly reported cholera cases in Haiti from October 2010 to January 2019 for each of the 10 administrative departments.
 }
@@
 
+<<FitModel1, echo=FALSE, message=FALSE, include=FALSE>>=
+
+# This code chunk is used to fit Model 1 using the IF2 algorithm.
+set.seed(636813)
+
+# Check RUN_LEVEL, and set hyper-parameters to IF2 and pfilter accordingly.
+#                      RUN_LEVEL =    1,    2,      3
+NP_H1        <- switch(RUN_LEVEL,   100, 1000,   2000)
+NMIF_H1      <- switch(RUN_LEVEL,    10,   50,    200)
+NUM_TREND    <- switch(RUN_LEVEL,     5,   10,     90)
+NPROF        <- switch(RUN_LEVEL,     4,    4,     24)
+NP_EVAL      <- switch(RUN_LEVEL,   200, 1000,   5000)
+NREPS_EVAL   <- switch(RUN_LEVEL, cores,   20,  cores)
+
+# Create the <RUN_LEVEL> version of the results and save them.
+h1_MIF <- bake(
+  file = paste0("model1/", rl_dir, "haiti1_fit.rds"), {
+    fit_haiti1(  # haitipkg function used to fit Model 1
+      NP = NP_H1,
+      NMIF = NMIF_H1,
+      NUM_TREND = NUM_TREND,
+      NPROF = NPROF,
+      NREPS_EVAL = NREPS_EVAL,
+      NP_EVAL = NP_EVAL,
+      ncores = cores
+    )
+  }
+)
+
+# The result is a data.frame containing many parameter values and their
+# corresponding likelihood evaluation. To perform a likelihood profile,
+# we calculate the maximum likelihood for each value of betat.
+h1_prof_res <- h1_MIF %>%
+  group_by(betat) %>%
+  summarize(logLik = max(ll))
+
+mcap_results <- pomp::mcap(h1_prof_res$logLik, h1_prof_res$betat, span = 1)
+
+# Get the best set of parameters from MIF2 search.
+p1 <- h1_MIF %>%
+  filter(ll == max(ll)) %>%
+  select(-ll, -ll.se) %>%
+  unlist()
+
+# Save the likelihood of the best parameters, for use later.
+mod1_ll <- h1_MIF %>%
+  filter(ll == max(ll)) %>%
+  select(ll) %>%
+  unlist()
+
+# Just a bit of tidying up: removing parameters that we won't use again.
+rm(NP_H1, NMIF_H1, NUM_TREND, NPROF, NREPS_EVAL, NP_EVAL, h1_MIF)
+gc()
+
+@
+
+<<FitModel2, echo=FALSE, include=FALSE, message=FALSE>>=
+
+# This code chunk fits Model 2. This is a quick caclulation compared to models
+# 1 and 3, so no run_level is used.
+
+h2_fit <- bake(
+  file = paste0("model2/model2_fit.rds"), {
+    fit_haiti2()
+  },
+  timing = FALSE
+)
+@
+
+<<FitModel3, echo=FALSE, include=FALSE, message=FALSE>>=
+
+# This code chunk is used to fit Model 3 using the IBPF algorithm.
+set.seed(8636712)
+
+# Create vectors for the unit and shared parameters
+unit_specific_names <- c("betaB", "foi_add", "aHur", "hHur", "Iinit")
+shared_param_names <- c(
+  "mu_B", "XthetaA", "thetaI", "lambdaR", "r", "std_W",
+  "epsilon", "k"
+)
+est_param_names <- c(
+  unit_specific_names, shared_param_names
+)
+
+# Add unit numbers to each parameter
+est_param_names_expanded <- paste0(rep(est_param_names, each = 10), 1:10)
+
+# Simple function that is used to set the rw_sd for the first search
+# This function is convenient so that we don't have to type out the rw_sd
+# for each unit parameter.
+set_rw_1 <- function(x) {
+  if (gsub("[[:digit:]]+$", "", x) %in% shared_param_names) {
+    return(0.01)
+  } else if (x %in% paste0(rep(c("aHur", 'hHur'), each = 2), c(3, 9))) {
+    return(expression(ifelse(time >= 2016.754 & time <= 2017, 0.015, 0)))
+  } else if (x %in% c("Iinit3", "Iinit4")) {
+    return(expression(ivp(0.25)))
+  } else if (grepl('^foi_add[[:digit:]]+$', x)) {
+    return(0.015)
+  } else if (grepl('^betaB[[:digit:]]+$', x)) {
+    return(0.01)
+  } else {
+    return(0)
+  }
+}
+
+set_rw_2 <- function(x) {
+  if (gsub("[[:digit:]]+$", "", x) %in% shared_param_names) {
+    return(0.003)
+  } else if (x %in% paste0(rep(c("aHur", 'hHur'), each = 2), c(3, 9))) {
+    return(expression(ifelse(time >= 2016.754 & time <= 2017, 0.005, 0)))
+  } else if (x %in% c("Iinit3", "Iinit4")) {
+    return(expression(ivp(0.125)))
+  } else if (grepl('^foi_add[[:digit:]]+$', x)) {
+    return(0.005)
+  } else if (grepl('^betaB[[:digit:]]+$', x)) {
+    return(0.003)
+  } else {
+    return(0)
+  }
+}
+
+# First Search RW
+reg_rw_1.sd <- lapply(est_param_names_expanded, set_rw_1)
+names(reg_rw_1.sd) <- est_param_names_expanded
+chol_rw_1.sd <- do.call(rw_sd, reg_rw_1.sd)
+
+# Second Search RW
+reg_rw_2.sd <- lapply(est_param_names_expanded, set_rw_2)
+names(reg_rw_2.sd) <- est_param_names_expanded
+chol_rw_2.sd <- do.call(rw_sd, reg_rw_2.sd)
+
+if (RUN_LEVEL == 1){
+  SEARCH1 = list(
+    NBPF = 10,
+    NP = 200,
+    SPAT_REGRESSION = 0.05,
+    NREPS = 20,
+    NP_EVAL = 200,
+    NREPS_EVAL = 3,
+    RW_SD = chol_rw_1.sd,
+    COOLING = 0.5,
+    KEEP_TRACES = TRUE
+  )
+
+  SEARCH2 = list(
+    TOP_N = 8,
+    NBPF = 10,
+    NP = 200,
+    SPAT_REGRESSION = 0.05,
+    NREPS = 2,
+    NP_EVAL = 200,
+    NREPS_EVAL = 3,
+    RW_SD = chol_rw_2.sd,
+    COOLING = 0.5,
+    KEEP_TRACES = TRUE
+  )
+
+} else if (RUN_LEVEL == 2) {
+  SEARCH1 = list(
+    NBPF = 100,
+    NP = 1000,
+    SPAT_REGRESSION = 0.05,
+    NREPS = 72,
+    NP_EVAL = 1000,
+    NREPS_EVAL = 8,
+    RW_SD = chol_rw_1.sd,
+    COOLING = 0.5,
+    KEEP_TRACES = TRUE,
+    KEEP_LIK_MAT = TRUE
+  )
+
+  SEARCH2 = list(
+    TOP_N = 9,
+    NBPF = 100,
+    NP = 1000,
+    SPAT_REGRESSION = 0.05,
+    NREPS = 4,
+    NP_EVAL = 1000,
+    NREPS_EVAL = 9,
+    RW_SD = chol_rw_2.sd,
+    COOLING = 0.5,
+    KEEP_TRACES = TRUE,
+    KEEP_LIK_MAT = TRUE
+  )
+
+} else if (RUN_LEVEL == 3) {
+
+  SEARCH1 = list(
+    NBPF = 100,
+    NP = 2500,
+    SPAT_REGRESSION = 0.05,
+    NREPS = 144,
+    NP_EVAL = 2500,
+    NREPS_EVAL = 18,
+    RW_SD = chol_rw_1.sd,
+    COOLING = 0.5,
+    KEEP_TRACES = FALSE,
+    KEEP_LIKE_MAT = FALSE
+  )
+
+  SEARCH2 = list(
+    TOP_N = 12,
+    NBPF = 100,
+    NP = 2500,
+    SPAT_REGRESSION = 0.05,
+    NREPS = 6,
+    NP_EVAL = 2500,
+    NREPS_EVAL = 18,
+    RW_SD = chol_rw_2.sd,
+    COOLING = 0.5,
+    KEEP_TRACES = FALSE,
+    KEEP_LIK_MAT = TRUE
+  )
+}
+
+# Search hyper-parameters for each run-level and search. This sets
+# parameters like number of particles, number of BPF iterations, and
+# number of replications for each search.
+# h3_fit_parms <- set_h3_fit_parms(
+#   RUN_LEVEL = RUN_LEVEL,
+#   rw_1.sd = chol_rw_1.sd,
+#   rw_2.sd = chol_rw_2.sd
+# )
+
+# Calculate the number of parameters that are estimated:
+# This is calculated by the number of non-zero rw_sd values (including expressions), but then subtracting of repeated shared parameters.
+mod3_n_params <- sum(reg_rw_1.sd != 0, na.rm = TRUE) + sum(sapply(reg_rw_1.sd, is.expression)) - length(shared_param_names) * 9
+
+# haiti3_fit <- readRDS("model3/run_level_3/h3_revision.rds")
+
+# Fit Model 3
+haiti3_fit <- bake(
+  file = paste0("model3/", rl_dir, "haiti3_fit.rds"), {
+    fit_haiti3(
+      nsearches = 2L,
+      search1 = SEARCH1,
+      search2 = SEARCH2,
+      search_rho = FALSE,
+      search_gamma = FALSE,
+      search_Iinit = TRUE,
+      search_hur = TRUE,
+      ncores = cores
+    )
+  }
+)
+
+# haiti3_fit contains ll estimate for many parameter sets, so we want to find
+# the highest value
+mod3_ll <- max(
+  haiti3_fit[[length(haiti3_fit)]]$logLiks$logLik,
+  na.rm = TRUE
+)
+
+# Just a bit of tidying up: removing parameters that we won't use again.
+rm(
+  SEARCH1, SEARCH2, SEARCH3, N_SEARCHES,
+  est_param_names, est_param_names_expanded,
+  set_rw_1, set_rw_2, set_rw_3, reg_rw_1.sd, reg_rw_2.sd, reg_rw_3.sd,
+  chol_rw_1.sd, chol_rw_2.sd, chol_rw_3.sd
+)
+gc()
+
+@
+
+<<table-input,echo=FALSE,eval=T, include=FALSE, message=FALSE>>=
+
+# Find the index for parameters of model 3 that correspond to the MLE.
+best_m3 <- which.max(haiti3_fit[[length(haiti3_fit)]]$logLiks$logLik)
+
+# Save the parameters that correspond to the MLE of Model 3.
+p3 <- haiti3_fit[[length(haiti3_fit)]]$params[best_m3, ]
+
+p3u <- matrix(nrow = length(unit_specific_names), ncol = 10)
+rownames(p3u) <- unit_specific_names
+p3s <- numeric(length(shared_param_names))
+names(p3s) <- shared_param_names
+
+for (p in shared_param_names) {
+  p3s[p] <- p3[paste0(p, 1)]  # They are the same for 1:10
+}
+
+for (p in unit_specific_names) {
+  pat <- paste0("^", p, "[[:digit:]]+$")
+  p3u[p, ] <- p3[grepl(pat, names(p3))]
+}
+
+rm(unit_specific_names, shared_param_names, p, pat)
+
+stew(file = "models.rda", {
+  h1 <- haiti1_joint()
+  coef(h1) <- p1
+
+  h2_epi <- haiti2(region = 'before', cutoff = 10000, measure = 'log')
+  coef(h2_epi) <- h2_fit$h2_params
+
+  h3Spat <- haiti3_spatPomp()
+  coef(h3Spat) <- p3
+})
+@
+
+<<AICtableInput, echo=FALSE, eval=TRUE>>=
+NREPS_EVAL <- switch(RUN_LEVEL,   3,   8,    36)
+NP_EVAL    <- switch(RUN_LEVEL, 100, 500,  5000)
+
+# Objective function of model 2, epidemic phase
+epi_ofun <- traj_objfun(
+  h2_epi,
+  params = h2_fit$h2_params
+)
+
+mod2_epi_ll <- -epi_ofun(par = h2_fit$h2_params)
+mod2_ll <- mod2_epi_ll - sum(log(h2_epi@data + 1), na.rm = TRUE)
+
+arima_lik2 <- 0
+
+# model 2 AIC
+mod2_n_params <- h2_fit$n_fit_params
+mod2_aic <- 2 * mod2_n_params - 2 * mod2_ll
+
+# model 1 AIC
+mod1_n_params <- length(
+  c('beta1', 'beta2', 'beta3', 'beta4', 'beta5', 'beta6',
+    'rho', 'nu', 'E_0', 'I_0', 'tau_epi',
+    'tau_end', 'sig_sq_epi', 'sig_sq_end', 'betat')
+)
+mod1_aic <- 2 * mod1_n_params - 2 * mod1_ll
+
+# model 3 AIC
+mod3_aic <- 2 * mod3_n_params - 2 * mod3_ll
+
+# Calculate likelihood via block-particle filter
+mod3_eval_res <- bake(
+  file = paste0('model3/', rl_dir, 'mod3_eval_res.rds'), {
+
+    # Create a list to store results
+    results <- list()
+
+    coef(h3Spat) <- p3
+
+    # Perform block-particle filter
+    h3_bpf <- foreach(i = 1:NREPS_EVAL, .combine = c) %dopar% {
+      bpfilter(h3Spat, Np = NP_EVAL, block_size = 1)
+    }
+
+    ll <- logLik(h3_bpf)
+    results$ll <- logmeanexp(ll[!is.na(ll)], se = FALSE)
+
+    ### Get likelihood for subset of the data
+
+    # Find appropriate sub-set
+    in_subset_cols <- h3_bpf[[1]]@times >= as.numeric(lubridate::decimal_date(as.Date("2014-03-01")))
+    in_subset_rows <- h3_bpf[[1]]@unit_names != 'Ouest'
+
+    mod3_coupled_subset_evals <- c()
+    for (i in 1:length(h3_bpf)) {
+      mod3_coupled_subset_evals <- c(
+        mod3_coupled_subset_evals,
+        sum(h3_bpf[[i]]@block.cond.loglik[in_subset_rows, in_subset_cols])
+      )
+    }
+
+    results$subset_ll <- logmeanexp(mod3_coupled_subset_evals)
+    results
+  },
+  timing = FALSE
+)
+
+mod3_eval_res$subset_aic <- 2 * mod3_n_params - 2 * mod3_eval_res$subset_ll
+
+rm(NREPS_EVAL, NP_EVAL)
+@
+
+<<CalculateARMALikelihood, echo=FALSE, message=FALSE, include=FALSE, cache=TRUE>>=
+
+ARMA_benchmarks <- list()
+
+###
+### Model 1 ###
+###
+
+# Load aggregated data
+m1_agg_data <- haiti1_agg_data()
+
+# Fit ARMA(2, 1) on log-cases
+m1_log_arma <- arima(log(m1_agg_data$cases + 1), order = c(2, 0, 1))
+
+# Fit ARMA(2, 1) on natural scale of cases
+m1_arma <- arima(m1_agg_data$cases, order = c(2, 0, 1))
+
+# Check which one is better, and save benchmark as the better one
+if (m1_log_arma$loglik - sum(log(m1_agg_data$cases + 1), na.rm = TRUE) >= m1_arma$loglik) {
+  ARMA_benchmarks[['m1']] <- m1_log_arma$loglik - sum(log(m1_agg_data$cases + 1), na.rm = TRUE)
+} else {
+  ARMA_benchmarks[['m1']] <- m1_arma$loglik
+}
+
+rm(m1_agg_data, m1_log_arma, m1_arma)
+gc()
+
+###
+### Model 2 ###
+###
+
+h2_data <- haiti2_data(recreate = FALSE)
+
+# model 2 ARIMA benchmark
+h2_dep_names <- h2_data$department %>% unique()
+m2_benchmark <- 0
+for (i in 1:10){
+  cases <- h2_data[h2_data$department == h2_dep_names[i], ]$cases
+  log_cases <- log(cases + 1)
+
+  m2_arma_dep_ll <- arima(cases, order=c(2, 0, 1))$loglik
+  m2_log_arma_dep_ll <- arima(log_cases, order=c(2, 0, 1))$loglik - sum(log_cases, na.rm = TRUE)
+
+  m2_benchmark <- m2_benchmark + max(m2_arma_dep_ll, m2_log_arma_dep_ll)
+}
+
+ARMA_benchmarks[['m2']] <- m2_benchmark
+
+rm(m2_benchmark, h2_dep_names, cases, log_cases, m2_arma_dep_ll, m2_log_arma_dep_ll, h2_data)
+gc()
+
+###
+### Model 3 ###
+###
+
+mod3_cases <- as.data.frame(t(h3Spat@data))
+
+m3_benchmark <- 0
+for (dep in 1:10) {
+
+  # Get department cases
+  cases <- as.numeric(mod3_cases[, dep])
+
+  # Fit model on natural scale of cases
+  m3_a201 <- arima(cases, order = c(2, 0, 1))
+
+  # Fit model on log-scale of cases
+  m3_log_a201 <- arima(log(cases + 1), order = c(2, 0, 1))
+
+  m3_benchmark <- m3_benchmark + max(
+    m3_log_a201$loglik - sum(log(cases + 1), na.rm = TRUE),
+    m3_a201$loglik
+  )
+}
+
+ARMA_benchmarks[['m3']] <- m3_benchmark
+rm(cases, m3_a201, m3_log_a201, m3_benchmark, dep,
+   mod3_cases, temp, jumps, MODEL3_CASES, i, j)
+gc()
+@
+
+<<AIC Lee et al, echo=FALSE, include=FALSE, message=FALSE>>=
+lee3_NP   <- switch(RUN_LEVEL, 50, 250, 5000)
+lee3_NREP <- switch(RUN_LEVEL,  3,   8,   36)
+
+mod1_lee_n_params <- 20
+mod1_lee_aic <- 2 * mod1_lee_n_params - 2 * mod1_lee_ll
+
+mod2_lee_aic <- 2 * h2_fit$leeFit_n_params - 2 * h2_fit$leeFit_ll
+
+registerDoRNG(49637421)
+
+# Create Lee et al (2020a) version of model 3 on sub-data
+lee3_sub <- lee3_spatPomp()
+
+# This model is only fit for a subset of the data, so we will set data
+# that are not part of this subset as NA. This will allow us to estimate
+# the likelihood for only the values that were considered during the
+# parameter estimation phase.
+# set_as_na <- lee3_sub@times < lubridate::decimal_date(as.Date('2017-06-10'))
+# temp_ouest_cases <- lee3_sub@data['casesOuest', ]
+# temp_ouest_cases[set_as_na] <- NA_real_
+# lee3_sub@data['casesOuest', ] <- temp_ouest_cases
+
+lee3_all <- lee3_spatPomp(start_date = "2010-10-23")
+
+lee3_results <- bake(
+  file = paste0('model3/', rl_dir, 'lee3_results.rds'), {
+
+    # Create lists to save results
+    lee3_results <- list()
+
+    # Save how long it takes for sub-data
+    lee3_results[["sub_bpf_time"]] <- system.time({
+
+      lee3_sub_ll_evals <- foreach(
+        i = 1:lee3_NREP, .combine = rbind
+      ) %dopar% {
+
+        liks <- numeric(10L)
+
+        lee_bpf_out <- bpfilter(lee3_sub, block_size = 1, Np = lee3_NP)
+        Ouest_ind <- which(lee_bpf_out@unit_names == "Ouest")
+
+        for (u in 1:10) {
+          if (u == Ouest_ind) {
+            liks[u] <- sum(lee_bpf_out@block.cond.loglik[u, lee3_sub@times >= lubridate::decimal_date(as.Date('2017-06-10'))])
+          } else {
+            liks[u] <- sum(lee_bpf_out@block.cond.loglik[u, ])
+          }
+        }
+
+        liks
+      }
+    })
+
+    # Save all likelihood evaluations
+    lee3_results[["sub_ll_eval"]] <- logmeanexp(rowSums(lee3_sub_ll_evals)[!is.na(rowSums(lee3_sub_ll_evals))])
+
+    registerDoRNG(2789131)
+
+    # Save how long it takes for all data
+    lee3_results[["all_pf_time"]] <- system.time({
+      # Save all likelihood evaluations
+
+      lee3_all_ll_evals <- foreach(
+        i = 1:lee3_NREP, .combine = c
+      ) %dopar% {
+        logLik(bpfilter(lee3_all, block_size = 1, Np = lee3_NP))  # Get likelihood evaluations
+      }
+
+      lee3_results[["all_ll_eval"]] <- logmeanexp(lee3_all_ll_evals[!is.na(lee3_all_ll_evals)])
+
+    })
+
+    lee3_results[["n_params"]] <- 29
+
+    lee3_results
+  },
+  timing = FALSE
+)
+
+rm(lee3_sub, lee3_all, lee3_NP, lee3_NREP)
+gc()
+
+mod3_lee_coupled_aic <- 2 * lee3_results$n_params - 2 * lee3_results$all_ll_eval
+
+mod3_lee_coupled_partial_aic <- 2 * lee3_results$n_params - 2 * lee3_results$sub_ll_eval
+@
+
 \section*{Materials and methods}
@@
 However, it led to differences in notation, and in subsets of the data chosen for analysis, that hinder direct comparison.
-Here, we have put all three models into a common notational framework.
-Translations back to the original notation of Lee et al.~\cite{lee20} are given in Table~S-1 of the supplement.
+Here, we have created a common notational framework that facilitates model comparison, and put all comparable model parameters---including parameters that were estimated or held constant---into Table~\ref{tab:params}.
+% Here, we have put all three models into a common notational framework.
+Translations back to the original notation of Lee et al.~\cite{lee20} are given in Table~S-1 of the supplement. % TODO: Check S-1 table matches.
+
+\begin{table}
+\begin{adjustwidth}{-2.25in}{0in}
+\centering
+% \small
+\begin{tabular}{|p{0.15\linewidth}|p{0.14\linewidth}t|p{0.13\linewidth}t|p{0.17\linewidth}t|}
+\hline
+Mechanism & Model 1 && Model 2 && Model 3 &
+\\
+\hline
+\hline
+Infection (day)
+  & ${\fixed \muIR^{-1}=\Sexpr{myround(7/p1["gamma"],1)}}$
+  & \eqref{model1:toR}
+  & ${\fixed \muIR^{-1}=\Sexpr{myround(365/h2_fit$h2_params["gamma"],1)}}$
+  & \eqref{model2:mu_IR}
+  & ${\fixed \muIR^{-1}=\Sexpr{myround(365/p3["gamma1"],1)}}$
+  & \eqref{eq:model3:IR}
+\\
+Latency (day)
+  & ${\fixed \muEI^{-1}=\Sexpr{myround(7/p1["sigma"],1)}}$
+  & \eqref{model1:EA}
+  & ${\fixed \muEI^{-1}=\Sexpr{myround(365/h2_fit$h2_params["gammaE"],1)}}$
+  & \eqref{model2:mu_EI}
+  & \missing
+  % & \multicolumn{1}{c}{\hspace{16pt}\missing}
+  &
+\\
+\hline
+\multirow{3}{*}{Seasonality}
+  & $\transmission_{1:6}=(
+      \Sexpr{myround(p1["beta1"],1)},\allowbreak
+      \Sexpr{myround(p1["beta2"],1)},\allowbreak
+      \Sexpr{myround(p1["beta3"],1)},\allowbreak
+      \Sexpr{myround(p1["beta4"],1)},\allowbreak
+      \Sexpr{myround(p1["beta5"],1)},\allowbreak
+      \Sexpr{myround(p1["beta6"],1)})$
+  & \eqref{model1:beta}
+  & \multirow{2}{*}{${\fixed \seasAmplitude=0.4}$}
+  & \multirow{2}{*}{\eqref{model2:lambda}}
+  & \multirow{2}{*}{$\seasAmplitude = \Sexpr{myround(p3s["lambdaR"],2)}$}
+  & \multirow{2}{*}{\eqref{eq:model3:water}}
+  \\
+  & $\transmissionTrend = \Sexpr{myround(p1["betat"],2)}^*$
+  & \eqref{model1:betat}
+  & \multirow{-2}{*}{$\phaseParm = \Sexpr{myround(h2_fit$h2_params['phase'], 2)}^*$}
+  & \multirow{-2}{*}{\eqref{model2:lambda}}
+  & \multirow{-2}{*}{$\rainfallExponent = \Sexpr{signif(p3s["r"],3)}$}
+  & \multirow{-2}{*}{\eqref{eq:model3:water}}
+  \\\hline
+  \multirow{3}{*}{Immunity (yr)}
+  & \multirow{3}{*}{${\fixed \muRS^{-1}=\Sexpr{myround(1/p1["alpha"]/52,1)}}$}
+  & \multirow{3}{*}{\eqref{model1:RS}}
+  & $\muRS^{-1}= \Sexpr{signif(1/h2_fit$h2_params["sigma"],2)}$
+  & \eqref{model2:RS}
+  & \multirow{3}{*}{${\fixed \muRS^{-1}=\Sexpr{myround(1/p3['rho1'],1)}}$}
+  & \multirow{3}{*}{\eqref{eq:model3:RRnext}}
+\\
+& & & ${\fixed \omega_1^{-1} = \Sexpr{myround(1/h2_fit$h2_params["Omega1"],1)}}$ & \eqref{model2:omega1} & &\\
+& & & ${\fixed \omega_2^{-1} = \Sexpr{myround(1/h2_fit$h2_params["Omega2"],1)}}$ & \eqref{model2:omega2} & &\\
+\hline
+\multirow{2}{*}{Birth/death (yr)}
+  & ${\fixed \muBirth^{-1} = \Sexpr{myround(1/p1["mu"]/52,1)}}$
+    %
+  & \eqref{model1:death}
+  & \multirow{2}{*}{\missing}
+  &
+  & \multirow{2}{*}{${\fixed \muDeath^{-1} = \Sexpr{myround(1/p3["mu1"],1)}}$}
+  & \multirow{2}{*}{\eqref{eq:model3:IS}}
+\\
+& ${\fixed \muDeath^{-1} = \Sexpr{myround(1/p1["delta"]/52,1)}}$
+& \eqref{model1:death}
+&
+&
+& &
+\\
+Sympt. frac.
+  & ${\fixed \symptomFrac_z(t)=c\vaccineEfficacy^*(t-\tau_d)}$
+  & (\ref{model1:EI}-\ref{model1:EA})
+  & ${\fixed \symptomFrac=\Sexpr{myround(h2_fit$h2_params["k"],1)}}$
+  & \eqref{model2:mu_EI}
+  & ${\fixed \symptomFrac=\Sexpr{myround(p3["sigma1"],2)}}$
+  & \eqref{eq:model3:SA}
+\\
+Asympt.
+  & \multirow{2}{*}{${\fixed \asymptomRelativeInfect=0.05}$}
+  & \multirow{2}{*}{\eqref{model1:lambda}}
+  & ${\fixed \asymptomRelativeInfect = 0.001}$
+  & \eqref{model2:lambda}
+  &  ${\fixed \asymptomRelativeInfect = 1}$
+  & \eqref{eq:model3:foi}
+\\
+infectivity
+&
+&
+& ${\fixed \asymptomRelativeShed = \Sexpr{1e-7} }$
+& \eqref{model2:to_W}
+& $\asymptomRelativeShed = \Sexpr{myround(p3s["XthetaA"],3)}$
+& \eqref{eq:model3:water}
+\\
+\hline
+\multirow{4}{*}{Human to human}
+  & \multirow{4}{*}{$\transmission_{1:6}$ as above}
+  & \multirow{4}{*}{\eqref{model1:lambda}}
+  & \multirow{4}{*}{$\transmission=$\Sexpr{signif(h2_fit$h2_params["Beta"],3)}}
+  & \multirow{4}{*}{\eqref{model2:lambda}}
+  & $\transmission_{1:10}=(
+  \Sexpr{myround(p3u["foi_add",1]*1e6,2)}, \allowbreak
+	\Sexpr{myround(p3u["foi_add",2]*1e6,2)}, \allowbreak
+	\Sexpr{myround(p3u["foi_add",3]*1e6,2)}, \allowbreak
+  \Sexpr{myround(p3u["foi_add",4]*1e6,2)}, \allowbreak
+	\Sexpr{myround(p3u["foi_add",5]*1e6,2)}, \allowbreak
+	\Sexpr{myround(p3u["foi_add",6]*1e6,2)}, \allowbreak
+	\Sexpr{myround(p3u["foi_add",7]*1e6,2)}, \allowbreak
+	\Sexpr{myround(p3u["foi_add",8]*1e6,2)}, \allowbreak
+	\Sexpr{myround(p3u["foi_add",9]*1e6,2)}, \allowbreak
+	\Sexpr{myround(p3u["foi_add",10]*1e6,2)}
+	) \allowbreak \times 10^{-6}$
+  & \multirow{4}{*}{\eqref{eq:model3:foi}}
+\\
+\hline
+\multirow{5}{*}{Water to human}
+  & \multirow{5}{*}{\missing}
+  % & \multicolumn{1}{c}{\hspace{16pt}\missing}
+  &
+  & \multirow{3}{*}{${\fixed \Wsat = \Sexpr{signif(h2_fit$h2_params["Sat"],2)}}$}
+  & \multirow{3}{*}{\eqref{model2:lambda}}
+  % \eqref{model2:lambda}
+  & $\Wbeta{_{1:10}}= (
+        \Sexpr{myround(p3u["betaB",1],2)}, \allowbreak
+        \Sexpr{myround(p3u["betaB",2],2)}, \allowbreak
+        \Sexpr{myround(p3u["betaB",3],2)}, \allowbreak
+        \Sexpr{myround(p3u["betaB",4],2)}, \allowbreak
+        \Sexpr{myround(p3u["betaB",5],2)}, \allowbreak
+        \Sexpr{myround(p3u["betaB",6],2)}, \allowbreak
+        \Sexpr{myround(p3u["betaB",7],2)}, \allowbreak
+        \Sexpr{myround(p3u["betaB",8],2)}, \allowbreak
+        \Sexpr{myround(p3u["betaB",9],2)}, \allowbreak
+        \Sexpr{myround(p3u["betaB",10],2)})$
+  & \multirow{4}{*}{\eqref{eq:model3:foi}}
+\\
+& & & \multirow{-3}{*}{$\beta_W= \Sexpr{signif(h2_fit$h2_params["BetaW"],3)}$} & \multirow{-3}{*}{\eqref{model2:lambda}} & &
+\\
+\hline
+Human to water
+  & \missing
+  % & \multicolumn{1}{c}{\hspace{16pt}\missing}
+  &
+  & $\Wshed = $ \Sexpr{signif(h2_fit$h2_params["Mu"],3)}
+  & \eqref{model2:to_W}
+  & $\Wshed= \Sexpr{signif(p3s["thetaI"],3)}$
+  & \eqref{eq:model3:water}
+\\
+Water survival \newline (wk)
+  & \multirow{2}{*}{\missing}
+  % & \multicolumn{1}{c}{\hspace{16pt}\missing}
+  &
+  & \multirow{2}{*}{${\fixed \Wremoval^{-1} = \Sexpr{signif(52/h2_fit$h2_params["Delta"],3)}}$}
+  & \multirow{2}{*}{\eqref{model2:from_W}}
+  & \multirow{2}{*}{$\Wremoval^{-1}=\Sexpr{myround(52/p3s["mu_B"],2)}$}
+  & \multirow{2}{*}{\eqref{eq:model3:Decay}}
+\\
+Mixing exponent
+  & $\mixExponent=\Sexpr{myround(p1["nu"],2)}$
+  & \eqref{model1:lambda}
+  & \missing
+  % & \multicolumn{1}{c}{\hspace{16pt}\missing}
+  &
+  & \missing
+  % & \multicolumn{1}{c}{\hspace{16pt}\missing}
+  &
+\\
+\hline
+Process noise \newline (wk\textsuperscript{1/2})
+  & $\sigmaProc =$ \newline $(\Sexpr{myround(sqrt(p1["sig_sq_epi"]),2)},
+  \Sexpr{myround(sqrt(p1["sig_sq_end"]),2)})^*$
+  & \multirow{2}{*}{\eqref{model1:lambda}}
+  & \multirow{2}{*}{\missing}
+  % & \multicolumn{1}{c}{\hspace{16pt}\missing}
+  &
+  & \multirow{2}{*}{$\sigmaProc= \Sexpr{myround(p3s["std_W"],3)}$}
+  & \multirow{2}{*}{\eqref{eq:model3:SA}}
+\\
+Reporting rate
+  & $\reportRate=\Sexpr{myround(p1["rho"], 3)}$
+  & (S15) % TODO: Check
+  & ${\fixed \reportRate=\Sexpr{myround(h2_fit$h2_params["Rho"],2)}}$
+  & (S16) % TODO: Check
+  & $\reportRate=\Sexpr{myround(p3s["epsilon"],2)}$
+  & (S19) % TODO: Check
+\\
+Observation \newline variance
+  & $\obsOverdispersion =$ \newline $(\Sexpr{myround(p1["tau_epi"],2)}, \Sexpr{round(p1["tau_end"],2)})$
+  & \multirow{2}{*}{(S15)} % TODO: Check
+  & \multirow{2}{*}{$\obsOverdispersion = \Sexpr{signif(h2_fit$h2_params['sigma'], 3)}$}
+  & \multirow{2}{*}{(S16)} % TODO: Check
+  & \multirow{2}{*}{$\obsOverdispersion = \Sexpr{myround(p3s["k"],2)}$}
+  & \multirow{2}{*}{(S19)} % TODO: Check
+\\
+\hline
+\multirow{2}{*}{Initial Values} & $I_{0,0} = \Sexpr{round(prod(p1[c('pop_0', 'I_0')]))}$
+  &
+  & \multirow{2}{*}{\missing}
+  % & \multicolumn{1}{c}{\multirow{2}{*}{\hspace{16pt}\missing}}
+  &
+  & \multirow{2}{*}{$I_{3,7,10;0}(0)= (
+        \Sexpr{round(prod(p3[c('H3', 'Iinit3')]))}, \Sexpr{round(prod(p3[c('H7', 'Iinit7')]))}, \Sexpr{round(prod(p3[c('H10', 'Iinit10')]))})^*$}
+  &  % TODO: \multirow{2}{*}{EQUATION NUMBER}
+  \\
+  & $E_{0,0} = \Sexpr{round(prod(p1[c('pop_0', 'E_0')]))}$ & & \multicolumn{1}{c}{}
+  &
+  &
+  &
+\\\hline
+\multirow{3}{*}{Hurricane}
+  & \multirow{4}{*}{\missing}
+% & \multicolumn{1}{c}{\multirow{2}{*}{\hspace{16pt}\missing}}
+  &
+  & \multirow{4}{*}{\missing}
+  % & \multicolumn{1}{c}{\multirow{2}{*}{\hspace{16pt}\missing}}
+  &
+  & $\Whur{3, 9} =$\newline $(\Sexpr{myround(p3['aHur3'], 2)}, \Sexpr{myround(p3['aHur9'], 2)})^*$
+  & \multirow{2}{*}{\eqref{eq:model3:foi}}
+\\
+ \multirow{-1}{*}{Parameters} &
+  % & \multicolumn{1}{c}{}
+  &
+  &
+  % & \multicolumn{1}{c}{}
+  &
+  & $\hHur{3, 9} =$\newline $(\Sexpr{myround(p3['hHur3'], 2)}, \Sexpr{myround(p3['hHur9'], 2)})^*$
+  & \multirow{2}{*}{\eqref{eq:model3:foi}}
+\\\hline
+\end{tabular}
+\caption{\label{tab:params}References to the relevant equation are given in parentheses.
+Parameters in blue were fixed based on scientific reasoning and not fitted to the data.
+$^*$ denotes parameters added during our re-analysis, not considered by Lee et al.
+Translations back into the notation of \cite{lee20} are given in Table~S1.
+}
+\end{adjustwidth}
+\end{table}
 
@@
 
-<<FitModel1, echo=FALSE, message=FALSE, include=FALSE>>=
-
-# This code chunk is used to fit Model 1 using the IF2 algorithm.
-set.seed(636813)
-
-# Check RUN_LEVEL, and set hyper-parameters to IF2 and pfilter accordingly.
-#                      RUN_LEVEL =    1,    2,      3
-NP_H1        <- switch(RUN_LEVEL,   100, 1000,   2000)
-NMIF_H1      <- switch(RUN_LEVEL,    10,   50,    200)
-NUM_TREND    <- switch(RUN_LEVEL,     5,   10,     90)
-NPROF        <- switch(RUN_LEVEL,     4,    4,     24)
-NP_EVAL      <- switch(RUN_LEVEL,   200, 1000,   5000)
-NREPS_EVAL   <- switch(RUN_LEVEL, cores,   20,  cores)
-
-# Create the <RUN_LEVEL> version of the results and save them.
-h1_MIF <- bake(
-  file = paste0("model1/", rl_dir, "haiti1_fit.rds"), {
-    fit_haiti1(  # haitipkg function used to fit Model 1
-      NP = NP_H1,
-      NMIF = NMIF_H1,
-      NUM_TREND = NUM_TREND,
-      NPROF = NPROF,
-      NREPS_EVAL = NREPS_EVAL,
-      NP_EVAL = NP_EVAL,
-      ncores = cores
-    )
-  }
-)
-
-# The result is a data.frame containing many parameter values and their
-# corresponding likelihood evaluation. To perform a likelihood profile,
-# we calculate the maximum likelihood for each value of betat.
-h1_prof_res <- h1_MIF %>%
-  group_by(betat) %>%
-  summarize(logLik = max(ll))
-
-mcap_results <- pomp::mcap(h1_prof_res$logLik, h1_prof_res$betat, span = 1)
-
-# Get the best set of parameters from MIF2 search.
-p1 <- h1_MIF %>%
-  filter(ll == max(ll)) %>%
-  select(-ll, -ll.se) %>%
-  unlist()
-
-# Save the likelihood of the best parameters, for use later.
-mod1_ll <- h1_MIF %>%
-  filter(ll == max(ll)) %>%
-  select(ll) %>%
-  unlist()
-
-# Just a bit of tidying up: removing parameters that we won't use again.
-rm(NP_H1, NMIF_H1, NUM_TREND, NPROF, NREPS_EVAL, NP_EVAL, h1_MIF)
-gc()
-
-@
-
-<<FitModel2, echo=FALSE, include=FALSE, message=FALSE>>=
-
-# This code chunk fits Model 2. This is a quick caclulation compared to models
-# 1 and 3, so no run_level is used.
-
-h2_fit <- bake(
-  file = paste0("model2/model2_fit.rds"), {
-    fit_haiti2()
-  },
-  timing = FALSE
-)
-@
-
-<<FitModel3, echo=FALSE, include=FALSE, message=FALSE>>=
-
-# This code chunk is used to fit Model 3 using the IBPF algorithm.
-set.seed(8636712)
-
-# Create vectors for the unit and shared parameters
-unit_specific_names <- c("betaB", "foi_add", "aHur", "hHur", "Iinit")
-shared_param_names <- c(
-  "mu_B", "XthetaA", "thetaI", "lambdaR", "r", "std_W",
-  "epsilon", "k"
-)
-est_param_names <- c(
-  unit_specific_names, shared_param_names
-)
-
-# Add unit numbers to each parameter
-est_param_names_expanded <- paste0(rep(est_param_names, each = 10), 1:10)
-
-# Simple function that is used to set the rw_sd for the first search
-# This function is convenient so that we don't have to type out the rw_sd
-# for each unit parameter.
-set_rw_1 <- function(x) {
-  if (gsub("[[:digit:]]+$", "", x) %in% shared_param_names) {
-    return(0.01)
-  } else if (x %in% paste0(rep(c("aHur", 'hHur'), each = 2), c(3, 9))) {
-    return(expression(ifelse(time >= 2016.754 & time <= 2017, 0.015, 0)))
-  } else if (x %in% c("Iinit3", "Iinit4")) {
-    return(expression(ivp(0.25)))
-  } else if (grepl('^foi_add[[:digit:]]+$', x)) {
-    return(0.015)
-  } else if (grepl('^betaB[[:digit:]]+$', x)) {
-    return(0.01)
-  } else {
-    return(0)
-  }
-}
-
-set_rw_2 <- function(x) {
-  if (gsub("[[:digit:]]+$", "", x) %in% shared_param_names) {
-    return(0.003)
-  } else if (x %in% paste0(rep(c("aHur", 'hHur'), each = 2), c(3, 9))) {
-    return(expression(ifelse(time >= 2016.754 & time <= 2017, 0.005, 0)))
-  } else if (x %in% c("Iinit3", "Iinit4")) {
-    return(expression(ivp(0.125)))
-  } else if (grepl('^foi_add[[:digit:]]+$', x)) {
-    return(0.005)
-  } else if (grepl('^betaB[[:digit:]]+$', x)) {
-    return(0.003)
-  } else {
-    return(0)
-  }
-}
-
-# First Search RW
-reg_rw_1.sd <- lapply(est_param_names_expanded, set_rw_1)
-names(reg_rw_1.sd) <- est_param_names_expanded
-chol_rw_1.sd <- do.call(rw_sd, reg_rw_1.sd)
-
-# Second Search RW
-reg_rw_2.sd <- lapply(est_param_names_expanded, set_rw_2)
-names(reg_rw_2.sd) <- est_param_names_expanded
-chol_rw_2.sd <- do.call(rw_sd, reg_rw_2.sd)
-
-if (RUN_LEVEL == 1){
-  SEARCH1 = list(
-    NBPF = 10,
-    NP = 200,
-    SPAT_REGRESSION = 0.05,
-    NREPS = 20,
-    NP_EVAL = 200,
-    NREPS_EVAL = 3,
-    RW_SD = chol_rw_1.sd,
-    COOLING = 0.5,
-    KEEP_TRACES = TRUE
-  )
-
-  SEARCH2 = list(
-    TOP_N = 8,
-    NBPF = 10,
-    NP = 200,
-    SPAT_REGRESSION = 0.05,
-    NREPS = 2,
-    NP_EVAL = 200,
-    NREPS_EVAL = 3,
-    RW_SD = chol_rw_2.sd,
-    COOLING = 0.5,
-    KEEP_TRACES = TRUE
-  )
-
-} else if (RUN_LEVEL == 2) {
-  SEARCH1 = list(
-    NBPF = 100,
-    NP = 1000,
-    SPAT_REGRESSION = 0.05,
-    NREPS = 72,
-    NP_EVAL = 1000,
-    NREPS_EVAL = 8,
-    RW_SD = chol_rw_1.sd,
-    COOLING = 0.5,
-    KEEP_TRACES = TRUE,
-    KEEP_LIK_MAT = TRUE
-  )
-
-  SEARCH2 = list(
-    TOP_N = 9,
-    NBPF = 100,
-    NP = 1000,
-    SPAT_REGRESSION = 0.05,
-    NREPS = 4,
-    NP_EVAL = 1000,
-    NREPS_EVAL = 9,
-    RW_SD = chol_rw_2.sd,
-    COOLING = 0.5,
-    KEEP_TRACES = TRUE,
-    KEEP_LIK_MAT = TRUE
-  )
-
-} else if (RUN_LEVEL == 3) {
-
-  SEARCH1 = list(
-    NBPF = 100,
-    NP = 2500,
-    SPAT_REGRESSION = 0.05,
-    NREPS = 144,
-    NP_EVAL = 2500,
-    NREPS_EVAL = 18,
-    RW_SD = chol_rw_1.sd,
-    COOLING = 0.5,
-    KEEP_TRACES = FALSE,
-    KEEP_LIKE_MAT = FALSE
-  )
-
-  SEARCH2 = list(
-    TOP_N = 12,
-    NBPF = 100,
-    NP = 2500,
-    SPAT_REGRESSION = 0.05,
-    NREPS = 6,
-    NP_EVAL = 2500,
-    NREPS_EVAL = 18,
-    RW_SD = chol_rw_2.sd,
-    COOLING = 0.5,
-    KEEP_TRACES = FALSE,
-    KEEP_LIK_MAT = TRUE
-  )
-
-  # SEARCH3 = list(  # Not used, but was used for some preliminary results.
-  #   TOP_N = 9,
-  #   NBPF = 100,
-  #   NP = 2000,
-  #   SPAT_REGRESSION = 0.05,
-  #   NREPS = 8,
-  #   NP_EVAL = 2000,
-  #   NREPS_EVAL = 18,
-  #   RW_SD = rw_3.sd,
-  #   COOLING = 0.5,
-  #   KEEP_TRACES = FALSE,
-  #   KEEP_LIK_MAT = TRUE
-  # )
-}
-
-# Search hyper-parameters for each run-level and search. This sets
-# parameters like number of particles, number of BPF iterations, and
-# number of replications for each search.
-# h3_fit_parms <- set_h3_fit_parms(
-#   RUN_LEVEL = RUN_LEVEL,
-#   rw_1.sd = chol_rw_1.sd,
-#   rw_2.sd = chol_rw_2.sd
-# )
-
-# Calculate the number of parameters that are estimated:
-# This is calculated by the number of non-zero rw_sd values (including expressions), but then subtracting of repeated shared parameters.
-mod3_n_params <- sum(reg_rw_1.sd != 0, na.rm = TRUE) + sum(sapply(reg_rw_1.sd, is.expression)) - length(shared_param_names) * 9
-
-# haiti3_fit <- readRDS("model3/run_level_3/h3_revision.rds")
-
-# Fit Model 3
-haiti3_fit <- bake(
-  file = paste0("model3/", rl_dir, "haiti3_fit.rds"), {
-    fit_haiti3(
-      nsearches = 2L,
-      search1 = SEARCH1,
-      search2 = SEARCH2,
-      search_rho = FALSE,
-      search_gamma = FALSE,
-      search_Iinit = TRUE,
-      search_hur = TRUE,
-      ncores = cores
-    )
-  }
-)
-
-# haiti3_fit contains ll estimate for many parameter sets, so we want to find
-# the highest value
-mod3_ll <- max(
-  haiti3_fit[[length(haiti3_fit)]]$logLiks$logLik,
-  na.rm = TRUE
-)
-
-# Just a bit of tidying up: removing parameters that we won't use again.
-rm(
-  SEARCH1, SEARCH2, SEARCH3, N_SEARCHES,
-  est_param_names, est_param_names_expanded,
-  set_rw_1, set_rw_2, set_rw_3, reg_rw_1.sd, reg_rw_2.sd, reg_rw_3.sd,
-  chol_rw_1.sd, chol_rw_2.sd, chol_rw_3.sd
-)
-gc()
-
-@
-
-<<table-input,echo=FALSE,eval=T, include=FALSE, message=FALSE>>=
-
-# Find the index for parameters of model 3 that correspond to the MLE.
-best_m3 <- which.max(haiti3_fit[[length(haiti3_fit)]]$logLiks$logLik)
-
-# Save the parameters that correspond to the MLE of Model 3.
-p3 <- haiti3_fit[[length(haiti3_fit)]]$params[best_m3, ]
-
-p3u <- matrix(nrow = length(unit_specific_names), ncol = 10)
-rownames(p3u) <- unit_specific_names
-p3s <- numeric(length(shared_param_names))
-names(p3s) <- shared_param_names
-
-for (p in shared_param_names) {
-  p3s[p] <- p3[paste0(p, 1)]  # They are the same for 1:10
-}
-
-for (p in unit_specific_names) {
-  pat <- paste0("^", p, "[[:digit:]]+$")
-  p3u[p, ] <- p3[grepl(pat, names(p3))]
-}
-
-rm(unit_specific_names, shared_param_names, p, pat)
-
-stew(file = "models.rda", {
-  h1 <- haiti1_joint()
-  coef(h1) <- p1
-
-  h2_epi <- haiti2(region = 'before', cutoff = 10000, measure = 'log')
-  coef(h2_epi) <- h2_fit$h2_params
-
-  h3Spat <- haiti3_spatPomp()
-  coef(h3Spat) <- p3
-})
-@
-
 \subsection*{Model Fitting}\label{sec:model_fitting}
@@
 In this article we calibrate the parameters of each of our models by maximizing the likelihood, as described in Eq.~\myeqref{eq:likedef}.
+The likelihood for each of the fitted models---and the corresponding AIC values for model comparisons that include an adjustment for the number of calibrated parameters---is provided in Table~\ref{tab:likes}
 In the following subsections we describe in detail our approach to calibrating the three proposed mechanistic models to observed cholera incidence data.
+
+\begin{table}[!h]
+\centering
+
+\begin{tabular}{|c|c|c|c|}
+\hline
+ & \thead{Model~1} & \thead{Model~2} & \thead{Model~3}
+\\
+\hline
+\hline
+\multirow{2}{*}{Log-likelihood} &
+  $\Sexpr{myround(mod1_ll, 1)}$ &
+  $\Sexpr{myround(mod2_ll, 1)}$ &
+  $\Sexpr{myround(mod3_ll, 1)}$ \\
+    & ($\Sexpr{myround(mod1_lee_ll, 1)}$)\footnotemark[1] &
+  ($\Sexpr{myround(h2_fit$leeFit_ll, 1)}$) &
+  ($\Sexpr{myround(lee3_results[["all_ll_eval"]], 1)}$)\footnotemark[2]
+\\
+\hline
+Number of &
+  $\Sexpr{as.character(mod1_n_params)}$ &
+  $\Sexpr{as.character(mod2_n_params)}$ &
+  $\Sexpr{as.character(mod3_n_params)}$ \\
+ Fit Parameters &
+ (\Sexpr{as.character(mod1_lee_n_params)}) &
+ ($\Sexpr{as.character(h2_fit$leeFit_n_params)}$) &
+ ($\Sexpr{as.character(lee3_results$n_params, 1)}$)
+\\
+\hline
+\multirow{2}{*}{AIC} &
+  $\Sexpr{myround(mod1_aic, 1)}$ &
+  $\Sexpr{myround(mod2_aic, 1)}$ &
+  $\Sexpr{myround(mod3_aic, 1)}$ \\
+  & ($\Sexpr{myround(mod1_lee_aic, 1)}$)\footnotemark[1] &
+  ($\Sexpr{myround(mod2_lee_aic, 1)}$) &
+  ($\Sexpr{myround(mod3_lee_coupled_aic, 1)}$)\footnotemark[2]
+\\
+\hline
+\thead{Log-ARMA(2,1) \\ AIC} &
+  $\Sexpr{myround(-2 * ARMA_benchmarks$m1 + 2 * 5, 1)}$ &
+  $\Sexpr{myround(-2 * ARMA_benchmarks$m2 + 2 * 5, 1)}$ &
+  $\Sexpr{myround(-2 * ARMA_benchmarks$m3 + 2 * 5, 1)}$
+\\
+\hline
+\end{tabular}
+\caption{\label{tab:likes}AIC values for each model compared to their ARMA benchmark.
+Values in parentheses are corresponding values using Lee et al.~\cite{lee20} parameter estimates.
+$\textsuperscript{1}$The reported likelihood is an upper bound of the likelihood of the Lee et al.~\cite{lee20} model.
+$\textsuperscript{2}$ Lee et al.~\cite{lee20} fit Model~3 to a subset of the data (March 2014 onward, excluding data from Ouest in 2015-2016).
+On this subset, their model has a likelihood of $\Sexpr{myround(lee3_results$sub_ll_eval, 1)}$.
+On this same subset, our model has a likelihood of $\Sexpr{myround(mod3_eval_res$subset_ll, 1)}$.
+See Sec.~S6 of the supplement material for more details on estimating the likelihood of the Lee et al.~\cite{lee20} models.
+}
+\end{table}
 
@@
 \caption{\label{fig:betat}
-Monte Carlo adjusted profile of $\transmissionTrend$.
+{\figTitle Confidence inverval for linear trend in transmission.}
+Monte Carlo adjusted profile of $\transmissionTrend$ for Model~1.
 The blue curve is the profile, the blue line indicates the MLE, and the dashed lines indicate the $95\%$ confidence interval.
@@
 \caption{\label{fig:mod1fit}
-Simulations from Model~1 compared to reported cholera cases.
+{\figTitle Simulations from Model~1 compared to reported cholera cases.}
 The black curve is observed data, the blue curve is median of 500 simulations from the fitted model, and the vertical dashed line represents break-point when parameters are refit.
@@
 \caption{\label{fig:mod2Traj}
-Simulated trajectory of Model~2 (blue curve) and projections under the various vaccination scenarios.
-Reported cholera incidence is shown in black.}
+{\figTitle Simulated trajectory of Model~2.}
+The black line shows the nationally aggregated weekly cholera incidence data.
+The blue curve from 2012-2019 is the trajectory of the calibrated version of Model~2.
+Projections under the various vaccination scenarios, which are discussed in the \bf{Forecasts} subsection are also included.}
 \end{figure}
@@
 \caption{\label{fig:h3spatsims}
+{\figTitle Simulations from Model~3 compared to reported cholera cases.}
 Simulations from initial conditions using the spatially coupled version of Model~3.
@@
 Large advantages (or shortcomings) in model likelihood relative to the benchmark for a given spatial scale that are not present in other spatial scales may provide weak evidence for (or against) the statistical fit of models across a range of spatial resolutions.
-
-<<Table 2 Input, echo=FALSE, eval=TRUE>>=
-NREPS_EVAL <- switch(RUN_LEVEL,   3,   8,    36)
-NP_EVAL    <- switch(RUN_LEVEL, 100, 500,  5000)
-
-# Objective function of model 2, epidemic phase
-epi_ofun <- traj_objfun(
-  h2_epi,
-  params = h2_fit$h2_params
-)
-
-mod2_epi_ll <- -epi_ofun(par = h2_fit$h2_params)
-mod2_ll <- mod2_epi_ll - sum(log(h2_epi@data + 1), na.rm = TRUE)
-
-arima_lik2 <- 0
-
-# model 2 AIC
-mod2_n_params <- h2_fit$n_fit_params
-mod2_aic <- 2 * mod2_n_params - 2 * mod2_ll
-
-# model 1 AIC
-mod1_n_params <- length(
-  c('beta1', 'beta2', 'beta3', 'beta4', 'beta5', 'beta6',
-    'rho', 'nu', 'E_0', 'I_0', 'tau_epi',
-    'tau_end', 'sig_sq_epi', 'sig_sq_end', 'betat')
-)
-mod1_aic <- 2 * mod1_n_params - 2 * mod1_ll
-
-# model 3 AIC
-mod3_aic <- 2 * mod3_n_params - 2 * mod3_ll
-
-# Calculate likelihood via block-particle filter
-mod3_eval_res <- bake(
-  file = paste0('model3/', rl_dir, 'mod3_eval_res.rds'), {
-
-    # Create a list to store results
-    results <- list()
-
-    coef(h3Spat) <- p3
-
-    # Perform block-particle filter
-    h3_bpf <- foreach(i = 1:NREPS_EVAL, .combine = c) %dopar% {
-      bpfilter(h3Spat, Np = NP_EVAL, block_size = 1)
-    }
-
-    ll <- logLik(h3_bpf)
-    results$ll <- logmeanexp(ll[!is.na(ll)], se = FALSE)
-
-    ### Get likelihood for subset of the data
-
-    # Find appropriate sub-set
-    in_subset_cols <- h3_bpf[[1]]@times >= as.numeric(lubridate::decimal_date(as.Date("2014-03-01")))
-    in_subset_rows <- h3_bpf[[1]]@unit_names != 'Ouest'
-
-    mod3_coupled_subset_evals <- c()
-    for (i in 1:length(h3_bpf)) {
-      mod3_coupled_subset_evals <- c(
-        mod3_coupled_subset_evals,
-        sum(h3_bpf[[i]]@block.cond.loglik[in_subset_rows, in_subset_cols])
-      )
-    }
-
-    results$subset_ll <- logmeanexp(mod3_coupled_subset_evals)
-    results
-  },
-  timing = FALSE
-)
-
-mod3_eval_res$subset_aic <- 2 * mod3_n_params - 2 * mod3_eval_res$subset_ll
-
-rm(NREPS_EVAL, NP_EVAL)
-@
-
-<<CalculateARMA, echo=FALSE, message=FALSE, include=FALSE, cache=TRUE>>=
-
-ARMA_benchmarks <- list()
-
-###
-### Model 1 ###
-###
-
-# Load aggregated data
-m1_agg_data <- haiti1_agg_data()
-
-# Fit ARMA(2, 1) on log-cases
-m1_log_arma <- arima(log(m1_agg_data$cases + 1), order = c(2, 0, 1))
-
-# Fit ARMA(2, 1) on natural scale of cases
-m1_arma <- arima(m1_agg_data$cases, order = c(2, 0, 1))
-
-# Check which one is better, and save benchmark as the better one
-if (m1_log_arma$loglik - sum(log(m1_agg_data$cases + 1), na.rm = TRUE) >= m1_arma$loglik) {
-  ARMA_benchmarks[['m1']] <- m1_log_arma$loglik - sum(log(m1_agg_data$cases + 1), na.rm = TRUE)
-} else {
-  ARMA_benchmarks[['m1']] <- m1_arma$loglik
-}
-
-rm(m1_agg_data, m1_log_arma, m1_arma)
-gc()
-
-###
-### Model 2 ###
-###
-
-h2_data <- haiti2_data(recreate = FALSE)
-
-# model 2 ARIMA benchmark
-h2_dep_names <- h2_data$department %>% unique()
-m2_benchmark <- 0
-for (i in 1:10){
-  cases <- h2_data[h2_data$department == h2_dep_names[i], ]$cases
-  log_cases <- log(cases + 1)
-
-  m2_arma_dep_ll <- arima(cases, order=c(2, 0, 1))$loglik
-  m2_log_arma_dep_ll <- arima(log_cases, order=c(2, 0, 1))$loglik - sum(log_cases, na.rm = TRUE)
-
-  m2_benchmark <- m2_benchmark + max(m2_arma_dep_ll, m2_log_arma_dep_ll)
-}
-
-ARMA_benchmarks[['m2']] <- m2_benchmark
-
-rm(m2_benchmark, h2_dep_names, cases, log_cases, m2_arma_dep_ll, m2_log_arma_dep_ll, h2_data)
-gc()
-
-###
-### Model 3 ###
-###
-
-mod3_cases <- as.data.frame(t(h3Spat@data))
-
-m3_benchmark <- 0
-for (dep in 1:10) {
-
-  # Get department cases
-  cases <- as.numeric(mod3_cases[, dep])
-
-  # Fit model on natural scale of cases
-  m3_a201 <- arima(cases, order = c(2, 0, 1))
-
-  # Fit model on log-scale of cases
-  m3_log_a201 <- arima(log(cases + 1), order = c(2, 0, 1))
-
-  m3_benchmark <- m3_benchmark + max(
-    m3_log_a201$loglik - sum(log(cases + 1), na.rm = TRUE),
-    m3_a201$loglik
-  )
-}
-
-ARMA_benchmarks[['m3']] <- m3_benchmark
-rm(cases, m3_a201, m3_log_a201, m3_benchmark, dep,
-   mod3_cases, temp, jumps, MODEL3_CASES, i, j)
-gc()
-@
-
-<<Table 2 Lee Input, echo=FALSE, include=FALSE>>=
-mod1_lee_n_params <- 20
-mod1_lee_aic <- 2 * mod1_lee_n_params - 2 * mod1_lee_ll
-
-mod2_lee_aic <- 2 * h2_fit$leeFit_n_params - 2 * h2_fit$leeFit_ll
-@
-
-<<Lee3 Sub Calculations, echo=FALSE, include=FALSE, message=FALSE>>=
-lee3_NP   <- switch(RUN_LEVEL, 50, 250, 5000)
-lee3_NREP <- switch(RUN_LEVEL,  3,   8,   36)
-
-registerDoRNG(49637421)
-
-# Create Lee et al (2020a) version of model 3 on sub-data
-lee3_sub <- lee3_spatPomp()
-
-# This model is only fit for a subset of the data, so we will set data
-# that are not part of this subset as NA. This will allow us to estimate
-# the likelihood for only the values that were considered during the
-# parameter estimation phase.
-# set_as_na <- lee3_sub@times < lubridate::decimal_date(as.Date('2017-06-10'))
-# temp_ouest_cases <- lee3_sub@data['casesOuest', ]
-# temp_ouest_cases[set_as_na] <- NA_real_
-# lee3_sub@data['casesOuest', ] <- temp_ouest_cases
-
-lee3_all <- lee3_spatPomp(start_date = "2010-10-23")
-
-lee3_results <- bake(
-  file = paste0('model3/', rl_dir, 'lee3_results.rds'), {
-
-    # Create lists to save results
-    lee3_results <- list()
-
-    # Save how long it takes for sub-data
-    lee3_results[["sub_bpf_time"]] <- system.time({
-
-      lee3_sub_ll_evals <- foreach(
-        i = 1:lee3_NREP, .combine = rbind
-      ) %dopar% {
-
-        liks <- numeric(10L)
-
-        lee_bpf_out <- bpfilter(lee3_sub, block_size = 1, Np = lee3_NP)
-        Ouest_ind <- which(lee_bpf_out@unit_names == "Ouest")
-
-        for (u in 1:10) {
-          if (u == Ouest_ind) {
-            liks[u] <- sum(lee_bpf_out@block.cond.loglik[u, lee3_sub@times >= lubridate::decimal_date(as.Date('2017-06-10'))])
-          } else {
-            liks[u] <- sum(lee_bpf_out@block.cond.loglik[u, ])
-          }
-        }
-
-        liks
-      }
-    })
-
-    # Save all likelihood evaluations
-    lee3_results[["sub_ll_eval"]] <- logmeanexp(rowSums(lee3_sub_ll_evals)[!is.na(rowSums(lee3_sub_ll_evals))])
-
-    registerDoRNG(2789131)
-
-    # Save how long it takes for all data
-    lee3_results[["all_pf_time"]] <- system.time({
-      # Save all likelihood evaluations
-
-      lee3_all_ll_evals <- foreach(
-        i = 1:lee3_NREP, .combine = c
-      ) %dopar% {
-        logLik(bpfilter(lee3_all, block_size = 1, Np = lee3_NP))  # Get likelihood evaluations
-      }
-
-      lee3_results[["all_ll_eval"]] <- logmeanexp(lee3_all_ll_evals[!is.na(lee3_all_ll_evals)])
-
-    })
-
-    lee3_results[["n_params"]] <- 29
-
-    lee3_results
-  },
-  timing = FALSE
-)
-
-rm(lee3_sub, lee3_all, lee3_NP, lee3_NREP)
-gc()
-
-mod3_lee_coupled_aic <- 2 * lee3_results$n_params - 2 * lee3_results$all_ll_eval
-
-mod3_lee_coupled_partial_aic <- 2 * lee3_results$n_params - 2 * lee3_results$sub_ll_eval
-@
-
-\begin{table}
-\centering
-
-\begin{tabular}{|c|c|c|c|}
-\hline
- & \thead{Model~1} & \thead{Model~2} & \thead{Model~3}
-\\
-\hline
-\hline
-\multirow{2}{*}{Log-likelihood} &
-  $\Sexpr{myround(mod1_ll, 1)}$ &
-  $\Sexpr{myround(mod2_ll, 1)}$ &
-  $\Sexpr{myround(mod3_ll, 1)}$ \\
-    & ($\Sexpr{myround(mod1_lee_ll, 1)}$)\footnotemark[1] &
-  ($\Sexpr{myround(h2_fit$leeFit_ll, 1)}$) &
-  ($\Sexpr{myround(lee3_results[["all_ll_eval"]], 1)}$)\footnotemark[2]
-\\
-\hline
-Number of &
-  $\Sexpr{as.character(mod1_n_params)}$ &
-  $\Sexpr{as.character(mod2_n_params)}$ &
-  $\Sexpr{as.character(mod3_n_params)}$ \\
- Fit Parameters &
- (\Sexpr{as.character(mod1_lee_n_params)}) &
- ($\Sexpr{as.character(h2_fit$leeFit_n_params)}$) &
- ($\Sexpr{as.character(lee3_results$n_params, 1)}$)
-\\
-\hline
-\multirow{2}{*}{AIC} &
-  $\Sexpr{myround(mod1_aic, 1)}$ &
-  $\Sexpr{myround(mod2_aic, 1)}$ &
-  $\Sexpr{myround(mod3_aic, 1)}$ \\
-  & ($\Sexpr{myround(mod1_lee_aic, 1)}$)\footnotemark[1] &
-  ($\Sexpr{myround(mod2_lee_aic, 1)}$) &
-  ($\Sexpr{myround(mod3_lee_coupled_aic, 1)}$)\footnotemark[2]
-\\
-\hline
-\thead{Log-ARMA(2,1) \\ AIC} &
-  $\Sexpr{myround(-2 * ARMA_benchmarks$m1 + 2 * 5, 1)}$ &
-  $\Sexpr{myround(-2 * ARMA_benchmarks$m2 + 2 * 5, 1)}$ &
-  $\Sexpr{myround(-2 * ARMA_benchmarks$m3 + 2 * 5, 1)}$
-\\
-\hline
-\end{tabular}
-\caption{\label{tab:likes}AIC values for each model compared to their ARMA benchmark.
-Values in parentheses are corresponding values using \cite{lee20} parameter estimates.
-$\textsuperscript{1}$The reported likelihood is an upper bound of the likelihood of the \cite{lee20} model.
-$\textsuperscript{2}$\cite{lee20} fit Model~3 to a subset of the data (March 2014 onward, excluding data from Ouest in 2015-2016).
-On this subset, their model has a likelihood of $\Sexpr{myround(lee3_results$sub_ll_eval, 1)}$.
-On this same subset, our model has a likelihood of $\Sexpr{myround(mod3_eval_res$subset_ll, 1)}$.
-See Sec.~S6 of the supplement material for more details on estimating the likelihood of the Lee et al.~\cite{lee20} models.
-}
-\end{table}
 
@@
 \caption{\label{fig:h1SeasRain}
+{\figTitle Seasonality of Model~1 transmission compared to rainfall data.}
 (Top) weekly rainfall in Haiti, lighter colors representing more recent years.
@@
 The robust statistical conclusion is that a model which allows for change fits better than one which does not---we argue that a decreasing transmission rate is a plausible way to explain this, but the incidence data themselves do not provide enough information to pin down the mechanism.
-
-% Place figure captions after the first paragraph in which they are cited.
-\begin{figure}[!h]
-\caption{{\bf Bold the figure title.}
-Figure caption text here, please use this space for the figure panel descriptions instead of using subfigure commands. A: Lorem ipsum dolor sit amet. B: Consectetur adipiscing elit.}
-\label{fig1}
-\end{figure}
 
@@
 \caption{\label{fig:elimProbs}
-Probability of elimination across simulations for a 10 year period.
+{\figTitle Simulated probability of elimination using Models~1 and 3.}
+Probability of cholera elimination, defined as having zero cholera infectious for at least 52 consecutive weeks, based on 10 year simulations from calibrated versions of Models~1 and 3.
 Compare to Fig.~3A of Lee et al.~\cite{lee20}.}
@@
 We therefore do not provide probability of elimination estimates under Model~2, but show trajectories under the various vaccination scenarios using this model (Fig.~\ref{fig:mod2Traj}).
-
-%%% table tttttttttt
-
-\begin{table}
-\begin{adjustwidth}{-2.25in}{0in}
-\centering
-% \small
-\begin{tabular}{|p{0.15\linewidth}|p{0.14\linewidth}t|p{0.13\linewidth}t|p{0.17\linewidth}t|}
-\hline
-Mechanism & Model 1 && Model 2 && Model 3 &
-\\
-\hline
-\hline
-Infection (day)
-  & ${\fixed \muIR^{-1}=\Sexpr{myround(7/p1["gamma"],1)}}$
-  & \eqref{model1:toR}
-  & ${\fixed \muIR^{-1}=\Sexpr{myround(365/h2_fit$h2_params["gamma"],1)}}$
-  & \eqref{model2:mu_IR}
-  & ${\fixed \muIR^{-1}=\Sexpr{myround(365/p3["gamma1"],1)}}$
-  & \eqref{eq:model3:IR}
-\\
-Latency (day)
-  & ${\fixed \muEI^{-1}=\Sexpr{myround(7/p1["sigma"],1)}}$
-  & \eqref{model1:EA}
-  & ${\fixed \muEI^{-1}=\Sexpr{myround(365/h2_fit$h2_params["gammaE"],1)}}$
-  & \eqref{model2:mu_EI}
-  & \missing
-  % & \multicolumn{1}{c}{\hspace{16pt}\missing}
-  &
-\\
-\hline
-\multirow{3}{*}{Seasonality}
-  & $\transmission_{1:6}=(
-      \Sexpr{myround(p1["beta1"],1)},\allowbreak
-      \Sexpr{myround(p1["beta2"],1)},\allowbreak
-      \Sexpr{myround(p1["beta3"],1)},\allowbreak
-      \Sexpr{myround(p1["beta4"],1)},\allowbreak
-      \Sexpr{myround(p1["beta5"],1)},\allowbreak
-      \Sexpr{myround(p1["beta6"],1)})$
-  & \eqref{model1:beta}
-  & \multirow{2}{*}{${\fixed \seasAmplitude=0.4}$}
-  & \multirow{2}{*}{\eqref{model2:lambda}}
-  & \multirow{2}{*}{$\seasAmplitude = \Sexpr{myround(p3s["lambdaR"],2)}$}
-  & \multirow{2}{*}{\eqref{eq:model3:water}}
-  \\
-  & $\transmissionTrend = \Sexpr{myround(p1["betat"],2)}^*$
-  & \eqref{model1:betat}
-  & \multirow{-2}{*}{$\phaseParm = \Sexpr{myround(h2_fit$h2_params['phase'], 2)}^*$}
-  & \multirow{-2}{*}{\eqref{model2:lambda}}
-  & \multirow{-2}{*}{$\rainfallExponent = \Sexpr{signif(p3s["r"],3)}$}
-  & \multirow{-2}{*}{\eqref{eq:model3:water}}
-  \\\hline
-  \multirow{3}{*}{Immunity (yr)}
-  & \multirow{3}{*}{${\fixed \muRS^{-1}=\Sexpr{myround(1/p1["alpha"]/52,1)}}$}
-  & \multirow{3}{*}{\eqref{model1:RS}}
-  & $\muRS^{-1}= \Sexpr{signif(1/h2_fit$h2_params["sigma"],2)}$
-  & \eqref{model2:RS}
-  & \multirow{3}{*}{${\fixed \muRS^{-1}=\Sexpr{myround(1/p3['rho1'],1)}}$}
-  & \multirow{3}{*}{\eqref{eq:model3:RRnext}}
-\\
-& & & ${\fixed \omega_1^{-1} = \Sexpr{myround(1/h2_fit$h2_params["Omega1"],1)}}$ & \eqref{model2:omega1} & &\\
-& & & ${\fixed \omega_2^{-1} = \Sexpr{myround(1/h2_fit$h2_params["Omega2"],1)}}$ & \eqref{model2:omega2} & &\\
-\hline
-\multirow{2}{*}{Birth/death (yr)}
-  & ${\fixed \muBirth^{-1} = \Sexpr{myround(1/p1["mu"]/52,1)}}$
-    %
-  & \eqref{model1:death}
-  & \multirow{2}{*}{\missing}
-  &
-  & \multirow{2}{*}{${\fixed \muDeath^{-1} = \Sexpr{myround(1/p3["mu1"],1)}}$}
-  & \multirow{2}{*}{\eqref{eq:model3:IS}}
-\\
-& ${\fixed \muDeath^{-1} = \Sexpr{myround(1/p1["delta"]/52,1)}}$
-& \eqref{model1:death}
-&
-&
-& &
-\\
-Sympt. frac.
-  & ${\fixed \symptomFrac_z(t)=c\vaccineEfficacy^*(t-\tau_d)}$
-  & (\ref{model1:EI}-\ref{model1:EA})
-  & ${\fixed \symptomFrac=\Sexpr{myround(h2_fit$h2_params["k"],1)}}$
-  & \eqref{model2:mu_EI}
-  & ${\fixed \symptomFrac=\Sexpr{myround(p3["sigma1"],2)}}$
-  & \eqref{eq:model3:SA}
-\\
-Asympt.
-  & \multirow{2}{*}{${\fixed \asymptomRelativeInfect=0.05}$}
-  & \multirow{2}{*}{\eqref{model1:lambda}}
-  & ${\fixed \asymptomRelativeInfect = 0.001}$
-  & \eqref{model2:lambda}
-  &  ${\fixed \asymptomRelativeInfect = 1}$
-  & \eqref{eq:model3:foi}
-\\
-infectivity
-&
-&
-& ${\fixed \asymptomRelativeShed = \Sexpr{1e-7} }$
-& \eqref{model2:to_W}
-& $\asymptomRelativeShed = \Sexpr{myround(p3s["XthetaA"],3)}$
-& \eqref{eq:model3:water}
-\\
-\hline
-\multirow{4}{*}{Human to human}
-  & \multirow{4}{*}{$\transmission_{1:6}$ as above}
-  & \multirow{4}{*}{\eqref{model1:lambda}}
-  & \multirow{4}{*}{$\transmission=$\Sexpr{signif(h2_fit$h2_params["Beta"],3)}}
-  & \multirow{4}{*}{\eqref{model2:lambda}}
-  & $\transmission_{1:10}=(
-  \Sexpr{myround(p3u["foi_add",1]*1e6,2)}, \allowbreak
-	\Sexpr{myround(p3u["foi_add",2]*1e6,2)}, \allowbreak
-	\Sexpr{myround(p3u["foi_add",3]*1e6,2)}, \allowbreak
-  \Sexpr{myround(p3u["foi_add",4]*1e6,2)}, \allowbreak
-	\Sexpr{myround(p3u["foi_add",5]*1e6,2)}, \allowbreak
-	\Sexpr{myround(p3u["foi_add",6]*1e6,2)}, \allowbreak
-	\Sexpr{myround(p3u["foi_add",7]*1e6,2)}, \allowbreak
-	\Sexpr{myround(p3u["foi_add",8]*1e6,2)}, \allowbreak
-	\Sexpr{myround(p3u["foi_add",9]*1e6,2)}, \allowbreak
-	\Sexpr{myround(p3u["foi_add",10]*1e6,2)}
-	) \allowbreak \times 10^{-6}$
-  & \multirow{4}{*}{\eqref{eq:model3:foi}}
-\\
-\hline
-\multirow{5}{*}{Water to human}
-  & \multirow{5}{*}{\missing}
-  % & \multicolumn{1}{c}{\hspace{16pt}\missing}
-  &
-  & \multirow{3}{*}{${\fixed \Wsat = \Sexpr{signif(h2_fit$h2_params["Sat"],2)}}$}
-  & \multirow{3}{*}{\eqref{model2:lambda}}
-  % \eqref{model2:lambda}
-  & $\Wbeta{_{1:10}}= (
-        \Sexpr{myround(p3u["betaB",1],2)}, \allowbreak
-        \Sexpr{myround(p3u["betaB",2],2)}, \allowbreak
-        \Sexpr{myround(p3u["betaB",3],2)}, \allowbreak
-        \Sexpr{myround(p3u["betaB",4],2)}, \allowbreak
-        \Sexpr{myround(p3u["betaB",5],2)}, \allowbreak
-        \Sexpr{myround(p3u["betaB",6],2)}, \allowbreak
-        \Sexpr{myround(p3u["betaB",7],2)}, \allowbreak
-        \Sexpr{myround(p3u["betaB",8],2)}, \allowbreak
-        \Sexpr{myround(p3u["betaB",9],2)}, \allowbreak
-        \Sexpr{myround(p3u["betaB",10],2)})$
-  & \multirow{4}{*}{\eqref{eq:model3:foi}}
-\\
-& & & \multirow{-3}{*}{$\beta_W= \Sexpr{signif(h2_fit$h2_params["BetaW"],3)}$} & \multirow{-3}{*}{\eqref{model2:lambda}} & &
-\\
-\hline
-Human to water
-  & \missing
-  % & \multicolumn{1}{c}{\hspace{16pt}\missing}
-  &
-  & $\Wshed = $ \Sexpr{signif(h2_fit$h2_params["Mu"],3)}
-  & \eqref{model2:to_W}
-  & $\Wshed= \Sexpr{signif(p3s["thetaI"],3)}$
-  & \eqref{eq:model3:water}
-\\
-Water survival \newline (wk)
-  & \multirow{2}{*}{\missing}
-  % & \multicolumn{1}{c}{\hspace{16pt}\missing}
-  &
-  & \multirow{2}{*}{${\fixed \Wremoval^{-1} = \Sexpr{signif(52/h2_fit$h2_params["Delta"],3)}}$}
-  & \multirow{2}{*}{\eqref{model2:from_W}}
-  & \multirow{2}{*}{$\Wremoval^{-1}=\Sexpr{myround(52/p3s["mu_B"],2)}$}
-  & \multirow{2}{*}{\eqref{eq:model3:Decay}}
-\\
-Mixing exponent
-  & $\mixExponent=\Sexpr{myround(p1["nu"],2)}$
-  & \eqref{model1:lambda}
-  & \missing
-  % & \multicolumn{1}{c}{\hspace{16pt}\missing}
-  &
-  & \missing
-  % & \multicolumn{1}{c}{\hspace{16pt}\missing}
-  &
-\\
-\hline
-Process noise \newline (wk\textsuperscript{1/2})
-  & $\sigmaProc =$ \newline $(\Sexpr{myround(sqrt(p1["sig_sq_epi"]),2)},
-  \Sexpr{myround(sqrt(p1["sig_sq_end"]),2)})^*$
-  & \multirow{2}{*}{\eqref{model1:lambda}}
-  & \multirow{2}{*}{\missing}
-  % & \multicolumn{1}{c}{\hspace{16pt}\missing}
-  &
-  & \multirow{2}{*}{$\sigmaProc= \Sexpr{myround(p3s["std_W"],3)}$}
-  & \multirow{2}{*}{\eqref{eq:model3:SA}}
-\\
-Reporting rate
-  & $\reportRate=\Sexpr{myround(p1["rho"], 3)}$
-  & (S15) % TODO: Check
-  & ${\fixed \reportRate=\Sexpr{myround(h2_fit$h2_params["Rho"],2)}}$
-  & (S16) % TODO: Check
-  & $\reportRate=\Sexpr{myround(p3s["epsilon"],2)}$
-  & (S19) % TODO: Check
-\\
-Observation \newline variance
-  & $\obsOverdispersion =$ \newline $(\Sexpr{myround(p1["tau_epi"],2)}, \Sexpr{round(p1["tau_end"],2)})$
-  & \multirow{2}{*}{(S15)} % TODO: Check
-  & \multirow{2}{*}{$\obsOverdispersion = \Sexpr{signif(h2_fit$h2_params['sigma'], 3)}$}
-  & \multirow{2}{*}{(S16)} % TODO: Check
-  & \multirow{2}{*}{$\obsOverdispersion = \Sexpr{myround(p3s["k"],2)}$}
-  & \multirow{2}{*}{(S19)} % TODO: Check
-\\
-\hline
-\multirow{2}{*}{Initial Values} & $I_{0,0} = \Sexpr{round(prod(p1[c('pop_0', 'I_0')]))}$
-  &
-  & \multirow{2}{*}{\missing}
-  % & \multicolumn{1}{c}{\multirow{2}{*}{\hspace{16pt}\missing}}
-  &
-  & \multirow{2}{*}{$I_{3,7,10;0}(0)= (
-        \Sexpr{round(prod(p3[c('H3', 'Iinit3')]))}, \Sexpr{round(prod(p3[c('H7', 'Iinit7')]))}, \Sexpr{round(prod(p3[c('H10', 'Iinit10')]))})^*$}
-  &  % TODO: \multirow{2}{*}{EQUATION NUMBER}
-  \\
-  & $E_{0,0} = \Sexpr{round(prod(p1[c('pop_0', 'E_0')]))}$ & & \multicolumn{1}{c}{}
-  &
-  &
-  &
-\\\hline
-\multirow{3}{*}{Hurricane}
-  & \multirow{4}{*}{\missing}
-% & \multicolumn{1}{c}{\multirow{2}{*}{\hspace{16pt}\missing}}
-  &
-  & \multirow{4}{*}{\missing}
-  % & \multicolumn{1}{c}{\multirow{2}{*}{\hspace{16pt}\missing}}
-  &
-  & $\Whur{3, 9} =$\newline $(\Sexpr{myround(p3['aHur3'], 2)}, \Sexpr{myround(p3['aHur9'], 2)})^*$
-  & \multirow{2}{*}{\eqref{eq:model3:foi}}
-\\
- \multirow{-1}{*}{Parameters} &
-  % & \multicolumn{1}{c}{}
-  &
-  &
-  % & \multicolumn{1}{c}{}
-  &
-  & $\hHur{3, 9} =$\newline $(\Sexpr{myround(p3['hHur3'], 2)}, \Sexpr{myround(p3['hHur9'], 2)})^*$
-  & \multirow{2}{*}{\eqref{eq:model3:foi}}
-\\\hline
-\end{tabular}
-\caption{\label{tab:params}References to the relevant equation are given in parentheses.
-Parameters in blue were fixed based on scientific reasoning and not fitted to the data.
-$^*$ denotes parameters added during our re-analysis, not considered by Lee et al.
-Translations back into the notation of \cite{lee20} are given in Table~S1.
-}
-\end{adjustwidth}
-\end{table}
 

### 2023-08-22 ms-plos.Rnw
removed parallel back-end from haitipkg

@@
 if(is.na(cores)) cores <- detectCores()
+
+# Register parallel back-end. This code below creates a fork cluster, so we are assuming
+# that the code is run on a mac or linux machine; this doesn't matter if you
+# plan on simply loading the results, but if you would like to redo the expensive
+# calculations (e.g., test minor changes to the code), you need to make sure you
+# register the correct parallel back-end.
 registerDoParallel(cores)

### 2023-08-23 ms-plos.Rnw
Updated SI w/ haitipkg, minor edits

@@
 \newcommand\tHur{t_{hm}}
+\newcommand\Iinit{I_{0,0}}
+\newcommand\Einit{E_{0, 0}}
 \newcommand\Wremoval{\delta_W}
@@
 rm(
-  SEARCH1, SEARCH2, SEARCH3, N_SEARCHES,
+  SEARCH1, SEARCH2, N_SEARCHES,
   est_param_names, est_param_names_expanded,
@@
 \hline
-\multirow{2}{*}{Initial Values} & $I_{0,0} = \Sexpr{round(prod(p1[c('pop_0', 'I_0')]))}$
+\multirow{2}{*}{Initial Values} & $\Iinit = \Sexpr{round(prod(p1[c('pop_0', 'I_0')]))}$
   &
@@
   &
-  & \multirow{2}{*}{$I_{3,7,10;0}(0)= (
-        \Sexpr{round(prod(p3[c('H3', 'Iinit3')]))}, \Sexpr{round(prod(p3[c('H7', 'Iinit7')]))}, \Sexpr{round(prod(p3[c('H10', 'Iinit10')]))})^*$}
+  & \multirow{2}{*}{$\Iinit^{3,7,10} = (
+        \Sexpr{round(prod(p3[c('H3', 'Iinit3')]))}, \Sexpr{round(prod(p3[c('H7', 'Iinit7')]))}, \Sexpr{round(prod(p3[c('H10', 'Iinit10')]))})^*$} % TODO: Check the unit numbers
   &  % TODO: \multirow{2}{*}{EQUATION NUMBER}

### 2023-08-23 si/si.Rnw
Updated SI w/ haitipkg, minor edits

@@
 \newcommand\hHur[1]{h_{#1}^{hm}}
+\newcommand\Iinit{I_{0,0}}
+\newcommand\Einit{E_{0, 0}}
+\newcommand\reportChange{\delta}
 
@@
 The numerical solutions to the model are arguably of more direct scientific interest than the exact solutions to the postulated equations.
-For ODE models, numerical methods are well studied and a standard numerical solution package such as \code{deSolve} in \code{R} is adequate for our purposes.
+For ODE models, numerical methods are well studied and a standard numerical solution package such as \code{deSolve} in \code{R} is adequate for many purposes.
 For SMCS and ODMCS models, exact schemes are feasible when the number of events is small, which may be the case for small populations.
@@
 Each POMP requires specification of a measurement model, which is a statistical description of how observations on the system are obtained.
-In general, we used the same measurement models that were reported by \cite{lee20}.
+In general, we used the same measurement models that were reported by \cite{lee20} unless specifically noted in the following subsections.
 
@@
   \label{model1:meas}
-  Y_{n} \mid \Delta N_{EI}(n) = z(n) \sim \text{NB}\left(\reportRate z(n), \obsOverdispersion \right),
+  Y_{n} \mid \Delta N_{EI}(n) = \reportChange \sim \text{NB}\left(\reportRate \reportChange, \obsOverdispersion \right),
 \end{equation}
-where $Y_n$ is the reported cholera cases at time $n \in 1:N$ and $\Delta N_{EI}(n)$ is the sum total of individuals across vaccination compartment $\vaccCounter$ who moved from compartment $E_\vaccCounter$ to $I_\vaccCounter$ since observation $n-1 \in 0:N-1$, and $\text{NB}\left(\reportRate z, \obsOverdispersion \right)$ is the negative binomial distribution with mean $\reportRate z$ and variance $\reportRate z \big(1 + \frac{\reportRate z}{\obsOverdispersion}\big)$.
+where $Y_n$ is the reported cholera cases at time $t_n \in t_1:t_N$ and $\Delta N_{EI}(n)$ is the sum total of individuals across each vaccination compartment $\vaccCounter \in 1:\vaccClass$ who moved from compartment $E_\vaccCounter$ to $I_\vaccCounter$ since observation $t_{n-1}$, and $\text{NB}\left(\reportRate \reportChange, \obsOverdispersion \right)$ is the negative binomial distribution with mean $\reportRate \reportChange$ and variance $\reportRate \reportChange \big(1 + \frac{\reportRate \reportChange}{\obsOverdispersion}\big)$.
 
@@
   \label{model2:meas}
-  \log\big(Y_{u, n} + 1\big) \mid \Delta N_{EI}(n) = z(n) \sim \text{N}\left(\log \big(\reportRate z(n) + 1\big), \obsOverdispersion^2 \right),
+  \log\big(Y_{u, n} + 1\big) \mid \Delta N_{EI}(n) = \reportChange \sim \text{N}\left(\log \big(\reportRate \reportChange + 1\big), \obsOverdispersion^2 \right),
 \end{equation}
-where $\Delta N_{EI}(n)$ is the sum total of individuals across vaccination compartment $\vaccCounter$ and unit $u$ who moved from compartment $E_{u\vaccCounter}$ to $I_{u\vaccCounter}$ since observation $n-1 \in 0:N-1$.
+where $\Delta N_{EI}(n)$ is the sum total of individuals across vaccination compartment $\vaccCounter \in 0:4$ and unit $u$ who moved from compartment $E_{u\vaccCounter}$ to $I_{u\vaccCounter}$ since observation $t_{n-1}$.
 Therefore, because the natural logarithm of observed case counts (plus one, to avoid taking the logarithm of zero) has a normal distribution, $Y_{u, n} + 1$ is assumed to follow a log-normal distribution with log-mean parameter $\mu = \log\big(\Delta N_{EI}(n) + 1\big)$ and log-variance $\obsOverdispersion^2$.
@@
 Although their text and supplement material do not explicitly describe the measurement model used, inspection of the code provided with their submission suggests a change in measurement model between the epidemic and endemic phases.
-In the file \code{choleraEqsPublic.py}, \citet{lee20} create several functions, where each function represents a set of coupled differential equations that could potentially be used to model cholera incidence data.
-Each function returns a vector (or \code{numpy} array) that represents the change in each state variable for a single time step, including the variable \code{dC}, which tracks the number of new infections and is used to obtain the reported case counts.
-Following their comments and code, it appears that the function \code{choleraEqs10WithoutVaccinationNetwork} was used to describe the dynamics of the epidemic phase, and \code{choleraEqs11WithoutVaccinationNetwork} was used for the endemic stage (see Sec~\ref{sec:mod2rep} for more details).
-Because their models were fit using least squares, the code in these functions suggest that the measurement model for the epidemic phase is
+% In the file \code{choleraEqsPublic.py}, \citet{lee20} create several functions, where each function represents a set of coupled differential equations that could potentially be used to model cholera incidence data.
+% Each function returns a vector (or \code{numpy} array) that represents the change in each state variable for a single time step, including the variable \code{dC}, which tracks the number of new infections and is used to obtain the reported case counts.
+% Following their comments and code, it appears that the function \code{choleraEqs10WithoutVaccinationNetwork} was used to describe the dynamics of the epidemic phase, and \code{choleraEqs11WithoutVaccinationNetwork} was used for the endemic stage (see Sec~\ref{sec:mod2rep} for more details).
+% Because their models were fit using least squares, the code in these functions suggest that the measurement model for the epidemic phase is
+The measurement model they used for the epidemic phase is
 \begin{equation}
   \label{model2:measEpi}
-  Y_{u, n} \mid \Delta N_{EI}(n) = z(n) \sim \text{N}\left(\reportRate z(n), \obsOverdispersion^2 \right),
+  Y_{u, n} \mid \Delta N_{EI}(n) = \reportChange \sim \text{N}\left(\reportRate \reportChange, \obsOverdispersion^2 \right),
 \end{equation}
-which is similar to our measurement model, the primary difference being that the measurement model is applied to raw case counts rather than log-transformed case counts.
-The measurement model for the endemic phase modifies the epidemic model by counting both asymptomatically infected (A) and symptomatically infected (I) individuals in the case counts:
+% which is similar to our measurement model, the primary difference being that the measurement model is applied to raw case counts rather than log-transformed case counts.
+The measurement model they used for the endemic phase modifies the epidemic model by counting both asymptomatically infected (A) and symptomatically infected (I) individuals in the case counts:
 \begin{equation}
   \label{model2:measEnd}
-  Y_{u, n} \mid \Delta N_{EI}(n) = z_1(n), \Delta N_{EA}(n) = z_2(n) \sim \text{N}\left(\reportRate \big(z_1(n) + z_2(n)\big), \obsOverdispersion^2 \right),
+  Y_{u, n} \mid \Delta N_{EI}(n) = \reportChange_1, \Delta N_{EA}(n) = \reportChange_2 \sim \text{N}\left(\reportRate \big(\reportChange_1 + \reportChange_2\big), \obsOverdispersion^2 \right),
 \end{equation}
@@
   \label{model3:meas}
-  Y_{u, n} \mid \Delta N_{S_{u \vaccCounter}I_{u \vaccCounter}}(t) = z_{u}(n) \sim \text{NB}\left(\reportRate z_{u}(n), \obsOverdispersion \right),
+  Y_{u, n} \mid \Delta N_{S_{u \vaccCounter}I_{u \vaccCounter}}(t) = \reportChange_u \sim \text{NB}\left(\reportRate \reportChange_{u}, \obsOverdispersion \right),
 \end{equation}
-where $\Delta N_{S_{u \vaccCounter}I_{u \vaccCounter}}(n)$ is the number of individuals who moved from compartment $S_{u \vaccCounter}$ to $I_{u \vaccCounter}$ since observation $n-1 \in 0:N-1$.
+where $\Delta N_{S_{u \vaccCounter}I_{u \vaccCounter}}(n)$ is the number of individuals who moved from compartment $S_{u \vaccCounter}$ to $I_{u \vaccCounter}$ since observation $t_{n-1}$.
 
@@
 
-To perform inference on POMP models, it is necessary to propose an initial probability density for the latent process $f_{X_0}(x_0;\theta)$, including the possibility that the initial values could be known, or could be a non-random function of the unknown parameter vector, $\theta$.
+To perform inference on POMP models, it is necessary to propose an initial probability density for the latent process $f_{X_0}(x_0;\theta)$, including the possibility that the initial values of the latent states are known, or are a non-random function of the unknown parameter vector, $\theta$.
 This density is used to obtain initial values of the latent state when fitting and evaluating the model.
@@
 
-For this model, the number of individuals in the Recovered and Asymptomatic compartments are set to zero, but the initial proportion of Infected and Exposed individuals is estimated as initial value parameters ($I(0)$ and $E(0)$, respectively) using the IF2 algorithm, implemented as \texttt{mif2} in the \texttt{pomp} package.
-Finally, the initial proportion of Susceptible individuals $S_0$ is calculated as $S(0) = 1 - I(0) - E(0)$.
+For this model, the number of individuals in the Recovered and Asymptomatic compartments are set to zero, but the initial proportion of Infected and Exposed individuals is estimated as initial value parameters ($\Iinit$ and $\Einit$, respectively) using the IF2 algorithm, implemented as \texttt{mif2} in the \texttt{pomp} package.
+Finally, the initial proportion of Susceptible individuals $S_{0, 0}$ is calculated as $S_{0, 0} = 1 - \Iinit - \Einit$.
+This model for the initial values of the latent states matches that which was used by \citet{lee20}.
 
@@
 Specifically, the model sets $I_{u0}(0) = y^*_{1u} / \reportRate$ for each unit $u \in 1:10$, where $I_{u0}(t)$ is the number of infected individuals in unit $u$ at time $t$, vaccination scenario $z = 0$, $y^*_{tu}$ is the reported number of cases, and $\reportRate$ is the reporting rate.
+This model for the initial values of the latent states matches that which was used by \citet{lee20}.
 
@@
 
-As described in the main test, the latent states are initialized by estimating of the number of infected, asymptomatic, and recovered individuals at time $t_0$ by using observed case counts at time $t_1$.
+The latent states of this model are initialized by estimating of the number of infected, asymptomatic, and recovered individuals at time $t_0$ by using observed case counts at time $t_1$.
 The initial value for the bacteria compartment is then determined using the rainfall dynamics of Eq.~23.
@@
 The remaining hidden states are estimated by enforcing model dynamics using this result.
-
-% \section{Block Panel Iterated Filter}
-%
-% The panel iterated filter (PIF) algorithm of \citet{breto20} has been used to fit parameters of panelPOMP models in several studies \citep{domeyer22,wale2019,ranjeva17,ranjeva19}.
-% Like other iterated filtering algorithms, PIF iteratively combines particle filter calculations with parameter perturbations.
-% Theoretical results show that, as the perturbations are reduced over successive iterations, the perturbed parameters converge to the MLE.
-%
-% In the case of a panelPOMP, the parameter vector $\theta$ may be composed of a set of shared parameters $\phi$, and a set of unit-specific parameters $\psi_{u}$, for units $u \in \seq{1}{U}$.
-% Here, we say that a parameter $\psi_{u}$ for $u \in \seq{1}{U}$ is unit-specific if it is not involved in the one-step transition density or the measurement model of any other unit $k \neq u \in \seq{1}{U}$, and shared parameters are those that are not unit-specific.
-% The PIF algorithm treats the parameter vector $\theta = (\phi, \psi_{1:U})$ as a complete entity and gives no distinction between shared and unit-specific parameters.
-% Even though the random walk standard deviation is set to zero for unit-specific parameters when considering data from other units, this lack of distinction may result in particle depletion for unit-specific parameters as the number of units $U$ grows large.
-% This is because it is likely that, when updating the parameters for a given unit, several parameter vector $\theta$ particles will not be resampled due to having low weights.
-% When the number of units is large, this will lead to a decreasing number of ancestral particles to sample unit-specific parameters $\psi_k$ as $k$ increases.
-%
-% As a solution to this issue, we propose the block panel iterated filter (BPIF), described in Algorithm~\ref{alg:bpif}.
-% In this algorithm, shared $\phi$ and unit-specific parameters $\psi_{u}$ are handled differently when updating the parameter vector $\theta = (\phi, \psi_{1:U})$.
-% The BPIF updates $\theta$ by updating the shared vector $\phi$ in the same way that is done with PIF, but index resampling for unit-specific parameters $\psi_{1:U}$ only occurs when data from the corresponding
-% unit are being considered.
-% We have found that this modification leads to improved likelihood maximization performance, especially as the number of units increases.
-% No theoretical properties of this algorithm have yet been derived, but the approach has similarities to the block particle filter of \citet{rebeschini15} and its extension to parameter estimation \citep{ionides22,ning21ibpf}.
-%
-% \begin{algorithm}[ht]
-%    \caption[BPIF]{\textbf{BPIF}. \\
-%     {\bf Inputs}:\\\hspace{\textwidth}
-%     Simulator of initial density, $f_{X_{u, 0}}(x_{u, 0}; \theta)$ for $u$ in $\seq{1}{U}$.\\\hspace{\textwidth}
-%     Simulator of transition density, $f_{X_{u, n}|X_{u, n-1}}(x_{u, n}|x_{u, n-1}; \theta)$ for $u$ in $\seq{1}{U}$, $n$ in $\seq{1}{N_u}$.\\\hspace{\textwidth}
-%     Evaluator of measurement density, $f_{Y_{u, n}|X_{u, n}}(y_{u, n}|x_{u, n} ; \theta)$ for $u$ in $\seq{1}{U}$, $n$ in $\seq{1}{N_u}$.\\\hspace{\textwidth}
-%     Data $y_{u, n}^*$,for $u$ in $\seq{1}{U}$, $n$ in $\seq{1}{N_u}$.\\\hspace{\textwidth}
-%     Number of iterations, $M$.\\\hspace{\textwidth}
-%     Number of particles, $J$.\\\hspace{\textwidth}
-%     Starting parameter swarm, $\Theta_j^0 = \big(\Phi_j^0, \Psi_{1:U, j}^0\big)$ for $j \in \seq{1}{J}$, $u \in \seq{1}{U}$.\\\hspace{\textwidth}
-%     Simulator of perturbation density, $h_{u, n}(\theta | \phi; \sigma)$ for $u \in \seq{1}{U}$, $n \in 0:N_u$.\\\hspace{\textwidth}
-%     Perturbation sequence, $\sigma_m$ for $m \in 1:M$.\\\hspace{\textwidth}
-%     {\bf Output:} \\\hspace{\textwidth}
-%     Final parameter swarm, $\Theta_{j}^{m} = \big(\Phi_j^m, \Psi_{\seq{1}{U}, j}^m\big)$ for $j \in \seq{1}{J}$, $u \in \seq{1}{U}$.
-%     \label{alg:bpif}}
-% %\BlankLine
-% \For{$m \in 1:M$}{
-%   Set $\Phi_{0, j}^m = \Phi_j^{m-1}$ for $j \in \seq{1}{J}$\;
-%     \For{$u \in \seq{1}{U}$}{
-%       Set $\Phi_{u, 0, j}^{F, m} \sim h_{u, 0}(\phi | \Phi_{0, j}^m; \sigma_m)$ for $j \in \seq{1}{J}$\;
-%       Set $\Psi_{u, 0, j}^{F, m} \sim h_{u, 0}(\psi | \Psi_{u, j}^{m - 1}; \sigma_m)$ for $j \in \seq{1}{J}$\;
-%       Set $\Theta_{u, 0, j}^{F, m} = \big(\Phi_{u, 0, j}^{F, m}, \Psi_{u, 0, j}^{F, m}\big)$ for $j \in \seq{1}{J}$\;
-%       Initialize $X_{u, 0, j}^{F, m} \sim f_{X_{u, 0}}(x_{u, 0}; \Theta_{u, 0, j}^{F, m})$ for $j \in \seq{1}{J}$\;
-%         \For{$n \in \seq{1}{N_u}$} {
-%           Set $\Theta_{u, n, j}^{P, m} \sim h_{u, n}(\theta|\Theta_{u, n-1, j}^{F, m}, \sigma_m)$ for $j \in \seq{1}{J}$\;
-%           $X_{u, n, j}^{P, m} \sim f_{X_{u, n}|X_{u, n-1}}(x_{u, n}|X_{u, n-1, j}^{F, m}; \Theta_{u, n, j}^{P, m})$ for $j \in \seq{1}{J}$\;
-%           $w_{u, n, j}^m = f_{Y_{u, n}|X_{u, n}}(y_{u, n}^*|X_{u, n, j}^{P, m})$ for $j \in \seq{1}{J}$\;
-%           Draw $k_{\seq{1}{j}}$ with $P(k_j = i) = w_{u, n, i}^m / \sum_{v = 1}^J w_{u, n, v}^m$ for $i, j \in \seq{1}{J}$\;
-%           Set $\Theta_{u, n, j}^{F, m} = \Theta_{u, n, k_j}^{P, m} = \big(\Phi_{u, n, k_j}^{P, m}, \Psi_{u, n, k_j}^{P, m}\big)$ and $X_{u, n, j}^{F, m} = X_{u, n, k_j}^{P, m}$ for $j \in \seq{1}{J}$\;
-%         }
-%       Set $\Theta_{u, j}^m = \big(\Phi_{u, N_u, j}^{F, m}, \Psi_{u, N_u, j}^{F, m}\big)$ for $j \in \seq{1}{J}$\;
-%     }
-%   Set $\Theta_j^m = \big(\Phi_{U, j}^m, \Psi_{1:U, j}^m\big)$ for $j \in \seq{1}{J}$\;
-% }
-% \end{algorithm}
+% TODO: Check this section.
 
@@
 
-# This function sets the rw.sd for named parameters, first search.
+# This function sets the rw_sd for named parameters, first search.
 set_rw_1 <- function(x) {
@@
 
-# This function sets the rw.sd for named parameters, second search.
+# This function sets the rw_sd for named parameters, second search.
 set_rw_2 <- function(x) {
@@
 
-# This function sets the rw.sd for named parameters, final search.
+# This function sets the rw_sd for named parameters, final search.
 set_rw_3 <- function(x) {
@@
 # Fit the model and save the results
+# TODO: Remove SEARCH3 when re-running everything.
 haiti3_orig_fit <- bake(
@@
 
-# Create dataset that will be used to fit ARMA models. To do this, we
-# need to make sure that missing values are marked as NA, so we check
-# for any jumps in time.
-MODEL3_CASES <- haitiCholera %>%
+MODEL3_CASES <- haitiCholera |>
   dplyr::rename(
@@
     Sud_Est = Sud.Est
-  ) %>%
-  dplyr::mutate(date = as.Date(date)) %>%
-  dplyr::select(-report) %>%
-  dplyr::mutate(time_diff = date - lag(date))
+  ) |>
+  dplyr::mutate(date = as.Date(date)) |>
+  dplyr::select(-report) |>
+  dplyr::mutate(time_diff = date - lag(date)) |>
+  dplyr::filter(date >= "2010-11-20")
+
+# Create dataset that will be used to fit ARMA models. To do this, we
+# need to make sure that missing values are marked as NA, so we check
+# for any jumps in time.
+# MODEL3_CASES <- haitiCholera %>%
+#   dplyr::rename(
+#     date = date_saturday, Grande_Anse = Grand.Anse,
+#     Nord_Est = Nord.Est, Nord_Ouest = Nord.Ouest,
+#     Sud_Est = Sud.Est
+#   ) %>%
+#   dplyr::mutate(date = as.Date(date)) %>%
+#   dplyr::select(-report) %>%
 
@@
 In this section, we provide more detail on the process that was used to estimate the coefficients used in Model~3.
-In particular, we discuss why we decided to include additional model parameters that were not considered by \cite{lee20} (parameters related to initial conditions and Hurricane Matthew).
+In particular, we discuss why we decided to include additional model parameters---those that are associated with the initial conditions of the model and the behavior of the system during Hurricane Matthew---that were not considered by \cite{lee20}.
 To calibrate this model, we used the iterated block particle filter (IBPF) method of \cite{ionides22}.
-Due to the novelty of this algorithm, there does not exist published examples the IBPF algorithm used for data analysis outside of the papers in which the algorithm was introduced \citep{ning21ibpf,ionides22}, which at least partially motivates the creation of this section.
+Due to the novelty of this algorithm, there does not exist published examples the IBPF algorithm used for data analysis outside of the papers in which the algorithm was introduced \citep{ning21ibpf,ionides22}, which at least partially motivates the inclusion additional details in this section.
 
@@
 The first collection of searches was performed by obtaining initial values for the parameters by uniformly sampling values from a predefined hypercube.
-Two subsequent refinement searches used starting values corresponding to the largest model likelihoods obtained in the previous round.
+A subsequent refinement search used starting values corresponding to the largest model likelihoods obtained in the previous round.
 This is similar to the approach used by \cite{ionides22}.
@@
 We use the technique described above to fit the fully coupled version of Model~3 proposed by \cite{lee20} to all observed cholera data.
-The maximum likelihood we obtained with the model was $\Sexpr{round(max(haiti3_orig_fit$search3$logLiks$logLik))}$, which is higher than the log-ARMA benchmark ($\Sexpr{round(sum(unitLikesARMA$ll))}$).
+The maximum likelihood we obtained with the model was $\Sexpr{round(max(haiti3_orig_fit$search2$logLiks$logLik))}$, which is higher than the log-ARMA benchmark ($\Sexpr{round(sum(unitLikesARMA$ll))}$).
+% TODO: Check this, numbers don't match up.
 For meta-population models, it is worth considering how well the model fits the data to each spatial unit.
@@
 # This result was used in the manuscript, so it doesn't need to be recalculated.
-h3_final_results <- readRDS("../model3/run_level_3/haiti3_coupled_fit.rds")
+h3_final_results <- readRDS("../model3/run_level_3/haiti3_fit.rds")
 
@@
 # Then, save the model parameter values and the likelihood evaluations
-best_m <- which.max(h3_final_results$search3$logLiks$logLik)
-best_params <- h3_final_results$search3$params[best_m, ]
-all_liks <- as.data.frame(h3_final_results$search3$likMat)
+best_m <- which.max(h3_final_results$search2$logLiks$logLik)
+best_params <- h3_final_results$search2$params[best_m, ]
+all_liks <- as.data.frame(h3_final_results$search2$likMat)
 all_liks$which <- rep(1:72, each = 18)
@@
 We refit Model~3 after introducing these seven parameters---three initial value parameters and four parameters related to Hurricane Matthew---using the same successive search scheme described above.
-The resulting model has a log-likelihood value of $\Sexpr{round(max(h3_final_results$search3$logLiks$logLik), 1)}$.
+The resulting model has a log-likelihood value of $\Sexpr{round(max(h3_final_results$search2$logLiks$logLik), 1)}$.
 We also provide figures for department specific log-likelihoods and conditional log-likelihoods in Figs.~\ref{fig:finalUnitLL} and \ref{fig:finalCondLL}.
@@
 This could be due to many factors, including the possibility that the coupling between units forces low initial value parameters in a given department in order for the model to fit the data in the remaining departments.
-Whatever the case may be, the inclusion of the seven parameters resulted in a $\Sexpr{round(max(h3_final_results$search3$logLiks$logLik) - max(haiti3_orig_fit$search3$logLiks$logLik), 1)}$ increase in log-likelihood, which is a highly statistically significant difference.
+Whatever the case may be, the inclusion of the seven parameters resulted in a $\Sexpr{round(max(h3_final_results$search2$logLiks$logLik) - max(haiti3_orig_fit$search2$logLiks$logLik), 1)}$ increase in log-likelihood, which is a highly statistically significant difference.
 One could similarly test whether the inclusion of the initial value parameters resulted in a significantly better fit to the observed data, but we do not do this here due to the large computational effort that would be required.
@@
 In this article we claimed that we were able to obtain better fits to the observed data using the same models that were proposed by \cite{lee20}.
-Along with visual comparisons to the data, this claim was supported by comparing likelihoods and AIC values in Table~1.
+Along with visual comparisons to the data, this claim was supported by comparing likelihoods and AIC values in Table~2 in the manuscript. % TODO: Double check the table number matches
 Because model likelihoods were not provided by \cite{lee20}, it is necessary to replicate these models in order to obtain likelihood estimates.
@@
 In the following subsections, we use our \code{R} package \code{haitipkg} to reproduce some of the results of \cite{lee20}.
-This reproduction allows us to estimate the likelihoods of the \cite{lee20} version of Models~1--3, and also provides a demonstration of reproducibility.
+This reproduction allows us to estimate the likelihoods of the \cite{lee20} version of Models~1--3, and also provides a demonstration of the importance and usefulness of reproducible research.
 
@@
 According to the supplement materials, this was because model realizations from a single parameter set retained substantial variability, but multiple realizations from a collection of parameter sets resulted in a reasonable visual fit to the data.
-We are also inclined to believe that the use of multiple parameter values was in part intended to account for parameter uncertainty (as mentioned in our main text), an effort by the Model~1 authors that we applaud.
+We are also inclined to believe that the use of multiple parameter values was in part intended to account for parameter uncertainty---the importance of which was discussed in the main text---an effort by the Model~1 authors that we applaud.
 Simulations from each of the parameter sets, however, were treated with equal importance when being used to diagnose the model fit and make inference on the system.
-This is problematic given Figures S8 and S9 of the supplement material, which suggest that some parameter sets that were used for inference were several hundred log-likelihood units lower than other parameter sets.
-Such a large difference in log-likelihoods is well beyond the threshold of statistical uncertainty determined by Wilks' theorem.
-
-To fully reproduce the results of the Model~1 authors, it is necessary to use the exact same set of model parameters.
-Because the parameters used to obtain the result given by \cite{lee20} are not publicly available, we relied on the source code provided by the Model~1 authors to approximately recreate the parameter set.
+This is problematic given Figures S8 and S9 of the supplement material, which suggest that some parameter sets that were used for inference may have been several hundred units of log-likelihood lower than other parameter sets that were simultaneously used to make forecasts.
+Such a large difference in log-likelihoods is well beyond the threshold of statistical uncertainty determined by Wilks' theorem, resulting in the equal use of statistically inferior parameter sets in order to make forecasts and conduct inference on the system.
+
+To fully reproduce the results of the Model~1 authors, it is necessary to use the exact same set of model parameters that were originally used to obtain the results presented by \cite{lee20}.
+Because these parameter sets were not made publicly available, we relied on the source code provided by the Model~1 authors to approximately recreate the parameter set.
 Due to software updates since the publication of the source code, we were unable to produce the exact same set of parameters.
-Still, running the publicly available source code resulted in a set of parameters that are visually similar to those used by the Model~1 authors (See Figures~\ref{fig:PlotEpiDist} and \ref{fig:plotEndParams}).
+Running the publicly available source code, however, resulted in a set of parameters that are visually similar to those used by the Model~1 authors (See Figures~\ref{fig:PlotEpiDist} and \ref{fig:plotEndParams}).
 Furthermore, simulations using the set of parameters produced by the source code appear practically equivalent to those displayed by \cite{lee20} (See Figure~\ref{fig:plotMod1Sims}).
@@
 While the discrepancy appears minor, the deterministic nature of Model~2 implies that an exact replication of model trajectories should be possible.
-In this case, these discrepancies may be attributed to implementing the model in two different programing languages.
+In this case, these discrepancies may possibly be attributed to implementing the model and plotting the model trajectory in two different programming languages.
 Another potential explanation for the discrepancy is that the parameters that we used are only approximately the same as those used by \citet{lee20sup}.
@@
 In Fig.~\ref{fig:mod3rep}, we plot simulations from this model.
-This figure should be compared to Figure S18 of \cite{lee20}.
+This figure can be compared to Figure S18 of \cite{lee20}.
 We note that slight differences may be accounted for due to variance in the model simulations and the difference in programming language used to produce the figure.
+Overall, the high standard of reproducibility that was achieved by the Model~3 authors facilitated the ability to readily replicate their model and results.
 
@@
 In practice, because the likelihood maximization routines of iterated filtering methods are stochastic, it is common to run the iterated filtering method multiple times for each model in order to obtain a maximum likelihood estimate for model parameters.
-This results in a natural set of parameters near the MLE that could be used in in the discrete prior distribution.
+This results in a natural set of parameters near the MLE that could be used as the discrete prior distribution.
 

### 2023-08-23 ms-plos.Rnw
modified figure path

@@
     ),
-    fig.path = "testFigs/"
+    fig.path = "figure/"
 )

### 2023-08-24 si/si.Rnw
corrected notation for measurement models

@@
   \label{model1:meas}
-  Y_{n} \mid \Delta N_{EI}(n) = \reportChange \sim \text{NB}\left(\reportRate \reportChange, \obsOverdispersion \right),
+  Y_{n} \mid \Delta N_{E_{\cdot}I_{\cdot}}(n) = \reportChange \sim \text{NB}\left(\reportRate \reportChange, \obsOverdispersion \right),
 \end{equation}
-where $Y_n$ is the reported cholera cases at time $t_n \in t_1:t_N$ and $\Delta N_{EI}(n)$ is the sum total of individuals across each vaccination compartment $\vaccCounter \in 1:\vaccClass$ who moved from compartment $E_\vaccCounter$ to $I_\vaccCounter$ since observation $t_{n-1}$, and $\text{NB}\left(\reportRate \reportChange, \obsOverdispersion \right)$ is the negative binomial distribution with mean $\reportRate \reportChange$ and variance $\reportRate \reportChange \big(1 + \frac{\reportRate \reportChange}{\obsOverdispersion}\big)$.
+where $Y_n$ is the reported cholera cases at time $t_n \in t_1:t_N$ and $\Delta N_{E_{\cdot}I_{\cdot}}(n)$ is the sum total of individuals across each vaccination compartment $\vaccCounter \in 1:\vaccClass$ who moved from compartment $E_\vaccCounter$ to $I_\vaccCounter$ since observation $t_{n-1}$.
+Here, $\text{NB}\left(\reportRate \reportChange, \obsOverdispersion \right)$ denotes a negative binomial distribution with mean $\reportRate \reportChange$ and variance $\reportRate \reportChange \big(1 + \frac{\reportRate \reportChange}{\obsOverdispersion}\big)$.
 
@@
 
-As mentioned in the main text, Model~2 was fit using reported case counts that were transformed using the natural logarithm.
+Model~2 was fit using reported case counts that were transformed using the natural logarithm.
 We fit Model~2 using the subplex algorithm in the \code{subplex} package, using
-a Gaussian measurement model (Eq.~\myeqref{model2:meas}) on the log transformed cases as the loss function.
+a Gaussian measurement model (Eq.~\myeqref{model2:meas}) on the log transformed cases within each unit.
+The final loss function that is maximized is the product of the likelihoods of the individual units, or the sum of the log-likelihood of the individual units.
+The measurement model for individual units is given in Eq.~\myeqref{model2:meas}.
 \begin{equation}
   \label{model2:meas}
-  \log\big(Y_{u, n} + 1\big) \mid \Delta N_{EI}(n) = \reportChange \sim \text{N}\left(\log \big(\reportRate \reportChange + 1\big), \obsOverdispersion^2 \right),
+  \log\big(Y_{u, n} + 1\big) \mid \Delta N_{E_{u\cdot}I_{u\cdot}}(n) = \reportChange_u \sim \text{N}\left(\log \big(\reportRate \reportChange_u + 1\big), \obsOverdispersion^2 \right),
 \end{equation}
-where $\Delta N_{EI}(n)$ is the sum total of individuals across vaccination compartment $\vaccCounter \in 0:4$ and unit $u$ who moved from compartment $E_{u\vaccCounter}$ to $I_{u\vaccCounter}$ since observation $t_{n-1}$.
-Therefore, because the natural logarithm of observed case counts (plus one, to avoid taking the logarithm of zero) has a normal distribution, $Y_{u, n} + 1$ is assumed to follow a log-normal distribution with log-mean parameter $\mu = \log\big(\Delta N_{EI}(n) + 1\big)$ and log-variance $\obsOverdispersion^2$.
+where $\Delta N_{E_{u\cdot}I_{u\cdot}}(n)$ is the sum total of individuals across vaccination compartment $\vaccCounter \in 0:4$ within unit $u$ who moved from compartment $E_{u\vaccCounter}$ to $I_{u\vaccCounter}$ since observation $t_{n-1}$.
+Therefore, because the natural logarithm of observed case counts (plus one, to avoid taking the logarithm of zero) has a normal distribution, $Y_{u, n} + 1$ is assumed to follow a log-normal distribution with log-mean parameter $\log\big(\reportRate \Delta N_{E_{u\cdot}I_{u\cdot}}(n) + 1\big)$ and log-variance $\obsOverdispersion^2$.
 We note that fitting a model with this measurement model is equivalent to fitting using least squares, with $\log(Y_{u, n} + 1)$ as the response variable.
@@
   \label{model2:measEpi}
-  Y_{u, n} \mid \Delta N_{EI}(n) = \reportChange \sim \text{N}\left(\reportRate \reportChange, \obsOverdispersion^2 \right),
+  Y_{u, n} \mid \Delta N_{E_{u\cdot}I_{u\cdot}}(n) = \reportChange_u \sim \text{N}\left(\reportRate \reportChange_u, \obsOverdispersion^2 \right),
 \end{equation}
@@
   \label{model2:measEnd}
-  Y_{u, n} \mid \Delta N_{EI}(n) = \reportChange_1, \Delta N_{EA}(n) = \reportChange_2 \sim \text{N}\left(\reportRate \big(\reportChange_1 + \reportChange_2\big), \obsOverdispersion^2 \right),
+  Y_{u, n} \mid \Delta N_{E_{u\cdot}I_{u\cdot}}(n) = \reportChange_{u1}, \Delta N_{E_{u\cdot}A_{u\cdot}}(n) = \reportChange_{u2} \sim \text{N}\left(\reportRate \big(\reportChange_{u1} + \reportChange_{u2}\big), \obsOverdispersion^2 \right),
 \end{equation}
-where the notation for $\Delta N_{EA}(n)$ is similar to $\Delta N_{EI}(n)$, described above.
+where the notation for $\Delta N_{E_{u\cdot}A_{u\cdot}}(n)$ is similar to $\Delta N_{E_{u\cdot}I_{u\cdot}}(n)$, described above.
 
@@
   \label{model3:meas}
-  Y_{u, n} \mid \Delta N_{S_{u \vaccCounter}I_{u \vaccCounter}}(t) = \reportChange_u \sim \text{NB}\left(\reportRate \reportChange_{u}, \obsOverdispersion \right),
+  Y_{u, n} \mid \Delta N_{S_{u \cdot}I_{u}}(t) = \reportChange_u \sim \text{NB}\left(\reportRate \reportChange_{u}, \obsOverdispersion \right),
 \end{equation}
-where $\Delta N_{S_{u \vaccCounter}I_{u \vaccCounter}}(n)$ is the number of individuals who moved from compartment $S_{u \vaccCounter}$ to $I_{u \vaccCounter}$ since observation $t_{n-1}$.
+where $\Delta N_{S_{u \cdot}I_{u \vaccCounter}}(n)$ is the number of individuals who moved from compartment $S_{u \vaccCounter}$ to $I_{u}$ since observation $t_{n-1}$.
 

### 2023-08-25 ms-plos.Rnw
Changed to NB benchmark, a little commentary on rinit3

@@
 \newcommand\childReduce{q}
+\newcommand\NBintercept{\alpha}
+\newcommand\NBar{\beta}
+\newcommand\NBsize{\varphi}
 
@@
 
-<<CalculateARMALikelihood, echo=FALSE, message=FALSE, include=FALSE, cache=TRUE>>=
-
-ARMA_benchmarks <- list()
+<<CalculateBenchLikelihood, echo=FALSE, message=FALSE, include=FALSE, cache=TRUE>>=
+
+benchmarks <- list()
 
@@
 
-# Fit ARMA(2, 1) on log-cases
-m1_log_arma <- arima(log(m1_agg_data$cases + 1), order = c(2, 0, 1))
-
-# Fit ARMA(2, 1) on natural scale of cases
-m1_arma <- arima(m1_agg_data$cases, order = c(2, 0, 1))
-
-# Check which one is better, and save benchmark as the better one
-if (m1_log_arma$loglik - sum(log(m1_agg_data$cases + 1), na.rm = TRUE) >= m1_arma$loglik) {
-  ARMA_benchmarks[['m1']] <- m1_log_arma$loglik - sum(log(m1_agg_data$cases + 1), na.rm = TRUE)
-} else {
-  ARMA_benchmarks[['m1']] <- m1_arma$loglik
-}
-
-rm(m1_agg_data, m1_log_arma, m1_arma)
+# Fit Model 1 bechmark:
+m1_bench <- ar1_NegBinom(m1_agg_data$cases, init = 7364)
+benchmarks[['m1']] <- m1_bench$ll
+
+rm(m1_agg_data, m1_bench)
 gc()
@@
 h2_data <- haiti2_data(recreate = FALSE)
+
+# This are initial values for Y_u(0). They were estimated, but because they are
+# integers, close to the value of the first observation, it was easiest to fit these
+# by hand.
+m2_inits <- c(6449, 542, 0, 0, 16, 0, 0, 522, 0, 0)  # TODO: These were for the original initialization model, may need to fit these again by hand (just for-loop through possible inits.)
 
@@
   cases <- h2_data[h2_data$department == h2_dep_names[i], ]$cases
-  log_cases <- log(cases + 1)
-
-  m2_arma_dep_ll <- arima(cases, order=c(2, 0, 1))$loglik
-  m2_log_arma_dep_ll <- arima(log_cases, order=c(2, 0, 1))$loglik - sum(log_cases, na.rm = TRUE)
-
-  m2_benchmark <- m2_benchmark + max(m2_arma_dep_ll, m2_log_arma_dep_ll)
-}
-
-ARMA_benchmarks[['m2']] <- m2_benchmark
-
-rm(m2_benchmark, h2_dep_names, cases, log_cases, m2_arma_dep_ll, m2_log_arma_dep_ll, h2_data)
+  m2_bench <- ar1_NegBinom(data = cases, init = m2_inits[i])
+
+  m2_benchmark <- m2_benchmark + m2_bench$ll
+}
+
+benchmarks[['m2']] <- m2_benchmark
+
+rm(m2_benchmark, h2_dep_names, cases, h2_data, m2_bench, i)
 gc()
@@
 mod3_cases <- as.data.frame(t(h3Spat@data))
+m3_inits <- c(7362, 1914, 22, 8, 2818, 82, 2499, 6953, 44, 7) # TODO: Also fit by hand, but could be changed to initializing to the observation before.
 
@@
 
-  # Fit model on natural scale of cases
-  m3_a201 <- arima(cases, order = c(2, 0, 1))
-
-  # Fit model on log-scale of cases
-  m3_log_a201 <- arima(log(cases + 1), order = c(2, 0, 1))
-
-  m3_benchmark <- m3_benchmark + max(
-    m3_log_a201$loglik - sum(log(cases + 1), na.rm = TRUE),
-    m3_a201$loglik
-  )
-}
-
-ARMA_benchmarks[['m3']] <- m3_benchmark
-rm(cases, m3_a201, m3_log_a201, m3_benchmark, dep,
-   mod3_cases, temp, jumps, MODEL3_CASES, i, j)
+  m3_benchmark <- m3_benchmark + ar1_NegBinom(data = cases, init = m3_inits[dep])$ll
+}
+
+benchmarks[['m3']] <- m3_benchmark
+rm(cases, m3_benchmark, m3_bench, dep, mod3_cases)
 gc()
@@
 \hline
-\thead{Log-ARMA(2,1) \\ AIC} &
-  $\Sexpr{myround(-2 * ARMA_benchmarks$m1 + 2 * 5, 1)}$ &
-  $\Sexpr{myround(-2 * ARMA_benchmarks$m2 + 2 * 5, 1)}$ &
-  $\Sexpr{myround(-2 * ARMA_benchmarks$m3 + 2 * 5, 1)}$
+\thead{Benchmark \\ AIC} &
+  $\Sexpr{myround(-2 * benchmarks$m1 + 2 * 4, 1)}$ &
+  $\Sexpr{myround(-2 * benchmarks$m2 + 2 * 40, 1)}$ &
+  $\Sexpr{myround(-2 * benchmarks$m3 + 2 * 40, 1)}$
 \\
@@
 \end{tabular}
-\caption{\label{tab:likes}AIC values for each model compared to their ARMA benchmark.
+\caption{\label{tab:likes}AIC values for each model compared to their respective benchmarks.
 Values in parentheses are corresponding values using Lee et al.~\cite{lee20} parameter estimates.
@@
 In some cases, a possible benchmark model could be a generally accepted mechanistic model, but often no such model is available.
-Because of this, we use a log-linear Gaussian ARMA model as an associative benchmark, as recommended by He et al.~\cite{he10}.
-The theory and practice of ARMA models is well developed, and these linear models are appropriate on a log scale due to the exponential growth and decay characteristic of biological dynamics.
+Because of this, we use a simple negative binomial model with an auto regressive mean as our associative benchmark; this model is described in \myeqref{eq:AR_NB}.
+\begin{eqnarray}
+\label{eq:AR_NB}
+Y_n|Y_{n - 1} \sim \text{NB}\big(\NBintercept + \NBar Y_{n-1}, \NBsize \big),
+\end{eqnarray}
+where $\mathrm{E}\big(Y_n|Y_{n - 1}\big) = \NBintercept + \NBar Y_{n-1}$, and $\Var\big(Y_n|Y_{n - 1}\big) = \mathrm{E}\big(Y_n|Y_{n - 1}\big) + \mathrm{E}\big(Y_n|Y_{n - 1}\big)^2 / \NBsize$.
+In the case where the case counts are large, an alternative benchmark recommended by He et al.~\cite{he10} is a log-linear Gaussian ARMA model;
+the theory and practice of ARMA models is well developed, and these linear models are appropriate on a log scale due to the exponential growth and decay characteristic of biological dynamics.
+We chose to use the auto-regressive negative binomial model, however, because the large number of weeks with zero recorded cholera cases in department level data makes a benchmark based on a continuous distribution problematic.
 Likelihoods of Models~1--3 and their respective ARMA benchmark models are provided in Table~\ref{tab:likes}.

### 2023-08-25 si/si.Rnw
Changed to NB benchmark, a little commentary on rinit3

@@
 
-The latent states of this model are initialized by estimating of the number of infected, asymptomatic, and recovered individuals at time $t_0$ by using observed case counts at time $t_1$.
-The initial value for the bacteria compartment is then determined using the rainfall dynamics of Eq.~23.
+The latent states of this model are initialized by enforcing the model dynamics on the incidence data from the start of the recorded cases until time $t_0$, requiring that some of the available data be used to determine the initial values of the latent states.
+This is the same approach that was taken by \citet{lee20}, who used the value $t_0 = \text{2014-02-22}$; this choice of $t_0$ results in modeling roughly only $60\%$ of the available data, some of which is later discarded for alternative reasons \citep{lee20sup}.
+
+We do not see any immediate reason that this model could not be extended to cover a larger range of the data, and chose the value $t_0 = \text{2010-11-13}$.
+This choice of $t_0$ corresponds to using approximately $1\%$ of the available data to determine initial values of the latent states.
+In addition to modeling a larger portion of the available data, this choice of $t_0$ corresponds to an important real-world event, as daily reports from each of the departments were not being sent to the health ministry until November 10, 2010; % TODO: Cite
+this choice of $t_0$ therefore makes $\bm{Y}\big(t_1\big)$ the first week of complete data.
+
+% TODO: Talk about IVPs for 0 counts at t_0
+% Give more details on the initial model
+% Argue why we didn't need to do this for Model 2.
+% TODO: Change the benchmark model from ARMA -> NB
+
+While
+In order to retain a similar model for the initial values of the latent state, it was necessary to pick a value of $t_0$ that was greater than 2010-10-23 (the first available data points).
+
 For the departments of Grande'Anse, Nord-Ouest, and Sud-Est, estimating the latent states at time $t_0$ using the observed data at time $t_1$ is problematic (see Sec.~\ref{sec:mod3Cal}), so we instead estimate initial value of the latent state $I_{u0}(t_0)$ for the integers $u \in \{3, 7, 9\}$, which correspond to the aforementioned departments.

### 2023-08-29 si/si.Rnw
minor edit

@@
 In addition to modeling a larger portion of the available data, this choice of $t_0$ corresponds to an important real-world event, as daily reports from each of the departments were not being sent to the health ministry until November 10, 2010; % TODO: Cite
-this choice of $t_0$ therefore makes $\bm{Y}\big(t_1\big)$ the first week of complete data.
+this choice of $t_0$ therefore makes $\bm{Y}\big(t_1\big)$ the first week of data once daily reports are being sent to the health ministry.
 
@@
 % TODO: Change the benchmark model from ARMA -> NB
-
-While
-In order to retain a similar model for the initial values of the latent state, it was necessary to pick a value of $t_0$ that was greater than 2010-10-23 (the first available data points).
 

### 2023-08-29 si/si.Rnw
Model 3 initialization model fixed

@@
 this choice of $t_0$ therefore makes $\bm{Y}\big(t_1\big)$ the first week of data once daily reports are being sent to the health ministry.
-
-% TODO: Talk about IVPs for 0 counts at t_0
-% Give more details on the initial model
+The few observation times that exist before $t_0$ are used to initialize the model by enforcing model dynamics on these preliminary observations.
+For convenience, we denote these observations as $t_{-3}, t_{-2}$ and $t_{-1}$; as before, we let $y_{u,-k} = Y_{u}\big(t_{-k}\big)$ denote the observed case count for unit $u$ at time point $t_{-k}$, where $k \in 1:3$.
+Equations for the initial values of non-zero latent states are provided in Eqs.~\myeqref{eq:model3:I0}--\myeqref{eq:model3:B0}.
+
+\begin{eqnarray}
+\label{eq:model3:I0}
+I_{u}\big(t_0\big) &=& \frac{y^*_{u, -1}}{7\reportRate\left(\muIR + \left(\muDeath + \choleraDeath\right)/365\right)},
+\\%[-3pt]
+\label{eq:model3:A0}
+A_{u}\big(t_0\big) &=& \frac{I_{u0}\big(t_0\big)\left(1 - \symptomFrac\right)}{\symptomFrac},
+\\
+\label{eq:model3:R0}
+R_{u01}\big(t_0\big) &=& R_{u02}\big(t_0\big) = R_{u03}\big(t_0\big) = \left(\frac{\sum_{k = -3}^0 y^*_{u, k}}{\reportRate \symptomFrac} - \left(I_{u0}\big(t_0\big) + A_{u0}\big(t_0\big)\right)\right) / 3
+\\
+\label{eq:model3:S0}
+S_{u0}\big(t_0\big) &=& \mathrm{Pop}_u - I_{u}\big(t_0\big) - A_{u}\big(t_0\big) - \sum_{k=1}^3 R_{u0k}\big(t_0\big)
+\\
+\label{eq:model3:B0}
+B_{u}\big(t_0\big) &=& \big[1 + \seasAmplitude \tilde{J}^r \big] \mathrm{Den}_u \, \Wshed \big[ I_{u}(t_0)+ \asymptomRelativeShed A_{u_0}(t) \big]/\Wshed.
+\end{eqnarray}
+
+In Eq.~\myeqref{eq:model3:B0}, $\tilde{J} = 0.002376$ is the median adjusted rainfall over the observation period.
+One important consideration to make with this parameter initialization model is when $y_{u, -1}^* = 0$, which occurs for units $u \in \{3, 4\}$, which correspond to the Grand'Anse and Nippes departments, respectively.
+When this is the case, each of the infectious $I_u(t_0)$, asymptomatic $A_u(t_0)$, and bacterial reservoir $W_u(t_0)$ compartments have a value of zero.
+Because Model~3 models cholera transmission primarily by means of the bacterial reservoir, this makes it nearly impossible for an outbreak to occur.
+Therefore for units $u \in \{3, 4\}$, we introduce initial value parameters $I_{0,0}^3$ and $I_{0,0}^4$, and calibrate these parameter values using the data.
+The resulting parameter estimates are used to obtain the remaining non-zero initial values of the latent states using Eqs.~\myeqref{eq:model3:A0}--\myeqref{eq:model3:B0}.
+
 % Argue why we didn't need to do this for Model 2.
@@
 
-For the departments of Grande'Anse, Nord-Ouest, and Sud-Est, estimating the latent states at time $t_0$ using the observed data at time $t_1$ is problematic (see Sec.~\ref{sec:mod3Cal}), so we instead estimate initial value of the latent state $I_{u0}(t_0)$ for the integers $u \in \{3, 7, 9\}$, which correspond to the aforementioned departments.
-The remaining hidden states are estimated by enforcing model dynamics using this result.
 % TODO: Check this section.

### 2023-08-31 si/si.Rnw
Minor text edit

@@
 For convenience, we denote these observations as $t_{-3}, t_{-2}$ and $t_{-1}$; as before, we let $y_{u,-k} = Y_{u}\big(t_{-k}\big)$ denote the observed case count for unit $u$ at time point $t_{-k}$, where $k \in 1:3$.
-Equations for the initial values of non-zero latent states are provided in Eqs.~\myeqref{eq:model3:I0}--\myeqref{eq:model3:B0}.
+Equations for the initial values of non-zero latent states are provided in Eqs.~\myeqref{eq:model3:I0}--\myeqref{eq:model3:B0}; these equations match those that were used by \citet{lee20}, the primary change being a change to the value of $t_0$.
 

### 2023-08-31 si/si.Rnw
ARMA -> NB, plus changes to model3 fitting

--- 2023-08-29 si/si.Rnw
@@
 % Argue why we didn't need to do this for Model 2.
-% TODO: Change the benchmark model from ARMA -> NB
 
@@
 <<CalibrateOrig3>>=
-set.seed(8636712)
-
 # Create vectors for the unit and shared parameters
@@
 )
-
-# Vector containing all of the estimate parameter names.
 est_param_names <- c(
@@
 
-# This function sets the rw_sd for named parameters, first search.
+# Simple function that is used to set the rw_sd for the first search
+# This function is convenient so that we don't have to type out the rw_sd
+# for each unit parameter.
 set_rw_1 <- function(x) {
@@
     return(0.01)
-  } else if (x %in% paste0("k", 1:10)) {
-    return(0.01)
-  } else if (grepl('^foi_add[[:digit:]]+$', x) || grepl('^betaB[[:digit:]]+$', x)) {
+  # } else if (x %in% paste0(rep(c("aHur", 'hHur'), each = 2), c(3, 9))) {
+  #   return(expression(ifelse(time >= 2016.754 & time <= 2017, 0.015, 0)))
+  # } else if (x %in% c("Iinit3", "Iinit4")) {
+  #   return(expression(ivp(0.25)))
+  } else if (grepl('^foi_add[[:digit:]]+$', x)) {
+    return(0.015)
+  } else if (grepl('^betaB[[:digit:]]+$', x)) {
     return(0.01)
@@
 
-# This function sets the rw_sd for named parameters, second search.
 set_rw_2 <- function(x) {
@@
     return(0.003)
-  } else if (x %in% paste0("k", 1:10)) {
-    return(0.003)
-  } else if (grepl('^foi_add[[:digit:]]+$', x) || grepl('^betaB[[:digit:]]+$', x)) {
+  # } else if (x %in% paste0(rep(c("aHur", 'hHur'), each = 2), c(3, 9))) {
+  #   return(expression(ifelse(time >= 2016.754 & time <= 2017, 0.005, 0)))
+  # } else if (x %in% c("Iinit3", "Iinit4")) {
+  #   return(expression(ivp(0.125)))
+  } else if (grepl('^foi_add[[:digit:]]+$', x)) {
+    return(0.005)
+  } else if (grepl('^betaB[[:digit:]]+$', x)) {
     return(0.003)
@@
 
-# This function sets the rw_sd for named parameters, final search.
-set_rw_3 <- function(x) {
-  if (gsub("[[:digit:]]+$", "", x) %in% shared_param_names) {
-    return(0.0025)
-  } else if (x %in% paste0("k", 1:10)) {
-    return(0.0025)
-  } else if (grepl('^foi_add[[:digit:]]+$', x) || grepl('^betaB[[:digit:]]+$', x)) {
-    return(0.0025)
-  } else {
-    return(0)
-  }
-}
-
 # First Search RW
@@
 
-# Third Search RW
-reg_rw_3.sd <- lapply(est_param_names_expanded, set_rw_3)
-names(reg_rw_3.sd) <- est_param_names_expanded
-chol_rw_3.sd <- do.call(rw_sd, reg_rw_3.sd)
-
-# Search1 hyper-parameters
 SEARCH1 = list(
   NBPF = 100,
-  NP = 1000,
+  NP = 2500,
   SPAT_REGRESSION = 0.05,
-  NREPS = 72,
-  NP_EVAL = 1000,
-  NREPS_EVAL = 8,
+  NREPS = 144,
+  NP_EVAL = 2500,
+  NREPS_EVAL = 18,
   RW_SD = chol_rw_1.sd,
@@
   KEEP_TRACES = FALSE,
-  KEEP_LIK_MAT = FALSE
-)
-
-# Search2 hyper-parameters
+  KEEP_LIKE_MAT = FALSE
+)
+
 SEARCH2 = list(
-  TOP_N = 9,
+  TOP_N = 12,
   NBPF = 100,
-  NP = 1000,
+  NP = 2500,
   SPAT_REGRESSION = 0.05,
-  NREPS = 4,
-  NP_EVAL = 1000,
-  NREPS_EVAL = 9,
+  NREPS = 6,
+  NP_EVAL = 2500,
+  NREPS_EVAL = 18,
   RW_SD = chol_rw_2.sd,
@@
   KEEP_TRACES = FALSE,
-  KEEP_LIK_MAT = FALSE
-)
-
-# Search3 hyper-parameters
-SEARCH3 = list(
-  TOP_N = 6,
-  NBPF = 100,
-  NP = 1000,
-  SPAT_REGRESSION = 0.05,
-  NREPS = 6,
-  NP_EVAL = 1000,
-  NREPS_EVAL = 9,
-  RW_SD = chol_rw_3.sd,
-  COOLING = 0.5,
-  KEEP_TRACES = FALSE,
   KEEP_LIK_MAT = TRUE
@@
 
-N_SEARCHES <- 3L
-
-save_file <- "../model3/si/fit_orig_model.rds"
-
-# Fit the model and save the results
-# TODO: Remove SEARCH3 when re-running everything.
-haiti3_orig_fit <- bake(
-  file = save_file, {
-    fit_coupled_haiti3(
-      nsearches = N_SEARCHES,
+haiti3_sub_fit <- bake(
+  file = "../model3/si/subset_noparams.rds", {
+    fit_haiti3(
+      nsearches = 2L,
       search1 = SEARCH1,
       search2 = SEARCH2,
-      search3 = SEARCH3,
       search_rho = FALSE,
       search_gamma = FALSE,
+      search_Iinit = FALSE,
+      search_hur = FALSE,
       ncores = cores
@@
 # Find model index with best log-likelihood
-best_m <- which.max(haiti3_orig_fit$search3$logLiks$logLik)
+best_m <- which.max(haiti3_sub_fit$search2$logLiks$logLik)
 
 # Find parameters using best_m index
-best_params <- haiti3_orig_fit$search3$params[best_m, ]
+sub_params <- haiti3_sub_fit$search2$params[best_m, ]
 
 # Get unit log-likelihoods for the best model.
-all_liks <- as.data.frame(haiti3_orig_fit$search3$likMat)
-all_liks$which <- rep(1:(SEARCH3$TOP_N * SEARCH3$NREPS), each = SEARCH3$NREPS_EVAL)
+all_liks <- as.data.frame(haiti3_sub_fit$search2$likMat)
+# all_liks$which <- rep(1:(SEARCH3$TOP_N * SEARCH3$NREPS), each = SEARCH3$NREPS_EVAL)
+all_liks$which <- rep(1:(12 * 6), each = 18)
 best_mod <- filter(all_liks, which == best_m)
@@
 <<M3_bench, echo=FALSE, message=FALSE, include=FALSE, cache=TRUE>>=
-
-MODEL3_CASES <- haitiCholera |>
-  dplyr::rename(
-    date = date_saturday, Grande_Anse = Grand.Anse,
-    Nord_Est = Nord.Est, Nord_Ouest = Nord.Ouest,
-    Sud_Est = Sud.Est
-  ) |>
-  dplyr::mutate(date = as.Date(date)) |>
-  dplyr::select(-report) |>
-  dplyr::mutate(time_diff = date - lag(date)) |>
-  dplyr::filter(date >= "2010-11-20")
-
-# Create dataset that will be used to fit ARMA models. To do this, we
-# need to make sure that missing values are marked as NA, so we check
-# for any jumps in time.
-# MODEL3_CASES <- haitiCholera %>%
-#   dplyr::rename(
-#     date = date_saturday, Grande_Anse = Grand.Anse,
-#     Nord_Est = Nord.Est, Nord_Ouest = Nord.Ouest,
-#     Sud_Est = Sud.Est
-#   ) %>%
-#   dplyr::mutate(date = as.Date(date)) %>%
-#   dplyr::select(-report) %>%
-
-# Find jumps
-jumps <- which(MODEL3_CASES$time_diff > lubridate::make_difftime(day = 7))
-
-# If there is a jump, insert NA between observations
-temp <- MODEL3_CASES
-for (j in 1:length(jumps)) {
-  temp <- rbind(
-    temp[1:(jumps[j] + j - 2), ],
-    rep(NA, ncol(temp)),
-    temp[(jumps[j] + j - 1):nrow(temp), ]
-  )
-}
-
-mod3_cases <- temp %>% dplyr::select(-time_diff)
-
-mod3_departements <-
-  c(
-    'Artibonite',
-    'Centre',
-    'Grande_Anse',
-    'Nippes',
-    'Nord',
-    'Nord-Est',
-    'Nord-Ouest',
-    'Ouest',
-    'Sud',
-    'Sud-Est'
-  )
-
-m3_benchmark <- c()
-for (dep in mod3_departements) {
-
-  # Get department cases
-  cases <- as.numeric(mod3_cases[, gsub("-", "_", dep)])
-
-  # Fit model on natural scale of cases
-  m3_a201 <- arima(cases, order = c(2, 0, 1))
-
-  # Fit model on log-scale of cases
-  m3_log_a201 <- arima(log(cases + 1), order = c(2, 0, 1))
-
-  m3_benchmark <- c(m3_benchmark, max(
-    m3_log_a201$loglik - sum(log(cases + 1), na.rm = TRUE),
-    m3_a201$loglik
-  )
-  )
-}
-
-rm(cases, m3_a201, m3_log_a201, mod3_departements, dep,
-   mod3_cases, temp, jumps, MODEL3_CASES, j)
-
-unitLikesARMA <- data.frame(
-  unit = 1:10,
-  ll = m3_benchmark
-) %>%
-  as_tibble()
-@
-
-<<combineARMA_M3, echo=FALSE>>=
 h3 <- haiti3_spatPomp()
@@
 
+mod3_cases <- as.data.frame(t(h3@data))
+colnames(mod3_cases) <- deps
+m3_inits <- c(7362, 1914, 22, 8, 2818, 82, 2499, 6953, 44, 7) # TODO: Also fit by hand, but could be changed to initializing to the observation before.
+names(m3_inits) <- deps
+
+bench_cond_lls <- list()
+for (dep in deps) {
+
+  # Get department cases
+  cases <- as.numeric(mod3_cases[, dep])
+  mod <- ar1_NegBinom(data = cases, init = m3_inits[dep])
+
+  bench_cond_lls[[dep]] <- haitipkg:::.ar1_NegBinom_cond_ll(
+    theta = mod$theta, init = m3_inits[dep], data = cases, transform = FALSE
+  )
+}
+
+bench_cond_lls <- as.data.frame(bench_cond_lls)
+
+bench_lls <- apply(bench_cond_lls, 2, sum)
+
+unitLikes_bench <- data.frame(
+  unit = 1:10,
+  # dep = names(bench_lls),
+  ll = bench_lls
+) %>%
+  as_tibble()
+
+bench_cond_lls$date <- lubridate::round_date(lubridate::date_decimal(h3@times), unit = 'day')
+@
+
+<<combine_bench_M3, echo=FALSE>>=
+
 # Combine tables with model 3 and benchmark likelihoods.
-ARMA_M3_joined <- dplyr::left_join(
+bench_M3_joined <- dplyr::left_join(
   x = unitLikesM3,
-  y = unitLikesARMA,
+  y = unitLikes_bench,
   by = "unit"
@@
 # search
-coef(h3)[names(best_params)] <- best_params
+coef(h3)[names(sub_params)] <- sub_params
 
@@
 # log-likelihoods.
-bpf_orig3_out <- bake(
-  file = "../model3/si/bpf_orig3_cond_LL.rds", {
+bpf_sub3_out <- bake(
+  file = "../model3/si/bpf_sub_cond_LL.rds", {
 
@@
     bpf_out <- h3 %>% bpfilter(
-      Np = 2000,
+      Np = 5000,
       block_size = 1,
@@
 In this section, we provide more detail on the process that was used to estimate the coefficients used in Model~3.
-In particular, we discuss why we decided to include additional model parameters---those that are associated with the initial conditions of the model and the behavior of the system during Hurricane Matthew---that were not considered by \cite{lee20}.
+In particular, we discuss why we decided to include additional model parameters---those that are associated with the behavior of the system during Hurricane Matthew---that were not considered by \cite{lee20}.
 To calibrate this model, we used the iterated block particle filter (IBPF) method of \cite{ionides22}.
@@
 \cite{lee20} were only able to estimate model parameters to a simplified version of Model~3 on a subset of the available data, as no method existed at the time of their publication to fit a fully coupled meta-population model to disease incidence data.
-Building on their results, we fit the fully coupled version of Model~3 to all available data.
+Building on their results, we fit the fully coupled version of Model~3 to nearly all available data, reserving only a few observations to use to calibrate the initial conditions of the model (see the previous section for more details).
 In order to properly maximize the model likelihood, we found that it was necessary to use multiple searches for the MLE, periodically pruning away less successful searches.
@@
 
-We use the technique described above to fit the fully coupled version of Model~3 proposed by \cite{lee20} to all observed cholera data.
-The maximum likelihood we obtained with the model was $\Sexpr{round(max(haiti3_orig_fit$search2$logLiks$logLik))}$, which is higher than the log-ARMA benchmark ($\Sexpr{round(sum(unitLikesARMA$ll))}$).
-% TODO: Check this, numbers don't match up.
+We use the technique described above to fit the fully coupled version of Model~3 proposed by \cite{lee20}.
+The maximum likelihood we obtained with the model was $\Sexpr{round(max(haiti3_sub_fit$search2$logLiks$logLik))}$, which is higher than the benchmark model ($\Sexpr{round(sum(unitLikes_bench$ll))}$).
+While beating a simple associative benchmark is promising, this does not immediately rule out the possibility that the model is a poor description of the system.
+Additional investigation of parameters estimates and their corresponding implications on model based conclusions should always be conducted.
 For meta-population models, it is worth considering how well the model fits the data to each spatial unit.
 This can be done by looking at conditional log-likelihoods, which is part of the output of the \texttt{bpfilter} algorithm in the \texttt{spatPomp} package.
-The likelihoods for each department, compared to the corresponding log-ARMA benchmark, are displayed in Fig.~\ref{fig:h3UnitLikes}.
-The figure demonstrates that while the log-likelihood of the fitted model outperforms the log-ARMA benchmark, the model has lower likelihoods than the benchmark for some departments.
-This result warrants further investigation of the model fit.
+The likelihoods for each department, compared to the corresponding benchmark model, are displayed in Fig.~\ref{fig:h3UnitLikes}.
+The figure demonstrates that while the log-likelihood of the fitted model outperforms the auto-regressive negative binomial benchmark model, Model~3 has lower likelihoods than the benchmark for some departments.
 
@@
 <<h3UnitLikes, fig.height=3, fig.width=5.5, echo=FALSE>>=
-ggplot(ARMA_M3_joined, aes(x = ll, y = reorder(unit, ll))) +
+ggplot(bench_M3_joined, aes(x = ll, y = reorder(unit, ll))) +
   geom_point(aes(col = type)) +
@@
   scale_color_manual(
-    labels = c("ll" = "ARMA", "ll_est" = "Model 3"),
+    labels = c("ll" = "Benchmark", "ll_est" = "Model 3"),
     values = c("ll" = "#1b9e77", "ll_est" = "#d95f02")
@@
 @
-\caption{\label{fig:h3UnitLikes}Log-likelihoods of Model~3 for each department compared to the corresponding log-ARMA benchmark prior to the inclusion of parameters for initial values and Hurricane Matthew.}
+\caption{\label{fig:h3UnitLikes}Log-likelihoods of Model~3 for each department compared to the corresponding benchmark model prior to the inclusion of parameters for initial values and Hurricane Matthew.}
 \end{figure}
 
-In an attempt to determine why the model is underperforming in some departments, we plot the conditional log likelihoods of each observation given the previous observations for each department in Fig.~\ref{fig:condLL}.
+In addition to considering the conditional log-likelihoods of each unit, one can consider conditional log-likelihoods of each observation.
+When compared to a benchmark, this level of detail can provide useful information about which observations are well described by the model and which are poorly described.
+In Fig.~\ref{fig:condLL}, we plot the conditional log-likelihoods of each observation.
 This figure reveals that the fitted model has difficult time explaining certain features of the data.
@@
 # Plot conditional log-likelihoods for each unit
-t(bpf_orig3_out$block.cond.loglik) %>%
+t(bpf_sub3_out$block.cond.loglik) %>%
   as.data.frame() %>%
-  mutate(t = bpf_orig3_out$times) %>%
+  mutate(t = bpf_sub3_out$times) %>%
   pivot_longer(
@@
 # Combine model likelihood table to benchmarks
-ARMA_M3_joined <- dplyr::left_join(
+bench_M3_joined <- dplyr::left_join(
   x = unitLikesM3,
-  y = unitLikesARMA,
+  y = unitLikes_bench,
   by = "unit"
@@
 <<finalUnitLLFig, fig.height=3, fig.width=5.5, echo=FALSE>>=
-ggplot(ARMA_M3_joined, aes(x = ll, y = reorder(unit, ll))) +
+ggplot(bench_M3_joined, aes(x = ll, y = reorder(unit, ll))) +
   geom_point(aes(col = type)) +
@@
 @
-\caption{\label{fig:finalUnitLL}Log-likelihoods of Model~3 for each department compared to the corresponding log-ARMA benchmark after adding and estimating parameters related to initial values and Hurricane Matthew.}
+\caption{\label{fig:finalUnitLL}Log-likelihoods of Model~3 for each department compared to the corresponding benchmark model after adding and estimating parameters related to initial values and Hurricane Matthew.}
 \end{figure}
@@
 This could be due to many factors, including the possibility that the coupling between units forces low initial value parameters in a given department in order for the model to fit the data in the remaining departments.
-Whatever the case may be, the inclusion of the seven parameters resulted in a $\Sexpr{round(max(h3_final_results$search2$logLiks$logLik) - max(haiti3_orig_fit$search2$logLiks$logLik), 1)}$ increase in log-likelihood, which is a highly statistically significant difference.
+Whatever the case may be, the inclusion of the seven parameters resulted in a $\Sexpr{round(max(h3_final_results$search2$logLiks$logLik) - max(haiti3_sub_fit$search2$logLiks$logLik), 1)}$ increase in log-likelihood, which is a highly statistically significant difference.
 One could similarly test whether the inclusion of the initial value parameters resulted in a significantly better fit to the observed data, but we do not do this here due to the large computational effort that would be required.

### 2023-08-31 ms-plos.Rnw
Model 3 initialization model fixed
Made most of final changes, need to adjust how supplement and ms interact

@@
 # by hand.
-m2_inits <- c(6449, 542, 0, 0, 16, 0, 0, 522, 0, 0)  # TODO: These were for the original initialization model, may need to fit these again by hand (just for-loop through possible inits.)
+m2_inits <- c(6449, 542, 0, 0, 16, 0, 0, 522, 0, 0)  # fit by hand
 
@@
 mod3_cases <- as.data.frame(t(h3Spat@data))
-m3_inits <- c(7362, 1914, 22, 8, 2818, 82, 2499, 6953, 44, 7) # TODO: Also fit by hand, but could be changed to initializing to the observation before.
+m3_inits <- c(7362, 1914, 22, 8, 2818, 82, 2499, 6953, 44, 7) # fit by hand
 
@@
 % Here, we have put all three models into a common notational framework.
-Translations back to the original notation of Lee et al.~\cite{lee20} are given in Table~S-1 of the supplement. % TODO: Check S-1 table matches.
+Translations back to the original notation of Lee et al.~\cite{lee20} are given in the supplement (\nameref{S1_Table}).
 
@@
 \centering
+\caption{\label{tab:params} {\bf Model parameters.}}
 % \small
@@
   &
-  & \multirow{2}{*}{${\fixed \muDeath^{-1} = \Sexpr{myround(1/p3["mu1"],1)}}$}
+  & ${\fixed \muDeath^{-1} = \Sexpr{myround(1/p3["mu1"],1)}}$
   & \multirow{2}{*}{\eqref{eq:model3:IS}}
@@
 &
-& &
+& ${\fixed \choleraDeath^{-1} = \Sexpr{myround(1/p3["alpha1"],2)}}$
+&
 \\
@@
   & $\reportRate=\Sexpr{myround(p1["rho"], 3)}$
-  & (S15) % TODO: Check
+  & (S15)
   & ${\fixed \reportRate=\Sexpr{myround(h2_fit$h2_params["Rho"],2)}}$
-  & (S16) % TODO: Check
+  & (S16)
   & $\reportRate=\Sexpr{myround(p3s["epsilon"],2)}$
-  & (S19) % TODO: Check
+  & (S19)
 \\
@@
   & $\obsOverdispersion =$ \newline $(\Sexpr{myround(p1["tau_epi"],2)}, \Sexpr{round(p1["tau_end"],2)})$
-  & \multirow{2}{*}{(S15)} % TODO: Check
+  & \multirow{2}{*}{(S15)}
   & \multirow{2}{*}{$\obsOverdispersion = \Sexpr{signif(h2_fit$h2_params['sigma'], 3)}$}
-  & \multirow{2}{*}{(S16)} % TODO: Check
+  & \multirow{2}{*}{(S16)}
   & \multirow{2}{*}{$\obsOverdispersion = \Sexpr{myround(p3s["k"],2)}$}
-  & \multirow{2}{*}{(S19)} % TODO: Check
+  & \multirow{2}{*}{(S19)}
 \\
@@
   &
-  & \multirow{2}{*}{$\Iinit^{3,7,10} = (
-        \Sexpr{round(prod(p3[c('H3', 'Iinit3')]))}, \Sexpr{round(prod(p3[c('H7', 'Iinit7')]))}, \Sexpr{round(prod(p3[c('H10', 'Iinit10')]))})^*$} % TODO: Check the unit numbers
-  &  % TODO: \multirow{2}{*}{EQUATION NUMBER}
+  & \multirow{2}{*}{$\Iinit^{3,4} = (
+        \Sexpr{round(prod(p3[c('H3', 'Iinit3')]))}, \Sexpr{round(prod(p3[c('H4', 'Iinit4')]))})^*$}
+  &  \multirow{2}{*}{(S20)}
   \\
@@
 \end{tabular}
-\caption{\label{tab:params}References to the relevant equation are given in parentheses.
+\begin{flushleft} References to the relevant equation are given in parentheses.
 Parameters in blue were fixed based on scientific reasoning and not fitted to the data.
@@
 Translations back into the notation of \cite{lee20} are given in Table~S1.
-}
+\end{flushleft}
 \end{adjustwidth}
@@
 A first-time reader may choose to skim through the rest of this section, and return later.
-Additional details about the numeric implementation of these models are provided in Sec.~S1 and S2 of the supplement.
+Additional details about the numeric implementation of these models are provided in Sec.~S1 and S2 of the supplemental text (\nameref{S1_Text}).
 
@@
 \label{eq:model3:water}
-\mu_{{\demography}W_u} &=& \big[1 + \seasAmplitude \big(J_u(t))^r \big] \mathrm{Den}_u \, \Wshed \big[ I_{u}(t)+ \asymptomRelativeShed A_{u}(t) \big],
+\mu_{{\demography}W_u} &=& \big[1 + \seasAmplitude \big(J_u(t)\big)^r \big] \mathrm{Den}_u \, \Wshed \big[ I_{u}(t)+ \asymptomRelativeShed A_{u}(t) \big],
 \\%[-3pt]
@@
 \centering
-
+\caption{\label{tab:likes}{\bf AIC values for Models~1--3 and their benchmarks.}}
 \begin{tabular}{|c|c|c|c|}
@@
 \end{tabular}
-\caption{\label{tab:likes}AIC values for each model compared to their respective benchmarks.
+\begin{flushleft}
 Values in parentheses are corresponding values using Lee et al.~\cite{lee20} parameter estimates.
-$\textsuperscript{1}$The reported likelihood is an upper bound of the likelihood of the Lee et al.~\cite{lee20} model.
-$\textsuperscript{2}$ Lee et al.~\cite{lee20} fit Model~3 to a subset of the data (March 2014 onward, excluding data from Ouest in 2015-2016).
+\textsuperscript{1}The reported likelihood is an upper bound of the likelihood of the Lee et al.~\cite{lee20} model.
+\textsuperscript{2}Lee et al.~\cite{lee20} fit Model~3 to a subset of the data (March 2014 onward, excluding data from Ouest in 2015-2016).
 On this subset, their model has a likelihood of $\Sexpr{myround(lee3_results$sub_ll_eval, 1)}$.
@@
 See Sec.~S6 of the supplement material for more details on estimating the likelihood of the Lee et al.~\cite{lee20} models.
-}
+\end{flushleft}
 \end{table}
@@
 
-% TODO: Remove the unused code below.
-% % Place tables after the first paragraph in which they are cited.
-% \begin{table}[!ht]
-% \begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
-% \centering
-% \caption{
-% {\bf Table caption Nulla mi mi, venenatis sed ipsum varius, volutpat euismod diam.}}
-% \begin{tabular}{|l+l|l|l|l|l|l|l|}
-% \hline
-% \multicolumn{4}{|l|}{\bf Heading1} & \multicolumn{4}{|l|}{\bf Heading2}\\ \thickhline
-% $cell1 row1$ & cell2 row 1 & cell3 row 1 & cell4 row 1 & cell5 row 1 & cell6 row 1 & cell7 row 1 & cell8 row 1\\ \hline
-% $cell1 row2$ & cell2 row 2 & cell3 row 2 & cell4 row 2 & cell5 row 2 & cell6 row 2 & cell7 row 2 & cell8 row 2\\ \hline
-% $cell1 row3$ & cell2 row 3 & cell3 row 3 & cell4 row 3 & cell5 row 3 & cell6 row 3 & cell7 row 3 & cell8 row 3\\ \hline
-% \end{tabular}
-% \begin{flushleft} Table notes Phasellus venenatis, tortor nec vestibulum mattis, massa tortor interdum felis, nec pellentesque metus tortor nec nisl.
-Ut ornare mauris tellus, vel dapibus arcu suscipit sed.
-% \end{flushleft}
-% \label{table1}
-% \end{adjustwidth}
-% \end{table}
-
 \section*{Discussion}\label{sec:discussion}
@@
 % Include only the SI item label in the paragraph heading. Use the \nameref{label} command to cite SI items in the text.
-\paragraph*{S1 Fig.}
-\label{S1_Fig}
-{\bf Bold the title sentence.} Add descriptive text after the title of the item (optional).
+\paragraph*{S1 Table}
+\label{S1_Table}
+{\bf Notation conversion table.} Conversions between the notation used here and the notation of Lee et al. (2020) \cite{lee20}.
+
+\paragraph*{S1 Text}
+\label{S1_Text}
+{\bf Supplemental Material.} In depth description of the numeric implementation of models.
 
@@
 
-\paragraph*{S1 Table.}
-\label{S1_Table}
-{\bf Lorem ipsum.} Maecenas convallis mauris sit amet sem ultrices gravida.
+% \paragraph*{S1 Table.}
+% \label{S1_Table}
+% {\bf Lorem ipsum.} Maecenas convallis mauris sit amet sem ultrices gravida.
 Etiam eget sapien nibh.

### 2023-08-31 si/si.Rnw
Made most of final changes, need to adjust how supplement and ms interact

@@
 \usepackage{amsmath,amsfonts,amssymb,graphicx}
+\usepackage{pict2e}
+\usepackage{curve2e}
 
@@
 \newcommand\reportChange{\delta}
+\newcommand\NBintercept{\alpha}
+\newcommand\NBsize{\varphi}
 
@@
 \usepackage{color}
+\usepackage[section]{placeins}
 \usepackage[normalem]{ulem}% to use \sout in feedback commands
@@
 \newcommand\ar[2]{\sout{#1} \textcolor{cyan}{#2}}
-\newcommand\arc[1]{\textcolor{cyan}{#1}}
+% \newcommand\arc[1]{\textcolor{cyan}{#1}}
 % light brown for KT
@@
 
+\makeatletter
+\AtBeginDocument{%
+  \expandafter\renewcommand\expandafter\subsection\expandafter{%
+    \expandafter\@fb@secFB\subsection
+  }%
+}
+\makeatother
+
 %% FOR PSEUDOCODE SETUP
@@
 \usepackage{fullpage}
+% \usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,justification=raggedright,singlelinecheck=off]{caption}
+% \usepackage[labelfont=bf,labelsep=period]{caption} % For some reason this doesn't work with .Rnw file, but it does with just the tex file, so after compiling you can edit this so that it works.
+\renewcommand{\figurename}{Fig}
 
@@
 \newpage
+
+\section{Model Diagrams}
+
+Each of the dynamic models considered in this manuscript can be fully described using the model descriptions in the manuscript, coupled with the additional information described in Sections 2 and 3 of this supplement.
+Despite this, diagrams of dynamic systems are often helpful to understand the equations.
+In this section, we give three diagrams representing Models~1--3, respectively.
+Because the models are defined by their mathematical equations and numeric implementation, these diagrams are not unique visual representations of the model.
+Alternative representations that may be helpful in understanding the models explored in this paper were made by \citet{lee20sup}.
+
+\subsection{Model~1}
+
+\begin{figure}[!h]
+%%%%% SEAIR diagram
+\begin{center}
+  \resizebox{0.9\textwidth}{!}{
+    \Large
+    \setlength{\unitlength}{5pt}
+    \begin{picture}(100,95)(0,15)
+      \thicklines
+
+      % Unvaccinated Compartments
+      \put(10,63){\framebox(6,6){$\mathrm{S}_0$}}
+      \put(34,63){\framebox(6,6){$\mathrm{E}_0$}}
+      \put(58,71){\framebox(6,6){$\mathrm{A}_0$}}
+      \put(82,63){\framebox(6,6){$\mathrm{R}_0$}}
+      \put(58,55){\framebox(6,6){$\mathrm{I}_0$}}
+
+      % Births
+      \put(1, 66){\vector(1, 0){8}}
+      \put(4, 68){$\muBirth$}
+
+      % S_0 -> E_0
+      \put(17,66){\vector(1,0){16}}
+      \put(22,67){$\lambda(t)$}
+
+      % E_0 -> A_0
+      \put(41,67.5){\vector(2.1,1){16}}
+      % \put(37, 72){$\muEI\big(1-\symptomFrac_{\vaccCounter}(t)\big)$}
+
+      % E_0 -> I_0
+      \put(41,65.5){\vector(2.1,-1){16}}
+
+      % A_0 -> R_0
+      \put(65,74){\vector(2.1,-1){16}}
+
+      % I_0 -> R_0
+      \put(65,59){\vector(2.4,1){16}}
+
+      % R_0 -> S_0
+      \cbezier(84,70)(75, 95)(23,95)(14,70)
+      \put(14,70){\vector(-1, -2){0.3}}
+
+      % Unvaccinated death rates
+      \put(13, 70){\vector(0, 1){10}}
+      \put(11, 74){$\muDeath$}
+
+      \put(37, 70){\vector(0, 1){10}}
+      \put(35, 74){$\muDeath$}
+
+      \put(61, 78){\vector(0, 1){9}}
+      \put(59, 80){$\muDeath$}
+
+      \put(61, 62){\vector(1.5, 2){6}}
+      \put(61.5, 65.5){$\muDeath$}
+
+      \put(85, 70){\vector(0, 1){10}}
+      \put(83, 74){$\muDeath$}
+
+      % Vaccinated Compartments
+      \put(10,37){\framebox(6,6){$\mathrm{S}_\vaccCounter$}}
+      \put(34,37){\framebox(6,6){$\mathrm{E}_\vaccCounter$}}
+      \put(58,45){\framebox(6,6){$\mathrm{A}_\vaccCounter$}}
+      \put(82,37){\framebox(6,6){$\mathrm{R}_\vaccCounter$}}
+      \put(58,29){\framebox(6,6){$\mathrm{I}_\vaccCounter$}}
+
+      % S -> E
+      \put(17,40){\vector(1,0){16}}
+      \put(22,41){$\lambda(t)$}
+
+      % E -> A
+      \put(41,40.5){\vector(2.1,1){16}}
+
+      % E -> I
+      \put(41,39.5){\vector(2.1,-1){16}}
+
+      % A -> R
+      \put(65,48){\vector(2.1,-1){16}}
+
+      % I -> R
+      \put(65,32){\vector(2.1,1){16}}
+
+      % R -> S
+      \cbezier(84, 36)(75, 10)(23, 10)(14, 36)
+      \put(14,36){\vector(-1, 2){0.3}}
+
+      % Vaccinated death rates
+      \put(13, 36){\vector(0, -1){10}}
+      \put(11, 31){$\muDeath$}
+
+      \put(37, 36){\vector(0, -1){10}}
+      \put(35, 31){$\muDeath$}
+
+      \put(61, 28.5){\vector(0, -1){9}}
+      \put(59, 24){$\muDeath$}
+
+      \put(61, 44){\vector(1.5, -2){6}}
+      \put(62, 38){$\muDeath$}
+
+      \put(85, 36){\vector(0, -1){10}}
+      \put(83.2, 30.2){$\muDeath$}
+
+      % S_0 -> S_\vaccCounter
+      \multiput(13,62)(0,-2){9}{\line(0,-1){1}}
+      \put(13, 43.5){\vector(0, -1){0}}
+
+      % E_0 -> E_\vaccCounter
+      \multiput(37,62)(0,-2){9}{\line(0,-1){1}}
+      \put(37, 43.5){\vector(0, -1){0}}
+
+      % R_0 -> R_\vaccCounter
+      \multiput(85,62)(0,-2){9}{\line(0,-1){1}}
+      \put(85, 43.5){\vector(0, -1){0}}
+
+      % A_0 -> A_\vaccCounter
+      \multiput(57.5,72)(-1.9, 0){3}{\line(-1, 0){0.9}}
+      \put(52.2, 72){\line(-2.1, -1){1}}
+      \multiput(51,71)(0, -2){11}{\line(0, -1){1}}
+      \put(52.2, 49){\line(-2.1, 1){1}}
+      \multiput(54,49)(1.9, 0){2}{\line(-1, 0){0.9}}
+      \put(57, 49){\vector(1, 0){0}}
+
+      % I_0 -> I_\vaccCounter
+      \multiput(65, 57.5)(1.9, 0){3}{\line(1, 0){0.9}}
+      \put(70.6, 57.5){\line(2.1, -1){1}}
+      \multiput(72, 56.5)(0, -2){13}{\line(0, -1){1}}
+      \put(72, 30.9){\line(-2, -1.2){0.9}}
+      \multiput(65.6, 30.2)(1.9, 0){3}{\line(1, 0){0.9}}
+      \put(64.5, 30.2){\vector(-1, 0){0}}
+
+    \end{picture}
+  }
+\end{center}
+\caption{A flow diagram for the SEAIR model.}\label{fig:flow_diagram}
+\end{figure}
+
+\subsection{Model~2}
+
+\begin{figure}[!h]
+%%%%% SEAIR diagram
+\begin{center}
+  \resizebox{\textwidth}{!}{
+    \Large
+    \setlength{\unitlength}{5pt}
+    \begin{picture}(100,95)(0,15)
+      \thicklines
+
+      % Unvaccinated Compartments
+      \put(10,63){\framebox(6,6){$\mathrm{S}_{u0}$}}
+      \put(34,69){\framebox(6,6){$\mathrm{E}_{u0}$}}
+      \put(58,77){\framebox(6,6){$\mathrm{A}_{u0}$}}
+      \put(82,77){\framebox(6,6){$\mathrm{R^A}_{u0}$}}
+      \put(82,61){\framebox(6,6){$\mathrm{R^I}_{u0}$}}
+      \put(58,61){\framebox(6,6){$\mathrm{I}_{u0}$}}
+
+      % % S_0 -> E_0
+      \put(17,66){\vector(3,1.1){16}}
+      % \put(22,67){$\lambda(t)$}
+
+      % % E_0 -> A_0
+      \put(41,72.5){\vector(2.1,1){16}}
+      % % \put(37, 72){$\muEI\big(1-\symptomFrac_{\vaccCounter}(t)\big)$}
+
+      % % E_0 -> I_0
+      \put(41,71.5){\vector(2.1,-1){16}}
+
+      % % A_0 -> R_0
+      \put(65,80.5){\vector(1,0){16}}
+
+      % A_0 -> W_u
+      % \multiput(65, 79.8)(1, -1){8}{\line(1, -1){0.25}}
+
+      % I_0 -> W_u
+      % \multiput(65, 64.2)(1, 1){8}{\line(1, 1){0.25}}
+
+      % Long verticle dotted line
+      % \multiput(72.9, 72)(0, -1.41){13}{\line(0, -1){0.25}}
+
+      % Long backwards horizontal dotted line
+      % \multiput(73, 54)(-1.41, 0){22}{\line(0, -1){0.25}}
+      % \multiput(73, 34)(0, 1.41){15}{\line(0, 1){0.25}}
+      % \put(42, 53.9){\vector(-1, 0){0.2}}
+
+      % A_uz -> W_u
+      \put(58,39){\framebox(6,6){$\mathrm{A}_{u\vaccCounter}$}}
+      \qbezier[25](60.5, 46)(51, 52)(42, 53)
+      \put(42, 53){\vector(-1, 0.1){0.5}}
+
+      % A_uz -> W_u
+      \qbezier[25](60.5, 60)(51, 54)(42, 54)
+      \put(42, 54){\vector(-1, -0.1){0.5}}
+
+      % I_uz -> W_u
+      \qbezier[30](60.5,30)(48, 45)(40, 50)
+      \put(40, 50){\vector(-1, 0.8){0.5}}
+
+      % A_u0 -> W_u
+      \qbezier[29](60.5,76)(48, 61)(40.6, 57)
+      \put(40.8, 57.1){\vector(-1, -0.7){0.5}}
+
+      % A + I_\vaccCounter -> W_u
+      % \multiput(65, 41.8)(1, -1){8}{\line(1, -1){0.25}}
+      % \multiput(65, 26.2)(1, 1){8}{\line(1, 1){0.25}}
+
+      % W_u -> backwards diags.
+      % \multiput(33, 57)(-1, 1.2){10}{\line(-1, 1.2){0.1}}
+      % \put(24.5, 67){\vector(-1, 1.2){0.5}}
+      % \multiput(33, 51)(-1, -1.2){10}{\line(-1, -1.2){0.1}}
+      % \put(24, 40){\vector(-1, -1.2){0.5}}
+      %\put(65, 80){\vector(2, -1){8}}
+
+      % % I_0 -> R_0
+      \put(65,63.5){\vector(1,0){16}}
+
+      % % Vaccinated Compartments
+      \put(10,37){\framebox(6,6){$\mathrm{S}_{u\vaccCounter}$}}
+      \put(34,31){\framebox(6,6){$\mathrm{E}_{u\vaccCounter}$}}
+      \put(58,39){\framebox(6,6){$\mathrm{A}_{u\vaccCounter}$}}
+      \put(82,39){\framebox(6,6){$\mathrm{R^A}_{u\vaccCounter}$}}
+      \put(58,23){\framebox(6,6){$\mathrm{I}_{u\vaccCounter}$}}
+      \put(82,23){\framebox(6,6){$\mathrm{R^I}_{u\vaccCounter}$}}
+
+      % R_0 -> S_0
+      \cbezier(85,22)(80, 8)(18,8)(12,36)
+      \put(12.1,35.5){\vector(-1, 3.5){0.3}}
+
+      \cbezier(85,38)(80, 12)(18,15)(14,36)
+      \put(14.1,35.5){\vector(-1, 3.5){0.3}}
+
+      % R_0 -> S_0
+      \cbezier(85,84)(80, 100)(18,100)(12,70)
+      \put(12.1,70.5){\vector(-1, -3.5){0.3}}
+
+      \cbezier(85,68)(80, 95)(18, 92)(14,70)
+      \put(14.1,70.5){\vector(-1, -3.5){0.3}}
+
+      % % S -> E
+      \put(17,40){\vector(3,-1.1){16}}
+      % \put(22,41){$\lambda(t)$}
+
+      % % E -> A
+      \put(41,34.5){\vector(2.1,1){16}}
+
+      % % E -> I
+      \put(41,33.5){\vector(2.1,-1){16}}
+
+      % % A -> R
+      \put(65,42.5){\vector(1,0){16}}
+
+      % % I -> R
+      \put(65,25.5){\vector(1,0){16}}
+
+      % S_0 -> S_\vaccCounter
+      \multiput(12,62)(0,-2){9}{\line(0,-1){1}}
+      \put(12, 44){\vector(0, -1){0}}
+
+      % S_0 -> S_\vaccCounter
+      % Wanning vaccine efficacy
+      \multiput(14,44)(0,2){9}{\line(0,1){1}}
+      \put(14, 62){\vector(0, 1){0}}
+
+      \put(37, 54){\circle{7}}
+      \put(35, 53){$\mathrm{W_u}$}
+
+      \qbezier[10](32.5, 54)(29.5, 54)(26.5, 54)
+      \put(26, 54){\vector(-1, 0){0.1}}
+      \put(28, 55){$\Wremoval$}
+
+    \end{picture}
+  }
+\end{center}
+\caption{A flow diagram for the SEAIR model 2.
+This is a constant population model, there are no births/deaths.
+Vaccinations are assumed to only be given to susceptible individuals, and vaccine immunity wanes only with susceptible vaccinated individuals.}\label{fig:flow_diagram2}
+\end{figure}
+
+\newpage
+
+\subsection{Model~3}
+
+\begin{figure}[!h]
+%%%%% SEAIR diagram
+\begin{center}
+  \resizebox{\textwidth}{!}{
+    \Large
+    \setlength{\unitlength}{5pt}
+    \begin{picture}(100,85)(0,15)
+
+    % COMPARTMENTS
+    % \put(36, 47){\framebox(6, 6){$\mathrm{W}_{u}$}}
+
+    \put(39, 50){\circle{6}}
+    \put(37, 49){$\mathrm{W}_{u}$}
+
+    \put(8, 55){\framebox(6, 6){$\mathrm{S_{u0}}$}}
+    \put(8, 39){\framebox(6, 6){$\mathrm{S_{u\vaccCounter}}$}}
+    % \put(22, 45){\circle{2}}
+    % \put(22, 55){\circle{2}}
+    \put(36, 62.5){\framebox(6, 6){$\mathrm{A_{u}}$}}
+    \put(36, 31.5){\framebox(6, 6){$\mathrm{I_{u}}$}}
+    \put(56, 55){\framebox(6, 6){$\mathrm{R^1_{u0}}$}}
+    \put(71, 55){\framebox(6, 6){$\mathrm{R^2_{u0}}$}}
+    \put(86, 55){\framebox(6, 6){$\mathrm{R^3_{u0}}$}}
+    \put(56, 39){\framebox(6, 6){$\mathrm{R^1_{u\vaccCounter}}$}}
+    \put(71, 39){\framebox(6, 6){$\mathrm{R^2_{u\vaccCounter}}$}}
+    \put(86, 39){\framebox(6, 6){$\mathrm{R^3_{u\vaccCounter}}$}}
+    % \dottedline{3}(0,5)(70,5)
+
+    % INDIVIDUAL MOVEMENT
+    % \put(15, 58){\line(2.4, -1){5.5}}
+    % \put(15, 42){\line(2.4, 1){5.5}}
+
+    % \put(23, 46.3){\vector(1, 1.45){12}}
+    % \put(23, 53.7){\vector(1, -1.45){12}}
+
+    % \put(23.5, 55.7){\vector(1.1, 1){11}}
+    % \put(23.5, 44.3){\vector(1.1, -1){11}}
+
+    % Backward Diags, W_u
+    % \multiput(35, 51)(-1, .3){12}{\line(-1, .3){0.2}}
+    % \put(24, 54.4){\vector(-1, 0.3){0.5}}
+    % \multiput(35, 49)(-1, -0.3){12}{\line(-1, -0.3){0.15}}
+    % \put(24, 45.6){\vector(-1, -0.3){0.5}}
+
+    % S_u0 -> A_u
+    \put(15, 58.5){\vector(3, 1.2){20}}
+
+    % S_u0 -> I_u
+    \put(15, 57.5){\vector(1, -1.1){20}}
+
+    % S_uz -> I_u
+    \put(15, 41.5){\vector(3, -1.2){20}}
+
+    % S_uz -> A_u
+    \put(15, 42.5){\vector(1, 1.1){20}}
+
+    % I_u -> R_u0
+    \put(43, 66){\vector(2, -1){12}}
+
+    % A_u -> R_uz
+    \multiput(43, 64)(1.25, -1.6){12}{\line(1, -1.3){0.7}}
+    \put(57, 46){\vector(1, -1.3){0.2}}
+
+    \put(43, 36){\vector(1.7, 2){15}}
+
+    \multiput(60, 54)(0, -2){4}{\line(0, -1){1}}
+    \put(60, 46){\vector(0, -1){0.5}}
+
+    \multiput(74, 54)(0, -2){4}{\line(0, -1){1}}
+    \put(74, 46){\vector(0, -1){0.5}}
+
+    \multiput(89, 54)(0, -2){4}{\line(0, -1){1}}
+    \put(89, 46){\vector(0, -1){0.5}}
+
+    \put(63, 58){\vector(1, 0){7}}
+    \put(78, 58){\vector(1, 0){7}}
+
+    \put(63, 42){\vector(1, 0){7}}
+    \put(78, 42){\vector(1, 0){7}}
+
+    \multiput(39, 61.5)(0, -1){8}{\line(0, -1){0.3}}
+    \put(39,54){\vector(0, -1){0.5}}
+
+    \multiput(39, 38.5)(0, 1){8}{\line(0, 1){0.3}}
+    \put(39,46){\vector(0, 1){0.5}}
+
+    \cbezier(89,62)(80, 85)(20, 85)(12,62)
+    \put(12.2,62.5){\vector(-1, -3.5){0.3}}
+
+    \cbezier(89,38)(80, 15)(20, 15)(12,38)
+    \put(12.2,37.5){\vector(-1, 3.5){0.3}}
+
+    \multiput(11, 54)(0, -2){4}{\line(0, -1){1}}
+    \put(11, 46){\vector(0, -1){0.5}}
+
+    % DEMOGRAPHY
+    % -> S
+    \put(0, 58){\vector(1, 0){7}}
+    \put(1, 59){$\mathrm{\mu_{\demography S_{u0}}}$}
+
+    % S Death
+    \put(10, 62){\vector(0, 1){7}}
+    \put(8, 64){$\mathrm{\muDeath}$}
+    \put(10, 38){\vector(0, -1){7}}
+    \put(8, 34){$\mathrm{\muDeath}$}
+
+    % A + I death
+    \put(39, 69.5){\vector(0, 1){7}}
+    \put(37, 71.5){$\mathrm{\muDeath}$}
+
+    \put(39, 31.5){\vector(0, -1){7}}
+    \put(30, 27.5){$\mathrm{\muDeath + \choleraDeath}$}
+
+    % R Death
+    \put(59, 62){\vector(0, 1){7}}
+    \put(57, 64){$\mathrm{\muDeath}$}
+
+    \put(74, 62){\vector(0, 1){7}}
+    \put(72, 64){$\mathrm{\muDeath}$}
+
+    \put(59, 38){\vector(0, -1){7}}
+    \put(57, 34){$\mathrm{\muDeath}$}
+
+    \put(74, 38){\vector(0, -1){7}}
+    \put(72, 34){$\mathrm{\muDeath}$}
+
+    \put(93, 42){\vector(1, 0){7}}
+    \put(95.5, 43){$\mathrm{\muDeath}$}
+
+    \put(93, 58){\vector(1, 0){7}}
+    \put(95.5, 59){$\mathrm{\muDeath}$}
+
+    % W death
+    % \put(43, 50){\vector(1, 0){7}}
+    \multiput(43, 50)(1, 0){7}{\line(1, 0){0.2}}
+    \put(50.2, 50){\vector(1, 0){0.5}}
+    \put(44, 51){$\mathrm{\Wremoval}$}
+    % \qbezier[12](10, 20)(20, 30)(25, 20)
+
+    \end{picture}
+  }
+\end{center}
+\caption{A flow diagram for the SAIR model 3.
+This model assumes a constant population while also including a mechanism for births/deaths; all deaths are balanced by births into the unvaccinated susceptible compartment, so the birth rate $\mu_{\demography S_{u0}}$ corresponds to the sum total deaths from the remaining compartments.
+The model assumes that symptomatic individuals will not be vaccinated, hence no vaccination arrow exiting the $I_u0$ compartment.}\label{fig:flow_diagram3}
+\end{figure}
 
@@
 This choice of $t_0$ corresponds to using approximately $1\%$ of the available data to determine initial values of the latent states.
-In addition to modeling a larger portion of the available data, this choice of $t_0$ corresponds to an important real-world event, as daily reports from each of the departments were not being sent to the health ministry until November 10, 2010; % TODO: Cite
+In addition to modeling a larger portion of the available data, this choice of $t_0$ corresponds to an important real-world event, as daily reports from each of the departments were not being sent to the health ministry until November 10, 2010 \citep{barzilay13};
 this choice of $t_0$ therefore makes $\bm{Y}\big(t_1\big)$ the first week of data once daily reports are being sent to the health ministry.
@@
 The resulting parameter estimates are used to obtain the remaining non-zero initial values of the latent states using Eqs.~\myeqref{eq:model3:A0}--\myeqref{eq:model3:B0}.
-
-% Argue why we didn't need to do this for Model 2.
-
-% TODO: Check this section.
 
@@
 colnames(mod3_cases) <- deps
-m3_inits <- c(7362, 1914, 22, 8, 2818, 82, 2499, 6953, 44, 7) # TODO: Also fit by hand, but could be changed to initializing to the observation before.
+m3_inits <- c(7362, 1914, 22, 8, 2818, 82, 2499, 6953, 44, 7)  # fit by hand
 names(m3_inits) <- deps
@@
 
-In this section, we provide more detail on the process that was used to estimate the coefficients used in Model~3.
+In this section, we provide more detail on the process that was used to estimate the coefficients of Model~3.
 In particular, we discuss why we decided to include additional model parameters---those that are associated with the behavior of the system during Hurricane Matthew---that were not considered by \cite{lee20}.
 To calibrate this model, we used the iterated block particle filter (IBPF) method of \cite{ionides22}.
-Due to the novelty of this algorithm, there does not exist published examples the IBPF algorithm used for data analysis outside of the papers in which the algorithm was introduced \citep{ning21ibpf,ionides22}, which at least partially motivates the inclusion additional details in this section.
+Due to the novelty of this algorithm, there does not exist published examples the IBPF algorithm used for data analysis outside of the papers in which the algorithm was introduced \citep{ning21ibpf,ionides22}, which is one motivation of the inclusion additional details related to fitting and diagnosing the model fit provided here.
 
@@
 The first collection of searches was performed by obtaining initial values for the parameters by uniformly sampling values from a predefined hypercube.
-A subsequent refinement search used starting values corresponding to the largest model likelihoods obtained in the previous round.
-This is similar to the approach used by \cite{ionides22}.
+A subsequent refinement search used parameter values corresponding to the largest model likelihoods as starting parameter values.
+The need for multiple searches does not appear to be uncommon, as a similar approach was used by \cite{ionides22}.
 
 We use the technique described above to fit the fully coupled version of Model~3 proposed by \cite{lee20}.
-The maximum likelihood we obtained with the model was $\Sexpr{round(max(haiti3_sub_fit$search2$logLiks$logLik))}$, which is higher than the benchmark model ($\Sexpr{round(sum(unitLikes_bench$ll))}$).
+The maximum likelihood we obtained after two rounds of searching was $\Sexpr{round(max(haiti3_sub_fit$search2$logLiks$logLik))}$, which is higher than the benchmark model ($\Sexpr{round(sum(unitLikes_bench$ll))}$).
 While beating a simple associative benchmark is promising, this does not immediately rule out the possibility that the model is a poor description of the system.
@@
 For meta-population models, it is worth considering how well the model fits the data to each spatial unit.
-This can be done by looking at conditional log-likelihoods, which is part of the output of the \texttt{bpfilter} algorithm in the \texttt{spatPomp} package.
+% This can be done by looking at conditional log-likelihoods, which is part of the output of the \texttt{bpfilter} algorithm in the \texttt{spatPomp} package.
 The likelihoods for each department, compared to the corresponding benchmark model, are displayed in Fig.~\ref{fig:h3UnitLikes}.
@@
 
-\begin{figure}[ht]
+\begin{figure}[!h]
 <<h3UnitLikes, fig.height=3, fig.width=5.5, echo=FALSE>>=
@@
 @
-\caption{\label{fig:h3UnitLikes}Log-likelihoods of Model~3 for each department compared to the corresponding benchmark model prior to the inclusion of parameters for initial values and Hurricane Matthew.}
+\caption{\label{fig:h3UnitLikes}Log-likelihoods of Model~3 for each department compared to the corresponding benchmark model prior to the inclusion of parameters that account for Hurricane Matthew.}
 \end{figure}
@@
 In addition to considering the conditional log-likelihoods of each unit, one can consider conditional log-likelihoods of each observation.
-When compared to a benchmark, this level of detail can provide useful information about which observations are well described by the model and which are poorly described.
-In Fig.~\ref{fig:condLL}, we plot the conditional log-likelihoods of each observation.
-This figure reveals that the fitted model has difficult time explaining certain features of the data.
-For example, many departments (in particular Sud) have observations with lower conditional log-likelihoods near October 2016 than at other time points.
-Further investigation of the model output reveals that the model is struggling to explain the sudden surge in cholera cases that occurred at this time, which coincides with the time that Hurricane Matthew struck Haiti.
-While the model does include a mechanism to account for increased cholera transmission due to large rainfall events, the mechanism does not appear to be sufficient to capture the damaging effects of the hurricane, which had the greatest impact in the the Sud and Grand'Anse departments \citep{ferreirai16}.
-This result led us to include parameters $\Whur{u}$ and $\hHur{u}$ in Eq.~23 of the main text, which allows for an increase in the transmission rate between environmental cholera and humans for Sud and Grand'Anse during and after the hurricane.
-The effect of the hurricane on cholera transmission is assumed to have an exponential decay, where the magnitude is determined by $\Whur{u}$ and the duration of the effect determined by $\hHur{u}$.
-
-\begin{figure}[ht]
+When compared to a benchmark, this level of detail can provide useful information about which observations are well described by the model and which are not.
+In Fig.~\ref{fig:condLL}, we plot the conditional log-likelihoods of Model~3 for each observation.
+Typically it is most useful to compare the conditional log-likelihoods of the model under consideration to a benchmark, as plotting only conditional log-likelihoods without a comparison may not be helpful.
+In this case, however, the same insight can be drawn using a figure without a benchmark comparison, so we do not include the benchmark in order to avoid the issue of over-plotting.
+
+\begin{figure}[!h]
 <<CondLLFig, echo=FALSE, message=FALSE, fig.height=3>>=
@@
 @
-\caption{\label{fig:condLL}Conditional log-likelihoods of Model~3 prior to the inclusion of parameters for initial values and Hurricane Matthew.}
+\caption{\label{fig:condLL}Conditional log-likelihoods of Model~3 prior to the inclusion of the Hurricane Matthew related parameters.}
 \end{figure}
 
-Other points of concern include the observations at or near the start of the recorded cases, for example, in Nord-Ouest and Sud-Est.
-These low conditional log-likelihoods are a result of having zero or near zero confirmed cases in the first week of observed data, followed by a rapid growth of cases.
-The model cannot explain the rapid growth of cholera in the absence of starting cases because the initialization scheme forces departments with no recorded cases in the first week to have hidden states $I_{u0}(0) = A_{u0}(0) = B_{u}(0) = 0$.
-In reality, however, it is possible to have some infected or asymptomatic individuals in a department when no cases are recorded due to under-reporting.
-This may be particularly true at the onset of an infectious disease outbreak, as lower public awareness and inadequate testing infrastructure may result in many cases going undetected.
-We therefore include and estimate $I_{0u}$ that determines the initial number of Infected individuals at the start of the pandemic for the departments Grande'Anse, Nord-Ouest, and Sud-Est;
-each of these departments have both low starting case counts followed by rapid growth, and low initial values of the conditinoal log-likelihood.
+Fig.~\ref{fig:condLL} reveals that the fitted model poorly describes certain features of the data.
+For example, many departments (in particular Sud) have observations with lower conditional log-likelihoods near October 2016 than at other time points.
+Further investigation of the model output reveals that the model is struggling to explain the sudden surge in cholera cases that occurred at this time, which coincides with the time that Hurricane Matthew struck Haiti.
+While the model does include a mechanism to account for increased cholera transmission due to large rainfall events, the mechanism does not appear to be sufficient to capture the damaging effects of the hurricane, which had the greatest impact in the the Sud and Grand'Anse departments \citep{ferreirai16}.
+This result led us to include parameters $\Whur{u}$ and $\hHur{u}$ in Eq.~23 of the main text, which allows for an increase in the transmission rate between environmental cholera and humans for in Sud and Grand'Anse during and after the hurricane.
+The effect of the hurricane on cholera transmission is assumed to have an exponential decay, where the magnitude is determined by $\Whur{u}$ and the duration of the effect determined by $\hHur{u}$.
 
@@
   mutate(unit = paste0(unit, ": ", deps[as.numeric(unit)]))
-@
-
-We refit Model~3 after introducing these seven parameters---three initial value parameters and four parameters related to Hurricane Matthew---using the same successive search scheme described above.
+
+bench_diffs <- dplyr::left_join(
+  x = unitLikesM3,
+  y = unitLikes_bench,
+  by = "unit"
+) %>%
+  rename(ll_mod3 = ll_est, ll_bench = ll) %>%
+  mutate(unit = unlist(myLabeller(unit)),
+         ll_diff = ll_mod3 - ll_bench) %>%
+  arrange(ll_diff)
+
+floor(100 * sum(mod3_cases[, c(bench_diffs$unit[1:4])], na.rm = TRUE) / sum(mod3_cases, na.rm = TRUE))
+@
+
+We refit Model~3 after introducing these hurricane-related parameters using the same successive search scheme described above.
 The resulting model has a log-likelihood value of $\Sexpr{round(max(h3_final_results$search2$logLiks$logLik), 1)}$.
-We also provide figures for department specific log-likelihoods and conditional log-likelihoods in Figs.~\ref{fig:finalUnitLL} and \ref{fig:finalCondLL}.
-
-\begin{figure}[ht]
+% We also provide figures for department specific log-likelihoods and conditional log-likelihoods in Figs.~\ref{fig:finalUnitLL} and \ref{fig:finalCondLL}.
+The addition of the Hurricane parameters seems to have resulted in a dramatic increase in conditional likelihoods around and after October 2016, as seen in Fig.~\ref{fig:finalCondLL}.
+The inclusion of these parameters resulted in an overall increase of $\Sexpr{round(max(h3_final_results$search2$logLiks$logLik) - max(haiti3_sub_fit$search2$logLiks$logLik), 1)}$ log-likelihood units, which is a highly statistically significant difference.
+
+<<finalModel3BPF, echo=FALSE, message=FALSE, include=FALSE>>=
+
+# Set the model parameters to the best results from previous search
+coef(h3)[names(best_params)] <- best_params
+
+# block particle filter to get conditional log-likelihoods.
+bpf_final3_out <- bake(
+  file = "../model3/si/h3_final_bpf_cond_ll.rds", {
+
+    results <- list()
+
+    bpf_out <- h3 %>% bpfilter(
+      Np = 5000,
+      block_size = 1,
+      save_states = TRUE
+    )
+
+    results$Sus_df <- foreach(i=1:length(h3@times), .combine = bind_rows) %do% {
+      get_h3_S_quants(bpf_out, i)
+    }
+
+    results$block.cond.loglik <- bpf_out@block.cond.loglik
+    results$times <- bpf_out@times  # saved for convenience
+    results
+  },
+  seed = 786321
+)
+@
+
+\begin{figure}[!h]
+<<finalCondLLFig, echo=FALSE, message=FALSE, fig.height=3>>=
+t(bpf_final3_out$block.cond.loglik) %>%
+  as.data.frame() %>%
+  mutate(t = bpf_final3_out$times) %>%
+  pivot_longer(
+    cols = -t,
+    values_to = 'likelihood',
+    names_prefix = "V",
+    names_to = 'Unit'
+  ) %>%
+  mutate(date = as.Date(lubridate::date_decimal(t))) %>%
+  ggplot(aes(x = date, y = likelihood)) +
+  geom_line() +
+  geom_point() +
+  facet_wrap(~Unit, labeller = myLabeller, scales = 'free_y',
+             nrow = 2) +
+  theme_bw() +
+  theme(axis.title.x = element_blank(),
+        axis.text.x = element_text(angle = -45, hjust = 0)) +
+  ylab("Conditional log-likelihoods")
+@
+\caption{\label{fig:finalCondLL}Conditional log-likelihoods of Model~3 after adding and estimating the parameters related to Hurricane Matthew.}
+\end{figure}
+
+Now that the model with additional parameters has been calibrated to the incidence data, we plot the conditional log-likelihood of each department compared to a benchmark model in Fig.~\ref{fig:finalUnitLL}.
+Interestingly, the difference in log-likelihoods between Model~3 and the benchmark model is smallest in the departments \Sexpr{bench_diffs$unit[1]}, \Sexpr{bench_diffs$unit[2]}, \Sexpr{bench_diffs$unit[3]} and \Sexpr{bench_diffs$unit[4]}.
+Each of these departments also exhibited the most sustained cholera transmission, defined by having the fewest number of weeks with no recorded cholera cases.
+Specifically, these four departments have zero cholera cases recorded in less than $4\%$ of the available data, and all remaining departments---except for Nord-Ouest, which has approximately $9.5\%$ of cases that are zeros and also exhibits the next smallest difference in log-likelihoods---have zero cases recorded in at least $14\%$ of the available weekly data.
+This result suggests that the quantitative advantage Model~3 has over its respective benchmark is primarily due to the model's ability to describe a resurgence of cases after a department records a week of zero cholera cases.
+This result may be unsurprising in the context of the models that we are comparing.
+Because the cholera transmission in individual departments likely depends on the national prevalence of cholera and the Vibrio cholerae bacteria, our benchmark model that relies exclusively on the previous number of case within any given unit has a difficult time predicting a resurgence of cases.
+
+\begin{figure}[!h]
 <<finalUnitLLFig, fig.height=3, fig.width=5.5, echo=FALSE>>=
@@
 @
-\caption{\label{fig:finalUnitLL}Log-likelihoods of Model~3 for each department compared to the corresponding benchmark model after adding and estimating parameters related to initial values and Hurricane Matthew.}
+\caption{\label{fig:finalUnitLL}Log-likelihoods of Model~3 for each department compared to the corresponding benchmark model after adding and estimating parameters related to Hurricane Matthew.}
 \end{figure}
 
-<<finalModel3BPF, echo=FALSE, message=FALSE, include=FALSE>>=
-
-# Set the model parameters to the best results from previous search
-coef(h3)[names(best_params)] <- best_params
-
-# block particle filter to get conditional log-likelihoods.
-bpf_final3_out <- bake(
-  file = "../model3/si/h3_final_bpf_cond_ll.rds", {
-
-    results <- list()
-
-    bpf_out <- h3 %>% bpfilter(
-      Np = 5000,
-      block_size = 1,
-      save_states = TRUE
-    )
-
-    results$Sus_df <- foreach(i=1:length(h3@times), .combine = bind_rows) %do% {
-      get_h3_S_quants(bpf_out, i)
-    }
-
-    results$block.cond.loglik <- bpf_out@block.cond.loglik
-    results$times <- bpf_out@times  # saved for convenience
-    results
-  },
-  seed = 786321
-)
-@
-
-\begin{figure}[ht]
-<<finalCondLLFig, echo=FALSE, message=FALSE, fig.height=3>>=
-t(bpf_final3_out$block.cond.loglik) %>%
-  as.data.frame() %>%
-  mutate(t = bpf_final3_out$times) %>%
-  pivot_longer(
-    cols = -t,
-    values_to = 'likelihood',
-    names_prefix = "V",
-    names_to = 'Unit'
-  ) %>%
-  mutate(date = as.Date(lubridate::date_decimal(t))) %>%
-  ggplot(aes(x = date, y = likelihood)) +
-  geom_line() +
-  geom_point() +
-  facet_wrap(~Unit, labeller = myLabeller, scales = 'free_y',
-             nrow = 2) +
-  theme_bw() +
-  theme(axis.title.x = element_blank(),
-        axis.text.x = element_text(angle = -45, hjust = 0)) +
-  ylab("Conditional log-likelihoods")
-@
-\caption{\label{fig:finalCondLL}Conditional log-likelihoods of Model~3 after adding and estimating the parameters for initial values and Hurricane Matthew.}
-\end{figure}
-
-While the addition of the Hurricane parameters seems to have resulted in a dramatic increase of conditional likelihoods around and after October 2016, the inclusion of the initial value parameters does not appear to have fixed the low conditional log-likelihood values in Sud-Est and Nord-Ouest.
-This could be due to many factors, including the possibility that the coupling between units forces low initial value parameters in a given department in order for the model to fit the data in the remaining departments.
-Whatever the case may be, the inclusion of the seven parameters resulted in a $\Sexpr{round(max(h3_final_results$search2$logLiks$logLik) - max(haiti3_sub_fit$search2$logLiks$logLik), 1)}$ increase in log-likelihood, which is a highly statistically significant difference.
-One could similarly test whether the inclusion of the initial value parameters resulted in a significantly better fit to the observed data, but we do not do this here due to the large computational effort that would be required.
+The difference in log-likelihoods between Model~3 and its benchmark model for each individual units suggests that Model~3 has a relatively poor fit for the four units with the most sustained cholera transmission.
+The simple four parameter benchmark has a higher likelihood than Model~3 for the \Sexpr{bench_diffs$unit[1]} and \Sexpr{bench_diffs$unit[2]} departments, and also has log-likelihoods only a few units smaller than Model~3 for units \Sexpr{bench_diffs$unit[3]} and \Sexpr{bench_diffs$unit[4]}.
+This is particularly worrisome given that these four departments account for more than $\Sexpr{floor(100 * sum(mod3_cases[, c(bench_diffs$unit[1:4])], na.rm = TRUE) / sum(mod3_cases, na.rm = TRUE))}\%$ of the total number of reported cholera cases.
 
@@
 We demonstrate this principle by examining the latent state of the calibrated model.
-In particular, we examine the compartment of susceptible individuals under various scenarios, as this compartment drives transmission dynamics in SIR models for infectious disease outbreaks.
+In particular, we examine the compartment of susceptible individuals under various scenarios.
 
@@
 
-\begin{figure}[ht]
+\begin{figure}[!h]
 <<h3SusFig, echo=FALSE, message=FALSE, fig.height=3>>=
@@
 In this article we claimed that we were able to obtain better fits to the observed data using the same models that were proposed by \cite{lee20}.
-Along with visual comparisons to the data, this claim was supported by comparing likelihoods and AIC values in Table~2 in the manuscript. % TODO: Double check the table number matches
+Along with visual comparisons to the data, this claim was supported by comparing likelihoods and AIC values in Table~2 in the manuscript.
 Because model likelihoods were not provided by \cite{lee20}, it is necessary to replicate these models in order to obtain likelihood estimates.
@@
 
-\begin{figure}[ht]
+\begin{figure}[!h]
 <<plotModel2Rep, cache=TRUE, echo=FALSE, fig.width=6.5, fig.height=4.5>>=
@@
 
-\begin{figure}[ht]
+\begin{figure}[!h]
 <<PlotMod3Rep, echo=FALSE, message=FALSE, fig.height=3.75>>=

### 2023-09-01 si/calibrateMod3.Rnw
Started work to make individual si sections seperate

--- 2023-08-31 si/si.Rnw
@@
 
-\pdfminorversion=4
-
-\newcommand\secTitleSpace{\hspace{3mm}}
+\input{inputs/header}
+
+\setcounter{tocdepth}{1}
 
@@
 library(doRNG)
-
-cores <-  as.numeric(Sys.getenv('SLURM_NTASKS_PER_NODE', unset = NA))
-if(is.na(cores)) cores <- detectCores()
-# cores <- 20
-registerDoParallel(cores)
-theme_set(theme_bw())
-
-if (!dir.exists("../model3/si")) {
-  dir.create("../model3/si", recursive = TRUE)
-}
-@
-
-<<set-opts, include=FALSE>>=
+library(knitr)
+
 options(
@@
 )
-@
-
-<<knitr-opts, include=FALSE, purl=FALSE>>=
-library(knitr)
-opts_knit$set(concordance=TRUE)
+
 opts_chunk$set(
-  progress=TRUE,prompt=TRUE,highlight=FALSE,
-  tidy=TRUE,
   tidy.opts=list(
@@
   comment="",
-  warning=FALSE,
-  message=FALSE,
-  error=TRUE,
+  fig.path = 'figure/',
+  # warning=FALSE,
+  # message=FALSE,
   echo=FALSE,
-  cache=FALSE,
-  strip.white=TRUE,
-  results="markup",
-  fig.path="figure/",
-  fig.lp="fig:",
   fig.align="center",
-  fig.show="asis",
   dev="pdf",
@@
 
-myround <- function (x, digits = 1) {
-  # taken from the broman package
-  if (digits < 1)
-    stop("This is intended for the case digits >= 1.")
-  if (length(digits) > 1) {
-    digits <- digits[1]
-    warning("Using only digits[1]")
-  }
-  tmp <- sprintf(paste("%.", digits, "f", sep = ""), x)
-  zero <- paste0("0.", paste(rep("0", digits), collapse = ""))
-  tmp[tmp == paste0("-", zero)] <- zero
-  tmp
-}
-
-@
-
-%% copied from pnas-new.cls
-\usepackage{amsmath,amsfonts,amssymb,graphicx}
-\usepackage{pict2e}
-\usepackage{curve2e}
-
-%%%% parameters %%%%%%%%%%5
-\newcommand\Wsat{W_{\mathrm{sat}}}
-\newcommand\muIR{\mu_{IR}}
-\newcommand\muEI{\mu_{EI}}
-\newcommand\transmission{\beta}
-\newcommand\seasAmplitude{a}
-\newcommand\rainfallExponent{r}
-\newcommand\muRS{\mu_{RS}}
-\newcommand\vaccineEfficacy{\theta}
-\newcommand\muBirth{\mu_S}
-\newcommand\muDeath{\delta}
-\newcommand\choleraDeath{\delta_{C}}
-\newcommand\symptomFrac{f}
-\newcommand\asymptomRelativeInfect{\epsilon}
-\newcommand\asymptomRelativeShed{\epsilon_{W}}
-\newcommand\Wbeta[1]{\beta_{W#1}}
-\newcommand\Wremoval{\delta_W}
-\newcommand\Wshed{\mu_W}
-\newcommand\mixExponent{\nu}
-\newcommand\sigmaProc{\sigma_{\mathrm{proc}}}
-\newcommand\reportRate{\rho}
-\newcommand\obsOverdispersion{\psi}
-\newcommand\phaseParm{\phi}
-\newcommand\transmissionTrend{\zeta}
-\newcommand\Binit{\xi}
-\newcommand\vaccClass{Z}
-\newcommand\vaccCounter{z}
-\newcommand\modelCounter{m}
-\newcommand\Whur[1]{\beta_{W#1}^{hm}}
-\newcommand\hHur[1]{h_{#1}^{hm}}
-\newcommand\Iinit{I_{0,0}}
-\newcommand\Einit{E_{0, 0}}
-\newcommand\reportChange{\delta}
-\newcommand\NBintercept{\alpha}
-\newcommand\NBsize{\varphi}
-
-\newcommand\missing{\hspace{6mm} ---}
-\newcommand\fixed{\color{blue}}
-\newcommand\demography{\bullet}
-
-\newcommand\mytitle{Informing policy via dynamic models: Cholera in Haiti}
-\newcommand\comma{{\hspace{-0.25mm},\hspace{-0.2mm}}}
-
-%%%%%%%%%%%%%%%% START OF USER DEFINITIONS %%%%%%%%%%%%%%%5
-
-\newcommand\siOnly[1]{} %% redefined for the supplement
-\newcommand\msOnly[1]{#1} %% redefined for the supplement
-
-\newcommand\code[1]{\texttt{#1}}
-\usepackage{xurl}
-% \usepackage[breaklinks]{hyperref}
-\usepackage{makecell}
-\usepackage{multirow}
-\usepackage{multicol}
-\usepackage[ruled,noline,linesnumbered]{algorithm2e}
-% \usepackage{setspace}
-
-%% EDITING MACROS
-\usepackage{color}
-\usepackage[section]{placeins}
-\usepackage[normalem]{ulem}% to use \sout in feedback commands
-\usepackage{bm}
-% orange for EI
-\definecolor{orange}{rgb}{1,0.5,0}
-\newcommand\ei[2]{\sout{#1} \textcolor{orange}{#2}}
-\newcommand\eic[1]{\textcolor{orange}{[#1]}}
-% green for JW
-\definecolor{green}{rgb}{0,0.5,0}
-\newcommand\jw[2]{\sout{#1} \textcolor{green}{#2}}
-\newcommand\jwc[1]{\textcolor{green}{[#1]}}
-% purple for JJ
-\definecolor{purple}{rgb}{0.5,0,1}
-\newcommand\jj[2]{\sout{#1} \textcolor{purple}{#2}}
-\newcommand\jjc[1]{\textcolor{purple}{[#1]}}
-% cyan for AR
-\definecolor{cyan}{rgb}{0,.5,.5}
-\newcommand\ar[2]{\sout{#1} \textcolor{cyan}{#2}}
-% \newcommand\arc[1]{\textcolor{cyan}{#1}}
-% light brown for KT
-\definecolor{lightbrown}{rgb}{0.5,0.5,0}
-\newcommand\kt[2]{\sout{#1} \textcolor{lightbrown}{#2}}
-\newcommand\ktc[1]{\textcolor{lightbrown}{#1}}
-
-%% basic POMP definitions
-\newcommand\Xspace{{\mathbb X}}
-\newcommand\Yspace{{\mathbb Y}}
-\newcommand\Thetaspace{\R^{\Thetadim}}
-\newcommand\Xunitspace{{\scriptsize{\mathcal X}}}
-\newcommand\Yunitspace{{\tiny{\mathcal Y}}}
-\newcommand\hatTheta{\widehat{\Theta}}
-\newcommand\Xdim{{\mathrm{dim}}(\Xspace)}
-\newcommand\Ydim{{\mathrm{dim}}(\Yspace)}
-\newcommand\Thetadim{P}
-\newcommand\thetadim{p}
-\newcommand\timeSet{{\mathbb T}}
-\newcommand\data[1]{#1^*}
-
-\newcommand\unitSpecific{\hspace{0.1mm}\mathrm{us}}
-\newcommand\shared{\hspace{0.15mm}\mathrm{sh}}
-
-% for algorithm pseudocode
-\SetKwFor{For}{for}{:}{end}
-
-%% for particle filters
-\newcommand\unit{u}
-\newcommand\altUnit{\tilde u}
-\newcommand\Unit{U}
-\newcommand\rep{i}
-\newcommand\Rep{\mathcal{I}}
-\newcommand\island{\rep}
-\newcommand\Island{\Rep}
-%\newcommand\txtisland{replicate}
-\newcommand\altIsland{\tilde i}
-\usepackage[mathscr]{euscript}
-
-\renewcommand\time{n}
-\newcommand\myvec[1]{\boldsymbol{#1}}
-\newcommand\altTime{\tilde n}
-\newcommand\Time{N}
-\newcommand\Np{J}
-\newcommand\np{j}
-\newcommand\altNp{k}
-\newcommand\altAltNp{\tilde j}
-\newcommand\resampleIndex{r}
-\newcommand\unitWeight{w}
-
-%% for all iterated filtering algorithms
-\newcommand\Nit{M}
-\newcommand\nit{m}
-
-%% customized math macros
-\newcommand\prob{\mathbb{P}}
-\newcommand\E{\mathbb{E}}
-\newcommand\dd[1]{\mathrm{d}{#1}}
-\newcommand\given{{\,\vert\,}}
-\newcommand\equals{{{\,}={\,}}}
-\newcommand\myequals{\hspace{0.5mm}{=}\hspace{0.5mm}}
-\newcommand\myto{{\;:\;}}
-\newcommand\seq[2]{{#1}\!:\!{#2}}
-\newcommand\mydot{{\,\cdot\,}}
-\newcommand\cp[2]{N_{\mathrm{#1}\mathrm{#2}}}
-\newcommand\giventh{{\hspace{0.5mm};\hspace{0.5mm}}}
-\newcommand\normal{\mathcal{N}}
-\newcommand\argequals{{\,=\,}}
-\newcommand\lags{c}
-\newcommand\maxlag{\overline{c}}
-\newcommand\nlfList{C}
-\newcommand\bigO{\mathcal{O}}
-\newcommand\loglik{\lambda}
-\newcommand\loglikMC{\MC{\loglik}}
-\newcommand\R{\mathbb{R}}
-\newcommand\param{\,;}
-\newcommand\mycolon{{\hspace{0.6mm}:\hspace{0.6mm}}}
-\newcommand\MC[1]{#1^{\,\mbox{\tiny MC}}}
-\newcommand\EMC{{\E}}
-\newcommand\Var{\mathrm{Var}}
-\newcommand\var{\Var}
-\newcommand\Cov{\mathrm{Cov}}
-\newcommand\cov{\Cov}
-\newcommand\iid{\mathrm{iid}}
-\newcommand\dist{d}
-\newcommand\transpose{\top}
-
-\newcommand\gnsep{,}
-
-\def\lik{L}
-\def\loglik{\ell}
-
-%% THEOREM-LIKE ENVIRONMENTS
-\usepackage{amsthm}
-\newtheorem{prop}{Proposition}
-\newtheorem{theorem}{Theorem}
-\newtheorem{lemma}{Lemma}
-\newtheorem{example}{Example}
-\newtheorem{remark}{Remark}
-\newtheorem{cor}{Corrolary}
-\newtheorem{fact}{Fact}
-\newtheorem{assumption}{Assumption}
-\newtheorem{assumptionB}{Assumption}
-\renewcommand\theassumption{A\arabic{assumption}}
-\renewcommand\theassumptionB{B\arabic{assumptionB}}
-
-\makeatletter
-\AtBeginDocument{%
-  \expandafter\renewcommand\expandafter\subsection\expandafter{%
-    \expandafter\@fb@secFB\subsection
-  }%
-}
-\makeatother
-
-%% FOR PSEUDOCODE SETUP
-\newcommand\mystretch{\rule[-2mm]{0mm}{5mm} }
-\newcommand\asp{\hspace{4mm}}
-\newcommand\codeIndent{\hspace{4mm}}
-
-%%%%%%%%%%%%%%%% END OF USER DEFINITIONSS %%%%%%%%%%%%%%%
-
-\bibliographystyle{apalike}
-\usepackage{natbib}
-\usepackage{fullpage}
-% \usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,justification=raggedright,singlelinecheck=off]{caption}
-% \usepackage[labelfont=bf,labelsep=period]{caption} % For some reason this doesn't work with .Rnw file, but it does with just the tex file, so after compiling you can edit this so that it works.
-\renewcommand{\figurename}{Fig}
-
-\newcommand\myeqref[1]{(\ref{#1})}
-
-\renewcommand{\contentsname}{Supplementary Content}
-\renewcommand{\refname}{Supplementary References}
-\renewcommand\thefigure{S-\arabic{figure}}
-\renewcommand\thetable{S-\arabic{table}}
-\renewcommand\thesection{S\arabic{section}}
-\renewcommand\theequation{S\arabic{equation}}
-\renewcommand\theprop{S\arabic{prop}}
-\renewcommand\thelemma{S\arabic{lemma}}
-\renewcommand\thealgocf{S\arabic{algocf}}
-
-\setcounter{tocdepth}{1}
+cores <-  as.numeric(Sys.getenv('SLURM_NTASKS_PER_NODE', unset = NA))
+if(is.na(cores)) cores <- detectCores()
+# cores <- 20
+registerDoParallel(cores)
+theme_set(theme_bw())
+
+if (!dir.exists("../model3/si")) {
+  dir.create("../model3/si", recursive = TRUE)
+}
+@
 
@@
 
-\date{\today}
-\title{Supplement to ``{\mytitle}''}
-  \author{Jesse Wheeler, AnnaElaine L. Rosengart, Zhuoxun Jiang, \\ Kevin Tan, Noah Treutle and Edward L. Ionides
- \\ \hspace{.2cm}\\
-    Department of Statistics, University of Michigan\\
-}
-
-\newcommand{\blind}{1}
-
-\if1\blind
-{
-\maketitle
-}
-\fi
-
-\if0\blind
-{
-  \bigskip
-  \bigskip
-  \bigskip
-  \begin{center}
-    {\LARGE\bf \mytitle}
-\end{center}
-  \bigskip
-  \bigskip
-}\fi
-
-\tableofcontents
-
-\newpage
-
-\section{Model Diagrams}
-
-Each of the dynamic models considered in this manuscript can be fully described using the model descriptions in the manuscript, coupled with the additional information described in Sections 2 and 3 of this supplement.
-Despite this, diagrams of dynamic systems are often helpful to understand the equations.
-In this section, we give three diagrams representing Models~1--3, respectively.
-Because the models are defined by their mathematical equations and numeric implementation, these diagrams are not unique visual representations of the model.
-Alternative representations that may be helpful in understanding the models explored in this paper were made by \citet{lee20sup}.
-
-\subsection{Model~1}
-
-\begin{figure}[!h]
-%%%%% SEAIR diagram
-\begin{center}
-  \resizebox{0.9\textwidth}{!}{
-    \Large
-    \setlength{\unitlength}{5pt}
-    \begin{picture}(100,95)(0,15)
-      \thicklines
-
-      % Unvaccinated Compartments
-      \put(10,63){\framebox(6,6){$\mathrm{S}_0$}}
-      \put(34,63){\framebox(6,6){$\mathrm{E}_0$}}
-      \put(58,71){\framebox(6,6){$\mathrm{A}_0$}}
-      \put(82,63){\framebox(6,6){$\mathrm{R}_0$}}
-      \put(58,55){\framebox(6,6){$\mathrm{I}_0$}}
-
-      % Births
-      \put(1, 66){\vector(1, 0){8}}
-      \put(4, 68){$\muBirth$}
-
-      % S_0 -> E_0
-      \put(17,66){\vector(1,0){16}}
-      \put(22,67){$\lambda(t)$}
-
-      % E_0 -> A_0
-      \put(41,67.5){\vector(2.1,1){16}}
-      % \put(37, 72){$\muEI\big(1-\symptomFrac_{\vaccCounter}(t)\big)$}
-
-      % E_0 -> I_0
-      \put(41,65.5){\vector(2.1,-1){16}}
-
-      % A_0 -> R_0
-      \put(65,74){\vector(2.1,-1){16}}
-
-      % I_0 -> R_0
-      \put(65,59){\vector(2.4,1){16}}
-
-      % R_0 -> S_0
-      \cbezier(84,70)(75, 95)(23,95)(14,70)
-      \put(14,70){\vector(-1, -2){0.3}}
-
-      % Unvaccinated death rates
-      \put(13, 70){\vector(0, 1){10}}
-      \put(11, 74){$\muDeath$}
-
-      \put(37, 70){\vector(0, 1){10}}
-      \put(35, 74){$\muDeath$}
-
-      \put(61, 78){\vector(0, 1){9}}
-      \put(59, 80){$\muDeath$}
-
-      \put(61, 62){\vector(1.5, 2){6}}
-      \put(61.5, 65.5){$\muDeath$}
-
-      \put(85, 70){\vector(0, 1){10}}
-      \put(83, 74){$\muDeath$}
-
-      % Vaccinated Compartments
-      \put(10,37){\framebox(6,6){$\mathrm{S}_\vaccCounter$}}
-      \put(34,37){\framebox(6,6){$\mathrm{E}_\vaccCounter$}}
-      \put(58,45){\framebox(6,6){$\mathrm{A}_\vaccCounter$}}
-      \put(82,37){\framebox(6,6){$\mathrm{R}_\vaccCounter$}}
-      \put(58,29){\framebox(6,6){$\mathrm{I}_\vaccCounter$}}
-
-      % S -> E
-      \put(17,40){\vector(1,0){16}}
-      \put(22,41){$\lambda(t)$}
-
-      % E -> A
-      \put(41,40.5){\vector(2.1,1){16}}
-
-      % E -> I
-      \put(41,39.5){\vector(2.1,-1){16}}
-
-      % A -> R
-      \put(65,48){\vector(2.1,-1){16}}
-
-      % I -> R
-      \put(65,32){\vector(2.1,1){16}}
-
-      % R -> S
-      \cbezier(84, 36)(75, 10)(23, 10)(14, 36)
-      \put(14,36){\vector(-1, 2){0.3}}
-
-      % Vaccinated death rates
-      \put(13, 36){\vector(0, -1){10}}
-      \put(11, 31){$\muDeath$}
-
-      \put(37, 36){\vector(0, -1){10}}
-      \put(35, 31){$\muDeath$}
-
-      \put(61, 28.5){\vector(0, -1){9}}
-      \put(59, 24){$\muDeath$}
-
-      \put(61, 44){\vector(1.5, -2){6}}
-      \put(62, 38){$\muDeath$}
-
-      \put(85, 36){\vector(0, -1){10}}
-      \put(83.2, 30.2){$\muDeath$}
-
-      % S_0 -> S_\vaccCounter
-      \multiput(13,62)(0,-2){9}{\line(0,-1){1}}
-      \put(13, 43.5){\vector(0, -1){0}}
-
-      % E_0 -> E_\vaccCounter
-      \multiput(37,62)(0,-2){9}{\line(0,-1){1}}
-      \put(37, 43.5){\vector(0, -1){0}}
-
-      % R_0 -> R_\vaccCounter
-      \multiput(85,62)(0,-2){9}{\line(0,-1){1}}
-      \put(85, 43.5){\vector(0, -1){0}}
-
-      % A_0 -> A_\vaccCounter
-      \multiput(57.5,72)(-1.9, 0){3}{\line(-1, 0){0.9}}
-      \put(52.2, 72){\line(-2.1, -1){1}}
-      \multiput(51,71)(0, -2){11}{\line(0, -1){1}}
-      \put(52.2, 49){\line(-2.1, 1){1}}
-      \multiput(54,49)(1.9, 0){2}{\line(-1, 0){0.9}}
-      \put(57, 49){\vector(1, 0){0}}
-
-      % I_0 -> I_\vaccCounter
-      \multiput(65, 57.5)(1.9, 0){3}{\line(1, 0){0.9}}
-      \put(70.6, 57.5){\line(2.1, -1){1}}
-      \multiput(72, 56.5)(0, -2){13}{\line(0, -1){1}}
-      \put(72, 30.9){\line(-2, -1.2){0.9}}
-      \multiput(65.6, 30.2)(1.9, 0){3}{\line(1, 0){0.9}}
-      \put(64.5, 30.2){\vector(-1, 0){0}}
-
-    \end{picture}
-  }
-\end{center}
-\caption{A flow diagram for the SEAIR model.}\label{fig:flow_diagram}
-\end{figure}
-
-\subsection{Model~2}
-
-\begin{figure}[!h]
-%%%%% SEAIR diagram
-\begin{center}
-  \resizebox{\textwidth}{!}{
-    \Large
-    \setlength{\unitlength}{5pt}
-    \begin{picture}(100,95)(0,15)
-      \thicklines
-
-      % Unvaccinated Compartments
-      \put(10,63){\framebox(6,6){$\mathrm{S}_{u0}$}}
-      \put(34,69){\framebox(6,6){$\mathrm{E}_{u0}$}}
-      \put(58,77){\framebox(6,6){$\mathrm{A}_{u0}$}}
-      \put(82,77){\framebox(6,6){$\mathrm{R^A}_{u0}$}}
-      \put(82,61){\framebox(6,6){$\mathrm{R^I}_{u0}$}}
-      \put(58,61){\framebox(6,6){$\mathrm{I}_{u0}$}}
-
-      % % S_0 -> E_0
-      \put(17,66){\vector(3,1.1){16}}
-      % \put(22,67){$\lambda(t)$}
-
-      % % E_0 -> A_0
-      \put(41,72.5){\vector(2.1,1){16}}
-      % % \put(37, 72){$\muEI\big(1-\symptomFrac_{\vaccCounter}(t)\big)$}
-
-      % % E_0 -> I_0
-      \put(41,71.5){\vector(2.1,-1){16}}
-
-      % % A_0 -> R_0
-      \put(65,80.5){\vector(1,0){16}}
-
-      % A_0 -> W_u
-      % \multiput(65, 79.8)(1, -1){8}{\line(1, -1){0.25}}
-
-      % I_0 -> W_u
-      % \multiput(65, 64.2)(1, 1){8}{\line(1, 1){0.25}}
-
-      % Long verticle dotted line
-      % \multiput(72.9, 72)(0, -1.41){13}{\line(0, -1){0.25}}
-
-      % Long backwards horizontal dotted line
-      % \multiput(73, 54)(-1.41, 0){22}{\line(0, -1){0.25}}
-      % \multiput(73, 34)(0, 1.41){15}{\line(0, 1){0.25}}
-      % \put(42, 53.9){\vector(-1, 0){0.2}}
-
-      % A_uz -> W_u
-      \put(58,39){\framebox(6,6){$\mathrm{A}_{u\vaccCounter}$}}
-      \qbezier[25](60.5, 46)(51, 52)(42, 53)
-      \put(42, 53){\vector(-1, 0.1){0.5}}
-
-      % A_uz -> W_u
-      \qbezier[25](60.5, 60)(51, 54)(42, 54)
-      \put(42, 54){\vector(-1, -0.1){0.5}}
-
-      % I_uz -> W_u
-      \qbezier[30](60.5,30)(48, 45)(40, 50)
-      \put(40, 50){\vector(-1, 0.8){0.5}}
-
-      % A_u0 -> W_u
-      \qbezier[29](60.5,76)(48, 61)(40.6, 57)
-      \put(40.8, 57.1){\vector(-1, -0.7){0.5}}
-
-      % A + I_\vaccCounter -> W_u
-      % \multiput(65, 41.8)(1, -1){8}{\line(1, -1){0.25}}
-      % \multiput(65, 26.2)(1, 1){8}{\line(1, 1){0.25}}
-
-      % W_u -> backwards diags.
-      % \multiput(33, 57)(-1, 1.2){10}{\line(-1, 1.2){0.1}}
-      % \put(24.5, 67){\vector(-1, 1.2){0.5}}
-      % \multiput(33, 51)(-1, -1.2){10}{\line(-1, -1.2){0.1}}
-      % \put(24, 40){\vector(-1, -1.2){0.5}}
-      %\put(65, 80){\vector(2, -1){8}}
-
-      % % I_0 -> R_0
-      \put(65,63.5){\vector(1,0){16}}
-
-      % % Vaccinated Compartments
-      \put(10,37){\framebox(6,6){$\mathrm{S}_{u\vaccCounter}$}}
-      \put(34,31){\framebox(6,6){$\mathrm{E}_{u\vaccCounter}$}}
-      \put(58,39){\framebox(6,6){$\mathrm{A}_{u\vaccCounter}$}}
-      \put(82,39){\framebox(6,6){$\mathrm{R^A}_{u\vaccCounter}$}}
-      \put(58,23){\framebox(6,6){$\mathrm{I}_{u\vaccCounter}$}}
-      \put(82,23){\framebox(6,6){$\mathrm{R^I}_{u\vaccCounter}$}}
-
-      % R_0 -> S_0
-      \cbezier(85,22)(80, 8)(18,8)(12,36)
-      \put(12.1,35.5){\vector(-1, 3.5){0.3}}
-
-      \cbezier(85,38)(80, 12)(18,15)(14,36)
-      \put(14.1,35.5){\vector(-1, 3.5){0.3}}
-
-      % R_0 -> S_0
-      \cbezier(85,84)(80, 100)(18,100)(12,70)
-      \put(12.1,70.5){\vector(-1, -3.5){0.3}}
-
-      \cbezier(85,68)(80, 95)(18, 92)(14,70)
-      \put(14.1,70.5){\vector(-1, -3.5){0.3}}
-
-      % % S -> E
-      \put(17,40){\vector(3,-1.1){16}}
-      % \put(22,41){$\lambda(t)$}
-
-      % % E -> A
-      \put(41,34.5){\vector(2.1,1){16}}
-
-      % % E -> I
-      \put(41,33.5){\vector(2.1,-1){16}}
-
-      % % A -> R
-      \put(65,42.5){\vector(1,0){16}}
-
-      % % I -> R
-      \put(65,25.5){\vector(1,0){16}}
-
-      % S_0 -> S_\vaccCounter
-      \multiput(12,62)(0,-2){9}{\line(0,-1){1}}
-      \put(12, 44){\vector(0, -1){0}}
-
-      % S_0 -> S_\vaccCounter
-      % Wanning vaccine efficacy
-      \multiput(14,44)(0,2){9}{\line(0,1){1}}
-      \put(14, 62){\vector(0, 1){0}}
-
-      \put(37, 54){\circle{7}}
-      \put(35, 53){$\mathrm{W_u}$}
-
-      \qbezier[10](32.5, 54)(29.5, 54)(26.5, 54)
-      \put(26, 54){\vector(-1, 0){0.1}}
-      \put(28, 55){$\Wremoval$}
-
-    \end{picture}
-  }
-\end{center}
-\caption{A flow diagram for the SEAIR model 2.
-This is a constant population model, there are no births/deaths.
-Vaccinations are assumed to only be given to susceptible individuals, and vaccine immunity wanes only with susceptible vaccinated individuals.}\label{fig:flow_diagram2}
-\end{figure}
-
-\newpage
-
-\subsection{Model~3}
-
-\begin{figure}[!h]
-%%%%% SEAIR diagram
-\begin{center}
-  \resizebox{\textwidth}{!}{
-    \Large
-    \setlength{\unitlength}{5pt}
-    \begin{picture}(100,85)(0,15)
-
-    % COMPARTMENTS
-    % \put(36, 47){\framebox(6, 6){$\mathrm{W}_{u}$}}
-
-    \put(39, 50){\circle{6}}
-    \put(37, 49){$\mathrm{W}_{u}$}
-
-    \put(8, 55){\framebox(6, 6){$\mathrm{S_{u0}}$}}
-    \put(8, 39){\framebox(6, 6){$\mathrm{S_{u\vaccCounter}}$}}
-    % \put(22, 45){\circle{2}}
-    % \put(22, 55){\circle{2}}
-    \put(36, 62.5){\framebox(6, 6){$\mathrm{A_{u}}$}}
-    \put(36, 31.5){\framebox(6, 6){$\mathrm{I_{u}}$}}
-    \put(56, 55){\framebox(6, 6){$\mathrm{R^1_{u0}}$}}
-    \put(71, 55){\framebox(6, 6){$\mathrm{R^2_{u0}}$}}
-    \put(86, 55){\framebox(6, 6){$\mathrm{R^3_{u0}}$}}
-    \put(56, 39){\framebox(6, 6){$\mathrm{R^1_{u\vaccCounter}}$}}
-    \put(71, 39){\framebox(6, 6){$\mathrm{R^2_{u\vaccCounter}}$}}
-    \put(86, 39){\framebox(6, 6){$\mathrm{R^3_{u\vaccCounter}}$}}
-    % \dottedline{3}(0,5)(70,5)
-
-    % INDIVIDUAL MOVEMENT
-    % \put(15, 58){\line(2.4, -1){5.5}}
-    % \put(15, 42){\line(2.4, 1){5.5}}
-
-    % \put(23, 46.3){\vector(1, 1.45){12}}
-    % \put(23, 53.7){\vector(1, -1.45){12}}
-
-    % \put(23.5, 55.7){\vector(1.1, 1){11}}
-    % \put(23.5, 44.3){\vector(1.1, -1){11}}
-
-    % Backward Diags, W_u
-    % \multiput(35, 51)(-1, .3){12}{\line(-1, .3){0.2}}
-    % \put(24, 54.4){\vector(-1, 0.3){0.5}}
-    % \multiput(35, 49)(-1, -0.3){12}{\line(-1, -0.3){0.15}}
-    % \put(24, 45.6){\vector(-1, -0.3){0.5}}
-
-    % S_u0 -> A_u
-    \put(15, 58.5){\vector(3, 1.2){20}}
-
-    % S_u0 -> I_u
-    \put(15, 57.5){\vector(1, -1.1){20}}
-
-    % S_uz -> I_u
-    \put(15, 41.5){\vector(3, -1.2){20}}
-
-    % S_uz -> A_u
-    \put(15, 42.5){\vector(1, 1.1){20}}
-
-    % I_u -> R_u0
-    \put(43, 66){\vector(2, -1){12}}
-
-    % A_u -> R_uz
-    \multiput(43, 64)(1.25, -1.6){12}{\line(1, -1.3){0.7}}
-    \put(57, 46){\vector(1, -1.3){0.2}}
-
-    \put(43, 36){\vector(1.7, 2){15}}
-
-    \multiput(60, 54)(0, -2){4}{\line(0, -1){1}}
-    \put(60, 46){\vector(0, -1){0.5}}
-
-    \multiput(74, 54)(0, -2){4}{\line(0, -1){1}}
-    \put(74, 46){\vector(0, -1){0.5}}
-
-    \multiput(89, 54)(0, -2){4}{\line(0, -1){1}}
-    \put(89, 46){\vector(0, -1){0.5}}
-
-    \put(63, 58){\vector(1, 0){7}}
-    \put(78, 58){\vector(1, 0){7}}
-
-    \put(63, 42){\vector(1, 0){7}}
-    \put(78, 42){\vector(1, 0){7}}
-
-    \multiput(39, 61.5)(0, -1){8}{\line(0, -1){0.3}}
-    \put(39,54){\vector(0, -1){0.5}}
-
-    \multiput(39, 38.5)(0, 1){8}{\line(0, 1){0.3}}
-    \put(39,46){\vector(0, 1){0.5}}
-
-    \cbezier(89,62)(80, 85)(20, 85)(12,62)
-    \put(12.2,62.5){\vector(-1, -3.5){0.3}}
-
-    \cbezier(89,38)(80, 15)(20, 15)(12,38)
-    \put(12.2,37.5){\vector(-1, 3.5){0.3}}
-
-    \multiput(11, 54)(0, -2){4}{\line(0, -1){1}}
-    \put(11, 46){\vector(0, -1){0.5}}
-
-    % DEMOGRAPHY
-    % -> S
-    \put(0, 58){\vector(1, 0){7}}
-    \put(1, 59){$\mathrm{\mu_{\demography S_{u0}}}$}
-
-    % S Death
-    \put(10, 62){\vector(0, 1){7}}
-    \put(8, 64){$\mathrm{\muDeath}$}
-    \put(10, 38){\vector(0, -1){7}}
-    \put(8, 34){$\mathrm{\muDeath}$}
-
-    % A + I death
-    \put(39, 69.5){\vector(0, 1){7}}
-    \put(37, 71.5){$\mathrm{\muDeath}$}
-
-    \put(39, 31.5){\vector(0, -1){7}}
-    \put(30, 27.5){$\mathrm{\muDeath + \choleraDeath}$}
-
-    % R Death
-    \put(59, 62){\vector(0, 1){7}}
-    \put(57, 64){$\mathrm{\muDeath}$}
-
-    \put(74, 62){\vector(0, 1){7}}
-    \put(72, 64){$\mathrm{\muDeath}$}
-
-    \put(59, 38){\vector(0, -1){7}}
-    \put(57, 34){$\mathrm{\muDeath}$}
-
-    \put(74, 38){\vector(0, -1){7}}
-    \put(72, 34){$\mathrm{\muDeath}$}
-
-    \put(93, 42){\vector(1, 0){7}}
-    \put(95.5, 43){$\mathrm{\muDeath}$}
-
-    \put(93, 58){\vector(1, 0){7}}
-    \put(95.5, 59){$\mathrm{\muDeath}$}
-
-    % W death
-    % \put(43, 50){\vector(1, 0){7}}
-    \multiput(43, 50)(1, 0){7}{\line(1, 0){0.2}}
-    \put(50.2, 50){\vector(1, 0){0.5}}
-    \put(44, 51){$\mathrm{\Wremoval}$}
-    % \qbezier[12](10, 20)(20, 30)(25, 20)
-
-    \end{picture}
-  }
-\end{center}
-\caption{A flow diagram for the SAIR model 3.
-This model assumes a constant population while also including a mechanism for births/deaths; all deaths are balanced by births into the unvaccinated susceptible compartment, so the birth rate $\mu_{\demography S_{u0}}$ corresponds to the sum total deaths from the remaining compartments.
-The model assumes that symptomatic individuals will not be vaccinated, hence no vaccination arrow exiting the $I_u0$ compartment.}\label{fig:flow_diagram3}
-\end{figure}
-
-\section{Markov chain and differential equation interpretations of compartment flow rates}
-
-In Sections 2.1, 2.2 and 2.3 of the main article, we define compartment models in terms of their flow rates.
-For a discrete population model, these rates define a Markov chain.
-For a continuous and deterministic model, the rates define a system of ordinary differential equations.
-Here, we add additional details to clarify the mapping from a collection of rate functions to a fully specified process.
-Our treatment follows \citet{breto09}.
-
-A general compartment model is a vector-valued process $X(t)=(X_1(t),\dots,X_c(t))$ denoting the (integer or real-valued) counts in each of $c$ compartments, where $t$ is any continuous value in the interval  $\left[t_0, \infty\right)$ for some real valued starting time $t_0$.
-The compartments may also have names, but to set up general notation we simply refer to them by their numerical index.
-The basic characteristic of a compartment model is that $X(t)$ can be written in terms of the flows $N_{ij}(t)$ from $i$ to $j$.
-A flow into compartment $i$ from outside the system is denoted by $N_{\demography i}$, and a flow out of the system from compartment $i$ is denoted by $N_{i\demography}$.
-We call $\demography$ a source/sink compartment, though it is an irregular compartment since $X_{\demography}(t)$ is not defined.
-These flows are required to satisfy a ``conservation of mass'' identity:
-\begin{equation}
-X_i(t)=X_i(t_0)+N_{\demography i}(t) - N_{i\demography}(t) + \sum_{j\neq
-i}N_{ji}(t)-\sum_{j\neq i}N_{ij}(t). \label{eq:conservation}
-\end{equation}
-Each {\em flow} $N_{ij}(t)$ is associated with a {\em rate} function
-$\mu_{ij}=\mu_{ij}(t,X(t))$, where we include the possibility that $i$ or $j$ takes value $\demography$.
-
-There are different ways to use a collection of rate functions to build a fully specified model.
-We proceed to describe the ones we use in this paper: via a system of ordinary differential equations (Sec.~\ref{subsec:ode}), a simple Markov counting system (Sec.~\ref{subsec:smcs}), and an overdispersed Markov counting system (Sec.~\ref{subsec:odmcs}).
-Other representations include stochastic differential equations driven by Gaussian noise or Gamma noise \citep{bhadra11}.
-
-\subsection{Ordinary differential equation (ODE) interpretation}
-\label{subsec:ode}
-
-A basic deterministic specification is
-\begin{equation}
-\label{eq:ode1}
-dN_{ij}/dt = \mu_{ij}\big(t,X(t) \big) X_i(t), \hspace{3mm} i\in \seq{1}{c}, \hspace{3mm} j\in \seq{1}{c} \cup\{\demography\}, \hspace{3mm} i\neq j,
-\end{equation}
-where $\mu_{ij}\big(t,X(t)\big)$ is called a per-capita rate or a unit rate.
-Flows into the system require special treatment since $X_i(t)$ in \myeqref{eq:ode1} is not defined for $i=\demography$.
-Instead, we specify
-\begin{equation}
-\label{eq:ode2}
-dN_{\demography i}/dt = \mu_{\demography i}\big(t,X(t) \big).
-\end{equation}
-This is the the interpretation and implementation used for Model~2 in our study.
-
-\subsection{Simple Markov counting system interpretation}
-\label{subsec:smcs}
-A continuous time Markov chain can be specified via its infinitesimal transition probabilities.
-A basic approach to this is to define
-\begin{eqnarray}
-\label{eq:smcs1}
-\prob\big[ N_{ij}(t+\delta)-N_{ij}(t)=0 \given X(t)\big]
- &=& 1-\delta \mu_{ij}\big(t,X(t)\big)X_i(t) + o(\delta),
-\\
-\label{eq:smcs2}
-\prob\big[ N_{ij}(t+\delta)-N_{ij}(t)=1 \given X(t)\big]
- &=& \delta \mu_{ij}\big(t,X(t)\big)X_i(t) + o(\delta),
-\end{eqnarray}
-for $i\in \seq{1}{c}$ and $j\in\seq{1}{c}\cup\{\demography\}$ with $i\neq j$.
-As with the ODE case, we need special attention for flows into the system, and we define
-\begin{eqnarray}
-\label{eq:smcs3}
-\prob\big[ N_{\demography i}(t+\delta)-N_{\demography i}(t)=0 \given X(t)\big]
- &=& 1-\delta \mu_{\demography i}\big(t,X(t)\big) + o(\delta),
-\\
-\label{eq:smcs4}
-\prob\big[ N_{\demography i}(t+\delta)-N_{\demography i}(t)=1 \given X(t)\big]
- &=& \delta \mu_{\demography i}\big(t,X(t)\big) + o(\delta).
-\end{eqnarray}
-Together with the initial conditions $X(0)$, equations \myeqref{eq:smcs1}--\myeqref{eq:smcs4} define a Markov chain.
-Each flow is a simple counting process, meaning a non-decreasing integer-valued process that only has jumps of size one.
-We therefore call the Markov chain a simple Markov counting system (SMCS).
-The infinitesimal mean of every flow is equal to its infinitesimal variance \citep{breto11} and so an SMCS is called equidispersed.
-We note that the special case of Model~1 used by \cite{lee20} (with $\sigmaProc = 0$) is an SMCS.
-To permit more general mean-variance relationships for a Markov counting system, we must permit jumps of size greater than one.
-The utility of overdispersed models, where the infinitesimal variance of the flow exceeds the infinitesimal mean, has become widely recognized \citep{stocks20,he10}.
-
-\subsection{Overdispersed Markov counting system interpretation}
-\label{subsec:odmcs}
-
-Including white noise in the rate function enables the possibility of an overdispersed Markov counting system \citep{breto11,breto09,he10}.
-Since rates should be non-negative, Gaussian noise is not appropriate and gamma noise is a convenient option that has found various applications \citep{romero-severson15, subramanian20}.
-Specifically, we consider a model given by
-\begin{equation}
-\label{eq:odmcs1}
-\mu_{ij}\big(t,X(t)\big) = \bar\mu_{ij}\big(t,X(t)\big) \, d\Gamma_{ij}(t)/dt,
-\end{equation}
-where $\Gamma_{ij}(t)$ is a stochastic process having independent gamma distributed increments, with
-\begin{equation}
-\label{eq:odmcs2}
-\E\big[\Gamma_{ij}(t)\big] = t, \quad \var\big[\Gamma_{ij}(t)\big] = \sigma_{ij}^2 t.
-\end{equation}
-Formally interpreting the meaning of \myeqref{eq:odmcs1} is not trivial, and we do so by constructing a Markov process $X(t)$ as the limit of the Euler scheme descrived in Section~\ref{sec:numerics}, below.
-Therefore, the numerical scheme in Sec.~\ref{sec:numerics} can be taken as a definition of the meaning of \myeqref{eq:odmcs1}.
-The Markov chain defined by the limit of this Euler scheme as the step size decreases is an overdispersed Markov counting system, with the possibility of instantaneous jumps of size greater than one \citep{breto11}.
-
-\section{Numerical solutions to compartment models}
-\label{sec:numerics}
-
-Models may be fitted and their implications assessed via numerical solutions (i.e., simulations) from the model equations.
-All the analyses we consider have this simulation-based property, known as plug-and-play or equation-free or likelihood-free.
-The numerical solutions to the model are arguably of more direct scientific interest than the exact solutions to the postulated equations.
-For ODE models, numerical methods are well studied and a standard numerical solution package such as \code{deSolve} in \code{R} is adequate for many purposes.
-For SMCS and ODMCS models, exact schemes are feasible when the number of events is small, which may be the case for small populations.
-However, for applicability to larger populations, we use instead the following Euler scheme.
-Write $\delta$ for an Euler time step, and $\Delta N_{ij}$ for the numerical approximation to $N_{ij}(t+\delta)-N_{ij}(t)$ given $X(t)$.
-For each $i$ and $j$ in $\seq{1}{c} \cup \{\demography\}$ with $i \neq j$, we draw independent Gamma distributed noise increments with mean $\delta$ and variance $\sigma_{ij}^2 \delta$, denoted using a mean-variance parameterization of the gamma distribution as
-\begin{equation}
-\label{eq:numerics1}
-\Delta\Gamma_{ij} \sim \mathrm{gamma}(\delta, \sigma_{ij}^2 \delta).
-\end{equation}
-In the case of an SMCS model, $\sigma_{ij}=0$ for all $i$ and $j$, so we have $\Delta\Gamma_{ij}=\delta$.
-Then, for $i\neq \demography$ and $j\neq i$, and writing
-\begin{equation}
-\label{eq:numerics2}
-\mu_{ij}=\bar\mu_{ij}\big(t,X(t)\big) \Delta\Gamma_{ij} / \delta,
-\end{equation}
-we calculate transition probabilities
-\begin{eqnarray}
-\label{eq:numerics3}
-p_{ij} &=& \exp\left\{-\sum_{k\in 1:c \, \cup \{\demography\}} \mu_{ik} \, \delta \right\}
-\frac{\mu_{ij}}{\sum_{k\in 1:c\cup \{\demography\}} \mu_{ik}},
-\\
-\label{eq:numerics4}
-p_{ii} &=& 1 - \sum_{j\neq i} p_{ij}.
-\end{eqnarray}
-These probabilities correspond to competing hazards for every individual in compartment $i$ to transition to some compartment $j$, interpreting $j=i$ to mean that the individual remains in $i$.
-Then, $\big(\Delta N_{i1},\dots,\Delta N_{ic},\Delta N_{i\demography}\big)$ has the multinomial distribution where $X_i(t)$ individuals are allocated independently to $\seq{1}{c}\cup\{\demography\}$ with probabilities given by \eqref{eq:numerics3} and \eqref{eq:numerics4}.
-We use the \code{reulermultinom} function in the \code{pomp} package to draw from this multionomial distribution.
-
-Different treatments of demographic flows---such as birth, death, immigration and emigration---are possible.
-For the case $i=\demography$, the treatment used by Model~1 is to set
-\begin{equation}
-\label{eq:numerics6}
-\Delta N_{\demography j} \sim \mathrm{poisson}( \mu_{\demography j} \delta),
-\end{equation}
-an independent Poisson random variable with mean $\mu_{\demography j} \delta$.
-
-Models~2 and 3 used an alternative approach, balancing the total number of flows in and out of the compartment, i.e., $\sum_{i}N_{\demography i}(t) = \sum_{i}N_{i \demography}(t)$, in order to make the model consistent with the known total population.
-In this case, we formally model the death rate as a rate of returning to the susceptible class $S$, and use external transitions from $\demography$ into $S$ to describe only net population increase.
-
-\section{Measurement Models}
-
-Each POMP requires specification of a measurement model, which is a statistical description of how observations on the system are obtained.
-In general, we used the same measurement models that were reported by \cite{lee20} unless specifically noted in the following subsections.
-
-\subsection{Model~1}
-
-In this model, the advantage afforded by vaccination is an increased probability that an infection is asymptomatic.
-Therefore, under the assumptions of this model, all reported cases are assumed to be a fraction of individuals that transition from the exposed to the infected compartment, as noted in Eq.~\myeqref{model1:meas}:
-\begin{equation}
-  \label{model1:meas}
-  Y_{n} \mid \Delta N_{E_{\cdot}I_{\cdot}}(n) = \reportChange \sim \text{NB}\left(\reportRate \reportChange, \obsOverdispersion \right),
-\end{equation}
-where $Y_n$ is the reported cholera cases at time $t_n \in t_1:t_N$ and $\Delta N_{E_{\cdot}I_{\cdot}}(n)$ is the sum total of individuals across each vaccination compartment $\vaccCounter \in 1:\vaccClass$ who moved from compartment $E_\vaccCounter$ to $I_\vaccCounter$ since observation $t_{n-1}$.
-Here, $\text{NB}\left(\reportRate \reportChange, \obsOverdispersion \right)$ denotes a negative binomial distribution with mean $\reportRate \reportChange$ and variance $\reportRate \reportChange \big(1 + \frac{\reportRate \reportChange}{\obsOverdispersion}\big)$.
-
-\subsection{Model~2}\label{sec:mod2Meas}
-
-Model~2 was fit using reported case counts that were transformed using the natural logarithm.
-We fit Model~2 using the subplex algorithm in the \code{subplex} package, using
-a Gaussian measurement model (Eq.~\myeqref{model2:meas}) on the log transformed cases within each unit.
-The final loss function that is maximized is the product of the likelihoods of the individual units, or the sum of the log-likelihood of the individual units.
-The measurement model for individual units is given in Eq.~\myeqref{model2:meas}.
-\begin{equation}
-  \label{model2:meas}
-  \log\big(Y_{u, n} + 1\big) \mid \Delta N_{E_{u\cdot}I_{u\cdot}}(n) = \reportChange_u \sim \text{N}\left(\log \big(\reportRate \reportChange_u + 1\big), \obsOverdispersion^2 \right),
-\end{equation}
-where $\Delta N_{E_{u\cdot}I_{u\cdot}}(n)$ is the sum total of individuals across vaccination compartment $\vaccCounter \in 0:4$ within unit $u$ who moved from compartment $E_{u\vaccCounter}$ to $I_{u\vaccCounter}$ since observation $t_{n-1}$.
-Therefore, because the natural logarithm of observed case counts (plus one, to avoid taking the logarithm of zero) has a normal distribution, $Y_{u, n} + 1$ is assumed to follow a log-normal distribution with log-mean parameter $\log\big(\reportRate \Delta N_{E_{u\cdot}I_{u\cdot}}(n) + 1\big)$ and log-variance $\obsOverdispersion^2$.
-We note that fitting a model with this measurement model is equivalent to fitting using least squares, with $\log(Y_{u, n} + 1)$ as the response variable.
-
-This measurement model differs from that used by \citet{lee20}, who fit the model in two stages: epidemic and endemic phases.
-Although their text and supplement material do not explicitly describe the measurement model used, inspection of the code provided with their submission suggests a change in measurement model between the epidemic and endemic phases.
-% In the file \code{choleraEqsPublic.py}, \citet{lee20} create several functions, where each function represents a set of coupled differential equations that could potentially be used to model cholera incidence data.
-% Each function returns a vector (or \code{numpy} array) that represents the change in each state variable for a single time step, including the variable \code{dC}, which tracks the number of new infections and is used to obtain the reported case counts.
-% Following their comments and code, it appears that the function \code{choleraEqs10WithoutVaccinationNetwork} was used to describe the dynamics of the epidemic phase, and \code{choleraEqs11WithoutVaccinationNetwork} was used for the endemic stage (see Sec~\ref{sec:mod2rep} for more details).
-% Because their models were fit using least squares, the code in these functions suggest that the measurement model for the epidemic phase is
-The measurement model they used for the epidemic phase is
-\begin{equation}
-  \label{model2:measEpi}
-  Y_{u, n} \mid \Delta N_{E_{u\cdot}I_{u\cdot}}(n) = \reportChange_u \sim \text{N}\left(\reportRate \reportChange_u, \obsOverdispersion^2 \right),
-\end{equation}
-% which is similar to our measurement model, the primary difference being that the measurement model is applied to raw case counts rather than log-transformed case counts.
-The measurement model they used for the endemic phase modifies the epidemic model by counting both asymptomatically infected (A) and symptomatically infected (I) individuals in the case counts:
-\begin{equation}
-  \label{model2:measEnd}
-  Y_{u, n} \mid \Delta N_{E_{u\cdot}I_{u\cdot}}(n) = \reportChange_{u1}, \Delta N_{E_{u\cdot}A_{u\cdot}}(n) = \reportChange_{u2} \sim \text{N}\left(\reportRate \big(\reportChange_{u1} + \reportChange_{u2}\big), \obsOverdispersion^2 \right),
-\end{equation}
-where the notation for $\Delta N_{E_{u\cdot}A_{u\cdot}}(n)$ is similar to $\Delta N_{E_{u\cdot}I_{u\cdot}}(n)$, described above.
-
-\subsection{Model~3}
-
-In this model, reported cholera cases are assumed to stem from individuals who develop symptoms and seek healthcare.
-Therefore reported cases are assumed to come from an over-dispersed negative binomial model, given the increase in infected individuals:
-\begin{equation}
-  \label{model3:meas}
-  Y_{u, n} \mid \Delta N_{S_{u \cdot}I_{u}}(t) = \reportChange_u \sim \text{NB}\left(\reportRate \reportChange_{u}, \obsOverdispersion \right),
-\end{equation}
-where $\Delta N_{S_{u \cdot}I_{u \vaccCounter}}(n)$ is the number of individuals who moved from compartment $S_{u \vaccCounter}$ to $I_{u}$ since observation $t_{n-1}$.
-
-This measurement model is a minor change from that used by \cite{lee20}, which allowed for a change in the reporting rate on January 1st, 2018.
-The fitted values of the reporting rate---before and after January 2018---were $0.97$ and $0.097$, respectively.
-An instantaneous change from near perfect to almost non-existent reporting can be problematic, as it forces the model to explain the observed reduction in reported cases as a decrease in the reporting of cases, rather than a decrease in the prevalence of cholera.
-This shift was justified by a ``change of the case definition that occurred on January 1st, 2018";
-this claim was not cited, and we could find no evidence that such a drastic change in the reporting rate would be warranted.
-We therefore do not allow a change in reporting rate when fitting Model~3.
-
-\section{Initial Values}
-
-To perform inference on POMP models, it is necessary to propose an initial probability density for the latent process $f_{X_0}(x_0;\theta)$, including the possibility that the initial values of the latent states are known, or are a non-random function of the unknown parameter vector, $\theta$.
-This density is used to obtain initial values of the latent state when fitting and evaluating the model.
-For each of the models considered in this analysis, the initial conditions are derived by enforcing the model dynamics on reported cholera cases.
-It is also sometimes necessary to fit some initial value parameters in order to help determine initial values for weakly identifiable compartments.
-In the following subsections, we mention initial value parameters that were fit for each model.
-
-\subsection{Model~1}
-
-For this model, the number of individuals in the Recovered and Asymptomatic compartments are set to zero, but the initial proportion of Infected and Exposed individuals is estimated as initial value parameters ($\Iinit$ and $\Einit$, respectively) using the IF2 algorithm, implemented as \texttt{mif2} in the \texttt{pomp} package.
-Finally, the initial proportion of Susceptible individuals $S_{0, 0}$ is calculated as $S_{0, 0} = 1 - \Iinit - \Einit$.
-This model for the initial values of the latent states matches that which was used by \citet{lee20}.
-
-\subsection{Model~2}
-
-Model~2 assumes that the initial values are a deterministic function of the reporting rate and the initial case reports, and so no initial value parameters need to be estimated.
-Initial values for latent state compartments are chosen so as to enforce the model dynamics on the observed number of cases.
-Specifically, the model sets $I_{u0}(0) = y^*_{1u} / \reportRate$ for each unit $u \in 1:10$, where $I_{u0}(t)$ is the number of infected individuals in unit $u$ at time $t$, vaccination scenario $z = 0$, $y^*_{tu}$ is the reported number of cases, and $\reportRate$ is the reporting rate.
-This model for the initial values of the latent states matches that which was used by \citet{lee20}.
-
-\subsection{Model~3}
-
-The latent states of this model are initialized by enforcing the model dynamics on the incidence data from the start of the recorded cases until time $t_0$, requiring that some of the available data be used to determine the initial values of the latent states.
-This is the same approach that was taken by \citet{lee20}, who used the value $t_0 = \text{2014-02-22}$; this choice of $t_0$ results in modeling roughly only $60\%$ of the available data, some of which is later discarded for alternative reasons \citep{lee20sup}.
-
-We do not see any immediate reason that this model could not be extended to cover a larger range of the data, and chose the value $t_0 = \text{2010-11-13}$.
-This choice of $t_0$ corresponds to using approximately $1\%$ of the available data to determine initial values of the latent states.
-In addition to modeling a larger portion of the available data, this choice of $t_0$ corresponds to an important real-world event, as daily reports from each of the departments were not being sent to the health ministry until November 10, 2010 \citep{barzilay13};
-this choice of $t_0$ therefore makes $\bm{Y}\big(t_1\big)$ the first week of data once daily reports are being sent to the health ministry.
-The few observation times that exist before $t_0$ are used to initialize the model by enforcing model dynamics on these preliminary observations.
-For convenience, we denote these observations as $t_{-3}, t_{-2}$ and $t_{-1}$; as before, we let $y_{u,-k} = Y_{u}\big(t_{-k}\big)$ denote the observed case count for unit $u$ at time point $t_{-k}$, where $k \in 1:3$.
-Equations for the initial values of non-zero latent states are provided in Eqs.~\myeqref{eq:model3:I0}--\myeqref{eq:model3:B0}; these equations match those that were used by \citet{lee20}, the primary change being a change to the value of $t_0$.
-
-\begin{eqnarray}
-\label{eq:model3:I0}
-I_{u}\big(t_0\big) &=& \frac{y^*_{u, -1}}{7\reportRate\left(\muIR + \left(\muDeath + \choleraDeath\right)/365\right)},
-\\%[-3pt]
-\label{eq:model3:A0}
-A_{u}\big(t_0\big) &=& \frac{I_{u0}\big(t_0\big)\left(1 - \symptomFrac\right)}{\symptomFrac},
-\\
-\label{eq:model3:R0}
-R_{u01}\big(t_0\big) &=& R_{u02}\big(t_0\big) = R_{u03}\big(t_0\big) = \left(\frac{\sum_{k = -3}^0 y^*_{u, k}}{\reportRate \symptomFrac} - \left(I_{u0}\big(t_0\big) + A_{u0}\big(t_0\big)\right)\right) / 3
-\\
-\label{eq:model3:S0}
-S_{u0}\big(t_0\big) &=& \mathrm{Pop}_u - I_{u}\big(t_0\big) - A_{u}\big(t_0\big) - \sum_{k=1}^3 R_{u0k}\big(t_0\big)
-\\
-\label{eq:model3:B0}
-B_{u}\big(t_0\big) &=& \big[1 + \seasAmplitude \tilde{J}^r \big] \mathrm{Den}_u \, \Wshed \big[ I_{u}(t_0)+ \asymptomRelativeShed A_{u_0}(t) \big]/\Wshed.
-\end{eqnarray}
-
-In Eq.~\myeqref{eq:model3:B0}, $\tilde{J} = 0.002376$ is the median adjusted rainfall over the observation period.
-One important consideration to make with this parameter initialization model is when $y_{u, -1}^* = 0$, which occurs for units $u \in \{3, 4\}$, which correspond to the Grand'Anse and Nippes departments, respectively.
-When this is the case, each of the infectious $I_u(t_0)$, asymptomatic $A_u(t_0)$, and bacterial reservoir $W_u(t_0)$ compartments have a value of zero.
-Because Model~3 models cholera transmission primarily by means of the bacterial reservoir, this makes it nearly impossible for an outbreak to occur.
-Therefore for units $u \in \{3, 4\}$, we introduce initial value parameters $I_{0,0}^3$ and $I_{0,0}^4$, and calibrate these parameter values using the data.
-The resulting parameter estimates are used to obtain the remaining non-zero initial values of the latent states using Eqs.~\myeqref{eq:model3:A0}--\myeqref{eq:model3:B0}.
+%%%%%%%%%% START
 
@@
 The resulting model has a log-likelihood value of $\Sexpr{round(max(h3_final_results$search2$logLiks$logLik), 1)}$.
-% We also provide figures for department specific log-likelihoods and conditional log-likelihoods in Figs.~\ref{fig:finalUnitLL} and \ref{fig:finalCondLL}.
 The addition of the Hurricane parameters seems to have resulted in a dramatic increase in conditional likelihoods around and after October 2016, as seen in Fig.~\ref{fig:finalCondLL}.
@@
 
-\clearpage
-
-\section{Replication of \cite{lee20}}
-
-In this article we claimed that we were able to obtain better fits to the observed data using the same models that were proposed by \cite{lee20}.
-Along with visual comparisons to the data, this claim was supported by comparing likelihoods and AIC values in Table~2 in the manuscript.
-Because model likelihoods were not provided by \cite{lee20}, it is necessary to replicate these models in order to obtain likelihood estimates.
-Here we would like to thank the authors of \cite{lee20}, who provided detailed descriptions of their models, which enabled us to build on their work.
-In the following subsections, we use our \code{R} package \code{haitipkg} to reproduce some of the results of \cite{lee20}.
-This reproduction allows us to estimate the likelihoods of the \cite{lee20} version of Models~1--3, and also provides a demonstration of the importance and usefulness of reproducible research.
-
-\subsection{Model~1 Replication}
-
-The model was implemented by a team at Johns Hopkins Bloomberg School of Public Health (hereafter referred to as the Model~1 authors) in the \code{R} programming language using the \code{pomp} package \citep{king16}.
-Original source code is publicly available with DOI: 10.5281/zenodo.3360991.
-The final results reported by the Model~1 authors were obtained by using several different parameter sets rather than a single point estimate.
-According to the supplement materials, this was because model realizations from a single parameter set retained substantial variability, but multiple realizations from a collection of parameter sets resulted in a reasonable visual fit to the data.
-We are also inclined to believe that the use of multiple parameter values was in part intended to account for parameter uncertainty---the importance of which was discussed in the main text---an effort by the Model~1 authors that we applaud.
-Simulations from each of the parameter sets, however, were treated with equal importance when being used to diagnose the model fit and make inference on the system.
-This is problematic given Figures S8 and S9 of the supplement material, which suggest that some parameter sets that were used for inference may have been several hundred units of log-likelihood lower than other parameter sets that were simultaneously used to make forecasts.
-Such a large difference in log-likelihoods is well beyond the threshold of statistical uncertainty determined by Wilks' theorem, resulting in the equal use of statistically inferior parameter sets in order to make forecasts and conduct inference on the system.
-
-To fully reproduce the results of the Model~1 authors, it is necessary to use the exact same set of model parameters that were originally used to obtain the results presented by \cite{lee20}.
-Because these parameter sets were not made publicly available, we relied on the source code provided by the Model~1 authors to approximately recreate the parameter set.
-Due to software updates since the publication of the source code, we were unable to produce the exact same set of parameters.
-Running the publicly available source code, however, resulted in a set of parameters that are visually similar to those used by the Model~1 authors (See Figures~\ref{fig:PlotEpiDist} and \ref{fig:plotEndParams}).
-Furthermore, simulations using the set of parameters produced by the source code appear practically equivalent to those displayed by \cite{lee20} (See Figure~\ref{fig:plotMod1Sims}).
-
-<<Simulate Lee Model 1, echo=FALSE, cache=TRUE>>=
-h1_epi <- haiti1()
-N_LEE_SIMS <- 20
-
-registerDoRNG(18599687)
-foreach(
-  i = 1:length(h1LeeStartsEpi),
-  .combine = rbind
-) %dopar% {
-  guess <- h1LeeStartsEpi[i, ]
-  sims <- simulate(h1_epi, nsim = N_LEE_SIMS, format = 'data.frame',
-                   params = guess)  # 20 sims for each set of parameters
-  sims$param_set <- i
-  sims %>%
-    dplyr::select(.id, week, cases, param_set)
-} -> all_sims
-@
-
-<<PlotEpiDist, echo=FALSE, fig.cap="Compare to Figure S8 of Lee et al. (2020) supplement.", fig.width=7, fig.height=7, fig.align='center'>>=
-GGally::ggpairs(
-  h1LeeStartsEpi,
-  columns = c("rho", "tau", "beta1", "nu", "loglik")
-)
-@
-
-<<SimulateEpi, echo=FALSE, include=FALSE, message=FALSE, cache=TRUE>>=
-h1_end <- haiti1(period = "endemic")
-
-true_agg_cases <- haitiCholera %>%
-  select(-report) %>%
-  pivot_longer(
-    data = .,
-    cols = -date_saturday,
-    values_to = 'cases',
-    names_to = 'dep'
-  ) %>%
-  mutate(
-    date = as.Date(date_saturday),
-    dep = gsub("\\.", "_", dep)
-  ) %>%
-  mutate(
-    dep = case_when(dep == "Grand_Anse" ~ "Grande_Anse", TRUE ~ dep)
-  ) %>%
-  tidyr::pivot_wider(
-    data = .,
-    id_cols = c(date),
-    names_from = dep,
-    values_from = cases,
-    names_prefix = 'cases_'
-  ) %>%
-  mutate(
-    ReportedAll = cases_Artibonite + cases_Centre +
-      cases_Grande_Anse + cases_Nippes + cases_Nord +
-      cases_Nord_Est + cases_Ouest + cases_Sud +
-      cases_Sud_Est + cases_Nord_Ouest
-  )
-@
-
-<<FitEnd, echo=FALSE, message=FALSE, include=FALSE>>=
-foreach(
-  i = 1:nrow(h1LeeStartsEnd),
-  .combine = rbind
-) %dopar% {
-  r_params <- h1LeeStartsEnd[i, ][names(coef(h1_end))]
-  coef(h1_end) <- r_params
-
-  sims <- simulate(
-    h1_end, nsim = N_LEE_SIMS,
-    format = 'data.frame',
-    params = r_params
-  )
-  sims$param_set <- i
-  sims %>%
-    dplyr::select(.id, week, cases, param_set)
-} -> end_sims
-
-@
-
-<<plotEndParams, fig.width=7, fig.height=7, fig.align='center', fig.cap="Bivariate relationships between variables after fitting endemic period.
-Compare to S9 of Lee et al.
-(2020) supplement.", cache=TRUE, echo=FALSE>>=
-GGally::ggpairs(
-  h1LeeStartsEnd,
-  columns = c("rho", "tau", "beta1", "nu", "loglik")
-)
-@
-
-<<calcMod1LeeLikes, echo=FALSE, include=FALSE, message=FALSE>>=
-h1_epi_evals <- readRDS("../model1/run_level_3/lee1_epi_evals.rds")
-h1_end_evals <- readRDS("../model1/run_level_3/lee1_end_evals.rds")
-
-mod1_all_likes <- dplyr::inner_join(
-  x = h1_epi_evals,
-  y = h1_end_evals,
-  by = "parid"
-  ) %>%
-  dplyr::mutate(
-    joint_ll = pfLL.x + pfLL.y
-  )
-
-# Approximate the likelihood of the Lee et al (2020) version of Model 1 by
-# using the parameter sets that resulted in the maximum likelihood.
-mod1_lee_ll <- max(mod1_all_likes$joint_ll, na.rm = TRUE)
-@
-
-<<plotMod1Sims, echo=FALSE, fig.cap=paste0("Simulations using parameter sets that were generated by running source code provided by Lee et al (2020).
-Compare to Figure S7 of \\cite{lee20sup}.
-The upper bound for the likelihood of this model is ", round(mod1_lee_ll), "."), cache=TRUE, fig.width=6.5, fig.height=3.75>>=
-
-quants <- all_sims %>%
-  filter(!param_set %in% c(1, 2)) %>%  # Remove parameter sets inconsistent with Lee et al (2020).
-  group_by(week) %>%
-  summarize(
-    q025 = quantile(cases, probs = 0.025, na.rm = TRUE),
-    q50  = quantile(cases, probs = 0.500, na.rm = TRUE),
-    q975 = quantile(cases, probs = 0.975, na.rm = TRUE)
-  ) %>%
-  ungroup() %>%
-  mutate(date = lubridate::ymd("2010-10-16") + lubridate::weeks(week))
-
-quants_end <- end_sims %>%
-  group_by(week) %>%
-  summarize(
-    q025 = quantile(cases, probs = 0.025, na.rm = TRUE),
-    q50  = quantile(cases, probs = 0.500, na.rm = TRUE),
-    q975 = quantile(cases, probs = 0.975, na.rm = TRUE)
-  ) %>%
-  ungroup() %>%
-  mutate(date = lubridate::ymd("2010-10-16") + lubridate::weeks(week))
-
-ggplot() +
-  geom_point(data = true_agg_cases, aes(x = date, y = ReportedAll)) +
-  geom_line(data = quants, aes(x = date, y = q50), col = '#0b8a2b', size = 1) +
-  geom_ribbon(
-    data = quants,
-    aes(x = date, ymin = q025, ymax = q975),
-    alpha = 0.5, fill = '#0b9c30'
-  ) +
-  geom_line(data = quants_end, aes(x = date, y = q50), col = '#882871') +
-  geom_ribbon(data = quants_end, aes(x = date, ymin = q025, ymax = q975), alpha = 0.5, fill = '#882871') +
-  ylab("Reported, Symptomatic Incidence") +
-  scale_x_date(date_breaks = "1 year") +
-  theme(axis.title.x = element_blank(),
-        axis.text.x = element_text(hjust = 1, angle = 45))
-@
-
-Because the model forecasts provided by \cite{lee20} come from various sets of parameters---which each correspond to a unique log-likelihood value---it is not obvious how one would obtain an estimate for the log-likelihood of the model that was used for simulations by the Model~1 authors.
-One approach could be to calculate the logarithm of the weighted mean of the likelihoods for each parameter sets used to obtain the forecasts, where the weights are proportional to the number of times the parameter set was used.
-However, in an effort to not underestimate the likelihood of the model of the Model~1 authors, we report the estimated log-likelihood as the log-likelihood value corresponding to the parameter set with the largest likelihood value, even though the majority of simulations were obtained using parameter sets with lower likelihood values.
-In this sense, we consider the log-likelihood reported in Table~1 of the main text to be an upper-bound of the log-likelihood of the model used by \cite{lee20}.
-For each parameter set, the log-likelihood was estimated using a particle filter, implemented as the \texttt{pfilter} function in the \texttt{pomp} package.
-
-\subsection{Model~2 Replication}\label{sec:mod2rep}
-
-Model~2 was developed by a team that consisted of members from the Fred Hutchinson Cancer Research Center and the University of Florida (hereafter referred to as the Model~2 authors).
-While Model~2 is the only deterministic model we considered in our analysis, it contains perhaps the most complex descriptions of cholera in Haiti: Model~2 accounts for movement between spatial units; human-to-human and environment-to-human cholera infections; and transfer of water between spatial units based on elevation charts and river flows.
-
-The source code that the Model~2 authors used to generate their results was written in the \code{Python} programming language, and is publicly available at \url{10.5281/zenodo.3360857} and its accompanying GitHub repository \url{https://github.com/lulelita/HaitiCholeraMultiModelingProject}.
-In order to perform our analysis in a unified framework, we re-implemented this model in the \code{R} programming language using the \code{spatPomp} package \citep{asfaw23arxiv}, which facilitates the creation of meta-population models.
-We note that the travel and water matrices used to implement the complex dynamics in Model~2 \citep{lee20sup} are not available in either the Zenodo archive or the GitHub repository;
-instead, we obtained these matrices via personal correspondence with the Model~2 authors.
-Using these matrices, and the point estimates for model parameters provided by \citep{lee20sup}, we created trajectories of the cholera dynamics and compared this to available data.
-These trajectories, shown in Figure~\ref{fig:mod2rep}, are very similar to the trajectories shown in Figure~S15 of \cite{lee20sup}.
-
-<<Model2Replication, cache=TRUE, echo=FALSE>>=
-h2_epi <- haiti2()
-h2_end <- haiti2(region = 'after')
-
-h2_epi_traj <- trajectory(h2_epi, format = 'data.frame')
-h2_end_traj <- trajectory(h2_end, format = 'data.frame')
-
-h2_epi_traj$Ctotal <- rowSums(h2_epi_traj[, paste0("C", 1:10)]) * 0.2
-h2_end_traj$Ctotal <- rowSums(h2_end_traj[, paste0("C", 1:10)]) * 0.2
-
-dep_plot_df <- haitipkg::haitiCholera %>%
-  select(-report) %>%
-  tidyr::pivot_longer(
-    data = .,
-    cols = -date_saturday,
-    values_to = 'cases',
-    names_to = 'dep'
-  ) %>%
-  mutate(
-    date = as.Date(date_saturday),
-    dep = gsub("\\.", "_", dep)
-  ) %>%
-  mutate(
-    dep = case_when(dep == "Grand_Anse" ~ "Grande_Anse", TRUE ~ dep)
-  )
-
-true_agg_cases <- dep_plot_df %>%
-  tidyr::pivot_wider(
-    data = .,
-    id_cols = c(date),
-    names_from = dep,
-    values_from = cases,
-    names_prefix = 'cases_'
-  ) %>%
-  mutate(
-    ReportedAll = cases_Artibonite + cases_Centre +
-      cases_Grande_Anse + cases_Nippes + cases_Nord +
-      cases_Nord_Est + cases_Ouest + cases_Sud +
-      cases_Sud_Est + cases_Nord_Ouest
-  ) %>%
-  select(date, ReportedAll)
-
-h2_epi_traj$date <- as.Date(
-  lubridate::round_date(
-    lubridate::date_decimal(h2_epi_traj$year),
-    unit = 'day'
-  )
-)
-
-h2_end_traj$date <- as.Date(
-  lubridate::round_date(
-    lubridate::date_decimal(h2_end_traj$year),
-    unit = 'day'
-  )
-)
-
-all_traj <- bind_rows(
-  select(h2_epi_traj, date, Ctotal),
-  select(h2_end_traj, date, Ctotal)
-)
-
-projections <- left_join(
-  x = true_agg_cases,
-  y = all_traj,
-  by = 'date'
-)
-@
-
-\begin{figure}[!h]
-<<plotModel2Rep, cache=TRUE, echo=FALSE, fig.width=6.5, fig.height=4.5>>=
-gg_epi <- projections %>%
-  filter(date < as.Date("03-01-2014", format = '%m-%d-%Y')) %>%
-  ggplot(aes(x = date)) +
-  geom_line(aes(y = ReportedAll), col = 'black', linetype = 'dashed') +
-  geom_line(aes(y = Ctotal), col = 'red', size = 1) +
-  theme_bw() +
-  scale_y_continuous(limits = c(0, 30000), breaks = seq(0, 30000, 5000)) +
-  theme(axis.title.x = element_blank()) +
-  ylab("Reported number of cases")
-
-gg_end <- projections %>%
-  filter(date >= as.Date("03-01-2014", format = '%m-%d-%Y')) %>%
-  ggplot(aes(x = date)) +
-  geom_line(aes(y = ReportedAll), col = 'black', linetype = 'dashed') +
-  geom_line(aes(y = Ctotal), col = 'red', size = 1) +
-  theme_bw() +
-  scale_y_continuous(limits = c(0, 7000), breaks = seq(0, 7000, 1000)) +
-  theme(axis.title.x = element_blank()) +
-  ylab("Reported number of cases") +
-  scale_x_date(date_breaks = "1 year", date_labels = "%Y")
-
-cowplot::plot_grid(gg_epi, gg_end, ncol = 1)
-@
-\caption{\label{fig:mod2rep}
-Model 2 trajectories using the \code{haitipkg}. Compare to Figure S15 of \cite{lee20sup}.
-}
-\end{figure}
-
-There are minor differences between Figure~\ref{fig:mod2rep} and Figure~S15 of \cite{lee20sup}.
-While the discrepancy appears minor, the deterministic nature of Model~2 implies that an exact replication of model trajectories should be possible.
-In this case, these discrepancies may possibly be attributed to implementing the model and plotting the model trajectory in two different programming languages.
-Another potential explanation for the discrepancy is that the parameters that we used are only approximately the same as those used by \citet{lee20sup}.
-For example, the parameters $\transmission$, $\Wbeta{}$ (See Table~\ref{tab:translate}) had reported values of $9.9 \times 10^{-7}$ and $4.03 \times 10^{-2}$, respectively (Table~S13 of \cite{lee20sup}), but were actually fit to data and therefore likely these values have been rounded.
-Additionally, our implementation of Model~2 used a time scale of years and many of the parameters were reported on a weekly scale, so small differences may result due to unit conversions.
-The collective effect of these small differences in model parameters likely will result in small differences in model trajectories.
-
-Some additional concerns about being able to accurately replicate the results of \cite{lee20} are valid.
-Details about the measurement models and how latent states were initialized for the epidemic model were not provided by \citet{lee20sup} and therefore these details must be inferred by looking at the provided source code.
-According to repository comments, the files \code{fit\allowbreak In\allowbreak Pieces\allowbreak 3params\allowbreak Clean\allowbreak May2019\allowbreak Public.py} and \code{fit\allowbreak In\allowbreak Pieces\allowbreak Mu\allowbreak With\allowbreak Frac\allowbreak Sus\allowbreak Fixed\allowbreak All\allowbreak Infections\allowbreak Public.py} were used to fit the epidemic and endemic phases of the model respectively, although it is apparent that these exact files were not used to obtain the reported results since the files contain some variable-naming errors that make it impossible to run the files without making modifications \footnote{One example of why the code cannot be run that the file loads functions from a non-extant file named \code{choleraEqs.py} in line 13 rather than \code{cholera\allowbreak Eqs\allowbreak Public.py}.}.
-The inability to replicate the results by \citet{lee20} by running the provided source code makes checking whether or not a our numeric implementation faithfully represents their results very difficult.
-Additionally, the script that was said to been used to obtain the results reported by \cite{lee20} appears to use a different measurement model than what was described in the supplemental material, again making it difficult to fully replicate the result of \cite{lee20} without being able to easily run the provided source code.
-In this case, we chose to use measurement model described by Eq.~\ref{model2:measEpi} for both phases of the epidemic, as this seemed to visually match the results of \cite{lee20} most closely.
-
-\subsection{Model~3 Replication}
-
-Model~3 was developed by a team of researchers at the Laboratory of the Swiss Federal Institute of Technology in Lausanne, hereafter referred to as the Model~3 authors.
-The code that was originally used to implement Model~3 is archived with the DOI: \url{10.5281/zenodo.3360723}, and also available in the public GitHub repository: \code{jcblemai/haiti-mass-ocv-campaign}.
-Because the code was made publicly available, and final model parameters were reported in the supplementary material of \cite{lee20}, we were able to reproduce Model~3 by directly using the source code.
-In Fig.~\ref{fig:mod3rep}, we plot simulations from this model.
-This figure can be compared to Figure S18 of \cite{lee20}.
-We note that slight differences may be accounted for due to variance in the model simulations and the difference in programming language used to produce the figure.
-Overall, the high standard of reproducibility that was achieved by the Model~3 authors facilitated the ability to readily replicate their model and results.
-
-<<Model3Replication, echo=FALSE, cache=TRUE>>=
-NSIM <- 500
-
-plot_order <- c(
-  'Artibonite',
-  'Sud_Est',
-  'Nippes',
-  'Nord_Est',
-  'Ouest',
-  'Centre',
-  'Nord',
-  'Sud',
-  'Nord_Ouest',
-  'Grande_Anse'
-)
-
-dep_labeller <- as_labeller(
-  c(
-    'Artibonite' = 'Artibonite',
-    'Sud_Est' = 'Sud-Est',
-    'Sud.Est' = 'Sud-Est',
-    'Nippes' = 'Nippes',
-    'Nord_Est' = 'Nord-Est',
-    'Nord.Est' = 'Nord-Est',
-    'Ouest' = 'Ouest',
-    'Centre' = 'Centre',
-    'Nord' = 'Nord',
-    'Sud' = 'Sud',
-    'Nord_Ouest' = 'Nord-Ouest',
-    'Nord.Ouest' = 'Nord-Ouest',
-    'Grande_Anse' = 'Grand\'Anse',
-    'Grand.Anse' = 'Grand\'Anse'
-  )
-)
-
-lee3 <- lee3_spatPomp()
-lee3_sims_quants <- bake("../model3/si/lee3Sims.rds", {
-
-  # Simulate from lee et al 2020 model
-  sims_temp <- simulate(
-    lee3, nsim = NSIM,
-    seed = 321, format = 'data.frame'
-  )
-
-  # Get desired quantiles of reported cases from simulations
-  quants <- sims_temp %>%
-    dplyr::group_by(unitname, time) %>%
-    dplyr::summarise(
-      q05 = quantile(cases, 0.025, na.rm = T),
-      mean = mean(cases, na.rm = T),
-      q50 = quantile(cases, 0.5, na.rm = T),
-      q95 = quantile(cases, 0.975, na.rm = T)
-    ) %>%
-    dplyr::ungroup() %>%
-    dplyr::mutate(
-      date = lubridate::date_decimal(time),
-      date = as.Date(lubridate::round_date(date))
-    ) %>%
-    dplyr::rename(dep = unitname)
-
-  # Remove cases from Ouest that were not considered by Lee et al. (2020)
-  quants %>%
-    dplyr::filter(
-      !(dep == "Ouest" & date < '2017-06-10')
-    )
-
-})
-
-# Get the subset of cholera cases that match simulations from Lee et al. (2020) model.
-sub_cases <- haitiCholera %>%
-  dplyr::mutate(date = as.Date(date_saturday)) %>%
-  dplyr::select(-report, -date_saturday) %>%
-  dplyr::filter(lubridate::decimal_date(date) >= min(lee3@times)) %>%
-  tidyr::pivot_longer(
-    data = .,
-    cols = -date,
-    names_to = "dep",
-    values_to = "cases"
-  ) %>%
-  dplyr::mutate(
-    dep = dplyr::case_when(
-      dep == "Grand.Anse" ~ "Grande_Anse",
-      dep == "Nord.Est" ~ "Nord_Est",
-      dep == "Nord.Ouest" ~ "Nord_Ouest",
-      dep == "Sud.Est" ~ "Sud_Est",
-      TRUE ~ dep
-    )
-  )
-@
-
-\begin{figure}[!h]
-<<PlotMod3Rep, echo=FALSE, message=FALSE, fig.height=3.75>>=
-ggplot() +
-  # Region for upper and lower simulation bounds
-  geom_ribbon(
-    data = lee3_sims_quants,
-    aes(x = date, ymin = q05, ymax = q95),
-    fill = 'darkblue', alpha = 0.5
-  ) +
-  # Median line
-  geom_line(
-    data = lee3_sims_quants,
-    aes(x = date, y = q50),
-    col = 'darkblue'
-  ) +
-  # Observed data
-  geom_point(
-    data = dplyr::filter(sub_cases, cases < 500),
-    aes(x = date, y = cases), size = 0.2
-  ) +
-  facet_wrap(
-    ~factor(dep, levels = plot_order),
-    nrow = 2, labeller = dep_labeller
-  ) +
-  labs(y = 'Reported Cholera Cases') +
-  scale_x_date(date_breaks = "6 months", date_labels = "%Y-%m") +
-  theme(
-    axis.text.x = element_text(angle = 40, hjust = 1, vjust = 1, size = 7),
-    axis.title.x = element_blank()
-  )
-@
-\caption{\label{fig:mod3rep}
-Simulations from Model~3.
-Compare to Figure S18 of \cite{lee20}.
-}
-\end{figure}
-
-<<lee3Cleanup, echo=FALSE, include=FALSE, message=FALSE>>=
-rm(lee3, lee3_sims_quants, sub_cases)
-gc()
-@
-
-\section{Forecasting with parameter uncertainty}
-
-Let $f_{Y_{1:N}}(y_{1:N} | \theta)$ denote the pdf of the model under consideration, were $\theta$ is a parameter vector that indexes the model.
-Furthermore, denote the observed data as $y_{1:N}^*$.
-Because the uncertainty in just a single parameter can lead to drastically different forecasts \citep{saltelli20},
-parameter uncertainty should be considered when obtaining model forecasts when the goal is to influencing policy.
-In a Bayesian modelling paradigm, the most natural way to account for parameter uncertainty in model forecasts is to suppose that $\theta$ comes from a distribution $f_{\Theta}$, and then to obtain $J$ forecasts from the model where each forecast is obtained using parameters drawn from the posterior distribution $\Theta_{1:J} \mid Y_{1:N} = y_{1:N}^* \sim f_{\Theta}\big(\theta | Y_{1:N} = y_{1:N}^*\big)$.
-
-When frequentist methods are used, however, there does not exist a posterior distribution from which one could sample.
-A common approach could be to obtain a weighted average of the simulations from various models \citep{hoeting99}, but this can be problematic when forecasts from each model are very different from each other \citep{grueber11}.
-A similar approach that has been taken \citep{king15} is to obtain model forecasts using multiple sets of parameter values and then sample from the resulting forecasts using weights proportional to the corresponding likelihoods of the parameter values.
-This approach could be considered as empirical Bayes, as it is equivalent to using a discrete uniform prior where the set of values in the prior distribution is determined by a stochastic routine applied to the observed data, as discussed below.
-
-For each $i \in 1:k$, let $\Theta_i$ be a random vector of model parameters.
-Then, letting $\Theta$ denote the true model parameters, we endow the set $\{\Theta_1, \Theta_2, \ldots, \Theta_k\}$ with a discrete uniform distribution, such that $P\big(\Theta = \Theta_i\big) = \frac{1}{K}$ for all values $i \in \seq{1}{k}$.
-Using this as a prior distribution, the posterior distribution of $\Theta | Y_{1:N} = y_{1:N}^*$ can be expressed as:
-$P\big(\Theta = \Theta_k | Y_{1:N} = y_{1:N}^*\big) = \frac{f_{Y_{1:N}}(y_{1:N}^*| \Theta_k)}{\sum_{l = 1}^K f_{Y_{1:N}}(y_{1:N}^*| \theta_l)}$.
-In a standard empirical Bayes analysis, the values $\Theta_1, \ldots, \Theta_k$ of the prior distribution would be chosen using the observed data, resulting in a posterior distribution that weighs the prior parameter vectors proportional to their corresponding likelihoods.
-Instead, we choose $\Theta_i$ to be the output of a stochastic routine applied to the observed data by setting $\Theta_i$ to be the output of an iterated filtering algorithm.
-In practice, because the likelihood maximization routines of iterated filtering methods are stochastic, it is common to run the iterated filtering method multiple times for each model in order to obtain a maximum likelihood estimate for model parameters.
-This results in a natural set of parameters near the MLE that could be used as the discrete prior distribution.
-
-\section{Translating to \citet{lee20} notation}
-
-Since the models of \citet{lee20} were developed independently, the choice of notation varies inconsistently between models.
-For our reanalysis, we rename parameters to provide a unified notation facilitating comparison between models.
-Table~\ref{tab:translate} maps this notation back to the original notations, for reference.
-
-\begin{table}
-  \begin{center}
-  \begin{tabular}{|c|c|c|c|c|}\hline
-    \multirow{2}{*}{Parameter} & Our & \multicolumn{3}{c|}{Lee et al. (2020a)} \\\cline{3-5}
-     & Notation & 1 & 2 & 3 \\
-    \hline
-    \hline
-    % Table begins here
-    Reporting Rate & $\reportRate$ & $\rho$ & $\rho$ & $\epsilon_1, \epsilon_2$ \\\hline
-    Mixing Coefficient & $\mixExponent$ & $\nu$ & --- & --- \\\hline
-    Measurement Over-Dispersion & $\obsOverdispersion$ & $\tau$ & --- & $p$ \\\hline
-    Birth Rate & $\muBirth$ & $\mu$ & --- & --- \\\hline
-    Natural Mortality Rate & $\muDeath$ & $\delta$ & --- & $\mu$ \\\hline
-    Cholera Mortality Rate & $\choleraDeath$ & --- & --- & $\alpha$ \\\hline
-    Latent Period & $1/\muEI$ & $1/\sigma$ & $1/\gamma_E$ & --- \\\hline
-    Recovery Rate & $\muIR$ & $\gamma$ & $\gamma$ & $\gamma$ \\\hline
-    Loss of Immunity & $\muRS$ & $\alpha$ & $\sigma$ & $\rho$ \\\hline
-    Symptomatic Ratio & $\symptomFrac$ & $1 - \theta_0$ & $k$ & $\sigma$ \\\hline
-    Asymptomatic Relative Infectiousness & $\asymptomRelativeInfect$ & $\kappa$ & $red_\beta$ & --- \\\hline
-    Human-to-Water Shedding & $\Wshed$ & --- & $\mu$ & $\theta_I$ \\\hline
-    Asymptomatic Relative Shedding & $\asymptomRelativeShed$ & --- & $red_\mu$ & $\theta_A/\theta_I$ \\\hline
-    Seasonal Amplitude & $\seasAmplitude$ & --- & $\alpha_s$ & $\lambda$ \\\hline
-    Transmission & $\transmission$ & $\beta$ & $\beta$ & $c$ \\\hline
-    Water-to-Human & $\Wbeta{}$ & --- & $\beta_W$ & $\beta$ \\\hline
-    Bacteria Mortality & $\Wremoval$ & --- & $\delta$ & $\mu_\beta$ \\\hline
-    Vaccination Efficacy & $\vaccineEfficacy$ & $\theta_{vk}$ & $\theta_1, \theta_2, \theta_{1_5}, \theta_{2_5}$ & $\eta_{1d}, \eta_{2d}$ \\\hline
-    Process Over-dispersion & $\sigmaProc$ & --- & --- & $\sigma_w$\\\hline
-    % Half Saturation Constant & $\Wsat$ & --- & $Sat$ & $K$ \\\hline
-  \end{tabular}
-  \end{center}
-  \caption{
-  \label{tab:translate}Translations between our common notation and notation used by \cite{lee20}
-  }
-\end{table}
-
-\clearpage
+%%%%%%%%%% END
 

### 2023-09-01 si/inputs/initialValues.tex
Started work to make individual si sections seperate

--- 2023-08-31 si/si.Rnw
@@
-\documentclass[11pt]{article}
-
-\pdfminorversion=4
-
-\newcommand\secTitleSpace{\hspace{3mm}}
-
-<<packages, include=FALSE>>=
-library(tidyverse)
-library(pomp)
-library(haitipkg)
-library(spatPomp)
-library(doParallel)
-library(doRNG)
-
-cores <-  as.numeric(Sys.getenv('SLURM_NTASKS_PER_NODE', unset = NA))
-if(is.na(cores)) cores <- detectCores()
-# cores <- 20
-registerDoParallel(cores)
-theme_set(theme_bw())
-
-if (!dir.exists("../model3/si")) {
-  dir.create("../model3/si", recursive = TRUE)
-}
-@
-
-<<set-opts, include=FALSE>>=
-options(
-  scipen = 2,
-  help_type = "html",
-  stringsAsFactors = FALSE,
-  continue = "+  ",
-  width = 70,
-  useFancyQuotes = FALSE,
-  reindent.spaces = 2,
-  xtable.comment = FALSE
-)
-@
-
-<<knitr-opts, include=FALSE, purl=FALSE>>=
-library(knitr)
-opts_knit$set(concordance=TRUE)
-opts_chunk$set(
-  progress=TRUE,prompt=TRUE,highlight=FALSE,
-  tidy=TRUE,
-  tidy.opts=list(
-    keep.blank.line=FALSE
-  ),
-  comment="",
-  warning=FALSE,
-  message=FALSE,
-  error=TRUE,
-  echo=FALSE,
-  cache=FALSE,
-  strip.white=TRUE,
-  results="markup",
-  fig.path="figure/",
-  fig.lp="fig:",
-  fig.align="center",
-  fig.show="asis",
-  dev="pdf",
-  dev.args=list(
-    bg="transparent",
-    pointsize=9
-  )
-)
-
-myround <- function (x, digits = 1) {
-  # taken from the broman package
-  if (digits < 1)
-    stop("This is intended for the case digits >= 1.")
-  if (length(digits) > 1) {
-    digits <- digits[1]
-    warning("Using only digits[1]")
-  }
-  tmp <- sprintf(paste("%.", digits, "f", sep = ""), x)
-  zero <- paste0("0.", paste(rep("0", digits), collapse = ""))
-  tmp[tmp == paste0("-", zero)] <- zero
-  tmp
-}
-
-@
-
-%% copied from pnas-new.cls
-\usepackage{amsmath,amsfonts,amssymb,graphicx}
-
-%%%% parameters %%%%%%%%%%5
-\newcommand\Wsat{W_{\mathrm{sat}}}
-\newcommand\muIR{\mu_{IR}}
-\newcommand\muEI{\mu_{EI}}
-\newcommand\transmission{\beta}
-\newcommand\seasAmplitude{a}
-\newcommand\rainfallExponent{r}
-\newcommand\muRS{\mu_{RS}}
-\newcommand\vaccineEfficacy{\theta}
-\newcommand\muBirth{\mu_S}
-\newcommand\muDeath{\delta}
-\newcommand\choleraDeath{\delta_{C}}
-\newcommand\symptomFrac{f}
-\newcommand\asymptomRelativeInfect{\epsilon}
-\newcommand\asymptomRelativeShed{\epsilon_{W}}
-\newcommand\Wbeta[1]{\beta_{W#1}}
-\newcommand\Wremoval{\delta_W}
-\newcommand\Wshed{\mu_W}
-\newcommand\mixExponent{\nu}
-\newcommand\sigmaProc{\sigma_{\mathrm{proc}}}
-\newcommand\reportRate{\rho}
-\newcommand\obsOverdispersion{\psi}
-\newcommand\phaseParm{\phi}
-\newcommand\transmissionTrend{\zeta}
-\newcommand\Binit{\xi}
-\newcommand\vaccClass{Z}
-\newcommand\vaccCounter{z}
-\newcommand\modelCounter{m}
-\newcommand\Whur[1]{\beta_{W#1}^{hm}}
-\newcommand\hHur[1]{h_{#1}^{hm}}
-\newcommand\Iinit{I_{0,0}}
-\newcommand\Einit{E_{0, 0}}
-\newcommand\reportChange{\delta}
-
-\newcommand\missing{\hspace{6mm} ---}
-\newcommand\fixed{\color{blue}}
-\newcommand\demography{\bullet}
-
-\newcommand\mytitle{Informing policy via dynamic models: Cholera in Haiti}
-\newcommand\comma{{\hspace{-0.25mm},\hspace{-0.2mm}}}
-
-%%%%%%%%%%%%%%%% START OF USER DEFINITIONS %%%%%%%%%%%%%%%5
-
-\newcommand\siOnly[1]{} %% redefined for the supplement
-\newcommand\msOnly[1]{#1} %% redefined for the supplement
-
-\newcommand\code[1]{\texttt{#1}}
-\usepackage{xurl}
-% \usepackage[breaklinks]{hyperref}
-\usepackage{makecell}
-\usepackage{multirow}
-\usepackage{multicol}
-\usepackage[ruled,noline,linesnumbered]{algorithm2e}
-% \usepackage{setspace}
-
-%% EDITING MACROS
-\usepackage{color}
-\usepackage[normalem]{ulem}% to use \sout in feedback commands
-\usepackage{bm}
-% orange for EI
-\definecolor{orange}{rgb}{1,0.5,0}
-\newcommand\ei[2]{\sout{#1} \textcolor{orange}{#2}}
-\newcommand\eic[1]{\textcolor{orange}{[#1]}}
-% green for JW
-\definecolor{green}{rgb}{0,0.5,0}
-\newcommand\jw[2]{\sout{#1} \textcolor{green}{#2}}
-\newcommand\jwc[1]{\textcolor{green}{[#1]}}
-% purple for JJ
-\definecolor{purple}{rgb}{0.5,0,1}
-\newcommand\jj[2]{\sout{#1} \textcolor{purple}{#2}}
-\newcommand\jjc[1]{\textcolor{purple}{[#1]}}
-% cyan for AR
-\definecolor{cyan}{rgb}{0,.5,.5}
-\newcommand\ar[2]{\sout{#1} \textcolor{cyan}{#2}}
-\newcommand\arc[1]{\textcolor{cyan}{#1}}
-% light brown for KT
-\definecolor{lightbrown}{rgb}{0.5,0.5,0}
-\newcommand\kt[2]{\sout{#1} \textcolor{lightbrown}{#2}}
-\newcommand\ktc[1]{\textcolor{lightbrown}{#1}}
-
-%% basic POMP definitions
-\newcommand\Xspace{{\mathbb X}}
-\newcommand\Yspace{{\mathbb Y}}
-\newcommand\Thetaspace{\R^{\Thetadim}}
-\newcommand\Xunitspace{{\scriptsize{\mathcal X}}}
-\newcommand\Yunitspace{{\tiny{\mathcal Y}}}
-\newcommand\hatTheta{\widehat{\Theta}}
-\newcommand\Xdim{{\mathrm{dim}}(\Xspace)}
-\newcommand\Ydim{{\mathrm{dim}}(\Yspace)}
-\newcommand\Thetadim{P}
-\newcommand\thetadim{p}
-\newcommand\timeSet{{\mathbb T}}
-\newcommand\data[1]{#1^*}
-
-\newcommand\unitSpecific{\hspace{0.1mm}\mathrm{us}}
-\newcommand\shared{\hspace{0.15mm}\mathrm{sh}}
-
-% for algorithm pseudocode
-\SetKwFor{For}{for}{:}{end}
-
-%% for particle filters
-\newcommand\unit{u}
-\newcommand\altUnit{\tilde u}
-\newcommand\Unit{U}
-\newcommand\rep{i}
-\newcommand\Rep{\mathcal{I}}
-\newcommand\island{\rep}
-\newcommand\Island{\Rep}
-%\newcommand\txtisland{replicate}
-\newcommand\altIsland{\tilde i}
-\usepackage[mathscr]{euscript}
-
-\renewcommand\time{n}
-\newcommand\myvec[1]{\boldsymbol{#1}}
-\newcommand\altTime{\tilde n}
-\newcommand\Time{N}
-\newcommand\Np{J}
-\newcommand\np{j}
-\newcommand\altNp{k}
-\newcommand\altAltNp{\tilde j}
-\newcommand\resampleIndex{r}
-\newcommand\unitWeight{w}
-
-%% for all iterated filtering algorithms
-\newcommand\Nit{M}
-\newcommand\nit{m}
-
-%% customized math macros
-\newcommand\prob{\mathbb{P}}
-\newcommand\E{\mathbb{E}}
-\newcommand\dd[1]{\mathrm{d}{#1}}
-\newcommand\given{{\,\vert\,}}
-\newcommand\equals{{{\,}={\,}}}
-\newcommand\myequals{\hspace{0.5mm}{=}\hspace{0.5mm}}
-\newcommand\myto{{\;:\;}}
-\newcommand\seq[2]{{#1}\!:\!{#2}}
-\newcommand\mydot{{\,\cdot\,}}
-\newcommand\cp[2]{N_{\mathrm{#1}\mathrm{#2}}}
-\newcommand\giventh{{\hspace{0.5mm};\hspace{0.5mm}}}
-\newcommand\normal{\mathcal{N}}
-\newcommand\argequals{{\,=\,}}
-\newcommand\lags{c}
-\newcommand\maxlag{\overline{c}}
-\newcommand\nlfList{C}
-\newcommand\bigO{\mathcal{O}}
-\newcommand\loglik{\lambda}
-\newcommand\loglikMC{\MC{\loglik}}
-\newcommand\R{\mathbb{R}}
-\newcommand\param{\,;}
-\newcommand\mycolon{{\hspace{0.6mm}:\hspace{0.6mm}}}
-\newcommand\MC[1]{#1^{\,\mbox{\tiny MC}}}
-\newcommand\EMC{{\E}}
-\newcommand\Var{\mathrm{Var}}
-\newcommand\var{\Var}
-\newcommand\Cov{\mathrm{Cov}}
-\newcommand\cov{\Cov}
-\newcommand\iid{\mathrm{iid}}
-\newcommand\dist{d}
-\newcommand\transpose{\top}
-
-\newcommand\gnsep{,}
-
-\def\lik{L}
-\def\loglik{\ell}
-
-%% THEOREM-LIKE ENVIRONMENTS
-\usepackage{amsthm}
-\newtheorem{prop}{Proposition}
-\newtheorem{theorem}{Theorem}
-\newtheorem{lemma}{Lemma}
-\newtheorem{example}{Example}
-\newtheorem{remark}{Remark}
-\newtheorem{cor}{Corrolary}
-\newtheorem{fact}{Fact}
-\newtheorem{assumption}{Assumption}
-\newtheorem{assumptionB}{Assumption}
-\renewcommand\theassumption{A\arabic{assumption}}
-\renewcommand\theassumptionB{B\arabic{assumptionB}}
-
-%% FOR PSEUDOCODE SETUP
-\newcommand\mystretch{\rule[-2mm]{0mm}{5mm} }
-\newcommand\asp{\hspace{4mm}}
-\newcommand\codeIndent{\hspace{4mm}}
-
-%%%%%%%%%%%%%%%% END OF USER DEFINITIONSS %%%%%%%%%%%%%%%
-
-\bibliographystyle{apalike}
-\usepackage{natbib}
-\usepackage{fullpage}
-
-\newcommand\myeqref[1]{(\ref{#1})}
-
-\renewcommand{\contentsname}{Supplementary Content}
-\renewcommand{\refname}{Supplementary References}
-\renewcommand\thefigure{S-\arabic{figure}}
-\renewcommand\thetable{S-\arabic{table}}
-\renewcommand\thesection{S\arabic{section}}
-\renewcommand\theequation{S\arabic{equation}}
-\renewcommand\theprop{S\arabic{prop}}
-\renewcommand\thelemma{S\arabic{lemma}}
-\renewcommand\thealgocf{S\arabic{algocf}}
-
-\setcounter{tocdepth}{1}
-
-\begin{document}
-
-\date{\today}
-\title{Supplement to ``{\mytitle}''}
-  \author{Jesse Wheeler, AnnaElaine L. Rosengart, Zhuoxun Jiang, \\ Kevin Tan, Noah Treutle and Edward L. Ionides
- \\ \hspace{.2cm}\\
-    Department of Statistics, University of Michigan\\
-}
-
-\newcommand{\blind}{1}
-
-\if1\blind
-{
-\maketitle
-}
-\fi
-
-\if0\blind
-{
-  \bigskip
-  \bigskip
-  \bigskip
-  \begin{center}
-    {\LARGE\bf \mytitle}
-\end{center}
-  \bigskip
-  \bigskip
-}\fi
-
-\tableofcontents
-
-\newpage
-
-\section{Markov chain and differential equation interpretations of compartment flow rates}
-
-In Sections 2.1, 2.2 and 2.3 of the main article, we define compartment models in terms of their flow rates.
-For a discrete population model, these rates define a Markov chain.
-For a continuous and deterministic model, the rates define a system of ordinary differential equations.
-Here, we add additional details to clarify the mapping from a collection of rate functions to a fully specified process.
-Our treatment follows \citet{breto09}.
-
-A general compartment model is a vector-valued process $X(t)=(X_1(t),\dots,X_c(t))$ denoting the (integer or real-valued) counts in each of $c$ compartments, where $t$ is any continuous value in the interval  $\left[t_0, \infty\right)$ for some real valued starting time $t_0$.
-The compartments may also have names, but to set up general notation we simply refer to them by their numerical index.
-The basic characteristic of a compartment model is that $X(t)$ can be written in terms of the flows $N_{ij}(t)$ from $i$ to $j$.
-A flow into compartment $i$ from outside the system is denoted by $N_{\demography i}$, and a flow out of the system from compartment $i$ is denoted by $N_{i\demography}$.
-We call $\demography$ a source/sink compartment, though it is an irregular compartment since $X_{\demography}(t)$ is not defined.
-These flows are required to satisfy a ``conservation of mass'' identity:
-\begin{equation}
-X_i(t)=X_i(t_0)+N_{\demography i}(t) - N_{i\demography}(t) + \sum_{j\neq
-i}N_{ji}(t)-\sum_{j\neq i}N_{ij}(t). \label{eq:conservation}
-\end{equation}
-Each {\em flow} $N_{ij}(t)$ is associated with a {\em rate} function
-$\mu_{ij}=\mu_{ij}(t,X(t))$, where we include the possibility that $i$ or $j$ takes value $\demography$.
-
-There are different ways to use a collection of rate functions to build a fully specified model.
-We proceed to describe the ones we use in this paper: via a system of ordinary differential equations (Sec.~\ref{subsec:ode}), a simple Markov counting system (Sec.~\ref{subsec:smcs}), and an overdispersed Markov counting system (Sec.~\ref{subsec:odmcs}).
-Other representations include stochastic differential equations driven by Gaussian noise or Gamma noise \citep{bhadra11}.
-
-\subsection{Ordinary differential equation (ODE) interpretation}
-\label{subsec:ode}
-
-A basic deterministic specification is
-\begin{equation}
-\label{eq:ode1}
-dN_{ij}/dt = \mu_{ij}\big(t,X(t) \big) X_i(t), \hspace{3mm} i\in \seq{1}{c}, \hspace{3mm} j\in \seq{1}{c} \cup\{\demography\}, \hspace{3mm} i\neq j,
-\end{equation}
-where $\mu_{ij}\big(t,X(t)\big)$ is called a per-capita rate or a unit rate.
-Flows into the system require special treatment since $X_i(t)$ in \myeqref{eq:ode1} is not defined for $i=\demography$.
-Instead, we specify
-\begin{equation}
-\label{eq:ode2}
-dN_{\demography i}/dt = \mu_{\demography i}\big(t,X(t) \big).
-\end{equation}
-This is the the interpretation and implementation used for Model~2 in our study.
-
-\subsection{Simple Markov counting system interpretation}
-\label{subsec:smcs}
-A continuous time Markov chain can be specified via its infinitesimal transition probabilities.
-A basic approach to this is to define
-\begin{eqnarray}
-\label{eq:smcs1}
-\prob\big[ N_{ij}(t+\delta)-N_{ij}(t)=0 \given X(t)\big]
- &=& 1-\delta \mu_{ij}\big(t,X(t)\big)X_i(t) + o(\delta),
-\\
-\label{eq:smcs2}
-\prob\big[ N_{ij}(t+\delta)-N_{ij}(t)=1 \given X(t)\big]
- &=& \delta \mu_{ij}\big(t,X(t)\big)X_i(t) + o(\delta),
-\end{eqnarray}
-for $i\in \seq{1}{c}$ and $j\in\seq{1}{c}\cup\{\demography\}$ with $i\neq j$.
-As with the ODE case, we need special attention for flows into the system, and we define
-\begin{eqnarray}
-\label{eq:smcs3}
-\prob\big[ N_{\demography i}(t+\delta)-N_{\demography i}(t)=0 \given X(t)\big]
- &=& 1-\delta \mu_{\demography i}\big(t,X(t)\big) + o(\delta),
-\\
-\label{eq:smcs4}
-\prob\big[ N_{\demography i}(t+\delta)-N_{\demography i}(t)=1 \given X(t)\big]
- &=& \delta \mu_{\demography i}\big(t,X(t)\big) + o(\delta).
-\end{eqnarray}
-Together with the initial conditions $X(0)$, equations \myeqref{eq:smcs1}--\myeqref{eq:smcs4} define a Markov chain.
-Each flow is a simple counting process, meaning a non-decreasing integer-valued process that only has jumps of size one.
-We therefore call the Markov chain a simple Markov counting system (SMCS).
-The infinitesimal mean of every flow is equal to its infinitesimal variance \citep{breto11} and so an SMCS is called equidispersed.
-We note that the special case of Model~1 used by \cite{lee20} (with $\sigmaProc = 0$) is an SMCS.
-To permit more general mean-variance relationships for a Markov counting system, we must permit jumps of size greater than one.
-The utility of overdispersed models, where the infinitesimal variance of the flow exceeds the infinitesimal mean, has become widely recognized \citep{stocks20,he10}.
-
-\subsection{Overdispersed Markov counting system interpretation}
-\label{subsec:odmcs}
-
-Including white noise in the rate function enables the possibility of an overdispersed Markov counting system \citep{breto11,breto09,he10}.
-Since rates should be non-negative, Gaussian noise is not appropriate and gamma noise is a convenient option that has found various applications \citep{romero-severson15, subramanian20}.
-Specifically, we consider a model given by
-\begin{equation}
-\label{eq:odmcs1}
-\mu_{ij}\big(t,X(t)\big) = \bar\mu_{ij}\big(t,X(t)\big) \, d\Gamma_{ij}(t)/dt,
-\end{equation}
-where $\Gamma_{ij}(t)$ is a stochastic process having independent gamma distributed increments, with
-\begin{equation}
-\label{eq:odmcs2}
-\E\big[\Gamma_{ij}(t)\big] = t, \quad \var\big[\Gamma_{ij}(t)\big] = \sigma_{ij}^2 t.
-\end{equation}
-Formally interpreting the meaning of \myeqref{eq:odmcs1} is not trivial, and we do so by constructing a Markov process $X(t)$ as the limit of the Euler scheme descrived in Section~\ref{sec:numerics}, below.
-Therefore, the numerical scheme in Sec.~\ref{sec:numerics} can be taken as a definition of the meaning of \myeqref{eq:odmcs1}.
-The Markov chain defined by the limit of this Euler scheme as the step size decreases is an overdispersed Markov counting system, with the possibility of instantaneous jumps of size greater than one \citep{breto11}.
-
-\section{Numerical solutions to compartment models}
-\label{sec:numerics}
-
-Models may be fitted and their implications assessed via numerical solutions (i.e., simulations) from the model equations.
-All the analyses we consider have this simulation-based property, known as plug-and-play or equation-free or likelihood-free.
-The numerical solutions to the model are arguably of more direct scientific interest than the exact solutions to the postulated equations.
-For ODE models, numerical methods are well studied and a standard numerical solution package such as \code{deSolve} in \code{R} is adequate for many purposes.
-For SMCS and ODMCS models, exact schemes are feasible when the number of events is small, which may be the case for small populations.
-However, for applicability to larger populations, we use instead the following Euler scheme.
-Write $\delta$ for an Euler time step, and $\Delta N_{ij}$ for the numerical approximation to $N_{ij}(t+\delta)-N_{ij}(t)$ given $X(t)$.
-For each $i$ and $j$ in $\seq{1}{c} \cup \{\demography\}$ with $i \neq j$, we draw independent Gamma distributed noise increments with mean $\delta$ and variance $\sigma_{ij}^2 \delta$, denoted using a mean-variance parameterization of the gamma distribution as
-\begin{equation}
-\label{eq:numerics1}
-\Delta\Gamma_{ij} \sim \mathrm{gamma}(\delta, \sigma_{ij}^2 \delta).
-\end{equation}
-In the case of an SMCS model, $\sigma_{ij}=0$ for all $i$ and $j$, so we have $\Delta\Gamma_{ij}=\delta$.
-Then, for $i\neq \demography$ and $j\neq i$, and writing
-\begin{equation}
-\label{eq:numerics2}
-\mu_{ij}=\bar\mu_{ij}\big(t,X(t)\big) \Delta\Gamma_{ij} / \delta,
-\end{equation}
-we calculate transition probabilities
-\begin{eqnarray}
-\label{eq:numerics3}
-p_{ij} &=& \exp\left\{-\sum_{k\in 1:c \, \cup \{\demography\}} \mu_{ik} \, \delta \right\}
-\frac{\mu_{ij}}{\sum_{k\in 1:c\cup \{\demography\}} \mu_{ik}},
-\\
-\label{eq:numerics4}
-p_{ii} &=& 1 - \sum_{j\neq i} p_{ij}.
-\end{eqnarray}
-These probabilities correspond to competing hazards for every individual in compartment $i$ to transition to some compartment $j$, interpreting $j=i$ to mean that the individual remains in $i$.
-Then, $\big(\Delta N_{i1},\dots,\Delta N_{ic},\Delta N_{i\demography}\big)$ has the multinomial distribution where $X_i(t)$ individuals are allocated independently to $\seq{1}{c}\cup\{\demography\}$ with probabilities given by \eqref{eq:numerics3} and \eqref{eq:numerics4}.
-We use the \code{reulermultinom} function in the \code{pomp} package to draw from this multionomial distribution.
-
-Different treatments of demographic flows---such as birth, death, immigration and emigration---are possible.
-For the case $i=\demography$, the treatment used by Model~1 is to set
-\begin{equation}
-\label{eq:numerics6}
-\Delta N_{\demography j} \sim \mathrm{poisson}( \mu_{\demography j} \delta),
-\end{equation}
-an independent Poisson random variable with mean $\mu_{\demography j} \delta$.
-
-Models~2 and 3 used an alternative approach, balancing the total number of flows in and out of the compartment, i.e., $\sum_{i}N_{\demography i}(t) = \sum_{i}N_{i \demography}(t)$, in order to make the model consistent with the known total population.
-In this case, we formally model the death rate as a rate of returning to the susceptible class $S$, and use external transitions from $\demography$ into $S$ to describe only net population increase.
-
-\section{Measurement Models}
-
-Each POMP requires specification of a measurement model, which is a statistical description of how observations on the system are obtained.
-In general, we used the same measurement models that were reported by \cite{lee20} unless specifically noted in the following subsections.
-
-\subsection{Model~1}
-
-In this model, the advantage afforded by vaccination is an increased probability that an infection is asymptomatic.
-Therefore, under the assumptions of this model, all reported cases are assumed to be a fraction of individuals that transition from the exposed to the infected compartment, as noted in Eq.~\myeqref{model1:meas}:
-\begin{equation}
-  \label{model1:meas}
-  Y_{n} \mid \Delta N_{E_{\cdot}I_{\cdot}}(n) = \reportChange \sim \text{NB}\left(\reportRate \reportChange, \obsOverdispersion \right),
-\end{equation}
-where $Y_n$ is the reported cholera cases at time $t_n \in t_1:t_N$ and $\Delta N_{E_{\cdot}I_{\cdot}}(n)$ is the sum total of individuals across each vaccination compartment $\vaccCounter \in 1:\vaccClass$ who moved from compartment $E_\vaccCounter$ to $I_\vaccCounter$ since observation $t_{n-1}$.
-Here, $\text{NB}\left(\reportRate \reportChange, \obsOverdispersion \right)$ denotes a negative binomial distribution with mean $\reportRate \reportChange$ and variance $\reportRate \reportChange \big(1 + \frac{\reportRate \reportChange}{\obsOverdispersion}\big)$.
-
-\subsection{Model~2}\label{sec:mod2Meas}
-
-Model~2 was fit using reported case counts that were transformed using the natural logarithm.
-We fit Model~2 using the subplex algorithm in the \code{subplex} package, using
-a Gaussian measurement model (Eq.~\myeqref{model2:meas}) on the log transformed cases within each unit.
-The final loss function that is maximized is the product of the likelihoods of the individual units, or the sum of the log-likelihood of the individual units.
-The measurement model for individual units is given in Eq.~\myeqref{model2:meas}.
-\begin{equation}
-  \label{model2:meas}
-  \log\big(Y_{u, n} + 1\big) \mid \Delta N_{E_{u\cdot}I_{u\cdot}}(n) = \reportChange_u \sim \text{N}\left(\log \big(\reportRate \reportChange_u + 1\big), \obsOverdispersion^2 \right),
-\end{equation}
-where $\Delta N_{E_{u\cdot}I_{u\cdot}}(n)$ is the sum total of individuals across vaccination compartment $\vaccCounter \in 0:4$ within unit $u$ who moved from compartment $E_{u\vaccCounter}$ to $I_{u\vaccCounter}$ since observation $t_{n-1}$.
-Therefore, because the natural logarithm of observed case counts (plus one, to avoid taking the logarithm of zero) has a normal distribution, $Y_{u, n} + 1$ is assumed to follow a log-normal distribution with log-mean parameter $\log\big(\reportRate \Delta N_{E_{u\cdot}I_{u\cdot}}(n) + 1\big)$ and log-variance $\obsOverdispersion^2$.
-We note that fitting a model with this measurement model is equivalent to fitting using least squares, with $\log(Y_{u, n} + 1)$ as the response variable.
-
-This measurement model differs from that used by \citet{lee20}, who fit the model in two stages: epidemic and endemic phases.
-Although their text and supplement material do not explicitly describe the measurement model used, inspection of the code provided with their submission suggests a change in measurement model between the epidemic and endemic phases.
-% In the file \code{choleraEqsPublic.py}, \citet{lee20} create several functions, where each function represents a set of coupled differential equations that could potentially be used to model cholera incidence data.
-% Each function returns a vector (or \code{numpy} array) that represents the change in each state variable for a single time step, including the variable \code{dC}, which tracks the number of new infections and is used to obtain the reported case counts.
-% Following their comments and code, it appears that the function \code{choleraEqs10WithoutVaccinationNetwork} was used to describe the dynamics of the epidemic phase, and \code{choleraEqs11WithoutVaccinationNetwork} was used for the endemic stage (see Sec~\ref{sec:mod2rep} for more details).
-% Because their models were fit using least squares, the code in these functions suggest that the measurement model for the epidemic phase is
-The measurement model they used for the epidemic phase is
-\begin{equation}
-  \label{model2:measEpi}
-  Y_{u, n} \mid \Delta N_{E_{u\cdot}I_{u\cdot}}(n) = \reportChange_u \sim \text{N}\left(\reportRate \reportChange_u, \obsOverdispersion^2 \right),
-\end{equation}
-% which is similar to our measurement model, the primary difference being that the measurement model is applied to raw case counts rather than log-transformed case counts.
-The measurement model they used for the endemic phase modifies the epidemic model by counting both asymptomatically infected (A) and symptomatically infected (I) individuals in the case counts:
-\begin{equation}
-  \label{model2:measEnd}
-  Y_{u, n} \mid \Delta N_{E_{u\cdot}I_{u\cdot}}(n) = \reportChange_{u1}, \Delta N_{E_{u\cdot}A_{u\cdot}}(n) = \reportChange_{u2} \sim \text{N}\left(\reportRate \big(\reportChange_{u1} + \reportChange_{u2}\big), \obsOverdispersion^2 \right),
-\end{equation}
-where the notation for $\Delta N_{E_{u\cdot}A_{u\cdot}}(n)$ is similar to $\Delta N_{E_{u\cdot}I_{u\cdot}}(n)$, described above.
-
-\subsection{Model~3}
-
-In this model, reported cholera cases are assumed to stem from individuals who develop symptoms and seek healthcare.
-Therefore reported cases are assumed to come from an over-dispersed negative binomial model, given the increase in infected individuals:
-\begin{equation}
-  \label{model3:meas}
-  Y_{u, n} \mid \Delta N_{S_{u \cdot}I_{u}}(t) = \reportChange_u \sim \text{NB}\left(\reportRate \reportChange_{u}, \obsOverdispersion \right),
-\end{equation}
-where $\Delta N_{S_{u \cdot}I_{u \vaccCounter}}(n)$ is the number of individuals who moved from compartment $S_{u \vaccCounter}$ to $I_{u}$ since observation $t_{n-1}$.
-
-This measurement model is a minor change from that used by \cite{lee20}, which allowed for a change in the reporting rate on January 1st, 2018.
-The fitted values of the reporting rate---before and after January 2018---were $0.97$ and $0.097$, respectively.
-An instantaneous change from near perfect to almost non-existent reporting can be problematic, as it forces the model to explain the observed reduction in reported cases as a decrease in the reporting of cases, rather than a decrease in the prevalence of cholera.
-This shift was justified by a ``change of the case definition that occurred on January 1st, 2018";
-this claim was not cited, and we could find no evidence that such a drastic change in the reporting rate would be warranted.
-We therefore do not allow a change in reporting rate when fitting Model~3.
-
 \section{Initial Values}
@@
 The resulting parameter estimates are used to obtain the remaining non-zero initial values of the latent states using Eqs.~\myeqref{eq:model3:A0}--\myeqref{eq:model3:B0}.
-
-% Argue why we didn't need to do this for Model 2.
-
-% TODO: Check this section.
-
-\section{Calibrating Model~3 to observed cases}\label{sec:mod3Cal}
-
-<<CalibrateOrig3>>=
-# Create vectors for the unit and shared parameters
-unit_specific_names <- c("betaB", "foi_add")
-shared_param_names <- c(
-  "mu_B", "XthetaA", "thetaI", "lambdaR", "r", "std_W",
-  "epsilon", "k"
-)
-est_param_names <- c(
-  unit_specific_names, shared_param_names
-)
-
-# Add unit numbers to each parameter
-est_param_names_expanded <- paste0(rep(est_param_names, each = 10), 1:10)
-
-# Simple function that is used to set the rw_sd for the first search
-# This function is convenient so that we don't have to type out the rw_sd
-# for each unit parameter.
-set_rw_1 <- function(x) {
-  if (gsub("[[:digit:]]+$", "", x) %in% shared_param_names) {
-    return(0.01)
-  # } else if (x %in% paste0(rep(c("aHur", 'hHur'), each = 2), c(3, 9))) {
-  #   return(expression(ifelse(time >= 2016.754 & time <= 2017, 0.015, 0)))
-  # } else if (x %in% c("Iinit3", "Iinit4")) {
-  #   return(expression(ivp(0.25)))
-  } else if (grepl('^foi_add[[:digit:]]+$', x)) {
-    return(0.015)
-  } else if (grepl('^betaB[[:digit:]]+$', x)) {
-    return(0.01)
-  } else {
-    return(0)
-  }
-}
-
-set_rw_2 <- function(x) {
-  if (gsub("[[:digit:]]+$", "", x) %in% shared_param_names) {
-    return(0.003)
-  # } else if (x %in% paste0(rep(c("aHur", 'hHur'), each = 2), c(3, 9))) {
-  #   return(expression(ifelse(time >= 2016.754 & time <= 2017, 0.005, 0)))
-  # } else if (x %in% c("Iinit3", "Iinit4")) {
-  #   return(expression(ivp(0.125)))
-  } else if (grepl('^foi_add[[:digit:]]+$', x)) {
-    return(0.005)
-  } else if (grepl('^betaB[[:digit:]]+$', x)) {
-    return(0.003)
-  } else {
-    return(0)
-  }
-}
-
-# First Search RW
-reg_rw_1.sd <- lapply(est_param_names_expanded, set_rw_1)
-names(reg_rw_1.sd) <- est_param_names_expanded
-chol_rw_1.sd <- do.call(rw_sd, reg_rw_1.sd)
-
-# Second Search RW
-reg_rw_2.sd <- lapply(est_param_names_expanded, set_rw_2)
-names(reg_rw_2.sd) <- est_param_names_expanded
-chol_rw_2.sd <- do.call(rw_sd, reg_rw_2.sd)
-
-SEARCH1 = list(
-  NBPF = 100,
-  NP = 2500,
-  SPAT_REGRESSION = 0.05,
-  NREPS = 144,
-  NP_EVAL = 2500,
-  NREPS_EVAL = 18,
-  RW_SD = chol_rw_1.sd,
-  COOLING = 0.5,
-  KEEP_TRACES = FALSE,
-  KEEP_LIKE_MAT = FALSE
-)
-
-SEARCH2 = list(
-  TOP_N = 12,
-  NBPF = 100,
-  NP = 2500,
-  SPAT_REGRESSION = 0.05,
-  NREPS = 6,
-  NP_EVAL = 2500,
-  NREPS_EVAL = 18,
-  RW_SD = chol_rw_2.sd,
-  COOLING = 0.5,
-  KEEP_TRACES = FALSE,
-  KEEP_LIK_MAT = TRUE
-)
-
-haiti3_sub_fit <- bake(
-  file = "../model3/si/subset_noparams.rds", {
-    fit_haiti3(
-      nsearches = 2L,
-      search1 = SEARCH1,
-      search2 = SEARCH2,
-      search_rho = FALSE,
-      search_gamma = FALSE,
-      search_Iinit = FALSE,
-      search_hur = FALSE,
-      ncores = cores
-    )
-  }
-)
-
-# Find model index with best log-likelihood
-best_m <- which.max(haiti3_sub_fit$search2$logLiks$logLik)
-
-# Find parameters using best_m index
-sub_params <- haiti3_sub_fit$search2$params[best_m, ]
-
-# Get unit log-likelihoods for the best model.
-all_liks <- as.data.frame(haiti3_sub_fit$search2$likMat)
-# all_liks$which <- rep(1:(SEARCH3$TOP_N * SEARCH3$NREPS), each = SEARCH3$NREPS_EVAL)
-all_liks$which <- rep(1:(12 * 6), each = 18)
-best_mod <- filter(all_liks, which == best_m)
-
-# Pivot longer to get estimate of log-likelihood for each unit.
-unitLikesM3 <- best_mod %>%
-  pivot_longer(
-    cols = -which,
-    names_to = "unit",
-    names_prefix = "V",
-    values_to = "logLik"
-  ) %>%
-  select(-which) %>%
-  group_by(unit) %>%
-  summarize(
-    ll_est = logmeanexp(logLik),
-    ll_se = logmeanexp(logLik, se = TRUE)[2]
-  ) %>%
-  ungroup() %>%
-  mutate(unit = as.numeric(unit)) %>%
-  arrange(unit)
-@
-
-<<M3_bench, echo=FALSE, message=FALSE, include=FALSE, cache=TRUE>>=
-h3 <- haiti3_spatPomp()
-deps <- h3@unit_names
-
-mod3_cases <- as.data.frame(t(h3@data))
-colnames(mod3_cases) <- deps
-m3_inits <- c(7362, 1914, 22, 8, 2818, 82, 2499, 6953, 44, 7) # TODO: Also fit by hand, but could be changed to initializing to the observation before.
-names(m3_inits) <- deps
-
-bench_cond_lls <- list()
-for (dep in deps) {
-
-  # Get department cases
-  cases <- as.numeric(mod3_cases[, dep])
-  mod <- ar1_NegBinom(data = cases, init = m3_inits[dep])
-
-  bench_cond_lls[[dep]] <- haitipkg:::.ar1_NegBinom_cond_ll(
-    theta = mod$theta, init = m3_inits[dep], data = cases, transform = FALSE
-  )
-}
-
-bench_cond_lls <- as.data.frame(bench_cond_lls)
-
-bench_lls <- apply(bench_cond_lls, 2, sum)
-
-unitLikes_bench <- data.frame(
-  unit = 1:10,
-  # dep = names(bench_lls),
-  ll = bench_lls
-) %>%
-  as_tibble()
-
-bench_cond_lls$date <- lubridate::round_date(lubridate::date_decimal(h3@times), unit = 'day')
-@
-
-<<combine_bench_M3, echo=FALSE>>=
-
-# Combine tables with model 3 and benchmark likelihoods.
-bench_M3_joined <- dplyr::left_join(
-  x = unitLikesM3,
-  y = unitLikes_bench,
-  by = "unit"
-) %>%
-  select(-ll_se) %>%
-  mutate(unit = as.character(unit)) %>%
-  pivot_longer(
-    cols = -unit,
-    names_to = "type",
-    values_to = "ll"
-  ) %>%
-  mutate(unit = paste0(unit, ": ", deps[as.numeric(unit)]))
-@
-
-<<M3bpf, echo=FALSE, message=FALSE, include=FALSE>>=
-
-# Change model parameters to the best set of parameters from previous
-# search
-coef(h3)[names(sub_params)] <- sub_params
-
-registerDoRNG(40765101)  # for modular, reproducible results
-
-# Perform block partical filter with fitted model, to get conditional
-# log-likelihoods.
-bpf_sub3_out <- bake(
-  file = "../model3/si/bpf_sub_cond_LL.rds", {
-
-    results <- list()
-
-    bpf_out <- h3 %>% bpfilter(
-      Np = 5000,
-      block_size = 1,
-      save_states = TRUE
-    )
-
-    results$block.cond.loglik <- bpf_out@block.cond.loglik
-    results$times <- bpf_out@times  # Saved for convenience
-    results
-  },
-  seed = 432235
-)
-@
-
-In this section, we provide more detail on the process that was used to estimate the coefficients used in Model~3.
-In particular, we discuss why we decided to include additional model parameters---those that are associated with the behavior of the system during Hurricane Matthew---that were not considered by \cite{lee20}.
-To calibrate this model, we used the iterated block particle filter (IBPF) method of \cite{ionides22}.
-Due to the novelty of this algorithm, there does not exist published examples the IBPF algorithm used for data analysis outside of the papers in which the algorithm was introduced \citep{ning21ibpf,ionides22}, which at least partially motivates the inclusion additional details in this section.
-
-\cite{lee20} were only able to estimate model parameters to a simplified version of Model~3 on a subset of the available data, as no method existed at the time of their publication to fit a fully coupled meta-population model to disease incidence data.
-Building on their results, we fit the fully coupled version of Model~3 to nearly all available data, reserving only a few observations to use to calibrate the initial conditions of the model (see the previous section for more details).
-In order to properly maximize the model likelihood, we found that it was necessary to use multiple searches for the MLE, periodically pruning away less successful searches.
-The first collection of searches was performed by obtaining initial values for the parameters by uniformly sampling values from a predefined hypercube.
-A subsequent refinement search used starting values corresponding to the largest model likelihoods obtained in the previous round.
-This is similar to the approach used by \cite{ionides22}.
-
-We use the technique described above to fit the fully coupled version of Model~3 proposed by \cite{lee20}.
-The maximum likelihood we obtained with the model was $\Sexpr{round(max(haiti3_sub_fit$search2$logLiks$logLik))}$, which is higher than the benchmark model ($\Sexpr{round(sum(unitLikes_bench$ll))}$).
-While beating a simple associative benchmark is promising, this does not immediately rule out the possibility that the model is a poor description of the system.
-Additional investigation of parameters estimates and their corresponding implications on model based conclusions should always be conducted.
-For meta-population models, it is worth considering how well the model fits the data to each spatial unit.
-This can be done by looking at conditional log-likelihoods, which is part of the output of the \texttt{bpfilter} algorithm in the \texttt{spatPomp} package.
-The likelihoods for each department, compared to the corresponding benchmark model, are displayed in Fig.~\ref{fig:h3UnitLikes}.
-The figure demonstrates that while the log-likelihood of the fitted model outperforms the auto-regressive negative binomial benchmark model, Model~3 has lower likelihoods than the benchmark for some departments.
-
-\begin{figure}[ht]
-<<h3UnitLikes, fig.height=3, fig.width=5.5, echo=FALSE>>=
-ggplot(bench_M3_joined, aes(x = ll, y = reorder(unit, ll))) +
-  geom_point(aes(col = type)) +
-  ylab("Unit") +
-  xlab("Log Likelihood") +
-  scale_color_manual(
-    labels = c("ll" = "Benchmark", "ll_est" = "Model 3"),
-    values = c("ll" = "#1b9e77", "ll_est" = "#d95f02")
-  ) +
-  theme(legend.title = element_blank(),
-        axis.title.y = element_blank())
-@
-\caption{\label{fig:h3UnitLikes}Log-likelihoods of Model~3 for each department compared to the corresponding benchmark model prior to the inclusion of parameters for initial values and Hurricane Matthew.}
-\end{figure}
-
-In addition to considering the conditional log-likelihoods of each unit, one can consider conditional log-likelihoods of each observation.
-When compared to a benchmark, this level of detail can provide useful information about which observations are well described by the model and which are poorly described.
-In Fig.~\ref{fig:condLL}, we plot the conditional log-likelihoods of each observation.
-This figure reveals that the fitted model has difficult time explaining certain features of the data.
-For example, many departments (in particular Sud) have observations with lower conditional log-likelihoods near October 2016 than at other time points.
-Further investigation of the model output reveals that the model is struggling to explain the sudden surge in cholera cases that occurred at this time, which coincides with the time that Hurricane Matthew struck Haiti.
-While the model does include a mechanism to account for increased cholera transmission due to large rainfall events, the mechanism does not appear to be sufficient to capture the damaging effects of the hurricane, which had the greatest impact in the the Sud and Grand'Anse departments \citep{ferreirai16}.
-This result led us to include parameters $\Whur{u}$ and $\hHur{u}$ in Eq.~23 of the main text, which allows for an increase in the transmission rate between environmental cholera and humans for Sud and Grand'Anse during and after the hurricane.
-The effect of the hurricane on cholera transmission is assumed to have an exponential decay, where the magnitude is determined by $\Whur{u}$ and the duration of the effect determined by $\hHur{u}$.
-
-\begin{figure}[ht]
-<<CondLLFig, echo=FALSE, message=FALSE, fig.height=3>>=
-
-# Changing labels to human-readable labels.
-deps[grepl("_", deps)] <- gsub("_", "-", deps[grepl("_", deps)])
-deps[deps == "Grande-Anse"] <- "Grand\'Anse"
-names(deps) <- as.character(1:10)
-myLabeller <- as_labeller(deps)
-
-# Plot conditional log-likelihoods for each unit
-t(bpf_sub3_out$block.cond.loglik) %>%
-  as.data.frame() %>%
-  mutate(t = bpf_sub3_out$times) %>%
-  pivot_longer(
-    cols = -t,
-    values_to = 'likelihood',
-    names_prefix = "V",
-    names_to = 'Unit'
-  ) %>%
-  mutate(date = as.Date(lubridate::date_decimal(t))) %>%
-  ggplot(aes(x = date, y = likelihood)) +
-  geom_line() +
-  geom_point() +
-  facet_wrap(~Unit, labeller = myLabeller, scales = 'free_y',
-             nrow = 2) +
-  theme_bw() +
-  theme(axis.title.x = element_blank(),
-        axis.text.x = element_text(angle = -45, hjust = 0)) +
-  ylab("Conditional log-likelihoods")
-@
-\caption{\label{fig:condLL}Conditional log-likelihoods of Model~3 prior to the inclusion of parameters for initial values and Hurricane Matthew.}
-\end{figure}
-
-Other points of concern include the observations at or near the start of the recorded cases, for example, in Nord-Ouest and Sud-Est.
-These low conditional log-likelihoods are a result of having zero or near zero confirmed cases in the first week of observed data, followed by a rapid growth of cases.
-The model cannot explain the rapid growth of cholera in the absence of starting cases because the initialization scheme forces departments with no recorded cases in the first week to have hidden states $I_{u0}(0) = A_{u0}(0) = B_{u}(0) = 0$.
-In reality, however, it is possible to have some infected or asymptomatic individuals in a department when no cases are recorded due to under-reporting.
-This may be particularly true at the onset of an infectious disease outbreak, as lower public awareness and inadequate testing infrastructure may result in many cases going undetected.
-We therefore include and estimate $I_{0u}$ that determines the initial number of Infected individuals at the start of the pandemic for the departments Grande'Anse, Nord-Ouest, and Sud-Est;
-each of these departments have both low starting case counts followed by rapid growth, and low initial values of the conditinoal log-likelihood.
-
-<<loadFinalModel3, echo=FALSE, include=FALSE, message=FALSE>>=
-
-# This result was used in the manuscript, so it doesn't need to be recalculated.
-h3_final_results <- readRDS("../model3/run_level_3/haiti3_fit.rds")
-
-# Save best_m as index for the model parameters with the best likelihood.
-# Then, save the model parameter values and the likelihood evaluations
-best_m <- which.max(h3_final_results$search2$logLiks$logLik)
-best_params <- h3_final_results$search2$params[best_m, ]
-all_liks <- as.data.frame(h3_final_results$search2$likMat)
-all_liks$which <- rep(1:72, each = 18)
-best_mod <- dplyr::filter(all_liks, which == best_m)
-
-# Get repeated estimates of unit log-likelihoods
-unitLikesM3 <- best_mod %>%
-  pivot_longer(
-    cols = -which,
-    names_to = "unit",
-    names_prefix = "V",
-    values_to = "logLik"
-  ) %>%
-  select(-which) %>%
-  group_by(unit) %>%
-  summarize(
-    ll_est = pomp::logmeanexp(logLik),
-    ll_se = pomp::logmeanexp(logLik, se = TRUE)[2]
-  ) %>%
-  ungroup() %>%
-  mutate(unit = as.numeric(unit)) %>%
-  arrange(unit)
-
-# Combine model likelihood table to benchmarks
-bench_M3_joined <- dplyr::left_join(
-  x = unitLikesM3,
-  y = unitLikes_bench,
-  by = "unit"
-) %>%
-  select(-ll_se) %>%
-  mutate(unit = as.character(unit)) %>%
-  pivot_longer(
-    cols = -unit,
-    names_to = "type",
-    values_to = "ll"
-  ) %>%
-  mutate(unit = paste0(unit, ": ", deps[as.numeric(unit)]))
-@
-
-We refit Model~3 after introducing these seven parameters---three initial value parameters and four parameters related to Hurricane Matthew---using the same successive search scheme described above.
-The resulting model has a log-likelihood value of $\Sexpr{round(max(h3_final_results$search2$logLiks$logLik), 1)}$.
-We also provide figures for department specific log-likelihoods and conditional log-likelihoods in Figs.~\ref{fig:finalUnitLL} and \ref{fig:finalCondLL}.
-
-\begin{figure}[ht]
-<<finalUnitLLFig, fig.height=3, fig.width=5.5, echo=FALSE>>=
-ggplot(bench_M3_joined, aes(x = ll, y = reorder(unit, ll))) +
-  geom_point(aes(col = type)) +
-  ylab("Unit") +
-  xlab("Log Likelihood") +
-  scale_color_manual(
-    labels = c("ll" = "Benchmark", "ll_est" = "Model 3"),
-    values = c("ll" = "#1b9e77", "ll_est" = "#d95f02")
-  ) +
-  theme(legend.title = element_blank(),
-        axis.title.y = element_blank())
-@
-\caption{\label{fig:finalUnitLL}Log-likelihoods of Model~3 for each department compared to the corresponding benchmark model after adding and estimating parameters related to initial values and Hurricane Matthew.}
-\end{figure}
-
-<<finalModel3BPF, echo=FALSE, message=FALSE, include=FALSE>>=
-
-# Set the model parameters to the best results from previous search
-coef(h3)[names(best_params)] <- best_params
-
-# block particle filter to get conditional log-likelihoods.
-bpf_final3_out <- bake(
-  file = "../model3/si/h3_final_bpf_cond_ll.rds", {
-
-    results <- list()
-
-    bpf_out <- h3 %>% bpfilter(
-      Np = 5000,
-      block_size = 1,
-      save_states = TRUE
-    )
-
-    results$Sus_df <- foreach(i=1:length(h3@times), .combine = bind_rows) %do% {
-      get_h3_S_quants(bpf_out, i)
-    }
-
-    results$block.cond.loglik <- bpf_out@block.cond.loglik
-    results$times <- bpf_out@times  # saved for convenience
-    results
-  },
-  seed = 786321
-)
-@
-
-\begin{figure}[ht]
-<<finalCondLLFig, echo=FALSE, message=FALSE, fig.height=3>>=
-t(bpf_final3_out$block.cond.loglik) %>%
-  as.data.frame() %>%
-  mutate(t = bpf_final3_out$times) %>%
-  pivot_longer(
-    cols = -t,
-    values_to = 'likelihood',
-    names_prefix = "V",
-    names_to = 'Unit'
-  ) %>%
-  mutate(date = as.Date(lubridate::date_decimal(t))) %>%
-  ggplot(aes(x = date, y = likelihood)) +
-  geom_line() +
-  geom_point() +
-  facet_wrap(~Unit, labeller = myLabeller, scales = 'free_y',
-             nrow = 2) +
-  theme_bw() +
-  theme(axis.title.x = element_blank(),
-        axis.text.x = element_text(angle = -45, hjust = 0)) +
-  ylab("Conditional log-likelihoods")
-@
-\caption{\label{fig:finalCondLL}Conditional log-likelihoods of Model~3 after adding and estimating the parameters for initial values and Hurricane Matthew.}
-\end{figure}
-
-While the addition of the Hurricane parameters seems to have resulted in a dramatic increase of conditional likelihoods around and after October 2016, the inclusion of the initial value parameters does not appear to have fixed the low conditional log-likelihood values in Sud-Est and Nord-Ouest.
-This could be due to many factors, including the possibility that the coupling between units forces low initial value parameters in a given department in order for the model to fit the data in the remaining departments.
-Whatever the case may be, the inclusion of the seven parameters resulted in a $\Sexpr{round(max(h3_final_results$search2$logLiks$logLik) - max(haiti3_sub_fit$search2$logLiks$logLik), 1)}$ increase in log-likelihood, which is a highly statistically significant difference.
-One could similarly test whether the inclusion of the initial value parameters resulted in a significantly better fit to the observed data, but we do not do this here due to the large computational effort that would be required.
-
-\subsection{Examining the Hidden States of the Calibrated Model}
-
-For mechanistic models, beating a suitable statistical benchmark does not alone guarantee that the model provides an accurate description of a dynamic process.
-Indeed, a good statistical fit does not require the model to assert a causal explanation.
-For example, reconstructed latent variables should make sense in the context of alternative measurements of these variables \citep{grad12}.
-We demonstrate this principle by examining the latent state of the calibrated model.
-In particular, we examine the compartment of susceptible individuals under various scenarios, as this compartment drives transmission dynamics in SIR models for infectious disease outbreaks.
-
-Recall that the filtering distribution for the calibrated version of Model~3 at time $t_k$ is defined as the distribution of the latent state at time $t_k$ given the data from times $t_{1}:t_{k}$, i.e.
-$f^{(3)}_{\bm{X}_k|\bm{Y}_{1:k}}(\bm{x}_{k} | \bm{y}^*_{1:k} ; \hat\theta)$.
-In general, one may expect simulations from the filtering distribution of a model with a good statistical fit to result in hidden states that are highly consistent with the observed data because the filtering distribution is conditioned on the observed data.
-Fig.~\ref{fig:h3Sus} shows the percentage of individuals that are in the susceptible compartment from various simulations of the model:
-simulations from Model~3 under initial conditions are displayed in red; simulations from the filtering distribution of model are displayed in blue.
-This figure shows that simulations from initial conditions tends to result in a much more rapid depletion of susceptible individuals at the start of the epidemic than simulations from the filtering distribution, suggesting the calibrated model has a propensity to predict larger outbreaks than what is typically seen in the data.
-This also appears true in later stages of the time period of observed data, where the data suggests that there is a replenishment of the susceptible compartment, but the Model simulations typically retain a smaller proportion of susceptible individuals.
-This result demonstrates that the calibrated model favors a more rapid growth in cholera cases than what is typically seen in the observed data, which explains why the model fails to predict the absence of cases between February, 2019 and October, 2022.
-
-<<h3SusFigSetup, echo=FALSE, message=FALSE, include=FALSE>>=
-
-# This bake perfoms the following:
-#  (1) simulates the model from initial conditions
-#  (2) get the values of the hidden states at the end of each simulation, and
-#      saves these for later use.
-#  (3) Obtain quantiles for the Susceptible pool from the simulations.
-init_conditions_results <- bake(
-  file = "../model3/si/InitSimsQuants.rds", {
-
-    results <- list()  # To return final output of bake
-
-    # Simulate from the fitted model
-    h3Spat_sims <- simulate(
-      h3, nsim = 2000, format = 'data.frame'
-    )
-
-    # Get states at the end of the simulations
-    h3_spat_end <- h3Spat_sims %>%
-      filter(time == max(time))
-
-    # Save hidden state values at the end of the simulation. Saved as a matrix
-    # in the same format as the output from a block-particle filter, so that
-    # it can be used as the input to project_from_filter.
-    results[["end_conditions"]] <- h3_spat_end %>%
-      select(-time, -cases) %>%
-      mutate(
-        unit = case_when(
-          unitname == "Artibonite" ~ 1,
-          unitname == "Centre" ~ 2,
-          unitname == "Grande_Anse" ~ 3,
-          unitname == "Nippes" ~ 4,
-          unitname == "Nord" ~ 5,
-          unitname == "Nord_Est" ~ 6,
-          unitname == "Nord_Ouest" ~ 7,
-          unitname == "Ouest" ~ 8,
-          unitname == "Sud" ~ 9,
-          unitname == "Sud_Est" ~ 10
-        )
-      ) %>%
-      select(-unitname) %>%
-      pivot_wider(
-        data = .,
-        names_from = unit,
-        values_from = c(S, I, A, R_one, R_two, R_three, VSd, VR1d, VR2d,
-                        VR3d, VSdd, VR1dd, VR2dd, VR3dd, VSd_alt, VR1d_alt,
-                        VR2d_alt, VR3d_alt, VSdd_alt, VR1dd_alt, VR2dd_alt,
-                        VR3dd_alt, C, B, Doses, totInc),
-        names_sep = ""
-      ) %>%
-      select(-.id) %>%
-      as.matrix() %>%
-      t()
-
-    # Get the quantiles for the Susceptible pool from the simulations
-    results[["SQuants"]] <- h3Spat_sims %>%
-      rename(dep = unitname) %>%
-      group_by(dep, time) %>%
-      summarise(
-        Q025 = quantile(S + VSd + VSdd + VSd_alt + VSdd_alt, probs = 0.025),
-        Q50  = quantile(S + VSd + VSdd + VSd_alt + VSdd_alt, probs = 0.5),
-        Q975 = quantile(S + VSd + VSdd + VSd_alt + VSdd_alt, probs = 0.975)
-      ) %>%
-      ungroup() %>%
-      rowwise() %>%
-      mutate(
-        unit_num = which(h3@unit_names == dep),
-        date = lubridate::date_decimal(time),
-        Q025 = Q025 / h3@params[paste0("H", unit_num)],
-        Q50  = Q50  / h3@params[paste0("H", unit_num)],
-        Q975 = Q975 / h3@params[paste0("H", unit_num)]
-      ) %>%
-      filter(
-        date >= as.Date(
-          lubridate::round_date(lubridate::date_decimal(h3@t0), unit = 'day')
-        )
-      ) %>%
-      mutate(date = as.Date(lubridate::round_date(date, unit = 'day')))
-
-    results
-  },
-  seed = 834687
-)
-
-# Get the output of the "bake" call for plotting.
-h3Spat_quants <- init_conditions_results[["SQuants"]]
-
-# This bake call uses the ending hidden states from above to forecast cholera
-# cases in the future under the V0 vaccination scenario
-mod3_V0_initcond_res <- bake(
-  file = '../model3/si/vacc_sim_from_init.rds', {
-    h3_V0_res <- list()
-
-    # Get "noVacc" scenario parameters
-    h3_S0_par <- get_model3_vacc_scenario_params(scenario = "noVacc")
-    h3Spat <- h3
-    p3 <- best_params
-    coef(h3Spat) <- p3
-    coef(h3Spat)[names(h3_S0_par)] <- h3_S0_par
-
-    # store the end states that will be used as initial values
-    h3_bpf_end_states <- init_conditions_results$end_conditions
-
-    # Using parameters and end state values, obtain projections
-    h3_V0_sims <- project_from_filter(
-      h3Spat, end_states = h3_bpf_end_states, covarGen = project_rain,
-      nsims = 1000, seed = 184697
-    )
-
-    # Store nationally aggregated results
-    h3_V0_res$mod3_V0_sims <- agg_mod3_sims(h3_V0_sims)
-    h3_V0_res$mod3_V0_probs <- get_elimProbs(h3_V0_sims, model = 3)
-
-    # Get quantiles of observed cases for each department and time
-    h3_V0_res$UnitQuants <- h3_V0_sims %>%
-      mutate(
-        date = as.Date(  # Get date object from decimal time
-          lubridate::round_date(lubridate::date_decimal(time), unit = 'day')
-        )
-      ) %>%
-      group_by(unitname, date) %>%
-      summarize(
-        Q025 = quantile(cases, probs = 0.025),
-        Q50 = quantile(cases, probs = 0.5),
-        Q975 = quantile(cases, probs = 0.975)
-      ) %>%
-      ungroup() %>%
-      rename(dep = unitname)
-
-    # Get quantiles of susceptible individuals for each department and time
-    h3_V0_res$SQuants <- h3_V0_sims %>%
-      mutate(
-        date = as.Date(
-          lubridate::round_date(lubridate::date_decimal(time), unit = 'day')
-        )
-      ) %>%
-      group_by(unitname, date) %>%
-      mutate(
-        # Get the population for each unit
-        pop = h3@params[paste0("H", which(first(unitname) == unit_names(h3)))]
-      ) %>%
-      summarize(  # Calculate the quantiles
-        Q025 = quantile(
-          (S + VSd + VSdd + VSdd_alt + VSd_alt) / pop,
-          probs = 0.025
-        ),
-        Q50 = quantile(
-          (S + VSd + VSdd + VSdd_alt + VSd_alt) / pop,
-          probs = 0.5
-        ),
-        Q975 = quantile(
-          (S + VSd + VSdd + VSdd_alt + VSd_alt) / pop,
-          probs = 0.975
-        )
-      ) %>%
-      ungroup() %>%
-      rename(dep = unitname)
-
-    h3_V0_res
-  }
-)
-
-# This result was already used in the manuscript; it doesn't need to
-# be recalculated.
-mod3_V0_res <- readRDS("../model3/run_level_3/mod3_V0_res.rds")
-@
-
-\begin{figure}[ht]
-<<h3SusFig, echo=FALSE, message=FALSE, fig.height=3>>=
-ggplot() +
-  # Sims from initial condition
-  geom_line(data = h3Spat_quants, aes(x = date, y = Q50), col = 'red') +
-  geom_ribbon(
-    data = h3Spat_quants,
-    aes(x = date, ymin = Q025, ymax = Q975),
-    fill = 'red', alpha = 0.5
-  ) +
-  # projections from initial condition, using filtered sims function
-  geom_line(data = mod3_V0_initcond_res$SQuants, aes(x = date, y = Q50), col = 'red') +
-  geom_ribbon(
-    data = mod3_V0_initcond_res$SQuants,
-    aes(x = date, ymin = Q025, ymax = Q975),
-    fill = 'red', alpha = 0.5
-  ) +
-  # Sims from filtered distribution
-  geom_line(
-    data = bpf_final3_out$Sus_df,
-    aes(x = as.Date(lubridate::date_decimal(time)), y = Q50),
-    col = 'blue'
-  ) +
-  geom_ribbon(
-    data = bpf_final3_out$Sus_df,
-    aes(x = as.Date(lubridate::date_decimal(time)), ymin = Q025, ymax = Q975),
-    fill = 'blue', alpha = 0.5
-  ) +
-  # Projections from filtered distribution
-  geom_line(data = mod3_V0_res$SQuants, aes(x = date, y = Q50), col = 'blue') +
-  geom_ribbon(
-    data = mod3_V0_res$SQuants,
-    aes(x = date, ymin = Q025, ymax = Q975),
-    fill = 'blue', alpha = 0.5
-  ) +
-  geom_vline(xintercept = max(h3Spat_quants$date), linetype = 'dashed') +
-  facet_wrap(~dep, scales = 'free_y', nrow = 2) +
-  theme_bw() +
-  theme(axis.title.x = element_blank(),
-        axis.text.x = element_text(angle = -45, hjust = 0)) +
-  ylab("Total Susceptible (%)")
-@
-\caption{\label{fig:h3Sus}Percentage of individuals that are in the susceptible compartment.
-Simulations from Model~3 under initial conditions are displayed in red; simulations from the filtering distribution of model are displayed in blue.
-The dashed line represents the end of the observed data.}
-\end{figure}
-
-\clearpage
-
-\section{Replication of \cite{lee20}}
-
-In this article we claimed that we were able to obtain better fits to the observed data using the same models that were proposed by \cite{lee20}.
-Along with visual comparisons to the data, this claim was supported by comparing likelihoods and AIC values in Table~2 in the manuscript. % TODO: Double check the table number matches
-Because model likelihoods were not provided by \cite{lee20}, it is necessary to replicate these models in order to obtain likelihood estimates.
-Here we would like to thank the authors of \cite{lee20}, who provided detailed descriptions of their models, which enabled us to build on their work.
-In the following subsections, we use our \code{R} package \code{haitipkg} to reproduce some of the results of \cite{lee20}.
-This reproduction allows us to estimate the likelihoods of the \cite{lee20} version of Models~1--3, and also provides a demonstration of the importance and usefulness of reproducible research.
-
-\subsection{Model~1 Replication}
-
-The model was implemented by a team at Johns Hopkins Bloomberg School of Public Health (hereafter referred to as the Model~1 authors) in the \code{R} programming language using the \code{pomp} package \citep{king16}.
-Original source code is publicly available with DOI: 10.5281/zenodo.3360991.
-The final results reported by the Model~1 authors were obtained by using several different parameter sets rather than a single point estimate.
-According to the supplement materials, this was because model realizations from a single parameter set retained substantial variability, but multiple realizations from a collection of parameter sets resulted in a reasonable visual fit to the data.
-We are also inclined to believe that the use of multiple parameter values was in part intended to account for parameter uncertainty---the importance of which was discussed in the main text---an effort by the Model~1 authors that we applaud.
-Simulations from each of the parameter sets, however, were treated with equal importance when being used to diagnose the model fit and make inference on the system.
-This is problematic given Figures S8 and S9 of the supplement material, which suggest that some parameter sets that were used for inference may have been several hundred units of log-likelihood lower than other parameter sets that were simultaneously used to make forecasts.
-Such a large difference in log-likelihoods is well beyond the threshold of statistical uncertainty determined by Wilks' theorem, resulting in the equal use of statistically inferior parameter sets in order to make forecasts and conduct inference on the system.
-
-To fully reproduce the results of the Model~1 authors, it is necessary to use the exact same set of model parameters that were originally used to obtain the results presented by \cite{lee20}.
-Because these parameter sets were not made publicly available, we relied on the source code provided by the Model~1 authors to approximately recreate the parameter set.
-Due to software updates since the publication of the source code, we were unable to produce the exact same set of parameters.
-Running the publicly available source code, however, resulted in a set of parameters that are visually similar to those used by the Model~1 authors (See Figures~\ref{fig:PlotEpiDist} and \ref{fig:plotEndParams}).
-Furthermore, simulations using the set of parameters produced by the source code appear practically equivalent to those displayed by \cite{lee20} (See Figure~\ref{fig:plotMod1Sims}).
-
-<<Simulate Lee Model 1, echo=FALSE, cache=TRUE>>=
-h1_epi <- haiti1()
-N_LEE_SIMS <- 20
-
-registerDoRNG(18599687)
-foreach(
-  i = 1:length(h1LeeStartsEpi),
-  .combine = rbind
-) %dopar% {
-  guess <- h1LeeStartsEpi[i, ]
-  sims <- simulate(h1_epi, nsim = N_LEE_SIMS, format = 'data.frame',
-                   params = guess)  # 20 sims for each set of parameters
-  sims$param_set <- i
-  sims %>%
-    dplyr::select(.id, week, cases, param_set)
-} -> all_sims
-@
-
-<<PlotEpiDist, echo=FALSE, fig.cap="Compare to Figure S8 of Lee et al. (2020) supplement.", fig.width=7, fig.height=7, fig.align='center'>>=
-GGally::ggpairs(
-  h1LeeStartsEpi,
-  columns = c("rho", "tau", "beta1", "nu", "loglik")
-)
-@
-
-<<SimulateEpi, echo=FALSE, include=FALSE, message=FALSE, cache=TRUE>>=
-h1_end <- haiti1(period = "endemic")
-
-true_agg_cases <- haitiCholera %>%
-  select(-report) %>%
-  pivot_longer(
-    data = .,
-    cols = -date_saturday,
-    values_to = 'cases',
-    names_to = 'dep'
-  ) %>%
-  mutate(
-    date = as.Date(date_saturday),
-    dep = gsub("\\.", "_", dep)
-  ) %>%
-  mutate(
-    dep = case_when(dep == "Grand_Anse" ~ "Grande_Anse", TRUE ~ dep)
-  ) %>%
-  tidyr::pivot_wider(
-    data = .,
-    id_cols = c(date),
-    names_from = dep,
-    values_from = cases,
-    names_prefix = 'cases_'
-  ) %>%
-  mutate(
-    ReportedAll = cases_Artibonite + cases_Centre +
-      cases_Grande_Anse + cases_Nippes + cases_Nord +
-      cases_Nord_Est + cases_Ouest + cases_Sud +
-      cases_Sud_Est + cases_Nord_Ouest
-  )
-@
-
-<<FitEnd, echo=FALSE, message=FALSE, include=FALSE>>=
-foreach(
-  i = 1:nrow(h1LeeStartsEnd),
-  .combine = rbind
-) %dopar% {
-  r_params <- h1LeeStartsEnd[i, ][names(coef(h1_end))]
-  coef(h1_end) <- r_params
-
-  sims <- simulate(
-    h1_end, nsim = N_LEE_SIMS,
-    format = 'data.frame',
-    params = r_params
-  )
-  sims$param_set <- i
-  sims %>%
-    dplyr::select(.id, week, cases, param_set)
-} -> end_sims
-
-@
-
-<<plotEndParams, fig.width=7, fig.height=7, fig.align='center', fig.cap="Bivariate relationships between variables after fitting endemic period.
-Compare to S9 of Lee et al.
-(2020) supplement.", cache=TRUE, echo=FALSE>>=
-GGally::ggpairs(
-  h1LeeStartsEnd,
-  columns = c("rho", "tau", "beta1", "nu", "loglik")
-)
-@
-
-<<calcMod1LeeLikes, echo=FALSE, include=FALSE, message=FALSE>>=
-h1_epi_evals <- readRDS("../model1/run_level_3/lee1_epi_evals.rds")
-h1_end_evals <- readRDS("../model1/run_level_3/lee1_end_evals.rds")
-
-mod1_all_likes <- dplyr::inner_join(
-  x = h1_epi_evals,
-  y = h1_end_evals,
-  by = "parid"
-  ) %>%
-  dplyr::mutate(
-    joint_ll = pfLL.x + pfLL.y
-  )
-
-# Approximate the likelihood of the Lee et al (2020) version of Model 1 by
-# using the parameter sets that resulted in the maximum likelihood.
-mod1_lee_ll <- max(mod1_all_likes$joint_ll, na.rm = TRUE)
-@
-
-<<plotMod1Sims, echo=FALSE, fig.cap=paste0("Simulations using parameter sets that were generated by running source code provided by Lee et al (2020).
-Compare to Figure S7 of \\cite{lee20sup}.
-The upper bound for the likelihood of this model is ", round(mod1_lee_ll), "."), cache=TRUE, fig.width=6.5, fig.height=3.75>>=
-
-quants <- all_sims %>%
-  filter(!param_set %in% c(1, 2)) %>%  # Remove parameter sets inconsistent with Lee et al (2020).
-  group_by(week) %>%
-  summarize(
-    q025 = quantile(cases, probs = 0.025, na.rm = TRUE),
-    q50  = quantile(cases, probs = 0.500, na.rm = TRUE),
-    q975 = quantile(cases, probs = 0.975, na.rm = TRUE)
-  ) %>%
-  ungroup() %>%
-  mutate(date = lubridate::ymd("2010-10-16") + lubridate::weeks(week))
-
-quants_end <- end_sims %>%
-  group_by(week) %>%
-  summarize(
-    q025 = quantile(cases, probs = 0.025, na.rm = TRUE),
-    q50  = quantile(cases, probs = 0.500, na.rm = TRUE),
-    q975 = quantile(cases, probs = 0.975, na.rm = TRUE)
-  ) %>%
-  ungroup() %>%
-  mutate(date = lubridate::ymd("2010-10-16") + lubridate::weeks(week))
-
-ggplot() +
-  geom_point(data = true_agg_cases, aes(x = date, y = ReportedAll)) +
-  geom_line(data = quants, aes(x = date, y = q50), col = '#0b8a2b', size = 1) +
-  geom_ribbon(
-    data = quants,
-    aes(x = date, ymin = q025, ymax = q975),
-    alpha = 0.5, fill = '#0b9c30'
-  ) +
-  geom_line(data = quants_end, aes(x = date, y = q50), col = '#882871') +
-  geom_ribbon(data = quants_end, aes(x = date, ymin = q025, ymax = q975), alpha = 0.5, fill = '#882871') +
-  ylab("Reported, Symptomatic Incidence") +
-  scale_x_date(date_breaks = "1 year") +
-  theme(axis.title.x = element_blank(),
-        axis.text.x = element_text(hjust = 1, angle = 45))
-@
-
-Because the model forecasts provided by \cite{lee20} come from various sets of parameters---which each correspond to a unique log-likelihood value---it is not obvious how one would obtain an estimate for the log-likelihood of the model that was used for simulations by the Model~1 authors.
-One approach could be to calculate the logarithm of the weighted mean of the likelihoods for each parameter sets used to obtain the forecasts, where the weights are proportional to the number of times the parameter set was used.
-However, in an effort to not underestimate the likelihood of the model of the Model~1 authors, we report the estimated log-likelihood as the log-likelihood value corresponding to the parameter set with the largest likelihood value, even though the majority of simulations were obtained using parameter sets with lower likelihood values.
-In this sense, we consider the log-likelihood reported in Table~1 of the main text to be an upper-bound of the log-likelihood of the model used by \cite{lee20}.
-For each parameter set, the log-likelihood was estimated using a particle filter, implemented as the \texttt{pfilter} function in the \texttt{pomp} package.
-
-\subsection{Model~2 Replication}\label{sec:mod2rep}
-
-Model~2 was developed by a team that consisted of members from the Fred Hutchinson Cancer Research Center and the University of Florida (hereafter referred to as the Model~2 authors).
-While Model~2 is the only deterministic model we considered in our analysis, it contains perhaps the most complex descriptions of cholera in Haiti: Model~2 accounts for movement between spatial units; human-to-human and environment-to-human cholera infections; and transfer of water between spatial units based on elevation charts and river flows.
-
-The source code that the Model~2 authors used to generate their results was written in the \code{Python} programming language, and is publicly available at \url{10.5281/zenodo.3360857} and its accompanying GitHub repository \url{https://github.com/lulelita/HaitiCholeraMultiModelingProject}.
-In order to perform our analysis in a unified framework, we re-implemented this model in the \code{R} programming language using the \code{spatPomp} package \citep{asfaw23arxiv}, which facilitates the creation of meta-population models.
-We note that the travel and water matrices used to implement the complex dynamics in Model~2 \citep{lee20sup} are not available in either the Zenodo archive or the GitHub repository;
-instead, we obtained these matrices via personal correspondence with the Model~2 authors.
-Using these matrices, and the point estimates for model parameters provided by \citep{lee20sup}, we created trajectories of the cholera dynamics and compared this to available data.
-These trajectories, shown in Figure~\ref{fig:mod2rep}, are very similar to the trajectories shown in Figure~S15 of \cite{lee20sup}.
-
-<<Model2Replication, cache=TRUE, echo=FALSE>>=
-h2_epi <- haiti2()
-h2_end <- haiti2(region = 'after')
-
-h2_epi_traj <- trajectory(h2_epi, format = 'data.frame')
-h2_end_traj <- trajectory(h2_end, format = 'data.frame')
-
-h2_epi_traj$Ctotal <- rowSums(h2_epi_traj[, paste0("C", 1:10)]) * 0.2
-h2_end_traj$Ctotal <- rowSums(h2_end_traj[, paste0("C", 1:10)]) * 0.2
-
-dep_plot_df <- haitipkg::haitiCholera %>%
-  select(-report) %>%
-  tidyr::pivot_longer(
-    data = .,
-    cols = -date_saturday,
-    values_to = 'cases',
-    names_to = 'dep'
-  ) %>%
-  mutate(
-    date = as.Date(date_saturday),
-    dep = gsub("\\.", "_", dep)
-  ) %>%
-  mutate(
-    dep = case_when(dep == "Grand_Anse" ~ "Grande_Anse", TRUE ~ dep)
-  )
-
-true_agg_cases <- dep_plot_df %>%
-  tidyr::pivot_wider(
-    data = .,
-    id_cols = c(date),
-    names_from = dep,
-    values_from = cases,
-    names_prefix = 'cases_'
-  ) %>%
-  mutate(
-    ReportedAll = cases_Artibonite + cases_Centre +
-      cases_Grande_Anse + cases_Nippes + cases_Nord +
-      cases_Nord_Est + cases_Ouest + cases_Sud +
-      cases_Sud_Est + cases_Nord_Ouest
-  ) %>%
-  select(date, ReportedAll)
-
-h2_epi_traj$date <- as.Date(
-  lubridate::round_date(
-    lubridate::date_decimal(h2_epi_traj$year),
-    unit = 'day'
-  )
-)
-
-h2_end_traj$date <- as.Date(
-  lubridate::round_date(
-    lubridate::date_decimal(h2_end_traj$year),
-    unit = 'day'
-  )
-)
-
-all_traj <- bind_rows(
-  select(h2_epi_traj, date, Ctotal),
-  select(h2_end_traj, date, Ctotal)
-)
-
-projections <- left_join(
-  x = true_agg_cases,
-  y = all_traj,
-  by = 'date'
-)
-@
-
-\begin{figure}[ht]
-<<plotModel2Rep, cache=TRUE, echo=FALSE, fig.width=6.5, fig.height=4.5>>=
-gg_epi <- projections %>%
-  filter(date < as.Date("03-01-2014", format = '%m-%d-%Y')) %>%
-  ggplot(aes(x = date)) +
-  geom_line(aes(y = ReportedAll), col = 'black', linetype = 'dashed') +
-  geom_line(aes(y = Ctotal), col = 'red', size = 1) +
-  theme_bw() +
-  scale_y_continuous(limits = c(0, 30000), breaks = seq(0, 30000, 5000)) +
-  theme(axis.title.x = element_blank()) +
-  ylab("Reported number of cases")
-
-gg_end <- projections %>%
-  filter(date >= as.Date("03-01-2014", format = '%m-%d-%Y')) %>%
-  ggplot(aes(x = date)) +
-  geom_line(aes(y = ReportedAll), col = 'black', linetype = 'dashed') +
-  geom_line(aes(y = Ctotal), col = 'red', size = 1) +
-  theme_bw() +
-  scale_y_continuous(limits = c(0, 7000), breaks = seq(0, 7000, 1000)) +
-  theme(axis.title.x = element_blank()) +
-  ylab("Reported number of cases") +
-  scale_x_date(date_breaks = "1 year", date_labels = "%Y")
-
-cowplot::plot_grid(gg_epi, gg_end, ncol = 1)
-@
-\caption{\label{fig:mod2rep}
-Model 2 trajectories using the \code{haitipkg}. Compare to Figure S15 of \cite{lee20sup}.
-}
-\end{figure}
-
-There are minor differences between Figure~\ref{fig:mod2rep} and Figure~S15 of \cite{lee20sup}.
-While the discrepancy appears minor, the deterministic nature of Model~2 implies that an exact replication of model trajectories should be possible.
-In this case, these discrepancies may possibly be attributed to implementing the model and plotting the model trajectory in two different programming languages.
-Another potential explanation for the discrepancy is that the parameters that we used are only approximately the same as those used by \citet{lee20sup}.
-For example, the parameters $\transmission$, $\Wbeta{}$ (See Table~\ref{tab:translate}) had reported values of $9.9 \times 10^{-7}$ and $4.03 \times 10^{-2}$, respectively (Table~S13 of \cite{lee20sup}), but were actually fit to data and therefore likely these values have been rounded.
-Additionally, our implementation of Model~2 used a time scale of years and many of the parameters were reported on a weekly scale, so small differences may result due to unit conversions.
-The collective effect of these small differences in model parameters likely will result in small differences in model trajectories.
-
-Some additional concerns about being able to accurately replicate the results of \cite{lee20} are valid.
-Details about the measurement models and how latent states were initialized for the epidemic model were not provided by \citet{lee20sup} and therefore these details must be inferred by looking at the provided source code.
-According to repository comments, the files \code{fit\allowbreak In\allowbreak Pieces\allowbreak 3params\allowbreak Clean\allowbreak May2019\allowbreak Public.py} and \code{fit\allowbreak In\allowbreak Pieces\allowbreak Mu\allowbreak With\allowbreak Frac\allowbreak Sus\allowbreak Fixed\allowbreak All\allowbreak Infections\allowbreak Public.py} were used to fit the epidemic and endemic phases of the model respectively, although it is apparent that these exact files were not used to obtain the reported results since the files contain some variable-naming errors that make it impossible to run the files without making modifications \footnote{One example of why the code cannot be run that the file loads functions from a non-extant file named \code{choleraEqs.py} in line 13 rather than \code{cholera\allowbreak Eqs\allowbreak Public.py}.}.
-The inability to replicate the results by \citet{lee20} by running the provided source code makes checking whether or not a our numeric implementation faithfully represents their results very difficult.
-Additionally, the script that was said to been used to obtain the results reported by \cite{lee20} appears to use a different measurement model than what was described in the supplemental material, again making it difficult to fully replicate the result of \cite{lee20} without being able to easily run the provided source code.
-In this case, we chose to use measurement model described by Eq.~\ref{model2:measEpi} for both phases of the epidemic, as this seemed to visually match the results of \cite{lee20} most closely.
-
-\subsection{Model~3 Replication}
-
-Model~3 was developed by a team of researchers at the Laboratory of the Swiss Federal Institute of Technology in Lausanne, hereafter referred to as the Model~3 authors.
-The code that was originally used to implement Model~3 is archived with the DOI: \url{10.5281/zenodo.3360723}, and also available in the public GitHub repository: \code{jcblemai/haiti-mass-ocv-campaign}.
-Because the code was made publicly available, and final model parameters were reported in the supplementary material of \cite{lee20}, we were able to reproduce Model~3 by directly using the source code.
-In Fig.~\ref{fig:mod3rep}, we plot simulations from this model.
-This figure can be compared to Figure S18 of \cite{lee20}.
-We note that slight differences may be accounted for due to variance in the model simulations and the difference in programming language used to produce the figure.
-Overall, the high standard of reproducibility that was achieved by the Model~3 authors facilitated the ability to readily replicate their model and results.
-
-<<Model3Replication, echo=FALSE, cache=TRUE>>=
-NSIM <- 500
-
-plot_order <- c(
-  'Artibonite',
-  'Sud_Est',
-  'Nippes',
-  'Nord_Est',
-  'Ouest',
-  'Centre',
-  'Nord',
-  'Sud',
-  'Nord_Ouest',
-  'Grande_Anse'
-)
-
-dep_labeller <- as_labeller(
-  c(
-    'Artibonite' = 'Artibonite',
-    'Sud_Est' = 'Sud-Est',
-    'Sud.Est' = 'Sud-Est',
-    'Nippes' = 'Nippes',
-    'Nord_Est' = 'Nord-Est',
-    'Nord.Est' = 'Nord-Est',
-    'Ouest' = 'Ouest',
-    'Centre' = 'Centre',
-    'Nord' = 'Nord',
-    'Sud' = 'Sud',
-    'Nord_Ouest' = 'Nord-Ouest',
-    'Nord.Ouest' = 'Nord-Ouest',
-    'Grande_Anse' = 'Grand\'Anse',
-    'Grand.Anse' = 'Grand\'Anse'
-  )
-)
-
-lee3 <- lee3_spatPomp()
-lee3_sims_quants <- bake("../model3/si/lee3Sims.rds", {
-
-  # Simulate from lee et al 2020 model
-  sims_temp <- simulate(
-    lee3, nsim = NSIM,
-    seed = 321, format = 'data.frame'
-  )
-
-  # Get desired quantiles of reported cases from simulations
-  quants <- sims_temp %>%
-    dplyr::group_by(unitname, time) %>%
-    dplyr::summarise(
-      q05 = quantile(cases, 0.025, na.rm = T),
-      mean = mean(cases, na.rm = T),
-      q50 = quantile(cases, 0.5, na.rm = T),
-      q95 = quantile(cases, 0.975, na.rm = T)
-    ) %>%
-    dplyr::ungroup() %>%
-    dplyr::mutate(
-      date = lubridate::date_decimal(time),
-      date = as.Date(lubridate::round_date(date))
-    ) %>%
-    dplyr::rename(dep = unitname)
-
-  # Remove cases from Ouest that were not considered by Lee et al. (2020)
-  quants %>%
-    dplyr::filter(
-      !(dep == "Ouest" & date < '2017-06-10')
-    )
-
-})
-
-# Get the subset of cholera cases that match simulations from Lee et al. (2020) model.
-sub_cases <- haitiCholera %>%
-  dplyr::mutate(date = as.Date(date_saturday)) %>%
-  dplyr::select(-report, -date_saturday) %>%
-  dplyr::filter(lubridate::decimal_date(date) >= min(lee3@times)) %>%
-  tidyr::pivot_longer(
-    data = .,
-    cols = -date,
-    names_to = "dep",
-    values_to = "cases"
-  ) %>%
-  dplyr::mutate(
-    dep = dplyr::case_when(
-      dep == "Grand.Anse" ~ "Grande_Anse",
-      dep == "Nord.Est" ~ "Nord_Est",
-      dep == "Nord.Ouest" ~ "Nord_Ouest",
-      dep == "Sud.Est" ~ "Sud_Est",
-      TRUE ~ dep
-    )
-  )
-@
-
-\begin{figure}[ht]
-<<PlotMod3Rep, echo=FALSE, message=FALSE, fig.height=3.75>>=
-ggplot() +
-  # Region for upper and lower simulation bounds
-  geom_ribbon(
-    data = lee3_sims_quants,
-    aes(x = date, ymin = q05, ymax = q95),
-    fill = 'darkblue', alpha = 0.5
-  ) +
-  # Median line
-  geom_line(
-    data = lee3_sims_quants,
-    aes(x = date, y = q50),
-    col = 'darkblue'
-  ) +
-  # Observed data
-  geom_point(
-    data = dplyr::filter(sub_cases, cases < 500),
-    aes(x = date, y = cases), size = 0.2
-  ) +
-  facet_wrap(
-    ~factor(dep, levels = plot_order),
-    nrow = 2, labeller = dep_labeller
-  ) +
-  labs(y = 'Reported Cholera Cases') +
-  scale_x_date(date_breaks = "6 months", date_labels = "%Y-%m") +
-  theme(
-    axis.text.x = element_text(angle = 40, hjust = 1, vjust = 1, size = 7),
-    axis.title.x = element_blank()
-  )
-@
-\caption{\label{fig:mod3rep}
-Simulations from Model~3.
-Compare to Figure S18 of \cite{lee20}.
-}
-\end{figure}
-
-<<lee3Cleanup, echo=FALSE, include=FALSE, message=FALSE>>=
-rm(lee3, lee3_sims_quants, sub_cases)
-gc()
-@
-
-\section{Forecasting with parameter uncertainty}
-
-Let $f_{Y_{1:N}}(y_{1:N} | \theta)$ denote the pdf of the model under consideration, were $\theta$ is a parameter vector that indexes the model.
-Furthermore, denote the observed data as $y_{1:N}^*$.
-Because the uncertainty in just a single parameter can lead to drastically different forecasts \citep{saltelli20},
-parameter uncertainty should be considered when obtaining model forecasts when the goal is to influencing policy.
-In a Bayesian modelling paradigm, the most natural way to account for parameter uncertainty in model forecasts is to suppose that $\theta$ comes from a distribution $f_{\Theta}$, and then to obtain $J$ forecasts from the model where each forecast is obtained using parameters drawn from the posterior distribution $\Theta_{1:J} \mid Y_{1:N} = y_{1:N}^* \sim f_{\Theta}\big(\theta | Y_{1:N} = y_{1:N}^*\big)$.
-
-When frequentist methods are used, however, there does not exist a posterior distribution from which one could sample.
-A common approach could be to obtain a weighted average of the simulations from various models \citep{hoeting99}, but this can be problematic when forecasts from each model are very different from each other \citep{grueber11}.
-A similar approach that has been taken \citep{king15} is to obtain model forecasts using multiple sets of parameter values and then sample from the resulting forecasts using weights proportional to the corresponding likelihoods of the parameter values.
-This approach could be considered as empirical Bayes, as it is equivalent to using a discrete uniform prior where the set of values in the prior distribution is determined by a stochastic routine applied to the observed data, as discussed below.
-
-For each $i \in 1:k$, let $\Theta_i$ be a random vector of model parameters.
-Then, letting $\Theta$ denote the true model parameters, we endow the set $\{\Theta_1, \Theta_2, \ldots, \Theta_k\}$ with a discrete uniform distribution, such that $P\big(\Theta = \Theta_i\big) = \frac{1}{K}$ for all values $i \in \seq{1}{k}$.
-Using this as a prior distribution, the posterior distribution of $\Theta | Y_{1:N} = y_{1:N}^*$ can be expressed as:
-$P\big(\Theta = \Theta_k | Y_{1:N} = y_{1:N}^*\big) = \frac{f_{Y_{1:N}}(y_{1:N}^*| \Theta_k)}{\sum_{l = 1}^K f_{Y_{1:N}}(y_{1:N}^*| \theta_l)}$.
-In a standard empirical Bayes analysis, the values $\Theta_1, \ldots, \Theta_k$ of the prior distribution would be chosen using the observed data, resulting in a posterior distribution that weighs the prior parameter vectors proportional to their corresponding likelihoods.
-Instead, we choose $\Theta_i$ to be the output of a stochastic routine applied to the observed data by setting $\Theta_i$ to be the output of an iterated filtering algorithm.
-In practice, because the likelihood maximization routines of iterated filtering methods are stochastic, it is common to run the iterated filtering method multiple times for each model in order to obtain a maximum likelihood estimate for model parameters.
-This results in a natural set of parameters near the MLE that could be used as the discrete prior distribution.
-
-\section{Translating to \citet{lee20} notation}
-
-Since the models of \citet{lee20} were developed independently, the choice of notation varies inconsistently between models.
-For our reanalysis, we rename parameters to provide a unified notation facilitating comparison between models.
-Table~\ref{tab:translate} maps this notation back to the original notations, for reference.
-
-\begin{table}
-  \begin{center}
-  \begin{tabular}{|c|c|c|c|c|}\hline
-    \multirow{2}{*}{Parameter} & Our & \multicolumn{3}{c|}{Lee et al. (2020a)} \\\cline{3-5}
-     & Notation & 1 & 2 & 3 \\
-    \hline
-    \hline
-    % Table begins here
-    Reporting Rate & $\reportRate$ & $\rho$ & $\rho$ & $\epsilon_1, \epsilon_2$ \\\hline
-    Mixing Coefficient & $\mixExponent$ & $\nu$ & --- & --- \\\hline
-    Measurement Over-Dispersion & $\obsOverdispersion$ & $\tau$ & --- & $p$ \\\hline
-    Birth Rate & $\muBirth$ & $\mu$ & --- & --- \\\hline
-    Natural Mortality Rate & $\muDeath$ & $\delta$ & --- & $\mu$ \\\hline
-    Cholera Mortality Rate & $\choleraDeath$ & --- & --- & $\alpha$ \\\hline
-    Latent Period & $1/\muEI$ & $1/\sigma$ & $1/\gamma_E$ & --- \\\hline
-    Recovery Rate & $\muIR$ & $\gamma$ & $\gamma$ & $\gamma$ \\\hline
-    Loss of Immunity & $\muRS$ & $\alpha$ & $\sigma$ & $\rho$ \\\hline
-    Symptomatic Ratio & $\symptomFrac$ & $1 - \theta_0$ & $k$ & $\sigma$ \\\hline
-    Asymptomatic Relative Infectiousness & $\asymptomRelativeInfect$ & $\kappa$ & $red_\beta$ & --- \\\hline
-    Human-to-Water Shedding & $\Wshed$ & --- & $\mu$ & $\theta_I$ \\\hline
-    Asymptomatic Relative Shedding & $\asymptomRelativeShed$ & --- & $red_\mu$ & $\theta_A/\theta_I$ \\\hline
-    Seasonal Amplitude & $\seasAmplitude$ & --- & $\alpha_s$ & $\lambda$ \\\hline
-    Transmission & $\transmission$ & $\beta$ & $\beta$ & $c$ \\\hline
-    Water-to-Human & $\Wbeta{}$ & --- & $\beta_W$ & $\beta$ \\\hline
-    Bacteria Mortality & $\Wremoval$ & --- & $\delta$ & $\mu_\beta$ \\\hline
-    Vaccination Efficacy & $\vaccineEfficacy$ & $\theta_{vk}$ & $\theta_1, \theta_2, \theta_{1_5}, \theta_{2_5}$ & $\eta_{1d}, \eta_{2d}$ \\\hline
-    Process Over-dispersion & $\sigmaProc$ & --- & --- & $\sigma_w$\\\hline
-    % Half Saturation Constant & $\Wsat$ & --- & $Sat$ & $K$ \\\hline
-  \end{tabular}
-  \end{center}
-  \caption{
-  \label{tab:translate}Translations between our common notation and notation used by \cite{lee20}
-  }
-\end{table}
-
-\clearpage
-
-\bibliography{../bib-haiti}
-
-\end{document}

### 2023-09-01 si/inputs/measurementModels.tex
Started work to make individual si sections seperate

--- 2023-08-31 si/si.Rnw
@@
-\documentclass[11pt]{article}
-
-\pdfminorversion=4
-
-\newcommand\secTitleSpace{\hspace{3mm}}
-
-<<packages, include=FALSE>>=
-library(tidyverse)
-library(pomp)
-library(haitipkg)
-library(spatPomp)
-library(doParallel)
-library(doRNG)
-
-cores <-  as.numeric(Sys.getenv('SLURM_NTASKS_PER_NODE', unset = NA))
-if(is.na(cores)) cores <- detectCores()
-# cores <- 20
-registerDoParallel(cores)
-theme_set(theme_bw())
-
-if (!dir.exists("../model3/si")) {
-  dir.create("../model3/si", recursive = TRUE)
-}
-@
-
-<<set-opts, include=FALSE>>=
-options(
-  scipen = 2,
-  help_type = "html",
-  stringsAsFactors = FALSE,
-  continue = "+  ",
-  width = 70,
-  useFancyQuotes = FALSE,
-  reindent.spaces = 2,
-  xtable.comment = FALSE
-)
-@
-
-<<knitr-opts, include=FALSE, purl=FALSE>>=
-library(knitr)
-opts_knit$set(concordance=TRUE)
-opts_chunk$set(
-  progress=TRUE,prompt=TRUE,highlight=FALSE,
-  tidy=TRUE,
-  tidy.opts=list(
-    keep.blank.line=FALSE
-  ),
-  comment="",
-  warning=FALSE,
-  message=FALSE,
-  error=TRUE,
-  echo=FALSE,
-  cache=FALSE,
-  strip.white=TRUE,
-  results="markup",
-  fig.path="figure/",
-  fig.lp="fig:",
-  fig.align="center",
-  fig.show="asis",
-  dev="pdf",
-  dev.args=list(
-    bg="transparent",
-    pointsize=9
-  )
-)
-
-myround <- function (x, digits = 1) {
-  # taken from the broman package
-  if (digits < 1)
-    stop("This is intended for the case digits >= 1.")
-  if (length(digits) > 1) {
-    digits <- digits[1]
-    warning("Using only digits[1]")
-  }
-  tmp <- sprintf(paste("%.", digits, "f", sep = ""), x)
-  zero <- paste0("0.", paste(rep("0", digits), collapse = ""))
-  tmp[tmp == paste0("-", zero)] <- zero
-  tmp
-}
-
-@
-
-%% copied from pnas-new.cls
-\usepackage{amsmath,amsfonts,amssymb,graphicx}
-
-%%%% parameters %%%%%%%%%%5
-\newcommand\Wsat{W_{\mathrm{sat}}}
-\newcommand\muIR{\mu_{IR}}
-\newcommand\muEI{\mu_{EI}}
-\newcommand\transmission{\beta}
-\newcommand\seasAmplitude{a}
-\newcommand\rainfallExponent{r}
-\newcommand\muRS{\mu_{RS}}
-\newcommand\vaccineEfficacy{\theta}
-\newcommand\muBirth{\mu_S}
-\newcommand\muDeath{\delta}
-\newcommand\choleraDeath{\delta_{C}}
-\newcommand\symptomFrac{f}
-\newcommand\asymptomRelativeInfect{\epsilon}
-\newcommand\asymptomRelativeShed{\epsilon_{W}}
-\newcommand\Wbeta[1]{\beta_{W#1}}
-\newcommand\Wremoval{\delta_W}
-\newcommand\Wshed{\mu_W}
-\newcommand\mixExponent{\nu}
-\newcommand\sigmaProc{\sigma_{\mathrm{proc}}}
-\newcommand\reportRate{\rho}
-\newcommand\obsOverdispersion{\psi}
-\newcommand\phaseParm{\phi}
-\newcommand\transmissionTrend{\zeta}
-\newcommand\Binit{\xi}
-\newcommand\vaccClass{Z}
-\newcommand\vaccCounter{z}
-\newcommand\modelCounter{m}
-\newcommand\Whur[1]{\beta_{W#1}^{hm}}
-\newcommand\hHur[1]{h_{#1}^{hm}}
-\newcommand\Iinit{I_{0,0}}
-\newcommand\Einit{E_{0, 0}}
-\newcommand\reportChange{\delta}
-
-\newcommand\missing{\hspace{6mm} ---}
-\newcommand\fixed{\color{blue}}
-\newcommand\demography{\bullet}
-
-\newcommand\mytitle{Informing policy via dynamic models: Cholera in Haiti}
-\newcommand\comma{{\hspace{-0.25mm},\hspace{-0.2mm}}}
-
-%%%%%%%%%%%%%%%% START OF USER DEFINITIONS %%%%%%%%%%%%%%%5
-
-\newcommand\siOnly[1]{} %% redefined for the supplement
-\newcommand\msOnly[1]{#1} %% redefined for the supplement
-
-\newcommand\code[1]{\texttt{#1}}
-\usepackage{xurl}
-% \usepackage[breaklinks]{hyperref}
-\usepackage{makecell}
-\usepackage{multirow}
-\usepackage{multicol}
-\usepackage[ruled,noline,linesnumbered]{algorithm2e}
-% \usepackage{setspace}
-
-%% EDITING MACROS
-\usepackage{color}
-\usepackage[normalem]{ulem}% to use \sout in feedback commands
-\usepackage{bm}
-% orange for EI
-\definecolor{orange}{rgb}{1,0.5,0}
-\newcommand\ei[2]{\sout{#1} \textcolor{orange}{#2}}
-\newcommand\eic[1]{\textcolor{orange}{[#1]}}
-% green for JW
-\definecolor{green}{rgb}{0,0.5,0}
-\newcommand\jw[2]{\sout{#1} \textcolor{green}{#2}}
-\newcommand\jwc[1]{\textcolor{green}{[#1]}}
-% purple for JJ
-\definecolor{purple}{rgb}{0.5,0,1}
-\newcommand\jj[2]{\sout{#1} \textcolor{purple}{#2}}
-\newcommand\jjc[1]{\textcolor{purple}{[#1]}}
-% cyan for AR
-\definecolor{cyan}{rgb}{0,.5,.5}
-\newcommand\ar[2]{\sout{#1} \textcolor{cyan}{#2}}
-\newcommand\arc[1]{\textcolor{cyan}{#1}}
-% light brown for KT
-\definecolor{lightbrown}{rgb}{0.5,0.5,0}
-\newcommand\kt[2]{\sout{#1} \textcolor{lightbrown}{#2}}
-\newcommand\ktc[1]{\textcolor{lightbrown}{#1}}
-
-%% basic POMP definitions
-\newcommand\Xspace{{\mathbb X}}
-\newcommand\Yspace{{\mathbb Y}}
-\newcommand\Thetaspace{\R^{\Thetadim}}
-\newcommand\Xunitspace{{\scriptsize{\mathcal X}}}
-\newcommand\Yunitspace{{\tiny{\mathcal Y}}}
-\newcommand\hatTheta{\widehat{\Theta}}
-\newcommand\Xdim{{\mathrm{dim}}(\Xspace)}
-\newcommand\Ydim{{\mathrm{dim}}(\Yspace)}
-\newcommand\Thetadim{P}
-\newcommand\thetadim{p}
-\newcommand\timeSet{{\mathbb T}}
-\newcommand\data[1]{#1^*}
-
-\newcommand\unitSpecific{\hspace{0.1mm}\mathrm{us}}
-\newcommand\shared{\hspace{0.15mm}\mathrm{sh}}
-
-% for algorithm pseudocode
-\SetKwFor{For}{for}{:}{end}
-
-%% for particle filters
-\newcommand\unit{u}
-\newcommand\altUnit{\tilde u}
-\newcommand\Unit{U}
-\newcommand\rep{i}
-\newcommand\Rep{\mathcal{I}}
-\newcommand\island{\rep}
-\newcommand\Island{\Rep}
-%\newcommand\txtisland{replicate}
-\newcommand\altIsland{\tilde i}
-\usepackage[mathscr]{euscript}
-
-\renewcommand\time{n}
-\newcommand\myvec[1]{\boldsymbol{#1}}
-\newcommand\altTime{\tilde n}
-\newcommand\Time{N}
-\newcommand\Np{J}
-\newcommand\np{j}
-\newcommand\altNp{k}
-\newcommand\altAltNp{\tilde j}
-\newcommand\resampleIndex{r}
-\newcommand\unitWeight{w}
-
-%% for all iterated filtering algorithms
-\newcommand\Nit{M}
-\newcommand\nit{m}
-
-%% customized math macros
-\newcommand\prob{\mathbb{P}}
-\newcommand\E{\mathbb{E}}
-\newcommand\dd[1]{\mathrm{d}{#1}}
-\newcommand\given{{\,\vert\,}}
-\newcommand\equals{{{\,}={\,}}}
-\newcommand\myequals{\hspace{0.5mm}{=}\hspace{0.5mm}}
-\newcommand\myto{{\;:\;}}
-\newcommand\seq[2]{{#1}\!:\!{#2}}
-\newcommand\mydot{{\,\cdot\,}}
-\newcommand\cp[2]{N_{\mathrm{#1}\mathrm{#2}}}
-\newcommand\giventh{{\hspace{0.5mm};\hspace{0.5mm}}}
-\newcommand\normal{\mathcal{N}}
-\newcommand\argequals{{\,=\,}}
-\newcommand\lags{c}
-\newcommand\maxlag{\overline{c}}
-\newcommand\nlfList{C}
-\newcommand\bigO{\mathcal{O}}
-\newcommand\loglik{\lambda}
-\newcommand\loglikMC{\MC{\loglik}}
-\newcommand\R{\mathbb{R}}
-\newcommand\param{\,;}
-\newcommand\mycolon{{\hspace{0.6mm}:\hspace{0.6mm}}}
-\newcommand\MC[1]{#1^{\,\mbox{\tiny MC}}}
-\newcommand\EMC{{\E}}
-\newcommand\Var{\mathrm{Var}}
-\newcommand\var{\Var}
-\newcommand\Cov{\mathrm{Cov}}
-\newcommand\cov{\Cov}
-\newcommand\iid{\mathrm{iid}}
-\newcommand\dist{d}
-\newcommand\transpose{\top}
-
-\newcommand\gnsep{,}
-
-\def\lik{L}
-\def\loglik{\ell}
-
-%% THEOREM-LIKE ENVIRONMENTS
-\usepackage{amsthm}
-\newtheorem{prop}{Proposition}
-\newtheorem{theorem}{Theorem}
-\newtheorem{lemma}{Lemma}
-\newtheorem{example}{Example}
-\newtheorem{remark}{Remark}
-\newtheorem{cor}{Corrolary}
-\newtheorem{fact}{Fact}
-\newtheorem{assumption}{Assumption}
-\newtheorem{assumptionB}{Assumption}
-\renewcommand\theassumption{A\arabic{assumption}}
-\renewcommand\theassumptionB{B\arabic{assumptionB}}
-
-%% FOR PSEUDOCODE SETUP
-\newcommand\mystretch{\rule[-2mm]{0mm}{5mm} }
-\newcommand\asp{\hspace{4mm}}
-\newcommand\codeIndent{\hspace{4mm}}
-
-%%%%%%%%%%%%%%%% END OF USER DEFINITIONSS %%%%%%%%%%%%%%%
-
-\bibliographystyle{apalike}
-\usepackage{natbib}
-\usepackage{fullpage}
-
-\newcommand\myeqref[1]{(\ref{#1})}
-
-\renewcommand{\contentsname}{Supplementary Content}
-\renewcommand{\refname}{Supplementary References}
-\renewcommand\thefigure{S-\arabic{figure}}
-\renewcommand\thetable{S-\arabic{table}}
-\renewcommand\thesection{S\arabic{section}}
-\renewcommand\theequation{S\arabic{equation}}
-\renewcommand\theprop{S\arabic{prop}}
-\renewcommand\thelemma{S\arabic{lemma}}
-\renewcommand\thealgocf{S\arabic{algocf}}
-
-\setcounter{tocdepth}{1}
-
-\begin{document}
-
-\date{\today}
-\title{Supplement to ``{\mytitle}''}
-  \author{Jesse Wheeler, AnnaElaine L. Rosengart, Zhuoxun Jiang, \\ Kevin Tan, Noah Treutle and Edward L. Ionides
- \\ \hspace{.2cm}\\
-    Department of Statistics, University of Michigan\\
-}
-
-\newcommand{\blind}{1}
-
-\if1\blind
-{
-\maketitle
-}
-\fi
-
-\if0\blind
-{
-  \bigskip
-  \bigskip
-  \bigskip
-  \begin{center}
-    {\LARGE\bf \mytitle}
-\end{center}
-  \bigskip
-  \bigskip
-}\fi
-
-\tableofcontents
-
-\newpage
-
-\section{Markov chain and differential equation interpretations of compartment flow rates}
-
-In Sections 2.1, 2.2 and 2.3 of the main article, we define compartment models in terms of their flow rates.
-For a discrete population model, these rates define a Markov chain.
-For a continuous and deterministic model, the rates define a system of ordinary differential equations.
-Here, we add additional details to clarify the mapping from a collection of rate functions to a fully specified process.
-Our treatment follows \citet{breto09}.
-
-A general compartment model is a vector-valued process $X(t)=(X_1(t),\dots,X_c(t))$ denoting the (integer or real-valued) counts in each of $c$ compartments, where $t$ is any continuous value in the interval  $\left[t_0, \infty\right)$ for some real valued starting time $t_0$.
-The compartments may also have names, but to set up general notation we simply refer to them by their numerical index.
-The basic characteristic of a compartment model is that $X(t)$ can be written in terms of the flows $N_{ij}(t)$ from $i$ to $j$.
-A flow into compartment $i$ from outside the system is denoted by $N_{\demography i}$, and a flow out of the system from compartment $i$ is denoted by $N_{i\demography}$.
-We call $\demography$ a source/sink compartment, though it is an irregular compartment since $X_{\demography}(t)$ is not defined.
-These flows are required to satisfy a ``conservation of mass'' identity:
-\begin{equation}
-X_i(t)=X_i(t_0)+N_{\demography i}(t) - N_{i\demography}(t) + \sum_{j\neq
-i}N_{ji}(t)-\sum_{j\neq i}N_{ij}(t). \label{eq:conservation}
-\end{equation}
-Each {\em flow} $N_{ij}(t)$ is associated with a {\em rate} function
-$\mu_{ij}=\mu_{ij}(t,X(t))$, where we include the possibility that $i$ or $j$ takes value $\demography$.
-
-There are different ways to use a collection of rate functions to build a fully specified model.
-We proceed to describe the ones we use in this paper: via a system of ordinary differential equations (Sec.~\ref{subsec:ode}), a simple Markov counting system (Sec.~\ref{subsec:smcs}), and an overdispersed Markov counting system (Sec.~\ref{subsec:odmcs}).
-Other representations include stochastic differential equations driven by Gaussian noise or Gamma noise \citep{bhadra11}.
-
-\subsection{Ordinary differential equation (ODE) interpretation}
-\label{subsec:ode}
-
-A basic deterministic specification is
-\begin{equation}
-\label{eq:ode1}
-dN_{ij}/dt = \mu_{ij}\big(t,X(t) \big) X_i(t), \hspace{3mm} i\in \seq{1}{c}, \hspace{3mm} j\in \seq{1}{c} \cup\{\demography\}, \hspace{3mm} i\neq j,
-\end{equation}
-where $\mu_{ij}\big(t,X(t)\big)$ is called a per-capita rate or a unit rate.
-Flows into the system require special treatment since $X_i(t)$ in \myeqref{eq:ode1} is not defined for $i=\demography$.
-Instead, we specify
-\begin{equation}
-\label{eq:ode2}
-dN_{\demography i}/dt = \mu_{\demography i}\big(t,X(t) \big).
-\end{equation}
-This is the the interpretation and implementation used for Model~2 in our study.
-
-\subsection{Simple Markov counting system interpretation}
-\label{subsec:smcs}
-A continuous time Markov chain can be specified via its infinitesimal transition probabilities.
-A basic approach to this is to define
-\begin{eqnarray}
-\label{eq:smcs1}
-\prob\big[ N_{ij}(t+\delta)-N_{ij}(t)=0 \given X(t)\big]
- &=& 1-\delta \mu_{ij}\big(t,X(t)\big)X_i(t) + o(\delta),
-\\
-\label{eq:smcs2}
-\prob\big[ N_{ij}(t+\delta)-N_{ij}(t)=1 \given X(t)\big]
- &=& \delta \mu_{ij}\big(t,X(t)\big)X_i(t) + o(\delta),
-\end{eqnarray}
-for $i\in \seq{1}{c}$ and $j\in\seq{1}{c}\cup\{\demography\}$ with $i\neq j$.
-As with the ODE case, we need special attention for flows into the system, and we define
-\begin{eqnarray}
-\label{eq:smcs3}
-\prob\big[ N_{\demography i}(t+\delta)-N_{\demography i}(t)=0 \given X(t)\big]
- &=& 1-\delta \mu_{\demography i}\big(t,X(t)\big) + o(\delta),
-\\
-\label{eq:smcs4}
-\prob\big[ N_{\demography i}(t+\delta)-N_{\demography i}(t)=1 \given X(t)\big]
- &=& \delta \mu_{\demography i}\big(t,X(t)\big) + o(\delta).
-\end{eqnarray}
-Together with the initial conditions $X(0)$, equations \myeqref{eq:smcs1}--\myeqref{eq:smcs4} define a Markov chain.
-Each flow is a simple counting process, meaning a non-decreasing integer-valued process that only has jumps of size one.
-We therefore call the Markov chain a simple Markov counting system (SMCS).
-The infinitesimal mean of every flow is equal to its infinitesimal variance \citep{breto11} and so an SMCS is called equidispersed.
-We note that the special case of Model~1 used by \cite{lee20} (with $\sigmaProc = 0$) is an SMCS.
-To permit more general mean-variance relationships for a Markov counting system, we must permit jumps of size greater than one.
-The utility of overdispersed models, where the infinitesimal variance of the flow exceeds the infinitesimal mean, has become widely recognized \citep{stocks20,he10}.
-
-\subsection{Overdispersed Markov counting system interpretation}
-\label{subsec:odmcs}
-
-Including white noise in the rate function enables the possibility of an overdispersed Markov counting system \citep{breto11,breto09,he10}.
-Since rates should be non-negative, Gaussian noise is not appropriate and gamma noise is a convenient option that has found various applications \citep{romero-severson15, subramanian20}.
-Specifically, we consider a model given by
-\begin{equation}
-\label{eq:odmcs1}
-\mu_{ij}\big(t,X(t)\big) = \bar\mu_{ij}\big(t,X(t)\big) \, d\Gamma_{ij}(t)/dt,
-\end{equation}
-where $\Gamma_{ij}(t)$ is a stochastic process having independent gamma distributed increments, with
-\begin{equation}
-\label{eq:odmcs2}
-\E\big[\Gamma_{ij}(t)\big] = t, \quad \var\big[\Gamma_{ij}(t)\big] = \sigma_{ij}^2 t.
-\end{equation}
-Formally interpreting the meaning of \myeqref{eq:odmcs1} is not trivial, and we do so by constructing a Markov process $X(t)$ as the limit of the Euler scheme descrived in Section~\ref{sec:numerics}, below.
-Therefore, the numerical scheme in Sec.~\ref{sec:numerics} can be taken as a definition of the meaning of \myeqref{eq:odmcs1}.
-The Markov chain defined by the limit of this Euler scheme as the step size decreases is an overdispersed Markov counting system, with the possibility of instantaneous jumps of size greater than one \citep{breto11}.
-
-\section{Numerical solutions to compartment models}
-\label{sec:numerics}
-
-Models may be fitted and their implications assessed via numerical solutions (i.e., simulations) from the model equations.
-All the analyses we consider have this simulation-based property, known as plug-and-play or equation-free or likelihood-free.
-The numerical solutions to the model are arguably of more direct scientific interest than the exact solutions to the postulated equations.
-For ODE models, numerical methods are well studied and a standard numerical solution package such as \code{deSolve} in \code{R} is adequate for many purposes.
-For SMCS and ODMCS models, exact schemes are feasible when the number of events is small, which may be the case for small populations.
-However, for applicability to larger populations, we use instead the following Euler scheme.
-Write $\delta$ for an Euler time step, and $\Delta N_{ij}$ for the numerical approximation to $N_{ij}(t+\delta)-N_{ij}(t)$ given $X(t)$.
-For each $i$ and $j$ in $\seq{1}{c} \cup \{\demography\}$ with $i \neq j$, we draw independent Gamma distributed noise increments with mean $\delta$ and variance $\sigma_{ij}^2 \delta$, denoted using a mean-variance parameterization of the gamma distribution as
-\begin{equation}
-\label{eq:numerics1}
-\Delta\Gamma_{ij} \sim \mathrm{gamma}(\delta, \sigma_{ij}^2 \delta).
-\end{equation}
-In the case of an SMCS model, $\sigma_{ij}=0$ for all $i$ and $j$, so we have $\Delta\Gamma_{ij}=\delta$.
-Then, for $i\neq \demography$ and $j\neq i$, and writing
-\begin{equation}
-\label{eq:numerics2}
-\mu_{ij}=\bar\mu_{ij}\big(t,X(t)\big) \Delta\Gamma_{ij} / \delta,
-\end{equation}
-we calculate transition probabilities
-\begin{eqnarray}
-\label{eq:numerics3}
-p_{ij} &=& \exp\left\{-\sum_{k\in 1:c \, \cup \{\demography\}} \mu_{ik} \, \delta \right\}
-\frac{\mu_{ij}}{\sum_{k\in 1:c\cup \{\demography\}} \mu_{ik}},
-\\
-\label{eq:numerics4}
-p_{ii} &=& 1 - \sum_{j\neq i} p_{ij}.
-\end{eqnarray}
-These probabilities correspond to competing hazards for every individual in compartment $i$ to transition to some compartment $j$, interpreting $j=i$ to mean that the individual remains in $i$.
-Then, $\big(\Delta N_{i1},\dots,\Delta N_{ic},\Delta N_{i\demography}\big)$ has the multinomial distribution where $X_i(t)$ individuals are allocated independently to $\seq{1}{c}\cup\{\demography\}$ with probabilities given by \eqref{eq:numerics3} and \eqref{eq:numerics4}.
-We use the \code{reulermultinom} function in the \code{pomp} package to draw from this multionomial distribution.
-
-Different treatments of demographic flows---such as birth, death, immigration and emigration---are possible.
-For the case $i=\demography$, the treatment used by Model~1 is to set
-\begin{equation}
-\label{eq:numerics6}
-\Delta N_{\demography j} \sim \mathrm{poisson}( \mu_{\demography j} \delta),
-\end{equation}
-an independent Poisson random variable with mean $\mu_{\demography j} \delta$.
-
-Models~2 and 3 used an alternative approach, balancing the total number of flows in and out of the compartment, i.e., $\sum_{i}N_{\demography i}(t) = \sum_{i}N_{i \demography}(t)$, in order to make the model consistent with the known total population.
-In this case, we formally model the death rate as a rate of returning to the susceptible class $S$, and use external transitions from $\demography$ into $S$ to describe only net population increase.
-
 \section{Measurement Models}
@@
 We therefore do not allow a change in reporting rate when fitting Model~3.
-
-\section{Initial Values}
-
-To perform inference on POMP models, it is necessary to propose an initial probability density for the latent process $f_{X_0}(x_0;\theta)$, including the possibility that the initial values of the latent states are known, or are a non-random function of the unknown parameter vector, $\theta$.
-This density is used to obtain initial values of the latent state when fitting and evaluating the model.
-For each of the models considered in this analysis, the initial conditions are derived by enforcing the model dynamics on reported cholera cases.
-It is also sometimes necessary to fit some initial value parameters in order to help determine initial values for weakly identifiable compartments.
-In the following subsections, we mention initial value parameters that were fit for each model.
-
-\subsection{Model~1}
-
-For this model, the number of individuals in the Recovered and Asymptomatic compartments are set to zero, but the initial proportion of Infected and Exposed individuals is estimated as initial value parameters ($\Iinit$ and $\Einit$, respectively) using the IF2 algorithm, implemented as \texttt{mif2} in the \texttt{pomp} package.
-Finally, the initial proportion of Susceptible individuals $S_{0, 0}$ is calculated as $S_{0, 0} = 1 - \Iinit - \Einit$.
-This model for the initial values of the latent states matches that which was used by \citet{lee20}.
-
-\subsection{Model~2}
-
-Model~2 assumes that the initial values are a deterministic function of the reporting rate and the initial case reports, and so no initial value parameters need to be estimated.
-Initial values for latent state compartments are chosen so as to enforce the model dynamics on the observed number of cases.
-Specifically, the model sets $I_{u0}(0) = y^*_{1u} / \reportRate$ for each unit $u \in 1:10$, where $I_{u0}(t)$ is the number of infected individuals in unit $u$ at time $t$, vaccination scenario $z = 0$, $y^*_{tu}$ is the reported number of cases, and $\reportRate$ is the reporting rate.
-This model for the initial values of the latent states matches that which was used by \citet{lee20}.
-
-\subsection{Model~3}
-
-The latent states of this model are initialized by enforcing the model dynamics on the incidence data from the start of the recorded cases until time $t_0$, requiring that some of the available data be used to determine the initial values of the latent states.
-This is the same approach that was taken by \citet{lee20}, who used the value $t_0 = \text{2014-02-22}$; this choice of $t_0$ results in modeling roughly only $60\%$ of the available data, some of which is later discarded for alternative reasons \citep{lee20sup}.
-
-We do not see any immediate reason that this model could not be extended to cover a larger range of the data, and chose the value $t_0 = \text{2010-11-13}$.
-This choice of $t_0$ corresponds to using approximately $1\%$ of the available data to determine initial values of the latent states.
-In addition to modeling a larger portion of the available data, this choice of $t_0$ corresponds to an important real-world event, as daily reports from each of the departments were not being sent to the health ministry until November 10, 2010; % TODO: Cite
-this choice of $t_0$ therefore makes $\bm{Y}\big(t_1\big)$ the first week of data once daily reports are being sent to the health ministry.
-The few observation times that exist before $t_0$ are used to initialize the model by enforcing model dynamics on these preliminary observations.
-For convenience, we denote these observations as $t_{-3}, t_{-2}$ and $t_{-1}$; as before, we let $y_{u,-k} = Y_{u}\big(t_{-k}\big)$ denote the observed case count for unit $u$ at time point $t_{-k}$, where $k \in 1:3$.
-Equations for the initial values of non-zero latent states are provided in Eqs.~\myeqref{eq:model3:I0}--\myeqref{eq:model3:B0}.
-
-\begin{eqnarray}
-\label{eq:model3:I0}
-I_{u}\big(t_0\big) &=& \frac{y^*_{u, -1}}{7\reportRate\left(\muIR + \left(\muDeath + \choleraDeath\right)/365\right)},
-\\%[-3pt]
-\label{eq:model3:A0}
-A_{u}\big(t_0\big) &=& \frac{I_{u0}\big(t_0\big)\left(1 - \symptomFrac\right)}{\symptomFrac},
-\\
-\label{eq:model3:R0}
-R_{u01}\big(t_0\big) &=& R_{u02}\big(t_0\big) = R_{u03}\big(t_0\big) = \left(\frac{\sum_{k = -3}^0 y^*_{u, k}}{\reportRate \symptomFrac} - \left(I_{u0}\big(t_0\big) + A_{u0}\big(t_0\big)\right)\right) / 3
-\\
-\label{eq:model3:S0}
-S_{u0}\big(t_0\big) &=& \mathrm{Pop}_u - I_{u}\big(t_0\big) - A_{u}\big(t_0\big) - \sum_{k=1}^3 R_{u0k}\big(t_0\big)
-\\
-\label{eq:model3:B0}
-B_{u}\big(t_0\big) &=& \big[1 + \seasAmplitude \tilde{J}^r \big] \mathrm{Den}_u \, \Wshed \big[ I_{u}(t_0)+ \asymptomRelativeShed A_{u_0}(t) \big]/\Wshed.
-\end{eqnarray}
-
-In Eq.~\myeqref{eq:model3:B0}, $\tilde{J} = 0.002376$ is the median adjusted rainfall over the observation period.
-One important consideration to make with this parameter initialization model is when $y_{u, -1}^* = 0$, which occurs for units $u \in \{3, 4\}$, which correspond to the Grand'Anse and Nippes departments, respectively.
-When this is the case, each of the infectious $I_u(t_0)$, asymptomatic $A_u(t_0)$, and bacterial reservoir $W_u(t_0)$ compartments have a value of zero.
-Because Model~3 models cholera transmission primarily by means of the bacterial reservoir, this makes it nearly impossible for an outbreak to occur.
-Therefore for units $u \in \{3, 4\}$, we introduce initial value parameters $I_{0,0}^3$ and $I_{0,0}^4$, and calibrate these parameter values using the data.
-The resulting parameter estimates are used to obtain the remaining non-zero initial values of the latent states using Eqs.~\myeqref{eq:model3:A0}--\myeqref{eq:model3:B0}.
-
-% Argue why we didn't need to do this for Model 2.
-
-% TODO: Check this section.
-
-\section{Calibrating Model~3 to observed cases}\label{sec:mod3Cal}
-
-<<CalibrateOrig3>>=
-# Create vectors for the unit and shared parameters
-unit_specific_names <- c("betaB", "foi_add")
-shared_param_names <- c(
-  "mu_B", "XthetaA", "thetaI", "lambdaR", "r", "std_W",
-  "epsilon", "k"
-)
-est_param_names <- c(
-  unit_specific_names, shared_param_names
-)
-
-# Add unit numbers to each parameter
-est_param_names_expanded <- paste0(rep(est_param_names, each = 10), 1:10)
-
-# Simple function that is used to set the rw_sd for the first search
-# This function is convenient so that we don't have to type out the rw_sd
-# for each unit parameter.
-set_rw_1 <- function(x) {
-  if (gsub("[[:digit:]]+$", "", x) %in% shared_param_names) {
-    return(0.01)
-  # } else if (x %in% paste0(rep(c("aHur", 'hHur'), each = 2), c(3, 9))) {
-  #   return(expression(ifelse(time >= 2016.754 & time <= 2017, 0.015, 0)))
-  # } else if (x %in% c("Iinit3", "Iinit4")) {
-  #   return(expression(ivp(0.25)))
-  } else if (grepl('^foi_add[[:digit:]]+$', x)) {
-    return(0.015)
-  } else if (grepl('^betaB[[:digit:]]+$', x)) {
-    return(0.01)
-  } else {
-    return(0)
-  }
-}
-
-set_rw_2 <- function(x) {
-  if (gsub("[[:digit:]]+$", "", x) %in% shared_param_names) {
-    return(0.003)
-  # } else if (x %in% paste0(rep(c("aHur", 'hHur'), each = 2), c(3, 9))) {
-  #   return(expression(ifelse(time >= 2016.754 & time <= 2017, 0.005, 0)))
-  # } else if (x %in% c("Iinit3", "Iinit4")) {
-  #   return(expression(ivp(0.125)))
-  } else if (grepl('^foi_add[[:digit:]]+$', x)) {
-    return(0.005)
-  } else if (grepl('^betaB[[:digit:]]+$', x)) {
-    return(0.003)
-  } else {
-    return(0)
-  }
-}
-
-# First Search RW
-reg_rw_1.sd <- lapply(est_param_names_expanded, set_rw_1)
-names(reg_rw_1.sd) <- est_param_names_expanded
-chol_rw_1.sd <- do.call(rw_sd, reg_rw_1.sd)
-
-# Second Search RW
-reg_rw_2.sd <- lapply(est_param_names_expanded, set_rw_2)
-names(reg_rw_2.sd) <- est_param_names_expanded
-chol_rw_2.sd <- do.call(rw_sd, reg_rw_2.sd)
-
-SEARCH1 = list(
-  NBPF = 100,
-  NP = 2500,
-  SPAT_REGRESSION = 0.05,
-  NREPS = 144,
-  NP_EVAL = 2500,
-  NREPS_EVAL = 18,
-  RW_SD = chol_rw_1.sd,
-  COOLING = 0.5,
-  KEEP_TRACES = FALSE,
-  KEEP_LIKE_MAT = FALSE
-)
-
-SEARCH2 = list(
-  TOP_N = 12,
-  NBPF = 100,
-  NP = 2500,
-  SPAT_REGRESSION = 0.05,
-  NREPS = 6,
-  NP_EVAL = 2500,
-  NREPS_EVAL = 18,
-  RW_SD = chol_rw_2.sd,
-  COOLING = 0.5,
-  KEEP_TRACES = FALSE,
-  KEEP_LIK_MAT = TRUE
-)
-
-haiti3_sub_fit <- bake(
-  file = "../model3/si/subset_noparams.rds", {
-    fit_haiti3(
-      nsearches = 2L,
-      search1 = SEARCH1,
-      search2 = SEARCH2,
-      search_rho = FALSE,
-      search_gamma = FALSE,
-      search_Iinit = FALSE,
-      search_hur = FALSE,
-      ncores = cores
-    )
-  }
-)
-
-# Find model index with best log-likelihood
-best_m <- which.max(haiti3_sub_fit$search2$logLiks$logLik)
-
-# Find parameters using best_m index
-sub_params <- haiti3_sub_fit$search2$params[best_m, ]
-
-# Get unit log-likelihoods for the best model.
-all_liks <- as.data.frame(haiti3_sub_fit$search2$likMat)
-# all_liks$which <- rep(1:(SEARCH3$TOP_N * SEARCH3$NREPS), each = SEARCH3$NREPS_EVAL)
-all_liks$which <- rep(1:(12 * 6), each = 18)
-best_mod <- filter(all_liks, which == best_m)
-
-# Pivot longer to get estimate of log-likelihood for each unit.
-unitLikesM3 <- best_mod %>%
-  pivot_longer(
-    cols = -which,
-    names_to = "unit",
-    names_prefix = "V",
-    values_to = "logLik"
-  ) %>%
-  select(-which) %>%
-  group_by(unit) %>%
-  summarize(
-    ll_est = logmeanexp(logLik),
-    ll_se = logmeanexp(logLik, se = TRUE)[2]
-  ) %>%
-  ungroup() %>%
-  mutate(unit = as.numeric(unit)) %>%
-  arrange(unit)
-@
-
-<<M3_bench, echo=FALSE, message=FALSE, include=FALSE, cache=TRUE>>=
-h3 <- haiti3_spatPomp()
-deps <- h3@unit_names
-
-mod3_cases <- as.data.frame(t(h3@data))
-colnames(mod3_cases) <- deps
-m3_inits <- c(7362, 1914, 22, 8, 2818, 82, 2499, 6953, 44, 7) # TODO: Also fit by hand, but could be changed to initializing to the observation before.
-names(m3_inits) <- deps
-
-bench_cond_lls <- list()
-for (dep in deps) {
-
-  # Get department cases
-  cases <- as.numeric(mod3_cases[, dep])
-  mod <- ar1_NegBinom(data = cases, init = m3_inits[dep])
-
-  bench_cond_lls[[dep]] <- haitipkg:::.ar1_NegBinom_cond_ll(
-    theta = mod$theta, init = m3_inits[dep], data = cases, transform = FALSE
-  )
-}
-
-bench_cond_lls <- as.data.frame(bench_cond_lls)
-
-bench_lls <- apply(bench_cond_lls, 2, sum)
-
-unitLikes_bench <- data.frame(
-  unit = 1:10,
-  # dep = names(bench_lls),
-  ll = bench_lls
-) %>%
-  as_tibble()
-
-bench_cond_lls$date <- lubridate::round_date(lubridate::date_decimal(h3@times), unit = 'day')
-@
-
-<<combine_bench_M3, echo=FALSE>>=
-
-# Combine tables with model 3 and benchmark likelihoods.
-bench_M3_joined <- dplyr::left_join(
-  x = unitLikesM3,
-  y = unitLikes_bench,
-  by = "unit"
-) %>%
-  select(-ll_se) %>%
-  mutate(unit = as.character(unit)) %>%
-  pivot_longer(
-    cols = -unit,
-    names_to = "type",
-    values_to = "ll"
-  ) %>%
-  mutate(unit = paste0(unit, ": ", deps[as.numeric(unit)]))
-@
-
-<<M3bpf, echo=FALSE, message=FALSE, include=FALSE>>=
-
-# Change model parameters to the best set of parameters from previous
-# search
-coef(h3)[names(sub_params)] <- sub_params
-
-registerDoRNG(40765101)  # for modular, reproducible results
-
-# Perform block partical filter with fitted model, to get conditional
-# log-likelihoods.
-bpf_sub3_out <- bake(
-  file = "../model3/si/bpf_sub_cond_LL.rds", {
-
-    results <- list()
-
-    bpf_out <- h3 %>% bpfilter(
-      Np = 5000,
-      block_size = 1,
-      save_states = TRUE
-    )
-
-    results$block.cond.loglik <- bpf_out@block.cond.loglik
-    results$times <- bpf_out@times  # Saved for convenience
-    results
-  },
-  seed = 432235
-)
-@
-
-In this section, we provide more detail on the process that was used to estimate the coefficients used in Model~3.
-In particular, we discuss why we decided to include additional model parameters---those that are associated with the behavior of the system during Hurricane Matthew---that were not considered by \cite{lee20}.
-To calibrate this model, we used the iterated block particle filter (IBPF) method of \cite{ionides22}.
-Due to the novelty of this algorithm, there does not exist published examples the IBPF algorithm used for data analysis outside of the papers in which the algorithm was introduced \citep{ning21ibpf,ionides22}, which at least partially motivates the inclusion additional details in this section.
-
-\cite{lee20} were only able to estimate model parameters to a simplified version of Model~3 on a subset of the available data, as no method existed at the time of their publication to fit a fully coupled meta-population model to disease incidence data.
-Building on their results, we fit the fully coupled version of Model~3 to nearly all available data, reserving only a few observations to use to calibrate the initial conditions of the model (see the previous section for more details).
-In order to properly maximize the model likelihood, we found that it was necessary to use multiple searches for the MLE, periodically pruning away less successful searches.
-The first collection of searches was performed by obtaining initial values for the parameters by uniformly sampling values from a predefined hypercube.
-A subsequent refinement search used starting values corresponding to the largest model likelihoods obtained in the previous round.
-This is similar to the approach used by \cite{ionides22}.
-
-We use the technique described above to fit the fully coupled version of Model~3 proposed by \cite{lee20}.
-The maximum likelihood we obtained with the model was $\Sexpr{round(max(haiti3_sub_fit$search2$logLiks$logLik))}$, which is higher than the benchmark model ($\Sexpr{round(sum(unitLikes_bench$ll))}$).
-While beating a simple associative benchmark is promising, this does not immediately rule out the possibility that the model is a poor description of the system.
-Additional investigation of parameters estimates and their corresponding implications on model based conclusions should always be conducted.
-For meta-population models, it is worth considering how well the model fits the data to each spatial unit.
-This can be done by looking at conditional log-likelihoods, which is part of the output of the \texttt{bpfilter} algorithm in the \texttt{spatPomp} package.
-The likelihoods for each department, compared to the corresponding benchmark model, are displayed in Fig.~\ref{fig:h3UnitLikes}.
-The figure demonstrates that while the log-likelihood of the fitted model outperforms the auto-regressive negative binomial benchmark model, Model~3 has lower likelihoods than the benchmark for some departments.
-
-\begin{figure}[ht]
-<<h3UnitLikes, fig.height=3, fig.width=5.5, echo=FALSE>>=
-ggplot(bench_M3_joined, aes(x = ll, y = reorder(unit, ll))) +
-  geom_point(aes(col = type)) +
-  ylab("Unit") +
-  xlab("Log Likelihood") +
-  scale_color_manual(
-    labels = c("ll" = "Benchmark", "ll_est" = "Model 3"),
-    values = c("ll" = "#1b9e77", "ll_est" = "#d95f02")
-  ) +
-  theme(legend.title = element_blank(),
-        axis.title.y = element_blank())
-@
-\caption{\label{fig:h3UnitLikes}Log-likelihoods of Model~3 for each department compared to the corresponding benchmark model prior to the inclusion of parameters for initial values and Hurricane Matthew.}
-\end{figure}
-
-In addition to considering the conditional log-likelihoods of each unit, one can consider conditional log-likelihoods of each observation.
-When compared to a benchmark, this level of detail can provide useful information about which observations are well described by the model and which are poorly described.
-In Fig.~\ref{fig:condLL}, we plot the conditional log-likelihoods of each observation.
-This figure reveals that the fitted model has difficult time explaining certain features of the data.
-For example, many departments (in particular Sud) have observations with lower conditional log-likelihoods near October 2016 than at other time points.
-Further investigation of the model output reveals that the model is struggling to explain the sudden surge in cholera cases that occurred at this time, which coincides with the time that Hurricane Matthew struck Haiti.
-While the model does include a mechanism to account for increased cholera transmission due to large rainfall events, the mechanism does not appear to be sufficient to capture the damaging effects of the hurricane, which had the greatest impact in the the Sud and Grand'Anse departments \citep{ferreirai16}.
-This result led us to include parameters $\Whur{u}$ and $\hHur{u}$ in Eq.~23 of the main text, which allows for an increase in the transmission rate between environmental cholera and humans for Sud and Grand'Anse during and after the hurricane.
-The effect of the hurricane on cholera transmission is assumed to have an exponential decay, where the magnitude is determined by $\Whur{u}$ and the duration of the effect determined by $\hHur{u}$.
-
-\begin{figure}[ht]
-<<CondLLFig, echo=FALSE, message=FALSE, fig.height=3>>=
-
-# Changing labels to human-readable labels.
-deps[grepl("_", deps)] <- gsub("_", "-", deps[grepl("_", deps)])
-deps[deps == "Grande-Anse"] <- "Grand\'Anse"
-names(deps) <- as.character(1:10)
-myLabeller <- as_labeller(deps)
-
-# Plot conditional log-likelihoods for each unit
-t(bpf_sub3_out$block.cond.loglik) %>%
-  as.data.frame() %>%
-  mutate(t = bpf_sub3_out$times) %>%
-  pivot_longer(
-    cols = -t,
-    values_to = 'likelihood',
-    names_prefix = "V",
-    names_to = 'Unit'
-  ) %>%
-  mutate(date = as.Date(lubridate::date_decimal(t))) %>%
-  ggplot(aes(x = date, y = likelihood)) +
-  geom_line() +
-  geom_point() +
-  facet_wrap(~Unit, labeller = myLabeller, scales = 'free_y',
-             nrow = 2) +
-  theme_bw() +
-  theme(axis.title.x = element_blank(),
-        axis.text.x = element_text(angle = -45, hjust = 0)) +
-  ylab("Conditional log-likelihoods")
-@
-\caption{\label{fig:condLL}Conditional log-likelihoods of Model~3 prior to the inclusion of parameters for initial values and Hurricane Matthew.}
-\end{figure}
-
-Other points of concern include the observations at or near the start of the recorded cases, for example, in Nord-Ouest and Sud-Est.
-These low conditional log-likelihoods are a result of having zero or near zero confirmed cases in the first week of observed data, followed by a rapid growth of cases.
-The model cannot explain the rapid growth of cholera in the absence of starting cases because the initialization scheme forces departments with no recorded cases in the first week to have hidden states $I_{u0}(0) = A_{u0}(0) = B_{u}(0) = 0$.
-In reality, however, it is possible to have some infected or asymptomatic individuals in a department when no cases are recorded due to under-reporting.
-This may be particularly true at the onset of an infectious disease outbreak, as lower public awareness and inadequate testing infrastructure may result in many cases going undetected.
-We therefore include and estimate $I_{0u}$ that determines the initial number of Infected individuals at the start of the pandemic for the departments Grande'Anse, Nord-Ouest, and Sud-Est;
-each of these departments have both low starting case counts followed by rapid growth, and low initial values of the conditinoal log-likelihood.
-
-<<loadFinalModel3, echo=FALSE, include=FALSE, message=FALSE>>=
-
-# This result was used in the manuscript, so it doesn't need to be recalculated.
-h3_final_results <- readRDS("../model3/run_level_3/haiti3_fit.rds")
-
-# Save best_m as index for the model parameters with the best likelihood.
-# Then, save the model parameter values and the likelihood evaluations
-best_m <- which.max(h3_final_results$search2$logLiks$logLik)
-best_params <- h3_final_results$search2$params[best_m, ]
-all_liks <- as.data.frame(h3_final_results$search2$likMat)
-all_liks$which <- rep(1:72, each = 18)
-best_mod <- dplyr::filter(all_liks, which == best_m)
-
-# Get repeated estimates of unit log-likelihoods
-unitLikesM3 <- best_mod %>%
-  pivot_longer(
-    cols = -which,
-    names_to = "unit",
-    names_prefix = "V",
-    values_to = "logLik"
-  ) %>%
-  select(-which) %>%
-  group_by(unit) %>%
-  summarize(
-    ll_est = pomp::logmeanexp(logLik),
-    ll_se = pomp::logmeanexp(logLik, se = TRUE)[2]
-  ) %>%
-  ungroup() %>%
-  mutate(unit = as.numeric(unit)) %>%
-  arrange(unit)
-
-# Combine model likelihood table to benchmarks
-bench_M3_joined <- dplyr::left_join(
-  x = unitLikesM3,
-  y = unitLikes_bench,
-  by = "unit"
-) %>%
-  select(-ll_se) %>%
-  mutate(unit = as.character(unit)) %>%
-  pivot_longer(
-    cols = -unit,
-    names_to = "type",
-    values_to = "ll"
-  ) %>%
-  mutate(unit = paste0(unit, ": ", deps[as.numeric(unit)]))
-@
-
-We refit Model~3 after introducing these seven parameters---three initial value parameters and four parameters related to Hurricane Matthew---using the same successive search scheme described above.
-The resulting model has a log-likelihood value of $\Sexpr{round(max(h3_final_results$search2$logLiks$logLik), 1)}$.
-We also provide figures for department specific log-likelihoods and conditional log-likelihoods in Figs.~\ref{fig:finalUnitLL} and \ref{fig:finalCondLL}.
-
-\begin{figure}[ht]
-<<finalUnitLLFig, fig.height=3, fig.width=5.5, echo=FALSE>>=
-ggplot(bench_M3_joined, aes(x = ll, y = reorder(unit, ll))) +
-  geom_point(aes(col = type)) +
-  ylab("Unit") +
-  xlab("Log Likelihood") +
-  scale_color_manual(
-    labels = c("ll" = "Benchmark", "ll_est" = "Model 3"),
-    values = c("ll" = "#1b9e77", "ll_est" = "#d95f02")
-  ) +
-  theme(legend.title = element_blank(),
-        axis.title.y = element_blank())
-@
-\caption{\label{fig:finalUnitLL}Log-likelihoods of Model~3 for each department compared to the corresponding benchmark model after adding and estimating parameters related to initial values and Hurricane Matthew.}
-\end{figure}
-
-<<finalModel3BPF, echo=FALSE, message=FALSE, include=FALSE>>=
-
-# Set the model parameters to the best results from previous search
-coef(h3)[names(best_params)] <- best_params
-
-# block particle filter to get conditional log-likelihoods.
-bpf_final3_out <- bake(
-  file = "../model3/si/h3_final_bpf_cond_ll.rds", {
-
-    results <- list()
-
-    bpf_out <- h3 %>% bpfilter(
-      Np = 5000,
-      block_size = 1,
-      save_states = TRUE
-    )
-
-    results$Sus_df <- foreach(i=1:length(h3@times), .combine = bind_rows) %do% {
-      get_h3_S_quants(bpf_out, i)
-    }
-
-    results$block.cond.loglik <- bpf_out@block.cond.loglik
-    results$times <- bpf_out@times  # saved for convenience
-    results
-  },
-  seed = 786321
-)
-@
-
-\begin{figure}[ht]
-<<finalCondLLFig, echo=FALSE, message=FALSE, fig.height=3>>=
-t(bpf_final3_out$block.cond.loglik) %>%
-  as.data.frame() %>%
-  mutate(t = bpf_final3_out$times) %>%
-  pivot_longer(
-    cols = -t,
-    values_to = 'likelihood',
-    names_prefix = "V",
-    names_to = 'Unit'
-  ) %>%
-  mutate(date = as.Date(lubridate::date_decimal(t))) %>%
-  ggplot(aes(x = date, y = likelihood)) +
-  geom_line() +
-  geom_point() +
-  facet_wrap(~Unit, labeller = myLabeller, scales = 'free_y',
-             nrow = 2) +
-  theme_bw() +
-  theme(axis.title.x = element_blank(),
-        axis.text.x = element_text(angle = -45, hjust = 0)) +
-  ylab("Conditional log-likelihoods")
-@
-\caption{\label{fig:finalCondLL}Conditional log-likelihoods of Model~3 after adding and estimating the parameters for initial values and Hurricane Matthew.}
-\end{figure}
-
-While the addition of the Hurricane parameters seems to have resulted in a dramatic increase of conditional likelihoods around and after October 2016, the inclusion of the initial value parameters does not appear to have fixed the low conditional log-likelihood values in Sud-Est and Nord-Ouest.
-This could be due to many factors, including the possibility that the coupling between units forces low initial value parameters in a given department in order for the model to fit the data in the remaining departments.
-Whatever the case may be, the inclusion of the seven parameters resulted in a $\Sexpr{round(max(h3_final_results$search2$logLiks$logLik) - max(haiti3_sub_fit$search2$logLiks$logLik), 1)}$ increase in log-likelihood, which is a highly statistically significant difference.
-One could similarly test whether the inclusion of the initial value parameters resulted in a significantly better fit to the observed data, but we do not do this here due to the large computational effort that would be required.
-
-\subsection{Examining the Hidden States of the Calibrated Model}
-
-For mechanistic models, beating a suitable statistical benchmark does not alone guarantee that the model provides an accurate description of a dynamic process.
-Indeed, a good statistical fit does not require the model to assert a causal explanation.
-For example, reconstructed latent variables should make sense in the context of alternative measurements of these variables \citep{grad12}.
-We demonstrate this principle by examining the latent state of the calibrated model.
-In particular, we examine the compartment of susceptible individuals under various scenarios, as this compartment drives transmission dynamics in SIR models for infectious disease outbreaks.
-
-Recall that the filtering distribution for the calibrated version of Model~3 at time $t_k$ is defined as the distribution of the latent state at time $t_k$ given the data from times $t_{1}:t_{k}$, i.e.
-$f^{(3)}_{\bm{X}_k|\bm{Y}_{1:k}}(\bm{x}_{k} | \bm{y}^*_{1:k} ; \hat\theta)$.
-In general, one may expect simulations from the filtering distribution of a model with a good statistical fit to result in hidden states that are highly consistent with the observed data because the filtering distribution is conditioned on the observed data.
-Fig.~\ref{fig:h3Sus} shows the percentage of individuals that are in the susceptible compartment from various simulations of the model:
-simulations from Model~3 under initial conditions are displayed in red; simulations from the filtering distribution of model are displayed in blue.
-This figure shows that simulations from initial conditions tends to result in a much more rapid depletion of susceptible individuals at the start of the epidemic than simulations from the filtering distribution, suggesting the calibrated model has a propensity to predict larger outbreaks than what is typically seen in the data.
-This also appears true in later stages of the time period of observed data, where the data suggests that there is a replenishment of the susceptible compartment, but the Model simulations typically retain a smaller proportion of susceptible individuals.
-This result demonstrates that the calibrated model favors a more rapid growth in cholera cases than what is typically seen in the observed data, which explains why the model fails to predict the absence of cases between February, 2019 and October, 2022.
-
-<<h3SusFigSetup, echo=FALSE, message=FALSE, include=FALSE>>=
-
-# This bake perfoms the following:
-#  (1) simulates the model from initial conditions
-#  (2) get the values of the hidden states at the end of each simulation, and
-#      saves these for later use.
-#  (3) Obtain quantiles for the Susceptible pool from the simulations.
-init_conditions_results <- bake(
-  file = "../model3/si/InitSimsQuants.rds", {
-
-    results <- list()  # To return final output of bake
-
-    # Simulate from the fitted model
-    h3Spat_sims <- simulate(
-      h3, nsim = 2000, format = 'data.frame'
-    )
-
-    # Get states at the end of the simulations
-    h3_spat_end <- h3Spat_sims %>%
-      filter(time == max(time))
-
-    # Save hidden state values at the end of the simulation. Saved as a matrix
-    # in the same format as the output from a block-particle filter, so that
-    # it can be used as the input to project_from_filter.
-    results[["end_conditions"]] <- h3_spat_end %>%
-      select(-time, -cases) %>%
-      mutate(
-        unit = case_when(
-          unitname == "Artibonite" ~ 1,
-          unitname == "Centre" ~ 2,
-          unitname == "Grande_Anse" ~ 3,
-          unitname == "Nippes" ~ 4,
-          unitname == "Nord" ~ 5,
-          unitname == "Nord_Est" ~ 6,
-          unitname == "Nord_Ouest" ~ 7,
-          unitname == "Ouest" ~ 8,
-          unitname == "Sud" ~ 9,
-          unitname == "Sud_Est" ~ 10
-        )
-      ) %>%
-      select(-unitname) %>%
-      pivot_wider(
-        data = .,
-        names_from = unit,
-        values_from = c(S, I, A, R_one, R_two, R_three, VSd, VR1d, VR2d,
-                        VR3d, VSdd, VR1dd, VR2dd, VR3dd, VSd_alt, VR1d_alt,
-                        VR2d_alt, VR3d_alt, VSdd_alt, VR1dd_alt, VR2dd_alt,
-                        VR3dd_alt, C, B, Doses, totInc),
-        names_sep = ""
-      ) %>%
-      select(-.id) %>%
-      as.matrix() %>%
-      t()
-
-    # Get the quantiles for the Susceptible pool from the simulations
-    results[["SQuants"]] <- h3Spat_sims %>%
-      rename(dep = unitname) %>%
-      group_by(dep, time) %>%
-      summarise(
-        Q025 = quantile(S + VSd + VSdd + VSd_alt + VSdd_alt, probs = 0.025),
-        Q50  = quantile(S + VSd + VSdd + VSd_alt + VSdd_alt, probs = 0.5),
-        Q975 = quantile(S + VSd + VSdd + VSd_alt + VSdd_alt, probs = 0.975)
-      ) %>%
-      ungroup() %>%
-      rowwise() %>%
-      mutate(
-        unit_num = which(h3@unit_names == dep),
-        date = lubridate::date_decimal(time),
-        Q025 = Q025 / h3@params[paste0("H", unit_num)],
-        Q50  = Q50  / h3@params[paste0("H", unit_num)],
-        Q975 = Q975 / h3@params[paste0("H", unit_num)]
-      ) %>%
-      filter(
-        date >= as.Date(
-          lubridate::round_date(lubridate::date_decimal(h3@t0), unit = 'day')
-        )
-      ) %>%
-      mutate(date = as.Date(lubridate::round_date(date, unit = 'day')))
-
-    results
-  },
-  seed = 834687
-)
-
-# Get the output of the "bake" call for plotting.
-h3Spat_quants <- init_conditions_results[["SQuants"]]
-
-# This bake call uses the ending hidden states from above to forecast cholera
-# cases in the future under the V0 vaccination scenario
-mod3_V0_initcond_res <- bake(
-  file = '../model3/si/vacc_sim_from_init.rds', {
-    h3_V0_res <- list()
-
-    # Get "noVacc" scenario parameters
-    h3_S0_par <- get_model3_vacc_scenario_params(scenario = "noVacc")
-    h3Spat <- h3
-    p3 <- best_params
-    coef(h3Spat) <- p3
-    coef(h3Spat)[names(h3_S0_par)] <- h3_S0_par
-
-    # store the end states that will be used as initial values
-    h3_bpf_end_states <- init_conditions_results$end_conditions
-
-    # Using parameters and end state values, obtain projections
-    h3_V0_sims <- project_from_filter(
-      h3Spat, end_states = h3_bpf_end_states, covarGen = project_rain,
-      nsims = 1000, seed = 184697
-    )
-
-    # Store nationally aggregated results
-    h3_V0_res$mod3_V0_sims <- agg_mod3_sims(h3_V0_sims)
-    h3_V0_res$mod3_V0_probs <- get_elimProbs(h3_V0_sims, model = 3)
-
-    # Get quantiles of observed cases for each department and time
-    h3_V0_res$UnitQuants <- h3_V0_sims %>%
-      mutate(
-        date = as.Date(  # Get date object from decimal time
-          lubridate::round_date(lubridate::date_decimal(time), unit = 'day')
-        )
-      ) %>%
-      group_by(unitname, date) %>%
-      summarize(
-        Q025 = quantile(cases, probs = 0.025),
-        Q50 = quantile(cases, probs = 0.5),
-        Q975 = quantile(cases, probs = 0.975)
-      ) %>%
-      ungroup() %>%
-      rename(dep = unitname)
-
-    # Get quantiles of susceptible individuals for each department and time
-    h3_V0_res$SQuants <- h3_V0_sims %>%
-      mutate(
-        date = as.Date(
-          lubridate::round_date(lubridate::date_decimal(time), unit = 'day')
-        )
-      ) %>%
-      group_by(unitname, date) %>%
-      mutate(
-        # Get the population for each unit
-        pop = h3@params[paste0("H", which(first(unitname) == unit_names(h3)))]
-      ) %>%
-      summarize(  # Calculate the quantiles
-        Q025 = quantile(
-          (S + VSd + VSdd + VSdd_alt + VSd_alt) / pop,
-          probs = 0.025
-        ),
-        Q50 = quantile(
-          (S + VSd + VSdd + VSdd_alt + VSd_alt) / pop,
-          probs = 0.5
-        ),
-        Q975 = quantile(
-          (S + VSd + VSdd + VSdd_alt + VSd_alt) / pop,
-          probs = 0.975
-        )
-      ) %>%
-      ungroup() %>%
-      rename(dep = unitname)
-
-    h3_V0_res
-  }
-)
-
-# This result was already used in the manuscript; it doesn't need to
-# be recalculated.
-mod3_V0_res <- readRDS("../model3/run_level_3/mod3_V0_res.rds")
-@
-
-\begin{figure}[ht]
-<<h3SusFig, echo=FALSE, message=FALSE, fig.height=3>>=
-ggplot() +
-  # Sims from initial condition
-  geom_line(data = h3Spat_quants, aes(x = date, y = Q50), col = 'red') +
-  geom_ribbon(
-    data = h3Spat_quants,
-    aes(x = date, ymin = Q025, ymax = Q975),
-    fill = 'red', alpha = 0.5
-  ) +
-  # projections from initial condition, using filtered sims function
-  geom_line(data = mod3_V0_initcond_res$SQuants, aes(x = date, y = Q50), col = 'red') +
-  geom_ribbon(
-    data = mod3_V0_initcond_res$SQuants,
-    aes(x = date, ymin = Q025, ymax = Q975),
-    fill = 'red', alpha = 0.5
-  ) +
-  # Sims from filtered distribution
-  geom_line(
-    data = bpf_final3_out$Sus_df,
-    aes(x = as.Date(lubridate::date_decimal(time)), y = Q50),
-    col = 'blue'
-  ) +
-  geom_ribbon(
-    data = bpf_final3_out$Sus_df,
-    aes(x = as.Date(lubridate::date_decimal(time)), ymin = Q025, ymax = Q975),
-    fill = 'blue', alpha = 0.5
-  ) +
-  # Projections from filtered distribution
-  geom_line(data = mod3_V0_res$SQuants, aes(x = date, y = Q50), col = 'blue') +
-  geom_ribbon(
-    data = mod3_V0_res$SQuants,
-    aes(x = date, ymin = Q025, ymax = Q975),
-    fill = 'blue', alpha = 0.5
-  ) +
-  geom_vline(xintercept = max(h3Spat_quants$date), linetype = 'dashed') +
-  facet_wrap(~dep, scales = 'free_y', nrow = 2) +
-  theme_bw() +
-  theme(axis.title.x = element_blank(),
-        axis.text.x = element_text(angle = -45, hjust = 0)) +
-  ylab("Total Susceptible (%)")
-@
-\caption{\label{fig:h3Sus}Percentage of individuals that are in the susceptible compartment.
-Simulations from Model~3 under initial conditions are displayed in red; simulations from the filtering distribution of model are displayed in blue.
-The dashed line represents the end of the observed data.}
-\end{figure}
-
-\clearpage
-
-\section{Replication of \cite{lee20}}
-
-In this article we claimed that we were able to obtain better fits to the observed data using the same models that were proposed by \cite{lee20}.
-Along with visual comparisons to the data, this claim was supported by comparing likelihoods and AIC values in Table~2 in the manuscript. % TODO: Double check the table number matches
-Because model likelihoods were not provided by \cite{lee20}, it is necessary to replicate these models in order to obtain likelihood estimates.
-Here we would like to thank the authors of \cite{lee20}, who provided detailed descriptions of their models, which enabled us to build on their work.
-In the following subsections, we use our \code{R} package \code{haitipkg} to reproduce some of the results of \cite{lee20}.
-This reproduction allows us to estimate the likelihoods of the \cite{lee20} version of Models~1--3, and also provides a demonstration of the importance and usefulness of reproducible research.
-
-\subsection{Model~1 Replication}
-
-The model was implemented by a team at Johns Hopkins Bloomberg School of Public Health (hereafter referred to as the Model~1 authors) in the \code{R} programming language using the \code{pomp} package \citep{king16}.
-Original source code is publicly available with DOI: 10.5281/zenodo.3360991.
-The final results reported by the Model~1 authors were obtained by using several different parameter sets rather than a single point estimate.
-According to the supplement materials, this was because model realizations from a single parameter set retained substantial variability, but multiple realizations from a collection of parameter sets resulted in a reasonable visual fit to the data.
-We are also inclined to believe that the use of multiple parameter values was in part intended to account for parameter uncertainty---the importance of which was discussed in the main text---an effort by the Model~1 authors that we applaud.
-Simulations from each of the parameter sets, however, were treated with equal importance when being used to diagnose the model fit and make inference on the system.
-This is problematic given Figures S8 and S9 of the supplement material, which suggest that some parameter sets that were used for inference may have been several hundred units of log-likelihood lower than other parameter sets that were simultaneously used to make forecasts.
-Such a large difference in log-likelihoods is well beyond the threshold of statistical uncertainty determined by Wilks' theorem, resulting in the equal use of statistically inferior parameter sets in order to make forecasts and conduct inference on the system.
-
-To fully reproduce the results of the Model~1 authors, it is necessary to use the exact same set of model parameters that were originally used to obtain the results presented by \cite{lee20}.
-Because these parameter sets were not made publicly available, we relied on the source code provided by the Model~1 authors to approximately recreate the parameter set.
-Due to software updates since the publication of the source code, we were unable to produce the exact same set of parameters.
-Running the publicly available source code, however, resulted in a set of parameters that are visually similar to those used by the Model~1 authors (See Figures~\ref{fig:PlotEpiDist} and \ref{fig:plotEndParams}).
-Furthermore, simulations using the set of parameters produced by the source code appear practically equivalent to those displayed by \cite{lee20} (See Figure~\ref{fig:plotMod1Sims}).
-
-<<Simulate Lee Model 1, echo=FALSE, cache=TRUE>>=
-h1_epi <- haiti1()
-N_LEE_SIMS <- 20
-
-registerDoRNG(18599687)
-foreach(
-  i = 1:length(h1LeeStartsEpi),
-  .combine = rbind
-) %dopar% {
-  guess <- h1LeeStartsEpi[i, ]
-  sims <- simulate(h1_epi, nsim = N_LEE_SIMS, format = 'data.frame',
-                   params = guess)  # 20 sims for each set of parameters
-  sims$param_set <- i
-  sims %>%
-    dplyr::select(.id, week, cases, param_set)
-} -> all_sims
-@
-
-<<PlotEpiDist, echo=FALSE, fig.cap="Compare to Figure S8 of Lee et al. (2020) supplement.", fig.width=7, fig.height=7, fig.align='center'>>=
-GGally::ggpairs(
-  h1LeeStartsEpi,
-  columns = c("rho", "tau", "beta1", "nu", "loglik")
-)
-@
-
-<<SimulateEpi, echo=FALSE, include=FALSE, message=FALSE, cache=TRUE>>=
-h1_end <- haiti1(period = "endemic")
-
-true_agg_cases <- haitiCholera %>%
-  select(-report) %>%
-  pivot_longer(
-    data = .,
-    cols = -date_saturday,
-    values_to = 'cases',
-    names_to = 'dep'
-  ) %>%
-  mutate(
-    date = as.Date(date_saturday),
-    dep = gsub("\\.", "_", dep)
-  ) %>%
-  mutate(
-    dep = case_when(dep == "Grand_Anse" ~ "Grande_Anse", TRUE ~ dep)
-  ) %>%
-  tidyr::pivot_wider(
-    data = .,
-    id_cols = c(date),
-    names_from = dep,
-    values_from = cases,
-    names_prefix = 'cases_'
-  ) %>%
-  mutate(
-    ReportedAll = cases_Artibonite + cases_Centre +
-      cases_Grande_Anse + cases_Nippes + cases_Nord +
-      cases_Nord_Est + cases_Ouest + cases_Sud +
-      cases_Sud_Est + cases_Nord_Ouest
-  )
-@
-
-<<FitEnd, echo=FALSE, message=FALSE, include=FALSE>>=
-foreach(
-  i = 1:nrow(h1LeeStartsEnd),
-  .combine = rbind
-) %dopar% {
-  r_params <- h1LeeStartsEnd[i, ][names(coef(h1_end))]
-  coef(h1_end) <- r_params
-
-  sims <- simulate(
-    h1_end, nsim = N_LEE_SIMS,
-    format = 'data.frame',
-    params = r_params
-  )
-  sims$param_set <- i
-  sims %>%
-    dplyr::select(.id, week, cases, param_set)
-} -> end_sims
-
-@
-
-<<plotEndParams, fig.width=7, fig.height=7, fig.align='center', fig.cap="Bivariate relationships between variables after fitting endemic period.
-Compare to S9 of Lee et al.
-(2020) supplement.", cache=TRUE, echo=FALSE>>=
-GGally::ggpairs(
-  h1LeeStartsEnd,
-  columns = c("rho", "tau", "beta1", "nu", "loglik")
-)
-@
-
-<<calcMod1LeeLikes, echo=FALSE, include=FALSE, message=FALSE>>=
-h1_epi_evals <- readRDS("../model1/run_level_3/lee1_epi_evals.rds")
-h1_end_evals <- readRDS("../model1/run_level_3/lee1_end_evals.rds")
-
-mod1_all_likes <- dplyr::inner_join(
-  x = h1_epi_evals,
-  y = h1_end_evals,
-  by = "parid"
-  ) %>%
-  dplyr::mutate(
-    joint_ll = pfLL.x + pfLL.y
-  )
-
-# Approximate the likelihood of the Lee et al (2020) version of Model 1 by
-# using the parameter sets that resulted in the maximum likelihood.
-mod1_lee_ll <- max(mod1_all_likes$joint_ll, na.rm = TRUE)
-@
-
-<<plotMod1Sims, echo=FALSE, fig.cap=paste0("Simulations using parameter sets that were generated by running source code provided by Lee et al (2020).
-Compare to Figure S7 of \\cite{lee20sup}.
-The upper bound for the likelihood of this model is ", round(mod1_lee_ll), "."), cache=TRUE, fig.width=6.5, fig.height=3.75>>=
-
-quants <- all_sims %>%
-  filter(!param_set %in% c(1, 2)) %>%  # Remove parameter sets inconsistent with Lee et al (2020).
-  group_by(week) %>%
-  summarize(
-    q025 = quantile(cases, probs = 0.025, na.rm = TRUE),
-    q50  = quantile(cases, probs = 0.500, na.rm = TRUE),
-    q975 = quantile(cases, probs = 0.975, na.rm = TRUE)
-  ) %>%
-  ungroup() %>%
-  mutate(date = lubridate::ymd("2010-10-16") + lubridate::weeks(week))
-
-quants_end <- end_sims %>%
-  group_by(week) %>%
-  summarize(
-    q025 = quantile(cases, probs = 0.025, na.rm = TRUE),
-    q50  = quantile(cases, probs = 0.500, na.rm = TRUE),
-    q975 = quantile(cases, probs = 0.975, na.rm = TRUE)
-  ) %>%
-  ungroup() %>%
-  mutate(date = lubridate::ymd("2010-10-16") + lubridate::weeks(week))
-
-ggplot() +
-  geom_point(data = true_agg_cases, aes(x = date, y = ReportedAll)) +
-  geom_line(data = quants, aes(x = date, y = q50), col = '#0b8a2b', size = 1) +
-  geom_ribbon(
-    data = quants,
-    aes(x = date, ymin = q025, ymax = q975),
-    alpha = 0.5, fill = '#0b9c30'
-  ) +
-  geom_line(data = quants_end, aes(x = date, y = q50), col = '#882871') +
-  geom_ribbon(data = quants_end, aes(x = date, ymin = q025, ymax = q975), alpha = 0.5, fill = '#882871') +
-  ylab("Reported, Symptomatic Incidence") +
-  scale_x_date(date_breaks = "1 year") +
-  theme(axis.title.x = element_blank(),
-        axis.text.x = element_text(hjust = 1, angle = 45))
-@
-
-Because the model forecasts provided by \cite{lee20} come from various sets of parameters---which each correspond to a unique log-likelihood value---it is not obvious how one would obtain an estimate for the log-likelihood of the model that was used for simulations by the Model~1 authors.
-One approach could be to calculate the logarithm of the weighted mean of the likelihoods for each parameter sets used to obtain the forecasts, where the weights are proportional to the number of times the parameter set was used.
-However, in an effort to not underestimate the likelihood of the model of the Model~1 authors, we report the estimated log-likelihood as the log-likelihood value corresponding to the parameter set with the largest likelihood value, even though the majority of simulations were obtained using parameter sets with lower likelihood values.
-In this sense, we consider the log-likelihood reported in Table~1 of the main text to be an upper-bound of the log-likelihood of the model used by \cite{lee20}.
-For each parameter set, the log-likelihood was estimated using a particle filter, implemented as the \texttt{pfilter} function in the \texttt{pomp} package.
-
-\subsection{Model~2 Replication}\label{sec:mod2rep}
-
-Model~2 was developed by a team that consisted of members from the Fred Hutchinson Cancer Research Center and the University of Florida (hereafter referred to as the Model~2 authors).
-While Model~2 is the only deterministic model we considered in our analysis, it contains perhaps the most complex descriptions of cholera in Haiti: Model~2 accounts for movement between spatial units; human-to-human and environment-to-human cholera infections; and transfer of water between spatial units based on elevation charts and river flows.
-
-The source code that the Model~2 authors used to generate their results was written in the \code{Python} programming language, and is publicly available at \url{10.5281/zenodo.3360857} and its accompanying GitHub repository \url{https://github.com/lulelita/HaitiCholeraMultiModelingProject}.
-In order to perform our analysis in a unified framework, we re-implemented this model in the \code{R} programming language using the \code{spatPomp} package \citep{asfaw23arxiv}, which facilitates the creation of meta-population models.
-We note that the travel and water matrices used to implement the complex dynamics in Model~2 \citep{lee20sup} are not available in either the Zenodo archive or the GitHub repository;
-instead, we obtained these matrices via personal correspondence with the Model~2 authors.
-Using these matrices, and the point estimates for model parameters provided by \citep{lee20sup}, we created trajectories of the cholera dynamics and compared this to available data.
-These trajectories, shown in Figure~\ref{fig:mod2rep}, are very similar to the trajectories shown in Figure~S15 of \cite{lee20sup}.
-
-<<Model2Replication, cache=TRUE, echo=FALSE>>=
-h2_epi <- haiti2()
-h2_end <- haiti2(region = 'after')
-
-h2_epi_traj <- trajectory(h2_epi, format = 'data.frame')
-h2_end_traj <- trajectory(h2_end, format = 'data.frame')
-
-h2_epi_traj$Ctotal <- rowSums(h2_epi_traj[, paste0("C", 1:10)]) * 0.2
-h2_end_traj$Ctotal <- rowSums(h2_end_traj[, paste0("C", 1:10)]) * 0.2
-
-dep_plot_df <- haitipkg::haitiCholera %>%
-  select(-report) %>%
-  tidyr::pivot_longer(
-    data = .,
-    cols = -date_saturday,
-    values_to = 'cases',
-    names_to = 'dep'
-  ) %>%
-  mutate(
-    date = as.Date(date_saturday),
-    dep = gsub("\\.", "_", dep)
-  ) %>%
-  mutate(
-    dep = case_when(dep == "Grand_Anse" ~ "Grande_Anse", TRUE ~ dep)
-  )
-
-true_agg_cases <- dep_plot_df %>%
-  tidyr::pivot_wider(
-    data = .,
-    id_cols = c(date),
-    names_from = dep,
-    values_from = cases,
-    names_prefix = 'cases_'
-  ) %>%
-  mutate(
-    ReportedAll = cases_Artibonite + cases_Centre +
-      cases_Grande_Anse + cases_Nippes + cases_Nord +
-      cases_Nord_Est + cases_Ouest + cases_Sud +
-      cases_Sud_Est + cases_Nord_Ouest
-  ) %>%
-  select(date, ReportedAll)
-
-h2_epi_traj$date <- as.Date(
-  lubridate::round_date(
-    lubridate::date_decimal(h2_epi_traj$year),
-    unit = 'day'
-  )
-)
-
-h2_end_traj$date <- as.Date(
-  lubridate::round_date(
-    lubridate::date_decimal(h2_end_traj$year),
-    unit = 'day'
-  )
-)
-
-all_traj <- bind_rows(
-  select(h2_epi_traj, date, Ctotal),
-  select(h2_end_traj, date, Ctotal)
-)
-
-projections <- left_join(
-  x = true_agg_cases,
-  y = all_traj,
-  by = 'date'
-)
-@
-
-\begin{figure}[ht]
-<<plotModel2Rep, cache=TRUE, echo=FALSE, fig.width=6.5, fig.height=4.5>>=
-gg_epi <- projections %>%
-  filter(date < as.Date("03-01-2014", format = '%m-%d-%Y')) %>%
-  ggplot(aes(x = date)) +
-  geom_line(aes(y = ReportedAll), col = 'black', linetype = 'dashed') +
-  geom_line(aes(y = Ctotal), col = 'red', size = 1) +
-  theme_bw() +
-  scale_y_continuous(limits = c(0, 30000), breaks = seq(0, 30000, 5000)) +
-  theme(axis.title.x = element_blank()) +
-  ylab("Reported number of cases")
-
-gg_end <- projections %>%
-  filter(date >= as.Date("03-01-2014", format = '%m-%d-%Y')) %>%
-  ggplot(aes(x = date)) +
-  geom_line(aes(y = ReportedAll), col = 'black', linetype = 'dashed') +
-  geom_line(aes(y = Ctotal), col = 'red', size = 1) +
-  theme_bw() +
-  scale_y_continuous(limits = c(0, 7000), breaks = seq(0, 7000, 1000)) +
-  theme(axis.title.x = element_blank()) +
-  ylab("Reported number of cases") +
-  scale_x_date(date_breaks = "1 year", date_labels = "%Y")
-
-cowplot::plot_grid(gg_epi, gg_end, ncol = 1)
-@
-\caption{\label{fig:mod2rep}
-Model 2 trajectories using the \code{haitipkg}. Compare to Figure S15 of \cite{lee20sup}.
-}
-\end{figure}
-
-There are minor differences between Figure~\ref{fig:mod2rep} and Figure~S15 of \cite{lee20sup}.
-While the discrepancy appears minor, the deterministic nature of Model~2 implies that an exact replication of model trajectories should be possible.
-In this case, these discrepancies may possibly be attributed to implementing the model and plotting the model trajectory in two different programming languages.
-Another potential explanation for the discrepancy is that the parameters that we used are only approximately the same as those used by \citet{lee20sup}.
-For example, the parameters $\transmission$, $\Wbeta{}$ (See Table~\ref{tab:translate}) had reported values of $9.9 \times 10^{-7}$ and $4.03 \times 10^{-2}$, respectively (Table~S13 of \cite{lee20sup}), but were actually fit to data and therefore likely these values have been rounded.
-Additionally, our implementation of Model~2 used a time scale of years and many of the parameters were reported on a weekly scale, so small differences may result due to unit conversions.
-The collective effect of these small differences in model parameters likely will result in small differences in model trajectories.
-
-Some additional concerns about being able to accurately replicate the results of \cite{lee20} are valid.
-Details about the measurement models and how latent states were initialized for the epidemic model were not provided by \citet{lee20sup} and therefore these details must be inferred by looking at the provided source code.
-According to repository comments, the files \code{fit\allowbreak In\allowbreak Pieces\allowbreak 3params\allowbreak Clean\allowbreak May2019\allowbreak Public.py} and \code{fit\allowbreak In\allowbreak Pieces\allowbreak Mu\allowbreak With\allowbreak Frac\allowbreak Sus\allowbreak Fixed\allowbreak All\allowbreak Infections\allowbreak Public.py} were used to fit the epidemic and endemic phases of the model respectively, although it is apparent that these exact files were not used to obtain the reported results since the files contain some variable-naming errors that make it impossible to run the files without making modifications \footnote{One example of why the code cannot be run that the file loads functions from a non-extant file named \code{choleraEqs.py} in line 13 rather than \code{cholera\allowbreak Eqs\allowbreak Public.py}.}.
-The inability to replicate the results by \citet{lee20} by running the provided source code makes checking whether or not a our numeric implementation faithfully represents their results very difficult.
-Additionally, the script that was said to been used to obtain the results reported by \cite{lee20} appears to use a different measurement model than what was described in the supplemental material, again making it difficult to fully replicate the result of \cite{lee20} without being able to easily run the provided source code.
-In this case, we chose to use measurement model described by Eq.~\ref{model2:measEpi} for both phases of the epidemic, as this seemed to visually match the results of \cite{lee20} most closely.
-
-\subsection{Model~3 Replication}
-
-Model~3 was developed by a team of researchers at the Laboratory of the Swiss Federal Institute of Technology in Lausanne, hereafter referred to as the Model~3 authors.
-The code that was originally used to implement Model~3 is archived with the DOI: \url{10.5281/zenodo.3360723}, and also available in the public GitHub repository: \code{jcblemai/haiti-mass-ocv-campaign}.
-Because the code was made publicly available, and final model parameters were reported in the supplementary material of \cite{lee20}, we were able to reproduce Model~3 by directly using the source code.
-In Fig.~\ref{fig:mod3rep}, we plot simulations from this model.
-This figure can be compared to Figure S18 of \cite{lee20}.
-We note that slight differences may be accounted for due to variance in the model simulations and the difference in programming language used to produce the figure.
-Overall, the high standard of reproducibility that was achieved by the Model~3 authors facilitated the ability to readily replicate their model and results.
-
-<<Model3Replication, echo=FALSE, cache=TRUE>>=
-NSIM <- 500
-
-plot_order <- c(
-  'Artibonite',
-  'Sud_Est',
-  'Nippes',
-  'Nord_Est',
-  'Ouest',
-  'Centre',
-  'Nord',
-  'Sud',
-  'Nord_Ouest',
-  'Grande_Anse'
-)
-
-dep_labeller <- as_labeller(
-  c(
-    'Artibonite' = 'Artibonite',
-    'Sud_Est' = 'Sud-Est',
-    'Sud.Est' = 'Sud-Est',
-    'Nippes' = 'Nippes',
-    'Nord_Est' = 'Nord-Est',
-    'Nord.Est' = 'Nord-Est',
-    'Ouest' = 'Ouest',
-    'Centre' = 'Centre',
-    'Nord' = 'Nord',
-    'Sud' = 'Sud',
-    'Nord_Ouest' = 'Nord-Ouest',
-    'Nord.Ouest' = 'Nord-Ouest',
-    'Grande_Anse' = 'Grand\'Anse',
-    'Grand.Anse' = 'Grand\'Anse'
-  )
-)
-
-lee3 <- lee3_spatPomp()
-lee3_sims_quants <- bake("../model3/si/lee3Sims.rds", {
-
-  # Simulate from lee et al 2020 model
-  sims_temp <- simulate(
-    lee3, nsim = NSIM,
-    seed = 321, format = 'data.frame'
-  )
-
-  # Get desired quantiles of reported cases from simulations
-  quants <- sims_temp %>%
-    dplyr::group_by(unitname, time) %>%
-    dplyr::summarise(
-      q05 = quantile(cases, 0.025, na.rm = T),
-      mean = mean(cases, na.rm = T),
-      q50 = quantile(cases, 0.5, na.rm = T),
-      q95 = quantile(cases, 0.975, na.rm = T)
-    ) %>%
-    dplyr::ungroup() %>%
-    dplyr::mutate(
-      date = lubridate::date_decimal(time),
-      date = as.Date(lubridate::round_date(date))
-    ) %>%
-    dplyr::rename(dep = unitname)
-
-  # Remove cases from Ouest that were not considered by Lee et al. (2020)
-  quants %>%
-    dplyr::filter(
-      !(dep == "Ouest" & date < '2017-06-10')
-    )
-
-})
-
-# Get the subset of cholera cases that match simulations from Lee et al. (2020) model.
-sub_cases <- haitiCholera %>%
-  dplyr::mutate(date = as.Date(date_saturday)) %>%
-  dplyr::select(-report, -date_saturday) %>%
-  dplyr::filter(lubridate::decimal_date(date) >= min(lee3@times)) %>%
-  tidyr::pivot_longer(
-    data = .,
-    cols = -date,
-    names_to = "dep",
-    values_to = "cases"
-  ) %>%
-  dplyr::mutate(
-    dep = dplyr::case_when(
-      dep == "Grand.Anse" ~ "Grande_Anse",
-      dep == "Nord.Est" ~ "Nord_Est",
-      dep == "Nord.Ouest" ~ "Nord_Ouest",
-      dep == "Sud.Est" ~ "Sud_Est",
-      TRUE ~ dep
-    )
-  )
-@
-
-\begin{figure}[ht]
-<<PlotMod3Rep, echo=FALSE, message=FALSE, fig.height=3.75>>=
-ggplot() +
-  # Region for upper and lower simulation bounds
-  geom_ribbon(
-    data = lee3_sims_quants,
-    aes(x = date, ymin = q05, ymax = q95),
-    fill = 'darkblue', alpha = 0.5
-  ) +
-  # Median line
-  geom_line(
-    data = lee3_sims_quants,
-    aes(x = date, y = q50),
-    col = 'darkblue'
-  ) +
-  # Observed data
-  geom_point(
-    data = dplyr::filter(sub_cases, cases < 500),
-    aes(x = date, y = cases), size = 0.2
-  ) +
-  facet_wrap(
-    ~factor(dep, levels = plot_order),
-    nrow = 2, labeller = dep_labeller
-  ) +
-  labs(y = 'Reported Cholera Cases') +
-  scale_x_date(date_breaks = "6 months", date_labels = "%Y-%m") +
-  theme(
-    axis.text.x = element_text(angle = 40, hjust = 1, vjust = 1, size = 7),
-    axis.title.x = element_blank()
-  )
-@
-\caption{\label{fig:mod3rep}
-Simulations from Model~3.
-Compare to Figure S18 of \cite{lee20}.
-}
-\end{figure}
-
-<<lee3Cleanup, echo=FALSE, include=FALSE, message=FALSE>>=
-rm(lee3, lee3_sims_quants, sub_cases)
-gc()
-@
-
-\section{Forecasting with parameter uncertainty}
-
-Let $f_{Y_{1:N}}(y_{1:N} | \theta)$ denote the pdf of the model under consideration, were $\theta$ is a parameter vector that indexes the model.
-Furthermore, denote the observed data as $y_{1:N}^*$.
-Because the uncertainty in just a single parameter can lead to drastically different forecasts \citep{saltelli20},
-parameter uncertainty should be considered when obtaining model forecasts when the goal is to influencing policy.
-In a Bayesian modelling paradigm, the most natural way to account for parameter uncertainty in model forecasts is to suppose that $\theta$ comes from a distribution $f_{\Theta}$, and then to obtain $J$ forecasts from the model where each forecast is obtained using parameters drawn from the posterior distribution $\Theta_{1:J} \mid Y_{1:N} = y_{1:N}^* \sim f_{\Theta}\big(\theta | Y_{1:N} = y_{1:N}^*\big)$.
-
-When frequentist methods are used, however, there does not exist a posterior distribution from which one could sample.
-A common approach could be to obtain a weighted average of the simulations from various models \citep{hoeting99}, but this can be problematic when forecasts from each model are very different from each other \citep{grueber11}.
-A similar approach that has been taken \citep{king15} is to obtain model forecasts using multiple sets of parameter values and then sample from the resulting forecasts using weights proportional to the corresponding likelihoods of the parameter values.
-This approach could be considered as empirical Bayes, as it is equivalent to using a discrete uniform prior where the set of values in the prior distribution is determined by a stochastic routine applied to the observed data, as discussed below.
-
-For each $i \in 1:k$, let $\Theta_i$ be a random vector of model parameters.
-Then, letting $\Theta$ denote the true model parameters, we endow the set $\{\Theta_1, \Theta_2, \ldots, \Theta_k\}$ with a discrete uniform distribution, such that $P\big(\Theta = \Theta_i\big) = \frac{1}{K}$ for all values $i \in \seq{1}{k}$.
-Using this as a prior distribution, the posterior distribution of $\Theta | Y_{1:N} = y_{1:N}^*$ can be expressed as:
-$P\big(\Theta = \Theta_k | Y_{1:N} = y_{1:N}^*\big) = \frac{f_{Y_{1:N}}(y_{1:N}^*| \Theta_k)}{\sum_{l = 1}^K f_{Y_{1:N}}(y_{1:N}^*| \theta_l)}$.
-In a standard empirical Bayes analysis, the values $\Theta_1, \ldots, \Theta_k$ of the prior distribution would be chosen using the observed data, resulting in a posterior distribution that weighs the prior parameter vectors proportional to their corresponding likelihoods.
-Instead, we choose $\Theta_i$ to be the output of a stochastic routine applied to the observed data by setting $\Theta_i$ to be the output of an iterated filtering algorithm.
-In practice, because the likelihood maximization routines of iterated filtering methods are stochastic, it is common to run the iterated filtering method multiple times for each model in order to obtain a maximum likelihood estimate for model parameters.
-This results in a natural set of parameters near the MLE that could be used as the discrete prior distribution.
-
-\section{Translating to \citet{lee20} notation}
-
-Since the models of \citet{lee20} were developed independently, the choice of notation varies inconsistently between models.
-For our reanalysis, we rename parameters to provide a unified notation facilitating comparison between models.
-Table~\ref{tab:translate} maps this notation back to the original notations, for reference.
-
-\begin{table}
-  \begin{center}
-  \begin{tabular}{|c|c|c|c|c|}\hline
-    \multirow{2}{*}{Parameter} & Our & \multicolumn{3}{c|}{Lee et al. (2020a)} \\\cline{3-5}
-     & Notation & 1 & 2 & 3 \\
-    \hline
-    \hline
-    % Table begins here
-    Reporting Rate & $\reportRate$ & $\rho$ & $\rho$ & $\epsilon_1, \epsilon_2$ \\\hline
-    Mixing Coefficient & $\mixExponent$ & $\nu$ & --- & --- \\\hline
-    Measurement Over-Dispersion & $\obsOverdispersion$ & $\tau$ & --- & $p$ \\\hline
-    Birth Rate & $\muBirth$ & $\mu$ & --- & --- \\\hline
-    Natural Mortality Rate & $\muDeath$ & $\delta$ & --- & $\mu$ \\\hline
-    Cholera Mortality Rate & $\choleraDeath$ & --- & --- & $\alpha$ \\\hline
-    Latent Period & $1/\muEI$ & $1/\sigma$ & $1/\gamma_E$ & --- \\\hline
-    Recovery Rate & $\muIR$ & $\gamma$ & $\gamma$ & $\gamma$ \\\hline
-    Loss of Immunity & $\muRS$ & $\alpha$ & $\sigma$ & $\rho$ \\\hline
-    Symptomatic Ratio & $\symptomFrac$ & $1 - \theta_0$ & $k$ & $\sigma$ \\\hline
-    Asymptomatic Relative Infectiousness & $\asymptomRelativeInfect$ & $\kappa$ & $red_\beta$ & --- \\\hline
-    Human-to-Water Shedding & $\Wshed$ & --- & $\mu$ & $\theta_I$ \\\hline
-    Asymptomatic Relative Shedding & $\asymptomRelativeShed$ & --- & $red_\mu$ & $\theta_A/\theta_I$ \\\hline
-    Seasonal Amplitude & $\seasAmplitude$ & --- & $\alpha_s$ & $\lambda$ \\\hline
-    Transmission & $\transmission$ & $\beta$ & $\beta$ & $c$ \\\hline
-    Water-to-Human & $\Wbeta{}$ & --- & $\beta_W$ & $\beta$ \\\hline
-    Bacteria Mortality & $\Wremoval$ & --- & $\delta$ & $\mu_\beta$ \\\hline
-    Vaccination Efficacy & $\vaccineEfficacy$ & $\theta_{vk}$ & $\theta_1, \theta_2, \theta_{1_5}, \theta_{2_5}$ & $\eta_{1d}, \eta_{2d}$ \\\hline
-    Process Over-dispersion & $\sigmaProc$ & --- & --- & $\sigma_w$\\\hline
-    % Half Saturation Constant & $\Wsat$ & --- & $Sat$ & $K$ \\\hline
-  \end{tabular}
-  \end{center}
-  \caption{
-  \label{tab:translate}Translations between our common notation and notation used by \cite{lee20}
-  }
-\end{table}
-
-\clearpage
-
-\bibliography{../bib-haiti}
-
-\end{document}

### 2023-09-01 si/inputs/modelDetails.tex
Started work to make individual si sections seperate

--- 2023-08-31 si/si.Rnw
@@
-\documentclass[11pt]{article}
-
-\pdfminorversion=4
-
-\newcommand\secTitleSpace{\hspace{3mm}}
-
-<<packages, include=FALSE>>=
-library(tidyverse)
-library(pomp)
-library(haitipkg)
-library(spatPomp)
-library(doParallel)
-library(doRNG)
-
-cores <-  as.numeric(Sys.getenv('SLURM_NTASKS_PER_NODE', unset = NA))
-if(is.na(cores)) cores <- detectCores()
-# cores <- 20
-registerDoParallel(cores)
-theme_set(theme_bw())
-
-if (!dir.exists("../model3/si")) {
-  dir.create("../model3/si", recursive = TRUE)
-}
-@
-
-<<set-opts, include=FALSE>>=
-options(
-  scipen = 2,
-  help_type = "html",
-  stringsAsFactors = FALSE,
-  continue = "+  ",
-  width = 70,
-  useFancyQuotes = FALSE,
-  reindent.spaces = 2,
-  xtable.comment = FALSE
-)
-@
-
-<<knitr-opts, include=FALSE, purl=FALSE>>=
-library(knitr)
-opts_knit$set(concordance=TRUE)
-opts_chunk$set(
-  progress=TRUE,prompt=TRUE,highlight=FALSE,
-  tidy=TRUE,
-  tidy.opts=list(
-    keep.blank.line=FALSE
-  ),
-  comment="",
-  warning=FALSE,
-  message=FALSE,
-  error=TRUE,
-  echo=FALSE,
-  cache=FALSE,
-  strip.white=TRUE,
-  results="markup",
-  fig.path="figure/",
-  fig.lp="fig:",
-  fig.align="center",
-  fig.show="asis",
-  dev="pdf",
-  dev.args=list(
-    bg="transparent",
-    pointsize=9
-  )
-)
-
-myround <- function (x, digits = 1) {
-  # taken from the broman package
-  if (digits < 1)
-    stop("This is intended for the case digits >= 1.")
-  if (length(digits) > 1) {
-    digits <- digits[1]
-    warning("Using only digits[1]")
-  }
-  tmp <- sprintf(paste("%.", digits, "f", sep = ""), x)
-  zero <- paste0("0.", paste(rep("0", digits), collapse = ""))
-  tmp[tmp == paste0("-", zero)] <- zero
-  tmp
-}
-
-@
-
-%% copied from pnas-new.cls
-\usepackage{amsmath,amsfonts,amssymb,graphicx}
-
-%%%% parameters %%%%%%%%%%5
-\newcommand\Wsat{W_{\mathrm{sat}}}
-\newcommand\muIR{\mu_{IR}}
-\newcommand\muEI{\mu_{EI}}
-\newcommand\transmission{\beta}
-\newcommand\seasAmplitude{a}
-\newcommand\rainfallExponent{r}
-\newcommand\muRS{\mu_{RS}}
-\newcommand\vaccineEfficacy{\theta}
-\newcommand\muBirth{\mu_S}
-\newcommand\muDeath{\delta}
-\newcommand\choleraDeath{\delta_{C}}
-\newcommand\symptomFrac{f}
-\newcommand\asymptomRelativeInfect{\epsilon}
-\newcommand\asymptomRelativeShed{\epsilon_{W}}
-\newcommand\Wbeta[1]{\beta_{W#1}}
-\newcommand\Wremoval{\delta_W}
-\newcommand\Wshed{\mu_W}
-\newcommand\mixExponent{\nu}
-\newcommand\sigmaProc{\sigma_{\mathrm{proc}}}
-\newcommand\reportRate{\rho}
-\newcommand\obsOverdispersion{\psi}
-\newcommand\phaseParm{\phi}
-\newcommand\transmissionTrend{\zeta}
-\newcommand\Binit{\xi}
-\newcommand\vaccClass{Z}
-\newcommand\vaccCounter{z}
-\newcommand\modelCounter{m}
-\newcommand\Whur[1]{\beta_{W#1}^{hm}}
-\newcommand\hHur[1]{h_{#1}^{hm}}
-\newcommand\Iinit{I_{0,0}}
-\newcommand\Einit{E_{0, 0}}
-\newcommand\reportChange{\delta}
-
-\newcommand\missing{\hspace{6mm} ---}
-\newcommand\fixed{\color{blue}}
-\newcommand\demography{\bullet}
-
-\newcommand\mytitle{Informing policy via dynamic models: Cholera in Haiti}
-\newcommand\comma{{\hspace{-0.25mm},\hspace{-0.2mm}}}
-
-%%%%%%%%%%%%%%%% START OF USER DEFINITIONS %%%%%%%%%%%%%%%5
-
-\newcommand\siOnly[1]{} %% redefined for the supplement
-\newcommand\msOnly[1]{#1} %% redefined for the supplement
-
-\newcommand\code[1]{\texttt{#1}}
-\usepackage{xurl}
-% \usepackage[breaklinks]{hyperref}
-\usepackage{makecell}
-\usepackage{multirow}
-\usepackage{multicol}
-\usepackage[ruled,noline,linesnumbered]{algorithm2e}
-% \usepackage{setspace}
-
-%% EDITING MACROS
-\usepackage{color}
-\usepackage[normalem]{ulem}% to use \sout in feedback commands
-\usepackage{bm}
-% orange for EI
-\definecolor{orange}{rgb}{1,0.5,0}
-\newcommand\ei[2]{\sout{#1} \textcolor{orange}{#2}}
-\newcommand\eic[1]{\textcolor{orange}{[#1]}}
-% green for JW
-\definecolor{green}{rgb}{0,0.5,0}
-\newcommand\jw[2]{\sout{#1} \textcolor{green}{#2}}
-\newcommand\jwc[1]{\textcolor{green}{[#1]}}
-% purple for JJ
-\definecolor{purple}{rgb}{0.5,0,1}
-\newcommand\jj[2]{\sout{#1} \textcolor{purple}{#2}}
-\newcommand\jjc[1]{\textcolor{purple}{[#1]}}
-% cyan for AR
-\definecolor{cyan}{rgb}{0,.5,.5}
-\newcommand\ar[2]{\sout{#1} \textcolor{cyan}{#2}}
-\newcommand\arc[1]{\textcolor{cyan}{#1}}
-% light brown for KT
-\definecolor{lightbrown}{rgb}{0.5,0.5,0}
-\newcommand\kt[2]{\sout{#1} \textcolor{lightbrown}{#2}}
-\newcommand\ktc[1]{\textcolor{lightbrown}{#1}}
-
-%% basic POMP definitions
-\newcommand\Xspace{{\mathbb X}}
-\newcommand\Yspace{{\mathbb Y}}
-\newcommand\Thetaspace{\R^{\Thetadim}}
-\newcommand\Xunitspace{{\scriptsize{\mathcal X}}}
-\newcommand\Yunitspace{{\tiny{\mathcal Y}}}
-\newcommand\hatTheta{\widehat{\Theta}}
-\newcommand\Xdim{{\mathrm{dim}}(\Xspace)}
-\newcommand\Ydim{{\mathrm{dim}}(\Yspace)}
-\newcommand\Thetadim{P}
-\newcommand\thetadim{p}
-\newcommand\timeSet{{\mathbb T}}
-\newcommand\data[1]{#1^*}
-
-\newcommand\unitSpecific{\hspace{0.1mm}\mathrm{us}}
-\newcommand\shared{\hspace{0.15mm}\mathrm{sh}}
-
-% for algorithm pseudocode
-\SetKwFor{For}{for}{:}{end}
-
-%% for particle filters
-\newcommand\unit{u}
-\newcommand\altUnit{\tilde u}
-\newcommand\Unit{U}
-\newcommand\rep{i}
-\newcommand\Rep{\mathcal{I}}
-\newcommand\island{\rep}
-\newcommand\Island{\Rep}
-%\newcommand\txtisland{replicate}
-\newcommand\altIsland{\tilde i}
-\usepackage[mathscr]{euscript}
-
-\renewcommand\time{n}
-\newcommand\myvec[1]{\boldsymbol{#1}}
-\newcommand\altTime{\tilde n}
-\newcommand\Time{N}
-\newcommand\Np{J}
-\newcommand\np{j}
-\newcommand\altNp{k}
-\newcommand\altAltNp{\tilde j}
-\newcommand\resampleIndex{r}
-\newcommand\unitWeight{w}
-
-%% for all iterated filtering algorithms
-\newcommand\Nit{M}
-\newcommand\nit{m}
-
-%% customized math macros
-\newcommand\prob{\mathbb{P}}
-\newcommand\E{\mathbb{E}}
-\newcommand\dd[1]{\mathrm{d}{#1}}
-\newcommand\given{{\,\vert\,}}
-\newcommand\equals{{{\,}={\,}}}
-\newcommand\myequals{\hspace{0.5mm}{=}\hspace{0.5mm}}
-\newcommand\myto{{\;:\;}}
-\newcommand\seq[2]{{#1}\!:\!{#2}}
-\newcommand\mydot{{\,\cdot\,}}
-\newcommand\cp[2]{N_{\mathrm{#1}\mathrm{#2}}}
-\newcommand\giventh{{\hspace{0.5mm};\hspace{0.5mm}}}
-\newcommand\normal{\mathcal{N}}
-\newcommand\argequals{{\,=\,}}
-\newcommand\lags{c}
-\newcommand\maxlag{\overline{c}}
-\newcommand\nlfList{C}
-\newcommand\bigO{\mathcal{O}}
-\newcommand\loglik{\lambda}
-\newcommand\loglikMC{\MC{\loglik}}
-\newcommand\R{\mathbb{R}}
-\newcommand\param{\,;}
-\newcommand\mycolon{{\hspace{0.6mm}:\hspace{0.6mm}}}
-\newcommand\MC[1]{#1^{\,\mbox{\tiny MC}}}
-\newcommand\EMC{{\E}}
-\newcommand\Var{\mathrm{Var}}
-\newcommand\var{\Var}
-\newcommand\Cov{\mathrm{Cov}}
-\newcommand\cov{\Cov}
-\newcommand\iid{\mathrm{iid}}
-\newcommand\dist{d}
-\newcommand\transpose{\top}
-
-\newcommand\gnsep{,}
-
-\def\lik{L}
-\def\loglik{\ell}
-
-%% THEOREM-LIKE ENVIRONMENTS
-\usepackage{amsthm}
-\newtheorem{prop}{Proposition}
-\newtheorem{theorem}{Theorem}
-\newtheorem{lemma}{Lemma}
-\newtheorem{example}{Example}
-\newtheorem{remark}{Remark}
-\newtheorem{cor}{Corrolary}
-\newtheorem{fact}{Fact}
-\newtheorem{assumption}{Assumption}
-\newtheorem{assumptionB}{Assumption}
-\renewcommand\theassumption{A\arabic{assumption}}
-\renewcommand\theassumptionB{B\arabic{assumptionB}}
-
-%% FOR PSEUDOCODE SETUP
-\newcommand\mystretch{\rule[-2mm]{0mm}{5mm} }
-\newcommand\asp{\hspace{4mm}}
-\newcommand\codeIndent{\hspace{4mm}}
-
-%%%%%%%%%%%%%%%% END OF USER DEFINITIONSS %%%%%%%%%%%%%%%
-
-\bibliographystyle{apalike}
-\usepackage{natbib}
-\usepackage{fullpage}
-
-\newcommand\myeqref[1]{(\ref{#1})}
-
-\renewcommand{\contentsname}{Supplementary Content}
-\renewcommand{\refname}{Supplementary References}
-\renewcommand\thefigure{S-\arabic{figure}}
-\renewcommand\thetable{S-\arabic{table}}
-\renewcommand\thesection{S\arabic{section}}
-\renewcommand\theequation{S\arabic{equation}}
-\renewcommand\theprop{S\arabic{prop}}
-\renewcommand\thelemma{S\arabic{lemma}}
-\renewcommand\thealgocf{S\arabic{algocf}}
-
-\setcounter{tocdepth}{1}
-
-\begin{document}
-
-\date{\today}
-\title{Supplement to ``{\mytitle}''}
-  \author{Jesse Wheeler, AnnaElaine L. Rosengart, Zhuoxun Jiang, \\ Kevin Tan, Noah Treutle and Edward L. Ionides
- \\ \hspace{.2cm}\\
-    Department of Statistics, University of Michigan\\
-}
-
-\newcommand{\blind}{1}
-
-\if1\blind
-{
-\maketitle
-}
-\fi
-
-\if0\blind
-{
-  \bigskip
-  \bigskip
-  \bigskip
-  \begin{center}
-    {\LARGE\bf \mytitle}
-\end{center}
-  \bigskip
-  \bigskip
-}\fi
-
-\tableofcontents
-
-\newpage
-
 \section{Markov chain and differential equation interpretations of compartment flow rates}
@@
 In this case, we formally model the death rate as a rate of returning to the susceptible class $S$, and use external transitions from $\demography$ into $S$ to describe only net population increase.
-
-\section{Measurement Models}
-
-Each POMP requires specification of a measurement model, which is a statistical description of how observations on the system are obtained.
-In general, we used the same measurement models that were reported by \cite{lee20} unless specifically noted in the following subsections.
-
-\subsection{Model~1}
-
-In this model, the advantage afforded by vaccination is an increased probability that an infection is asymptomatic.
-Therefore, under the assumptions of this model, all reported cases are assumed to be a fraction of individuals that transition from the exposed to the infected compartment, as noted in Eq.~\myeqref{model1:meas}:
-\begin{equation}
-  \label{model1:meas}
-  Y_{n} \mid \Delta N_{E_{\cdot}I_{\cdot}}(n) = \reportChange \sim \text{NB}\left(\reportRate \reportChange, \obsOverdispersion \right),
-\end{equation}
-where $Y_n$ is the reported cholera cases at time $t_n \in t_1:t_N$ and $\Delta N_{E_{\cdot}I_{\cdot}}(n)$ is the sum total of individuals across each vaccination compartment $\vaccCounter \in 1:\vaccClass$ who moved from compartment $E_\vaccCounter$ to $I_\vaccCounter$ since observation $t_{n-1}$.
-Here, $\text{NB}\left(\reportRate \reportChange, \obsOverdispersion \right)$ denotes a negative binomial distribution with mean $\reportRate \reportChange$ and variance $\reportRate \reportChange \big(1 + \frac{\reportRate \reportChange}{\obsOverdispersion}\big)$.
-
-\subsection{Model~2}\label{sec:mod2Meas}
-
-Model~2 was fit using reported case counts that were transformed using the natural logarithm.
-We fit Model~2 using the subplex algorithm in the \code{subplex} package, using
-a Gaussian measurement model (Eq.~\myeqref{model2:meas}) on the log transformed cases within each unit.
-The final loss function that is maximized is the product of the likelihoods of the individual units, or the sum of the log-likelihood of the individual units.
-The measurement model for individual units is given in Eq.~\myeqref{model2:meas}.
-\begin{equation}
-  \label{model2:meas}
-  \log\big(Y_{u, n} + 1\big) \mid \Delta N_{E_{u\cdot}I_{u\cdot}}(n) = \reportChange_u \sim \text{N}\left(\log \big(\reportRate \reportChange_u + 1\big), \obsOverdispersion^2 \right),
-\end{equation}
-where $\Delta N_{E_{u\cdot}I_{u\cdot}}(n)$ is the sum total of individuals across vaccination compartment $\vaccCounter \in 0:4$ within unit $u$ who moved from compartment $E_{u\vaccCounter}$ to $I_{u\vaccCounter}$ since observation $t_{n-1}$.
-Therefore, because the natural logarithm of observed case counts (plus one, to avoid taking the logarithm of zero) has a normal distribution, $Y_{u, n} + 1$ is assumed to follow a log-normal distribution with log-mean parameter $\log\big(\reportRate \Delta N_{E_{u\cdot}I_{u\cdot}}(n) + 1\big)$ and log-variance $\obsOverdispersion^2$.
-We note that fitting a model with this measurement model is equivalent to fitting using least squares, with $\log(Y_{u, n} + 1)$ as the response variable.
-
-This measurement model differs from that used by \citet{lee20}, who fit the model in two stages: epidemic and endemic phases.
-Although their text and supplement material do not explicitly describe the measurement model used, inspection of the code provided with their submission suggests a change in measurement model between the epidemic and endemic phases.
-% In the file \code{choleraEqsPublic.py}, \citet{lee20} create several functions, where each function represents a set of coupled differential equations that could potentially be used to model cholera incidence data.
-% Each function returns a vector (or \code{numpy} array) that represents the change in each state variable for a single time step, including the variable \code{dC}, which tracks the number of new infections and is used to obtain the reported case counts.
-% Following their comments and code, it appears that the function \code{choleraEqs10WithoutVaccinationNetwork} was used to describe the dynamics of the epidemic phase, and \code{choleraEqs11WithoutVaccinationNetwork} was used for the endemic stage (see Sec~\ref{sec:mod2rep} for more details).
-% Because their models were fit using least squares, the code in these functions suggest that the measurement model for the epidemic phase is
-The measurement model they used for the epidemic phase is
-\begin{equation}
-  \label{model2:measEpi}
-  Y_{u, n} \mid \Delta N_{E_{u\cdot}I_{u\cdot}}(n) = \reportChange_u \sim \text{N}\left(\reportRate \reportChange_u, \obsOverdispersion^2 \right),
-\end{equation}
-% which is similar to our measurement model, the primary difference being that the measurement model is applied to raw case counts rather than log-transformed case counts.
-The measurement model they used for the endemic phase modifies the epidemic model by counting both asymptomatically infected (A) and symptomatically infected (I) individuals in the case counts:
-\begin{equation}
-  \label{model2:measEnd}
-  Y_{u, n} \mid \Delta N_{E_{u\cdot}I_{u\cdot}}(n) = \reportChange_{u1}, \Delta N_{E_{u\cdot}A_{u\cdot}}(n) = \reportChange_{u2} \sim \text{N}\left(\reportRate \big(\reportChange_{u1} + \reportChange_{u2}\big), \obsOverdispersion^2 \right),
-\end{equation}
-where the notation for $\Delta N_{E_{u\cdot}A_{u\cdot}}(n)$ is similar to $\Delta N_{E_{u\cdot}I_{u\cdot}}(n)$, described above.
-
-\subsection{Model~3}
-
-In this model, reported cholera cases are assumed to stem from individuals who develop symptoms and seek healthcare.
-Therefore reported cases are assumed to come from an over-dispersed negative binomial model, given the increase in infected individuals:
-\begin{equation}
-  \label{model3:meas}
-  Y_{u, n} \mid \Delta N_{S_{u \cdot}I_{u}}(t) = \reportChange_u \sim \text{NB}\left(\reportRate \reportChange_{u}, \obsOverdispersion \right),
-\end{equation}
-where $\Delta N_{S_{u \cdot}I_{u \vaccCounter}}(n)$ is the number of individuals who moved from compartment $S_{u \vaccCounter}$ to $I_{u}$ since observation $t_{n-1}$.
-
-This measurement model is a minor change from that used by \cite{lee20}, which allowed for a change in the reporting rate on January 1st, 2018.
-The fitted values of the reporting rate---before and after January 2018---were $0.97$ and $0.097$, respectively.
-An instantaneous change from near perfect to almost non-existent reporting can be problematic, as it forces the model to explain the observed reduction in reported cases as a decrease in the reporting of cases, rather than a decrease in the prevalence of cholera.
-This shift was justified by a ``change of the case definition that occurred on January 1st, 2018";
-this claim was not cited, and we could find no evidence that such a drastic change in the reporting rate would be warranted.
-We therefore do not allow a change in reporting rate when fitting Model~3.
-
-\section{Initial Values}
-
-To perform inference on POMP models, it is necessary to propose an initial probability density for the latent process $f_{X_0}(x_0;\theta)$, including the possibility that the initial values of the latent states are known, or are a non-random function of the unknown parameter vector, $\theta$.
-This density is used to obtain initial values of the latent state when fitting and evaluating the model.
-For each of the models considered in this analysis, the initial conditions are derived by enforcing the model dynamics on reported cholera cases.
-It is also sometimes necessary to fit some initial value parameters in order to help determine initial values for weakly identifiable compartments.
-In the following subsections, we mention initial value parameters that were fit for each model.
-
-\subsection{Model~1}
-
-For this model, the number of individuals in the Recovered and Asymptomatic compartments are set to zero, but the initial proportion of Infected and Exposed individuals is estimated as initial value parameters ($\Iinit$ and $\Einit$, respectively) using the IF2 algorithm, implemented as \texttt{mif2} in the \texttt{pomp} package.
-Finally, the initial proportion of Susceptible individuals $S_{0, 0}$ is calculated as $S_{0, 0} = 1 - \Iinit - \Einit$.
-This model for the initial values of the latent states matches that which was used by \citet{lee20}.
-
-\subsection{Model~2}
-
-Model~2 assumes that the initial values are a deterministic function of the reporting rate and the initial case reports, and so no initial value parameters need to be estimated.
-Initial values for latent state compartments are chosen so as to enforce the model dynamics on the observed number of cases.
-Specifically, the model sets $I_{u0}(0) = y^*_{1u} / \reportRate$ for each unit $u \in 1:10$, where $I_{u0}(t)$ is the number of infected individuals in unit $u$ at time $t$, vaccination scenario $z = 0$, $y^*_{tu}$ is the reported number of cases, and $\reportRate$ is the reporting rate.
-This model for the initial values of the latent states matches that which was used by \citet{lee20}.
-
-\subsection{Model~3}
-
-The latent states of this model are initialized by enforcing the model dynamics on the incidence data from the start of the recorded cases until time $t_0$, requiring that some of the available data be used to determine the initial values of the latent states.
-This is the same approach that was taken by \citet{lee20}, who used the value $t_0 = \text{2014-02-22}$; this choice of $t_0$ results in modeling roughly only $60\%$ of the available data, some of which is later discarded for alternative reasons \citep{lee20sup}.
-
-We do not see any immediate reason that this model could not be extended to cover a larger range of the data, and chose the value $t_0 = \text{2010-11-13}$.
-This choice of $t_0$ corresponds to using approximately $1\%$ of the available data to determine initial values of the latent states.
-In addition to modeling a larger portion of the available data, this choice of $t_0$ corresponds to an important real-world event, as daily reports from each of the departments were not being sent to the health ministry until November 10, 2010; % TODO: Cite
-this choice of $t_0$ therefore makes $\bm{Y}\big(t_1\big)$ the first week of data once daily reports are being sent to the health ministry.
-The few observation times that exist before $t_0$ are used to initialize the model by enforcing model dynamics on these preliminary observations.
-For convenience, we denote these observations as $t_{-3}, t_{-2}$ and $t_{-1}$; as before, we let $y_{u,-k} = Y_{u}\big(t_{-k}\big)$ denote the observed case count for unit $u$ at time point $t_{-k}$, where $k \in 1:3$.
-Equations for the initial values of non-zero latent states are provided in Eqs.~\myeqref{eq:model3:I0}--\myeqref{eq:model3:B0}.
-
-\begin{eqnarray}
-\label{eq:model3:I0}
-I_{u}\big(t_0\big) &=& \frac{y^*_{u, -1}}{7\reportRate\left(\muIR + \left(\muDeath + \choleraDeath\right)/365\right)},
-\\%[-3pt]
-\label{eq:model3:A0}
-A_{u}\big(t_0\big) &=& \frac{I_{u0}\big(t_0\big)\left(1 - \symptomFrac\right)}{\symptomFrac},
-\\
-\label{eq:model3:R0}
-R_{u01}\big(t_0\big) &=& R_{u02}\big(t_0\big) = R_{u03}\big(t_0\big) = \left(\frac{\sum_{k = -3}^0 y^*_{u, k}}{\reportRate \symptomFrac} - \left(I_{u0}\big(t_0\big) + A_{u0}\big(t_0\big)\right)\right) / 3
-\\
-\label{eq:model3:S0}
-S_{u0}\big(t_0\big) &=& \mathrm{Pop}_u - I_{u}\big(t_0\big) - A_{u}\big(t_0\big) - \sum_{k=1}^3 R_{u0k}\big(t_0\big)
-\\
-\label{eq:model3:B0}
-B_{u}\big(t_0\big) &=& \big[1 + \seasAmplitude \tilde{J}^r \big] \mathrm{Den}_u \, \Wshed \big[ I_{u}(t_0)+ \asymptomRelativeShed A_{u_0}(t) \big]/\Wshed.
-\end{eqnarray}
-
-In Eq.~\myeqref{eq:model3:B0}, $\tilde{J} = 0.002376$ is the median adjusted rainfall over the observation period.
-One important consideration to make with this parameter initialization model is when $y_{u, -1}^* = 0$, which occurs for units $u \in \{3, 4\}$, which correspond to the Grand'Anse and Nippes departments, respectively.
-When this is the case, each of the infectious $I_u(t_0)$, asymptomatic $A_u(t_0)$, and bacterial reservoir $W_u(t_0)$ compartments have a value of zero.
-Because Model~3 models cholera transmission primarily by means of the bacterial reservoir, this makes it nearly impossible for an outbreak to occur.
-Therefore for units $u \in \{3, 4\}$, we introduce initial value parameters $I_{0,0}^3$ and $I_{0,0}^4$, and calibrate these parameter values using the data.
-The resulting parameter estimates are used to obtain the remaining non-zero initial values of the latent states using Eqs.~\myeqref{eq:model3:A0}--\myeqref{eq:model3:B0}.
-
-% Argue why we didn't need to do this for Model 2.
-
-% TODO: Check this section.
-
-\section{Calibrating Model~3 to observed cases}\label{sec:mod3Cal}
-
-<<CalibrateOrig3>>=
-# Create vectors for the unit and shared parameters
-unit_specific_names <- c("betaB", "foi_add")
-shared_param_names <- c(
-  "mu_B", "XthetaA", "thetaI", "lambdaR", "r", "std_W",
-  "epsilon", "k"
-)
-est_param_names <- c(
-  unit_specific_names, shared_param_names
-)
-
-# Add unit numbers to each parameter
-est_param_names_expanded <- paste0(rep(est_param_names, each = 10), 1:10)
-
-# Simple function that is used to set the rw_sd for the first search
-# This function is convenient so that we don't have to type out the rw_sd
-# for each unit parameter.
-set_rw_1 <- function(x) {
-  if (gsub("[[:digit:]]+$", "", x) %in% shared_param_names) {
-    return(0.01)
-  # } else if (x %in% paste0(rep(c("aHur", 'hHur'), each = 2), c(3, 9))) {
-  #   return(expression(ifelse(time >= 2016.754 & time <= 2017, 0.015, 0)))
-  # } else if (x %in% c("Iinit3", "Iinit4")) {
-  #   return(expression(ivp(0.25)))
-  } else if (grepl('^foi_add[[:digit:]]+$', x)) {
-    return(0.015)
-  } else if (grepl('^betaB[[:digit:]]+$', x)) {
-    return(0.01)
-  } else {
-    return(0)
-  }
-}
-
-set_rw_2 <- function(x) {
-  if (gsub("[[:digit:]]+$", "", x) %in% shared_param_names) {
-    return(0.003)
-  # } else if (x %in% paste0(rep(c("aHur", 'hHur'), each = 2), c(3, 9))) {
-  #   return(expression(ifelse(time >= 2016.754 & time <= 2017, 0.005, 0)))
-  # } else if (x %in% c("Iinit3", "Iinit4")) {
-  #   return(expression(ivp(0.125)))
-  } else if (grepl('^foi_add[[:digit:]]+$', x)) {
-    return(0.005)
-  } else if (grepl('^betaB[[:digit:]]+$', x)) {
-    return(0.003)
-  } else {
-    return(0)
-  }
-}
-
-# First Search RW
-reg_rw_1.sd <- lapply(est_param_names_expanded, set_rw_1)
-names(reg_rw_1.sd) <- est_param_names_expanded
-chol_rw_1.sd <- do.call(rw_sd, reg_rw_1.sd)
-
-# Second Search RW
-reg_rw_2.sd <- lapply(est_param_names_expanded, set_rw_2)
-names(reg_rw_2.sd) <- est_param_names_expanded
-chol_rw_2.sd <- do.call(rw_sd, reg_rw_2.sd)
-
-SEARCH1 = list(
-  NBPF = 100,
-  NP = 2500,
-  SPAT_REGRESSION = 0.05,
-  NREPS = 144,
-  NP_EVAL = 2500,
-  NREPS_EVAL = 18,
-  RW_SD = chol_rw_1.sd,
-  COOLING = 0.5,
-  KEEP_TRACES = FALSE,
-  KEEP_LIKE_MAT = FALSE
-)
-
-SEARCH2 = list(
-  TOP_N = 12,
-  NBPF = 100,
-  NP = 2500,
-  SPAT_REGRESSION = 0.05,
-  NREPS = 6,
-  NP_EVAL = 2500,
-  NREPS_EVAL = 18,
-  RW_SD = chol_rw_2.sd,
-  COOLING = 0.5,
-  KEEP_TRACES = FALSE,
-  KEEP_LIK_MAT = TRUE
-)
-
-haiti3_sub_fit <- bake(
-  file = "../model3/si/subset_noparams.rds", {
-    fit_haiti3(
-      nsearches = 2L,
-      search1 = SEARCH1,
-      search2 = SEARCH2,
-      search_rho = FALSE,
-      search_gamma = FALSE,
-      search_Iinit = FALSE,
-      search_hur = FALSE,
-      ncores = cores
-    )
-  }
-)
-
-# Find model index with best log-likelihood
-best_m <- which.max(haiti3_sub_fit$search2$logLiks$logLik)
-
-# Find parameters using best_m index
-sub_params <- haiti3_sub_fit$search2$params[best_m, ]
-
-# Get unit log-likelihoods for the best model.
-all_liks <- as.data.frame(haiti3_sub_fit$search2$likMat)
-# all_liks$which <- rep(1:(SEARCH3$TOP_N * SEARCH3$NREPS), each = SEARCH3$NREPS_EVAL)
-all_liks$which <- rep(1:(12 * 6), each = 18)
-best_mod <- filter(all_liks, which == best_m)
-
-# Pivot longer to get estimate of log-likelihood for each unit.
-unitLikesM3 <- best_mod %>%
-  pivot_longer(
-    cols = -which,
-    names_to = "unit",
-    names_prefix = "V",
-    values_to = "logLik"
-  ) %>%
-  select(-which) %>%
-  group_by(unit) %>%
-  summarize(
-    ll_est = logmeanexp(logLik),
-    ll_se = logmeanexp(logLik, se = TRUE)[2]
-  ) %>%
-  ungroup() %>%
-  mutate(unit = as.numeric(unit)) %>%
-  arrange(unit)
-@
-
-<<M3_bench, echo=FALSE, message=FALSE, include=FALSE, cache=TRUE>>=
-h3 <- haiti3_spatPomp()
-deps <- h3@unit_names
-
-mod3_cases <- as.data.frame(t(h3@data))
-colnames(mod3_cases) <- deps
-m3_inits <- c(7362, 1914, 22, 8, 2818, 82, 2499, 6953, 44, 7) # TODO: Also fit by hand, but could be changed to initializing to the observation before.
-names(m3_inits) <- deps
-
-bench_cond_lls <- list()
-for (dep in deps) {
-
-  # Get department cases
-  cases <- as.numeric(mod3_cases[, dep])
-  mod <- ar1_NegBinom(data = cases, init = m3_inits[dep])
-
-  bench_cond_lls[[dep]] <- haitipkg:::.ar1_NegBinom_cond_ll(
-    theta = mod$theta, init = m3_inits[dep], data = cases, transform = FALSE
-  )
-}
-
-bench_cond_lls <- as.data.frame(bench_cond_lls)
-
-bench_lls <- apply(bench_cond_lls, 2, sum)
-
-unitLikes_bench <- data.frame(
-  unit = 1:10,
-  # dep = names(bench_lls),
-  ll = bench_lls
-) %>%
-  as_tibble()
-
-bench_cond_lls$date <- lubridate::round_date(lubridate::date_decimal(h3@times), unit = 'day')
-@
-
-<<combine_bench_M3, echo=FALSE>>=
-
-# Combine tables with model 3 and benchmark likelihoods.
-bench_M3_joined <- dplyr::left_join(
-  x = unitLikesM3,
-  y = unitLikes_bench,
-  by = "unit"
-) %>%
-  select(-ll_se) %>%
-  mutate(unit = as.character(unit)) %>%
-  pivot_longer(
-    cols = -unit,
-    names_to = "type",
-    values_to = "ll"
-  ) %>%
-  mutate(unit = paste0(unit, ": ", deps[as.numeric(unit)]))
-@
-
-<<M3bpf, echo=FALSE, message=FALSE, include=FALSE>>=
-
-# Change model parameters to the best set of parameters from previous
-# search
-coef(h3)[names(sub_params)] <- sub_params
-
-registerDoRNG(40765101)  # for modular, reproducible results
-
-# Perform block partical filter with fitted model, to get conditional
-# log-likelihoods.
-bpf_sub3_out <- bake(
-  file = "../model3/si/bpf_sub_cond_LL.rds", {
-
-    results <- list()
-
-    bpf_out <- h3 %>% bpfilter(
-      Np = 5000,
-      block_size = 1,
-      save_states = TRUE
-    )
-
-    results$block.cond.loglik <- bpf_out@block.cond.loglik
-    results$times <- bpf_out@times  # Saved for convenience
-    results
-  },
-  seed = 432235
-)
-@
-
-In this section, we provide more detail on the process that was used to estimate the coefficients used in Model~3.
-In particular, we discuss why we decided to include additional model parameters---those that are associated with the behavior of the system during Hurricane Matthew---that were not considered by \cite{lee20}.
-To calibrate this model, we used the iterated block particle filter (IBPF) method of \cite{ionides22}.
-Due to the novelty of this algorithm, there does not exist published examples the IBPF algorithm used for data analysis outside of the papers in which the algorithm was introduced \citep{ning21ibpf,ionides22}, which at least partially motivates the inclusion additional details in this section.
-
-\cite{lee20} were only able to estimate model parameters to a simplified version of Model~3 on a subset of the available data, as no method existed at the time of their publication to fit a fully coupled meta-population model to disease incidence data.
-Building on their results, we fit the fully coupled version of Model~3 to nearly all available data, reserving only a few observations to use to calibrate the initial conditions of the model (see the previous section for more details).
-In order to properly maximize the model likelihood, we found that it was necessary to use multiple searches for the MLE, periodically pruning away less successful searches.
-The first collection of searches was performed by obtaining initial values for the parameters by uniformly sampling values from a predefined hypercube.
-A subsequent refinement search used starting values corresponding to the largest model likelihoods obtained in the previous round.
-This is similar to the approach used by \cite{ionides22}.
-
-We use the technique described above to fit the fully coupled version of Model~3 proposed by \cite{lee20}.
-The maximum likelihood we obtained with the model was $\Sexpr{round(max(haiti3_sub_fit$search2$logLiks$logLik))}$, which is higher than the benchmark model ($\Sexpr{round(sum(unitLikes_bench$ll))}$).
-While beating a simple associative benchmark is promising, this does not immediately rule out the possibility that the model is a poor description of the system.
-Additional investigation of parameters estimates and their corresponding implications on model based conclusions should always be conducted.
-For meta-population models, it is worth considering how well the model fits the data to each spatial unit.
-This can be done by looking at conditional log-likelihoods, which is part of the output of the \texttt{bpfilter} algorithm in the \texttt{spatPomp} package.
-The likelihoods for each department, compared to the corresponding benchmark model, are displayed in Fig.~\ref{fig:h3UnitLikes}.
-The figure demonstrates that while the log-likelihood of the fitted model outperforms the auto-regressive negative binomial benchmark model, Model~3 has lower likelihoods than the benchmark for some departments.
-
-\begin{figure}[ht]
-<<h3UnitLikes, fig.height=3, fig.width=5.5, echo=FALSE>>=
-ggplot(bench_M3_joined, aes(x = ll, y = reorder(unit, ll))) +
-  geom_point(aes(col = type)) +
-  ylab("Unit") +
-  xlab("Log Likelihood") +
-  scale_color_manual(
-    labels = c("ll" = "Benchmark", "ll_est" = "Model 3"),
-    values = c("ll" = "#1b9e77", "ll_est" = "#d95f02")
-  ) +
-  theme(legend.title = element_blank(),
-        axis.title.y = element_blank())
-@
-\caption{\label{fig:h3UnitLikes}Log-likelihoods of Model~3 for each department compared to the corresponding benchmark model prior to the inclusion of parameters for initial values and Hurricane Matthew.}
-\end{figure}
-
-In addition to considering the conditional log-likelihoods of each unit, one can consider conditional log-likelihoods of each observation.
-When compared to a benchmark, this level of detail can provide useful information about which observations are well described by the model and which are poorly described.
-In Fig.~\ref{fig:condLL}, we plot the conditional log-likelihoods of each observation.
-This figure reveals that the fitted model has difficult time explaining certain features of the data.
-For example, many departments (in particular Sud) have observations with lower conditional log-likelihoods near October 2016 than at other time points.
-Further investigation of the model output reveals that the model is struggling to explain the sudden surge in cholera cases that occurred at this time, which coincides with the time that Hurricane Matthew struck Haiti.
-While the model does include a mechanism to account for increased cholera transmission due to large rainfall events, the mechanism does not appear to be sufficient to capture the damaging effects of the hurricane, which had the greatest impact in the the Sud and Grand'Anse departments \citep{ferreirai16}.
-This result led us to include parameters $\Whur{u}$ and $\hHur{u}$ in Eq.~23 of the main text, which allows for an increase in the transmission rate between environmental cholera and humans for Sud and Grand'Anse during and after the hurricane.
-The effect of the hurricane on cholera transmission is assumed to have an exponential decay, where the magnitude is determined by $\Whur{u}$ and the duration of the effect determined by $\hHur{u}$.
-
-\begin{figure}[ht]
-<<CondLLFig, echo=FALSE, message=FALSE, fig.height=3>>=
-
-# Changing labels to human-readable labels.
-deps[grepl("_", deps)] <- gsub("_", "-", deps[grepl("_", deps)])
-deps[deps == "Grande-Anse"] <- "Grand\'Anse"
-names(deps) <- as.character(1:10)
-myLabeller <- as_labeller(deps)
-
-# Plot conditional log-likelihoods for each unit
-t(bpf_sub3_out$block.cond.loglik) %>%
-  as.data.frame() %>%
-  mutate(t = bpf_sub3_out$times) %>%
-  pivot_longer(
-    cols = -t,
-    values_to = 'likelihood',
-    names_prefix = "V",
-    names_to = 'Unit'
-  ) %>%
-  mutate(date = as.Date(lubridate::date_decimal(t))) %>%
-  ggplot(aes(x = date, y = likelihood)) +
-  geom_line() +
-  geom_point() +
-  facet_wrap(~Unit, labeller = myLabeller, scales = 'free_y',
-             nrow = 2) +
-  theme_bw() +
-  theme(axis.title.x = element_blank(),
-        axis.text.x = element_text(angle = -45, hjust = 0)) +
-  ylab("Conditional log-likelihoods")
-@
-\caption{\label{fig:condLL}Conditional log-likelihoods of Model~3 prior to the inclusion of parameters for initial values and Hurricane Matthew.}
-\end{figure}
-
-Other points of concern include the observations at or near the start of the recorded cases, for example, in Nord-Ouest and Sud-Est.
-These low conditional log-likelihoods are a result of having zero or near zero confirmed cases in the first week of observed data, followed by a rapid growth of cases.
-The model cannot explain the rapid growth of cholera in the absence of starting cases because the initialization scheme forces departments with no recorded cases in the first week to have hidden states $I_{u0}(0) = A_{u0}(0) = B_{u}(0) = 0$.
-In reality, however, it is possible to have some infected or asymptomatic individuals in a department when no cases are recorded due to under-reporting.
-This may be particularly true at the onset of an infectious disease outbreak, as lower public awareness and inadequate testing infrastructure may result in many cases going undetected.
-We therefore include and estimate $I_{0u}$ that determines the initial number of Infected individuals at the start of the pandemic for the departments Grande'Anse, Nord-Ouest, and Sud-Est;
-each of these departments have both low starting case counts followed by rapid growth, and low initial values of the conditinoal log-likelihood.
-
-<<loadFinalModel3, echo=FALSE, include=FALSE, message=FALSE>>=
-
-# This result was used in the manuscript, so it doesn't need to be recalculated.
-h3_final_results <- readRDS("../model3/run_level_3/haiti3_fit.rds")
-
-# Save best_m as index for the model parameters with the best likelihood.
-# Then, save the model parameter values and the likelihood evaluations
-best_m <- which.max(h3_final_results$search2$logLiks$logLik)
-best_params <- h3_final_results$search2$params[best_m, ]
-all_liks <- as.data.frame(h3_final_results$search2$likMat)
-all_liks$which <- rep(1:72, each = 18)
-best_mod <- dplyr::filter(all_liks, which == best_m)
-
-# Get repeated estimates of unit log-likelihoods
-unitLikesM3 <- best_mod %>%
-  pivot_longer(
-    cols = -which,
-    names_to = "unit",
-    names_prefix = "V",
-    values_to = "logLik"
-  ) %>%
-  select(-which) %>%
-  group_by(unit) %>%
-  summarize(
-    ll_est = pomp::logmeanexp(logLik),
-    ll_se = pomp::logmeanexp(logLik, se = TRUE)[2]
-  ) %>%
-  ungroup() %>%
-  mutate(unit = as.numeric(unit)) %>%
-  arrange(unit)
-
-# Combine model likelihood table to benchmarks
-bench_M3_joined <- dplyr::left_join(
-  x = unitLikesM3,
-  y = unitLikes_bench,
-  by = "unit"
-) %>%
-  select(-ll_se) %>%
-  mutate(unit = as.character(unit)) %>%
-  pivot_longer(
-    cols = -unit,
-    names_to = "type",
-    values_to = "ll"
-  ) %>%
-  mutate(unit = paste0(unit, ": ", deps[as.numeric(unit)]))
-@
-
-We refit Model~3 after introducing these seven parameters---three initial value parameters and four parameters related to Hurricane Matthew---using the same successive search scheme described above.
-The resulting model has a log-likelihood value of $\Sexpr{round(max(h3_final_results$search2$logLiks$logLik), 1)}$.
-We also provide figures for department specific log-likelihoods and conditional log-likelihoods in Figs.~\ref{fig:finalUnitLL} and \ref{fig:finalCondLL}.
-
-\begin{figure}[ht]
-<<finalUnitLLFig, fig.height=3, fig.width=5.5, echo=FALSE>>=
-ggplot(bench_M3_joined, aes(x = ll, y = reorder(unit, ll))) +
-  geom_point(aes(col = type)) +
-  ylab("Unit") +
-  xlab("Log Likelihood") +
-  scale_color_manual(
-    labels = c("ll" = "Benchmark", "ll_est" = "Model 3"),
-    values = c("ll" = "#1b9e77", "ll_est" = "#d95f02")
-  ) +
-  theme(legend.title = element_blank(),
-        axis.title.y = element_blank())
-@
-\caption{\label{fig:finalUnitLL}Log-likelihoods of Model~3 for each department compared to the corresponding benchmark model after adding and estimating parameters related to initial values and Hurricane Matthew.}
-\end{figure}
-
-<<finalModel3BPF, echo=FALSE, message=FALSE, include=FALSE>>=
-
-# Set the model parameters to the best results from previous search
-coef(h3)[names(best_params)] <- best_params
-
-# block particle filter to get conditional log-likelihoods.
-bpf_final3_out <- bake(
-  file = "../model3/si/h3_final_bpf_cond_ll.rds", {
-
-    results <- list()
-
-    bpf_out <- h3 %>% bpfilter(
-      Np = 5000,
-      block_size = 1,
-      save_states = TRUE
-    )
-
-    results$Sus_df <- foreach(i=1:length(h3@times), .combine = bind_rows) %do% {
-      get_h3_S_quants(bpf_out, i)
-    }
-
-    results$block.cond.loglik <- bpf_out@block.cond.loglik
-    results$times <- bpf_out@times  # saved for convenience
-    results
-  },
-  seed = 786321
-)
-@
-
-\begin{figure}[ht]
-<<finalCondLLFig, echo=FALSE, message=FALSE, fig.height=3>>=
-t(bpf_final3_out$block.cond.loglik) %>%
-  as.data.frame() %>%
-  mutate(t = bpf_final3_out$times) %>%
-  pivot_longer(
-    cols = -t,
-    values_to = 'likelihood',
-    names_prefix = "V",
-    names_to = 'Unit'
-  ) %>%
-  mutate(date = as.Date(lubridate::date_decimal(t))) %>%
-  ggplot(aes(x = date, y = likelihood)) +
-  geom_line() +
-  geom_point() +
-  facet_wrap(~Unit, labeller = myLabeller, scales = 'free_y',
-             nrow = 2) +
-  theme_bw() +
-  theme(axis.title.x = element_blank(),
-        axis.text.x = element_text(angle = -45, hjust = 0)) +
-  ylab("Conditional log-likelihoods")
-@
-\caption{\label{fig:finalCondLL}Conditional log-likelihoods of Model~3 after adding and estimating the parameters for initial values and Hurricane Matthew.}
-\end{figure}
-
-While the addition of the Hurricane parameters seems to have resulted in a dramatic increase of conditional likelihoods around and after October 2016, the inclusion of the initial value parameters does not appear to have fixed the low conditional log-likelihood values in Sud-Est and Nord-Ouest.
-This could be due to many factors, including the possibility that the coupling between units forces low initial value parameters in a given department in order for the model to fit the data in the remaining departments.
-Whatever the case may be, the inclusion of the seven parameters resulted in a $\Sexpr{round(max(h3_final_results$search2$logLiks$logLik) - max(haiti3_sub_fit$search2$logLiks$logLik), 1)}$ increase in log-likelihood, which is a highly statistically significant difference.
-One could similarly test whether the inclusion of the initial value parameters resulted in a significantly better fit to the observed data, but we do not do this here due to the large computational effort that would be required.
-
-\subsection{Examining the Hidden States of the Calibrated Model}
-
-For mechanistic models, beating a suitable statistical benchmark does not alone guarantee that the model provides an accurate description of a dynamic process.
-Indeed, a good statistical fit does not require the model to assert a causal explanation.
-For example, reconstructed latent variables should make sense in the context of alternative measurements of these variables \citep{grad12}.
-We demonstrate this principle by examining the latent state of the calibrated model.
-In particular, we examine the compartment of susceptible individuals under various scenarios, as this compartment drives transmission dynamics in SIR models for infectious disease outbreaks.
-
-Recall that the filtering distribution for the calibrated version of Model~3 at time $t_k$ is defined as the distribution of the latent state at time $t_k$ given the data from times $t_{1}:t_{k}$, i.e.
-$f^{(3)}_{\bm{X}_k|\bm{Y}_{1:k}}(\bm{x}_{k} | \bm{y}^*_{1:k} ; \hat\theta)$.
-In general, one may expect simulations from the filtering distribution of a model with a good statistical fit to result in hidden states that are highly consistent with the observed data because the filtering distribution is conditioned on the observed data.
-Fig.~\ref{fig:h3Sus} shows the percentage of individuals that are in the susceptible compartment from various simulations of the model:
-simulations from Model~3 under initial conditions are displayed in red; simulations from the filtering distribution of model are displayed in blue.
-This figure shows that simulations from initial conditions tends to result in a much more rapid depletion of susceptible individuals at the start of the epidemic than simulations from the filtering distribution, suggesting the calibrated model has a propensity to predict larger outbreaks than what is typically seen in the data.
-This also appears true in later stages of the time period of observed data, where the data suggests that there is a replenishment of the susceptible compartment, but the Model simulations typically retain a smaller proportion of susceptible individuals.
-This result demonstrates that the calibrated model favors a more rapid growth in cholera cases than what is typically seen in the observed data, which explains why the model fails to predict the absence of cases between February, 2019 and October, 2022.
-
-<<h3SusFigSetup, echo=FALSE, message=FALSE, include=FALSE>>=
-
-# This bake perfoms the following:
-#  (1) simulates the model from initial conditions
-#  (2) get the values of the hidden states at the end of each simulation, and
-#      saves these for later use.
-#  (3) Obtain quantiles for the Susceptible pool from the simulations.
-init_conditions_results <- bake(
-  file = "../model3/si/InitSimsQuants.rds", {
-
-    results <- list()  # To return final output of bake
-
-    # Simulate from the fitted model
-    h3Spat_sims <- simulate(
-      h3, nsim = 2000, format = 'data.frame'
-    )
-
-    # Get states at the end of the simulations
-    h3_spat_end <- h3Spat_sims %>%
-      filter(time == max(time))
-
-    # Save hidden state values at the end of the simulation. Saved as a matrix
-    # in the same format as the output from a block-particle filter, so that
-    # it can be used as the input to project_from_filter.
-    results[["end_conditions"]] <- h3_spat_end %>%
-      select(-time, -cases) %>%
-      mutate(
-        unit = case_when(
-          unitname == "Artibonite" ~ 1,
-          unitname == "Centre" ~ 2,
-          unitname == "Grande_Anse" ~ 3,
-          unitname == "Nippes" ~ 4,
-          unitname == "Nord" ~ 5,
-          unitname == "Nord_Est" ~ 6,
-          unitname == "Nord_Ouest" ~ 7,
-          unitname == "Ouest" ~ 8,
-          unitname == "Sud" ~ 9,
-          unitname == "Sud_Est" ~ 10
-        )
-      ) %>%
-      select(-unitname) %>%
-      pivot_wider(
-        data = .,
-        names_from = unit,
-        values_from = c(S, I, A, R_one, R_two, R_three, VSd, VR1d, VR2d,
-                        VR3d, VSdd, VR1dd, VR2dd, VR3dd, VSd_alt, VR1d_alt,
-                        VR2d_alt, VR3d_alt, VSdd_alt, VR1dd_alt, VR2dd_alt,
-                        VR3dd_alt, C, B, Doses, totInc),
-        names_sep = ""
-      ) %>%
-      select(-.id) %>%
-      as.matrix() %>%
-      t()
-
-    # Get the quantiles for the Susceptible pool from the simulations
-    results[["SQuants"]] <- h3Spat_sims %>%
-      rename(dep = unitname) %>%
-      group_by(dep, time) %>%
-      summarise(
-        Q025 = quantile(S + VSd + VSdd + VSd_alt + VSdd_alt, probs = 0.025),
-        Q50  = quantile(S + VSd + VSdd + VSd_alt + VSdd_alt, probs = 0.5),
-        Q975 = quantile(S + VSd + VSdd + VSd_alt + VSdd_alt, probs = 0.975)
-      ) %>%
-      ungroup() %>%
-      rowwise() %>%
-      mutate(
-        unit_num = which(h3@unit_names == dep),
-        date = lubridate::date_decimal(time),
-        Q025 = Q025 / h3@params[paste0("H", unit_num)],
-        Q50  = Q50  / h3@params[paste0("H", unit_num)],
-        Q975 = Q975 / h3@params[paste0("H", unit_num)]
-      ) %>%
-      filter(
-        date >= as.Date(
-          lubridate::round_date(lubridate::date_decimal(h3@t0), unit = 'day')
-        )
-      ) %>%
-      mutate(date = as.Date(lubridate::round_date(date, unit = 'day')))
-
-    results
-  },
-  seed = 834687
-)
-
-# Get the output of the "bake" call for plotting.
-h3Spat_quants <- init_conditions_results[["SQuants"]]
-
-# This bake call uses the ending hidden states from above to forecast cholera
-# cases in the future under the V0 vaccination scenario
-mod3_V0_initcond_res <- bake(
-  file = '../model3/si/vacc_sim_from_init.rds', {
-    h3_V0_res <- list()
-
-    # Get "noVacc" scenario parameters
-    h3_S0_par <- get_model3_vacc_scenario_params(scenario = "noVacc")
-    h3Spat <- h3
-    p3 <- best_params
-    coef(h3Spat) <- p3
-    coef(h3Spat)[names(h3_S0_par)] <- h3_S0_par
-
-    # store the end states that will be used as initial values
-    h3_bpf_end_states <- init_conditions_results$end_conditions
-
-    # Using parameters and end state values, obtain projections
-    h3_V0_sims <- project_from_filter(
-      h3Spat, end_states = h3_bpf_end_states, covarGen = project_rain,
-      nsims = 1000, seed = 184697
-    )
-
-    # Store nationally aggregated results
-    h3_V0_res$mod3_V0_sims <- agg_mod3_sims(h3_V0_sims)
-    h3_V0_res$mod3_V0_probs <- get_elimProbs(h3_V0_sims, model = 3)
-
-    # Get quantiles of observed cases for each department and time
-    h3_V0_res$UnitQuants <- h3_V0_sims %>%
-      mutate(
-        date = as.Date(  # Get date object from decimal time
-          lubridate::round_date(lubridate::date_decimal(time), unit = 'day')
-        )
-      ) %>%
-      group_by(unitname, date) %>%
-      summarize(
-        Q025 = quantile(cases, probs = 0.025),
-        Q50 = quantile(cases, probs = 0.5),
-        Q975 = quantile(cases, probs = 0.975)
-      ) %>%
-      ungroup() %>%
-      rename(dep = unitname)
-
-    # Get quantiles of susceptible individuals for each department and time
-    h3_V0_res$SQuants <- h3_V0_sims %>%
-      mutate(
-        date = as.Date(
-          lubridate::round_date(lubridate::date_decimal(time), unit = 'day')
-        )
-      ) %>%
-      group_by(unitname, date) %>%
-      mutate(
-        # Get the population for each unit
-        pop = h3@params[paste0("H", which(first(unitname) == unit_names(h3)))]
-      ) %>%
-      summarize(  # Calculate the quantiles
-        Q025 = quantile(
-          (S + VSd + VSdd + VSdd_alt + VSd_alt) / pop,
-          probs = 0.025
-        ),
-        Q50 = quantile(
-          (S + VSd + VSdd + VSdd_alt + VSd_alt) / pop,
-          probs = 0.5
-        ),
-        Q975 = quantile(
-          (S + VSd + VSdd + VSdd_alt + VSd_alt) / pop,
-          probs = 0.975
-        )
-      ) %>%
-      ungroup() %>%
-      rename(dep = unitname)
-
-    h3_V0_res
-  }
-)
-
-# This result was already used in the manuscript; it doesn't need to
-# be recalculated.
-mod3_V0_res <- readRDS("../model3/run_level_3/mod3_V0_res.rds")
-@
-
-\begin{figure}[ht]
-<<h3SusFig, echo=FALSE, message=FALSE, fig.height=3>>=
-ggplot() +
-  # Sims from initial condition
-  geom_line(data = h3Spat_quants, aes(x = date, y = Q50), col = 'red') +
-  geom_ribbon(
-    data = h3Spat_quants,
-    aes(x = date, ymin = Q025, ymax = Q975),
-    fill = 'red', alpha = 0.5
-  ) +
-  # projections from initial condition, using filtered sims function
-  geom_line(data = mod3_V0_initcond_res$SQuants, aes(x = date, y = Q50), col = 'red') +
-  geom_ribbon(
-    data = mod3_V0_initcond_res$SQuants,
-    aes(x = date, ymin = Q025, ymax = Q975),
-    fill = 'red', alpha = 0.5
-  ) +
-  # Sims from filtered distribution
-  geom_line(
-    data = bpf_final3_out$Sus_df,
-    aes(x = as.Date(lubridate::date_decimal(time)), y = Q50),
-    col = 'blue'
-  ) +
-  geom_ribbon(
-    data = bpf_final3_out$Sus_df,
-    aes(x = as.Date(lubridate::date_decimal(time)), ymin = Q025, ymax = Q975),
-    fill = 'blue', alpha = 0.5
-  ) +
-  # Projections from filtered distribution
-  geom_line(data = mod3_V0_res$SQuants, aes(x = date, y = Q50), col = 'blue') +
-  geom_ribbon(
-    data = mod3_V0_res$SQuants,
-    aes(x = date, ymin = Q025, ymax = Q975),
-    fill = 'blue', alpha = 0.5
-  ) +
-  geom_vline(xintercept = max(h3Spat_quants$date), linetype = 'dashed') +
-  facet_wrap(~dep, scales = 'free_y', nrow = 2) +
-  theme_bw() +
-  theme(axis.title.x = element_blank(),
-        axis.text.x = element_text(angle = -45, hjust = 0)) +
-  ylab("Total Susceptible (%)")
-@
-\caption{\label{fig:h3Sus}Percentage of individuals that are in the susceptible compartment.
-Simulations from Model~3 under initial conditions are displayed in red; simulations from the filtering distribution of model are displayed in blue.
-The dashed line represents the end of the observed data.}
-\end{figure}
-
-\clearpage
-
-\section{Replication of \cite{lee20}}
-
-In this article we claimed that we were able to obtain better fits to the observed data using the same models that were proposed by \cite{lee20}.
-Along with visual comparisons to the data, this claim was supported by comparing likelihoods and AIC values in Table~2 in the manuscript. % TODO: Double check the table number matches
-Because model likelihoods were not provided by \cite{lee20}, it is necessary to replicate these models in order to obtain likelihood estimates.
-Here we would like to thank the authors of \cite{lee20}, who provided detailed descriptions of their models, which enabled us to build on their work.
-In the following subsections, we use our \code{R} package \code{haitipkg} to reproduce some of the results of \cite{lee20}.
-This reproduction allows us to estimate the likelihoods of the \cite{lee20} version of Models~1--3, and also provides a demonstration of the importance and usefulness of reproducible research.
-
-\subsection{Model~1 Replication}
-
-The model was implemented by a team at Johns Hopkins Bloomberg School of Public Health (hereafter referred to as the Model~1 authors) in the \code{R} programming language using the \code{pomp} package \citep{king16}.
-Original source code is publicly available with DOI: 10.5281/zenodo.3360991.
-The final results reported by the Model~1 authors were obtained by using several different parameter sets rather than a single point estimate.
-According to the supplement materials, this was because model realizations from a single parameter set retained substantial variability, but multiple realizations from a collection of parameter sets resulted in a reasonable visual fit to the data.
-We are also inclined to believe that the use of multiple parameter values was in part intended to account for parameter uncertainty---the importance of which was discussed in the main text---an effort by the Model~1 authors that we applaud.
-Simulations from each of the parameter sets, however, were treated with equal importance when being used to diagnose the model fit and make inference on the system.
-This is problematic given Figures S8 and S9 of the supplement material, which suggest that some parameter sets that were used for inference may have been several hundred units of log-likelihood lower than other parameter sets that were simultaneously used to make forecasts.
-Such a large difference in log-likelihoods is well beyond the threshold of statistical uncertainty determined by Wilks' theorem, resulting in the equal use of statistically inferior parameter sets in order to make forecasts and conduct inference on the system.
-
-To fully reproduce the results of the Model~1 authors, it is necessary to use the exact same set of model parameters that were originally used to obtain the results presented by \cite{lee20}.
-Because these parameter sets were not made publicly available, we relied on the source code provided by the Model~1 authors to approximately recreate the parameter set.
-Due to software updates since the publication of the source code, we were unable to produce the exact same set of parameters.
-Running the publicly available source code, however, resulted in a set of parameters that are visually similar to those used by the Model~1 authors (See Figures~\ref{fig:PlotEpiDist} and \ref{fig:plotEndParams}).
-Furthermore, simulations using the set of parameters produced by the source code appear practically equivalent to those displayed by \cite{lee20} (See Figure~\ref{fig:plotMod1Sims}).
-
-<<Simulate Lee Model 1, echo=FALSE, cache=TRUE>>=
-h1_epi <- haiti1()
-N_LEE_SIMS <- 20
-
-registerDoRNG(18599687)
-foreach(
-  i = 1:length(h1LeeStartsEpi),
-  .combine = rbind
-) %dopar% {
-  guess <- h1LeeStartsEpi[i, ]
-  sims <- simulate(h1_epi, nsim = N_LEE_SIMS, format = 'data.frame',
-                   params = guess)  # 20 sims for each set of parameters
-  sims$param_set <- i
-  sims %>%
-    dplyr::select(.id, week, cases, param_set)
-} -> all_sims
-@
-
-<<PlotEpiDist, echo=FALSE, fig.cap="Compare to Figure S8 of Lee et al. (2020) supplement.", fig.width=7, fig.height=7, fig.align='center'>>=
-GGally::ggpairs(
-  h1LeeStartsEpi,
-  columns = c("rho", "tau", "beta1", "nu", "loglik")
-)
-@
-
-<<SimulateEpi, echo=FALSE, include=FALSE, message=FALSE, cache=TRUE>>=
-h1_end <- haiti1(period = "endemic")
-
-true_agg_cases <- haitiCholera %>%
-  select(-report) %>%
-  pivot_longer(
-    data = .,
-    cols = -date_saturday,
-    values_to = 'cases',
-    names_to = 'dep'
-  ) %>%
-  mutate(
-    date = as.Date(date_saturday),
-    dep = gsub("\\.", "_", dep)
-  ) %>%
-  mutate(
-    dep = case_when(dep == "Grand_Anse" ~ "Grande_Anse", TRUE ~ dep)
-  ) %>%
-  tidyr::pivot_wider(
-    data = .,
-    id_cols = c(date),
-    names_from = dep,
-    values_from = cases,
-    names_prefix = 'cases_'
-  ) %>%
-  mutate(
-    ReportedAll = cases_Artibonite + cases_Centre +
-      cases_Grande_Anse + cases_Nippes + cases_Nord +
-      cases_Nord_Est + cases_Ouest + cases_Sud +
-      cases_Sud_Est + cases_Nord_Ouest
-  )
-@
-
-<<FitEnd, echo=FALSE, message=FALSE, include=FALSE>>=
-foreach(
-  i = 1:nrow(h1LeeStartsEnd),
-  .combine = rbind
-) %dopar% {
-  r_params <- h1LeeStartsEnd[i, ][names(coef(h1_end))]
-  coef(h1_end) <- r_params
-
-  sims <- simulate(
-    h1_end, nsim = N_LEE_SIMS,
-    format = 'data.frame',
-    params = r_params
-  )
-  sims$param_set <- i
-  sims %>%
-    dplyr::select(.id, week, cases, param_set)
-} -> end_sims
-
-@
-
-<<plotEndParams, fig.width=7, fig.height=7, fig.align='center', fig.cap="Bivariate relationships between variables after fitting endemic period.
-Compare to S9 of Lee et al.
-(2020) supplement.", cache=TRUE, echo=FALSE>>=
-GGally::ggpairs(
-  h1LeeStartsEnd,
-  columns = c("rho", "tau", "beta1", "nu", "loglik")
-)
-@
-
-<<calcMod1LeeLikes, echo=FALSE, include=FALSE, message=FALSE>>=
-h1_epi_evals <- readRDS("../model1/run_level_3/lee1_epi_evals.rds")
-h1_end_evals <- readRDS("../model1/run_level_3/lee1_end_evals.rds")
-
-mod1_all_likes <- dplyr::inner_join(
-  x = h1_epi_evals,
-  y = h1_end_evals,
-  by = "parid"
-  ) %>%
-  dplyr::mutate(
-    joint_ll = pfLL.x + pfLL.y
-  )
-
-# Approximate the likelihood of the Lee et al (2020) version of Model 1 by
-# using the parameter sets that resulted in the maximum likelihood.
-mod1_lee_ll <- max(mod1_all_likes$joint_ll, na.rm = TRUE)
-@
-
-<<plotMod1Sims, echo=FALSE, fig.cap=paste0("Simulations using parameter sets that were generated by running source code provided by Lee et al (2020).
-Compare to Figure S7 of \\cite{lee20sup}.
-The upper bound for the likelihood of this model is ", round(mod1_lee_ll), "."), cache=TRUE, fig.width=6.5, fig.height=3.75>>=
-
-quants <- all_sims %>%
-  filter(!param_set %in% c(1, 2)) %>%  # Remove parameter sets inconsistent with Lee et al (2020).
-  group_by(week) %>%
-  summarize(
-    q025 = quantile(cases, probs = 0.025, na.rm = TRUE),
-    q50  = quantile(cases, probs = 0.500, na.rm = TRUE),
-    q975 = quantile(cases, probs = 0.975, na.rm = TRUE)
-  ) %>%
-  ungroup() %>%
-  mutate(date = lubridate::ymd("2010-10-16") + lubridate::weeks(week))
-
-quants_end <- end_sims %>%
-  group_by(week) %>%
-  summarize(
-    q025 = quantile(cases, probs = 0.025, na.rm = TRUE),
-    q50  = quantile(cases, probs = 0.500, na.rm = TRUE),
-    q975 = quantile(cases, probs = 0.975, na.rm = TRUE)
-  ) %>%
-  ungroup() %>%
-  mutate(date = lubridate::ymd("2010-10-16") + lubridate::weeks(week))
-
-ggplot() +
-  geom_point(data = true_agg_cases, aes(x = date, y = ReportedAll)) +
-  geom_line(data = quants, aes(x = date, y = q50), col = '#0b8a2b', size = 1) +
-  geom_ribbon(
-    data = quants,
-    aes(x = date, ymin = q025, ymax = q975),
-    alpha = 0.5, fill = '#0b9c30'
-  ) +
-  geom_line(data = quants_end, aes(x = date, y = q50), col = '#882871') +
-  geom_ribbon(data = quants_end, aes(x = date, ymin = q025, ymax = q975), alpha = 0.5, fill = '#882871') +
-  ylab("Reported, Symptomatic Incidence") +
-  scale_x_date(date_breaks = "1 year") +
-  theme(axis.title.x = element_blank(),
-        axis.text.x = element_text(hjust = 1, angle = 45))
-@
-
-Because the model forecasts provided by \cite{lee20} come from various sets of parameters---which each correspond to a unique log-likelihood value---it is not obvious how one would obtain an estimate for the log-likelihood of the model that was used for simulations by the Model~1 authors.
-One approach could be to calculate the logarithm of the weighted mean of the likelihoods for each parameter sets used to obtain the forecasts, where the weights are proportional to the number of times the parameter set was used.
-However, in an effort to not underestimate the likelihood of the model of the Model~1 authors, we report the estimated log-likelihood as the log-likelihood value corresponding to the parameter set with the largest likelihood value, even though the majority of simulations were obtained using parameter sets with lower likelihood values.
-In this sense, we consider the log-likelihood reported in Table~1 of the main text to be an upper-bound of the log-likelihood of the model used by \cite{lee20}.
-For each parameter set, the log-likelihood was estimated using a particle filter, implemented as the \texttt{pfilter} function in the \texttt{pomp} package.
-
-\subsection{Model~2 Replication}\label{sec:mod2rep}
-
-Model~2 was developed by a team that consisted of members from the Fred Hutchinson Cancer Research Center and the University of Florida (hereafter referred to as the Model~2 authors).
-While Model~2 is the only deterministic model we considered in our analysis, it contains perhaps the most complex descriptions of cholera in Haiti: Model~2 accounts for movement between spatial units; human-to-human and environment-to-human cholera infections; and transfer of water between spatial units based on elevation charts and river flows.
-
-The source code that the Model~2 authors used to generate their results was written in the \code{Python} programming language, and is publicly available at \url{10.5281/zenodo.3360857} and its accompanying GitHub repository \url{https://github.com/lulelita/HaitiCholeraMultiModelingProject}.
-In order to perform our analysis in a unified framework, we re-implemented this model in the \code{R} programming language using the \code{spatPomp} package \citep{asfaw23arxiv}, which facilitates the creation of meta-population models.
-We note that the travel and water matrices used to implement the complex dynamics in Model~2 \citep{lee20sup} are not available in either the Zenodo archive or the GitHub repository;
-instead, we obtained these matrices via personal correspondence with the Model~2 authors.
-Using these matrices, and the point estimates for model parameters provided by \citep{lee20sup}, we created trajectories of the cholera dynamics and compared this to available data.
-These trajectories, shown in Figure~\ref{fig:mod2rep}, are very similar to the trajectories shown in Figure~S15 of \cite{lee20sup}.
-
-<<Model2Replication, cache=TRUE, echo=FALSE>>=
-h2_epi <- haiti2()
-h2_end <- haiti2(region = 'after')
-
-h2_epi_traj <- trajectory(h2_epi, format = 'data.frame')
-h2_end_traj <- trajectory(h2_end, format = 'data.frame')
-
-h2_epi_traj$Ctotal <- rowSums(h2_epi_traj[, paste0("C", 1:10)]) * 0.2
-h2_end_traj$Ctotal <- rowSums(h2_end_traj[, paste0("C", 1:10)]) * 0.2
-
-dep_plot_df <- haitipkg::haitiCholera %>%
-  select(-report) %>%
-  tidyr::pivot_longer(
-    data = .,
-    cols = -date_saturday,
-    values_to = 'cases',
-    names_to = 'dep'
-  ) %>%
-  mutate(
-    date = as.Date(date_saturday),
-    dep = gsub("\\.", "_", dep)
-  ) %>%
-  mutate(
-    dep = case_when(dep == "Grand_Anse" ~ "Grande_Anse", TRUE ~ dep)
-  )
-
-true_agg_cases <- dep_plot_df %>%
-  tidyr::pivot_wider(
-    data = .,
-    id_cols = c(date),
-    names_from = dep,
-    values_from = cases,
-    names_prefix = 'cases_'
-  ) %>%
-  mutate(
-    ReportedAll = cases_Artibonite + cases_Centre +
-      cases_Grande_Anse + cases_Nippes + cases_Nord +
-      cases_Nord_Est + cases_Ouest + cases_Sud +
-      cases_Sud_Est + cases_Nord_Ouest
-  ) %>%
-  select(date, ReportedAll)
-
-h2_epi_traj$date <- as.Date(
-  lubridate::round_date(
-    lubridate::date_decimal(h2_epi_traj$year),
-    unit = 'day'
-  )
-)
-
-h2_end_traj$date <- as.Date(
-  lubridate::round_date(
-    lubridate::date_decimal(h2_end_traj$year),
-    unit = 'day'
-  )
-)
-
-all_traj <- bind_rows(
-  select(h2_epi_traj, date, Ctotal),
-  select(h2_end_traj, date, Ctotal)
-)
-
-projections <- left_join(
-  x = true_agg_cases,
-  y = all_traj,
-  by = 'date'
-)
-@
-
-\begin{figure}[ht]
-<<plotModel2Rep, cache=TRUE, echo=FALSE, fig.width=6.5, fig.height=4.5>>=
-gg_epi <- projections %>%
-  filter(date < as.Date("03-01-2014", format = '%m-%d-%Y')) %>%
-  ggplot(aes(x = date)) +
-  geom_line(aes(y = ReportedAll), col = 'black', linetype = 'dashed') +
-  geom_line(aes(y = Ctotal), col = 'red', size = 1) +
-  theme_bw() +
-  scale_y_continuous(limits = c(0, 30000), breaks = seq(0, 30000, 5000)) +
-  theme(axis.title.x = element_blank()) +
-  ylab("Reported number of cases")
-
-gg_end <- projections %>%
-  filter(date >= as.Date("03-01-2014", format = '%m-%d-%Y')) %>%
-  ggplot(aes(x = date)) +
-  geom_line(aes(y = ReportedAll), col = 'black', linetype = 'dashed') +
-  geom_line(aes(y = Ctotal), col = 'red', size = 1) +
-  theme_bw() +
-  scale_y_continuous(limits = c(0, 7000), breaks = seq(0, 7000, 1000)) +
-  theme(axis.title.x = element_blank()) +
-  ylab("Reported number of cases") +
-  scale_x_date(date_breaks = "1 year", date_labels = "%Y")
-
-cowplot::plot_grid(gg_epi, gg_end, ncol = 1)
-@
-\caption{\label{fig:mod2rep}
-Model 2 trajectories using the \code{haitipkg}. Compare to Figure S15 of \cite{lee20sup}.
-}
-\end{figure}
-
-There are minor differences between Figure~\ref{fig:mod2rep} and Figure~S15 of \cite{lee20sup}.
-While the discrepancy appears minor, the deterministic nature of Model~2 implies that an exact replication of model trajectories should be possible.
-In this case, these discrepancies may possibly be attributed to implementing the model and plotting the model trajectory in two different programming languages.
-Another potential explanation for the discrepancy is that the parameters that we used are only approximately the same as those used by \citet{lee20sup}.
-For example, the parameters $\transmission$, $\Wbeta{}$ (See Table~\ref{tab:translate}) had reported values of $9.9 \times 10^{-7}$ and $4.03 \times 10^{-2}$, respectively (Table~S13 of \cite{lee20sup}), but were actually fit to data and therefore likely these values have been rounded.
-Additionally, our implementation of Model~2 used a time scale of years and many of the parameters were reported on a weekly scale, so small differences may result due to unit conversions.
-The collective effect of these small differences in model parameters likely will result in small differences in model trajectories.
-
-Some additional concerns about being able to accurately replicate the results of \cite{lee20} are valid.
-Details about the measurement models and how latent states were initialized for the epidemic model were not provided by \citet{lee20sup} and therefore these details must be inferred by looking at the provided source code.
-According to repository comments, the files \code{fit\allowbreak In\allowbreak Pieces\allowbreak 3params\allowbreak Clean\allowbreak May2019\allowbreak Public.py} and \code{fit\allowbreak In\allowbreak Pieces\allowbreak Mu\allowbreak With\allowbreak Frac\allowbreak Sus\allowbreak Fixed\allowbreak All\allowbreak Infections\allowbreak Public.py} were used to fit the epidemic and endemic phases of the model respectively, although it is apparent that these exact files were not used to obtain the reported results since the files contain some variable-naming errors that make it impossible to run the files without making modifications \footnote{One example of why the code cannot be run that the file loads functions from a non-extant file named \code{choleraEqs.py} in line 13 rather than \code{cholera\allowbreak Eqs\allowbreak Public.py}.}.
-The inability to replicate the results by \citet{lee20} by running the provided source code makes checking whether or not a our numeric implementation faithfully represents their results very difficult.
-Additionally, the script that was said to been used to obtain the results reported by \cite{lee20} appears to use a different measurement model than what was described in the supplemental material, again making it difficult to fully replicate the result of \cite{lee20} without being able to easily run the provided source code.
-In this case, we chose to use measurement model described by Eq.~\ref{model2:measEpi} for both phases of the epidemic, as this seemed to visually match the results of \cite{lee20} most closely.
-
-\subsection{Model~3 Replication}
-
-Model~3 was developed by a team of researchers at the Laboratory of the Swiss Federal Institute of Technology in Lausanne, hereafter referred to as the Model~3 authors.
-The code that was originally used to implement Model~3 is archived with the DOI: \url{10.5281/zenodo.3360723}, and also available in the public GitHub repository: \code{jcblemai/haiti-mass-ocv-campaign}.
-Because the code was made publicly available, and final model parameters were reported in the supplementary material of \cite{lee20}, we were able to reproduce Model~3 by directly using the source code.
-In Fig.~\ref{fig:mod3rep}, we plot simulations from this model.
-This figure can be compared to Figure S18 of \cite{lee20}.
-We note that slight differences may be accounted for due to variance in the model simulations and the difference in programming language used to produce the figure.
-Overall, the high standard of reproducibility that was achieved by the Model~3 authors facilitated the ability to readily replicate their model and results.
-
-<<Model3Replication, echo=FALSE, cache=TRUE>>=
-NSIM <- 500
-
-plot_order <- c(
-  'Artibonite',
-  'Sud_Est',
-  'Nippes',
-  'Nord_Est',
-  'Ouest',
-  'Centre',
-  'Nord',
-  'Sud',
-  'Nord_Ouest',
-  'Grande_Anse'
-)
-
-dep_labeller <- as_labeller(
-  c(
-    'Artibonite' = 'Artibonite',
-    'Sud_Est' = 'Sud-Est',
-    'Sud.Est' = 'Sud-Est',
-    'Nippes' = 'Nippes',
-    'Nord_Est' = 'Nord-Est',
-    'Nord.Est' = 'Nord-Est',
-    'Ouest' = 'Ouest',
-    'Centre' = 'Centre',
-    'Nord' = 'Nord',
-    'Sud' = 'Sud',
-    'Nord_Ouest' = 'Nord-Ouest',
-    'Nord.Ouest' = 'Nord-Ouest',
-    'Grande_Anse' = 'Grand\'Anse',
-    'Grand.Anse' = 'Grand\'Anse'
-  )
-)
-
-lee3 <- lee3_spatPomp()
-lee3_sims_quants <- bake("../model3/si/lee3Sims.rds", {
-
-  # Simulate from lee et al 2020 model
-  sims_temp <- simulate(
-    lee3, nsim = NSIM,
-    seed = 321, format = 'data.frame'
-  )
-
-  # Get desired quantiles of reported cases from simulations
-  quants <- sims_temp %>%
-    dplyr::group_by(unitname, time) %>%
-    dplyr::summarise(
-      q05 = quantile(cases, 0.025, na.rm = T),
-      mean = mean(cases, na.rm = T),
-      q50 = quantile(cases, 0.5, na.rm = T),
-      q95 = quantile(cases, 0.975, na.rm = T)
-    ) %>%
-    dplyr::ungroup() %>%
-    dplyr::mutate(
-      date = lubridate::date_decimal(time),
-      date = as.Date(lubridate::round_date(date))
-    ) %>%
-    dplyr::rename(dep = unitname)
-
-  # Remove cases from Ouest that were not considered by Lee et al. (2020)
-  quants %>%
-    dplyr::filter(
-      !(dep == "Ouest" & date < '2017-06-10')
-    )
-
-})
-
-# Get the subset of cholera cases that match simulations from Lee et al. (2020) model.
-sub_cases <- haitiCholera %>%
-  dplyr::mutate(date = as.Date(date_saturday)) %>%
-  dplyr::select(-report, -date_saturday) %>%
-  dplyr::filter(lubridate::decimal_date(date) >= min(lee3@times)) %>%
-  tidyr::pivot_longer(
-    data = .,
-    cols = -date,
-    names_to = "dep",
-    values_to = "cases"
-  ) %>%
-  dplyr::mutate(
-    dep = dplyr::case_when(
-      dep == "Grand.Anse" ~ "Grande_Anse",
-      dep == "Nord.Est" ~ "Nord_Est",
-      dep == "Nord.Ouest" ~ "Nord_Ouest",
-      dep == "Sud.Est" ~ "Sud_Est",
-      TRUE ~ dep
-    )
-  )
-@
-
-\begin{figure}[ht]
-<<PlotMod3Rep, echo=FALSE, message=FALSE, fig.height=3.75>>=
-ggplot() +
-  # Region for upper and lower simulation bounds
-  geom_ribbon(
-    data = lee3_sims_quants,
-    aes(x = date, ymin = q05, ymax = q95),
-    fill = 'darkblue', alpha = 0.5
-  ) +
-  # Median line
-  geom_line(
-    data = lee3_sims_quants,
-    aes(x = date, y = q50),
-    col = 'darkblue'
-  ) +
-  # Observed data
-  geom_point(
-    data = dplyr::filter(sub_cases, cases < 500),
-    aes(x = date, y = cases), size = 0.2
-  ) +
-  facet_wrap(
-    ~factor(dep, levels = plot_order),
-    nrow = 2, labeller = dep_labeller
-  ) +
-  labs(y = 'Reported Cholera Cases') +
-  scale_x_date(date_breaks = "6 months", date_labels = "%Y-%m") +
-  theme(
-    axis.text.x = element_text(angle = 40, hjust = 1, vjust = 1, size = 7),
-    axis.title.x = element_blank()
-  )
-@
-\caption{\label{fig:mod3rep}
-Simulations from Model~3.
-Compare to Figure S18 of \cite{lee20}.
-}
-\end{figure}
-
-<<lee3Cleanup, echo=FALSE, include=FALSE, message=FALSE>>=
-rm(lee3, lee3_sims_quants, sub_cases)
-gc()
-@
-
-\section{Forecasting with parameter uncertainty}
-
-Let $f_{Y_{1:N}}(y_{1:N} | \theta)$ denote the pdf of the model under consideration, were $\theta$ is a parameter vector that indexes the model.
-Furthermore, denote the observed data as $y_{1:N}^*$.
-Because the uncertainty in just a single parameter can lead to drastically different forecasts \citep{saltelli20},
-parameter uncertainty should be considered when obtaining model forecasts when the goal is to influencing policy.
-In a Bayesian modelling paradigm, the most natural way to account for parameter uncertainty in model forecasts is to suppose that $\theta$ comes from a distribution $f_{\Theta}$, and then to obtain $J$ forecasts from the model where each forecast is obtained using parameters drawn from the posterior distribution $\Theta_{1:J} \mid Y_{1:N} = y_{1:N}^* \sim f_{\Theta}\big(\theta | Y_{1:N} = y_{1:N}^*\big)$.
-
-When frequentist methods are used, however, there does not exist a posterior distribution from which one could sample.
-A common approach could be to obtain a weighted average of the simulations from various models \citep{hoeting99}, but this can be problematic when forecasts from each model are very different from each other \citep{grueber11}.
-A similar approach that has been taken \citep{king15} is to obtain model forecasts using multiple sets of parameter values and then sample from the resulting forecasts using weights proportional to the corresponding likelihoods of the parameter values.
-This approach could be considered as empirical Bayes, as it is equivalent to using a discrete uniform prior where the set of values in the prior distribution is determined by a stochastic routine applied to the observed data, as discussed below.
-
-For each $i \in 1:k$, let $\Theta_i$ be a random vector of model parameters.
-Then, letting $\Theta$ denote the true model parameters, we endow the set $\{\Theta_1, \Theta_2, \ldots, \Theta_k\}$ with a discrete uniform distribution, such that $P\big(\Theta = \Theta_i\big) = \frac{1}{K}$ for all values $i \in \seq{1}{k}$.
-Using this as a prior distribution, the posterior distribution of $\Theta | Y_{1:N} = y_{1:N}^*$ can be expressed as:
-$P\big(\Theta = \Theta_k | Y_{1:N} = y_{1:N}^*\big) = \frac{f_{Y_{1:N}}(y_{1:N}^*| \Theta_k)}{\sum_{l = 1}^K f_{Y_{1:N}}(y_{1:N}^*| \theta_l)}$.
-In a standard empirical Bayes analysis, the values $\Theta_1, \ldots, \Theta_k$ of the prior distribution would be chosen using the observed data, resulting in a posterior distribution that weighs the prior parameter vectors proportional to their corresponding likelihoods.
-Instead, we choose $\Theta_i$ to be the output of a stochastic routine applied to the observed data by setting $\Theta_i$ to be the output of an iterated filtering algorithm.
-In practice, because the likelihood maximization routines of iterated filtering methods are stochastic, it is common to run the iterated filtering method multiple times for each model in order to obtain a maximum likelihood estimate for model parameters.
-This results in a natural set of parameters near the MLE that could be used as the discrete prior distribution.
-
-\section{Translating to \citet{lee20} notation}
-
-Since the models of \citet{lee20} were developed independently, the choice of notation varies inconsistently between models.
-For our reanalysis, we rename parameters to provide a unified notation facilitating comparison between models.
-Table~\ref{tab:translate} maps this notation back to the original notations, for reference.
-
-\begin{table}
-  \begin{center}
-  \begin{tabular}{|c|c|c|c|c|}\hline
-    \multirow{2}{*}{Parameter} & Our & \multicolumn{3}{c|}{Lee et al. (2020a)} \\\cline{3-5}
-     & Notation & 1 & 2 & 3 \\
-    \hline
-    \hline
-    % Table begins here
-    Reporting Rate & $\reportRate$ & $\rho$ & $\rho$ & $\epsilon_1, \epsilon_2$ \\\hline
-    Mixing Coefficient & $\mixExponent$ & $\nu$ & --- & --- \\\hline
-    Measurement Over-Dispersion & $\obsOverdispersion$ & $\tau$ & --- & $p$ \\\hline
-    Birth Rate & $\muBirth$ & $\mu$ & --- & --- \\\hline
-    Natural Mortality Rate & $\muDeath$ & $\delta$ & --- & $\mu$ \\\hline
-    Cholera Mortality Rate & $\choleraDeath$ & --- & --- & $\alpha$ \\\hline
-    Latent Period & $1/\muEI$ & $1/\sigma$ & $1/\gamma_E$ & --- \\\hline
-    Recovery Rate & $\muIR$ & $\gamma$ & $\gamma$ & $\gamma$ \\\hline
-    Loss of Immunity & $\muRS$ & $\alpha$ & $\sigma$ & $\rho$ \\\hline
-    Symptomatic Ratio & $\symptomFrac$ & $1 - \theta_0$ & $k$ & $\sigma$ \\\hline
-    Asymptomatic Relative Infectiousness & $\asymptomRelativeInfect$ & $\kappa$ & $red_\beta$ & --- \\\hline
-    Human-to-Water Shedding & $\Wshed$ & --- & $\mu$ & $\theta_I$ \\\hline
-    Asymptomatic Relative Shedding & $\asymptomRelativeShed$ & --- & $red_\mu$ & $\theta_A/\theta_I$ \\\hline
-    Seasonal Amplitude & $\seasAmplitude$ & --- & $\alpha_s$ & $\lambda$ \\\hline
-    Transmission & $\transmission$ & $\beta$ & $\beta$ & $c$ \\\hline
-    Water-to-Human & $\Wbeta{}$ & --- & $\beta_W$ & $\beta$ \\\hline
-    Bacteria Mortality & $\Wremoval$ & --- & $\delta$ & $\mu_\beta$ \\\hline
-    Vaccination Efficacy & $\vaccineEfficacy$ & $\theta_{vk}$ & $\theta_1, \theta_2, \theta_{1_5}, \theta_{2_5}$ & $\eta_{1d}, \eta_{2d}$ \\\hline
-    Process Over-dispersion & $\sigmaProc$ & --- & --- & $\sigma_w$\\\hline
-    % Half Saturation Constant & $\Wsat$ & --- & $Sat$ & $K$ \\\hline
-  \end{tabular}
-  \end{center}
-  \caption{
-  \label{tab:translate}Translations between our common notation and notation used by \cite{lee20}
-  }
-\end{table}
-
-\clearpage
-
-\bibliography{../bib-haiti}
-
-\end{document}

### 2023-09-01 si/inputs/paramUncertainty.tex
Started work to make individual si sections seperate

--- 2023-08-31 si/si.Rnw
@@
-\documentclass[11pt]{article}
-
-\pdfminorversion=4
-
-\newcommand\secTitleSpace{\hspace{3mm}}
-
-<<packages, include=FALSE>>=
-library(tidyverse)
-library(pomp)
-library(haitipkg)
-library(spatPomp)
-library(doParallel)
-library(doRNG)
-
-cores <-  as.numeric(Sys.getenv('SLURM_NTASKS_PER_NODE', unset = NA))
-if(is.na(cores)) cores <- detectCores()
-# cores <- 20
-registerDoParallel(cores)
-theme_set(theme_bw())
-
-if (!dir.exists("../model3/si")) {
-  dir.create("../model3/si", recursive = TRUE)
-}
-@
-
-<<set-opts, include=FALSE>>=
-options(
-  scipen = 2,
-  help_type = "html",
-  stringsAsFactors = FALSE,
-  continue = "+  ",
-  width = 70,
-  useFancyQuotes = FALSE,
-  reindent.spaces = 2,
-  xtable.comment = FALSE
-)
-@
-
-<<knitr-opts, include=FALSE, purl=FALSE>>=
-library(knitr)
-opts_knit$set(concordance=TRUE)
-opts_chunk$set(
-  progress=TRUE,prompt=TRUE,highlight=FALSE,
-  tidy=TRUE,
-  tidy.opts=list(
-    keep.blank.line=FALSE
-  ),
-  comment="",
-  warning=FALSE,
-  message=FALSE,
-  error=TRUE,
-  echo=FALSE,
-  cache=FALSE,
-  strip.white=TRUE,
-  results="markup",
-  fig.path="figure/",
-  fig.lp="fig:",
-  fig.align="center",
-  fig.show="asis",
-  dev="pdf",
-  dev.args=list(
-    bg="transparent",
-    pointsize=9
-  )
-)
-
-myround <- function (x, digits = 1) {
-  # taken from the broman package
-  if (digits < 1)
-    stop("This is intended for the case digits >= 1.")
-  if (length(digits) > 1) {
-    digits <- digits[1]
-    warning("Using only digits[1]")
-  }
-  tmp <- sprintf(paste("%.", digits, "f", sep = ""), x)
-  zero <- paste0("0.", paste(rep("0", digits), collapse = ""))
-  tmp[tmp == paste0("-", zero)] <- zero
-  tmp
-}
-
-@
-
-%% copied from pnas-new.cls
-\usepackage{amsmath,amsfonts,amssymb,graphicx}
-
-%%%% parameters %%%%%%%%%%5
-\newcommand\Wsat{W_{\mathrm{sat}}}
-\newcommand\muIR{\mu_{IR}}
-\newcommand\muEI{\mu_{EI}}
-\newcommand\transmission{\beta}
-\newcommand\seasAmplitude{a}
-\newcommand\rainfallExponent{r}
-\newcommand\muRS{\mu_{RS}}
-\newcommand\vaccineEfficacy{\theta}
-\newcommand\muBirth{\mu_S}
-\newcommand\muDeath{\delta}
-\newcommand\choleraDeath{\delta_{C}}
-\newcommand\symptomFrac{f}
-\newcommand\asymptomRelativeInfect{\epsilon}
-\newcommand\asymptomRelativeShed{\epsilon_{W}}
-\newcommand\Wbeta[1]{\beta_{W#1}}
-\newcommand\Wremoval{\delta_W}
-\newcommand\Wshed{\mu_W}
-\newcommand\mixExponent{\nu}
-\newcommand\sigmaProc{\sigma_{\mathrm{proc}}}
-\newcommand\reportRate{\rho}
-\newcommand\obsOverdispersion{\psi}
-\newcommand\phaseParm{\phi}
-\newcommand\transmissionTrend{\zeta}
-\newcommand\Binit{\xi}
-\newcommand\vaccClass{Z}
-\newcommand\vaccCounter{z}
-\newcommand\modelCounter{m}
-\newcommand\Whur[1]{\beta_{W#1}^{hm}}
-\newcommand\hHur[1]{h_{#1}^{hm}}
-\newcommand\Iinit{I_{0,0}}
-\newcommand\Einit{E_{0, 0}}
-\newcommand\reportChange{\delta}
-
-\newcommand\missing{\hspace{6mm} ---}
-\newcommand\fixed{\color{blue}}
-\newcommand\demography{\bullet}
-
-\newcommand\mytitle{Informing policy via dynamic models: Cholera in Haiti}
-\newcommand\comma{{\hspace{-0.25mm},\hspace{-0.2mm}}}
-
-%%%%%%%%%%%%%%%% START OF USER DEFINITIONS %%%%%%%%%%%%%%%5
-
-\newcommand\siOnly[1]{} %% redefined for the supplement
-\newcommand\msOnly[1]{#1} %% redefined for the supplement
-
-\newcommand\code[1]{\texttt{#1}}
-\usepackage{xurl}
-% \usepackage[breaklinks]{hyperref}
-\usepackage{makecell}
-\usepackage{multirow}
-\usepackage{multicol}
-\usepackage[ruled,noline,linesnumbered]{algorithm2e}
-% \usepackage{setspace}
-
-%% EDITING MACROS
-\usepackage{color}
-\usepackage[normalem]{ulem}% to use \sout in feedback commands
-\usepackage{bm}
-% orange for EI
-\definecolor{orange}{rgb}{1,0.5,0}
-\newcommand\ei[2]{\sout{#1} \textcolor{orange}{#2}}
-\newcommand\eic[1]{\textcolor{orange}{[#1]}}
-% green for JW
-\definecolor{green}{rgb}{0,0.5,0}
-\newcommand\jw[2]{\sout{#1} \textcolor{green}{#2}}
-\newcommand\jwc[1]{\textcolor{green}{[#1]}}
-% purple for JJ
-\definecolor{purple}{rgb}{0.5,0,1}
-\newcommand\jj[2]{\sout{#1} \textcolor{purple}{#2}}
-\newcommand\jjc[1]{\textcolor{purple}{[#1]}}
-% cyan for AR
-\definecolor{cyan}{rgb}{0,.5,.5}
-\newcommand\ar[2]{\sout{#1} \textcolor{cyan}{#2}}
-\newcommand\arc[1]{\textcolor{cyan}{#1}}
-% light brown for KT
-\definecolor{lightbrown}{rgb}{0.5,0.5,0}
-\newcommand\kt[2]{\sout{#1} \textcolor{lightbrown}{#2}}
-\newcommand\ktc[1]{\textcolor{lightbrown}{#1}}
-
-%% basic POMP definitions
-\newcommand\Xspace{{\mathbb X}}
-\newcommand\Yspace{{\mathbb Y}}
-\newcommand\Thetaspace{\R^{\Thetadim}}
-\newcommand\Xunitspace{{\scriptsize{\mathcal X}}}
-\newcommand\Yunitspace{{\tiny{\mathcal Y}}}
-\newcommand\hatTheta{\widehat{\Theta}}
-\newcommand\Xdim{{\mathrm{dim}}(\Xspace)}
-\newcommand\Ydim{{\mathrm{dim}}(\Yspace)}
-\newcommand\Thetadim{P}
-\newcommand\thetadim{p}
-\newcommand\timeSet{{\mathbb T}}
-\newcommand\data[1]{#1^*}
-
-\newcommand\unitSpecific{\hspace{0.1mm}\mathrm{us}}
-\newcommand\shared{\hspace{0.15mm}\mathrm{sh}}
-
-% for algorithm pseudocode
-\SetKwFor{For}{for}{:}{end}
-
-%% for particle filters
-\newcommand\unit{u}
-\newcommand\altUnit{\tilde u}
-\newcommand\Unit{U}
-\newcommand\rep{i}
-\newcommand\Rep{\mathcal{I}}
-\newcommand\island{\rep}
-\newcommand\Island{\Rep}
-%\newcommand\txtisland{replicate}
-\newcommand\altIsland{\tilde i}
-\usepackage[mathscr]{euscript}
-
-\renewcommand\time{n}
-\newcommand\myvec[1]{\boldsymbol{#1}}
-\newcommand\altTime{\tilde n}
-\newcommand\Time{N}
-\newcommand\Np{J}
-\newcommand\np{j}
-\newcommand\altNp{k}
-\newcommand\altAltNp{\tilde j}
-\newcommand\resampleIndex{r}
-\newcommand\unitWeight{w}
-
-%% for all iterated filtering algorithms
-\newcommand\Nit{M}
-\newcommand\nit{m}
-
-%% customized math macros
-\newcommand\prob{\mathbb{P}}
-\newcommand\E{\mathbb{E}}
-\newcommand\dd[1]{\mathrm{d}{#1}}
-\newcommand\given{{\,\vert\,}}
-\newcommand\equals{{{\,}={\,}}}
-\newcommand\myequals{\hspace{0.5mm}{=}\hspace{0.5mm}}
-\newcommand\myto{{\;:\;}}
-\newcommand\seq[2]{{#1}\!:\!{#2}}
-\newcommand\mydot{{\,\cdot\,}}
-\newcommand\cp[2]{N_{\mathrm{#1}\mathrm{#2}}}
-\newcommand\giventh{{\hspace{0.5mm};\hspace{0.5mm}}}
-\newcommand\normal{\mathcal{N}}
-\newcommand\argequals{{\,=\,}}
-\newcommand\lags{c}
-\newcommand\maxlag{\overline{c}}
-\newcommand\nlfList{C}
-\newcommand\bigO{\mathcal{O}}
-\newcommand\loglik{\lambda}
-\newcommand\loglikMC{\MC{\loglik}}
-\newcommand\R{\mathbb{R}}
-\newcommand\param{\,;}
-\newcommand\mycolon{{\hspace{0.6mm}:\hspace{0.6mm}}}
-\newcommand\MC[1]{#1^{\,\mbox{\tiny MC}}}
-\newcommand\EMC{{\E}}
-\newcommand\Var{\mathrm{Var}}
-\newcommand\var{\Var}
-\newcommand\Cov{\mathrm{Cov}}
-\newcommand\cov{\Cov}
-\newcommand\iid{\mathrm{iid}}
-\newcommand\dist{d}
-\newcommand\transpose{\top}
-
-\newcommand\gnsep{,}
-
-\def\lik{L}
-\def\loglik{\ell}
-
-%% THEOREM-LIKE ENVIRONMENTS
-\usepackage{amsthm}
-\newtheorem{prop}{Proposition}
-\newtheorem{theorem}{Theorem}
-\newtheorem{lemma}{Lemma}
-\newtheorem{example}{Example}
-\newtheorem{remark}{Remark}
-\newtheorem{cor}{Corrolary}
-\newtheorem{fact}{Fact}
-\newtheorem{assumption}{Assumption}
-\newtheorem{assumptionB}{Assumption}
-\renewcommand\theassumption{A\arabic{assumption}}
-\renewcommand\theassumptionB{B\arabic{assumptionB}}
-
-%% FOR PSEUDOCODE SETUP
-\newcommand\mystretch{\rule[-2mm]{0mm}{5mm} }
-\newcommand\asp{\hspace{4mm}}
-\newcommand\codeIndent{\hspace{4mm}}
-
-%%%%%%%%%%%%%%%% END OF USER DEFINITIONSS %%%%%%%%%%%%%%%
-
-\bibliographystyle{apalike}
-\usepackage{natbib}
-\usepackage{fullpage}
-
-\newcommand\myeqref[1]{(\ref{#1})}
-
-\renewcommand{\contentsname}{Supplementary Content}
-\renewcommand{\refname}{Supplementary References}
-\renewcommand\thefigure{S-\arabic{figure}}
-\renewcommand\thetable{S-\arabic{table}}
-\renewcommand\thesection{S\arabic{section}}
-\renewcommand\theequation{S\arabic{equation}}
-\renewcommand\theprop{S\arabic{prop}}
-\renewcommand\thelemma{S\arabic{lemma}}
-\renewcommand\thealgocf{S\arabic{algocf}}
-
-\setcounter{tocdepth}{1}
-
-\begin{document}
-
-\date{\today}
-\title{Supplement to ``{\mytitle}''}
-  \author{Jesse Wheeler, AnnaElaine L. Rosengart, Zhuoxun Jiang, \\ Kevin Tan, Noah Treutle and Edward L. Ionides
- \\ \hspace{.2cm}\\
-    Department of Statistics, University of Michigan\\
-}
-
-\newcommand{\blind}{1}
-
-\if1\blind
-{
-\maketitle
-}
-\fi
-
-\if0\blind
-{
-  \bigskip
-  \bigskip
-  \bigskip
-  \begin{center}
-    {\LARGE\bf \mytitle}
-\end{center}
-  \bigskip
-  \bigskip
-}\fi
-
-\tableofcontents
-
-\newpage
-
-\section{Markov chain and differential equation interpretations of compartment flow rates}
-
-In Sections 2.1, 2.2 and 2.3 of the main article, we define compartment models in terms of their flow rates.
-For a discrete population model, these rates define a Markov chain.
-For a continuous and deterministic model, the rates define a system of ordinary differential equations.
-Here, we add additional details to clarify the mapping from a collection of rate functions to a fully specified process.
-Our treatment follows \citet{breto09}.
-
-A general compartment model is a vector-valued process $X(t)=(X_1(t),\dots,X_c(t))$ denoting the (integer or real-valued) counts in each of $c$ compartments, where $t$ is any continuous value in the interval  $\left[t_0, \infty\right)$ for some real valued starting time $t_0$.
-The compartments may also have names, but to set up general notation we simply refer to them by their numerical index.
-The basic characteristic of a compartment model is that $X(t)$ can be written in terms of the flows $N_{ij}(t)$ from $i$ to $j$.
-A flow into compartment $i$ from outside the system is denoted by $N_{\demography i}$, and a flow out of the system from compartment $i$ is denoted by $N_{i\demography}$.
-We call $\demography$ a source/sink compartment, though it is an irregular compartment since $X_{\demography}(t)$ is not defined.
-These flows are required to satisfy a ``conservation of mass'' identity:
-\begin{equation}
-X_i(t)=X_i(t_0)+N_{\demography i}(t) - N_{i\demography}(t) + \sum_{j\neq
-i}N_{ji}(t)-\sum_{j\neq i}N_{ij}(t). \label{eq:conservation}
-\end{equation}
-Each {\em flow} $N_{ij}(t)$ is associated with a {\em rate} function
-$\mu_{ij}=\mu_{ij}(t,X(t))$, where we include the possibility that $i$ or $j$ takes value $\demography$.
-
-There are different ways to use a collection of rate functions to build a fully specified model.
-We proceed to describe the ones we use in this paper: via a system of ordinary differential equations (Sec.~\ref{subsec:ode}), a simple Markov counting system (Sec.~\ref{subsec:smcs}), and an overdispersed Markov counting system (Sec.~\ref{subsec:odmcs}).
-Other representations include stochastic differential equations driven by Gaussian noise or Gamma noise \citep{bhadra11}.
-
-\subsection{Ordinary differential equation (ODE) interpretation}
-\label{subsec:ode}
-
-A basic deterministic specification is
-\begin{equation}
-\label{eq:ode1}
-dN_{ij}/dt = \mu_{ij}\big(t,X(t) \big) X_i(t), \hspace{3mm} i\in \seq{1}{c}, \hspace{3mm} j\in \seq{1}{c} \cup\{\demography\}, \hspace{3mm} i\neq j,
-\end{equation}
-where $\mu_{ij}\big(t,X(t)\big)$ is called a per-capita rate or a unit rate.
-Flows into the system require special treatment since $X_i(t)$ in \myeqref{eq:ode1} is not defined for $i=\demography$.
-Instead, we specify
-\begin{equation}
-\label{eq:ode2}
-dN_{\demography i}/dt = \mu_{\demography i}\big(t,X(t) \big).
-\end{equation}
-This is the the interpretation and implementation used for Model~2 in our study.
-
-\subsection{Simple Markov counting system interpretation}
-\label{subsec:smcs}
-A continuous time Markov chain can be specified via its infinitesimal transition probabilities.
-A basic approach to this is to define
-\begin{eqnarray}
-\label{eq:smcs1}
-\prob\big[ N_{ij}(t+\delta)-N_{ij}(t)=0 \given X(t)\big]
- &=& 1-\delta \mu_{ij}\big(t,X(t)\big)X_i(t) + o(\delta),
-\\
-\label{eq:smcs2}
-\prob\big[ N_{ij}(t+\delta)-N_{ij}(t)=1 \given X(t)\big]
- &=& \delta \mu_{ij}\big(t,X(t)\big)X_i(t) + o(\delta),
-\end{eqnarray}
-for $i\in \seq{1}{c}$ and $j\in\seq{1}{c}\cup\{\demography\}$ with $i\neq j$.
-As with the ODE case, we need special attention for flows into the system, and we define
-\begin{eqnarray}
-\label{eq:smcs3}
-\prob\big[ N_{\demography i}(t+\delta)-N_{\demography i}(t)=0 \given X(t)\big]
- &=& 1-\delta \mu_{\demography i}\big(t,X(t)\big) + o(\delta),
-\\
-\label{eq:smcs4}
-\prob\big[ N_{\demography i}(t+\delta)-N_{\demography i}(t)=1 \given X(t)\big]
- &=& \delta \mu_{\demography i}\big(t,X(t)\big) + o(\delta).
-\end{eqnarray}
-Together with the initial conditions $X(0)$, equations \myeqref{eq:smcs1}--\myeqref{eq:smcs4} define a Markov chain.
-Each flow is a simple counting process, meaning a non-decreasing integer-valued process that only has jumps of size one.
-We therefore call the Markov chain a simple Markov counting system (SMCS).
-The infinitesimal mean of every flow is equal to its infinitesimal variance \citep{breto11} and so an SMCS is called equidispersed.
-We note that the special case of Model~1 used by \cite{lee20} (with $\sigmaProc = 0$) is an SMCS.
-To permit more general mean-variance relationships for a Markov counting system, we must permit jumps of size greater than one.
-The utility of overdispersed models, where the infinitesimal variance of the flow exceeds the infinitesimal mean, has become widely recognized \citep{stocks20,he10}.
-
-\subsection{Overdispersed Markov counting system interpretation}
-\label{subsec:odmcs}
-
-Including white noise in the rate function enables the possibility of an overdispersed Markov counting system \citep{breto11,breto09,he10}.
-Since rates should be non-negative, Gaussian noise is not appropriate and gamma noise is a convenient option that has found various applications \citep{romero-severson15, subramanian20}.
-Specifically, we consider a model given by
-\begin{equation}
-\label{eq:odmcs1}
-\mu_{ij}\big(t,X(t)\big) = \bar\mu_{ij}\big(t,X(t)\big) \, d\Gamma_{ij}(t)/dt,
-\end{equation}
-where $\Gamma_{ij}(t)$ is a stochastic process having independent gamma distributed increments, with
-\begin{equation}
-\label{eq:odmcs2}
-\E\big[\Gamma_{ij}(t)\big] = t, \quad \var\big[\Gamma_{ij}(t)\big] = \sigma_{ij}^2 t.
-\end{equation}
-Formally interpreting the meaning of \myeqref{eq:odmcs1} is not trivial, and we do so by constructing a Markov process $X(t)$ as the limit of the Euler scheme descrived in Section~\ref{sec:numerics}, below.
-Therefore, the numerical scheme in Sec.~\ref{sec:numerics} can be taken as a definition of the meaning of \myeqref{eq:odmcs1}.
-The Markov chain defined by the limit of this Euler scheme as the step size decreases is an overdispersed Markov counting system, with the possibility of instantaneous jumps of size greater than one \citep{breto11}.
-
-\section{Numerical solutions to compartment models}
-\label{sec:numerics}
-
-Models may be fitted and their implications assessed via numerical solutions (i.e., simulations) from the model equations.
-All the analyses we consider have this simulation-based property, known as plug-and-play or equation-free or likelihood-free.
-The numerical solutions to the model are arguably of more direct scientific interest than the exact solutions to the postulated equations.
-For ODE models, numerical methods are well studied and a standard numerical solution package such as \code{deSolve} in \code{R} is adequate for many purposes.
-For SMCS and ODMCS models, exact schemes are feasible when the number of events is small, which may be the case for small populations.
-However, for applicability to larger populations, we use instead the following Euler scheme.
-Write $\delta$ for an Euler time step, and $\Delta N_{ij}$ for the numerical approximation to $N_{ij}(t+\delta)-N_{ij}(t)$ given $X(t)$.
-For each $i$ and $j$ in $\seq{1}{c} \cup \{\demography\}$ with $i \neq j$, we draw independent Gamma distributed noise increments with mean $\delta$ and variance $\sigma_{ij}^2 \delta$, denoted using a mean-variance parameterization of the gamma distribution as
-\begin{equation}
-\label{eq:numerics1}
-\Delta\Gamma_{ij} \sim \mathrm{gamma}(\delta, \sigma_{ij}^2 \delta).
-\end{equation}
-In the case of an SMCS model, $\sigma_{ij}=0$ for all $i$ and $j$, so we have $\Delta\Gamma_{ij}=\delta$.
-Then, for $i\neq \demography$ and $j\neq i$, and writing
-\begin{equation}
-\label{eq:numerics2}
-\mu_{ij}=\bar\mu_{ij}\big(t,X(t)\big) \Delta\Gamma_{ij} / \delta,
-\end{equation}
-we calculate transition probabilities
-\begin{eqnarray}
-\label{eq:numerics3}
-p_{ij} &=& \exp\left\{-\sum_{k\in 1:c \, \cup \{\demography\}} \mu_{ik} \, \delta \right\}
-\frac{\mu_{ij}}{\sum_{k\in 1:c\cup \{\demography\}} \mu_{ik}},
-\\
-\label{eq:numerics4}
-p_{ii} &=& 1 - \sum_{j\neq i} p_{ij}.
-\end{eqnarray}
-These probabilities correspond to competing hazards for every individual in compartment $i$ to transition to some compartment $j$, interpreting $j=i$ to mean that the individual remains in $i$.
-Then, $\big(\Delta N_{i1},\dots,\Delta N_{ic},\Delta N_{i\demography}\big)$ has the multinomial distribution where $X_i(t)$ individuals are allocated independently to $\seq{1}{c}\cup\{\demography\}$ with probabilities given by \eqref{eq:numerics3} and \eqref{eq:numerics4}.
-We use the \code{reulermultinom} function in the \code{pomp} package to draw from this multionomial distribution.
-
-Different treatments of demographic flows---such as birth, death, immigration and emigration---are possible.
-For the case $i=\demography$, the treatment used by Model~1 is to set
-\begin{equation}
-\label{eq:numerics6}
-\Delta N_{\demography j} \sim \mathrm{poisson}( \mu_{\demography j} \delta),
-\end{equation}
-an independent Poisson random variable with mean $\mu_{\demography j} \delta$.
-
-Models~2 and 3 used an alternative approach, balancing the total number of flows in and out of the compartment, i.e., $\sum_{i}N_{\demography i}(t) = \sum_{i}N_{i \demography}(t)$, in order to make the model consistent with the known total population.
-In this case, we formally model the death rate as a rate of returning to the susceptible class $S$, and use external transitions from $\demography$ into $S$ to describe only net population increase.
-
-\section{Measurement Models}
-
-Each POMP requires specification of a measurement model, which is a statistical description of how observations on the system are obtained.
-In general, we used the same measurement models that were reported by \cite{lee20} unless specifically noted in the following subsections.
-
-\subsection{Model~1}
-
-In this model, the advantage afforded by vaccination is an increased probability that an infection is asymptomatic.
-Therefore, under the assumptions of this model, all reported cases are assumed to be a fraction of individuals that transition from the exposed to the infected compartment, as noted in Eq.~\myeqref{model1:meas}:
-\begin{equation}
-  \label{model1:meas}
-  Y_{n} \mid \Delta N_{E_{\cdot}I_{\cdot}}(n) = \reportChange \sim \text{NB}\left(\reportRate \reportChange, \obsOverdispersion \right),
-\end{equation}
-where $Y_n$ is the reported cholera cases at time $t_n \in t_1:t_N$ and $\Delta N_{E_{\cdot}I_{\cdot}}(n)$ is the sum total of individuals across each vaccination compartment $\vaccCounter \in 1:\vaccClass$ who moved from compartment $E_\vaccCounter$ to $I_\vaccCounter$ since observation $t_{n-1}$.
-Here, $\text{NB}\left(\reportRate \reportChange, \obsOverdispersion \right)$ denotes a negative binomial distribution with mean $\reportRate \reportChange$ and variance $\reportRate \reportChange \big(1 + \frac{\reportRate \reportChange}{\obsOverdispersion}\big)$.
-
-\subsection{Model~2}\label{sec:mod2Meas}
-
-Model~2 was fit using reported case counts that were transformed using the natural logarithm.
-We fit Model~2 using the subplex algorithm in the \code{subplex} package, using
-a Gaussian measurement model (Eq.~\myeqref{model2:meas}) on the log transformed cases within each unit.
-The final loss function that is maximized is the product of the likelihoods of the individual units, or the sum of the log-likelihood of the individual units.
-The measurement model for individual units is given in Eq.~\myeqref{model2:meas}.
-\begin{equation}
-  \label{model2:meas}
-  \log\big(Y_{u, n} + 1\big) \mid \Delta N_{E_{u\cdot}I_{u\cdot}}(n) = \reportChange_u \sim \text{N}\left(\log \big(\reportRate \reportChange_u + 1\big), \obsOverdispersion^2 \right),
-\end{equation}
-where $\Delta N_{E_{u\cdot}I_{u\cdot}}(n)$ is the sum total of individuals across vaccination compartment $\vaccCounter \in 0:4$ within unit $u$ who moved from compartment $E_{u\vaccCounter}$ to $I_{u\vaccCounter}$ since observation $t_{n-1}$.
-Therefore, because the natural logarithm of observed case counts (plus one, to avoid taking the logarithm of zero) has a normal distribution, $Y_{u, n} + 1$ is assumed to follow a log-normal distribution with log-mean parameter $\log\big(\reportRate \Delta N_{E_{u\cdot}I_{u\cdot}}(n) + 1\big)$ and log-variance $\obsOverdispersion^2$.
-We note that fitting a model with this measurement model is equivalent to fitting using least squares, with $\log(Y_{u, n} + 1)$ as the response variable.
-
-This measurement model differs from that used by \citet{lee20}, who fit the model in two stages: epidemic and endemic phases.
-Although their text and supplement material do not explicitly describe the measurement model used, inspection of the code provided with their submission suggests a change in measurement model between the epidemic and endemic phases.
-% In the file \code{choleraEqsPublic.py}, \citet{lee20} create several functions, where each function represents a set of coupled differential equations that could potentially be used to model cholera incidence data.
-% Each function returns a vector (or \code{numpy} array) that represents the change in each state variable for a single time step, including the variable \code{dC}, which tracks the number of new infections and is used to obtain the reported case counts.
-% Following their comments and code, it appears that the function \code{choleraEqs10WithoutVaccinationNetwork} was used to describe the dynamics of the epidemic phase, and \code{choleraEqs11WithoutVaccinationNetwork} was used for the endemic stage (see Sec~\ref{sec:mod2rep} for more details).
-% Because their models were fit using least squares, the code in these functions suggest that the measurement model for the epidemic phase is
-The measurement model they used for the epidemic phase is
-\begin{equation}
-  \label{model2:measEpi}
-  Y_{u, n} \mid \Delta N_{E_{u\cdot}I_{u\cdot}}(n) = \reportChange_u \sim \text{N}\left(\reportRate \reportChange_u, \obsOverdispersion^2 \right),
-\end{equation}
-% which is similar to our measurement model, the primary difference being that the measurement model is applied to raw case counts rather than log-transformed case counts.
-The measurement model they used for the endemic phase modifies the epidemic model by counting both asymptomatically infected (A) and symptomatically infected (I) individuals in the case counts:
-\begin{equation}
-  \label{model2:measEnd}
-  Y_{u, n} \mid \Delta N_{E_{u\cdot}I_{u\cdot}}(n) = \reportChange_{u1}, \Delta N_{E_{u\cdot}A_{u\cdot}}(n) = \reportChange_{u2} \sim \text{N}\left(\reportRate \big(\reportChange_{u1} + \reportChange_{u2}\big), \obsOverdispersion^2 \right),
-\end{equation}
-where the notation for $\Delta N_{E_{u\cdot}A_{u\cdot}}(n)$ is similar to $\Delta N_{E_{u\cdot}I_{u\cdot}}(n)$, described above.
-
-\subsection{Model~3}
-
-In this model, reported cholera cases are assumed to stem from individuals who develop symptoms and seek healthcare.
-Therefore reported cases are assumed to come from an over-dispersed negative binomial model, given the increase in infected individuals:
-\begin{equation}
-  \label{model3:meas}
-  Y_{u, n} \mid \Delta N_{S_{u \cdot}I_{u}}(t) = \reportChange_u \sim \text{NB}\left(\reportRate \reportChange_{u}, \obsOverdispersion \right),
-\end{equation}
-where $\Delta N_{S_{u \cdot}I_{u \vaccCounter}}(n)$ is the number of individuals who moved from compartment $S_{u \vaccCounter}$ to $I_{u}$ since observation $t_{n-1}$.
-
-This measurement model is a minor change from that used by \cite{lee20}, which allowed for a change in the reporting rate on January 1st, 2018.
-The fitted values of the reporting rate---before and after January 2018---were $0.97$ and $0.097$, respectively.
-An instantaneous change from near perfect to almost non-existent reporting can be problematic, as it forces the model to explain the observed reduction in reported cases as a decrease in the reporting of cases, rather than a decrease in the prevalence of cholera.
-This shift was justified by a ``change of the case definition that occurred on January 1st, 2018";
-this claim was not cited, and we could find no evidence that such a drastic change in the reporting rate would be warranted.
-We therefore do not allow a change in reporting rate when fitting Model~3.
-
-\section{Initial Values}
-
-To perform inference on POMP models, it is necessary to propose an initial probability density for the latent process $f_{X_0}(x_0;\theta)$, including the possibility that the initial values of the latent states are known, or are a non-random function of the unknown parameter vector, $\theta$.
-This density is used to obtain initial values of the latent state when fitting and evaluating the model.
-For each of the models considered in this analysis, the initial conditions are derived by enforcing the model dynamics on reported cholera cases.
-It is also sometimes necessary to fit some initial value parameters in order to help determine initial values for weakly identifiable compartments.
-In the following subsections, we mention initial value parameters that were fit for each model.
-
-\subsection{Model~1}
-
-For this model, the number of individuals in the Recovered and Asymptomatic compartments are set to zero, but the initial proportion of Infected and Exposed individuals is estimated as initial value parameters ($\Iinit$ and $\Einit$, respectively) using the IF2 algorithm, implemented as \texttt{mif2} in the \texttt{pomp} package.
-Finally, the initial proportion of Susceptible individuals $S_{0, 0}$ is calculated as $S_{0, 0} = 1 - \Iinit - \Einit$.
-This model for the initial values of the latent states matches that which was used by \citet{lee20}.
-
-\subsection{Model~2}
-
-Model~2 assumes that the initial values are a deterministic function of the reporting rate and the initial case reports, and so no initial value parameters need to be estimated.
-Initial values for latent state compartments are chosen so as to enforce the model dynamics on the observed number of cases.
-Specifically, the model sets $I_{u0}(0) = y^*_{1u} / \reportRate$ for each unit $u \in 1:10$, where $I_{u0}(t)$ is the number of infected individuals in unit $u$ at time $t$, vaccination scenario $z = 0$, $y^*_{tu}$ is the reported number of cases, and $\reportRate$ is the reporting rate.
-This model for the initial values of the latent states matches that which was used by \citet{lee20}.
-
-\subsection{Model~3}
-
-The latent states of this model are initialized by enforcing the model dynamics on the incidence data from the start of the recorded cases until time $t_0$, requiring that some of the available data be used to determine the initial values of the latent states.
-This is the same approach that was taken by \citet{lee20}, who used the value $t_0 = \text{2014-02-22}$; this choice of $t_0$ results in modeling roughly only $60\%$ of the available data, some of which is later discarded for alternative reasons \citep{lee20sup}.
-
-We do not see any immediate reason that this model could not be extended to cover a larger range of the data, and chose the value $t_0 = \text{2010-11-13}$.
-This choice of $t_0$ corresponds to using approximately $1\%$ of the available data to determine initial values of the latent states.
-In addition to modeling a larger portion of the available data, this choice of $t_0$ corresponds to an important real-world event, as daily reports from each of the departments were not being sent to the health ministry until November 10, 2010; % TODO: Cite
-this choice of $t_0$ therefore makes $\bm{Y}\big(t_1\big)$ the first week of data once daily reports are being sent to the health ministry.
-The few observation times that exist before $t_0$ are used to initialize the model by enforcing model dynamics on these preliminary observations.
-For convenience, we denote these observations as $t_{-3}, t_{-2}$ and $t_{-1}$; as before, we let $y_{u,-k} = Y_{u}\big(t_{-k}\big)$ denote the observed case count for unit $u$ at time point $t_{-k}$, where $k \in 1:3$.
-Equations for the initial values of non-zero latent states are provided in Eqs.~\myeqref{eq:model3:I0}--\myeqref{eq:model3:B0}.
-
-\begin{eqnarray}
-\label{eq:model3:I0}
-I_{u}\big(t_0\big) &=& \frac{y^*_{u, -1}}{7\reportRate\left(\muIR + \left(\muDeath + \choleraDeath\right)/365\right)},
-\\%[-3pt]
-\label{eq:model3:A0}
-A_{u}\big(t_0\big) &=& \frac{I_{u0}\big(t_0\big)\left(1 - \symptomFrac\right)}{\symptomFrac},
-\\
-\label{eq:model3:R0}
-R_{u01}\big(t_0\big) &=& R_{u02}\big(t_0\big) = R_{u03}\big(t_0\big) = \left(\frac{\sum_{k = -3}^0 y^*_{u, k}}{\reportRate \symptomFrac} - \left(I_{u0}\big(t_0\big) + A_{u0}\big(t_0\big)\right)\right) / 3
-\\
-\label{eq:model3:S0}
-S_{u0}\big(t_0\big) &=& \mathrm{Pop}_u - I_{u}\big(t_0\big) - A_{u}\big(t_0\big) - \sum_{k=1}^3 R_{u0k}\big(t_0\big)
-\\
-\label{eq:model3:B0}
-B_{u}\big(t_0\big) &=& \big[1 + \seasAmplitude \tilde{J}^r \big] \mathrm{Den}_u \, \Wshed \big[ I_{u}(t_0)+ \asymptomRelativeShed A_{u_0}(t) \big]/\Wshed.
-\end{eqnarray}
-
-In Eq.~\myeqref{eq:model3:B0}, $\tilde{J} = 0.002376$ is the median adjusted rainfall over the observation period.
-One important consideration to make with this parameter initialization model is when $y_{u, -1}^* = 0$, which occurs for units $u \in \{3, 4\}$, which correspond to the Grand'Anse and Nippes departments, respectively.
-When this is the case, each of the infectious $I_u(t_0)$, asymptomatic $A_u(t_0)$, and bacterial reservoir $W_u(t_0)$ compartments have a value of zero.
-Because Model~3 models cholera transmission primarily by means of the bacterial reservoir, this makes it nearly impossible for an outbreak to occur.
-Therefore for units $u \in \{3, 4\}$, we introduce initial value parameters $I_{0,0}^3$ and $I_{0,0}^4$, and calibrate these parameter values using the data.
-The resulting parameter estimates are used to obtain the remaining non-zero initial values of the latent states using Eqs.~\myeqref{eq:model3:A0}--\myeqref{eq:model3:B0}.
-
-% Argue why we didn't need to do this for Model 2.
-
-% TODO: Check this section.
-
-\section{Calibrating Model~3 to observed cases}\label{sec:mod3Cal}
-
-<<CalibrateOrig3>>=
-# Create vectors for the unit and shared parameters
-unit_specific_names <- c("betaB", "foi_add")
-shared_param_names <- c(
-  "mu_B", "XthetaA", "thetaI", "lambdaR", "r", "std_W",
-  "epsilon", "k"
-)
-est_param_names <- c(
-  unit_specific_names, shared_param_names
-)
-
-# Add unit numbers to each parameter
-est_param_names_expanded <- paste0(rep(est_param_names, each = 10), 1:10)
-
-# Simple function that is used to set the rw_sd for the first search
-# This function is convenient so that we don't have to type out the rw_sd
-# for each unit parameter.
-set_rw_1 <- function(x) {
-  if (gsub("[[:digit:]]+$", "", x) %in% shared_param_names) {
-    return(0.01)
-  # } else if (x %in% paste0(rep(c("aHur", 'hHur'), each = 2), c(3, 9))) {
-  #   return(expression(ifelse(time >= 2016.754 & time <= 2017, 0.015, 0)))
-  # } else if (x %in% c("Iinit3", "Iinit4")) {
-  #   return(expression(ivp(0.25)))
-  } else if (grepl('^foi_add[[:digit:]]+$', x)) {
-    return(0.015)
-  } else if (grepl('^betaB[[:digit:]]+$', x)) {
-    return(0.01)
-  } else {
-    return(0)
-  }
-}
-
-set_rw_2 <- function(x) {
-  if (gsub("[[:digit:]]+$", "", x) %in% shared_param_names) {
-    return(0.003)
-  # } else if (x %in% paste0(rep(c("aHur", 'hHur'), each = 2), c(3, 9))) {
-  #   return(expression(ifelse(time >= 2016.754 & time <= 2017, 0.005, 0)))
-  # } else if (x %in% c("Iinit3", "Iinit4")) {
-  #   return(expression(ivp(0.125)))
-  } else if (grepl('^foi_add[[:digit:]]+$', x)) {
-    return(0.005)
-  } else if (grepl('^betaB[[:digit:]]+$', x)) {
-    return(0.003)
-  } else {
-    return(0)
-  }
-}
-
-# First Search RW
-reg_rw_1.sd <- lapply(est_param_names_expanded, set_rw_1)
-names(reg_rw_1.sd) <- est_param_names_expanded
-chol_rw_1.sd <- do.call(rw_sd, reg_rw_1.sd)
-
-# Second Search RW
-reg_rw_2.sd <- lapply(est_param_names_expanded, set_rw_2)
-names(reg_rw_2.sd) <- est_param_names_expanded
-chol_rw_2.sd <- do.call(rw_sd, reg_rw_2.sd)
-
-SEARCH1 = list(
-  NBPF = 100,
-  NP = 2500,
-  SPAT_REGRESSION = 0.05,
-  NREPS = 144,
-  NP_EVAL = 2500,
-  NREPS_EVAL = 18,
-  RW_SD = chol_rw_1.sd,
-  COOLING = 0.5,
-  KEEP_TRACES = FALSE,
-  KEEP_LIKE_MAT = FALSE
-)
-
-SEARCH2 = list(
-  TOP_N = 12,
-  NBPF = 100,
-  NP = 2500,
-  SPAT_REGRESSION = 0.05,
-  NREPS = 6,
-  NP_EVAL = 2500,
-  NREPS_EVAL = 18,
-  RW_SD = chol_rw_2.sd,
-  COOLING = 0.5,
-  KEEP_TRACES = FALSE,
-  KEEP_LIK_MAT = TRUE
-)
-
-haiti3_sub_fit <- bake(
-  file = "../model3/si/subset_noparams.rds", {
-    fit_haiti3(
-      nsearches = 2L,
-      search1 = SEARCH1,
-      search2 = SEARCH2,
-      search_rho = FALSE,
-      search_gamma = FALSE,
-      search_Iinit = FALSE,
-      search_hur = FALSE,
-      ncores = cores
-    )
-  }
-)
-
-# Find model index with best log-likelihood
-best_m <- which.max(haiti3_sub_fit$search2$logLiks$logLik)
-
-# Find parameters using best_m index
-sub_params <- haiti3_sub_fit$search2$params[best_m, ]
-
-# Get unit log-likelihoods for the best model.
-all_liks <- as.data.frame(haiti3_sub_fit$search2$likMat)
-# all_liks$which <- rep(1:(SEARCH3$TOP_N * SEARCH3$NREPS), each = SEARCH3$NREPS_EVAL)
-all_liks$which <- rep(1:(12 * 6), each = 18)
-best_mod <- filter(all_liks, which == best_m)
-
-# Pivot longer to get estimate of log-likelihood for each unit.
-unitLikesM3 <- best_mod %>%
-  pivot_longer(
-    cols = -which,
-    names_to = "unit",
-    names_prefix = "V",
-    values_to = "logLik"
-  ) %>%
-  select(-which) %>%
-  group_by(unit) %>%
-  summarize(
-    ll_est = logmeanexp(logLik),
-    ll_se = logmeanexp(logLik, se = TRUE)[2]
-  ) %>%
-  ungroup() %>%
-  mutate(unit = as.numeric(unit)) %>%
-  arrange(unit)
-@
-
-<<M3_bench, echo=FALSE, message=FALSE, include=FALSE, cache=TRUE>>=
-h3 <- haiti3_spatPomp()
-deps <- h3@unit_names
-
-mod3_cases <- as.data.frame(t(h3@data))
-colnames(mod3_cases) <- deps
-m3_inits <- c(7362, 1914, 22, 8, 2818, 82, 2499, 6953, 44, 7) # TODO: Also fit by hand, but could be changed to initializing to the observation before.
-names(m3_inits) <- deps
-
-bench_cond_lls <- list()
-for (dep in deps) {
-
-  # Get department cases
-  cases <- as.numeric(mod3_cases[, dep])
-  mod <- ar1_NegBinom(data = cases, init = m3_inits[dep])
-
-  bench_cond_lls[[dep]] <- haitipkg:::.ar1_NegBinom_cond_ll(
-    theta = mod$theta, init = m3_inits[dep], data = cases, transform = FALSE
-  )
-}
-
-bench_cond_lls <- as.data.frame(bench_cond_lls)
-
-bench_lls <- apply(bench_cond_lls, 2, sum)
-
-unitLikes_bench <- data.frame(
-  unit = 1:10,
-  # dep = names(bench_lls),
-  ll = bench_lls
-) %>%
-  as_tibble()
-
-bench_cond_lls$date <- lubridate::round_date(lubridate::date_decimal(h3@times), unit = 'day')
-@
-
-<<combine_bench_M3, echo=FALSE>>=
-
-# Combine tables with model 3 and benchmark likelihoods.
-bench_M3_joined <- dplyr::left_join(
-  x = unitLikesM3,
-  y = unitLikes_bench,
-  by = "unit"
-) %>%
-  select(-ll_se) %>%
-  mutate(unit = as.character(unit)) %>%
-  pivot_longer(
-    cols = -unit,
-    names_to = "type",
-    values_to = "ll"
-  ) %>%
-  mutate(unit = paste0(unit, ": ", deps[as.numeric(unit)]))
-@
-
-<<M3bpf, echo=FALSE, message=FALSE, include=FALSE>>=
-
-# Change model parameters to the best set of parameters from previous
-# search
-coef(h3)[names(sub_params)] <- sub_params
-
-registerDoRNG(40765101)  # for modular, reproducible results
-
-# Perform block partical filter with fitted model, to get conditional
-# log-likelihoods.
-bpf_sub3_out <- bake(
-  file = "../model3/si/bpf_sub_cond_LL.rds", {
-
-    results <- list()
-
-    bpf_out <- h3 %>% bpfilter(
-      Np = 5000,
-      block_size = 1,
-      save_states = TRUE
-    )
-
-    results$block.cond.loglik <- bpf_out@block.cond.loglik
-    results$times <- bpf_out@times  # Saved for convenience
-    results
-  },
-  seed = 432235
-)
-@
-
-In this section, we provide more detail on the process that was used to estimate the coefficients used in Model~3.
-In particular, we discuss why we decided to include additional model parameters---those that are associated with the behavior of the system during Hurricane Matthew---that were not considered by \cite{lee20}.
-To calibrate this model, we used the iterated block particle filter (IBPF) method of \cite{ionides22}.
-Due to the novelty of this algorithm, there does not exist published examples the IBPF algorithm used for data analysis outside of the papers in which the algorithm was introduced \citep{ning21ibpf,ionides22}, which at least partially motivates the inclusion additional details in this section.
-
-\cite{lee20} were only able to estimate model parameters to a simplified version of Model~3 on a subset of the available data, as no method existed at the time of their publication to fit a fully coupled meta-population model to disease incidence data.
-Building on their results, we fit the fully coupled version of Model~3 to nearly all available data, reserving only a few observations to use to calibrate the initial conditions of the model (see the previous section for more details).
-In order to properly maximize the model likelihood, we found that it was necessary to use multiple searches for the MLE, periodically pruning away less successful searches.
-The first collection of searches was performed by obtaining initial values for the parameters by uniformly sampling values from a predefined hypercube.
-A subsequent refinement search used starting values corresponding to the largest model likelihoods obtained in the previous round.
-This is similar to the approach used by \cite{ionides22}.
-
-We use the technique described above to fit the fully coupled version of Model~3 proposed by \cite{lee20}.
-The maximum likelihood we obtained with the model was $\Sexpr{round(max(haiti3_sub_fit$search2$logLiks$logLik))}$, which is higher than the benchmark model ($\Sexpr{round(sum(unitLikes_bench$ll))}$).
-While beating a simple associative benchmark is promising, this does not immediately rule out the possibility that the model is a poor description of the system.
-Additional investigation of parameters estimates and their corresponding implications on model based conclusions should always be conducted.
-For meta-population models, it is worth considering how well the model fits the data to each spatial unit.
-This can be done by looking at conditional log-likelihoods, which is part of the output of the \texttt{bpfilter} algorithm in the \texttt{spatPomp} package.
-The likelihoods for each department, compared to the corresponding benchmark model, are displayed in Fig.~\ref{fig:h3UnitLikes}.
-The figure demonstrates that while the log-likelihood of the fitted model outperforms the auto-regressive negative binomial benchmark model, Model~3 has lower likelihoods than the benchmark for some departments.
-
-\begin{figure}[ht]
-<<h3UnitLikes, fig.height=3, fig.width=5.5, echo=FALSE>>=
-ggplot(bench_M3_joined, aes(x = ll, y = reorder(unit, ll))) +
-  geom_point(aes(col = type)) +
-  ylab("Unit") +
-  xlab("Log Likelihood") +
-  scale_color_manual(
-    labels = c("ll" = "Benchmark", "ll_est" = "Model 3"),
-    values = c("ll" = "#1b9e77", "ll_est" = "#d95f02")
-  ) +
-  theme(legend.title = element_blank(),
-        axis.title.y = element_blank())
-@
-\caption{\label{fig:h3UnitLikes}Log-likelihoods of Model~3 for each department compared to the corresponding benchmark model prior to the inclusion of parameters for initial values and Hurricane Matthew.}
-\end{figure}
-
-In addition to considering the conditional log-likelihoods of each unit, one can consider conditional log-likelihoods of each observation.
-When compared to a benchmark, this level of detail can provide useful information about which observations are well described by the model and which are poorly described.
-In Fig.~\ref{fig:condLL}, we plot the conditional log-likelihoods of each observation.
-This figure reveals that the fitted model has difficult time explaining certain features of the data.
-For example, many departments (in particular Sud) have observations with lower conditional log-likelihoods near October 2016 than at other time points.
-Further investigation of the model output reveals that the model is struggling to explain the sudden surge in cholera cases that occurred at this time, which coincides with the time that Hurricane Matthew struck Haiti.
-While the model does include a mechanism to account for increased cholera transmission due to large rainfall events, the mechanism does not appear to be sufficient to capture the damaging effects of the hurricane, which had the greatest impact in the the Sud and Grand'Anse departments \citep{ferreirai16}.
-This result led us to include parameters $\Whur{u}$ and $\hHur{u}$ in Eq.~23 of the main text, which allows for an increase in the transmission rate between environmental cholera and humans for Sud and Grand'Anse during and after the hurricane.
-The effect of the hurricane on cholera transmission is assumed to have an exponential decay, where the magnitude is determined by $\Whur{u}$ and the duration of the effect determined by $\hHur{u}$.
-
-\begin{figure}[ht]
-<<CondLLFig, echo=FALSE, message=FALSE, fig.height=3>>=
-
-# Changing labels to human-readable labels.
-deps[grepl("_", deps)] <- gsub("_", "-", deps[grepl("_", deps)])
-deps[deps == "Grande-Anse"] <- "Grand\'Anse"
-names(deps) <- as.character(1:10)
-myLabeller <- as_labeller(deps)
-
-# Plot conditional log-likelihoods for each unit
-t(bpf_sub3_out$block.cond.loglik) %>%
-  as.data.frame() %>%
-  mutate(t = bpf_sub3_out$times) %>%
-  pivot_longer(
-    cols = -t,
-    values_to = 'likelihood',
-    names_prefix = "V",
-    names_to = 'Unit'
-  ) %>%
-  mutate(date = as.Date(lubridate::date_decimal(t))) %>%
-  ggplot(aes(x = date, y = likelihood)) +
-  geom_line() +
-  geom_point() +
-  facet_wrap(~Unit, labeller = myLabeller, scales = 'free_y',
-             nrow = 2) +
-  theme_bw() +
-  theme(axis.title.x = element_blank(),
-        axis.text.x = element_text(angle = -45, hjust = 0)) +
-  ylab("Conditional log-likelihoods")
-@
-\caption{\label{fig:condLL}Conditional log-likelihoods of Model~3 prior to the inclusion of parameters for initial values and Hurricane Matthew.}
-\end{figure}
-
-Other points of concern include the observations at or near the start of the recorded cases, for example, in Nord-Ouest and Sud-Est.
-These low conditional log-likelihoods are a result of having zero or near zero confirmed cases in the first week of observed data, followed by a rapid growth of cases.
-The model cannot explain the rapid growth of cholera in the absence of starting cases because the initialization scheme forces departments with no recorded cases in the first week to have hidden states $I_{u0}(0) = A_{u0}(0) = B_{u}(0) = 0$.
-In reality, however, it is possible to have some infected or asymptomatic individuals in a department when no cases are recorded due to under-reporting.
-This may be particularly true at the onset of an infectious disease outbreak, as lower public awareness and inadequate testing infrastructure may result in many cases going undetected.
-We therefore include and estimate $I_{0u}$ that determines the initial number of Infected individuals at the start of the pandemic for the departments Grande'Anse, Nord-Ouest, and Sud-Est;
-each of these departments have both low starting case counts followed by rapid growth, and low initial values of the conditinoal log-likelihood.
-
-<<loadFinalModel3, echo=FALSE, include=FALSE, message=FALSE>>=
-
-# This result was used in the manuscript, so it doesn't need to be recalculated.
-h3_final_results <- readRDS("../model3/run_level_3/haiti3_fit.rds")
-
-# Save best_m as index for the model parameters with the best likelihood.
-# Then, save the model parameter values and the likelihood evaluations
-best_m <- which.max(h3_final_results$search2$logLiks$logLik)
-best_params <- h3_final_results$search2$params[best_m, ]
-all_liks <- as.data.frame(h3_final_results$search2$likMat)
-all_liks$which <- rep(1:72, each = 18)
-best_mod <- dplyr::filter(all_liks, which == best_m)
-
-# Get repeated estimates of unit log-likelihoods
-unitLikesM3 <- best_mod %>%
-  pivot_longer(
-    cols = -which,
-    names_to = "unit",
-    names_prefix = "V",
-    values_to = "logLik"
-  ) %>%
-  select(-which) %>%
-  group_by(unit) %>%
-  summarize(
-    ll_est = pomp::logmeanexp(logLik),
-    ll_se = pomp::logmeanexp(logLik, se = TRUE)[2]
-  ) %>%
-  ungroup() %>%
-  mutate(unit = as.numeric(unit)) %>%
-  arrange(unit)
-
-# Combine model likelihood table to benchmarks
-bench_M3_joined <- dplyr::left_join(
-  x = unitLikesM3,
-  y = unitLikes_bench,
-  by = "unit"
-) %>%
-  select(-ll_se) %>%
-  mutate(unit = as.character(unit)) %>%
-  pivot_longer(
-    cols = -unit,
-    names_to = "type",
-    values_to = "ll"
-  ) %>%
-  mutate(unit = paste0(unit, ": ", deps[as.numeric(unit)]))
-@
-
-We refit Model~3 after introducing these seven parameters---three initial value parameters and four parameters related to Hurricane Matthew---using the same successive search scheme described above.
-The resulting model has a log-likelihood value of $\Sexpr{round(max(h3_final_results$search2$logLiks$logLik), 1)}$.
-We also provide figures for department specific log-likelihoods and conditional log-likelihoods in Figs.~\ref{fig:finalUnitLL} and \ref{fig:finalCondLL}.
-
-\begin{figure}[ht]
-<<finalUnitLLFig, fig.height=3, fig.width=5.5, echo=FALSE>>=
-ggplot(bench_M3_joined, aes(x = ll, y = reorder(unit, ll))) +
-  geom_point(aes(col = type)) +
-  ylab("Unit") +
-  xlab("Log Likelihood") +
-  scale_color_manual(
-    labels = c("ll" = "Benchmark", "ll_est" = "Model 3"),
-    values = c("ll" = "#1b9e77", "ll_est" = "#d95f02")
-  ) +
-  theme(legend.title = element_blank(),
-        axis.title.y = element_blank())
-@
-\caption{\label{fig:finalUnitLL}Log-likelihoods of Model~3 for each department compared to the corresponding benchmark model after adding and estimating parameters related to initial values and Hurricane Matthew.}
-\end{figure}
-
-<<finalModel3BPF, echo=FALSE, message=FALSE, include=FALSE>>=
-
-# Set the model parameters to the best results from previous search
-coef(h3)[names(best_params)] <- best_params
-
-# block particle filter to get conditional log-likelihoods.
-bpf_final3_out <- bake(
-  file = "../model3/si/h3_final_bpf_cond_ll.rds", {
-
-    results <- list()
-
-    bpf_out <- h3 %>% bpfilter(
-      Np = 5000,
-      block_size = 1,
-      save_states = TRUE
-    )
-
-    results$Sus_df <- foreach(i=1:length(h3@times), .combine = bind_rows) %do% {
-      get_h3_S_quants(bpf_out, i)
-    }
-
-    results$block.cond.loglik <- bpf_out@block.cond.loglik
-    results$times <- bpf_out@times  # saved for convenience
-    results
-  },
-  seed = 786321
-)
-@
-
-\begin{figure}[ht]
-<<finalCondLLFig, echo=FALSE, message=FALSE, fig.height=3>>=
-t(bpf_final3_out$block.cond.loglik) %>%
-  as.data.frame() %>%
-  mutate(t = bpf_final3_out$times) %>%
-  pivot_longer(
-    cols = -t,
-    values_to = 'likelihood',
-    names_prefix = "V",
-    names_to = 'Unit'
-  ) %>%
-  mutate(date = as.Date(lubridate::date_decimal(t))) %>%
-  ggplot(aes(x = date, y = likelihood)) +
-  geom_line() +
-  geom_point() +
-  facet_wrap(~Unit, labeller = myLabeller, scales = 'free_y',
-             nrow = 2) +
-  theme_bw() +
-  theme(axis.title.x = element_blank(),
-        axis.text.x = element_text(angle = -45, hjust = 0)) +
-  ylab("Conditional log-likelihoods")
-@
-\caption{\label{fig:finalCondLL}Conditional log-likelihoods of Model~3 after adding and estimating the parameters for initial values and Hurricane Matthew.}
-\end{figure}
-
-While the addition of the Hurricane parameters seems to have resulted in a dramatic increase of conditional likelihoods around and after October 2016, the inclusion of the initial value parameters does not appear to have fixed the low conditional log-likelihood values in Sud-Est and Nord-Ouest.
-This could be due to many factors, including the possibility that the coupling between units forces low initial value parameters in a given department in order for the model to fit the data in the remaining departments.
-Whatever the case may be, the inclusion of the seven parameters resulted in a $\Sexpr{round(max(h3_final_results$search2$logLiks$logLik) - max(haiti3_sub_fit$search2$logLiks$logLik), 1)}$ increase in log-likelihood, which is a highly statistically significant difference.
-One could similarly test whether the inclusion of the initial value parameters resulted in a significantly better fit to the observed data, but we do not do this here due to the large computational effort that would be required.
-
-\subsection{Examining the Hidden States of the Calibrated Model}
-
-For mechanistic models, beating a suitable statistical benchmark does not alone guarantee that the model provides an accurate description of a dynamic process.
-Indeed, a good statistical fit does not require the model to assert a causal explanation.
-For example, reconstructed latent variables should make sense in the context of alternative measurements of these variables \citep{grad12}.
-We demonstrate this principle by examining the latent state of the calibrated model.
-In particular, we examine the compartment of susceptible individuals under various scenarios, as this compartment drives transmission dynamics in SIR models for infectious disease outbreaks.
-
-Recall that the filtering distribution for the calibrated version of Model~3 at time $t_k$ is defined as the distribution of the latent state at time $t_k$ given the data from times $t_{1}:t_{k}$, i.e.
-$f^{(3)}_{\bm{X}_k|\bm{Y}_{1:k}}(\bm{x}_{k} | \bm{y}^*_{1:k} ; \hat\theta)$.
-In general, one may expect simulations from the filtering distribution of a model with a good statistical fit to result in hidden states that are highly consistent with the observed data because the filtering distribution is conditioned on the observed data.
-Fig.~\ref{fig:h3Sus} shows the percentage of individuals that are in the susceptible compartment from various simulations of the model:
-simulations from Model~3 under initial conditions are displayed in red; simulations from the filtering distribution of model are displayed in blue.
-This figure shows that simulations from initial conditions tends to result in a much more rapid depletion of susceptible individuals at the start of the epidemic than simulations from the filtering distribution, suggesting the calibrated model has a propensity to predict larger outbreaks than what is typically seen in the data.
-This also appears true in later stages of the time period of observed data, where the data suggests that there is a replenishment of the susceptible compartment, but the Model simulations typically retain a smaller proportion of susceptible individuals.
-This result demonstrates that the calibrated model favors a more rapid growth in cholera cases than what is typically seen in the observed data, which explains why the model fails to predict the absence of cases between February, 2019 and October, 2022.
-
-<<h3SusFigSetup, echo=FALSE, message=FALSE, include=FALSE>>=
-
-# This bake perfoms the following:
-#  (1) simulates the model from initial conditions
-#  (2) get the values of the hidden states at the end of each simulation, and
-#      saves these for later use.
-#  (3) Obtain quantiles for the Susceptible pool from the simulations.
-init_conditions_results <- bake(
-  file = "../model3/si/InitSimsQuants.rds", {
-
-    results <- list()  # To return final output of bake
-
-    # Simulate from the fitted model
-    h3Spat_sims <- simulate(
-      h3, nsim = 2000, format = 'data.frame'
-    )
-
-    # Get states at the end of the simulations
-    h3_spat_end <- h3Spat_sims %>%
-      filter(time == max(time))
-
-    # Save hidden state values at the end of the simulation. Saved as a matrix
-    # in the same format as the output from a block-particle filter, so that
-    # it can be used as the input to project_from_filter.
-    results[["end_conditions"]] <- h3_spat_end %>%
-      select(-time, -cases) %>%
-      mutate(
-        unit = case_when(
-          unitname == "Artibonite" ~ 1,
-          unitname == "Centre" ~ 2,
-          unitname == "Grande_Anse" ~ 3,
-          unitname == "Nippes" ~ 4,
-          unitname == "Nord" ~ 5,
-          unitname == "Nord_Est" ~ 6,
-          unitname == "Nord_Ouest" ~ 7,
-          unitname == "Ouest" ~ 8,
-          unitname == "Sud" ~ 9,
-          unitname == "Sud_Est" ~ 10
-        )
-      ) %>%
-      select(-unitname) %>%
-      pivot_wider(
-        data = .,
-        names_from = unit,
-        values_from = c(S, I, A, R_one, R_two, R_three, VSd, VR1d, VR2d,
-                        VR3d, VSdd, VR1dd, VR2dd, VR3dd, VSd_alt, VR1d_alt,
-                        VR2d_alt, VR3d_alt, VSdd_alt, VR1dd_alt, VR2dd_alt,
-                        VR3dd_alt, C, B, Doses, totInc),
-        names_sep = ""
-      ) %>%
-      select(-.id) %>%
-      as.matrix() %>%
-      t()
-
-    # Get the quantiles for the Susceptible pool from the simulations
-    results[["SQuants"]] <- h3Spat_sims %>%
-      rename(dep = unitname) %>%
-      group_by(dep, time) %>%
-      summarise(
-        Q025 = quantile(S + VSd + VSdd + VSd_alt + VSdd_alt, probs = 0.025),
-        Q50  = quantile(S + VSd + VSdd + VSd_alt + VSdd_alt, probs = 0.5),
-        Q975 = quantile(S + VSd + VSdd + VSd_alt + VSdd_alt, probs = 0.975)
-      ) %>%
-      ungroup() %>%
-      rowwise() %>%
-      mutate(
-        unit_num = which(h3@unit_names == dep),
-        date = lubridate::date_decimal(time),
-        Q025 = Q025 / h3@params[paste0("H", unit_num)],
-        Q50  = Q50  / h3@params[paste0("H", unit_num)],
-        Q975 = Q975 / h3@params[paste0("H", unit_num)]
-      ) %>%
-      filter(
-        date >= as.Date(
-          lubridate::round_date(lubridate::date_decimal(h3@t0), unit = 'day')
-        )
-      ) %>%
-      mutate(date = as.Date(lubridate::round_date(date, unit = 'day')))
-
-    results
-  },
-  seed = 834687
-)
-
-# Get the output of the "bake" call for plotting.
-h3Spat_quants <- init_conditions_results[["SQuants"]]
-
-# This bake call uses the ending hidden states from above to forecast cholera
-# cases in the future under the V0 vaccination scenario
-mod3_V0_initcond_res <- bake(
-  file = '../model3/si/vacc_sim_from_init.rds', {
-    h3_V0_res <- list()
-
-    # Get "noVacc" scenario parameters
-    h3_S0_par <- get_model3_vacc_scenario_params(scenario = "noVacc")
-    h3Spat <- h3
-    p3 <- best_params
-    coef(h3Spat) <- p3
-    coef(h3Spat)[names(h3_S0_par)] <- h3_S0_par
-
-    # store the end states that will be used as initial values
-    h3_bpf_end_states <- init_conditions_results$end_conditions
-
-    # Using parameters and end state values, obtain projections
-    h3_V0_sims <- project_from_filter(
-      h3Spat, end_states = h3_bpf_end_states, covarGen = project_rain,
-      nsims = 1000, seed = 184697
-    )
-
-    # Store nationally aggregated results
-    h3_V0_res$mod3_V0_sims <- agg_mod3_sims(h3_V0_sims)
-    h3_V0_res$mod3_V0_probs <- get_elimProbs(h3_V0_sims, model = 3)
-
-    # Get quantiles of observed cases for each department and time
-    h3_V0_res$UnitQuants <- h3_V0_sims %>%
-      mutate(
-        date = as.Date(  # Get date object from decimal time
-          lubridate::round_date(lubridate::date_decimal(time), unit = 'day')
-        )
-      ) %>%
-      group_by(unitname, date) %>%
-      summarize(
-        Q025 = quantile(cases, probs = 0.025),
-        Q50 = quantile(cases, probs = 0.5),
-        Q975 = quantile(cases, probs = 0.975)
-      ) %>%
-      ungroup() %>%
-      rename(dep = unitname)
-
-    # Get quantiles of susceptible individuals for each department and time
-    h3_V0_res$SQuants <- h3_V0_sims %>%
-      mutate(
-        date = as.Date(
-          lubridate::round_date(lubridate::date_decimal(time), unit = 'day')
-        )
-      ) %>%
-      group_by(unitname, date) %>%
-      mutate(
-        # Get the population for each unit
-        pop = h3@params[paste0("H", which(first(unitname) == unit_names(h3)))]
-      ) %>%
-      summarize(  # Calculate the quantiles
-        Q025 = quantile(
-          (S + VSd + VSdd + VSdd_alt + VSd_alt) / pop,
-          probs = 0.025
-        ),
-        Q50 = quantile(
-          (S + VSd + VSdd + VSdd_alt + VSd_alt) / pop,
-          probs = 0.5
-        ),
-        Q975 = quantile(
-          (S + VSd + VSdd + VSdd_alt + VSd_alt) / pop,
-          probs = 0.975
-        )
-      ) %>%
-      ungroup() %>%
-      rename(dep = unitname)
-
-    h3_V0_res
-  }
-)
-
-# This result was already used in the manuscript; it doesn't need to
-# be recalculated.
-mod3_V0_res <- readRDS("../model3/run_level_3/mod3_V0_res.rds")
-@
-
-\begin{figure}[ht]
-<<h3SusFig, echo=FALSE, message=FALSE, fig.height=3>>=
-ggplot() +
-  # Sims from initial condition
-  geom_line(data = h3Spat_quants, aes(x = date, y = Q50), col = 'red') +
-  geom_ribbon(
-    data = h3Spat_quants,
-    aes(x = date, ymin = Q025, ymax = Q975),
-    fill = 'red', alpha = 0.5
-  ) +
-  # projections from initial condition, using filtered sims function
-  geom_line(data = mod3_V0_initcond_res$SQuants, aes(x = date, y = Q50), col = 'red') +
-  geom_ribbon(
-    data = mod3_V0_initcond_res$SQuants,
-    aes(x = date, ymin = Q025, ymax = Q975),
-    fill = 'red', alpha = 0.5
-  ) +
-  # Sims from filtered distribution
-  geom_line(
-    data = bpf_final3_out$Sus_df,
-    aes(x = as.Date(lubridate::date_decimal(time)), y = Q50),
-    col = 'blue'
-  ) +
-  geom_ribbon(
-    data = bpf_final3_out$Sus_df,
-    aes(x = as.Date(lubridate::date_decimal(time)), ymin = Q025, ymax = Q975),
-    fill = 'blue', alpha = 0.5
-  ) +
-  # Projections from filtered distribution
-  geom_line(data = mod3_V0_res$SQuants, aes(x = date, y = Q50), col = 'blue') +
-  geom_ribbon(
-    data = mod3_V0_res$SQuants,
-    aes(x = date, ymin = Q025, ymax = Q975),
-    fill = 'blue', alpha = 0.5
-  ) +
-  geom_vline(xintercept = max(h3Spat_quants$date), linetype = 'dashed') +
-  facet_wrap(~dep, scales = 'free_y', nrow = 2) +
-  theme_bw() +
-  theme(axis.title.x = element_blank(),
-        axis.text.x = element_text(angle = -45, hjust = 0)) +
-  ylab("Total Susceptible (%)")
-@
-\caption{\label{fig:h3Sus}Percentage of individuals that are in the susceptible compartment.
-Simulations from Model~3 under initial conditions are displayed in red; simulations from the filtering distribution of model are displayed in blue.
-The dashed line represents the end of the observed data.}
-\end{figure}
-
-\clearpage
-
-\section{Replication of \cite{lee20}}
-
-In this article we claimed that we were able to obtain better fits to the observed data using the same models that were proposed by \cite{lee20}.
-Along with visual comparisons to the data, this claim was supported by comparing likelihoods and AIC values in Table~2 in the manuscript. % TODO: Double check the table number matches
-Because model likelihoods were not provided by \cite{lee20}, it is necessary to replicate these models in order to obtain likelihood estimates.
-Here we would like to thank the authors of \cite{lee20}, who provided detailed descriptions of their models, which enabled us to build on their work.
-In the following subsections, we use our \code{R} package \code{haitipkg} to reproduce some of the results of \cite{lee20}.
-This reproduction allows us to estimate the likelihoods of the \cite{lee20} version of Models~1--3, and also provides a demonstration of the importance and usefulness of reproducible research.
-
-\subsection{Model~1 Replication}
-
-The model was implemented by a team at Johns Hopkins Bloomberg School of Public Health (hereafter referred to as the Model~1 authors) in the \code{R} programming language using the \code{pomp} package \citep{king16}.
-Original source code is publicly available with DOI: 10.5281/zenodo.3360991.
-The final results reported by the Model~1 authors were obtained by using several different parameter sets rather than a single point estimate.
-According to the supplement materials, this was because model realizations from a single parameter set retained substantial variability, but multiple realizations from a collection of parameter sets resulted in a reasonable visual fit to the data.
-We are also inclined to believe that the use of multiple parameter values was in part intended to account for parameter uncertainty---the importance of which was discussed in the main text---an effort by the Model~1 authors that we applaud.
-Simulations from each of the parameter sets, however, were treated with equal importance when being used to diagnose the model fit and make inference on the system.
-This is problematic given Figures S8 and S9 of the supplement material, which suggest that some parameter sets that were used for inference may have been several hundred units of log-likelihood lower than other parameter sets that were simultaneously used to make forecasts.
-Such a large difference in log-likelihoods is well beyond the threshold of statistical uncertainty determined by Wilks' theorem, resulting in the equal use of statistically inferior parameter sets in order to make forecasts and conduct inference on the system.
-
-To fully reproduce the results of the Model~1 authors, it is necessary to use the exact same set of model parameters that were originally used to obtain the results presented by \cite{lee20}.
-Because these parameter sets were not made publicly available, we relied on the source code provided by the Model~1 authors to approximately recreate the parameter set.
-Due to software updates since the publication of the source code, we were unable to produce the exact same set of parameters.
-Running the publicly available source code, however, resulted in a set of parameters that are visually similar to those used by the Model~1 authors (See Figures~\ref{fig:PlotEpiDist} and \ref{fig:plotEndParams}).
-Furthermore, simulations using the set of parameters produced by the source code appear practically equivalent to those displayed by \cite{lee20} (See Figure~\ref{fig:plotMod1Sims}).
-
-<<Simulate Lee Model 1, echo=FALSE, cache=TRUE>>=
-h1_epi <- haiti1()
-N_LEE_SIMS <- 20
-
-registerDoRNG(18599687)
-foreach(
-  i = 1:length(h1LeeStartsEpi),
-  .combine = rbind
-) %dopar% {
-  guess <- h1LeeStartsEpi[i, ]
-  sims <- simulate(h1_epi, nsim = N_LEE_SIMS, format = 'data.frame',
-                   params = guess)  # 20 sims for each set of parameters
-  sims$param_set <- i
-  sims %>%
-    dplyr::select(.id, week, cases, param_set)
-} -> all_sims
-@
-
-<<PlotEpiDist, echo=FALSE, fig.cap="Compare to Figure S8 of Lee et al. (2020) supplement.", fig.width=7, fig.height=7, fig.align='center'>>=
-GGally::ggpairs(
-  h1LeeStartsEpi,
-  columns = c("rho", "tau", "beta1", "nu", "loglik")
-)
-@
-
-<<SimulateEpi, echo=FALSE, include=FALSE, message=FALSE, cache=TRUE>>=
-h1_end <- haiti1(period = "endemic")
-
-true_agg_cases <- haitiCholera %>%
-  select(-report) %>%
-  pivot_longer(
-    data = .,
-    cols = -date_saturday,
-    values_to = 'cases',
-    names_to = 'dep'
-  ) %>%
-  mutate(
-    date = as.Date(date_saturday),
-    dep = gsub("\\.", "_", dep)
-  ) %>%
-  mutate(
-    dep = case_when(dep == "Grand_Anse" ~ "Grande_Anse", TRUE ~ dep)
-  ) %>%
-  tidyr::pivot_wider(
-    data = .,
-    id_cols = c(date),
-    names_from = dep,
-    values_from = cases,
-    names_prefix = 'cases_'
-  ) %>%
-  mutate(
-    ReportedAll = cases_Artibonite + cases_Centre +
-      cases_Grande_Anse + cases_Nippes + cases_Nord +
-      cases_Nord_Est + cases_Ouest + cases_Sud +
-      cases_Sud_Est + cases_Nord_Ouest
-  )
-@
-
-<<FitEnd, echo=FALSE, message=FALSE, include=FALSE>>=
-foreach(
-  i = 1:nrow(h1LeeStartsEnd),
-  .combine = rbind
-) %dopar% {
-  r_params <- h1LeeStartsEnd[i, ][names(coef(h1_end))]
-  coef(h1_end) <- r_params
-
-  sims <- simulate(
-    h1_end, nsim = N_LEE_SIMS,
-    format = 'data.frame',
-    params = r_params
-  )
-  sims$param_set <- i
-  sims %>%
-    dplyr::select(.id, week, cases, param_set)
-} -> end_sims
-
-@
-
-<<plotEndParams, fig.width=7, fig.height=7, fig.align='center', fig.cap="Bivariate relationships between variables after fitting endemic period.
-Compare to S9 of Lee et al.
-(2020) supplement.", cache=TRUE, echo=FALSE>>=
-GGally::ggpairs(
-  h1LeeStartsEnd,
-  columns = c("rho", "tau", "beta1", "nu", "loglik")
-)
-@
-
-<<calcMod1LeeLikes, echo=FALSE, include=FALSE, message=FALSE>>=
-h1_epi_evals <- readRDS("../model1/run_level_3/lee1_epi_evals.rds")
-h1_end_evals <- readRDS("../model1/run_level_3/lee1_end_evals.rds")
-
-mod1_all_likes <- dplyr::inner_join(
-  x = h1_epi_evals,
-  y = h1_end_evals,
-  by = "parid"
-  ) %>%
-  dplyr::mutate(
-    joint_ll = pfLL.x + pfLL.y
-  )
-
-# Approximate the likelihood of the Lee et al (2020) version of Model 1 by
-# using the parameter sets that resulted in the maximum likelihood.
-mod1_lee_ll <- max(mod1_all_likes$joint_ll, na.rm = TRUE)
-@
-
-<<plotMod1Sims, echo=FALSE, fig.cap=paste0("Simulations using parameter sets that were generated by running source code provided by Lee et al (2020).
-Compare to Figure S7 of \\cite{lee20sup}.
-The upper bound for the likelihood of this model is ", round(mod1_lee_ll), "."), cache=TRUE, fig.width=6.5, fig.height=3.75>>=
-
-quants <- all_sims %>%
-  filter(!param_set %in% c(1, 2)) %>%  # Remove parameter sets inconsistent with Lee et al (2020).
-  group_by(week) %>%
-  summarize(
-    q025 = quantile(cases, probs = 0.025, na.rm = TRUE),
-    q50  = quantile(cases, probs = 0.500, na.rm = TRUE),
-    q975 = quantile(cases, probs = 0.975, na.rm = TRUE)
-  ) %>%
-  ungroup() %>%
-  mutate(date = lubridate::ymd("2010-10-16") + lubridate::weeks(week))
-
-quants_end <- end_sims %>%
-  group_by(week) %>%
-  summarize(
-    q025 = quantile(cases, probs = 0.025, na.rm = TRUE),
-    q50  = quantile(cases, probs = 0.500, na.rm = TRUE),
-    q975 = quantile(cases, probs = 0.975, na.rm = TRUE)
-  ) %>%
-  ungroup() %>%
-  mutate(date = lubridate::ymd("2010-10-16") + lubridate::weeks(week))
-
-ggplot() +
-  geom_point(data = true_agg_cases, aes(x = date, y = ReportedAll)) +
-  geom_line(data = quants, aes(x = date, y = q50), col = '#0b8a2b', size = 1) +
-  geom_ribbon(
-    data = quants,
-    aes(x = date, ymin = q025, ymax = q975),
-    alpha = 0.5, fill = '#0b9c30'
-  ) +
-  geom_line(data = quants_end, aes(x = date, y = q50), col = '#882871') +
-  geom_ribbon(data = quants_end, aes(x = date, ymin = q025, ymax = q975), alpha = 0.5, fill = '#882871') +
-  ylab("Reported, Symptomatic Incidence") +
-  scale_x_date(date_breaks = "1 year") +
-  theme(axis.title.x = element_blank(),
-        axis.text.x = element_text(hjust = 1, angle = 45))
-@
-
-Because the model forecasts provided by \cite{lee20} come from various sets of parameters---which each correspond to a unique log-likelihood value---it is not obvious how one would obtain an estimate for the log-likelihood of the model that was used for simulations by the Model~1 authors.
-One approach could be to calculate the logarithm of the weighted mean of the likelihoods for each parameter sets used to obtain the forecasts, where the weights are proportional to the number of times the parameter set was used.
-However, in an effort to not underestimate the likelihood of the model of the Model~1 authors, we report the estimated log-likelihood as the log-likelihood value corresponding to the parameter set with the largest likelihood value, even though the majority of simulations were obtained using parameter sets with lower likelihood values.
-In this sense, we consider the log-likelihood reported in Table~1 of the main text to be an upper-bound of the log-likelihood of the model used by \cite{lee20}.
-For each parameter set, the log-likelihood was estimated using a particle filter, implemented as the \texttt{pfilter} function in the \texttt{pomp} package.
-
-\subsection{Model~2 Replication}\label{sec:mod2rep}
-
-Model~2 was developed by a team that consisted of members from the Fred Hutchinson Cancer Research Center and the University of Florida (hereafter referred to as the Model~2 authors).
-While Model~2 is the only deterministic model we considered in our analysis, it contains perhaps the most complex descriptions of cholera in Haiti: Model~2 accounts for movement between spatial units; human-to-human and environment-to-human cholera infections; and transfer of water between spatial units based on elevation charts and river flows.
-
-The source code that the Model~2 authors used to generate their results was written in the \code{Python} programming language, and is publicly available at \url{10.5281/zenodo.3360857} and its accompanying GitHub repository \url{https://github.com/lulelita/HaitiCholeraMultiModelingProject}.
-In order to perform our analysis in a unified framework, we re-implemented this model in the \code{R} programming language using the \code{spatPomp} package \citep{asfaw23arxiv}, which facilitates the creation of meta-population models.
-We note that the travel and water matrices used to implement the complex dynamics in Model~2 \citep{lee20sup} are not available in either the Zenodo archive or the GitHub repository;
-instead, we obtained these matrices via personal correspondence with the Model~2 authors.
-Using these matrices, and the point estimates for model parameters provided by \citep{lee20sup}, we created trajectories of the cholera dynamics and compared this to available data.
-These trajectories, shown in Figure~\ref{fig:mod2rep}, are very similar to the trajectories shown in Figure~S15 of \cite{lee20sup}.
-
-<<Model2Replication, cache=TRUE, echo=FALSE>>=
-h2_epi <- haiti2()
-h2_end <- haiti2(region = 'after')
-
-h2_epi_traj <- trajectory(h2_epi, format = 'data.frame')
-h2_end_traj <- trajectory(h2_end, format = 'data.frame')
-
-h2_epi_traj$Ctotal <- rowSums(h2_epi_traj[, paste0("C", 1:10)]) * 0.2
-h2_end_traj$Ctotal <- rowSums(h2_end_traj[, paste0("C", 1:10)]) * 0.2
-
-dep_plot_df <- haitipkg::haitiCholera %>%
-  select(-report) %>%
-  tidyr::pivot_longer(
-    data = .,
-    cols = -date_saturday,
-    values_to = 'cases',
-    names_to = 'dep'
-  ) %>%
-  mutate(
-    date = as.Date(date_saturday),
-    dep = gsub("\\.", "_", dep)
-  ) %>%
-  mutate(
-    dep = case_when(dep == "Grand_Anse" ~ "Grande_Anse", TRUE ~ dep)
-  )
-
-true_agg_cases <- dep_plot_df %>%
-  tidyr::pivot_wider(
-    data = .,
-    id_cols = c(date),
-    names_from = dep,
-    values_from = cases,
-    names_prefix = 'cases_'
-  ) %>%
-  mutate(
-    ReportedAll = cases_Artibonite + cases_Centre +
-      cases_Grande_Anse + cases_Nippes + cases_Nord +
-      cases_Nord_Est + cases_Ouest + cases_Sud +
-      cases_Sud_Est + cases_Nord_Ouest
-  ) %>%
-  select(date, ReportedAll)
-
-h2_epi_traj$date <- as.Date(
-  lubridate::round_date(
-    lubridate::date_decimal(h2_epi_traj$year),
-    unit = 'day'
-  )
-)
-
-h2_end_traj$date <- as.Date(
-  lubridate::round_date(
-    lubridate::date_decimal(h2_end_traj$year),
-    unit = 'day'
-  )
-)
-
-all_traj <- bind_rows(
-  select(h2_epi_traj, date, Ctotal),
-  select(h2_end_traj, date, Ctotal)
-)
-
-projections <- left_join(
-  x = true_agg_cases,
-  y = all_traj,
-  by = 'date'
-)
-@
-
-\begin{figure}[ht]
-<<plotModel2Rep, cache=TRUE, echo=FALSE, fig.width=6.5, fig.height=4.5>>=
-gg_epi <- projections %>%
-  filter(date < as.Date("03-01-2014", format = '%m-%d-%Y')) %>%
-  ggplot(aes(x = date)) +
-  geom_line(aes(y = ReportedAll), col = 'black', linetype = 'dashed') +
-  geom_line(aes(y = Ctotal), col = 'red', size = 1) +
-  theme_bw() +
-  scale_y_continuous(limits = c(0, 30000), breaks = seq(0, 30000, 5000)) +
-  theme(axis.title.x = element_blank()) +
-  ylab("Reported number of cases")
-
-gg_end <- projections %>%
-  filter(date >= as.Date("03-01-2014", format = '%m-%d-%Y')) %>%
-  ggplot(aes(x = date)) +
-  geom_line(aes(y = ReportedAll), col = 'black', linetype = 'dashed') +
-  geom_line(aes(y = Ctotal), col = 'red', size = 1) +
-  theme_bw() +
-  scale_y_continuous(limits = c(0, 7000), breaks = seq(0, 7000, 1000)) +
-  theme(axis.title.x = element_blank()) +
-  ylab("Reported number of cases") +
-  scale_x_date(date_breaks = "1 year", date_labels = "%Y")
-
-cowplot::plot_grid(gg_epi, gg_end, ncol = 1)
-@
-\caption{\label{fig:mod2rep}
-Model 2 trajectories using the \code{haitipkg}. Compare to Figure S15 of \cite{lee20sup}.
-}
-\end{figure}
-
-There are minor differences between Figure~\ref{fig:mod2rep} and Figure~S15 of \cite{lee20sup}.
-While the discrepancy appears minor, the deterministic nature of Model~2 implies that an exact replication of model trajectories should be possible.
-In this case, these discrepancies may possibly be attributed to implementing the model and plotting the model trajectory in two different programming languages.
-Another potential explanation for the discrepancy is that the parameters that we used are only approximately the same as those used by \citet{lee20sup}.
-For example, the parameters $\transmission$, $\Wbeta{}$ (See Table~\ref{tab:translate}) had reported values of $9.9 \times 10^{-7}$ and $4.03 \times 10^{-2}$, respectively (Table~S13 of \cite{lee20sup}), but were actually fit to data and therefore likely these values have been rounded.
-Additionally, our implementation of Model~2 used a time scale of years and many of the parameters were reported on a weekly scale, so small differences may result due to unit conversions.
-The collective effect of these small differences in model parameters likely will result in small differences in model trajectories.
-
-Some additional concerns about being able to accurately replicate the results of \cite{lee20} are valid.
-Details about the measurement models and how latent states were initialized for the epidemic model were not provided by \citet{lee20sup} and therefore these details must be inferred by looking at the provided source code.
-According to repository comments, the files \code{fit\allowbreak In\allowbreak Pieces\allowbreak 3params\allowbreak Clean\allowbreak May2019\allowbreak Public.py} and \code{fit\allowbreak In\allowbreak Pieces\allowbreak Mu\allowbreak With\allowbreak Frac\allowbreak Sus\allowbreak Fixed\allowbreak All\allowbreak Infections\allowbreak Public.py} were used to fit the epidemic and endemic phases of the model respectively, although it is apparent that these exact files were not used to obtain the reported results since the files contain some variable-naming errors that make it impossible to run the files without making modifications \footnote{One example of why the code cannot be run that the file loads functions from a non-extant file named \code{choleraEqs.py} in line 13 rather than \code{cholera\allowbreak Eqs\allowbreak Public.py}.}.
-The inability to replicate the results by \citet{lee20} by running the provided source code makes checking whether or not a our numeric implementation faithfully represents their results very difficult.
-Additionally, the script that was said to been used to obtain the results reported by \cite{lee20} appears to use a different measurement model than what was described in the supplemental material, again making it difficult to fully replicate the result of \cite{lee20} without being able to easily run the provided source code.
-In this case, we chose to use measurement model described by Eq.~\ref{model2:measEpi} for both phases of the epidemic, as this seemed to visually match the results of \cite{lee20} most closely.
-
-\subsection{Model~3 Replication}
-
-Model~3 was developed by a team of researchers at the Laboratory of the Swiss Federal Institute of Technology in Lausanne, hereafter referred to as the Model~3 authors.
-The code that was originally used to implement Model~3 is archived with the DOI: \url{10.5281/zenodo.3360723}, and also available in the public GitHub repository: \code{jcblemai/haiti-mass-ocv-campaign}.
-Because the code was made publicly available, and final model parameters were reported in the supplementary material of \cite{lee20}, we were able to reproduce Model~3 by directly using the source code.
-In Fig.~\ref{fig:mod3rep}, we plot simulations from this model.
-This figure can be compared to Figure S18 of \cite{lee20}.
-We note that slight differences may be accounted for due to variance in the model simulations and the difference in programming language used to produce the figure.
-Overall, the high standard of reproducibility that was achieved by the Model~3 authors facilitated the ability to readily replicate their model and results.
-
-<<Model3Replication, echo=FALSE, cache=TRUE>>=
-NSIM <- 500
-
-plot_order <- c(
-  'Artibonite',
-  'Sud_Est',
-  'Nippes',
-  'Nord_Est',
-  'Ouest',
-  'Centre',
-  'Nord',
-  'Sud',
-  'Nord_Ouest',
-  'Grande_Anse'
-)
-
-dep_labeller <- as_labeller(
-  c(
-    'Artibonite' = 'Artibonite',
-    'Sud_Est' = 'Sud-Est',
-    'Sud.Est' = 'Sud-Est',
-    'Nippes' = 'Nippes',
-    'Nord_Est' = 'Nord-Est',
-    'Nord.Est' = 'Nord-Est',
-    'Ouest' = 'Ouest',
-    'Centre' = 'Centre',
-    'Nord' = 'Nord',
-    'Sud' = 'Sud',
-    'Nord_Ouest' = 'Nord-Ouest',
-    'Nord.Ouest' = 'Nord-Ouest',
-    'Grande_Anse' = 'Grand\'Anse',
-    'Grand.Anse' = 'Grand\'Anse'
-  )
-)
-
-lee3 <- lee3_spatPomp()
-lee3_sims_quants <- bake("../model3/si/lee3Sims.rds", {
-
-  # Simulate from lee et al 2020 model
-  sims_temp <- simulate(
-    lee3, nsim = NSIM,
-    seed = 321, format = 'data.frame'
-  )
-
-  # Get desired quantiles of reported cases from simulations
-  quants <- sims_temp %>%
-    dplyr::group_by(unitname, time) %>%
-    dplyr::summarise(
-      q05 = quantile(cases, 0.025, na.rm = T),
-      mean = mean(cases, na.rm = T),
-      q50 = quantile(cases, 0.5, na.rm = T),
-      q95 = quantile(cases, 0.975, na.rm = T)
-    ) %>%
-    dplyr::ungroup() %>%
-    dplyr::mutate(
-      date = lubridate::date_decimal(time),
-      date = as.Date(lubridate::round_date(date))
-    ) %>%
-    dplyr::rename(dep = unitname)
-
-  # Remove cases from Ouest that were not considered by Lee et al. (2020)
-  quants %>%
-    dplyr::filter(
-      !(dep == "Ouest" & date < '2017-06-10')
-    )
-
-})
-
-# Get the subset of cholera cases that match simulations from Lee et al. (2020) model.
-sub_cases <- haitiCholera %>%
-  dplyr::mutate(date = as.Date(date_saturday)) %>%
-  dplyr::select(-report, -date_saturday) %>%
-  dplyr::filter(lubridate::decimal_date(date) >= min(lee3@times)) %>%
-  tidyr::pivot_longer(
-    data = .,
-    cols = -date,
-    names_to = "dep",
-    values_to = "cases"
-  ) %>%
-  dplyr::mutate(
-    dep = dplyr::case_when(
-      dep == "Grand.Anse" ~ "Grande_Anse",
-      dep == "Nord.Est" ~ "Nord_Est",
-      dep == "Nord.Ouest" ~ "Nord_Ouest",
-      dep == "Sud.Est" ~ "Sud_Est",
-      TRUE ~ dep
-    )
-  )
-@
-
-\begin{figure}[ht]
-<<PlotMod3Rep, echo=FALSE, message=FALSE, fig.height=3.75>>=
-ggplot() +
-  # Region for upper and lower simulation bounds
-  geom_ribbon(
-    data = lee3_sims_quants,
-    aes(x = date, ymin = q05, ymax = q95),
-    fill = 'darkblue', alpha = 0.5
-  ) +
-  # Median line
-  geom_line(
-    data = lee3_sims_quants,
-    aes(x = date, y = q50),
-    col = 'darkblue'
-  ) +
-  # Observed data
-  geom_point(
-    data = dplyr::filter(sub_cases, cases < 500),
-    aes(x = date, y = cases), size = 0.2
-  ) +
-  facet_wrap(
-    ~factor(dep, levels = plot_order),
-    nrow = 2, labeller = dep_labeller
-  ) +
-  labs(y = 'Reported Cholera Cases') +
-  scale_x_date(date_breaks = "6 months", date_labels = "%Y-%m") +
-  theme(
-    axis.text.x = element_text(angle = 40, hjust = 1, vjust = 1, size = 7),
-    axis.title.x = element_blank()
-  )
-@
-\caption{\label{fig:mod3rep}
-Simulations from Model~3.
-Compare to Figure S18 of \cite{lee20}.
-}
-\end{figure}
-
-<<lee3Cleanup, echo=FALSE, include=FALSE, message=FALSE>>=
-rm(lee3, lee3_sims_quants, sub_cases)
-gc()
-@
-
 \section{Forecasting with parameter uncertainty}
@@
 This results in a natural set of parameters near the MLE that could be used as the discrete prior distribution.
-
-\section{Translating to \citet{lee20} notation}
-
-Since the models of \citet{lee20} were developed independently, the choice of notation varies inconsistently between models.
-For our reanalysis, we rename parameters to provide a unified notation facilitating comparison between models.
-Table~\ref{tab:translate} maps this notation back to the original notations, for reference.
-
-\begin{table}
-  \begin{center}
-  \begin{tabular}{|c|c|c|c|c|}\hline
-    \multirow{2}{*}{Parameter} & Our & \multicolumn{3}{c|}{Lee et al. (2020a)} \\\cline{3-5}
-     & Notation & 1 & 2 & 3 \\
-    \hline
-    \hline
-    % Table begins here
-    Reporting Rate & $\reportRate$ & $\rho$ & $\rho$ & $\epsilon_1, \epsilon_2$ \\\hline
-    Mixing Coefficient & $\mixExponent$ & $\nu$ & --- & --- \\\hline
-    Measurement Over-Dispersion & $\obsOverdispersion$ & $\tau$ & --- & $p$ \\\hline
-    Birth Rate & $\muBirth$ & $\mu$ & --- & --- \\\hline
-    Natural Mortality Rate & $\muDeath$ & $\delta$ & --- & $\mu$ \\\hline
-    Cholera Mortality Rate & $\choleraDeath$ & --- & --- & $\alpha$ \\\hline
-    Latent Period & $1/\muEI$ & $1/\sigma$ & $1/\gamma_E$ & --- \\\hline
-    Recovery Rate & $\muIR$ & $\gamma$ & $\gamma$ & $\gamma$ \\\hline
-    Loss of Immunity & $\muRS$ & $\alpha$ & $\sigma$ & $\rho$ \\\hline
-    Symptomatic Ratio & $\symptomFrac$ & $1 - \theta_0$ & $k$ & $\sigma$ \\\hline
-    Asymptomatic Relative Infectiousness & $\asymptomRelativeInfect$ & $\kappa$ & $red_\beta$ & --- \\\hline
-    Human-to-Water Shedding & $\Wshed$ & --- & $\mu$ & $\theta_I$ \\\hline
-    Asymptomatic Relative Shedding & $\asymptomRelativeShed$ & --- & $red_\mu$ & $\theta_A/\theta_I$ \\\hline
-    Seasonal Amplitude & $\seasAmplitude$ & --- & $\alpha_s$ & $\lambda$ \\\hline
-    Transmission & $\transmission$ & $\beta$ & $\beta$ & $c$ \\\hline
-    Water-to-Human & $\Wbeta{}$ & --- & $\beta_W$ & $\beta$ \\\hline
-    Bacteria Mortality & $\Wremoval$ & --- & $\delta$ & $\mu_\beta$ \\\hline
-    Vaccination Efficacy & $\vaccineEfficacy$ & $\theta_{vk}$ & $\theta_1, \theta_2, \theta_{1_5}, \theta_{2_5}$ & $\eta_{1d}, \eta_{2d}$ \\\hline
-    Process Over-dispersion & $\sigmaProc$ & --- & --- & $\sigma_w$\\\hline
-    % Half Saturation Constant & $\Wsat$ & --- & $Sat$ & $K$ \\\hline
-  \end{tabular}
-  \end{center}
-  \caption{
-  \label{tab:translate}Translations between our common notation and notation used by \cite{lee20}
-  }
-\end{table}
-
-\clearpage
-
-\bibliography{../bib-haiti}
-
-\end{document}

### 2023-09-05 ms-plos.Rnw
revised author summary

@@
 \section*{Author summary}
-Scientists often use data from an infectious disease outbreak to build statistical models.
-Among other things, these models are often used to guide governmental policy, as the models can provide insight into the dynamics of an outbreak.
-Despite recent computational and methodological advancements, fitting dynamic models to observed incidence data remains a challenging task.
-In this article, we discuss common mistakes that may lead to models that do not adequately describe reality, and make suggestions that can lead to more useful statistical models.
+Quantitative understanding of infectious disease transmission dynamics relies upon mathematical models informed by scientific knowledge and relevant data.
+The models aim to provide a statistical description of the trajectory of an epidemic and its uncertainty, together with a representation of the underlying biological mechanisms.
+Evaluation of success at these goals is necessary in order for a model to provide a reliable tool for guiding evidence-based public policy interventions.
+In this article, we discuss common modeling decisions that may not provide an adequate statistical explanation of the data, and we show that attention to statistical fit can lead to more accurate policy evaluations.
 We do this by conducting a re-analysis of the 2012-2019 cholera outbreak in Haiti.
-We focus on the importance of appropriately diagnosing how well a model describes observed data, correctly interpreting model based results, and stressing the importance of reproducibility.
-An important contribution of this article is demonstrating a use-case of a newly developed computational algorithm to fit spatially coupled meta-population models.
-We assert that raising the currently accepted standards of infectious disease modeling will result in a greater ability of scientists and policy makers to understand and respond to future infectious disease outbreaks.
+Our analysis presents methodology for diagnosing how well a model describes observed data, and provides careful justification of valid conclusions from the fitted model.
+Objective measures are used to benchmark model fit; when these are combined with reproducibility, a framework emerges for continual improvment when revisiting the data and models.
+Our data analysis workflow is supported by recent advances in algorithms, software and hardware, which facilitate statistical fitting of nonlinear stochastic dynamic models to observed incidence data.
+However, inference for high-dimensional systems remains a methodological challenge. 
+One of the models under consideration involves spatially coupled stochastic meta-populations, and we demonstrate how a recently developed algorithm permits likelihood-based inference and model diagnostics in this setting.
+We contend that raising the currently accepted standards of infectious disease modeling will result in a greater ability of scientists and policy makers to understand and respond to future infectious disease outbreaks.
 

### 2023-09-05 si/ReplicateLee20Out.tex
new Makefile for creating supplement material

--- 2023-08-31 si/si.Rnw
@@
-\documentclass[11pt]{article}
-
-\pdfminorversion=4
-
-\newcommand\secTitleSpace{\hspace{3mm}}
-
-<<packages, include=FALSE>>=
-library(tidyverse)
-library(pomp)
-library(haitipkg)
-library(spatPomp)
-library(doParallel)
-library(doRNG)
-
-cores <-  as.numeric(Sys.getenv('SLURM_NTASKS_PER_NODE', unset = NA))
-if(is.na(cores)) cores <- detectCores()
-# cores <- 20
-registerDoParallel(cores)
-theme_set(theme_bw())
-
-if (!dir.exists("../model3/si")) {
-  dir.create("../model3/si", recursive = TRUE)
-}
-@
-
-<<set-opts, include=FALSE>>=
-options(
-  scipen = 2,
-  help_type = "html",
-  stringsAsFactors = FALSE,
-  continue = "+  ",
-  width = 70,
-  useFancyQuotes = FALSE,
-  reindent.spaces = 2,
-  xtable.comment = FALSE
-)
-@
-
-<<knitr-opts, include=FALSE, purl=FALSE>>=
-library(knitr)
-opts_knit$set(concordance=TRUE)
-opts_chunk$set(
-  progress=TRUE,prompt=TRUE,highlight=FALSE,
-  tidy=TRUE,
-  tidy.opts=list(
-    keep.blank.line=FALSE
-  ),
-  comment="",
-  warning=FALSE,
-  message=FALSE,
-  error=TRUE,
-  echo=FALSE,
-  cache=FALSE,
-  strip.white=TRUE,
-  results="markup",
-  fig.path="figure/",
-  fig.lp="fig:",
-  fig.align="center",
-  fig.show="asis",
-  dev="pdf",
-  dev.args=list(
-    bg="transparent",
-    pointsize=9
-  )
-)
-
-myround <- function (x, digits = 1) {
-  # taken from the broman package
-  if (digits < 1)
-    stop("This is intended for the case digits >= 1.")
-  if (length(digits) > 1) {
-    digits <- digits[1]
-    warning("Using only digits[1]")
-  }
-  tmp <- sprintf(paste("%.", digits, "f", sep = ""), x)
-  zero <- paste0("0.", paste(rep("0", digits), collapse = ""))
-  tmp[tmp == paste0("-", zero)] <- zero
-  tmp
-}
-
-@
-
-%% copied from pnas-new.cls
-\usepackage{amsmath,amsfonts,amssymb,graphicx}
-
-%%%% parameters %%%%%%%%%%5
-\newcommand\Wsat{W_{\mathrm{sat}}}
-\newcommand\muIR{\mu_{IR}}
-\newcommand\muEI{\mu_{EI}}
-\newcommand\transmission{\beta}
-\newcommand\seasAmplitude{a}
-\newcommand\rainfallExponent{r}
-\newcommand\muRS{\mu_{RS}}
-\newcommand\vaccineEfficacy{\theta}
-\newcommand\muBirth{\mu_S}
-\newcommand\muDeath{\delta}
-\newcommand\choleraDeath{\delta_{C}}
-\newcommand\symptomFrac{f}
-\newcommand\asymptomRelativeInfect{\epsilon}
-\newcommand\asymptomRelativeShed{\epsilon_{W}}
-\newcommand\Wbeta[1]{\beta_{W#1}}
-\newcommand\Wremoval{\delta_W}
-\newcommand\Wshed{\mu_W}
-\newcommand\mixExponent{\nu}
-\newcommand\sigmaProc{\sigma_{\mathrm{proc}}}
-\newcommand\reportRate{\rho}
-\newcommand\obsOverdispersion{\psi}
-\newcommand\phaseParm{\phi}
-\newcommand\transmissionTrend{\zeta}
-\newcommand\Binit{\xi}
-\newcommand\vaccClass{Z}
-\newcommand\vaccCounter{z}
-\newcommand\modelCounter{m}
-\newcommand\Whur[1]{\beta_{W#1}^{hm}}
-\newcommand\hHur[1]{h_{#1}^{hm}}
-\newcommand\Iinit{I_{0,0}}
-\newcommand\Einit{E_{0, 0}}
-\newcommand\reportChange{\delta}
-
-\newcommand\missing{\hspace{6mm} ---}
-\newcommand\fixed{\color{blue}}
-\newcommand\demography{\bullet}
-
-\newcommand\mytitle{Informing policy via dynamic models: Cholera in Haiti}
-\newcommand\comma{{\hspace{-0.25mm},\hspace{-0.2mm}}}
-
-%%%%%%%%%%%%%%%% START OF USER DEFINITIONS %%%%%%%%%%%%%%%5
-
-\newcommand\siOnly[1]{} %% redefined for the supplement
-\newcommand\msOnly[1]{#1} %% redefined for the supplement
-
-\newcommand\code[1]{\texttt{#1}}
-\usepackage{xurl}
-% \usepackage[breaklinks]{hyperref}
-\usepackage{makecell}
-\usepackage{multirow}
-\usepackage{multicol}
-\usepackage[ruled,noline,linesnumbered]{algorithm2e}
-% \usepackage{setspace}
-
-%% EDITING MACROS
-\usepackage{color}
-\usepackage[normalem]{ulem}% to use \sout in feedback commands
-\usepackage{bm}
-% orange for EI
-\definecolor{orange}{rgb}{1,0.5,0}
-\newcommand\ei[2]{\sout{#1} \textcolor{orange}{#2}}
-\newcommand\eic[1]{\textcolor{orange}{[#1]}}
-% green for JW
-\definecolor{green}{rgb}{0,0.5,0}
-\newcommand\jw[2]{\sout{#1} \textcolor{green}{#2}}
-\newcommand\jwc[1]{\textcolor{green}{[#1]}}
-% purple for JJ
-\definecolor{purple}{rgb}{0.5,0,1}
-\newcommand\jj[2]{\sout{#1} \textcolor{purple}{#2}}
-\newcommand\jjc[1]{\textcolor{purple}{[#1]}}
-% cyan for AR
-\definecolor{cyan}{rgb}{0,.5,.5}
-\newcommand\ar[2]{\sout{#1} \textcolor{cyan}{#2}}
-\newcommand\arc[1]{\textcolor{cyan}{#1}}
-% light brown for KT
-\definecolor{lightbrown}{rgb}{0.5,0.5,0}
-\newcommand\kt[2]{\sout{#1} \textcolor{lightbrown}{#2}}
-\newcommand\ktc[1]{\textcolor{lightbrown}{#1}}
-
-%% basic POMP definitions
-\newcommand\Xspace{{\mathbb X}}
-\newcommand\Yspace{{\mathbb Y}}
-\newcommand\Thetaspace{\R^{\Thetadim}}
-\newcommand\Xunitspace{{\scriptsize{\mathcal X}}}
-\newcommand\Yunitspace{{\tiny{\mathcal Y}}}
-\newcommand\hatTheta{\widehat{\Theta}}
-\newcommand\Xdim{{\mathrm{dim}}(\Xspace)}
-\newcommand\Ydim{{\mathrm{dim}}(\Yspace)}
-\newcommand\Thetadim{P}
-\newcommand\thetadim{p}
-\newcommand\timeSet{{\mathbb T}}
-\newcommand\data[1]{#1^*}
-
-\newcommand\unitSpecific{\hspace{0.1mm}\mathrm{us}}
-\newcommand\shared{\hspace{0.15mm}\mathrm{sh}}
-
-% for algorithm pseudocode
-\SetKwFor{For}{for}{:}{end}
-
-%% for particle filters
-\newcommand\unit{u}
-\newcommand\altUnit{\tilde u}
-\newcommand\Unit{U}
-\newcommand\rep{i}
-\newcommand\Rep{\mathcal{I}}
-\newcommand\island{\rep}
-\newcommand\Island{\Rep}
-%\newcommand\txtisland{replicate}
-\newcommand\altIsland{\tilde i}
-\usepackage[mathscr]{euscript}
-
-\renewcommand\time{n}
-\newcommand\myvec[1]{\boldsymbol{#1}}
-\newcommand\altTime{\tilde n}
-\newcommand\Time{N}
-\newcommand\Np{J}
-\newcommand\np{j}
-\newcommand\altNp{k}
-\newcommand\altAltNp{\tilde j}
-\newcommand\resampleIndex{r}
-\newcommand\unitWeight{w}
-
-%% for all iterated filtering algorithms
-\newcommand\Nit{M}
-\newcommand\nit{m}
-
-%% customized math macros
-\newcommand\prob{\mathbb{P}}
-\newcommand\E{\mathbb{E}}
-\newcommand\dd[1]{\mathrm{d}{#1}}
-\newcommand\given{{\,\vert\,}}
-\newcommand\equals{{{\,}={\,}}}
-\newcommand\myequals{\hspace{0.5mm}{=}\hspace{0.5mm}}
-\newcommand\myto{{\;:\;}}
-\newcommand\seq[2]{{#1}\!:\!{#2}}
-\newcommand\mydot{{\,\cdot\,}}
-\newcommand\cp[2]{N_{\mathrm{#1}\mathrm{#2}}}
-\newcommand\giventh{{\hspace{0.5mm};\hspace{0.5mm}}}
-\newcommand\normal{\mathcal{N}}
-\newcommand\argequals{{\,=\,}}
-\newcommand\lags{c}
-\newcommand\maxlag{\overline{c}}
-\newcommand\nlfList{C}
-\newcommand\bigO{\mathcal{O}}
-\newcommand\loglik{\lambda}
-\newcommand\loglikMC{\MC{\loglik}}
-\newcommand\R{\mathbb{R}}
-\newcommand\param{\,;}
-\newcommand\mycolon{{\hspace{0.6mm}:\hspace{0.6mm}}}
-\newcommand\MC[1]{#1^{\,\mbox{\tiny MC}}}
-\newcommand\EMC{{\E}}
-\newcommand\Var{\mathrm{Var}}
-\newcommand\var{\Var}
-\newcommand\Cov{\mathrm{Cov}}
-\newcommand\cov{\Cov}
-\newcommand\iid{\mathrm{iid}}
-\newcommand\dist{d}
-\newcommand\transpose{\top}
-
-\newcommand\gnsep{,}
-
-\def\lik{L}
-\def\loglik{\ell}
-
-%% THEOREM-LIKE ENVIRONMENTS
-\usepackage{amsthm}
-\newtheorem{prop}{Proposition}
-\newtheorem{theorem}{Theorem}
-\newtheorem{lemma}{Lemma}
-\newtheorem{example}{Example}
-\newtheorem{remark}{Remark}
-\newtheorem{cor}{Corrolary}
-\newtheorem{fact}{Fact}
-\newtheorem{assumption}{Assumption}
-\newtheorem{assumptionB}{Assumption}
-\renewcommand\theassumption{A\arabic{assumption}}
-\renewcommand\theassumptionB{B\arabic{assumptionB}}
-
-%% FOR PSEUDOCODE SETUP
-\newcommand\mystretch{\rule[-2mm]{0mm}{5mm} }
-\newcommand\asp{\hspace{4mm}}
-\newcommand\codeIndent{\hspace{4mm}}
-
-%%%%%%%%%%%%%%%% END OF USER DEFINITIONSS %%%%%%%%%%%%%%%
-
-\bibliographystyle{apalike}
-\usepackage{natbib}
-\usepackage{fullpage}
-
-\newcommand\myeqref[1]{(\ref{#1})}
-
-\renewcommand{\contentsname}{Supplementary Content}
-\renewcommand{\refname}{Supplementary References}
-\renewcommand\thefigure{S-\arabic{figure}}
-\renewcommand\thetable{S-\arabic{table}}
-\renewcommand\thesection{S\arabic{section}}
-\renewcommand\theequation{S\arabic{equation}}
-\renewcommand\theprop{S\arabic{prop}}
-\renewcommand\thelemma{S\arabic{lemma}}
-\renewcommand\thealgocf{S\arabic{algocf}}
-
-\setcounter{tocdepth}{1}
-
-\begin{document}
-
-\date{\today}
-\title{Supplement to ``{\mytitle}''}
-  \author{Jesse Wheeler, AnnaElaine L. Rosengart, Zhuoxun Jiang, \\ Kevin Tan, Noah Treutle and Edward L. Ionides
- \\ \hspace{.2cm}\\
-    Department of Statistics, University of Michigan\\
-}
-
-\newcommand{\blind}{1}
-
-\if1\blind
-{
-\maketitle
-}
-\fi
-
-\if0\blind
-{
-  \bigskip
-  \bigskip
-  \bigskip
-  \begin{center}
-    {\LARGE\bf \mytitle}
-\end{center}
-  \bigskip
-  \bigskip
-}\fi
-
-\tableofcontents
-
-\newpage
-
-\section{Markov chain and differential equation interpretations of compartment flow rates}
-
-In Sections 2.1, 2.2 and 2.3 of the main article, we define compartment models in terms of their flow rates.
-For a discrete population model, these rates define a Markov chain.
-For a continuous and deterministic model, the rates define a system of ordinary differential equations.
-Here, we add additional details to clarify the mapping from a collection of rate functions to a fully specified process.
-Our treatment follows \citet{breto09}.
-
-A general compartment model is a vector-valued process $X(t)=(X_1(t),\dots,X_c(t))$ denoting the (integer or real-valued) counts in each of $c$ compartments, where $t$ is any continuous value in the interval  $\left[t_0, \infty\right)$ for some real valued starting time $t_0$.
-The compartments may also have names, but to set up general notation we simply refer to them by their numerical index.
-The basic characteristic of a compartment model is that $X(t)$ can be written in terms of the flows $N_{ij}(t)$ from $i$ to $j$.
-A flow into compartment $i$ from outside the system is denoted by $N_{\demography i}$, and a flow out of the system from compartment $i$ is denoted by $N_{i\demography}$.
-We call $\demography$ a source/sink compartment, though it is an irregular compartment since $X_{\demography}(t)$ is not defined.
-These flows are required to satisfy a ``conservation of mass'' identity:
-\begin{equation}
-X_i(t)=X_i(t_0)+N_{\demography i}(t) - N_{i\demography}(t) + \sum_{j\neq
-i}N_{ji}(t)-\sum_{j\neq i}N_{ij}(t). \label{eq:conservation}
-\end{equation}
-Each {\em flow} $N_{ij}(t)$ is associated with a {\em rate} function
-$\mu_{ij}=\mu_{ij}(t,X(t))$, where we include the possibility that $i$ or $j$ takes value $\demography$.
-
-There are different ways to use a collection of rate functions to build a fully specified model.
-We proceed to describe the ones we use in this paper: via a system of ordinary differential equations (Sec.~\ref{subsec:ode}), a simple Markov counting system (Sec.~\ref{subsec:smcs}), and an overdispersed Markov counting system (Sec.~\ref{subsec:odmcs}).
-Other representations include stochastic differential equations driven by Gaussian noise or Gamma noise \citep{bhadra11}.
-
-\subsection{Ordinary differential equation (ODE) interpretation}
-\label{subsec:ode}
-
-A basic deterministic specification is
-\begin{equation}
-\label{eq:ode1}
-dN_{ij}/dt = \mu_{ij}\big(t,X(t) \big) X_i(t), \hspace{3mm} i\in \seq{1}{c}, \hspace{3mm} j\in \seq{1}{c} \cup\{\demography\}, \hspace{3mm} i\neq j,
-\end{equation}
-where $\mu_{ij}\big(t,X(t)\big)$ is called a per-capita rate or a unit rate.
-Flows into the system require special treatment since $X_i(t)$ in \myeqref{eq:ode1} is not defined for $i=\demography$.
-Instead, we specify
-\begin{equation}
-\label{eq:ode2}
-dN_{\demography i}/dt = \mu_{\demography i}\big(t,X(t) \big).
-\end{equation}
-This is the the interpretation and implementation used for Model~2 in our study.
-
-\subsection{Simple Markov counting system interpretation}
-\label{subsec:smcs}
-A continuous time Markov chain can be specified via its infinitesimal transition probabilities.
-A basic approach to this is to define
-\begin{eqnarray}
-\label{eq:smcs1}
-\prob\big[ N_{ij}(t+\delta)-N_{ij}(t)=0 \given X(t)\big]
- &=& 1-\delta \mu_{ij}\big(t,X(t)\big)X_i(t) + o(\delta),
-\\
-\label{eq:smcs2}
-\prob\big[ N_{ij}(t+\delta)-N_{ij}(t)=1 \given X(t)\big]
- &=& \delta \mu_{ij}\big(t,X(t)\big)X_i(t) + o(\delta),
-\end{eqnarray}
-for $i\in \seq{1}{c}$ and $j\in\seq{1}{c}\cup\{\demography\}$ with $i\neq j$.
-As with the ODE case, we need special attention for flows into the system, and we define
-\begin{eqnarray}
-\label{eq:smcs3}
-\prob\big[ N_{\demography i}(t+\delta)-N_{\demography i}(t)=0 \given X(t)\big]
- &=& 1-\delta \mu_{\demography i}\big(t,X(t)\big) + o(\delta),
-\\
-\label{eq:smcs4}
-\prob\big[ N_{\demography i}(t+\delta)-N_{\demography i}(t)=1 \given X(t)\big]
- &=& \delta \mu_{\demography i}\big(t,X(t)\big) + o(\delta).
-\end{eqnarray}
-Together with the initial conditions $X(0)$, equations \myeqref{eq:smcs1}--\myeqref{eq:smcs4} define a Markov chain.
-Each flow is a simple counting process, meaning a non-decreasing integer-valued process that only has jumps of size one.
-We therefore call the Markov chain a simple Markov counting system (SMCS).
-The infinitesimal mean of every flow is equal to its infinitesimal variance \citep{breto11} and so an SMCS is called equidispersed.
-We note that the special case of Model~1 used by \cite{lee20} (with $\sigmaProc = 0$) is an SMCS.
-To permit more general mean-variance relationships for a Markov counting system, we must permit jumps of size greater than one.
-The utility of overdispersed models, where the infinitesimal variance of the flow exceeds the infinitesimal mean, has become widely recognized \citep{stocks20,he10}.
-
-\subsection{Overdispersed Markov counting system interpretation}
-\label{subsec:odmcs}
-
-Including white noise in the rate function enables the possibility of an overdispersed Markov counting system \citep{breto11,breto09,he10}.
-Since rates should be non-negative, Gaussian noise is not appropriate and gamma noise is a convenient option that has found various applications \citep{romero-severson15, subramanian20}.
-Specifically, we consider a model given by
-\begin{equation}
-\label{eq:odmcs1}
-\mu_{ij}\big(t,X(t)\big) = \bar\mu_{ij}\big(t,X(t)\big) \, d\Gamma_{ij}(t)/dt,
-\end{equation}
-where $\Gamma_{ij}(t)$ is a stochastic process having independent gamma distributed increments, with
-\begin{equation}
-\label{eq:odmcs2}
-\E\big[\Gamma_{ij}(t)\big] = t, \quad \var\big[\Gamma_{ij}(t)\big] = \sigma_{ij}^2 t.
-\end{equation}
-Formally interpreting the meaning of \myeqref{eq:odmcs1} is not trivial, and we do so by constructing a Markov process $X(t)$ as the limit of the Euler scheme descrived in Section~\ref{sec:numerics}, below.
-Therefore, the numerical scheme in Sec.~\ref{sec:numerics} can be taken as a definition of the meaning of \myeqref{eq:odmcs1}.
-The Markov chain defined by the limit of this Euler scheme as the step size decreases is an overdispersed Markov counting system, with the possibility of instantaneous jumps of size greater than one \citep{breto11}.
-
-\section{Numerical solutions to compartment models}
-\label{sec:numerics}
-
-Models may be fitted and their implications assessed via numerical solutions (i.e., simulations) from the model equations.
-All the analyses we consider have this simulation-based property, known as plug-and-play or equation-free or likelihood-free.
-The numerical solutions to the model are arguably of more direct scientific interest than the exact solutions to the postulated equations.
-For ODE models, numerical methods are well studied and a standard numerical solution package such as \code{deSolve} in \code{R} is adequate for many purposes.
-For SMCS and ODMCS models, exact schemes are feasible when the number of events is small, which may be the case for small populations.
-However, for applicability to larger populations, we use instead the following Euler scheme.
-Write $\delta$ for an Euler time step, and $\Delta N_{ij}$ for the numerical approximation to $N_{ij}(t+\delta)-N_{ij}(t)$ given $X(t)$.
-For each $i$ and $j$ in $\seq{1}{c} \cup \{\demography\}$ with $i \neq j$, we draw independent Gamma distributed noise increments with mean $\delta$ and variance $\sigma_{ij}^2 \delta$, denoted using a mean-variance parameterization of the gamma distribution as
-\begin{equation}
-\label{eq:numerics1}
-\Delta\Gamma_{ij} \sim \mathrm{gamma}(\delta, \sigma_{ij}^2 \delta).
-\end{equation}
-In the case of an SMCS model, $\sigma_{ij}=0$ for all $i$ and $j$, so we have $\Delta\Gamma_{ij}=\delta$.
-Then, for $i\neq \demography$ and $j\neq i$, and writing
-\begin{equation}
-\label{eq:numerics2}
-\mu_{ij}=\bar\mu_{ij}\big(t,X(t)\big) \Delta\Gamma_{ij} / \delta,
-\end{equation}
-we calculate transition probabilities
-\begin{eqnarray}
-\label{eq:numerics3}
-p_{ij} &=& \exp\left\{-\sum_{k\in 1:c \, \cup \{\demography\}} \mu_{ik} \, \delta \right\}
-\frac{\mu_{ij}}{\sum_{k\in 1:c\cup \{\demography\}} \mu_{ik}},
-\\
-\label{eq:numerics4}
-p_{ii} &=& 1 - \sum_{j\neq i} p_{ij}.
-\end{eqnarray}
-These probabilities correspond to competing hazards for every individual in compartment $i$ to transition to some compartment $j$, interpreting $j=i$ to mean that the individual remains in $i$.
-Then, $\big(\Delta N_{i1},\dots,\Delta N_{ic},\Delta N_{i\demography}\big)$ has the multinomial distribution where $X_i(t)$ individuals are allocated independently to $\seq{1}{c}\cup\{\demography\}$ with probabilities given by \eqref{eq:numerics3} and \eqref{eq:numerics4}.
-We use the \code{reulermultinom} function in the \code{pomp} package to draw from this multionomial distribution.
-
-Different treatments of demographic flows---such as birth, death, immigration and emigration---are possible.
-For the case $i=\demography$, the treatment used by Model~1 is to set
-\begin{equation}
-\label{eq:numerics6}
-\Delta N_{\demography j} \sim \mathrm{poisson}( \mu_{\demography j} \delta),
-\end{equation}
-an independent Poisson random variable with mean $\mu_{\demography j} \delta$.
-
-Models~2 and 3 used an alternative approach, balancing the total number of flows in and out of the compartment, i.e., $\sum_{i}N_{\demography i}(t) = \sum_{i}N_{i \demography}(t)$, in order to make the model consistent with the known total population.
-In this case, we formally model the death rate as a rate of returning to the susceptible class $S$, and use external transitions from $\demography$ into $S$ to describe only net population increase.
-
-\section{Measurement Models}
-
-Each POMP requires specification of a measurement model, which is a statistical description of how observations on the system are obtained.
-In general, we used the same measurement models that were reported by \cite{lee20} unless specifically noted in the following subsections.
-
-\subsection{Model~1}
-
-In this model, the advantage afforded by vaccination is an increased probability that an infection is asymptomatic.
-Therefore, under the assumptions of this model, all reported cases are assumed to be a fraction of individuals that transition from the exposed to the infected compartment, as noted in Eq.~\myeqref{model1:meas}:
-\begin{equation}
-  \label{model1:meas}
-  Y_{n} \mid \Delta N_{E_{\cdot}I_{\cdot}}(n) = \reportChange \sim \text{NB}\left(\reportRate \reportChange, \obsOverdispersion \right),
-\end{equation}
-where $Y_n$ is the reported cholera cases at time $t_n \in t_1:t_N$ and $\Delta N_{E_{\cdot}I_{\cdot}}(n)$ is the sum total of individuals across each vaccination compartment $\vaccCounter \in 1:\vaccClass$ who moved from compartment $E_\vaccCounter$ to $I_\vaccCounter$ since observation $t_{n-1}$.
-Here, $\text{NB}\left(\reportRate \reportChange, \obsOverdispersion \right)$ denotes a negative binomial distribution with mean $\reportRate \reportChange$ and variance $\reportRate \reportChange \big(1 + \frac{\reportRate \reportChange}{\obsOverdispersion}\big)$.
-
-\subsection{Model~2}\label{sec:mod2Meas}
-
-Model~2 was fit using reported case counts that were transformed using the natural logarithm.
-We fit Model~2 using the subplex algorithm in the \code{subplex} package, using
-a Gaussian measurement model (Eq.~\myeqref{model2:meas}) on the log transformed cases within each unit.
-The final loss function that is maximized is the product of the likelihoods of the individual units, or the sum of the log-likelihood of the individual units.
-The measurement model for individual units is given in Eq.~\myeqref{model2:meas}.
-\begin{equation}
-  \label{model2:meas}
-  \log\big(Y_{u, n} + 1\big) \mid \Delta N_{E_{u\cdot}I_{u\cdot}}(n) = \reportChange_u \sim \text{N}\left(\log \big(\reportRate \reportChange_u + 1\big), \obsOverdispersion^2 \right),
-\end{equation}
-where $\Delta N_{E_{u\cdot}I_{u\cdot}}(n)$ is the sum total of individuals across vaccination compartment $\vaccCounter \in 0:4$ within unit $u$ who moved from compartment $E_{u\vaccCounter}$ to $I_{u\vaccCounter}$ since observation $t_{n-1}$.
-Therefore, because the natural logarithm of observed case counts (plus one, to avoid taking the logarithm of zero) has a normal distribution, $Y_{u, n} + 1$ is assumed to follow a log-normal distribution with log-mean parameter $\log\big(\reportRate \Delta N_{E_{u\cdot}I_{u\cdot}}(n) + 1\big)$ and log-variance $\obsOverdispersion^2$.
-We note that fitting a model with this measurement model is equivalent to fitting using least squares, with $\log(Y_{u, n} + 1)$ as the response variable.
-
-This measurement model differs from that used by \citet{lee20}, who fit the model in two stages: epidemic and endemic phases.
-Although their text and supplement material do not explicitly describe the measurement model used, inspection of the code provided with their submission suggests a change in measurement model between the epidemic and endemic phases.
-% In the file \code{choleraEqsPublic.py}, \citet{lee20} create several functions, where each function represents a set of coupled differential equations that could potentially be used to model cholera incidence data.
-% Each function returns a vector (or \code{numpy} array) that represents the change in each state variable for a single time step, including the variable \code{dC}, which tracks the number of new infections and is used to obtain the reported case counts.
-% Following their comments and code, it appears that the function \code{choleraEqs10WithoutVaccinationNetwork} was used to describe the dynamics of the epidemic phase, and \code{choleraEqs11WithoutVaccinationNetwork} was used for the endemic stage (see Sec~\ref{sec:mod2rep} for more details).
-% Because their models were fit using least squares, the code in these functions suggest that the measurement model for the epidemic phase is
-The measurement model they used for the epidemic phase is
-\begin{equation}
-  \label{model2:measEpi}
-  Y_{u, n} \mid \Delta N_{E_{u\cdot}I_{u\cdot}}(n) = \reportChange_u \sim \text{N}\left(\reportRate \reportChange_u, \obsOverdispersion^2 \right),
-\end{equation}
-% which is similar to our measurement model, the primary difference being that the measurement model is applied to raw case counts rather than log-transformed case counts.
-The measurement model they used for the endemic phase modifies the epidemic model by counting both asymptomatically infected (A) and symptomatically infected (I) individuals in the case counts:
-\begin{equation}
-  \label{model2:measEnd}
-  Y_{u, n} \mid \Delta N_{E_{u\cdot}I_{u\cdot}}(n) = \reportChange_{u1}, \Delta N_{E_{u\cdot}A_{u\cdot}}(n) = \reportChange_{u2} \sim \text{N}\left(\reportRate \big(\reportChange_{u1} + \reportChange_{u2}\big), \obsOverdispersion^2 \right),
-\end{equation}
-where the notation for $\Delta N_{E_{u\cdot}A_{u\cdot}}(n)$ is similar to $\Delta N_{E_{u\cdot}I_{u\cdot}}(n)$, described above.
-
-\subsection{Model~3}
-
-In this model, reported cholera cases are assumed to stem from individuals who develop symptoms and seek healthcare.
-Therefore reported cases are assumed to come from an over-dispersed negative binomial model, given the increase in infected individuals:
-\begin{equation}
-  \label{model3:meas}
-  Y_{u, n} \mid \Delta N_{S_{u \cdot}I_{u}}(t) = \reportChange_u \sim \text{NB}\left(\reportRate \reportChange_{u}, \obsOverdispersion \right),
-\end{equation}
-where $\Delta N_{S_{u \cdot}I_{u \vaccCounter}}(n)$ is the number of individuals who moved from compartment $S_{u \vaccCounter}$ to $I_{u}$ since observation $t_{n-1}$.
-
-This measurement model is a minor change from that used by \cite{lee20}, which allowed for a change in the reporting rate on January 1st, 2018.
-The fitted values of the reporting rate---before and after January 2018---were $0.97$ and $0.097$, respectively.
-An instantaneous change from near perfect to almost non-existent reporting can be problematic, as it forces the model to explain the observed reduction in reported cases as a decrease in the reporting of cases, rather than a decrease in the prevalence of cholera.
-This shift was justified by a ``change of the case definition that occurred on January 1st, 2018";
-this claim was not cited, and we could find no evidence that such a drastic change in the reporting rate would be warranted.
-We therefore do not allow a change in reporting rate when fitting Model~3.
-
-\section{Initial Values}
-
-To perform inference on POMP models, it is necessary to propose an initial probability density for the latent process $f_{X_0}(x_0;\theta)$, including the possibility that the initial values of the latent states are known, or are a non-random function of the unknown parameter vector, $\theta$.
-This density is used to obtain initial values of the latent state when fitting and evaluating the model.
-For each of the models considered in this analysis, the initial conditions are derived by enforcing the model dynamics on reported cholera cases.
-It is also sometimes necessary to fit some initial value parameters in order to help determine initial values for weakly identifiable compartments.
-In the following subsections, we mention initial value parameters that were fit for each model.
-
-\subsection{Model~1}
-
-For this model, the number of individuals in the Recovered and Asymptomatic compartments are set to zero, but the initial proportion of Infected and Exposed individuals is estimated as initial value parameters ($\Iinit$ and $\Einit$, respectively) using the IF2 algorithm, implemented as \texttt{mif2} in the \texttt{pomp} package.
-Finally, the initial proportion of Susceptible individuals $S_{0, 0}$ is calculated as $S_{0, 0} = 1 - \Iinit - \Einit$.
-This model for the initial values of the latent states matches that which was used by \citet{lee20}.
-
-\subsection{Model~2}
-
-Model~2 assumes that the initial values are a deterministic function of the reporting rate and the initial case reports, and so no initial value parameters need to be estimated.
-Initial values for latent state compartments are chosen so as to enforce the model dynamics on the observed number of cases.
-Specifically, the model sets $I_{u0}(0) = y^*_{1u} / \reportRate$ for each unit $u \in 1:10$, where $I_{u0}(t)$ is the number of infected individuals in unit $u$ at time $t$, vaccination scenario $z = 0$, $y^*_{tu}$ is the reported number of cases, and $\reportRate$ is the reporting rate.
-This model for the initial values of the latent states matches that which was used by \citet{lee20}.
-
-\subsection{Model~3}
-
-The latent states of this model are initialized by enforcing the model dynamics on the incidence data from the start of the recorded cases until time $t_0$, requiring that some of the available data be used to determine the initial values of the latent states.
-This is the same approach that was taken by \citet{lee20}, who used the value $t_0 = \text{2014-02-22}$; this choice of $t_0$ results in modeling roughly only $60\%$ of the available data, some of which is later discarded for alternative reasons \citep{lee20sup}.
-
-We do not see any immediate reason that this model could not be extended to cover a larger range of the data, and chose the value $t_0 = \text{2010-11-13}$.
-This choice of $t_0$ corresponds to using approximately $1\%$ of the available data to determine initial values of the latent states.
-In addition to modeling a larger portion of the available data, this choice of $t_0$ corresponds to an important real-world event, as daily reports from each of the departments were not being sent to the health ministry until November 10, 2010; % TODO: Cite
-this choice of $t_0$ therefore makes $\bm{Y}\big(t_1\big)$ the first week of data once daily reports are being sent to the health ministry.
-The few observation times that exist before $t_0$ are used to initialize the model by enforcing model dynamics on these preliminary observations.
-For convenience, we denote these observations as $t_{-3}, t_{-2}$ and $t_{-1}$; as before, we let $y_{u,-k} = Y_{u}\big(t_{-k}\big)$ denote the observed case count for unit $u$ at time point $t_{-k}$, where $k \in 1:3$.
-Equations for the initial values of non-zero latent states are provided in Eqs.~\myeqref{eq:model3:I0}--\myeqref{eq:model3:B0}.
-
-\begin{eqnarray}
-\label{eq:model3:I0}
-I_{u}\big(t_0\big) &=& \frac{y^*_{u, -1}}{7\reportRate\left(\muIR + \left(\muDeath + \choleraDeath\right)/365\right)},
-\\%[-3pt]
-\label{eq:model3:A0}
-A_{u}\big(t_0\big) &=& \frac{I_{u0}\big(t_0\big)\left(1 - \symptomFrac\right)}{\symptomFrac},
-\\
-\label{eq:model3:R0}
-R_{u01}\big(t_0\big) &=& R_{u02}\big(t_0\big) = R_{u03}\big(t_0\big) = \left(\frac{\sum_{k = -3}^0 y^*_{u, k}}{\reportRate \symptomFrac} - \left(I_{u0}\big(t_0\big) + A_{u0}\big(t_0\big)\right)\right) / 3
-\\
-\label{eq:model3:S0}
-S_{u0}\big(t_0\big) &=& \mathrm{Pop}_u - I_{u}\big(t_0\big) - A_{u}\big(t_0\big) - \sum_{k=1}^3 R_{u0k}\big(t_0\big)
-\\
-\label{eq:model3:B0}
-B_{u}\big(t_0\big) &=& \big[1 + \seasAmplitude \tilde{J}^r \big] \mathrm{Den}_u \, \Wshed \big[ I_{u}(t_0)+ \asymptomRelativeShed A_{u_0}(t) \big]/\Wshed.
-\end{eqnarray}
-
-In Eq.~\myeqref{eq:model3:B0}, $\tilde{J} = 0.002376$ is the median adjusted rainfall over the observation period.
-One important consideration to make with this parameter initialization model is when $y_{u, -1}^* = 0$, which occurs for units $u \in \{3, 4\}$, which correspond to the Grand'Anse and Nippes departments, respectively.
-When this is the case, each of the infectious $I_u(t_0)$, asymptomatic $A_u(t_0)$, and bacterial reservoir $W_u(t_0)$ compartments have a value of zero.
-Because Model~3 models cholera transmission primarily by means of the bacterial reservoir, this makes it nearly impossible for an outbreak to occur.
-Therefore for units $u \in \{3, 4\}$, we introduce initial value parameters $I_{0,0}^3$ and $I_{0,0}^4$, and calibrate these parameter values using the data.
-The resulting parameter estimates are used to obtain the remaining non-zero initial values of the latent states using Eqs.~\myeqref{eq:model3:A0}--\myeqref{eq:model3:B0}.
-
-% Argue why we didn't need to do this for Model 2.
-
-% TODO: Check this section.
-
-\section{Calibrating Model~3 to observed cases}\label{sec:mod3Cal}
-
-<<CalibrateOrig3>>=
-# Create vectors for the unit and shared parameters
-unit_specific_names <- c("betaB", "foi_add")
-shared_param_names <- c(
-  "mu_B", "XthetaA", "thetaI", "lambdaR", "r", "std_W",
-  "epsilon", "k"
-)
-est_param_names <- c(
-  unit_specific_names, shared_param_names
-)
-
-# Add unit numbers to each parameter
-est_param_names_expanded <- paste0(rep(est_param_names, each = 10), 1:10)
-
-# Simple function that is used to set the rw_sd for the first search
-# This function is convenient so that we don't have to type out the rw_sd
-# for each unit parameter.
-set_rw_1 <- function(x) {
-  if (gsub("[[:digit:]]+$", "", x) %in% shared_param_names) {
-    return(0.01)
-  # } else if (x %in% paste0(rep(c("aHur", 'hHur'), each = 2), c(3, 9))) {
-  #   return(expression(ifelse(time >= 2016.754 & time <= 2017, 0.015, 0)))
-  # } else if (x %in% c("Iinit3", "Iinit4")) {
-  #   return(expression(ivp(0.25)))
-  } else if (grepl('^foi_add[[:digit:]]+$', x)) {
-    return(0.015)
-  } else if (grepl('^betaB[[:digit:]]+$', x)) {
-    return(0.01)
-  } else {
-    return(0)
-  }
-}
-
-set_rw_2 <- function(x) {
-  if (gsub("[[:digit:]]+$", "", x) %in% shared_param_names) {
-    return(0.003)
-  # } else if (x %in% paste0(rep(c("aHur", 'hHur'), each = 2), c(3, 9))) {
-  #   return(expression(ifelse(time >= 2016.754 & time <= 2017, 0.005, 0)))
-  # } else if (x %in% c("Iinit3", "Iinit4")) {
-  #   return(expression(ivp(0.125)))
-  } else if (grepl('^foi_add[[:digit:]]+$', x)) {
-    return(0.005)
-  } else if (grepl('^betaB[[:digit:]]+$', x)) {
-    return(0.003)
-  } else {
-    return(0)
-  }
-}
-
-# First Search RW
-reg_rw_1.sd <- lapply(est_param_names_expanded, set_rw_1)
-names(reg_rw_1.sd) <- est_param_names_expanded
-chol_rw_1.sd <- do.call(rw_sd, reg_rw_1.sd)
-
-# Second Search RW
-reg_rw_2.sd <- lapply(est_param_names_expanded, set_rw_2)
-names(reg_rw_2.sd) <- est_param_names_expanded
-chol_rw_2.sd <- do.call(rw_sd, reg_rw_2.sd)
-
-SEARCH1 = list(
-  NBPF = 100,
-  NP = 2500,
-  SPAT_REGRESSION = 0.05,
-  NREPS = 144,
-  NP_EVAL = 2500,
-  NREPS_EVAL = 18,
-  RW_SD = chol_rw_1.sd,
-  COOLING = 0.5,
-  KEEP_TRACES = FALSE,
-  KEEP_LIKE_MAT = FALSE
-)
-
-SEARCH2 = list(
-  TOP_N = 12,
-  NBPF = 100,
-  NP = 2500,
-  SPAT_REGRESSION = 0.05,
-  NREPS = 6,
-  NP_EVAL = 2500,
-  NREPS_EVAL = 18,
-  RW_SD = chol_rw_2.sd,
-  COOLING = 0.5,
-  KEEP_TRACES = FALSE,
-  KEEP_LIK_MAT = TRUE
-)
-
-haiti3_sub_fit <- bake(
-  file = "../model3/si/subset_noparams.rds", {
-    fit_haiti3(
-      nsearches = 2L,
-      search1 = SEARCH1,
-      search2 = SEARCH2,
-      search_rho = FALSE,
-      search_gamma = FALSE,
-      search_Iinit = FALSE,
-      search_hur = FALSE,
-      ncores = cores
-    )
-  }
-)
-
-# Find model index with best log-likelihood
-best_m <- which.max(haiti3_sub_fit$search2$logLiks$logLik)
-
-# Find parameters using best_m index
-sub_params <- haiti3_sub_fit$search2$params[best_m, ]
-
-# Get unit log-likelihoods for the best model.
-all_liks <- as.data.frame(haiti3_sub_fit$search2$likMat)
-# all_liks$which <- rep(1:(SEARCH3$TOP_N * SEARCH3$NREPS), each = SEARCH3$NREPS_EVAL)
-all_liks$which <- rep(1:(12 * 6), each = 18)
-best_mod <- filter(all_liks, which == best_m)
-
-# Pivot longer to get estimate of log-likelihood for each unit.
-unitLikesM3 <- best_mod %>%
-  pivot_longer(
-    cols = -which,
-    names_to = "unit",
-    names_prefix = "V",
-    values_to = "logLik"
-  ) %>%
-  select(-which) %>%
-  group_by(unit) %>%
-  summarize(
-    ll_est = logmeanexp(logLik),
-    ll_se = logmeanexp(logLik, se = TRUE)[2]
-  ) %>%
-  ungroup() %>%
-  mutate(unit = as.numeric(unit)) %>%
-  arrange(unit)
-@
-
-<<M3_bench, echo=FALSE, message=FALSE, include=FALSE, cache=TRUE>>=
-h3 <- haiti3_spatPomp()
-deps <- h3@unit_names
-
-mod3_cases <- as.data.frame(t(h3@data))
-colnames(mod3_cases) <- deps
-m3_inits <- c(7362, 1914, 22, 8, 2818, 82, 2499, 6953, 44, 7) # TODO: Also fit by hand, but could be changed to initializing to the observation before.
-names(m3_inits) <- deps
-
-bench_cond_lls <- list()
-for (dep in deps) {
-
-  # Get department cases
-  cases <- as.numeric(mod3_cases[, dep])
-  mod <- ar1_NegBinom(data = cases, init = m3_inits[dep])
-
-  bench_cond_lls[[dep]] <- haitipkg:::.ar1_NegBinom_cond_ll(
-    theta = mod$theta, init = m3_inits[dep], data = cases, transform = FALSE
-  )
-}
-
-bench_cond_lls <- as.data.frame(bench_cond_lls)
-
-bench_lls <- apply(bench_cond_lls, 2, sum)
-
-unitLikes_bench <- data.frame(
-  unit = 1:10,
-  # dep = names(bench_lls),
-  ll = bench_lls
-) %>%
-  as_tibble()
-
-bench_cond_lls$date <- lubridate::round_date(lubridate::date_decimal(h3@times), unit = 'day')
-@
-
-<<combine_bench_M3, echo=FALSE>>=
-
-# Combine tables with model 3 and benchmark likelihoods.
-bench_M3_joined <- dplyr::left_join(
-  x = unitLikesM3,
-  y = unitLikes_bench,
-  by = "unit"
-) %>%
-  select(-ll_se) %>%
-  mutate(unit = as.character(unit)) %>%
-  pivot_longer(
-    cols = -unit,
-    names_to = "type",
-    values_to = "ll"
-  ) %>%
-  mutate(unit = paste0(unit, ": ", deps[as.numeric(unit)]))
-@
-
-<<M3bpf, echo=FALSE, message=FALSE, include=FALSE>>=
-
-# Change model parameters to the best set of parameters from previous
-# search
-coef(h3)[names(sub_params)] <- sub_params
-
-registerDoRNG(40765101)  # for modular, reproducible results
-
-# Perform block partical filter with fitted model, to get conditional
-# log-likelihoods.
-bpf_sub3_out <- bake(
-  file = "../model3/si/bpf_sub_cond_LL.rds", {
-
-    results <- list()
-
-    bpf_out <- h3 %>% bpfilter(
-      Np = 5000,
-      block_size = 1,
-      save_states = TRUE
-    )
-
-    results$block.cond.loglik <- bpf_out@block.cond.loglik
-    results$times <- bpf_out@times  # Saved for convenience
-    results
-  },
-  seed = 432235
-)
-@
-
-In this section, we provide more detail on the process that was used to estimate the coefficients used in Model~3.
-In particular, we discuss why we decided to include additional model parameters---those that are associated with the behavior of the system during Hurricane Matthew---that were not considered by \cite{lee20}.
-To calibrate this model, we used the iterated block particle filter (IBPF) method of \cite{ionides22}.
-Due to the novelty of this algorithm, there does not exist published examples the IBPF algorithm used for data analysis outside of the papers in which the algorithm was introduced \citep{ning21ibpf,ionides22}, which at least partially motivates the inclusion additional details in this section.
-
-\cite{lee20} were only able to estimate model parameters to a simplified version of Model~3 on a subset of the available data, as no method existed at the time of their publication to fit a fully coupled meta-population model to disease incidence data.
-Building on their results, we fit the fully coupled version of Model~3 to nearly all available data, reserving only a few observations to use to calibrate the initial conditions of the model (see the previous section for more details).
-In order to properly maximize the model likelihood, we found that it was necessary to use multiple searches for the MLE, periodically pruning away less successful searches.
-The first collection of searches was performed by obtaining initial values for the parameters by uniformly sampling values from a predefined hypercube.
-A subsequent refinement search used starting values corresponding to the largest model likelihoods obtained in the previous round.
-This is similar to the approach used by \cite{ionides22}.
-
-We use the technique described above to fit the fully coupled version of Model~3 proposed by \cite{lee20}.
-The maximum likelihood we obtained with the model was $\Sexpr{round(max(haiti3_sub_fit$search2$logLiks$logLik))}$, which is higher than the benchmark model ($\Sexpr{round(sum(unitLikes_bench$ll))}$).
-While beating a simple associative benchmark is promising, this does not immediately rule out the possibility that the model is a poor description of the system.
-Additional investigation of parameters estimates and their corresponding implications on model based conclusions should always be conducted.
-For meta-population models, it is worth considering how well the model fits the data to each spatial unit.
-This can be done by looking at conditional log-likelihoods, which is part of the output of the \texttt{bpfilter} algorithm in the \texttt{spatPomp} package.
-The likelihoods for each department, compared to the corresponding benchmark model, are displayed in Fig.~\ref{fig:h3UnitLikes}.
-The figure demonstrates that while the log-likelihood of the fitted model outperforms the auto-regressive negative binomial benchmark model, Model~3 has lower likelihoods than the benchmark for some departments.
-
-\begin{figure}[ht]
-<<h3UnitLikes, fig.height=3, fig.width=5.5, echo=FALSE>>=
-ggplot(bench_M3_joined, aes(x = ll, y = reorder(unit, ll))) +
-  geom_point(aes(col = type)) +
-  ylab("Unit") +
-  xlab("Log Likelihood") +
-  scale_color_manual(
-    labels = c("ll" = "Benchmark", "ll_est" = "Model 3"),
-    values = c("ll" = "#1b9e77", "ll_est" = "#d95f02")
-  ) +
-  theme(legend.title = element_blank(),
-        axis.title.y = element_blank())
-@
-\caption{\label{fig:h3UnitLikes}Log-likelihoods of Model~3 for each department compared to the corresponding benchmark model prior to the inclusion of parameters for initial values and Hurricane Matthew.}
-\end{figure}
-
-In addition to considering the conditional log-likelihoods of each unit, one can consider conditional log-likelihoods of each observation.
-When compared to a benchmark, this level of detail can provide useful information about which observations are well described by the model and which are poorly described.
-In Fig.~\ref{fig:condLL}, we plot the conditional log-likelihoods of each observation.
-This figure reveals that the fitted model has difficult time explaining certain features of the data.
-For example, many departments (in particular Sud) have observations with lower conditional log-likelihoods near October 2016 than at other time points.
-Further investigation of the model output reveals that the model is struggling to explain the sudden surge in cholera cases that occurred at this time, which coincides with the time that Hurricane Matthew struck Haiti.
-While the model does include a mechanism to account for increased cholera transmission due to large rainfall events, the mechanism does not appear to be sufficient to capture the damaging effects of the hurricane, which had the greatest impact in the the Sud and Grand'Anse departments \citep{ferreirai16}.
-This result led us to include parameters $\Whur{u}$ and $\hHur{u}$ in Eq.~23 of the main text, which allows for an increase in the transmission rate between environmental cholera and humans for Sud and Grand'Anse during and after the hurricane.
-The effect of the hurricane on cholera transmission is assumed to have an exponential decay, where the magnitude is determined by $\Whur{u}$ and the duration of the effect determined by $\hHur{u}$.
-
-\begin{figure}[ht]
-<<CondLLFig, echo=FALSE, message=FALSE, fig.height=3>>=
-
-# Changing labels to human-readable labels.
-deps[grepl("_", deps)] <- gsub("_", "-", deps[grepl("_", deps)])
-deps[deps == "Grande-Anse"] <- "Grand\'Anse"
-names(deps) <- as.character(1:10)
-myLabeller <- as_labeller(deps)
-
-# Plot conditional log-likelihoods for each unit
-t(bpf_sub3_out$block.cond.loglik) %>%
-  as.data.frame() %>%
-  mutate(t = bpf_sub3_out$times) %>%
-  pivot_longer(
-    cols = -t,
-    values_to = 'likelihood',
-    names_prefix = "V",
-    names_to = 'Unit'
-  ) %>%
-  mutate(date = as.Date(lubridate::date_decimal(t))) %>%
-  ggplot(aes(x = date, y = likelihood)) +
-  geom_line() +
-  geom_point() +
-  facet_wrap(~Unit, labeller = myLabeller, scales = 'free_y',
-             nrow = 2) +
-  theme_bw() +
-  theme(axis.title.x = element_blank(),
-        axis.text.x = element_text(angle = -45, hjust = 0)) +
-  ylab("Conditional log-likelihoods")
-@
-\caption{\label{fig:condLL}Conditional log-likelihoods of Model~3 prior to the inclusion of parameters for initial values and Hurricane Matthew.}
-\end{figure}
-
-Other points of concern include the observations at or near the start of the recorded cases, for example, in Nord-Ouest and Sud-Est.
-These low conditional log-likelihoods are a result of having zero or near zero confirmed cases in the first week of observed data, followed by a rapid growth of cases.
-The model cannot explain the rapid growth of cholera in the absence of starting cases because the initialization scheme forces departments with no recorded cases in the first week to have hidden states $I_{u0}(0) = A_{u0}(0) = B_{u}(0) = 0$.
-In reality, however, it is possible to have some infected or asymptomatic individuals in a department when no cases are recorded due to under-reporting.
-This may be particularly true at the onset of an infectious disease outbreak, as lower public awareness and inadequate testing infrastructure may result in many cases going undetected.
-We therefore include and estimate $I_{0u}$ that determines the initial number of Infected individuals at the start of the pandemic for the departments Grande'Anse, Nord-Ouest, and Sud-Est;
-each of these departments have both low starting case counts followed by rapid growth, and low initial values of the conditinoal log-likelihood.
-
-<<loadFinalModel3, echo=FALSE, include=FALSE, message=FALSE>>=
-
-# This result was used in the manuscript, so it doesn't need to be recalculated.
-h3_final_results <- readRDS("../model3/run_level_3/haiti3_fit.rds")
-
-# Save best_m as index for the model parameters with the best likelihood.
-# Then, save the model parameter values and the likelihood evaluations
-best_m <- which.max(h3_final_results$search2$logLiks$logLik)
-best_params <- h3_final_results$search2$params[best_m, ]
-all_liks <- as.data.frame(h3_final_results$search2$likMat)
-all_liks$which <- rep(1:72, each = 18)
-best_mod <- dplyr::filter(all_liks, which == best_m)
-
-# Get repeated estimates of unit log-likelihoods
-unitLikesM3 <- best_mod %>%
-  pivot_longer(
-    cols = -which,
-    names_to = "unit",
-    names_prefix = "V",
-    values_to = "logLik"
-  ) %>%
-  select(-which) %>%
-  group_by(unit) %>%
-  summarize(
-    ll_est = pomp::logmeanexp(logLik),
-    ll_se = pomp::logmeanexp(logLik, se = TRUE)[2]
-  ) %>%
-  ungroup() %>%
-  mutate(unit = as.numeric(unit)) %>%
-  arrange(unit)
-
-# Combine model likelihood table to benchmarks
-bench_M3_joined <- dplyr::left_join(
-  x = unitLikesM3,
-  y = unitLikes_bench,
-  by = "unit"
-) %>%
-  select(-ll_se) %>%
-  mutate(unit = as.character(unit)) %>%
-  pivot_longer(
-    cols = -unit,
-    names_to = "type",
-    values_to = "ll"
-  ) %>%
-  mutate(unit = paste0(unit, ": ", deps[as.numeric(unit)]))
-@
-
-We refit Model~3 after introducing these seven parameters---three initial value parameters and four parameters related to Hurricane Matthew---using the same successive search scheme described above.
-The resulting model has a log-likelihood value of $\Sexpr{round(max(h3_final_results$search2$logLiks$logLik), 1)}$.
-We also provide figures for department specific log-likelihoods and conditional log-likelihoods in Figs.~\ref{fig:finalUnitLL} and \ref{fig:finalCondLL}.
-
-\begin{figure}[ht]
-<<finalUnitLLFig, fig.height=3, fig.width=5.5, echo=FALSE>>=
-ggplot(bench_M3_joined, aes(x = ll, y = reorder(unit, ll))) +
-  geom_point(aes(col = type)) +
-  ylab("Unit") +
-  xlab("Log Likelihood") +
-  scale_color_manual(
-    labels = c("ll" = "Benchmark", "ll_est" = "Model 3"),
-    values = c("ll" = "#1b9e77", "ll_est" = "#d95f02")
-  ) +
-  theme(legend.title = element_blank(),
-        axis.title.y = element_blank())
-@
-\caption{\label{fig:finalUnitLL}Log-likelihoods of Model~3 for each department compared to the corresponding benchmark model after adding and estimating parameters related to initial values and Hurricane Matthew.}
-\end{figure}
-
-<<finalModel3BPF, echo=FALSE, message=FALSE, include=FALSE>>=
-
-# Set the model parameters to the best results from previous search
-coef(h3)[names(best_params)] <- best_params
-
-# block particle filter to get conditional log-likelihoods.
-bpf_final3_out <- bake(
-  file = "../model3/si/h3_final_bpf_cond_ll.rds", {
-
-    results <- list()
-
-    bpf_out <- h3 %>% bpfilter(
-      Np = 5000,
-      block_size = 1,
-      save_states = TRUE
-    )
-
-    results$Sus_df <- foreach(i=1:length(h3@times), .combine = bind_rows) %do% {
-      get_h3_S_quants(bpf_out, i)
-    }
-
-    results$block.cond.loglik <- bpf_out@block.cond.loglik
-    results$times <- bpf_out@times  # saved for convenience
-    results
-  },
-  seed = 786321
-)
-@
-
-\begin{figure}[ht]
-<<finalCondLLFig, echo=FALSE, message=FALSE, fig.height=3>>=
-t(bpf_final3_out$block.cond.loglik) %>%
-  as.data.frame() %>%
-  mutate(t = bpf_final3_out$times) %>%
-  pivot_longer(
-    cols = -t,
-    values_to = 'likelihood',
-    names_prefix = "V",
-    names_to = 'Unit'
-  ) %>%
-  mutate(date = as.Date(lubridate::date_decimal(t))) %>%
-  ggplot(aes(x = date, y = likelihood)) +
-  geom_line() +
-  geom_point() +
-  facet_wrap(~Unit, labeller = myLabeller, scales = 'free_y',
-             nrow = 2) +
-  theme_bw() +
-  theme(axis.title.x = element_blank(),
-        axis.text.x = element_text(angle = -45, hjust = 0)) +
-  ylab("Conditional log-likelihoods")
-@
-\caption{\label{fig:finalCondLL}Conditional log-likelihoods of Model~3 after adding and estimating the parameters for initial values and Hurricane Matthew.}
-\end{figure}
-
-While the addition of the Hurricane parameters seems to have resulted in a dramatic increase of conditional likelihoods around and after October 2016, the inclusion of the initial value parameters does not appear to have fixed the low conditional log-likelihood values in Sud-Est and Nord-Ouest.
-This could be due to many factors, including the possibility that the coupling between units forces low initial value parameters in a given department in order for the model to fit the data in the remaining departments.
-Whatever the case may be, the inclusion of the seven parameters resulted in a $\Sexpr{round(max(h3_final_results$search2$logLiks$logLik) - max(haiti3_sub_fit$search2$logLiks$logLik), 1)}$ increase in log-likelihood, which is a highly statistically significant difference.
-One could similarly test whether the inclusion of the initial value parameters resulted in a significantly better fit to the observed data, but we do not do this here due to the large computational effort that would be required.
-
-\subsection{Examining the Hidden States of the Calibrated Model}
-
-For mechanistic models, beating a suitable statistical benchmark does not alone guarantee that the model provides an accurate description of a dynamic process.
-Indeed, a good statistical fit does not require the model to assert a causal explanation.
-For example, reconstructed latent variables should make sense in the context of alternative measurements of these variables \citep{grad12}.
-We demonstrate this principle by examining the latent state of the calibrated model.
-In particular, we examine the compartment of susceptible individuals under various scenarios, as this compartment drives transmission dynamics in SIR models for infectious disease outbreaks.
-
-Recall that the filtering distribution for the calibrated version of Model~3 at time $t_k$ is defined as the distribution of the latent state at time $t_k$ given the data from times $t_{1}:t_{k}$, i.e.
-$f^{(3)}_{\bm{X}_k|\bm{Y}_{1:k}}(\bm{x}_{k} | \bm{y}^*_{1:k} ; \hat\theta)$.
-In general, one may expect simulations from the filtering distribution of a model with a good statistical fit to result in hidden states that are highly consistent with the observed data because the filtering distribution is conditioned on the observed data.
-Fig.~\ref{fig:h3Sus} shows the percentage of individuals that are in the susceptible compartment from various simulations of the model:
-simulations from Model~3 under initial conditions are displayed in red; simulations from the filtering distribution of model are displayed in blue.
-This figure shows that simulations from initial conditions tends to result in a much more rapid depletion of susceptible individuals at the start of the epidemic than simulations from the filtering distribution, suggesting the calibrated model has a propensity to predict larger outbreaks than what is typically seen in the data.
-This also appears true in later stages of the time period of observed data, where the data suggests that there is a replenishment of the susceptible compartment, but the Model simulations typically retain a smaller proportion of susceptible individuals.
-This result demonstrates that the calibrated model favors a more rapid growth in cholera cases than what is typically seen in the observed data, which explains why the model fails to predict the absence of cases between February, 2019 and October, 2022.
-
-<<h3SusFigSetup, echo=FALSE, message=FALSE, include=FALSE>>=
-
-# This bake perfoms the following:
-#  (1) simulates the model from initial conditions
-#  (2) get the values of the hidden states at the end of each simulation, and
-#      saves these for later use.
-#  (3) Obtain quantiles for the Susceptible pool from the simulations.
-init_conditions_results <- bake(
-  file = "../model3/si/InitSimsQuants.rds", {
-
-    results <- list()  # To return final output of bake
-
-    # Simulate from the fitted model
-    h3Spat_sims <- simulate(
-      h3, nsim = 2000, format = 'data.frame'
-    )
-
-    # Get states at the end of the simulations
-    h3_spat_end <- h3Spat_sims %>%
-      filter(time == max(time))
-
-    # Save hidden state values at the end of the simulation. Saved as a matrix
-    # in the same format as the output from a block-particle filter, so that
-    # it can be used as the input to project_from_filter.
-    results[["end_conditions"]] <- h3_spat_end %>%
-      select(-time, -cases) %>%
-      mutate(
-        unit = case_when(
-          unitname == "Artibonite" ~ 1,
-          unitname == "Centre" ~ 2,
-          unitname == "Grande_Anse" ~ 3,
-          unitname == "Nippes" ~ 4,
-          unitname == "Nord" ~ 5,
-          unitname == "Nord_Est" ~ 6,
-          unitname == "Nord_Ouest" ~ 7,
-          unitname == "Ouest" ~ 8,
-          unitname == "Sud" ~ 9,
-          unitname == "Sud_Est" ~ 10
-        )
-      ) %>%
-      select(-unitname) %>%
-      pivot_wider(
-        data = .,
-        names_from = unit,
-        values_from = c(S, I, A, R_one, R_two, R_three, VSd, VR1d, VR2d,
-                        VR3d, VSdd, VR1dd, VR2dd, VR3dd, VSd_alt, VR1d_alt,
-                        VR2d_alt, VR3d_alt, VSdd_alt, VR1dd_alt, VR2dd_alt,
-                        VR3dd_alt, C, B, Doses, totInc),
-        names_sep = ""
-      ) %>%
-      select(-.id) %>%
-      as.matrix() %>%
-      t()
-
-    # Get the quantiles for the Susceptible pool from the simulations
-    results[["SQuants"]] <- h3Spat_sims %>%
-      rename(dep = unitname) %>%
-      group_by(dep, time) %>%
-      summarise(
-        Q025 = quantile(S + VSd + VSdd + VSd_alt + VSdd_alt, probs = 0.025),
-        Q50  = quantile(S + VSd + VSdd + VSd_alt + VSdd_alt, probs = 0.5),
-        Q975 = quantile(S + VSd + VSdd + VSd_alt + VSdd_alt, probs = 0.975)
-      ) %>%
-      ungroup() %>%
-      rowwise() %>%
-      mutate(
-        unit_num = which(h3@unit_names == dep),
-        date = lubridate::date_decimal(time),
-        Q025 = Q025 / h3@params[paste0("H", unit_num)],
-        Q50  = Q50  / h3@params[paste0("H", unit_num)],
-        Q975 = Q975 / h3@params[paste0("H", unit_num)]
-      ) %>%
-      filter(
-        date >= as.Date(
-          lubridate::round_date(lubridate::date_decimal(h3@t0), unit = 'day')
-        )
-      ) %>%
-      mutate(date = as.Date(lubridate::round_date(date, unit = 'day')))
-
-    results
-  },
-  seed = 834687
-)
-
-# Get the output of the "bake" call for plotting.
-h3Spat_quants <- init_conditions_results[["SQuants"]]
-
-# This bake call uses the ending hidden states from above to forecast cholera
-# cases in the future under the V0 vaccination scenario
-mod3_V0_initcond_res <- bake(
-  file = '../model3/si/vacc_sim_from_init.rds', {
-    h3_V0_res <- list()
-
-    # Get "noVacc" scenario parameters
-    h3_S0_par <- get_model3_vacc_scenario_params(scenario = "noVacc")
-    h3Spat <- h3
-    p3 <- best_params
-    coef(h3Spat) <- p3
-    coef(h3Spat)[names(h3_S0_par)] <- h3_S0_par
-
-    # store the end states that will be used as initial values
-    h3_bpf_end_states <- init_conditions_results$end_conditions
-
-    # Using parameters and end state values, obtain projections
-    h3_V0_sims <- project_from_filter(
-      h3Spat, end_states = h3_bpf_end_states, covarGen = project_rain,
-      nsims = 1000, seed = 184697
-    )
-
-    # Store nationally aggregated results
-    h3_V0_res$mod3_V0_sims <- agg_mod3_sims(h3_V0_sims)
-    h3_V0_res$mod3_V0_probs <- get_elimProbs(h3_V0_sims, model = 3)
-
-    # Get quantiles of observed cases for each department and time
-    h3_V0_res$UnitQuants <- h3_V0_sims %>%
-      mutate(
-        date = as.Date(  # Get date object from decimal time
-          lubridate::round_date(lubridate::date_decimal(time), unit = 'day')
-        )
-      ) %>%
-      group_by(unitname, date) %>%
-      summarize(
-        Q025 = quantile(cases, probs = 0.025),
-        Q50 = quantile(cases, probs = 0.5),
-        Q975 = quantile(cases, probs = 0.975)
-      ) %>%
-      ungroup() %>%
-      rename(dep = unitname)
-
-    # Get quantiles of susceptible individuals for each department and time
-    h3_V0_res$SQuants <- h3_V0_sims %>%
-      mutate(
-        date = as.Date(
-          lubridate::round_date(lubridate::date_decimal(time), unit = 'day')
-        )
-      ) %>%
-      group_by(unitname, date) %>%
-      mutate(
-        # Get the population for each unit
-        pop = h3@params[paste0("H", which(first(unitname) == unit_names(h3)))]
-      ) %>%
-      summarize(  # Calculate the quantiles
-        Q025 = quantile(
-          (S + VSd + VSdd + VSdd_alt + VSd_alt) / pop,
-          probs = 0.025
-        ),
-        Q50 = quantile(
-          (S + VSd + VSdd + VSdd_alt + VSd_alt) / pop,
-          probs = 0.5
-        ),
-        Q975 = quantile(
-          (S + VSd + VSdd + VSdd_alt + VSd_alt) / pop,
-          probs = 0.975
-        )
-      ) %>%
-      ungroup() %>%
-      rename(dep = unitname)
-
-    h3_V0_res
-  }
-)
-
-# This result was already used in the manuscript; it doesn't need to
-# be recalculated.
-mod3_V0_res <- readRDS("../model3/run_level_3/mod3_V0_res.rds")
-@
-
-\begin{figure}[ht]
-<<h3SusFig, echo=FALSE, message=FALSE, fig.height=3>>=
-ggplot() +
-  # Sims from initial condition
-  geom_line(data = h3Spat_quants, aes(x = date, y = Q50), col = 'red') +
-  geom_ribbon(
-    data = h3Spat_quants,
-    aes(x = date, ymin = Q025, ymax = Q975),
-    fill = 'red', alpha = 0.5
-  ) +
-  # projections from initial condition, using filtered sims function
-  geom_line(data = mod3_V0_initcond_res$SQuants, aes(x = date, y = Q50), col = 'red') +
-  geom_ribbon(
-    data = mod3_V0_initcond_res$SQuants,
-    aes(x = date, ymin = Q025, ymax = Q975),
-    fill = 'red', alpha = 0.5
-  ) +
-  # Sims from filtered distribution
-  geom_line(
-    data = bpf_final3_out$Sus_df,
-    aes(x = as.Date(lubridate::date_decimal(time)), y = Q50),
-    col = 'blue'
-  ) +
-  geom_ribbon(
-    data = bpf_final3_out$Sus_df,
-    aes(x = as.Date(lubridate::date_decimal(time)), ymin = Q025, ymax = Q975),
-    fill = 'blue', alpha = 0.5
-  ) +
-  # Projections from filtered distribution
-  geom_line(data = mod3_V0_res$SQuants, aes(x = date, y = Q50), col = 'blue') +
-  geom_ribbon(
-    data = mod3_V0_res$SQuants,
-    aes(x = date, ymin = Q025, ymax = Q975),
-    fill = 'blue', alpha = 0.5
-  ) +
-  geom_vline(xintercept = max(h3Spat_quants$date), linetype = 'dashed') +
-  facet_wrap(~dep, scales = 'free_y', nrow = 2) +
-  theme_bw() +
-  theme(axis.title.x = element_blank(),
-        axis.text.x = element_text(angle = -45, hjust = 0)) +
-  ylab("Total Susceptible (%)")
-@
-\caption{\label{fig:h3Sus}Percentage of individuals that are in the susceptible compartment.
-Simulations from Model~3 under initial conditions are displayed in red; simulations from the filtering distribution of model are displayed in blue.
-The dashed line represents the end of the observed data.}
-\end{figure}
-
-\clearpage
-
 \section{Replication of \cite{lee20}}
@@
 
-<<Simulate Lee Model 1, echo=FALSE, cache=TRUE>>=
-h1_epi <- haiti1()
-N_LEE_SIMS <- 20
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{figure}
 
-registerDoRNG(18599687)
-foreach(
-  i = 1:length(h1LeeStartsEpi),
-  .combine = rbind
-) %dopar% {
-  guess <- h1LeeStartsEpi[i, ]
-  sims <- simulate(h1_epi, nsim = N_LEE_SIMS, format = 'data.frame',
-                   params = guess)  # 20 sims for each set of parameters
-  sims$param_set <- i
-  sims %>%
-    dplyr::select(.id, week, cases, param_set)
-} -> all_sims
-@
+{\centering \includegraphics[width=\maxwidth]{figure/PlotEpiDist-1}
 
-<<PlotEpiDist, echo=FALSE, fig.cap="Compare to Figure S8 of Lee et al. (2020) supplement.", fig.width=7, fig.height=7, fig.align='center'>>=
-GGally::ggpairs(
-  h1LeeStartsEpi,
-  columns = c("rho", "tau", "beta1", "nu", "loglik")
-)
-@
+}
 
-<<SimulateEpi, echo=FALSE, include=FALSE, message=FALSE, cache=TRUE>>=
-h1_end <- haiti1(period = "endemic")
+\caption[Compare to Figure S8 of Lee et al]{Compare to Figure S8 of Lee et al. (2020) supplement.}\label{fig:PlotEpiDist}
+\end{figure}
 
-true_agg_cases <- haitiCholera %>%
-  select(-report) %>%
-  pivot_longer(
-    data = .,
-    cols = -date_saturday,
-    values_to = 'cases',
-    names_to = 'dep'
-  ) %>%
-  mutate(
-    date = as.Date(date_saturday),
-    dep = gsub("\\.", "_", dep)
-  ) %>%
-  mutate(
-    dep = case_when(dep == "Grand_Anse" ~ "Grande_Anse", TRUE ~ dep)
-  ) %>%
-  tidyr::pivot_wider(
-    data = .,
-    id_cols = c(date),
-    names_from = dep,
-    values_from = cases,
-    names_prefix = 'cases_'
-  ) %>%
-  mutate(
-    ReportedAll = cases_Artibonite + cases_Centre +
-      cases_Grande_Anse + cases_Nippes + cases_Nord +
-      cases_Nord_Est + cases_Ouest + cases_Sud +
-      cases_Sud_Est + cases_Nord_Ouest
-  )
-@
+\end{knitrout}
 
-<<FitEnd, echo=FALSE, message=FALSE, include=FALSE>>=
-foreach(
-  i = 1:nrow(h1LeeStartsEnd),
-  .combine = rbind
-) %dopar% {
-  r_params <- h1LeeStartsEnd[i, ][names(coef(h1_end))]
-  coef(h1_end) <- r_params
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{figure}
 
-  sims <- simulate(
-    h1_end, nsim = N_LEE_SIMS,
-    format = 'data.frame',
-    params = r_params
-  )
-  sims$param_set <- i
-  sims %>%
-    dplyr::select(.id, week, cases, param_set)
-} -> end_sims
+{\centering \includegraphics[width=\maxwidth]{figure/plotEndParams-1}
 
-@
+}
 
-<<plotEndParams, fig.width=7, fig.height=7, fig.align='center', fig.cap="Bivariate relationships between variables after fitting endemic period.
+\caption[Bivariate relationships between variables after fitting endemic period]{Bivariate relationships between variables after fitting endemic period.
 Compare to S9 of Lee et al.
-(2020) supplement.", cache=TRUE, echo=FALSE>>=
-GGally::ggpairs(
-  h1LeeStartsEnd,
-  columns = c("rho", "tau", "beta1", "nu", "loglik")
-)
-@
+(2020) supplement.}\label{fig:plotEndParams}
+\end{figure}
 
-<<calcMod1LeeLikes, echo=FALSE, include=FALSE, message=FALSE>>=
-h1_epi_evals <- readRDS("../model1/run_level_3/lee1_epi_evals.rds")
-h1_end_evals <- readRDS("../model1/run_level_3/lee1_end_evals.rds")
+\end{knitrout}
 
-mod1_all_likes <- dplyr::inner_join(
-  x = h1_epi_evals,
-  y = h1_end_evals,
-  by = "parid"
-  ) %>%
-  dplyr::mutate(
-    joint_ll = pfLL.x + pfLL.y
-  )
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{figure}
 
-# Approximate the likelihood of the Lee et al (2020) version of Model 1 by
-# using the parameter sets that resulted in the maximum likelihood.
-mod1_lee_ll <- max(mod1_all_likes$joint_ll, na.rm = TRUE)
-@
+{\centering \includegraphics[width=\maxwidth]{figure/plotMod1Sims-1}
 
-<<plotMod1Sims, echo=FALSE, fig.cap=paste0("Simulations using parameter sets that were generated by running source code provided by Lee et al (2020).
-Compare to Figure S7 of \\cite{lee20sup}.
-The upper bound for the likelihood of this model is ", round(mod1_lee_ll), "."), cache=TRUE, fig.width=6.5, fig.height=3.75>>=
+}
 
-quants <- all_sims %>%
-  filter(!param_set %in% c(1, 2)) %>%  # Remove parameter sets inconsistent with Lee et al (2020).
-  group_by(week) %>%
-  summarize(
-    q025 = quantile(cases, probs = 0.025, na.rm = TRUE),
-    q50  = quantile(cases, probs = 0.500, na.rm = TRUE),
-    q975 = quantile(cases, probs = 0.975, na.rm = TRUE)
-  ) %>%
-  ungroup() %>%
-  mutate(date = lubridate::ymd("2010-10-16") + lubridate::weeks(week))
+\caption[Simulations using parameter sets that were generated by running source code provided by Lee et al (2020)]{Simulations using parameter sets that were generated by running source code provided by Lee et al (2020).
+Compare to Figure S7 of \cite{lee20sup}.
+The upper bound for the likelihood of this model is -3031.}\label{fig:plotMod1Sims}
+\end{figure}
 
-quants_end <- end_sims %>%
-  group_by(week) %>%
-  summarize(
-    q025 = quantile(cases, probs = 0.025, na.rm = TRUE),
-    q50  = quantile(cases, probs = 0.500, na.rm = TRUE),
-    q975 = quantile(cases, probs = 0.975, na.rm = TRUE)
-  ) %>%
-  ungroup() %>%
-  mutate(date = lubridate::ymd("2010-10-16") + lubridate::weeks(week))
-
-ggplot() +
-  geom_point(data = true_agg_cases, aes(x = date, y = ReportedAll)) +
-  geom_line(data = quants, aes(x = date, y = q50), col = '#0b8a2b', size = 1) +
-  geom_ribbon(
-    data = quants,
-    aes(x = date, ymin = q025, ymax = q975),
-    alpha = 0.5, fill = '#0b9c30'
-  ) +
-  geom_line(data = quants_end, aes(x = date, y = q50), col = '#882871') +
-  geom_ribbon(data = quants_end, aes(x = date, ymin = q025, ymax = q975), alpha = 0.5, fill = '#882871') +
-  ylab("Reported, Symptomatic Incidence") +
-  scale_x_date(date_breaks = "1 year") +
-  theme(axis.title.x = element_blank(),
-        axis.text.x = element_text(hjust = 1, angle = 45))
-@
+\end{knitrout}
 
@@
 
-<<Model2Replication, cache=TRUE, echo=FALSE>>=
-h2_epi <- haiti2()
-h2_end <- haiti2(region = 'after')
+\begin{figure}[!h]
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
 
-h2_epi_traj <- trajectory(h2_epi, format = 'data.frame')
-h2_end_traj <- trajectory(h2_end, format = 'data.frame')
+{\centering \includegraphics[width=\maxwidth]{figure/plotModel2Rep-1}
 
-h2_epi_traj$Ctotal <- rowSums(h2_epi_traj[, paste0("C", 1:10)]) * 0.2
-h2_end_traj$Ctotal <- rowSums(h2_end_traj[, paste0("C", 1:10)]) * 0.2
+}
 
-dep_plot_df <- haitipkg::haitiCholera %>%
-  select(-report) %>%
-  tidyr::pivot_longer(
-    data = .,
-    cols = -date_saturday,
-    values_to = 'cases',
-    names_to = 'dep'
-  ) %>%
-  mutate(
-    date = as.Date(date_saturday),
-    dep = gsub("\\.", "_", dep)
-  ) %>%
-  mutate(
-    dep = case_when(dep == "Grand_Anse" ~ "Grande_Anse", TRUE ~ dep)
-  )
-
-true_agg_cases <- dep_plot_df %>%
-  tidyr::pivot_wider(
-    data = .,
-    id_cols = c(date),
-    names_from = dep,
-    values_from = cases,
-    names_prefix = 'cases_'
-  ) %>%
-  mutate(
-    ReportedAll = cases_Artibonite + cases_Centre +
-      cases_Grande_Anse + cases_Nippes + cases_Nord +
-      cases_Nord_Est + cases_Ouest + cases_Sud +
-      cases_Sud_Est + cases_Nord_Ouest
-  ) %>%
-  select(date, ReportedAll)
-
-h2_epi_traj$date <- as.Date(
-  lubridate::round_date(
-    lubridate::date_decimal(h2_epi_traj$year),
-    unit = 'day'
-  )
-)
-
-h2_end_traj$date <- as.Date(
-  lubridate::round_date(
-    lubridate::date_decimal(h2_end_traj$year),
-    unit = 'day'
-  )
-)
-
-all_traj <- bind_rows(
-  select(h2_epi_traj, date, Ctotal),
-  select(h2_end_traj, date, Ctotal)
-)
-
-projections <- left_join(
-  x = true_agg_cases,
-  y = all_traj,
-  by = 'date'
-)
-@
-
-\begin{figure}[ht]
-<<plotModel2Rep, cache=TRUE, echo=FALSE, fig.width=6.5, fig.height=4.5>>=
-gg_epi <- projections %>%
-  filter(date < as.Date("03-01-2014", format = '%m-%d-%Y')) %>%
-  ggplot(aes(x = date)) +
-  geom_line(aes(y = ReportedAll), col = 'black', linetype = 'dashed') +
-  geom_line(aes(y = Ctotal), col = 'red', size = 1) +
-  theme_bw() +
-  scale_y_continuous(limits = c(0, 30000), breaks = seq(0, 30000, 5000)) +
-  theme(axis.title.x = element_blank()) +
-  ylab("Reported number of cases")
-
-gg_end <- projections %>%
-  filter(date >= as.Date("03-01-2014", format = '%m-%d-%Y')) %>%
-  ggplot(aes(x = date)) +
-  geom_line(aes(y = ReportedAll), col = 'black', linetype = 'dashed') +
-  geom_line(aes(y = Ctotal), col = 'red', size = 1) +
-  theme_bw() +
-  scale_y_continuous(limits = c(0, 7000), breaks = seq(0, 7000, 1000)) +
-  theme(axis.title.x = element_blank()) +
-  ylab("Reported number of cases") +
-  scale_x_date(date_breaks = "1 year", date_labels = "%Y")
-
-cowplot::plot_grid(gg_epi, gg_end, ncol = 1)
-@
+\end{knitrout}
 \caption{\label{fig:mod2rep}
@@
 
-<<Model3Replication, echo=FALSE, cache=TRUE>>=
-NSIM <- 500
+\begin{figure}[!h]
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
 
-plot_order <- c(
-  'Artibonite',
-  'Sud_Est',
-  'Nippes',
-  'Nord_Est',
-  'Ouest',
-  'Centre',
-  'Nord',
-  'Sud',
-  'Nord_Ouest',
-  'Grande_Anse'
-)
+{\centering \includegraphics[width=\maxwidth]{figure/PlotMod3Rep-1}
 
-dep_labeller <- as_labeller(
-  c(
-    'Artibonite' = 'Artibonite',
-    'Sud_Est' = 'Sud-Est',
-    'Sud.Est' = 'Sud-Est',
-    'Nippes' = 'Nippes',
-    'Nord_Est' = 'Nord-Est',
-    'Nord.Est' = 'Nord-Est',
-    'Ouest' = 'Ouest',
-    'Centre' = 'Centre',
-    'Nord' = 'Nord',
-    'Sud' = 'Sud',
-    'Nord_Ouest' = 'Nord-Ouest',
-    'Nord.Ouest' = 'Nord-Ouest',
-    'Grande_Anse' = 'Grand\'Anse',
-    'Grand.Anse' = 'Grand\'Anse'
-  )
-)
+}
 
-lee3 <- lee3_spatPomp()
-lee3_sims_quants <- bake("../model3/si/lee3Sims.rds", {
-
-  # Simulate from lee et al 2020 model
-  sims_temp <- simulate(
-    lee3, nsim = NSIM,
-    seed = 321, format = 'data.frame'
-  )
-
-  # Get desired quantiles of reported cases from simulations
-  quants <- sims_temp %>%
-    dplyr::group_by(unitname, time) %>%
-    dplyr::summarise(
-      q05 = quantile(cases, 0.025, na.rm = T),
-      mean = mean(cases, na.rm = T),
-      q50 = quantile(cases, 0.5, na.rm = T),
-      q95 = quantile(cases, 0.975, na.rm = T)
-    ) %>%
-    dplyr::ungroup() %>%
-    dplyr::mutate(
-      date = lubridate::date_decimal(time),
-      date = as.Date(lubridate::round_date(date))
-    ) %>%
-    dplyr::rename(dep = unitname)
-
-  # Remove cases from Ouest that were not considered by Lee et al. (2020)
-  quants %>%
-    dplyr::filter(
-      !(dep == "Ouest" & date < '2017-06-10')
-    )
-
-})
-
-# Get the subset of cholera cases that match simulations from Lee et al. (2020) model.
-sub_cases <- haitiCholera %>%
-  dplyr::mutate(date = as.Date(date_saturday)) %>%
-  dplyr::select(-report, -date_saturday) %>%
-  dplyr::filter(lubridate::decimal_date(date) >= min(lee3@times)) %>%
-  tidyr::pivot_longer(
-    data = .,
-    cols = -date,
-    names_to = "dep",
-    values_to = "cases"
-  ) %>%
-  dplyr::mutate(
-    dep = dplyr::case_when(
-      dep == "Grand.Anse" ~ "Grande_Anse",
-      dep == "Nord.Est" ~ "Nord_Est",
-      dep == "Nord.Ouest" ~ "Nord_Ouest",
-      dep == "Sud.Est" ~ "Sud_Est",
-      TRUE ~ dep
-    )
-  )
-@
-
-\begin{figure}[ht]
-<<PlotMod3Rep, echo=FALSE, message=FALSE, fig.height=3.75>>=
-ggplot() +
-  # Region for upper and lower simulation bounds
-  geom_ribbon(
-    data = lee3_sims_quants,
-    aes(x = date, ymin = q05, ymax = q95),
-    fill = 'darkblue', alpha = 0.5
-  ) +
-  # Median line
-  geom_line(
-    data = lee3_sims_quants,
-    aes(x = date, y = q50),
-    col = 'darkblue'
-  ) +
-  # Observed data
-  geom_point(
-    data = dplyr::filter(sub_cases, cases < 500),
-    aes(x = date, y = cases), size = 0.2
-  ) +
-  facet_wrap(
-    ~factor(dep, levels = plot_order),
-    nrow = 2, labeller = dep_labeller
-  ) +
-  labs(y = 'Reported Cholera Cases') +
-  scale_x_date(date_breaks = "6 months", date_labels = "%Y-%m") +
-  theme(
-    axis.text.x = element_text(angle = 40, hjust = 1, vjust = 1, size = 7),
-    axis.title.x = element_blank()
-  )
-@
+\end{knitrout}
 \caption{\label{fig:mod3rep}
@@
 \end{figure}
-
-<<lee3Cleanup, echo=FALSE, include=FALSE, message=FALSE>>=
-rm(lee3, lee3_sims_quants, sub_cases)
-gc()
-@
-
-\section{Forecasting with parameter uncertainty}
-
-Let $f_{Y_{1:N}}(y_{1:N} | \theta)$ denote the pdf of the model under consideration, were $\theta$ is a parameter vector that indexes the model.
-Furthermore, denote the observed data as $y_{1:N}^*$.
-Because the uncertainty in just a single parameter can lead to drastically different forecasts \citep{saltelli20},
-parameter uncertainty should be considered when obtaining model forecasts when the goal is to influencing policy.
-In a Bayesian modelling paradigm, the most natural way to account for parameter uncertainty in model forecasts is to suppose that $\theta$ comes from a distribution $f_{\Theta}$, and then to obtain $J$ forecasts from the model where each forecast is obtained using parameters drawn from the posterior distribution $\Theta_{1:J} \mid Y_{1:N} = y_{1:N}^* \sim f_{\Theta}\big(\theta | Y_{1:N} = y_{1:N}^*\big)$.
-
-When frequentist methods are used, however, there does not exist a posterior distribution from which one could sample.
-A common approach could be to obtain a weighted average of the simulations from various models \citep{hoeting99}, but this can be problematic when forecasts from each model are very different from each other \citep{grueber11}.
-A similar approach that has been taken \citep{king15} is to obtain model forecasts using multiple sets of parameter values and then sample from the resulting forecasts using weights proportional to the corresponding likelihoods of the parameter values.
-This approach could be considered as empirical Bayes, as it is equivalent to using a discrete uniform prior where the set of values in the prior distribution is determined by a stochastic routine applied to the observed data, as discussed below.
-
-For each $i \in 1:k$, let $\Theta_i$ be a random vector of model parameters.
-Then, letting $\Theta$ denote the true model parameters, we endow the set $\{\Theta_1, \Theta_2, \ldots, \Theta_k\}$ with a discrete uniform distribution, such that $P\big(\Theta = \Theta_i\big) = \frac{1}{K}$ for all values $i \in \seq{1}{k}$.
-Using this as a prior distribution, the posterior distribution of $\Theta | Y_{1:N} = y_{1:N}^*$ can be expressed as:
-$P\big(\Theta = \Theta_k | Y_{1:N} = y_{1:N}^*\big) = \frac{f_{Y_{1:N}}(y_{1:N}^*| \Theta_k)}{\sum_{l = 1}^K f_{Y_{1:N}}(y_{1:N}^*| \theta_l)}$.
-In a standard empirical Bayes analysis, the values $\Theta_1, \ldots, \Theta_k$ of the prior distribution would be chosen using the observed data, resulting in a posterior distribution that weighs the prior parameter vectors proportional to their corresponding likelihoods.
-Instead, we choose $\Theta_i$ to be the output of a stochastic routine applied to the observed data by setting $\Theta_i$ to be the output of an iterated filtering algorithm.
-In practice, because the likelihood maximization routines of iterated filtering methods are stochastic, it is common to run the iterated filtering method multiple times for each model in order to obtain a maximum likelihood estimate for model parameters.
-This results in a natural set of parameters near the MLE that could be used as the discrete prior distribution.
-
-\section{Translating to \citet{lee20} notation}
-
-Since the models of \citet{lee20} were developed independently, the choice of notation varies inconsistently between models.
-For our reanalysis, we rename parameters to provide a unified notation facilitating comparison between models.
-Table~\ref{tab:translate} maps this notation back to the original notations, for reference.
-
-\begin{table}
-  \begin{center}
-  \begin{tabular}{|c|c|c|c|c|}\hline
-    \multirow{2}{*}{Parameter} & Our & \multicolumn{3}{c|}{Lee et al. (2020a)} \\\cline{3-5}
-     & Notation & 1 & 2 & 3 \\
-    \hline
-    \hline
-    % Table begins here
-    Reporting Rate & $\reportRate$ & $\rho$ & $\rho$ & $\epsilon_1, \epsilon_2$ \\\hline
-    Mixing Coefficient & $\mixExponent$ & $\nu$ & --- & --- \\\hline
-    Measurement Over-Dispersion & $\obsOverdispersion$ & $\tau$ & --- & $p$ \\\hline
-    Birth Rate & $\muBirth$ & $\mu$ & --- & --- \\\hline
-    Natural Mortality Rate & $\muDeath$ & $\delta$ & --- & $\mu$ \\\hline
-    Cholera Mortality Rate & $\choleraDeath$ & --- & --- & $\alpha$ \\\hline
-    Latent Period & $1/\muEI$ & $1/\sigma$ & $1/\gamma_E$ & --- \\\hline
-    Recovery Rate & $\muIR$ & $\gamma$ & $\gamma$ & $\gamma$ \\\hline
-    Loss of Immunity & $\muRS$ & $\alpha$ & $\sigma$ & $\rho$ \\\hline
-    Symptomatic Ratio & $\symptomFrac$ & $1 - \theta_0$ & $k$ & $\sigma$ \\\hline
-    Asymptomatic Relative Infectiousness & $\asymptomRelativeInfect$ & $\kappa$ & $red_\beta$ & --- \\\hline
-    Human-to-Water Shedding & $\Wshed$ & --- & $\mu$ & $\theta_I$ \\\hline
-    Asymptomatic Relative Shedding & $\asymptomRelativeShed$ & --- & $red_\mu$ & $\theta_A/\theta_I$ \\\hline
-    Seasonal Amplitude & $\seasAmplitude$ & --- & $\alpha_s$ & $\lambda$ \\\hline
-    Transmission & $\transmission$ & $\beta$ & $\beta$ & $c$ \\\hline
-    Water-to-Human & $\Wbeta{}$ & --- & $\beta_W$ & $\beta$ \\\hline
-    Bacteria Mortality & $\Wremoval$ & --- & $\delta$ & $\mu_\beta$ \\\hline
-    Vaccination Efficacy & $\vaccineEfficacy$ & $\theta_{vk}$ & $\theta_1, \theta_2, \theta_{1_5}, \theta_{2_5}$ & $\eta_{1d}, \eta_{2d}$ \\\hline
-    Process Over-dispersion & $\sigmaProc$ & --- & --- & $\sigma_w$\\\hline
-    % Half Saturation Constant & $\Wsat$ & --- & $Sat$ & $K$ \\\hline
-  \end{tabular}
-  \end{center}
-  \caption{
-  \label{tab:translate}Translations between our common notation and notation used by \cite{lee20}
-  }
-\end{table}
-
-\clearpage
-
-\bibliography{../bib-haiti}
-
-\end{document}

### 2023-09-05 si/calibrateMod3.Rnw
new Makefile for creating supplement material

@@
 
-\begin{figure}[!h]
+\begin{figure}[!ht]
 <<h3UnitLikes, fig.height=3, fig.width=5.5, echo=FALSE>>=
@@
 
-\begin{figure}[!h]
+\begin{figure}[!ht]
 <<CondLLFig, echo=FALSE, message=FALSE, fig.height=3>>=
@@
 
-\begin{figure}[!h]
+\begin{figure}[!ht]
 <<finalCondLLFig, echo=FALSE, message=FALSE, fig.height=3>>=
@@
 
-\begin{figure}[!h]
+\begin{figure}[!ht]
 <<finalUnitLLFig, fig.height=3, fig.width=5.5, echo=FALSE>>=
@@
 
-\begin{figure}[!h]
+\begin{figure}[!ht]
 <<h3SusFig, echo=FALSE, message=FALSE, fig.height=3>>=

### 2023-09-05 si/calibrateMod3Out.tex
new Makefile for creating supplement material

--- 2023-08-31 si/si.Rnw
@@
-\documentclass[11pt]{article}
-
-\pdfminorversion=4
-
-\newcommand\secTitleSpace{\hspace{3mm}}
-
-<<packages, include=FALSE>>=
-library(tidyverse)
-library(pomp)
-library(haitipkg)
-library(spatPomp)
-library(doParallel)
-library(doRNG)
-
-cores <-  as.numeric(Sys.getenv('SLURM_NTASKS_PER_NODE', unset = NA))
-if(is.na(cores)) cores <- detectCores()
-# cores <- 20
-registerDoParallel(cores)
-theme_set(theme_bw())
-
-if (!dir.exists("../model3/si")) {
-  dir.create("../model3/si", recursive = TRUE)
-}
-@
-
-<<set-opts, include=FALSE>>=
-options(
-  scipen = 2,
-  help_type = "html",
-  stringsAsFactors = FALSE,
-  continue = "+  ",
-  width = 70,
-  useFancyQuotes = FALSE,
-  reindent.spaces = 2,
-  xtable.comment = FALSE
-)
-@
-
-<<knitr-opts, include=FALSE, purl=FALSE>>=
-library(knitr)
-opts_knit$set(concordance=TRUE)
-opts_chunk$set(
-  progress=TRUE,prompt=TRUE,highlight=FALSE,
-  tidy=TRUE,
-  tidy.opts=list(
-    keep.blank.line=FALSE
-  ),
-  comment="",
-  warning=FALSE,
-  message=FALSE,
-  error=TRUE,
-  echo=FALSE,
-  cache=FALSE,
-  strip.white=TRUE,
-  results="markup",
-  fig.path="figure/",
-  fig.lp="fig:",
-  fig.align="center",
-  fig.show="asis",
-  dev="pdf",
-  dev.args=list(
-    bg="transparent",
-    pointsize=9
-  )
-)
-
-myround <- function (x, digits = 1) {
-  # taken from the broman package
-  if (digits < 1)
-    stop("This is intended for the case digits >= 1.")
-  if (length(digits) > 1) {
-    digits <- digits[1]
-    warning("Using only digits[1]")
-  }
-  tmp <- sprintf(paste("%.", digits, "f", sep = ""), x)
-  zero <- paste0("0.", paste(rep("0", digits), collapse = ""))
-  tmp[tmp == paste0("-", zero)] <- zero
-  tmp
-}
-
-@
-
-%% copied from pnas-new.cls
-\usepackage{amsmath,amsfonts,amssymb,graphicx}
-
-%%%% parameters %%%%%%%%%%5
-\newcommand\Wsat{W_{\mathrm{sat}}}
-\newcommand\muIR{\mu_{IR}}
-\newcommand\muEI{\mu_{EI}}
-\newcommand\transmission{\beta}
-\newcommand\seasAmplitude{a}
-\newcommand\rainfallExponent{r}
-\newcommand\muRS{\mu_{RS}}
-\newcommand\vaccineEfficacy{\theta}
-\newcommand\muBirth{\mu_S}
-\newcommand\muDeath{\delta}
-\newcommand\choleraDeath{\delta_{C}}
-\newcommand\symptomFrac{f}
-\newcommand\asymptomRelativeInfect{\epsilon}
-\newcommand\asymptomRelativeShed{\epsilon_{W}}
-\newcommand\Wbeta[1]{\beta_{W#1}}
-\newcommand\Wremoval{\delta_W}
-\newcommand\Wshed{\mu_W}
-\newcommand\mixExponent{\nu}
-\newcommand\sigmaProc{\sigma_{\mathrm{proc}}}
-\newcommand\reportRate{\rho}
-\newcommand\obsOverdispersion{\psi}
-\newcommand\phaseParm{\phi}
-\newcommand\transmissionTrend{\zeta}
-\newcommand\Binit{\xi}
-\newcommand\vaccClass{Z}
-\newcommand\vaccCounter{z}
-\newcommand\modelCounter{m}
-\newcommand\Whur[1]{\beta_{W#1}^{hm}}
-\newcommand\hHur[1]{h_{#1}^{hm}}
-\newcommand\Iinit{I_{0,0}}
-\newcommand\Einit{E_{0, 0}}
-\newcommand\reportChange{\delta}
-
-\newcommand\missing{\hspace{6mm} ---}
-\newcommand\fixed{\color{blue}}
-\newcommand\demography{\bullet}
-
-\newcommand\mytitle{Informing policy via dynamic models: Cholera in Haiti}
-\newcommand\comma{{\hspace{-0.25mm},\hspace{-0.2mm}}}
-
-%%%%%%%%%%%%%%%% START OF USER DEFINITIONS %%%%%%%%%%%%%%%5
-
-\newcommand\siOnly[1]{} %% redefined for the supplement
-\newcommand\msOnly[1]{#1} %% redefined for the supplement
-
-\newcommand\code[1]{\texttt{#1}}
-\usepackage{xurl}
-% \usepackage[breaklinks]{hyperref}
-\usepackage{makecell}
-\usepackage{multirow}
-\usepackage{multicol}
-\usepackage[ruled,noline,linesnumbered]{algorithm2e}
-% \usepackage{setspace}
-
-%% EDITING MACROS
-\usepackage{color}
-\usepackage[normalem]{ulem}% to use \sout in feedback commands
-\usepackage{bm}
-% orange for EI
-\definecolor{orange}{rgb}{1,0.5,0}
-\newcommand\ei[2]{\sout{#1} \textcolor{orange}{#2}}
-\newcommand\eic[1]{\textcolor{orange}{[#1]}}
-% green for JW
-\definecolor{green}{rgb}{0,0.5,0}
-\newcommand\jw[2]{\sout{#1} \textcolor{green}{#2}}
-\newcommand\jwc[1]{\textcolor{green}{[#1]}}
-% purple for JJ
-\definecolor{purple}{rgb}{0.5,0,1}
-\newcommand\jj[2]{\sout{#1} \textcolor{purple}{#2}}
-\newcommand\jjc[1]{\textcolor{purple}{[#1]}}
-% cyan for AR
-\definecolor{cyan}{rgb}{0,.5,.5}
-\newcommand\ar[2]{\sout{#1} \textcolor{cyan}{#2}}
-\newcommand\arc[1]{\textcolor{cyan}{#1}}
-% light brown for KT
-\definecolor{lightbrown}{rgb}{0.5,0.5,0}
-\newcommand\kt[2]{\sout{#1} \textcolor{lightbrown}{#2}}
-\newcommand\ktc[1]{\textcolor{lightbrown}{#1}}
-
-%% basic POMP definitions
-\newcommand\Xspace{{\mathbb X}}
-\newcommand\Yspace{{\mathbb Y}}
-\newcommand\Thetaspace{\R^{\Thetadim}}
-\newcommand\Xunitspace{{\scriptsize{\mathcal X}}}
-\newcommand\Yunitspace{{\tiny{\mathcal Y}}}
-\newcommand\hatTheta{\widehat{\Theta}}
-\newcommand\Xdim{{\mathrm{dim}}(\Xspace)}
-\newcommand\Ydim{{\mathrm{dim}}(\Yspace)}
-\newcommand\Thetadim{P}
-\newcommand\thetadim{p}
-\newcommand\timeSet{{\mathbb T}}
-\newcommand\data[1]{#1^*}
-
-\newcommand\unitSpecific{\hspace{0.1mm}\mathrm{us}}
-\newcommand\shared{\hspace{0.15mm}\mathrm{sh}}
-
-% for algorithm pseudocode
-\SetKwFor{For}{for}{:}{end}
-
-%% for particle filters
-\newcommand\unit{u}
-\newcommand\altUnit{\tilde u}
-\newcommand\Unit{U}
-\newcommand\rep{i}
-\newcommand\Rep{\mathcal{I}}
-\newcommand\island{\rep}
-\newcommand\Island{\Rep}
-%\newcommand\txtisland{replicate}
-\newcommand\altIsland{\tilde i}
-\usepackage[mathscr]{euscript}
-
-\renewcommand\time{n}
-\newcommand\myvec[1]{\boldsymbol{#1}}
-\newcommand\altTime{\tilde n}
-\newcommand\Time{N}
-\newcommand\Np{J}
-\newcommand\np{j}
-\newcommand\altNp{k}
-\newcommand\altAltNp{\tilde j}
-\newcommand\resampleIndex{r}
-\newcommand\unitWeight{w}
-
-%% for all iterated filtering algorithms
-\newcommand\Nit{M}
-\newcommand\nit{m}
-
-%% customized math macros
-\newcommand\prob{\mathbb{P}}
-\newcommand\E{\mathbb{E}}
-\newcommand\dd[1]{\mathrm{d}{#1}}
-\newcommand\given{{\,\vert\,}}
-\newcommand\equals{{{\,}={\,}}}
-\newcommand\myequals{\hspace{0.5mm}{=}\hspace{0.5mm}}
-\newcommand\myto{{\;:\;}}
-\newcommand\seq[2]{{#1}\!:\!{#2}}
-\newcommand\mydot{{\,\cdot\,}}
-\newcommand\cp[2]{N_{\mathrm{#1}\mathrm{#2}}}
-\newcommand\giventh{{\hspace{0.5mm};\hspace{0.5mm}}}
-\newcommand\normal{\mathcal{N}}
-\newcommand\argequals{{\,=\,}}
-\newcommand\lags{c}
-\newcommand\maxlag{\overline{c}}
-\newcommand\nlfList{C}
-\newcommand\bigO{\mathcal{O}}
-\newcommand\loglik{\lambda}
-\newcommand\loglikMC{\MC{\loglik}}
-\newcommand\R{\mathbb{R}}
-\newcommand\param{\,;}
-\newcommand\mycolon{{\hspace{0.6mm}:\hspace{0.6mm}}}
-\newcommand\MC[1]{#1^{\,\mbox{\tiny MC}}}
-\newcommand\EMC{{\E}}
-\newcommand\Var{\mathrm{Var}}
-\newcommand\var{\Var}
-\newcommand\Cov{\mathrm{Cov}}
-\newcommand\cov{\Cov}
-\newcommand\iid{\mathrm{iid}}
-\newcommand\dist{d}
-\newcommand\transpose{\top}
-
-\newcommand\gnsep{,}
-
-\def\lik{L}
-\def\loglik{\ell}
-
-%% THEOREM-LIKE ENVIRONMENTS
-\usepackage{amsthm}
-\newtheorem{prop}{Proposition}
-\newtheorem{theorem}{Theorem}
-\newtheorem{lemma}{Lemma}
-\newtheorem{example}{Example}
-\newtheorem{remark}{Remark}
-\newtheorem{cor}{Corrolary}
-\newtheorem{fact}{Fact}
-\newtheorem{assumption}{Assumption}
-\newtheorem{assumptionB}{Assumption}
-\renewcommand\theassumption{A\arabic{assumption}}
-\renewcommand\theassumptionB{B\arabic{assumptionB}}
-
-%% FOR PSEUDOCODE SETUP
-\newcommand\mystretch{\rule[-2mm]{0mm}{5mm} }
-\newcommand\asp{\hspace{4mm}}
-\newcommand\codeIndent{\hspace{4mm}}
-
-%%%%%%%%%%%%%%%% END OF USER DEFINITIONSS %%%%%%%%%%%%%%%
-
-\bibliographystyle{apalike}
-\usepackage{natbib}
-\usepackage{fullpage}
-
-\newcommand\myeqref[1]{(\ref{#1})}
-
-\renewcommand{\contentsname}{Supplementary Content}
-\renewcommand{\refname}{Supplementary References}
-\renewcommand\thefigure{S-\arabic{figure}}
-\renewcommand\thetable{S-\arabic{table}}
-\renewcommand\thesection{S\arabic{section}}
-\renewcommand\theequation{S\arabic{equation}}
-\renewcommand\theprop{S\arabic{prop}}
-\renewcommand\thelemma{S\arabic{lemma}}
-\renewcommand\thealgocf{S\arabic{algocf}}
-
-\setcounter{tocdepth}{1}
-
-\begin{document}
-
-\date{\today}
-\title{Supplement to ``{\mytitle}''}
-  \author{Jesse Wheeler, AnnaElaine L. Rosengart, Zhuoxun Jiang, \\ Kevin Tan, Noah Treutle and Edward L. Ionides
- \\ \hspace{.2cm}\\
-    Department of Statistics, University of Michigan\\
-}
-
-\newcommand{\blind}{1}
-
-\if1\blind
-{
-\maketitle
-}
-\fi
-
-\if0\blind
-{
-  \bigskip
-  \bigskip
-  \bigskip
-  \begin{center}
-    {\LARGE\bf \mytitle}
-\end{center}
-  \bigskip
-  \bigskip
-}\fi
-
-\tableofcontents
-
-\newpage
-
-\section{Markov chain and differential equation interpretations of compartment flow rates}
-
-In Sections 2.1, 2.2 and 2.3 of the main article, we define compartment models in terms of their flow rates.
-For a discrete population model, these rates define a Markov chain.
-For a continuous and deterministic model, the rates define a system of ordinary differential equations.
-Here, we add additional details to clarify the mapping from a collection of rate functions to a fully specified process.
-Our treatment follows \citet{breto09}.
-
-A general compartment model is a vector-valued process $X(t)=(X_1(t),\dots,X_c(t))$ denoting the (integer or real-valued) counts in each of $c$ compartments, where $t$ is any continuous value in the interval  $\left[t_0, \infty\right)$ for some real valued starting time $t_0$.
-The compartments may also have names, but to set up general notation we simply refer to them by their numerical index.
-The basic characteristic of a compartment model is that $X(t)$ can be written in terms of the flows $N_{ij}(t)$ from $i$ to $j$.
-A flow into compartment $i$ from outside the system is denoted by $N_{\demography i}$, and a flow out of the system from compartment $i$ is denoted by $N_{i\demography}$.
-We call $\demography$ a source/sink compartment, though it is an irregular compartment since $X_{\demography}(t)$ is not defined.
-These flows are required to satisfy a ``conservation of mass'' identity:
-\begin{equation}
-X_i(t)=X_i(t_0)+N_{\demography i}(t) - N_{i\demography}(t) + \sum_{j\neq
-i}N_{ji}(t)-\sum_{j\neq i}N_{ij}(t). \label{eq:conservation}
-\end{equation}
-Each {\em flow} $N_{ij}(t)$ is associated with a {\em rate} function
-$\mu_{ij}=\mu_{ij}(t,X(t))$, where we include the possibility that $i$ or $j$ takes value $\demography$.
-
-There are different ways to use a collection of rate functions to build a fully specified model.
-We proceed to describe the ones we use in this paper: via a system of ordinary differential equations (Sec.~\ref{subsec:ode}), a simple Markov counting system (Sec.~\ref{subsec:smcs}), and an overdispersed Markov counting system (Sec.~\ref{subsec:odmcs}).
-Other representations include stochastic differential equations driven by Gaussian noise or Gamma noise \citep{bhadra11}.
-
-\subsection{Ordinary differential equation (ODE) interpretation}
-\label{subsec:ode}
-
-A basic deterministic specification is
-\begin{equation}
-\label{eq:ode1}
-dN_{ij}/dt = \mu_{ij}\big(t,X(t) \big) X_i(t), \hspace{3mm} i\in \seq{1}{c}, \hspace{3mm} j\in \seq{1}{c} \cup\{\demography\}, \hspace{3mm} i\neq j,
-\end{equation}
-where $\mu_{ij}\big(t,X(t)\big)$ is called a per-capita rate or a unit rate.
-Flows into the system require special treatment since $X_i(t)$ in \myeqref{eq:ode1} is not defined for $i=\demography$.
-Instead, we specify
-\begin{equation}
-\label{eq:ode2}
-dN_{\demography i}/dt = \mu_{\demography i}\big(t,X(t) \big).
-\end{equation}
-This is the the interpretation and implementation used for Model~2 in our study.
-
-\subsection{Simple Markov counting system interpretation}
-\label{subsec:smcs}
-A continuous time Markov chain can be specified via its infinitesimal transition probabilities.
-A basic approach to this is to define
-\begin{eqnarray}
-\label{eq:smcs1}
-\prob\big[ N_{ij}(t+\delta)-N_{ij}(t)=0 \given X(t)\big]
- &=& 1-\delta \mu_{ij}\big(t,X(t)\big)X_i(t) + o(\delta),
-\\
-\label{eq:smcs2}
-\prob\big[ N_{ij}(t+\delta)-N_{ij}(t)=1 \given X(t)\big]
- &=& \delta \mu_{ij}\big(t,X(t)\big)X_i(t) + o(\delta),
-\end{eqnarray}
-for $i\in \seq{1}{c}$ and $j\in\seq{1}{c}\cup\{\demography\}$ with $i\neq j$.
-As with the ODE case, we need special attention for flows into the system, and we define
-\begin{eqnarray}
-\label{eq:smcs3}
-\prob\big[ N_{\demography i}(t+\delta)-N_{\demography i}(t)=0 \given X(t)\big]
- &=& 1-\delta \mu_{\demography i}\big(t,X(t)\big) + o(\delta),
-\\
-\label{eq:smcs4}
-\prob\big[ N_{\demography i}(t+\delta)-N_{\demography i}(t)=1 \given X(t)\big]
- &=& \delta \mu_{\demography i}\big(t,X(t)\big) + o(\delta).
-\end{eqnarray}
-Together with the initial conditions $X(0)$, equations \myeqref{eq:smcs1}--\myeqref{eq:smcs4} define a Markov chain.
-Each flow is a simple counting process, meaning a non-decreasing integer-valued process that only has jumps of size one.
-We therefore call the Markov chain a simple Markov counting system (SMCS).
-The infinitesimal mean of every flow is equal to its infinitesimal variance \citep{breto11} and so an SMCS is called equidispersed.
-We note that the special case of Model~1 used by \cite{lee20} (with $\sigmaProc = 0$) is an SMCS.
-To permit more general mean-variance relationships for a Markov counting system, we must permit jumps of size greater than one.
-The utility of overdispersed models, where the infinitesimal variance of the flow exceeds the infinitesimal mean, has become widely recognized \citep{stocks20,he10}.
-
-\subsection{Overdispersed Markov counting system interpretation}
-\label{subsec:odmcs}
-
-Including white noise in the rate function enables the possibility of an overdispersed Markov counting system \citep{breto11,breto09,he10}.
-Since rates should be non-negative, Gaussian noise is not appropriate and gamma noise is a convenient option that has found various applications \citep{romero-severson15, subramanian20}.
-Specifically, we consider a model given by
-\begin{equation}
-\label{eq:odmcs1}
-\mu_{ij}\big(t,X(t)\big) = \bar\mu_{ij}\big(t,X(t)\big) \, d\Gamma_{ij}(t)/dt,
-\end{equation}
-where $\Gamma_{ij}(t)$ is a stochastic process having independent gamma distributed increments, with
-\begin{equation}
-\label{eq:odmcs2}
-\E\big[\Gamma_{ij}(t)\big] = t, \quad \var\big[\Gamma_{ij}(t)\big] = \sigma_{ij}^2 t.
-\end{equation}
-Formally interpreting the meaning of \myeqref{eq:odmcs1} is not trivial, and we do so by constructing a Markov process $X(t)$ as the limit of the Euler scheme descrived in Section~\ref{sec:numerics}, below.
-Therefore, the numerical scheme in Sec.~\ref{sec:numerics} can be taken as a definition of the meaning of \myeqref{eq:odmcs1}.
-The Markov chain defined by the limit of this Euler scheme as the step size decreases is an overdispersed Markov counting system, with the possibility of instantaneous jumps of size greater than one \citep{breto11}.
-
-\section{Numerical solutions to compartment models}
-\label{sec:numerics}
-
-Models may be fitted and their implications assessed via numerical solutions (i.e., simulations) from the model equations.
-All the analyses we consider have this simulation-based property, known as plug-and-play or equation-free or likelihood-free.
-The numerical solutions to the model are arguably of more direct scientific interest than the exact solutions to the postulated equations.
-For ODE models, numerical methods are well studied and a standard numerical solution package such as \code{deSolve} in \code{R} is adequate for many purposes.
-For SMCS and ODMCS models, exact schemes are feasible when the number of events is small, which may be the case for small populations.
-However, for applicability to larger populations, we use instead the following Euler scheme.
-Write $\delta$ for an Euler time step, and $\Delta N_{ij}$ for the numerical approximation to $N_{ij}(t+\delta)-N_{ij}(t)$ given $X(t)$.
-For each $i$ and $j$ in $\seq{1}{c} \cup \{\demography\}$ with $i \neq j$, we draw independent Gamma distributed noise increments with mean $\delta$ and variance $\sigma_{ij}^2 \delta$, denoted using a mean-variance parameterization of the gamma distribution as
-\begin{equation}
-\label{eq:numerics1}
-\Delta\Gamma_{ij} \sim \mathrm{gamma}(\delta, \sigma_{ij}^2 \delta).
-\end{equation}
-In the case of an SMCS model, $\sigma_{ij}=0$ for all $i$ and $j$, so we have $\Delta\Gamma_{ij}=\delta$.
-Then, for $i\neq \demography$ and $j\neq i$, and writing
-\begin{equation}
-\label{eq:numerics2}
-\mu_{ij}=\bar\mu_{ij}\big(t,X(t)\big) \Delta\Gamma_{ij} / \delta,
-\end{equation}
-we calculate transition probabilities
-\begin{eqnarray}
-\label{eq:numerics3}
-p_{ij} &=& \exp\left\{-\sum_{k\in 1:c \, \cup \{\demography\}} \mu_{ik} \, \delta \right\}
-\frac{\mu_{ij}}{\sum_{k\in 1:c\cup \{\demography\}} \mu_{ik}},
-\\
-\label{eq:numerics4}
-p_{ii} &=& 1 - \sum_{j\neq i} p_{ij}.
-\end{eqnarray}
-These probabilities correspond to competing hazards for every individual in compartment $i$ to transition to some compartment $j$, interpreting $j=i$ to mean that the individual remains in $i$.
-Then, $\big(\Delta N_{i1},\dots,\Delta N_{ic},\Delta N_{i\demography}\big)$ has the multinomial distribution where $X_i(t)$ individuals are allocated independently to $\seq{1}{c}\cup\{\demography\}$ with probabilities given by \eqref{eq:numerics3} and \eqref{eq:numerics4}.
-We use the \code{reulermultinom} function in the \code{pomp} package to draw from this multionomial distribution.
-
-Different treatments of demographic flows---such as birth, death, immigration and emigration---are possible.
-For the case $i=\demography$, the treatment used by Model~1 is to set
-\begin{equation}
-\label{eq:numerics6}
-\Delta N_{\demography j} \sim \mathrm{poisson}( \mu_{\demography j} \delta),
-\end{equation}
-an independent Poisson random variable with mean $\mu_{\demography j} \delta$.
-
-Models~2 and 3 used an alternative approach, balancing the total number of flows in and out of the compartment, i.e., $\sum_{i}N_{\demography i}(t) = \sum_{i}N_{i \demography}(t)$, in order to make the model consistent with the known total population.
-In this case, we formally model the death rate as a rate of returning to the susceptible class $S$, and use external transitions from $\demography$ into $S$ to describe only net population increase.
-
-\section{Measurement Models}
-
-Each POMP requires specification of a measurement model, which is a statistical description of how observations on the system are obtained.
-In general, we used the same measurement models that were reported by \cite{lee20} unless specifically noted in the following subsections.
-
-\subsection{Model~1}
-
-In this model, the advantage afforded by vaccination is an increased probability that an infection is asymptomatic.
-Therefore, under the assumptions of this model, all reported cases are assumed to be a fraction of individuals that transition from the exposed to the infected compartment, as noted in Eq.~\myeqref{model1:meas}:
-\begin{equation}
-  \label{model1:meas}
-  Y_{n} \mid \Delta N_{E_{\cdot}I_{\cdot}}(n) = \reportChange \sim \text{NB}\left(\reportRate \reportChange, \obsOverdispersion \right),
-\end{equation}
-where $Y_n$ is the reported cholera cases at time $t_n \in t_1:t_N$ and $\Delta N_{E_{\cdot}I_{\cdot}}(n)$ is the sum total of individuals across each vaccination compartment $\vaccCounter \in 1:\vaccClass$ who moved from compartment $E_\vaccCounter$ to $I_\vaccCounter$ since observation $t_{n-1}$.
-Here, $\text{NB}\left(\reportRate \reportChange, \obsOverdispersion \right)$ denotes a negative binomial distribution with mean $\reportRate \reportChange$ and variance $\reportRate \reportChange \big(1 + \frac{\reportRate \reportChange}{\obsOverdispersion}\big)$.
-
-\subsection{Model~2}\label{sec:mod2Meas}
-
-Model~2 was fit using reported case counts that were transformed using the natural logarithm.
-We fit Model~2 using the subplex algorithm in the \code{subplex} package, using
-a Gaussian measurement model (Eq.~\myeqref{model2:meas}) on the log transformed cases within each unit.
-The final loss function that is maximized is the product of the likelihoods of the individual units, or the sum of the log-likelihood of the individual units.
-The measurement model for individual units is given in Eq.~\myeqref{model2:meas}.
-\begin{equation}
-  \label{model2:meas}
-  \log\big(Y_{u, n} + 1\big) \mid \Delta N_{E_{u\cdot}I_{u\cdot}}(n) = \reportChange_u \sim \text{N}\left(\log \big(\reportRate \reportChange_u + 1\big), \obsOverdispersion^2 \right),
-\end{equation}
-where $\Delta N_{E_{u\cdot}I_{u\cdot}}(n)$ is the sum total of individuals across vaccination compartment $\vaccCounter \in 0:4$ within unit $u$ who moved from compartment $E_{u\vaccCounter}$ to $I_{u\vaccCounter}$ since observation $t_{n-1}$.
-Therefore, because the natural logarithm of observed case counts (plus one, to avoid taking the logarithm of zero) has a normal distribution, $Y_{u, n} + 1$ is assumed to follow a log-normal distribution with log-mean parameter $\log\big(\reportRate \Delta N_{E_{u\cdot}I_{u\cdot}}(n) + 1\big)$ and log-variance $\obsOverdispersion^2$.
-We note that fitting a model with this measurement model is equivalent to fitting using least squares, with $\log(Y_{u, n} + 1)$ as the response variable.
-
-This measurement model differs from that used by \citet{lee20}, who fit the model in two stages: epidemic and endemic phases.
-Although their text and supplement material do not explicitly describe the measurement model used, inspection of the code provided with their submission suggests a change in measurement model between the epidemic and endemic phases.
-% In the file \code{choleraEqsPublic.py}, \citet{lee20} create several functions, where each function represents a set of coupled differential equations that could potentially be used to model cholera incidence data.
-% Each function returns a vector (or \code{numpy} array) that represents the change in each state variable for a single time step, including the variable \code{dC}, which tracks the number of new infections and is used to obtain the reported case counts.
-% Following their comments and code, it appears that the function \code{choleraEqs10WithoutVaccinationNetwork} was used to describe the dynamics of the epidemic phase, and \code{choleraEqs11WithoutVaccinationNetwork} was used for the endemic stage (see Sec~\ref{sec:mod2rep} for more details).
-% Because their models were fit using least squares, the code in these functions suggest that the measurement model for the epidemic phase is
-The measurement model they used for the epidemic phase is
-\begin{equation}
-  \label{model2:measEpi}
-  Y_{u, n} \mid \Delta N_{E_{u\cdot}I_{u\cdot}}(n) = \reportChange_u \sim \text{N}\left(\reportRate \reportChange_u, \obsOverdispersion^2 \right),
-\end{equation}
-% which is similar to our measurement model, the primary difference being that the measurement model is applied to raw case counts rather than log-transformed case counts.
-The measurement model they used for the endemic phase modifies the epidemic model by counting both asymptomatically infected (A) and symptomatically infected (I) individuals in the case counts:
-\begin{equation}
-  \label{model2:measEnd}
-  Y_{u, n} \mid \Delta N_{E_{u\cdot}I_{u\cdot}}(n) = \reportChange_{u1}, \Delta N_{E_{u\cdot}A_{u\cdot}}(n) = \reportChange_{u2} \sim \text{N}\left(\reportRate \big(\reportChange_{u1} + \reportChange_{u2}\big), \obsOverdispersion^2 \right),
-\end{equation}
-where the notation for $\Delta N_{E_{u\cdot}A_{u\cdot}}(n)$ is similar to $\Delta N_{E_{u\cdot}I_{u\cdot}}(n)$, described above.
-
-\subsection{Model~3}
-
-In this model, reported cholera cases are assumed to stem from individuals who develop symptoms and seek healthcare.
-Therefore reported cases are assumed to come from an over-dispersed negative binomial model, given the increase in infected individuals:
-\begin{equation}
-  \label{model3:meas}
-  Y_{u, n} \mid \Delta N_{S_{u \cdot}I_{u}}(t) = \reportChange_u \sim \text{NB}\left(\reportRate \reportChange_{u}, \obsOverdispersion \right),
-\end{equation}
-where $\Delta N_{S_{u \cdot}I_{u \vaccCounter}}(n)$ is the number of individuals who moved from compartment $S_{u \vaccCounter}$ to $I_{u}$ since observation $t_{n-1}$.
-
-This measurement model is a minor change from that used by \cite{lee20}, which allowed for a change in the reporting rate on January 1st, 2018.
-The fitted values of the reporting rate---before and after January 2018---were $0.97$ and $0.097$, respectively.
-An instantaneous change from near perfect to almost non-existent reporting can be problematic, as it forces the model to explain the observed reduction in reported cases as a decrease in the reporting of cases, rather than a decrease in the prevalence of cholera.
-This shift was justified by a ``change of the case definition that occurred on January 1st, 2018";
-this claim was not cited, and we could find no evidence that such a drastic change in the reporting rate would be warranted.
-We therefore do not allow a change in reporting rate when fitting Model~3.
-
-\section{Initial Values}
-
-To perform inference on POMP models, it is necessary to propose an initial probability density for the latent process $f_{X_0}(x_0;\theta)$, including the possibility that the initial values of the latent states are known, or are a non-random function of the unknown parameter vector, $\theta$.
-This density is used to obtain initial values of the latent state when fitting and evaluating the model.
-For each of the models considered in this analysis, the initial conditions are derived by enforcing the model dynamics on reported cholera cases.
-It is also sometimes necessary to fit some initial value parameters in order to help determine initial values for weakly identifiable compartments.
-In the following subsections, we mention initial value parameters that were fit for each model.
-
-\subsection{Model~1}
-
-For this model, the number of individuals in the Recovered and Asymptomatic compartments are set to zero, but the initial proportion of Infected and Exposed individuals is estimated as initial value parameters ($\Iinit$ and $\Einit$, respectively) using the IF2 algorithm, implemented as \texttt{mif2} in the \texttt{pomp} package.
-Finally, the initial proportion of Susceptible individuals $S_{0, 0}$ is calculated as $S_{0, 0} = 1 - \Iinit - \Einit$.
-This model for the initial values of the latent states matches that which was used by \citet{lee20}.
-
-\subsection{Model~2}
-
-Model~2 assumes that the initial values are a deterministic function of the reporting rate and the initial case reports, and so no initial value parameters need to be estimated.
-Initial values for latent state compartments are chosen so as to enforce the model dynamics on the observed number of cases.
-Specifically, the model sets $I_{u0}(0) = y^*_{1u} / \reportRate$ for each unit $u \in 1:10$, where $I_{u0}(t)$ is the number of infected individuals in unit $u$ at time $t$, vaccination scenario $z = 0$, $y^*_{tu}$ is the reported number of cases, and $\reportRate$ is the reporting rate.
-This model for the initial values of the latent states matches that which was used by \citet{lee20}.
-
-\subsection{Model~3}
-
-The latent states of this model are initialized by enforcing the model dynamics on the incidence data from the start of the recorded cases until time $t_0$, requiring that some of the available data be used to determine the initial values of the latent states.
-This is the same approach that was taken by \citet{lee20}, who used the value $t_0 = \text{2014-02-22}$; this choice of $t_0$ results in modeling roughly only $60\%$ of the available data, some of which is later discarded for alternative reasons \citep{lee20sup}.
-
-We do not see any immediate reason that this model could not be extended to cover a larger range of the data, and chose the value $t_0 = \text{2010-11-13}$.
-This choice of $t_0$ corresponds to using approximately $1\%$ of the available data to determine initial values of the latent states.
-In addition to modeling a larger portion of the available data, this choice of $t_0$ corresponds to an important real-world event, as daily reports from each of the departments were not being sent to the health ministry until November 10, 2010; % TODO: Cite
-this choice of $t_0$ therefore makes $\bm{Y}\big(t_1\big)$ the first week of data once daily reports are being sent to the health ministry.
-The few observation times that exist before $t_0$ are used to initialize the model by enforcing model dynamics on these preliminary observations.
-For convenience, we denote these observations as $t_{-3}, t_{-2}$ and $t_{-1}$; as before, we let $y_{u,-k} = Y_{u}\big(t_{-k}\big)$ denote the observed case count for unit $u$ at time point $t_{-k}$, where $k \in 1:3$.
-Equations for the initial values of non-zero latent states are provided in Eqs.~\myeqref{eq:model3:I0}--\myeqref{eq:model3:B0}.
-
-\begin{eqnarray}
-\label{eq:model3:I0}
-I_{u}\big(t_0\big) &=& \frac{y^*_{u, -1}}{7\reportRate\left(\muIR + \left(\muDeath + \choleraDeath\right)/365\right)},
-\\%[-3pt]
-\label{eq:model3:A0}
-A_{u}\big(t_0\big) &=& \frac{I_{u0}\big(t_0\big)\left(1 - \symptomFrac\right)}{\symptomFrac},
-\\
-\label{eq:model3:R0}
-R_{u01}\big(t_0\big) &=& R_{u02}\big(t_0\big) = R_{u03}\big(t_0\big) = \left(\frac{\sum_{k = -3}^0 y^*_{u, k}}{\reportRate \symptomFrac} - \left(I_{u0}\big(t_0\big) + A_{u0}\big(t_0\big)\right)\right) / 3
-\\
-\label{eq:model3:S0}
-S_{u0}\big(t_0\big) &=& \mathrm{Pop}_u - I_{u}\big(t_0\big) - A_{u}\big(t_0\big) - \sum_{k=1}^3 R_{u0k}\big(t_0\big)
-\\
-\label{eq:model3:B0}
-B_{u}\big(t_0\big) &=& \big[1 + \seasAmplitude \tilde{J}^r \big] \mathrm{Den}_u \, \Wshed \big[ I_{u}(t_0)+ \asymptomRelativeShed A_{u_0}(t) \big]/\Wshed.
-\end{eqnarray}
-
-In Eq.~\myeqref{eq:model3:B0}, $\tilde{J} = 0.002376$ is the median adjusted rainfall over the observation period.
-One important consideration to make with this parameter initialization model is when $y_{u, -1}^* = 0$, which occurs for units $u \in \{3, 4\}$, which correspond to the Grand'Anse and Nippes departments, respectively.
-When this is the case, each of the infectious $I_u(t_0)$, asymptomatic $A_u(t_0)$, and bacterial reservoir $W_u(t_0)$ compartments have a value of zero.
-Because Model~3 models cholera transmission primarily by means of the bacterial reservoir, this makes it nearly impossible for an outbreak to occur.
-Therefore for units $u \in \{3, 4\}$, we introduce initial value parameters $I_{0,0}^3$ and $I_{0,0}^4$, and calibrate these parameter values using the data.
-The resulting parameter estimates are used to obtain the remaining non-zero initial values of the latent states using Eqs.~\myeqref{eq:model3:A0}--\myeqref{eq:model3:B0}.
-
-% Argue why we didn't need to do this for Model 2.
-
-% TODO: Check this section.
-
 \section{Calibrating Model~3 to observed cases}\label{sec:mod3Cal}
 
-<<CalibrateOrig3>>=
-# Create vectors for the unit and shared parameters
-unit_specific_names <- c("betaB", "foi_add")
-shared_param_names <- c(
-  "mu_B", "XthetaA", "thetaI", "lambdaR", "r", "std_W",
-  "epsilon", "k"
-)
-est_param_names <- c(
-  unit_specific_names, shared_param_names
-)
-
-# Add unit numbers to each parameter
-est_param_names_expanded <- paste0(rep(est_param_names, each = 10), 1:10)
-
-# Simple function that is used to set the rw_sd for the first search
-# This function is convenient so that we don't have to type out the rw_sd
-# for each unit parameter.
-set_rw_1 <- function(x) {
-  if (gsub("[[:digit:]]+$", "", x) %in% shared_param_names) {
-    return(0.01)
-  # } else if (x %in% paste0(rep(c("aHur", 'hHur'), each = 2), c(3, 9))) {
-  #   return(expression(ifelse(time >= 2016.754 & time <= 2017, 0.015, 0)))
-  # } else if (x %in% c("Iinit3", "Iinit4")) {
-  #   return(expression(ivp(0.25)))
-  } else if (grepl('^foi_add[[:digit:]]+$', x)) {
-    return(0.015)
-  } else if (grepl('^betaB[[:digit:]]+$', x)) {
-    return(0.01)
-  } else {
-    return(0)
-  }
-}
-
-set_rw_2 <- function(x) {
-  if (gsub("[[:digit:]]+$", "", x) %in% shared_param_names) {
-    return(0.003)
-  # } else if (x %in% paste0(rep(c("aHur", 'hHur'), each = 2), c(3, 9))) {
-  #   return(expression(ifelse(time >= 2016.754 & time <= 2017, 0.005, 0)))
-  # } else if (x %in% c("Iinit3", "Iinit4")) {
-  #   return(expression(ivp(0.125)))
-  } else if (grepl('^foi_add[[:digit:]]+$', x)) {
-    return(0.005)
-  } else if (grepl('^betaB[[:digit:]]+$', x)) {
-    return(0.003)
-  } else {
-    return(0)
-  }
-}
-
-# First Search RW
-reg_rw_1.sd <- lapply(est_param_names_expanded, set_rw_1)
-names(reg_rw_1.sd) <- est_param_names_expanded
-chol_rw_1.sd <- do.call(rw_sd, reg_rw_1.sd)
-
-# Second Search RW
-reg_rw_2.sd <- lapply(est_param_names_expanded, set_rw_2)
-names(reg_rw_2.sd) <- est_param_names_expanded
-chol_rw_2.sd <- do.call(rw_sd, reg_rw_2.sd)
-
-SEARCH1 = list(
-  NBPF = 100,
-  NP = 2500,
-  SPAT_REGRESSION = 0.05,
-  NREPS = 144,
-  NP_EVAL = 2500,
-  NREPS_EVAL = 18,
-  RW_SD = chol_rw_1.sd,
-  COOLING = 0.5,
-  KEEP_TRACES = FALSE,
-  KEEP_LIKE_MAT = FALSE
-)
-
-SEARCH2 = list(
-  TOP_N = 12,
-  NBPF = 100,
-  NP = 2500,
-  SPAT_REGRESSION = 0.05,
-  NREPS = 6,
-  NP_EVAL = 2500,
-  NREPS_EVAL = 18,
-  RW_SD = chol_rw_2.sd,
-  COOLING = 0.5,
-  KEEP_TRACES = FALSE,
-  KEEP_LIK_MAT = TRUE
-)
-
-haiti3_sub_fit <- bake(
-  file = "../model3/si/subset_noparams.rds", {
-    fit_haiti3(
-      nsearches = 2L,
-      search1 = SEARCH1,
-      search2 = SEARCH2,
-      search_rho = FALSE,
-      search_gamma = FALSE,
-      search_Iinit = FALSE,
-      search_hur = FALSE,
-      ncores = cores
-    )
-  }
-)
-
-# Find model index with best log-likelihood
-best_m <- which.max(haiti3_sub_fit$search2$logLiks$logLik)
-
-# Find parameters using best_m index
-sub_params <- haiti3_sub_fit$search2$params[best_m, ]
-
-# Get unit log-likelihoods for the best model.
-all_liks <- as.data.frame(haiti3_sub_fit$search2$likMat)
-# all_liks$which <- rep(1:(SEARCH3$TOP_N * SEARCH3$NREPS), each = SEARCH3$NREPS_EVAL)
-all_liks$which <- rep(1:(12 * 6), each = 18)
-best_mod <- filter(all_liks, which == best_m)
-
-# Pivot longer to get estimate of log-likelihood for each unit.
-unitLikesM3 <- best_mod %>%
-  pivot_longer(
-    cols = -which,
-    names_to = "unit",
-    names_prefix = "V",
-    values_to = "logLik"
-  ) %>%
-  select(-which) %>%
-  group_by(unit) %>%
-  summarize(
-    ll_est = logmeanexp(logLik),
-    ll_se = logmeanexp(logLik, se = TRUE)[2]
-  ) %>%
-  ungroup() %>%
-  mutate(unit = as.numeric(unit)) %>%
-  arrange(unit)
-@
-
-<<M3_bench, echo=FALSE, message=FALSE, include=FALSE, cache=TRUE>>=
-h3 <- haiti3_spatPomp()
-deps <- h3@unit_names
-
-mod3_cases <- as.data.frame(t(h3@data))
-colnames(mod3_cases) <- deps
-m3_inits <- c(7362, 1914, 22, 8, 2818, 82, 2499, 6953, 44, 7) # TODO: Also fit by hand, but could be changed to initializing to the observation before.
-names(m3_inits) <- deps
-
-bench_cond_lls <- list()
-for (dep in deps) {
-
-  # Get department cases
-  cases <- as.numeric(mod3_cases[, dep])
-  mod <- ar1_NegBinom(data = cases, init = m3_inits[dep])
-
-  bench_cond_lls[[dep]] <- haitipkg:::.ar1_NegBinom_cond_ll(
-    theta = mod$theta, init = m3_inits[dep], data = cases, transform = FALSE
-  )
-}
-
-bench_cond_lls <- as.data.frame(bench_cond_lls)
-
-bench_lls <- apply(bench_cond_lls, 2, sum)
-
-unitLikes_bench <- data.frame(
-  unit = 1:10,
-  # dep = names(bench_lls),
-  ll = bench_lls
-) %>%
-  as_tibble()
-
-bench_cond_lls$date <- lubridate::round_date(lubridate::date_decimal(h3@times), unit = 'day')
-@
-
-<<combine_bench_M3, echo=FALSE>>=
-
-# Combine tables with model 3 and benchmark likelihoods.
-bench_M3_joined <- dplyr::left_join(
-  x = unitLikesM3,
-  y = unitLikes_bench,
-  by = "unit"
-) %>%
-  select(-ll_se) %>%
-  mutate(unit = as.character(unit)) %>%
-  pivot_longer(
-    cols = -unit,
-    names_to = "type",
-    values_to = "ll"
-  ) %>%
-  mutate(unit = paste0(unit, ": ", deps[as.numeric(unit)]))
-@
-
-<<M3bpf, echo=FALSE, message=FALSE, include=FALSE>>=
-
-# Change model parameters to the best set of parameters from previous
-# search
-coef(h3)[names(sub_params)] <- sub_params
-
-registerDoRNG(40765101)  # for modular, reproducible results
-
-# Perform block partical filter with fitted model, to get conditional
-# log-likelihoods.
-bpf_sub3_out <- bake(
-  file = "../model3/si/bpf_sub_cond_LL.rds", {
-
-    results <- list()
-
-    bpf_out <- h3 %>% bpfilter(
-      Np = 5000,
-      block_size = 1,
-      save_states = TRUE
-    )
-
-    results$block.cond.loglik <- bpf_out@block.cond.loglik
-    results$times <- bpf_out@times  # Saved for convenience
-    results
-  },
-  seed = 432235
-)
-@
-
-In this section, we provide more detail on the process that was used to estimate the coefficients used in Model~3.
+In this section, we provide more detail on the process that was used to estimate the coefficients of Model~3.
 In particular, we discuss why we decided to include additional model parameters---those that are associated with the behavior of the system during Hurricane Matthew---that were not considered by \cite{lee20}.
 To calibrate this model, we used the iterated block particle filter (IBPF) method of \cite{ionides22}.
-Due to the novelty of this algorithm, there does not exist published examples the IBPF algorithm used for data analysis outside of the papers in which the algorithm was introduced \citep{ning21ibpf,ionides22}, which at least partially motivates the inclusion additional details in this section.
+Due to the novelty of this algorithm, there does not exist published examples the IBPF algorithm used for data analysis outside of the papers in which the algorithm was introduced \citep{ning21ibpf,ionides22}, which is one motivation of the inclusion additional details related to fitting and diagnosing the model fit provided here.
 
@@
 The first collection of searches was performed by obtaining initial values for the parameters by uniformly sampling values from a predefined hypercube.
-A subsequent refinement search used starting values corresponding to the largest model likelihoods obtained in the previous round.
-This is similar to the approach used by \cite{ionides22}.
+A subsequent refinement search used parameter values corresponding to the largest model likelihoods as starting parameter values.
+The need for multiple searches does not appear to be uncommon, as a similar approach was used by \cite{ionides22}.
 
 We use the technique described above to fit the fully coupled version of Model~3 proposed by \cite{lee20}.
-The maximum likelihood we obtained with the model was $\Sexpr{round(max(haiti3_sub_fit$search2$logLiks$logLik))}$, which is higher than the benchmark model ($\Sexpr{round(sum(unitLikes_bench$ll))}$).
+The maximum likelihood we obtained after two rounds of searching was $-17559$, which is higher than the benchmark model ($-17933$).
 While beating a simple associative benchmark is promising, this does not immediately rule out the possibility that the model is a poor description of the system.
@@
 
-\begin{figure}[ht]
-<<h3UnitLikes, fig.height=3, fig.width=5.5, echo=FALSE>>=
-ggplot(bench_M3_joined, aes(x = ll, y = reorder(unit, ll))) +
-  geom_point(aes(col = type)) +
-  ylab("Unit") +
-  xlab("Log Likelihood") +
-  scale_color_manual(
-    labels = c("ll" = "Benchmark", "ll_est" = "Model 3"),
-    values = c("ll" = "#1b9e77", "ll_est" = "#d95f02")
-  ) +
-  theme(legend.title = element_blank(),
-        axis.title.y = element_blank())
-@
-\caption{\label{fig:h3UnitLikes}Log-likelihoods of Model~3 for each department compared to the corresponding benchmark model prior to the inclusion of parameters for initial values and Hurricane Matthew.}
+\begin{figure}[!ht]
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
+
+{\centering \includegraphics[width=\maxwidth]{figure/h3UnitLikes-1}
+
+}
+
+\end{knitrout}
+\caption{\label{fig:h3UnitLikes}Log-likelihoods of Model~3 for each department compared to the corresponding benchmark model prior to the inclusion of parameters that account for Hurricane Matthew.}
 \end{figure}
@@
 In addition to considering the conditional log-likelihoods of each unit, one can consider conditional log-likelihoods of each observation.
-When compared to a benchmark, this level of detail can provide useful information about which observations are well described by the model and which are poorly described.
-In Fig.~\ref{fig:condLL}, we plot the conditional log-likelihoods of each observation.
-This figure reveals that the fitted model has difficult time explaining certain features of the data.
+When compared to a benchmark, this level of detail can provide useful information about which observations are well described by the model and which are not.
+In Fig.~\ref{fig:condLL}, we plot the conditional log-likelihoods of Model~3 for each observation.
+Typically it is most useful to compare the conditional log-likelihoods of the model under consideration to a benchmark, as plotting only conditional log-likelihoods without a comparison may not be helpful.
+In this case, however, the same insight can be drawn using a figure without a benchmark comparison, so we do not include the benchmark in order to avoid the issue of over-plotting.
+
+\begin{figure}[!ht]
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
+
+{\centering \includegraphics[width=\maxwidth]{figure/CondLLFig-1}
+
+}
+
+\end{knitrout}
+\caption{\label{fig:condLL}Conditional log-likelihoods of Model~3 prior to the inclusion of the Hurricane Matthew related parameters.}
+\end{figure}
+
+Fig.~\ref{fig:condLL} reveals that the fitted model poorly describes certain features of the data.
 For example, many departments (in particular Sud) have observations with lower conditional log-likelihoods near October 2016 than at other time points.
@@
 While the model does include a mechanism to account for increased cholera transmission due to large rainfall events, the mechanism does not appear to be sufficient to capture the damaging effects of the hurricane, which had the greatest impact in the the Sud and Grand'Anse departments \citep{ferreirai16}.
-This result led us to include parameters $\Whur{u}$ and $\hHur{u}$ in Eq.~23 of the main text, which allows for an increase in the transmission rate between environmental cholera and humans for Sud and Grand'Anse during and after the hurricane.
+This result led us to include parameters $\Whur{u}$ and $\hHur{u}$ in Eq.~23 of the main text, which allows for an increase in the transmission rate between environmental cholera and humans for in Sud and Grand'Anse during and after the hurricane.
 The effect of the hurricane on cholera transmission is assumed to have an exponential decay, where the magnitude is determined by $\Whur{u}$ and the duration of the effect determined by $\hHur{u}$.
 
-\begin{figure}[ht]
-<<CondLLFig, echo=FALSE, message=FALSE, fig.height=3>>=
+We refit Model~3 after introducing these hurricane-related parameters using the same successive search scheme described above.
+The resulting model has a log-likelihood value of $-17402.4$.
+The addition of the Hurricane parameters seems to have resulted in a dramatic increase in conditional likelihoods around and after October 2016, as seen in Fig.~\ref{fig:finalCondLL}.
+The inclusion of these parameters resulted in an overall increase of $157.1$ log-likelihood units, which is a highly statistically significant difference.
 
-# Changing labels to human-readable labels.
-deps[grepl("_", deps)] <- gsub("_", "-", deps[grepl("_", deps)])
-deps[deps == "Grande-Anse"] <- "Grand\'Anse"
-names(deps) <- as.character(1:10)
-myLabeller <- as_labeller(deps)
+\begin{figure}[!ht]
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
 
-# Plot conditional log-likelihoods for each unit
-t(bpf_sub3_out$block.cond.loglik) %>%
-  as.data.frame() %>%
-  mutate(t = bpf_sub3_out$times) %>%
-  pivot_longer(
-    cols = -t,
-    values_to = 'likelihood',
-    names_prefix = "V",
-    names_to = 'Unit'
-  ) %>%
-  mutate(date = as.Date(lubridate::date_decimal(t))) %>%
-  ggplot(aes(x = date, y = likelihood)) +
-  geom_line() +
-  geom_point() +
-  facet_wrap(~Unit, labeller = myLabeller, scales = 'free_y',
-             nrow = 2) +
-  theme_bw() +
-  theme(axis.title.x = element_blank(),
-        axis.text.x = element_text(angle = -45, hjust = 0)) +
-  ylab("Conditional log-likelihoods")
-@
-\caption{\label{fig:condLL}Conditional log-likelihoods of Model~3 prior to the inclusion of parameters for initial values and Hurricane Matthew.}
+{\centering \includegraphics[width=\maxwidth]{figure/finalCondLLFig-1}
+
+}
+
+\end{knitrout}
+\caption{\label{fig:finalCondLL}Conditional log-likelihoods of Model~3 after adding and estimating the parameters related to Hurricane Matthew.}
 \end{figure}
 
-Other points of concern include the observations at or near the start of the recorded cases, for example, in Nord-Ouest and Sud-Est.
-These low conditional log-likelihoods are a result of having zero or near zero confirmed cases in the first week of observed data, followed by a rapid growth of cases.
-The model cannot explain the rapid growth of cholera in the absence of starting cases because the initialization scheme forces departments with no recorded cases in the first week to have hidden states $I_{u0}(0) = A_{u0}(0) = B_{u}(0) = 0$.
-In reality, however, it is possible to have some infected or asymptomatic individuals in a department when no cases are recorded due to under-reporting.
-This may be particularly true at the onset of an infectious disease outbreak, as lower public awareness and inadequate testing infrastructure may result in many cases going undetected.
-We therefore include and estimate $I_{0u}$ that determines the initial number of Infected individuals at the start of the pandemic for the departments Grande'Anse, Nord-Ouest, and Sud-Est;
-each of these departments have both low starting case counts followed by rapid growth, and low initial values of the conditinoal log-likelihood.
+Now that the model with additional parameters has been calibrated to the incidence data, we plot the conditional log-likelihood of each department compared to a benchmark model in Fig.~\ref{fig:finalUnitLL}.
+Interestingly, the difference in log-likelihoods between Model~3 and the benchmark model is smallest in the departments Artibonite, Nord, Centre and Ouest.
+Each of these departments also exhibited the most sustained cholera transmission, defined by having the fewest number of weeks with no recorded cholera cases.
+Specifically, these four departments have zero cholera cases recorded in less than $4\%$ of the available data, and all remaining departments---except for Nord-Ouest, which has approximately $9.5\%$ of cases that are zeros and also exhibits the next smallest difference in log-likelihoods---have zero cases recorded in at least $14\%$ of the available weekly data.
+This result suggests that the quantitative advantage Model~3 has over its respective benchmark is primarily due to the model's ability to describe a resurgence of cases after a department records a week of zero cholera cases.
+This result may be unsurprising in the context of the models that we are comparing.
+Because the cholera transmission in individual departments likely depends on the national prevalence of cholera and the Vibrio cholerae bacteria, our benchmark model that relies exclusively on the previous number of case within any given unit has a difficult time predicting a resurgence of cases.
 
-<<loadFinalModel3, echo=FALSE, include=FALSE, message=FALSE>>=
+\begin{figure}[!ht]
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
 
-# This result was used in the manuscript, so it doesn't need to be recalculated.
-h3_final_results <- readRDS("../model3/run_level_3/haiti3_fit.rds")
+{\centering \includegraphics[width=\maxwidth]{figure/finalUnitLLFig-1}
 
-# Save best_m as index for the model parameters with the best likelihood.
-# Then, save the model parameter values and the likelihood evaluations
-best_m <- which.max(h3_final_results$search2$logLiks$logLik)
-best_params <- h3_final_results$search2$params[best_m, ]
-all_liks <- as.data.frame(h3_final_results$search2$likMat)
-all_liks$which <- rep(1:72, each = 18)
-best_mod <- dplyr::filter(all_liks, which == best_m)
+}
 
-# Get repeated estimates of unit log-likelihoods
-unitLikesM3 <- best_mod %>%
-  pivot_longer(
-    cols = -which,
-    names_to = "unit",
-    names_prefix = "V",
-    values_to = "logLik"
-  ) %>%
-  select(-which) %>%
-  group_by(unit) %>%
-  summarize(
-    ll_est = pomp::logmeanexp(logLik),
-    ll_se = pomp::logmeanexp(logLik, se = TRUE)[2]
-  ) %>%
-  ungroup() %>%
-  mutate(unit = as.numeric(unit)) %>%
-  arrange(unit)
-
-# Combine model likelihood table to benchmarks
-bench_M3_joined <- dplyr::left_join(
-  x = unitLikesM3,
-  y = unitLikes_bench,
-  by = "unit"
-) %>%
-  select(-ll_se) %>%
-  mutate(unit = as.character(unit)) %>%
-  pivot_longer(
-    cols = -unit,
-    names_to = "type",
-    values_to = "ll"
-  ) %>%
-  mutate(unit = paste0(unit, ": ", deps[as.numeric(unit)]))
-@
-
-We refit Model~3 after introducing these seven parameters---three initial value parameters and four parameters related to Hurricane Matthew---using the same successive search scheme described above.
-The resulting model has a log-likelihood value of $\Sexpr{round(max(h3_final_results$search2$logLiks$logLik), 1)}$.
-We also provide figures for department specific log-likelihoods and conditional log-likelihoods in Figs.~\ref{fig:finalUnitLL} and \ref{fig:finalCondLL}.
-
-\begin{figure}[ht]
-<<finalUnitLLFig, fig.height=3, fig.width=5.5, echo=FALSE>>=
-ggplot(bench_M3_joined, aes(x = ll, y = reorder(unit, ll))) +
-  geom_point(aes(col = type)) +
-  ylab("Unit") +
-  xlab("Log Likelihood") +
-  scale_color_manual(
-    labels = c("ll" = "Benchmark", "ll_est" = "Model 3"),
-    values = c("ll" = "#1b9e77", "ll_est" = "#d95f02")
-  ) +
-  theme(legend.title = element_blank(),
-        axis.title.y = element_blank())
-@
-\caption{\label{fig:finalUnitLL}Log-likelihoods of Model~3 for each department compared to the corresponding benchmark model after adding and estimating parameters related to initial values and Hurricane Matthew.}
+\end{knitrout}
+\caption{\label{fig:finalUnitLL}Log-likelihoods of Model~3 for each department compared to the corresponding benchmark model after adding and estimating parameters related to Hurricane Matthew.}
 \end{figure}
 
-<<finalModel3BPF, echo=FALSE, message=FALSE, include=FALSE>>=
-
-# Set the model parameters to the best results from previous search
-coef(h3)[names(best_params)] <- best_params
-
-# block particle filter to get conditional log-likelihoods.
-bpf_final3_out <- bake(
-  file = "../model3/si/h3_final_bpf_cond_ll.rds", {
-
-    results <- list()
-
-    bpf_out <- h3 %>% bpfilter(
-      Np = 5000,
-      block_size = 1,
-      save_states = TRUE
-    )
-
-    results$Sus_df <- foreach(i=1:length(h3@times), .combine = bind_rows) %do% {
-      get_h3_S_quants(bpf_out, i)
-    }
-
-    results$block.cond.loglik <- bpf_out@block.cond.loglik
-    results$times <- bpf_out@times  # saved for convenience
-    results
-  },
-  seed = 786321
-)
-@
-
-\begin{figure}[ht]
-<<finalCondLLFig, echo=FALSE, message=FALSE, fig.height=3>>=
-t(bpf_final3_out$block.cond.loglik) %>%
-  as.data.frame() %>%
-  mutate(t = bpf_final3_out$times) %>%
-  pivot_longer(
-    cols = -t,
-    values_to = 'likelihood',
-    names_prefix = "V",
-    names_to = 'Unit'
-  ) %>%
-  mutate(date = as.Date(lubridate::date_decimal(t))) %>%
-  ggplot(aes(x = date, y = likelihood)) +
-  geom_line() +
-  geom_point() +
-  facet_wrap(~Unit, labeller = myLabeller, scales = 'free_y',
-             nrow = 2) +
-  theme_bw() +
-  theme(axis.title.x = element_blank(),
-        axis.text.x = element_text(angle = -45, hjust = 0)) +
-  ylab("Conditional log-likelihoods")
-@
-\caption{\label{fig:finalCondLL}Conditional log-likelihoods of Model~3 after adding and estimating the parameters for initial values and Hurricane Matthew.}
-\end{figure}
-
-While the addition of the Hurricane parameters seems to have resulted in a dramatic increase of conditional likelihoods around and after October 2016, the inclusion of the initial value parameters does not appear to have fixed the low conditional log-likelihood values in Sud-Est and Nord-Ouest.
-This could be due to many factors, including the possibility that the coupling between units forces low initial value parameters in a given department in order for the model to fit the data in the remaining departments.
-Whatever the case may be, the inclusion of the seven parameters resulted in a $\Sexpr{round(max(h3_final_results$search2$logLiks$logLik) - max(haiti3_sub_fit$search2$logLiks$logLik), 1)}$ increase in log-likelihood, which is a highly statistically significant difference.
-One could similarly test whether the inclusion of the initial value parameters resulted in a significantly better fit to the observed data, but we do not do this here due to the large computational effort that would be required.
+The difference in log-likelihoods between Model~3 and its benchmark model for each individual units suggests that Model~3 has a relatively poor fit for the four units with the most sustained cholera transmission.
+The simple four parameter benchmark has a higher likelihood than Model~3 for the Artibonite and Nord departments, and also has log-likelihoods only a few units smaller than Model~3 for units Centre and Ouest.
+This is particularly worrisome given that these four departments account for more than $77\%$ of the total number of reported cholera cases.
 
@@
 
-<<h3SusFigSetup, echo=FALSE, message=FALSE, include=FALSE>>=
+\begin{figure}[!ht]
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
 
-# This bake perfoms the following:
-#  (1) simulates the model from initial conditions
-#  (2) get the values of the hidden states at the end of each simulation, and
-#      saves these for later use.
-#  (3) Obtain quantiles for the Susceptible pool from the simulations.
-init_conditions_results <- bake(
-  file = "../model3/si/InitSimsQuants.rds", {
+{\centering \includegraphics[width=\maxwidth]{figure/h3SusFig-1}
 
-    results <- list()  # To return final output of bake
+}
 
-    # Simulate from the fitted model
-    h3Spat_sims <- simulate(
-      h3, nsim = 2000, format = 'data.frame'
-    )
-
-    # Get states at the end of the simulations
-    h3_spat_end <- h3Spat_sims %>%
-      filter(time == max(time))
-
-    # Save hidden state values at the end of the simulation. Saved as a matrix
-    # in the same format as the output from a block-particle filter, so that
-    # it can be used as the input to project_from_filter.
-    results[["end_conditions"]] <- h3_spat_end %>%
-      select(-time, -cases) %>%
-      mutate(
-        unit = case_when(
-          unitname == "Artibonite" ~ 1,
-          unitname == "Centre" ~ 2,
-          unitname == "Grande_Anse" ~ 3,
-          unitname == "Nippes" ~ 4,
-          unitname == "Nord" ~ 5,
-          unitname == "Nord_Est" ~ 6,
-          unitname == "Nord_Ouest" ~ 7,
-          unitname == "Ouest" ~ 8,
-          unitname == "Sud" ~ 9,
-          unitname == "Sud_Est" ~ 10
-        )
-      ) %>%
-      select(-unitname) %>%
-      pivot_wider(
-        data = .,
-        names_from = unit,
-        values_from = c(S, I, A, R_one, R_two, R_three, VSd, VR1d, VR2d,
-                        VR3d, VSdd, VR1dd, VR2dd, VR3dd, VSd_alt, VR1d_alt,
-                        VR2d_alt, VR3d_alt, VSdd_alt, VR1dd_alt, VR2dd_alt,
-                        VR3dd_alt, C, B, Doses, totInc),
-        names_sep = ""
-      ) %>%
-      select(-.id) %>%
-      as.matrix() %>%
-      t()
-
-    # Get the quantiles for the Susceptible pool from the simulations
-    results[["SQuants"]] <- h3Spat_sims %>%
-      rename(dep = unitname) %>%
-      group_by(dep, time) %>%
-      summarise(
-        Q025 = quantile(S + VSd + VSdd + VSd_alt + VSdd_alt, probs = 0.025),
-        Q50  = quantile(S + VSd + VSdd + VSd_alt + VSdd_alt, probs = 0.5),
-        Q975 = quantile(S + VSd + VSdd + VSd_alt + VSdd_alt, probs = 0.975)
-      ) %>%
-      ungroup() %>%
-      rowwise() %>%
-      mutate(
-        unit_num = which(h3@unit_names == dep),
-        date = lubridate::date_decimal(time),
-        Q025 = Q025 / h3@params[paste0("H", unit_num)],
-        Q50  = Q50  / h3@params[paste0("H", unit_num)],
-        Q975 = Q975 / h3@params[paste0("H", unit_num)]
-      ) %>%
-      filter(
-        date >= as.Date(
-          lubridate::round_date(lubridate::date_decimal(h3@t0), unit = 'day')
-        )
-      ) %>%
-      mutate(date = as.Date(lubridate::round_date(date, unit = 'day')))
-
-    results
-  },
-  seed = 834687
-)
-
-# Get the output of the "bake" call for plotting.
-h3Spat_quants <- init_conditions_results[["SQuants"]]
-
-# This bake call uses the ending hidden states from above to forecast cholera
-# cases in the future under the V0 vaccination scenario
-mod3_V0_initcond_res <- bake(
-  file = '../model3/si/vacc_sim_from_init.rds', {
-    h3_V0_res <- list()
-
-    # Get "noVacc" scenario parameters
-    h3_S0_par <- get_model3_vacc_scenario_params(scenario = "noVacc")
-    h3Spat <- h3
-    p3 <- best_params
-    coef(h3Spat) <- p3
-    coef(h3Spat)[names(h3_S0_par)] <- h3_S0_par
-
-    # store the end states that will be used as initial values
-    h3_bpf_end_states <- init_conditions_results$end_conditions
-
-    # Using parameters and end state values, obtain projections
-    h3_V0_sims <- project_from_filter(
-      h3Spat, end_states = h3_bpf_end_states, covarGen = project_rain,
-      nsims = 1000, seed = 184697
-    )
-
-    # Store nationally aggregated results
-    h3_V0_res$mod3_V0_sims <- agg_mod3_sims(h3_V0_sims)
-    h3_V0_res$mod3_V0_probs <- get_elimProbs(h3_V0_sims, model = 3)
-
-    # Get quantiles of observed cases for each department and time
-    h3_V0_res$UnitQuants <- h3_V0_sims %>%
-      mutate(
-        date = as.Date(  # Get date object from decimal time
-          lubridate::round_date(lubridate::date_decimal(time), unit = 'day')
-        )
-      ) %>%
-      group_by(unitname, date) %>%
-      summarize(
-        Q025 = quantile(cases, probs = 0.025),
-        Q50 = quantile(cases, probs = 0.5),
-        Q975 = quantile(cases, probs = 0.975)
-      ) %>%
-      ungroup() %>%
-      rename(dep = unitname)
-
-    # Get quantiles of susceptible individuals for each department and time
-    h3_V0_res$SQuants <- h3_V0_sims %>%
-      mutate(
-        date = as.Date(
-          lubridate::round_date(lubridate::date_decimal(time), unit = 'day')
-        )
-      ) %>%
-      group_by(unitname, date) %>%
-      mutate(
-        # Get the population for each unit
-        pop = h3@params[paste0("H", which(first(unitname) == unit_names(h3)))]
-      ) %>%
-      summarize(  # Calculate the quantiles
-        Q025 = quantile(
-          (S + VSd + VSdd + VSdd_alt + VSd_alt) / pop,
-          probs = 0.025
-        ),
-        Q50 = quantile(
-          (S + VSd + VSdd + VSdd_alt + VSd_alt) / pop,
-          probs = 0.5
-        ),
-        Q975 = quantile(
-          (S + VSd + VSdd + VSdd_alt + VSd_alt) / pop,
-          probs = 0.975
-        )
-      ) %>%
-      ungroup() %>%
-      rename(dep = unitname)
-
-    h3_V0_res
-  }
-)
-
-# This result was already used in the manuscript; it doesn't need to
-# be recalculated.
-mod3_V0_res <- readRDS("../model3/run_level_3/mod3_V0_res.rds")
-@
-
-\begin{figure}[ht]
-<<h3SusFig, echo=FALSE, message=FALSE, fig.height=3>>=
-ggplot() +
-  # Sims from initial condition
-  geom_line(data = h3Spat_quants, aes(x = date, y = Q50), col = 'red') +
-  geom_ribbon(
-    data = h3Spat_quants,
-    aes(x = date, ymin = Q025, ymax = Q975),
-    fill = 'red', alpha = 0.5
-  ) +
-  # projections from initial condition, using filtered sims function
-  geom_line(data = mod3_V0_initcond_res$SQuants, aes(x = date, y = Q50), col = 'red') +
-  geom_ribbon(
-    data = mod3_V0_initcond_res$SQuants,
-    aes(x = date, ymin = Q025, ymax = Q975),
-    fill = 'red', alpha = 0.5
-  ) +
-  # Sims from filtered distribution
-  geom_line(
-    data = bpf_final3_out$Sus_df,
-    aes(x = as.Date(lubridate::date_decimal(time)), y = Q50),
-    col = 'blue'
-  ) +
-  geom_ribbon(
-    data = bpf_final3_out$Sus_df,
-    aes(x = as.Date(lubridate::date_decimal(time)), ymin = Q025, ymax = Q975),
-    fill = 'blue', alpha = 0.5
-  ) +
-  # Projections from filtered distribution
-  geom_line(data = mod3_V0_res$SQuants, aes(x = date, y = Q50), col = 'blue') +
-  geom_ribbon(
-    data = mod3_V0_res$SQuants,
-    aes(x = date, ymin = Q025, ymax = Q975),
-    fill = 'blue', alpha = 0.5
-  ) +
-  geom_vline(xintercept = max(h3Spat_quants$date), linetype = 'dashed') +
-  facet_wrap(~dep, scales = 'free_y', nrow = 2) +
-  theme_bw() +
-  theme(axis.title.x = element_blank(),
-        axis.text.x = element_text(angle = -45, hjust = 0)) +
-  ylab("Total Susceptible (%)")
-@
+\end{knitrout}
 \caption{\label{fig:h3Sus}Percentage of individuals that are in the susceptible compartment.
@@
 \end{figure}
-
-\clearpage
-
-\section{Replication of \cite{lee20}}
-
-In this article we claimed that we were able to obtain better fits to the observed data using the same models that were proposed by \cite{lee20}.
-Along with visual comparisons to the data, this claim was supported by comparing likelihoods and AIC values in Table~2 in the manuscript. % TODO: Double check the table number matches
-Because model likelihoods were not provided by \cite{lee20}, it is necessary to replicate these models in order to obtain likelihood estimates.
-Here we would like to thank the authors of \cite{lee20}, who provided detailed descriptions of their models, which enabled us to build on their work.
-In the following subsections, we use our \code{R} package \code{haitipkg} to reproduce some of the results of \cite{lee20}.
-This reproduction allows us to estimate the likelihoods of the \cite{lee20} version of Models~1--3, and also provides a demonstration of the importance and usefulness of reproducible research.
-
-\subsection{Model~1 Replication}
-
-The model was implemented by a team at Johns Hopkins Bloomberg School of Public Health (hereafter referred to as the Model~1 authors) in the \code{R} programming language using the \code{pomp} package \citep{king16}.
-Original source code is publicly available with DOI: 10.5281/zenodo.3360991.
-The final results reported by the Model~1 authors were obtained by using several different parameter sets rather than a single point estimate.
-According to the supplement materials, this was because model realizations from a single parameter set retained substantial variability, but multiple realizations from a collection of parameter sets resulted in a reasonable visual fit to the data.
-We are also inclined to believe that the use of multiple parameter values was in part intended to account for parameter uncertainty---the importance of which was discussed in the main text---an effort by the Model~1 authors that we applaud.
-Simulations from each of the parameter sets, however, were treated with equal importance when being used to diagnose the model fit and make inference on the system.
-This is problematic given Figures S8 and S9 of the supplement material, which suggest that some parameter sets that were used for inference may have been several hundred units of log-likelihood lower than other parameter sets that were simultaneously used to make forecasts.
-Such a large difference in log-likelihoods is well beyond the threshold of statistical uncertainty determined by Wilks' theorem, resulting in the equal use of statistically inferior parameter sets in order to make forecasts and conduct inference on the system.
-
-To fully reproduce the results of the Model~1 authors, it is necessary to use the exact same set of model parameters that were originally used to obtain the results presented by \cite{lee20}.
-Because these parameter sets were not made publicly available, we relied on the source code provided by the Model~1 authors to approximately recreate the parameter set.
-Due to software updates since the publication of the source code, we were unable to produce the exact same set of parameters.
-Running the publicly available source code, however, resulted in a set of parameters that are visually similar to those used by the Model~1 authors (See Figures~\ref{fig:PlotEpiDist} and \ref{fig:plotEndParams}).
-Furthermore, simulations using the set of parameters produced by the source code appear practically equivalent to those displayed by \cite{lee20} (See Figure~\ref{fig:plotMod1Sims}).
-
-<<Simulate Lee Model 1, echo=FALSE, cache=TRUE>>=
-h1_epi <- haiti1()
-N_LEE_SIMS <- 20
-
-registerDoRNG(18599687)
-foreach(
-  i = 1:length(h1LeeStartsEpi),
-  .combine = rbind
-) %dopar% {
-  guess <- h1LeeStartsEpi[i, ]
-  sims <- simulate(h1_epi, nsim = N_LEE_SIMS, format = 'data.frame',
-                   params = guess)  # 20 sims for each set of parameters
-  sims$param_set <- i
-  sims %>%
-    dplyr::select(.id, week, cases, param_set)
-} -> all_sims
-@
-
-<<PlotEpiDist, echo=FALSE, fig.cap="Compare to Figure S8 of Lee et al. (2020) supplement.", fig.width=7, fig.height=7, fig.align='center'>>=
-GGally::ggpairs(
-  h1LeeStartsEpi,
-  columns = c("rho", "tau", "beta1", "nu", "loglik")
-)
-@
-
-<<SimulateEpi, echo=FALSE, include=FALSE, message=FALSE, cache=TRUE>>=
-h1_end <- haiti1(period = "endemic")
-
-true_agg_cases <- haitiCholera %>%
-  select(-report) %>%
-  pivot_longer(
-    data = .,
-    cols = -date_saturday,
-    values_to = 'cases',
-    names_to = 'dep'
-  ) %>%
-  mutate(
-    date = as.Date(date_saturday),
-    dep = gsub("\\.", "_", dep)
-  ) %>%
-  mutate(
-    dep = case_when(dep == "Grand_Anse" ~ "Grande_Anse", TRUE ~ dep)
-  ) %>%
-  tidyr::pivot_wider(
-    data = .,
-    id_cols = c(date),
-    names_from = dep,
-    values_from = cases,
-    names_prefix = 'cases_'
-  ) %>%
-  mutate(
-    ReportedAll = cases_Artibonite + cases_Centre +
-      cases_Grande_Anse + cases_Nippes + cases_Nord +
-      cases_Nord_Est + cases_Ouest + cases_Sud +
-      cases_Sud_Est + cases_Nord_Ouest
-  )
-@
-
-<<FitEnd, echo=FALSE, message=FALSE, include=FALSE>>=
-foreach(
-  i = 1:nrow(h1LeeStartsEnd),
-  .combine = rbind
-) %dopar% {
-  r_params <- h1LeeStartsEnd[i, ][names(coef(h1_end))]
-  coef(h1_end) <- r_params
-
-  sims <- simulate(
-    h1_end, nsim = N_LEE_SIMS,
-    format = 'data.frame',
-    params = r_params
-  )
-  sims$param_set <- i
-  sims %>%
-    dplyr::select(.id, week, cases, param_set)
-} -> end_sims
-
-@
-
-<<plotEndParams, fig.width=7, fig.height=7, fig.align='center', fig.cap="Bivariate relationships between variables after fitting endemic period.
-Compare to S9 of Lee et al.
-(2020) supplement.", cache=TRUE, echo=FALSE>>=
-GGally::ggpairs(
-  h1LeeStartsEnd,
-  columns = c("rho", "tau", "beta1", "nu", "loglik")
-)
-@
-
-<<calcMod1LeeLikes, echo=FALSE, include=FALSE, message=FALSE>>=
-h1_epi_evals <- readRDS("../model1/run_level_3/lee1_epi_evals.rds")
-h1_end_evals <- readRDS("../model1/run_level_3/lee1_end_evals.rds")
-
-mod1_all_likes <- dplyr::inner_join(
-  x = h1_epi_evals,
-  y = h1_end_evals,
-  by = "parid"
-  ) %>%
-  dplyr::mutate(
-    joint_ll = pfLL.x + pfLL.y
-  )
-
-# Approximate the likelihood of the Lee et al (2020) version of Model 1 by
-# using the parameter sets that resulted in the maximum likelihood.
-mod1_lee_ll <- max(mod1_all_likes$joint_ll, na.rm = TRUE)
-@
-
-<<plotMod1Sims, echo=FALSE, fig.cap=paste0("Simulations using parameter sets that were generated by running source code provided by Lee et al (2020).
-Compare to Figure S7 of \\cite{lee20sup}.
-The upper bound for the likelihood of this model is ", round(mod1_lee_ll), "."), cache=TRUE, fig.width=6.5, fig.height=3.75>>=
-
-quants <- all_sims %>%
-  filter(!param_set %in% c(1, 2)) %>%  # Remove parameter sets inconsistent with Lee et al (2020).
-  group_by(week) %>%
-  summarize(
-    q025 = quantile(cases, probs = 0.025, na.rm = TRUE),
-    q50  = quantile(cases, probs = 0.500, na.rm = TRUE),
-    q975 = quantile(cases, probs = 0.975, na.rm = TRUE)
-  ) %>%
-  ungroup() %>%
-  mutate(date = lubridate::ymd("2010-10-16") + lubridate::weeks(week))
-
-quants_end <- end_sims %>%
-  group_by(week) %>%
-  summarize(
-    q025 = quantile(cases, probs = 0.025, na.rm = TRUE),
-    q50  = quantile(cases, probs = 0.500, na.rm = TRUE),
-    q975 = quantile(cases, probs = 0.975, na.rm = TRUE)
-  ) %>%
-  ungroup() %>%
-  mutate(date = lubridate::ymd("2010-10-16") + lubridate::weeks(week))
-
-ggplot() +
-  geom_point(data = true_agg_cases, aes(x = date, y = ReportedAll)) +
-  geom_line(data = quants, aes(x = date, y = q50), col = '#0b8a2b', size = 1) +
-  geom_ribbon(
-    data = quants,
-    aes(x = date, ymin = q025, ymax = q975),
-    alpha = 0.5, fill = '#0b9c30'
-  ) +
-  geom_line(data = quants_end, aes(x = date, y = q50), col = '#882871') +
-  geom_ribbon(data = quants_end, aes(x = date, ymin = q025, ymax = q975), alpha = 0.5, fill = '#882871') +
-  ylab("Reported, Symptomatic Incidence") +
-  scale_x_date(date_breaks = "1 year") +
-  theme(axis.title.x = element_blank(),
-        axis.text.x = element_text(hjust = 1, angle = 45))
-@
-
-Because the model forecasts provided by \cite{lee20} come from various sets of parameters---which each correspond to a unique log-likelihood value---it is not obvious how one would obtain an estimate for the log-likelihood of the model that was used for simulations by the Model~1 authors.
-One approach could be to calculate the logarithm of the weighted mean of the likelihoods for each parameter sets used to obtain the forecasts, where the weights are proportional to the number of times the parameter set was used.
-However, in an effort to not underestimate the likelihood of the model of the Model~1 authors, we report the estimated log-likelihood as the log-likelihood value corresponding to the parameter set with the largest likelihood value, even though the majority of simulations were obtained using parameter sets with lower likelihood values.
-In this sense, we consider the log-likelihood reported in Table~1 of the main text to be an upper-bound of the log-likelihood of the model used by \cite{lee20}.
-For each parameter set, the log-likelihood was estimated using a particle filter, implemented as the \texttt{pfilter} function in the \texttt{pomp} package.
-
-\subsection{Model~2 Replication}\label{sec:mod2rep}
-
-Model~2 was developed by a team that consisted of members from the Fred Hutchinson Cancer Research Center and the University of Florida (hereafter referred to as the Model~2 authors).
-While Model~2 is the only deterministic model we considered in our analysis, it contains perhaps the most complex descriptions of cholera in Haiti: Model~2 accounts for movement between spatial units; human-to-human and environment-to-human cholera infections; and transfer of water between spatial units based on elevation charts and river flows.
-
-The source code that the Model~2 authors used to generate their results was written in the \code{Python} programming language, and is publicly available at \url{10.5281/zenodo.3360857} and its accompanying GitHub repository \url{https://github.com/lulelita/HaitiCholeraMultiModelingProject}.
-In order to perform our analysis in a unified framework, we re-implemented this model in the \code{R} programming language using the \code{spatPomp} package \citep{asfaw23arxiv}, which facilitates the creation of meta-population models.
-We note that the travel and water matrices used to implement the complex dynamics in Model~2 \citep{lee20sup} are not available in either the Zenodo archive or the GitHub repository;
-instead, we obtained these matrices via personal correspondence with the Model~2 authors.
-Using these matrices, and the point estimates for model parameters provided by \citep{lee20sup}, we created trajectories of the cholera dynamics and compared this to available data.
-These trajectories, shown in Figure~\ref{fig:mod2rep}, are very similar to the trajectories shown in Figure~S15 of \cite{lee20sup}.
-
-<<Model2Replication, cache=TRUE, echo=FALSE>>=
-h2_epi <- haiti2()
-h2_end <- haiti2(region = 'after')
-
-h2_epi_traj <- trajectory(h2_epi, format = 'data.frame')
-h2_end_traj <- trajectory(h2_end, format = 'data.frame')
-
-h2_epi_traj$Ctotal <- rowSums(h2_epi_traj[, paste0("C", 1:10)]) * 0.2
-h2_end_traj$Ctotal <- rowSums(h2_end_traj[, paste0("C", 1:10)]) * 0.2
-
-dep_plot_df <- haitipkg::haitiCholera %>%
-  select(-report) %>%
-  tidyr::pivot_longer(
-    data = .,
-    cols = -date_saturday,
-    values_to = 'cases',
-    names_to = 'dep'
-  ) %>%
-  mutate(
-    date = as.Date(date_saturday),
-    dep = gsub("\\.", "_", dep)
-  ) %>%
-  mutate(
-    dep = case_when(dep == "Grand_Anse" ~ "Grande_Anse", TRUE ~ dep)
-  )
-
-true_agg_cases <- dep_plot_df %>%
-  tidyr::pivot_wider(
-    data = .,
-    id_cols = c(date),
-    names_from = dep,
-    values_from = cases,
-    names_prefix = 'cases_'
-  ) %>%
-  mutate(
-    ReportedAll = cases_Artibonite + cases_Centre +
-      cases_Grande_Anse + cases_Nippes + cases_Nord +
-      cases_Nord_Est + cases_Ouest + cases_Sud +
-      cases_Sud_Est + cases_Nord_Ouest
-  ) %>%
-  select(date, ReportedAll)
-
-h2_epi_traj$date <- as.Date(
-  lubridate::round_date(
-    lubridate::date_decimal(h2_epi_traj$year),
-    unit = 'day'
-  )
-)
-
-h2_end_traj$date <- as.Date(
-  lubridate::round_date(
-    lubridate::date_decimal(h2_end_traj$year),
-    unit = 'day'
-  )
-)
-
-all_traj <- bind_rows(
-  select(h2_epi_traj, date, Ctotal),
-  select(h2_end_traj, date, Ctotal)
-)
-
-projections <- left_join(
-  x = true_agg_cases,
-  y = all_traj,
-  by = 'date'
-)
-@
-
-\begin{figure}[ht]
-<<plotModel2Rep, cache=TRUE, echo=FALSE, fig.width=6.5, fig.height=4.5>>=
-gg_epi <- projections %>%
-  filter(date < as.Date("03-01-2014", format = '%m-%d-%Y')) %>%
-  ggplot(aes(x = date)) +
-  geom_line(aes(y = ReportedAll), col = 'black', linetype = 'dashed') +
-  geom_line(aes(y = Ctotal), col = 'red', size = 1) +
-  theme_bw() +
-  scale_y_continuous(limits = c(0, 30000), breaks = seq(0, 30000, 5000)) +
-  theme(axis.title.x = element_blank()) +
-  ylab("Reported number of cases")
-
-gg_end <- projections %>%
-  filter(date >= as.Date("03-01-2014", format = '%m-%d-%Y')) %>%
-  ggplot(aes(x = date)) +
-  geom_line(aes(y = ReportedAll), col = 'black', linetype = 'dashed') +
-  geom_line(aes(y = Ctotal), col = 'red', size = 1) +
-  theme_bw() +
-  scale_y_continuous(limits = c(0, 7000), breaks = seq(0, 7000, 1000)) +
-  theme(axis.title.x = element_blank()) +
-  ylab("Reported number of cases") +
-  scale_x_date(date_breaks = "1 year", date_labels = "%Y")
-
-cowplot::plot_grid(gg_epi, gg_end, ncol = 1)
-@
-\caption{\label{fig:mod2rep}
-Model 2 trajectories using the \code{haitipkg}. Compare to Figure S15 of \cite{lee20sup}.
-}
-\end{figure}
-
-There are minor differences between Figure~\ref{fig:mod2rep} and Figure~S15 of \cite{lee20sup}.
-While the discrepancy appears minor, the deterministic nature of Model~2 implies that an exact replication of model trajectories should be possible.
-In this case, these discrepancies may possibly be attributed to implementing the model and plotting the model trajectory in two different programming languages.
-Another potential explanation for the discrepancy is that the parameters that we used are only approximately the same as those used by \citet{lee20sup}.
-For example, the parameters $\transmission$, $\Wbeta{}$ (See Table~\ref{tab:translate}) had reported values of $9.9 \times 10^{-7}$ and $4.03 \times 10^{-2}$, respectively (Table~S13 of \cite{lee20sup}), but were actually fit to data and therefore likely these values have been rounded.
-Additionally, our implementation of Model~2 used a time scale of years and many of the parameters were reported on a weekly scale, so small differences may result due to unit conversions.
-The collective effect of these small differences in model parameters likely will result in small differences in model trajectories.
-
-Some additional concerns about being able to accurately replicate the results of \cite{lee20} are valid.
-Details about the measurement models and how latent states were initialized for the epidemic model were not provided by \citet{lee20sup} and therefore these details must be inferred by looking at the provided source code.
-According to repository comments, the files \code{fit\allowbreak In\allowbreak Pieces\allowbreak 3params\allowbreak Clean\allowbreak May2019\allowbreak Public.py} and \code{fit\allowbreak In\allowbreak Pieces\allowbreak Mu\allowbreak With\allowbreak Frac\allowbreak Sus\allowbreak Fixed\allowbreak All\allowbreak Infections\allowbreak Public.py} were used to fit the epidemic and endemic phases of the model respectively, although it is apparent that these exact files were not used to obtain the reported results since the files contain some variable-naming errors that make it impossible to run the files without making modifications \footnote{One example of why the code cannot be run that the file loads functions from a non-extant file named \code{choleraEqs.py} in line 13 rather than \code{cholera\allowbreak Eqs\allowbreak Public.py}.}.
-The inability to replicate the results by \citet{lee20} by running the provided source code makes checking whether or not a our numeric implementation faithfully represents their results very difficult.
-Additionally, the script that was said to been used to obtain the results reported by \cite{lee20} appears to use a different measurement model than what was described in the supplemental material, again making it difficult to fully replicate the result of \cite{lee20} without being able to easily run the provided source code.
-In this case, we chose to use measurement model described by Eq.~\ref{model2:measEpi} for both phases of the epidemic, as this seemed to visually match the results of \cite{lee20} most closely.
-
-\subsection{Model~3 Replication}
-
-Model~3 was developed by a team of researchers at the Laboratory of the Swiss Federal Institute of Technology in Lausanne, hereafter referred to as the Model~3 authors.
-The code that was originally used to implement Model~3 is archived with the DOI: \url{10.5281/zenodo.3360723}, and also available in the public GitHub repository: \code{jcblemai/haiti-mass-ocv-campaign}.
-Because the code was made publicly available, and final model parameters were reported in the supplementary material of \cite{lee20}, we were able to reproduce Model~3 by directly using the source code.
-In Fig.~\ref{fig:mod3rep}, we plot simulations from this model.
-This figure can be compared to Figure S18 of \cite{lee20}.
-We note that slight differences may be accounted for due to variance in the model simulations and the difference in programming language used to produce the figure.
-Overall, the high standard of reproducibility that was achieved by the Model~3 authors facilitated the ability to readily replicate their model and results.
-
-<<Model3Replication, echo=FALSE, cache=TRUE>>=
-NSIM <- 500
-
-plot_order <- c(
-  'Artibonite',
-  'Sud_Est',
-  'Nippes',
-  'Nord_Est',
-  'Ouest',
-  'Centre',
-  'Nord',
-  'Sud',
-  'Nord_Ouest',
-  'Grande_Anse'
-)
-
-dep_labeller <- as_labeller(
-  c(
-    'Artibonite' = 'Artibonite',
-    'Sud_Est' = 'Sud-Est',
-    'Sud.Est' = 'Sud-Est',
-    'Nippes' = 'Nippes',
-    'Nord_Est' = 'Nord-Est',
-    'Nord.Est' = 'Nord-Est',
-    'Ouest' = 'Ouest',
-    'Centre' = 'Centre',
-    'Nord' = 'Nord',
-    'Sud' = 'Sud',
-    'Nord_Ouest' = 'Nord-Ouest',
-    'Nord.Ouest' = 'Nord-Ouest',
-    'Grande_Anse' = 'Grand\'Anse',
-    'Grand.Anse' = 'Grand\'Anse'
-  )
-)
-
-lee3 <- lee3_spatPomp()
-lee3_sims_quants <- bake("../model3/si/lee3Sims.rds", {
-
-  # Simulate from lee et al 2020 model
-  sims_temp <- simulate(
-    lee3, nsim = NSIM,
-    seed = 321, format = 'data.frame'
-  )
-
-  # Get desired quantiles of reported cases from simulations
-  quants <- sims_temp %>%
-    dplyr::group_by(unitname, time) %>%
-    dplyr::summarise(
-      q05 = quantile(cases, 0.025, na.rm = T),
-      mean = mean(cases, na.rm = T),
-      q50 = quantile(cases, 0.5, na.rm = T),
-      q95 = quantile(cases, 0.975, na.rm = T)
-    ) %>%
-    dplyr::ungroup() %>%
-    dplyr::mutate(
-      date = lubridate::date_decimal(time),
-      date = as.Date(lubridate::round_date(date))
-    ) %>%
-    dplyr::rename(dep = unitname)
-
-  # Remove cases from Ouest that were not considered by Lee et al. (2020)
-  quants %>%
-    dplyr::filter(
-      !(dep == "Ouest" & date < '2017-06-10')
-    )
-
-})
-
-# Get the subset of cholera cases that match simulations from Lee et al. (2020) model.
-sub_cases <- haitiCholera %>%
-  dplyr::mutate(date = as.Date(date_saturday)) %>%
-  dplyr::select(-report, -date_saturday) %>%
-  dplyr::filter(lubridate::decimal_date(date) >= min(lee3@times)) %>%
-  tidyr::pivot_longer(
-    data = .,
-    cols = -date,
-    names_to = "dep",
-    values_to = "cases"
-  ) %>%
-  dplyr::mutate(
-    dep = dplyr::case_when(
-      dep == "Grand.Anse" ~ "Grande_Anse",
-      dep == "Nord.Est" ~ "Nord_Est",
-      dep == "Nord.Ouest" ~ "Nord_Ouest",
-      dep == "Sud.Est" ~ "Sud_Est",
-      TRUE ~ dep
-    )
-  )
-@
-
-\begin{figure}[ht]
-<<PlotMod3Rep, echo=FALSE, message=FALSE, fig.height=3.75>>=
-ggplot() +
-  # Region for upper and lower simulation bounds
-  geom_ribbon(
-    data = lee3_sims_quants,
-    aes(x = date, ymin = q05, ymax = q95),
-    fill = 'darkblue', alpha = 0.5
-  ) +
-  # Median line
-  geom_line(
-    data = lee3_sims_quants,
-    aes(x = date, y = q50),
-    col = 'darkblue'
-  ) +
-  # Observed data
-  geom_point(
-    data = dplyr::filter(sub_cases, cases < 500),
-    aes(x = date, y = cases), size = 0.2
-  ) +
-  facet_wrap(
-    ~factor(dep, levels = plot_order),
-    nrow = 2, labeller = dep_labeller
-  ) +
-  labs(y = 'Reported Cholera Cases') +
-  scale_x_date(date_breaks = "6 months", date_labels = "%Y-%m") +
-  theme(
-    axis.text.x = element_text(angle = 40, hjust = 1, vjust = 1, size = 7),
-    axis.title.x = element_blank()
-  )
-@
-\caption{\label{fig:mod3rep}
-Simulations from Model~3.
-Compare to Figure S18 of \cite{lee20}.
-}
-\end{figure}
-
-<<lee3Cleanup, echo=FALSE, include=FALSE, message=FALSE>>=
-rm(lee3, lee3_sims_quants, sub_cases)
-gc()
-@
-
-\section{Forecasting with parameter uncertainty}
-
-Let $f_{Y_{1:N}}(y_{1:N} | \theta)$ denote the pdf of the model under consideration, were $\theta$ is a parameter vector that indexes the model.
-Furthermore, denote the observed data as $y_{1:N}^*$.
-Because the uncertainty in just a single parameter can lead to drastically different forecasts \citep{saltelli20},
-parameter uncertainty should be considered when obtaining model forecasts when the goal is to influencing policy.
-In a Bayesian modelling paradigm, the most natural way to account for parameter uncertainty in model forecasts is to suppose that $\theta$ comes from a distribution $f_{\Theta}$, and then to obtain $J$ forecasts from the model where each forecast is obtained using parameters drawn from the posterior distribution $\Theta_{1:J} \mid Y_{1:N} = y_{1:N}^* \sim f_{\Theta}\big(\theta | Y_{1:N} = y_{1:N}^*\big)$.
-
-When frequentist methods are used, however, there does not exist a posterior distribution from which one could sample.
-A common approach could be to obtain a weighted average of the simulations from various models \citep{hoeting99}, but this can be problematic when forecasts from each model are very different from each other \citep{grueber11}.
-A similar approach that has been taken \citep{king15} is to obtain model forecasts using multiple sets of parameter values and then sample from the resulting forecasts using weights proportional to the corresponding likelihoods of the parameter values.
-This approach could be considered as empirical Bayes, as it is equivalent to using a discrete uniform prior where the set of values in the prior distribution is determined by a stochastic routine applied to the observed data, as discussed below.
-
-For each $i \in 1:k$, let $\Theta_i$ be a random vector of model parameters.
-Then, letting $\Theta$ denote the true model parameters, we endow the set $\{\Theta_1, \Theta_2, \ldots, \Theta_k\}$ with a discrete uniform distribution, such that $P\big(\Theta = \Theta_i\big) = \frac{1}{K}$ for all values $i \in \seq{1}{k}$.
-Using this as a prior distribution, the posterior distribution of $\Theta | Y_{1:N} = y_{1:N}^*$ can be expressed as:
-$P\big(\Theta = \Theta_k | Y_{1:N} = y_{1:N}^*\big) = \frac{f_{Y_{1:N}}(y_{1:N}^*| \Theta_k)}{\sum_{l = 1}^K f_{Y_{1:N}}(y_{1:N}^*| \theta_l)}$.
-In a standard empirical Bayes analysis, the values $\Theta_1, \ldots, \Theta_k$ of the prior distribution would be chosen using the observed data, resulting in a posterior distribution that weighs the prior parameter vectors proportional to their corresponding likelihoods.
-Instead, we choose $\Theta_i$ to be the output of a stochastic routine applied to the observed data by setting $\Theta_i$ to be the output of an iterated filtering algorithm.
-In practice, because the likelihood maximization routines of iterated filtering methods are stochastic, it is common to run the iterated filtering method multiple times for each model in order to obtain a maximum likelihood estimate for model parameters.
-This results in a natural set of parameters near the MLE that could be used as the discrete prior distribution.
-
-\section{Translating to \citet{lee20} notation}
-
-Since the models of \citet{lee20} were developed independently, the choice of notation varies inconsistently between models.
-For our reanalysis, we rename parameters to provide a unified notation facilitating comparison between models.
-Table~\ref{tab:translate} maps this notation back to the original notations, for reference.
-
-\begin{table}
-  \begin{center}
-  \begin{tabular}{|c|c|c|c|c|}\hline
-    \multirow{2}{*}{Parameter} & Our & \multicolumn{3}{c|}{Lee et al. (2020a)} \\\cline{3-5}
-     & Notation & 1 & 2 & 3 \\
-    \hline
-    \hline
-    % Table begins here
-    Reporting Rate & $\reportRate$ & $\rho$ & $\rho$ & $\epsilon_1, \epsilon_2$ \\\hline
-    Mixing Coefficient & $\mixExponent$ & $\nu$ & --- & --- \\\hline
-    Measurement Over-Dispersion & $\obsOverdispersion$ & $\tau$ & --- & $p$ \\\hline
-    Birth Rate & $\muBirth$ & $\mu$ & --- & --- \\\hline
-    Natural Mortality Rate & $\muDeath$ & $\delta$ & --- & $\mu$ \\\hline
-    Cholera Mortality Rate & $\choleraDeath$ & --- & --- & $\alpha$ \\\hline
-    Latent Period & $1/\muEI$ & $1/\sigma$ & $1/\gamma_E$ & --- \\\hline
-    Recovery Rate & $\muIR$ & $\gamma$ & $\gamma$ & $\gamma$ \\\hline
-    Loss of Immunity & $\muRS$ & $\alpha$ & $\sigma$ & $\rho$ \\\hline
-    Symptomatic Ratio & $\symptomFrac$ & $1 - \theta_0$ & $k$ & $\sigma$ \\\hline
-    Asymptomatic Relative Infectiousness & $\asymptomRelativeInfect$ & $\kappa$ & $red_\beta$ & --- \\\hline
-    Human-to-Water Shedding & $\Wshed$ & --- & $\mu$ & $\theta_I$ \\\hline
-    Asymptomatic Relative Shedding & $\asymptomRelativeShed$ & --- & $red_\mu$ & $\theta_A/\theta_I$ \\\hline
-    Seasonal Amplitude & $\seasAmplitude$ & --- & $\alpha_s$ & $\lambda$ \\\hline
-    Transmission & $\transmission$ & $\beta$ & $\beta$ & $c$ \\\hline
-    Water-to-Human & $\Wbeta{}$ & --- & $\beta_W$ & $\beta$ \\\hline
-    Bacteria Mortality & $\Wremoval$ & --- & $\delta$ & $\mu_\beta$ \\\hline
-    Vaccination Efficacy & $\vaccineEfficacy$ & $\theta_{vk}$ & $\theta_1, \theta_2, \theta_{1_5}, \theta_{2_5}$ & $\eta_{1d}, \eta_{2d}$ \\\hline
-    Process Over-dispersion & $\sigmaProc$ & --- & --- & $\sigma_w$\\\hline
-    % Half Saturation Constant & $\Wsat$ & --- & $Sat$ & $K$ \\\hline
-  \end{tabular}
-  \end{center}
-  \caption{
-  \label{tab:translate}Translations between our common notation and notation used by \cite{lee20}
-  }
-\end{table}
-
-\clearpage
-
-\bibliography{../bib-haiti}
-
-\end{document}

### 2023-09-05 si/ReplicateLee20.Rnw

--- 2023-08-31 si/si.Rnw
@@
 
-\pdfminorversion=4
+\input{inputs/header}
 
-\newcommand\secTitleSpace{\hspace{3mm}}
+\setcounter{tocdepth}{1}
 
@@
 library(doRNG)
+library(knitr)
+
+options(
+  scipen = 2,
+  help_type = "html",
+  stringsAsFactors = FALSE,
+  continue = "+  ",
+  width = 70,
+  useFancyQuotes = FALSE,
+  reindent.spaces = 2,
+  xtable.comment = FALSE
+)
+
+opts_chunk$set(
+  tidy.opts=list(
+    keep.blank.line=FALSE
+  ),
+  comment="",
+  fig.path = 'figure/',
+  # warning=FALSE,
+  # message=FALSE,
+  echo=FALSE,
+  fig.align="center",
+  dev="pdf",
+  dev.args=list(
+    bg="transparent",
+    pointsize=9
+  )
+)
 
@@
 
-<<set-opts, include=FALSE>>=
-options(
-  scipen = 2,
-  help_type = "html",
-  stringsAsFactors = FALSE,
-  continue = "+  ",
-  width = 70,
-  useFancyQuotes = FALSE,
-  reindent.spaces = 2,
-  xtable.comment = FALSE
-)
-@
-
-<<knitr-opts, include=FALSE, purl=FALSE>>=
-library(knitr)
-opts_knit$set(concordance=TRUE)
-opts_chunk$set(
-  progress=TRUE,prompt=TRUE,highlight=FALSE,
-  tidy=TRUE,
-  tidy.opts=list(
-    keep.blank.line=FALSE
-  ),
-  comment="",
-  warning=FALSE,
-  message=FALSE,
-  error=TRUE,
-  echo=FALSE,
-  cache=FALSE,
-  strip.white=TRUE,
-  results="markup",
-  fig.path="figure/",
-  fig.lp="fig:",
-  fig.align="center",
-  fig.show="asis",
-  dev="pdf",
-  dev.args=list(
-    bg="transparent",
-    pointsize=9
-  )
-)
-
-myround <- function (x, digits = 1) {
-  # taken from the broman package
-  if (digits < 1)
-    stop("This is intended for the case digits >= 1.")
-  if (length(digits) > 1) {
-    digits <- digits[1]
-    warning("Using only digits[1]")
-  }
-  tmp <- sprintf(paste("%.", digits, "f", sep = ""), x)
-  zero <- paste0("0.", paste(rep("0", digits), collapse = ""))
-  tmp[tmp == paste0("-", zero)] <- zero
-  tmp
-}
-
-@
-
-%% copied from pnas-new.cls
-\usepackage{amsmath,amsfonts,amssymb,graphicx}
-
-%%%% parameters %%%%%%%%%%5
-\newcommand\Wsat{W_{\mathrm{sat}}}
-\newcommand\muIR{\mu_{IR}}
-\newcommand\muEI{\mu_{EI}}
-\newcommand\transmission{\beta}
-\newcommand\seasAmplitude{a}
-\newcommand\rainfallExponent{r}
-\newcommand\muRS{\mu_{RS}}
-\newcommand\vaccineEfficacy{\theta}
-\newcommand\muBirth{\mu_S}
-\newcommand\muDeath{\delta}
-\newcommand\choleraDeath{\delta_{C}}
-\newcommand\symptomFrac{f}
-\newcommand\asymptomRelativeInfect{\epsilon}
-\newcommand\asymptomRelativeShed{\epsilon_{W}}
-\newcommand\Wbeta[1]{\beta_{W#1}}
-\newcommand\Wremoval{\delta_W}
-\newcommand\Wshed{\mu_W}
-\newcommand\mixExponent{\nu}
-\newcommand\sigmaProc{\sigma_{\mathrm{proc}}}
-\newcommand\reportRate{\rho}
-\newcommand\obsOverdispersion{\psi}
-\newcommand\phaseParm{\phi}
-\newcommand\transmissionTrend{\zeta}
-\newcommand\Binit{\xi}
-\newcommand\vaccClass{Z}
-\newcommand\vaccCounter{z}
-\newcommand\modelCounter{m}
-\newcommand\Whur[1]{\beta_{W#1}^{hm}}
-\newcommand\hHur[1]{h_{#1}^{hm}}
-\newcommand\Iinit{I_{0,0}}
-\newcommand\Einit{E_{0, 0}}
-\newcommand\reportChange{\delta}
-
-\newcommand\missing{\hspace{6mm} ---}
-\newcommand\fixed{\color{blue}}
-\newcommand\demography{\bullet}
-
-\newcommand\mytitle{Informing policy via dynamic models: Cholera in Haiti}
-\newcommand\comma{{\hspace{-0.25mm},\hspace{-0.2mm}}}
-
-%%%%%%%%%%%%%%%% START OF USER DEFINITIONS %%%%%%%%%%%%%%%5
-
-\newcommand\siOnly[1]{} %% redefined for the supplement
-\newcommand\msOnly[1]{#1} %% redefined for the supplement
-
-\newcommand\code[1]{\texttt{#1}}
-\usepackage{xurl}
-% \usepackage[breaklinks]{hyperref}
-\usepackage{makecell}
-\usepackage{multirow}
-\usepackage{multicol}
-\usepackage[ruled,noline,linesnumbered]{algorithm2e}
-% \usepackage{setspace}
-
-%% EDITING MACROS
-\usepackage{color}
-\usepackage[normalem]{ulem}% to use \sout in feedback commands
-\usepackage{bm}
-% orange for EI
-\definecolor{orange}{rgb}{1,0.5,0}
-\newcommand\ei[2]{\sout{#1} \textcolor{orange}{#2}}
-\newcommand\eic[1]{\textcolor{orange}{[#1]}}
-% green for JW
-\definecolor{green}{rgb}{0,0.5,0}
-\newcommand\jw[2]{\sout{#1} \textcolor{green}{#2}}
-\newcommand\jwc[1]{\textcolor{green}{[#1]}}
-% purple for JJ
-\definecolor{purple}{rgb}{0.5,0,1}
-\newcommand\jj[2]{\sout{#1} \textcolor{purple}{#2}}
-\newcommand\jjc[1]{\textcolor{purple}{[#1]}}
-% cyan for AR
-\definecolor{cyan}{rgb}{0,.5,.5}
-\newcommand\ar[2]{\sout{#1} \textcolor{cyan}{#2}}
-\newcommand\arc[1]{\textcolor{cyan}{#1}}
-% light brown for KT
-\definecolor{lightbrown}{rgb}{0.5,0.5,0}
-\newcommand\kt[2]{\sout{#1} \textcolor{lightbrown}{#2}}
-\newcommand\ktc[1]{\textcolor{lightbrown}{#1}}
-
-%% basic POMP definitions
-\newcommand\Xspace{{\mathbb X}}
-\newcommand\Yspace{{\mathbb Y}}
-\newcommand\Thetaspace{\R^{\Thetadim}}
-\newcommand\Xunitspace{{\scriptsize{\mathcal X}}}
-\newcommand\Yunitspace{{\tiny{\mathcal Y}}}
-\newcommand\hatTheta{\widehat{\Theta}}
-\newcommand\Xdim{{\mathrm{dim}}(\Xspace)}
-\newcommand\Ydim{{\mathrm{dim}}(\Yspace)}
-\newcommand\Thetadim{P}
-\newcommand\thetadim{p}
-\newcommand\timeSet{{\mathbb T}}
-\newcommand\data[1]{#1^*}
-
-\newcommand\unitSpecific{\hspace{0.1mm}\mathrm{us}}
-\newcommand\shared{\hspace{0.15mm}\mathrm{sh}}
-
-% for algorithm pseudocode
-\SetKwFor{For}{for}{:}{end}
-
-%% for particle filters
-\newcommand\unit{u}
-\newcommand\altUnit{\tilde u}
-\newcommand\Unit{U}
-\newcommand\rep{i}
-\newcommand\Rep{\mathcal{I}}
-\newcommand\island{\rep}
-\newcommand\Island{\Rep}
-%\newcommand\txtisland{replicate}
-\newcommand\altIsland{\tilde i}
-\usepackage[mathscr]{euscript}
-
-\renewcommand\time{n}
-\newcommand\myvec[1]{\boldsymbol{#1}}
-\newcommand\altTime{\tilde n}
-\newcommand\Time{N}
-\newcommand\Np{J}
-\newcommand\np{j}
-\newcommand\altNp{k}
-\newcommand\altAltNp{\tilde j}
-\newcommand\resampleIndex{r}
-\newcommand\unitWeight{w}
-
-%% for all iterated filtering algorithms
-\newcommand\Nit{M}
-\newcommand\nit{m}
-
-%% customized math macros
-\newcommand\prob{\mathbb{P}}
-\newcommand\E{\mathbb{E}}
-\newcommand\dd[1]{\mathrm{d}{#1}}
-\newcommand\given{{\,\vert\,}}
-\newcommand\equals{{{\,}={\,}}}
-\newcommand\myequals{\hspace{0.5mm}{=}\hspace{0.5mm}}
-\newcommand\myto{{\;:\;}}
-\newcommand\seq[2]{{#1}\!:\!{#2}}
-\newcommand\mydot{{\,\cdot\,}}
-\newcommand\cp[2]{N_{\mathrm{#1}\mathrm{#2}}}
-\newcommand\giventh{{\hspace{0.5mm};\hspace{0.5mm}}}
-\newcommand\normal{\mathcal{N}}
-\newcommand\argequals{{\,=\,}}
-\newcommand\lags{c}
-\newcommand\maxlag{\overline{c}}
-\newcommand\nlfList{C}
-\newcommand\bigO{\mathcal{O}}
-\newcommand\loglik{\lambda}
-\newcommand\loglikMC{\MC{\loglik}}
-\newcommand\R{\mathbb{R}}
-\newcommand\param{\,;}
-\newcommand\mycolon{{\hspace{0.6mm}:\hspace{0.6mm}}}
-\newcommand\MC[1]{#1^{\,\mbox{\tiny MC}}}
-\newcommand\EMC{{\E}}
-\newcommand\Var{\mathrm{Var}}
-\newcommand\var{\Var}
-\newcommand\Cov{\mathrm{Cov}}
-\newcommand\cov{\Cov}
-\newcommand\iid{\mathrm{iid}}
-\newcommand\dist{d}
-\newcommand\transpose{\top}
-
-\newcommand\gnsep{,}
-
-\def\lik{L}
-\def\loglik{\ell}
-
-%% THEOREM-LIKE ENVIRONMENTS
-\usepackage{amsthm}
-\newtheorem{prop}{Proposition}
-\newtheorem{theorem}{Theorem}
-\newtheorem{lemma}{Lemma}
-\newtheorem{example}{Example}
-\newtheorem{remark}{Remark}
-\newtheorem{cor}{Corrolary}
-\newtheorem{fact}{Fact}
-\newtheorem{assumption}{Assumption}
-\newtheorem{assumptionB}{Assumption}
-\renewcommand\theassumption{A\arabic{assumption}}
-\renewcommand\theassumptionB{B\arabic{assumptionB}}
-
-%% FOR PSEUDOCODE SETUP
-\newcommand\mystretch{\rule[-2mm]{0mm}{5mm} }
-\newcommand\asp{\hspace{4mm}}
-\newcommand\codeIndent{\hspace{4mm}}
-
-%%%%%%%%%%%%%%%% END OF USER DEFINITIONSS %%%%%%%%%%%%%%%
-
-\bibliographystyle{apalike}
-\usepackage{natbib}
-\usepackage{fullpage}
-
-\newcommand\myeqref[1]{(\ref{#1})}
-
-\renewcommand{\contentsname}{Supplementary Content}
-\renewcommand{\refname}{Supplementary References}
-\renewcommand\thefigure{S-\arabic{figure}}
-\renewcommand\thetable{S-\arabic{table}}
-\renewcommand\thesection{S\arabic{section}}
-\renewcommand\theequation{S\arabic{equation}}
-\renewcommand\theprop{S\arabic{prop}}
-\renewcommand\thelemma{S\arabic{lemma}}
-\renewcommand\thealgocf{S\arabic{algocf}}
-
-\setcounter{tocdepth}{1}
-
 \begin{document}
 
-\date{\today}
-\title{Supplement to ``{\mytitle}''}
-  \author{Jesse Wheeler, AnnaElaine L. Rosengart, Zhuoxun Jiang, \\ Kevin Tan, Noah Treutle and Edward L. Ionides
- \\ \hspace{.2cm}\\
-    Department of Statistics, University of Michigan\\
-}
-
-\newcommand{\blind}{1}
-
-\if1\blind
-{
-\maketitle
-}
-\fi
-
-\if0\blind
-{
-  \bigskip
-  \bigskip
-  \bigskip
-  \begin{center}
-    {\LARGE\bf \mytitle}
-\end{center}
-  \bigskip
-  \bigskip
-}\fi
-
-\tableofcontents
-
-\newpage
-
-\section{Markov chain and differential equation interpretations of compartment flow rates}
-
-In Sections 2.1, 2.2 and 2.3 of the main article, we define compartment models in terms of their flow rates.
-For a discrete population model, these rates define a Markov chain.
-For a continuous and deterministic model, the rates define a system of ordinary differential equations.
-Here, we add additional details to clarify the mapping from a collection of rate functions to a fully specified process.
-Our treatment follows \citet{breto09}.
-
-A general compartment model is a vector-valued process $X(t)=(X_1(t),\dots,X_c(t))$ denoting the (integer or real-valued) counts in each of $c$ compartments, where $t$ is any continuous value in the interval  $\left[t_0, \infty\right)$ for some real valued starting time $t_0$.
-The compartments may also have names, but to set up general notation we simply refer to them by their numerical index.
-The basic characteristic of a compartment model is that $X(t)$ can be written in terms of the flows $N_{ij}(t)$ from $i$ to $j$.
-A flow into compartment $i$ from outside the system is denoted by $N_{\demography i}$, and a flow out of the system from compartment $i$ is denoted by $N_{i\demography}$.
-We call $\demography$ a source/sink compartment, though it is an irregular compartment since $X_{\demography}(t)$ is not defined.
-These flows are required to satisfy a ``conservation of mass'' identity:
-\begin{equation}
-X_i(t)=X_i(t_0)+N_{\demography i}(t) - N_{i\demography}(t) + \sum_{j\neq
-i}N_{ji}(t)-\sum_{j\neq i}N_{ij}(t). \label{eq:conservation}
-\end{equation}
-Each {\em flow} $N_{ij}(t)$ is associated with a {\em rate} function
-$\mu_{ij}=\mu_{ij}(t,X(t))$, where we include the possibility that $i$ or $j$ takes value $\demography$.
-
-There are different ways to use a collection of rate functions to build a fully specified model.
-We proceed to describe the ones we use in this paper: via a system of ordinary differential equations (Sec.~\ref{subsec:ode}), a simple Markov counting system (Sec.~\ref{subsec:smcs}), and an overdispersed Markov counting system (Sec.~\ref{subsec:odmcs}).
-Other representations include stochastic differential equations driven by Gaussian noise or Gamma noise \citep{bhadra11}.
-
-\subsection{Ordinary differential equation (ODE) interpretation}
-\label{subsec:ode}
-
-A basic deterministic specification is
-\begin{equation}
-\label{eq:ode1}
-dN_{ij}/dt = \mu_{ij}\big(t,X(t) \big) X_i(t), \hspace{3mm} i\in \seq{1}{c}, \hspace{3mm} j\in \seq{1}{c} \cup\{\demography\}, \hspace{3mm} i\neq j,
-\end{equation}
-where $\mu_{ij}\big(t,X(t)\big)$ is called a per-capita rate or a unit rate.
-Flows into the system require special treatment since $X_i(t)$ in \myeqref{eq:ode1} is not defined for $i=\demography$.
-Instead, we specify
-\begin{equation}
-\label{eq:ode2}
-dN_{\demography i}/dt = \mu_{\demography i}\big(t,X(t) \big).
-\end{equation}
-This is the the interpretation and implementation used for Model~2 in our study.
-
-\subsection{Simple Markov counting system interpretation}
-\label{subsec:smcs}
-A continuous time Markov chain can be specified via its infinitesimal transition probabilities.
-A basic approach to this is to define
-\begin{eqnarray}
-\label{eq:smcs1}
-\prob\big[ N_{ij}(t+\delta)-N_{ij}(t)=0 \given X(t)\big]
- &=& 1-\delta \mu_{ij}\big(t,X(t)\big)X_i(t) + o(\delta),
-\\
-\label{eq:smcs2}
-\prob\big[ N_{ij}(t+\delta)-N_{ij}(t)=1 \given X(t)\big]
- &=& \delta \mu_{ij}\big(t,X(t)\big)X_i(t) + o(\delta),
-\end{eqnarray}
-for $i\in \seq{1}{c}$ and $j\in\seq{1}{c}\cup\{\demography\}$ with $i\neq j$.
-As with the ODE case, we need special attention for flows into the system, and we define
-\begin{eqnarray}
-\label{eq:smcs3}
-\prob\big[ N_{\demography i}(t+\delta)-N_{\demography i}(t)=0 \given X(t)\big]
- &=& 1-\delta \mu_{\demography i}\big(t,X(t)\big) + o(\delta),
-\\
-\label{eq:smcs4}
-\prob\big[ N_{\demography i}(t+\delta)-N_{\demography i}(t)=1 \given X(t)\big]
- &=& \delta \mu_{\demography i}\big(t,X(t)\big) + o(\delta).
-\end{eqnarray}
-Together with the initial conditions $X(0)$, equations \myeqref{eq:smcs1}--\myeqref{eq:smcs4} define a Markov chain.
-Each flow is a simple counting process, meaning a non-decreasing integer-valued process that only has jumps of size one.
-We therefore call the Markov chain a simple Markov counting system (SMCS).
-The infinitesimal mean of every flow is equal to its infinitesimal variance \citep{breto11} and so an SMCS is called equidispersed.
-We note that the special case of Model~1 used by \cite{lee20} (with $\sigmaProc = 0$) is an SMCS.
-To permit more general mean-variance relationships for a Markov counting system, we must permit jumps of size greater than one.
-The utility of overdispersed models, where the infinitesimal variance of the flow exceeds the infinitesimal mean, has become widely recognized \citep{stocks20,he10}.
-
-\subsection{Overdispersed Markov counting system interpretation}
-\label{subsec:odmcs}
-
-Including white noise in the rate function enables the possibility of an overdispersed Markov counting system \citep{breto11,breto09,he10}.
-Since rates should be non-negative, Gaussian noise is not appropriate and gamma noise is a convenient option that has found various applications \citep{romero-severson15, subramanian20}.
-Specifically, we consider a model given by
-\begin{equation}
-\label{eq:odmcs1}
-\mu_{ij}\big(t,X(t)\big) = \bar\mu_{ij}\big(t,X(t)\big) \, d\Gamma_{ij}(t)/dt,
-\end{equation}
-where $\Gamma_{ij}(t)$ is a stochastic process having independent gamma distributed increments, with
-\begin{equation}
-\label{eq:odmcs2}
-\E\big[\Gamma_{ij}(t)\big] = t, \quad \var\big[\Gamma_{ij}(t)\big] = \sigma_{ij}^2 t.
-\end{equation}
-Formally interpreting the meaning of \myeqref{eq:odmcs1} is not trivial, and we do so by constructing a Markov process $X(t)$ as the limit of the Euler scheme descrived in Section~\ref{sec:numerics}, below.
-Therefore, the numerical scheme in Sec.~\ref{sec:numerics} can be taken as a definition of the meaning of \myeqref{eq:odmcs1}.
-The Markov chain defined by the limit of this Euler scheme as the step size decreases is an overdispersed Markov counting system, with the possibility of instantaneous jumps of size greater than one \citep{breto11}.
-
-\section{Numerical solutions to compartment models}
-\label{sec:numerics}
-
-Models may be fitted and their implications assessed via numerical solutions (i.e., simulations) from the model equations.
-All the analyses we consider have this simulation-based property, known as plug-and-play or equation-free or likelihood-free.
-The numerical solutions to the model are arguably of more direct scientific interest than the exact solutions to the postulated equations.
-For ODE models, numerical methods are well studied and a standard numerical solution package such as \code{deSolve} in \code{R} is adequate for many purposes.
-For SMCS and ODMCS models, exact schemes are feasible when the number of events is small, which may be the case for small populations.
-However, for applicability to larger populations, we use instead the following Euler scheme.
-Write $\delta$ for an Euler time step, and $\Delta N_{ij}$ for the numerical approximation to $N_{ij}(t+\delta)-N_{ij}(t)$ given $X(t)$.
-For each $i$ and $j$ in $\seq{1}{c} \cup \{\demography\}$ with $i \neq j$, we draw independent Gamma distributed noise increments with mean $\delta$ and variance $\sigma_{ij}^2 \delta$, denoted using a mean-variance parameterization of the gamma distribution as
-\begin{equation}
-\label{eq:numerics1}
-\Delta\Gamma_{ij} \sim \mathrm{gamma}(\delta, \sigma_{ij}^2 \delta).
-\end{equation}
-In the case of an SMCS model, $\sigma_{ij}=0$ for all $i$ and $j$, so we have $\Delta\Gamma_{ij}=\delta$.
-Then, for $i\neq \demography$ and $j\neq i$, and writing
-\begin{equation}
-\label{eq:numerics2}
-\mu_{ij}=\bar\mu_{ij}\big(t,X(t)\big) \Delta\Gamma_{ij} / \delta,
-\end{equation}
-we calculate transition probabilities
-\begin{eqnarray}
-\label{eq:numerics3}
-p_{ij} &=& \exp\left\{-\sum_{k\in 1:c \, \cup \{\demography\}} \mu_{ik} \, \delta \right\}
-\frac{\mu_{ij}}{\sum_{k\in 1:c\cup \{\demography\}} \mu_{ik}},
-\\
-\label{eq:numerics4}
-p_{ii} &=& 1 - \sum_{j\neq i} p_{ij}.
-\end{eqnarray}
-These probabilities correspond to competing hazards for every individual in compartment $i$ to transition to some compartment $j$, interpreting $j=i$ to mean that the individual remains in $i$.
-Then, $\big(\Delta N_{i1},\dots,\Delta N_{ic},\Delta N_{i\demography}\big)$ has the multinomial distribution where $X_i(t)$ individuals are allocated independently to $\seq{1}{c}\cup\{\demography\}$ with probabilities given by \eqref{eq:numerics3} and \eqref{eq:numerics4}.
-We use the \code{reulermultinom} function in the \code{pomp} package to draw from this multionomial distribution.
-
-Different treatments of demographic flows---such as birth, death, immigration and emigration---are possible.
-For the case $i=\demography$, the treatment used by Model~1 is to set
-\begin{equation}
-\label{eq:numerics6}
-\Delta N_{\demography j} \sim \mathrm{poisson}( \mu_{\demography j} \delta),
-\end{equation}
-an independent Poisson random variable with mean $\mu_{\demography j} \delta$.
-
-Models~2 and 3 used an alternative approach, balancing the total number of flows in and out of the compartment, i.e., $\sum_{i}N_{\demography i}(t) = \sum_{i}N_{i \demography}(t)$, in order to make the model consistent with the known total population.
-In this case, we formally model the death rate as a rate of returning to the susceptible class $S$, and use external transitions from $\demography$ into $S$ to describe only net population increase.
-
-\section{Measurement Models}
-
-Each POMP requires specification of a measurement model, which is a statistical description of how observations on the system are obtained.
-In general, we used the same measurement models that were reported by \cite{lee20} unless specifically noted in the following subsections.
-
-\subsection{Model~1}
-
-In this model, the advantage afforded by vaccination is an increased probability that an infection is asymptomatic.
-Therefore, under the assumptions of this model, all reported cases are assumed to be a fraction of individuals that transition from the exposed to the infected compartment, as noted in Eq.~\myeqref{model1:meas}:
-\begin{equation}
-  \label{model1:meas}
-  Y_{n} \mid \Delta N_{E_{\cdot}I_{\cdot}}(n) = \reportChange \sim \text{NB}\left(\reportRate \reportChange, \obsOverdispersion \right),
-\end{equation}
-where $Y_n$ is the reported cholera cases at time $t_n \in t_1:t_N$ and $\Delta N_{E_{\cdot}I_{\cdot}}(n)$ is the sum total of individuals across each vaccination compartment $\vaccCounter \in 1:\vaccClass$ who moved from compartment $E_\vaccCounter$ to $I_\vaccCounter$ since observation $t_{n-1}$.
-Here, $\text{NB}\left(\reportRate \reportChange, \obsOverdispersion \right)$ denotes a negative binomial distribution with mean $\reportRate \reportChange$ and variance $\reportRate \reportChange \big(1 + \frac{\reportRate \reportChange}{\obsOverdispersion}\big)$.
-
-\subsection{Model~2}\label{sec:mod2Meas}
-
-Model~2 was fit using reported case counts that were transformed using the natural logarithm.
-We fit Model~2 using the subplex algorithm in the \code{subplex} package, using
-a Gaussian measurement model (Eq.~\myeqref{model2:meas}) on the log transformed cases within each unit.
-The final loss function that is maximized is the product of the likelihoods of the individual units, or the sum of the log-likelihood of the individual units.
-The measurement model for individual units is given in Eq.~\myeqref{model2:meas}.
-\begin{equation}
-  \label{model2:meas}
-  \log\big(Y_{u, n} + 1\big) \mid \Delta N_{E_{u\cdot}I_{u\cdot}}(n) = \reportChange_u \sim \text{N}\left(\log \big(\reportRate \reportChange_u + 1\big), \obsOverdispersion^2 \right),
-\end{equation}
-where $\Delta N_{E_{u\cdot}I_{u\cdot}}(n)$ is the sum total of individuals across vaccination compartment $\vaccCounter \in 0:4$ within unit $u$ who moved from compartment $E_{u\vaccCounter}$ to $I_{u\vaccCounter}$ since observation $t_{n-1}$.
-Therefore, because the natural logarithm of observed case counts (plus one, to avoid taking the logarithm of zero) has a normal distribution, $Y_{u, n} + 1$ is assumed to follow a log-normal distribution with log-mean parameter $\log\big(\reportRate \Delta N_{E_{u\cdot}I_{u\cdot}}(n) + 1\big)$ and log-variance $\obsOverdispersion^2$.
-We note that fitting a model with this measurement model is equivalent to fitting using least squares, with $\log(Y_{u, n} + 1)$ as the response variable.
-
-This measurement model differs from that used by \citet{lee20}, who fit the model in two stages: epidemic and endemic phases.
-Although their text and supplement material do not explicitly describe the measurement model used, inspection of the code provided with their submission suggests a change in measurement model between the epidemic and endemic phases.
-% In the file \code{choleraEqsPublic.py}, \citet{lee20} create several functions, where each function represents a set of coupled differential equations that could potentially be used to model cholera incidence data.
-% Each function returns a vector (or \code{numpy} array) that represents the change in each state variable for a single time step, including the variable \code{dC}, which tracks the number of new infections and is used to obtain the reported case counts.
-% Following their comments and code, it appears that the function \code{choleraEqs10WithoutVaccinationNetwork} was used to describe the dynamics of the epidemic phase, and \code{choleraEqs11WithoutVaccinationNetwork} was used for the endemic stage (see Sec~\ref{sec:mod2rep} for more details).
-% Because their models were fit using least squares, the code in these functions suggest that the measurement model for the epidemic phase is
-The measurement model they used for the epidemic phase is
-\begin{equation}
-  \label{model2:measEpi}
-  Y_{u, n} \mid \Delta N_{E_{u\cdot}I_{u\cdot}}(n) = \reportChange_u \sim \text{N}\left(\reportRate \reportChange_u, \obsOverdispersion^2 \right),
-\end{equation}
-% which is similar to our measurement model, the primary difference being that the measurement model is applied to raw case counts rather than log-transformed case counts.
-The measurement model they used for the endemic phase modifies the epidemic model by counting both asymptomatically infected (A) and symptomatically infected (I) individuals in the case counts:
-\begin{equation}
-  \label{model2:measEnd}
-  Y_{u, n} \mid \Delta N_{E_{u\cdot}I_{u\cdot}}(n) = \reportChange_{u1}, \Delta N_{E_{u\cdot}A_{u\cdot}}(n) = \reportChange_{u2} \sim \text{N}\left(\reportRate \big(\reportChange_{u1} + \reportChange_{u2}\big), \obsOverdispersion^2 \right),
-\end{equation}
-where the notation for $\Delta N_{E_{u\cdot}A_{u\cdot}}(n)$ is similar to $\Delta N_{E_{u\cdot}I_{u\cdot}}(n)$, described above.
-
-\subsection{Model~3}
-
-In this model, reported cholera cases are assumed to stem from individuals who develop symptoms and seek healthcare.
-Therefore reported cases are assumed to come from an over-dispersed negative binomial model, given the increase in infected individuals:
-\begin{equation}
-  \label{model3:meas}
-  Y_{u, n} \mid \Delta N_{S_{u \cdot}I_{u}}(t) = \reportChange_u \sim \text{NB}\left(\reportRate \reportChange_{u}, \obsOverdispersion \right),
-\end{equation}
-where $\Delta N_{S_{u \cdot}I_{u \vaccCounter}}(n)$ is the number of individuals who moved from compartment $S_{u \vaccCounter}$ to $I_{u}$ since observation $t_{n-1}$.
-
-This measurement model is a minor change from that used by \cite{lee20}, which allowed for a change in the reporting rate on January 1st, 2018.
-The fitted values of the reporting rate---before and after January 2018---were $0.97$ and $0.097$, respectively.
-An instantaneous change from near perfect to almost non-existent reporting can be problematic, as it forces the model to explain the observed reduction in reported cases as a decrease in the reporting of cases, rather than a decrease in the prevalence of cholera.
-This shift was justified by a ``change of the case definition that occurred on January 1st, 2018";
-this claim was not cited, and we could find no evidence that such a drastic change in the reporting rate would be warranted.
-We therefore do not allow a change in reporting rate when fitting Model~3.
-
-\section{Initial Values}
-
-To perform inference on POMP models, it is necessary to propose an initial probability density for the latent process $f_{X_0}(x_0;\theta)$, including the possibility that the initial values of the latent states are known, or are a non-random function of the unknown parameter vector, $\theta$.
-This density is used to obtain initial values of the latent state when fitting and evaluating the model.
-For each of the models considered in this analysis, the initial conditions are derived by enforcing the model dynamics on reported cholera cases.
-It is also sometimes necessary to fit some initial value parameters in order to help determine initial values for weakly identifiable compartments.
-In the following subsections, we mention initial value parameters that were fit for each model.
-
-\subsection{Model~1}
-
-For this model, the number of individuals in the Recovered and Asymptomatic compartments are set to zero, but the initial proportion of Infected and Exposed individuals is estimated as initial value parameters ($\Iinit$ and $\Einit$, respectively) using the IF2 algorithm, implemented as \texttt{mif2} in the \texttt{pomp} package.
-Finally, the initial proportion of Susceptible individuals $S_{0, 0}$ is calculated as $S_{0, 0} = 1 - \Iinit - \Einit$.
-This model for the initial values of the latent states matches that which was used by \citet{lee20}.
-
-\subsection{Model~2}
-
-Model~2 assumes that the initial values are a deterministic function of the reporting rate and the initial case reports, and so no initial value parameters need to be estimated.
-Initial values for latent state compartments are chosen so as to enforce the model dynamics on the observed number of cases.
-Specifically, the model sets $I_{u0}(0) = y^*_{1u} / \reportRate$ for each unit $u \in 1:10$, where $I_{u0}(t)$ is the number of infected individuals in unit $u$ at time $t$, vaccination scenario $z = 0$, $y^*_{tu}$ is the reported number of cases, and $\reportRate$ is the reporting rate.
-This model for the initial values of the latent states matches that which was used by \citet{lee20}.
-
-\subsection{Model~3}
-
-The latent states of this model are initialized by enforcing the model dynamics on the incidence data from the start of the recorded cases until time $t_0$, requiring that some of the available data be used to determine the initial values of the latent states.
-This is the same approach that was taken by \citet{lee20}, who used the value $t_0 = \text{2014-02-22}$; this choice of $t_0$ results in modeling roughly only $60\%$ of the available data, some of which is later discarded for alternative reasons \citep{lee20sup}.
-
-We do not see any immediate reason that this model could not be extended to cover a larger range of the data, and chose the value $t_0 = \text{2010-11-13}$.
-This choice of $t_0$ corresponds to using approximately $1\%$ of the available data to determine initial values of the latent states.
-In addition to modeling a larger portion of the available data, this choice of $t_0$ corresponds to an important real-world event, as daily reports from each of the departments were not being sent to the health ministry until November 10, 2010; % TODO: Cite
-this choice of $t_0$ therefore makes $\bm{Y}\big(t_1\big)$ the first week of data once daily reports are being sent to the health ministry.
-The few observation times that exist before $t_0$ are used to initialize the model by enforcing model dynamics on these preliminary observations.
-For convenience, we denote these observations as $t_{-3}, t_{-2}$ and $t_{-1}$; as before, we let $y_{u,-k} = Y_{u}\big(t_{-k}\big)$ denote the observed case count for unit $u$ at time point $t_{-k}$, where $k \in 1:3$.
-Equations for the initial values of non-zero latent states are provided in Eqs.~\myeqref{eq:model3:I0}--\myeqref{eq:model3:B0}.
-
-\begin{eqnarray}
-\label{eq:model3:I0}
-I_{u}\big(t_0\big) &=& \frac{y^*_{u, -1}}{7\reportRate\left(\muIR + \left(\muDeath + \choleraDeath\right)/365\right)},
-\\%[-3pt]
-\label{eq:model3:A0}
-A_{u}\big(t_0\big) &=& \frac{I_{u0}\big(t_0\big)\left(1 - \symptomFrac\right)}{\symptomFrac},
-\\
-\label{eq:model3:R0}
-R_{u01}\big(t_0\big) &=& R_{u02}\big(t_0\big) = R_{u03}\big(t_0\big) = \left(\frac{\sum_{k = -3}^0 y^*_{u, k}}{\reportRate \symptomFrac} - \left(I_{u0}\big(t_0\big) + A_{u0}\big(t_0\big)\right)\right) / 3
-\\
-\label{eq:model3:S0}
-S_{u0}\big(t_0\big) &=& \mathrm{Pop}_u - I_{u}\big(t_0\big) - A_{u}\big(t_0\big) - \sum_{k=1}^3 R_{u0k}\big(t_0\big)
-\\
-\label{eq:model3:B0}
-B_{u}\big(t_0\big) &=& \big[1 + \seasAmplitude \tilde{J}^r \big] \mathrm{Den}_u \, \Wshed \big[ I_{u}(t_0)+ \asymptomRelativeShed A_{u_0}(t) \big]/\Wshed.
-\end{eqnarray}
-
-In Eq.~\myeqref{eq:model3:B0}, $\tilde{J} = 0.002376$ is the median adjusted rainfall over the observation period.
-One important consideration to make with this parameter initialization model is when $y_{u, -1}^* = 0$, which occurs for units $u \in \{3, 4\}$, which correspond to the Grand'Anse and Nippes departments, respectively.
-When this is the case, each of the infectious $I_u(t_0)$, asymptomatic $A_u(t_0)$, and bacterial reservoir $W_u(t_0)$ compartments have a value of zero.
-Because Model~3 models cholera transmission primarily by means of the bacterial reservoir, this makes it nearly impossible for an outbreak to occur.
-Therefore for units $u \in \{3, 4\}$, we introduce initial value parameters $I_{0,0}^3$ and $I_{0,0}^4$, and calibrate these parameter values using the data.
-The resulting parameter estimates are used to obtain the remaining non-zero initial values of the latent states using Eqs.~\myeqref{eq:model3:A0}--\myeqref{eq:model3:B0}.
-
-% Argue why we didn't need to do this for Model 2.
-
-% TODO: Check this section.
-
-\section{Calibrating Model~3 to observed cases}\label{sec:mod3Cal}
-
-<<CalibrateOrig3>>=
-# Create vectors for the unit and shared parameters
-unit_specific_names <- c("betaB", "foi_add")
-shared_param_names <- c(
-  "mu_B", "XthetaA", "thetaI", "lambdaR", "r", "std_W",
-  "epsilon", "k"
-)
-est_param_names <- c(
-  unit_specific_names, shared_param_names
-)
-
-# Add unit numbers to each parameter
-est_param_names_expanded <- paste0(rep(est_param_names, each = 10), 1:10)
-
-# Simple function that is used to set the rw_sd for the first search
-# This function is convenient so that we don't have to type out the rw_sd
-# for each unit parameter.
-set_rw_1 <- function(x) {
-  if (gsub("[[:digit:]]+$", "", x) %in% shared_param_names) {
-    return(0.01)
-  # } else if (x %in% paste0(rep(c("aHur", 'hHur'), each = 2), c(3, 9))) {
-  #   return(expression(ifelse(time >= 2016.754 & time <= 2017, 0.015, 0)))
-  # } else if (x %in% c("Iinit3", "Iinit4")) {
-  #   return(expression(ivp(0.25)))
-  } else if (grepl('^foi_add[[:digit:]]+$', x)) {
-    return(0.015)
-  } else if (grepl('^betaB[[:digit:]]+$', x)) {
-    return(0.01)
-  } else {
-    return(0)
-  }
-}
-
-set_rw_2 <- function(x) {
-  if (gsub("[[:digit:]]+$", "", x) %in% shared_param_names) {
-    return(0.003)
-  # } else if (x %in% paste0(rep(c("aHur", 'hHur'), each = 2), c(3, 9))) {
-  #   return(expression(ifelse(time >= 2016.754 & time <= 2017, 0.005, 0)))
-  # } else if (x %in% c("Iinit3", "Iinit4")) {
-  #   return(expression(ivp(0.125)))
-  } else if (grepl('^foi_add[[:digit:]]+$', x)) {
-    return(0.005)
-  } else if (grepl('^betaB[[:digit:]]+$', x)) {
-    return(0.003)
-  } else {
-    return(0)
-  }
-}
-
-# First Search RW
-reg_rw_1.sd <- lapply(est_param_names_expanded, set_rw_1)
-names(reg_rw_1.sd) <- est_param_names_expanded
-chol_rw_1.sd <- do.call(rw_sd, reg_rw_1.sd)
-
-# Second Search RW
-reg_rw_2.sd <- lapply(est_param_names_expanded, set_rw_2)
-names(reg_rw_2.sd) <- est_param_names_expanded
-chol_rw_2.sd <- do.call(rw_sd, reg_rw_2.sd)
-
-SEARCH1 = list(
-  NBPF = 100,
-  NP = 2500,
-  SPAT_REGRESSION = 0.05,
-  NREPS = 144,
-  NP_EVAL = 2500,
-  NREPS_EVAL = 18,
-  RW_SD = chol_rw_1.sd,
-  COOLING = 0.5,
-  KEEP_TRACES = FALSE,
-  KEEP_LIKE_MAT = FALSE
-)
-
-SEARCH2 = list(
-  TOP_N = 12,
-  NBPF = 100,
-  NP = 2500,
-  SPAT_REGRESSION = 0.05,
-  NREPS = 6,
-  NP_EVAL = 2500,
-  NREPS_EVAL = 18,
-  RW_SD = chol_rw_2.sd,
-  COOLING = 0.5,
-  KEEP_TRACES = FALSE,
-  KEEP_LIK_MAT = TRUE
-)
-
-haiti3_sub_fit <- bake(
-  file = "../model3/si/subset_noparams.rds", {
-    fit_haiti3(
-      nsearches = 2L,
-      search1 = SEARCH1,
-      search2 = SEARCH2,
-      search_rho = FALSE,
-      search_gamma = FALSE,
-      search_Iinit = FALSE,
-      search_hur = FALSE,
-      ncores = cores
-    )
-  }
-)
-
-# Find model index with best log-likelihood
-best_m <- which.max(haiti3_sub_fit$search2$logLiks$logLik)
-
-# Find parameters using best_m index
-sub_params <- haiti3_sub_fit$search2$params[best_m, ]
-
-# Get unit log-likelihoods for the best model.
-all_liks <- as.data.frame(haiti3_sub_fit$search2$likMat)
-# all_liks$which <- rep(1:(SEARCH3$TOP_N * SEARCH3$NREPS), each = SEARCH3$NREPS_EVAL)
-all_liks$which <- rep(1:(12 * 6), each = 18)
-best_mod <- filter(all_liks, which == best_m)
-
-# Pivot longer to get estimate of log-likelihood for each unit.
-unitLikesM3 <- best_mod %>%
-  pivot_longer(
-    cols = -which,
-    names_to = "unit",
-    names_prefix = "V",
-    values_to = "logLik"
-  ) %>%
-  select(-which) %>%
-  group_by(unit) %>%
-  summarize(
-    ll_est = logmeanexp(logLik),
-    ll_se = logmeanexp(logLik, se = TRUE)[2]
-  ) %>%
-  ungroup() %>%
-  mutate(unit = as.numeric(unit)) %>%
-  arrange(unit)
-@
-
-<<M3_bench, echo=FALSE, message=FALSE, include=FALSE, cache=TRUE>>=
-h3 <- haiti3_spatPomp()
-deps <- h3@unit_names
-
-mod3_cases <- as.data.frame(t(h3@data))
-colnames(mod3_cases) <- deps
-m3_inits <- c(7362, 1914, 22, 8, 2818, 82, 2499, 6953, 44, 7) # TODO: Also fit by hand, but could be changed to initializing to the observation before.
-names(m3_inits) <- deps
-
-bench_cond_lls <- list()
-for (dep in deps) {
-
-  # Get department cases
-  cases <- as.numeric(mod3_cases[, dep])
-  mod <- ar1_NegBinom(data = cases, init = m3_inits[dep])
-
-  bench_cond_lls[[dep]] <- haitipkg:::.ar1_NegBinom_cond_ll(
-    theta = mod$theta, init = m3_inits[dep], data = cases, transform = FALSE
-  )
-}
-
-bench_cond_lls <- as.data.frame(bench_cond_lls)
-
-bench_lls <- apply(bench_cond_lls, 2, sum)
-
-unitLikes_bench <- data.frame(
-  unit = 1:10,
-  # dep = names(bench_lls),
-  ll = bench_lls
-) %>%
-  as_tibble()
-
-bench_cond_lls$date <- lubridate::round_date(lubridate::date_decimal(h3@times), unit = 'day')
-@
-
-<<combine_bench_M3, echo=FALSE>>=
-
-# Combine tables with model 3 and benchmark likelihoods.
-bench_M3_joined <- dplyr::left_join(
-  x = unitLikesM3,
-  y = unitLikes_bench,
-  by = "unit"
-) %>%
-  select(-ll_se) %>%
-  mutate(unit = as.character(unit)) %>%
-  pivot_longer(
-    cols = -unit,
-    names_to = "type",
-    values_to = "ll"
-  ) %>%
-  mutate(unit = paste0(unit, ": ", deps[as.numeric(unit)]))
-@
-
-<<M3bpf, echo=FALSE, message=FALSE, include=FALSE>>=
-
-# Change model parameters to the best set of parameters from previous
-# search
-coef(h3)[names(sub_params)] <- sub_params
-
-registerDoRNG(40765101)  # for modular, reproducible results
-
-# Perform block partical filter with fitted model, to get conditional
-# log-likelihoods.
-bpf_sub3_out <- bake(
-  file = "../model3/si/bpf_sub_cond_LL.rds", {
-
-    results <- list()
-
-    bpf_out <- h3 %>% bpfilter(
-      Np = 5000,
-      block_size = 1,
-      save_states = TRUE
-    )
-
-    results$block.cond.loglik <- bpf_out@block.cond.loglik
-    results$times <- bpf_out@times  # Saved for convenience
-    results
-  },
-  seed = 432235
-)
-@
-
-In this section, we provide more detail on the process that was used to estimate the coefficients used in Model~3.
-In particular, we discuss why we decided to include additional model parameters---those that are associated with the behavior of the system during Hurricane Matthew---that were not considered by \cite{lee20}.
-To calibrate this model, we used the iterated block particle filter (IBPF) method of \cite{ionides22}.
-Due to the novelty of this algorithm, there does not exist published examples the IBPF algorithm used for data analysis outside of the papers in which the algorithm was introduced \citep{ning21ibpf,ionides22}, which at least partially motivates the inclusion additional details in this section.
-
-\cite{lee20} were only able to estimate model parameters to a simplified version of Model~3 on a subset of the available data, as no method existed at the time of their publication to fit a fully coupled meta-population model to disease incidence data.
-Building on their results, we fit the fully coupled version of Model~3 to nearly all available data, reserving only a few observations to use to calibrate the initial conditions of the model (see the previous section for more details).
-In order to properly maximize the model likelihood, we found that it was necessary to use multiple searches for the MLE, periodically pruning away less successful searches.
-The first collection of searches was performed by obtaining initial values for the parameters by uniformly sampling values from a predefined hypercube.
-A subsequent refinement search used starting values corresponding to the largest model likelihoods obtained in the previous round.
-This is similar to the approach used by \cite{ionides22}.
-
-We use the technique described above to fit the fully coupled version of Model~3 proposed by \cite{lee20}.
-The maximum likelihood we obtained with the model was $\Sexpr{round(max(haiti3_sub_fit$search2$logLiks$logLik))}$, which is higher than the benchmark model ($\Sexpr{round(sum(unitLikes_bench$ll))}$).
-While beating a simple associative benchmark is promising, this does not immediately rule out the possibility that the model is a poor description of the system.
-Additional investigation of parameters estimates and their corresponding implications on model based conclusions should always be conducted.
-For meta-population models, it is worth considering how well the model fits the data to each spatial unit.
-This can be done by looking at conditional log-likelihoods, which is part of the output of the \texttt{bpfilter} algorithm in the \texttt{spatPomp} package.
-The likelihoods for each department, compared to the corresponding benchmark model, are displayed in Fig.~\ref{fig:h3UnitLikes}.
-The figure demonstrates that while the log-likelihood of the fitted model outperforms the auto-regressive negative binomial benchmark model, Model~3 has lower likelihoods than the benchmark for some departments.
-
-\begin{figure}[ht]
-<<h3UnitLikes, fig.height=3, fig.width=5.5, echo=FALSE>>=
-ggplot(bench_M3_joined, aes(x = ll, y = reorder(unit, ll))) +
-  geom_point(aes(col = type)) +
-  ylab("Unit") +
-  xlab("Log Likelihood") +
-  scale_color_manual(
-    labels = c("ll" = "Benchmark", "ll_est" = "Model 3"),
-    values = c("ll" = "#1b9e77", "ll_est" = "#d95f02")
-  ) +
-  theme(legend.title = element_blank(),
-        axis.title.y = element_blank())
-@
-\caption{\label{fig:h3UnitLikes}Log-likelihoods of Model~3 for each department compared to the corresponding benchmark model prior to the inclusion of parameters for initial values and Hurricane Matthew.}
-\end{figure}
-
-In addition to considering the conditional log-likelihoods of each unit, one can consider conditional log-likelihoods of each observation.
-When compared to a benchmark, this level of detail can provide useful information about which observations are well described by the model and which are poorly described.
-In Fig.~\ref{fig:condLL}, we plot the conditional log-likelihoods of each observation.
-This figure reveals that the fitted model has difficult time explaining certain features of the data.
-For example, many departments (in particular Sud) have observations with lower conditional log-likelihoods near October 2016 than at other time points.
-Further investigation of the model output reveals that the model is struggling to explain the sudden surge in cholera cases that occurred at this time, which coincides with the time that Hurricane Matthew struck Haiti.
-While the model does include a mechanism to account for increased cholera transmission due to large rainfall events, the mechanism does not appear to be sufficient to capture the damaging effects of the hurricane, which had the greatest impact in the the Sud and Grand'Anse departments \citep{ferreirai16}.
-This result led us to include parameters $\Whur{u}$ and $\hHur{u}$ in Eq.~23 of the main text, which allows for an increase in the transmission rate between environmental cholera and humans for Sud and Grand'Anse during and after the hurricane.
-The effect of the hurricane on cholera transmission is assumed to have an exponential decay, where the magnitude is determined by $\Whur{u}$ and the duration of the effect determined by $\hHur{u}$.
-
-\begin{figure}[ht]
-<<CondLLFig, echo=FALSE, message=FALSE, fig.height=3>>=
-
-# Changing labels to human-readable labels.
-deps[grepl("_", deps)] <- gsub("_", "-", deps[grepl("_", deps)])
-deps[deps == "Grande-Anse"] <- "Grand\'Anse"
-names(deps) <- as.character(1:10)
-myLabeller <- as_labeller(deps)
-
-# Plot conditional log-likelihoods for each unit
-t(bpf_sub3_out$block.cond.loglik) %>%
-  as.data.frame() %>%
-  mutate(t = bpf_sub3_out$times) %>%
-  pivot_longer(
-    cols = -t,
-    values_to = 'likelihood',
-    names_prefix = "V",
-    names_to = 'Unit'
-  ) %>%
-  mutate(date = as.Date(lubridate::date_decimal(t))) %>%
-  ggplot(aes(x = date, y = likelihood)) +
-  geom_line() +
-  geom_point() +
-  facet_wrap(~Unit, labeller = myLabeller, scales = 'free_y',
-             nrow = 2) +
-  theme_bw() +
-  theme(axis.title.x = element_blank(),
-        axis.text.x = element_text(angle = -45, hjust = 0)) +
-  ylab("Conditional log-likelihoods")
-@
-\caption{\label{fig:condLL}Conditional log-likelihoods of Model~3 prior to the inclusion of parameters for initial values and Hurricane Matthew.}
-\end{figure}
-
-Other points of concern include the observations at or near the start of the recorded cases, for example, in Nord-Ouest and Sud-Est.
-These low conditional log-likelihoods are a result of having zero or near zero confirmed cases in the first week of observed data, followed by a rapid growth of cases.
-The model cannot explain the rapid growth of cholera in the absence of starting cases because the initialization scheme forces departments with no recorded cases in the first week to have hidden states $I_{u0}(0) = A_{u0}(0) = B_{u}(0) = 0$.
-In reality, however, it is possible to have some infected or asymptomatic individuals in a department when no cases are recorded due to under-reporting.
-This may be particularly true at the onset of an infectious disease outbreak, as lower public awareness and inadequate testing infrastructure may result in many cases going undetected.
-We therefore include and estimate $I_{0u}$ that determines the initial number of Infected individuals at the start of the pandemic for the departments Grande'Anse, Nord-Ouest, and Sud-Est;
-each of these departments have both low starting case counts followed by rapid growth, and low initial values of the conditinoal log-likelihood.
-
-<<loadFinalModel3, echo=FALSE, include=FALSE, message=FALSE>>=
-
-# This result was used in the manuscript, so it doesn't need to be recalculated.
-h3_final_results <- readRDS("../model3/run_level_3/haiti3_fit.rds")
-
-# Save best_m as index for the model parameters with the best likelihood.
-# Then, save the model parameter values and the likelihood evaluations
-best_m <- which.max(h3_final_results$search2$logLiks$logLik)
-best_params <- h3_final_results$search2$params[best_m, ]
-all_liks <- as.data.frame(h3_final_results$search2$likMat)
-all_liks$which <- rep(1:72, each = 18)
-best_mod <- dplyr::filter(all_liks, which == best_m)
-
-# Get repeated estimates of unit log-likelihoods
-unitLikesM3 <- best_mod %>%
-  pivot_longer(
-    cols = -which,
-    names_to = "unit",
-    names_prefix = "V",
-    values_to = "logLik"
-  ) %>%
-  select(-which) %>%
-  group_by(unit) %>%
-  summarize(
-    ll_est = pomp::logmeanexp(logLik),
-    ll_se = pomp::logmeanexp(logLik, se = TRUE)[2]
-  ) %>%
-  ungroup() %>%
-  mutate(unit = as.numeric(unit)) %>%
-  arrange(unit)
-
-# Combine model likelihood table to benchmarks
-bench_M3_joined <- dplyr::left_join(
-  x = unitLikesM3,
-  y = unitLikes_bench,
-  by = "unit"
-) %>%
-  select(-ll_se) %>%
-  mutate(unit = as.character(unit)) %>%
-  pivot_longer(
-    cols = -unit,
-    names_to = "type",
-    values_to = "ll"
-  ) %>%
-  mutate(unit = paste0(unit, ": ", deps[as.numeric(unit)]))
-@
-
-We refit Model~3 after introducing these seven parameters---three initial value parameters and four parameters related to Hurricane Matthew---using the same successive search scheme described above.
-The resulting model has a log-likelihood value of $\Sexpr{round(max(h3_final_results$search2$logLiks$logLik), 1)}$.
-We also provide figures for department specific log-likelihoods and conditional log-likelihoods in Figs.~\ref{fig:finalUnitLL} and \ref{fig:finalCondLL}.
-
-\begin{figure}[ht]
-<<finalUnitLLFig, fig.height=3, fig.width=5.5, echo=FALSE>>=
-ggplot(bench_M3_joined, aes(x = ll, y = reorder(unit, ll))) +
-  geom_point(aes(col = type)) +
-  ylab("Unit") +
-  xlab("Log Likelihood") +
-  scale_color_manual(
-    labels = c("ll" = "Benchmark", "ll_est" = "Model 3"),
-    values = c("ll" = "#1b9e77", "ll_est" = "#d95f02")
-  ) +
-  theme(legend.title = element_blank(),
-        axis.title.y = element_blank())
-@
-\caption{\label{fig:finalUnitLL}Log-likelihoods of Model~3 for each department compared to the corresponding benchmark model after adding and estimating parameters related to initial values and Hurricane Matthew.}
-\end{figure}
-
-<<finalModel3BPF, echo=FALSE, message=FALSE, include=FALSE>>=
-
-# Set the model parameters to the best results from previous search
-coef(h3)[names(best_params)] <- best_params
-
-# block particle filter to get conditional log-likelihoods.
-bpf_final3_out <- bake(
-  file = "../model3/si/h3_final_bpf_cond_ll.rds", {
-
-    results <- list()
-
-    bpf_out <- h3 %>% bpfilter(
-      Np = 5000,
-      block_size = 1,
-      save_states = TRUE
-    )
-
-    results$Sus_df <- foreach(i=1:length(h3@times), .combine = bind_rows) %do% {
-      get_h3_S_quants(bpf_out, i)
-    }
-
-    results$block.cond.loglik <- bpf_out@block.cond.loglik
-    results$times <- bpf_out@times  # saved for convenience
-    results
-  },
-  seed = 786321
-)
-@
-
-\begin{figure}[ht]
-<<finalCondLLFig, echo=FALSE, message=FALSE, fig.height=3>>=
-t(bpf_final3_out$block.cond.loglik) %>%
-  as.data.frame() %>%
-  mutate(t = bpf_final3_out$times) %>%
-  pivot_longer(
-    cols = -t,
-    values_to = 'likelihood',
-    names_prefix = "V",
-    names_to = 'Unit'
-  ) %>%
-  mutate(date = as.Date(lubridate::date_decimal(t))) %>%
-  ggplot(aes(x = date, y = likelihood)) +
-  geom_line() +
-  geom_point() +
-  facet_wrap(~Unit, labeller = myLabeller, scales = 'free_y',
-             nrow = 2) +
-  theme_bw() +
-  theme(axis.title.x = element_blank(),
-        axis.text.x = element_text(angle = -45, hjust = 0)) +
-  ylab("Conditional log-likelihoods")
-@
-\caption{\label{fig:finalCondLL}Conditional log-likelihoods of Model~3 after adding and estimating the parameters for initial values and Hurricane Matthew.}
-\end{figure}
-
-While the addition of the Hurricane parameters seems to have resulted in a dramatic increase of conditional likelihoods around and after October 2016, the inclusion of the initial value parameters does not appear to have fixed the low conditional log-likelihood values in Sud-Est and Nord-Ouest.
-This could be due to many factors, including the possibility that the coupling between units forces low initial value parameters in a given department in order for the model to fit the data in the remaining departments.
-Whatever the case may be, the inclusion of the seven parameters resulted in a $\Sexpr{round(max(h3_final_results$search2$logLiks$logLik) - max(haiti3_sub_fit$search2$logLiks$logLik), 1)}$ increase in log-likelihood, which is a highly statistically significant difference.
-One could similarly test whether the inclusion of the initial value parameters resulted in a significantly better fit to the observed data, but we do not do this here due to the large computational effort that would be required.
-
-\subsection{Examining the Hidden States of the Calibrated Model}
-
-For mechanistic models, beating a suitable statistical benchmark does not alone guarantee that the model provides an accurate description of a dynamic process.
-Indeed, a good statistical fit does not require the model to assert a causal explanation.
-For example, reconstructed latent variables should make sense in the context of alternative measurements of these variables \citep{grad12}.
-We demonstrate this principle by examining the latent state of the calibrated model.
-In particular, we examine the compartment of susceptible individuals under various scenarios, as this compartment drives transmission dynamics in SIR models for infectious disease outbreaks.
-
-Recall that the filtering distribution for the calibrated version of Model~3 at time $t_k$ is defined as the distribution of the latent state at time $t_k$ given the data from times $t_{1}:t_{k}$, i.e.
-$f^{(3)}_{\bm{X}_k|\bm{Y}_{1:k}}(\bm{x}_{k} | \bm{y}^*_{1:k} ; \hat\theta)$.
-In general, one may expect simulations from the filtering distribution of a model with a good statistical fit to result in hidden states that are highly consistent with the observed data because the filtering distribution is conditioned on the observed data.
-Fig.~\ref{fig:h3Sus} shows the percentage of individuals that are in the susceptible compartment from various simulations of the model:
-simulations from Model~3 under initial conditions are displayed in red; simulations from the filtering distribution of model are displayed in blue.
-This figure shows that simulations from initial conditions tends to result in a much more rapid depletion of susceptible individuals at the start of the epidemic than simulations from the filtering distribution, suggesting the calibrated model has a propensity to predict larger outbreaks than what is typically seen in the data.
-This also appears true in later stages of the time period of observed data, where the data suggests that there is a replenishment of the susceptible compartment, but the Model simulations typically retain a smaller proportion of susceptible individuals.
-This result demonstrates that the calibrated model favors a more rapid growth in cholera cases than what is typically seen in the observed data, which explains why the model fails to predict the absence of cases between February, 2019 and October, 2022.
-
-<<h3SusFigSetup, echo=FALSE, message=FALSE, include=FALSE>>=
-
-# This bake perfoms the following:
-#  (1) simulates the model from initial conditions
-#  (2) get the values of the hidden states at the end of each simulation, and
-#      saves these for later use.
-#  (3) Obtain quantiles for the Susceptible pool from the simulations.
-init_conditions_results <- bake(
-  file = "../model3/si/InitSimsQuants.rds", {
-
-    results <- list()  # To return final output of bake
-
-    # Simulate from the fitted model
-    h3Spat_sims <- simulate(
-      h3, nsim = 2000, format = 'data.frame'
-    )
-
-    # Get states at the end of the simulations
-    h3_spat_end <- h3Spat_sims %>%
-      filter(time == max(time))
-
-    # Save hidden state values at the end of the simulation. Saved as a matrix
-    # in the same format as the output from a block-particle filter, so that
-    # it can be used as the input to project_from_filter.
-    results[["end_conditions"]] <- h3_spat_end %>%
-      select(-time, -cases) %>%
-      mutate(
-        unit = case_when(
-          unitname == "Artibonite" ~ 1,
-          unitname == "Centre" ~ 2,
-          unitname == "Grande_Anse" ~ 3,
-          unitname == "Nippes" ~ 4,
-          unitname == "Nord" ~ 5,
-          unitname == "Nord_Est" ~ 6,
-          unitname == "Nord_Ouest" ~ 7,
-          unitname == "Ouest" ~ 8,
-          unitname == "Sud" ~ 9,
-          unitname == "Sud_Est" ~ 10
-        )
-      ) %>%
-      select(-unitname) %>%
-      pivot_wider(
-        data = .,
-        names_from = unit,
-        values_from = c(S, I, A, R_one, R_two, R_three, VSd, VR1d, VR2d,
-                        VR3d, VSdd, VR1dd, VR2dd, VR3dd, VSd_alt, VR1d_alt,
-                        VR2d_alt, VR3d_alt, VSdd_alt, VR1dd_alt, VR2dd_alt,
-                        VR3dd_alt, C, B, Doses, totInc),
-        names_sep = ""
-      ) %>%
-      select(-.id) %>%
-      as.matrix() %>%
-      t()
-
-    # Get the quantiles for the Susceptible pool from the simulations
-    results[["SQuants"]] <- h3Spat_sims %>%
-      rename(dep = unitname) %>%
-      group_by(dep, time) %>%
-      summarise(
-        Q025 = quantile(S + VSd + VSdd + VSd_alt + VSdd_alt, probs = 0.025),
-        Q50  = quantile(S + VSd + VSdd + VSd_alt + VSdd_alt, probs = 0.5),
-        Q975 = quantile(S + VSd + VSdd + VSd_alt + VSdd_alt, probs = 0.975)
-      ) %>%
-      ungroup() %>%
-      rowwise() %>%
-      mutate(
-        unit_num = which(h3@unit_names == dep),
-        date = lubridate::date_decimal(time),
-        Q025 = Q025 / h3@params[paste0("H", unit_num)],
-        Q50  = Q50  / h3@params[paste0("H", unit_num)],
-        Q975 = Q975 / h3@params[paste0("H", unit_num)]
-      ) %>%
-      filter(
-        date >= as.Date(
-          lubridate::round_date(lubridate::date_decimal(h3@t0), unit = 'day')
-        )
-      ) %>%
-      mutate(date = as.Date(lubridate::round_date(date, unit = 'day')))
-
-    results
-  },
-  seed = 834687
-)
-
-# Get the output of the "bake" call for plotting.
-h3Spat_quants <- init_conditions_results[["SQuants"]]
-
-# This bake call uses the ending hidden states from above to forecast cholera
-# cases in the future under the V0 vaccination scenario
-mod3_V0_initcond_res <- bake(
-  file = '../model3/si/vacc_sim_from_init.rds', {
-    h3_V0_res <- list()
-
-    # Get "noVacc" scenario parameters
-    h3_S0_par <- get_model3_vacc_scenario_params(scenario = "noVacc")
-    h3Spat <- h3
-    p3 <- best_params
-    coef(h3Spat) <- p3
-    coef(h3Spat)[names(h3_S0_par)] <- h3_S0_par
-
-    # store the end states that will be used as initial values
-    h3_bpf_end_states <- init_conditions_results$end_conditions
-
-    # Using parameters and end state values, obtain projections
-    h3_V0_sims <- project_from_filter(
-      h3Spat, end_states = h3_bpf_end_states, covarGen = project_rain,
-      nsims = 1000, seed = 184697
-    )
-
-    # Store nationally aggregated results
-    h3_V0_res$mod3_V0_sims <- agg_mod3_sims(h3_V0_sims)
-    h3_V0_res$mod3_V0_probs <- get_elimProbs(h3_V0_sims, model = 3)
-
-    # Get quantiles of observed cases for each department and time
-    h3_V0_res$UnitQuants <- h3_V0_sims %>%
-      mutate(
-        date = as.Date(  # Get date object from decimal time
-          lubridate::round_date(lubridate::date_decimal(time), unit = 'day')
-        )
-      ) %>%
-      group_by(unitname, date) %>%
-      summarize(
-        Q025 = quantile(cases, probs = 0.025),
-        Q50 = quantile(cases, probs = 0.5),
-        Q975 = quantile(cases, probs = 0.975)
-      ) %>%
-      ungroup() %>%
-      rename(dep = unitname)
-
-    # Get quantiles of susceptible individuals for each department and time
-    h3_V0_res$SQuants <- h3_V0_sims %>%
-      mutate(
-        date = as.Date(
-          lubridate::round_date(lubridate::date_decimal(time), unit = 'day')
-        )
-      ) %>%
-      group_by(unitname, date) %>%
-      mutate(
-        # Get the population for each unit
-        pop = h3@params[paste0("H", which(first(unitname) == unit_names(h3)))]
-      ) %>%
-      summarize(  # Calculate the quantiles
-        Q025 = quantile(
-          (S + VSd + VSdd + VSdd_alt + VSd_alt) / pop,
-          probs = 0.025
-        ),
-        Q50 = quantile(
-          (S + VSd + VSdd + VSdd_alt + VSd_alt) / pop,
-          probs = 0.5
-        ),
-        Q975 = quantile(
-          (S + VSd + VSdd + VSdd_alt + VSd_alt) / pop,
-          probs = 0.975
-        )
-      ) %>%
-      ungroup() %>%
-      rename(dep = unitname)
-
-    h3_V0_res
-  }
-)
-
-# This result was already used in the manuscript; it doesn't need to
-# be recalculated.
-mod3_V0_res <- readRDS("../model3/run_level_3/mod3_V0_res.rds")
-@
-
-\begin{figure}[ht]
-<<h3SusFig, echo=FALSE, message=FALSE, fig.height=3>>=
-ggplot() +
-  # Sims from initial condition
-  geom_line(data = h3Spat_quants, aes(x = date, y = Q50), col = 'red') +
-  geom_ribbon(
-    data = h3Spat_quants,
-    aes(x = date, ymin = Q025, ymax = Q975),
-    fill = 'red', alpha = 0.5
-  ) +
-  # projections from initial condition, using filtered sims function
-  geom_line(data = mod3_V0_initcond_res$SQuants, aes(x = date, y = Q50), col = 'red') +
-  geom_ribbon(
-    data = mod3_V0_initcond_res$SQuants,
-    aes(x = date, ymin = Q025, ymax = Q975),
-    fill = 'red', alpha = 0.5
-  ) +
-  # Sims from filtered distribution
-  geom_line(
-    data = bpf_final3_out$Sus_df,
-    aes(x = as.Date(lubridate::date_decimal(time)), y = Q50),
-    col = 'blue'
-  ) +
-  geom_ribbon(
-    data = bpf_final3_out$Sus_df,
-    aes(x = as.Date(lubridate::date_decimal(time)), ymin = Q025, ymax = Q975),
-    fill = 'blue', alpha = 0.5
-  ) +
-  # Projections from filtered distribution
-  geom_line(data = mod3_V0_res$SQuants, aes(x = date, y = Q50), col = 'blue') +
-  geom_ribbon(
-    data = mod3_V0_res$SQuants,
-    aes(x = date, ymin = Q025, ymax = Q975),
-    fill = 'blue', alpha = 0.5
-  ) +
-  geom_vline(xintercept = max(h3Spat_quants$date), linetype = 'dashed') +
-  facet_wrap(~dep, scales = 'free_y', nrow = 2) +
-  theme_bw() +
-  theme(axis.title.x = element_blank(),
-        axis.text.x = element_text(angle = -45, hjust = 0)) +
-  ylab("Total Susceptible (%)")
-@
-\caption{\label{fig:h3Sus}Percentage of individuals that are in the susceptible compartment.
-Simulations from Model~3 under initial conditions are displayed in red; simulations from the filtering distribution of model are displayed in blue.
-The dashed line represents the end of the observed data.}
-\end{figure}
-
-\clearpage
+%%%%%%%%%% START
 
@@
 
-<<PlotEpiDist, echo=FALSE, fig.cap="Compare to Figure S8 of Lee et al. (2020) supplement.", fig.width=7, fig.height=7, fig.align='center'>>=
+<<PlotEpiDist, echo=FALSE, fig.cap="Compare to Figure S8 of Lee et al. (2020) supplement.", fig.width=7, fig.height=7, fig.align='center',message=FALSE,warning=FALSE>>=
 GGally::ggpairs(
@@
 Compare to S9 of Lee et al.
-(2020) supplement.", cache=TRUE, echo=FALSE>>=
+(2020) supplement.", cache=TRUE, echo=FALSE,warning=FALSE>>=
 GGally::ggpairs(
@@
 Compare to Figure S7 of \\cite{lee20sup}.
-The upper bound for the likelihood of this model is ", round(mod1_lee_ll), "."), cache=TRUE, fig.width=6.5, fig.height=3.75>>=
+The upper bound for the likelihood of this model is ", round(mod1_lee_ll), "."), cache=TRUE, fig.width=6.5, fig.height=3.75,warning=FALSE>>=
 
@@
   geom_point(data = true_agg_cases, aes(x = date, y = ReportedAll)) +
-  geom_line(data = quants, aes(x = date, y = q50), col = '#0b8a2b', size = 1) +
+  geom_line(data = quants, aes(x = date, y = q50), col = '#0b8a2b', linewidth = 1) +
   geom_ribbon(
@@
 
-\section{Forecasting with parameter uncertainty}
-
-Let $f_{Y_{1:N}}(y_{1:N} | \theta)$ denote the pdf of the model under consideration, were $\theta$ is a parameter vector that indexes the model.
-Furthermore, denote the observed data as $y_{1:N}^*$.
-Because the uncertainty in just a single parameter can lead to drastically different forecasts \citep{saltelli20},
-parameter uncertainty should be considered when obtaining model forecasts when the goal is to influencing policy.
-In a Bayesian modelling paradigm, the most natural way to account for parameter uncertainty in model forecasts is to suppose that $\theta$ comes from a distribution $f_{\Theta}$, and then to obtain $J$ forecasts from the model where each forecast is obtained using parameters drawn from the posterior distribution $\Theta_{1:J} \mid Y_{1:N} = y_{1:N}^* \sim f_{\Theta}\big(\theta | Y_{1:N} = y_{1:N}^*\big)$.
-
-When frequentist methods are used, however, there does not exist a posterior distribution from which one could sample.
-A common approach could be to obtain a weighted average of the simulations from various models \citep{hoeting99}, but this can be problematic when forecasts from each model are very different from each other \citep{grueber11}.
-A similar approach that has been taken \citep{king15} is to obtain model forecasts using multiple sets of parameter values and then sample from the resulting forecasts using weights proportional to the corresponding likelihoods of the parameter values.
-This approach could be considered as empirical Bayes, as it is equivalent to using a discrete uniform prior where the set of values in the prior distribution is determined by a stochastic routine applied to the observed data, as discussed below.
-
-For each $i \in 1:k$, let $\Theta_i$ be a random vector of model parameters.
-Then, letting $\Theta$ denote the true model parameters, we endow the set $\{\Theta_1, \Theta_2, \ldots, \Theta_k\}$ with a discrete uniform distribution, such that $P\big(\Theta = \Theta_i\big) = \frac{1}{K}$ for all values $i \in \seq{1}{k}$.
-Using this as a prior distribution, the posterior distribution of $\Theta | Y_{1:N} = y_{1:N}^*$ can be expressed as:
-$P\big(\Theta = \Theta_k | Y_{1:N} = y_{1:N}^*\big) = \frac{f_{Y_{1:N}}(y_{1:N}^*| \Theta_k)}{\sum_{l = 1}^K f_{Y_{1:N}}(y_{1:N}^*| \theta_l)}$.
-In a standard empirical Bayes analysis, the values $\Theta_1, \ldots, \Theta_k$ of the prior distribution would be chosen using the observed data, resulting in a posterior distribution that weighs the prior parameter vectors proportional to their corresponding likelihoods.
-Instead, we choose $\Theta_i$ to be the output of a stochastic routine applied to the observed data by setting $\Theta_i$ to be the output of an iterated filtering algorithm.
-In practice, because the likelihood maximization routines of iterated filtering methods are stochastic, it is common to run the iterated filtering method multiple times for each model in order to obtain a maximum likelihood estimate for model parameters.
-This results in a natural set of parameters near the MLE that could be used as the discrete prior distribution.
-
-\section{Translating to \citet{lee20} notation}
-
-Since the models of \citet{lee20} were developed independently, the choice of notation varies inconsistently between models.
-For our reanalysis, we rename parameters to provide a unified notation facilitating comparison between models.
-Table~\ref{tab:translate} maps this notation back to the original notations, for reference.
-
-\begin{table}
-  \begin{center}
-  \begin{tabular}{|c|c|c|c|c|}\hline
-    \multirow{2}{*}{Parameter} & Our & \multicolumn{3}{c|}{Lee et al. (2020a)} \\\cline{3-5}
-     & Notation & 1 & 2 & 3 \\
-    \hline
-    \hline
-    % Table begins here
-    Reporting Rate & $\reportRate$ & $\rho$ & $\rho$ & $\epsilon_1, \epsilon_2$ \\\hline
-    Mixing Coefficient & $\mixExponent$ & $\nu$ & --- & --- \\\hline
-    Measurement Over-Dispersion & $\obsOverdispersion$ & $\tau$ & --- & $p$ \\\hline
-    Birth Rate & $\muBirth$ & $\mu$ & --- & --- \\\hline
-    Natural Mortality Rate & $\muDeath$ & $\delta$ & --- & $\mu$ \\\hline
-    Cholera Mortality Rate & $\choleraDeath$ & --- & --- & $\alpha$ \\\hline
-    Latent Period & $1/\muEI$ & $1/\sigma$ & $1/\gamma_E$ & --- \\\hline
-    Recovery Rate & $\muIR$ & $\gamma$ & $\gamma$ & $\gamma$ \\\hline
-    Loss of Immunity & $\muRS$ & $\alpha$ & $\sigma$ & $\rho$ \\\hline
-    Symptomatic Ratio & $\symptomFrac$ & $1 - \theta_0$ & $k$ & $\sigma$ \\\hline
-    Asymptomatic Relative Infectiousness & $\asymptomRelativeInfect$ & $\kappa$ & $red_\beta$ & --- \\\hline
-    Human-to-Water Shedding & $\Wshed$ & --- & $\mu$ & $\theta_I$ \\\hline
-    Asymptomatic Relative Shedding & $\asymptomRelativeShed$ & --- & $red_\mu$ & $\theta_A/\theta_I$ \\\hline
-    Seasonal Amplitude & $\seasAmplitude$ & --- & $\alpha_s$ & $\lambda$ \\\hline
-    Transmission & $\transmission$ & $\beta$ & $\beta$ & $c$ \\\hline
-    Water-to-Human & $\Wbeta{}$ & --- & $\beta_W$ & $\beta$ \\\hline
-    Bacteria Mortality & $\Wremoval$ & --- & $\delta$ & $\mu_\beta$ \\\hline
-    Vaccination Efficacy & $\vaccineEfficacy$ & $\theta_{vk}$ & $\theta_1, \theta_2, \theta_{1_5}, \theta_{2_5}$ & $\eta_{1d}, \eta_{2d}$ \\\hline
-    Process Over-dispersion & $\sigmaProc$ & --- & --- & $\sigma_w$\\\hline
-    % Half Saturation Constant & $\Wsat$ & --- & $Sat$ & $K$ \\\hline
-  \end{tabular}
-  \end{center}
-  \caption{
-  \label{tab:translate}Translations between our common notation and notation used by \cite{lee20}
-  }
-\end{table}
-
-\clearpage
+%%%%%%%%%% END
 

### 2023-09-05 si/si.Rnw

--- 2023-08-31 si/si.Rnw
@@
 library(doRNG)
+library(knitr)
+
+options(
+  scipen = 2,
+  help_type = "html",
+  stringsAsFactors = FALSE,
+  continue = "+  ",
+  width = 70,
+  useFancyQuotes = FALSE,
+  reindent.spaces = 2,
+  xtable.comment = FALSE
+)
+
+opts_knit$set(concordance=TRUE)
+opts_chunk$set(
+  tidy.opts=list(
+    keep.blank.line=FALSE
+  ),
+  comment="",
+  # warning=FALSE,
+  # message=FALSE,
+  echo=FALSE,
+  fig.align="center",
+  dev="pdf",
+  dev.args=list(
+    bg="transparent",
+    pointsize=9
+  )
+)
 
@@
 @
-
-<<set-opts, include=FALSE>>=
-options(
-  scipen = 2,
-  help_type = "html",
-  stringsAsFactors = FALSE,
-  continue = "+  ",
-  width = 70,
-  useFancyQuotes = FALSE,
-  reindent.spaces = 2,
-  xtable.comment = FALSE
-)
-@
-
-<<knitr-opts, include=FALSE, purl=FALSE>>=
-library(knitr)
-opts_knit$set(concordance=TRUE)
-opts_chunk$set(
-  progress=TRUE,prompt=TRUE,highlight=FALSE,
-  tidy=TRUE,
-  tidy.opts=list(
-    keep.blank.line=FALSE
-  ),
-  comment="",
-  warning=FALSE,
-  message=FALSE,
-  error=TRUE,
-  echo=FALSE,
-  cache=FALSE,
-  strip.white=TRUE,
-  results="markup",
-  fig.path="figure/",
-  fig.lp="fig:",
-  fig.align="center",
-  fig.show="asis",
-  dev="pdf",
-  dev.args=list(
-    bg="transparent",
-    pointsize=9
-  )
-)
-
-myround <- function (x, digits = 1) {
-  # taken from the broman package
-  if (digits < 1)
-    stop("This is intended for the case digits >= 1.")
-  if (length(digits) > 1) {
-    digits <- digits[1]
-    warning("Using only digits[1]")
-  }
-  tmp <- sprintf(paste("%.", digits, "f", sep = ""), x)
-  zero <- paste0("0.", paste(rep("0", digits), collapse = ""))
-  tmp[tmp == paste0("-", zero)] <- zero
-  tmp
-}
-
-@
-
-%% copied from pnas-new.cls
-\usepackage{amsmath,amsfonts,amssymb,graphicx}
-
-%%%% parameters %%%%%%%%%%5
-\newcommand\Wsat{W_{\mathrm{sat}}}
-\newcommand\muIR{\mu_{IR}}
-\newcommand\muEI{\mu_{EI}}
-\newcommand\transmission{\beta}
-\newcommand\seasAmplitude{a}
-\newcommand\rainfallExponent{r}
-\newcommand\muRS{\mu_{RS}}
-\newcommand\vaccineEfficacy{\theta}
-\newcommand\muBirth{\mu_S}
-\newcommand\muDeath{\delta}
-\newcommand\choleraDeath{\delta_{C}}
-\newcommand\symptomFrac{f}
-\newcommand\asymptomRelativeInfect{\epsilon}
-\newcommand\asymptomRelativeShed{\epsilon_{W}}
-\newcommand\Wbeta[1]{\beta_{W#1}}
-\newcommand\Wremoval{\delta_W}
-\newcommand\Wshed{\mu_W}
-\newcommand\mixExponent{\nu}
-\newcommand\sigmaProc{\sigma_{\mathrm{proc}}}
-\newcommand\reportRate{\rho}
-\newcommand\obsOverdispersion{\psi}
-\newcommand\phaseParm{\phi}
-\newcommand\transmissionTrend{\zeta}
-\newcommand\Binit{\xi}
-\newcommand\vaccClass{Z}
-\newcommand\vaccCounter{z}
-\newcommand\modelCounter{m}
-\newcommand\Whur[1]{\beta_{W#1}^{hm}}
-\newcommand\hHur[1]{h_{#1}^{hm}}
-\newcommand\Iinit{I_{0,0}}
-\newcommand\Einit{E_{0, 0}}
-\newcommand\reportChange{\delta}
-
-\newcommand\missing{\hspace{6mm} ---}
-\newcommand\fixed{\color{blue}}
-\newcommand\demography{\bullet}
-
-\newcommand\mytitle{Informing policy via dynamic models: Cholera in Haiti}
-\newcommand\comma{{\hspace{-0.25mm},\hspace{-0.2mm}}}
-
-%%%%%%%%%%%%%%%% START OF USER DEFINITIONS %%%%%%%%%%%%%%%5
-
-\newcommand\siOnly[1]{} %% redefined for the supplement
-\newcommand\msOnly[1]{#1} %% redefined for the supplement
-
-\newcommand\code[1]{\texttt{#1}}
-\usepackage{xurl}
-% \usepackage[breaklinks]{hyperref}
-\usepackage{makecell}
-\usepackage{multirow}
-\usepackage{multicol}
-\usepackage[ruled,noline,linesnumbered]{algorithm2e}
-% \usepackage{setspace}
-
-%% EDITING MACROS
-\usepackage{color}
-\usepackage[normalem]{ulem}% to use \sout in feedback commands
-\usepackage{bm}
-% orange for EI
-\definecolor{orange}{rgb}{1,0.5,0}
-\newcommand\ei[2]{\sout{#1} \textcolor{orange}{#2}}
-\newcommand\eic[1]{\textcolor{orange}{[#1]}}
-% green for JW
-\definecolor{green}{rgb}{0,0.5,0}
-\newcommand\jw[2]{\sout{#1} \textcolor{green}{#2}}
-\newcommand\jwc[1]{\textcolor{green}{[#1]}}
-% purple for JJ
-\definecolor{purple}{rgb}{0.5,0,1}
-\newcommand\jj[2]{\sout{#1} \textcolor{purple}{#2}}
-\newcommand\jjc[1]{\textcolor{purple}{[#1]}}
-% cyan for AR
-\definecolor{cyan}{rgb}{0,.5,.5}
-\newcommand\ar[2]{\sout{#1} \textcolor{cyan}{#2}}
-\newcommand\arc[1]{\textcolor{cyan}{#1}}
-% light brown for KT
-\definecolor{lightbrown}{rgb}{0.5,0.5,0}
-\newcommand\kt[2]{\sout{#1} \textcolor{lightbrown}{#2}}
-\newcommand\ktc[1]{\textcolor{lightbrown}{#1}}
-
-%% basic POMP definitions
-\newcommand\Xspace{{\mathbb X}}
-\newcommand\Yspace{{\mathbb Y}}
-\newcommand\Thetaspace{\R^{\Thetadim}}
-\newcommand\Xunitspace{{\scriptsize{\mathcal X}}}
-\newcommand\Yunitspace{{\tiny{\mathcal Y}}}
-\newcommand\hatTheta{\widehat{\Theta}}
-\newcommand\Xdim{{\mathrm{dim}}(\Xspace)}
-\newcommand\Ydim{{\mathrm{dim}}(\Yspace)}
-\newcommand\Thetadim{P}
-\newcommand\thetadim{p}
-\newcommand\timeSet{{\mathbb T}}
-\newcommand\data[1]{#1^*}
-
-\newcommand\unitSpecific{\hspace{0.1mm}\mathrm{us}}
-\newcommand\shared{\hspace{0.15mm}\mathrm{sh}}
-
-% for algorithm pseudocode
-\SetKwFor{For}{for}{:}{end}
-
-%% for particle filters
-\newcommand\unit{u}
-\newcommand\altUnit{\tilde u}
-\newcommand\Unit{U}
-\newcommand\rep{i}
-\newcommand\Rep{\mathcal{I}}
-\newcommand\island{\rep}
-\newcommand\Island{\Rep}
-%\newcommand\txtisland{replicate}
-\newcommand\altIsland{\tilde i}
-\usepackage[mathscr]{euscript}
-
-\renewcommand\time{n}
-\newcommand\myvec[1]{\boldsymbol{#1}}
-\newcommand\altTime{\tilde n}
-\newcommand\Time{N}
-\newcommand\Np{J}
-\newcommand\np{j}
-\newcommand\altNp{k}
-\newcommand\altAltNp{\tilde j}
-\newcommand\resampleIndex{r}
-\newcommand\unitWeight{w}
-
-%% for all iterated filtering algorithms
-\newcommand\Nit{M}
-\newcommand\nit{m}
-
-%% customized math macros
-\newcommand\prob{\mathbb{P}}
-\newcommand\E{\mathbb{E}}
-\newcommand\dd[1]{\mathrm{d}{#1}}
-\newcommand\given{{\,\vert\,}}
-\newcommand\equals{{{\,}={\,}}}
-\newcommand\myequals{\hspace{0.5mm}{=}\hspace{0.5mm}}
-\newcommand\myto{{\;:\;}}
-\newcommand\seq[2]{{#1}\!:\!{#2}}
-\newcommand\mydot{{\,\cdot\,}}
-\newcommand\cp[2]{N_{\mathrm{#1}\mathrm{#2}}}
-\newcommand\giventh{{\hspace{0.5mm};\hspace{0.5mm}}}
-\newcommand\normal{\mathcal{N}}
-\newcommand\argequals{{\,=\,}}
-\newcommand\lags{c}
-\newcommand\maxlag{\overline{c}}
-\newcommand\nlfList{C}
-\newcommand\bigO{\mathcal{O}}
-\newcommand\loglik{\lambda}
-\newcommand\loglikMC{\MC{\loglik}}
-\newcommand\R{\mathbb{R}}
-\newcommand\param{\,;}
-\newcommand\mycolon{{\hspace{0.6mm}:\hspace{0.6mm}}}
-\newcommand\MC[1]{#1^{\,\mbox{\tiny MC}}}
-\newcommand\EMC{{\E}}
-\newcommand\Var{\mathrm{Var}}
-\newcommand\var{\Var}
-\newcommand\Cov{\mathrm{Cov}}
-\newcommand\cov{\Cov}
-\newcommand\iid{\mathrm{iid}}
-\newcommand\dist{d}
-\newcommand\transpose{\top}
-
-\newcommand\gnsep{,}
-
-\def\lik{L}
-\def\loglik{\ell}
-
-%% THEOREM-LIKE ENVIRONMENTS
-\usepackage{amsthm}
-\newtheorem{prop}{Proposition}
-\newtheorem{theorem}{Theorem}
-\newtheorem{lemma}{Lemma}
-\newtheorem{example}{Example}
-\newtheorem{remark}{Remark}
-\newtheorem{cor}{Corrolary}
-\newtheorem{fact}{Fact}
-\newtheorem{assumption}{Assumption}
-\newtheorem{assumptionB}{Assumption}
-\renewcommand\theassumption{A\arabic{assumption}}
-\renewcommand\theassumptionB{B\arabic{assumptionB}}
-
-%% FOR PSEUDOCODE SETUP
-\newcommand\mystretch{\rule[-2mm]{0mm}{5mm} }
-\newcommand\asp{\hspace{4mm}}
-\newcommand\codeIndent{\hspace{4mm}}
-
-%%%%%%%%%%%%%%%% END OF USER DEFINITIONSS %%%%%%%%%%%%%%%
-
-\bibliographystyle{apalike}
-\usepackage{natbib}
-\usepackage{fullpage}
-
-\newcommand\myeqref[1]{(\ref{#1})}
-
-\renewcommand{\contentsname}{Supplementary Content}
-\renewcommand{\refname}{Supplementary References}
-\renewcommand\thefigure{S-\arabic{figure}}
-\renewcommand\thetable{S-\arabic{table}}
-\renewcommand\thesection{S\arabic{section}}
-\renewcommand\theequation{S\arabic{equation}}
-\renewcommand\theprop{S\arabic{prop}}
-\renewcommand\thelemma{S\arabic{lemma}}
-\renewcommand\thealgocf{S\arabic{algocf}}
-
-\setcounter{tocdepth}{1}
 
@@
 
-\section{Markov chain and differential equation interpretations of compartment flow rates}
+\section{Model Diagrams}
 
-In Sections 2.1, 2.2 and 2.3 of the main article, we define compartment models in terms of their flow rates.
-For a discrete population model, these rates define a Markov chain.
-For a continuous and deterministic model, the rates define a system of ordinary differential equations.
-Here, we add additional details to clarify the mapping from a collection of rate functions to a fully specified process.
-Our treatment follows \citet{breto09}.
-
-A general compartment model is a vector-valued process $X(t)=(X_1(t),\dots,X_c(t))$ denoting the (integer or real-valued) counts in each of $c$ compartments, where $t$ is any continuous value in the interval  $\left[t_0, \infty\right)$ for some real valued starting time $t_0$.
-The compartments may also have names, but to set up general notation we simply refer to them by their numerical index.
-The basic characteristic of a compartment model is that $X(t)$ can be written in terms of the flows $N_{ij}(t)$ from $i$ to $j$.
-A flow into compartment $i$ from outside the system is denoted by $N_{\demography i}$, and a flow out of the system from compartment $i$ is denoted by $N_{i\demography}$.
-We call $\demography$ a source/sink compartment, though it is an irregular compartment since $X_{\demography}(t)$ is not defined.
-These flows are required to satisfy a ``conservation of mass'' identity:
-\begin{equation}
-X_i(t)=X_i(t_0)+N_{\demography i}(t) - N_{i\demography}(t) + \sum_{j\neq
-i}N_{ji}(t)-\sum_{j\neq i}N_{ij}(t). \label{eq:conservation}
-\end{equation}
-Each {\em flow} $N_{ij}(t)$ is associated with a {\em rate} function
-$\mu_{ij}=\mu_{ij}(t,X(t))$, where we include the possibility that $i$ or $j$ takes value $\demography$.
-
-There are different ways to use a collection of rate functions to build a fully specified model.
-We proceed to describe the ones we use in this paper: via a system of ordinary differential equations (Sec.~\ref{subsec:ode}), a simple Markov counting system (Sec.~\ref{subsec:smcs}), and an overdispersed Markov counting system (Sec.~\ref{subsec:odmcs}).
-Other representations include stochastic differential equations driven by Gaussian noise or Gamma noise \citep{bhadra11}.
-
-\subsection{Ordinary differential equation (ODE) interpretation}
-\label{subsec:ode}
-
-A basic deterministic specification is
-\begin{equation}
-\label{eq:ode1}
-dN_{ij}/dt = \mu_{ij}\big(t,X(t) \big) X_i(t), \hspace{3mm} i\in \seq{1}{c}, \hspace{3mm} j\in \seq{1}{c} \cup\{\demography\}, \hspace{3mm} i\neq j,
-\end{equation}
-where $\mu_{ij}\big(t,X(t)\big)$ is called a per-capita rate or a unit rate.
-Flows into the system require special treatment since $X_i(t)$ in \myeqref{eq:ode1} is not defined for $i=\demography$.
-Instead, we specify
-\begin{equation}
-\label{eq:ode2}
-dN_{\demography i}/dt = \mu_{\demography i}\big(t,X(t) \big).
-\end{equation}
-This is the the interpretation and implementation used for Model~2 in our study.
-
-\subsection{Simple Markov counting system interpretation}
-\label{subsec:smcs}
-A continuous time Markov chain can be specified via its infinitesimal transition probabilities.
-A basic approach to this is to define
-\begin{eqnarray}
-\label{eq:smcs1}
-\prob\big[ N_{ij}(t+\delta)-N_{ij}(t)=0 \given X(t)\big]
- &=& 1-\delta \mu_{ij}\big(t,X(t)\big)X_i(t) + o(\delta),
-\\
-\label{eq:smcs2}
-\prob\big[ N_{ij}(t+\delta)-N_{ij}(t)=1 \given X(t)\big]
- &=& \delta \mu_{ij}\big(t,X(t)\big)X_i(t) + o(\delta),
-\end{eqnarray}
-for $i\in \seq{1}{c}$ and $j\in\seq{1}{c}\cup\{\demography\}$ with $i\neq j$.
-As with the ODE case, we need special attention for flows into the system, and we define
-\begin{eqnarray}
-\label{eq:smcs3}
-\prob\big[ N_{\demography i}(t+\delta)-N_{\demography i}(t)=0 \given X(t)\big]
- &=& 1-\delta \mu_{\demography i}\big(t,X(t)\big) + o(\delta),
-\\
-\label{eq:smcs4}
-\prob\big[ N_{\demography i}(t+\delta)-N_{\demography i}(t)=1 \given X(t)\big]
- &=& \delta \mu_{\demography i}\big(t,X(t)\big) + o(\delta).
-\end{eqnarray}
-Together with the initial conditions $X(0)$, equations \myeqref{eq:smcs1}--\myeqref{eq:smcs4} define a Markov chain.
-Each flow is a simple counting process, meaning a non-decreasing integer-valued process that only has jumps of size one.
-We therefore call the Markov chain a simple Markov counting system (SMCS).
-The infinitesimal mean of every flow is equal to its infinitesimal variance \citep{breto11} and so an SMCS is called equidispersed.
-We note that the special case of Model~1 used by \cite{lee20} (with $\sigmaProc = 0$) is an SMCS.
-To permit more general mean-variance relationships for a Markov counting system, we must permit jumps of size greater than one.
-The utility of overdispersed models, where the infinitesimal variance of the flow exceeds the infinitesimal mean, has become widely recognized \citep{stocks20,he10}.
-
-\subsection{Overdispersed Markov counting system interpretation}
-\label{subsec:odmcs}
-
-Including white noise in the rate function enables the possibility of an overdispersed Markov counting system \citep{breto11,breto09,he10}.
-Since rates should be non-negative, Gaussian noise is not appropriate and gamma noise is a convenient option that has found various applications \citep{romero-severson15, subramanian20}.
-Specifically, we consider a model given by
-\begin{equation}
-\label{eq:odmcs1}
-\mu_{ij}\big(t,X(t)\big) = \bar\mu_{ij}\big(t,X(t)\big) \, d\Gamma_{ij}(t)/dt,
-\end{equation}
-where $\Gamma_{ij}(t)$ is a stochastic process having independent gamma distributed increments, with
-\begin{equation}
-\label{eq:odmcs2}
-\E\big[\Gamma_{ij}(t)\big] = t, \quad \var\big[\Gamma_{ij}(t)\big] = \sigma_{ij}^2 t.
-\end{equation}
-Formally interpreting the meaning of \myeqref{eq:odmcs1} is not trivial, and we do so by constructing a Markov process $X(t)$ as the limit of the Euler scheme descrived in Section~\ref{sec:numerics}, below.
-Therefore, the numerical scheme in Sec.~\ref{sec:numerics} can be taken as a definition of the meaning of \myeqref{eq:odmcs1}.
-The Markov chain defined by the limit of this Euler scheme as the step size decreases is an overdispersed Markov counting system, with the possibility of instantaneous jumps of size greater than one \citep{breto11}.
-
-\section{Numerical solutions to compartment models}
-\label{sec:numerics}
-
-Models may be fitted and their implications assessed via numerical solutions (i.e., simulations) from the model equations.
-All the analyses we consider have this simulation-based property, known as plug-and-play or equation-free or likelihood-free.
-The numerical solutions to the model are arguably of more direct scientific interest than the exact solutions to the postulated equations.
-For ODE models, numerical methods are well studied and a standard numerical solution package such as \code{deSolve} in \code{R} is adequate for many purposes.
-For SMCS and ODMCS models, exact schemes are feasible when the number of events is small, which may be the case for small populations.
-However, for applicability to larger populations, we use instead the following Euler scheme.
-Write $\delta$ for an Euler time step, and $\Delta N_{ij}$ for the numerical approximation to $N_{ij}(t+\delta)-N_{ij}(t)$ given $X(t)$.
-For each $i$ and $j$ in $\seq{1}{c} \cup \{\demography\}$ with $i \neq j$, we draw independent Gamma distributed noise increments with mean $\delta$ and variance $\sigma_{ij}^2 \delta$, denoted using a mean-variance parameterization of the gamma distribution as
-\begin{equation}
-\label{eq:numerics1}
-\Delta\Gamma_{ij} \sim \mathrm{gamma}(\delta, \sigma_{ij}^2 \delta).
-\end{equation}
-In the case of an SMCS model, $\sigma_{ij}=0$ for all $i$ and $j$, so we have $\Delta\Gamma_{ij}=\delta$.
-Then, for $i\neq \demography$ and $j\neq i$, and writing
-\begin{equation}
-\label{eq:numerics2}
-\mu_{ij}=\bar\mu_{ij}\big(t,X(t)\big) \Delta\Gamma_{ij} / \delta,
-\end{equation}
-we calculate transition probabilities
-\begin{eqnarray}
-\label{eq:numerics3}
-p_{ij} &=& \exp\left\{-\sum_{k\in 1:c \, \cup \{\demography\}} \mu_{ik} \, \delta \right\}
-\frac{\mu_{ij}}{\sum_{k\in 1:c\cup \{\demography\}} \mu_{ik}},
-\\
-\label{eq:numerics4}
-p_{ii} &=& 1 - \sum_{j\neq i} p_{ij}.
-\end{eqnarray}
-These probabilities correspond to competing hazards for every individual in compartment $i$ to transition to some compartment $j$, interpreting $j=i$ to mean that the individual remains in $i$.
-Then, $\big(\Delta N_{i1},\dots,\Delta N_{ic},\Delta N_{i\demography}\big)$ has the multinomial distribution where $X_i(t)$ individuals are allocated independently to $\seq{1}{c}\cup\{\demography\}$ with probabilities given by \eqref{eq:numerics3} and \eqref{eq:numerics4}.
-We use the \code{reulermultinom} function in the \code{pomp} package to draw from this multionomial distribution.
-
-Different treatments of demographic flows---such as birth, death, immigration and emigration---are possible.
-For the case $i=\demography$, the treatment used by Model~1 is to set
-\begin{equation}
-\label{eq:numerics6}
-\Delta N_{\demography j} \sim \mathrm{poisson}( \mu_{\demography j} \delta),
-\end{equation}
-an independent Poisson random variable with mean $\mu_{\demography j} \delta$.
-
-Models~2 and 3 used an alternative approach, balancing the total number of flows in and out of the compartment, i.e., $\sum_{i}N_{\demography i}(t) = \sum_{i}N_{i \demography}(t)$, in order to make the model consistent with the known total population.
-In this case, we formally model the death rate as a rate of returning to the susceptible class $S$, and use external transitions from $\demography$ into $S$ to describe only net population increase.
-
-\section{Measurement Models}
-
-Each POMP requires specification of a measurement model, which is a statistical description of how observations on the system are obtained.
-In general, we used the same measurement models that were reported by \cite{lee20} unless specifically noted in the following subsections.
+Each of the dynamic models considered in this manuscript can be fully described using the model descriptions in the manuscript, coupled with the additional information described in Sections 2 and 3 of this supplement.
+Despite this, diagrams of dynamic systems are often helpful to understand the equations.
+In this section, we give three diagrams representing Models~1--3, respectively.
+Because the models are defined by their mathematical equations and numeric implementation, these diagrams are not unique visual representations of the model.
+Alternative representations that may be helpful in understanding the models explored in this paper were made by \citet{lee20sup}.
 
@@
 
-In this model, the advantage afforded by vaccination is an increased probability that an infection is asymptomatic.
-Therefore, under the assumptions of this model, all reported cases are assumed to be a fraction of individuals that transition from the exposed to the infected compartment, as noted in Eq.~\myeqref{model1:meas}:
-\begin{equation}
-  \label{model1:meas}
-  Y_{n} \mid \Delta N_{E_{\cdot}I_{\cdot}}(n) = \reportChange \sim \text{NB}\left(\reportRate \reportChange, \obsOverdispersion \right),
-\end{equation}
-where $Y_n$ is the reported cholera cases at time $t_n \in t_1:t_N$ and $\Delta N_{E_{\cdot}I_{\cdot}}(n)$ is the sum total of individuals across each vaccination compartment $\vaccCounter \in 1:\vaccClass$ who moved from compartment $E_\vaccCounter$ to $I_\vaccCounter$ since observation $t_{n-1}$.
-Here, $\text{NB}\left(\reportRate \reportChange, \obsOverdispersion \right)$ denotes a negative binomial distribution with mean $\reportRate \reportChange$ and variance $\reportRate \reportChange \big(1 + \frac{\reportRate \reportChange}{\obsOverdispersion}\big)$.
+\input{inputs/mod1diagram}
 
-\subsection{Model~2}\label{sec:mod2Meas}
+\subsection{Model~2}
 
-Model~2 was fit using reported case counts that were transformed using the natural logarithm.
-We fit Model~2 using the subplex algorithm in the \code{subplex} package, using
-a Gaussian measurement model (Eq.~\myeqref{model2:meas}) on the log transformed cases within each unit.
-The final loss function that is maximized is the product of the likelihoods of the individual units, or the sum of the log-likelihood of the individual units.
-The measurement model for individual units is given in Eq.~\myeqref{model2:meas}.
-\begin{equation}
-  \label{model2:meas}
-  \log\big(Y_{u, n} + 1\big) \mid \Delta N_{E_{u\cdot}I_{u\cdot}}(n) = \reportChange_u \sim \text{N}\left(\log \big(\reportRate \reportChange_u + 1\big), \obsOverdispersion^2 \right),
-\end{equation}
-where $\Delta N_{E_{u\cdot}I_{u\cdot}}(n)$ is the sum total of individuals across vaccination compartment $\vaccCounter \in 0:4$ within unit $u$ who moved from compartment $E_{u\vaccCounter}$ to $I_{u\vaccCounter}$ since observation $t_{n-1}$.
-Therefore, because the natural logarithm of observed case counts (plus one, to avoid taking the logarithm of zero) has a normal distribution, $Y_{u, n} + 1$ is assumed to follow a log-normal distribution with log-mean parameter $\log\big(\reportRate \Delta N_{E_{u\cdot}I_{u\cdot}}(n) + 1\big)$ and log-variance $\obsOverdispersion^2$.
-We note that fitting a model with this measurement model is equivalent to fitting using least squares, with $\log(Y_{u, n} + 1)$ as the response variable.
+\input{inputs/mod2diagram}
 
-This measurement model differs from that used by \citet{lee20}, who fit the model in two stages: epidemic and endemic phases.
-Although their text and supplement material do not explicitly describe the measurement model used, inspection of the code provided with their submission suggests a change in measurement model between the epidemic and endemic phases.
-% In the file \code{choleraEqsPublic.py}, \citet{lee20} create several functions, where each function represents a set of coupled differential equations that could potentially be used to model cholera incidence data.
-% Each function returns a vector (or \code{numpy} array) that represents the change in each state variable for a single time step, including the variable \code{dC}, which tracks the number of new infections and is used to obtain the reported case counts.
-% Following their comments and code, it appears that the function \code{choleraEqs10WithoutVaccinationNetwork} was used to describe the dynamics of the epidemic phase, and \code{choleraEqs11WithoutVaccinationNetwork} was used for the endemic stage (see Sec~\ref{sec:mod2rep} for more details).
-% Because their models were fit using least squares, the code in these functions suggest that the measurement model for the epidemic phase is
-The measurement model they used for the epidemic phase is
-\begin{equation}
-  \label{model2:measEpi}
-  Y_{u, n} \mid \Delta N_{E_{u\cdot}I_{u\cdot}}(n) = \reportChange_u \sim \text{N}\left(\reportRate \reportChange_u, \obsOverdispersion^2 \right),
-\end{equation}
-% which is similar to our measurement model, the primary difference being that the measurement model is applied to raw case counts rather than log-transformed case counts.
-The measurement model they used for the endemic phase modifies the epidemic model by counting both asymptomatically infected (A) and symptomatically infected (I) individuals in the case counts:
-\begin{equation}
-  \label{model2:measEnd}
-  Y_{u, n} \mid \Delta N_{E_{u\cdot}I_{u\cdot}}(n) = \reportChange_{u1}, \Delta N_{E_{u\cdot}A_{u\cdot}}(n) = \reportChange_{u2} \sim \text{N}\left(\reportRate \big(\reportChange_{u1} + \reportChange_{u2}\big), \obsOverdispersion^2 \right),
-\end{equation}
-where the notation for $\Delta N_{E_{u\cdot}A_{u\cdot}}(n)$ is similar to $\Delta N_{E_{u\cdot}I_{u\cdot}}(n)$, described above.
+\newpage
 
@@
 
-In this model, reported cholera cases are assumed to stem from individuals who develop symptoms and seek healthcare.
-Therefore reported cases are assumed to come from an over-dispersed negative binomial model, given the increase in infected individuals:
-\begin{equation}
-  \label{model3:meas}
-  Y_{u, n} \mid \Delta N_{S_{u \cdot}I_{u}}(t) = \reportChange_u \sim \text{NB}\left(\reportRate \reportChange_{u}, \obsOverdispersion \right),
-\end{equation}
-where $\Delta N_{S_{u \cdot}I_{u \vaccCounter}}(n)$ is the number of individuals who moved from compartment $S_{u \vaccCounter}$ to $I_{u}$ since observation $t_{n-1}$.
+\input{inputs/mod3diagram}
 
-This measurement model is a minor change from that used by \cite{lee20}, which allowed for a change in the reporting rate on January 1st, 2018.
-The fitted values of the reporting rate---before and after January 2018---were $0.97$ and $0.097$, respectively.
-An instantaneous change from near perfect to almost non-existent reporting can be problematic, as it forces the model to explain the observed reduction in reported cases as a decrease in the reporting of cases, rather than a decrease in the prevalence of cholera.
-This shift was justified by a ``change of the case definition that occurred on January 1st, 2018";
-this claim was not cited, and we could find no evidence that such a drastic change in the reporting rate would be warranted.
-We therefore do not allow a change in reporting rate when fitting Model~3.
+\input{inputs/modelDetails}
 
-\section{Initial Values}
+\input{inputs/measurementModels}
 
-To perform inference on POMP models, it is necessary to propose an initial probability density for the latent process $f_{X_0}(x_0;\theta)$, including the possibility that the initial values of the latent states are known, or are a non-random function of the unknown parameter vector, $\theta$.
-This density is used to obtain initial values of the latent state when fitting and evaluating the model.
-For each of the models considered in this analysis, the initial conditions are derived by enforcing the model dynamics on reported cholera cases.
-It is also sometimes necessary to fit some initial value parameters in order to help determine initial values for weakly identifiable compartments.
-In the following subsections, we mention initial value parameters that were fit for each model.
+\input{inputs/initialValues}
 
-\subsection{Model~1}
-
-For this model, the number of individuals in the Recovered and Asymptomatic compartments are set to zero, but the initial proportion of Infected and Exposed individuals is estimated as initial value parameters ($\Iinit$ and $\Einit$, respectively) using the IF2 algorithm, implemented as \texttt{mif2} in the \texttt{pomp} package.
-Finally, the initial proportion of Susceptible individuals $S_{0, 0}$ is calculated as $S_{0, 0} = 1 - \Iinit - \Einit$.
-This model for the initial values of the latent states matches that which was used by \citet{lee20}.
-
-\subsection{Model~2}
-
-Model~2 assumes that the initial values are a deterministic function of the reporting rate and the initial case reports, and so no initial value parameters need to be estimated.
-Initial values for latent state compartments are chosen so as to enforce the model dynamics on the observed number of cases.
-Specifically, the model sets $I_{u0}(0) = y^*_{1u} / \reportRate$ for each unit $u \in 1:10$, where $I_{u0}(t)$ is the number of infected individuals in unit $u$ at time $t$, vaccination scenario $z = 0$, $y^*_{tu}$ is the reported number of cases, and $\reportRate$ is the reporting rate.
-This model for the initial values of the latent states matches that which was used by \citet{lee20}.
-
-\subsection{Model~3}
-
-The latent states of this model are initialized by enforcing the model dynamics on the incidence data from the start of the recorded cases until time $t_0$, requiring that some of the available data be used to determine the initial values of the latent states.
-This is the same approach that was taken by \citet{lee20}, who used the value $t_0 = \text{2014-02-22}$; this choice of $t_0$ results in modeling roughly only $60\%$ of the available data, some of which is later discarded for alternative reasons \citep{lee20sup}.
-
-We do not see any immediate reason that this model could not be extended to cover a larger range of the data, and chose the value $t_0 = \text{2010-11-13}$.
-This choice of $t_0$ corresponds to using approximately $1\%$ of the available data to determine initial values of the latent states.
-In addition to modeling a larger portion of the available data, this choice of $t_0$ corresponds to an important real-world event, as daily reports from each of the departments were not being sent to the health ministry until November 10, 2010; % TODO: Cite
-this choice of $t_0$ therefore makes $\bm{Y}\big(t_1\big)$ the first week of data once daily reports are being sent to the health ministry.
-The few observation times that exist before $t_0$ are used to initialize the model by enforcing model dynamics on these preliminary observations.
-For convenience, we denote these observations as $t_{-3}, t_{-2}$ and $t_{-1}$; as before, we let $y_{u,-k} = Y_{u}\big(t_{-k}\big)$ denote the observed case count for unit $u$ at time point $t_{-k}$, where $k \in 1:3$.
-Equations for the initial values of non-zero latent states are provided in Eqs.~\myeqref{eq:model3:I0}--\myeqref{eq:model3:B0}.
-
-\begin{eqnarray}
-\label{eq:model3:I0}
-I_{u}\big(t_0\big) &=& \frac{y^*_{u, -1}}{7\reportRate\left(\muIR + \left(\muDeath + \choleraDeath\right)/365\right)},
-\\%[-3pt]
-\label{eq:model3:A0}
-A_{u}\big(t_0\big) &=& \frac{I_{u0}\big(t_0\big)\left(1 - \symptomFrac\right)}{\symptomFrac},
-\\
-\label{eq:model3:R0}
-R_{u01}\big(t_0\big) &=& R_{u02}\big(t_0\big) = R_{u03}\big(t_0\big) = \left(\frac{\sum_{k = -3}^0 y^*_{u, k}}{\reportRate \symptomFrac} - \left(I_{u0}\big(t_0\big) + A_{u0}\big(t_0\big)\right)\right) / 3
-\\
-\label{eq:model3:S0}
-S_{u0}\big(t_0\big) &=& \mathrm{Pop}_u - I_{u}\big(t_0\big) - A_{u}\big(t_0\big) - \sum_{k=1}^3 R_{u0k}\big(t_0\big)
-\\
-\label{eq:model3:B0}
-B_{u}\big(t_0\big) &=& \big[1 + \seasAmplitude \tilde{J}^r \big] \mathrm{Den}_u \, \Wshed \big[ I_{u}(t_0)+ \asymptomRelativeShed A_{u_0}(t) \big]/\Wshed.
-\end{eqnarray}
-
-In Eq.~\myeqref{eq:model3:B0}, $\tilde{J} = 0.002376$ is the median adjusted rainfall over the observation period.
-One important consideration to make with this parameter initialization model is when $y_{u, -1}^* = 0$, which occurs for units $u \in \{3, 4\}$, which correspond to the Grand'Anse and Nippes departments, respectively.
-When this is the case, each of the infectious $I_u(t_0)$, asymptomatic $A_u(t_0)$, and bacterial reservoir $W_u(t_0)$ compartments have a value of zero.
-Because Model~3 models cholera transmission primarily by means of the bacterial reservoir, this makes it nearly impossible for an outbreak to occur.
-Therefore for units $u \in \{3, 4\}$, we introduce initial value parameters $I_{0,0}^3$ and $I_{0,0}^4$, and calibrate these parameter values using the data.
-The resulting parameter estimates are used to obtain the remaining non-zero initial values of the latent states using Eqs.~\myeqref{eq:model3:A0}--\myeqref{eq:model3:B0}.
-
-% Argue why we didn't need to do this for Model 2.
-
-% TODO: Check this section.
-
-\section{Calibrating Model~3 to observed cases}\label{sec:mod3Cal}
-
-<<CalibrateOrig3>>=
-# Create vectors for the unit and shared parameters
-unit_specific_names <- c("betaB", "foi_add")
-shared_param_names <- c(
-  "mu_B", "XthetaA", "thetaI", "lambdaR", "r", "std_W",
-  "epsilon", "k"
-)
-est_param_names <- c(
-  unit_specific_names, shared_param_names
-)
-
-# Add unit numbers to each parameter
-est_param_names_expanded <- paste0(rep(est_param_names, each = 10), 1:10)
-
-# Simple function that is used to set the rw_sd for the first search
-# This function is convenient so that we don't have to type out the rw_sd
-# for each unit parameter.
-set_rw_1 <- function(x) {
-  if (gsub("[[:digit:]]+$", "", x) %in% shared_param_names) {
-    return(0.01)
-  # } else if (x %in% paste0(rep(c("aHur", 'hHur'), each = 2), c(3, 9))) {
-  #   return(expression(ifelse(time >= 2016.754 & time <= 2017, 0.015, 0)))
-  # } else if (x %in% c("Iinit3", "Iinit4")) {
-  #   return(expression(ivp(0.25)))
-  } else if (grepl('^foi_add[[:digit:]]+$', x)) {
-    return(0.015)
-  } else if (grepl('^betaB[[:digit:]]+$', x)) {
-    return(0.01)
-  } else {
-    return(0)
-  }
-}
-
-set_rw_2 <- function(x) {
-  if (gsub("[[:digit:]]+$", "", x) %in% shared_param_names) {
-    return(0.003)
-  # } else if (x %in% paste0(rep(c("aHur", 'hHur'), each = 2), c(3, 9))) {
-  #   return(expression(ifelse(time >= 2016.754 & time <= 2017, 0.005, 0)))
-  # } else if (x %in% c("Iinit3", "Iinit4")) {
-  #   return(expression(ivp(0.125)))
-  } else if (grepl('^foi_add[[:digit:]]+$', x)) {
-    return(0.005)
-  } else if (grepl('^betaB[[:digit:]]+$', x)) {
-    return(0.003)
-  } else {
-    return(0)
-  }
-}
-
-# First Search RW
-reg_rw_1.sd <- lapply(est_param_names_expanded, set_rw_1)
-names(reg_rw_1.sd) <- est_param_names_expanded
-chol_rw_1.sd <- do.call(rw_sd, reg_rw_1.sd)
-
-# Second Search RW
-reg_rw_2.sd <- lapply(est_param_names_expanded, set_rw_2)
-names(reg_rw_2.sd) <- est_param_names_expanded
-chol_rw_2.sd <- do.call(rw_sd, reg_rw_2.sd)
-
-SEARCH1 = list(
-  NBPF = 100,
-  NP = 2500,
-  SPAT_REGRESSION = 0.05,
-  NREPS = 144,
-  NP_EVAL = 2500,
-  NREPS_EVAL = 18,
-  RW_SD = chol_rw_1.sd,
-  COOLING = 0.5,
-  KEEP_TRACES = FALSE,
-  KEEP_LIKE_MAT = FALSE
-)
-
-SEARCH2 = list(
-  TOP_N = 12,
-  NBPF = 100,
-  NP = 2500,
-  SPAT_REGRESSION = 0.05,
-  NREPS = 6,
-  NP_EVAL = 2500,
-  NREPS_EVAL = 18,
-  RW_SD = chol_rw_2.sd,
-  COOLING = 0.5,
-  KEEP_TRACES = FALSE,
-  KEEP_LIK_MAT = TRUE
-)
-
-haiti3_sub_fit <- bake(
-  file = "../model3/si/subset_noparams.rds", {
-    fit_haiti3(
-      nsearches = 2L,
-      search1 = SEARCH1,
-      search2 = SEARCH2,
-      search_rho = FALSE,
-      search_gamma = FALSE,
-      search_Iinit = FALSE,
-      search_hur = FALSE,
-      ncores = cores
-    )
-  }
-)
-
-# Find model index with best log-likelihood
-best_m <- which.max(haiti3_sub_fit$search2$logLiks$logLik)
-
-# Find parameters using best_m index
-sub_params <- haiti3_sub_fit$search2$params[best_m, ]
-
-# Get unit log-likelihoods for the best model.
-all_liks <- as.data.frame(haiti3_sub_fit$search2$likMat)
-# all_liks$which <- rep(1:(SEARCH3$TOP_N * SEARCH3$NREPS), each = SEARCH3$NREPS_EVAL)
-all_liks$which <- rep(1:(12 * 6), each = 18)
-best_mod <- filter(all_liks, which == best_m)
-
-# Pivot longer to get estimate of log-likelihood for each unit.
-unitLikesM3 <- best_mod %>%
-  pivot_longer(
-    cols = -which,
-    names_to = "unit",
-    names_prefix = "V",
-    values_to = "logLik"
-  ) %>%
-  select(-which) %>%
-  group_by(unit) %>%
-  summarize(
-    ll_est = logmeanexp(logLik),
-    ll_se = logmeanexp(logLik, se = TRUE)[2]
-  ) %>%
-  ungroup() %>%
-  mutate(unit = as.numeric(unit)) %>%
-  arrange(unit)
-@
-
-<<M3_bench, echo=FALSE, message=FALSE, include=FALSE, cache=TRUE>>=
-h3 <- haiti3_spatPomp()
-deps <- h3@unit_names
-
-mod3_cases <- as.data.frame(t(h3@data))
-colnames(mod3_cases) <- deps
-m3_inits <- c(7362, 1914, 22, 8, 2818, 82, 2499, 6953, 44, 7) # TODO: Also fit by hand, but could be changed to initializing to the observation before.
-names(m3_inits) <- deps
-
-bench_cond_lls <- list()
-for (dep in deps) {
-
-  # Get department cases
-  cases <- as.numeric(mod3_cases[, dep])
-  mod <- ar1_NegBinom(data = cases, init = m3_inits[dep])
-
-  bench_cond_lls[[dep]] <- haitipkg:::.ar1_NegBinom_cond_ll(
-    theta = mod$theta, init = m3_inits[dep], data = cases, transform = FALSE
-  )
-}
-
-bench_cond_lls <- as.data.frame(bench_cond_lls)
-
-bench_lls <- apply(bench_cond_lls, 2, sum)
-
-unitLikes_bench <- data.frame(
-  unit = 1:10,
-  # dep = names(bench_lls),
-  ll = bench_lls
-) %>%
-  as_tibble()
-
-bench_cond_lls$date <- lubridate::round_date(lubridate::date_decimal(h3@times), unit = 'day')
-@
-
-<<combine_bench_M3, echo=FALSE>>=
-
-# Combine tables with model 3 and benchmark likelihoods.
-bench_M3_joined <- dplyr::left_join(
-  x = unitLikesM3,
-  y = unitLikes_bench,
-  by = "unit"
-) %>%
-  select(-ll_se) %>%
-  mutate(unit = as.character(unit)) %>%
-  pivot_longer(
-    cols = -unit,
-    names_to = "type",
-    values_to = "ll"
-  ) %>%
-  mutate(unit = paste0(unit, ": ", deps[as.numeric(unit)]))
-@
-
-<<M3bpf, echo=FALSE, message=FALSE, include=FALSE>>=
-
-# Change model parameters to the best set of parameters from previous
-# search
-coef(h3)[names(sub_params)] <- sub_params
-
-registerDoRNG(40765101)  # for modular, reproducible results
-
-# Perform block partical filter with fitted model, to get conditional
-# log-likelihoods.
-bpf_sub3_out <- bake(
-  file = "../model3/si/bpf_sub_cond_LL.rds", {
-
-    results <- list()
-
-    bpf_out <- h3 %>% bpfilter(
-      Np = 5000,
-      block_size = 1,
-      save_states = TRUE
-    )
-
-    results$block.cond.loglik <- bpf_out@block.cond.loglik
-    results$times <- bpf_out@times  # Saved for convenience
-    results
-  },
-  seed = 432235
-)
-@
-
-In this section, we provide more detail on the process that was used to estimate the coefficients used in Model~3.
-In particular, we discuss why we decided to include additional model parameters---those that are associated with the behavior of the system during Hurricane Matthew---that were not considered by \cite{lee20}.
-To calibrate this model, we used the iterated block particle filter (IBPF) method of \cite{ionides22}.
-Due to the novelty of this algorithm, there does not exist published examples the IBPF algorithm used for data analysis outside of the papers in which the algorithm was introduced \citep{ning21ibpf,ionides22}, which at least partially motivates the inclusion additional details in this section.
-
-\cite{lee20} were only able to estimate model parameters to a simplified version of Model~3 on a subset of the available data, as no method existed at the time of their publication to fit a fully coupled meta-population model to disease incidence data.
-Building on their results, we fit the fully coupled version of Model~3 to nearly all available data, reserving only a few observations to use to calibrate the initial conditions of the model (see the previous section for more details).
-In order to properly maximize the model likelihood, we found that it was necessary to use multiple searches for the MLE, periodically pruning away less successful searches.
-The first collection of searches was performed by obtaining initial values for the parameters by uniformly sampling values from a predefined hypercube.
-A subsequent refinement search used starting values corresponding to the largest model likelihoods obtained in the previous round.
-This is similar to the approach used by \cite{ionides22}.
-
-We use the technique described above to fit the fully coupled version of Model~3 proposed by \cite{lee20}.
-The maximum likelihood we obtained with the model was $\Sexpr{round(max(haiti3_sub_fit$search2$logLiks$logLik))}$, which is higher than the benchmark model ($\Sexpr{round(sum(unitLikes_bench$ll))}$).
-While beating a simple associative benchmark is promising, this does not immediately rule out the possibility that the model is a poor description of the system.
-Additional investigation of parameters estimates and their corresponding implications on model based conclusions should always be conducted.
-For meta-population models, it is worth considering how well the model fits the data to each spatial unit.
-This can be done by looking at conditional log-likelihoods, which is part of the output of the \texttt{bpfilter} algorithm in the \texttt{spatPomp} package.
-The likelihoods for each department, compared to the corresponding benchmark model, are displayed in Fig.~\ref{fig:h3UnitLikes}.
-The figure demonstrates that while the log-likelihood of the fitted model outperforms the auto-regressive negative binomial benchmark model, Model~3 has lower likelihoods than the benchmark for some departments.
-
-\begin{figure}[ht]
-<<h3UnitLikes, fig.height=3, fig.width=5.5, echo=FALSE>>=
-ggplot(bench_M3_joined, aes(x = ll, y = reorder(unit, ll))) +
-  geom_point(aes(col = type)) +
-  ylab("Unit") +
-  xlab("Log Likelihood") +
-  scale_color_manual(
-    labels = c("ll" = "Benchmark", "ll_est" = "Model 3"),
-    values = c("ll" = "#1b9e77", "ll_est" = "#d95f02")
-  ) +
-  theme(legend.title = element_blank(),
-        axis.title.y = element_blank())
-@
-\caption{\label{fig:h3UnitLikes}Log-likelihoods of Model~3 for each department compared to the corresponding benchmark model prior to the inclusion of parameters for initial values and Hurricane Matthew.}
-\end{figure}
-
-In addition to considering the conditional log-likelihoods of each unit, one can consider conditional log-likelihoods of each observation.
-When compared to a benchmark, this level of detail can provide useful information about which observations are well described by the model and which are poorly described.
-In Fig.~\ref{fig:condLL}, we plot the conditional log-likelihoods of each observation.
-This figure reveals that the fitted model has difficult time explaining certain features of the data.
-For example, many departments (in particular Sud) have observations with lower conditional log-likelihoods near October 2016 than at other time points.
-Further investigation of the model output reveals that the model is struggling to explain the sudden surge in cholera cases that occurred at this time, which coincides with the time that Hurricane Matthew struck Haiti.
-While the model does include a mechanism to account for increased cholera transmission due to large rainfall events, the mechanism does not appear to be sufficient to capture the damaging effects of the hurricane, which had the greatest impact in the the Sud and Grand'Anse departments \citep{ferreirai16}.
-This result led us to include parameters $\Whur{u}$ and $\hHur{u}$ in Eq.~23 of the main text, which allows for an increase in the transmission rate between environmental cholera and humans for Sud and Grand'Anse during and after the hurricane.
-The effect of the hurricane on cholera transmission is assumed to have an exponential decay, where the magnitude is determined by $\Whur{u}$ and the duration of the effect determined by $\hHur{u}$.
-
-\begin{figure}[ht]
-<<CondLLFig, echo=FALSE, message=FALSE, fig.height=3>>=
-
-# Changing labels to human-readable labels.
-deps[grepl("_", deps)] <- gsub("_", "-", deps[grepl("_", deps)])
-deps[deps == "Grande-Anse"] <- "Grand\'Anse"
-names(deps) <- as.character(1:10)
-myLabeller <- as_labeller(deps)
-
-# Plot conditional log-likelihoods for each unit
-t(bpf_sub3_out$block.cond.loglik) %>%
-  as.data.frame() %>%
-  mutate(t = bpf_sub3_out$times) %>%
-  pivot_longer(
-    cols = -t,
-    values_to = 'likelihood',
-    names_prefix = "V",
-    names_to = 'Unit'
-  ) %>%
-  mutate(date = as.Date(lubridate::date_decimal(t))) %>%
-  ggplot(aes(x = date, y = likelihood)) +
-  geom_line() +
-  geom_point() +
-  facet_wrap(~Unit, labeller = myLabeller, scales = 'free_y',
-             nrow = 2) +
-  theme_bw() +
-  theme(axis.title.x = element_blank(),
-        axis.text.x = element_text(angle = -45, hjust = 0)) +
-  ylab("Conditional log-likelihoods")
-@
-\caption{\label{fig:condLL}Conditional log-likelihoods of Model~3 prior to the inclusion of parameters for initial values and Hurricane Matthew.}
-\end{figure}
-
-Other points of concern include the observations at or near the start of the recorded cases, for example, in Nord-Ouest and Sud-Est.
-These low conditional log-likelihoods are a result of having zero or near zero confirmed cases in the first week of observed data, followed by a rapid growth of cases.
-The model cannot explain the rapid growth of cholera in the absence of starting cases because the initialization scheme forces departments with no recorded cases in the first week to have hidden states $I_{u0}(0) = A_{u0}(0) = B_{u}(0) = 0$.
-In reality, however, it is possible to have some infected or asymptomatic individuals in a department when no cases are recorded due to under-reporting.
-This may be particularly true at the onset of an infectious disease outbreak, as lower public awareness and inadequate testing infrastructure may result in many cases going undetected.
-We therefore include and estimate $I_{0u}$ that determines the initial number of Infected individuals at the start of the pandemic for the departments Grande'Anse, Nord-Ouest, and Sud-Est;
-each of these departments have both low starting case counts followed by rapid growth, and low initial values of the conditinoal log-likelihood.
-
-<<loadFinalModel3, echo=FALSE, include=FALSE, message=FALSE>>=
-
-# This result was used in the manuscript, so it doesn't need to be recalculated.
-h3_final_results <- readRDS("../model3/run_level_3/haiti3_fit.rds")
-
-# Save best_m as index for the model parameters with the best likelihood.
-# Then, save the model parameter values and the likelihood evaluations
-best_m <- which.max(h3_final_results$search2$logLiks$logLik)
-best_params <- h3_final_results$search2$params[best_m, ]
-all_liks <- as.data.frame(h3_final_results$search2$likMat)
-all_liks$which <- rep(1:72, each = 18)
-best_mod <- dplyr::filter(all_liks, which == best_m)
-
-# Get repeated estimates of unit log-likelihoods
-unitLikesM3 <- best_mod %>%
-  pivot_longer(
-    cols = -which,
-    names_to = "unit",
-    names_prefix = "V",
-    values_to = "logLik"
-  ) %>%
-  select(-which) %>%
-  group_by(unit) %>%
-  summarize(
-    ll_est = pomp::logmeanexp(logLik),
-    ll_se = pomp::logmeanexp(logLik, se = TRUE)[2]
-  ) %>%
-  ungroup() %>%
-  mutate(unit = as.numeric(unit)) %>%
-  arrange(unit)
-
-# Combine model likelihood table to benchmarks
-bench_M3_joined <- dplyr::left_join(
-  x = unitLikesM3,
-  y = unitLikes_bench,
-  by = "unit"
-) %>%
-  select(-ll_se) %>%
-  mutate(unit = as.character(unit)) %>%
-  pivot_longer(
-    cols = -unit,
-    names_to = "type",
-    values_to = "ll"
-  ) %>%
-  mutate(unit = paste0(unit, ": ", deps[as.numeric(unit)]))
-@
-
-We refit Model~3 after introducing these seven parameters---three initial value parameters and four parameters related to Hurricane Matthew---using the same successive search scheme described above.
-The resulting model has a log-likelihood value of $\Sexpr{round(max(h3_final_results$search2$logLiks$logLik), 1)}$.
-We also provide figures for department specific log-likelihoods and conditional log-likelihoods in Figs.~\ref{fig:finalUnitLL} and \ref{fig:finalCondLL}.
-
-\begin{figure}[ht]
-<<finalUnitLLFig, fig.height=3, fig.width=5.5, echo=FALSE>>=
-ggplot(bench_M3_joined, aes(x = ll, y = reorder(unit, ll))) +
-  geom_point(aes(col = type)) +
-  ylab("Unit") +
-  xlab("Log Likelihood") +
-  scale_color_manual(
-    labels = c("ll" = "Benchmark", "ll_est" = "Model 3"),
-    values = c("ll" = "#1b9e77", "ll_est" = "#d95f02")
-  ) +
-  theme(legend.title = element_blank(),
-        axis.title.y = element_blank())
-@
-\caption{\label{fig:finalUnitLL}Log-likelihoods of Model~3 for each department compared to the corresponding benchmark model after adding and estimating parameters related to initial values and Hurricane Matthew.}
-\end{figure}
-
-<<finalModel3BPF, echo=FALSE, message=FALSE, include=FALSE>>=
-
-# Set the model parameters to the best results from previous search
-coef(h3)[names(best_params)] <- best_params
-
-# block particle filter to get conditional log-likelihoods.
-bpf_final3_out <- bake(
-  file = "../model3/si/h3_final_bpf_cond_ll.rds", {
-
-    results <- list()
-
-    bpf_out <- h3 %>% bpfilter(
-      Np = 5000,
-      block_size = 1,
-      save_states = TRUE
-    )
-
-    results$Sus_df <- foreach(i=1:length(h3@times), .combine = bind_rows) %do% {
-      get_h3_S_quants(bpf_out, i)
-    }
-
-    results$block.cond.loglik <- bpf_out@block.cond.loglik
-    results$times <- bpf_out@times  # saved for convenience
-    results
-  },
-  seed = 786321
-)
-@
-
-\begin{figure}[ht]
-<<finalCondLLFig, echo=FALSE, message=FALSE, fig.height=3>>=
-t(bpf_final3_out$block.cond.loglik) %>%
-  as.data.frame() %>%
-  mutate(t = bpf_final3_out$times) %>%
-  pivot_longer(
-    cols = -t,
-    values_to = 'likelihood',
-    names_prefix = "V",
-    names_to = 'Unit'
-  ) %>%
-  mutate(date = as.Date(lubridate::date_decimal(t))) %>%
-  ggplot(aes(x = date, y = likelihood)) +
-  geom_line() +
-  geom_point() +
-  facet_wrap(~Unit, labeller = myLabeller, scales = 'free_y',
-             nrow = 2) +
-  theme_bw() +
-  theme(axis.title.x = element_blank(),
-        axis.text.x = element_text(angle = -45, hjust = 0)) +
-  ylab("Conditional log-likelihoods")
-@
-\caption{\label{fig:finalCondLL}Conditional log-likelihoods of Model~3 after adding and estimating the parameters for initial values and Hurricane Matthew.}
-\end{figure}
-
-While the addition of the Hurricane parameters seems to have resulted in a dramatic increase of conditional likelihoods around and after October 2016, the inclusion of the initial value parameters does not appear to have fixed the low conditional log-likelihood values in Sud-Est and Nord-Ouest.
-This could be due to many factors, including the possibility that the coupling between units forces low initial value parameters in a given department in order for the model to fit the data in the remaining departments.
-Whatever the case may be, the inclusion of the seven parameters resulted in a $\Sexpr{round(max(h3_final_results$search2$logLiks$logLik) - max(haiti3_sub_fit$search2$logLiks$logLik), 1)}$ increase in log-likelihood, which is a highly statistically significant difference.
-One could similarly test whether the inclusion of the initial value parameters resulted in a significantly better fit to the observed data, but we do not do this here due to the large computational effort that would be required.
-
-\subsection{Examining the Hidden States of the Calibrated Model}
-
-For mechanistic models, beating a suitable statistical benchmark does not alone guarantee that the model provides an accurate description of a dynamic process.
-Indeed, a good statistical fit does not require the model to assert a causal explanation.
-For example, reconstructed latent variables should make sense in the context of alternative measurements of these variables \citep{grad12}.
-We demonstrate this principle by examining the latent state of the calibrated model.
-In particular, we examine the compartment of susceptible individuals under various scenarios, as this compartment drives transmission dynamics in SIR models for infectious disease outbreaks.
-
-Recall that the filtering distribution for the calibrated version of Model~3 at time $t_k$ is defined as the distribution of the latent state at time $t_k$ given the data from times $t_{1}:t_{k}$, i.e.
-$f^{(3)}_{\bm{X}_k|\bm{Y}_{1:k}}(\bm{x}_{k} | \bm{y}^*_{1:k} ; \hat\theta)$.
-In general, one may expect simulations from the filtering distribution of a model with a good statistical fit to result in hidden states that are highly consistent with the observed data because the filtering distribution is conditioned on the observed data.
-Fig.~\ref{fig:h3Sus} shows the percentage of individuals that are in the susceptible compartment from various simulations of the model:
-simulations from Model~3 under initial conditions are displayed in red; simulations from the filtering distribution of model are displayed in blue.
-This figure shows that simulations from initial conditions tends to result in a much more rapid depletion of susceptible individuals at the start of the epidemic than simulations from the filtering distribution, suggesting the calibrated model has a propensity to predict larger outbreaks than what is typically seen in the data.
-This also appears true in later stages of the time period of observed data, where the data suggests that there is a replenishment of the susceptible compartment, but the Model simulations typically retain a smaller proportion of susceptible individuals.
-This result demonstrates that the calibrated model favors a more rapid growth in cholera cases than what is typically seen in the observed data, which explains why the model fails to predict the absence of cases between February, 2019 and October, 2022.
-
-<<h3SusFigSetup, echo=FALSE, message=FALSE, include=FALSE>>=
-
-# This bake perfoms the following:
-#  (1) simulates the model from initial conditions
-#  (2) get the values of the hidden states at the end of each simulation, and
-#      saves these for later use.
-#  (3) Obtain quantiles for the Susceptible pool from the simulations.
-init_conditions_results <- bake(
-  file = "../model3/si/InitSimsQuants.rds", {
-
-    results <- list()  # To return final output of bake
-
-    # Simulate from the fitted model
-    h3Spat_sims <- simulate(
-      h3, nsim = 2000, format = 'data.frame'
-    )
-
-    # Get states at the end of the simulations
-    h3_spat_end <- h3Spat_sims %>%
-      filter(time == max(time))
-
-    # Save hidden state values at the end of the simulation. Saved as a matrix
-    # in the same format as the output from a block-particle filter, so that
-    # it can be used as the input to project_from_filter.
-    results[["end_conditions"]] <- h3_spat_end %>%
-      select(-time, -cases) %>%
-      mutate(
-        unit = case_when(
-          unitname == "Artibonite" ~ 1,
-          unitname == "Centre" ~ 2,
-          unitname == "Grande_Anse" ~ 3,
-          unitname == "Nippes" ~ 4,
-          unitname == "Nord" ~ 5,
-          unitname == "Nord_Est" ~ 6,
-          unitname == "Nord_Ouest" ~ 7,
-          unitname == "Ouest" ~ 8,
-          unitname == "Sud" ~ 9,
-          unitname == "Sud_Est" ~ 10
-        )
-      ) %>%
-      select(-unitname) %>%
-      pivot_wider(
-        data = .,
-        names_from = unit,
-        values_from = c(S, I, A, R_one, R_two, R_three, VSd, VR1d, VR2d,
-                        VR3d, VSdd, VR1dd, VR2dd, VR3dd, VSd_alt, VR1d_alt,
-                        VR2d_alt, VR3d_alt, VSdd_alt, VR1dd_alt, VR2dd_alt,
-                        VR3dd_alt, C, B, Doses, totInc),
-        names_sep = ""
-      ) %>%
-      select(-.id) %>%
-      as.matrix() %>%
-      t()
-
-    # Get the quantiles for the Susceptible pool from the simulations
-    results[["SQuants"]] <- h3Spat_sims %>%
-      rename(dep = unitname) %>%
-      group_by(dep, time) %>%
-      summarise(
-        Q025 = quantile(S + VSd + VSdd + VSd_alt + VSdd_alt, probs = 0.025),
-        Q50  = quantile(S + VSd + VSdd + VSd_alt + VSdd_alt, probs = 0.5),
-        Q975 = quantile(S + VSd + VSdd + VSd_alt + VSdd_alt, probs = 0.975)
-      ) %>%
-      ungroup() %>%
-      rowwise() %>%
-      mutate(
-        unit_num = which(h3@unit_names == dep),
-        date = lubridate::date_decimal(time),
-        Q025 = Q025 / h3@params[paste0("H", unit_num)],
-        Q50  = Q50  / h3@params[paste0("H", unit_num)],
-        Q975 = Q975 / h3@params[paste0("H", unit_num)]
-      ) %>%
-      filter(
-        date >= as.Date(
-          lubridate::round_date(lubridate::date_decimal(h3@t0), unit = 'day')
-        )
-      ) %>%
-      mutate(date = as.Date(lubridate::round_date(date, unit = 'day')))
-
-    results
-  },
-  seed = 834687
-)
-
-# Get the output of the "bake" call for plotting.
-h3Spat_quants <- init_conditions_results[["SQuants"]]
-
-# This bake call uses the ending hidden states from above to forecast cholera
-# cases in the future under the V0 vaccination scenario
-mod3_V0_initcond_res <- bake(
-  file = '../model3/si/vacc_sim_from_init.rds', {
-    h3_V0_res <- list()
-
-    # Get "noVacc" scenario parameters
-    h3_S0_par <- get_model3_vacc_scenario_params(scenario = "noVacc")
-    h3Spat <- h3
-    p3 <- best_params
-    coef(h3Spat) <- p3
-    coef(h3Spat)[names(h3_S0_par)] <- h3_S0_par
-
-    # store the end states that will be used as initial values
-    h3_bpf_end_states <- init_conditions_results$end_conditions
-
-    # Using parameters and end state values, obtain projections
-    h3_V0_sims <- project_from_filter(
-      h3Spat, end_states = h3_bpf_end_states, covarGen = project_rain,
-      nsims = 1000, seed = 184697
-    )
-
-    # Store nationally aggregated results
-    h3_V0_res$mod3_V0_sims <- agg_mod3_sims(h3_V0_sims)
-    h3_V0_res$mod3_V0_probs <- get_elimProbs(h3_V0_sims, model = 3)
-
-    # Get quantiles of observed cases for each department and time
-    h3_V0_res$UnitQuants <- h3_V0_sims %>%
-      mutate(
-        date = as.Date(  # Get date object from decimal time
-          lubridate::round_date(lubridate::date_decimal(time), unit = 'day')
-        )
-      ) %>%
-      group_by(unitname, date) %>%
-      summarize(
-        Q025 = quantile(cases, probs = 0.025),
-        Q50 = quantile(cases, probs = 0.5),
-        Q975 = quantile(cases, probs = 0.975)
-      ) %>%
-      ungroup() %>%
-      rename(dep = unitname)
-
-    # Get quantiles of susceptible individuals for each department and time
-    h3_V0_res$SQuants <- h3_V0_sims %>%
-      mutate(
-        date = as.Date(
-          lubridate::round_date(lubridate::date_decimal(time), unit = 'day')
-        )
-      ) %>%
-      group_by(unitname, date) %>%
-      mutate(
-        # Get the population for each unit
-        pop = h3@params[paste0("H", which(first(unitname) == unit_names(h3)))]
-      ) %>%
-      summarize(  # Calculate the quantiles
-        Q025 = quantile(
-          (S + VSd + VSdd + VSdd_alt + VSd_alt) / pop,
-          probs = 0.025
-        ),
-        Q50 = quantile(
-          (S + VSd + VSdd + VSdd_alt + VSd_alt) / pop,
-          probs = 0.5
-        ),
-        Q975 = quantile(
-          (S + VSd + VSdd + VSdd_alt + VSd_alt) / pop,
-          probs = 0.975
-        )
-      ) %>%
-      ungroup() %>%
-      rename(dep = unitname)
-
-    h3_V0_res
-  }
-)
-
-# This result was already used in the manuscript; it doesn't need to
-# be recalculated.
-mod3_V0_res <- readRDS("../model3/run_level_3/mod3_V0_res.rds")
-@
-
-\begin{figure}[ht]
-<<h3SusFig, echo=FALSE, message=FALSE, fig.height=3>>=
-ggplot() +
-  # Sims from initial condition
-  geom_line(data = h3Spat_quants, aes(x = date, y = Q50), col = 'red') +
-  geom_ribbon(
-    data = h3Spat_quants,
-    aes(x = date, ymin = Q025, ymax = Q975),
-    fill = 'red', alpha = 0.5
-  ) +
-  # projections from initial condition, using filtered sims function
-  geom_line(data = mod3_V0_initcond_res$SQuants, aes(x = date, y = Q50), col = 'red') +
-  geom_ribbon(
-    data = mod3_V0_initcond_res$SQuants,
-    aes(x = date, ymin = Q025, ymax = Q975),
-    fill = 'red', alpha = 0.5
-  ) +
-  # Sims from filtered distribution
-  geom_line(
-    data = bpf_final3_out$Sus_df,
-    aes(x = as.Date(lubridate::date_decimal(time)), y = Q50),
-    col = 'blue'
-  ) +
-  geom_ribbon(
-    data = bpf_final3_out$Sus_df,
-    aes(x = as.Date(lubridate::date_decimal(time)), ymin = Q025, ymax = Q975),
-    fill = 'blue', alpha = 0.5
-  ) +
-  # Projections from filtered distribution
-  geom_line(data = mod3_V0_res$SQuants, aes(x = date, y = Q50), col = 'blue') +
-  geom_ribbon(
-    data = mod3_V0_res$SQuants,
-    aes(x = date, ymin = Q025, ymax = Q975),
-    fill = 'blue', alpha = 0.5
-  ) +
-  geom_vline(xintercept = max(h3Spat_quants$date), linetype = 'dashed') +
-  facet_wrap(~dep, scales = 'free_y', nrow = 2) +
-  theme_bw() +
-  theme(axis.title.x = element_blank(),
-        axis.text.x = element_text(angle = -45, hjust = 0)) +
-  ylab("Total Susceptible (%)")
-@
-\caption{\label{fig:h3Sus}Percentage of individuals that are in the susceptible compartment.
-Simulations from Model~3 under initial conditions are displayed in red; simulations from the filtering distribution of model are displayed in blue.
-The dashed line represents the end of the observed data.}
-\end{figure}
+\input{calibrateMod3Out}
 
@@
 
-\section{Replication of \cite{lee20}}
+\input{ReplicateLee20Out}
 
-In this article we claimed that we were able to obtain better fits to the observed data using the same models that were proposed by \cite{lee20}.
-Along with visual comparisons to the data, this claim was supported by comparing likelihoods and AIC values in Table~2 in the manuscript. % TODO: Double check the table number matches
-Because model likelihoods were not provided by \cite{lee20}, it is necessary to replicate these models in order to obtain likelihood estimates.
-Here we would like to thank the authors of \cite{lee20}, who provided detailed descriptions of their models, which enabled us to build on their work.
-In the following subsections, we use our \code{R} package \code{haitipkg} to reproduce some of the results of \cite{lee20}.
-This reproduction allows us to estimate the likelihoods of the \cite{lee20} version of Models~1--3, and also provides a demonstration of the importance and usefulness of reproducible research.
-
-\subsection{Model~1 Replication}
-
-The model was implemented by a team at Johns Hopkins Bloomberg School of Public Health (hereafter referred to as the Model~1 authors) in the \code{R} programming language using the \code{pomp} package \citep{king16}.
-Original source code is publicly available with DOI: 10.5281/zenodo.3360991.
-The final results reported by the Model~1 authors were obtained by using several different parameter sets rather than a single point estimate.
-According to the supplement materials, this was because model realizations from a single parameter set retained substantial variability, but multiple realizations from a collection of parameter sets resulted in a reasonable visual fit to the data.
-We are also inclined to believe that the use of multiple parameter values was in part intended to account for parameter uncertainty---the importance of which was discussed in the main text---an effort by the Model~1 authors that we applaud.
-Simulations from each of the parameter sets, however, were treated with equal importance when being used to diagnose the model fit and make inference on the system.
-This is problematic given Figures S8 and S9 of the supplement material, which suggest that some parameter sets that were used for inference may have been several hundred units of log-likelihood lower than other parameter sets that were simultaneously used to make forecasts.
-Such a large difference in log-likelihoods is well beyond the threshold of statistical uncertainty determined by Wilks' theorem, resulting in the equal use of statistically inferior parameter sets in order to make forecasts and conduct inference on the system.
-
-To fully reproduce the results of the Model~1 authors, it is necessary to use the exact same set of model parameters that were originally used to obtain the results presented by \cite{lee20}.
-Because these parameter sets were not made publicly available, we relied on the source code provided by the Model~1 authors to approximately recreate the parameter set.
-Due to software updates since the publication of the source code, we were unable to produce the exact same set of parameters.
-Running the publicly available source code, however, resulted in a set of parameters that are visually similar to those used by the Model~1 authors (See Figures~\ref{fig:PlotEpiDist} and \ref{fig:plotEndParams}).
-Furthermore, simulations using the set of parameters produced by the source code appear practically equivalent to those displayed by \cite{lee20} (See Figure~\ref{fig:plotMod1Sims}).
-
-<<Simulate Lee Model 1, echo=FALSE, cache=TRUE>>=
-h1_epi <- haiti1()
-N_LEE_SIMS <- 20
-
-registerDoRNG(18599687)
-foreach(
-  i = 1:length(h1LeeStartsEpi),
-  .combine = rbind
-) %dopar% {
-  guess <- h1LeeStartsEpi[i, ]
-  sims <- simulate(h1_epi, nsim = N_LEE_SIMS, format = 'data.frame',
-                   params = guess)  # 20 sims for each set of parameters
-  sims$param_set <- i
-  sims %>%
-    dplyr::select(.id, week, cases, param_set)
-} -> all_sims
-@
-
-<<PlotEpiDist, echo=FALSE, fig.cap="Compare to Figure S8 of Lee et al. (2020) supplement.", fig.width=7, fig.height=7, fig.align='center'>>=
-GGally::ggpairs(
-  h1LeeStartsEpi,
-  columns = c("rho", "tau", "beta1", "nu", "loglik")
-)
-@
-
-<<SimulateEpi, echo=FALSE, include=FALSE, message=FALSE, cache=TRUE>>=
-h1_end <- haiti1(period = "endemic")
-
-true_agg_cases <- haitiCholera %>%
-  select(-report) %>%
-  pivot_longer(
-    data = .,
-    cols = -date_saturday,
-    values_to = 'cases',
-    names_to = 'dep'
-  ) %>%
-  mutate(
-    date = as.Date(date_saturday),
-    dep = gsub("\\.", "_", dep)
-  ) %>%
-  mutate(
-    dep = case_when(dep == "Grand_Anse" ~ "Grande_Anse", TRUE ~ dep)
-  ) %>%
-  tidyr::pivot_wider(
-    data = .,
-    id_cols = c(date),
-    names_from = dep,
-    values_from = cases,
-    names_prefix = 'cases_'
-  ) %>%
-  mutate(
-    ReportedAll = cases_Artibonite + cases_Centre +
-      cases_Grande_Anse + cases_Nippes + cases_Nord +
-      cases_Nord_Est + cases_Ouest + cases_Sud +
-      cases_Sud_Est + cases_Nord_Ouest
-  )
-@
-
-<<FitEnd, echo=FALSE, message=FALSE, include=FALSE>>=
-foreach(
-  i = 1:nrow(h1LeeStartsEnd),
-  .combine = rbind
-) %dopar% {
-  r_params <- h1LeeStartsEnd[i, ][names(coef(h1_end))]
-  coef(h1_end) <- r_params
-
-  sims <- simulate(
-    h1_end, nsim = N_LEE_SIMS,
-    format = 'data.frame',
-    params = r_params
-  )
-  sims$param_set <- i
-  sims %>%
-    dplyr::select(.id, week, cases, param_set)
-} -> end_sims
-
-@
-
-<<plotEndParams, fig.width=7, fig.height=7, fig.align='center', fig.cap="Bivariate relationships between variables after fitting endemic period.
-Compare to S9 of Lee et al.
-(2020) supplement.", cache=TRUE, echo=FALSE>>=
-GGally::ggpairs(
-  h1LeeStartsEnd,
-  columns = c("rho", "tau", "beta1", "nu", "loglik")
-)
-@
-
-<<calcMod1LeeLikes, echo=FALSE, include=FALSE, message=FALSE>>=
-h1_epi_evals <- readRDS("../model1/run_level_3/lee1_epi_evals.rds")
-h1_end_evals <- readRDS("../model1/run_level_3/lee1_end_evals.rds")
-
-mod1_all_likes <- dplyr::inner_join(
-  x = h1_epi_evals,
-  y = h1_end_evals,
-  by = "parid"
-  ) %>%
-  dplyr::mutate(
-    joint_ll = pfLL.x + pfLL.y
-  )
-
-# Approximate the likelihood of the Lee et al (2020) version of Model 1 by
-# using the parameter sets that resulted in the maximum likelihood.
-mod1_lee_ll <- max(mod1_all_likes$joint_ll, na.rm = TRUE)
-@
-
-<<plotMod1Sims, echo=FALSE, fig.cap=paste0("Simulations using parameter sets that were generated by running source code provided by Lee et al (2020).
-Compare to Figure S7 of \\cite{lee20sup}.
-The upper bound for the likelihood of this model is ", round(mod1_lee_ll), "."), cache=TRUE, fig.width=6.5, fig.height=3.75>>=
-
-quants <- all_sims %>%
-  filter(!param_set %in% c(1, 2)) %>%  # Remove parameter sets inconsistent with Lee et al (2020).
-  group_by(week) %>%
-  summarize(
-    q025 = quantile(cases, probs = 0.025, na.rm = TRUE),
-    q50  = quantile(cases, probs = 0.500, na.rm = TRUE),
-    q975 = quantile(cases, probs = 0.975, na.rm = TRUE)
-  ) %>%
-  ungroup() %>%
-  mutate(date = lubridate::ymd("2010-10-16") + lubridate::weeks(week))
-
-quants_end <- end_sims %>%
-  group_by(week) %>%
-  summarize(
-    q025 = quantile(cases, probs = 0.025, na.rm = TRUE),
-    q50  = quantile(cases, probs = 0.500, na.rm = TRUE),
-    q975 = quantile(cases, probs = 0.975, na.rm = TRUE)
-  ) %>%
-  ungroup() %>%
-  mutate(date = lubridate::ymd("2010-10-16") + lubridate::weeks(week))
-
-ggplot() +
-  geom_point(data = true_agg_cases, aes(x = date, y = ReportedAll)) +
-  geom_line(data = quants, aes(x = date, y = q50), col = '#0b8a2b', size = 1) +
-  geom_ribbon(
-    data = quants,
-    aes(x = date, ymin = q025, ymax = q975),
-    alpha = 0.5, fill = '#0b9c30'
-  ) +
-  geom_line(data = quants_end, aes(x = date, y = q50), col = '#882871') +
-  geom_ribbon(data = quants_end, aes(x = date, ymin = q025, ymax = q975), alpha = 0.5, fill = '#882871') +
-  ylab("Reported, Symptomatic Incidence") +
-  scale_x_date(date_breaks = "1 year") +
-  theme(axis.title.x = element_blank(),
-        axis.text.x = element_text(hjust = 1, angle = 45))
-@
-
-Because the model forecasts provided by \cite{lee20} come from various sets of parameters---which each correspond to a unique log-likelihood value---it is not obvious how one would obtain an estimate for the log-likelihood of the model that was used for simulations by the Model~1 authors.
-One approach could be to calculate the logarithm of the weighted mean of the likelihoods for each parameter sets used to obtain the forecasts, where the weights are proportional to the number of times the parameter set was used.
-However, in an effort to not underestimate the likelihood of the model of the Model~1 authors, we report the estimated log-likelihood as the log-likelihood value corresponding to the parameter set with the largest likelihood value, even though the majority of simulations were obtained using parameter sets with lower likelihood values.
-In this sense, we consider the log-likelihood reported in Table~1 of the main text to be an upper-bound of the log-likelihood of the model used by \cite{lee20}.
-For each parameter set, the log-likelihood was estimated using a particle filter, implemented as the \texttt{pfilter} function in the \texttt{pomp} package.
-
-\subsection{Model~2 Replication}\label{sec:mod2rep}
-
-Model~2 was developed by a team that consisted of members from the Fred Hutchinson Cancer Research Center and the University of Florida (hereafter referred to as the Model~2 authors).
-While Model~2 is the only deterministic model we considered in our analysis, it contains perhaps the most complex descriptions of cholera in Haiti: Model~2 accounts for movement between spatial units; human-to-human and environment-to-human cholera infections; and transfer of water between spatial units based on elevation charts and river flows.
-
-The source code that the Model~2 authors used to generate their results was written in the \code{Python} programming language, and is publicly available at \url{10.5281/zenodo.3360857} and its accompanying GitHub repository \url{https://github.com/lulelita/HaitiCholeraMultiModelingProject}.
-In order to perform our analysis in a unified framework, we re-implemented this model in the \code{R} programming language using the \code{spatPomp} package \citep{asfaw23arxiv}, which facilitates the creation of meta-population models.
-We note that the travel and water matrices used to implement the complex dynamics in Model~2 \citep{lee20sup} are not available in either the Zenodo archive or the GitHub repository;
-instead, we obtained these matrices via personal correspondence with the Model~2 authors.
-Using these matrices, and the point estimates for model parameters provided by \citep{lee20sup}, we created trajectories of the cholera dynamics and compared this to available data.
-These trajectories, shown in Figure~\ref{fig:mod2rep}, are very similar to the trajectories shown in Figure~S15 of \cite{lee20sup}.
-
-<<Model2Replication, cache=TRUE, echo=FALSE>>=
-h2_epi <- haiti2()
-h2_end <- haiti2(region = 'after')
-
-h2_epi_traj <- trajectory(h2_epi, format = 'data.frame')
-h2_end_traj <- trajectory(h2_end, format = 'data.frame')
-
-h2_epi_traj$Ctotal <- rowSums(h2_epi_traj[, paste0("C", 1:10)]) * 0.2
-h2_end_traj$Ctotal <- rowSums(h2_end_traj[, paste0("C", 1:10)]) * 0.2
-
-dep_plot_df <- haitipkg::haitiCholera %>%
-  select(-report) %>%
-  tidyr::pivot_longer(
-    data = .,
-    cols = -date_saturday,
-    values_to = 'cases',
-    names_to = 'dep'
-  ) %>%
-  mutate(
-    date = as.Date(date_saturday),
-    dep = gsub("\\.", "_", dep)
-  ) %>%
-  mutate(
-    dep = case_when(dep == "Grand_Anse" ~ "Grande_Anse", TRUE ~ dep)
-  )
-
-true_agg_cases <- dep_plot_df %>%
-  tidyr::pivot_wider(
-    data = .,
-    id_cols = c(date),
-    names_from = dep,
-    values_from = cases,
-    names_prefix = 'cases_'
-  ) %>%
-  mutate(
-    ReportedAll = cases_Artibonite + cases_Centre +
-      cases_Grande_Anse + cases_Nippes + cases_Nord +
-      cases_Nord_Est + cases_Ouest + cases_Sud +
-      cases_Sud_Est + cases_Nord_Ouest
-  ) %>%
-  select(date, ReportedAll)
-
-h2_epi_traj$date <- as.Date(
-  lubridate::round_date(
-    lubridate::date_decimal(h2_epi_traj$year),
-    unit = 'day'
-  )
-)
-
-h2_end_traj$date <- as.Date(
-  lubridate::round_date(
-    lubridate::date_decimal(h2_end_traj$year),
-    unit = 'day'
-  )
-)
-
-all_traj <- bind_rows(
-  select(h2_epi_traj, date, Ctotal),
-  select(h2_end_traj, date, Ctotal)
-)
-
-projections <- left_join(
-  x = true_agg_cases,
-  y = all_traj,
-  by = 'date'
-)
-@
-
-\begin{figure}[ht]
-<<plotModel2Rep, cache=TRUE, echo=FALSE, fig.width=6.5, fig.height=4.5>>=
-gg_epi <- projections %>%
-  filter(date < as.Date("03-01-2014", format = '%m-%d-%Y')) %>%
-  ggplot(aes(x = date)) +
-  geom_line(aes(y = ReportedAll), col = 'black', linetype = 'dashed') +
-  geom_line(aes(y = Ctotal), col = 'red', size = 1) +
-  theme_bw() +
-  scale_y_continuous(limits = c(0, 30000), breaks = seq(0, 30000, 5000)) +
-  theme(axis.title.x = element_blank()) +
-  ylab("Reported number of cases")
-
-gg_end <- projections %>%
-  filter(date >= as.Date("03-01-2014", format = '%m-%d-%Y')) %>%
-  ggplot(aes(x = date)) +
-  geom_line(aes(y = ReportedAll), col = 'black', linetype = 'dashed') +
-  geom_line(aes(y = Ctotal), col = 'red', size = 1) +
-  theme_bw() +
-  scale_y_continuous(limits = c(0, 7000), breaks = seq(0, 7000, 1000)) +
-  theme(axis.title.x = element_blank()) +
-  ylab("Reported number of cases") +
-  scale_x_date(date_breaks = "1 year", date_labels = "%Y")
-
-cowplot::plot_grid(gg_epi, gg_end, ncol = 1)
-@
-\caption{\label{fig:mod2rep}
-Model 2 trajectories using the \code{haitipkg}. Compare to Figure S15 of \cite{lee20sup}.
-}
-\end{figure}
-
-There are minor differences between Figure~\ref{fig:mod2rep} and Figure~S15 of \cite{lee20sup}.
-While the discrepancy appears minor, the deterministic nature of Model~2 implies that an exact replication of model trajectories should be possible.
-In this case, these discrepancies may possibly be attributed to implementing the model and plotting the model trajectory in two different programming languages.
-Another potential explanation for the discrepancy is that the parameters that we used are only approximately the same as those used by \citet{lee20sup}.
-For example, the parameters $\transmission$, $\Wbeta{}$ (See Table~\ref{tab:translate}) had reported values of $9.9 \times 10^{-7}$ and $4.03 \times 10^{-2}$, respectively (Table~S13 of \cite{lee20sup}), but were actually fit to data and therefore likely these values have been rounded.
-Additionally, our implementation of Model~2 used a time scale of years and many of the parameters were reported on a weekly scale, so small differences may result due to unit conversions.
-The collective effect of these small differences in model parameters likely will result in small differences in model trajectories.
-
-Some additional concerns about being able to accurately replicate the results of \cite{lee20} are valid.
-Details about the measurement models and how latent states were initialized for the epidemic model were not provided by \citet{lee20sup} and therefore these details must be inferred by looking at the provided source code.
-According to repository comments, the files \code{fit\allowbreak In\allowbreak Pieces\allowbreak 3params\allowbreak Clean\allowbreak May2019\allowbreak Public.py} and \code{fit\allowbreak In\allowbreak Pieces\allowbreak Mu\allowbreak With\allowbreak Frac\allowbreak Sus\allowbreak Fixed\allowbreak All\allowbreak Infections\allowbreak Public.py} were used to fit the epidemic and endemic phases of the model respectively, although it is apparent that these exact files were not used to obtain the reported results since the files contain some variable-naming errors that make it impossible to run the files without making modifications \footnote{One example of why the code cannot be run that the file loads functions from a non-extant file named \code{choleraEqs.py} in line 13 rather than \code{cholera\allowbreak Eqs\allowbreak Public.py}.}.
-The inability to replicate the results by \citet{lee20} by running the provided source code makes checking whether or not a our numeric implementation faithfully represents their results very difficult.
-Additionally, the script that was said to been used to obtain the results reported by \cite{lee20} appears to use a different measurement model than what was described in the supplemental material, again making it difficult to fully replicate the result of \cite{lee20} without being able to easily run the provided source code.
-In this case, we chose to use measurement model described by Eq.~\ref{model2:measEpi} for both phases of the epidemic, as this seemed to visually match the results of \cite{lee20} most closely.
-
-\subsection{Model~3 Replication}
-
-Model~3 was developed by a team of researchers at the Laboratory of the Swiss Federal Institute of Technology in Lausanne, hereafter referred to as the Model~3 authors.
-The code that was originally used to implement Model~3 is archived with the DOI: \url{10.5281/zenodo.3360723}, and also available in the public GitHub repository: \code{jcblemai/haiti-mass-ocv-campaign}.
-Because the code was made publicly available, and final model parameters were reported in the supplementary material of \cite{lee20}, we were able to reproduce Model~3 by directly using the source code.
-In Fig.~\ref{fig:mod3rep}, we plot simulations from this model.
-This figure can be compared to Figure S18 of \cite{lee20}.
-We note that slight differences may be accounted for due to variance in the model simulations and the difference in programming language used to produce the figure.
-Overall, the high standard of reproducibility that was achieved by the Model~3 authors facilitated the ability to readily replicate their model and results.
-
-<<Model3Replication, echo=FALSE, cache=TRUE>>=
-NSIM <- 500
-
-plot_order <- c(
-  'Artibonite',
-  'Sud_Est',
-  'Nippes',
-  'Nord_Est',
-  'Ouest',
-  'Centre',
-  'Nord',
-  'Sud',
-  'Nord_Ouest',
-  'Grande_Anse'
-)
-
-dep_labeller <- as_labeller(
-  c(
-    'Artibonite' = 'Artibonite',
-    'Sud_Est' = 'Sud-Est',
-    'Sud.Est' = 'Sud-Est',
-    'Nippes' = 'Nippes',
-    'Nord_Est' = 'Nord-Est',
-    'Nord.Est' = 'Nord-Est',
-    'Ouest' = 'Ouest',
-    'Centre' = 'Centre',
-    'Nord' = 'Nord',
-    'Sud' = 'Sud',
-    'Nord_Ouest' = 'Nord-Ouest',
-    'Nord.Ouest' = 'Nord-Ouest',
-    'Grande_Anse' = 'Grand\'Anse',
-    'Grand.Anse' = 'Grand\'Anse'
-  )
-)
-
-lee3 <- lee3_spatPomp()
-lee3_sims_quants <- bake("../model3/si/lee3Sims.rds", {
-
-  # Simulate from lee et al 2020 model
-  sims_temp <- simulate(
-    lee3, nsim = NSIM,
-    seed = 321, format = 'data.frame'
-  )
-
-  # Get desired quantiles of reported cases from simulations
-  quants <- sims_temp %>%
-    dplyr::group_by(unitname, time) %>%
-    dplyr::summarise(
-      q05 = quantile(cases, 0.025, na.rm = T),
-      mean = mean(cases, na.rm = T),
-      q50 = quantile(cases, 0.5, na.rm = T),
-      q95 = quantile(cases, 0.975, na.rm = T)
-    ) %>%
-    dplyr::ungroup() %>%
-    dplyr::mutate(
-      date = lubridate::date_decimal(time),
-      date = as.Date(lubridate::round_date(date))
-    ) %>%
-    dplyr::rename(dep = unitname)
-
-  # Remove cases from Ouest that were not considered by Lee et al. (2020)
-  quants %>%
-    dplyr::filter(
-      !(dep == "Ouest" & date < '2017-06-10')
-    )
-
-})
-
-# Get the subset of cholera cases that match simulations from Lee et al. (2020) model.
-sub_cases <- haitiCholera %>%
-  dplyr::mutate(date = as.Date(date_saturday)) %>%
-  dplyr::select(-report, -date_saturday) %>%
-  dplyr::filter(lubridate::decimal_date(date) >= min(lee3@times)) %>%
-  tidyr::pivot_longer(
-    data = .,
-    cols = -date,
-    names_to = "dep",
-    values_to = "cases"
-  ) %>%
-  dplyr::mutate(
-    dep = dplyr::case_when(
-      dep == "Grand.Anse" ~ "Grande_Anse",
-      dep == "Nord.Est" ~ "Nord_Est",
-      dep == "Nord.Ouest" ~ "Nord_Ouest",
-      dep == "Sud.Est" ~ "Sud_Est",
-      TRUE ~ dep
-    )
-  )
-@
-
-\begin{figure}[ht]
-<<PlotMod3Rep, echo=FALSE, message=FALSE, fig.height=3.75>>=
-ggplot() +
-  # Region for upper and lower simulation bounds
-  geom_ribbon(
-    data = lee3_sims_quants,
-    aes(x = date, ymin = q05, ymax = q95),
-    fill = 'darkblue', alpha = 0.5
-  ) +
-  # Median line
-  geom_line(
-    data = lee3_sims_quants,
-    aes(x = date, y = q50),
-    col = 'darkblue'
-  ) +
-  # Observed data
-  geom_point(
-    data = dplyr::filter(sub_cases, cases < 500),
-    aes(x = date, y = cases), size = 0.2
-  ) +
-  facet_wrap(
-    ~factor(dep, levels = plot_order),
-    nrow = 2, labeller = dep_labeller
-  ) +
-  labs(y = 'Reported Cholera Cases') +
-  scale_x_date(date_breaks = "6 months", date_labels = "%Y-%m") +
-  theme(
-    axis.text.x = element_text(angle = 40, hjust = 1, vjust = 1, size = 7),
-    axis.title.x = element_blank()
-  )
-@
-\caption{\label{fig:mod3rep}
-Simulations from Model~3.
-Compare to Figure S18 of \cite{lee20}.
-}
-\end{figure}
-
-<<lee3Cleanup, echo=FALSE, include=FALSE, message=FALSE>>=
-rm(lee3, lee3_sims_quants, sub_cases)
-gc()
-@
-
-\section{Forecasting with parameter uncertainty}
-
-Let $f_{Y_{1:N}}(y_{1:N} | \theta)$ denote the pdf of the model under consideration, were $\theta$ is a parameter vector that indexes the model.
-Furthermore, denote the observed data as $y_{1:N}^*$.
-Because the uncertainty in just a single parameter can lead to drastically different forecasts \citep{saltelli20},
-parameter uncertainty should be considered when obtaining model forecasts when the goal is to influencing policy.
-In a Bayesian modelling paradigm, the most natural way to account for parameter uncertainty in model forecasts is to suppose that $\theta$ comes from a distribution $f_{\Theta}$, and then to obtain $J$ forecasts from the model where each forecast is obtained using parameters drawn from the posterior distribution $\Theta_{1:J} \mid Y_{1:N} = y_{1:N}^* \sim f_{\Theta}\big(\theta | Y_{1:N} = y_{1:N}^*\big)$.
-
-When frequentist methods are used, however, there does not exist a posterior distribution from which one could sample.
-A common approach could be to obtain a weighted average of the simulations from various models \citep{hoeting99}, but this can be problematic when forecasts from each model are very different from each other \citep{grueber11}.
-A similar approach that has been taken \citep{king15} is to obtain model forecasts using multiple sets of parameter values and then sample from the resulting forecasts using weights proportional to the corresponding likelihoods of the parameter values.
-This approach could be considered as empirical Bayes, as it is equivalent to using a discrete uniform prior where the set of values in the prior distribution is determined by a stochastic routine applied to the observed data, as discussed below.
-
-For each $i \in 1:k$, let $\Theta_i$ be a random vector of model parameters.
-Then, letting $\Theta$ denote the true model parameters, we endow the set $\{\Theta_1, \Theta_2, \ldots, \Theta_k\}$ with a discrete uniform distribution, such that $P\big(\Theta = \Theta_i\big) = \frac{1}{K}$ for all values $i \in \seq{1}{k}$.
-Using this as a prior distribution, the posterior distribution of $\Theta | Y_{1:N} = y_{1:N}^*$ can be expressed as:
-$P\big(\Theta = \Theta_k | Y_{1:N} = y_{1:N}^*\big) = \frac{f_{Y_{1:N}}(y_{1:N}^*| \Theta_k)}{\sum_{l = 1}^K f_{Y_{1:N}}(y_{1:N}^*| \theta_l)}$.
-In a standard empirical Bayes analysis, the values $\Theta_1, \ldots, \Theta_k$ of the prior distribution would be chosen using the observed data, resulting in a posterior distribution that weighs the prior parameter vectors proportional to their corresponding likelihoods.
-Instead, we choose $\Theta_i$ to be the output of a stochastic routine applied to the observed data by setting $\Theta_i$ to be the output of an iterated filtering algorithm.
-In practice, because the likelihood maximization routines of iterated filtering methods are stochastic, it is common to run the iterated filtering method multiple times for each model in order to obtain a maximum likelihood estimate for model parameters.
-This results in a natural set of parameters near the MLE that could be used as the discrete prior distribution.
+\input{inputs/paramUncertainty}
 
@@
 
-\begin{table}
-  \begin{center}
-  \begin{tabular}{|c|c|c|c|c|}\hline
-    \multirow{2}{*}{Parameter} & Our & \multicolumn{3}{c|}{Lee et al. (2020a)} \\\cline{3-5}
-     & Notation & 1 & 2 & 3 \\
-    \hline
-    \hline
-    % Table begins here
-    Reporting Rate & $\reportRate$ & $\rho$ & $\rho$ & $\epsilon_1, \epsilon_2$ \\\hline
-    Mixing Coefficient & $\mixExponent$ & $\nu$ & --- & --- \\\hline
-    Measurement Over-Dispersion & $\obsOverdispersion$ & $\tau$ & --- & $p$ \\\hline
-    Birth Rate & $\muBirth$ & $\mu$ & --- & --- \\\hline
-    Natural Mortality Rate & $\muDeath$ & $\delta$ & --- & $\mu$ \\\hline
-    Cholera Mortality Rate & $\choleraDeath$ & --- & --- & $\alpha$ \\\hline
-    Latent Period & $1/\muEI$ & $1/\sigma$ & $1/\gamma_E$ & --- \\\hline
-    Recovery Rate & $\muIR$ & $\gamma$ & $\gamma$ & $\gamma$ \\\hline
-    Loss of Immunity & $\muRS$ & $\alpha$ & $\sigma$ & $\rho$ \\\hline
-    Symptomatic Ratio & $\symptomFrac$ & $1 - \theta_0$ & $k$ & $\sigma$ \\\hline
-    Asymptomatic Relative Infectiousness & $\asymptomRelativeInfect$ & $\kappa$ & $red_\beta$ & --- \\\hline
-    Human-to-Water Shedding & $\Wshed$ & --- & $\mu$ & $\theta_I$ \\\hline
-    Asymptomatic Relative Shedding & $\asymptomRelativeShed$ & --- & $red_\mu$ & $\theta_A/\theta_I$ \\\hline
-    Seasonal Amplitude & $\seasAmplitude$ & --- & $\alpha_s$ & $\lambda$ \\\hline
-    Transmission & $\transmission$ & $\beta$ & $\beta$ & $c$ \\\hline
-    Water-to-Human & $\Wbeta{}$ & --- & $\beta_W$ & $\beta$ \\\hline
-    Bacteria Mortality & $\Wremoval$ & --- & $\delta$ & $\mu_\beta$ \\\hline
-    Vaccination Efficacy & $\vaccineEfficacy$ & $\theta_{vk}$ & $\theta_1, \theta_2, \theta_{1_5}, \theta_{2_5}$ & $\eta_{1d}, \eta_{2d}$ \\\hline
-    Process Over-dispersion & $\sigmaProc$ & --- & --- & $\sigma_w$\\\hline
-    % Half Saturation Constant & $\Wsat$ & --- & $Sat$ & $K$ \\\hline
-  \end{tabular}
-  \end{center}
-  \caption{
-  \label{tab:translate}Translations between our common notation and notation used by \cite{lee20}
-  }
-\end{table}
+\input{inputs/translateTab}
 

### 2023-09-06 si/ReplicateLee20.Rnw
Modified makefile because sed is not the same for mac and linux

@@
   geom_line(aes(y = ReportedAll), col = 'black', linetype = 'dashed') +
-  geom_line(aes(y = Ctotal), col = 'red', size = 1) +
+  geom_line(aes(y = Ctotal), col = 'red', linewidth = 1) +
   theme_bw() +

### 2023-09-06 si/clean_tex.R

@@ -1,0 +1,3 @@
+args <- commandArgs(trailingOnly = TRUE)
+source('utilityFuns.R')
+remove_start_end(args[1])

### 2023-09-06 si/utilityFuns.R

@@ -1,0 +1,8 @@
+remove_start_end <- function(file) {
+  tmp_file <- readLines(file)
+  start_line <- which(grepl("%* START", tmp_file))
+  end_line <- which(grepl("%* END", tmp_file))
+  tmp_file[start_line:end_line]
+  writeLines(tmp_file[(start_line+1):(end_line-1)], con = gsub("\\.tex", "Out.tex", file))
+}
+

### 2023-09-07 si/make_si_file.R
finished making Makefile to create supplement files

@@ -1,0 +1,3 @@
+args <- commandArgs(trailingOnly = TRUE)
+source('utilityFuns.R')
+make_si_file(args[1])

### 2023-09-07 si/utilityFuns.R

@@ -6,3 +6,21 @@ remove_start_end <- function(file) {
   writeLines(tmp_file[(start_line+1):(end_line-1)], con = gsub("\\.tex", "Out.tex", file))
 }
 
+make_si_file <- function(file) {
+  template <- readLines('inputs/si_files_template.tex')
+  input_file <- readLines(file)
+
+  if (grepl('Tab', file)) {
+    tab_caption <- which(grepl('\\caption', input_file))
+    input_file <- input_file[-tab_caption]
+  }
+
+  start_line <- which(grepl("%* START", template))
+  end_line <- which(grepl("%* END", template))
+  file_loc <- gsub("inputs/", "", file)
+  new_file <- c(template[1:(start_line - 1)], input_file, template[(end_line + 1):length(template)])
+  writeLines(new_file, con = file_loc)
+
+  has_citation <- any(grepl("\\cite", new_file))
+  if (has_citation) writeLines(as.character(has_citation), con = gsub('.tex', '.tmp', file_loc))
+}

### 2023-09-07 ms-plos.Rnw
Make-file updated to create tif images

@@
-% Template for PLoS
-% Version 3.6 Aug 2022
-%
-% % % % % % % % % % % % % % % % % % % % % %
-%
-% -- IMPORTANT NOTE
-%
-% This template contains comments intended
-% to minimize problems and delays during our production
-% process. Please follow the template instructions
-% whenever possible.
-%
-% % % % % % % % % % % % % % % % % % % % % % %
-%
-% Once your paper is accepted for publication,
-% PLEASE REMOVE ALL TRACKED CHANGES in this file
-% and leave only the final text of your manuscript.
-% PLOS recommends the use of latexdiff to track changes during review, as this will help to maintain a clean tex file.
-% Visit https://www.ctan.org/pkg/latexdiff?lang=en for info or contact us at latex@plos.org.
-%
-%
-% There are no restrictions on package use within the LaTeX files except that no packages listed in the template may be deleted.
-%
+% TODO: Remove colored text from table, instead, create a new symbol.
 % Please do not include colors or graphics in the text.
-%
-% The manuscript LaTeX source should be contained within a single file (do not use \input, \externaldocument, or similar commands).
 %
@@
 \begin{figure}[!h]
-\centering
 <<Model1_Seasonality_Figure, fig.height=2.9, fig.width=5, fig.align='center'>>=

### 2023-09-08 ms-plos.Rnw
Removed unnecessary comments from .Rnw + a few small edits

@@
 % % % % % % % % % % % % % % % % % % % % % % %
-%
-% -- FIGURES AND TABLES
-%
-% Please include tables/figure captions directly after the paragraph where they are first cited in the text.
-%
-% DO NOT INCLUDE GRAPHICS IN YOUR MANUSCRIPT
-% - Figures should be uploaded separately from your manuscript file.
-% - Figures generated using LaTeX should be extracted and removed from the PDF before submission.
-% - Figures containing multiple panels/subfigures must be combined into one image file before submission.
-% For figure citations, please use "Fig" instead of "Figure".
-% See http://journals.plos.org/plosone/s/figures for PLOS figure guidelines.
 %
@@
 %
-% For tables that exceed the width of the text column, use the adjustwidth environment as illustrated in the example table in text below.
-%
-% % % % % % % % % % % % % % % % % % % % % % % %
-%
 % -- EQUATIONS, MATH SYMBOLS, SUBSCRIPTS, AND SUPERSCRIPTS
@@
 %
-% For inline equations, please be sure to include all portions of an equation in the math environment.
-For example, x$^2$ is incorrect; this should be formatted as $x^2$ (or $\mathrm{x}^2$ if the romanized font is desired).
-%
+% TODO: See below
 % Do not include text that is not math in the math environment. For example, CO2 should be written as CO\textsubscript{2} instead of CO$_2$.
 %
-% Please add line breaks to long display equations when possible in order to fit size of the column.
-%
+% TODO: see below.
 % For inline equations, please do not include punctuation (commas, etc) within the math environment unless this is part of the equation.
 %
+% TODO: see below
 % When adding superscript or subscripts outside of brackets/braces, please group using {}.  For example, change "[U(D,E,\gamma)]^2" to "{[U(D,E,\gamma)]}^2".
 %
-% Do not use \cal for caligraphic font.  Instead, use \mathcal{}
-%
-% % % % % % % % % % % % % % % % % % % % % % % %
-%
-% Please contact latex@plos.org with any questions.
-%
-% % % % % % % % % % % % % % % % % % % % % % % %
 
@@
 
-%% START CUSTOM MACROS
-
-%% END CUSTOM MACROS
-
 \begin{document}
@@
 \\
-\textbf{2} Affiliation Dept/Program/Center, Institution Name, City, State, Country
-\\
-\textbf{3} Affiliation Dept/Program/Center, Institution Name, City, State, Country
+\textbf{2} Statistics and Data Science, Carnegie Mellon University, Pittsburgh, Pennsylvania, USA
+\\
+\textbf{3} Wharton Statistics and Data Science, University of Pennsylvania, Philadelphia, Pennsylvania, USA
 \\
@@
 % \textcurrency c Insert third current address
-
-% Deceased author note
-% \dag Deceased
-
-% Group/Consortium Author Note
-% \textpilcrow Membership list can be found in the Acknowledgments section.
 
@@
 
-% Please keep the abstract below 300 words
 \section*{Abstract}
@@
 
-% Please keep the Author Summary between 150 and 200 words
-% Use first person. PLOS ONE authors please skip this step.
-% Author Summary not valid for PLOS ONE submissions.
 \section*{Author summary}
@@
 
-% Use "Eq" instead of "Equation" for equation citations.
 \section*{Introduction}
@@
 Finally, we conclude with a discussion of the results.
-% This is followed by a consideration of the robustness of model based policy recommendations in Sec.~\ref{sec:advice}, and a discussion in Sec.~\ref{sec:discussion}.
 
@@
 
-# Search hyper-parameters for each run-level and search. This sets
-# parameters like number of particles, number of BPF iterations, and
-# number of replications for each search.
-# h3_fit_parms <- set_h3_fit_parms(
-#   RUN_LEVEL = RUN_LEVEL,
-#   rw_1.sd = chol_rw_1.sd,
-#   rw_2.sd = chol_rw_2.sd
-# )
-
 # Calculate the number of parameters that are estimated:
@@
 mod3_n_params <- sum(reg_rw_1.sd != 0, na.rm = TRUE) + sum(sapply(reg_rw_1.sd, is.expression)) - length(shared_param_names) * 9
-
-# haiti3_fit <- readRDS("model3/run_level_3/h3_revision.rds")
 
@@
 
-# This model is only fit for a subset of the data, so we will set data
-# that are not part of this subset as NA. This will allow us to estimate
-# the likelihood for only the values that were considered during the
-# parameter estimation phase.
-# set_as_na <- lee3_sub@times < lubridate::decimal_date(as.Date('2017-06-10'))
-# temp_ouest_cases <- lee3_sub@data['casesOuest', ]
-# temp_ouest_cases[set_as_na] <- NA_real_
-# lee3_sub@data['casesOuest', ] <- temp_ouest_cases
-
 lee3_all <- lee3_spatPomp(start_date = "2010-10-23")
@@
 Here, we have created a common notational framework that facilitates model comparison, and put all comparable model parameters---including parameters that were estimated or held constant---into Table~\ref{tab:params}.
-% Here, we have put all three models into a common notational framework.
 Translations back to the original notation of Lee et al.~\cite{lee20} are given in the supplement (\nameref{S1_Table}).
@@
 \caption{\label{tab:params} {\bf Model parameters.}}
-% \small
 \begin{tabular}{|p{0.15\linewidth}|p{0.14\linewidth}t|p{0.13\linewidth}t|p{0.17\linewidth}t|}
@@
   & \missing
-  % & \multicolumn{1}{c}{\hspace{16pt}\missing}
   &
@@
   & \multirow{5}{*}{\missing}
-  % & \multicolumn{1}{c}{\hspace{16pt}\missing}
   &
@@
   & \multirow{3}{*}{\eqref{model2:lambda}}
-  % \eqref{model2:lambda}
   & $\Wbeta{_{1:10}}= (
@@
   & \multirow{2}{*}{\missing}
-  % & \multicolumn{1}{c}{\hspace{16pt}\missing}
   &
@@
   & \missing
-  % & \multicolumn{1}{c}{\hspace{16pt}\missing}
   &
   & \missing
-  % & \multicolumn{1}{c}{\hspace{16pt}\missing}
   &
@@
   & \multirow{2}{*}{\missing}
-  % & \multicolumn{1}{c}{\multirow{2}{*}{\hspace{16pt}\missing}}
   &
@@
   & \multirow{4}{*}{\missing}
-% & \multicolumn{1}{c}{\multirow{2}{*}{\hspace{16pt}\missing}}
   &
   & \multirow{4}{*}{\missing}
-  % & \multicolumn{1}{c}{\multirow{2}{*}{\hspace{16pt}\missing}}
   &
@@
  \multirow{-1}{*}{Parameters} &
-  % & \multicolumn{1}{c}{}
   &
   &
-  % & \multicolumn{1}{c}{}
   &
@@
 Additional details about the numeric implementation of these models are provided in Sec.~S1 and S2 of the supplemental text (\nameref{S1_Text}).
-
-%%%%%%%%% 11111111111 %%%%%%%%%%
 
@@
 \mu_{S_{\vaccCounter}E_{\vaccCounter}} &=& \lambda(t),
-\\%[-3pt]
+\\
 \label{model1:EI}
 \mu_{E_{\vaccCounter}I_{\vaccCounter}} &=& \muEI\big(1-\vaccineEfficacy_\vaccCounter(t)\big),
-\\%[-3pt]
+\\
 \label{model1:EA}
 \mu_{E_{\vaccCounter}A_{\vaccCounter}} &=& \muEI\, \vaccineEfficacy_\vaccCounter(t),
-\\%[-3pt]
+\\
 \label{model1:toR}
 \mu_{I_{\vaccCounter}R_{\vaccCounter}} &=& \mu_{A_{\vaccCounter}R_{\vaccCounter}} = \muIR,
-\\%[-3pt]
+\\
 \label{model1:RS}
 \mu_{R_{\vaccCounter}S_{\vaccCounter}} &=& \muRS,
-\\%[-3pt]
+\\
 \label{model1:vacc}
 \mu_{S_0S_{\vaccCounter}} &=& \mu_{E_0E_{\vaccCounter}} = \mu_{I_0I_{\vaccCounter}} = \mu_{A_0A_{\vaccCounter}} = \mu_{R_0R_{\vaccCounter}} = \eta_{\vaccCounter}(t),
-\\%[-3pt]
+\\
 \label{model1:death}
@@
 See Sec.~S3 of the supplement material for more details.
-
-%%%%%%% 222222222222 %%%%%%%%%%
 
@@
 \mu_{S_{u\vaccCounter}E_{u\vaccCounter}} &=& \vaccineEfficacy_\vaccCounter \, \lambda,
-\\%[-3pt]
+\\
 \label{model2:mu_EI}
 \mu_{E_{u\vaccCounter}I_{u\vaccCounter}} &=& \symptomFrac\muEI, \quad \mu_{E_{u\vaccCounter}A_{u\vaccCounter}} = (1-\symptomFrac)\muEI,
-\\%[-3pt]
+\\
 \label{model2:mu_IR}
 \mu_{I_{u\vaccCounter}R_{u\vaccCounter}} &=& \mu_{A_{u\vaccCounter}R^A_{u\vaccCounter}} = \muIR,
-\\%[-3pt]
+\\
 \label{model2:RS}
 \mu_{R_{u\vaccCounter}S_{u\vaccCounter}} &=& \mu_{R^A_{u\vaccCounter}S_{u\vaccCounter}} = \muRS,
-\\%[-3pt]
+\\
 \label{model2:transport}
 \mu_{S_{u\vaccCounter}S_{{\varv}\vaccCounter}} &=& \mu_{E_{u\vaccCounter}E_{{\varv}\vaccCounter}} = \mu_{I_{u\vaccCounter} I_{{\varv}\vaccCounter}} = \mu_{A_{u\vaccCounter}A_{{\varv}\vaccCounter}} = \mu_{R_{u\vaccCounter}R_{{\varv}\vaccCounter}} = \mu_{R^A_{u\vaccCounter} R^A_{{\varv}\vaccCounter}} = T_{u\varv},
-\\%[-3pt]
+\\
 \label{model2:omega1}
 \mu_{S_{u1}S_{u0}} &=& \mu_{S_{u3}S_{u0}} = \omega_1,
-\\%[-3pt]
+\\
 \label{model2:omega2}
 \mu_{S_{u2}S_{u0}} &=& \mu_{S_{u4}S_{u0}} = \omega_2,
-\\%[-3pt]
+\\
 \label{model2:to_W}
 \mu_{\demography W_u} &=& \Wshed \left\{ \sum_{\vaccCounter=0}^4 I_{u\vaccCounter}(t) + \asymptomRelativeShed \sum_{\vaccCounter=0}^4 A_{u\vaccCounter}(t) \right\},
-\\%[-3pt]
+\\
 \label{model2:from_W}
 \mu_{W_u\demography} &=& \Wremoval,
-\\%[-3pt]
+\\
 \label{model2:water_transport}
@@
 
-%%%%%%% 333333333 %%%%%%%%%%
-
 \subsubsection*{Model~3}
@@
 \mu_{S_{u\vaccCounter}I_{u}} &=& \symptomFrac \,  \lambda_u \big(1-\vaccineEfficacy_{u\vaccCounter}(t)\big) \, d\Gamma/dt,
-\\%[-3pt]
+\\
 \label{eq:model3:SA}
 \mu_{S_{u\vaccCounter}A_{u}} &=& (1-\symptomFrac) \,  \lambda_u \big(1-\vaccineEfficacy_{u\vaccCounter}(t)\big) \,  d\Gamma/dt,
-\\%[-3pt]
+\\
 \label{eq:model3:IR}
 \mu_{I_{u}R_{u\vaccCounter 1}} &=& \mu_{A_{u}R_{u\vaccCounter 1}} = \muIR,
-\\%[-3pt]
+\\
 \label{eq:model3:IS}
 \mu_{I_{u}S_{u0}} &=& \muDeath + \choleraDeath, \hspace{2mm} \mu_{A_{u}S_{u0}} = \muDeath
-\\%[-3pt]
+\\
 \label{eq:model3:RRnext}
 \mu_{R_{u\vaccCounter 1}R_{u\vaccCounter 2}} &=& \mu_{R_{u\vaccCounter 2}R_{u\vaccCounter 3}} = 3\muRS,
-\\%[-3pt]
+\\
 \label{eq:model3:RS}
 \mu_{R_{u\vaccCounter k}S_{u0}} &=& \muDeath + 3\muRS \, \mathbf{1}_{\{k=3\}},
-\\%[-3pt]
+\\
 \label{eq:model3:water}
 \mu_{{\demography}W_u} &=& \big[1 + \seasAmplitude \big(J_u(t)\big)^r \big] \mathrm{Den}_u \, \Wshed \big[ I_{u}(t)+ \asymptomRelativeShed A_{u}(t) \big],
-\\%[-3pt]
+\\
 \label{eq:model3:Decay}
@@
                                 "V4" = '#4393c3')) +
-  # geom_vline(xintercept = yearsToDate(2014.161), linetype = 'dashed') +
   scale_y_log10(labels = scales::trans_format("log10", scales::math_format(10^.x))) +
@@
 The blue curve from 2012-2019 is the trajectory of the calibrated version of Model~2.
-Projections under the various vaccination scenarios, which are discussed in the \bf{Forecasts} subsection are also included.}
+Projections under the various vaccination scenarios, which are discussed in detail in the  {\bf Forecasts} subsection are also included.}
 \end{figure}
@@
 
-% Results and Discussion can be combined.
 \section*{Results}
@@
 
-% Include only the SI item label in the paragraph heading. Use the \nameref{label} command to cite SI items in the text.
 \paragraph*{S1 Table}
@@
 
-% Either type in your references using
-% \begin{thebibliography}{}
-% \bibitem{}
-% Text
-% \end{thebibliography}
-%
-% or
-%
-% Compile your BiBTeX database using our plos2015.bst
-% style file and paste the contents of your .bbl file
-% here. See http://journals.plos.org/plosone/s/latex for
-% step-by-step instructions.
-%
-% \begin{thebibliography}{10}
-%
-% \bibitem{bib1}
-% Conant GC, Wolfe KH.
-% \newblock {{T}urning a hobby into a job: how duplicated genes find new
-%   functions}.
-% \newblock Nat Rev Genet. 2008 Dec;9(12):938--950.
-%
-% \bibitem{bib2}
-% Ohno S.
-% \newblock Evolution by gene duplication.
-% \newblock London: George Alien \& Unwin Ltd. Berlin, Heidelberg and New York:
-%   Springer-Verlag.; 1970.
-%
-% \bibitem{bib3}
-% Magwire MM, Bayer F, Webster CL, Cao C, Jiggins FM.
-% \newblock {{S}uccessive increases in the resistance of {D}rosophila to viral
-%   infection through a transposon insertion followed by a {D}uplication}.
-% \newblock PLoS Genet. 2011 Oct;7(10):e1002337.
-%
-% \end{thebibliography}
-
-% TODO: we have to remove this and copy the output from the .bbl file.
 \bibliography{bib-haiti}

### 2023-09-08 ms-plos.Rnw
Finished acknoledgements, supplemental info, removed table colors

@@
-% TODO: Remove colored text from table, instead, create a new symbol.
 % Please do not include colors or graphics in the text.
@@
 % Do not include text that is not math in the math environment. For example, CO2 should be written as CO\textsubscript{2} instead of CO$_2$.
-%
-% TODO: see below.
-% For inline equations, please do not include punctuation (commas, etc) within the math environment unless this is part of the equation.
 %
@@
 \newcommand\missing{---}
-\newcommand\fixed{\color{blue}}
+\newcommand\fixed{^\dagger}
 \newcommand\demography{\bullet}
@@
 Infection (day)
-  & ${\fixed \muIR^{-1}=\Sexpr{myround(7/p1["gamma"],1)}}$
+  & ${\muIR^{-1}=\Sexpr{myround(7/p1["gamma"],1)}}\fixed$
   & \eqref{model1:toR}
-  & ${\fixed \muIR^{-1}=\Sexpr{myround(365/h2_fit$h2_params["gamma"],1)}}$
+  & ${\muIR^{-1}=\Sexpr{myround(365/h2_fit$h2_params["gamma"],1)}}\fixed $
   & \eqref{model2:mu_IR}
-  & ${\fixed \muIR^{-1}=\Sexpr{myround(365/p3["gamma1"],1)}}$
+  & ${\muIR^{-1}=\Sexpr{myround(365/p3["gamma1"],1)}}\fixed $
   & \eqref{eq:model3:IR}
@@
 Latency (day)
-  & ${\fixed \muEI^{-1}=\Sexpr{myround(7/p1["sigma"],1)}}$
+  & ${\muEI^{-1}=\Sexpr{myround(7/p1["sigma"],1)}}\fixed$
   & \eqref{model1:EA}
-  & ${\fixed \muEI^{-1}=\Sexpr{myround(365/h2_fit$h2_params["gammaE"],1)}}$
+  & ${\muEI^{-1}=\Sexpr{myround(365/h2_fit$h2_params["gammaE"],1)}}\fixed$
   & \eqref{model2:mu_EI}
@@
   & \eqref{model1:beta}
-  & \multirow{2}{*}{${\fixed \seasAmplitude=0.4}$}
+  & \multirow{2}{*}{${\seasAmplitude=0.4}\fixed$}
   & \multirow{2}{*}{\eqref{model2:lambda}}
@@
   \multirow{3}{*}{Immunity (yr)}
-  & \multirow{3}{*}{${\fixed \muRS^{-1}=\Sexpr{myround(1/p1["alpha"]/52,1)}}$}
+  & \multirow{3}{*}{${\muRS^{-1}=\Sexpr{myround(1/p1["alpha"]/52,1)}}\fixed$}
   & \multirow{3}{*}{\eqref{model1:RS}}
@@
   & \eqref{model2:RS}
-  & \multirow{3}{*}{${\fixed \muRS^{-1}=\Sexpr{myround(1/p3['rho1'],1)}}$}
+  & \multirow{3}{*}{${\muRS^{-1}=\Sexpr{myround(1/p3['rho1'],1)}}\fixed$}
   & \multirow{3}{*}{\eqref{eq:model3:RRnext}}
 \\
-& & & ${\fixed \omega_1^{-1} = \Sexpr{myround(1/h2_fit$h2_params["Omega1"],1)}}$ & \eqref{model2:omega1} & &\\
-& & & ${\fixed \omega_2^{-1} = \Sexpr{myround(1/h2_fit$h2_params["Omega2"],1)}}$ & \eqref{model2:omega2} & &\\
+& & & ${\omega_1^{-1} = \Sexpr{myround(1/h2_fit$h2_params["Omega1"],1)}}\fixed$ & \eqref{model2:omega1} & &\\
+& & & ${\omega_2^{-1} = \Sexpr{myround(1/h2_fit$h2_params["Omega2"],1)}}\fixed$ & \eqref{model2:omega2} & &\\
 \hline
 \multirow{2}{*}{Birth/death (yr)}
-  & ${\fixed \muBirth^{-1} = \Sexpr{myround(1/p1["mu"]/52,1)}}$
+  & ${\muBirth^{-1} = \Sexpr{myround(1/p1["mu"]/52,1)}}\fixed$
     %
@@
   &
-  & ${\fixed \muDeath^{-1} = \Sexpr{myround(1/p3["mu1"],1)}}$
+  & ${\muDeath^{-1} = \Sexpr{myround(1/p3["mu1"],1)}}\fixed$
   & \multirow{2}{*}{\eqref{eq:model3:IS}}
 \\
-& ${\fixed \muDeath^{-1} = \Sexpr{myround(1/p1["delta"]/52,1)}}$
+& ${\muDeath^{-1} = \Sexpr{myround(1/p1["delta"]/52,1)}}\fixed$
 & \eqref{model1:death}
@@
 &
-& ${\fixed \choleraDeath^{-1} = \Sexpr{myround(1/p3["alpha1"],2)}}$
+& ${\choleraDeath^{-1} = \Sexpr{myround(1/p3["alpha1"],2)}}\fixed$
 &
@@
 Sympt. frac.
-  & ${\fixed \symptomFrac_z(t)=c\vaccineEfficacy^*(t-\tau_d)}$
+  & ${\symptomFrac_z(t)=c\vaccineEfficacy^*(t-\tau_d)}\fixed$
   & (\ref{model1:EI}-\ref{model1:EA})
-  & ${\fixed \symptomFrac=\Sexpr{myround(h2_fit$h2_params["k"],1)}}$
+  & ${\symptomFrac=\Sexpr{myround(h2_fit$h2_params["k"],1)}}\fixed$
   & \eqref{model2:mu_EI}
-  & ${\fixed \symptomFrac=\Sexpr{myround(p3["sigma1"],2)}}$
+  & ${\symptomFrac=\Sexpr{myround(p3["sigma1"],2)}}\fixed$
   & \eqref{eq:model3:SA}
@@
 Asympt.
-  & \multirow{2}{*}{${\fixed \asymptomRelativeInfect=0.05}$}
+  & \multirow{2}{*}{${\asymptomRelativeInfect=0.05}\fixed$}
   & \multirow{2}{*}{\eqref{model1:lambda}}
-  & ${\fixed \asymptomRelativeInfect = 0.001}$
+  & ${\asymptomRelativeInfect = 0.001}\fixed$
   & \eqref{model2:lambda}
-  &  ${\fixed \asymptomRelativeInfect = 1}$
+  &  ${\asymptomRelativeInfect = 1}\fixed$
   & \eqref{eq:model3:foi}
@@
 &
-& ${\fixed \asymptomRelativeShed = \Sexpr{1e-7} }$
+& ${\asymptomRelativeShed = \Sexpr{1e-7}}\fixed$
 & \eqref{model2:to_W}
@@
   &
-  & \multirow{3}{*}{${\fixed \Wsat = \Sexpr{signif(h2_fit$h2_params["Sat"],2)}}$}
+  & \multirow{3}{*}{${\Wsat = \Sexpr{signif(h2_fit$h2_params["Sat"],2)}}\fixed$}
   & \multirow{3}{*}{\eqref{model2:lambda}}
@@
   &
-  & \multirow{2}{*}{${\fixed \Wremoval^{-1} = \Sexpr{signif(52/h2_fit$h2_params["Delta"],3)}}$}
+  & \multirow{2}{*}{${\Wremoval^{-1} = \Sexpr{signif(52/h2_fit$h2_params["Delta"],3)}}\fixed$}
   & \multirow{2}{*}{\eqref{model2:from_W}}
@@
   & (S15)
-  & ${\fixed \reportRate=\Sexpr{myround(h2_fit$h2_params["Rho"],2)}}$
+  & ${\reportRate=\Sexpr{myround(h2_fit$h2_params["Rho"],2)}}\fixed$
   & (S16)
@@
 \begin{flushleft} References to the relevant equation are given in parentheses.
-Parameters in blue were fixed based on scientific reasoning and not fitted to the data.
-$^*$ denotes parameters added during our re-analysis, not considered by Lee et al.
-Translations back into the notation of \cite{lee20} are given in Table~S1.
+Parameters that were fixed and not calibrated using the data are indicated with $\fixed$;
+parameters that were added during our re-analysis and were not considered by Lee et al. are indicated with $^*$.
+Translations back into the notation of \cite{lee20} are given in \nameref{S1_Table}.
 \end{flushleft}
@@
 A first-time reader may choose to skim through the rest of this section, and return later.
-Additional details about the numeric implementation of these models are provided in Sec.~S1 and S2 of the supplemental text (\nameref{S1_Text}).
+Additional details about the numeric implementation of these models are provided in a supplemental text (\nameref{S1_Text}).
+While each of the dynamic models considered in this manuscript can be fully described using the mathematical equations provided in the following section, diagrams of dynamic systems are often helpful to understand the equations.
+For this reason, we provide flow chart diagrams for Models~1--3 in supplement figures (\nameref{S_Mod1}, \nameref{S_Mod2} and \nameref{S_Mod3}).
+Because flow diagrams do not uniquely describe a dynamic model, these diagrams are not unique visual representations of each model.
+Alternative representations that may be helpful in understanding the models explored in this paper were made by Lee et al. \cite{lee20sup}.
 
@@
 The measurement model assumes reported cholera cases at time point $n$ come from a negative binomial distribution, where only a fraction ($\reportRate$) of new weekly cases are reported.
-See Sec.~S3 of the supplement material for more details.
+More details about the initialization model $f_{\bm{X}_0}\big(\bm{x}_0;\paramVec\big)$ and the \emph{measurement model} $f_{\bm{Y}_n|\bm{X}_{n}}\big(\bm{y}_n|\bm{x}_{n}\big)$ for Models~1--3 are provided a supplement text (\nameref{S_init} and \nameref{S_meas}).
 
@@
 On this same subset, our model has a likelihood of $\Sexpr{myround(mod3_eval_res$subset_ll, 1)}$.
+Details of estimating the likelihood of the Lee et al.~\cite{lee20} models is provided in \nameref{S_lee20}
 See Sec.~S6 of the supplement material for more details on estimating the likelihood of the Lee et al.~\cite{lee20} models.
@@
 Additional techniques include the inspection of conditional log-likelihoods of each observation given the previous observations in order to understand how well the model describes each data point (Sec.~S5).
-Other tools include plotting the effective sample size of each observation \cite{liu01}; plotting the values of the hidden states from simulations (supplement Sec.~5.1);
+Other tools include plotting the effective sample size of each observation \cite{liu01}; plotting the values of the hidden states from simulations (\nameref{S_mod3cal});
 and comparing summary statistics of the observed data to simulations from the model \cite{wood10,king15}.
@@
 For example, reconstructed latent variables (such as numbers of susceptible and recovered individuals) should make sense in the context of alternative measurements of these variables \cite{grad12}.
-In the context of our case study, by reconstructing the latent states of Model~3, we notice that the calibrated model favors higher levels of cholera transmission than what was typically observed in the incidence data (Sec.~S5.1 of the supplement).
+In the context of our case study, by reconstructing the latent states of Model~3, we notice that the calibrated model favors higher levels of cholera transmission than what was typically observed in the incidence data (\nameref{S_mod3cal}).
 This result hints at the possibility of model mispecification, and should prompt us to be sceptical of the reliability of forecasts from this model.
@@
 If a Bayesian technique is used for parameter estimation, a natural way to account for parameter uncertainty is to obtain simulations from the model where each simulation is obtained using parameters drawn from the estimated posterior distribution.
-For frequentist inference, one possible approach is obtaining model forecasts from various values of $\paramVec$, where the values of $\paramVec$ are sampled proportionally according to their corresponding likelihoods \cite{king15}; see Sec.~S7 of the supplement material for more details.
+For frequentist inference, one possible approach is obtaining model forecasts from various values of $\paramVec$, where the values of $\paramVec$ are sampled proportionally according to their corresponding likelihoods \cite{king15} (\nameref{S_uncertain}).
 Both of these approaches share the similarity that parameters are chosen for the forecast approximately in proportion to their corresponding value of the likelihood function, $f_{\bm{Y}_{1:N}}(\bm{y}_{1:N}^*; \paramVec)$.
@@
 
+\paragraph*{S1 Fig.}
+\label{S_Mod1}
+{\bf Model~1.} Flow chart representation of Model~1.
+
+\paragraph*{S2 Fig.}
+\label{S_Mod2}
+{\bf Model~2.} Flow chart representation of Model~2.
+
+\paragraph*{S3 Fig.}
+\label{S_Mod3}
+{\bf Model~3.} Flow chart representation of Model~3.
+
 \paragraph*{S1 Table}
@@
 \label{S1_Text}
-{\bf Supplemental Material.} In depth description of the numeric implementation of models.
-
-\paragraph*{S2 Fig.}
-\label{S2_Fig}
-{\bf Lorem ipsum.} Maecenas convallis mauris sit amet sem ultrices gravida.
-Etiam eget sapien nibh.
-Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula.
-Curabitur fringilla pulvinar lectus consectetur pellentesque.
-
-\paragraph*{S1 File.}
-\label{S1_File}
-{\bf Lorem ipsum.}  Maecenas convallis mauris sit amet sem ultrices gravida.
-Etiam eget sapien nibh.
-Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula.
-Curabitur fringilla pulvinar lectus consectetur pellentesque.
-
-\paragraph*{S1 Video.}
-\label{S1_Video}
-{\bf Lorem ipsum.}  Maecenas convallis mauris sit amet sem ultrices gravida.
-Etiam eget sapien nibh.
-Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula.
-Curabitur fringilla pulvinar lectus consectetur pellentesque.
-
-\paragraph*{S1 Appendix.}
-\label{S1_Appendix}
-{\bf Lorem ipsum.} Maecenas convallis mauris sit amet sem ultrices gravida.
-Etiam eget sapien nibh.
-Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula.
-Curabitur fringilla pulvinar lectus consectetur pellentesque.
-
-% \paragraph*{S1 Table.}
-% \label{S1_Table}
-% {\bf Lorem ipsum.} Maecenas convallis mauris sit amet sem ultrices gravida.
-Etiam eget sapien nibh.
-Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula.
-Curabitur fringilla pulvinar lectus consectetur pellentesque.
+{\bf Model details.} In depth description of the Markov chain and differential equation interpretations of compartment flow rates, as well as details on the numeric implementation of these models.
+
+\paragraph*{S2 Text}
+\label{S_init}
+{\bf Initialization models.} Additional details of the initialization model that were used for Models~1--3.
+
+\paragraph*{S3 Text}
+\label{S_meas}
+{\bf Measurement models.} Additional details of the measurement model that were used for Models~1--3.
+
+\paragraph*{S4 Text}
+\label{S_lee20}
+{\bf Likelihood of Lee et al. models.} Details of how the log-likelihood of the Lee et al. models were calculated.
+
+\paragraph*{S5 Text}
+\label{S_mod3cal}
+{\bf Likelihood of Lee et al. models.} Additional details on the procedures for fitting and diagnosing Model~3.
+
+\paragraph*{S6 Text}
+\label{S_uncertain}
+{\bf Accounting for parameter uncertainty} A description of a empirical-Bayes approach that can be used to account for parameter uncertainty.
 
 \section*{Acknowledgments}
-Cras egestas velit mauris, eu mollis turpis pellentesque sit amet.
-Interdum et malesuada fames ac ante ipsum primis in faucibus.
-Nam id pretium nisi.
-Sed ac quam id nisi malesuada congue.
-Sed interdum aliquet augue, at pellentesque quam rhoncus vitae.
+
+This work was supported by National Science Foundation grants DMS-1761603 and DMS-1646108.
+The authors would like to thank Mercedes Pascual and Betz Halloran for helpful discussions. Laura Matrajt provided additional data for the Model~2 analysis.
 

### 2023-09-08 removeTrailingNums.R
Created two R files that help create plos comp bio submission

@@ -1,0 +1,8 @@
+args <- commandArgs(trailingOnly = TRUE)
+
+removeTrailingNums <- function(file_name) {
+  new_name <- gsub("-1-1", '', file_name)
+  file.rename(from = file_name, to = new_name)
+}
+
+removeTrailingNums(args[1])

### 2023-09-08 remove_knitrout.R

@@ -1,0 +1,27 @@
+args <- commandArgs(trailingOnly = TRUE)
+
+remove_knitrout <- function(file_name) {
+  tmp_file <- readLines(file_name)
+  knitrout_env <- which(grepl("\\newenvironment{knitrout}{}{}", tmp_file, fixed = TRUE))
+  begin_env <- which(grepl("\\begin{knitrout}", tmp_file, fixed = TRUE))
+  end_env <- which(grepl("\\end{knitrout}", tmp_file, fixed = TRUE))
+
+  if (length(begin_env) != length(end_env)) {
+    stop("Number of \\begin{knitrout} does not match number of \\end{knitrout}.")
+  }
+
+  row_ids <- c(knitrout_env)
+
+  for (i in 1:length(begin_env)) {
+    if (begin_env[i] > end_env[i]) {
+      stop("Problem matching \\begin{knitrout} with \\end{knitrout}")
+    }
+
+    row_ids <- c(row_ids, begin_env[i]:end_env[i])
+  }
+
+  writeLines(tmp_file[-row_ids], con = gsub("\\.tex", "-submission.tex", file_name))
+  # file.rename(from = file_name, to = new_name)
+}
+
+remove_knitrout(args[1])

### 2023-09-11 ms-plos.Rnw
Implemented some of Ed's minor suggestions

@@
 For more tips to help reduce the possibility of formatting errors during conversion, please see our LaTeX guidelines at http://journals.plos.org/plosone/s/latex
-%
-% TODO: See below
-% Do not include text that is not math in the math environment. For example, CO2 should be written as CO\textsubscript{2} instead of CO$_2$.
-%
-% TODO: see below
-% When adding superscript or subscripts outside of brackets/braces, please group using {}.  For example, change "[U(D,E,\gamma)]^2" to "{[U(D,E,\gamma)]}^2".
 %
@@
 \newcommand\modelCounter{m}
-\newcommand\missing{---}
+\newcommand\missing{}
 \newcommand\fixed{^\dagger}
@@
 We build on a multi-group modeling exercise by Lee et al.~\cite{lee20} in which four expert modeling teams developed models to the same dataset with the goal of comparing conclusions on the feasibility of eliminating cholera by a vaccination campaign.
-The data that was used for this study, and that is the subject of our reanalysis, is displayed in Fig.~\ref{fig:CholeraData}.
+The data that were used for this study, and that is the subject of our reanalysis, is displayed in Fig.~\ref{fig:CholeraData}.
 Model~1 is stochastic and describes cholera at the national level;
@@
 The models make various choices on the dichotomies considered by Lee et al.~\cite{lee20}; deterministic versus stochastic; compartment model versus agent-based; aggregated versus spatially explicit.
-This is no accident, since Lee et al.~\cite{lee20} purposefully designed their study to cover the range of current modeling practice.
 
@@
 Asympt.
-  & \multirow{2}{*}{${\asymptomRelativeInfect=0.05}\fixed$}
-  & \multirow{2}{*}{\eqref{model1:lambda}}
+  & ${\asymptomRelativeInfect=0.05}\fixed$
+  & \eqref{model1:lambda}
   & ${\asymptomRelativeInfect = 0.001}\fixed$
@@
 infectivity
-&
+& \missing
 &
@@
 Additional details about the numeric implementation of these models are provided in a supplemental text (\nameref{S1_Text}).
-While each of the dynamic models considered in this manuscript can be fully described using the mathematical equations provided in the following section, diagrams of dynamic systems are often helpful to understand the equations.
+While each of the dynamic models considered in this manuscript can be fully described using the mathematical equations provided in the following section, diagrams of dynamic systems can be helpful to understand the equations.
 For this reason, we provide flow chart diagrams for Models~1--3 in supplement figures (\nameref{S_Mod1}, \nameref{S_Mod2} and \nameref{S_Mod3}).
-Because flow diagrams do not uniquely describe a dynamic model, these diagrams are not unique visual representations of each model.
-Alternative representations that may be helpful in understanding the models explored in this paper were made by Lee et al. \cite{lee20sup}.
 
@@
 where $\vaccineEfficacy^*(t)$ is efficacy at time $t$ since vaccination for adults, taken from \cite{lee20}, Table~S4, and $c=\big(1-(1-0.4688)\times 0.11\big)$ is a correction to allow for reduced efficacy in the 11\% of the population aged under 5 years.
-Single and double vaccine doses were modeled by changing the waning of protection; protection was assumed to be equal between single and double dose until 52 weeks after vaccination, at which point the single dose becomes ineffective.
+Single and double vaccine doses were modeled by changing the waning of protection; protection was modeled as equal between single and double dose until 52 weeks after vaccination, at which point the single dose becomes ineffective.
 
 The latent state vector $\bm{X}(t)$ is initialized by setting the counts for each compartment and vaccination scenario $\vaccCounter \neq 0$ as zero, and introducing initial-value parameters $I_{0,0}$ and $E_{0, 0}$ such that $R_{0}(0) = 0$, $I_{0}(0) = \mathrm{Pop} \times I_{0, 0}$, $E_{0}(0) = \mathrm{Pop} \times E_{0, 0}$ and $S_{0}(0) = \mathrm{Pop} \times (1 - I_{0, 0} - E_{0, 0})$, where $\mathrm{Pop}$ is the total population of Haiti.
-The measurement model assumes reported cholera cases at time point $n$ come from a negative binomial distribution, where only a fraction ($\reportRate$) of new weekly cases are reported.
+The measurement model describes reported cholera cases at time point $n$ come from a negative binomial distribution, where only a fraction ($\reportRate$) of new weekly cases are reported.
 More details about the initialization model $f_{\bm{X}_0}\big(\bm{x}_0;\paramVec\big)$ and the \emph{measurement model} $f_{\bm{Y}_n|\bm{X}_{n}}\big(\bm{y}_n|\bm{x}_{n}\big)$ for Models~1--3 are provided a supplement text (\nameref{S_init} and \nameref{S_meas}).
@@
 The starting value for each element of the latent state vector $\bm{X}(0)$ are set to zero except for $I_{u0}(0) = y_u^*(0)/\reportRate$ and $R_{u0}(0) = \mathrm{Pop}_u - I_{u0}(0)$, where $y_u^*(0)$ is the reported number of cholera cases in department $u$ at time $t = 0$.
-Reported cases are assumed to come from a log-normal distribution, with the log-scale mean equal to the reporting rate $\reportRate$ times the number of newly infected individuals.
+Reported cases are described using a log-normal distribution, with the log-scale mean equal to the reporting rate $\reportRate$ times the number of newly infected individuals.
 See the section describing how Model~2 is fit to data and Sec.~S3.2 of the supplement for more details.
@@
 Therefore, for departments with zero case counts in week 1, we consider estimating the number of infected individuals rather then treating this value as a constant (see supplement Sec.~S5 for more details).
-Finally, reported cholera cases are assumed to come from a negative binomial measurement model with mean equal to a fraction ($\reportRate$) of individuals in each unit who develop symptoms and seek healthcare
+Finally, reported cholera cases are modeled using a negative binomial measurement model with mean equal to a fraction ($\reportRate$) of individuals in each unit who develop symptoms and seek healthcare
 (see Sec.~S3.3 of the supplement).
@@
 Therefore, the choice of $\paramVec$ used to make inference about the system can greatly affect model based conclusions.
-Elements of $\paramVec$ can be fixed at a constant value based on scientific understanding of the system, but parameters can also be calibrated to data by maximizing a measure of congruency between the observed data and the assumed mechanistic structure.
-Calibrating model parameters to observed data does not guarantee that the resulting model successfully approximates real-world mechanisms, since model assumptions may be incorrect, and do not change as the model is calibrated to data.
+Elements of $\paramVec$ can be fixed at a constant value based on scientific understanding of the system, but parameters can also be calibrated to data by maximizing a measure of congruency between the observed data and the model's mechanistic structure.
+Calibrating model parameters to observed data does not guarantee that the resulting model successfully approximates real-world mechanisms, since the model description of the dynamic system may be incorrect and do not change as the model is calibrated to data.
 However, the congruency between the model and observed data serves as a proxy for the congruency between the model and the true underlying dynamic system.
@@
 The particle filter is often used to obtain values of latent states, but can also be used to obtain likelihood estimates of a POMP model.
-Iterated filtering algorithms, like IF2 \cite{ionides15}, extend the particle filter by performing a random walk for each parameter and particle;
+Iterated filtering algorithms such as IF2 \cite{ionides15} extend the particle filter by performing a random walk for each parameter and particle;
 these perturbations are carried out iteratively over multiple filtering operations, using the collection of parameters from the previous filtering pass as the parameter initialization for the next iteration, and decreasing the random walk variance at each step.
@@
 as parameters that maximize the likelihood of the decoupled model almost certainly do not maximize the likelihood of the fully coupled model.
-These two concerns may explain the unrealistic forecasts and low likelihood of Model~3 from Lee et al.~\cite{lee20} (see Table~\ref{tab:likes}).
-
-At the time Lee et al.~\cite{lee20} was published, there was no known algorithm that could readily be used to maximize the likelihood of an arbitrary meta-population POMP model with coupled spatial dynamics, which justifies the spatial decoupling approximation that was used to calibrate model parameters.
+These two concerns may explain the unrealistic forecasts and low likelihood of Model~3 of Lee et al.~\cite{lee20} (see Table~\ref{tab:likes}).
+
+At the time Lee et al.~\cite{lee20} conducted their study, there was no known algorithm that could readily be used to maximize the likelihood of an arbitrary meta-population POMP model with coupled spatial dynamics, which justifies the spatial decoupling approximation that was used to calibrate model parameters.
 For our analysis, we calibrate the parameters of the spatially coupled version of Model~3 using the iterated block particle filter (IBPF) algorithm \cite{ionides22}.
@@
 \subsection*{Robust interpretation of model based conclusions}\label{sec:advice}
+
+In this section, we make some general points on deriving scientific conclusions from statistical analysis of mechanistic models, and connect these points to our specific case study.
 

### 2023-09-11 si/ReplicateLee20.Rnw
Implemented some of Ed's minor suggestions

@@
 
-<<PlotEpiDist, echo=FALSE, fig.cap="Compare to Figure S8 of Lee et al. (2020) supplement.", fig.width=7, fig.height=7, fig.align='center',message=FALSE,warning=FALSE>>=
+<<PlotEpiDist, echo=FALSE, fig.cap="Bivariate distributions of parameter estimates after fitting epidemic phase of the Model~1 following the procedure described by Lee et al.
+(2020).
+Compare to Figure S8 in the supplement of Lee et al.
+(2020).", fig.width=7, fig.height=7, fig.align='center',message=FALSE,warning=FALSE>>=
 GGally::ggpairs(
@@
 
-<<plotEndParams, fig.width=7, fig.height=7, fig.align='center', fig.cap="Bivariate relationships between variables after fitting endemic period.
-Compare to S9 of Lee et al.
-(2020) supplement.", cache=TRUE, echo=FALSE,warning=FALSE>>=
+<<plotEndParams, fig.width=7, fig.height=7, fig.align='center', fig.cap="Bivariate distributions of parameter estimates after fitting endemic phase of the Model~1 following the procedure described by Lee et al.
+(2020).
+Compare to Figure S9 in the supplement of Lee et al.
+(2020).", cache=TRUE, echo=FALSE,warning=FALSE>>=
 GGally::ggpairs(
@@
 <<plotMod1Sims, echo=FALSE, fig.cap=paste0("Simulations using parameter sets that were generated by running source code provided by Lee et al (2020).
-Compare to Figure S7 of \\cite{lee20sup}.
+Compare to Figure S7 in the supplement of Lee et al.
+(2020).
 The upper bound for the likelihood of this model is ", round(mod1_lee_ll), "."), cache=TRUE, fig.width=6.5, fig.height=3.75,warning=FALSE>>=
@@
 \caption{\label{fig:mod2rep}
-Model 2 trajectories using the \code{haitipkg}. Compare to Figure S15 of \cite{lee20sup}.
+Model 2 trajectories using the \code{haitipkg}. Compare to Figure S15 in the supplement of \cite{lee20sup}.
 }
@@
 Simulations from Model~3.
-Compare to Figure S18 of \cite{lee20}.
+Compare to Figure S18 in the supplement of \cite{lee20}.
 }

### 2023-09-11 si/ReplicateLee20Out.tex
Modified makefile because sed is not the same for mac and linux
Implemented some of Ed's minor suggestions

@@
 
-\caption[Compare to Figure S8 of Lee et al]{Compare to Figure S8 of Lee et al. (2020) supplement.}\label{fig:PlotEpiDist}
+\caption[Bivariate distributions of parameter estimates after fitting epidemic phase of the Model~1 following the procedure described by Lee et al]{Bivariate distributions of parameter estimates after fitting epidemic phase of the Model~1 following the procedure described by Lee et al.
+(2020).
+Compare to Figure S8 in the supplement of Lee et al.
+(2020).}\label{fig:PlotEpiDist}
 \end{figure}
@@
 
-\caption[Bivariate relationships between variables after fitting endemic period]{Bivariate relationships between variables after fitting endemic period.
-Compare to S9 of Lee et al.
-(2020) supplement.}\label{fig:plotEndParams}
+\caption[Bivariate distributions of parameter estimates after fitting endemic phase of the Model~1 following the procedure described by Lee et al]{Bivariate distributions of parameter estimates after fitting endemic phase of the Model~1 following the procedure described by Lee et al.
+(2020).
+Compare to Figure S9 in the supplement of Lee et al.
+(2020).}\label{fig:plotEndParams}
 \end{figure}
@@
 \caption[Simulations using parameter sets that were generated by running source code provided by Lee et al (2020)]{Simulations using parameter sets that were generated by running source code provided by Lee et al (2020).
-Compare to Figure S7 of \cite{lee20sup}.
+Compare to Figure S7 in the supplement of Lee et al.
+(2020).
 The upper bound for the likelihood of this model is -3031.}\label{fig:plotMod1Sims}

### 2023-09-11 si/calibrateMod3.Rnw
Implemented some of Ed's minor suggestions

@@
 The addition of the Hurricane parameters seems to have resulted in a dramatic increase in conditional likelihoods around and after October 2016, as seen in Fig.~\ref{fig:finalCondLL}.
-The inclusion of these parameters resulted in an overall increase of $\Sexpr{round(max(h3_final_results$search2$logLiks$logLik) - max(haiti3_sub_fit$search2$logLiks$logLik), 1)}$ log-likelihood units, which is a highly statistically significant difference.
+The inclusion of these parameters resulted in an overall increase of $\Sexpr{round(max(h3_final_results$search2$logLiks$logLik) - max(haiti3_sub_fit$search2$logLiks$logLik), 1)}$ log-likelihood units.
+Such a large difference in log-likelihoods is well beyond the threshold of statistical uncertainty determined by Wilks' theorem, suggesting that the data highly favor the inclusion of the additional parameters.
 

### 2023-09-11 si/calibrateMod3Out.tex
Implemented some of Ed's minor suggestions

@@
 The addition of the Hurricane parameters seems to have resulted in a dramatic increase in conditional likelihoods around and after October 2016, as seen in Fig.~\ref{fig:finalCondLL}.
-The inclusion of these parameters resulted in an overall increase of $157.1$ log-likelihood units, which is a highly statistically significant difference.
+The inclusion of these parameters resulted in an overall increase of $157.1$ log-likelihood units.
+Such a large difference in log-likelihoods is well beyond the threshold of statistical uncertainty determined by Wilks' theorem, suggesting that the data highly favor the inclusion of the additional parameters.
 

### 2023-09-11 ms.Rnw
Removed unused ncores argument

@@
-%% Based on the template for the submission to:
-%%   The Annals of Applied Statistics [AOAS]
-
-%\documentclass[aoas]{imsart}
-\documentclass[aoas]{imsart}
-
-%% Packages
+% Please do not include colors or graphics in the text.
+%
+% % % % % % % % % % % % % % % % % % % % % % %
+%
+% Tables should be cell-based and may not contain:
+% - spacing/line breaks within cells to alter layout or alignment
+% - do not nest tabular environments (no tabular environments within tabular environments)
+% - no graphics or colored text (cell background color/shading OK)
+% See http://journals.plos.org/plosone/s/tables for table guidelines.
+%
+% -- EQUATIONS, MATH SYMBOLS, SUBSCRIPTS, AND SUPERSCRIPTS
+%
+% IMPORTANT
+% Below are a few tips to help format your equations and other special characters according to our specifications.
+For more tips to help reduce the possibility of formatting errors during conversion, please see our LaTeX guidelines at http://journals.plos.org/plosone/s/latex
+%
+
+\documentclass[10pt,letterpaper]{article}
+\usepackage[top=0.85in,left=2.75in,footskip=0.75in]{geometry}
+
+% amsmath and amssymb packages, useful for mathematical formulas and symbols
+\usepackage{amsmath,amssymb}
+
+% Use adjustwidth environment to exceed column width (see example table in text)
+\usepackage{changepage}
+
+% textcomp package and marvosym package for additional characters
+\usepackage{textcomp,marvosym}
+
+% cite package, to clean up citations in the main text. Do not remove.
+\usepackage{cite}
+
+% Use nameref to cite supporting information files (see Supporting Information section for more info)
+\usepackage{nameref,hyperref}
+
+% line numbers
+\usepackage[right]{lineno}
+
+% ligatures disabled
+\usepackage[nopatch=eqnum]{microtype}
+\DisableLigatures[f]{encoding = *, family = * }
+
+% color can be used to apply background shading to table cells only
+\usepackage[table]{xcolor}
+
+% array package and thick rules for tables
+\usepackage{array}
+
+% create "+" rule type for thick vertical lines
+\newcolumntype{+}{!{\vrule width 2pt}}
+
+% create \thickcline for thick horizontal lines of variable length
+\newlength\savedwidth
+\newcommand\thickcline[1]{%
+  \noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
+  \cline{#1}%
+  \noalign{\vskip\arrayrulewidth}%
+  \noalign{\global\arrayrulewidth\savedwidth}%
+}
+
+% \thickhline command for thick horizontal lines that span the table
+\newcommand\thickhline{\noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
+\hline
+\noalign{\global\arrayrulewidth\savedwidth}}
+
+% Remove comment for double spacing
+%\usepackage{setspace}
+%\doublespacing
+
+% Text layout
+\raggedright
+\setlength{\parindent}{0.5cm}
+\textwidth 5.25in
+\textheight 8.75in
+
+% Bold the 'Figure #' in the caption and separate it from the title/caption with a period
+% Captions will be left justified
+\usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,justification=raggedright,singlelinecheck=off]{caption}
+\renewcommand{\figurename}{Fig}
+
+% Use the PLoS provided BiBTeX style
+\bibliographystyle{plos2015}
+
+% Remove brackets from numbering in List of References
+\makeatletter
+\renewcommand{\@biblabel}[1]{\quad#1.}
+\makeatother
+
+% Header and Footer with logo
+\usepackage{lastpage,fancyhdr,graphicx}
+\usepackage{epstopdf}
+%\pagestyle{myheadings}
+\pagestyle{fancy}
+\fancyhf{}
+%\setlength{\headheight}{27.023pt}
+%\lhead{\includegraphics[width=2.0in]{PLOS-submission.eps}}
+\rfoot{\thepage/\pageref{LastPage}}
+\renewcommand{\headrulewidth}{0pt}
+\renewcommand{\footrule}{\hrule height 2pt \vspace{2mm}}
+\fancyheadoffset[L]{2.25in}
+\fancyfootoffset[L]{2.25in}
+\lfoot{\today}
+
+%% Include all macros below
+
+\newcommand{\lorem}{{\bf LOREM}}
+\newcommand{\ipsum}{{\bf IPSUM}}
+
+%% END MACROS SECTION
+
 \RequirePackage{amsthm,amsmath,amsfonts,amssymb,graphicx,enumerate,url,xr,lmodern}
-\RequirePackage[authoryear]{natbib}
 \usepackage{url}
@@
 \newcommand\modelCounter{m}
-\newcommand\missing{---}
-\newcommand\fixed{\color{blue}}
+\newcommand\missing{}
+\newcommand\fixed{^\dagger}
 \newcommand\demography{\bullet}
 \newcommand\code[1]{\texttt{#1}}
+\newcommand\figTitle{\bf}
 \newcommand\paramVec{\theta}
 \newcommand\childReduce{q}
+\newcommand\NBintercept{\alpha}
+\newcommand\NBar{\beta}
+\newcommand\NBsize{\varphi}
 
@@
 
-\endlocaldefs
-
 \begin{document}
+\vspace*{0.2in}
+
+% Title must be 250 characters or less.
+\begin{flushleft}
+{\Large
+\textbf\newline{Informing policy via dynamic models: Cholera in Haiti} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
+}
+\newline
+% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
+\\
+Jesse Wheeler\textsuperscript{1*},
+AnnaElaine L. Rosengart\textsuperscript{2},
+Zhuoxun Jiang\textsuperscript{1},
+Kevin Tan\textsuperscript{3},
+Noah Treutle\textsuperscript{1},
+Edward L. Ionides\textsuperscript{1}
+\\
+\bigskip
+\textbf{1} Statistics Department, University of Michigan, Ann Arbor, Michigan, USA
+\\
+\textbf{2} Statistics and Data Science, Carnegie Mellon University, Pittsburgh, Pennsylvania, USA
+\\
+\textbf{3} Wharton Statistics and Data Science, University of Pennsylvania, Philadelphia, Pennsylvania, USA
+\\
+\bigskip
+
+% Insert additional author notes using the symbols described below. Insert symbol callouts after author names as necessary.
+%
+% Remove or comment out the author notes below if they aren't used.
+%
+% Primary Equal Contribution Note
+% \Yinyang These authors contributed equally to this work.
+
+% Additional Equal Contribution Note
+% Also use this double-dagger symbol for special authorship notes, such as senior authorship.
+% \ddag These authors also contributed equally to this work.
+
+% Current address notes
+% \textcurrency Current Address: Dept/Program/Center, Institution Name, City, State, Country % change symbol to "\textcurrency a" if more than one current address note
+% \textcurrency b Insert second current address
+% \textcurrency c Insert third current address
+
+% Use the asterisk to denote corresponding authorship and provide email address in note below.
+* jeswheel@umich.edu
+
+\end{flushleft}
 
@@
 opts_chunk$set(
-    progress = TRUE,
     concordance = TRUE,
-    prompt = TRUE,
-    highlight = FALSE,
     tidy = TRUE,
+    message = FALSE,
+    warning = FALSE,
     tidy.opts = list(
@@
         pointsize = 9
-    )
+    ),
+    fig.path = "figure/"
 )
@@
 
-\begin{frontmatter}
-
-\title{Informing policy via dynamic models: Cholera in Haiti}
-\runtitle{Cholera in Haiti}
-
-\begin{aug}
-
-\author[A]{\fnms{Jesse}~\snm{Wheeler}\ead[label=e1]{jeswheel@umich.edu}\orcid{0000-0003-3941-3884}},
-\author[B]{\fnms{AnnaElaine L.}~\snm{Rosengart}\ead[label=e2]{arosenga@andrew.cmu.edu}}
-\author[A]{\fnms{Zhuoxun}~\snm{Jiang}\ead[label=e3]{zhuoxunj@umich.edu}},
-\author[A]{\fnms{Kevin}~\snm{Tan}\ead[label=e4]{kevtan@umich.edu}},
-\author[A]{\fnms{Noah}~\snm{Treutle}\ead[label=e5]{ntreutle@umich.edu}}
-\and
-\author[A]{\fnms{Edward L.}~\snm{Ionides}\ead[label=e6]{ionides@umich.edu}}
-
-\address[A]{Statistics Department, University of Michigan\printead[presep={,\ }]{e1,e3,e4,e5,e6}}
-
-\address[B]{Statistics \& Data Science Department, Carnegie Mellon University\printead[presep={,\ }]{e2}}
-
-\end{aug}
-
-\begin{abstract}
-% \begin{center}
-% Working draft, \today
-% \end{center}
-% \vspace{3mm}
-
+\section*{Abstract}
 Public health decisions must be made about when and how to implement interventions to control an infectious disease epidemic.
@@
 Our workflow is reproducible and extendable, facilitating future investigations of this disease system.
-\end{abstract}
-
-\begin{keyword}
-  \kwd{Partially observed Markov process}
-  \kwd{Hidden Markov model}
-  \kwd{infectious disease}
-  \kwd{cholera}
-  \kwd{sequential Monte Carlo}
-\end{keyword}
-
-\end{frontmatter}
-
-\section{Introduction}
+
+\section*{Author summary}
+Quantitative understanding of infectious disease transmission dynamics relies upon mathematical models informed by scientific knowledge and relevant data.
+The models aim to provide a statistical description of the trajectory of an epidemic and its uncertainty, together with a representation of the underlying biological mechanisms.
+Evaluation of success at these goals is necessary in order for a model to provide a reliable tool for guiding evidence-based public policy interventions.
+In this article, we discuss common modeling decisions that may not provide an adequate statistical explanation of the data, and we show that attention to statistical fit can lead to more accurate policy evaluations.
+We do this by conducting a re-analysis of the 2012-2019 cholera outbreak in Haiti.
+Our analysis presents methodology for diagnosing how well a model describes observed data, and provides careful justification of valid conclusions from the fitted model.
+Objective measures are used to benchmark model fit; when these are combined with reproducibility, a framework emerges for continual improvment when revisiting the data and models.
+Our data analysis workflow is supported by recent advances in algorithms, software and hardware, which facilitate statistical fitting of nonlinear stochastic dynamic models to observed incidence data.
+However, inference for high-dimensional systems remains a methodological challenge.
+One of the models under consideration involves spatially coupled stochastic meta-populations, and we demonstrate how a recently developed algorithm permits likelihood-based inference and model diagnostics in this setting.
+We contend that raising the currently accepted standards of infectious disease modeling will result in a greater ability of scientists and policy makers to understand and respond to future infectious disease outbreaks.
+
+\linenumbers
+
+\section*{Introduction}
 
@@
 This case study investigates the use of dynamic models and spatiotemporal data to inform a policy decision in the context of the cholera outbreak in Haiti, which started in 2010.
-We build on a multi-group modeling exercise by \citet{lee20} in which four expert modeling teams developed models to the same dataset with the goal of comparing conclusions on the feasibility of eliminating cholera by a vaccination campaign.
+We build on a multi-group modeling exercise by Lee et al.~\cite{lee20} in which four expert modeling teams developed models to the same dataset with the goal of comparing conclusions on the feasibility of eliminating cholera by a vaccination campaign.
+The data that were used for this study, and that is the subject of our reanalysis, is displayed in Fig.~\ref{fig:CholeraData}.
 Model~1 is stochastic and describes cholera at the national level;
@@
 Model~4 has an agent-based construction, featuring considerable mechanistic detail but limited ability to calibrate these details to data.
-We focus on Models~1--3, as the strengths and weaknesses of the agent-based modeling approach  \citep{tracy18} are outside the scope of this article.
-
-The four independent teams were given the task of estimating the
-potential effect of prospective oral cholera vaccine (OCV) programs.
-While OCV is accepted as a safe and effective tool for controlling the spread of cholera, the global stockpile of OCV doses remains limited \citep{pezzoli20}.
-Advances in OCV technology and vaccine availability, however, raised the possibility of planning a national vaccination program \citep{lee20}.
-In the study, certain data were shared between the groups, including demography and vaccination history; vaccine efficacy was also fixed at a shared value between groups.
-Beyond this, the groups made autonomous decisions on what to include and exclude from their models.
-The groups largely adhered to existing guidelines on creating models to inform policy \citep{behrend20,saltelli20} and, despite their autonomy, obtained a consensus that an extensive nationwide vaccination campaign would be necessary to eliminate cholera from Haiti.
-Their conclusion is inconsistent with the prolonged period with no confirmed cholera cases between February, 2019 and September, 2022 \citep{trevisin22}, though cholera has recently reemerged in Haiti \citep{rubin22}.
-
-The discrepancy between the model based conclusions of \cite{lee20} and the prolonged absence of cholera in Haiti has been debated \citep{francois20,rebaudetComment20,henrys20,leeReply20}.
-Suggested origins of this discrepancy include the use of unrealistic models \citep{rebaudetComment20} and unrealistic criteria for cholera elimination \citep{henrys20}.
-We find a more nuanced conclusion: attention to methodological details in model fitting, diagnosis and forecasting can improve each of the proposed model's ability to quantitatively describe observed data.
-This improved ability may result in more accurate forecasts and facilitates the exploration of model assumptions.
-Based on this retrospective analysis, we offer suggestions on fitting mechanistic models to dynamic systems for future studies.
-
-While this work is focused on the models proposed by \cite{lee20}, our suggestions have broader relevance.
-To investigate the extent to which \citet{lee20} is typical of the substantial body of work on Haiti cholera dynamics, we preformed a literature review by searching PubMed with keywords: Haiti, cholera, model.
-The search resulted in 66 papers, of which 32 used dynamic models to describe the cholera epidemic in Haiti.
-The models make various choices on the dichotomies considered by \citet{lee20}; deterministic versus stochastic; compartment model versus agent-based; aggregated versus spatially explicit.
-This is no accident, since \citet{lee20} purposefully designed their study to cover the range of current modeling practice.
-
-We proceed by introducing Models~1--3 in Sec.~\ref{sec:models};
-in Sec.~\ref{sec:methods}, we present a systematic approach to examining and refining these models and then use improved model fits to project cholera incidence in Haiti under various vaccination scenarios.
-This is followed by a consideration of the robustness of model based policy recommendations in Sec.~\ref{sec:advice}, and a discussion in Sec.~\ref{sec:discussion}.
-
-\begin{figure}
+We focus on Models~1--3, as the strengths and weaknesses of the agent-based modeling approach \cite{tracy18} are outside the scope of this article.
+
+\begin{figure}[!h]
 <<Plot_Reported_Cases, echo=FALSE, fig.height=3.5>>=
@@
 @
-\caption{\label{CholeraData}
-\small
-Weekly reported cholera cases in Haiti from October 2010 to January 2019.
+\caption{\label{fig:CholeraData}
+{\figTitle Weekly cholera cases.}
+Weekly reported cholera cases in Haiti from October 2010 to January 2019 for each of the 10 administrative departments.
 }
@@
 
-\section{Mechanistic models for cholera in Haiti}\label{sec:models}
+The four independent teams were given the task of estimating the
+potential effect of prospective oral cholera vaccine (OCV) programs.
+While OCV is accepted as a safe and effective tool for controlling the spread of cholera, the global stockpile of OCV doses remains limited \cite{pezzoli20}.
+Advances in OCV technology and vaccine availability, however, raised the possibility of planning a national vaccination program \cite{lee20}.
+In the study, certain data were shared between the groups, including demography and vaccination history; vaccine efficacy was also fixed at a shared value between groups.
+Beyond this, the groups made autonomous decisions on what to include and exclude from their models.
+The groups largely adhered to existing guidelines for creating models to inform policy \cite{behrend20,saltelli20} and, despite their autonomy, obtained a consensus that an extensive nationwide vaccination campaign would be necessary to eliminate cholera from Haiti.
+Their forecasts, however, were inconsistent with the prolonged period with no confirmed cholera cases between February, 2019 and September, 2022 \cite{trevisin22}.
+Though cholera has recently reemerged in Haiti \cite{rubin22}, the inability to accurately forecast cholera incidence from 2019-2022---while diligently adhering to existing scientific standards for infectious disease modeling---demonstrates that existing standards of disease modeling and forecasting may be insufficient.
+
+The discrepancy between the model based conclusions of Lee et al.~\cite{lee20} and the prolonged absence of cholera in Haiti has been debated \cite{francois20,rebaudetComment20,henrys20,leeReply20}.
+Suggested origins of this discrepancy include the use of unrealistic models \cite{rebaudetComment20} and unrealistic criteria for cholera elimination \cite{henrys20}.
+We find a more nuanced conclusion: attention to methodological details in model fitting, diagnosis and forecasting can improve each of the proposed model's ability to quantitatively describe observed data.
+This improved ability may result in more accurate forecasts and facilitates the exploration of model assumptions.
+Based on this retrospective analysis, we offer suggestions on fitting mechanistic models to dynamic systems for future studies.
+
+While this work is focused on the models proposed by Lee et al.~\cite{lee20}, our suggestions have broader relevance.
+To investigate the extent to which \cite{lee20} is typical of the substantial body of work on Haiti cholera dynamics, we preformed a literature review by searching PubMed with keywords: Haiti, cholera, model.
+The search resulted in 66 papers, of which 32 used dynamic models to describe the cholera epidemic in Haiti.
+The models make various choices on the dichotomies considered by Lee et al.~\cite{lee20}; deterministic versus stochastic; compartment model versus agent-based; aggregated versus spatially explicit.
+
+We proceed by introducing the general modeling scheme employed by Models~1--3 and provide details of each individual model;
+we then describe how each model is calibrated to data, and present a systematic approach to examining and refining these models.
+We then use improved model fits to project cholera incidence in Haiti under various vaccination scenarios considered by Lee et al.~\cite{lee20}.
+Finally, we conclude with a discussion of the results.
+
+<<FitModel1, echo=FALSE, message=FALSE, include=FALSE>>=
+
+# This code chunk is used to fit Model 1 using the IF2 algorithm.
+set.seed(636813)
+
+# Check RUN_LEVEL, and set hyper-parameters to IF2 and pfilter accordingly.
+#                      RUN_LEVEL =    1,    2,      3
+NP_H1        <- switch(RUN_LEVEL,   100, 1000,   2000)
+NMIF_H1      <- switch(RUN_LEVEL,    10,   50,    200)
+NUM_TREND    <- switch(RUN_LEVEL,     5,   10,     90)
+NPROF        <- switch(RUN_LEVEL,     4,    4,     24)
+NP_EVAL      <- switch(RUN_LEVEL,   200, 1000,   5000)
+NREPS_EVAL   <- switch(RUN_LEVEL, cores,   20,  cores)
+
+# Create the <RUN_LEVEL> version of the results and save them.
+h1_MIF <- bake(
+  file = paste0("model1/", rl_dir, "haiti1_fit.rds"), {
+    fit_haiti1(  # haitipkg function used to fit Model 1
+      NP = NP_H1,
+      NMIF = NMIF_H1,
+      NUM_TREND = NUM_TREND,
+      NPROF = NPROF,
+      NREPS_EVAL = NREPS_EVAL,
+      NP_EVAL = NP_EVAL
+    )
+  }
+)
+
+# The result is a data.frame containing many parameter values and their
+# corresponding likelihood evaluation. To perform a likelihood profile,
+# we calculate the maximum likelihood for each value of betat.
+h1_prof_res <- h1_MIF %>%
+  group_by(betat) %>%
+  summarize(logLik = max(ll))
+
+mcap_results <- pomp::mcap(h1_prof_res$logLik, h1_prof_res$betat, span = 1)
+
+# Get the best set of parameters from MIF2 search.
+p1 <- h1_MIF %>%
+  filter(ll == max(ll)) %>%
+  select(-ll, -ll.se) %>%
+  unlist()
+
+# Save the likelihood of the best parameters, for use later.
+mod1_ll <- h1_MIF %>%
+  filter(ll == max(ll)) %>%
+  select(ll) %>%
+  unlist()
+
+# Just a bit of tidying up: removing parameters that we won't use again.
+rm(NP_H1, NMIF_H1, NUM_TREND, NPROF, NREPS_EVAL, NP_EVAL, h1_MIF)
+gc()
+
+@
+
+<<FitModel2, echo=FALSE, include=FALSE, message=FALSE>>=
+
+# This code chunk fits Model 2. This is a quick caclulation compared to models
+# 1 and 3, so no run_level is used.
+
+h2_fit <- bake(
+  file = paste0("model2/model2_fit.rds"), {
+    fit_haiti2()
+  },
+  timing = FALSE
+)
+@
+
+<<FitModel3, echo=FALSE, include=FALSE, message=FALSE>>=
+
+# This code chunk is used to fit Model 3 using the IBPF algorithm.
+set.seed(8636712)
+
+# Create vectors for the unit and shared parameters
+unit_specific_names <- c("betaB", "foi_add", "aHur", "hHur", "Iinit")
+shared_param_names <- c(
+  "mu_B", "XthetaA", "thetaI", "lambdaR", "r", "std_W",
+  "epsilon", "k"
+)
+est_param_names <- c(
+  unit_specific_names, shared_param_names
+)
+
+# Add unit numbers to each parameter
+est_param_names_expanded <- paste0(rep(est_param_names, each = 10), 1:10)
+
+# Simple function that is used to set the rw_sd for the first search
+# This function is convenient so that we don't have to type out the rw_sd
+# for each unit parameter.
+set_rw_1 <- function(x) {
+  if (gsub("[[:digit:]]+$", "", x) %in% shared_param_names) {
+    return(0.01)
+  } else if (x %in% paste0(rep(c("aHur", 'hHur'), each = 2), c(3, 9))) {
+    return(expression(ifelse(time >= 2016.754 & time <= 2017, 0.015, 0)))
+  } else if (x %in% c("Iinit3", "Iinit4")) {
+    return(expression(ivp(0.25)))
+  } else if (grepl('^foi_add[[:digit:]]+$', x)) {
+    return(0.015)
+  } else if (grepl('^betaB[[:digit:]]+$', x)) {
+    return(0.01)
+  } else {
+    return(0)
+  }
+}
+
+set_rw_2 <- function(x) {
+  if (gsub("[[:digit:]]+$", "", x) %in% shared_param_names) {
+    return(0.003)
+  } else if (x %in% paste0(rep(c("aHur", 'hHur'), each = 2), c(3, 9))) {
+    return(expression(ifelse(time >= 2016.754 & time <= 2017, 0.005, 0)))
+  } else if (x %in% c("Iinit3", "Iinit4")) {
+    return(expression(ivp(0.125)))
+  } else if (grepl('^foi_add[[:digit:]]+$', x)) {
+    return(0.005)
+  } else if (grepl('^betaB[[:digit:]]+$', x)) {
+    return(0.003)
+  } else {
+    return(0)
+  }
+}
+
+# First Search RW
+reg_rw_1.sd <- lapply(est_param_names_expanded, set_rw_1)
+names(reg_rw_1.sd) <- est_param_names_expanded
+chol_rw_1.sd <- do.call(rw_sd, reg_rw_1.sd)
+
+# Second Search RW
+reg_rw_2.sd <- lapply(est_param_names_expanded, set_rw_2)
+names(reg_rw_2.sd) <- est_param_names_expanded
+chol_rw_2.sd <- do.call(rw_sd, reg_rw_2.sd)
+
+if (RUN_LEVEL == 1){
+  SEARCH1 = list(
+    NBPF = 10,
+    NP = 200,
+    SPAT_REGRESSION = 0.05,
+    NREPS = 20,
+    NP_EVAL = 200,
+    NREPS_EVAL = 3,
+    RW_SD = chol_rw_1.sd,
+    COOLING = 0.5,
+    KEEP_TRACES = TRUE
+  )
+
+  SEARCH2 = list(
+    TOP_N = 8,
+    NBPF = 10,
+    NP = 200,
+    SPAT_REGRESSION = 0.05,
+    NREPS = 2,
+    NP_EVAL = 200,
+    NREPS_EVAL = 3,
+    RW_SD = chol_rw_2.sd,
+    COOLING = 0.5,
+    KEEP_TRACES = TRUE
+  )
+
+} else if (RUN_LEVEL == 2) {
+  SEARCH1 = list(
+    NBPF = 100,
+    NP = 1000,
+    SPAT_REGRESSION = 0.05,
+    NREPS = 72,
+    NP_EVAL = 1000,
+    NREPS_EVAL = 8,
+    RW_SD = chol_rw_1.sd,
+    COOLING = 0.5,
+    KEEP_TRACES = TRUE,
+    KEEP_LIK_MAT = TRUE
+  )
+
+  SEARCH2 = list(
+    TOP_N = 9,
+    NBPF = 100,
+    NP = 1000,
+    SPAT_REGRESSION = 0.05,
+    NREPS = 4,
+    NP_EVAL = 1000,
+    NREPS_EVAL = 9,
+    RW_SD = chol_rw_2.sd,
+    COOLING = 0.5,
+    KEEP_TRACES = TRUE,
+    KEEP_LIK_MAT = TRUE
+  )
+
+} else if (RUN_LEVEL == 3) {
+
+  SEARCH1 = list(
+    NBPF = 100,
+    NP = 2500,
+    SPAT_REGRESSION = 0.05,
+    NREPS = 144,
+    NP_EVAL = 2500,
+    NREPS_EVAL = 18,
+    RW_SD = chol_rw_1.sd,
+    COOLING = 0.5,
+    KEEP_TRACES = FALSE,
+    KEEP_LIKE_MAT = FALSE
+  )
+
+  SEARCH2 = list(
+    TOP_N = 12,
+    NBPF = 100,
+    NP = 2500,
+    SPAT_REGRESSION = 0.05,
+    NREPS = 6,
+    NP_EVAL = 2500,
+    NREPS_EVAL = 18,
+    RW_SD = chol_rw_2.sd,
+    COOLING = 0.5,
+    KEEP_TRACES = FALSE,
+    KEEP_LIK_MAT = TRUE
+  )
+}
+
+# Calculate the number of parameters that are estimated:
+# This is calculated by the number of non-zero rw_sd values (including expressions), but then subtracting of repeated shared parameters.
+mod3_n_params <- sum(reg_rw_1.sd != 0, na.rm = TRUE) + sum(sapply(reg_rw_1.sd, is.expression)) - length(shared_param_names) * 9
+
+# Fit Model 3
+haiti3_fit <- bake(
+  file = paste0("model3/", rl_dir, "haiti3_fit.rds"), {
+    fit_haiti3(
+      nsearches = 2L,
+      search1 = SEARCH1,
+      search2 = SEARCH2,
+      search_rho = FALSE,
+      search_gamma = FALSE,
+      search_Iinit = TRUE,
+      search_hur = TRUE
+    )
+  }
+)
+
+# haiti3_fit contains ll estimate for many parameter sets, so we want to find
+# the highest value
+mod3_ll <- max(
+  haiti3_fit[[length(haiti3_fit)]]$logLiks$logLik,
+  na.rm = TRUE
+)
+
+# Just a bit of tidying up: removing parameters that we won't use again.
+rm(
+  SEARCH1, SEARCH2, N_SEARCHES,
+  est_param_names, est_param_names_expanded,
+  set_rw_1, set_rw_2, set_rw_3, reg_rw_1.sd, reg_rw_2.sd, reg_rw_3.sd,
+  chol_rw_1.sd, chol_rw_2.sd, chol_rw_3.sd
+)
+gc()
+
+@
+
+<<table-input,echo=FALSE,eval=T, include=FALSE, message=FALSE>>=
+
+# Find the index for parameters of model 3 that correspond to the MLE.
+best_m3 <- which.max(haiti3_fit[[length(haiti3_fit)]]$logLiks$logLik)
+
+# Save the parameters that correspond to the MLE of Model 3.
+p3 <- haiti3_fit[[length(haiti3_fit)]]$params[best_m3, ]
+
+p3u <- matrix(nrow = length(unit_specific_names), ncol = 10)
+rownames(p3u) <- unit_specific_names
+p3s <- numeric(length(shared_param_names))
+names(p3s) <- shared_param_names
+
+for (p in shared_param_names) {
+  p3s[p] <- p3[paste0(p, 1)]  # They are the same for 1:10
+}
+
+for (p in unit_specific_names) {
+  pat <- paste0("^", p, "[[:digit:]]+$")
+  p3u[p, ] <- p3[grepl(pat, names(p3))]
+}
+
+rm(unit_specific_names, shared_param_names, p, pat)
+
+stew(file = "models.rda", {
+  h1 <- haiti1_joint()
+  coef(h1) <- p1
+
+  h2_epi <- haiti2(region = 'before', cutoff = 10000, measure = 'log')
+  coef(h2_epi) <- h2_fit$h2_params
+
+  h3Spat <- haiti3_spatPomp()
+  coef(h3Spat) <- p3
+})
+@
+
+<<AICtableInput, echo=FALSE, eval=TRUE>>=
+NREPS_EVAL <- switch(RUN_LEVEL,   3,   8,    36)
+NP_EVAL    <- switch(RUN_LEVEL, 100, 500,  5000)
+
+# Objective function of model 2, epidemic phase
+epi_ofun <- traj_objfun(
+  h2_epi,
+  params = h2_fit$h2_params
+)
+
+mod2_epi_ll <- -epi_ofun(par = h2_fit$h2_params)
+mod2_ll <- mod2_epi_ll - sum(log(h2_epi@data + 1), na.rm = TRUE)
+
+arima_lik2 <- 0
+
+# model 2 AIC
+mod2_n_params <- h2_fit$n_fit_params
+mod2_aic <- 2 * mod2_n_params - 2 * mod2_ll
+
+# model 1 AIC
+mod1_n_params <- length(
+  c('beta1', 'beta2', 'beta3', 'beta4', 'beta5', 'beta6',
+    'rho', 'nu', 'E_0', 'I_0', 'tau_epi',
+    'tau_end', 'sig_sq_epi', 'sig_sq_end', 'betat')
+)
+mod1_aic <- 2 * mod1_n_params - 2 * mod1_ll
+
+# model 3 AIC
+mod3_aic <- 2 * mod3_n_params - 2 * mod3_ll
+
+# Calculate likelihood via block-particle filter
+mod3_eval_res <- bake(
+  file = paste0('model3/', rl_dir, 'mod3_eval_res.rds'), {
+
+    # Create a list to store results
+    results <- list()
+
+    coef(h3Spat) <- p3
+
+    # Perform block-particle filter
+    h3_bpf <- foreach(i = 1:NREPS_EVAL, .combine = c) %dopar% {
+      bpfilter(h3Spat, Np = NP_EVAL, block_size = 1)
+    }
+
+    ll <- logLik(h3_bpf)
+    results$ll <- logmeanexp(ll[!is.na(ll)], se = FALSE)
+
+    ### Get likelihood for subset of the data
+
+    # Find appropriate sub-set
+    in_subset_cols <- h3_bpf[[1]]@times >= as.numeric(lubridate::decimal_date(as.Date("2014-03-01")))
+    in_subset_rows <- h3_bpf[[1]]@unit_names != 'Ouest'
+
+    mod3_coupled_subset_evals <- c()
+    for (i in 1:length(h3_bpf)) {
+      mod3_coupled_subset_evals <- c(
+        mod3_coupled_subset_evals,
+        sum(h3_bpf[[i]]@block.cond.loglik[in_subset_rows, in_subset_cols])
+      )
+    }
+
+    results$subset_ll <- logmeanexp(mod3_coupled_subset_evals)
+    results
+  },
+  timing = FALSE
+)
+
+mod3_eval_res$subset_aic <- 2 * mod3_n_params - 2 * mod3_eval_res$subset_ll
+
+rm(NREPS_EVAL, NP_EVAL)
+@
+
+<<CalculateBenchLikelihood, echo=FALSE, message=FALSE, include=FALSE, cache=TRUE>>=
+
+benchmarks <- list()
+
+###
+### Model 1 ###
+###
+
+# Load aggregated data
+m1_agg_data <- haiti1_agg_data()
+
+# Fit Model 1 bechmark:
+m1_bench <- ar1_NegBinom(m1_agg_data$cases, init = 7364)
+benchmarks[['m1']] <- m1_bench$ll
+
+rm(m1_agg_data, m1_bench)
+gc()
+
+###
+### Model 2 ###
+###
+
+h2_data <- haiti2_data(recreate = FALSE)
+
+# This are initial values for Y_u(0). They were estimated, but because they are
+# integers, close to the value of the first observation, it was easiest to fit these
+# by hand.
+m2_inits <- c(6449, 542, 0, 0, 16, 0, 0, 522, 0, 0)  # fit by hand
+
+# model 2 ARIMA benchmark
+h2_dep_names <- h2_data$department %>% unique()
+m2_benchmark <- 0
+for (i in 1:10){
+  cases <- h2_data[h2_data$department == h2_dep_names[i], ]$cases
+  m2_bench <- ar1_NegBinom(data = cases, init = m2_inits[i])
+
+  m2_benchmark <- m2_benchmark + m2_bench$ll
+}
+
+benchmarks[['m2']] <- m2_benchmark
+
+rm(m2_benchmark, h2_dep_names, cases, h2_data, m2_bench, i)
+gc()
+
+###
+### Model 3 ###
+###
+
+mod3_cases <- as.data.frame(t(h3Spat@data))
+m3_inits <- c(7362, 1914, 22, 8, 2818, 82, 2499, 6953, 44, 7) # fit by hand
+
+m3_benchmark <- 0
+for (dep in 1:10) {
+
+  # Get department cases
+  cases <- as.numeric(mod3_cases[, dep])
+
+  m3_benchmark <- m3_benchmark + ar1_NegBinom(data = cases, init = m3_inits[dep])$ll
+}
+
+benchmarks[['m3']] <- m3_benchmark
+rm(cases, m3_benchmark, m3_bench, dep, mod3_cases)
+gc()
+@
+
+<<AIC Lee et al, echo=FALSE, include=FALSE, message=FALSE>>=
+lee3_NP   <- switch(RUN_LEVEL, 50, 250, 5000)
+lee3_NREP <- switch(RUN_LEVEL,  3,   8,   36)
+
+mod1_lee_n_params <- 20
+mod1_lee_aic <- 2 * mod1_lee_n_params - 2 * mod1_lee_ll
+
+mod2_lee_aic <- 2 * h2_fit$leeFit_n_params - 2 * h2_fit$leeFit_ll
+
+registerDoRNG(49637421)
+
+# Create Lee et al (2020a) version of model 3 on sub-data
+lee3_sub <- lee3_spatPomp()
+
+lee3_all <- lee3_spatPomp(start_date = "2010-10-23")
+
+lee3_results <- bake(
+  file = paste0('model3/', rl_dir, 'lee3_results.rds'), {
+
+    # Create lists to save results
+    lee3_results <- list()
+
+    # Save how long it takes for sub-data
+    lee3_results[["sub_bpf_time"]] <- system.time({
+
+      lee3_sub_ll_evals <- foreach(
+        i = 1:lee3_NREP, .combine = rbind
+      ) %dopar% {
+
+        liks <- numeric(10L)
+
+        lee_bpf_out <- bpfilter(lee3_sub, block_size = 1, Np = lee3_NP)
+        Ouest_ind <- which(lee_bpf_out@unit_names == "Ouest")
+
+        for (u in 1:10) {
+          if (u == Ouest_ind) {
+            liks[u] <- sum(lee_bpf_out@block.cond.loglik[u, lee3_sub@times >= lubridate::decimal_date(as.Date('2017-06-10'))])
+          } else {
+            liks[u] <- sum(lee_bpf_out@block.cond.loglik[u, ])
+          }
+        }
+
+        liks
+      }
+    })
+
+    # Save all likelihood evaluations
+    lee3_results[["sub_ll_eval"]] <- logmeanexp(rowSums(lee3_sub_ll_evals)[!is.na(rowSums(lee3_sub_ll_evals))])
+
+    registerDoRNG(2789131)
+
+    # Save how long it takes for all data
+    lee3_results[["all_pf_time"]] <- system.time({
+      # Save all likelihood evaluations
+
+      lee3_all_ll_evals <- foreach(
+        i = 1:lee3_NREP, .combine = c
+      ) %dopar% {
+        logLik(bpfilter(lee3_all, block_size = 1, Np = lee3_NP))  # Get likelihood evaluations
+      }
+
+      lee3_results[["all_ll_eval"]] <- logmeanexp(lee3_all_ll_evals[!is.na(lee3_all_ll_evals)])
+
+    })
+
+    lee3_results[["n_params"]] <- 29
+
+    lee3_results
+  },
+  timing = FALSE
+)
+
+rm(lee3_sub, lee3_all, lee3_NP, lee3_NREP)
+gc()
+
+mod3_lee_coupled_aic <- 2 * lee3_results$n_params - 2 * lee3_results$all_ll_eval
+
+mod3_lee_coupled_partial_aic <- 2 * lee3_results$n_params - 2 * lee3_results$sub_ll_eval
+@
+
+\section*{Materials and methods}
+\subsection*{Mechanistic models for disease modeling}
 Models that focus on learning relationships between variables in a dataset are called {\it associative}, whereas models that incorporate a known scientific property of the system are called {\it causal} or {\it mechanistic}.
 The danger in using forecasting techniques which rely on associative models to predict the consequence of interventions is called the Lucas critique in an econometric context.
-\citet{lucas76} pointed out that it is naive to predict the effects of an intervention on a given system based entirely on historical associations.
+Lucas et al.~\cite{lucas76} pointed out that it is naive to predict the effects of an intervention on a given system based entirely on historical associations.
 To successfully predict the effect of an intervention, a model should therefore both provide a quantitative explanation of existing data and should have a causal interpretation: a manipulation of the system should correspond quantitatively with the corresponding change to the model.
@@
 
-The four mechanistic models of \citet{lee20} were deliberately developed with limited coordination.
+The four mechanistic models of Lee et al.~\cite{lee20} were deliberately developed with limited coordination.
 This allows us to treat the models as fairly independently developed expert approaches to understanding cholera transmission.
 However, it led to differences in notation, and in subsets of the data chosen for analysis, that hinder direct comparison.
-Here, we have put all three models into a common notational framework.
-Translations back to the original notation of \citet{lee20} are given in Table~S-1 of the supplement.
+Here, we have created a common notational framework that facilitates model comparison, and put all comparable model parameters---including parameters that were estimated or held constant---into Table~\ref{tab:params}.
+Translations back to the original notation of Lee et al.~\cite{lee20} are given in the supplement (\nameref{S1_Table}).
+
+\begin{table}
+\begin{adjustwidth}{-2.25in}{0in}
+\centering
+\caption{\label{tab:params} {\bf Model parameters.}}
+\begin{tabular}{|p{0.15\linewidth}|p{0.14\linewidth}t|p{0.13\linewidth}t|p{0.17\linewidth}t|}
+\hline
+Mechanism & Model 1 && Model 2 && Model 3 &
+\\
+\hline
+\hline
+Infection (day)
+  & ${\muIR^{-1}=\Sexpr{myround(7/p1["gamma"],1)}}\fixed$
+  & \eqref{model1:toR}
+  & ${\muIR^{-1}=\Sexpr{myround(365/h2_fit$h2_params["gamma"],1)}}\fixed $
+  & \eqref{model2:mu_IR}
+  & ${\muIR^{-1}=\Sexpr{myround(365/p3["gamma1"],1)}}\fixed $
+  & \eqref{eq:model3:IR}
+\\
+Latency (day)
+  & ${\muEI^{-1}=\Sexpr{myround(7/p1["sigma"],1)}}\fixed$
+  & \eqref{model1:EA}
+  & ${\muEI^{-1}=\Sexpr{myround(365/h2_fit$h2_params["gammaE"],1)}}\fixed$
+  & \eqref{model2:mu_EI}
+  & \missing
+  &
+\\
+\hline
+\multirow{3}{*}{Seasonality}
+  & $\transmission_{1:6}=(
+      \Sexpr{myround(p1["beta1"],1)},\allowbreak
+      \Sexpr{myround(p1["beta2"],1)},\allowbreak
+      \Sexpr{myround(p1["beta3"],1)},\allowbreak
+      \Sexpr{myround(p1["beta4"],1)},\allowbreak
+      \Sexpr{myround(p1["beta5"],1)},\allowbreak
+      \Sexpr{myround(p1["beta6"],1)})$
+  & \eqref{model1:beta}
+  & \multirow{2}{*}{${\seasAmplitude=0.4}\fixed$}
+  & \multirow{2}{*}{\eqref{model2:lambda}}
+  & \multirow{2}{*}{$\seasAmplitude = \Sexpr{myround(p3s["lambdaR"],2)}$}
+  & \multirow{2}{*}{\eqref{eq:model3:water}}
+  \\
+  & $\transmissionTrend = \Sexpr{myround(p1["betat"],2)}^*$
+  & \eqref{model1:betat}
+  & \multirow{-2}{*}{$\phaseParm = \Sexpr{myround(h2_fit$h2_params['phase'], 2)}^*$}
+  & \multirow{-2}{*}{\eqref{model2:lambda}}
+  & \multirow{-2}{*}{$\rainfallExponent = \Sexpr{signif(p3s["r"],3)}$}
+  & \multirow{-2}{*}{\eqref{eq:model3:water}}
+  \\\hline
+  \multirow{3}{*}{Immunity (yr)}
+  & \multirow{3}{*}{${\muRS^{-1}=\Sexpr{myround(1/p1["alpha"]/52,1)}}\fixed$}
+  & \multirow{3}{*}{\eqref{model1:RS}}
+  & $\muRS^{-1}= \Sexpr{signif(1/h2_fit$h2_params["sigma"],2)}$
+  & \eqref{model2:RS}
+  & \multirow{3}{*}{${\muRS^{-1}=\Sexpr{myround(1/p3['rho1'],1)}}\fixed$}
+  & \multirow{3}{*}{\eqref{eq:model3:RRnext}}
+\\
+& & & ${\omega_1^{-1} = \Sexpr{myround(1/h2_fit$h2_params["Omega1"],1)}}\fixed$ & \eqref{model2:omega1} & &\\
+& & & ${\omega_2^{-1} = \Sexpr{myround(1/h2_fit$h2_params["Omega2"],1)}}\fixed$ & \eqref{model2:omega2} & &\\
+\hline
+\multirow{2}{*}{Birth/death (yr)}
+  & ${\muBirth^{-1} = \Sexpr{myround(1/p1["mu"]/52,1)}}\fixed$
+    %
+  & \eqref{model1:death}
+  & \multirow{2}{*}{\missing}
+  &
+  & ${\muDeath^{-1} = \Sexpr{myround(1/p3["mu1"],1)}}\fixed$
+  & \multirow{2}{*}{\eqref{eq:model3:IS}}
+\\
+& ${\muDeath^{-1} = \Sexpr{myround(1/p1["delta"]/52,1)}}\fixed$
+& \eqref{model1:death}
+&
+&
+& ${\choleraDeath^{-1} = \Sexpr{myround(1/p3["alpha1"],2)}}\fixed$
+&
+\\
+Sympt. frac.
+  & ${\symptomFrac_z(t)=c\vaccineEfficacy^*(t-\tau_d)}\fixed$
+  & (\ref{model1:EI}-\ref{model1:EA})
+  & ${\symptomFrac=\Sexpr{myround(h2_fit$h2_params["k"],1)}}\fixed$
+  & \eqref{model2:mu_EI}
+  & ${\symptomFrac=\Sexpr{myround(p3["sigma1"],2)}}\fixed$
+  & \eqref{eq:model3:SA}
+\\
+Asympt.
+  & ${\asymptomRelativeInfect=0.05}\fixed$
+  & \eqref{model1:lambda}
+  & ${\asymptomRelativeInfect = 0.001}\fixed$
+  & \eqref{model2:lambda}
+  &  ${\asymptomRelativeInfect = 1}\fixed$
+  & \eqref{eq:model3:foi}
+\\
+infectivity
+& \missing
+&
+& ${\asymptomRelativeShed = \Sexpr{1e-7}}\fixed$
+& \eqref{model2:to_W}
+& $\asymptomRelativeShed = \Sexpr{myround(p3s["XthetaA"],3)}$
+& \eqref{eq:model3:water}
+\\
+\hline
+\multirow{4}{*}{Human to human}
+  & \multirow{4}{*}{$\transmission_{1:6}$ as above}
+  & \multirow{4}{*}{\eqref{model1:lambda}}
+  & \multirow{4}{*}{$\transmission=$\Sexpr{signif(h2_fit$h2_params["Beta"],3)}}
+  & \multirow{4}{*}{\eqref{model2:lambda}}
+  & $\transmission_{1:10}=(
+  \Sexpr{myround(p3u["foi_add",1]*1e6,2)}, \allowbreak
+	\Sexpr{myround(p3u["foi_add",2]*1e6,2)}, \allowbreak
+	\Sexpr{myround(p3u["foi_add",3]*1e6,2)}, \allowbreak
+  \Sexpr{myround(p3u["foi_add",4]*1e6,2)}, \allowbreak
+	\Sexpr{myround(p3u["foi_add",5]*1e6,2)}, \allowbreak
+	\Sexpr{myround(p3u["foi_add",6]*1e6,2)}, \allowbreak
+	\Sexpr{myround(p3u["foi_add",7]*1e6,2)}, \allowbreak
+	\Sexpr{myround(p3u["foi_add",8]*1e6,2)}, \allowbreak
+	\Sexpr{myround(p3u["foi_add",9]*1e6,2)}, \allowbreak
+	\Sexpr{myround(p3u["foi_add",10]*1e6,2)}
+	) \allowbreak \times 10^{-6}$
+  & \multirow{4}{*}{\eqref{eq:model3:foi}}
+\\
+\hline
+\multirow{5}{*}{Water to human}
+  & \multirow{5}{*}{\missing}
+  &
+  & \multirow{3}{*}{${\Wsat = \Sexpr{signif(h2_fit$h2_params["Sat"],2)}}\fixed$}
+  & \multirow{3}{*}{\eqref{model2:lambda}}
+  & $\Wbeta{_{1:10}}= (
+        \Sexpr{myround(p3u["betaB",1],2)}, \allowbreak
+        \Sexpr{myround(p3u["betaB",2],2)}, \allowbreak
+        \Sexpr{myround(p3u["betaB",3],2)}, \allowbreak
+        \Sexpr{myround(p3u["betaB",4],2)}, \allowbreak
+        \Sexpr{myround(p3u["betaB",5],2)}, \allowbreak
+        \Sexpr{myround(p3u["betaB",6],2)}, \allowbreak
+        \Sexpr{myround(p3u["betaB",7],2)}, \allowbreak
+        \Sexpr{myround(p3u["betaB",8],2)}, \allowbreak
+        \Sexpr{myround(p3u["betaB",9],2)}, \allowbreak
+        \Sexpr{myround(p3u["betaB",10],2)})$
+  & \multirow{4}{*}{\eqref{eq:model3:foi}}
+\\
+& & & \multirow{-3}{*}{$\beta_W= \Sexpr{signif(h2_fit$h2_params["BetaW"],3)}$} & \multirow{-3}{*}{\eqref{model2:lambda}} & &
+\\
+\hline
+Human to water
+  & \missing
+  &
+  & $\Wshed = $ \Sexpr{signif(h2_fit$h2_params["Mu"],3)}
+  & \eqref{model2:to_W}
+  & $\Wshed= \Sexpr{signif(p3s["thetaI"],3)}$
+  & \eqref{eq:model3:water}
+\\
+Water survival \newline (wk)
+  & \multirow{2}{*}{\missing}
+  &
+  & \multirow{2}{*}{${\Wremoval^{-1} = \Sexpr{signif(52/h2_fit$h2_params["Delta"],3)}}\fixed$}
+  & \multirow{2}{*}{\eqref{model2:from_W}}
+  & \multirow{2}{*}{$\Wremoval^{-1}=\Sexpr{myround(52/p3s["mu_B"],2)}$}
+  & \multirow{2}{*}{\eqref{eq:model3:Decay}}
+\\
+Mixing exponent
+  & $\mixExponent=\Sexpr{myround(p1["nu"],2)}$
+  & \eqref{model1:lambda}
+  & \missing
+  &
+  & \missing
+  &
+\\
+\hline
+Process noise \newline (wk\textsuperscript{1/2})
+  & $\sigmaProc =$ \newline $(\Sexpr{myround(sqrt(p1["sig_sq_epi"]),2)},
+  \Sexpr{myround(sqrt(p1["sig_sq_end"]),2)})^*$
+  & \multirow{2}{*}{\eqref{model1:lambda}}
+  & \multirow{2}{*}{\missing}
+  &
+  & \multirow{2}{*}{$\sigmaProc= \Sexpr{myround(p3s["std_W"],3)}$}
+  & \multirow{2}{*}{\eqref{eq:model3:SA}}
+\\
+Reporting rate
+  & $\reportRate=\Sexpr{myround(p1["rho"], 3)}$
+  & (S15)
+  & ${\reportRate=\Sexpr{myround(h2_fit$h2_params["Rho"],2)}}\fixed$
+  & (S16)
+  & $\reportRate=\Sexpr{myround(p3s["epsilon"],2)}$
+  & (S19)
+\\
+Observation \newline variance
+  & $\obsOverdispersion =$ \newline $(\Sexpr{myround(p1["tau_epi"],2)}, \Sexpr{round(p1["tau_end"],2)})$
+  & \multirow{2}{*}{(S15)}
+  & \multirow{2}{*}{$\obsOverdispersion = \Sexpr{signif(h2_fit$h2_params['sigma'], 3)}$}
+  & \multirow{2}{*}{(S16)}
+  & \multirow{2}{*}{$\obsOverdispersion = \Sexpr{myround(p3s["k"],2)}$}
+  & \multirow{2}{*}{(S19)}
+\\
+\hline
+\multirow{2}{*}{Initial Values} & $\Iinit = \Sexpr{round(prod(p1[c('pop_0', 'I_0')]))}$
+  &
+  & \multirow{2}{*}{\missing}
+  &
+  & \multirow{2}{*}{$\Iinit^{3,4} = (
+        \Sexpr{round(prod(p3[c('H3', 'Iinit3')]))}, \Sexpr{round(prod(p3[c('H4', 'Iinit4')]))})^*$}
+  &  \multirow{2}{*}{(S20)}
+  \\
+  & $E_{0,0} = \Sexpr{round(prod(p1[c('pop_0', 'E_0')]))}$ & & \multicolumn{1}{c}{}
+  &
+  &
+  &
+\\\hline
+\multirow{3}{*}{Hurricane}
+  & \multirow{4}{*}{\missing}
+  &
+  & \multirow{4}{*}{\missing}
+  &
+  & $\Whur{3, 9} =$\newline $(\Sexpr{myround(p3['aHur3'], 2)}, \Sexpr{myround(p3['aHur9'], 2)})^*$
+  & \multirow{2}{*}{\eqref{eq:model3:foi}}
+\\
+ \multirow{-1}{*}{Parameters} &
+  &
+  &
+  &
+  & $\hHur{3, 9} =$\newline $(\Sexpr{myround(p3['hHur3'], 2)}, \Sexpr{myround(p3['hHur9'], 2)})^*$
+  & \multirow{2}{*}{\eqref{eq:model3:foi}}
+\\\hline
+\end{tabular}
+\begin{flushleft} References to the relevant equation are given in parentheses.
+Parameters that were fixed and not calibrated using the data are indicated with $\fixed$;
+parameters that were added during our re-analysis and were not considered by Lee et al. are indicated with $^*$.
+Translations back into the notation of \cite{lee20} are given in \nameref{S1_Table}.
+\end{flushleft}
+\end{adjustwidth}
+\end{table}
 
@@
 A first-time reader may choose to skim through the rest of this section, and return later.
-Additional details about the numeric implementation of these models are provided in Sec.~S1 and S2 of the supplement.
-
-%%%%%%%%% 11111111111 %%%%%%%%%%
-
-\subsection{Model~1}
+Additional details about the numeric implementation of these models are provided in a supplemental text (\nameref{S1_Text}).
+While each of the dynamic models considered in this manuscript can be fully described using the mathematical equations provided in the following section, diagrams of dynamic systems can be helpful to understand the equations.
+For this reason, we provide flow chart diagrams for Models~1--3 in supplement figures (\nameref{S_Mod1}, \nameref{S_Mod2} and \nameref{S_Mod3}).
+
+\subsubsection*{Model~1}
 \label{sec:model1}
@@
 The process noise $d\Gamma(t)/dt$ is multiplicative Gamma-distributed white noise, with infinitesimal variance parameter $\sigmaProc^2$.
-\cite{lee20} included process noise in Model~3 but not in Model~1, i.e., they fixed $\sigmaProc^2 = 0$.
-Gamma white noise in the transmission rate gives rise to an over-dispersed latent Markov process \citep{breto11} which has been found to improve the statistical fit of disease transmission models \citep{stocks20,he10}.
-
-For any time point in $t_{1:N}$, the process model describes the movement of individuals from one latent state compartment to another.
-Per-capita transition rates are given in Equations~\ref{model1:SE}-\ref{model1:birth}:
+Lee et al.~\cite{lee20} included process noise in Model~3 but not in Model~1, i.e., they fixed $\sigmaProc^2 = 0$.
+Gamma white noise in the transmission rate gives rise to an over-dispersed latent Markov process \cite{breto11} which has been found to improve the statistical fit of disease transmission models \cite{stocks20,he10}.
+
+For any time point in $t_{1:N}$, the process model $f_{\bm{X}_n|\bm{X}_{n-1}}\big(\bm{x}_{n}|\bm{x}_{n-1}; \paramVec\big)$ is defined by the describing how individuals move from one latent state compartment to another.
+Per-capita transition rates are given in Eqs.~\myeqref{model1:SE}-\myeqref{model1:birth}:
 \begin{eqnarray}
@@
 \mu_{S_{\vaccCounter}E_{\vaccCounter}} &=& \lambda(t),
-\\[-3pt]
+\\
 \label{model1:EI}
 \mu_{E_{\vaccCounter}I_{\vaccCounter}} &=& \muEI\big(1-\vaccineEfficacy_\vaccCounter(t)\big),
-\\[-3pt]
+\\
 \label{model1:EA}
 \mu_{E_{\vaccCounter}A_{\vaccCounter}} &=& \muEI\, \vaccineEfficacy_\vaccCounter(t),
-\\[-3pt]
+\\
 \label{model1:toR}
 \mu_{I_{\vaccCounter}R_{\vaccCounter}} &=& \mu_{A_{\vaccCounter}R_{\vaccCounter}} = \muIR,
-\\[-3pt]
+\\
 \label{model1:RS}
 \mu_{R_{\vaccCounter}S_{\vaccCounter}} &=& \muRS,
-\\[-3pt]
+\\
 \label{model1:vacc}
 \mu_{S_0S_{\vaccCounter}} &=& \mu_{E_0E_{\vaccCounter}} = \mu_{I_0I_{\vaccCounter}} = \mu_{A_0A_{\vaccCounter}} = \mu_{R_0R_{\vaccCounter}} = \eta_{\vaccCounter}(t),
-\\[-3pt]
+\\
 \label{model1:death}
@@
 Conditional on infection status, vaccinated individuals are also less infectious than their non-vaccinated counterparts by a rate of $\asymptomRelativeInfect = 0.05$ in Eq.~\eqref{model1:lambda}.
-In \eqref{model1:EA} and~\eqref{model1:EI} the asymptomatic ratio for non-vaccinated individuals is set $\symptomFrac_0(t)=0$, so that the asymptomatic route is reserved for vaccinated individuals.
+In Eqs.~\myeqref{model1:EA} and~\myeqref{model1:EI} the asymptomatic ratio for non-vaccinated individuals is set $\symptomFrac_0(t)=0$, so that the asymptomatic route is reserved for vaccinated individuals.
 For $\vaccCounter\in\seq{1}{\vaccClass}$, the vaccination cohort $\vaccCounter$ is assigned a time $\tau_{\vaccCounter}$, and we take $\vaccineEfficacy_\vaccCounter(t) = c \, \vaccineEfficacy^*(t-\tau_{\vaccCounter})$
 where $\vaccineEfficacy^*(t)$ is efficacy at time $t$ since vaccination for adults, taken from \cite{lee20}, Table~S4, and $c=\big(1-(1-0.4688)\times 0.11\big)$ is a correction to allow for reduced efficacy in the 11\% of the population aged under 5 years.
-Single and double vaccine doses were modeled by changing the waning of protection; protection was assumed to be equal between single and double dose until 52 weeks after vaccination, at which point the single dose becomes ineffective.
-
-The latent state vector $\bm{X}(t)$ is initialized by setting each compartment count for each vaccination scenario $\vaccCounter \neq 0$ as zero, and introducing initial-value parameters $I_{0,0}$ and $E_{0, 0}$ such that $R_{0}(0) = 0$, $I_{0}(0) = \mathrm{Pop} \times I_{0, 0}$, $E_{0}(0) = \mathrm{Pop} \times E_{0, 0}$ and $S_{0}(0) = \mathrm{Pop} \times (1 - I_{0, 0} - E_{0, 0})$, where $\mathrm{Pop}$ is the total population of Haiti.
-The measurement model assumes reported cholera cases at time point $n$ come from a negative binomial distribution, where only a fraction ($\reportRate$) of new weekly cases are reported.
-See Sec.~S3 of the supplement material for more details.
-
-%%%%%%% 222222222222 %%%%%%%%%%
-
-\subsection{Model~2}
+Single and double vaccine doses were modeled by changing the waning of protection; protection was modeled as equal between single and double dose until 52 weeks after vaccination, at which point the single dose becomes ineffective.
+
+The latent state vector $\bm{X}(t)$ is initialized by setting the counts for each compartment and vaccination scenario $\vaccCounter \neq 0$ as zero, and introducing initial-value parameters $I_{0,0}$ and $E_{0, 0}$ such that $R_{0}(0) = 0$, $I_{0}(0) = \mathrm{Pop} \times I_{0, 0}$, $E_{0}(0) = \mathrm{Pop} \times E_{0, 0}$ and $S_{0}(0) = \mathrm{Pop} \times (1 - I_{0, 0} - E_{0, 0})$, where $\mathrm{Pop}$ is the total population of Haiti.
+The measurement model describes reported cholera cases at time point $n$ come from a negative binomial distribution, where only a fraction ($\reportRate$) of new weekly cases are reported.
+More details about the initialization model $f_{\bm{X}_0}\big(\bm{x}_0;\paramVec\big)$ and the \emph{measurement model} $f_{\bm{Y}_n|\bm{X}_{n}}\big(\bm{y}_n|\bm{x}_{n}\big)$ for Models~1--3 are provided a supplement text (\nameref{S_init} and \nameref{S_meas}).
+
+\subsubsection*{Model~2}
 \label{sec:model2}
@@
 \mu_{S_{u\vaccCounter}E_{u\vaccCounter}} &=& \vaccineEfficacy_\vaccCounter \, \lambda,
-\\[-3pt]
+\\
 \label{model2:mu_EI}
 \mu_{E_{u\vaccCounter}I_{u\vaccCounter}} &=& \symptomFrac\muEI, \quad \mu_{E_{u\vaccCounter}A_{u\vaccCounter}} = (1-\symptomFrac)\muEI,
-\\[-3pt]
+\\
 \label{model2:mu_IR}
 \mu_{I_{u\vaccCounter}R_{u\vaccCounter}} &=& \mu_{A_{u\vaccCounter}R^A_{u\vaccCounter}} = \muIR,
-\\[-3pt]
+\\
 \label{model2:RS}
 \mu_{R_{u\vaccCounter}S_{u\vaccCounter}} &=& \mu_{R^A_{u\vaccCounter}S_{u\vaccCounter}} = \muRS,
-\\[-3pt]
+\\
 \label{model2:transport}
 \mu_{S_{u\vaccCounter}S_{{\varv}\vaccCounter}} &=& \mu_{E_{u\vaccCounter}E_{{\varv}\vaccCounter}} = \mu_{I_{u\vaccCounter} I_{{\varv}\vaccCounter}} = \mu_{A_{u\vaccCounter}A_{{\varv}\vaccCounter}} = \mu_{R_{u\vaccCounter}R_{{\varv}\vaccCounter}} = \mu_{R^A_{u\vaccCounter} R^A_{{\varv}\vaccCounter}} = T_{u\varv},
-\\[-3pt]
+\\
 \label{model2:omega1}
 \mu_{S_{u1}S_{u0}} &=& \mu_{S_{u3}S_{u0}} = \omega_1,
-\\[-3pt]
+\\
 \label{model2:omega2}
 \mu_{S_{u2}S_{u0}} &=& \mu_{S_{u4}S_{u0}} = \omega_2,
-\\[-3pt]
+\\
 \label{model2:to_W}
 \mu_{\demography W_u} &=& \Wshed \left\{ \sum_{\vaccCounter=0}^4 I_{u\vaccCounter}(t) + \asymptomRelativeShed \sum_{\vaccCounter=0}^4 A_{u\vaccCounter}(t) \right\},
-\\[-3pt]
+\\
 \label{model2:from_W}
 \mu_{W_u\demography} &=& \Wremoval,
-\\[-3pt]
+\\
 \label{model2:water_transport}
@@
 $D_{u\varv}$ is a distance measure estimating average road distance between randomly chosen members of each population, and $v_{\mathrm{rate}}= 10^{-12} \,\mbox{km$^2$yr$^{-1}$}$  was treated as a fixed constant.
-In \eqref{model2:water_transport}, $T^W_{u\varv}$ is a measure of river flow between departments.
-The unit of $W_u(t)$ is cells per ml, with dose response modeled via a saturation constant of $\Wsat$ in \eqref{model2:lambda}.
-In \eqref{model2:mu_SE}, $\vaccineEfficacy_\vaccCounter$ denotes the vaccine efficacy for each vaccination campaign $\vaccCounter \in \vaccClass$, with $\vaccineEfficacy_0 = 1$, $\vaccineEfficacy_1 = 1 - 0.429 \childReduce$, $\vaccineEfficacy_2 = 1 - 0.519 \childReduce$, $\vaccineEfficacy_3 = 1 - 0.429$, and $\vaccineEfficacy_4 = 1 - 0.519$
-Here, $\childReduce = 0.4688$ represents the reduced efficacy of the vaccination for children under the age of five years, and the values $0.429$ and $0.519$ are the median effectiveness of one and two doses over their effective period respectively, according to Table S4 in the supplement material of \citet{lee20}.
+In Eq.~\myeqref{model2:water_transport}, $T^W_{u\varv}$ is a measure of river flow between departments.
+The unit of $W_u(t)$ is cells per ml, with dose response modeled via a saturation constant of $\Wsat$ in Eq.~\myeqref{model2:lambda}.
+In Eq.~\myeqref{model2:mu_SE}, $\vaccineEfficacy_\vaccCounter$ denotes the vaccine efficacy for each vaccination campaign $\vaccCounter \in \vaccClass$, with $\vaccineEfficacy_0 = 1$, $\vaccineEfficacy_1 = 1 - 0.429 \childReduce$, $\vaccineEfficacy_2 = 1 - 0.519 \childReduce$, $\vaccineEfficacy_3 = 1 - 0.429$, and $\vaccineEfficacy_4 = 1 - 0.519$
+Here, $\childReduce = 0.4688$ represents the reduced efficacy of the vaccination for children under the age of five years, and the values $0.429$ and $0.519$ are the median effectiveness of one and two doses over their effective period respectively, according to Table S4 in the supplement material of Lee et al.~\cite{lee20}.
 Because vaccine efficacy remains constant, individuals in this model transition from a vaccinated compartment to the susceptible compartment at the end of the vaccine coverage period.
@@
 The starting value for each element of the latent state vector $\bm{X}(0)$ are set to zero except for $I_{u0}(0) = y_u^*(0)/\reportRate$ and $R_{u0}(0) = \mathrm{Pop}_u - I_{u0}(0)$, where $y_u^*(0)$ is the reported number of cholera cases in department $u$ at time $t = 0$.
-Reported cases are assumed to come from a log-normal distribution, with the log-scale mean equal to the reporting rate $\reportRate$ times the number of newly infected individuals.
-See Sec.~\ref{sec:fit2} and Sec.~S3.2 of the supplement for more details.
-
-%%%%%%% 333333333 %%%%%%%%%%
-
-\subsection{Model~3}
+Reported cases are described using a log-normal distribution, with the log-scale mean equal to the reporting rate $\reportRate$ times the number of newly infected individuals.
+See the section describing how Model~2 is fit to data and Sec.~S3.2 of the supplement for more details.
+
+\subsubsection*{Model~3}
 \label{sec:model3}
@@
 
-Per-capita transition rates are given in Equations~\ref{eq:model3:SI}--\ref{eq:model3:Decay}.
+Per-capita transition rates are used for both compartments representing human counts and the aquatic reservoir of bacteria; these rates  are given in Eqs.~\myeqref{eq:model3:SI}--\myeqref{eq:model3:Decay}.
 \begin{eqnarray}
@@
 \mu_{S_{u\vaccCounter}I_{u}} &=& \symptomFrac \,  \lambda_u \big(1-\vaccineEfficacy_{u\vaccCounter}(t)\big) \, d\Gamma/dt,
-\\[-3pt]
+\\
 \label{eq:model3:SA}
 \mu_{S_{u\vaccCounter}A_{u}} &=& (1-\symptomFrac) \,  \lambda_u \big(1-\vaccineEfficacy_{u\vaccCounter}(t)\big) \,  d\Gamma/dt,
-\\[-3pt]
+\\
 \label{eq:model3:IR}
 \mu_{I_{u}R_{u\vaccCounter 1}} &=& \mu_{A_{u}R_{u\vaccCounter 1}} = \muIR,
-\\[-3pt]
+\\
 \label{eq:model3:IS}
 \mu_{I_{u}S_{u0}} &=& \muDeath + \choleraDeath, \hspace{2mm} \mu_{A_{u}S_{u0}} = \muDeath
-\\[-3pt]
+\\
 \label{eq:model3:RRnext}
 \mu_{R_{u\vaccCounter 1}R_{u\vaccCounter 2}} &=& \mu_{R_{u\vaccCounter 2}R_{u\vaccCounter 3}} = 3\muRS,
-\\[-3pt]
+\\
 \label{eq:model3:RS}
 \mu_{R_{u\vaccCounter k}S_{u0}} &=& \muDeath + 3\muRS \, \mathbf{1}_{\{k=3\}},
-\\[-3pt]
+\\
 \label{eq:model3:water}
-\mu_{{\demography}W_u} &=& \big[1 + \seasAmplitude \big(J_u(t))^r \big] \mathrm{Den}_u \, \Wshed \big[ I_{u}(t)+ \asymptomRelativeShed A_{u}(t) \big],
-\\[-3pt]
+\mu_{{\demography}W_u} &=& \big[1 + \seasAmplitude \big(J_u(t)\big)^r \big] \mathrm{Den}_u \, \Wshed \big[ I_{u}(t)+ \asymptomRelativeShed A_{u}(t) \big],
+\\
 \label{eq:model3:Decay}
@@
 Therefore, for departments with zero case counts in week 1, we consider estimating the number of infected individuals rather then treating this value as a constant (see supplement Sec.~S5 for more details).
-Finally, reported cholera cases are assumed to come from a negative binomial measurement model with mean equal to a fraction ($\reportRate$) of individuals in each unit who develop symptoms and seek healthcare
+Finally, reported cholera cases are modeled using a negative binomial measurement model with mean equal to a fraction ($\reportRate$) of individuals in each unit who develop symptoms and seek healthcare
 (see Sec.~S3.3 of the supplement).
 
-\section{Statistical Analysis}\label{sec:methods}
-
-We consider model fitting (Sec.~\ref{sec:model_fitting}) followed by diagnostic investigations (Sec.~\ref{sec:model_diagnostics}), forecasting (Sec.~\ref{sec:filter}) and external scientific corroboration of model fit (Sec.~\ref{sec:science}).
-
-<<FitModel1, echo=FALSE, message=FALSE, include=FALSE>>=
-
-# This code chunk is used to fit Model 1 using the IF2 algorithm.
-set.seed(636813)
-
-# Check RUN_LEVEL, and set hyper-parameters to IF2 and pfilter accordingly.
-#                      RUN_LEVEL =    1,    2,      3
-NP_H1        <- switch(RUN_LEVEL,   100, 1000,   2000)
-NMIF_H1      <- switch(RUN_LEVEL,    10,   50,    200)
-NUM_TREND    <- switch(RUN_LEVEL,     5,   10,     90)
-NPROF        <- switch(RUN_LEVEL,     4,    4,     24)
-NP_EVAL      <- switch(RUN_LEVEL,   200, 1000,   5000)
-NREPS_EVAL   <- switch(RUN_LEVEL, cores,   20,  cores)
-
-# Create the <RUN_LEVEL> version of the results and save them.
-  h1_MIF <- bake(
-    file = paste0("model1/", rl_dir, "haiti1_fit.rds"), {
-      fit_haiti1(  # haitipkg function used to fit Model 1
-        NP = NP_H1,
-        NMIF = NMIF_H1,
-        NUM_TREND = NUM_TREND,
-        NPROF = NPROF,
-        NREPS_EVAL = NREPS_EVAL,
-        NP_EVAL = NP_EVAL,
-        ncores = cores
-      )
-    }
-  )
-}
-
-# The result is a data.frame containing many parameter values and their
-# corresponding likelihood evaluation. To perform a likelihood profile,
-# we calculate the maximum likelihood for each value of betat.
-h1_prof_res <- h1_MIF %>%
-  group_by(betat) %>%
-  summarize(logLik = max(ll))
-
-mcap_results <- pomp::mcap(h1_prof_res$logLik, h1_prof_res$betat, span = 1)
-
-# Get the best set of parameters from MIF2 search.
-p1 <- h1_MIF %>%
-  filter(ll == max(ll)) %>%
-  select(-ll, -ll.se) %>%
-  unlist()
-
-# Save the likelihood of the best parameters, for use later.
-mod1_ll <- h1_MIF %>%
-  filter(ll == max(ll)) %>%
-  select(ll) %>%
-  unlist()
-
-# Just a bit of tidying up: removing parameters that we won't use again.
-rm(NP_H1, NMIF_H1, NUM_TREND, NPROF, NREPS_EVAL, NP_EVAL, h1_MIF)
-gc()
-
-@
-
-<<FitModel2, echo=FALSE, include=FALSE, message=FALSE>>=
-
-# This code chunk fits Model 2. This is a quick caclulation compared to models
-# 1 and 3, so no run_level is used.
-
-h2_fit <- bake(
-  file = paste0("model2/model2_fit.rds"), {
-    fit_haiti2()
-  },
-  timing = FALSE
-)
-@
-
-<<FitModel3, echo=FALSE, include=FALSE, message=FALSE>>=
-
-# This code chunk is used to fit Model 3 using the IBPF algorithm.
-set.seed(1859687)
-
-# Create vectors for the unit and shared parameters that are estimated
-unit_specific_names <- c("betaB", "foi_add", "aHur", "hHur", "Iinit")
-shared_param_names <- c(
-  "mu_B", "XthetaA", "thetaI", "lambdaR", "r", "std_W", "epsilon", "k"
-)
-
-est_param_names <- c(unit_specific_names, shared_param_names)
-
-# Add unit numbers to each parameter
-est_param_names_expanded <- paste0(rep(est_param_names, each = 10), 1:10)
-
-# Simple function that is used to set the rw_sd for the first search
-# This function is convenient so that we don't have to type out the rw_sd
-# for each unit parameter.
-set_rw_1 <- function(x) {
-  if (gsub("[[:digit:]]+$", "", x) %in% shared_param_names) {
-    return(0.01)
-  } else if (x %in% paste0(rep(c("aHur", 'hHur'), each = 2), c(3, 9))) {
-    return(expression(ifelse(time >= 2016.754 & time <= 2017, 0.015, 0)))
-  } else if (x %in% c("Iinit3", "Iinit7", "Iinit10")) {
-    return(expression(ivp(0.25)))
-  } else if (grepl('^foi_add[[:digit:]]+$', x)) {
-    return(0.015)
-  } else if (grepl('^betaB[[:digit:]]+$', x)) {
-    return(0.01)
-  } else {
-    return(0)
-  }
-}
-
-# Simple function that is used to set the rw_sd for the second search
-# This function is convenient so that we don't have to type out the rw_sd
-# for each unit parameter.
-set_rw_2 <- function(x) {
-  if (gsub("[[:digit:]]+$", "", x) %in% shared_param_names) {
-    return(0.003)
-  } else if (x %in% paste0(rep(c("aHur", 'hHur'), each = 2), c(3, 9))) {
-    return(expression(ifelse(time >= 2016.754 & time <= 2017, 0.005, 0)))
-  } else if (x %in% c("Iinit3", "Iinit7", "Iinit10")) {
-    return(expression(ivp(0.125)))
-  } else if (grepl('^foi_add[[:digit:]]+$', x)) {
-    return(0.005)
-  } else if (grepl('^betaB[[:digit:]]+$', x)) {
-    return(0.003)
-  } else {
-    return(0)
-  }
-}
-
-# Simple function that is used to set the rw_sd for the third search
-# This function is convenient so that we don't have to type out the rw_sd
-# for each unit parameter.
-set_rw_3 <- function(x) {
-  if (gsub("[[:digit:]]+$", "", x) %in% shared_param_names) {
-    return(0.0025)
-  } else if (x %in% paste0(rep(c("aHur", 'hHur'), each = 2), c(3, 9))) {
-    return(expression(ifelse(time >= 2016.754 & time <= 2017, 0.005, 0)))
-  } else if (x %in% c("Iinit3", "Iinit7", "Iinit10")) {
-    return(expression(ivp(0.125)))
-  } else if (grepl('^foi_add[[:digit:]]+$', x)) {
-    return(0.003)
-  } else if (grepl('^betaB[[:digit:]]+$', x)) {
-    return(0.0025)
-  } else {
-    return(0)
-  }
-}
-
-# First Search RW
-reg_rw_1.sd <- lapply(est_param_names_expanded, set_rw_1)
-names(reg_rw_1.sd) <- est_param_names_expanded
-chol_rw_1.sd <- do.call(rw_sd, reg_rw_1.sd)
-
-# Second Search RW
-reg_rw_2.sd <- lapply(est_param_names_expanded, set_rw_2)
-names(reg_rw_2.sd) <- est_param_names_expanded
-chol_rw_2.sd <- do.call(rw_sd, reg_rw_2.sd)
-
-# Third Search RW
-reg_rw_3.sd <- lapply(est_param_names_expanded, set_rw_3)
-names(reg_rw_3.sd) <- est_param_names_expanded
-chol_rw_3.sd <- do.call(rw_sd, reg_rw_3.sd)
-
-# Search hyper-parameters for each run-level and search. This sets
-# parameters like number of particles, number of BPF iterations, and
-# number of replications for each search.
-h3_fit_parms <- set_h3_fit_parms(
-  RUN_LEVEL = RUN_LEVEL,
-  rw_1.sd = chol_rw_1.sd,
-  rw_2.sd = chol_rw_2.sd,
-  rw_3.sd = chol_rw_3.sd
-)
-
-# Calculate the number of parameters that are estimated:
-# This is calculated by the number of non-zero rw_sd values (including expressions), but then subtracting of repeated shared parameters.
-mod3_n_params <- sum(reg_rw_1.sd != 0, na.rm = TRUE) + sum(sapply(reg_rw_1.sd, is.expression)) - length(shared_param_names) * 9
-
-# Fit Model 3
-haiti3_fit <- bake(
-  file = paste0("model3/", rl_dir, "haiti3_coupled_fit.rds"), {
-    fit_coupled_haiti3(
-      nsearches = h3_fit_parms$N_SEARCHES,
-      search1 = h3_fit_parms$SEARCH1,
-      search2 = h3_fit_parms$SEARCH2,
-      search3 = h3_fit_parms$SEARCH3,
-      search_rho = FALSE,
-      search_gamma = FALSE,
-      search_Iinit = TRUE,
-      search_hur = TRUE,
-      ncores = cores
-    )
-  }
-)
-
-# haiti3_fit contains ll estimate for many parameter sets, so we want to find
-# the highest value
-mod3_ll <- max(
-  haiti3_fit[[length(haiti3_fit)]]$logLiks$logLik,
-  na.rm = TRUE
-)
-
-# Just a bit of tidying up: removing parameters that we won't use again.
-rm(
-  SEARCH1, SEARCH2, SEARCH3, N_SEARCHES,
-  est_param_names, est_param_names_expanded,
-  set_rw_1, set_rw_2, set_rw_3, reg_rw_1.sd, reg_rw_2.sd, reg_rw_3.sd,
-  chol_rw_1.sd, chol_rw_2.sd, chol_rw_3.sd
-)
-gc()
-
-@
-
-<<table-input,echo=FALSE,eval=T, include=FALSE, message=FALSE>>=
-
-# Find the index for parameters of model 3 that correspond to the MLE.
-best_m3 <- which.max(haiti3_fit[[length(haiti3_fit)]]$logLiks$logLik)
-
-# Save the parameters that correspond to the MLE of Model 3.
-p3 <- haiti3_fit[[length(haiti3_fit)]]$params[best_m3, ]
-
-p3u <- matrix(nrow = length(unit_specific_names), ncol = 10)
-rownames(p3u) <- unit_specific_names
-p3s <- numeric(length(shared_param_names))
-names(p3s) <- shared_param_names
-
-for (p in shared_param_names) {
-  p3s[p] <- p3[paste0(p, 1)]  # They are the same for 1:10
-}
-
-for (p in unit_specific_names) {
-  pat <- paste0("^", p, "[[:digit:]]+$")
-  p3u[p, ] <- p3[grepl(pat, names(p3))]
-}
-
-rm(unit_specific_names, shared_param_names, p, pat)
-
-stew(file = "models.rda", {
-  h1 <- haiti1_joint()
-  coef(h1) <- p1
-
-  h2_epi <- haiti2(region = 'before', cutoff = 10000, measure = 'log')
-  coef(h2_epi) <- h2_fit$h2_params
-
-  h3Spat <- haiti3_spatPomp()
-  coef(h3Spat) <- p3
-})
-@
-
-\subsection{Model Fitting}\label{sec:model_fitting}
+\subsection*{Model Fitting}\label{sec:model_fitting}
 
@@
 Therefore, the choice of $\paramVec$ used to make inference about the system can greatly affect model based conclusions.
-Elements of $\paramVec$ can be fixed at a constant value based on scientific understanding of the system, but parameters can also be calibrated to data by maximizing a measure of congruency between the observed data and the assumed mechanistic structure.
-Calibrating model parameters to observed data does not guarantee that the resulting model successfully approximates real-world mechanisms, since model assumptions may be incorrect, and do not change as the model is calibrated to data.
+Elements of $\paramVec$ can be fixed at a constant value based on scientific understanding of the system, but parameters can also be calibrated to data by maximizing a measure of congruency between the observed data and the model's mechanistic structure.
+Calibrating model parameters to observed data does not guarantee that the resulting model successfully approximates real-world mechanisms, since the model description of the dynamic system may be incorrect and do not change as the model is calibrated to data.
 However, the congruency between the model and observed data serves as a proxy for the congruency between the model and the true underlying dynamic system.
 As such, it is desirable to obtain the best possible fit of the proposed mechanistic structure to the observed data.
-In the following subsections we describe our approach to calibrating the three proposed mechanistic models to observed cholera incidence data.
-
-\subsubsection{Calibrating Model~1 Parameters}\label{sec:fit1}
+
+In this article we calibrate the parameters of each of our models by maximizing the likelihood, as described in Eq.~\myeqref{eq:likedef}.
+The likelihood for each of the fitted models---and the corresponding AIC values for model comparisons that include an adjustment for the number of calibrated parameters---is provided in Table~\ref{tab:likes}
+In the following subsections we describe in detail our approach to calibrating the three proposed mechanistic models to observed cholera incidence data.
+
+\begin{table}[!h]
+\centering
+\caption{\label{tab:likes}{\bf AIC values for Models~1--3 and their benchmarks.}}
+\begin{tabular}{|c|c|c|c|}
+\hline
+ & \thead{Model~1} & \thead{Model~2} & \thead{Model~3}
+\\
+\hline
+\hline
+\multirow{2}{*}{Log-likelihood} &
+  $\Sexpr{myround(mod1_ll, 1)}$ &
+  $\Sexpr{myround(mod2_ll, 1)}$ &
+  $\Sexpr{myround(mod3_ll, 1)}$ \\
+    & ($\Sexpr{myround(mod1_lee_ll, 1)}$)\footnotemark[1] &
+  ($\Sexpr{myround(h2_fit$leeFit_ll, 1)}$) &
+  ($\Sexpr{myround(lee3_results[["all_ll_eval"]], 1)}$)\footnotemark[2]
+\\
+\hline
+Number of &
+  $\Sexpr{as.character(mod1_n_params)}$ &
+  $\Sexpr{as.character(mod2_n_params)}$ &
+  $\Sexpr{as.character(mod3_n_params)}$ \\
+ Fit Parameters &
+ (\Sexpr{as.character(mod1_lee_n_params)}) &
+ ($\Sexpr{as.character(h2_fit$leeFit_n_params)}$) &
+ ($\Sexpr{as.character(lee3_results$n_params, 1)}$)
+\\
+\hline
+\multirow{2}{*}{AIC} &
+  $\Sexpr{myround(mod1_aic, 1)}$ &
+  $\Sexpr{myround(mod2_aic, 1)}$ &
+  $\Sexpr{myround(mod3_aic, 1)}$ \\
+  & ($\Sexpr{myround(mod1_lee_aic, 1)}$)\footnotemark[1] &
+  ($\Sexpr{myround(mod2_lee_aic, 1)}$) &
+  ($\Sexpr{myround(mod3_lee_coupled_aic, 1)}$)\footnotemark[2]
+\\
+\hline
+\thead{Benchmark \\ AIC} &
+  $\Sexpr{myround(-2 * benchmarks$m1 + 2 * 4, 1)}$ &
+  $\Sexpr{myround(-2 * benchmarks$m2 + 2 * 40, 1)}$ &
+  $\Sexpr{myround(-2 * benchmarks$m3 + 2 * 40, 1)}$
+\\
+\hline
+\end{tabular}
+\begin{flushleft}
+Values in parentheses are corresponding values using Lee et al.~\cite{lee20} parameter estimates.
+\textsuperscript{1}The reported likelihood is an upper bound of the likelihood of the Lee et al.~\cite{lee20} model.
+\textsuperscript{2}Lee et al.~\cite{lee20} fit Model~3 to a subset of the data (March 2014 onward, excluding data from Ouest in 2015-2016).
+On this subset, their model has a likelihood of $\Sexpr{myround(lee3_results$sub_ll_eval, 1)}$.
+On this same subset, our model has a likelihood of $\Sexpr{myround(mod3_eval_res$subset_ll, 1)}$.
+Details of estimating the likelihood of the Lee et al.~\cite{lee20} models is provided in \nameref{S_lee20}
+See Sec.~S6 of the supplement material for more details on estimating the likelihood of the Lee et al.~\cite{lee20} models.
+\end{flushleft}
+\end{table}
+
+\subsubsection*{Calibrating Model~1 Parameters}\label{sec:fit1}
 
 One of the most frequently used approaches to calibrate the parameters of a statistical model is maximum likelihood estimation.
-In the case of a POMP model, the likelihood function---as described in Sec.~\ref{sec:models}---is written as $\mathcal{L}(\paramVec) = \int f_{\bm{X}_{0:N}, \bm{Y}_{1:N}}\big(\bm{x}_{0:N}, \bm{y}_{1:N};\paramVec\big) d\bm{x}_{0:N}$.
-The high-dimensional integral in the likelihood, and the potentially non-linear nature of the process model, makes it impossible to obtain a closed form expression of the likelihood function of POMP models, except in the most basic cases.
-For example, the process model that defines Model~1 was developed with the intent of providing a plausible description of cholera dynamics rather than a providing statistically convenient description.
+In the case of a POMP model, the likelihood function---as described in Eq.~\myeqref{eq:likedef}---is written as $\mathcal{L}(\paramVec) = \int f_{\bm{X}_{0:N}, \bm{Y}_{1:N}}\big(\bm{x}_{0:N}, \bm{y}_{1:N};\paramVec\big) d\bm{x}_{0:N}$.
+The high-dimensional integral in the likelihood, and the potentially non-linear nature of the process model, make it impossible to obtain a closed form expression of the likelihood function of POMP models except in the most simple cases.
+The process model that defines Model~1 was developed with the intent of providing a plausible description of cholera dynamics rather than a providing statistically convenient description.
 Because of this, a closed form expression of the likelihood of Model~1 is not readily available.
 
-In order to retain the ability to propose models that are scientifically meaningful rather than only those that are simply statistically convenient, we restrict ourselves to parameter estimation techniques that have the plug-and-play property, which is that the fitting procedure only requires the ability to simulate the latent process instead of evaluating transition densities \citep{breto09,he10}.
-Plug-and-play algorithms include Bayesian approaches like ABC and PMCMC \citep{toni09,andrieu10}, but here we use algorithms that enable maximum likelihood estimation.
-To our knowledge, the only plug-and-play methods that can maximize the likelihood for POMP models of this complexity are iterated filtering algorithms \citep{ionides15}, which modify the well-known \emph{particle filter} \citep{arulampalam02}.
+In order to retain the ability to propose models that are scientifically meaningful rather than only those that are simply statistically convenient, we restrict ourselves to parameter estimation techniques that have the plug-and-play property, which is that the fitting procedure only requires the ability to simulate the latent process instead of evaluating transition densities \cite{breto09,he10};
+in the context of the notation and definitions employed in this article, this means that we only require the ability to simulate from $f_{\bm{X}_0}\big(\bm{x}_0;\paramVec\big)$ and $f_{\bm{X}_n|\bm{X}_{n-1}}\big(\bm{x}_{n}|\bm{x}_{n-1}; \paramVec\big)$ rather than needing to evaluate these densities.
+Plug-and-play algorithms include Bayesian approaches like ABC and PMCMC \cite{toni09,andrieu10}, but here we use algorithms that enable maximum likelihood estimation.
+To our knowledge, the only plug-and-play methods that can be used to maximize the likelihood for POMP models of this complexity are iterated filtering algorithms \cite{ionides15}, which modify the well-known \emph{particle filter} \cite{arulampalam02}.
 The particle filter is often used to obtain values of latent states, but can also be used to obtain likelihood estimates of a POMP model.
-Iterated filtering algorithms, like IF2 \citep{ionides15}, extend the particle filter by performing a random walk for each parameter and particle;
+Iterated filtering algorithms such as IF2 \cite{ionides15} extend the particle filter by performing a random walk for each parameter and particle;
 these perturbations are carried out iteratively over multiple filtering operations, using the collection of parameters from the previous filtering pass as the parameter initialization for the next iteration, and decreasing the random walk variance at each step.
@@
 where $\bar{t} = \frac{t - (t_N + t_0) / 2}{t_N - (t_N + t_0) / 2}$, so that $\bar{t} \in [-1, 1]$.
-The proposal of a linear trend in transmission is a result of observing an apparent decrease in reported cholera infections from 2012-2019 in Fig.~\ref{CholeraData}.
-While several factors may contribute to this decrease, one explanation is that case-area targeted interventions (CATIs), which included education sessions, increased monitoring, household decontamination, soap distribution, and water chlorination in infected areas \citep{rebaudet19CATI}, may have substantially reduced cholera transmission \citep{rebaudet21}.
+The proposal of a linear trend in transmission is a result of observing an apparent decrease in reported cholera infections from 2012-2019 in Fig.~\ref{fig:CholeraData}.
+While several factors may contribute to this decrease, one explanation is that case-area targeted interventions (CATIs), which included education sessions, increased monitoring, household decontamination, soap distribution, and water chlorination in infected areas \cite{rebaudet19CATI}, may have substantially reduced cholera transmission over time \cite{rebaudet21}.
 
 We perform a statistical test to determine whether or not the data indicate the presence of a linear trend in transmissibility.
-To do this, we perform a profile-likelihood search on the parameter $\transmissionTrend$ and obtain a $95\%$ confidence interval via a Monte Carlo Adjusted Profile (MCAP) \citep{ionides17}.
-\cite{lee20} implemented Model~1 by fitting two distinct phases: an epidemic phase from October 2010 through March 2015, and an endemic phase from March 2015 onward.
+To do this, we perform a profile-likelihood search on the parameter $\transmissionTrend$ and obtain a $95\%$ confidence interval via a Monte Carlo Adjusted Profile (MCAP) \cite{ionides17}.
+Lee et al.~\cite{lee20} implemented Model~1 by fitting two distinct phases: an epidemic phase from October 2010 through March 2015, and an endemic phase from March 2015 onward.
 We similarly allow the re-estimation of process and measurement overdispersion parameters ($\sigmaProc^2$ and $\obsOverdispersion$), and require that the latent Markov process $X(t)$ carry over from one phase into the next.
 The resulting confidence interval for $\transmissionTrend$ is $(\Sexpr{myround(mcap_results$ci[1], digits = 3)}, \Sexpr{myround(mcap_results$ci[2], digits = 3)})$, with the full results displayed in Fig.~\ref{fig:betat}.
-These results are suggestive that the inclusion of a trend in transmission rate improves the quantitative ability of Model~1 to describe the observed data.
+These results are suggestive that the inclusion of a trend in the transmission rate improves the quantitative ability of Model~1 to describe the observed data.
 The reported results for Model~1 in the remainder of this article were obtained with the inclusion of the parameter $\transmissionTrend$.
 
-\begin{figure}
+\begin{figure}[!h]
 \centering
@@
 
-We implemented Model~1 using the \code{pomp} package \citep{king16}, relying heavily on the \code{pomp} source code provided by \cite{lee20}.
+We implemented Model~1 using the \code{pomp} package \cite{king16}, relying heavily on the \code{pomp} source code provided by Lee et al.~\cite{lee20}.
 Both analyses used the \code{mif2} implementation of the IF2 algorithm to estimate $\paramVec$ by maximum likelihood.
@@
 Many Monte Carlo replications are then required to quantify and further reduce the error.
-The large difference in likelihoods that were obtain by these two approaches (see Table~\ref{tab:likes}) highlights the importance of carefully determining the necessary computational effort needed to maximize model likelihoods and acting accordingly.
-
-\begin{figure}
+The large increase in the log-likelihood for Model~1 (see Table~\ref{tab:likes}) can primarily be attributed to increasing the computational effort in fitting the model; this result highlights the importance of carefully determining the necessary computational effort needed to maximize model likelihoods and acting accordingly.
+Simulations from the initial conditions of our fitted model are plotted against the observed incidence data in Fig.~\ref{fig:mod1fit}.
+
+\begin{figure}[!h]
 \centering
@@
 
-\subsubsection{Calibrating Model~2 Parameters}\label{sec:fit2}
+\subsubsection*{Calibrating Model~2 Parameters}\label{sec:fit2}
 
 Model~2 is a deterministic compartmental model defined by a set of coupled differential equations.
-The use of deterministic compartment models have a long history in the field of infectious disease epidemiology \citep{kermack1927,brauer2017,giordano20},
-and can be justified by asymptotic considerations in a large-population limit \citep{dadlani2020,ndii17}.
+The use of deterministic compartment models have a long history in the field of infectious disease epidemiology \cite{kermack1927,brauer2017,giordano20},
+and can be justified by asymptotic considerations in a large-population limit \cite{dadlani2020,ndii17}.
 Because the process model of Model~2 is deterministic, maximum likelihood estimation reduces to a least squares calculation when combined with a Gaussian measurement model (see Sec.~S3.2 of the supplement material for details).
-\cite{lee20} fit two versions of Model~2 based on a presupposed change in cholera transmission from a epidemic phase to endemic phase that occurred in March, 2014.
+Lee et al.~\cite{lee20} fit two versions of Model~2 based on a presupposed change in cholera transmission from a epidemic phase to endemic phase that occurred in March, 2014.
 The inclusion of a change-point in model states and parameters increases the flexibility of the model and hence the ability to fit the observed data.
 The increase in model flexibility, however, results in hidden states that are inconsistent between model phases.
-The inclusion of a model break-point by \cite{lee20} is perhaps due to a challenging feature of fitting a deterministic model via least squares: discrepancies between model trajectories and observed case counts in highly infectious periods of a disease outbreak will result in greater penalty than the discrepancies between model trajectories and observed case counts in times of relatively low infectiousness, resulting in a bias towards accurately describing periods of high infectiousness.
+The inclusion of a model break-point by Lee et al.~\cite{lee20} is perhaps due to a challenging feature of fitting a deterministic model via least squares: discrepancies between model trajectories and observed case counts in highly infectious periods of a disease outbreak will result in greater penalty than the discrepancies between model trajectories and observed case counts in times of relatively low infectiousness, resulting in a bias towards accurately describing periods of high infectiousness.
 This issue is particularly troublesome for modeling cholera dynamics in Haiti: the inability to accurately fit times of low infectiousness may result in poor model forecasts, as few cases of cholera were observed in the last few years of the epidemic.
@@
 This permits the consideration of demographic stochasticity, which is dominant for small infected populations, together with log scale stochasticity (also called multiplicative, or environmental, or extra-demographic) which is dominant at high population counts.
-Here we chose to fit the model to transformed case counts rather than adding overdispersion to the measurement model with the goal of minimizing the changes to the model proposed by \cite{lee20}.
-
-We implemented this model using the \code{spatPomp} \code{R} package \citep{asfaw23arxiv}.
-The model was then fit using the subplex algorithm, implemented in the  \code{subplex} package \citep{king2020Subplex}.
-A comparison of the trajectory of the fitted model to the data is given in Fig~\ref{fig:mod2Traj}.
+Here we chose to fit the model to transformed case counts rather than adding overdispersion to the measurement model with the goal of minimizing the changes to the model proposed by Lee et al.~\cite{lee20}.
+
+We implemented this model using the \code{spatPomp} \code{R} package \cite{asfaw23arxiv}.
+The model was then fit using the subplex algorithm, implemented in the  \code{subplex} package \cite{king2020Subplex}.
+A comparison of the trajectory of the fitted model to the data is given in Fig.~\ref{fig:mod2Traj}.
 
@@
 
-\subsubsection{Calibrating Model~3 Parameters}\label{sec:fit3}
+<<Load Model 2 VaccScenarios, message=FALSE, include=FALSE>>=
+
+# Run Vaccination scenarios for Model 2. Note that the code runs the same
+# speed for each run-level.
+mod2_VaccScenarios <- bake(
+  file = paste0("model2/VaccinationScenarios.rds"), {
+    haiti2_vaccScenario(
+      h2_params = h2_fit$h2_params
+    )
+  },
+  timing = FALSE
+)
+@
+
+\begin{figure}[!h]
+<<Mod2Fit_and_Scenarios_Figure, fig.height=1.9, fig.width=4.8, fig.align='center'>>=
+ggplot() +
+  geom_line(data = h2_traj, aes(x = date, y = Ctotal + 1), col = 'blue') +
+  geom_line(data = true_agg_cases, aes(x = date, y = ReportedAll + 1)) +
+  geom_line(data = filter(mod2_VaccScenarios$all_trajs,
+                          year >= max(h2_traj$year)),
+            aes(x = as.Date(date), y = ReportedCases + 1, color = scenario)) +
+  theme(axis.title.x = element_blank(),
+        legend.title = element_blank(),
+        axis.title.y = element_text(size = 9),
+        axis.text = element_text(size = 8)) +
+  ylab("Reported cholera cases") +
+  scale_color_manual(values = c("V0" = '#f4a582',
+                                "V1" = '#d6604d',
+                                "V2" = '#b2182b',
+                                "V3" = '#92c5de',
+                                "V4" = '#4393c3')) +
+  scale_y_log10(labels = scales::trans_format("log10", scales::math_format(10^.x))) +
+  scale_x_date(date_labels = "'%y", breaks = seq.Date(from = as.Date("2010-01-01"), as.Date("2030-01-01"), by = '2 years'))
+@
+\caption{\label{fig:mod2Traj}
+{\figTitle Simulated trajectory of Model~2.}
+The black line shows the nationally aggregated weekly cholera incidence data.
+The blue curve from 2012-2019 is the trajectory of the calibrated version of Model~2.
+Projections under the various vaccination scenarios, which are discussed in detail in the  {\bf Forecasts} subsection are also included.}
+\end{figure}
+
+\subsubsection*{Calibrating Model~3 Parameters}\label{sec:fit3}
 Model~3 describes cholera dynamics in Haiti using a metapopulation model, where the hidden states in each administrative department has an effect on the dynamics in other departments.
 The decision to address metapopulation dynamics using a spatially explicit model, rather than to aggregate over space, is double-edged.
-Evidence for the former approach has been provided in previous studies \citep{king15}, including the specific case of heterogeneity between Haitian departments in cholera transmission \citep{collins14}.
-However, a legitimate preference for simplicity \citep{saltelli20,green15} can support a decision to consider nationally aggregated models.
+Evidence for the former approach has been provided in previous studies \cite{king15}, including the specific case of heterogeneity between Haitian departments in cholera transmission \cite{collins14}.
+However, a legitimate preference for simplicity can support a decision to consider nationally aggregated models \cite{saltelli20,green15}.
 One issue that arises when fitting spatially explicit models is that parameter estimation techniques based on the particle filter become computationally intractable as the number of spatial units increases.
-This is a result of the approximation error of particle filters growing exponentially in the dimension of the model \citep{rebeschini15,park20}.
-
-To avoid the approximation error present in high-dimensional models, \citet{lee20} simplified the problem of estimating the parameters of Model~3 by creating an approximate version of the model where the units are independent given the observed data.
+This is a result of the approximation error of particle filters growing exponentially in the dimension of the model \cite{rebeschini15,park20}.
+
+To avoid the approximation error present in high-dimensional models, Lee et al.~\cite{lee20} simplified the problem of estimating the parameters of Model~3 by creating an approximate version of the model where the units are independent given the observed data.
 Reducing a spatially coupled model to individual units in this fashion requires special treatment of any interactive mechanisms between spatial units, such as found in Eq.~\myeqref{eq:model3:foi}.
-Because the simplified, spatially-decoupled version of Model~3 implemented by \cite{lee20} relies on the observed cholera cases, the calibrated model cannot readily be used to obtain forecasts.
-Therefore, in order to obtain model forecasts, \cite{lee20} used the parameters estimates from the spatially-decoupled approximation of Model~3 to obtain forecasts using the fully coupled version of the model.
+Because the simplified, spatially-decoupled version of Model~3 implemented in \cite{lee20} relies on the observed cholera cases, the calibrated model cannot readily be used to obtain forecasts.
+Therefore, in order to obtain model forecasts, Lee et al.~\cite{lee20} used the parameters estimates from the spatially-decoupled approximation of Model~3 to obtain forecasts using the fully coupled version of the model.
 This approach of model calibration and forecasting avoids the issue of particle depletion, but may also be problematic.
@@
 as parameters that maximize the likelihood of the decoupled model almost certainly do not maximize the likelihood of the fully coupled model.
-These two concerns may explain the unrealistic forecasts and low likelihood of the model (see Table~\ref{tab:likes}).
-
-For our analysis, we calibrate the parameters of the spatially coupled version of Model~3 using the iterated block particle filter (IBPF) algorithm of \cite{ionides22}.
-This algorithm extends the work of \cite{ning21ibpf}, who provided theoretic justification for the version of the algorithm that only estimates unit-specific parameters.
+These two concerns may explain the unrealistic forecasts and low likelihood of Model~3 of Lee et al.~\cite{lee20} (see Table~\ref{tab:likes}).
+
+At the time Lee et al.~\cite{lee20} conducted their study, there was no known algorithm that could readily be used to maximize the likelihood of an arbitrary meta-population POMP model with coupled spatial dynamics, which justifies the spatial decoupling approximation that was used to calibrate model parameters.
+For our analysis, we calibrate the parameters of the spatially coupled version of Model~3 using the iterated block particle filter (IBPF) algorithm \cite{ionides22}.
+This algorithm extends the work of Ning and Ionides \cite{ning21ibpf}, who provided theoretic justification for the version of the algorithm that only estimates unit-specific parameters.
 The IBPF algorithm enables us to directly estimate the parameters of models describing high-dimensional partially-observed nonlinear dynamic systems via likelihood maximization.
-The ability to directly estimate parameters of Model~3 is responsible for the large increase in model likelihoods reported in Table~\ref{tab:likes}, as no algorithm with similar capabilities existed when \cite{lee20} published their results.
+The ability to directly estimate parameters of Model~3 is responsible for the large increase in model likelihoods reported in Table~\ref{tab:likes}.
 Simulations from the fitted model are displayed in Fig.~\ref{fig:h3spatsims}.
@@
 
-\begin{figure}
+\begin{figure}[!h]
 <<Model3_SpatPOMP_sims_Figure, fig.height=3>>=
@@
 
-\subsection{Model Diagnostics} \label{sec:model_diagnostics}
-
-Parameter calibration (whether Bayesian or frequentist) aims to find the best description of the observed data under the assumptions of the model.
-Obtaining the best fitting set of parameters for a given model does not, however, guarantee that the model provides an accurate representation of the system in question.
+\subsection*{Model Diagnostics}\label{sec:model_diagnostics}
+
+The goal of parameter calibration---whether done using Bayesian or frequentist methods---is to find the best description of the observed data under the assumptions of the model.
+Obtaining the best fitting set of parameters for a given model does not, however, guarantee that the model provides an accurate representation of the system under investigation.
 Model misspecification, which may be thought of as the omission of a mechanism in the model that is an important feature of the dynamic system, is inevitable at all levels of model complexity.
-To make progress, while accepting proper limitations, one must bear in mind the much-quoted observation of \citet{box79} that ``all models are wrong but some are useful.''
-Beyond being good practical advice for applied statistics, this assertion is relevant for the philosophical justification of statistical inference as severe testing \citep[][Sec.~4.8]{mayo18}.
+To make progress, while accepting proper limitations, one must bear in mind the much-quoted observation of George Box \cite{box79} that ``all models are wrong but some are useful.''
+Beyond being good practical advice for applied statistics, this assertion is relevant for the philosophical justification of statistical inference as severe testing \cite{mayo18}.
+% \citep[][Sec.~4.8]{mayo18}.
 In this section, we discuss some tools for diagnosing mechanistic models with the goal of making the subjective assessment of model ``usefulness'' more objective.
-To do this, we will rely on the quantitative and statistical ability of the model to match the observed data, which we call the model's {\it goodness-of-fit}, with the guiding principle that a model which cannot adequately describe observed data may not be reliable for useful purposes.
+To do this, we will rely on the quantitative ability of the model to match the observed data, which we call the model's {\it goodness-of-fit}, with the guiding principle that a model which cannot adequately describe observed data may not be reliable for useful purposes.
 Goodness-of-fit may provide evidence supporting the causal interpretation of one model versus another, but cannot by itself rule out the possibility of alternative explanations.
@@
 While visual comparisons can be informative, they provide only a weak and informal measure of the goodness-of-fit of a model.
-The study by \citet{lee20} provides an example of this: their models and parameter estimates resulted in simulations that visually resembled the observed data, yet resulted in model likelihoods that were---in some cases---considerably smaller than likelihoods that can be achieved via the likelihood based optimization techniques that were used (see Table~\ref{tab:likes}).
+The study by Lee et al.~\cite{lee20} provides an example of this: their models and parameter estimates resulted in simulations that visually resembled the observed data, yet resulted in model likelihoods that were considerably smaller than likelihoods that can be achieved via the likelihood based optimization techniques that were used (see Table~\ref{tab:likes}).
 Alternative forms of model validation should therefore be used in conjunction with visual comparisons of simulations to observed data.
@@
 Ideally, the measure should also have favorable theoretical properties.
-Consequently, we focus on log-likelihood as a measure of goodness of fit, and we adjust for the degrees of freedom of the models to be compared by using the Akaike information criterion (AIC) \citep{aic74}.
+Consequently, we focus on log-likelihood as a measure of goodness of fit, and we adjust for the degrees of freedom of the models to be compared by using the Akaike information criterion (AIC) \cite{aic74}.
 
 In some cases, a possible benchmark model could be a generally accepted mechanistic model, but often no such model is available.
-Because of this, we use a log-linear Gaussian ARMA model as an associative benchmark, as recommended by \citet{he10}.
-The theory and practice of ARMA models is well developed, and these linear models are appropriate on a log scale due to the exponential growth and decay characteristic of biological dynamics.
+Because of this, we use a simple negative binomial model with an auto regressive mean as our associative benchmark; this model is described in \myeqref{eq:AR_NB}.
+\begin{eqnarray}
+\label{eq:AR_NB}
+Y_n|Y_{n - 1} \sim \text{NB}\big(\NBintercept + \NBar Y_{n-1}, \NBsize \big),
+\end{eqnarray}
+where $\mathrm{E}\big(Y_n|Y_{n - 1}\big) = \NBintercept + \NBar Y_{n-1}$, and $\Var\big(Y_n|Y_{n - 1}\big) = \mathrm{E}\big(Y_n|Y_{n - 1}\big) + \mathrm{E}\big(Y_n|Y_{n - 1}\big)^2 / \NBsize$.
+In the case where the case counts are large, an alternative benchmark recommended by He et al.~\cite{he10} is a log-linear Gaussian ARMA model;
+the theory and practice of ARMA models is well developed, and these linear models are appropriate on a log scale due to the exponential growth and decay characteristic of biological dynamics.
+We chose to use the auto-regressive negative binomial model, however, because the large number of weeks with zero recorded cholera cases in department level data makes a benchmark based on a continuous distribution problematic.
 Likelihoods of Models~1--3 and their respective ARMA benchmark models are provided in Table~\ref{tab:likes}.
@@
 In addition, the published goodness of fit provides a concrete point of comparison for subsequent scientific investigations.
-When combined with online availability of data and code, objective measures of fit provide a powerful tool to accelerate scientific progress, following the paradigm of the {\it common task framework} \citep[][Sec.~6]{donoho17}.
+When combined with online availability of data and code, objective measures of fit provide a powerful tool to accelerate scientific progress, following the paradigm of the {\it common task framework} \cite{donoho17}.
+% \citep[][Sec.~6]{donoho17}.
 
@@
 
-<<Table 2 Input, echo=FALSE, eval=TRUE>>=
-NREPS_EVAL <- switch(RUN_LEVEL,   3,   8,    36)
-NP_EVAL    <- switch(RUN_LEVEL, 100, 500,  5000)
-
-# Objective function of model 2, epidemic phase
-epi_ofun <- traj_objfun(
-  h2_epi,
-  params = h2_fit$h2_params
-)
-
-mod2_epi_ll <- -epi_ofun(par = h2_fit$h2_params)
-mod2_ll <- mod2_epi_ll - sum(log(h2_epi@data + 1), na.rm = TRUE)
-
-arima_lik2 <- 0
-
-# model 2 AIC
-mod2_n_params <- h2_fit$n_fit_params
-mod2_aic <- 2 * mod2_n_params - 2 * mod2_ll
-
-# model 1 AIC
-mod1_n_params <- length(
-  c('beta1', 'beta2', 'beta3', 'beta4', 'beta5', 'beta6',
-    'rho', 'nu', 'E_0', 'I_0', 'tau_epi',
-    'tau_end', 'sig_sq_epi', 'sig_sq_end', 'betat')
-)
-mod1_aic <- 2 * mod1_n_params - 2 * mod1_ll
-
-# model 3 AIC
-mod3_aic <- 2 * mod3_n_params - 2 * mod3_ll
-
-# Calculate likelihood via block-particle filter
-mod3_eval_res <- bake(
-  file = paste0('model3/', rl_dir, 'mod3_eval_res.rds'), {
-
-    # Create a list to store results
-    results <- list()
-
-    coef(h3Spat) <- p3
-
-    # Perform block-particle filter
-    h3_bpf <- foreach(i = 1:NREPS_EVAL, .combine = c) %dopar% {
-      bpfilter(h3Spat, Np = NP_EVAL, block_size = 1)
-    }
-
-    ll <- logLik(h3_bpf)
-    results$ll <- logmeanexp(ll[!is.na(ll)], se = FALSE)
-
-    ### Get likelihood for subset of the data
-
-    # Find appropriate sub-set
-    in_subset_cols <- h3_bpf[[1]]@times >= as.numeric(lubridate::decimal_date(as.Date("2014-03-01")))
-    in_subset_rows <- h3_bpf[[1]]@unit_names != 'Ouest'
-
-    mod3_coupled_subset_evals <- c()
-    for (i in 1:length(h3_bpf)) {
-      mod3_coupled_subset_evals <- c(
-        mod3_coupled_subset_evals,
-        sum(h3_bpf[[i]]@block.cond.loglik[in_subset_rows, in_subset_cols])
-      )
-    }
-
-    results$subset_ll <- logmeanexp(mod3_coupled_subset_evals)
-    results
-  },
-  timing = FALSE
-)
-
-mod3_eval_res$subset_aic <- 2 * mod3_n_params - 2 * mod3_eval_res$subset_ll
-
-rm(NREPS_EVAL, NP_EVAL)
-@
-
-<<CalculateARMA, echo=FALSE, message=FALSE, include=FALSE, cache=TRUE>>=
-
-ARMA_benchmarks <- list()
-
-###
-### Model 1 ###
-###
-
-# Load aggregated data
-m1_agg_data <- haiti1_agg_data()
-
-# Fit ARMA(2, 1) on log-cases
-m1_log_arma <- arima(log(m1_agg_data$cases + 1), order = c(2, 0, 1))
-
-# Fit ARMA(2, 1) on natural scale of cases
-m1_arma <- arima(m1_agg_data$cases, order = c(2, 0, 1))
-
-# Check which one is better, and save benchmark as the better one
-if (m1_log_arma$loglik - sum(log(m1_agg_data$cases + 1), na.rm = TRUE) >= m1_arma$loglik) {
-  ARMA_benchmarks[['m1']] <- m1_log_arma$loglik - sum(log(m1_agg_data$cases + 1), na.rm = TRUE)
-} else {
-  ARMA_benchmarks[['m1']] <- m1_arma$loglik
-}
-
-rm(m1_agg_data, m1_log_arma, m1_arma)
-gc()
-
-###
-### Model 2 ###
-###
-
-h2_data <- haiti2_data(recreate = FALSE)
-
-# model 2 ARIMA benchmark
-h2_dep_names <- h2_data$department %>% unique()
-m2_benchmark <- 0
-for (i in 1:10){
-  cases <- h2_data[h2_data$department == h2_dep_names[i], ]$cases
-  log_cases <- log(cases + 1)
-
-  m2_arma_dep_ll <- arima(cases, order=c(2, 0, 1))$loglik
-  m2_log_arma_dep_ll <- arima(log_cases, order=c(2, 0, 1))$loglik - sum(log_cases, na.rm = TRUE)
-
-  m2_benchmark <- m2_benchmark + max(m2_arma_dep_ll, m2_log_arma_dep_ll)
-}
-
-ARMA_benchmarks[['m2']] <- m2_benchmark
-
-rm(m2_benchmark, h2_dep_names, cases, log_cases, m2_arma_dep_ll, m2_log_arma_dep_ll, h2_data)
-gc()
-
-###
-### Model 3 ###
-###
-
-mod3_cases <- as.data.frame(t(h3Spat@data))
-
-m3_benchmark <- 0
-for (dep in 1:10) {
-
-  # Get department cases
-  cases <- as.numeric(mod3_cases[, dep])
-
-  # Fit model on natural scale of cases
-  m3_a201 <- arima(cases, order = c(2, 0, 1))
-
-  # Fit model on log-scale of cases
-  m3_log_a201 <- arima(log(cases + 1), order = c(2, 0, 1))
-
-  m3_benchmark <- m3_benchmark + max(
-    m3_log_a201$loglik - sum(log(cases + 1), na.rm = TRUE),
-    m3_a201$loglik
-  )
-}
-
-ARMA_benchmarks[['m3']] <- m3_benchmark
-rm(cases, m3_a201, m3_log_a201, m3_benchmark, dep,
-   mod3_cases, temp, jumps, MODEL3_CASES, i, j)
-gc()
-@
-
-<<Table 2 Lee Input, echo=FALSE, include=FALSE>>=
-mod1_lee_n_params <- 20
-mod1_lee_aic <- 2 * mod1_lee_n_params - 2 * mod1_lee_ll
-
-mod2_lee_aic <- 2 * h2_fit$leeFit_n_params - 2 * h2_fit$leeFit_ll
-@
-
-<<Lee3 Sub Calculations, echo=FALSE, include=FALSE, message=FALSE>>=
-lee3_NP   <- switch(RUN_LEVEL, 50, 250, 5000)
-lee3_NREP <- switch(RUN_LEVEL,  3,   8,   36)
-
-registerDoRNG(49637421)
-
-# Create Lee et al (2020a) version of model 3 on sub-data
-lee3_sub <- lee3_spatPomp()
-
-# This model is only fit for a subset of the data, so we will set data
-# that are not part of this subset as NA. This will allow us to estimate
-# the likelihood for only the values that were considered during the
-# parameter estimation phase.
-# set_as_na <- lee3_sub@times < lubridate::decimal_date(as.Date('2017-06-10'))
-# temp_ouest_cases <- lee3_sub@data['casesOuest', ]
-# temp_ouest_cases[set_as_na] <- NA_real_
-# lee3_sub@data['casesOuest', ] <- temp_ouest_cases
-
-lee3_all <- lee3_spatPomp(start_date = "2010-10-23")
-
-lee3_results <- bake(
-  file = paste0('model3/', rl_dir, 'lee3_results.rds'), {
-
-    # Create lists to save results
-    lee3_results <- list()
-
-    # Save how long it takes for sub-data
-    lee3_results[["sub_bpf_time"]] <- system.time({
-
-      lee3_sub_ll_evals <- foreach(
-        i = 1:lee3_NREP, .combine = rbind
-      ) %dopar% {
-
-        liks <- numeric(10L)
-
-        lee_bpf_out <- bpfilter(lee3_sub, block_size = 1, Np = lee3_NP)
-        Ouest_ind <- which(lee_bpf_out@unit_names == "Ouest")
-
-        for (u in 1:10) {
-          if (u == Ouest_ind) {
-            liks[u] <- sum(lee_bpf_out@block.cond.loglik[u, lee3_sub@times >= lubridate::decimal_date(as.Date('2017-06-10'))])
-          } else {
-            liks[u] <- sum(lee_bpf_out@block.cond.loglik[u, ])
-          }
-        }
-
-        liks
-      }
-    })
-
-    # Save all likelihood evaluations
-    lee3_results[["sub_ll_eval"]] <- logmeanexp(rowSums(lee3_sub_ll_evals)[!is.na(rowSums(lee3_sub_ll_evals))])
-
-    registerDoRNG(2789131)
-
-    # Save how long it takes for all data
-    lee3_results[["all_pf_time"]] <- system.time({
-      # Save all likelihood evaluations
-
-      lee3_all_ll_evals <- foreach(
-        i = 1:lee3_NREP, .combine = c
-      ) %dopar% {
-        logLik(bpfilter(lee3_all, block_size = 1, Np = lee3_NP))  # Get likelihood evaluations
-      }
-
-      lee3_results[["all_ll_eval"]] <- logmeanexp(lee3_all_ll_evals[!is.na(lee3_all_ll_evals)])
-
-    })
-
-    lee3_results[["n_params"]] <- 29
-
-    lee3_results
-  },
-  timing = FALSE
-)
-
-rm(lee3_sub, lee3_all, lee3_NP, lee3_NREP)
-gc()
-
-mod3_lee_coupled_aic <- 2 * lee3_results$n_params - 2 * lee3_results$all_ll_eval
-
-mod3_lee_coupled_partial_aic <- 2 * lee3_results$n_params - 2 * lee3_results$sub_ll_eval
-@
-
-\begin{table}
-\centering
-
-\begin{tabular}{|c|c|c|c|}
-\hline
- & \thead{Model~1} & \thead{Model~2} & \thead{Model~3}
-\\
-\hline
-\hline
-\multirow{2}{*}{Log-likelihood} &
-  $\Sexpr{myround(mod1_ll, 1)}$ &
-  $\Sexpr{myround(mod2_ll, 1)}$ &
-  $\Sexpr{myround(mod3_ll, 1)}$ \\
-    & ($\Sexpr{myround(mod1_lee_ll, 1)}$)\footnotemark[1] &
-  ($\Sexpr{myround(h2_fit$leeFit_ll, 1)}$) &
-  ($\Sexpr{myround(lee3_results[["all_ll_eval"]], 1)}$)\footnotemark[2]
-\\
-\hline
-Number of &
-  $\Sexpr{as.character(mod1_n_params)}$ &
-  $\Sexpr{as.character(mod2_n_params)}$ &
-  $\Sexpr{as.character(mod3_n_params)}$ \\
- Fit Parameters &
- (\Sexpr{as.character(mod1_lee_n_params)}) &
- ($\Sexpr{as.character(h2_fit$leeFit_n_params)}$) &
- ($\Sexpr{as.character(lee3_results$n_params, 1)}$)
-\\
-\hline
-\multirow{2}{*}{AIC} &
-  $\Sexpr{myround(mod1_aic, 1)}$ &
-  $\Sexpr{myround(mod2_aic, 1)}$ &
-  $\Sexpr{myround(mod3_aic, 1)}$ \\
-  & ($\Sexpr{myround(mod1_lee_aic, 1)}$)\footnotemark[1] &
-  ($\Sexpr{myround(mod2_lee_aic, 1)}$) &
-  ($\Sexpr{myround(mod3_lee_coupled_aic, 1)}$)\footnotemark[2]
-\\
-\hline
-\thead{Log-ARMA(2,1) \\ AIC} &
-  $\Sexpr{myround(-2 * ARMA_benchmarks$m1 + 2 * 5, 1)}$ &
-  $\Sexpr{myround(-2 * ARMA_benchmarks$m2 + 2 * 5, 1)}$ &
-  $\Sexpr{myround(-2 * ARMA_benchmarks$m3 + 2 * 5, 1)}$
-\\
-\hline
-\end{tabular}
-\caption{\label{tab:likes}AIC values for each model compared to their ARMA benchmark.
-Values in parentheses are corresponding values using \citet{lee20} parameter estimates.
-$\textsuperscript{1}$The reported likelihood is an upper bound of the likelihood of the \citet{lee20} model.
-$\textsuperscript{2}$\cite{lee20} fit Model~3 to a subset of the data (March 2014 onward, excluding data from Ouest in 2015-2016).
-On this subset, their model has a likelihood of $\Sexpr{myround(lee3_results$sub_ll_eval, 1)}$.
-On this same subset, our model has a likelihood of $\Sexpr{myround(mod3_eval_res$subset_ll, 1)}$.
-See Sec.~S6 of the supplement material for more details on estimating the likelihood of the \cite{lee20} models.
-}
-\end{table}
-
 Comparing model log-likelihoods to a suitable benchmark may not be sufficient to identify all the strengths and weaknesses of a given model.
 Additional techniques include the inspection of conditional log-likelihoods of each observation given the previous observations in order to understand how well the model describes each data point (Sec.~S5).
-Other tools include plotting the effective sample size of each observation \citep{liu01}; plotting the values of the hidden states from simulations (supplement Sec.~5.1);
-and comparing summary statistics of the observed data to simulations from the model \citep{wood10,king15}.
-
-\subsection{Forecasts}\label{sec:filter}
+Other tools include plotting the effective sample size of each observation \cite{liu01}; plotting the values of the hidden states from simulations (\nameref{S_mod3cal});
+and comparing summary statistics of the observed data to simulations from the model \cite{wood10,king15}.
+
+\subsection*{Corroborating Fitted Models with Scientific Knowledge}
+
+The resulting mechanisms in a fitted model can be compared to current scientific knowledge about a system.
+Agreement between model based inference and our current understanding of a system may be taken as a confirmation of both model based conclusions and our scientific understanding.
+On the other hand, comparisons may generate unexpected results that have the potential to spark new scientific knowledge \cite{ganusov16}.
+
+In the context of our case study, we demonstrate how the fit of Model~1 corroborates other evidence concerning the role of rainfall in cholera epidemics.
+Specifically, we examine the results of fitting the flexible cubic spline term in Model~1 (Eqs.~\myeqref{model1:lambda}--\myeqref{model1:beta}).
+The cubic splines permit flexible estimation of seasonality in the force of infection, $\transmission(t)$.
+Fig.~\ref{fig:h1SeasRain} shows that the estimated seasonal transmission rate $\transmission$ mimics the rainfall dynamics in Haiti, despite Model~1 not having access to rainfall data.
+This is consistent with previous studies finding that rainfall played an important role in cholera transmission in Haiti \cite{lemaitre19,eisenberg13}.
+The estimated seasonality also features an increased transmission rate during the fall, which was noticed at an earlier stage of the epidemic \cite{rinaldo12}; the high transmission rate in the fall may also be a result of the increase transmission that occurred in the fall of 2016, when hurricane Matthew struck Haiti \cite{ferreirai16}.
+
+<<Get Model 1 Seasonality>>=
+h1_seas_df <- get_h1_seasonality(h1)
+
+df_rain <- haitiRainfall %>%
+  pivot_longer(  # Convert data to "long" format
+    data = .,
+    cols = -date,
+    names_to = 'dep',
+    values_to = 'rainfall'
+  ) %>%
+  mutate(year = lubridate::year(date), week = lubridate::week(date)) %>%
+  group_by(year, week) %>%
+  summarize(nwr = sum(rainfall)) %>%
+  ungroup() %>%  # Not needed, but good practice
+  mutate(
+    date_week_start = lubridate::ymd(paste0(2014, "-01-01")) +
+      lubridate::weeks(week - 1),
+    week_date = as.Date(date_week_start, format = "%m-%d"),
+    std_national_rain = (nwr - min(nwr)) / (max(nwr) - min(nwr))
+  )
+
+# Calculate the average national weekly rainfall.
+mean_rain <- df_rain %>%
+  group_by(week_date) %>%
+  summarize(mean_rain = mean(std_national_rain)) %>%
+  ungroup() %>%
+  mutate(mean_rain = (mean_rain - min(mean_rain)) / ((max(mean_rain) - min(mean_rain))))
+
+# Create a figure showing annual rainfal, scaled from 0-1. This is top figure
+gg_rain <- df_rain %>%
+  filter(week != 1, week != max(week)) %>%
+  ggplot(aes(x = week_date, y = std_national_rain, group = year, col = year)) +
+  geom_line() +
+  theme_bw() +
+  guides(color = 'none') +
+  ylab('Standardized Weekly\nNational Rain') +
+  theme(  # Remove elements that are un-unsed becuase this is top figure
+    axis.title.x = element_blank(),
+    axis.ticks.x = element_blank(),
+    axis.text.x = element_blank(),
+    axis.text.y = element_text(size = 7.5),
+    axis.title.y = element_text(size = 9)
+  )
+
+# Create plot that shows the seasonality of Model 1, alongside the average
+# weekly national rainfall.
+gg_trans <- ggplot() +
+  geom_line(data = h1_seas_df, aes(x = date, y = trans_std), linetype = 'dashed') +
+  geom_line(
+    data = mean_rain,
+    aes(x = week_date + lubridate::years(6), y = mean_rain),
+    col = '#2166ac'
+  ) +
+  scale_x_date(date_labels = "%b", date_breaks = '1 month') +
+  ylab('Standardized Seasonal\nContact Rate') +
+  theme(axis.title.x = element_blank(),
+        axis.text = element_text(size = 7.5),
+        axis.title.y = element_text(size = 9))
+
+@
+
+\begin{figure}[!h]
+<<Model1_Seasonality_Figure, fig.height=2.9, fig.width=5, fig.align='center'>>=
+cowplot::plot_grid(gg_rain, gg_trans, align = "v", ncol = 1)
+@
+\caption{\label{fig:h1SeasRain}
+{\figTitle Seasonality of Model~1 transmission compared to rainfall data.}
+(Top) weekly rainfall in Haiti, lighter colors representing more recent years.
+(Bottom) estimated seasonality in the transmission rate (dashed line) plotted alongside mean rainfall (solid line).
+The outsided effect of rainfall in the fall may be due to Hurricane Matthew, which struck Haiti in October of 2016 and resulted in an increase of cholera cases in the nation.
+}
+\end{figure}
+
+For any model-based inference, it is important to recognize and assess the modeling simplifications and assumptions that were used in order to arrive at the conclusions.
+In epidemiological studies, for example, quantitative understanding of individual-level processes may not perfectly match model parameters that were fit to population-level case counts, even when the model provides a strong statistical fit \cite{he10}.
+This makes direct interpretation of estimated parameters delicate.
+
+Our case study provides an example of this in the parameter estimate for the duration of natural immunity due to cholera infection, $\muRS^{-1}$.
+Under the framework of Model~2, the best estimate for this parameter is $\Sexpr{signif(1/h2_fit$h2_params["sigma"],3)} \, \mathrm{yr}$, suggesting that individuals have effectively permanent immunity to cholera once infected.
+To interpret this result, we bear in mind that the data ranged from 2010-2019, and therefore estimates of immunity longer than $10\, \mathrm{yr}$---the upper end of previous estimates of natural immunity \cite{king08}---effectively result in the same model dynamics.
+The depletion of susceptible individuals may also be attributed to confounding mechanisms---such as localized vaccination programs and non-pharmaceutical interventions that reduce cholera transmission \cite{trevisin22, rebaudet21}---that were not accounted for in the model.
+Perhaps the best interpretation of the estimated parameter, then, is that under model assumptions, the model most adequately describes the observed data by having a steady decrease in the number of susceptible individuals.
+The weak statistical fit of Model~2 compared to a log-linear benchmark (see Table~\ref{tab:likes}) cautions us against drawing quantitative conclusions from this model.
+A model that has a poor statistical fit may nevertheless provide a useful conceptual framework for thinking about the system under investigation.
+However, a claim that the model has been validated against data should be reserved for situations where the model provides a statistical fit that is competitive against alternative explanations.
+
+\subsection*{Robust interpretation of model based conclusions}\label{sec:advice}
+
+In this section, we make some general points on deriving scientific conclusions from statistical analysis of mechanistic models, and connect these points to our specific case study.
+
+A model which aspires to provide quantitative guidance for assessing interventions should provide a quantitative statistical fit for available data.
+However, strong statistical fit does not guarantee a correct causal structure: it does not even necessarily require the model to assert a causal explanation.
+A causal interpretation is strengthened by corroborative evidence.
+For example, reconstructed latent variables (such as numbers of susceptible and recovered individuals) should make sense in the context of alternative measurements of these variables \cite{grad12}.
+In the context of our case study, by reconstructing the latent states of Model~3, we notice that the calibrated model favors higher levels of cholera transmission than what was typically observed in the incidence data (\nameref{S_mod3cal}).
+This result hints at the possibility of model mispecification, and should prompt us to be sceptical of the reliability of forecasts from this model.
+Similarly, parameters that have been calibrated to data should make sense in the context of alternative lines of evidence about the phenomena being modeled, while making allowance for the possibility that the interpretations of parameters may vary when modeling across differing spatial scales.
+
+If a mechanistic model including a feature (such as a representation of a mechanism, or the inclusion of a covariate) fits better than mechanistic models without that feature, and also has competitive fit compared to associative benchmarks, this may be taken as evidence supporting the scientific relevance of the feature.
+As for any analysis of observational data, we must be alert to the possibility of confounding.
+For a covariate, this shows up in a similar way to regression analysis: the covariate under investigation could be a proxy for some other unmodeled or unmeasured covariate.
+For a mechanism, the model feature could in principle explain the data by helping to account for some different unmodeled phenomenon.
+In the context of our analysis, the statistical evidence of a trend in transmission rate in Model~1 could be explained by any trending variable (such as hygiene improvements, or changes in population behavior), resulting in confounding from colinear covariates.
+Alternatively, the trend could be attributed to a decreasing reporting rate rather than decreasing transmission rate, resulting in confounded mechanisms.
+The robust statistical conclusion is that a model which allows for change fits better than one which does not---we argue that a decreasing transmission rate is a plausible way to explain this, but the incidence data themselves do not provide enough information to pin down the mechanism.
+
+\section*{Results}
+
+\subsection*{Forecasts}\label{sec:filter}
 
@@
 This assertion may seem self-evident, but it is not the case for deterministic models, for which the initial conditions together with the parameters are sufficient for forecasting, and so recent data do not have special importance.
-Epidemiological forecasts based on deterministic models are not uncommon in practice, despite their limitations \citep{king15}.
-That may explain why \citet{lee20} chose to obtain forecasts by simulating the calibrated models forward from initial conditions.
-Here, we compare with a forecast projected from the filtering distribution: to obtain a forecast using POMP model $m$ at a collection of times $t_{N+1:N+s}$, where $N$ is the index for the last available data, we simulate forward from a draw from
-$f_{\bm{X}_N|\bm{Y}_{1:N}}(\bm{x}_{N} | \bm{y}^*_{1:N} ; \hat\paramVec)$
+Epidemiological forecasts based on deterministic models are not uncommon in practice, despite their limitations \cite{king15}.
+That may explain why Lee et al.~\cite{lee20} chose to obtain forecasts by simulating the calibrated models forward from initial conditions.
+Here, we obtain forecasts by simulating future values using latent states that are consistent with the most recent observations by drawing latent states at time the last observation time $t_N$ from the filtering distribution $f_{\bm{X}_N|\bm{Y}_{1:N}}(\bm{x}_{N} | \bm{y}^*_{1:N} ; \hat\paramVec)$,
 where $\hat\paramVec$ is a vector of calibrated parameters.
-The decision not to do this partially explains the unsuccessful forecasts of \cite{lee20}: their Table~S7 shows that the subset of their simulations which were consistent with observing zero cases in 2019 also accurately predicted the prolonged absence of detected cholera.
-
-Uncertainty in just a single parameter can lead to drastically different forecasts \citep{saltelli20}.
+The decision not to do this partially explains the unsuccessful forecasts of Lee et al.~\cite{lee20}: their Table~S7 shows that the subset of their simulations which were consistent with observing zero cases in 2019 also accurately predicted the prolonged absence of detected cholera.
+
+Uncertainty in just a single parameter can lead to drastically different forecasts \cite{saltelli20}.
 Therefore, parameter uncertainty should also be considered when obtaining model forecasts to influence policy.
 If a Bayesian technique is used for parameter estimation, a natural way to account for parameter uncertainty is to obtain simulations from the model where each simulation is obtained using parameters drawn from the estimated posterior distribution.
-For frequentist inference, one possible approach is obtaining model forecasts from various values of $\paramVec$, where the values of $\paramVec$ are sampled proportionally according to their corresponding likelihoods \citep{king15}; see Sec.~S7 of the supplement material for more details.
+For frequentist inference, one possible approach is obtaining model forecasts from various values of $\paramVec$, where the values of $\paramVec$ are sampled proportionally according to their corresponding likelihoods \cite{king15} (\nameref{S_uncertain}).
 Both of these approaches share the similarity that parameters are chosen for the forecast approximately in proportion to their corresponding value of the likelihood function, $f_{\bm{Y}_{1:N}}(\bm{y}_{1:N}^*; \paramVec)$.
@@
 
-The primary forecasting goal of \citet{lee20} was to investigate the potential consequences of vaccination interventions on a system to inform policy.
+The primary forecasting goal of Lee et al.~\cite{lee20} was to investigate the potential consequences of vaccination interventions on a system to inform policy.
 One outcome of their study include estimates for the probability of cholera elimination under several possible vaccination scenarios.
@@
     results
-  },
-  timing = FALSE
-)
-@
-
-<<Load Model 2 VaccScenarios, message=FALSE, include=FALSE>>=
-
-# Run Vaccination scenarios for Model 2. Note that the code runs the same
-# speed for each run-level.
-mod2_VaccScenarios <- bake(
-  file = paste0("model2/VaccinationScenarios.rds"), {
-    haiti2_vaccScenario(
-      h2_params = h2_fit$h2_params
-    )
   },
@@
 Estimates of the probability of cholera elimination can therefore be obtained as the proportion of simulations from these models that result in cholera elimination.
-The results of these projections are summarized in Figs.~\ref{fig:Mod1Scenarios}--\ref{fig:elimProbs}.
-
-Probability of elimination estimates of this form are not meaningful for deterministic models, as the trajectory of these models only represent the mean behavior of the system rather than individual potential outcomes.
-We therefore do not provide probability of elimination estimates under Model~2, but show trajectories under the various vaccination scenarios using this model (Fig.~\ref{fig:mod2Traj}).
-
-<<Create Model 1 VaccScen plots>>=
-gg_m1_V0 <- ggplot() +
-  geom_line(data = h1_S0_results$mod1_V0_sims, aes(x = date, y = q50 + 1), col = 'blue') +
-  geom_ribbon(data = h1_S0_results$mod1_V0_sims, aes(x = date, ymin = q05 + 1, ymax = q95 + 1), alpha = 0.5) +
-  geom_line(data = true_agg_cases, aes(x = date, y = ReportedAll + 1)) +
-  ylab("Log reported cases") +
-  theme(axis.title.x = element_blank()) +
-  ggtitle("V0") +
-  scale_x_date(date_labels = "'%y") +
-  scale_y_log10(labels = scales::trans_format("log10", scales::math_format(10^.x)))
-
-gg_m1_V1 <- ggplot() +
-  geom_line(data = h1_S2_results$mod1_V1_sims, aes(x = date, y = q50 + 1), col = 'blue') +
-  geom_ribbon(data = h1_S2_results$mod1_V1_sims, aes(x = date, ymin = q05 + 1, ymax = q95 + 1), alpha = 0.5) +
-  geom_line(data = true_agg_cases, aes(x = date, y = ReportedAll + 1)) +
-  ylab("Log reported cases") +
-  theme(axis.title = element_blank(),
-        axis.ticks.y = element_blank(),
-        axis.text.y = element_blank()) +
-  ggtitle("V1") +
-  scale_x_date(date_labels = "'%y") +
-  scale_y_log10(labels = scales::trans_format("log10", scales::math_format(10^.x)))
-
-gg_m1_V2 <- ggplot() +
-  geom_line(data = h1_S4_results$mod1_V2_sims, aes(x = date, y = q50 + 1), col = 'blue') +
-  geom_ribbon(data = h1_S4_results$mod1_V2_sims, aes(x = date, ymin = q05 + 1, ymax = q95 + 1), alpha = 0.5) +
-  geom_line(data = true_agg_cases, aes(x = date, y = ReportedAll + 1)) +
-  ylab("Log reported cases") +
-    theme(axis.title = element_blank(),
-        axis.ticks.y = element_blank(),
-        axis.text.y = element_blank()) +
-  ggtitle("V2") +
-  scale_x_date(date_labels = "'%y") +
-  scale_y_log10(labels = scales::trans_format("log10", scales::math_format(10^.x)))
-
-gg_m1_V3 <- ggplot() +
-  geom_line(data = h1_S3_results$mod1_V3_sims, aes(x = date, y = q50 + 1), col = 'blue') +
-  geom_ribbon(data = h1_S3_results$mod1_V3_sims, aes(x = date, ymin = q05 + 1, ymax = q95 + 1), alpha = 0.5) +
-  geom_line(data = true_agg_cases, aes(x = date, y = ReportedAll + 1)) +
-  ylab("Log reported cases") +
-    theme(axis.title = element_blank(),
-        axis.ticks.y = element_blank(),
-        axis.text.y = element_blank()) +
-  ggtitle("V3") +
-  scale_x_date(date_labels = "'%y") +
-  scale_y_log10(labels = scales::trans_format("log10", scales::math_format(10^.x)))
-
-gg_m1_V4 <- ggplot() +
-  geom_line(data = h1_S1_results$mod1_V4_sims, aes(x = date, y = q50 + 1), col = 'blue') +
-  geom_ribbon(data = h1_S1_results$mod1_V4_sims, aes(x = date, ymin = q05 + 1, ymax = q95 + 1), alpha = 0.5) +
-  geom_line(data = true_agg_cases, aes(x = date, y = ReportedAll + 1)) +
-  ylab("Log reported cases") +
-    theme(axis.title = element_blank(),
-        axis.ticks.y = element_blank(),
-        axis.text.y = element_blank()) +
-  ggtitle("V4") +
-  scale_x_date(date_labels = "'%y") +
-  scale_y_log10(labels = scales::trans_format("log10", scales::math_format(10^.x)))
-
-@
-
-\begin{figure}
-<<Plot_Model1_Scenarios, fig.height=2>>=
-lay <- rbind(c(1, 1, 1, 1, 2, 2, 2, 3, 3, 3, 4, 4, 4, 5, 5, 5))
-gridExtra::grid.arrange(
-  gg_m1_V0, gg_m1_V1, gg_m1_V2, gg_m1_V3, gg_m1_V4,
-  layout_matrix = lay
-)
-@
-\caption{\label{fig:Mod1Scenarios}
-Simulations of Model~1 under each vaccination scenario.
-Blue line indicates the simulated median of reported cases, and the ribbon represents $95\%$ of simulations.
-% The various vaccination campaigns made no practical difference in the median scenario, but a drastic difference in the extreme cases.
-}
-\end{figure}
-
-\begin{figure}
-<<Mod2Fit_and_Scenarios_Figure, fig.height=1.9, fig.width=4.8, fig.align='center'>>=
-ggplot() +
-  geom_line(data = h2_traj, aes(x = date, y = Ctotal + 1), col = 'blue') +
-  geom_line(data = true_agg_cases, aes(x = date, y = ReportedAll + 1)) +
-  geom_line(data = filter(mod2_VaccScenarios$all_trajs,
-                          year >= max(h2_traj$year)),
-            aes(x = as.Date(date), y = ReportedCases + 1, color = scenario)) +
-  theme(axis.title.x = element_blank(),
-        legend.title = element_blank(),
-        axis.title.y = element_text(size = 9),
-        axis.text = element_text(size = 8)) +
-  ylab("Reported cholera cases") +
-  scale_color_manual(values = c("V0" = '#f4a582',
-                                "V1" = '#d6604d',
-                                "V2" = '#b2182b',
-                                "V3" = '#92c5de',
-                                "V4" = '#4393c3')) +
-  # geom_vline(xintercept = yearsToDate(2014.161), linetype = 'dashed') +
-  scale_y_log10(labels = scales::trans_format("log10", scales::math_format(10^.x))) +
-  scale_x_date(date_labels = "'%y", breaks = seq.Date(from = as.Date("2010-01-01"), as.Date("2030-01-01"), by = '2 years'))
-@
-\caption{\label{fig:mod2Traj}
-Simulated trajectory of Model~2 (blue curve) and projections under the various vaccination scenarios.
-Reported cholera incidence is shown in black.}
-\end{figure}
-
-<<Create Model 3 VaccScen Plots>>=
-gg_m3_V0 <- ggplot() +
-  geom_line(data = mod3_V0_res$mod3_V0_sims, aes(x = date, y = q50 + 1), col = 'blue') +
-  geom_ribbon(data = mod3_V0_res$mod3_V0_sims, aes(x = date, ymin = q05 + 1, ymax = q95 + 1), alpha = 0.5) +
-  geom_line(data = true_agg_cases, aes(x = date, y = ReportedAll + 1)) +
-  ylab("Log reported cases") +
-  theme(axis.title.x = element_blank()) +
-  ggtitle("V0") +
-  scale_x_date(date_labels = "'%y") +
-  scale_y_log10(labels = scales::trans_format("log10", scales::math_format(10^.x)))
-
-gg_m3_V1 <- ggplot() +
-  geom_line(data = mod3_V1_res$mod3_V1_sims, aes(x = date, y = q50 + 1), col = 'blue') +
-  geom_ribbon(data = mod3_V1_res$mod3_V1_sims, aes(x = date, ymin = q05 + 1, ymax = q95 + 1), alpha = 0.5) +
-  geom_line(data = true_agg_cases, aes(x = date, y = ReportedAll + 1)) +
-  ylab("Log reported cases") +
-  theme(axis.title = element_blank(),
-        axis.ticks.y = element_blank(),
-        axis.text.y = element_blank()) +
-  ggtitle("V1") +
-  scale_x_date(date_labels = "'%y") +
-  scale_y_log10(labels = scales::trans_format("log10", scales::math_format(10^.x)))
-
-gg_m3_V2 <- ggplot() +
-  geom_line(data = mod3_V2_res$mod3_V2_sims, aes(x = date, y = q50 + 1), col = 'blue') +
-  geom_ribbon(data = mod3_V2_res$mod3_V2_sims, aes(x = date, ymin = q05 + 1, ymax = q95 + 1), alpha = 0.5) +
-  geom_line(data = true_agg_cases, aes(x = date, y = ReportedAll + 1)) +
-  ylab("Log reported cases") +
-    theme(axis.title = element_blank(),
-        axis.ticks.y = element_blank(),
-        axis.text.y = element_blank()) +
-  ggtitle("V2") +
-  scale_x_date(date_labels = "'%y") +
-  scale_y_log10(labels = scales::trans_format("log10", scales::math_format(10^.x)))
-
-gg_m3_V3 <- ggplot() +
-  geom_line(data = mod3_V3_res$mod3_V3_sims, aes(x = date, y = q50 + 1), col = 'blue') +
-  geom_ribbon(data = mod3_V3_res$mod3_V3_sims, aes(x = date, ymin = q05 + 1, ymax = q95 + 1), alpha = 0.5) +
-  geom_line(data = true_agg_cases, aes(x = date, y = ReportedAll + 1)) +
-  ylab("Log reported cases") +
-    theme(axis.title = element_blank(),
-        axis.ticks.y = element_blank(),
-        axis.text.y = element_blank()) +
-  ggtitle("V3") +
-  scale_x_date(date_labels = "'%y") +
-  scale_y_log10(labels = scales::trans_format("log10", scales::math_format(10^.x)))
-
-gg_m3_V4 <- ggplot() +
-  geom_line(data = mod3_V4_res$mod3_V4_sims, aes(x = date, y = q50 + 1), col = 'blue') +
-  geom_ribbon(data = mod3_V4_res$mod3_V4_sims, aes(x = date, ymin = q05 + 1, ymax = q95 + 1), alpha = 0.5) +
-  geom_line(data = true_agg_cases, aes(x = date, y = ReportedAll + 1)) +
-  ylab("Log reported cases") +
-    theme(axis.title = element_blank(),
-        axis.ticks.y = element_blank(),
-        axis.text.y = element_blank()) +
-  ggtitle("V4") +
-  scale_x_date(date_labels = "'%y") +
-  scale_y_log10(labels = scales::trans_format("log10", scales::math_format(10^.x)))
-@
-
-\begin{figure}
-<<Plot_Model3_Scenarios, fig.height=2>>=
-lay <- rbind(c(1, 1, 1, 1, 2, 2, 2, 3, 3, 3, 4, 4, 4, 5, 5, 5))
-gridExtra::grid.arrange(
-  gg_m3_V0, gg_m3_V1, gg_m3_V2, gg_m3_V3, gg_m3_V4,
-  layout_matrix = lay
-)
-@
-\caption{\label{fig:Mod3Scenarios}
-Simulations of Model~3 under each vaccination scenario.
-Blue line indicates the simulated median of reported cases, and ribbon represents $95\%$ of simulations.
-% The various vaccination campaigns made no practical difference in the median scenario, but a drastic difference in the extreme cases.
-}
-\end{figure}
+The results of these projections are summarized in Figs. \ref{fig:elimProbs}.
 
@@
 \caption{\label{fig:elimProbs}
-Probability of elimination across simulations for a 10 year period.
-Compare to Figure 3A of \cite{lee20}.}
+{\figTitle Simulated probability of elimination using Models~1 and 3.}
+Probability of cholera elimination, defined as having zero cholera infectious for at least 52 consecutive weeks, based on 10 year simulations from calibrated versions of Models~1 and 3.
+Compare to Fig.~3A of Lee et al.~\cite{lee20}.}
 \end{figure}
 
-%%% table tttttttttt
-
-\begin{table}
-\centering
-% \small
-\begin{tabular}{|l@{}|l@{}t@{}|l@{}t@{}|l@{}t@{}|}
-\hline
-Mechanism & Model 1 && Model 2 && Model 3 &
-\\
-\hline
-\hline
-Infection (day)
-  & {\fixed $\muIR^{-1}=\Sexpr{myround(7/p1["gamma"],1)}$}
-  & \eqref{model1:toR}
-  & {\fixed $\muIR^{-1}=\Sexpr{myround(365/h2_fit$h2_params["gamma"],1)}$}
-  & \eqref{model2:mu_IR}
-  & {\fixed $\muIR^{-1}=\Sexpr{myround(365/p3["gamma1"],1)}$ }
-  & \eqref{eq:model3:IR}
-\\
-Latency (day)
-  & {\fixed $\muEI^{-1}=\Sexpr{myround(7/p1["sigma"],1)}$}
-  & \eqref{model1:EA}
-  & {\fixed $\muEI^{-1}=\Sexpr{myround(365/h2_fit$h2_params["gammaE"],1)}$}
-  & \eqref{model2:mu_EI}
-  & \multicolumn{1}{c}{\hspace{16pt}\missing}
-  &
-\\
-\hline
-\multirow{2}{*}{Seasonality}
-  & $\begin{array}{l}
-    \hspace{-1mm} \transmission_{1:6}=(
-      \Sexpr{myround(p1["beta1"],1)},
-      \Sexpr{myround(p1["beta2"],1)},
-    \\
-      \Sexpr{myround(p1["beta3"],1)},
-      \Sexpr{myround(p1["beta4"],1)},
-      \Sexpr{myround(p1["beta5"],1)},
-      \Sexpr{myround(p1["beta6"],1)})
-    \end{array}$
-  & \eqref{model1:beta}
-  & {\fixed \seasAmplitude=0.4}
-  & \eqref{model2:lambda}
-  & $\seasAmplitude = \Sexpr{myround(p3s["lambdaR"],2)}$
-  & \eqref{eq:model3:water} \\
-  & $\transmissionTrend = \Sexpr{myround(p1["betat"],2)}^*$
-  & \eqref{model1:betat}
-  & $\phaseParm = \Sexpr{myround(h2_fit$h2_params['phase'], 2)}^*$
-  & \eqref{model2:lambda}
-  & $r = \Sexpr{signif(p3s["r"],3)}$
-  & \eqref{eq:model3:water}
-  \\\hline
-  Immunity (yr)
-  & {\fixed $\muRS^{-1}=\Sexpr{myround(1/p1["alpha"]/52,1)}$}
-  & \eqref{model1:RS}
-  &
-    $\begin{array}{l}
-    \muRS^{-1}= \Sexpr{signif(1/h2_fit$h2_params["sigma"],2)} \\
-    {\fixed \omega_1^{-1} = \Sexpr{myround(1/h2_fit$h2_params["Omega1"],1)}} \\
-    {\fixed \omega_2^{-1} = \Sexpr{myround(1/h2_fit$h2_params["Omega2"],1)}}
-    \end{array}$
-  & $\begin{array}{l}
-    \eqref{model2:RS} \\
-    \eqref{model2:omega1} \\
-    \eqref{model2:omega1}
-    \end{array} $
-  & {\fixed $\muRS^{-1}=\Sexpr{myround(1/p3['rho1'],1)}$}
-  & \eqref{eq:model3:RRnext}
-\\
-\hline
-Birth/death (yr)
-  & \fixed{$\begin{array}{l}
-    \muBirth^{-1} = \Sexpr{myround(1/p1["mu"]/52,1)} \\
-    \muDeath^{-1} = \Sexpr{myround(1/p1["delta"]/52,1)}
-    \end{array}$}
-  & \eqref{model1:death}
-  & \multicolumn{1}{c}{\hspace{16pt}\missing}
-  &
-  & {\fixed $\muDeath^{-1} = \Sexpr{myround(1/p3["mu1"],1)}$}
-  & \eqref{eq:model3:IS}
-\\
-Sympt. frac.
-  & {\fixed $\symptomFrac_z(t)=c\vaccineEfficacy^*(t-\tau_d)$}
-  & (\ref{model1:EI}-\ref{model1:EA})
-  & {\fixed $\symptomFrac=\Sexpr{myround(h2_fit$h2_params["k"],1)}$}
-  & \eqref{model2:mu_EI}
-  & {\fixed $\symptomFrac=\Sexpr{myround(p3["sigma1"],2)}$}
-  & \eqref{eq:model3:SA}
-\\
-$\begin{array}{l}
-\text{Asympt.} \\
-\text{infectivity}
-\end{array}$
-  & {\fixed $\asymptomRelativeInfect=0.05$ }
-  & \eqref{model1:lambda}
-  & $\begin{array}{l}
-      {\fixed \asymptomRelativeInfect =0.001 } \\
-      {\fixed \asymptomRelativeShed = \Sexpr{1e-7} }
-    \end{array}$
-  & \begin{tabular}{l}
-      \eqref{model2:lambda} \\
-      \eqref{model2:to_W}
-    \end{tabular}
-  &  $\begin{array}{l}
-      {\fixed \asymptomRelativeInfect =1 } \\
-      \asymptomRelativeShed = \Sexpr{myround(p3s["XthetaA"],3)}
-    \end{array}$
-  & \begin{tabular}{l}
-      \eqref{eq:model3:foi} \\
-      \eqref{eq:model3:water}
-    \end{tabular}
-\\
-\hline
-$\begin{array}{l}
-\text{Human to}\\
-\text{human}\end{array}$
-  & $\transmission_{1:6}$ as above
-  & \eqref{model1:lambda}
-  & $\transmission=$\Sexpr{signif(h2_fit$h2_params["Beta"],3)}
-  & \eqref{model2:lambda}
-  & $\begin{array}{l}
-      \hspace{-1mm} \transmission_{1:10}=(
-        \Sexpr{myround(p3u["foi_add",1]*1e6,2)},
-	\Sexpr{myround(p3u["foi_add",2]*1e6,2)},
-      \\
-	\Sexpr{myround(p3u["foi_add",3]*1e6,2)},
-        \Sexpr{myround(p3u["foi_add",4]*1e6,2)},
-	\Sexpr{myround(p3u["foi_add",5]*1e6,2)},
-	\Sexpr{myround(p3u["foi_add",6]*1e6,2)},
-      \\
-	\Sexpr{myround(p3u["foi_add",7]*1e6,2)},
-	\Sexpr{myround(p3u["foi_add",8]*1e6,2)},
-	\Sexpr{myround(p3u["foi_add",9]*1e6,2)},
-	\Sexpr{myround(p3u["foi_add",10]*1e6,2)}
-	)
-      \\
-      \times 10^{-6}
-    \end{array}$
-  & \eqref{eq:model3:foi}
-\\
-\hline
-$\begin{array}{l}
-\text{Water to}\\ \text{human}
-\end{array}$
-  & \multicolumn{1}{c}{\hspace{16pt}\missing}
-  &
-  & $\begin{array}{l}
-    {\fixed \Wsat = \Sexpr{signif(h2_fit$h2_params["Sat"],2)} }
-    \\
-    \beta_W= \Sexpr{signif(h2_fit$h2_params["BetaW"],3)}
-    \end{array}$
-  & \eqref{model2:lambda}
-  & $\begin{array}{l}
-      \hspace{-1mm} \Wbeta{_{1:10}}= (
-        \Sexpr{myround(p3u["betaB",1],2)}, \Sexpr{myround(p3u["betaB",2],2)},
-       \\
-        \Sexpr{myround(p3u["betaB",3],2)}, \Sexpr{myround(p3u["betaB",4],2)},
-        \Sexpr{myround(p3u["betaB",5],2)}, \Sexpr{myround(p3u["betaB",6],2)},
-       \\
-        \Sexpr{myround(p3u["betaB",7],2)}, \Sexpr{myround(p3u["betaB",8],2)},
-        \Sexpr{myround(p3u["betaB",9],2)}, \Sexpr{myround(p3u["betaB",10],2)}
-      ) \end{array}$
-  & \eqref{eq:model3:foi}
-\\
-\hline
-$\begin{array}{l}
-\text{Human to}\\
-\text{water}\end{array}$
-  & \multicolumn{1}{c}{\hspace{16pt}\missing}
-  &
-  & $\Wshed = $ \Sexpr{signif(h2_fit$h2_params["Mu"],3)}
-  & \eqref{model2:to_W}
-  & $\Wshed= \Sexpr{signif(p3s["thetaI"],3)}$
-  & \eqref{eq:model3:water}
-\\
-$\begin{array}{l}
-\text{Water} \\
-\text{survival (wk)}
-\end{array}$
-  & \multicolumn{1}{c}{\hspace{16pt}\missing}
-  &
-  & {\fixed $\Wremoval^{-1} = $ \Sexpr{signif(52/h2_fit$h2_params["Delta"],3)}}
-  & \eqref{model2:from_W}
-  & $\Wremoval^{-1}=\Sexpr{myround(52/p3s["mu_B"],2)}$
-  & \eqref{eq:model3:Decay}
-\\
-Mixing exponent
-  & $\mixExponent=\Sexpr{myround(p1["nu"],2)}$
-  & \eqref{model1:lambda}
-  & \multicolumn{1}{c}{\hspace{16pt}\missing}
-  &
-  & \multicolumn{1}{c}{\hspace{16pt}\missing}
-  &
-\\
-$\begin{array}{l}
-\text{Process} \\
-\text{noise} (\text{wk}^{1/2})
-\end{array}$
-  & $\sigmaProc=(\Sexpr{myround(sqrt(p1["sig_sq_epi"]),2)}, \Sexpr{myround(sqrt(p1["sig_sq_end"]),2)})^*$
-  & \eqref{model1:lambda}
-  & \multicolumn{1}{c}{\hspace{16pt}\missing}
-  &
-  & $\sigmaProc= \Sexpr{myround(p3s["std_W"],3)}$
-  & \eqref{eq:model3:SA}
-\\
-Reporting rate
-  & $\reportRate=\Sexpr{myround(p1["rho"], 3)}$
-  & (S15)
-  & {\fixed $\reportRate=\Sexpr{myround(h2_fit$h2_params["Rho"],2)}$}
-  & (S16)
-  & $\reportRate=\Sexpr{myround(p3s["epsilon"],2)}$
-  & (S19)
-\\
-$\begin{array}{l}
-\text{Observation} \\
-\text{variance}
-\end{array}$
-  & $\obsOverdispersion = (\Sexpr{myround(p1["tau_epi"],2)}, \Sexpr{round(p1["tau_end"],2)})$
-  & (S15)
-  & $\obsOverdispersion = \Sexpr{signif(h2_fit$h2_params['sigma'], 3)}$
-  & (S16)
-  & $\obsOverdispersion = \Sexpr{myround(p3s["k"],2)}$
-  & (S19)
-\\
-\hline
-\multirow{2}{*}{Initial Values} & $I_{0,0} = \Sexpr{round(prod(p1[c('pop_0', 'I_0')]))}$ & & \multicolumn{1}{c}{\multirow{2}{*}{\hspace{16pt}\missing}} & & \multirow{2}{*}{$I_{3,7,10;0}(0)= (
-        \Sexpr{round(prod(p3[c('H3', 'Iinit3')]))}, \Sexpr{round(prod(p3[c('H7', 'Iinit7')]))}, \Sexpr{round(prod(p3[c('H10', 'Iinit10')]))})^*$} & \\
-& $E_{0,0} = \Sexpr{round(prod(p1[c('pop_0', 'E_0')]))}$ & & \multicolumn{1}{c}{} & & & \\\hline
-Hurricane & \multicolumn{1}{c}{\multirow{2}{*}{\hspace{16pt}\missing}} & & \multicolumn{1}{c}{\multirow{2}{*}{\hspace{16pt}\missing}} & & $\Whur{3, 9} = (\Sexpr{myround(p3['aHur3'], 2)}, \Sexpr{myround(p3['aHur9'], 2)})^*$ & \multirow{2}{*}{\eqref{eq:model3:foi}} \\
-Parameters & \multicolumn{1}{c}{} & & \multicolumn{1}{c}{} & & $\hHur{3, 9} = (\Sexpr{myround(p3['hHur3'], 2)}, \Sexpr{myround(p3['hHur9'], 2)})^*$ & \\\hline
-
-\end{tabular}
-\caption{References to the relevant equation are given in parentheses.
-Parameters in blue were fixed based on scientific reasoning and not fitted to the data.
-$^*$ denotes parameters added during our re-analysis, not considered by Lee et al.
-Translations back into the notation of \citet{lee20} are given in Table~S1.
-}
-\end{table}
-
-\subsection{Corroborating Fitted Models with Previous Scientific Knowledge}\label{sec:science}
-
-The resulting mechanisms in a fitted model can be compared to current scientific knowledge about a system.
-Agreement between model based inference and our current understanding of a system may be taken as a confirmation of both model based conclusions and our scientific understanding.
-On the other hand, comparisons may generate unexpected results that have the potential to spark new scientific knowledge \citep{ganusov16}.
-
-In the context of our case study, we demonstrate how the fit of Model~1 corroborates other evidence concerning the role of rainfall in cholera epidemics.
-Specifically, we examine the results of fitting the flexible cubic spline term in Model~1 (Eq.~\myeqref{model1:lambda}--\myeqref{model1:beta}).
-The cubic splines permit flexible estimation of seasonality in the force of infection, $\transmission(t)$.
-Fig.~\ref{fig:h1SeasRain} shows that the estimated seasonal transmission rate $\transmission$ mimics the rainfall dynamics in Haiti, despite Model~1 not having access to rainfall data.
-This is consistent with previous studies finding that rainfall played an important role in cholera transmission in Haiti \citep{lemaitre19,eisenberg13}.
-The estimated seasonality also features an increased transmission rate during the fall, which was noticed at an earlier stage of the epidemic \citep{rinaldo12}; the high transimission rate in the fall may also be a result of the increase transission that occured in the fall of 2016, when hurricane Matthew struck Haiti.
-
-<<Get Model 1 Seasonality>>=
-
-h1_seas_df <- get_h1_seasonality(h1)
-
-df_rain <- haitiRainfall %>%
-  pivot_longer(  # Convert data to "long" format
-    data = .,
-    cols = -date,
-    names_to = 'dep',
-    values_to = 'rainfall'
-  ) %>%
-  mutate(year = lubridate::year(date), week = lubridate::week(date)) %>%
-  group_by(year, week) %>%
-  summarize(nwr = sum(rainfall)) %>%
-  ungroup() %>%  # Not needed, but good practice
-  mutate(
-    date_week_start = lubridate::ymd(paste0(2014, "-01-01")) +
-      lubridate::weeks(week - 1),
-    week_date = as.Date(date_week_start, format = "%m-%d"),
-    std_national_rain = (nwr - min(nwr)) / (max(nwr) - min(nwr))
-  )
-
-# Calculate the average national weekly rainfall.
-mean_rain <- df_rain %>%
-  group_by(week_date) %>%
-  summarize(mean_rain = mean(std_national_rain)) %>%
-  ungroup() %>%
-  mutate(mean_rain = (mean_rain - min(mean_rain)) / ((max(mean_rain) - min(mean_rain))))
-
-# Create a figure showing annual rainfal, scaled from 0-1. This is top figure
-gg_rain <- df_rain %>%
-  filter(week != 1, week != max(week)) %>%
-  ggplot(aes(x = week_date, y = std_national_rain, group = year, col = year)) +
-  geom_line() +
-  theme_bw() +
-  guides(color = 'none') +
-  ylab('Standardized Weekly\nNational Rain') +
-  theme(  # Remove elements that are un-unsed becuase this is top figure
-    axis.title.x = element_blank(),
-    axis.ticks.x = element_blank(),
-    axis.text.x = element_blank(),
-    axis.text.y = element_text(size = 7.5),
-    axis.title.y = element_text(size = 9)
-  )
-
-# Create plot that shows the seasonality of Model 1, alongside the average
-# weekly national rainfall.
-gg_trans <- ggplot() +
-  geom_line(data = h1_seas_df, aes(x = date, y = trans_std), linetype = 'dashed') +
-  geom_line(
-    data = mean_rain,
-    aes(x = week_date + lubridate::years(6), y = mean_rain),
-    col = '#2166ac'
-  ) +
-  scale_x_date(date_labels = "%b", date_breaks = '1 month') +
-  ylab('Standardized Seasonal\nContact Rate') +
-  theme(axis.title.x = element_blank(),
-        axis.text = element_text(size = 7.5),
-        axis.title.y = element_text(size = 9))
-
-@
-
-\begin{figure}
-\centering
-<<Model1_Seasonality_Figure, fig.height=2.9, fig.width=5, fig.align='center'>>=
-cowplot::plot_grid(gg_rain, gg_trans, align = "v", ncol = 1)
-@
-\caption{\label{fig:h1SeasRain}
-(Top) weekly rainfall in Haiti, lighter colors representing more recent years.
-(Bottom) estimated seasonality in the transmission rate (dashed line) plotted alongside mean rainfall (solid line).
-The outsided effect of rainfall in the fall may be due to Hurricane Matthew, which struck Haiti in October of 2016 and resulted in an increase of cholera cases in the nation.
-}
-\end{figure}
-
-For any model-based inference, it is important to recognize and assess the modeling simplifications and assumptions that were used in order to arrive at the conclusions.
-In epidemiological studies, for example, quantitative understanding of individual-level processes may not perfectly match model parameters that were fit to population-level case counts, even when the model provides a strong statistical fit \citep{he10}.
-This makes direct interpretation of estimated parameters delicate.
-
-Our case study provides an example of this in the parameter estimate for the duration of natural immunity due to cholera infection, $\muRS^{-1}$.
-Under the framework of Model~2, the best estimate for this parameter is $\Sexpr{signif(1/h2_fit$h2_params["sigma"],3)} \, \mathrm{yr}$, suggesting that individuals have effectively permanent immunity to cholera once infected.
-To interpret this result, we bear in mind that the data ranged from 2010-2019, and therefore estimates of immunity longer than $10\, \mathrm{yr}$---the upper end of previous estimates of natural immunity \citep{king08}---effectively result in the same model dynamics.
-The depletion of susceptible individuals may also be attributed to confounding mechanisms---such as localized vaccination programs and non-pharmaceutical interventions that reduce cholera transmission \citep{trevisin22, rebaudet21}---that were not accounted for in the model.
-Perhaps the best interpretation of the estimated parameter, then, is that under model assumptions, the model most adequately describes the observed data by having a steady decrease in the number of susceptible individuals.
-The weak statistical fit of Model~2 compared to a log-linear benchmark (see Table~\ref{tab:likes}) cautions us against drawing quantitative conclusions from this model.
-A model that has a poor statistical fit may nevertheless provide a useful conceptual framework for thinking about the system under investigation.
-However, a claim that the model has been validated against data should be reserved for situations where the model provides a statistical fit that is competitive against alternative explanations.
-
-\section{Robust interpretation of model based conclusions}\label{sec:advice}
-
-A model which aspires to provide quantitative guidance for assessing interventions should provide a quantitative statistical fit for available data.
-However, strong statistical fit does not guarantee a correct causal structure: it does not even necessarily require the model to assert a causal explanation.
-A causal interpretation is strengthened by corroborative evidence.
-For example, reconstructed latent variables (such as numbers of susceptible and recovered individuals) should make sense in the context of alternative measurements of these variables \citep{grad12}.
-In the context of our case study, by reconstructing the latent states of Model~3, we notice that the calibrated model favors higher levels of cholera transmission than what was typically observed in the incidence data (Sec.~S5.1 of the supplement).
-This result hints at the possibility of model mispecification, and should prompt us to be sceptical of the reliability of forecasts from this model.
-Similarly, parameters that have been calibrated to data should make sense in the context of alternative lines of evidence about the phenomena being modeled, while making allowance for the possibility that the interpretations of parameters may vary when modeling across differing spatial scales.
-
-If a mechanistic model including a feature (such as a representation of a mechanism, or the inclusion of a covariate) fits better than mechanistic models without that feature, and also has competitive fit compared to associative benchmarks, this may be taken as evidence supporting the scientific relevance of the feature.
-As for any analysis of observational data, we must be alert to the possibility of confounding.
-For a covariate, this shows up in a similar way to regression analysis: the covariate under investigation could be a proxy for some other unmodeled or unmeasured covariate.
-For a mechanism, the model feature could in principle explain the data by helping to account for some different unmodeled phenomenon.
-In the context of our analysis, the estimated trend in transmission rate could be explained by any trending variable (such as hygiene improvements, or changes in population behavior), resulting in confounding from collinear covariates.
-Alternatively, the trend could be attributed to a decreasing reporting rate rather than decreasing transmission rate, resulting in confounded mechanisms.
-The robust statistical conclusion is that a model which allows for change fits better than one which does not---we argue that a decreasing transmission rate is a plausible way to explain this, but the incidence data themselves do not provide enough information to pin down the mechanism.
-
-\section{Discussion}\label{sec:discussion}
-
-The ongoing global COVID-19 pandemic has provided a clear example on how government policy may be affected by the conclusions of scientific models \citep{saltelli20}.
+Probability of elimination estimates of this form are not meaningful for deterministic models, as the trajectory of these models only represent the mean behavior of the system rather than individual potential outcomes.
+We therefore do not provide probability of elimination estimates under Model~2, but show trajectories under the various vaccination scenarios using this model (Fig.~\ref{fig:mod2Traj}).
+
+\section*{Discussion}\label{sec:discussion}
+
+The ongoing global COVID-19 pandemic has provided a clear example on how government policy may be affected by the conclusions of scientific models \cite{saltelli20}.
 This article demonstrates that fitting appropriate scientific models to guide policy is a challenging statistical task.
 In our case study, we found that additional attention to statistical details could have resulted in improved statistical fits to the observed incidence data, leading to improvements in the accuracy of the resulting policy guidance.
-We used the same data and models, and even much of the same code, as \citet{lee20}, and yet ended up with drastically different conclusions.
+We used the same data and models, and even much of the same code, as Lee et al.~\cite{lee20}, and yet ended up with drastically different conclusions.
 We acknowledge the benefit of hindsight: our demonstration of a statistically principled route to obtain better-fitting models resulting in more robust insights does not rule out the possibility of discovering other models that fit well yet predict poorly.
@@
 Studies that combine model development with thoughtful data analysis, supported by a high standard of reproducibility, build knowledge about the system under investigation.
-Cautionary warnings about the difficulties inherent in understanding complex systems \citep{saltelli20,ioannidis20,ganusov16} should motivate us to follow best practices in data analysis, rather than avoiding the challenge.
-
-\subsection{Reproducibility and Extendability}
-
-\citet{lee20} published their code and data online, and this reproducibility facilitated our work.
+Cautionary warnings about the difficulties inherent in understanding complex systems \cite{saltelli20,ioannidis20,ganusov16} should motivate us to follow best practices in data analysis, rather than avoiding the challenge.
+
+\subsection*{Reproducibility and Extendability}
+
+Lee et al.~\cite{lee20} published their code and data online, and this reproducibility facilitated our work.
 Robust data analysis requires not only reproducibility but also extendability: if one wishes to try new model variations, or new approaches to fitting the existing models, or plotting the results in a different way, this should not be excessively burdensome.
-Scientific results are only trustworthy so far as they can be critically questioned, and an extendable analysis should facilitate such examination \citep{gentleman07}.
-
-We provide a strong form of reproducibility, as well as extendability, by developing our analysis in the context of a software package, \code{haitipkg}, written in the R language \citep{r}.
+Scientific results are only trustworthy so far as they can be critically questioned, and an extendable analysis should facilitate such examination \cite{gentleman07}.
+
+We provide a strong form of reproducibility, as well as extendability, by developing our analysis in the context of a software package, \code{haitipkg}, written in the R language \cite{r}.
 Using a software package mechanism supports documentation, standardization and portability that promote extendability.
-In the terminology of \citet{gentleman07}, the source code for this article is a {\it dynamic document} combining code chunks with text.
+In the terminology of Gentleman and Temple Lang~\cite{gentleman07}, the source code for this article is a {\it dynamic document} combining code chunks with text.
 In addition to reproducing the article, the code can be extended to examine alternative analysis to that presented.
-The dynamic document, together with the R packages, form a {\it compendium}, defined by \citet{gentleman07} as a distributable and executable unit which combines data, text and auxiliary software (the latter meaning code written to run in a general-purpose, portable programming environment, which in this case is R).
-
-\begin{acks}[Acknowledgments]
+The dynamic document, together with the R packages, form a {\it compendium}, defined by Gentleman and Temple Lang~\cite{gentleman07} as a distributable and executable unit which combines data, text and auxiliary software (the latter meaning code written to run in a general-purpose, portable programming environment, which in this case is R).
+
+\section*{Supporting information}
+
+\paragraph*{S1 Fig.}
+\label{S_Mod1}
+{\bf Model~1.} Flow chart representation of Model~1.
+
+\paragraph*{S2 Fig.}
+\label{S_Mod2}
+{\bf Model~2.} Flow chart representation of Model~2.
+
+\paragraph*{S3 Fig.}
+\label{S_Mod3}
+{\bf Model~3.} Flow chart representation of Model~3.
+
+\paragraph*{S1 Table}
+\label{S1_Table}
+{\bf Notation conversion table.} Conversions between the notation used here and the notation of Lee et al. (2020) \cite{lee20}.
+
+\paragraph*{S1 Text}
+\label{S1_Text}
+{\bf Model details.} In depth description of the Markov chain and differential equation interpretations of compartment flow rates, as well as details on the numeric implementation of these models.
+
+\paragraph*{S2 Text}
+\label{S_init}
+{\bf Initialization models.} Additional details of the initialization model that were used for Models~1--3.
+
+\paragraph*{S3 Text}
+\label{S_meas}
+{\bf Measurement models.} Additional details of the measurement model that were used for Models~1--3.
+
+\paragraph*{S4 Text}
+\label{S_lee20}
+{\bf Likelihood of Lee et al. models.} Details of how the log-likelihood of the Lee et al. models were calculated.
+
+\paragraph*{S5 Text}
+\label{S_mod3cal}
+{\bf Likelihood of Lee et al. models.} Additional details on the procedures for fitting and diagnosing Model~3.
+
+\paragraph*{S6 Text}
+\label{S_uncertain}
+{\bf Accounting for parameter uncertainty} A description of a empirical-Bayes approach that can be used to account for parameter uncertainty.
+
+\section*{Acknowledgments}
+
+This work was supported by National Science Foundation grants DMS-1761603 and DMS-1646108.
 The authors would like to thank Mercedes Pascual and Betz Halloran for helpful discussions. Laura Matrajt provided additional data for the Model~2 analysis.
-\end{acks}
-
-\begin{funding}
-This work was supported by National Science Foundation grants DMS-1761603 and DMS-1646108.
-%
-% The second author was supported in part by ...
-\end{funding}
-
-\begin{supplement}
-\stitle{Eliminating cholera in Haiti: Supplement}
-\sdescription{
-This document contains additional details about the numerical implementations of Models~1--3, as well as a translation table that facilitates comparisons between these models and those described by \cite{lee20}.
-The supplement also demonstrates our capability to faithfully estimate the likelihood of the models of \cite{lee20}.
-}
-\end{supplement}
-
-\begin{supplement}
-\stitle{Auxiliary software: \code{haitipkg}}
-\sdescription{
-The \code{haitipkg} R package is maintained in a GitHub repository, \code{jeswheel/haitipkg}. The submitted version of this package is archived on Zenodo: \code{10.5281/zenodo.7557099}
-The package contains all of the data and code used to create, fit and explore the models presented in this article.
-}
-\end{supplement}
-
-\begin{supplement}
-\stitle{Dynamic document: \code{haiti\_article}}
-\sdescription{
-The Rnw \citep[R noweb,][]{ramsey94} files generating this article and its supplement are maintained in a GitHub repository, \code{jeswheel/haiti\_article}.
-The submitted version will be archived on Zenodo.
-}
-\end{supplement}
-
-\bibliographystyle{imsart-nameyear}
+
+\nolinenumbers
+
 \bibliography{bib-haiti}

### 2023-09-11 si/calibrateMod3.Rnw
Removed unused ncores argument

@@
       search_Iinit = FALSE,
-      search_hur = FALSE,
-      ncores = cores
+      search_hur = FALSE
     )

### 2023-09-25 checkRes.R
removed unused files

@@ -1,83 +1,0 @@
-# Checking RL2 results.
-#
-# Jull 11
-
-library(pomp)
-library(spatPomp)
-library(tidyverse)
-
-results <- readRDS("model3/run_level_2/h3_revision.rds")
-# results3 <- readRDS("model3/run_level_3/h3_revision.rds")
-
-s1_traces <- melt(results$search1$traces) %>%
-  rename(L1 = `.L1`) %>%
-  mutate(iteration = as.numeric(iteration))
-
-s2_traces <- melt(results$search2$traces) %>%
-  rename(L1 = `.L1`) %>%
-  # filter(as.numeric(iteration) != 0) %>%
-  mutate(iteration = as.numeric(iteration) + 100)
-
-s3_traces <- melt(results$search3$traces) %>%
-  rename(L1 = `.L1`) %>%
-  # filter(as.numeric(iteration) != 0) %>%
-  mutate(iteration = as.numeric(iteration) + 200)
-
-unit_specific <- c("betaB", "foi_add", "aHur", "hHur")
-
-shared <- c(paste0(c(
-  "mu_B", "XthetaA", "thetaI", "lambdaR", "r", "std_W",
-  "epsilon", "k"
-), "1"), 'loglik')
-
-s1_topn <- order(-results$search1$logLiks$logLik)[1:9]
-s2_topn <- order(-results$search2$logLiks$logLik)[1:6]
-s3_top10 <- order(-results$search3$logLiks$logLik)[1:10]
-
-s2_ancestor <- function(id) {
-  s1_topn[((id - 1) %/% 4) + 1]
-}
-
-s3_ancestor2 <- function(id) {
-  s2_topn[((id - 1) %/% 6) + 1]
-}
-
-s3_ancestor1 <- function(id) {
-  s2_ancestor(s3_ancestor2(id))
-}
-
-s2_traces %>%
-  mutate(ancest_1 = s2_ancestor(L1)) -> s2_traces
-
-s3_traces %>%
-  mutate(ancest_2 = s3_ancestor2(L1),
-         ancest_1 = s3_ancestor1(L1)) -> s3_traces
-
-merged_results <- bind_rows(
-  filter(s1_traces, L1 %in% unique(s3_traces$ancest_1)) %>%
-    mutate(L2 = L1),
-  filter(s2_traces, L1 %in% unique(s3_traces$ancest_2)) %>%
-    mutate(L2 = ancest_1) %>%
[…]

### 2023-09-28 ms.Rnw
Finished addressing Anna's + Kevin's comments

@@
 In this article, we discuss common modeling decisions that may not provide an adequate statistical explanation of the data, and we show that attention to statistical fit can lead to more accurate policy evaluations.
-We do this by conducting a re-analysis of the 2012-2019 cholera outbreak in Haiti.
+We do this by conducting a re-analysis of the 2010 cholera outbreak in Haiti.
 Our analysis presents methodology for diagnosing how well a model describes observed data, and provides careful justification of valid conclusions from the fitted model.
-Objective measures are used to benchmark model fit; when these are combined with reproducibility, a framework emerges for continual improvment when revisiting the data and models.
+Objective measures are used to benchmark model fit; when these are combined with reproducibility, a framework emerges for continual improvement when revisiting the data and models.
 Our data analysis workflow is supported by recent advances in algorithms, software and hardware, which facilitate statistical fitting of nonlinear stochastic dynamic models to observed incidence data.
@@
 Their forecasts, however, were inconsistent with the prolonged period with no confirmed cholera cases between February, 2019 and September, 2022 \cite{trevisin22}.
-Though cholera has recently reemerged in Haiti \cite{rubin22}, the inability to accurately forecast cholera incidence from 2019-2022---while diligently adhering to existing scientific standards for infectious disease modeling---demonstrates that existing standards of disease modeling and forecasting may be insufficient.
+Though cholera has recently reemerged in Haiti \cite{rubin22,PAHO23}, the inability to accurately forecast cholera incidence from 2019-2022---while diligently adhering to existing scientific standards for infectious disease modeling---demonstrates that existing standards of disease modeling and forecasting may be insufficient.
 
@@
 While this work is focused on the models proposed by Lee et al.~\cite{lee20}, our suggestions have broader relevance.
-To investigate the extent to which \cite{lee20} is typical of the substantial body of work on Haiti cholera dynamics, we preformed a literature review by searching PubMed with keywords: Haiti, cholera, model.
+To investigate the extent to which \cite{lee20} is typical of the substantial body of work on Haiti cholera dynamics, we performed a literature review by searching PubMed with keywords: Haiti, cholera, model.
 The search resulted in 66 papers, of which 32 used dynamic models to describe the cholera epidemic in Haiti.
@@
 \hline
-\multirow{2}{*}{Birth/death (yr)}
-  & ${\muBirth^{-1} = \Sexpr{myround(1/p1["mu"]/52,1)}}\fixed$
+\multirow{2}{*}{Birth/death (yr\textsuperscript{-1})}
+  & ${\muBirth =  10^{-2}\times \Sexpr{myround(100 * 52 * p1["mu"],2)}}\fixed$
     %
@@
   &
-  & ${\muDeath^{-1} = \Sexpr{myround(1/p3["mu1"],1)}}\fixed$
+  & ${\muDeath = 10^{-2}\times\Sexpr{myround(100 * p3["mu1"],2)}}\fixed$
   & \multirow{2}{*}{\eqref{eq:model3:IS}}
 \\
-& ${\muDeath^{-1} = \Sexpr{myround(1/p1["delta"]/52,1)}}\fixed$
+& ${\muDeath = 10^{-3}\times\Sexpr{myround(1000*52*p1["delta"],1)}}\fixed$
 & \eqref{model1:death}
@@
 &
-& ${\choleraDeath^{-1} = \Sexpr{myround(1/p3["alpha1"],2)}}\fixed$
+& ${\choleraDeath = \Sexpr{myround(p3["alpha1"],2)}}\fixed$
 &
@@
 \begin{flushleft} References to the relevant equation are given in parentheses.
-Parameters that were fixed and not calibrated using the data are indicated with $\fixed$;
-parameters that were added during our re-analysis and were not considered by Lee et al. are indicated with $^*$.
+Parameters that were fixed and not calibrated using the data are indicated with $\fixed$; all fixed parameters values were chosen to match the fixed parameter values of \cite{lee20}.
+Parameters that were added during our re-analysis and were not considered by Lee et al. are indicated with $^*$.
 Translations back into the notation of \cite{lee20} are given in \nameref{S1_Table}.
@@
 In this article we calibrate the parameters of each of our models by maximizing the likelihood, as described in Eq.~\myeqref{eq:likedef}.
-The likelihood for each of the fitted models---and the corresponding AIC values for model comparisons that include an adjustment for the number of calibrated parameters---is provided in Table~\ref{tab:likes}
+The likelihood for each of the fitted models---and the corresponding AIC values for model comparisons that include an adjustment for the number of calibrated parameters---is provided in Table~\ref{tab:likes}.
 In the following subsections we describe in detail our approach to calibrating the three proposed mechanistic models to observed cholera incidence data.
@@
 
-One of the most frequently used approaches to calibrate the parameters of a statistical model is maximum likelihood estimation.
-In the case of a POMP model, the likelihood function---as described in Eq.~\myeqref{eq:likedef}---is written as $\mathcal{L}(\paramVec) = \int f_{\bm{X}_{0:N}, \bm{Y}_{1:N}}\big(\bm{x}_{0:N}, \bm{y}_{1:N};\paramVec\big) d\bm{x}_{0:N}$.
-The high-dimensional integral in the likelihood, and the potentially non-linear nature of the process model, make it impossible to obtain a closed form expression of the likelihood function of POMP models except in the most simple cases.
-The process model that defines Model~1 was developed with the intent of providing a plausible description of cholera dynamics rather than a providing statistically convenient description.
-Because of this, a closed form expression of the likelihood of Model~1 is not readily available.
-
-In order to retain the ability to propose models that are scientifically meaningful rather than only those that are simply statistically convenient, we restrict ourselves to parameter estimation techniques that have the plug-and-play property, which is that the fitting procedure only requires the ability to simulate the latent process instead of evaluating transition densities \cite{breto09,he10};
+Model~1 was developed with the intent of providing a description of cholera dynamics that is scientifically plausible rather than one that is statistically convenient.
+This results in the inability to obtain a closed form expression of the non-linear joint model density---described in Eq.~\myeqref{eq:jointLik}.
+Therefore in order to perform likelihood based inference on this model, we are restricted to use parameter estimation techniques that have the \emph{plug-and-play} property, which is that the fitting procedure only requires the ability to simulate the latent process rather than evaluating transition densities \cite{breto09,he10};
 in the context of the notation and definitions employed in this article, this means that we only require the ability to simulate from $f_{\bm{X}_0}\big(\bm{x}_0;\paramVec\big)$ and $f_{\bm{X}_n|\bm{X}_{n-1}}\big(\bm{x}_{n}|\bm{x}_{n-1}; \paramVec\big)$ rather than needing to evaluate these densities.
@@
 To our knowledge, the only plug-and-play methods that can be used to maximize the likelihood for POMP models of this complexity are iterated filtering algorithms \cite{ionides15}, which modify the well-known \emph{particle filter} \cite{arulampalam02}.
-The particle filter is often used to obtain values of latent states, but can also be used to obtain likelihood estimates of a POMP model.
-Iterated filtering algorithms such as IF2 \cite{ionides15} extend the particle filter by performing a random walk for each parameter and particle;
+
+The particle filter, also referred to as sequential Monte Carlo, is a simulation based method that is frequently used in Bayesian inference to approximate the posterior distribution of latent states.
+This algorithm can also be used to accurately approximate the log-likelihood of a POMP model, defined as the integral in Eq.~\myeqref{eq:likedef}.
+Iterated filtering---algorithms such as IF2 \cite{ionides15}---extend the particle filter by performing a random walk for each parameter and particle;
 these perturbations are carried out iteratively over multiple filtering operations, using the collection of parameters from the previous filtering pass as the parameter initialization for the next iteration, and decreasing the random walk variance at each step.
+With a sufficient number of iterations, the resulting parameter values converge to a region of the parameter space that maximizes the model likelihood.
+
+% One of the most frequently used approaches to calibrate the parameters of a statistical model is maximum likelihood estimation.
+% In particular, a closed form expression of the joint model density---described in Eq.~\myeqref{eq:jointLik}---is not readily available, making likelihood based inference a challenging task.
+% In the case of any POMP model, the likelihood function---as described in Eq.~\myeqref{eq:likedef}---is written as $\mathcal{L}(\paramVec) = \int f_{\bm{X}_{0:N}, \bm{Y}_{1:N}}\big(\bm{x}_{0:N}, \bm{y}_{1:N};\paramVec\big) d\bm{x}_{0:N}$.
+% The high-dimensional integral in the likelihood, and the potentially non-linear nature of the process model, make it impossible to obtain a closed form expression of the likelihood function of POMP models except in the most simple cases.
+% Model~1 was developed with the intent of providing a description of cholera dynamics that is scientifically plausible rather than on that is simply statistically convenient.
+% Because of this, a closed form expression of the likelihood of Model~1 is not readily available.
+
+% In order to retain the ability to propose models that are scientifically meaningful rather than only those that are simply statistically convenient, we restrict ourselves to parameter estimation techniques that have the plug-and-play property, which is that the fitting procedure only requires the ability to simulate the latent process instead of evaluating transition densities \cite{breto09,he10};
+% in the context of the notation and definitions employed in this article, this means that we only require the ability to simulate from $f_{\bm{X}_0}\big(\bm{x}_0;\paramVec\big)$ and $f_{\bm{X}_n|\bm{X}_{n-1}}\big(\bm{x}_{n}|\bm{x}_{n-1}; \paramVec\big)$ rather than needing to evaluate these densities.
+% Plug-and-play algorithms include Bayesian approaches like ABC and PMCMC \cite{toni09,andrieu10}, but here we use algorithms that enable maximum likelihood estimation.
+% To our knowledge, the only plug-and-play methods that can be used to maximize the likelihood for POMP models of this complexity are iterated filtering algorithms \cite{ionides15}, which modify the well-known \emph{particle filter} \cite{arulampalam02}.
+% The particle filter is often used to obtain values of latent states, but can also be used to obtain likelihood estimates of a POMP model.
+% Iterated filtering algorithms such as IF2 \cite{ionides15} extend the particle filter by performing a random walk for each parameter and particle;
+% these perturbations are carried out iteratively over multiple filtering operations, using the collection of parameters from the previous filtering pass as the parameter initialization for the next iteration, and decreasing the random walk variance at each step.
 
@@
 \end{figure}
+
+If a mechanistic model including a feature (such as a representation of a mechanism, or the inclusion of a covariate) fits better than mechanistic models without that feature, and also has competitive fit compared to associative benchmarks, this may be taken as evidence supporting the scientific relevance of the feature.
+As for any analysis of observational data, we must be alert to the possibility of confounding.
+For a covariate, this shows up in a similar way to regression analysis: the covariate under investigation could be a proxy for some other unmodeled or unmeasured covariate.
+For a mechanism, the model feature could in principle explain the data by helping to account for some different unmodeled phenomenon.
+The statistical evidence of a trend in transmission rate in this model could be explained by any trending variable (such as hygiene improvements, or changes in population behavior), resulting in confounding from collinear covariates.
+Alternatively, the trend could be attributed to a decreasing reporting rate rather than decreasing transmission rate, resulting in confounded mechanisms.
+The robust statistical conclusion is that a model which allows for change fits better than one which does not---we argue that a decreasing transmission rate is a plausible way to explain this, but the incidence data themselves do not provide enough information to pin down the mechanism.
 
@@
 Standard statistical models, interpreted as associative models without requiring any mechanistic interpretation of their parameters, provide suitable benchmarks.
-Examples include linear regression, auto-regressive moving average time series models, or even independent and identically distributed measurements.
+Examples include linear regression, auto regressive moving average time series models, or even independent and identically distributed measurements.
 Benchmarks enable us to evaluate the goodness of fit that can be expected of a suitable mechanistic model.
@@
 the theory and practice of ARMA models is well developed, and these linear models are appropriate on a log scale due to the exponential growth and decay characteristic of biological dynamics.
-We chose to use the auto-regressive negative binomial model, however, because the large number of weeks with zero recorded cholera cases in department level data makes a benchmark based on a continuous distribution problematic.
+We chose to use the auto regressive negative binomial model, however, because the large number of weeks with zero recorded cholera cases in department level data makes a benchmark based on a continuous distribution problematic.
 Likelihoods of Models~1--3 and their respective ARMA benchmark models are provided in Table~\ref{tab:likes}.
@@
 The depletion of susceptible individuals may also be attributed to confounding mechanisms---such as localized vaccination programs and non-pharmaceutical interventions that reduce cholera transmission \cite{trevisin22, rebaudet21}---that were not accounted for in the model.
-Perhaps the best interpretation of the estimated parameter, then, is that under model assumptions, the model most adequately describes the observed data by having a steady decrease in the number of susceptible individuals.
+Perhaps the best interpretation of the estimated parameter, then, is that under the modeling framework that was used, the model most adequately describes the observed data by having a steady decrease in the number of susceptible individuals.
 The weak statistical fit of Model~2 compared to a log-linear benchmark (see Table~\ref{tab:likes}) cautions us against drawing quantitative conclusions from this model.
@@
 However, a claim that the model has been validated against data should be reserved for situations where the model provides a statistical fit that is competitive against alternative explanations.
-
-\subsection*{Robust interpretation of model based conclusions}\label{sec:advice}
-
-In this section, we make some general points on deriving scientific conclusions from statistical analysis of mechanistic models, and connect these points to our specific case study.
 
@@
 
-If a mechanistic model including a feature (such as a representation of a mechanism, or the inclusion of a covariate) fits better than mechanistic models without that feature, and also has competitive fit compared to associative benchmarks, this may be taken as evidence supporting the scientific relevance of the feature.
-As for any analysis of observational data, we must be alert to the possibility of confounding.
-For a covariate, this shows up in a similar way to regression analysis: the covariate under investigation could be a proxy for some other unmodeled or unmeasured covariate.
-For a mechanism, the model feature could in principle explain the data by helping to account for some different unmodeled phenomenon.
-In the context of our analysis, the statistical evidence of a trend in transmission rate in Model~1 could be explained by any trending variable (such as hygiene improvements, or changes in population behavior), resulting in confounding from colinear covariates.
-Alternatively, the trend could be attributed to a decreasing reporting rate rather than decreasing transmission rate, resulting in confounded mechanisms.
-The robust statistical conclusion is that a model which allows for change fits better than one which does not---we argue that a decreasing transmission rate is a plausible way to explain this, but the incidence data themselves do not provide enough information to pin down the mechanism.
-
 \section*{Results}
@@
 That may explain why Lee et al.~\cite{lee20} chose to obtain forecasts by simulating the calibrated models forward from initial conditions.
-Here, we obtain forecasts by simulating future values using latent states that are consistent with the most recent observations by drawing latent states at time the last observation time $t_N$ from the filtering distribution $f_{\bm{X}_N|\bm{Y}_{1:N}}(\bm{x}_{N} | \bm{y}^*_{1:N} ; \hat\paramVec)$,
-where $\hat\paramVec$ is a vector of calibrated parameters.
+We obtain forecasts by simulating future values using latent states that are consistent with the most recent observations; this is done by drawing latent states at the last observation time ($t_N$) from the filtering distribution $f_{\bm{X}_N|\bm{Y}_{1:N}}(\bm{x}_{N} | \bm{y}^*_{1:N} ; \hat\paramVec)$.
 The decision not to do this partially explains the unsuccessful forecasts of Lee et al.~\cite{lee20}: their Table~S7 shows that the subset of their simulations which were consistent with observing zero cases in 2019 also accurately predicted the prolonged absence of detected cholera.

### 2023-09-28 append_bib.R
commiting R file used by Makefile

@@ -1,0 +1,15 @@
+args <- commandArgs(trailingOnly = TRUE)
+
+append_bib <- function(file_name) {
+  tex_file <- readLines(file_name)
+
+  bib_spot <- which(grepl("\\bibliography{bib-haiti}", tex_file, fixed = TRUE))
+  bib_content <- readLines(gsub("\\.tex", "\\.bbl", file_name))
+
+  new_file <- c(tex_file[1:(bib_spot - 1)], bib_content, tex_file[(bib_spot + 1):length(tex_file)])
+
+  writeLines(new_file, con = file_name)
+  # file.rename(from = file_name, to = new_name)
+}
+
+append_bib(args[1])

### 2023-10-04 ArXiv.R
created make command for ArXiv submission

@@ -1,0 +1,47 @@
+args <- commandArgs(trailingOnly = TRUE)
+
+ms_to_ArXiv <- function(file_name) {
+  tmp_file <- readLines(file_name)
+  tmp_file <- gsub("figure/", "", tmp_file)
+  tmp_file <- gsub("\\linenumbers", "", tmp_file, fixed = TRUE)
+  tmp_file <- gsub("\\nolinenumber", "", tmp_file, fixed = TRUE)
+  lineno_pkg <- which(grepl('lineno', tmp_file))
+  tmp_file <- c(tmp_file[1:(lineno_pkg - 1)], tmp_file[(lineno_pkg + 1):length(tmp_file)])
+
+  # geometry_line <- which(grepl("{geometry}", tmp_file, fixed = TRUE))
+  # tmp_file[geometry_line] <- "\\usepackage[top=0.85in,left=1.5in,footskip=0.75in]{geometry}"
+  # tmp_file <- gsub("bib-haiti", 'ms', tmp_file)
+
+  bib_spot <- which(grepl("\\bibliography{bib-haiti}", tmp_file, fixed = TRUE))
+  bib_content <- readLines(gsub("\\.tex", "\\.bbl", file_name))
+
+  new_file <- c(tmp_file[1:(bib_spot - 1)], bib_content, tmp_file[(bib_spot + 1):length(tmp_file)])
+
+  writeLines(new_file, con = file_name)
+}
+
+si_to_ArXiv <- function() {
+  tmp_file <- readLines("ArXiv/si.tex")
+  tmp_file <- gsub("inputs/", "", tmp_file)
+  tmp_file <- gsub("figure/", "", tmp_file)
+  # tmp_file <- gsub("../bib-haiti", 'ms', tmp_file)
+
+  input_lines <- which(grepl("\\\\input\\{([^}]*)\\}", tmp_file))
+  input_files <- paste0(gsub("\\\\input\\{|\\}", "", tmp_file[input_lines]), ".tex")
+
+  for (i in 1:length(input_files)) {
+    tmp_input_file <- readLines(paste0("ArXiv/", input_files[i]))
+    tmp_input_file <- gsub("figure/", "", tmp_input_file)
+    writeLines(tmp_input_file, con = paste0("ArXiv/", input_files[i]))
+  }
+
+  bib_spot <- which(grepl("bib-haiti", tmp_file, fixed = TRUE))
[…]

### 2023-10-06 ms.Rnw
Minor aesthetic changes to param table
Fixed minor typos

@@
-% Please do not include colors or graphics in the text.
-%
-% % % % % % % % % % % % % % % % % % % % % % %
-%
-% Tables should be cell-based and may not contain:
-% - spacing/line breaks within cells to alter layout or alignment
-% - do not nest tabular environments (no tabular environments within tabular environments)
-% - no graphics or colored text (cell background color/shading OK)
-% See http://journals.plos.org/plosone/s/tables for table guidelines.
-%
-% -- EQUATIONS, MATH SYMBOLS, SUBSCRIPTS, AND SUPERSCRIPTS
-%
-% IMPORTANT
-% Below are a few tips to help format your equations and other special characters according to our specifications.
-For more tips to help reduce the possibility of formatting errors during conversion, please see our LaTeX guidelines at http://journals.plos.org/plosone/s/latex
-%
-
 \documentclass[10pt,letterpaper]{article}
@@
 \\
-Water survival \newline (wk)
+\raggedright Water survival (wk)
   & \multirow{2}{*}{\missing}
@@
 \hline
-Process noise \newline (wk\textsuperscript{1/2})
-  & $\sigmaProc =$ \newline $(\Sexpr{myround(sqrt(p1["sig_sq_epi"]),2)},
+\raggedright Process noise (wk\textsuperscript{1/2})
+  & \raggedright $\sigmaProc = (\Sexpr{myround(sqrt(p1["sig_sq_epi"]),2)},
   \Sexpr{myround(sqrt(p1["sig_sq_end"]),2)})^*$
@@
 \\
-Observation \newline variance
-  & $\obsOverdispersion =$ \newline $(\Sexpr{myround(p1["tau_epi"],2)}, \Sexpr{round(p1["tau_end"],2)})$
+\raggedright Observation variance
+  & \raggedright $\obsOverdispersion = (\Sexpr{myround(p1["tau_epi"],2)}, \Sexpr{round(p1["tau_end"],2)})$
   & \multirow{2}{*}{(S15)}
@@
   &
-  & $\Whur{3, 9} =$\newline $(\Sexpr{myround(p3['aHur3'], 2)}, \Sexpr{myround(p3['aHur9'], 2)})^*$
+  & \raggedright $\Whur{3, 9} = (\Sexpr{myround(p3['aHur3'], 2)}, \Sexpr{myround(p3['aHur9'], 2)})^*$
   & \multirow{2}{*}{\eqref{eq:model3:foi}}
@@
   &
-  & $\hHur{3, 9} =$\newline $(\Sexpr{myround(p3['hHur3'], 2)}, \Sexpr{myround(p3['hHur9'], 2)})^*$
+  & \raggedright $\hHur{3, 9} = (\Sexpr{myround(p3['hHur3'], 2)}, \Sexpr{myround(p3['hHur9'], 2)})^*$
   & \multirow{2}{*}{\eqref{eq:model3:foi}}
@@
 \caption{\label{fig:betat}
-{\figTitle Confidence inverval for linear trend in transmission.}
+{\figTitle Confidence interval for linear trend in transmission.}
 Monte Carlo adjusted profile of $\transmissionTrend$ for Model~1.
@@
 \label{S_mod3cal}
-{\bf Likelihood of Lee et al. models.} Additional details on the procedures for fitting and diagnosing Model~3.
+{\bf Calibrating Model 3.} Additional details on the procedures for fitting and diagnosing Model~3.
 

### 2024-01-10 r1/response.tex
First commit with the response to reviewers

\documentclass[11pt]{article}

\input{../edits.tex}
%\usepackage[sectionbib]{natbib}

%\usepackage{xr}
%\externaldocument{../ms}

\renewcommand\thefigure{L-\arabic{figure}}
\renewcommand\thetable{L-\arabic{table}}
\renewcommand\thepage{L-\arabic{page}}
\renewcommand\theequation{L\arabic{equation}}

\newcommand\vsp{\vspace{2mm}}

\usepackage{fullpage}
\bibliographystyle{apalike}

\usepackage{natbib,amsmath,amsthm}
\usepackage{verbatim}
\usepackage{graphicx}
\usepackage{amsmath,amssymb}
\usepackage{amsfonts}
\usepackage{natbib}
 \usepackage{url}
\usepackage{color}

%\newcommand\ifLetter[2]{{#1}}

\parskip 7pt
\parindent 0pt
\newcommand\report[1]{{\color{mygreen} \vspace{1mm}\hspace{0.25in}\parbox{6in}{\em #1}}}
\newcommand\article[1]{{\color{blue} \vspace{1mm}\hspace{0.25in}\parbox{6in}{\em #1}}}
\newcommand\blue[1]{{\color{blue}{#1}}}

%\topmargin -0.3in % for EI dvips

\begin{document}

\rule{0cm}{0.3cm}

%\vspace{-28mm}
\vspace{-8mm}

% \includegraphics[width=0.85\textwidth,trim={1.7cm 25.6cm 10cm 1.3cm},clip]{/Users/ionides/Dropbox/computing/latex/umich-stat-letterhead-2019.pdf}

\vspace{10mm}

\begin{flushright}
\today
\end{flushright}

\noindent {\bf Re:  PCOMPBIOL-D-23-01609. ``Informing policy via dynamic models: Cholera in Haiti''}

\vspace{5mm}

Dear PLOS Computational Biology editorial board,

\TODO{Edit this letter}
Thank you for arranging the review of our manuscript, and for the invitation to submit a revision.
We are grateful for the thoughtful comments from the referees.
Our revised manuscript has various edits for clarity, following the referee suggestions, as well as some small corrections in the text.
The numerical results are unchanged.

To make space for some additional sentences, we cut a few sentences that seemed least essential. We also placed some of the new material in the supplement.

Below are point-by-point responses to the referee reports.
The reviewer comments are shown in green, italic type.
Our responses are in black, and material copied verbatim from the revision is in blue.

\vspace{3mm}

Sincerely,

% \vspace{-3mm}

% \hspace{3cm}\includegraphics[width=2cm,angle=90]{/Users/ionides/Dropbox/computing/latex/ei-signature1.pdf}

% \vspace{-3mm}

Jesse Wheeler \\
PhD Student, Statistics Department \\
University of Michigan\\
jeswheel@umich.edu

\vspace{5mm}

\newpage

\begin{center}
{\large{\textbf{\underline{Referee 1}}}} %1111111111111111
\end{center}

\report{
  The authors' manuscript aims to explain how public health decisions can best be informed by combining an understanding of the underlying pathogen transmission dynamics with epidemiological data. In particular, the authors make a number of recommendations of best practices that modellers can adopt when using mechanistic models to make statistical inferences from time series data. To make their recommendations more concrete, they perform a re-analysis of cholera incidence data from Haiti spanning the years 2014-2019. These data were previously analysed in a modelling study performed by a separate team of scientists, published in Lee et al. (2020), see Ref. 1. Using refinements of three models from Lee et al., the authors demonstrate a number of ways in which the analysis could have been improved. Furthermore, the authors present differences between the findings of their reanalysis and the original analysis of Lee et al., for instance in the predicted probability of eliminating cholera from Haiti (see Figure 8).
}

\report{
  My most substantial concern is with the framing of the paper, and whether that might mislead potential readers. To me, the paper serves more of an educational purpose, presenting ideas on how to best use mathematical models to gain insight from time series data. While the case study into cholera in Haiti is methodologically sound, it does not seem to be the main focus of the manuscript. To such an extent that, in the discussion, the only comment to on the Haiti analysis is that the authors ``used the same data and models, and even much of the same code, as Lee et al. [1], and yet ended up with drastically different conclusions."
}

\report{
  This major concern can be resolved in two ways, both of which seem acceptable to me, either a) reframe the paper as a more general series of recommendations for how to better perform statistical inference on time series data using mechanistic models (including a longer introduction and discussion of existing methodological deficiencies) or b) perform a more in-depth analysis and commentary on cholera in Haiti (some of which has already been done but is buried in supplementary information files that are only briefly mentioned in the main text, e.g. S5).
}

\report{
  I found the methodology used to be well-suited to addressing the aims of the paper.
}

\report{
  Overall, I found the paper interesting and informative, but a little unfocused. I expect with some heavy rewriting in places (especially the introduction and discussion), the paper will be acceptable for publication in PLoS Computational Biology. I do not anticipate the authors having to perform additional computational analysis.
}

\report{
  \textbf{\underline{Abstract}}

  The authors state they ``develop data analysis strategies leading to improved statistical fit." It's not clear to me exactly what novel strategies have been developed, it seems the authors exclusively use pre-existing techniques from the literature. This is fine, if the focus of the paper is either a) pedagogical or b) on understanding a specific disease system (cholera in Haiti). If any methods are novel, this should be more clearly stated.
}

\report{
  \textbf{\underline{Introduction}}

  To make clear that the issues being addressed by this paper go beyond one modelling study of Haiti, the authors should provide examples and references to support the (accurate in my opinion) statement about, ``common modeling decisions that may not provide an adequate statistical explanation of the data".
}

\report{
  Given the focus of the paper on how to better perform inference, explicitly commenting on what are the ``existing guidelines for creating models to inform policy [4, 5]". It would then also be helpful in the discussion to comment on how the recommendations of this study extend these pre-existing guidelines.
}

\report{
  \textbf{\underline{Methods}}

  Fig 2. Model parameters. I think this should be ``Table 1"?
}

\report{
    ``and $z \in 1:Z$ describes hypothetical vaccination programs" It would help to give an example of the different vaccine programs here. E.g. does this mean the number of doses the individual has received, or differences in the vaccine administered?
}

\report{
  Eq 6 and line 1: I think the notation for the asymptomatic fraction might be clearer and more consistent with the other models if you used $f_z(t)$ instead of $\vartheta_z(t)$. In fact, this is the notation used in the table in Figure 2.
}

\report{
  Line 198: Reference for where value of $v_{rate}$ comes from.
}

\report{
  Usually, vaccine efficacy = 1 implies the vaccine completely blocks transmission and 0 implies no effect. In model 2 the definition of vaccine efficacy (lines 200-205 and Eq 14) implies the opposite. I suggest re-parameterising the model, or renaming the variable. For what its worth, models 1 and 3 seem to follow the usual convention.
}

\report{
  I found the commentary on comparing fitted mechanistic models to a (statistical) benchmark model to be informative and helpful. I was wondering if the authors had any comments on whether (or not) to directly compare the AIC values of the different fitted models directly. Am I right in thinking this should be possible for models 2 and 3 given (I think) they are fitted to the same data? Model 1 is fitted to a different data set, which I understand makes such a comparison impossible.
}

\report{
  Lines 308-309: Isn't this a log-linear trend in the transmission rate?
}

\report{
  While the data support a linear trend at the $95\%$ confidence level, it's worth commenting on the magnitude of the trend, e.g. the total reduction in $\beta(t)$ over the period of study.
}

\report{
  I might have misunderstood it slightly, but I don't entirely follow the reasoning of lines 340-346. It seems that two different claims are being made: A) the model can’t identify which mechanism underlies a trend in $\beta(t)$ and B) the model can't definitively state there is a trend in $\beta(t)$ compared with some other time-varying parameter (e.g. reporting probability). For model 1, I agree with the authors that claims A and B are valid. What I take issue with is the final statement, which seems to be stronger and apply to both claims and mechanistic models more generally: ``we argue that a decreasing transmission rate is a plausible way to explain this, but the incidence data themselves do not provide enough information to pin down the mechanism." I agree that the ``incidence data themselves" (which I take to imply ``without additional covariates data") make claim A valid regardless of the model. However, I don’t think claim B holds regardless of the model. By changing model 1 to include time variation in the reporting rate I don't see \emph{a priori} why the ``the incidence data themselves" might not provide enough information to distinguish between time variation in transmission compared with time variation in reporting without the need for additional data. The deficiency is in model 1 rather than the information content of the data. I suggest the authors rework the paragraph to make the reasoning clearer.
}

\report{
  Lines 53-56: I push back against this point. The models assume an exponential distribution for vaccine-derived immunity. Assuming a mean duration of 10 years, the proportion of individuals who remain immune 9 years after vaccination is . Even for much larger durations of immunity a non-negligible fraction of the population will lose immunity after 9 years, e.g. . I therefore don't expect values for the duration of immunity around 10 years to ``effectively result in the same model dynamics".
}

\report{
  \textbf{\underline{Discussion}}

In contrast with the introduction, which almost exclusively focuses on cholera in Haiti, the discussion only mentions it once, briefly: ``We used the same data and models, and even much of the same code, as Lee et al. [1], and yet ended up with drastically different conclusions." It would help the reader to have those discrepancies summarised. It would be helpful to summarise the recommendations the authors make attempting to fit mechanistic models to time series data (even as a list). Along similar lines, it would be helpful to present a more concise summary of what the ``more accurate policy evaluations" found by using the approaches outlined in this study.
}

\report{
  \textbf{underline{Typos}}

  Eq. 1: Is there a missing star on $y_{1:N}$ on the right-hand-side?
}

\report{
  Line 146: delete ``the" before describing
}

\newpage

\begin{center}
{\large{\textbf{\underline{Referee 2}}}} %22222222222222
\end{center}

\report{
  This well-written article is a strong piece of work that will be useful to the readers of this journal and, in particular, to researchers who perform statistical inference on compartmental models of infectious diseases. Furthermore, this paper is enhanced by its transparency and reproducibility, as the provided code is well-documented and organised in an R package. Additionally, the supplementary information is a valuable resource for researchers in this field.
}

\report{
  In a nutshell, the authors argue that existing criteria to evaluate the validity of a disease model are insufficient. Therefore, they propose more stringent standards for evaluating models’ ability to fit the available data in order to obtain more reliable forecasts. The authors use a Cholera case study to outline their suggestions. To highlight, a key contribution from this work is the recommendation of employing inductive (associative) models as a goodness-of-fit benchmark, as evidenced by this sentence: “It should be universal practice to present measures of goodness of fit for published models, and mechanistic models should be compared against benchmarks”. Undoubtedly, this approach provides an objective measure to judge the ability of mechanistic models to fit the data. In the following sections, I express my opinion on how this paper may be improved.
}

\report{
\textbf{\underline{Major Concerns}}
}

\report{
  \textbf{\underline{1. Literature}}

While the arguments provided throughout the article are well-articulated, there needs to be more supporting literature at the beginning of major sections. For instance, I don’t need to be convinced that the structure of a model should be based on a realistic theory about the observed phenomenon; namely, models should be a white box. However, not everyone is on board with this premise, and supporting literature that argues in favour of this approach should be mentioned. In short, more citations should be added at the beginning of each major section.
}

\report{
\textbf{\underline{2. Limitations}}

In various passages of this paper, it is hinted that modellers should refrain from deterministic models and instead opt for more realistic stochastic representations. While the critique of ODE models is valid, the shift to stochastic structures is not a free choice. For example, introducing extra-demographic variability adds one additional parameter (infinitesimal variance). In ODE models, one extra parameter can lead to unidentifiability, and there's no apparent reason why this would differ in a stochastic version. Moreover, transitioning to stochastic models involves abandoning well-established MCMC algorithms in favour of methods still in development. Therefore, modellers should not assume that more realistic models with additional parameters are necessarily better without proper caution. In my experience, diagnosing unidentifiability is easier in ODE models than in POMP structures. Hence, there is a trade-off between benefits and costs.
}

\report{
Moreover, Monte Carlo methods, such as the Particle Filter and, by extension, Iterated Filtering, aim to approximate integrals (the posterior or filtering distributions). However, one cannot take for granted that these methods provide accurate descriptions of these targets without proper validation. In more ‘traditional’ MCMC methods, diagnostics like the potential scale reduction factor and effective sample size play a crucial role in the inference process. Unsatisfactory values of these diagnostics render inferences unreliable, often necessitating model reformulation. In contrast, in the literature on POMP models (including this paper), diagnostics are tangentially mentioned. Users (like me) sometimes face uncertainty about whether the lack of fit is due to model misspecification or problems with the Monte Carlo algorithm exploring the parameter space.

For example, in Figure 5, department Ouest exhibits substantial uncertainty from 2014 onwards, and this figure is on a log scale. Essentially, the inference suggests that 'anything can happen'. I would like to pinpoint the nature of this collapse in uncertainty. Identifiability issues might be at play, given the possibility of more estimated parameters than the incidence data can inform. Sometimes, we ask too much from the data. Observe the discrepancy in trends between the average behaviour and the uncertainty ribbons.

In summary, the authors should elaborate on the limitations of the proposed approach.
}

\report{
  \textbf{\underline{3. Conclusions}}

  I find that the conclusions are somewhat disconnected from the introduction and abstract, which state that the paper presents a methodology to diagnose model misspecification, develop alternative models, and make computational improvements. It would improve readability to include a summary in this section. Specifically, link each contribution to a particular example. For instance, in the case of model 1, computational improvements increased the log-likelihood. In short, connect the findings more explicitly to the research question and stated goals.
}

\report{
  \textbf{\underline{Minor Comments:}}
}

\report{
  1. In the author summary, this part is hard to follow: ``and provides careful justification of valid conclusions from the fitted model. Objective measures are used to benchmark model fit; when these are combined with reproducibility, a framework emerges for continual improvement when revisiting the data and models." Please rephrase.
}

\report{
  2. Lines 1-8. Please add more citations.
}

\report{
  3. Line 36. Can you be explicit about what the forecasts predicted? Did the models predict a rise in cases?
}

\report{
  4. Line 42. Add hyphen: ``Model-based conclusions".
}

\report{
  5. Lines 67-78. Add more citations.
}

\report{
  6. Please add uncertainty intervals to the estimated parameters in Table 1. If necessary, consider splitting the table into two or including this additional information in the supplementary material. I suggest this update because there is an indication of unidentifiability when parameter estimates are fairly broad.
}

\report{
  7. In line 159, it is stated that ``$\upsilon^\star(t)$ is efficacy at time $t$ since vaccination for adults" and then ``single and double vaccine doses were modeled by changing the waning of protection; protection was modeled as equal between single and double dose until 52 weeks after vaccination, at which point the single dose becomes ineffective". I examined the reference from which this function is based but found only a numeric table. Please provide the equation of this function or a detailed description in the supplementary information. It would be beneficial for readers to understand how to model this complex feature.
}

\report{
  8. Lines 209-211. In model 2, an incidence measurement is employed to configure a prevalence compartment. As the authors may know, initial values severely condition the dynamics of a model. Please explain this decision. Is it because that was the approach followed in the original formulation (Lee at al's paper)?
}

\report{
  9. Lines 245-247. Same comment as before. What’s the justification for assuming incidence measurements as the basis for prevalence states? What are the risks?
}

\report{
  10. Line 262. ``deterministic Model 2 is a degenerate case of a stochastic model". Please explain why or provide a reference.
}

\report{
  11. Lines 260-274. Please add more citations.
}

\report{
  12. Lines 343-345. This sentence is key for this paper, but it’s hard to follow: ``The robust statistical conclusion is that a model which allows for change fits better than one which does not—we argue that a decreasing transmission rate is a plausible way to explain this, but the incidence data themselves do not provide enough information to pin down the mechanism". Please rephrase.
}

\report{
  13. Line 357. ``Determining the necessary computational effort needed to maximize model likelihoods and acting accordingly" How do we determine the necessary computational effort?
}

\report{
  14. Please add the predicted intervals to Fig 4. I would like to see the effect of the log-normal measurement model.
}

\report{
  15. Lines 513-515. Please clarify the comparison between disaggregated models and a benchmark. Let’s say I have spatial units 1 and 2, for which I have observations y1 and y2. Should I fit the disaggregated mechanistic model to y1 and y2 simultaneously (as usual) but keep a record of the individual log-likelihoods (log-lik y1 and log-lik y2). In parallel, fit the benchmark independently to y1 and y2, and then compare by log-likelihoods or information criteria by spatial unit.
}

\report{
  16. Line 528. Please add hyphen: ``Model-based inference".
}

\report{
  17. Lines 576-577. ``We notice that the calibrated model favors higher levels of cholera transmission than what was typically observed in the incidence data (S5 Text)". After this fragment, please summarise in one or two sentences what it will be found in S5 Text.
}

\report{
  18. Lines 599-601. ``The decision not to do this partially explains the unsuccessful forecasts of Lee et al. [1]: their Table S7 shows that the subset of their simulations which were consistent with observing zero cases in 2019 also accurately predicted the prolonged absence of detected cholera". The fragment before the colon says that Lee was unsuccessful. However, the fragment after the colon says that was in part successful. Please clarify.
}

\report{
  19. Line 631. Since Model 1 accounts for infections at the national level, how are scenarios V1 and V2 handled?
}

\bibliography{../bib-haiti.bib}

\end{document}

### 2024-01-10 edits.tex
First commit with the response to reviewers

\usepackage[dvipsnames]{xcolor}
\definecolor{mygreen}{RGB}{0, 80, 0}
\newcommand\original[1]{\textcolor{red}{#1}}
\newcommand\TODO[1]{\textcolor{red}{[TODO: #1]}}

### 2024-01-10 ms.Rnw
Fixed typos mentioned by reviewers

@@
     utils::capture.output(sessionInfo()),
-    file = "sessionInfo.txt",
+    file = "sessionInfo_tmp.txt",
     sep="\n"
@@
 
-The discrepancy between the model based conclusions of Lee et al.~\cite{lee20} and the prolonged absence of cholera in Haiti has been debated \cite{francois20,rebaudetComment20,henrys20,leeReply20}.
+The discrepancy between the model-based conclusions of Lee et al.~\cite{lee20} and the prolonged absence of cholera in Haiti has been debated \cite{francois20,rebaudetComment20,henrys20,leeReply20}.
 Suggested origins of this discrepancy include the use of unrealistic models \cite{rebaudetComment20} and unrealistic criteria for cholera elimination \cite{henrys20}.
@@
 \label{eq:likedef}
-f_{\bm{Y}_{1:N}}\big(\bm{y}_{1:N}^*; \paramVec\big) = \int f_{\bm{X}_{0:N}, \bm{Y}_{1:N}}\big(\bm{x}_{0:N}, \bm{y}_{1:N};\paramVec\big) d\bm{x}_{0:N}.
+f_{\bm{Y}_{1:N}}\big(\bm{y}_{1:N}^*; \paramVec\big) = \int f_{\bm{X}_{0:N}, \bm{Y}_{1:N}}\big(\bm{x}_{0:N}, \bm{y}_{1:N}^*;\paramVec\big) d\bm{x}_{0:N}.
 \end{eqnarray}
@@
 
-For any time point in $t_{1:N}$, the process model $f_{\bm{X}_n|\bm{X}_{n-1}}\big(\bm{x}_{n}|\bm{x}_{n-1}; \paramVec\big)$ is defined by the describing how individuals move from one latent state compartment to another.
+For any time point in $t_{1:N}$, the process model $f_{\bm{X}_n|\bm{X}_{n-1}}\big(\bm{x}_{n}|\bm{x}_{n-1}; \paramVec\big)$ is defined by describing how individuals move from one latent state compartment to another.
 Per-capita transition rates are given in Eqs.~\myeqref{model1:SE}-\myeqref{model1:birth}:
@@
 Each model is indexed by a parameter vector, $\paramVec$, and different values of $\paramVec$ can result in qualitative differences in the predicted behavior of the system.
-Therefore, the choice of $\paramVec$ used to make inference about the system can greatly affect model based conclusions.
+Therefore, the choice of $\paramVec$ used to make inference about the system can greatly affect model-based conclusions.
 Elements of $\paramVec$ can be fixed at a constant value based on scientific understanding of the system, but parameters can also be calibrated to data by maximizing a measure of congruency between the observed data and the model's mechanistic structure.
@@
 The resulting mechanisms in a fitted model can be compared to current scientific knowledge about a system.
-Agreement between model based inference and our current understanding of a system may be taken as a confirmation of both model based conclusions and our scientific understanding.
+Agreement between model-based inference and our current understanding of a system may be taken as a confirmation of both model-based conclusions and our scientific understanding.
 On the other hand, comparisons may generate unexpected results that have the potential to spark new scientific knowledge \cite{ganusov16}.

### 2024-01-11 H3profile.R
Created a new file to attempt profiles with batchtools

@@ -1,0 +1,324 @@
+# Fitting profiles for Haiti 3.
+#
+# The fit_haiti3 function is a very simple helper that will take as input
+# some starting parameters, and hyperparameters for the IBPF algorithm, and
+# will return parameter estimates along with corresponding estimates of their
+# log-likelihoods.
+#
+# In this script we will use this function to do a profile likelihood search
+# of all model parameters using the batchtools package. To do this, we will
+# create a grid of parameter values, and submit jobs where each job will use
+# a row from the grid as a starting point to fit the model.
+
+
+# Load Packages -----------------------------------------------------------
+
+library(batchtools)
+library(tidyverse)
+# library(data.table)
+library(haitipkg)
+
+RUN_LEVEL <- 2
+
+NBPF <- switch(RUN_LEVEL, 5, 50, 100)
+NP <- switch(RUN_LEVEL, 50, 500, 1000)
+NP_EVAL <- switch(RUN_LEVEL, 100, 1000, 2000)
+NREPS_EVAL <- switch(RUN_LEVEL, 3, 6, 10)
+SPAT_REGRESSION <- 0.05
+COOLING <- 0.05
+
+# Create vectors for the unit and shared parameters
+unit_specific_names <- c("betaB", "foi_add", "aHur", "hHur")
+shared_param_names <- c(
+  "mu_B", "XthetaA", "thetaI", "lambdaR", "r", "std_W",
+  "epsilon", "k"
+)
+
+est_param_names <- c(
+  unit_specific_names, shared_param_names
+)
+
+# Add unit numbers to each parameter
+est_param_names_expanded <- paste0(rep(est_param_names, each = 10), 1:10)
+
+# Simple function that is used to set the rw_sd for the first search
+# This function is convenient so that we don't have to type out the rw_sd
+# for each unit parameter.
+set_rw_1 <- function(x) {
+  if (gsub("[[:digit:]]+$", "", x) %in% shared_param_names) {
+    return(0.01)
+  } else if (x %in% paste0(rep(c("aHur", 'hHur'), each = 2), c(3, 9))) {
[…]

### 2024-01-11 H3profile.R
Fixed bug, passing data into fitting function

@@ -18,68 +18,15 @@ library(tidyverse)
 # library(data.table)
 library(haitipkg)
 
-RUN_LEVEL <- 2
+RUN_LEVEL <- 1
 
+nprof <- switch(RUN_LEVEL, 2, 12, 20)
 NBPF <- switch(RUN_LEVEL, 5, 50, 100)
 NP <- switch(RUN_LEVEL, 50, 500, 1000)
 NP_EVAL <- switch(RUN_LEVEL, 100, 1000, 2000)
 NREPS_EVAL <- switch(RUN_LEVEL, 3, 6, 10)
 SPAT_REGRESSION <- 0.05
 COOLING <- 0.05
-
-# Create vectors for the unit and shared parameters
-unit_specific_names <- c("betaB", "foi_add", "aHur", "hHur")
-shared_param_names <- c(
-  "mu_B", "XthetaA", "thetaI", "lambdaR", "r", "std_W",
-  "epsilon", "k"
-)
-
-est_param_names <- c(
-  unit_specific_names, shared_param_names
-)
-
-# Add unit numbers to each parameter
-est_param_names_expanded <- paste0(rep(est_param_names, each = 10), 1:10)
-
-# Simple function that is used to set the rw_sd for the first search
-# This function is convenient so that we don't have to type out the rw_sd
-# for each unit parameter.
-set_rw_1 <- function(x) {
-  if (gsub("[[:digit:]]+$", "", x) %in% shared_param_names) {
-    return(0.01)
-  } else if (x %in% paste0(rep(c("aHur", 'hHur'), each = 2), c(3, 9))) {
-    return(expression(ifelse(time >= 2016.754 & time <= 2017, 0.015, 0)))
-  } else if (x %in% c("Iinit3", "Iinit4")) {
-    return(expression(ivp(0.25)))
-  } else if (grepl('^foi_add[[:digit:]]+$', x)) {
-    return(0.015)
-  } else if (grepl('^betaB[[:digit:]]+$', x)) {
-    return(0.01)
-  } else {
-    return(0)
-  }
-}
-
-set_rw_2 <- function(x) {
-  if (gsub("[[:digit:]]+$", "", x) %in% shared_param_names) {
-    return(0.003)
-  } else if (x %in% paste0(rep(c("aHur", 'hHur'), each = 2), c(3, 9))) {
-    return(expression(ifelse(time >= 2016.754 & time <= 2017, 0.005, 0)))
[…]

### 2024-01-12 H3profile.R
changed run-level to 2 and ran the code

@@ -18,7 +18,7 @@ library(tidyverse)
 # library(data.table)
 library(haitipkg)
 
-RUN_LEVEL <- 1
+RUN_LEVEL <- 2
 
 nprof <- switch(RUN_LEVEL, 2, 12, 20)
 NBPF <- switch(RUN_LEVEL, 5, 50, 100)
@@ -321,8 +321,8 @@ addExperiments(prob.designs = pdes, algo.designs = ades)
 
 # Submit Jobs -------------------------------------------------------------
 
-resources1 <- list(account = 'stats_dept1', walltime = '10:00', memory = '5000m', ncpus = 1)
-# resources1 <- list(account = 'stats_dept1', walltime = '90:00', memory = '5000m', ncpus = 1)
+# resources1 <- list(account = 'stats_dept1', walltime = '10:00', memory = '5000m', ncpus = 1)
+resources1 <- list(account = 'stats_dept1', walltime = '90:00', memory = '5000m', ncpus = 1)
 
-# submitJobs(resources = resources1)
-submitJobs(ids = 1)
+submitJobs(resources = resources1)
+# submitJobs(ids = 1, resources = resources1)

### 2024-01-12 H3profile.R
Modified RL based on previous results

@@ -20,9 +20,9 @@ library(haitipkg)
 
 RUN_LEVEL <- 1
 
-nprof <- switch(RUN_LEVEL, 2, 12, 20)
+nprof <- switch(RUN_LEVEL, 2, 14, 20)
 NBPF <- switch(RUN_LEVEL, 5, 50, 100)
-NP <- switch(RUN_LEVEL, 50, 500, 1000)
+NP <- switch(RUN_LEVEL, 50, 750, 1000)
 NP_EVAL <- switch(RUN_LEVEL, 100, 1000, 2000)
 NREPS_EVAL <- switch(RUN_LEVEL, 3, 6, 10)
 SPAT_REGRESSION <- 0.05
@@ -63,21 +63,21 @@ prof_vars <- c()
 for (pp in prof_params) {
 
   if (pp == "mu_B") {
-    prof_values <- seq(375, 750, length.out = 20)
+    prof_values <- seq(300, 700, length.out = 25)
   } else if (pp == 'XthetaA') {
-    prof_values <- seq(0.02, 0.25, length.out = 20)
+    prof_values <- seq(0.03, 0.25, length.out = 21)
   } else if (pp == 'thetaI') {
-    prof_values <- seq(2.5e-05, 1e-04, length.out = 20)
+    prof_values <- seq(2.5e-05, 1e-04, length.out = 21)
   } else if (pp == 'lambdaR') {
-    prof_values <- seq(0.5, 3.5, length.out = 15)
+    prof_values <- seq(0.5, 3.2, length.out = 15)
   } else if (pp == 'r') {
-    prof_values <- seq(0.5, 1.75, length.out = 15)
+    prof_values <- seq(0.5, 1.75, length.out = 16)
   } else if (pp == 'std_W') {
-    prof_values <- seq(0.02, 0.03, length.out = 20)
+    prof_values <- seq(0.02, 0.04, length.out = 25)
   } else if (pp == 'epsilon') {
     prof_values <- seq(0.65, 0.99, length.out = 20)
   } else if (pp == 'k') {
-    prof_values <- seq(75, 175, length.out = 20)
+    prof_values <- seq(70, 175, length.out = 25)
   }
 
   tmp_pars <- matrix(
@@ -167,7 +167,7 @@ for (pp in prof_params) {
     "r" = 0.5,
     "std_W" = 0.02,
     "epsilon" = 0.65,
-    "k" = 75
+    "k" = 70
[…]

### 2024-01-12 H3profile.R
Finished RL2 modifications

@@ -31,14 +31,14 @@ COOLING <- 0.05
 # Create Experiment Registry ----------------------------------------------
 
 reg <- makeExperimentRegistry(
-  file.dir = paste0('model3/profileReg_RL', RUN_LEVEL),
+  file.dir = paste0('model3/profileReg_RL', RUN_LEVEL, '_v2'),
   seed = 739164,
   packages = c("spatPomp", 'haitipkg', 'pomp')
 )
 
 # Create Profile Design Matrix --------------------------------------------
 
-set.seed(998877)
+set.seed(665544)
 
 # Read in previous best results
 haiti3_fit <- readRDS("model3/run_level_3/haiti3_fit.rds")
@@ -322,7 +322,7 @@ addExperiments(prob.designs = pdes, algo.designs = ades)
 # Submit Jobs -------------------------------------------------------------
 
 # resources1 <- list(account = 'stats_dept1', walltime = '10:00', memory = '5000m', ncpus = 1)
-resources1 <- list(account = 'stats_dept1', walltime = '90:00', memory = '5000m', ncpus = 1)
+resources1 <- list(account = 'stats_dept1', walltime = '2:00:00', memory = '5000m', ncpus = 1)
 
 submitJobs(resources = resources1)
 # submitJobs(ids = 1, resources = resources1)

### 2024-01-12 H3profile.R
Fixed cooling rate 0.05 -> 0.5

@@ -26,12 +26,12 @@ NP <- switch(RUN_LEVEL, 50, 750, 1000)
 NP_EVAL <- switch(RUN_LEVEL, 100, 1000, 2000)
 NREPS_EVAL <- switch(RUN_LEVEL, 3, 6, 10)
 SPAT_REGRESSION <- 0.05
-COOLING <- 0.05
+COOLING <- 0.50
 
 # Create Experiment Registry ----------------------------------------------
 
 reg <- makeExperimentRegistry(
-  file.dir = paste0('model3/profileReg_RL', RUN_LEVEL, '_v2'),
+  file.dir = paste0('model3/profileReg_RL', RUN_LEVEL, '_v3'),
   seed = 739164,
   packages = c("spatPomp", 'haitipkg', 'pomp')
 )

### 2024-01-12 H3profile.R
Fixed minor bug for unit-specific profiles, not yet tested

@@ -96,7 +96,7 @@ for (pp in prof_params) {
   } else {
     prof_cols <- matrix(rep(prof_values, each = nprof), ncol = 1)
     colnames(prof_cols) <- pp
-    tmp_pars[pp] <- prof_values
+    tmp_pars[, pp] <- prof_values
   }
 
   # Unit-lower bounds

### 2024-01-12 r1/response.tex
Fixed typos mentioned by reviewers
some letter edits

@@
 
+WE CAN TRIM DOWN THE RESPONSE LATER. HERE ARE SOME INITIAL BRAINSTORMING THOUGHTS.
+
+Thanks for the feedback on the framing of our contribution.
+The original goal of our manuscript was to combine the two foci identified by the referee: 
+we use an in-depth case study to motivate general procedures for fitting mathematical models to time series data, while explaining how these procedures help to address limitations of previous approaches.
+The case study was intended both to help motivate the advice and to make sure that the advice is pertinent and practical on at least one suite of models.
+Since our approach is predicated on the possibility of carrying out some nontrivial computational procedures, we are led to address computational issues in parallel with scientific and statistical issues.
+
+The referee would prefer us to emphasize one of these foci, and this feedback has helped us see how readers might find themselves struggling to orient themselves while reading the manuscript.
+We agree that the ``educational'' purpose of the paper is primary to the specific conclusions about the Haiti cholera epidemic.
+However, the validity of our reanalysis of the Haiti data and models provides critical support to our methodological points.
+\TODO{For the revision, we have edited the abstract and introduction to make explicit how the case study fits into the overall goals of the paper.}
+\TODO{In the methods and results sections, we have made explicit how the decisions about what to present in the main text and in the supplement relate to these goals.}
+In the discussion, we return from the concrete consideration of the case study, placing it back in the more general setting of inferring complex dynamics from time series data.
+\TODO{In the revision, we have added more signposts to help the reader see how the points that arise in the discussion generalize the corresponding points arising in the case study.}
+
 \report{
@@
 
+Fitting models 1 and 2 does not require any methodological innovation.
+Model 3 has a stochastic spatiotemporal partially-observed structure that leads to methodological challenges.
+\citet{lee20} made some approximations to avoid carrying out inference for the full coupled system.
+We show some consequences of those approximations, taking advantage of a newly developed method which has not yet been applied in the context of scientific investigations.
+So, although the method is not entirely new, its applicability to this task is novel.
+\TODO{We have clarified this in the manuscript WHERE?}
+
 \report{
@@
 
+1.
 To make clear that the issues being addressed by this paper go beyond one modelling study of Haiti, the authors should provide examples and references to support the (accurate in my opinion) statement about, ``common modeling decisions that may not provide an adequate statistical explanation of the data".
@@
 
-\report{
+\report{2.
   Given the focus of the paper on how to better perform inference, explicitly commenting on what are the ``existing guidelines for creating models to inform policy [4, 5]".
@@
 
+Thanks for this suggestion.
+\TODO{We have implemented this in the revision.}
+
+THIS IS DISCUSSED MORE EXTENSIVELY IN JIFAN'S COVID PAPER. THERE IS SOME DANGER THAT THESE PAPERS COULD CONVERGE TOO MUCH. ANOTHER OPTION IS TO REFER TO THAT PAPER. SO, AS ANOTHER OPTION:
+
+\TODO{A subsequent manuscript addresses [4,5] more directly \citep{li23} and so we have referred the reader to that paper (currently available on arXiv) for additional discussion of this point.}
+
 \report{
@@
 
-  Fig 2. Model parameters. I think this should be ``Table 1"?
+3.  Fig 2. Model parameters. I think this should be ``Table 1"?
 }
 
-\report{
+\report{4.
     ``and $z \in 1:Z$ describes hypothetical vaccination programs" It would help to give an example of the different vaccine programs here.
@@
 
-\report{
+\report{5.
   Eq 6 and line 1: I think the notation for the asymptomatic fraction might be clearer and more consistent with the other models if you used $f_z(t)$ instead of $\vartheta_z(t)$.
@@
 
-\report{
+\report{6.
   Line 198: Reference for where value of $v_{rate}$ comes from.
@@
 
-\report{
+\report{7.
   Usually, vaccine efficacy = 1 implies the vaccine completely blocks transmission and 0 implies no effect.
@@
 
-\report{
+\report{8.
   I found the commentary on comparing fitted mechanistic models to a (statistical) benchmark model to be informative and helpful.
@@
 
-\report{
+\TODO{This is right, as long as there are not any initialization issues...}
+
+\report{9. 
   Lines 308-309: Isn't this a log-linear trend in the transmission rate?
@@
 
-\report{
+\report{10.
   While the data support a linear trend at the $95\%$ confidence level, it's worth commenting on the magnitude of the trend, e.g. the total reduction in $\beta(t)$ over the period of study.
@@
 
-\report{
+\report{11.
   I might have misunderstood it slightly, but I don't entirely follow the reasoning of lines 340-346.
@@
 
-\report{
+Our views align with the referee's comments, and we have rewritten the article to make this clearer.
+
+WE COULD ADD THE SUBMITTED PDF TO THE REVISION REPO SO WE CAN CHECK LINE NUMBERS.
+
+\report{12.  [I THINK THIS MEANS LINES 553-556, ASSUMING THE NUMBERING MATCHES]
   Lines 53-56: I push back against this point.
@@
 Even for much larger durations of immunity a non-negligible fraction of the population will lose immunity after 9 years, e.g.
-.
+$e^{-9/10}\approx 0.41$.
 I therefore don't expect values for the duration of immunity around 10 years to ``effectively result in the same model dynamics".
 }
+
+The qualitative comparison we had in mind was the regime with short immunity where many people become infected multiple times during the ten year epidemic, allowing multiple epidemic peaks due to susceptible replenishment.
+\TODO{We have reworded the sentence to avoid making too strong a claim about 10yr immunity being a precise point where the dynamics transition to a regime where reinfection events become negligible.}
 
@@
 
+13.
 In contrast with the introduction, which almost exclusively focuses on cholera in Haiti, the discussion only mentions it once, briefly: ``We used the same data and models, and even much of the same code, as Lee et al.
@@
 
+Yes, there should be an asterisk. This has been fixed.
+
 \report{
@@
 }
+
+Fixed.
 
@@
 
+A hyphen has been added for all instances of the phrase: ``model-based".
+
 \report{

### 2024-01-13 H1profile.R
Changed to RL2

@@ -18,13 +18,13 @@ library(tidyverse)
 # library(data.table)
 library(haitipkg)
 
-RUN_LEVEL <- 1
+RUN_LEVEL <- 2
 
-nprof      <- switch(RUN_LEVEL, 2, 14, 20)
-NMIF       <- switch(RUN_LEVEL, 5, 100, 200)
-NP         <- switch(RUN_LEVEL, 50, 1000, 1000)
-NP_EVAL    <- switch(RUN_LEVEL, 100, 5000, 10000)
-NREPS_EVAL <- switch(RUN_LEVEL, 3, 6, 10)
+nprof      <- switch(RUN_LEVEL,  2,    14, 20)
+NMIF       <- switch(RUN_LEVEL,  5,   100, 200)
+NP         <- switch(RUN_LEVEL,  50, 1000, 1000)
+NP_EVAL    <- switch(RUN_LEVEL, 100, 2500, 10000)
+NREPS_EVAL <- switch(RUN_LEVEL,   3,   10, 10)
 COOLING    <- 0.5
 
 # Create Experiment Registry ----------------------------------------------
@@ -222,8 +222,7 @@ addAlgorithm(name = 'fitMod', fun = fit_model)
 pdes <- list('profile' = data.frame(i = 1:nrow(final_pars)))
 ades <- list(
   'fitMod' = data.frame(
-    nbpf = NBPF, np = NP, spat_regression = SPAT_REGRESSION,
-    np_eval = NP_EVAL, nreps_eval = NREPS_EVAL, cooling = COOLING
+    nmif = NMIF, np = NP, np_eval = NP_EVAL, nreps_eval = NREPS_EVAL, cooling = COOLING
   )
 )
 
@@ -231,8 +230,13 @@ addExperiments(prob.designs = pdes, algo.designs = ades)
 
 # Submit Jobs -------------------------------------------------------------
 
-resources1 <- list(account = 'stats_dept1', walltime = '10:00', memory = '1000m', ncpus = 1)
-# resources1 <- list(account = 'stats_dept1', walltime = '50:00', memory = '5000m', ncpus = 1)
+# resources1 <- list(account = 'stats_dept1', walltime = '10:00', memory = '1000m', ncpus = 1)
+resources1 <- list(account = 'stats_dept1', walltime = '50:00', memory = '5000m', ncpus = 1)
 
-# submitJobs(resources = resources1)
-submitJobs(ids = 1, resources = resources1)
+submitJobs(
[…]

### 2024-01-13 H1profile.R
Added library(data.table)

@@ -15,7 +15,7 @@
 
 library(batchtools)
 library(tidyverse)
-# library(data.table)
+library(data.table)
 library(haitipkg)
 
 RUN_LEVEL <- 2

### 2024-01-14 H3profile.R
Adjusted profile search based on previous results

@@ -26,12 +26,12 @@ NP <- switch(RUN_LEVEL, 50, 750, 1000)
 NP_EVAL <- switch(RUN_LEVEL, 100, 1000, 2000)
 NREPS_EVAL <- switch(RUN_LEVEL, 3, 6, 10)
 SPAT_REGRESSION <- 0.05
-COOLING <- 0.50
+COOLING <- 0.4
 
 # Create Experiment Registry ----------------------------------------------
 
 reg <- makeExperimentRegistry(
-  file.dir = paste0('model3/profileReg_RL', RUN_LEVEL, '_v3'),
+  file.dir = paste0('model3/profileReg_RL', RUN_LEVEL, '_v4'),
   seed = 739164,
   packages = c("spatPomp", 'haitipkg', 'pomp')
 )
@@ -63,21 +63,21 @@ prof_vars <- c()
 for (pp in prof_params) {
 
   if (pp == "mu_B") {
-    prof_values <- seq(300, 700, length.out = 25)
+    prof_values <- seq(200, 650, length.out = 25)
   } else if (pp == 'XthetaA') {
-    prof_values <- seq(0.03, 0.25, length.out = 21)
+    prof_values <- seq(0.025, 0.25, length.out = 21)
   } else if (pp == 'thetaI') {
-    prof_values <- seq(2.5e-05, 1e-04, length.out = 21)
+    prof_values <- seq(4e-05, 1.75e-04, length.out = 21)
   } else if (pp == 'lambdaR') {
-    prof_values <- seq(0.5, 3.2, length.out = 15)
+    prof_values <- seq(1, 3.5, length.out = 21)
   } else if (pp == 'r') {
-    prof_values <- seq(0.5, 1.75, length.out = 16)
+    prof_values <- seq(0.5, 1.5, length.out = 21)
   } else if (pp == 'std_W') {
-    prof_values <- seq(0.02, 0.04, length.out = 25)
+    prof_values <- seq(0.025, 0.05, length.out = 25)
   } else if (pp == 'epsilon') {
-    prof_values <- seq(0.65, 0.99, length.out = 20)
+    prof_values <- seq(0.8, 0.999, length.out = 20)
   } else if (pp == 'k') {
-    prof_values <- seq(70, 175, length.out = 25)
+    prof_values <- seq(30, 120, length.out = 25)
   }
 
   tmp_pars <- matrix(
[…]

### 2024-01-14 H3profile.R
Lowered cooling rate

@@ -26,7 +26,7 @@ NP <- switch(RUN_LEVEL, 50, 750, 1000)
 NP_EVAL <- switch(RUN_LEVEL, 100, 1000, 2000)
 NREPS_EVAL <- switch(RUN_LEVEL, 3, 6, 10)
 SPAT_REGRESSION <- 0.05
-COOLING <- 0.4
+COOLING <- 0.3
 
 # Create Experiment Registry ----------------------------------------------
 

### 2024-01-14 ms.Rnw
add lit review in bib-lit.bib

@@
 
-This case study investigates the use of dynamic models and spatiotemporal data to inform a policy decision in the context of the cholera outbreak in Haiti, which started in 2010.
-We build on a multi-group modeling exercise by Lee et al.~\cite{lee20} in which four expert modeling teams developed models to the same dataset with the goal of comparing conclusions on the feasibility of eliminating cholera by a vaccination campaign.
-The data that were used for this study, and that is the subject of our reanalysis, is displayed in Fig.~\ref{fig:CholeraData}.
+This case study investigates the use of dynamic models and spatiotemporal data to inform public health policy in the context of the cholera outbreak in Haiti, which started in 2010.
+Various dynamic models were developed to study this outbreak: searching PubMed with keywords ``Haiti, cholera, model'' we obtained 20 deterministic models \cite{kirpich16,kirpich17,andrews11,botelho21,fitzgibbon20,eisenberg13,rinaldo12,chao11,date11,lin19,collins21,akman15,trevisin22,mavian20,collins14,kelly16,capone14,bengtsson15,mari15,gatto12} and 8 stochastic models \cite{tuite11,mukandvire13,lewnard16,kunke17,sallah17,azman12,kuhn14,azman15}.
+Here we focus on a multi-group modeling exercise by Lee et al.~\cite{lee20} in which four expert modeling teams developed models to the same dataset with the goal of comparing conclusions on the feasibility of eliminating cholera by a vaccination campaign.
 Model~1 is stochastic and describes cholera at the national level;
@@
 Model~4 has an agent-based construction, featuring considerable mechanistic detail but limited ability to calibrate these details to data.
+These modeling strategies were selected by \cite{li20} to represent the range of approaches used in the research community.
 We focus on Models~1--3, as the strengths and weaknesses of the agent-based modeling approach \cite{tracy18} are outside the scope of this article.
+\eic{In addition, agent-based models were less widely used [NO OTHERS SHOW UP IN OUR LITERATURE REVIEW???].}
+The data that were used for this study, and that is the subject of our reanalysis, are displayed in Fig.~\ref{fig:CholeraData}.
 
@@
 
-While this work is focused on the models proposed by Lee et al.~\cite{lee20}, our suggestions have broader relevance.
-To investigate the extent to which \cite{lee20} is typical of the substantial body of work on Haiti cholera dynamics, we performed a literature review by searching PubMed with keywords: Haiti, cholera, model.
-The search resulted in 66 papers, of which 32 used dynamic models to describe the cholera epidemic in Haiti.
-The models make various choices on the dichotomies considered by Lee et al.~\cite{lee20}; deterministic versus stochastic; compartment model versus agent-based; aggregated versus spatially explicit.
+%% IT SEEMS THIS PARAGRAPH IS DEPRECATED IN FAVOR OF CITING THESE REFERENCES WEHRE APPROPRIAT
+%% While this work is focused on the models proposed by Lee et al.~\cite{lee20}, our suggestions have broader relevance.
+%% To investigate the extent to which \cite{lee20} is typical of the substantial body of work on Haiti cholera dynamics, we performed a literature review by searching PubMed with keywords: Haiti, cholera, model.
+%% The search resulted in 66 papers, of which 32 used dynamic models to describe the cholera epidemic in Haiti.
+%% The models make various choices on the dichotomies considered by Lee et al.~\cite{lee20}; deterministic versus stochastic; compartment model versus agent-based; aggregated versus spatially explicit.
 
@@
 
-In this article we calibrate the parameters of each of our models by maximizing the likelihood, as described in Eq.~\myeqref{eq:likedef}.
+In this article we follow \cite{lee20) by calibrating the parameters of each of our models using maximizing the likelihood, as described in Eq.~\myeqref{eq:likedef}.
 The likelihood for each of the fitted models---and the corresponding AIC values for model comparisons that include an adjustment for the number of calibrated parameters---is provided in Table~\ref{tab:likes}.
 In the following subsections we describe in detail our approach to calibrating the three proposed mechanistic models to observed cholera incidence data.
+\eic{I'VE STARTED TO PUT IN SOME DISCUSSION OF BAYESIAN METHODS HERE, THOUGH IT COULD ALSO GO IN THE DISCUSSION}
+The main alternative to maximum likelihood estimation is Bayesian inference via Markov chain Monte Carlo, used to analyze the Haiti cholera epidemic by \cite{andrews11,rinaldo12,lewnard16,trevisin22,sallah17,azman12,kuhn14,mari15,gatto12}.
+Alternatively, some investigations obtained parameters from previous sources without calibrating to data.
 
@@
 
-\bibliography{bib-haiti}
+\bibliography{bib-haiti,bib-lit}
 

### 2024-01-15 r1/response.tex
response to R2 about mcmc, etc

@@
 \documentclass[11pt]{article}
-
-\input{../edits.tex}
-%\usepackage[sectionbib]{natbib}
-
-%\usepackage{xr}
-%\externaldocument{../ms}
 
@@
 \usepackage{color}
+
+\input{../edits.tex}
+%\usepackage[sectionbib]{natbib}
+
+%\usepackage{xr}
+%\externaldocument{../ms}
 
@@
 
+The referee raises several points in this paragraph, and we consider them one at a time.
+
+We agree that adding to model complexity has costs, including loss of identifability, AIC and other formal statistical methods can assess whether the additional complexity is justified by a sufficiently large improvement of statistical fit.
+Identifiability and model simplicity have practical scientific value which goes beyond their statistical role in reducing generalization error and hence improving out-of-fit predictions.
+This is a variation on the issue of "statistical significance versus practical significance" familiar from introductory statistics textbooks.
+HOW DOES THIS PLAY OUT IN THE MS? HOW DO WE CLARIFY IT?
+
+Moving from an ODE model to a stochastic model may not involve any additional parameters, if only demographic stochasticity is included.
+Generally, it is advisable to consider the possibility of over-dispersed variation on the transmission rate \cite{breto09,he10} leading to one extra parameter compared to a deterministic model.
+In the context of cholera dynamics, \cite{lemaitre19} found a substantial improvement in model fit by including stochasticity, involving one over-dispersion parameter.
+stocks20 found similar results, favoring a stochastic dynamic model with over-dispersion, when investigating transmission dynamics of a different enteric pathogen, rotavirus.
+In other contexts, king15 noted various scientific advantages for including stochasticity in the model; it is undoubtedly part of the biological dynamics, and its omission can lead to over-confident predictions.
+In our study, we find that the deterministic model (model 2) has poor statistical fit.
+This can lead to a risk of identifying spurious scientific phenomena: addition deterministic complexity added to the model may help to describe unmodeled stochastic phenomena present in the data, and may therefore spuriously be assessed as a highly statistically significant model improvement.
+
+The decision on whether to choose Markov chain Monte Carlo (MCMC) methods is not fully aligned with the decision on whether to include stochasticity in the model.
+In our literature review, of the 9 previous Haiti cholera investigations including stochastic dynamic models \cite{tuite11,mukandavire13,lewnard16,kunkel17,sallah17,azman12,kuhn14,azman15,lee20}, four of them carried out inference via MCMC \cite{lewnard16,sallah17,azman12,kuhn14}.
+WHAT SORTS OF MCMC ALGORITHMS DID THEY USE?
+We agree with the referee that MCMC is more routine for determinstic models than for stochastic models, and advanced MCMC methods such as Hamiltonian Monte Carlo become more readily applicable. 
+Sequential Monte Carlo (SMC) was used for only two of the investigations in our literature review \cite{lee20,azman15}.
+In principle, MCMC methods can be applied for inference in over-dispersed stochastic dynamic models via so-called particle MCMC (PMCMC) \cite{andrieu10}.
+In practice, these models have been fitted by maximum likelihood, some examples including \cite{lemaitre19,fox22,molodecky23,pons-salort18,subramanian20,stocks20}.
+In this case, identifiability is generally addressed via profile likelihood methods.
+
+We have included some additional sentences to address these comments in the discussion section of the revision.
+TODO
+
+We acknowledge the existence of useful mathematical tools for studying ODE models, their identifiability and stability and limiting properties.
+Stochastic dynamic models can be reduced to a deterministic model (a so-called determinsitic skeleton) which can usefully be analyzed to help understand the stochastic dynamics \cite{coulson04}.
+We therefore encourage the use of ODE analysis tools whether or not a stochastic model is adopted.
+We have not followed that approach in the current manuscript.
+
+In summary, maximum likelihood (ML) inference via SMC nowaways provides a fairly well estabilished alternative to Bayesian MCMC inference.
+ML appears to be methodologically more convenient for stochastic dynamic models with overdispersion.
+When analyzing long epidemiological datasets, this stochasticity appears to be important for statistical model fit.
+That alone is not enough to show that models with overdispersed stochastic dynamics are scientifically preferable.
+However, discovery of a  model feature which generates a large improvement in statistcal fit from adding one more parameter is typically treated as scientific evidence favoring the inclusion of that feature in a scientific model for the phenomenon.
+
 \report{
@@
 
-\bibliography{../bib-haiti.bib}
+\bibliography{../bib-haiti.bib,../bib-lit.bib}
 

### 2024-01-15 ms.Rnw
response to R2 about mcmc, etc

@@
 
-%%%%%% EDITING MACROS %%%%%%%%%
-% orange for EI
-\definecolor{orange}{rgb}{1,0.5,0}
-\newcommand\ei[2]{\sout{#1} \textcolor{orange}{#2}}
-\newcommand\eic[1]{\textcolor{orange}{[#1]}}
-% green for JW
-\definecolor{green}{rgb}{0,0.5,0}
-\newcommand\jw[2]{\sout{#1} \textcolor{green}{#2}}
-\newcommand\jwc[1]{\textcolor{green}{[#1]}}
-% purple for JJ
-\definecolor{purple}{rgb}{0.5,0,1}
-\newcommand\jj[2]{\sout{#1} \textcolor{purple}{#2}}
-\newcommand\jjc[1]{\textcolor{purple}{[#1]}}
-% cyan for AR
-\definecolor{cyan}{rgb}{0,.5,.5}
-\newcommand\ar[2]{\sout{#1} \textcolor{cyan}{#2}}
-\newcommand\arc[1]{\textcolor{cyan}{#1}}
-% light brown for KT
-\definecolor{lightbrown}{rgb}{0.5,0.5,0}
-\newcommand\kt[2]{\sout{#1} \textcolor{lightbrown}{#2}}
-\newcommand\ktc[1]{\textcolor{lightbrown}{#1}}
-
 \newcolumntype{t}{>{\tiny}c}
@@
 This necessitates addressing the tradeoff between biological fidelity and model simplicity, and the reality of misspecification for models at all levels of complexity.
-As a case study, we consider the 2010-2019 cholera epidemic in Haiti.
-We study three dynamic models developed by expert teams to advise on vaccination policies.
-We assess previous methods used for fitting and evaluating these models, and we develop data analysis strategies leading to improved statistical fit.
+\ei{I'VE RE-WRITTEN HERE, LOOKING TO BE MORE SPECIFIC ABOUT THE GOALS AND CONTRIBUTION OF THE PAPER.} 
+We assess current methodological approaches to these issues via a case study of the 2010-2019 cholera epidemic in Haiti.
+We consider three dynamic models developed by expert teams to advise on vaccination policies.
+We evaluate previous methods used for fitting these models, and we demonstrate modified data analysis strategies leading to improved statistical fit.
 Specifically, we present approaches to diagnosis of model misspecification, development of alternative models, and computational improvements in optimization, in the context of likelihood-based inference on nonlinear dynamic systems.

### 2024-01-15 ms.Rnw
Changed xcolor .tex options, recompiled

@@
 \usepackage[top=0.85in,left=2.75in,footskip=0.75in]{geometry}
+
+\input{edits.tex}
 
@@
 % color can be used to apply background shading to table cells only
-\usepackage[table]{xcolor}
+% \usepackage[table]{xcolor}
 
@@
 
-In this article we follow \cite{lee20) by calibrating the parameters of each of our models using maximizing the likelihood, as described in Eq.~\myeqref{eq:likedef}.
+In this article we follow \cite{lee20} by calibrating the parameters of each of our models using maximizing the likelihood, as described in Eq.~\myeqref{eq:likedef}.
 The likelihood for each of the fitted models---and the corresponding AIC values for model comparisons that include an adjustment for the number of calibrated parameters---is provided in Table~\ref{tab:likes}.

### 2024-01-15 H3profile.R
Modified RL2 profile bounds, only profiling some params

@@ -26,12 +26,12 @@ NP <- switch(RUN_LEVEL, 50, 750, 1000)
 NP_EVAL <- switch(RUN_LEVEL, 100, 1000, 2000)
 NREPS_EVAL <- switch(RUN_LEVEL, 3, 6, 10)
 SPAT_REGRESSION <- 0.05
-COOLING <- 0.3
+COOLING <- 0.25
 
 # Create Experiment Registry ----------------------------------------------
 
 reg <- makeExperimentRegistry(
-  file.dir = paste0('model3/profileReg_RL', RUN_LEVEL, '_v4'),
+  file.dir = paste0('model3/profileReg_RL', RUN_LEVEL, '_v5'),
   seed = 739164,
   packages = c("spatPomp", 'haitipkg', 'pomp')
 )
@@ -56,7 +56,10 @@ est_param_names <- c(
   unit_specific_names, shared_param_names
 )
 
-prof_params <- shared_param_names
+# prof_params <- shared_param_names
+prof_params <- c(
+  "epsilon", "k", "lambdaR", "r", "thetaI", "XthetaA"
+  )
 
 final_pars <- best_pars
 prof_vars <- c()
@@ -65,19 +68,19 @@ for (pp in prof_params) {
   if (pp == "mu_B") {
     prof_values <- seq(200, 650, length.out = 25)
   } else if (pp == 'XthetaA') {
-    prof_values <- seq(0.025, 0.25, length.out = 21)
+    prof_values <- seq(0.005, 0.1, length.out = 21)
   } else if (pp == 'thetaI') {
-    prof_values <- seq(4e-05, 1.75e-04, length.out = 21)
+    prof_values <- seq(0.000075, 0.00016, length.out = 20)
   } else if (pp == 'lambdaR') {
-    prof_values <- seq(1, 3.5, length.out = 21)
+    prof_values <- seq(1, 3, length.out = 20)
   } else if (pp == 'r') {
-    prof_values <- seq(0.5, 1.5, length.out = 21)
+    prof_values <- seq(0.15, 1, length.out = 21)
   } else if (pp == 'std_W') {
     prof_values <- seq(0.025, 0.05, length.out = 25)
   } else if (pp == 'epsilon') {
-    prof_values <- seq(0.8, 0.999, length.out = 20)
[…]

### 2024-01-15 H1profile.R
Redoing H1 profiles after fixing E_0, I_0, S_0

@@ -30,7 +30,7 @@ COOLING    <- 0.5
 # Create Experiment Registry ----------------------------------------------
 
 reg <- makeExperimentRegistry(
-  file.dir = paste0('model1/profileReg_RL', RUN_LEVEL, '_v1'),
+  file.dir = paste0('model1/profileReg_RL', RUN_LEVEL, '_v2'),
   seed = 739164,
   packages = c("spatPomp", 'haitipkg', 'pomp')
 )
@@ -66,19 +66,19 @@ prof_vars <- c()
 for (pp in prof_params) {
 
   if (pp == "betat") {
-    prof_values <- seq(-0.15, 0.05, length.out = 30)
+    prof_values <- seq(-0.15, 0.05, length.out = 31)
   } else if (pp == 'tau_epi') {
-    prof_values <- seq(180, 1800, length.out = 30)
+    prof_values <- seq(180, 1800, length.out = 31)
   } else if (pp == 'tau_end') {
-    prof_values <- seq(50, 1800, length.out = 30)
+    prof_values <- seq(50, 1600, length.out = 30)
   } else if (pp == 'rho') {
-    prof_values <- seq(0.1, 1, length.out = 30)
+    prof_values <- seq(0.25, 1, length.out = 30)
   } else if (pp == 'nu') {
-    prof_values <- seq(0.94, 1, length.out = 20)
+    prof_values <- seq(0.94, 1, length.out = 21)
   } else if (pp == 'sig_sq_epi') {
-    prof_values <- seq(0.075, 0.125, length.out = 30)
+    prof_values <- seq(0.075, 0.125, length.out = 31)
   } else if (pp == 'sig_sq_end') {
-    prof_values <- seq(0.05, 0.2, length.out = 30)
+    prof_values <- seq(0.05, 0.25, length.out = 30)
   } else if (pp == 'E_0') {
     prof_values <- seq(10 / 10911819, 5000 / 10911819, 100 / 10911819)
   } else if (pp == 'I_0') {
@@ -97,6 +97,10 @@ for (pp in prof_params) {
   prof_cols <- matrix(rep(prof_values, each = nprof), ncol = 1)
   colnames(prof_cols) <- pp
   tmp_pars[, pp] <- prof_values
+
+  if (pp == "E_0" || pp == 'I_0') {
+    tmp_pars[, 'S_0'] <- unname(1 - tmp_pars[, 'E_0'] - tmp_pars[, 'I_0'])
+  }
 
   bounds <- tibble::tribble(
[…]

### 2024-01-16 H1profile.R
Redoing profiles for I_0 and E_0

@@ -30,7 +30,7 @@ COOLING    <- 0.5
 # Create Experiment Registry ----------------------------------------------
 
 reg <- makeExperimentRegistry(
-  file.dir = paste0('model1/profileReg_RL', RUN_LEVEL, '_v2'),
+  file.dir = paste0('model1/profileReg_RL', RUN_LEVEL, '_v3'),
   seed = 739164,
   packages = c("spatPomp", 'haitipkg', 'pomp')
 )
@@ -50,13 +50,13 @@ h1 <- haiti1_joint()
 coef(h1) <- best_pars
 
 prof_params <- c(
-  'betat',
-  'tau_epi',
-  'tau_end',
-  'rho',
-  'nu',
-  'sig_sq_epi',
-  'sig_sq_end',
+  # 'betat',
+  # 'tau_epi',
+  # 'tau_end',
+  # 'rho',
+  # 'nu',
+  # 'sig_sq_epi',
+  # 'sig_sq_end',
   'E_0',
   'I_0'
 )
@@ -80,9 +80,9 @@ for (pp in prof_params) {
   } else if (pp == 'sig_sq_end') {
     prof_values <- seq(0.05, 0.25, length.out = 30)
   } else if (pp == 'E_0') {
-    prof_values <- seq(10 / 10911819, 5000 / 10911819, 100 / 10911819)
+    prof_values <- seq(1000 / 10911819, 8000 / 10911819, 50 / 10911819)
   } else if (pp == 'I_0') {
-    prof_values <- seq(10 / 10911819, 25000 / 10911819, 500 / 10911819)
+    prof_values <- seq(1 / 10911819, 5000 / 10911819, 50 / 10911819)
   }
 
   tmp_pars <- matrix(

### 2024-01-16 H3profile.R
Redoing profile for XthetaA

@@ -26,12 +26,12 @@ NP <- switch(RUN_LEVEL, 50, 750, 1000)
 NP_EVAL <- switch(RUN_LEVEL, 100, 1000, 2000)
 NREPS_EVAL <- switch(RUN_LEVEL, 3, 6, 10)
 SPAT_REGRESSION <- 0.05
-COOLING <- 0.25
+COOLING <- 0.3
 
 # Create Experiment Registry ----------------------------------------------
 
 reg <- makeExperimentRegistry(
-  file.dir = paste0('model3/profileReg_RL', RUN_LEVEL, '_v5'),
+  file.dir = paste0('model3/profileReg_RL', RUN_LEVEL, '_v6'),
   seed = 739164,
   packages = c("spatPomp", 'haitipkg', 'pomp')
 )
@@ -58,8 +58,8 @@ est_param_names <- c(
 
 # prof_params <- shared_param_names
 prof_params <- c(
-  "epsilon", "k", "lambdaR", "r", "thetaI", "XthetaA"
-  )
+  "XthetaA"
+)
 
 final_pars <- best_pars
 prof_vars <- c()
@@ -68,7 +68,7 @@ for (pp in prof_params) {
   if (pp == "mu_B") {
     prof_values <- seq(200, 650, length.out = 25)
   } else if (pp == 'XthetaA') {
-    prof_values <- seq(0.005, 0.1, length.out = 21)
+    prof_values <- seq(0.002, 0.125, length.out = 31)
   } else if (pp == 'thetaI') {
     prof_values <- seq(0.000075, 0.00016, length.out = 20)
   } else if (pp == 'lambdaR') {
@@ -167,10 +167,10 @@ for (pp in prof_params) {
     "XthetaA" = 0.01,
     "thetaI" = 2.5e-05,
     "lambdaR" = 0.75,
-    "r" = 0.25,
-    "std_W" = 0.025,
+    "r" = 0.2,
+    "std_W" = 0.0275,
     "epsilon" = 0.8,
-    "k" = 30
+    "k" = 25
   )
 
   shared_upper_bounds <- c(
@@ -178,10 +178,10 @@ for (pp in prof_params) {
     "XthetaA" = 0.15,
     "thetaI" = 1e-04,
     "lambdaR" = 3.5,
-    "r" = 1.25,
-    "std_W" = 0.04,
+    "r" = 1,
+    "std_W" = 0.0375,
     "epsilon" = 0.99,
[…]

### 2024-01-16 r1/response.tex

@@
 
+Our starting point was the formulation of \cite{lee20}.
+Judging by the referee's comment, they will not be suprised to learn that this structure led to difficulties in some situations. \TODO{}
+
 \report{
@@
 }
+
+The cholera pathogen was recently introduced to Haiti, which facilitates the specification of initial conditions.
+The goals of inferring initial prevalence is to reduce the number of parameters.
+The risk is that the form of this assumption could have substantial effects on the dynamics, and hence on the consequences of the analysis.
+Our approaches to addressing this issue involve: (i) looking at residual diagnostic plots to assess time points at which the model is misspecified; (ii) if the initial values seem problematic, calculate the room for improvement by estimating the initial values as free parameters.
 
@@
 }
+
+This is similar to point 11 of referee 1.
+LET'S DISCUSS THIS.
+Our point is that there are many other ways one might include time-varying effects in the model, and a strongly supported conclusion should avoid ruling out untested hypotheses.
+We can say that the best available model with trending parameters is statistically superior to the best available model with time-constant parameters.
+This frames the evidence in a way that avoids asserting clear support for any one particular way of modeling this effect.
+
+R1 mentioned the possibility of reporting rate decreasing instead.
+I'M TRYING TO RECALL OUR REASONING THAT MODEL 3 IN LEE ET AL OVER-EMPHASIZED THIS.
+APART FROM THE POSSIBILITY OF A CASE DEFINITION CHANGE, ONE MIGHT EXPECT REPORTING RATE TO INCREASE?
 
@@
 }
+
+The benchmarks we have looked at operate independently on each unit, and in this case the joint log-likelihood benchmark for units 1 and 2 is the sum of the marginal log-likelihoods.
+A simple statistical benchmark does not need to have this structure; for example, one could consider an autoregressive model including spatial dependence.
+
+In general, a joint model for units 1 and 2 may not have well-defined individual log-likelihoods log-lik y1 and log-lik y2 that sum to the full likelihood.
+An exception arises in the case of likelihood evaluation by the block particle filter, in which conditinal log likelihoods for each block of units do sum to the total log likelihood estimate.
+If the block approximation is found to be appropriate \cite{ionides21,ionides22} then fits can be compared against an independent benchmark for each block, and then summed to compare on the entire dataset.
+
+Since the value of benchmarks is a major point in this paper, we have added these extra details in a supplementary section \TODO{}.
 

### 2024-01-16 ms.Rnw
Temporarily changed bake calls to "readRDS". Also fixed a few citation bugs.
pomp version: 5.5
spatPomp version: 0.34.1
haitipkg version: 3.0.0.9002 (the updates to the package shouldn't matter removing the bake calls)
R Version: 4.2.3

@@
 # Compute the run_level_<RUN_LEVEL> results and save them.
-lee_epi_ll <- bake(
-  file = paste0("model1/", rl_dir, "lee1_epi_evals.rds"), {
-
-    # Randomly sample from starting params used by Lee et al (2020).
-    # If RUN_LEVEL == 3, all starting parameters are used.
-    which_params <- sample(
-      1:nrow(h1LeeStartsEpi),
-      size = N_PARAMS,
-      replace = FALSE
-    )
-    epi_params <- h1LeeStartsEpi[which_params, ]
-
-    # Create data.frame to save results
-    lee_epi_ll <- data.frame(
-      "pfLL" = NA_real_,
-      "pfse" = NA_real_,
-      "parid" = epi_params$parid
-    )
-
-    for (j in 1:nrow(lee_epi_ll)) {
-      # Get parameters of interest
-      pf_params <- epi_params[j, -1][names(coef(h1_epi))]
-
-      # make results reproducible
-      registerDoRNG((j * 687383921) %% 7919)
-
-      # Calculate log likelihoods
-      ll_evals <- foreach(i=1:NREPS_EVAL, .combine = c) %dopar% {
-        logLik(pfilter(h1_epi, params = pf_params, Np = NP_EVAL))
-      }
-      lee_epi_ll[j, 'pfLL'] <- logmeanexp(ll_evals, se = TRUE)[1]  # Log-lik
-      lee_epi_ll[j, 'pfse'] <- logmeanexp(ll_evals, se = TRUE)[2]  # SE
-    }
-
-    lee_epi_ll
-  }
-)
+lee_epi_ll <- readRDS("model1/run_level_3/lee1_epi_evals.rds")
+# lee_epi_ll <- bake(
+#   file = paste0("model1/", rl_dir, "lee1_epi_evals.rds"), {
+#
+#     # Randomly sample from starting params used by Lee et al (2020).
+#     # If RUN_LEVEL == 3, all starting parameters are used.
+#     which_params <- sample(
+#       1:nrow(h1LeeStartsEpi),
+#       size = N_PARAMS,
+#       replace = FALSE
+#     )
+#     epi_params <- h1LeeStartsEpi[which_params, ]
+#
+#     # Create data.frame to save results
+#     lee_epi_ll <- data.frame(
+#       "pfLL" = NA_real_,
+#       "pfse" = NA_real_,
+#       "parid" = epi_params$parid
+#     )
+#
+#     for (j in 1:nrow(lee_epi_ll)) {
+#       # Get parameters of interest
+#       pf_params <- epi_params[j, -1][names(coef(h1_epi))]
+#
+#       # make results reproducible
+#       registerDoRNG((j * 687383921) %% 7919)
+#
+#       # Calculate log likelihoods
+#       ll_evals <- foreach(i=1:NREPS_EVAL, .combine = c) %dopar% {
+#         logLik(pfilter(h1_epi, params = pf_params, Np = NP_EVAL))
+#       }
+#       lee_epi_ll[j, 'pfLL'] <- logmeanexp(ll_evals, se = TRUE)[1]  # Log-lik
+#       lee_epi_ll[j, 'pfse'] <- logmeanexp(ll_evals, se = TRUE)[2]  # SE
+#     }
+#
+#     lee_epi_ll
+#   }
+# )
 
@@
 
-lee_end_ll <- bake(
-  file = paste0("model1/", rl_dir, "lee1_end_evals.rds"), {
-
-    end_params <- h1LeeStartsEnd[h1LeeStartsEnd$parid %in% lee_epi_ll$parid, ]
-
-    # Create data.frame to save results
-    lee_end_ll <- data.frame(
-      "pfLL" = NA_real_,
-      "pfse" = NA_real_,
-      'parid' = end_params$parid
-    )
-
-    # Perform particle filter to get likelihood estimates
-    for (j in 1:nrow(lee_end_ll)) {
-      # Get parameters of interest
-      pf_params <- end_params[j, ][names(coef(h1_end))]
-
-      # make results reproducible
-      registerDoRNG((j * 687383921) %% 7919)
-
-      # Calculate log likelihoods
-      ll_evals <- foreach(i=1:NREPS_EVAL, .combine = c) %dopar% {
-        logLik(pfilter(h1_end, params = pf_params, Np = NP_EVAL))
-      }
-
-      lee_end_ll[j, 'pfLL'] <- logmeanexp(ll_evals, se = TRUE)[1]
-      lee_end_ll[j, 'pfse'] <- logmeanexp(ll_evals, se = TRUE)[2]
-    }
-
-    lee_end_ll
-  }
-)
+lee_end_ll <- readRDS("model1/run_level_3/lee1_end_evals.rds")
+# lee_end_ll <- bake(
+#   file = paste0("model1/", rl_dir, "lee1_end_evals.rds"), {
+#
+#     end_params <- h1LeeStartsEnd[h1LeeStartsEnd$parid %in% lee_epi_ll$parid, ]
+#
+#     # Create data.frame to save results
+#     lee_end_ll <- data.frame(
+#       "pfLL" = NA_real_,
+#       "pfse" = NA_real_,
+#       'parid' = end_params$parid
+#     )
+#
+#     # Perform particle filter to get likelihood estimates
+#     for (j in 1:nrow(lee_end_ll)) {
+#       # Get parameters of interest
+#       pf_params <- end_params[j, ][names(coef(h1_end))]
+#
+#       # make results reproducible
+#       registerDoRNG((j * 687383921) %% 7919)
+#
+#       # Calculate log likelihoods
+#       ll_evals <- foreach(i=1:NREPS_EVAL, .combine = c) %dopar% {
+#         logLik(pfilter(h1_end, params = pf_params, Np = NP_EVAL))
+#       }
+#
+#       lee_end_ll[j, 'pfLL'] <- logmeanexp(ll_evals, se = TRUE)[1]
+#       lee_end_ll[j, 'pfse'] <- logmeanexp(ll_evals, se = TRUE)[2]
+#     }
+#
+#     lee_end_ll
+#   }
+# )
 
@@
 This case study investigates the use of dynamic models and spatiotemporal data to inform public health policy in the context of the cholera outbreak in Haiti, which started in 2010.
-Various dynamic models were developed to study this outbreak: searching PubMed with keywords ``Haiti, cholera, model'' we obtained 20 deterministic models \cite{kirpich16,kirpich17,andrews11,botelho21,fitzgibbon20,eisenberg13,rinaldo12,chao11,date11,lin19,collins21,akman15,trevisin22,mavian20,collins14,kelly16,capone14,bengtsson15,mari15,gatto12} and 8 stochastic models \cite{tuite11,mukandvire13,lewnard16,kunke17,sallah17,azman12,kuhn14,azman15}.
+Various dynamic models were developed to study this outbreak: searching PubMed with keywords ``Haiti, cholera, model'' we obtained 20 deterministic models \cite{kirpich16,kirpich17,andrews11,botelho21,fitzgibbon20,eisenberg13,rinaldo12,chao11,date11,lin19,collins21,akman15,trevisin22,mavian20,collins14,kelly16,capone15,bengtsson15,mari15,gatto12} and 8 stochastic models \cite{tuite11,mukandavire13,lewnard16,kunkel17,sallah17,azman12,kuhn14,azman15}.
 Here we focus on a multi-group modeling exercise by Lee et al.~\cite{lee20} in which four expert modeling teams developed models to the same dataset with the goal of comparing conclusions on the feasibility of eliminating cholera by a vaccination campaign.
@@
 # Create the <RUN_LEVEL> version of the results and save them.
-h1_MIF <- bake(
-  file = paste0("model1/", rl_dir, "haiti1_fit.rds"), {
-    fit_haiti1(  # haitipkg function used to fit Model 1
-      NP = NP_H1,
-      NMIF = NMIF_H1,
-      NUM_TREND = NUM_TREND,
-      NPROF = NPROF,
-      NREPS_EVAL = NREPS_EVAL,
-      NP_EVAL = NP_EVAL
-    )
-  }
-)
+h1_MIF <- readRDS('model1/run_level_3/haiti1_fit.rds')
+
+# h1_MIF <- bake(
+#   file = paste0("model1/", rl_dir, "haiti1_fit.rds"), {
+#     fit_haiti1(  # haitipkg function used to fit Model 1
+#       NP = NP_H1,
+#       NMIF = NMIF_H1,
+#       NUM_TREND = NUM_TREND,
+#       NPROF = NPROF,
+#       NREPS_EVAL = NREPS_EVAL,
+#       NP_EVAL = NP_EVAL
+#     )
+#   }
+# )
 
@@
 
-h2_fit <- bake(
-  file = paste0("model2/model2_fit.rds"), {
-    fit_haiti2()
-  },
-  timing = FALSE
-)
+h2_fit <- readRDS("model2/model2_fit.rds")
+
+# h2_fit <- bake(
+#   file = paste0("model2/model2_fit.rds"), {
+#     fit_haiti2()
+#   },
+#   timing = FALSE
+# )
 @
@@
 # Fit Model 3
-haiti3_fit <- bake(
-  file = paste0("model3/", rl_dir, "haiti3_fit.rds"), {
-    fit_haiti3(
-      nsearches = 2L,
-      search1 = SEARCH1,
-      search2 = SEARCH2,
-      search_rho = FALSE,
-      search_gamma = FALSE,
-      search_Iinit = TRUE,
-      search_hur = TRUE
-    )
-  }
-)
+haiti3_fit <- readRDS("model3/run_level_3/haiti3_fit.rds")
+# haiti3_fit <- bake(
+#   file = paste0("model3/", rl_dir, "haiti3_fit.rds"), {
+#     fit_haiti3(
+#       nsearches = 2L,
+#       search1 = SEARCH1,
+#       search2 = SEARCH2,
+#       search_rho = FALSE,
+#       search_gamma = FALSE,
+#       search_Iinit = TRUE,
+#       search_hur = TRUE
+#     )
+#   }
+# )
 
@@
 # Calculate likelihood via block-particle filter
-mod3_eval_res <- bake(
-  file = paste0('model3/', rl_dir, 'mod3_eval_res.rds'), {
-
-    # Create a list to store results
-    results <- list()
-
-    coef(h3Spat) <- p3
-
-    # Perform block-particle filter
-    h3_bpf <- foreach(i = 1:NREPS_EVAL, .combine = c) %dopar% {
-      bpfilter(h3Spat, Np = NP_EVAL, block_size = 1)
-    }
-
-    ll <- logLik(h3_bpf)
-    results$ll <- logmeanexp(ll[!is.na(ll)], se = FALSE)
-
-    ### Get likelihood for subset of the data
-
-    # Find appropriate sub-set
-    in_subset_cols <- h3_bpf[[1]]@times >= as.numeric(lubridate::decimal_date(as.Date("2014-03-01")))
-    in_subset_rows <- h3_bpf[[1]]@unit_names != 'Ouest'
-
-    mod3_coupled_subset_evals <- c()
-    for (i in 1:length(h3_bpf)) {
-      mod3_coupled_subset_evals <- c(
-        mod3_coupled_subset_evals,
-        sum(h3_bpf[[i]]@block.cond.loglik[in_subset_rows, in_subset_cols])
-      )
-    }
-
-    results$subset_ll <- logmeanexp(mod3_coupled_subset_evals)
-    results
-  },
-  timing = FALSE
-)
+mod3_eval_res <- readRDS("model3/run_level_3/mod3_eval_res.rds")
+# mod3_eval_res <- bake(
+#   file = paste0('model3/', rl_dir, 'mod3_eval_res.rds'), {
+#
+#     # Create a list to store results
+#     results <- list()
+#
+#     coef(h3Spat) <- p3
+#
+#     # Perform block-particle filter
+#     h3_bpf <- foreach(i = 1:NREPS_EVAL, .combine = c) %dopar% {
+#       bpfilter(h3Spat, Np = NP_EVAL, block_size = 1)
+#     }
+#
+#     ll <- logLik(h3_bpf)
+#     results$ll <- logmeanexp(ll[!is.na(ll)], se = FALSE)
+#
+#     ### Get likelihood for subset of the data
+#
+#     # Find appropriate sub-set
+#     in_subset_cols <- h3_bpf[[1]]@times >= as.numeric(lubridate::decimal_date(as.Date("2014-03-01")))
+#     in_subset_rows <- h3_bpf[[1]]@unit_names != 'Ouest'
+#
+#     mod3_coupled_subset_evals <- c()
+#     for (i in 1:length(h3_bpf)) {
+#       mod3_coupled_subset_evals <- c(
+#         mod3_coupled_subset_evals,
+#         sum(h3_bpf[[i]]@block.cond.loglik[in_subset_rows, in_subset_cols])
+#       )
+#     }
+#
+#     results$subset_ll <- logmeanexp(mod3_coupled_subset_evals)
+#     results
+#   },
+#   timing = FALSE
+# )
 
@@
 
-lee3_results <- bake(
-  file = paste0('model3/', rl_dir, 'lee3_results.rds'), {
-
-    # Create lists to save results
-    lee3_results <- list()
-
-    # Save how long it takes for sub-data
-    lee3_results[["sub_bpf_time"]] <- system.time({
-
-      lee3_sub_ll_evals <- foreach(
-        i = 1:lee3_NREP, .combine = rbind
-      ) %dopar% {
-
-        liks <- numeric(10L)
-
-        lee_bpf_out <- bpfilter(lee3_sub, block_size = 1, Np = lee3_NP)
-        Ouest_ind <- which(lee_bpf_out@unit_names == "Ouest")
-
-        for (u in 1:10) {
-          if (u == Ouest_ind) {
-            liks[u] <- sum(lee_bpf_out@block.cond.loglik[u, lee3_sub@times >= lubridate::decimal_date(as.Date('2017-06-10'))])
-          } else {
-            liks[u] <- sum(lee_bpf_out@block.cond.loglik[u, ])
-          }
-        }
-
-        liks
-      }
-    })
-
-    # Save all likelihood evaluations
-    lee3_results[["sub_ll_eval"]] <- logmeanexp(rowSums(lee3_sub_ll_evals)[!is.na(rowSums(lee3_sub_ll_evals))])
-
-    registerDoRNG(2789131)
-
-    # Save how long it takes for all data
-    lee3_results[["all_pf_time"]] <- system.time({
-      # Save all likelihood evaluations
-
-      lee3_all_ll_evals <- foreach(
-        i = 1:lee3_NREP, .combine = c
-      ) %dopar% {
-        logLik(bpfilter(lee3_all, block_size = 1, Np = lee3_NP))  # Get likelihood evaluations
-      }
-
-      lee3_results[["all_ll_eval"]] <- logmeanexp(lee3_all_ll_evals[!is.na(lee3_all_ll_evals)])
-
-    })
-
-    lee3_results[["n_params"]] <- 29
-
-    lee3_results
-  },
-  timing = FALSE
-)
+lee3_results <- readRDS("model3/run_level_3/lee3_results.rds")
+# lee3_results <- bake(
+#   file = paste0('model3/', rl_dir, 'lee3_results.rds'), {
+#
+#     # Create lists to save results
+#     lee3_results <- list()
+#
+#     # Save how long it takes for sub-data
+#     lee3_results[["sub_bpf_time"]] <- system.time({
+#
+#       lee3_sub_ll_evals <- foreach(
+#         i = 1:lee3_NREP, .combine = rbind
+#       ) %dopar% {
+#
+#         liks <- numeric(10L)
+#
+#         lee_bpf_out <- bpfilter(lee3_sub, block_size = 1, Np = lee3_NP)
+#         Ouest_ind <- which(lee_bpf_out@unit_names == "Ouest")
+#
+#         for (u in 1:10) {
+#           if (u == Ouest_ind) {
+#             liks[u] <- sum(lee_bpf_out@block.cond.loglik[u, lee3_sub@times >= lubridate::decimal_date(as.Date('2017-06-10'))])
+#           } else {
+#             liks[u] <- sum(lee_bpf_out@block.cond.loglik[u, ])
+#           }
+#         }
+#
+#         liks
+#       }
+#     })
+#
+#     # Save all likelihood evaluations
+#     lee3_results[["sub_ll_eval"]] <- logmeanexp(rowSums(lee3_sub_ll_evals)[!is.na(rowSums(lee3_sub_ll_evals))])
+#
+#     registerDoRNG(2789131)
+#
+#     # Save how long it takes for all data
+#     lee3_results[["all_pf_time"]] <- system.time({
+#       # Save all likelihood evaluations
+#
+#       lee3_all_ll_evals <- foreach(
+#         i = 1:lee3_NREP, .combine = c
+#       ) %dopar% {
+#         logLik(bpfilter(lee3_all, block_size = 1, Np = lee3_NP))  # Get likelihood evaluations
+#       }
+#
+#       lee3_results[["all_ll_eval"]] <- logmeanexp(lee3_all_ll_evals[!is.na(lee3_all_ll_evals)])
+#
+#     })
+#
+#     lee3_results[["n_params"]] <- 29
+#
+#     lee3_results
+#   },
+#   timing = FALSE
+# )
 
@@
 # speed for each run-level.
-mod2_VaccScenarios <- bake(
-  file = paste0("model2/VaccinationScenarios.rds"), {
-    haiti2_vaccScenario(
-      h2_params = h2_fit$h2_params
-    )
-  },
-  timing = FALSE
-)
+mod2_VaccScenarios <- readRDS("model2/VaccinationScenarios.rds")
+# mod2_VaccScenarios <- bake(
+#   file = paste0("model2/VaccinationScenarios.rds"), {
+#     haiti2_vaccScenario(
+#       h2_params = h2_fit$h2_params
+#     )
+#   },
+#   timing = FALSE
+# )
 @
@@
 # Get quantiles of simulations from the calibrated Model 3.
-h3_quants <- bake(
-  file = paste0('model3/', rl_dir, 'h3Sims.rds'), {
-
-    coef(h3Spat) <- p3
-    # Simulate from the fitted version of Model 3
-    h3_sims <- simulate(
-      h3Spat, nsim = h3_nsim, format = 'data.frame'
-    )
-
-    # For each department and time-point, calculate percentiles of the
-    # simulations
-    h3_sims %>%
-      rename(dep = unitname) %>%
-      group_by(dep, time) %>%
-      summarise(
-        q05 = quantile(cases, 0.025, na.rm = T),
-        mean = mean(cases, na.rm = T),
-        q50 = quantile(cases, 0.5, na.rm = T),
-        q95 = quantile(cases, 0.975, na.rm = T)
-      ) %>%
-      ungroup() %>%
-      mutate(
-        date = lubridate::date_decimal(time)
-      ) %>%
-      mutate(date = as.Date(lubridate::round_date(date, unit = 'day')))
-  },
-  timing = FALSE
-)
+h3_quants <- readRDS("model3/run_level_3/h3Sims.rds")
+# h3_quants <- bake(
+#   file = paste0('model3/', rl_dir, 'h3Sims.rds'), {
+#
+#     coef(h3Spat) <- p3
+#     # Simulate from the fitted version of Model 3
+#     h3_sims <- simulate(
+#       h3Spat, nsim = h3_nsim, format = 'data.frame'
+#     )
+#
+#     # For each department and time-point, calculate percentiles of the
+#     # simulations
+#     h3_sims %>%
+#       rename(dep = unitname) %>%
+#       group_by(dep, time) %>%
+#       summarise(
+#         q05 = quantile(cases, 0.025, na.rm = T),
+#         mean = mean(cases, na.rm = T),
+#         q50 = quantile(cases, 0.5, na.rm = T),
+#         q95 = quantile(cases, 0.975, na.rm = T)
+#       ) %>%
+#       ungroup() %>%
+#       mutate(
+#         date = lubridate::date_decimal(time)
+#       ) %>%
+#       mutate(date = as.Date(lubridate::round_date(date, unit = 'day')))
+#   },
+#   timing = FALSE
+# )
 
@@
 # projects the vaccination scenario, under the specified model.
-h1_S0_results <- bake(
-  file = paste0("model1/", rl_dir, "S0_sims.rds"), {
-
-    # Create the model
-    s0 <- haiti1_joint(vacscen = 'id0')
-
-    # Load coefficients
-    coef(s0) <- p1
-
-    # Get samples from filtering distribution
-    s0_pf <- pfilter(s0, save.state = TRUE, Np = h1_scen_NP)
-    end_states <- saved_states(s0_pf)
-    end_states <- end_states[[length(end_states)]]
-
-    # project vacc scenario, with filtering dist, and model.
-    h1_S0_sims <- project_from_filter(mod = s0, end_states = end_states, nsims = h1_scen_sims)
-
-    results <- list()
-    results$mod1_V0_sims <- agg_mod1_sims(h1_S0_sims)
-    results$mod1_V0_probs <- get_elimProbs(h1_S0_sims, model = 1)
-
-    results
-  },
-  timing = FALSE
-)
+h1_S0_results <- readRDS("model1/run_level_3/S0_sims.rds")
+# h1_S0_results <- bake(
+#   file = paste0("model1/", rl_dir, "S0_sims.rds"), {
+#
+#     # Create the model
+#     s0 <- haiti1_joint(vacscen = 'id0')
+#
+#     # Load coefficients
+#     coef(s0) <- p1
+#
+#     # Get samples from filtering distribution
+#     s0_pf <- pfilter(s0, save.state = TRUE, Np = h1_scen_NP)
+#     end_states <- saved_states(s0_pf)
+#     end_states <- end_states[[length(end_states)]]
+#
+#     # project vacc scenario, with filtering dist, and model.
+#     h1_S0_sims <- project_from_filter(mod = s0, end_states = end_states, nsims = h1_scen_sims)
+#
+#     results <- list()
+#     results$mod1_V0_sims <- agg_mod1_sims(h1_S0_sims)
+#     results$mod1_V0_probs <- get_elimProbs(h1_S0_sims, model = 1)
+#
+#     results
+#   },
+#   timing = FALSE
+# )
 
@@
 # projects the vaccination scenario, under the specified model.
-h1_S2_results <- bake(
-  file = paste0('model1/', rl_dir, 'S2_sims.rds'), {
-
-    # Create the model
-    s2 <- haiti1_joint(vacscen = "id2")
-
-    # Set model parameters that need to be adjusted
-    depts <- 2
-    h1_par_names <- names(p1)
-    h1_params_temp <- c(p1, rep(0.0, 5 * depts))
-    names(h1_params_temp) <- c(
-      h1_par_names,
-      paste0("S", 1:depts, "_0"),
-      paste0("E", 1:depts, "_0"),
-      paste0("I", 1:depts, "_0"),
-      paste0("A", 1:depts, "_0"),
-      paste0("R", 1:depts, "_0")
-    )
-    coef(s2) <- h1_params_temp  # Set model parameters
-
-    # Get samples from filtering distribution for Scenario 2 model:
-    s2_pf <- pfilter(s2, save.state = TRUE, Np = h1_scen_NP)
-    end_states <- saved_states(s2_pf)
-    end_states <- end_states[[length(end_states)]]
-
-    # project vacc scenario, with filtering dist, and model.
-    h1_S2_sims <- project_from_filter(mod = s2, end_states = end_states, nsims = h1_scen_sims)
-
-    results <- list()
-    results$mod1_V1_sims <- agg_mod1_sims(h1_S2_sims)
-    results$mod1_V1_probs <- get_elimProbs(h1_S2_sims, model = 1)
-
-    results
-  },
-  timing = FALSE
-)
+h1_S2_results <- readRDS("model1/run_level_3/S2_sims.rds")
+# h1_S2_results <- bake(
+#   file = paste0('model1/', rl_dir, 'S2_sims.rds'), {
+#
+#     # Create the model
+#     s2 <- haiti1_joint(vacscen = "id2")
+#
+#     # Set model parameters that need to be adjusted
+#     depts <- 2
+#     h1_par_names <- names(p1)
+#     h1_params_temp <- c(p1, rep(0.0, 5 * depts))
+#     names(h1_params_temp) <- c(
+#       h1_par_names,
+#       paste0("S", 1:depts, "_0"),
+#       paste0("E", 1:depts, "_0"),
+#       paste0("I", 1:depts, "_0"),
+#       paste0("A", 1:depts, "_0"),
+#       paste0("R", 1:depts, "_0")
+#     )
+#     coef(s2) <- h1_params_temp  # Set model parameters
+#
+#     # Get samples from filtering distribution for Scenario 2 model:
+#     s2_pf <- pfilter(s2, save.state = TRUE, Np = h1_scen_NP)
+#     end_states <- saved_states(s2_pf)
+#     end_states <- end_states[[length(end_states)]]
+#
+#     # project vacc scenario, with filtering dist, and model.
+#     h1_S2_sims <- project_from_filter(mod = s2, end_states = end_states, nsims = h1_scen_sims)
+#
+#     results <- list()
+#     results$mod1_V1_sims <- agg_mod1_sims(h1_S2_sims)
+#     results$mod1_V1_probs <- get_elimProbs(h1_S2_sims, model = 1)
+#
+#     results
+#   },
+#   timing = FALSE
+# )
 
@@
 # projects the vaccination scenario, under the specified model.
-h1_S4_results <- bake(
-  file = paste0('model1/', rl_dir, 'S4_sims.rds'), {
-
-    # Create the model
-    s4 <- haiti1_joint(vacscen = "id4")
-
-    # Set model parameters that need to be adjusted
-    depts <- 3
-    h1_par_names <- names(p1)
-    h1_params_temp <- c(p1, rep(0.0, 5 * depts))
-    names(h1_params_temp) <- c(
-      h1_par_names,
-      paste0("S", 1:depts, "_0"),
-      paste0("E", 1:depts, "_0"),
-      paste0("I", 1:depts, "_0"),
-      paste0("A", 1:depts, "_0"),
-      paste0("R", 1:depts, "_0")
-    )
-    coef(s4) <- h1_params_temp  # Set model parameters
-
-    # Get samples from filtering distribution for Scenario 2 model:
-    s4_pf <- pfilter(s4, save.state = TRUE, Np = h1_scen_NP)
-    end_states <- saved_states(s4_pf)
-    end_states <- end_states[[length(end_states)]]
-
-    # project vacc scenario, with filtering dist, and model.
-    h1_S4_sims <- project_from_filter(mod = s4, end_states = end_states, nsims = h1_scen_sims)
-
-    results <- list()
-    results$mod1_V2_sims <- agg_mod1_sims(h1_S4_sims)
-    results$mod1_V2_probs <- get_elimProbs(h1_S4_sims, model = 1)
-
-    results
-  },
-  timing = FALSE
-)
+h1_S4_results <- readRDS("model1/run_level_3/S4_sims.rds")
+# h1_S4_results <- bake(
+#   file = paste0('model1/', rl_dir, 'S4_sims.rds'), {
+#
+#     # Create the model
+#     s4 <- haiti1_joint(vacscen = "id4")
+#
+#     # Set model parameters that need to be adjusted
+#     depts <- 3
+#     h1_par_names <- names(p1)
+#     h1_params_temp <- c(p1, rep(0.0, 5 * depts))
+#     names(h1_params_temp) <- c(
+#       h1_par_names,
+#       paste0("S", 1:depts, "_0"),
+#       paste0("E", 1:depts, "_0"),
+#       paste0("I", 1:depts, "_0"),
+#       paste0("A", 1:depts, "_0"),
+#       paste0("R", 1:depts, "_0")
+#     )
+#     coef(s4) <- h1_params_temp  # Set model parameters
+#
+#     # Get samples from filtering distribution for Scenario 2 model:
+#     s4_pf <- pfilter(s4, save.state = TRUE, Np = h1_scen_NP)
+#     end_states <- saved_states(s4_pf)
+#     end_states <- end_states[[length(end_states)]]
+#
+#     # project vacc scenario, with filtering dist, and model.
+#     h1_S4_sims <- project_from_filter(mod = s4, end_states = end_states, nsims = h1_scen_sims)
+#
+#     results <- list()
+#     results$mod1_V2_sims <- agg_mod1_sims(h1_S4_sims)
+#     results$mod1_V2_probs <- get_elimProbs(h1_S4_sims, model = 1)
+#
+#     results
+#   },
+#   timing = FALSE
+# )
 
@@
 # projects the vaccination scenario, under the specified model.
-h1_S3_results <- bake(
-  file = paste0('model1/', rl_dir, 'S3_sims.rds'), {
-
-    # Create the model
-    s3 <- haiti1_joint(vacscen = "id3")
-
-    # Set model parameters that need to be adjusted
-    depts <- 10
-    h1_par_names <- names(p1)
-    h1_params_temp <- c(p1, rep(0.0, 5 * depts))
-    names(h1_params_temp) <- c(
-      h1_par_names,
-      paste0("S", 1:depts, "_0"),
-      paste0("E", 1:depts, "_0"),
-      paste0("I", 1:depts, "_0"),
-      paste0("A", 1:depts, "_0"),
-      paste0("R", 1:depts, "_0")
-    )
-    coef(s3) <- h1_params_temp  # Set model parameters
-
-    # Get samples from filtering distribution for Scenario 2 model:
-    s3_pf <- pfilter(s3, save.state = TRUE, Np = h1_scen_NP)
-    end_states <- saved_states(s3_pf)
-    end_states <- end_states[[length(end_states)]]
-
-    # project vacc scenario, with filtering dist, and model.
-    h1_S3_sims <- project_from_filter(mod = s3, end_states = end_states, nsims = h1_scen_sims)
-
-    results <- list()
-    results$mod1_V3_sims <- agg_mod1_sims(h1_S3_sims)
-    results$mod1_V3_probs <- get_elimProbs(h1_S3_sims, model = 1)
-
-    results
-  },
-  timing = FALSE
-)
+h1_S3_results <- readRDS("model1/run_level_3/S3_sims.rds")
+# h1_S3_results <- bake(
+#   file = paste0('model1/', rl_dir, 'S3_sims.rds'), {
+#
+#     # Create the model
+#     s3 <- haiti1_joint(vacscen = "id3")
+#
+#     # Set model parameters that need to be adjusted
+#     depts <- 10
+#     h1_par_names <- names(p1)
+#     h1_params_temp <- c(p1, rep(0.0, 5 * depts))
+#     names(h1_params_temp) <- c(
+#       h1_par_names,
+#       paste0("S", 1:depts, "_0"),
+#       paste0("E", 1:depts, "_0"),
+#       paste0("I", 1:depts, "_0"),
+#       paste0("A", 1:depts, "_0"),
+#       paste0("R", 1:depts, "_0")
+#     )
+#     coef(s3) <- h1_params_temp  # Set model parameters
+#
+#     # Get samples from filtering distribution for Scenario 2 model:
+#     s3_pf <- pfilter(s3, save.state = TRUE, Np = h1_scen_NP)
+#     end_states <- saved_states(s3_pf)
+#     end_states <- end_states[[length(end_states)]]
+#
+#     # project vacc scenario, with filtering dist, and model.
+#     h1_S3_sims <- project_from_filter(mod = s3, end_states = end_states, nsims = h1_scen_sims)
+#
+#     results <- list()
+#     results$mod1_V3_sims <- agg_mod1_sims(h1_S3_sims)
+#     results$mod1_V3_probs <- get_elimProbs(h1_S3_sims, model = 1)
+#
+#     results
+#   },
+#   timing = FALSE
+# )
 
@@
 # projects the vaccination scenario, under the specified model.
-h1_S1_results <- bake(
-  file = paste0('model1/', rl_dir, 'S1_sims.rds'), {
-
-    # Create the model
-    s1 <- haiti1_joint(vacscen = "id1")
-
-    # Set model parameters that need to be adjusted
-    depts <- 10
-    h1_par_names <- names(p1)
-    h1_params_temp <- c(p1, rep(0.0, 5 * depts))
-    names(h1_params_temp) <- c(
-      h1_par_names,
-      paste0("S", 1:depts, "_0"),
-      paste0("E", 1:depts, "_0"),
-      paste0("I", 1:depts, "_0"),
-      paste0("A", 1:depts, "_0"),
-      paste0("R", 1:depts, "_0")
-    )
-    coef(s1) <- h1_params_temp  # Set model parameters
-
-    # Get samples from filtering distribution for Scenario 2 model:
-    s1_pf <- pfilter(s1, save.state = TRUE, Np = h1_scen_NP)
-    end_states <- saved_states(s1_pf)
-    end_states <- end_states[[length(end_states)]]
-
-    # project vacc scenario, with filtering dist, and model.
-    h1_S1_sims <- project_from_filter(mod = s1, end_states = end_states, nsims = h1_scen_sims)
-
-    results <- list()
-    results$mod1_V4_sims <- agg_mod1_sims(h1_S1_sims)
-    results$mod1_V4_probs <- get_elimProbs(h1_S1_sims, model = 1)
-
-    results
-  },
-  timing = FALSE
-)
+h1_S1_results <- readRDS("model1/run_level_3/S1_sims.rds")
+# h1_S1_results <- bake(
+#   file = paste0('model1/', rl_dir, 'S1_sims.rds'), {
+#
+#     # Create the model
+#     s1 <- haiti1_joint(vacscen = "id1")
+#
+#     # Set model parameters that need to be adjusted
+#     depts <- 10
+#     h1_par_names <- names(p1)
+#     h1_params_temp <- c(p1, rep(0.0, 5 * depts))
+#     names(h1_params_temp) <- c(
+#       h1_par_names,
+#       paste0("S", 1:depts, "_0"),
+#       paste0("E", 1:depts, "_0"),
+#       paste0("I", 1:depts, "_0"),
+#       paste0("A", 1:depts, "_0"),
+#       paste0("R", 1:depts, "_0")
+#     )
+#     coef(s1) <- h1_params_temp  # Set model parameters
+#
+#     # Get samples from filtering distribution for Scenario 2 model:
+#     s1_pf <- pfilter(s1, save.state = TRUE, Np = h1_scen_NP)
+#     end_states <- saved_states(s1_pf)
+#     end_states <- end_states[[length(end_states)]]
+#
+#     # project vacc scenario, with filtering dist, and model.
+#     h1_S1_sims <- project_from_filter(mod = s1, end_states = end_states, nsims = h1_scen_sims)
+#
+#     results <- list()
+#     results$mod1_V4_sims <- agg_mod1_sims(h1_S1_sims)
+#     results$mod1_V4_probs <- get_elimProbs(h1_S1_sims, model = 1)
+#
+#     results
+#   },
+#   timing = FALSE
+# )
 @
@@
 
-h3_bpf_end_states <- bake(
-  file = paste0("model3/", rl_dir, "h3_bpf_end_states.rds"),
-  expr = {
-    coef(h3Spat) <- p3
-    h3_bpf <- bpfilter(h3Spat, Np = NP_BPF, block_size = 1, save_states = TRUE)
-    ss <- h3_bpf@saved.states
-    ss[[length(ss)]]
-  },
-  timing = FALSE
-)
-
-mod3_V0_res <- bake(
-  file = paste0("model3/", rl_dir, "mod3_V0_res.rds"), {
-
-    # Create a list to store results
-    h3_V0_res <- list()
-
-    # Get "noVacc" scenario parameters
-    h3_S0_par <- get_model3_vacc_scenario_params(scenario = "noVacc")
-    coef(h3Spat) <- p3
-    coef(h3Spat)[names(h3_S0_par)] <- h3_S0_par
-
-    # Using parameters and filtered distribution, project system
-    h3_V0_sims <- project_from_filter(
-      h3Spat, end_states = h3_bpf_end_states, covarGen = project_rain,
-      nsims = NSIM, seed = 6897154
-    )
-
-    # Store nationally aggregated results
-    h3_V0_res$mod3_V0_sims <- agg_mod3_sims(h3_V0_sims)
-    h3_V0_res$mod3_V0_probs <- get_elimProbs(h3_V0_sims, model = 3)
-
-    # For the supplement material, save quantiles of the Susceptible individuals
-    h3_V0_res$SQuants <- h3_V0_sims %>%
-      mutate(
-        date = as.Date(lubridate::round_date(lubridate::date_decimal(time), unit = 'day'))
-      ) %>%
-      group_by(unitname, date) %>%
-      mutate(
-        pop = h3Spat@params[paste0("H", which(first(unitname) == unit_names(h3Spat)))]
-      ) %>%
-      summarize(
-        Q025 = quantile((S + VSd + VSdd + VSdd_alt + VSd_alt) / pop, probs = 0.025),
-        Q50 = quantile((S + VSd + VSdd + VSdd_alt + VSd_alt) / pop, probs = 0.5),
-        Q975 = quantile((S + VSd + VSdd + VSdd_alt + VSd_alt) / pop, probs = 0.975)
-      ) %>%
-      ungroup() %>%
-      rename(dep = unitname)
-
-    # Return results
-    h3_V0_res
-
-  }
-)
-
-mod3_V1_res <- bake(
-  file = paste0("model3/", rl_dir, "mod3_V1_res.rds"), {
-
-    # Create a list to store results
-    h3_V1_res <- list()
-
-    # Get "noVacc" scenario parameters
-    h3_V1_par <- get_model3_vacc_scenario_params(scenario = "2dep")
-    coef(h3Spat) <- p3
-    coef(h3Spat)[names(h3_V1_par)] <- h3_V1_par
-
-    # Using parameters and filtered distribution, project system
-    h3_V1_sims <- project_from_filter(
-      h3Spat, end_states = h3_bpf_end_states, covarGen = project_rain,
-      nsims = NSIM, seed = 6897154
-    )
-
-    # Store nationally aggregated results
-    h3_V1_res$mod3_V1_sims <- agg_mod3_sims(h3_V1_sims)
-    h3_V1_res$mod3_V1_probs <- get_elimProbs(h3_V1_sims, model = 3)
-
-    # Return results
-    h3_V1_res
-
-  }
-)
-
-mod3_V2_res <- bake(
-  file = paste0("model3/", rl_dir, "mod3_V2_res.rds"), {
-
-    # Create a list to store results
-    h3_V2_res <- list()
-
-    # Get "noVacc" scenario parameters
-    h3_V2_par <- get_model3_vacc_scenario_params(scenario = "3dep")
-    coef(h3Spat) <- p3
-    coef(h3Spat)[names(h3_V2_par)] <- h3_V2_par
-
-    # Using parameters and filtered distribution, project system
-    h3_V2_sims <- project_from_filter(
-      h3Spat, end_states = h3_bpf_end_states, covarGen = project_rain,
-      nsims = NSIM, seed = 6897154
-    )
-
-    # Store nationally aggregated results
-    h3_V2_res$mod3_V2_sims <- agg_mod3_sims(h3_V2_sims)
-    h3_V2_res$mod3_V2_probs <- get_elimProbs(h3_V2_sims, model = 3)
-
-    # Return results
-    h3_V2_res
-  }
-)
-
-mod3_V3_res <- bake(
-  file = paste0("model3/", rl_dir, "mod3_V3_res.rds"), {
-
-    # Create a list to store results
-    h3_V3_res <- list()
-
-    # Get "noVacc" scenario parameters
-    h3_V3_par <- get_model3_vacc_scenario_params(scenario = "slowNation")
-    coef(h3Spat) <- p3
-    coef(h3Spat)[names(h3_V3_par)] <- h3_V3_par
-
-    # Using parameters and filtered distribution, project system
-    h3_V3_sims <- project_from_filter(
-      h3Spat, end_states = h3_bpf_end_states, covarGen = project_rain,
-      nsims = NSIM, seed = 6897154
-    )
-
-    # Store nationally aggregated results
-    h3_V3_res$mod3_V3_sims <- agg_mod3_sims(h3_V3_sims)
-    h3_V3_res$mod3_V3_probs <- get_elimProbs(h3_V3_sims, model = 3)
-
-    # Return results
-    h3_V3_res
-
-  }
-)
-
-mod3_V4_res <- bake(
-  file = paste0("model3/", rl_dir, "mod3_V4_res.rds"), {
-
-    # Create a list to store results
-    h3_V4_res <- list()
-
-    # Get "noVacc" scenario parameters
-    h3_V4_par <- get_model3_vacc_scenario_params(scenario = "fastNation")
-    coef(h3Spat) <- p3
-    coef(h3Spat)[names(h3_V4_par)] <- h3_V4_par
-
-    # Using parameters and filtered distribution, project system
-    h3_V4_sims <- project_from_filter(
-      h3Spat, end_states = h3_bpf_end_states, covarGen = project_rain,
-      nsims = NSIM, seed = 6897154
-    )
-
-    # Store nationally aggregated results
-    h3_V4_res$mod3_V4_sims <- agg_mod3_sims(h3_V4_sims)
-    h3_V4_res$mod3_V4_probs <- get_elimProbs(h3_V4_sims, model = 3)
-
-    # Return results
-    h3_V4_res
-
-  }
-)
+h3_bpf_end_stats <- readRDS("model3/run_level_3/h3_bpf_end_states.rds")
+# h3_bpf_end_states <- bake(
+#   file = paste0("model3/", rl_dir, "h3_bpf_end_states.rds"),
+#   expr = {
+#     coef(h3Spat) <- p3
+#     h3_bpf <- bpfilter(h3Spat, Np = NP_BPF, block_size = 1, save_states = TRUE)
+#     ss <- h3_bpf@saved.states
+#     ss[[length(ss)]]
+#   },
+#   timing = FALSE
+# )
+
+mod3_V0_res <- readRDS("model3/run_level_3/mod3_V0_res.rds")
+# mod3_V0_res <- bake(
+#   file = paste0("model3/", rl_dir, "mod3_V0_res.rds"), {
+#
+#     # Create a list to store results
+#     h3_V0_res <- list()
+#
+#     # Get "noVacc" scenario parameters
+#     h3_S0_par <- get_model3_vacc_scenario_params(scenario = "noVacc")
+#     coef(h3Spat) <- p3
+#     coef(h3Spat)[names(h3_S0_par)] <- h3_S0_par
+#
+#     # Using parameters and filtered distribution, project system
+#     h3_V0_sims <- project_from_filter(
+#       h3Spat, end_states = h3_bpf_end_states, covarGen = project_rain,
+#       nsims = NSIM, seed = 6897154
+#     )
+#
+#     # Store nationally aggregated results
+#     h3_V0_res$mod3_V0_sims <- agg_mod3_sims(h3_V0_sims)
+#     h3_V0_res$mod3_V0_probs <- get_elimProbs(h3_V0_sims, model = 3)
+#
+#     # For the supplement material, save quantiles of the Susceptible individuals
+#     h3_V0_res$SQuants <- h3_V0_sims %>%
+#       mutate(
+#         date = as.Date(lubridate::round_date(lubridate::date_decimal(time), unit = 'day'))
+#       ) %>%
+#       group_by(unitname, date) %>%
+#       mutate(
+#         pop = h3Spat@params[paste0("H", which(first(unitname) == unit_names(h3Spat)))]
+#       ) %>%
+#       summarize(
+#         Q025 = quantile((S + VSd + VSdd + VSdd_alt + VSd_alt) / pop, probs = 0.025),
+#         Q50 = quantile((S + VSd + VSdd + VSdd_alt + VSd_alt) / pop, probs = 0.5),
+#         Q975 = quantile((S + VSd + VSdd + VSdd_alt + VSd_alt) / pop, probs = 0.975)
+#       ) %>%
+#       ungroup() %>%
+#       rename(dep = unitname)
+#
+#     # Return results
+#     h3_V0_res
+#
+#   }
+# )
+
+mod3_V1_res <- readRDS("model3/run_level_3/mod3_V1_res.rds")
+# mod3_V1_res <- bake(
+#   file = paste0("model3/", rl_dir, "mod3_V1_res.rds"), {
+#
+#     # Create a list to store results
+#     h3_V1_res <- list()
+#
+#     # Get "noVacc" scenario parameters
+#     h3_V1_par <- get_model3_vacc_scenario_params(scenario = "2dep")
+#     coef(h3Spat) <- p3
+#     coef(h3Spat)[names(h3_V1_par)] <- h3_V1_par
+#
+#     # Using parameters and filtered distribution, project system
+#     h3_V1_sims <- project_from_filter(
+#       h3Spat, end_states = h3_bpf_end_states, covarGen = project_rain,
+#       nsims = NSIM, seed = 6897154
+#     )
+#
+#     # Store nationally aggregated results
+#     h3_V1_res$mod3_V1_sims <- agg_mod3_sims(h3_V1_sims)
+#     h3_V1_res$mod3_V1_probs <- get_elimProbs(h3_V1_sims, model = 3)
+#
+#     # Return results
+#     h3_V1_res
+#
+#   }
+# )
+
+mod3_V2_res <- readRDS("model3/run_level_3/mod3_V2_res.rds")
+# mod3_V2_res <- bake(
+#   file = paste0("model3/", rl_dir, "mod3_V2_res.rds"), {
+#
+#     # Create a list to store results
+#     h3_V2_res <- list()
+#
+#     # Get "noVacc" scenario parameters
+#     h3_V2_par <- get_model3_vacc_scenario_params(scenario = "3dep")
+#     coef(h3Spat) <- p3
+#     coef(h3Spat)[names(h3_V2_par)] <- h3_V2_par
+#
+#     # Using parameters and filtered distribution, project system
+#     h3_V2_sims <- project_from_filter(
+#       h3Spat, end_states = h3_bpf_end_states, covarGen = project_rain,
+#       nsims = NSIM, seed = 6897154
+#     )
+#
+#     # Store nationally aggregated results
+#     h3_V2_res$mod3_V2_sims <- agg_mod3_sims(h3_V2_sims)
+#     h3_V2_res$mod3_V2_probs <- get_elimProbs(h3_V2_sims, model = 3)
+#
+#     # Return results
+#     h3_V2_res
+#   }
+# )
+
+mod3_V3_res <- readRDS("model3/run_level_3/mod3_V3_res.rds")
+# mod3_V3_res <- bake(
+#   file = paste0("model3/", rl_dir, "mod3_V3_res.rds"), {
+#
+#     # Create a list to store results
+#     h3_V3_res <- list()
+#
+#     # Get "noVacc" scenario parameters
+#     h3_V3_par <- get_model3_vacc_scenario_params(scenario = "slowNation")
+#     coef(h3Spat) <- p3
+#     coef(h3Spat)[names(h3_V3_par)] <- h3_V3_par
+#
+#     # Using parameters and filtered distribution, project system
+#     h3_V3_sims <- project_from_filter(
+#       h3Spat, end_states = h3_bpf_end_states, covarGen = project_rain,
+#       nsims = NSIM, seed = 6897154
+#     )
+#
+#     # Store nationally aggregated results
+#     h3_V3_res$mod3_V3_sims <- agg_mod3_sims(h3_V3_sims)
+#     h3_V3_res$mod3_V3_probs <- get_elimProbs(h3_V3_sims, model = 3)
+#
+#     # Return results
+#     h3_V3_res
+#
+#   }
+# )
+
+mod3_V4_res <- readRDS("model3/run_level_3/mod3_V4_res.rds")
+# mod3_V4_res <- bake(
+#   file = paste0("model3/", rl_dir, "mod3_V4_res.rds"), {
+#
+#     # Create a list to store results
+#     h3_V4_res <- list()
+#
+#     # Get "noVacc" scenario parameters
+#     h3_V4_par <- get_model3_vacc_scenario_params(scenario = "fastNation")
+#     coef(h3Spat) <- p3
+#     coef(h3Spat)[names(h3_V4_par)] <- h3_V4_par
+#
+#     # Using parameters and filtered distribution, project system
+#     h3_V4_sims <- project_from_filter(
+#       h3Spat, end_states = h3_bpf_end_states, covarGen = project_rain,
+#       nsims = NSIM, seed = 6897154
+#     )
+#
+#     # Store nationally aggregated results
+#     h3_V4_res$mod3_V4_sims <- agg_mod3_sims(h3_V4_sims)
+#     h3_V4_res$mod3_V4_probs <- get_elimProbs(h3_V4_sims, model = 3)
+#
+#     # Return results
+#     h3_V4_res
+#
+#   }
+# )
 @

### 2024-01-16 H1profile.R
Ready to run Model 1 RL 3 profiles

@@ -18,19 +18,19 @@ library(tidyverse)
 library(data.table)
 library(haitipkg)
 
-RUN_LEVEL <- 2
+RUN_LEVEL <- 3
 
-nprof      <- switch(RUN_LEVEL,  2,    14, 20)
+nprof      <- switch(RUN_LEVEL,  2,    14, 25)
 NMIF       <- switch(RUN_LEVEL,  5,   100, 200)
-NP         <- switch(RUN_LEVEL,  50, 1000, 1000)
-NP_EVAL    <- switch(RUN_LEVEL, 100, 2500, 10000)
-NREPS_EVAL <- switch(RUN_LEVEL,   3,   10, 10)
+NP         <- switch(RUN_LEVEL,  50, 1000, 1500)
+NP_EVAL    <- switch(RUN_LEVEL, 100, 2500, 5000)
+NREPS_EVAL <- switch(RUN_LEVEL,   3,   10, 15)
 COOLING    <- 0.5
 
 # Create Experiment Registry ----------------------------------------------
 
 reg <- makeExperimentRegistry(
-  file.dir = paste0('model1/profileReg_RL', RUN_LEVEL, '_v3'),
+  file.dir = paste0('model1/profileReg_RL', RUN_LEVEL, '_v1'),
   seed = 739164,
   packages = c("spatPomp", 'haitipkg', 'pomp')
 )
@@ -39,75 +39,71 @@ reg <- makeExperimentRegistry(
 
 set.seed(665544)
 
-# Read in previous best results
-haiti1_fit <- readRDS("model1/run_level_3/haiti1_fit.rds")
-best_pars <- haiti1_fit %>%
-  filter(ll == max(ll)) %>%
-  select(-ll, -ll.se) %>%
-  unlist()
-
 h1 <- haiti1_joint()
-coef(h1) <- best_pars
 
 prof_params <- c(
-  # 'betat',
-  # 'tau_epi',
-  # 'tau_end',
-  # 'rho',
-  # 'nu',
-  # 'sig_sq_epi',
-  # 'sig_sq_end',
+  'betat',
+  'tau_epi',
+  'tau_end',
+  'rho',
+  'nu',
+  'sig_sq_epi',
+  'sig_sq_end',
   'E_0',
-  'I_0'
+  'I_0',
+  'beta1',
+  'beta2',
+  'beta3',
+  'beta4',
+  'beta5',
+  'beta6'
 )
 
-final_pars <- best_pars
[…]

### 2024-01-16 r1/response.tex

@@
 
+That statement arose in the author summary, where references are not expected.
+However, we appreciate that the point should be expanded on, and we have done so in the revision \TODO{}.
+We have cited \cite{he10,stocks20,li23} for examples demonstrating situations where overdispersed dynamic noise is present, and \cite{king15} for an investigation of the statistical limitations of fitting deterministic models to data.
+
 \report{2.
@@
 
-Thanks for this suggestion.
-\TODO{We have implemented this in the revision.}
+We have added some material in the discussion connecting our results to the stated goals of the paper from the abstract and introduction.
+We could have gone further and discussed in more detail the relationship to the references [4, 5], following this recommendation.
+However, a subsequent manuscript addresses [4,5] more directly \citep{li23} and so we have referred the reader to that paper (currently available on arXiv) for additional discussion of this point.
 
-THIS IS DISCUSSED MORE EXTENSIVELY IN JIFAN'S COVID PAPER. THERE IS SOME DANGER THAT THESE PAPERS COULD CONVERGE TOO MUCH. ANOTHER OPTION IS TO REFER TO THAT PAPER. SO, AS ANOTHER OPTION:
+\eic{IF THE P0INTS OF 4, 5 CAN BE CONNECTED DIRECTLY TO THIS DATA ANALYSIS, WE OUGHT TO BE ABLE TO DISCUSS THAT WITHOUT CONFLICTING WITH LI23.
+WE SHOULD PROBABLY RE-READ THESE REFERENCES AND FIGURE IT OUT.}
 
-\TODO{A subsequent manuscript addresses [4,5] more directly \citep{li23} and so we have referred the reader to that paper (currently available on arXiv) for additional discussion of this point.}
+\eic{ANOTHER IDEA: WE CAN VIEW OUR PAPER AND LI23 AS SYNERGISTIC RATHER THAN IN COMPETITION.
+WE COULD EVEN TALK ABOUT HOW THAT DIFFERENT SETTING (COVID, WITH AN EMPHASIS ON SPATIOTEMPORAL MODELING AND COMPARING PRIMARILY WITH ENSEMBLE FILTER METHODS) LEADS TO SIMILAR CONCLUSIONS TO HAITI CHOLERA.
+FROM THAT PERSPECTIVE, THE PAPERS STRENGTHEN EACH OTHER.
+AS LONG AS BOTH LISTS OF RECOMMENDATIONS ARE FIRMLY TIED TO THE RESPECTIVE DATA ANALYSES, IT SHOULD BE A STRENGTH NOT A WEAKNESS IF THEY OVERLAP.
+BUT IF WE GO THAT ROUTE WE'LL HAVE TO EXPLAIN THIS CONVINCINGLY.}
 

### 2024-01-17 H1profile.R
Adjusted RL3 for some params, will re-run

@@ -30,7 +30,7 @@ COOLING    <- 0.5
 # Create Experiment Registry ----------------------------------------------
 
 reg <- makeExperimentRegistry(
-  file.dir = paste0('model1/profileReg_RL', RUN_LEVEL, '_v1'),
+  file.dir = paste0('model1/profileReg_RL', RUN_LEVEL, '_v2'),
   seed = 739164,
   packages = c("spatPomp", 'haitipkg', 'pomp')
 )
@@ -42,21 +42,21 @@ set.seed(665544)
 h1 <- haiti1_joint()
 
 prof_params <- c(
-  'betat',
+  # 'betat',
   'tau_epi',
   'tau_end',
   'rho',
-  'nu',
+  # 'nu',
   'sig_sq_epi',
-  'sig_sq_end',
+  # 'sig_sq_end',
   'E_0',
-  'I_0',
-  'beta1',
-  'beta2',
-  'beta3',
-  'beta4',
-  'beta5',
-  'beta6'
+  'I_0'
+  # 'beta1',
+  # 'beta2',
+  # 'beta3',
+  # 'beta4',
+  # 'beta5',
+  # 'beta6'
 )
 
 final_pars <- coef(h1)
@@ -66,21 +66,21 @@ for (pp in prof_params) {
   if (pp == "betat") {
     prof_values <- seq(-0.15, 0.05, length.out = 30)
   } else if (pp == 'tau_epi') {
-    prof_values <- seq(150, 2000, length.out = 30)
+    prof_values <- seq(5, 600, length.out = 30)
   } else if (pp == 'tau_end') {
-    prof_values <- seq(20, 1800, length.out = 30)
+    prof_values <- seq(1, 500, length.out = 30)
   } else if (pp == 'rho') {
-    prof_values <- seq(0.18, 1, length.out = 30)
+    prof_values <- seq(0.15, 0.75, length.out = 30)
   } else if (pp == 'nu') {
-    prof_values <- seq(0.85, 1, 0.01)
+    prof_values <- seq(0.9, 1, 0.01)
   } else if (pp == 'sig_sq_epi') {
-    prof_values <- seq(0.08, 0.14, length.out = 30)
+    prof_values <- seq(0.025, 0.12, length.out = 30)
   } else if (pp == 'sig_sq_end') {
-    prof_values <- seq(0.05, 0.25, length.out = 30)
+    prof_values <- seq(0.06, 0.225, length.out = 30)
   } else if (pp == 'E_0') {
[…]

### 2024-01-17 H3profile.R
Getting ready to run RL3

@@ -18,7 +18,7 @@ library(tidyverse)
 # library(data.table)
 library(haitipkg)
 
-RUN_LEVEL <- 2
+RUN_LEVEL <- 3
 
 nprof <- switch(RUN_LEVEL, 2, 14, 20)
 NBPF <- switch(RUN_LEVEL, 5, 50, 100)
@@ -26,7 +26,7 @@ NP <- switch(RUN_LEVEL, 50, 750, 1000)
 NP_EVAL <- switch(RUN_LEVEL, 100, 1000, 2000)
 NREPS_EVAL <- switch(RUN_LEVEL, 3, 6, 10)
 SPAT_REGRESSION <- 0.05
-COOLING <- 0.3
+COOLING <- 0.35
 
 # Create Experiment Registry ----------------------------------------------
 
@@ -41,9 +41,8 @@ reg <- makeExperimentRegistry(
 set.seed(665544)
 
 # Read in previous best results
-haiti3_fit <- readRDS("model3/run_level_3/haiti3_fit.rds")
-best_pars <- haiti3_fit$search2$params[which.max(haiti3_fit$search2$logLiks$logLik), ]
 h3_spat <- haiti3_spatPomp()
+final_pars <- coef(h3_spat)
 
 # Create vectors for the unit and shared parameters
 unit_specific_names <- c("betaB", "foi_add", "aHur", "hHur")
@@ -52,69 +51,54 @@ shared_param_names <- c(
   "epsilon", "k"
 )
 
-est_param_names <- c(
-  unit_specific_names, shared_param_names
-)
+prof_params <- shared_param_names
+# If you wanted to profile over the six added unit-specific
+# parameters
+# prof_params <- c(
+#   shared_param_names, 'aHur3', 'hHur3',
+#   'aHur9', 'hHur9', 'Iinit3', 'Iinit4'
+# )
 
-# prof_params <- shared_param_names
-prof_params <- c(
-  'aHur3', 'hHur3', 'aHur9', 'hHur9',
-  'Iinit3', "Iinit4", 'betaB1'
-)
-
-final_pars <- best_pars
+best_pars <- final_pars
 prof_vars <- c()
 for (pp in prof_params) {
 
   if (pp == "mu_B") {
-    prof_values <- seq(200, 650, length.out = 25)
+    prof_values <- seq(300, 750, length.out = 30)
   } else if (pp == 'XthetaA') {
[…]

### 2024-01-18 H2profile.R
First attempt at H2 profile

@@ -1,0 +1,217 @@
+# Fitting profiles for Haiti 2.
+#
+# The fit_haiti2 function is a very simple helper that will take as input
+# some starting parameters and an objective function for Model 2 and will
+# return parameters estimates along with corresponding estimates of their
+# log-likelihood.
+#
+# This this script we use this helper function to do a profile likelihood search
+# of the estimated parameters using the batchtools package. To do this, we will
+# create a grid of parameter values and submit jobs where each job will use a
+# row from the grid as a starting point to fit the model.
+
+
+# Load Packages -----------------------------------------------------------
+
+library(batchtools)
+library(tidyverse)
+library(data.table)
+library(haitipkg)
+library(pomp)
+library(spatPomp)
+library(subplex)
+
+
+# Create Experiment Registry ----------------------------------------------
+
+reg <- makeExperimentRegistry(
+  file.dir = paste0('model2/profileReg_RL', RUN_LEVEL, '_v1'),
+  seed = 739164,
+  packages = c("spatPomp", 'haitipkg', 'pomp')
+)
+
+n_starts <- 5
+
+# Create Profile Design Matrix --------------------------------------------
+
+set.seed(22446688)
+h2 <- haiti2(cutoff = 10000, measure = "log")
+
+prof_params <- c(
+  "Mu",
+  "Beta",
+  "BetaW",
+  "v",
+  "sigma",
+  "phase"
+)
+
+final_pars <- coef(h2)
+prof_vars <- c()
+for (pp in prof_params) {
+
+  if (pp == 'Mu') {
+    prof_values <- seq(9000, 9500, length.out = 30)
+  } else if (pp == "Beta") {
+    prof_values <- seq(2e-16, 5e-16, length.out = 30)
+  } else if (pp == "BetaW") {
+    prof_values <- seq(0.5, 1.5, length.out = 20)
+  } else if (pp == "v") {
+    prof_values <- seq(0.75, 2, length.out = 20)
+  } else if (pp == "sigma") {
+    prof_values <- seq(2e-13, 4e-13, length.out = 20)
+  } else if (pp == "phase") {
+    prof_values <- seq(0, 12, length.out = 30)
+  }
+
+  prof_cols <- matrix(rep(prof_values, each = n_starts), ncol = 1)
+  colnames(prof_cols) <- pp
+
+  bounds <- tibble::tribble(
[…]

### 2024-01-18 H2profile.R
Added RUN_LEVEL param

@@ -24,6 +24,8 @@ library(subplex)
 
 # Create Experiment Registry ----------------------------------------------
 
+RUN_LEVEL = 2
+
 reg <- makeExperimentRegistry(
   file.dir = paste0('model2/profileReg_RL', RUN_LEVEL, '_v1'),
   seed = 739164,

### 2024-01-18 H1profile.R
Minor adjustments to RL3 profile search bounds

@@ -30,7 +30,7 @@ COOLING    <- 0.5
 # Create Experiment Registry ----------------------------------------------
 
 reg <- makeExperimentRegistry(
-  file.dir = paste0('model1/profileReg_RL', RUN_LEVEL, '_v2'),
+  file.dir = paste0('model1/profileReg_RL', RUN_LEVEL, '_v3'),
   seed = 739164,
   packages = c("spatPomp", 'haitipkg', 'pomp')
 )
@@ -66,21 +66,21 @@ for (pp in prof_params) {
   if (pp == "betat") {
     prof_values <- seq(-0.15, 0.05, length.out = 30)
   } else if (pp == 'tau_epi') {
-    prof_values <- seq(5, 600, length.out = 30)
+    prof_values <- seq(1, 800, length.out = 30)
   } else if (pp == 'tau_end') {
-    prof_values <- seq(1, 500, length.out = 30)
+    prof_values <- seq(1, 350, length.out = 30)
   } else if (pp == 'rho') {
-    prof_values <- seq(0.15, 0.75, length.out = 30)
+    prof_values <- seq(0.1, 0.9, length.out = 30)
   } else if (pp == 'nu') {
     prof_values <- seq(0.9, 1, 0.01)
   } else if (pp == 'sig_sq_epi') {
-    prof_values <- seq(0.025, 0.12, length.out = 30)
+    prof_values <- seq(0.035, 0.2, length.out = 30)
   } else if (pp == 'sig_sq_end') {
-    prof_values <- seq(0.06, 0.225, length.out = 30)
+    prof_values <- seq(0.02, 0.3, length.out = 30)
   } else if (pp == 'E_0') {
-    prof_values <- seq(1 / 10911819, 16000 / 10911819, length.out = 50)
+    prof_values <- seq(1 / 10911819, 25000 / 10911819, length.out = 50)
   } else if (pp == 'I_0') {
-    prof_values <- seq(1 / 10911819, 12000 / 10911819, length.out = 50)
+    prof_values <- seq(1 / 10911819, 25000 / 10911819, length.out = 50)
   } else if (pp == 'beta1') {
     prof_values <- seq(0.5, 2.5, length.out = 20)
   } else if (pp == 'beta2') {

### 2024-01-18 r1/response.tex
more responses

@@
 \usepackage{fullpage}
-\bibliographystyle{apalike}
 
-\usepackage{natbib,amsmath,amsthm}
+\usepackage{amsmath,amsthm}
+
+% \usepackage{natbib}
+%\bibliographystyle{apalike}
+
+% Use the PLoS provided BiBTeX style
+\bibliographystyle{../plos2015}
+\usepackage{cite}
+
+% Remove brackets from numbering in List of References
+\makeatletter
+\renewcommand{\@biblabel}[1]{\quad#1.}
+\makeatother
+
 \usepackage{verbatim}
@@
 \usepackage{amsfonts}
-\usepackage{natbib}
  \usepackage{url}
@@
 
+\newcommand\code[1]{\texttt{#1}}
 \input{../edits.tex}
@@
 Model 3 has a stochastic spatiotemporal partially-observed structure that leads to methodological challenges.
-\citet{lee20} made some approximations to avoid carrying out inference for the full coupled system.
+\cite{lee20} made some approximations to avoid carrying out inference for the full coupled system.
 We show some consequences of those approximations, taking advantage of a newly developed method which has not yet been applied in the context of scientific investigations.
@@
 We could have gone further and discussed in more detail the relationship to the references [4, 5], following this recommendation.
-However, a subsequent manuscript addresses [4,5] more directly \citep{li23} and so we have referred the reader to that paper (currently available on arXiv) for additional discussion of this point.
+However, a subsequent manuscript addresses [4,5] more directly \cite{li23} and so we have referred the reader to that paper (currently available on arXiv) for additional discussion of this point.
 
@@
 
+The referee's suggestions have led to the addition of various references in the revision.
+For the particular issue of black box versus white box models, we have included the following paragraph
+\TODO{where?}
+
+\article{\editMechModels}
+
 \report{
@@
 Users (like me) sometimes face uncertainty about whether the lack of fit is due to model misspecification or problems with the Monte Carlo algorithm exploring the parameter space.
+}
 
+We agree with the referee that diagnostic are an important aspect of careful data analysis.
+The most widely used diagnostic tools for particle filter and iterated particle filter methods probably match those provided by default in the \code{pomp} R package \cite{king16} when objects generated by \code{pfilter()} or \code{mif2()} are plotted.
+For filtering, one tracks the effective sample size, the conditional log likelihood for each observation, and the values of the filtered states.
+For iterated filtering, one checks the filtering diagnostics for the last iteration and also how the parameter estimates and the log-likelihood estimate evolve through iterations, typically superimposing the results of 10-100 independent searches.
+
+The particle filter provides access to the likelihood of the data, and conditional likelhood of each individual data point.
+By contrast, obtaining reliable likelihood values from MCMC is not straightforward.
+Thus, particle filter methods facilitate the use of various likelihood-based methods for diagnostics and model selection.
+We demonstrate this in Figs S4, S5, S6, S7.
+
+The particle filter is unbiased for the likelihood, and therefore negatively biased for the log-likelihood with bias approaching zero as the number of particles increases and the Monte Carlo variance decreases.
+%%Log-likelihood is a proper scoring rule for probabilistic forecasts \cite{gneiting07}, so any error in filtering is expected to lead to a loss in log-likelihood.
+Thus, log-likelihood is an appropriate way to evaluate filters as well as models.
+If optimization searches from diverse starting pionts provide a consensus on the maximum log-likelihood, and the Monte Carlo variance is small, we can have reasonable confidence in our maximization.
+Monte Carlo adusted profile (MCAP) methods \cite{ionides17,ning21} provide methodology which reduces and evaluates the consequences of this Monte Carlo error.
+MCAP methods have been used in various recent investigations using particle filters \eic{REFS}.
+
+We conclude that suitable diagnostic methods are available for the POMP methods we employ, and some of them are demonstrated and discussed in our paper.
+We would like to do what we can, within the constraints of this paper, to help build awareness of these diagnostic methods and ways in which they can be useful.
+A full exposition of this topic is out of the scope of this paper.
+Howerver, we can alert the reader to the topic, and provide some additional references.
+Not infrequently, diagnostic issues are relegated to supplementary material and so we have indicated to interested readers some papers with extensive supplements demonstrating these diagnostic issues in action.
+\TODO{}
+
+\report{
 For example, in Figure 5, department Ouest exhibits substantial uncertainty from 2014 onwards, and this figure is on a log scale.
@@
 }
+
+We agree with the referee that uncertainty quantification and evaluation of identifiability are fundamental tasks for model-based data analysis.
+
+Wide prediction intervals are not necessarily a sign of a problem.
+For example, nobody knew for sure whether cholera was banished from Haiti after it was declared eliminated in February 2022.
+Cholera reemerged in September 2022, but a reasonable model might be expected to cover the possiblity of this reemergence either happening or not happening, leading to wide prediction intervals.
+
+In the revision, we have included profile confidence intervals on all parameters, except those taking separate values on each unit.
+For these unit-specific parameters, the spread of values across units provides an alternative measure of uncertainty.
+We find that all parameters are identifiable, though some are more tightly estimable than others.
+
+We had not done this in the original submission since it was only tangentially relevant to the comparison with \cite{lee20} that lies at the center of our paper.
+However, these additional results are worthwhile extension to our analysis and we are grateful for the encouragement to include it.
 

### 2024-01-18 edits.tex
Fixed typos mentioned by reviewers
response to R2 about mcmc, etc
Changed xcolor .tex options, recompiled
more responses

@@
-\usepackage[dvipsnames]{xcolor}
-\definecolor{mygreen}{RGB}{0, 80, 0}
+\usepackage[table]{xcolor}
+\definecolor{mygreen}{RGB}{0, 90, 0}
 \newcommand\original[1]{\textcolor{red}{#1}}
 \newcommand\TODO[1]{\textcolor{red}{[TODO: #1]}}
+
+%%%%%% EDITING MACROS %%%%%%%%%
+% orange for EI
+\definecolor{orange}{rgb}{1,0.5,0}
+\newcommand\ei[2]{\sout{#1} \textcolor{orange}{#2}}
+\newcommand\eic[1]{\textcolor{orange}{[#1]}}
+% green for JW
+\definecolor{green}{rgb}{0,0.5,0}
+\newcommand\jw[2]{\sout{#1} \textcolor{green}{#2}}
+\newcommand\jwc[1]{\textcolor{green}{[#1]}}
+% purple for JJ
+\definecolor{purple}{rgb}{0.5,0,1}
+\newcommand\jj[2]{\sout{#1} \textcolor{purple}{#2}}
+\newcommand\jjc[1]{\textcolor{purple}{[#1]}}
+% cyan for AR
+\definecolor{cyan}{rgb}{0,.5,.5}
+\newcommand\ar[2]{\sout{#1} \textcolor{cyan}{#2}}
+\newcommand\arc[1]{\textcolor{cyan}{#1}}
+% light brown for KT
+\definecolor{lightbrown}{rgb}{0.5,0.5,0}
+\newcommand\kt[2]{\sout{#1} \textcolor{lightbrown}{#2}}
+\newcommand\ktc[1]{\textcolor{lightbrown}{#1}}
+
+\newcommand\editMechModels{Mathematical models representing biological phenomena are valuable for epidemiology and consequently for public health policy \cite{lofgren14,mccabe21}.
+More broadly, they have useful roles throughout biology, expecially when combined with statistical methods that properly account for stochasticity and nonlinearity \cite{may04}.
+In some situations, modern machine learning methods can outperform mechanistic models on epidemiological forecasting tasks \cite{lau22}.
+The predictive skill of non-mechansitic models can reveal limitations in mechanistic models, but cannot readily replace the scientific understanding obtained by describing the biological dynamics of the system in a mathematical model.}

### 2024-01-18 H2profile.R
Modified search bounds for Beta and sigma

@@ -27,12 +27,12 @@ library(subplex)
 RUN_LEVEL = 2
 
 reg <- makeExperimentRegistry(
-  file.dir = paste0('model2/profileReg_RL', RUN_LEVEL, '_v2'),
+  file.dir = paste0('model2/profileReg_RL', RUN_LEVEL, '_v3'),
   seed = 739164,
   packages = c("spatPomp", 'haitipkg', 'pomp')
 )
 
-n_starts <- 5
+n_starts <- 3
 
 # Create Profile Design Matrix --------------------------------------------
 
@@ -40,12 +40,12 @@ set.seed(22446688)
 h2 <- haiti2(cutoff = 10000, measure = "log")
 
 prof_params <- c(
-  "Mu",
+  # "Mu",
   "Beta",
-  "BetaW",
-  "v",
-  "sigma",
-  "phase"
+  # "BetaW",
+  # "v",
+  "sigma"
+  # "phase"
 )
 
 final_pars <- coef(h2)
@@ -53,17 +53,17 @@ prof_vars <- c()
 for (pp in prof_params) {
 
   if (pp == 'Mu') {
-    prof_values <- seq(7000, 12000, length.out = 50)
+    prof_values <- seq(6500, 12500, length.out = 30)
   } else if (pp == "Beta") {
-    prof_values <- seq(2e-17, 5e-15, length.out = 50)
+    prof_values <- seq(.Machine$double.eps, 5e-05, length.out = 50)
   } else if (pp == "BetaW") {
     prof_values <- seq(0.9, 1.25, length.out = 20)
   } else if (pp == "v") {
-    prof_values <- seq(1.1, 1.5, length.out = 20)
+    prof_values <- seq(1.2, 1.5, length.out = 20)
   } else if (pp == "sigma") {
-    prof_values <- seq(2e-14, 4e-12, length.out = 20)
+    prof_values <- seq(2e-14, 1, length.out = 50)
   } else if (pp == "phase") {
-    prof_values <- seq(6, 9.1, length.out = 30)
+    prof_values <- seq(6, 8.5, length.out = 30)
   }
 
   prof_cols <- matrix(rep(prof_values, each = n_starts), ncol = 1)

### 2024-01-18 H2profile.R
Closer search for Beta and sigma

@@ -32,7 +32,7 @@ reg <- makeExperimentRegistry(
   packages = c("spatPomp", 'haitipkg', 'pomp')
 )
 
-n_starts <- 3
+n_starts <- 1
 
 # Create Profile Design Matrix --------------------------------------------
 
@@ -41,7 +41,7 @@ h2 <- haiti2(cutoff = 10000, measure = "log")
 
 prof_params <- c(
   # "Mu",
-  "Beta",
+  # "Beta",
   # "BetaW",
   # "v",
   "sigma"
@@ -55,13 +55,13 @@ for (pp in prof_params) {
   if (pp == 'Mu') {
     prof_values <- seq(6500, 12500, length.out = 30)
   } else if (pp == "Beta") {
-    prof_values <- seq(.Machine$double.eps, 5e-05, length.out = 50)
+    prof_values <- seq(.Machine$double.eps, 5e-06, length.out = 100)
   } else if (pp == "BetaW") {
     prof_values <- seq(0.9, 1.25, length.out = 20)
   } else if (pp == "v") {
     prof_values <- seq(1.2, 1.5, length.out = 20)
   } else if (pp == "sigma") {
-    prof_values <- seq(2e-14, 1, length.out = 50)
+    prof_values <- seq(2e-14, 0.0005, length.out = 100)
   } else if (pp == "phase") {
     prof_values <- seq(6, 8.5, length.out = 30)
   }

### 2024-01-18 H2profile.R
changed registry name

@@ -27,7 +27,7 @@ library(subplex)
 RUN_LEVEL = 2
 
 reg <- makeExperimentRegistry(
-  file.dir = paste0('model2/profileReg_RL', RUN_LEVEL, '_v3'),
+  file.dir = paste0('model2/profileReg_RL', RUN_LEVEL, '_v4'),
   seed = 739164,
   packages = c("spatPomp", 'haitipkg', 'pomp')
 )

### 2024-01-18 H2profile.R
One more adjustment to Beta, Sigma, running all profs again

@@ -27,7 +27,7 @@ library(subplex)
 RUN_LEVEL = 2
 
 reg <- makeExperimentRegistry(
-  file.dir = paste0('model2/profileReg_RL', RUN_LEVEL, '_v4'),
+  file.dir = paste0('model2/profileReg_RL', RUN_LEVEL, '_v6'),
   seed = 739164,
   packages = c("spatPomp", 'haitipkg', 'pomp')
 )
@@ -40,12 +40,12 @@ set.seed(22446688)
 h2 <- haiti2(cutoff = 10000, measure = "log")
 
 prof_params <- c(
-  # "Mu",
-  # "Beta",
-  # "BetaW",
-  # "v",
-  "sigma"
-  # "phase"
+  "Mu",
+  "Beta",
+  "BetaW",
+  "v",
+  "sigma",
+  "phase"
 )
 
 final_pars <- coef(h2)
@@ -55,13 +55,13 @@ for (pp in prof_params) {
   if (pp == 'Mu') {
     prof_values <- seq(6500, 12500, length.out = 30)
   } else if (pp == "Beta") {
-    prof_values <- seq(.Machine$double.eps, 5e-06, length.out = 100)
+    prof_values <- seq((.Machine$double.eps)^(1/5), (5e-05)^(1/5), length.out = 100)^5
   } else if (pp == "BetaW") {
     prof_values <- seq(0.9, 1.25, length.out = 20)
   } else if (pp == "v") {
     prof_values <- seq(1.2, 1.5, length.out = 20)
   } else if (pp == "sigma") {
-    prof_values <- seq(2e-14, 0.0005, length.out = 100)
+    prof_values <- seq((.Machine$double.eps)^(1/5), (1e-03)^(1/5), length.out = 100)^5
   } else if (pp == "phase") {
     prof_values <- seq(6, 8.5, length.out = 30)
   }

### 2024-01-19 H3profile.R
Redoing ThetaI and LambdaR, better bounds

@@ -31,7 +31,7 @@ COOLING <- 0.35
 # Create Experiment Registry ----------------------------------------------
 
 reg <- makeExperimentRegistry(
-  file.dir = paste0('model3/profileReg_RL', RUN_LEVEL, '_v7'),
+  file.dir = paste0('model3/profileReg_RL', RUN_LEVEL, '_thetaI_lambdaR'),
   seed = 739164,
   packages = c("spatPomp", 'haitipkg', 'pomp')
 )
@@ -51,7 +51,12 @@ shared_param_names <- c(
   "epsilon", "k"
 )
 
-prof_params <- shared_param_names
+prof_params <- c(
+  'thetaI',
+  'lambdaR'
+)
+
+# prof_params <- shared_param_names
 # If you wanted to profile over the six added unit-specific
 # parameters
 # prof_params <- c(
@@ -68,9 +73,9 @@ for (pp in prof_params) {
   } else if (pp == 'XthetaA') {
     prof_values <- seq(0.0001, 0.15, length.out = 30)
   } else if (pp == 'thetaI') {
-    prof_values <- seq(2e-05, 0.00015, length.out = 30)
+    prof_values <- seq(7.5e-06, 0.0001, length.out = 30)
   } else if (pp == 'lambdaR') {
-    prof_values <- seq(0.5, 3.6, length.out = 30)
+    prof_values <- seq(0.15, 1.75, length.out = 30)
   } else if (pp == 'r') {
     prof_values <- seq(0.05, 1.3, length.out = 30)
   } else if (pp == 'std_W') {
@@ -323,7 +328,7 @@ addExperiments(prob.designs = pdes, algo.designs = ades)
 # Submit Jobs -------------------------------------------------------------
 
 # resources1 <- list(account = 'stats_dept1', walltime = '10:00', memory = '5000m', ncpus = 1)
-resources1 <- list(account = 'ionides2', walltime = '6:00:00', memory = '5000m', ncpus = 1)
+resources1 <- list(account = 'ionides2', walltime = '8:00:00', memory = '9000m', ncpus = 1)
 
 submitJobs(resources = resources1)
 # submitJobs(ids = 1, resources = resources1)

### 2024-01-20 H1profile.R
Trying large values of tau

@@ -18,7 +18,7 @@ library(tidyverse)
 library(data.table)
 library(haitipkg)
 
-RUN_LEVEL <- 3
+RUN_LEVEL <- 2
 
 nprof      <- switch(RUN_LEVEL,  2,    14, 25)
 NMIF       <- switch(RUN_LEVEL,  5,   100, 200)
@@ -30,7 +30,7 @@ COOLING    <- 0.5
 # Create Experiment Registry ----------------------------------------------
 
 reg <- makeExperimentRegistry(
-  file.dir = paste0('model1/profileReg_RL', RUN_LEVEL, '_v3'),
+  file.dir = paste0('model1/profileReg_RL', RUN_LEVEL, '_tau'),
   seed = 739164,
   packages = c("spatPomp", 'haitipkg', 'pomp')
 )
@@ -44,13 +44,13 @@ h1 <- haiti1_joint()
 prof_params <- c(
   # 'betat',
   'tau_epi',
-  'tau_end',
-  'rho',
+  'tau_end'
+  # 'rho',
   # 'nu',
-  'sig_sq_epi',
+  # 'sig_sq_epi',
   # 'sig_sq_end',
-  'E_0',
-  'I_0'
+  # 'E_0',
+  # 'I_0'
   # 'beta1',
   # 'beta2',
   # 'beta3',
@@ -66,21 +66,21 @@ for (pp in prof_params) {
   if (pp == "betat") {
     prof_values <- seq(-0.15, 0.05, length.out = 30)
   } else if (pp == 'tau_epi') {
-    prof_values <- seq(1, 800, length.out = 30)
+    prof_values <- c(exp(seq(log(5), log(5000), length.out = 40)), 5e+12)
   } else if (pp == 'tau_end') {
-    prof_values <- seq(1, 350, length.out = 30)
+    prof_values <- c(exp(seq(log(5), log(5000), length.out = 40)), 5e+12)
   } else if (pp == 'rho') {
     prof_values <- seq(0.1, 0.9, length.out = 30)
   } else if (pp == 'nu') {
     prof_values <- seq(0.9, 1, 0.01)
   } else if (pp == 'sig_sq_epi') {
-    prof_values <- seq(0.035, 0.2, length.out = 30)
+    prof_values <- seq(0.04, 0.2, length.out = 30)
   } else if (pp == 'sig_sq_end') {
-    prof_values <- seq(0.02, 0.3, length.out = 30)
[…]

### 2024-01-20 H1profile.R
Larger values of tau

@@ -20,10 +20,10 @@ library(haitipkg)
 
 RUN_LEVEL <- 2
 
-nprof      <- switch(RUN_LEVEL,  2,    14, 25)
-NMIF       <- switch(RUN_LEVEL,  5,   100, 200)
+nprof      <- switch(RUN_LEVEL,  2,    15, 25)
+NMIF       <- switch(RUN_LEVEL,  5,   150, 200)
 NP         <- switch(RUN_LEVEL,  50, 1000, 1500)
-NP_EVAL    <- switch(RUN_LEVEL, 100, 2500, 5000)
+NP_EVAL    <- switch(RUN_LEVEL, 100, 3000, 5000)
 NREPS_EVAL <- switch(RUN_LEVEL,   3,   10, 15)
 COOLING    <- 0.5
 
@@ -235,7 +235,7 @@ addExperiments(prob.designs = pdes, algo.designs = ades)
 
 # resources1 <- list(account = 'stats_dept1', walltime = '10:00', memory = '1000m', ncpus = 1)
 resources1 <- list(
-  account = 'stats_dept1', walltime = '30:00',
+  account = 'stats_dept1', walltime = '45:00',
   memory = '5000m', ncpus = 1
   )
 

### 2024-01-22 H1profile.R
Created final bounds for RL3 Model 1 submission

@@ -66,13 +66,13 @@ for (pp in prof_params) {
   if (pp == "betat") {
     prof_values <- seq(-0.15, 0.05, length.out = 30)
   } else if (pp == 'tau_epi') {
-    prof_values <- c(exp(seq(log(5), log(5000), length.out = 40)), 5e+12)
+    prof_values <- c(exp(seq(2.75, 9, length.out = 40)), 5e+12)
   } else if (pp == 'tau_end') {
-    prof_values <- c(exp(seq(log(5), log(5000), length.out = 40)), 5e+12)
+    prof_values <- c(exp(seq(2, 9, length.out = 40)), 5e+12)
   } else if (pp == 'rho') {
     prof_values <- seq(0.1, 0.9, length.out = 30)
   } else if (pp == 'nu') {
-    prof_values <- seq(0.9, 1, 0.01)
+    prof_values <- seq(0.9, 1, length.out = 15)
   } else if (pp == 'sig_sq_epi') {
     prof_values <- seq(0.04, 0.2, length.out = 30)
   } else if (pp == 'sig_sq_end') {

### 2024-01-22 H2profile.R
Created final set of profile bounds for Model 2

@@ -24,10 +24,8 @@ library(subplex)
 
 # Create Experiment Registry ----------------------------------------------
 
-RUN_LEVEL = 2
-
 reg <- makeExperimentRegistry(
-  file.dir = paste0('model2/profileReg_RL', RUN_LEVEL, '_v6'),
+  file.dir ='model2/profileReg_RL',
   seed = 739164,
   packages = c("spatPomp", 'haitipkg', 'pomp')
 )
@@ -53,17 +51,17 @@ prof_vars <- c()
 for (pp in prof_params) {
 
   if (pp == 'Mu') {
-    prof_values <- seq(6500, 12500, length.out = 30)
+    prof_values <- seq(6750, 13000, length.out = 30)
   } else if (pp == "Beta") {
-    prof_values <- seq((.Machine$double.eps)^(1/5), (5e-05)^(1/5), length.out = 100)^5
+    prof_values <- seq((.Machine$double.eps)^(1/5), (6.5e-05)^(1/5), length.out = 50)^5
   } else if (pp == "BetaW") {
-    prof_values <- seq(0.9, 1.25, length.out = 20)
+    prof_values <- seq(1, 1.2, length.out = 30)
   } else if (pp == "v") {
-    prof_values <- seq(1.2, 1.5, length.out = 20)
+    prof_values <- seq(1.225, 1.44, length.out = 25)
   } else if (pp == "sigma") {
-    prof_values <- seq((.Machine$double.eps)^(1/5), (1e-03)^(1/5), length.out = 100)^5
+    prof_values <- seq((.Machine$double.eps)^(1/5), (2e-03)^(1/5), length.out = 50)^5
   } else if (pp == "phase") {
-    prof_values <- seq(6, 8.5, length.out = 30)
+    prof_values <- seq(6.5, 7.9, length.out = 30)
   }
 
   prof_cols <- matrix(rep(prof_values, each = n_starts), ncol = 1)
@@ -73,10 +71,10 @@ for (pp in prof_params) {
     ~param, ~lower, ~upper,
     "Mu",     8000,  10000,
     "Beta",  2e-17,  5e-15,
-    "BetaW",  0.75,    1.3,
-    "v",      0.75,      2,
+    "BetaW",  0.85,   1.25,
+    "v",      1.2,     1.5,
     "sigma", 2e-13,  4e-13,
-    "phase",  4.36,  10.64
+    "phase",     6,    8.5
   )
 
   bounds <- bounds %>%

### 2024-01-22 H3profile.R
Final modifications for Model 3 profile search

@@ -31,7 +31,7 @@ COOLING <- 0.35
 # Create Experiment Registry ----------------------------------------------
 
 reg <- makeExperimentRegistry(
-  file.dir = paste0('model3/profileReg_RL', RUN_LEVEL, '_thetaI_lambdaR'),
+  file.dir = paste0('model3/profileReg_RL', RUN_LEVEL),
   seed = 739164,
   packages = c("spatPomp", 'haitipkg', 'pomp')
 )
@@ -69,33 +69,21 @@ prof_vars <- c()
 for (pp in prof_params) {
 
   if (pp == "mu_B") {
-    prof_values <- seq(300, 750, length.out = 30)
+    prof_values <- seq(350, 700, length.out = 25)
   } else if (pp == 'XthetaA') {
-    prof_values <- seq(0.0001, 0.15, length.out = 30)
+    prof_values <- seq(0, sqrt(0.15), length.out = 25)^2
   } else if (pp == 'thetaI') {
-    prof_values <- seq(7.5e-06, 0.0001, length.out = 30)
+    prof_values <- seq(1.05e-05, 1.2e-04, length.out = 30)
   } else if (pp == 'lambdaR') {
-    prof_values <- seq(0.15, 1.75, length.out = 30)
+    prof_values <- c(seq(0, 2, length.out = 24), 1)
   } else if (pp == 'r') {
-    prof_values <- seq(0.05, 1.3, length.out = 30)
+    prof_values <- c(seq(0.55, 1.5, length.out = 24), 1)
   } else if (pp == 'std_W') {
-    prof_values <- seq(0.025, 0.045, length.out = 30)
+    prof_values <- seq(0.024, 0.04, length.out = 27)
   } else if (pp == 'epsilon') {
-    prof_values <- seq(0.66, 1, 0.02)
+    prof_values <- seq(0.84^5, 1, length.out = 14)^(1/5)
   } else if (pp == 'k') {
-    prof_values <- seq(10, 115, length.out = 30)
-  } else if (pp == 'aHur3') {
-    prof_values <- seq(0, 100, length.out = 30)
-  } else if (pp == 'aHur9') {
-    prof_values <- seq(0, 100, length.out = 30)
-  } else if (pp == 'hHur3') {
-    prof_values <- seq(0, 55, length.out = 30)
-  } else if (pp == 'hHur9') {
[…]

### 2024-01-22 H1profile.R
Double checked files are ready for RL3

@@ -18,7 +18,7 @@ library(tidyverse)
 library(data.table)
 library(haitipkg)
 
-RUN_LEVEL <- 2
+RUN_LEVEL <- 3
 
 nprof      <- switch(RUN_LEVEL,  2,    15, 25)
 NMIF       <- switch(RUN_LEVEL,  5,   150, 200)
@@ -30,7 +30,7 @@ COOLING    <- 0.5
 # Create Experiment Registry ----------------------------------------------
 
 reg <- makeExperimentRegistry(
-  file.dir = paste0('model1/profileReg_RL', RUN_LEVEL, '_tau'),
+  file.dir = paste0('model1/profileReg_RL', RUN_LEVEL),
   seed = 739164,
   packages = c("spatPomp", 'haitipkg', 'pomp')
 )
@@ -39,30 +39,34 @@ reg <- makeExperimentRegistry(
 
 set.seed(665544)
 
+# Create Model 1 pomp object, used to get fixed parameters in the searches.
 h1 <- haiti1_joint()
 
+# Vector of parameters to profile over.
 prof_params <- c(
-  # 'betat',
+  'betat',
   'tau_epi',
-  'tau_end'
-  # 'rho',
-  # 'nu',
-  # 'sig_sq_epi',
-  # 'sig_sq_end',
-  # 'E_0',
-  # 'I_0'
-  # 'beta1',
-  # 'beta2',
-  # 'beta3',
-  # 'beta4',
-  # 'beta5',
-  # 'beta6'
+  'tau_end',
+  'rho',
+  'nu',
+  'sig_sq_epi',
+  'sig_sq_end',
+  'E_0',
+  'I_0',
+  'beta1',
+  'beta2',
+  'beta3',
+  'beta4',
+  'beta5',
+  'beta6'
 )
 
+# Initialize parameter matrix using default parameter values.
 final_pars <- coef(h1)
 prof_vars <- c()
-for (pp in prof_params) {
+for (pp in prof_params) {  # Loop through all profile parameters
 
+  # Profile bounds estimated using RL = 2
   if (pp == "betat") {
     prof_values <- seq(-0.15, 0.05, length.out = 30)
   } else if (pp == 'tau_epi') {
@@ -98,6 +102,7 @@ for (pp in prof_params) {
   prof_cols <- matrix(rep(prof_values, each = nprof), ncol = 1)
   colnames(prof_cols) <- pp
 
+  # Parameter bounds found using RL 2.
[…]

### 2024-01-22 H3profile.R

@@ -40,7 +40,7 @@ reg <- makeExperimentRegistry(
 
 set.seed(665544)
 
-# Read in previous best results
+# Create H3 model object to get fixed parameters
 h3_spat <- haiti3_spatPomp()
 final_pars <- coef(h3_spat)
 
@@ -51,18 +51,7 @@ shared_param_names <- c(
   "epsilon", "k"
 )
 
-prof_params <- c(
-  'thetaI',
-  'lambdaR'
-)
-
-# prof_params <- shared_param_names
-# If you wanted to profile over the six added unit-specific
-# parameters
-# prof_params <- c(
-#   shared_param_names, 'aHur3', 'hHur3',
-#   'aHur9', 'hHur9', 'Iinit3', 'Iinit4'
-# )
+prof_params <- shared_param_names
 
 best_pars <- final_pars
 prof_vars <- c()
@@ -222,13 +211,14 @@ for (pp in prof_params) {
   prof_vars <- c(prof_vars, rep(pp, nrow(guesses_all)))
 }
 
+# Remove the first row which contains the original params as a place-holder.
 final_pars <- final_pars[-1, names(coef(h3_spat))]
-
 data_obj <- list(starts = final_pars, prof_vars = prof_vars)
 # dim(final_pars)
 
 # Define profile problem for registry -------------------------------------
 
+# Function to sample a starting point for a given profile search
 sample_starting_point <- function(data = data_obj, job, i, ...) {
   list(
     start_par = unlist(data$starts[i, ]),
@@ -240,6 +230,9 @@ addProblem(name = 'profile', data = data_obj, fun = sample_starting_point)
 
 # Define algorithm for parameter estimation -------------------------------
 
+# Function used to fit the model under the constraints of the profile search.
+# If parameter "PARAM" is getting profiled over, the RW_SD is set to zero
+# for "PARAM" and all other fixed parameters.
 fit_model <- function(
     data, job, instance, nbpf, np, spat_regression, np_eval, nreps_eval,
     cooling, start_date = "2010-11-20", ...
@@ -315,8 +308,6 @@ addExperiments(prob.designs = pdes, algo.designs = ades)
 
 # Submit Jobs -------------------------------------------------------------
 
-# resources1 <- list(account = 'stats_dept1', walltime = '10:00', memory = '5000m', ncpus = 1)
 resources1 <- list(account = 'ionides2', walltime = '8:00:00', memory = '9000m', ncpus = 1)
 
[…]

### 2024-01-23 H1profile.R
Added more documentation and comments at the beginning and end of profile scripts

@@ -1,15 +1,52 @@
 # Fitting profiles for Haiti 1.
 #
-# The fit_haiti1 function is a very simple helper that will take as input
-# some starting parameters, and hyperparameters for the IF2 algorithm, and
+# ********
+# NOTE:
+#
+#   If you would like to run this code but have never used the `batchtools` R
+#   package, then we strongly recommend completing both Examples 1 and 2 from
+#   the batchtools vignette, by first installing the `batchtools` package and
+#   running: `vignette('batchtools')`, or visiting:
+#
+#         https://mllg.github.io/batchtools/articles/batchtools.html
+#
+# WARNING:
+#
+#   This script was ran using a high-performance computing cluster. In total,
+#   the RUN_LEVEL == 3 version of this script took TODO compute hours. Partial
+#   results can be obtained by lowering the required computational effort
+#   (params: NP, NMIF, nprof, NP_EVAL, NREPS_EVAL), which can be done by
+#   reducing the RUN_LEVEL to 2 or 1, or alternatively by manually changing
+#   these values.
+# ********
+#
+# The haitipkg::fit_haiti1 function is a very simple helper that will take as
+# input some starting parameters, and hyperparameters for the IF2 algorithm, and
 # will return parameter estimates along with corresponding estimates of their
 # log-likelihoods.
 #
 # In this script we will use this function to do a profile likelihood search
-# of all model parameters using the batchtools package. To do this, we will
+# of all model parameters using the `batchtools` package. To do this, we will
 # create a grid of parameter values, and submit jobs where each job will use
 # a row from the grid as a starting point to fit the model.
-
+#
+# The number of rows in the grid (which is the `final_pars` object that is
+# created using a for-loop over the profile parameters) correspond to the total
+# number of jobs that will be submitted using `batchtools`. Each job takes
+# roughly the same amount of time to compute, and hence the run-time complexity
+# is linear with respect to the total number of jobs. The total number of jobs
+# can be controlled by either (1) changing the number of variables to profile
+# over (2) changing the number of unique values for each profile parameter (3)
+# changing the number of points per profile point.
+#
+# The time-complexity of each job is roughly linear in the parameters NP, and
+# NMIF. The fitting algorithm (IF2) has a computational complexity that is
+# linear with these parameters, and the fitting procedure takes the majority
[…]

### 2024-01-23 H2profile.R

@@ -1,5 +1,17 @@
 # Fitting profiles for Haiti 2.
 #
+# ********
+# NOTE:
+#
+#   If you would like to run this code but have never used the `batchtools` R
+#   package, then we strongly recommend completing both Examples 1 and 2 from
+#   the batchtools vignette, by first installing the `batchtools` package and
+#   running: `vignette('batchtools')`, or visiting:
+#
+#         https://mllg.github.io/batchtools/articles/batchtools.html
+#
+# ********
+#
 # The fit_haiti2 function is a very simple helper that will take as input
 # some starting parameters and an objective function for Model 2 and will
 # return parameters estimates along with corresponding estimates of their
@@ -25,7 +37,7 @@ library(subplex)
 # Create Experiment Registry ----------------------------------------------
 
 reg <- makeExperimentRegistry(
-  file.dir ='model2/profileReg_RL',
+  file.dir ='model2/profileReg',
   seed = 739164,
   packages = c("spatPomp", 'haitipkg', 'pomp')
 )
@@ -215,3 +227,48 @@ submitJobs(
   ), resources = resources1
 )
 
+# Get Results -------------------------------------------------------------
+
+# You can check the status of the jobs (how many are queued, how many have
+# started, how many have finished, and how many have errors / ran out of time)
+# by running:
+#
+# getJobStatus()
+#
+# You can ensure that you won't summarize the results until everything has
+# finished by running:
+#
+# waitForJobs()
+#
+# Because there are many long jobs, however, it may not be the best idea to do
+# this interactively. Instead, it's wise to periodically check on the status
+# of the jobs when you think they may have finished. Therefore if you would
+# like to check on your results after closing your R session, you need to
+# re-load your registry (after opening up a new R-session in the same
+# working directory that was used to create the registry) using:
+#
+# library(batchtools)
+#
+# reg = loadRegistry('model2/profileReg')
+# getStatus()
+#
+# Once all of the jobs are finished, we can summarize the results using:
+#
+# h2_profile_results = unwrap(reduceResultsDataTable())
+# h2_pars = unwrap(getJobPars())
+# h2_results = ijoin(h2_pars, h2_profile_results)
+#
+# saveRDS(
+#   h2_profile_results,
+#   'model2/h2_profiles.rds')
+# )
+#
+# We can also get information about each job, which can be helpful to pick the
+# size of each run level and to estimate the cost of the RUN_LEVEL_3 search.
[…]

### 2024-01-23 H3profile.R

@@ -1,5 +1,25 @@
 # Fitting profiles for Haiti 3.
 #
+# ********
+# NOTE:
+#
+#   If you would like to run this code but have never used the `batchtools` R
+#   package, then we strongly recommend completing both Examples 1 and 2 from
+#   the batchtools vignette, by first installing the `batchtools` package and
+#   running: `vignette('batchtools')`, or visiting:
+#
+#         https://mllg.github.io/batchtools/articles/batchtools.html
+#
+# WARNING:
+#
+#   This script was ran using a high-performance computing cluster. In total,
+#   the RUN_LEVEL == 3 version of this script took TODO compute hours. Partial
+#   results can be obtained by lowering the required computational effort
+#   (params: NP, NBPF, nprof, NP_EVAL, NREPS_EVAL), which can be done by
+#   reducing the RUN_LEVEL to 2 or 1, or alternatively by manually changing
+#   these values.
+# ********
+#
 # The fit_haiti3 function is a very simple helper that will take as input
 # some starting parameters, and hyperparameters for the IBPF algorithm, and
 # will return parameter estimates along with corresponding estimates of their
@@ -9,6 +29,24 @@
 # of all model parameters using the batchtools package. To do this, we will
 # create a grid of parameter values, and submit jobs where each job will use
 # a row from the grid as a starting point to fit the model.
+#
+# The number of rows in the grid (which is the `final_pars` object that is
+# created using a for-loop over the profile parameters) correspond to the total
+# number of jobs that will be submitted using `batchtools`. Each job takes
+# roughly the same amount of time to compute, and hence the run-time complexity
+# is linear with respect to the total number of jobs. The total number of jobs
+# can be controlled by either (1) changing the number of variables to profile
+# over (2) changing the number of unique values for each profile parameter (3)
+# changing the number of points per profile point.
+#
+# The time-complexity of each job is roughly linear in the parameters NP, and
+# NBPF. The fitting algorithm (IBPF) has a computational complexity that is
+# linear with these parameters, and the fitting procedure takes the majority
+# of the time for each job. The model fit is evaluated in each job using the
+# block particle filter algorithm, which has a computational complexity that is
+# linear in NP_EVAL and NREPS_EVAL, so modifying these parameters will also
+# have an effect on the total computational effort required, but is typically
[…]

### 2024-01-23 ms.Rnw
Incorporated H1 profile results into the manuscript

@@
 #                    RUN_LEVEL =  1,   2,    3
-N_PARAMS   <- switch(RUN_LEVEL,  10,  20,  nrow(h1LeeStartsEpi))  # Number of parameters to sample
-N_LEE_SIMS <- switch(RUN_LEVEL,  10,  20,   20)  # Number of simulations per parameter
+N_PARAMS   <- switch(RUN_LEVEL,   5,  20,  nrow(h1LeeStartsEpi))  # Number of parameters to sample
+N_LEE_SIMS <- switch(RUN_LEVEL,   5,  20,   20)  # Number of simulations per parameter
 NP_EVAL    <- switch(RUN_LEVEL,  50, 200, 2000)  # Number of particles used to evaluate the likelihood
-NREPS_EVAL <- switch(RUN_LEVEL,   3,   5,   36)  # Number of replicates of likelihood evaluation.
+NREPS_EVAL <- switch(RUN_LEVEL,   2,   5,   36)  # Number of replicates of likelihood evaluation.
 
@@
 
-# Compute the run_level_<RUN_LEVEL> results and save them.
+if (RUN_LEVEL == 1 && file.exists('model1/run_level_3/lee1_epi_evals.rds')) {
+
+  # If RUN_LEVEL == 1, and RUN_LEVEL == 3 results exist, just load the RL3
+  # results.
   lee_epi_ll <- readRDS("model1/run_level_3/lee1_epi_evals.rds")
-# lee_epi_ll <- bake(
-#   file = paste0("model1/", rl_dir, "lee1_epi_evals.rds"), {
-#
-#     # Randomly sample from starting params used by Lee et al (2020).
-#     # If RUN_LEVEL == 3, all starting parameters are used.
-#     which_params <- sample(
-#       1:nrow(h1LeeStartsEpi),
-#       size = N_PARAMS,
-#       replace = FALSE
-#     )
-#     epi_params <- h1LeeStartsEpi[which_params, ]
-#
-#     # Create data.frame to save results
-#     lee_epi_ll <- data.frame(
-#       "pfLL" = NA_real_,
-#       "pfse" = NA_real_,
-#       "parid" = epi_params$parid
-#     )
-#
-#     for (j in 1:nrow(lee_epi_ll)) {
-#       # Get parameters of interest
-#       pf_params <- epi_params[j, -1][names(coef(h1_epi))]
-#
-#       # make results reproducible
-#       registerDoRNG((j * 687383921) %% 7919)
-#
-#       # Calculate log likelihoods
-#       ll_evals <- foreach(i=1:NREPS_EVAL, .combine = c) %dopar% {
-#         logLik(pfilter(h1_epi, params = pf_params, Np = NP_EVAL))
-#       }
-#       lee_epi_ll[j, 'pfLL'] <- logmeanexp(ll_evals, se = TRUE)[1]  # Log-lik
-#       lee_epi_ll[j, 'pfse'] <- logmeanexp(ll_evals, se = TRUE)[2]  # SE
-#     }
-#
-#     lee_epi_ll
-#   }
-# )
+} else {
+
+  # Compute the run_level_<RUN_LEVEL> results and save them, or simply load
+  # the results if the bake load conditions are met.
+  lee_epi_ll <- bake(
+    file = paste0("model1/", rl_dir, "lee1_epi_evals.rds"), {
+
+      # Randomly sample from starting params used by Lee et al (2020).
+      # If RUN_LEVEL == 3, all starting parameters are used.
+      which_params <- sample(
+        1:nrow(h1LeeStartsEpi),
+        size = N_PARAMS,
+        replace = FALSE
+      )
+      epi_params <- h1LeeStartsEpi[which_params, ]
+
+      # Create data.frame to save results
+      lee_epi_ll <- data.frame(
+        "pfLL" = NA_real_,
+        "pfse" = NA_real_,
+        "parid" = epi_params$parid
+      )
+
+      for (j in 1:nrow(lee_epi_ll)) {
+        # Get parameters of interest
+        pf_params <- epi_params[j, -1][names(coef(h1_epi))]
+
+        # make results reproducible
+        registerDoRNG((j * 687383921) %% 7919)
+
+        # Calculate log likelihoods
+        ll_evals <- foreach(i=1:NREPS_EVAL, .combine = c) %dopar% {
+          logLik(pfilter(h1_epi, params = pf_params, Np = NP_EVAL))
+        }
+        lee_epi_ll[j, 'pfLL'] <- logmeanexp(ll_evals, se = TRUE)[1]  # Log-lik
+        lee_epi_ll[j, 'pfse'] <- logmeanexp(ll_evals, se = TRUE)[2]  # SE
+      }
+
+      lee_epi_ll
+    }
+  )
+}
 
@@
 
+if (RUN_LEVEL == 1 && file.exists('model1/run_level_3/lee1_end_evals.rds')) {
+
+  # If RUN_LEVEL == 1, and RUN_LEVEL == 3 results exist, just load the RL3
+  # results.
   lee_end_ll <- readRDS("model1/run_level_3/lee1_end_evals.rds")
-# lee_end_ll <- bake(
-#   file = paste0("model1/", rl_dir, "lee1_end_evals.rds"), {
-#
-#     end_params <- h1LeeStartsEnd[h1LeeStartsEnd$parid %in% lee_epi_ll$parid, ]
-#
-#     # Create data.frame to save results
-#     lee_end_ll <- data.frame(
-#       "pfLL" = NA_real_,
-#       "pfse" = NA_real_,
-#       'parid' = end_params$parid
-#     )
-#
-#     # Perform particle filter to get likelihood estimates
-#     for (j in 1:nrow(lee_end_ll)) {
-#       # Get parameters of interest
-#       pf_params <- end_params[j, ][names(coef(h1_end))]
-#
-#       # make results reproducible
-#       registerDoRNG((j * 687383921) %% 7919)
-#
-#       # Calculate log likelihoods
-#       ll_evals <- foreach(i=1:NREPS_EVAL, .combine = c) %dopar% {
-#         logLik(pfilter(h1_end, params = pf_params, Np = NP_EVAL))
-#       }
-#
-#       lee_end_ll[j, 'pfLL'] <- logmeanexp(ll_evals, se = TRUE)[1]
-#       lee_end_ll[j, 'pfse'] <- logmeanexp(ll_evals, se = TRUE)[2]
-#     }
-#
-#     lee_end_ll
-#   }
-# )
-
-# Combine epidemic and endemic likelihoods by parid to get the likelihood of
+} else {
+
+  # Compute the run_level_<RUN_LEVEL> results and save them, or simply load
+  # the results if the bake load conditions are met.
+  lee_end_ll <- bake(
+    file = paste0("model1/", rl_dir, "lee1_end_evals.rds"), {
+
+      end_params <- h1LeeStartsEnd[h1LeeStartsEnd$parid %in% lee_epi_ll$parid, ]
+
+      # Create data.frame to save results
+      lee_end_ll <- data.frame(
+        "pfLL" = NA_real_,
+        "pfse" = NA_real_,
+        'parid' = end_params$parid
+      )
+
+      # Perform particle filter to get likelihood estimates
+      for (j in 1:nrow(lee_end_ll)) {
+        # Get parameters of interest
+        pf_params <- end_params[j, ][names(coef(h1_end))]
+
+        # make results reproducible
+        registerDoRNG((j * 687383921) %% 7919)
+
+        # Calculate log likelihoods
+        ll_evals <- foreach(i=1:NREPS_EVAL, .combine = c) %dopar% {
+          logLik(pfilter(h1_end, params = pf_params, Np = NP_EVAL))
+        }
+
+        lee_end_ll[j, 'pfLL'] <- logmeanexp(ll_evals, se = TRUE)[1]
+        lee_end_ll[j, 'pfse'] <- logmeanexp(ll_evals, se = TRUE)[2]
+      }
+
+      lee_end_ll
+    }
+  )
+}
+
+# Combine epidemic and endemic likelihoods by `parid` to get the likelihood of
 # the complete model.
@@
 
+# Doing some clean-up. The goal is that each major chunk can be run independent
+# of other chunks. Removing saved objects is not entirely necessary, but can
+# help new users know what objects are needed later on and which were only
+# temporarily needed for each particular chunk. Here we suppress warnings
+# because some of the temporary objects only exist if the `bake` code needs to
+# be recomputed, and hence `rm` will throw warnings if it tries to remove
+# objects that don't exist.
+suppressWarnings(
   rm(
@@
   )
-gc()
+)
+gc()  # Garbage Collection to free un-used memory
 @
@@
 
-<<FitModel1, echo=FALSE, message=FALSE, include=FALSE>>=
+<<LoadModel1Params, echo=FALSE, message=FALSE, include=FALSE>>=
 
@@
 
-# Check RUN_LEVEL, and set hyper-parameters to IF2 and pfilter accordingly.
-#                      RUN_LEVEL =    1,    2,      3
-NP_H1        <- switch(RUN_LEVEL,   100, 1000,   2000)
-NMIF_H1      <- switch(RUN_LEVEL,    10,   50,    200)
-NUM_TREND    <- switch(RUN_LEVEL,     5,   10,     90)
-NPROF        <- switch(RUN_LEVEL,     4,    4,     24)
-NP_EVAL      <- switch(RUN_LEVEL,   200, 1000,   5000)
-NREPS_EVAL   <- switch(RUN_LEVEL, cores,   20,  cores)
-
-# Create the <RUN_LEVEL> version of the results and save them.
-h1_MIF <- readRDS('model1/run_level_3/haiti1_fit.rds')
-
-# h1_MIF <- bake(
-#   file = paste0("model1/", rl_dir, "haiti1_fit.rds"), {
-#     fit_haiti1(  # haitipkg function used to fit Model 1
-#       NP = NP_H1,
-#       NMIF = NMIF_H1,
-#       NUM_TREND = NUM_TREND,
-#       NPROF = NPROF,
-#       NREPS_EVAL = NREPS_EVAL,
-#       NP_EVAL = NP_EVAL
-#     )
-#   }
-# )
+# Load the file containing profiles of Model 1 Parameters.
+if (RUN_LEVEL != 3 && file.exists("model1/run_level_3/h1_profiles.rds")) {
+  h1_prof <- readRDS("model1/run_level_3/h1_profiles.rds")
+} else if (file.exists(paste0('model1', rl_dir, 'h1_profiles.rds'))) {
+  h1_prof <- readRDS(paste0('model1', rl_dir, 'h1_profiles.rds'))
+} else {
+  stop(paste0("Missing file: `model1/", rl_dir, 'h1_profiles.rds`;\nThe file can be obtained by running the script `H1profile.R`'))
+}
 
@@
 # we calculate the maximum likelihood for each value of betat.
-h1_prof_res <- h1_MIF %>%
+h1_prof_res <- h1_prof %>%
+  filter(prof_var == 'betat') %>%
   group_by(betat) %>%
-  summarize(logLik = max(ll))
-
+  summarize(logLik = max(logLik))
+
+# TODO: This is used later in figure 2 (TODO)
 mcap_results <- pomp::mcap(h1_prof_res$logLik, h1_prof_res$betat, span = 1)
 
-# Get the best set of parameters from MIF2 search.
-p1 <- h1_MIF %>%
-  filter(ll == max(ll)) %>%
-  select(-ll, -ll.se) %>%
+# Find the set of parameters that correspond to the maximum likelihood
+p1 <- h1_prof %>%
+  slice_max(order_by = logLik, n = 1) %>%
+  # Remove non-parameters from the data.frame (hyperparameters + likelihood)
+  select(-(1:12)) %>%
   unlist()
@@
 # Save the likelihood of the best parameters, for use later.
-mod1_ll <- h1_MIF %>%
-  filter(ll == max(ll)) %>%
-  select(ll) %>%
+mod1_ll <- h1_prof %>%
+  slice_max(order_by = logLik, n = 1) %>%
+  select(logLik) %>%
   unlist()
@@
 # Just a bit of tidying up: removing parameters that we won't use again.
-rm(NP_H1, NMIF_H1, NUM_TREND, NPROF, NREPS_EVAL, NP_EVAL, h1_MIF)
+rm(h1_prof)
 gc()
-
 @
@@
 <<FitModel2, echo=FALSE, include=FALSE, message=FALSE>>=
+
+# TODO: Model 2
 
@@
 
-<<table-input,echo=FALSE,eval=T, include=FALSE, message=FALSE>>=
+<<createModelObjects,echo=FALSE,eval=TRUE, include=FALSE, message=FALSE>>=
 
@@
   geom_line(data = mcap_results$fit, aes(x = parameter, y = smoothed), col = 'blue') +
+  geom_line(data = mcap_results$fit, aes(x = parameter, y = quadratic), col = 'red') +
   geom_vline(xintercept = mcap_results$ci[1], linetype = 'dashed') +
@@
 {\figTitle Confidence interval for linear trend in transmission.}
-Monte Carlo adjusted profile of $\transmissionTrend$ for Model~1.
-The blue curve is the profile, the blue line indicates the MLE, and the dashed lines indicate the $95\%$ confidence interval.
+Monte Carlo adjusted profile (MCAP) of $\transmissionTrend$ for Model~1.
+The blue curve is the MCAP, the vertical blue line indicates the MLE, and the vertical dashed lines indicate the $95\%$ confidence interval.
 }
 \end{figure}
+
+<<cleanUpBetatFig, echo=FALSE, message=FALSE, include=FALSE, warning=FALSE>>=
+suppressWarnings(
+  rm(h1_prof_res, mcap_results)
+)
+gc()
+@
 
@@
 # projects the vaccination scenario, under the specified model.
-h1_S0_results <- readRDS("model1/run_level_3/S0_sims.rds")
-# h1_S0_results <- bake(
-#   file = paste0("model1/", rl_dir, "S0_sims.rds"), {
-#
-#     # Create the model
-#     s0 <- haiti1_joint(vacscen = 'id0')
-#
-#     # Load coefficients
-#     coef(s0) <- p1
-#
-#     # Get samples from filtering distribution
-#     s0_pf <- pfilter(s0, save.state = TRUE, Np = h1_scen_NP)
-#     end_states <- saved_states(s0_pf)
-#     end_states <- end_states[[length(end_states)]]
-#
-#     # project vacc scenario, with filtering dist, and model.
-#     h1_S0_sims <- project_from_filter(mod = s0, end_states = end_states, nsims = h1_scen_sims)
-#
-#     results <- list()
-#     results$mod1_V0_sims <- agg_mod1_sims(h1_S0_sims)
-#     results$mod1_V0_probs <- get_elimProbs(h1_S0_sims, model = 1)
-#
-#     results
-#   },
-#   timing = FALSE
-# )
+h1_S0_results <- bake(
+  file = paste0("model1/", rl_dir, "S0_sims.rds"), {
+
+    # Create the model
+    s0 <- haiti1_joint(vacscen = 'id0')
+
+    # Load coefficients
+    coef(s0) <- p1
+
+    # Get samples from filtering distribution
+    s0_pf <- pfilter(s0, save.state = TRUE, Np = h1_scen_NP)
+    end_states <- saved_states(s0_pf)
+    end_states <- end_states[[length(end_states)]]
+
+    # project vacc scenario, with filtering dist, and model.
+    h1_S0_sims <- project_from_filter(mod = s0, end_states = end_states, nsims = h1_scen_sims)
+
+    results <- list()
+    results$mod1_V0_sims <- agg_mod1_sims(h1_S0_sims)
+    results$mod1_V0_probs <- get_elimProbs(h1_S0_sims, model = 1)
+
+    results
+  },
+  timing = FALSE
+)
 
@@
 # projects the vaccination scenario, under the specified model.
-h1_S2_results <- readRDS("model1/run_level_3/S2_sims.rds")
-# h1_S2_results <- bake(
-#   file = paste0('model1/', rl_dir, 'S2_sims.rds'), {
-#
-#     # Create the model
-#     s2 <- haiti1_joint(vacscen = "id2")
-#
-#     # Set model parameters that need to be adjusted
-#     depts <- 2
-#     h1_par_names <- names(p1)
-#     h1_params_temp <- c(p1, rep(0.0, 5 * depts))
-#     names(h1_params_temp) <- c(
-#       h1_par_names,
-#       paste0("S", 1:depts, "_0"),
-#       paste0("E", 1:depts, "_0"),
-#       paste0("I", 1:depts, "_0"),
-#       paste0("A", 1:depts, "_0"),
-#       paste0("R", 1:depts, "_0")
-#     )
-#     coef(s2) <- h1_params_temp  # Set model parameters
-#
-#     # Get samples from filtering distribution for Scenario 2 model:
-#     s2_pf <- pfilter(s2, save.state = TRUE, Np = h1_scen_NP)
-#     end_states <- saved_states(s2_pf)
-#     end_states <- end_states[[length(end_states)]]
-#
-#     # project vacc scenario, with filtering dist, and model.
-#     h1_S2_sims <- project_from_filter(mod = s2, end_states = end_states, nsims = h1_scen_sims)
-#
-#     results <- list()
-#     results$mod1_V1_sims <- agg_mod1_sims(h1_S2_sims)
-#     results$mod1_V1_probs <- get_elimProbs(h1_S2_sims, model = 1)
-#
-#     results
-#   },
-#   timing = FALSE
-# )
+h1_S2_results <- bake(
+  file = paste0('model1/', rl_dir, 'S2_sims.rds'), {
+
+    # Create the model
+    s2 <- haiti1_joint(vacscen = "id2")
+
+    # Set model parameters that need to be adjusted
+    depts <- 2
+    h1_par_names <- names(p1)
+    h1_params_temp <- c(p1, rep(0.0, 5 * depts))
+    names(h1_params_temp) <- c(
+      h1_par_names,
+      paste0("S", 1:depts, "_0"),
+      paste0("E", 1:depts, "_0"),
+      paste0("I", 1:depts, "_0"),
+      paste0("A", 1:depts, "_0"),
+      paste0("R", 1:depts, "_0")
+    )
+    coef(s2) <- h1_params_temp  # Set model parameters
+
+    # Get samples from filtering distribution for Scenario 2 model:
+    s2_pf <- pfilter(s2, save.state = TRUE, Np = h1_scen_NP)
+    end_states <- saved_states(s2_pf)
+    end_states <- end_states[[length(end_states)]]
+
+    # project vacc scenario, with filtering dist, and model.
+    h1_S2_sims <- project_from_filter(mod = s2, end_states = end_states, nsims = h1_scen_sims)
+
+    results <- list()
+    results$mod1_V1_sims <- agg_mod1_sims(h1_S2_sims)
+    results$mod1_V1_probs <- get_elimProbs(h1_S2_sims, model = 1)
+
+    results
+  },
+  timing = FALSE
+)
 
@@
 # projects the vaccination scenario, under the specified model.
-h1_S4_results <- readRDS("model1/run_level_3/S4_sims.rds")
-# h1_S4_results <- bake(
-#   file = paste0('model1/', rl_dir, 'S4_sims.rds'), {
-#
-#     # Create the model
-#     s4 <- haiti1_joint(vacscen = "id4")
-#
-#     # Set model parameters that need to be adjusted
-#     depts <- 3
-#     h1_par_names <- names(p1)
-#     h1_params_temp <- c(p1, rep(0.0, 5 * depts))
-#     names(h1_params_temp) <- c(
-#       h1_par_names,
-#       paste0("S", 1:depts, "_0"),
-#       paste0("E", 1:depts, "_0"),
-#       paste0("I", 1:depts, "_0"),
-#       paste0("A", 1:depts, "_0"),
-#       paste0("R", 1:depts, "_0")
-#     )
-#     coef(s4) <- h1_params_temp  # Set model parameters
-#
-#     # Get samples from filtering distribution for Scenario 2 model:
-#     s4_pf <- pfilter(s4, save.state = TRUE, Np = h1_scen_NP)
-#     end_states <- saved_states(s4_pf)
-#     end_states <- end_states[[length(end_states)]]
-#
-#     # project vacc scenario, with filtering dist, and model.
-#     h1_S4_sims <- project_from_filter(mod = s4, end_states = end_states, nsims = h1_scen_sims)
-#
-#     results <- list()
-#     results$mod1_V2_sims <- agg_mod1_sims(h1_S4_sims)
-#     results$mod1_V2_probs <- get_elimProbs(h1_S4_sims, model = 1)
-#
-#     results
-#   },
-#   timing = FALSE
-# )
+h1_S4_results <- bake(
+  file = paste0('model1/', rl_dir, 'S4_sims.rds'), {
+
+    # Create the model
+    s4 <- haiti1_joint(vacscen = "id4")
+
+    # Set model parameters that need to be adjusted
+    depts <- 3
+    h1_par_names <- names(p1)
+    h1_params_temp <- c(p1, rep(0.0, 5 * depts))
+    names(h1_params_temp) <- c(
+      h1_par_names,
+      paste0("S", 1:depts, "_0"),
+      paste0("E", 1:depts, "_0"),
+      paste0("I", 1:depts, "_0"),
+      paste0("A", 1:depts, "_0"),
+      paste0("R", 1:depts, "_0")
+    )
+    coef(s4) <- h1_params_temp  # Set model parameters
+
+    # Get samples from filtering distribution for Scenario 2 model:
+    s4_pf <- pfilter(s4, save.state = TRUE, Np = h1_scen_NP)
+    end_states <- saved_states(s4_pf)
+    end_states <- end_states[[length(end_states)]]
+
+    # project vacc scenario, with filtering dist, and model.
+    h1_S4_sims <- project_from_filter(mod = s4, end_states = end_states, nsims = h1_scen_sims)
+
+    results <- list()
+    results$mod1_V2_sims <- agg_mod1_sims(h1_S4_sims)
+    results$mod1_V2_probs <- get_elimProbs(h1_S4_sims, model = 1)
+
+    results
+  },
+  timing = FALSE
+)
 
@@
 # projects the vaccination scenario, under the specified model.
-h1_S3_results <- readRDS("model1/run_level_3/S3_sims.rds")
-# h1_S3_results <- bake(
-#   file = paste0('model1/', rl_dir, 'S3_sims.rds'), {
-#
-#     # Create the model
-#     s3 <- haiti1_joint(vacscen = "id3")
-#
-#     # Set model parameters that need to be adjusted
-#     depts <- 10
-#     h1_par_names <- names(p1)
-#     h1_params_temp <- c(p1, rep(0.0, 5 * depts))
-#     names(h1_params_temp) <- c(
-#       h1_par_names,
-#       paste0("S", 1:depts, "_0"),
-#       paste0("E", 1:depts, "_0"),
-#       paste0("I", 1:depts, "_0"),
-#       paste0("A", 1:depts, "_0"),
-#       paste0("R", 1:depts, "_0")
-#     )
-#     coef(s3) <- h1_params_temp  # Set model parameters
-#
-#     # Get samples from filtering distribution for Scenario 2 model:
-#     s3_pf <- pfilter(s3, save.state = TRUE, Np = h1_scen_NP)
-#     end_states <- saved_states(s3_pf)
-#     end_states <- end_states[[length(end_states)]]
-#
-#     # project vacc scenario, with filtering dist, and model.
-#     h1_S3_sims <- project_from_filter(mod = s3, end_states = end_states, nsims = h1_scen_sims)
-#
-#     results <- list()
-#     results$mod1_V3_sims <- agg_mod1_sims(h1_S3_sims)
-#     results$mod1_V3_probs <- get_elimProbs(h1_S3_sims, model = 1)
-#
-#     results
-#   },
-#   timing = FALSE
-# )
+h1_S3_results <- bake(
+  file = paste0('model1/', rl_dir, 'S3_sims.rds'), {
+
+    # Create the model
+    s3 <- haiti1_joint(vacscen = "id3")
+
+    # Set model parameters that need to be adjusted
+    depts <- 10
+    h1_par_names <- names(p1)
+    h1_params_temp <- c(p1, rep(0.0, 5 * depts))
+    names(h1_params_temp) <- c(
+      h1_par_names,
+      paste0("S", 1:depts, "_0"),
+      paste0("E", 1:depts, "_0"),
+      paste0("I", 1:depts, "_0"),
+      paste0("A", 1:depts, "_0"),
+      paste0("R", 1:depts, "_0")
+    )
+    coef(s3) <- h1_params_temp  # Set model parameters
+
+    # Get samples from filtering distribution for Scenario 2 model:
+    s3_pf <- pfilter(s3, save.state = TRUE, Np = h1_scen_NP)
+    end_states <- saved_states(s3_pf)
+    end_states <- end_states[[length(end_states)]]
+
+    # project vacc scenario, with filtering dist, and model.
+    h1_S3_sims <- project_from_filter(mod = s3, end_states = end_states, nsims = h1_scen_sims)
+
+    results <- list()
+    results$mod1_V3_sims <- agg_mod1_sims(h1_S3_sims)
+    results$mod1_V3_probs <- get_elimProbs(h1_S3_sims, model = 1)
+
+    results
+  },
+  timing = FALSE
+)
 
@@
 # projects the vaccination scenario, under the specified model.
-h1_S1_results <- readRDS("model1/run_level_3/S1_sims.rds")
-# h1_S1_results <- bake(
-#   file = paste0('model1/', rl_dir, 'S1_sims.rds'), {
-#
-#     # Create the model
-#     s1 <- haiti1_joint(vacscen = "id1")
-#
-#     # Set model parameters that need to be adjusted
-#     depts <- 10
-#     h1_par_names <- names(p1)
-#     h1_params_temp <- c(p1, rep(0.0, 5 * depts))
-#     names(h1_params_temp) <- c(
-#       h1_par_names,
-#       paste0("S", 1:depts, "_0"),
-#       paste0("E", 1:depts, "_0"),
-#       paste0("I", 1:depts, "_0"),
-#       paste0("A", 1:depts, "_0"),
-#       paste0("R", 1:depts, "_0")
-#     )
-#     coef(s1) <- h1_params_temp  # Set model parameters
-#
-#     # Get samples from filtering distribution for Scenario 2 model:
-#     s1_pf <- pfilter(s1, save.state = TRUE, Np = h1_scen_NP)
-#     end_states <- saved_states(s1_pf)
-#     end_states <- end_states[[length(end_states)]]
-#
-#     # project vacc scenario, with filtering dist, and model.
-#     h1_S1_sims <- project_from_filter(mod = s1, end_states = end_states, nsims = h1_scen_sims)
-#
-#     results <- list()
-#     results$mod1_V4_sims <- agg_mod1_sims(h1_S1_sims)
-#     results$mod1_V4_probs <- get_elimProbs(h1_S1_sims, model = 1)
-#
-#     results
-#   },
-#   timing = FALSE
-# )
+h1_S1_results <- bake(
+  file = paste0('model1/', rl_dir, 'S1_sims.rds'), {
+
+    # Create the model
+    s1 <- haiti1_joint(vacscen = "id1")
+
+    # Set model parameters that need to be adjusted
+    depts <- 10
+    h1_par_names <- names(p1)
+    h1_params_temp <- c(p1, rep(0.0, 5 * depts))
+    names(h1_params_temp) <- c(
+      h1_par_names,
+      paste0("S", 1:depts, "_0"),
+      paste0("E", 1:depts, "_0"),
+      paste0("I", 1:depts, "_0"),
+      paste0("A", 1:depts, "_0"),
+      paste0("R", 1:depts, "_0")
+    )
+    coef(s1) <- h1_params_temp  # Set model parameters
+
+    # Get samples from filtering distribution for Scenario 2 model:
+    s1_pf <- pfilter(s1, save.state = TRUE, Np = h1_scen_NP)
+    end_states <- saved_states(s1_pf)
+    end_states <- end_states[[length(end_states)]]
+
+    # project vacc scenario, with filtering dist, and model.
+    h1_S1_sims <- project_from_filter(mod = s1, end_states = end_states, nsims = h1_scen_sims)
+
+    results <- list()
+    results$mod1_V4_sims <- agg_mod1_sims(h1_S1_sims)
+    results$mod1_V4_probs <- get_elimProbs(h1_S1_sims, model = 1)
+
+    results
+  },
+  timing = FALSE
+)
 @

### 2024-01-23 ms.Rnw
Incorporating changes to Model 2 fitting process

@@
 
-<<FitModel2, echo=FALSE, include=FALSE, message=FALSE>>=
+<<FitModel2, echo=FALSE, include=FALSE, message=FALSE, warning=FALSE>>=
+
+# Calculate log-likelihood of Lee et al (2020) model.
+# We have parameter estimates for everything except the variance of the
+# measurement model, so here we fix all parameters except the variance, which
+# is estimated for both endemic and epidemic periods.
+lee20_replication <- bake(
+  file = 'model2/lee20_replication.rds', {
+    h2_lee_epi <- haiti2()
+    lee_epi_params <- h2_lee_epi@params
+    epi_ofun_v <- traj_objfun(
+      h2_lee_epi, est = c('v'), paramnames = c('v')
+    )
+
+    # Get starting values for the estimated parameter
+    start_v_epi <- lee_epi_params[c('v')]
+
+    # Use the subplex package to calibrate 'v' for the epi-model with fixed coefs.
+    epi_fit_v <- subplex::subplex(par = log(start_v_epi), fn = epi_ofun_v)
+    epi_params_update_v <- lee_epi_params
+    epi_params_update_v['v'] <- exp(epi_fit_v$par)
+
+    # Now we repeat this for the endemic portion of the model:
+    h2_lee_end <- haiti2(region = 'after')
+    lee_end_params <- h2_lee_end@params
+    end_ofun_v <- traj_objfun(
+      h2_lee_end, est = 'v', paramnames = 'v'
+    )
+
+    # Get starting values for the estimated parameter
+    start_v_end <- lee_end_params[c('v')]
+
+    # Use the subplex package to calibrate 'v' for the epi-model with fixed coefs.
+    end_fit_v <- subplex::subplex(par = log(start_v_end), fn = end_ofun_v)
+    end_params_update_v <- lee_end_params
+    end_params_update_v['v'] <- exp(epi_fit_v$par)
+
+    # Return the results
+    c(
+      "leeFit_ll" = -epi_fit_v$value - end_fit_v$value,
+      "leeFit_n_params" = 6
+    )
+  }, timing = FALSE
+)
 
@@
 
-# This code chunk fits Model 2. This is a quick caclulation compared to models
-# 1 and 3, so no run_level is used.
-
-h2_fit <- readRDS("model2/model2_fit.rds")
-
-# h2_fit <- bake(
-#   file = paste0("model2/model2_fit.rds"), {
-#     fit_haiti2()
-#   },
-#   timing = FALSE
-# )
+h2_fit <- bake(
+  file = 'model2/model2_fit.rds', {
+
+    # Create model and save initial values for each parameter
+    h2 <- haiti2(cutoff = 10000, measure = "log")  # Default is epidemic period
+    h2_params <- h2@params
+    est_params <- c('Mu', 'Beta', 'BetaW', 'v', 'sigma', 'phase')
+    n_fit_params <- length(est_params)
+
+    # Get objective function to minimize
+    epi_ofun <- pomp::traj_objfun(h2, est = est_params, params = h2_params)
+
+    # Vector of starting values of the parameters that are estimated.
+    # phase is not included because it's the only non log-transformed variable
+    theta <- h2_params[c("Mu", "Beta", "BetaW", "v", "sigma")]
+
+    # Fit the epi model
+    h2_fit_tmp <- subplex::subplex(
+      par = c(log(theta), h2_params['phase']),
+      fn = epi_ofun
+    )
+
+    # Save the fitted parameters to the h2_params vector
+    h2_params[c("Mu","Beta","BetaW","v", "sigma")] <- exp(h2_fit_tmp$par[c("Mu", "Beta", "BetaW", "v", "sigma")])
+    h2_params['phase'] <- h2_fit_tmp$par['phase']
+
+    model2_fit_results <- list()
+    model2_fit_results[['h2_params']] <- h2_params
+    model2_fit_results[["n_fit_params"]] <- n_fit_params
+    model2_fit_results[['logLik']] <- h2_fit_tmp$value
+
+    rm(h2, epi_ofun, theta, h2_fit_tmp)
+
+    model2_fit_results
+  }, timing = FALSE
+)
 @
@@
 
-# Objective function of model 2, epidemic phase
-epi_ofun <- traj_objfun(
-  h2_epi,
-  params = h2_fit$h2_params
-)
-
-mod2_epi_ll <- -epi_ofun(par = h2_fit$h2_params)
-mod2_ll <- mod2_epi_ll - sum(log(h2_epi@data + 1), na.rm = TRUE)
+mod2_ll <- h2_fit$logLik
 
@@
 
-mod2_lee_aic <- 2 * h2_fit$leeFit_n_params - 2 * h2_fit$leeFit_ll
+mod2_lee_aic <- 2 * unname(lee20_replication['leeFit_n_params']) - 2 * unname(lee20_replication['leeFit_ll'])
 
@@
     & ($\Sexpr{myround(mod1_lee_ll, 1)}$)\footnotemark[1] &
-  ($\Sexpr{myround(h2_fit$leeFit_ll, 1)}$) &
+  ($\Sexpr{myround(unname(lee20_replication['leeFit_ll']), 1)}$) &
   ($\Sexpr{myround(lee3_results[["all_ll_eval"]], 1)}$)\footnotemark[2]
@@
  (\Sexpr{as.character(mod1_lee_n_params)}) &
- ($\Sexpr{as.character(h2_fit$leeFit_n_params)}$) &
+ ($\Sexpr{as.character(unname(lee20_replication['leeFit_n_params']))}$) &
  ($\Sexpr{as.character(lee3_results$n_params, 1)}$)
@@
 # Simulate Model 1 from initial conditions.
-sims <- simulate(h1, nsim = 500, format = 'data.frame', seed = 3448931)
+sims <- simulate(
+  h1, nsim = switch(RUN_LEVEL, 50, 100, 500),
+  format = 'data.frame', seed = 3448931
+)
 
@@
 # speed for each run-level.
-mod2_VaccScenarios <- readRDS("model2/VaccinationScenarios.rds")
-# mod2_VaccScenarios <- bake(
-#   file = paste0("model2/VaccinationScenarios.rds"), {
-#     haiti2_vaccScenario(
-#       h2_params = h2_fit$h2_params
-#     )
-#   },
-#   timing = FALSE
-# )
+mod2_VaccScenarios <- bake(
+  file = paste0("model2/VaccinationScenarios.rds"), {
+    haiti2_vaccScenario(
+      h2_params = h2_fit$h2_params
+    )
+  },
+  timing = FALSE
+)
 @

### 2024-01-23 ms.Rnw
Incorporated changes to Model3 fitting process

@@
 
-<<CalculateLee1Likelihoods, echo=FALSE, message=FALSE, include=FALSE>>=
+<<Calculate Lee1 Likelihoods, echo=FALSE, message=FALSE, include=FALSE>>=
 
@@
 
-<<LoadModel1Params, echo=FALSE, message=FALSE, include=FALSE>>=
-
-# This code chunk is used to fit Model 1 using the IF2 algorithm.
-set.seed(636813)
+<<Load Model1 Params, echo=FALSE, message=FALSE, include=FALSE>>=
 
@@
 
-<<FitModel2, echo=FALSE, include=FALSE, message=FALSE, warning=FALSE>>=
+<<Fit Model 2, echo=FALSE, include=FALSE, message=FALSE, warning=FALSE>>=
 
@@
 
-# TODO: Model 2
-
 h2_fit <- bake(
@@
     model2_fit_results[["n_fit_params"]] <- n_fit_params
-    model2_fit_results[['logLik']] <- h2_fit_tmp$value
+    model2_fit_results[['logLik']] <- -h2_fit_tmp$value
 
@@
 
-<<FitModel3, echo=FALSE, include=FALSE, message=FALSE>>=
-
-# This code chunk is used to fit Model 3 using the IBPF algorithm.
-set.seed(8636712)
-
-# Create vectors for the unit and shared parameters
+<<Load Model3 Parameters, echo=FALSE, include=FALSE, message=FALSE>>=
+
+# Calculate the number of parameters that are estimated:
+mod3_n_params <- length(c(
+  "mu_B", "XthetaA", "thetaI", "lambdaR", "r", "std_W", "epsilon", "k",
+  paste0('betaB', 1:10), paste0('foi_add', 1:10),
+  'aHur3', 'aHur9',
+  'hHur3', 'aHur9',
+  'Iinit3', 'Iinit4'
+))
+
+# Load the file containing profiles of Model 1 Parameters.
+if (RUN_LEVEL != 3 && file.exists("model3/run_level_3/h3_profiles.rds")) {
+
+  # If RUN_LEVEL != 3, and RUN_LEVEL == 3 results exist, just load the RL3
+  # results.
+  h3_prof <- readRDS("model3/run_level_3/h3_profiles.rds")
+} else if (file.exists(paste0('model3', rl_dir, 'h3_profiles.rds'))) {
+
+  # If the `model3/run_level_<RUN_LEVEL>/h3_profiles.rds` file exists, load it
+  h3_prof <- readRDS(paste0('model3', rl_dir, 'h3_profiles.rds'))
+} else {
+
+  # Otherwise (missing RL3 file, and the corresponding <RUN_LEVEL> file):
+  # Throw an error, letting users know how to create the file.
+  stop(paste0("Missing file: `model3/", rl_dir, 'h3_profiles.rds`;\nThe file can be obtained by running the script `H3profile.R`'))
+}
+
+# haiti3_fit contains ll estimate for many parameter sets, so we want to find
+# the highest value
+mod3_ll <- h3_prof %>%
+  slice_max(order_by = logLik, n = 1) %>%
+  pull(logLik)
+@
+
+<<Create Model Objects,echo=FALSE,eval=TRUE, include=FALSE, message=FALSE>>=
+
+# Save the parameters that correspond to the MLE of Model 3.
+p3 <- h3_prof %>%
+  slice_max(logLik, n = 1) %>%
+  select(-(job.id:logLik_se)) %>%
+  unlist()
+
 unit_specific_names <- c("betaB", "foi_add", "aHur", "hHur", "Iinit")
@@
 )
-est_param_names <- c(
-  unit_specific_names, shared_param_names
-)
-
-# Add unit numbers to each parameter
-est_param_names_expanded <- paste0(rep(est_param_names, each = 10), 1:10)
-
-# Simple function that is used to set the rw_sd for the first search
-# This function is convenient so that we don't have to type out the rw_sd
-# for each unit parameter.
-set_rw_1 <- function(x) {
-  if (gsub("[[:digit:]]+$", "", x) %in% shared_param_names) {
-    return(0.01)
-  } else if (x %in% paste0(rep(c("aHur", 'hHur'), each = 2), c(3, 9))) {
-    return(expression(ifelse(time >= 2016.754 & time <= 2017, 0.015, 0)))
-  } else if (x %in% c("Iinit3", "Iinit4")) {
-    return(expression(ivp(0.25)))
-  } else if (grepl('^foi_add[[:digit:]]+$', x)) {
-    return(0.015)
-  } else if (grepl('^betaB[[:digit:]]+$', x)) {
-    return(0.01)
-  } else {
-    return(0)
-  }
-}
-
-set_rw_2 <- function(x) {
-  if (gsub("[[:digit:]]+$", "", x) %in% shared_param_names) {
-    return(0.003)
-  } else if (x %in% paste0(rep(c("aHur", 'hHur'), each = 2), c(3, 9))) {
-    return(expression(ifelse(time >= 2016.754 & time <= 2017, 0.005, 0)))
-  } else if (x %in% c("Iinit3", "Iinit4")) {
-    return(expression(ivp(0.125)))
-  } else if (grepl('^foi_add[[:digit:]]+$', x)) {
-    return(0.005)
-  } else if (grepl('^betaB[[:digit:]]+$', x)) {
-    return(0.003)
-  } else {
-    return(0)
-  }
-}
-
-# First Search RW
-reg_rw_1.sd <- lapply(est_param_names_expanded, set_rw_1)
-names(reg_rw_1.sd) <- est_param_names_expanded
-chol_rw_1.sd <- do.call(rw_sd, reg_rw_1.sd)
-
-# Second Search RW
-reg_rw_2.sd <- lapply(est_param_names_expanded, set_rw_2)
-names(reg_rw_2.sd) <- est_param_names_expanded
-chol_rw_2.sd <- do.call(rw_sd, reg_rw_2.sd)
-
-if (RUN_LEVEL == 1){
-  SEARCH1 = list(
-    NBPF = 10,
-    NP = 200,
-    SPAT_REGRESSION = 0.05,
-    NREPS = 20,
-    NP_EVAL = 200,
-    NREPS_EVAL = 3,
-    RW_SD = chol_rw_1.sd,
-    COOLING = 0.5,
-    KEEP_TRACES = TRUE
-  )
-
-  SEARCH2 = list(
-    TOP_N = 8,
-    NBPF = 10,
-    NP = 200,
-    SPAT_REGRESSION = 0.05,
-    NREPS = 2,
-    NP_EVAL = 200,
-    NREPS_EVAL = 3,
-    RW_SD = chol_rw_2.sd,
-    COOLING = 0.5,
-    KEEP_TRACES = TRUE
-  )
-
-} else if (RUN_LEVEL == 2) {
-  SEARCH1 = list(
-    NBPF = 100,
-    NP = 1000,
-    SPAT_REGRESSION = 0.05,
-    NREPS = 72,
-    NP_EVAL = 1000,
-    NREPS_EVAL = 8,
-    RW_SD = chol_rw_1.sd,
-    COOLING = 0.5,
-    KEEP_TRACES = TRUE,
-    KEEP_LIK_MAT = TRUE
-  )
-
-  SEARCH2 = list(
-    TOP_N = 9,
-    NBPF = 100,
-    NP = 1000,
-    SPAT_REGRESSION = 0.05,
-    NREPS = 4,
-    NP_EVAL = 1000,
-    NREPS_EVAL = 9,
-    RW_SD = chol_rw_2.sd,
-    COOLING = 0.5,
-    KEEP_TRACES = TRUE,
-    KEEP_LIK_MAT = TRUE
-  )
-
-} else if (RUN_LEVEL == 3) {
-
-  SEARCH1 = list(
-    NBPF = 100,
-    NP = 2500,
-    SPAT_REGRESSION = 0.05,
-    NREPS = 144,
-    NP_EVAL = 2500,
-    NREPS_EVAL = 18,
-    RW_SD = chol_rw_1.sd,
-    COOLING = 0.5,
-    KEEP_TRACES = FALSE,
-    KEEP_LIKE_MAT = FALSE
-  )
-
-  SEARCH2 = list(
-    TOP_N = 12,
-    NBPF = 100,
-    NP = 2500,
-    SPAT_REGRESSION = 0.05,
-    NREPS = 6,
-    NP_EVAL = 2500,
-    NREPS_EVAL = 18,
-    RW_SD = chol_rw_2.sd,
-    COOLING = 0.5,
-    KEEP_TRACES = FALSE,
-    KEEP_LIK_MAT = TRUE
-  )
-}
-
-# Calculate the number of parameters that are estimated:
-# This is calculated by the number of non-zero rw_sd values (including expressions), but then subtracting of repeated shared parameters.
-mod3_n_params <- sum(reg_rw_1.sd != 0, na.rm = TRUE) + sum(sapply(reg_rw_1.sd, is.expression)) - length(shared_param_names) * 9
-
-# Fit Model 3
-haiti3_fit <- readRDS("model3/run_level_3/haiti3_fit.rds")
-# haiti3_fit <- bake(
-#   file = paste0("model3/", rl_dir, "haiti3_fit.rds"), {
-#     fit_haiti3(
-#       nsearches = 2L,
-#       search1 = SEARCH1,
-#       search2 = SEARCH2,
-#       search_rho = FALSE,
-#       search_gamma = FALSE,
-#       search_Iinit = TRUE,
-#       search_hur = TRUE
-#     )
-#   }
-# )
-
-# haiti3_fit contains ll estimate for many parameter sets, so we want to find
-# the highest value
-mod3_ll <- max(
-  haiti3_fit[[length(haiti3_fit)]]$logLiks$logLik,
-  na.rm = TRUE
-)
-
-# Just a bit of tidying up: removing parameters that we won't use again.
-rm(
-  SEARCH1, SEARCH2, N_SEARCHES,
-  est_param_names, est_param_names_expanded,
-  set_rw_1, set_rw_2, set_rw_3, reg_rw_1.sd, reg_rw_2.sd, reg_rw_3.sd,
-  chol_rw_1.sd, chol_rw_2.sd, chol_rw_3.sd
-)
-gc()
-
-@
-
-<<createModelObjects,echo=FALSE,eval=TRUE, include=FALSE, message=FALSE>>=
-
-# Find the index for parameters of model 3 that correspond to the MLE.
-best_m3 <- which.max(haiti3_fit[[length(haiti3_fit)]]$logLiks$logLik)
-
-# Save the parameters that correspond to the MLE of Model 3.
-p3 <- haiti3_fit[[length(haiti3_fit)]]$params[best_m3, ]
 
@@
 
-<<AICtableInput, echo=FALSE, eval=TRUE>>=
+<<Get AIC table Input, echo=FALSE, eval=TRUE>>=
 NREPS_EVAL <- switch(RUN_LEVEL,   3,   8,    36)
-NP_EVAL    <- switch(RUN_LEVEL, 100, 500,  5000)
+NP_EVAL    <- switch(RUN_LEVEL,  50, 500,  5000)
 
@@
 # Calculate likelihood via block-particle filter
-mod3_eval_res <- readRDS("model3/run_level_3/mod3_eval_res.rds")
-# mod3_eval_res <- bake(
-#   file = paste0('model3/', rl_dir, 'mod3_eval_res.rds'), {
-#
-#     # Create a list to store results
-#     results <- list()
-#
-#     coef(h3Spat) <- p3
-#
-#     # Perform block-particle filter
-#     h3_bpf <- foreach(i = 1:NREPS_EVAL, .combine = c) %dopar% {
-#       bpfilter(h3Spat, Np = NP_EVAL, block_size = 1)
-#     }
-#
-#     ll <- logLik(h3_bpf)
-#     results$ll <- logmeanexp(ll[!is.na(ll)], se = FALSE)
-#
-#     ### Get likelihood for subset of the data
-#
-#     # Find appropriate sub-set
-#     in_subset_cols <- h3_bpf[[1]]@times >= as.numeric(lubridate::decimal_date(as.Date("2014-03-01")))
-#     in_subset_rows <- h3_bpf[[1]]@unit_names != 'Ouest'
-#
-#     mod3_coupled_subset_evals <- c()
-#     for (i in 1:length(h3_bpf)) {
-#       mod3_coupled_subset_evals <- c(
-#         mod3_coupled_subset_evals,
-#         sum(h3_bpf[[i]]@block.cond.loglik[in_subset_rows, in_subset_cols])
-#       )
-#     }
-#
-#     results$subset_ll <- logmeanexp(mod3_coupled_subset_evals)
-#     results
-#   },
-#   timing = FALSE
-# )
+mod3_eval_res <- bake(
+  file = paste0('model3/', rl_dir, 'mod3_eval_res.rds'), {
+
+    # Create a list to store results
+    results <- list()
+
+    coef(h3Spat) <- p3
+
+    # Perform block-particle filter
+    h3_bpf <- foreach(i = 1:NREPS_EVAL, .combine = c) %dopar% {
+      bpfilter(h3Spat, Np = NP_EVAL, block_size = 1)
+    }
+
+    ll <- logLik(h3_bpf)
+    results$ll <- logmeanexp(ll[!is.na(ll)], se = FALSE)
+
+    ### Get likelihood for subset of the data
+
+    # Find appropriate sub-set
+    in_subset_cols <- h3_bpf[[1]]@times >= as.numeric(lubridate::decimal_date(as.Date("2014-03-01")))
+    in_subset_rows <- h3_bpf[[1]]@unit_names != 'Ouest'
+
+    mod3_coupled_subset_evals <- c()
+    for (i in 1:length(h3_bpf)) {
+      mod3_coupled_subset_evals <- c(
+        mod3_coupled_subset_evals,
+        sum(h3_bpf[[i]]@block.cond.loglik[in_subset_rows, in_subset_cols])
+      )
+    }
+
+    results$subset_ll <- logmeanexp(mod3_coupled_subset_evals)
+    results
+  },
+  timing = FALSE
+)
 
@@
 
-<<CalculateBenchLikelihood, echo=FALSE, message=FALSE, include=FALSE, cache=TRUE>>=
+<<Calculate Bench Likelihood, echo=FALSE, message=FALSE, include=FALSE, cache=TRUE>>=
 
@@
 benchmarks[['m3']] <- m3_benchmark
-rm(cases, m3_benchmark, m3_bench, dep, mod3_cases)
+rm(cases, m3_benchmark, dep, mod3_cases)
 gc()
@@
 
-<<AIC Lee et al, echo=FALSE, include=FALSE, message=FALSE>>=
+<<Computing AIC of Lee et al, echo=FALSE, include=FALSE, message=FALSE>>=
 lee3_NP   <- switch(RUN_LEVEL, 50, 250, 5000)
@@
 lee3_sub <- lee3_spatPomp()
-
 lee3_all <- lee3_spatPomp(start_date = "2010-10-23")
 
-lee3_results <- readRDS("model3/run_level_3/lee3_results.rds")
-# lee3_results <- bake(
-#   file = paste0('model3/', rl_dir, 'lee3_results.rds'), {
-#
-#     # Create lists to save results
-#     lee3_results <- list()
-#
-#     # Save how long it takes for sub-data
-#     lee3_results[["sub_bpf_time"]] <- system.time({
-#
-#       lee3_sub_ll_evals <- foreach(
-#         i = 1:lee3_NREP, .combine = rbind
-#       ) %dopar% {
-#
-#         liks <- numeric(10L)
-#
-#         lee_bpf_out <- bpfilter(lee3_sub, block_size = 1, Np = lee3_NP)
-#         Ouest_ind <- which(lee_bpf_out@unit_names == "Ouest")
-#
-#         for (u in 1:10) {
-#           if (u == Ouest_ind) {
-#             liks[u] <- sum(lee_bpf_out@block.cond.loglik[u, lee3_sub@times >= lubridate::decimal_date(as.Date('2017-06-10'))])
-#           } else {
-#             liks[u] <- sum(lee_bpf_out@block.cond.loglik[u, ])
-#           }
-#         }
-#
-#         liks
-#       }
-#     })
-#
-#     # Save all likelihood evaluations
-#     lee3_results[["sub_ll_eval"]] <- logmeanexp(rowSums(lee3_sub_ll_evals)[!is.na(rowSums(lee3_sub_ll_evals))])
-#
-#     registerDoRNG(2789131)
-#
-#     # Save how long it takes for all data
-#     lee3_results[["all_pf_time"]] <- system.time({
-#       # Save all likelihood evaluations
-#
-#       lee3_all_ll_evals <- foreach(
-#         i = 1:lee3_NREP, .combine = c
-#       ) %dopar% {
-#         logLik(bpfilter(lee3_all, block_size = 1, Np = lee3_NP))  # Get likelihood evaluations
-#       }
-#
-#       lee3_results[["all_ll_eval"]] <- logmeanexp(lee3_all_ll_evals[!is.na(lee3_all_ll_evals)])
-#
-#     })
-#
-#     lee3_results[["n_params"]] <- 29
-#
-#     lee3_results
-#   },
-#   timing = FALSE
-# )
+lee3_results <- bake(
+  file = paste0('model3/', rl_dir, 'lee3_results.rds'), {
+
+    # Create lists to save results
+    lee3_results <- list()
+
+    # Save how long it takes for sub-data
+    lee3_results[["sub_bpf_time"]] <- system.time({
+
+      lee3_sub_ll_evals <- foreach(
+        i = 1:lee3_NREP, .combine = rbind
+      ) %dopar% {
+
+        liks <- numeric(10L)
+
+        lee_bpf_out <- bpfilter(lee3_sub, block_size = 1, Np = lee3_NP)
+        Ouest_ind <- which(lee_bpf_out@unit_names == "Ouest")
+
+        for (u in 1:10) {
+          if (u == Ouest_ind) {
+            liks[u] <- sum(lee_bpf_out@block.cond.loglik[u, lee3_sub@times >= lubridate::decimal_date(as.Date('2017-06-10'))])
+          } else {
+            liks[u] <- sum(lee_bpf_out@block.cond.loglik[u, ])
+          }
+        }
+
+        liks
+      }
+    })
+
+    # Save all likelihood evaluations
+    lee3_results[["sub_ll_eval"]] <- logmeanexp(rowSums(lee3_sub_ll_evals)[!is.na(rowSums(lee3_sub_ll_evals))])
+
+    registerDoRNG(2789131)
+
+    # Save how long it takes for all data
+    lee3_results[["all_pf_time"]] <- system.time({
+      # Save all likelihood evaluations
+
+      lee3_all_ll_evals <- foreach(
+        i = 1:lee3_NREP, .combine = c
+      ) %dopar% {
+        logLik(bpfilter(lee3_all, block_size = 1, Np = lee3_NP))  # Get likelihood evaluations
+      }
+
+      lee3_results[["all_ll_eval"]] <- logmeanexp(lee3_all_ll_evals[!is.na(lee3_all_ll_evals)])
+
+    })
+
+    lee3_results[["n_params"]] <- 29
+
+    lee3_results
+  },
+  timing = FALSE
+)
 
@@
 mod3_lee_coupled_aic <- 2 * lee3_results$n_params - 2 * lee3_results$all_ll_eval
-
 mod3_lee_coupled_partial_aic <- 2 * lee3_results$n_params - 2 * lee3_results$sub_ll_eval
@@
 
-<<cleanUpBetatFig, echo=FALSE, message=FALSE, include=FALSE, warning=FALSE>>=
+<<clean Up of Betat data, echo=FALSE, message=FALSE, include=FALSE, warning=FALSE>>=
 suppressWarnings(
@@
 # Get quantiles of simulations from the calibrated Model 3.
-h3_quants <- readRDS("model3/run_level_3/h3Sims.rds")
-# h3_quants <- bake(
-#   file = paste0('model3/', rl_dir, 'h3Sims.rds'), {
-#
-#     coef(h3Spat) <- p3
-#     # Simulate from the fitted version of Model 3
-#     h3_sims <- simulate(
-#       h3Spat, nsim = h3_nsim, format = 'data.frame'
-#     )
-#
-#     # For each department and time-point, calculate percentiles of the
-#     # simulations
-#     h3_sims %>%
-#       rename(dep = unitname) %>%
-#       group_by(dep, time) %>%
-#       summarise(
-#         q05 = quantile(cases, 0.025, na.rm = T),
-#         mean = mean(cases, na.rm = T),
-#         q50 = quantile(cases, 0.5, na.rm = T),
-#         q95 = quantile(cases, 0.975, na.rm = T)
-#       ) %>%
-#       ungroup() %>%
-#       mutate(
-#         date = lubridate::date_decimal(time)
-#       ) %>%
-#       mutate(date = as.Date(lubridate::round_date(date, unit = 'day')))
-#   },
-#   timing = FALSE
-# )
-
+h3_quants <- bake(
+  file = paste0('model3/', rl_dir, 'h3Sims.rds'), {
+
+    coef(h3Spat) <- p3
+    # Simulate from the fitted version of Model 3
+    h3_sims <- simulate(
+      h3Spat, nsim = h3_nsim, format = 'data.frame'
+    )
+
+    # For each department and time-point, calculate percentiles of the
+    # simulations
+    h3_sims %>%
+      rename(dep = unitname) %>%
+      group_by(dep, time) %>%
+      summarise(
+        q05 = quantile(cases, 0.025, na.rm = T),
+        mean = mean(cases, na.rm = T),
+        q50 = quantile(cases, 0.5, na.rm = T),
+        q95 = quantile(cases, 0.975, na.rm = T)
+      ) %>%
+      ungroup() %>%
+      mutate(
+        date = lubridate::date_decimal(time)
+      ) %>%
+      mutate(date = as.Date(lubridate::round_date(date, unit = 'day')))
+  },
+  timing = FALSE
+)
 @
@@
 
-<<Load Model 3 Vacc Scenarios, message=FALSE, include=FALSE>>=
-NP_BPF <- switch(RUN_LEVEL, 50, 200, 2000)
-NSIM   <- switch(RUN_LEVEL, 10,  50, 1000)
-
-h3_bpf_end_stats <- readRDS("model3/run_level_3/h3_bpf_end_states.rds")
-# h3_bpf_end_states <- bake(
-#   file = paste0("model3/", rl_dir, "h3_bpf_end_states.rds"),
-#   expr = {
-#     coef(h3Spat) <- p3
-#     h3_bpf <- bpfilter(h3Spat, Np = NP_BPF, block_size = 1, save_states = TRUE)
-#     ss <- h3_bpf@saved.states
-#     ss[[length(ss)]]
-#   },
-#   timing = FALSE
-# )
-
-mod3_V0_res <- readRDS("model3/run_level_3/mod3_V0_res.rds")
-# mod3_V0_res <- bake(
-#   file = paste0("model3/", rl_dir, "mod3_V0_res.rds"), {
-#
-#     # Create a list to store results
-#     h3_V0_res <- list()
-#
-#     # Get "noVacc" scenario parameters
-#     h3_S0_par <- get_model3_vacc_scenario_params(scenario = "noVacc")
-#     coef(h3Spat) <- p3
-#     coef(h3Spat)[names(h3_S0_par)] <- h3_S0_par
-#
-#     # Using parameters and filtered distribution, project system
-#     h3_V0_sims <- project_from_filter(
-#       h3Spat, end_states = h3_bpf_end_states, covarGen = project_rain,
-#       nsims = NSIM, seed = 6897154
-#     )
-#
-#     # Store nationally aggregated results
-#     h3_V0_res$mod3_V0_sims <- agg_mod3_sims(h3_V0_sims)
-#     h3_V0_res$mod3_V0_probs <- get_elimProbs(h3_V0_sims, model = 3)
-#
-#     # For the supplement material, save quantiles of the Susceptible individuals
-#     h3_V0_res$SQuants <- h3_V0_sims %>%
-#       mutate(
-#         date = as.Date(lubridate::round_date(lubridate::date_decimal(time), unit = 'day'))
-#       ) %>%
-#       group_by(unitname, date) %>%
-#       mutate(
-#         pop = h3Spat@params[paste0("H", which(first(unitname) == unit_names(h3Spat)))]
-#       ) %>%
-#       summarize(
-#         Q025 = quantile((S + VSd + VSdd + VSdd_alt + VSd_alt) / pop, probs = 0.025),
-#         Q50 = quantile((S + VSd + VSdd + VSdd_alt + VSd_alt) / pop, probs = 0.5),
-#         Q975 = quantile((S + VSd + VSdd + VSdd_alt + VSd_alt) / pop, probs = 0.975)
-#       ) %>%
-#       ungroup() %>%
-#       rename(dep = unitname)
-#
-#     # Return results
-#     h3_V0_res
-#
-#   }
-# )
-
-mod3_V1_res <- readRDS("model3/run_level_3/mod3_V1_res.rds")
-# mod3_V1_res <- bake(
-#   file = paste0("model3/", rl_dir, "mod3_V1_res.rds"), {
-#
-#     # Create a list to store results
-#     h3_V1_res <- list()
-#
-#     # Get "noVacc" scenario parameters
-#     h3_V1_par <- get_model3_vacc_scenario_params(scenario = "2dep")
-#     coef(h3Spat) <- p3
-#     coef(h3Spat)[names(h3_V1_par)] <- h3_V1_par
-#
-#     # Using parameters and filtered distribution, project system
-#     h3_V1_sims <- project_from_filter(
-#       h3Spat, end_states = h3_bpf_end_states, covarGen = project_rain,
-#       nsims = NSIM, seed = 6897154
-#     )
-#
-#     # Store nationally aggregated results
-#     h3_V1_res$mod3_V1_sims <- agg_mod3_sims(h3_V1_sims)
-#     h3_V1_res$mod3_V1_probs <- get_elimProbs(h3_V1_sims, model = 3)
-#
-#     # Return results
-#     h3_V1_res
-#
-#   }
-# )
-
-mod3_V2_res <- readRDS("model3/run_level_3/mod3_V2_res.rds")
-# mod3_V2_res <- bake(
-#   file = paste0("model3/", rl_dir, "mod3_V2_res.rds"), {
-#
-#     # Create a list to store results
-#     h3_V2_res <- list()
-#
-#     # Get "noVacc" scenario parameters
-#     h3_V2_par <- get_model3_vacc_scenario_params(scenario = "3dep")
-#     coef(h3Spat) <- p3
-#     coef(h3Spat)[names(h3_V2_par)] <- h3_V2_par
-#
-#     # Using parameters and filtered distribution, project system
-#     h3_V2_sims <- project_from_filter(
-#       h3Spat, end_states = h3_bpf_end_states, covarGen = project_rain,
-#       nsims = NSIM, seed = 6897154
-#     )
-#
-#     # Store nationally aggregated results
-#     h3_V2_res$mod3_V2_sims <- agg_mod3_sims(h3_V2_sims)
-#     h3_V2_res$mod3_V2_probs <- get_elimProbs(h3_V2_sims, model = 3)
-#
-#     # Return results
-#     h3_V2_res
-#   }
-# )
-
-mod3_V3_res <- readRDS("model3/run_level_3/mod3_V3_res.rds")
-# mod3_V3_res <- bake(
-#   file = paste0("model3/", rl_dir, "mod3_V3_res.rds"), {
-#
-#     # Create a list to store results
-#     h3_V3_res <- list()
-#
-#     # Get "noVacc" scenario parameters
-#     h3_V3_par <- get_model3_vacc_scenario_params(scenario = "slowNation")
-#     coef(h3Spat) <- p3
-#     coef(h3Spat)[names(h3_V3_par)] <- h3_V3_par
-#
-#     # Using parameters and filtered distribution, project system
-#     h3_V3_sims <- project_from_filter(
-#       h3Spat, end_states = h3_bpf_end_states, covarGen = project_rain,
-#       nsims = NSIM, seed = 6897154
-#     )
-#
-#     # Store nationally aggregated results
-#     h3_V3_res$mod3_V3_sims <- agg_mod3_sims(h3_V3_sims)
-#     h3_V3_res$mod3_V3_probs <- get_elimProbs(h3_V3_sims, model = 3)
-#
-#     # Return results
-#     h3_V3_res
-#
-#   }
-# )
-
-mod3_V4_res <- readRDS("model3/run_level_3/mod3_V4_res.rds")
-# mod3_V4_res <- bake(
-#   file = paste0("model3/", rl_dir, "mod3_V4_res.rds"), {
-#
-#     # Create a list to store results
-#     h3_V4_res <- list()
-#
-#     # Get "noVacc" scenario parameters
-#     h3_V4_par <- get_model3_vacc_scenario_params(scenario = "fastNation")
-#     coef(h3Spat) <- p3
-#     coef(h3Spat)[names(h3_V4_par)] <- h3_V4_par
-#
-#     # Using parameters and filtered distribution, project system
-#     h3_V4_sims <- project_from_filter(
-#       h3Spat, end_states = h3_bpf_end_states, covarGen = project_rain,
-#       nsims = NSIM, seed = 6897154
-#     )
-#
-#     # Store nationally aggregated results
-#     h3_V4_res$mod3_V4_sims <- agg_mod3_sims(h3_V4_sims)
-#     h3_V4_res$mod3_V4_probs <- get_elimProbs(h3_V4_sims, model = 3)
-#
-#     # Return results
-#     h3_V4_res
-#
-#   }
-# )
+<<Simulate Model 3 Vacc Scenarios, message=FALSE, include=FALSE>>=
+NP_BPF <- switch(RUN_LEVEL, 20, 200, 2000)
+NSIM   <- switch(RUN_LEVEL, 5,  50, 1000)
+
+h3_bpf_end_states <- bake(
+  file = paste0("model3/", rl_dir, "h3_bpf_end_states.rds"),
+  expr = {
+    coef(h3Spat) <- p3
+    h3_bpf <- bpfilter(h3Spat, Np = NP_BPF, block_size = 1, save_states = TRUE)
+    ss <- h3_bpf@saved.states
+    ss[[length(ss)]]
+  },
+  timing = FALSE
+)
+
+mod3_V0_res <- bake(
+  file = paste0("model3/", rl_dir, "mod3_V0_res.rds"), {
+
+    # Create a list to store results
+    h3_V0_res <- list()
+
+    # Get "noVacc" scenario parameters
+    h3_S0_par <- get_model3_vacc_scenario_params(scenario = "noVacc")
+    coef(h3Spat) <- p3
+    coef(h3Spat)[names(h3_S0_par)] <- h3_S0_par
+
+    # Using parameters and filtered distribution, project system
+    h3_V0_sims <- project_from_filter(
+      h3Spat, end_states = h3_bpf_end_states, covarGen = project_rain,
+      nsims = NSIM, seed = 6897154
+    )
+
+    # Store nationally aggregated results
+    h3_V0_res$mod3_V0_sims <- agg_mod3_sims(h3_V0_sims)
+    h3_V0_res$mod3_V0_probs <- get_elimProbs(h3_V0_sims, model = 3)
+
+    # For the supplement material, save quantiles of the Susceptible individuals
+    h3_V0_res$SQuants <- h3_V0_sims %>%
+      mutate(
+        date = as.Date(lubridate::round_date(lubridate::date_decimal(time), unit = 'day'))
+      ) %>%
+      group_by(unitname, date) %>%
+      mutate(
+        pop = h3Spat@params[paste0("H", which(first(unitname) == unit_names(h3Spat)))]
+      ) %>%
+      summarize(
+        Q025 = quantile((S + VSd + VSdd + VSdd_alt + VSd_alt) / pop, probs = 0.025),
+        Q50 = quantile((S + VSd + VSdd + VSdd_alt + VSd_alt) / pop, probs = 0.5),
+        Q975 = quantile((S + VSd + VSdd + VSdd_alt + VSd_alt) / pop, probs = 0.975)
+      ) %>%
+      ungroup() %>%
+      rename(dep = unitname)
+
+    # Return results
+    h3_V0_res
+  }
+)
+
+mod3_V1_res <- bake(
+  file = paste0("model3/", rl_dir, "mod3_V1_res.rds"), {
+
+    # Create a list to store results
+    h3_V1_res <- list()
+
+    # Get "noVacc" scenario parameters
+    h3_V1_par <- get_model3_vacc_scenario_params(scenario = "2dep")
+    coef(h3Spat) <- p3
+    coef(h3Spat)[names(h3_V1_par)] <- h3_V1_par
+
+    # Using parameters and filtered distribution, project system
+    h3_V1_sims <- project_from_filter(
+      h3Spat, end_states = h3_bpf_end_states, covarGen = project_rain,
+      nsims = NSIM, seed = 6897154
+    )
+
+    # Store nationally aggregated results
+    h3_V1_res$mod3_V1_sims <- agg_mod3_sims(h3_V1_sims)
+    h3_V1_res$mod3_V1_probs <- get_elimProbs(h3_V1_sims, model = 3)
+
+    # Return results
+    h3_V1_res
+  }
+)
+
+mod3_V2_res <- bake(
+  file = paste0("model3/", rl_dir, "mod3_V2_res.rds"), {
+
+    # Create a list to store results
+    h3_V2_res <- list()
+
+    # Get "noVacc" scenario parameters
+    h3_V2_par <- get_model3_vacc_scenario_params(scenario = "3dep")
+    coef(h3Spat) <- p3
+    coef(h3Spat)[names(h3_V2_par)] <- h3_V2_par
+
+    # Using parameters and filtered distribution, project system
+    h3_V2_sims <- project_from_filter(
+      h3Spat, end_states = h3_bpf_end_states, covarGen = project_rain,
+      nsims = NSIM, seed = 6897154
+    )
+
+    # Store nationally aggregated results
+    h3_V2_res$mod3_V2_sims <- agg_mod3_sims(h3_V2_sims)
+    h3_V2_res$mod3_V2_probs <- get_elimProbs(h3_V2_sims, model = 3)
+
+    # Return results
+    h3_V2_res
+  }
+)
+
+mod3_V3_res <- bake(
+  file = paste0("model3/", rl_dir, "mod3_V3_res.rds"), {
+
+    # Create a list to store results
+    h3_V3_res <- list()
+
+    # Get "noVacc" scenario parameters
+    h3_V3_par <- get_model3_vacc_scenario_params(scenario = "slowNation")
+    coef(h3Spat) <- p3
+    coef(h3Spat)[names(h3_V3_par)] <- h3_V3_par
+
+    # Using parameters and filtered distribution, project system
+    h3_V3_sims <- project_from_filter(
+      h3Spat, end_states = h3_bpf_end_states, covarGen = project_rain,
+      nsims = NSIM, seed = 6897154
+    )
+
+    # Store nationally aggregated results
+    h3_V3_res$mod3_V3_sims <- agg_mod3_sims(h3_V3_sims)
+    h3_V3_res$mod3_V3_probs <- get_elimProbs(h3_V3_sims, model = 3)
+
+    # Return results
+    h3_V3_res
+  }
+)
+
+mod3_V4_res <- bake(
+  file = paste0("model3/", rl_dir, "mod3_V4_res.rds"), {
+
+    # Create a list to store results
+    h3_V4_res <- list()
+
+    # Get "noVacc" scenario parameters
+    h3_V4_par <- get_model3_vacc_scenario_params(scenario = "fastNation")
+    coef(h3Spat) <- p3
+    coef(h3Spat)[names(h3_V4_par)] <- h3_V4_par
+
+    # Using parameters and filtered distribution, project system
+    h3_V4_sims <- project_from_filter(
+      h3Spat, end_states = h3_bpf_end_states, covarGen = project_rain,
+      nsims = NSIM, seed = 6897154
+    )
+
+    # Store nationally aggregated results
+    h3_V4_res$mod3_V4_sims <- agg_mod3_sims(h3_V4_sims)
+    h3_V4_res$mod3_V4_probs <- get_elimProbs(h3_V4_sims, model = 3)
+
+    # Return results
+    h3_V4_res
+  }
+)
 @

### 2024-01-24 ms.Rnw
Fixed parameter typo

--- 2024-01-23 ms.Rnw
@@
 #                    RUN_LEVEL =  1,   2,    3
-N_PARAMS   <- switch(RUN_LEVEL,   5,  20,  nrow(h1LeeStartsEpi))  # Number of parameters to sample
-N_LEE_SIMS <- switch(RUN_LEVEL,   5,  20,   20)  # Number of simulations per parameter
+N_PARAMS   <- switch(RUN_LEVEL,  10,  20,  nrow(h1LeeStartsEpi))  # Number of parameters to sample
+N_LEE_SIMS <- switch(RUN_LEVEL,  10,  20,   20)  # Number of simulations per parameter
 NP_EVAL    <- switch(RUN_LEVEL,  50, 200, 2000)  # Number of particles used to evaluate the likelihood
-NREPS_EVAL <- switch(RUN_LEVEL,   2,   5,   36)  # Number of replicates of likelihood evaluation.
+NREPS_EVAL <- switch(RUN_LEVEL,   3,   5,   36)  # Number of replicates of likelihood evaluation.
 
@@
 
-if (RUN_LEVEL == 1 && file.exists('model1/run_level_3/lee1_epi_evals.rds')) {
-
-  # If RUN_LEVEL == 1, and RUN_LEVEL == 3 results exist, just load the RL3
-  # results.
+# Compute the run_level_<RUN_LEVEL> results and save them.
 lee_epi_ll <- readRDS("model1/run_level_3/lee1_epi_evals.rds")
-} else {
-
-  # Compute the run_level_<RUN_LEVEL> results and save them, or simply load
-  # the results if the bake load conditions are met.
-  lee_epi_ll <- bake(
-    file = paste0("model1/", rl_dir, "lee1_epi_evals.rds"), {
-
-      # Randomly sample from starting params used by Lee et al (2020).
-      # If RUN_LEVEL == 3, all starting parameters are used.
-      which_params <- sample(
-        1:nrow(h1LeeStartsEpi),
-        size = N_PARAMS,
-        replace = FALSE
-      )
-      epi_params <- h1LeeStartsEpi[which_params, ]
-
-      # Create data.frame to save results
-      lee_epi_ll <- data.frame(
-        "pfLL" = NA_real_,
-        "pfse" = NA_real_,
-        "parid" = epi_params$parid
-      )
-
-      for (j in 1:nrow(lee_epi_ll)) {
-        # Get parameters of interest
-        pf_params <- epi_params[j, -1][names(coef(h1_epi))]
-
-        # make results reproducible
-        registerDoRNG((j * 687383921) %% 7919)
-
-        # Calculate log likelihoods
-        ll_evals <- foreach(i=1:NREPS_EVAL, .combine = c) %dopar% {
-          logLik(pfilter(h1_epi, params = pf_params, Np = NP_EVAL))
-        }
-        lee_epi_ll[j, 'pfLL'] <- logmeanexp(ll_evals, se = TRUE)[1]  # Log-lik
-        lee_epi_ll[j, 'pfse'] <- logmeanexp(ll_evals, se = TRUE)[2]  # SE
-      }
-
-      lee_epi_ll
-    }
-  )
-}
+# lee_epi_ll <- bake(
+#   file = paste0("model1/", rl_dir, "lee1_epi_evals.rds"), {
+#
+#     # Randomly sample from starting params used by Lee et al (2020).
+#     # If RUN_LEVEL == 3, all starting parameters are used.
+#     which_params <- sample(
+#       1:nrow(h1LeeStartsEpi),
+#       size = N_PARAMS,
+#       replace = FALSE
+#     )
+#     epi_params <- h1LeeStartsEpi[which_params, ]
+#
+#     # Create data.frame to save results
+#     lee_epi_ll <- data.frame(
+#       "pfLL" = NA_real_,
+#       "pfse" = NA_real_,
+#       "parid" = epi_params$parid
+#     )
+#
+#     for (j in 1:nrow(lee_epi_ll)) {
+#       # Get parameters of interest
+#       pf_params <- epi_params[j, -1][names(coef(h1_epi))]
+#
+#       # make results reproducible
+#       registerDoRNG((j * 687383921) %% 7919)
+#
+#       # Calculate log likelihoods
+#       ll_evals <- foreach(i=1:NREPS_EVAL, .combine = c) %dopar% {
+#         logLik(pfilter(h1_epi, params = pf_params, Np = NP_EVAL))
+#       }
+#       lee_epi_ll[j, 'pfLL'] <- logmeanexp(ll_evals, se = TRUE)[1]  # Log-lik
+#       lee_epi_ll[j, 'pfse'] <- logmeanexp(ll_evals, se = TRUE)[2]  # SE
+#     }
+#
+#     lee_epi_ll
+#   }
+# )
 
@@
 
-if (RUN_LEVEL == 1 && file.exists('model1/run_level_3/lee1_end_evals.rds')) {
-
-  # If RUN_LEVEL == 1, and RUN_LEVEL == 3 results exist, just load the RL3
-  # results.
 lee_end_ll <- readRDS("model1/run_level_3/lee1_end_evals.rds")
-} else {
-
-  # Compute the run_level_<RUN_LEVEL> results and save them, or simply load
-  # the results if the bake load conditions are met.
-  lee_end_ll <- bake(
-    file = paste0("model1/", rl_dir, "lee1_end_evals.rds"), {
-
-      end_params <- h1LeeStartsEnd[h1LeeStartsEnd$parid %in% lee_epi_ll$parid, ]
-
-      # Create data.frame to save results
-      lee_end_ll <- data.frame(
-        "pfLL" = NA_real_,
-        "pfse" = NA_real_,
-        'parid' = end_params$parid
-      )
-
-      # Perform particle filter to get likelihood estimates
-      for (j in 1:nrow(lee_end_ll)) {
-        # Get parameters of interest
-        pf_params <- end_params[j, ][names(coef(h1_end))]
-
-        # make results reproducible
-        registerDoRNG((j * 687383921) %% 7919)
-
-        # Calculate log likelihoods
-        ll_evals <- foreach(i=1:NREPS_EVAL, .combine = c) %dopar% {
-          logLik(pfilter(h1_end, params = pf_params, Np = NP_EVAL))
-        }
-
-        lee_end_ll[j, 'pfLL'] <- logmeanexp(ll_evals, se = TRUE)[1]
-        lee_end_ll[j, 'pfse'] <- logmeanexp(ll_evals, se = TRUE)[2]
-      }
-
-      lee_end_ll
-    }
-  )
-}
-
-# Combine epidemic and endemic likelihoods by `parid` to get the likelihood of
+# lee_end_ll <- bake(
+#   file = paste0("model1/", rl_dir, "lee1_end_evals.rds"), {
+#
+#     end_params <- h1LeeStartsEnd[h1LeeStartsEnd$parid %in% lee_epi_ll$parid, ]
+#
+#     # Create data.frame to save results
+#     lee_end_ll <- data.frame(
+#       "pfLL" = NA_real_,
+#       "pfse" = NA_real_,
+#       'parid' = end_params$parid
+#     )
+#
+#     # Perform particle filter to get likelihood estimates
+#     for (j in 1:nrow(lee_end_ll)) {
+#       # Get parameters of interest
+#       pf_params <- end_params[j, ][names(coef(h1_end))]
+#
+#       # make results reproducible
+#       registerDoRNG((j * 687383921) %% 7919)
+#
+#       # Calculate log likelihoods
+#       ll_evals <- foreach(i=1:NREPS_EVAL, .combine = c) %dopar% {
+#         logLik(pfilter(h1_end, params = pf_params, Np = NP_EVAL))
+#       }
+#
+#       lee_end_ll[j, 'pfLL'] <- logmeanexp(ll_evals, se = TRUE)[1]
+#       lee_end_ll[j, 'pfse'] <- logmeanexp(ll_evals, se = TRUE)[2]
+#     }
+#
+#     lee_end_ll
+#   }
+# )
+
+# Combine epidemic and endemic likelihoods by parid to get the likelihood of
 # the complete model.
@@
 
-# Doing some clean-up. The goal is that each major chunk can be run independent
-# of other chunks. Removing saved objects is not entirely necessary, but can
-# help new users know what objects are needed later on and which were only
-# temporarily needed for each particular chunk. Here we suppress warnings
-# because some of the temporary objects only exist if the `bake` code needs to
-# be recomputed, and hence `rm` will throw warnings if it tries to remove
-# objects that don't exist.
-suppressWarnings(
 rm(
@@
 )
-)
-gc()  # Garbage Collection to free un-used memory
+gc()
 @
@@
 
-<<LoadModel1Params, echo=FALSE, message=FALSE, include=FALSE>>=
+<<FitModel1, echo=FALSE, message=FALSE, include=FALSE>>=
 
@@
 
-# Load the file containing profiles of Model 1 Parameters.
-if (RUN_LEVEL != 3 && file.exists("model1/run_level_3/h1_profiles.rds")) {
-  h1_prof <- readRDS("model1/run_level_3/h1_profiles.rds")
-} else if (file.exists(paste0('model1', rl_dir, 'h1_profiles.rds'))) {
-  h1_prof <- readRDS(paste0('model1', rl_dir, 'h1_profiles.rds'))
-} else {
-  stop(paste0("Missing file: `model1/", rl_dir, 'h1_profiles.rds`;\nThe file can be obtained by running the script `H1profile.R`'))
-}
+# Check RUN_LEVEL, and set hyper-parameters to IF2 and pfilter accordingly.
+#                      RUN_LEVEL =    1,    2,      3
+NP_H1        <- switch(RUN_LEVEL,   100, 1000,   2000)
+NMIF_H1      <- switch(RUN_LEVEL,    10,   50,    200)
+NUM_TREND    <- switch(RUN_LEVEL,     5,   10,     90)
+NPROF        <- switch(RUN_LEVEL,     4,    4,     24)
+NP_EVAL      <- switch(RUN_LEVEL,   200, 1000,   5000)
+NREPS_EVAL   <- switch(RUN_LEVEL, cores,   20,  cores)
+
+# Create the <RUN_LEVEL> version of the results and save them.
+h1_MIF <- readRDS('model1/run_level_3/haiti1_fit.rds')
+
+# h1_MIF <- bake(
+#   file = paste0("model1/", rl_dir, "haiti1_fit.rds"), {
+#     fit_haiti1(  # haitipkg function used to fit Model 1
+#       NP = NP_H1,
+#       NMIF = NMIF_H1,
+#       NUM_TREND = NUM_TREND,
+#       NPROF = NPROF,
+#       NREPS_EVAL = NREPS_EVAL,
+#       NP_EVAL = NP_EVAL
+#     )
+#   }
+# )
 
@@
 # we calculate the maximum likelihood for each value of betat.
-h1_prof_res <- h1_prof %>%
-  filter(prof_var == 'betat') %>%
+h1_prof_res <- h1_MIF %>%
   group_by(betat) %>%
-  summarize(logLik = max(logLik))
-
-# TODO: This is used later in figure 2 (TODO)
+  summarize(logLik = max(ll))
+
 mcap_results <- pomp::mcap(h1_prof_res$logLik, h1_prof_res$betat, span = 1)
 
-# Find the set of parameters that correspond to the maximum likelihood
-p1 <- h1_prof %>%
-  slice_max(order_by = logLik, n = 1) %>%
-  # Remove non-parameters from the data.frame (hyperparameters + likelihood)
-  select(-(1:12)) %>%
+# Get the best set of parameters from MIF2 search.
+p1 <- h1_MIF %>%
+  filter(ll == max(ll)) %>%
+  select(-ll, -ll.se) %>%
   unlist()
@@
 # Save the likelihood of the best parameters, for use later.
-mod1_ll <- h1_prof %>%
-  slice_max(order_by = logLik, n = 1) %>%
-  select(logLik) %>%
+mod1_ll <- h1_MIF %>%
+  filter(ll == max(ll)) %>%
+  select(ll) %>%
   unlist()
@@
 # Just a bit of tidying up: removing parameters that we won't use again.
-rm(h1_prof)
+rm(NP_H1, NMIF_H1, NUM_TREND, NPROF, NREPS_EVAL, NP_EVAL, h1_MIF)
 gc()
+
 @
@@
 <<FitModel2, echo=FALSE, include=FALSE, message=FALSE>>=
-
-# TODO: Model 2
 
@@
 
-<<createModelObjects,echo=FALSE,eval=TRUE, include=FALSE, message=FALSE>>=
+<<table-input,echo=FALSE,eval=T, include=FALSE, message=FALSE>>=
 
@@
   & \multirow{2}{*}{(S15)}
-  & \multirow{2}{*}{$\obsOverdispersion = \Sexpr{signif(h2_fit$h2_params['sigma'], 3)}$}
+  & \multirow{2}{*}{$\obsOverdispersion = \Sexpr{myround(h2_fit$h2_params['v'], 3)}$}
   & \multirow{2}{*}{(S16)}
@@
   geom_line(data = mcap_results$fit, aes(x = parameter, y = smoothed), col = 'blue') +
-  geom_line(data = mcap_results$fit, aes(x = parameter, y = quadratic), col = 'red') +
   geom_vline(xintercept = mcap_results$ci[1], linetype = 'dashed') +
@@
 {\figTitle Confidence interval for linear trend in transmission.}
-Monte Carlo adjusted profile (MCAP) of $\transmissionTrend$ for Model~1.
-The blue curve is the MCAP, the vertical blue line indicates the MLE, and the vertical dashed lines indicate the $95\%$ confidence interval.
+Monte Carlo adjusted profile of $\transmissionTrend$ for Model~1.
+The blue curve is the profile, the blue line indicates the MLE, and the dashed lines indicate the $95\%$ confidence interval.
 }
 \end{figure}
-
-<<cleanUpBetatFig, echo=FALSE, message=FALSE, include=FALSE, warning=FALSE>>=
-suppressWarnings(
-  rm(h1_prof_res, mcap_results)
-)
-gc()
-@
 

### 2024-01-24 H1profile.R
Modified comments at the bottom on how to correctly save results

@@ -325,7 +325,7 @@ submitJobs(
 # h1_results = ijoin(h1_pars, h1_profile_results)
 #
 # saveRDS(
-#   h1_profile_results,
+#   h1_results,
 #   paste0('model1/run_level_', RUN_LEVEL, '/', 'h1_profiles.rds')
 # )
 #

### 2024-01-24 H2profile.R

@@ -259,7 +259,7 @@ submitJobs(
 # h2_results = ijoin(h2_pars, h2_profile_results)
 #
 # saveRDS(
-#   h2_profile_results,
+#   h2_results,
 #   'model2/h2_profiles.rds')
 # )
 #

### 2024-01-24 H3profile.R

@@ -383,7 +383,7 @@ submitJobs(resources = resources1)
 # h3_results = ijoin(h3_pars, h3_profile_results)
 #
 # saveRDS(
-#   h3_profile_results,
+#   h3_results,
 #   paste0('model3/run_level_', RUN_LEVEL, '/', 'h3_profiles.rds')
 # )
 #

### 2024-01-24 ms.Rnw
minor edits to the manuscript

--- 2024-01-23 ms.Rnw
@@
     ll <- logLik(h3_bpf)
-    results$ll <- logmeanexp(ll[!is.na(ll)], se = FALSE)
+    results$ll <- logmeanexp(ll[!is.na(ll)], se = TRUE)
 
@@
 
-    results$subset_ll <- logmeanexp(mod3_coupled_subset_evals)
+    results$subset_ll <- logmeanexp(mod3_coupled_subset_evals, se = TRUE)
     results
@@
 
-mod3_eval_res$subset_aic <- 2 * mod3_n_params - 2 * mod3_eval_res$subset_ll
+mod3_eval_res$subset_aic <- 2 * mod3_n_params - 2 * mod3_eval_res$subset_ll['se']
 
@@
 On this subset, their model has a likelihood of $\Sexpr{myround(lee3_results$sub_ll_eval, 1)}$.
-On this same subset, our model has a likelihood of $\Sexpr{myround(mod3_eval_res$subset_ll, 1)}$.
+On this same subset, our model has a likelihood of $\Sexpr{myround(mod3_eval_res$subset_ll['est'], 1)}$.
 Details of estimating the likelihood of the Lee et al.~\cite{lee20} models is provided in \nameref{S_lee20}
@@
 
-<<Model 3 SpatPOMP sims, echo=FALSE>>=
+<<Model 3 SpatPOMP Sims, echo=FALSE>>=
 

### 2024-01-24 ms.Rnw
Fixed broken file path

@@
   h1_prof <- readRDS("model1/run_level_3/h1_profiles.rds")
-} else if (file.exists(paste0('model1', rl_dir, 'h1_profiles.rds'))) {
-  h1_prof <- readRDS(paste0('model1', rl_dir, 'h1_profiles.rds'))
+} else if (file.exists(paste0('model1/', rl_dir, 'h1_profiles.rds'))) {
+  h1_prof <- readRDS(paste0('model1/', rl_dir, 'h1_profiles.rds'))
 } else {
@@
   h3_prof <- readRDS("model3/run_level_3/h3_profiles.rds")
-} else if (file.exists(paste0('model3', rl_dir, 'h3_profiles.rds'))) {
+} else if (file.exists(paste0('model3/', rl_dir, 'h3_profiles.rds'))) {
 
   # If the `model3/run_level_<RUN_LEVEL>/h3_profiles.rds` file exists, load it
-  h3_prof <- readRDS(paste0('model3', rl_dir, 'h3_profiles.rds'))
+  h3_prof <- readRDS(paste0('model3/', rl_dir, 'h3_profiles.rds'))
 } else {

### 2024-01-25 ms.Rnw
Knitted file at RL3 with new code
Recording all diffs in RNW file

@@
 This necessitates addressing the tradeoff between biological fidelity and model simplicity, and the reality of misspecification for models at all levels of complexity.
-\ei{I'VE RE-WRITTEN HERE, LOOKING TO BE MORE SPECIFIC ABOUT THE GOALS AND CONTRIBUTION OF THE PAPER.}
-We assess current methodological approaches to these issues via a case study of the 2010-2019 cholera epidemic in Haiti.
+\eic{I'VE RE-WRITTEN HERE, LOOKING TO BE MORE SPECIFIC ABOUT THE GOALS AND CONTRIBUTION OF THE PAPER.}
+\old{As a case study, we consider the 2010-2019 cholera epidemic in Haiti.
+We study three dynamic models developed by expert teams to advise on vaccination policies.
+We assess previous methods used for fitting and evaluating these models, and we develop data analysis strategies leading to improved statistical fit.}
+\new{We assess current methodological approaches to these issues via a case study of the 2010-2019 cholera epidemic in Haiti.
 We consider three dynamic models developed by expert teams to advise on vaccination policies.
-We evaluate previous methods used for fitting these models, and we demonstrate modified data analysis strategies leading to improved statistical fit.
+We evaluate previous methods used for fitting these models, and we demonstrate modified data analysis strategies leading to improved statistical fit.}
 Specifically, we present approaches to diagnosis of model misspecification, development of alternative models, and computational improvements in optimization, in the context of likelihood-based inference on nonlinear dynamic systems.
@@
 
-This case study investigates the use of dynamic models and spatiotemporal data to inform public health policy in the context of the cholera outbreak in Haiti, which started in 2010.
-Various dynamic models were developed to study this outbreak: searching PubMed with keywords ``Haiti, cholera, model'' we obtained 20 deterministic models \cite{kirpich16,kirpich17,andrews11,botelho21,fitzgibbon20,eisenberg13,rinaldo12,chao11,date11,lin19,collins21,akman15,trevisin22,mavian20,collins14,kelly16,capone15,bengtsson15,mari15,gatto12} and 8 stochastic models \cite{tuite11,mukandavire13,lewnard16,kunkel17,sallah17,azman12,kuhn14,azman15}.
-Here we focus on a multi-group modeling exercise by Lee et al.~\cite{lee20} in which four expert modeling teams developed models to the same dataset with the goal of comparing conclusions on the feasibility of eliminating cholera by a vaccination campaign.
+This case study investigates the use of dynamic models and spatiotemporal data to inform \old{a policy decision} \new{public health policy} in the context of the cholera outbreak in Haiti, which started in 2010.
+\old{We build on a multi-group modeling exercise by Lee et al.
+[1] in which four expert modeling teams developed models to the same dataset with the goal of comparing conclusions on the feasibility of eliminating cholera by a vaccination campaign}.
+\old{The data that were used for this study, and that is the subject of our reanalysis, is displayed in Fig.~[1].}
+\new{Various dynamic models were developed to study this outbreak: searching PubMed with keywords ``Haiti, cholera, model'' we obtained 20 deterministic models \cite{kirpich16,kirpich17,andrews11,botelho21,fitzgibbon20,eisenberg13,rinaldo12,chao11,date11,lin19,collins21,akman15,trevisin22,mavian20,collins14,kelly16,capone15,bengtsson15,mari15,gatto12} and 8 stochastic models \cite{tuite11,mukandavire13,lewnard16,kunkel17,sallah17,azman12,kuhn14,azman15}.
+Here we focus on a multi-group modeling exercise by Lee et al.~\cite{lee20} in which four expert modeling teams developed models to the same dataset with the goal of comparing conclusions on the feasibility of eliminating cholera by a vaccination campaign.}
 Model~1 is stochastic and describes cholera at the national level;
@@
 Model~4 has an agent-based construction, featuring considerable mechanistic detail but limited ability to calibrate these details to data.
-These modeling strategies were selected by \cite{li20} to represent the range of approaches used in the research community.
+\new{These modeling strategies were selected by \cite{li20} to represent the range of approaches used in the research community.}
 We focus on Models~1--3, as the strengths and weaknesses of the agent-based modeling approach \cite{tracy18} are outside the scope of this article.
 \eic{In addition, agent-based models were less widely used [NO OTHERS SHOW UP IN OUR LITERATURE REVIEW???].}
-The data that were used for this study, and that is the subject of our reanalysis, are displayed in Fig.~\ref{fig:CholeraData}.
+\new{The data that were used for this study, and that is the subject of our reanalysis, are displayed in Fig.~\ref{fig:CholeraData}.}
 
@@
 
-%% IT SEEMS THIS PARAGRAPH IS DEPRECATED IN FAVOR OF CITING THESE REFERENCES WEHRE APPROPRIAT
-%% While this work is focused on the models proposed by Lee et al.~\cite{lee20}, our suggestions have broader relevance.
-%% To investigate the extent to which \cite{lee20} is typical of the substantial body of work on Haiti cholera dynamics, we performed a literature review by searching PubMed with keywords: Haiti, cholera, model.
-%% The search resulted in 66 papers, of which 32 used dynamic models to describe the cholera epidemic in Haiti.
-%% The models make various choices on the dichotomies considered by Lee et al.~\cite{lee20}; deterministic versus stochastic; compartment model versus agent-based; aggregated versus spatially explicit.
+%% IT SEEMS THIS PARAGRAPH IS DEPRECATED IN FAVOR OF CITING THESE REFERENCES WHERE APPROPRIATE
+\old{While this work is focused on the models proposed by Lee et al.~[1], our suggestions have broader relevance.
+To investigate the extent to which [1] is typical of the substantial body of work on Haiti cholera dynamics, we performed a literature review by searching PubMed with keywords: Haiti, cholera, model.
+The search resulted in 66 papers, of which 32 used dynamic models to describe the cholera epidemic in Haiti.
+The models make various choices on the dichotomies considered by Lee et al.~[1]; deterministic versus stochastic; compartment model versus agent-based; aggregated versus spatially explicit.}
 
@@
 
-In this article we follow \cite{lee20} by calibrating the parameters of each of our models using maximizing the likelihood, as described in Eq.~\myeqref{eq:likedef}.
+In this article we \old{calibrate} \new{follow \cite{lee20} by calibrating} the parameters of each of our models \old{by maximizing the likelihood} \new{using maximum likelihood}, as described in Eq.~\myeqref{eq:likedef}.
 The likelihood for each of the fitted models---and the corresponding AIC values for model comparisons that include an adjustment for the number of calibrated parameters---is provided in Table~\ref{tab:likes}.
@@
 \eic{I'VE STARTED TO PUT IN SOME DISCUSSION OF BAYESIAN METHODS HERE, THOUGH IT COULD ALSO GO IN THE DISCUSSION}
-The main alternative to maximum likelihood estimation is Bayesian inference via Markov chain Monte Carlo, used to analyze the Haiti cholera epidemic by \cite{andrews11,rinaldo12,lewnard16,trevisin22,sallah17,azman12,kuhn14,mari15,gatto12}.
-Alternatively, some investigations obtained parameters from previous sources without calibrating to data.
+\new{The main alternative to maximum likelihood estimation is Bayesian inference via Markov chain Monte Carlo, used to analyze the Haiti cholera epidemic by \cite{andrews11,rinaldo12,lewnard16,trevisin22,sallah17,azman12,kuhn14,mari15,gatto12}.
+Alternatively, some investigations obtained parameters from previous sources without calibrating to data.}
 

### 2024-01-25 r1/response.tex
Minor edits + editions to response to reviewers

@@
 
+Yes, that is correct.
+This has been fixed in the revision.
+In other instances where the linear trend was previously mentioned, we simply refer to it now as a trend in transmision rate, unless the additional detail of the trend being log-linear is needed for clarity.
+
 \report{10.
@@
 }
+
+Thank you for this suggestion, a sentence on the overall effect of the parameter estimate has been included.
+For reference, the maxmum likelihood estimate of the parameter corresponds with a 7.3\% reduction to the transmission rate over the course of the outbreak, with a $95\%$ confidence interval $(1.8\%, 17.9\%)$ for the overall reduction in transmission.
 
@@
 For the particular issue of black box versus white box models, we have included the following paragraph
-\TODO{where?}
+\TODO{where?} \jwc{How does the Materials and Methods: Mechanistic models for disease modeling sub-section sound?}
 
@@
 
-We agree that adding to model complexity has costs, including loss of identifability, AIC and other formal statistical methods can assess whether the additional complexity is justified by a sufficiently large improvement of statistical fit.
+We agree that adding to model complexity has costs, including loss of identifability.
+AIC and other formal statistical methods can assess whether the additional complexity is justified by a sufficiently large improvement of statistical fit.
 Identifiability and model simplicity have practical scientific value which goes beyond their statistical role in reducing generalization error and hence improving out-of-fit predictions.
-This is a variation on the issue of "statistical significance versus practical significance" familiar from introductory statistics textbooks.
-HOW DOES THIS PLAY OUT IN THE MS? HOW DO WE CLARIFY IT?
+This is a variation on the issue of ``statistical significance versus practical significance" familiar from introductory statistics textbooks.
+\eic{HOW DOES THIS PLAY OUT IN THE MS? HOW DO WE CLARIFY IT?}
 
@@
 In the context of cholera dynamics, \cite{lemaitre19} found a substantial improvement in model fit by including stochasticity, involving one over-dispersion parameter.
-stocks20 found similar results, favoring a stochastic dynamic model with over-dispersion, when investigating transmission dynamics of a different enteric pathogen, rotavirus.
-In other contexts, king15 noted various scientific advantages for including stochasticity in the model; it is undoubtedly part of the biological dynamics, and its omission can lead to over-confident predictions.
+\cite{stocks20} found similar results, favoring a stochastic dynamic model with over-dispersion, when investigating transmission dynamics of a different enteric pathogen, rotavirus.
+In other contexts, \cite{king15} noted various scientific advantages for including stochasticity in the model; it is undoubtedly part of the biological dynamics, and its omission can lead to over-confident predictions.
 In our study, we find that the deterministic model (model 2) has poor statistical fit.
-This can lead to a risk of identifying spurious scientific phenomena: addition deterministic complexity added to the model may help to describe unmodeled stochastic phenomena present in the data, and may therefore spuriously be assessed as a highly statistically significant model improvement.
+This can lead to a risk of identifying spurious scientific phenomena: additional deterministic complexity added to the model may help to describe unmodeled stochastic phenomena present in the data, and may therefore spuriously be assessed as a highly statistically significant model improvement.
 
@@
 In our literature review, of the 9 previous Haiti cholera investigations including stochastic dynamic models \cite{tuite11,mukandavire13,lewnard16,kunkel17,sallah17,azman12,kuhn14,azman15,lee20}, four of them carried out inference via MCMC \cite{lewnard16,sallah17,azman12,kuhn14}.
-WHAT SORTS OF MCMC ALGORITHMS DID THEY USE?
+\eic{WHAT SORTS OF MCMC ALGORITHMS DID THEY USE?}
 We agree with the referee that MCMC is more routine for determinstic models than for stochastic models, and advanced MCMC methods such as Hamiltonian Monte Carlo become more readily applicable.
@@
 
-We have included some additional sentences to address these comments in the discussion section of the revision.
-TODO
+\TODO{We have included some additional sentences to address these comments in the discussion section of the revision.}
 

### 2024-01-25 edits.tex
Minor edits + editions to response to reviewers

@@
 %%%%%% EDITING MACROS %%%%%%%%%
+
+\newcommand\new[1]{\textcolor{blue}{#1}}
+\newcommand\old[1]{\sout{#1}}
 % orange for EI
@@
 % green for JW
-\definecolor{green}{rgb}{0,0.5,0}
-\newcommand\jw[2]{\sout{#1} \textcolor{green}{#2}}
-\newcommand\jwc[1]{\textcolor{green}{[#1]}}
+% \definecolor{green}{rgb}{0,0.5,0}
+\newcommand\jw[2]{\sout{#1} \textcolor{purple}{#2}}
+\newcommand\jwc[1]{\textcolor{purple}{[#1]}}
 % purple for JJ
@@
 \newcommand\jj[2]{\sout{#1} \textcolor{purple}{#2}}
-\newcommand\jjc[1]{\textcolor{purple}{[#1]}}
+\newcommand\jjc[1]{\textcolor{purple}{[Josh: #1]}}
 % cyan for AR
@@
 
-\newcommand\editMechModels{Mathematical models representing biological phenomena are valuable for epidemiology and consequently for public health policy \cite{lofgren14,mccabe21}.
+\newcommand\editMechModels{Mechanistic models representing biological phenomena are valuable for epidemiology and consequently for public health policy \cite{lofgren14,mccabe21}.
 More broadly, they have useful roles throughout biology, expecially when combined with statistical methods that properly account for stochasticity and nonlinearity \cite{may04}.

### 2024-01-25 ms.Rnw
Minor edits + editions to response to reviewers

@@
 This motivates the development of mechanistic models, which provides a statistical fit to the available data while also supporting a causal interpretation.
+\jwc{This paragraph and the one below seem to go well together, so perhaps this is a natural place to put this new paragraph.}
+
+\new{\editMechModels}
 
@@
 The ability to maximize the likelihood allows for likelihood-based inference, such as performing statistical tests for potential model improvements.
-We demonstrate this capability by proposing a linear trend $\transmissionTrend$ in transmission in Eq.~\myeqref{model1:beta}:
+We demonstrate this capability by proposing a \new{log-}linear trend $\transmissionTrend$ in transmission in Eq.~\myeqref{model1:beta}:
 \begin{equation}
@@
 where $\bar{t} = \frac{t - (t_N + t_0) / 2}{t_N - (t_N + t_0) / 2}$, so that $\bar{t} \in [-1, 1]$.
-The proposal of a linear trend in transmission is a result of observing an apparent decrease in reported cholera infections from 2012-2019 in Fig.~\ref{fig:CholeraData}.
+The proposal of a \old{linear} trend in transmission is a result of observing an apparent decrease in reported cholera infections from 2012-2019 in Fig.~\ref{fig:CholeraData}.
 While several factors may contribute to this decrease, one explanation is that case-area targeted interventions (CATIs), which included education sessions, increased monitoring, household decontamination, soap distribution, and water chlorination in infected areas \cite{rebaudet19CATI}, may have substantially reduced cholera transmission over time \cite{rebaudet21}.
 
-We perform a statistical test to determine whether or not the data indicate the presence of a linear trend in transmissibility.
+We perform a statistical test to determine whether or not the data indicate the presence of a \old{linear} trend in transmissibility.
 To do this, we perform a profile-likelihood search on the parameter $\transmissionTrend$ and obtain a $95\%$ confidence interval via a Monte Carlo Adjusted Profile (MCAP) \cite{ionides17}.
@@
 We similarly allow the re-estimation of process and measurement overdispersion parameters ($\sigmaProc^2$ and $\obsOverdispersion$), and require that the latent Markov process $X(t)$ carry over from one phase into the next.
-The resulting confidence interval for $\transmissionTrend$ is $(\Sexpr{myround(mcap_results$ci[1], digits = 3)}, \Sexpr{myround(mcap_results$ci[2], digits = 3)})$, with the full results displayed in Fig.~\ref{fig:betat}.
+The resulting 95\% confidence interval for $\transmissionTrend$ is $(\Sexpr{myround(mcap_results$ci[1], digits = 3)}, \Sexpr{myround(mcap_results$ci[2], digits = 3)})$, with the full results displayed in Fig.~\ref{fig:betat}.
 These results are suggestive that the inclusion of a trend in the transmission rate improves the quantitative ability of Model~1 to describe the observed data.
+\new{The maximum likelihood estimate for the trend corresponds with a \Sexpr{paste0(myround(100 - 100*exp(mcap_results$mle * 430/215), 1), '\\%')} reduction to the transmission rate over the course of the outbreak, with a $95\%$ confidence interval (\Sexpr{paste0(rev(myround(100 - 100*exp(mcap_results$ci * 430/215), 1)), collapse = ', ', sep = '\\%')}) for the overall reduction in transmission.}.
 The reported results for Model~1 in the remainder of this article were obtained with the inclusion of the parameter $\transmissionTrend$.
@@
   geom_vline(xintercept = mcap_results$mle, col = 'blue') +
-  labs(x = "Linear Trend in Transmission", y = 'Log Likelihood') +
+  labs(x = "Log-Linear Trend in Transmission", y = 'Log Likelihood') +
   theme(axis.text = element_text(size = 8),
@@
 \caption{\label{fig:betat}
-{\figTitle Confidence interval for linear trend in transmission.}
+{\figTitle Confidence interval for \new{the log-}linear trend in transmission.}
 Monte Carlo adjusted profile (MCAP) of $\transmissionTrend$ for Model~1.

### 2024-01-25 ms.Rnw
Added details in discussion

@@
 We evaluate previous methods used for fitting these models, and we demonstrate modified data analysis strategies leading to improved statistical fit.}
-Specifically, we present approaches to diagnosis of model misspecification, development of alternative models, and computational improvements in optimization, in the context of likelihood-based inference on nonlinear dynamic systems.
+Specifically, we present approaches \old{to diagnosis of model misspecification, development of alternative models, and computational improvements in optimization, in the context of likelihood-based inference on nonlinear dynamic systems.} \new{of diagnosing model misspecification, which facilitates the development of model improvements.
+We further demonstrate the usefulness of recent advancements in likelihood optimization for high-dimensional nonlinear dynamic models, resulting in increased concurrence between high dimensional models and observed incidence data.}
 Our workflow is reproducible and extendable, facilitating future investigations of this disease system.
@@
 Evaluation of success at these goals is necessary in order for a model to provide a reliable tool for guiding evidence-based public policy interventions.
-In this article, we discuss common modeling decisions that may not provide an adequate statistical explanation of the data, and we show that attention to statistical fit can lead to more accurate policy evaluations.
+In this article, we discuss common modeling decisions that may not provide an adequate statistical explanation of the data, and we show that attention to statistical fit can lead to more accurate policy \old{evaluations} \new{recommendations}.
 We do this by conducting a re-analysis of the 2010 cholera outbreak in Haiti.
@@
 We contend that raising the currently accepted standards of infectious disease modeling will result in a greater ability of scientists and policy makers to understand and respond to future infectious disease outbreaks.
+\TODO{I think we can and should directly state something here that we discuss with the reviewers: the article has both educational purposes, arguing how a data analysis using dynamic models should happen, and as an in-depth analysis of cholera in Haiti.}
+\new{This analysis serves both as a pedagogical...}
 
@@
 \new{These modeling strategies were selected by \cite{li20} to represent the range of approaches used in the research community.}
+\jwc{I think the citation above is incorrect? Should it be lee20, not li20?}.
 We focus on Models~1--3, as the strengths and weaknesses of the agent-based modeling approach \cite{tracy18} are outside the scope of this article.
-\eic{In addition, agent-based models were less widely used [NO OTHERS SHOW UP IN OUR LITERATURE REVIEW???].}
+\new{In addition, agent-based models were less widely used, the agent based model of \cite{lee20} being the only model of this class that was found in our literature review.}
 \new{The data that were used for this study, and that is the subject of our reanalysis, are displayed in Fig.~\ref{fig:CholeraData}.}
+
+\new{\editMechModels}
 
@@
 The groups largely adhered to existing guidelines for creating models to inform policy \cite{behrend20,saltelli20} and, despite their autonomy, obtained a consensus that an extensive nationwide vaccination campaign would be necessary to eliminate cholera from Haiti.
+\TODO{How?
+This statement should be epanded or re-evaluated.}
 Their forecasts, however, were inconsistent with the prolonged period with no confirmed cholera cases between February, 2019 and September, 2022 \cite{trevisin22}.
@@
 we then describe how each model is calibrated to data, and present a systematic approach to examining and refining these models.
-We then use improved model fits to project cholera incidence in Haiti under various vaccination scenarios considered by Lee et al.~\cite{lee20}.
+\new{Specifically, we focus on how to develop and test variations to these proposed models, as well as diagnosing the models once they have been assimilated to incidence reports.} \jwc{make a statement about fitting Model 3 (stochastic meta-pop) here?}
+We then use \new{the} improved model fits to project cholera incidence in Haiti under various vaccination scenarios considered by Lee et al.~\cite{lee20}.
 Finally, we conclude with a discussion of the results.
@@
 This motivates the development of mechanistic models, which provides a statistical fit to the available data while also supporting a causal interpretation.
-\jwc{This paragraph and the one below seem to go well together, so perhaps this is a natural place to put this new paragraph.}
-
-\new{\editMechModels}
+\jwc{Maybe right here is another place for the new paragraph on mechanistic models?}
 
@@
 These results are suggestive that the inclusion of a trend in the transmission rate improves the quantitative ability of Model~1 to describe the observed data.
-\new{The maximum likelihood estimate for the trend corresponds with a \Sexpr{paste0(myround(100 - 100*exp(mcap_results$mle * 430/215), 1), '\\%')} reduction to the transmission rate over the course of the outbreak, with a $95\%$ confidence interval (\Sexpr{paste0(rev(myround(100 - 100*exp(mcap_results$ci * 430/215), 1)), collapse = ', ', sep = '\\%')}) for the overall reduction in transmission.}.
+\new{The maximum likelihood estimate for $\transmissionTrend$ corresponds with a \Sexpr{paste0(myround(100 - 100*exp(mcap_results$mle * 430/215), 1), '\\%')} reduction to the transmission rate over the course of the outbreak, with a $95\%$ confidence interval (\Sexpr{paste0(rev(myround(100 - 100*exp(mcap_results$ci * 430/215), 1)), collapse = ', ', sep = '\\%')}) for the overall reduction in transmission.}
 The reported results for Model~1 in the remainder of this article were obtained with the inclusion of the parameter $\transmissionTrend$.
+
+\new{This example demonstrates the type of model variation that should be considered in the course of a data analysis: the model variation---the addition of a trend in transmission rate---has a plausible scientific justification, and is easily testable using likelihood based inference methods and observed data.
+This class of model variations are commonly considered in epidemiological modeling \cite{lemaitre19,fitzgibbon20,eisenberg13,rinaldo12,kunkel17}} , \new{but their adaptation are sometimes based on subjective assessments \cite{fitzgibbon20,eisenberg13,kunkel17}} \TODO{get citations}.
+\jwc{Alternatively, we could emphasize the importance of considering model variations, and cite studies from the literature review that only considered a single model, or only reported their final chosen model.
+Although models that were not ultimately selected are not as important, a discussion of which models you tried and why you didn't use them can be beneficial to future researchers who may want to consider similar variations in the future.
+I got to article 35 checking for those with multiple models considered.}
 
@@
 The statistical evidence of a trend in transmission rate in this model could be explained by any trending variable (such as hygiene improvements, or changes in population behavior), resulting in confounding from collinear covariates.
-Alternatively, the trend could be attributed to a decreasing reporting rate rather than decreasing transmission rate, resulting in confounded mechanisms.
-The robust statistical conclusion is that a model which allows for change fits better than one which does not---we argue that a decreasing transmission rate is a plausible way to explain this, but the incidence data themselves do not provide enough information to pin down the mechanism.
+\jwc{HERE'S SOME INITIAL THOUGHTS ON HOW TO CLARIFY THIS FOR THE REVIEWER}.
+\old{Alternatively, the trend could be attributed to a decreasing reporting rate rather than decreasing transmission rate, resulting in confounded mechanisms.}
+\new{Because we only have access to incidence data, the decrease in observed cases over the course of the outbreak may alternatively be attributed to a trend in other model parameters, such as a decrease in the reporting rate over time.
+This results in confounded mechanisms, as likelihood-based comparisons of a model that permits a change in reporting rate against a model that permits a change in transmission rate may not yield definitive results.}
+The robust statistical conclusion is that a model which allows for change fits better than one which does not\old{---we argue that a decreasing transmission rate is a plausible way to explain this}, but the incidence data themselves \old{do} \new{may} not provide enough information to pin down the mechanism.
+\new{In this case study, we argue that a decreasing transmission rate is a plausible way to explain the decrease in cases over time \cite{rebaudet19CATI,rebaudet21,michel19}; there is little scientific evidence to support a decrease in reporting rate, as the rate of testing increased throughout this time period.} \TODO{cite: Wasn't there a study we thought demonstrated the testing effort increased}?
 
@@
 Simulations from the fitted model are displayed in Fig.~\ref{fig:h3spatsims}.
+\jwc{Should there be more added / said to this paragraph, to demonstrate the novelty of being able to fit a model like Model 3?
+Maybe add a paragraph between the first and second paragraphs of this sub-section.}
 
@@
 
+\jwc{Should we make benchmark modeling a sub-section of model-diagnostics?
+I think this could (A) add some organization to the paper so that readers / reviewers can clearly see how benchmarks help answer the question we posed: what indications should we look for in data to assess trustworthiness of data (B) further emphasize the importance of benchmarks, which is something the reviewers seemed to have found useful.
+}
 One common approach to assess a mechanistic model's goodness-of-fit is to compare simulations from the fitted model to the observed data.
@@
 
-The ongoing global COVID-19 pandemic has provided a clear example on how government policy may be affected by the conclusions of scientific models \cite{saltelli20}.
-This article demonstrates that fitting appropriate scientific models to guide policy is a challenging statistical task.
+The ongoing global COVID-19 pandemic has provided a clear example on how government policy may be affected by the conclusions of \old{scientific} \new{mathematical} models \cite{saltelli20}.
+\old{This article demonstrates that fitting appropriate scientific models to guide policy is a challenging statistical task.}
+\new{Despite their importance, creating realistic, data-informed models that can be used to reliably and accurately influence policy is a challenging task.
+In this article, we demonstrate that increasing the standards of mathematical modeling of ecological system can improve our ability to use models to understand the dynamics of a complex, non-linear system, and make data informed inferences on such systems.
+We suggest a greater effort be placed on evaluating dynamic models that will be used to make inference on a system.
+Specifically, we argue that researchers can and should look for indications in available data in order to assess whether model based inferences are trustworthy.
+Here, we suggest that this can be done by comparing fitted models to a suitable statistical benchmark, and verifying that model outcomes---including estimated model parameters, latent states, and seasonal trends---make sense in the context of the system in question.}
+
+\new{In the context of our cholera case study, we demonstrate the importance of model diagnostics by using them to improve model fits, resulting in more reliable conclusions.
+We compared the Models 1--3 of \cite{lee20} to simple statistical benchmarks, which enabled us to find deficiencies in the models that could be improved.
+In the case of Model 3, the process of comparing a fitted model to a benchmark and investigating why the model did not describe the data as well as a simple linear benchmark}
 In our case study, we found that additional attention to statistical details could have resulted in improved statistical fits to the observed incidence data, leading to improvements in the accuracy of the resulting policy guidance.

### 2024-01-26 ms.Rnw
Finished first draft of discussion revision.

@@
 \old{This article demonstrates that fitting appropriate scientific models to guide policy is a challenging statistical task.}
-\new{Despite their importance, creating realistic, data-informed models that can be used to reliably and accurately influence policy is a challenging task.
-In this article, we demonstrate that increasing the standards of mathematical modeling of ecological system can improve our ability to use models to understand the dynamics of a complex, non-linear system, and make data informed inferences on such systems.
+\new{Despite their importance, creating scientifically plausible, data-informed models that can be used to reliably gain insight into a dynamic system is a challenging task.
+In this article, we demonstrate that increasing the standards used for the mathematical modeling of ecological systems can improve our ability to use models to understand the dynamics of a complex, non-linear system, and make data informed inferences.
 We suggest a greater effort be placed on evaluating dynamic models that will be used to make inference on a system.
 Specifically, we argue that researchers can and should look for indications in available data in order to assess whether model based inferences are trustworthy.
-Here, we suggest that this can be done by comparing fitted models to a suitable statistical benchmark, and verifying that model outcomes---including estimated model parameters, latent states, and seasonal trends---make sense in the context of the system in question.}
-
-\new{In the context of our cholera case study, we demonstrate the importance of model diagnostics by using them to improve model fits, resulting in more reliable conclusions.
+Here, we show that this can be done by comparing fitted models to a suitable statistical benchmark, and verifying that model outcomes---including estimated model parameters, latent states, and seasonal trends---make sense in the context of the system in question.}
+
+\new{In the cholera case study, we demonstrate the importance of model diagnostics by using them to improve model fits, resulting in more reliable conclusions.
 We compared the Models 1--3 of \cite{lee20} to simple statistical benchmarks, which enabled us to find deficiencies in the models that could be improved.
-In the case of Model 3, the process of comparing a fitted model to a benchmark and investigating why the model did not describe the data as well as a simple linear benchmark}
+In the case of Model~3, the process of comparing a fitted model to a benchmark and investigating why the model did not describe the data as well as its benchmark helped us recognize that the model could not adequately explain the increase in cholera transmission that resulted from Hurricane Matthew, leading us to propose a model alteration.
+Because Model~1 outperformed its benchmark model, we were able to confidently inspect implications from the fitted model; here we found that the seasonality term of Model~1 closely follows seasonal rainfall patterns in Haiti, both confirming previously held beliefs about the importance of rainfall as a driver of cholera infection, as well as bolstering our confidence that Model~1 captures important dynamics of the cholera outbreak.}
+
+\new{In any study using dynamic models, more than one model framework---or model variation---should be considered, tested, and reported.
+In our literature review of modeling cholera in Haiti, only a few papers that used mechanistic models reported more than a single model variation. (TODO: Cite).
+We recognize the possibility that a study that reports only a single model may have considered others and reported only the model that they used for the final analysis; details about models that were tried but not used can be of great benefit to the scientific community, however, as they may help future researchers to avoid frameworks that have previously been shown to be less favorable.}
+
+\new{While likelihood based inference enables researchers to chose between any model that is evaluated on the same dataset using a metric like AIC, nested model variations are particularly useful as they enable formal statistical testing of the nested features.
+In the cholera case study, we demonstrate the benefit of this type of model variation in each of the three models.
+In Model~1, we tested the inclusion of a trend in transmission rate over time; in Model~2, we considered adding an additional phase parameter, which enables a shift in the seasonal peaks of cholera infections; in Model~3, we included additional parameters such as the Hurricane Matthew related parameters.
+In each of these cases, we found strong statistical evidence that favored these model variations over their more simple counterparts.}
+
+\new{Another important consideration when fitting a mechanistic model to a dynamic system is the complexity of the model.
+Mathematical models necessarily simplify complex dynamical systems in order to make inference possible.
+Decisions about which aspects of a system to include in a model and which to exclude can have significant impacts on model based conclusions, but can be tested under a likelihood-based framework.
+Common modeling decisions include deterministic vs stochastic modeling, and whether or not to include a meta-population structure.
+Model benchmarks can be useful in determining the level of modeling complexity to use, allowing researchers to calibrate the relative performance of a model of any given complexity against a better understood statistical model.}
+
+\new{The three models in this case study each represent a unique combination of decisions of model complexity.
+Model~3 provides an interesting example in that it is both stochastic and has a meta-population structure.
+This class of models is notoriously difficult to perform likelihood based inference on (TODO: Cite).
+Here we demonstrate how this class of model can be calibrated to incidence data using the novel IBPF algorithm.
+This case study demonstrates the potential benefit of this algorithm, as it is one of the examples that uses this algorithm.
+}
+
 In our case study, we found that additional attention to statistical details could have resulted in improved statistical fits to the observed incidence data, leading to improvements in the accuracy of the resulting policy guidance.
 We used the same data and models, and even much of the same code, as Lee et al.~\cite{lee20}, and yet ended up with drastically different conclusions.
+\new{Specifically, each of the re-calibrated models predict that cholera would disappear from Haiti with moderate probably.
+While new cases of cholera have since been detected in Haiti, this conclusion is more consistent with the prolonged absence of cholera cases that occurred from 2019-2022.}
 We acknowledge the benefit of hindsight: our demonstration of a statistically principled route to obtain better-fitting models resulting in more robust insights does not rule out the possibility of discovering other models that fit well yet predict poorly.

### 2024-01-26 ms.Rnw
Made some edits to discussion section

--- 2024-01-25 ms.Rnw
@@
 
-The ongoing global COVID-19 pandemic has provided a clear example on how government policy may be affected by the conclusions of \old{scientific} \new{mathematical} models \cite{saltelli20}.
+The ongoing global COVID-19 pandemic has \old{provided a clear example on} \new{demonstrated} how government policy may be affected by the \old{conclusions of scientific models} \new{inferences drawn from mathematical modeling} \cite{saltelli20}.
 \old{This article demonstrates that fitting appropriate scientific models to guide policy is a challenging statistical task.}
-\new{Despite their importance, creating realistic, data-informed models that can be used to reliably and accurately influence policy is a challenging task.
-In this article, we demonstrate that increasing the standards of mathematical modeling of ecological system can improve our ability to use models to understand the dynamics of a complex, non-linear system, and make data informed inferences on such systems.
-We suggest a greater effort be placed on evaluating dynamic models that will be used to make inference on a system.
-Specifically, we argue that researchers can and should look for indications in available data in order to assess whether model based inferences are trustworthy.
-Here, we suggest that this can be done by comparing fitted models to a suitable statistical benchmark, and verifying that model outcomes---including estimated model parameters, latent states, and seasonal trends---make sense in the context of the system in question.}
-
-\new{In the context of our cholera case study, we demonstrate the importance of model diagnostics by using them to improve model fits, resulting in more reliable conclusions.
-We compared the Models 1--3 of \cite{lee20} to simple statistical benchmarks, which enabled us to find deficiencies in the models that could be improved.
-In the case of Model 3, the process of comparing a fitted model to a benchmark and investigating why the model did not describe the data as well as a simple linear benchmark}
-In our case study, we found that additional attention to statistical details could have resulted in improved statistical fits to the observed incidence data, leading to improvements in the accuracy of the resulting policy guidance.
-We used the same data and models, and even much of the same code, as Lee et al.~\cite{lee20}, and yet ended up with drastically different conclusions.
+\new{However, developing credible models---which are supported by data and can lend insights into a dynamic system---remains a challenging task.
+In this article, we demonstrate that increasing the current standards of mathematical modeling of ecological systems can increase our capacity to make sense of a complex, nonlinear system's dynamics and make data-driven inferences.
+We suggest devoting more effort to assess dynamic models intended for system inference.
+Explicitly, we urge researchers to search for indications in available data that quantify the reliability of model-based inferences.
+We illustrate that this can be achieved by comparing fitted models to a suitable statistical benchmark and verifying that model outcomes---including estimated parameters, latent states, and seasonal trends---are justified in the system context.}
+
+\new{In our cholera case study, we highlight the importance of model diagnostics; they are instrumental in enhancing model fits, thus leading to more trustworthy conclusions.
+We compared the Models 1--3 of \cite{lee20} to simple statistical benchmarks, enabling us to identify deficiencies in the models that could be improved.
+In the case of Model~3, comparing a fitted model to a benchmark and then following up by questioning why the model did not describe the data as efficiently as its benchmark helped us discern that the model fell short in explaining the surge in cholera transmission following Hurricane Matthew.
+This led us to suggest a model modification.
+Because Model~1 outperformed its benchmark model, we were able to confidently scrutinize the implications of the fitted model.
+We found that the seasonality term of Model~1 closely mirrors the seasonal rainfall patterns in Haiti, reaffirming alternative evidence of the importance of rainfall as a driver of cholera infection and bolstering our confidence in Model~1's ability to capture the critical dynamics of the cholera outbreak.
+}
+
+\new{In any study involving dynamic models, it is important to consider, test, and present more than one model framework or variant.
+In our review of the literature featuring cholera modeling in Haiti, we found that only a handful of papers that utilized mechanistic models reported more than a single model variation. (TODO: Cite).
+We acknowledged the possibility that a study reporting merely a single model may have explored others and only reported the one used in the final analysis.
+However, divulging details about the models that were considered, but not used, may greatly benefit the broader scientific community.
+These details may guide future researchers to avoid model variations that have previously demonstrated to perform poorly on incidence data.
+Furthermore, it can enhance confidence in the reported model's ability to describe the system under investigation.
+This confidence stems from the demonstrated evidence that the chosen model offers a better description of the observed data compared to other reasonable model alternatives.}
+
+\new{Likelihood based inference enables researchers to choose between any model that is evaluated on the same dataset using a metric like AIC.
+Nested model variations are particularly useful as they enable formal statistical testing of the nested features.
+The cholera case study demonstrated the advantage of such a model variation in all three models.
+For instance, we tested the inclusion of a trend in transmission rate over time in Model~1; in Model~2, we considered adding an additional phase parameter, which enables a shift in the seasonal peaks of cholera infections; in Model~3, we included additional parameters such as those associated with Hurricane Matthew.
+In all these cases, we found strong statistical evidence in favor of these model variations over their simpler counterparts.}
+
+\new{Another important consideration when fitting a mechanistic model to a dynamic system is the complexity of the model.
+Mathematical models necessarily simplify complex dynamical systems in order to make inference possible.
+Decision-making about which features of a system need to be incorporated in a model and which should be excluded can significantly influence the model-based conclusions.
+However, these decisions can be tested using a likelihood-based framework.
+Standard model decisions cover deterministic versus stochastic modeling and the decision to include or exclude a meta-population structure.
+Model benchmarks can assist in determining the level of complexity to be used in the model, enabling researchers to gauge a model's relative performance against a better-understood statistical model.}
+
+\new{The three models in this case study offer a unique combination of decisions of model complexity.
+Model~3 provides an interesting example in that it is both stochastic and has a meta-population structure, making it challenging to draw likelihood-based inferences.
+In this paper, we demonstrated how this model class could be calibrated to incidence data using the innovative IBPF algorithm.
+One of only a few examples of fitting a model with this level of complexity, this case study exemplifies the algorithm's potential benefits and provides an example for future researchers on a possible approach to fitting a high-dimensional non-linear model.}
+
+In our case study, we found that additional attention to statistical details could have resulted in \old{improved} \new{an enhanced} statistical fit to the observed incidence data\old{, leading to improvements in the accuracy of the resulting policy guidance.} \new{This would have improved the accuracy of the policy guidance resulting from the study.}
+\old{We used the same data and models, and even much of the same code, as Lee et al.~[29], and yet ended up with drastically different conclusions.}
+\new{We used the same data, models, and much of the same code used by Lee et al.~\cite{lee20}, but we arrived at drastically different conclusions.}
+\new{Specifically, each of the re-calibrated models predicted with moderate probability that cholera would disappear from Haiti.
+Although there have been new cases of cholera in Haiti, this conclusion aligns more with the prolonged absence of cholera cases from 2019-2022.}
 We acknowledge the benefit of hindsight: our demonstration of a statistically principled route to obtain better-fitting models resulting in more robust insights does not rule out the possibility of discovering other models that fit well yet predict poorly.
 
-Inference for mechanistic time series models offers opportunities for understanding and controlling complex dynamic systems.
+\old{Inference for mechanistic time series} \new{Mechanistic} models offer opportunities for understanding and controlling complex dynamic systems.
 This case study has investigated issues requiring attention when applying powerful new statistical techniques that can enable statistically efficient inference for a general class of partially observed Markov process models.
-Researchers should check that the computationally intensive numerical calculations are carried out adequately.
-Comparison against benchmarks and alternative model specifications should be considered to evaluate the statistical goodness-of-fit.
-Once that is accomplished, care is required to assess what causal conclusions can properly be inferred given the possibility of alternative explanations consistent with the data.
+\old{Researchers should check that the computationally intensive numerical calculations are carried out adequately.}
+\new{Researchers should ensure that intensive numerical calculations are adequately executed.}
+\old{Comparison against benchmarks and alternative model specifications should be considered to evaluate the statistical goodness-of-fit.}
+\new{Using benchmarks and alternative model specifications to assess statistical goodness-of-fit should also should be common practice.}
+Once \old{that is accomplished} \new{a model has been adequately callibrated to data}, care is required to assess what causal conclusions can properly be inferred given the possibility of alternative explanations consistent with the data.
 Studies that combine model development with thoughtful data analysis, supported by a high standard of reproducibility, build knowledge about the system under investigation.

### 2024-01-27 H3profile-r.R
Extra script to do profile for r

@@ -1,0 +1,399 @@
+# Fitting profiles for Haiti 3.
+#
+# ********
+# NOTE:
+#
+#   If you would like to run this code but have never used the `batchtools` R
+#   package, then we strongly recommend completing both Examples 1 and 2 from
+#   the batchtools vignette, by first installing the `batchtools` package and
+#   running: `vignette('batchtools')`, or visiting:
+#
+#         https://mllg.github.io/batchtools/articles/batchtools.html
+#
+# WARNING:
+#
+#   This script was ran using a high-performance computing cluster. In total,
+#   the RUN_LEVEL == 3 version of this script took TODO compute hours. Partial
+#   results can be obtained by lowering the required computational effort
+#   (params: NP, NBPF, nprof, NP_EVAL, NREPS_EVAL), which can be done by
+#   reducing the RUN_LEVEL to 2 or 1, or alternatively by manually changing
+#   these values.
+# ********
+#
+# The fit_haiti3 function is a very simple helper that will take as input
+# some starting parameters, and hyperparameters for the IBPF algorithm, and
+# will return parameter estimates along with corresponding estimates of their
+# log-likelihoods.
+#
+# In this script we will use this function to do a profile likelihood search
+# of all model parameters using the batchtools package. To do this, we will
+# create a grid of parameter values, and submit jobs where each job will use
+# a row from the grid as a starting point to fit the model.
+#
+# The number of rows in the grid (which is the `final_pars` object that is
+# created using a for-loop over the profile parameters) correspond to the total
+# number of jobs that will be submitted using `batchtools`. Each job takes
+# roughly the same amount of time to compute, and hence the run-time complexity
+# is linear with respect to the total number of jobs. The total number of jobs
+# can be controlled by either (1) changing the number of variables to profile
[…]

### 2024-01-30 ms.Rnw
a few edits, mostly the abstract

@@
 This necessitates addressing the tradeoff between biological fidelity and model simplicity, and the reality of misspecification for models at all levels of complexity.
-\eic{I'VE RE-WRITTEN HERE, LOOKING TO BE MORE SPECIFIC ABOUT THE GOALS AND CONTRIBUTION OF THE PAPER.}
 \old{As a case study, we consider the 2010-2019 cholera epidemic in Haiti.
@@
 We evaluate previous methods used for fitting these models, and we demonstrate modified data analysis strategies leading to improved statistical fit.}
-Specifically, we present approaches \old{to diagnosis of model misspecification, development of alternative models, and computational improvements in optimization, in the context of likelihood-based inference on nonlinear dynamic systems.} \new{of diagnosing model misspecification, which facilitates the development of model improvements.
-We further demonstrate the usefulness of recent advancements in likelihood optimization for high-dimensional nonlinear dynamic models, resulting in increased concurrence between high dimensional models and observed incidence data.}
+Specifically, we present approaches to diagnosis of model misspecification \old{, development of alternative models, and computational improvements in optimization, in the context of likelihood-based inference on nonlinear dynamic systems.} \new{and the consequent development of improved models.
+We further demonstrate the utility of recent advances in likelihood maximization for high-dimensional nonlinear dynamic models, enabling likelihood-based inference for high-dimensional models of spatiotemporal incidence data.}
 Our workflow is reproducible and extendable, facilitating future investigations of this disease system.
@@
 \new{In any study involving dynamic models, it is important to consider, test, and present more than one model framework or variant.
+\eic{I'M NOT SURE THIS IS A CONCLUSION.
+LEE ET AL STUDIED MULTIPLE MODELS BECAUSE THAT WAS THE GOAL OF THAT PAPER.
+MOST OTHER PAPERS STUDY ONE MODEL.
+WE WOULD LIKE RESULTS FROM DIFFERENT PAPERS TO HAVE OBJECTIVE MEASURES OF FIT THAT CAN BE COMPARED BETWEEN PAPERS, BUT I DON'T THINK WE WANT TO SAY THAT ALL PAPERS HAVE TO ANALYZE MANY MODELS.}
 In our review of the literature featuring cholera modeling in Haiti, we found that only a handful of papers that utilized mechanistic models reported more than a single model variation. (TODO: Cite).

### 2024-01-31 r1/response.tex
Minor edits addressing reviewer concerns

@@
 
+While Fig 2 really is a table, here we are following PLOS Computational Biology Submission guidelines (\url{https://journals.plos.org/ploscompbiol/s/tables}):
+\begin{quote}
+If the table has a very complex structure or contains graphics, the safest solution is to make it into a figure. Export the table as a TIFF, and cite and re-label it as a figure.
+\end{quote}
+The Table in this case is necessarily complex, as we are trying to put all comparable model parameters from three seperate models into a single table.
+If the journal requests that we make this into a table rather than a figure we are happy to comply.
+
 \report{4.
@@
 }
+
+Thank you for this suggestion, we have made this change.
 
@@
 Our views align with the referee's comments, and we have rewritten the article to make this clearer.
-
-WE COULD ADD THE SUBMITTED PDF TO THE REVISION REPO SO WE CAN CHECK LINE NUMBERS.
 

### 2024-01-31 ms.Rnw
Minor edits addressing reviewer concerns

--- 2024-01-26 ms.Rnw
@@
 Specifically, we present approaches \old{to diagnosis of model misspecification, development of alternative models, and computational improvements in optimization, in the context of likelihood-based inference on nonlinear dynamic systems.} \new{of diagnosing model misspecification, which facilitates the development of model improvements.
-We further demonstrate the usefulness of recent advancements in likelihood optimization for high-dimensional nonlinear dynamic models, resulting in increased concurrence between high dimensional models and observed incidence data.}
-Our workflow is reproducible and extendable, facilitating future investigations of this disease system.
+We further demonstrate the usefulness of recent advancements in likelihood optimization for high-dimensional nonlinear dynamic models, which enable improved consistency between high dimensional models and observed incidence data.}
+\old{Our workflow is} \new{The presented case study is} reproducible and extendable, facilitating future investigations of this disease system.
 
@@
 Evaluation of success at these goals is necessary in order for a model to provide a reliable tool for guiding evidence-based public policy interventions.
-In this article, we discuss common modeling decisions that may not provide an adequate statistical explanation of the data, and we show that attention to statistical fit can lead to more accurate policy \old{evaluations} \new{recommendations}.
+\old{In this article, we discuss common modeling decisions that may not provide an adequate statistical explanation of the data, and we show that attention to statistical fit can lead to more accurate policy evaluations}.
+\new{In this article, we use an in-depth case study to motivate general procedures for fitting mathematical models to time series data, while explaining how these procedures help to address limitations of previous approaches.}
 We do this by conducting a re-analysis of the 2010 cholera outbreak in Haiti.
@@
 We contend that raising the currently accepted standards of infectious disease modeling will result in a greater ability of scientists and policy makers to understand and respond to future infectious disease outbreaks.
-\TODO{I think we can and should directly state something here that we discuss with the reviewers: the article has both educational purposes, arguing how a data analysis using dynamic models should happen, and as an in-depth analysis of cholera in Haiti.}
-\new{This analysis serves both as a pedagogical...}
 
@@
 \label{model1:EI}
-\mu_{E_{\vaccCounter}I_{\vaccCounter}} &=& \muEI\big(1-\vaccineEfficacy_\vaccCounter(t)\big),
+\mu_{E_{\vaccCounter}I_{\vaccCounter}} &=& \muEI\big(1-\symptomFrac_\vaccCounter(t)\big),
 \\
 \label{model1:EA}
-\mu_{E_{\vaccCounter}A_{\vaccCounter}} &=& \muEI\, \vaccineEfficacy_\vaccCounter(t),
+\mu_{E_{\vaccCounter}A_{\vaccCounter}} &=& \muEI\, \symptomFrac_\vaccCounter(t),
 \\
@@
 In Eqs.~\myeqref{model1:EA} and~\myeqref{model1:EI} the asymptomatic ratio for non-vaccinated individuals is set $\symptomFrac_0(t)=0$, so that the asymptomatic route is reserved for vaccinated individuals.
-For $\vaccCounter\in\seq{1}{\vaccClass}$, the vaccination cohort $\vaccCounter$ is assigned a time $\tau_{\vaccCounter}$, and we take $\vaccineEfficacy_\vaccCounter(t) = c \, \vaccineEfficacy^*(t-\tau_{\vaccCounter})$
+For $\vaccCounter\in\seq{1}{\vaccClass}$, the vaccination cohort $\vaccCounter$ is assigned a time $\tau_{\vaccCounter}$, and we take $\symptomFrac_\vaccCounter(t) = c \, \vaccineEfficacy^*(t-\tau_{\vaccCounter})$
 where $\vaccineEfficacy^*(t)$ is efficacy at time $t$ since vaccination for adults, taken from \cite{lee20}, Table~S4, and $c=\big(1-(1-0.4688)\times 0.11\big)$ is a correction to allow for reduced efficacy in the 11\% of the population aged under 5 years.
@@
 We chose to use the auto regressive negative binomial model, however, because the large number of weeks with zero recorded cholera cases in department level data makes a benchmark based on a continuous distribution problematic.
-Likelihoods of Models~1--3 and their respective ARMA benchmark models are provided in Table~\ref{tab:likes}.
+Likelihoods of Models~1--3 and their respective \old{ARMA} benchmark models are provided in Table~\ref{tab:likes}.
 

### 2024-01-31 H3profile-r.R

@@ -374,7 +374,7 @@ submitJobs(resources = resources1)
 # library(batchtools)
 #
 # RUN_LEVEL = 3  # Or adjust RUN_LEVEL as needed.
-# reg = loadRegistry(paste0('model3/profileReg_RL', RUN_LEVEL))
+# reg = loadRegistry(paste0('model3/profileReg_r_RL', RUN_LEVEL))
 # getStatus()
 #
 # Once all of the jobs are finished, we can summarize the results using:

### 2024-01-31 ms.Rnw

@@
 We evaluate previous methods used for fitting these models, and we demonstrate modified data analysis strategies leading to improved statistical fit.}
-Specifically, we present approaches \old{to diagnosis of model misspecification, development of alternative models, and computational improvements in optimization, in the context of likelihood-based inference on nonlinear dynamic systems.} \new{of diagnosing model misspecification, which facilitates the development of model improvements.
-We further demonstrate the usefulness of recent advancements in likelihood optimization for high-dimensional nonlinear dynamic models, which enable improved consistency between high dimensional models and observed incidence data.}
-\old{Our workflow is} \new{The presented case study is} reproducible and extendable, facilitating future investigations of this disease system.
+Specifically, we present approaches to diagnosis of model misspecification \old{, development of alternative models, and computational improvements in optimization, in the context of likelihood-based inference on nonlinear dynamic systems.} \new{and the consequent development of improved models.
+We further demonstrate the utility of recent advances in likelihood maximization for high-dimensional nonlinear dynamic models, enabling likelihood-based inference for high-dimensional models of spatiotemporal incidence data.}
+Our workflow is reproducible and extendable, facilitating future investigations of this disease system.
 

### 2024-01-31 H3profile-r.R
Changed all profile parameters to just 'r', not doing all computations again

@@ -89,7 +89,7 @@ shared_param_names <- c(
   "epsilon", "k"
 )
 
-prof_params <- shared_param_names
+prof_params <- 'r'
 
 best_pars <- final_pars
 prof_vars <- c()

### 2024-02-01 ms.Rnw
Added some TODO statements in the article itself

@@
 We evaluate previous methods used for fitting these models, and we demonstrate modified data analysis strategies leading to improved statistical fit.}
-Specifically, we present approaches to diagnosis of model misspecification \old{, development of alternative models, and computational improvements in optimization, in the context of likelihood-based inference on nonlinear dynamic systems.} \new{and the consequent development of improved models.
+Specifically, we present approaches of diagnosing model misspecification \old{, development of alternative models, and computational improvements in optimization, in the context of likelihood-based inference on nonlinear dynamic systems.} \new{and the consequent development of improved models.
 We further demonstrate the utility of recent advances in likelihood maximization for high-dimensional nonlinear dynamic models, enabling likelihood-based inference for high-dimensional models of spatiotemporal incidence data.}
@@
 These comparisons, which provide insight into how each model performs relative to the others, are quite common \cite{rinaldo12,sallah17}.
+\TODO{Add citations from lit-review.}
 To calibrate relative measures of fit, it is useful to compare against a model that has well-understood statistical ability to fit data, and we call this model a {\it benchmark}.
@@
 This is consistent with previous studies finding that rainfall played an important role in cholera transmission in Haiti \cite{lemaitre19,eisenberg13}.
+\TODO{Citations from lit-review}.
 The estimated seasonality also features an increased transmission rate during the fall, which was noticed at an earlier stage of the epidemic \cite{rinaldo12}; the high transmission rate in the fall may also be a result of the increase transmission that occurred in the fall of 2016, when hurricane Matthew struck Haiti \cite{ferreirai16}.
@@
 
+\TODO{Rethink this paragraph.}
 Our case study provides an example of this in the parameter estimate for the duration of natural immunity due to cholera infection, $\muRS^{-1}$.
@@
 
+\TODO{Re-phrase this paragraph, as suggested by one of the reviewers.}
 Recent information about a dynamic system should be more relevant for a forecast than older information.
@@
 That may explain why Lee et al.~\cite{lee20} chose to obtain forecasts by simulating the calibrated models forward from initial conditions.
-We obtain forecasts by simulating future values using latent states that are consistent with the most recent observations; this is done by drawing latent states at the last observation time ($t_N$) from the filtering distribution $f_{\bm{X}_N|\bm{Y}_{1:N}}(\bm{x}_{N} | \bm{y}^*_{1:N} ; \hat\paramVec)$.
+\new{For Models~1 and 3---which are non-deterministic---}we obtain forecasts by simulating future values using latent states that are consistent with the most recent observations; this is done by drawing latent states at the last observation time ($t_N$) from the filtering distribution $f_{\bm{X}_N|\bm{Y}_{1:N}}(\bm{x}_{N} | \bm{y}^*_{1:N} ; \hat\paramVec)$.
 The decision not to do this partially explains the unsuccessful forecasts of Lee et al.~\cite{lee20}: their Table~S7 shows that the subset of their simulations which were consistent with observing zero cases in 2019 also accurately predicted the prolonged absence of detected cholera.

### 2024-02-06 si/confidenceIntervals.Rnw
Created a script to calculate / display MCAP confidence intervals

\documentclass[11pt]{article}

\input{inputs/header}

\setcounter{tocdepth}{1}

<<packages, include=FALSE>>=
library(tidyverse)
library(pomp)
library(haitipkg)
library(spatPomp)
library(doParallel)
library(doRNG)
library(knitr)
library(latex2exp)

options(
  scipen = 2,
  help_type = "html",
  stringsAsFactors = FALSE,
  continue = "+  ",
  width = 70,
  useFancyQuotes = FALSE,
  reindent.spaces = 2,
  xtable.comment = FALSE
)

opts_chunk$set(
  tidy.opts=list(
    keep.blank.line=FALSE
  ),
  comment="",
  fig.path = 'figure/',
  echo=FALSE,
  fig.align="center",
  dev="pdf",
  dev.args=list(
    bg="transparent",
    pointsize=9
  )
)

if (!dir.exists("../model3/si")) {
  dir.create("../model3/si", recursive = TRUE)
}
@

\begin{document}

%%%%%%%%%% START

\section{Confidence Intervals for Model Parameters}\label{sec:ci}

In this section we provide confidence intervals for all model parameters, excluding those that take unique values for each spatial unit.
For each model and parameter, we use profile log-likelihoods to obtain confidence intervals. TODO: Cite.
Due to the non-linear and stochastic nature of Models~1 and 3, exact evaluation of the profile log-likelihood is unattainable.
Instead, the log-likelihood at each point of the profile is estimated using via Monte-Carlo based particle filter methods.
We therefore obtain confidence intervals for the parameters of Model~1 and Model~3 using the Monte Carlo adjust profile (MCAP) algorithm. TODO: Cite.

\subsection{Model~1 parameters}

<<Load Model 1 Confidence Intervals, include=FALSE, message=FALSE, echo=FALSE>>=
H1_profiles <- readRDS("../model1/run_level_3/h1_profiles.rds")

# TAKEN FROM STACK OVERFLOW:
# https://stackoverflow.com/a/44958156
midcut<-function(x,from,to,by) {
  ## cut the data into bins...
  x=cut(x,seq(from,to,by),include.lowest=T)
  ## make a named vector of the midpoints, names=binnames
  vec=seq(from+by/2,to-by/2,by)
  names(vec)=levels(x)
  ## use the vector to map the names of the bins to the midpoint values
  unname(vec[x])
}

prof_params <- c(
  'betat', 'tau_epi', 'tau_end', 'rho', 'nu', 'sig_sq_epi',
  'sig_sq_end', 'E_0', 'I_0', 'beta1', 'beta2', 'beta3',
  'beta4', 'beta5', 'beta6'
)

H1_profiles_long <- H1_profiles %>%
  select(prof_var, logLik, logLik_se, all_of(prof_params), -E_0, -I_0, -tau_end, -tau_epi, -rho) %>%
  pivot_longer(
    cols = -c(prof_var, logLik, logLik_se),
    names_to = 'variable',
    values_to = 'value'
  ) %>%
  filter(variable == prof_var) %>%
  group_by(prof_var, value) %>%
  slice_max(order_by = logLik, n = 1)

# Tau done on log-scale
H1_long_tau <- H1_profiles %>%
  select(prof_var, logLik, logLik_se, tau_end, tau_epi) %>%
  pivot_longer(
    cols = -c(prof_var, logLik, logLik_se),
    names_to = 'variable',
    values_to = 'value'
  ) %>%
  filter(value <= 1e+12) %>%
  mutate(value = log(value)) %>%
  filter(variable == prof_var) %>%
  group_by(prof_var, value) %>%
  slice_max(order_by = logLik, n = 1)

H1_long_rho <- H1_profiles %>%
  select(prof_var, logLik, logLik_se, rho) %>%
  pivot_longer(
    cols = -c(prof_var, logLik, logLik_se),
    names_to = 'variable',
    values_to = 'value'
  ) %>%
  filter(value > 0.15) %>%
  filter(variable == prof_var) %>%
  group_by(prof_var, value) %>%
  slice_max(order_by = logLik, n = 1)

# E_0, I_0 need to be treated slightly differently
H1_E0_I0_long <- H1_profiles %>%
  select(prof_var, logLik, logLik_se, E_0, I_0) %>%
  pivot_longer(
    cols = -c(prof_var, logLik, logLik_se),
    names_to = 'variable',
    values_to = 'value'
  ) %>%
  filter(variable == prof_var) %>%
  mutate(value = value * 10911819) %>%
  mutate(value = midcut(value, from = -499, to = 24501, by = 1000)) %>%
  # mutate(value = value / 10911819) %>% 
  group_by(prof_var, value) %>%
  slice_max(order_by = logLik, n = 1)

H1_profiles_long <- bind_rows(H1_profiles_long, H1_E0_I0_long, H1_long_tau, H1_long_rho)
@

<<Create H1 Profile CIs>>=
all_mcaps <- list()
all_ci <- list()
for (p in prof_params) {

  mcap_tmp <- mcap(
    logLik = H1_profiles_long %>% filter(prof_var == p) %>% pull(logLik),
    parameter = H1_profiles_long %>% filter(prof_var == p) %>% pull(value)
  )

  all_mcaps[[p]] <- mcap_tmp
  tmp_ci <- mcap_tmp$ci
  names(tmp_ci) <- c("lower", "upper")

  all_ci[[p]] <- c(tmp_ci, 'mle' = mcap_tmp$mle)
}

all_mcaps <- purrr::map_df(
  prof_params,
  function(x) all_mcaps[[x]]$fit %>% mutate(prof_var = x)
)

all_ci <- all_ci |> as.data.frame() |> t() |> as.data.frame()
all_ci$prof_var <- rownames(all_ci)

H1_profiles %>% slice_max(order_by = logLik) %>% 
  select(logLik, all_of(prof_params)) %>% 
  mutate(E_0 = E_0 * 10911819, I_0 = I_0 * 10911819,
         tau_epi = log(tau_epi), tau_end = log(tau_end)) %>%
  pivot_longer(
    cols = -logLik,
    names_to = 'prof_var',
    values_to = 'value'
  ) -> mles
@

\begin{figure}[ht]
<<m1Profs, fig.height=4.6, fig.width=5.8, echo=FALSE, warning=FALSE, message=FALSE>>=
# Create a mapping between department names and what should be shown in figures
h1_plots <- list()
for (i in 1:length(prof_params)) {
  p = prof_params[i]
  if (p == 'betat') {
    my_lab <- TeX('$\\zeta')
  } else if (grepl('^beta[[:digit:]]$', p)) {
    num <- gsub('^beta', '', p)
    my_lab <- TeX(paste0('$\\beta_', num, '$'))
  } else if (p == 'E_0') {
    my_lab <- TeX('$E_0(0)$')
  } else if (p == 'I_0') {
    my_lab <- TeX('$I_0(0)$')
  } else if (p == 'nu' || p == 'rho') {
    my_lab <- TeX(paste0('$\\', p, '$'))
  } else if (p == 'sig_sq_epi') {
    my_lab <- TeX('$\\sigma$: Epidemic')
  } else if (p == 'sig_sq_end') {
    my_lab <- TeX('$\\sigma$: Endemic')
  } else if (p == 'tau_end') {
    my_lab <- TeX('$\\log(\\psi)$: Endemic')
  } else if (p == 'tau_epi') {
    my_lab <- TeX('$\\log(\\psi)$: Epidemic')
  }

  gg_tmp <- ggplot() +
    geom_point(
      data = H1_profiles_long %>% filter(prof_var == p), 
      aes(x = value, y = logLik)
    ) +
    geom_line(
      data = all_mcaps %>% filter(prof_var == p), 
      aes(x = parameter, y = smoothed), 
      col = 'blue'
    ) +
    # geom_line(data = all_mcaps, aes(x = parameter, y = quadratic), col = 'red') +
    geom_vline(
      data = all_ci %>% filter(prof_var == p), 
      aes(xintercept = lower), linetype = 'dashed'
    ) +
    geom_vline(
      data = all_ci %>% filter(prof_var == p), 
      aes(xintercept = upper), 
      linetype = 'dashed'
    ) +
    geom_vline(
      data = all_ci %>% filter(prof_var == p), 
      aes(xintercept = mle), col = 'blue'
    ) +
    # This code chunk below will plot the MLE along with the smoothed marginal 
    # MLEs and the confidence interval. 
    # geom_point(
    #   data = mles %>% filter(prof_var == p), 
    #   aes(x = value, y = logLik), col = 'red'
    # ) +
    labs(y = 'Log-Likelihood', title = my_lab) +
    theme(
      axis.text = element_text(size = 8),
      axis.title = element_text(size = 10)
    ) +
    theme_bw() +
    scale_y_continuous(limits = c(-2770, max(H1_profiles$logLik))) + 
    theme(
      axis.title.x = element_blank(), 
      plot.title = element_text(
        hjust = 0.5, size = 8, margin = margin(t = 0, unit = "pt")
        ),
      axis.title.y = element_text(size = 8),
      axis.text.y = element_text(size = 7),
      axis.text.x = element_text(size = 6.5),
      plot.margin = margin(5.5, 3, 0, -1, unit = 'pt')
    )

  if (p == 'E_0') {
    gg_tmp <- gg_tmp + 
      scale_x_continuous(breaks = c(0, round(all_ci[p, 'upper'])))
  } else if (p == 'I_0') {
    gg_tmp <- gg_tmp + 
      scale_x_continuous(
        breaks = c(round(all_ci[p, 'lower']), round(all_ci[p, 'upper']))
      )
  } else if (p == 'nu') {
    gg_tmp <- gg_tmp + 
      scale_x_continuous(breaks = c(0.9, 0.95, 1.0))
  }

  if (i %in% seq(1, 16, 4)) {
    h1_plots[[i]] <- gg_tmp
  } else {
    h1_plots[[i]] <- gg_tmp + theme(
      axis.text.y = element_blank(), axis.ticks.y = element_blank(),
      axis.title.y = element_blank()
      )
  }
}

cowplot::plot_grid(
  plotlist = h1_plots,
  align = 'h', ncol = 4,
  rel_widths = c(1.25, 1, 1, 1)
)
@
\caption{\label{m1Profs}MCAP confidence intervals for Model 1 parameters.}
\end{figure}

<<Load Model 2 Confidence Intervals, include=FALSE, message=FALSE, echo=FALSE>>=
NULL
@

<<Load Model 3 Profiles, include=FALSE, message=FALSE, echo=FALSE>>=
H3_profiles1 <- readRDS('../model3/run_level_3/h3_profiles.rds') %>% 
  filter(prof_var == 'std_W')
H3_profiles2 <- readRDS("../model3/run_level_3/h3_profiles2.rds")
H3_profiles3 <- readRDS("../model3/run_level_3/h3_profiles3.rds")

H3_MLE_evals <- readRDS('../model3/run_level_3/h3_MLE_eval.rds')

H3_profiles <- bind_rows(H3_profiles1, H3_profiles2, H3_profiles3)

prof_params <- unique(H3_profiles$prof_var)

H3_profiles_long <- H3_profiles %>%
  select(prof_var, logLik, logLik_se, all_of(paste0(prof_params, '1'))) %>%
  rename_with(~ sub("1$", "", .x)) %>%
  select(-epsilon, -XthetaA) %>% 
  pivot_longer(
    cols = -c(prof_var, logLik, logLik_se),
    names_to = 'variable',
    values_to = 'value'
  ) %>%
  filter(variable == prof_var) %>%
  group_by(prof_var, value) %>%
  slice_max(order_by = logLik, n = 1)

H3_remaining_long <- H3_profiles %>% 
  select(prof_var, logLik, logLik_se, epsilon1, XthetaA1) %>% 
  rename(epsilon = epsilon1, XthetaA = XthetaA1) %>% 
  mutate(XthetaA = sqrt(XthetaA)) %>%
  mutate(epsilon = epsilon^2) %>%
  pivot_longer(
    cols = -c(prof_var, logLik, logLik_se),
    names_to = 'variable',
    values_to = 'value'
  ) %>%
  filter(variable == prof_var) %>%
  group_by(prof_var, value) %>%
  slice_max(order_by = logLik, n = 2)

H3_profiles_long <- bind_rows(
  H3_profiles_long, H3_remaining_long
)
@

<<Create Confidence intervals>>=
all_mcaps <- list()
all_ci <- list()
for (p in prof_params) {

  mcap_tmp <- mcap(
    logLik = H3_profiles_long %>% filter(prof_var == p) %>% pull(logLik),
    parameter = H3_profiles_long %>% filter(prof_var == p) %>% pull(value),
    span = case_when(
      p == 'r' ~ 1,
      p == 'XthetaA' ~ 0.7,
      TRUE ~ 0.75
    )
  )

  all_mcaps[[p]] <- mcap_tmp
  tmp_ci <- mcap_tmp$ci
  names(tmp_ci) <- c("lower", "upper")

  all_ci[[p]] <- c(tmp_ci, 'mle' = mcap_tmp$mle)
}

all_mcaps <- purrr::map_df(
  prof_params,
  function(x) all_mcaps[[x]]$fit %>% mutate(prof_var = x)
)

all_ci <- all_ci |> as.data.frame() |> t() |> as.data.frame()
all_ci$prof_var <- rownames(all_ci)

# Transform epsilong (rho) to natural scale
all_ci <- all_ci %>% 
  mutate(
    lower = case_when(
      prof_var == 'epsilon' ~ sqrt(lower), 
      # prof_var == 'XthetaA' ~ lower^2,
      TRUE ~ lower
    ),
    upper = case_when(
      prof_var == 'epsilon' ~ sqrt(upper), 
      # prof_var == 'XthetaA' ~ upper^2,
      TRUE ~ upper
    ),
    mle = case_when(
      prof_var == 'epsilon' ~ sqrt(mle), 
      # prof_var == 'XthetaA' ~ mle^2,
      TRUE ~ mle
    )
  )

all_mcaps <- all_mcaps %>% 
  mutate(
    parameter = case_when(
      prof_var == 'epsilon' ~ sqrt(parameter), 
      # prof_var == 'XthetaA' ~ parameter^2,
      TRUE ~ parameter
      )
  )

H3_profiles_long <- H3_profiles_long %>% 
  mutate(
    value = case_when(
      prof_var == 'epsilon'~ sqrt(value), 
      # prof_var == 'XthetaA' ~ value^2,
      TRUE ~ value
    )
  )

# This chunk of code returns the MLE for model 3
H3_MLE_evals %>%
  slice_max(order_by = logLik, n = 1) %>%
  select(logLik, all_of(paste0(prof_params, '1'))) %>%
  rename_with(~ sub("1$", "", .x)) %>%
  mutate(XthetaA = sqrt(XthetaA)) %>%
  pivot_longer(
    cols = -c(logLik),
    names_to = 'prof_var',
    values_to = 'value'
  ) -> mles
@

\begin{figure}[ht]
<<m3Profs, fig.height=4.6, fig.width=5.8, echo=FALSE, warning=FALSE, message=FALSE>>=
h3_plots <- list()
for (i in 1:length(prof_params)) {
  p = prof_params[i]
  if (p == 'mu_B') {
    my_lab <- TeX('$\\zeta')
  } else if (p == 'XthetaA') {
    my_lab <- TeX('$\\sqrt{\\epsilon_W}$')
  } else if (p == 'thetaI') {
    my_lab <- TeX('$\\mu_W$')
  } else if (p == 'lambdaR') {
    my_lab <- TeX('$a$')
  } else if (p == 'r') {
    my_lab <- TeX('$r$')
  } else if (p == 'std_W') {
    my_lab <- TeX('$\\sigma$')
  } else if (p == 'epsilon') {
    my_lab <- TeX('$\\rho$')
  } else if (p == 'k') {
    my_lab <- TeX('$\\psi$')
  }

  gg_tmp <- ggplot() +
    geom_point(
      data = H3_profiles_long %>% filter(prof_var == p), 
      aes(x = value, y = logLik)
    ) +
    geom_line(
      data = all_mcaps %>% filter(prof_var == p), 
      aes(x = parameter, y = smoothed), 
      col = 'blue'
    ) +
    # geom_line(data = all_mcaps, aes(x = parameter, y = quadratic), col = 'red') +
    geom_vline(
      data = all_ci %>% filter(prof_var == p), 
      aes(xintercept = lower), linetype = 'dashed'
    ) +
    geom_vline(
      data = all_ci %>% filter(prof_var == p), 
      aes(xintercept = upper), 
      linetype = 'dashed'
    ) +
    geom_vline(
      data = all_ci %>% filter(prof_var == p), 
      aes(xintercept = mle), col = 'blue'
    ) +
    # This commented out code below will plot the MLE along with the confidence 
    # intervals for the given parameter. 
    # geom_point(
    #   data = mles %>% filter(prof_var == p),
    #   aes(x = value, y = logLik), col = 'red'
    # ) +
    labs(y = 'Log-Likelihood', title = my_lab) +
    theme(
      axis.text = element_text(size = 8),
      axis.title = element_text(size = 10)
    ) +
    theme_bw() +
    # scale_y_continuous(limits = c(-17450, max(H3_profiles$logLik))) + 
    scale_y_continuous(limits = c(-17450, max(H3_MLE_evals$logLik))) +
    theme(
      axis.title.x = element_blank(), 
      plot.title = element_text(
        hjust = 0.5, size = 8, margin = margin(t = 0, unit = "pt")
        ),
      axis.title.y = element_text(size = 8),
      axis.text.y = element_text(size = 7),
      axis.text.x = element_text(size = 6.5),
      plot.margin = margin(5.5, 3, 0, -1, unit = 'pt')
    ) + 
    theme(legend.position = 'none')

  if (i %in% seq(1, 9, 3)) {
    h3_plots[[i]] <- gg_tmp
  } else {
    h3_plots[[i]] <- gg_tmp + theme(
      axis.text.y = element_blank(), axis.ticks.y = element_blank(),
      axis.title.y = element_blank()
      )
  }
}

cowplot::plot_grid(
  plotlist = h3_plots,
  align = 'h', ncol = 3,
  rel_widths = c(1.25, 1, 1, 1)
)
@
\caption{\label{m1Profs}MCAP confidence intervals for Model 1 parameters.}
\end{figure}

%%%%%%%%%% END

\bibliography{../bib-haiti}

\end{document}

### 2024-02-06 H3_MLE_Eval.R
Created and ran script to evaluate log-likelihood for parameters in confidence region

@@ -1,0 +1,297 @@
+# Finding and evaluating the MLE for model 3 using the profile likelihood 
+# searches. 
+#
+# ********
+# NOTE:
+#
+#   If you would like to run this code but have never used the `batchtools` R
+#   package, then we strongly recommend completing both Examples 1 and 2 from
+#   the batchtools vignette, by first installing the `batchtools` package and
+#   running: `vignette('batchtools')`, or visiting:
+#
+#         https://mllg.github.io/batchtools/articles/batchtools.html
+#
+# WARNING:
+#
+#   This script was ran using a high-performance computing cluster. In total,
+#   the RUN_LEVEL == 3 version of this script took TODO compute hours. Partial
+#   results can be obtained by lowering the required computational effort
+#   (params: NP, NBPF, nprof, NP_EVAL, NREPS_EVAL), which can be done by
+#   reducing the RUN_LEVEL to 2 or 1, or alternatively by manually changing
+#   these values.
+# ********
+#
+# In this script we use the results of the profile likelihood searches to find 
+# and evaluate the MLE of Model 3. The profile likelihood searches (done using 
+# the scripts H3profile.R, H3profile2.R, and H3profile3.R) contain the results 
+# of a large number of searches and computation. These scripts result in 
+# data.frame objects, where each row is a different parameter estimate. 
+# Here, we consider only parameter estimates that are within the smoothed MCAP
+# confidence intervals for all profile likelihoods, and evaluate their 
+# likelihood. The time-complexity of each job is roughly linear in the 
+# parameters NP_EVAL, and NREPS_EVAL. 
+
+
+# Load Packages -----------------------------------------------------------
+
+library(batchtools)
+library(tidyverse)
+library(haitipkg)
+
+RUN_LEVEL <- 3
+
+NP_EVAL <- switch(RUN_LEVEL, 100, 1000, 5000)
+NREPS_EVAL <- switch(RUN_LEVEL, 3, 6, 15)
+
+# Create Experiment Registry ----------------------------------------------
+
+reg <- makeExperimentRegistry(
+  file.dir = paste0('model3/H3_MLE_Eval_RL', RUN_LEVEL),
+  seed = 739164,
+  packages = c("spatPomp", 'haitipkg', 'pomp')
+)
+
[…]

### 2024-02-06 ms.Rnw
incorporated H3 profiles into manuscript, needs to be re-ran.

@@
 
-# TODO: This is used later in figure 2 (TODO)
-mcap_results <- pomp::mcap(h1_prof_res$logLik, h1_prof_res$betat, span = 1)
+# TODO: update this comment: This is used later in figure 2 (TODO)
+beta_t_mcap_results <- pomp::mcap(h1_prof_res$logLik, h1_prof_res$betat, span = 1)
 
@@
 # Load the file containing profiles of Model 1 Parameters.
-if (RUN_LEVEL != 3 && file.exists("model3/run_level_3/h3_profiles.rds")) {
+if (RUN_LEVEL != 3 && file.exists("model3/run_level_3/h3_MLE_eval.rds")) {
 
@@
   # results.
-  h3_prof <- readRDS("model3/run_level_3/h3_profiles.rds")
-} else if (file.exists(paste0('model3/', rl_dir, 'h3_profiles.rds'))) {
+  h3_prof <- readRDS("model3/run_level_3/h3_MLE_eval.rds")
+} else if (file.exists(paste0('model3/', rl_dir, 'h3_MLE_eval.rds'))) {
 
   # If the `model3/run_level_<RUN_LEVEL>/h3_profiles.rds` file exists, load it
-  h3_prof <- readRDS(paste0('model3/', rl_dir, 'h3_profiles.rds'))
+  h3_prof <- readRDS(paste0('model3/', rl_dir, 'h3_MLE_eval.rds'))
 } else {
@@
   # Throw an error, letting users know how to create the file.
-  stop(paste0("Missing file: `model3/", rl_dir, 'h3_profiles.rds`;\nThe file can be obtained by running the script `H3profile.R`'))
+  stop(paste0("Missing file: `model3/", rl_dir, 'h3_MLE_eval.rds`;\nThe file can be obtained by running the script `H3_MLE_Eval.R`.'))
 }
@@
 We similarly allow the re-estimation of process and measurement overdispersion parameters ($\sigmaProc^2$ and $\obsOverdispersion$), and require that the latent Markov process $X(t)$ carry over from one phase into the next.
-The resulting 95\% confidence interval for $\transmissionTrend$ is $(\Sexpr{myround(mcap_results$ci[1], digits = 3)}, \Sexpr{myround(mcap_results$ci[2], digits = 3)})$, with the full results displayed in Fig.~\ref{fig:betat}.
+The resulting 95\% confidence interval for $\transmissionTrend$ is $(\Sexpr{myround(beta_t_mcap_results$ci[1], digits = 3)}, \Sexpr{myround(beta_t_mcap_results$ci[2], digits = 3)})$, with the full results displayed in Fig.~\ref{fig:betat}.
 These results are suggestive that the inclusion of a trend in the transmission rate improves the quantitative ability of Model~1 to describe the observed data.
-\new{The maximum likelihood estimate for $\transmissionTrend$ corresponds with a \Sexpr{paste0(myround(100 - 100*exp(mcap_results$mle * 430/215), 1), '\\%')} reduction to the transmission rate over the course of the outbreak, with a $95\%$ confidence interval (\Sexpr{paste0(rev(myround(100 - 100*exp(mcap_results$ci * 430/215), 1)), collapse = ', ', sep = '\\%')}) for the overall reduction in transmission.}
+\new{The maximum likelihood estimate for $\transmissionTrend$ corresponds with a \Sexpr{paste0(myround(100 - 100*exp(beta_t_mcap_results$mle * 430/215), 1), '\\%')} reduction to the transmission rate over the course of the outbreak, with a $95\%$ confidence interval (\Sexpr{paste0(rev(myround(100 - 100*exp(beta_t_mcap_results$ci * 430/215), 1)), collapse = ', ', sep = '\\%')}) for the overall reduction in transmission.}
 The reported results for Model~1 in the remainder of this article were obtained with the inclusion of the parameter $\transmissionTrend$.
@@
   geom_point(data = h1_prof_res, aes(x = betat, y = logLik)) +
-  geom_line(data = mcap_results$fit, aes(x = parameter, y = smoothed), col = 'blue') +
-  geom_line(data = mcap_results$fit, aes(x = parameter, y = quadratic), col = 'red') +
-  geom_vline(xintercept = mcap_results$ci[1], linetype = 'dashed') +
-  geom_vline(xintercept = mcap_results$ci[2], linetype = 'dashed') +
-  geom_vline(xintercept = mcap_results$mle, col = 'blue') +
+  geom_line(data = beta_t_mcap_results$fit, aes(x = parameter, y = smoothed), col = 'blue') +
+  geom_line(data = beta_t_mcap_results$fit, aes(x = parameter, y = quadratic), col = 'red') +
+  geom_vline(xintercept = beta_t_mcap_results$ci[1], linetype = 'dashed') +
+  geom_vline(xintercept = beta_t_mcap_results$ci[2], linetype = 'dashed') +
+  geom_vline(xintercept = beta_t_mcap_results$mle, col = 'blue') +
   labs(x = "Log-Linear Trend in Transmission", y = 'Log Likelihood') +
@@
 suppressWarnings(
-  rm(h1_prof_res, mcap_results)
+  rm(h1_prof_res, beta_t_mcap_results)
 )

### 2024-02-06 H3_MLE_Eval.R
Added some additional comments documenting how long the script take

@@ -14,7 +14,7 @@
 # WARNING:
 #
 #   This script was ran using a high-performance computing cluster. In total,
-#   the RUN_LEVEL == 3 version of this script took TODO compute hours. Partial
+#   the RUN_LEVEL == 3 version of this script took 320 compute hours. Partial
 #   results can be obtained by lowering the required computational effort
 #   (params: NP, NBPF, nprof, NP_EVAL, NREPS_EVAL), which can be done by
 #   reducing the RUN_LEVEL to 2 or 1, or alternatively by manually changing
@@ -244,6 +244,9 @@ ades <- list(
 addExperiments(prob.designs = pdes, algo.designs = ades)
 
 # Submit Jobs -------------------------------------------------------------
+
+# The maximum run-time of the 168 jobs was 2.20 hours, with a mean time of 
+# 1.91 hours
 
 resources1 <- list(account = 'stats_dept1', walltime = '3:00:00', memory = '5000m', ncpus = 1)
 submitJobs(resources = resources1)

### 2024-02-06 knit_ms.sh
tracking knit slurm script for convinience

@@ -1,0 +1,31 @@
+#!/bin/bash
+
+#SBATCH --job-name=ms_RL3
+#SBATCH --mail-user=jeswheel@umich.edu
+#SBATCH --mail-type=BEGIN,END,FAIL
+
+#SBATCH --account=ionides2
+#SBATCH --partition=standard
+
+#SBATCH --nodes=1
+#SBATCH --ntasks-per-node=36
+#SBATCH --cpus-per-task=1
+
+## 5GB/cpu is the basic share
+#SBATCH --mem-per-cpu=2GB
+
+## wall time hours:minutes:seconds
+#SBATCH --time=6-00:00:00
+
+###   Load software modules
+
+module load R
+module list
+
+####  Commands your job should run follow this line
+
+echo "Running on $SLURM_JOB_NODELIST"
+echo "Running in $(pwd)"
+
+Rscript -e "library(knitr); knit2pdf('ms.Rnw')"
+

### 2024-02-06 knit_ms.sh
Updating knit script with reduce computational needs

@@ -4,7 +4,7 @@
 #SBATCH --mail-user=jeswheel@umich.edu
 #SBATCH --mail-type=BEGIN,END,FAIL
 
-#SBATCH --account=ionides2
+#SBATCH --account=stats_dept1
 #SBATCH --partition=standard
 
 #SBATCH --nodes=1
@@ -12,10 +12,10 @@
 #SBATCH --cpus-per-task=1
 
 ## 5GB/cpu is the basic share
-#SBATCH --mem-per-cpu=2GB
+#SBATCH --mem-per-cpu=1GB
 
 ## wall time hours:minutes:seconds
-#SBATCH --time=6-00:00:00
+#SBATCH --time=05:00:00
 
 ###   Load software modules
 

### 2024-02-06 ms.Rnw
updated many references and other minor edits

@@
 \old{The data that were used for this study, and that is the subject of our reanalysis, is displayed in Fig.~[1].}
-\new{Various dynamic models were developed to study this outbreak: searching PubMed with keywords ``Haiti, cholera, model'' we obtained 20 deterministic models \cite{kirpich16,kirpich17,andrews11,botelho21,fitzgibbon20,eisenberg13,rinaldo12,chao11,date11,lin19,collins21,akman15,trevisin22,mavian20,collins14,kelly16,capone15,bengtsson15,mari15,gatto12} and 8 stochastic models \cite{tuite11,mukandavire13,lewnard16,kunkel17,sallah17,azman12,kuhn14,azman15}.
+\new{Various dynamic models were developed to study this outbreak: searching PubMed with keywords ``Haiti, cholera, model'' we obtained 22 studies that utilized deterministic models \cite{lee20,tuite11,andrews11,botelho21,fitzgibbon20,eisenberg13,rinaldo12,chao11,date11,lin19,abrams13,collins21,akman15,trevisin22,mavian20,collins14,kelly16,capone15,leung22,mari15,gatto12,kuhn14}
+and 11 studies that used stochastic models \cite{kirpich16,lee20,pasetto18,mukandavire13,kirpich17,lewnard16,kunkel17,mukandavire15,sallah17,azman12,azman15}.
+Incidence data on the outbreak were available at various spatial scales, which resulted with 17 studies that considered spatially explicit dynamic models \cite{lee20,tuite11,pasetto18,fitzgibbon20,eisenberg13,rinaldo12,chao11,abrams13,trevisin22,sallah17,collins14,kelly16,azman12,leung22,kuhn14,mari15,gatto12}.
 Here we focus on a multi-group modeling exercise by Lee et al.~\cite{lee20} in which four expert modeling teams developed models to the same dataset with the goal of comparing conclusions on the feasibility of eliminating cholera by a vaccination campaign.}
@@
 Model~4 has an agent-based construction, featuring considerable mechanistic detail but limited ability to calibrate these details to data.
-\new{These modeling strategies were selected by \cite{li20} to represent the range of approaches used in the research community.}
-\jwc{I think the citation above is incorrect? Should it be lee20, not li20?}.
+\new{These modeling strategies were selected by \cite{lee20} to represent the range of approaches used in the research community.}
 We focus on Models~1--3, as the strengths and weaknesses of the agent-based modeling approach \cite{tracy18} are outside the scope of this article.
 \new{In addition, agent-based models were less widely used, the agent based model of \cite{lee20} being the only model of this class that was found in our literature review.}
-\new{The data that were used for this study, and that is the subject of our reanalysis, are displayed in Fig.~\ref{fig:CholeraData}.}
-
-\new{\editMechModels}
+\new{The data that were used in \cite{lee20}, and that is the subject of our reanalysis, are displayed in Fig.~\ref{fig:CholeraData}.}
 
@@
 
+\new{\editMechModels}
+
 %% IT SEEMS THIS PARAGRAPH IS DEPRECATED IN FAVOR OF CITING THESE REFERENCES WHERE APPROPRIATE
@@
 we then describe how each model is calibrated to data, and present a systematic approach to examining and refining these models.
-\new{Specifically, we focus on how to develop and test variations to these proposed models, as well as diagnosing the models once they have been assimilated to incidence reports.} \jwc{make a statement about fitting Model 3 (stochastic meta-pop) here?}
+\new{Specifically, we focus on how to develop and test variations of the proposed models, as well as diagnosing the models once they have been assimilated to incidence reports.
+This includes a comprehensive tutorial on performing inference with Model~3, a highly non-linear, spatially explicit stochastic model, a challenging task that is possible due to recent methodological advancements.}
 We then use \new{the} improved model fits to project cholera incidence in Haiti under various vaccination scenarios considered by Lee et al.~\cite{lee20}.
-Finally, we conclude with a discussion of the results.
+Finally, we conclude with a discussion of the results\new{, in which we relate our general recommendations for model based inference of biological systems to the case study of the Haiti cholera outbreak.}
 
@@
 This motivates the development of mechanistic models, which provides a statistical fit to the available data while also supporting a causal interpretation.
-\jwc{Maybe right here is another place for the new paragraph on mechanistic models?}
+\new{Despite their limited ability to project the effect of interventions on a system, associative models can be effectively used to make inference on an certain features of a system. 
+In our literature review, there were 14 studies that used associative models to describe various aspects of the cholera epidemic \cite{hulland19,moise20,piarroux11,matias17,eisenberg13,rebaudet19CATI,ivers15,raila17,cuneo17,charles14,richterman19,richterman19inverse,bengtsson15,li16adjusting}.}
 
@@
 \eic{I'VE STARTED TO PUT IN SOME DISCUSSION OF BAYESIAN METHODS HERE, THOUGH IT COULD ALSO GO IN THE DISCUSSION}
-\new{The main alternative to maximum likelihood estimation is Bayesian inference via Markov chain Monte Carlo, used to analyze the Haiti cholera epidemic by \cite{andrews11,rinaldo12,lewnard16,trevisin22,sallah17,azman12,kuhn14,mari15,gatto12}.
+\new{The main alternative to maximum likelihood estimation is Bayesian inference via Markov chain Monte Carlo, used to analyze the Haiti cholera epidemic by \cite{andrews11,pasetto18,rinaldo12,lewnard16,trevisin22,sallah17,azman12,kuhn14,mari15,gatto12}.
 Alternatively, some investigations obtained parameters from previous sources without calibrating to data.}
@@
 These results are suggestive that the inclusion of a trend in the transmission rate improves the quantitative ability of Model~1 to describe the observed data.
-\new{The maximum likelihood estimate for $\transmissionTrend$ corresponds with a \Sexpr{paste0(myround(100 - 100*exp(beta_t_mcap_results$mle * 430/215), 1), '\\%')} reduction to the transmission rate over the course of the outbreak, with a $95\%$ confidence interval (\Sexpr{paste0(rev(myround(100 - 100*exp(beta_t_mcap_results$ci * 430/215), 1)), collapse = ', ', sep = '\\%')}) for the overall reduction in transmission.}
+\new{The maximum likelihood estimate for $\transmissionTrend$ corresponds with a \Sexpr{paste0(myround(100 - 100*exp(beta_t_mcap_results$mle * 430/215), 1), '\\%')} reduction to the transmission rate over the course of the outbreak, with a $95\%$ confidence interval of (\Sexpr{paste0(rev(myround(100 - 100*exp(beta_t_mcap_results$ci * 430/215), 1)), collapse = ', ', sep = '\\%')}) for the overall reduction in transmission.}
 The reported results for Model~1 in the remainder of this article were obtained with the inclusion of the parameter $\transmissionTrend$.
@@
 Fig.~\ref{fig:h1SeasRain} shows that the estimated seasonal transmission rate $\transmission$ mimics the rainfall dynamics in Haiti, despite Model~1 not having access to rainfall data.
-This is consistent with previous studies finding that rainfall played an important role in cholera transmission in Haiti \cite{lemaitre19,eisenberg13}.
-\TODO{Citations from lit-review}.
+This is consistent with previous studies \old{finding that rainfall played an important role in cholera transmission in Haiti [50, 51]} \new{that incorporated rainfall as an important part of their mechanistic model or otherwise argue that rainfall is an important driver of cholera dynamics in Haiti \cite{hulland19,kirpich17,lee20,moise20,pasetto18,kirpich16,eisenberg13,rinaldo12,mavian20}.}
 The estimated seasonality also features an increased transmission rate during the fall, which was noticed at an earlier stage of the epidemic \cite{rinaldo12}; the high transmission rate in the fall may also be a result of the increase transmission that occurred in the fall of 2016, when hurricane Matthew struck Haiti \cite{ferreirai16}.

### 2024-02-06 ms.Rnw
Recompiling article after incorporating new model3 fits

@@
 
+# TODO: Remove
 # haiti3_fit contains ll estimate for many parameter sets, so we want to find
 # the highest value
-mod3_ll <- h3_prof %>%
-  slice_max(order_by = logLik, n = 1) %>%
-  pull(logLik)
+# mod3_ll <- h3_prof %>%
+#   slice_max(order_by = logLik, n = 1) %>%
+#   pull(logLik)
 @
@@
 # model 3 AIC
-mod3_aic <- 2 * mod3_n_params - 2 * mod3_ll
+# TODO: Remove
+# mod3_aic <- 2 * mod3_n_params - 2 * mod3_ll
 
@@
 
+mod3_ll <- mod3_eval_res$ll[1]
+mod3_aic <- 2 * mod3_n_params - 2 * mod3_ll
 mod3_eval_res$subset_aic <- 2 * mod3_n_params - 2 * mod3_eval_res$subset_ll['se']
@@
   geom_line(data = beta_t_mcap_results$fit, aes(x = parameter, y = smoothed), col = 'blue') +
-  geom_line(data = beta_t_mcap_results$fit, aes(x = parameter, y = quadratic), col = 'red') +
+  # geom_line(data = beta_t_mcap_results$fit, aes(x = parameter, y = quadratic), col = 'red') +
   geom_vline(xintercept = beta_t_mcap_results$ci[1], linetype = 'dashed') +

### 2024-02-07 H2profileAll.R
Created file to profile over all h2 parameters

@@ -1,0 +1,379 @@
+# Fitting profiles for Haiti 2.
+#
+# ********
+# NOTE:
+#
+#   If you would like to run this code but have never used the `batchtools` R
+#   package, then we strongly recommend completing both Examples 1 and 2 from
+#   the batchtools vignette, by first installing the `batchtools` package and
+#   running: `vignette('batchtools')`, or visiting:
+#
+#         https://mllg.github.io/batchtools/articles/batchtools.html
+#
+# ********
+#
+# The fit_haiti2 function is a very simple helper that will take as input
+# some starting parameters and an objective function for Model 2 and will
+# return parameters estimates along with corresponding estimates of their
+# log-likelihood.
+#
+# This this script we use this helper function to do a profile likelihood search
+# of the estimated parameters using the batchtools package. To do this, we will
+# create a grid of parameter values and submit jobs where each job will use a
+# row from the grid as a starting point to fit the model.
+
+
+# Load Packages -----------------------------------------------------------
+
+library(batchtools)
+library(tidyverse)
+library(data.table)
+library(haitipkg)
+library(pomp)
+library(spatPomp)
+library(subplex)
+
+
+# Create Experiment Registry ----------------------------------------------
+
+reg <- makeExperimentRegistry(
+  file.dir ='model2/profileAll',
+  seed = 739164,
+  packages = c("spatPomp", 'haitipkg', 'pomp')
+)
+
+# More than one start could be considered, but our preliminary results 
+# suggest that only one is needed. 
+n_starts <- 1
+
+# Create Profile Design Matrix --------------------------------------------
+
+set.seed(22446688)
+h2 <- haiti2(cutoff = 10000, measure = "log")
+
+prof_params <- c(
+  "Mu",
+  "Beta",
+  "BetaW",
+  "v",
+  "sigma",
+  "phase",
+  paste0("InitInfected", 1:10),
+  'Delta',
+  'Rho', 
+  'redmu', 
+  'AlphaS'
+)
+
+final_pars <- coef(h2)
+prof_vars <- c()
+for (pp in prof_params) {
+
+  if (pp == 'Mu') {
+    prof_values <- seq(6750, 13000, length.out = 30)
[…]

### 2024-02-07 H2profileAll.R
temporarily checking only initialization pars

@@ -33,36 +33,39 @@ library(pomp)
 library(spatPomp)
 library(subplex)
 
-
 # Create Experiment Registry ----------------------------------------------
 
 reg <- makeExperimentRegistry(
-  file.dir ='model2/profileAll',
+  file.dir ='model2/profileInit',
   seed = 739164,
   packages = c("spatPomp", 'haitipkg', 'pomp')
 )
 
 # More than one start could be considered, but our preliminary results 
 # suggest that only one is needed. 
-n_starts <- 1
+n_starts <- 3
 
 # Create Profile Design Matrix --------------------------------------------
 
 set.seed(22446688)
 h2 <- haiti2(cutoff = 10000, measure = "log")
 
+# prof_params <- c(
+#   "Mu",
+#   "Beta",
+#   "BetaW",
+#   "v",
+#   "sigma",
+#   "phase",
+#   paste0("InitInfected", 1:10),
+#   'Delta',
+#   'Rho', 
+#   'redmu', 
+#   'AlphaS'
+# )
+
 prof_params <- c(
-  "Mu",
-  "Beta",
-  "BetaW",
-  "v",
-  "sigma",
-  "phase",
-  paste0("InitInfected", 1:10),
-  'Delta',
-  'Rho', 
-  'redmu', 
-  'AlphaS'
+  paste0("InitInfected", 1:10)
 )
 
 final_pars <- coef(h2)
@@ -82,24 +85,24 @@ for (pp in prof_params) {
   } else if (pp == "phase") {
     prof_values <- seq(6.5, 7.9, length.out = 30)
   } else if (pp == 'InitInfected1') {
-    prof_values <- seq(5000, 40000, length.out = 50)
+    prof_values <- seq(5000, 15000, length.out = 50)
   } else if (pp == 'InitInfected2') {
     prof_values <- seq(1000, 7000, length.out = 40)
-  } else if (pp == "initInfected 3") {
+  } else if (pp == "InitInfected3") {
     prof_values <- seq(0.5, 1000, length.out = 30)
-  } else if (pp == "initInfected 4") {
+  } else if (pp == "InitInfected4") {
     prof_values <- seq(0.5, 1000, length.out = 30)
-  } else if (pp == "initInfected 5") {
+  } else if (pp == "InitInfected5") {
     prof_values <- seq(0.5, 1500, length.out = 30)
-  } else if (pp == "initInfected 6") {
+  } else if (pp == "InitInfected6") {
     prof_values <- seq(0.5, 1000, length.out = 30)
[…]

### 2024-02-07 H2profile.R
added minor comment

@@ -42,6 +42,8 @@ reg <- makeExperimentRegistry(
   packages = c("spatPomp", 'haitipkg', 'pomp')
 )
 
+# More than one start could be considered, but our preliminary results 
+# suggest that only one is needed. 
 n_starts <- 1
 
 # Create Profile Design Matrix --------------------------------------------

### 2024-02-07 H2profile.R
Modified profile CI bounds

@@ -65,17 +65,17 @@ prof_vars <- c()
 for (pp in prof_params) {
 
   if (pp == 'Mu') {
-    prof_values <- seq(6750, 13000, length.out = 30)
+    prof_values <- seq(6520, 14750, length.out = 30)
   } else if (pp == "Beta") {
-    prof_values <- seq((.Machine$double.eps)^(1/5), (6.5e-05)^(1/5), length.out = 50)^5
+    prof_values <- seq((.Machine$double.eps)^(1/5), 0.095, length.out = 50)^5
   } else if (pp == "BetaW") {
-    prof_values <- seq(1, 1.2, length.out = 30)
+    prof_values <- seq(1.04, 1.155, length.out = 30)
   } else if (pp == "v") {
-    prof_values <- seq(1.225, 1.44, length.out = 25)
+    prof_values <- seq(1.265, 1.375, length.out = 30)
   } else if (pp == "sigma") {
-    prof_values <- seq((.Machine$double.eps)^(1/5), (2e-03)^(1/5), length.out = 50)^5
+    prof_values <- seq((.Machine$double.eps)^(1/5), 0.29, length.out = 50)^5
   } else if (pp == "phase") {
-    prof_values <- seq(6.5, 7.9, length.out = 30)
+    prof_values <- seq(7.01, 7.5, length.out = 30)
   }
 
   prof_cols <- matrix(rep(prof_values, each = n_starts), ncol = 1)
@@ -262,7 +262,7 @@ submitJobs(
 #
 # saveRDS(
 #   h2_results,
-#   'model2/h2_profiles.rds')
+#   'model2/h2_profiles.rds'
 # )
 #
 # We can also get information about each job, which can be helpful to pick the

### 2024-02-07 r1/response.tex
minor edits to response, added H2 profiles to si

@@
 \begin{quote}
-If the table has a very complex structure or contains graphics, the safest solution is to make it into a figure. Export the table as a TIFF, and cite and re-label it as a figure.
+``If the table has a very complex structure or contains graphics, the safest solution is to make it into a figure. Export the table as a TIFF, and cite and re-label it as a figure."
 \end{quote}

### 2024-02-07 si/confidenceIntervals.Rnw
minor edits to response, added H2 profiles to si

@@
 For each model and parameter, we use profile log-likelihoods to obtain confidence intervals. TODO: Cite.
-Due to the non-linear and stochastic nature of Models~1 and 3, exact evaluation of the profile log-likelihood is unattainable.
+Due to the non-linear and stochastic nature of Models~1 and 3, exact evaluations of the profile log-likelihood are difficult to obtain.
 Instead, the log-likelihood at each point of the profile is estimated using via Monte-Carlo based particle filter methods.
@@
 \begin{figure}[ht]
-<<m1Profs, fig.height=4.6, fig.width=5.8, echo=FALSE, warning=FALSE, message=FALSE>>=
+<<m1Profs, fig.height=5, fig.width=6, echo=FALSE, warning=FALSE, message=FALSE>>=
 # Create a mapping between department names and what should be shown in figures
@@
 <<Load Model 2 Confidence Intervals, include=FALSE, message=FALSE, echo=FALSE>>=
-NULL
+h2_profiles <- readRDS("../model2/h2_profiles.rds") %>% as.data.frame()
+
+prof_params <- c(
+  "Mu",
+  "BetaW",
+  "v",
+  "phase",
+  "Beta",
+  "sigma"
+)
+
+h2_plots <- list()
+for (i in 1:length(prof_params)) {
+  p = prof_params[i]
+  if (p == 'Mu') {
+    my_lab <- TeX('$\\mu_W')
+  } else if (p == 'Beta') {
+    my_lab <- TeX('$\\beta$')
+  } else if (p == 'BetaW') {
+    my_lab <- TeX('$\\beta_W$')
+  } else if (p == 'v') {
+    my_lab <- TeX("$\\psi")
+  } else if (p == 'sigma') {
+    my_lab <- TeX('$\\mu_{\\text{RS}}$')
+  } else if (p == 'phase') {
+    my_lab <- TeX('$\\phi')
+  }
+
+  if (p == 'Beta') {
+    tmp_df_beta <- h2_profiles %>% 
+      filter(prof_var == 'Beta') %>% 
+      # filter(logLik >= -21965) %>%
+      arrange(p)
+
+    vals_beta <- tmp_df_beta[, 'Beta']^(1/5)
+
+    profile_like_beta <- splinefun(x = vals_beta, y = tmp_df_beta[, 'logLik'])
+    # max_profile <- optimize(profile_like, interval = range(vals), maximum = TRUE)
+    max_profile_beta <- max(h2_profiles$logLik)
+    threshold_beta <- max_profile_beta - 1.92
+    plot_threshold_beta <- max_profile_beta - 6
+    profile_roots_beta <- function(x) {profile_like_beta(x) - threshold_beta}
+    plot_limits_beta <- function(x) {profile_like_beta(x) - plot_threshold_beta}
+
+    upper_bound_beta <- uniroot(
+      profile_roots_beta, 
+      interval = c(vals_beta[which.max(tmp_df_beta$logLik)], max(vals_beta))
+    )$root
+
+    max_x_beta <- uniroot(
+      plot_limits_beta, 
+      interval = c(0, 7e-06^(1/5))
+    )$root
+
+    gg_tmp_beta <- ggplot() + 
+      geom_point(
+        data = tmp_df_beta %>% filter(prof_var == p) %>% filter(Beta <= max_x_beta^5),
+        aes(x = Beta, y = logLik)
+      ) + 
+      geom_function(fun = function(x) profile_like_beta(x^(1/5))) + 
+      geom_vline(xintercept = upper_bound_beta^5, linetype = 'dashed') + 
+      geom_vline(xintercept = 0, col = 'blue') + 
+      labs(y = 'Log-Likelihood', title = my_lab) + 
+      theme(
+        axis.text = element_text(size = 8),
+        axis.title = element_text(size = 10)
+      ) +
+      theme_bw() +
+      scale_y_continuous(limits = c(max_profile_beta-6, max_profile_beta)) +
+      scale_x_continuous(limits = c(0, max_x_beta^5)) +
+      theme(
+        axis.title.x = element_blank(), 
+        plot.title = element_text(
+          hjust = 0.5, size = 8, margin = margin(t = 0, unit = "pt")
+        ),
+        axis.title.y = element_text(size = 8),
+        axis.text.y = element_text(size = 7),
+        axis.text.x = element_text(size = 6.5),
+        plot.margin = margin(5.5, 3, 0, -1, unit = 'pt')
+      )
+
+    if (i %in% c(1, 4)) {
+      h2_plots[['beta']] <- gg_tmp_beta
+    } else {
+      h2_plots[['beta']] <- gg_tmp_beta + theme(
+        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
+        axis.title.y = element_blank()
+      )
+    }
+
+  } else if (p == 'sigma') {
+    tmp_df_sigma <- h2_profiles %>% 
+      filter(prof_var == 'sigma') %>% 
+      # filter(logLik >= -21965) %>%
+      arrange(p)
+
+    vals_sigma <- tmp_df_sigma[, 'sigma']^(1/5)
+
+    profile_like_sigma <- splinefun(x = vals_sigma, y = tmp_df_sigma[, 'logLik'])
+    # max_profile <- optimize(profile_like, interval = range(vals), maximum = TRUE)
+    max_profile_sigma <- max(h2_profiles$logLik)
+    threshold_sigma <- max_profile_sigma - 1.92
+    plot_threshold_sigma <- max_profile_sigma - 6
+    profile_roots_sigma <- function(x) {profile_like_sigma(x) - threshold_sigma}
+    plot_limits_sigma <- function(x) {profile_like_sigma(x) - plot_threshold_sigma}
+
+    upper_bound_sigma <- uniroot(
+      profile_roots_sigma, 
+      interval = c(vals_sigma[which.max(tmp_df_sigma$logLik)], max(vals_sigma))
+    )$root
+
+    max_x_sigma <- uniroot(
+      plot_limits_sigma, 
+      interval = c(0, 0.0022^(1/5))
+    )$root
+
+    gg_tmp_sigma <- ggplot() + 
+      geom_point(
+        data = tmp_df_sigma %>% filter(prof_var == 'sigma') %>% filter(sigma <= max_x_sigma^5),
+        aes(x = sigma, y = logLik)
+      ) + 
+      geom_function(fun = function(x) profile_like_sigma(x^(1/5))) + 
+      geom_vline(xintercept = upper_bound_sigma^5, linetype = 'dashed') + 
+      geom_vline(xintercept = 0, col = 'blue') + 
+      labs(y = 'Log-Likelihood', title = my_lab) + 
+      theme(
+        axis.text = element_text(size = 8),
+        axis.title = element_text(size = 10)
+      ) +
+      theme_bw() +
+      scale_y_continuous(limits = c(max_profile_sigma-6, max_profile_sigma)) +
+      scale_x_continuous(limits = c(0, max_x_sigma^5)) +
+      theme(
+        axis.title.x = element_blank(), 
+        plot.title = element_text(
+          hjust = 0.5, size = 8, margin = margin(t = 0, unit = "pt")
+        ),
+        axis.title.y = element_text(size = 8),
+        axis.text.y = element_text(size = 7),
+        axis.text.x = element_text(size = 6.5),
+        plot.margin = margin(5.5, 3, 0, -1, unit = 'pt')
+      )
+
+    if (i %in% c(1, 4)) {
+      h2_plots[['sigma']] <- gg_tmp_sigma
+    } else {
+      h2_plots[['sigma']] <- gg_tmp_sigma + theme(
+        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
+        axis.title.y = element_blank()
+      )
+    }
+
+  } else {
+    tmp_df <- h2_profiles %>% 
+      filter(prof_var == p) %>% 
+      # filter(logLik >= -21957.26 - 4) %>%
+      arrange(p)
+
+    vals <- tmp_df[, p]
+
+    profile_like <- splinefun(x = tmp_df[, p], y = tmp_df[, 'logLik'])
+    max_profile <- optimize(profile_like, interval = range(vals), maximum = TRUE)
+    threshold <- max_profile$objective - 1.92
+    plot_threshold <- max_profile$objective - 6
+    profile_roots <- function(x) {profile_like(x) - threshold}
+    plot_limits <- function(x) {profile_like(x) - plot_threshold}
+
+    lower_bound <- uniroot(
+      profile_roots, 
+      interval = c(min(vals), max_profile$maximum)
+    )$root
+
+    upper_bound <- uniroot(
+      profile_roots, 
+      interval = c(vals[which.max(tmp_df$logLik)], max(vals))
+    )$root
+
+    add_piece <- c(
+      'Mu' = 3000, 'BetaW' = 0.3, 'phase' = 0.2, v = 0.5 
+    )
+
+    max_x <- uniroot(
+      plot_limits, 
+      interval = c(max_profile$maximum, max(vals) + add_piece[p])
+    )$root
+
+    min_x <- uniroot(
+      plot_limits, 
+      interval = c(min(vals) - add_piece[p], max_profile$maximum)
+    )$root
+
+    gg_tmp <- ggplot() + 
+      geom_point(
+        data = h2_profiles %>% 
+          filter(prof_var == p) %>% 
+          filter(.data[[p]] <= max_x & .data[[p]] >= min_x) ,
+        aes(x = .data[[p]], y = logLik)
+      ) + 
+      geom_function(fun = profile_like) + 
+      geom_vline(xintercept = lower_bound, linetype = 'dashed') + 
+      geom_vline(xintercept = upper_bound, linetype = 'dashed') + 
+      geom_vline(xintercept = max_profile$maximum, col = 'blue') + 
+      labs(y = 'Log-Likelihood', title = my_lab) + 
+      theme(
+        axis.text = element_text(size = 8),
+        axis.title = element_text(size = 10)
+      ) +
+      theme_bw() +
+      # scale_y_continuous(limits = c(-21963, -21957)) + 
+      scale_y_continuous(limits = c(max(h2_profiles$logLik)-6, max(h2_profiles$logLik))) +
+      scale_x_continuous(limits = c(min_x, max_x)) + 
+      theme(
+        axis.title.x = element_blank(), 
+        plot.title = element_text(
+          hjust = 0.5, size = 8, margin = margin(t = 0, unit = "pt")
+        ),
+        axis.title.y = element_text(size = 8),
+        axis.text.y = element_text(size = 7),
+        axis.text.x = element_text(size = 6.5),
+        plot.margin = margin(5.5, 3, 0, -1, unit = 'pt')
+      )
+
+    if (i %in% c(1, 4)) {
+      h2_plots[[p]] <- gg_tmp
+    } else {
+      h2_plots[[p]] <- gg_tmp + theme(
+        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
+        axis.title.y = element_blank()
+      )
+    }
+  }
+}
 @
+
+\begin{figure}[ht]
+<<m2Profs, fig.height=3, fig.width=5, echo=FALSE, warning=FALSE, message=FALSE>>=
+cowplot::plot_grid(
+  plotlist = h2_plots, ncol = 3,
+  align = 'h',
+  rel_widths = c(1.15, 1, 1)
+)
+@
+\caption{\label{m2Profs}MCAP confidence intervals for Model 2 parameters.}
+\end{figure}
 
@@
     theme_bw() +
-    # scale_y_continuous(limits = c(-17450, max(H3_profiles$logLik))) + 
-    scale_y_continuous(limits = c(-17450, max(H3_MLE_evals$logLik))) +
+    scale_y_continuous(limits = c(-17450, max(H3_profiles$logLik))) +
+    # scale_y_continuous(limits = c(-17450, max(H3_MLE_evals$logLik))) +
     theme(
@@
 @
-\caption{\label{m1Profs}MCAP confidence intervals for Model 1 parameters.}
+\caption{\label{m3Profs}MCAP confidence intervals for Model 1 parameters.}
 \end{figure}

### 2024-02-07 H2profileAll.R
Removed unused file

@@ -1,382 +1,0 @@
-# Fitting profiles for Haiti 2.
-#
-# ********
-# NOTE:
-#
-#   If you would like to run this code but have never used the `batchtools` R
-#   package, then we strongly recommend completing both Examples 1 and 2 from
-#   the batchtools vignette, by first installing the `batchtools` package and
-#   running: `vignette('batchtools')`, or visiting:
-#
-#         https://mllg.github.io/batchtools/articles/batchtools.html
-#
-# ********
-#
-# The fit_haiti2 function is a very simple helper that will take as input
-# some starting parameters and an objective function for Model 2 and will
-# return parameters estimates along with corresponding estimates of their
-# log-likelihood.
-#
-# This this script we use this helper function to do a profile likelihood search
-# of the estimated parameters using the batchtools package. To do this, we will
-# create a grid of parameter values and submit jobs where each job will use a
-# row from the grid as a starting point to fit the model.
-
-
-# Load Packages -----------------------------------------------------------
-
-library(batchtools)
-library(tidyverse)
-library(data.table)
-library(haitipkg)
-library(pomp)
-library(spatPomp)
-library(subplex)
-
-# Create Experiment Registry ----------------------------------------------
-
-reg <- makeExperimentRegistry(
-  file.dir ='model2/profileInit',
-  seed = 739164,
-  packages = c("spatPomp", 'haitipkg', 'pomp')
-)
-
-# More than one start could be considered, but our preliminary results 
-# suggest that only one is needed. 
-n_starts <- 3
-
-# Create Profile Design Matrix --------------------------------------------
-
-set.seed(22446688)
-h2 <- haiti2(cutoff = 10000, measure = "log")
-
-# prof_params <- c(
-#   "Mu",
-#   "Beta",
-#   "BetaW",
-#   "v",
-#   "sigma",
-#   "phase",
-#   paste0("InitInfected", 1:10),
-#   'Delta',
-#   'Rho', 
-#   'redmu', 
-#   'AlphaS'
-# )
-
-prof_params <- c(
-  paste0("InitInfected", 1:10)
-)
-
-final_pars <- coef(h2)
-prof_vars <- c()
[…]

### 2024-02-08 edits.tex
Major edits to the introduction

@@
 
+%They recommend five guiding principles: stakeholder engagement, complete model documentation, complete description of data used, communicating uncerainty, and testable model outcomes. 
+\newcommand\editModelingGuidanceFirst{There are many guidelines available regarding the use of mechanistic models for policy influence.
+For instance, Behrend et al.~\cite{behrend20} conducted an extensive literature review on modeling principles and standards.
+They used this review to develop general recommendations for policy-guiding modeling.
+This review is particularly relevant given that their recommendations are specifically intended for the modeling of neglected tropical diseases, which includes cholera.
+Their resulting recommendations for communicating model results align with other manifestos on the proper use of models intended to inform policy \cite{saltelli20,donnelly18}, each emphasizing the importance of involving stakeholders in the modeling process, transparency, and communicating uncertainty.
+Other recommendations that focus on issues of model calibration and assessment also exist \cite{dahabreh17,egger17,penaloza15}, demonstrating the considerable amount of published guidance on effectively modeling dynamic systems.}
+
+\newcommand\editModelingGuidanceSecond{Based on our assessment and literature review, \cite{lee20} was among the studies that best followed these general recommendations for policy-driven modeling.
+Concretly, \cite{lee20} follows at least four of the five principles outlined in \cite{behrend20}: \emph{complete model documentation}, \emph{complete description of data used}, \emph{communicating uncertainty}, and \emph{testable model outcomes}.
+Determining the level of adherence to the first principle, \emph{stakeholder engagement}, is difficult based solely on the article.
+Despite this, the inconsistency between their forecasts and the cholera incidence from 2019 to 2022 suggests that existing recommendations and standards related to policy-driven modeling may be insufficient.
+\cite{saltelli19} suggests that improvements in model model-based outputs may be obtained by developing structures and standards based on statistical principles.
+Our general recommendations therefore complement and extend existing guidelines by focusing on the methodological tasks of calibrating and evaluating dynamic models in a rigorous statistical framework.
+}
+
 \newcommand\editMechModels{Mechanistic models representing biological phenomena are valuable for epidemiology and consequently for public health policy \cite{lofgren14,mccabe21}.

### 2024-02-08 ms.Rnw
Major edits to the introduction

@@
 and 11 studies that used stochastic models \cite{kirpich16,lee20,pasetto18,mukandavire13,kirpich17,lewnard16,kunkel17,mukandavire15,sallah17,azman12,azman15}.
-Incidence data on the outbreak were available at various spatial scales, which resulted with 17 studies that considered spatially explicit dynamic models \cite{lee20,tuite11,pasetto18,fitzgibbon20,eisenberg13,rinaldo12,chao11,abrams13,trevisin22,sallah17,collins14,kelly16,azman12,leung22,kuhn14,mari15,gatto12}.
+Incidence data on the outbreak are available at various spatial scales, motivating 17 studies in our literature review to consider spatially explicit dynamic models \cite{lee20,tuite11,pasetto18,fitzgibbon20,eisenberg13,rinaldo12,chao11,abrams13,trevisin22,sallah17,collins14,kelly16,azman12,leung22,kuhn14,mari15,gatto12}.
 Here we focus on a multi-group modeling exercise by Lee et al.~\cite{lee20} in which four expert modeling teams developed models to the same dataset with the goal of comparing conclusions on the feasibility of eliminating cholera by a vaccination campaign.}
@@
 Model~4 has an agent-based construction, featuring considerable mechanistic detail but limited ability to calibrate these details to data.
-\new{These modeling strategies were selected by \cite{lee20} to represent the range of approaches used in the research community.}
+\new{These modeling strategies were selected by \cite{lee20} to represent the range of approaches used in the research community}
 We focus on Models~1--3, as the strengths and weaknesses of the agent-based modeling approach \cite{tracy18} are outside the scope of this article.
 \new{In addition, agent-based models were less widely used, the agent based model of \cite{lee20} being the only model of this class that was found in our literature review.}
-\new{The data that were used in \cite{lee20}, and that is the subject of our reanalysis, are displayed in Fig.~\ref{fig:CholeraData}.}
+\new{The data that were used in \cite{lee20}, and that we reanalyze, are displayed in Fig.~\ref{fig:CholeraData}.}
 
@@
 Advances in OCV technology and vaccine availability, however, raised the possibility of planning a national vaccination program \cite{lee20}.
-In the study, certain data were shared between the groups, including demography and vaccination history; vaccine efficacy was also fixed at a shared value between groups.
+\new{The possibility of controling the Haiti cholera outbreak via OCV was considered by various research groups.} \TODO{Citations}. 
+In the \new{Lee et al.~\cite{lee20}} study, certain data were shared between the groups, including demography and vaccination history; vaccine efficacy was also fixed at a shared value between groups.
 Beyond this, the groups made autonomous decisions on what to include and exclude from their models.
-The groups largely adhered to existing guidelines for creating models to inform policy \cite{behrend20,saltelli20} and, despite their autonomy, obtained a consensus that an extensive nationwide vaccination campaign would be necessary to eliminate cholera from Haiti.
-\TODO{How?
-This statement should be epanded or re-evaluated.}
+\old{The groups largely adhered to existing guidelines for creating models to inform policy [35, 36] and, d}\new{D}espite their autonomy, \new{the four independent teams} obtained a consensus that an extensive nationwide vaccination campaign would be necessary to eliminate cholera from Haiti.
 Their forecasts, however, were inconsistent with the prolonged period with no confirmed cholera cases between February, 2019 and September, 2022 \cite{trevisin22}.
-Though cholera has recently reemerged in Haiti \cite{rubin22,PAHO23}, the inability to accurately forecast cholera incidence from 2019-2022---while diligently adhering to existing scientific standards for infectious disease modeling---demonstrates that existing standards of disease modeling and forecasting may be insufficient.
+Though cholera has recently reemerged in Haiti \cite{rubin22,PAHO23}, the inability to accurately forecast cholera incidence from 2019-2022\old{---while diligently adhering to existing scientific standards for infectious disease modeling---demonstrates that existing standards of disease modeling and forecasting may be insufficient.} \new{prompts us to consider retrospectively what may have been done differently in order to obtain more reliable conclusions.}
 
@@
 
-\new{\editMechModels}
-
 %% IT SEEMS THIS PARAGRAPH IS DEPRECATED IN FAVOR OF CITING THESE REFERENCES WHERE APPROPRIATE
@@
 
+\new{\editModelingGuidanceFirst}
+
+\new{\editModelingGuidanceSecond}
+
+\new{Our recommendations are presented in the context of a case study, with the goal of demonstrating how careful adherence to statistical principles may result in improved model fits.}
 We proceed by introducing the general modeling scheme employed by Models~1--3 and provide details of each individual model;
@@
 In our literature review, there were 14 studies that used associative models to describe various aspects of the cholera epidemic \cite{hulland19,moise20,piarroux11,matias17,eisenberg13,rebaudet19CATI,ivers15,raila17,cuneo17,charles14,richterman19,richterman19inverse,bengtsson15,li16adjusting}.}
+
+\jwc{This doesn't quite feel like the right place for this either, but I can't seem to pin down exactly where that would be yet. Perhaps the discussion?}
+\new{\editMechModels}
 
@@
 \eic{I'VE STARTED TO PUT IN SOME DISCUSSION OF BAYESIAN METHODS HERE, THOUGH IT COULD ALSO GO IN THE DISCUSSION}
-\new{The main alternative to maximum likelihood estimation is Bayesian inference via Markov chain Monte Carlo, used to analyze the Haiti cholera epidemic by \cite{andrews11,pasetto18,rinaldo12,lewnard16,trevisin22,sallah17,azman12,kuhn14,mari15,gatto12}.
-Alternatively, some investigations obtained parameters from previous sources without calibrating to data.}
+\new{The main alternative to maximum likelihood estimation is Bayesian inference via Markov chain Monte Carlo, used to analyze the Haiti cholera epidemic by \cite{andrews11,pasetto18,rinaldo12,lewnard16,trevisin22,sallah17,azman12,kuhn14,mari15,gatto12}.}
 
@@
 
-\new{In any study involving dynamic models, it is important to consider, test, and present more than one model framework or variant.
-\eic{I'M NOT SURE THIS IS A CONCLUSION.
+% \new{In any study involving dynamic models, it is important to consider, test, and present more than one model framework or variant.
+% \eic{I'M NOT SURE THIS IS A CONCLUSION.
 LEE ET AL STUDIED MULTIPLE MODELS BECAUSE THAT WAS THE GOAL OF THAT PAPER.
@@
 WE WOULD LIKE RESULTS FROM DIFFERENT PAPERS TO HAVE OBJECTIVE MEASURES OF FIT THAT CAN BE COMPARED BETWEEN PAPERS, BUT I DON'T THINK WE WANT TO SAY THAT ALL PAPERS HAVE TO ANALYZE MANY MODELS.}
-In our review of the literature featuring cholera modeling in Haiti, we found that only a handful of papers that utilized mechanistic models reported more than a single model variation. (TODO: Cite).
-We acknowledged the possibility that a study reporting merely a single model may have explored others and only reported the one used in the final analysis.
-However, divulging details about the models that were considered, but not used, may greatly benefit the broader scientific community.
-These details may guide future researchers to avoid model variations that have previously demonstrated to perform poorly on incidence data.
-Furthermore, it can enhance confidence in the reported model's ability to describe the system under investigation.
-This confidence stems from the demonstrated evidence that the chosen model offers a better description of the observed data compared to other reasonable model alternatives.}
+% In our review of the literature featuring cholera modeling in Haiti, we found that only a handful of papers that utilized mechanistic models reported more than a single model variation.
+(TODO: Cite).
+% We acknowledged the possibility that a study reporting merely a single model may have explored others and only reported the one used in the final analysis.
+% However, divulging details about the models that were considered, but not used, may greatly benefit the broader scientific community.
+% These details may guide future researchers to avoid model variations that have previously demonstrated to perform poorly on incidence data.
+% Furthermore, it can enhance confidence in the reported model's ability to describe the system under investigation.
+% This confidence stems from the demonstrated evidence that the chosen model offers a better description of the observed data compared to other reasonable model alternatives.}
 

### 2024-02-09 r1/response.tex
Changed discussion section to use edits, also made edits and response to reviewers comments for clarity

@@
 \usepackage{fullpage}
+\usepackage{bm}
 
@@
 \newcommand\code[1]{\texttt{#1}}
+\newcommand\paramVec{\theta}
 \input{../edits.tex}
@@
 
+In their article, Lee et al.~\cite{lee20} only presented forecasts that were based on simulations from initial conditions.
+These forecasts were generally inconsistent with the absence of cholera cases from 2019-2022, and do not account for recent evidence of cholera dynamics.
+In their supplement material, however, they provide a table (S7) in which they show that if they condition the simulations on most recent data (0 cases from 2019-2020), their forecasts predict cholera elimination.
+This second effort that was only presented in the supplement material is more consistent with what we suggest is correct practice: forecasts of a system should start using latent states that are consistent with available data.
+We have expanded this argument in order to clarify this point:
+
+\article{\editForecastOne}
+
+\article{\editForecastTwo}
+
 \report{
@@
 
+We followed the same approach as Lee et al.~\cite{lee20}, which was requiring that the same number of vaccinations are administered at the national level as they would have been in the spatially-explicit vaccination campaigns.
+The following sentence has been added to clarify this:
+
+\article{\editModVacc}
+
 \bibliography{../bib-haiti.bib,../bib-lit.bib}

### 2024-02-09 edits.tex
Changed discussion section to use edits, also made edits and response to reviewers comments for clarity

@@
 \cite{saltelli19} suggests that improvements in model model-based outputs may be obtained by developing structures and standards based on statistical principles.
-Our general recommendations therefore complement and extend existing guidelines by focusing on the methodological tasks of calibrating and evaluating dynamic models in a rigorous statistical framework.
-}
+Our general recommendations therefore complement and extend existing guidelines by focusing on the methodological tasks of calibrating and evaluating dynamic models in a rigorous statistical framework.}
 
@@
 The predictive skill of non-mechansitic models can reveal limitations in mechanistic models, but cannot readily replace the scientific understanding obtained by describing the biological dynamics of the system in a mathematical model.}
+
+\newcommand\editDiscussionOne{In our cholera case study, we highlight the importance of model diagnostics; they are instrumental in enhancing model fits, thus leading to more trustworthy conclusions.
+We compared the Models 1--3 of \cite{lee20} to simple statistical benchmarks, enabling us to identify deficiencies in the models that could be improved.
+In the case of Model~3, comparing a fitted model to a benchmark and then following up by questioning why the model did not describe the data as efficiently as its benchmark helped us discern that the model fell short in explaining the surge in cholera transmission following Hurricane Matthew.
+This led us to suggest a model modification.
+Because Model~1 outperformed its benchmark model, we were able to confidently scrutinize the implications of the fitted model.
+We found that the seasonality term of Model~1 closely mirrors the seasonal rainfall patterns in Haiti, reaffirming alternative evidence of the importance of rainfall as a driver of cholera infection and bolstering our confidence in Model~1's ability to capture the critical dynamics of the cholera outbreak.}
+
+\newcommand\editDiscussionTwo{Likelihood based inference enables researchers to choose between any model that is evaluated on the same dataset using a metric like AIC.
+Nested model variations are particularly useful as they enable formal statistical testing of the nested features.
+The cholera case study demonstrated the advantage of such a model variation in all three models.
+For instance, we tested the inclusion of a trend in transmission rate over time in Model~1; in Model~2, we considered adding an additional phase parameter, which enables a shift in the seasonal peaks of cholera infections; in Model~3, we included additional parameters such as those associated with Hurricane Matthew.
+In all these cases, we found strong statistical evidence in favor of these model variations over their simpler counterparts.}
+
+\newcommand\editDiscussionThree{Another important consideration when fitting a mechanistic model to a dynamic system is the complexity of the model.
+Mathematical models necessarily simplify complex dynamical systems in order to make inference possible.
+Decision-making about which features of a system need to be incorporated in a model and which should be excluded can significantly influence the model-based conclusions.
+However, these decisions can be tested using a likelihood-based framework.
+Standard model decisions cover deterministic versus stochastic modeling and the decision to include or exclude a meta-population structure.
+Model benchmarks can assist in determining the level of complexity to be used in the model, enabling researchers to gauge a model's relative performance against a better-understood statistical model.}
+
+\newcommand\editDiscussionFour{The three models in this case study offer a unique combination of decisions of model complexity.
+Model~3 provides an interesting example in that it is both stochastic and has a meta-population structure, making it challenging to draw likelihood-based inferences.
+In this paper, we demonstrated how this model class could be calibrated to incidence data using the innovative IBPF algorithm.
+One of only a few examples of fitting a model with this level of complexity, this case study exemplifies the algorithm's potential benefits and provides an example for future researchers on a possible approach to fitting a high-dimensional non-linear model.}
+
+\newcommand\editDiscussionFive{If forecasts are an important component of a modeling task, the most recent information on the dynamic system should be considered with increased importance.
+This is particularly true for long time series, as the latent states from simulations from initial conditions may be more likely to diverge from the truth as time increases.
+One way to address this issue is by simulating forward from the filtering distribution at the last available time point, as this proceedure considers available evidence about the latent states of a system under the framework of the dynamic model.
+This type of forecasting, however, is not possible using a deterministic model.
+Deterministic simulations may also result in over-confidence in model forecasts, as they can only account for uncertainty due to the parameter estimation proceedure \cite{king15}.
+Despite these limitations, forecasts from deterministic models can provide valuable insights on potential future trends, though researchers should be transparent about these limitations when presenting forecasts from deterministic models.}
+
+\newcommand\editModVacc{Because Model~1 only accounts for national level disease dynamics, the pre-determined department-specific vaccination campaigns are carried out by assuming the vaccines are administered in one week to the same number of individuals that would have obtained vaccines if explicitly administered to the specific departments.
+We refer readers to \cite{lee20} and the accompanying supplement material for more details.}
+
+\newcommand\editForecastOne{Forecasts of a dynamic system should ideally be based on the most recent information available as it is likely to be more relevant than older data.
+While this assertion may seem self-evident, it is not the case for deterministic models, for which the initial conditions together with the parameters are sufficient for forecasting, and so recent data do not have special importance.
+Epidemiological forecasts based on deterministic models are not uncommon in practice, despite their limitations \cite{king15}.
+Lee et al.~\cite{lee20} chose to obtain forecasts from all of their models by simulating forward from initial conditions.
+This decision is possibly as a result of using a deterministic model: forecasts from different models may only be considered comparable only if they are obtained in the same way, which must be by simulating from initial conditions because Model~2 is deterministic.}
+
+\newcommand\editForecastTwo{In contrast, for non-deterministic Models~1 and 3, we obtain forecasts by simulating future values using latent states that are harmonious with the most recent data.
+This is done by drawing latent states at the last observation time ($t_N$) from the filtering distribution $f_{\bm{X}_N|\bm{Y}_{1:N}}(\bm{x}_{N} | \bm{y}^*_{1:N} ; \hat\paramVec)$.
+The decision to obtain model forecasts from initial conditions partially explains the unsuccessful forecasts of Lee et al.~\cite{lee20}.
+Table~S7 in their supplement material, which contains results that were not discussed in their main article, shows that the subset of their simulations which were consistent with observing zero cases from 2019-2020 were also more consistent with the disappearance of cholera from Haiti from 2019-2022.
+These results support our argument: forecasts of a dynamic system should start with latent states that are consistent with available data.}

### 2024-02-09 ms.Rnw
Changed discussion section to use edits, also made edits and response to reviewers comments for clarity

@@
 
-\TODO{Re-phrase this paragraph, as suggested by one of the reviewers.}
-Recent information about a dynamic system should be more relevant for a forecast than older information.
+\old{Recent information about a dynamic system should be more relevant for a forecast than older information.
 This assertion may seem self-evident, but it is not the case for deterministic models, for which the initial conditions together with the parameters are sufficient for forecasting, and so recent data do not have special importance.
-Epidemiological forecasts based on deterministic models are not uncommon in practice, despite their limitations \cite{king15}.
-That may explain why Lee et al.~\cite{lee20} chose to obtain forecasts by simulating the calibrated models forward from initial conditions.
-\new{For Models~1 and 3---which are non-deterministic---}we obtain forecasts by simulating future values using latent states that are consistent with the most recent observations; this is done by drawing latent states at the last observation time ($t_N$) from the filtering distribution $f_{\bm{X}_N|\bm{Y}_{1:N}}(\bm{x}_{N} | \bm{y}^*_{1:N} ; \hat\paramVec)$.
-The decision not to do this partially explains the unsuccessful forecasts of Lee et al.~\cite{lee20}: their Table~S7 shows that the subset of their simulations which were consistent with observing zero cases in 2019 also accurately predicted the prolonged absence of detected cholera.
+Epidemiological forecasts based on deterministic models are not uncommon in practice, despite their limitations [34].
+That may explain why Lee et al.~[1] chose to obtain forecasts by simulating the calibrated models forward from initial conditions.
+We obtain forecasts by simulating future values using latent states that are consistent with the most recent observations; this is done by drawing latent states at the last observation time ($t_N$) from the filtering distribution $f_{\bm{X}_N|\bm{Y}_{1:N}}(\bm{x}_{N} | \bm{y}^*_{1:N} ; \hat\paramVec)$.
+The decision not to do this partially explains the unsuccessful forecasts of Lee et al.~[1]: their Table~S7 shows that the subset of their simulations which were consistent with observing zero cases in 2019 also accurately predicted the prolonged absence of detected cholera.}
+
+\new{\editForecastOne}
+
+\new{\editForecastTwo}
 
@@
 Simulations from probabilistic models (Models~1 and~3) represent possible trajectories of the dynamic system under the scientific assumptions of the models.
+\new{\editModVacc}
 Estimates of the probability of cholera elimination can therefore be obtained as the proportion of simulations from these models that result in cholera elimination. 
@@
 
-\new{In our cholera case study, we highlight the importance of model diagnostics; they are instrumental in enhancing model fits, thus leading to more trustworthy conclusions.
-We compared the Models 1--3 of \cite{lee20} to simple statistical benchmarks, enabling us to identify deficiencies in the models that could be improved.
-In the case of Model~3, comparing a fitted model to a benchmark and then following up by questioning why the model did not describe the data as efficiently as its benchmark helped us discern that the model fell short in explaining the surge in cholera transmission following Hurricane Matthew.
-This led us to suggest a model modification.
-Because Model~1 outperformed its benchmark model, we were able to confidently scrutinize the implications of the fitted model.
-We found that the seasonality term of Model~1 closely mirrors the seasonal rainfall patterns in Haiti, reaffirming alternative evidence of the importance of rainfall as a driver of cholera infection and bolstering our confidence in Model~1's ability to capture the critical dynamics of the cholera outbreak.
-}
-
-% \new{In any study involving dynamic models, it is important to consider, test, and present more than one model framework or variant.
-% \eic{I'M NOT SURE THIS IS A CONCLUSION.
-LEE ET AL STUDIED MULTIPLE MODELS BECAUSE THAT WAS THE GOAL OF THAT PAPER.
-MOST OTHER PAPERS STUDY ONE MODEL.
-WE WOULD LIKE RESULTS FROM DIFFERENT PAPERS TO HAVE OBJECTIVE MEASURES OF FIT THAT CAN BE COMPARED BETWEEN PAPERS, BUT I DON'T THINK WE WANT TO SAY THAT ALL PAPERS HAVE TO ANALYZE MANY MODELS.}
-% In our review of the literature featuring cholera modeling in Haiti, we found that only a handful of papers that utilized mechanistic models reported more than a single model variation.
-(TODO: Cite).
-% We acknowledged the possibility that a study reporting merely a single model may have explored others and only reported the one used in the final analysis.
-% However, divulging details about the models that were considered, but not used, may greatly benefit the broader scientific community.
-% These details may guide future researchers to avoid model variations that have previously demonstrated to perform poorly on incidence data.
-% Furthermore, it can enhance confidence in the reported model's ability to describe the system under investigation.
-% This confidence stems from the demonstrated evidence that the chosen model offers a better description of the observed data compared to other reasonable model alternatives.}
-
-\new{Likelihood based inference enables researchers to choose between any model that is evaluated on the same dataset using a metric like AIC.
-Nested model variations are particularly useful as they enable formal statistical testing of the nested features.
-The cholera case study demonstrated the advantage of such a model variation in all three models.
-For instance, we tested the inclusion of a trend in transmission rate over time in Model~1; in Model~2, we considered adding an additional phase parameter, which enables a shift in the seasonal peaks of cholera infections; in Model~3, we included additional parameters such as those associated with Hurricane Matthew.
-In all these cases, we found strong statistical evidence in favor of these model variations over their simpler counterparts.}
-
-\new{Another important consideration when fitting a mechanistic model to a dynamic system is the complexity of the model.
-Mathematical models necessarily simplify complex dynamical systems in order to make inference possible.
-Decision-making about which features of a system need to be incorporated in a model and which should be excluded can significantly influence the model-based conclusions.
-However, these decisions can be tested using a likelihood-based framework.
-Standard model decisions cover deterministic versus stochastic modeling and the decision to include or exclude a meta-population structure.
-Model benchmarks can assist in determining the level of complexity to be used in the model, enabling researchers to gauge a model's relative performance against a better-understood statistical model.}
-
-\new{The three models in this case study offer a unique combination of decisions of model complexity.
-Model~3 provides an interesting example in that it is both stochastic and has a meta-population structure, making it challenging to draw likelihood-based inferences.
-In this paper, we demonstrated how this model class could be calibrated to incidence data using the innovative IBPF algorithm.
-One of only a few examples of fitting a model with this level of complexity, this case study exemplifies the algorithm's potential benefits and provides an example for future researchers on a possible approach to fitting a high-dimensional non-linear model.}
+\new{\editDiscussionOne}
+
+\new{\editDiscussionTwo}
+
+\new{\editDiscussionThree}
+
+\new{\editDiscussionFour}
+
+\new{\editDiscussionFive}
 

### 2024-02-09 r1/response.tex
Finished addressing most of the concerns of reviewer 1

@@
 \newcommand\paramVec{\theta}
+
+\newcommand\seq[2]{{#1}\!:\!{#2}}
+\newcommand\vaccClass{Z}
+\newcommand\vaccCounter{z}
 \input{../edits.tex}
@@
 
-WE CAN TRIM DOWN THE RESPONSE LATER. HERE ARE SOME INITIAL BRAINSTORMING THOUGHTS.
-
 Thanks for the feedback on the framing of our contribution.
@@
 However, the validity of our reanalysis of the Haiti data and models provides critical support to our methodological points.
-\TODO{For the revision, we have edited the abstract and introduction to make explicit how the case study fits into the overall goals of the paper.}
+For the revision, we have edited the abstract and introduction to make explicit how the case study fits into the overall goals of the paper.
 \TODO{In the methods and results sections, we have made explicit how the decisions about what to present in the main text and in the supplement relate to these goals.}
 In the discussion, we return from the concrete consideration of the case study, placing it back in the more general setting of inferring complex dynamics from time series data.
-\TODO{In the revision, we have added more signposts to help the reader see how the points that arise in the discussion generalize the corresponding points arising in the case study.}
+In the revision, we have added more signposts to help the reader see how the points that arise in the discussion generalize the corresponding points arising in the case study.
 
@@
 }
+
+Thank you for this feedback.
+We have made substantial changes to the abstract, author summary, and introduction in order to make the focus of the paper more clear. 
+We have also done a nearly complete rewritting of the discussion section. 
+Specifically, though the content and arguments made remain the same, we specifically point out ways that our case study demonstrates the usefulness of our general suggestions for model based inference of dynamic systems.
 
@@
 So, although the method is not entirely new, its applicability to this task is novel.
-\TODO{We have clarified this in the manuscript WHERE?}
+Additional commentary to this point has been added in the abstract, introduction, and \TODO{the subsection on fitting Model 3}.
+
+We have further modified the statement in the abstract that led to this confusion. Rather than stating that we ``develop data analysis strategies leading to improved statistical fit", we now say:
+
+\article{We assess current methodological approaches to these issues via a case study of the 2010-2019 cholera epidemic in Haiti.
+We consider three dynamic models developed by expert teams to advise on vaccination policies.
+We evaluate previous methods used for fitting these models, and we demonstrate modified data analysis strategies leading to improved statistical fit.
+Speficially, we present appraoches of diagnosing model misspecification and the consequent development of improved models.
+We futher demonstrate the utility of recent advances in likelihood maximization for high-dimensional nonlinear dynamic models, enabling likelihood-based inference for high-dimensional models for spatiotemporal incidence data.}
+
+% We have clarified this in the abstract and introduction by mentioning that we demonstrate the utility of recent advances in likelihood maximization for high-dimensional nonlinear dynamic models. 
+% \TODO{Furthermore, we have added additional commentary that clearly indicates that although this case study is among the first to use the nov
 
@@
 That statement arose in the author summary, where references are not expected.
-However, we appreciate that the point should be expanded on, and we have done so in the revision \TODO{}.
-We have cited \cite{he10,stocks20,li23} for examples demonstrating situations where overdispersed dynamic noise is present, and \cite{king15} for an investigation of the statistical limitations of fitting deterministic models to data.
+However, we appreciate that the point should be expanded on, and we have done so in the revision.
+Specifically, we cite papers from the literature review we performed on the use of dynamic models for analyzing the Haiti cholera oubreak.
+The literature review was previously mentioned, but citations from the review were not made.
+Furthermore we modified the quoted statement in the author summary to be the following:
+
+\article{we use an in-depth case study to motivate general procedures for fitting mathematical models to time series data, while explaining how these procedures help to address limitations of previous approaches.}
+
+Among the most common mistakes in mechanistic modeling, in our opinion, is not properly assessing / evaluating a fitted model. 
+We originally tried to make this point by having an entire section dedicated to model diagnostics, though we admit that the connection between the statement in the author summary and a section later in the article was not clear.
+\TODO{We have addressed this by adding sign-posts to references and suggestions back to this statement, and other similar statements in the abtract / introduction.} 
+We have also added citations to \cite{he10,stocks20} \TODO{and \cite{li23}} for examples demonstrating situations where overdispersed dynamic noise is present, and \cite{king15} for an investigation of the statistical limitations of fitting deterministic models to data.
 
@@
 
-We have added some material in the discussion connecting our results to the stated goals of the paper from the abstract and introduction.
-We could have gone further and discussed in more detail the relationship to the references [4, 5], following this recommendation.
-However, a subsequent manuscript addresses [4,5] more directly \cite{li23} and so we have referred the reader to that paper (currently available on arXiv) for additional discussion of this point.
+We have added two paragraphs in the introduction discussing how recommendations we make in this article fit in the broader category of papers that make recommendations on fitting mechanistic models to dynamic systems.
+This includes discussing these two papers that we originally cited in more detail, as well as other articles on similar topics:
 
-\eic{IF THE P0INTS OF 4, 5 CAN BE CONNECTED DIRECTLY TO THIS DATA ANALYSIS, WE OUGHT TO BE ABLE TO DISCUSS THAT WITHOUT CONFLICTING WITH LI23.
-WE SHOULD PROBABLY RE-READ THESE REFERENCES AND FIGURE IT OUT.}
+\article{\editModelingGuidanceFirst}
 
-\eic{ANOTHER IDEA: WE CAN VIEW OUR PAPER AND LI23 AS SYNERGISTIC RATHER THAN IN COMPETITION.
-WE COULD EVEN TALK ABOUT HOW THAT DIFFERENT SETTING (COVID, WITH AN EMPHASIS ON SPATIOTEMPORAL MODELING AND COMPARING PRIMARILY WITH ENSEMBLE FILTER METHODS) LEADS TO SIMILAR CONCLUSIONS TO HAITI CHOLERA.
-FROM THAT PERSPECTIVE, THE PAPERS STRENGTHEN EACH OTHER.
-AS LONG AS BOTH LISTS OF RECOMMENDATIONS ARE FIRMLY TIED TO THE RESPECTIVE DATA ANALYSES, IT SHOULD BE A STRENGTH NOT A WEAKNESS IF THEY OVERLAP.
-BUT IF WE GO THAT ROUTE WE'LL HAVE TO EXPLAIN THIS CONVINCINGLY.}
+\article{\indent\editModelingGuidanceSecond}
+
+We have also added some material in the discussion connecting our results to the stated goals of the paper from the abstract and introduction.
+We could have gone further and discussed in more detail the relationship to the references \cite{behrend20,saltelli20}, following your recommendation.
+However, our newly added reference \cite{li23} more directly addresses \cite{behrend20,saltelli20} and so we have referred the reader to that paper (currently available on arXiv) for additional discussion of this point.
 
@@
 
+It could be either of the above, though in the analysis we only consider compartments based on the vaccination scenarios outlined in the Forecasts section. 
+For Model~1, the assumption made by Lee et al.~\cite{lee20}, and reused in our analysis, is that the added dynamics of the vaccination compartments is an increased probability that an infection is asymptomatic, as described in a paragraph that follows the quoted text.
+a clause has been added to this sentence in order to clarify this point:
+
+\article{Here, $\vaccCounter=0$ corresponds to unvaccinated individuals, and $\vaccCounter \in \seq{1}{\vaccClass}$ describes hypothetical vaccination programs, where each program $\vaccCounter$ indexes differences in both the number of doses administered (one versus two doses per individual) and the round of vaccine administration.}
+
 \report{5.
@@
 
+This parameter comes from Lee et al.~\cite{lee20}, and the only details that they provide is a subscript comment in one of their tables (S13) in their supplement: ``$v_{rate}$ was chosen to mimic the distribution of infections in Haiti".
+They do not, however, list it as a parameter that was ``fit" in the same table, making it unclear where the value comes from. 
+Presumably, it was chosen somewhat subjectively based on their understanding of cholera dynamics in Haiti, because in order to obtain a fit for their model they had to have fixed the value at something without actually estimating it, and perhaps only considered a very small set of possible values.
+Notably, this parameter can mathematically be estimated like all other parameters in the model, but their specific implementation of Model~2 does not permit its estimation.
+Parameters that are partially estimated via subjective measures is outside the scope of our article, so to avoid this the revised manuscript simply reads:
+
+\article{$v_{\mathrm{rate}}= 10^{-12} \,\mbox{km$^2$yr$^{-1}$}$ was fixed at the value used by \cite{lee20}.}
+
 \report{7.
@@
 }
+
+This change has been made.
 
@@
 
-\TODO{This is right, as long as there are not any initialization issues...}
+It is correct that models fit to the same datasets can be directly compared using the AIC values.
+Models~2 and 3 are not fit to the same datasets, however, due to the way Model~3 is initialized.
+To initialize the latent states of Model~3, Lee et al~\cite{lee20} used approximately $60\%$ of the available data as ``burn-in" observations to initialize latent states.
+According to their supplement, this was done primarily because they believed the dynamics from the start of the outbreak were significantly different enough from the dynamics at the end of the outbreak to suggest that two seperate models would be needed to appropriately describe the data.
+We feel that a properly designed and calibrated model should be able to explain the entire dynamics of the outbreak, including the change from what Lee et al~\cite{lee20} referred to as the ``epidemic" and ``endemic" periods of the outbreak.
+
+One possible approach to re-analyzing their model would then be to fit all available observations; this would make Models~2 and 3 directly comparable, but would require a significant change to Model~3 itself, as the initialization model used by Lee et al~\cite{lee20} requires a ``burn-in" period.
+One of our goals, however, was to demonstrate that the models provided by Lee et al.~\cite{lee20} were not the issue, and a complete refactorization of the initialization could potentially contradict this objective.
+Therefore instead we decided to substantially shrink the ``burn-in" period to only the first three available observations, which corresponds to a natural change in the observations themselves \cite{barzilay13}.
+These details are available in the Initial Values section in the supplement material.
+
+In short, though models that are fit to the same dataset are directly comparable using AIC, this is not the case for Models~2 and 3, as the first few observations in the dataset were not considered likelihood evalution of Model~3.
+This also explains the slight discrepency in AIC for the benchmark models of these two models (Table 2).
+Without going into too much detail in the main article, we added the following sentence to clarify this point:
+
+\article{Models that are fit to the same datasets can be directly compared using AIC values, making it a useful tool to compare to benchmark models.
+Though Models~2 and 3 are both fit to department level incidence reports, their AIC values are not directly comparable due to the way Model~3 initializes latent states, which is described in detail in Sec.
+S5 of the supplement.}
 
@@
 
-Our views align with the referee's comments, and we have rewritten the article to make this clearer.
+Our views align with the referee's comments, and we have rewritten the article to make this clearer. The revised paragraph is now:
+
+\article{If a mechanistic model including a feature (such as a representation of a mechanism, or the inclusion of a covariate) fits better than mechanistic models without that feature, and also has competitive fit compared to associative benchmarks, this may be taken as evidence supporting the scientific relevance of the feature.
+As for any analysis of observational data, we must be alert to the possibility of confounding.
+For a covariate, this shows up in a similar way to regression analysis: the covariate under investigation could be a proxy for some other unmodeled or unmeasured covariate.
+For a mechanism, the model feature could in principle explain the data by helping to account for some different unmodeled phenomenon.}
+
+\article{The statistical evidence of a trend in transmission rate in this model could be explained by any trending variable (such as hygiene improvements, or changes in population behavior), resulting in confounding from collinear covariates.
+Because we only have access to incidence data, the decrease in observed cases over the course of the outbreak may alternatively be attributed to a trend in other model parameters, such as a decrease in the reporting rate over time.
+This results in confounded mechanisms, as likelihood-based comparisons of a model that permits a change in reporting rate against a model that permits a change in transmission rate may not yield definitive results.
+In such a case, the robust statistical conclusion is that a model which allows for change fits better than one which does not, but the incidence data themselves may not provide enough information to pin down the mechanism.
+In this case study, we did not explicitly test if the incidence data favor a trend in transmission over a trend in reporting rate.
+We argue that a decreasing transmission rate is a plausible way to explain the decrease in cases over time, as there is alternative evidence that supports this model \cite{rebaudet19CATI,rebaudet21,michel19}.
+Additionally there is little scientific evidence to support a decrease in reporting rate (for example, see Figure~1 of \cite{rebaudet21}).}
 
@@
 }
+
+The revised version of the paper contains significant edits to the introduction and discussion sections that we feel address this concern.
 
@@
 For the particular issue of black box versus white box models, we have included the following paragraph
-\TODO{where?} \jwc{How does the Materials and Methods: Mechanistic models for disease modeling sub-section sound?}
 
 \article{\editMechModels}
+
+This addition is intended to compliment the discussion about mechanistic (white-box) vs associative (black-box) models that was provided at the start of the ``Materials and methods" section.
 
@@
 \eic{HOW DOES THIS PLAY OUT IN THE MS? HOW DO WE CLARIFY IT?}
+\jwc{It doesn't immediately come into play in our analysis a lot, so it might be good to point to a reference suce as saltelli19, who has a sketch of what they believe on dynamic modeling: adding model complexity is good because it becomes more realistic, but too much complexity results in more ``propogation error"}
 

### 2024-02-09 edits.tex
Finished addressing most of the concerns of reviewer 1

@@
 Their resulting recommendations for communicating model results align with other manifestos on the proper use of models intended to inform policy \cite{saltelli20,donnelly18}, each emphasizing the importance of involving stakeholders in the modeling process, transparency, and communicating uncertainty.
-Other recommendations that focus on issues of model calibration and assessment also exist \cite{dahabreh17,egger17,penaloza15}, demonstrating the considerable amount of published guidance on effectively modeling dynamic systems.}
+Other recommendations that focus on issues of model calibration and assessment also exist \cite{dahabreh17,egger17,penaloza15}, demonstrating the considerable amount of published guidance on how to effectively model dynamic systems.}
 
@@
 \cite{saltelli19} suggests that improvements in model model-based outputs may be obtained by developing structures and standards based on statistical principles.
-Our general recommendations therefore complement and extend existing guidelines by focusing on the methodological tasks of calibrating and evaluating dynamic models in a rigorous statistical framework.}
+Our general recommendations therefore complement and extend existing guidelines by focusing on the methodological tasks of calibrating and evaluating dynamic models in a rigorous statistical framework.
+We specifically emphasize principles that proved essential in our case study; complementary methodological suggestions are given in \cite{li23}.}
 

### 2024-02-09 ms.Rnw
Finished addressing most of the concerns of reviewer 1

@@
 Specifically, we present approaches of diagnosing model misspecification \old{, development of alternative models, and computational improvements in optimization, in the context of likelihood-based inference on nonlinear dynamic systems.} \new{and the consequent development of improved models.
-We further demonstrate the utility of recent advances in likelihood maximization for high-dimensional nonlinear dynamic models, enabling likelihood-based inference for high-dimensional models of spatiotemporal incidence data.}
+We further demonstrate the utility of recent advances in likelihood maximization for high-dimensional nonlinear dynamic models, enabling likelihood-based inference for high-dimensional models for spatiotemporal incidence data.}
 Our workflow is reproducible and extendable, facilitating future investigations of this disease system.
@@
 The latent state vector $\bm{X}(t) = \big(S_{\vaccCounter}(t),  E_{\vaccCounter}(t), I_{\vaccCounter}(t), A_{\vaccCounter}(t), R_{\vaccCounter}(t), \vaccCounter \in 0:\vaccClass\big)$ describes susceptible, latent (exposed), infected (and symptomatic), asymptomatic, and recovered individuals in vaccine cohort $\vaccCounter$ at time $t$.
-Here, $\vaccCounter=0$ corresponds to unvaccinated individuals, and $\vaccCounter \in \seq{1}{\vaccClass}$ describes hypothetical vaccination programs.
+Here, $\vaccCounter=0$ corresponds to unvaccinated individuals, and $\vaccCounter \in \seq{1}{\vaccClass}$ describes hypothetical vaccination programs\new{, where each program $\vaccCounter$ indexes differences in both the number of doses administered (one versus two doses per individual) and the round of vaccine administration.}
 The force of infection is
@@
 \label{model2:mu_SE}
-\mu_{S_{u\vaccCounter}E_{u\vaccCounter}} &=& \vaccineEfficacy_\vaccCounter \, \lambda,
+\mu_{S_{u\vaccCounter}E_{u\vaccCounter}} &=& \new{(1 - \vaccineEfficacy_\vaccCounter) \, \lambda_u(t)},
 \\
@@
 where $\mathrm{Pop}_u$ is the mean population for department $u$,
-$D_{u\varv}$ is a distance measure estimating average road distance between randomly chosen members of each population, and $v_{\mathrm{rate}}= 10^{-12} \,\mbox{km$^2$yr$^{-1}$}$  was treated as a fixed constant.
+$D_{u\varv}$ is a distance measure estimating average road distance between randomly chosen members of each population, and $v_{\mathrm{rate}}= 10^{-12} \,\mbox{km$^2$yr$^{-1}$}$ \old{was treated as a fixed constant.}\new{was fixed at the value used by \cite{lee20}.}
 In Eq.~\myeqref{model2:water_transport}, $T^W_{u\varv}$ is a measure of river flow between departments.
 The unit of $W_u(t)$ is cells per ml, with dose response modeled via a saturation constant of $\Wsat$ in Eq.~\myeqref{model2:lambda}.
-In Eq.~\myeqref{model2:mu_SE}, $\vaccineEfficacy_\vaccCounter$ denotes the vaccine efficacy for each vaccination campaign $\vaccCounter \in \vaccClass$, with $\vaccineEfficacy_0 = 1$, $\vaccineEfficacy_1 = 1 - 0.429 \childReduce$, $\vaccineEfficacy_2 = 1 - 0.519 \childReduce$, $\vaccineEfficacy_3 = 1 - 0.429$, and $\vaccineEfficacy_4 = 1 - 0.519$
+In Eq.~\myeqref{model2:mu_SE}, $\vaccineEfficacy_\vaccCounter$ denotes the vaccine efficacy for each vaccination campaign $\vaccCounter \in \vaccClass$, with \old{$\vaccineEfficacy_0 = 1$, $\vaccineEfficacy_1 = 1 - 0.429 \childReduce$, $\vaccineEfficacy_2 = 1 - 0.519 \childReduce$, $\vaccineEfficacy_3 = 1 - 0.429$, and $\vaccineEfficacy_4 = 1 - 0.519$} \new{$\vaccineEfficacy_0 = 0$, $\vaccineEfficacy_1 = 0.429 \childReduce$, $\vaccineEfficacy_2 = 0.519 \childReduce$, $\vaccineEfficacy_3 = 0.429$, and $\vaccineEfficacy_4 = 0.519$}
 Here, $\childReduce = 0.4688$ represents the reduced efficacy of the vaccination for children under the age of five years, and the values $0.429$ and $0.519$ are the median effectiveness of one and two doses over their effective period respectively, according to Table S4 in the supplement material of Lee et al.~\cite{lee20}.
@@
 In reality, it is plausible that some bacteria or infected individuals were present in unit $u$ but went unreported.
-Therefore, for departments with zero case counts in week 1, we consider estimating the number of infected individuals rather then treating this value as a constant (see supplement Sec.~S5 for more details).
+Therefore, for departments with zero case counts in week 1, we \old{consider estimating}\new{estimate} the number of infected individuals rather then treating this value as a constant (see supplement Sec.~S5 for more details).
 Finally, reported cholera cases are modeled using a negative binomial measurement model with mean equal to a fraction ($\reportRate$) of individuals in each unit who develop symptoms and seek healthcare
@@
 In the following subsections we describe in detail our approach to calibrating the three proposed mechanistic models to observed cholera incidence data.
-\eic{I'VE STARTED TO PUT IN SOME DISCUSSION OF BAYESIAN METHODS HERE, THOUGH IT COULD ALSO GO IN THE DISCUSSION}
 \new{The main alternative to maximum likelihood estimation is Bayesian inference via Markov chain Monte Carlo, used to analyze the Haiti cholera epidemic by \cite{andrews11,pasetto18,rinaldo12,lewnard16,trevisin22,sallah17,azman12,kuhn14,mari15,gatto12}.}
@@
 The reported results for Model~1 in the remainder of this article were obtained with the inclusion of the parameter $\transmissionTrend$.
-
-\new{This example demonstrates the type of model variation that should be considered in the course of a data analysis: the model variation---the addition of a trend in transmission rate---has a plausible scientific justification, and is easily testable using likelihood based inference methods and observed data.
-This class of model variations are commonly considered in epidemiological modeling \cite{lemaitre19,fitzgibbon20,eisenberg13,rinaldo12,kunkel17}} , \new{but their adaptation are sometimes based on subjective assessments \cite{fitzgibbon20,eisenberg13,kunkel17}} \TODO{get citations}.
-\jwc{Alternatively, we could emphasize the importance of considering model variations, and cite studies from the literature review that only considered a single model, or only reported their final chosen model.
-Although models that were not ultimately selected are not as important, a discussion of which models you tried and why you didn't use them can be beneficial to future researchers who may want to consider similar variations in the future.
-I got to article 35 checking for those with multiple models considered.}
+\new{This example demonstrates the type of model variation that should be considered in the course of a data analysis: the model variation---the addition of a trend in transmission rate---has a plausible scientific justification, and is easily testable using likelihood based inference methods and observed data.}
 
@@
 For a mechanism, the model feature could in principle explain the data by helping to account for some different unmodeled phenomenon.
+
 The statistical evidence of a trend in transmission rate in this model could be explained by any trending variable (such as hygiene improvements, or changes in population behavior), resulting in confounding from collinear covariates.
-\jwc{HERE'S SOME INITIAL THOUGHTS ON HOW TO CLARIFY THIS FOR THE REVIEWER}.
 \old{Alternatively, the trend could be attributed to a decreasing reporting rate rather than decreasing transmission rate, resulting in confounded mechanisms.}
 \new{Because we only have access to incidence data, the decrease in observed cases over the course of the outbreak may alternatively be attributed to a trend in other model parameters, such as a decrease in the reporting rate over time.
-This results in confounded mechanisms, as likelihood-based comparisons of a model that permits a change in reporting rate against a model that permits a change in transmission rate may not yield definitive results.}
-The robust statistical conclusion is that a model which allows for change fits better than one which does not\old{---we argue that a decreasing transmission rate is a plausible way to explain this}, but the incidence data themselves \old{do} \new{may} not provide enough information to pin down the mechanism.
-\new{In this case study, we argue that a decreasing transmission rate is a plausible way to explain the decrease in cases over time \cite{rebaudet19CATI,rebaudet21,michel19}; there is little scientific evidence to support a decrease in reporting rate, as the rate of testing increased throughout this time period.} \TODO{cite: Wasn't there a study we thought demonstrated the testing effort increased}?
+This results in confounded mechanisms, as likelihood-based comparisons of a model that permits a change in reporting rate against a model that permits a change in transmission rate may not yield definitive results.
+In such a case, t}\old{T}he robust statistical conclusion is that a model which allows for change fits better than one which does not\old{---we argue that a decreasing transmission rate is a plausible way to explain this}, but the incidence data themselves \old{do} \new{may} not provide enough information to pin down the mechanism.
+\new{In this case study, we did not explicitly test if the incidence data favor a trend in transmission over a trend in reporting rate.
+We argue that a decreasing transmission rate is a plausible way to explain the decrease in cases over time, as there is alternative evidence that supports this model \cite{rebaudet19CATI,rebaudet21,michel19}.
+Additionally there is little scientific evidence to support a decrease in reporting rate (for example, see Figure~1 of \cite{rebaudet21}).}
 
@@
 We chose to use the auto regressive negative binomial model, however, because the large number of weeks with zero recorded cholera cases in department level data makes a benchmark based on a continuous distribution problematic.
-Likelihoods of Models~1--3 and their respective \old{ARMA} benchmark models are provided in Table~\ref{tab:likes}.
+Log-likelihoods and AIC values of Models~1--3 and of their respective \old{ARMA} benchmark models are provided in Table~\ref{tab:likes}.
+\new{Models that are fit to the same datasets can be directly compared using AIC values, making it a useful tool to compare to benchmark models.
+Though Models~2 and 3 are both fit to department level incidence reports, their AIC values are not directly comparable due to the way Model~3 initializes latent states, which is described in detail in Sec.
+S5 of the supplement.}
 
@@
 
-\TODO{Rethink this paragraph.}
+\TODO{The profile confidence interval for the parameter suggests that the true estimate should be zero, but it was unable to get this value due to the parameter transformation.
+Should we manually set the value to zero?
+This result will also affect this paragraph, but only in ways that we already anticipated with the estimate close to zero.}
 Our case study provides an example of this in the parameter estimate for the duration of natural immunity due to cholera infection, $\muRS^{-1}$.

### 2024-02-13 r1/response.tex
Adressing comments from reviewers

@@
 \newcommand\vaccCounter{z}
+\newcommand\muRS{\mu_{RS}}
 \input{../edits.tex}
@@
 Because we only have access to incidence data, the decrease in observed cases over the course of the outbreak may alternatively be attributed to a trend in other model parameters, such as a decrease in the reporting rate over time.
-This results in confounded mechanisms, as likelihood-based comparisons of a model that permits a change in reporting rate against a model that permits a change in transmission rate may not yield definitive results.
-In such a case, the robust statistical conclusion is that a model which allows for change fits better than one which does not, but the incidence data themselves may not provide enough information to pin down the mechanism.
+This results in confounded mechanisms, as one may not formally test all possible ways to include a trending parameter.
+In such a case, the robust statistical conclusion is that a model which allows for change fits better than one which does not.
+Because the incidence data themselves may not provide enough information to pin down the mechanism, a strongly supported conclusion should avoid ruling out untested hypotheses.
 In this case study, we did not explicitly test if the incidence data favor a trend in transmission over a trend in reporting rate.
 We argue that a decreasing transmission rate is a plausible way to explain the decrease in cases over time, as there is alternative evidence that supports this model \cite{rebaudet19CATI,rebaudet21,michel19}.
-Additionally there is little scientific evidence to support a decrease in reporting rate (for example, see Figure~1 of \cite{rebaudet21}).}
+Additionally, there is little scientific evidence to support a decrease in reporting rate (for example, see Figure~1 of \cite{rebaudet21}).}
 
-\report{12.  [I THINK THIS MEANS LINES 553-556, ASSUMING THE NUMBERING MATCHES]
-  Lines 53-56: I push back against this point.
+\report{12.
+  Lines 553-556: I push back against this point.
 The models assume an exponential distribution for vaccine-derived immunity.
@@
 The qualitative comparison we had in mind was the regime with short immunity where many people become infected multiple times during the ten year epidemic, allowing multiple epidemic peaks due to susceptible replenishment.
-\TODO{We have reworded the sentence to avoid making too strong a claim about 10yr immunity being a precise point where the dynamics transition to a regime where reinfection events become negligible.}
+We have reworded the sentence to avoid making too strong a claim about 10yr immunity being a precise point where the dynamics transition to a regime where reinfection events become negligible. 
+The revised paragraph is provided below.
+
+\article{Our case study provides an example of this in the parameter estimate for the duration of natural immunity due to cholera infection, $\muRS^{-1}$.
+Under the framework of Model~2, the best estimate for this parameter is $1.4 \times 10^{11} \, \mathrm{yr}$, suggesting that individuals have effectively permanent immunity to cholera once infected.
+Rather than interpreting this as scientific evidence that individuals have permanent immunity from cholera, this result suggests that Model~2 favors a regime where reinfection events are a negligble part of the dynamics.
+The depletion of susceptible individuals may be attributed to confounding mechanisms---such as localized vaccination programs and non-pharmaceutical interventions that reduce cholera transmission \cite{trevisin22, rebaudet21}---that were not accounted for in the model.
+Perhaps the best interpretation of the estimated parameter, then, is that under the modeling framework that was used, the model most adequately describes the observed data by having a steady decrease in the number of susceptible individuals.
+The weak statistical fit of Model~2 compared to a log-linear benchmark (see Table~2) cautions us against drawing quantitative conclusions from this model.
+A model that has a poor statistical fit may nevertheless provide a useful conceptual framework for thinking about the system under investigation.
+However, a claim that the model has been validated against data should be reserved for situations where the model provides a statistical fit that is competitive against alternative explanations.}
 
@@
 \report{
-  \textbf{underline{Typos}}
+  \textbf{\underline{Typos}}
 
@@
 This is a variation on the issue of ``statistical significance versus practical significance" familiar from introductory statistics textbooks.
-\eic{HOW DOES THIS PLAY OUT IN THE MS? HOW DO WE CLARIFY IT?}
-\jwc{It doesn't immediately come into play in our analysis a lot, so it might be good to point to a reference suce as saltelli19, who has a sketch of what they believe on dynamic modeling: adding model complexity is good because it becomes more realistic, but too much complexity results in more ``propogation error"}
 
@@
 In other contexts, \cite{king15} noted various scientific advantages for including stochasticity in the model; it is undoubtedly part of the biological dynamics, and its omission can lead to over-confident predictions.
+
 In our study, we find that the deterministic model (model 2) has poor statistical fit.
 This can lead to a risk of identifying spurious scientific phenomena: additional deterministic complexity added to the model may help to describe unmodeled stochastic phenomena present in the data, and may therefore spuriously be assessed as a highly statistically significant model improvement.
+We see some evidence of this in our analysis.
+Aside from not being stochastic, the mechanistic structure of Model~2 is far more complex than the other two models considered in our study; we also see the most clear cases of spurious scientific findings, as some estimated parameters favor regions of the parameter space that we would not expect scientifically (see the added supplement material for confidence intervals of model parameters).
+This may partially explain the choice of Lee et al~\cite{lee20} to consider a scientifically unjustifiable jump in latent state values from their epidemic to endemic periods: without this additional flexibility, Model~2 cannot adequately describe cholera dynamics in Haiti while keeping parameter values in scientifically plausible ranges.
+From our experience, this type of a result is quite common: for a deterministic model to adequately account for unmodeled features of a dynamic system, parameters that are estimated are often forced to unexpected regions of the parameter space.
+Adding stochasticity to the model often fixes this issue by capturing unmodeled features in the inherent stochasticity of the model rather than in estimated parameters.
 
 The decision on whether to choose Markov chain Monte Carlo (MCMC) methods is not fully aligned with the decision on whether to include stochasticity in the model.
-In our literature review, of the 9 previous Haiti cholera investigations including stochastic dynamic models \cite{tuite11,mukandavire13,lewnard16,kunkel17,sallah17,azman12,kuhn14,azman15,lee20}, four of them carried out inference via MCMC \cite{lewnard16,sallah17,azman12,kuhn14}.
-\eic{WHAT SORTS OF MCMC ALGORITHMS DID THEY USE?}
+In our literature review, of the 12 previous Haiti cholera investigations including stochastic dynamic models \cite{kirpich17,lee20,pasett018,mukandavire13,kirpich16,lewnard16,kunkel17,mukandavire15,sallah17,azman12,azman15}, four of them carried out inference via MCMC \cite{pasetto18,lewnard16,sallah17,azman12}.
 We agree with the referee that MCMC is more routine for determinstic models than for stochastic models, and advanced MCMC methods such as Hamiltonian Monte Carlo become more readily applicable.
@@
 We acknowledge the existence of useful mathematical tools for studying ODE models, their identifiability and stability and limiting properties.
-Stochastic dynamic models can be reduced to a deterministic model (a so-called determinsitic skeleton) which can usefully be analyzed to help understand the stochastic dynamics \cite{coulson04}.
+Perhaps one example demonstrating the usefulness of ODE models is the Lee et al.~\cite{lee20} study: of the four models considered, the forecasts from Model~2 were most consistent with the actual cholera situation in Haiti.
+Despite their inability to describe incidence data, deterministic models can do a good job at detecting overal trends, like the gradual decline of cholera cases in Haiti. 
+This, coupled with the ease in fitting ODEs to data, makes ODEs a useful tool to gain preliminary insight on a dynamic system. 
+Furthermore, stochastic dynamic models can be reduced to a deterministic model (a so-called determinsitic skeleton) which can usefully be analyzed to help understand the stochastic dynamics \cite{coulson04}.
 We therefore encourage the use of ODE analysis tools whether or not a stochastic model is adopted.
@@
 
-In summary, maximum likelihood (ML) inference via SMC nowaways provides a fairly well estabilished alternative to Bayesian MCMC inference.
+In summary, maximum likelihood (ML) inference via SMC nowadays provides a fairly well estabilished alternative to Bayesian MCMC inference.
 ML appears to be methodologically more convenient for stochastic dynamic models with overdispersion.
@@
 
+We have added several paragraphs in the discussion to address this point. These paragraphs are listed below for reference.
+
+\article{\editDiscussionOne}
+
+\article{\editDiscussionTwo}
+
+\article{\editDiscussionThree}
+
+\article{\editDiscussionFour}
+
+\article{\editDiscussionFive}
+
 \report{
@@
 
+The previous statement has been changed to the following:
+
+\article{Our analysis presents methodology for diagnosing how well a model describes observed data.
+Objective measures are used to evaluate model fit; when objective measures of the goodness-of-fit of a model are combined with reproducibility, a framework emerges for continual improvement to model based inference when revisiting the data.
+This framework promotes scientific discovery and bolsters model-based policy recommendations by simplifying the process for researchers to build off of previous results.}
+
 \report{
@@
 
+\TODO{Add citations based for this paragraph, given below}
+
+Regulation of biological populations is a fundamental topic in epidemiology, ecology, fisheries and agriculture.
+Population dynamics may be nonlinear and stochastic, with the resulting complexities compounded by incomplete understanding of the underlying biological mechanisms and by partial observability of the system variables.
+Quantitative models for these dynamic systems offer potential for designing effective control measures.
+Developing and testing models for dynamic systems, and assessing their fitness for guiding policy, is a challenging statistical task.
+Questions of interest include: What indications should we look for in the data to assess whether the model-based inferences are trustworthy?
+What diagnostic tests and model variations can and should be considered in the course of the data analysis?
+What are the possible trade-offs of increasing model complexity, such as the inclusion of interactions across spatial units?
+
 \report{
@@
 }
+
+This statement has been modified to include more details. It now reads:
+
+\article{Despite their autonomy, the four independent teams obtained a consensus that an extensive nationwide vaccination campaign would be necessary to eliminate cholera from Haiti.
+They also provided a range for the forecasted number of cumulative cases from February 2019 to Februaruy 2024 based on model simulations, with median estimates from each model measured in hundreds of thousands.
+These forecasts are inconsistent with the prolonged period with no confirmed cholera cases between February, 2019 and September, 2022 \cite{trevisin22}.}
 
@@
 
+It is difficult to directly add citations as suggested, as the text that was contained in lines 67-78 was original thought or argued in \cite{lucas76}, which was already cited.
+However, we appreciate the point that an improved discussion about the role of mechanistic vs associative modeling would be beneficial, and would be strengthened by other opinions / evidence on the topic.
+
+To that effect, we have modified the original text to be more clear that the definitions of \emph{associative} versus \emph{mechanistic} models are our own, as the precise difference between these may be different in alternative contexts.
+Furthermore, we provide references from our literature review that demonstrate the usefulness of associative models in some cases, whereas we typically focus on the benefits of mechanistic models in the article.
+Finally, we added a paragraph that includes a discussion and references on mechanistic / associative modeling, which also addresses the concerns of Reviewer 1.
+This paragraph contains several citations, and is included below for reference.
+
+\article{\editMechModels}
+
 \report{
@@
 }
+
+Due to the size and complexity of the table we have elected to put uncertainty estimates elsewhere.
+Specifically, have created an additional supplement material document that outlines how confidence intervals are obtained, along with tables and figures that show parameter uncertainty estimates.
+In the caption for Table~1, we include a reference to this supplement material.
+\TODO{finish this supplement material}.
 
@@
 }
+
+The table you found is the function $\upsilon^\star(t)$, as that is what is used by each of their models.
+Lee et al~\cite{lee20} provide a smoothed version of this function in Figure~2 (B), but the actual function implemented by each of their models (and consequently, our models) is the numeric table itself; in this sense, the smoothed function they show in Figure~2 (B) is an approximation of the true vaccination function that they use, which is provided in Table S4 in their supplement material.
+
+According to Lee et al~\cite{lee20}, this function was obtained by ``fitting a log-linear weighted regression model to the raw data from a published meta-analysis on killed OCV efficacy against medically attended culter-confirmed cholera (figure 2, appendix 3 p 6) \cite{bi17}".
+Aside from this statement, they provided limited details on how they came up with this function.
+Because of this, we unfortunately can't provide much more additional information other than the statement that was already included in our article:
+
+\article{$\upsilon^\star(t)$ is efficacy at time $t$ since vaccination for adults, taken from \cite{lee20}, Table S4, ...}
 
@@
 Our starting point was the formulation of \cite{lee20}.
-Judging by the referee's comment, they will not be suprised to learn that this structure led to difficulties in some situations. \TODO{}
+Judging by the referee's comment, they will not be suprised to learn that this structure led to difficulties in some situations. 
+We did consider the effect of not calibrating initial values. 
+Unsurprisingly, fitting initial values results in a quantitative improvement to model-fit, measured by AIC, but we also found that this had negligable qualitative effects on model-based conclusions. 
+We have included additional material in the supplement for Initialization Models that demonstrates this.
 
@@
 
-The cholera pathogen was recently introduced to Haiti, which facilitates the specification of initial conditions.
+% The cholera pathogen was recently introduced to Haiti, which facilitates the specification of initial conditions.
+Our experience with mechanistic modeling reflects that of the reviewer: calibrating initial values often results in improved dynamics. 
+As such, we would recommend calibrating initial values in most cases; here, doing so may constitute a major change from the initialization model used by Lee et al.~\cite{lee20}.
+Because one of our goals was to demonstrate that the models of \cite{lee20} were alone not responsible for their poor forecasts, we elected to use the same initialization approach to the extent possible.
+
 The goals of inferring initial prevalence is to reduce the number of parameters. 
 The risk is that the form of this assumption could have substantial effects on the dynamics, and hence on the consequences of the analysis.
-Our approaches to addressing this issue involve: (i) looking at residual diagnostic plots to assess time points at which the model is misspecified; (ii) if the initial values seem problematic, calculate the room for improvement by estimating the initial values as free parameters.
+In our supplement material, we described our approach to addressing this issue: (i) looking at conditional log-likelihoods to assess time points at which the model is misspecified; (ii) if the initial values seem problematic, calculate the room for improvement by estimating the initial values as free parameters.
+We found that enforcing model dynamics for initial states worked well for many of the departments, but was problematic for departments where this resulted in zero counts in Infected, Asymptomatic, and Bacterial compartments, leading to the estimation of initial counts of Infected individuals in two departments (Grand'Anse and Nippes).
+Initial values for Asymptomatic and Bacterial compartments were then obtained by enforcing the dynamics of the model using the estimates for Infected individuals, as described in the supplement material.
 
@@
 
+The text has been modified to add additional clarity. The text now reads:
+
+\article{deterministic Model~2 is a special case of a Markov processes solving a stochastic differential equation in the limit as the noise parameter goes to zero.}
+
 \report{
@@
 }
+
+We have added a citation to \cite{king16}, which discusses POMP models and their implementation.
 
@@
 
-This is similar to point 11 of referee 1.
-LET'S DISCUSS THIS.
-Our point is that there are many other ways one might include time-varying effects in the model, and a strongly supported conclusion should avoid ruling out untested hypotheses.
-We can say that the best available model with trending parameters is statistically superior to the best available model with time-constant parameters.
-This frames the evidence in a way that avoids asserting clear support for any one particular way of modeling this effect.
+This suggestion is similar to one posed by Reviewer 1. We have made substantial edits to the paragraph that included this statement that clarifies our point:
 
-R1 mentioned the possibility of reporting rate decreasing instead.
-I'M TRYING TO RECALL OUR REASONING THAT MODEL 3 IN LEE ET AL OVER-EMPHASIZED THIS.
-APART FROM THE POSSIBILITY OF A CASE DEFINITION CHANGE, ONE MIGHT EXPECT REPORTING RATE TO INCREASE?
+\article{The statistical evidence of a trend in transmission rate in this model could be explained by any trending variable (such as hygiene improvements, or changes in population behavior), resulting in confounding from collinear covariates.
+Because we only have access to incidence data, the decrease in observed cases over the course of the outbreak may alternatively be attributed to a trend in other model parameters, such as a decrease in the reporting rate over time.
+This results in confounded mechanisms, as one may not formally test all possible ways to include a trending parameter.
+In such a case, the robust statistical conclusion is that a model which allows for change fits better than one which does not.
+Because the incidence data themselves may not provide enough information to pin down the mechanism, a strongly supported conclusion should avoid ruling out untested hypotheses.
+In this case study, we did not explicitly test if the incidence data favor a trend in transmission over a trend in reporting rate.
+We argue that a decreasing transmission rate is a plausible way to explain the decrease in cases over time, as there is alternative evidence that supports this model \cite{rebaudet19CATI,rebaudet21,michel19}.
+Additionally, there is little scientific evidence to support a decrease in reporting rate (for example, see Figure~1 of \cite{rebaudet21}).}
 
@@
 
+\TODO{Add citations to papers that do a good job discussing specific diagnostics (maybe leneary Malaria plos comp bio paper, and li23.
+Then we can also make a small mention of run levels: SMC + IF2 based methods scale linearly, so try first with small number of computations to help roughly evaluate what computational effort is needed.}
+
 \report{
@@
 }
+
+\TODO{saltelli had a paper about sensitivity analysis, check this and maybe cite / use it.}
 

### 2024-02-13 edits.tex
Adressing comments from reviewers

@@
 Our general recommendations therefore complement and extend existing guidelines by focusing on the methodological tasks of calibrating and evaluating dynamic models in a rigorous statistical framework.
-We specifically emphasize principles that proved essential in our case study; complementary methodological suggestions are given in \cite{li23}.}
+We specifically emphasize principles that proved essential in our case study; complementary methodological suggestions arising from a spatio-temporal analysis of COVID-19 are given in \cite{li23}.}
 
@@
 More broadly, they have useful roles throughout biology, expecially when combined with statistical methods that properly account for stochasticity and nonlinearity \cite{may04}.
-In some situations, modern machine learning methods can outperform mechanistic models on epidemiological forecasting tasks \cite{lau22}.
-The predictive skill of non-mechansitic models can reveal limitations in mechanistic models, but cannot readily replace the scientific understanding obtained by describing the biological dynamics of the system in a mathematical model.}
+In some situations, modern machine learning methods, which are associative by nature, can outperform mechanistic models on epidemiological forecasting tasks \cite{lau22,baker18}.
+The predictive skill of non-mechansitic models can reveal limitations in mechanistic models, but cannot readily replace the scientific understanding obtained by describing the biological dynamics of the system in a mathematical model \cite{baker18}.}
 

### 2024-02-13 ms.Rnw
Adressing comments from reviewers

@@
 We do this by conducting a re-analysis of the 2010 cholera outbreak in Haiti.
-Our analysis presents methodology for diagnosing how well a model describes observed data, and provides careful justification of valid conclusions from the fitted model.
-Objective measures are used to benchmark model fit; when these are combined with reproducibility, a framework emerges for continual improvement when revisiting the data and models.
+Our analysis presents methodology for diagnosing how well a model describes observed data\new{.}\old{, and provides careful justification of valid conclusions from the fitted model.}
+Objective measures are used to \old{benchmark} \new{evaluate} model fit; when \old{these}\new{objective measures of the goodness-of-fit of a model} are combined with reproducibility, a framework emerges for continual improvement \new{to model based inference} when revisiting the data\old{ and models}.
+\new{This framework promotes scientific discovery and bolsters model-based policy recommendations by simplifying the process for researchers to build off of previous results.}
 Our data analysis workflow is supported by recent advances in algorithms, software and hardware, which facilitate statistical fitting of nonlinear stochastic dynamic models to observed incidence data.
@@
 \old{The groups largely adhered to existing guidelines for creating models to inform policy [35, 36] and, d}\new{D}espite their autonomy, \new{the four independent teams} obtained a consensus that an extensive nationwide vaccination campaign would be necessary to eliminate cholera from Haiti.
-Their forecasts, however, were inconsistent with the prolonged period with no confirmed cholera cases between February, 2019 and September, 2022 \cite{trevisin22}.
+\new{They also provided a range for the forecasted number of cumulative cases from February 2019 to Februaruy 2024 based on model simulations, with median estimates from each model measured in hundreds of thousands.}
+\old{Their forecasts, however, were} \new{These forecasts are} inconsistent with the prolonged period with no confirmed cholera cases between February, 2019 and September, 2022 \cite{trevisin22}.
 Though cholera has recently reemerged in Haiti \cite{rubin22,PAHO23}, the inability to accurately forecast cholera incidence from 2019-2022\old{---while diligently adhering to existing scientific standards for infectious disease modeling---demonstrates that existing standards of disease modeling and forecasting may be insufficient.} \new{prompts us to consider retrospectively what may have been done differently in order to obtain more reliable conclusions.}
@@
 \subsection*{Mechanistic models for disease modeling}
-Models that focus on learning relationships between variables in a dataset are called {\it associative}, whereas models that incorporate a known scientific property of the system are called {\it causal} or {\it mechanistic}.
+\new{In this article, we refer to} models that focus on learning relationships between variables in a dataset \new{as} \old{are called} {\it associative}, whereas models that incorporate a known scientific property of the system \old{are called} \new{we call} {\it causal} or {\it mechanistic}.
 The danger in using forecasting techniques which rely on associative models to predict the consequence of interventions is called the Lucas critique in an econometric context.
@@
 \raggedright Process noise (wk\textsuperscript{1/2})
-  & \raggedright $\sigmaProc = (\Sexpr{myround(sqrt(p1["sig_sq_epi"]),2)},
-  \Sexpr{myround(sqrt(p1["sig_sq_end"]),2)})^*$
+  & \raggedright $\sigmaProc = (\Sexpr{myround(p1["sig_sq_epi"],2)},
+  \Sexpr{myround(p1["sig_sq_end"],2)})^*$
   & \multirow{2}{*}{\eqref{model1:lambda}}
@@
 
-Each of the three models considered in this study describes cholera dynamics as a partially observed Markov process (POMP), with the understanding that the deterministic Model~2 is a degenerate case of a stochastic model.
+Each of the three models considered in this study describes cholera dynamics as a partially observed Markov process (POMP) \cite{king16}, with the understanding that the deterministic Model~2 is a \old{degenerate case of a stochastic model} \new{special case of a Markov processes solving a stochastic differential equation in the limit as the noise parameter goes to zero.}
 Each model is indexed by a parameter vector, $\paramVec$, and different values of $\paramVec$ can result in qualitative differences in the predicted behavior of the system.
@@
 Elements of $\paramVec$ can be fixed at a constant value based on scientific understanding of the system, but parameters can also be calibrated to data by maximizing a measure of congruency between the observed data and the model's mechanistic structure.
-Calibrating model parameters to observed data does not guarantee that the resulting model successfully approximates real-world mechanisms, since the model description of the dynamic system may be incorrect and do not change as the model is calibrated to data.
+Calibrating model parameters to observed data does not guarantee that the resulting model successfully approximates real-world mechanisms, since the model description of the dynamic system may be incorrect and \old{do} \new{does} not change as the model is calibrated to data.
 However, the congruency between the model and observed data serves as a proxy for the congruency between the model and the true underlying dynamic system.
@@
 \new{Because we only have access to incidence data, the decrease in observed cases over the course of the outbreak may alternatively be attributed to a trend in other model parameters, such as a decrease in the reporting rate over time.
-This results in confounded mechanisms, as likelihood-based comparisons of a model that permits a change in reporting rate against a model that permits a change in transmission rate may not yield definitive results.
-In such a case, t}\old{T}he robust statistical conclusion is that a model which allows for change fits better than one which does not\old{---we argue that a decreasing transmission rate is a plausible way to explain this}, but the incidence data themselves \old{do} \new{may} not provide enough information to pin down the mechanism.
+This results in confounded mechanisms, as one may not formally test all possible ways to include a trending parameter.
+In such a case, t}\old{T}he robust statistical conclusion is that a model which allows for change fits better than one which does not.\old{---we argue that a decreasing transmission rate is a plausible way to explain this, but} \new{Because} the incidence data themselves \old{do} \new{may} not provide enough information to pin down the mechanism\new{, a strongly supported conclusion should avoid ruling out untested hypotheses.}
 \new{In this case study, we did not explicitly test if the incidence data favor a trend in transmission over a trend in reporting rate.
 We argue that a decreasing transmission rate is a plausible way to explain the decrease in cases over time, as there is alternative evidence that supports this model \cite{rebaudet19CATI,rebaudet21,michel19}.
-Additionally there is little scientific evidence to support a decrease in reporting rate (for example, see Figure~1 of \cite{rebaudet21}).}
+Additionally, there is little scientific evidence to support a decrease in reporting rate (for example, see Figure~1 of \cite{rebaudet21}).}
 
@@
 
-\TODO{The profile confidence interval for the parameter suggests that the true estimate should be zero, but it was unable to get this value due to the parameter transformation.
-Should we manually set the value to zero?
-This result will also affect this paragraph, but only in ways that we already anticipated with the estimate close to zero.}
 Our case study provides an example of this in the parameter estimate for the duration of natural immunity due to cholera infection, $\muRS^{-1}$.
 Under the framework of Model~2, the best estimate for this parameter is $\Sexpr{signif(1/h2_fit$h2_params["sigma"],3)} \, \mathrm{yr}$, suggesting that individuals have effectively permanent immunity to cholera once infected.
-To interpret this result, we bear in mind that the data ranged from 2010-2019, and therefore estimates of immunity longer than $10\, \mathrm{yr}$---the upper end of previous estimates of natural immunity \cite{king08}---effectively result in the same model dynamics.
-The depletion of susceptible individuals may also be attributed to confounding mechanisms---such as localized vaccination programs and non-pharmaceutical interventions that reduce cholera transmission \cite{trevisin22, rebaudet21}---that were not accounted for in the model.
+\new{Rather than interpreting this as scientific evidence that individuals have permanent immunity from cholera, this result suggests that Model~2 favors a regime where reinfection events are a negligble part of the dynamics.}
+\old{To interpret this result, we bear in mind that the data ranged from 2010-2019, and therefore estimates of immunity longer than $10\, \mathrm{yr}$---the upper end of previous estimates of natural immunity [101]---effectively result in the same model dynamics.}
+The depletion of susceptible individuals may \old{also} be attributed to confounding mechanisms---such as localized vaccination programs and non-pharmaceutical interventions that reduce cholera transmission \cite{trevisin22, rebaudet21}---that were not accounted for in the model.
 Perhaps the best interpretation of the estimated parameter, then, is that under the modeling framework that was used, the model most adequately describes the observed data by having a steady decrease in the number of susceptible individuals.

### 2024-02-13 si/confidenceIntervals.Rnw
Refactored SI + Makefile

@@
 options(
-  scipen = 2,
+  scipen = 0,
   help_type = "html",
@@
 }
+
+myround <- function(x, digits = 1) {
+  # taken from the broman package
+  if (digits < 1)
+    stop("This is intended for the case digits >= 1.")
+  if (length(digits) > 1) {
+    digits <- digits[1]
+    warning("Using only digits[1]")
+  }
+  tmp <- sprintf(paste("%.", digits, "f", sep = ""), x)
+  zero <- paste0("0.", paste(rep("0", digits), collapse = ""))
+  tmp[tmp == paste0("-", zero)] <- zero
+  tmp
+}
 @
@@
 In this section we provide confidence intervals for all model parameters, excluding those that take unique values for each spatial unit.
-For each model and parameter, we use profile log-likelihoods to obtain confidence intervals. TODO: Cite.
+For each model and parameter, we use principles of profile likelihood to obtain confidence intervals \cite{pawitan01}.
 Due to the non-linear and stochastic nature of Models~1 and 3, exact evaluations of the profile log-likelihood are difficult to obtain.
 Instead, the log-likelihood at each point of the profile is estimated using via Monte-Carlo based particle filter methods.
-We therefore obtain confidence intervals for the parameters of Model~1 and Model~3 using the Monte Carlo adjust profile (MCAP) algorithm. TODO: Cite.
+We therefore obtain confidence intervals for the parameters of Model~1 and Model~3 using the Monte Carlo adjust profile (MCAP) algorithm \cite{ionides17}.
+
+In each subsection, we provide figures that show the curvature of the profile log-likelihood near the MLE (Figures~\ref{fig:m1Profs}--\ref{fig:m3Profs}). 
+In these figures, the parameter values are shown on the transformed scale in which the profile was calculated.
 
 \subsection{Model~1 parameters}
+
+Parameter estimates for Model~1, along with the MCAP confidence intervals for the estimate, are given in Table~\ref{tab:mod1CI}. 
+Figure~\ref{fig:m1Profs} displays the Monte Carlo evaluations of the profile likelihood values, obtained using a particle filter.
 
@@
 
-H1_profiles %>% slice_max(order_by = logLik) %>% 
+h1_mles <- H1_profiles %>% slice_max(order_by = logLik) %>% 
   select(logLik, all_of(prof_params)) %>% 
@@
     values_to = 'value'
-  ) -> mles
+  )
 @
@@
     # geom_point(
-    #   data = mles %>% filter(prof_var == p), 
+    #   data = h1_mles %>% filter(prof_var == p), 
     #   aes(x = value, y = logLik), col = 'red'
@@
 @
-\caption{\label{m1Profs}MCAP confidence intervals for Model 1 parameters.}
+\caption{\label{fig:m1Profs}MCAP confidence intervals for Model 1 parameters.}
 \end{figure}
+
+\begin{table}[!h]
+\centering
+\caption{\label{tab:mod1CI}Model~1 parameter estimates and their corresponding confidence intervals, obtained via the MCAP algorithm.}
+\vspace{2mm}
+\begin{tabular}{|c|c|c|c|}
+\hline
+\textbf{Mechanism} & \textbf{Parameter} & \textbf{MLE} & $\bm{95\%}$ \textbf{Confidence Interval}
+\\
+\hline
+\hline
+
+ Seasonality & $\transmissionTrend$ & $\Sexpr{myround(h1_mles %>% filter(prof_var == 'betat') %>% pull(value), 3)}$
+   &
+  $(\Sexpr{myround(all_ci %>% filter(prof_var == 'betat') %>% pull(lower), 3)}, \Sexpr{myround(all_ci %>% filter(prof_var == 'betat') %>% pull(upper), 3)})$ 
+\\
+\hline
+
+ Seasonality & $\transmission_1$ & $\Sexpr{myround(h1_mles %>% filter(prof_var == 'beta1') %>% pull(value), 3)}$
+   &
+  $(\Sexpr{myround(all_ci %>% filter(prof_var == 'beta1') %>% pull(lower), 3)}, \Sexpr{myround(all_ci %>% filter(prof_var == 'beta1') %>% pull(upper), 3)})$ 
+\\
+\hline
+
+ Seasonality & $\transmission_2$ & $\Sexpr{myround(h1_mles %>% filter(prof_var == 'beta2') %>% pull(value), 3)}$
+   &
+  $(\Sexpr{myround(all_ci %>% filter(prof_var == 'beta2') %>% pull(lower), 3)}, \Sexpr{myround(all_ci %>% filter(prof_var == 'beta2') %>% pull(upper), 3)})$ 
+\\
+\hline
+
+ Seasonality & $\transmission_3$ & $\Sexpr{myround(h1_mles %>% filter(prof_var == 'beta3') %>% pull(value), 3)}$
+   &
+  $(\Sexpr{myround(all_ci %>% filter(prof_var == 'beta3') %>% pull(lower), 3)}, \Sexpr{myround(all_ci %>% filter(prof_var == 'beta3') %>% pull(upper), 3)})$ 
+\\
+\hline
+
+ Seasonality & $\transmission_4$ & $\Sexpr{myround(h1_mles %>% filter(prof_var == 'beta4') %>% pull(value), 3)}$
+   &
+  $(\Sexpr{myround(all_ci %>% filter(prof_var == 'beta4') %>% pull(lower), 3)}, \Sexpr{myround(all_ci %>% filter(prof_var == 'beta4') %>% pull(upper), 3)})$ 
+\\
+\hline
+
+ Seasonality & $\transmission_5$ & $\Sexpr{myround(h1_mles %>% filter(prof_var == 'beta5') %>% pull(value), 3)}$
+   &
+  $(\Sexpr{myround(all_ci %>% filter(prof_var == 'beta5') %>% pull(lower), 3)}, \Sexpr{myround(all_ci %>% filter(prof_var == 'beta5') %>% pull(upper), 3)})$ 
+\\
+\hline
+
+ Seasonality & $\transmission_6$ & $\Sexpr{myround(h1_mles %>% filter(prof_var == 'beta6') %>% pull(value), 3)}$
+   &
+  $(\Sexpr{myround(all_ci %>% filter(prof_var == 'beta6') %>% pull(lower), 3)}, \Sexpr{myround(all_ci %>% filter(prof_var == 'beta6') %>% pull(upper), 3)})$ 
+\\
+\hline
+
+ Observation Variance & $\obsOverdispersion: \mathrm{Epi}$ & $\Sexpr{myround(h1_mles %>% filter(prof_var == 'tau_epi') %>% pull(value) %>% exp(), 3)}$
+   &
+  $(\Sexpr{myround(all_ci %>% filter(prof_var == 'tau_epi') %>% pull(lower) %>% exp(), 3)}, \Sexpr{myround(all_ci %>% filter(prof_var == 'tau_epi') %>% pull(upper) %>% exp(), 3)})$ 
+\\
+\hline
+
+ Observation Variance & $\obsOverdispersion: \mathrm{End}$ & $\Sexpr{myround(h1_mles %>% filter(prof_var == 'tau_end') %>% pull(value) %>% exp(), 3)}$
+   &
+  $(\Sexpr{myround(all_ci %>% filter(prof_var == 'tau_end') %>% pull(lower) %>% exp(), 3)}, \Sexpr{myround(all_ci %>% filter(prof_var == 'tau_end') %>% pull(upper) %>% exp(), 3)})$ 
+\\
+\hline
+
+  Reporting Rate & $\reportRate$ & $\Sexpr{myround(h1_mles %>% filter(prof_var == 'rho') %>% pull(value), 3)}$
+   &
+  $(\Sexpr{myround(all_ci %>% filter(prof_var == 'rho') %>% pull(lower), 3)}, \Sexpr{myround(all_ci %>% filter(prof_var == 'rho') %>% pull(upper), 3)})$ 
+\\
+\hline
+
+  Mixing Exponent & $\mixExponent$ & $\Sexpr{myround(h1_mles %>% filter(prof_var == 'nu') %>% pull(value), 3)}$
+   &
+  $(\Sexpr{myround(all_ci %>% filter(prof_var == 'nu') %>% pull(lower), 3)}, \Sexpr{myround(all_ci %>% filter(prof_var == 'nu') %>% pull(upper), 3)})$ 
+\\
+\hline
+
+  Process noise & $\sigmaProc: \mathrm{Epi}$ & $\Sexpr{myround(h1_mles %>% filter(prof_var == 'sig_sq_epi') %>% pull(value), 3)}$
+   &
+  $(\Sexpr{myround(all_ci %>% filter(prof_var == 'sig_sq_epi') %>% pull(lower), 3)}, \Sexpr{myround(all_ci %>% filter(prof_var == 'sig_sq_epi') %>% pull(upper), 3)})$ 
+\\
+\hline
+
+  Process noise & $\sigmaProc: \mathrm{End}$ & $\Sexpr{myround(h1_mles %>% filter(prof_var == 'sig_sq_end') %>% pull(value), 3)}$
+   &
+  $(\Sexpr{myround(all_ci %>% filter(prof_var == 'sig_sq_end') %>% pull(lower), 3)}, \Sexpr{myround(all_ci %>% filter(prof_var == 'sig_sq_end') %>% pull(upper), 3)})$ 
+\\
+\hline
+
+  Initial Values & $I_{0}(0)$ & $\Sexpr{round(h1_mles %>% filter(prof_var == 'I_0') %>% pull(value))}$
+   &
+  $(\Sexpr{round(all_ci %>% filter(prof_var == 'I_0') %>% pull(lower))}, \Sexpr{round(all_ci %>% filter(prof_var == 'I_0') %>% pull(upper))})$ 
+\\
+\hline
+
+  Initial Values & $E_{0}(0)$ & $\Sexpr{round(h1_mles %>% filter(prof_var == 'E_0') %>% pull(value))}$
+   &
+  $(\Sexpr{round(all_ci %>% filter(prof_var == 'E_0') %>% pull(lower))}, \Sexpr{round(all_ci %>% filter(prof_var == 'E_0') %>% pull(upper))})$ 
+\\
+\hline
+
+\end{tabular}
+\end{table}
+
+\subsection{Model~2 parameters}
+
+Parameter estimates for Model~2, along with the profile likelihood confidence intervals for each estimate, are given in Table~\ref{tab:mod2CI}. 
+Figure~\ref{fig:m2Profs} displays the profile log-likelihood curve near the MLE. 
+In Table~\ref{tab:mod2CI}, the confidence interval for $\muRS^{-1}$, the duration of natural immunity due to cholera infection, is arbitrarily large (going to infinity). 
+This is possible because the parameter that was estimated was $\muRS$, and the true MLE for this parameter is zero (see Figure~\ref{fig:m2Profs}). 
+This suggests that the fitted model favors a regime where reinfection events are not possible. 
+Similarly, the MLE for the parameter $\transmission$, which controls the amount of cholera transmission from human to human, is zero. 
+Because Model~2 fails to describe the incidence data as well as a simple statistical benchmark, we must be careful to not interpret these results as evidence that reinfections and human-to-human infection events do not occur.
+Instead, we may consider this as additional evidence of model mispecification.
 
@@
 h2_profiles <- readRDS("../model2/h2_profiles.rds") %>% as.data.frame()
+h2_mles <- readRDS("../model2/model2_fit.rds")$h2_params
 
@@
 
+h2_plots_lb <- numeric(length(prof_params))
+h2_plots_ub <- numeric(length(prof_params))
+
+names(h2_plots_lb) <- prof_params
+names(h2_plots_ub) <- prof_params
+
 h2_plots <- list()
@@
   } else if (p == 'Beta') {
-    my_lab <- TeX('$\\beta$')
+    my_lab <- TeX('$\\beta^{1/5}$')
   } else if (p == 'BetaW') {
@@
   } else if (p == 'sigma') {
-    my_lab <- TeX('$\\mu_{\\text{RS}}$')
+    my_lab <- TeX('$\\mu^{1/5}_{RS}$')
   } else if (p == 'phase') {
@@
         data = tmp_df_beta %>% filter(prof_var == p) %>% filter(Beta <= max_x_beta^5),
-        aes(x = Beta, y = logLik)
+        aes(x = Beta^(1/5), y = logLik)
       ) + 
-      geom_function(fun = function(x) profile_like_beta(x^(1/5))) + 
-      geom_vline(xintercept = upper_bound_beta^5, linetype = 'dashed') + 
+      geom_function(fun = profile_like_beta) + 
+      # geom_function(fun = function(x) profile_like_beta(x^(1/5))) + 
+      # geom_vline(xintercept = upper_bound_beta^5, linetype = 'dashed') + 
+      geom_vline(xintercept = upper_bound_beta, linetype = 'dashed') +
       geom_vline(xintercept = 0, col = 'blue') + 
@@
       scale_y_continuous(limits = c(max_profile_beta-6, max_profile_beta)) +
-      scale_x_continuous(limits = c(0, max_x_beta^5)) +
+      # scale_x_continuous(limits = c(0, max_x_beta^5)) +
+      scale_x_continuous(limits = c(0, max_x_beta)) + 
       theme(
@@
       )
+
+    h2_plots_lb['Beta'] <- 0
+    h2_plots_ub['Beta'] <- upper_bound_beta^5
 
@@
         data = tmp_df_sigma %>% filter(prof_var == 'sigma') %>% filter(sigma <= max_x_sigma^5),
-        aes(x = sigma, y = logLik)
+        aes(x = sigma^{1/5}, y = logLik)
       ) + 
-      geom_function(fun = function(x) profile_like_sigma(x^(1/5))) + 
-      geom_vline(xintercept = upper_bound_sigma^5, linetype = 'dashed') + 
+      # geom_function(fun = function(x) profile_like_sigma(x^(1/5))) + 
+      # geom_vline(xintercept = upper_bound_sigma^5, linetype = 'dashed') + 
+      geom_function(fun = profile_like_sigma) + 
+      geom_vline(xintercept = upper_bound_sigma, linetype = 'dashed') + 
       geom_vline(xintercept = 0, col = 'blue') + 
@@
       scale_y_continuous(limits = c(max_profile_sigma-6, max_profile_sigma)) +
-      scale_x_continuous(limits = c(0, max_x_sigma^5)) +
+      # scale_x_continuous(limits = c(0, max_x_sigma^5)) +
+      scale_x_continuous(limits = c(0, max_x_sigma)) +
       theme(
@@
       )
+
+    h2_plots_lb['sigma'] <- 0
+    h2_plots_ub['sigma'] <- upper_bound_sigma^5
 
@@
     )$root
+
+    h2_plots_lb[p] <- lower_bound
+    h2_plots_ub[p] <- upper_bound
 
@@
 @
-\caption{\label{m2Profs}MCAP confidence intervals for Model 2 parameters.}
+\caption{\label{fig:m2Profs}MCAP confidence intervals for Model 2 parameters.}
 \end{figure}
+
+\begin{table}[!h]
+\centering
+\caption{\label{tab:mod2CI}Model~2 parameter estimates and their corresponding confidence intervals, obtained via profile likelihood.}
+\vspace{2mm}
+\begin{tabular}{|c|c|c|c|}
+\hline
+\textbf{Mechanism} & \textbf{Parameter} & \textbf{MLE} & $\bm{95\%}$ \textbf{Confidence Interval}
+\\
+\hline
+\hline
+
+ Human to water shedding & $\Wshed$ & $\Sexpr{myround(h2_mles['Mu'], 1)}$
+   &
+  $(\Sexpr{myround(h2_plots_lb['Mu'], 1)}, \Sexpr{myround(h2_plots_ub['Mu'], 1)})$ 
+\\
+\hline
+
+ Water to Human Infection & $\beta_W$ & $\Sexpr{myround(h2_mles['BetaW'], 3)}$
+   &
+  $(\Sexpr{myround(h2_plots_lb['BetaW'], 3)}, \Sexpr{myround(h2_plots_ub['BetaW'], 3)})$ 
+\\
+\hline
+
+ Observation Variance & $\obsOverdispersion$ & $\Sexpr{myround(h2_mles['v'], 3)}$
+   &
+  $(\Sexpr{myround(h2_plots_lb['v'], 3)}, \Sexpr{myround(h2_plots_ub['v'], 3)})$ 
+\\
+\hline
+
+ Seasonality & $\phaseParm$ & $\Sexpr{myround(h2_mles['phase'], 3)}$
+   &
+  $(\Sexpr{myround(h2_plots_lb['phase'], 3)}, \Sexpr{myround(h2_plots_ub['phase'], 3)})$ 
+\\
+\hline
+
+Human to Human Infection & $\transmission$ & $\Sexpr{signif(h2_mles['Beta'], 3)}$\textsuperscript{*}
+   &
+  $[0, \Sexpr{signif(h2_plots_ub['Beta'], 3)})$ 
+\\
+\hline
+
+Immunity (yr) & $\muRS^{-1}$ & $\Sexpr{signif(1/h2_mles['sigma'], 3)}$\textsuperscript{*}
+   &
+  $(\Sexpr{signif(1/h2_plots_ub['sigma'], 3)}, \inf)$ 
+\\
+\hline
+
+\end{tabular}
+\begin{flushleft} 
+\textsuperscript{*}As evident in Figure~\ref{fig:m2Profs}, the true MLE for these parameters is $0$ and $\infty$, respectively; this value could not be obtained numerically due to the parameter transformation applied to the parameter for the model fitting processes.
+\end{flushleft}
+\end{table}
+
+\subsection{Model~3 parameters}
+
+Parameter estimates for Model~3, along with the MCAP confidence intervals for the estimate, are given in Table~\ref{tab:mod3CI}. 
+Figure~\ref{fig:m3Profs} displays the Monte Carlo evaluations of the profile likelihood values, obtained using a particle filter.
 
@@
   rename_with(~ sub("1$", "", .x)) %>%
+  mutate(mu_B = mu_B / 52) %>% 
   select(-epsilon, -XthetaA) %>% 
@@
 # This chunk of code returns the MLE for model 3
-H3_MLE_evals %>%
+h3_mles <- H3_MLE_evals %>%
   slice_max(order_by = logLik, n = 1) %>%
@@
   if (p == 'mu_B') {
-    my_lab <- TeX('$\\zeta')
+    my_lab <- TeX('$\\delta_W$')
   } else if (p == 'XthetaA') {
@@
     # geom_point(
-    #   data = mles %>% filter(prof_var == p),
+    #   data = h3_mles %>% filter(prof_var == p),
     #   aes(x = value, y = logLik), col = 'red'
@@
 @
-\caption{\label{m3Profs}MCAP confidence intervals for Model 1 parameters.}
+\caption{\label{fig:m3Profs}MCAP confidence intervals for Model 3 parameters.}
 \end{figure}
+
+\begin{table}[!h]
+\centering
+\caption{\label{tab:mod3CI}Model~3 parameter estimates and their corresponding confidence intervals, obtained via the MCAP algorithm.}
+\vspace{2mm}
+\begin{tabular}{|c|c|c|c|}
+\hline
+\textbf{Mechanism} & \textbf{Parameter} & \textbf{MLE} & $\bm{95\%}$ \textbf{Confidence Interval}
+\\
+\hline
+\hline
+
+ Process Noise & $\sigmaProc$ & $\Sexpr{myround(h3_mles %>% filter(prof_var == 'std_W') %>% pull(value), 3)}$
+   &
+  $(\Sexpr{myround(all_ci %>% filter(prof_var == 'std_W') %>% pull(lower), 3)}, \Sexpr{myround(all_ci %>% filter(prof_var == 'std_W') %>% pull(upper), 3)})$ 
+\\
+\hline
+
+ Water Survival (wk) & $\Wremoval^{-1}$ & $\Sexpr{myround(52 / (h3_mles %>% filter(prof_var == 'mu_B') %>% pull(value)), 3)}$
+   &
+  $(\Sexpr{myround(1 / (all_ci %>% filter(prof_var == 'mu_B') %>% pull(upper)), 3)}, \Sexpr{myround(1 / (all_ci %>% filter(prof_var == 'mu_B') %>% pull(lower)), 3)})$ 
+\\
+\hline
+
+ Human to Water Shedding & $\Wshed$ & $\Sexpr{signif(h3_mles %>% filter(prof_var == 'thetaI') %>% pull(value), 3)}$
+   &
+  $(\Sexpr{signif(all_ci %>% filter(prof_var == 'thetaI') %>% pull(lower), 3)}, \Sexpr{signif(all_ci %>% filter(prof_var == 'thetaI') %>% pull(upper), 3)})$ 
+\\
+\hline
+
+ Asymptotic Shedding & $\asymptomRelativeShed$ & $\Sexpr{myround((h3_mles %>% filter(prof_var == 'XthetaA') %>% pull(value))^2, 3)}$
+   &
+  $(\Sexpr{myround((all_ci %>% filter(prof_var == 'XthetaA') %>% pull(lower))^2, 1)}, \Sexpr{myround((all_ci %>% filter(prof_var == 'XthetaA') %>% pull(upper))^2, 3)})$ 
+\\
+\hline
+
+ Seasonality & $\seasAmplitude$ & $\Sexpr{myround(h3_mles %>% filter(prof_var == 'lambdaR') %>% pull(value), 3)}$
+   &
+  $(\Sexpr{myround(all_ci %>% filter(prof_var == 'lambdaR') %>% pull(lower), 3)}, \Sexpr{myround(all_ci %>% filter(prof_var == 'lambdaR') %>% pull(upper), 3)})$ 
+\\
+\hline
+
+ Seasonality & $\rainfallExponent$ & $\Sexpr{myround(h3_mles %>% filter(prof_var == 'r') %>% pull(value), 3)}$
+   &
+  $(\Sexpr{myround(all_ci %>% filter(prof_var == 'r') %>% pull(lower), 3)}, \Sexpr{myround(all_ci %>% filter(prof_var == 'r') %>% pull(upper), 3)})$ 
+\\
+\hline
+
+ Reporting Rate & $\reportRate$ & $\Sexpr{myround(h3_mles %>% filter(prof_var == 'epsilon') %>% pull(value), 3)}$
+   &
+  $(\Sexpr{myround(all_ci %>% filter(prof_var == 'epsilon') %>% pull(lower), 3)}, \Sexpr{myround(all_ci %>% filter(prof_var == 'epsilon') %>% pull(upper), 3)})$ 
+\\
+\hline
+
+ Observation Variance & $\obsOverdispersion$ & $\Sexpr{myround(h3_mles %>% filter(prof_var == 'k') %>% pull(value), 3)}$
+   &
+  $(\Sexpr{myround(all_ci %>% filter(prof_var == 'k') %>% pull(lower), 3)}, \Sexpr{myround(all_ci %>% filter(prof_var == 'k') %>% pull(upper), 3)})$ 
+\\
+\hline
+
+\end{tabular}
+\end{table}
 

### 2024-02-13 si/initialValues.Rnw
Refactored SI + Makefile

\documentclass[11pt]{article}

\input{inputs/header}

\setcounter{tocdepth}{1}

<<packages, include=FALSE>>=
library(tidyverse)
library(pomp)
library(haitipkg)
library(spatPomp)
library(doParallel)
library(doRNG)
library(knitr)

options(
  scipen = 2,
  help_type = "html",
  stringsAsFactors = FALSE,
  continue = "+  ",
  width = 70,
  useFancyQuotes = FALSE,
  reindent.spaces = 2,
  xtable.comment = FALSE
)

opts_chunk$set(
  tidy.opts=list(
    keep.blank.line=FALSE
  ),
  comment="",
  fig.path = 'figure/',
  # warning=FALSE,
  # message=FALSE,
  echo=FALSE,
  fig.align="center",
  dev="pdf",
  dev.args=list(
    bg="transparent",
    pointsize=9
  )
)

cores <-  as.numeric(Sys.getenv('SLURM_NTASKS_PER_NODE', unset = NA))
if(is.na(cores)) cores <- detectCores()
# cores <- 20
registerDoParallel(cores)
theme_set(theme_bw())

if (!dir.exists("../model3/si")) {
  dir.create("../model3/si", recursive = TRUE)
}

myround <- function(x, digits = 1) {
  # taken from the broman package
  if (digits < 1)
    stop("This is intended for the case digits >= 1.")
  if (length(digits) > 1) {
    digits <- digits[1]
    warning("Using only digits[1]")
  }
  tmp <- sprintf(paste("%.", digits, "f", sep = ""), x)
  zero <- paste0("0.", paste(rep("0", digits), collapse = ""))
  tmp[tmp == paste0("-", zero)] <- zero
  tmp
}

theme_set(theme_bw() + theme(plot.title = element_text(hjust = 0.5)))

dep_labeller <- as_labeller(
  c(
    'Artibonite' = 'Artibonite',
    'Sud_Est' = 'Sud-Est',
    'Sud.Est' = 'Sud-Est',
    'Nippes' = 'Nippes',
    'Nord_Est' = 'Nord-Est',
    'Nord.Est' = 'Nord-Est',
    'Ouest' = 'Ouest',
    'Centre' = 'Centre',
    'Nord' = 'Nord',
    'Sud' = 'Sud',
    'Nord_Ouest' = 'Nord-Ouest',
    'Nord.Ouest' = 'Nord-Ouest',
    'Grande_Anse' = 'Grand\'Anse',
    'Grand.Anse' = 'Grand\'Anse'
  )
)
@

\begin{document}

%%%%%%%%%% START

\section{Initial Values}

To perform inference on POMP models, it is necessary to propose an initial probability density for the latent process $f_{X_0}(x_0;\theta)$, including the possibility that the initial values of the latent states are known, or are a non-random function of the unknown parameter vector, $\theta$.
This density is used to obtain initial values of the latent state when fitting and evaluating the model.
For each of the models considered in this analysis, the initial conditions are derived by enforcing the model dynamics on reported cholera cases.
It is also sometimes necessary to fit some initial value parameters in order to help determine initial values for weakly identifiable compartments.
In the following subsections, we mention initial value parameters that were fit for each model.

\subsection{Model~1}

For this model, the number of individuals in the Recovered and Asymptomatic compartments are set to zero, but the initial proportion of Infected and Exposed individuals is estimated as initial value parameters ($\Iinit$ and $\Einit$, respectively) using the IF2 algorithm, implemented as \texttt{mif2} in the \texttt{pomp} package.
Finally, the initial proportion of Susceptible individuals $S_{0, 0}$ is calculated as $S_{0, 0} = 1 - \Iinit - \Einit$.
This model for the initial values of the latent states matches that which was used by \citet{lee20}.

\subsection{Model~2}

Model~2 assumes that the initial values are a deterministic function of the reporting rate and the initial case reports, and so no initial value parameters need to be estimated.
Initial values for latent state compartments are chosen so as to enforce the model dynamics on the observed number of cases.
Specifically, the model sets $I_{u0}(0) = y^*_{1u} / \reportRate$ for each unit $u \in 1:10$, where $I_{u0}(t)$ is the number of infected individuals in unit $u$ at time $t$, vaccination scenario $z = 0$, $y^*_{tu}$ is the reported number of cases, and $\reportRate$ is the reporting rate.
It is futher assumed that there are no individuals in the recovered compartment, as the epidemic has just begun. 
This model for the initial values of the latent states matches that which was used by \citet{lee20}.

The decision to fix initial values so that they satisfy the dynamics of the model has the benefit of reducing the number of estimated parameters and enforcing latent states at time $t_0$ to be consistent with the calibrated model. 
The risk of fixing initial values rather than estimating them is doing so may have substantial effects on the model dynamics, and hence on the consequences of the analysis. 
To consider the impact of the chosen model approach, we consider an alternative initialization model that enables flexible estimation of certain latent states. Specifically, we initialize $I_{u0}(0) = \tilde{I}_{u0}$ for $u \in 1:10$, where $\{\tilde{I}_{u0}\}_{u = 1}^{10}$ is a set of additional model parameters. 
We then fix $S_{u0}(0) = \text{pop}_u - I_{u0}(0)$, and all other starting values are set to zero, as with the fixed value approach.

<<Fit H2 Initialization Params, include=FALSE, message=FALSE, warning=FALSE>>=
h2_main_fit <- readRDS("../model2/model2_fit.rds")

h2Init_fit <- bake(
  file = '../model2/model2_all_fit.rds', {

    # Create model and save initial values for each parameter
    h2 <- haiti2(cutoff = 10000, measure = "log")  # Default is epidemic period
    h2_params <- h2@params
    est_params <- c(
      'Mu', 'Beta', 'BetaW', 'v', 'sigma', 'phase',
      paste0('InitInfected', 1:10)
      )
    n_fit_params <- length(est_params)

    # Get objective function to minimize
    epi_ofun <- pomp::traj_objfun(h2, est = est_params, params = h2_params)

    # Vector of starting values of the parameters that are estimated.
    # phase is not included because it's the only non log-transformed variable
    log_parms <- c(
      'Mu', 'Beta', 'BetaW', 'v', 'sigma', paste0('InitInfected', 1:10)
    )

    log_theta <- h2_params[est_params[est_params %in% log_parms]]
    log_theta[log_theta == 0] <- 1

    # Fit the epi model
    h2_fit_tmp <- subplex::subplex(
      par = c(
        log(log_theta), 
        h2_params[setdiff(est_params, log_parms)]
      )[est_params],
      fn = epi_ofun
    )

    # Save the fitted parameters to the h2_params vector
    h2_params[log_parms] <- exp(h2_fit_tmp$par[log_parms])
    h2_params[setdiff(est_params, log_parms)] <- h2_fit_tmp$par[setdiff(est_params, log_parms)]

    model2_fit_results <- list()
    model2_fit_results[['h2_params']] <- h2_params
    model2_fit_results[["n_fit_params"]] <- n_fit_params
    model2_fit_results[['logLik']] <- -h2_fit_tmp$value

    model2_fit_results
  }, timing = FALSE
)

mod2_AIC <- -2 * h2_main_fit$logLik + 2 * h2_main_fit$n_fit_params
mod2_newInit_AIC <- -2 * h2Init_fit$logLik + 2 * h2Init_fit$n_fit_params
@

The AIC of this alternative approach is $\Sexpr{myround(mod2_newInit_AIC, 1)}$, compared to the fixed approach with an AIC of $\Sexpr{myround(mod2_AIC, 1)}$. 
This alternative initialization approach results in quantitative improvement to the model-fit, but does not result in qualitative differences in the conclusions made using this model. Figure~\ref{fig:h2InitialTraj} displays the trajectory of the model with this alternate initialization model. 
Table~\ref{tab:InitValues} gives the estimated initial values. 
The estimated value of the latent state is similar to the fixed value in Artibonite and Centre, where the largest number of cases are present at the start of the epidemic. 
Because of this, the qualitative dynamics do not differ by much when the parameters are estimated versus held constant, despite the improvement in model fit, measured by AIC.

<<Get H2 trajectories, include=FALSE, echo=FALSE, message=FALSE>>=

h2_Init <- haiti2(region = 'before', cutoff = 10000, measure = 'log')
coef(h2_Init) <- h2Init_fit$h2_params

h2_Init_traj <- trajectory(
  h2_Init,  # Model
  params = h2Init_fit$h2_params,  # estimated parameters
  format = 'data.frame'
)

# Display the output as nationally aggregated data, as was done in Lee et al (2020)
h2_Init_traj$Ctotal <- rowSums(
  h2_Init_traj[, paste0("C", 1:10)]  # Infections in each department
) * h2Init_fit$h2_params['Rho']  # multiply by reporting rate, to compare to reported cases.

h2_traj <- h2_Init_traj %>%
  select(year, Ctotal) %>%
  mutate(
    date = as.Date(lubridate::round_date(lubridate::date_decimal(year), unit = 'day'))
  )

mod2_VaccScenarios <- bake(
  file = paste0("../model2/VaccinationScenarios_InitPars.rds"), {
    haiti2_vaccScenario(
      h2_params = h2Init_fit$h2_params
    )
  },
  timing = FALSE
)

# Create "long" version of the observed cholera cases, for plotting purposes.
dep_plot_df <- haitiCholera %>%
  select(-report) %>%
  pivot_longer(
    data = .,
    cols = -date_saturday,
    values_to = 'cases',
    names_to = 'dep'
  ) %>%
  mutate(
    date = as.Date(date_saturday),
    dep = gsub("\\.", "_", dep)
  ) %>%
  mutate(
    dep = case_when(dep == "Grand_Anse" ~ "Grande_Anse", TRUE ~ dep)
  )

# Calculate total number of nationally aggregated cholera cases, using previous
# data.
true_agg_cases <- dep_plot_df %>%
  tidyr::pivot_wider(
    data = .,
    id_cols = c(date),
    names_from = dep,
    values_from = cases,
    names_prefix = 'cases_'
  ) %>%
  mutate(
    ReportedAll = cases_Artibonite + cases_Centre +
      cases_Grande_Anse + cases_Nippes + cases_Nord +
      cases_Nord_Est + cases_Ouest + cases_Sud +
      cases_Sud_Est + cases_Nord_Ouest
  )
@

\begin{figure}[!ht]
<<h2InitialTraj, echo=FALSE, message=FALSE, fig.height=1.9, fig.width=4.8, fig.align='center'>>=
ggplot() +
  geom_line(data = h2_traj, aes(x = date, y = Ctotal + 1), col = 'blue') +
  geom_line(data = true_agg_cases, aes(x = date, y = ReportedAll + 1)) +
  geom_line(data = filter(mod2_VaccScenarios$all_trajs,
                          year >= max(h2_traj$year)),
            aes(x = as.Date(date), y = ReportedCases + 1, color = scenario)) +
  theme(axis.title.x = element_blank(),
        legend.title = element_blank(),
        axis.title.y = element_text(size = 9),
        axis.text = element_text(size = 8)) +
  ylab("Reported cholera cases") +
  scale_color_manual(values = c("V0" = '#f4a582',
                                "V1" = '#d6604d',
                                "V2" = '#b2182b',
                                "V3" = '#92c5de',
                                "V4" = '#4393c3')) +
  scale_y_log10(labels = scales::trans_format("log10", scales::math_format(10^.x))) +
  scale_x_date(date_labels = "'%y", breaks = seq.Date(from = as.Date("2010-01-01"), as.Date("2030-01-01"), by = '2 years'))
@
\caption{\label{fig:h2InitialTraj}
{\figTitle Simulated trajectory of alternate initialization of Model~2.}
The black line shows the nationally aggregated weekly cholera incidence data.
The blue curve from 2012-2019 is the trajectory of the calibrated version of Model~2.
Compare to Fig.~4 of the main article.}
\end{figure}

\begin{table}[!h]
\centering
\caption{\label{tab:InitValues}Initial values estimated for alternative initialization model for Model~2 compared to the fixed-value initialization model.}
\vspace{2mm}
\begin{tabular}{|c|c|c|c|}
\hline
Latent State & Department & Calibrated Model Value & Fixed Value
\\
\hline
\hline

$I_{1,0}(0)$ & \Sexpr{unlist(dep_labeller(unit_names(h2_Init)[1]))} &
  $\Sexpr{round(h2Init_fit$h2_params["InitInfected1"], 0)}$ &
  $\Sexpr{round(h2_main_fit$h2_params["InitInfected1"], 0)}$ 
\\
\hline

$I_{2,0}(0)$ & \Sexpr{unlist(dep_labeller(unit_names(h2_Init)[2]))} &
  $\Sexpr{round(h2Init_fit$h2_params["InitInfected2"], 0)}$ &
  $\Sexpr{round(h2_main_fit$h2_params["InitInfected2"], 0)}$ 
\\
\hline

$I_{3,0}(0)$ & \Sexpr{unlist(dep_labeller(unit_names(h2_Init)[3]))} &
  $\Sexpr{round(h2Init_fit$h2_params["InitInfected3"], 0)}$ &
  $\Sexpr{round(h2_main_fit$h2_params["InitInfected3"], 0)}$ 
\\
\hline

$I_{4,0}(0)$ & \Sexpr{unlist(dep_labeller(unit_names(h2_Init)[4]))} &
  $\Sexpr{round(h2Init_fit$h2_params["InitInfected4"], 0)}$ &
  $\Sexpr{round(h2_main_fit$h2_params["InitInfected4"], 0)}$ 
\\
\hline

$I_{5,0}(0)$ & \Sexpr{unlist(dep_labeller(unit_names(h2_Init)[5]))} &
  $\Sexpr{round(h2Init_fit$h2_params["InitInfected5"], 0)}$ &
  $\Sexpr{round(h2_main_fit$h2_params["InitInfected5"], 0)}$ 
\\
\hline

$I_{6,0}(0)$ & \Sexpr{unlist(dep_labeller(unit_names(h2_Init)[6]))} &
  $\Sexpr{round(h2Init_fit$h2_params["InitInfected6"], 0)}$ &
  $\Sexpr{round(h2_main_fit$h2_params["InitInfected6"], 0)}$ 
\\
\hline

$I_{7,0}(0)$ & \Sexpr{unlist(dep_labeller(unit_names(h2_Init)[7]))} &
  $\Sexpr{round(h2Init_fit$h2_params["InitInfected7"], 0)}$ &
  $\Sexpr{round(h2_main_fit$h2_params["InitInfected7"], 0)}$ 
\\
\hline

$I_{8,0}(0)$ & \Sexpr{unlist(dep_labeller(unit_names(h2_Init)[8]))} &
  $\Sexpr{round(h2Init_fit$h2_params["InitInfected8"], 0)}$ &
  $\Sexpr{round(h2_main_fit$h2_params["InitInfected8"], 0)}$ 
\\
\hline

$I_{9,0}(0)$ & \Sexpr{unlist(dep_labeller(unit_names(h2_Init)[9]))} &
  $\Sexpr{round(h2Init_fit$h2_params["InitInfected9"], 0)}$ &
  $\Sexpr{round(h2_main_fit$h2_params["InitInfected9"], 0)}$ 
\\
\hline

$I_{10,0}(0)$ & \Sexpr{unlist(dep_labeller(unit_names(h2_Init)[10]))} &
  $\Sexpr{round(h2Init_fit$h2_params["InitInfected10"], 0)}$ &
  $\Sexpr{round(h2_main_fit$h2_params["InitInfected10"], 0)}$ 
\\
\hline

\end{tabular}
\end{table}

\subsection{Model~3}

The latent states of this model are initialized by enforcing the model dynamics on the incidence data from the start of the recorded cases until time $t_0$, requiring that some of the available data be used to determine the initial values of the latent states.
This is the same approach that was taken by \citet{lee20}, who used the value $t_0 = \text{2014-02-22}$; this choice of $t_0$ results in modeling roughly only $60\%$ of the available data, some of which is later discarded for alternative reasons \citep{lee20sup}.

We do not see any immediate reason that this model could not be extended to cover a larger range of the data, and chose the value $t_0 = \text{2010-11-13}$.
This choice of $t_0$ corresponds to using approximately $1\%$ of the available data to determine initial values of the latent states.
In addition to modeling a larger portion of the available data, this choice of $t_0$ corresponds to an important real-world event, as daily reports from each of the departments were not being sent to the health ministry until November 10, 2010 \citep{barzilay13};
this choice of $t_0$ therefore makes $\bm{Y}\big(t_1\big)$ the first week of data once daily reports are being sent to the health ministry.
The few observation times that exist before $t_0$ are used to initialize the model by enforcing model dynamics on these preliminary observations.
For convenience, we denote these observations as $t_{-3}, t_{-2}$ and $t_{-1}$; as before, we let $y_{u,-k} = Y_{u}\big(t_{-k}\big)$ denote the observed case count for unit $u$ at time point $t_{-k}$, where $k \in 1:3$.
Equations for the initial values of non-zero latent states are provided in Eqs.~\myeqref{eq:model3:I0}--\myeqref{eq:model3:B0}; these equations match those that were used by \citet{lee20}, the primary change being a change to the value of $t_0$.

\begin{eqnarray}
\label{eq:model3:I0}
I_{u}\big(t_0\big) &=& \frac{y^*_{u, -1}}{7\reportRate\left(\muIR + \left(\muDeath + \choleraDeath\right)/365\right)},
\\%[-3pt]
\label{eq:model3:A0}
A_{u}\big(t_0\big) &=& \frac{I_{u0}\big(t_0\big)\left(1 - \symptomFrac\right)}{\symptomFrac},
\\
\label{eq:model3:R0}
R_{u01}\big(t_0\big) &=& R_{u02}\big(t_0\big) = R_{u03}\big(t_0\big) = \left(\frac{\sum_{k = -3}^0 y^*_{u, k}}{\reportRate \symptomFrac} - \left(I_{u0}\big(t_0\big) + A_{u0}\big(t_0\big)\right)\right) / 3
\\
\label{eq:model3:S0}
S_{u0}\big(t_0\big) &=& \mathrm{Pop}_u - I_{u}\big(t_0\big) - A_{u}\big(t_0\big) - \sum_{k=1}^3 R_{u0k}\big(t_0\big)
\\
\label{eq:model3:B0}
B_{u}\big(t_0\big) &=& \big[1 + \seasAmplitude \tilde{J}^r \big] \mathrm{Den}_u \, \Wshed \big[ I_{u}(t_0)+ \asymptomRelativeShed A_{u_0}(t) \big]/\Wshed.
\end{eqnarray}

In Eq.~\myeqref{eq:model3:B0}, $\tilde{J} = 0.002376$ is the median adjusted rainfall over the observation period.
One important consideration to make with this parameter initialization model is when $y_{u, -1}^* = 0$, which occurs for units $u \in \{3, 4\}$, which correspond to the Grand'Anse and Nippes departments, respectively.
When this is the case, each of the infectious $I_u(t_0)$, asymptomatic $A_u(t_0)$, and bacterial reservoir $W_u(t_0)$ compartments have a value of zero.
Because Model~3 models cholera transmission primarily by means of the bacterial reservoir, this makes it nearly impossible for an outbreak to occur.
Therefore for units $u \in \{3, 4\}$, we introduce initial value parameters $I_{0,0}^3$ and $I_{0,0}^4$, and calibrate these parameter values using the data.
The resulting parameter estimates are used to obtain the remaining non-zero initial values of the latent states using Eqs.~\myeqref{eq:model3:A0}--\myeqref{eq:model3:B0}.

%%%%%%%%%% END

\bibliography{../bib-haiti}

\end{document}

### 2024-02-13 si/calibrateMod3.Rnw
refactored fitting model 3 without hurricane parameters. INCOMPLETE

@@
 
+<<DefineH3FitFunction>>=
+
+# This function is how the model was fit before doing profiles. The commit id
+# that this function was copied from, for reference, is 49b52e3 of haitipkg. 
+fit_haiti3_si <- function(
+    search1 = NULL,
+    search2 = NULL,
+    nsearches = 1,
+    search_rho = FALSE,
+    search_gamma = FALSE,
+    search_hur = FALSE,
+    search_Iinit = FALSE,
+    start_date = "2010-11-20"
+    ) {
+
+  #
+  ##
+  ### Search1: Global search
+  ##
+  #
+
+  if (is.null(search1$COOLING)) {
+    search1$COOLING <- 0.5
+  }
+
+  if (is.null(search1$KEEP_LIK_MAT)) {
+    search1$KEEP_LIK_MAT <- TRUE
+  }
+
+  if (nsearches >= 2 && is.null(search2$COOLING)) {
+    search2$COOLING <- 0.5
+  }
+
+  if (nsearches >= 2 && is.null(search2$KEEP_LIK_MAT)) {
+    search2$KEEP_LIK_MAT <- TRUE
+  }
+
+  # Create the model that will be fit to cholera incidence data
+  h3_spat <- haitipkg::haiti3_spatPomp(start_date = start_date)
+
+  # Create a list to save all of the results.
+  results <- list()
+
+  # Create vectors for the unit and shared parameters
+  unit_specific_names <- c("betaB", "foi_add", "aHur", "hHur")
+
+  shared_param_names <- c(
+    "mu_B", "XthetaA", "thetaI", "lambdaR", "r", "std_W",
+    "epsilon", "k"
+  )
+
+  if (search_rho) {
+    shared_param_names <- c(shared_param_names, "rho")
+  }
+
+  if (search_gamma) {
+    shared_param_names <- c(shared_param_names, "gamma")
+  }
+
+  est_param_names <- c(
+    unit_specific_names, shared_param_names
+  )
+
+  # Add unit numbers to each parameter
+  est_param_names_expanded <- paste0(rep(est_param_names, each = 10), 1:10)
+
+  if (is.null(search1$RW_SD)) {
+    stop("RW_SD must be supplied for search 1.")
+  } else {
+    chol_rw_1.sd <- search1$RW_SD
+  }
+
+  if (nsearches >= 2 && is.null(search2$RW_SD)) {
+    stop("RW_SD must be supplied for search 2.")
+  } else {
+    chol_rw_2.sd <- search2$RW_SD
+  }
+
+  # Get lower bound for unit parameters (global search)
+  min_val <- 1e-8
+  unit_lb <- rep(c(1, 1e-9, 0, 0), each = 10)
+  names(unit_lb) <- paste0(rep(unit_specific_names[1:4], each = 10), 1:10)
+
+  # Get upper bound for unit parameters (global search)
+  unit_ub <- rep(c(50, 1e-6, 0, 0), each = 10)
+  names(unit_ub) <- paste0(rep(unit_specific_names[1:4], each = 10), 1:10)
+
+  # Set unique upper-bounds for betaB, based on run_level_2 search results.
+  unit_ub['betaB1'] <- 15
+  unit_ub['betaB2'] <- 80
+  unit_ub['betaB3'] <- 75
+  unit_ub['betaB4'] <- 70
+  unit_ub['betaB10'] <- 35
+  unit_ub['betaB5'] <- 15
+  unit_ub['betaB6'] <- 80
+  unit_ub['betaB7'] <- 20
+  unit_ub['betaB8'] <- 5
+  unit_ub['betaB9'] <- 30
+  unit_ub['foi_add10'] <- 2e-7
+  unit_ub['foi_add3'] <- 2e-7
+  unit_ub['foi_add4'] <- 2e-7
+  unit_ub['foi_add6'] <- 5e-7
+  unit_ub['foi_add9'] <- 5e-7
+
+  if (search_hur) {
+    unit_ub['aHur3'] <- 50
+    unit_ub['aHur9'] <- 25
+    unit_ub['hHur3'] <- 120
+    unit_ub['hHur9'] <- 120
+
+    unit_lb['aHur3'] <- 0.01
+    unit_lb['aHur9'] <- 0.01
+    unit_lb['hHur3'] <- 30
+    unit_lb['hHur9'] <- 30
+  }
+
+  if (search_Iinit) {
+    unit_lb['Iinit3'] <- 0.8 / 468301  # H3
+    unit_lb['Iinit4'] <- 0.8 / 342525  # H4
+
+    unit_ub['Iinit3']  <- 40 / 468301
+    unit_ub['Iinit4']  <- 40 / 342525
+  }
+
+  # Get lower bound for shared parameters (global search)
+  shared_lb <- c(
+    25, 0.02, min_val, min_val,
+    0.75, 0.005, 0.5, 1
+  )
+
+  if (search_rho) {
+    shared_lb <- c(shared_lb, 1/8)
+  }
+
+  if (search_gamma) {
+    shared_lb <- c(shared_lb, 365.25 / 5)
+  }
+
+  names(shared_lb) <- shared_param_names
+
+  # Get upper bound for shared parameters (global search)
+  shared_ub <- c(500, 0.8, 0.0001, 5, 1.75, 0.075, 0.99, 200)
+
+  if (search_rho) {
+    shared_ub <- c(shared_ub, 1/(0.2))
+  }
+
+  if (search_gamma) {
+    shared_ub <- c(shared_ub, 365.25 / 2)
+  }
+
+  names(shared_ub) <- shared_param_names
+
+  # Create data.frame with random unit parameters
+  guesses_unit <- pomp::runif_design(
+    lower = unit_lb,
+    upper = unit_ub,
+    nseq = search1$NREPS
+  )
+
+  # Create data.frame with random shared parameters
+  guesses_shared <- pomp::runif_design(
+    lower = shared_lb,
+    upper = shared_ub,
+    nseq = search1$NREPS
+  )
+
+  # Need to duplicate each of the shared parameter columns
+  guesses_shared <- guesses_shared[, rep(1:length(shared_param_names), each = 10)]
+  colnames(guesses_shared) <- paste0(rep(shared_param_names, each = 10), 1:10)
+
+  # Combine the unit and shared parameters
+  guesses <- cbind(guesses_shared, guesses_unit)
+
+  # We need to add fixed parameters
+  all_params <- coef(h3_spat)
+  fixed_params <- all_params[!names(all_params) %in% colnames(guesses)]
+  fixed_mat <- matrix(
+    rep(fixed_params, search1$NREPS),
+    byrow = TRUE, nrow = search1$NREPS
+  )
+  colnames(fixed_mat) <- names(all_params[!names(all_params) %in% colnames(guesses)])
+
+  # Combine estimated and fixed parameters, and reorder based on original order.
+  guesses_all <- cbind(guesses, fixed_mat)[, names(coef(h3_spat))]
+
+  # Memory clean-up
+  rm(guesses_unit, guesses_shared, fixed_mat, min_val, shared_lb,
+     shared_ub, unit_lb, unit_ub,  all_params, fixed_params,
+     est_param_names)
+  gc()
+
+  # It is assumed that a parallel backend has been registered prior to calling
+  # this function.
+  # doParallel::registerDoParallel(ncores)
+  doRNG::registerDoRNG(2198635)
+
+  t_global <- system.time(
+    foreach(
+      i = 1:search1$NREPS,
+      .packages = c("spatPomp"),
+      .combine = c
+    ) %dopar% {
+      r_params <- unlist(guesses_all[i, ])
+      coef(h3_spat) <- r_params
+
+      ibpf_out <- ibpf(
+        h3_spat,
+        Nbpf = search1$NBPF,
+        Np = search1$NP,
+        sharedParNames = shared_param_names,
+        unitParNames = unit_specific_names,
+        spat_regression = search1$SPAT_REGRESSION,
+        rw.sd = chol_rw_1.sd,
+        cooling.fraction.50 = search1$COOLING,
+        block_size = 1
+        # params = r_params  Not yet implemented in spatPomp
+      )
+    } -> Global_ibpf
+  )
+
+  rm(guesses_all)
+
+  pfilterLikes <- data.frame(
+    "ll" = rep(0, search1$NREPS_EVAL*search1$NREPS),
+    "which" = rep(1:search1$NREPS, each = search1$NREPS_EVAL)
+  )
+
+  ibpf_params <- t(coef(Global_ibpf))
+
+  for (i in 1:nrow(ibpf_params)) {
+    for (sp in shared_param_names) {
+      ibpf_params[i, paste0(sp, 1:10)] <- mean(ibpf_params[i, paste0(sp, 1:10)])
+    }
+  }
+
+  t_global_bpf <- system.time(
+    likMat <- foreach(  # Get log-likelihood for each unit and set of parameters, NREPS_EVAL times each
+      i = 1:(search1$NREPS_EVAL*search1$NREPS), .combine = rbind, .packages = 'spatPomp'
+    ) %dopar% {
+      p3 <- ibpf_params[(i-1) %/% search1$NREPS_EVAL + 1, ]
+
+      coef(h3_spat) <- p3
+      apply(bpfilter(
+        h3_spat, Np = search1$NP_EVAL,
+        block_size = 1
+      )@block.cond.loglik, 1, sum)
+    }
+  )
+
+  # Condense unit likelihoods into model likelihood
+  # pfilterLikes$ll <- apply(likMat, 1, sum)
+  pfilterLikes$ll <- rowSums(likMat)  # Faster
+
+  # Group by model parameter set.
+  ibpf_logLik <- pfilterLikes |>
+    dplyr::group_by(which) |>
+    dplyr::summarize(logLik = logmeanexp(ll),
+              se = logmeanexp(ll, se = TRUE)[2])
+
+  search1_results <- list()
+  search1_results$logLiks <- ibpf_logLik
+  search1_results$params <- ibpf_params
+  search1_results$ibpf_time <- t_global
+  search1_results$bpf_time <- t_global_bpf
+
+  if (isTRUE(search1$KEEP_LIK_MAT)) {
+    search1_results$likMat <- likMat
+  }
+
+  if (isTRUE(search1$KEEP_TRACES)) {
+    search1_results$traces <- lapply(Global_ibpf, function(x) x@traces[, c('loglik', names(chol_rw_1.sd@call)[-1])])
+  }
+
+  results$search1 <- search1_results
+
+  if (nsearches == 1) {
+    return(results)
+  }
+
+  rm(ibpf_logLik, guesses, likMat, pfilterLikes,
+     Global_ibpf, chol_rw_1.sd, t_global, t_global_bpf,
+     i, sp, ibpf_params)
+  gc()
+
+  #
+  ##
+  ### Search 2: Local Search
+  ##
+  #
+
+  top_n_global <- order(-search1_results$logLiks$logLik)[1:search2$TOP_N]
+
+  # Contains the best set of parameters, measure for the coupled model
+  params <- search1_results$params[rep(top_n_global, each = search2$NREPS), ]
+
+  all_params <- params
+
+  rm(params)
+
+  doRNG::registerDoRNG(327498615)
+
+  t_ibpf_local <- system.time(
+    foreach(
+      i = 1:(search2$TOP_N * search2$NREPS),
+      .packages = c("spatPomp"),
+      .combine = c
+    ) %dopar% {
+      r_params <- all_params[i, ]
+      coef(h3_spat) <- r_params
+
+      ibpf_out <- ibpf(
+        h3_spat,
+        Nbpf = search2$NBPF,
+        Np = search2$NP,
+        sharedParNames = shared_param_names,
+        unitParNames = unit_specific_names,
+        spat_regression = search2$SPAT_REGRESSION,
+        rw.sd = chol_rw_2.sd,
+        cooling.fraction.50 = search2$COOLING,
+        block_size = 1
+        # params = r_params  Not yet implemented in spatPomp
+      )
+    } -> local_ibpf
+  )
+
+  rm(all_params)
+
+  pfilterLikes <- data.frame(
+    "ll" = NA_real_,
+    "starting_set" = rep(top_n_global, each = search2$NREPS * search2$NREPS_EVAL),
+    "which" = rep(1:(search2$NREPS * search2$TOP_N), each = search2$NREPS_EVAL)
+  )
+
+  ibpf_params <- t(coef(local_ibpf))
+
+  for (i in 1:nrow(ibpf_params)) {
+    for (sp in shared_param_names) {
+      ibpf_params[i, paste0(sp, 1:10)] <- mean(ibpf_params[i, paste0(sp, 1:10)])
+    }
+  }
+
+  t_local_bpf <- system.time(
+    likMat <- foreach(  # Get log-likelihood for each unit and set of parameters, NREPS_EVAL times each
+      i = 1:(search2$NREPS_EVAL*search2$NREPS*search2$TOP_N), .combine = rbind, .packages = 'spatPomp'
+    ) %dopar% {
+      # p3 <- coef(local_ibpf[[(i-1) %/% search2$NREPS_EVAL + 1]])
+      p3 <- ibpf_params[(i-1) %/% search2$NREPS_EVAL + 1, ]
+      coef(h3_spat) <- p3
+      apply(bpfilter(
+        h3_spat, Np = search2$NP_EVAL,
+        block_size = 1
+      )@block.cond.loglik, 1, sum)
+    }
+  )
+
+  # Condense unit likelihoods into model likelihood
+  pfilterLikes$ll <- rowSums(likMat)
+
+  # Group by model parameter set.
+  ibpf_logLik_temp <- pfilterLikes |>
+    dplyr::group_by(which) |>
+    dplyr::summarize(logLik = logmeanexp(ll),
+                     se = logmeanexp(ll, se = TRUE)[2])
+
+  ibpf_logLik <- pfilterLikes |>
+    dplyr::group_by(which) |>
+    dplyr::summarise(starting_set = dplyr::first(starting_set)) |>
+    dplyr::right_join(
+      y = ibpf_logLik_temp,
+      by = "which"
+    ) |>
+    dplyr::select(which, starting_set, logLik, se)
+
+  # ibpf_logLik$init_method <- (ibpf_logLik$which - 1) %/% (search2$TOP_N * search2$NREPS) + 1
+  search2_results <- list()
+  search2_results$logLiks <- ibpf_logLik
+  search2_results$params <- ibpf_params
+  search2_results$ibpf_time <- t_ibpf_local
+  search2_results$bpf_time <- t_local_bpf
+
+  if (isTRUE(search2$KEEP_LIK_MAT)) {
+    search2_results$likMat <- likMat
+  }
+
+  if (isTRUE(search2$KEEP_TRACES)) {
+    search2_results$traces <- lapply(local_ibpf, function(x) x@traces[, c('loglik', names(chol_rw_2.sd@call)[-1])])
+  }
+
+  results$search2 <- search2_results
+
+  if (nsearches == 2) {
+    return(results)
+  }
+
+  rm(
+    search1, search1_results, top_n_global, t_ibpf_local, t_local_bpf,
+    pfilterLikes, ibpf_logLik, chol_rw_2.sd, local_ibpf, likMat, ibpf_logLik_temp
+  )
+
+  gc()
+
+  results
+}
+@
+
 <<CalibrateOrig3>>=
-# Create vectors for the unit and shared parameters
 unit_specific_names <- c("betaB", "foi_add")
@@
 haiti3_sub_fit <- bake(
-  file = "../model3/si/subset_noparams.rds", {
-    fit_haiti3(
+  file = "../model3/si/h3_no_hurricane.rds", {
+    fit_haiti3_si(
       nsearches = 2L,
@@
 )
+@
+
+<<LoadOrig3>>=
 
@@
 bpf_sub3_out <- bake(
-  file = "../model3/si/bpf_sub_cond_LL.rds", {
+  file = "../model3/si/noHur_cond_LL.rds", {
 
@@
 # This result was used in the manuscript, so it doesn't need to be recalculated.
-h3_final_results <- readRDS("../model3/run_level_3/haiti3_fit.rds")
+h3_final_mod_pars <- readRDS("../model3/run_level_3/h3_MLE_eval.rds") %>%
+  slice_max(logLik, n = 1) %>%
+  select(-(job.id:logLik_se)) %>%
+  unlist()
+
+# h3_final_results <- readRDS("../model3/run_level_3/haiti3_fit.rds")
 
@@
 # Then, save the model parameter values and the likelihood evaluations
-best_m <- which.max(h3_final_results$search2$logLiks$logLik)
-best_params <- h3_final_results$search2$params[best_m, ]
-all_liks <- as.data.frame(h3_final_results$search2$likMat)
-all_liks$which <- rep(1:72, each = 18)
-best_mod <- dplyr::filter(all_liks, which == best_m)
+# best_m <- which.max(h3_final_results$search2$logLiks$logLik)
+# best_params <- h3_final_results$search2$params[best_m, ]
+# all_liks <- as.data.frame(h3_final_results$search2$likMat)
+# all_liks$which <- rep(1:72, each = 18)
+# best_mod <- dplyr::filter(all_liks, which == best_m)
+
+# Change model parameters to the best set of parameters from previous
+# search
+coef(h3)[names(h3_final_mod_pars)] <- h3_final_mod_pars
+
+registerDoRNG(731955)  # for modular, reproducible results
+
+# Perform block partical filter with fitted model, to get conditional
+# log-likelihoods.
+bpf_final3_out <- bake(
+  file = "../model3/si/h3_cond_LL.rds", {
+
+    results <- list()
+
+    bpf_out <- h3 %>% bpfilter(
+      Np = 5000,
+      block_size = 1,
+      save_states = TRUE
+    )
+
+    results$block.cond.loglik <- bpf_out@block.cond.loglik
+    results$times <- bpf_out@times  # Saved for convenience
+    results
+  },
+  seed = 432235
+)
 
@@
 
+# # A tibble: 10 × 3
+ #    unit ll_est ll_se
+ #   <dbl>  <dbl> <dbl>
+ # 1     1 -2122. 0.221
+ # 2     2 -2135. 0.419
+ # 3     3 -1529. 0.882
+ # 4     4 -1134. 0.501
+ # 5     5 -1902. 0.213
+ # 6     6 -1423. 0.303
+ # 7     7 -1615. 0.633
+
 # Combine model likelihood table to benchmarks
@@
 ) %>%
-  select(-ll_se) %>%
+  select(-se) %>%
   mutate(unit = as.character(unit)) %>%
@@
 ) %>%
-  rename(ll_mod3 = ll_est, ll_bench = ll) %>%
+  rename(ll_mod3 = est, ll_bench = ll) %>%
   mutate(unit = unlist(myLabeller(unit)),
@@
 
-floor(100 * sum(mod3_cases[, c(bench_diffs$unit[1:4])], na.rm = TRUE) / sum(mod3_cases, na.rm = TRUE))
+# floor(100 * sum(mod3_cases[, c(bench_diffs$unit[1:4])], na.rm = TRUE) / sum(mod3_cases, na.rm = TRUE))
 @

### 2024-02-14 edits.tex
Added paragraph about fitting meta-pop stochastic model

@@
 These results support our argument: forecasts of a dynamic system should start with latent states that are consistent with available data.}
+
+\newcommand\editMetaPop{In our literature review, 17 articles considered dynamic models that incorporate spatial heterogeneity \cite{lee20,tuite11,pasetto18,fitzgibbon20,eisenberg13,rinaldo12,chao11,abrams13,trevisin22,sallah17,collins14,kelly16,azman12,leung22,kuhn14,mari15,gatto12}.
+All but four \cite{lee20,pasetto18,sallah17,azman12} of these studies used deterministic dynamic models, which greatly simplifies the process of calibrating model parameters to incidence data.
+The model in \cite{pasetto18} was fit using an Ensemble Kalman Filter (EnKF) \cite{evensen09}; though EnKF scales favorably with the number of spatial units, it relies on linearization of latent states which can be problematic for highly nonlinear systems \cite{evensen22,ionides21}.
+Alternative approaches used to fit stochastic models included making additional simplifying assumptions to aid in the fitting process \cite{lee20}, and using MCMC algorithms \cite{sallah17,azman12} which require specific structures in the latent dynamics, making these algorithms non plug-and-play.
+In this subsection, we present how the recently developed iterated block particle filter (IBPF) algorithm \cite{ning21ibpf,ionides22} can be used to fit a spatially explicit stochastic dynamic model to incidence data.}

### 2024-02-14 ms.Rnw
Save conditional unit log-likelihoods obtained with the block-particle filter.
Added paragraph about fitting meta-pop stochastic model

@@
 
+    u_cond_ll <- t(sapply(h3_bpf, function(x) rowSums(x@block.cond.loglik)))
+    results$u_cond_ll <- apply(u_cond_ll, 2, logmeanexp, se = TRUE)
+
     ### Get likelihood for subset of the data
@@
 
-\jwc{This doesn't quite feel like the right place for this either, but I can't seem to pin down exactly where that would be yet. Perhaps the discussion?}
 \new{\editMechModels}
@@
 However, a legitimate preference for simplicity can support a decision to consider nationally aggregated models \cite{saltelli20,green15}.
+
+\new{\editMetaPop}
+
 One issue that arises when fitting spatially explicit models is that parameter estimation techniques based on the particle filter become computationally intractable as the number of spatial units increases.
 This is a result of the approximation error of particle filters growing exponentially in the dimension of the model \cite{rebeschini15,park20}.
-
 To avoid the approximation error present in high-dimensional models, Lee et al.~\cite{lee20} simplified the problem of estimating the parameters of Model~3 by creating an approximate version of the model where the units are independent given the observed data.
@@
 At the time Lee et al.~\cite{lee20} conducted their study, there was no known algorithm that could readily be used to maximize the likelihood of an arbitrary meta-population POMP model with coupled spatial dynamics, which justifies the spatial decoupling approximation that was used to calibrate model parameters.
-For our analysis, we calibrate the parameters of the spatially coupled version of Model~3 using the iterated block particle filter (IBPF) algorithm \cite{ionides22}.
+For our analysis, we calibrate the parameters of the spatially coupled version of Model~3 using the \old{iterated block particle filter (IBPF)} \new{IBPF} algorithm \cite{ionides22}.
 This algorithm extends the work of Ning and Ionides \cite{ning21ibpf}, who provided theoretic justification for the version of the algorithm that only estimates unit-specific parameters.
@@
 Simulations from the fitted model are displayed in Fig.~\ref{fig:h3spatsims}.
-\jwc{Should there be more added / said to this paragraph, to demonstrate the novelty of being able to fit a model like Model 3?
-Maybe add a paragraph between the first and second paragraphs of this sub-section.}
 
@@
 
-\jwc{Should we make benchmark modeling a sub-section of model-diagnostics?
-I think this could (A) add some organization to the paper so that readers / reviewers can clearly see how benchmarks help answer the question we posed: what indications should we look for in data to assess trustworthiness of data (B) further emphasize the importance of benchmarks, which is something the reviewers seemed to have found useful.
-}
 One common approach to assess a mechanistic model's goodness-of-fit is to compare simulations from the fitted model to the observed data.

### 2024-02-15 r1/response.tex
Added paragraph about fitting meta-pop stochastic model
Addressed several minor TODO items

@@
 For the revision, we have edited the abstract and introduction to make explicit how the case study fits into the overall goals of the paper.
-\TODO{In the methods and results sections, we have made explicit how the decisions about what to present in the main text and in the supplement relate to these goals.}
 In the discussion, we return from the concrete consideration of the case study, placing it back in the more general setting of inferring complex dynamics from time series data.
@@
 So, although the method is not entirely new, its applicability to this task is novel.
-Additional commentary to this point has been added in the abstract, introduction, and \TODO{the subsection on fitting Model 3}.
+Additional commentary to this point has been added in the abstract, introduction, and the subsection on fitting Model 3.
 
@@
 We futher demonstrate the utility of recent advances in likelihood maximization for high-dimensional nonlinear dynamic models, enabling likelihood-based inference for high-dimensional models for spatiotemporal incidence data.}
-
-% We have clarified this in the abstract and introduction by mentioning that we demonstrate the utility of recent advances in likelihood maximization for high-dimensional nonlinear dynamic models. 
-% \TODO{Furthermore, we have added additional commentary that clearly indicates that although this case study is among the first to use the nov
 
@@
 The literature review was previously mentioned, but citations from the review were not made.
-Furthermore we modified the quoted statement in the author summary to be the following:
+Furthermore, we modified the quoted statement in the author summary to be the following:
 
@@
 \TODO{We have addressed this by adding sign-posts to references and suggestions back to this statement, and other similar statements in the abtract / introduction.}
-We have also added citations to \cite{he10,stocks20} \TODO{and \cite{li23}} for examples demonstrating situations where overdispersed dynamic noise is present, and \cite{king15} for an investigation of the statistical limitations of fitting deterministic models to data.
+Our modified discussion also includes detailed examples of principles that were very useful in our case study but are not universal practice.
 
@@
 From our experience, this type of a result is quite common: for a deterministic model to adequately account for unmodeled features of a dynamic system, parameters that are estimated are often forced to unexpected regions of the parameter space.
-Adding stochasticity to the model often fixes this issue by capturing unmodeled features in the inherent stochasticity of the model rather than in estimated parameters.
+Adding stochasticity to the model helps address this issue by capturing unmodeled features in the inherent stochasticity of the model rather than in estimated parameters.
+Stochasticity in the model alone may not fully solve this problem as evidence suggests that the failure to account for extra-demographic variablility may lead to biases in parameter estimates \cite{he10}.
+Our profile likelihood results for this case study also provide strong evidence for the inclusion of extra-demographic stochasticity in Models~1 and 3 (see supplement material section S8).
 
 The decision on whether to choose Markov chain Monte Carlo (MCMC) methods is not fully aligned with the decision on whether to include stochasticity in the model.
-In our literature review, of the 12 previous Haiti cholera investigations including stochastic dynamic models \cite{kirpich17,lee20,pasett018,mukandavire13,kirpich16,lewnard16,kunkel17,mukandavire15,sallah17,azman12,azman15}, four of them carried out inference via MCMC \cite{pasetto18,lewnard16,sallah17,azman12}.
+In our literature review, of the 12 previous Haiti cholera investigations including stochastic dynamic models \cite{kirpich17,lee20,pasetto18,mukandavire13,kirpich16,lewnard16,kunkel17,mukandavire15,sallah17,azman12,azman15}, four of them carried out inference via MCMC \cite{pasetto18,lewnard16,sallah17,azman12}.
 We agree with the referee that MCMC is more routine for determinstic models than for stochastic models, and advanced MCMC methods such as Hamiltonian Monte Carlo become more readily applicable.
@@
 In this case, identifiability is generally addressed via profile likelihood methods.
-
-\TODO{We have included some additional sentences to address these comments in the discussion section of the revision.}
 
@@
 \article{\editDiscussionFive}
+
+\article{\editDiscussionSix}
 

### 2024-02-15 edits.tex
Addressed several minor TODO items

@@
 More broadly, they have useful roles throughout biology, expecially when combined with statistical methods that properly account for stochasticity and nonlinearity \cite{may04}.
-In some situations, modern machine learning methods, which are associative by nature, can outperform mechanistic models on epidemiological forecasting tasks \cite{lau22,baker18}.
+In some situations, modern machine learning methods can outperform mechanistic models on epidemiological forecasting tasks \cite{lau22,baker18}.
 The predictive skill of non-mechansitic models can reveal limitations in mechanistic models, but cannot readily replace the scientific understanding obtained by describing the biological dynamics of the system in a mathematical model \cite{baker18}.}
@@
 
-\newcommand\editDiscussionFour{The three models in this case study offer a unique combination of decisions of model complexity.
+\newcommand\editDiscussionFour{Unmodeled features of a dynamic system can lead to spurious or misleading parameter estimates if the feature greatly impacts observed data.
+This is because the model must compensate for these unaccounted-for features using its inherent flexibility, particularly in the calibration of parameters.
+In deterministic models, parameter estimation is the sole source of such flexibility.
+When the influence of unmodeled dynamics is significant, it can skew parameter estimates toward scientifically implausible values, a phenomenon observed with Model~2 in our case study.
+Incorporating demographic and environmental stochasticity into models can mitigate the impact of these unmodeled features.
+These stochastic processes are not only arguably present in most dynamic systems, but their inclusion in a model also allows observed data variations to be attributed to inherent randomness rather than to distorted parameter values.
+Furthermore, incorporating extra-demographic stochasticity in a model has also been shown to be beneficial \cite{he10,stocks20,li23}, which aligns with our findings: Models~1 and 3 suggest the presence of extra-demographic stochsticity, as evidenced by the confidence intervals provided in Suplement S8.}
+
+\newcommand\editDiscussionFive{The three models in this case study offer a unique combination of decisions of model complexity.
 Model~3 provides an interesting example in that it is both stochastic and has a meta-population structure, making it challenging to draw likelihood-based inferences.
@@
 
-\newcommand\editDiscussionFive{If forecasts are an important component of a modeling task, the most recent information on the dynamic system should be considered with increased importance.
+\newcommand\editDiscussionSix{If forecasts are an important component of a modeling task, the most recent information on the dynamic system should be considered with increased importance.
 This is particularly true for long time series, as the latent states from simulations from initial conditions may be more likely to diverge from the truth as time increases.

### 2024-02-15 ms.Rnw
Addressed several minor TODO items

@@
 Advances in OCV technology and vaccine availability, however, raised the possibility of planning a national vaccination program \cite{lee20}.
-\new{The possibility of controling the Haiti cholera outbreak via OCV was considered by various research groups.} \TODO{Citations}. 
+\new{The possibility of controlling the Haiti cholera outbreak via OCV was considered by various research groups \cite{ivers17,andrews11,walton11,matias17,chao11,fung13,date11,leung22,azman15}}.
 In the \new{Lee et al.~\cite{lee20}} study, certain data were shared between the groups, including demography and vaccination history; vaccine efficacy was also fixed at a shared value between groups.
@@
 
-# TODO: update this comment: This is used later in figure 2 (TODO)
+# This is used later in figure 2
 beta_t_mcap_results <- pomp::mcap(h1_prof_res$logLik, h1_prof_res$betat, span = 1)
@@
 }
-
-# TODO: Remove
-# haiti3_fit contains ll estimate for many parameter sets, so we want to find
-# the highest value
-# mod3_ll <- h3_prof %>%
-#   slice_max(order_by = logLik, n = 1) %>%
-#   pull(logLik)
 @
@@
 mod1_aic <- 2 * mod1_n_params - 2 * mod1_ll
-
-# model 3 AIC
-# TODO: Remove
-# mod3_aic <- 2 * mod3_n_params - 2 * mod3_ll
 
@@
 \subsection*{Mechanistic models for disease modeling}
+
+\new{\editMechModels}
+
 \new{In this article, we refer to} models that focus on learning relationships between variables in a dataset \new{as} \old{are called} {\it associative}, whereas models that incorporate a known scientific property of the system \old{are called} \new{we call} {\it causal} or {\it mechanistic}.
@@
 In our literature review, there were 14 studies that used associative models to describe various aspects of the cholera epidemic \cite{hulland19,moise20,piarroux11,matias17,eisenberg13,rebaudet19CATI,ivers15,raila17,cuneo17,charles14,richterman19,richterman19inverse,bengtsson15,li16adjusting}.}
-
-\new{\editMechModels}
 
@@
 Many Monte Carlo replications are then required to quantify and further reduce the error.
-The large increase in the log-likelihood for Model~1 (see Table~\ref{tab:likes}) can primarily be attributed to increasing the computational effort in fitting the model; this result highlights the importance of carefully determining the necessary computational effort needed to maximize model likelihoods and acting accordingly.
+The large increase in the log-likelihood for Model~1 (see Table~\ref{tab:likes}) can primarily be attributed to increasing the computational effort in fitting the model.
+This result highlights the importance of carefully determining the necessary computational effort needed to maximize model likelihoods and acting accordingly.
+\TODO{Possibly add more citations to the diagnostics part, maybe li23, laneri10 (Forcing Versus Feedback: Epidemic Malaria and Monsoon Rains in Northwest India)}
+\new{In this case study, this was done by performing standard diagnostics for the IF2 and particle filter algorithms \cite{king16}.
+Given the considerable computational costs of simulation-based algorithms, we find it useful to perform an initial assessment using hyperparameter values that enable relatively quick calculations.
+The insights obtained from this preliminary analysis help in accurately determining the amount of computation that is required to achieve reliable outcomes.}
 Simulations from the initial conditions of our fitted model are plotted against the observed incidence data in Fig.~\ref{fig:mod1fit}.
@@
 These comparisons, which provide insight into how each model performs relative to the others, are quite common \cite{rinaldo12,sallah17}.
-\TODO{Add citations from lit-review.}
 To calibrate relative measures of fit, it is useful to compare against a model that has well-understood statistical ability to fit data, and we call this model a {\it benchmark}.
@@
 
+\new{\editDiscussionSix}
+
 In our case study, we found that additional attention to statistical details could have resulted in \old{improved} \new{an enhanced} statistical fit to the observed incidence data\old{, leading to improvements in the accuracy of the resulting policy guidance.} \new{This would have improved the accuracy of the policy guidance resulting from the study.}

### 2024-02-16 r1/response.tex
Finished addressing reviewer comments

@@
 \usepackage{amsmath,amsthm}
+\usepackage[normalem]{ulem}% to use \sout in feedback commands
 
@@
 We originally tried to make this point by having an entire section dedicated to model diagnostics, though we admit that the connection between the statement in the author summary and a section later in the article was not clear.
-\TODO{We have addressed this by adding sign-posts to references and suggestions back to this statement, and other similar statements in the abtract / introduction.}
+We have addressed this by adding sign-posts to references and suggestions back to this statement, and other similar statements in the abtract / introduction.
 Our modified discussion also includes detailed examples of principles that were very useful in our case study but are not universal practice.
@@
 
-We agree with the referee that diagnostic are an important aspect of careful data analysis.
+We agree with the referee that diagnostics are an important aspect of careful data analysis.
 The most widely used diagnostic tools for particle filter and iterated particle filter methods probably match those provided by default in the \code{pomp} R package \cite{king16} when objects generated by \code{pfilter()} or \code{mif2()} are plotted.
@@
 
-The particle filter provides access to the likelihood of the data, and conditional likelhood of each individual data point.
+The particle filter provides access to the likelihood of the data, and conditional likelihood of each individual data point.
 By contrast, obtaining reliable likelihood values from MCMC is not straightforward.
 Thus, particle filter methods facilitate the use of various likelihood-based methods for diagnostics and model selection.
-We demonstrate this in Figs S4, S5, S6, S7.
+We demonstrate this in Figs S4, S5, S6, S7. \TODO{Check all figure numbers, especially to the supplement.}
 
+As a proper scoring rule \cite{gneiting07}, log-likelihood can be used as an effective way to evaluate filters as well as models. 
 The particle filter is unbiased for the likelihood, and therefore negatively biased for the log-likelihood with bias approaching zero as the number of particles increases and the Monte Carlo variance decreases.
@@
 Thus, log-likelihood is an appropriate way to evaluate filters as well as models.
-If optimization searches from diverse starting pionts provide a consensus on the maximum log-likelihood, and the Monte Carlo variance is small, we can have reasonable confidence in our maximization.
-Monte Carlo adusted profile (MCAP) methods \cite{ionides17,ning21} provide methodology which reduces and evaluates the consequences of this Monte Carlo error.
-MCAP methods have been used in various recent investigations using particle filters \eic{REFS}.
+If optimization searches from diverse starting points provide a consensus on the maximum log-likelihood, and the Monte Carlo variance is small, we can have reasonable confidence in our maximization.
+Monte Carlo adjusted profile (MCAP) methods \cite{ionides17,ning21} provide methodology which reduces and evaluates the consequences of this Monte Carlo error.
+MCAP methods have been used in various recent investigations using particle filters \cite{subramanian21,ali22,pons-salort18}.
 
@@
 A full exposition of this topic is out of the scope of this paper.
+
+\jwc{This text below is something that Ed added.
+I think it's a good idea.
+However, simply because of the overal length of the paper and total number of citations, I think it's probably okay to just leave this out:}
 Howerver, we can alert the reader to the topic, and provide some additional references.
 Not infrequently, diagnostic issues are relegated to supplementary material and so we have indicated to interested readers some papers with extensive supplements demonstrating these diagnostic issues in action.
-\TODO{}
+% \cite{pons-salort18}
 
@@
 We agree with the referee that uncertainty quantification and evaluation of identifiability are fundamental tasks for model-based data analysis.
-
-Wide prediction intervals are not necessarily a sign of a problem.
-For example, nobody knew for sure whether cholera was banished from Haiti after it was declared eliminated in February 2022.
-Cholera reemerged in September 2022, but a reasonable model might be expected to cover the possiblity of this reemergence either happening or not happening, leading to wide prediction intervals.
-
 In the revision, we have included profile confidence intervals on all parameters, except those taking separate values on each unit.
@@
 However, these additional results are worthwhile extension to our analysis and we are grateful for the encouragement to include it.
+
+Wide prediction intervals are not necessarily a sign of a problem.
+For example, nobody knew for sure whether cholera was banished from Haiti after it was declared eliminated in February 2022.
+Cholera reemerged in September 2022, but a reasonable model might be expected to cover the possiblity of this reemergence either happening or not happening, leading to wide prediction intervals.
+It is also critical to recognize that the simulations in Figure~5 are obtained from initial conditions, effectively making simulations at the end of the time series (January 2019) 8 year forecasts from the starting point in November 2010.
+The $95\%$ confidence interval for Ouest in this figure is from $0-1094$ reported cases.
+This range may be considered small given the inherent demographic and environmental stochasticity present in the dynamic system and all of the many possible outcomes of the outbreak from its initial state in 2010.
+% Shorter term forecasts obtained by conditioning on previous data can be obtained by simulating from the filtering distribution.
 
@@
 
-\TODO{Add citations based for this paragraph, given below}
-
-Regulation of biological populations is a fundamental topic in epidemiology, ecology, fisheries and agriculture.
-Population dynamics may be nonlinear and stochastic, with the resulting complexities compounded by incomplete understanding of the underlying biological mechanisms and by partial observability of the system variables.
-Quantitative models for these dynamic systems offer potential for designing effective control measures.
-Developing and testing models for dynamic systems, and assessing their fitness for guiding policy, is a challenging statistical task.
-Questions of interest include: What indications should we look for in the data to assess whether the model-based inferences are trustworthy?
-What diagnostic tests and model variations can and should be considered in the course of the data analysis?
-What are the possible trade-offs of increasing model complexity, such as the inclusion of interactions across spatial units?
+Citations have been added.
 
@@
 In the caption for Table~1, we include a reference to this supplement material.
-\TODO{finish this supplement material}.
 
@@
 Our starting point was the formulation of \cite{lee20}.
-Judging by the referee's comment, they will not be suprised to learn that this structure led to difficulties in some situations.
+% Judging by the referee's comment, they will not be suprised to learn that this structure led to difficulties in some situations.
 We did consider the effect of not calibrating initial values.
 Unsurprisingly, fitting initial values results in a quantitative improvement to model-fit, measured by AIC, but we also found that this had negligable qualitative effects on model-based conclusions.
+On a nationally aggregated scale, the total number of initial infectious individuals in the fixed scenario was 36500, and 33480 when the initial states were estimated.
 We have included additional material in the supplement for Initialization Models that demonstrates this.
@@
 
-\TODO{Add citations to papers that do a good job discussing specific diagnostics (maybe leneary Malaria plos comp bio paper, and li23.
-Then we can also make a small mention of run levels: SMC + IF2 based methods scale linearly, so try first with small number of computations to help roughly evaluate what computational effort is needed.}
+We have added the following text to clarify this point:
+
+\article{The large increase in the log-likelihood for Model~1 (see Table~2) can primarily be attributed to increasing the computational effort in fitting the model.
+This result highlights the importance of carefully determining the necessary computational effort needed to maximize model likelihoods and acting accordingly.
+In this case study, this was done by performing standard diagnostics for the IF2 and particle filter algorithms \cite{king16}.
+Given the considerable computational costs of simulation-based algorithms, we find it useful to perform an initial assessment using hyperparameter values that enable relatively quick calculations.
+The insights obtained from this preliminary analysis help in accurately determining the amount of computation that is required to achieve reliable outcomes.}
 
@@
 
-\TODO{saltelli had a paper about sensitivity analysis, check this and maybe cite / use it.}
+We intentionally do not include predictive intervals in Fig~4. There are several reasons for this, one of which is mentioned in the article:
+
+\article{In this analysis, we do not construct forecasts accounting for parameter uncertainty as our focus is on the estimation and diagnosis of mechanistic models.
+Furthermore, we use the projections from a single point estimate to highlight the deficiency of deterministic models that the only variability in model projections is a result of parameter uncertainty, which can lead to over-confidence in forecasts \cite{king15}.}
+
+\jwc{Below are some thoughts I had. We don't have to include all / any of this, but I thought it was something to think about on the topic of parameter uncertainty}
+
+Parameter uncertainty in model forecasts is an important topic, one that we would have liked to have addressed in our article but felt that we could not since the article is already becoming longer than one would expect in this journal.
+In our literature review of dynamic models for cholera in Haiti, and our experience with literature with dynamic modeling outside of this problem, the most common approaches to accounting for parameter uncertainty are the following:
+
+\begin{itemize}
+  \item Maximum likelihood etimation theory states that under certain conditions, the difference between the MLE and the generating model parameter is asymptoticly normally distributed with mean zero and variance determined by the inverse Fisher information matrix.
+Many numeric optimization techniques often approximate the curvature of the likelihood surface, thereby Fisher information matrix.
+  This can then be used to obtain approximate confidence intervals using the asymptotic distribution of the MLE and the approximation to Fisher's information matrix.
+While mathematically and computationally convinient, the validity of these confidence intervals depends on asymptotics and assumptions that are often not appropriate for non-linear models.
+In fact, there is evidence that these types of confidence intervals are often far too narrow even in the most idealistic cases for dynamic models: linear Gaussian models fit with a large number of observations \cite{wheeler23}.
+  Because of this, confidence regions created in this way likely too narrow, resulting in over-confidence in parameter estimates. 
+  \item Alternatively, it is common to account for parameter uncertainty by performing a sensitivity analysis.
+This is often done when parameters were selected via non-likelihood based approaches, but our review contained instances (e.g.
+\cite{lee20}) where this was done even if parameters were calibrated by maximizing likelihoods.
+  To perform a sensitivity anaylsis, researchers typically pick ranges of parameter estimates, creating a hypercube from which they obtain sample parameter values. 
+  The bounds of this hypercube, however, are often obtained using approximations of Fisher's Information Matrix (the deficiencies of which are mentioned above), or by some non-data driven approach. 
+  For example, bounds are often selected by arbitrarily adding / subtracting a percentage of the estimated parameter (where the bounds can be either larger or smaller than what should be used), or by using parameter estimates from other studies (in which case the parameter estimates may not make sense in the context of the model and data being considered, again resulting in intervals that may be arbitrarily wide or narrow).
+  A persistent theme with all of the above approaches is that they enable researchers to effectively pick the level of uncertainty to present in their reports. 
+  \item Sampling parameters from a Bayesian posterior.
+For many researchers, this may be one of the primary reasons they prefer Baysian methods over maximum likelihood estimates, as these methods do have an advantage in accounting for parameter uncertainty.
+  Bayesian methods to estimate model parameters may not be possible for all scientific models of interest.
+We have been made aware of recent work that demonstrates cases where the authors would like to have a posterior distribution for their model, but the scientific model they propose is unable to be calibrated using state-of-the-art Bayesian approaches; they therefore use IF2 to obtain parameter estimates that result in good behaviour of particle filters, and then arbitrarily select a uniform distribution centered around the MLE as a prior distribution to Bayesian methods (e.g.
+\cite{smith22,rivera22}).
+This approach is related to Empirical Bayes, where the prior distribution is selected using information in the available data.
+We propose something similar in our supplement material, but our proposition does not require additional runs of pMCMC which can be computationally expensive.
+  One drawback of our proposed Empirical Bayes approach is that when the Monte Carlo variance in likelihood evaluation is high, it is possible that the posterior mass function has the majority of its mass assigned to only a few parameter sets.
+\end{itemize}
+
+A full investigation / discussion on parameter uncertainty is outside of the scope of this article.
 
@@
 
+Thank you for this suggestion.
+We have added the following details to help clarify how the benchmarks for the spatially explicit models were created, and how to appropriately compare model likelihoods to these benchmarks:
+
+\article{To obtain a benchmark for models with a meta-population structure, we fit independent auto-regressive negative binomial models to each spatial unit.
+Under the assumption of independence, the log-likelihood of the benchmark on the entire collection of data can be obtained by summing up the log-likelihood for each independent model.
+In general, a spatially explicit model may not have well-defined individual log-likelihoods, and therefore comparisons to benchmarks should be made at the level of the joint model.}
+
 The benchmarks we have looked at operate independently on each unit, and in this case the joint log-likelihood benchmark for units 1 and 2 is the sum of the marginal log-likelihoods.
@@
 
-Since the value of benchmarks is a major point in this paper, we have added these extra details in a supplementary section \TODO{}.
+\jwc{I think it may be easier to just add an additional sentence to the manuscript.}
+Since the value of benchmarks is a major point in this paper, we have added these extra details in a supplementary section.
 
@@
 }
+
+\jwc{I'm not exactly sure how to respond to this point.
+The relevant information about what can be found in S5 text is precisly what is already stated: ``by reconstructing the latent states of Model~3, we notice that the calibrated model favors higher levels of cholera transmission than what was typically observed in the incidence data (S5 Text)".}
+
+\jwc{Maybe we could edit this part to the following:}
+
+\old{In the context of our case study, by reconstructing the latent states of Model~3, we notice that the calibrated model favors higher levels of cholera transmission than what was typically observed in the incidence data (S5 Text).}
+
+In the supplement material (S5 Text), we explore in more detail the process of model fitting and diagnostics.
+Here we demonstrate that although Model~3 does outperform its benchmark model on the aggregate scale, the model performance compared to the benchmark on units with the most cholera cases is poor.
+Furthermore, by comparing simulations from the fitted model to the filtering distribution, we see that the reconstructed latent states of the model favor higher levels of cholera transmission than what is typically observed in the incidence data.
+These result hints at the possibility of model mispecification, and should prompt us to be skeptical of the reliability of conclusions drawn from this model.
 

### 2024-02-16 ms.Rnw
Finished addressing reviewer comments

@@
 Population dynamics may be nonlinear and stochastic, with the resulting complexities compounded by incomplete understanding of the underlying biological mechanisms and by partial observability of the system variables.
-Quantitative models for these dynamic systems offer potential for designing effective control measures.
-Developing and testing models for dynamic systems, and assessing their fitness for guiding policy, is a challenging statistical task.
+Quantitative models for these dynamic systems offer potential for designing effective control measures \new{\cite{vandermeer13,he10}}.
+Developing and testing models for dynamic systems, and assessing their fitness for guiding policy, is a challenging statistical task \new{\cite{king16}}.
 Questions of interest include: What indications should we look for in the data to assess whether the model-based inferences are trustworthy?
@@
 Parameters that were added during our re-analysis and were not considered by Lee et al. are indicated with $^*$.
+\new{Confidence intervals for model parameters are given in Supplement S8.}
 Translations back into the notation of \cite{lee20} are given in \nameref{S1_Table}.
@@
 This result highlights the importance of carefully determining the necessary computational effort needed to maximize model likelihoods and acting accordingly.
-\TODO{Possibly add more citations to the diagnostics part, maybe li23, laneri10 (Forcing Versus Feedback: Epidemic Malaria and Monsoon Rains in Northwest India)}
+% \TODO{Possibly add more citations to the diagnostics part, maybe li23, laneri10 (Forcing Versus Feedback: Epidemic Malaria and Monsoon Rains in Northwest India), pons-salort18}
 \new{In this case study, this was done by performing standard diagnostics for the IF2 and particle filter algorithms \cite{king16}.
@@
 where $\mathrm{E}\big(Y_n|Y_{n - 1}\big) = \NBintercept + \NBar Y_{n-1}$, and $\Var\big(Y_n|Y_{n - 1}\big) = \mathrm{E}\big(Y_n|Y_{n - 1}\big) + \mathrm{E}\big(Y_n|Y_{n - 1}\big)^2 / \NBsize$.
+\new{To obtain a benchmark for models with a meta-population structure, we fit independent auto-regressive negative binomial models to each spatial unit.
+Under the assumption of independence, the log-likelihood of the benchmark on the entire collection of data can be obtained by summing up the log-likelihood for each independent model.
+In general, a spatially explicit model may not have well-defined individual log-likelihoods, and therefore comparisons to benchmarks should be made at the level of the joint model.}
+
 In the case where the case counts are large, an alternative benchmark recommended by He et al.~\cite{he10} is a log-linear Gaussian ARMA model;
@@
 In the context of our case study, by reconstructing the latent states of Model~3, we notice that the calibrated model favors higher levels of cholera transmission than what was typically observed in the incidence data (\nameref{S_mod3cal}).
-This result hints at the possibility of model mispecification, and should prompt us to be sceptical of the reliability of forecasts from this model.
+This result hints at the possibility of model mispecification, and should prompt us to be \old{sceptical} \new{skeptical} of the reliability of forecasts from this model.
 Similarly, parameters that have been calibrated to data should make sense in the context of alternative lines of evidence about the phenomena being modeled, while making allowance for the possibility that the interpretations of parameters may vary when modeling across differing spatial scales.

### 2024-02-16 si/calibrateMod3.Rnw
Finished addressing reviewer comments

@@
 
+unitLikesM3_final <- readRDS("../model3/run_level_3/mod3_eval_res.rds")[['u_cond_ll']] %>%
+  t() %>% 
+  as_tibble() %>% 
+  rename(ll_est = est, ll_se = se) %>% 
+  mutate(unit = 1:10) %>% 
+  select(unit, ll_est, ll_se)
+
 # Get repeated estimates of unit log-likelihoods
-unitLikesM3 <- best_mod %>%
-  pivot_longer(
-    cols = -which,
-    names_to = "unit",
-    names_prefix = "V",
-    values_to = "logLik"
-  ) %>%
-  select(-which) %>%
-  group_by(unit) %>%
-  summarize(
-    ll_est = pomp::logmeanexp(logLik),
-    ll_se = pomp::logmeanexp(logLik, se = TRUE)[2]
-  ) %>%
-  ungroup() %>%
-  mutate(unit = as.numeric(unit)) %>%
-  arrange(unit)
+# unitLikesM3 <- best_mod %>%
+#   pivot_longer(
+#     cols = -which,
+#     names_to = "unit",
+#     names_prefix = "V",
+#     values_to = "logLik"
+#   ) %>%
+#   select(-which) %>%
+#   group_by(unit) %>%
+#   summarize(
+#     ll_est = pomp::logmeanexp(logLik),
+#     ll_se = pomp::logmeanexp(logLik, se = TRUE)[2]
+#   ) %>%
+#   ungroup() %>%
+#   mutate(unit = as.numeric(unit)) %>%
+#   arrange(unit)
 
@@
 # Combine model likelihood table to benchmarks
-bench_M3_joined <- dplyr::left_join(
-  x = unitLikesM3,
+bench_M3_joined_final <- dplyr::left_join(
+  x = unitLikesM3_final,
   y = unitLikes_bench,
@@
 ) %>%
-  select(-se) %>%
+  select(-ll_se) %>%
   mutate(unit = as.character(unit)) %>%
@@
 
-bench_diffs <- dplyr::left_join(
-  x = unitLikesM3,
+bench_diffs_final <- dplyr::left_join(
+  x = unitLikesM3_final,
   y = unitLikes_bench,
@@
 ) %>%
-  rename(ll_mod3 = est, ll_bench = ll) %>%
+  rename(ll_mod3 = ll_est, ll_bench = ll) %>%
   mutate(unit = unlist(myLabeller(unit)),
@@
 We refit Model~3 after introducing these hurricane-related parameters using the same successive search scheme described above.
-The resulting model has a log-likelihood value of $\Sexpr{round(max(h3_final_results$search2$logLiks$logLik), 1)}$.
+The resulting model has a log-likelihood value of $\Sexpr{round(readRDS("../model3/run_level_3/mod3_eval_res.rds")[['ll']]['est'], 1)}$.
 The addition of the Hurricane parameters seems to have resulted in a dramatic increase in conditional likelihoods around and after October 2016, as seen in Fig.~\ref{fig:finalCondLL}.
-The inclusion of these parameters resulted in an overall increase of $\Sexpr{round(max(h3_final_results$search2$logLiks$logLik) - max(haiti3_sub_fit$search2$logLiks$logLik), 1)}$ log-likelihood units.
+The inclusion of these parameters resulted in an overall increase of $\Sexpr{round(readRDS("../model3/run_level_3/mod3_eval_res.rds")[['ll']]['est'] - max(haiti3_sub_fit$search2$logLiks$logLik), 1)}$ log-likelihood units.
 Such a large difference in log-likelihoods is well beyond the threshold of statistical uncertainty determined by Wilks' theorem, suggesting that the data highly favor the inclusion of the additional parameters.
@@
 <<finalModel3BPF, echo=FALSE, message=FALSE, include=FALSE>>=
-
-# Set the model parameters to the best results from previous search
-coef(h3)[names(best_params)] <- best_params
 
@@
   file = "../model3/si/h3_final_bpf_cond_ll.rds", {
+
+    # Set the model parameters to the best results from previous search
+    coef(h3)[names(h3_final_mod_pars)] <- h3_final_mod_pars
 
@@
 <<finalUnitLLFig, fig.height=3, fig.width=5.5, echo=FALSE>>=
-ggplot(bench_M3_joined, aes(x = ll, y = reorder(unit, ll))) +
+ggplot(bench_M3_joined_final, aes(x = ll, y = reorder(unit, ll))) +
   geom_point(aes(col = type)) +
@@
 The difference in log-likelihoods between Model~3 and its benchmark model for each individual units suggests that Model~3 has a relatively poor fit for the four units with the most sustained cholera transmission.
-The simple four parameter benchmark has a higher likelihood than Model~3 for the \Sexpr{bench_diffs$unit[1]} and \Sexpr{bench_diffs$unit[2]} departments, and also has log-likelihoods only a few units smaller than Model~3 for units \Sexpr{bench_diffs$unit[3]} and \Sexpr{bench_diffs$unit[4]}.
+The simple four parameter benchmark has a higher likelihood than Model~3 for the \Sexpr{bench_diffs$unit[1]} and \Sexpr{bench_diffs$unit[2]} departments, and also has log-likelihoods only a few units smaller than Model~3 for the departments \Sexpr{bench_diffs$unit[3]} and \Sexpr{bench_diffs$unit[4]}.
 This is particularly worrisome given that these four departments account for more than $\Sexpr{floor(100 * sum(mod3_cases[, c(bench_diffs$unit[1:4])], na.rm = TRUE) / sum(mod3_cases, na.rm = TRUE))}\%$ of the total number of reported cholera cases.
@@
 This also appears true in later stages of the time period of observed data, where the data suggests that there is a replenishment of the susceptible compartment, but the Model simulations typically retain a smaller proportion of susceptible individuals.
-This result demonstrates that the calibrated model favors a more rapid growth in cholera cases than what is typically seen in the observed data, which explains why the model fails to predict the absence of cases between February, 2019 and October, 2022.
+This result demonstrates that the calibrated model favors a more rapid growth in cholera cases than what is typically seen in the observed data. 
+This, coupled with the fact that the model fails to beat a simple statistical benchmark on data on the four departments that experienced the majority of observed cholera cases, suggests that one should be careful when making inference using this model.
 
@@
   file = "../model3/si/InitSimsQuants.rds", {
+
+    # Set the model parameters to the best results from previous search
+    coef(h3)[names(h3_final_mod_pars)] <- h3_final_mod_pars
 
@@
     h3Spat <- h3
-    p3 <- best_params
+    p3 <- h3_final_mod_pars
+
     coef(h3Spat) <- p3
@@
       h3Spat, end_states = h3_bpf_end_states, covarGen = project_rain,
-      nsims = 1000, seed = 184697
+      nsims = 500, seed = 184697
     )

### 2024-02-16 si/initialValues.Rnw
Finished addressing reviewer comments

@@
 The estimated value of the latent state is similar to the fixed value in Artibonite and Centre, where the largest number of cases are present at the start of the epidemic. 
-Because of this, the qualitative dynamics do not differ by much when the parameters are estimated versus held constant, despite the improvement in model fit, measured by AIC.
+Because of this, the qualitative dynamics do not differ by much when the parameters are estimated versus held constant, despite the improvement in model fit measured by AIC.
 
@@
 The blue curve from 2012-2019 is the trajectory of the calibrated version of Model~2.
-Compare to Fig.~4 of the main article.}
+Compare to Fig.~4 of the article.}
 \end{figure}

### 2024-02-19 si/calibrateMod3.Rnw
Fixed minor Sexpr typo

@@
 Now that the model with additional parameters has been calibrated to the incidence data, we plot the conditional log-likelihood of each department compared to a benchmark model in Fig.~\ref{fig:finalUnitLL}.
-Interestingly, the difference in log-likelihoods between Model~3 and the benchmark model is smallest in the departments \Sexpr{bench_diffs$unit[1]}, \Sexpr{bench_diffs$unit[2]}, \Sexpr{bench_diffs$unit[3]} and \Sexpr{bench_diffs$unit[4]}.
+Interestingly, the difference in log-likelihoods between Model~3 and the benchmark model is smallest in the departments \Sexpr{bench_diffs_final$unit[1]}, \Sexpr{bench_diffs_final$unit[2]}, \Sexpr{bench_diffs_final$unit[3]} and \Sexpr{bench_diffs_final$unit[4]}.
 Each of these departments also exhibited the most sustained cholera transmission, defined by having the fewest number of weeks with no recorded cholera cases.

### 2024-02-19 si/calibrateMod3.Rnw
Fixed minor Sexpr typo, missed some previous typos

@@
 The difference in log-likelihoods between Model~3 and its benchmark model for each individual units suggests that Model~3 has a relatively poor fit for the four units with the most sustained cholera transmission.
-The simple four parameter benchmark has a higher likelihood than Model~3 for the \Sexpr{bench_diffs$unit[1]} and \Sexpr{bench_diffs$unit[2]} departments, and also has log-likelihoods only a few units smaller than Model~3 for the departments \Sexpr{bench_diffs$unit[3]} and \Sexpr{bench_diffs$unit[4]}.
-This is particularly worrisome given that these four departments account for more than $\Sexpr{floor(100 * sum(mod3_cases[, c(bench_diffs$unit[1:4])], na.rm = TRUE) / sum(mod3_cases, na.rm = TRUE))}\%$ of the total number of reported cholera cases.
+The simple four parameter benchmark has a higher likelihood than Model~3 for the \Sexpr{bench_diffs_final$unit[1]} and \Sexpr{bench_diffs_final$unit[2]} departments, and also has log-likelihoods only a few units smaller than Model~3 for the departments \Sexpr{bench_diffs_final$unit[3]} and \Sexpr{bench_diffs_final$unit[4]}.
+This is particularly worrisome given that these four departments account for more than $\Sexpr{floor(100 * sum(mod3_cases[, c(bench_diffs_final$unit[1:4])], na.rm = TRUE) / sum(mod3_cases, na.rm = TRUE))}\%$ of the total number of reported cholera cases.
 

### 2024-02-20 r1/response.tex
Added measurement prediction intervals for Mod2

@@
 
-We intentionally do not include predictive intervals in Fig~4. There are several reasons for this, one of which is mentioned in the article:
+We have added prediction intervals due to the error in the log-normal measurement model.
 
-\article{In this analysis, we do not construct forecasts accounting for parameter uncertainty as our focus is on the estimation and diagnosis of mechanistic models.
-Furthermore, we use the projections from a single point estimate to highlight the deficiency of deterministic models that the only variability in model projections is a result of parameter uncertainty, which can lead to over-confidence in forecasts \cite{king15}.}
-
-\jwc{Below are some thoughts I had. We don't have to include all / any of this, but I thought it was something to think about on the topic of parameter uncertainty}
-
-Parameter uncertainty in model forecasts is an important topic, one that we would have liked to have addressed in our article but felt that we could not since the article is already becoming longer than one would expect in this journal.
-In our literature review of dynamic models for cholera in Haiti, and our experience with literature with dynamic modeling outside of this problem, the most common approaches to accounting for parameter uncertainty are the following:
-
-\begin{itemize}
-  \item Maximum likelihood etimation theory states that under certain conditions, the difference between the MLE and the generating model parameter is asymptoticly normally distributed with mean zero and variance determined by the inverse Fisher information matrix.
+% We intentionally do not include predictive intervals in Fig~4. There are several reasons for this, one of which is mentioned in the article:
+% 
+% \article{In this analysis, we do not construct forecasts accounting for parameter uncertainty as our focus is on the estimation and diagnosis of mechanistic models.
+% Furthermore, we use the projections from a single point estimate to highlight the deficiency of deterministic models that the only variability in model projections is a result of parameter uncertainty, which can lead to over-confidence in forecasts \cite{king15}.}
+% 
+% \jwc{Below are some thoughts I had. We don't have to include all / any of this, but I thought it was something to think about on the topic of parameter uncertainty}
+% 
+% Parameter uncertainty in model forecasts is an important topic, one that we would have liked to have addressed in our article but felt that we could not since the article is already becoming longer than one would expect in this journal.
+% In our literature review of dynamic models for cholera in Haiti, and our experience with literature with dynamic modeling outside of this problem, the most common approaches to accounting for parameter uncertainty are the following:
+% 
+% \begin{itemize}
+%   \item Maximum likelihood etimation theory states that under certain conditions, the difference between the MLE and the generating model parameter is asymptoticly normally distributed with mean zero and variance determined by the inverse Fisher information matrix.
 Many numeric optimization techniques often approximate the curvature of the likelihood surface, thereby Fisher information matrix.
-  This can then be used to obtain approximate confidence intervals using the asymptotic distribution of the MLE and the approximation to Fisher's information matrix.
+%   This can then be used to obtain approximate confidence intervals using the asymptotic distribution of the MLE and the approximation to Fisher's information matrix.
 While mathematically and computationally convinient, the validity of these confidence intervals depends on asymptotics and assumptions that are often not appropriate for non-linear models.
 In fact, there is evidence that these types of confidence intervals are often far too narrow even in the most idealistic cases for dynamic models: linear Gaussian models fit with a large number of observations \cite{wheeler23}.
-  Because of this, confidence regions created in this way likely too narrow, resulting in over-confidence in parameter estimates. 
-  \item Alternatively, it is common to account for parameter uncertainty by performing a sensitivity analysis.
+%   Because of this, confidence regions created in this way likely too narrow, resulting in over-confidence in parameter estimates. 
+%   \item Alternatively, it is common to account for parameter uncertainty by performing a sensitivity analysis.
 This is often done when parameters were selected via non-likelihood based approaches, but our review contained instances (e.g.
 \cite{lee20}) where this was done even if parameters were calibrated by maximizing likelihoods.
-  To perform a sensitivity anaylsis, researchers typically pick ranges of parameter estimates, creating a hypercube from which they obtain sample parameter values. 
-  The bounds of this hypercube, however, are often obtained using approximations of Fisher's Information Matrix (the deficiencies of which are mentioned above), or by some non-data driven approach. 
-  For example, bounds are often selected by arbitrarily adding / subtracting a percentage of the estimated parameter (where the bounds can be either larger or smaller than what should be used), or by using parameter estimates from other studies (in which case the parameter estimates may not make sense in the context of the model and data being considered, again resulting in intervals that may be arbitrarily wide or narrow).
-  A persistent theme with all of the above approaches is that they enable researchers to effectively pick the level of uncertainty to present in their reports. 
-  \item Sampling parameters from a Bayesian posterior.
+%   To perform a sensitivity anaylsis, researchers typically pick ranges of parameter estimates, creating a hypercube from which they obtain sample parameter values. 
+%   The bounds of this hypercube, however, are often obtained using approximations of Fisher's Information Matrix (the deficiencies of which are mentioned above), or by some non-data driven approach. 
+%   For example, bounds are often selected by arbitrarily adding / subtracting a percentage of the estimated parameter (where the bounds can be either larger or smaller than what should be used), or by using parameter estimates from other studies (in which case the parameter estimates may not make sense in the context of the model and data being considered, again resulting in intervals that may be arbitrarily wide or narrow).
+%   A persistent theme with all of the above approaches is that they enable researchers to effectively pick the level of uncertainty to present in their reports. 
+%   \item Sampling parameters from a Bayesian posterior.
 For many researchers, this may be one of the primary reasons they prefer Baysian methods over maximum likelihood estimates, as these methods do have an advantage in accounting for parameter uncertainty.
-  Bayesian methods to estimate model parameters may not be possible for all scientific models of interest.
+%   Bayesian methods to estimate model parameters may not be possible for all scientific models of interest.
 We have been made aware of recent work that demonstrates cases where the authors would like to have a posterior distribution for their model, but the scientific model they propose is unable to be calibrated using state-of-the-art Bayesian approaches; they therefore use IF2 to obtain parameter estimates that result in good behaviour of particle filters, and then arbitrarily select a uniform distribution centered around the MLE as a prior distribution to Bayesian methods (e.g.
@@
 We propose something similar in our supplement material, but our proposition does not require additional runs of pMCMC which can be computationally expensive.
-  One drawback of our proposed Empirical Bayes approach is that when the Monte Carlo variance in likelihood evaluation is high, it is possible that the posterior mass function has the majority of its mass assigned to only a few parameter sets.
-\end{itemize}
-
-A full investigation / discussion on parameter uncertainty is outside of the scope of this article.
+%   One drawback of our proposed Empirical Bayes approach is that when the Monte Carlo variance in likelihood evaluation is high, it is possible that the posterior mass function has the majority of its mass assigned to only a few parameter sets.
+% \end{itemize}
+% 
+% A full investigation / discussion on parameter uncertainty is outside of the scope of this article.
 

### 2024-02-20 ms.Rnw
Added measurement prediction intervals for Mod2

@@
   geom_line(data = h2_traj, aes(x = date, y = Ctotal + 1), col = 'blue') +
+  geom_ribbon(
+    data = h2_traj, 
+    aes(
+      x = date, 
+      ymin = (Ctotal + 1) * exp(-1.96*h2_fit$h2_params['v']),
+      ymax = (Ctotal + 1) * exp(1.96*h2_fit$h2_params['v'])
+    ), alpha = 0.5
+  ) + 
   geom_line(data = true_agg_cases, aes(x = date, y = ReportedAll + 1)) +
-  geom_line(data = filter(mod2_VaccScenarios$all_trajs,
-                          year >= max(h2_traj$year)),
-            aes(x = as.Date(date), y = ReportedCases + 1, color = scenario)) +
+  geom_line(
+    data = filter(mod2_VaccScenarios$all_trajs, year >= max(h2_traj$year)),
+    aes(x = as.Date(date), y = ReportedCases + 1, color = scenario)
+  ) +
+  geom_ribbon(
+    data = filter(mod2_VaccScenarios$all_trajs, year >= max(h2_traj$year)) %>% filter(scenario == 'V0'),
+    aes(
+      x = as.Date(date),
+      ymin = pmax((ReportedCases + 1) * exp(-1.96*h2_fit$h2_params['v']), 1),
+      ymax = pmax((ReportedCases + 1) * exp(1.96*h2_fit$h2_params['v']), 1)
+    ), linetype = 'dashed', alpha = 0.3
+  ) + 
   theme(axis.title.x = element_blank(),
@@
 The blue curve from 2012-2019 is the trajectory of the calibrated version of Model~2.
-Projections under the various vaccination scenarios, which are discussed in detail in the  {\bf Forecasts} subsection are also included.}
+Projections under the various vaccination scenarios, which are discussed in detail in the  {\bf Forecasts} subsection are also included.
+The gray ribbons represent a $95\%$ interval obtained from the log-normal measurement model.
+To avoid over-plotting, measurement variance is only plotted for the V0 vaccination scenario.}
 \end{figure}

### 2024-02-20 si/calibrateMod3.Rnw
Added measurement prediction intervals for Mod2

@@
 Building on their results, we fit the fully coupled version of Model~3 to nearly all available data, reserving only a few observations to use to calibrate the initial conditions of the model (see the previous section for more details).
-In order to properly maximize the model likelihood, we found that it was necessary to use multiple searches for the MLE, periodically pruning away less successful searches.
-The first collection of searches was performed by obtaining initial values for the parameters by uniformly sampling values from a predefined hypercube.
+To maximize model likelihoods, we relied on parameter estimates obtained via the profile-likelihood confidence intervals provided in section S8.
+Prior to obtaining estimates via the profile likelihood search, we found that it was necessary to use multiple searches for the MLE, periodically pruning away less successful searches.
+The first collection of searches is performed by obtaining initial values for the parameters by uniformly sampling values from a predefined hypercube.
 A subsequent refinement search used parameter values corresponding to the largest model likelihoods as starting parameter values.
 The need for multiple searches does not appear to be uncommon, as a similar approach was used by \cite{ionides22}.
+While computationally inefficient, profile likelihoods proved to be an effective alternative to maximizing model likelihoods without the need to apply this multistage heuristic.
 
@@
 
-# h3_final_results <- readRDS("../model3/run_level_3/haiti3_fit.rds")
-
-# Save best_m as index for the model parameters with the best likelihood.
-# Then, save the model parameter values and the likelihood evaluations
-# best_m <- which.max(h3_final_results$search2$logLiks$logLik)
-# best_params <- h3_final_results$search2$params[best_m, ]
-# all_liks <- as.data.frame(h3_final_results$search2$likMat)
-# all_liks$which <- rep(1:72, each = 18)
-# best_mod <- dplyr::filter(all_liks, which == best_m)
-
 # Change model parameters to the best set of parameters from previous
@@
 
-# Get repeated estimates of unit log-likelihoods
-# unitLikesM3 <- best_mod %>%
-#   pivot_longer(
-#     cols = -which,
-#     names_to = "unit",
-#     names_prefix = "V",
-#     values_to = "logLik"
-#   ) %>%
-#   select(-which) %>%
-#   group_by(unit) %>%
-#   summarize(
-#     ll_est = pomp::logmeanexp(logLik),
-#     ll_se = pomp::logmeanexp(logLik, se = TRUE)[2]
-#   ) %>%
-#   ungroup() %>%
-#   mutate(unit = as.numeric(unit)) %>%
-#   arrange(unit)
-
-# # A tibble: 10 × 3
- #    unit ll_est ll_se
- #   <dbl>  <dbl> <dbl>
- # 1     1 -2122. 0.221
- # 2     2 -2135. 0.419
- # 3     3 -1529. 0.882
- # 4     4 -1134. 0.501
- # 5     5 -1902. 0.213
- # 6     6 -1423. 0.303
- # 7     7 -1615. 0.633
-
 # Combine model likelihood table to benchmarks
@@
 
-# floor(100 * sum(mod3_cases[, c(bench_diffs$unit[1:4])], na.rm = TRUE) / sum(mod3_cases, na.rm = TRUE))
 @

### 2024-02-26 r1/response.tex
First round of edits of revision based on Ed's comments

@@
 We evaluate previous methods used for fitting these models, and we demonstrate modified data analysis strategies leading to improved statistical fit.
-Speficially, we present appraoches of diagnosing model misspecification and the consequent development of improved models.
+Speficially, we present appraoches for diagnosing model misspecification and the consequent development of improved models.
 We futher demonstrate the utility of recent advances in likelihood maximization for high-dimensional nonlinear dynamic models, enabling likelihood-based inference for high-dimensional models for spatiotemporal incidence data.}
@@
 
-Among the most common mistakes in mechanistic modeling, in our opinion, is not properly assessing / evaluating a fitted model.
+Among the most common mistakes in mechanistic modeling, in our opinion, is not critically evaluating the statistical fit of the model.
 We originally tried to make this point by having an entire section dedicated to model diagnostics, though we admit that the connection between the statement in the author summary and a section later in the article was not clear.
 We have addressed this by adding sign-posts to references and suggestions back to this statement, and other similar statements in the abtract / introduction.
-Our modified discussion also includes detailed examples of principles that were very useful in our case study but are not universal practice.
+Our modified discussion also includes detailed examples of principles that were very useful in our case study but are not yet universal practice.
 
@@
 
-\article{\editModelingGuidanceFirst}
+\jwc{I combined the two paragraphs into a single paragraph that I think addresses Ed's concern of "brevity and clarity".
+This paragraph is still quite long and could use some editing, so revisit this later.}
 
-\article{\indent\editModelingGuidanceSecond}
+\article{\editModelingGuidance}
+
+% \article{\indent\editModelingGuidanceSecond}
 
@@
 
-It is correct that models fit to the same datasets can be directly compared using the AIC values.
-Models~2 and 3 are not fit to the same datasets, however, due to the way Model~3 is initialized.
-To initialize the latent states of Model~3, Lee et al~\cite{lee20} used approximately $60\%$ of the available data as ``burn-in" observations to initialize latent states.
+It is correct that models fit to the same datasets can be directly compared using the AIC values. Models~2 and 3 are not fit to the same datasets, however, due to the way Model~3 is initialized. 
+Details on how they are initialized are provided in the section with the model descriptions, as well as supplement~S5. 
+% To initialize the latent states of Model~3, Lee et al~\cite{lee20} used approximately $60\%$ of the available data as ``burn-in" observations to initialize latent states.
 According to their supplement, this was done primarily because they believed the dynamics from the start of the outbreak were significantly different enough from the dynamics at the end of the outbreak to suggest that two seperate models would be needed to appropriately describe the data.
-We feel that a properly designed and calibrated model should be able to explain the entire dynamics of the outbreak, including the change from what Lee et al~\cite{lee20} referred to as the ``epidemic" and ``endemic" periods of the outbreak.
+% We feel that a properly designed and calibrated model should be able to explain the entire dynamics of the outbreak, including the change from what Lee et al~\cite{lee20} referred to as the ``epidemic" and ``endemic" periods of the outbreak.
 
-One possible approach to re-analyzing their model would then be to fit all available observations; this would make Models~2 and 3 directly comparable, but would require a significant change to Model~3 itself, as the initialization model used by Lee et al~\cite{lee20} requires a ``burn-in" period.
+% One possible approach to re-analyzing their model would then be to fit all available observations; this would make Models~2 and 3 directly comparable, but would require a significant change to Model~3 itself, as the initialization model used by Lee et al~\cite{lee20} requires a ``burn-in" period.
 One of our goals, however, was to demonstrate that the models provided by Lee et al.~\cite{lee20} were not the issue, and a complete refactorization of the initialization could potentially contradict this objective.
@@
 
-In short, though models that are fit to the same dataset are directly comparable using AIC, this is not the case for Models~2 and 3, as the first few observations in the dataset were not considered likelihood evalution of Model~3.
+% In short, though models that are fit to the same dataset are directly comparable using AIC, this is not the case for Models~2 and 3, as the first few observations in the dataset were not considered likelihood evalution of Model~3.
 This also explains the slight discrepency in AIC for the benchmark models of these two models (Table 2).
-Without going into too much detail in the main article, we added the following sentence to clarify this point:
+We added the following sentence to clarify this point:
 
 \article{Models that are fit to the same datasets can be directly compared using AIC values, making it a useful tool to compare to benchmark models.
-Though Models~2 and 3 are both fit to department level incidence reports, their AIC values are not directly comparable due to the way Model~3 initializes latent states, which is described in detail in Sec.
-S5 of the supplement.}
+Though Models~2 and 3 are both fit to department level incidence reports, their AIC values are not directly comparable due to the way Model~3 initializes latent states, which is described in detail in Sec.~S5 of the supplement.}
 
@@
 This has been fixed in the revision.
-In other instances where the linear trend was previously mentioned, we simply refer to it now as a trend in transmision rate, unless the additional detail of the trend being log-linear is needed for clarity.
+In other instances where the linear trend was previously mentioned, we simply refer to it now as a trend in transmision rate, except where the additional detail of the trend being log-linear is needed for clarity.
 
@@
 
-Our views align with the referee's comments, and we have rewritten the article to make this clearer. The revised paragraph is now:
+Our views align with the referee's comments, and we have rewritten the article to make this clearer. The revised paragraphs with this statement are now:
 
@@
 As for any analysis of observational data, we must be alert to the possibility of confounding.
-For a covariate, this shows up in a similar way to regression analysis: the covariate under investigation could be a proxy for some other unmodeled or unmeasured covariate.
-For a mechanism, the model feature could in principle explain the data by helping to account for some different unmodeled phenomenon.}
+For a covariate, this shows up in a similar way to regression analysis: the covariate under investigation could be a proxy for some other unmodeled \new{phenomenon} or unmeasured covariate.}
 
@@
 \article{Our case study provides an example of this in the parameter estimate for the duration of natural immunity due to cholera infection, $\muRS^{-1}$.
-Under the framework of Model~2, the best estimate for this parameter is $1.4 \times 10^{11} \, \mathrm{yr}$, suggesting that individuals have effectively permanent immunity to cholera once infected.
+Under the framework of Model~2, the best estimate for this parameter is $1.4 \times 10^{11} \, \mathrm{yr}$, suggesting that individuals have permanent immunity to cholera once infected.
 Rather than interpreting this as scientific evidence that individuals have permanent immunity from cholera, this result suggests that Model~2 favors a regime where reinfection events are a negligble part of the dynamics.
@@
 The referee's suggestions have led to the addition of various references in the revision.
-For the particular issue of black box versus white box models, we have included the following paragraph
+For the particular issue of black box versus white box models, we have included the following paragraph:
 
@@
 
-We agree that adding to model complexity has costs, including loss of identifability.
-AIC and other formal statistical methods can assess whether the additional complexity is justified by a sufficiently large improvement of statistical fit.
-Identifiability and model simplicity have practical scientific value which goes beyond their statistical role in reducing generalization error and hence improving out-of-fit predictions.
-This is a variation on the issue of ``statistical significance versus practical significance" familiar from introductory statistics textbooks.
+Moving from an ODE model to a stochastic model may not involve any additional parameters, if only demographic stochasticity is included.
+We agree, however, that adding to model complexity has costs, including loss of identifiability.
+In our revised manuscript, we address identifiability via profile likelihood methods.
+% AIC and other formal statistical methods can assess whether the additional complexity is justified by a sufficiently large improvement of statistical fit.
+% Identifiability and model simplicity have practical scientific value which goes beyond their statistical role in reducing generalization error and hence improving out-of-fit predictions.
+% This is a variation on the issue of ``statistical significance versus practical significance"  often covered in introductory statistics textbooks.
 
-Moving from an ODE model to a stochastic model may not involve any additional parameters, if only demographic stochasticity is included.
+In our study, we find that the deterministic model (model 2) has poor statistical fit.
+A poor fit can lead to a risk of identifying spurious scientific phenomena: additional deterministic complexity added to the model may help to fit the model to unmodeled stochastic phenomena present in the data, and may therefore spuriously be assessed as a highly statistically significant model improvement.
+We see some evidence of this in our analysis.
+Aside from not being stochastic, the mechanistic structure of Model~2 is far more complex than the other two models considered in our study.
+Model~2 also demonstrates the most clear cases of spurious scientific findings, as some estimated parameters favor regions of the parameter space that we would not expect scientifically.
+This may partially explain the choice of Lee et al~\cite{lee20} to consider a scientifically unjustifiable jump in latent state values from their epidemic to endemic periods: without this additional flexibility, Model~2 cannot adequately describe cholera dynamics in Haiti while keeping parameter values in scientifically plausible ranges.
+From our experience, this type of a result is quite common: for a deterministic model to adequately account for unmodeled features of a dynamic system, parameters that are estimated are often forced to unexpected regions of the parameter space.
+
+Adding stochasticity to the model helps address this issue by capturing unmodeled features in the inherent stochasticity of the model rather than in estimated parameters.
+Stochasticity in the model alone may not fully solve this problem as evidence suggests that the failure to account for extra-demographic variablility may lead to biases in parameter estimates \cite{he10}.
+Our revision includes the addition of profile likelihood for model parameters, which also provide strong evidence for the inclusion of extra-demographic stochasticity in Models~1 and 3 (see supplement material section S8).
 Generally, it is advisable to consider the possibility of over-dispersed variation on the transmission rate \cite{breto09,he10} leading to one extra parameter compared to a deterministic model.
@@
 In other contexts, \cite{king15} noted various scientific advantages for including stochasticity in the model; it is undoubtedly part of the biological dynamics, and its omission can lead to over-confident predictions.
+Our revised manuscript includes several references demonstrating the importance of modeling extra-demonstraphic stochasticity, including our revised discussion:
 
-In our study, we find that the deterministic model (model 2) has poor statistical fit.
-This can lead to a risk of identifying spurious scientific phenomena: additional deterministic complexity added to the model may help to describe unmodeled stochastic phenomena present in the data, and may therefore spuriously be assessed as a highly statistically significant model improvement.
-We see some evidence of this in our analysis.
-Aside from not being stochastic, the mechanistic structure of Model~2 is far more complex than the other two models considered in our study; we also see the most clear cases of spurious scientific findings, as some estimated parameters favor regions of the parameter space that we would not expect scientifically (see the added supplement material for confidence intervals of model parameters).
-This may partially explain the choice of Lee et al~\cite{lee20} to consider a scientifically unjustifiable jump in latent state values from their epidemic to endemic periods: without this additional flexibility, Model~2 cannot adequately describe cholera dynamics in Haiti while keeping parameter values in scientifically plausible ranges.
-From our experience, this type of a result is quite common: for a deterministic model to adequately account for unmodeled features of a dynamic system, parameters that are estimated are often forced to unexpected regions of the parameter space.
-Adding stochasticity to the model helps address this issue by capturing unmodeled features in the inherent stochasticity of the model rather than in estimated parameters.
-Stochasticity in the model alone may not fully solve this problem as evidence suggests that the failure to account for extra-demographic variablility may lead to biases in parameter estimates \cite{he10}.
-Our profile likelihood results for this case study also provide strong evidence for the inclusion of extra-demographic stochasticity in Models~1 and 3 (see supplement material section S8).
+\article{\editDiscussionFour}
+
+We acknowledge the existence of useful mathematical tools for studying ODE models, their identifiability and stability and limiting properties.
+Perhaps one example demonstrating the usefulness of ODE models is the Lee et al.~\cite{lee20} study: of the four models considered, the forecasts from Model~2 were most consistent with the actual cholera situation in Haiti.
+Despite their inability to describe incidence data, deterministic models can do a good job at detecting overal trends, like the gradual decline of cholera cases in Haiti.
+This, coupled with the ease in fitting ODEs to data, makes ODEs a useful tool to gain preliminary insight on a dynamic system.
+Furthermore, stochastic dynamic models can be reduced to a deterministic model (a so-called determinsitic skeleton) which can usefully be analyzed to help understand the stochastic dynamics \cite{coulson04}.
+We therefore encourage the use of ODE analysis tools whether or not a stochastic model is adopted.
+Because this approach is tangential to the points we make in our analysis, and it was not done by Lee et al.~\cite{lee20}, we have not followed that approach in the current manuscript.
+A full discussion on deterministic vs simple stochastic models with and without over-dispersion is outside the scope of this article. 
+However, we do make a few comments on the benefits of including overdispersion in a model in several locations, including the description of Model~1 and the conclusion. 
+Because of their usefulness, we do not go as far as to explicitly forbid the use of deterministic models, but instead point out some of their limitations that should be considered. 
+Our revised discussion includes a comment that demonstrates this belief:
+
+\article{Despite their limitations, deterministic models can offer valuable insights into dynamic systems; researchers should carefully weigh these limitations when using these models and presenting their results.}
 
@@
 In practice, these models have been fitted by maximum likelihood, some examples including \cite{lemaitre19,fox22,molodecky23,pons-salort18,subramanian20,stocks20}.
-In this case, identifiability is generally addressed via profile likelihood methods.
-
-We acknowledge the existence of useful mathematical tools for studying ODE models, their identifiability and stability and limiting properties.
-Perhaps one example demonstrating the usefulness of ODE models is the Lee et al.~\cite{lee20} study: of the four models considered, the forecasts from Model~2 were most consistent with the actual cholera situation in Haiti.
-Despite their inability to describe incidence data, deterministic models can do a good job at detecting overal trends, like the gradual decline of cholera cases in Haiti.
-This, coupled with the ease in fitting ODEs to data, makes ODEs a useful tool to gain preliminary insight on a dynamic system.
-Furthermore, stochastic dynamic models can be reduced to a deterministic model (a so-called determinsitic skeleton) which can usefully be analyzed to help understand the stochastic dynamics \cite{coulson04}.
-We therefore encourage the use of ODE analysis tools whether or not a stochastic model is adopted.
-We have not followed that approach in the current manuscript.
 
@@
 I think it's a good idea.
-However, simply because of the overal length of the paper and total number of citations, I think it's probably okay to just leave this out:}
+However, simply because of the overal length of the paper and total number of citations, I think it's probably okay to just leave this out}
+
 Howerver, we can alert the reader to the topic, and provide some additional references.
@@
 We had not done this in the original submission since it was only tangentially relevant to the comparison with \cite{lee20} that lies at the center of our paper.
+In particular, \cite{lee20} did not present confidence intervals.
 However, these additional results are worthwhile extension to our analysis and we are grateful for the encouragement to include it.
@@
 The $95\%$ confidence interval for Ouest in this figure is from $0-1094$ reported cases.
-This range may be considered small given the inherent demographic and environmental stochasticity present in the dynamic system and all of the many possible outcomes of the outbreak from its initial state in 2010.
+This range may not be considered unreasonably given the inherent demographic and environmental stochasticity present in the dynamic system and all of the many possible outcomes of the outbreak from its initial state in 2010.
 % Shorter term forecasts obtained by conditioning on previous data can be obtained by simulating from the filtering distribution.
@@
 Objective measures are used to evaluate model fit; when objective measures of the goodness-of-fit of a model are combined with reproducibility, a framework emerges for continual improvement to model based inference when revisiting the data.
-This framework promotes scientific discovery and bolsters model-based policy recommendations by simplifying the process for researchers to build off of previous results.}
+This framework promotes scientific discovery and supports model-based policy recommendations by simplifying the process for researchers to build off of previous results.}
 
@@
 
-It is difficult to directly add citations as suggested, as the text that was contained in lines 67-78 was original thought or argued in \cite{lucas76}, which was already cited.
-However, we appreciate the point that an improved discussion about the role of mechanistic vs associative modeling would be beneficial, and would be strengthened by other opinions / evidence on the topic.
-
-To that effect, we have modified the original text to be more clear that the definitions of \emph{associative} versus \emph{mechanistic} models are our own, as the precise difference between these may be different in alternative contexts.
-Furthermore, we provide references from our literature review that demonstrate the usefulness of associative models in some cases, whereas we typically focus on the benefits of mechanistic models in the article.
-Finally, we added a paragraph that includes a discussion and references on mechanistic / associative modeling, which also addresses the concerns of Reviewer 1.
-This paragraph contains several citations, and is included below for reference.
+We have added a paragraph that includes a discussion and references on mechanistic vs associative modeling, which also addresses the concerns of Reviewer 1.
 
@@
 Thank you for this suggestion.
-We have added the following details to help clarify how the benchmarks for the spatially explicit models were created, and how to appropriately compare model likelihoods to these benchmarks:
-
-\article{To obtain a benchmark for models with a meta-population structure, we fit independent auto-regressive negative binomial models to each spatial unit.
-Under the assumption of independence, the log-likelihood of the benchmark on the entire collection of data can be obtained by summing up the log-likelihood for each independent model.
-In general, a spatially explicit model may not have well-defined individual log-likelihoods, and therefore comparisons to benchmarks should be made at the level of the joint model.}
-
-The benchmarks we have looked at operate independently on each unit, and in this case the joint log-likelihood benchmark for units 1 and 2 is the sum of the marginal log-likelihoods.
-A simple statistical benchmark does not need to have this structure; for example, one could consider an autoregressive model including spatial dependence.
-
 In general, a joint model for units 1 and 2 may not have well-defined individual log-likelihoods log-lik y1 and log-lik y2 that sum to the full likelihood.
@@
 
-\jwc{I think it may be easier to just add an additional sentence to the manuscript.}
-Since the value of benchmarks is a major point in this paper, we have added these extra details in a supplementary section.
+We have added the following details to help clarify how the benchmarks for the spatially explicit models were created, and how to appropriately compare model likelihoods to these benchmarks:
+
+\article{\editSpatBench}
 
@@
 In their article, Lee et al.~\cite{lee20} only presented forecasts that were based on simulations from initial conditions.
-These forecasts were generally inconsistent with the absence of cholera cases from 2019-2022, and do not account for recent evidence of cholera dynamics.
+These forecasts were generally inconsistent with the absence of cholera cases from 2019-2022, and do not account for the particular relevance of recent data for forecasting cholera dynamics.
 In their supplement material, however, they provide a table (S7) in which they show that if they condition the simulations on most recent data (0 cases from 2019-2020), their forecasts predict cholera elimination.

### 2024-02-26 edits.tex
First round of edits of revision based on Ed's comments

@@
 \definecolor{mygreen}{RGB}{0, 90, 0}
+% \definecolor{purple}{RGB}{128,0,128}
 \newcommand\original[1]{\textcolor{red}{#1}}
@@
 %They recommend five guiding principles: stakeholder engagement, complete model documentation, complete description of data used, communicating uncerainty, and testable model outcomes.
-\newcommand\editModelingGuidanceFirst{There are many guidelines available regarding the use of mechanistic models for policy influence.
+% \newcommand\editModelingGuidanceFirst{There are many guidelines available regarding the use of mechanistic models for policy influence.
 For instance, Behrend et al.~\cite{behrend20} conducted an extensive literature review on modeling principles and standards.
@@
 
-\newcommand\editModelingGuidanceSecond{Based on our assessment and literature review, \cite{lee20} was among the studies that best followed these general recommendations for policy-driven modeling.
+\newcommand\editModelingGuidance{There are many guidelines available regarding the use of mechanistic models for policy influence.
+For instance, Behrend et al.~\cite{behrend20} conducted an extensive literature review on modeling principles and standards.
+Their recommendations generally align with other manifestos on the proper use of models intended to inform policy \cite{saltelli20,donnelly18}, each emphasizing the importance of involving stakeholders in the modeling process, transparency, and communicating uncertainty.
+Other recommendations that focus on issues of model calibration and assessment also exist \cite{dahabreh17,egger17,penaloza15}, demonstrating the considerable amount of published guidance on how to effectively model dynamic systems.
+Based on our assessment and literature review, \cite{lee20} was among the studies that best followed these general recommendations for policy-driven modeling.
 Concretly, \cite{lee20} follows at least four of the five principles outlined in \cite{behrend20}: \emph{complete model documentation}, \emph{complete description of data used}, \emph{communicating uncertainty}, and \emph{testable model outcomes}.
@@
 
-\newcommand\editDiscussionTwo{Likelihood based inference enables researchers to choose between any model that is evaluated on the same dataset using a metric like AIC.
+\newcommand\editDiscussionTwo{Likelihood based inference enables researchers to choose between any model that is evaluated on the same dataset using a metric such as AIC.
 Nested model variations are particularly useful as they enable formal statistical testing of the nested features.
@@
 Mathematical models necessarily simplify complex dynamical systems in order to make inference possible.
-Decision-making about which features of a system need to be incorporated in a model and which should be excluded can significantly influence the model-based conclusions.
+Decisions about which features of a system to incorporate in a model and which should be excluded can significantly influence the model-based conclusions.
 However, these decisions can be tested using a likelihood-based framework.
@@
 \newcommand\editDiscussionFour{Unmodeled features of a dynamic system can lead to spurious or misleading parameter estimates if the feature greatly impacts observed data.
-This is because the model must compensate for these unaccounted-for features using its inherent flexibility, particularly in the calibration of parameters.
+This is because the fitted model must compensate for these unaccounted-for features using its inherent flexibility, particularly in the calibration of parameters.
 In deterministic models, parameter estimation is the sole source of such flexibility.
-When the influence of unmodeled dynamics is significant, it can skew parameter estimates toward scientifically implausible values, a phenomenon observed with Model~2 in our case study.
+When the influence of unmodeled dynamics is substantial, it can skew parameter estimates toward scientifically implausible values, a phenomenon observed with Model~2 in our case study.
 Incorporating demographic and environmental stochasticity into models can mitigate the impact of these unmodeled features.
-These stochastic processes are not only arguably present in most dynamic systems, but their inclusion in a model also allows observed data variations to be attributed to inherent randomness rather than to distorted parameter values.
-Furthermore, incorporating extra-demographic stochasticity in a model has also been shown to be beneficial \cite{he10,stocks20,li23}, which aligns with our findings: Models~1 and 3 suggest the presence of extra-demographic stochsticity, as evidenced by the confidence intervals provided in Suplement S8.}
+These stochastic phenomena are not only arguably present in most biological systems, but their inclusion in a model also allows observed data variations to be attributed to inherent randomness rather than to distorted parameter values.
+Furthermore, incorporating extra-demographic stochasticity in a model has also been shown to be beneficial \cite{he10,stocks20,li23}, which is consistent with our results: Models~1 and 3 suggest the presence of extra-demographic stochsticity, as evidenced by the confidence intervals provided in Suplement S8.}
 
-\newcommand\editDiscussionFive{The three models in this case study offer a unique combination of decisions of model complexity.
-Model~3 provides an interesting example in that it is both stochastic and has a meta-population structure, making it challenging to draw likelihood-based inferences.
+\newcommand\editDiscussionFive{The three models in this case study consider a range of possible model complexity.
+Of these models, Model~3 is distinct in that it is both stochastic and has a meta-population structure, making it challenging to draw likelihood-based inferences.
 In this paper, we demonstrated how this model class could be calibrated to incidence data using the innovative IBPF algorithm.
-One of only a few examples of fitting a model with this level of complexity, this case study exemplifies the algorithm's potential benefits and provides an example for future researchers on a possible approach to fitting a high-dimensional non-linear model.}
+One of only a few examples of fitting a model with this level of complexity via maximum likelihood, this case study exemplifies the algorithm's potential benefits and provides an example for future researchers on a possible approach to fitting a high-dimensional non-linear model.}
 
@@
 Deterministic simulations may also result in over-confidence in model forecasts, as they can only account for uncertainty due to the parameter estimation proceedure \cite{king15}.
-Despite these limitations, forecasts from deterministic models can provide valuable insights on potential future trends, though researchers should be transparent about these limitations when presenting forecasts from deterministic models.}
+Despite their limitations, deterministic models can offer valuable insights into dynamic systems; researchers should carefully weigh these limitations when using these models and presenting their results.}
 
@@
 
-\newcommand\editForecastOne{Forecasts of a dynamic system should ideally be based on the most recent information available as it is likely to be more relevant than older data.
-While this assertion may seem self-evident, it is not the case for deterministic models, for which the initial conditions together with the parameters are sufficient for forecasting, and so recent data do not have special importance.
+\newcommand\editForecastOne{Forecasts of a dynamic system should should be consistent with the available data.
+It is particularly important that forecasts are consistent with the most recent information available, as recent data is likely to be more relevant than older data.
+While this assertion may seem self-evident, it is not the case for deterministic models, for which the initial conditions together with the parameters are sufficient for forecasting, and so recent data may not be consistent with model trajectories.
 Epidemiological forecasts based on deterministic models are not uncommon in practice, despite their limitations \cite{king15}.
-Lee et al.~\cite{lee20} chose to obtain forecasts from all of their models by simulating forward from initial conditions.
-This decision is possibly as a result of using a deterministic model: forecasts from different models may only be considered comparable only if they are obtained in the same way, which must be by simulating from initial conditions because Model~2 is deterministic.}
+Lee et al.~\cite{lee20} chose to obtain forecasts from all of their models by simulating forward from initial conditions, rather than conditioning forecasts based on the available data.
+This decision is possibly as a result of using a deterministic model, as forecasts from different models may only be considered comparable if they are obtained in the same way, which must be by simulating from initial conditions because Model~2 is deterministic.}
 
@@
 Table~S7 in their supplement material, which contains results that were not discussed in their main article, shows that the subset of their simulations which were consistent with observing zero cases from 2019-2020 were also more consistent with the disappearance of cholera from Haiti from 2019-2022.
-These results support our argument: forecasts of a dynamic system should start with latent states that are consistent with available data.}
+These results support our argument that forecasts should be made by ensuring the starting point for the forecast is consistent with available data.}
 
@@
 In this subsection, we present how the recently developed iterated block particle filter (IBPF) algorithm \cite{ning21ibpf,ionides22} can be used to fit a spatially explicit stochastic dynamic model to incidence data.}
+
+\newcommand\editSpatBench{To obtain a benchmark for models with a meta-population structure, we fit independent auto-regressive negative binomial models to each spatial unit.
+Under the assumption of independence, the log-likelihood of the benchmark on the entire collection of data can be obtained by summing up the log-likelihood for each independent model.
+In general, a spatially explicit model may not have well-defined individual log-likelihoods, and, in this case, comparisons to benchmarks must be made at the level of the joint model.}

### 2024-02-26 ms.Rnw
First round of edits of revision based on Ed's comments

@@
 We evaluate previous methods used for fitting these models, and we demonstrate modified data analysis strategies leading to improved statistical fit.}
-Specifically, we present approaches of diagnosing model misspecification \old{, development of alternative models, and computational improvements in optimization, in the context of likelihood-based inference on nonlinear dynamic systems.} \new{and the consequent development of improved models.
+Specifically, we present approaches for diagnosing model misspecification \old{, development of alternative models, and computational improvements in optimization, in the context of likelihood-based inference on nonlinear dynamic systems.} \new{and the consequent development of improved models.
 We further demonstrate the utility of recent advances in likelihood maximization for high-dimensional nonlinear dynamic models, enabling likelihood-based inference for high-dimensional models for spatiotemporal incidence data.}
@@
 Objective measures are used to \old{benchmark} \new{evaluate} model fit; when \old{these}\new{objective measures of the goodness-of-fit of a model} are combined with reproducibility, a framework emerges for continual improvement \new{to model based inference} when revisiting the data\old{ and models}.
-\new{This framework promotes scientific discovery and bolsters model-based policy recommendations by simplifying the process for researchers to build off of previous results.}
+\new{This framework promotes scientific discovery and supports model-based policy recommendations by simplifying the process for researchers to build off of previous results.}
 Our data analysis workflow is supported by recent advances in algorithms, software and hardware, which facilitate statistical fitting of nonlinear stochastic dynamic models to observed incidence data.
@@
 
-\new{\editModelingGuidanceFirst}
-
-\new{\editModelingGuidanceSecond}
+\new{\editModelingGuidance}
+
+% \new{\editModelingGuidanceSecond}
 
@@
 As for any analysis of observational data, we must be alert to the possibility of confounding.
-For a covariate, this shows up in a similar way to regression analysis: the covariate under investigation could be a proxy for some other unmodeled or unmeasured covariate.
-For a mechanism, the model feature could in principle explain the data by helping to account for some different unmodeled phenomenon.
+For a covariate, this shows up in a similar way to regression analysis: the covariate under investigation could be a proxy for some other unmodeled \new{phenomenon} or unmeasured covariate.
+\old{For a mechanism, the model feature could in principle explain the data by helping to account for some different unmodeled phenomenon.}
 
@@
 where $\mathrm{E}\big(Y_n|Y_{n - 1}\big) = \NBintercept + \NBar Y_{n-1}$, and $\Var\big(Y_n|Y_{n - 1}\big) = \mathrm{E}\big(Y_n|Y_{n - 1}\big) + \mathrm{E}\big(Y_n|Y_{n - 1}\big)^2 / \NBsize$.
-\new{To obtain a benchmark for models with a meta-population structure, we fit independent auto-regressive negative binomial models to each spatial unit.
-Under the assumption of independence, the log-likelihood of the benchmark on the entire collection of data can be obtained by summing up the log-likelihood for each independent model.
-In general, a spatially explicit model may not have well-defined individual log-likelihoods, and therefore comparisons to benchmarks should be made at the level of the joint model.}
+\new{\editSpatBench}
 
@@
 Our case study provides an example of this in the parameter estimate for the duration of natural immunity due to cholera infection, $\muRS^{-1}$.
-Under the framework of Model~2, the best estimate for this parameter is $\Sexpr{signif(1/h2_fit$h2_params["sigma"],3)} \, \mathrm{yr}$, suggesting that individuals have effectively permanent immunity to cholera once infected.
+Under the framework of Model~2, the best estimate for this parameter is $\Sexpr{signif(1/h2_fit$h2_params["sigma"],3)} \, \mathrm{yr}$, suggesting that individuals have \old{effectively} permanent immunity to cholera once infected.
 \new{Rather than interpreting this as scientific evidence that individuals have permanent immunity from cholera, this result suggests that Model~2 favors a regime where reinfection events are a negligble part of the dynamics.}

### 2024-02-26 si/ReplicateLee20.Rnw
First round of edits of revision based on Ed's comments

@@
 
-<<plotMod1Sims, echo=FALSE, fig.cap=paste0("Simulations using parameter sets that were generated by running source code provided by Lee et al (2020).
+<<plotMod1Sims, echo=FALSE, fig.cap=paste0("Simulations from Model~1 using parameter sets that were generated by running source code provided by Lee et al (2020).
 Compare to Figure S7 in the supplement of Lee et al.

### 2024-02-27 edits.tex
some edits to page 16

@@
 
-\newcommand\editModelingGuidance{There are many guidelines available regarding the use of mechanistic models for policy influence.
-For instance, Behrend et al.~\cite{behrend20} conducted an extensive literature review on modeling principles and standards.
-Their recommendations generally align with other manifestos on the proper use of models intended to inform policy \cite{saltelli20,donnelly18}, each emphasizing the importance of involving stakeholders in the modeling process, transparency, and communicating uncertainty.
-Other recommendations that focus on issues of model calibration and assessment also exist \cite{dahabreh17,egger17,penaloza15}, demonstrating the considerable amount of published guidance on how to effectively model dynamic systems.
-Based on our assessment and literature review, \cite{lee20} was among the studies that best followed these general recommendations for policy-driven modeling.
-Concretly, \cite{lee20} follows at least four of the five principles outlined in \cite{behrend20}: \emph{complete model documentation}, \emph{complete description of data used}, \emph{communicating uncertainty}, and \emph{testable model outcomes}.
+\newcommand\editModelingGuidance{Many guidelines have been proposed regarding the use of mechanistic models to inform policy, reviewed by Behrend et al.~\cite{behrend20}.
+The recommendations of \cite{behrend20} generally align with other manifestos on the proper use of models for informing policy \cite{saltelli20,donnelly18}, each emphasizing the importance of involving stakeholders in the modeling process, transparency, and communication of uncertainty.
+Other recommendations focus on issues of model calibration and assessment \cite{dahabreh17,egger17,penaloza15}.
+Based on our literature review, \cite{lee20} was among the studies that best followed these general recommendations for policy-driven modeling.
+Concretly, \cite{lee20} follows four of the five principles outlined in \cite{behrend20}: \emph{complete model documentation}, \emph{complete description of data used}, \emph{communicating uncertainty}, and \emph{testable model outcomes}.
 Determining the level of adherence to the first principle, \emph{stakeholder engagement}, is difficult based solely on the article.
@@
 Our general recommendations therefore complement and extend existing guidelines by focusing on the methodological tasks of calibrating and evaluating dynamic models in a rigorous statistical framework.
-We specifically emphasize principles that proved essential in our case study; complementary methodological suggestions arising from a spatio-temporal analysis of COVID-19 are given in \cite{li23}.}
+We specifically emphasize principles that proved relevant to our case study; complementary methodological suggestions arising from a spatio-temporal analysis of COVID-19 are given in \cite{li23}.}
 
@@
 \newcommand\editMetaPop{In our literature review, 17 articles considered dynamic models that incorporate spatial heterogeneity \cite{lee20,tuite11,pasetto18,fitzgibbon20,eisenberg13,rinaldo12,chao11,abrams13,trevisin22,sallah17,collins14,kelly16,azman12,leung22,kuhn14,mari15,gatto12}.
-All but four \cite{lee20,pasetto18,sallah17,azman12} of these studies used deterministic dynamic models, which greatly simplifies the process of calibrating model parameters to incidence data.
+All but four \cite{lee20,pasetto18,sallah17,azman12} of these studies used deterministic dynamic models: this greatly simplifies the process of calibrating model parameters to incidence data, but unfortunately that is not enough to guarantee that the calibrated model provides an adequate statistical explanation of the data [I'VE EDITED THIS...
+IT MAY NEED MORE WORK, BUT WE MAY WANT TO REMIND THE READER THAT EASIER FITTING IS NOT SUCH A GOOD THING IF THE FIT IS JUNK].
 The model in \cite{pasetto18} was fit using an Ensemble Kalman Filter (EnKF) \cite{evensen09}; though EnKF scales favorably with the number of spatial units, it relies on linearization of latent states which can be problematic for highly nonlinear systems \cite{evensen22,ionides21}.

### 2024-02-27 ms.Rnw
some edits to page 16

@@
 Specifically, we present approaches for diagnosing model misspecification \old{, development of alternative models, and computational improvements in optimization, in the context of likelihood-based inference on nonlinear dynamic systems.} \new{and the consequent development of improved models.
-We further demonstrate the utility of recent advances in likelihood maximization for high-dimensional nonlinear dynamic models, enabling likelihood-based inference for high-dimensional models for spatiotemporal incidence data.}
+Additionally, we demonstrate the utility of recent advances in likelihood maximization for high-dimensional nonlinear dynamic models, enabling likelihood-based inference for high-dimensional models for spatiotemporal incidence data.}
 Our workflow is reproducible and extendable, facilitating future investigations of this disease system.
@@
 \old{In this article, we discuss common modeling decisions that may not provide an adequate statistical explanation of the data, and we show that attention to statistical fit can lead to more accurate policy evaluations}.
-\new{In this article, we use an in-depth case study to motivate general procedures for fitting mathematical models to time series data, while explaining how these procedures help to address limitations of previous approaches.}
-We do this by conducting a re-analysis of the 2010 cholera outbreak in Haiti.
-Our analysis presents methodology for diagnosing how well a model describes observed data\new{.}\old{, and provides careful justification of valid conclusions from the fitted model.}
+In this article, we conduct a re-analysis of the 2010 cholera outbreak in Haiti.
+\new{We use this case study to investigate current procedures for fitting mathematical models to time series data, while identifying limitations of these methodologies and proposing remedies.}
+We present methodology for diagnosing how well a model describes observed data\new{.}\old{, and provides careful justification of valid conclusions from the fitted model.}
 Objective measures are used to \old{benchmark} \new{evaluate} model fit; when \old{these}\new{objective measures of the goodness-of-fit of a model} are combined with reproducibility, a framework emerges for continual improvement \new{to model based inference} when revisiting the data\old{ and models}.
@@
 \old{The data that were used for this study, and that is the subject of our reanalysis, is displayed in Fig.~[1].}
-\new{Various dynamic models were developed to study this outbreak: searching PubMed with keywords ``Haiti, cholera, model'' we obtained 22 studies that utilized deterministic models \cite{lee20,tuite11,andrews11,botelho21,fitzgibbon20,eisenberg13,rinaldo12,chao11,date11,lin19,abrams13,collins21,akman15,trevisin22,mavian20,collins14,kelly16,capone15,leung22,mari15,gatto12,kuhn14}
+\new{Various dynamic models were developed to study this outbreak: searching PubMed with keywords ``Haiti, cholera, model'' we obtained 22 studies that utilized deterministic mechanistic dynamic models \cite{lee20,tuite11,andrews11,botelho21,fitzgibbon20,eisenberg13,rinaldo12,chao11,date11,lin19,abrams13,collins21,akman15,trevisin22,mavian20,collins14,kelly16,capone15,leung22,mari15,gatto12,kuhn14}
 and 11 studies that used stochastic models \cite{kirpich16,lee20,pasetto18,mukandavire13,kirpich17,lewnard16,kunkel17,mukandavire15,sallah17,azman12,azman15}.
@@
 Parameters that were added during our re-analysis and were not considered by Lee et al. are indicated with $^*$.
-\new{Confidence intervals for model parameters are given in Supplement S8.}
+\new{Confidence intervals for model parameters are given in the supplement (Sec.~S8).}
 Translations back into the notation of \cite{lee20} are given in \nameref{S1_Table}.
@@
 The observable random variables $\bm{Y}_{1:N}$ are assumed to be conditionally independent given $\bm{X}_{0:N}$.
-Together, with the density for the initial value of the latent state $\bm{X}_{0} = \bm{X}(t_0)$, each model assumes a joint density $f_{\bm{X}_{0:N}, \bm{Y}_{1:N}}\big(\bm{x}_{0:N}, \bm{y}_{1:N};\paramVec\big)$, where $\paramVec$ is a parameter vector that indexes the model.
+Together, with the density for the initial value of the latent state $\bm{X}_{0} = \bm{X}(t_0)$, each model defines a joint density $f_{\bm{X}_{0:N}, \bm{Y}_{1:N}}\big(\bm{x}_{0:N}, \bm{y}_{1:N};\paramVec\big)$, where $\paramVec$ is a parameter vector that indexes the model.
 The observed data $\bm{y}_{1:N}^*$, along with the unobserved true value of the latent state, are modeled as a realization of this joint distribution.
@@
 These results are suggestive that the inclusion of a trend in the transmission rate improves the quantitative ability of Model~1 to describe the observed data.
-\new{The maximum likelihood estimate for $\transmissionTrend$ corresponds with a \Sexpr{paste0(myround(100 - 100*exp(beta_t_mcap_results$mle * 430/215), 1), '\\%')} reduction to the transmission rate over the course of the outbreak, with a $95\%$ confidence interval of (\Sexpr{paste0(rev(myround(100 - 100*exp(beta_t_mcap_results$ci * 430/215), 1)), collapse = ', ', sep = '\\%')}) for the overall reduction in transmission.}
+\new{The maximum likelihood estimate for $\transmissionTrend$ corresponds to a \Sexpr{paste0(myround(100 - 100*exp(beta_t_mcap_results$mle * 430/215), 1), '\\%')} reduction to the transmission rate over the course of the outbreak, with a $95\%$ confidence interval of (\Sexpr{paste0(rev(myround(100 - 100*exp(beta_t_mcap_results$ci * 430/215), 1)), collapse = ', ', sep = '\\%')}) for the overall reduction in transmission.}
 The reported results for Model~1 in the remainder of this article were obtained with the inclusion of the parameter $\transmissionTrend$.

### 2024-02-27 r1/response.tex
Second round of edits based on Ed's comments

@@
 
-\article{Despite their limitations, deterministic models can offer valuable insights into dynamic systems; researchers should carefully weigh these limitations when using these models and presenting their results.}
+\article{Despite their limitations, deterministic models can offer valuable insights into dynamic systems.
+As with all models, researchers should consider and acknoweledge these limitations when applying deterministic models and interpreting their outcomes.}
 
@@
 Aside from this statement, they provided limited details on how they came up with this function.
-Because of this, we unfortunately can't provide much more additional information other than the statement that was already included in our article:
+Our revised manuscript clarifies this point
 
-\article{$\upsilon^\star(t)$ is efficacy at time $t$ since vaccination for adults, taken from \cite{lee20}, Table S4, ...}
+\article{$\upsilon^\star(t)$ is efficacy at time $t$ since vaccination for adults, a step-function represented in Table~S4 of \cite{lee20}, ...}
 

### 2024-02-27 edits.tex
Second round of edits based on Ed's comments

--- 2024-02-26 edits.tex
@@
 
-\newcommand\editModelingGuidance{There are many guidelines available regarding the use of mechanistic models for policy influence.
+% \newcommand\editModelingGuidance{There are many guidelines available regarding the use of mechanistic models for policy influence.
 For instance, Behrend et al.~\cite{behrend20} conducted an extensive literature review on modeling principles and standards.
@@
 
+% \newcommand\editModelingGuidance{Numerous guidelines for using mechanistic models to inform policy exist, yet they often lack implementation specifics.
+Behrend et al.~\cite{behrend20}, for instance, stress the importance stakeholder engagement, transparency, and uncertainty communication.
+Other guidelines for mechanistic modeling emphasize similar principles \cite{saltelli20,donnelly18}, demonstrating that the importance of these ideas is well known.
+Additional literature discusses model calibration and evaluation techniques \cite{dahabreh17,egger17,penaloza15}, but do not make specific recommendations on how to effectively implement these techniques in practice.
+Our literature review reveals gaps in applying principles found in \cite{behrend20,saltelli20,donnelly18,dahabreh17,egger17,penaloza15} effectively.
+For example, despite ostensibly adhering to four out of five key recommendations from \cite{behrend20}---determining the level of adherence to the remaining principle, \emph{stakeholder engagement}, is difficult based solely on the article---the projections of \cite{lee20} show a clear misalignment with actual cholera incidence data from 2019 to 2022.
+Such findings underscore the limitations of current standards.
+Saltelli et al.
+\cite{saltelli19} suggests that principles and structres based on statistical principles can improve model based outcomes.
+Hence, our recommendations advocate for rigorous statistical calibration and evaluation of dynamic models while complimenting existing recommendations such as \cite{behrend20}.
+We specifically emphasize principles that proved essential in our case study; complementary methodological suggestions arising from a spatio-temporal analysis of COVID-19 are detailed in \cite{li23}.}
+
+\newcommand\editModelingGuidance{Numerous guidelines for using mechanistic models to inform policy exist, yet they often lack implementation specifics.
+Behrend et al.~\cite{behrend20}, for instance, stress the importance stakeholder engagement, transparency, and uncertainty communication.
+These and other related principles are echoed by other well-cited studies \cite{saltelli20,donnelly18}, suggesting that the importance of these ideas is well known.
+Additional literature discusses model calibration and evaluation techniques \cite{dahabreh17,egger17,penaloza15}, but do not make specific recommendations on how to effectively implement these techniques in practice.
+Our literature review reveals gaps in applying principles found in \cite{behrend20,saltelli20,donnelly18,dahabreh17,egger17,penaloza15} effectively.
+For example, \cite{lee20} largely adhere to the principles of \cite{behrend20}---though assessing the extent of stackolder engagement is challenging---yet their projections are inconsistent with actual cholera incidence data from 2019 to 2022.
+Such findings underscore the limitations of current standards.
+Saltelli et al.
+\cite{saltelli19} suggests that principles and structres based on statistical principles can improve model based outcomes.
+Hence, our recommendations advocate for rigorous statistical calibration and evaluation of dynamic models, thereby expanding on the prevailing guidance from sources like \cite{behrend20}.
+We specifically emphasize principles that proved essential in our case study.
+Complementary methodological suggestions arising from a spatio-temporal analysis of COVID-19 are detailed in \cite{li23}.}
+
 \newcommand\editMechModels{Mechanistic models representing biological phenomena are valuable for epidemiology and consequently for public health policy \cite{lofgren14,mccabe21}.
@@
 In some situations, modern machine learning methods can outperform mechanistic models on epidemiological forecasting tasks \cite{lau22,baker18}.
-The predictive skill of non-mechansitic models can reveal limitations in mechanistic models, but cannot readily replace the scientific understanding obtained by describing the biological dynamics of the system in a mathematical model \cite{baker18}.}
+The predictive skill of non-mechansitic models can reveal limitations in mechanistic models, but cannot readily replace the scientific understanding obtained by describing the biological dynamics of the system in a mathematical model \cite{baker18,prosperi20}.}
 
-\newcommand\editDiscussionOne{In our cholera case study, we highlight the importance of model diagnostics; they are instrumental in enhancing model fits, thus leading to more trustworthy conclusions.
-We compared the Models 1--3 of \cite{lee20} to simple statistical benchmarks, enabling us to identify deficiencies in the models that could be improved.
-In the case of Model~3, comparing a fitted model to a benchmark and then following up by questioning why the model did not describe the data as efficiently as its benchmark helped us discern that the model fell short in explaining the surge in cholera transmission following Hurricane Matthew.
-This led us to suggest a model modification.
-Because Model~1 outperformed its benchmark model, we were able to confidently scrutinize the implications of the fitted model.
-We found that the seasonality term of Model~1 closely mirrors the seasonal rainfall patterns in Haiti, reaffirming alternative evidence of the importance of rainfall as a driver of cholera infection and bolstering our confidence in Model~1's ability to capture the critical dynamics of the cholera outbreak.}
+% \newcommand\editDiscussionOne{Model diagnostics are instrumental in enhancing model fits and identifying limitations of models that cannot readily be fixed.
+In our case study, we compared the Models~1--3 of \cite{lee20} to simple statistical benchmarks, enabling us to identify deficiencies in the models that could be improved.
+In the case of Model~3, comparing a fitted model to a benchmark and then following up by questioning why the model did not describe the data as efficiently as a simple statistical model helped us discern that the model fell short in explaining the surge in cholera transmission following Hurricane Matthew.
+This led us to suggest a model modification that enchanced model fit.
+When a mechanistic model beats a statistical benchmark, we can begin to critically evalutate the causal implications of the fitted model, carefully acknowledging the fact that beating a simple benchmark alone does not imply that the model is both a useful quantitative and qualitative description of the data.
+In Model~1, we found that the flexible seasonality term closely mirrors the seasonal rainfall patterns in Haiti, reaffirming alternative evidence of the importance of rainfall as a driver of cholera infection and increasing our confidence in Model~1's ability to capture the critical dynamics of the cholera outbreak.}
+% 
+\newcommand\editDiscussionOne{Model diagnostics are instrumental for improving model fit and exposing unresolvable limitations.
+In our case study, we compared the three models from Lee et al.
+(2020) to statistical benchmarks, revealing areas for enhancement.
+For example, comparisons of Model~3 to a benchmark revealed its inadequacy in accounting for the post-hurricane increase in transmission, leading to a beneficial model refinement.
+When a mechanistic model beats a statistical benchmark, we can begin to critically evalutate causal implications, carefully acknowledging that beating a simple benchmark alone does not imply the model is a useful qualitative description of the data.
+The recalibrated version of Model~1 outperforms its benchmark, enabling evaluation of causal implications.
+When doing so, we find that the fitted model provides a causal description of the dynamic system that is consitent with known features of the system, such as the importance of rainfall as a driver of cholera infection.
+The congruency between causal implications of the model and our belief about the dynamic system, coupled with a strong quantitative description of obsered data relative to a benchmark, increases our confidence that the model adequately captures critical dynamics of the system.}
 
-\newcommand\editDiscussionTwo{Likelihood based inference enables researchers to choose between any model that is evaluated on the same dataset using a metric such as AIC.
+% \newcommand\editDiscussionTwo{Likelihood based inference enables researchers to choose between any model that is evaluated on the same dataset using a metric such as AIC.
 Nested model variations are particularly useful as they enable formal statistical testing of the nested features.
@@
 
-\newcommand\editDiscussionThree{Another important consideration when fitting a mechanistic model to a dynamic system is the complexity of the model.
+% \newcommand\editDiscussionThree{Another important consideration when fitting a mechanistic model to a dynamic system is the complexity of the model.
 Mathematical models necessarily simplify complex dynamical systems in order to make inference possible.
@@
 Model benchmarks can assist in determining the level of complexity to be used in the model, enabling researchers to gauge a model's relative performance against a better-understood statistical model.}
+% 
+\newcommand\editDiscussionTwo{When fitting a mechanistic model to a dynamic system, the complexity of the model warrants careful attention.
+Mathematical models simplify complex dynamical systems, highlighting essential features that both facilitate scientific understanding and enable inference.
+In this simplification process, careful consideration must be given to which facets of the system are incorporated into the model and which are omitted, as these choices can shape the conclusions drawn from the model.
+These modeling decisions are driven both by the need for tractable inference and by the pursuit of scientific parsimony.
+Standard modeling choices—such as adopting a deterministic or stochastic approach, or incorporating a meta-population structure—reflect the balance between realism and simplicity.} %Benchmarking models against established and simpler statistical models in a likelihood based framework can help researchers determine an appropriate level of complexity, ensuring that models are both manageable and meaningful in their portrayal of the underlying dynamics.}
+
+\newcommand\editDiscussionThree{Along with enabling comparisons to statistical benchmarks, likelihood-based methods aid in determining an appropriate level of model complexity.
+Models fit to the same data can be compared using a criteria such as AIC.
+Nested model variations are particularly useful as they enable formal statistical testing of the nested features.
+We illustrated the benefits of examining nested model features, enhancing all three models.
+Model~1 included a time-varying transmission rate; Model~2 assessed a phase-shift parameter in seasonal cholera peaks; Model~3 incorporated hurricane-related parameters.
+Each addition received strong statistical support over their simpler predecessors.}
 
 \newcommand\editDiscussionFour{Unmodeled features of a dynamic system can lead to spurious or misleading parameter estimates if the feature greatly impacts observed data.
-This is because the fitted model must compensate for these unaccounted-for features using its inherent flexibility, particularly in the calibration of parameters.
+This is because the model must compensate for these unaccounted features using its inherent flexibility.
 In deterministic models, parameter estimation is the sole source of such flexibility.
-When the influence of unmodeled dynamics is substantial, it can skew parameter estimates toward scientifically implausible values, a phenomenon observed with Model~2 in our case study.
+When the influence of unmodeled dynamics is substantial, it can skew parameter estimates toward scientifically implausible values, a phenomenon observed in the estimated parameter values of Model~2.
 Incorporating demographic and environmental stochasticity into models can mitigate the impact of these unmodeled features.
-These stochastic phenomena are not only arguably present in most biological systems, but their inclusion in a model also allows observed data variations to be attributed to inherent randomness rather than to distorted parameter values.
-Furthermore, incorporating extra-demographic stochasticity in a model has also been shown to be beneficial \cite{he10,stocks20,li23}, which is consistent with our results: Models~1 and 3 suggest the presence of extra-demographic stochsticity, as evidenced by the confidence intervals provided in Suplement S8.}
+These stochastic phenomena are not only arguably present in biological systems, but their inclusion in a model also allows observed data variations to be attributed to inherent randomness rather than to distorted parameter values.
+Incorporating extra-demographic stochasticity in a model has also been shown to be beneficial \cite{he10,stocks20,li23}, which is consistent with our results: Models~1 and 3 suggest the presence of extra-demographic stochsticity, as evidenced by the confidence intervals provided in Suplement~S8.}
 
@@
 
-\newcommand\editDiscussionSix{If forecasts are an important component of a modeling task, the most recent information on the dynamic system should be considered with increased importance.
+% \newcommand\editDiscussionSix{If forecasts are an important component of a modeling task, the most recent information on the dynamic system should be considered with increased importance.
 This is particularly true for long time series, as the latent states from simulations from initial conditions may be more likely to diverge from the truth as time increases.
@@
 Despite their limitations, deterministic models can offer valuable insights into dynamic systems; researchers should carefully weigh these limitations when using these models and presenting their results.}
+
+\newcommand\editDiscussionSix{If forecasts are an important component of a modeling task, the forecasts should be consistent with the available data, particularly at the most recently available time points.
+In our case study, we do this by simulating forward from the filtering distribution, as this proceedure conditions latent variables on the available data.
+This type of forecasting, however, is not possible using a deterministic model, where only initial conditions and parameter values can influence forecasts.
+Deterministic simulations may also result in over-confidence in model forecasts, as they can only account for uncertainty due to the parameter estimation proceedure \cite{king15}.
+Despite their limitations, deterministic models can offer valuable insights into dynamic systems.
+As with all models, researchers should consider and acknoweledge these limitations when applying deterministic models and interpreting their outcomes.}
 

### 2024-02-27 ms.Rnw
Second round of edits based on Ed's comments

--- 2024-02-26 ms.Rnw
@@
 For $\vaccCounter\in\seq{1}{\vaccClass}$, the vaccination cohort $\vaccCounter$ is assigned a time $\tau_{\vaccCounter}$, and we take $\symptomFrac_\vaccCounter(t) = c \, \vaccineEfficacy^*(t-\tau_{\vaccCounter})$
-where $\vaccineEfficacy^*(t)$ is efficacy at time $t$ since vaccination for adults, taken from \cite{lee20}, Table~S4, and $c=\big(1-(1-0.4688)\times 0.11\big)$ is a correction to allow for reduced efficacy in the 11\% of the population aged under 5 years.
+where $\vaccineEfficacy^*(t)$ is efficacy at time $t$ since vaccination for adults, \old{taken from [4], Table~S4,} \new{a step-function represented in Table~S4 of \cite{lee20}}, and $c=\big(1-(1-0.4688)\times 0.11\big)$ is a correction to allow for reduced efficacy in the 11\% of the population aged under 5 years.
 Single and double vaccine doses were modeled by changing the waning of protection; protection was modeled as equal between single and double dose until 52 weeks after vaccination, at which point the single dose becomes ineffective.

### 2024-02-27 r1/response.tex

@@
 Thank you for this suggestion, a sentence on the overall effect of the parameter estimate has been included.
-For reference, the maxmum likelihood estimate of the parameter corresponds with a 7.3\% reduction to the transmission rate over the course of the outbreak, with a $95\%$ confidence interval $(1.8\%, 17.9\%)$ for the overall reduction in transmission.
+For reference, the maxmum likelihood estimate of the parameter corresponds to a 7.3\% reduction to the transmission rate over the course of the outbreak, with a $95\%$ confidence interval $(1.8\%, 17.9\%)$ for the overall reduction in transmission.
 

### 2024-02-27 edits.tex

@@
 
-%They recommend five guiding principles: stakeholder engagement, complete model documentation, complete description of data used, communicating uncerainty, and testable model outcomes.
-% \newcommand\editModelingGuidanceFirst{There are many guidelines available regarding the use of mechanistic models for policy influence.
-For instance, Behrend et al.~\cite{behrend20} conducted an extensive literature review on modeling principles and standards.
-They used this review to develop general recommendations for policy-guiding modeling.
-This review is particularly relevant given that their recommendations are specifically intended for the modeling of neglected tropical diseases, which includes cholera.
-Their resulting recommendations for communicating model results align with other manifestos on the proper use of models intended to inform policy \cite{saltelli20,donnelly18}, each emphasizing the importance of involving stakeholders in the modeling process, transparency, and communicating uncertainty.
-Other recommendations that focus on issues of model calibration and assessment also exist \cite{dahabreh17,egger17,penaloza15}, demonstrating the considerable amount of published guidance on how to effectively model dynamic systems.}
-
-% \newcommand\editModelingGuidance{There are many guidelines available regarding the use of mechanistic models for policy influence.
-For instance, Behrend et al.~\cite{behrend20} conducted an extensive literature review on modeling principles and standards.
-Their recommendations generally align with other manifestos on the proper use of models intended to inform policy \cite{saltelli20,donnelly18}, each emphasizing the importance of involving stakeholders in the modeling process, transparency, and communicating uncertainty.
-Other recommendations that focus on issues of model calibration and assessment also exist \cite{dahabreh17,egger17,penaloza15}, demonstrating the considerable amount of published guidance on how to effectively model dynamic systems.
-Based on our assessment and literature review, \cite{lee20} was among the studies that best followed these general recommendations for policy-driven modeling.
-Concretly, \cite{lee20} follows at least four of the five principles outlined in \cite{behrend20}: \emph{complete model documentation}, \emph{complete description of data used}, \emph{communicating uncertainty}, and \emph{testable model outcomes}.
+\newcommand\editModelingGuidanceEd{Many guidelines have been proposed regarding the use of mechanistic models to inform policy, reviewed by Behrend et al.~\cite{behrend20}.
+The recommendations of \cite{behrend20} generally align with other manifestos on the proper use of models for informing policy \cite{saltelli20,donnelly18}, each emphasizing the importance of involving stakeholders in the modeling process, transparency, and communication of uncertainty.
+Other recommendations focus on issues of model calibration and assessment \cite{dahabreh17,egger17,penaloza15}.
+Based on our literature review, \cite{lee20} was among the studies that best followed these general recommendations for policy-driven modeling.
+Concretly, \cite{lee20} follows four of the five principles outlined in \cite{behrend20}: \emph{complete model documentation}, \emph{complete description of data used}, \emph{communicating uncertainty}, and \emph{testable model outcomes}.
 Determining the level of adherence to the first principle, \emph{stakeholder engagement}, is difficult based solely on the article.
@@
 Our general recommendations therefore complement and extend existing guidelines by focusing on the methodological tasks of calibrating and evaluating dynamic models in a rigorous statistical framework.
-We specifically emphasize principles that proved essential in our case study; complementary methodological suggestions arising from a spatio-temporal analysis of COVID-19 are given in \cite{li23}.}
-
-% \newcommand\editModelingGuidance{Numerous guidelines for using mechanistic models to inform policy exist, yet they often lack implementation specifics.
-Behrend et al.~\cite{behrend20}, for instance, stress the importance stakeholder engagement, transparency, and uncertainty communication.
-Other guidelines for mechanistic modeling emphasize similar principles \cite{saltelli20,donnelly18}, demonstrating that the importance of these ideas is well known.
-Additional literature discusses model calibration and evaluation techniques \cite{dahabreh17,egger17,penaloza15}, but do not make specific recommendations on how to effectively implement these techniques in practice.
-Our literature review reveals gaps in applying principles found in \cite{behrend20,saltelli20,donnelly18,dahabreh17,egger17,penaloza15} effectively.
-For example, despite ostensibly adhering to four out of five key recommendations from \cite{behrend20}---determining the level of adherence to the remaining principle, \emph{stakeholder engagement}, is difficult based solely on the article---the projections of \cite{lee20} show a clear misalignment with actual cholera incidence data from 2019 to 2022.
-Such findings underscore the limitations of current standards.
-Saltelli et al.
-\cite{saltelli19} suggests that principles and structres based on statistical principles can improve model based outcomes.
-Hence, our recommendations advocate for rigorous statistical calibration and evaluation of dynamic models while complimenting existing recommendations such as \cite{behrend20}.
-We specifically emphasize principles that proved essential in our case study; complementary methodological suggestions arising from a spatio-temporal analysis of COVID-19 are detailed in \cite{li23}.}
+We specifically emphasize principles that proved relevant to our case study; complementary methodological suggestions arising from a spatio-temporal analysis of COVID-19 are given in \cite{li23}.}
 
@@
 
-% \newcommand\editDiscussionSix{If forecasts are an important component of a modeling task, the most recent information on the dynamic system should be considered with increased importance.
-This is particularly true for long time series, as the latent states from simulations from initial conditions may be more likely to diverge from the truth as time increases.
-One way to address this issue is by simulating forward from the filtering distribution at the last available time point, as this proceedure considers available evidence about the latent states of a system under the framework of the dynamic model.
-This type of forecasting, however, is not possible using a deterministic model.
-Deterministic simulations may also result in over-confidence in model forecasts, as they can only account for uncertainty due to the parameter estimation proceedure \cite{king15}.
-Despite their limitations, deterministic models can offer valuable insights into dynamic systems; researchers should carefully weigh these limitations when using these models and presenting their results.}
-
 \newcommand\editDiscussionSix{If forecasts are an important component of a modeling task, the forecasts should be consistent with the available data, particularly at the most recently available time points.

### 2024-02-27 ms.Rnw

--- 2024-02-27 ms.Rnw
@@
 \old{In this article, we discuss common modeling decisions that may not provide an adequate statistical explanation of the data, and we show that attention to statistical fit can lead to more accurate policy evaluations}.
-In this article, we conduct a re-analysis of the 2010 cholera outbreak in Haiti.
-\new{We use this case study to investigate current procedures for fitting mathematical models to time series data, while identifying limitations of these methodologies and proposing remedies.}
-We present methodology for diagnosing how well a model describes observed data\new{.}\old{, and provides careful justification of valid conclusions from the fitted model.}
+\new{In this article, we conduct a re-analysis of the 2010 cholera outbreak in Haiti.
+We use this case study to investigate current procedures for fitting mathematical models to time series data, while identifying limitations of these methodologies and proposing remedies.}
+\new{We present} \old{Our analysis presents} methodology for diagnosing how well a model describes observed data\new{.}\old{, and provides careful justification of valid conclusions from the fitted model.}
 Objective measures are used to \old{benchmark} \new{evaluate} model fit; when \old{these}\new{objective measures of the goodness-of-fit of a model} are combined with reproducibility, a framework emerges for continual improvement \new{to model based inference} when revisiting the data\old{ and models}.
@@
 Parameters that were added during our re-analysis and were not considered by Lee et al. are indicated with $^*$.
-\new{Confidence intervals for model parameters are given in the supplement (Sec.~S8).}
+\new{Confidence intervals for model parameters are given in supplement~S8.}
 Translations back into the notation of \cite{lee20} are given in \nameref{S1_Table}.
@@
 The observable random variables $\bm{Y}_{1:N}$ are assumed to be conditionally independent given $\bm{X}_{0:N}$.
-Together, with the density for the initial value of the latent state $\bm{X}_{0} = \bm{X}(t_0)$, each model defines a joint density $f_{\bm{X}_{0:N}, \bm{Y}_{1:N}}\big(\bm{x}_{0:N}, \bm{y}_{1:N};\paramVec\big)$, where $\paramVec$ is a parameter vector that indexes the model.
+Together, with the density for the initial value of the latent state $\bm{X}_{0} = \bm{X}(t_0)$, each model \new{defines} \old{assumes} a joint density $f_{\bm{X}_{0:N}, \bm{Y}_{1:N}}\big(\bm{x}_{0:N}, \bm{y}_{1:N};\paramVec\big)$, where $\paramVec$ is a parameter vector that indexes the model.
 The observed data $\bm{y}_{1:N}^*$, along with the unobserved true value of the latent state, are modeled as a realization of this joint distribution.

### 2024-02-27 r1/response.tex
letterhead for response
implementing changes discussed in meeting

@@
 As for any analysis of observational data, we must be alert to the possibility of confounding.
-For a covariate, this shows up in a similar way to regression analysis: the covariate under investigation could be a proxy for some other unmodeled \new{phenomenon} or unmeasured covariate.}
+For a covariate, this shows up in a similar way to regression analysis: the covariate under investigation could be a proxy for some other unmodeled phenomenon or unmeasured covariate.
+The statistical evidence of a trend in transmission rate in this model could be explained by any trending variable (such as hygiene improvements, or changes in population behavior), resulting in confounding from collinear covariates.
+Alternatively, the trend could be attributed to a decreasing reporting rate rather than decreasing transmission rate.
+This could be formally tested by comparing models with either trend specification.
+We did not do this because evidence suggests that reporting rate was maintained or increased (Figure~1 of \cite{rebaudet21}).
+We instead argue that a decreasing transmission rate is a plausible way to explain the decrease in cases over time, as there is alternative evidence that supports this model \cite{rebaudet19CATI,rebaudet21,michel19}.
+It is not practical to test all remotely plausible model variations, yet a strongly supported conclusion should avoid ruling out untested hypotheses.
+The robust statistical conclusion for our analysis is that a model which allows for change fits better than one which does not, and a trend in transmission is a plausible way to do this.}
 

### 2024-02-27 edits.tex
implementing changes discussed in meeting

@@
 
-\newcommand\editModelingGuidance{Numerous guidelines for using mechanistic models to inform policy exist, yet they often lack implementation specifics.
-Behrend et al.~\cite{behrend20}, for instance, stress the importance stakeholder engagement, transparency, and uncertainty communication.
-These and other related principles are echoed by other well-cited studies \cite{saltelli20,donnelly18}, suggesting that the importance of these ideas is well known.
-Additional literature discusses model calibration and evaluation techniques \cite{dahabreh17,egger17,penaloza15}, but do not make specific recommendations on how to effectively implement these techniques in practice.
-Our literature review reveals gaps in applying principles found in \cite{behrend20,saltelli20,donnelly18,dahabreh17,egger17,penaloza15} effectively.
-For example, \cite{lee20} largely adhere to the principles of \cite{behrend20}---though assessing the extent of stackolder engagement is challenging---yet their projections are inconsistent with actual cholera incidence data from 2019 to 2022.
-Such findings underscore the limitations of current standards.
-Saltelli et al.
-\cite{saltelli19} suggests that principles and structres based on statistical principles can improve model based outcomes.
-Hence, our recommendations advocate for rigorous statistical calibration and evaluation of dynamic models, thereby expanding on the prevailing guidance from sources like \cite{behrend20}.
-We specifically emphasize principles that proved essential in our case study.
+\newcommand\editModelingGuidance{Numerous guidelines have been proposed for using mechanistic models to inform policy, reviewed in \cite{behrend20}.
+Behrend et al.~\cite{behrend20} identify the importance of stakeholder engagement, transparency, reproducibility, uncertainty communication, and testable model outcomes.
+These and related principles are echoed by other influential articles \cite{saltelli20,donnelly18}.
+Additional literature emphasizes model calibration and evaluation techniques \cite{dahabreh17,egger17,penaloza15}.
+These guidelines often lack implementation specifics.
+As an example, \cite{lee20} largely adhere to the principles of \cite{behrend20}---though assessing the extent of stake-holder engagement is challenging---yet their projections are inconsistent with actual cholera incidence data from 2019 to 2022, demonstrating the limitations of current standards.
+We provide methodology for rigorous statistical calibration and evaluation of dynamic models (as advocated by \cite{saltelli19}), thereby expanding on the prevailing guidance.
+We specifically emphasize principles that prove essential in our case study.
 Complementary methodological suggestions arising from a spatio-temporal analysis of COVID-19 are detailed in \cite{li23}.}
@@
 \newcommand\editMetaPop{In our literature review, 17 articles considered dynamic models that incorporate spatial heterogeneity \cite{lee20,tuite11,pasetto18,fitzgibbon20,eisenberg13,rinaldo12,chao11,abrams13,trevisin22,sallah17,collins14,kelly16,azman12,leung22,kuhn14,mari15,gatto12}.
-All but four \cite{lee20,pasetto18,sallah17,azman12} of these studies used deterministic dynamic models: this greatly simplifies the process of calibrating model parameters to incidence data, but unfortunately that is not enough to guarantee that the calibrated model provides an adequate statistical explanation of the data [I'VE EDITED THIS...
-IT MAY NEED MORE WORK, BUT WE MAY WANT TO REMIND THE READER THAT EASIER FITTING IS NOT SUCH A GOOD THING IF THE FIT IS JUNK].
+All but four \cite{lee20,pasetto18,sallah17,azman12} of these studies used deterministic dynamic models: this greatly simplifies the process of calibrating model parameters to incidence data, though deterministic models can struggle to describe complex stochastic dynamics.
 The model in \cite{pasetto18} was fit using an Ensemble Kalman Filter (EnKF) \cite{evensen09}; though EnKF scales favorably with the number of spatial units, it relies on linearization of latent states which can be problematic for highly nonlinear systems \cite{evensen22,ionides21}.

### 2024-02-27 ms.Rnw
implementing changes discussed in meeting

@@
 For a covariate, this shows up in a similar way to regression analysis: the covariate under investigation could be a proxy for some other unmodeled \new{phenomenon} or unmeasured covariate.
-\old{For a mechanism, the model feature could in principle explain the data by helping to account for some different unmodeled phenomenon.}
-
+\old{For a mechanism, the model feature could in principle explain the data by helping to account for some different unmodeled phenomenon.
 The statistical evidence of a trend in transmission rate in this model could be explained by any trending variable (such as hygiene improvements, or changes in population behavior), resulting in confounding from collinear covariates.
-\old{Alternatively, the trend could be attributed to a decreasing reporting rate rather than decreasing transmission rate, resulting in confounded mechanisms.}
-\new{Because we only have access to incidence data, the decrease in observed cases over the course of the outbreak may alternatively be attributed to a trend in other model parameters, such as a decrease in the reporting rate over time.
-This results in confounded mechanisms, as one may not formally test all possible ways to include a trending parameter.
+Alternatively, the trend could be attributed to a decreasing reporting rate rather than decreasing transmission rate, resulting in confounded mechanisms.
+The robust statistical conclusion is that a model which allows for change fits better than one which does not---we argue that a decreasing transmission rate is a plausible way to explain this, but the incidence data themselves do not provide enough information to pin down the mechanism.}
+\new{The statistical evidence of a trend in transmission rate in this model could be explained by any trending variable (such as hygiene improvements, or changes in population behavior), resulting in confounding from collinear covariates.
+Alternatively, the trend could be attributed to a decreasing reporting rate rather than decreasing transmission rate.
+This could be formally tested by comparing models with either trend specification.
+We did not do this because evidence suggests that reporting rate was maintained or increased (Figure~1 of \cite{rebaudet21}).
+We instead argue that a decreasing transmission rate is a plausible way to explain the decrease in cases over time, as there is alternative evidence that supports this model \cite{rebaudet19CATI,rebaudet21,michel19}.
+It is not practical to test all remotely plausible model variations, yet a strongly supported conclusion should avoid ruling out untested hypotheses.
+The robust statistical conclusion for our analysis is that a model which allows for change fits better than one which does not, and a trend in transmission is a plausible way to do this.}
+
+% The statistical evidence of a trend in transmission rate in this model could be explained by any trending variable (such as hygiene improvements, or changes in population behavior), resulting in confounding from collinear covariates.
+% \old{Alternatively, the trend could be attributed to a decreasing reporting rate rather than decreasing transmission rate, resulting in confounded mechanisms.}
+% \new{Because we only have access to incidence data, the decrease in observed cases over the course of the outbreak may alternatively be attributed to a trend in other model parameters, such as a decrease in the reporting rate over time.
+% This results in confounded mechanisms, as one may not formally test all possible ways to include a trending parameter.
 In such a case, t}\old{T}he robust statistical conclusion is that a model which allows for change fits better than one which does not.\old{---we argue that a decreasing transmission rate is a plausible way to explain this, but} \new{Because} the incidence data themselves \old{do} \new{may} not provide enough information to pin down the mechanism\new{, a strongly supported conclusion should avoid ruling out untested hypotheses.}
-\new{In this case study, we did not explicitly test if the incidence data favor a trend in transmission over a trend in reporting rate.
+% \new{In this case study, we did not explicitly test if the incidence data favor a trend in transmission over a trend in reporting rate.
 We argue that a decreasing transmission rate is a plausible way to explain the decrease in cases over time, as there is alternative evidence that supports this model \cite{rebaudet19CATI,rebaudet21,michel19}.

### 2024-02-29 edits.tex
discussion edits

@@
 % 
-\newcommand\editDiscussionOne{Model diagnostics are instrumental for improving model fit and exposing unresolvable limitations.
+\newcommand\editDiscussionOne{Model diagnostics are a key tool for exposing unresolved model limitations and improving model fit.
 In our case study, we compared the three models from Lee et al.
@@
 For example, comparisons of Model~3 to a benchmark revealed its inadequacy in accounting for the post-hurricane increase in transmission, leading to a beneficial model refinement.
-When a mechanistic model beats a statistical benchmark, we can begin to critically evalutate causal implications, carefully acknowledging that beating a simple benchmark alone does not imply the model is a useful qualitative description of the data.
-The recalibrated version of Model~1 outperforms its benchmark, enabling evaluation of causal implications.
-When doing so, we find that the fitted model provides a causal description of the dynamic system that is consitent with known features of the system, such as the importance of rainfall as a driver of cholera infection.
-The congruency between causal implications of the model and our belief about the dynamic system, coupled with a strong quantitative description of obsered data relative to a benchmark, increases our confidence that the model adequately captures critical dynamics of the system.}
+When a mechanistic model is competitive with statistical benchmarks, we have a license to begin critical evaluation of its causal implications.
+If a model falls far behind simple benchmarks, there is likely to be substantial limitation in the data analysis that should be identified and remedied.
+In our case study, the recalibrated version of Model~1 outperforms its benchmark, enabling evaluation of causal implications.
+When doing so, we find that the fitted model provides a causal description of the dynamic system that is consistent with known features of the system, such as the importance of rainfall as a driver of cholera infection.
+The congruency between causal implications of the model and our belief about the dynamic system, coupled with a strong quantitative description of observed data relative to a benchmark, provides support for viewing the model as a plausible quantitative representation of the system under investigation.}
 
@@
 % 
-\newcommand\editDiscussionTwo{When fitting a mechanistic model to a dynamic system, the complexity of the model warrants careful attention.
-Mathematical models simplify complex dynamical systems, highlighting essential features that both facilitate scientific understanding and enable inference.
-In this simplification process, careful consideration must be given to which facets of the system are incorporated into the model and which are omitted, as these choices can shape the conclusions drawn from the model.
+\newcommand\editDiscussionTwo{When fitting a mechanistic model to a dynamic system, the complexity of the model warrants consideration.
+Mathematical models provide simplified representations of complex systems, with the simplicity serving both to facilitate scientific understanding and to enable statistical inference on unknown parameters.
+In our case study, employing a deterministic dynamic model was found to be an over-simplification by comparing model fit with benchmarks and stochastic models.
+Including stochastic metapopulation dynamics permitted investigation of spatiotemporal data, the main cost being additional complexity of numerical computations.}
+%In this simplification process, careful consideration must be given to which facets of the system are incorporated into the model and which are omitted, as these choices can shape the conclusions drawn from the model.
 These modeling decisions are driven both by the need for tractable inference and by the pursuit of scientific parsimony.
-Standard modeling choices—such as adopting a deterministic or stochastic approach, or incorporating a meta-population structure—reflect the balance between realism and simplicity.} %Benchmarking models against established and simpler statistical models in a likelihood based framework can help researchers determine an appropriate level of complexity, ensuring that models are both manageable and meaningful in their portrayal of the underlying dynamics.}
+%Standard modeling choices—such as adopting a deterministic or stochastic approach, or incorporating a meta-population structure—reflect the balance between realism and simplicity.} %Benchmarking models against established and simpler statistical models in a likelihood based framework can help researchers determine an appropriate level of complexity, ensuring that models are both manageable and meaningful in their portrayal of the underlying dynamics.}
 
@@
 Models fit to the same data can be compared using a criteria such as AIC.
-Nested model variations are particularly useful as they enable formal statistical testing of the nested features.
-We illustrated the benefits of examining nested model features, enhancing all three models.
-Model~1 included a time-varying transmission rate; Model~2 assessed a phase-shift parameter in seasonal cholera peaks; Model~3 incorporated hurricane-related parameters.
-Each addition received strong statistical support over their simpler predecessors.}
+Nested model variations are particularly useful as they enable formal statistical testing of the nested features via likelihood ratio tests.
+Our case study demonstrated the examination of nested model features for all three models.
+Model~1 investigated a time-varying transmission rate; Model~2 assessed a phase-shift parameter in seasonal cholera peaks; Model~3 incorporated hurricane-related parameters.}
 
-\newcommand\editDiscussionFour{Unmodeled features of a dynamic system can lead to spurious or misleading parameter estimates if the feature greatly impacts observed data.
+\newcommand\editDiscussionFour{Unmodeled features of a dynamic system can lead to spurious or misleading parameter estimates if the features substantially impact observed data.
 This is because the model must compensate for these unaccounted features using its inherent flexibility.
-In deterministic models, parameter estimation is the sole source of such flexibility.
-When the influence of unmodeled dynamics is substantial, it can skew parameter estimates toward scientifically implausible values, a phenomenon observed in the estimated parameter values of Model~2.
-Incorporating demographic and environmental stochasticity into models can mitigate the impact of these unmodeled features.
-These stochastic phenomena are not only arguably present in biological systems, but their inclusion in a model also allows observed data variations to be attributed to inherent randomness rather than to distorted parameter values.
-Incorporating extra-demographic stochasticity in a model has also been shown to be beneficial \cite{he10,stocks20,li23}, which is consistent with our results: Models~1 and 3 suggest the presence of extra-demographic stochsticity, as evidenced by the confidence intervals provided in Suplement~S8.}
+In deterministic models, the choice of parameters is the sole source of such flexibility.
+For our case study, some of the parameter estimates for the deterministic Model~2 are implausible [WHICH ONES?] and this may be explained by compensation for model misspecification.
+Incorporating demographic and environmental stochasticity into models can mitigate the impact of unmodeled features.
+Stochastic phenomena are not only arguably present in biological systems, but their inclusion in a model also allows observed data variations to be attributed to inherent uncertainty rather than to distorted parameter values.
+Models~1 and 3 suggest the presence of extra-demographic stochsticity \cite{he10,stocks20,li23}, as evidenced by the confidence intervals provided in Supplement~S8.
+}
 
@@
 In this paper, we demonstrated how this model class could be calibrated to incidence data using the innovative IBPF algorithm.
-One of only a few examples of fitting a model with this level of complexity via maximum likelihood, this case study exemplifies the algorithm's potential benefits and provides an example for future researchers on a possible approach to fitting a high-dimensional non-linear model.}
+One of only a few examples of fitting a model with this level of complexity via maximum likelihood, this case study exemplifies the algorithm's potential benefits and provides an example for future researchers on a possible approach to fitting a high-dimensional non-linear model.
+THIS PARAGRAPH SEEMS TO OVERLAP OTHERS. MAYBE FIT WHATEVER IS NEEDED INTO OTHER PARAGRAPHS AND DELETE?}
 
@@
 Despite their limitations, deterministic models can offer valuable insights into dynamic systems.
-As with all models, researchers should consider and acknoweledge these limitations when applying deterministic models and interpreting their outcomes.}
+As with all models, researchers should consider and acknoweledge these limitations when applying deterministic models and interpreting their outcomes.
+I WONDER WHETHER PEOPLE SERIOUSLY FORECASTING WITH DETERMINISTIC MODELS HAVE OTHER TRICKS.
+FOR EXAMPLE, WITH WEATHER FORECASTING MODELS, THEY ARE OFTEN WRITTEN AS DETERMINISTIC BUT PEOPLE DO SOMETHING ENKF-LIKE ANYHOW.
+JEFF SHAMAN HAS ALSO DONE THINGS LIKE THIS WITH EPI MODELS.
+ANOTHER OPTION COULD BE TO FIT IN A MOVING WINDOW.
+ANOTHER THING: THE PENULTIMATE SENTENCE MAKES A FAIRLY STRONG CLAIM THAT IS NOT SUPPORTED BY EVIDENCE HERE.
+PERHAPS WE CAN REMIND READERS THAT DETERMINISTIC MODELS REMAIN WIDELY USED, BUT IT WOULD BE NICE TO CLARIFY WHAT THEY ARE GOOD FOR SINCE WE SPEND SO MUCH TIME ON THEIR LIMITATIONS.}
 

### 2024-02-29 ms.Rnw
discussion edits

@@
 \old{This article demonstrates that fitting appropriate scientific models to guide policy is a challenging statistical task.}
-\new{However, developing credible models---which are supported by data and can lend insights into a dynamic system---remains a challenging task.
-In this article, we demonstrate that increasing the current standards of mathematical modeling of ecological systems can increase our capacity to make sense of a complex, nonlinear system's dynamics and make data-driven inferences.
-We suggest devoting more effort to assess dynamic models intended for system inference.
-Explicitly, we urge researchers to search for indications in available data that quantify the reliability of model-based inferences.
-We illustrate that this can be achieved by comparing fitted models to a suitable statistical benchmark and verifying that model outcomes---including estimated parameters, latent states, and seasonal trends---are justified in the system context.}
+\new{However, the development of credible models---which are supported by data and can lend quantitative insights into a dynamic system---remains a challenging task.
+In this article, we demonstrate opportunites available for raising the current standards of statistical inference for mathematical models of biological systems.
+
+We present methodology consistent with existing guidelines but exceeding standard practice.  
+In particular, we show the value of comparing the likelihood of fitted mechanistic models versus non-mechanistic benchmarks.
+Further, we use likelihood-based diagnostic methods to identify specific limitations of proposed models.
+We use likelihood profile methods to assess parameter identifiability and to construct confidence intervals for parameter estimates. 
+We assess the accuracy of numerical approximations involved in our calculations.
+When reaching conclusions, we identify potential consequences of confounded variables and model misspecification.
+}
 

### 2024-02-29 r1/response.tex
Recompiling after merge
Minor edits for clarity

--- 2024-02-27 r1/response.tex
@@
 \newcommand\muRS{\mu_{RS}}
+\newcommand\transmissionTrend{\zeta}
 \input{../edits.tex}
@@
 
-% \includegraphics[width=0.85\textwidth,trim={1.7cm 25.6cm 10cm 1.3cm},clip]{/Users/ionides/Dropbox/computing/latex/umich-stat-letterhead-2019.pdf}
+\includegraphics[width=0.85\textwidth,trim={1.7cm 25.6cm 10cm 1.3cm},clip]{umich-stat-letterhead-2019.pdf}
 
@@
 
-\article{we use an in-depth case study to motivate general procedures for fitting mathematical models to time series data, while explaining how these procedures help to address limitations of previous approaches.}
+\article{In this article, we conduct a re-analysis of the 2010 cholera outbreak in Haiti.
+We use this case study to investigate current procedures for fitting mathematical models to time series data, while identifying limitations of these methodologies and proposing remedies.}
 
@@
 
-We have added two paragraphs in the introduction discussing how recommendations we make in this article fit in the broader category of papers that make recommendations on fitting mechanistic models to dynamic systems.
+We have added a paragraph in the introduction discussing how recommendations we make in this article fit in the broader category of papers that make recommendations on fitting mechanistic models to dynamic systems.
 This includes discussing these two papers that we originally cited in more detail, as well as other articles on similar topics:
-
-\jwc{I combined the two paragraphs into a single paragraph that I think addresses Ed's concern of "brevity and clarity".
-This paragraph is still quite long and could use some editing, so revisit this later.}
 
@@
 
-We have also added some material in the discussion connecting our results to the stated goals of the paper from the abstract and introduction.
-We could have gone further and discussed in more detail the relationship to the references \cite{behrend20,saltelli20}, following your recommendation.
-However, our newly added reference \cite{li23} more directly addresses \cite{behrend20,saltelli20} and so we have referred the reader to that paper (currently available on arXiv) for additional discussion of this point.
+We have also added some material in the discussion that connects our results to the stated goals of the paper in the abstract and introduction.
 
@@
 
-It could be either of the above, though in the analysis we only consider compartments based on the vaccination scenarios outlined in the Forecasts section.
+It could be either of the above, the main point being that the subscript seperates individuals into compartments that have distinct dynamics.
+We try to make this more clear by adding the following sentences:
+
+\article{Here, $\vaccCounter=0$ corresponds to unvaccinated individuals, and $\vaccCounter \in \seq{1}{\vaccClass}$ describes hypothetical vaccination programs.
+Each program $\vaccCounter$ indexes differences in both the number of doses administered (one versus two doses per individual) and the round of vaccine administration, seperating individuals into compartments with distinct dynamics based on vaccination status.}
+
+In our analysis we only consider vaccination dynamics based on the scenarios outlined in the Forecasts section.
 For Model~1, the assumption made by Lee et al.~\cite{lee20}, and reused in our analysis, is that the added dynamics of the vaccination compartments is an increased probability that an infection is asymptomatic, as described in a paragraph that follows the quoted text.
-a clause has been added to this sentence in order to clarify this point:
-
-\article{Here, $\vaccCounter=0$ corresponds to unvaccinated individuals, and $\vaccCounter \in \seq{1}{\vaccClass}$ describes hypothetical vaccination programs, where each program $\vaccCounter$ indexes differences in both the number of doses administered (one versus two doses per individual) and the round of vaccine administration.}
 
@@
 Notably, this parameter can mathematically be estimated like all other parameters in the model, but their specific implementation of Model~2 does not permit its estimation.
-Parameters that are partially estimated via subjective measures is outside the scope of our article, so to avoid this the revised manuscript simply reads:
+A full discussion on parameters that are partially estimated via subjective measures is outside the scope of our article, so the revised manuscript simply reads:
 
-\article{$v_{\mathrm{rate}}= 10^{-12} \,\mbox{km$^2$yr$^{-1}$}$ was fixed at the value used by \cite{lee20}.}
+\article{$v_{\mathrm{rate}}= 10^{-12} \,\mbox{km$^2$yr$^{-1}$}$ was fixed at the value used in \cite{lee20}.}
 
@@
 
-Thank you for this suggestion, a sentence on the overall effect of the parameter estimate has been included.
-For reference, the maxmum likelihood estimate of the parameter corresponds to a 7.3\% reduction to the transmission rate over the course of the outbreak, with a $95\%$ confidence interval $(1.8\%, 17.9\%)$ for the overall reduction in transmission.
+Thank you for this suggestion, a sentence on the overall effect of the parameter estimate has been included in the revised manuscript:
+
+\article{The maximum likelihood estimate for $\transmissionTrend$ corresponds to a 7.3\% reduction to the transmission rate over the course of the outbreak, with a $95\%$ confidence interval of $(1.8\%, 17.9\%)$ for the overall reduction in transmission.}
 
@@
 As for any analysis of observational data, we must be alert to the possibility of confounding.
-For a covariate, this shows up in a similar way to regression analysis: the covariate under investigation could be a proxy for some other unmodeled \new{phenomenon} or unmeasured covariate.}
+For a covariate, this shows up in a similar way to regression analysis: the covariate under investigation could be a proxy for some other unmodeled phenomenon or unmeasured covariate.}
 
-\article{The statistical evidence of a trend in transmission rate in this model could be explained by any trending variable (such as hygiene improvements, or changes in population behavior), resulting in confounding from collinear covariates.
-Because we only have access to incidence data, the decrease in observed cases over the course of the outbreak may alternatively be attributed to a trend in other model parameters, such as a decrease in the reporting rate over time.
-This results in confounded mechanisms, as one may not formally test all possible ways to include a trending parameter.
-In such a case, the robust statistical conclusion is that a model which allows for change fits better than one which does not.
-Because the incidence data themselves may not provide enough information to pin down the mechanism, a strongly supported conclusion should avoid ruling out untested hypotheses.
-In this case study, we did not explicitly test if the incidence data favor a trend in transmission over a trend in reporting rate.
-We argue that a decreasing transmission rate is a plausible way to explain the decrease in cases over time, as there is alternative evidence that supports this model \cite{rebaudet19CATI,rebaudet21,michel19}.
-Additionally, there is little scientific evidence to support a decrease in reporting rate (for example, see Figure~1 of \cite{rebaudet21}).}
+\article{\editConfound}
 
@@
 
-\jwc{This text below is something that Ed added.
+% \jwc{This text below is something that Ed added.
 I think it's a good idea.
 However, simply because of the overal length of the paper and total number of citations, I think it's probably okay to just leave this out}
-
-Howerver, we can alert the reader to the topic, and provide some additional references.
-Not infrequently, diagnostic issues are relegated to supplementary material and so we have indicated to interested readers some papers with extensive supplements demonstrating these diagnostic issues in action.
+%
+% Howerver, we can alert the reader to the topic, and provide some additional references.
+% Not infrequently, diagnostic issues are relegated to supplementary material and so we have indicated to interested readers some papers with extensive supplements demonstrating these diagnostic issues in action.
 % \cite{pons-salort18}
@@
 The $95\%$ confidence interval for Ouest in this figure is from $0-1094$ reported cases.
-This range may not be considered unreasonably given the inherent demographic and environmental stochasticity present in the dynamic system and all of the many possible outcomes of the outbreak from its initial state in 2010.
+This range may not be considered unreasonably large given the inherent stochasticity present in the dynamic system and all of the many possible outcomes of the outbreak from its initial state in 2010.
 % Shorter term forecasts obtained by conditioning on previous data can be obtained by simulating from the filtering distribution.
@@
 Objective measures are used to evaluate model fit; when objective measures of the goodness-of-fit of a model are combined with reproducibility, a framework emerges for continual improvement to model based inference when revisiting the data.
-This framework promotes scientific discovery and supports model-based policy recommendations by simplifying the process for researchers to build off of previous results.}
+This framework promotes scientific discovery and supports model-driven policy recommendations by simplifying the process of understanding and improving previous results.}
 
@@
 According to Lee et al~\cite{lee20}, this function was obtained by ``fitting a log-linear weighted regression model to the raw data from a published meta-analysis on killed OCV efficacy against medically attended culter-confirmed cholera (figure 2, appendix 3 p 6) \cite{bi17}".
-Aside from this statement, they provided limited details on how they came up with this function.
+Aside from this statement and the table, they no additional details on how they came up with this function.
 Our revised manuscript clarifies this point
@@
 Our starting point was the formulation of \cite{lee20}.
-% Judging by the referee's comment, they will not be suprised to learn that this structure led to difficulties in some situations.
 We did consider the effect of not calibrating initial values.
@@
 On a nationally aggregated scale, the total number of initial infectious individuals in the fixed scenario was 36500, and 33480 when the initial states were estimated.
-We have included additional material in the supplement for Initialization Models that demonstrates this.
+We have included additional material in the supplement for Initialization Models that demonstrates this. % TODO: Specifically link / cite supplement.
 
@@
 
+% TODO: Revisit this a bit more, giving explicit details in the manuscript and supplement that address this concern.
 % The cholera pathogen was recently introduced to Haiti, which facilitates the specification of initial conditions.
@@
 
-\article{The statistical evidence of a trend in transmission rate in this model could be explained by any trending variable (such as hygiene improvements, or changes in population behavior), resulting in confounding from collinear covariates.
-Because we only have access to incidence data, the decrease in observed cases over the course of the outbreak may alternatively be attributed to a trend in other model parameters, such as a decrease in the reporting rate over time.
-This results in confounded mechanisms, as one may not formally test all possible ways to include a trending parameter.
-In such a case, the robust statistical conclusion is that a model which allows for change fits better than one which does not.
-Because the incidence data themselves may not provide enough information to pin down the mechanism, a strongly supported conclusion should avoid ruling out untested hypotheses.
-In this case study, we did not explicitly test if the incidence data favor a trend in transmission over a trend in reporting rate.
-We argue that a decreasing transmission rate is a plausible way to explain the decrease in cases over time, as there is alternative evidence that supports this model \cite{rebaudet19CATI,rebaudet21,michel19}.
-Additionally, there is little scientific evidence to support a decrease in reporting rate (for example, see Figure~1 of \cite{rebaudet21}).}
+\article{If a mechanistic model including a feature (such as a representation of a mechanism, or the inclusion of a covariate) fits better than mechanistic models without that feature, and also has competitive fit compared to associative benchmarks, this may be taken as evidence supporting the scientific relevance of the feature.
+As for any analysis of observational data, we must be alert to the possibility of confounding.
+For a covariate, this shows up in a similar way to regression analysis: the covariate under investigation could be a proxy for some other unmodeled phenomenon or unmeasured covariate.}
+
+\article{\editConfound}
 
@@
 We have added prediction intervals due to the error in the log-normal measurement model.
-
-% We intentionally do not include predictive intervals in Fig~4. There are several reasons for this, one of which is mentioned in the article:
-% 
-% \article{In this analysis, we do not construct forecasts accounting for parameter uncertainty as our focus is on the estimation and diagnosis of mechanistic models.
-% Furthermore, we use the projections from a single point estimate to highlight the deficiency of deterministic models that the only variability in model projections is a result of parameter uncertainty, which can lead to over-confidence in forecasts \cite{king15}.}
-% 
-% \jwc{Below are some thoughts I had. We don't have to include all / any of this, but I thought it was something to think about on the topic of parameter uncertainty}
-% 
-% Parameter uncertainty in model forecasts is an important topic, one that we would have liked to have addressed in our article but felt that we could not since the article is already becoming longer than one would expect in this journal.
-% In our literature review of dynamic models for cholera in Haiti, and our experience with literature with dynamic modeling outside of this problem, the most common approaches to accounting for parameter uncertainty are the following:
-% 
-% \begin{itemize}
-%   \item Maximum likelihood etimation theory states that under certain conditions, the difference between the MLE and the generating model parameter is asymptoticly normally distributed with mean zero and variance determined by the inverse Fisher information matrix.
-Many numeric optimization techniques often approximate the curvature of the likelihood surface, thereby Fisher information matrix.
-%   This can then be used to obtain approximate confidence intervals using the asymptotic distribution of the MLE and the approximation to Fisher's information matrix.
-While mathematically and computationally convinient, the validity of these confidence intervals depends on asymptotics and assumptions that are often not appropriate for non-linear models.
-In fact, there is evidence that these types of confidence intervals are often far too narrow even in the most idealistic cases for dynamic models: linear Gaussian models fit with a large number of observations \cite{wheeler23}.
-%   Because of this, confidence regions created in this way likely too narrow, resulting in over-confidence in parameter estimates. 
-%   \item Alternatively, it is common to account for parameter uncertainty by performing a sensitivity analysis.
-This is often done when parameters were selected via non-likelihood based approaches, but our review contained instances (e.g.
-\cite{lee20}) where this was done even if parameters were calibrated by maximizing likelihoods.
-%   To perform a sensitivity anaylsis, researchers typically pick ranges of parameter estimates, creating a hypercube from which they obtain sample parameter values. 
-%   The bounds of this hypercube, however, are often obtained using approximations of Fisher's Information Matrix (the deficiencies of which are mentioned above), or by some non-data driven approach. 
-%   For example, bounds are often selected by arbitrarily adding / subtracting a percentage of the estimated parameter (where the bounds can be either larger or smaller than what should be used), or by using parameter estimates from other studies (in which case the parameter estimates may not make sense in the context of the model and data being considered, again resulting in intervals that may be arbitrarily wide or narrow).
-%   A persistent theme with all of the above approaches is that they enable researchers to effectively pick the level of uncertainty to present in their reports. 
-%   \item Sampling parameters from a Bayesian posterior.
-For many researchers, this may be one of the primary reasons they prefer Baysian methods over maximum likelihood estimates, as these methods do have an advantage in accounting for parameter uncertainty.
-%   Bayesian methods to estimate model parameters may not be possible for all scientific models of interest.
-We have been made aware of recent work that demonstrates cases where the authors would like to have a posterior distribution for their model, but the scientific model they propose is unable to be calibrated using state-of-the-art Bayesian approaches; they therefore use IF2 to obtain parameter estimates that result in good behaviour of particle filters, and then arbitrarily select a uniform distribution centered around the MLE as a prior distribution to Bayesian methods (e.g.
-\cite{smith22,rivera22}).
-This approach is related to Empirical Bayes, where the prior distribution is selected using information in the available data.
-We propose something similar in our supplement material, but our proposition does not require additional runs of pMCMC which can be computationally expensive.
-%   One drawback of our proposed Empirical Bayes approach is that when the Monte Carlo variance in likelihood evaluation is high, it is possible that the posterior mass function has the majority of its mass assigned to only a few parameter sets.
-% \end{itemize}
-% 
-% A full investigation / discussion on parameter uncertainty is outside of the scope of this article.
 
@@
 These forecasts were generally inconsistent with the absence of cholera cases from 2019-2022, and do not account for the particular relevance of recent data for forecasting cholera dynamics.
-In their supplement material, however, they provide a table (S7) in which they show that if they condition the simulations on most recent data (0 cases from 2019-2020), their forecasts predict cholera elimination.
+In their supplement material, however, they provide a table (S7) in which they show that if they conditionally select only the simulations that were consistent with recent data (0 cases from 2019-2020), their forecasts predict cholera elimination.
 This second effort that was only presented in the supplement material is more consistent with what we suggest is correct practice: forecasts of a system should start using latent states that are consistent with available data.

### 2024-02-29 edits.tex
Minor edits for clarity

@@
 Lee et al.~\cite{lee20} chose to obtain forecasts from all of their models by simulating forward from initial conditions, rather than conditioning forecasts based on the available data.
-This decision is possibly as a result of using a deterministic model, as forecasts from different models may only be considered comparable if they are obtained in the same way, which must be by simulating from initial conditions because Model~2 is deterministic.}
+This decision is possibly as a result of using a deterministic model, as forecasts from different models may only be considered comparable if they are obtained in the same way, which is most easily done be by simulating from initial conditions because Model~2 is deterministic.}
 
 \newcommand\editForecastTwo{In contrast, for non-deterministic Models~1 and 3, we obtain forecasts by simulating future values using latent states that are harmonious with the most recent data.
-This is done by drawing latent states at the last observation time ($t_N$) from the filtering distribution $f_{\bm{X}_N|\bm{Y}_{1:N}}(\bm{x}_{N} | \bm{y}^*_{1:N} ; \hat\paramVec)$.
+This is done by simulating forward from latent states drawn at the last observation time ($t_N$) from the filtering distrubion $f_{\bm{X}_N|\bm{Y}_{1:N}}(\bm{x}_{N} | \bm{y}^*_{1:N} ; \hat\paramVec)$.
 The decision to obtain model forecasts from initial conditions partially explains the unsuccessful forecasts of Lee et al.~\cite{lee20}.
-Table~S7 in their supplement material, which contains results that were not discussed in their main article, shows that the subset of their simulations which were consistent with observing zero cases from 2019-2020 were also more consistent with the disappearance of cholera from Haiti from 2019-2022.
+Table~S7 in their supplement material, which contains results that were not discussed in their main article, shows that the subset of their simulations with zero cholera cases from 2019-2020 also correspond with its disappearance until 2022.
 These results support our argument that forecasts should be made by ensuring the starting point for the forecast is consistent with available data.}
@@
 In general, a spatially explicit model may not have well-defined individual log-likelihoods, and, in this case, comparisons to benchmarks must be made at the level of the joint model.}
+
+\newcommand\editConfound{The statistical evidence of a trend in transmission rate in this model could be explained by any trending variable (such as hygiene improvements, or changes in population behavior), resulting in confounding from collinear covariates.
+Alternatively, the trend could be attributed to a decreasing reporting rate rather than decreasing transmission rate.
+This could be formally tested by comparing models with either trend specification.
+We did not do this because evidence suggests that reporting rate was maintained or increased (Figure~1 of \cite{rebaudet21}).
+We instead argue that a decreasing transmission rate is a plausible way to explain the decrease in cases over time, as there is alternative evidence that supports this model \cite{rebaudet19CATI,rebaudet21,michel19}.
+It is not practical to test all remotely plausible model variations, yet a strongly supported conclusion should avoid ruling out untested hypotheses.
+The robust statistical conclusion for our analysis is that a model which allows for change fits better than one which does not, and a trend in transmission is a plausible way to do this.}

### 2024-02-29 ms.Rnw
Minor edits for clarity

@@
     utils::capture.output(sessionInfo()),
-    file = "sessionInfo_tmp.txt",
+    file = "sessionInfo.txt",
     sep="\n"
@@
 Objective measures are used to \old{benchmark} \new{evaluate} model fit; when \old{these}\new{objective measures of the goodness-of-fit of a model} are combined with reproducibility, a framework emerges for continual improvement \new{to model based inference} when revisiting the data\old{ and models}.
-\new{This framework promotes scientific discovery and supports model-based policy recommendations by simplifying the process for researchers to build off of previous results.}
+\new{This framework promotes scientific discovery and supports model-driven policy recommendations by simplifying the process of understanding and improving previous results.}
 Our data analysis workflow is supported by recent advances in algorithms, software and hardware, which facilitate statistical fitting of nonlinear stochastic dynamic models to observed incidence data.
@@
 Model~4 has an agent-based construction, featuring considerable mechanistic detail but limited ability to calibrate these details to data.
-\new{These modeling strategies were selected by \cite{lee20} to represent the range of approaches used in the research community}
+\new{These modeling strategies were selected by Lee et a.~\cite{lee20} to represent the range of approaches used in the research community}
 We focus on Models~1--3, as the strengths and weaknesses of the agent-based modeling approach \cite{tracy18} are outside the scope of this article.
-\new{In addition, agent-based models were less widely used, the agent based model of \cite{lee20} being the only model of this class that was found in our literature review.}
+\new{In addition, agent-based models were less widely used, the agent based model in \cite{lee20} being the only model of this class that was found in our literature review.}
 \new{The data that were used in \cite{lee20}, and that we reanalyze, are displayed in Fig.~\ref{fig:CholeraData}.}
@@
 While OCV is accepted as a safe and effective tool for controlling the spread of cholera, the global stockpile of OCV doses remains limited \cite{pezzoli20}.
-Advances in OCV technology and vaccine availability, however, raised the possibility of planning a national vaccination program \cite{lee20}.
-\new{The possibility of controlling the Haiti cholera outbreak via OCV was considered by various research groups \cite{ivers17,andrews11,walton11,matias17,chao11,fung13,date11,leung22,azman15}}.
+Advances in OCV technology and vaccine availability, however, raised the possibility of planning a national vaccination program.
+\new{The possibility of controlling the Haiti cholera outbreak via OCV was considered by various research groups \cite{lee20,ivers17,andrews11,walton11,matias17,chao11,fung13,date11,leung22,azman15}}.
 In the \new{Lee et al.~\cite{lee20}} study, certain data were shared between the groups, including demography and vaccination history; vaccine efficacy was also fixed at a shared value between groups.
@@
 Parameters that were added during our re-analysis and were not considered by Lee et al. are indicated with $^*$.
-\new{Confidence intervals for model parameters are given in supplement~S8.}
+\new{Confidence intervals for model parameters are given in supplement~S8.} %TODO: nameref this
 Translations back into the notation of \cite{lee20} are given in \nameref{S1_Table}.
@@
 The latent state vector $\bm{X}(t) = \big(S_{\vaccCounter}(t),  E_{\vaccCounter}(t), I_{\vaccCounter}(t), A_{\vaccCounter}(t), R_{\vaccCounter}(t), \vaccCounter \in 0:\vaccClass\big)$ describes susceptible, latent (exposed), infected (and symptomatic), asymptomatic, and recovered individuals in vaccine cohort $\vaccCounter$ at time $t$.
-Here, $\vaccCounter=0$ corresponds to unvaccinated individuals, and $\vaccCounter \in \seq{1}{\vaccClass}$ describes hypothetical vaccination programs\new{, where each program $\vaccCounter$ indexes differences in both the number of doses administered (one versus two doses per individual) and the round of vaccine administration.}
+Here, $\vaccCounter=0$ corresponds to unvaccinated individuals, and $\vaccCounter \in \seq{1}{\vaccClass}$ describes hypothetical vaccination programs.
+\new{Each program $\vaccCounter$ indexes differences in both the number of doses administered (one versus two doses per individual) and the round of vaccine administration, seperating individuals into compartments with distinct dynamics based on vaccination status.}
 The force of infection is
@@
 where $\mathrm{Pop}_u$ is the mean population for department $u$,
-$D_{u\varv}$ is a distance measure estimating average road distance between randomly chosen members of each population, and $v_{\mathrm{rate}}= 10^{-12} \,\mbox{km$^2$yr$^{-1}$}$ \old{was treated as a fixed constant.}\new{was fixed at the value used by \cite{lee20}.}
+$D_{u\varv}$ is a distance measure estimating average road distance between randomly chosen members of each population, and $v_{\mathrm{rate}}= 10^{-12} \,\mbox{km$^2$yr$^{-1}$}$ \old{was treated as a fixed constant.}\new{was fixed at the value used in \cite{lee20}.}
 In Eq.~\myeqref{model2:water_transport}, $T^W_{u\varv}$ is a measure of river flow between departments.
@@
 \begin{flushleft}
-Values in parentheses are corresponding values using Lee et al.~\cite{lee20} parameter estimates.
-\textsuperscript{1}The reported likelihood is an upper bound of the likelihood of the Lee et al.~\cite{lee20} model.
-\textsuperscript{2}Lee et al.~\cite{lee20} fit Model~3 to a subset of the data (March 2014 onward, excluding data from Ouest in 2015-2016).
+Values in parentheses are corresponding values \new{obtained} using \new{the models of \cite{lee20}} \old{Lee et al.~[4] parameter estimates}.
+\textsuperscript{1}The reported likelihood is an upper bound of the likelihood of the \old{Lee et al.~[4]} model \new{in \cite{lee20}}.
+\old{Lee et al.~[4] fit} \textsuperscript{2}\new{In \cite{lee20},} Model~3 \new{was fit} to a subset of the data (March 2014 onward, excluding data from Ouest in 2015-2016).
 On this subset, their model has a likelihood of $\Sexpr{myround(lee3_results$sub_ll_eval, 1)}$.
 On this same subset, our model has a likelihood of $\Sexpr{myround(mod3_eval_res$subset_ll['est'], 1)}$.
-Details of estimating the likelihood of the Lee et al.~\cite{lee20} models is provided in \nameref{S_lee20}
-See Sec.~S6 of the supplement material for more details on estimating the likelihood of the Lee et al.~\cite{lee20} models.
+Details of estimating the likelihood of the \old{Lee et al.~[4]} models \new{used in \cite{lee20} are} \old{is} provided in \nameref{S_lee20}
+\old{See Sec.~S6 of the supplement material for more details on estimating the likelihood of the Lee et al.~[4] models.}
 \end{flushleft}
@@
 With a sufficient number of iterations, the resulting parameter values converge to a region of the parameter space that maximizes the model likelihood.
-
-% One of the most frequently used approaches to calibrate the parameters of a statistical model is maximum likelihood estimation.
-% In particular, a closed form expression of the joint model density---described in Eq.~\myeqref{eq:jointLik}---is not readily available, making likelihood based inference a challenging task.
-% In the case of any POMP model, the likelihood function---as described in Eq.~\myeqref{eq:likedef}---is written as $\mathcal{L}(\paramVec) = \int f_{\bm{X}_{0:N}, \bm{Y}_{1:N}}\big(\bm{x}_{0:N}, \bm{y}_{1:N};\paramVec\big) d\bm{x}_{0:N}$.
-% The high-dimensional integral in the likelihood, and the potentially non-linear nature of the process model, make it impossible to obtain a closed form expression of the likelihood function of POMP models except in the most simple cases.
-% Model~1 was developed with the intent of providing a description of cholera dynamics that is scientifically plausible rather than on that is simply statistically convenient.
-% Because of this, a closed form expression of the likelihood of Model~1 is not readily available.
-
-% In order to retain the ability to propose models that are scientifically meaningful rather than only those that are simply statistically convenient, we restrict ourselves to parameter estimation techniques that have the plug-and-play property, which is that the fitting procedure only requires the ability to simulate the latent process instead of evaluating transition densities \cite{breto09,he10};
-% in the context of the notation and definitions employed in this article, this means that we only require the ability to simulate from $f_{\bm{X}_0}\big(\bm{x}_0;\paramVec\big)$ and $f_{\bm{X}_n|\bm{X}_{n-1}}\big(\bm{x}_{n}|\bm{x}_{n-1}; \paramVec\big)$ rather than needing to evaluate these densities.
-% Plug-and-play algorithms include Bayesian approaches like ABC and PMCMC \cite{toni09,andrieu10}, but here we use algorithms that enable maximum likelihood estimation.
-% To our knowledge, the only plug-and-play methods that can be used to maximize the likelihood for POMP models of this complexity are iterated filtering algorithms \cite{ionides15}, which modify the well-known \emph{particle filter} \cite{arulampalam02}.
-% The particle filter is often used to obtain values of latent states, but can also be used to obtain likelihood estimates of a POMP model.
-% Iterated filtering algorithms such as IF2 \cite{ionides15} extend the particle filter by performing a random walk for each parameter and particle;
-% these perturbations are carried out iteratively over multiple filtering operations, using the collection of parameters from the previous filtering pass as the parameter initialization for the next iteration, and decreasing the random walk variance at each step.
 
@@
 The robust statistical conclusion is that a model which allows for change fits better than one which does not---we argue that a decreasing transmission rate is a plausible way to explain this, but the incidence data themselves do not provide enough information to pin down the mechanism.}
-\new{The statistical evidence of a trend in transmission rate in this model could be explained by any trending variable (such as hygiene improvements, or changes in population behavior), resulting in confounding from collinear covariates.
-Alternatively, the trend could be attributed to a decreasing reporting rate rather than decreasing transmission rate.
-This could be formally tested by comparing models with either trend specification.
-We did not do this because evidence suggests that reporting rate was maintained or increased (Figure~1 of \cite{rebaudet21}).
-We instead argue that a decreasing transmission rate is a plausible way to explain the decrease in cases over time, as there is alternative evidence that supports this model \cite{rebaudet19CATI,rebaudet21,michel19}.
-It is not practical to test all remotely plausible model variations, yet a strongly supported conclusion should avoid ruling out untested hypotheses.
-The robust statistical conclusion for our analysis is that a model which allows for change fits better than one which does not, and a trend in transmission is a plausible way to do this.}
-
-% The statistical evidence of a trend in transmission rate in this model could be explained by any trending variable (such as hygiene improvements, or changes in population behavior), resulting in confounding from collinear covariates.
-% \old{Alternatively, the trend could be attributed to a decreasing reporting rate rather than decreasing transmission rate, resulting in confounded mechanisms.}
-% \new{Because we only have access to incidence data, the decrease in observed cases over the course of the outbreak may alternatively be attributed to a trend in other model parameters, such as a decrease in the reporting rate over time.
-% This results in confounded mechanisms, as one may not formally test all possible ways to include a trending parameter.
-In such a case, t}\old{T}he robust statistical conclusion is that a model which allows for change fits better than one which does not.\old{---we argue that a decreasing transmission rate is a plausible way to explain this, but} \new{Because} the incidence data themselves \old{do} \new{may} not provide enough information to pin down the mechanism\new{, a strongly supported conclusion should avoid ruling out untested hypotheses.}
-% \new{In this case study, we did not explicitly test if the incidence data favor a trend in transmission over a trend in reporting rate.
-We argue that a decreasing transmission rate is a plausible way to explain the decrease in cases over time, as there is alternative evidence that supports this model \cite{rebaudet19CATI,rebaudet21,michel19}.
-Additionally, there is little scientific evidence to support a decrease in reporting rate (for example, see Figure~1 of \cite{rebaudet21}).}
+
+\new{\editConfound}
 
@@
 as parameters that maximize the likelihood of the decoupled model almost certainly do not maximize the likelihood of the fully coupled model.
-These two concerns may explain the unrealistic forecasts and low likelihood of Model~3 of Lee et al.~\cite{lee20} (see Table~\ref{tab:likes}).
+These two concerns may explain the unrealistic forecasts and low likelihood of Model~3 \new{in \cite{lee20}} \old{of Lee et al.~[4]} (see Table~\ref{tab:likes}).
 
@@
 Probability of cholera elimination, defined as having zero cholera infectious for at least 52 consecutive weeks, based on 10 year simulations from calibrated versions of Models~1 and 3.
-Compare to Fig.~3A of Lee et al.~\cite{lee20}.}
+Compare to Fig.~3A of \cite{lee20}.}
 \end{figure}
@@
 \new{However, the development of credible models---which are supported by data and can lend quantitative insights into a dynamic system---remains a challenging task.
-In this article, we demonstrate opportunites available for raising the current standards of statistical inference for mathematical models of biological systems.
-
-We present methodology consistent with existing guidelines but exceeding standard practice.  
+In this article, we demonstrate opportunities available for raising the current standards of statistical inference for mathematical models of biological systems.}
+
+\new{We present methodology consistent with existing guidelines but exceeding standard practice.
 In particular, we show the value of comparing the likelihood of fitted mechanistic models versus non-mechanistic benchmarks.

### 2024-02-29 H1profile.R
Added computation times for future reference

@@ -13,7 +13,7 @@
 # WARNING:
 #
 #   This script was ran using a high-performance computing cluster. In total,
-#   the RUN_LEVEL == 3 version of this script took TODO compute hours. Partial
+#   the RUN_LEVEL == 3 version of this script took 3631 compute hours. Partial
 #   results can be obtained by lowering the required computational effort
 #   (params: NP, NMIF, nprof, NP_EVAL, NREPS_EVAL), which can be done by
 #   reducing the RUN_LEVEL to 2 or 1, or alternatively by manually changing

### 2024-02-29 H3profile.R

@@ -13,7 +13,7 @@
 # WARNING:
 #
 #   This script was ran using a high-performance computing cluster. In total,
-#   the RUN_LEVEL == 3 version of this script took TODO compute hours. Partial
+#   the RUN_LEVEL == 3 version of this script took 14761 compute hours. Partial
 #   results can be obtained by lowering the required computational effort
 #   (params: NP, NBPF, nprof, NP_EVAL, NREPS_EVAL), which can be done by
 #   reducing the RUN_LEVEL to 2 or 1, or alternatively by manually changing

### 2024-02-29 H3profile-r.R
Removed unused file

@@ -1,399 +1,0 @@
-# Fitting profiles for Haiti 3.
-#
-# ********
-# NOTE:
-#
-#   If you would like to run this code but have never used the `batchtools` R
-#   package, then we strongly recommend completing both Examples 1 and 2 from
-#   the batchtools vignette, by first installing the `batchtools` package and
-#   running: `vignette('batchtools')`, or visiting:
-#
-#         https://mllg.github.io/batchtools/articles/batchtools.html
-#
-# WARNING:
-#
-#   This script was ran using a high-performance computing cluster. In total,
-#   the RUN_LEVEL == 3 version of this script took TODO compute hours. Partial
-#   results can be obtained by lowering the required computational effort
-#   (params: NP, NBPF, nprof, NP_EVAL, NREPS_EVAL), which can be done by
-#   reducing the RUN_LEVEL to 2 or 1, or alternatively by manually changing
-#   these values.
-# ********
-#
-# The fit_haiti3 function is a very simple helper that will take as input
-# some starting parameters, and hyperparameters for the IBPF algorithm, and
-# will return parameter estimates along with corresponding estimates of their
-# log-likelihoods.
-#
-# In this script we will use this function to do a profile likelihood search
-# of all model parameters using the batchtools package. To do this, we will
-# create a grid of parameter values, and submit jobs where each job will use
-# a row from the grid as a starting point to fit the model.
-#
-# The number of rows in the grid (which is the `final_pars` object that is
-# created using a for-loop over the profile parameters) correspond to the total
-# number of jobs that will be submitted using `batchtools`. Each job takes
-# roughly the same amount of time to compute, and hence the run-time complexity
-# is linear with respect to the total number of jobs. The total number of jobs
-# can be controlled by either (1) changing the number of variables to profile
[…]

### 2024-02-29 ms.Rnw
Edits related to name-ref of supplement files

@@
 Parameters that were added during our re-analysis and were not considered by Lee et al. are indicated with $^*$.
-\new{Confidence intervals for model parameters are given in supplement~S8.} %TODO: nameref this
+\new{Confidence intervals for model parameters are given in the supplement (\nameref{S_CI}).}
 Translations back into the notation of \cite{lee20} are given in \nameref{S1_Table}.
@@
 Reported cases are described using a log-normal distribution, with the log-scale mean equal to the reporting rate $\reportRate$ times the number of newly infected individuals.
-See the section describing how Model~2 is fit to data and Sec.~S3.2 of the supplement for more details.
+\old{See the section describing how Model~2 is fit to data and Sec.~S3.2 of the supplement for more details.}
+\new{See the supplement material on model initializations for more details (\nameref{S_init})}
 
@@
 On this same subset, our model has a likelihood of $\Sexpr{myround(mod3_eval_res$subset_ll['est'], 1)}$.
-Details of estimating the likelihood of the \old{Lee et al.~[4]} models \new{used in \cite{lee20} are} \old{is} provided in \nameref{S_lee20}
+Details of estimating the likelihood of the \old{Lee et al.~[4]} models \new{used in \cite{lee20} are} \old{is} provided in \new{the supplement} (\nameref{S_lee20}).
 \old{See Sec.~S6 of the supplement material for more details on estimating the likelihood of the Lee et al.~[4] models.}
@@
 
+\paragraph*{S7 Text}
+\label{S_CI}
+{\bf Confidence Intervals} Tables and figures describing 95\% confidence intervals for estimated parameters.
+
 \section*{Acknowledgments}

### 2024-02-29 si/ReplicateLee20.Rnw
Fixing section numbers and other minor details in supplement material

@@
 
-\setcounter{tocdepth}{1}
+\usepackage{titlesec}
+\usepackage{chngcntr}
+\usepackage{etoolbox}
+
+% Remove the numbering from sections
+\titleformat{\section}{\normalfont\Large\bfseries}{}{0pt}{}
+
+% Reset the subsection counter at the start of each section
+\counterwithin*{subsection}{section}
+
+% Change the numbering for subsections to have an S prefix and remove chapter numbering
+\renewcommand{\thesubsection}{S\arabic{subsection}}
+
+% \setcounter{tocdepth}{1}
 
@@
 Another potential explanation for the discrepancy is that the parameters that we used are only approximately the same as those used by \citet{lee20sup}.
-For example, the parameters $\transmission$, $\Wbeta{}$ (See Table~\ref{tab:translate}) had reported values of $9.9 \times 10^{-7}$ and $4.03 \times 10^{-2}$, respectively (Table~S13 of \cite{lee20sup}), but were actually fit to data and therefore likely these values have been rounded.
+For example, the parameters $\transmission$, $\Wbeta{}$ had reported values of $9.9 \times 10^{-7}$ and $4.03 \times 10^{-2}$, respectively (Table~S13 of \cite{lee20sup}), but were actually fit to data and therefore likely these values have been rounded.
 Additionally, our implementation of Model~2 used a time scale of years and many of the parameters were reported on a weekly scale, so small differences may result due to unit conversions.
@@
 Additionally, the script that was said to been used to obtain the results reported by \cite{lee20} appears to use a different measurement model than what was described in the supplemental material, again making it difficult to fully replicate the result of \cite{lee20} without being able to easily run the provided source code.
-In this case, we chose to use measurement model described by Eq.~\ref{model2:measEpi} for both phases of the epidemic, as this seemed to visually match the results of \cite{lee20} most closely.
+In this case, we chose to use measurement model that considers only symptomatic individuals for both phases of the epidemic, as this seemed to visually match the results of \cite{lee20} most closely.
 

### 2024-02-29 edits.tex
Adding name-ref to edits to make the correct link to supplement file

@@
 Stochastic phenomena are not only arguably present in biological systems, but their inclusion in a model also allows observed data variations to be attributed to inherent uncertainty rather than to distorted parameter values.
-Models~1 and 3 suggest the presence of extra-demographic stochsticity \cite{he10,stocks20,li23}, as evidenced by the confidence intervals provided in Supplement~S8.
-}
+Models~1 and 3 suggest the presence of extra-demographic stochsticity \cite{he10,stocks20,li23}, as evidenced by the confidence intervals (\nameref{S_CI}).} % TODO: The nameref won't work in the response.
 

### 2024-03-01 r1/response.tex
Adding name-ref to edits to make the correct link to supplement file
Joint edits from Ed / Jesse meeting

@@
 \usepackage{amsmath,amsthm}
+\usepackage{nameref,hyperref}
 \usepackage[normalem]{ulem}% to use \sout in feedback commands
+\usepackage{comment}
 
+\excludecomment{hidden}
 % \usepackage{natbib}
@@
 \newcommand\transmissionTrend{\zeta}
+\newcommand\transmission{\beta}
 \input{../edits.tex}
@@
 
-\TODO{Edit this letter}
 Thank you for arranging the review of our manuscript, and for the invitation to submit a revision.
@@
 Our revised manuscript has various edits for clarity, following the referee suggestions, as well as some small corrections in the text.
-The numerical results are unchanged.
 
-To make space for some additional sentences, we cut a few sentences that seemed least essential. We also placed some of the new material in the supplement.
+One of the reviewers recommended the inclusion of confidence intervals for model parameters, which was not originally included because the paper that we re-analyze did not include them. 
+However, we agree that the inclusion of confidence intervals is beneficial for readers, so we have done so in the revised manuscript.
+The additional computational effort required to obtain confidence intervals resulted in slightly improved parameter estimates, though qualitative conclusions drawn from the fitted models remain unchanged.
+
+The other main changes revisions to both the introduction and discussion, as recommended by both reviewers.
 
@@
 We evaluate previous methods used for fitting these models, and we demonstrate modified data analysis strategies leading to improved statistical fit.
-Speficially, we present appraoches for diagnosing model misspecification and the consequent development of improved models.
+Specifically, we present appraoches for diagnosing model misspecification and the consequent development of improved models.
 We futher demonstrate the utility of recent advances in likelihood maximization for high-dimensional nonlinear dynamic models, enabling likelihood-based inference for high-dimensional models for spatiotemporal incidence data.}
@@
 We agree, however, that adding to model complexity has costs, including loss of identifiability.
-In our revised manuscript, we address identifiability via profile likelihood methods.
+In our revised manuscript, we address identifiability via profile likelihood methods (\nameref{S_CI}).
 % AIC and other formal statistical methods can assess whether the additional complexity is justified by a sufficiently large improvement of statistical fit.
@@
 We see some evidence of this in our analysis.
-Aside from not being stochastic, the mechanistic structure of Model~2 is far more complex than the other two models considered in our study.
-Model~2 also demonstrates the most clear cases of spurious scientific findings, as some estimated parameters favor regions of the parameter space that we would not expect scientifically.
+Aside from not being stochastic, the mechanistic structure of Model~2 is more complex than the other two models considered in our study.
+Model~2 also demonstrates the most clear cases of spurious scientific findings, as some estimated parameters favor regions of the parameter space that we would not expect scientifically---such as permanent immunity from infection once recovered from an infection ($\muRS^{-1}$), and zero human-to-human transmission ($\transmission$).
 This may partially explain the choice of Lee et al~\cite{lee20} to consider a scientifically unjustifiable jump in latent state values from their epidemic to endemic periods: without this additional flexibility, Model~2 cannot adequately describe cholera dynamics in Haiti while keeping parameter values in scientifically plausible ranges.
@@
 
-\article{Despite their limitations, deterministic models can offer valuable insights into dynamic systems.
-As with all models, researchers should consider and acknoweledge these limitations when applying deterministic models and interpreting their outcomes.}
+\article{Despite their limitations, deterministic models can offer valuable insights into dynamic systems \cite{may04}.
+In \cite{lee20}, the forecasts from the deterministic Model~2 were qualitatively more consistent with the observed dissapearance of cholera than the stochastic models.}
 
@@
 Thus, particle filter methods facilitate the use of various likelihood-based methods for diagnostics and model selection.
-We demonstrate this in Figs S4, S5, S6, S7. \TODO{Check all figure numbers, especially to the supplement.}
+We demonstrate this in the supplement (\nameref{S_CI}, \nameref{S_mod3cal}). 
+% Figs S4, S5, S6, S7. \TODO{Check all figure numbers, especially to the supplement.}
 
@@
 MCAP methods have been used in various recent investigations using particle filters \cite{subramanian21,ali22,pons-salort18}.
+Our revised manuscript includes MCAP confidence intervals for both Model~1 and Model~3 (\nameref{S_CI}).
 
@@
 
-\article{\editDiscussionSix}
+% \article{\editDiscussionSix}
 
@@
 
+\paragraph*{S5 Text}
+\label{S_mod3cal}
+{\bf Calibrating Model 3.} Additional details on the procedures for fitting and diagnosing Model~3.
+
+% \begin{hidden}
+\paragraph*{S7 Text}
+\label{S_CI}
+{\bf Confidence Intervals} Tables and figures describing 95\% confidence intervals for estimated parameters.
+% \end{hidden}
+
 \end{document}

### 2024-03-01 edits.tex
Joint edits from Ed / Jesse meeting

@@
 If a model falls far behind simple benchmarks, there is likely to be substantial limitation in the data analysis that should be identified and remedied.
-In our case study, the recalibrated version of Model~1 outperforms its benchmark, enabling evaluation of causal implications.
+In our case study, the recalibrated version of Model~1 outperforms its benchmark, so we proceeded to examine causal implications.
 When doing so, we find that the fitted model provides a causal description of the dynamic system that is consistent with known features of the system, such as the importance of rainfall as a driver of cholera infection.
@@
 Mathematical models provide simplified representations of complex systems, with the simplicity serving both to facilitate scientific understanding and to enable statistical inference on unknown parameters.
-In our case study, employing a deterministic dynamic model was found to be an over-simplification by comparing model fit with benchmarks and stochastic models.
-Including stochastic metapopulation dynamics permitted investigation of spatiotemporal data, the main cost being additional complexity of numerical computations.}
+In our case study, employing deterministic dynamics in Model~2 was found to be an over-simplification by comparing model fit with benchmarks.
+% Including stochastic metapopulation dynamics permitted investigation of spatiotemporal data, the main cost being additional complexity of numerical computations.
+Model~3 is distinct in that it is both stochastic and has a meta-population structure, making it challenging to draw likelihood-based inferences.
+In this paper, we demonstrated how this model class could be calibrated to incidence data using the innovative IBPF algorithm.
+One of only a few examples of fitting a nonlinear non-Gaussian meta-population model via maximum likelihood \cite{li23,ionides22}, this case study exemplifies the algorithm's potential benefits and provides an example for future researchers on a possible approach to fitting a high-dimensional non-linear model.
+}
 %In this simplification process, careful consideration must be given to which facets of the system are incorporated into the model and which are omitted, as these choices can shape the conclusions drawn from the model.
@@
 
-\newcommand\editDiscussionThree{Along with enabling comparisons to statistical benchmarks, likelihood-based methods aid in determining an appropriate level of model complexity.
+\newcommand\editDiscussionThree{Likelihood-based methods aid in determining an appropriate level of model complexity.
 Models fit to the same data can be compared using a criteria such as AIC.
@@
 \newcommand\editDiscussionFour{Unmodeled features of a dynamic system can lead to spurious or misleading parameter estimates if the features substantially impact observed data. 
-This is because the model must compensate for these unaccounted features using its inherent flexibility.
-In deterministic models, the choice of parameters is the sole source of such flexibility.
-For our case study, some of the parameter estimates for the deterministic Model~2 are implausible [WHICH ONES?] and this may be explained by compensation for model misspecification.
+% A fitted model must compensate for these unaccounted features using its inherent flexibility. 
+In deterministic models, features that cannot be explained by measurement error must be accounted for by the choice of parameters. 
+% the choice of parameters is the primary source of flexibility to compensate for these unaccounted features.
+For our case study, some of the parameter estimates for the deterministic Model~2 are implausible, such as the infinite immunity discussed above, and this may be explained by compensation for model misspecification.
 Incorporating demographic and environmental stochasticity into models can mitigate the impact of unmodeled features.
@@
 
-\newcommand\editDiscussionFive{The three models in this case study consider a range of possible model complexity.
-Of these models, Model~3 is distinct in that it is both stochastic and has a meta-population structure, making it challenging to draw likelihood-based inferences.
+% \newcommand\editDiscussionFive{Model~3 is distinct in that it is both stochastic and has a meta-population structure, making it challenging to draw likelihood-based inferences.
 In this paper, we demonstrated how this model class could be calibrated to incidence data using the innovative IBPF algorithm.
-One of only a few examples of fitting a model with this level of complexity via maximum likelihood, this case study exemplifies the algorithm's potential benefits and provides an example for future researchers on a possible approach to fitting a high-dimensional non-linear model.
-THIS PARAGRAPH SEEMS TO OVERLAP OTHERS. MAYBE FIT WHATEVER IS NEEDED INTO OTHER PARAGRAPHS AND DELETE?}
+One of only a few examples of fitting a model with this level of complexity via maximum likelihood, this case study exemplifies the algorithm's potential benefits and provides an example for future researchers on a possible approach to fitting a high-dimensional non-linear model.}
 
-\newcommand\editDiscussionSix{If forecasts are an important component of a modeling task, the forecasts should be consistent with the available data, particularly at the most recently available time points.
+\newcommand\editDiscussionFive{If forecasts are an important component of a modeling task, the forecasts should be consistent with the available data, particularly at the most recently available time points.
 In our case study, we do this by simulating forward from the filtering distribution, as this proceedure conditions latent variables on the available data.
-This type of forecasting, however, is not possible using a deterministic model, where only initial conditions and parameter values can influence forecasts.
-Deterministic simulations may also result in over-confidence in model forecasts, as they can only account for uncertainty due to the parameter estimation proceedure \cite{king15}.
-Despite their limitations, deterministic models can offer valuable insights into dynamic systems.
-As with all models, researchers should consider and acknoweledge these limitations when applying deterministic models and interpreting their outcomes.
-I WONDER WHETHER PEOPLE SERIOUSLY FORECASTING WITH DETERMINISTIC MODELS HAVE OTHER TRICKS.
-FOR EXAMPLE, WITH WEATHER FORECASTING MODELS, THEY ARE OFTEN WRITTEN AS DETERMINISTIC BUT PEOPLE DO SOMETHING ENKF-LIKE ANYHOW.
-JEFF SHAMAN HAS ALSO DONE THINGS LIKE THIS WITH EPI MODELS.
-ANOTHER OPTION COULD BE TO FIT IN A MOVING WINDOW.
-ANOTHER THING: THE PENULTIMATE SENTENCE MAKES A FAIRLY STRONG CLAIM THAT IS NOT SUPPORTED BY EVIDENCE HERE.
-PERHAPS WE CAN REMIND READERS THAT DETERMINISTIC MODELS REMAIN WIDELY USED, BUT IT WOULD BE NICE TO CLARIFY WHAT THEY ARE GOOD FOR SINCE WE SPEND SO MUCH TIME ON THEIR LIMITATIONS.}
+This type of forecasting, however, is not directly available using a deterministic model, where future dynamics are fully determined by initial conditions and parameter values.
+This can result in over-confident model forecasts \cite{king15}.
+Despite their limitations, deterministic models can offer valuable insights into dynamic systems \cite{may04}.
+In \cite{lee20}, the forecasts from the deterministic Model~2 were qualitatively more consistent with the observed dissapearance of cholera than the stochastic models.
+In our case study, we found improvements to Models~1 and 3 that resulted in improved forecasts for these models.}
 

### 2024-03-01 ms.Rnw
Joint edits from Ed / Jesse meeting

@@
   &
-  & $\Wshed = $ \Sexpr{signif(h2_fit$h2_params["Mu"],3)}
+  & $\Wshed = $ \Sexpr{signif(h2_fit$h2_params["Mu"]/52,3)}
   & \eqref{model2:to_W}
@@
   & $\reportRate=\Sexpr{myround(p1["rho"], 3)}$
-  & (S15)
+  & (\nameref{S_meas})
   & ${\reportRate=\Sexpr{myround(h2_fit$h2_params["Rho"],2)}}\fixed$
-  & (S16)
+  & (\nameref{S_meas})
   & $\reportRate=\Sexpr{myround(p3s["epsilon"],2)}$
-  & (S19)
+  & (\nameref{S_meas})
 \\
@@
   & \raggedright $\obsOverdispersion = (\Sexpr{myround(p1["tau_epi"],2)}, \Sexpr{round(p1["tau_end"],2)})$
-  & \multirow{2}{*}{(S15)}
+  & \multirow{2}{*}{(\nameref{S_meas})}
   & \multirow{2}{*}{$\obsOverdispersion = \Sexpr{myround(h2_fit$h2_params['v'], 3)}$}
-  & \multirow{2}{*}{(S16)}
+  & \multirow{2}{*}{(\nameref{S_meas})}
   & \multirow{2}{*}{$\obsOverdispersion = \Sexpr{myround(p3s["k"],2)}$}
-  & \multirow{2}{*}{(S19)}
+  & \multirow{2}{*}{(\nameref{S_meas})}
 \\
@@
         \Sexpr{round(prod(p3[c('H3', 'Iinit3')]))}, \Sexpr{round(prod(p3[c('H4', 'Iinit4')]))})^*$}
-  &  \multirow{2}{*}{(S20)}
+  &  \multirow{2}{*}{(\nameref{S_init})}
   \\
@@
 \label{model1:beta}
-\log\transmission(t) = \sum_{j=1}^6 \transmission_j s_j(t).
+\transmission(t) = \bar{\transmission} \exp\Big(\sum_{j=1}^6 \transmission_j s_j(t)\Big),
 \end{equation}
+\new{where $\bar{\transmission} = 1 (\mathrm{wk})^{-1}$ is a dimensionality constant.}
 The process noise $d\Gamma(t)/dt$ is multiplicative Gamma-distributed white noise, with infinitesimal variance parameter $\sigmaProc^2$.
@@
 \label{model1:betat}
-\log\transmission(t) = \sum_{j=1}^6 \transmission_s s_j(t) + \transmissionTrend\bar{t},
+\transmission(t) = \bar{\transmission} \exp\Big(\sum_{j=1}^6 \transmission_s s_j(t) + \transmissionTrend\bar{t}\Big),
 \end{equation}
@@
 
-\new{\editDiscussionSix}
+% \new{\editDiscussionSix}
 

### 2024-03-01 si/confidenceIntervals.Rnw
Fixing section numbers and other minor details in supplement material
Joint edits from Ed / Jesse meeting

@@
   if (p == 'Mu') {
-    my_lab <- TeX('$\\mu_W')
+    my_lab <- TeX('$52\\times\\mu_W')
   } else if (p == 'Beta') {
@@
 
- Human to water shedding & $\Wshed$ & $\Sexpr{myround(h2_mles['Mu'], 1)}$
-   &
-  $(\Sexpr{myround(h2_plots_lb['Mu'], 1)}, \Sexpr{myround(h2_plots_ub['Mu'], 1)})$ 
+ Human to water shedding & $\Wshed$ & $\Sexpr{myround(h2_mles['Mu']/52, 1)}$
+   &
+  $(\Sexpr{myround(h2_plots_lb['Mu']/52, 1)}, \Sexpr{myround(h2_plots_ub['Mu']/52, 1)})$ 
 \\

### 2024-03-02 r1/response.tex
Fixed units for parameters, other minor typos, and response to editors

@@
 
-The other main changes revisions to both the introduction and discussion, as recommended by both reviewers.
+The other main change in the revised manuscript is edits to both the introduction and discussion, as recommended by both reviewers.
 
@@
 Thank you for this feedback.
-We have made substantial changes to the abstract, author summary, and introduction in order to make the focus of the paper more clear.
-We have also done a nearly complete rewritting of the discussion section.
-Specifically, though the content and arguments made remain the same, we specifically point out ways that our case study demonstrates the usefulness of our general suggestions for model based inference of dynamic systems.
+Our primary changes in the revised manuscript include edits to the abstract, author summary, introduction, in order to make the focus of the paper more clear.
+We have also done a nearly complete rewritting of the discussion section, improving the connection between the our overal goals with the cholera case study. 
+Specifically, though the content and arguments made in the discussion remain the same, we specifically point out ways that our case study demonstrates the usefulness of our general suggestions for model based inference of dynamic systems.
 
@@
 So, although the method is not entirely new, its applicability to this task is novel.
-Additional commentary to this point has been added in the abstract, introduction, and the subsection on fitting Model 3.
+Additional commentary to this point has been added in the abstract, introduction, and discussion sections.
 
@@
 We evaluate previous methods used for fitting these models, and we demonstrate modified data analysis strategies leading to improved statistical fit.
-Specifically, we present appraoches for diagnosing model misspecification and the consequent development of improved models.
-We futher demonstrate the utility of recent advances in likelihood maximization for high-dimensional nonlinear dynamic models, enabling likelihood-based inference for high-dimensional models for spatiotemporal incidence data.}
+Specifically, we present approaches for diagnosing model misspecification and the consequent development of improved models.
+Additionally, we demonstrate the utility of recent advances in likelihood maximization for high-dimensional nonlinear dynamic models, enabling likelihood-based inference for high-dimensional models for spatiotemporal incidence data.}
 
@@
 Specifically, we cite papers from the literature review we performed on the use of dynamic models for analyzing the Haiti cholera oubreak.
-The literature review was previously mentioned, but citations from the review were not made.
+Our literature review was previously mentioned, but citations from the review were not made.
 Furthermore, we modified the quoted statement in the author summary to be the following:
@@
 
-It could be either of the above, the main point being that the subscript seperates individuals into compartments that have distinct dynamics.
-We try to make this more clear by adding the following sentences:
+It could be either of the above, the main point being that the subscript seperates individuals into compartments that have distinct dynamics. We clarified this by adding the following sentences:
 
@@
 In other contexts, \cite{king15} noted various scientific advantages for including stochasticity in the model; it is undoubtedly part of the biological dynamics, and its omission can lead to over-confident predictions.
-Our revised manuscript includes several references demonstrating the importance of modeling extra-demonstraphic stochasticity, including our revised discussion:
+Our revised manuscript includes several references demonstrating the importance of modeling extra-demographic stochasticity, including our revised discussion:
 
 \article{\editDiscussionFour}
-
-We acknowledge the existence of useful mathematical tools for studying ODE models, their identifiability and stability and limiting properties.
-Perhaps one example demonstrating the usefulness of ODE models is the Lee et al.~\cite{lee20} study: of the four models considered, the forecasts from Model~2 were most consistent with the actual cholera situation in Haiti.
-Despite their inability to describe incidence data, deterministic models can do a good job at detecting overal trends, like the gradual decline of cholera cases in Haiti.
-This, coupled with the ease in fitting ODEs to data, makes ODEs a useful tool to gain preliminary insight on a dynamic system.
-Furthermore, stochastic dynamic models can be reduced to a deterministic model (a so-called determinsitic skeleton) which can usefully be analyzed to help understand the stochastic dynamics \cite{coulson04}.
-We therefore encourage the use of ODE analysis tools whether or not a stochastic model is adopted.
-Because this approach is tangential to the points we make in our analysis, and it was not done by Lee et al.~\cite{lee20}, we have not followed that approach in the current manuscript.
-A full discussion on deterministic vs simple stochastic models with and without over-dispersion is outside the scope of this article.
-However, we do make a few comments on the benefits of including overdispersion in a model in several locations, including the description of Model~1 and the conclusion.
-Because of their usefulness, we do not go as far as to explicitly forbid the use of deterministic models, but instead point out some of their limitations that should be considered.
-Our revised discussion includes a comment that demonstrates this belief:
-
-\article{Despite their limitations, deterministic models can offer valuable insights into dynamic systems \cite{may04}.
-In \cite{lee20}, the forecasts from the deterministic Model~2 were qualitatively more consistent with the observed dissapearance of cholera than the stochastic models.}
 
@@
 In practice, these models have been fitted by maximum likelihood, some examples including \cite{lemaitre19,fox22,molodecky23,pons-salort18,subramanian20,stocks20}.
+
+Despite their inability to describe incidence data, deterministic models can do a good job at detecting overal trends, like the gradual decline of cholera cases in Haiti.
+This, coupled with the ease in fitting ODEs to data, makes ODEs a useful tool to gain preliminary insight on a dynamic system.
+They are also a useful tool to understand a stochastic model \cite{coulson04}.
+Because of their usefulness, we do not go as far as to suggest that deterministic models should not be used. 
+Our revised discussion includes a comment that demonstrates this belief:
+
+\article{Despite their limitations, deterministic models can offer valuable insights into dynamic systems \cite{may04}.
+In \cite{lee20}, the forecasts from the deterministic Model~2 were qualitatively more consistent with the observed dissapearance of cholera than the stochastic models.}
 
@@
 
-\jwc{I'm not exactly sure how to respond to this point.
-The relevant information about what can be found in S5 text is precisly what is already stated: ``by reconstructing the latent states of Model~3, we notice that the calibrated model favors higher levels of cholera transmission than what was typically observed in the incidence data (S5 Text)".}
+We have expanded this paragraph to provide more details about the supplement material:
 
-\jwc{Maybe we could edit this part to the following:}
-
-\old{In the context of our case study, by reconstructing the latent states of Model~3, we notice that the calibrated model favors higher levels of cholera transmission than what was typically observed in the incidence data (S5 Text).}
-
-In the supplement material (S5 Text), we explore in more detail the process of model fitting and diagnostics.
-Here we demonstrate that although Model~3 does outperform its benchmark model on the aggregate scale, the model performance compared to the benchmark on units with the most cholera cases is poor.
-Furthermore, by comparing simulations from the fitted model to the filtering distribution, we see that the reconstructed latent states of the model favor higher levels of cholera transmission than what is typically observed in the incidence data.
-These result hints at the possibility of model mispecification, and should prompt us to be skeptical of the reliability of conclusions drawn from this model.
+\article{\editSuppText}
 

### 2024-03-02 edits.tex
Fixed units for parameters, other minor typos, and response to editors

@@
 In our case study, we compared the three models from Lee et al.
-(2020) to statistical benchmarks, revealing areas for enhancement.
+(2020) to statistical benchmarks, revealing areas for improvement.
 For example, comparisons of Model~3 to a benchmark revealed its inadequacy in accounting for the post-hurricane increase in transmission, leading to a beneficial model refinement.
@@
 If a model falls far behind simple benchmarks, there is likely to be substantial limitation in the data analysis that should be identified and remedied.
-In our case study, the recalibrated version of Model~1 outperforms its benchmark, so we proceeded to examine causal implications.
-When doing so, we find that the fitted model provides a causal description of the dynamic system that is consistent with known features of the system, such as the importance of rainfall as a driver of cholera infection.
+In our case study, the recalibrated version of Model~1 outperformed its benchmark, so we proceeded to examine causal implications.
+When doing so, we found that the fitted model provides a causal description of the dynamic system that is consistent with known features of the system, such as the importance of rainfall as a driver of cholera infection.
 The congruency between causal implications of the model and our belief about the dynamic system, coupled with a strong quantitative description of observed data relative to a benchmark, provides support for viewing the model as a plausible quantitative representation of the system under investigation.}
@@
 Model~3 is distinct in that it is both stochastic and has a meta-population structure, making it challenging to draw likelihood-based inferences.
-In this paper, we demonstrated how this model class could be calibrated to incidence data using the innovative IBPF algorithm.
+In this paper, we demonstrated how this model class can be calibrated to incidence data using the innovative IBPF algorithm.
 One of only a few examples of fitting a nonlinear non-Gaussian meta-population model via maximum likelihood \cite{li23,ionides22}, this case study exemplifies the algorithm's potential benefits and provides an example for future researchers on a possible approach to fitting a high-dimensional non-linear model.
@@
 \newcommand\editDiscussionFive{If forecasts are an important component of a modeling task, the forecasts should be consistent with the available data, particularly at the most recently available time points.
-In our case study, we do this by simulating forward from the filtering distribution, as this proceedure conditions latent variables on the available data.
+In our case study, we did this by simulating forward from the filtering distribution, as this proceedure conditions latent variables on the available data.
 This type of forecasting, however, is not directly available using a deterministic model, where future dynamics are fully determined by initial conditions and parameter values.
@@
 The robust statistical conclusion for our analysis is that a model which allows for change fits better than one which does not, and a trend in transmission is a plausible way to do this.}
+
+\newcommand\editSuppText{In the supplement material (\nameref{S_mod3cal}), we explore in more detail the process of model fitting and diagnostics for Model~3.
+Here we demonstrate that the model outperforms its benchmark model on the aggregate scale.
+However, when focusing on the spatial units with the highest incidence of cholera, Model~3 performs roughly the same as a simple benchmark.
+Furthermore, by comparing simulations from the fitted model to the filtering distribution, we see that the reconstructed latent states of the model favor higher levels of cholera transmission than what is typically observed in the incidence data.
+These results hint at the possibility of model mispecification, and warrant a degree of caution in intepreting the model's outputs.}

### 2024-03-02 ms.Rnw
Fixed units for parameters, other minor typos, and response to editors

@@
 \caption{\label{tab:params} {\bf Model parameters.}}
-\begin{tabular}{|p{0.15\linewidth}|p{0.14\linewidth}t|p{0.13\linewidth}t|p{0.17\linewidth}t|}
+\begin{tabular}{|p{0.15\linewidth}|p{0.14\linewidth}t|p{0.15\linewidth}t|p{0.17\linewidth}t|}
 \hline
@@
 \hline
-Infection (day)
+Infection {\small (day)}
   & ${\muIR^{-1}=\Sexpr{myround(7/p1["gamma"],1)}}\fixed$
@@
   & \eqref{eq:model3:IR}
-\\
-Latency (day)
+\\\hline
+Latency {\small (day)}
   & ${\muEI^{-1}=\Sexpr{myround(7/p1["sigma"],1)}}\fixed$
@@
   \\\hline
-  \multirow{3}{*}{Immunity (yr)}
+  \multirow{3}{*}{Immunity {\small (yr)}}
   & \multirow{3}{*}{${\muRS^{-1}=\Sexpr{myround(1/p1["alpha"]/52,1)}}\fixed$}
@@
 \hline
-\multirow{2}{*}{Birth/death (yr\textsuperscript{-1})}
+\multirow{2}{*}{Birth/death {\small (yr\textsuperscript{-1})}}
   & ${\muBirth =  10^{-2}\times \Sexpr{myround(100 * 52 * p1["mu"],2)}}\fixed$
@@
 &
-\\
+\\\hline
 Sympt. frac.
@@
   & \eqref{eq:model3:SA}
-\\
+\\\hline
 Asympt.
-  & ${\asymptomRelativeInfect=0.05}\fixed$
-  & \eqref{model1:lambda}
+  & \multirow{2}{*}{${\asymptomRelativeInfect=0.05}\fixed$}
+  & \multirow{2}{*}{\eqref{model1:lambda}}
   & ${\asymptomRelativeInfect = 0.001}\fixed$
@@
 infectivity
-& \missing
+&
 &
@@
   & \multirow{4}{*}{\eqref{model1:lambda}}
-  & \multirow{4}{*}{$\transmission=$\Sexpr{signif(h2_fit$h2_params["Beta"],3)}}
+  & \multirow{4}{*}{$\transmission=$\Sexpr{signif(h2_fit$h2_params["Beta"],3)} {\small yr\textsuperscript{-1}}}
   & \multirow{4}{*}{\eqref{model2:lambda}}
@@
 	\Sexpr{myround(p3u["foi_add",10]*1e6,2)}
-	) \allowbreak \times 10^{-6}$
+	) \allowbreak \times 10^{-6}$ {\small yr\textsuperscript{-1}}
   & \multirow{4}{*}{\eqref{eq:model3:foi}}
@@
         \Sexpr{myround(p3u["betaB",9],2)}, \allowbreak
-        \Sexpr{myround(p3u["betaB",10],2)})$
+        \Sexpr{myround(p3u["betaB",10],2)})$ {\small yr\textsuperscript{-1}}
   & \multirow{4}{*}{\eqref{eq:model3:foi}}
 \\
-& & & \multirow{-3}{*}{$\beta_W= \Sexpr{signif(h2_fit$h2_params["BetaW"],3)}$} & \multirow{-3}{*}{\eqref{model2:lambda}} & &
+& & & \multirow{-3}{*}{$\beta_W= \Sexpr{signif(h2_fit$h2_params["BetaW"],3)}$ {\small yr\textsuperscript{-1}}} & \multirow{-3}{*}{\eqref{model2:lambda}} & &
 \\
@@
   &
-  & $\Wshed = $ \Sexpr{signif(h2_fit$h2_params["Mu"]/52,3)}
+  & $\Wshed = $ \Sexpr{signif(h2_fit$h2_params["Mu"]/52,3)} {\small wk\textsuperscript{-1}}
   & \eqref{model2:to_W}
-  & $\Wshed= \Sexpr{signif(p3s["thetaI"],3)}$
+  & $\Wshed= \Sexpr{signif(p3s["thetaI"]/52,3)}$ {\tiny	$\frac{\mathrm{km^2}}{\mathrm{wk}}$}
   & \eqref{eq:model3:water}
 \\
-\raggedright Water survival (wk)
-  & \multirow{2}{*}{\missing}
+Water survival {\footnotesize (wk)}
+  & \missing
   &
-  & \multirow{2}{*}{${\Wremoval^{-1} = \Sexpr{signif(52/h2_fit$h2_params["Delta"],3)}}\fixed$}
-  & \multirow{2}{*}{\eqref{model2:from_W}}
-  & \multirow{2}{*}{$\Wremoval^{-1}=\Sexpr{myround(52/p3s["mu_B"],2)}$}
-  & \multirow{2}{*}{\eqref{eq:model3:Decay}}
-\\
+  & ${\Wremoval^{-1} = \Sexpr{signif(52/h2_fit$h2_params["Delta"],3)}}\fixed$
+  & \eqref{model2:from_W}
+  & $\Wremoval^{-1}=\Sexpr{myround(52/p3s["mu_B"],2)}$
+  & \eqref{eq:model3:Decay}
+\\\hline
 Mixing exponent
@@
 \hline
-\raggedright Process noise (wk\textsuperscript{1/2})
+\raggedright Process noise {\small (wk\textsuperscript{1/2})}
   & \raggedright $\sigmaProc = (\Sexpr{myround(p1["sig_sq_epi"],2)},
@@
   &
-  & \multirow{2}{*}{$\sigmaProc= \Sexpr{myround(p3s["std_W"],3)}$}
+  & \multirow{2}{*}{$\sigmaProc= \Sexpr{myround(p3s["std_W"] * sqrt(52),3)}$}
   & \multirow{2}{*}{\eqref{eq:model3:SA}}
@@
 \end{equation}
-\new{where $\bar{\transmission} = 1 (\mathrm{wk})^{-1}$ is a dimensionality constant.}
+\new{where $\bar{\transmission} = 1 (wk)\textsuperscript{-1}$ is a dimensionality constant.}
 The process noise $d\Gamma(t)/dt$ is multiplicative Gamma-distributed white noise, with infinitesimal variance parameter $\sigmaProc^2$.
@@
 Our case study provides an example of this in the parameter estimate for the duration of natural immunity due to cholera infection, $\muRS^{-1}$.
-Under the framework of Model~2, the best estimate for this parameter is $\Sexpr{signif(1/h2_fit$h2_params["sigma"],3)} \, \mathrm{yr}$, suggesting that individuals have \old{effectively} permanent immunity to cholera once infected.
+Under the framework of Model~2, the best estimate for this parameter is $\Sexpr{signif(1/h2_fit$h2_params["sigma"],3)}$ yr, suggesting that individuals have \old{effectively} permanent immunity to cholera once infected.
 \new{Rather than interpreting this as scientific evidence that individuals have permanent immunity from cholera, this result suggests that Model~2 favors a regime where reinfection events are a negligble part of the dynamics.}
@@
 For example, reconstructed latent variables (such as numbers of susceptible and recovered individuals) should make sense in the context of alternative measurements of these variables \cite{grad12}.
-In the context of our case study, by reconstructing the latent states of Model~3, we notice that the calibrated model favors higher levels of cholera transmission than what was typically observed in the incidence data (\nameref{S_mod3cal}).
-This result hints at the possibility of model mispecification, and should prompt us to be \old{sceptical} \new{skeptical} of the reliability of forecasts from this model.
+\old{In the context of our case study, by reconstructing the latent states of Model~3, we notice that the calibrated model favors higher levels of cholera transmission than what was typically observed in the incidence data (S5 Text).
+This result hints at the possibility of model mispecification, and should prompt us to be skeptical of the reliability of forecasts from this model.}
 Similarly, parameters that have been calibrated to data should make sense in the context of alternative lines of evidence about the phenomena being modeled, while making allowance for the possibility that the interpretations of parameters may vary when modeling across differing spatial scales.
+
+\new{\editSuppText}
 
@@
 \old{This article demonstrates that fitting appropriate scientific models to guide policy is a challenging statistical task.}
-\new{However, the development of credible models---which are supported by data and can lend quantitative insights into a dynamic system---remains a challenging task.
-In this article, we demonstrate opportunities available for raising the current standards of statistical inference for mathematical models of biological systems.}
-
-\new{We present methodology consistent with existing guidelines but exceeding standard practice.
-In particular, we show the value of comparing the likelihood of fitted mechanistic models versus non-mechanistic benchmarks.
-Further, we use likelihood-based diagnostic methods to identify specific limitations of proposed models.
-We use likelihood profile methods to assess parameter identifiability and to construct confidence intervals for parameter estimates.
-We assess the accuracy of numerical approximations involved in our calculations.
+\new{However, the development of credible models---which are supported by data and can provide quantitative insights into a dynamic system---remains a challenging task.
+In this article, we demonstrated opportunities available for raising the current standards of statistical inference for mathematical models of biological systems.}
+
+\new{We presented methodology consistent with existing guidelines but going beyond standard practice.
+In particular, we showed the value of comparing the likelihood of fitted mechanistic models versus non-mechanistic benchmarks.
+Further, we used likelihood-based diagnostic methods to identify specific limitations of proposed models.
+We used likelihood profile methods to assess parameter identifiability and to construct confidence intervals for parameter estimates.
+% We assessed the accuracy of numerical approximations involved in our calculations.
 When reaching conclusions, we identify potential consequences of confounded variables and model misspecification.

### 2024-03-02 si/confidenceIntervals.Rnw
Fixed units for parameters, other minor typos, and response to editors

--- 2024-02-13 si/confidenceIntervals.Rnw
@@
 
-  Process noise & $\sigmaProc: \mathrm{Epi}$ & $\Sexpr{myround(h1_mles %>% filter(prof_var == 'sig_sq_epi') %>% pull(value), 3)}$
+  Process noise {\small (wk\textsuperscript{1/2})} & $\sigmaProc: \mathrm{Epi}$ & $\Sexpr{myround(h1_mles %>% filter(prof_var == 'sig_sq_epi') %>% pull(value), 3)}$
    &
@@
 
-  Process noise & $\sigmaProc: \mathrm{End}$ & $\Sexpr{myround(h1_mles %>% filter(prof_var == 'sig_sq_end') %>% pull(value), 3)}$
+  Process noise {\small (wk\textsuperscript{1/2})} & $\sigmaProc: \mathrm{End}$ & $\Sexpr{myround(h1_mles %>% filter(prof_var == 'sig_sq_end') %>% pull(value), 3)}$
    &
@@
 
+  } else if (p == 'Mu') {
+
+    tmp_df <- h2_profiles %>% 
+      filter(prof_var == p) %>% 
+      # filter(logLik >= -21957.26 - 4) %>%
+      arrange(p)
+
+    vals <- tmp_df[, p]
+
+    profile_like <- splinefun(x = tmp_df[, p], y = tmp_df[, 'logLik'])
+    max_profile <- optimize(profile_like, interval = range(vals), maximum = TRUE)
+    threshold <- max_profile$objective - 1.92
+    plot_threshold <- max_profile$objective - 6
+    profile_roots <- function(x) {profile_like(x) - threshold}
+    plot_limits <- function(x) {profile_like(x) - plot_threshold}
+
+    lower_bound <- uniroot(
+      profile_roots, 
+      interval = c(min(vals), max_profile$maximum)
+    )$root
+
+    upper_bound <- uniroot(
+      profile_roots, 
+      interval = c(vals[which.max(tmp_df$logLik)], max(vals))
+    )$root
+
+    add_piece <- c(
+      'Mu' = 3000, 'BetaW' = 0.3, 'phase' = 0.2, v = 0.5 
+    )
+
+    max_x <- uniroot(
+      plot_limits, 
+      interval = c(max_profile$maximum, max(vals) + add_piece[p])
+    )$root
+
+    min_x <- uniroot(
+      plot_limits, 
+      interval = c(min(vals) - add_piece[p], max_profile$maximum)
+    )$root
+
+    h2_plots_lb[p] <- lower_bound
+    h2_plots_ub[p] <- upper_bound
+
+    gg_tmp <- ggplot() + 
+      geom_point(
+        data = h2_profiles %>% 
+          filter(prof_var == p) %>% 
+          filter(.data[[p]] <= max_x & .data[[p]] >= min_x) ,
+        aes(x = .data[[p]] / 52, y = logLik)
+      ) + 
+      geom_function(fun = profile_like) + 
+      geom_vline(xintercept = lower_bound / 52, linetype = 'dashed') + 
+      geom_vline(xintercept = upper_bound / 52, linetype = 'dashed') + 
+      geom_vline(xintercept = max_profile$maximum / 52, col = 'blue') + 
+      labs(y = 'Log-Likelihood', title = my_lab) + 
+      theme(
+        axis.text = element_text(size = 8),
+        axis.title = element_text(size = 10)
+      ) +
+      theme_bw() +
+      # scale_y_continuous(limits = c(-21963, -21957)) + 
+      scale_y_continuous(limits = c(max(h2_profiles$logLik)-6, max(h2_profiles$logLik))) +
+      scale_x_continuous(limits = c(min_x/52, max_x/52)) + 
+      theme(
+        axis.title.x = element_blank(), 
+        plot.title = element_text(
+          hjust = 0.5, size = 8, margin = margin(t = 0, unit = "pt")
+        ),
+        axis.title.y = element_text(size = 8),
+        axis.text.y = element_text(size = 7),
+        axis.text.x = element_text(size = 6.5),
+        plot.margin = margin(5.5, 3, 0, -1, unit = 'pt')
+      )
+
+    if (i %in% c(1, 4)) {
+      h2_plots[[p]] <- gg_tmp
+    } else {
+      h2_plots[[p]] <- gg_tmp + theme(
+        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
+        axis.title.y = element_blank()
+      )
+    }
   } else {
@@
 
- Human to water shedding & $\Wshed$ & $\Sexpr{myround(h2_mles['Mu'], 1)}$
-   &
-  $(\Sexpr{myround(h2_plots_lb['Mu'], 1)}, \Sexpr{myround(h2_plots_ub['Mu'], 1)})$ 
-\\
-\hline
-
- Water to Human Infection & $\beta_W$ & $\Sexpr{myround(h2_mles['BetaW'], 3)}$
+ Human to water shedding {\small (wk\textsuperscript{-1})} & $\Wshed$ & $\Sexpr{myround(h2_mles['Mu']/52, 1)}$
+   &
+  $(\Sexpr{myround(h2_plots_lb['Mu']/52, 1)}, \Sexpr{myround(h2_plots_ub['Mu']/52, 1)})$ 
+\\
+\hline
+
+ Water to Human Infection {\small (yr\textsuperscript{-1})} & $\beta_W$ & $\Sexpr{myround(h2_mles['BetaW'], 3)}$
    &
@@
 
-Human to Human Infection & $\transmission$ & $\Sexpr{signif(h2_mles['Beta'], 3)}$\textsuperscript{*}
+Human to Human Infection {\small (yr\textsuperscript{-1})} & $\transmission$ & $\Sexpr{signif(h2_mles['Beta'], 3)}$\textsuperscript{*}
    &
@@
 
-Immunity (yr) & $\muRS^{-1}$ & $\Sexpr{signif(1/h2_mles['sigma'], 3)}$\textsuperscript{*}
+Immunity {\small (yr)} & $\muRS^{-1}$ & $\Sexpr{signif(1/h2_mles['sigma'], 3)}$\textsuperscript{*}
    &
@@
   mutate(mu_B = mu_B / 52) %>% 
+  mutate(thetaI = thetaI / 52) %>% 
+  mutate(std_W =  std_W * sqrt(52)) %>%
   select(-epsilon, -XthetaA) %>% 
@@
   mutate(epsilon = epsilon^2) %>%
+  # mutate(std_W =  std_W * sqrt(52)) %>% 
   pivot_longer(
@@
   } else if (p == 'std_W') {
-    my_lab <- TeX('$\\sigma$')
+    my_lab <- TeX('$\\sigma_{proc}$')
   } else if (p == 'epsilon') {
@@
 
- Process Noise & $\sigmaProc$ & $\Sexpr{myround(h3_mles %>% filter(prof_var == 'std_W') %>% pull(value), 3)}$
-   &
-  $(\Sexpr{myround(all_ci %>% filter(prof_var == 'std_W') %>% pull(lower), 3)}, \Sexpr{myround(all_ci %>% filter(prof_var == 'std_W') %>% pull(upper), 3)})$ 
-\\
-\hline
-
- Water Survival (wk) & $\Wremoval^{-1}$ & $\Sexpr{myround(52 / (h3_mles %>% filter(prof_var == 'mu_B') %>% pull(value)), 3)}$
+ Process Noise {\small (wk\textsuperscript{1/2})} & $\sigmaProc$ & $\Sexpr{myround((h3_mles %>% filter(prof_var == 'std_W') %>% pull(value)) * sqrt(52), 3)}$
+   &
+  $(\Sexpr{myround((all_ci %>% filter(prof_var == 'std_W') %>% pull(lower)), 3)}, \Sexpr{myround((all_ci %>% filter(prof_var == 'std_W') %>% pull(upper)), 3)})$ 
+\\
+\hline
+
+ Water Survival {\small (wk)} & $\Wremoval^{-1}$ & $\Sexpr{myround(52 / (h3_mles %>% filter(prof_var == 'mu_B') %>% pull(value)), 3)}$
    &
@@
 
- Human to Water Shedding & $\Wshed$ & $\Sexpr{signif(h3_mles %>% filter(prof_var == 'thetaI') %>% pull(value), 3)}$
+ Human to Water Shedding {\small $\frac{\mathrm{km^2}}{\mathrm{wk}}$} & $\Wshed$ & $\Sexpr{signif((h3_mles %>% filter(prof_var == 'thetaI') %>% pull(value))/52, 3)}$
    &

### 2024-03-04 r1/response.tex
final edits to MS and response, ready to clean and submit

@@
 \newcommand\muRS{\mu_{RS}}
+\newcommand\sigmaProc{\sigma_{\mathrm{proc}}}
 \newcommand\transmissionTrend{\zeta}
@@
 
-The other main change in the revised manuscript is edits to both the introduction and discussion, as recommended by both reviewers.
+The other main change in the revised manuscript is edits to the introduction and discussion, as recommended by both reviewers.
+These changes help make the goals of the mansucript clear, and connect the specific case study that was conducted to these goals.
+This resulted in improvements to the article without modifying the overall content of the original submission.
 
@@
 Thank you for this feedback.
-Our primary changes in the revised manuscript include edits to the abstract, author summary, introduction, in order to make the focus of the paper more clear.
+Our primary changes in the revised manuscript include edits to the abstract, author summary, and introduction, with the goal of making the focus of the paper more clear.
 We have also done a nearly complete rewritting of the discussion section, improving the connection between the our overal goals with the cholera case study. 
-Specifically, though the content and arguments made in the discussion remain the same, we specifically point out ways that our case study demonstrates the usefulness of our general suggestions for model based inference of dynamic systems.
+Specifically, though the content and arguments made in the discussion remain the same, we point out ways that our case study demonstrates the usefulness of our general suggestions for model based inference of dynamic systems.
 
@@
 
-\article{In this article, we conduct a re-analysis of the 2010 cholera outbreak in Haiti.
-We use this case study to investigate current procedures for fitting mathematical models to time series data, while identifying limitations of these methodologies and proposing remedies.}
+\article{In this article, we conduct a re-analysis of the 2010-2019 cholera outbreak in Haiti.
+We use this case study to investigate current procedures for fitting mechanistic models to time series data, while identifying limitations of these methodologies and proposing remedies.}
 
@@
 In our analysis we only consider vaccination dynamics based on the scenarios outlined in the Forecasts section.
-For Model~1, the assumption made by Lee et al.~\cite{lee20}, and reused in our analysis, is that the added dynamics of the vaccination compartments is an increased probability that an infection is asymptomatic, as described in a paragraph that follows the quoted text.
+For Model~1, the assumption made by Lee et al.~\cite{lee20}, and reused in our analysis, is that the added dynamics of the vaccination compartments is an increased probability that an infection is asymptomatic, as described in an unmodified paragraph that follows the quoted text.
 
@@
 This parameter comes from Lee et al.~\cite{lee20}, and the only details that they provide is a subscript comment in one of their tables (S13) in their supplement: ``$v_{rate}$ was chosen to mimic the distribution of infections in Haiti".
-They do not, however, list it as a parameter that was ``fit" in the same table, making it unclear where the value comes from.
+In the same table, however, they do not list $v_{rate}$ as a parameter that was ``fit", making it unclear where the value comes from.
 Presumably, it was chosen somewhat subjectively based on their understanding of cholera dynamics in Haiti, because in order to obtain a fit for their model they had to have fixed the value at something without actually estimating it, and perhaps only considered a very small set of possible values.
 Notably, this parameter can mathematically be estimated like all other parameters in the model, but their specific implementation of Model~2 does not permit its estimation.
-A full discussion on parameters that are partially estimated via subjective measures is outside the scope of our article, so the revised manuscript simply reads:
+A full discussion on parameters that are partially estimated via subjective measures is outside the scope of our article. To clarify where this fixed value comes from the revised manuscript reads:
 
@@
 Details on how they are initialized are provided in the section with the model descriptions, as well as supplement~S5.
-% To initialize the latent states of Model~3, Lee et al~\cite{lee20} used approximately $60\%$ of the available data as ``burn-in" observations to initialize latent states.
-According to their supplement, this was done primarily because they believed the dynamics from the start of the outbreak were significantly different enough from the dynamics at the end of the outbreak to suggest that two seperate models would be needed to appropriately describe the data.
-% We feel that a properly designed and calibrated model should be able to explain the entire dynamics of the outbreak, including the change from what Lee et al~\cite{lee20} referred to as the ``epidemic" and ``endemic" periods of the outbreak.
 
-% One possible approach to re-analyzing their model would then be to fit all available observations; this would make Models~2 and 3 directly comparable, but would require a significant change to Model~3 itself, as the initialization model used by Lee et al~\cite{lee20} requires a ``burn-in" period.
-One of our goals, however, was to demonstrate that the models provided by Lee et al.~\cite{lee20} were not the issue, and a complete refactorization of the initialization could potentially contradict this objective.
-Therefore instead we decided to substantially shrink the ``burn-in" period to only the first three available observations, which corresponds to a natural change in the observations themselves \cite{barzilay13}.
-These details are available in the Initial Values section in the supplement material.
-
-% In short, though models that are fit to the same dataset are directly comparable using AIC, this is not the case for Models~2 and 3, as the first few observations in the dataset were not considered likelihood evalution of Model~3.
-This also explains the slight discrepency in AIC for the benchmark models of these two models (Table 2).
 We added the following sentence to clarify this point:
@@
 \article{Models that are fit to the same datasets can be directly compared using AIC values, making it a useful tool to compare to benchmark models.
-Though Models~2 and 3 are both fit to department level incidence reports, their AIC values are not directly comparable due to the way Model~3 initializes latent states, which is described in detail in Sec.~S5 of the supplement.}
+Though Models~2 and 3 are both fit to department level incidence reports, their AIC values are not directly comparable due to the way Model~3 initializes latent states (\nameref{S_init}).}
 
@@
 The revised version of the paper contains significant edits to the introduction and discussion sections that we feel address this concern.
+In the discussion, the revised mansucript specifically mentions how the recalibrated models suggest a moderately high probability of cholera elimination from 2019-2022, which is consistent with what actually occured.
+We also discuss scientific insights that can made with the recalbrated models, such as the effect of Hurricane Matthew on cholera incidence, the importance of rainfall as a driver of cholera infection in Haiti, and the presence of overdispersion in the cholera dynamics.
 
@@
 
-The referee's suggestions have led to the addition of various references in the revision.
+The referee's suggestions have led to the addition of various references in the revision, which now includes over 100 citations.
 For the particular issue of black box versus white box models, we have included the following paragraph:
@@
 
-This addition is intended to compliment the discussion about mechanistic (white-box) vs associative (black-box) models that was provided at the start of the ``Materials and methods" section.
+This addition is intended to compliment the discussion about mechanistic (white-box) vs associative (black-box) models that was part of the ``Materials and methods" section.
 
@@
 In our revised manuscript, we address identifiability via profile likelihood methods (\nameref{S_CI}).
-% AIC and other formal statistical methods can assess whether the additional complexity is justified by a sufficiently large improvement of statistical fit.
-% Identifiability and model simplicity have practical scientific value which goes beyond their statistical role in reducing generalization error and hence improving out-of-fit predictions.
-% This is a variation on the issue of ``statistical significance versus practical significance"  often covered in introductory statistics textbooks.
 
-In our study, we find that the deterministic model (model 2) has poor statistical fit.
+In our study, we find that the deterministic model (model 2) has poor statistical fit (Table~2).
 A poor fit can lead to a risk of identifying spurious scientific phenomena: additional deterministic complexity added to the model may help to fit the model to unmodeled stochastic phenomena present in the data, and may therefore spuriously be assessed as a highly statistically significant model improvement.
@@
 \article{Despite their limitations, deterministic models can offer valuable insights into dynamic systems \cite{may04}.
-In \cite{lee20}, the forecasts from the deterministic Model~2 were qualitatively more consistent with the observed dissapearance of cholera than the stochastic models.}
+In \cite{lee20}, the forecasts from the deterministic Model~2 were qualitatively more consistent with the observed disappearance of cholera than the stochastic models.}
 
@@
 Thus, particle filter methods facilitate the use of various likelihood-based methods for diagnostics and model selection.
-We demonstrate this in the supplement (\nameref{S_CI}, \nameref{S_mod3cal}). 
+We demonstrate this in the supplement (\nameref{S_mod3cal}, \nameref{S_CI}). 
 % Figs S4, S5, S6, S7. \TODO{Check all figure numbers, especially to the supplement.}
@@
 Monte Carlo adjusted profile (MCAP) methods \cite{ionides17,ning21} provide methodology which reduces and evaluates the consequences of this Monte Carlo error.
-MCAP methods have been used in various recent investigations using particle filters \cite{subramanian21,ali22,pons-salort18}.
+MCAP methods have been used in various recent investigations using particle filters \cite{subramanian21,ali22,pons-salort18,simpson23}.
 Our revised manuscript includes MCAP confidence intervals for both Model~1 and Model~3 (\nameref{S_CI}).
@@
 We would like to do what we can, within the constraints of this paper, to help build awareness of these diagnostic methods and ways in which they can be useful.
-A full exposition of this topic is out of the scope of this paper.
+A full exposition of this topic is out of the scope of this paper, though some of these diagnostics are included in supplement material (\nameref{S_mod3cal}, \nameref{S_CI}).
+To further help the reader in this regard, we have added additional information in the Model~1 calibration subsection, which includes citations to papers that discuss these diagnostics or have extensive supplements that demonstrate these diagnostic issues in action.
 
-% \jwc{This text below is something that Ed added.
-I think it's a good idea.
-However, simply because of the overal length of the paper and total number of citations, I think it's probably okay to just leave this out}
-%
-% Howerver, we can alert the reader to the topic, and provide some additional references.
-% Not infrequently, diagnostic issues are relegated to supplementary material and so we have indicated to interested readers some papers with extensive supplements demonstrating these diagnostic issues in action.
-% \cite{pons-salort18}
+\article{One change we made in the statistical analysis that led to larger model likelihoods was increasing the computational effort in the numerical maximization.
+While IF2 enables parameter estimation for a large class of models, the theoretic ability to maximize the likelihood depends on asymptotics in both the number of particles and the number of filtering iterations.
+Many Monte Carlo replications are then required to quantify and further reduce the error.
+The large increase in the log-likelihood for Model~1 (Table~2) can primarily be attributed to increasing the computational effort used to calibrate the model.
+This result highlights the importance of carefully determining the necessary computational effort needed to maximize model likelihoods and acting accordingly.
+In this case study, this was done by performing standard diagnostics for the IF2 and particle filter algorithms \cite{king16,li23,pons-salort18,laneri10}.
+Given the considerable computational costs of simulation-based algorithms, we find it useful to perform an initial assessment using hyperparameter values---such as the number of particles, filtering iterations, and replicates based on different parameter initializations---that enable relatively quick calculations.
+The insights obtained from this preliminary analysis help in accurately determining the amount of computation that is required to achieve reliable outcomes.}
 
@@
 We agree with the referee that uncertainty quantification and evaluation of identifiability are fundamental tasks for model-based data analysis.
-In the revision, we have included profile confidence intervals on all parameters, except those taking separate values on each unit.
+In the revision, we have included profile confidence intervals on all parameters, except those taking separate values on each unit (\nameref{S_CI}).
 For these unit-specific parameters, the spread of values across units provides an alternative measure of uncertainty.
@@
 This range may not be considered unreasonably large given the inherent stochasticity present in the dynamic system and all of the many possible outcomes of the outbreak from its initial state in 2010.
+
+% Finally, we agree that we must be careful in drawing conclusions from any model, and that it is important acknowledge possible limitations.
+In our original manuscript, we explicitly noted that we should be careful drawing strong conclusions drawn from Model~3 due to some hints of model misspecification (\nameref{S_mod3cal}).
+This is clarified further in the revised manuscript:
+
+% \article{\editSuppText}
+
 % Shorter term forecasts obtained by conditioning on previous data can be obtained by simulating from the filtering distribution.
@@
 
-We have added several paragraphs in the discussion to address this point. These paragraphs are listed below for reference.
+We have added several paragraphs in the discussion to address this point. These paragraphs are provided below for reference.
 
@@
 
-\article{Our analysis presents methodology for diagnosing how well a model describes observed data.
-Objective measures are used to evaluate model fit; when objective measures of the goodness-of-fit of a model are combined with reproducibility, a framework emerges for continual improvement to model based inference when revisiting the data.
-This framework promotes scientific discovery and supports model-driven policy recommendations by simplifying the process of understanding and improving previous results.}
+\article{\editAuthorSum}
 
@@
 
-\article{Despite their autonomy, the four independent teams obtained a consensus that an extensive nationwide vaccination campaign would be necessary to eliminate cholera from Haiti.
-They also provided a range for the forecasted number of cumulative cases from February 2019 to Februaruy 2024 based on model simulations, with median estimates from each model measured in hundreds of thousands.
+\article{Despite their autonomy, the four independent teams obtained a consensus that an extensive nationwide vaccination campaign would be necessary to eliminate cholera from Haiti, estimating that a large number of cumulative cholera cases would be observed in the absense of additional vaccination efforts (Figure~3 and 4 of \cite{lee20}).
 These forecasts are inconsistent with the prolonged period with no confirmed cholera cases between February, 2019 and September, 2022 \cite{trevisin22}.}
@@
 
-Due to the size and complexity of the table we have elected to put uncertainty estimates elsewhere.
-Specifically, have created an additional supplement material document that outlines how confidence intervals are obtained, along with tables and figures that show parameter uncertainty estimates.
+Due to the size and complexity of the table we have elected to put uncertainty estimates elsewhere (\nameref{S_CI}).
+This additional supplement material document outlines how confidence intervals are obtained, and includes tables and figures that show parameter uncertainty estimates.
 In the caption for Table~1, we include a reference to this supplement material.
@@
 Unsurprisingly, fitting initial values results in a quantitative improvement to model-fit, measured by AIC, but we also found that this had negligable qualitative effects on model-based conclusions.
-On a nationally aggregated scale, the total number of initial infectious individuals in the fixed scenario was 36500, and 33480 when the initial states were estimated.
-We have included additional material in the supplement for Initialization Models that demonstrates this. % TODO: Specifically link / cite supplement.
+Importantly, the fitted model still falls short of the simple statistical benchmark used for comparison (\nameref{S_init}).
+We have included additional material in the supplement for Initialization Models that demonstrates this. 
+On a nationally aggregated scale, the total number of initial infectious individuals in the fixed scenario was 36500, and 33480 when the initial states were estimated, which perhaps explains why calibrating initial parameters had little effect on qualitative conclusions.
 
@@
 
-% TODO: Revisit this a bit more, giving explicit details in the manuscript and supplement that address this concern.
-% The cholera pathogen was recently introduced to Haiti, which facilitates the specification of initial conditions.
 Our experience with mechanistic modeling reflects that of the reviewer: calibrating initial values often results in improved dynamics.
 As such, we would recommend calibrating initial values in most cases; here, doing so may constitute a major change from the initialization model used by Lee et al.~\cite{lee20}.
-Because one of our goals was to demonstrate that the models of \cite{lee20} were alone not responsible for their poor forecasts, we elected to use the same initialization approach to the extent possible.
+Because one of our goals was to demonstrate that the models of \cite{lee20} were alone not responsible for their poor forecasts, we elected to use the same initialization approach to the extent possible, as noted in the manuscript.
 
-The goals of inferring initial prevalence is to reduce the number of parameters.
+The goal of inferring initial prevalence is to reduce the number of parameters.
 The risk is that the form of this assumption could have substantial effects on the dynamics, and hence on the consequences of the analysis.
-In our supplement material, we described our approach to addressing this issue: (i) looking at conditional log-likelihoods to assess time points at which the model is misspecified; (ii) if the initial values seem problematic, calculate the room for improvement by estimating the initial values as free parameters.
-We found that enforcing model dynamics for initial states worked well for many of the departments, but was problematic for departments where this resulted in zero counts in Infected, Asymptomatic, and Bacterial compartments, leading to the estimation of initial counts of Infected individuals in two departments (Grand'Anse and Nippes).
-Initial values for Asymptomatic and Bacterial compartments were then obtained by enforcing the dynamics of the model using the estimates for Infected individuals, as described in the supplement material.
+In our supplement material (\nameref{S_mod3cal}), we discuss investigating conditional log-likelihoods of each observation in order to detect model misspecification.
+Here, we do not see clear evidence that fixing initial values in the same way that was done by \cite{lee20} results in poor fits of the first few available observations.
+Specifically, we found that enforcing model dynamics for initial states worked well for many of the departments, but was problematic for departments where this resulted in zero counts in Infected, Asymptomatic, and Bacterial compartments, leading to the estimation of initial counts of Infected individuals in two departments (Grand'Anse and Nippes) (units 3 and 4, respectively).
+See Table~1 for estimates of initial values for these two units.
 
@@
 
-\article{The large increase in the log-likelihood for Model~1 (see Table~2) can primarily be attributed to increasing the computational effort in fitting the model.
+\article{The large increase in the log-likelihood for Model~1 (Table~2) can primarily be attributed to increasing the computational effort used to calibrate the model.
 This result highlights the importance of carefully determining the necessary computational effort needed to maximize model likelihoods and acting accordingly.
-In this case study, this was done by performing standard diagnostics for the IF2 and particle filter algorithms \cite{king16}.
-Given the considerable computational costs of simulation-based algorithms, we find it useful to perform an initial assessment using hyperparameter values that enable relatively quick calculations.
+In this case study, this was done by performing standard diagnostics for the IF2 and particle filter algorithms \cite{king16,li23,pons-salort18,laneri10}.
+Given the considerable computational costs of simulation-based algorithms, we find it useful to perform an initial assessment using hyperparameter values---such as the number of particles, filtering iterations, and replicates based on different parameter initializations---that enable relatively quick calculations.
 The insights obtained from this preliminary analysis help in accurately determining the amount of computation that is required to achieve reliable outcomes.}
@@
 In their supplement material, however, they provide a table (S7) in which they show that if they conditionally select only the simulations that were consistent with recent data (0 cases from 2019-2020), their forecasts predict cholera elimination.
-This second effort that was only presented in the supplement material is more consistent with what we suggest is correct practice: forecasts of a system should start using latent states that are consistent with available data.
+This second effort was only presented in the supplement material, but is more consistent with what we suggest is correct practice: forecasts of a system should start using latent states that are consistent with available data.
 We have expanded this argument in order to clarify this point:
@@
 
+\paragraph*{S2 Text}
+\label{S_init}
+{\bf Initialization models.} Additional details of the initialization model that were used for Models~1--3.
+
 \paragraph*{S5 Text}

### 2024-03-04 edits.tex
final edits to MS and response, ready to clean and submit

@@
 \newcommand\editDiscussionOne{Model diagnostics are a key tool for exposing unresolved model limitations and improving model fit.
-In our case study, we compared the three models from Lee et al.
-(2020) to statistical benchmarks, revealing areas for improvement.
+In our case study, we compared the three models from Lee et al.~\cite{lee20} to statistical benchmarks, revealing areas for improvement.
 For example, comparisons of Model~3 to a benchmark revealed its inadequacy in accounting for the post-hurricane increase in transmission, leading to a beneficial model refinement.
@@
 Stochastic phenomena are not only arguably present in biological systems, but their inclusion in a model also allows observed data variations to be attributed to inherent uncertainty rather than to distorted parameter values.
-Models~1 and 3 suggest the presence of extra-demographic stochsticity \cite{he10,stocks20,li23}, as evidenced by the confidence intervals (\nameref{S_CI}).} % TODO: The nameref won't work in the response.
+Models~1 and 3 suggest the presence of extra-demographic stochasticity \cite{he10,stocks20,li23}, as evidenced by the confidence intervals for the corresponding parameter $\sigmaProc$ (\nameref{S_CI}).} % TODO: The nameref won't work in the response.
 
@@
 Despite their limitations, deterministic models can offer valuable insights into dynamic systems \cite{may04}.
-In \cite{lee20}, the forecasts from the deterministic Model~2 were qualitatively more consistent with the observed dissapearance of cholera than the stochastic models.
+In \cite{lee20}, the forecasts from the deterministic Model~2 were qualitatively more consistent with the observed disappearance of cholera than the stochastic models.
 In our case study, we found improvements to Models~1 and 3 that resulted in improved forecasts for these models.}
@@
 Alternative approaches used to fit stochastic models included making additional simplifying assumptions to aid in the fitting process \cite{lee20}, and using MCMC algorithms \cite{sallah17,azman12} which require specific structures in the latent dynamics, making these algorithms non plug-and-play.
-In this subsection, we present how the recently developed iterated block particle filter (IBPF) algorithm \cite{ning21ibpf,ionides22} can be used to fit a spatially explicit stochastic dynamic model to incidence data.}
+In this subsection, we present how the recently developed iterated block particle filter (IBPF) algorithm \cite{ning23,ionides22} can be used to fit a spatially explicit stochastic dynamic model to incidence data.}
 
@@
 \newcommand\editConfound{The statistical evidence of a trend in transmission rate in this model could be explained by any trending variable (such as hygiene improvements, or changes in population behavior), resulting in confounding from collinear covariates.
-Alternatively, the trend could be attributed to a decreasing reporting rate rather than decreasing transmission rate.
+Alternatively, it is possible that the negative trend observed in the incidence data could be attributed to a decreasing reporting rate rather than decreasing transmission rate.
 This could be formally tested by comparing models with either trend specification.
@@
 However, when focusing on the spatial units with the highest incidence of cholera, Model~3 performs roughly the same as a simple benchmark.
-Furthermore, by comparing simulations from the fitted model to the filtering distribution, we see that the reconstructed latent states of the model favor higher levels of cholera transmission than what is typically observed in the incidence data.
-These results hint at the possibility of model mispecification, and warrant a degree of caution in intepreting the model's outputs.}
+By comparing simulations from the fitted model to the filtering distribution, we see that the reconstructed latent states of the model favor higher levels of cholera transmission than what is typically observed in the incidence data.
+These results hint at the possibility of model mispecification, and warrant a degree of caution in interpreting the model's outputs.}
+
+\newcommand\editAuthorSum{Our analysis presents methodology for diagnosing how well a model describes observed data.
+Using objective measures to assess model fit ensures that our evaluation is based on quantifiable criteria.
+Incorporating reproducibility into this assessment results in a framework that enables the validation or refinement of model based inferences when revisiting the data, facilitating scientific discovery.}

### 2024-03-04 si/calibrateMod3.Rnw
Fixing section numbers and other minor details in supplement material
final edits to MS and response, ready to clean and submit

@@
 To calibrate this model, we used the iterated block particle filter (IBPF) method of \cite{ionides22}.
-Due to the novelty of this algorithm, there does not exist published examples the IBPF algorithm used for data analysis outside of the papers in which the algorithm was introduced \citep{ning21ibpf,ionides22}, which is one motivation of the inclusion additional details related to fitting and diagnosing the model fit provided here.
-
-\cite{lee20} were only able to estimate model parameters to a simplified version of Model~3 on a subset of the available data, as no method existed at the time of their publication to fit a fully coupled meta-population model to disease incidence data.
-Building on their results, we fit the fully coupled version of Model~3 to nearly all available data, reserving only a few observations to use to calibrate the initial conditions of the model (see the previous section for more details).
-To maximize model likelihoods, we relied on parameter estimates obtained via the profile-likelihood confidence intervals provided in section S8.
-Prior to obtaining estimates via the profile likelihood search, we found that it was necessary to use multiple searches for the MLE, periodically pruning away less successful searches.
-The first collection of searches is performed by obtaining initial values for the parameters by uniformly sampling values from a predefined hypercube.
+Due to the novelty of this algorithm, there exists only a few examples where the IBPF algorithm is used for data analysis \citep{li23,ionides22}, which is one motivation of the inclusion additional details related to fitting and diagnosing the model fit provided here.
+
+\cite{lee20} only estimated model parameters to a simplified version of Model~3 on a subset of the available data, as no method existed at the time of their publication to fit a fully coupled meta-population model to disease incidence data.
+Building on their results, we fit the fully coupled version of Model~3 to (nearly) all available data, reserving only a few observations to use to calibrate the initial conditions of the model (see the supplement for initialization models for more details).
+To maximize model likelihoods, we relied on parameter estimates obtained while calculating profile-likelihood confidence intervals, as this calculation requires many replicated IBPF searches.
+In our preliminary investigations that were done prior to conducting a profile likelihood search, we found that it was necessary to use multiple searches for the MLE, periodically pruning away less successful searches.
+To do this, the first collection of searches is performed by obtaining initial values for the parameters by uniformly sampling values from a predefined hypercube.
 A subsequent refinement search used parameter values corresponding to the largest model likelihoods as starting parameter values.
 The need for multiple searches does not appear to be uncommon, as a similar approach was used by \cite{ionides22}.
-While computationally inefficient, profile likelihoods proved to be an effective alternative to maximizing model likelihoods without the need to apply this multistage heuristic.
-
-We use the technique described above to fit the fully coupled version of Model~3 proposed by \cite{lee20}.
+While computationally intensive, profile likelihoods proved to be an effective alternative to maximizing model likelihoods without the need to apply this multistage heuristic.
+
+We use the iterative fitting / pruning technique described above to fit the fully coupled version of Model~3 proposed by \cite{lee20}.
 The maximum likelihood we obtained after two rounds of searching was $\Sexpr{round(max(haiti3_sub_fit$search2$logLiks$logLik))}$, which is higher than the benchmark model ($\Sexpr{round(sum(unitLikes_bench$ll))}$).
-While beating a simple associative benchmark is promising, this does not immediately rule out the possibility that the model is a poor description of the system.
+While beating a simple associative benchmark is promising, this does not immediately imply that the model is a good description of the system.
 Additional investigation of parameters estimates and their corresponding implications on model based conclusions should always be conducted.
@@
 The likelihoods for each department, compared to the corresponding benchmark model, are displayed in Fig.~\ref{fig:h3UnitLikes}.
-The figure demonstrates that while the log-likelihood of the fitted model outperforms the auto-regressive negative binomial benchmark model, Model~3 has lower likelihoods than the benchmark for some departments.
+The figure demonstrates that while the log-likelihood of the fitted model outperforms the auto-regressive negative binomial benchmark model at the aggregate level, Model~3 has lower likelihoods for some departments.
 
@@
 
-We refit Model~3 after introducing these hurricane-related parameters using the same successive search scheme described above.
+We refit Model~3 after introducing these hurricane-related parameters.
 The resulting model has a log-likelihood value of $\Sexpr{round(readRDS("../model3/run_level_3/mod3_eval_res.rds")[['ll']]['est'], 1)}$.
-The addition of the Hurricane parameters seems to have resulted in a dramatic increase in conditional likelihoods around and after October 2016, as seen in Fig.~\ref{fig:finalCondLL}.
 The inclusion of these parameters resulted in an overall increase of $\Sexpr{round(readRDS("../model3/run_level_3/mod3_eval_res.rds")[['ll']]['est'] - max(haiti3_sub_fit$search2$logLiks$logLik), 1)}$ log-likelihood units.
 Such a large difference in log-likelihoods is well beyond the threshold of statistical uncertainty determined by Wilks' theorem, suggesting that the data highly favor the inclusion of the additional parameters.
+The addition of the Hurricane parameters also increases in conditional likelihoods for each observation, particularly around October 2016 (Fig.~\ref{fig:finalCondLL}).
 
@@
 Now that the model with additional parameters has been calibrated to the incidence data, we plot the conditional log-likelihood of each department compared to a benchmark model in Fig.~\ref{fig:finalUnitLL}.
-Interestingly, the difference in log-likelihoods between Model~3 and the benchmark model is smallest in the departments \Sexpr{bench_diffs_final$unit[1]}, \Sexpr{bench_diffs_final$unit[2]}, \Sexpr{bench_diffs_final$unit[3]} and \Sexpr{bench_diffs_final$unit[4]}.
+The difference in log-likelihoods between Model~3 and the benchmark model is smallest in the departments \Sexpr{bench_diffs_final$unit[1]}, \Sexpr{bench_diffs_final$unit[2]}, \Sexpr{bench_diffs_final$unit[3]} and \Sexpr{bench_diffs_final$unit[4]}.
 Each of these departments also exhibited the most sustained cholera transmission, defined by having the fewest number of weeks with no recorded cholera cases.
@@
 This result may be unsurprising in the context of the models that we are comparing.
-Because the cholera transmission in individual departments likely depends on the national prevalence of cholera and the Vibrio cholerae bacteria, our benchmark model that relies exclusively on the previous number of case within any given unit has a difficult time predicting a resurgence of cases.
+Because the cholera transmission in individual departments likely depends on the national prevalence of cholera and the Vibrio cholerae bacteria, our spatially-independnet benchmark model that relies exclusively on the previous number of case within any given unit has a difficult time predicting a resurgence of cases.
 
@@
 In particular, we examine the compartment of susceptible individuals under various scenarios.
+This analysis can also provide insight into why the calibrated model fails to outperform the benchmark model on the four departments with the most sustained cholera transmission.
 
@@
 This figure shows that simulations from initial conditions tends to result in a much more rapid depletion of susceptible individuals at the start of the epidemic than simulations from the filtering distribution, suggesting the calibrated model has a propensity to predict larger outbreaks than what is typically seen in the data.
-This also appears true in later stages of the time period of observed data, where the data suggests that there is a replenishment of the susceptible compartment, but the Model simulations typically retain a smaller proportion of susceptible individuals.
-This result demonstrates that the calibrated model favors a more rapid growth in cholera cases than what is typically seen in the observed data. 
-This, coupled with the fact that the model fails to beat a simple statistical benchmark on data on the four departments that experienced the majority of observed cholera cases, suggests that one should be careful when making inference using this model.
+This result demonstrates that the calibrated model favors a more rapid growth in cholera cases than what is typically seen in the observed data, providing a possible explanation as to why the model fails to beat the simple benchmark for each spatial unit.
+This results hints at the possibility of model mispecification, and warrants a degree of caution in interpreting the model's outputs.
 

### 2024-03-04 si/confidenceIntervals.Rnw
final edits to MS and response, ready to clean and submit

@@
   } else if (p == 'sig_sq_epi') {
-    my_lab <- TeX('$\\sigma$: Epidemic')
+    my_lab <- TeX('$\\sigma_{proc}$: Epidemic')
   } else if (p == 'sig_sq_end') {
-    my_lab <- TeX('$\\sigma$: Endemic')
+    my_lab <- TeX('$\\sigma_{proc}$: Endemic')
   } else if (p == 'tau_end') {
@@
 @
-\caption{\label{fig:m1Profs}MCAP confidence intervals for Model 1 parameters.}
+\caption{\label{fig:m1Profs}MCAP confidence intervals for Model 1 parameters. The vertical blue line indicates the smoothed MLE.}
 \end{figure}
@@
 @
-\caption{\label{fig:m2Profs}MCAP confidence intervals for Model 2 parameters.}
+\caption{\label{fig:m2Profs}MCAP confidence intervals for Model 2 parameters. The vertical blue line indicates the MLE.}
 \end{figure}
@@
 @
-\caption{\label{fig:m3Profs}MCAP confidence intervals for Model 3 parameters.}
+\caption{\label{fig:m3Profs}MCAP confidence intervals for Model 3 parameters. The vertical blue line indicates the smoothed MLE.}
 \end{figure}
@@
 
- Asymptotic Shedding & $\asymptomRelativeShed$ & $\Sexpr{myround((h3_mles %>% filter(prof_var == 'XthetaA') %>% pull(value))^2, 3)}$
+ Asymptomatic Shedding & $\asymptomRelativeShed$ & $\Sexpr{myround((h3_mles %>% filter(prof_var == 'XthetaA') %>% pull(value))^2, 3)}$
    &

### 2024-03-04 trackedMS.Rnw
Creating tracked version of the MS

--- 2024-03-02 ms.Rnw
@@
 \old{In this article, we discuss common modeling decisions that may not provide an adequate statistical explanation of the data, and we show that attention to statistical fit can lead to more accurate policy evaluations}.
-\new{In this article, we conduct a re-analysis of the 2010 cholera outbreak in Haiti.
-We use this case study to investigate current procedures for fitting mathematical models to time series data, while identifying limitations of these methodologies and proposing remedies.}
-\new{We present} \old{Our analysis presents} methodology for diagnosing how well a model describes observed data\new{.}\old{, and provides careful justification of valid conclusions from the fitted model.}
-Objective measures are used to \old{benchmark} \new{evaluate} model fit; when \old{these}\new{objective measures of the goodness-of-fit of a model} are combined with reproducibility, a framework emerges for continual improvement \new{to model based inference} when revisiting the data\old{ and models}.
-\new{This framework promotes scientific discovery and supports model-driven policy recommendations by simplifying the process of understanding and improving previous results.}
+\new{In this article, we conduct a re-analysis of the 2010-2019 cholera outbreak in Haiti.
+We use this case study to investigate current procedures for fitting mechanistic models to time series data, while identifying limitations of these methodologies and proposing remedies.}
+\old{Our analysis presents methodology for diagnosing how well a model describes observed data, and provides careful justification of valid conclusions from the fitted model.
+Objective measures are used to benchmark model fit; when these are combined with reproducibility, a framework emerges for continual improvement when revisiting the data and models.}
+\new{\editAuthorSum}
 Our data analysis workflow is supported by recent advances in algorithms, software and hardware, which facilitate statistical fitting of nonlinear stochastic dynamic models to observed incidence data.
@@
 Beyond this, the groups made autonomous decisions on what to include and exclude from their models.
-\old{The groups largely adhered to existing guidelines for creating models to inform policy [35, 36] and, d}\new{D}espite their autonomy, \new{the four independent teams} obtained a consensus that an extensive nationwide vaccination campaign would be necessary to eliminate cholera from Haiti.
-\new{They also provided a range for the forecasted number of cumulative cases from February 2019 to Februaruy 2024 based on model simulations, with median estimates from each model measured in hundreds of thousands.}
+\old{The groups largely adhered to existing guidelines for creating models to inform policy [35, 36] and, d}\new{D}espite their autonomy, \new{the four independent teams} obtained a consensus that an extensive nationwide vaccination campaign would be necessary to eliminate cholera from Haiti\new{, estimating that a large number of cumulative cholera cases would be observed in the absense of additional vaccination efforts (Figure~3 and 4 of \cite{lee20}).}
 \old{Their forecasts, however, were} \new{These forecasts are} inconsistent with the prolonged period with no confirmed cholera cases between February, 2019 and September, 2022 \cite{trevisin22}.
-Though cholera has recently reemerged in Haiti \cite{rubin22,PAHO23}, the inability to accurately forecast cholera incidence from 2019-2022\old{---while diligently adhering to existing scientific standards for infectious disease modeling---demonstrates that existing standards of disease modeling and forecasting may be insufficient.} \new{prompts us to consider retrospectively what may have been done differently in order to obtain more reliable conclusions.}
+Though cholera has recently reemerged in Haiti \cite{rubin22,PAHO23}, the inability to accurately forecast cholera incidence from 2019-2022\old{---while diligently adhering to existing scientific standards for infectious disease modeling---demonstrates that existing standards of disease modeling and forecasting may be insufficient.} \new{prompts us to consider retrospectively what may have been done differently in order to obtain more reliable conclusions, leading to recommendations for future studies.}
 
@@
 \new{Specifically, we focus on how to develop and test variations of the proposed models, as well as diagnosing the models once they have been assimilated to incidence reports.
-This includes a comprehensive tutorial on performing inference with Model~3, a highly non-linear, spatially explicit stochastic model, a challenging task that is possible due to recent methodological advancements.}
+This includes a comprehensive tutorial on performing inference with Model~3 (\nameref{S_mod3cal}), a highly non-linear, spatially explicit stochastic model, a challenging task that is possible due to recent methodological advancements.}
 We then use \new{the} improved model fits to project cholera incidence in Haiti under various vaccination scenarios considered by Lee et al.~\cite{lee20}.
@@
   & \multirow{2}{*}{\eqref{eq:model3:SA}}
-\\
+\\\hline
 Reporting rate
@@
   & (\nameref{S_meas})
-\\
+\\\hline
 \raggedright Observation variance
@@
 \end{equation}
-\new{where $\bar{\transmission} = 1 (wk)\textsuperscript{-1}$ is a dimensionality constant.}
+\new{where $\bar{\transmission} = 1$ (wk)\textsuperscript{-1} is a dimensionality constant.}
 The process noise $d\Gamma(t)/dt$ is multiplicative Gamma-distributed white noise, with infinitesimal variance parameter $\sigmaProc^2$.
@@
 The measurement model describes reported cholera cases at time point $n$ come from a negative binomial distribution, where only a fraction ($\reportRate$) of new weekly cases are reported.
-More details about the initialization model $f_{\bm{X}_0}\big(\bm{x}_0;\paramVec\big)$ and the \emph{measurement model} $f_{\bm{Y}_n|\bm{X}_{n}}\big(\bm{y}_n|\bm{x}_{n}\big)$ for Models~1--3 are provided a supplement text (\nameref{S_init} and \nameref{S_meas}).
+More details about the initialization model $f_{\bm{X}_0}\big(\bm{x}_0;\paramVec\big)$ and the measurement model $f_{\bm{Y}_n|\bm{X}_{n}}\big(\bm{y}_n|\bm{x}_{n}\big)$ for Models~1--3 are provided a supplement text (\nameref{S_init} and \nameref{S_meas}).
 
@@
 In reality, it is plausible that some bacteria or infected individuals were present in unit $u$ but went unreported.
-Therefore, for departments with zero case counts in week 1, we \old{consider estimating}\new{estimate} the number of infected individuals rather then treating this value as a constant (see supplement Sec.~S5 for more details).
-Finally, reported cholera cases are modeled using a negative binomial measurement model with mean equal to a fraction ($\reportRate$) of individuals in each unit who develop symptoms and seek healthcare
-(see Sec.~S3.3 of the supplement).
+Therefore, for departments with zero case counts in week 1, we \old{consider estimating}\new{estimate} the number of infected individuals rather then treating this value as a constant (\nameref{S_init}).
+Finally, reported cholera cases are modeled using a negative binomial \old{measurement model} \new{distribution} with mean equal to a fraction ($\reportRate$) of individuals in each unit who develop symptoms and seek healthcare\new{, and with over-dispersion parameter $\obsOverdispersion$} (\nameref{S_meas}).
 
@@
 Each model is indexed by a parameter vector, $\paramVec$, and different values of $\paramVec$ can result in qualitative differences in the predicted behavior of the system.
-Therefore, the choice of $\paramVec$ used to make inference about the system can greatly affect model-based conclusions.
+Therefore, the choice of $\paramVec$ used to make inference about the system can greatly affect model-based conclusions  \new{\cite{saltelli20}}.
 Elements of $\paramVec$ can be fixed at a constant value based on scientific understanding of the system, but parameters can also be calibrated to data by maximizing a measure of congruency between the observed data and the model's mechanistic structure.
@@
 
-Model~1 was developed with the intent of providing a description of cholera dynamics that is scientifically plausible rather than one that is statistically convenient.
-This results in the inability to obtain a closed form expression of the non-linear joint model density---described in Eq.~\myeqref{eq:jointLik}.
+\old{Model~1 was developed with the intent of providing a description of cholera dynamics that is scientifically plausible rather than one that is statistically convenient.}
+\new{Model~1 is a highly nonlinear over-dispersed stochastic dynamic model, favoring a scientifically plausible description of cholera dynamics rather than one that is statistically convenient \cite{he10}.}
+This results in the inability to obtain a closed form expression of the \old{non-linear} joint model density---described in Eq.~\myeqref{eq:jointLik}.
 Therefore in order to perform likelihood based inference on this model, we are restricted to use parameter estimation techniques that have the \emph{plug-and-play} property, which is that the fitting procedure only requires the ability to simulate the latent process rather than evaluating transition densities \cite{breto09,he10};
@@
 Plug-and-play algorithms include Bayesian approaches like ABC and PMCMC \cite{toni09,andrieu10}, but here we use algorithms that enable maximum likelihood estimation.
-To our knowledge, the only plug-and-play methods that can be used to maximize the likelihood for POMP models of this complexity are iterated filtering algorithms \cite{ionides15}, which modify the well-known \emph{particle filter} \cite{arulampalam02}.
+To our knowledge, the only plug-and-play methods that \old{can be} \new{have been effectively} used to maximize the likelihood for \new{arbitrary nonlinear} POMP models \old{of this complexity} are iterated filtering algorithms \cite{ionides15}, which modify the well-known \emph{particle filter} \cite{arulampalam02}.
 
@@
 This algorithm can also be used to accurately approximate the log-likelihood of a POMP model, defined as the integral in Eq.~\myeqref{eq:likedef}.
-Iterated filtering---algorithms such as IF2 \cite{ionides15}---extend the particle filter by performing a random walk for each parameter and particle;
+Iterated filtering algorithms, such as IF2 \cite{ionides15}, extend the particle filter by performing a random walk for each parameter and particle;
 these perturbations are carried out iteratively over multiple filtering operations, using the collection of parameters from the previous filtering pass as the parameter initialization for the next iteration, and decreasing the random walk variance at each step.
@@
 The reported results for Model~1 in the remainder of this article were obtained with the inclusion of the parameter $\transmissionTrend$.
-\new{This example demonstrates the type of model variation that should be considered in the course of a data analysis: the model variation---the addition of a trend in transmission rate---has a plausible scientific justification, and is easily testable using likelihood based inference methods and observed data.}
+\new{The inclusion of a trend in transmission rate demonstrates a class of model variation that can be highly beneficial to consider: the model variation has a plausible scientific justification, and is easily testable using likelihood based methods.}
 
@@
 
-We implemented Model~1 using the \code{pomp} package \cite{king16}, relying heavily on the \code{pomp} source code provided by Lee et al.~\cite{lee20}.
+We implemented Model~1 using the \code{pomp} package \cite{king16}, relying heavily on the \old{\code{pomp}} source code provided by Lee et al.~\cite{lee20}.
 Both analyses used the \code{mif2} implementation of the IF2 algorithm to estimate $\paramVec$ by maximum likelihood.
@@
 Many Monte Carlo replications are then required to quantify and further reduce the error.
-The large increase in the log-likelihood for Model~1 (see Table~\ref{tab:likes}) can primarily be attributed to increasing the computational effort in fitting the model.
+The large increase in the log-likelihood for Model~1 (Table~\ref{tab:likes}) can primarily be attributed to increasing the computational effort \old{in fitting} \new{used to calibrate} the model.
 This result highlights the importance of carefully determining the necessary computational effort needed to maximize model likelihoods and acting accordingly.
-% \TODO{Possibly add more citations to the diagnostics part, maybe li23, laneri10 (Forcing Versus Feedback: Epidemic Malaria and Monsoon Rains in Northwest India), pons-salort18}
-\new{In this case study, this was done by performing standard diagnostics for the IF2 and particle filter algorithms \cite{king16}.
-Given the considerable computational costs of simulation-based algorithms, we find it useful to perform an initial assessment using hyperparameter values that enable relatively quick calculations.
+\new{In this case study, this was done by performing standard diagnostics for the IF2 and particle filter algorithms\cite{king16,li23,pons-salort18,laneri10}.
+Given the considerable computational costs of simulation-based algorithms, we find it useful to perform an initial assessment using hyperparameter values---such as the number of particles, filtering iterations, and replicates based on different parameter initializations---that enable relatively quick calculations.
 The insights obtained from this preliminary analysis help in accurately determining the amount of computation that is required to achieve reliable outcomes.}
@@
 {\figTitle Simulations from Model~1 compared to reported cholera cases.}
-The black curve is observed data, the blue curve is median of 500 simulations from the fitted model, and the vertical dashed line represents break-point when parameters are refit.
+The black curve is observed data, the blue curve is median of 500 simulations from \old{the fitted model} \new{initial conditions using estimated parameters}, and the vertical dashed line represents break-point when parameters are refit.
 }
@@
 and can be justified by asymptotic considerations in a large-population limit \cite{dadlani2020,ndii17}.
-Because the process model of Model~2 is deterministic, maximum likelihood estimation reduces to a least squares calculation when combined with a Gaussian measurement model (see Sec.~S3.2 of the supplement material for details).
+Because the process model of Model~2 is deterministic, maximum likelihood estimation reduces to a least squares calculation when combined with a Gaussian measurement model \old{(see Sec.~S3.2 of the supplement material for details)} \new{(\nameref{S_meas})}.
 Lee et al.~\cite{lee20} fit two versions of Model~2 based on a presupposed change in cholera transmission from a epidemic phase to endemic phase that occurred in March, 2014.
-The inclusion of a change-point in model states and parameters increases the flexibility of the model and hence the ability to fit the observed data.
-The increase in model flexibility, however, results in hidden states that are inconsistent between model phases.
-The inclusion of a model break-point by Lee et al.~\cite{lee20} is perhaps due to a challenging feature of fitting a deterministic model via least squares: discrepancies between model trajectories and observed case counts in highly infectious periods of a disease outbreak will result in greater penalty than the discrepancies between model trajectories and observed case counts in times of relatively low infectiousness, resulting in a bias towards accurately describing periods of high infectiousness.
-This issue is particularly troublesome for modeling cholera dynamics in Haiti: the inability to accurately fit times of low infectiousness may result in poor model forecasts, as few cases of cholera were observed in the last few years of the epidemic.
+The inclusion of a change-point in model states and parameters increased the flexibility of the model and hence the ability to fit the observed data.
+The increase in model flexibility, however, resulted in hidden states that were inconsistent between model phases.
+The inclusion of a model break-point by Lee et al.~\cite{lee20} is perhaps due to a challenging feature of fitting a deterministic model via least squares: discrepancies between model trajectories and observed case counts in highly infectious periods of a disease outbreak will result in greater penalty than the discrepancies between model trajectories and observed case counts in times of relatively low infectiousness\new{.
+This results} \old{resulting} in a bias towards accurately describing periods of high infectiousness.
+This \old{issue} \new{bias} is particularly troublesome for modeling cholera dynamics in Haiti: the inability to accurately fit times of low infectiousness may result in poor model forecasts, as few cases of cholera were observed in the last few years of the epidemic.
 
@@
 We implemented this model using the \code{spatPomp} \code{R} package \cite{asfaw23arxiv}.
-The model was then fit using the subplex algorithm, implemented in the  \code{subplex} package \cite{king2020Subplex}.
+The model was then fit using the subplex algorithm \old{, implemented in the  \code{subplex} package} \cite{king2020Subplex}.
 A comparison of the trajectory of the fitted model to the data is given in Fig.~\ref{fig:mod2Traj}.
@@
 Projections under the various vaccination scenarios, which are discussed in detail in the  {\bf Forecasts} subsection are also included.
-The gray ribbons represent a $95\%$ interval obtained from the log-normal measurement model.
-To avoid over-plotting, measurement variance is only plotted for the V0 vaccination scenario.}
+\new{The gray ribbons represent a $95\%$ interval obtained from the log-normal measurement model.
+To avoid over-plotting, measurement variance is only plotted for the V0 vaccination scenario.}}
 \end{figure}
@@
 For our analysis, we calibrate the parameters of the spatially coupled version of Model~3 using the \old{iterated block particle filter (IBPF)} \new{IBPF} algorithm \cite{ionides22}.
-This algorithm extends the work of Ning and Ionides \cite{ning21ibpf}, who provided theoretic justification for the version of the algorithm that only estimates unit-specific parameters.
+This algorithm extends the work of Ning and Ionides \cite{ning23}, who provided theoretic justification for the version of the algorithm that only estimates unit-specific parameters.
 The IBPF algorithm enables us to directly estimate the parameters of models describing high-dimensional partially-observed nonlinear dynamic systems via likelihood maximization.
@@
 
-The goal of parameter calibration---whether done using Bayesian or frequentist methods---is to find the best description of the observed data under the assumptions of the model.
+The goal of parameter calibration---whether done using Bayesian or frequentist methods---is to find the best description of the observed data \old{under the assumptions} \new{in the context} of the model.
 Obtaining the best fitting set of parameters for a given model does not, however, guarantee that the model provides an accurate representation of the system under investigation.
@@
 While visual comparisons can be informative, they provide only a weak and informal measure of the goodness-of-fit of a model.
-The study by Lee et al.~\cite{lee20} provides an example of this: their models and parameter estimates resulted in simulations that visually resembled the observed data, yet resulted in model likelihoods that were considerably smaller than likelihoods that can be achieved via the likelihood based optimization techniques that were used (see Table~\ref{tab:likes}).
+The study by Lee et al.~\cite{lee20} provides an example of this: their models and parameter estimates resulted in simulations that visually resembled the observed data, yet resulted in model likelihoods that were considerably smaller than likelihoods that can be achieved \old{via the likelihood based optimization techniques that were used} (see Table~\ref{tab:likes}).
 Alternative forms of model validation should therefore be used in conjunction with visual comparisons of simulations to observed data.
@@
 Standard statistical models, interpreted as associative models without requiring any mechanistic interpretation of their parameters, provide suitable benchmarks.
-Examples include linear regression, auto regressive moving average time series models, or even independent and identically distributed measurements.
+Examples include linear regression, auto regressive moving average \new{(ARMA)} time series models, or even independent and identically distributed measurements.
 Benchmarks enable us to evaluate the goodness of fit that can be expected of a suitable mechanistic model.
@@
 the theory and practice of ARMA models is well developed, and these linear models are appropriate on a log scale due to the exponential growth and decay characteristic of biological dynamics.
-We chose to use the auto regressive negative binomial model, however, because the large number of weeks with zero recorded cholera cases in department level data makes a benchmark based on a continuous distribution problematic.
+We \old{chose to} use the auto regressive negative binomial model, however, because the large number of weeks with zero recorded cholera cases in department level data makes a benchmark based on a continuous distribution problematic.
 Log-likelihoods and AIC values of Models~1--3 and of their respective \old{ARMA} benchmark models are provided in Table~\ref{tab:likes}.
 \new{Models that are fit to the same datasets can be directly compared using AIC values, making it a useful tool to compare to benchmark models.
-Though Models~2 and 3 are both fit to department level incidence reports, their AIC values are not directly comparable due to the way Model~3 initializes latent states, which is described in detail in Sec.
-S5 of the supplement.}
+Though Models~2 and 3 are both fit to department level incidence reports, their AIC values are not directly comparable due to the way Model~3 initializes latent states (\nameref{S_init}).}
 
@@
 Comparing model log-likelihoods to a suitable benchmark may not be sufficient to identify all the strengths and weaknesses of a given model.
-Additional techniques include the inspection of conditional log-likelihoods of each observation given the previous observations in order to understand how well the model describes each data point (Sec.~S5).
+Additional techniques include the inspection of conditional log-likelihoods of each observation given the previous observations in order to understand how well the model describes each data point (\nameref{S_mod3cal}).
 Other tools include plotting the effective sample size of each observation \cite{liu01}; plotting the values of the hidden states from simulations (\nameref{S_mod3cal});
@@
 This is consistent with previous studies \old{finding that rainfall played an important role in cholera transmission in Haiti [50, 51]} \new{that incorporated rainfall as an important part of their mechanistic model or otherwise argue that rainfall is an important driver of cholera dynamics in Haiti \cite{hulland19,kirpich17,lee20,moise20,pasetto18,kirpich16,eisenberg13,rinaldo12,mavian20}.}
-The estimated seasonality also features an increased transmission rate during the fall, which was noticed at an earlier stage of the epidemic \cite{rinaldo12}; the high transmission rate in the fall may also be a result of the increase transmission that occurred in the fall of 2016, when hurricane Matthew struck Haiti \cite{ferreirai16}.
+The estimated seasonality also features an increased transmission rate during the fall, which was noticed at an earlier stage of the epidemic \cite{rinaldo12}\new{.
+T}he high transmission rate in the fall may \old{also} be a result of the increase transmission that occurred in the fall of 2016, when hurricane Matthew struck Haiti \cite{ferreirai16}.
 
@@
 Both of these approaches share the similarity that parameters are chosen for the forecast approximately in proportion to their corresponding value of the likelihood function, $f_{\bm{Y}_{1:N}}(\bm{y}_{1:N}^*; \paramVec)$.
-In this analysis, we do not construct forecasts accounting for parameter uncertainty as our focus is on the estimation and diagnosis of mechanistic models.
-Furthermore, we use the projections from a single point estimate to highlight the deficiency of deterministic models that the only variability in model projections is a result of parameter uncertainty, which can lead to over-confidence in forecasts \cite{king15}.
+In this analysis, we do not construct forecasts accounting for parameter uncertainty as our focus is on the estimation and diagnosis of mechanistic models\new{, rather than providing forecasts intended to influence policy.}
+Furthermore, we use the projections from a single point estimate to highlight the deficiency of deterministic models that the only variability in model projections is a result of parameter \new{and measurement} uncertainty, which can lead to over-confidence in forecasts \cite{king15}.
 
@@
 
-\new{We presented methodology consistent with existing guidelines but going beyond standard practice.
-In particular, we showed the value of comparing the likelihood of fitted mechanistic models versus non-mechanistic benchmarks.
-Further, we used likelihood-based diagnostic methods to identify specific limitations of proposed models.
-We used likelihood profile methods to assess parameter identifiability and to construct confidence intervals for parameter estimates.
+\new{We presented methodology consistent with existing guidelines \cite{behrend20} but going beyond standard practice.
+In particular, we showed the value of comparing the likelihood of fitted mechanistic models versus non-mechanistic benchmarks, a practice that has been previously advocated for \cite{he10} but was not done by any of the studies in our literature review.
+These comparisons, along with other likelihood based diagnostics, help identify specific limitations of proposed models.
+Diagnostic tools include likelihood profile methods, which help to assess parameter identifiability and enable the construction of confidence intervals for parameter estimates \cite{ionides17, simpson23}.
 % We assessed the accuracy of numerical approximations involved in our calculations.
-When reaching conclusions, we identify potential consequences of confounded variables and model misspecification.
+When reaching conclusions, it is important to consider potential consequences of confounded variables and model misspecification.
 }
@@
 
-\paragraph*{S7 Text}
+\paragraph*{\new{S7 Text}}
 \label{S_CI}
-{\bf Confidence Intervals} Tables and figures describing 95\% confidence intervals for estimated parameters.
+{\bf \new{Confidence Intervals}} \new{Tables and figures describing 95\% confidence intervals for estimated parameters.}
 

### 2024-03-04 ms.Rnw
Removed all calls to 'old'

@@
 This necessitates addressing the tradeoff between biological fidelity and model simplicity, and the reality of misspecification for models at all levels of complexity.
-\old{As a case study, we consider the 2010-2019 cholera epidemic in Haiti.
-We study three dynamic models developed by expert teams to advise on vaccination policies.
-We assess previous methods used for fitting and evaluating these models, and we develop data analysis strategies leading to improved statistical fit.}
-\new{We assess current methodological approaches to these issues via a case study of the 2010-2019 cholera epidemic in Haiti.
+We assess current methodological approaches to these issues via a case study of the 2010-2019 cholera epidemic in Haiti.
 We consider three dynamic models developed by expert teams to advise on vaccination policies.
-We evaluate previous methods used for fitting these models, and we demonstrate modified data analysis strategies leading to improved statistical fit.}
-Specifically, we present approaches for diagnosing model misspecification \old{, development of alternative models, and computational improvements in optimization, in the context of likelihood-based inference on nonlinear dynamic systems.} \new{and the consequent development of improved models.
-Additionally, we demonstrate the utility of recent advances in likelihood maximization for high-dimensional nonlinear dynamic models, enabling likelihood-based inference for high-dimensional models for spatiotemporal incidence data.}
+We evaluate previous methods used for fitting these models, and we demonstrate modified data analysis strategies leading to improved statistical fit.
+Specifically, we present approaches for diagnosing model misspecification and the consequent development of improved models.
+Additionally, we demonstrate the utility of recent advances in likelihood maximization for high-dimensional nonlinear dynamic models, enabling likelihood-based inference for high-dimensional models for spatiotemporal incidence data.
 Our workflow is reproducible and extendable, facilitating future investigations of this disease system.
@@
 Evaluation of success at these goals is necessary in order for a model to provide a reliable tool for guiding evidence-based public policy interventions.
-\old{In this article, we discuss common modeling decisions that may not provide an adequate statistical explanation of the data, and we show that attention to statistical fit can lead to more accurate policy evaluations}.
-\new{In this article, we conduct a re-analysis of the 2010 cholera outbreak in Haiti.
-We use this case study to investigate current procedures for fitting mathematical models to time series data, while identifying limitations of these methodologies and proposing remedies.}
-\new{We present} \old{Our analysis presents} methodology for diagnosing how well a model describes observed data\new{.}\old{, and provides careful justification of valid conclusions from the fitted model.}
-Objective measures are used to \old{benchmark} \new{evaluate} model fit; when \old{these}\new{objective measures of the goodness-of-fit of a model} are combined with reproducibility, a framework emerges for continual improvement \new{to model based inference} when revisiting the data\old{ and models}.
-\new{This framework promotes scientific discovery and supports model-driven policy recommendations by simplifying the process of understanding and improving previous results.}
+In this article, we conduct a re-analysis of the 2010-2019 cholera outbreak in Haiti.
+We use this case study to investigate current procedures for fitting mechanistic models to time series data, while identifying limitations of these methodologies and proposing remedies.
+\editAuthorSum
 Our data analysis workflow is supported by recent advances in algorithms, software and hardware, which facilitate statistical fitting of nonlinear stochastic dynamic models to observed incidence data.
@@
 
-This case study investigates the use of dynamic models and spatiotemporal data to inform \old{a policy decision} \new{public health policy} in the context of the cholera outbreak in Haiti, which started in 2010.
-\old{We build on a multi-group modeling exercise by Lee et al.
-[1] in which four expert modeling teams developed models to the same dataset with the goal of comparing conclusions on the feasibility of eliminating cholera by a vaccination campaign}.
-\old{The data that were used for this study, and that is the subject of our reanalysis, is displayed in Fig.~[1].}
-\new{Various dynamic models were developed to study this outbreak: searching PubMed with keywords ``Haiti, cholera, model'' we obtained 22 studies that utilized deterministic mechanistic dynamic models \cite{lee20,tuite11,andrews11,botelho21,fitzgibbon20,eisenberg13,rinaldo12,chao11,date11,lin19,abrams13,collins21,akman15,trevisin22,mavian20,collins14,kelly16,capone15,leung22,mari15,gatto12,kuhn14}
+This case study investigates the use of dynamic models and spatiotemporal data to inform public health policy in the context of the cholera outbreak in Haiti, which started in 2010.
+Various dynamic models were developed to study this outbreak: searching PubMed with keywords ``Haiti, cholera, model'' we obtained 22 studies that utilized deterministic mechanistic dynamic models \cite{lee20,tuite11,andrews11,botelho21,fitzgibbon20,eisenberg13,rinaldo12,chao11,date11,lin19,abrams13,collins21,akman15,trevisin22,mavian20,collins14,kelly16,capone15,leung22,mari15,gatto12,kuhn14}
 and 11 studies that used stochastic models \cite{kirpich16,lee20,pasetto18,mukandavire13,kirpich17,lewnard16,kunkel17,mukandavire15,sallah17,azman12,azman15}.
 Incidence data on the outbreak are available at various spatial scales, motivating 17 studies in our literature review to consider spatially explicit dynamic models \cite{lee20,tuite11,pasetto18,fitzgibbon20,eisenberg13,rinaldo12,chao11,abrams13,trevisin22,sallah17,collins14,kelly16,azman12,leung22,kuhn14,mari15,gatto12}.
-Here we focus on a multi-group modeling exercise by Lee et al.~\cite{lee20} in which four expert modeling teams developed models to the same dataset with the goal of comparing conclusions on the feasibility of eliminating cholera by a vaccination campaign.}
+Here we focus on a multi-group modeling exercise by Lee et al.~\cite{lee20} in which four expert modeling teams developed models to the same dataset with the goal of comparing conclusions on the feasibility of eliminating cholera by a vaccination campaign.
 Model~1 is stochastic and describes cholera at the national level;
@@
 Beyond this, the groups made autonomous decisions on what to include and exclude from their models.
-\old{The groups largely adhered to existing guidelines for creating models to inform policy [35, 36] and, d}\new{D}espite their autonomy, \new{the four independent teams} obtained a consensus that an extensive nationwide vaccination campaign would be necessary to eliminate cholera from Haiti.
-\new{They also provided a range for the forecasted number of cumulative cases from February 2019 to Februaruy 2024 based on model simulations, with median estimates from each model measured in hundreds of thousands.}
-\old{Their forecasts, however, were} \new{These forecasts are} inconsistent with the prolonged period with no confirmed cholera cases between February, 2019 and September, 2022 \cite{trevisin22}.
-Though cholera has recently reemerged in Haiti \cite{rubin22,PAHO23}, the inability to accurately forecast cholera incidence from 2019-2022\old{---while diligently adhering to existing scientific standards for infectious disease modeling---demonstrates that existing standards of disease modeling and forecasting may be insufficient.} \new{prompts us to consider retrospectively what may have been done differently in order to obtain more reliable conclusions.}
+Despite their autonomy, the four independent teams obtained a consensus that an extensive nationwide vaccination campaign would be necessary to eliminate cholera from Haiti, estimating that a large number of cumulative cholera cases would be observed in the absense of additional vaccination efforts (Figure~3 and 4 of \cite{lee20}).
+These forecasts are inconsistent with the prolonged period with no confirmed cholera cases between February, 2019 and September, 2022 \cite{trevisin22}.
+Though cholera has recently reemerged in Haiti \cite{rubin22,PAHO23}, the inability to accurately forecast cholera incidence from 2019-2022 prompts us to consider retrospectively what may have been done differently in order to obtain more reliable conclusions, leading to recommendations for future studies.
 
@@
 
-%% IT SEEMS THIS PARAGRAPH IS DEPRECATED IN FAVOR OF CITING THESE REFERENCES WHERE APPROPRIATE
-\old{While this work is focused on the models proposed by Lee et al.~[1], our suggestions have broader relevance.
-To investigate the extent to which [1] is typical of the substantial body of work on Haiti cholera dynamics, we performed a literature review by searching PubMed with keywords: Haiti, cholera, model.
-The search resulted in 66 papers, of which 32 used dynamic models to describe the cholera epidemic in Haiti.
-The models make various choices on the dichotomies considered by Lee et al.~[1]; deterministic versus stochastic; compartment model versus agent-based; aggregated versus spatially explicit.}
-
 \new{\editModelingGuidance}
@@
 
-\new{In this article, we refer to} models that focus on learning relationships between variables in a dataset \new{as} \old{are called} {\it associative}, whereas models that incorporate a known scientific property of the system \old{are called} \new{we call} {\it causal} or {\it mechanistic}.
+In this article, we refer to models that focus on learning relationships between variables in a dataset as {\it associative}, whereas models that incorporate a known scientific property of the system we call {\it causal} or {\it mechanistic}.
 The danger in using forecasting techniques which rely on associative models to predict the consequence of interventions is called the Lucas critique in an econometric context.
@@
 The observable random variables $\bm{Y}_{1:N}$ are assumed to be conditionally independent given $\bm{X}_{0:N}$.
-Together, with the density for the initial value of the latent state $\bm{X}_{0} = \bm{X}(t_0)$, each model \new{defines} \old{assumes} a joint density $f_{\bm{X}_{0:N}, \bm{Y}_{1:N}}\big(\bm{x}_{0:N}, \bm{y}_{1:N};\paramVec\big)$, where $\paramVec$ is a parameter vector that indexes the model.
+Together, with the density for the initial value of the latent state $\bm{X}_{0} = \bm{X}(t_0)$, each model defines a joint density $f_{\bm{X}_{0:N}, \bm{Y}_{1:N}}\big(\bm{x}_{0:N}, \bm{y}_{1:N};\paramVec\big)$, where $\paramVec$ is a parameter vector that indexes the model.
 The observed data $\bm{y}_{1:N}^*$, along with the unobserved true value of the latent state, are modeled as a realization of this joint distribution.
@@
 For $\vaccCounter\in\seq{1}{\vaccClass}$, the vaccination cohort $\vaccCounter$ is assigned a time $\tau_{\vaccCounter}$, and we take $\symptomFrac_\vaccCounter(t) = c \, \vaccineEfficacy^*(t-\tau_{\vaccCounter})$
-where $\vaccineEfficacy^*(t)$ is efficacy at time $t$ since vaccination for adults, \old{taken from [4], Table~S4,} \new{a step-function represented in Table~S4 of \cite{lee20}}, and $c=\big(1-(1-0.4688)\times 0.11\big)$ is a correction to allow for reduced efficacy in the 11\% of the population aged under 5 years.
+where $\vaccineEfficacy^*(t)$ is efficacy at time $t$ since vaccination for adults, a step-function represented in Table~S4 of \cite{lee20}, and $c=\big(1-(1-0.4688)\times 0.11\big)$ is a correction to allow for reduced efficacy in the 11\% of the population aged under 5 years.
 Single and double vaccine doses were modeled by changing the waning of protection; protection was modeled as equal between single and double dose until 52 weeks after vaccination, at which point the single dose becomes ineffective.
@@
 where $\mathrm{Pop}_u$ is the mean population for department $u$,
-$D_{u\varv}$ is a distance measure estimating average road distance between randomly chosen members of each population, and $v_{\mathrm{rate}}= 10^{-12} \,\mbox{km$^2$yr$^{-1}$}$ \old{was treated as a fixed constant.}\new{was fixed at the value used in \cite{lee20}.}
+$D_{u\varv}$ is a distance measure estimating average road distance between randomly chosen members of each population, and $v_{\mathrm{rate}}= 10^{-12} \,\mbox{km$^2$yr$^{-1}$}$ was fixed at the value used in \cite{lee20}.
 In Eq.~\myeqref{model2:water_transport}, $T^W_{u\varv}$ is a measure of river flow between departments.
 The unit of $W_u(t)$ is cells per ml, with dose response modeled via a saturation constant of $\Wsat$ in Eq.~\myeqref{model2:lambda}.
-In Eq.~\myeqref{model2:mu_SE}, $\vaccineEfficacy_\vaccCounter$ denotes the vaccine efficacy for each vaccination campaign $\vaccCounter \in \vaccClass$, with \old{$\vaccineEfficacy_0 = 1$, $\vaccineEfficacy_1 = 1 - 0.429 \childReduce$, $\vaccineEfficacy_2 = 1 - 0.519 \childReduce$, $\vaccineEfficacy_3 = 1 - 0.429$, and $\vaccineEfficacy_4 = 1 - 0.519$} \new{$\vaccineEfficacy_0 = 0$, $\vaccineEfficacy_1 = 0.429 \childReduce$, $\vaccineEfficacy_2 = 0.519 \childReduce$, $\vaccineEfficacy_3 = 0.429$, and $\vaccineEfficacy_4 = 0.519$}
+In Eq.~\myeqref{model2:mu_SE}, $\vaccineEfficacy_\vaccCounter$ denotes the vaccine efficacy for each vaccination campaign $\vaccCounter \in \vaccClass$, with $\vaccineEfficacy_0 = 0$, $\vaccineEfficacy_1 = 0.429 \childReduce$, $\vaccineEfficacy_2 = 0.519 \childReduce$, $\vaccineEfficacy_3 = 0.429$, and $\vaccineEfficacy_4 = 0.519$
 Here, $\childReduce = 0.4688$ represents the reduced efficacy of the vaccination for children under the age of five years, and the values $0.429$ and $0.519$ are the median effectiveness of one and two doses over their effective period respectively, according to Table S4 in the supplement material of Lee et al.~\cite{lee20}.
@@
 Reported cases are described using a log-normal distribution, with the log-scale mean equal to the reporting rate $\reportRate$ times the number of newly infected individuals.
-\old{See the section describing how Model~2 is fit to data and Sec.~S3.2 of the supplement for more details.}
-\new{See the supplement material on model initializations for more details (\nameref{S_init})}
+See the supplement material on model initializations for more details (\nameref{S_init}).
 
@@
 In reality, it is plausible that some bacteria or infected individuals were present in unit $u$ but went unreported.
-Therefore, for departments with zero case counts in week 1, we \old{consider estimating}\new{estimate} the number of infected individuals rather then treating this value as a constant (see supplement Sec.~S5 for more details).
-Finally, reported cholera cases are modeled using a negative binomial measurement model with mean equal to a fraction ($\reportRate$) of individuals in each unit who develop symptoms and seek healthcare
-(see Sec.~S3.3 of the supplement).
+Therefore, for departments with zero case counts in week 1, we estimate the number of infected individuals rather then treating this value as a constant (\nameref{S_init}).
+Finally, reported cholera cases are modeled using a negative binomial distribution with mean equal to a fraction ($\reportRate$) of individuals in each unit who develop symptoms and seek healthcare\new{, and with over-dispersion parameter $\obsOverdispersion$} (\nameref{S_meas}).
 
@@
 
-Each of the three models considered in this study describes cholera dynamics as a partially observed Markov process (POMP) \cite{king16}, with the understanding that the deterministic Model~2 is a \old{degenerate case of a stochastic model} \new{special case of a Markov processes solving a stochastic differential equation in the limit as the noise parameter goes to zero.}
+Each of the three models considered in this study describes cholera dynamics as a partially observed Markov process (POMP) \cite{king16}, with the understanding that the deterministic Model~2 is a special case of a Markov processes solving a stochastic differential equation in the limit as the noise parameter goes to zero.
 Each model is indexed by a parameter vector, $\paramVec$, and different values of $\paramVec$ can result in qualitative differences in the predicted behavior of the system.
-Therefore, the choice of $\paramVec$ used to make inference about the system can greatly affect model-based conclusions.
+Therefore, the choice of $\paramVec$ used to make inference about the system can greatly affect model-based conclusions  \new{\cite{saltelli20}}.
 Elements of $\paramVec$ can be fixed at a constant value based on scientific understanding of the system, but parameters can also be calibrated to data by maximizing a measure of congruency between the observed data and the model's mechanistic structure.
-Calibrating model parameters to observed data does not guarantee that the resulting model successfully approximates real-world mechanisms, since the model description of the dynamic system may be incorrect and \old{do} \new{does} not change as the model is calibrated to data.
+Calibrating model parameters to observed data does not guarantee that the resulting model successfully approximates real-world mechanisms, since the model description of the dynamic system may be incorrect and does not change as the model is calibrated to data.
 However, the congruency between the model and observed data serves as a proxy for the congruency between the model and the true underlying dynamic system.
@@
 
-In this article we \old{calibrate} \new{follow \cite{lee20} by calibrating} the parameters of each of our models \old{by maximizing the likelihood} \new{using maximum likelihood}, as described in Eq.~\myeqref{eq:likedef}.
+In this article we follow \cite{lee20} by calibrating the parameters of each of our models using maximum likelihood, as described in Eq.~\myeqref{eq:likedef}.
 The likelihood for each of the fitted models---and the corresponding AIC values for model comparisons that include an adjustment for the number of calibrated parameters---is provided in Table~\ref{tab:likes}.
@@
 \begin{flushleft}
-Values in parentheses are corresponding values \new{obtained} using \new{the models of \cite{lee20}} \old{Lee et al.~[4] parameter estimates}.
-\textsuperscript{1}The reported likelihood is an upper bound of the likelihood of the \old{Lee et al.~[4]} model \new{in \cite{lee20}}.
-\old{Lee et al.~[4] fit} \textsuperscript{2}\new{In \cite{lee20},} Model~3 \new{was fit} to a subset of the data (March 2014 onward, excluding data from Ouest in 2015-2016).
+Values in parentheses are corresponding values obtained using the models of \cite{lee20}.
+\textsuperscript{1}The reported likelihood is an upper bound of the likelihood of the model in \cite{lee20}.
+\textsuperscript{2}In \cite{lee20}, Model~3 was fit to a subset of the data (March 2014 onward, excluding data from Ouest in 2015-2016).
 On this subset, their model has a likelihood of $\Sexpr{myround(lee3_results$sub_ll_eval, 1)}$.
 On this same subset, our model has a likelihood of $\Sexpr{myround(mod3_eval_res$subset_ll['est'], 1)}$.
-Details of estimating the likelihood of the \old{Lee et al.~[4]} models \new{used in \cite{lee20} are} \old{is} provided in \new{the supplement} (\nameref{S_lee20}).
-\old{See Sec.~S6 of the supplement material for more details on estimating the likelihood of the Lee et al.~[4] models.}
+Details of estimating the likelihood of the models used in \cite{lee20} are provided in the supplement (\nameref{S_lee20}).
 \end{flushleft}
@@
 
-Model~1 was developed with the intent of providing a description of cholera dynamics that is scientifically plausible rather than one that is statistically convenient.
-This results in the inability to obtain a closed form expression of the non-linear joint model density---described in Eq.~\myeqref{eq:jointLik}.
+Model~1 is a highly nonlinear over-dispersed stochastic dynamic model, favoring a scientifically plausible description of cholera dynamics rather than one that is statistically convenient \cite{he10}.
+This results in the inability to obtain a closed form expression of the joint model density---described in Eq.~\myeqref{eq:jointLik}.
 Therefore in order to perform likelihood based inference on this model, we are restricted to use parameter estimation techniques that have the \emph{plug-and-play} property, which is that the fitting procedure only requires the ability to simulate the latent process rather than evaluating transition densities \cite{breto09,he10};
@@
 Plug-and-play algorithms include Bayesian approaches like ABC and PMCMC \cite{toni09,andrieu10}, but here we use algorithms that enable maximum likelihood estimation.
-To our knowledge, the only plug-and-play methods that can be used to maximize the likelihood for POMP models of this complexity are iterated filtering algorithms \cite{ionides15}, which modify the well-known \emph{particle filter} \cite{arulampalam02}.
+To our knowledge, the only plug-and-play methods that have been effectively used to maximize the likelihood for arbitrary nonlinear POMP models are iterated filtering algorithms \cite{ionides15}, which modify the well-known \emph{particle filter} \cite{arulampalam02}.
 
@@
 where $\bar{t} = \frac{t - (t_N + t_0) / 2}{t_N - (t_N + t_0) / 2}$, so that $\bar{t} \in [-1, 1]$.
-The proposal of a \old{linear} trend in transmission is a result of observing an apparent decrease in reported cholera infections from 2012-2019 in Fig.~\ref{fig:CholeraData}.
+The proposal of a trend in transmission is a result of observing an apparent decrease in reported cholera infections from 2012-2019 in Fig.~\ref{fig:CholeraData}.
 While several factors may contribute to this decrease, one explanation is that case-area targeted interventions (CATIs), which included education sessions, increased monitoring, household decontamination, soap distribution, and water chlorination in infected areas \cite{rebaudet19CATI}, may have substantially reduced cholera transmission over time \cite{rebaudet21}.
 
-We perform a statistical test to determine whether or not the data indicate the presence of a \old{linear} trend in transmissibility.
+We perform a statistical test to determine whether or not the data indicate the presence of a trend in transmissibility.
 To do this, we perform a profile-likelihood search on the parameter $\transmissionTrend$ and obtain a $95\%$ confidence interval via a Monte Carlo Adjusted Profile (MCAP) \cite{ionides17}.
@@
 As for any analysis of observational data, we must be alert to the possibility of confounding.
-For a covariate, this shows up in a similar way to regression analysis: the covariate under investigation could be a proxy for some other unmodeled \new{phenomenon} or unmeasured covariate.
-\old{For a mechanism, the model feature could in principle explain the data by helping to account for some different unmodeled phenomenon.
-The statistical evidence of a trend in transmission rate in this model could be explained by any trending variable (such as hygiene improvements, or changes in population behavior), resulting in confounding from collinear covariates.
-Alternatively, the trend could be attributed to a decreasing reporting rate rather than decreasing transmission rate, resulting in confounded mechanisms.
-The robust statistical conclusion is that a model which allows for change fits better than one which does not---we argue that a decreasing transmission rate is a plausible way to explain this, but the incidence data themselves do not provide enough information to pin down the mechanism.}
+For a covariate, this shows up in a similar way to regression analysis: the covariate under investigation could be a proxy for some other unmodeled phenomenon or unmeasured covariate.
 
@@
 
-We implemented Model~1 using the \code{pomp} package \cite{king16}, relying heavily on the \code{pomp} source code provided by Lee et al.~\cite{lee20}.
+We implemented Model~1 using the \code{pomp} package \cite{king16}, relying heavily on the source code provided by Lee et al.~\cite{lee20}.
 Both analyses used the \code{mif2} implementation of the IF2 algorithm to estimate $\paramVec$ by maximum likelihood.
@@
 Many Monte Carlo replications are then required to quantify and further reduce the error.
-The large increase in the log-likelihood for Model~1 (see Table~\ref{tab:likes}) can primarily be attributed to increasing the computational effort in fitting the model.
+The large increase in the log-likelihood for Model~1 (Table~\ref{tab:likes}) can primarily be attributed to increasing the computational effort used to calibrate the model.
 This result highlights the importance of carefully determining the necessary computational effort needed to maximize model likelihoods and acting accordingly.
-% \TODO{Possibly add more citations to the diagnostics part, maybe li23, laneri10 (Forcing Versus Feedback: Epidemic Malaria and Monsoon Rains in Northwest India), pons-salort18}
-\new{In this case study, this was done by performing standard diagnostics for the IF2 and particle filter algorithms \cite{king16}.
-Given the considerable computational costs of simulation-based algorithms, we find it useful to perform an initial assessment using hyperparameter values that enable relatively quick calculations.
+\new{In this case study, this was done by performing standard diagnostics for the IF2 and particle filter algorithms\cite{king16,li23,pons-salort18,laneri10}.
+Given the considerable computational costs of simulation-based algorithms, we find it useful to perform an initial assessment using hyperparameter values---such as the number of particles, filtering iterations, and replicates based on different parameter initializations---that enable relatively quick calculations.
 The insights obtained from this preliminary analysis help in accurately determining the amount of computation that is required to achieve reliable outcomes.}
@@
 {\figTitle Simulations from Model~1 compared to reported cholera cases.}
-The black curve is observed data, the blue curve is median of 500 simulations from the fitted model, and the vertical dashed line represents break-point when parameters are refit.
+The black curve is observed data, the blue curve is median of 500 simulations from initial conditions using estimated parameters, and the vertical dashed line represents break-point when parameters are refit.
 }
@@
 and can be justified by asymptotic considerations in a large-population limit \cite{dadlani2020,ndii17}.
-Because the process model of Model~2 is deterministic, maximum likelihood estimation reduces to a least squares calculation when combined with a Gaussian measurement model (see Sec.~S3.2 of the supplement material for details).
+Because the process model of Model~2 is deterministic, maximum likelihood estimation reduces to a least squares calculation when combined with a Gaussian measurement model (\nameref{S_meas}).
 Lee et al.~\cite{lee20} fit two versions of Model~2 based on a presupposed change in cholera transmission from a epidemic phase to endemic phase that occurred in March, 2014.
-The inclusion of a change-point in model states and parameters increases the flexibility of the model and hence the ability to fit the observed data.
-The increase in model flexibility, however, results in hidden states that are inconsistent between model phases.
-The inclusion of a model break-point by Lee et al.~\cite{lee20} is perhaps due to a challenging feature of fitting a deterministic model via least squares: discrepancies between model trajectories and observed case counts in highly infectious periods of a disease outbreak will result in greater penalty than the discrepancies between model trajectories and observed case counts in times of relatively low infectiousness, resulting in a bias towards accurately describing periods of high infectiousness.
-This issue is particularly troublesome for modeling cholera dynamics in Haiti: the inability to accurately fit times of low infectiousness may result in poor model forecasts, as few cases of cholera were observed in the last few years of the epidemic.
+The inclusion of a change-point in model states and parameters increased the flexibility of the model and hence the ability to fit the observed data.
+The increase in model flexibility, however, resulted in hidden states that were inconsistent between model phases.
+The inclusion of a model break-point by Lee et al.~\cite{lee20} is perhaps due to a challenging feature of fitting a deterministic model via least squares: discrepancies between model trajectories and observed case counts in highly infectious periods of a disease outbreak will result in greater penalty than the discrepancies between model trajectories and observed case counts in times of relatively low infectiousness.
+This results in a bias towards accurately describing periods of high infectiousness.
+This bias is particularly troublesome for modeling cholera dynamics in Haiti: the inability to accurately fit times of low infectiousness may result in poor model forecasts, as few cases of cholera were observed in the last few years of the epidemic.
 
@@
 We implemented this model using the \code{spatPomp} \code{R} package \cite{asfaw23arxiv}.
-The model was then fit using the subplex algorithm, implemented in the  \code{subplex} package \cite{king2020Subplex}.
+The model was then fit using the subplex algorithm \cite{king2020Subplex}.
 A comparison of the trajectory of the fitted model to the data is given in Fig.~\ref{fig:mod2Traj}.
@@
 as parameters that maximize the likelihood of the decoupled model almost certainly do not maximize the likelihood of the fully coupled model.
-These two concerns may explain the unrealistic forecasts and low likelihood of Model~3 \new{in \cite{lee20}} \old{of Lee et al.~[4]} (see Table~\ref{tab:likes}).
+These two concerns may explain the unrealistic forecasts and low likelihood of Model~3 in \cite{lee20} (Table~\ref{tab:likes}).
 
 At the time Lee et al.~\cite{lee20} conducted their study, there was no known algorithm that could readily be used to maximize the likelihood of an arbitrary meta-population POMP model with coupled spatial dynamics, which justifies the spatial decoupling approximation that was used to calibrate model parameters.
-For our analysis, we calibrate the parameters of the spatially coupled version of Model~3 using the \old{iterated block particle filter (IBPF)} \new{IBPF} algorithm \cite{ionides22}.
-This algorithm extends the work of Ning and Ionides \cite{ning21ibpf}, who provided theoretic justification for the version of the algorithm that only estimates unit-specific parameters.
+For our analysis, we calibrate the parameters of the spatially coupled version of Model~3 using the IBPF algorithm \cite{ionides22}.
+This algorithm extends the work of Ning and Ionides \cite{ning23}, who provided theoretic justification for the version of the algorithm that only estimates unit-specific parameters.
 The IBPF algorithm enables us to directly estimate the parameters of models describing high-dimensional partially-observed nonlinear dynamic systems via likelihood maximization.
@@
 
-The goal of parameter calibration---whether done using Bayesian or frequentist methods---is to find the best description of the observed data under the assumptions of the model.
+The goal of parameter calibration---whether done using Bayesian or frequentist methods---is to find the best description of the observed data in the context of the model.
 Obtaining the best fitting set of parameters for a given model does not, however, guarantee that the model provides an accurate representation of the system under investigation.
@@
 While visual comparisons can be informative, they provide only a weak and informal measure of the goodness-of-fit of a model.
-The study by Lee et al.~\cite{lee20} provides an example of this: their models and parameter estimates resulted in simulations that visually resembled the observed data, yet resulted in model likelihoods that were considerably smaller than likelihoods that can be achieved via the likelihood based optimization techniques that were used (see Table~\ref{tab:likes}).
+The study by Lee et al.~\cite{lee20} provides an example of this: their models and parameter estimates resulted in simulations that visually resembled the observed data, yet resulted in model likelihoods that were considerably smaller than likelihoods that can be achieved (see Table~\ref{tab:likes}).
 Alternative forms of model validation should therefore be used in conjunction with visual comparisons of simulations to observed data.
@@
 the theory and practice of ARMA models is well developed, and these linear models are appropriate on a log scale due to the exponential growth and decay characteristic of biological dynamics.
-We chose to use the auto regressive negative binomial model, however, because the large number of weeks with zero recorded cholera cases in department level data makes a benchmark based on a continuous distribution problematic.
-Log-likelihoods and AIC values of Models~1--3 and of their respective \old{ARMA} benchmark models are provided in Table~\ref{tab:likes}.
-\new{Models that are fit to the same datasets can be directly compared using AIC values, making it a useful tool to compare to benchmark models.
-Though Models~2 and 3 are both fit to department level incidence reports, their AIC values are not directly comparable due to the way Model~3 initializes latent states, which is described in detail in Sec.
-S5 of the supplement.}
+We use the auto regressive negative binomial model, however, because the large number of weeks with zero recorded cholera cases in department level data makes a benchmark based on a continuous distribution problematic.
+Log-likelihoods and AIC values of Models~1--3 and of their respective benchmark models are provided in Table~\ref{tab:likes}.
+Models that are fit to the same datasets can be directly compared using AIC values, making it a useful tool to compare to benchmark models.
+Though Models~2 and 3 are both fit to department level incidence reports, their AIC values are not directly comparable due to the way Model~3 initializes latent states (\nameref{S_init}).
 
@@
 Fig.~\ref{fig:h1SeasRain} shows that the estimated seasonal transmission rate $\transmission$ mimics the rainfall dynamics in Haiti, despite Model~1 not having access to rainfall data.
-This is consistent with previous studies \old{finding that rainfall played an important role in cholera transmission in Haiti [50, 51]} \new{that incorporated rainfall as an important part of their mechanistic model or otherwise argue that rainfall is an important driver of cholera dynamics in Haiti \cite{hulland19,kirpich17,lee20,moise20,pasetto18,kirpich16,eisenberg13,rinaldo12,mavian20}.}
-The estimated seasonality also features an increased transmission rate during the fall, which was noticed at an earlier stage of the epidemic \cite{rinaldo12}; the high transmission rate in the fall may also be a result of the increase transmission that occurred in the fall of 2016, when hurricane Matthew struck Haiti \cite{ferreirai16}.
+This is consistent with previous studies that incorporated rainfall as an important part of their mechanistic model or otherwise argue that rainfall is an important driver of cholera dynamics in Haiti \cite{hulland19,kirpich17,lee20,moise20,pasetto18,kirpich16,eisenberg13,rinaldo12,mavian20}.
+The estimated seasonality also features an increased transmission rate during the fall, which was noticed at an earlier stage of the epidemic \cite{rinaldo12}.
+The high transmission rate in the fall may be a result of the increase transmission that occurred in the fall of 2016, when hurricane Matthew struck Haiti \cite{ferreirai16}.
 
@@
 Our case study provides an example of this in the parameter estimate for the duration of natural immunity due to cholera infection, $\muRS^{-1}$.
-Under the framework of Model~2, the best estimate for this parameter is $\Sexpr{signif(1/h2_fit$h2_params["sigma"],3)}$ yr, suggesting that individuals have \old{effectively} permanent immunity to cholera once infected.
-\new{Rather than interpreting this as scientific evidence that individuals have permanent immunity from cholera, this result suggests that Model~2 favors a regime where reinfection events are a negligble part of the dynamics.}
-\old{To interpret this result, we bear in mind that the data ranged from 2010-2019, and therefore estimates of immunity longer than $10\, \mathrm{yr}$---the upper end of previous estimates of natural immunity [101]---effectively result in the same model dynamics.}
-The depletion of susceptible individuals may \old{also} be attributed to confounding mechanisms---such as localized vaccination programs and non-pharmaceutical interventions that reduce cholera transmission \cite{trevisin22, rebaudet21}---that were not accounted for in the model.
+Under the framework of Model~2, the best estimate for this parameter is $\Sexpr{signif(1/h2_fit$h2_params["sigma"],3)}$ yr, suggesting that individuals have permanent immunity to cholera once infected.
+Rather than interpreting this as scientific evidence that individuals have permanent immunity from cholera, this result suggests that Model~2 favors a regime where reinfection events are a negligble part of the dynamics.
+% TODO: fix typo
+The depletion of susceptible individuals may be attributed to confounding mechanisms---such as localized vaccination programs and non-pharmaceutical interventions that reduce cholera transmission \cite{trevisin22, rebaudet21}---that were not accounted for in the model.
 Perhaps the best interpretation of the estimated parameter, then, is that under the modeling framework that was used, the model most adequately describes the observed data by having a steady decrease in the number of susceptible individuals.
@@
 For example, reconstructed latent variables (such as numbers of susceptible and recovered individuals) should make sense in the context of alternative measurements of these variables \cite{grad12}.
-\old{In the context of our case study, by reconstructing the latent states of Model~3, we notice that the calibrated model favors higher levels of cholera transmission than what was typically observed in the incidence data (S5 Text).
-This result hints at the possibility of model mispecification, and should prompt us to be skeptical of the reliability of forecasts from this model.}
 Similarly, parameters that have been calibrated to data should make sense in the context of alternative lines of evidence about the phenomena being modeled, while making allowance for the possibility that the interpretations of parameters may vary when modeling across differing spatial scales.
@@
 Further, they may enable real-time testing of new scientific hypotheses \cite{lewis22}.
-
-\old{Recent information about a dynamic system should be more relevant for a forecast than older information.
-This assertion may seem self-evident, but it is not the case for deterministic models, for which the initial conditions together with the parameters are sufficient for forecasting, and so recent data do not have special importance.
-Epidemiological forecasts based on deterministic models are not uncommon in practice, despite their limitations [34].
-That may explain why Lee et al.~[1] chose to obtain forecasts by simulating the calibrated models forward from initial conditions.
-We obtain forecasts by simulating future values using latent states that are consistent with the most recent observations; this is done by drawing latent states at the last observation time ($t_N$) from the filtering distribution $f_{\bm{X}_N|\bm{Y}_{1:N}}(\bm{x}_{N} | \bm{y}^*_{1:N} ; \hat\paramVec)$.
-The decision not to do this partially explains the unsuccessful forecasts of Lee et al.~[1]: their Table~S7 shows that the subset of their simulations which were consistent with observing zero cases in 2019 also accurately predicted the prolonged absence of detected cholera.}
 
@@
 
-The ongoing global COVID-19 pandemic has \old{provided a clear example on} \new{demonstrated} how government policy may be affected by the \old{conclusions of scientific models} \new{inferences drawn from mathematical modeling} \cite{saltelli20}.
-\old{This article demonstrates that fitting appropriate scientific models to guide policy is a challenging statistical task.}
-\new{However, the development of credible models---which are supported by data and can provide quantitative insights into a dynamic system---remains a challenging task.
-In this article, we demonstrated opportunities available for raising the current standards of statistical inference for mathematical models of biological systems.}
-
-\new{We presented methodology consistent with existing guidelines but going beyond standard practice.
-In particular, we showed the value of comparing the likelihood of fitted mechanistic models versus non-mechanistic benchmarks.
-Further, we used likelihood-based diagnostic methods to identify specific limitations of proposed models.
-We used likelihood profile methods to assess parameter identifiability and to construct confidence intervals for parameter estimates.
+The ongoing global COVID-19 pandemic has demonstrated how government policy may be affected by the inferences drawn from mathematical modeling \cite{saltelli20}.
+However, the development of credible models---which are supported by data and can provide quantitative insights into a dynamic system---remains a challenging task.
+In this article, we demonstrated opportunities available for raising the current standards of statistical inference for mathematical models of biological systems.
+
+\new{We presented methodology consistent with existing guidelines \cite{behrend20} but going beyond standard practice.
+In particular, we showed the value of comparing the likelihood of fitted mechanistic models versus non-mechanistic benchmarks, a practice that has been previously advocated for \cite{he10} but was not done by any of the studies in our literature review.
+These comparisons, along with other likelihood based diagnostics, help identify specific limitations of proposed models.
+Diagnostic tools include likelihood profile methods, which help to assess parameter identifiability and enable the construction of confidence intervals for parameter estimates \cite{ionides17, simpson23}.
 % We assessed the accuracy of numerical approximations involved in our calculations.
-When reaching conclusions, we identify potential consequences of confounded variables and model misspecification.
+When reaching conclusions, it is important to consider potential consequences of confounded variables and model misspecification.
 }
@@
 
-In our case study, we found that additional attention to statistical details could have resulted in \old{improved} \new{an enhanced} statistical fit to the observed incidence data\old{, leading to improvements in the accuracy of the resulting policy guidance.} \new{This would have improved the accuracy of the policy guidance resulting from the study.}
-\old{We used the same data and models, and even much of the same code, as Lee et al.~[29], and yet ended up with drastically different conclusions.}
-\new{We used the same data, models, and much of the same code used by Lee et al.~\cite{lee20}, but we arrived at drastically different conclusions.}
-\new{Specifically, each of the re-calibrated models predicted with moderate probability that cholera would disappear from Haiti.
-Although there have been new cases of cholera in Haiti, this conclusion aligns more with the prolonged absence of cholera cases from 2019-2022.}
+In our case study, we found that additional attention to statistical details could have resulted in an enhanced statistical fit to the observed incidence data.
+This would have improved the accuracy of the policy guidance resulting from the study.
+We used the same data, models, and much of the same code used by Lee et al.~\cite{lee20}, but we arrived at drastically different conclusions.
+Specifically, each of the re-calibrated models predicted with moderate probability that cholera would disappear from Haiti.
+Although there have been new cases of cholera in Haiti, this conclusion aligns more with the prolonged absence of cholera cases from 2019-2022.
 We acknowledge the benefit of hindsight: our demonstration of a statistically principled route to obtain better-fitting models resulting in more robust insights does not rule out the possibility of discovering other models that fit well yet predict poorly.
 
-\old{Inference for mechanistic time series} \new{Mechanistic} models offer opportunities for understanding and controlling complex dynamic systems.
+Mechanistic models offer opportunities for understanding and controlling complex dynamic systems.
 This case study has investigated issues requiring attention when applying powerful new statistical techniques that can enable statistically efficient inference for a general class of partially observed Markov process models.
-\old{Researchers should check that the computationally intensive numerical calculations are carried out adequately.}
-\new{Researchers should ensure that intensive numerical calculations are adequately executed.}
-\old{Comparison against benchmarks and alternative model specifications should be considered to evaluate the statistical goodness-of-fit.}
-\new{Using benchmarks and alternative model specifications to assess statistical goodness-of-fit should also should be common practice.}
-Once \old{that is accomplished} \new{a model has been adequately callibrated to data}, care is required to assess what causal conclusions can properly be inferred given the possibility of alternative explanations consistent with the data.
+Researchers should ensure that intensive numerical calculations are adequately executed.
+Using benchmarks and alternative model specifications to assess statistical goodness-of-fit should also should be common practice.
+Once a model has been adequately callibrated to data, care is required to assess what causal conclusions can properly be inferred given the possibility of alternative explanations consistent with the data.
 Studies that combine model development with thoughtful data analysis, supported by a high standard of reproducibility, build knowledge about the system under investigation.

### 2024-03-04 ms.Rnw
Removed all calls to 'new'

@@
 Population dynamics may be nonlinear and stochastic, with the resulting complexities compounded by incomplete understanding of the underlying biological mechanisms and by partial observability of the system variables.
-Quantitative models for these dynamic systems offer potential for designing effective control measures \new{\cite{vandermeer13,he10}}.
-Developing and testing models for dynamic systems, and assessing their fitness for guiding policy, is a challenging statistical task \new{\cite{king16}}.
+Quantitative models for these dynamic systems offer potential for designing effective control measures \cite{vandermeer13,he10}.
+Developing and testing models for dynamic systems, and assessing their fitness for guiding policy, is a challenging statistical task \cite{king16}.
 Questions of interest include: What indications should we look for in the data to assess whether the model-based inferences are trustworthy?
@@
 Model~4 has an agent-based construction, featuring considerable mechanistic detail but limited ability to calibrate these details to data.
-\new{These modeling strategies were selected by Lee et a.~\cite{lee20} to represent the range of approaches used in the research community}
+These modeling strategies were selected by Lee et a.~\cite{lee20} to represent the range of approaches used in the research community.
 We focus on Models~1--3, as the strengths and weaknesses of the agent-based modeling approach \cite{tracy18} are outside the scope of this article.
-\new{In addition, agent-based models were less widely used, the agent based model in \cite{lee20} being the only model of this class that was found in our literature review.}
-\new{The data that were used in \cite{lee20}, and that we reanalyze, are displayed in Fig.~\ref{fig:CholeraData}.}
+In addition, agent-based models were less widely used, the agent based model in \cite{lee20} being the only model of this class that was found in our literature review.
+The data that were used in \cite{lee20}, and that we reanalyze, are displayed in Fig.~\ref{fig:CholeraData}.
 
@@
 Advances in OCV technology and vaccine availability, however, raised the possibility of planning a national vaccination program.
-\new{The possibility of controlling the Haiti cholera outbreak via OCV was considered by various research groups \cite{lee20,ivers17,andrews11,walton11,matias17,chao11,fung13,date11,leung22,azman15}}.
-In the \new{Lee et al.~\cite{lee20}} study, certain data were shared between the groups, including demography and vaccination history; vaccine efficacy was also fixed at a shared value between groups.
+The possibility of controlling the Haiti cholera outbreak via OCV was considered by various research groups \cite{lee20,ivers17,andrews11,walton11,matias17,chao11,fung13,date11,leung22,azman15}.
+In the Lee et al.~\cite{lee20} study, certain data were shared between the groups, including demography and vaccination history; vaccine efficacy was also fixed at a shared value between groups.
 Beyond this, the groups made autonomous decisions on what to include and exclude from their models.
@@
 
-\new{\editModelingGuidance}
-
-% \new{\editModelingGuidanceSecond}
-
-\new{Our recommendations are presented in the context of a case study, with the goal of demonstrating how careful adherence to statistical principles may result in improved model fits.}
+\editModelingGuidance
+
+Our recommendations are presented in the context of a case study, with the goal of demonstrating how careful adherence to statistical principles may result in improved model fits.
 We proceed by introducing the general modeling scheme employed by Models~1--3 and provide details of each individual model;
 we then describe how each model is calibrated to data, and present a systematic approach to examining and refining these models.
-\new{Specifically, we focus on how to develop and test variations of the proposed models, as well as diagnosing the models once they have been assimilated to incidence reports.
-This includes a comprehensive tutorial on performing inference with Model~3 (\nameref{S_mod3cal}), a highly non-linear, spatially explicit stochastic model, a challenging task that is possible due to recent methodological advancements.}
-We then use \new{the} improved model fits to project cholera incidence in Haiti under various vaccination scenarios considered by Lee et al.~\cite{lee20}.
-Finally, we conclude with a discussion of the results\new{, in which we relate our general recommendations for model based inference of biological systems to the case study of the Haiti cholera outbreak.}
+Specifically, we focus on how to develop and test variations of the proposed models, as well as diagnosing the models once they have been assimilated to incidence reports.
+This includes a comprehensive tutorial on performing inference with Model~3 (\nameref{S_mod3cal}), a highly non-linear, spatially explicit stochastic model, a challenging task that is possible due to recent methodological advancements.
+We then use the improved model fits to project cholera incidence in Haiti under various vaccination scenarios considered by Lee et al.~\cite{lee20}.
+Finally, we conclude with a discussion of the results, in which we relate our general recommendations for model based inference of biological systems to the case study of the Haiti cholera outbreak.
 
@@
 
-\new{\editMechModels}
+\editMechModels
 
@@
 This motivates the development of mechanistic models, which provides a statistical fit to the available data while also supporting a causal interpretation.
-\new{Despite their limited ability to project the effect of interventions on a system, associative models can be effectively used to make inference on an certain features of a system.
-In our literature review, there were 14 studies that used associative models to describe various aspects of the cholera epidemic \cite{hulland19,moise20,piarroux11,matias17,eisenberg13,rebaudet19CATI,ivers15,raila17,cuneo17,charles14,richterman19,richterman19inverse,bengtsson15,li16adjusting}.}
+Despite their limited ability to project the effect of interventions on a system, associative models can be effectively used to make inference on an certain features of a system.
+In our literature review, there were 14 studies that used associative models to describe various aspects of the cholera epidemic \cite{hulland19,moise20,piarroux11,matias17,eisenberg13,rebaudet19CATI,ivers15,raila17,cuneo17,charles14,richterman19,richterman19inverse,bengtsson15,li16adjusting}.
 
@@
 Parameters that were added during our re-analysis and were not considered by Lee et al. are indicated with $^*$.
-\new{Confidence intervals for model parameters are given in the supplement (\nameref{S_CI}).}
+Confidence intervals for model parameters are given in the supplement (\nameref{S_CI}).
 Translations back into the notation of \cite{lee20} are given in \nameref{S1_Table}.
@@
 Here, $\vaccCounter=0$ corresponds to unvaccinated individuals, and $\vaccCounter \in \seq{1}{\vaccClass}$ describes hypothetical vaccination programs.
-\new{Each program $\vaccCounter$ indexes differences in both the number of doses administered (one versus two doses per individual) and the round of vaccine administration, seperating individuals into compartments with distinct dynamics based on vaccination status.}
+Each program $\vaccCounter$ indexes differences in both the number of doses administered (one versus two doses per individual) and the round of vaccine administration, seperating individuals into compartments with distinct dynamics based on vaccination status.
 The force of infection is
@@
 \end{equation}
-\new{where $\bar{\transmission} = 1$ (wk)\textsuperscript{-1} is a dimensionality constant.}
+where $\bar{\transmission} = 1$ (wk)\textsuperscript{-1} is a dimensionality constant.
 The process noise $d\Gamma(t)/dt$ is multiplicative Gamma-distributed white noise, with infinitesimal variance parameter $\sigmaProc^2$.
@@
 \label{model2:mu_SE}
-\mu_{S_{u\vaccCounter}E_{u\vaccCounter}} &=& \new{(1 - \vaccineEfficacy_\vaccCounter) \, \lambda_u(t)},
+\mu_{S_{u\vaccCounter}E_{u\vaccCounter}} &=& (1 - \vaccineEfficacy_\vaccCounter) \, \lambda_u(t),
 \\
@@
 Therefore, for departments with zero case counts in week 1, we estimate the number of infected individuals rather then treating this value as a constant (\nameref{S_init}).
-Finally, reported cholera cases are modeled using a negative binomial distribution with mean equal to a fraction ($\reportRate$) of individuals in each unit who develop symptoms and seek healthcare\new{, and with over-dispersion parameter $\obsOverdispersion$} (\nameref{S_meas}).
+Finally, reported cholera cases are modeled using a negative binomial distribution with mean equal to a fraction ($\reportRate$) of individuals in each unit who develop symptoms and seek healthcare, and with over-dispersion parameter $\obsOverdispersion$ (\nameref{S_meas}).
 
@@
 Each model is indexed by a parameter vector, $\paramVec$, and different values of $\paramVec$ can result in qualitative differences in the predicted behavior of the system.
-Therefore, the choice of $\paramVec$ used to make inference about the system can greatly affect model-based conclusions  \new{\cite{saltelli20}}.
+Therefore, the choice of $\paramVec$ used to make inference about the system can greatly affect model-based conclusions \cite{saltelli20}.
 Elements of $\paramVec$ can be fixed at a constant value based on scientific understanding of the system, but parameters can also be calibrated to data by maximizing a measure of congruency between the observed data and the model's mechanistic structure.
@@
 In the following subsections we describe in detail our approach to calibrating the three proposed mechanistic models to observed cholera incidence data.
-\new{The main alternative to maximum likelihood estimation is Bayesian inference via Markov chain Monte Carlo, used to analyze the Haiti cholera epidemic by \cite{andrews11,pasetto18,rinaldo12,lewnard16,trevisin22,sallah17,azman12,kuhn14,mari15,gatto12}.}
+The main alternative to maximum likelihood estimation is Bayesian inference via Markov chain Monte Carlo, used to analyze the Haiti cholera epidemic by \cite{andrews11,pasetto18,rinaldo12,lewnard16,trevisin22,sallah17,azman12,kuhn14,mari15,gatto12}.
 
@@
 The ability to maximize the likelihood allows for likelihood-based inference, such as performing statistical tests for potential model improvements.
-We demonstrate this capability by proposing a \new{log-}linear trend $\transmissionTrend$ in transmission in Eq.~\myeqref{model1:beta}:
+We demonstrate this capability by proposing a log-linear trend $\transmissionTrend$ in transmission in Eq.~\myeqref{model1:beta}:
 \begin{equation}
@@
 These results are suggestive that the inclusion of a trend in the transmission rate improves the quantitative ability of Model~1 to describe the observed data.
-\new{The maximum likelihood estimate for $\transmissionTrend$ corresponds to a \Sexpr{paste0(myround(100 - 100*exp(beta_t_mcap_results$mle * 430/215), 1), '\\%')} reduction to the transmission rate over the course of the outbreak, with a $95\%$ confidence interval of (\Sexpr{paste0(rev(myround(100 - 100*exp(beta_t_mcap_results$ci * 430/215), 1)), collapse = ', ', sep = '\\%')}) for the overall reduction in transmission.}
+The maximum likelihood estimate for $\transmissionTrend$ corresponds to a \Sexpr{paste0(myround(100 - 100*exp(beta_t_mcap_results$mle * 430/215), 1), '\\%')} reduction to the transmission rate over the course of the outbreak, with a $95\%$ confidence interval of (\Sexpr{paste0(rev(myround(100 - 100*exp(beta_t_mcap_results$ci * 430/215), 1)), collapse = ', ', sep = '\\%')}) for the overall reduction in transmission.
 The reported results for Model~1 in the remainder of this article were obtained with the inclusion of the parameter $\transmissionTrend$.
-\new{The inclusion of a trend in transmission rate demonstrates a class of model variation that can be highly beneficial to consider: the model variation has a plausible scientific justification, and is easily testable using likelihood based methods.}
+The inclusion of a trend in transmission rate demonstrates a class of model variation that can be highly beneficial to consider: the model variation has a plausible scientific justification, and is easily testable using likelihood based methods.
 
@@
 \caption{\label{fig:betat}
-{\figTitle Confidence interval for \new{the log-}linear trend in transmission.}
+{\figTitle Confidence interval for the log-linear trend in transmission.}
 Monte Carlo adjusted profile (MCAP) of $\transmissionTrend$ for Model~1.
@@
 
-\new{\editConfound}
+\editConfound
 
@@
 This result highlights the importance of carefully determining the necessary computational effort needed to maximize model likelihoods and acting accordingly.
-\new{In this case study, this was done by performing standard diagnostics for the IF2 and particle filter algorithms\cite{king16,li23,pons-salort18,laneri10}.
+In this case study, this was done by performing standard diagnostics for the IF2 and particle filter algorithms\cite{king16,li23,pons-salort18,laneri10}.
 Given the considerable computational costs of simulation-based algorithms, we find it useful to perform an initial assessment using hyperparameter values---such as the number of particles, filtering iterations, and replicates based on different parameter initializations---that enable relatively quick calculations.
-The insights obtained from this preliminary analysis help in accurately determining the amount of computation that is required to achieve reliable outcomes.}
+The insights obtained from this preliminary analysis help in accurately determining the amount of computation that is required to achieve reliable outcomes.
 Simulations from the initial conditions of our fitted model are plotted against the observed incidence data in Fig.~\ref{fig:mod1fit}.
@@
 Projections under the various vaccination scenarios, which are discussed in detail in the  {\bf Forecasts} subsection are also included.
-\new{The gray ribbons represent a $95\%$ interval obtained from the log-normal measurement model.
-To avoid over-plotting, measurement variance is only plotted for the V0 vaccination scenario.}}
+The gray ribbons represent a $95\%$ interval obtained from the log-normal measurement model.
+To avoid over-plotting, measurement variance is only plotted for the V0 vaccination scenario.}
 \end{figure}
@@
 
-\new{\editMetaPop}
+\editMetaPop
 
@@
 Standard statistical models, interpreted as associative models without requiring any mechanistic interpretation of their parameters, provide suitable benchmarks.
-Examples include linear regression, auto regressive moving average \new{(ARMA)} time series models, or even independent and identically distributed measurements.
+Examples include linear regression, auto regressive moving average (ARMA) time series models, or even independent and identically distributed measurements.
 Benchmarks enable us to evaluate the goodness of fit that can be expected of a suitable mechanistic model.
@@
 where $\mathrm{E}\big(Y_n|Y_{n - 1}\big) = \NBintercept + \NBar Y_{n-1}$, and $\Var\big(Y_n|Y_{n - 1}\big) = \mathrm{E}\big(Y_n|Y_{n - 1}\big) + \mathrm{E}\big(Y_n|Y_{n - 1}\big)^2 / \NBsize$.
-\new{\editSpatBench}
+\editSpatBench
 
@@
 
-\new{\editSuppText}
+\editSuppText
 
@@
 
-\new{\editForecastOne}
-
-\new{\editForecastTwo}
+\editForecastOne
+
+\editForecastTwo
 
@@
 Both of these approaches share the similarity that parameters are chosen for the forecast approximately in proportion to their corresponding value of the likelihood function, $f_{\bm{Y}_{1:N}}(\bm{y}_{1:N}^*; \paramVec)$.
-In this analysis, we do not construct forecasts accounting for parameter uncertainty as our focus is on the estimation and diagnosis of mechanistic models\new{, rather than providing forecasts intended to influence policy.}
-Furthermore, we use the projections from a single point estimate to highlight the deficiency of deterministic models that the only variability in model projections is a result of parameter \new{and measurement} uncertainty, which can lead to over-confidence in forecasts \cite{king15}.
+In this analysis, we do not construct forecasts accounting for parameter uncertainty as our focus is on the estimation and diagnosis of mechanistic models, rather than providing forecasts intended to influence policy.
+Furthermore, we use the projections from a single point estimate to highlight the deficiency of deterministic models that the only variability in model projections is a result of parameter and measurement uncertainty, which can lead to over-confidence in forecasts \cite{king15}.
 
@@
 Simulations from probabilistic models (Models~1 and~3) represent possible trajectories of the dynamic system under the scientific assumptions of the models.
-\new{\editModVacc}
+\editModVacc
 Estimates of the probability of cholera elimination can therefore be obtained as the proportion of simulations from these models that result in cholera elimination.
@@
 
-\new{We presented methodology consistent with existing guidelines \cite{behrend20} but going beyond standard practice.
+We presented methodology consistent with existing guidelines \cite{behrend20} but going beyond standard practice.
 In particular, we showed the value of comparing the likelihood of fitted mechanistic models versus non-mechanistic benchmarks, a practice that has been previously advocated for \cite{he10} but was not done by any of the studies in our literature review.
@@
 Diagnostic tools include likelihood profile methods, which help to assess parameter identifiability and enable the construction of confidence intervals for parameter estimates \cite{ionides17, simpson23}.
-% We assessed the accuracy of numerical approximations involved in our calculations.
 When reaching conclusions, it is important to consider potential consequences of confounded variables and model misspecification.
-}
-
-\new{\editDiscussionOne}
-
-\new{\editDiscussionTwo}
-
-\new{\editDiscussionThree}
-
-\new{\editDiscussionFour}
-
-\new{\editDiscussionFive}
-
-% \new{\editDiscussionSix}
+
+\editDiscussionOne
+
+\editDiscussionTwo
+
+\editDiscussionThree
+
+\editDiscussionFour
+
+\editDiscussionFive
 
@@
 
-\paragraph*{\new{S7 Text}}
+\paragraph*{S7 Text}
 \label{S_CI}
-{\bf \new{Confidence Intervals}} \new{Tables and figures describing 95\% confidence intervals for estimated parameters.}
+{\bf Confidence Intervals} Tables and figures describing 95\% confidence intervals for estimated parameters.
 

### 2024-03-04 edits.tex
ran files through a spell-checker

@@
 Based on our literature review, \cite{lee20} was among the studies that best followed these general recommendations for policy-driven modeling.
-Concretly, \cite{lee20} follows four of the five principles outlined in \cite{behrend20}: \emph{complete model documentation}, \emph{complete description of data used}, \emph{communicating uncertainty}, and \emph{testable model outcomes}.
+Concretely, \cite{lee20} follows four of the five principles outlined in \cite{behrend20}: \emph{complete model documentation}, \emph{complete description of data used}, \emph{communicating uncertainty}, and \emph{testable model outcomes}.
 Determining the level of adherence to the first principle, \emph{stakeholder engagement}, is difficult based solely on the article.
@@
 \newcommand\editMechModels{Mechanistic models representing biological phenomena are valuable for epidemiology and consequently for public health policy \cite{lofgren14,mccabe21}.
-More broadly, they have useful roles throughout biology, expecially when combined with statistical methods that properly account for stochasticity and nonlinearity \cite{may04}.
+More broadly, they have useful roles throughout biology, especially when combined with statistical methods that properly account for stochasticity and nonlinearity \cite{may04}.
 In some situations, modern machine learning methods can outperform mechanistic models on epidemiological forecasting tasks \cite{lau22,baker18}.
-The predictive skill of non-mechansitic models can reveal limitations in mechanistic models, but cannot readily replace the scientific understanding obtained by describing the biological dynamics of the system in a mathematical model \cite{baker18,prosperi20}.}
+The predictive skill of non-mechanistic models can reveal limitations in mechanistic models, but cannot readily replace the scientific understanding obtained by describing the biological dynamics of the system in a mathematical model \cite{baker18,prosperi20}.}
 
@@
 In the case of Model~3, comparing a fitted model to a benchmark and then following up by questioning why the model did not describe the data as efficiently as a simple statistical model helped us discern that the model fell short in explaining the surge in cholera transmission following Hurricane Matthew.
-This led us to suggest a model modification that enchanced model fit.
-When a mechanistic model beats a statistical benchmark, we can begin to critically evalutate the causal implications of the fitted model, carefully acknowledging the fact that beating a simple benchmark alone does not imply that the model is both a useful quantitative and qualitative description of the data.
+This led us to suggest a model modification that enhanced model fit.
+When a mechanistic model beats a statistical benchmark, we can begin to critically evaluate the causal implications of the fitted model, carefully acknowledging the fact that beating a simple benchmark alone does not imply that the model is both a useful quantitative and qualitative description of the data.
 In Model~1, we found that the flexible seasonality term closely mirrors the seasonal rainfall patterns in Haiti, reaffirming alternative evidence of the importance of rainfall as a driver of cholera infection and increasing our confidence in Model~1's ability to capture the critical dynamics of the cholera outbreak.}
@@
 If a model falls far behind simple benchmarks, there is likely to be substantial limitation in the data analysis that should be identified and remedied.
-In our case study, the recalibrated version of Model~1 outperformed its benchmark, so we proceeded to examine causal implications.
+In our case study, the re-calibrated version of Model~1 outperformed its benchmark, so we proceeded to examine causal implications.
 When doing so, we found that the fitted model provides a causal description of the dynamic system that is consistent with known features of the system, such as the importance of rainfall as a driver of cholera infection.
@@
 \newcommand\editDiscussionFive{If forecasts are an important component of a modeling task, the forecasts should be consistent with the available data, particularly at the most recently available time points.
-In our case study, we did this by simulating forward from the filtering distribution, as this proceedure conditions latent variables on the available data.
+In our case study, we did this by simulating forward from the filtering distribution, as this procedure conditions latent variables on the available data.
 This type of forecasting, however, is not directly available using a deterministic model, where future dynamics are fully determined by initial conditions and parameter values.
@@
 \newcommand\editForecastTwo{In contrast, for non-deterministic Models~1 and 3, we obtain forecasts by simulating future values using latent states that are harmonious with the most recent data.
-This is done by simulating forward from latent states drawn at the last observation time ($t_N$) from the filtering distrubion $f_{\bm{X}_N|\bm{Y}_{1:N}}(\bm{x}_{N} | \bm{y}^*_{1:N} ; \hat\paramVec)$.
+This is done by simulating forward from latent states drawn at the last observation time ($t_N$) from the filtering distribution $f_{\bm{X}_N|\bm{Y}_{1:N}}(\bm{x}_{N} | \bm{y}^*_{1:N} ; \hat\paramVec)$.
 The decision to obtain model forecasts from initial conditions partially explains the unsuccessful forecasts of Lee et al.~\cite{lee20}.

### 2024-03-04 ms.Rnw
ran files through a spell-checker

@@
 Beyond this, the groups made autonomous decisions on what to include and exclude from their models.
-Despite their autonomy, the four independent teams obtained a consensus that an extensive nationwide vaccination campaign would be necessary to eliminate cholera from Haiti, estimating that a large number of cumulative cholera cases would be observed in the absense of additional vaccination efforts (Figure~3 and 4 of \cite{lee20}).
+Despite their autonomy, the four independent teams obtained a consensus that an extensive nationwide vaccination campaign would be necessary to eliminate cholera from Haiti, estimating that a large number of cumulative cholera cases would be observed in the absence of additional vaccination efforts (Figure~3 and 4 of \cite{lee20}).
 These forecasts are inconsistent with the prolonged period with no confirmed cholera cases between February, 2019 and September, 2022 \cite{trevisin22}.
@@
 
-# Fit Model 1 bechmark:
+# Fit Model 1 benchmark:
 m1_bench <- ar1_NegBinom(m1_agg_data$cases, init = 7364)
@@
 
-Using the conditional independence of $\bm{Y}_{1:N}$ given $\bm{X}_{0:N}$ and the Markov property of $\bm{X}_{0:N}$, the joint density can be refactored into the useful form given in Eq.~\myeqref{eq:jointLik}:
+Using the conditional independence of $\bm{Y}_{1:N}$ given $\bm{X}_{0:N}$ and the Markov property of $\bm{X}_{0:N}$, the joint density can be re-factored into the useful form given in Eq.~\myeqref{eq:jointLik}:
 
@@
 Here, $\vaccCounter=0$ corresponds to unvaccinated individuals, and $\vaccCounter \in \seq{1}{\vaccClass}$ describes hypothetical vaccination programs.
-Each program $\vaccCounter$ indexes differences in both the number of doses administered (one versus two doses per individual) and the round of vaccine administration, seperating individuals into compartments with distinct dynamics based on vaccination status.
+Each program $\vaccCounter$ indexes differences in both the number of doses administered (one versus two doses per individual) and the round of vaccine administration, separating individuals into compartments with distinct dynamics based on vaccination status.
 The force of infection is
@@
 
-# Create a figure showing annual rainfal, scaled from 0-1. This is top figure
+# Create a figure showing annual rainfall, scaled from 0-1. This is top figure
 gg_rain <- df_rain %>%
@@
   ylab('Standardized Weekly\nNational Rain') +
-  theme(  # Remove elements that are un-unsed becuase this is top figure
+  theme(  # Remove elements that are un-unused because this is top figure
     axis.title.x = element_blank(),
@@
 (Bottom) estimated seasonality in the transmission rate (dashed line) plotted alongside mean rainfall (solid line).
-The outsided effect of rainfall in the fall may be due to Hurricane Matthew, which struck Haiti in October of 2016 and resulted in an increase of cholera cases in the nation.
+The outsized effect of rainfall in the fall may be due to Hurricane Matthew, which struck Haiti in October of 2016 and resulted in an increase of cholera cases in the nation.
 }
@@
 Under the framework of Model~2, the best estimate for this parameter is $\Sexpr{signif(1/h2_fit$h2_params["sigma"],3)}$ yr, suggesting that individuals have permanent immunity to cholera once infected.
-Rather than interpreting this as scientific evidence that individuals have permanent immunity from cholera, this result suggests that Model~2 favors a regime where reinfection events are a negligble part of the dynamics.
+Rather than interpreting this as scientific evidence that individuals have permanent immunity from cholera, this result suggests that Model~2 favors a regime where reinfection events are a negligible part of the dynamics.
 % TODO: fix typo
@@
 Using benchmarks and alternative model specifications to assess statistical goodness-of-fit should also should be common practice.
-Once a model has been adequately callibrated to data, care is required to assess what causal conclusions can properly be inferred given the possibility of alternative explanations consistent with the data.
+Once a model has been adequately calibrated to data, care is required to assess what causal conclusions can properly be inferred given the possibility of alternative explanations consistent with the data.
 Studies that combine model development with thoughtful data analysis, supported by a high standard of reproducibility, build knowledge about the system under investigation.

### 2024-03-04 si/calibrateMod3.Rnw
ran files through a spell-checker

@@
 
-# Perform block partical filter with fitted model, to get conditional
+# Perform block particle filter with fitted model, to get conditional
 # log-likelihoods.
@@
 This result may be unsurprising in the context of the models that we are comparing.
-Because the cholera transmission in individual departments likely depends on the national prevalence of cholera and the Vibrio cholerae bacteria, our spatially-independnet benchmark model that relies exclusively on the previous number of case within any given unit has a difficult time predicting a resurgence of cases.
+Because the cholera transmission in individual departments likely depends on the national prevalence of cholera and the Vibrio cholerae bacteria, our spatially-independent benchmark model that relies exclusively on the previous number of case within any given unit has a difficult time predicting a resurgence of cases.
 
@@
 
-# This bake perfoms the following:
+# This bake performs the following:
 #  (1) simulates the model from initial conditions

### 2024-03-04 si/inputs/modelDetails.tex
ran files through a spell-checker

@@
 There are different ways to use a collection of rate functions to build a fully specified model.
-We proceed to describe the ones we use in this paper: via a system of ordinary differential equations (Sec.~\ref{subsec:ode}), a simple Markov counting system (Sec.~\ref{subsec:smcs}), and an overdispersed Markov counting system (Sec.~\ref{subsec:odmcs}).
+We proceed to describe the ones we use in this paper: via a system of ordinary differential equations (Sec.~\ref{subsec:ode}), a simple Markov counting system (Sec.~\ref{subsec:smcs}), and an over-dispersed Markov counting system (Sec.~\ref{subsec:odmcs}).
 Other representations include stochastic differential equations driven by Gaussian noise or Gamma noise \citep{bhadra11}.
@@
 To permit more general mean-variance relationships for a Markov counting system, we must permit jumps of size greater than one.
-The utility of overdispersed models, where the infinitesimal variance of the flow exceeds the infinitesimal mean, has become widely recognized \citep{stocks20,he10}.
+The utility of over-dispersed models, where the infinitesimal variance of the flow exceeds the infinitesimal mean, has become widely recognized \citep{stocks20,he10}.
 
@@
 
-Including white noise in the rate function enables the possibility of an overdispersed Markov counting system \citep{breto11,breto09,he10}.
+Including white noise in the rate function enables the possibility of an over-dispersed Markov counting system \citep{breto11,breto09,he10}.
 Since rates should be non-negative, Gaussian noise is not appropriate and gamma noise is a convenient option that has found various applications \citep{romero-severson15, subramanian20}.
@@
 \end{equation}
-Formally interpreting the meaning of \myeqref{eq:odmcs1} is not trivial, and we do so by constructing a Markov process $X(t)$ as the limit of the Euler scheme descrived in Section~\ref{sec:numerics}, below.
+Formally interpreting the meaning of \myeqref{eq:odmcs1} is not trivial, and we do so by constructing a Markov process $X(t)$ as the limit of the Euler scheme described in Section~\ref{sec:numerics}, below.
 Therefore, the numerical scheme in Sec.~\ref{sec:numerics} can be taken as a definition of the meaning of \myeqref{eq:odmcs1}.
-The Markov chain defined by the limit of this Euler scheme as the step size decreases is an overdispersed Markov counting system, with the possibility of instantaneous jumps of size greater than one \citep{breto11}.
+The Markov chain defined by the limit of this Euler scheme as the step size decreases is an over-dispersed Markov counting system, with the possibility of instantaneous jumps of size greater than one \citep{breto11}.
 
@@
 Then, $\big(\Delta N_{i1},\dots,\Delta N_{ic},\Delta N_{i\demography}\big)$ has the multinomial distribution where $X_i(t)$ individuals are allocated independently to $\seq{1}{c}\cup\{\demography\}$ with probabilities given by \eqref{eq:numerics3} and \eqref{eq:numerics4}.
-We use the \code{reulermultinom} function in the \code{pomp} package to draw from this multionomial distribution.
+We use the \code{reulermultinom} function in the \code{pomp} package to draw from this multinomial distribution.
 

### 2024-03-04 r1/trackedMS.Rnw
moved tracked MS files to r1

--- 2024-03-04 trackedMS.Rnw
@@
 Beyond this, the groups made autonomous decisions on what to include and exclude from their models.
-\old{The groups largely adhered to existing guidelines for creating models to inform policy [35, 36] and, d}\new{D}espite their autonomy, \new{the four independent teams} obtained a consensus that an extensive nationwide vaccination campaign would be necessary to eliminate cholera from Haiti\new{, estimating that a large number of cumulative cholera cases would be observed in the absense of additional vaccination efforts (Figure~3 and 4 of \cite{lee20}).}
+\old{The groups largely adhered to existing guidelines for creating models to inform policy [35, 36] and, d}\new{D}espite their autonomy, \new{the four independent teams} obtained a consensus that an extensive nationwide vaccination campaign would be necessary to eliminate cholera from Haiti\new{, estimating that a large number of cumulative cholera cases would be observed in the absence of additional vaccination efforts (Figure~3 and 4 of \cite{lee20}).}
 \old{Their forecasts, however, were} \new{These forecasts are} inconsistent with the prolonged period with no confirmed cholera cases between February, 2019 and September, 2022 \cite{trevisin22}.
@@
 
-Using the conditional independence of $\bm{Y}_{1:N}$ given $\bm{X}_{0:N}$ and the Markov property of $\bm{X}_{0:N}$, the joint density can be refactored into the useful form given in Eq.~\myeqref{eq:jointLik}:
+Using the conditional independence of $\bm{Y}_{1:N}$ given $\bm{X}_{0:N}$ and the Markov property of $\bm{X}_{0:N}$, the joint density can be \old{refactored} \new{re-factored} into the useful form given in Eq.~\myeqref{eq:jointLik}:
 
@@
 Under the framework of Model~2, the best estimate for this parameter is $\Sexpr{signif(1/h2_fit$h2_params["sigma"],3)}$ yr, suggesting that individuals have \old{effectively} permanent immunity to cholera once infected.
-\new{Rather than interpreting this as scientific evidence that individuals have permanent immunity from cholera, this result suggests that Model~2 favors a regime where reinfection events are a negligble part of the dynamics.}
+\new{Rather than interpreting this as scientific evidence that individuals have permanent immunity from cholera, this result suggests that Model~2 favors a regime where reinfection events are a negligible part of the dynamics.}
 \old{To interpret this result, we bear in mind that the data ranged from 2010-2019, and therefore estimates of immunity longer than $10\, \mathrm{yr}$---the upper end of previous estimates of natural immunity [101]---effectively result in the same model dynamics.}
@@
 \new{Using benchmarks and alternative model specifications to assess statistical goodness-of-fit should also should be common practice.}
-Once \old{that is accomplished} \new{a model has been adequately callibrated to data}, care is required to assess what causal conclusions can properly be inferred given the possibility of alternative explanations consistent with the data.
+Once \old{that is accomplished} \new{a model has been adequately calibrated to data}, care is required to assess what causal conclusions can properly be inferred given the possibility of alternative explanations consistent with the data.
 Studies that combine model development with thoughtful data analysis, supported by a high standard of reproducibility, build knowledge about the system under investigation.

### 2024-03-04 r1/trackedMS.tex
moved tracked MS files to r1

--- 2024-03-04 r1/trackedMS.Rnw
@@
-\documentclass[10pt,letterpaper]{article}
+\documentclass[10pt,letterpaper]{article}\usepackage[]{graphicx}\usepackage[table]{xcolor}
+% maxwidth is the original width if it is less than linewidth
+% otherwise use linewidth (to make sure the graphics do not exceed the margin)
+\makeatletter
+\def\maxwidth{ %
+  \ifdim\Gin@nat@width>\linewidth
+    \linewidth
+  \else
+    \Gin@nat@width
+  \fi
+}
+\makeatother
+
+\definecolor{fgcolor}{rgb}{0.345, 0.345, 0.345}
+\newcommand{\hlnum}[1]{\textcolor[rgb]{0.686,0.059,0.569}{#1}}%
+\newcommand{\hlstr}[1]{\textcolor[rgb]{0.192,0.494,0.8}{#1}}%
+\newcommand{\hlcom}[1]{\textcolor[rgb]{0.678,0.584,0.686}{\textit{#1}}}%
+\newcommand{\hlopt}[1]{\textcolor[rgb]{0,0,0}{#1}}%
+\newcommand{\hlstd}[1]{\textcolor[rgb]{0.345,0.345,0.345}{#1}}%
+\newcommand{\hlkwa}[1]{\textcolor[rgb]{0.161,0.373,0.58}{\textbf{#1}}}%
+\newcommand{\hlkwb}[1]{\textcolor[rgb]{0.69,0.353,0.396}{#1}}%
+\newcommand{\hlkwc}[1]{\textcolor[rgb]{0.333,0.667,0.333}{#1}}%
+\newcommand{\hlkwd}[1]{\textcolor[rgb]{0.737,0.353,0.396}{\textbf{#1}}}%
+\let\hlipl\hlkwb
+
+\usepackage{framed}
+\makeatletter
+\newenvironment{kframe}{%
+ \def\at@end@of@kframe{}%
+ \ifinner\ifhmode%
+  \def\at@end@of@kframe{\end{minipage}}%
+  \begin{minipage}{\columnwidth}%
+ \fi\fi%
+ \def\FrameCommand##1{\hskip\@totalleftmargin \hskip-\fboxsep
+ \colorbox{shadecolor}{##1}\hskip-\fboxsep
+     % There is no \\@totalrightmargin, so:
+     \hskip-\linewidth \hskip-\@totalleftmargin \hskip\columnwidth}%
+ \MakeFramed {\advance\hsize-\width
+   \@totalleftmargin\z@ \linewidth\hsize
+   \@setminipage}}%
+ {\par\unskip\endMakeFramed%
+ \at@end@of@kframe}
+\makeatother
+
+\definecolor{shadecolor}{rgb}{.97, .97, .97}
+\definecolor{messagecolor}{rgb}{0, 0, 0}
+\definecolor{warningcolor}{rgb}{1, 0, 1}
+\definecolor{errorcolor}{rgb}{1, 0, 0}
+\newenvironment{knitrout}{}{} % an empty environment to be redefined in TeX
+
+\usepackage{alltt}
 \usepackage[top=0.85in,left=2.75in,footskip=0.75in]{geometry}
@@
 \newcolumntype{t}{>{\tiny}c}
-
+\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
 \begin{document}
@@
 \end{flushleft}
-
-<<Setup, include=FALSE,echo=FALSE,results='hide'>>=
-
-# Load Required packages
-library(knitr)
-library(pomp)
-library(spatPomp)
-library(foreach)
-library(doParallel)
-library(doRNG)
-library(haitipkg)
-library(tidyverse)
-
-# Set run level:
-#   1: Used primarily for debugging
-#   2: Used to test various results in a way that provides meaningful results, yet doesn't require an extensive amount of time to run.
-#   3: Used to get final results.
-#
-# Throughout the document, if run_level_3 results exist already exist,
-# then these results are used.
-RUN_LEVEL <- 3
-rl_dir <- paste0("run_level_", RUN_LEVEL, "/")
-
-# Check if run_level_* directories exist; if not, create them.
-for (i in c(1, 3)) {
-  if (!dir.exists(paste0("model", i, '/', rl_dir))) {
-    dir.create(paste0("model", i, '/', rl_dir), recursive = TRUE)
-  }
-}
-
-# Model 2 doesn't need a run_level, since all computations are the same.
-if (!dir.exists('model2')) dir.create("model2/")
-
-if (RUN_LEVEL == 3) {
-  # Print session info into a txt file. Only update if RUN_LEVEL == 3, which
-  # is when the final results are calculated.
-  cat(
-    utils::capture.output(sessionInfo()),
-    file = "sessionInfo.txt",
-    sep="\n"
-  )
-}
-
-opts_knit$set(concordance=TRUE)
-opts_chunk$set(
-    concordance = TRUE,
-    tidy = TRUE,
-    message = FALSE,
-    warning = FALSE,
-    tidy.opts = list(
-        keep.blank.line = FALSE
-    ),
-    comment = "",
-    echo = FALSE,
-    dev.args = list(
-        bg = "transparent",
-        pointsize = 9
-    ),
-    fig.path = "figure/"
-)
-
-myround <- function(x, digits = 1) {
-  # taken from the broman package
-  if (digits < 1)
-    stop("This is intended for the case digits >= 1.")
-  if (length(digits) > 1) {
-    digits <- digits[1]
-    warning("Using only digits[1]")
-  }
-  tmp <- sprintf(paste("%.", digits, "f", sep = ""), x)
-  zero <- paste0("0.", paste(rep("0", digits), collapse = ""))
-  tmp[tmp == paste0("-", zero)] <- zero
-  tmp
-}
-
-# Setting black and white ggplot2 theme for entire document.
-theme_set(theme_bw() + theme(plot.title = element_text(hjust = 0.5)))
-
-# Get the number of cores. The high performance computing cluster that we
-# used has 36 cores.
-cores <-  as.numeric(Sys.getenv('SLURM_NTASKS_PER_NODE', unset=NA))
-if(is.na(cores)) cores <- detectCores()
-
-# Register parallel back-end. This code below creates a fork cluster, so we are assuming
-# that the code is run on a mac or linux machine; this doesn't matter if you
-# plan on simply loading the results, but if you would like to redo the expensive
-# calculations (e.g., test minor changes to the code), you need to make sure you
-# register the correct parallel back-end.
-registerDoParallel(cores)
-
-# Create a mapping between department names and what should be shown in figures
-dep_labeller <- as_labeller(
-  c(
-    'Artibonite' = 'Artibonite',
-    'Sud_Est' = 'Sud-Est',
-    'Sud.Est' = 'Sud-Est',
-    'Nippes' = 'Nippes',
-    'Nord_Est' = 'Nord-Est',
-    'Nord.Est' = 'Nord-Est',
-    'Ouest' = 'Ouest',
-    'Centre' = 'Centre',
-    'Nord' = 'Nord',
-    'Sud' = 'Sud',
-    'Nord_Ouest' = 'Nord-Ouest',
-    'Nord.Ouest' = 'Nord-Ouest',
-    'Grande_Anse' = 'Grand\'Anse',
-    'Grand.Anse' = 'Grand\'Anse'
-  )
-)
-
-# Vector used to arrange figures in the same order that was used in Lee et al (2020)
-plot_order <- c(
-  'Artibonite',
-  'Sud_Est',
-  'Nippes',
-  'Nord_Est',
-  'Ouest',
-  'Centre',
-  'Nord',
-  'Sud',
-  'Nord_Ouest',
-  'Grande_Anse'
-)
-
-options(
-  scipen = 0,
-  help_type = "html",
-  stringsAsFactors = FALSE,
-  continue = "+  ",
-  width = 70,
-  useFancyQuotes = FALSE,
-  reindent.spaces = 2,
-  xtable.comment = FALSE
-)
-
-# Create "long" version of the observed cholera cases, for plotting purposes.
-dep_plot_df <- haitiCholera %>%
-  select(-report) %>%
-  pivot_longer(
-    data = .,
-    cols = -date_saturday,
-    values_to = 'cases',
-    names_to = 'dep'
-  ) %>%
-  mutate(
-    date = as.Date(date_saturday),
-    dep = gsub("\\.", "_", dep)
-  ) %>%
-  mutate(
-    dep = case_when(dep == "Grand_Anse" ~ "Grande_Anse", TRUE ~ dep)
-  )
-
-# Calculate total number of nationally aggregated cholera cases, using previous
-# data.
-true_agg_cases <- dep_plot_df %>%
-  tidyr::pivot_wider(
-    data = .,
-    id_cols = c(date),
-    names_from = dep,
-    values_from = cases,
-    names_prefix = 'cases_'
-  ) %>%
-  mutate(
-    ReportedAll = cases_Artibonite + cases_Centre +
-      cases_Grande_Anse + cases_Nippes + cases_Nord +
-      cases_Nord_Est + cases_Ouest + cases_Sud +
-      cases_Sud_Est + cases_Nord_Ouest
-  )
-@
-
-<<Calculate Lee1 Likelihoods, echo=FALSE, message=FALSE, include=FALSE>>=
-
-# In this code chunk, we attempt to estimate the likelihood of Model 1 that
-# is described by Lee et al (2020). Because this chunk produces some
-# computationally intensive calculations that must also be done in the
-# supplement, we save some additional information in this chunk that is only
-# used in the supplement (si/si.Rnw) and not in the manuscript.
-
-# Setting algorithmic hyper-parameters based on RUN_LEVEL.
-#
-#                    RUN_LEVEL =  1,   2,    3
-N_PARAMS   <- switch(RUN_LEVEL,   5,  20,  nrow(h1LeeStartsEpi))  # Number of parameters to sample
-N_LEE_SIMS <- switch(RUN_LEVEL,   5,  20,   20)  # Number of simulations per parameter
-NP_EVAL    <- switch(RUN_LEVEL,  50, 200, 2000)  # Number of particles used to evaluate the likelihood
-NREPS_EVAL <- switch(RUN_LEVEL,   2,   5,   36)  # Number of replicates of likelihood evaluation.
-
-h1_epi <- haiti1()
-
-set.seed(141837)
-
-if (RUN_LEVEL == 1 && file.exists('model1/run_level_3/lee1_epi_evals.rds')) {
-
-  # If RUN_LEVEL == 1, and RUN_LEVEL == 3 results exist, just load the RL3
-  # results.
-  lee_epi_ll <- readRDS("model1/run_level_3/lee1_epi_evals.rds")
-} else {
-
-  # Compute the run_level_<RUN_LEVEL> results and save them, or simply load
-  # the results if the bake load conditions are met.
-  lee_epi_ll <- bake(
-    file = paste0("model1/", rl_dir, "lee1_epi_evals.rds"), {
-
-      # Randomly sample from starting params used by Lee et al (2020).
-      # If RUN_LEVEL == 3, all starting parameters are used.
-      which_params <- sample(
-        1:nrow(h1LeeStartsEpi),
-        size = N_PARAMS,
-        replace = FALSE
-      )
-      epi_params <- h1LeeStartsEpi[which_params, ]
-
-      # Create data.frame to save results
-      lee_epi_ll <- data.frame(
-        "pfLL" = NA_real_,
-        "pfse" = NA_real_,
-        "parid" = epi_params$parid
-      )
-
-      for (j in 1:nrow(lee_epi_ll)) {
-        # Get parameters of interest
-        pf_params <- epi_params[j, -1][names(coef(h1_epi))]
-
-        # make results reproducible
-        registerDoRNG((j * 687383921) %% 7919)
-
-        # Calculate log likelihoods
-        ll_evals <- foreach(i=1:NREPS_EVAL, .combine = c) %dopar% {
-          logLik(pfilter(h1_epi, params = pf_params, Np = NP_EVAL))
-        }
-        lee_epi_ll[j, 'pfLL'] <- logmeanexp(ll_evals, se = TRUE)[1]  # Log-lik
-        lee_epi_ll[j, 'pfse'] <- logmeanexp(ll_evals, se = TRUE)[2]  # SE
-      }
-
-      lee_epi_ll
-    }
-  )
-}
-
-# Create the endemic version of the model
-h1_end <- haiti1(period = "endemic")
-
-if (RUN_LEVEL == 1 && file.exists('model1/run_level_3/lee1_end_evals.rds')) {
-
-  # If RUN_LEVEL == 1, and RUN_LEVEL == 3 results exist, just load the RL3
-  # results.
-  lee_end_ll <- readRDS("model1/run_level_3/lee1_end_evals.rds")
-} else {
-
-  # Compute the run_level_<RUN_LEVEL> results and save them, or simply load
-  # the results if the bake load conditions are met.
-  lee_end_ll <- bake(
-    file = paste0("model1/", rl_dir, "lee1_end_evals.rds"), {
-
-      end_params <- h1LeeStartsEnd[h1LeeStartsEnd$parid %in% lee_epi_ll$parid, ]
-
-      # Create data.frame to save results
-      lee_end_ll <- data.frame(
-        "pfLL" = NA_real_,
-        "pfse" = NA_real_,
-        'parid' = end_params$parid
-      )
-
-      # Perform particle filter to get likelihood estimates
-      for (j in 1:nrow(lee_end_ll)) {
-        # Get parameters of interest
-        pf_params <- end_params[j, ][names(coef(h1_end))]
-
-        # make results reproducible
-        registerDoRNG((j * 687383921) %% 7919)
-
-        # Calculate log likelihoods
-        ll_evals <- foreach(i=1:NREPS_EVAL, .combine = c) %dopar% {
-          logLik(pfilter(h1_end, params = pf_params, Np = NP_EVAL))
-        }
-
-        lee_end_ll[j, 'pfLL'] <- logmeanexp(ll_evals, se = TRUE)[1]
-        lee_end_ll[j, 'pfse'] <- logmeanexp(ll_evals, se = TRUE)[2]
-      }
-
-      lee_end_ll
-    }
-  )
-}
-
-# Combine epidemic and endemic likelihoods by `parid` to get the likelihood of
-# the complete model.
-mod1_all_likes <- dplyr::inner_join(
-  x = lee_epi_ll,
-  y = lee_end_ll,
-  by = "parid"
-  ) %>%
-  dplyr::mutate(
-    joint_ll = pfLL.x + pfLL.y
-  )
-
-# Approximate the likelihood of the Lee et al (2020) version of Model 1 by
-# using the parameter sets that resulted in the maximum likelihood.
-mod1_lee_ll <- max(mod1_all_likes$joint_ll, na.rm = TRUE)
-
-# Doing some clean-up. The goal is that each major chunk can be run independent
-# of other chunks. Removing saved objects is not entirely necessary, but can
-# help new users know what objects are needed later on and which were only
-# temporarily needed for each particular chunk. Here we suppress warnings
-# because some of the temporary objects only exist if the `bake` code needs to
-# be recomputed, and hence `rm` will throw warnings if it tries to remove
-# objects that don't exist.
-suppressWarnings(
-  rm(
-    lee_end_ll, lee_epi_ll, h1_end, h1_epi, i, mod1_all_likes,
-    N_LEE_SIMS, N_PARAMS, NP_EVAL, NREPS_EVAL, end_params, epi_params,
-    pf_params, j, ll_evals, which_params
-  )
-)
-gc()  # Garbage Collection to free un-used memory
-@
 
@@
 \begin{figure}[!h]
-<<Plot_Reported_Cases, echo=FALSE, fig.height=3.5>>=
-
-# Plot reported cholera cases, by department. Creates Figure 1.
-ggplot(dep_plot_df, aes(x = date, y = cases + 1)) +
-  facet_wrap(~dep, nrow = 2, labeller = dep_labeller) +
-  geom_line() +
-  theme(axis.title.x = element_blank()) +
-  ylab('Reported Cases') +
-  scale_y_log10(
-    labels = scales::trans_format("log10", scales::math_format(10^.x))
-  ) +
-  scale_x_date(
-    date_labels = "'%y",
-    breaks = seq.Date(
-      from = as.Date("2011-01-01"),
-      to = as.Date("2019-01-01"),
-      by = '2 years'
-    )
-  )
-@
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
+\includegraphics[width=\maxwidth]{figure/Plot_Reported_Cases-1} 
+\end{knitrout}
 \caption{\label{fig:CholeraData}
@@
 
-<<Load Model1 Params, echo=FALSE, message=FALSE, include=FALSE>>=
-
-# Load the file containing profiles of Model 1 Parameters.
-if (RUN_LEVEL != 3 && file.exists("model1/run_level_3/h1_profiles.rds")) {
-  h1_prof <- readRDS("model1/run_level_3/h1_profiles.rds")
-} else if (file.exists(paste0('model1/', rl_dir, 'h1_profiles.rds'))) {
-  h1_prof <- readRDS(paste0('model1/', rl_dir, 'h1_profiles.rds'))
-} else {
-  stop(paste0("Missing file: `model1/", rl_dir, 'h1_profiles.rds`;\nThe file can be obtained by running the script `H1profile.R`'))
-}
-
-# The result is a data.frame containing many parameter values and their
-# corresponding likelihood evaluation. To perform a likelihood profile,
-# we calculate the maximum likelihood for each value of betat.
-h1_prof_res <- h1_prof %>%
-  filter(prof_var == 'betat') %>%
-  group_by(betat) %>%
-  summarize(logLik = max(logLik))
-
-# This is used later in figure 2
-beta_t_mcap_results <- pomp::mcap(h1_prof_res$logLik, h1_prof_res$betat, span = 1)
-
-# Find the set of parameters that correspond to the maximum likelihood
-p1 <- h1_prof %>%
-  slice_max(order_by = logLik, n = 1) %>%
-  # Remove non-parameters from the data.frame (hyperparameters + likelihood)
-  select(-(1:12)) %>%
-  unlist()
-
-# Save the likelihood of the best parameters, for use later.
-mod1_ll <- h1_prof %>%
-  slice_max(order_by = logLik, n = 1) %>%
-  select(logLik) %>%
-  unlist()
-
-# Just a bit of tidying up: removing parameters that we won't use again.
-rm(h1_prof)
-gc()
-@
-
-<<Fit Model 2, echo=FALSE, include=FALSE, message=FALSE, warning=FALSE>>=
-
-# Calculate log-likelihood of Lee et al (2020) model.
-# We have parameter estimates for everything except the variance of the
-# measurement model, so here we fix all parameters except the variance, which
-# is estimated for both endemic and epidemic periods.
-lee20_replication <- bake(
-  file = 'model2/lee20_replication.rds', {
-    h2_lee_epi <- haiti2()
-    lee_epi_params <- h2_lee_epi@params
-    epi_ofun_v <- traj_objfun(
-      h2_lee_epi, est = c('v'), paramnames = c('v')
-    )
-
-    # Get starting values for the estimated parameter
-    start_v_epi <- lee_epi_params[c('v')]
-
-    # Use the subplex package to calibrate 'v' for the epi-model with fixed coefs.
-    epi_fit_v <- subplex::subplex(par = log(start_v_epi), fn = epi_ofun_v)
-    epi_params_update_v <- lee_epi_params
-    epi_params_update_v['v'] <- exp(epi_fit_v$par)
-
-    # Now we repeat this for the endemic portion of the model:
-    h2_lee_end <- haiti2(region = 'after')
-    lee_end_params <- h2_lee_end@params
-    end_ofun_v <- traj_objfun(
-      h2_lee_end, est = 'v', paramnames = 'v'
-    )
-
-    # Get starting values for the estimated parameter
-    start_v_end <- lee_end_params[c('v')]
-
-    # Use the subplex package to calibrate 'v' for the epi-model with fixed coefs.
-    end_fit_v <- subplex::subplex(par = log(start_v_end), fn = end_ofun_v)
-    end_params_update_v <- lee_end_params
-    end_params_update_v['v'] <- exp(epi_fit_v$par)
-
-    # Return the results
-    c(
-      "leeFit_ll" = -epi_fit_v$value - end_fit_v$value,
-      "leeFit_n_params" = 6
-    )
-  }, timing = FALSE
-)
-
-h2_fit <- bake(
-  file = 'model2/model2_fit.rds', {
-
-    # Create model and save initial values for each parameter
-    h2 <- haiti2(cutoff = 10000, measure = "log")  # Default is epidemic period
-    h2_params <- h2@params
-    est_params <- c('Mu', 'Beta', 'BetaW', 'v', 'sigma', 'phase')
-    n_fit_params <- length(est_params)
-
-    # Get objective function to minimize
-    epi_ofun <- pomp::traj_objfun(h2, est = est_params, params = h2_params)
-
-    # Vector of starting values of the parameters that are estimated.
-    # phase is not included because it's the only non log-transformed variable
-    theta <- h2_params[c("Mu", "Beta", "BetaW", "v", "sigma")]
-
-    # Fit the epi model
-    h2_fit_tmp <- subplex::subplex(
-      par = c(log(theta), h2_params['phase']),
-      fn = epi_ofun
-    )
-
-    # Save the fitted parameters to the h2_params vector
-    h2_params[c("Mu","Beta","BetaW","v", "sigma")] <- exp(h2_fit_tmp$par[c("Mu", "Beta", "BetaW", "v", "sigma")])
-    h2_params['phase'] <- h2_fit_tmp$par['phase']
-
-    model2_fit_results <- list()
-    model2_fit_results[['h2_params']] <- h2_params
-    model2_fit_results[["n_fit_params"]] <- n_fit_params
-    model2_fit_results[['logLik']] <- -h2_fit_tmp$value
-
-    rm(h2, epi_ofun, theta, h2_fit_tmp)
-
-    model2_fit_results
-  }, timing = FALSE
-)
-@
-
-<<Load Model3 Parameters, echo=FALSE, include=FALSE, message=FALSE>>=
-
-# Calculate the number of parameters that are estimated:
-mod3_n_params <- length(c(
-  "mu_B", "XthetaA", "thetaI", "lambdaR", "r", "std_W", "epsilon", "k",
-  paste0('betaB', 1:10), paste0('foi_add', 1:10),
-  'aHur3', 'aHur9',
-  'hHur3', 'aHur9',
-  'Iinit3', 'Iinit4'
-))
-
-# Load the file containing profiles of Model 1 Parameters.
-if (RUN_LEVEL != 3 && file.exists("model3/run_level_3/h3_MLE_eval.rds")) {
-
-  # If RUN_LEVEL != 3, and RUN_LEVEL == 3 results exist, just load the RL3
-  # results.
-  h3_prof <- readRDS("model3/run_level_3/h3_MLE_eval.rds")
-} else if (file.exists(paste0('model3/', rl_dir, 'h3_MLE_eval.rds'))) {
-
-  # If the `model3/run_level_<RUN_LEVEL>/h3_profiles.rds` file exists, load it
-  h3_prof <- readRDS(paste0('model3/', rl_dir, 'h3_MLE_eval.rds'))
-} else {
-
-  # Otherwise (missing RL3 file, and the corresponding <RUN_LEVEL> file):
-  # Throw an error, letting users know how to create the file.
-  stop(paste0("Missing file: `model3/", rl_dir, 'h3_MLE_eval.rds`;\nThe file can be obtained by running the script `H3_MLE_Eval.R`.'))
-}
-@
-
-<<Create Model Objects,echo=FALSE,eval=TRUE, include=FALSE, message=FALSE>>=
-
-# Save the parameters that correspond to the MLE of Model 3.
-p3 <- h3_prof %>%
-  slice_max(logLik, n = 1) %>%
-  select(-(job.id:logLik_se)) %>%
-  unlist()
-
-unit_specific_names <- c("betaB", "foi_add", "aHur", "hHur", "Iinit")
-shared_param_names <- c(
-  "mu_B", "XthetaA", "thetaI", "lambdaR", "r", "std_W",
-  "epsilon", "k"
-)
-
-p3u <- matrix(nrow = length(unit_specific_names), ncol = 10)
-rownames(p3u) <- unit_specific_names
-p3s <- numeric(length(shared_param_names))
-names(p3s) <- shared_param_names
-
-for (p in shared_param_names) {
-  p3s[p] <- p3[paste0(p, 1)]  # They are the same for 1:10
-}
-
-for (p in unit_specific_names) {
-  pat <- paste0("^", p, "[[:digit:]]+$")
-  p3u[p, ] <- p3[grepl(pat, names(p3))]
-}
-
-rm(unit_specific_names, shared_param_names, p, pat)
-
-stew(file = "models.rda", {
-  h1 <- haiti1_joint()
-  coef(h1) <- p1
-
-  h2_epi <- haiti2(region = 'before', cutoff = 10000, measure = 'log')
-  coef(h2_epi) <- h2_fit$h2_params
-
-  h3Spat <- haiti3_spatPomp()
-  coef(h3Spat) <- p3
-})
-@
-
-<<Get AIC table Input, echo=FALSE, eval=TRUE>>=
-NREPS_EVAL <- switch(RUN_LEVEL,   3,   8,    36)
-NP_EVAL    <- switch(RUN_LEVEL,  50, 500,  5000)
-
-mod2_ll <- h2_fit$logLik
-
-arima_lik2 <- 0
-
-# model 2 AIC
-mod2_n_params <- h2_fit$n_fit_params
-mod2_aic <- 2 * mod2_n_params - 2 * mod2_ll
-
-# model 1 AIC
-mod1_n_params <- length(
-  c('beta1', 'beta2', 'beta3', 'beta4', 'beta5', 'beta6',
-    'rho', 'nu', 'E_0', 'I_0', 'tau_epi',
-    'tau_end', 'sig_sq_epi', 'sig_sq_end', 'betat')
-)
-mod1_aic <- 2 * mod1_n_params - 2 * mod1_ll
-
-# Calculate likelihood via block-particle filter
-mod3_eval_res <- bake(
-  file = paste0('model3/', rl_dir, 'mod3_eval_res.rds'), {
-
-    # Create a list to store results
-    results <- list()
-
-    coef(h3Spat) <- p3
-
-    # Perform block-particle filter
-    h3_bpf <- foreach(i = 1:NREPS_EVAL, .combine = c) %dopar% {
-      bpfilter(h3Spat, Np = NP_EVAL, block_size = 1)
-    }
-
-    ll <- logLik(h3_bpf)
-    results$ll <- logmeanexp(ll[!is.na(ll)], se = TRUE)
-
-    u_cond_ll <- t(sapply(h3_bpf, function(x) rowSums(x@block.cond.loglik)))
-    results$u_cond_ll <- apply(u_cond_ll, 2, logmeanexp, se = TRUE)
-
-    ### Get likelihood for subset of the data
-
-    # Find appropriate sub-set
-    in_subset_cols <- h3_bpf[[1]]@times >= as.numeric(lubridate::decimal_date(as.Date("2014-03-01")))
-    in_subset_rows <- h3_bpf[[1]]@unit_names != 'Ouest'
-
-    mod3_coupled_subset_evals <- c()
-    for (i in 1:length(h3_bpf)) {
-      mod3_coupled_subset_evals <- c(
-        mod3_coupled_subset_evals,
-        sum(h3_bpf[[i]]@block.cond.loglik[in_subset_rows, in_subset_cols])
-      )
-    }
-
-    results$subset_ll <- logmeanexp(mod3_coupled_subset_evals, se = TRUE)
-    results
-  },
-  timing = FALSE
-)
-
-mod3_ll <- mod3_eval_res$ll[1]
-mod3_aic <- 2 * mod3_n_params - 2 * mod3_ll
-mod3_eval_res$subset_aic <- 2 * mod3_n_params - 2 * mod3_eval_res$subset_ll['se']
-
-rm(NREPS_EVAL, NP_EVAL)
-@
-
-<<Calculate Bench Likelihood, echo=FALSE, message=FALSE, include=FALSE, cache=TRUE>>=
-
-benchmarks <- list()
-
-###
-### Model 1 ###
-###
-
-# Load aggregated data
-m1_agg_data <- haiti1_agg_data()
-
-# Fit Model 1 benchmark:
-m1_bench <- ar1_NegBinom(m1_agg_data$cases, init = 7364)
-benchmarks[['m1']] <- m1_bench$ll
-
-rm(m1_agg_data, m1_bench)
-gc()
-
-###
-### Model 2 ###
-###
-
-h2_data <- haiti2_data(recreate = FALSE)
-
-# This are initial values for Y_u(0). They were estimated, but because they are
-# integers, close to the value of the first observation, it was easiest to fit these
-# by hand.
-m2_inits <- c(6449, 542, 0, 0, 16, 0, 0, 522, 0, 0)  # fit by hand
-
-# model 2 ARIMA benchmark
-h2_dep_names <- h2_data$department %>% unique()
-m2_benchmark <- 0
-for (i in 1:10){
-  cases <- h2_data[h2_data$department == h2_dep_names[i], ]$cases
-  m2_bench <- ar1_NegBinom(data = cases, init = m2_inits[i])
-
-  m2_benchmark <- m2_benchmark + m2_bench$ll
-}
-
-benchmarks[['m2']] <- m2_benchmark
-
-rm(m2_benchmark, h2_dep_names, cases, h2_data, m2_bench, i)
-gc()
-
-###
-### Model 3 ###
-###
-
-mod3_cases <- as.data.frame(t(h3Spat@data))
-m3_inits <- c(7362, 1914, 22, 8, 2818, 82, 2499, 6953, 44, 7) # fit by hand
-
-m3_benchmark <- 0
-for (dep in 1:10) {
-
-  # Get department cases
-  cases <- as.numeric(mod3_cases[, dep])
-
-  m3_benchmark <- m3_benchmark + ar1_NegBinom(data = cases, init = m3_inits[dep])$ll
-}
-
-benchmarks[['m3']] <- m3_benchmark
-rm(cases, m3_benchmark, dep, mod3_cases)
-gc()
-@
-
-<<Computing AIC of Lee et al, echo=FALSE, include=FALSE, message=FALSE>>=
-lee3_NP   <- switch(RUN_LEVEL, 50, 250, 5000)
-lee3_NREP <- switch(RUN_LEVEL,  3,   8,   36)
-
-mod1_lee_n_params <- 20
-mod1_lee_aic <- 2 * mod1_lee_n_params - 2 * mod1_lee_ll
-
-mod2_lee_aic <- 2 * unname(lee20_replication['leeFit_n_params']) - 2 * unname(lee20_replication['leeFit_ll'])
-
-registerDoRNG(49637421)
-
-# Create Lee et al (2020a) version of model 3 on sub-data
-lee3_sub <- lee3_spatPomp()
-lee3_all <- lee3_spatPomp(start_date = "2010-10-23")
-
-lee3_results <- bake(
-  file = paste0('model3/', rl_dir, 'lee3_results.rds'), {
-
-    # Create lists to save results
-    lee3_results <- list()
-
-    # Save how long it takes for sub-data
-    lee3_results[["sub_bpf_time"]] <- system.time({
-
-      lee3_sub_ll_evals <- foreach(
-        i = 1:lee3_NREP, .combine = rbind
-      ) %dopar% {
-
-        liks <- numeric(10L)
-
-        lee_bpf_out <- bpfilter(lee3_sub, block_size = 1, Np = lee3_NP)
-        Ouest_ind <- which(lee_bpf_out@unit_names == "Ouest")
-
-        for (u in 1:10) {
-          if (u == Ouest_ind) {
-            liks[u] <- sum(lee_bpf_out@block.cond.loglik[u, lee3_sub@times >= lubridate::decimal_date(as.Date('2017-06-10'))])
-          } else {
-            liks[u] <- sum(lee_bpf_out@block.cond.loglik[u, ])
-          }
-        }
-
-        liks
-      }
-    })
-
-    # Save all likelihood evaluations
-    lee3_results[["sub_ll_eval"]] <- logmeanexp(rowSums(lee3_sub_ll_evals)[!is.na(rowSums(lee3_sub_ll_evals))])
-
-    registerDoRNG(2789131)
-
-    # Save how long it takes for all data
-    lee3_results[["all_pf_time"]] <- system.time({
-      # Save all likelihood evaluations
-
-      lee3_all_ll_evals <- foreach(
-        i = 1:lee3_NREP, .combine = c
-      ) %dopar% {
-        logLik(bpfilter(lee3_all, block_size = 1, Np = lee3_NP))  # Get likelihood evaluations
-      }
-
-      lee3_results[["all_ll_eval"]] <- logmeanexp(lee3_all_ll_evals[!is.na(lee3_all_ll_evals)])
-
-    })
-
-    lee3_results[["n_params"]] <- 29
-
-    lee3_results
-  },
-  timing = FALSE
-)
-
-rm(lee3_sub, lee3_all, lee3_NP, lee3_NREP)
-gc()
-
-mod3_lee_coupled_aic <- 2 * lee3_results$n_params - 2 * lee3_results$all_ll_eval
-mod3_lee_coupled_partial_aic <- 2 * lee3_results$n_params - 2 * lee3_results$sub_ll_eval
-@
-
 \section*{Materials and methods}
@@
 Infection {\small (day)}
-  & ${\muIR^{-1}=\Sexpr{myround(7/p1["gamma"],1)}}\fixed$
+  & ${\muIR^{-1}=2.0}\fixed$
   & \eqref{model1:toR}
-  & ${\muIR^{-1}=\Sexpr{myround(365/h2_fit$h2_params["gamma"],1)}}\fixed $
+  & ${\muIR^{-1}=7.0}\fixed $
   & \eqref{model2:mu_IR}
-  & ${\muIR^{-1}=\Sexpr{myround(365/p3["gamma1"],1)}}\fixed $
+  & ${\muIR^{-1}=5.0}\fixed $
   & \eqref{eq:model3:IR}
@@
 Latency {\small (day)}
-  & ${\muEI^{-1}=\Sexpr{myround(7/p1["sigma"],1)}}\fixed$
+  & ${\muEI^{-1}=1.4}\fixed$
   & \eqref{model1:EA}
-  & ${\muEI^{-1}=\Sexpr{myround(365/h2_fit$h2_params["gammaE"],1)}}\fixed$
+  & ${\muEI^{-1}=1.3}\fixed$
   & \eqref{model2:mu_EI}
@@
   & $\transmission_{1:6}=(
-      \Sexpr{myround(p1["beta1"],1)},\allowbreak
-      \Sexpr{myround(p1["beta2"],1)},\allowbreak
-      \Sexpr{myround(p1["beta3"],1)},\allowbreak
-      \Sexpr{myround(p1["beta4"],1)},\allowbreak
-      \Sexpr{myround(p1["beta5"],1)},\allowbreak
-      \Sexpr{myround(p1["beta6"],1)})$
+      1.4,\allowbreak
+      1.2,\allowbreak
+      1.1,\allowbreak
+      1.1,\allowbreak
+      1.4,\allowbreak
+      1.0)$
   & \eqref{model1:beta}
@@
   & \multirow{2}{*}{\eqref{model2:lambda}}
-  & \multirow{2}{*}{$\seasAmplitude = \Sexpr{myround(p3s["lambdaR"],2)}$}
+  & \multirow{2}{*}{$\seasAmplitude = 1.00$}
   & \multirow{2}{*}{\eqref{eq:model3:water}}
   \\
-  & $\transmissionTrend = \Sexpr{myround(p1["betat"],2)}^*$
+  & $\transmissionTrend = -0.04^*$
   & \eqref{model1:betat}
-  & \multirow{-2}{*}{$\phaseParm = \Sexpr{myround(h2_fit$h2_params['phase'], 2)}^*$}
+  & \multirow{-2}{*}{$\phaseParm = 0.97^*$}
   & \multirow{-2}{*}{\eqref{model2:lambda}}
-  & \multirow{-2}{*}{$\rainfallExponent = \Sexpr{signif(p3s["r"],3)}$}
+  & \multirow{-2}{*}{$\rainfallExponent = 0.78$}
   & \multirow{-2}{*}{\eqref{eq:model3:water}}
@@
   \multirow{3}{*}{Immunity {\small (yr)}}
-  & \multirow{3}{*}{${\muRS^{-1}=\Sexpr{myround(1/p1["alpha"]/52,1)}}\fixed$}
+  & \multirow{3}{*}{${\muRS^{-1}=8.0}\fixed$}
   & \multirow{3}{*}{\eqref{model1:RS}}
-  & $\muRS^{-1}= \Sexpr{signif(1/h2_fit$h2_params["sigma"],2)}$
+  & $\muRS^{-1}= \ensuremath{1.4\times 10^{11}}$
   & \eqref{model2:RS}
-  & \multirow{3}{*}{${\muRS^{-1}=\Sexpr{myround(1/p3['rho1'],1)}}\fixed$}
+  & \multirow{3}{*}{${\muRS^{-1}=8.0}\fixed$}
   & \multirow{3}{*}{\eqref{eq:model3:RRnext}}
 \\
-& & & ${\omega_1^{-1} = \Sexpr{myround(1/h2_fit$h2_params["Omega1"],1)}}\fixed$ & \eqref{model2:omega1} & &\\
-& & & ${\omega_2^{-1} = \Sexpr{myround(1/h2_fit$h2_params["Omega2"],1)}}\fixed$ & \eqref{model2:omega2} & &\\
+& & & ${\omega_1^{-1} = 1.0}\fixed$ & \eqref{model2:omega1} & &\\
+& & & ${\omega_2^{-1} = 5.0}\fixed$ & \eqref{model2:omega2} & &\\
 \hline
 \multirow{2}{*}{Birth/death {\small (yr\textsuperscript{-1})}}
-  & ${\muBirth =  10^{-2}\times \Sexpr{myround(100 * 52 * p1["mu"],2)}}\fixed$
+  & ${\muBirth =  10^{-2}\times 2.23}\fixed$
     %
@@
   &
-  & ${\muDeath = 10^{-2}\times\Sexpr{myround(100 * p3["mu1"],2)}}\fixed$
+  & ${\muDeath = 10^{-2}\times1.59}\fixed$
   & \multirow{2}{*}{\eqref{eq:model3:IS}}
 \\
-& ${\muDeath = 10^{-3}\times\Sexpr{myround(1000*52*p1["delta"],1)}}\fixed$
+& ${\muDeath = 10^{-3}\times7.5}\fixed$
 & \eqref{model1:death}
@@
 &
-& ${\choleraDeath = \Sexpr{myround(p3["alpha1"],2)}}\fixed$
+& ${\choleraDeath = 1.46}\fixed$
 &
@@
   & (\ref{model1:EI}-\ref{model1:EA})
-  & ${\symptomFrac=\Sexpr{myround(h2_fit$h2_params["k"],1)}}\fixed$
+  & ${\symptomFrac=0.2}\fixed$
   & \eqref{model2:mu_EI}
-  & ${\symptomFrac=\Sexpr{myround(p3["sigma1"],2)}}\fixed$
+  & ${\symptomFrac=0.25}\fixed$
   & \eqref{eq:model3:SA}
@@
 &
-& ${\asymptomRelativeShed = \Sexpr{1e-7}}\fixed$
+& ${\asymptomRelativeShed = \ensuremath{10^{-7}}}\fixed$
 & \eqref{model2:to_W}
-& $\asymptomRelativeShed = \Sexpr{myround(p3s["XthetaA"],3)}$
+& $\asymptomRelativeShed = 0.008$
 & \eqref{eq:model3:water}
@@
   & \multirow{4}{*}{\eqref{model1:lambda}}
-  & \multirow{4}{*}{$\transmission=$\Sexpr{signif(h2_fit$h2_params["Beta"],3)} {\small yr\textsuperscript{-1}}}
+  & \multirow{4}{*}{$\transmission=$\ensuremath{5.97\times 10^{-15}} {\small yr\textsuperscript{-1}}}
   & \multirow{4}{*}{\eqref{model2:lambda}}
   & $\transmission_{1:10}=(
-  \Sexpr{myround(p3u["foi_add",1]*1e6,2)}, \allowbreak
-	\Sexpr{myround(p3u["foi_add",2]*1e6,2)}, \allowbreak
-	\Sexpr{myround(p3u["foi_add",3]*1e6,2)}, \allowbreak
-  \Sexpr{myround(p3u["foi_add",4]*1e6,2)}, \allowbreak
-	\Sexpr{myround(p3u["foi_add",5]*1e6,2)}, \allowbreak
-	\Sexpr{myround(p3u["foi_add",6]*1e6,2)}, \allowbreak
-	\Sexpr{myround(p3u["foi_add",7]*1e6,2)}, \allowbreak
-	\Sexpr{myround(p3u["foi_add",8]*1e6,2)}, \allowbreak
-	\Sexpr{myround(p3u["foi_add",9]*1e6,2)}, \allowbreak
-	\Sexpr{myround(p3u["foi_add",10]*1e6,2)}
+  0.82, \allowbreak
+	0.02, \allowbreak
+	0.38, \allowbreak
+  0.21, \allowbreak
+	0.51, \allowbreak
+	0.51, \allowbreak
+	0.35, \allowbreak
+	0.12, \allowbreak
+	0.26, \allowbreak
+	0.10
 	) \allowbreak \times 10^{-6}$ {\small yr\textsuperscript{-1}}
@@
   &
-  & \multirow{3}{*}{${\Wsat = \Sexpr{signif(h2_fit$h2_params["Sat"],2)}}\fixed$}
+  & \multirow{3}{*}{${\Wsat = \ensuremath{10^{5}}}\fixed$}
   & \multirow{3}{*}{\eqref{model2:lambda}}
   & $\Wbeta{_{1:10}}= (
-        \Sexpr{myround(p3u["betaB",1],2)}, \allowbreak
-        \Sexpr{myround(p3u["betaB",2],2)}, \allowbreak
-        \Sexpr{myround(p3u["betaB",3],2)}, \allowbreak
-        \Sexpr{myround(p3u["betaB",4],2)}, \allowbreak
-        \Sexpr{myround(p3u["betaB",5],2)}, \allowbreak
-        \Sexpr{myround(p3u["betaB",6],2)}, \allowbreak
-        \Sexpr{myround(p3u["betaB",7],2)}, \allowbreak
-        \Sexpr{myround(p3u["betaB",8],2)}, \allowbreak
-        \Sexpr{myround(p3u["betaB",9],2)}, \allowbreak
-        \Sexpr{myround(p3u["betaB",10],2)})$ {\small yr\textsuperscript{-1}}
+        4.70, \allowbreak
+        21.00, \allowbreak
+        24.97, \allowbreak
+        27.14, \allowbreak
+        5.28, \allowbreak
+        30.70, \allowbreak
+        10.17, \allowbreak
+        0.99, \allowbreak
+        11.89, \allowbreak
+        12.82)$ {\small yr\textsuperscript{-1}}
   & \multirow{4}{*}{\eqref{eq:model3:foi}}
 \\
-& & & \multirow{-3}{*}{$\beta_W= \Sexpr{signif(h2_fit$h2_params["BetaW"],3)}$ {\small yr\textsuperscript{-1}}} & \multirow{-3}{*}{\eqref{model2:lambda}} & &
+& & & \multirow{-3}{*}{$\beta_W= 1.1$ {\small yr\textsuperscript{-1}}} & \multirow{-3}{*}{\eqref{model2:lambda}} & &
 \\
@@
   &
-  & $\Wshed = $ \Sexpr{signif(h2_fit$h2_params["Mu"]/52,3)} {\small wk\textsuperscript{-1}}
+  & $\Wshed = $ 179 {\small wk\textsuperscript{-1}}
   & \eqref{model2:to_W}
-  & $\Wshed= \Sexpr{signif(p3s["thetaI"]/52,3)}$ {\tiny	$\frac{\mathrm{km^2}}{\mathrm{wk}}$}
+  & $\Wshed= \ensuremath{9.77\times 10^{-7}}$ {\tiny	$\frac{\mathrm{km^2}}{\mathrm{wk}}$}
   & \eqref{eq:model3:water}
@@
   &
-  & ${\Wremoval^{-1} = \Sexpr{signif(52/h2_fit$h2_params["Delta"],3)}}\fixed$
+  & ${\Wremoval^{-1} = 3}\fixed$
   & \eqref{model2:from_W}
-  & $\Wremoval^{-1}=\Sexpr{myround(52/p3s["mu_B"],2)}$
+  & $\Wremoval^{-1}=0.11$
   & \eqref{eq:model3:Decay}
@@
 Mixing exponent
-  & $\mixExponent=\Sexpr{myround(p1["nu"],2)}$
+  & $\mixExponent=0.98$
   & \eqref{model1:lambda}
@@
 \raggedright Process noise {\small (wk\textsuperscript{1/2})}
-  & \raggedright $\sigmaProc = (\Sexpr{myround(p1["sig_sq_epi"],2)},
-  \Sexpr{myround(p1["sig_sq_end"],2)})^*$
+  & \raggedright $\sigmaProc = (0.09,
+  0.12)^*$
   & \multirow{2}{*}{\eqref{model1:lambda}}
@@
   &
-  & \multirow{2}{*}{$\sigmaProc= \Sexpr{myround(p3s["std_W"] * sqrt(52),3)}$}
+  & \multirow{2}{*}{$\sigmaProc= 0.218$}
   & \multirow{2}{*}{\eqref{eq:model3:SA}}
@@
 Reporting rate
-  & $\reportRate=\Sexpr{myround(p1["rho"], 3)}$
+  & $\reportRate=0.679$
   & (\nameref{S_meas})
-  & ${\reportRate=\Sexpr{myround(h2_fit$h2_params["Rho"],2)}}\fixed$
+  & ${\reportRate=0.20}\fixed$
   & (\nameref{S_meas})
-  & $\reportRate=\Sexpr{myround(p3s["epsilon"],2)}$
+  & $\reportRate=0.98$
   & (\nameref{S_meas})
@@
 \raggedright Observation variance
-  & \raggedright $\obsOverdispersion = (\Sexpr{myround(p1["tau_epi"],2)}, \Sexpr{round(p1["tau_end"],2)})$
+  & \raggedright $\obsOverdispersion = (279.15, 78.33)$
   & \multirow{2}{*}{(\nameref{S_meas})}
-  & \multirow{2}{*}{$\obsOverdispersion = \Sexpr{myround(h2_fit$h2_params['v'], 3)}$}
+  & \multirow{2}{*}{$\obsOverdispersion = 1.319$}
   & \multirow{2}{*}{(\nameref{S_meas})}
-  & \multirow{2}{*}{$\obsOverdispersion = \Sexpr{myround(p3s["k"],2)}$}
+  & \multirow{2}{*}{$\obsOverdispersion = 88.58$}
   & \multirow{2}{*}{(\nameref{S_meas})}
@@
 \hline
-\multirow{2}{*}{Initial Values} & $\Iinit = \Sexpr{round(prod(p1[c('pop_0', 'I_0')]))}$
+\multirow{2}{*}{Initial Values} & $\Iinit = 7298$
   &
@@
   & \multirow{2}{*}{$\Iinit^{3,4} = (
-        \Sexpr{round(prod(p3[c('H3', 'Iinit3')]))}, \Sexpr{round(prod(p3[c('H4', 'Iinit4')]))})^*$}
+        21, 6)^*$}
   &  \multirow{2}{*}{(\nameref{S_init})}
   \\
-  & $E_{0,0} = \Sexpr{round(prod(p1[c('pop_0', 'E_0')]))}$ & & \multicolumn{1}{c}{}
+  & $E_{0,0} = 350$ & & \multicolumn{1}{c}{}
   &
@@
   &
-  & \raggedright $\Whur{3, 9} = (\Sexpr{myround(p3['aHur3'], 2)}, \Sexpr{myround(p3['aHur9'], 2)})^*$
+  & \raggedright $\Whur{3, 9} = (36.88, 31.64)^*$
   & \multirow{2}{*}{\eqref{eq:model3:foi}}
@@
   &
-  & \raggedright $\hHur{3, 9} = (\Sexpr{myround(p3['hHur3'], 2)}, \Sexpr{myround(p3['hHur9'], 2)})^*$
+  & \raggedright $\hHur{3, 9} = (98.98, 58.43)^*$
   & \multirow{2}{*}{\eqref{eq:model3:foi}}
@@
 \multirow{2}{*}{Log-likelihood} &
-  $\Sexpr{myround(mod1_ll, 1)}$ &
-  $\Sexpr{myround(mod2_ll, 1)}$ &
-  $\Sexpr{myround(mod3_ll, 1)}$ \\
-    & ($\Sexpr{myround(mod1_lee_ll, 1)}$)\footnotemark[1] &
-  ($\Sexpr{myround(unname(lee20_replication['leeFit_ll']), 1)}$) &
-  ($\Sexpr{myround(lee3_results[["all_ll_eval"]], 1)}$)\footnotemark[2]
+  $-2728.1$ &
+  $-21957.3$ &
+  $-17332.9$ \\
+    & ($-3030.9$)\footnotemark[1] &
+  ($-29367.4$) &
+  ($-33832.6$)\footnotemark[2]
 \\
@@
 Number of &
-  $\Sexpr{as.character(mod1_n_params)}$ &
-  $\Sexpr{as.character(mod2_n_params)}$ &
-  $\Sexpr{as.character(mod3_n_params)}$ \\
+  $15$ &
+  $6$ &
+  $34$ \\
  Fit Parameters &
- (\Sexpr{as.character(mod1_lee_n_params)}) &
- ($\Sexpr{as.character(unname(lee20_replication['leeFit_n_params']))}$) &
- ($\Sexpr{as.character(lee3_results$n_params, 1)}$)
+ (20) &
+ ($6$) &
+ ($29$)
 \\
@@
 \multirow{2}{*}{AIC} &
-  $\Sexpr{myround(mod1_aic, 1)}$ &
-  $\Sexpr{myround(mod2_aic, 1)}$ &
-  $\Sexpr{myround(mod3_aic, 1)}$ \\
-  & ($\Sexpr{myround(mod1_lee_aic, 1)}$)\footnotemark[1] &
-  ($\Sexpr{myround(mod2_lee_aic, 1)}$) &
-  ($\Sexpr{myround(mod3_lee_coupled_aic, 1)}$)\footnotemark[2]
+  $5486.3$ &
+  $43926.5$ &
+  $34733.9$ \\
+  & ($6101.8$)\footnotemark[1] &
+  ($58746.9$) &
+  ($67723.2$)\footnotemark[2]
 \\
@@
 \thead{Benchmark \\ AIC} &
-  $\Sexpr{myround(-2 * benchmarks$m1 + 2 * 4, 1)}$ &
-  $\Sexpr{myround(-2 * benchmarks$m2 + 2 * 40, 1)}$ &
-  $\Sexpr{myround(-2 * benchmarks$m3 + 2 * 40, 1)}$
+  $5585.3$ &
+  $36961.0$ &
+  $35945.2$
 \\
@@
 \old{Lee et al.~[4] fit} \textsuperscript{2}\new{In \cite{lee20},} Model~3 \new{was fit} to a subset of the data (March 2014 onward, excluding data from Ouest in 2015-2016).
-On this subset, their model has a likelihood of $\Sexpr{myround(lee3_results$sub_ll_eval, 1)}$.
-On this same subset, our model has a likelihood of $\Sexpr{myround(mod3_eval_res$subset_ll['est'], 1)}$.
+On this subset, their model has a likelihood of $-9721.2$.
+On this same subset, our model has a likelihood of $-7219.5$.
 Details of estimating the likelihood of the \old{Lee et al.~[4]} models \new{used in \cite{lee20} are} \old{is} provided in \new{the supplement} (\nameref{S_lee20}).
@@
 We similarly allow the re-estimation of process and measurement overdispersion parameters ($\sigmaProc^2$ and $\obsOverdispersion$), and require that the latent Markov process $X(t)$ carry over from one phase into the next.
-The resulting 95\% confidence interval for $\transmissionTrend$ is $(\Sexpr{myround(beta_t_mcap_results$ci[1], digits = 3)}, \Sexpr{myround(beta_t_mcap_results$ci[2], digits = 3)})$, with the full results displayed in Fig.~\ref{fig:betat}.
+The resulting 95\% confidence interval for $\transmissionTrend$ is $(-0.098, -0.009)$, with the full results displayed in Fig.~\ref{fig:betat}.
 These results are suggestive that the inclusion of a trend in the transmission rate improves the quantitative ability of Model~1 to describe the observed data.
-\new{The maximum likelihood estimate for $\transmissionTrend$ corresponds to a \Sexpr{paste0(myround(100 - 100*exp(beta_t_mcap_results$mle * 430/215), 1), '\\%')} reduction to the transmission rate over the course of the outbreak, with a $95\%$ confidence interval of (\Sexpr{paste0(rev(myround(100 - 100*exp(beta_t_mcap_results$ci * 430/215), 1)), collapse = ', ', sep = '\\%')}) for the overall reduction in transmission.}
+\new{The maximum likelihood estimate for $\transmissionTrend$ corresponds to a 7.3\% reduction to the transmission rate over the course of the outbreak, with a $95\%$ confidence interval of (1.8\%, 17.9\%) for the overall reduction in transmission.}
 The reported results for Model~1 in the remainder of this article were obtained with the inclusion of the parameter $\transmissionTrend$.
@@
 \centering
-<<Beta_trend_Figure, fig.height=2.4, fig.width=3.6, fig.align='center'>>=
-ggplot() +
-  geom_point(data = h1_prof_res, aes(x = betat, y = logLik)) +
-  geom_line(data = beta_t_mcap_results$fit, aes(x = parameter, y = smoothed), col = 'blue') +
-  # geom_line(data = beta_t_mcap_results$fit, aes(x = parameter, y = quadratic), col = 'red') +
-  geom_vline(xintercept = beta_t_mcap_results$ci[1], linetype = 'dashed') +
-  geom_vline(xintercept = beta_t_mcap_results$ci[2], linetype = 'dashed') +
-  geom_vline(xintercept = beta_t_mcap_results$mle, col = 'blue') +
-  labs(x = "Log-Linear Trend in Transmission", y = 'Log Likelihood') +
-  theme(axis.text = element_text(size = 8),
-        axis.title = element_text(size = 10))
-@
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
+
+{\centering \includegraphics[width=\maxwidth]{figure/Beta_trend_Figure-1}
+
+}
+
+\end{knitrout}
 \caption{\label{fig:betat}
@@
 \end{figure}
-
-<<clean Up of Betat data, echo=FALSE, message=FALSE, include=FALSE, warning=FALSE>>=
-suppressWarnings(
-  rm(h1_prof_res, beta_t_mcap_results)
-)
-gc()
-@
 
@@
 \centering
-<<Model_1_Sims_Figure, fig.height=2.35, fig.width=4.6, fig.align='center'>>=
-# This code chunk creates the figure containing simulations from Model 1
-
-# Simulate Model 1 from initial conditions.
-sims <- simulate(
-  h1, nsim = switch(RUN_LEVEL, 50, 100, 500),
-  format = 'data.frame', seed = 3448931
-)
-
-# Calculate 0.025 and 0.975 percentiles of simulations, at each time point.
-quants <- sims %>%
-  mutate(is_data = .id == 'data') %>%
-  filter(!is_data) %>%
-  select(.id, week, cases) %>%
-  group_by(week) %>%
-  summarize(
-    q025 = quantile(cases, probs = 0.025, na.rm = TRUE),
-    q50  = quantile(cases, probs = 0.500, na.rm = TRUE),
-    q975 = quantile(cases, probs = 0.975, na.rm = TRUE)
-  ) %>%
-  ungroup() %>%
-  mutate(date = lubridate::ymd("2010-10-16") + lubridate::weeks(week))
-
-# Plot the data, median value of the simulation, and estimated quantiles.
-ggplot() +
-  geom_line(data = true_agg_cases, aes(x = date, y = ReportedAll + 1)) +
-  geom_line(data = quants, aes(x = date, y = q50 + 1), col = 'blue') +
-  geom_ribbon(  # Plot confidence region
-    data = quants,
-    aes(x = date, ymin = q025 + 1, ymax = q975 + 1),
-    alpha = 0.5
-  ) +
-  theme(
-    axis.title.x = element_blank(),
-    axis.text = element_text(size = 8),
-    axis.title.y = element_text(size = 10)
-  )+
-  ylab('Reported cholera cases') +
-  geom_vline(  # Show change from epidemic -> endemic models.
-    xintercept = lubridate::weeks(232) + lubridate::ymd("2010-10-16"),
-    linetype = 'dashed'
-  ) +
-  scale_y_log10(  # Plot on the log-scale
-    labels = scales::trans_format("log10", scales::math_format(10^.x))
-  ) +
-  scale_x_date(  # Format the date on x-axis.
-    date_labels = "'%y",
-    breaks = seq.Date(
-      from = as.Date("2011-01-01"),
-      as.Date("2019-01-01"),
-      by = '1 year'
-    )
-  )
-@
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
+
+{\centering \includegraphics[width=\maxwidth]{figure/Model_1_Sims_Figure-1}
+
+}
+
+\end{knitrout}
 \caption{\label{fig:mod1fit}
@@
 \end{figure}
-
-<<Clean-up Model 1 Simulations, echo=FALSE, include=FALSE>>=
-rm(sims, quants)
-gc()
-@
 
@@
 
-<<Model 2 Trajectory, echo=FALSE>>=
-
-# This code chunk calculates the "trajectory" of the differential equations
-# for Model 2.
-
-h2_epi_traj <- trajectory(
-  h2_epi,  # Model
-  params = h2_fit$h2_params,  # estimated parameters
-  format = 'data.frame'
-)
-
-# Display the output as nationally aggregated data, as was done in Lee et al (2020)
-h2_epi_traj$Ctotal <- rowSums(
-  h2_epi_traj[, paste0("C", 1:10)]  # Infections in each department
-) * h2_fit$h2_params['Rho']  # multiply by reporting rate, to compare to reported cases.
-
-h2_traj <- h2_epi_traj %>%
-  select(year, Ctotal) %>%
-  mutate(
-    date = as.Date(lubridate::round_date(lubridate::date_decimal(year), unit = 'day'))
-  )
-@
-
-<<Load Model 2 VaccScenarios, message=FALSE, include=FALSE>>=
-
-# Run Vaccination scenarios for Model 2. Note that the code runs the same
-# speed for each run-level.
-mod2_VaccScenarios <- bake(
-  file = paste0("model2/VaccinationScenarios.rds"), {
-    haiti2_vaccScenario(
-      h2_params = h2_fit$h2_params
-    )
-  },
-  timing = FALSE
-)
-@
-
 \begin{figure}[!h]
-<<Mod2Fit_and_Scenarios_Figure, fig.height=1.9, fig.width=4.8, fig.align='center'>>=
-ggplot() +
-  geom_line(data = h2_traj, aes(x = date, y = Ctotal + 1), col = 'blue') +
-  geom_ribbon(
-    data = h2_traj,
-    aes(
-      x = date,
-      ymin = (Ctotal + 1) * exp(-1.96*h2_fit$h2_params['v']),
-      ymax = (Ctotal + 1) * exp(1.96*h2_fit$h2_params['v'])
-    ), alpha = 0.5
-  ) +
-  geom_line(data = true_agg_cases, aes(x = date, y = ReportedAll + 1)) +
-  geom_line(
-    data = filter(mod2_VaccScenarios$all_trajs, year >= max(h2_traj$year)),
-    aes(x = as.Date(date), y = ReportedCases + 1, color = scenario)
-  ) +
-  geom_ribbon(
-    data = filter(mod2_VaccScenarios$all_trajs, year >= max(h2_traj$year)) %>% filter(scenario == 'V0'),
-    aes(
-      x = as.Date(date),
-      ymin = pmax((ReportedCases + 1) * exp(-1.96*h2_fit$h2_params['v']), 1),
-      ymax = pmax((ReportedCases + 1) * exp(1.96*h2_fit$h2_params['v']), 1)
-    ), linetype = 'dashed', alpha = 0.3
-  ) +
-  theme(axis.title.x = element_blank(),
-        legend.title = element_blank(),
-        axis.title.y = element_text(size = 9),
-        axis.text = element_text(size = 8)) +
-  ylab("Reported cholera cases") +
-  scale_color_manual(values = c("V0" = '#f4a582',
-                                "V1" = '#d6604d',
-                                "V2" = '#b2182b',
-                                "V3" = '#92c5de',
-                                "V4" = '#4393c3')) +
-  scale_y_log10(labels = scales::trans_format("log10", scales::math_format(10^.x))) +
-  scale_x_date(date_labels = "'%y", breaks = seq.Date(from = as.Date("2010-01-01"), as.Date("2030-01-01"), by = '2 years'))
-@
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
+
+{\centering \includegraphics[width=\maxwidth]{figure/Mod2Fit_and_Scenarios_Figure-1}
+
+}
+
+\end{knitrout}
 \caption{\label{fig:mod2Traj}
@@
 
-<<Model 3 SpatPOMP Sims, echo=FALSE>>=
-
-# Number of simulations from Model 3. Because it's high-dimensional, we include
-# a run level to avoid time/memory issues when editing. Final results shown
-# with RUN_LEVEL = 3.
-
-h3_nsim <- switch(RUN_LEVEL, 20, 100, 500)
-
-# Get quantiles of simulations from the calibrated Model 3.
-h3_quants <- bake(
-  file = paste0('model3/', rl_dir, 'h3Sims.rds'), {
-
-    coef(h3Spat) <- p3
-    # Simulate from the fitted version of Model 3
-    h3_sims <- simulate(
-      h3Spat, nsim = h3_nsim, format = 'data.frame'
-    )
-
-    # For each department and time-point, calculate percentiles of the
-    # simulations
-    h3_sims %>%
-      rename(dep = unitname) %>%
-      group_by(dep, time) %>%
-      summarise(
-        q05 = quantile(cases, 0.025, na.rm = T),
-        mean = mean(cases, na.rm = T),
-        q50 = quantile(cases, 0.5, na.rm = T),
-        q95 = quantile(cases, 0.975, na.rm = T)
-      ) %>%
-      ungroup() %>%
-      mutate(
-        date = lubridate::date_decimal(time)
-      ) %>%
-      mutate(date = as.Date(lubridate::round_date(date, unit = 'day')))
-  },
-  timing = FALSE
-)
-@
-
 \begin{figure}[!h]
-<<Model3_SpatPOMP_sims_Figure, fig.height=3>>=
-
-# Plot the reported cases and confidence region for each department.
-ggplot() +
-  geom_line(data = h3_quants, aes(x = date, y = q50 + 1), col = 'blue') +
-  geom_line(data = dep_plot_df, aes(x = date, y = cases + 1), col = 'black') +
-  geom_ribbon(  # plot confidence region.
-    data = h3_quants,
-    aes(x = date, ymin = q05 + 1, ymax = q95 + 1),
-    alpha = 0.4
-  ) +
-  facet_wrap(
-    ~factor(dep, levels = plot_order), nrow = 2, labeller = dep_labeller
-  ) +
-  labs(y = 'Log Reported Cholera Cases') +
-  scale_x_date(  # Format the date on the X-axis.
-    date_labels = "'%y",
-    breaks = seq.Date(
-      from = as.Date("2011-01-01"),
-      to = as.Date("2019-01-01"),
-      by = '2 years'
-    )
-  ) +
-  theme(axis.title.x = element_blank()) +
-  scale_y_log10(  # Plot on log-scale
-    labels = scales::trans_format("log10", scales::math_format(10^.x))
-  )
-@
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
+\includegraphics[width=\maxwidth]{figure/Model3_SpatPOMP_sims_Figure-1} 
+\end{knitrout}
 \caption{\label{fig:h3spatsims}
@@
 \end{figure}
-
-<<Remove Model3 spatSims, cache=FALSE, include=FALSE, message=FALSE, echo=FALSE>>=
-rm(h3_quants, h3_nsim)
-gc()
-@
 
@@
 
-<<Get Model 1 Seasonality>>=
-h1_seas_df <- get_h1_seasonality(h1)
-
-df_rain <- haitiRainfall %>%
-  pivot_longer(  # Convert data to "long" format
-    data = .,
-    cols = -date,
-    names_to = 'dep',
-    values_to = 'rainfall'
-  ) %>%
-  mutate(year = lubridate::year(date), week = lubridate::week(date)) %>%
-  group_by(year, week) %>%
-  summarize(nwr = sum(rainfall)) %>%
-  ungroup() %>%  # Not needed, but good practice
-  mutate(
-    date_week_start = lubridate::ymd(paste0(2014, "-01-01")) +
-      lubridate::weeks(week - 1),
-    week_date = as.Date(date_week_start, format = "%m-%d"),
-    std_national_rain = (nwr - min(nwr)) / (max(nwr) - min(nwr))
-  )
-
-# Calculate the average national weekly rainfall.
-mean_rain <- df_rain %>%
-  group_by(week_date) %>%
-  summarize(mean_rain = mean(std_national_rain)) %>%
-  ungroup() %>%
-  mutate(mean_rain = (mean_rain - min(mean_rain)) / ((max(mean_rain) - min(mean_rain))))
-
-# Create a figure showing annual rainfall, scaled from 0-1. This is top figure
-gg_rain <- df_rain %>%
-  filter(week != 1, week != max(week)) %>%
-  ggplot(aes(x = week_date, y = std_national_rain, group = year, col = year)) +
-  geom_line() +
-  theme_bw() +
-  guides(color = 'none') +
-  ylab('Standardized Weekly\nNational Rain') +
-  theme(  # Remove elements that are un-unused because this is top figure
-    axis.title.x = element_blank(),
-    axis.ticks.x = element_blank(),
-    axis.text.x = element_blank(),
-    axis.text.y = element_text(size = 7.5),
-    axis.title.y = element_text(size = 9)
-  )
-
-# Create plot that shows the seasonality of Model 1, alongside the average
-# weekly national rainfall.
-gg_trans <- ggplot() +
-  geom_line(data = h1_seas_df, aes(x = date, y = trans_std), linetype = 'dashed') +
-  geom_line(
-    data = mean_rain,
-    aes(x = week_date + lubridate::years(6), y = mean_rain),
-    col = '#2166ac'
-  ) +
-  scale_x_date(date_labels = "%b", date_breaks = '1 month') +
-  ylab('Standardized Seasonal\nContact Rate') +
-  theme(axis.title.x = element_blank(),
-        axis.text = element_text(size = 7.5),
-        axis.title.y = element_text(size = 9))
-
-@
-
 \begin{figure}[!h]
-<<Model1_Seasonality_Figure, fig.height=2.9, fig.width=5, fig.align='center'>>=
-cowplot::plot_grid(gg_rain, gg_trans, align = "v", ncol = 1)
-@
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
+
+{\centering \includegraphics[width=\maxwidth]{figure/Model1_Seasonality_Figure-1}
+
+}
+
+\end{knitrout}
 \caption{\label{fig:h1SeasRain}
@@
 Our case study provides an example of this in the parameter estimate for the duration of natural immunity due to cholera infection, $\muRS^{-1}$.
-Under the framework of Model~2, the best estimate for this parameter is $\Sexpr{signif(1/h2_fit$h2_params["sigma"],3)}$ yr, suggesting that individuals have \old{effectively} permanent immunity to cholera once infected.
+Under the framework of Model~2, the best estimate for this parameter is $\ensuremath{1.4\times 10^{11}}$ yr, suggesting that individuals have \old{effectively} permanent immunity to cholera once infected.
 \new{Rather than interpreting this as scientific evidence that individuals have permanent immunity from cholera, this result suggests that Model~2 favors a regime where reinfection events are a negligible part of the dynamics.}
@@
 
-<<Load Model 1 VaccScen Sims, message=FALSE, include=FALSE>>=
-
-# Number of particles and simulations to use for pfilter and project from filter, respectively.
-h1_scen_NP   <- switch(RUN_LEVEL, 50, 500, 5000)
-h1_scen_sims <- switch(RUN_LEVEL, 20, 100, 1000)
-
-# Simulate vaccination scenario from filtering distribution. The code
-# below creates a model, samples from the filtering distribution, and then
-# projects the vaccination scenario, under the specified model.
-h1_S0_results <- bake(
-  file = paste0("model1/", rl_dir, "S0_sims.rds"), {
-
-    # Create the model
-    s0 <- haiti1_joint(vacscen = 'id0')
-
-    # Load coefficients
-    coef(s0) <- p1
-
-    # Get samples from filtering distribution
-    s0_pf <- pfilter(s0, save.state = TRUE, Np = h1_scen_NP)
-    end_states <- saved_states(s0_pf)
-    end_states <- end_states[[length(end_states)]]
-
-    # project vacc scenario, with filtering dist, and model.
-    h1_S0_sims <- project_from_filter(mod = s0, end_states = end_states, nsims = h1_scen_sims)
-
-    results <- list()
-    results$mod1_V0_sims <- agg_mod1_sims(h1_S0_sims)
-    results$mod1_V0_probs <- get_elimProbs(h1_S0_sims, model = 1)
-
-    results
-  },
-  timing = FALSE
-)
-
-# Simulate vaccination scenario from filtering distribution. The code
-# below creates a model, samples from the filtering distribution, and then
-# projects the vaccination scenario, under the specified model.
-h1_S2_results <- bake(
-  file = paste0('model1/', rl_dir, 'S2_sims.rds'), {
-
-    # Create the model
-    s2 <- haiti1_joint(vacscen = "id2")
-
-    # Set model parameters that need to be adjusted
-    depts <- 2
-    h1_par_names <- names(p1)
-    h1_params_temp <- c(p1, rep(0.0, 5 * depts))
-    names(h1_params_temp) <- c(
-      h1_par_names,
-      paste0("S", 1:depts, "_0"),
-      paste0("E", 1:depts, "_0"),
-      paste0("I", 1:depts, "_0"),
-      paste0("A", 1:depts, "_0"),
-      paste0("R", 1:depts, "_0")
-    )
-    coef(s2) <- h1_params_temp  # Set model parameters
-
-    # Get samples from filtering distribution for Scenario 2 model:
-    s2_pf <- pfilter(s2, save.state = TRUE, Np = h1_scen_NP)
-    end_states <- saved_states(s2_pf)
-    end_states <- end_states[[length(end_states)]]
-
-    # project vacc scenario, with filtering dist, and model.
-    h1_S2_sims <- project_from_filter(mod = s2, end_states = end_states, nsims = h1_scen_sims)
-
-    results <- list()
-    results$mod1_V1_sims <- agg_mod1_sims(h1_S2_sims)
-    results$mod1_V1_probs <- get_elimProbs(h1_S2_sims, model = 1)
-
-    results
-  },
-  timing = FALSE
-)
-
-# Simulate vaccination scenario from filtering distribution. The code
-# below creates a model, samples from the filtering distribution, and then
-# projects the vaccination scenario, under the specified model.
-h1_S4_results <- bake(
-  file = paste0('model1/', rl_dir, 'S4_sims.rds'), {
-
-    # Create the model
-    s4 <- haiti1_joint(vacscen = "id4")
-
-    # Set model parameters that need to be adjusted
-    depts <- 3
-    h1_par_names <- names(p1)
-    h1_params_temp <- c(p1, rep(0.0, 5 * depts))
-    names(h1_params_temp) <- c(
-      h1_par_names,
-      paste0("S", 1:depts, "_0"),
-      paste0("E", 1:depts, "_0"),
-      paste0("I", 1:depts, "_0"),
-      paste0("A", 1:depts, "_0"),
-      paste0("R", 1:depts, "_0")
-    )
-    coef(s4) <- h1_params_temp  # Set model parameters
-
-    # Get samples from filtering distribution for Scenario 2 model:
-    s4_pf <- pfilter(s4, save.state = TRUE, Np = h1_scen_NP)
-    end_states <- saved_states(s4_pf)
-    end_states <- end_states[[length(end_states)]]
-
-    # project vacc scenario, with filtering dist, and model.
-    h1_S4_sims <- project_from_filter(mod = s4, end_states = end_states, nsims = h1_scen_sims)
-
-    results <- list()
-    results$mod1_V2_sims <- agg_mod1_sims(h1_S4_sims)
-    results$mod1_V2_probs <- get_elimProbs(h1_S4_sims, model = 1)
-
-    results
-  },
-  timing = FALSE
-)
-
-# Simulate vaccination scenario from filtering distribution. The code
-# below creates a model, samples from the filtering distribution, and then
-# projects the vaccination scenario, under the specified model.
-h1_S3_results <- bake(
-  file = paste0('model1/', rl_dir, 'S3_sims.rds'), {
-
-    # Create the model
-    s3 <- haiti1_joint(vacscen = "id3")
-
-    # Set model parameters that need to be adjusted
-    depts <- 10
-    h1_par_names <- names(p1)
-    h1_params_temp <- c(p1, rep(0.0, 5 * depts))
-    names(h1_params_temp) <- c(
-      h1_par_names,
-      paste0("S", 1:depts, "_0"),
-      paste0("E", 1:depts, "_0"),
-      paste0("I", 1:depts, "_0"),
-      paste0("A", 1:depts, "_0"),
-      paste0("R", 1:depts, "_0")
-    )
-    coef(s3) <- h1_params_temp  # Set model parameters
-
-    # Get samples from filtering distribution for Scenario 2 model:
-    s3_pf <- pfilter(s3, save.state = TRUE, Np = h1_scen_NP)
-    end_states <- saved_states(s3_pf)
-    end_states <- end_states[[length(end_states)]]
-
-    # project vacc scenario, with filtering dist, and model.
-    h1_S3_sims <- project_from_filter(mod = s3, end_states = end_states, nsims = h1_scen_sims)
-
-    results <- list()
-    results$mod1_V3_sims <- agg_mod1_sims(h1_S3_sims)
-    results$mod1_V3_probs <- get_elimProbs(h1_S3_sims, model = 1)
-
-    results
-  },
-  timing = FALSE
-)
-
-# Simulate vaccination scenario from filtering distribution. The code
-# below creates a model, samples from the filtering distribution, and then
-# projects the vaccination scenario, under the specified model.
-h1_S1_results <- bake(
-  file = paste0('model1/', rl_dir, 'S1_sims.rds'), {
-
-    # Create the model
-    s1 <- haiti1_joint(vacscen = "id1")
-
-    # Set model parameters that need to be adjusted
-    depts <- 10
-    h1_par_names <- names(p1)
-    h1_params_temp <- c(p1, rep(0.0, 5 * depts))
-    names(h1_params_temp) <- c(
-      h1_par_names,
-      paste0("S", 1:depts, "_0"),
-      paste0("E", 1:depts, "_0"),
-      paste0("I", 1:depts, "_0"),
-      paste0("A", 1:depts, "_0"),
-      paste0("R", 1:depts, "_0")
-    )
-    coef(s1) <- h1_params_temp  # Set model parameters
-
-    # Get samples from filtering distribution for Scenario 2 model:
-    s1_pf <- pfilter(s1, save.state = TRUE, Np = h1_scen_NP)
-    end_states <- saved_states(s1_pf)
-    end_states <- end_states[[length(end_states)]]
-
-    # project vacc scenario, with filtering dist, and model.
-    h1_S1_sims <- project_from_filter(mod = s1, end_states = end_states, nsims = h1_scen_sims)
-
-    results <- list()
-    results$mod1_V4_sims <- agg_mod1_sims(h1_S1_sims)
-    results$mod1_V4_probs <- get_elimProbs(h1_S1_sims, model = 1)
-
-    results
-  },
-  timing = FALSE
-)
-@
-
-<<Simulate Model 3 Vacc Scenarios, message=FALSE, include=FALSE>>=
-NP_BPF <- switch(RUN_LEVEL, 20, 200, 2000)
-NSIM   <- switch(RUN_LEVEL, 5,  50, 1000)
-
-h3_bpf_end_states <- bake(
-  file = paste0("model3/", rl_dir, "h3_bpf_end_states.rds"),
-  expr = {
-    coef(h3Spat) <- p3
-    h3_bpf <- bpfilter(h3Spat, Np = NP_BPF, block_size = 1, save_states = TRUE)
-    ss <- h3_bpf@saved.states
-    ss[[length(ss)]]
-  },
-  timing = FALSE
-)
-
-mod3_V0_res <- bake(
-  file = paste0("model3/", rl_dir, "mod3_V0_res.rds"), {
-
-    # Create a list to store results
-    h3_V0_res <- list()
-
-    # Get "noVacc" scenario parameters
-    h3_S0_par <- get_model3_vacc_scenario_params(scenario = "noVacc")
-    coef(h3Spat) <- p3
-    coef(h3Spat)[names(h3_S0_par)] <- h3_S0_par
-
-    # Using parameters and filtered distribution, project system
-    h3_V0_sims <- project_from_filter(
-      h3Spat, end_states = h3_bpf_end_states, covarGen = project_rain,
-      nsims = NSIM, seed = 6897154
-    )
-
-    # Store nationally aggregated results
-    h3_V0_res$mod3_V0_sims <- agg_mod3_sims(h3_V0_sims)
-    h3_V0_res$mod3_V0_probs <- get_elimProbs(h3_V0_sims, model = 3)
-
-    # For the supplement material, save quantiles of the Susceptible individuals
-    h3_V0_res$SQuants <- h3_V0_sims %>%
-      mutate(
-        date = as.Date(lubridate::round_date(lubridate::date_decimal(time), unit = 'day'))
-      ) %>%
-      group_by(unitname, date) %>%
-      mutate(
-        pop = h3Spat@params[paste0("H", which(first(unitname) == unit_names(h3Spat)))]
-      ) %>%
-      summarize(
-        Q025 = quantile((S + VSd + VSdd + VSdd_alt + VSd_alt) / pop, probs = 0.025),
-        Q50 = quantile((S + VSd + VSdd + VSdd_alt + VSd_alt) / pop, probs = 0.5),
-        Q975 = quantile((S + VSd + VSdd + VSdd_alt + VSd_alt) / pop, probs = 0.975)
-      ) %>%
-      ungroup() %>%
-      rename(dep = unitname)
-
-    # Return results
-    h3_V0_res
-  }
-)
-
-mod3_V1_res <- bake(
-  file = paste0("model3/", rl_dir, "mod3_V1_res.rds"), {
-
-    # Create a list to store results
-    h3_V1_res <- list()
-
-    # Get "noVacc" scenario parameters
-    h3_V1_par <- get_model3_vacc_scenario_params(scenario = "2dep")
-    coef(h3Spat) <- p3
-    coef(h3Spat)[names(h3_V1_par)] <- h3_V1_par
-
-    # Using parameters and filtered distribution, project system
-    h3_V1_sims <- project_from_filter(
-      h3Spat, end_states = h3_bpf_end_states, covarGen = project_rain,
-      nsims = NSIM, seed = 6897154
-    )
-
-    # Store nationally aggregated results
-    h3_V1_res$mod3_V1_sims <- agg_mod3_sims(h3_V1_sims)
-    h3_V1_res$mod3_V1_probs <- get_elimProbs(h3_V1_sims, model = 3)
-
-    # Return results
-    h3_V1_res
-  }
-)
-
-mod3_V2_res <- bake(
-  file = paste0("model3/", rl_dir, "mod3_V2_res.rds"), {
-
-    # Create a list to store results
-    h3_V2_res <- list()
-
-    # Get "noVacc" scenario parameters
-    h3_V2_par <- get_model3_vacc_scenario_params(scenario = "3dep")
-    coef(h3Spat) <- p3
-    coef(h3Spat)[names(h3_V2_par)] <- h3_V2_par
-
-    # Using parameters and filtered distribution, project system
-    h3_V2_sims <- project_from_filter(
-      h3Spat, end_states = h3_bpf_end_states, covarGen = project_rain,
-      nsims = NSIM, seed = 6897154
-    )
-
-    # Store nationally aggregated results
-    h3_V2_res$mod3_V2_sims <- agg_mod3_sims(h3_V2_sims)
-    h3_V2_res$mod3_V2_probs <- get_elimProbs(h3_V2_sims, model = 3)
-
-    # Return results
-    h3_V2_res
-  }
-)
-
-mod3_V3_res <- bake(
-  file = paste0("model3/", rl_dir, "mod3_V3_res.rds"), {
-
-    # Create a list to store results
-    h3_V3_res <- list()
-
-    # Get "noVacc" scenario parameters
-    h3_V3_par <- get_model3_vacc_scenario_params(scenario = "slowNation")
-    coef(h3Spat) <- p3
-    coef(h3Spat)[names(h3_V3_par)] <- h3_V3_par
-
-    # Using parameters and filtered distribution, project system
-    h3_V3_sims <- project_from_filter(
-      h3Spat, end_states = h3_bpf_end_states, covarGen = project_rain,
-      nsims = NSIM, seed = 6897154
-    )
-
-    # Store nationally aggregated results
-    h3_V3_res$mod3_V3_sims <- agg_mod3_sims(h3_V3_sims)
-    h3_V3_res$mod3_V3_probs <- get_elimProbs(h3_V3_sims, model = 3)
-
-    # Return results
-    h3_V3_res
-  }
-)
-
-mod3_V4_res <- bake(
-  file = paste0("model3/", rl_dir, "mod3_V4_res.rds"), {
-
-    # Create a list to store results
-    h3_V4_res <- list()
-
-    # Get "noVacc" scenario parameters
-    h3_V4_par <- get_model3_vacc_scenario_params(scenario = "fastNation")
-    coef(h3Spat) <- p3
-    coef(h3Spat)[names(h3_V4_par)] <- h3_V4_par
-
-    # Using parameters and filtered distribution, project system
-    h3_V4_sims <- project_from_filter(
-      h3Spat, end_states = h3_bpf_end_states, covarGen = project_rain,
-      nsims = NSIM, seed = 6897154
-    )
-
-    # Store nationally aggregated results
-    h3_V4_res$mod3_V4_sims <- agg_mod3_sims(h3_V4_sims)
-    h3_V4_res$mod3_V4_probs <- get_elimProbs(h3_V4_sims, model = 3)
-
-    # Return results
-    h3_V4_res
-  }
-)
-@
-
 Simulations from probabilistic models (Models~1 and~3) represent possible trajectories of the dynamic system under the scientific assumptions of the models.
@@
 
-<<Compute Elimination Probs, echo=FALSE>>=
-h1_S0_results$mod1_V0_probs$ElimTime$scenario = "V0"
-h1_S2_results$mod1_V1_probs$ElimTime$scenario = "V1"
-h1_S4_results$mod1_V2_probs$ElimTime$scenario = "V2"
-h1_S3_results$mod1_V3_probs$ElimTime$scenario = "V3"
-h1_S1_results$mod1_V4_probs$ElimTime$scenario = "V4"
-
-mod1_probElims <- rbind(
-  h1_S0_results$mod1_V0_probs$ElimTime,
-  h1_S2_results$mod1_V1_probs$ElimTime,
-  h1_S4_results$mod1_V2_probs$ElimTime,
-  h1_S3_results$mod1_V3_probs$ElimTime,
-  h1_S1_results$mod1_V4_probs$ElimTime
-)
-
-mod3_V0_res$mod3_V0_probs$ElimTime$scenario = "V0"
-mod3_V1_res$mod3_V1_probs$ElimTime$scenario = "V1"
-mod3_V2_res$mod3_V2_probs$ElimTime$scenario = "V2"
-mod3_V3_res$mod3_V3_probs$ElimTime$scenario = "V3"
-mod3_V4_res$mod3_V4_probs$ElimTime$scenario = "V4"
-
-mod3_probElims <- rbind(
-  mod3_V0_res$mod3_V0_probs$ElimTime,
-  mod3_V1_res$mod3_V1_probs$ElimTime,
-  mod3_V2_res$mod3_V2_probs$ElimTime,
-  mod3_V3_res$mod3_V3_probs$ElimTime,
-  mod3_V4_res$mod3_V4_probs$ElimTime
-)
-
-mod1_probElims$mod <- "Model 1"
-mod3_probElims$mod <- "Model 3"
-mod3_probElims$time <- as.Date(lubridate::round_date(lubridate::date_decimal(mod3_probElims$time), unit = 'day'))
-
-all_prob_elims <- dplyr::bind_rows(mod1_probElims, mod3_probElims)
-@
-
 \begin{figure}[!h]
-<<Elimination_Probs_Figure, fig.height=2.2>>=
-ggplot(tidyr::drop_na(all_prob_elims), aes(x = time, y = elim_prob, col = mod)) +
-  geom_line() +
-  facet_wrap(~scenario, nrow = 1) +
-  scale_color_manual(values = c("Model 1" = "#377eb8", "Model 3" = "#e41a1c")) +
-  ylab("Probability of Elimination (%)") +
-  theme(axis.title.x = element_blank(),
-        legend.position = 'bottom',
-        legend.title = element_blank(),
-        axis.text.x = element_text(angle = 35, hjust = 1),
-        legend.margin=margin(c(-5, 0, -3, 0)),
-        legend.box.margin = margin(c(-5, 0, -3, 0))) +
-  scale_x_date(date_labels = "'%y", breaks = seq.Date(from = as.Date("2010-01-01"), as.Date("2030-01-01"), by = '2 years'))
-@
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
+\includegraphics[width=\maxwidth]{figure/Elimination_Probs_Figure-1} 
+\end{knitrout}
 \caption{\label{fig:elimProbs}

### 2024-03-04 ms.Rnw
Replaced edit commands with the actual text

@@
 We use this case study to investigate current procedures for fitting mechanistic models to time series data, while identifying limitations of these methodologies and proposing remedies.
-\editAuthorSum
+Our analysis presents methodology for diagnosing how well a model describes observed data.
+Using objective measures to assess model fit ensures that our evaluation is based on quantifiable criteria.
+Incorporating reproducibility into this assessment results in a framework that enables the validation or refinement of model based inferences when revisiting the data, facilitating scientific discovery.
 Our data analysis workflow is supported by recent advances in algorithms, software and hardware, which facilitate statistical fitting of nonlinear stochastic dynamic models to observed incidence data.
@@
 
-\editModelingGuidance
+Numerous guidelines have been proposed for using mechanistic models to inform policy, reviewed in \cite{behrend20}.
+Behrend et al.~\cite{behrend20} identify the importance of stakeholder engagement, transparency, reproducibility, uncertainty communication, and testable model outcomes.
+These and related principles are echoed by other influential articles \cite{saltelli20,donnelly18}.
+Additional literature emphasizes model calibration and evaluation techniques \cite{dahabreh17,egger17,penaloza15}.
+These guidelines often lack implementation specifics.
+As an example, \cite{lee20} largely adhere to the principles of \cite{behrend20}---though assessing the extent of stake-holder engagement is challenging---yet their projections are inconsistent with actual cholera incidence data from 2019 to 2022, demonstrating the limitations of current standards.
+We provide methodology for rigorous statistical calibration and evaluation of dynamic models (as advocated by \cite{saltelli19}), thereby expanding on the prevailing guidance.
+We specifically emphasize principles that prove essential in our case study.
+Complementary methodological suggestions arising from a spatio-temporal analysis of COVID-19 are detailed in \cite{li23}.
 
@@
 
-\editMechModels
+Mechanistic models representing biological phenomena are valuable for epidemiology and consequently for public health policy \cite{lofgren14,mccabe21}.
+More broadly, they have useful roles throughout biology, especially when combined with statistical methods that properly account for stochasticity and nonlinearity \cite{may04}.
+In some situations, modern machine learning methods can outperform mechanistic models on epidemiological forecasting tasks \cite{lau22,baker18}.
+The predictive skill of non-mechanistic models can reveal limitations in mechanistic models, but cannot readily replace the scientific understanding obtained by describing the biological dynamics of the system in a mathematical model \cite{baker18,prosperi20}.
 
@@
 
-\editConfound
+The statistical evidence of a trend in transmission rate in this model could be explained by any trending variable (such as hygiene improvements, or changes in population behavior), resulting in confounding from collinear covariates.
+Alternatively, it is possible that the negative trend observed in the incidence data could be attributed to a decreasing reporting rate rather than decreasing transmission rate.
+This could be formally tested by comparing models with either trend specification.
+We did not do this because evidence suggests that reporting rate was maintained or increased (Figure~1 of \cite{rebaudet21}).
+We instead argue that a decreasing transmission rate is a plausible way to explain the decrease in cases over time, as there is alternative evidence that supports this model \cite{rebaudet19CATI,rebaudet21,michel19}.
+It is not practical to test all remotely plausible model variations, yet a strongly supported conclusion should avoid ruling out untested hypotheses.
+The robust statistical conclusion for our analysis is that a model which allows for change fits better than one which does not, and a trend in transmission is a plausible way to do this.
 
@@
 
-\editMetaPop
+In our literature review, 17 articles considered dynamic models that incorporate spatial heterogeneity \cite{lee20,tuite11,pasetto18,fitzgibbon20,eisenberg13,rinaldo12,chao11,abrams13,trevisin22,sallah17,collins14,kelly16,azman12,leung22,kuhn14,mari15,gatto12}.
+All but four \cite{lee20,pasetto18,sallah17,azman12} of these studies used deterministic dynamic models: this greatly simplifies the process of calibrating model parameters to incidence data, though deterministic models can struggle to describe complex stochastic dynamics.
+The model in \cite{pasetto18} was fit using an Ensemble Kalman Filter (EnKF) \cite{evensen09}; though EnKF scales favorably with the number of spatial units, it relies on linearization of latent states which can be problematic for highly nonlinear systems \cite{evensen22,ionides21}.
+Alternative approaches used to fit stochastic models included making additional simplifying assumptions to aid in the fitting process \cite{lee20}, and using MCMC algorithms \cite{sallah17,azman12} which require specific structures in the latent dynamics, making these algorithms non plug-and-play.
+In this subsection, we present how the recently developed iterated block particle filter (IBPF) algorithm \cite{ning23,ionides22} can be used to fit a spatially explicit stochastic dynamic model to incidence data.
 
@@
 where $\mathrm{E}\big(Y_n|Y_{n - 1}\big) = \NBintercept + \NBar Y_{n-1}$, and $\Var\big(Y_n|Y_{n - 1}\big) = \mathrm{E}\big(Y_n|Y_{n - 1}\big) + \mathrm{E}\big(Y_n|Y_{n - 1}\big)^2 / \NBsize$.
-\editSpatBench
+To obtain a benchmark for models with a meta-population structure, we fit independent auto-regressive negative binomial models to each spatial unit.
+Under the assumption of independence, the log-likelihood of the benchmark on the entire collection of data can be obtained by summing up the log-likelihood for each independent model.
+In general, a spatially explicit model may not have well-defined individual log-likelihoods, and, in this case, comparisons to benchmarks must be made at the level of the joint model.
 
@@
 
-\editSuppText
+In the supplement material (\nameref{S_mod3cal}), we explore in more detail the process of model fitting and diagnostics for Model~3.
+Here we demonstrate that the model outperforms its benchmark model on the aggregate scale.
+However, when focusing on the spatial units with the highest incidence of cholera, Model~3 performs roughly the same as a simple benchmark.
+By comparing simulations from the fitted model to the filtering distribution, we see that the reconstructed latent states of the model favor higher levels of cholera transmission than what is typically observed in the incidence data.
+These results hint at the possibility of model mispecification, and warrant a degree of caution in interpreting the model's outputs.
 
@@
 
-\editForecastOne
-
-\editForecastTwo
+Forecasts of a dynamic system should should be consistent with the available data.
+It is particularly important that forecasts are consistent with the most recent information available, as recent data is likely to be more relevant than older data.
+While this assertion may seem self-evident, it is not the case for deterministic models, for which the initial conditions together with the parameters are sufficient for forecasting, and so recent data may not be consistent with model trajectories.
+Epidemiological forecasts based on deterministic models are not uncommon in practice, despite their limitations \cite{king15}.
+Lee et al.~\cite{lee20} chose to obtain forecasts from all of their models by simulating forward from initial conditions, rather than conditioning forecasts based on the available data.
+This decision is possibly as a result of using a deterministic model, as forecasts from different models may only be considered comparable if they are obtained in the same way, which is most easily done be by simulating from initial conditions because Model~2 is deterministic.
+
+In contrast, for non-deterministic Models~1 and 3, we obtain forecasts by simulating future values using latent states that are harmonious with the most recent data.
+This is done by simulating forward from latent states drawn at the last observation time ($t_N$) from the filtering distribution $f_{\bm{X}_N|\bm{Y}_{1:N}}(\bm{x}_{N} | \bm{y}^*_{1:N} ; \hat\paramVec)$.
+The decision to obtain model forecasts from initial conditions partially explains the unsuccessful forecasts of Lee et al.~\cite{lee20}.
+Table~S7 in their supplement material, which contains results that were not discussed in their main article, shows that the subset of their simulations with zero cholera cases from 2019-2020 also correspond with its disappearance until 2022.
+These results support our argument that forecasts should be made by ensuring the starting point for the forecast is consistent with available data.
 
@@
 Simulations from probabilistic models (Models~1 and~3) represent possible trajectories of the dynamic system under the scientific assumptions of the models.
-\editModVacc
+Because Model~1 only accounts for national level disease dynamics, the pre-determined department-specific vaccination campaigns are carried out by assuming the vaccines are administered in one week to the same number of individuals that would have obtained vaccines if explicitly administered to the specific departments.
+We refer readers to \cite{lee20} and the accompanying supplement material for more details.
 Estimates of the probability of cholera elimination can therefore be obtained as the proportion of simulations from these models that result in cholera elimination.
@@
 
-\editDiscussionOne
-
-\editDiscussionTwo
-
-\editDiscussionThree
-
-\editDiscussionFour
-
-\editDiscussionFive
+Model diagnostics are a key tool for exposing unresolved model limitations and improving model fit.
+In our case study, we compared the three models from Lee et al.~\cite{lee20} to statistical benchmarks, revealing areas for improvement.
+For example, comparisons of Model~3 to a benchmark revealed its inadequacy in accounting for the post-hurricane increase in transmission, leading to a beneficial model refinement.
+When a mechanistic model is competitive with statistical benchmarks, we have a license to begin critical evaluation of its causal implications.
+If a model falls far behind simple benchmarks, there is likely to be substantial limitation in the data analysis that should be identified and remedied.
+In our case study, the re-calibrated version of Model~1 outperformed its benchmark, so we proceeded to examine causal implications.
+When doing so, we found that the fitted model provides a causal description of the dynamic system that is consistent with known features of the system, such as the importance of rainfall as a driver of cholera infection.
+The congruency between causal implications of the model and our belief about the dynamic system, coupled with a strong quantitative description of observed data relative to a benchmark, provides support for viewing the model as a plausible quantitative representation of the system under investigation.
+
+When fitting a mechanistic model to a dynamic system, the complexity of the model warrants consideration.
+Mathematical models provide simplified representations of complex systems, with the simplicity serving both to facilitate scientific understanding and to enable statistical inference on unknown parameters.
+In our case study, employing deterministic dynamics in Model~2 was found to be an over-simplification by comparing model fit with benchmarks.
+% Including stochastic metapopulation dynamics permitted investigation of spatiotemporal data, the main cost being additional complexity of numerical computations.
+Model~3 is distinct in that it is both stochastic and has a meta-population structure, making it challenging to draw likelihood-based inferences.
+In this paper, we demonstrated how this model class can be calibrated to incidence data using the innovative IBPF algorithm.
+One of only a few examples of fitting a nonlinear non-Gaussian meta-population model via maximum likelihood \cite{li23,ionides22}, this case study exemplifies the algorithm's potential benefits and provides an example for future researchers on a possible approach to fitting a high-dimensional non-linear model.
+
+Likelihood-based methods aid in determining an appropriate level of model complexity.
+Models fit to the same data can be compared using a criteria such as AIC.
+Nested model variations are particularly useful as they enable formal statistical testing of the nested features via likelihood ratio tests.
+Our case study demonstrated the examination of nested model features for all three models.
+Model~1 investigated a time-varying transmission rate; Model~2 assessed a phase-shift parameter in seasonal cholera peaks; Model~3 incorporated hurricane-related parameters.
+
+Unmodeled features of a dynamic system can lead to spurious or misleading parameter estimates if the features substantially impact observed data. 
+% A fitted model must compensate for these unaccounted features using its inherent flexibility. 
+In deterministic models, features that cannot be explained by measurement error must be accounted for by the choice of parameters. 
+% the choice of parameters is the primary source of flexibility to compensate for these unaccounted features.
+For our case study, some of the parameter estimates for the deterministic Model~2 are implausible, such as the infinite immunity discussed above, and this may be explained by compensation for model misspecification.
+Incorporating demographic and environmental stochasticity into models can mitigate the impact of unmodeled features.
+Stochastic phenomena are not only arguably present in biological systems, but their inclusion in a model also allows observed data variations to be attributed to inherent uncertainty rather than to distorted parameter values.
+Models~1 and 3 suggest the presence of extra-demographic stochasticity \cite{he10,stocks20,li23}, as evidenced by the confidence intervals for the corresponding parameter $\sigmaProc$ (\nameref{S_CI}).
+
+If forecasts are an important component of a modeling task, the forecasts should be consistent with the available data, particularly at the most recently available time points.
+In our case study, we did this by simulating forward from the filtering distribution, as this procedure conditions latent variables on the available data.
+This type of forecasting, however, is not directly available using a deterministic model, where future dynamics are fully determined by initial conditions and parameter values.
+This can result in over-confident model forecasts \cite{king15}.
+Despite their limitations, deterministic models can offer valuable insights into dynamic systems \cite{may04}.
+In \cite{lee20}, the forecasts from the deterministic Model~2 were qualitatively more consistent with the observed disappearance of cholera than the stochastic models.
+In our case study, we found improvements to Models~1 and 3 that resulted in improved forecasts for these models.
 

### 2024-03-04 si/inputs/paramUncertainty.tex

@@
 parameter uncertainty should be considered when obtaining model forecasts when the goal is to influencing policy.
-In a Bayesian modelling paradigm, the most natural way to account for parameter uncertainty in model forecasts is to suppose that $\theta$ comes from a distribution $f_{\Theta}$, and then to obtain $J$ forecasts from the model where each forecast is obtained using parameters drawn from the posterior distribution $\Theta_{1:J} \mid Y_{1:N} = y_{1:N}^* \sim f_{\Theta}\big(\theta | Y_{1:N} = y_{1:N}^*\big)$.
+In a Bayesian modelling paradigm, the most natural way to account for parameter uncertainty in model forecasts is to suppose that $\theta$ comes from a distribution $f_{\Theta}$, and then to obtain $J$ forecasts from the model where each forecast is obtained using parameters drawn from the posterior distribution $\theta_{1:J} \mid Y_{1:N} = y_{1:N}^* \sim f_{\Theta}\big(\theta | Y_{1:N} = y_{1:N}^*\big)$.
 
@@
 
-For each $i \in 1:k$, let $\Theta_i$ be a random vector of model parameters.
-Then, letting $\Theta$ denote the true model parameters, we endow the set $\{\Theta_1, \Theta_2, \ldots, \Theta_k\}$ with a discrete uniform distribution, such that $P\big(\Theta = \Theta_i\big) = \frac{1}{K}$ for all values $i \in \seq{1}{k}$.
+% Let $\Theta$ denote a random vector of model parameters.
+For each $k \in 1:K$, let $\theta_k$ be a unique set of model parameters.
+Letting $\Theta$ denote a random vector of model parameters, we endow $\Theta$ with a discrete uniform distribution on the set $\{\theta_1, \theta_2, \ldots, \theta_K\}$, such that $P\big(\Theta = \theta_k\big) = \frac{1}{K}$ for all values $k \in \seq{1}{K}$.
 Using this as a prior distribution, the posterior distribution of $\Theta | Y_{1:N} = y_{1:N}^*$ can be expressed as:
-$P\big(\Theta = \Theta_k | Y_{1:N} = y_{1:N}^*\big) = \frac{f_{Y_{1:N}}(y_{1:N}^*| \Theta_k)}{\sum_{l = 1}^K f_{Y_{1:N}}(y_{1:N}^*| \theta_l)}$.
-In a standard empirical Bayes analysis, the values $\Theta_1, \ldots, \Theta_k$ of the prior distribution would be chosen using the observed data, resulting in a posterior distribution that weighs the prior parameter vectors proportional to their corresponding likelihoods.
-Instead, we choose $\Theta_i$ to be the output of a stochastic routine applied to the observed data by setting $\Theta_i$ to be the output of an iterated filtering algorithm.
-In practice, because the likelihood maximization routines of iterated filtering methods are stochastic, it is common to run the iterated filtering method multiple times for each model in order to obtain a maximum likelihood estimate for model parameters.
+$P\big(\Theta = \theta_k | Y_{1:N} = y_{1:N}^*\big) = \frac{f_{Y_{1:N}}(y_{1:N}^*| \theta_k)}{\sum_{l = 1}^K f_{Y_{1:N}}(y_{1:N}^*| \theta_l)}$.
+In a standard empirical Bayes analysis, the values $\theta_1, \ldots, \theta_k$ of the prior distribution would be chosen using the observed data, resulting in a posterior distribution that weighs the prior parameter vectors proportional to their corresponding likelihoods.
+We choose $\theta_k$ to be the output of a stochastic routine applied to the observed data by setting $\theta_k$ to be the output of an iterated filtering algorithm.
+In practice, because the likelihood maximization routines of iterated filtering methods are stochastic, it is common to run the iterated filtering method multiple times $(K)$ for each model in order to obtain a maximum likelihood estimate for model parameters.
 This results in a natural set of parameters near the MLE that could be used as the discrete prior distribution.
+Alternatively, the set $\{\theta_1, \theta_2, \ldots, \theta_K\}$ could be determined by first obtaining marginal confidence intervals for each element of the parameter vector $\Theta$, and then creating a hypercube using the combination of marginal confidence intervals.
+The set $\{\theta_1, \theta_2, \ldots, \theta_K\}$ is then obtained by sampling uniformly $K$ values from the resulting hypercube, as was done by \citep{king15}.

### 2024-03-04 ms.Rnw
Improved wording in abstract

@@
 
-\input{edits.tex}
+% \input{edits.tex}
 
@@
 Specifically, we present approaches for diagnosing model misspecification and the consequent development of improved models.
-Additionally, we demonstrate the utility of recent advances in likelihood maximization for high-dimensional nonlinear dynamic models, enabling likelihood-based inference for high-dimensional models for spatiotemporal incidence data.
+Additionally, we demonstrate the utility of recent advances in likelihood maximization for high-dimensional nonlinear dynamic models, enabling likelihood-based inference for spatiotemporal incidence data using this class of models.
 Our workflow is reproducible and extendable, facilitating future investigations of this disease system.

### 2024-03-05 ArXiv.R
modified helper script with additional .bib file

@@ -12,7 +12,7 @@ ms_to_ArXiv <- function(file_name) {
   # tmp_file[geometry_line] <- "\\usepackage[top=0.85in,left=1.5in,footskip=0.75in]{geometry}"
   # tmp_file <- gsub("bib-haiti", 'ms', tmp_file)
 
-  bib_spot <- which(grepl("\\bibliography{bib-haiti}", tmp_file, fixed = TRUE))
+  bib_spot <- which(grepl("\\bibliography{bib-haiti,bib-lit}", tmp_file, fixed = TRUE))
   bib_content <- readLines(gsub("\\.tex", "\\.bbl", file_name))
 
   new_file <- c(tmp_file[1:(bib_spot - 1)], bib_content, tmp_file[(bib_spot + 1):length(tmp_file)])

### 2024-03-05 append_bib.R

@@ -3,7 +3,7 @@ args <- commandArgs(trailingOnly = TRUE)
 append_bib <- function(file_name) {
   tex_file <- readLines(file_name)
 
-  bib_spot <- which(grepl("\\bibliography{bib-haiti}", tex_file, fixed = TRUE))
+  bib_spot <- which(grepl("\\bibliography{bib-haiti,bib-lit}", tex_file, fixed = TRUE))
   bib_content <- readLines(gsub("\\.tex", "\\.bbl", file_name))
 
   new_file <- c(tex_file[1:(bib_spot - 1)], bib_content, tex_file[(bib_spot + 1):length(tex_file)])

### 2024-03-05 si/ReplicateLee20.Rnw
Removed cached statements

@@
 
-<<Simulate Lee Model 1, echo=FALSE, cache=TRUE>>=
+<<Simulate Lee Model 1, echo=FALSE, cache=FALSE>>=
 h1_epi <- haiti1()
@@
 
-<<SimulateEpi, echo=FALSE, include=FALSE, message=FALSE, cache=TRUE>>=
+<<SimulateEpi, echo=FALSE, include=FALSE, message=FALSE, cache=FALSE>>=
 h1_end <- haiti1(period = "endemic")
@@
 Compare to Figure S9 in the supplement of Lee et al.
-(2020).", cache=TRUE, echo=FALSE,warning=FALSE>>=
+(2020).", cache=FALSE, echo=FALSE,warning=FALSE>>=
 GGally::ggpairs(
@@
 (2020).
-The upper bound for the likelihood of this model is ", round(mod1_lee_ll), "."), cache=TRUE, fig.width=6.5, fig.height=3.75,warning=FALSE>>=
+The upper bound for the likelihood of this model is ", round(mod1_lee_ll), "."), cache=FALSE, fig.width=6.5, fig.height=3.75,warning=FALSE>>=
 
@@
 
-<<Model2Replication, cache=TRUE, echo=FALSE>>=
+<<Model2Replication, cache=FALSE, echo=FALSE>>=
 h2_epi <- haiti2()
@@
 \begin{figure}[!h]
-<<plotModel2Rep, cache=TRUE, echo=FALSE, fig.width=6.5, fig.height=4.5>>=
+<<plotModel2Rep, cache=FALSE, echo=FALSE, fig.width=6.5, fig.height=4.5>>=
 gg_epi <- projections %>%
@@
 
-<<Model3Replication, echo=FALSE, cache=TRUE>>=
+<<Model3Replication, echo=FALSE, cache=FALSE>>=
 NSIM <- 500

### 2024-04-03 ms.Rnw
removed cache=TRUE
Removed middle initial from Anna's name, per her request

@@
 Jesse Wheeler\textsuperscript{1*},
-AnnaElaine L. Rosengart\textsuperscript{2},
+AnnaElaine Rosengart\textsuperscript{2},
 Zhuoxun Jiang\textsuperscript{1},
@@
 
-<<Calculate Bench Likelihood, echo=FALSE, message=FALSE, include=FALSE, cache=TRUE>>=
+<<Calculate Bench Likelihood, echo=FALSE, message=FALSE, include=FALSE, cache=FALSE>>=
 

### 2024-04-03 si/confidenceIntervals.Rnw
Added computation times to supplement, as requested by a reviewer

@@
 In this section we provide confidence intervals for all model parameters, excluding those that take unique values for each spatial unit.
-For each model and parameter, we use principles of profile likelihood to obtain confidence intervals \cite{pawitan01}.
+For each model and parameter, we use principles of profile likelihood to obtain confidence intervals \citep{pawitan01}.
 Due to the non-linear and stochastic nature of Models~1 and 3, exact evaluations of the profile log-likelihood are difficult to obtain.
 Instead, the log-likelihood at each point of the profile is estimated using via Monte-Carlo based particle filter methods.
-We therefore obtain confidence intervals for the parameters of Model~1 and Model~3 using the Monte Carlo adjust profile (MCAP) algorithm \cite{ionides17}.
-
-In each subsection, we provide figures that show the curvature of the profile log-likelihood near the MLE (Figures~\ref{fig:m1Profs}--\ref{fig:m3Profs}). 
+We therefore obtain confidence intervals for the parameters of Model~1 and Model~3 using the Monte Carlo adjust profile (MCAP) algorithm \citep{ionides17}.
+
+Profile confidence intervals for nonlinear POMP models are require a large number of computations.
+In the Model~1 and Model~3 subsections, we mention the total computational expense of each profile log-likelihood evaluation.
+Each subsection also provide figures that show the curvature of the profile log-likelihood near the MLE (Figures~\ref{fig:m1Profs}--\ref{fig:m3Profs}).
 In these figures, the parameter values are shown on the transformed scale in which the profile was calculated.
@@
 Figure~\ref{fig:m1Profs} displays the Monte Carlo evaluations of the profile likelihood values, obtained using a particle filter.
+The total computational burden of this profile likelihood search was 3631 hours, which was computed in parallel using 9675 separate jobs via the \texttt{batchtools} R package \citep{batchtools}.
 
@@
 Figure~\ref{fig:m3Profs} displays the Monte Carlo evaluations of the profile likelihood values, obtained using a particle filter.
+The total computational burden of this profile likelihood search was 28938 hours, which was computed in parallel using 7568 separate jobs via the \texttt{batchtools} R package \citep{batchtools}.
 

### 2024-04-15 si/ReplicateLee20.Rnw
Changed reference style based on journal requirement

@@
 
-\section{Replication of \cite{lee20}}
+\section{Replication of Lee et al.~(2020)}
 
-In this article we claimed that we were able to obtain better fits to the observed data using the same models that were proposed by \cite{lee20}.
+In this article we claimed that we were able to obtain better fits to the observed data using the same models that were proposed by Lee et al.~(2020) \cite{lee20}.
 Along with visual comparisons to the data, this claim was supported by comparing likelihoods and AIC values in Table~2 in the manuscript.
-Because model likelihoods were not provided by \cite{lee20}, it is necessary to replicate these models in order to obtain likelihood estimates.
-Here we would like to thank the authors of \cite{lee20}, who provided detailed descriptions of their models, which enabled us to build on their work.
+Because model likelihoods were not provided by Lee et al.~(2020) \cite{lee20}, it is necessary to replicate these models in order to obtain likelihood estimates.
+Here we would like to thank the authors of Lee et al.~(2020) \cite{lee20}, who provided detailed descriptions of their models, which enabled us to build on their work.
 In the following subsections, we use our \code{R} package \code{haitipkg} to reproduce some of the results of \cite{lee20}.
-This reproduction allows us to estimate the likelihoods of the \cite{lee20} version of Models~1--3, and also provides a demonstration of the importance and usefulness of reproducible research.
+This reproduction allows us to estimate the likelihoods of the Lee et al.~(2020) \cite{lee20} version of Models~1--3, and also provides a demonstration of the importance and usefulness of reproducible research.
 
@@
 
-The model was implemented by a team at Johns Hopkins Bloomberg School of Public Health (hereafter referred to as the Model~1 authors) in the \code{R} programming language using the \code{pomp} package \citep{king16}.
+The model was implemented by a team at Johns Hopkins Bloomberg School of Public Health (hereafter referred to as the Model~1 authors) in the \code{R} programming language using the \code{pomp} package \cite{king16}.
 Original source code is publicly available with DOI: 10.5281/zenodo.3360991.
@@
 The source code that the Model~2 authors used to generate their results was written in the \code{Python} programming language, and is publicly available at \url{10.5281/zenodo.3360857} and its accompanying GitHub repository \url{https://github.com/lulelita/HaitiCholeraMultiModelingProject}.
-In order to perform our analysis in a unified framework, we re-implemented this model in the \code{R} programming language using the \code{spatPomp} package \citep{asfaw23arxiv}, which facilitates the creation of meta-population models.
-We note that the travel and water matrices used to implement the complex dynamics in Model~2 \citep{lee20sup} are not available in either the Zenodo archive or the GitHub repository;
+In order to perform our analysis in a unified framework, we re-implemented this model in the \code{R} programming language using the \code{spatPomp} package \cite{asfaw23arxiv}, which facilitates the creation of meta-population models.
+We note that the travel and water matrices used to implement the complex dynamics in Model~2 \cite{lee20sup} are not available in either the Zenodo archive or the GitHub repository;
 instead, we obtained these matrices via personal correspondence with the Model~2 authors.
-Using these matrices, and the point estimates for model parameters provided by \citep{lee20sup}, we created trajectories of the cholera dynamics and compared this to available data.
+Using these matrices, and the point estimates for model parameters provided by \cite{lee20sup}, we created trajectories of the cholera dynamics and compared this to available data.
 These trajectories, shown in Figure~\ref{fig:mod2rep}, are very similar to the trajectories shown in Figure~S15 of \cite{lee20sup}.
@@
 In this case, these discrepancies may possibly be attributed to implementing the model and plotting the model trajectory in two different programming languages.
-Another potential explanation for the discrepancy is that the parameters that we used are only approximately the same as those used by \citet{lee20sup}.
+Another potential explanation for the discrepancy is that the parameters that we used are only approximately the same as those used by Lee et al. (2020) \cite{lee20sup}.
 For example, the parameters $\transmission$, $\Wbeta{}$ had reported values of $9.9 \times 10^{-7}$ and $4.03 \times 10^{-2}$, respectively (Table~S13 of \cite{lee20sup}), but were actually fit to data and therefore likely these values have been rounded.
@@
 Some additional concerns about being able to accurately replicate the results of \cite{lee20} are valid.
-Details about the measurement models and how latent states were initialized for the epidemic model were not provided by \citet{lee20sup} and therefore these details must be inferred by looking at the provided source code.
+Details about the measurement models and how latent states were initialized for the epidemic model were not provided by Lee et al.
+(2020) \cite{lee20sup} and therefore these details must be inferred by looking at the provided source code.
 According to repository comments, the files \code{fit\allowbreak In\allowbreak Pieces\allowbreak 3params\allowbreak Clean\allowbreak May2019\allowbreak Public.py} and \code{fit\allowbreak In\allowbreak Pieces\allowbreak Mu\allowbreak With\allowbreak Frac\allowbreak Sus\allowbreak Fixed\allowbreak All\allowbreak Infections\allowbreak Public.py} were used to fit the epidemic and endemic phases of the model respectively, although it is apparent that these exact files were not used to obtain the reported results since the files contain some variable-naming errors that make it impossible to run the files without making modifications \footnote{One example of why the code cannot be run that the file loads functions from a non-extant file named \code{choleraEqs.py} in line 13 rather than \code{cholera\allowbreak Eqs\allowbreak Public.py}.}.
-The inability to replicate the results by \citet{lee20} by running the provided source code makes checking whether or not a our numeric implementation faithfully represents their results very difficult.
+The inability to replicate the results by Lee et al.
+(2020) \cite{lee20} by running the provided source code makes checking whether or not a our numeric implementation faithfully represents their results very difficult.
 Additionally, the script that was said to been used to obtain the results reported by \cite{lee20} appears to use a different measurement model than what was described in the supplemental material, again making it difficult to fully replicate the result of \cite{lee20} without being able to easily run the provided source code.

### 2024-04-15 si/calibrateMod3.Rnw
Removed cached statements
Changed reference style based on journal requirement

@@
 
-<<M3_bench, echo=FALSE, message=FALSE, include=FALSE, cache=TRUE>>=
+<<M3_bench, echo=FALSE, message=FALSE, include=FALSE, cache=FALSE>>=
 h3 <- haiti3_spatPomp()
@@
 To calibrate this model, we used the iterated block particle filter (IBPF) method of \cite{ionides22}.
-Due to the novelty of this algorithm, there exists only a few examples where the IBPF algorithm is used for data analysis \citep{li23,ionides22}, which is one motivation of the inclusion additional details related to fitting and diagnosing the model fit provided here.
-
-\cite{lee20} only estimated model parameters to a simplified version of Model~3 on a subset of the available data, as no method existed at the time of their publication to fit a fully coupled meta-population model to disease incidence data.
+Due to the novelty of this algorithm, there exists only a few examples where the IBPF algorithm is used for data analysis \cite{li23,ionides22}, which is one motivation of the inclusion additional details related to fitting and diagnosing the model fit provided here.
+
+Lee et al.
+(2020)~\cite{lee20} only estimated model parameters to a simplified version of Model~3 on a subset of the available data, as no method existed at the time of their publication to fit a fully coupled meta-population model to disease incidence data.
 Building on their results, we fit the fully coupled version of Model~3 to (nearly) all available data, reserving only a few observations to use to calibrate the initial conditions of the model (see the supplement for initialization models for more details).
@@
 A subsequent refinement search used parameter values corresponding to the largest model likelihoods as starting parameter values.
-The need for multiple searches does not appear to be uncommon, as a similar approach was used by \cite{ionides22}.
+The need for multiple searches does not appear to be uncommon, as a similar approach was used in \cite{ionides22}.
 While computationally intensive, profile likelihoods proved to be an effective alternative to maximizing model likelihoods without the need to apply this multistage heuristic.
 
-We use the iterative fitting / pruning technique described above to fit the fully coupled version of Model~3 proposed by \cite{lee20}.
+We use the iterative fitting / pruning technique described above to fit the fully coupled version of Model~3 proposed in \cite{lee20}.
 The maximum likelihood we obtained after two rounds of searching was $\Sexpr{round(max(haiti3_sub_fit$search2$logLiks$logLik))}$, which is higher than the benchmark model ($\Sexpr{round(sum(unitLikes_bench$ll))}$).
@@
 Further investigation of the model output reveals that the model is struggling to explain the sudden surge in cholera cases that occurred at this time, which coincides with the time that Hurricane Matthew struck Haiti.
-While the model does include a mechanism to account for increased cholera transmission due to large rainfall events, the mechanism does not appear to be sufficient to capture the damaging effects of the hurricane, which had the greatest impact in the the Sud and Grand'Anse departments \citep{ferreirai16}.
+While the model does include a mechanism to account for increased cholera transmission due to large rainfall events, the mechanism does not appear to be sufficient to capture the damaging effects of the hurricane, which had the greatest impact in the the Sud and Grand'Anse departments \cite{ferreirai16}.
 This result led us to include parameters $\Whur{u}$ and $\hHur{u}$ in Eq.~23 of the main text, which allows for an increase in the transmission rate between environmental cholera and humans for in Sud and Grand'Anse during and after the hurricane.
@@
 Indeed, a good statistical fit does not require the model to assert a causal explanation.
-For example, reconstructed latent variables should make sense in the context of alternative measurements of these variables \citep{grad12}.
+For example, reconstructed latent variables should make sense in the context of alternative measurements of these variables \cite{grad12}.
 We demonstrate this principle by examining the latent state of the calibrated model.

### 2024-04-15 si/confidenceIntervals.Rnw
Changed reference style based on journal requirement

--- 2024-03-04 si/confidenceIntervals.Rnw
@@
 
-In each subsection, we provide figures that show the curvature of the profile log-likelihood near the MLE (Figures~\ref{fig:m1Profs}--\ref{fig:m3Profs}). 
+Profile confidence intervals for nonlinear POMP models are require a large number of computations.
+In the Model~1 and Model~3 subsections, we mention the total computational expense of each profile log-likelihood evaluation.
+Each subsection also provide figures that show the curvature of the profile log-likelihood near the MLE (Figures~\ref{fig:m1Profs}--\ref{fig:m3Profs}).
 In these figures, the parameter values are shown on the transformed scale in which the profile was calculated.
@@
 Figure~\ref{fig:m1Profs} displays the Monte Carlo evaluations of the profile likelihood values, obtained using a particle filter.
+The total computational burden of this profile likelihood search was 3631 hours, which was computed in parallel using 9675 separate jobs via the \texttt{batchtools} R package \cite{batchtools}.
 
@@
 Figure~\ref{fig:m3Profs} displays the Monte Carlo evaluations of the profile likelihood values, obtained using a particle filter.
+The total computational burden of this profile likelihood search was 28938 hours, which was computed in parallel using 7568 separate jobs via the \texttt{batchtools} R package \cite{batchtools}.
 

### 2024-04-15 si/initialValues.Rnw
Changed reference style based on journal requirement

@@
 Finally, the initial proportion of Susceptible individuals $S_{0, 0}$ is calculated as $S_{0, 0} = 1 - \Iinit - \Einit$.
-This model for the initial values of the latent states matches that which was used by \citet{lee20}.
+This model for the initial values of the latent states matches that which was used by Lee et al. (2020) \cite{lee20}.
 
@@
 Specifically, the model sets $I_{u0}(0) = y^*_{1u} / \reportRate$ for each unit $u \in 1:10$, where $I_{u0}(t)$ is the number of infected individuals in unit $u$ at time $t$, vaccination scenario $z = 0$, $y^*_{tu}$ is the reported number of cases, and $\reportRate$ is the reporting rate.
-It is futher assumed that there are no individuals in the recovered compartment, as the epidemic has just begun. 
-This model for the initial values of the latent states matches that which was used by \citet{lee20}.
+It is further assumed that there are no individuals in the recovered compartment, as the epidemic has just begun.
+This model for the initial values of the latent states matches that which was used by Lee et al. (2020) \cite{lee20}.
 
@@
 The latent states of this model are initialized by enforcing the model dynamics on the incidence data from the start of the recorded cases until time $t_0$, requiring that some of the available data be used to determine the initial values of the latent states.
-This is the same approach that was taken by \citet{lee20}, who used the value $t_0 = \text{2014-02-22}$; this choice of $t_0$ results in modeling roughly only $60\%$ of the available data, some of which is later discarded for alternative reasons \citep{lee20sup}.
+This is the same approach that was taken by Lee et al.
+(2020) \cite{lee20}, who used the value $t_0 = \text{2014-02-22}$; this choice of $t_0$ results in modeling roughly only $60\%$ of the available data, some of which is later discarded for alternative reasons \cite{lee20sup}.
 
@@
 This choice of $t_0$ corresponds to using approximately $1\%$ of the available data to determine initial values of the latent states.
-In addition to modeling a larger portion of the available data, this choice of $t_0$ corresponds to an important real-world event, as daily reports from each of the departments were not being sent to the health ministry until November 10, 2010 \citep{barzilay13};
+In addition to modeling a larger portion of the available data, this choice of $t_0$ corresponds to an important real-world event, as daily reports from each of the departments were not being sent to the health ministry until November 10, 2010 \cite{barzilay13};
 this choice of $t_0$ therefore makes $\bm{Y}\big(t_1\big)$ the first week of data once daily reports are being sent to the health ministry.
@@
 For convenience, we denote these observations as $t_{-3}, t_{-2}$ and $t_{-1}$; as before, we let $y_{u,-k} = Y_{u}\big(t_{-k}\big)$ denote the observed case count for unit $u$ at time point $t_{-k}$, where $k \in 1:3$.
-Equations for the initial values of non-zero latent states are provided in Eqs.~\myeqref{eq:model3:I0}--\myeqref{eq:model3:B0}; these equations match those that were used by \citet{lee20}, the primary change being a change to the value of $t_0$.
+Equations for the initial values of non-zero latent states are provided in Eqs.~\myeqref{eq:model3:I0}--\myeqref{eq:model3:B0}; these equations match those that were used by Lee et al.
+(2020) \cite{lee20}, the primary change being a change to the value of $t_0$.
 

### 2024-04-15 si/inputs/initialValues.tex
Changed reference style based on journal requirement

@@
 Specifically, the model sets $I_{u0}(0) = y^*_{1u} / \reportRate$ for each unit $u \in 1:10$, where $I_{u0}(t)$ is the number of infected individuals in unit $u$ at time $t$, vaccination scenario $z = 0$, $y^*_{tu}$ is the reported number of cases, and $\reportRate$ is the reporting rate.
-This model for the initial values of the latent states matches that which was used by \citet{lee20}.
+This model for the initial values of the latent states matches that which was used by Lee et al. (2020) \cite{lee20}.
 

### 2024-04-15 si/inputs/modelDetails.tex
Changed reference style based on journal requirement

@@
 
-In Sections 2.1, 2.2 and 2.3 of the main article, we define compartment models in terms of their flow rates.
+In the Materials and methods Section of the main article, we define compartment models in terms of their flow rates.
 For a discrete population model, these rates define a Markov chain.
@@
 Here, we add additional details to clarify the mapping from a collection of rate functions to a fully specified process.
-Our treatment follows \citet{breto09}.
+Our treatment follows \cite{breto09}.
 
@@
 We proceed to describe the ones we use in this paper: via a system of ordinary differential equations (Sec.~\ref{subsec:ode}), a simple Markov counting system (Sec.~\ref{subsec:smcs}), and an over-dispersed Markov counting system (Sec.~\ref{subsec:odmcs}).
-Other representations include stochastic differential equations driven by Gaussian noise or Gamma noise \citep{bhadra11}.
+Other representations include stochastic differential equations driven by Gaussian noise or Gamma noise \cite{bhadra11}.
 
@@
 We therefore call the Markov chain a simple Markov counting system (SMCS).
-The infinitesimal mean of every flow is equal to its infinitesimal variance \citep{breto11} and so an SMCS is called equidispersed.
+The infinitesimal mean of every flow is equal to its infinitesimal variance \cite{breto11} and so an SMCS is called equidispersed.
 We note that the special case of Model~1 used by \cite{lee20} (with $\sigmaProc = 0$) is an SMCS.
 To permit more general mean-variance relationships for a Markov counting system, we must permit jumps of size greater than one.
-The utility of over-dispersed models, where the infinitesimal variance of the flow exceeds the infinitesimal mean, has become widely recognized \citep{stocks20,he10}.
+The utility of over-dispersed models, where the infinitesimal variance of the flow exceeds the infinitesimal mean, has become widely recognized \cite{stocks20,he10}.
 
@@
 
-Including white noise in the rate function enables the possibility of an over-dispersed Markov counting system \citep{breto11,breto09,he10}.
-Since rates should be non-negative, Gaussian noise is not appropriate and gamma noise is a convenient option that has found various applications \citep{romero-severson15, subramanian20}.
+Including white noise in the rate function enables the possibility of an over-dispersed Markov counting system \cite{breto11,breto09,he10}.
+Since rates should be non-negative, Gaussian noise is not appropriate and gamma noise is a convenient option that has found various applications \cite{romero-severson15, subramanian20}.
 Specifically, we consider a model given by
@@
 Therefore, the numerical scheme in Sec.~\ref{sec:numerics} can be taken as a definition of the meaning of \myeqref{eq:odmcs1}.
-The Markov chain defined by the limit of this Euler scheme as the step size decreases is an over-dispersed Markov counting system, with the possibility of instantaneous jumps of size greater than one \citep{breto11}.
+The Markov chain defined by the limit of this Euler scheme as the step size decreases is an over-dispersed Markov counting system, with the possibility of instantaneous jumps of size greater than one \cite{breto11}.
 

### 2024-04-15 si/inputs/paramUncertainty.tex
Changed reference style based on journal requirement

@@
 Furthermore, denote the observed data as $y_{1:N}^*$.
-Because the uncertainty in just a single parameter can lead to drastically different forecasts \citep{saltelli20},
+Because the uncertainty in just a single parameter can lead to drastically different forecasts \cite{saltelli20},
 parameter uncertainty should be considered when obtaining model forecasts when the goal is to influencing policy.
@@
 When frequentist methods are used, however, there does not exist a posterior distribution from which one could sample.
-A common approach could be to obtain a weighted average of the simulations from various models \citep{hoeting99}, but this can be problematic when forecasts from each model are very different from each other \citep{grueber11}.
-A similar approach that has been taken \citep{king15} is to obtain model forecasts using multiple sets of parameter values and then sample from the resulting forecasts using weights proportional to the corresponding likelihoods of the parameter values.
+A common approach could be to obtain a weighted average of the simulations from various models \cite{hoeting99}, but this can be problematic when forecasts from each model are very different from each other \cite{grueber11}.
+A similar approach that has been taken \cite{king15} is to obtain model forecasts using multiple sets of parameter values and then sample from the resulting forecasts using weights proportional to the corresponding likelihoods of the parameter values.
 This approach could be considered as empirical Bayes, as it is equivalent to using a discrete uniform prior where the set of values in the prior distribution is determined by a stochastic routine applied to the observed data, as discussed below.
@@
 Alternatively, the set $\{\theta_1, \theta_2, \ldots, \theta_K\}$ could be determined by first obtaining marginal confidence intervals for each element of the parameter vector $\Theta$, and then creating a hypercube using the combination of marginal confidence intervals.
-The set $\{\theta_1, \theta_2, \ldots, \theta_K\}$ is then obtained by sampling uniformly $K$ values from the resulting hypercube, as was done by \citep{king15}.
+The set $\{\theta_1, \theta_2, \ldots, \theta_K\}$ is then obtained by sampling uniformly $K$ values from the resulting hypercube, as was done by \cite{king15}.

### 2024-04-15 si/si.Rnw
Refactored SI + Makefile
Changed reference style based on journal requirement

@@
 Because the models are defined by their mathematical equations and numeric implementation, these diagrams are not unique visual representations of the model.
-Alternative representations that may be helpful in understanding the models explored in this paper were made by \citet{lee20sup}.
+Alternative representations that may be helpful in understanding the models explored in this paper were made by Lee et al. (2020) \cite{lee20sup}.
 
@@
 
-\input{inputs/initialValues}
+\input{initialValuesOut}
 
@@
 
+\input{confidenceIntervalsOut}
+
 \input{inputs/paramUncertainty}
 
-\section{Translating to \citet{lee20} notation}
+\section{Translating to Lee et al. (2020) \cite{lee20} notation}
 
-Since the models of \citet{lee20} were developed independently, the choice of notation varies inconsistently between models.
+Since the models of Lee et al. (2020) \cite{lee20} were developed independently, the choice of notation varies inconsistently between models.
 For our reanalysis, we rename parameters to provide a unified notation facilitating comparison between models.

### 2024-04-18 si/ReplicateLee20.Rnw
Modified references in supplement file, as requested by PLOS COMP BIO editorial office

@@
 
-\section{Replication of Lee et al.~(2020)}
+\section{Replication of Lee et al.~(2020)~\cite{lee20}}
 
@@
 <<PlotEpiDist, echo=FALSE, fig.cap="Bivariate distributions of parameter estimates after fitting epidemic phase of the Model~1 following the procedure described by Lee et al.
-(2020).
+(2020)~\\cite{lee20}.
 Compare to Figure S8 in the supplement of Lee et al.
-(2020).", fig.width=7, fig.height=7, fig.align='center',message=FALSE,warning=FALSE>>=
+(2020)~\\cite{lee20}.", fig.width=7, fig.height=7, fig.align='center',message=FALSE,warning=FALSE>>=
 GGally::ggpairs(
@@
 <<plotEndParams, fig.width=7, fig.height=7, fig.align='center', fig.cap="Bivariate distributions of parameter estimates after fitting endemic phase of the Model~1 following the procedure described by Lee et al.
-(2020).
+(2020)~\\cite{lee20}.
 Compare to Figure S9 in the supplement of Lee et al.
-(2020).", cache=FALSE, echo=FALSE,warning=FALSE>>=
+(2020)~\\cite{lee20}.", cache=FALSE, echo=FALSE,warning=FALSE>>=
 GGally::ggpairs(
@@
 
-<<plotMod1Sims, echo=FALSE, fig.cap=paste0("Simulations from Model~1 using parameter sets that were generated by running source code provided by Lee et al (2020).
+<<plotMod1Sims, echo=FALSE, fig.cap=paste0("Simulations from Model~1 using parameter sets that were generated by running source code provided by Lee et al (2020)~\\cite{lee20}.
 Compare to Figure S7 in the supplement of Lee et al.
-(2020).
+(2020)~\\cite{lee20}.
 The upper bound for the likelihood of this model is ", round(mod1_lee_ll), "."), cache=FALSE, fig.width=6.5, fig.height=3.75,warning=FALSE>>=
@@
 In order to perform our analysis in a unified framework, we re-implemented this model in the \code{R} programming language using the \code{spatPomp} package \cite{asfaw23arxiv}, which facilitates the creation of meta-population models.
-We note that the travel and water matrices used to implement the complex dynamics in Model~2 \cite{lee20sup} are not available in either the Zenodo archive or the GitHub repository;
+We note that the travel and water matrices used to implement the complex dynamics in Model~2 \cite{lee20} are not available in either the Zenodo archive or the GitHub repository;
 instead, we obtained these matrices via personal correspondence with the Model~2 authors.
-Using these matrices, and the point estimates for model parameters provided by \cite{lee20sup}, we created trajectories of the cholera dynamics and compared this to available data.
-These trajectories, shown in Figure~\ref{fig:mod2rep}, are very similar to the trajectories shown in Figure~S15 of \cite{lee20sup}.
+Using these matrices, and the point estimates for model parameters provided by \cite{lee20}, we created trajectories of the cholera dynamics and compared this to available data.
+These trajectories, shown in Figure~\ref{fig:mod2rep}, are very similar to the trajectories shown in Figure~S15 of the supplement of \cite{lee20}.
 
@@
 \caption{\label{fig:mod2rep}
-Model 2 trajectories using the \code{haitipkg}. Compare to Figure S15 in the supplement of \cite{lee20sup}.
+Model 2 trajectories using the \code{haitipkg}. Compare to Figure S15 in the supplement of \cite{lee20}.
 }
@@
 
-There are minor differences between Figure~\ref{fig:mod2rep} and Figure~S15 of \cite{lee20sup}.
+There are minor differences between Figure~\ref{fig:mod2rep} and Figure~S15 of \cite{lee20}.
 While the discrepancy appears minor, the deterministic nature of Model~2 implies that an exact replication of model trajectories should be possible.
 In this case, these discrepancies may possibly be attributed to implementing the model and plotting the model trajectory in two different programming languages.
-Another potential explanation for the discrepancy is that the parameters that we used are only approximately the same as those used by Lee et al. (2020) \cite{lee20sup}.
-For example, the parameters $\transmission$, $\Wbeta{}$ had reported values of $9.9 \times 10^{-7}$ and $4.03 \times 10^{-2}$, respectively (Table~S13 of \cite{lee20sup}), but were actually fit to data and therefore likely these values have been rounded.
+Another potential explanation for the discrepancy is that the parameters that we used are only approximately the same as those used by Lee et al. (2020) \cite{lee20}.
+For example, the parameters $\transmission$, $\Wbeta{}$ had reported values of $9.9 \times 10^{-7}$ and $4.03 \times 10^{-2}$, respectively (Table~S13 of the supplement material of \cite{lee20}), but were actually fit to data and therefore likely these values have been rounded.
 Additionally, our implementation of Model~2 used a time scale of years and many of the parameters were reported on a weekly scale, so small differences may result due to unit conversions.
@@
 Details about the measurement models and how latent states were initialized for the epidemic model were not provided by Lee et al.
-(2020) \cite{lee20sup} and therefore these details must be inferred by looking at the provided source code.
+(2020) \cite{lee20} and therefore these details must be inferred by looking at the provided source code.
 According to repository comments, the files \code{fit\allowbreak In\allowbreak Pieces\allowbreak 3params\allowbreak Clean\allowbreak May2019\allowbreak Public.py} and \code{fit\allowbreak In\allowbreak Pieces\allowbreak Mu\allowbreak With\allowbreak Frac\allowbreak Sus\allowbreak Fixed\allowbreak All\allowbreak Infections\allowbreak Public.py} were used to fit the epidemic and endemic phases of the model respectively, although it is apparent that these exact files were not used to obtain the reported results since the files contain some variable-naming errors that make it impossible to run the files without making modifications \footnote{One example of why the code cannot be run that the file loads functions from a non-extant file named \code{choleraEqs.py} in line 13 rather than \code{cholera\allowbreak Eqs\allowbreak Public.py}.}.
